### 2019-09-10 ResearchProposal.tex
Initial commit of research proposal - much of it copied from other proposals

\documentclass[12pt]{article}

\usepackage[round]{natbib}

\usepackage{hyperref}
\hypersetup{
    colorlinks=true,       % false: boxed links; true: colored links
    linkcolor=red,          % color of internal links (change box color with linkbordercolor)
    citecolor=green,       % color of links to bibliography
    filecolor=magenta,   % color of file links
    urlcolor=cyan           % color of external links
}

%% Comments
\newif\ifcomments\commentstrue

\ifcomments
\newcommand{\authornote}[3]{\textcolor{#1}{[#3 ---#2]}}
\newcommand{\todo}[1]{\textcolor{red}{[TODO: #1]}}
\else
\newcommand{\authornote}[3]{}
\newcommand{\todo}[1]{}
\fi

\newcommand{\wss}[1]{\authornote{blue}{SS}{#1}} %Spencer Smith
\newcommand{\jc}[1]{\authornote{red}{JC}{#1}} %Jacques Carette
\newcommand{\oo}[1]{\authornote{magenta}{OO}{#1}} %Olu Owojaiye
\newcommand{\pmi}[1]{\authornote{green}{PM}{#1}} %Peter Michalski
\newcommand{\ad}[1]{\authornote{cyan}{AD}{#1}} %Ao Dong

%\oddsidemargin 0mm
%\evensidemargin 0mm
%\textwidth 160mm
%\textheight 200mm

\begin{document}

\title{Assessing the Impact of MDE and Code Generation on the Sustainability of
  Scientific Computing Software: A Research Proposal} 
\author{Spencer Smith}
\date{\today}

\maketitle

\begin{abstract}
  ...
\end{abstract}

\tableofcontents

\section{Introduction}

Without dramatic intervention, our collective confidence in Scientific Computing
Software (SCS) is due for a catastrophic collapse.  We are increasingly trusting
ever more complex and ambitious computations, but our software foundations are
built on sand. Areas of concern include nuclear safety analysis and
computational medicine. Successfully building such software requires
communication between software developers and experts from multiple
domains. Collaboration is difficult at the best of times, and is made worse
because developers favour handcrafted solutions over adapting software
engineering processes, methods and tools \citep{FaulkEtAl2009}. Handcrafted
solutions do not account for the inevitable changes in requirements, design and
implementation. The twin challenges of changing requirements and inadequate
documentation conspire to make computational results notoriously difficult to
reproduce, especially in the important endeavour of one researcher independently
replicating the results of another \citep{Smith2018}.  Inadequate traceability
within documentation means re­certification of previously certified engineering
software is as expensive and time consuming as the original certification
exercise, because of a lack of explicit, effort reducing, traceability
information.

Quality of SCS improved through documentation.  SCS developers do not always
value documentation, but grants such as, NSERC's CRDPJ 447533-13, have allowed
me to demonstrate its value and to provide techniques to improve its quality.
For instance, \citet{SmithEtAl2015-SS-TR}
shows that the quality of statistical software for psychology is generally
improved when developed using the structured CRAN (Comprehensive R Archive
Network) process and tools, versus an ad hoc process.
\citet{SmithAndKoothoor2016}
highlights the value of proper documentation by redeveloping nuclear safety
analysis software.  Twenty seven (27) worrisome documentation problems were
found, including incompleteness, ambiguity, inconsistency, verifiability,
modifiability, traceability and a lack of abstraction.  A redevelopment
experiment with five existing projects \citep{SmithJegatheesanAndKelly2016}
enabled the code owners (as ascertained through interviews) to clearly see the
value of documentation.  However, mirroring other studies
\citep{CarverEtAl2007}, the code owners felt documentation takes too much time.

One promising approach to avoid the impending collapse is to use a Model Driven
Engineering (MDE) where MDE is an engineering approach that exploits structured representations
(metamodels for abstract syntax, and constraints to capture static semantics) to
automate repetitive, tedious and error prone tasks.  In particular, code
generation.   approach together with code generation.  Citation supporting
this idea.  My research will dramatically reduce the cost of developing reliable,
reproducible and re-certifiable software by: i) creating infrastructure for a
knowledge base of scientific and computing models; and, ii) writing explicit
``recipes'' that weave together this knowledge to generate theoretical models,
design documents, code, test cases and build scripts. The generator will render
to multiple languages, such as html, LaTeX, RTF for documentation and Python,
Java, C, and Matlab, for code. One source of knowledge, with rules for
transformation, means completeness, consistency and traceability can be achieved by
construction. Moreover, these qualities can be maintained as requirements are
modified, design decisions are changed, documentation standards are varied, and
software is re­certified.  The knowledge will be built up incrementally.  New
projects will reuse existing knowledge and expand as necessary.

Scientists are awakening to the critical importance of SCS sustainability, as
evidenced by the success of Software Carpentry and the Software
Sustainability Institute, but the current solutions are not bold enough.
Although valuable, we need more than studies of developers, training sessions
and project support tools.  \emph{We need transformative change (enabled by
  creating new tools for knowledge capture and document/code generation) where
  documentation, design and verification are expected for all SCS.}

- background and context

- problem statement and research question
- research question: What ideas (if any) from Drasil and Drasil procedures are perceived
by graduate students and end user developers with improving software quality and
productivity - for long lived (safety related?) projects

roadmap of document

\section{Literature Review}

roadmap of this section

\subsection{Current State of Practice for SCS Development}

- how currently done, qualities of interest, add sustainability

The SCS community is finally realizing that current practices are not
sustainable.  \citet{FaulkEtAl2009} observe, ``growing concern about the
reliability of scientific results based on ... software.''  Embarrassing
failures have occurred, like a retraction of derived molecular protein
structures \citep{Miller2006}, false reproduction of sonoluminescent fusion
\citep{PostAndVotta2005}, and fixing and then reintroducing the same error in a
large code base three times in 20 year \citep{MilewiczAndRaybourn2018}.  A
recent report on directions for SCS research and education states: ``While the
volume and complexity of [SCS] have grown substantially in recent decades, [SCS]
traditionally has not received the focused attention it so desperately needs
... to fulfill this key role as a cornerstone of long-term collaboration and
scientific progress'' \citep{RudeEtAl2018}.  Estimates suggest that the number
of released faults per thousand executable lines of code during a given
program’s life cycle is at best 0.1, and more likely 10 to 100 times worse
\citep{Hatton2007}.

Although reproducibility is the cornerstone of the scientific method, until
recently it has not been treated seriously in software
\citep{BenureauAndRougier2017}.  Fortunately, in recent years multiple
conferences, workshops and individuals are calling for dramatic change
\citep{BaileyEtAl2016}.  The need for action is highlighted by a study of 402
computer systems papers - only 48.3\% of the code was both available and
compilable \citep{CollbergEtAl2015}.  (Drasil addresses this problem because as
programming languages evolve the code renderers in Drasil can be updated.)
Reproducibility problems are even more extreme when the goal is replicability.
A third party should be able to repeat a study using only the description of the
methodology from a published article %, with no access to the original code or
%computing environment 
\citep{BenureauAndRougier2017}.  However, replicability is rarely achieved, as
shown for microarray gene expression \citep{IoannidisEtAl2009} and for economics
modelling \citep{IonescuAndJansson2013}.  Drasil addresses completeness and
ambiguity problems, since it emphasizes capturing and documenting all of the
required knowledge, including derivation of equations and
rationales. \citet{CrickAndHall2014} point out potential roadblocks for
reproducibility, including page length constraints and differing detail needs
depending on the audience.  Again Drasil addresses these concerns because the
recipes used to generate the documentation can be tailored to the level of
detail required.

\emph{Although scientists recognize the seriousness of their SCS problems, their
  corrective steps are too incremental.}  For instance, a recent proposal for
the future of High Energy Physics (HEP) software \cite{StewartEtAl2017} uses
words like sustainability, maintainability and reproducibility, but is almost
completely silent on how these qualities are to be achieved.  The proposal
mentions developing new and improved algorithms (including parallel computing
and machine learning), programming tools, recruitment and training, but there is
little on documentation, design or verification techniques (other than unit
testing).  Similarly, a recent proposal on future directions for SCS research
and education \citep{RudeEtAl2018} recognizes the desperate need for change, but
then only suggests training on project management tools, open sharing and
ethics.  \emph{Incremental change is not adequate; we need transformative
  change.}

Thankfully SCS leaders recognize that an interdisciplinary approach provides the
path forward.  They believe that the solution to SCS quality problems is
applying, adapting and developing SE methods, tools and techniques.  However,
typical software processes are a barrier to progress.  ``To break the gridlock,
we must establish a degree of cooperation and collaboration with the [SE]
community that does not yet exist'' \citep{FaulkEtAl2009}.  ``There is a need to
improve the transfer of existing practices and tools from ... [SE] to scientific
programming. In addition, ... there is a need for research to specifically
develop methods and tools that are tailored to the domain'' \citep{Storer2017}.
\emph{This tailored research will require individuals, like myself, that have a
  multidisciplinary background.}

\subsection{MDE and Code Generation}

DSLs and code/document generation provide a transformative technology for
documentation, design and verification \citep{JohansonAndHasselbring2018,
  Smith2018}.  DSLs allow scientists their preferred approach of focusing on
science not software \citep{Kelly2007}.  A generative approach removes the
maintenance nightmare of documentation duplicates and near duplicates
\citep{LucivEtAl2018}, since knowledge is only captured once and automatically
transformed as needed.  Code generation has previously been applied to improve
SCS \citep{WhaleyEtAl2001, Veldhuizen1998, Pueschel2001}.
%   For instance,
% ATLAS (Automatically Tuned Linear Algebra Software) \citep{WhaleyEtAl2001} 
%and
%Blitz++ \citep{Veldhuizen1998} 
% produces efficient and portable linear algebra
% software.  
% Spiral \citep{Pueschel2001} uses software/hardware generation for
% digital signal processing.  
\citet{Carette2008} shows how to generate a family
of efficient, type-safe Gaussian elimination algorithms.  FEniCS (Finite Element
and Computational Software) \citep{LoggEtAl2012} uses code generation when
solving differential equations.  Unlike previous work on SCS code generation,
Drasil will focus on generating all software artifacts (requirements, design
etc.), not just code.

\citet{SzymczakEtAl2016} removes excuses for avoiding documentation by providing
transformative SCS development technology.  Drasil provides an infrastructure
for knowledge capture and document/code generation.  Jacques Carette and I,
along with our students, have developed a prototype (available on GitHub),
implemented via Domain Specific Languages (DSLs) embedded in Haskell.  Drasil
already generates requirements documentation and code for several case studies,
including simulating the temperature of a solar water heating system, glass
breakage and two-dimensional game physics.  The full documentation
(requirements, design etc), code and test cases have been created manually for
each case study.  These manual case studies (many from \citet{SmithJegatheesanAndKelly2016})
provide the `gold standard' against which Drasil is tested.  All of these
artifacts are available publicly.  The generation techniques for Drasil began to
take shape during my work on generating geometric data structures and functions
\citep{CaretteEtAl2011}.

\subsection{Empirical Methods for Software Engineering}

empirical software engineering paper, case study paper

\section{Research Design and Methodologies}

roadmap - fit the different pieces together

\subsection{State of the Practice}

A survey of existing SCS software will identify patterns, find opportunities for
improvement and identify candidate program families for future implementation.
The survey will require developing terminology and a classification system to
answer such questions as: What are the current best practises?  What software
packages are reused and why these packages?  What do practitioners look for in
the tools that support their work?

To achieve objective S2, a measurement template, possibly using pairwise
comparisons of quality metrics between software packages, will be developed for
assessing the quality of the documentation, code, test cases and development
processes for existing SCS program families.  The quality measurement template
will be used to assess the quality of approximately 30 different SCS families.
About half of the families will be model oriented and the other half tool
oriented.  Additional details on best practices will be collected by students
interviewing code owners.  Based on preliminary discussions, Dr.\ Jeffrey Carver
from the University of Alabama and some of his colleagues may also participate
in collecting interview data. Once all the surveys and interviews are
completed, a meta-analysis will draw conclusions for each domain and between
domains.  Knowing the state of practice for family development will highlight
which domains are using best practices that should be emulated in LSS.

The plan here is to return to the task of measuring/assessing the state of
software development practice in several scientific computing domains.  We will
update the work that was done previously for domains such as Geographic
Information Systems~\cite{SmithEtAl2018_arXivGIS}, Mesh
Generators~\cite{SmithEtAl2016}, Seismology software~\cite{SmithEtAl2018}, and
Statistical software for psychology~\cite{SmithEtAl2018_StatSoft}.  We could
return to these domains and/or introduce new domains.  Potential domains/collaborators
are listed in the subsections.

MEng students will be asked to assess the state of practice in each domain.
Last time each student measured two domains, so we would start off with this
model again.  If the scope becomes large enough, we might switch to one domain
per student.

In the previous project, we measured 30 software projects for each domain.  With
the increased scrutiny required in this re-boot, we won't likely be able to
measure the details on this high a number of software projects.  We may still
start with 30 software examples in each domain, but then use some criteria to
create a shorter sub-list for detailed measurement.  The details of how to
proceed here would be determined and documented in the measurement protocol.

In the previous state of practice assessing exercise, a series of questions and
simple metrics was devised to measure the quality of the documentation and
adherence to best software development practices.  The new project will
critically assess the previous set of questions and revise as necessary.  In
addition, the following data will be collected/developed for each domain:

\begin{itemize}
\item Characterization of the functionality provided by the software in the
  domain.  What services does each member of the domain provide?  Ideally this
  information would be summarized in a commonality analysis document.  The
  document will summarize the domain software via its commonalities,
  variabilities and parameters of variation.
\item Usability testing of each software in the domain.  Specific metrics for
  assessing usability still need to be researched, but measures that involve the
  time to complete a task and the users overall impression would be considered.
\item Collect empirical software engineering related data, such as the number of
  files, number of lines of code, cyclomatic complexity, number of open issues,
  etc.  As a starting point for tool support, HubListener
  (\url{https://github.com/pjmc-oliveira/HubListener}) could be used.
\item Collect at least one empirical measure of the quality of the documentation
  - the number of lines of documentation.  [Has anyone previously looked at this
  metric?]
\end{itemize}

The above data will be combined to rank the software in the domain.  The
specifics are yet to be determined, but in the past the Analytic Hierarchy
Process proved helpful in this.

Key to the success of this exercise will be to involve and engage a partner for
each state of practice project.  In the previous effort we didn't involve domain
experts with the rationale of excluding potential bias, but this advantage is
not worth not being able to evaluate the functionality/usability of the software.
Moreover, not having an expert makes publication more difficult, since there is
no one to advice on how best to approach journals and publishers.  The domain
expert will be asked to help in the following ways:

\begin{itemize}
\item Review the protocol for assessing/measuring quality.  This same protocol
  will be used in each domain.
\item Provide their expertise on potential publication.  Specifically, they will
  recommend a suitable journal and act as the corresponding author for any paper
  submissions. 
\item Provide an authoritative list of domain software, possibly augmented by
  existing on-line lists.
\item Provide some assistance from their own team of supervised students to
  facilitate software testing and domain characterization.  Assessing the
  functionality and measuring usability will require an individual with domain
  knowledge.  Multiple measurements will be necessary to have confidence.  The
  measurements will likely take a few hours over a few days of time from each
  student volunteer.
\end{itemize}

There is no budget for this project, but the student volunteers will be
considered co-authors in the resulting paper.  Having them as co-authors also
means that ethics approval should not be necessary.

GitHub will be used to coordinate the work of the large team of people that will
be involved in this project.  In addition to the project record left on GitHub,
the final data will be exported to Mendeley.

Potential collaborators/domains are listed in the following subsections.

\subsection{Medical Imaging}

Medical imaging and analysis software (Dr.\ Michael Noseworthy).

\subsection{Climate Modelling Software}

Climate modelling software (Dr.\ Zoe Li or Dr.\ Xander Wang).

\subsection{Computational Medicine}

Extraction of 3D geometries from medical images (Dr.\ Zahra Motamed).

\subsection{Finite Element Analysis}

Finite element solvers and/or mesh generators (Dr.\ Dieter Stolle).  Potential
software includes FeNiCS, MOOSE, FreeFEM++, Dolphin, etc.  Install the packages
and use them to do something non-trivial.

\subsection{Psychology Software}

Statistics software for psychology (Dr.\ Karin Humphreys).

\subsection{Chemical Engineering Software}

Li Xi (\url{https://www.eng.mcmaster.ca/chemeng/people/faculty/li-xi}).

\subsection{Geographic Information Systems}

Geographic Information Systems (Jason Brodeur and/or Patrick DeLuca).

\subsection{Solidification and Casting Software}

Solidification and casting software (Dr.\ Andr\'e Phillion).

\subsection{Quantum Chemistry Software}

Quantum chemistry software (Dr.\ Paul Ayers).

\subsection{Impact of MDE on SCS}

roadmap

\subsection{Fitting Drasil into the Scientific Software Development Process}

Scientists generally use an approximation of agile methods for their software
development.  In addition, documentation is generally not emphasized.  How
should Drasil be fit into the scientific software development process?  The
Drasil process should encourage an early intensity of thought and effort, but at
the same time, it shouldn't require waterfall development.  Drasil supports
frequent chance and incremental development, but where should developers start
and how should the work proceed?  How does one create an almost empty project in
Drasil and then borrow from existing knowledge and create new knowledge?  This
project could start with something like the task described in~\ref{SecPDF}.

Drasil process should move importance of requirements and domain analysis
earlier in the process.  The quality of the resulting software depends on the
quality of the requirements.  As is known from historical data, the greatest
return on investment is at the requirements state (Boehm papers).
Unfortunately, developers don't like requirements.  Quite possibly they don't
like the intensity of thought that is necessary to get the requirements right.
Inspection techniques, such as task based inspection (Kelly et al) hold promise
for getting the scientist to think deeply at an earlier stage in the development
process.

\subsection{etc.}

Experimentally Determine Appropriate Documentation Template: The best approach
for documenting requirements and design is an open question.  Drasil could be
used to attempt to address this question.  Different recipes for documentation
could be developed and compared.  Some comparison could be automated between the
documents, such as empirical measures of their quality
(Section~\ref{SecEmpiricalMeasureDoc}).  Other measures would require some kind
of usability experiments.

Empirical Measurement of Documentation: Considerable effort has been invested
into developing metrics, like cyclomatic complexity, to attempt to quantify code
quality, but no measures (as far as I know) exist for quantifying the quality of
documentation.  (Khedri does have some work with Bhahati (sp?) Sanga that
quantifies the connectivity of the documentation via the traceability graph.)
Likely the reason that effort has not been invested to quantifying the quality
of documentation is that is not represented formally.  With Drasil much of the
documentation is formal.  This should facilitate automatic measures of
documentation quality, or at least metrics that might be correlated with
quality.

Measuring the Impact of Drasil on Developing Scientific Software: This project
is challenging to design, but would be HUGE if we were successful.  We assume
that Drasil will improve the quality of scientific software and the efficiency
of its production, but data to back this up would go a long way to convincing
the scientific community.  Experiments where a given project is developed in
parallel by two independent teams come to mind, but this would be expensive, and
difficult to control.  Some additional thought, and resources, are necessary
here.  If we could do a project where the scientist, or a student under their
supervision, is a true partner in the development, we might get more convincing
anecdotal evidence.

To quantify the quality of the generated documentation, measures of information
compression will be used.  The measure will be based on the length of the
(family of) generated documentation, code and test cases, compared to the length
of the generator itself.  Our measure will be based on the Minimum Description
Length (MDL) Principle, which states that ``the more we are able to compress the
data, the more we have learned about the data'' \citep{Grunwald2004}.  Another
important measure will be whether the generated documentation achieves
reproducibility. The proposed measure is reproducibility depth
\citep{Soergel2014}.  A depth of 1 means that an independent researcher can
start from the generated code and calculate the same results.  Achieving deeper
reproducibility will mean the same results can be calculated for an independent
researcher starting with the design documentation.  Each higher measure gets
further from the source code, until the highest depth (replicability), which
starts from the requirements and leads to identical (within acceptable error)
results.

\subsection{Impact of Assurance Cases for Building Confidence}

Some SCS needs to be verifiably safe.  We will achieve this by building a safety
assurance case. An assurance case is an explicit structured argument pertaining
to specific properties, such as trustworthiness

Assurance cases have potential to be used in SC to improve our collective
confidence in the developed software~\cite{SmithEtAl2018_ICSEPoster}.  Further
work is necessary to demonstrate the value of assurance cases, but they
certainly hold promise.  Drasil could go a long way to supporting development of
assurance cases.  Some potential Drasil related support ideas include the
following:

\begin{itemize}
\item Automatic generation of visual depiction of the assurance case from the
  goals, claims, evidence etc.
\item Automated reporting of missing evidence, incomplete arguments, etc.
\item Coordination of requirements, design decisions, etc.  Assurance cases use
  the same information as recommended in ``document driven'' design, but the
  traceability and connection between the data is even more important.  A tool
  like Drasil could go a long way in facilitating the presentation and
  navigation of an assurance case.
\end{itemize}

Assurance cases in Drasil could potentially go beyond the scientific context.
For instance, they could also incorporate safety cases for
control systems.  This could potentially be of interest to McSCert.

\subsection{Impact of Computational Variability Testing}

This activity supports objective S3.  The applicant's students will apply LSS
and code generation to Computational Variability Testing (CVT), using the
example of a family of finite element programs.  As proposed by the applicant,
CVT uses code generation to build confidence in the generated code in an
analogous way to the use of grid refinement studies.  Grid refinement looks at
how the solution changes by varying the run-time parameter of grid density and
comparing the results to theoretical expectations.  CVT, on the other hand, can
generate code to ``refine'' build time parameters, such as order of
interpolation, or degree of implicitness.  These parameters can be
systematically varied and the results compared against the expected trend.
Varying computational variabilities is not a new idea, but it is only recently
that the option has become available of coupling this with code generation, so
that the validity of an entire program family can be assessed at the same time.
The experimentation with CVT will initially use the FEniCS
package~\citep{LoggEtAl2012} for elliptical partial differential equations to
compare elements with different orders of interpolation to verify that the
convergence to the solution agrees with the theoretical expectations.

{Design and Documentation For Family of Data Fitting Algorithms}: The fitting
used in GlassBR and in SFS (Software for Solidification) could be made much more
generic.  We could have a family of fitting algorithms that could be used in any
situation where fitting is required.  A proper commonality analysis of this
domain could potentially show the potential design decisions that bridge between
the requirements and the design.  In the SFS example many different fitting
routines were tried.  If the experiments could have been done easily via a
declarative specification, considerable time would have been saved.  If the
experiments are combined with automated testing and ``properties of a correct
solution'' the human involvement could be reduced, so that we have partially
automated algorithm selection.

\section{Implications and Contributions to Knowledge}

We are currently putting too much trust in SCS.  Examples such as nuclear safety
analysis and computational medicine show the significance of SCS for health and
safety.  Although SCS developers do excellent work, we do not currently have
enough checks and balances in place for full confidence.  More should be
expected for documentation, design and testing.  Fortunately, SE offers
techniques, like DSLs and code generation, that can be employed to not only
improve SCS quality, but also reduce development costs.  Now is the time for
change.  SCS practitioners have recognized problems with the status quo and
multidisciplinary researchers are needed to bridge the gap between SE and SCS.

Codifying medical, biomechanical and computational knowledge is challenging, but
success will completely transform the development of safety-related software in
medicine, science and engineering.  MDE will remove the drudgery of generating
and maintaining documentation, code, test cases and build environments, enabling
scientists to focus on science, engineers to focus on engineering and medical
professionals to focus on medicine.  MDE will detect inconsistencies between
models via inter-model consistency constraints.  As a consequence, mistakes like
inconsistent units or assumptions cannot occur.  MDE will raise the bar so that
we can expect an explicit argument for safety, not just code that we are
expected to blindly trust.  With the right up-front investment of knowledge
capture, we can have software that is long lived because the stable knowledge is
separated from the rapidly changing modelling assumptions and design decisions.

\section{Research Schedule}

\bibliographystyle {plainnat}
\bibliography {ResearchProposal}

\end{document}

### 2019-09-11 ResearchProposal.tex
Additional changes to structure, additions to background, editing of introduction

@@
 \usepackage[round]{natbib}
+\usepackage{booktabs}
 
@@
     linkcolor=red,          % color of internal links (change box color with linkbordercolor)
-    citecolor=green,       % color of links to bibliography
+    citecolor=blue,       % color of links to bibliography
     filecolor=magenta,   % color of file links
@@
 
+\newpage
+
 \section{Introduction}
@@
 
-Quality of SCS improved through documentation.  SCS developers do not always
-value documentation, but grants such as, NSERC's CRDPJ 447533-13, have allowed
-me to demonstrate its value and to provide techniques to improve its quality.
-For instance, \citet{SmithEtAl2015-SS-TR}
-shows that the quality of statistical software for psychology is generally
-improved when developed using the structured CRAN (Comprehensive R Archive
-Network) process and tools, versus an ad hoc process.
-\citet{SmithAndKoothoor2016}
-highlights the value of proper documentation by redeveloping nuclear safety
-analysis software.  Twenty seven (27) worrisome documentation problems were
-found, including incompleteness, ambiguity, inconsistency, verifiability,
-modifiability, traceability and a lack of abstraction.  A redevelopment
-experiment with five existing projects \citep{SmithJegatheesanAndKelly2016}
-enabled the code owners (as ascertained through interviews) to clearly see the
-value of documentation.  However, mirroring other studies
-\citep{CarverEtAl2007}, the code owners felt documentation takes too much time.
+Documentation can address the confidence problem mentioned above.  For software
+in general, documentation, written before and during development, provides many
+benefits~\cite{Parnas2010}: easier reuse of old designs, better communication
+about requirements, more useful design reviews, easier integration of separately
+written modules, more effective code inspection, more effective testing, and
+more efficient corrections and improvements.  For SCS in particular, developers
+do not always value documentation, but documentation has been shown to help
+improve SCS.  For instance, \citet{SmithEtAl2015-SS-TR} shows that the quality
+of statistical software for psychology is generally improved when developed
+using the structured CRAN (Comprehensive R Archive Network) process and tools,
+versus an ad hoc process.  \citet{SmithAndKoothoor2016} highlights the value of
+proper documentation by redeveloping nuclear safety analysis software.  Twenty
+seven (27) worrisome documentation problems were found, including
+incompleteness, ambiguity, inconsistency, verifiability, modifiability,
+traceability and a lack of abstraction.  A redevelopment experiment with five
+existing projects \citep{SmithJegatheesanAndKelly2016} enabled the code owners
+(as ascertained through interviews) to clearly see the value of documentation.
+However, mirroring other studies \citep{CarverEtAl2007}, the code owners felt
+documentation takes too much time.
 
-One promising approach to avoid the impending collapse is to use a Model Driven
-Engineering (MDE) where MDE is an engineering approach that exploits structured representations
+One promising approach to gain the benefits of documentation, without the often
+infeasible time commitment, is to use a Model Driven Engineering (MDE), where
+MDE is an engineering approach that exploits structured representations
 (metamodels for abstract syntax, and constraints to capture static semantics) to
-automate repetitive, tedious and error prone tasks.  In particular, code
-generation.   approach together with code generation.  Citation supporting
-this idea.  My research will dramatically reduce the cost of developing reliable,
-reproducible and re-certifiable software by: i) creating infrastructure for a
-knowledge base of scientific and computing models; and, ii) writing explicit
-``recipes'' that weave together this knowledge to generate theoretical models,
-design documents, code, test cases and build scripts. The generator will render
-to multiple languages, such as html, LaTeX, RTF for documentation and Python,
-Java, C, and Matlab, for code. One source of knowledge, with rules for
-transformation, means completeness, consistency and traceability can be achieved by
-construction. Moreover, these qualities can be maintained as requirements are
-modified, design decisions are changed, documentation standards are varied, and
-software is re­certified.  The knowledge will be built up incrementally.  New
-projects will reuse existing knowledge and expand as necessary.
+automate repetitive, tedious and error prone tasks.  In this research proposal,
+we will use MDE as a short-hand to include model transformation, Domain Specific
+Languages (DSLs) and code generation.  Several current papers point to DSLs and
+code/document generation as a transformative technology for documentation,
+design and verification of SCS \citep{JohansonAndHasselbring2018, Smith2018}.
+The benefit of an MDE approach is to dramatically reduce the cost of developing
+reliable, reproducible and re-certifiable software by: i) creating
+infrastructure for a knowledge base of scientific and computing models; and, ii)
+writing explicit ``recipes'' that weave together this knowledge to generate
+theoretical models, design documents, code, test cases and build scripts. The
+generator can render to multiple languages, such as html, LaTeX, RTF for
+documentation and Python, Java, C, and Matlab, for code. One source of
+knowledge, with rules for transformation, means completeness, consistency and
+traceability can be achieved by construction. Moreover, these qualities can be
+maintained as requirements are modified, design decisions are changed,
+documentation standards are varied, and software is re­certified.  The knowledge
+will be built up incrementally.  New projects will reuse existing knowledge and
+expand as necessary.
 
@@
 
-\section{Literature Review}
+\section{Background and Literature Review}
 
-roadmap of this section
+roadmap of this section.  Some overview of the topics themselves, but an
+emphasis on how they appear in the SCS context.
+
+\subsection{Software Qualities of Interest}
+
+For instance,
+a rational design process can provide quality improvements, and Quality
+Assurance (QA), as shown in Table~\ref{Tbl_Qualities}.  Moreover, documentation,
+written before and during development, can provide many
+benefits~\cite{Parnas2010}: easier reuse of old designs, better communication
+about requirements, more useful design reviews, easier integration of separately
+written modules, more effective code inspection, more effective testing, and
+more efficient corrections and improvements.
+
+- qualities of interest, add sustainability
+
+\begin{description}%[leftmargin=\the\parindent]
+
+\item [\textbf{Verifiability}] involves ``solving the equations
+  right''~\cite[p.~23]{Roache1998}; it benefits from rational documentation
+  that systematically shows, with explicit traceability, how the governing
+  equations are transformed into code.
+
+\item [\textbf{Validatability}] means ``solving the right
+  equations''~\cite[p.~23]{Roache1998}.  Validatability is improved by a
+  rational process via clear documentation of the theory and assumptions, along
+  with an explicit statement of the systematic steps required for experimental
+  validation.
+
+% \item [\textbf{Reliability}] is a critical quality for scientific software,
+%   since the results of computations are meaningless, if they are not dependable.
+%   Reliability is closely tied to verifiability, since the key quality to verify
+%   is reliability, while the act of verification itself improves reliability.
+
+% \item [\textbf{Performance}] considerations can make certification challenging, since QA
+%   becomes more difficult for more complex code.  However,
+%   as Roache~\cite[p.~355]{Roache1998} points out, using simpler algorithms and
+%   reducing the number of options in general purpose code, is not always a
+%   practical option.
+
+\item [\textbf{Usability}] can be a problem.  Different users, solving the same
+  physical problem, using the same software, can come up with different answers,
+  due to differences in parameter selection~\cite[p.~370]{Roache1998}.  To
+  reduce misuse, a rational process must state expected user characteristics,
+  modelling assumptions, definitions and the range of applicability of the code.
+
+\item [\textbf{Maintainability}] is necessary in scientific software, since change,
+  through iteration, experimentation and exploration, is inevitable.  Models of
+  physical phenomena and numerical techniques necessarily evolve over
+  time~\cite{CarverEtAl2007, SegalAndMorris2008}.  Proper documentation,
+  designed with change in mind, can greatly assist with change management.%   QA
+  % activities need to take the need for creativity into account, while not
+  % smothering it~\cite[p.~352]{Roache1998}.
+
+\item [\textbf{Reusability}] provides support for the quality of reliability,
+  since reliability is improved by reusing trusted components~\cite{Dubois2005}.
+  (Care must still be taken with reusing trusted components, since blind reuse
+  in a new context can lead to errors, as dramatically shown in the Ariane 5
+  disaster~\cite[p.~37--38]{OliveiraAndStewart2006}.)  The odds of reuse are
+  improved when it is considered right from the start.
+
+\item [\textbf{Understandability}] is necessary, since reviewers can only certify
+  something they understand.  Scientific software developers have the
+  view ``that the science a developer embeds in the code must be apparent to
+  another scientist, even ten years later''~\cite{Kelly2013}.
+  Understandability applies to the documentation and code, while usability
+  refers to the executable software.  Documentation that follows a rational
+  process is the easiest to follow.
+
+\item [\textbf{Reproducibility}] is a required component of the scientific
+  method~\cite{Davison2012}.  Although QA has, ``a bad name among
+  creative scientists and engineers''~\cite[p.~352]{Roache1998}, the community
+  need to recognize that participating in QA management also improves
+  reproducibility.  Reproducibility, like QA, benefits from a consistent and
+  repeatable computing environment, version control and separating code from
+  configuration/parameters~\cite{Davison2012}.
+\end{description}
+
+\subsection{Desirable Qualities of Documentation}
+
+To address the challenges for adequate documentation for QA and certification
+efforts for SCS, the qualities described below need to be a priority.  All but
+the final quality listed (abstraction), are adapted from the IEEE recommended
+practise for producing good software requirements~\cite{IEEE1998}.
+
+\begin{itemize}
+
+\item \emph{Completeness}: Documentation is said to be complete when all the
+  requirements of the software are detailed. That is, each goal, functionality,
+  attribute, design constraint, value, data, model, symbol, term (with its unit
+  of measurement if applicable), abbreviation, acronym, assumption and
+  performance requirement of the software is defined.  The software's response
+  to all classes of inputs, both valid and invalid and for both desired and
+  undesired events, also needs to be specified.
+
+\item \emph{Consistency}: Documentation is said to be consistent when no subset
+  of individual statements are in conflict with each other. That is, a
+  specification of an item made at one place in the document should not
+  contradict the specification of the same item at another location.
+
+\item \emph{Modifiability}: The documentation should be developed in such a way
+  that it is easily modifiable so that likely future changes do not destroy the
+  structure of the document. Also it should be easy to reflect the change,
+  wherever needed in the document to maintain consistency, traceability and
+  completeness. For documentation to be modifiable, its format must be
+  structured in a way that repetition is avoided and cross-referencing is
+  employed.
+
+\item \emph{Traceability}: Documentation should be traceable, as this
+  facilitates maintenance and review. If a change is made to the design or code
+  of the software, then all the documentation relating to those segments have to
+  be modified.  This property is also important for recertification.
+
+\item \emph{Unambiguity}: Documentation is said to be unambiguous only when
+  every requirement's specification has a unique interpretation.  The
+  documentation should be unambiguous to all audiences, including developers,
+  users and reviewers.
+
+\item \emph{Correctness}: There is no direct tool or method for measuring
+  correctness. One way of building confidence in correctness is by reviewing to
+  ensure that each requirement stated is one that the stakeholders and experts
+  desire.  By maintaining traceability, consistency and unambiguity, we can
+  reduce the occurrence of errors and make the goal of reviewing for correctness
+  easier.
+
+\item \emph{Verifiability}: Every requirement in the documentation must be the
+  one fulfilled by the implemented software. Therefore all the requirements
+  should be clear, unambiguous and testable, so that a person or a machine can
+  verify whether the software product meets the requirements.
+
+\item \emph{Abstract}: Documented requirements are said to be abstract if they
+  state what the software must do and the properties it must possess, but do not
+  speak about how these are to be achieved. For example, a requirement can
+  specify that an Ordinary Differential Equation (ODE) must be solved, but it
+  should not mention that Euler's method should be used to solve the ODE. How to
+  accomplish the requirement is a design decision, which is documented during
+  the design phase.
+
+\end{itemize}
 
 \subsection{Current State of Practice for SCS Development}
-
-- how currently done, qualities of interest, add sustainability
 
@@
   multidisciplinary background.}
+
+\subsection{Documentation for SCS}
+
+Table~\ref{TblDocuments} shows the recommended documentation for a scientific
+software project.  The documents are typical of what is suggested for scientific
+software certification, where certification consists of official recognition by
+an authority, or regulatory body, that the software is fit for its intended use.
+For instance, the Canadian Standards Association (CSA) requires a similar set of
+documents for quality assurance of scientific programs for nuclear power
+plants~\cite{CSA1999}.
+
+\begin{table}[!h]
+  \caption{Recommended Documentation} \label{TblDocuments}
+
+  \begin{tabular}{l p{8cm}}
+    %\hline
+    %\toprule
+    \textit{Problem Statement} & Description of problem to be solved\\
+    \textit{Development Plan} & Overview of development process/infrastructure\\
+    \textit{Requirements} & Desired functions and qualities of the software\\
+    \textit{V\&V Plan} & Verification that all documentation artifacts,
+                           including the code, are internally correct.
+                           Validation, from an external viewpoint, that the right
+                           problem, or model, is being solved.\\
+    \textit{Design Specification} & Documentation of how the requirements are to
+                                    be realized, through both a software
+                                    architecture and detailed design of modules
+                                    and their interfaces\\
+    \textit{Code} & Implementation of the design in code\\
+    \textit{V\&V Report} & Summary of the V\&V efforts,
+                             including testing\\
+    \textit{User Manual} & Instructions on installation, usage; worked examples\\
+    %\hline
+    %\bottomrule
+  \end{tabular}
+\end{table}
+
+To achieve the qualities listed in Table~\ref{Tbl_Qualities}, the documentation
+in Table~\ref{TblDocuments} should have the following qualities:
+\emph{complete}, \emph{correct}, \emph{consistent}, \emph{modifiable},
+\emph{traceable}, \emph{unambiguous}, and \emph{verifiable}.  All of these
+qualities are listed in the IEEE recommended practice for software
+requirements~\cite{IEEE1998}.  The IEEE guidelines are for requirements, but
+most qualities are relevant for all documentation artifacts.  Another relevant
+quality, which is not on the IEEE list, is \emph{abstract}.  Requirements should
+state what is to be achieved, but be silent on how it is to be achieved.
+Abstraction is an important software development principle for dealing with
+complexity~\cite[p.~40]{GhezziEtAl2003}.  Smith and Koothoor present further
+details on the qualities of documentation for scientific
+software~\cite{SmithAndKoothoor2016}.
 
@@
 
+Determine criteria for what makes a domain a good fit.
+
 Potential collaborators/domains are listed in the following subsections.
 
-\subsection{Medical Imaging}
+\subsubsection{Medical Imaging}
 
-Medical imaging and analysis software (Dr.\ Michael Noseworthy).
+Medical imaging and analysis software (Dr.\ Michael Noseworthy)?
 
-\subsection{Climate Modelling Software}
+\subsubsection{Climate Modelling Software}
 
-Climate modelling software (Dr.\ Zoe Li or Dr.\ Xander Wang).
+Climate modelling software (Dr.\ Zoe Li or Dr.\ Xander Wang)?
 
-\subsection{Computational Medicine}
+\subsubsection{Computational Medicine}
 
-Extraction of 3D geometries from medical images (Dr.\ Zahra Motamed).
+Extraction of 3D geometries from medical images (Dr.\ Zahra Motamed)?  Lattice
+Boltzmann Solver?
 
-\subsection{Finite Element Analysis}
+\subsubsection{Finite Element Analysis}
 
-Finite element solvers and/or mesh generators (Dr.\ Dieter Stolle).  Potential
+Finite element solvers and/or mesh generators (Dr.\ Dieter Stolle)?  Potential
 software includes FeNiCS, MOOSE, FreeFEM++, Dolphin, etc.  Install the packages
@@
 
-\subsection{Psychology Software}
+\subsubsection{Psychology Software}
 
-Statistics software for psychology (Dr.\ Karin Humphreys).
+Statistics software for psychology (Dr.\ Karin Humphreys)?
 
-\subsection{Chemical Engineering Software}
+\subsubsection{Chemical Engineering Software}
 
-Li Xi (\url{https://www.eng.mcmaster.ca/chemeng/people/faculty/li-xi}).
+Li Xi (\url{https://www.eng.mcmaster.ca/chemeng/people/faculty/li-xi})?
 
-\subsection{Geographic Information Systems}
+\subsubsection{Geographic Information Systems}
 
-Geographic Information Systems (Jason Brodeur and/or Patrick DeLuca).
+Geographic Information Systems (Jason Brodeur and/or Patrick DeLuca)?
 
-\subsection{Solidification and Casting Software}
+\subsubsection{Physical Metallurgy Software}
 
-Solidification and casting software (Dr.\ Andr\'e Phillion).
+Solidification and casting software (Dr.\ Andr\'e Phillion)?
 
@@
 
-Quantum chemistry software (Dr.\ Paul Ayers).
+Quantum chemistry software (Dr.\ Paul Ayers)?
 
@@
 
+\newpage
+
 \section{Research Schedule}

### 2019-09-16 ResearchProposal.tex
Additional edits to research proposal - fleshing out introductory sections

@@
 exercise, because of a lack of explicit, effort reducing, traceability
-information.
+information.  The existing challenges for modification, maintenance and
+extension point to problems with the sustainability of SCS.
 
-Documentation can address the confidence problem mentioned above.  For software
+Documentation can improve software sustainability.  For software
 in general, documentation, written before and during development, provides many
@@
 written modules, more effective code inspection, more effective testing, and
-more efficient corrections and improvements.  For SCS in particular, developers
-do not always value documentation, but documentation has been shown to help
-improve SCS.  For instance, \citet{SmithEtAl2015-SS-TR} shows that the quality
-of statistical software for psychology is generally improved when developed
-using the structured CRAN (Comprehensive R Archive Network) process and tools,
-versus an ad hoc process.  \citet{SmithAndKoothoor2016} highlights the value of
-proper documentation by redeveloping nuclear safety analysis software.  Twenty
-seven (27) worrisome documentation problems were found, including
-incompleteness, ambiguity, inconsistency, verifiability, modifiability,
-traceability and a lack of abstraction.  A redevelopment experiment with five
-existing projects \citep{SmithJegatheesanAndKelly2016} enabled the code owners
-(as ascertained through interviews) to clearly see the value of documentation.
-However, mirroring other studies \citep{CarverEtAl2007}, the code owners felt
-documentation takes too much time.
+more efficient corrections and improvements. For SCS in particular,
+documentation provides benefits.  For instance, \citet{SmithEtAl2015-SS-TR}
+shows that the quality of statistical software for psychology is generally
+improved when developed using the structured CRAN (Comprehensive R Archive
+Network) process and tools, versus an ad hoc process.
+\citet{SmithAndKoothoor2016} highlights the value of proper documentation by
+redeveloping nuclear safety analysis software.  Twenty seven (27) worrisome
+documentation problems were found, including incompleteness, ambiguity,
+inconsistency, verifiability, modifiability, traceability and a lack of
+abstraction.  A redevelopment experiment with five existing projects
+\citep{SmithJegatheesanAndKelly2016} enabled the code owners (as ascertained
+through interviews) to clearly see the value of documentation.  However,
+mirroring other studies \citep{CarverEtAl2007}, the code owners felt
+documentation takes too much time.  Due to the time commitment and resource
+limitation, documentation is rarely emphasized in SCS.
 
-One promising approach to gain the benefits of documentation, without the often
-infeasible time commitment, is to use a Model Driven Engineering (MDE), where
-MDE is an engineering approach that exploits structured representations
-(metamodels for abstract syntax, and constraints to capture static semantics) to
-automate repetitive, tedious and error prone tasks.  In this research proposal,
-we will use MDE as a short-hand to include model transformation, Domain Specific
-Languages (DSLs) and code generation.  Several current papers point to DSLs and
-code/document generation as a transformative technology for documentation,
-design and verification of SCS \citep{JohansonAndHasselbring2018, Smith2018}.
-The benefit of an MDE approach is to dramatically reduce the cost of developing
-reliable, reproducible and re-certifiable software by: i) creating
-infrastructure for a knowledge base of scientific and computing models; and, ii)
-writing explicit ``recipes'' that weave together this knowledge to generate
-theoretical models, design documents, code, test cases and build scripts. The
-generator can render to multiple languages, such as html, LaTeX, RTF for
-documentation and Python, Java, C, and Matlab, for code. One source of
+One promising approach to gain the benefits of documentation, but significantly
+reduce the time commitment, is to use a Model Driven Engineering (MDE).  MDE is
+an engineering approach that exploits structured representations (metamodels for
+abstract syntax, and constraints to capture static semantics) to automate
+repetitive, tedious and error prone tasks.  In this research proposal, we will
+use MDE as a short-hand to include knowledge models, model transformation,
+Domain Specific Languages (DSLs) and code generation.  Several current papers
+point to DSLs and code/document generation as a transformative technology for
+documentation, design and verification of SCS \citep{JohansonAndHasselbring2018,
+  Smith2018}.  The benefit of an MDE approach is to dramatically reduce the cost
+of developing reliable, reproducible and re-certifiable software.
+
+One example of applying MDE to SCS is Drasil, which is introduced in
+\citet{SzymczakEtAl2016} and available at
+\href{https://github.com/JacquesCarette/Drasil}
+{https://github.com/JacquesCarette/Drasil}.  Drasil is implemented as an open
+source set of Domain Specific Languages (DSLs).  The Drasil approach involves:
+i) creating infrastructure for a knowledge base of scientific and computing
+models; and, ii) writing explicit ``recipes'' that weave together this knowledge
+to generate theoretical models, design documents, code, test cases and build
+scripts. The generator can render to multiple languages, such as html, LaTeX,
+RTF for documentation and Python, Java, C, and Matlab, for code. One source of
 knowledge, with rules for transformation, means completeness, consistency and
@@
 maintained as requirements are modified, design decisions are changed,
-documentation standards are varied, and software is re­certified.  The knowledge
-will be built up incrementally.  New projects will reuse existing knowledge and
-expand as necessary.
+documentation standards are varied, and software is re­certified.  Although
+creating the knowledge base is time consuming, the knowledge can be built up
+incrementally.  New projects will reuse existing knowledge and expand as
+necessary.
 
-Scientists are awakening to the critical importance of SCS sustainability, as
-evidenced by the success of Software Carpentry and the Software
-Sustainability Institute, but the current solutions are not bold enough.
-Although valuable, we need more than studies of developers, training sessions
-and project support tools.  \emph{We need transformative change (enabled by
-  creating new tools for knowledge capture and document/code generation) where
-  documentation, design and verification are expected for all SCS.}
+As described above, sustainability is a concern for SCS, and MDE holds promise
+for addressing this concern.  However, a transformative change in the
+development of SCS requires more than promise.  Empirical evidence is necessary
+to demonstrate that an MDE process can improve the sustainability of SCS.  This
+leads to the long-term objective for this research proposal: \emph{assess and
+  measure the impact of MDE on the sustainability of SCS.}
 
-- background and context
+The scope of SCS software for this proposal includes long-lived software, as
+implied by the sustainability objective.  To gain the full benefits of
+generating documentation, the scope also includes software that interests
+multiple stakeholders, with different interests and backgrounds.  Finally, the
+scope of this proposal emphasizes safety related software, such as software for
+nuclear safety analysis, medical imaging and computational medicine.  Safety
+related software is prioritized because documentation is often expected as part
+of a certification exercise.  Although this study is not specifically restricted
+to open-source software, for practical reasons, commercial software will not be
+strongly emphasized.
 
-- problem statement and research question
-- research question: What ideas (if any) from Drasil and Drasil procedures are perceived
-by graduate students and end user developers with improving software quality and
-productivity - for long lived (safety related?) projects
+Given the large number of potential MDE processes, techniques and technologies,
+together with the wide variety in purpose, scope and context for SCS, the
+long-term objective of measuring the impact of MDE is decomposed into four
+short-term objectives.  For each objective, a reference is given to the
+corresponding section in this document that provides the details.
 
-roadmap of document
+\begin{enumerate}
+\item Assess the current state of software development practice for SCS.  How is
+  the software developed?  Are there existing development methods that lead to
+  higher software sustainability? (Section~\ref{SecStateOfThePractice})
+\item Assess the impact of an MDE process on end user developed SCS with respect
+  to developer productivity and software sustainability.  
+  (Section~\ref{SecImpactMDE})
+\item Assess the impact on safety when the MDE generated documentation targets a
+  safety assurance case.  An assurance case is an explicit structured argument
+  pertaining to specific properties, such as trustworthiness.  The evidence for
+  the argument will come from the traceability between artifacts generated by
+  model transformations, expert reviews, test cases,
+  etc. (Section~\ref{SecAssuranceCase})
+\item Assess the impact on software quality when using MDE to facilitate
+  Computational Variability Testing (CVT).  CVT uses code generation to build
+  confidence in the generated code in an analogous way to the use of grid
+  refinement studies.  Grid refinement looks at how the solution changes by
+  varying the run-time parameter of grid density and comparing the results to
+  theoretical expectations.  CVT, on the other hand, can generate code to
+  ``refine'' build time parameters, such as order of interpolation, or degree of
+  implicitness.  These parameters can be systematically varied and the results
+  compared against the expected trend.  (Section~\ref{SecCVT})
+\end{enumerate}
 
-\section{Background and Literature Review}
+Before presenting the details of the short-term objectives
+(Section~\ref{SecMethodologies}), the necessary background information is
+provided, along with a literature review (Section~\ref{SecBackgroundLitReview}).
+The literature review covers software quality (Sections~\ref{SecSoftwareQuality}
+and~\ref{SecDesirableQs}), the literature on the current state of the practice
+for SCS development (Section~\ref{SecCurrStateOfThePractice}), potential target
+documentation artifacts for SCS (Section~\ref{SecDocumentation}), MDE and code
+generation (Section~\ref{SecMDE}) and empirical methods for software engineering
+(Section~\ref{SecEmpirical}).  This document ends with a summary of the
+potential contributions to knowledge from this study
+(Section~\ref{SecContribKnowledge}) and a research schedule
+(Section~\ref{SecSchedule}).
 
-roadmap of this section.  Some overview of the topics themselves, but an
-emphasis on how they appear in the SCS context.
+\section{Background and Literature Review} \label{SecBackgroundLitReview}
 
-\subsection{Software Qualities of Interest}
+To assess the impact of MDE on SCS quality, we need a clear definition of what
+we mean by quality.  Section~\ref{SecSoftwareQuality} shows how the concept of
+quality is decomposed into a set of separate qualities.  This set of qualities
+can be applied to the software artifacts (documentation, test cases, etc) and to
+the software development process itself.  Several of the qualities from the list
+in Section~\ref{SecSoftwareQuality} cannot be measured directly, such as
+maintainability.  Therefore, Section~\ref{SecDesirableQs} introduces measurable
+documentation qualities that are believed to contribute to software quality.  In
+some cases it may be necessary to use the qualities in
+Section~\ref{SecDesirableQs} to indirectly measure qualities listed in
+Section~\ref{SecSoftwareQuality}.
 
-For instance,
-a rational design process can provide quality improvements, and Quality
-Assurance (QA), as shown in Table~\ref{Tbl_Qualities}.  Moreover, documentation,
-written before and during development, can provide many
-benefits~\cite{Parnas2010}: easier reuse of old designs, better communication
-about requirements, more useful design reviews, easier integration of separately
-written modules, more effective code inspection, more effective testing, and
-more efficient corrections and improvements.
+focuses on presents it, roadmap of this
+section.  Some overview of the topics themselves, but an emphasis on how they
+appear in the SCS context.
 
-- qualities of interest, add sustainability
+\subsection{Software Qualities of Interest} \label{SecSoftwareQuality}
+
+Our analysis is centred around a set of software qualities.  Quality is not
+considered as a single measure, but a collection of different qualities, often
+called ``ilities.''  These qualities highlight the desirable nonfunctional
+properties for software artifacts, which include both documentation and
+code. Some qualities, such as visibility and productivity, apply to the process
+used for developing the software. The following list of qualities is based on
+\cite{GhezziEtAl2003}. To the list from \cite{GhezziEtAl2003}, we have added
+three qualities important for SC: installability, reproducibility and
+sustainability.
 
 \begin{description}%[leftmargin=\the\parindent]
+
+\item [\textbf{Installability}] A measure of the ease of installation.
+
+\item [\textbf{Correctness}] Software is correct if it matches its specification.
 
@@
 
-% \item [\textbf{Reliability}] is a critical quality for scientific software,
-%   since the results of computations are meaningless, if they are not dependable.
-%   Reliability is closely tied to verifiability, since the key quality to verify
-%   is reliability, while the act of verification itself improves reliability.
+\item [\textbf{Reliability}] is a critical quality for scientific software,
+  since the results of computations are meaningless, if they are not dependable.
+  Reliability is closely tied to verifiability, since the key quality to verify
+  is reliability, while the act of verification itself improves reliability.
 
-% \item [\textbf{Performance}] considerations can make certification challenging, since QA
-%   becomes more difficult for more complex code.  However,
-%   as Roache~\cite[p.~355]{Roache1998} points out, using simpler algorithms and
-%   reducing the number of options in general purpose code, is not always a
-%   practical option.
+\item [\textbf{Performance}] considerations can make certification challenging, since QA
+  becomes more difficult for more complex code.  However,
+  as Roache~\cite[p.~355]{Roache1998} points out, using simpler algorithms and
+  reducing the number of options in general purpose code, is not always a
+  practical option.
 
@@
   configuration/parameters~\cite{Davison2012}.
+
+\item [\textbf{Productivity}] [Fill in]
+
+\item [\textbf{Sustainability}] A combination of other qualities.
+
 \end{description}
 
-\subsection{Desirable Qualities of Documentation}
+\subsection{Desirable Qualities of Documentation} \label{SecDesirableQs}
 
-To address the challenges for adequate documentation for QA and certification
-efforts for SCS, the qualities described below need to be a priority.  All but
-the final quality listed (abstraction), are adapted from the IEEE recommended
+To achieve the qualities listed in Section~\ref{SecSoftwareQuality}, the
+documentation should achieve the qualities listed in this section.  All but the
+final quality listed (abstraction), are adapted from the IEEE recommended
 practise for producing good software requirements~\cite{IEEE1998}.
 
-\begin{itemize}
+\begin{description}
 
-\item \emph{Completeness}: Documentation is said to be complete when all the
+\item [\textbf{Completeness}] Documentation is said to be complete when all the
   requirements of the software are detailed. That is, each goal, functionality,
@@
 
-\item \emph{Consistency}: Documentation is said to be consistent when no subset
+\item [\textbf{Consistency}] Documentation is said to be consistent when no subset
   of individual statements are in conflict with each other. That is, a
@@
 
-\item \emph{Modifiability}: The documentation should be developed in such a way
+\item [\textbf{Modifiability}] The documentation should be developed in such a way
   that it is easily modifiable so that likely future changes do not destroy the
@@
 
-\item \emph{Traceability}: Documentation should be traceable, as this
+\item [\textbf{Traceability}] Documentation should be traceable, as this
   facilitates maintenance and review. If a change is made to the design or code
@@
 
-\item \emph{Unambiguity}: Documentation is said to be unambiguous only when
+\item [\textbf{Unambiguity}] Documentation is said to be unambiguous only when
   every requirement's specification has a unique interpretation.  The
@@
 
-\item \emph{Correctness}: There is no direct tool or method for measuring
+\item [\textbf{Correctness}] There is no direct tool or method for measuring
   correctness. One way of building confidence in correctness is by reviewing to
@@
 
-\item \emph{Verifiability}: Every requirement in the documentation must be the
+\item [\textbf{Verifiability}] Every requirement in the documentation must be the
   one fulfilled by the implemented software. Therefore all the requirements
@@
 
-\item \emph{Abstract}: Documented requirements are said to be abstract if they
+\item [\textbf{Abstract}] Documented requirements are said to be abstract if they
   state what the software must do and the properties it must possess, but do not
@@
 
-\end{itemize}
+\end{description}
 
-\subsection{Current State of Practice for SCS Development}
+\subsection{Literature Review on Current State of the Practice for SCS
+  Development} \label{SecCurrStateOfThePractice}
 
@@
 
-\subsection{Documentation for SCS}
+\subsection{Documentation for SCS} \label{SecDocumentation}
 
@@
 
-\subsection{MDE and Code Generation}
+\subsection{MDE and Code Generation} \label{SecMDE}
 
@@
 transformed as needed.  Code generation has previously been applied to improve
-SCS \citep{WhaleyEtAl2001, Veldhuizen1998, Pueschel2001}.
-%   For instance,
-% ATLAS (Automatically Tuned Linear Algebra Software) \citep{WhaleyEtAl2001} 
-%and
-%Blitz++ \citep{Veldhuizen1998} 
-% produces efficient and portable linear algebra
-% software.  
-% Spiral \citep{Pueschel2001} uses software/hardware generation for
-% digital signal processing.  
-\citet{Carette2008} shows how to generate a family
-of efficient, type-safe Gaussian elimination algorithms.  FEniCS (Finite Element
-and Computational Software) \citep{LoggEtAl2012} uses code generation when
-solving differential equations.  Unlike previous work on SCS code generation,
-Drasil will focus on generating all software artifacts (requirements, design
-etc.), not just code.
+SCS \citep{WhaleyEtAl2001, Veldhuizen1998, Pueschel2001}.  For instance, ATLAS
+(Automatically Tuned Linear Algebra Software) \citep{WhaleyEtAl2001} and Blitz++
+\citep{Veldhuizen1998} produces efficient and portable linear algebra software.
+Spiral \citep{Pueschel2001} uses software/hardware generation for digital signal
+processing.  \citet{Carette2008} shows how to generate a family of efficient,
+type-safe Gaussian elimination algorithms.  FEniCS (Finite Element and
+Computational Software) \citep{LoggEtAl2012} uses code generation when solving
+differential equations.  Unlike previous work on SCS code generation, Drasil
+will focus on generating all software artifacts (requirements, design etc.), not
+just code.
 
@@
 
-\subsection{Empirical Methods for Software Engineering}
+\subsection{Empirical Methods for Software Engineering} \label{SecEmpirical}
 
@@
 
-\section{Research Design and Methodologies}
+\section{Research Design and Methodologies} \label{SecMethodologies}
 
@@
 
-\subsection{State of the Practice}
+\subsection{State of the Practice} \label{SecStateOfThePractice}
 
@@
 
-\subsection{Quantum Chemistry Software}
+\subsubsection{Quantum Chemistry Software}
 
@@
 
-\subsection{Impact of MDE on SCS}
+\subsection{Impact of MDE on SCS} \label{SecImpactMDE}
 
@@
 
-\subsection{Fitting Drasil into the Scientific Software Development Process}
+Research Question: What tools and methods from MDE are perceived by end user developers to
+  improve their productivity and the sustainability of their software.
+
+\subsubsection{Fitting Drasil into the Scientific Software Development Process} \label{Sec}
 
@@
 
-\subsection{etc.}
+\subsubsection{etc.}
 
@@
 
-\subsection{Impact of Assurance Cases for Building Confidence}
+\subsection{Impact of Assurance Cases for Building Confidence} \label{SecAssuranceCase}
 
@@
 
-\subsection{Impact of Computational Variability Testing}
+\subsection{Impact of Computational Variability Testing} \label{SecCVT}
 
@@
 
-\section{Implications and Contributions to Knowledge}
+\section{Implications and Contributions to Knowledge} \label{SecContribKnowledge}
 
@@
 
-\section{Research Schedule}
+\section{Research Schedule} \label{SecSchedule}
 

### 2019-09-17 ResearchProposal.tex
Updates to background and literature review section

@@
 
-Documentation can improve software sustainability.  For software
-in general, documentation, written before and during development, provides many
+Documentation can improve software sustainability.  For software in general,
+documentation, written before and during development, provides many
 benefits~\cite{Parnas2010}: easier reuse of old designs, better communication
@@
 more efficient corrections and improvements. For SCS in particular,
-documentation provides benefits.  For instance, \citet{SmithEtAl2015-SS-TR}
-shows that the quality of statistical software for psychology is generally
-improved when developed using the structured CRAN (Comprehensive R Archive
-Network) process and tools, versus an ad hoc process.
+documentation provides significant benefits.  For instance,
+\citet{SmithEtAl2015-SS-TR} shows that the quality of statistical software for
+psychology is generally improved when developed using the structured CRAN
+(Comprehensive R Archive Network) process and tools, versus an ad hoc approach.
 \citet{SmithAndKoothoor2016} highlights the value of proper documentation by
 redeveloping nuclear safety analysis software.  Twenty seven (27) worrisome
-documentation problems were found, including incompleteness, ambiguity,
-inconsistency, verifiability, modifiability, traceability and a lack of
-abstraction.  A redevelopment experiment with five existing projects
+documentation problems were found, including problems with incompleteness,
+ambiguity, inconsistency, verifiability, modifiability, traceability and a lack
+of abstraction.  A redevelopment experiment with five existing projects
 \citep{SmithJegatheesanAndKelly2016} enabled the code owners (as ascertained
@@
 
-One promising approach to gain the benefits of documentation, but significantly
-reduce the time commitment, is to use a Model Driven Engineering (MDE).  MDE is
-an engineering approach that exploits structured representations (metamodels for
-abstract syntax, and constraints to capture static semantics) to automate
-repetitive, tedious and error prone tasks.  In this research proposal, we will
-use MDE as a short-hand to include knowledge models, model transformation,
-Domain Specific Languages (DSLs) and code generation.  Several current papers
-point to DSLs and code/document generation as a transformative technology for
-documentation, design and verification of SCS \citep{JohansonAndHasselbring2018,
-  Smith2018}.  The benefit of an MDE approach is to dramatically reduce the cost
-of developing reliable, reproducible and re-certifiable software.
+One promising approach to gain the sustainability benefits of documentation, but
+significantly reduce the time commitment, is to use a Model Driven Engineering
+(MDE) process.  MDE is an engineering approach that exploits structured
+representations (metamodels for abstract syntax, and constraints to capture
+static semantics) to automate repetitive, tedious and error prone tasks.  In
+this research proposal, we will use MDE as a short-hand to include knowledge
+models, model transformation, Domain Specific Languages (DSLs) and code
+generation.  Several current papers point to DSLs and code/document generation
+as a transformative technology for documentation, design and verification of SCS
+\citep{JohansonAndHasselbring2018, Smith2018}.  MDE has the potential to
+dramatically reduce the cost of developing reliable, reproducible and
+re-certifiable software.
 
@@
 to generate theoretical models, design documents, code, test cases and build
-scripts. The generator can render to multiple languages, such as html, LaTeX,
-RTF for documentation and Python, Java, C, and Matlab, for code. One source of
+scripts. The generator can render to multiple languages, such as html, and LaTeX
+for documentation and Python, Java, C++ and Lua, for code. One source of
 knowledge, with rules for transformation, means completeness, consistency and
@@
 
-As described above, sustainability is a concern for SCS, and MDE holds promise
+As described above, sustainability is a concern for SCS; MDE holds promise
 for addressing this concern.  However, a transformative change in the
@@
 of a certification exercise.  Although this study is not specifically restricted
-to open-source software, for practical reasons, commercial software will not be
-strongly emphasized.
+to open-source software, for practical reasons (especially the lack of access to
+the code), commercial software will not be strongly emphasized.
 
@@
 short-term objectives.  For each objective, a reference is given to the
-corresponding section in this document that provides the details.
+corresponding section where details are provided.
 
 \begin{enumerate}
-\item Assess the current state of software development practice for SCS.  How is
-  the software developed?  Are there existing development methods that lead to
-  higher software sustainability? (Section~\ref{SecStateOfThePractice})
+\item Assess the current state of the software development practice for SCS.
+  How is the software developed in different sub-domains of SCS?  Are there
+  existing development methods that lead to higher software sustainability?
+  (Section~\ref{SecStateOfThePractice})
 \item Assess the impact of an MDE process on end user developed SCS with respect
-  to developer productivity and software sustainability.  
+  to developer productivity and software sustainability for long-lived lived software.  
   (Section~\ref{SecImpactMDE})
@@
   the argument will come from the traceability between artifacts generated by
-  model transformations, expert reviews, test cases,
-  etc. (Section~\ref{SecAssuranceCase})
+  model transformations, expert reviews, test cases, etc.
+  (Section~\ref{SecAssuranceCase})
 \item Assess the impact on software quality when using MDE to facilitate
@@
 maintainability.  Therefore, Section~\ref{SecDesirableQs} introduces measurable
-documentation qualities that are believed to contribute to software quality.  In
-some cases it may be necessary to use the qualities in
+documentation qualities that are believed to contribute to the overall software
+qualities.  In some cases it may be necessary to use the qualities in
 Section~\ref{SecDesirableQs} to indirectly measure qualities listed in
@@
 
-focuses on presents it, roadmap of this
-section.  Some overview of the topics themselves, but an emphasis on how they
-appear in the SCS context.
+The other subsection in this background section provide an overview of the
+literature on the current approaches to developing SCS software
+\ref{SecCurrStateOfThePractice}, recommended document artifacts for SCS
+\ref{SecDocumentation}, an overview of MDE and Code Generation in an SCS context
+\ref{SecMDE}, and a summary of existing literature on empirical methods for
+software engineering \ref{SecEmpirical}.
 
@@
 
-\item [\textbf{Performance}] considerations can make certification challenging, since QA
-  becomes more difficult for more complex code.  However,
-  as Roache~\cite[p.~355]{Roache1998} points out, using simpler algorithms and
+\item [\textbf{Performance}] considerations can make certification challenging,
+  since QA becomes more difficult for more complex code.  However, as
+  Roache~\cite[p.~355]{Roache1998} points out, using simpler algorithms and
   reducing the number of options in general purpose code, is not always a
@@
 
-\item [\textbf{Sustainability}] A combination of other qualities.
+\item [\textbf{Sustainability}] A combination of other qualities, likely
+  maintainability and productivity.  Reference authorless paper found on-line.
 
@@
 final quality listed (abstraction), are adapted from the IEEE recommended
-practise for producing good software requirements~\cite{IEEE1998}.
+practise for producing good software requirements~\cite{IEEE1998}.  Abstraction
+means only revealing relevant details, which in a requirements document means
+stating what is to be achieved, but remaining silent on how it is to be
+achieved.  Abstraction is an important software development principle for
+dealing with complexity~\cite[p.~40]{GhezziEtAl2003}.
+\citet{SmithAndKoothoor2016} present further details on the qualities of
+documentation for SCS.
 
@@
 A third party should be able to repeat a study using only the description of the
-methodology from a published article %, with no access to the original code or
-%computing environment 
-\citep{BenureauAndRougier2017}.  However, replicability is rarely achieved, as
-shown for microarray gene expression \citep{IoannidisEtAl2009} and for economics
-modelling \citep{IonescuAndJansson2013}.  Drasil addresses completeness and
-ambiguity problems, since it emphasizes capturing and documenting all of the
-required knowledge, including derivation of equations and
+methodology from a published article, with no access to the original code or
+computing environment \citep{BenureauAndRougier2017}.  However, replicability is
+rarely achieved, as shown for microarray gene expression
+\citep{IoannidisEtAl2009} and for economics modelling
+\citep{IonescuAndJansson2013}.  Drasil addresses completeness and ambiguity
+problems, since it emphasizes capturing and documenting all of the required
+knowledge, including derivation of equations and
 rationales. \citet{CrickAndHall2014} point out potential roadblocks for
@@
 develop methods and tools that are tailored to the domain'' \citep{Storer2017}.
-\emph{This tailored research will require individuals, like myself, that have a
-  multidisciplinary background.}
+This tailored research will require teams with a multidisciplinary background.
 
@@
 
-To achieve the qualities listed in Table~\ref{Tbl_Qualities}, the documentation
-in Table~\ref{TblDocuments} should have the following qualities:
-\emph{complete}, \emph{correct}, \emph{consistent}, \emph{modifiable},
-\emph{traceable}, \emph{unambiguous}, and \emph{verifiable}.  All of these
-qualities are listed in the IEEE recommended practice for software
-requirements~\cite{IEEE1998}.  The IEEE guidelines are for requirements, but
-most qualities are relevant for all documentation artifacts.  Another relevant
-quality, which is not on the IEEE list, is \emph{abstract}.  Requirements should
-state what is to be achieved, but be silent on how it is to be achieved.
-Abstraction is an important software development principle for dealing with
-complexity~\cite[p.~40]{GhezziEtAl2003}.  Smith and Koothoor present further
-details on the qualities of documentation for scientific
-software~\cite{SmithAndKoothoor2016}.
+\wss{Add figure showing the V-model of the documentation.}
 
@@
 differential equations.  Unlike previous work on SCS code generation, Drasil
-will focus on generating all software artifacts (requirements, design etc.), not
-just code.
+(\href{https://github.com/JacquesCarette/Drasil}
+{https://github.com/JacquesCarette/Drasil}) focus on generating all software
+artifacts (requirements, design etc.), not just code.
 
-\citet{SzymczakEtAl2016} removes excuses for avoiding documentation by providing
-transformative SCS development technology.  Drasil provides an infrastructure
-for knowledge capture and document/code generation.  Jacques Carette and I,
-along with our students, have developed a prototype (available on GitHub),
-implemented via Domain Specific Languages (DSLs) embedded in Haskell.  Drasil
-already generates requirements documentation and code for several case studies,
+Drasil, as presented in \citet{SzymczakEtAl2016} removes excuses for avoiding
+documentation by providing transformative SCS development technology.  Drasil
+provides an infrastructure for knowledge capture and document/code generation.
+Drasil is implemented via Domain Specific Languages (DSLs) embedded in Haskell.
+Drasil generates requirements documentation and code for several case studies,
 including simulating the temperature of a solar water heating system, glass
@@
 (requirements, design etc), code and test cases have been created manually for
-each case study.  These manual case studies (many from \citet{SmithJegatheesanAndKelly2016})
-provide the `gold standard' against which Drasil is tested.  All of these
-artifacts are available publicly.  The generation techniques for Drasil began to
-take shape during my work on generating geometric data structures and functions
-\citep{CaretteEtAl2011}.
+each case study.  These manual case studies (many from
+\citet{SmithJegatheesanAndKelly2016}) provide the `gold standard' against which
+Drasil is tested.  All of these artifacts are available publicly.  The
+generation techniques for Drasil began to take shape during work on generating
+geometric data structures and functions \citep{CaretteEtAl2011}.
 
@@
 
-empirical software engineering paper, case study paper
+The empirical methods in this study will follow the guidelines given in
+\citet{KitchenhamEtAl2002}.  Since case studies will be part of the methodology,
+the guidelines for conducting and reporting case study research in software
+engineering \citep{RunesonAndHost2009} will be followed.
 
@@
 
+Rule 4: Keep knowledge up to date and findable \citep{ShollerEtAl2019}.
+
 \subsubsection{Medical Imaging}
@@
 Drasil and then borrow from existing knowledge and create new knowledge?  This
-project could start with something like the task described in~\ref{SecPDF}.
+project could start with something like the
+\href{https://jacquescarette.github.io/Drasil/examples/Projectile/srs/Projectile_SRS.html}
+{Projectile} example currently in Drasil.
 
@@
 could be developed and compared.  Some comparison could be automated between the
-documents, such as empirical measures of their quality
-(Section~\ref{SecEmpiricalMeasureDoc}).  Other measures would require some kind
-of usability experiments.
+documents, such as empirical measures of their quality.  Other measures would
+require some kind of usability experiments.
 

### 2019-09-18 ResearchProposal.tex
Edits to research proposal

@@
 documentation, design and verification \citep{JohansonAndHasselbring2018,
-  Smith2018}.  DSLs allow scientists their preferred approach of focusing on
-science not software \citep{Kelly2007}.  A generative approach removes the
-maintenance nightmare of documentation duplicates and near duplicates
+  Smith2018}.  Scientists prefer focusing on science not software
+\citep{Kelly2007}.  DSLs allow them to do this.  A generative approach removes
+the maintenance nightmare of documentation duplicates and near duplicates
 \citep{LucivEtAl2018}, since knowledge is only captured once and automatically
@@
 \citet{SmithJegatheesanAndKelly2016}) provide the `gold standard' against which
-Drasil is tested.  All of these artifacts are available publicly.  The
-generation techniques for Drasil began to take shape during work on generating
-geometric data structures and functions \citep{CaretteEtAl2011}.
+Drasil is tested.  All of these artifacts are available publicly at
+\href{https://jacquescarette.github.io/Drasil/}
+{https://jacquescarette.github.io/Drasil/}.  The generation techniques for
+Drasil began to take shape during work on generating geometric data structures
+and functions \citep{CaretteEtAl2011}.
 
@@
 
-roadmap - fit the different pieces together
+This section is decomposed into 4 subsections -- one for each of the different
+approaches for assessing the impact of MDE on the sustainability of SCS.  The
+goal of the first section is to establish a ``baseline'' on the current state of
+the practice for SCS.  This is done by identifying domains within SCS where
+significant software development work has been done and then systematically
+assessing the quality of the existing work.  The next section looks specifically
+at the impact of MDE on SCS through a series of case studies.  An ``ideal'' MDE
+process is proposed and then followed with different stakeholders and
+qualitative data is collected.  The last two sub-sections discuss assessing the
+impact on SCS of assurance cases and computational variability testing,
+respectively.
 
@@
 Key to the success of this exercise will be to involve and engage a partner for
-each state of practice project.  In the previous effort we didn't involve domain
+each state of practice project.  In the previous effort we did not involve domain
 experts with the rationale of excluding potential bias, but this advantage is
@@
 Determine criteria for what makes a domain a good fit.
+
+Write up experimental procedure as a publication - on arxiv, but hopefully
+somewhere else too.
+
+Start from previous work and modify.  Brainstorm how to do measures.
+
+Schedule meeting.
 

### 2019-09-23 ResearchProposal.tex
Updates to research proposal

@@
 
-\section{Introduction}
+\section{Introduction} \label{SecIntroduction}
 
@@
 This section is decomposed into 4 subsections -- one for each of the different
-approaches for assessing the impact of MDE on the sustainability of SCS.  The
-goal of the first section is to establish a ``baseline'' on the current state of
-the practice for SCS.  This is done by identifying domains within SCS where
-significant software development work has been done and then systematically
-assessing the quality of the existing work.  The next section looks specifically
-at the impact of MDE on SCS through a series of case studies.  An ``ideal'' MDE
-process is proposed and then followed with different stakeholders and
-qualitative data is collected.  The last two sub-sections discuss assessing the
-impact on SCS of assurance cases and computational variability testing,
-respectively.
+approaches for assessing the impact of MDE on the sustainability of SCS, as
+outlined in the Introduction (Section~\ref{SecIntroduction}).  The goal of the
+first subsection is to establish a ``baseline'' on the current state of the
+practice for SCS.  This is done by identifying domains within SCS and then
+systematically assessing the quality of the work in those domains.  The second
+subsection looks specifically at the impact of MDE on SCS through a series of
+case studies.  An ``ideal'' MDE process is proposed and then followed with
+different stakeholders and qualitative data is collected.  The last two
+sub-sections discuss assessing the impact on SCS of assurance cases and
+computational variability testing, respectively.
 
@@
 
-A survey of existing SCS software will identify patterns, find opportunities for
-improvement and identify candidate program families for future implementation.
-The survey will require developing terminology and a classification system to
-answer such questions as: What are the current best practises?  What software
-packages are reused and why these packages?  What do practitioners look for in
-the tools that support their work?
-
-To achieve objective S2, a measurement template, possibly using pairwise
-comparisons of quality metrics between software packages, will be developed for
-assessing the quality of the documentation, code, test cases and development
-processes for existing SCS program families.  The quality measurement template
-will be used to assess the quality of approximately 30 different SCS families.
-About half of the families will be model oriented and the other half tool
-oriented.  Additional details on best practices will be collected by students
-interviewing code owners.  Based on preliminary discussions, Dr.\ Jeffrey Carver
-from the University of Alabama and some of his colleagues may also participate
-in collecting interview data. Once all the surveys and interviews are
-completed, a meta-analysis will draw conclusions for each domain and between
-domains.  Knowing the state of practice for family development will highlight
-which domains are using best practices that should be emulated in LSS.
-
-The plan here is to return to the task of measuring/assessing the state of
-software development practice in several scientific computing domains.  We will
-update the work that was done previously for domains such as Geographic
-Information Systems~\cite{SmithEtAl2018_arXivGIS}, Mesh
-Generators~\cite{SmithEtAl2016}, Seismology software~\cite{SmithEtAl2018}, and
-Statistical software for psychology~\cite{SmithEtAl2018_StatSoft}.  We could
-return to these domains and/or introduce new domains.  Potential domains/collaborators
-are listed in the subsections.
+To understand the current quality of SCS and to assist in developing methods for
+measuring qualities, we will identify several domains within SCS and endeavour
+to summarize the current state of the practice within those domains.  This
+``state of the practice'' exercise will build off of prior work on
+measuring/assessing the state of software development practice in several SCS
+domains.  We will update the work that was done previously for domains such as
+Geographic Information Systems \citep{SmithEtAl2018_arXivGIS}, Mesh Generators
+\citep{SmithEtAl2016}, Seismology software \citep{SmithEtAl2018}, and
+Statistical software for psychology \citep{SmithEtAl2018_StatSoft}.  We may
+return to these domains and/or introduce new domains.  Potential
+domains/collaborators are listed in the subsections below.
 
 MEng students will be asked to assess the state of practice in each domain.
-Last time each student measured two domains, so we would start off with this
-model again.  If the scope becomes large enough, we might switch to one domain
-per student.
+During the last state of the practice exercise each student measured two
+domains, but with the higher demands from the new measurement protocol, we will
+likely switch to one domain per student.
 
 In the previous project, we measured 30 software projects for each domain.  With
-the increased scrutiny required in this re-boot, we won't likely be able to
+the increased scrutiny required in this re-boot, we will not likely be able to
 measure the details on this high a number of software projects.  We may still
@@
 
-In the previous state of practice assessing exercise, a series of questions and
-simple metrics was devised to measure the quality of the documentation and
-adherence to best software development practices.  The new project will
-critically assess the previous set of questions and revise as necessary.  In
-addition, the following data will be collected/developed for each domain:
+In the previous state of the practice assessing exercise
+\citep{SmithEtAl2018_arXivGIS}, a series of questions and simple metrics was
+devised to measure the quality of the documentation and adherence to best
+software development practices.  The previous set of questions is provided
+\href{https://github.com/adamlazz/DomainX/blob/master/TemplateToGradeSCSoft.pdf}
+{here}. The new project will critically assess the previous set of questions and
+revise as necessary.  In addition, the following data will also be considered
+for collection for each domain:
 
@@
   assessing usability still need to be researched, but measures that involve the
-  time to complete a task and the users overall impression would be considered.
+  time to complete a task and the users overall impression will be considered.
 \item Collect empirical software engineering related data, such as the number of
@@
 \item Collect at least one empirical measure of the quality of the documentation
-  - the number of lines of documentation.  [Has anyone previously looked at this
-  metric?]
+  - the number of lines of documentation.  We should investigate to see if this
+  is a metric that has previously been collected.  So many of the existing
+  metrics are completely focused on the code, or possibly on issue creation and
+  management; it seems that little work has been done to assess the
+  documentation quality.
+\item One of the guidelines for successfully on-boarding new developers is to
+  ``Keep knowledge up to date and findable'' \citep{ShollerEtAl2019}.  The
+  measurements selected should be able to assist with judging whether a project
+  succeeds on this front.
 \end{itemize}
@@
 specifics are yet to be determined, but in the past the Analytic Hierarchy
-Process proved helpful in this.
+Process (AHP) \citep{Saaty1980} proved helpful in this.  In the past the AHP
+results were calculated by a grade out of 10 that each student evaluator gave to
+each quality for each measured product.  A better approach may be to have a
+group of people familiar with the software do a pairwise comparison between each
+product to come up with a number.  The number would be based on the data
+collected, but not in a numerical way.  The data collected would sever the same
+role as a decision support system.  It would provided insight, but human beings
+would do the final pairwise comparisons.
 
@@
 be involved in this project.  In addition to the project record left on GitHub,
-the final data will be exported to Mendeley.
+the final data will be exported to Mendeley (as was done for the previous
+quality measurement exercise, for instance the grades for GIS software are
+summarized at \href{https://data.mendeley.com/datasets/6kprpvv7r7/1}
+{https://data.mendeley.com/datasets/6kprpvv7r7/1}.)
 
-Determine criteria for what makes a domain a good fit.
+As a next step for this work, the procedure for measuring software qualities in
+SCS will be brainstormed, discussed and written up.  The procedure will be
+applied to each domain with SCS that is assessed.  The write-up on the procedure
+will be published.  Ideally at a relevant conference, but if that is not a
+feasible option, it will be hosted on the arxiv.
 
-Write up experimental procedure as a publication - on arxiv, but hopefully
-somewhere else too.
+Potential domains for future measurement are listed below.  Determination of
+these domains are good candidates will require evaluation of the following
+criteria:
 
-Start from previous work and modify.  Brainstorm how to do measures.
+\begin{itemize}
+\item Availability of a domain expert that is willing to assist, and who has at
+  least a few graduate students that can assist with the measurement tasks.
+\item A domain where end-user developed software is common
+\item A domain where open-source software is available.  Commercial products are
+  not as effective for measurement, since we cannot generally gain access to the
+  source code, developer documentation, or issue tracking system.
+\item A domain where scientific computing is used.  The selected software should
+  fit in the larger domain of SCS.
+\end{itemize}
 
-Schedule meeting.
-
-Potential collaborators/domains are listed in the following subsections.
-
-Rule 4: Keep knowledge up to date and findable \citep{ShollerEtAl2019}.
+The data for each of the different subdomains within SCS will be standardized to
+facilitate a meta-analysis that combines all of the data.
 
@@
 
-Medical imaging and analysis software (Dr.\ Michael Noseworthy)?
+Medical imaging and analysis software.  This work could build off of the
+previous project on this topic completed by Vasudha Kapil.  (Vasudha's report is
+located in the \texttt{mmsc} svn repo) .  % (Dr.\ Michael Noseworthy)?
 
@@
 
-Climate modelling software (Dr.\ Zoe Li or Dr.\ Xander Wang)?
+Climate modelling software.  This could be software used for multi-model
+ensemble aggregating \citep{SamoulyEtAl2018} or possibly with software that uses
+physical models to predict climate change.
+% (Dr.\ Zoe Li or Dr.\ Xander Wang)?
 
@@
 
-Extraction of 3D geometries from medical images (Dr.\ Zahra Motamed)?  Lattice
-Boltzmann Solver?
+Potential sub-domains with computational medicine could include: extraction of
+3D geometries from medical images, or the use of Lattice Boltzmann
+Solvers.  % (Dr.\ Zahra Motamed)?
 
@@
 
-Finite element solvers and/or mesh generators (Dr.\ Dieter Stolle)?  Potential
+Finite element solvers and/or mesh generators.  Potential
 software includes FeNiCS, MOOSE, FreeFEM++, Dolphin, etc.  Install the packages
-and use them to do something non-trivial.
+and use them to do something non-trivial.  %(Dr.\ Dieter Stolle)?
 
@@
 
-Statistics software for psychology (Dr.\ Karin Humphreys)?
+Statistics software for psychology.  A project in this domain would allow us to
+update \citet{SmithEtAl2018_StatSoft}. %(Dr.\ Karin Humphreys?)
 
@@
 
-Li Xi (\url{https://www.eng.mcmaster.ca/chemeng/people/faculty/li-xi})?
+Software is used for many purposes in chemical engineering.  We would want to
+find an application where non-commercial software is available.
+% Li Xi? (\url{https://www.eng.mcmaster.ca/chemeng/people/faculty/li-xi})?
 
-\subsubsection{Geographic Information Systems}
+\subsubsection{Geographic Information Systems (GIS)}
 
-Geographic Information Systems (Jason Brodeur and/or Patrick DeLuca)?
+We studied GIS software in the past \citep{SmithEtAl2018_arXivGIS}, so this
+would be an interesting area to update.  The key for the updated effort will be
+to identify a domain expert available for collaboration. %(Jason Brodeur and/or Patrick DeLuca)?
 
@@
 
-Solidification and casting software (Dr.\ Andr\'e Phillion)?
+Solidification and casting software and other software used for physical
+metallurgy.  An example of software written for analyzing the solidification of
+a molten alloy is given in \citet{SmithEtAl2019_arXiv}.  % (Dr.\ Andr\'e Phillion)?
 
@@
 
-Quantum chemistry software (Dr.\ Paul Ayers)?
+Quantum chemistry software.  One example is software for searching for
+conformers, like
+\href{https://github.com/PeaWagon/Kaplan}{Kaplan}.  %(Dr.\ Paul Ayers)?
 
-\subsection{Impact of MDE on SCS} \label{SecImpactMDE}
+\subsection{Impact of MDE on the Sustainability of SCS} \label{SecImpactMDE}
 
-roadmap
+Once we have details on the current state of the practice in SCS, we can begin
+to apply MDE to SCS projects and assess the ramifications of this change.  This
+section starts by laying out the details on how MDE can be applied to SCS
+(Section~\ref{SecMDEProcess}).  We plan to start with an ideal process and then
+approximate it through different experiments to measure and assess the impact of
+MDE.  The second subsection (Section~\ref{SecQOfMDEArtifacts}) outlines how the
+quality of MDE artifacts can be assessed.  These measurements will be used in
+the various experiments to assess the impact of MDE.  The third and fourth
+subsections describe two experiments for collecting data: i) conducting case
+studies with real SCS projects (Section~\ref{SecCaseStudy}); and, ii) performing
+an experiment where multiple developers create a small SCS project in one day
+(Section~\ref{SecSmallPhysExample}).
 
-Research Question: What tools and methods from MDE are perceived by end user developers to
-  improve their productivity and the sustainability of their software.
+\subsubsection{Fitting MDE into the Scientific Software Development
+  Process} \label{SecMDEProcess}
 
-\subsubsection{Fitting Drasil into the Scientific Software Development Process} \label{Sec}
+Many researchers have stated that a document driven process is not used by, nor
+suitable for, scientific software.  These researchers argue that scientific
+developers naturally use an agile philosophy~\cite{AckroydEtAl2008,
+  CarverEtAl2007, EasterbrookAndJohns2009, Segal2005}, or an amethododical
+process~\cite{Kelly2013}, or a knowledge acquisition driven
+process~\cite{Kelly2015}.  However, as discussed in the introduction to this
+proposal (Section~\ref{SecIntroduction}, documentation provides many benefits.
+An MDE approach holds promise to provide these benefits, while removing the
+drudgery from the process of documentation and facilitating inevitable changes
+in the documentation and the software.
 
-Scientists generally use an approximation of agile methods for their software
-development.  In addition, documentation is generally not emphasized.  How
-should Drasil be fit into the scientific software development process?  The
-Drasil process should encourage an early intensity of thought and effort, but at
-the same time, it shouldn't require waterfall development.  Drasil supports
-frequent chance and incremental development, but where should developers start
+The general idea that MDE will be helpful is appealing, but to study the impact
+of MDE, we need to know how it will actually be used.  How exactly would MDE be
+fit into the scientific software development process?  The new process should
+encourage an early intensity of thought and effort, but at the same time, it
+should not impose waterfall development, since waterfall is an unrealistic
+expectation for a problem that is not yet well understood.  MDE supports
+frequent change and incremental development, but where should developers start
 and how should the work proceed?  How does one create an almost empty project in
-Drasil and then borrow from existing knowledge and create new knowledge?  This
-project could start with something like the
+MDE and then borrow from existing knowledge and create new knowledge?  To create
+a specific MDE process for SCS, the Drasil tools will be utilized.  A good
+starting project could be
 \href{https://jacquescarette.github.io/Drasil/examples/Projectile/srs/Projectile_SRS.html}
-{Projectile} example currently in Drasil.
+{Projectile}, which models simple projectile motion.
 
-Drasil process should move importance of requirements and domain analysis
-earlier in the process.  The quality of the resulting software depends on the
-quality of the requirements.  As is known from historical data, the greatest
-return on investment is at the requirements state (Boehm papers).
-Unfortunately, developers don't like requirements.  Quite possibly they don't
-like the intensity of thought that is necessary to get the requirements right.
-Inspection techniques, such as task based inspection (Kelly et al) hold promise
+Compared to existing SCS development techniques, the new Drasil process should
+emphasize requirements and domain analysis earlier in the process.  The quality
+of the resulting software depends on the quality of the requirements.  As
+Boehm's data shows~\cite{Boehm1981}, early identification of errors implies a
+significant return on investment, since the cost of fixing errors increases
+dramatically for later development stages.  An SRS is invaluable for
+verification, since the quality of software cannot be assessed without a
+standard against which to judge it. Better design decisions are facilitated by
+the information captured in an SRS.  Moreover, the SRS aids the software
+lifecycle by facilitating incremental development.  This last advantage is
+important for SCS software, since changes frequently occur as developers explore
+the problem domain.  Unfortunately, developers do not generally write
+requirements.  Quite possibly they do not like the intensity of thought that is
+necessary to get the requirements right.  Inspection techniques, such as task
+based inspection \citep{KellyAndShepard2000, KellyAndShepard2004} hold promise
 for getting the scientist to think deeply at an earlier stage in the development
@@
 
-\subsubsection{etc.}
+Since Drasil is not the only tool that can be used for MDE, we will also
+consider what an MDE approach would look like using other tools.  As a starting
+point, we will look at \href{https://www.eclipse.org/epsilon/}{Epsilon}.
+Epsilon is a family of languages and tools for code generation, model-to-model
+transformation, model validation, comparison, migration and refactoring that
+work out of the box with EMF, UML, Simulink, XML and other types of models.
 
-Experimentally Determine Appropriate Documentation Template: The best approach
-for documenting requirements and design is an open question.  Drasil could be
-used to attempt to address this question.  Different recipes for documentation
-could be developed and compared.  Some comparison could be automated between the
-documents, such as empirical measures of their quality.  Other measures would
-require some kind of usability experiments.
+\subsubsection{Measuring the Qualities of MDE Developed
+  Artifacts} \label{SecQOfMDEArtifacts}
 
-Empirical Measurement of Documentation: Considerable effort has been invested
-into developing metrics, like cyclomatic complexity, to attempt to quantify code
-quality, but no measures (as far as I know) exist for quantifying the quality of
-documentation.  (Khedri does have some work with Bhahati (sp?) Sanga that
-quantifies the connectivity of the documentation via the traceability graph.)
-Likely the reason that effort has not been invested to quantifying the quality
-of documentation is that is not represented formally.  With Drasil much of the
-documentation is formal.  This should facilitate automatic measures of
-documentation quality, or at least metrics that might be correlated with
-quality.
-
-Measuring the Impact of Drasil on Developing Scientific Software: This project
-is challenging to design, but would be HUGE if we were successful.  We assume
-that Drasil will improve the quality of scientific software and the efficiency
-of its production, but data to back this up would go a long way to convincing
-the scientific community.  Experiments where a given project is developed in
-parallel by two independent teams come to mind, but this would be expensive, and
-difficult to control.  Some additional thought, and resources, are necessary
-here.  If we could do a project where the scientist, or a student under their
-supervision, is a true partner in the development, we might get more convincing
-anecdotal evidence.
+Some of the quality measures developed in Section~\ref{SecStateOfThePractice}
+for the ``state of the practice'' exercise will also be used to measure the
+quality of the generated documentation and code.
+Section~\ref{SecStateOfThePractice} mentions using code quality metrics, like
+cyclomatic complexity, to attempt to quantify code quality.  Given the
+importance of documentation, it seems reasonable to also assess its quality.
+Measuring documentation quality is not something that is often promoted.
+(\citet{Sanga2003} does discuss quantifying the connectivity within an SRS
+document by analyzing the traceability graph.)  Part of the reason for the lack
+of effort on quantifying the quality of documentation is that is not represented
+formally.  With Drasil much of the documentation is formal.  This should
+facilitate automatic measures of documentation quality, or at least metrics that
+might be correlated with quality.
 
@@
 
-\subsection{Impact of Assurance Cases for Building Confidence} \label{SecAssuranceCase}
+\subsubsection{Case Study} \label{SecCaseStudy}
+
+For assessing the impact of MDE, experiments where a given project is developed
+in parallel by two independent teams are infeasible due to expense and
+challenges for having a truly controlled experiment.  Instead, we will focus on
+the more practical, but still very useful, case study approach.  Guidelines for
+conducting and reporting case study research in software engineering are given
+in \citet{RunesonAndHost2009}.  As explained there, a case study should be built
+around a research question.  The research question should be specific and
+measurable.  For the current case study exercise, we propose the question:
+``What tools and methods from MDE are perceived by end user developers to
+improve their productivity and the sustainability of their software in the long
+term?''
+
+We will identify multiple case studies, with potential projects coming from the
+domains listed for the ``state of the practice'' exercise
+(Section~\ref{SecStateOfThePractice}), like computational medicine, medical
+imaging or quantum chemistry software.  We will first identify the partners and
+look for feasible projects that can be developed using an MDE approach.  We will
+use the ideal process identified in Section~\ref{SecMDEProcess}.  Compare to Some additional thought, and resources, are necessary
+here.  If we could do a project where the scientist, or a student under their
+supervision, is a true partner in the development, we might get more convincing
+anecdotal evidence.  Pre and post interviews.  Measurement of quality following the
+techniques developed in the state of the practice exercise.
+
+List of projects above could also be used for case studies.
+
+Observe - how answer questions when problems come up.  Should have so
+pre and post interviews.
+
+\subsubsection{Development of Small Physics
+  Examples} \label{SecSmallPhysExample}
+
+describe a different experiment.  conference multiple develop physics code.
+Like projectile.  Survey before and after.  Quantify success of efforts and time required.
+
+\subsection{Impact of Assurance Cases for Building
+  Confidence} \label{SecAssuranceCase}
+
+Sketching out an Assurance Case in advance of developing the software could be
+just what is needed to manufacture the up-front intensity that is needed for
+successful projects.
+
+Experimentally Determine Appropriate Documentation Template: The best approach
+for documenting requirements and design is an open question.  Drasil could be
+used to attempt to address this question.  Different recipes for documentation
+could be developed and compared.  Some comparison could be automated between the
+documents, such as empirical measures of their quality.  Other measures would
+require some kind of usability experiments.
 
@@
 
+\section{Research Schedule} \label{SecSchedule}
+
 \newpage
-
-\section{Research Schedule} \label{SecSchedule}
 

### 2019-09-25 OverallResearchProposal/ResearchProposal.tex
Reorg folder structure to put research proposal in its own folder

--- 2019-09-23 ResearchProposal.tex
@@
 
-The scope of SCS software for this proposal includes long-lived software, as
-implied by the sustainability objective.  To gain the full benefits of
-generating documentation, the scope also includes software that interests
-multiple stakeholders, with different interests and backgrounds.  Finally, the
-scope of this proposal emphasizes safety related software, such as software for
-nuclear safety analysis, medical imaging and computational medicine.  Safety
-related software is prioritized because documentation is often expected as part
-of a certification exercise.  Although this study is not specifically restricted
-to open-source software, for practical reasons (especially the lack of access to
-the code), commercial software will not be strongly emphasized.
+The scope of SCS software for this proposal includes software intended for a
+long life (10+ years), as implied by the sustainability objective.  The effort
+involved in MDE does not provide enough of a payoff for short-term projects.  To
+gain the full benefits of generating documentation, the scope also includes
+software that interests multiple stakeholders, with different interests and
+backgrounds.  Finally, the scope of this proposal emphasizes safety related
+software, such as software for nuclear safety analysis, medical imaging and
+computational medicine.  Safety related software is prioritized because
+documentation is often expected as part of a certification exercise.  Although
+this study is not specifically restricted to open-source software, for practical
+reasons (especially the lack of access to the code), commercial software will
+not be strongly emphasized.
+
+Implicit in the scope is that we are interested in software for domains that are
+well-understood.  This is necessary for knowledge capture and the associated
+construction of models.  We can refine the notion of well-understood by using
+the fact that MDE is simply a technology based realization of the older concept
+of program families.  Success with an MDE approach requires
+satisfaction of the same three hypotheses as success for program family
+development \citep{Weiss1997}:
+
+\begin{description}
+\item [Redevelopement Hypothesis] -- most software development involved in
+  producing an individual family members should be redevelopement because the
+  family members have so much in common.
+
+\item[Oracle Hypothesis] -- the changes that are likely to occur during the
+  software's lifetime are predictable.
+
+\item [Organizational Hypothesis] -- designers can organize the software and the
+  development effort so that predicted changes can be made independently.
+\end{description}
+
+SCS can be shown to satisfy these three hypotheses \citep{SmithAndChen2004,
+  SmithMcCutchanAndCao2007}; therefore, SCS is suitable for a program family, or
+MDE software development approach.  Although our focus is on SCS, an MDE
+approach could be applied to any domain that satisfies the above three
+hypotheses.
 
@@
 
-\item [\textbf{Productivity}] [Fill in]
+\item [\textbf{Productivity}] \wss{This needs to be filled in.  Productivity is
+    an important quality for this proposal, since part of what MDE promises is
+    increased productivity.}
 
-\item [\textbf{Sustainability}] A combination of other qualities, likely
-  maintainability and productivity.  Reference authorless paper found on-line.
+\item [\textbf{Sustainability}] \wss{This needs to be filled in.  A search needs
+    to be done to find a good definition.  A starting point might be the recent
+    paper that defines sustainability as a combination of other qualities.}
 
@@
 MEng students will be asked to assess the state of practice in each domain.
-During the last state of the practice exercise each student measured two
+During the previous state of the practice exercise each student measured two
 domains, but with the higher demands from the new measurement protocol, we will
@@
 create a shorter sub-list for detailed measurement.  The details of how to
-proceed here would be determined and documented in the measurement protocol.
+proceed here would be determined and documented in the measurement procedures.
+We still would like to keep the number of measured software projects above 10.
 
@@
   files, number of lines of code, cyclomatic complexity, number of open issues,
-  etc.  As a starting point for tool support, HubListener
-  (\url{https://github.com/pjmc-oliveira/HubListener}) could be used.
+  number of closed issues, etc.  As a starting point for tool support,
+  HubListener (\url{https://github.com/pjmc-oliveira/HubListener}) could be
+  used.
 \item Collect at least one empirical measure of the quality of the documentation
@@
   ``Keep knowledge up to date and findable'' \citep{ShollerEtAl2019}.  The
-  measurements selected should be able to assist with judging whether a project
+  selected measurements should be able to assist with judging whether a project
   succeeds on this front.
 \end{itemize}
+
+To focus the revised measurement procedure, we will go through each of the
+quality criteria listed in Section~\ref{SecSoftwareQuality}.  For each criteria
+we will look at how it was previously measured by us, but also at the literature
+and current best practices for its measurement.
 
@@
 Process (AHP) \citep{Saaty1980} proved helpful in this.  In the past the AHP
-results were calculated by a grade out of 10 that each student evaluator gave to
+results were calculated by a grade out of 10 from each student evaluator for
 each quality for each measured product.  A better approach may be to have a
@@
 product to come up with a number.  The number would be based on the data
-collected, but not in a numerical way.  The data collected would sever the same
+collected, but not in a numerical way.  The data collected would serve the same
 role as a decision support system.  It would provided insight, but human beings
-would do the final pairwise comparisons.
+would do the final pairwise comparisons, and thus the final ranking.
 
@@
 SCS will be brainstormed, discussed and written up.  The procedure will be
-applied to each domain with SCS that is assessed.  The write-up on the procedure
+applied to each domain within SCS that is assessed.  The write-up on the procedure
 will be published.  Ideally at a relevant conference, but if that is not a
-feasible option, it will be hosted on the arxiv.
+feasible option, it will be posted on the arxiv.
 
@@
   least a few graduate students that can assist with the measurement tasks.
-\item A domain where end-user developed software is common
+\item A domain where end-user developed software is common.
 \item A domain where open-source software is available.  Commercial products are
@@
   source code, developer documentation, or issue tracking system.
+\item A domain where at least 10 software projects are available for
+  measurement.
 \item A domain where scientific computing is used.  The selected software should
@@
 of MDE, we need to know how it will actually be used.  How exactly would MDE be
-fit into the scientific software development process?  The new process should
-encourage an early intensity of thought and effort, but at the same time, it
-should not impose waterfall development, since waterfall is an unrealistic
+fit into the scientific software development process?  The new ideal process
+should encourage an early intensity of thought and effort, but at the same time,
+it should not impose waterfall development, since waterfall is an unrealistic
 expectation for a problem that is not yet well understood.  MDE supports
@@
 process.
+
+The ideal process using the Drasil tool will focus on the ideal version of
+Drasil, since Drasil is still a work on progress.  The design of the ideal
+process for the ideal Drasil, will help to guide future development of Drasil.
 
@@
 look for feasible projects that can be developed using an MDE approach.  We will
-use the ideal process identified in Section~\ref{SecMDEProcess}.  Compare to Some additional thought, and resources, are necessary
-here.  If we could do a project where the scientist, or a student under their
-supervision, is a true partner in the development, we might get more convincing
-anecdotal evidence.  Pre and post interviews.  Measurement of quality following the
-techniques developed in the state of the practice exercise.
+use the ideal process identified in Section~\ref{SecMDEProcess}.  The respective
+advantages and disadvantages of Drasil versus Epsilon will be compared.
 
-List of projects above could also be used for case studies.
+To assess the impact of MDE, we will use techniques like pre and post
+interviews, the measurements of quality used for the ``state of the practice''
+exercise, ethnographic study,  etc.
 
-Observe - how answer questions when problems come up.  Should have so
-pre and post interviews.
+\wss{More information is needed in this section.}
 
-\subsubsection{Development of Small Physics
-  Examples} \label{SecSmallPhysExample}
+\subsubsection {Development of Small Physics Examples} \label{SecSmallPhysExample}
 
-describe a different experiment.  conference multiple develop physics code.
-Like projectile.  Survey before and after.  Quantify success of efforts and time required.
+To learn more about MDE for SCS, and to assess the current and future state of
+Drasil, a workshop could be conducted where the participants are all expected to
+build a small physics application, together with full documentation and test
+cases, in a one day, or possibly half day workshop.  The participants would be
+surveyed before and after the development exercise, and observations would be
+made throughout the exercise.
+
+\wss{More detail is needed.}
 
@@
 
+\wss{Schedules are incomplete}
+
+\subsection{State of Practice Exercise}
+
+\begin{itemize}
+\item Fall 2019
+  \begin{itemize}
+  \item Review previous measurement template
+  \item Update measurement process
+  \item Meet with stakeholders
+  \item Identify list of software to be measured
+  \end{itemize}
+\item Winter 2020
+  \begin{itemize}
+  \item Complete measurements
+  \item Use AHP to rank software products for each quality, and for overall
+    quality
+  \end{itemize}
+\item Summer 2020
+  \begin{itemize}
+  \item Write up results, MEng reports, papers
+  \end{itemize}
+\end{itemize}
+
+\subsection{Impact of MDE on the Sustainability Timeline}
+
+A high-level timeline of the research plan for assessing the impact of MDE on
+sustainability (see Section~\ref{SecImpactMDE} ) follows:
+
+\begin{itemize}
+\item Fall 2019
+  \begin{itemize}
+  \item Write research protocol
+  \end{itemize}
+\item Winter 2020
+  \begin{itemize}
+  \item 
+  \end{itemize}
+\item Summer 2020
+  \begin{itemize}
+  \item 
+  \end{itemize}
+\item Fall 2020
+\item Winter 2021
+\item Summer 2021
+\item Fall 2021
+\end{itemize}
+
 \newpage

### 2019-09-25 StateOfPractice/QualityDefinitionsOfQuality/QualityDefinitionsOfQuality.tex
Start of state of practice quality definitions

--- 2019-09-18 ResearchProposal.tex
@@
 
-\title{Assessing the Impact of MDE and Code Generation on the Sustainability of
-  Scientific Computing Software: A Research Proposal} 
-\author{Spencer Smith}
+\title{Definining Quality} 
+\author{Spencer Smith, Jacques Carette, Olu Owojaiye, Peter Michalski and Ao Dong}
 \date{\today}
@@
 
-\section{Introduction}
+\section{Introduction} \label{SecIntroduction}
 
-Without dramatic intervention, our collective confidence in Scientific Computing
-Software (SCS) is due for a catastrophic collapse.  We are increasingly trusting
-ever more complex and ambitious computations, but our software foundations are
-built on sand. Areas of concern include nuclear safety analysis and
-computational medicine. Successfully building such software requires
-communication between software developers and experts from multiple
-domains. Collaboration is difficult at the best of times, and is made worse
-because developers favour handcrafted solutions over adapting software
-engineering processes, methods and tools \citep{FaulkEtAl2009}. Handcrafted
-solutions do not account for the inevitable changes in requirements, design and
-implementation. The twin challenges of changing requirements and inadequate
-documentation conspire to make computational results notoriously difficult to
-reproduce, especially in the important endeavour of one researcher independently
-replicating the results of another \citep{Smith2018}.  Inadequate traceability
-within documentation means re­certification of previously certified engineering
-software is as expensive and time consuming as the original certification
-exercise, because of a lack of explicit, effort reducing, traceability
-information.  The existing challenges for modification, maintenance and
-extension point to problems with the sustainability of SCS.
+Purpose and scope of the document.
 
-Documentation can improve software sustainability.  For software in general,
-documentation, written before and during development, provides many
-benefits~\cite{Parnas2010}: easier reuse of old designs, better communication
-about requirements, more useful design reviews, easier integration of separately
-written modules, more effective code inspection, more effective testing, and
-more efficient corrections and improvements. For SCS in particular,
-documentation provides significant benefits.  For instance,
-\citet{SmithEtAl2015-SS-TR} shows that the quality of statistical software for
-psychology is generally improved when developed using the structured CRAN
-(Comprehensive R Archive Network) process and tools, versus an ad hoc approach.
-\citet{SmithAndKoothoor2016} highlights the value of proper documentation by
-redeveloping nuclear safety analysis software.  Twenty seven (27) worrisome
-documentation problems were found, including problems with incompleteness,
-ambiguity, inconsistency, verifiability, modifiability, traceability and a lack
-of abstraction.  A redevelopment experiment with five existing projects
-\citep{SmithJegatheesanAndKelly2016} enabled the code owners (as ascertained
-through interviews) to clearly see the value of documentation.  However,
-mirroring other studies \citep{CarverEtAl2007}, the code owners felt
-documentation takes too much time.  Due to the time commitment and resource
-limitation, documentation is rarely emphasized in SCS.
-
-One promising approach to gain the sustainability benefits of documentation, but
-significantly reduce the time commitment, is to use a Model Driven Engineering
-(MDE) process.  MDE is an engineering approach that exploits structured
-representations (metamodels for abstract syntax, and constraints to capture
-static semantics) to automate repetitive, tedious and error prone tasks.  In
-this research proposal, we will use MDE as a short-hand to include knowledge
-models, model transformation, Domain Specific Languages (DSLs) and code
-generation.  Several current papers point to DSLs and code/document generation
-as a transformative technology for documentation, design and verification of SCS
-\citep{JohansonAndHasselbring2018, Smith2018}.  MDE has the potential to
-dramatically reduce the cost of developing reliable, reproducible and
-re-certifiable software.
-
-One example of applying MDE to SCS is Drasil, which is introduced in
-\citet{SzymczakEtAl2016} and available at
-\href{https://github.com/JacquesCarette/Drasil}
-{https://github.com/JacquesCarette/Drasil}.  Drasil is implemented as an open
-source set of Domain Specific Languages (DSLs).  The Drasil approach involves:
-i) creating infrastructure for a knowledge base of scientific and computing
-models; and, ii) writing explicit ``recipes'' that weave together this knowledge
-to generate theoretical models, design documents, code, test cases and build
-scripts. The generator can render to multiple languages, such as html, and LaTeX
-for documentation and Python, Java, C++ and Lua, for code. One source of
-knowledge, with rules for transformation, means completeness, consistency and
-traceability can be achieved by construction. Moreover, these qualities can be
-maintained as requirements are modified, design decisions are changed,
-documentation standards are varied, and software is re­certified.  Although
-creating the knowledge base is time consuming, the knowledge can be built up
-incrementally.  New projects will reuse existing knowledge and expand as
-necessary.
-
-As described above, sustainability is a concern for SCS; MDE holds promise
-for addressing this concern.  However, a transformative change in the
-development of SCS requires more than promise.  Empirical evidence is necessary
-to demonstrate that an MDE process can improve the sustainability of SCS.  This
-leads to the long-term objective for this research proposal: \emph{assess and
-  measure the impact of MDE on the sustainability of SCS.}
-
-The scope of SCS software for this proposal includes long-lived software, as
-implied by the sustainability objective.  To gain the full benefits of
-generating documentation, the scope also includes software that interests
-multiple stakeholders, with different interests and backgrounds.  Finally, the
-scope of this proposal emphasizes safety related software, such as software for
-nuclear safety analysis, medical imaging and computational medicine.  Safety
-related software is prioritized because documentation is often expected as part
-of a certification exercise.  Although this study is not specifically restricted
-to open-source software, for practical reasons (especially the lack of access to
-the code), commercial software will not be strongly emphasized.
-
-Given the large number of potential MDE processes, techniques and technologies,
-together with the wide variety in purpose, scope and context for SCS, the
-long-term objective of measuring the impact of MDE is decomposed into four
-short-term objectives.  For each objective, a reference is given to the
-corresponding section where details are provided.
-
-\begin{enumerate}
-\item Assess the current state of the software development practice for SCS.
-  How is the software developed in different sub-domains of SCS?  Are there
-  existing development methods that lead to higher software sustainability?
-  (Section~\ref{SecStateOfThePractice})
-\item Assess the impact of an MDE process on end user developed SCS with respect
-  to developer productivity and software sustainability for long-lived lived software.  
-  (Section~\ref{SecImpactMDE})
-\item Assess the impact on safety when the MDE generated documentation targets a
-  safety assurance case.  An assurance case is an explicit structured argument
-  pertaining to specific properties, such as trustworthiness.  The evidence for
-  the argument will come from the traceability between artifacts generated by
-  model transformations, expert reviews, test cases, etc.
-  (Section~\ref{SecAssuranceCase})
-\item Assess the impact on software quality when using MDE to facilitate
-  Computational Variability Testing (CVT).  CVT uses code generation to build
-  confidence in the generated code in an analogous way to the use of grid
-  refinement studies.  Grid refinement looks at how the solution changes by
-  varying the run-time parameter of grid density and comparing the results to
-  theoretical expectations.  CVT, on the other hand, can generate code to
-  ``refine'' build time parameters, such as order of interpolation, or degree of
-  implicitness.  These parameters can be systematically varied and the results
-  compared against the expected trend.  (Section~\ref{SecCVT})
-\end{enumerate}
-
-Before presenting the details of the short-term objectives
-(Section~\ref{SecMethodologies}), the necessary background information is
-provided, along with a literature review (Section~\ref{SecBackgroundLitReview}).
-The literature review covers software quality (Sections~\ref{SecSoftwareQuality}
-and~\ref{SecDesirableQs}), the literature on the current state of the practice
-for SCS development (Section~\ref{SecCurrStateOfThePractice}), potential target
-documentation artifacts for SCS (Section~\ref{SecDocumentation}), MDE and code
-generation (Section~\ref{SecMDE}) and empirical methods for software engineering
-(Section~\ref{SecEmpirical}).  This document ends with a summary of the
-potential contributions to knowledge from this study
-(Section~\ref{SecContribKnowledge}) and a research schedule
-(Section~\ref{SecSchedule}).
-
-\section{Background and Literature Review} \label{SecBackgroundLitReview}
+\section{Qualities} \label{SecQualities}
 
 To assess the impact of MDE on SCS quality, we need a clear definition of what
-we mean by quality.  Section~\ref{SecSoftwareQuality} shows how the concept of
+we mean by quality.  The concept of
 quality is decomposed into a set of separate qualities.  This set of qualities
@@
 Section~\ref{SecSoftwareQuality}.
-
-The other subsection in this background section provide an overview of the
-literature on the current approaches to developing SCS software
-\ref{SecCurrStateOfThePractice}, recommended document artifacts for SCS
-\ref{SecDocumentation}, an overview of MDE and Code Generation in an SCS context
-\ref{SecMDE}, and a summary of existing literature on empirical methods for
-software engineering \ref{SecEmpirical}.
-
-\subsection{Software Qualities of Interest} \label{SecSoftwareQuality}
 
@@
 
-\begin{description}%[leftmargin=\the\parindent]
+\subsection*{Installability}
 
-\item [\textbf{Installability}] A measure of the ease of installation.
+A measure of the ease of installation.
 
@@
 
-\subsection{Literature Review on Current State of the Practice for SCS
-  Development} \label{SecCurrStateOfThePractice}
-
-The SCS community is finally realizing that current practices are not
-sustainable.  \citet{FaulkEtAl2009} observe, ``growing concern about the
-reliability of scientific results based on ... software.''  Embarrassing
-failures have occurred, like a retraction of derived molecular protein
-structures \citep{Miller2006}, false reproduction of sonoluminescent fusion
-\citep{PostAndVotta2005}, and fixing and then reintroducing the same error in a
-large code base three times in 20 year \citep{MilewiczAndRaybourn2018}.  A
-recent report on directions for SCS research and education states: ``While the
-volume and complexity of [SCS] have grown substantially in recent decades, [SCS]
-traditionally has not received the focused attention it so desperately needs
-... to fulfill this key role as a cornerstone of long-term collaboration and
-scientific progress'' \citep{RudeEtAl2018}.  Estimates suggest that the number
-of released faults per thousand executable lines of code during a given
-program’s life cycle is at best 0.1, and more likely 10 to 100 times worse
-\citep{Hatton2007}.
-
-Although reproducibility is the cornerstone of the scientific method, until
-recently it has not been treated seriously in software
-\citep{BenureauAndRougier2017}.  Fortunately, in recent years multiple
-conferences, workshops and individuals are calling for dramatic change
-\citep{BaileyEtAl2016}.  The need for action is highlighted by a study of 402
-computer systems papers - only 48.3\% of the code was both available and
-compilable \citep{CollbergEtAl2015}.  (Drasil addresses this problem because as
-programming languages evolve the code renderers in Drasil can be updated.)
-Reproducibility problems are even more extreme when the goal is replicability.
-A third party should be able to repeat a study using only the description of the
-methodology from a published article, with no access to the original code or
-computing environment \citep{BenureauAndRougier2017}.  However, replicability is
-rarely achieved, as shown for microarray gene expression
-\citep{IoannidisEtAl2009} and for economics modelling
-\citep{IonescuAndJansson2013}.  Drasil addresses completeness and ambiguity
-problems, since it emphasizes capturing and documenting all of the required
-knowledge, including derivation of equations and
-rationales. \citet{CrickAndHall2014} point out potential roadblocks for
-reproducibility, including page length constraints and differing detail needs
-depending on the audience.  Again Drasil addresses these concerns because the
-recipes used to generate the documentation can be tailored to the level of
-detail required.
-
-\emph{Although scientists recognize the seriousness of their SCS problems, their
-  corrective steps are too incremental.}  For instance, a recent proposal for
-the future of High Energy Physics (HEP) software \cite{StewartEtAl2017} uses
-words like sustainability, maintainability and reproducibility, but is almost
-completely silent on how these qualities are to be achieved.  The proposal
-mentions developing new and improved algorithms (including parallel computing
-and machine learning), programming tools, recruitment and training, but there is
-little on documentation, design or verification techniques (other than unit
-testing).  Similarly, a recent proposal on future directions for SCS research
-and education \citep{RudeEtAl2018} recognizes the desperate need for change, but
-then only suggests training on project management tools, open sharing and
-ethics.  \emph{Incremental change is not adequate; we need transformative
-  change.}
-
-Thankfully SCS leaders recognize that an interdisciplinary approach provides the
-path forward.  They believe that the solution to SCS quality problems is
-applying, adapting and developing SE methods, tools and techniques.  However,
-typical software processes are a barrier to progress.  ``To break the gridlock,
-we must establish a degree of cooperation and collaboration with the [SE]
-community that does not yet exist'' \citep{FaulkEtAl2009}.  ``There is a need to
-improve the transfer of existing practices and tools from ... [SE] to scientific
-programming. In addition, ... there is a need for research to specifically
-develop methods and tools that are tailored to the domain'' \citep{Storer2017}.
-This tailored research will require teams with a multidisciplinary background.
-
-\subsection{Documentation for SCS} \label{SecDocumentation}
-
-Table~\ref{TblDocuments} shows the recommended documentation for a scientific
-software project.  The documents are typical of what is suggested for scientific
-software certification, where certification consists of official recognition by
-an authority, or regulatory body, that the software is fit for its intended use.
-For instance, the Canadian Standards Association (CSA) requires a similar set of
-documents for quality assurance of scientific programs for nuclear power
-plants~\cite{CSA1999}.
-
-\begin{table}[!h]
-  \caption{Recommended Documentation} \label{TblDocuments}
-
-  \begin{tabular}{l p{8cm}}
-    %\hline
-    %\toprule
-    \textit{Problem Statement} & Description of problem to be solved\\
-    \textit{Development Plan} & Overview of development process/infrastructure\\
-    \textit{Requirements} & Desired functions and qualities of the software\\
-    \textit{V\&V Plan} & Verification that all documentation artifacts,
-                           including the code, are internally correct.
-                           Validation, from an external viewpoint, that the right
-                           problem, or model, is being solved.\\
-    \textit{Design Specification} & Documentation of how the requirements are to
-                                    be realized, through both a software
-                                    architecture and detailed design of modules
-                                    and their interfaces\\
-    \textit{Code} & Implementation of the design in code\\
-    \textit{V\&V Report} & Summary of the V\&V efforts,
-                             including testing\\
-    \textit{User Manual} & Instructions on installation, usage; worked examples\\
-    %\hline
-    %\bottomrule
-  \end{tabular}
-\end{table}
-
-\wss{Add figure showing the V-model of the documentation.}
-
-\subsection{MDE and Code Generation} \label{SecMDE}
-
-DSLs and code/document generation provide a transformative technology for
-documentation, design and verification \citep{JohansonAndHasselbring2018,
-  Smith2018}.  Scientists prefer focusing on science not software
-\citep{Kelly2007}.  DSLs allow them to do this.  A generative approach removes
-the maintenance nightmare of documentation duplicates and near duplicates
-\citep{LucivEtAl2018}, since knowledge is only captured once and automatically
-transformed as needed.  Code generation has previously been applied to improve
-SCS \citep{WhaleyEtAl2001, Veldhuizen1998, Pueschel2001}.  For instance, ATLAS
-(Automatically Tuned Linear Algebra Software) \citep{WhaleyEtAl2001} and Blitz++
-\citep{Veldhuizen1998} produces efficient and portable linear algebra software.
-Spiral \citep{Pueschel2001} uses software/hardware generation for digital signal
-processing.  \citet{Carette2008} shows how to generate a family of efficient,
-type-safe Gaussian elimination algorithms.  FEniCS (Finite Element and
-Computational Software) \citep{LoggEtAl2012} uses code generation when solving
-differential equations.  Unlike previous work on SCS code generation, Drasil
-(\href{https://github.com/JacquesCarette/Drasil}
-{https://github.com/JacquesCarette/Drasil}) focus on generating all software
-artifacts (requirements, design etc.), not just code.
-
-Drasil, as presented in \citet{SzymczakEtAl2016} removes excuses for avoiding
-documentation by providing transformative SCS development technology.  Drasil
-provides an infrastructure for knowledge capture and document/code generation.
-Drasil is implemented via Domain Specific Languages (DSLs) embedded in Haskell.
-Drasil generates requirements documentation and code for several case studies,
-including simulating the temperature of a solar water heating system, glass
-breakage and two-dimensional game physics.  The full documentation
-(requirements, design etc), code and test cases have been created manually for
-each case study.  These manual case studies (many from
-\citet{SmithJegatheesanAndKelly2016}) provide the `gold standard' against which
-Drasil is tested.  All of these artifacts are available publicly at
-\href{https://jacquescarette.github.io/Drasil/}
-{https://jacquescarette.github.io/Drasil/}.  The generation techniques for
-Drasil began to take shape during work on generating geometric data structures
-and functions \citep{CaretteEtAl2011}.
-
-\subsection{Empirical Methods for Software Engineering} \label{SecEmpirical}
-
-The empirical methods in this study will follow the guidelines given in
-\citet{KitchenhamEtAl2002}.  Since case studies will be part of the methodology,
-the guidelines for conducting and reporting case study research in software
-engineering \citep{RunesonAndHost2009} will be followed.
-
-\section{Research Design and Methodologies} \label{SecMethodologies}
-
-This section is decomposed into 4 subsections -- one for each of the different
-approaches for assessing the impact of MDE on the sustainability of SCS.  The
-goal of the first section is to establish a ``baseline'' on the current state of
-the practice for SCS.  This is done by identifying domains within SCS where
-significant software development work has been done and then systematically
-assessing the quality of the existing work.  The next section looks specifically
-at the impact of MDE on SCS through a series of case studies.  An ``ideal'' MDE
-process is proposed and then followed with different stakeholders and
-qualitative data is collected.  The last two sub-sections discuss assessing the
-impact on SCS of assurance cases and computational variability testing,
-respectively.
-
-\subsection{State of the Practice} \label{SecStateOfThePractice}
-
-A survey of existing SCS software will identify patterns, find opportunities for
-improvement and identify candidate program families for future implementation.
-The survey will require developing terminology and a classification system to
-answer such questions as: What are the current best practises?  What software
-packages are reused and why these packages?  What do practitioners look for in
-the tools that support their work?
-
-To achieve objective S2, a measurement template, possibly using pairwise
-comparisons of quality metrics between software packages, will be developed for
-assessing the quality of the documentation, code, test cases and development
-processes for existing SCS program families.  The quality measurement template
-will be used to assess the quality of approximately 30 different SCS families.
-About half of the families will be model oriented and the other half tool
-oriented.  Additional details on best practices will be collected by students
-interviewing code owners.  Based on preliminary discussions, Dr.\ Jeffrey Carver
-from the University of Alabama and some of his colleagues may also participate
-in collecting interview data. Once all the surveys and interviews are
-completed, a meta-analysis will draw conclusions for each domain and between
-domains.  Knowing the state of practice for family development will highlight
-which domains are using best practices that should be emulated in LSS.
-
-The plan here is to return to the task of measuring/assessing the state of
-software development practice in several scientific computing domains.  We will
-update the work that was done previously for domains such as Geographic
-Information Systems~\cite{SmithEtAl2018_arXivGIS}, Mesh
-Generators~\cite{SmithEtAl2016}, Seismology software~\cite{SmithEtAl2018}, and
-Statistical software for psychology~\cite{SmithEtAl2018_StatSoft}.  We could
-return to these domains and/or introduce new domains.  Potential domains/collaborators
-are listed in the subsections.
-
-MEng students will be asked to assess the state of practice in each domain.
-Last time each student measured two domains, so we would start off with this
-model again.  If the scope becomes large enough, we might switch to one domain
-per student.
-
-In the previous project, we measured 30 software projects for each domain.  With
-the increased scrutiny required in this re-boot, we won't likely be able to
-measure the details on this high a number of software projects.  We may still
-start with 30 software examples in each domain, but then use some criteria to
-create a shorter sub-list for detailed measurement.  The details of how to
-proceed here would be determined and documented in the measurement protocol.
-
-In the previous state of practice assessing exercise, a series of questions and
-simple metrics was devised to measure the quality of the documentation and
-adherence to best software development practices.  The new project will
-critically assess the previous set of questions and revise as necessary.  In
-addition, the following data will be collected/developed for each domain:
-
-\begin{itemize}
-\item Characterization of the functionality provided by the software in the
-  domain.  What services does each member of the domain provide?  Ideally this
-  information would be summarized in a commonality analysis document.  The
-  document will summarize the domain software via its commonalities,
-  variabilities and parameters of variation.
-\item Usability testing of each software in the domain.  Specific metrics for
-  assessing usability still need to be researched, but measures that involve the
-  time to complete a task and the users overall impression would be considered.
-\item Collect empirical software engineering related data, such as the number of
-  files, number of lines of code, cyclomatic complexity, number of open issues,
-  etc.  As a starting point for tool support, HubListener
-  (\url{https://github.com/pjmc-oliveira/HubListener}) could be used.
-\item Collect at least one empirical measure of the quality of the documentation
-  - the number of lines of documentation.  [Has anyone previously looked at this
-  metric?]
-\end{itemize}
-
-The above data will be combined to rank the software in the domain.  The
-specifics are yet to be determined, but in the past the Analytic Hierarchy
-Process proved helpful in this.
-
-Key to the success of this exercise will be to involve and engage a partner for
-each state of practice project.  In the previous effort we did not involve domain
-experts with the rationale of excluding potential bias, but this advantage is
-not worth not being able to evaluate the functionality/usability of the software.
-Moreover, not having an expert makes publication more difficult, since there is
-no one to advice on how best to approach journals and publishers.  The domain
-expert will be asked to help in the following ways:
-
-\begin{itemize}
-\item Review the protocol for assessing/measuring quality.  This same protocol
-  will be used in each domain.
-\item Provide their expertise on potential publication.  Specifically, they will
-  recommend a suitable journal and act as the corresponding author for any paper
-  submissions. 
-\item Provide an authoritative list of domain software, possibly augmented by
-  existing on-line lists.
-\item Provide some assistance from their own team of supervised students to
-  facilitate software testing and domain characterization.  Assessing the
-  functionality and measuring usability will require an individual with domain
-  knowledge.  Multiple measurements will be necessary to have confidence.  The
-  measurements will likely take a few hours over a few days of time from each
-  student volunteer.
-\end{itemize}
-
-There is no budget for this project, but the student volunteers will be
-considered co-authors in the resulting paper.  Having them as co-authors also
-means that ethics approval should not be necessary.
-
-GitHub will be used to coordinate the work of the large team of people that will
-be involved in this project.  In addition to the project record left on GitHub,
-the final data will be exported to Mendeley.
-
-Determine criteria for what makes a domain a good fit.
-
-Write up experimental procedure as a publication - on arxiv, but hopefully
-somewhere else too.
-
-Start from previous work and modify.  Brainstorm how to do measures.
-
-Schedule meeting.
-
-Potential collaborators/domains are listed in the following subsections.
-
-Rule 4: Keep knowledge up to date and findable \citep{ShollerEtAl2019}.
-
-\subsubsection{Medical Imaging}
-
-Medical imaging and analysis software (Dr.\ Michael Noseworthy)?
-
-\subsubsection{Climate Modelling Software}
-
-Climate modelling software (Dr.\ Zoe Li or Dr.\ Xander Wang)?
-
-\subsubsection{Computational Medicine}
-
-Extraction of 3D geometries from medical images (Dr.\ Zahra Motamed)?  Lattice
-Boltzmann Solver?
-
-\subsubsection{Finite Element Analysis}
-
-Finite element solvers and/or mesh generators (Dr.\ Dieter Stolle)?  Potential
-software includes FeNiCS, MOOSE, FreeFEM++, Dolphin, etc.  Install the packages
-and use them to do something non-trivial.
-
-\subsubsection{Psychology Software}
-
-Statistics software for psychology (Dr.\ Karin Humphreys)?
-
-\subsubsection{Chemical Engineering Software}
-
-Li Xi (\url{https://www.eng.mcmaster.ca/chemeng/people/faculty/li-xi})?
-
-\subsubsection{Geographic Information Systems}
-
-Geographic Information Systems (Jason Brodeur and/or Patrick DeLuca)?
-
-\subsubsection{Physical Metallurgy Software}
-
-Solidification and casting software (Dr.\ Andr\'e Phillion)?
-
-\subsubsection{Quantum Chemistry Software}
-
-Quantum chemistry software (Dr.\ Paul Ayers)?
-
-\subsection{Impact of MDE on SCS} \label{SecImpactMDE}
-
-roadmap
-
-Research Question: What tools and methods from MDE are perceived by end user developers to
-  improve their productivity and the sustainability of their software.
-
-\subsubsection{Fitting Drasil into the Scientific Software Development Process} \label{Sec}
-
-Scientists generally use an approximation of agile methods for their software
-development.  In addition, documentation is generally not emphasized.  How
-should Drasil be fit into the scientific software development process?  The
-Drasil process should encourage an early intensity of thought and effort, but at
-the same time, it shouldn't require waterfall development.  Drasil supports
-frequent chance and incremental development, but where should developers start
-and how should the work proceed?  How does one create an almost empty project in
-Drasil and then borrow from existing knowledge and create new knowledge?  This
-project could start with something like the
-\href{https://jacquescarette.github.io/Drasil/examples/Projectile/srs/Projectile_SRS.html}
-{Projectile} example currently in Drasil.
-
-Drasil process should move importance of requirements and domain analysis
-earlier in the process.  The quality of the resulting software depends on the
-quality of the requirements.  As is known from historical data, the greatest
-return on investment is at the requirements state (Boehm papers).
-Unfortunately, developers don't like requirements.  Quite possibly they don't
-like the intensity of thought that is necessary to get the requirements right.
-Inspection techniques, such as task based inspection (Kelly et al) hold promise
-for getting the scientist to think deeply at an earlier stage in the development
-process.
-
-\subsubsection{etc.}
-
-Experimentally Determine Appropriate Documentation Template: The best approach
-for documenting requirements and design is an open question.  Drasil could be
-used to attempt to address this question.  Different recipes for documentation
-could be developed and compared.  Some comparison could be automated between the
-documents, such as empirical measures of their quality.  Other measures would
-require some kind of usability experiments.
-
-Empirical Measurement of Documentation: Considerable effort has been invested
-into developing metrics, like cyclomatic complexity, to attempt to quantify code
-quality, but no measures (as far as I know) exist for quantifying the quality of
-documentation.  (Khedri does have some work with Bhahati (sp?) Sanga that
-quantifies the connectivity of the documentation via the traceability graph.)
-Likely the reason that effort has not been invested to quantifying the quality
-of documentation is that is not represented formally.  With Drasil much of the
-documentation is formal.  This should facilitate automatic measures of
-documentation quality, or at least metrics that might be correlated with
-quality.
-
-Measuring the Impact of Drasil on Developing Scientific Software: This project
-is challenging to design, but would be HUGE if we were successful.  We assume
-that Drasil will improve the quality of scientific software and the efficiency
-of its production, but data to back this up would go a long way to convincing
-the scientific community.  Experiments where a given project is developed in
-parallel by two independent teams come to mind, but this would be expensive, and
-difficult to control.  Some additional thought, and resources, are necessary
-here.  If we could do a project where the scientist, or a student under their
-supervision, is a true partner in the development, we might get more convincing
-anecdotal evidence.
-
-To quantify the quality of the generated documentation, measures of information
-compression will be used.  The measure will be based on the length of the
-(family of) generated documentation, code and test cases, compared to the length
-of the generator itself.  Our measure will be based on the Minimum Description
-Length (MDL) Principle, which states that ``the more we are able to compress the
-data, the more we have learned about the data'' \citep{Grunwald2004}.  Another
-important measure will be whether the generated documentation achieves
-reproducibility. The proposed measure is reproducibility depth
-\citep{Soergel2014}.  A depth of 1 means that an independent researcher can
-start from the generated code and calculate the same results.  Achieving deeper
-reproducibility will mean the same results can be calculated for an independent
-researcher starting with the design documentation.  Each higher measure gets
-further from the source code, until the highest depth (replicability), which
-starts from the requirements and leads to identical (within acceptable error)
-results.
-
-\subsection{Impact of Assurance Cases for Building Confidence} \label{SecAssuranceCase}
-
-Some SCS needs to be verifiably safe.  We will achieve this by building a safety
-assurance case. An assurance case is an explicit structured argument pertaining
-to specific properties, such as trustworthiness
-
-Assurance cases have potential to be used in SC to improve our collective
-confidence in the developed software~\cite{SmithEtAl2018_ICSEPoster}.  Further
-work is necessary to demonstrate the value of assurance cases, but they
-certainly hold promise.  Drasil could go a long way to supporting development of
-assurance cases.  Some potential Drasil related support ideas include the
-following:
-
-\begin{itemize}
-\item Automatic generation of visual depiction of the assurance case from the
-  goals, claims, evidence etc.
-\item Automated reporting of missing evidence, incomplete arguments, etc.
-\item Coordination of requirements, design decisions, etc.  Assurance cases use
-  the same information as recommended in ``document driven'' design, but the
-  traceability and connection between the data is even more important.  A tool
-  like Drasil could go a long way in facilitating the presentation and
-  navigation of an assurance case.
-\end{itemize}
-
-Assurance cases in Drasil could potentially go beyond the scientific context.
-For instance, they could also incorporate safety cases for
-control systems.  This could potentially be of interest to McSCert.
-
-\subsection{Impact of Computational Variability Testing} \label{SecCVT}
-
-This activity supports objective S3.  The applicant's students will apply LSS
-and code generation to Computational Variability Testing (CVT), using the
-example of a family of finite element programs.  As proposed by the applicant,
-CVT uses code generation to build confidence in the generated code in an
-analogous way to the use of grid refinement studies.  Grid refinement looks at
-how the solution changes by varying the run-time parameter of grid density and
-comparing the results to theoretical expectations.  CVT, on the other hand, can
-generate code to ``refine'' build time parameters, such as order of
-interpolation, or degree of implicitness.  These parameters can be
-systematically varied and the results compared against the expected trend.
-Varying computational variabilities is not a new idea, but it is only recently
-that the option has become available of coupling this with code generation, so
-that the validity of an entire program family can be assessed at the same time.
-The experimentation with CVT will initially use the FEniCS
-package~\citep{LoggEtAl2012} for elliptical partial differential equations to
-compare elements with different orders of interpolation to verify that the
-convergence to the solution agrees with the theoretical expectations.
-
-{Design and Documentation For Family of Data Fitting Algorithms}: The fitting
-used in GlassBR and in SFS (Software for Solidification) could be made much more
-generic.  We could have a family of fitting algorithms that could be used in any
-situation where fitting is required.  A proper commonality analysis of this
-domain could potentially show the potential design decisions that bridge between
-the requirements and the design.  In the SFS example many different fitting
-routines were tried.  If the experiments could have been done easily via a
-declarative specification, considerable time would have been saved.  If the
-experiments are combined with automated testing and ``properties of a correct
-solution'' the human involvement could be reduced, so that we have partially
-automated algorithm selection.
-
-\section{Implications and Contributions to Knowledge} \label{SecContribKnowledge}
-
-We are currently putting too much trust in SCS.  Examples such as nuclear safety
-analysis and computational medicine show the significance of SCS for health and
-safety.  Although SCS developers do excellent work, we do not currently have
-enough checks and balances in place for full confidence.  More should be
-expected for documentation, design and testing.  Fortunately, SE offers
-techniques, like DSLs and code generation, that can be employed to not only
-improve SCS quality, but also reduce development costs.  Now is the time for
-change.  SCS practitioners have recognized problems with the status quo and
-multidisciplinary researchers are needed to bridge the gap between SE and SCS.
-
-Codifying medical, biomechanical and computational knowledge is challenging, but
-success will completely transform the development of safety-related software in
-medicine, science and engineering.  MDE will remove the drudgery of generating
-and maintaining documentation, code, test cases and build environments, enabling
-scientists to focus on science, engineers to focus on engineering and medical
-professionals to focus on medicine.  MDE will detect inconsistencies between
-models via inter-model consistency constraints.  As a consequence, mistakes like
-inconsistent units or assumptions cannot occur.  MDE will raise the bar so that
-we can expect an explicit argument for safety, not just code that we are
-expected to blindly trust.  With the right up-front investment of knowledge
-capture, we can have software that is long lived because the stable knowledge is
-separated from the rapidly changing modelling assumptions and design decisions.
-
 \newpage
-
-\section{Research Schedule} \label{SecSchedule}
 

### 2019-09-26 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
Mostly empty document for adding our definitions of qualities

--- 2019-09-25 StateOfPractice/QualityDefinitionsOfQuality/QualityDefinitionsOfQuality.tex
@@
 
-\title{Definining Quality} 
-\author{Spencer Smith, Jacques Carette, Olu Owojaiye, Peter Michalski and Ao Dong}
+\title{Quality Definitions of Qualities} 
+\author{Spencer Smith, Jacques Carette, Olu Owojaiye,\\Peter Michalski and Ao Dong}
 \date{\today}
@@
 
-Purpose and scope of the document.
+Purpose and scope of the document.  \wss{Needs to be filled in.  Should
+  reference the overal research proposal, and the ``state of the practice''
+  exercise in particular.}
 
-\section{Qualities} \label{SecQualities}
+The presentation is divided into two main sections: i) qualities that apply to software products,
+software artifacts and software development processes, and ii) qualities that
+are considered important for good specifications.  The specification could be a
+specification of requirements, design or a test plan.
 
-To assess the impact of MDE on SCS quality, we need a clear definition of what
-we mean by quality.  The concept of
-quality is decomposed into a set of separate qualities.  This set of qualities
-can be applied to the software artifacts (documentation, test cases, etc) and to
-the software development process itself.  Several of the qualities from the list
-in Section~\ref{SecSoftwareQuality} cannot be measured directly, such as
-maintainability.  Therefore, Section~\ref{SecDesirableQs} introduces measurable
-documentation qualities that are believed to contribute to the overall software
-qualities.  In some cases it may be necessary to use the qualities in
-Section~\ref{SecDesirableQs} to indirectly measure qualities listed in
-Section~\ref{SecSoftwareQuality}.
+\section{Qualities of Software Products, Artifacts and
+  Processes} \label{SecQualities}
+
+To assess the current state of software development, and to understand how
+future changes impact software development, we need a clear definition of what
+we mean by quality.  The concept of quality is decomposed into a set of separate
+qualities.  This set of qualities can be applied to the software product, the
+software artifacts (documentation, test cases, etc) and to the software
+development process itself.
 
@@
 
-\subsection*{Installability}
+\subsection{Installability}
 
@@
 
-\item [\textbf{Correctness}] Software is correct if it matches its specification.
+\subsection{Correctness}
 
-\item [\textbf{Verifiability}] involves ``solving the equations
-  right''~\cite[p.~23]{Roache1998}; it benefits from rational documentation
-  that systematically shows, with explicit traceability, how the governing
-  equations are transformed into code.
+Software is correct if it matches its specification.
 
-\item [\textbf{Validatability}] means ``solving the right
-  equations''~\cite[p.~23]{Roache1998}.  Validatability is improved by a
-  rational process via clear documentation of the theory and assumptions, along
-  with an explicit statement of the systematic steps required for experimental
-  validation.
+\subsection{Verifiability}
 
-\item [\textbf{Reliability}] is a critical quality for scientific software,
-  since the results of computations are meaningless, if they are not dependable.
-  Reliability is closely tied to verifiability, since the key quality to verify
-  is reliability, while the act of verification itself improves reliability.
+Verifiability involves ``solving the equations right''~\cite[p.~23]{Roache1998};
+it benefits from rational documentation that systematically shows, with explicit
+traceability, how the governing equations are transformed into code.
 
-\item [\textbf{Performance}] considerations can make certification challenging,
-  since QA becomes more difficult for more complex code.  However, as
+\subsection{Validatability}
+
+Validatability means ``solving the right equations''~\cite[p.~23]{Roache1998}.
+Validatability is improved by a rational process via clear documentation of the
+theory and assumptions, along with an explicit statement of the systematic steps
+required for experimental validation.
+
+\subsection{Reliability}
+
+Reliability is a critical quality for scientific software, since the results of
+computations are meaningless, if they are not dependable.  Reliability is
+closely tied to verifiability, since the key quality to verify is reliability,
+while the act of verification itself improves reliability.
+
+\subsection{Robustness}
+
+\subsection{Performance}
+
+Performance considerations can make certification challenging, since QA becomes
+more difficult for more complex code.  However, as
 Roache~\cite[p.~355]{Roache1998} points out, using simpler algorithms and
@@
 
-\item [\textbf{Usability}] can be a problem.  Different users, solving the same
-  physical problem, using the same software, can come up with different answers,
-  due to differences in parameter selection~\cite[p.~370]{Roache1998}.  To
-  reduce misuse, a rational process must state expected user characteristics,
-  modelling assumptions, definitions and the range of applicability of the code.
+\subsection{Usability}
 
-\item [\textbf{Maintainability}] is necessary in scientific software, since change,
-  through iteration, experimentation and exploration, is inevitable.  Models of
-  physical phenomena and numerical techniques necessarily evolve over
-  time~\cite{CarverEtAl2007, SegalAndMorris2008}.  Proper documentation,
-  designed with change in mind, can greatly assist with change management.%   QA
-  % activities need to take the need for creativity into account, while not
-  % smothering it~\cite[p.~352]{Roache1998}.
+Usability can be a problem.  Different users, solving the same physical problem,
+using the same software, can come up with different answers, due to differences
+in parameter selection~\cite[p.~370]{Roache1998}.  To reduce misuse, a rational
+process must state expected user characteristics, modelling assumptions,
+definitions and the range of applicability of the code.
 
-\item [\textbf{Reusability}] provides support for the quality of reliability,
-  since reliability is improved by reusing trusted components~\cite{Dubois2005}.
-  (Care must still be taken with reusing trusted components, since blind reuse
-  in a new context can lead to errors, as dramatically shown in the Ariane 5
+\subsection{Maintainability}
+
+Maintainability is necessary in scientific software, since change, through
+iteration, experimentation and exploration, is inevitable.  Models of physical
+phenomena and numerical techniques necessarily evolve over
+time~\cite{CarverEtAl2007, SegalAndMorris2008}.  Proper documentation, designed
+with change in mind, can greatly assist with change management.  QA activities
+need to take the need for creativity into account, while not smothering
+it~\cite[p.~352]{Roache1998}.
+
+\subsection{Reusability}
+
+Reusability provides support for the quality of reliability, since reliability
+is improved by reusing trusted components~\cite{Dubois2005}.  (Care must still
+be taken with reusing trusted components, since blind reuse in a new context can
+lead to errors, as dramatically shown in the Ariane 5
 disaster~\cite[p.~37--38]{OliveiraAndStewart2006}.)  The odds of reuse are
@@
 
-\item [\textbf{Understandability}] is necessary, since reviewers can only certify
-  something they understand.  Scientific software developers have the
-  view ``that the science a developer embeds in the code must be apparent to
-  another scientist, even ten years later''~\cite{Kelly2013}.
-  Understandability applies to the documentation and code, while usability
-  refers to the executable software.  Documentation that follows a rational
-  process is the easiest to follow.
+\subsection{Portability}
 
-\item [\textbf{Reproducibility}] is a required component of the scientific
-  method~\cite{Davison2012}.  Although QA has, ``a bad name among
-  creative scientists and engineers''~\cite[p.~352]{Roache1998}, the community
-  need to recognize that participating in QA management also improves
-  reproducibility.  Reproducibility, like QA, benefits from a consistent and
-  repeatable computing environment, version control and separating code from
+\subsection{Understandability}
+
+Understandability is necessary, since reviewers can only certify something they
+understand.  Scientific software developers have the view ``that the science a
+developer embeds in the code must be apparent to another scientist, even ten
+years later''~\cite{Kelly2013}.  Understandability applies to the documentation
+and code, while usability refers to the executable software.  Documentation that
+follows a rational process is the easiest to follow.
+
+\subsection{Interoperability}
+
+\subsection{Visibility/Transparency}
+
+\subsection{Reproducibility}
+
+Reproducibility is a required component of the scientific
+method~\cite{Davison2012}.  Although QA has, ``a bad name among creative
+scientists and engineers''~\cite[p.~352]{Roache1998}, the community need to
+recognize that participating in QA management also improves reproducibility.
+Reproducibility, like QA, benefits from a consistent and repeatable computing
+environment, version control and separating code from
 configuration/parameters~\cite{Davison2012}.
 
-\item [\textbf{Productivity}] \wss{This needs to be filled in.  Productivity is
-    an important quality for this proposal, since part of what MDE promises is
-    increased productivity.}
+\subsection{Productivity}
 
-\item [\textbf{Sustainability}] \wss{This needs to be filled in.  A search needs
-    to be done to find a good definition.  A starting point might be the recent
-    paper that defines sustainability as a combination of other qualities.}
+\subsection{Sustainability}
 
-\end{description}
+\section{Desirable Qualities of Good Specifications} \label{SecDesirableQs}
 
-\subsection{Desirable Qualities of Documentation} \label{SecDesirableQs}
+To achieve the qualities listed in Section~\ref{SecQualities}, the documentation
+should achieve the qualities listed in this section.  All but the final quality
+listed (abstraction), are adapted from the IEEE recommended practise for
+producing good software requirements~\cite{IEEE1998}.  Abstraction means only
+revealing relevant details, which in a requirements document means stating what
+is to be achieved, but remaining silent on how it is to be achieved.
+Abstraction is an important software development principle for dealing with
+complexity~\cite[p.~40]{GhezziEtAl2003}.  \citet{SmithAndKoothoor2016} present
+further details on the qualities of documentation for SCS.
 
-To achieve the qualities listed in Section~\ref{SecSoftwareQuality}, the
-documentation should achieve the qualities listed in this section.  All but the
-final quality listed (abstraction), are adapted from the IEEE recommended
-practise for producing good software requirements~\cite{IEEE1998}.  Abstraction
-means only revealing relevant details, which in a requirements document means
-stating what is to be achieved, but remaining silent on how it is to be
-achieved.  Abstraction is an important software development principle for
-dealing with complexity~\cite[p.~40]{GhezziEtAl2003}.
-\citet{SmithAndKoothoor2016} present further details on the qualities of
-documentation for SCS.
+\subsection{Completeness}
 
-\begin{description}
+Documentation is said to be complete when all the requirements of the software
+are detailed. That is, each goal, functionality, attribute, design constraint,
+value, data, model, symbol, term (with its unit of measurement if applicable),
+abbreviation, acronym, assumption and performance requirement of the software is
+defined.  The software's response to all classes of inputs, both valid and
+invalid and for both desired and undesired events, also needs to be specified.
 
-\item [\textbf{Completeness}] Documentation is said to be complete when all the
-  requirements of the software are detailed. That is, each goal, functionality,
-  attribute, design constraint, value, data, model, symbol, term (with its unit
-  of measurement if applicable), abbreviation, acronym, assumption and
-  performance requirement of the software is defined.  The software's response
-  to all classes of inputs, both valid and invalid and for both desired and
-  undesired events, also needs to be specified.
+\subsection{Consistency}
 
-\item [\textbf{Consistency}] Documentation is said to be consistent when no subset
-  of individual statements are in conflict with each other. That is, a
-  specification of an item made at one place in the document should not
-  contradict the specification of the same item at another location.
+Documentation is said to be consistent when no subset of individual statements
+are in conflict with each other. That is, a specification of an item made at one
+place in the document should not contradict the specification of the same item
+at another location.
 
-\item [\textbf{Modifiability}] The documentation should be developed in such a way
-  that it is easily modifiable so that likely future changes do not destroy the
-  structure of the document. Also it should be easy to reflect the change,
-  wherever needed in the document to maintain consistency, traceability and
-  completeness. For documentation to be modifiable, its format must be
-  structured in a way that repetition is avoided and cross-referencing is
-  employed.
+\subsection{Modifiability}
 
-\item [\textbf{Traceability}] Documentation should be traceable, as this
-  facilitates maintenance and review. If a change is made to the design or code
-  of the software, then all the documentation relating to those segments have to
-  be modified.  This property is also important for recertification.
+The documentation should be developed in such a way that it is easily modifiable
+so that likely future changes do not destroy the structure of the document. Also
+it should be easy to reflect the change, wherever needed in the document to
+maintain consistency, traceability and completeness. For documentation to be
+modifiable, its format must be structured in a way that repetition is avoided
+and cross-referencing is employed.
 
-\item [\textbf{Unambiguity}] Documentation is said to be unambiguous only when
-  every requirement's specification has a unique interpretation.  The
-  documentation should be unambiguous to all audiences, including developers,
-  users and reviewers.
+\subsection{Traceability}
 
-\item [\textbf{Correctness}] There is no direct tool or method for measuring
-  correctness. One way of building confidence in correctness is by reviewing to
-  ensure that each requirement stated is one that the stakeholders and experts
-  desire.  By maintaining traceability, consistency and unambiguity, we can
-  reduce the occurrence of errors and make the goal of reviewing for correctness
-  easier.
+Documentation should be traceable, as this facilitates maintenance and
+review. If a change is made to the design or code of the software, then all the
+documentation relating to those segments have to be modified.  This property is
+also important for recertification.
 
-\item [\textbf{Verifiability}] Every requirement in the documentation must be the
-  one fulfilled by the implemented software. Therefore all the requirements
-  should be clear, unambiguous and testable, so that a person or a machine can
-  verify whether the software product meets the requirements.
+\subsection{Unambiguity}
 
-\item [\textbf{Abstract}] Documented requirements are said to be abstract if they
-  state what the software must do and the properties it must possess, but do not
-  speak about how these are to be achieved. For example, a requirement can
-  specify that an Ordinary Differential Equation (ODE) must be solved, but it
-  should not mention that Euler's method should be used to solve the ODE. How to
-  accomplish the requirement is a design decision, which is documented during
-  the design phase.
+Documentation is said to be unambiguous only when every requirement's
+specification has a unique interpretation.  The documentation should be
+unambiguous to all audiences, including developers, users and reviewers.
 
-\end{description}
+\subsection{Correctness}
+
+There is no direct tool or method for measuring correctness. One way of building
+confidence in correctness is by reviewing to ensure that each requirement stated
+is one that the stakeholders and experts desire.  By maintaining traceability,
+consistency and unambiguity, we can reduce the occurrence of errors and make the
+goal of reviewing for correctness easier.
+
+\subsection{Verifiability}
+
+Every requirement in the documentation must be the one fulfilled by the
+implemented software. Therefore all the requirements should be clear,
+unambiguous and testable, so that a person or a machine can verify whether the
+software product meets the requirements.
+
+\subsection{Abstract}
+
+Documented requirements are said to be abstract if they state what the software
+must do and the properties it must possess, but do not speak about how these are
+to be achieved. For example, a requirement can specify that an Ordinary
+Differential Equation (ODE) must be solved, but it should not mention that
+Euler's method should be used to solve the ODE. How to accomplish the
+requirement is a design decision, which is documented during the design phase.
 
@@
 \bibliographystyle {plainnat}
-\bibliography {ResearchProposal}
+\bibliography {../../CommonFiles/ResearchProposal}
 

### 2019-10-01 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
Updates to quality definitions document
Updates on quality of sustainability definitions

@@
 
-Software is correct if it matches its specification.
+Software is correct if it matches its specification.  There is no direct tool or
+method for measuring correctness. One way of building confidence in correctness
+is by reviewing to ensure that each requirement stated is one that the
+stakeholders and experts desire.  By maintaining traceability, consistency and
+unambiguity, we can reduce the occurrence of errors and make the goal of
+reviewing for correctness easier.
 
@@
 
+One of the original definitions of sustainability (for systems, not software
+specific), and still often quoted, is:
+
+\begin{quotation}
+\noindent The ability to meet the needs of the present without compromising the ability of
+future generations to meet their own needs \cite{Brundtland1987}.
+\end{quotation}
+
+This is the definition used by~\cite{IISD2019}.
+
+To make it more useful, this definition is often split into three dimensions:
+social, economic and environmental. \wss{cite UN paper [9] in
+  \cite{PenzenstadlerAndHenning2013}}  To this list Penzenstadler and Henning
+(2013) have added technical sustainability \cite{PenzenstadlerAndHenning2013}.
+Where technical sustainability for software is defined as:
+
+\begin{quotation}
+\noindent Technical sustainability has the central objective of long-time
+usage of systems and their adequate evolution with changing surrounding
+conditions and respective requirements \cite{PenzenstadlerAndHenning2013}.
+\end{quotation}
+
+The fourth dimension of technical sustainability is also added
+by~\cite{WolframEtAl2017}.  Technical sustainability is the focus on the thesis
+by~\cite{Hygerth2016}.
+
+\begin{quotation}
+  \noindent Sustainable development is a mindset (principles) and an
+  accompanying set of practices that enable a team to achieve and maintain an
+  optimal development pace indefinitely~\cite{Tate2005}.
+\end{quotation}
+
+Parnas discusses as software aging~\cite{Parnas1994a}.
+
+SCS specific definitions:
+
+\begin{quotation}
+  \noindent The concept of sustainability is based on three pillars: the
+  ecological, the economical and the social. This means that for a software to
+  be sustainable, we must take all of its effects -- direct and indirect -- on
+  the environment, the economy and the society into account. In addition, the
+  entire life cycle of a software has to be considered: from planning and
+  conception to programming, distribution, installation, usage and
+  disposal~\cite{Heine2017}.
+\end{quotation}
+
+\begin{quotation}
+  \noindent The capacity of the software to endure. In other words,
+  sustainability means that the software will continue to be available in the
+  future, on new platforms, meeting new needs~\cite{Katz2016}.
+\end{quotation}
+
+Definition from Neil Chue Hong:
+\begin{quotation}
+\noindent Sustainable software is software which is:
+-- Easy to evolve and maintain
+-- Fulfils its intent over time
+-- Survives uncertainty
+-- Supports relevant concerns (Political, Economic, Social, Technical,
+Legal, Environmental)~\cite{Katz2016}.
+\end{quotation}
+
+Paper critical of a lack of a definition~\cite{VentersEtAl2014}.
+
+Sounds like definition of maintainability.
+
+Find paper that combines nonfunctional qualities into sustainability.
+
+Sustainability depends on the software artifacts AND the software team AND the
+development process.
+
 \section{Desirable Qualities of Good Specifications} \label{SecDesirableQs}
@@
 Abstraction is an important software development principle for dealing with
-complexity~\cite[p.~40]{GhezziEtAl2003}.  \citet{SmithAndKoothoor2016} present
-further details on the qualities of documentation for SCS.
+complexity~\cite[p.~40]{GhezziEtAl2003}.  Correctness was in the above list, so
+it is not repeated here.  \citet{SmithAndKoothoor2016} present further details
+on the qualities of documentation for SCS.
 
@@
 
-\subsection{Correctness}
-
-There is no direct tool or method for measuring correctness. One way of building
-confidence in correctness is by reviewing to ensure that each requirement stated
-is one that the stakeholders and experts desire.  By maintaining traceability,
-consistency and unambiguity, we can reduce the occurrence of errors and make the
-goal of reviewing for correctness easier.
-
 \subsection{Verifiability}

### 2019-10-05 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
Added Interoperability, Visibility/Transparency, Productivity, Completness, Consistency

@@
 \usepackage{booktabs}
+\usepackage{amsmath}
 
@@
 
+\begin{quotation}
+	\noindent Interoperability is the ability of two or more systems or components to exchange information and to use the information that has been exchanged~\cite{IEEE1991}.
+	\noindent The degree to which two or more systems, products or components can exchange information and use the information that has been exchanged~\cite{ISO/IEC25010}.
+	\noindent The capability to communicate, execute programs, and transfer data among various functional units in a manner that requires the user to have little or no knowledge of the unique characteristics of those units~\cite{ISO/IEC/IEEE24765}.
+	\noindent Interoperability is a characteristic of a product or system, whose interfaces are completely understood, to work with other products or systems, present or future, in either implementation or access, without any restrictions~\cite{AFUL}.
+\end{quotation}
+
 \subsection{Visibility/Transparency}
-
+    \ad{I found little contents related to visibility or transparency. The following definition is the only one I could find, but may be irrelevant to this project.}
+    \begin{quotation}
+        1. the degree to which a transaction can access object state concurrently with other transactions. 2. the specification, for a property, of "who can see it?"~\cite{ISO/IEC/IEEE24765}
+    \end{quotation}
 \subsection{Reproducibility}
@@
 \subsection{Productivity}
+
+\begin{quotation}
+    The best definition of the productivity of a process is
+    \[Productivity = \dfrac{Outputs\ produced\ by\ the\ process}{Inputs\ consumed\ by\ the\ process}\]
+    Defining inputs.
+For the software process, providing a meaningful definition of inputs is a nontrivial but generally workable problem.
+Inputs to the software process generally comprise labor, computers, supplies, and other support facilities and equipment.
+Defining outputs.
+The big problem in defining software productivity is defining outputs.
+Here we find a paradox.
+Most sources say that defining delivered source instructions (DSI) or lines of code as the output of the software process is totally inadequate, and they argue that there are a number of deficiencies in using DSI.
+However, most organizations doing practical productivity measurement still use DSI as their primary metric~\cite{Boehm1987}.\\
+
+    \smallskip
+    Productivity is the amount of output (what is produced) per unit of input used.If we can measure the size of the software product and the effort required to develop the product, we have:
+    \begin{align}
+        productivity = size/effort
+    \end{align}
+    Equation (1) assumes that size is the output of the software production process and effort is the input to the process.
+This can be contrasted with the viewpoint of software cost models where we use size as an independent variable (i.e., an input) to predict effort which is treated as an output.
+Equation (1) is simple to operationalize if we have a single dominant size measure, for example, product size measured in lines of code~\cite{Kitchenham2004}.
+\end{quotation}
 
@@
 
+\begin{quotation}
+    A specification is complete to the extent that all of its parts are present and each part is fully developed. A software specification must exhibit several properties to assure its completeness:
+    \begin{itemize}
+        \item No TBDs. TBDs are places in the specification where decisions have been postponed by writing "To be Determined" or "TBD."
+        \item No nonexistent references. These are references in the specification to functions, inputs, or outputs (including databases) not defined in the specification.
+        \item No missing specification items. These are items that should be present as part of the standard format of the specification, but are not present.
+        \item No missing functions. These are functions that should be part of the software product but are not called for in the specification.
+        \item No missing products. These are products that should be part of the delivered software but are not called for in the specification.~\cite{Boehm1984}.
+    \end{itemize}
+
+\end{quotation}
+
 \subsection{Consistency}
@@
 at another location.
+
+\begin{quotation}
+    A specification is consistent to the extent that its provisions do not conflict with each other or with governing specifications and objectives. Specifications require consistency in several ways.
+    \begin{itemize}
+        \item Internal consistency. Items within the specification do not conflict with each other.
+        \item External consistency. Items in the specification do not conflict with external specifications or entities.
+        \item Traceability. Items in the specification have clear antecedents in earlier specifications or statements of system objectives~\cite{Boehm1984}.
+    \end{itemize}
+    Consistency requires that no two or more requirements in a specification contradict each other.
+It is also often regarded as the case where words and terms have the same meaning throughout the requirements specifications (consistent use of terminology).
+These two views of consistency imply that mutually exclusive statements and clashes in terminology should be avoided~\cite{ZOWGHI2003}.
+Consistency: 1.
+the degree of uniformity, standardization, and freedom from contradiction among the documents or parts of a system or component 2.
+software attributes that provide uniform design and implementation techniques and notations~\cite{ISO/IEC/IEEE24765}
+\end{quotation}
 

### 2019-10-05 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
updated Performance, Maintainability, Reusability

@@
 
-Performance considerations can make certification challenging, since QA becomes
-more difficult for more complex code.  However, as
-Roache~\cite[p.~355]{Roache1998} points out, using simpler algorithms and
-reducing the number of options in general purpose code, is not always a
-practical option.
+The degree to which a system or component accomplishes its designated functions within given constraints, such as speed, accuracy, or memory usage \cite{IEEEStdGlossarySET1990}.
+Performance considerations can make certification challenging, since QA becomes more difficult for more complex code.
+However, as Roache~\cite[p.~355]{Roache1998} points out, using simpler algorithms and reducing the number of options in general purpose code, is not always a practical option.
 
@@
 
-Maintainability is necessary in scientific software, since change, through
-iteration, experimentation and exploration, is inevitable.  Models of physical
-phenomena and numerical techniques necessarily evolve over
-time~\cite{CarverEtAl2007, SegalAndMorris2008}.  Proper documentation, designed
-with change in mind, can greatly assist with change management.  QA activities
-need to take the need for creativity into account, while not smothering
+The ease with which a software system or component can be modified to correct faults, improve performance or other attributes, or adapt to a changed environment \cite{IEEEStdGlossarySET1990}.
+ISO/IEC 25010 refers to maintainability as the degree of effectiveness and efficiency with which a product or system can be modified by the intended maintainers \cite{ISO/IEC25010}.
+Maintainability is necessary in scientific software, since change, through iteration, experimentation and exploration, is inevitable.
+Models of physical phenomena and numerical techniques necessarily evolve over time~\cite{CarverEtAl2007, SegalAndMorris2008}.
+Proper documentation, designed with change in mind, can greatly assist with change management.
+QA activities need to take the need for creativity into account, while not smothering
 it~\cite[p.~352]{Roache1998}.
@@
 
+The degree to which a software module or other work product can be used in more than one software system \cite{IEEEStdGlossarySET1990}.
 Reusability provides support for the quality of reliability, since reliability
 is improved by reusing trusted components~\cite{Dubois2005}.  (Care must still
-be taken with reusing trusted components, since blind reuse in a new context can
-lead to errors, as dramatically shown in the Ariane 5
-disaster~\cite[p.~37--38]{OliveiraAndStewart2006}.)  The odds of reuse are
-improved when it is considered right from the start.
+be taken with reusing trusted components, since blind reuse in a new context can lead to errors, as dramatically shown in the Ariane 5 disaster~\cite[p.~37--38]{OliveiraAndStewart2006}.)  The odds of reuse are improved when it is considered right from the start.
 

### 2019-10-06 StateOfPractice/QDefOfQualities/QDefOfQualities.tex

@@
 \subsection{Robustness}
+The degree to which a system or component can function correctly in the presence of invalid inputs or stressful environmental conditions \cite{IEEEStdGlossarySET1990}.
+The quality can be further informally refined as the ability of a software to keep an acceptable behavior, expressed in terms of robustness requirements, in spite of exceptional or unforeseen execution conditions (such as the unavailability of system resources, communication failures, invalid or stressful inputs, etc.) \cite{fernandez2005model}.
 
@@
 \subsection{Portability}
+
+The ease with which a system or component can be transferred from one hardware or software environment to another \cite{IEEEStdGlossarySET1990}.
+An application is portable across a class of environments to the degree that the effort required to transport and adapt it to a new environment in the class is less than the effort of redevelopment \cite{mooney1990strategies}.
 

### 2019-10-08 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
added installability and correctness

@@
 
-A measure of the ease of installation.
+A measure of the ease of installation.\\
+Installability is related to the effort required to install a software in a specified environment. In other words, its capability to be installed in a specified environment.
 
@@
 unambiguity, we can reduce the occurrence of errors and make the goal of
-reviewing for correctness easier.
+reviewing for correctness easier.\\
+ The extent to which a software's specification, design and implementation is free from errors is its correctness.
+The quality of a software's operation is dependent on the degree of correctness.
+Correctness and reliability are said to have dependencies, such that if a system exhibits a high degree of correctness then it tends to be reliable.
 

### 2019-10-08 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
added reliability

@@
 closely tied to verifiability, since the key quality to verify is reliability,
-while the act of verification itself improves reliability.
+while the act of verification itself improves reliability.\\
+Reliability models can be used to predict reliability of a software product. For example measuring Mean Time to Fail (MTTF) can be a good measure for reliability
 

### 2019-10-08 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
corrected typo
added citation to definitions

@@
  The extent to which a software's specification, design and implementation is free from errors is its correctness.
-The quality of a software's operation is dependent on the degree of correctness.
+The quality of a software's operation is dependent on the degree of correctness \cite{berander2005software}.
 Correctness and reliability are said to have dependencies, such that if a system exhibits a high degree of correctness then it tends to be reliable.
@@
 while the act of verification itself improves reliability.\\
-Reliability models can be used to predict reliability of a software product. For example measuring Mean Time to Fail (MTTF) can be a good measure for reliability
+Reliability models can be used to predict reliability of a software product. For example measuring Mean Time to Fail (MTTF) can be a good measure of reliability \cite{berander2005software}
 

### 2019-10-09 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
convert to LIPIcs format

@@
-\documentclass[12pt]{article}
+\documentclass[letterpaper,cleveref]{lipics-v2019}
 
@@
 
-\begin{document}
+\title{Quality Definitions of Qualities} 
+\author{Spencer Smith}{McMaster University, Canada}{smiths@mcmaster.ca}{}{}
+\author{Jacques Carette}{McMaster University, Canada}{carette@mcmaster.ca}{}{}
+\author{Olu Owojaiye}{McMaster University, Canada}{}{}{}
+\author{Peter Michalski}{McMaster University, Canada}{}{}{}
+\author{Ao Dong}{McMaster University, Canada}{}{}{}
 
-\title{Quality Definitions of Qualities} 
-\author{Spencer Smith, Jacques Carette, Olu Owojaiye,\\Peter Michalski and Ao Dong}
+\authorrunning{Smith et al.}
+\Copyright{Spencer Smith and Jacques Carette and Olu Owojaiye and Peter Michalski and Ao Dong}
+
 \date{\today}
 
+\hideLIPIcs
+\nolinenumbers
+
+\begin{document}
 \maketitle
@@
 \tableofcontents
-
-\newpage
 
@@
 
-\bibliographystyle {plainnat}
+\bibliographystyle {plainurl}
 \bibliography {../../CommonFiles/ResearchProposal}

### 2019-10-09 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
my 4 definitions.

@@
 \usepackage{booktabs}
-\usepackage{amsmath}
+\usepackage{amsmath,amsthm}
 
@@
 %\textheight 200mm
+
+\theoremstyle{definition}
+\newtheorem{defn}{Definition}
 
@@
 we mean by quality.  The concept of quality is decomposed into a set of separate
-qualities.  This set of qualities can be applied to the software product, the
-software artifacts (documentation, test cases, etc) and to the software
-development process itself.
+components that together make up ``quality''. Unfortunately, these are
+called \emph{qualities}. These are associated to the software product,
+the software artifacts (documentation, test cases, etc) and to the software
+development process itself, and combinations thereof.
 
@@
 
-The degree to which a system or component accomplishes its designated functions within given constraints, such as speed, accuracy, or memory usage \cite{IEEEStdGlossarySET1990}.
-Performance considerations can make certification challenging, since QA becomes more difficult for more complex code.
-However, as Roache~\cite[p.~355]{Roache1998} points out, using simpler algorithms and reducing the number of options in general purpose code, is not always a practical option.
+The degree to which a system or component accomplishes its designated functions
+within given constraints, such as speed, accuracy, or memory usage
+\cite{IEEEStdGlossarySET1990}.  Performance considerations can make
+certification challenging, since QA becomes more difficult for more complex
+code.  However, as Roache~\cite[p.~355]{Roache1998} points out, using simpler
+algorithms and reducing the number of options in general purpose code, is not
+always a practical option.
 
@@
 
-Usability can be a problem.  Different users, solving the same physical problem,
-using the same software, can come up with different answers, due to differences
-in parameter selection~\cite[p.~370]{Roache1998}.  To reduce misuse, a rational
-process must state expected user characteristics, modelling assumptions,
-definitions and the range of applicability of the code.
+ISO defines usability as
+\begin{quote}
+The extent to which a product can be used by specified users to achieve
+specified goals with effectiveness, efficiency, and satisfaction in a specified
+context of use.
+\end{quote}
+
+Nielsen and (separately) Schneidermann have defined usability as part of usefulness and
+is composed of
+\begin{itemize}
+\item Learnability: How easy is it for users to accomplish basic tasks the
+  first time they encounter the design?
+\item Efficiency: Once users have learned the design, how quickly can they perform tasks?
+\item Memorability: When users return to the design after a period of not using
+  it, how easily can they re-establish proficiency?
+\item Errors: How many errors do users make, how severe are these errors, and
+  how easily can they recover from the errors?
+\item Satisfaction: How pleasant is it to use the design?
+\end{itemize}
+In that context, it makes sense to separate \emph{usefulness} into
+\emph{usability} (purely an interface concern) and \emph{utility} (in the economics
+sense of the word).
+
+There are two ISO standards covering this, namely ISO/TR 16982:202 and ISO 9241.
+
+The Interaction Design Foundation~\url{https://www.interaction-design.org/literature/topics/usability}
+further lists the following desirable outcomes:
+\begin{enumerate}
+\item It should be easy for the user to become familiar with and competent in using
+the user interface during the first contact with the website. For example, if a
+travel agent’s website is a well-designed one, the user should be able to move
+through the sequence of actions to book a ticket quickly.
+\item It should be easy for users to achieve their objective through using the
+website. If a user has the goal of booking a flight, a good design will guide
+him/her through the easiest process to purchase that ticket.
+\item It should be easy to recall the user interface and how to use it on
+subsequent visits. So, a good design on the travel agent’s site means the user
+should learn from the first time and book a second ticket just as easily.
+\end{enumerate}
+
+One core reference, for definitions and metrics, is Bevan~\cite{bevan1995measuring}.
 
@@
 
-Understandability is necessary, since reviewers can only certify something they
-understand.  Scientific software developers have the view ``that the science a
-developer embeds in the code must be apparent to another scientist, even ten
-years later''~\cite{Kelly2013}.  Understandability applies to the documentation
-and code, while usability refers to the executable software.  Documentation that
-follows a rational process is the easiest to follow.
+Understandability is artifact-dependent. What it means for a user-interface (graphical
+or otherwise) to be understandable is wildly different than what it means for the code,
+and even the user documentation.
+
+The literature here is thin and scattered.  More work will need to be done to find
+something useful.
+
+Interestingly, the business literature seems to have taken more care to define this.
+Here we encounter
+\begin{quote}
+Understandability is the concept that X should be presented
+so that a reader can easily comprehend it.
+\end{quote}
+At least this brings in the idea that the \emph{reader} is actively involved, and
+indirectly that the reader's knowledge may be relevant, as well as the
+``clarity of exposition'' of X.
+
+Section 11.2 of~\cite{adams2015nonfunctional} does have a full definition.
 
@@
 
-The documentation should be developed in such a way that it is easily modifiable
-so that likely future changes do not destroy the structure of the document. Also
-it should be easy to reflect the change, wherever needed in the document to
-maintain consistency, traceability and completeness. For documentation to be
-modifiable, its format must be structured in a way that repetition is avoided
-and cross-referencing is employed.
+Here we do seem to have a simple, if somewhat uninformative, definition:
+
+\begin{defn}
+Modifiability is the degree of ease at which changes can be made to a system,
+and the flexibility with which the system adapts to such changes.
+\end{defn}
+
+IEEE Standard 610 seems to speak about this. (which is superceded?)
 
@@
 
-Documentation should be traceable, as this facilitates maintenance and
-review. If a change is made to the design or code of the software, then all the
-documentation relating to those segments have to be modified.  This property is
-also important for recertification.
+Here the Wikipedia page~\url{https://en.wikipedia.org/wiki/Traceability} is actually
+rather informative, especially as it also lists how this concept is used in other domains.
+A generic definition that is still quite useful is
+\begin{defn}
+The capability (and implementation) of keeping track of a given set or type of
+information to a given degree, or the ability to chronologically interrelate
+uniquely identifiable entities in a way that is verifiable.
+\end{defn}
+By specializing the above to software artifacts, ``interrelate'' to 
+``why is this here'' (for forward tracing from requirements), this does
+indeed give what is meant in SE.
+
+Various standards (DO178C, ISO 26262, and IEC61508) explicitly mention it.
+
+24765-2017 - ISO/IEC/IEEE International Standard - Systems and software engineering--Vocabulary
+has a full definition, namely
+\begin{enumerate}
+\item the degree to which a relationship can be established between two or more
+products of the development process, especially products having a
+predecessor-successor or master-subordinate relationship to one another;
+\item
+the identification and documentation of derivation paths (upward) and
+allocation or flowdown paths (downward) of work products in the work product
+hierarchy;
+\item the degree to which each element in a software development
+product establishes its reason for existing; and discernible association
+among two or more logical entities, such as requirements, system elements,
+verifications, or tasks.
+\end{enumerate}
 

### 2019-10-09 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
Updates lipics-v2019 so can compile, also switched back to natbib to get bib to work
Added initial definitions for reproducibility, unambiguity, verifiability and abstract.  Closes #5

@@
     \end{quotation}
+
 \subsection{Reproducibility}
@@
 
+Reproducibility is defined as:
+
+A result is said to be reproducible if another researcher can take the original
+code and input data, execute it, and re-obtain the same result (Peng, Dominici,
+and Zeger, 2006), as cited in \citet{BenureauAndRougier2017}.
+
+The related concept of replicable is defined as:
+
+Documentation achieves replicability if the description it provides of the
+algorithms is sufficiently precise and complete for an independent researcher to
+re-obtain the results it presents.  \citep{BenureauAndRougier2017}
+
+It would be worthwhile to look for some additional definitions.
+
 \subsection{Productivity}
@@
     The best definition of the productivity of a process is
-    \[Productivity = \dfrac{Outputs\ produced\ by\ the\ process}{Inputs\ consumed\ by\ the\ process}\]
-    Defining inputs.
-For the software process, providing a meaningful definition of inputs is a nontrivial but generally workable problem.
-Inputs to the software process generally comprise labor, computers, supplies, and other support facilities and equipment.
-Defining outputs.
-The big problem in defining software productivity is defining outputs.
-Here we find a paradox.
-Most sources say that defining delivered source instructions (DSI) or lines of code as the output of the software process is totally inadequate, and they argue that there are a number of deficiencies in using DSI.
-However, most organizations doing practical productivity measurement still use DSI as their primary metric~\cite{Boehm1987}.\\
+    \[\text{Productivity} = \dfrac{\text{Outputs produced by the
+          process}}{\text{Inputs consumed by the process}}\]
+    Defining inputs. For the software process, providing a meaningful definition
+    of inputs is a nontrivial but generally workable problem. Inputs to the
+    software process generally comprise labor, computers, supplies, and other
+    support facilities and equipment. Defining outputs. The big problem in
+    defining software productivity is defining outputs. Here we find a
+    paradox. Most sources say that defining delivered source instructions (DSI)
+    or lines of code as the output of the software process is totally
+    inadequate, and they argue that there are a number of deficiencies in using
+    DSI. However, most organizations doing practical productivity measurement
+    still use DSI as their primary metric~\cite{Boehm1987}.
 
@@
     \begin{align}
-        productivity = size/effort
+        \text{productivity} = \text{size}/\text{effort}
     \end{align}
-    Equation (1) assumes that size is the output of the software production process and effort is the input to the process.
-This can be contrasted with the viewpoint of software cost models where we use size as an independent variable (i.e., an input) to predict effort which is treated as an output.
-Equation (1) is simple to operationalize if we have a single dominant size measure, for example, product size measured in lines of code~\cite{Kitchenham2004}.
+    Equation (1) assumes that size is the output of the software production
+    process and effort is the input to the process. This can be contrasted with
+    the viewpoint of software cost models where we use size as an independent
+    variable (i.e., an input) to predict effort which is treated as an
+    output. Equation (1) is simple to operationalize if we have a single
+    dominant size measure, for example, product size measured in lines of
+    code~\cite{Kitchenham2004}.
 \end{quotation}
@@
 
-Documentation is said to be unambiguous only when every requirement's
-specification has a unique interpretation.  The documentation should be
-unambiguous to all audiences, including developers, users and reviewers.
+A specification is unambiguous when it has a unique interpretation.  If there is
+a possibility that two readers will have two different interpretations, than the
+specification is ambiguous.  \wss{When I get the Ghezzi text back from Olu, I'll
+  check to see if they have anything to add to this definition.}
+
+A Software Requirements Specification (SRS) is unambiguous if, and only if,
+every requirement stated therein has only one interpretation \citep{IEEE1998}.
 
@@
 
-Every requirement in the documentation must be the one fulfilled by the
-implemented software. Therefore all the requirements should be clear,
-unambiguous and testable, so that a person or a machine can verify whether the
-software product meets the requirements.
+\begin{itemize}
+
+\item Verification - Are we building the product right?  Are we implementing the
+  requirements correctly (internal)
+\item Validation - Are we building the right product? Are we getting the right
+  requirements (external)
+\item According to
+  \href{https://en.wikipedia.org/wiki/Software_verification_and_validation}{Capability
+    Maturity Model (CMM)}
+\begin{itemize}
+\item 
+    Software Verification: The process of evaluating software to determine
+    whether the products of a given development phase satisfy the conditions
+    imposed at the start of that phase. [IEEE-STD-610]
+  \item Software Validation: The process of evaluating software during or at the
+    end of the development process to determine whether it satisfies specified
+    requirements. [IEEE-STD-610] 
+\end{itemize}
+\end{itemize}
+
+``An SRS is verifiable if, and only if, every requirement stated therein is
+verifiable. A requirement is verifiable if, and only if, there exists some
+finite cost-effective process with which a person or machine can check that the
+software product meets the requirement. In general any ambiguous requirement is
+not verifiable.'' \citep{IEEE1998}
+
+Verifiability is related to testability, which is defined by McCall et al.\ as
+``The effort required to test a program to ensure that it performs its intended
+function'' \cite{VanVliet2000}.
+
+\wss{When I get the Ghezzi text back from Olu,
+  I'll check to see if they have anything to add to this definition.}
 
@@
 must do and the properties it must possess, but do not speak about how these are
-to be achieved. For example, a requirement can specify that an Ordinary
-Differential Equation (ODE) must be solved, but it should not mention that
-Euler's method should be used to solve the ODE. How to accomplish the
-requirement is a design decision, which is documented during the design phase.
+to be achieved \citep{GhezziEtAl2003}.
+
+``An abstraction for a software artifact is a succinct description that suppresses
+the details that are unimportant to a software developer and emphasizes the
+information that is important.'' \cite{Krueger1992}
+
+``Abstraction means that we concentrate on the essential features and ignore,
+abstract from, details that are not relevant at the level we are currently
+working.''  \cite[p.\ 296]{VanVliet2000}
+
+``Abstraction in mathematics is the process of extracting the underlying essence
+of a mathematical concept, removing any dependence on real world objects with
+which it might originally have been connected, and generalizing it so that it
+has wider applications or matching among other abstract descriptions of
+equivalent phenomena.''
+\href{https://en.wikipedia.org/wiki/Abstraction_(mathematics)} {Wikipedia
+  Definition}
+
+Abstraction is related to reusability (and other qualities).
+
+\wss{When I get the Ghezzi text back from Olu, I'll check to see if they have
+  anything to add to this definition.}
 
@@
 
-\bibliographystyle {plainurl}
+\bibliographystyle {plainnat}
 \bibliography {../../CommonFiles/ResearchProposal}

### 2019-10-09 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
added some more definitions to robustness, performance, maintainability, reusability, portability

@@
 \subsection{Robustness}
-The degree to which a system or component can function correctly in the presence of invalid inputs or stressful environmental conditions \cite{IEEEStdGlossarySET1990}.
-The quality can be further informally refined as the ability of a software to keep an acceptable behavior, expressed in terms of robustness requirements, in spite of exceptional or unforeseen execution conditions (such as the unavailability of system resources, communication failures, invalid or stressful inputs, etc.) \cite{fernandez2005model}.
+\begin{defn}
+The degree to which a system or component can function correctly in the presence of invalid inputs or stressful environmental conditions \cite{IEEEStdGlossarySET1990}.\\
+\end{defn}
+\begin{defn}
+The quality can be further informally refined as the ability of a software to keep an acceptable behavior, expressed in terms of robustness requirements, in spite of exceptional or unforeseen execution conditions (such as the unavailability of system resources, communication failures, invalid or stressful inputs, etc.) \cite{fernandez2005model}.\\
+\end{defn}
+\begin{defn}
+Code possesses the characteristic of robustness to the extent that it can continue to perform despite some violation of the assumptions in its specification \cite{boehm2007software}.
+\end{defn}
 
 \subsection{Performance}
-
-The degree to which a system or component accomplishes its designated functions
-within given constraints, such as speed, accuracy, or memory usage
-\cite{IEEEStdGlossarySET1990}.  Performance considerations can make
-certification challenging, since QA becomes more difficult for more complex
-code.  However, as Roache~\cite[p.~355]{Roache1998} points out, using simpler
-algorithms and reducing the number of options in general purpose code, is not
-always a practical option.
+\begin{defn}
+The degree to which a system or component accomplishes its designated functions within given constraints, such as speed, accuracy, or memory usage \cite{IEEEStdGlossarySET1990}.\\
+\end{defn}
+\begin{defn}
+How well or how rapidly the system must perform specific functions.
+Performance requirements encompass speed (database response times, for instance), throughput ( transactions per second), capacity (con-current usage loads), and timing (hard real-time demands) \cite{wiegers2003softreq}.
+\end{defn}
 
@@
 \subsection{Maintainability}
-
+\begin{defn}
 The ease with which a software system or component can be modified to correct faults, improve performance or other attributes, or adapt to a changed environment \cite{IEEEStdGlossarySET1990}. 
+\end{defn}\
+\begin{defn}
 ISO/IEC 25010 refers to maintainability as the degree of effectiveness and efficiency with which a product or system can be modified by the intended maintainers \cite{ISO/IEC25010}.
-Maintainability is necessary in scientific software, since change, through iteration, experimentation and exploration, is inevitable.
-Models of physical phenomena and numerical techniques necessarily evolve over time~\cite{CarverEtAl2007, SegalAndMorris2008}.
-Proper documentation, designed with change in mind, can greatly assist with change management.
-QA activities need to take the need for creativity into account, while not smothering
-it~\cite[p.~352]{Roache1998}.
+\end{defn}
+\begin{defn}
+Effort required to locate and fix an error in a program
+\cite{pressman2005software}.\\
+\end{defn}
+\begin{defn}
+A set of attributes that bear on the effort needed to make specified modifications (which may include corrections, improvements, or adaptations of software to environmental changes and changes in the requirements and functional specifications)\cite{pfleeger2006software}.
+\end{defn}
 
 \subsection{Reusability}
-
-The degree to which a software module or other work product can be used in more than one software system \cite{IEEEStdGlossarySET1990}.
-Reusability provides support for the quality of reliability, since reliability
-is improved by reusing trusted components~\cite{Dubois2005}.  (Care must still
-be taken with reusing trusted components, since blind reuse in a new context can lead to errors, as dramatically shown in the Ariane 5 disaster~\cite[p.~37--38]{OliveiraAndStewart2006}.)  The odds of reuse are improved when it is considered right from the start.
+\begin{defn}
+The degree to which a software module or other work product can be used in more than one software system \cite{IEEEStdGlossarySET1990}. \\
+\end{defn}
+\begin{defn}
+Extent to which a program [or parts of a program] can be reused in other applications - related to the packaging and scope of the functions that the program performs \cite{pressman2005software}.\\
+\end{defn}
+\begin{defn}
+Code possesses the characteristic of maintainability to the extent that it facilitates updating to satisfy new requirements or to correct deficiencies \cite{boehm2007software}.
+\end{defn}
 
 \subsection{Portability}
-
-The ease with which a system or component can be transferred from one hardware or software environment to another \cite{IEEEStdGlossarySET1990}.
-An application is portable across a class of environments to the degree that the effort required to transport and adapt it to a new environment in the class is less than the effort of redevelopment \cite{mooney1990strategies}.
+\begin{defn}
+The ease with which a system or component can be transferred from one hardware or software environment to another \cite{IEEEStdGlossarySET1990}. \\
+\end{defn}
+\begin{defn}
+An application is portable across a class of environments to the degree that the effort required to transport and adapt it to a new environment in the class is less than the effort of redevelopment \cite{mooney1990strategies}.\\
+\end{defn}
+\begin{defn}
+Effort required to transfer the program from one hardware and/or software system environment to another \cite{pressman2005software}.\\
+\end{defn}
+\begin{defn}
+A set of attributes that bear on the ability of software to be transferred from one environment to another (including the organizational, hardware, of software environment)\cite{pfleeger2006software}.\\
+\end{defn}
+\begin{defn}
+Code possesses the characteristic of portability to the extent that it can be operated easily and well on computer configurations other than its current one.
+This implies that special function features, not easily available at other facilities, are not used, that standard library functions and subroutines are selected for universal applicability, and so on \cite{boehm2007software}.
+\end{defn}
 

### 2019-10-10 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
selected/reworded/added more definitions to robustness, performance, maintainability, reusability, portability

@@
 \end{defn}
+\begin{defn}
+A program is robust if it behaves "reasonably", even in circumstances that were not anticipated in the requirements specification - for example, when it encounters incorrect input data or some hardware malfunction \cite{ghezzi1991fundamentals}\\
+\end{defn}
+
+\textbf{Chosen definition:}
+To do: Boehm and Ghezzi have good definitions, will consider rewording them into one\\
 
@@
 \end{defn}
+\begin{defn}
+In software engineering we often equate performance with efficiency. A software system is efficient if it uses computing resources economically \cite{ghezzi1991fundamentals}\\
+\end{defn}
+
+\textbf{Chosen definition:}
+Wiegers - but need to reword.\\
+
+note1: may need to consider removing performance and instead using other more specific qualities that fall under the performance umbrella - to be added\\
 
@@
 \end{defn}
+\begin{defn}
+We will view maintainability as two separate qualities: repairability and evolvability.
+Software is repairable if it allows the fixing of defects; it is evolvable if it allows changes that enable it to satisfy new requirements \cite{ghezzi1991fundamentals}.
+\end{defn}
+\begin{defn}
+Code possesses the characteristic of maintainability to the extent that it facilitates updating to satisfy new requirements or to correct deficiencies \cite{boehm2007software}.
+\end{defn}
+
+\textbf{Chosen definition:}
+Will try to combine IEEE and Boehm. IEEE clearly covers everything except meeting new requirements - I do not like the wording of changed environment - does not address new requirements well imo \\
+
+note1: \\
 
@@
 \end{defn}
-\begin{defn}
-Code possesses the characteristic of maintainability to the extent that it facilitates updating to satisfy new requirements or to correct deficiencies \cite{boehm2007software}.
-\end{defn}
+
+\textbf{Chosen definition:}
+New definition: The degree to which a software product can be modularized for easy transferal into another system.\\
+
+note1: "degree to which" needs to be further defined\\
 
@@
 \end{defn}
+\begin{defn}
+Portability refers to the ability to run a system on different hardware platforms \cite{ghezzi1991fundamentals}.
+\end{defn}
+\textbf{Chosen definition:} Effort required to transfer the program from one hardware and/or software system environment to another \cite{pressman2005software}.\\
+what is effort? metric? how to measure?\\
+
+note1: "transfer" can be expanded\\
+note2: I also like Mooney definition, but the measurement of effort of redevelopment is less feasible and out of scope imo
+\newline
 

### 2019-10-12 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
first draft of chosen definitions of robustness, performance, maintainability, reusability, portability

--- 2019-10-09 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
@@
 \end{defn}
+\begin{defn}
+A program is robust if it behaves "reasonably", even in circumstances that were not anticipated in the requirements specification - for example, when it encounters incorrect input data or some hardware malfunction \cite{ghezzi1991fundamentals}\\
+\end{defn}
+
+\textbf{Chosen definition:}\\
+Combined Boehm and GhezziL\\
+Software possesses the characteristic of robustness if it fulfills its requirements correctly or behaves reasonably in circumstances not anticipated in the requirements specification.\\
 
@@
 \end{defn}
+\begin{defn}
+In software engineering we often equate performance with efficiency. A software system is efficient if it uses computing resources economically \cite{ghezzi1991fundamentals}\\
+\end{defn}
+
+\textbf{Chosen definition:}\\
+How well or how rapidly the system must perform specific functions.
+Performance requirements encompass speed (database response times, for instance), throughput ( transactions per second), capacity (con-current usage loads), and timing (hard real-time demands) \cite{wiegers2003softreq}.\\
 
@@
 \end{defn}
+\begin{defn}
+We will view maintainability as two separate qualities: repairability and evolvability.
+Software is repairable if it allows the fixing of defects; it is evolvable if it allows changes that enable it to satisfy new requirements \cite{ghezzi1991fundamentals}.
+\end{defn}
+\begin{defn}
+Code possesses the characteristic of maintainability to the extent that it facilitates updating to satisfy new requirements or to correct deficiencies \cite{boehm2007software}.
+\end{defn}
+
+\textbf{Chosen definition:}\\
+Combined IEEE and Boehm:\\
+The ease with which a software system or component can be modified to correct faults, improve performance or other attributes, or satisfy new requirements.\\
 
@@
 \end{defn}
-\begin{defn}
-Code possesses the characteristic of maintainability to the extent that it facilitates updating to satisfy new requirements or to correct deficiencies \cite{boehm2007software}.
-\end{defn}
+
+\textbf{Chosen definition:}\\
+New definition:\\ The degree to which a software product can be modularized for easy transferal into other systems.\\
 
@@
 \end{defn}
+\begin{defn}
+Portability refers to the ability to run a system on different hardware platforms \cite{ghezzi1991fundamentals}.
+\end{defn}
+\textbf{Chosen definition:}\\ Effort required to transfer the program from one hardware and/or software system environment to another \cite{pressman2005software}.\\
+Rephrased: Effort required to transfer a program between system environments.\\
+Alternatively: Ease with which a program can be transferred between system environments.
+How is effort defined here? metric? measure?\\
 

### 2019-10-22 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
added definitions for correctness and installability

@@
 \author{Jacques Carette}{McMaster University, Canada}{carette@mcmaster.ca}{}{}
-\author{Olu Owojaiye}{McMaster University, Canada}{}{}{}
+\author{Olu Owojaiye}{McMaster University, Canada}{owojaiyo@mcmaster.ca}{}{}
 \author{Peter Michalski}{McMaster University, Canada}{}{}{}
@@
 \subsection{Installability}
-
-A measure of the ease of installation.\\
-Installability is related to the effort required to install a software in a specified environment. In other words, its capability to be installed in a specified environment.
+\begin{itemize}
+    \item{Definition 1:} A measure of the ease of installation.\\
+	\item{Definition 2:} Installability is related to the amount of human effort required to install a software in a designated environment \cite{berander2005software}. \\
+	\item{Definition 3:} ``The capability of a software product to be installed in a specified environment'' \cite{berander2005software}.\\
+	\item{Definition 4:}  ``The efficacy of the installation, uninstallation or reinstallation process of a software product in a specified enviroment'' (\cite{iso201017043})\\
+	\item {\textbf{Proposed definition}:} (Definition 2)
+\end{itemize}
 
 \subsection{Correctness}
+\begin{itemize}
+	\item{Definition 1:} The degree to which a software's specification is satisfied (\cite{berander2005software}).\\
+	\item{Definition 2:} A software is correct if it functions according to its specified functional and non-functional specification(\cite{GhezziEtAl2003})\\.
+	\item{Definition 3:} According to \cite{wilson2009quality}, ``the degree to which a system is free from defects in its specification, design, and implementation, is its correctness.\\
+	\item{Definition 4:} ``The ability of software products to perform their exact tasks, as defined by their specification'' (\cite{meyer1988object}). \\
+	\item{\textbf{Proposed definition}:} Definition 2
+\end{itemize}
 
-Software is correct if it matches its specification.  There is no direct tool or
+\subsubsection{Discussion}
+There is no direct tool or
 method for measuring correctness. One way of building confidence in correctness
@@
 reviewing for correctness easier.\\
- The extent to which a software's specification, design and implementation is free from errors is its correctness.
  The quality of a software's operation is dependent on the degree of correctness \cite{berander2005software}.
-Correctness and reliability are said to have dependencies, such that if a system exhibits a high degree of correctness then it tends to be reliable.
+Correctness and reliability are said to have dependencies, such that if a system exhibits a high degree of correctness then it tends to be reliable (\cite{GhezziEtAl2003}).
 

### 2019-10-23 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
Updated bib and QDefOfQualities

@@
 
-\begin{quotation}
-	\noindent Interoperability is the ability of two or more systems or components to exchange information and to use the information that has been exchanged~\cite{IEEE1991}.
-	\noindent The degree to which two or more systems, products or components can exchange information and use the information that has been exchanged~\cite{ISO/IEC25010}.
-	\noindent The capability to communicate, execute programs, and transfer data among various functional units in a manner that requires the user to have little or no knowledge of the unique characteristics of those units~\cite{ISO/IEC/IEEE24765}.
-	\noindent Interoperability is a characteristic of a product or system, whose interfaces are completely understood, to work with other products or systems, present or future, in either implementation or access, without any restrictions~\cite{AFUL}.
-\end{quotation}
+\begin{enumerate}[(1).]
+    \item Interoperability is the ability of two or more systems or components to exchange information and to use the information that has been exchanged~\cite{IEEE1991}.
+	\item The degree to which two or more systems, products or components can exchange information and use the information that has been exchanged~\cite{ISO/IEC25010}.
+	\item The capability to communicate, execute programs, and transfer data among various functional units in a manner that requires the user to have little or no knowledge of the unique characteristics of those units~\cite{ISO/IEC/IEEE24765}.
+	\item Interoperability is a characteristic of a product or system, whose interfaces are completely understood, to work with other products or systems, present or future, in either implementation or access, without any restrictions~\cite{AFUL}.
+    \item \ad{My recommendation is the following one}\\
+    Interoperability is the ability of different information systems, devices and applications (‘systems’) to access, exchange, integrate and cooperatively use data in a coordinated manner, within and across organizational, regional and national boundaries, to provide timely and seamless portability of information~\cite{HIMSS}.
+
+    Four Levels of Interoperability:
+    \begin{itemize}
+        \item Foundational (Level 1) – establishes the inter-connectivity requirements needed for one system or application to securely communicate data to and receive data from another
+
+        \item Structural (Level 2) – defines the format, syntax, and organization of data exchange including at the data field level for interpretation
+
+        \item Semantic (Level 3) – provides for common underlying models and codification of the data including the use of data elements with standardized definitions from publicly available value sets and coding vocabularies, providing shared understanding and meaning to the user
+
+        \item “New” Organizational (Level 4) – includes governance, policy, social, legal and organizational considerations to facilitate the secure, seamless and timely communication and use of data both within and between organizations, entities and individuals.
+These components enable shared consent, trust and integrated end-user processes and workflows
+    \end{itemize}
+\end{enumerate}
 
@@
 \subsection{Productivity}
+\begin{enumerate}[(1).]
+    \item
 
-\begin{quotation}
     The best definition of the productivity of a process is
@@
     Defining inputs. For the software process, providing a meaningful definition
-    of inputs is a nontrivial but generally workable problem. Inputs to the
-    software process generally comprise labor, computers, supplies, and other
-    support facilities and equipment. Defining outputs. The big problem in
+    of inputs is a nontrivial but generally workable problem.
+Inputs to the software process generally comprise labor, computers, supplies, and other support facilities and equipment.
+Defining outputs.
+The big problem in
     defining software productivity is defining outputs. Here we find a
@@
 
-    \smallskip
+    \item \ad{My recommendation is the following one}\\
     Productivity is the amount of output (what is produced) per unit of input used.If we can measure the size of the software product and the effort required to develop the product, we have:
@@
     code~\cite{Kitchenham2004}.
-\end{quotation}
+\end{enumerate}
 
@@
 \subsection{Completeness}
+\begin{enumerate}[(1).]
+    \item
+    \ad{This part was here before my modification}\\
+    Documentation is said to be complete when all the requirements of the software are detailed.
+That is, each goal, functionality, attribute, design constraint, value, data, model, symbol, term (with its unit of measurement if applicable), abbreviation, acronym, assumption and performance requirement of the software is defined.
+The software's response to all classes of inputs, both valid and invalid and for both desired and undesired events, also needs to be specified.
 
-Documentation is said to be complete when all the requirements of the software
-are detailed. That is, each goal, functionality, attribute, design constraint,
-value, data, model, symbol, term (with its unit of measurement if applicable),
-abbreviation, acronym, assumption and performance requirement of the software is
-defined.  The software's response to all classes of inputs, both valid and
-invalid and for both desired and undesired events, also needs to be specified.
-
-\begin{quotation}
+    \item
+    \ad{My recommendation is the following one}\\
     A specification is complete to the extent that all of its parts are present and each part is fully developed. A software specification must exhibit several properties to assure its completeness:
@@
     \end{itemize}
-
-\end{quotation}
+\end{enumerate}
 
 \subsection{Consistency}
+\begin{enumerate}[(1).]
+    \item
+    \ad{This part was here before my modification}\\
+    Documentation is said to be consistent when no subset of individual statements are in conflict with each other.
+That is, a specification of an item made at one place in the document should not contradict the specification of the same item at another location.
 
-Documentation is said to be consistent when no subset of individual statements
-are in conflict with each other. That is, a specification of an item made at one
-place in the document should not contradict the specification of the same item
-at another location.
-
-\begin{quotation}
-    A specification is consistent to the extent that its provisions do not conflict with each other or with governing specifications and objectives. Specifications require consistency in several ways.
+    \item    A specification is consistent to the extent that its provisions do not conflict with each other or with governing specifications and objectives.
+Specifications require consistency in several ways.
     \begin{itemize}
@@
     \end{itemize}
-    Consistency requires that no two or more requirements in a specification contradict each other.
+    \item Consistency requires that no two or more requirements in a specification contradict each other.
 It is also often regarded as the case where words and terms have the same meaning throughout the requirements specifications (consistent use of terminology).
 These two views of consistency imply that mutually exclusive statements and clashes in terminology should be avoided~\cite{ZOWGHI2003}.
-Consistency: 1.
+    \item Consistency: 1.
 the degree of uniformity, standardization, and freedom from contradiction among the documents or parts of a system or component 2.
 software attributes that provide uniform design and implementation techniques and notations~\cite{ISO/IEC/IEEE24765}
-\end{quotation}
+\end{enumerate}
 

### 2019-10-23 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
more quality definitions added

--- 2019-10-22 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
@@
 \subsection{Verifiability}
-
+\begin{itemize}
+	\item{Definition 1:} A software is verifiable if its properties can be verified (\cite{GhezziEtAl2003}).\\
+	\item{Definition 2:} Attributes of software that relate to the effort needed for validating the modified software (\cite{berander2005software}).\\
+	\item{Definition 3:}) The degree to which a system or component facilitates the establishment of test criteria and the performance of tests to determine whether those criteria have been met (\cite{IEEEStdGlossarySET1990}).\\
+	\item{Definition 4:}  The degree to which a requirement is stated in terms that permit establishment of test criteria and performance (\cite{IEEEStdGlossarySET1990}).\\
+	\item{Definition 5:} Ease of performing testing on a software product or system (\cite{IEEEStdGlossarySET1990}).\\
+	\item{\textbf{Proposed definition}:} Combo of definitions 4 and 5 ?\\
+\end{itemize}
+\subsubsection{Discussion}
 Verifiability involves ``solving the equations right''~\cite[p.~23]{Roache1998};
@@
 traceability, how the governing equations are transformed into code.
-
+\oo{Verifiability is sometimes referred to as testability, so I culled some testability definitions here}\\
 \subsection{Validatability}
-
-Validatability means ``solving the right equations''~\cite[p.~23]{Roache1998}.
+\begin{itemize}
+\item{Definition 1:} Validatability of a software is the degree of ease in validating(checking) that software meets user needs.\\
+\item{Definition 2:} Validatability means ``solving the right equations''~\cite[p.~23]{Roache1998}.
+\oo{question about this definition}\\
+\item{Definition 3:} The degree of ease of evaluating software products or system to determine whether it satisfies specified business requirements.
+\url{http://softwaretestingfundamentals.com/verification-vs-validation/}\\
+\oo{Here I tweaked the meaning of validation to suit validatability}\\
+\oo{Only a few resource on validatability}\\
+\item{\textbf{Proposed definition}:} Definition 3
+\end{itemize}
+\subsubsection{Discussion}
 Validatability is improved by a rational process via clear documentation of the
@@
 \subsection{Reliability}
+\begin{itemize}
+	\item{Definition 1:} The probability that the software will operate as expected over a specified time interval (\cite{GhezziEtAl2003}).\\
+	\item{Definition 2:} A set of attributes that relate to the capability of software to maintain its level of performance under stated conditions for a stated period of time (\cite{berander2005software}).\\
+	\item{Definition 3:} “The capability of the software product to maintain a specified level of performance when used under specified conditions (\cite{international2001iso}).\\
+	\item{Definition 4:} Code possesses the characteristic reliability to the extent that it can be expected to perform its intended functions satisfactorily (\cite{boehm2007software})\\
+	\item{Definition 5:} A concern encompassing correctness and robustness (\cite{meyer1988object}).\\
+	\item{Proposed definition}\\
+\end{itemize}
 
+\subsubsection{Discussion}
 Reliability is a critical quality for scientific software, since the results of

### 2019-10-23 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
Added begin{defn} ... \end{defn} to several definitions

@@
 
+\begin{defn}
 A result is said to be reproducible if another researcher can take the original
@@
 and Zeger, 2006), as cited in \citet{BenureauAndRougier2017}.
+\end{defn}
 
@@
 
+\begin{defn}
 Documentation achieves replicability if the description it provides of the
@@
 re-obtain the results it presents.  \citep{BenureauAndRougier2017}
+\end{defn}
 
@@
 
-\begin{quotation}
+\begin{defn}
 \noindent The ability to meet the needs of the present without compromising the ability of
 future generations to meet their own needs \cite{Brundtland1987}.
-\end{quotation}
+\end{defn}
 
@@
 
-\begin{quotation}
+\begin{defn}
 \noindent Technical sustainability has the central objective of long-time
@@
 conditions and respective requirements \cite{PenzenstadlerAndHenning2013}.
-\end{quotation}
+\end{defn}
 
@@
 
-\begin{quotation}
+\begin{defn}
   \noindent Sustainable development is a mindset (principles) and an
@@
   optimal development pace indefinitely~\cite{Tate2005}.
-\end{quotation}
+\end{defn}
 
@@
 
-\begin{quotation}
-  \noindent The concept of sustainability is based on three pillars: the
+\begin{defn}
+  The concept of sustainability is based on three pillars: the
   ecological, the economical and the social. This means that for a software to
@@
   disposal~\cite{Heine2017}.
-\end{quotation}
+\end{defn}
 
-\begin{quotation}
+\begin{defn}
   \noindent The capacity of the software to endure. In other words,
@@
   future, on new platforms, meeting new needs~\cite{Katz2016}.
-\end{quotation}
+\end{defn}
 
 Definition from Neil Chue Hong:
-\begin{quotation}
-\noindent Sustainable software is software which is:
+\begin{defn}
+Sustainable software is software which is:
 -- Easy to evolve and maintain
@@
 Legal, Environmental)~\cite{Katz2016}.
-\end{quotation}
+\end{defn}
 
@@
 
+\begin{defn}
 Documented requirements are said to be abstract if they state what the software
@@
 to be achieved \citep{GhezziEtAl2003}.  
+\end{defn}
 
+\begin{defn}
 ``An abstraction for a software artifact is a succinct description that suppresses
@@
 information that is important.'' \cite{Krueger1992}
+\end{defn}
 
+\begin{defn}
 ``Abstraction means that we concentrate on the essential features and ignore,
@@
 working.''  \cite[p.\ 296]{VanVliet2000}
+\end{defn}
 
+\begin{defn}
 ``Abstraction in mathematics is the process of extracting the underlying essence
@@
   Definition}
+\end{defn}
 

### 2019-10-23 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
Correctness definitions - fixed formatting as definitions, natbib citations, added new def

@@
 \subsection{Correctness}
-\begin{itemize}
-	\item{Definition 1:} The degree to which a software's specification is satisfied (\cite{berander2005software}).\\
-	\item{Definition 2:} A software is correct if it functions according to its specified functional and non-functional specification(\cite{GhezziEtAl2003})\\.
-	\item{Definition 3:} According to \cite{wilson2009quality}, ``the degree to which a system is free from defects in its specification, design, and implementation, is its correctness.\\
-	\item{Definition 4:} ``The ability of software products to perform their exact tasks, as defined by their specification'' (\cite{meyer1988object}). \\
-	\item{\textbf{Proposed definition}:} Definition 2
-\end{itemize}
+
+\begin{defn}
+  The degree to which a software's specification is satisfied
+  \citep{berander2005software}.
+\end{defn}
+
+\begin{defn} \label{CorrectDefnSelected}
+  A software is correct if it functions according to its specified functional
+  and non-functional specification \cite{GhezziEtAl2003}.
+\end{defn}
+
+\begin{defn}
+  According to \citet{wilson2009quality}, ``the degree to which a system is free
+  from defects in its specification, design, and implementation, is its
+  correctness.''
+\end{defn}
+
+\begin{defn}
+  ``The ability of software products to perform their exact tasks, as defined by
+  their specification'' \citep{meyer1988object}.
+\end{defn}
+
+\begin{defn}
+  The extent to which a program satisfies its specifications and fulfills the
+  user's mission objectives \citep{McCallEtAl1977}.
+\end{defn}
+
+\noindent \textbf{Proposed definition}: Definition~\ref{CorrectDefnSelected}
 

### 2019-10-23 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
Added van Vliet reference to new correctness def, to clarify the source
Some reformatting to fix overlong lines

@@
 
-The presentation is divided into two main sections: i) qualities that apply to software products,
-software artifacts and software development processes, and ii) qualities that
-are considered important for good specifications.  The specification could be a
-specification of requirements, design or a test plan.
+The presentation is divided into two main sections: i) qualities that apply to
+software products, software artifacts and software development processes, and
+ii) qualities that are considered important for good specifications.  The
+specification could be a specification of requirements, design or a test plan.
 
@@
   A software is correct if it functions according to its specified functional
-  and non-functional specification \cite{GhezziEtAl2003}.
+  and non-functional specification \citep{GhezziEtAl2003}.
 \end{defn}
@@
   The extent to which a program satisfies its specifications and fulfills the
-  user's mission objectives \citep{McCallEtAl1977}.
+  user's mission objectives \citep{McCallEtAl1977}. (As summarized in
+  \citet{VanVliet2000}.)
 \end{defn}
@@
 \subsubsection{Discussion}
-There is no direct tool or
-method for measuring correctness. One way of building confidence in correctness
-is by reviewing to ensure that each requirement stated is one that the
-stakeholders and experts desire.  By maintaining traceability, consistency and
-unambiguity, we can reduce the occurrence of errors and make the goal of
-reviewing for correctness easier.\\
- The quality of a software's operation is dependent on the degree of correctness \cite{berander2005software}.
-Correctness and reliability are said to have dependencies, such that if a system exhibits a high degree of correctness then it tends to be reliable (\cite{GhezziEtAl2003}).
+
+There is no direct tool or method for measuring correctness. One way of building
+confidence in correctness is by reviewing to ensure that each requirement stated
+is one that the stakeholders and experts desire.  By maintaining traceability,
+consistency and unambiguity, we can reduce the occurrence of errors and make the
+goal of reviewing for correctness easier.
+
+The quality of a software's operation is dependent on the degree of correctness
+\cite{berander2005software}. Correctness and reliability are said to have
+dependencies, such that if a system exhibits a high degree of correctness then
+it tends to be reliable (\cite{GhezziEtAl2003}).
 
@@
 \begin{itemize}
-	\item{Definition 1:} A software is verifiable if its properties can be verified (\cite{GhezziEtAl2003}).\\
-	\item{Definition 2:} Attributes of software that relate to the effort needed for validating the modified software (\cite{berander2005software}).\\
-	\item{Definition 3:}) The degree to which a system or component facilitates the establishment of test criteria and the performance of tests to determine whether those criteria have been met (\cite{IEEEStdGlossarySET1990}).\\
-	\item{Definition 4:}  The degree to which a requirement is stated in terms that permit establishment of test criteria and performance (\cite{IEEEStdGlossarySET1990}).\\
-	\item{Definition 5:} Ease of performing testing on a software product or system (\cite{IEEEStdGlossarySET1990}).\\
-	\item{\textbf{Proposed definition}:} Combo of definitions 4 and 5 ?\\
+\item{Definition 1:} A software is verifiable if its properties can be verified
+  (\cite{GhezziEtAl2003}).
+\item{Definition 2:} Attributes of software that relate to the effort needed for
+  validating the modified software (\cite{berander2005software}).
+\item{Definition 3:}) The degree to which a system or component facilitates the
+  establishment of test criteria and the performance of tests to determine
+  whether those criteria have been met (\cite{IEEEStdGlossarySET1990}).
+\item{Definition 4:} The degree to which a requirement is stated in terms that
+  permit establishment of test criteria and performance
+  (\cite{IEEEStdGlossarySET1990}).
+\item{Definition 5:} Ease of performing testing on a software product or system
+  (\cite{IEEEStdGlossarySET1990}).
+\item{\textbf{Proposed definition}:} Combo of definitions 4 and 5 ?
 \end{itemize}
@@
 traceability, how the governing equations are transformed into code.
-\oo{Verifiability is sometimes referred to as testability, so I culled some testability definitions here}\\
+\oo{Verifiability is sometimes referred to as testability, so I culled some
+  testability definitions here}
+
 \subsection{Validatability}
+
 \begin{itemize}
-\item{Definition 1:} Validatability of a software is the degree of ease in validating(checking) that software meets user needs.\\
+\item{Definition 1:} Validatability of a software is the degree of ease in
+  validating(checking) that software meets user needs.
 \item{Definition 2:} Validatability means ``solving the right equations''~\cite[p.~23]{Roache1998}.
-\oo{question about this definition}\\
-\item{Definition 3:} The degree of ease of evaluating software products or system to determine whether it satisfies specified business requirements.
-\url{http://softwaretestingfundamentals.com/verification-vs-validation/}\\
-\oo{Here I tweaked the meaning of validation to suit validatability}\\
-\oo{Only a few resource on validatability}\\
+\oo{question about this definition}
+\item{Definition 3:} The degree of ease of evaluating software products or
+  system to determine whether it satisfies specified business
+  requirements. \url{http://softwaretestingfundamentals.com/verification-vs-validation/}
+  \oo{Here I tweaked the meaning of validation to suit validatability}
+  \oo{Only a few resource on validatability}
 \item{\textbf{Proposed definition}:} Definition 3
 \end{itemize}
+
 \subsubsection{Discussion}
+
 Validatability is improved by a rational process via clear documentation of the
@@
 while the act of verification itself improves reliability.\\
-Reliability models can be used to predict reliability of a software product. For example measuring Mean Time to Fail (MTTF) can be a good measure of reliability \cite{berander2005software}
+Reliability models can be used to predict reliability of a software product. For
+example measuring Mean Time to Fail (MTTF) can be a good measure of reliability
+\cite{berander2005software}
 
@@
 \begin{defn}
-The degree to which a system or component can function correctly in the presence of invalid inputs or stressful environmental conditions \cite{IEEEStdGlossarySET1990}.\\
+  The degree to which a system or component can function correctly in the
+  presence of invalid inputs or stressful environmental conditions
+  \cite{IEEEStdGlossarySET1990}.
 \end{defn}
 \begin{defn}
-The quality can be further informally refined as the ability of a software to keep an acceptable behavior, expressed in terms of robustness requirements, in spite of exceptional or unforeseen execution conditions (such as the unavailability of system resources, communication failures, invalid or stressful inputs, etc.) \cite{fernandez2005model}.\\
+  The quality can be further informally refined as the ability of a software to
+  keep an acceptable behavior, expressed in terms of robustness requirements, in
+  spite of exceptional or unforeseen execution conditions (such as the
+  unavailability of system resources, communication failures, invalid or
+  stressful inputs, etc.) \cite{fernandez2005model}.
 \end{defn}
 \begin{defn}
-Code possesses the characteristic of robustness to the extent that it can continue to perform despite some violation of the assumptions in its specification \cite{boehm2007software}.
+  Code possesses the characteristic of robustness to the extent that it can
+  continue to perform despite some violation of the assumptions in its
+  specification \cite{boehm2007software}.
 \end{defn}
 \begin{defn}
-A program is robust if it behaves "reasonably", even in circumstances that were not anticipated in the requirements specification - for example, when it encounters incorrect input data or some hardware malfunction \cite{ghezzi1991fundamentals}\\
+  A program is robust if it behaves "reasonably", even in circumstances that
+  were not anticipated in the requirements specification - for example, when it
+  encounters incorrect input data or some hardware malfunction
+  \cite{ghezzi1991fundamentals}
 \end{defn}

### 2019-10-25 StateOfPractice/QDefOfQualities/QDefOfQualities.tex

@@
 \author{Olu Owojaiye}{McMaster University, Canada}{owojaiyo@mcmaster.ca}{}{}
-\author{Peter Michalski}{McMaster University, Canada}{}{}{}
+\author{Peter Michalski}{McMaster University, Canada}{michap@mcmaster.ca}{}{}
 \author{Ao Dong}{McMaster University, Canada}{}{}{}
@@
   presence of invalid inputs or stressful environmental conditions
-  \cite{IEEEStdGlossarySET1990}.
+  \citep{IEEEStdGlossarySET1990}.
 \end{defn}
@@
   unavailability of system resources, communication failures, invalid or
-  stressful inputs, etc.) \cite{fernandez2005model}.
+  stressful inputs, etc.) \citep{fernandez2005model}.
 \end{defn}
@@
   continue to perform despite some violation of the assumptions in its
-  specification \cite{boehm2007software}.
+  specification \citep{boehm2007software}.
 \end{defn}
-\begin{defn}
+\begin{defn} \label{RobustnessDefnSelected}
   A program is robust if it behaves "reasonably", even in circumstances that
@@
   encounters incorrect input data or some hardware malfunction
-  \cite{ghezzi1991fundamentals}
+  \citep{ghezzi1991fundamentals}.
 \end{defn}
 
-\textbf{Chosen definition:}\\
-Combined Boehm and GhezziL\\
-Software possesses the characteristic of robustness if it fulfills its requirements correctly or behaves reasonably in circumstances not anticipated in the requirements specification.\\
+\noindent \textbf{Chosen definition:} Definition \ref{RobustnessDefnSelected} rephrased:
+
+Software possesses the characteristic of robustness if it fulfills its requirements correctly or behaves reasonably in circumstances not anticipated in the requirements specification.
+
+\noindent \textbf{Reasoning:}
+
+This definition indicates that robustness is related to the quality of correctness (both within and outside of it).
+
+\pmi{I think this is done.}
 
@@
 \begin{defn}
-The degree to which a system or component accomplishes its designated functions within given constraints, such as speed, accuracy, or memory usage \cite{IEEEStdGlossarySET1990}.\\
+The degree to which a system or component accomplishes its designated functions within given constraints, such as speed, accuracy, or memory usage \citep{IEEEStdGlossarySET1990}.
+\end{defn}
+\begin{defn} \label{PerformanceDefnSelected}
+How well or how rapidly the system must perform specific functions.
+Performance requirements encompass speed (database response times, for instance), throughput ( transactions per second), capacity (con-current usage loads), and timing (hard real-time demands) \citep{wiegers2003softreq}.
 \end{defn}
 \begin{defn}
-How well or how rapidly the system must perform specific functions.
-Performance requirements encompass speed (database response times, for instance), throughput ( transactions per second), capacity (con-current usage loads), and timing (hard real-time demands) \cite{wiegers2003softreq}.
-\end{defn}
-\begin{defn}
-In software engineering we often equate performance with efficiency. A software system is efficient if it uses computing resources economically \cite{ghezzi1991fundamentals}\\
+In software engineering we often equate performance with efficiency. A software system is efficient if it uses computing resources economically \citep{ghezzi1991fundamentals}.
 \end{defn}
 
-\textbf{Chosen definition:}\\
+\noindent \textbf{Chosen definition:} Definition \ref{PerformanceDefnSelected}:
+
 How well or how rapidly the system must perform specific functions.
-Performance requirements encompass speed (database response times, for instance), throughput ( transactions per second), capacity (con-current usage loads), and timing (hard real-time demands) \cite{wiegers2003softreq}.\\
+Performance requirements encompass speed (database response times, for instance), throughput ( transactions per second), capacity (con-current usage loads), and timing (hard real-time demands).
+
+\noindent \textbf{Reasoning:}
+
+This definition includes all important categories of performance.
+
+\pmi{I think this is done.}
 
@@
 \subsection{Maintainability}
-\begin{defn}
-The ease with which a software system or component can be modified to correct faults, improve performance or other attributes, or adapt to a changed environment \cite{IEEEStdGlossarySET1990}. 
+\begin{defn} \label{MaintainabilityDefnSelected1}
+The ease with which a software system or component can be modified to correct faults, improve performance or other attributes, or adapt to a changed environment \citep{IEEEStdGlossarySET1990}. 
 \end{defn}\
 \begin{defn}
-ISO/IEC 25010 refers to maintainability as the degree of effectiveness and efficiency with which a product or system can be modified by the intended maintainers \cite{ISO/IEC25010}.
+ISO/IEC 25010 refers to maintainability as the degree of effectiveness and efficiency with which a product or system can be modified by the intended maintainers \citep{ISO/IEC25010}.
 \end{defn}
@@
 Effort required to locate and fix an error in a program
-\cite{pressman2005software}.\\
+\citep{pressman2005software}.
 \end{defn}
 \begin{defn}
-A set of attributes that bear on the effort needed to make specified modifications (which may include corrections, improvements, or adaptations of software to environmental changes and changes in the requirements and functional specifications)\cite{pfleeger2006software}.
+A set of attributes that bear on the effort needed to make specified modifications (which may include corrections, improvements, or adaptations of software to environmental changes and changes in the requirements and functional specifications)\citep{pfleeger2006software}.
 \end{defn}
@@
 We will view maintainability as two separate qualities: repairability and evolvability.
-Software is repairable if it allows the fixing of defects; it is evolvable if it allows changes that enable it to satisfy new requirements \cite{ghezzi1991fundamentals}.
+Software is repairable if it allows the fixing of defects; it is evolvable if it allows changes that enable it to satisfy new requirements \citep{ghezzi1991fundamentals}.
+\end{defn}
+\begin{defn} \label{MaintainabilityDefnSelected2}
+Code possesses the characteristic of maintainability to the extent that it facilitates updating to satisfy new requirements or to correct deficiencies \citep{boehm2007software}.
+\end{defn}
+
+\noindent \textbf{Chosen definition:} Combined Definition \ref{MaintainabilityDefnSelected1} and Definition \ref{MaintainabilityDefnSelected2}:
+
+The ease with which a software system or component can be modified to correct faults, improve performance or other attributes, or satisfy new requirements.
+
+\noindent \textbf{Reasoning:}
+
+This definition includes all potential reasons to modify the software.
+
+\pmi{I think this is done.}
+
+\subsection{Reusability}
+\begin{defn} \label{ReusabilityDefnSelected}
+The degree to which a software module or other work product can be used in more than one software system \citep{IEEEStdGlossarySET1990}. 
 \end{defn}
 \begin{defn}
-Code possesses the characteristic of maintainability to the extent that it facilitates updating to satisfy new requirements or to correct deficiencies \cite{boehm2007software}.
+Extent to which a program [or parts of a program] can be reused in other applications - related to the packaging and scope of the functions that the program performs \citep{pressman2005software}.
 \end{defn}
 
-\textbf{Chosen definition:}\\
-Combined IEEE and Boehm:\\
-The ease with which a software system or component can be modified to correct faults, improve performance or other attributes, or satisfy new requirements.\\
+\noindent \textbf{Chosen definition:} Definition \ref{ReusabilityDefnSelected} rephrased:
 
-\subsection{Reusability}
-\begin{defn}
-The degree to which a software module or other work product can be used in more than one software system \cite{IEEEStdGlossarySET1990}. \\
-\end{defn}
-\begin{defn}
-Extent to which a program [or parts of a program] can be reused in other applications - related to the packaging and scope of the functions that the program performs \cite{pressman2005software}.\\
-\end{defn}
+The degree to which a software product can be modularized for easy transferal into other systems.
 
-\textbf{Chosen definition:}\\
-New definition:\\ The degree to which a software product can be modularized for easy transferal into other systems.\\
+\noindent \textbf{Reasoning:}
+
+This definition highlights transferal of parts of the system, or the whole system as a module into another system.
+Transferal of the whole system onto another environment would be covered in portability.
+
+\pmi{I am still not comfortable with this.}
 
@@
 \begin{defn}
-The ease with which a system or component can be transferred from one hardware or software environment to another \cite{IEEEStdGlossarySET1990}. \\
+The ease with which a system or component can be transferred from one hardware or software environment to another \citep{IEEEStdGlossarySET1990}.
 \end{defn}
 \begin{defn}
-An application is portable across a class of environments to the degree that the effort required to transport and adapt it to a new environment in the class is less than the effort of redevelopment \cite{mooney1990strategies}.\\
+An application is portable across a class of environments to the degree that the effort required to transport and adapt it to a new environment in the class is less than the effort of redevelopment \citep{mooney1990strategies}.
+\end{defn}
+\begin{defn} \label{PortabilityDefnSelected}
+Effort required to transfer the program from one hardware and/or software system environment to another \citep{pressman2005software}.
 \end{defn}
 \begin{defn}
-Effort required to transfer the program from one hardware and/or software system environment to another \cite{pressman2005software}.\\
-\end{defn}
-\begin{defn}
-A set of attributes that bear on the ability of software to be transferred from one environment to another (including the organizational, hardware, of software environment)\cite{pfleeger2006software}.\\
+A set of attributes that bear on the ability of software to be transferred from one environment to another (including the organizational, hardware, of software environment)\citep{pfleeger2006software}.
 \end{defn}
@@
 Code possesses the characteristic of portability to the extent that it can be operated easily and well on computer configurations other than its current one.
-This implies that special function features, not easily available at other facilities, are not used, that standard library functions and subroutines are selected for universal applicability, and so on \cite{boehm2007software}.
+This implies that special function features, not easily available at other facilities, are not used, that standard library functions and subroutines are selected for universal applicability, and so on \citep{boehm2007software}.
 \end{defn}
 \begin{defn}
-Portability refers to the ability to run a system on different hardware platforms \cite{ghezzi1991fundamentals}.
+Portability refers to the ability to run a system on different hardware platforms \citep{ghezzi1991fundamentals}.
 \end{defn}
-\textbf{Chosen definition:}\\ Effort required to transfer the program from one hardware and/or software system environment to another \cite{pressman2005software}.\\
-Rephrased: Effort required to transfer a program between system environments.\\
-Alternatively: Ease with which a program can be transferred between system environments.
-How is effort defined here? metric? measure?\\
+
+\noindent \textbf{Chosen definition:} Definition \ref{PortabilityDefnSelected} rephrased:
+
+Effort required to transfer a program between system environments.
+
+\noindent \textbf{Reasoning:}
+
+This is measurable and succinct.
+
+\pmi{I think this is done.}
 

### 2019-11-02 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
updated Interoperability,Visibility/Transparency,Productivity,Completness,Consistency and bib

@@
 \subsection{Interoperability}
+\noindent \textbf{Chosen definition:} Def \ref{InteroperabilitySelected} rephrased: Interoperability is the ability of different information systems, devices and applications (‘systems’) to access, exchange, integrate and cooperatively use data in a coordinated manner, within and across organizational, regional and national boundaries, to provide timely and seamless portability of information.
 
-\begin{enumerate}[(1).]
-    \item Interoperability is the ability of two or more systems or components to exchange information and to use the information that has been exchanged~\cite{IEEE1991}.
-	\item The degree to which two or more systems, products or components can exchange information and use the information that has been exchanged~\cite{ISO/IEC25010}.
-	\item The capability to communicate, execute programs, and transfer data among various functional units in a manner that requires the user to have little or no knowledge of the unique characteristics of those units~\cite{ISO/IEC/IEEE24765}.
-	\item Interoperability is a characteristic of a product or system, whose interfaces are completely understood, to work with other products or systems, present or future, in either implementation or access, without any restrictions~\cite{AFUL}.
-    \item \ad{My recommendation is the following one}\\
-    Interoperability is the ability of different information systems, devices and applications (‘systems’) to access, exchange, integrate and cooperatively use data in a coordinated manner, within and across organizational, regional and national boundaries, to provide timely and seamless portability of information~\cite{HIMSS}.
+\begin{defn}
+Interoperability is the ability of two or more systems or components to exchange information and to use the information that has been exchanged \citep{IEEEComputerDictionary1991}.
+\end{defn}
+\begin{defn}
+The degree to which two or more systems, products or components can exchange information and use the information that has been exchanged \citep{ISO/IEC25010}.
+\end{defn}
+\begin{defn}
+The capability to communicate, execute programs, and transfer data among various functional units in a manner that requires the user to have little or no knowledge of the unique characteristics of those units \citep{ISO/IEC/IEEE24765}.
+\end{defn}
+\begin{defn}Interoperability is a characteristic of a product or system, whose interfaces are completely understood, to work with other products or systems, present or future, in either implementation or access, without any restrictions \citep{AFUL2019}.
+\end{defn}
+\begin{defn}
+\label{InteroperabilitySelected}
+Interoperability is the ability of different information systems, devices and applications (‘systems’) to access, exchange, integrate and cooperatively use data in a coordinated manner, within and across organizational, regional and national boundaries, to provide timely and seamless portability of information and optimize the health of individuals and populations globally.
+Health data exchange architectures, application interfaces and standards enable data to be accessed and shared appropriately and securely across the complete spectrum of care, within all applicable settings and with relevant stakeholders, including by the individual \citep{HIMSS2019}.
 
@@
 
-        \item “New” Organizational (Level 4) – includes governance, policy, social, legal and organizational considerations to facilitate the secure, seamless and timely communication and use of data both within and between organizations, entities and individuals.
+\item Organizational (Level 4) – includes governance, policy, social, legal and organizational considerations to facilitate the secure, seamless and timely communication and use of data both within and between organizations, entities and individuals.
 These components enable shared consent, trust and integrated end-user processes and workflows
 \end{itemize}
-\end{enumerate}
+\end{defn}
 
 \subsection{Visibility/Transparency}
-    \ad{I found little contents related to visibility or transparency. The following definition is the only one I could find, but may be irrelevant to this project.}
-    \begin{quotation}
-        1. the degree to which a transaction can access object state concurrently with other transactions. 2. the specification, for a property, of "who can see it?"~\cite{ISO/IEC/IEEE24765}
-    \end{quotation}
+\noindent \textbf{Chosen definition:} Def \ref{VisibilitySelected}.
+
+\begin{defn}
+\label{VisibilitySelected}
+A software development process is visible if all of its steps and its current status are documented clearly.
+Another term used to characterize this property is transparency \citep{ghezzi1991fundamentals}.
+\end{defn}
 
@@
 \subsection{Productivity}
-\begin{enumerate}[(1).]
-    \item
-
+\noindent \textbf{Chosen definition:} The first sentence of Def \ref{ProductivitySelected}: Productivity is the amount of output (what is produced) per unit of input used.
+\begin{defn}
 The best definition of the productivity of a process is
-    \[\text{Productivity} = \dfrac{\text{Outputs produced by the
-          process}}{\text{Inputs consumed by the process}}\]
-    Defining inputs. For the software process, providing a meaningful definition
-    of inputs is a nontrivial but generally workable problem.
+\[\text{Productivity} = \dfrac{\text{Outputs produced by the process}}{\text{Inputs consumed by the process}}\]
+Defining inputs.
+For the software process, providing a meaningful definition of inputs is a nontrivial but generally workable problem.
 Inputs to the software process generally comprise labor, computers, supplies, and other support facilities and equipment.
 Defining outputs.
-The big problem in
-    defining software productivity is defining outputs. Here we find a
-    paradox. Most sources say that defining delivered source instructions (DSI)
-    or lines of code as the output of the software process is totally
-    inadequate, and they argue that there are a number of deficiencies in using
-    DSI. However, most organizations doing practical productivity measurement
-    still use DSI as their primary metric~\cite{Boehm1987}.
-
-    \item \ad{My recommendation is the following one}\\
+The big problem in defining software productivity is defining outputs.
+Here we find a defining delivered source instructions (DSI) or lines of code as the output of the software process is totally inadequate, and they argue that there are a number of deficiencies in using DSI.
+However, most organizations doing practical productivity measurement still use DSI as their primary metric \citep{Boehm1987}.
+\end{defn}
+\begin{defn}
+\label{ProductivitySelected}
 Productivity is the amount of output (what is produced) per unit of input used.If we can measure the size of the software product and the effort required to develop the product, we have:
@@
 \end{align}
-    Equation (1) assumes that size is the output of the software production
-    process and effort is the input to the process. This can be contrasted with
-    the viewpoint of software cost models where we use size as an independent
-    variable (i.e., an input) to predict effort which is treated as an
-    output. Equation (1) is simple to operationalize if we have a single
-    dominant size measure, for example, product size measured in lines of
-    code~\cite{Kitchenham2004}.
-\end{enumerate}
+Equation (1) assumes that size is the output of the software production process and effort is the input to the process.
+This can be contrasted with the viewpoint of software cost models where we use size as an independent variable (i.e., an input) to predict effort which is treated as an output.
+Equation (1) is simple to operationalize if we have a single dominant size measure, for example, product size measured in lines of code \citep{Kitchenham2004}.
+\end{defn}
 
@@
 \subsection{Completeness}
-\begin{enumerate}[(1).]
-    \item
-    \ad{This part was here before my modification}\\
-    Documentation is said to be complete when all the requirements of the software are detailed.
-That is, each goal, functionality, attribute, design constraint, value, data, model, symbol, term (with its unit of measurement if applicable), abbreviation, acronym, assumption and performance requirement of the software is defined.
-The software's response to all classes of inputs, both valid and invalid and for both desired and undesired events, also needs to be specified.
+\noindent \textbf{Chosen definition:} The first sentence of Def \ref{CompletenessSelected}: A specification is complete to the extent that all of its parts are present and each part is fully developed.
 
-    \item
-    \ad{My recommendation is the following one}\\
-    A specification is complete to the extent that all of its parts are present and each part is fully developed. A software specification must exhibit several properties to assure its completeness:
+\begin{defn}
+\label{CompletenessSelected}
+A specification is complete to the extent that all of its parts are present and each part is fully developed.
+A software specification must exhibit several properties to assure its completeness \citep{Boehm1984}:
 \begin{itemize}
@@
 \item No missing functions. These are functions that should be part of the software product but are not called for in the specification.
-        \item No missing products. These are products that should be part of the delivered software but are not called for in the specification.~\cite{Boehm1984}.
+\item No missing products. These are products that should be part of the delivered software but are not called for in the specification.
 \end{itemize}
-\end{enumerate}
+\end{defn}
+
+\ad{The following part was here before my modification}
+
+Documentation is said to be complete when all the requirements of the software are detailed.
+That is, each goal, functionality, attribute, design constraint, value, data, model, symbol, term (with its unit of measurement if applicable), abbreviation, acronym, assumption and performance requirement of the software is defined.
+The software's response to all classes of inputs, both valid and invalid and for both desired and undesired events, also needs to be specified.
 
 \subsection{Consistency}
-\begin{enumerate}[(1).]
-    \item
-    \ad{This part was here before my modification}\\
-    Documentation is said to be consistent when no subset of individual statements are in conflict with each other.
-That is, a specification of an item made at one place in the document should not contradict the specification of the same item at another location.
+\noindent \textbf{Chosen definition:} Def \ref{ConsistencySelected}.
 
-    \item    A specification is consistent to the extent that its provisions do not conflict with each other or with governing specifications and objectives.
-Specifications require consistency in several ways.
+\begin{defn}
+A specification is consistent to the extent that its provisions do not conflict with each other or with governing specifications and objectives.
+Specifications require consistency in several ways \citep{Boehm1984}.
 \begin{itemize}
@@
 \item External consistency. Items in the specification do not conflict with external specifications or entities.
-        \item Traceability. Items in the specification have clear antecedents in earlier specifications or statements of system objectives~\cite{Boehm1984}.
+\item Traceability. Items in the specification have clear antecedents in earlier specifications or statements of system objectives.
 \end{itemize}
-    \item Consistency requires that no two or more requirements in a specification contradict each other.
+\end{defn}
+\begin{defn}
+Consistency requires that no two or more requirements in a specification contradict each other.
 It is also often regarded as the case where words and terms have the same meaning throughout the requirements specifications (consistent use of terminology).
-These two views of consistency imply that mutually exclusive statements and clashes in terminology should be avoided~\cite{ZOWGHI2003}.
-    \item Consistency: 1.
+These two views of consistency imply that mutually exclusive statements and clashes in terminology should be avoided \citep{ZOWGHI2003}.
+\end{defn}
+\begin{defn}
+\label{ConsistencySelected}
+Consistency: 1.
 the degree of uniformity, standardization, and freedom from contradiction among the documents or parts of a system or component 2.
-software attributes that provide uniform design and implementation techniques and notations~\cite{ISO/IEC/IEEE24765}
-\end{enumerate}
+software attributes that provide uniform design and implementation techniques and notations \citep{ISO/IEC/IEEE24765}.
+\end{defn}
+
+\ad{The following part was here before my modification}
+
+Documentation is said to be consistent when no subset of individual statements are in conflict with each other.
+That is, a specification of an item made at one place in the document should not contradict the specification of the same item at another location.
 

### 2019-11-02 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
80 text width for Interoperability,Visibility/Transparency,Productivity,Completness,Consistency

@@
 \subsection{Interoperability}
-\noindent \textbf{Chosen definition:} Def \ref{InteroperabilitySelected} rephrased: Interoperability is the ability of different information systems, devices and applications (‘systems’) to access, exchange, integrate and cooperatively use data in a coordinated manner, within and across organizational, regional and national boundaries, to provide timely and seamless portability of information.
+\noindent \textbf{Chosen definition:} Def \ref{InteroperabilitySelected}
+rephrased: Interoperability is the ability of different information systems,
+devices and applications (‘systems’) to access, exchange, integrate and
+cooperatively use data in a coordinated manner, within and across
+organizational, regional and national boundaries, to provide timely and seamless
+portability of information.
 
 \begin{defn}
-Interoperability is the ability of two or more systems or components to exchange information and to use the information that has been exchanged \citep{IEEEComputerDictionary1991}.
+Interoperability is the ability of two or more systems or components to exchange
+information and to use the information that has been exchanged
+\citep{IEEEComputerDictionary1991}.
 \end{defn}
 \begin{defn}
-The degree to which two or more systems, products or components can exchange information and use the information that has been exchanged \citep{ISO/IEC25010}.
+The degree to which two or more systems, products or components can exchange
+information and use the information that has been exchanged
+\citep{ISO/IEC25010}.
 \end{defn}
 \begin{defn}
-The capability to communicate, execute programs, and transfer data among various functional units in a manner that requires the user to have little or no knowledge of the unique characteristics of those units \citep{ISO/IEC/IEEE24765}.
+The capability to communicate, execute programs, and transfer data among various
+functional units in a manner that requires the user to have little or no
+knowledge of the unique characteristics of those units
+\citep{ISO/IEC/IEEE24765}.
 \end{defn}
-\begin{defn}Interoperability is a characteristic of a product or system, whose interfaces are completely understood, to work with other products or systems, present or future, in either implementation or access, without any restrictions \citep{AFUL2019}.
+\begin{defn}Interoperability is a characteristic of a product or system, whose
+interfaces are completely understood, to work with other products or systems,
+present or future, in either implementation or access, without any restrictions
+\citep{AFUL2019}.
 \end{defn}
@@
 \label{InteroperabilitySelected}
-Interoperability is the ability of different information systems, devices and applications (‘systems’) to access, exchange, integrate and cooperatively use data in a coordinated manner, within and across organizational, regional and national boundaries, to provide timely and seamless portability of information and optimize the health of individuals and populations globally.
-Health data exchange architectures, application interfaces and standards enable data to be accessed and shared appropriately and securely across the complete spectrum of care, within all applicable settings and with relevant stakeholders, including by the individual \citep{HIMSS2019}.
+Interoperability is the ability of different information systems, devices and
+applications (‘systems’) to access, exchange, integrate and cooperatively use
+data in a coordinated manner, within and across organizational, regional and
+national boundaries, to provide timely and seamless portability of information
+and optimize the health of individuals and populations globally. Health data
+exchange architectures, application interfaces and standards enable data to be
+accessed and shared appropriately and securely across the complete spectrum of
+care, within all applicable settings and with relevant stakeholders, including
+by the individual \citep{HIMSS2019}.
 
@@
 \begin{itemize}
-\item Foundational (Level 1) – establishes the inter-connectivity requirements needed for one system or application to securely communicate data to and receive data from another
+\item Foundational (Level 1) – establishes the inter-connectivity requirements
+needed for one system or application to securely communicate data to and receive
+data from another
 
-\item Structural (Level 2) – defines the format, syntax, and organization of data exchange including at the data field level for interpretation
+\item Structural (Level 2) – defines the format, syntax, and organization of
+data exchange including at the data field level for interpretation
 
-\item Semantic (Level 3) – provides for common underlying models and codification of the data including the use of data elements with standardized definitions from publicly available value sets and coding vocabularies, providing shared understanding and meaning to the user
+\item Semantic (Level 3) – provides for common underlying models and
+codification of the data including the use of data elements with standardized
+definitions from publicly available value sets and coding vocabularies,
+providing shared understanding and meaning to the user
 
-\item Organizational (Level 4) – includes governance, policy, social, legal and organizational considerations to facilitate the secure, seamless and timely communication and use of data both within and between organizations, entities and individuals.
-These components enable shared consent, trust and integrated end-user processes and workflows
+\item Organizational (Level 4) – includes governance, policy, social, legal and
+organizational considerations to facilitate the secure, seamless and timely
+communication and use of data both within and between organizations, entities
+and individuals. These components enable shared consent, trust and integrated
+end-user processes and workflows
 \end{itemize}
@@
 \label{VisibilitySelected}
-A software development process is visible if all of its steps and its current status are documented clearly.
-Another term used to characterize this property is transparency \citep{ghezzi1991fundamentals}.
+A software development process is visible if all of its steps and its current
+status are documented clearly. Another term used to characterize this property
+is transparency \citep{ghezzi1991fundamentals}.
 \end{defn}
@@
 \subsection{Productivity}
-\noindent \textbf{Chosen definition:} The first sentence of Def \ref{ProductivitySelected}: Productivity is the amount of output (what is produced) per unit of input used.
+\noindent \textbf{Chosen definition:} The first sentence of Def
+\ref{ProductivitySelected}: Productivity is the amount of output (what is
+produced) per unit of input used.
 \begin{defn}
 The best definition of the productivity of a process is
-\[\text{Productivity} = \dfrac{\text{Outputs produced by the process}}{\text{Inputs consumed by the process}}\]
-Defining inputs.
-For the software process, providing a meaningful definition of inputs is a nontrivial but generally workable problem.
-Inputs to the software process generally comprise labor, computers, supplies, and other support facilities and equipment.
-Defining outputs.
-The big problem in defining software productivity is defining outputs.
-Here we find a defining delivered source instructions (DSI) or lines of code as the output of the software process is totally inadequate, and they argue that there are a number of deficiencies in using DSI.
-However, most organizations doing practical productivity measurement still use DSI as their primary metric \citep{Boehm1987}.
+\[\text{Productivity} = \dfrac{\text{Outputs produced by the
+process}}{\text{Inputs consumed by the process}}\]
+Defining inputs. For the software process, providing a meaningful definition of
+inputs is a nontrivial but generally workable problem. Inputs to the software
+process generally comprise labor, computers, supplies, and other support
+facilities and equipment. Defining outputs. The big problem in defining software
+productivity is defining outputs. Here we find a defining delivered source
+instructions (DSI) or lines of code as the output of the software process is
+totally inadequate, and they argue that there are a number of deficiencies in
+using DSI. However, most organizations doing practical productivity measurement
+still use DSI as their primary metric \citep{Boehm1987}.
 \end{defn}
@@
 \label{ProductivitySelected}
-Productivity is the amount of output (what is produced) per unit of input used.If we can measure the size of the software product and the effort required to develop the product, we have:
+Productivity is the amount of output (what is produced) per unit of input
+used.If we can measure the size of the software product and the effort required
+to develop the product, we have:
 \begin{align}
@@
 \end{align}
-Equation (1) assumes that size is the output of the software production process and effort is the input to the process.
-This can be contrasted with the viewpoint of software cost models where we use size as an independent variable (i.e., an input) to predict effort which is treated as an output.
-Equation (1) is simple to operationalize if we have a single dominant size measure, for example, product size measured in lines of code \citep{Kitchenham2004}.
+Equation (1) assumes that size is the output of the software production process
+and effort is the input to the process. This can be contrasted with the
+viewpoint of software cost models where we use size as an independent variable
+(i.e., an input) to predict effort which is treated as an output. Equation (1)
+is simple to operationalize if we have a single dominant size measure, for
+example, product size measured in lines of code \citep{Kitchenham2004}.
 \end{defn}
@@
 \subsection{Completeness}
-\noindent \textbf{Chosen definition:} The first sentence of Def \ref{CompletenessSelected}: A specification is complete to the extent that all of its parts are present and each part is fully developed.
+\noindent \textbf{Chosen definition:} The first sentence of Def
+\ref{CompletenessSelected}: A specification is complete to the extent that all
+of its parts are present and each part is fully developed.
 
@@
 \label{CompletenessSelected}
-A specification is complete to the extent that all of its parts are present and each part is fully developed.
-A software specification must exhibit several properties to assure its completeness \citep{Boehm1984}:
+A specification is complete to the extent that all of its parts are present and
+each part is fully developed. A software specification must exhibit several
+properties to assure its completeness \citep{Boehm1984}:
 \begin{itemize}
-\item No TBDs. TBDs are places in the specification where decisions have been postponed by writing "To be Determined" or "TBD."
-\item No nonexistent references. These are references in the specification to functions, inputs, or outputs (including databases) not defined in the specification.
-\item No missing specification items. These are items that should be present as part of the standard format of the specification, but are not present.
-\item No missing functions. These are functions that should be part of the software product but are not called for in the specification.
-\item No missing products. These are products that should be part of the delivered software but are not called for in the specification.
+\item No TBDs. TBDs are places in the specification where decisions have been
+postponed by writing "To be Determined" or "TBD."
+\item No nonexistent references. These are references in the specification to
+functions, inputs, or outputs (including databases) not defined in the
+specification.
+\item No missing specification items. These are items that should be present as
+part of the standard format of the specification, but are not present.
+\item No missing functions. These are functions that should be part of the
+software product but are not called for in the specification.
+\item No missing products. These are products that should be part of the
+delivered software but are not called for in the specification.
 \end{itemize}
@@
 
-Documentation is said to be complete when all the requirements of the software are detailed.
-That is, each goal, functionality, attribute, design constraint, value, data, model, symbol, term (with its unit of measurement if applicable), abbreviation, acronym, assumption and performance requirement of the software is defined.
-The software's response to all classes of inputs, both valid and invalid and for both desired and undesired events, also needs to be specified.
+Documentation is said to be complete when all the requirements of the software
+are detailed. That is, each goal, functionality, attribute, design constraint,
+value, data, model, symbol, term (with its unit of measurement if applicable),
+abbreviation, acronym, assumption and performance requirement of the software is
+defined. The software's response to all classes of inputs, both valid and
+invalid and for both desired and undesired events, also needs to be specified.
 
@@
 \begin{defn}
-A specification is consistent to the extent that its provisions do not conflict with each other or with governing specifications and objectives.
-Specifications require consistency in several ways \citep{Boehm1984}.
+A specification is consistent to the extent that its provisions do not conflict
+with each other or with governing specifications and objectives. Specifications
+require consistency in several ways \citep{Boehm1984}.
 \begin{itemize}
-\item Internal consistency. Items within the specification do not conflict with each other.
-\item External consistency. Items in the specification do not conflict with external specifications or entities.
-\item Traceability. Items in the specification have clear antecedents in earlier specifications or statements of system objectives.
+\item Internal consistency. Items within the specification do not conflict with
+each other.
+\item External consistency. Items in the specification do not conflict with
+external specifications or entities.
+\item Traceability. Items in the specification have clear antecedents in earlier
+specifications or statements of system objectives.
 \end{itemize}
@@
 \begin{defn}
-Consistency requires that no two or more requirements in a specification contradict each other.
-It is also often regarded as the case where words and terms have the same meaning throughout the requirements specifications (consistent use of terminology).
-These two views of consistency imply that mutually exclusive statements and clashes in terminology should be avoided \citep{ZOWGHI2003}.
+Consistency requires that no two or more requirements in a specification
+contradict each other. It is also often regarded as the case where words and
+terms have the same meaning throughout the requirements specifications
+(consistent use of terminology). These two views of consistency imply that
+mutually exclusive statements and clashes in terminology should be avoided
+\citep{ZOWGHI2003}.
 \end{defn}
@@
 \label{ConsistencySelected}
-Consistency: 1.
-the degree of uniformity, standardization, and freedom from contradiction among the documents or parts of a system or component 2.
-software attributes that provide uniform design and implementation techniques and notations \citep{ISO/IEC/IEEE24765}.
+Consistency: 1. the degree of uniformity, standardization, and freedom from
+contradiction among the documents or parts of a system or component 2. software
+attributes that provide uniform design and implementation techniques and
+notations \citep{ISO/IEC/IEEE24765}.
 \end{defn}
@@
 
-Documentation is said to be consistent when no subset of individual statements are in conflict with each other.
-That is, a specification of an item made at one place in the document should not contradict the specification of the same item at another location.
+Documentation is said to be consistent when no subset of individual statements
+are in conflict with each other. That is, a specification of an item made at one
+place in the document should not contradict the specification of the same item
+at another location.
 

### 2019-11-05 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
completed defn reusability

@@
 \subsection{Reusability}
-\begin{defn} \label{ReusabilityDefnSelected}
+\begin{defn} 
 The degree to which a software module or other work product can be used in more than one software system \citep{IEEEStdGlossarySET1990}. 
@@
 \end{defn}
+\begin{defn} \label{ReusabilityDefnSelected}
+The extent to which a software component can be used with or without adaptation in a problem solution other than the one for which it was originally developed \citep{kalagiakos2003non}.
+\end{defn}
+\begin{defn}
+Reusability is the likelihood a segment of source code that can be used again to add new functionalities with slight or no modification \citep{sandhu2010survey}.
+\end{defn}
 
-\noindent \textbf{Chosen definition:} Definition \ref{ReusabilityDefnSelected} rephrased:
+\noindent \textbf{Chosen definition:} Definition \ref{ReusabilityDefnSelected}:
 
-The degree to which a software product can be modularized for easy transferal into other systems.
+The extent to which a software component can be used with or without adaptation in a problem solution other than the one for which it was originally developed.
 
@@
 
-This definition highlights transferal of parts of the system, or the whole system as a module into another system.
-Transferal of the whole system onto another environment would be covered in portability.
+This definition highlights the possible but not necessary adaptation of the software component(s) being transferred.
 
-\pmi{I am still not comfortable with this.}
+\pmi{I think this is done.}
 

### 2019-11-06 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
Reformat tex file for 80 character width

@@
 
-\authorrunning{Smith et al.}
-\Copyright{Spencer Smith and Jacques Carette and Olu Owojaiye and Peter Michalski and Ao Dong}
+\authorrunning{Smith et al.}  \Copyright{Spencer Smith and Jacques Carette and
+  Olu Owojaiye and Peter Michalski and Ao Dong}
 
@@
 \begin{itemize}
-\item{Definition 1:} A measure of the ease of installation.\\
-\item{Definition 2:} Installability is related to the amount of human effort required to install a software in a designated environment \cite{berander2005software}. \\
-\item{Definition 3:} ``The capability of a software product to be installed in a specified environment'' \cite{berander2005software}.\\
-\item{Definition 4:}  ``The efficacy of the installation, uninstallation or reinstallation process of a software product in a specified enviroment'' (\cite{iso201017043})\\
+\item{Definition 1:} A measure of the ease of installation.
+\item{Definition 2:} Installability is related to the amount of human effort
+  required to install a software in a designated environment
+  \cite{berander2005software}.
+\item{Definition 3:} ``The capability of a software product to be installed in a
+  specified environment'' \cite{berander2005software}.
+\item{Definition 4:} ``The efficacy of the installation, uninstallation or
+  reinstallation process of a software product in a specified enviroment''
+  (\cite{iso201017043})
 \item {\textbf{Proposed definition}:} (Definition 2)
@@
 \subsection{Verifiability}
+
 \begin{itemize}
@@
 \end{itemize}
+
 \subsubsection{Discussion}
+
 Verifiability involves ``solving the equations right''~\cite[p.~23]{Roache1998};
@@
   validating(checking) that software meets user needs.
-\item{Definition 2:} Validatability means ``solving the right equations''~\cite[p.~23]{Roache1998}.
+\item{Definition 2:} Validatability means ``solving the right
+  equations''~\cite[p.~23]{Roache1998}.
 \oo{question about this definition}
@@
 \subsection{Reliability}
+
 \begin{itemize}
-\item{Definition 1:} The probability that the software will operate as expected over a specified time interval (\cite{GhezziEtAl2003}).\\
-\item{Definition 2:} A set of attributes that relate to the capability of software to maintain its level of performance under stated conditions for a stated period of time (\cite{berander2005software}).\\
-\item{Definition 3:} “The capability of the software product to maintain a specified level of performance when used under specified conditions (\cite{international2001iso}).\\
-\item{Definition 4:} Code possesses the characteristic reliability to the extent that it can be expected to perform its intended functions satisfactorily (\cite{boehm2007software})\\
-\item{Definition 5:} A concern encompassing correctness and robustness (\cite{meyer1988object}).\\
-\item{Proposed definition}\\
+\item{Definition 1:} The probability that the software will operate as expected
+  over a specified time interval (\cite{GhezziEtAl2003}).
+\item{Definition 2:} A set of attributes that relate to the capability of
+  software to maintain its level of performance under stated conditions for a
+  stated period of time (\cite{berander2005software}).
+\item{Definition 3:} “The capability of the software product to maintain a
+  specified level of performance when used under specified conditions
+  (\cite{international2001iso}).
+\item{Definition 4:} Code possesses the characteristic reliability to the extent
+  that it can be expected to perform its intended functions satisfactorily
+  (\cite{boehm2007software})
+\item{Definition 5:} A concern encompassing correctness and robustness
+  (\cite{meyer1988object}).
+\item{Proposed definition}
 \end{itemize}
@@
 \subsubsection{Discussion}
+
 Reliability is a critical quality for scientific software, since the results of
@@
 closely tied to verifiability, since the key quality to verify is reliability,
-while the act of verification itself improves reliability.\\
+while the act of verification itself improves reliability.
+
 Reliability models can be used to predict reliability of a software product. For
 example measuring Mean Time to Fail (MTTF) can be a good measure of reliability
-\cite{berander2005software}
+\cite{berander2005software}.
 
@@
 
-Software possesses the characteristic of robustness if it fulfills its requirements correctly or behaves reasonably in circumstances not anticipated in the requirements specification.
+Software possesses the characteristic of robustness if it fulfills its
+requirements correctly or behaves reasonably in circumstances not anticipated in
+the requirements specification.
 
@@
 
-This definition indicates that robustness is related to the quality of correctness (both within and outside of it).
+This definition indicates that robustness is related to the quality of
+correctness (both within and outside of it).
 
@@
 \begin{defn}
-The degree to which a system or component accomplishes its designated functions within given constraints, such as speed, accuracy, or memory usage \citep{IEEEStdGlossarySET1990}.
+The degree to which a system or component accomplishes its designated functions
+within given constraints, such as speed, accuracy, or memory usage
+\citep{IEEEStdGlossarySET1990}.
 \end{defn}
 \begin{defn} \label{PerformanceDefnSelected}
-How well or how rapidly the system must perform specific functions.
-Performance requirements encompass speed (database response times, for instance), throughput ( transactions per second), capacity (con-current usage loads), and timing (hard real-time demands) \citep{wiegers2003softreq}.
+How well or how rapidly the system must perform specific functions. Performance
+requirements encompass speed (database response times, for instance), throughput
+( transactions per second), capacity (con-current usage loads), and timing (hard
+real-time demands) \citep{wiegers2003softreq}.
 \end{defn}
 \begin{defn}
-In software engineering we often equate performance with efficiency. A software system is efficient if it uses computing resources economically \citep{ghezzi1991fundamentals}.
+In software engineering we often equate performance with efficiency. A software
+system is efficient if it uses computing resources economically
+\citep{ghezzi1991fundamentals}.
 \end{defn}
@@
 
-How well or how rapidly the system must perform specific functions.
-Performance requirements encompass speed (database response times, for instance), throughput ( transactions per second), capacity (con-current usage loads), and timing (hard real-time demands).
+How well or how rapidly the system must perform specific functions. Performance
+requirements encompass speed (database response times, for instance), throughput
+( transactions per second), capacity (con-current usage loads), and timing (hard
+real-time demands).
 
@@
 Nielsen and (separately) Schneidermann have defined usability as part of usefulness and
-is composed of
+is composed of:
 \begin{itemize}
@@
 further lists the following desirable outcomes:
+
 \begin{enumerate}
@@
 \subsection{Maintainability}
+
 \begin{defn} \label{MaintainabilityDefnSelected1}
-The ease with which a software system or component can be modified to correct faults, improve performance or other attributes, or adapt to a changed environment \citep{IEEEStdGlossarySET1990}. 
+The ease with which a software system or component can be modified to correct
+faults, improve performance or other attributes, or adapt to a changed
+environment \citep{IEEEStdGlossarySET1990}. 
 \end{defn}\
 \begin{defn}
-ISO/IEC 25010 refers to maintainability as the degree of effectiveness and efficiency with which a product or system can be modified by the intended maintainers \citep{ISO/IEC25010}.
+ISO/IEC 25010 refers to maintainability as the degree of effectiveness and
+efficiency with which a product or system can be modified by the intended
+maintainers \citep{ISO/IEC25010}.
 \end{defn}
@@
 \begin{defn}
-A set of attributes that bear on the effort needed to make specified modifications (which may include corrections, improvements, or adaptations of software to environmental changes and changes in the requirements and functional specifications)\citep{pfleeger2006software}.
+A set of attributes that bear on the effort needed to make specified
+modifications (which may include corrections, improvements, or adaptations of
+software to environmental changes and changes in the requirements and functional
+specifications)\citep{pfleeger2006software}.
 \end{defn}
 \begin{defn}
-We will view maintainability as two separate qualities: repairability and evolvability.
-Software is repairable if it allows the fixing of defects; it is evolvable if it allows changes that enable it to satisfy new requirements \citep{ghezzi1991fundamentals}.
+We will view maintainability as two separate qualities: repairability and
+evolvability. Software is repairable if it allows the fixing of defects; it is
+evolvable if it allows changes that enable it to satisfy new requirements
+\citep{ghezzi1991fundamentals}.
 \end{defn}
 \begin{defn} \label{MaintainabilityDefnSelected2}
-Code possesses the characteristic of maintainability to the extent that it facilitates updating to satisfy new requirements or to correct deficiencies \citep{boehm2007software}.
+Code possesses the characteristic of maintainability to the extent that it
+facilitates updating to satisfy new requirements or to correct deficiencies
+\citep{boehm2007software}.
 \end{defn}
 
-\noindent \textbf{Chosen definition:} Combined Definition \ref{MaintainabilityDefnSelected1} and Definition \ref{MaintainabilityDefnSelected2}:
+\noindent \textbf{Chosen definition:} Combined Definition
+\ref{MaintainabilityDefnSelected1} and Definition
+\ref{MaintainabilityDefnSelected2}:
 
-The ease with which a software system or component can be modified to correct faults, improve performance or other attributes, or satisfy new requirements.
+The ease with which a software system or component can be modified to correct
+faults, improve performance or other attributes, or satisfy new requirements.
 
@@
 \begin{defn} 
-The degree to which a software module or other work product can be used in more than one software system \citep{IEEEStdGlossarySET1990}. 
+The degree to which a software module or other work product can be used in more
+than one software system \citep{IEEEStdGlossarySET1990}. 
 \end{defn}
 \begin{defn}
-Extent to which a program [or parts of a program] can be reused in other applications - related to the packaging and scope of the functions that the program performs \citep{pressman2005software}.
+Extent to which a program [or parts of a program] can be reused in other
+applications - related to the packaging and scope of the functions that the
+program performs \citep{pressman2005software}.
 \end{defn}
 \begin{defn} \label{ReusabilityDefnSelected}
-The extent to which a software component can be used with or without adaptation in a problem solution other than the one for which it was originally developed \citep{kalagiakos2003non}.
+The extent to which a software component can be used with or without adaptation
+in a problem solution other than the one for which it was originally developed
+\citep{kalagiakos2003non}.
 \end{defn}
 \begin{defn}
-Reusability is the likelihood a segment of source code that can be used again to add new functionalities with slight or no modification \citep{sandhu2010survey}.
+Reusability is the likelihood a segment of source code that can be used again to
+add new functionalities with slight or no modification \citep{sandhu2010survey}.
 \end{defn}
@@
 
-The extent to which a software component can be used with or without adaptation in a problem solution other than the one for which it was originally developed.
+The extent to which a software component can be used with or without adaptation
+in a problem solution other than the one for which it was originally developed.
 
@@
 
-This definition highlights the possible but not necessary adaptation of the software component(s) being transferred.
+This definition highlights the possible but not necessary adaptation of the
+software component(s) being transferred.
 
@@
 \begin{defn}
-The ease with which a system or component can be transferred from one hardware or software environment to another \citep{IEEEStdGlossarySET1990}.
+The ease with which a system or component can be transferred from one hardware
+or software environment to another \citep{IEEEStdGlossarySET1990}.
 \end{defn}
 \begin{defn}
-An application is portable across a class of environments to the degree that the effort required to transport and adapt it to a new environment in the class is less than the effort of redevelopment \citep{mooney1990strategies}.
+An application is portable across a class of environments to the degree that the
+effort required to transport and adapt it to a new environment in the class is
+less than the effort of redevelopment \citep{mooney1990strategies}.
 \end{defn}
 \begin{defn} \label{PortabilityDefnSelected}
-Effort required to transfer the program from one hardware and/or software system environment to another \citep{pressman2005software}.
+Effort required to transfer the program from one hardware and/or software system
+environment to another \citep{pressman2005software}.
 \end{defn}
 \begin{defn}
-A set of attributes that bear on the ability of software to be transferred from one environment to another (including the organizational, hardware, of software environment)\citep{pfleeger2006software}.
+A set of attributes that bear on the ability of software to be transferred from
+one environment to another (including the organizational, hardware, of software
+environment)\citep{pfleeger2006software}.
 \end{defn}
 \begin{defn}
-Code possesses the characteristic of portability to the extent that it can be operated easily and well on computer configurations other than its current one.
-This implies that special function features, not easily available at other facilities, are not used, that standard library functions and subroutines are selected for universal applicability, and so on \citep{boehm2007software}.
+Code possesses the characteristic of portability to the extent that it can be
+operated easily and well on computer configurations other than its current
+one. This implies that special function features, not easily available at other
+facilities, are not used, that standard library functions and subroutines are
+selected for universal applicability, and so on \citep{boehm2007software}.
 \end{defn}
 \begin{defn}
-Portability refers to the ability to run a system on different hardware platforms \citep{ghezzi1991fundamentals}.
+Portability refers to the ability to run a system on different hardware
+platforms \citep{ghezzi1991fundamentals}.
 \end{defn}
 
-\noindent \textbf{Chosen definition:} Definition \ref{PortabilityDefnSelected} rephrased:
+\noindent \textbf{Chosen definition:} Definition \ref{PortabilityDefnSelected}
+rephrased:
 
@@
 \subsection{Interoperability}
+
 \noindent \textbf{Chosen definition:} Def \ref{InteroperabilitySelected}
@@
 \subsection{Productivity}
+
 \noindent \textbf{Chosen definition:} The first sentence of Def
@@
 \subsection{Completeness}
+
 \noindent \textbf{Chosen definition:} The first sentence of Def
@@
 \subsection{Consistency}
+
 \noindent \textbf{Chosen definition:} Def \ref{ConsistencySelected}.
@@
 
-Here the Wikipedia page~\url{https://en.wikipedia.org/wiki/Traceability} is actually
-rather informative, especially as it also lists how this concept is used in other domains.
-A generic definition that is still quite useful is
+Here the Wikipedia page~\url{https://en.wikipedia.org/wiki/Traceability} is
+actually rather informative, especially as it also lists how this concept is
+used in other domains.  A generic definition that is still quite useful is
 \begin{defn}

### 2019-11-06 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
Addition of owner information to each quality
Added explicit Proposed Definition to each quality.

@@
 Purpose and scope of the document.  \wss{Needs to be filled in.  Should
-  reference the overal research proposal, and the ``state of the practice''
+  reference the overall research proposal, and the ``state of the practice''
   exercise in particular.}
@@
 
-\subsection{Installability}
-\begin{itemize}
-\item{Definition 1:} A measure of the ease of installation.
-\item{Definition 2:} Installability is related to the amount of human effort
-  required to install a software in a designated environment
-  \cite{berander2005software}.
-\item{Definition 3:} ``The capability of a software product to be installed in a
-  specified environment'' \cite{berander2005software}.
-\item{Definition 4:} ``The efficacy of the installation, uninstallation or
-  reinstallation process of a software product in a specified enviroment''
-  (\cite{iso201017043})
-\item {\textbf{Proposed definition}:} (Definition 2)
-\end{itemize}
+\subsection{Installability \oo{owner}}
 
-\subsection{Correctness}
+\begin{defn}
+  A measure of the ease of installation.  \wss{Where does this definition come
+    from? Please add a citation.}
+\end{defn}
+
+\begin{defn} \label{DefnInstallability}
+  Installability is related to the amount of human effort required to install a
+  software in a designated environment \cite{berander2005software}.
+\end{defn}
+
+\begin{defn}
+  ``The capability of a software product to be installed in a specified
+  environment'' \cite{berander2005software}.
+\end{defn}
+
+\begin{defn}
+  ``The efficacy of the installation, uninstallation or reinstallation process
+  of a software product in a specified environment'' (\cite{iso201017043})
+\end{defn}
+
+\noindent \textbf{Proposed Definition}
+
+Definition~\ref{DefnInstallability}.
+
+\wss{The rationale is missing.  Please add.}
+
+\subsection{Correctness \oo{owner}}
 
@@
 
-\noindent \textbf{Proposed definition}: Definition~\ref{CorrectDefnSelected}
+\noindent \textbf{Proposed Definition}
 
-\subsubsection{Discussion}
+Definition~\ref{CorrectDefnSelected}
+
+\noindent \textbf{Reasoning}
 
@@
 
-\subsection{Verifiability}
+\subsection{Verifiability \oo{owner}}
 
@@
 
+\noindent \textbf{Proposed Definition}
+
 \subsubsection{Discussion}
@@
 
-\subsection{Validatability}
+\subsection{Validatability \oo{owner}}
 
@@
 
-\subsection{Reliability}
+\subsection{Reliability \oo{owner}}
 
@@
 
-\subsection{Robustness}
+\subsection{Robustness \pmi{owner}}
 \begin{defn}
@@
 
-\noindent \textbf{Chosen definition:} Definition \ref{RobustnessDefnSelected} rephrased:
+\noindent \textbf{Proposed Definition}
+
+Definition \ref{RobustnessDefnSelected} rephrased:
 
@@
 
-\subsection{Performance}
+\subsection{Performance \pmi{owner}}
+
 \begin{defn}
@@
 
-\noindent \textbf{Chosen definition:} Definition \ref{PerformanceDefnSelected}:
+\noindent \textbf{Proposed Definition}
+
+Definition \ref{PerformanceDefnSelected}:
 
@@
 
-\subsection{Usability}
+\subsection{Usability \jc{owner}}
 
@@
 
-\subsection{Maintainability}
+\noindent \textbf{Proposed Definition}
+
+\wss{Still needs to be completed}
+
+\subsection{Maintainability \pmi{owner}}
 
@@
 
-\noindent \textbf{Chosen definition:} Combined Definition
+\noindent \textbf{Proposed Definition}
+
+Combined Definition
 \ref{MaintainabilityDefnSelected1} and Definition
@@
 
-\subsection{Reusability}
+\subsection{Reusability \pmi{owner}}
+
 \begin{defn} 
@@
 
-\noindent \textbf{Chosen definition:} Definition \ref{ReusabilityDefnSelected}:
+\noindent \textbf{Proposed Definition}
+
+Definition \ref{ReusabilityDefnSelected}:
 
@@
 
-\subsection{Portability}
+\subsection{Portability \pmi{owner}}
 \begin{defn}
@@
 
-\noindent \textbf{Chosen definition:} Definition \ref{PortabilityDefnSelected}
-rephrased:
+\noindent \textbf{Proposed Definition}
+
+Definition \ref{PortabilityDefnSelected} rephrased:
 
@@
 
-\subsection{Understandability}
+\subsection{Understandability \jc{owner}}
 
@@
 
-\subsection{Interoperability}
+\noindent \textbf{Proposed Definition}
 
-\noindent \textbf{Chosen definition:} Def \ref{InteroperabilitySelected}
-rephrased: Interoperability is the ability of different information systems,
-devices and applications (‘systems’) to access, exchange, integrate and
-cooperatively use data in a coordinated manner, within and across
-organizational, regional and national boundaries, to provide timely and seamless
-portability of information.
+\wss{Still needs to be completed}
+
+\subsection{Interoperability \ad{owner}}
 
@@
 
-\subsection{Visibility/Transparency}
-\noindent \textbf{Chosen definition:} Def \ref{VisibilitySelected}.
+\noindent \textbf{Proposed Definition}
+
+Definition \ref{InteroperabilitySelected}
+rephrased: Interoperability is the ability of different information systems,
+devices and applications (‘systems’) to access, exchange, integrate and
+cooperatively use data in a coordinated manner, within and across
+organizational, regional and national boundaries, to provide timely and seamless
+portability of information.
+
+\wss{I moved the proposed definition to the end, to match the other sections.
+  Hopefully I did not change the meaning.}
+
+\subsection{Visibility/Transparency \ad{owner}}
 
@@
 
-\subsection{Reproducibility}
+\noindent \textbf{Proposed definition}
+
+Definition \ref{VisibilitySelected}.
+
+\subsection{Reproducibility \wss{owner}}
 
@@
 
-\subsection{Productivity}
+\noindent \textbf{Proposed definition}
 
-\noindent \textbf{Chosen definition:} The first sentence of Def
-\ref{ProductivitySelected}: Productivity is the amount of output (what is
-produced) per unit of input used.
+\wss{Needs to be completed}
+
+\subsection{Productivity \ad{owner}}
+
 \begin{defn}
@@
 
-\subsection{Sustainability}
+\noindent \textbf{Proposed definition}
+
+The first sentence of Definition \ref{ProductivitySelected}: Productivity is the
+amount of output (what is produced) per unit of input used.
+
+\subsection{Sustainability \wss{owner}}
 
@@
 
+\noindent \textbf{Proposed definition}
+
+\wss{Needs to be completed}
+
 \section{Desirable Qualities of Good Specifications} \label{SecDesirableQs}
@@
 
-\subsection{Completeness}
-
-\noindent \textbf{Chosen definition:} The first sentence of Def
-\ref{CompletenessSelected}: A specification is complete to the extent that all
-of its parts are present and each part is fully developed.
+\subsection{Completeness \ad{owner}}
 
@@
 
-\ad{The following part was here before my modification}
+\noindent \textbf{Proposed definition}
 
-Documentation is said to be complete when all the requirements of the software
-are detailed. That is, each goal, functionality, attribute, design constraint,
-value, data, model, symbol, term (with its unit of measurement if applicable),
-abbreviation, acronym, assumption and performance requirement of the software is
-defined. The software's response to all classes of inputs, both valid and
-invalid and for both desired and undesired events, also needs to be specified.
+The first sentence of Definition \ref{CompletenessSelected}: A specification is
+complete to the extent that all of its parts are present and each part is fully
+developed.
 
-\subsection{Consistency}
-
-\noindent \textbf{Chosen definition:} Def \ref{ConsistencySelected}.
+\subsection{Consistency \ad{owner}}
 
@@
 
-\ad{The following part was here before my modification}
+\noindent \textbf{Proposed definition}
 
-Documentation is said to be consistent when no subset of individual statements
-are in conflict with each other. That is, a specification of an item made at one
-place in the document should not contradict the specification of the same item
-at another location.
+Definition \ref{ConsistencySelected}.
 
-\subsection{Modifiability}
+\subsection{Modifiability \jc{owner}}
 
@@
 
-\subsection{Traceability}
+\subsection{Traceability \jc{owner}}
 
@@
 
-\subsection{Unambiguity}
+\noindent \textbf{Proposed definition}
+
+\wss{Needs to be completed.}
+
+\subsection{Unambiguity \wss{owner}}
 
@@
 
-\subsection{Verifiability}
+\noindent \textbf{Proposed definition}
+
+\wss{Needs to be completed.}
+
+\subsection{Verifiability \wss{owner}}
 
@@
 
-\subsection{Abstract}
+\noindent \textbf{Proposed definition}
+
+\wss{Needs to be completed}
+
+\subsection{Abstract \wss{owner}}
 
@@
 
+\noindent \textbf{Proposed definition}
+
+\wss{Needs to be completed}
+
 \newpage

### 2019-11-06 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
Transformed Verifiability section to use definition environment

@@
 
+\noindent \textbf{Reasoning}
+
 \wss{The rationale is missing.  Please add.}
@@
 
-\begin{itemize}
-\item{Definition 1:} A software is verifiable if its properties can be verified
+\begin{defn}
+  A software is verifiable if its properties can be verified
   (\cite{GhezziEtAl2003}).
-\item{Definition 2:} Attributes of software that relate to the effort needed for
-  validating the modified software (\cite{berander2005software}).
-\item{Definition 3:}) The degree to which a system or component facilitates the
-  establishment of test criteria and the performance of tests to determine
-  whether those criteria have been met (\cite{IEEEStdGlossarySET1990}).
-\item{Definition 4:} The degree to which a requirement is stated in terms that
-  permit establishment of test criteria and performance
+\end{defn}
+
+\begin{defn}
+  Attributes of software that relate to the effort needed for validating the
+  modified software (\cite{berander2005software}).
+\end{defn}
+
+\begin{defn}
+  The degree to which a system or component facilitates the establishment of
+  test criteria and the performance of tests to determine whether those criteria
+  have been met (\cite{IEEEStdGlossarySET1990}).
+\end{defn}
+
+\begin{defn}
+  The degree to which a requirement is stated in terms that permit establishment
+  of test criteria and performance (\cite{IEEEStdGlossarySET1990}).
+\end{defn}
+
+\begin{defn}
+  Ease of performing testing on a software product or system
   (\cite{IEEEStdGlossarySET1990}).
-\item{Definition 5:} Ease of performing testing on a software product or system
-  (\cite{IEEEStdGlossarySET1990}).
-\item{\textbf{Proposed definition}:} Combo of definitions 4 and 5 ?
-\end{itemize}
-
-\noindent \textbf{Proposed Definition}
-
-\subsubsection{Discussion}
+\end{defn}
+
+\noindent \textbf{Proposed Definition}
+
+Combo of definitions 4 and 5 ?
+
+\noindent \textbf{Reasoning}
 
@@
   \oo{Only a few resource on validatability}
-\item{\textbf{Proposed definition}:} Definition 3
 \end{itemize}
@@
 
-\subsubsection{Discussion}
+Definition 3
+
+\noindent \textbf{Reasoning}
 
@@
   (\cite{meyer1988object}).
-\item{Proposed definition}
 \end{itemize}
@@
 
-\subsubsection{Discussion}
+\wss{Needs to be completed.}
+
+\noindent \textbf{Reasoning}
 
@@
   The quality can be further informally refined as the ability of a software to
-  keep an acceptable behavior, expressed in terms of robustness requirements, in
+  keep an acceptable behaviour, expressed in terms of robustness requirements, in
   spite of exceptional or unforeseen execution conditions (such as the
@@
 
-\noindent \textbf{Reasoning:}
+\noindent \textbf{Reasoning}
 
@@
 
+\noindent \textbf{Reasoning}
+
+\wss{Needs to be completed}
+
 \subsection{Maintainability \pmi{owner}}
@@
 \wss{Still needs to be completed}
+
+\noindent \textbf{Reasoning}
+
+\wss{Needs to be completed}
 
@@
 
+\noindent \textbf{Reasoning}
+
+\wss{Needs to be completed}
+
 \subsection{Visibility/Transparency \ad{owner}}
@@
 
-\noindent \textbf{Proposed definition}
+\noindent \textbf{Proposed Definition}
 
 Definition \ref{VisibilitySelected}.
+
+\noindent \textbf{Reasoning}
+
+\wss{Needs to be completed}
 
@@
 
-\noindent \textbf{Proposed definition}
+\noindent \textbf{Proposed Definition}
+
+\wss{Needs to be completed}
+
+\noindent \textbf{Reasoning}
 
@@
 
-\noindent \textbf{Proposed definition}
+\noindent \textbf{Proposed Definition}
 
@@
 amount of output (what is produced) per unit of input used.
+
+\noindent \textbf{Reasoning}
+
+\wss{Needs to be completed}
 
@@
 developed.
+
+\noindent \textbf{Reasoning}
+
+\wss{Needs to be completed}
 
@@
 
-\noindent \textbf{Proposed definition}
+\noindent \textbf{Proposed Definition}
 
 Definition \ref{ConsistencySelected}.
+
+\noindent \textbf{Reasoning}
+
+\wss{Needs to be completed}
 
@@
 
-IEEE Standard 610 seems to speak about this. (which is superceded?)
+IEEE Standard 610 seems to speak about this. (which is superseded?)
+
+\noindent \textbf{Reasoning}
+
+\wss{Needs to be completed}
 
@@
 
-\noindent \textbf{Proposed definition}
+\noindent \textbf{Proposed Definition}
 
 \wss{Needs to be completed.}
+
+\noindent \textbf{Reasoning}
+
+\wss{Needs to be completed}
 

### 2019-11-06 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
Transformed Validatability to use definition environments

@@
 
-\begin{defn}
+\begin{defn} \label{Defn_Verifiability1}
   The degree to which a requirement is stated in terms that permit establishment
@@
 
-\begin{defn}
+\begin{defn} \label{Defn_Verifiability2}
   Ease of performing testing on a software product or system
@@
 
-Combo of definitions 4 and 5 ?
+Combo of definitions~\ref{Defn_Verifiability1} and~\ref{Defn_Verifiability2} ?
 
@@
 
-\begin{itemize}
-\item{Definition 1:} Validatability of a software is the degree of ease in
-  validating(checking) that software meets user needs.
-\item{Definition 2:} Validatability means ``solving the right
-  equations''~\cite[p.~23]{Roache1998}.
+\begin{defn}
+  Validatability of a software is the degree of ease in validating (checking)
+  that software meets user needs.
+\end{defn}
+
+\begin{defn}
+  Validatability means ``solving the right equations''~\cite[p.~23]{Roache1998}.
   \oo{question about this definition}
-\item{Definition 3:} The degree of ease of evaluating software products or
-  system to determine whether it satisfies specified business
+\end{defn}
+
+\begin{defn} \label{DefnValidatability}
+  The degree of ease of evaluating software products or system to determine
+  whether it satisfies specified business
   requirements. \url{http://softwaretestingfundamentals.com/verification-vs-validation/}
-  \oo{Here I tweaked the meaning of validation to suit validatability}
-  \oo{Only a few resource on validatability}
-\end{itemize}
-
-\noindent \textbf{Proposed Definition}
-
-Definition 3
+  \oo{Here I tweaked the meaning of validation to suit validatability} \oo{Only
+    a few resource on validatability}
+\end{defn}
+
+\noindent \textbf{Proposed Definition}
+
+Definition~\ref{DefnValidatability}.
 

### 2019-11-06 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
Reformated Reliability to use definition environments

@@
 
-\begin{itemize}
-\item{Definition 1:} The probability that the software will operate as expected
-  over a specified time interval (\cite{GhezziEtAl2003}).
-\item{Definition 2:} A set of attributes that relate to the capability of
-  software to maintain its level of performance under stated conditions for a
-  stated period of time (\cite{berander2005software}).
-\item{Definition 3:} “The capability of the software product to maintain a
-  specified level of performance when used under specified conditions
+\begin{defn}
+  The probability that the software will operate as expected over a specified
+  time interval (\cite{GhezziEtAl2003}).
+\end{defn}
+
+\begin{defn}
+  A set of attributes that relate to the capability of software to maintain its
+  level of performance under stated conditions for a stated period of time
+  (\cite{berander2005software}).
+\end{defn}
+
+\begin{defn}
+  ``The capability of the software product to maintain a specified level of
+  performance when used under specified conditions''
   (\cite{international2001iso}).
-\item{Definition 4:} Code possesses the characteristic reliability to the extent
-  that it can be expected to perform its intended functions satisfactorily
-  (\cite{boehm2007software})
-\item{Definition 5:} A concern encompassing correctness and robustness
-  (\cite{meyer1988object}).
-\end{itemize}
+\end{defn}
+
+\begin{defn}
+  Code possesses the characteristic reliability to the extent that it can be
+  expected to perform its intended functions satisfactorily
+  (\cite{boehm2007software}).
+\end{defn}
+
+\begin{defn}
+  A concern encompassing correctness and robustness (\cite{meyer1988object}).
+\end{defn}
 

### 2019-11-06 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
Fixed cite commands to citet and citep, removed tildes

@@
   A measure of the ease of installation.  \wss{Where does this definition come
-    from? Please add a citation.}
+    from? Please add a citation.  I believe it is from \citet{McCallEtAl1977},
+    but you should verify this.}
 \end{defn}
@@
   Installability is related to the amount of human effort required to install a
-  software in a designated environment \cite{berander2005software}.
+  software in a designated environment \citep{berander2005software}.
 \end{defn}
@@
   ``The capability of a software product to be installed in a specified
-  environment'' \cite{berander2005software}.
+  environment'' \citep{berander2005software}.
 \end{defn}
@@
   ``The efficacy of the installation, uninstallation or reinstallation process
-  of a software product in a specified environment'' (\cite{iso201017043})
+  of a software product in a specified environment'' (\citep{iso201017043})
 \end{defn}
@@
 The quality of a software's operation is dependent on the degree of correctness
-\cite{berander2005software}. Correctness and reliability are said to have
+\citep{berander2005software}. Correctness and reliability are said to have
 dependencies, such that if a system exhibits a high degree of correctness then
-it tends to be reliable (\cite{GhezziEtAl2003}).
+it tends to be reliable (\citep{GhezziEtAl2003}).
 
@@
   A software is verifiable if its properties can be verified
-  (\cite{GhezziEtAl2003}).
+  \citep{GhezziEtAl2003}.
 \end{defn}
@@
   Attributes of software that relate to the effort needed for validating the
-  modified software (\cite{berander2005software}).
+  modified software \citep{berander2005software}.
 \end{defn}
@@
   test criteria and the performance of tests to determine whether those criteria
-  have been met (\cite{IEEEStdGlossarySET1990}).
+  have been met \citep{IEEEStdGlossarySET1990}.
 \end{defn}
@@
   The degree to which a requirement is stated in terms that permit establishment
-  of test criteria and performance (\cite{IEEEStdGlossarySET1990}).
+  of test criteria and performance \citep{IEEEStdGlossarySET1990}.
 \end{defn}
@@
   Ease of performing testing on a software product or system
-  (\cite{IEEEStdGlossarySET1990}).
+  \citep{IEEEStdGlossarySET1990}.
 \end{defn}
@@
 
-Verifiability involves ``solving the equations right''~\cite[p.~23]{Roache1998};
+Verifiability involves ``solving the equations right'' \citep[p.~23]{Roache1998};
 it benefits from rational documentation that systematically shows, with explicit
@@
 \begin{defn}
-  Validatability means ``solving the right equations''~\cite[p.~23]{Roache1998}.
+  Validatability means ``solving the right equations'' \citep[p.~23]{Roache1998}.
   \oo{question about this definition}
@@
   The probability that the software will operate as expected over a specified
-  time interval (\cite{GhezziEtAl2003}).
+  time interval (\citep{GhezziEtAl2003}).
 \end{defn}
@@
   level of performance under stated conditions for a stated period of time
-  (\cite{berander2005software}).
+  \citep{berander2005software}.
 \end{defn}
@@
   performance when used under specified conditions''
-  (\cite{international2001iso}).
+  \citep{international2001iso}.
 \end{defn}
@@
   expected to perform its intended functions satisfactorily
-  (\cite{boehm2007software}).
-\end{defn}
-
-\begin{defn}
-  A concern encompassing correctness and robustness (\cite{meyer1988object}).
+  \citep{boehm2007software}.
+\end{defn}
+
+\begin{defn}
+  A concern encompassing correctness and robustness \citep{meyer1988object}.
 \end{defn}
@@
 example measuring Mean Time to Fail (MTTF) can be a good measure of reliability
-\cite{berander2005software}.
+\citep{berander2005software}.
 
@@
 
-One core reference, for definitions and metrics, is Bevan~\cite{bevan1995measuring}.
+One core reference, for definitions and metrics, is
+\citet{bevan1995measuring}.
 
@@
 
-Section 11.2 of~\cite{adams2015nonfunctional} does have a full definition.
+Section 11.2 of \citet{adams2015nonfunctional} does have a full definition.
 
@@
 Reproducibility is a required component of the scientific
-method~\cite{Davison2012}.  Although QA has, ``a bad name among creative
-scientists and engineers''~\cite[p.~352]{Roache1998}, the community need to
+method \citep{Davison2012}.  Although QA has, ``a bad name among creative
+scientists and engineers'' \citep[p.~352]{Roache1998}, the community need to
 recognize that participating in QA management also improves reproducibility.
@@
 environment, version control and separating code from
-configuration/parameters~\cite{Davison2012}.
+configuration/parameters \citep{Davison2012}.
 
@@
 \noindent The ability to meet the needs of the present without compromising the ability of
-future generations to meet their own needs \cite{Brundtland1987}.
-\end{defn}
-
-This is the definition used by~\cite{IISD2019}.
+future generations to meet their own needs \citep{Brundtland1987}.
+\end{defn}
+
+This is the definition used by \citet{IISD2019}.
 
@@
 social, economic and environmental. \wss{cite UN paper [9] in
-  \cite{PenzenstadlerAndHenning2013}}  To this list Penzenstadler and Henning
-(2013) have added technical sustainability \cite{PenzenstadlerAndHenning2013}.
+  \citet{PenzenstadlerAndHenning2013}}  To this list Penzenstadler and Henning
+(2013) have added technical sustainability \citep{PenzenstadlerAndHenning2013}.
 Where technical sustainability for software is defined as:
@@
 usage of systems and their adequate evolution with changing surrounding
-conditions and respective requirements \cite{PenzenstadlerAndHenning2013}.
+conditions and respective requirements \citep{PenzenstadlerAndHenning2013}.
 \end{defn}
@@
 The fourth dimension of technical sustainability is also added
-by~\cite{WolframEtAl2017}.  Technical sustainability is the focus on the thesis
-by~\cite{Hygerth2016}.
+by \citep{WolframEtAl2017}.  Technical sustainability is the focus on the thesis
+by \citet{Hygerth2016}.
 
@@
   accompanying set of practices that enable a team to achieve and maintain an
-  optimal development pace indefinitely~\cite{Tate2005}.
-\end{defn}
-
-Parnas discusses as software aging~\cite{Parnas1994a}.
+  optimal development pace indefinitely \citep{Tate2005}.
+\end{defn}
+
+Parnas discusses as software aging \citep{Parnas1994a}.
 
@@
   conception to programming, distribution, installation, usage and
-  disposal~\cite{Heine2017}.
+  disposal \citep{Heine2017}.
 \end{defn}
@@
   sustainability means that the software will continue to be available in the
-  future, on new platforms, meeting new needs~\cite{Katz2016}.
+  future, on new platforms, meeting new needs \citep{Katz2016}.
 \end{defn}
@@
 -- Supports relevant concerns (Political, Economic, Social, Technical,
-Legal, Environmental)~\cite{Katz2016}.
-\end{defn}
-
-Paper critical of a lack of a definition~\cite{VentersEtAl2014}.
+Legal, Environmental) \citep{Katz2016}.
+\end{defn}
+
+Paper critical of a lack of a definition \citep{VentersEtAl2014}.
 
@@
 listed (abstraction), are adapted from the IEEE recommended practise for
-producing good software requirements~\cite{IEEE1998}.  Abstraction means only
+producing good software requirements \citep{IEEE1998}.  Abstraction means only
 revealing relevant details, which in a requirements document means stating what
@@
 Abstraction is an important software development principle for dealing with
-complexity~\cite[p.~40]{GhezziEtAl2003}.  Correctness was in the above list, so
+complexity \citep[p.~40]{GhezziEtAl2003}.  Correctness was in the above list, so
 it is not repeated here.  \citet{SmithAndKoothoor2016} present further details
@@
 ``The effort required to test a program to ensure that it performs its intended
-function'' \cite{VanVliet2000}.
+function'' \citep{VanVliet2000}.
 
@@
 the details that are unimportant to a software developer and emphasizes the
-information that is important.'' \cite{Krueger1992}
+information that is important.'' \citep{Krueger1992}
 \end{defn}
@@
 abstract from, details that are not relevant at the level we are currently
-working.''  \cite[p.\ 296]{VanVliet2000}
+working.''  \citep[p.\ 296]{VanVliet2000}
 \end{defn}

### 2019-11-06 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
Addition of McCall et al definitions (as found in Van  Vliet)

@@
   A measure of the ease of installation.  \wss{Where does this definition come
-    from? Please add a citation.  I believe it is from \citet{McCallEtAl1977},
-    but you should verify this.}
+    from? Please add a citation.  Is it from \citet{McCallEtAl1977}?  I thought
+    so originally, but after looking into it, I don't think so anymore.}
 \end{defn}
@@
 \begin{defn}
-  The extent to which a program satisfies its specifications and fulfills the
-  user's mission objectives \citep{McCallEtAl1977}. (As summarized in
+  ``The extent to which a program satisfies its specifications and fulfills the
+  user's mission objectives'' \citep{McCallEtAl1977}. (As summarized in
   \citet{VanVliet2000}.)
@@
 \oo{Verifiability is sometimes referred to as testability, so I culled some
-  testability definitions here}
+  testability definitions here}  \wss{culled means removed - is that your
+  intention?}
+
+\wss{In case we need it, here is a definition of testability.}
+
+\begin{defn}
+  ``The effort required to test a program to ensure that it performs its
+  intended function'' \citep{McCallEtAl1977}. (As summarized in
+  \citet{VanVliet2000}.)
+\end{defn}
 
@@
   A concern encompassing correctness and robustness \citep{meyer1988object}.
+\end{defn}
+
+\begin{defn}
+  ``The extent to which a program can be expected to perform its intended
+  function with required precision'' \citep{McCallEtAl1977}. (As summarized in
+  \citet{VanVliet2000}.)
 \end{defn}
@@
 
+\begin{defn}
+  ``The effort required to learn, operate, prepare input, and interpret output
+  of a program'' \citep{McCallEtAl1977}. (As summarized in
+  \citet{VanVliet2000}.)
+\end{defn}
+
 \noindent \textbf{Proposed Definition}
@@
 
+\begin{defn}
+  ``The effort required to locate and fix an error in an operational program''
+  \citep{McCallEtAl1977}. (As summarized in \citet{VanVliet2000}.) \wss{It looks
+    like \citet{pressman2005software} was using \citet{McCallEtAl1977} for their
+    definition.}
+\end{defn}
+
 \noindent \textbf{Proposed Definition}
@@
 add new functionalities with slight or no modification \citep{sandhu2010survey}.
+\end{defn}
+
+\begin{defn}
+  ``The extent to which a program (or parts thereof) can be reused in other
+  applications'' \citep{McCallEtAl1977}. (As summarized in
+  \citet{VanVliet2000}.)
 \end{defn}
@@
 platforms \citep{ghezzi1991fundamentals}.
+\end{defn}
+\begin{defn}
+  ``The effort required to transfer a program from one hardware and/or software
+  environment to another'' \citep{McCallEtAl1977}. (As summarized in
+  \citet{VanVliet2000}.)
 \end{defn}
@@
 \subsection{Interoperability \ad{owner}}
+
+\begin{defn}
+  ``The effort required to copule one system with another''
+  \citep{McCallEtAl1977}. (As summarized in \citet{VanVliet2000}.)
+\end{defn}
 

### 2019-11-06 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
Edits to definitions

--- 2019-11-06 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
@@
   ``The efficacy of the installation, uninstallation or reinstallation process
-  of a software product in a specified environment'' (\citep{iso201017043})
-\end{defn}
-
-\noindent \textbf{Proposed Definition}
-
-Definition~\ref{DefnInstallability}.
+  of a software product in a specified environment'' \citep{iso201017043}
+\end{defn}
+
+\noindent \textbf{Proposed Definition}
+
+Definition~\ref{DefnInstallability}.  \wss{This definition does not allow for
+  measuring the machine time.}
 
@@
   A software is correct if it functions according to its specified functional
-  and non-functional specification \citep{GhezziEtAl2003}.
+  and non-functional specification \citep{GhezziEtAl2003}.  \wss{This is not how
+    \citet{GhezziEtAl2003} defines correctness.  They only require satisfaction
+    of the functional requirements.}
+
 \end{defn}
@@
 dependencies, such that if a system exhibits a high degree of correctness then
-it tends to be reliable (\citep{GhezziEtAl2003}).
+it tends to be reliable \citep{GhezziEtAl2003}.
 
@@
   The probability that the software will operate as expected over a specified
-  time interval (\citep{GhezziEtAl2003}).
+  time interval \citep{GhezziEtAl2003}.
 \end{defn}
@@
 
-Definition \ref{RobustnessDefnSelected} rephrased:
-
-Software possesses the characteristic of robustness if it fulfills its
-requirements correctly or behaves reasonably in circumstances not anticipated in
-the requirements specification.
+Definition \ref{RobustnessDefnSelected} rephrased: Software possesses the
+characteristic of robustness if it fulfills its requirements correctly or
+behaves reasonably in circumstances not anticipated in the requirements
+specification. \wss{I think we might want to separate the definition of
+  robustness and correctness.  I'm thinking we just want the definition of
+  robustness to be how the software behaves in unanticipated circumstances.}
 
@@
 
-Definition \ref{PerformanceDefnSelected}:
-
-How well or how rapidly the system must perform specific functions. Performance
-requirements encompass speed (database response times, for instance), throughput
-( transactions per second), capacity (con-current usage loads), and timing (hard
-real-time demands).
+Definition \ref{PerformanceDefnSelected}: How well or how rapidly the system
+must perform specific functions. Performance requirements encompass speed
+(database response times, for instance), throughput ( transactions per second),
+capacity (con-current usage loads), and timing (hard real-time demands).
 
@@
 
-Combined Definition
-\ref{MaintainabilityDefnSelected1} and Definition
-\ref{MaintainabilityDefnSelected2}:
-
-The ease with which a software system or component can be modified to correct
-faults, improve performance or other attributes, or satisfy new requirements.
+Combined Definition \ref{MaintainabilityDefnSelected1} and Definition
+\ref{MaintainabilityDefnSelected2}: The ease with which a software system or
+component can be modified to correct faults, improve performance or other
+attributes, or satisfy new requirements.
 
@@
 
-Definition \ref{ReusabilityDefnSelected}:
-
-The extent to which a software component can be used with or without adaptation
-in a problem solution other than the one for which it was originally developed.
+Definition \ref{ReusabilityDefnSelected}: The extent to which a software
+component can be used with or without adaptation in a problem solution other
+than the one for which it was originally developed.
 
@@
 
-Definition \ref{PortabilityDefnSelected} rephrased:
-
-Effort required to transfer a program between system environments.
+Definition \ref{PortabilityDefnSelected} rephrased: Effort required to transfer
+a program between system environments.
 

### 2019-11-06 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
indicate which definitions are 'done'. Fix 4 citations to have authors to placate bibtex.

@@
 
-\usepackage[round]{natbib}
+\usepackage{natbib}
 \usepackage{booktabs}
@@
 \newcommand{\ad}[1]{\authornote{cyan}{AD}{#1}} %Ao Dong
+
+\newcommand{\notdone}[1]{\textcolor{red}{#1}}
 
@@
 
-\subsection{Installability \oo{owner}}
+\subsection{\notdone{Installability} \oo{owner}}
 
@@
 
-\subsection{Correctness \oo{owner}}
+\subsection{\notdone{Correctness} \oo{owner}}
 
@@
 
-\subsection{Verifiability \oo{owner}}
+\subsection{\notdone{Verifiability} \oo{owner}}
 
@@
 
-\subsection{Validatability \oo{owner}}
+\subsection{\notdone{Validatability} \oo{owner}}
 
@@
 
-\subsection{Reliability \oo{owner}}
+\subsection{\notdone{Reliability} \oo{owner}}
 
@@
 
-\subsection{Robustness \pmi{owner}}
+\subsection{\notdone{Robustness} \pmi{owner}}
 \begin{defn}
@@
 
-\subsection{Performance \pmi{owner}}
+\subsection{\notdone{Performance} \pmi{owner}}
 
@@
 
-\subsection{Usability \jc{owner}}
-
-ISO defines usability as
-\begin{quote}
+\subsection{\notdone{Usability} \jc{owner}}
+
+\begin{defn}
 The extent to which a product can be used by specified users to achieve
@@
 context of use.
-\end{quote}
+\end{defn}
+ISO defines usability as
 
@@
 
-\subsection{Maintainability \pmi{owner}}
+\subsection{\notdone{Maintainability} \pmi{owner}}
 
@@
 
-\subsection{Reusability \pmi{owner}}
+\subsection{\notdone{Reusability} \pmi{owner}}
 
@@
 
-\subsection{Portability \pmi{owner}}
+\subsection{\notdone{Portability} \pmi{owner}}
 \begin{defn}
@@
 
-\subsection{Understandability \jc{owner}}
+\subsection{\notdone{Understandability} \jc{owner}}
 
@@
 
-\subsection{Interoperability \ad{owner}}
+\subsection{\notdone{Interoperability} \ad{owner}}
 
@@
 
-\subsection{Visibility/Transparency \ad{owner}}
+\subsection{\notdone{Visibility/Transparency} \ad{owner}}
 
@@
 
-\subsection{Reproducibility \wss{owner}}
+\subsection{\notdone{Reproducibility} \wss{owner}}
 
@@
 
-\subsection{Productivity \ad{owner}}
+\subsection{\notdone{Productivity} \ad{owner}}
 
@@
 
-\subsection{Sustainability \wss{owner}}
+\subsection{\notdone{Sustainability} \wss{owner}}
 
@@
 
-\subsection{Completeness \ad{owner}}
+\subsection{\notdone{Completeness} \ad{owner}}
 
@@
 
-\subsection{Consistency \ad{owner}}
+\subsection{\notdone{Consistency} \ad{owner}}
 
@@
 
-\subsection{Modifiability \jc{owner}}
+\subsection{\notdone{Modifiability} \jc{owner}}
 
@@
 
-\subsection{Traceability \jc{owner}}
+\subsection{\notdone{Traceability} \jc{owner}}
 
@@
 
-\subsection{Unambiguity \wss{owner}}
+\subsection{\notdone{Unambiguity} \wss{owner}}
 
@@
 
-\subsection{Verifiability \wss{owner}}
+\subsection{\notdone{Verifiability} \wss{owner}}
 
@@
 
-\subsection{Abstract \wss{owner}}
+\subsection{\notdone{Abstract} \wss{owner}}
 

### 2019-11-06 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
Added some notes from meeting on Nov 6, 2019

--- 2019-11-06 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
@@
 specification could be a specification of requirements, design or a test plan.
+
+\wss{Introduce citation conventions.}
 
@@
 \begin{defn} \label{RobustnessDefnSelected}
-  A program is robust if it behaves "reasonably", even in circumstances that
+  A program is robust if it behaves ``reasonably'', even in circumstances that
   were not anticipated in the requirements specification - for example, when it
@@
 \begin{defn}
-  ``The effort required to copule one system with another''
+  ``The effort required to couple one system with another''
   \citep{McCallEtAl1977}. (As summarized in \citet{VanVliet2000}.)
@@
 \[\text{Productivity} = \dfrac{\text{Outputs produced by the
-process}}{\text{Inputs consumed by the process}}\]
-Defining inputs. For the software process, providing a meaningful definition of
-inputs is a nontrivial but generally workable problem. Inputs to the software
-process generally comprise labor, computers, supplies, and other support
-facilities and equipment. Defining outputs. The big problem in defining software
-productivity is defining outputs. Here we find a defining delivered source
-instructions (DSI) or lines of code as the output of the software process is
-totally inadequate, and they argue that there are a number of deficiencies in
-using DSI. However, most organizations doing practical productivity measurement
-still use DSI as their primary metric \citep{Boehm1987}.
+      process}}{\text{Inputs consumed by the process}}\] Defining inputs. For
+the software process, providing a meaningful definition of inputs is a
+nontrivial but generally workable problem. Inputs to the software process
+generally comprise labor, computers, supplies, and other support facilities and
+equipment. Defining outputs. The big problem in defining software productivity
+is defining outputs. Here we find a defining delivered source instructions (DSI)
+or lines of code as the output of the software process is totally inadequate,
+and they argue that there are a number of deficiencies in using DSI. However,
+most organizations doing practical productivity measurement still use DSI as
+their primary metric \citep{Boehm1987}.
 \end{defn}

### 2019-11-06 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
Addition to introduction section to clarify the "template" used for each quality summary/discussion

@@
 
-\wss{Introduce citation conventions.}
+The pattern in the presentation for each of the qualities is the same.  First we
+summarize all of the definitions that we could find in the literature.  To keep
+the clutter of quotation marks down, we have adopted the convention that each
+definition is given verbatim from the cited source, but without showing
+quotation marks.  After the summary of the existing definitions, we propose the
+definition that we would like to work with going forward.  This definition can
+either be our preference from the existing definitions, or a new definition,
+which is often found by combining existing definitions.  Following the proposed
+definition is an explanation for the reasoning that led to this choice.
 

### 2019-11-07 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
changes defn robustness and performance

@@
 Definition \ref{RobustnessDefnSelected} rephrased: Software possesses the
-characteristic of robustness if it fulfills its requirements correctly or
-behaves reasonably in circumstances not anticipated in the requirements
-specification. \wss{I think we might want to separate the definition of
-  robustness and correctness.  I'm thinking we just want the definition of
-  robustness to be how the software behaves in unanticipated circumstances.}
+characteristic of robustness if it behaves ``reasonably'' in circumstances that were not anticipated in the requirements specification or are in violation of it.
 
@@
 
-\pmi{I think this is done.}
+\pmi{Still want to refine what we mean by reasonable}
 
@@
 
-\begin{defn}
+\begin{defn} \label{PerformanceDefnSelected}
 The degree to which a system or component accomplishes its designated functions
@@
 \end{defn}
-\begin{defn} \label{PerformanceDefnSelected}
+\begin{defn} \label{PerformanceDefnSelected2}
 How well or how rapidly the system must perform specific functions. Performance
@@
 
-Definition \ref{PerformanceDefnSelected}: How well or how rapidly the system
-must perform specific functions. Performance requirements encompass speed
+Combined definition \ref{PerformanceDefnSelected} and \ref{PerformanceDefnSelected2}: The degree to which a system or component accomplishes its designated functions within given constraints, such as speed
 (database response times, for instance), throughput ( transactions per second),

### 2019-11-19 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
updated Ao's 5 qualities

@@
 Definition \ref{RobustnessDefnSelected} rephrased: Software possesses the
-characteristic of robustness if it behaves ``reasonably'' in circumstances that were not anticipated in the requirements specification or are in violation of it.
+characteristic of robustness if it behaves ``reasonably'' in circumstances that
+were not anticipated in the requirements specification or are in violation of
+it.
 
@@
 \begin{defn}
+\label{InteroperabilitySelected}
 The degree to which two or more systems, products or components can exchange
@@
 \citep{ISO/IEC25010}.
+\end{defn}
+\begin{defn}
+The ability of a system to coexist and cooperate with other systems.
+\citep{ghezzi1991fundamentals}.
 \end{defn}
@@
 \begin{defn}
-\label{InteroperabilitySelected}
 Interoperability is the ability of different information systems, devices and
@@
 Definition \ref{InteroperabilitySelected}
-rephrased: Interoperability is the ability of different information systems,
-devices and applications (‘systems’) to access, exchange, integrate and
-cooperatively use data in a coordinated manner, within and across
-organizational, regional and national boundaries, to provide timely and seamless
-portability of information.
-
-\wss{I moved the proposed definition to the end, to match the other sections.
-  Hopefully I did not change the meaning.}
-
-\noindent \textbf{Reasoning}
-
-\wss{Needs to be completed}
+rephrased: The effort to exchange information between two or more systems,
+products or components and use the information that has been exchanged.
+
+\noindent \textbf{Reasoning}
+
+This definition is concise and also detailed enough to show the concept not only
+on system, but also on products and components. Rephrased ``degree'' to
+``effort'' because the latter is more measurable.
 
@@
 
-Definition \ref{VisibilitySelected}.
-
-\noindent \textbf{Reasoning}
-
-\wss{Needs to be completed}
+Definition \ref{VisibilitySelected} rephrased: The clarity of documentation for
+all of the steps and the current status of a software development process.
+
+\noindent \textbf{Reasoning}
+
+Definition \ref{VisibilitySelected} is the only one could be found. It is
+rephrased because the original one might refer to binary status - ``visible'' or
+``unvisible''.
 
@@
 
+\begin{defn}
+A quality of the software production process, referring to its efficiency and
+performance \citep{ghezzi1991fundamentals}.
+\end{defn}
 \begin{defn}
@@
 
-\wss{Needs to be completed}
+It is concise and measurable.
 
@@
 \begin{defn}
+\label{ConsistencySelected}
 A specification is consistent to the extent that its provisions do not conflict
@@
 \begin{defn}
-\label{ConsistencySelected}
 Consistency: 1. the degree of uniformity, standardization, and freedom from
@@
 
-Definition \ref{ConsistencySelected}.
-
-\noindent \textbf{Reasoning}
-
-\wss{Needs to be completed}
+The first sentence of Definition \ref{ConsistencySelected}: A specification is
+consistent to the extent that its provisions do not conflict with each other or
+with governing specifications and objectives.
+
+\noindent \textbf{Reasoning}
+
+It is concise and measurable.
 

### 2019-11-20 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
Add correctness, reliability, usability and verifiability from Ghezzi1991 text book

@@
 
+\begin{defn}
+``A program is functionally correct if it behaves according to its stated
+functional specifications'' \citep{ghezzi1991fundamentals}.
+\end{defn}
+
 \noindent \textbf{Proposed Definition}
@@
 
+\begin{defn}
+``Informally, software is reliable if the user can depend on it''
+\citep{ghezzi1991fundamentals}.
+\end{defn}
+
 \noindent \textbf{Proposed Definition}
@@
 
-Combined definition \ref{PerformanceDefnSelected} and \ref{PerformanceDefnSelected2}: The degree to which a system or component accomplishes its designated functions within given constraints, such as speed
+Combined definition \ref{PerformanceDefnSelected} and
+\ref{PerformanceDefnSelected2}: The degree to which a system or component
+accomplishes its designated functions within given constraints, such as speed
 (database response times, for instance), throughput ( transactions per second),
@@
   \citet{VanVliet2000}.)
+\end{defn}
+
+\begin{defn}
+``A software is usable - or user fridendly - if its human users find it easy to
+use'' \citep{ghezzi1991fundamentals}.
 \end{defn}
@@
 
+\begin{defn}
+A software system is verifiable if its properties can be verified easily
+\citep{ghezzi1991fundamentals}.
+\end{defn}
+
 \noindent \textbf{Proposed Definition}

### 2019-11-20 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
Add resuability and understandability from Ghezzi1991 text boot
Review and comments on several qualities, in prep for group meeting on Nov 20, 2019

@@
   ``The capability of a software product to be installed in a specified
-  environment'' \citep{berander2005software}.
+  environment'' \citep{berander2005software}.  \wss{remove the quotation marks -
+    we are assuming all definitions are directly quoted from the source.  Please
+  remove quotes for all of the definitions that you ``own.''}
+
 \end{defn}
@@
   test criteria and the performance of tests to determine whether those criteria
-  have been met \citep{IEEEStdGlossarySET1990}.
+  have been met \citep{IEEEStdGlossarySET1990}. \wss{The author needs to be
+    fixed so that this displays properly.}
 \end{defn}
@@
   Validatability of a software is the degree of ease in validating (checking)
-  that software meets user needs.
+  that software meets user needs. \wss{Need a citation}
 \end{defn}
@@
 \end{defn}
-\begin{defn}
+\begin{defn} \label{RobustnessPartialDefn}
   Code possesses the characteristic of robustness to the extent that it can
@@
 
-Definition \ref{RobustnessDefnSelected} rephrased: Software possesses the
-characteristic of robustness if it behaves ``reasonably'' in circumstances that
-were not anticipated in the requirements specification or are in violation of
-it.
+Definition \ref{RobustnessDefnSelected} rephrased \wss{Aren't you really
+  combining Definition~\ref{RobustnessDefnSelected} with
+  Definition~\ref{RobustnessPartialDefn}?} : Software possesses the
+characteristic of robustness if it behaves ``reasonably'' in two situations: i)
+when it encounters circumstances not anticipated in the requirements
+specification; and ii) when the assumptions in its requirements specification
+are violated.
 
@@
 (database response times, for instance), throughput ( transactions per second),
-capacity (con-current usage loads), and timing (hard real-time demands).
-
-\noindent \textbf{Reasoning}
-
-This definition includes all important categories of performance.
+capacity (con-current usage loads), and timing (hard real-time
+demands). \wss{What is the difference between speed and timing?}
+
+\noindent \textbf{Reasoning}
+
+This definition includes all important categories of performance.  \wss{What are
+  the important categories?}
 
@@
 environment \citep{IEEEStdGlossarySET1990}. 
-\end{defn}\
+\end{defn}
 \begin{defn}
@@
 efficiency with which a product or system can be modified by the intended
-maintainers \citep{ISO/IEC25010}.
+maintainers \citep{ISO/IEC25010}. \wss{The author needs to be updated so that
+  this display properly.}
 \end{defn}
@@
 software to environmental changes and changes in the requirements and functional
-specifications)\citep{pfleeger2006software}.
+specifications) \citep{pfleeger2006software}.
 \end{defn}
@@
 \begin{defn}
-  ``The effort required to locate and fix an error in an operational program''
+  The effort required to locate and fix an error in an operational program
   \citep{McCallEtAl1977}. (As summarized in \citet{VanVliet2000}.) \wss{It looks
     like \citet{pressman2005software} was using \citet{McCallEtAl1977} for their
-    definition.}
+    definition.} \wss{PM - can you please look into whether
+    \citet{pressman2005software} cite \citet{McCallEtAl1977}?  Do they give a
+    reason for dropping the word operational?}
 \end{defn}
@@
 This definition includes all potential reasons to modify the software.
+\wss{What are the potential reasons?  How do you know that we haven't missed
+  any?}  \wss{What are the other attributes?  What is an attribute?  Is it the
+  same as a quality?  We are calling performance a quality in this document.}
 
@@
 \begin{defn}
-  ``The extent to which a program (or parts thereof) can be reused in other
-  applications'' \citep{McCallEtAl1977}. (As summarized in
+  The extent to which a program (or parts thereof) can be reused in other
+  applications \citep{McCallEtAl1977}. (As summarized in
   \citet{VanVliet2000}.)
@@
 
-\pmi{I think this is done.}
+\pmi{I think this is done.} \wss{Looks good at this time.}
 
@@
 one environment to another (including the organizational, hardware, of software
-environment)\citep{pfleeger2006software}.
+environment) \citep{pfleeger2006software}.
 \end{defn}
@@
 Definition \ref{PortabilityDefnSelected} rephrased: Effort required to transfer
-a program between system environments.
+a program between system environments. \wss{Should we clarify what a system
+  environment is?  What is a system environment?  Hardware and OS?}
 
@@
 knowledge of the unique characteristics of those units
-\citep{ISO/IEC/IEEE24765}.
+\citep{ISO/IEC/IEEE24765}. \wss{We need to fix the author so that this displays properly}
 \end{defn}
@@
 ``unvisible''.
+
+\wss{Did you also look for transparency?  You should be able to find definitions
+  for transparency, especially if you look outside of a software context.  For
+  instance, \url{https://www.macmillandictionary.com/dictionary/british/transparent}}
 
@@
 most organizations doing practical productivity measurement still use DSI as
-their primary metric \citep{Boehm1987}.
+their primary metric \citep{Boehm1987}. \wss{Is this a direct quote from
+  \citet{Boehm1987}?  The sentences seem incomplete?}
 \end{defn}
@@
 Productivity is the amount of output (what is produced) per unit of input
-used.If we can measure the size of the software product and the effort required
+used. If we can measure the size of the software product and the effort required
 to develop the product, we have:
@@
 
-It is concise and measurable.
+It is concise and measurable. \wss{What is the output?  What is the input?  I
+  think the definition needs to give more information on these.  In particular,
+  the above definitions focus on code as the output, but documentation, test
+  cases etc should also be part of the output.  If we are going to measure this,
+  we need a better idea of what we are measuring for outputs and inputs.}
 
@@
 
+\wss{Were there any other definitions of completeness?  You could add the
+  definition of completeness from \citet[p.\ 5--6]{IEEE1998}.  This definition is for
+  requirements, but maybe there is something we can generalize from the
+  definition?  We could also look for definitions outside of software development.}
+
 \noindent \textbf{Proposed Definition}
@@
 \end{defn}
+
+\wss{The definition from \citet{IEEE1998} might again be useful.  We could also
+  look for definitions outside of software development.  Even a dictionary
+  definition could be helpful.}
 
@@
     whether the products of a given development phase satisfy the conditions
-    imposed at the start of that phase. [IEEE-STD-610]
+    imposed at the start of that phase. [IEEE-STD-610] \wss{Need a proper citation}
   \item Software Validation: The process of evaluating software during or at the
     end of the development process to determine whether it satisfies specified
-    requirements. [IEEE-STD-610] 
+    requirements. [IEEE-STD-610] \wss{Need a proper citation}
 \end{itemize}

### 2019-11-20 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
updated installability

--- 2019-11-19 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
@@
 \begin{defn}
-  A measure of the ease of installation.  \wss{Where does this definition come
+  A measure of the ease of installation \citep{smith2015comparing}.\wss{Where does this definition come
     from? Please add a citation.  Is it from \citet{McCallEtAl1977}?  I thought
     so originally, but after looking into it, I don't think so anymore.}
-\end{defn}
-
-\begin{defn} \label{DefnInstallability}
+   \oo{@smith you added the definition in here but I guess it's from here}
+\end{defn}
+
+\begin{defn}
   Installability is related to the amount of human effort required to install a
@@
 
-\begin{defn}
+\begin{defn} \label{InstallabilityDefnSelected1}
   ``The efficacy of the installation, uninstallation or reinstallation process
@@
 
-\noindent \textbf{Proposed Definition}
-
-Definition~\ref{DefnInstallability}.  \wss{This definition does not allow for
+\begin{defn}
+ The degree of effectiveness and efficiency with which a product or system can be successfully installed and/or uninstalled in a specified environment \citep{ISO/IEC25010}.
+\end{defn}
+
+\begin{defn} \label{InstallabilityDefnSelected2}
+	 Installability refers to the cost or effort required for the installation of a software or product, given an installation is possible to begin with \citep{lenhard2013measuring}
+\end{defn}
+
+\noindent \textbf{Proposed Definition}
+
+Definition~ Combined definitions \ref{InstallabilityDefnSelected1} and \ref{InstallabilityDefnSelected2} rephrased.
+Installability refers to the cost or effort required for the installation, uninstallation or reinstallation of a software or product in a specified enviroment.  \wss{This definition does not allow for
   measuring the machine time.}
@@
 \wss{The rationale is missing.  Please add.}
+\oo{The definition allows to measure the cost both from system and human view points}
 

### 2019-11-20 StateOfPractice/QDefOfQualities/QDefOfQualities.tex

--- 2019-11-20 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
@@
 
+\begin{defn}
+A product is reusable if we can ``use the product - perhaps with minor changes -
+to build another product'' \citep{ghezzi1991fundamentals}.
+\end{defn}
+
 \noindent \textbf{Proposed Definition}
@@
 Section 11.2 of \citet{adams2015nonfunctional} does have a full definition.
+
+\begin{defn}
+The effort ``to uncover the logic of the application''
+\citep{ghezzi1991fundamentals}.
+\end{defn}
 

### 2019-11-20 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
updated Interoperability and Visibility/Transparency

@@
 
-Definition \ref{InteroperabilitySelected}
-rephrased: The effort to exchange information between two or more systems,
-products or components and use the information that has been exchanged.
+Definition \ref{InteroperabilitySelected}.
 
@@
 This definition is concise and also detailed enough to show the concept not only
-on system, but also on products and components. Rephrased ``degree'' to
-``effort'' because the latter is more measurable.
+on system, but also on products and components. It also covered the concept with
+more than 2 systems.
 
@@
 
-Definition \ref{VisibilitySelected} rephrased: The clarity of documentation for
-all of the steps and the current status of a software development process.
+Definition \ref{VisibilitySelected} rephrased: The extent to which all of the
+steps and the current status of a software development process are documented
+clearly.
 

### 2019-11-21 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
updated all of my qualities PM

@@
 
-\subsection{\notdone{Robustness} \pmi{owner}}
+\subsection{{Robustness} \pmi{owner}}
 \begin{defn}
@@
 
-Definition \ref{RobustnessDefnSelected} rephrased \wss{Aren't you really
-  combining Definition~\ref{RobustnessDefnSelected} with
-  Definition~\ref{RobustnessPartialDefn}?} : Software possesses the
+Definition \ref{RobustnessPartialDefn} and Definition \ref{RobustnessDefnSelected} rephrased: Software possesses the
 characteristic of robustness if it behaves ``reasonably'' in two situations: i)
@@
 
-\pmi{Still want to refine what we mean by reasonable}
+\pmi{Still may want to refine what we mean by reasonable}
 
@@
 
-This definition includes all important categories of performance.  \wss{What are
-  the important categories?}
-
-\pmi{I think this is done.}
+This definition offers a comprehensive list of constraints that are commonly associated with software performance, such as speed, throughput, capacity, and timing.
+
+\pmi{I will check if Wiegers discusses the difference between speed and timing.}
 
@@
 efficiency with which a product or system can be modified by the intended
-maintainers \citep{ISO/IEC25010}. \wss{The author needs to be updated so that
-  this display properly.}
+maintainers \citep{ISO/IEC25010}.
 \end{defn}
@@
     reason for dropping the word operational?}
+\pmi{Yes, \citet{pressman2005software} does cite the relevant section of the textbook with \citet{McCallEtAl1977}.
+While they do not give a reason for specifically dropping the word operational, they do indicate that their given definition ``is a very limited definition''.
+Perhaps their intent was breadth.}
 \end{defn}
@@
 
-This definition includes all potential reasons to modify the software.
-\wss{What are the potential reasons?  How do you know that we haven't missed
-  any?}  \wss{What are the other attributes?  What is an attribute?  Is it the
+This definition offers a comprehensive list of potential reasons for modifying software, such as correcting faults, improving performance or other attributes, or satisfying new requirements.
+
+ \wss{What are the other attributes?  What is an attribute?  Is it the
   same as a quality?  We are calling performance a quality in this document.}
-
-\pmi{I think this is done.}
-
-\subsection{\notdone{Reusability} \pmi{owner}}
+\pmi{The term quality attribute has come up often in the literature - qualities appear to be a subset of attributes.
+Quality attributes seem to be associated with non-functional requirements.
+The term resource attribute was encountered in a journal and implied an association with a functional requirement.
+The two might differ along those lines of NFR and FR.}
+
+\pmi{I think this is done}
+
+\subsection{{Reusability} \pmi{owner}}
 
@@
 software component(s) being transferred.
-
-\pmi{I think this is done.} \wss{Looks good at this time.}
 
@@
 Definition \ref{PortabilityDefnSelected} rephrased: Effort required to transfer
-a program between system environments. \wss{Should we clarify what a system
-  environment is?  What is a system environment?  Hardware and OS?}
+a program between system environments (including hardware and software).
 

### 2019-12-02 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
added notes to preformance quality
updated interoperability and visibility

@@
 \newcommand{\notdone}[1]{\textcolor{red}{#1}}
+\newcommand{\done}[1]{\textcolor{black}{#1}}
 
@@
 \pmi{I will check if Wiegers discusses the difference between speed and timing.}
+\pmi{Wiegers did not discuss this difference.}
+\pmi{Speed - change over time; timing - related to specification: when something should happen in relation to something else - state transitions}
 
@@
 
-\subsection{\notdone{Interoperability} \ad{owner}}
+\subsection{\done{Interoperability} \ad{owner}}
 
@@
 knowledge of the unique characteristics of those units
-\citep{ISO/IEC/IEEE24765}. \wss{We need to fix the author so that this displays properly}
+\citep{ISO/IEC/IEEE24765}.
 \end{defn}
@@
 
+\begin{defn}
+Visibility provides transparency into the development process. It is the ability
+to see progress at any point and determine the distance to completion of a goal.
+Visibility provides status of not only the progress of the project, but the
+product itself \citep{GSA2019}.
+\end{defn}
+
+\begin{defn}
+Business process visibility, also called process visibility, is the ability to
+accurately and completely view the processes, transactions and other activities
+operating within an enterprise \citep{Rouse2013}.
+\end{defn}
+
+\begin{defn}
+Process transparency refers to the ability to look inside. The “look inside”
+provides an in-depth and clear visibility into the business processes and how
+these operate \citep{PRIME2019}.
+\end{defn}
+
 \noindent \textbf{Proposed Definition}
@@
 
-Definition \ref{VisibilitySelected} is the only one could be found. It is
+Definition \ref{VisibilitySelected} points out that documentation is the way to improve visibility. It is
 rephrased because the original one might refer to binary status - ``visible'' or
 ``unvisible''.
-
-\wss{Did you also look for transparency?  You should be able to find definitions
-  for transparency, especially if you look outside of a software context.  For
-  instance, \url{https://www.macmillandictionary.com/dictionary/british/transparent}}
 

### 2019-12-03 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
updated reliability

@@
 colorlinks=true,       % false: boxed links; true: colored links
-    linkcolor=red,          % color of internal links (change box color with linkbordercolor)
+linkcolor=red,          % color of internal links (change box color with
+%linkbordercolor)
 citecolor=blue,       % color of links to bibliography
@@
 \begin{defn}
-  A measure of the ease of installation \citep{smith2015comparing}.\wss{Where does this definition come
-    from? Please add a citation.  Is it from \citet{McCallEtAl1977}?  I thought
-    so originally, but after looking into it, I don't think so anymore.}
+	A measure of the ease of installation \citep{smith2015comparing}.\wss{Where
+	does this definition come from? Please add a citation.  Is it from
+	\citet{McCallEtAl1977}? I thought so originally, but after looking into it, I
+	don't think so anymore.}
 	\oo{@smith you added the definition in here but I guess it's from here}
@@
 	Installability is related to the amount of human effort required to install a
-  software in a designated environment \citep{berander2005software}.
-\end{defn}
-
-\begin{defn}
-  ``The capability of a software product to be installed in a specified
-  environment'' \citep{berander2005software}.  \wss{remove the quotation marks -
+	software in a designated environment \citep{iso2001iec}
+		{ berander2005software}.
+\end{defn}
+
+\begin{defn}
+	The capability of a software product to be installed in a specified
+	environment \citep{berander2005software}.  \wss{remove the quotation marks -
 		we are assuming all definitions are directly quoted from the source.  Please
 		remove quotes for all of the definitions that you ``own.''} 
-
 \end{defn}
@@
 \begin{defn}
- The degree of effectiveness and efficiency with which a product or system can be successfully installed and/or uninstalled in a specified environment \citep{ISO/IEC25010}.
+	The degree of effectiveness and efficiency with which a product or system can
+	be successfully installed and/or uninstalled in a specified environment
+	\citep{ISO/IEC25010}.
 \end{defn}
@@
 \begin{defn} \label{InstallabilityDefnSelected2}
-	 Installability refers to the cost or effort required for the installation of a software or product, given an installation is possible to begin with \citep{lenhard2013measuring}
-\end{defn}
-
-\noindent \textbf{Proposed Definition}
-
-Definition~ Combined definitions \ref{InstallabilityDefnSelected1} and \ref{InstallabilityDefnSelected2} rephrased.
-Installability refers to the cost or effort required for the installation, uninstallation or reinstallation of a software or product in a specified enviroment.  \wss{This definition does not allow for
+	Installability refers to the cost or effort required for the installation of a
+	software or product, given an installation is possible to begin with
+	\citep{lenhard2013measuring}
+\end{defn}
+
+\noindent \textbf{Proposed Definition}
+
+Definition~ Combined definitions \ref{InstallabilityDefnSelected1} and
+\ref{InstallabilityDefnSelected2} rephrased.
+Installability refers to the cost or effort required for the installation,
+uninstallation or reinstallation of a software or product in a specified
+enviroment.  \wss{This definition does not allow for
 	measuring the machine time.}
@@
 \wss{The rationale is missing.  Please add.}
-\oo{The definition allows to measure the cost both from system and human view points}
+\oo{The definition allows to measure the cost both from system and human view
+	points}
 
@@
 
-Verifiability involves ``solving the equations right'' \citep[p.~23]{Roache1998};
+Verifiability involves ``solving the equations right''
+\citep[p.~23]{Roache1998};
 it benefits from rational documentation that systematically shows, with explicit
@@
 \begin{defn}
-  Validatability means ``solving the right equations'' \citep[p.~23]{Roache1998}.
+	Validatability means ``solving the right equations''
+	\citep[p.~23]{Roache1998}.
 	\oo{question about this definition}
@@
 	whether it satisfies specified business
-  requirements. \url{http://softwaretestingfundamentals.com/verification-vs-validation/}
+	requirements.
+	\url{http://softwaretestingfundamentals.com/verification-vs-validation/}
 	\oo{Here I tweaked the meaning of validation to suit validatability} \oo{Only
@@
 	A set of attributes that relate to the capability of software to maintain its
-  level of performance under stated conditions for a stated period of time
-  \citep{berander2005software}.
-\end{defn}
-
-\begin{defn}
-  ``The capability of the software product to maintain a specified level of
-  performance when used under specified conditions''
-  \citep{international2001iso}.
+	level of performance under stated conditions for a stated period of time \citep{international2001iso}. (As summarized in \citep{berander2005software}.)
+\end{defn}
+
+\begin{defn} \label{reliabilitySelected1}
+	Probability of failure-free operation of a computer program in a specified environment for a specified time \citep{musa1987software}. (As stated in \citep{pressman2005software}.)
 \end{defn}
@@
 	expected to perform its intended functions satisfactorily
-  \citep{boehm2007software}.
-\end{defn}
-
-\begin{defn}
-  A concern encompassing correctness and robustness \citep{meyer1988object}.
-\end{defn}
-
-\begin{defn}
-  ``The extent to which a program can be expected to perform its intended
-  function with required precision'' \citep{McCallEtAl1977}. (As summarized in
+	\citep{boehm1976quantitative}.
+\end{defn}
+
+\begin{defn} \label{reliabilitySelected2}
+   Reliability expresses the ability of the software to maintain a specified
+   level of fault tolerance, when used under specified condition \citep{singh2013different}.
+\end{defn}
+\begin{defn}
+	The extent to which a program can be expected to perform its intended
+	function with required precision \citep{McCallEtAl1977}. (As summarized in
     \citet{VanVliet2000}.)
@@
 
-\begin{defn}
-``Informally, software is reliable if the user can depend on it''
-\citep{ghezzi1991fundamentals}.
-\end{defn}
-
-\noindent \textbf{Proposed Definition}
-
-\wss{Needs to be completed.}
-
-\noindent \textbf{Reasoning}
-
+\noindent \textbf{Proposed Definition}
+Definition \ref{reliabilitySelected1} and \ref{reliabilitySelected2} rephrased : 	Probability of failure-free operation of a computer program in a specified environment for a specified time at a specified level of fault tolerance.
+
+\wss{Needs to be completed.}\oo{Done}
+
+\noindent \textbf{Reasoning}
+Reliability is defined in a way that the measurement metrics are visible - the specified enviroment, the specified time and the specified level of fault tolerance.
 Reliability is a critical quality for scientific software, since the results of
@@
 	The quality can be further informally refined as the ability of a software to
-  keep an acceptable behaviour, expressed in terms of robustness requirements, in
+	keep an acceptable behaviour, expressed in terms of robustness requirements,
+	in
 	spite of exceptional or unforeseen execution conditions (such as the
@@
 
-Definition \ref{RobustnessPartialDefn} and Definition \ref{RobustnessDefnSelected} rephrased: Software possesses the
+Definition \ref{RobustnessPartialDefn} and Definition
+\ref{RobustnessDefnSelected} rephrased: Software possesses the
 characteristic of robustness if it behaves ``reasonably'' in two situations: i)
@@
 
-This definition offers a comprehensive list of constraints that are commonly associated with software performance, such as speed, throughput, capacity, and timing.
+This definition offers a comprehensive list of constraints that are commonly
+associated with software performance, such as speed, throughput, capacity, and
+timing.
 
@@
 \pmi{Wiegers did not discuss this difference.}
-\pmi{Speed - change over time; timing - related to specification: when something should happen in relation to something else - state transitions}
+\pmi{Speed - change over time; timing - related to specification: when something
+	should happen in relation to something else - state transitions}
 
@@
 
-Nielsen and (separately) Schneidermann have defined usability as part of usefulness and
+Nielsen and (separately) Schneidermann have defined usability as part of
+usefulness and
 is composed of:
@@
 	first time they encounter the design?
-\item Efficiency: Once users have learned the design, how quickly can they perform tasks?
+	\item Efficiency: Once users have learned the design, how quickly can they
+	perform tasks?
 	\item Memorability: When users return to the design after a period of not using
@@
 In that context, it makes sense to separate \emph{usefulness} into
-\emph{usability} (purely an interface concern) and \emph{utility} (in the economics
+\emph{usability} (purely an interface concern) and \emph{utility} (in the
+economics
 sense of the word).
@@
 
-The Interaction Design Foundation~\url{https://www.interaction-design.org/literature/topics/usability}
+The Interaction Design
+Foundation~\url{https://www.interaction-design.org/literature/topics/usability}
 further lists the following desirable outcomes:
@@
 \begin{enumerate}
-\item It should be easy for the user to become familiar with and competent in using
+	\item It should be easy for the user to become familiar with and competent in
+	using
 	the user interface during the first contact with the website. For example, if a
@@
 		reason for dropping the word operational?}
-\pmi{Yes, \citet{pressman2005software} does cite the relevant section of the textbook with \citet{McCallEtAl1977}.
-While they do not give a reason for specifically dropping the word operational, they do indicate that their given definition ``is a very limited definition''.
-Perhaps their intent was breadth.}
+	\pmi{Yes, \citet{pressman2005software} does cite the relevant section of the
+		textbook with \citet{McCallEtAl1977}. While they do not give a reason for
+		specifically dropping the word operational, they do indicate that their given
+		definition ``is a very limited definition''. Perhaps their intent was breadth.}
 \end{defn}
@@
 
-This definition offers a comprehensive list of potential reasons for modifying software, such as correcting faults, improving performance or other attributes, or satisfying new requirements.
+This definition offers a comprehensive list of potential reasons for modifying
+software, such as correcting faults, improving performance or other attributes,
+or satisfying new requirements.
 
@@
 	same as a quality?  We are calling performance a quality in this document.}
-\pmi{The term quality attribute has come up often in the literature - qualities appear to be a subset of attributes.
-Quality attributes seem to be associated with non-functional requirements.
-The term resource attribute was encountered in a journal and implied an association with a functional requirement.
-The two might differ along those lines of NFR and FR.}
+\pmi{The term quality attribute has come up often in the literature - qualities
+	appear to be a subset of attributes. Quality attributes seem to be associated
+	with non-functional requirements. The term resource attribute was encountered in
+	a journal and implied an association with a functional requirement. The two
+	might differ along those lines of NFR and FR.}
 
@@
 
-Understandability is artifact-dependent. What it means for a user-interface (graphical
-or otherwise) to be understandable is wildly different than what it means for the code,
+Understandability is artifact-dependent. What it means for a user-interface
+(graphical
+or otherwise) to be understandable is wildly different than what it means for
+the code,
 and even the user documentation.
 
-The literature here is thin and scattered.  More work will need to be done to find
+The literature here is thin and scattered.  More work will need to be done to
+find
 something useful.
 
-Interestingly, the business literature seems to have taken more care to define this.
+Interestingly, the business literature seems to have taken more care to define
+this.
 Here we encounter
@@
 \end{quote}
-At least this brings in the idea that the \emph{reader} is actively involved, and
+At least this brings in the idea that the \emph{reader} is actively involved,
+and
 indirectly that the reader's knowledge may be relevant, as well as the
@@
 
-Definition \ref{VisibilitySelected} points out that documentation is the way to improve visibility. It is
+Definition \ref{VisibilitySelected} points out that documentation is the way to
+improve visibility. It is
 rephrased because the original one might refer to binary status - ``visible'' or
@@
 \begin{defn}
-\noindent The ability to meet the needs of the present without compromising the ability of
+	\noindent The ability to meet the needs of the present without compromising the
+	ability of
 	future generations to meet their own needs \citep{Brundtland1987}.
@@
 \wss{Were there any other definitions of completeness?  You could add the
-  definition of completeness from \citet[p.\ 5--6]{IEEE1998}.  This definition is for
+	definition of completeness from \citet[p.\ 5--6]{IEEE1998}.  This definition
+	is for
 	requirements, but maybe there is something we can generalize from the
-  definition?  We could also look for definitions outside of software development.}
+	definition?  We could also look for definitions outside of software
+	development.}
 
@@
 
-24765-2017 - ISO/IEC/IEEE International Standard - Systems and software engineering--Vocabulary
+24765-2017 - ISO/IEC/IEEE International Standard - Systems and software
+engineering--Vocabulary
 has a full definition, namely
@@
 	\item According to
+
 	\href{https://en.wikipedia.org/wiki/Software_verification_and_validation}{Capability
@@
 		whether the products of a given development phase satisfy the conditions
-    imposed at the start of that phase. [IEEE-STD-610] \wss{Need a proper citation}
+		imposed at the start of that phase. [IEEE-STD-610] \wss{Need a proper
+			citation}
 		\item Software Validation: The process of evaluating software during or at the
@@
 \begin{defn}
-``An abstraction for a software artifact is a succinct description that suppresses
+	``An abstraction for a software artifact is a succinct description that
+	suppresses
 	the details that are unimportant to a software developer and emphasizes the

### 2019-12-03 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
Done with Reliability and Correctness

@@
 \begin{defn}
-	Installability is related to the amount of human effort required to install a
-	software in a designated environment \citep{iso2001iec}
-		{ berander2005software}.
-\end{defn}
-
-\begin{defn}
 	The capability of a software product to be installed in a specified
-	environment \citep{berander2005software}.  \wss{remove the quotation marks -
+	environment \citep{iso2001iec}. (As summarized in \citep{berander2005software}.)  \wss{remove the quotation marks -
 	we are assuming all definitions are directly quoted from the source. Please
 	remove quotes for all of the definitions that you ``own.''}
-\end{defn}
-
-\begin{defn} \label{InstallabilityDefnSelected1}
-	``The efficacy of the installation, uninstallation or reinstallation process
-	of a software product in a specified environment'' \citep{iso201017043}
-\end{defn}
-
-\begin{defn}
+    \oo{Done} 
+\end{defn}
+
+\begin{defn} \label{InstallabilityDefnSelected2}
 	The degree of effectiveness and efficiency with which a product or system can
@@
 
-\begin{defn} \label{InstallabilityDefnSelected2}
-	Installability refers to the cost or effort required for the installation of a
-	software or product, given an installation is possible to begin with
-	\citep{lenhard2013measuring}
+\begin{defn} \label{InstallabilityDefnSelected1}
+	Installability refers to the cost or effort required for the installation, given an installation is possible to begin with
+	\citep{lenhard2013measuring}.
 \end{defn}
@@
 \wss{The rationale is missing.  Please add.}
-\oo{The definition allows to measure the cost both from system and human view
-	points}
-
+The definition captures the cost or effort - human or machine, and considers not just the initial installation but uninstallation and re-installation which are all related to installation activities.
+\oo{Done installability}
 \subsection{\notdone{Correctness} \oo{owner}}
@@
 \begin{defn}  
-	The degree to which a software's specification is satisfied
-	\citep{berander2005software}.
-\end{defn}
-
-\begin{defn} \label{CorrectDefnSelected} 
-	A software is correct if it functions according to its specified functional
-	and non-functional specification \citep{GhezziEtAl2003}.  \wss{This is not how
+	A program is functionally correct if it behaves according to its stated
+   functional specifications \citep{GhezziEtAl2003}.  \wss{This is not how
 		\citet{GhezziEtAl2003} defines correctness.  They only require satisfaction
-		of the functional requirements.}
-
-\end{defn}
-
-\begin{defn}
-	According to \citet{wilson2009quality}, ``the degree to which a system is free
+		of the functional requirements.} \oo{Updated}
+
+\end{defn}
+
+\begin{defn}
+	According to \citet{wilson2009quality}, the degree to which a system is free
 	from defects in its specification, design, and implementation, is its
-	correctness.''
-\end{defn}
-
-\begin{defn}
-	``The ability of software products to perform their exact tasks, as defined by
-	their specification'' \citep{meyer1988object}.
-\end{defn}
-
-\begin{defn}
-	``The extent to which a program satisfies its specifications and fulfills the
-	user's mission objectives'' \citep{McCallEtAl1977}. (As summarized in
+	correctness.
+\end{defn}
+
+\begin{defn}
+	The ability of software products to perform their exact tasks, as defined by
+	their specification \citep{meyer1988object}.
+\end{defn}
+
+\begin{defn} \label{CorrectnessSelected}
+	The extent to which a program satisfies its specifications and fulfills the
+	user's mission objectives \citep{McCallEtAl1977}. (As summarized in
 	\citet{VanVliet2000}.)
@@
 \begin{defn}
-	``A program is functionally correct if it behaves according to its stated
-	functional specifications'' \citep{ghezzi1991fundamentals}.
-\end{defn}
-
-\noindent \textbf{Proposed Definition}
-
-Definition~\ref{CorrectDefnSelected}
-
-\noindent \textbf{Reasoning}
-
+ The term correctness is often mentioned as a degree to which software meets the requirement specification \citep{IEEEStdGlossarySET1990}.
+\end{defn}
+
+\begin{defn}
+	Correctness from software engineering perspective can be defined as the adherence to the specifications that determine how users can interact with the software and how the software should behave when it is used correctly \citep{Tutorialspoint}.
+\end{defn}
+
+\noindent \textbf{Proposed Definition}
+
+Definition~\ref{CorrectnessSelected} rephrased: The extent to which a software product satisfies its functional and non-functional specifications and fulfills the
+user's mission objectives.
+
+\noindent \textbf{Reasoning}
+To determine whether a program is correct, we must be able to refer to both its functional and non-functional requirements to understand the intended purpose and based on defined metrics, determine the extent to which this is met.
+Most definitions of correctness refers only to functional specifications but this is rephrased to include the non-functionals.
 There is no direct tool or method for measuring correctness. One way of building
@@
 dependencies, such that if a system exhibits a high degree of correctness then
-it tends to be reliable \citep{GhezziEtAl2003}.
-
+it tends to be reliable \citep{GhezziEtAl2003}. \oo{Done Correctness}
 \subsection{\notdone{Verifiability} \oo{owner}}
@@
 	A set of attributes that relate to the capability of software to maintain its
-	level of performance under stated conditions for a stated period of time \citep{international2001iso}. (As summarized in \citep{berander2005software}.)
+	level of performance under stated conditions for a stated period of time \citep{iso2001iec}. (As summarized in \citet{berander2005software}.)
 \end{defn}
@@
 \begin{defn} \label{reliabilitySelected1}
-	Probability of failure-free operation of a computer program in a specified environment for a specified time \citep{musa1987software}. (As stated in \citep{pressman2005software}.)
+	Probability of failure-free operation of a computer program in a specified environment for a specified time \citep{musa1987software}. (As stated in \citet{pressman2005software}.)
 \end{defn}
@@
     \citet{VanVliet2000}.)
+\end{defn}
+
+\begin{defn}
+	Informally, the reliability of a system is the probability, over a given
+	period of time, that the system will correctly deliver services as expected by
+	the user \citep{sommerville}
 \end{defn}

### 2019-12-03 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
Completed Verifiability

@@
 
-\begin{defn} \label{InstallabilityDefnSelected2}
+\begin{defn} \label{Installability_Selected2}
 	The degree of effectiveness and efficiency with which a product or system can
@@
 
-\begin{defn} \label{InstallabilityDefnSelected1}
+\begin{defn} \label{Installability_Selected1}
 	Installability refers to the cost or effort required for the installation, given an installation is possible to begin with
@@
 
-Definition~ Combined definitions \ref{InstallabilityDefnSelected1} and
-\ref{InstallabilityDefnSelected2} rephrased.
+Combined definitions \ref{Installability_Selected1} and
+\ref{Installability_Selected2} rephrased.
 Installability refers to the cost or effort required for the installation,
@@
 
-\wss{The rationale is missing.  Please add.}
+\wss{The rationale is missing.  Please add.} \oo{Done}
 The definition captures the cost or effort - human or machine, and considers not just the initial installation but uninstallation and re-installation which are all related to installation activities.
@@
 
-\begin{defn} \label{CorrectnessSelected}
+\begin{defn} 
 	The extent to which a program satisfies its specifications and fulfills the
@@
 
-\begin{defn}
- The term correctness is often mentioned as a degree to which software meets the requirement specification \citep{IEEEStdGlossarySET1990}.
+\begin{defn} \label{Correctness_Selected}
+ The degree to which a system or component is free from faults in its specification, design and implementation \citep{IEEEComputerDictionary1991}.
 \end{defn}
@@
 
-Definition~\ref{CorrectnessSelected} rephrased: The extent to which a software product satisfies its functional and non-functional specifications and fulfills the
-user's mission objectives.
-
-\noindent \textbf{Reasoning}
-To determine whether a program is correct, we must be able to refer to both its functional and non-functional requirements to understand the intended purpose and based on defined metrics, determine the extent to which this is met.
-Most definitions of correctness refers only to functional specifications but this is rephrased to include the non-functionals.
+Definition~\ref{Correctness_Selected}: The degree to which a system or component is free from faults in its specification, design and implementation.
+
+\noindent \textbf{Reasoning}
+
+To determine whether a program is correct, we must be able to refer to both its its specification, design and implementation to understand the intended purpose and based on defined metrics, determine the extent to which this is met.
+Most definitions of correctness refers only to functional specifications but this definition is broad and is able to capture both functional and non-functional specification without explicitly specifying it.\oo{We could modify te definition to explicitly state functional and non-functional if we decide.}
 There is no direct tool or method for measuring correctness. One way of building
@@
 it tends to be reliable \citep{GhezziEtAl2003}. \oo{Done Correctness}
-\subsection{\notdone{Verifiability} \oo{owner}}
-
-\begin{defn}
-	A software is verifiable if its properties can be verified
+
+\subsection{\notdone{Verifiability/Testability} \oo{owner}}
+
+\begin{defn}
+	A software is verifiable if its properties can be verified easily
 	\citep{GhezziEtAl2003}.
@@
 \begin{defn}
-	Attributes of software that relate to the effort needed for validating the
-	modified software \citep{berander2005software}.
-\end{defn}
-
-\begin{defn}
-	The degree to which a system or component facilitates the establishment of
-	test criteria and the performance of tests to determine whether those criteria
-	have been met \citep{IEEEStdGlossarySET1990}. \wss{The author needs to be
-		fixed so that this displays properly.}
+	Attributes of software that relate to the effort needed for validating the modified software \citep{iso2001iec}.
+\end{defn}
+
+\begin{defn}
+	Effort required to test a program to ensure that it performs its intended
+	function \citep{McCallEtAl1977}. (As summarized in \citet{pressman2005software})
+\end{defn}
+
+\begin{defn}
+    Testability is the degree to which a requirement is stated in terms that permits establishment of test criteria and performance of tests to determine whether those criteria have been met \citep{IEEEStdGlossarySET1990}
+\end{defn}
+\begin{defn}
+	Code possesses the characteristic testability to the extent that it facilitates the establishment of verification criteria and supports evaluation of its performance \citep{boehm1976quantitative}.
 \end{defn}
@@
 \begin{defn} \label{Defn_Verifiability1}
-	The degree to which a requirement is stated in terms that permit establishment
-	of test criteria and performance \citep{IEEEStdGlossarySET1990}.
+	Testability is the degree to which a system or component facilitates the establishment of test criteria and the performance of tests to determine whether those criteria have been met \citep{ISO/IEC/IEEE24765}.
 \end{defn}
@@
 \begin{defn} \label{Defn_Verifiability2}
-	Ease of performing testing on a software product or system
-	\citep{IEEEStdGlossarySET1990}.
-\end{defn}
-
-\noindent \textbf{Proposed Definition}
-
-Combo of definitions~\ref{Defn_Verifiability1} and~\ref{Defn_Verifiability2} ?
-
-\noindent \textbf{Reasoning}
-
-Verifiability involves ``solving the equations right''
+	Testability is the effort required to test a software \citep{ISO/IEC/IEEE24765}.
+\end{defn}
+
+\noindent \textbf{Proposed Definition}
+
+Combined definitions~\ref{Defn_Verifiability1} and~\ref{Defn_Verifiability2} rephrased: Testability or verifiability is the degree to which a system or component facilitates the establishment of test criteria and the performance of tests to determine whether those criteria have been met and the effort required to test.
+
+\noindent \textbf{Reasoning}
+Definition is concise and measurable.
+Verifiability involves solving the equations right
 \citep[p.~23]{Roache1998};
@@
 \oo{Verifiability is sometimes referred to as testability, so I culled some
-	testability definitions here}  \wss{culled means removed - is that your
-	intention?}
+	testability definitions and added here}  \wss{culled means removed - is that your
+	intention?} \oo{culled could also mean  'to choose', so I meant, I choose from some definitions of testability and added to the list}
 
@@
 \begin{defn}
-	``The effort required to test a program to ensure that it performs its
-	intended function'' \citep{McCallEtAl1977}. (As summarized in
-	\citet{VanVliet2000}.)
+	The effort required to test a program to ensure that it performs its
+	intended function \citep{McCallEtAl1977}. (As summarized in
+	\citet{VanVliet2000}.) \oo{similar to \citep{pressman2005software}, this}
 \end{defn}
@@
 	the user \citep{sommerville}
+\end{defn}
+
+\begin{defn}
+	The ability of a system or component to to perform its required function under stated conditions for a specified period of time \citep{IEEEComputerDictionary1991}
 \end{defn}

### 2019-12-03 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
updated validatability

@@
 	Validatability of a software is the degree of ease in validating (checking)
-	that software meets user needs. \wss{Need a citation}
-\end{defn}
-
-\begin{defn}
-	Validatability means ``solving the right equations''
+	that software meets user needs. \wss{Need a citation} \oo{To be removed, I don't have any citation, definition was modified from the definition of verifiability}
+\end{defn}
+
+\begin{defn}
+	Validatability means solving the right equations
 	\citep[p.~23]{Roache1998}.
-	\oo{question about this definition}
-\end{defn}
-
-\begin{defn} \label{DefnValidatability}
-	The degree of ease of evaluating software products or system to determine
-	whether it satisfies specified business
-	requirements.
-	\url{http://softwaretestingfundamentals.com/verification-vs-validation/}
-	\oo{Here I tweaked the meaning of validation to suit validatability} \oo{Only
-		a few resource on validatability}
-\end{defn}
-
-\noindent \textbf{Proposed Definition}
-
-Definition~\ref{DefnValidatability}.
+	\oo{Validation?}
+\end{defn}
+
+\begin{defn} \label{Validatability_Selected}
+   Validation is the process of evaluating software during or at the end of the development
+   process to determine whether it satisfies specified business requirements \citep{softwaretestingfundamentals}.
+
+\end{defn}
+
+\begin{defn}
+	Validation refers to a different set
+	of tasks that ensure that the software that has been built is traceable to customer requirements.
+\end{defn}
+\noindent \textbf{Proposed Definition}
+
+\begin{defn}
+	Requirements validation examines the specification to ensure that all
+	software requirements have been stated unambiguously; that inconsistencies, omissions, and errors have been detected and corrected; and that the work products conform to the standards established for the process, the project, and the product \citep{pressman2005software}.
+\end{defn}
+
+\begin{defn}
+    Software validation checks that the software product satisfies or fits the intended use (high-level checking), i.e., the software meets the user requirements, not as specification artifacts or as needs of those who will operate the software only; but, as the needs of all the stakeholders (such as users, operators, administrators, managers, investors, etc.) \citep{WikipediaVnV}
+\end{defn}
+\noindent \textbf{Proposed Definition}: Definition~\ref{} \oo{need to discuss this quality, we might just define validatability from validation definitions above?
+Resources are limited for 'validatability'}.
 
@@
 required for experimental validation.
+It is entirely possible that a product passes when verified but fails when validated.
+This can happen when, say, a product is built as per the specifications but the specifications themselves fail to address the user’s needs
 

### 2019-12-04 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
Reformatted tex file for consistency, 80 character width

@@
     does this definition come from? Please add a citation.  Is it from
-	\citet{McCallEtAl1977}? I thought so originally, but after looking into it, I
-	don't think so anymore.}
-	\oo{@smith you added the definition in here but I guess it's from here}
+    \citet{McCallEtAl1977}? I thought so originally, but after looking into it,
+    I don't think so anymore.}  \oo{@smith you added the definition in here but
+    I guess it's from here} \wss{@olu, I was hoping you would verify the source
+    of the definition.  Can you please find a pdf of the McCall reference and
+    add it to our pub repo.  We should cite the primary source,
+    but we can't assume what it says.}
 \end{defn}
@@
   The capability of a software product to be installed in a specified
-	environment \citep{iso2001iec}. (As summarized in \citep{berander2005software}.)  \wss{remove the quotation marks -
-	we are assuming all definitions are directly quoted from the source. Please
-	remove quotes for all of the definitions that you ``own.''}
-    \oo{Done} 
+  environment \citep{iso2001iec}. (As summarized in
+  \citep{berander2005software}.)  \wss{remove the quotation marks - we are
+    assuming all definitions are directly quoted from the source. Please remove
+    quotes for all of the definitions that you ``own.''}  \oo{Done}
 \end{defn}
@@
 
-\begin{defn} \label{Installability_Selected1}
-	Installability refers to the cost or effort required for the installation, given an installation is possible to begin with
-	\citep{lenhard2013measuring}.
+\begin{defn} \label{Installability_Selected1} Installability refers to the cost
+  or effort required for the installation, given an installation is possible to
+  begin with \citep{lenhard2013measuring}.
 \end{defn}
@@
 
-\wss{The rationale is missing.  Please add.} \oo{Done}
-The definition captures the cost or effort - human or machine, and considers not just the initial installation but uninstallation and re-installation which are all related to installation activities.
-\oo{Done installability}
+\wss{The rationale is missing.  Please add.} \oo{Done} The definition captures
+the cost or effort - human or machine, and considers not just the initial
+installation but uninstallation and re-installation which are all related to
+installation activities.  \oo{Done installability}
+
 \subsection{\notdone{Correctness} \oo{owner}}
@@
 \begin{defn} \label{Correctness_Selected} 
- The degree to which a system or component is free from faults in its specification, design and implementation \citep{IEEEComputerDictionary1991}.
-\end{defn}
-
-\begin{defn}
-	Correctness from software engineering perspective can be defined as the adherence to the specifications that determine how users can interact with the software and how the software should behave when it is used correctly \citep{Tutorialspoint}.
-\end{defn}
-
-\noindent \textbf{Proposed Definition}
-
-Definition~\ref{Correctness_Selected}: The degree to which a system or component is free from faults in its specification, design and implementation.
-
-\noindent \textbf{Reasoning}
-
-To determine whether a program is correct, we must be able to refer to both its its specification, design and implementation to understand the intended purpose and based on defined metrics, determine the extent to which this is met.
-Most definitions of correctness refers only to functional specifications but this definition is broad and is able to capture both functional and non-functional specification without explicitly specifying it.\oo{We could modify te definition to explicitly state functional and non-functional if we decide.}
-There is no direct tool or method for measuring correctness. One way of building
-confidence in correctness is by reviewing to ensure that each requirement stated
-is one that the stakeholders and experts desire.  By maintaining traceability,
+  The degree to which a system or component is free from faults in its
+  specification, design and implementation \citep{IEEEComputerDictionary1991}.
+\end{defn}
+
+\begin{defn}
+  Correctness from software engineering perspective can be defined as the
+  adherence to the specifications that determine how users can interact with the
+  software and how the software should behave when it is used correctly
+  \citep{Tutorialspoint}.
+\end{defn}
+
+\noindent \textbf{Proposed Definition}
+
+Definition~\ref{Correctness_Selected}: The degree to which a system or component
+is free from faults in its specification, design and implementation.
+
+\noindent \textbf{Reasoning}
+
+To determine whether a program is correct, we must be able to refer to both its
+its specification, design and implementation to understand the intended purpose
+and based on defined metrics, determine the extent to which this is met. Most
+definitions of correctness refers only to functional specifications but this
+definition is broad and is able to capture both functional and non-functional
+specification without explicitly specifying it.\oo{We could modify te definition
+  to explicitly state functional and non-functional if we decide.}  There is no
+direct tool or method for measuring correctness. One way of building confidence
+in correctness is by reviewing to ensure that each requirement stated is one
+that the stakeholders and experts desire.  By maintaining traceability,
 consistency and unambiguity, we can reduce the occurrence of errors and make the
@@
 \begin{defn}
-	Attributes of software that relate to the effort needed for validating the modified software \citep{iso2001iec}.
+  Attributes of software that relate to the effort needed for validating the
+  modified software \citep{iso2001iec}.
 \end{defn}
@@
   Effort required to test a program to ensure that it performs its intended
-	function \citep{McCallEtAl1977}. (As summarized in \citet{pressman2005software})
-\end{defn}
-
-\begin{defn}
-    Testability is the degree to which a requirement is stated in terms that permits establishment of test criteria and performance of tests to determine whether those criteria have been met \citep{IEEEStdGlossarySET1990}
-\end{defn}
-\begin{defn}
-	Code possesses the characteristic testability to the extent that it facilitates the establishment of verification criteria and supports evaluation of its performance \citep{boehm1976quantitative}.
+  function \citep{McCallEtAl1977}. (As summarized in
+  \citet{pressman2005software})
+\end{defn}
+
+\begin{defn}
+  Testability is the degree to which a requirement is stated in terms that
+  permits establishment of test criteria and performance of tests to determine
+  whether those criteria have been met \citep{IEEEStdGlossarySET1990}
+\end{defn}
+
+\begin{defn}
+  Code possesses the characteristic testability to the extent that it
+  facilitates the establishment of verification criteria and supports evaluation
+  of its performance \citep{boehm1976quantitative}.
 \end{defn}
@@
 \begin{defn} \label{Defn_Verifiability1} 
-	Testability is the degree to which a system or component facilitates the establishment of test criteria and the performance of tests to determine whether those criteria have been met \citep{ISO/IEC/IEEE24765}.
+  Testability is the degree to which a system or component facilitates the
+  establishment of test criteria and the performance of tests to determine
+  whether those criteria have been met \citep{ISO/IEC/IEEE24765}.
 \end{defn}
@@
 \begin{defn} \label{Defn_Verifiability2} 
-	Testability is the effort required to test a software \citep{ISO/IEC/IEEE24765}.
-\end{defn}
-
-\noindent \textbf{Proposed Definition}
-
-Combined definitions~\ref{Defn_Verifiability1} and~\ref{Defn_Verifiability2} rephrased: Testability or verifiability is the degree to which a system or component facilitates the establishment of test criteria and the performance of tests to determine whether those criteria have been met and the effort required to test.
-
-\noindent \textbf{Reasoning}
-Definition is concise and measurable.
-Verifiability involves solving the equations right
-\citep[p.~23]{Roache1998};
-it benefits from rational documentation that systematically shows, with explicit
-traceability, how the governing equations are transformed into code.
-\oo{Verifiability is sometimes referred to as testability, so I culled some
-	testability definitions and added here}  \wss{culled means removed - is that your
-	intention?} \oo{culled could also mean  'to choose', so I meant, I choose from some definitions of testability and added to the list}
+  Testability is the effort required to test a software
+  \citep{ISO/IEC/IEEE24765}.
+\end{defn}
+
+\noindent \textbf{Proposed Definition}
+
+Combined definitions~\ref{Defn_Verifiability1} and~\ref{Defn_Verifiability2}
+rephrased: Testability or verifiability is the degree to which a system or
+component facilitates the establishment of test criteria and the performance of
+tests to determine whether those criteria have been met and the effort required
+to test.
+
+\noindent \textbf{Reasoning}
+
+Definition is concise and measurable.  Verifiability involves solving the
+equations right \citep[p.~23]{Roache1998}; it benefits from rational
+documentation that systematically shows, with explicit traceability, how the
+governing equations are transformed into code.  \oo{Verifiability is sometimes
+  referred to as testability, so I culled some testability definitions and added
+  here} \wss{culled means removed - is that your intention?} \oo{culled could
+  also mean 'to choose', so I meant, I choose from some definitions of
+  testability and added to the list}
 
@@
   Validatability of a software is the degree of ease in validating (checking)
-	that software meets user needs. \wss{Need a citation} \oo{To be removed, I don't have any citation, definition was modified from the definition of verifiability}
-\end{defn}
-
-\begin{defn}
-	Validatability means solving the right equations
-	\citep[p.~23]{Roache1998}.
+  that software meets user needs. \wss{Need a citation} \oo{To be removed, I
+    don't have any citation, definition was modified from the definition of
+    verifiability}
+\end{defn}
+
+\begin{defn}
+  Validatability means solving the right equations \citep[p.~23]{Roache1998}.
   \oo{Validation?}
@@
 \begin{defn} \label{Validatability_Selected} 
-   Validation is the process of evaluating software during or at the end of the development
-   process to determine whether it satisfies specified business requirements \citep{softwaretestingfundamentals}.
-
-\end{defn}
-
-\begin{defn}
-	Validation refers to a different set
-	of tasks that ensure that the software that has been built is traceable to customer requirements.
-\end{defn}
-\noindent \textbf{Proposed Definition}
-
-\begin{defn}
-	Requirements validation examines the specification to ensure that all
-	software requirements have been stated unambiguously; that inconsistencies, omissions, and errors have been detected and corrected; and that the work products conform to the standards established for the process, the project, and the product \citep{pressman2005software}.
-\end{defn}
-
-\begin{defn}
-    Software validation checks that the software product satisfies or fits the intended use (high-level checking), i.e., the software meets the user requirements, not as specification artifacts or as needs of those who will operate the software only; but, as the needs of all the stakeholders (such as users, operators, administrators, managers, investors, etc.) \citep{WikipediaVnV}
-\end{defn}
-\noindent \textbf{Proposed Definition}: Definition~\ref{} \oo{need to discuss this quality, we might just define validatability from validation definitions above?
-Resources are limited for 'validatability'}.
+  Validation is the process of evaluating software during or at the end of the
+  development process to determine whether it satisfies specified business
+  requirements \citep{softwaretestingfundamentals}.
+\end{defn}
+
+\begin{defn}
+  Validation refers to a different set of tasks that ensure that the software
+  that has been built is traceable to customer requirements.
+\end{defn}
+\noindent \textbf{Proposed Definition}
+
+\begin{defn}
+  Requirements validation examines the specification to ensure that all software
+  requirements have been stated unambiguously; that inconsistencies, omissions,
+  and errors have been detected and corrected; and that the work products
+  conform to the standards established for the process, the project, and the
+  product \citep{pressman2005software}.
+\end{defn}
+
+\begin{defn}
+  Software validation checks that the software product satisfies or fits the
+  intended use (high-level checking), i.e., the software meets the user
+  requirements, not as specification artifacts or as needs of those who will
+  operate the software only; but, as the needs of all the stakeholders (such as
+  users, operators, administrators, managers, investors, etc.)
+  \citep{WikipediaVnV}
+\end{defn}
+
+\noindent \textbf{Proposed Definition}
+
+Definition~\ref{} \oo{need to discuss this quality, we might just define
+  validatability from validation definitions above? Resources are limited for
+  'validatability'}.
 
@@
 theory and assumptions, along with an explicit statement of the systematic steps
-required for experimental validation.
-It is entirely possible that a product passes when verified but fails when validated.
-This can happen when, say, a product is built as per the specifications but the specifications themselves fail to address the user’s needs
+required for experimental validation. It is entirely possible that a product
+passes when verified but fails when validated. This can happen when, say, a
+product is built as per the specifications but the specifications themselves
+fail to address the user’s needs
 
@@
   A set of attributes that relate to the capability of software to maintain its
-	level of performance under stated conditions for a stated period of time \citep{iso2001iec}. (As summarized in \citet{berander2005software}.)
-\end{defn}
-
-\begin{defn} \label{reliabilitySelected1}
-	Probability of failure-free operation of a computer program in a specified environment for a specified time \citep{musa1987software}. (As stated in \citet{pressman2005software}.)
+  level of performance under stated conditions for a stated period of time
+  \citep{iso2001iec}. (As summarized in \citet{berander2005software}.)
+\end{defn}
+
+\begin{defn} \label{reliabilitySelected1} Probability of failure-free operation
+  of a computer program in a specified environment for a specified time
+  \citep{musa1987software}. (As stated in \citet{pressman2005software}.)
 \end{defn}
@@
 \begin{defn}
-	The ability of a system or component to to perform its required function under stated conditions for a specified period of time \citep{IEEEComputerDictionary1991}
-\end{defn}
-
-\noindent \textbf{Proposed Definition}
-Definition \ref{reliabilitySelected1} and \ref{reliabilitySelected2} rephrased : 	Probability of failure-free operation of a computer program in a specified environment for a specified time at a specified level of fault tolerance.
+  The ability of a system or component to to perform its required function under
+  stated conditions for a specified period of time
+  \citep{IEEEComputerDictionary1991}
+\end{defn}
+
+\noindent \textbf{Proposed Definition}
+
+Definition \ref{reliabilitySelected1} and \ref{reliabilitySelected2} rephrased:
+Probability of failure-free operation of a computer program in a specified
+environment for a specified time at a specified level of fault tolerance.
 
@@
 \noindent \textbf{Reasoning}
-Reliability is defined in a way that the measurement metrics are visible - the specified enviroment, the specified time and the specified level of fault tolerance.
-Reliability is a critical quality for scientific software, since the results of
-computations are meaningless, if they are not dependable.  Reliability is
-closely tied to verifiability, since the key quality to verify is reliability,
-while the act of verification itself improves reliability.
+
+Reliability is defined in a way that the measurement metrics are visible - the
+specified enviroment, the specified time and the specified level of fault
+tolerance. Reliability is a critical quality for scientific software, since the
+results of computations are meaningless, if they are not dependable.
+Reliability is closely tied to verifiability, since the key quality to verify is
+reliability, while the act of verification itself improves reliability.
 
@@
 Definition \ref{RobustnessPartialDefn} and Definition
-\ref{RobustnessDefnSelected} rephrased: Software possesses the
-characteristic of robustness if it behaves ``reasonably'' in two situations: i)
-when it encounters circumstances not anticipated in the requirements
-specification; and ii) when the assumptions in its requirements specification
-are violated.
+\ref{RobustnessDefnSelected} rephrased: Software possesses the characteristic of
+robustness if it behaves ``reasonably'' in two situations: i) when it encounters
+circumstances not anticipated in the requirements specification; and ii) when
+the assumptions in its requirements specification are violated.
 
@@
 
-The Interaction Design
-Foundation~\url{https://www.interaction-design.org/literature/topics/usability}
-further lists the following desirable outcomes:
+The Interaction Design Foundation
+\url{https://www.interaction-design.org/literature/topics/usability} further
+lists the following desirable outcomes:
 
@@
 \item It should be easy for the user to become familiar with and competent in
-	using
-	the user interface during the first contact with the website. For example, if a
-	travel agent’s website is a well-designed one, the user should be able to move
-	through the sequence of actions to book a ticket quickly.
+  using the user interface during the first contact with the website. For
+  example, if a travel agent’s website is a well-designed one, the user should
+  be able to move through the sequence of actions to book a ticket quickly.
 \item It should be easy for users to achieve their objective through using the
@@
 \begin{defn}
-	``A software is usable - or user fridendly - if its human users find it easy to
-	use'' \citep{ghezzi1991fundamentals}.
+  ``A software is usable - or user fridendly - if its human users find it easy
+  to use'' \citep{ghezzi1991fundamentals}.
 \end{defn}
@@
 \end{defn}
+
 \begin{defn}
@@
   modifications (which may include corrections, improvements, or adaptations of
-	software to environmental changes and changes in the requirements and functional
-	specifications) \citep{pfleeger2006software}.
-\end{defn}
+  software to environmental changes and changes in the requirements and
+  functional specifications) \citep{pfleeger2006software}.
+\end{defn}
+
 \begin{defn}
@@
 \end{defn}
+
 \begin{defn} \label{MaintainabilityDefnSelected2} 
@@
     \citet{pressman2005software} cite \citet{McCallEtAl1977}?  Do they give a
-		reason for dropping the word operational?}
-	\pmi{Yes, \citet{pressman2005software} does cite the relevant section of the
-		textbook with \citet{McCallEtAl1977}. While they do not give a reason for
-		specifically dropping the word operational, they do indicate that their given
-		definition ``is a very limited definition''. Perhaps their intent was breadth.}
+    reason for dropping the word operational?}  \pmi{Yes,
+    \citet{pressman2005software} does cite the relevant section of the textbook
+    with \citet{McCallEtAl1977}. While they do not give a reason for
+    specifically dropping the word operational, they do indicate that their
+    given definition ``is a very limited definition''. Perhaps their intent was
+    breadth.}
 \end{defn}
@@
 \begin{defn} 
-	The degree to which a software module or other work product can be used in more
-	than one software system \citep{IEEEStdGlossarySET1990}. 
+  The degree to which a software module or other work product can be used in
+  more than one software system \citep{IEEEStdGlossarySET1990}.
 \end{defn}
@@
 \begin{defn} \label{ReusabilityDefnSelected} 
-	The extent to which a software component can be used with or without adaptation
-	in a problem solution other than the one for which it was originally developed
-	\citep{kalagiakos2003non}.
-\end{defn}
-\begin{defn}
-	Reusability is the likelihood a segment of source code that can be used again to
-	add new functionalities with slight or no modification \citep{sandhu2010survey}.
+  The extent to which a software component can be used with or without
+  adaptation in a problem solution other than the one for which it was
+  originally developed \citep{kalagiakos2003non}.
+\end{defn}
+\begin{defn}
+  Reusability is the likelihood a segment of source code that can be used again
+  to add new functionalities with slight or no modification
+  \citep{sandhu2010survey}.
 \end{defn}
@@
   The extent to which a program (or parts thereof) can be reused in other
-	applications \citep{McCallEtAl1977}. (As summarized in
-	\citet{VanVliet2000}.)
-\end{defn}
-
-\begin{defn}
-	A product is reusable if we can ``use the product - perhaps with minor changes -
-	to build another product'' \citep{ghezzi1991fundamentals}.
+  applications \citep{McCallEtAl1977}. (As summarized in \citet{VanVliet2000}.)
+\end{defn}
+
+\begin{defn}
+  A product is reusable if we can ``use the product - perhaps with minor changes
+  - to build another product'' \citep{ghezzi1991fundamentals}.
 \end{defn}
@@
 \begin{defn}
-	An application is portable across a class of environments to the degree that the
-	effort required to transport and adapt it to a new environment in the class is
-	less than the effort of redevelopment \citep{mooney1990strategies}.
+  An application is portable across a class of environments to the degree that
+  the effort required to transport and adapt it to a new environment in the
+  class is less than the effort of redevelopment \citep{mooney1990strategies}.
 \end{defn}
 \begin{defn} \label{PortabilityDefnSelected} 
-	Effort required to transfer the program from one hardware and/or software system
-	environment to another \citep{pressman2005software}.
-\end{defn}
-\begin{defn}
-	A set of attributes that bear on the ability of software to be transferred from
-	one environment to another (including the organizational, hardware, of software
-	environment) \citep{pfleeger2006software}.
+  Effort required to transfer the program from one hardware and/or software
+  system environment to another \citep{pressman2005software}.
+\end{defn}
+\begin{defn}
+  A set of attributes that bear on the ability of software to be transferred
+  from one environment to another (including the organizational, hardware, of
+  software environment) \citep{pfleeger2006software}.
 \end{defn}
@@
   operated easily and well on computer configurations other than its current
-	one. This implies that special function features, not easily available at other
-	facilities, are not used, that standard library functions and subroutines are
-	selected for universal applicability, and so on \citep{boehm2007software}.
+  one. This implies that special function features, not easily available at
+  other facilities, are not used, that standard library functions and
+  subroutines are selected for universal applicability, and so on
+  \citep{boehm2007software}.
 \end{defn}
@@
 Understandability is artifact-dependent. What it means for a user-interface
-(graphical
-or otherwise) to be understandable is wildly different than what it means for
-the code,
-and even the user documentation.
+(graphical or otherwise) to be understandable is wildly different than what it
+means for the code, and even the user documentation.
 
 The literature here is thin and scattered.  More work will need to be done to
-find
-something useful.
+find something useful.
 
 Interestingly, the business literature seems to have taken more care to define
-this.
-Here we encounter
+this.  Here we encounter
 \begin{quote}
-	Understandability is the concept that X should be presented
-	so that a reader can easily comprehend it.
+  Understandability is the concept that X should be presented so that a reader
+  can easily comprehend it.
 \end{quote}
 At least this brings in the idea that the \emph{reader} is actively involved,
-and
-indirectly that the reader's knowledge may be relevant, as well as the
+and indirectly that the reader's knowledge may be relevant, as well as the
 ``clarity of exposition'' of X.
@@
 \begin{defn}
-	Interoperability is the ability of two or more systems or components to exchange
-	information and to use the information that has been exchanged
+  Interoperability is the ability of two or more systems or components to
+  exchange information and to use the information that has been exchanged
   \citep{IEEEComputerDictionary1991}.
 \end{defn}
+
 \begin{defn}
@@
 \end{defn}
-\begin{defn}
-	The capability to communicate, execute programs, and transfer data among various
-	functional units in a manner that requires the user to have little or no
-	knowledge of the unique characteristics of those units
+
+\begin{defn}
+  The capability to communicate, execute programs, and transfer data among
+  various functional units in a manner that requires the user to have little or
+  no knowledge of the unique characteristics of those units
   \citep{ISO/IEC/IEEE24765}.
 \end{defn}
-\begin{defn}Interoperability is a characteristic of a product or system, whose
-	interfaces are completely understood, to work with other products or systems,
-	present or future, in either implementation or access, without any restrictions
+
+\begin{defn}
+  Interoperability is a characteristic of a product or system, whose interfaces
+  are completely understood, to work with other products or systems, present or
+  future, in either implementation or access, without any restrictions
   \citep{AFUL2019}.
 \end{defn}
+
 \begin{defn}
@@
   \item Foundational (Level 1) – establishes the inter-connectivity requirements
-		needed for one system or application to securely communicate data to and receive
-		data from another
+    needed for one system or application to securely communicate data to and
+    receive data from another
 
@@
   \item Semantic (Level 3) – provides for common underlying models and
-		codification of the data including the use of data elements with standardized
-		definitions from publicly available value sets and coding vocabularies,
-		providing shared understanding and meaning to the user
-
-		\item Organizational (Level 4) – includes governance, policy, social, legal and
-		organizational considerations to facilitate the secure, seamless and timely
-		communication and use of data both within and between organizations, entities
-		and individuals. These components enable shared consent, trust and integrated
-		end-user processes and workflows
+    codification of the data including the use of data elements with
+    standardized definitions from publicly available value sets and coding
+    vocabularies, providing shared understanding and meaning to the user
+
+  \item Organizational (Level 4) – includes governance, policy, social, legal
+    and organizational considerations to facilitate the secure, seamless and
+    timely communication and use of data both within and between organizations,
+    entities and individuals. These components enable shared consent, trust and
+    integrated end-user processes and workflows
   \end{itemize}
@@
 \begin{defn}
-	Visibility provides transparency into the development process. It is the ability
-	to see progress at any point and determine the distance to completion of a goal.
-	Visibility provides status of not only the progress of the project, but the
-	product itself \citep{GSA2019}.
+  Visibility provides transparency into the development process. It is the
+  ability to see progress at any point and determine the distance to completion
+  of a goal.  Visibility provides status of not only the progress of the
+  project, but the product itself \citep{GSA2019}.
 \end{defn}
@@
   Business process visibility, also called process visibility, is the ability to
-	accurately and completely view the processes, transactions and other activities
-	operating within an enterprise \citep{Rouse2013}.
+  accurately and completely view the processes, transactions and other
+  activities operating within an enterprise \citep{Rouse2013}.
 \end{defn}
@@
 Definition \ref{VisibilitySelected} points out that documentation is the way to
-improve visibility. It is
-rephrased because the original one might refer to binary status - ``visible'' or
-``unvisible''.
+improve visibility. It is rephrased because the original one might refer to
+binary status - ``visible'' or ``unvisible''.
 
@@
 
-Reproducibility is a required component of the scientific
-method \citep{Davison2012}.  Although QA has, ``a bad name among creative
-scientists and engineers'' \citep[p.~352]{Roache1998}, the community need to
-recognize that participating in QA management also improves reproducibility.
-Reproducibility, like QA, benefits from a consistent and repeatable computing
-environment, version control and separating code from
-configuration/parameters \citep{Davison2012}.
+Reproducibility is a required component of the scientific method
+\citep{Davison2012}.  Although QA has, ``a bad name among creative scientists
+and engineers'' \citep[p.~352]{Roache1998}, the community need to recognize that
+participating in QA management also improves reproducibility.  Reproducibility,
+like QA, benefits from a consistent and repeatable computing environment,
+version control and separating code from configuration/parameters
+\citep{Davison2012}.
 
@@
 \begin{defn}
-	A result is said to be reproducible if another researcher can take the original
-	code and input data, execute it, and re-obtain the same result (Peng, Dominici,
-	and Zeger, 2006), as cited in \citet{BenureauAndRougier2017}.
+  A result is said to be reproducible if another researcher can take the
+  original code and input data, execute it, and re-obtain the same result (Peng,
+  Dominici, and Zeger, 2006), as cited in \citet{BenureauAndRougier2017}.
 \end{defn}
@@
   Documentation achieves replicability if the description it provides of the
-	algorithms is sufficiently precise and complete for an independent researcher to
-	re-obtain the results it presents.  \citep{BenureauAndRougier2017}
+  algorithms is sufficiently precise and complete for an independent researcher
+  to re-obtain the results it presents.  \citep{BenureauAndRougier2017}
 \end{defn}
@@
   nontrivial but generally workable problem. Inputs to the software process
-	generally comprise labor, computers, supplies, and other support facilities and
-	equipment. Defining outputs. The big problem in defining software productivity
-	is defining outputs. Here we find a defining delivered source instructions (DSI)
-	or lines of code as the output of the software process is totally inadequate,
-	and they argue that there are a number of deficiencies in using DSI. However,
-	most organizations doing practical productivity measurement still use DSI as
-	their primary metric \citep{Boehm1987}. \wss{Is this a direct quote from
-		\citet{Boehm1987}?  The sentences seem incomplete?}
-\end{defn}
+  generally comprise labor, computers, supplies, and other support facilities
+  and equipment. Defining outputs. The big problem in defining software
+  productivity is defining outputs. Here we find a defining delivered source
+  instructions (DSI) or lines of code as the output of the software process is
+  totally inadequate, and they argue that there are a number of deficiencies in
+  using DSI. However, most organizations doing practical productivity
+  measurement still use DSI as their primary metric \citep{Boehm1987}. \wss{Is
+    this a direct quote from \citet{Boehm1987}?  The sentences seem incomplete?}
+\end{defn}
+
 \begin{defn}
@@
   Productivity is the amount of output (what is produced) per unit of input
-	used. If we can measure the size of the software product and the effort required
-	to develop the product, we have:
+  used. If we can measure the size of the software product and the effort
+  required to develop the product, we have:
   \begin{align}
@@
   \end{align}
-	Equation (1) assumes that size is the output of the software production process
-	and effort is the input to the process. This can be contrasted with the
-	viewpoint of software cost models where we use size as an independent variable
-	(i.e., an input) to predict effort which is treated as an output. Equation (1)
-	is simple to operationalize if we have a single dominant size measure, for
-	example, product size measured in lines of code \citep{Kitchenham2004}.
+  Equation (1) assumes that size is the output of the software production
+  process and effort is the input to the process. This can be contrasted with
+  the viewpoint of software cost models where we use size as an independent
+  variable (i.e., an input) to predict effort which is treated as an
+  output. Equation (1) is simple to operationalize if we have a single dominant
+  size measure, for example, product size measured in lines of code
+  \citep{Kitchenham2004}.
 \end{defn}
@@
 \begin{defn}
-	\noindent The ability to meet the needs of the present without compromising the
-	ability of
-	future generations to meet their own needs \citep{Brundtland1987}.
+  \noindent The ability to meet the needs of the present without compromising
+  the ability of future generations to meet their own needs
+  \citep{Brundtland1987}.
 \end{defn}
@@
 
-The fourth dimension of technical sustainability is also added
-by \citep{WolframEtAl2017}.  Technical sustainability is the focus on the thesis
-by \citet{Hygerth2016}.
+The fourth dimension of technical sustainability is also added by
+\citep{WolframEtAl2017}.  Technical sustainability is the focus on the thesis by
+\citet{Hygerth2016}.
 
@@
 \begin{defn}
-	The concept of sustainability is based on three pillars: the
-	ecological, the economical and the social. This means that for a software to
-	be sustainable, we must take all of its effects -- direct and indirect -- on
-	the environment, the economy and the society into account. In addition, the
-	entire life cycle of a software has to be considered: from planning and
-	conception to programming, distribution, installation, usage and
-	disposal \citep{Heine2017}.
+  The concept of sustainability is based on three pillars: the ecological, the
+  economical and the social. This means that for a software to be sustainable,
+  we must take all of its effects -- direct and indirect -- on the environment,
+  the economy and the society into account. In addition, the entire life cycle
+  of a software has to be considered: from planning and conception to
+  programming, distribution, installation, usage and disposal \citep{Heine2017}.
 \end{defn}
@@
   \label{CompletenessSelected}
-	A specification is complete to the extent that all of its parts are present and
-	each part is fully developed. A software specification must exhibit several
-	properties to assure its completeness \citep{Boehm1984}:
+  A specification is complete to the extent that all of its parts are present
+  and each part is fully developed. A software specification must exhibit
+  several properties to assure its completeness \citep{Boehm1984}:
   \begin{itemize}
@@
     specification.
-		\item No missing specification items. These are items that should be present as
-		part of the standard format of the specification, but are not present.
+  \item No missing specification items. These are items that should be present
+    as part of the standard format of the specification, but are not present.
   \item No missing functions. These are functions that should be part of the
@@
   definition of completeness from \citet[p.\ 5--6]{IEEE1998}.  This definition
-	is for
-	requirements, but maybe there is something we can generalize from the
+  is for requirements, but maybe there is something we can generalize from the
   definition?  We could also look for definitions outside of software
@@
   \label{ConsistencySelected}
-	A specification is consistent to the extent that its provisions do not conflict
-	with each other or with governing specifications and objectives. Specifications
-	require consistency in several ways \citep{Boehm1984}.
+  A specification is consistent to the extent that its provisions do not
+  conflict with each other or with governing specifications and
+  objectives. Specifications require consistency in several ways
+  \citep{Boehm1984}.
   \begin{itemize}
-		\item Internal consistency. Items within the specification do not conflict with
-		each other.
+  \item Internal consistency. Items within the specification do not conflict
+    with each other.
   \item External consistency. Items in the specification do not conflict with
     external specifications or entities.
-		\item Traceability. Items in the specification have clear antecedents in earlier
-		specifications or statements of system objectives.
+  \item Traceability. Items in the specification have clear antecedents in
+    earlier specifications or statements of system objectives.
   \end{itemize}
 \end{defn}
+
 \begin{defn}
@@
   Consistency: 1. the degree of uniformity, standardization, and freedom from
-	contradiction among the documents or parts of a system or component 2. software
-	attributes that provide uniform design and implementation techniques and
-	notations \citep{ISO/IEC/IEEE24765}.
+  contradiction among the documents or parts of a system or component
+  2. software attributes that provide uniform design and implementation
+  techniques and notations \citep{ISO/IEC/IEEE24765}.
 \end{defn}
@@
 \end{defn}
-By specializing the above to software artifacts, ``interrelate'' to 
-``why is this here'' (for forward tracing from requirements), this does
-indeed give what is meant in SE.
+By specializing the above to software artifacts, ``interrelate'' to ``why is
+this here'' (for forward tracing from requirements), this does indeed give what
+is meant in SE.
 
@@
   predecessor-successor or master-subordinate relationship to one another;
-	\item
-	the identification and documentation of derivation paths (upward) and
+\item the identification and documentation of derivation paths (upward) and
   allocation or flowdown paths (downward) of work products in the work product
   hierarchy;
-	\item the degree to which each element in a software development
-	product establishes its reason for existing; and discernible association
-	among two or more logical entities, such as requirements, system elements,
-	verifications, or tasks.
+\item the degree to which each element in a software development product
+  establishes its reason for existing; and discernible association among two or
+  more logical entities, such as requirements, system elements, verifications,
+  or tasks.
 \end{enumerate}
@@
   \begin{itemize}
-		\item 
-		Software Verification: The process of evaluating software to determine
+  \item Software Verification: The process of evaluating software to determine
     whether the products of a given development phase satisfy the conditions
@@
 
-\wss{When I get the Ghezzi text back from Olu,
-	I'll check to see if they have anything to add to this definition.}
+\wss{When I get the Ghezzi text back from Olu, I'll check to see if they have
+  anything to add to this definition.}
 
@@
 \begin{defn}
-	Documented requirements are said to be abstract if they state what the software
-	must do and the properties it must possess, but do not speak about how these are
-	to be achieved \citep{GhezziEtAl2003}.  
+  Documented requirements are said to be abstract if they state what the
+  software must do and the properties it must possess, but do not speak about
+  how these are to be achieved \citep{GhezziEtAl2003}.
 \end{defn}
@@
   ``An abstraction for a software artifact is a succinct description that
-	suppresses
-	the details that are unimportant to a software developer and emphasizes the
-	information that is important.'' \citep{Krueger1992}
+  suppresses the details that are unimportant to a software developer and
+  emphasizes the information that is important.'' \citep{Krueger1992}
 \end{defn}
@@
 \begin{defn}
-	``Abstraction in mathematics is the process of extracting the underlying essence
-	of a mathematical concept, removing any dependence on real world objects with
-	which it might originally have been connected, and generalizing it so that it
-	has wider applications or matching among other abstract descriptions of
-	equivalent phenomena.''
+  ``Abstraction in mathematics is the process of extracting the underlying
+  essence of a mathematical concept, removing any dependence on real world
+  objects with which it might originally have been connected, and generalizing
+  it so that it has wider applications or matching among other abstract
+  descriptions of equivalent phenomena.''
   \href{https://en.wikipedia.org/wiki/Abstraction_(mathematics)} {Wikipedia

### 2019-12-04 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
Updates to quality definitions document - added mostly ISI 9126 definitions

@@
   The capability of a software product to be installed in a specified
-  environment \citep{iso2001iec}. (As summarized in
+  environment \citep{ISO9126}. (As summarized in
   \citep{berander2005software}.)  \wss{remove the quotation marks - we are
     assuming all definitions are directly quoted from the source. Please remove
-    quotes for all of the definitions that you ``own.''}  \oo{Done}
+    quotes for all of the definitions that you ``own.''}  \oo{Done} \wss{I've
+    verified that this definition is correct.  We do not have to cite the
+    textbook anymore, since we have the original source.}
 \end{defn}
@@
 
+\begin{defn}
+Testability is the capability of the software product to enable modified
+software to be validated \cite{ISO9126}.
+\end{defn}
+
 \noindent \textbf{Proposed Definition}
@@
   \citep{iso2001iec}. (As summarized in \citet{berander2005software}.)
+  \wss{This definition does not match the definition I found in the
+    \cite{ISO9126}.  @olu, can you please look into this?  My definition is
+    given below.  Also, I've added a pdf of the ISO standard to the pub repo.}
 \end{defn}
@@
   Reliability expresses the ability of the software to maintain a specified
-   level of fault tolerance, when used under specified condition \citep{singh2013different}.
-\end{defn}
-\begin{defn}
-	The extent to which a program can be expected to perform its intended
-	function with required precision \citep{McCallEtAl1977}. (As summarized in
-    \citet{VanVliet2000}.)
+  level of fault tolerance, when used under specified condition
+  \citep{singh2013different}.
+\end{defn}
+
+\begin{defn}
+  The extent to which a program can be expected to perform its intended function
+  with required precision \citep{McCallEtAl1977}. (As summarized in
+  \citet{VanVliet2000}.) \wss{When we have a copy of the McCall reference we can
+    verify this and simply cite the primary source.}
 \end{defn}
@@
 \begin{defn}
-  The ability of a system or component to to perform its required function under
+  The ability of a system or component to perform its required function under
   stated conditions for a specified period of time
   \citep{IEEEComputerDictionary1991}
+\end{defn}
+
+\begin{defn}
+ The capability of the software to maintain the level of performance of the
+ system when used under specified conditions \cite{ISO9126}.
 \end{defn}
@@
 Reliability is defined in a way that the measurement metrics are visible - the
-specified enviroment, the specified time and the specified level of fault
+specified environment, the specified time and the specified level of fault
 tolerance. Reliability is a critical quality for scientific software, since the
@@
   The extent to which a product can be used by specified users to achieve
-	specified goals with effectiveness, efficiency, and satisfaction in a specified
-	context of use.
+  specified goals with effectiveness, efficiency, and satisfaction in a
+  specified context of use.
 \end{defn}
@@
 \begin{itemize}
-	\item Learnability: How easy is it for users to accomplish basic tasks the
-	first time they encounter the design?
+\item Learnability: How easy is it for users to accomplish basic tasks the first
+  time they encounter the design?
 \item Efficiency: Once users have learned the design, how quickly can they
@@
 \emph{usability} (purely an interface concern) and \emph{utility} (in the
-economics
-sense of the word).
+economics sense of the word).
 
@@
 
+\begin{defn}
+  The capability of the software to be understood, learned, used and liked by
+  the user, when used under specified conditions \cite{ISO9126}.
+\end{defn}
 \noindent \textbf{Proposed Definition}
@@
 
+\begin{defn}
+  The capability of the software to be modified.  Modifications may include
+  corrections, improvements or adaptation of the software to changes in
+  environment, and in requirements and functional specifications. \cite{ISO9126}
+\end{defn}
+
 \noindent \textbf{Proposed Definition}
@@
 \subsection{\notdone{Portability} \pmi{owner}}
+
 \begin{defn}
@@
 \end{defn}
+
 \begin{defn} \label{PortabilityDefnSelected} 
@@
 
+\begin{defn}
+  The capability of software to be transferred from one environment to another
+  \cite{ISO9126}.
+\end{defn}
 \noindent \textbf{Proposed Definition}
@@
 
+\begin{defn}
+  The capability of the software product to enable the user to understand
+  whether the software is suitable, and how it can be used for particular tasks
+  and conditions of use \cite{ISO9126}.
+\end{defn}
+
 \noindent \textbf{Proposed Definition}
@@
 
+\begin{defn}
+  The capability of the software to interact with one or more specified systems
+  \cite{ISO9126}.
+\end{defn}
 \noindent \textbf{Proposed Definition}
@@
 
+\begin{defn}
+  The degree to which a full implementation of the required functionality has
+  been achieved \citep{McCallEtAl1977}. (As summarized in \citet{VanVliet2000}.)
+\end{defn}
+
 \wss{Were there any other definitions of completeness?  You could add the
@@
 
+\begin{defn}
+  The use of uniform design and implementation techniques and notations
+  throughout a project \citep{McCallEtAl1977}. (As summarized in
+  \citet{VanVliet2000}.)
+\end{defn}
+
 \wss{The definition from \citet{IEEE1998} might again be useful.  We could also
@@
 \item the identification and documentation of derivation paths (upward) and
-  allocation or flowdown paths (downward) of work products in the work product
+  allocation or flow-down paths (downward) of work products in the work product
   hierarchy;
@@
 \end{enumerate}
+
+\begin{defn}
+  The ability to link software components to requirements
+  \citep{McCallEtAl1977}. (As summarized in \citet{VanVliet2000}.)
+\end{defn}
 

### 2019-12-04 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
updated visibility
updated productivity

@@
 
+\begin{defn}
+The degree to which something is seen by the public
+\citep{CambridgeVisibility2019}.
+
+The degree to which something is seen or known about
+\citep{CambridgeVisibility2019}.
+\end{defn}
+
 \noindent \textbf{Proposed Definition}
@@
 \end{defn}
-
 \begin{defn}
@@
 	\[\text{Productivity} = \dfrac{\text{Outputs produced by the
-        process}}{\text{Inputs consumed by the process}}\] Defining inputs. For
-  the software process, providing a meaningful definition of inputs is a
-  nontrivial but generally workable problem. Inputs to the software process
-  generally comprise labor, computers, supplies, and other support facilities
-  and equipment. Defining outputs. The big problem in defining software
-  productivity is defining outputs. Here we find a defining delivered source
-  instructions (DSI) or lines of code as the output of the software process is
-  totally inadequate, and they argue that there are a number of deficiencies in
-  using DSI. However, most organizations doing practical productivity
-  measurement still use DSI as their primary metric \citep{Boehm1987}. \wss{Is
-    this a direct quote from \citet{Boehm1987}?  The sentences seem incomplete?}
-\end{defn}
-
-\begin{defn}
-  \label{ProductivitySelected}
-  Productivity is the amount of output (what is produced) per unit of input
-  used. If we can measure the size of the software product and the effort
-  required to develop the product, we have:
+			process}}{\text{Inputs consumed by the process}}\]
+
+Thus, we can improve the productivity of the software process by increasing its
+outputs, decreasing its inputs, or both. However, this means that we need to
+provide meaningful definitions of the inputs and outputs of the software
+process.
+
+\textbf{Defining inputs.} For the software process, providing a meaningful
+definition of inputs is a nontrivial but generally workable problem. Inputs to
+the software process generally comprise labor, computers, supplies, and other
+support facilities and equipment. However, one has to be careful which of
+various classes of items are to be counted as inputs. For example:
+\begin{itemize}
+\item Phases (just software development, or should we include system
+engineering, soft-ware requirements analysis, installation, or post development
+support?)
+\item Activities (to include documentation, project management, facilities
+management, conversion, training, database administration?)
+\item Personnel (to include secretaries, computer operators, business managers,
+contract administrators, line management?)
+\item Resources (to include facilities, equipment, communications, current
+versus future dollar payments?)
+\end{itemize}
+An organization can usually reach an agreement on which of the above are
+meaningful as inputs in their organizational context. Frequently, one can use
+present-value dollars as a uniform scale for various classes of resources.
+
+\textbf{Defining outputs.} The big problem in defining software productivity is
+defining outputs. Here we find a defining delivered source instructions (DSI) or
+lines of code as the output of the software process is totally inadequate, and
+they argue that there are a number of deficiencies in using DSI. However,
+	most organizations doing practical productivity measurement still use DSI as
+	their primary metric \citep{Boehm1987}. \wss{Is this a direct quote from
+		\citet{Boehm1987}?  The sentences seem incomplete?}
+\ad{I added some deleted parts, now it is a direct quote. After the ending of
+the quote, the following 2 pages discuss the flaws of DCI and a list of
+alternatives to DCI, so I ended the quote here.}
+\end{defn}
+\begin{defn}
+	\label{ProductivitySelected1}
+Productivity is the amount of output (what is produced) per unit of input used.
+In general, productivity is difficult to measure because outputs and inputs are
+typically quite diverse and are often themselves difficult to measure. In the
+context of software, productivity measurement is usually based on a simple ratio
+of product size to project effort. Thus,If we can measure the size of the
+software product and the effort required to develop the product, we have:
 	\begin{align}
@@
 	\end{align}
-  Equation (1) assumes that size is the output of the software production
-  process and effort is the input to the process. This can be contrasted with
-  the viewpoint of software cost models where we use size as an independent
-  variable (i.e., an input) to predict effort which is treated as an
-  output. Equation (1) is simple to operationalize if we have a single dominant
-  size measure, for example, product size measured in lines of code
-  \citep{Kitchenham2004}.
-\end{defn}
-
-\noindent \textbf{Proposed Definition}
-
-The first sentence of Definition \ref{ProductivitySelected}: Productivity is the
-amount of output (what is produced) per unit of input used.
+Equation (1) assumes that size is the output of the software production process
+and effort is the input to the process. This can be contrasted with the
+	viewpoint of software cost models where we use size as an independent variable
+	(i.e., an input) to predict effort which is treated as an output. Equation (1)
+	is simple to operationalize if we have a single dominant size measure, for
+	example, product size measured in lines of code \citep{Kitchenham2004}.
+\end{defn}
+\begin{defn}
+	\label{ProductivitySelected2}
+The number of lines of new code developed per person-day (an imperfect measure
+of productivity but one that could be measured consistently)
+\citep{MacCormack2003}.
+\end{defn}
+
+\noindent \textbf{Proposed Definition}
+
+The revision of the combination of Definition \ref{ProductivitySelected1} and
+Definition \ref{ProductivitySelected2}: Productivity is the
+amount of output per unit of input used, which can be measured by the summation
+of all output (such as the number of lines of new code, the number of pages of
+new documents and the number of new test cases) produced per person-day.
 
@@
 	we need a better idea of what we are measuring for outputs and inputs.}
+	\ad{I added another def and made it more measurable}
 

### 2019-12-04 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
updated completeness

@@
 
+\begin{defn}
+	An SRS is complete if, and only if, it includes the following elements:
+\begin{itemize}
+\item All significant requirements, whether relating to functionality,
+performance, design constraints, attributes, or external interfaces. In
+particular any external requirements imposed by a system specification should be
+acknowledged and treated.
+\item Definition of the responses of the software to all realizable classes of
+input data in all realizable classes of situations. Note that it is important to
+specify the responses to both valid and invalid input values.
+\item Full labels and references to all figures, tables, and diagrams in the SRS
+and definition of all terms and units of measure \citep{IEEE1998}.
+	\end{itemize}
+\end{defn}
+\begin{defn}
+The quality of being whole or perfect and having nothing missing.
+\citep{CambridgeCompleteness2019}.
+\end{defn}
+
 \wss{Were there any other definitions of completeness?  You could add the
 	definition of completeness from \citet[p.\ 5--6]{IEEE1998}.  This definition
-  is for requirements, but maybe there is something we can generalize from the
+	is for
+	requirements, but maybe there is something we can generalize from the
 	definition?  We could also look for definitions outside of software
 	development.}
+\ad{added IEEE and dictionary}
+\ad{\url{https://annals-csis.org/proceedings/2016/pliks/468.pdf} has a method of
+measuring completeness and consistency, but I haven't summarized any def from it
+yet}
 

### 2019-12-04 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
updated consistency

@@
 
+\begin{defn}
+Consistency refers to internal consistency. If an SRS does not agree with some
+higher-level document, such as a system requirements specification, then it is
+not correct \citep{IEEE1998}.
+\end{defn}
+\begin{defn}
+An SRS is internally consistent if, and only if, no subset of individual
+requirements described in it conflict. The three types of likely conflicts in an
+SRS are as follows \citep{IEEE1998}:
+	\begin{enumerate}[a)]
+\item The specified characteristics of real-world objects may conflict. For
+example,
+	\begin{enumerate}[1)]
+\item The format of an output report may be described in one requirement as
+tabular but in another as textual.
+\item One requirement may state that all lights shall be green while another may
+state that all lights shall be blue.
+	\end{enumerate}
+\item There may be logical or temporal conflict between two specified actions.
+For example,
+	\begin{enumerate}[1)]
+\item One requirement may specify that the program will add two inputs and
+another may specify that the program will multiply them.
+\item One requirement may state that “A” must always follow “B,” while another
+may require that “A and B” occur simultaneously.
+	\end{enumerate}
+\item Two or more requirements may describe the same real-world object but use
+different terms for that object. For example, a program’s request for a user
+input may be called a “prompt” in one requirement and a “cue” in another. The
+use of standard terminology and definitions promotes consistency.
+	\end{enumerate}
+\end{defn}
+\begin{defn}
+The state or condition of always happening or behaving in the same way
+\citep{CambridgeConsistency2019}.
+\end{defn}
+
 \wss{The definition from \citet{IEEE1998} might again be useful.  We could also
@@
   definition could be helpful.}
+\ad{added IEEE and dictionary}
 

### 2019-12-04 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
Incorporation of comments on some of the definitions

@@
 definition is given verbatim from the cited source, but without showing
-quotation marks.  After the summary of the existing definitions, we propose the
+quotation marks.  In cases where the definition has to be rephrased, quotation
+marks are used to show those portions that are taken verbatim for the original
+source.  After the summary of the existing definitions, we propose the
 definition that we would like to work with going forward.  This definition can
@@
 definition is an explanation for the reasoning that led to this choice.
+
+In determining our recommended definition for each quality, we used the
+following criteria:
+
+\begin{description}
+\item[Consistency] The final list of definitions should be consistent in the
+  terminology used.  For instance, we do not use synonyms for variety; we use
+  the same word to mean the same concept throughout the final list of
+  definitions.  This is why measurement is given by ``the effort required..''
+  and not the various synonyms for this that are seen in the literature. \wss{We
+    may change this phrase we use for measuring something.  I just wanted to put
+    this in here as a placeholder.}
+\item[Generality]
+\item[Measurability]
+\item[Etc.]
+\end{description}
 
@@
     of the definition.  Can you please find a pdf of the McCall reference and
-    add it to our pub repo.  We should cite the primary source,
-    but we can't assume what it says.}
+    add it to our pub repo.  We should cite the primary source, but we can't
+    assume what it says.  The source is definitely NOT
+    \citep{smith2015comparing}.  By the way, you should use the BibTeX entries
+    from References.bib in the pub repo when they are available.  This will save
+    you time and make the citations more accurate. }  \wss{Olu, once we have the
+    McCall reference, please verify all of the definitions where it is used, and
+    remove the reference to the Van Vliet textbook.}
 \end{defn}
@@
     verified that this definition is correct.  We do not have to cite the
-    textbook anymore, since we have the original source.}
+    textbook anymore, since we have the original source.  Olu, please remove the
+    \citep{berander2005software} in all instances where we have verified with
+    the original source.}
 \end{defn}
@@
 
-Combined definitions \ref{Installability_Selected1} and
-\ref{Installability_Selected2} rephrased.
-Installability refers to the cost or effort required for the installation,
-uninstallation or reinstallation of a software or product in a specified
-enviroment.  \wss{This definition does not allow for
-	measuring the machine time.}
+Combined definitions \ref{Installability_Selected2} and
+\ref{Installability_Selected1} rephrased.  Installability refers to the cost or
+effort required for the installation, uninstallation or reinstallation of a
+software or product in a specified environment.  \wss{This definition does not
+  allow for measuring the machine time.  Do we want to distinguish between the
+  human effort required and the total time of installation?  I think the human
+  effort is the important metric here.}
 
@@
     \citet{GhezziEtAl2003} defines correctness.  They only require satisfaction
-    of the functional requirements.} \oo{Updated}
+    of the functional requirements.} \oo{Updated} \wss{Is this verbatim from the
+    original source?  If not, we should show those parts that are verbatim in
+    quotes, so that readers know the parts that were developed by us.}
 
@@
   from defects in its specification, design, and implementation, is its
-  correctness.
+  correctness.  \wss{Olu, can you please rephrase this one to match how the
+    other definitions are presented.  That is, give the verbatim definition
+    followed by the citation.}
 \end{defn}
@@
 Definition~\ref{Correctness_Selected}: The degree to which a system or component
-is free from faults in its specification, design and implementation.
+is free from faults in its specification, design and implementation.  \wss{I
+  think I prefer the Ghezzi definition, but with respect to all requirements
+  (not just functional).}
 
@@
 To determine whether a program is correct, we must be able to refer to both its
-its specification, design and implementation to understand the intended purpose
-and based on defined metrics, determine the extent to which this is met. Most
+specification, design and implementation to understand the intended purpose and
+based on defined metrics, determine the extent to which this is met. Most
 definitions of correctness refers only to functional specifications but this
 definition is broad and is able to capture both functional and non-functional
-specification without explicitly specifying it.\oo{We could modify te definition
-  to explicitly state functional and non-functional if we decide.}  There is no
-direct tool or method for measuring correctness. One way of building confidence
-in correctness is by reviewing to ensure that each requirement stated is one
-that the stakeholders and experts desire.  By maintaining traceability,
+specification without explicitly specifying it.\oo{We could modify the
+  definition to explicitly state functional and non-functional if we decide.}
+There is no direct tool or method for measuring correctness. One way of building
+confidence in correctness is by reviewing to ensure that each requirement stated
+is one that the stakeholders and experts desire.  By maintaining traceability,
 consistency and unambiguity, we can reduce the occurrence of errors and make the
@@
 
+\wss{We probably do not need the extra bits about measuring correctness and the
+  dependency between correctness and other qualities.  We can discuss how to
+  measure correctness as the next exercise we are doing after concluding the
+  definitions exercise.  If the tradeoffs between qualities are important, we
+  should cover that in a more systematic way.}
+
 \subsection{\notdone{Verifiability/Testability} \oo{owner}}
@@
   A software is verifiable if its properties can be verified easily
-  \citep{GhezziEtAl2003}.
+  \citep{GhezziEtAl2003}.  \wss{Is this a direct quote?}
 \end{defn}
@@
   \citet{pressman2005software})
+\end{defn}
+
+\begin{defn}
+	The effort required to test a program to ensure that it performs its
+	intended function \citep{McCallEtAl1977}. (As summarized in
+	\citet{VanVliet2000}.) \oo{similar to \citep{pressman2005software}, this}
 \end{defn}
@@
 
-\wss{In case we need it, here is a definition of testability.}
-
-\begin{defn}
-	The effort required to test a program to ensure that it performs its
-	intended function \citep{McCallEtAl1977}. (As summarized in
-	\citet{VanVliet2000}.) \oo{similar to \citep{pressman2005software}, this}
-\end{defn}
-
 \subsection{\notdone{Validatability} \oo{owner}}
@@
   Validatability means solving the right equations \citep[p.~23]{Roache1998}.
-  \oo{Validation?}
+  \oo{Validation?}  \wss{We should also have the definition for verification
+    about solving the equations right.}
 \end{defn}
@@
   development process to determine whether it satisfies specified business
-  requirements \citep{softwaretestingfundamentals}.
+  requirements \citep{softwaretestingfundamentals}.  \wss{Please use the
+    author's name for this citation.  If that isn't known, use the name of the
+    organization that published this material.}
 \end{defn}
@@
 \end{defn}
-\noindent \textbf{Proposed Definition}
 
@@
   users, operators, administrators, managers, investors, etc.)
-  \citep{WikipediaVnV}
+  \citep{WikipediaVnV} \wss{This reference is not appearing in the Reference
+    list.  We also don't really want a Wikipedia citation in a scholarly paper.
+    Can you please find the source where this definition originally came from?}
 \end{defn}
@@
   The probability that the software will operate as expected over a specified
-  time interval \citep{GhezziEtAl2003}.
+  time interval \citep{GhezziEtAl2003}.  \wss{Is this verbatim from
+    \citet{GhezziEtAl2003}?}
 \end{defn}
@@
 
+\begin{defn}
+ The capability of the software to maintain the level of performance of the
+ system when used under specified conditions \cite{ISO9126}.
+\end{defn}
+
 \begin{defn} \label{reliabilitySelected1} Probability of failure-free operation
@@
 
-\begin{defn}
- The capability of the software to maintain the level of performance of the
- system when used under specified conditions \cite{ISO9126}.
-\end{defn}
+\wss{Olu, have you looked at the later chapters of the \citet{GhezziEtAl2003}?
+  I believe there is a statistical definition of reliability there.}
 
@@
 improve visibility. It is rephrased because the original one might refer to
-binary status - ``visible'' or ``unvisible''.
+binary status - ``visible'' or ``invisible''.
 

### 2019-12-14 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
updated installaility

@@
 \begin{defn}
-  A measure of the ease of installation \citep{smith2015comparing}.\wss{Where
-    does this definition come from? Please add a citation.  Is it from
-    \citet{McCallEtAl1977}? I thought so originally, but after looking into it,
-    I don't think so anymore.}  \oo{@smith you added the definition in here but
-    I guess it's from here} \wss{@olu, I was hoping you would verify the source
-    of the definition.  Can you please find a pdf of the McCall reference and
-    add it to our pub repo.  We should cite the primary source, but we can't
-    assume what it says.  The source is definitely NOT
-    \citep{smith2015comparing}.  By the way, you should use the BibTeX entries
-    from References.bib in the pub repo when they are available.  This will save
-    you time and make the citations more accurate. }  \wss{Olu, once we have the
-    McCall reference, please verify all of the definitions where it is used, and
-    remove the reference to the Van Vliet textbook.}
-\end{defn}
-
-\begin{defn}
-  The capability of a software product to be installed in a specified
-  environment \citep{ISO9126}. (As summarized in
-  \citep{berander2005software}.)  \wss{remove the quotation marks - we are
-    assuming all definitions are directly quoted from the source. Please remove
-    quotes for all of the definitions that you ``own.''}  \oo{Done} \wss{I've
-    verified that this definition is correct.  We do not have to cite the
-    textbook anymore, since we have the original source.  Olu, please remove the
-    \citep{berander2005software} in all instances where we have verified with
-    the original source.}
+  The capability of the software product to be installed in a specified
+  environment \citep{ISO9126}.\citep{ISO9126}
 \end{defn}
@@
 Combined definitions \ref{Installability_Selected2} and
-\ref{Installability_Selected1} rephrased.  Installability refers to the cost or
+\ref{Installability_Selected1} rephrased.  Installability refers to the
 effort required for the installation, uninstallation or reinstallation of a
-software or product in a specified environment.  \wss{This definition does not
-  allow for measuring the machine time.  Do we want to distinguish between the
-  human effort required and the total time of installation?  I think the human
+software or product in a specified environment.  \wss{I think the human
   effort is the important metric here.}
@@
 
-\wss{The rationale is missing.  Please add.} \oo{Done} The definition captures
-the cost or effort - human or machine, and considers not just the initial
-installation but uninstallation and re-installation which are all related to
-installation activities.  \oo{Done installability}
+ The definition captures a vital metric of measure for this quality - effort, and considers not just the installation but uninstallation and re-installation which are all related toinstallation activities.
+\oo{Done installability}
 
@@
   The extent to which a program satisfies its specifications and fulfills the
-  user's mission objectives \citep{McCallEtAl1977}. (As summarized in
-  \citet{VanVliet2000}.)
+  user's mission objectives \citep{McCallEtAl1977}. \oo{definition is checked with Mccall}
 \end{defn}
@@
 
-\subsection{\notdone{Verifiability/Testability} \oo{owner}}
+\subsection{\notdone{Verifiability} \oo{owner}}
 
@@
 
-\begin{defn}
+\begin{defn} \label{Defn_Verifiability1}
   Attributes of software that relate to the effort needed for validating the
@@
   function \citep{McCallEtAl1977}. (As summarized in
-  \citet{pressman2005software})
+  \citet{pressman2005software})\oo{McCall defined this as testability not verifiability}
 \end{defn}
@@
 	\citet{VanVliet2000}.) \oo{similar to \citep{pressman2005software}, this}
-\end{defn}
-
-\begin{defn}
-  Testability is the degree to which a requirement is stated in terms that
-  permits establishment of test criteria and performance of tests to determine
-  whether those criteria have been met \citep{IEEEStdGlossarySET1990}
-\end{defn}
-
-\begin{defn}
-  Code possesses the characteristic testability to the extent that it
-  facilitates the establishment of verification criteria and supports evaluation
-  of its performance \citep{boehm1976quantitative}.
-\end{defn}
-
-\begin{defn} \label{Defn_Verifiability1} 
-  Testability is the degree to which a system or component facilitates the
-  establishment of test criteria and the performance of tests to determine
-  whether those criteria have been met \citep{ISO/IEC/IEEE24765}.
-\end{defn}
-
-\begin{defn} \label{Defn_Verifiability2} 
-  Testability is the effort required to test a software
-  \citep{ISO/IEC/IEEE24765}.
-\end{defn}
-
-\begin{defn}
-Testability is the capability of the software product to enable modified
-software to be validated \cite{ISO9126}.
-\end{defn}
-
-\noindent \textbf{Proposed Definition}
-
-Combined definitions~\ref{Defn_Verifiability1} and~\ref{Defn_Verifiability2}
+\end{defn} \oo{McCall defined this as testability not verifiability}
+
+\noindent \textbf{Proposed Definition}
+
+Combined definitions~\ref{Defn_Verifiability1}
 rephrased: Testability or verifiability is the degree to which a system or
@@
   users, operators, administrators, managers, investors, etc.)
-  \citep{WikipediaVnV} \wss{This reference is not appearing in the Reference
+   \wss{This reference is not appearing in the Reference
     list.  We also don't really want a Wikipedia citation in a scholarly paper.
@@
 
-Definition~\ref{} \oo{need to discuss this quality, we might just define
+Definition~ \oo{need to discuss this quality, we might just define
   validatability from validation definitions above? Resources are limited for
@@
 \begin{defn}
- The capability of the software to maintain the level of performance of the
- system when used under specified conditions \cite{ISO9126}.
-\end{defn}
-
-\begin{defn} \label{reliabilitySelected1} Probability of failure-free operation
-  of a computer program in a specified environment for a specified time
-  \citep{musa1987software}. (As stated in \citet{pressman2005software}.)
+ The capability of the software product to maintain a specified level of performance when used under specified conditions \cite{ISO9126}.
+\end{defn}
+
+\begin{defn} \label{reliabilitySelected1} 
+	It is the probability of failure-free operation of a computer program in a specified environment for a specified time
+    \citep{musa1987software}.
 \end{defn}
@@
   The extent to which a program can be expected to perform its intended function
-  with required precision \citep{McCallEtAl1977}. (As summarized in
-  \citet{VanVliet2000}.) \wss{When we have a copy of the McCall reference we can
-    verify this and simply cite the primary source.}
+  with required precision \citep{McCallEtAl1977}. \oo{Check definition with pdf}
 \end{defn}
@@
 
-There are two ISO standards covering this, namely ISO/TR 16982:202 and ISO 9241.
+There are two ISO standards covering this, namely ISO/TR 16982:202 and ISO9241.
 

### 2019-12-16 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
update defnitions reliability, installability

@@
   A program is functionally correct if it behaves according to its stated
-  functional specifications \citep{GhezziEtAl2003}.  \wss{This is not how
-    \citet{GhezziEtAl2003} defines correctness.  They only require satisfaction
-    of the functional requirements.} \oo{Updated} \wss{Is this verbatim from the
-    original source?  If not, we should show those parts that are verbatim in
+  functional specifications \citep{GhezziEtAl2003}.   \wss{Is this verbatim from the original source?  If not, we should show those parts that are verbatim in
   quotes, so that readers know the parts that were developed by us.}
+  \oo{Yes, verbatim}
 
@@
 \begin{defn}
-  The probability that the software will operate as expected over a specified
-  time interval \citep{GhezziEtAl2003}.  \wss{Is this verbatim from
-    \citet{GhezziEtAl2003}?}
+  Informally, software is reliable if the user can depend on it \citep{GhezziEtAl2003}.
 \end{defn}
@@
 
-\begin{defn} \label{reliabilitySelected2} 
+\begin{defn}
   Reliability expresses the ability of the software to maintain a specified
@@
 \begin{defn}
-  The extent to which a program can be expected to perform its intended function
-  with required precision \citep{McCallEtAl1977}. \oo{Check definition with pdf}
+  Extent to which a program can be expected to perform its intended function
+  with required precision \citep{McCallEtAl1977}. \oo{Definition check pass}
 \end{defn}
@@
   period of time, that the system will correctly deliver services as expected by
-  the user \citep{sommerville}
+  the user \citep{sommerville}.
 \end{defn}
@@
   stated conditions for a specified period of time
-  \citep{IEEEComputerDictionary1991}
-\end{defn}
-
-\wss{Olu, have you looked at the later chapters of the \citet{GhezziEtAl2003}?
-  I believe there is a statistical definition of reliability there.}
+  \citep{IEEEComputerDictionary1991}.
+\end{defn}
+
+\begin{defn}
+  The ability of a system to perform its required functions under stated conditions whenever required—having a long mean time between failures \citep{mcconnell2004code}.
+\end{defn}
+\begin{defn} \label{reliabilitySelected2} 
+	Reliability can be estimated as ``the average time interval between two faliures, also called the mean time to faliure (MTTF).
+Clearly, MTTF (t) = 1/FI (t)''.
+Where FI (t) is the faliure intensity - the number of faliures per unit time \citep{GhezziEtAl2003}.
+\end{defn}
+\oo{Statistical definition from  \citep{GhezziEtAl2003} added}
 
@@
 Probability of failure-free operation of a computer program in a specified
-environment for a specified time at a specified level of fault tolerance.
+environment for a specified time, i.e. the average time interval between two faliures also known as the mean time to failure (MTTF).
 
@@
 
-Reliability is defined in a way that the measurement metrics are visible - the
-specified environment, the specified time and the specified level of fault
-tolerance. Reliability is a critical quality for scientific software, since the
-results of computations are meaningless, if they are not dependable.
-Reliability is closely tied to verifiability, since the key quality to verify is
-reliability, while the act of verification itself improves reliability.
-
-Reliability models can be used to predict reliability of a software product. For
-example measuring Mean Time to Fail (MTTF) can be a good measure of reliability
-\citep{berander2005software}.
+Reliability is defined such that the measurement metrics are visible - the
+specified environment and time.
 

### 2019-12-18 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
Fixed some long lines to match 80 character length constraint

@@
 
- The definition captures a vital metric of measure for this quality - effort, and considers not just the installation but uninstallation and re-installation which are all related toinstallation activities.
-\oo{Done installability}
+The definition captures a vital metric of measure for this quality - effort, and
+considers not just the installation but uninstallation and re-installation which
+are all related toinstallation activities.  \oo{Done installability}
 
@@
   A program is functionally correct if it behaves according to its stated
-  functional specifications \citep{GhezziEtAl2003}.   \wss{Is this verbatim from the original source?  If not, we should show those parts that are verbatim in
-  quotes, so that readers know the parts that were developed by us.}
+  functional specifications \citep{GhezziEtAl2003}.  \wss{Is this verbatim from
+    the original source?  If not, we should show those parts that are verbatim
+    in quotes, so that readers know the parts that were developed by us.}
   \oo{Yes, verbatim}
@@
 \begin{defn}
- The capability of the software product to maintain a specified level of performance when used under specified conditions \cite{ISO9126}.
-\end{defn}
-
-\begin{defn} \label{reliabilitySelected1} 
-	It is the probability of failure-free operation of a computer program in a specified environment for a specified time
-    \citep{musa1987software}.
+  The capability of the software product to maintain a specified level of
+  performance when used under specified conditions \cite{ISO9126}.
+\end{defn}
+
+\begin{defn} \label{reliabilitySelected1} It is the probability of failure-free
+  operation of a computer program in a specified environment for a specified
+  time \citep{musa1987software}.
 \end{defn}
@@
 \begin{defn}
-  The ability of a system to perform its required functions under stated conditions whenever required—having a long mean time between failures \citep{mcconnell2004code}.
-\end{defn}
-\begin{defn} \label{reliabilitySelected2} 
-	Reliability can be estimated as ``the average time interval between two faliures, also called the mean time to faliure (MTTF).
-Clearly, MTTF (t) = 1/FI (t)''.
-Where FI (t) is the faliure intensity - the number of faliures per unit time \citep{GhezziEtAl2003}.
+  The ability of a system to perform its required functions under stated
+  conditions whenever required—having a long mean time between failures
+  \citep{mcconnell2004code}.
+\end{defn}
+\begin{defn} \label{reliabilitySelected2} Reliability can be estimated as ``the
+  average time interval between two faliures, also called the mean time to
+  faliure (MTTF). Clearly, MTTF (t) = 1/FI (t)''. Where FI (t) is the faliure
+  intensity - the number of faliures per unit time \citep{GhezziEtAl2003}.
 \end{defn}
@@
 Probability of failure-free operation of a computer program in a specified
-environment for a specified time, i.e. the average time interval between two faliures also known as the mean time to failure (MTTF).
+environment for a specified time, i.e. the average time interval between two
+faliures also known as the mean time to failure (MTTF).
 

### 2019-12-18 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
Additions to sustainability section, clean up installability definition

@@
   The capability of the software product to be installed in a specified
-  environment \citep{ISO9126}.\citep{ISO9126}
+  environment \citep{ISO9126}.
 \end{defn}
@@
 effort required for the installation, uninstallation or reinstallation of a
-software or product in a specified environment.  \wss{I think the human
-  effort is the important metric here.}
+software or product in a specified environment.
 
@@
 The definition captures a vital metric of measure for this quality - effort, and
-considers not just the installation but uninstallation and re-installation which
-are all related toinstallation activities.  \oo{Done installability}
+considers not just the installation but uninstallation and re-installation, which
+are all related to installation activities.  \oo{Done installability}
 
@@
 
+Software Sustainability Institute proposal:
+
+\begin{defn}
+Capacity of the software to endure
+\end{defn}
+
+\citet{Katz2016} builds on this definition.
+
 \begin{defn}
@@
 \end{defn}
+
+\href{https://collegeville.github.io/CW3S19/WorkshopResources/Presentations/4-5-Katz_data-driven-software-sustainability.pdf}
+{Katz Presentation}
+
+\href{https://www.software.ac.uk/blog/2019-09-12-making-open-source-research-software-visible-path-better-sustainability}
+{Neil's blog}
 
@@
 
+\begin{defn}
+Sustainability encompasses cost efficient maintainability and evolvability \citep{SehestedtEtAl2014}.
+\end{defn}
+
+\citet{SehestedtEtAl2014} goes on to say that sustainability can be observed by
+evaluating the four criteria of the architectural model: completeness,
+consistency, correctness and clarity.
+
+Definitions for sustainability are often built by combining other definitions.
+\href{https://collegeville.github.io/CW3S19/WorkshopResources/Presentations/1-4-2019CollegevilleWorkshopWillenbring.pdf}
+{Willenbring.pdf}
+lists sustainability factors: extensible, interoperable, maintainable, portable,
+reusable, scalable and usable.  \wss{Sounds like they are listing almost all
+  software qualities.  It seems that sustainability is at least in part achieved
+  by having high quality software.}
+
+From
+\href{https://figshare.com/articles/Defining_Sustainability_through_Developers_Eyes_Recommendations_from_an_Interview_Study/1111925/1}
+{Rosado de Souza, et al.} there are two categories of software sustainability:
+
+\begin{description}
+\item[Intrinsic] Pertaining to characteristics of the software
+\item[Extrinsic] Pertaining to the software development environment
+\end{description}
+Find paper that combines nonfunctional qualities into sustainability.
+
+Sounds like definition of maintainability.
+
 Paper critical of a lack of a definition \citep{VentersEtAl2014}.
-
-Sounds like definition of maintainability.
-
-Find paper that combines nonfunctional qualities into sustainability.
 

### 2019-12-18 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
Updats to comments, fixed typos

@@
   The extent to which a program satisfies its specifications and fulfills the
-  user's mission objectives \citep{McCallEtAl1977}. \oo{definition is checked with Mccall}
+  user's mission objectives \citep{McCallEtAl1977}. \oo{definition is checked
+    with Mccall} \wss{Great!  Did you add the Mccall paper to the pub repo?}
 \end{defn}
@@
   function \citep{McCallEtAl1977}. (As summarized in
-  \citet{pressman2005software})\oo{McCall defined this as testability not verifiability}
+  \citet{pressman2005software})\oo{McCall defined this as testability not
+    verifiability}  \wss{I believe we discussed at the last meeting separating
+    the definitions of verifiability and testability.  Olu, please create a new
+    section for testability and appropriately copy and paste from this section to
+    populate that one.} \wss{Olu, since you have verified the McCall reference,
+    you can remove the pressman reference.}
 \end{defn}
@@
 	\citet{VanVliet2000}.) \oo{similar to \citep{pressman2005software}, this}
-\end{defn} \oo{McCall defined this as testability not verifiability}
+\end{defn} \oo{McCall defined this as testability not verifiability}  \wss{Since
+  you have verified the McCall reference, we can drop the VanVliet reference.}
 
@@
 \begin{defn}
-  Validatability of a software is the degree of ease in validating (checking)
-  that software meets user needs. \wss{Need a citation} \oo{To be removed, I
-    don't have any citation, definition was modified from the definition of
-    verifiability}
-\end{defn}
-
-\begin{defn}
   Validatability means solving the right equations \citep[p.~23]{Roache1998}.
@@
   conform to the standards established for the process, the project, and the
-  product \citep{pressman2005software}.
+  product \citep{pressman2005software}.  \wss{Olu, is this definition
+    originally from pressman, or should we track down the original source?}
 \end{defn}
@@
 accomplishes its designated functions within given constraints, such as speed
-(database response times, for instance), throughput ( transactions per second),
+(database response times, for instance), throughput (transactions per second),
 capacity (con-current usage loads), and timing (hard real-time
@@
 	should happen in relation to something else - state transitions}
+\wss{If we need both speed and timing, we should define them.  We aren't
+  including accuracy from the IEEE definition.  Is that on purpose?}
 
@@
   of a program'' \citep{McCallEtAl1977}. (As summarized in
-  \citet{VanVliet2000}.)
-\end{defn}
-
-\begin{defn}
-  ``A software is usable - or user fridendly - if its human users find it easy
+  \citet{VanVliet2000}.)  \wss{Once we have verified McCall said this, we can
+    remove the VanVliet reference.}
+\end{defn}
+
+\begin{defn}
+  ``A software is usable - or user friendly - if its human users find it easy
   to use'' \citep{ghezzi1991fundamentals}.
@@
   Effort required to locate and fix an error in a program
-  \citep{pressman2005software}.
+  \citep{pressman2005software}. \wss{Peter, if this is from
+    McCall, we should just cite the primary source and remove the pressman
+    reference.}
 \end{defn}
@@
     given definition ``is a very limited definition''. Perhaps their intent was
-    breadth.}
+    breadth.} \wss{Peter, please verify the McCall reference definition, and
+    then just cite that.  Olu has found the McCall reference, so you shouldn't
+    have to search for it.  We can remove both the VanVliet and Pressman
+    reference in  preference of the McCall definition.}
 \end{defn}
@@
 component can be modified to correct faults, improve performance or other
-attributes, or satisfy new requirements.
+attributes, or satisfy new requirements. \wss{We didn't agree as a group in the
+  last meeting on our ``measurement'' term, but I think ease is difficult to
+  measure.  Maybe we should rephrase to measuring the effort required?}
 
@@
   applications - related to the packaging and scope of the functions that the
-  program performs \citep{pressman2005software}.
+  program performs \citep{pressman2005software}. \wss{Peter, if this is from
+    McCall, we should just cite the primary source and remove the pressman
+    reference.}
 \end{defn}
@@
   applications \citep{McCallEtAl1977}. (As summarized in \citet{VanVliet2000}.)
+  \wss{Peter, if this is from McCall, we should just cite the primary source and
+    remove the VanVliet reference.}
 \end{defn}
@@
 
-\begin{defn} \label{PortabilityDefnSelected} 
-  Effort required to transfer the program from one hardware and/or software
-  system environment to another \citep{pressman2005software}.
+\begin{defn} \label{PortabilityDefnSelected} Effort required to transfer the
+  program from one hardware and/or software system environment to another
+  \citep{pressman2005software}.  \wss{Peter, if this is from McCall, we should
+    just cite the primary source and remove the pressman reference.}
 \end{defn}
@@
   environment to another'' \citep{McCallEtAl1977}. (As summarized in
-  \citet{VanVliet2000}.)
+  \citet{VanVliet2000}.)  \wss{Peter, if this is from McCall, we should just
+    cite the primary source and remove the VanVliet reference.}
 \end{defn}
@@
   ``The effort required to couple one system with another''
-  \citep{McCallEtAl1977}. (As summarized in \citet{VanVliet2000}.)
+  \citep{McCallEtAl1977}. (As summarized in \citet{VanVliet2000}.)  \wss{Ao, if
+    this is from McCall, we should just cite the primary source and remove the
+    Van Vliet reference.}
 \end{defn}
@@
 Definition \ref{VisibilitySelected} rephrased: The extent to which all of the
-steps and the current status of a software development process are documented
-clearly.
+steps \wss{What are all the steps?} and the current status of a software
+development process are documented clearly. \wss{I wonder if rather than
+  ``documented clearly'', we should be more abstract and talk about how these
+  things can be easily determined.  Documentation is probably how the
+  information will be conveyed, but it doesn't have to be how it is done.}
 
@@
   been achieved \citep{McCallEtAl1977}. (As summarized in \citet{VanVliet2000}.)
+  \wss{Ao, if this is from McCall, we should just cite the primary source and
+    remove the Van Vliet reference.}
 \end{defn}
@@
   throughout a project \citep{McCallEtAl1977}. (As summarized in
-  \citet{VanVliet2000}.)
+  \citet{VanVliet2000}.)  \wss{Ao, if this is from McCall, we should just cite
+    the primary source and remove the Van Vliet reference.}
 \end{defn}
@@
   \citep{McCallEtAl1977}. (As summarized in \citet{VanVliet2000}.)
-\end{defn}
+\end{defn} \wss{Once we have verified McCall, we can remove the Van Vliet
+  reference.}
 
@@
 ``The effort required to test a program to ensure that it performs its intended
-function'' \citep{VanVliet2000}.
+function'' \citep{VanVliet2000}. \wss{Need to verify McCall reference, then
+  delete VanVliet.}
 

### 2019-12-18 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
updated more definitions

--- 2019-12-18 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
@@
 
-\subsection{\notdone{Installability} \oo{owner}}
+\subsection{{Installability} \oo{owner}}
 
@@
 
-The definition captures a vital metric of measure for this quality - effort, and
-considers not just the installation but uninstallation and re-installation which
-are all related toinstallation activities.  \oo{Done installability}
-
-\subsection{\notdone{Correctness} \oo{owner}}
-
-\begin{defn}  
+ The definition captures a vital metric of measure for this quality - effort, and considers not just the installation but uninstallation and re-installation which are all related to installation activities.
+\oo{Done installability}
+
+\subsection{{Correctness} \oo{owner}}
+
+\begin{defn} \label{Correctness_Selected}
   A program is functionally correct if it behaves according to its stated
-  functional specifications \citep{GhezziEtAl2003}.  \wss{Is this verbatim from
-    the original source?  If not, we should show those parts that are verbatim
-    in quotes, so that readers know the parts that were developed by us.}
+  functional specifications \citep{GhezziEtAl2003}.   \wss{Is this verbatim from the original source?  If not, we should show those parts that are verbatim in
+  quotes, so that readers know the parts that were developed by us.}
   \oo{Yes, verbatim}
@@
 \begin{defn}
-  According to \citet{wilson2009quality}, the degree to which a system is free
-  from defects in its specification, design, and implementation, is its
-  correctness.  \wss{Olu, can you please rephrase this one to match how the
-    other definitions are presented.  That is, give the verbatim definition
-    followed by the citation.}
+  The degree to which a system is free from defects in its specification, design, and implementation, is its correctness \citep{wilson2009quality}.
+\oo{Need to verify this definition, still searching for this paper}
 \end{defn}
@@
   The extent to which a program satisfies its specifications and fulfills the
-  user's mission objectives \citep{McCallEtAl1977}. \oo{definition is checked with Mccall}
-\end{defn}
-
-\begin{defn} \label{Correctness_Selected} 
+  user's mission objectives \citep{McCallEtAl1977}. \oo{definition is checked with pdf}
+\end{defn}
+
+\begin{defn}  
   The degree to which a system or component is free from faults in its
@@
 
-Definition~\ref{Correctness_Selected}: The degree to which a system or component
-is free from faults in its specification, design and implementation.  \wss{I
-  think I prefer the Ghezzi definition, but with respect to all requirements
-  (not just functional).}
+Definition~\ref{Correctness_Selected} rephrased: A program is functionally correct if it behaves according to its stated specifications  \oo{Proposed definition updated to Ghezzi's}
 
@@
 specification, design and implementation to understand the intended purpose and
-based on defined metrics, determine the extent to which this is met. Most
-definitions of correctness refers only to functional specifications but this
-definition is broad and is able to capture both functional and non-functional
-specification without explicitly specifying it.\oo{We could modify the
-  definition to explicitly state functional and non-functional if we decide.}
-There is no direct tool or method for measuring correctness. One way of building
-confidence in correctness is by reviewing to ensure that each requirement stated
-is one that the stakeholders and experts desire.  By maintaining traceability,
-consistency and unambiguity, we can reduce the occurrence of errors and make the
-goal of reviewing for correctness easier.
-
-The quality of a software's operation is dependent on the degree of correctness
-\citep{berander2005software}. Correctness and reliability are said to have
-dependencies, such that if a system exhibits a high degree of correctness then
-it tends to be reliable \citep{GhezziEtAl2003}. \oo{Done Correctness}
-
-\wss{We probably do not need the extra bits about measuring correctness and the
-  dependency between correctness and other qualities.  We can discuss how to
-  measure correctness as the next exercise we are doing after concluding the
-  definitions exercise.  If the tradeoffs between qualities are important, we
-  should cover that in a more systematic way.}
+this definition captures it all.
 
@@
 \begin{defn}
+	Verification involves solving the equations right \citep[p.~23]{Roache1998}.
+\end{defn}
+
+\begin{defn}
   A software is verifiable if its properties can be verified easily
-  \citep{GhezziEtAl2003}.  \wss{Is this a direct quote?}
+  \citep{GhezziEtAl2003}.  \wss{Is this a direct quote?} \oo{yes}
 \end{defn}
@@
 \begin{defn} \label{Defn_Verifiability1}
-  Attributes of software that relate to the effort needed for validating the
-  modified software \citep{iso2001iec}.
-\end{defn}
-
-\begin{defn}
-  Effort required to test a program to ensure that it performs its intended
-  function \citep{McCallEtAl1977}. (As summarized in
-  \citet{pressman2005software})\oo{McCall defined this as testability not verifiability}
-\end{defn}
-
-\begin{defn}
-	The effort required to test a program to ensure that it performs its
-	intended function \citep{McCallEtAl1977}. (As summarized in
-	\citet{VanVliet2000}.) \oo{similar to \citep{pressman2005software}, this}
-\end{defn} \oo{McCall defined this as testability not verifiability}
-
-\noindent \textbf{Proposed Definition}
-
-Combined definitions~\ref{Defn_Verifiability1}
-rephrased: Testability or verifiability is the degree to which a system or
-component facilitates the establishment of test criteria and the performance of
-tests to determine whether those criteria have been met and the effort required
-to test.
+Verifiability ``means that you should be able to write a set of tests that can demonstrate that the delivered system meets each specified requirement'' \citep{sommerville}.	
+\end{defn}
+
+\noindent \textbf{Proposed Definition}
+
+Definition~\ref{Defn_Verifiability1}:
+Verifiability means that you should be able to write a set of tests that can demonstrate that the delivered system meets each specified requirement.
 
@@
 Definition is concise and measurable.  Verifiability involves solving the
-equations right \citep[p.~23]{Roache1998}; it benefits from rational
-documentation that systematically shows, with explicit traceability, how the
-governing equations are transformed into code.  \oo{Verifiability is sometimes
-  referred to as testability, so I culled some testability definitions and added
-  here} \wss{culled means removed - is that your intention?} \oo{culled could
-  also mean 'to choose', so I meant, I choose from some definitions of
-  testability and added to the list}
+equations right \citep[p.~23]{Roache1998}.
 
@@
 \begin{defn}
-  Validatability of a software is the degree of ease in validating (checking)
-  that software meets user needs. \wss{Need a citation} \oo{To be removed, I
-    don't have any citation, definition was modified from the definition of
-    verifiability}
-\end{defn}
-
-\begin{defn}
-  Validatability means solving the right equations \citep[p.~23]{Roache1998}.
-  \oo{Validation?}  \wss{We should also have the definition for verification
-    about solving the equations right.}
+  Validation means solving the right equations \citep[p.~23]{Roache1998}.
 \end{defn}
@@
   author's name for this citation.  If that isn't known, use the name of the
-    organization that published this material.}
+  organization that published this material.} \oo{This is a weblink, that's the organization's name I guess}
 \end{defn}
@@
   Validation refers to a different set of tasks that ensure that the software
-  that has been built is traceable to customer requirements.
+  that has been built is traceable to customer requirements.{to be removed, its from geeksforgeeks}
 \end{defn}
@@
   list.  We also don't really want a Wikipedia citation in a scholarly paper.
-    Can you please find the source where this definition originally came from?}
+  Can you please find the source where this definition originally came from?}\oo{Could not find a different source from Wikipedia}
 \end{defn}
@@
 
-Validatability is improved by a rational process via clear documentation of the
-theory and assumptions, along with an explicit statement of the systematic steps
-required for experimental validation. It is entirely possible that a product
-passes when verified but fails when validated. This can happen when, say, a
-product is built as per the specifications but the specifications themselves
-fail to address the user’s needs
-
 \subsection{\notdone{Reliability} \oo{owner}}
@@
 \begin{defn}
-  A set of attributes that relate to the capability of software to maintain its
-  level of performance under stated conditions for a stated period of time
-  \citep{iso2001iec}. (As summarized in \citet{berander2005software}.)
-  \wss{This definition does not match the definition I found in the
-    \cite{ISO9126}.  @olu, can you please look into this?  My definition is
-    given below.  Also, I've added a pdf of the ISO standard to the pub repo.}
-\end{defn}
-
-\begin{defn}
-  The capability of the software product to maintain a specified level of
-  performance when used under specified conditions \cite{ISO9126}.
-\end{defn}
-
-\begin{defn} \label{reliabilitySelected1} It is the probability of failure-free
-  operation of a computer program in a specified environment for a specified
-  time \citep{musa1987software}.
+ The capability of the software product to maintain a specified level of performance when used under specified conditions \cite{ISO9126}.
+\end{defn}
+
+\begin{defn} \label{reliabilitySelected1} 
+	It is the probability of failure-free operation of a computer program in a specified environment for a specified time
+    \citep{musa1987software}.
 \end{defn}
@@
 \begin{defn}
-  The ability of a system to perform its required functions under stated
-  conditions whenever required—having a long mean time between failures
-  \citep{mcconnell2004code}.
-\end{defn}
-\begin{defn} \label{reliabilitySelected2} Reliability can be estimated as ``the
-  average time interval between two faliures, also called the mean time to
-  faliure (MTTF). Clearly, MTTF (t) = 1/FI (t)''. Where FI (t) is the faliure
-  intensity - the number of faliures per unit time \citep{GhezziEtAl2003}.
+  The ability of a system to perform its required functions under stated conditions whenever required—having a long mean time between failures \citep{mcconnell2004code}.
+\end{defn}
+\begin{defn} \label{reliabilitySelected2} 
+	Reliability can be estimated as ``the average time interval between two faliures, also called the mean time to faliure (MTTF).
+Clearly, MTTF (t) = 1/FI (t)''.
+Where FI (t) is the faliure intensity - the number of faliures per unit time \citep{GhezziEtAl2003}.
 \end{defn}
@@
 Probability of failure-free operation of a computer program in a specified
-environment for a specified time, i.e. the average time interval between two
-faliures also known as the mean time to failure (MTTF).
+environment for a specified time, i.e. the average time interval between two faliures also known as the mean time to failure (MTTF).
 

### 2019-12-18 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
Minor edits
fixed errors

@@
   The extent to which a program satisfies its specifications and fulfills the
-  user's mission objectives \citep{McCallEtAl1977}. \oo{definition is checked with pdf}
+  user's mission objectives \citep{McCallEtAl1977}. \oo{definition is checked with pdf} \wss{Great!  Did you add the Mccall paper to the pub repo?}
+\oo{I will}
 \end{defn}
@@
 \end{defn}
-
-\begin{defn} \label{reliabilitySelected1} 
-	It is the probability of failure-free operation of a computer program in a specified environment for a specified time
-    \citep{musa1987software}.
+\begin{defn} \label{reliabilitySelected1} It is the probability of failure-free
+  operation of a computer program in a specified environment for a specified
+  time \citep{musa1987software}.
 \end{defn}
@@
 
-The Interaction Design Foundation
-\url{https://www.interaction-design.org/literature/topics/usability} further
+The Interaction Design Foundation \url{https://www.interaction-design.org/literature/topics/usability} further
 lists the following desirable outcomes:

### 2020-01-06 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
updated performance, maintainability, portability and made comments in regrads to McCall

@@
 	requirements encompass speed (database response times, for instance), throughput
-	( transactions per second), capacity (con-current usage loads), and timing (hard
+	(transactions per second), capacity (concurrent usage loads), and timing (hard
 	real-time demands) \citep{wiegers2003softreq}.
@@
 \begin{defn}
-	In software engineering we often equate performance with efficiency. A software
-	system is efficient if it uses computing resources economically
+	In software engineering we often equate performance with efficiency. A software system is efficient if it uses computing resources economically
 	\citep{ghezzi1991fundamentals}.
@@
 (database response times, for instance), throughput (transactions per second),
-capacity (con-current usage loads), and timing (hard real-time
+capacity (concurrent usage loads), and timing (hard real-time
 demands). \wss{What is the difference between speed and timing?}
@@
   including accuracy from the IEEE definition.  Is that on purpose?}
+\pmi{We have decided not to include accuracy. SS will find difference b/w speed and timing from a colleague.}
 
@@
   \citet{VanVliet2000}.)  \wss{Once we have verified McCall said this, we can
-    remove the VanVliet reference.}
+    remove the VanVliet reference.} \pmi{Yes this is almost exactly from McCall, except he did not have the word "The" at the beginning.}
 \end{defn}
@@
 \begin{defn}
-  Effort required to locate and fix an error in a program
-  \citep{pressman2005software}. \wss{Peter, if this is from
-    McCall, we should just cite the primary source and remove the pressman
-    reference.}
-\end{defn}
-
-\begin{defn}
   A set of attributes that bear on the effort needed to make specified
@@
 \begin{defn}
-  The effort required to locate and fix an error in an operational program
-  \citep{McCallEtAl1977}. (As summarized in \citet{VanVliet2000}.) \wss{It looks
+  Effort required to locate and fix an error in an operational program
+  \citep{McCallEtAl1977}. \wss{It looks
     like \citet{pressman2005software} was using \citet{McCallEtAl1977} for their
@@
     reference in  preference of the McCall definition.}
+\pmi{I have verified the McCall definition and removed the Pressman and van Vliet references above. I will leave these comments until all similar definitions in this document have been edited.}
 \end{defn}
@@
 
-Combined Definition \ref{MaintainabilityDefnSelected1} and Definition
-\ref{MaintainabilityDefnSelected2}: The ease with which a software system or
-component can be modified to correct faults, improve performance or other
-attributes, or satisfy new requirements. \wss{We didn't agree as a group in the
-  last meeting on our ``measurement'' term, but I think ease is difficult to
-  measure.  Maybe we should rephrase to measuring the effort required?}
+Combined altered Definition \ref{MaintainabilityDefnSelected1} and Definition
+\ref{MaintainabilityDefnSelected2}: The effort with which a software system or
+component can be modified to:
+
+\begin{enumerate}
+	\item correct faults
+	\item improve performance or other
+	attributes
+	\item satisfy new requirements
+\end{enumerate}
 
@@
 \begin{defn}
-  Extent to which a program [or parts of a program] can be reused in other
-  applications - related to the packaging and scope of the functions that the
-  program performs \citep{pressman2005software}. \wss{Peter, if this is from
-    McCall, we should just cite the primary source and remove the pressman
-    reference.}
+  Extent to which a program can be used in other applications - related to the packaging and scope of the functions that programs perform \citep{McCallEtAl1977}. 
 \end{defn}
@@
 \begin{defn}
-  The extent to which a program (or parts thereof) can be reused in other
-  applications \citep{McCallEtAl1977}. (As summarized in \citet{VanVliet2000}.)
-  \wss{Peter, if this is from McCall, we should just cite the primary source and
-    remove the VanVliet reference.}
-\end{defn}
-
-\begin{defn}
   A product is reusable if we can ``use the product - perhaps with minor changes
@@
 
-\subsection{\notdone{Portability} \pmi{owner}}
+\subsection{{Portability} \pmi{owner}}
 
@@
 
-\begin{defn} \label{PortabilityDefnSelected} Effort required to transfer the
-  program from one hardware and/or software system environment to another
-  \citep{pressman2005software}.  \wss{Peter, if this is from McCall, we should
-    just cite the primary source and remove the pressman reference.}
+\begin{defn} \label{PortabilityDefnSelected} Effort required to transfer a program from one hardware configuration and/or software system environment to another \citep{McCallEtAl1977}.
 \end{defn}
@@
 \begin{defn}
-  ``The effort required to transfer a program from one hardware and/or software
-  environment to another'' \citep{McCallEtAl1977}. (As summarized in
-  \citet{VanVliet2000}.)  \wss{Peter, if this is from McCall, we should just
-    cite the primary source and remove the VanVliet reference.}
-\end{defn}
-
-\begin{defn}
   The capability of software to be transferred from one environment to another
@@
 This is measurable and succinct.
-
-\pmi{I think this is done.}
 
@@
     this is from McCall, we should just cite the primary source and remove the
-    Van Vliet reference.}
+    Van Vliet reference.} \pmi{Yes this is almost exactly from McCall, except he did not have the word "The" at the beginning.}
 \end{defn}
@@
 provide meaningful definitions of the inputs and outputs of the software
-process.
+process. \wss{Where did this definition come from?  Is this all from Boehm?
+  This is quite a long quote.  We might actually want to paraphrase.}
 
@@
   \wss{Ao, if this is from McCall, we should just cite the primary source and
-    remove the Van Vliet reference.}
+    remove the Van Vliet reference.} \pmi{McCall's definition verbatim: Those attributes of the software that provide full implementation of the functions required.}
 \end{defn}
@@
   \citet{VanVliet2000}.)  \wss{Ao, if this is from McCall, we should just cite
-    the primary source and remove the Van Vliet reference.}
+    the primary source and remove the Van Vliet reference.} \pmi{McCall's definition verbatim: Those attributes of the software that provide uniform design and implementation techniques and notation.}
 \end{defn}
@@
 \end{defn} \wss{Once we have verified McCall, we can remove the Van Vliet
-  reference.}
+  reference.}  \pmi{McCall's definition verbatim: Those attributes of the software that provide a thread from the requirements to the implementation with respect to the specific development and operational environment.}
 
@@
 function'' \citep{VanVliet2000}. \wss{Need to verify McCall reference, then
-  delete VanVliet.}
+  delete VanVliet.} \pmi{McCall's definition verbatim: Effort required to test a program to insure it performs its intended function.}
 

### 2020-01-16 StateOfPractice/Peter-Notes.tex
added notes for LBM solutions and metrics/metric readings

\documentclass{article}

\usepackage{tabularx}
\usepackage{booktabs}

\title{Project Notes}

\author{Peter Michalski}

\date{}

\begin{document}

\maketitle

~\newpage

\section{List of LBM solutions}
\begin{enumerate}
	\item aromanro/LatticeBoltzmann

	\item Atruszkowska/LBM\_MATLAB

	\item ch4-project

	\item CPE490\_Lattice\_Boltzmann\_Project

	\item CUDA-LBM-simulator

	\item CudneLB (TCLB)

	\item DL\_Meso

	\item ESPResSo

	\item ESPResSo++

	\item firesim

	\item fvLBM

	\item HemeLB

	\item JavaCFD

	\item JFlowSim

	\item laboetie

	\item LatBo.jl

	\item lettuce

	\item LB2D\_Prime

	\item LB3D

	\item LBDEMcoupling-public

	\item LBSim

	\item Limbes

	\item listLBM

	\item LUMA

	\item MP-LABS

	\item MUPHY

	\item Openlb

	\item Palabos

	\item PyLBM

	\item Sailfish

	\item siramirsaman/LBM

	\item SunlightLB

	\item Taxila LBM

	\item Tensorflow

	\item turbulent\_lbm\_multigpu

	\item waLBerla

	\item wlb

	\item Zmhhaha/LBM-Cplusplus-A.A.Mohamad

\end{enumerate}

~\newpage
\section{Quality Measurement}
\subsection{Robustness}

Does not crash and provides (descriptive) error upon an invalid input.\\

Is software that crashes but provides an error more robust than a program that does not crash but does not provide an error?\\

Mean time to recovery from failures?\\

DOI: 10.1007/978-3-642-29032-9\_16:\\

The goal of robustness testing is to activate those faults (typically design or programming faults) or vulnerabilities in the system that result in incorrect operation, i.e., robustness failure, affecting the resilience of the system. Robustness testing mostly concentrates on the internal design faults activated through the system interface. The robustness failures are typically classified according to the CRASH criteria [540]: Catastrophic (the whole system crashes or reboots), Restart (the application has to be restarted), Abort (the application terminates abnormally), Silent (invalid operation is performed without error signal), and Hindering (incorrect error code is returned–note that returning a proper error code is considered as robust operation).\\

The measure of robustness can be given as the ratio of test cases that exposes robustness faults, or, from the system’s point of view, as the number of robustness faults exposed by a given test suite.\\

1) Injecting Random input 
2) Using invalid input 
3) Invalid inputs for each function in interface (valid and invalid inputs are type specific) 
4) applying mutation techniques 
\subsection{Performance}
How many concurrent loads (this will also require hardware parallelism)\\

How many application instances can be run?\\

CPU usage (monitor this)\\

Average Response Time (compared to other solutions)\\

Error rate (exceptions)\\

MTBF 
\subsection{Maintainability}
How much time it takes to add a new function (requirement)\\

Metrics for Assessing a Software System's Maintainability (Oman, Hagemeister)\\

Software Metrics for Predicting Maintainability (Frappier, Matwin, Mili) 
\subsection{Reusability}
Measuring Software Reusability (Poulin)\\

Chidamber and Kemerer object-oriented metrics:\\

https://www.aivosto.com/project/help/pm-oo-ck.html:\\

eg weighted methods per class, number of children, coupling per class\\

Software reusability metrics estimation: Algorithms, models and optimization techniques (Padhy)\\

Cyclomatic complexity: independent paths through source code\\

Software Reuse and Reusability Metrics and Models (Frakes, Terry)\\

A new reusability metric for object-oriented software (Barnard)\\

A metrics suite for measuring reusability of software components (Washizaki)\\

Reusability Index: A Measure for Assessing Software Assets Reusability (Ampatzoglou) 
\subsection{Portability}
compare effort to port to effort to redevelop\\

determine if the system can be ported: ram, processor, resolution, OS, browser\\

Issues in the Specification and Measurement of Software Portability (Mooney)\\

Designing a Measurement Method for the Portability Non-functional Requirement (Talib)\\

ISO 9126 breaks down portability testing: installability, compatability, adapatability, and replaceability.\\

https://www.softwaretestinghelp.com/what-is-portability-testing/ 
\end{document}

### 2020-01-19 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
added testability

--- 2019-12-18 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
@@
 
-\subsection{{Correctness} \oo{owner}}
+\subsection{\notdone{Correctness} \oo{owner}}
 
@@
   A program is functionally correct if it behaves according to its stated
-  functional specifications \citep{GhezziEtAl2003}.  \wss{Is this verbatim from
-    the original source?  If not, we should show those parts that are verbatim
-    in quotes, so that readers know the parts that were developed by us.}
-  \oo{Yes, verbatim}
-
-\end{defn}
-
-\begin{defn}
-  The degree to which a system is free from defects in its specification, design, and implementation, is its correctness \citep{wilson2009quality}.
-\oo{Need to verify this definition, still searching for this paper}
+  functional specifications \citep{GhezziEtAl2003}.  
 \end{defn}
@@
   The extent to which a program satisfies its specifications and fulfills the
-  user's mission objectives \citep{McCallEtAl1977}. \oo{definition is checked with pdf} \wss{Great!  Did you add the Mccall paper to the pub repo?}
-\oo{I will}
+  user's mission objectives \citep{McCallEtAl1977}. \oo{paper added to repo} 
 \end{defn}
@@
 
-\begin{defn}
-  Correctness from software engineering perspective can be defined as the
-  adherence to the specifications that determine how users can interact with the
-  software and how the software should behave when it is used correctly
-  \citep{Tutorialspoint}.
-\end{defn}
-
-\noindent \textbf{Proposed Definition}
-
-Definition~\ref{Correctness_Selected} rephrased: A program is functionally correct if it behaves according to its stated specifications  \oo{Proposed definition updated to Ghezzi's}
+\noindent \textbf{Proposed Definition}
+
+Definition~\ref{Correctness_Selected} rephrased: A program is correct if it behaves according to its stated specifications
 
@@
 this definition captures it all.
+
+	The degree to which a system is free from defects in its specification, design, and implementation, is its correctness \citep{wilson2009quality}.
+\oo{this book is unavailable to verify this definition}
 
@@
 specified environment and time.
+
+\subsection{\notdone{Testability} \oo{owner}}
+\begin{defn}
+	Code possesses the characteristic testability to the extent that it facilitates the establishment
+	of verification criteria and supports evaluation of its performance \citep{boehm1976quantitative}.
+\end{defn}
 

### 2020-01-19 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
definition clean ups

--- 2020-01-06 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
@@
   A program is functionally correct if it behaves according to its stated
-  functional specifications \citep{GhezziEtAl2003}.  \wss{Is this verbatim from
-    the original source?  If not, we should show those parts that are verbatim
-    in quotes, so that readers know the parts that were developed by us.}
-  \oo{Yes, verbatim}
-
-\end{defn}
-
-\begin{defn}
-  The degree to which a system is free from defects in its specification, design, and implementation, is its correctness \citep{wilson2009quality}.
-\oo{Need to verify this definition, still searching for this paper}
+  functional specifications \citep{GhezziEtAl2003}.  	
+\end{defn}
+
+\begin{defn}
+  The degree to which a system is free from faults in its specification, design, and 
+  implementation \citep{mcconnell2004code}.
 \end{defn}
@@
   The extent to which a program satisfies its specifications and fulfills the
-  user's mission objectives \citep{McCallEtAl1977}. \oo{definition is checked with pdf} \wss{Great!  Did you add the Mccall paper to the pub repo?}
-\oo{I will}
+  user's mission objectives \citep{McCallEtAl1977}. 
 \end{defn}
@@
 
-\begin{defn}
-  Correctness from software engineering perspective can be defined as the
-  adherence to the specifications that determine how users can interact with the
-  software and how the software should behave when it is used correctly
-  \citep{Tutorialspoint}.
-\end{defn}
-
-\noindent \textbf{Proposed Definition}
-
-Definition~\ref{Correctness_Selected} rephrased: A program is functionally correct if it behaves according to its stated specifications  \oo{Proposed definition updated to Ghezzi's}
+\noindent \textbf{Proposed Definition}
+
+Definition~\ref{Correctness_Selected} rephrased: A program is correct if it
+behaves according to its stated specifications.
 
@@
 \begin{defn} \label{Defn_Verifiability1}
-Verifiability ``means that you should be able to write a set of tests that can demonstrate that the delivered system meets each specified requirement'' \citep{sommerville}.	
-\end{defn}
-
-\noindent \textbf{Proposed Definition}
-
-Definition~\ref{Defn_Verifiability1}:
-Verifiability means that you should be able to write a set of tests that can demonstrate that the delivered system meets each specified requirement.
+Verifiability ``means that you should be able to write a set of tests that can
+demonstrate that the delivered system meets each specified requirement'' \citep{sommerville}.	
+\end{defn}
+
+\begin{defn}
+	The evaluation of whether or not a product, service, or result complies with
+	a regulation, requirement, specification, or imposed condition \citep{project2017guide}
+	\oo {verification definition}
+\end{defn}
+
+\begin{defn}
+	By verification, we mean all activities that are undertaken to ascertain that the
+	software meets its objectives \citep{GhezziEtAl2003}.
+\end{defn}
+\noindent \textbf{Proposed Definition}
+
+Definition~\ref{Defn_Verifiability1} rephrased:
+Verifiability is the extent to which a set of tests can be written and executed, to demonstrate
+that the delivered system meets each specified requirement.
 
@@
 \begin{defn}
-  Validation refers to a different set of tasks that ensure that the software
-  that has been built is traceable to customer requirements.{to be removed, its from geeksforgeeks}
+  The assurance that a product, service, or result meets the needs of the customer and other identified
+  stakeholders \citep{project2017guide}.
 \end{defn}
@@
 \noindent \textbf{Reasoning}
+
+\subsection{\notdone{Testability} \oo{owner}}
+
+\begin{defn}
+	The degree to which you can unit-test and system-test a system;
+	the degree to which you can verify that the system meets its requirements \citep{mcconnell2004code}.
+\end{defn}
+
+\begin{defn}
+	Code possesses the characteristic testability to the extent that it
+    facilitates the establishment of verification criteria and supports
+    evaluation of its performance \citep{boehm1976quantitative}
+\end{defn}
 

### 2020-01-19 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
updated validatin defn

@@
 \end{defn}
-
-\begin{defn}
-  Requirements validation examines the specification to ensure that all software
-  requirements have been stated unambiguously; that inconsistencies, omissions,
-  and errors have been detected and corrected; and that the work products
-  conform to the standards established for the process, the project, and the
-  product \citep{pressman2005software}.  \wss{Olu, is this definition
-    originally from pressman, or should we track down the original source?}
+\begin{defn}
+	Software validation is the process of checking that the system conforms to its specification and
+	that it meets the real needs of the users of the system \citep{sommerville}.
+
+\end{defn}
+\begin{defn}
+	Software validation, where the software is checked to ensure that it is what the
+	customer requires \citep{sommerville}.
+\end{defn}
+\begin{defn}
+  Software validation is achieved through a series of tests that demonstrate conformity
+  with requirements \citep{pressman2005software}.
+\end{defn}
+
+\begin{defn}
+	Validation is concerned with establishing that the product fulfills its
+	intended use \citep{Van vilet}.
 \end{defn}

### 2020-01-19 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
update contact for Ao
update interoperability

@@
 \author{Peter Michalski}{McMaster University, Canada}{michap@mcmaster.ca}{}{}
-\author{Ao Dong}{McMaster University, Canada}{}{}{}
+\author{Ao Dong}{McMaster University, Canada}{donga9@mcmaster.ca}{}{}
 
@@
 
-\subsection{\done{Interoperability} \ad{owner}}
-
-\begin{defn}
-  ``The effort required to couple one system with another''
-  \citep{McCallEtAl1977}. (As summarized in \citet{VanVliet2000}.)  \wss{Ao, if
+\subsection{\notdone{Interoperability} \ad{owner}}
+
+\begin{defn}
+  Effort required to couple one system with another
+  \citep{McCallEtAl1977}. \wss{Ao, if
     this is from McCall, we should just cite the primary source and remove the
     Van Vliet reference.} \pmi{Yes this is almost exactly from McCall, except he did not have the word "The" at the beginning.}
+    \ad{Someone with access to the book added it. I found the book in the pub folder and fixed it.}
 \end{defn}

### 2020-01-19 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
update interoperability
update visibility

@@
     Van Vliet reference.} \pmi{Yes this is almost exactly from McCall, except he did not have the word "The" at the beginning.}
-    \ad{Someone with access to the book added it. I found the book in the pub folder and fixed it.}
+    \ad{Someone with access to the book added it. I found the book in the pub
+    folder and fixed it.}
 \end{defn}
@@
 Definition \ref{VisibilitySelected} rephrased: The extent to which all of the
-steps \wss{What are all the steps?} and the current status of a software
-development process are documented clearly. \wss{I wonder if rather than
+steps of a software development process and the current status of it are 
+conveyed clearly. \wss{I wonder if rather than
   ``documented clearly'', we should be more abstract and talk about how these

### 2020-01-19 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
update Productivity
update completeness

@@
   This is quite a long quote.  We might actually want to paraphrase.}
+\ad{all from Boehm. I feel the best way is deleting everything after the 
+equation}
 
@@
 \begin{defn}
-  The degree to which a full implementation of the required functionality has
-  been achieved \citep{McCallEtAl1977}. (As summarized in \citet{VanVliet2000}.)
-  \wss{Ao, if this is from McCall, we should just cite the primary source and
-    remove the Van Vliet reference.} \pmi{McCall's definition verbatim: Those attributes of the software that provide full implementation of the functions required.}
+  Those attributes of the software that provide full implementation of the 
+  functions required. \citep{McCallEtAl1977}. \wss{Ao, if this is from McCall, 
+  we should just cite the primary source and remove the Van Vliet reference.} 
+  \pmi{McCall's definition verbatim: Those attributes of the software that 
+  provide full implementation of the functions required.} \ad{Originally added 
+  by someone else so I didn't have access to the book. But now it's available 
+  and I double checked, Peter got it accurately.}
 \end{defn}

### 2020-01-19 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
update consistency

@@
 \begin{defn}
-  The use of uniform design and implementation techniques and notations
-  throughout a project \citep{McCallEtAl1977}. (As summarized in
-  \citet{VanVliet2000}.)  \wss{Ao, if this is from McCall, we should just cite
-    the primary source and remove the Van Vliet reference.} \pmi{McCall's definition verbatim: Those attributes of the software that provide uniform design and implementation techniques and notation.}
+  Those attributes of the software that provide uniform design and 
+  implementation techniques and notation \citep{McCallEtAl1977}.  \wss{Ao, if 
+  this is from McCall, we should just cite
+    the primary source and remove the Van Vliet reference.} \pmi{McCall's 
+    definition verbatim: Those attributes of the software that provide uniform 
+    design and implementation techniques and notation.}\ad{Originally added 
+  by someone else so I didn't have access to the book. But now it's available 
+  and I double checked, Peter got it accurately.}
 \end{defn}

### 2020-01-20 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
updated more definitions

@@
 \end{defn}
+
+\begin{defn}
+	 Verification is the process of evaluating a system or component to determine whether the products
+	 of a given development phase satisfy the conditions imposed at the start of that phase \citep{IEEEStdGlossarySET1990}.
+\end{defn}
 \noindent \textbf{Proposed Definition}
@@
 Verifiability is the extent to which a set of tests can be written and executed, to demonstrate
-that the delivered system meets each specified requirement.
+that the delivered system meets the specification.
 
@@
 Definition is concise and measurable.  Verifiability involves solving the
-equations right \citep[p.~23]{Roache1998}.
+equations right \citep[p.~23]{Roache1998} and the definition states that system needs to be verifiable based on the specifications..
 
@@
   Validation means solving the right equations \citep[p.~23]{Roache1998}.
-\end{defn}
-
-\begin{defn} \label{Validatability_Selected} 
-  Validation is the process of evaluating software during or at the end of the
-  development process to determine whether it satisfies specified business
-  requirements \citep{softwaretestingfundamentals}.  \wss{Please use the
-  author's name for this citation.  If that isn't known, use the name of the
-  organization that published this material.} \oo{This is a weblink, that's the organization's name I guess}
 \end{defn}
@@
 	Validation is concerned with establishing that the product fulfills its
-	intended use \citep{Van vilet}.
-\end{defn}
-
-\begin{defn}
-  Software validation checks that the software product satisfies or fits the
-  intended use (high-level checking), i.e., the software meets the user
-  requirements, not as specification artifacts or as needs of those who will
-  operate the software only; but, as the needs of all the stakeholders (such as
-  users, operators, administrators, managers, investors, etc.)
-  \wss{This reference is not appearing in the Reference
-  list.  We also don't really want a Wikipedia citation in a scholarly paper.
-  Can you please find the source where this definition originally came from?}\oo{Could not find a different source from Wikipedia}
-\end{defn}
-
-\noindent \textbf{Proposed Definition}
-
-Definition~ \oo{need to discuss this quality, we might just define
-  validatability from validation definitions above? Resources are limited for
-  'validatability'}.
-
-\noindent \textbf{Reasoning}
-
+	intended use \citep{van2008software}.
+\end{defn}
+
+\begin{defn} \label{Defn_Validatability}
+  The process of evaluating a system or component during or at the end of the
+  development process to determine whether it satisfies specified requirements
+  \citep{IEEEStdGlossarySET1990}.
+\end{defn}
+
+\noindent \textbf{Proposed Definition}
+
+Definition \ref{Defn_Validatability} rephrased:  The effort required in evaluating
+a system or component during or at the end of the development process to determine
+whether it satisfies specified requirements
+
+\noindent \textbf{Reasoning}
+
+Definition describes that system should be measurable through the development life cycle,
+It is concise and can be measured.
 \subsection{\notdone{Testability} \oo{owner}}
@@
 
+\begin{defn}
+	Effort required to test a program to ensure it performs
+	its intended function \citep{McCallEtAl1977}.
+\end{defn}
+
+\begin{defn} \label{Defn_Testability}
+	 The degree to which a system or component facilitates the establishment of
+	 test criteria and the performance of tests to determine whether those
+	 criteria have been met \citep{IEEEStdGlossarySET1990}.
+\end{defn}
+
+\noindent \textbf{Proposed Definition}
+
+Definition \ref{Defn_Testability}: The degree to which a system or component
+ facilitates the establishment of test criteria and the performance of tests
+ to determine whether those criteria have been met.
+
+\noindent \textbf{Reasoning}
+
+Definition is concise and measurable
+
 \subsection{\notdone{Reliability} \oo{owner}}
@@
   Extent to which a program can be expected to perform its intended function
-  with required precision \citep{McCallEtAl1977}. \oo{Definition check pass}
+  with required precision \citep{McCallEtAl1977}.
 \end{defn}
@@
 \begin{defn} \label{reliabilitySelected2} Reliability can be estimated as ``the
-  average time interval between two faliures, also called the mean time to
-  faliure (MTTF). Clearly, MTTF (t) = 1/FI (t)''. Where FI (t) is the faliure
-  intensity - the number of faliures per unit time \citep{GhezziEtAl2003}.
+  average time interval between two failures, also called the mean time to
+  failure (MTTF). Clearly, MTTF (t) = 1/FI (t)''. Where FI (t) is the failure
+  intensity - the number of failures per unit time \citep{GhezziEtAl2003}.
 \end{defn}
@@
 environment for a specified time, i.e. the average time interval between two
-faliures also known as the mean time to failure (MTTF).
+failures also known as the mean time to failure (MTTF).
 

### 2020-01-20 StateOfPractice/Methodology/Methodology.tex
Initial draft of document for recording our "state of the practice" measurement methodology

--- 2020-01-19 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
@@
 
-\title{Quality Definitions of Qualities} 
+\title{Methodology for Assessing the State of the Practice for Domain X} 
 \author{Spencer Smith}{McMaster University, Canada}{smiths@mcmaster.ca}{}{}
@@
 \author{Peter Michalski}{McMaster University, Canada}{michap@mcmaster.ca}{}{}
-\author{Ao Dong}{McMaster University, Canada}{donga9@mcmaster.ca}{}{}
+\author{Ao Dong}{McMaster University, Canada}{}{}{}
 
@@
 
-The presentation is divided into two main sections: i) qualities that apply to
-software products, software artifacts and software development processes, and
-ii) qualities that are considered important for good specifications.  The
-specification could be a specification of requirements, design or a test plan.
+\section{Domain Analysis} \label{SecDomainAnalysis}
 
-The pattern in the presentation for each of the qualities is the same.  First we
-summarize all of the definitions that we could find in the literature.  To keep
-the clutter of quotation marks down, we have adopted the convention that each
-definition is given verbatim from the cited source, but without showing
-quotation marks.  In cases where the definition has to be rephrased, quotation
-marks are used to show those portions that are taken verbatim for the original
-source.  After the summary of the existing definitions, we propose the
-definition that we would like to work with going forward.  This definition can
-either be our preference from the existing definitions, or a new definition,
-which is often found by combining existing definitions.  Following the proposed
-definition is an explanation for the reasoning that led to this choice.
+Commonality analysis.  Follow as for mesh generator (likely with less detail).
 
-In determining our recommended definition for each quality, we used the
-following criteria:
+\section{Quality Measures}
 
-\begin{description}
-\item[Consistency] The final list of definitions should be consistent in the
-  terminology used.  For instance, we do not use synonyms for variety; we use
-  the same word to mean the same concept throughout the final list of
-  definitions.  This is why measurement is given by ``the effort required..''
-  and not the various synonyms for this that are seen in the literature. \wss{We
-    may change this phrase we use for measuring something.  I just wanted to put
-    this in here as a placeholder.}
-\item[Generality]
-\item[Measurability]
-\item[Etc.]
-\end{description}
+Go through each quality and describe how it will be measured.
 
-\section{Qualities of Software Products, Artifacts and
-	Processes} \label{SecQualities}
+Start with decided upon definition.  (If discussion is on-going, please select
+the ``best'' definition, but include a warning that it may change.)
 
-To assess the current state of software development, and to understand how
-future changes impact software development, we need a clear definition of what
-we mean by quality.  The concept of quality is decomposed into a set of separate
-components that together make up ``quality''. Unfortunately, these are
-called \emph{qualities}. These are associated to the software product,
-the software artifacts (documentation, test cases, etc) and to the software
-development process itself, and combinations thereof.
+\subsection{\notdone{Installability} \oo{owner}}
 
-Our analysis is centred around a set of software qualities.  Quality is not
-considered as a single measure, but a collection of different qualities, often
-called ``ilities.''  These qualities highlight the desirable nonfunctional
-properties for software artifacts, which include both documentation and
-code. Some qualities, such as visibility and productivity, apply to the process
-used for developing the software. The following list of qualities is based on
-\cite{GhezziEtAl2003}. To the list from \cite{GhezziEtAl2003}, we have added
-three qualities important for SC: installability, reproducibility and
-sustainability.
+\subsection{\notdone{Correctness} \oo{owner}}
 
-\subsection{{Installability} \oo{owner}}
-
-\begin{defn}
-  The capability of the software product to be installed in a specified
-  environment \citep{ISO9126}.
-\end{defn}
-
-\begin{defn} \label{Installability_Selected2}
-	The degree of effectiveness and efficiency with which a product or system can
-	be successfully installed and/or uninstalled in a specified environment
-	\citep{ISO/IEC25010}.
-\end{defn}
-
-\begin{defn} \label{Installability_Selected1} Installability refers to the cost
-  or effort required for the installation, given an installation is possible to
-  begin with \citep{lenhard2013measuring}.
-\end{defn}
-
-\noindent \textbf{Proposed Definition}
-
-Combined definitions \ref{Installability_Selected2} and
-\ref{Installability_Selected1} rephrased.  Installability refers to the
-effort required for the installation, uninstallation or reinstallation of a
-software or product in a specified environment.
-
-\noindent \textbf{Reasoning}
-
- The definition captures a vital metric of measure for this quality - effort, and considers not just the installation but uninstallation and re-installation which are all related to installation activities.
-\oo{Done installability}
-
-\subsection{{Correctness} \oo{owner}}
-
-\begin{defn} \label{Correctness_Selected}
-  A program is functionally correct if it behaves according to its stated
-  functional specifications \citep{GhezziEtAl2003}.  	
-\end{defn}
-
-\begin{defn}
-  The degree to which a system is free from faults in its specification, design, and 
-  implementation \citep{mcconnell2004code}.
-\end{defn}
-
-\begin{defn}
-  The ability of software products to perform their exact tasks, as defined by
-  their specification \citep{meyer1988object}.
-\end{defn}
-
-\begin{defn} 
-  The extent to which a program satisfies its specifications and fulfills the
-  user's mission objectives \citep{McCallEtAl1977}. 
-\end{defn}
-
-\begin{defn}  
-  The degree to which a system or component is free from faults in its
-  specification, design and implementation \citep{IEEEComputerDictionary1991}.
-\end{defn}
-
-\noindent \textbf{Proposed Definition}
-
-Definition~\ref{Correctness_Selected} rephrased: A program is correct if it
-behaves according to its stated specifications.
-
-\noindent \textbf{Reasoning}
-
-To determine whether a program is correct, we must be able to refer to both its
-specification, design and implementation to understand the intended purpose and
-this definition captures it all.
-
-\subsection{\notdone{Verifiability} \oo{owner}}
-
-\begin{defn}
-	Verification involves solving the equations right \citep[p.~23]{Roache1998}.
-\end{defn}
-
-\begin{defn}
-  A software is verifiable if its properties can be verified easily
-  \citep{GhezziEtAl2003}.  \wss{Is this a direct quote?} \oo{yes}
-\end{defn}
-
-\begin{defn} \label{Defn_Verifiability1}
-Verifiability ``means that you should be able to write a set of tests that can
-demonstrate that the delivered system meets each specified requirement'' \citep{sommerville}.	
-\end{defn}
-
-\begin{defn}
-	The evaluation of whether or not a product, service, or result complies with
-	a regulation, requirement, specification, or imposed condition \citep{project2017guide}
-	\oo {verification definition}
-\end{defn}
-
-\begin{defn}
-	By verification, we mean all activities that are undertaken to ascertain that the
-	software meets its objectives \citep{GhezziEtAl2003}.
-\end{defn}
-\noindent \textbf{Proposed Definition}
-
-Definition~\ref{Defn_Verifiability1} rephrased:
-Verifiability is the extent to which a set of tests can be written and executed, to demonstrate
-that the delivered system meets each specified requirement.
-
-\noindent \textbf{Reasoning}
-
-Definition is concise and measurable.  Verifiability involves solving the
-equations right \citep[p.~23]{Roache1998}.
+\subsection{\notdone{Verifiability/Testability} \oo{owner}}
 
@@
 
-\begin{defn}
-  Validation means solving the right equations \citep[p.~23]{Roache1998}.
-\end{defn}
-
-\begin{defn} \label{Validatability_Selected} 
-  Validation is the process of evaluating software during or at the end of the
-  development process to determine whether it satisfies specified business
-  requirements \citep{softwaretestingfundamentals}.  \wss{Please use the
-  author's name for this citation.  If that isn't known, use the name of the
-  organization that published this material.} \oo{This is a weblink, that's the organization's name I guess}
-\end{defn}
-
-\begin{defn}
-  The assurance that a product, service, or result meets the needs of the customer and other identified
-  stakeholders \citep{project2017guide}.
-\end{defn}
-\begin{defn}
-	Software validation is the process of checking that the system conforms to its specification and
-	that it meets the real needs of the users of the system \citep{sommerville}.
-
-\end{defn}
-\begin{defn}
-	Software validation, where the software is checked to ensure that it is what the
-	customer requires \citep{sommerville}.
-\end{defn}
-\begin{defn}
-  Software validation is achieved through a series of tests that demonstrate conformity
-  with requirements \citep{pressman2005software}.
-\end{defn}
-
-\begin{defn}
-	Validation is concerned with establishing that the product fulfills its
-	intended use \citep{Van vilet}.
-\end{defn}
-
-\begin{defn}
-  Software validation checks that the software product satisfies or fits the
-  intended use (high-level checking), i.e., the software meets the user
-  requirements, not as specification artifacts or as needs of those who will
-  operate the software only; but, as the needs of all the stakeholders (such as
-  users, operators, administrators, managers, investors, etc.)
-  \wss{This reference is not appearing in the Reference
-  list.  We also don't really want a Wikipedia citation in a scholarly paper.
-  Can you please find the source where this definition originally came from?}\oo{Could not find a different source from Wikipedia}
-\end{defn}
-
-\noindent \textbf{Proposed Definition}
-
-Definition~ \oo{need to discuss this quality, we might just define
-  validatability from validation definitions above? Resources are limited for
-  'validatability'}.
-
-\noindent \textbf{Reasoning}
-
-\subsection{\notdone{Testability} \oo{owner}}
-
-\begin{defn}
-	The degree to which you can unit-test and system-test a system;
-	the degree to which you can verify that the system meets its requirements \citep{mcconnell2004code}.
-\end{defn}
-
-\begin{defn}
-	Code possesses the characteristic testability to the extent that it
-    facilitates the establishment of verification criteria and supports
-    evaluation of its performance \citep{boehm1976quantitative}
-\end{defn}
-
 \subsection{\notdone{Reliability} \oo{owner}}
 
-\begin{defn}
-  Informally, software is reliable if the user can depend on it \citep{GhezziEtAl2003}.
-\end{defn}
-
-\begin{defn}
- The capability of the software product to maintain a specified level of performance when used under specified conditions \cite{ISO9126}.
-\end{defn}
-\begin{defn} \label{reliabilitySelected1} It is the probability of failure-free
-  operation of a computer program in a specified environment for a specified
-  time \citep{musa1987software}.
-\end{defn}
-
-\begin{defn}
-  Code possesses the characteristic reliability to the extent that it can be
-  expected to perform its intended functions satisfactorily
-  \citep{boehm1976quantitative}.
-\end{defn}
-
-\begin{defn}
-  Reliability expresses the ability of the software to maintain a specified
-  level of fault tolerance, when used under specified condition
-  \citep{singh2013different}.
-\end{defn}
-
-\begin{defn}
-  Extent to which a program can be expected to perform its intended function
-  with required precision \citep{McCallEtAl1977}. \oo{Definition check pass}
-\end{defn}
-
-\begin{defn}
-  Informally, the reliability of a system is the probability, over a given
-  period of time, that the system will correctly deliver services as expected by
-  the user \citep{sommerville}.
-\end{defn}
-
-\begin{defn}
-  The ability of a system or component to perform its required function under
-  stated conditions for a specified period of time
-  \citep{IEEEComputerDictionary1991}.
-\end{defn}
-
-\begin{defn}
-  The ability of a system to perform its required functions under stated
-  conditions whenever required—having a long mean time between failures
-  \citep{mcconnell2004code}.
-\end{defn}
-\begin{defn} \label{reliabilitySelected2} Reliability can be estimated as ``the
-  average time interval between two faliures, also called the mean time to
-  faliure (MTTF). Clearly, MTTF (t) = 1/FI (t)''. Where FI (t) is the faliure
-  intensity - the number of faliures per unit time \citep{GhezziEtAl2003}.
-\end{defn}
-\oo{Statistical definition from  \citep{GhezziEtAl2003} added}
-
-\noindent \textbf{Proposed Definition}
-
-Definition \ref{reliabilitySelected1} and \ref{reliabilitySelected2} rephrased:
-Probability of failure-free operation of a computer program in a specified
-environment for a specified time, i.e. the average time interval between two
-faliures also known as the mean time to failure (MTTF).
-
-\wss{Needs to be completed.}\oo{Done}
-
-\noindent \textbf{Reasoning}
-
-Reliability is defined such that the measurement metrics are visible - the
-specified environment and time.
-
-\subsection{{Robustness} \pmi{owner}}
-\begin{defn}
-	The degree to which a system or component can function correctly in the
-	presence of invalid inputs or stressful environmental conditions
-	\citep{IEEEStdGlossarySET1990}.
-\end{defn}
-\begin{defn}
-	The quality can be further informally refined as the ability of a software to
-	keep an acceptable behaviour, expressed in terms of robustness requirements,
-	in
-	spite of exceptional or unforeseen execution conditions (such as the
-	unavailability of system resources, communication failures, invalid or
-	stressful inputs, etc.) \citep{fernandez2005model}.
-\end{defn}
-\begin{defn} \label{RobustnessPartialDefn}
-	Code possesses the characteristic of robustness to the extent that it can
-	continue to perform despite some violation of the assumptions in its
-	specification \citep{boehm2007software}.
-\end{defn}
-\begin{defn} \label{RobustnessDefnSelected}
-	A program is robust if it behaves ``reasonably'', even in circumstances that
-	were not anticipated in the requirements specification - for example, when it
-	encounters incorrect input data or some hardware malfunction
-	\citep{ghezzi1991fundamentals}.
-\end{defn}
-
-\noindent \textbf{Proposed Definition}
-
-Definition \ref{RobustnessPartialDefn} and Definition
-\ref{RobustnessDefnSelected} rephrased: Software possesses the characteristic of
-robustness if it behaves ``reasonably'' in two situations: i) when it encounters
-circumstances not anticipated in the requirements specification; and ii) when
-the assumptions in its requirements specification are violated.
-
-\noindent \textbf{Reasoning}
-
-This definition indicates that robustness is related to the quality of
-correctness (both within and outside of it).
-
-\pmi{Still may want to refine what we mean by reasonable}
+\subsection{\notdone{Robustness} \pmi{owner}}
 
@@
 
-\begin{defn} \label{PerformanceDefnSelected}
-	The degree to which a system or component accomplishes its designated functions
-	within given constraints, such as speed, accuracy, or memory usage
-	\citep{IEEEStdGlossarySET1990}.
-\end{defn}
-\begin{defn} \label{PerformanceDefnSelected2}
-	How well or how rapidly the system must perform specific functions. Performance
-	requirements encompass speed (database response times, for instance), throughput
-	(transactions per second), capacity (concurrent usage loads), and timing (hard
-	real-time demands) \citep{wiegers2003softreq}.
-\end{defn}
-\begin{defn}
-	In software engineering we often equate performance with efficiency. A software system is efficient if it uses computing resources economically
-	\citep{ghezzi1991fundamentals}.
-\end{defn}
-
-\noindent \textbf{Proposed Definition}
-
-Combined definition \ref{PerformanceDefnSelected} and
-\ref{PerformanceDefnSelected2}: The degree to which a system or component
-accomplishes its designated functions within given constraints, such as speed
-(database response times, for instance), throughput (transactions per second),
-capacity (concurrent usage loads), and timing (hard real-time
-demands). \wss{What is the difference between speed and timing?}
-
-\noindent \textbf{Reasoning}
-
-This definition offers a comprehensive list of constraints that are commonly
-associated with software performance, such as speed, throughput, capacity, and
-timing.
-
-\pmi{I will check if Wiegers discusses the difference between speed and timing.}
-\pmi{Wiegers did not discuss this difference.}
-\pmi{Speed - change over time; timing - related to specification: when something
-	should happen in relation to something else - state transitions}
-\wss{If we need both speed and timing, we should define them.  We aren't
-  including accuracy from the IEEE definition.  Is that on purpose?}
-\pmi{We have decided not to include accuracy. SS will find difference b/w speed and timing from a colleague.}
-
 \subsection{\notdone{Usability} \jc{owner}}
-
-\begin{defn}
-  The extent to which a product can be used by specified users to achieve
-  specified goals with effectiveness, efficiency, and satisfaction in a
-  specified context of use.
-\end{defn}
-ISO defines usability as
-
-Nielsen and (separately) Schneidermann have defined usability as part of
-usefulness and
-is composed of:
-\begin{itemize}
-\item Learnability: How easy is it for users to accomplish basic tasks the first
-  time they encounter the design?
-\item Efficiency: Once users have learned the design, how quickly can they
-  perform tasks?
-\item Memorability: When users return to the design after a period of not using
-  it, how easily can they re-establish proficiency?
-\item Errors: How many errors do users make, how severe are these errors, and
-  how easily can they recover from the errors?
-\item Satisfaction: How pleasant is it to use the design?
-\end{itemize}
-In that context, it makes sense to separate \emph{usefulness} into
-\emph{usability} (purely an interface concern) and \emph{utility} (in the
-economics sense of the word).
-
-There are two ISO standards covering this, namely ISO/TR 16982:202 and ISO9241.
-
-The Interaction Design Foundation \url{https://www.interaction-design.org/literature/topics/usability} further
-lists the following desirable outcomes:
-
-\begin{enumerate}
-\item It should be easy for the user to become familiar with and competent in
-  using the user interface during the first contact with the website. For
-  example, if a travel agent’s website is a well-designed one, the user should
-  be able to move through the sequence of actions to book a ticket quickly.
-\item It should be easy for users to achieve their objective through using the
-  website. If a user has the goal of booking a flight, a good design will guide
-  him/her through the easiest process to purchase that ticket.
-\item It should be easy to recall the user interface and how to use it on
-  subsequent visits. So, a good design on the travel agent’s site means the user
-  should learn from the first time and book a second ticket just as easily.
-\end{enumerate}
-
-One core reference, for definitions and metrics, is
-\citet{bevan1995measuring}.
-
-\begin{defn}
-  ``The effort required to learn, operate, prepare input, and interpret output
-  of a program'' \citep{McCallEtAl1977}. (As summarized in
-  \citet{VanVliet2000}.)  \wss{Once we have verified McCall said this, we can
-    remove the VanVliet reference.} \pmi{Yes this is almost exactly from McCall, except he did not have the word "The" at the beginning.}
-\end{defn}
-
-\begin{defn}
-  ``A software is usable - or user friendly - if its human users find it easy
-  to use'' \citep{ghezzi1991fundamentals}.
-\end{defn}
-
-\begin{defn}
-  The capability of the software to be understood, learned, used and liked by
-  the user, when used under specified conditions \cite{ISO9126}.
-\end{defn}
-\noindent \textbf{Proposed Definition}
-
-\wss{Still needs to be completed}
-
-\noindent \textbf{Reasoning}
-
-\wss{Needs to be completed}
 
@@
 
-\begin{defn} \label{MaintainabilityDefnSelected1} 
-  The ease with which a software system or component can be modified to correct
-  faults, improve performance or other attributes, or adapt to a changed
-  environment \citep{IEEEStdGlossarySET1990}.
-\end{defn}
+\subsection{\notdone{Reusability} \pmi{owner}}
 
-\begin{defn}
-  ISO/IEC 25010 refers to maintainability as the degree of effectiveness and
-  efficiency with which a product or system can be modified by the intended
-  maintainers \citep{ISO/IEC25010}.
-\end{defn}
-
-\begin{defn}
-  A set of attributes that bear on the effort needed to make specified
-  modifications (which may include corrections, improvements, or adaptations of
-  software to environmental changes and changes in the requirements and
-  functional specifications) \citep{pfleeger2006software}.
-\end{defn}
-
-\begin{defn}
-  We will view maintainability as two separate qualities: repairability and
-  evolvability. Software is repairable if it allows the fixing of defects; it is
-  evolvable if it allows changes that enable it to satisfy new requirements
-  \citep{ghezzi1991fundamentals}.
-\end{defn}
-
-\begin{defn} \label{MaintainabilityDefnSelected2} 
-  Code possesses the characteristic of maintainability to the extent that it
-  facilitates updating to satisfy new requirements or to correct deficiencies
-  \citep{boehm2007software}.
-\end{defn}
-
-\begin{defn}
-  Effort required to locate and fix an error in an operational program
-  \citep{McCallEtAl1977}. \wss{It looks
-    like \citet{pressman2005software} was using \citet{McCallEtAl1977} for their
-    definition.} \wss{PM - can you please look into whether
-    \citet{pressman2005software} cite \citet{McCallEtAl1977}?  Do they give a
-    reason for dropping the word operational?}  \pmi{Yes,
-    \citet{pressman2005software} does cite the relevant section of the textbook
-    with \citet{McCallEtAl1977}. While they do not give a reason for
-    specifically dropping the word operational, they do indicate that their
-    given definition ``is a very limited definition''. Perhaps their intent was
-    breadth.} \wss{Peter, please verify the McCall reference definition, and
-    then just cite that.  Olu has found the McCall reference, so you shouldn't
-    have to search for it.  We can remove both the VanVliet and Pressman
-    reference in  preference of the McCall definition.}
-\pmi{I have verified the McCall definition and removed the Pressman and van Vliet references above. I will leave these comments until all similar definitions in this document have been edited.}
-\end{defn}
-
-\begin{defn}
-  The capability of the software to be modified.  Modifications may include
-  corrections, improvements or adaptation of the software to changes in
-  environment, and in requirements and functional specifications. \cite{ISO9126}
-\end{defn}
-
-\noindent \textbf{Proposed Definition}
-
-Combined altered Definition \ref{MaintainabilityDefnSelected1} and Definition
-\ref{MaintainabilityDefnSelected2}: The effort with which a software system or
-component can be modified to:
-
-\begin{enumerate}
-	\item correct faults
-	\item improve performance or other
-	attributes
-	\item satisfy new requirements
-\end{enumerate}
-
-\noindent \textbf{Reasoning}
-
-This definition offers a comprehensive list of potential reasons for modifying
-software, such as correcting faults, improving performance or other attributes,
-or satisfying new requirements.
-
-\wss{What are the other attributes?  What is an attribute?  Is it the
-	same as a quality?  We are calling performance a quality in this document.}
-\pmi{The term quality attribute has come up often in the literature - qualities
-	appear to be a subset of attributes. Quality attributes seem to be associated
-	with non-functional requirements. The term resource attribute was encountered in
-	a journal and implied an association with a functional requirement. The two
-	might differ along those lines of NFR and FR.}
-
-\pmi{I think this is done}
-
-\subsection{{Reusability} \pmi{owner}}
-
-\begin{defn} 
-  The degree to which a software module or other work product can be used in
-  more than one software system \citep{IEEEStdGlossarySET1990}.
-\end{defn}
-\begin{defn}
-  Extent to which a program can be used in other applications - related to the packaging and scope of the functions that programs perform \citep{McCallEtAl1977}. 
-\end{defn}
-\begin{defn} \label{ReusabilityDefnSelected} 
-  The extent to which a software component can be used with or without
-  adaptation in a problem solution other than the one for which it was
-  originally developed \citep{kalagiakos2003non}.
-\end{defn}
-\begin{defn}
-  Reusability is the likelihood a segment of source code that can be used again
-  to add new functionalities with slight or no modification
-  \citep{sandhu2010survey}.
-\end{defn}
-
-\begin{defn}
-  A product is reusable if we can ``use the product - perhaps with minor changes
-  - to build another product'' \citep{ghezzi1991fundamentals}.
-\end{defn}
-
-\noindent \textbf{Proposed Definition}
-
-Definition \ref{ReusabilityDefnSelected}: The extent to which a software
-component can be used with or without adaptation in a problem solution other
-than the one for which it was originally developed.
-
-\noindent \textbf{Reasoning}
-
-This definition highlights the possible but not necessary adaptation of the
-software component(s) being transferred.
-
-\subsection{{Portability} \pmi{owner}}
-
-\begin{defn}
-  The ease with which a system or component can be transferred from one hardware
-  or software environment to another \citep{IEEEStdGlossarySET1990}.
-\end{defn}
-
-\begin{defn}
-  An application is portable across a class of environments to the degree that
-  the effort required to transport and adapt it to a new environment in the
-  class is less than the effort of redevelopment \citep{mooney1990strategies}.
-\end{defn}
-
-\begin{defn} \label{PortabilityDefnSelected} Effort required to transfer a program from one hardware configuration and/or software system environment to another \citep{McCallEtAl1977}.
-\end{defn}
-
-\begin{defn}
-  A set of attributes that bear on the ability of software to be transferred
-  from one environment to another (including the organizational, hardware, of
-  software environment) \citep{pfleeger2006software}.
-\end{defn}
-
-\begin{defn}
-  Code possesses the characteristic of portability to the extent that it can be
-  operated easily and well on computer configurations other than its current
-  one. This implies that special function features, not easily available at
-  other facilities, are not used, that standard library functions and
-  subroutines are selected for universal applicability, and so on
-  \citep{boehm2007software}.
-\end{defn}
-
-\begin{defn}
-  Portability refers to the ability to run a system on different hardware
-  platforms \citep{ghezzi1991fundamentals}.
-\end{defn}
-
-\begin{defn}
-  The capability of software to be transferred from one environment to another
-  \cite{ISO9126}.
-\end{defn}
-\noindent \textbf{Proposed Definition}
-
-Definition \ref{PortabilityDefnSelected} rephrased: Effort required to transfer
-a program between system environments (including hardware and software).
-
-\noindent \textbf{Reasoning}
-
-This is measurable and succinct.
+\subsection{\notdone{Portability} \pmi{owner}}
 
@@
 
-Understandability is artifact-dependent. What it means for a user-interface
-(graphical or otherwise) to be understandable is wildly different than what it
-means for the code, and even the user documentation.
-
-The literature here is thin and scattered.  More work will need to be done to
-find something useful.
-
-Interestingly, the business literature seems to have taken more care to define
-this.  Here we encounter
-\begin{quote}
-  Understandability is the concept that X should be presented so that a reader
-  can easily comprehend it.
-\end{quote}
-At least this brings in the idea that the \emph{reader} is actively involved,
-and indirectly that the reader's knowledge may be relevant, as well as the
-``clarity of exposition'' of X.
-
-Section 11.2 of \citet{adams2015nonfunctional} does have a full definition.
-
-\begin{defn}
-  The effort ``to uncover the logic of the application''
-  \citep{ghezzi1991fundamentals}.
-\end{defn}
-
-\begin{defn}
-  The capability of the software product to enable the user to understand
-  whether the software is suitable, and how it can be used for particular tasks
-  and conditions of use \cite{ISO9126}.
-\end{defn}
-
-\noindent \textbf{Proposed Definition}
-
-\wss{Still needs to be completed}
-
-\noindent \textbf{Reasoning}
-
-\wss{Needs to be completed}
-
 \subsection{\notdone{Interoperability} \ad{owner}}
-
-\begin{defn}
-  Effort required to couple one system with another
-  \citep{McCallEtAl1977}. \wss{Ao, if
-    this is from McCall, we should just cite the primary source and remove the
-    Van Vliet reference.} \pmi{Yes this is almost exactly from McCall, except he did not have the word "The" at the beginning.}
-    \ad{Someone with access to the book added it. I found the book in the pub
-    folder and fixed it.}
-\end{defn}
-
-\begin{defn}
-  Interoperability is the ability of two or more systems or components to
-  exchange information and to use the information that has been exchanged
-  \citep{IEEEComputerDictionary1991}.
-\end{defn}
-
-\begin{defn}
-  \label{InteroperabilitySelected}
-  The degree to which two or more systems, products or components can exchange
-  information and use the information that has been exchanged
-  \citep{ISO/IEC25010}.
-\end{defn}
-
-\begin{defn}
-  The ability of a system to coexist and cooperate with other systems.
-  \citep{ghezzi1991fundamentals}.
-\end{defn}
-
-\begin{defn}
-  The capability to communicate, execute programs, and transfer data among
-  various functional units in a manner that requires the user to have little or
-  no knowledge of the unique characteristics of those units
-  \citep{ISO/IEC/IEEE24765}.
-\end{defn}
-
-\begin{defn}
-  Interoperability is a characteristic of a product or system, whose interfaces
-  are completely understood, to work with other products or systems, present or
-  future, in either implementation or access, without any restrictions
-  \citep{AFUL2019}.
-\end{defn}
-
-\begin{defn}
-  Interoperability is the ability of different information systems, devices and
-  applications (‘systems’) to access, exchange, integrate and cooperatively use
-  data in a coordinated manner, within and across organizational, regional and
-  national boundaries, to provide timely and seamless portability of information
-  and optimize the health of individuals and populations globally. Health data
-  exchange architectures, application interfaces and standards enable data to be
-  accessed and shared appropriately and securely across the complete spectrum of
-  care, within all applicable settings and with relevant stakeholders, including
-  by the individual \citep{HIMSS2019}.
-
-  Four Levels of Interoperability:
-  \begin{itemize}
-  \item Foundational (Level 1) – establishes the inter-connectivity requirements
-    needed for one system or application to securely communicate data to and
-    receive data from another
-
-  \item Structural (Level 2) – defines the format, syntax, and organization of
-    data exchange including at the data field level for interpretation
-
-  \item Semantic (Level 3) – provides for common underlying models and
-    codification of the data including the use of data elements with
-    standardized definitions from publicly available value sets and coding
-    vocabularies, providing shared understanding and meaning to the user
-
-  \item Organizational (Level 4) – includes governance, policy, social, legal
-    and organizational considerations to facilitate the secure, seamless and
-    timely communication and use of data both within and between organizations,
-    entities and individuals. These components enable shared consent, trust and
-    integrated end-user processes and workflows
-  \end{itemize}
-\end{defn}
-
-\begin{defn}
-  The capability of the software to interact with one or more specified systems
-  \cite{ISO9126}.
-\end{defn}
-\noindent \textbf{Proposed Definition}
-
-Definition \ref{InteroperabilitySelected}.
-
-\noindent \textbf{Reasoning}
-
-This definition is concise and also detailed enough to show the concept not only
-on system, but also on products and components. It also covered the concept with
-more than 2 systems.
 
@@
 
-\begin{defn}
-  \label{VisibilitySelected}
-  A software development process is visible if all of its steps and its current
-  status are documented clearly. Another term used to characterize this property
-  is transparency \citep{ghezzi1991fundamentals}.
-\end{defn}
-
-\begin{defn}
-  Visibility provides transparency into the development process. It is the
-  ability to see progress at any point and determine the distance to completion
-  of a goal.  Visibility provides status of not only the progress of the
-  project, but the product itself \citep{GSA2019}.
-\end{defn}
-
-\begin{defn}
-  Business process visibility, also called process visibility, is the ability to
-  accurately and completely view the processes, transactions and other
-  activities operating within an enterprise \citep{Rouse2013}.
-\end{defn}
-
-\begin{defn}
-  Process transparency refers to the ability to look inside. The “look inside”
-  provides an in-depth and clear visibility into the business processes and how
-  these operate \citep{PRIME2019}.
-\end{defn}
-
-\begin{defn}
-The degree to which something is seen by the public
-\citep{CambridgeVisibility2019}.
-
-The degree to which something is seen or known about
-\citep{CambridgeVisibility2019}.
-\end{defn}
-
-\noindent \textbf{Proposed Definition}
-
-Definition \ref{VisibilitySelected} rephrased: The extent to which all of the
-steps of a software development process and the current status of it are 
-conveyed clearly. \wss{I wonder if rather than
-  ``documented clearly'', we should be more abstract and talk about how these
-  things can be easily determined.  Documentation is probably how the
-  information will be conveyed, but it doesn't have to be how it is done.}
-
-\noindent \textbf{Reasoning}
-
-Definition \ref{VisibilitySelected} points out that documentation is the way to
-improve visibility. It is rephrased because the original one might refer to
-binary status - ``visible'' or ``invisible''.
-
 \subsection{\notdone{Reproducibility} \wss{owner}}
-
-Reproducibility is a required component of the scientific method
-\citep{Davison2012}.  Although QA has, ``a bad name among creative scientists
-and engineers'' \citep[p.~352]{Roache1998}, the community need to recognize that
-participating in QA management also improves reproducibility.  Reproducibility,
-like QA, benefits from a consistent and repeatable computing environment,
-version control and separating code from configuration/parameters
-\citep{Davison2012}.
-
-Reproducibility is defined as:
-
-\begin{defn}
-  A result is said to be reproducible if another researcher can take the
-  original code and input data, execute it, and re-obtain the same result (Peng,
-  Dominici, and Zeger, 2006), as cited in \citet{BenureauAndRougier2017}.
-\end{defn}
-
-The related concept of replicable is defined as:
-
-\begin{defn}
-  Documentation achieves replicability if the description it provides of the
-  algorithms is sufficiently precise and complete for an independent researcher
-  to re-obtain the results it presents.  \citep{BenureauAndRougier2017}
-\end{defn}
-
-It would be worthwhile to look for some additional definitions.
-
-\noindent \textbf{Proposed Definition}
-
-\wss{Needs to be completed}
-
-\noindent \textbf{Reasoning}
-
-\wss{Needs to be completed}
 
@@
 
-\begin{defn}
-  A quality of the software production process, referring to its efficiency and
-  performance \citep{ghezzi1991fundamentals}.
-\end{defn}
-\begin{defn}
-  The best definition of the productivity of a process is
-	\[\text{Productivity} = \dfrac{\text{Outputs produced by the
-              process}}{\text{Inputs consumed by the process}}\]
-
-Thus, we can improve the productivity of the software process by increasing its
-outputs, decreasing its inputs, or both. However, this means that we need to
-provide meaningful definitions of the inputs and outputs of the software
-process. \wss{Where did this definition come from?  Is this all from Boehm?
-  This is quite a long quote.  We might actually want to paraphrase.}
-\ad{all from Boehm. I feel the best way is deleting everything after the 
-equation}
-
-\textbf{Defining inputs.} For the software process, providing a meaningful
-definition of inputs is a nontrivial but generally workable problem. Inputs to
-the software process generally comprise labor, computers, supplies, and other
-support facilities and equipment. However, one has to be careful which of
-various classes of items are to be counted as inputs. For example:
-\begin{itemize}
-\item Phases (just software development, or should we include system
-engineering, soft-ware requirements analysis, installation, or post development
-support?)
-\item Activities (to include documentation, project management, facilities
-management, conversion, training, database administration?)
-\item Personnel (to include secretaries, computer operators, business managers,
-contract administrators, line management?)
-\item Resources (to include facilities, equipment, communications, current
-versus future dollar payments?)
-\end{itemize}
-An organization can usually reach an agreement on which of the above are
-meaningful as inputs in their organizational context. Frequently, one can use
-present-value dollars as a uniform scale for various classes of resources.
-
-\textbf{Defining outputs.} The big problem in defining software productivity is
-defining outputs. Here we find a defining delivered source instructions (DSI) or
-lines of code as the output of the software process is totally inadequate, and
-they argue that there are a number of deficiencies in using DSI. However,
-	most organizations doing practical productivity measurement still use DSI as
-	their primary metric \citep{Boehm1987}. \wss{Is this a direct quote from
-		\citet{Boehm1987}?  The sentences seem incomplete?}
-\ad{I added some deleted parts, now it is a direct quote. After the ending of
-the quote, the following 2 pages discuss the flaws of DCI and a list of
-alternatives to DCI, so I ended the quote here.}
-\end{defn}
-\begin{defn}
-	\label{ProductivitySelected1}
-Productivity is the amount of output (what is produced) per unit of input used.
-In general, productivity is difficult to measure because outputs and inputs are
-typically quite diverse and are often themselves difficult to measure. In the
-context of software, productivity measurement is usually based on a simple ratio
-of product size to project effort. Thus,If we can measure the size of the
-software product and the effort required to develop the product, we have:
-	\begin{align}
-	\text{productivity} = \text{size}/\text{effort}
-	\end{align}
-Equation (1) assumes that size is the output of the software production process
-and effort is the input to the process. This can be contrasted with the
-	viewpoint of software cost models where we use size as an independent variable
-	(i.e., an input) to predict effort which is treated as an output. Equation (1)
-	is simple to operationalize if we have a single dominant size measure, for
-	example, product size measured in lines of code \citep{Kitchenham2004}.
-\end{defn}
-\begin{defn}
-	\label{ProductivitySelected2}
-The number of lines of new code developed per person-day (an imperfect measure
-of productivity but one that could be measured consistently)
-\citep{MacCormack2003}.
-\end{defn}
-
-\noindent \textbf{Proposed Definition}
-
-The revision of the combination of Definition \ref{ProductivitySelected1} and
-Definition \ref{ProductivitySelected2}: Productivity is the
-amount of output per unit of input used, which can be measured by the summation
-of all output (such as the number of lines of new code, the number of pages of
-new documents and the number of new test cases) produced per person-day.
-
-\noindent \textbf{Reasoning}
-
-It is concise and measurable. \wss{What is the output?  What is the input?  I
-	think the definition needs to give more information on these.  In particular,
-	the above definitions focus on code as the output, but documentation, test
-	cases etc should also be part of the output.  If we are going to measure this,
-	we need a better idea of what we are measuring for outputs and inputs.}
-	\ad{I added another def and made it more measurable}
-
 \subsection{\notdone{Sustainability} \wss{owner}}
-
-One of the original definitions of sustainability (for systems, not software
-specific), and still often quoted, is:
-
-\begin{defn}
-  \noindent The ability to meet the needs of the present without compromising
-  the ability of future generations to meet their own needs
-  \citep{Brundtland1987}.
-\end{defn}
-
-This is the definition used by \citet{IISD2019}.
-
-To make it more useful, this definition is often split into three dimensions:
-social, economic and environmental. \wss{cite UN paper [9] in
-  \citet{PenzenstadlerAndHenning2013}} To this list Penzenstadler and Henning
-(2013) have added technical sustainability \citep{PenzenstadlerAndHenning2013}.
-Where technical sustainability for software is defined as:
-
-\begin{defn}
-  \noindent Technical sustainability has the central objective of long-time
-  usage of systems and their adequate evolution with changing surrounding
-  conditions and respective requirements \citep{PenzenstadlerAndHenning2013}.
-\end{defn}
-
-The fourth dimension of technical sustainability is also added by
-\citep{WolframEtAl2017}.  Technical sustainability is the focus on the thesis by
-\citet{Hygerth2016}.
-
-\begin{defn}
-  \noindent Sustainable development is a mindset (principles) and an
-  accompanying set of practices that enable a team to achieve and maintain an
-  optimal development pace indefinitely \citep{Tate2005}.
-\end{defn}
-
-Parnas discusses as software aging \citep{Parnas1994a}.
-
-SCS specific definitions:
-
-\begin{defn}
-  The concept of sustainability is based on three pillars: the ecological, the
-  economical and the social. This means that for a software to be sustainable,
-  we must take all of its effects -- direct and indirect -- on the environment,
-  the economy and the society into account. In addition, the entire life cycle
-  of a software has to be considered: from planning and conception to
-  programming, distribution, installation, usage and disposal \citep{Heine2017}.
-\end{defn}
-
-Software Sustainability Institute proposal:
-
-\begin{defn}
-Capacity of the software to endure
-\end{defn}
-
-\citet{Katz2016} builds on this definition.
-
-\begin{defn}
-  \noindent The capacity of the software to endure. In other words,
-  sustainability means that the software will continue to be available in the
-  future, on new platforms, meeting new needs \citep{Katz2016}.
-\end{defn}
-
-\href{https://collegeville.github.io/CW3S19/WorkshopResources/Presentations/4-5-Katz_data-driven-software-sustainability.pdf}
-{Katz Presentation}
-
-\href{https://www.software.ac.uk/blog/2019-09-12-making-open-source-research-software-visible-path-better-sustainability}
-{Neil's blog}
-
-Definition from Neil Chue Hong:
-\begin{defn}
-	Sustainable software is software which is:
-	-- Easy to evolve and maintain
-	-- Fulfils its intent over time
-	-- Survives uncertainty
-	-- Supports relevant concerns (Political, Economic, Social, Technical,
-	Legal, Environmental) \citep{Katz2016}.
-\end{defn}
-
-\begin{defn}
-Sustainability encompasses cost efficient maintainability and evolvability \citep{SehestedtEtAl2014}.
-\end{defn}
-
-\citet{SehestedtEtAl2014} goes on to say that sustainability can be observed by
-evaluating the four criteria of the architectural model: completeness,
-consistency, correctness and clarity.
-
-Definitions for sustainability are often built by combining other definitions.
-\href{https://collegeville.github.io/CW3S19/WorkshopResources/Presentations/1-4-2019CollegevilleWorkshopWillenbring.pdf}
-{Willenbring.pdf}
-lists sustainability factors: extensible, interoperable, maintainable, portable,
-reusable, scalable and usable.  \wss{Sounds like they are listing almost all
-  software qualities.  It seems that sustainability is at least in part achieved
-  by having high quality software.}
-
-From
-\href{https://figshare.com/articles/Defining_Sustainability_through_Developers_Eyes_Recommendations_from_an_Interview_Study/1111925/1}
-{Rosado de Souza, et al.} there are two categories of software sustainability:
-
-\begin{description}
-\item[Intrinsic] Pertaining to characteristics of the software
-\item[Extrinsic] Pertaining to the software development environment
-\end{description}
-Find paper that combines nonfunctional qualities into sustainability.
-
-Sounds like definition of maintainability.
-
-Paper critical of a lack of a definition \citep{VentersEtAl2014}.
-
-Sustainability depends on the software artifacts AND the software team AND the
-development process.
-
-\noindent \textbf{Proposed Definition}
-
-\wss{Needs to be completed}
-
-\noindent \textbf{Reasoning}
-
-\wss{Needs to be completed}
-
-\section{Desirable Qualities of Good Specifications} \label{SecDesirableQs}
-
-To achieve the qualities listed in Section~\ref{SecQualities}, the documentation
-should achieve the qualities listed in this section.  All but the final quality
-listed (abstraction), are adapted from the IEEE recommended practise for
-producing good software requirements \citep{IEEE1998}.  Abstraction means only
-revealing relevant details, which in a requirements document means stating what
-is to be achieved, but remaining silent on how it is to be achieved.
-Abstraction is an important software development principle for dealing with
-complexity \citep[p.~40]{GhezziEtAl2003}.  Correctness was in the above list, so
-it is not repeated here.  \citet{SmithAndKoothoor2016} present further details
-on the qualities of documentation for SCS.
 
@@
 
-\begin{defn}
-  \label{CompletenessSelected}
-  A specification is complete to the extent that all of its parts are present
-  and each part is fully developed. A software specification must exhibit
-  several properties to assure its completeness \citep{Boehm1984}:
-  \begin{itemize}
-  \item No TBDs. TBDs are places in the specification where decisions have been
-    postponed by writing "To be Determined" or "TBD."
-  \item No nonexistent references. These are references in the specification to
-    functions, inputs, or outputs (including databases) not defined in the
-    specification.
-  \item No missing specification items. These are items that should be present
-    as part of the standard format of the specification, but are not present.
-  \item No missing functions. These are functions that should be part of the
-    software product but are not called for in the specification.
-  \item No missing products. These are products that should be part of the
-    delivered software but are not called for in the specification.
-  \end{itemize}
-\end{defn}
-
-\begin{defn}
-  Those attributes of the software that provide full implementation of the 
-  functions required. \citep{McCallEtAl1977}. \wss{Ao, if this is from McCall, 
-  we should just cite the primary source and remove the Van Vliet reference.} 
-  \pmi{McCall's definition verbatim: Those attributes of the software that 
-  provide full implementation of the functions required.} \ad{Originally added 
-  by someone else so I didn't have access to the book. But now it's available 
-  and I double checked, Peter got it accurately.}
-\end{defn}
-
-\begin{defn}
-	An SRS is complete if, and only if, it includes the following elements:
-\begin{itemize}
-\item All significant requirements, whether relating to functionality,
-performance, design constraints, attributes, or external interfaces. In
-particular any external requirements imposed by a system specification should be
-acknowledged and treated.
-\item Definition of the responses of the software to all realizable classes of
-input data in all realizable classes of situations. Note that it is important to
-specify the responses to both valid and invalid input values.
-\item Full labels and references to all figures, tables, and diagrams in the SRS
-and definition of all terms and units of measure \citep{IEEE1998}.
-	\end{itemize}
-\end{defn}
-\begin{defn}
-The quality of being whole or perfect and having nothing missing.
-\citep{CambridgeCompleteness2019}.
-\end{defn}
-
-\wss{Were there any other definitions of completeness?  You could add the
-	definition of completeness from \citet[p.\ 5--6]{IEEE1998}.  This definition
-	is for
-	requirements, but maybe there is something we can generalize from the
-	definition?  We could also look for definitions outside of software
-	development.}
-\ad{added IEEE and dictionary}
-\ad{\url{https://annals-csis.org/proceedings/2016/pliks/468.pdf} has a method of
-measuring completeness and consistency, but I haven't summarized any def from it
-yet}
-
-\noindent \textbf{Proposed Definition}
-
-The first sentence of Definition \ref{CompletenessSelected}: A specification is
-complete to the extent that all of its parts are present and each part is fully
-developed.
-
-\noindent \textbf{Reasoning}
-
-It is concise and measurable.
-
 \subsection{\notdone{Consistency} \ad{owner}}
-
-\begin{defn}
-  \label{ConsistencySelected}
-  A specification is consistent to the extent that its provisions do not
-  conflict with each other or with governing specifications and
-  objectives. Specifications require consistency in several ways
-  \citep{Boehm1984}.
-  \begin{itemize}
-  \item Internal consistency. Items within the specification do not conflict
-    with each other.
-  \item External consistency. Items in the specification do not conflict with
-    external specifications or entities.
-  \item Traceability. Items in the specification have clear antecedents in
-    earlier specifications or statements of system objectives.
-  \end{itemize}
-\end{defn}
-
-\begin{defn}
-  Consistency requires that no two or more requirements in a specification
-  contradict each other. It is also often regarded as the case where words and
-  terms have the same meaning throughout the requirements specifications
-  (consistent use of terminology). These two views of consistency imply that
-  mutually exclusive statements and clashes in terminology should be avoided
-  \citep{ZOWGHI2003}.
-\end{defn}
-\begin{defn}
-  Consistency: 1. the degree of uniformity, standardization, and freedom from
-  contradiction among the documents or parts of a system or component
-  2. software attributes that provide uniform design and implementation
-  techniques and notations \citep{ISO/IEC/IEEE24765}.
-\end{defn}
-
-\begin{defn}
-  The use of uniform design and implementation techniques and notations
-  throughout a project \citep{McCallEtAl1977}. (As summarized in
-  \citet{VanVliet2000}.)  \wss{Ao, if this is from McCall, we should just cite
-    the primary source and remove the Van Vliet reference.} \pmi{McCall's definition verbatim: Those attributes of the software that provide uniform design and implementation techniques and notation.}
-\end{defn}
-
-\begin{defn}
-Consistency refers to internal consistency. If an SRS does not agree with some
-higher-level document, such as a system requirements specification, then it is
-not correct \citep{IEEE1998}.
-\end{defn}
-\begin{defn}
-An SRS is internally consistent if, and only if, no subset of individual
-requirements described in it conflict. The three types of likely conflicts in an
-SRS are as follows \citep{IEEE1998}:
-	\begin{enumerate}[a)]
-\item The specified characteristics of real-world objects may conflict. For
-example,
-	\begin{enumerate}[1)]
-\item The format of an output report may be described in one requirement as
-tabular but in another as textual.
-\item One requirement may state that all lights shall be green while another may
-state that all lights shall be blue.
-	\end{enumerate}
-\item There may be logical or temporal conflict between two specified actions.
-For example,
-	\begin{enumerate}[1)]
-\item One requirement may specify that the program will add two inputs and
-another may specify that the program will multiply them.
-\item One requirement may state that “A” must always follow “B,” while another
-may require that “A and B” occur simultaneously.
-	\end{enumerate}
-\item Two or more requirements may describe the same real-world object but use
-different terms for that object. For example, a program’s request for a user
-input may be called a “prompt” in one requirement and a “cue” in another. The
-use of standard terminology and definitions promotes consistency.
-	\end{enumerate}
-\end{defn}
-\begin{defn}
-The state or condition of always happening or behaving in the same way
-\citep{CambridgeConsistency2019}.
-\end{defn}
-
-\wss{The definition from \citet{IEEE1998} might again be useful.  We could also
-  look for definitions outside of software development.  Even a dictionary
-  definition could be helpful.}
-\ad{added IEEE and dictionary}
-
-\noindent \textbf{Proposed Definition}
-
-The first sentence of Definition \ref{ConsistencySelected}: A specification is
-consistent to the extent that its provisions do not conflict with each other or
-with governing specifications and objectives.
-
-\noindent \textbf{Reasoning}
-
-It is concise and measurable.
 
@@
 
-Here we do seem to have a simple, if somewhat uninformative, definition:
-
-\begin{defn}
-  Modifiability is the degree of ease at which changes can be made to a system,
-  and the flexibility with which the system adapts to such changes.
-\end{defn}
-
-IEEE Standard 610 seems to speak about this. (which is superseded?)
-
-\noindent \textbf{Reasoning}
-
-\wss{Needs to be completed}
-
 \subsection{\notdone{Traceability} \jc{owner}}
-
-Here the Wikipedia page~\url{https://en.wikipedia.org/wiki/Traceability} is
-actually rather informative, especially as it also lists how this concept is
-used in other domains.  A generic definition that is still quite useful is
-\begin{defn}
-  The capability (and implementation) of keeping track of a given set or type of
-  information to a given degree, or the ability to chronologically interrelate
-  uniquely identifiable entities in a way that is verifiable.
-\end{defn}
-By specializing the above to software artifacts, ``interrelate'' to ``why is
-this here'' (for forward tracing from requirements), this does indeed give what
-is meant in SE.
-
-Various standards (DO178C, ISO 26262, and IEC61508) explicitly mention it.
-
-24765-2017 - ISO/IEC/IEEE International Standard - Systems and software
-engineering--Vocabulary
-has a full definition, namely
-\begin{enumerate}
-\item the degree to which a relationship can be established between two or more
-  products of the development process, especially products having a
-  predecessor-successor or master-subordinate relationship to one another;
-\item the identification and documentation of derivation paths (upward) and
-  allocation or flow-down paths (downward) of work products in the work product
-  hierarchy;
-\item the degree to which each element in a software development product
-  establishes its reason for existing; and discernible association among two or
-  more logical entities, such as requirements, system elements, verifications,
-  or tasks.
-\end{enumerate}
-
-\begin{defn}
-  The ability to link software components to requirements
-  \citep{McCallEtAl1977}. (As summarized in \citet{VanVliet2000}.)
-\end{defn} \wss{Once we have verified McCall, we can remove the Van Vliet
-  reference.}  \pmi{McCall's definition verbatim: Those attributes of the software that provide a thread from the requirements to the implementation with respect to the specific development and operational environment.}
-
-\noindent \textbf{Proposed Definition}
-
-\wss{Needs to be completed.}
-
-\noindent \textbf{Reasoning}
-
-\wss{Needs to be completed}
 
@@
 
-A specification is unambiguous when it has a unique interpretation.  If there is
-a possibility that two readers will have two different interpretations, than the
-specification is ambiguous.  \wss{When I get the Ghezzi text back from Olu, I'll
-  check to see if they have anything to add to this definition.}
-
-A Software Requirements Specification (SRS) is unambiguous if, and only if,
-every requirement stated therein has only one interpretation \citep{IEEE1998}.
-
-\noindent \textbf{Proposed Definition}
-
-\wss{Needs to be completed.}
-
-\noindent \textbf{Reasoning}
-
-\wss{Needs to be completed}
-
 \subsection{\notdone{Verifiability} \wss{owner}}
-
-\begin{itemize}
-
-\item Verification - Are we building the product right?  Are we implementing the
-  requirements correctly (internal)
-\item Validation - Are we building the right product? Are we getting the right
-  requirements (external)
-\item According to
-
-  \href{https://en.wikipedia.org/wiki/Software_verification_and_validation}{Capability
-    Maturity Model (CMM)}
-  \begin{itemize}
-  \item Software Verification: The process of evaluating software to determine
-    whether the products of a given development phase satisfy the conditions
-    imposed at the start of that phase. [IEEE-STD-610] \wss{Need a proper
-      citation}
-  \item Software Validation: The process of evaluating software during or at the
-    end of the development process to determine whether it satisfies specified
-    requirements. [IEEE-STD-610] \wss{Need a proper citation}
-  \end{itemize}
-\end{itemize}
-
-``An SRS is verifiable if, and only if, every requirement stated therein is
-verifiable. A requirement is verifiable if, and only if, there exists some
-finite cost-effective process with which a person or machine can check that the
-software product meets the requirement. In general any ambiguous requirement is
-not verifiable.'' \citep{IEEE1998}
-
-Verifiability is related to testability, which is defined by McCall et al.\ as
-``The effort required to test a program to ensure that it performs its intended
-function'' \citep{VanVliet2000}. \wss{Need to verify McCall reference, then
-  delete VanVliet.} \pmi{McCall's definition verbatim: Effort required to test a program to insure it performs its intended function.}
-
-\wss{When I get the Ghezzi text back from Olu, I'll check to see if they have
-  anything to add to this definition.}
-
-\begin{defn}
-  A software system is verifiable if its properties can be verified easily
-  \citep{ghezzi1991fundamentals}.
-\end{defn}
-
-\noindent \textbf{Proposed Definition}
-
-\wss{Needs to be completed}
-
-\noindent \textbf{Reasoning}
-
-\wss{Needs to be completed}
 
@@
 
-\begin{defn}
-  Documented requirements are said to be abstract if they state what the
-  software must do and the properties it must possess, but do not speak about
-  how these are to be achieved \citep{GhezziEtAl2003}.
-\end{defn}
+\section{Using Data to Rank Family Members}
 
-\begin{defn}
-  ``An abstraction for a software artifact is a succinct description that
-  suppresses the details that are unimportant to a software developer and
-  emphasizes the information that is important.'' \citep{Krueger1992}
-\end{defn}
-
-\begin{defn}
-  ``Abstraction means that we concentrate on the essential features and ignore,
-  abstract from, details that are not relevant at the level we are currently
-  working.''  \citep[p.\ 296]{VanVliet2000}
-\end{defn}
-
-\begin{defn}
-  ``Abstraction in mathematics is the process of extracting the underlying
-  essence of a mathematical concept, removing any dependence on real world
-  objects with which it might originally have been connected, and generalizing
-  it so that it has wider applications or matching among other abstract
-  descriptions of equivalent phenomena.''
-  \href{https://en.wikipedia.org/wiki/Abstraction_(mathematics)} {Wikipedia
-    Definition}
-\end{defn}
-
-Abstraction is related to reusability (and other qualities).
-
-\wss{When I get the Ghezzi text back from Olu, I'll check to see if they have
-  anything to add to this definition.}
-
-\noindent \textbf{Proposed Definition}
-
-\wss{Needs to be completed}
-
-\noindent \textbf{Reasoning}
-
-\wss{Needs to be completed}
+Describe AHP process (or similar).
 

### 2020-02-01 StateOfPractice/Peter-Notes.tex
updated some of my notes - testing and solution descriptions

@@
 \usepackage{booktabs}
+\usepackage{hyperref}
 
@@
 \begin{enumerate}
-	\item aromanro/LatticeBoltzmann
+	\item aromanro/LatticeBoltzmann\\
+	\href{https://github.com/aromanro/LatticeBoltzmann}{https://github.com/aromanro/LatticeBoltzmann}\\
+	Found from list at \href{https://github.com/topics/lattice-boltzmann}{https://github.com/topics/lattice-boltzmann}\\
+	A 2D Lattice Boltzmann program
 
-	\item Atruszkowska/LBM\_MATLAB
+	\item Atruszkowska/LBM\_MATLAB\\
+	\href{https://github.com/atruszkowska/LBM_MATLAB}{https://github.com/atruszkowska/LBM\_MATLAB}\\
+		Found from list at \href{https://github.com/topics/lattice-boltzmann}{https://github.com/topics/lattice-boltzmann}\\
+	MPI-style parallelized Shan and Chen LBM with multiscale modeling extension
 
-	\item ch4-project
+	\item ch4-project\\
+	\href{https://github.com/ecalzavarini/ch4-project}{https://github.com/ecalzavarini/ch4-project}\\  
+	Found using Google search.\\
+	Eulerian-Lagrangian fluid dynamics platform
+	A general purpose Lattice-Boltzmann code for fluid-dynamics simulations. It includes :
+	fluid dynamics (with several volume forcing terms for Channel flow, Homogeneous Isotropic Turbulence, buoyancy)
+	temperature dynamics (advection, diffusion , sink/source or reaction terms)
+	phase change (enthalpy formulation for solid/liquid systems)
+	scalar transport (same functionalities as temperature)
+	lagrangian dynamics (tracers, heavy/light and active point-like particles; non-shperical Jeffery rotation, gyrotaxis)
+	large eddy simulation (Smagorinsky, Shear Improved Samgorinsky with Kalman Filter)
 
-	\item CPE490\_Lattice\_Boltzmann\_Project
+	\item CUDA-LBM-simulator\\
+	\href{https://github.com/henryfriedlander/CUDA-LBM-simulator}{https://github.com/henryfriedlander/CUDA-LBM-simulator}\\
+		Found from list at \href{https://github.com/topics/lattice-boltzmann}{https://github.com/topics/lattice-boltzmann}\\
+	This is a Lattice-Boltzmann simulation using CUDA GPU graphics optimization.
 
-	\item CUDA-LBM-simulator
+	\item CudneLB (TCLB)\\
+	\href{https://github.com/CFD-GO/TCLB}{https://github.com/CFD-GO/TCLB}\\
+		Found from list at \href{https://github.com/topics/lattice-boltzmann}{https://github.com/topics/lattice-boltzmann}\\
+	CudneLB is a MPI+CUDA or MPI+CPU high-performance CFD simulation code, based on Lattice Boltzmann Method.
 
-	\item CudneLB (TCLB)
-
-	\item DL\_Meso
+	\item DL\_Meso \\
+	\href{https://www.scd.stfc.ac.uk/Pages/DL_MESO.aspx}{https://www.scd.stfc.ac.uk/Pages/DL\_MESO.aspx} \\
+	Found using Google search.\\
+	DL\_MESO is a general purpose mesoscale simulation package developed by Michael Seaton for CCP5 and UKCOMES​ under a grant provided by EPSRC.
+It is written in Fortran 2003 and C++ and supports both Lattice Boltzmann Equation (LBE) and Dissipative Particle Dynamics (DPD) methods.
+It is supplied with its own Java-based Graphical User Interface (GUI) and is capable of both serial and parallel execution.
 
 	\item ESPResSo
+\\
+	\href{http://espressomd.org/html/doc/index.html}{http://espressomd.org/html/doc/index.html}
+\\
+		Found from list at \href{https://github.com/topics/lattice-boltzmann}{https://github.com/topics/lattice-boltzmann}\\
+	ESPResSo is a simulation package designed to perform Molecular Dynamics (MD) and Monte Carlo (MC) simulations.
+It is meant to be a universal tool for simulations of a variety of soft matter systems.
+It features a broad range of interaction potentials which opens up possibilities for performing simulations using models with different levels of coarse-graining.
+It also includes modern and efficient algorithms for treatment of Electrostatics (P3M, MMM-type algorithms, constant potential simulations, dielectric interfaces, …), hydrodynamic interactions (DPD, Lattice-Boltzmann), and magnetic interactions, only to name a few.
+It is designed to exploit the capabilities of parallel computational environments.
+The program is being continuously extended to keep the pace with current developments both in the algorithms and software.
 
 	\item ESPResSo++
+\\
+	\href{http://www.espresso-pp.de/}{http://www.espresso-pp.de/}
+\\
+			Found from list at \href{https://github.com/topics/lattice-boltzmann}{https://github.com/topics/lattice-boltzmann}\\
+	ESPResSo++ is a software package for the scientific simulation and analysis of coarse-grained atomistic or bead-spring models as they are used in soft matter research.\\
+	ESPResSo++ has a modern C++ core and flexible Python user interface.\\
+	ESPResSo and ESPResSo++ have common roots however their development is independent and they are different software packages.\\
+	ESPResSo++ is free, open-source software published under the GNU General Public License (GPL).
 
-	\item firesim
+	\item firesim \\ 
+	\href{https://github.com/kynan/firesim}{https://github.com/kynan/firesim}
+\\
+	Found using Google search.\\
+	Lattice Boltzmann Method (LBM) fluid solver driving a particle engine for the simulation and real-time visualization of fire
 
 	\item fvLBM
+\\
+	\href{https://github.com/zhulianhua/fvLBM}{https://github.com/zhulianhua/fvLBM}
+\\
+	Found using Google search.\\
+	finite volume lattice Boltzmann method
 
-	\item HemeLB
+	\item HemeLB \\
+	\href{https://github.com/UCL/hemelb}{https://github.com/UCL/hemelb}
+\\
+	Found using Google search.\\
+	A software pipeline that simulates the blood flow through a stent (or other flow diverting device) inserted in a patient’s brain.
 
 	\item JavaCFD
+\\
+	\href{https://github.com/SihaoHuang/JavaCFD}{https://github.com/SihaoHuang/JavaCFD}
+\\
+		Found from list at \href{https://github.com/topics/lattice-boltzmann}{https://github.com/topics/lattice-boltzmann}\\
+	Computational fluid dynamics software written in Java using the Lattice-Boltzmann method. Allows custom-defined, arbitrary geometries in 2D incompressible flow field.
 
-	\item JFlowSim
+	\item JFlowSim \\
+	\href{https://github.com/ChristianFJanssen/jflowsim}{https://github.com/ChristianFJanssen/jflowsim} \\
+		Found from list at \href{https://github.com/topics/lattice-boltzmann}{https://github.com/topics/lattice-boltzmann}\\
+	jFlowSim is an interactive, thread-parallel Lattice Boltzmann solver in two dimensions.
 
 	\item laboetie
+\\
+	\href{https://github.com/maxlevesque/laboetie}{https://github.com/maxlevesque/laboetie}
+\\
+		Found from list at \href{https://github.com/topics/lattice-boltzmann}{https://github.com/topics/lattice-boltzmann}\\
+	laboetie is a computational fluid dynamics code for chemical applications.
+	It uses the Lattice-Boltzmann algorithm.
 
-	\item LatBo.jl
+	\item LatBo.jl \\
+	\href{https://github.com/UCL/LatBo.jl}{https://github.com/UCL/LatBo.jl}
+\\
+	Found using Google search.\\
+	Lattice-Boltzmann implementation in Julia
 
 	\item lettuce
+\\
+	\href{https://github.com/Olllom/lettuce}{https://github.com/Olllom/lettuce}
+\\
+	Found using Google search.\\
+	GPU-acclerated Lattice Boltzmann in Python
 
 	\item LB2D\_Prime
+\\
+	\href{http://faculty.fiu.edu/~sukopm/LBnD_Prime/LBnD_Prime.html}{http://faculty.fiu.edu/~sukopm/LBnD\_Prime/LBnD\_Prime.html} \\
+	Found using Google search.\\
+	LB2D\_Prime is a lattice Boltzmann (LB) code capable of simulating single and multi-phase flows and solute/heat transport in geometrically complex domains.
 
-	\item LB3D
+	\item LB3D \\ 
+	\href{http://ccs.chem.ucl.ac.uk/lb3d}{http://ccs.chem.ucl.ac.uk/lb3d}
+\\
+	Found using Google search.\\
+	A parallel implementation of the Lattice-Boltzmann method for simulation of interacting amphiphilic fluids.
+LB3D provides functionality to simulate three-dimensional simple, binary oil/water and ternary oil/water/amphiphile fluids using the Shan-Chen model for binary fluid interactions.
 
 	\item LBDEMcoupling-public
+\\
+	\href{https://github.com/ParticulateFlow/LBDEMcoupling-public}{https://github.com/ParticulateFlow/LBDEMcoupling-public}
+\\
+	Found using Google search.\\
+	Coupling between the Lattice-Boltzmann code Palabos and the DEM code LIGGGHTS
 
 	\item LBSim
+\\
+	\href{https://github.com/noirb/lbsim}{https://github.com/noirb/lbsim}
+\\
+	Found using Google search.\\
+	A small and simple Lattice-Boltzmann Method fluid simulator supporting complex boundaries.
 
 	\item Limbes
+\\
+	\href{https://code.google.com/archive/p/limbes/}{https://code.google.com/archive/p/limbes/} \\
+	Found using Google search.\\
+	Open source (GPL) code in 2D based on Gauss-Hermite quadrature, parallel (openmp), fortran 90.
+LIMBES is the recursive acronym for LIMBES Is May be a Boltzmann Equation Solver.
+Version 1.0 solves numerically by a Lattice Boltzmann like method the BGK-Boltzmann equation for gas in two dimensions.
 
 	\item listLBM
+\\
+	\href{https://github.com/sorush-khajepor/listLBM}{https://github.com/sorush-khajepor/listLBM} \\
+		Found from list at \href{https://github.com/topics/lattice-boltzmann}{https://github.com/topics/lattice-boltzmann}\\
+	ListLBM is a sparse lattice Boltzmann solver for multiphase flow in porous media
 
 	\item LUMA
+\\
+	\href{https://github.com/ElsevierSoftwareX/SOFTX-D-18-00007}{https://github.com/ElsevierSoftwareX/SOFTX-D-18-00007}\\
+		Found from list at \href{https://github.com/topics/lattice-boltzmann}{https://github.com/topics/lattice-boltzmann}\\
+	LUMA: A many-core, Fluid–Structure Interaction solver based on the Lattice-Boltzmann Method
 
 	\item MP-LABS
+\\
+	\href{https://github.com/carlosrosales/mplabs}{https://github.com/carlosrosales/mplabs}
+\\
+	Found using Google search.\\
+	MP-LABS is a suite of numerical simulation tools for multiphase flows based on the free energy Lattice Boltzmann Method (LBM).
+The code allows for the simulation of quasi-incompressible two-phase flows, and uses multiphase models that allow for large density ratios.
+MP-LABS provides implementations that use periodic boundary conditions, but it is written in a way that allows for easy inclusion of different boundary conditions.
+The output from MP-LABS is in plain ASCII and VTK format, and can be analyzed using other Open Source tools such as Gnuplot and Paraview.
 
-	\item MUPHY
+	The objective of the MP-LABS project is to provide a core set of routines that are well documented, highly portable, and have proven to perform well in a variety of systems.
+The source code is written in Fortran 90 and MPI and uses separate subroutines for most tasks in order to make modifications easier.
 
-	\item Openlb
+	\item Openlb \\  
+	\href{https://www.openlb.net/}{https://www.openlb.net/}
+\\
+		Found from list at \href{https://github.com/topics/lattice-boltzmann}{https://github.com/topics/lattice-boltzmann}\\
+	The OpenLB project provides a C++ package for the implementation of lattice Boltzmann methods that is general enough to address a vast range of tansport problems, e.g.
+in computational fluid dynamics.
+The source code is publicly available and constructed in a well readable, modular way.
 
-	\item Palabos
+	\item Palabos \\
+	\href{https://palabos.unige.ch/}{https://palabos.unige.ch/}
+\\
+	Found using Google search.\\
+	The Palabos library is a framework for general-purpose computational fluid dynamics (CFD), with a kernel based on the lattice Boltzmann (LB) method.
+It is used both as a research and an engineering tool: its programming interface is straightforward and makes it possible to set up fluid flow simulations with relative ease, or, if you are knowledgeable of the lattice Boltzmann method, to extend the library with your own models.
+Palabos stands for Parallel Lattice Boltzmann Solver.
+	The library’s native programming interface in written in C++.
 
-	\item PyLBM
+	\item pyLBM
+\\
+	\href{https://github.com/pylbm/pylbm}{https://github.com/pylbm/pylbm}
+\\
+		Found from list at \href{https://github.com/topics/lattice-boltzmann}{https://github.com/topics/lattice-boltzmann}\\
+	pylbm is an all-in-one package for numerical simulations using Lattice Boltzmann solvers.
+	This package gives all the tools to describe your lattice Boltzmann scheme in 1D, 2D and 3D problems.
 
-	\item Sailfish
+	\item Sailfish \\
+	\href{https://github.com/sailfish-team/sailfish}{https://github.com/sailfish-team/sailfish}
+\\
+	Found using Google search.\\
+	Lattice Boltzmann (LBM) simulation package for GPUs (CUDA, OpenCL)
 
 	\item siramirsaman/LBM
+\\
+	\href{https://github.com/siramirsaman/LBM}{https://github.com/siramirsaman/LBM}
+\\
+		Found from list at \href{https://github.com/topics/lattice-boltzmann}{https://github.com/topics/lattice-boltzmann}\\
+	Lattice Boltzmann Method Implementation in MATLAB for Curved Boundaries
 
-	\item SunlightLB
+	\item SunlightLB \\
+	\href{http://sunlightlb.sourceforge.net/}{http://sunlightlb.sourceforge.net/}
+\\
+	Found using Google search.\\
+	SunlightLB is an open-source 3D lattice Boltzmann code which can be used to solve a variety of hydrodynamics problems, including passive scalar transport problems.
 
-	\item Taxila LBM
+	\item Taxila-LBM
+\\
+	\href{https://github.com/ecoon/Taxila-LBM}{https://github.com/ecoon/Taxila-LBM}
+\\
+	Found using Google search.\\
+	Taxila LBM is a parallel implementation of the Lattice Boltzmann Method for simulation of flow in porous and geometrically complex media.
 
-	\item Tensorflow
+	\item  loliverhennigh /
+Lattice-Boltzmann-fluid-flow-in-Tensorflow 
+\\
+	\href{https://github.com/loliverhennigh/Lattice-Boltzmann-fluid-flow-in-Tensorflow}{https://github.com/loliverhennigh/Lattice-Boltzmann-fluid-flow-in-Tensorflow}
+\\
+	Found using Google search.\\
+	A Lattice Boltzmann fluid flow simulation written in Tensorflow.
 
 	\item turbulent\_lbm\_multigpu
+\\
+	\href{https://github.com/arashb/turbulent_lbm_multigpu}{https://github.com/arashb/turbulent\_lbm\_multigpu}
+\\
+	Found using Google search.\\
+	Lattice Boltzmann simulation of turbulent fluid flow on GPU Cluster
 
 	\item waLBerla
+\\
+	\href{https://www.walberla.net/}{https://www.walberla.net/}
+\\
+	Found using Google search.\\
+	waLBerla uses the lattice Boltzmann method (LBM), which is an alternative to classical Navier-Stokes solvers for computational fluid dynamics simulations.
+All of the common LBM collision models are implemented (SRT, TRT, MRT).
+Additionally, a coupling to the rigid body physics engine pe is available.
 
 	\item wlb
+\\
+	\href{https://github.com/weierstrass/wlb}{https://github.com/weierstrass/wlb}
+\\
+	Found using Google search.\\
+	A Lattice-Boltzmann code for solving coupled equations in electrohydrodynamics. 
+	Three collision operators are implemented for the (incompressible) Navier-Stokes, 
+	Nernst-Planck (advection-diffusion) and Poission's equation for electrostatics 
+	respectively. Various implementations of Dirichlet/Neumann boundary conditions 
+	are also available. The code deals (so far) only with 2D systems.
+	This code is part of a  master thesis project carried out at Chalmers University, 
+	Gothenburg.
 
 	\item Zmhhaha/LBM-Cplusplus-A.A.Mohamad
+\\
+	\href{https://github.com/zmhhaha/LBM-Cplusplus-A.A.Mohamad}{https://github.com/zmhhaha/LBM-Cplusplus-A.A.Mohamad} \\
+		Found from list at \href{https://github.com/topics/lattice-boltzmann}{https://github.com/topics/lattice-boltzmann}\\
+	The C++ version code of "Lattice Boltzmann Method Fundamentals and Engineering Applications with Computer Codes".
 
+\end{enumerate}
+
+~\newpage
+\section{Variabilities}
+
+Input Variabilities
+\begin{enumerate}
+	\item boundary parameters
+	\item dimension
+	\item number of velocity directions
+\end{enumerate}
+
+\noindent Calculation Variabilities
+\begin{enumerate}
+	\item computational model
+	\item decomposition technique
+	\item coefficient weights
+	\item input check
+	\item encoding of output
+	\item exception check
+\end{enumerate}
+
+\noindent Other Variabilities
+\begin{enumerate}
+	\item language
+	\item license
 \end{enumerate}
@@
 \subsection{Robustness}
+According to our definition, we must place the system in a state not assumed or anticipated in its requirements specification.
+Some ideas are providing invalid input, using the software on different hardware or OS software than required, or not providing (or deleting) additional libraries or directories.
+Such changes should of course be analyzed within reason - a different OS version as opposed to a fundamentally different OS, for example.\\
 
-Does not crash and provides (descriptive) error upon an invalid input.\\
+Provide invalid input and observe behaviour: Does the system crash - is it recoverable? Does the system prevent you from providing invalid input? Does the system return an error message?\\
 
-Is software that crashes but provides an error more robust than a program that does not crash but does not provide an error?\\
+Using software on different hardware/OS software than required (would need to make a reasonable list of variance): Ask similar questions as above. \\
 
-Mean time to recovery from failures?\\
+Altering/removing/not providing external libraries and/or directories: Ask similar questions as above. \\
+
+Furthermore, in regard to these questions, is software that crashes easily but provides an error more robust than a program that rarely crash but does not provide an error message?\\
+
+We can additionally measure mean time to recovery from failures\\
 
 DOI: 10.1007/978-3-642-29032-9\_16:\\ 
-
-The goal of robustness testing is to activate those faults (typically design or programming faults) or vulnerabilities in the system that result in incorrect operation, i.e., robustness failure, affecting the resilience of the system.
-Robustness testing mostly concentrates on the internal design faults activated through the system interface.
-The robustness failures are typically classified according to the CRASH criteria [540]: Catastrophic (the whole system crashes or reboots), Restart (the application has to be restarted), Abort (the application terminates abnormally), Silent (invalid operation is performed without error signal), and Hindering (incorrect error code is returned–note that returning a proper error code is considered as robust operation).\\
-
+The robustness failures are typically classified according to the CRASH criteria [540]: Catastrophic (the whole system crashes or reboots), Restart (the application has to be restarted), Abort (the application terminates abnormally), Silent (invalid operation is performed without error signal), and Hindering (incorrect error code is returned–note that returning a proper error code is considered as robust operation).
 The measure of robustness can be given as the ratio of test cases that exposes robustness faults, or, from the system’s point of view, as the number of robustness faults exposed by a given test suite.\\
 
-1) Injecting Random input 
-2) Using invalid input 
-3) Invalid inputs for each function in interface (valid and invalid inputs are type specific) 
-4) applying mutation techniques 
+\begin{enumerate}
+	\item Injecting Random input 
+	\item Using invalid input 
+	\item Invalid inputs for each function in interface (valid and invalid inputs are type specific) 
+	\item applying mutation techniques 
+\end{enumerate}
 \subsection{Performance}
-How many concurrent loads (this will also require hardware parallelism)\\
+\begin{enumerate}
+	\item How many concurrent loads (this may also require hardware parallelism) - Try to run the multiple tests in one instance of the application (pass/fail)?
+	\item How many application instances can be run? - Try to open and run multiple instances of the application (yes/no)?
+	\item CPU usage (find a tool to monitor the applications use of CPU), calculate average over the run of tests. Compare CPU usage between solutions - keep in mind the solutions vary significantly.
+	\item Memory usage - similar to CPU usage
+	\item Average Response Time (compared to other solutions) - run a pre determined set of test
+	\item Error rate (exceptions) - would need to automate tests
+	\item MTBF - would need to automate tests
+\end{enumerate}
 
-How many application instances can be run?\\
-
-CPU usage (monitor this)\\
-
-Average Response Time (compared to other solutions)\\
-
-Error rate (exceptions)\\
-
-MTBF 
 \subsection{Maintainability}
@@
 Software Metrics for Predicting Maintainability (Frappier, Matwin, Mili)
+
 \subsection{Reusability}

### 2020-02-01 StateOfPractice/QDefOfQualities/QDefOfQualities.tex

@@
 Combined definitions \ref{Installability_Selected2} and
-\ref{Installability_Selected1} rephrased.  Installability refers to the
+\ref{Installability_Selected1} rephrased:
+ Installability refers to the
 effort required for the installation, uninstallation or reinstallation of a
@@
 
- The definition captures a vital metric of measure for this quality - effort, and considers not just the installation but uninstallation and re-installation which are all related to installation activities.
-\oo{Done installability}
+ The definition captures a vital metric of measure for this quality - 
+ effort, and considers not just the installation but uninstallation
+ and re-installation which are all related to installation activities.
 
@@
 \end{defn}
-\begin{defn} \label{reliabilitySelected2} Reliability can be estimated as ``the
-  average time interval between two failures, also called the mean time to
-  failure (MTTF). Clearly, MTTF (t) = 1/FI (t)''. Where FI (t) is the failure
-  intensity - the number of failures per unit time \citep{GhezziEtAl2003}.
+\begin{defn} \label{reliabilitySelected2} Reliability is concerned with measuring
+	the probability of occurence of faliure - that is, the probability of the
+	observable effects of errors \citep{GhezziEtAl2003}.
 \end{defn}

### 2020-02-02 StateOfPractice/Ao-Notes.tex
Create Ao-Notes

\documentclass{article}

\usepackage{natbib}
\usepackage{booktabs}
\usepackage{amsmath,amsthm}
\usepackage{tabularx}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{longtable}

\hypersetup{
colorlinks=true,       % false: boxed links; true: colored links
linkcolor=red,          % color of internal links (change box color with
%linkbordercolor)
citecolor=blue,       % color of links to bibliography
filecolor=magenta,   % color of file links
urlcolor=cyan           % color of external links
}

\title{Ao's Notes on Medical Imaging Software}

\author{Ao Dong}

\date{\today}

\begin{document}

\maketitle

\begin{table}[hp]
\caption{Revision History} \label{TblRevisionHistory}
\begin{tabularx}{\textwidth}{llX}
\toprule
\textbf{Date} & \textbf{Author(s)} & \textbf{Change}\\
\midrule
Jan/30/2020 & Ao Dong & Initial draft\\
\bottomrule
\end{tabularx}
\end{table}

\section*{Software List}
PACS: Picture Archiving and Communication System\\
TK: Tool kit\\
V: Visualization\\
A: Analysis\\
C: Convertion

\begin{longtable}{|l|l|l|l|l|l|}
\hline
Functions & Sftwr Name & \begin{tabular}[c]{@{}l@{}}Latest\\ Release\end{tabular} & Link & Source \& reference & Notes \\ \hline
TK & VTK & 2019 & \href{https://vtk.org/}{link} & \citep{Samala2014} &  \\ \hline
TK & ITK & 2019 & \href{https://itk.org/}{link} & \citep{Samala2014} &  \\ \hline
TK & DCMTK & 2019 & \href{https://www.dcmtk.org/dcmtk.php.en}{link} & \begin{tabular}[c]{@{}l@{}}\citep{Hasan2020}\\ \citep{Mu2019}\end{tabular} &  \\ \hline
TK & dcm4che & 2020 & \href{https://www.dcm4che.org/}{link} & \begin{tabular}[c]{@{}l@{}}\citep{Emms2019}\\ \citep{Bjorn2017}\\ \citep{Hasan2020}\end{tabular} &  \\ \hline
PACS & orthanc & 2019 & \href{https://www.orthanc-server.com/index.php}{link} & \begin{tabular}[c]{@{}l@{}}\citep{Bjorn2017}\\ \citep{Hasan2020}\end{tabular} &  \\ \hline
\begin{tabular}[c]{@{}l@{}}PACS\\ V\end{tabular} & ClearCanvas & 2015 & \href{http://clearcanvas.github.io/}{link} & \citep{Bjorn2017} &  \\ \hline
PACS & Conquest & 2020 & \href{https://ingenium.home.xs4all.nl/dicom.html}{link} & \citep{Bjorn2017} &  \\ \hline
V & Aliza & 2020 & \href{https://www.aliza-dicom-viewer.com/}{link} & \begin{tabular}[c]{@{}l@{}}\citep{Emms2019}\\ \citep{Mu2019}\end{tabular} & Maybe not open source \\ \hline
V & 3D Slicer & 2019 & \href{https://www.slicer.org/}{link} & \begin{tabular}[c]{@{}l@{}}\citep{Emms2019}\\ \citep{Hasan2020}\\ \citep{Samala2014}\\ \citep{Haak2015}\end{tabular} &  \\ \hline
V & Mango & 2019 & \href{http://ric.uthscsa.edu/mango/}{link} & \begin{tabular}[c]{@{}l@{}}\citep{Emms2019}\\ \citep{Hasan2020}\\ \citep{Mu2019}\end{tabular} & \begin{tabular}[c]{@{}l@{}}Only open source for\\ iPad and web app version\end{tabular} \\ \hline
V & Ginkgo CADx & 2019 & \href{http://ginkgo-cadx.com/en/}{link} & \begin{tabular}[c]{@{}l@{}}\citep{Emms2019}\\ \citep{Bjorn2017}\\ \citep{Hasan2020}\\ \citep{Bruhschwein2019}\\ \citep{Mu2019}\\ \citep{Haak2015}\end{tabular} &  \\ \hline
V & XMedCon & 2019 & \href{https://xmedcon.sourceforge.io/Main/HomePage}{link} & \begin{tabular}[c]{@{}l@{}}\citep{Emms2019}\\ \citep{Hasan2020}\\ \citep{Mu2019}\\ \citep{Haak2015}\end{tabular} &  \\ \hline
V & WEASIS & 2019 & \href{https://nroduit.github.io/en/}{link} & \begin{tabular}[c]{@{}l@{}}\citep{Emms2019}\\ \citep{Hasan2020}\\ \citep{Mu2019}\\ \citep{Haak2015}\end{tabular} &  \\ \hline
V & MRIcroGL & 2019 & \begin{tabular}[c]{@{}l@{}}\href{https://www.nitrc.org/projects/mricrogl/}{link1}\\ \href{https://www.mccauslandcenter.sc.edu/mricrogl/home}{link2}\end{tabular} & \begin{tabular}[c]{@{}l@{}}\citep{Emms2019}\\ \citep{Hasan2020}\\ \citep{Samala2014}\\ \citep{Haak2015}\end{tabular} &  \\ \hline
V & Amide & 2017 & \href{http://amide.sourceforge.net/}{link} & \begin{tabular}[c]{@{}l@{}}\citep{Emms2019}\\ \citep{Hasan2020}\\ \citep{Mu2019}\end{tabular} &  \\ \hline
V & SMILI & 2019 & \href{https://smili-project.sourceforge.io/}{link} & \begin{tabular}[c]{@{}l@{}}\citep{Emms2019}\\ \citep{Hasan2020}\\ \citep{Mu2019}\end{tabular} &  \\ \hline
V & ImageJ & 2019 & \href{https://imagej.nih.gov/ij/index.html}{link} & \begin{tabular}[c]{@{}l@{}}\citep{Emms2019}\\ \citep{Hasan2020}\\ \citep{Samala2014}\\ \citep{Haak2015}\end{tabular} &  \\ \hline
V & DicomBrowser & 2016 & \href{https://wiki.xnat.org/xnat-tools/dicombrowser}{link} & \begin{tabular}[c]{@{}l@{}}\citep{Emms2019}\\ \citep{Hasan2020}\end{tabular} &  \\ \hline
V & 3DimViewer & 2019 & \href{https://bitbucket.org/3dimlab/3dimviewer/src/master/}{link} & \begin{tabular}[c]{@{}l@{}}\citep{Bjorn2017}\\ \citep{Hasan2020}\\ \citep{Mu2019}\end{tabular} &  \\ \hline
V & MicroDicom & 2020 & \href{http://www.microdicom.com/}{link} & \begin{tabular}[c]{@{}l@{}}\citep{Mu2019}\\ \citep{Bjorn2017}\end{tabular} & Maybe not open source \\ \hline
V & Horos & 2019 & \href{https://github.com/horosproject/horos}{link} & \begin{tabular}[c]{@{}l@{}}\citep{Bjorn2017}\\ \citep{Bruhschwein2019}\end{tabular} &  \\ \hline
V & Osimis & 2020 & \href{https://bitbucket.org/osimis/osimis-webviewer-plugin/src/master/}{link} & \citep{Bjorn2017} &  \\ \hline
V & OsiriX Lite & 2013 & \href{https://github.com/pixmeo/osirix}{link} & \begin{tabular}[c]{@{}l@{}}\citep{Bjorn2017}\\ \citep{Bruhschwein2019}\\ \\ \citep{Haak2015}\end{tabular} &  \\ \hline
V & jivex & 2019 & \href{https://www.visus.com/en/downloads/jivex-dicom-viewer.html}{link} & \begin{tabular}[c]{@{}l@{}}\citep{Mu2019}\\ \citep{Haak2015}\end{tabular} & Maybe not open source \\ \hline
V & Open Dicom Viewer & 2015 & \href{https://sourceforge.net/projects/opendicomviewer/}{link} & \begin{tabular}[c]{@{}l@{}}\citep{Mu2019}\\ \citep{Haak2015}\end{tabular} &  \\ \hline
V & dwv & 2019 & \href{https://ivmartel.github.io/dwv/}{link} & \citep{Bjorn2017} &  \\ \hline
V & OHIF & 2020 & \href{http://ohif.org/}{link} & \citep{Bjorn2017} &  \\ \hline
V & vv & 2019 & \href{https://github.com/open-vv/vv}{link} & \citep{Bjorn2017} &  \\ \hline
V & drishti & 2018 & \href{https://github.com/nci/drishti}{link} & \citep{Bjorn2017} &  \\ \hline
V & BioImage Suite & 2020 & \href{https://bioimagesuiteweb.github.io/webapp/}{link} & \citep{Haak2015} &  \\ \hline
V & cornerstone & 2018 & \href{https://github.com/cornerstonejs/cornerstone/releases}{link} & \citep{Haak2015} &  \\ \hline
V & dwv & 2019 & \href{https://ivmartel.github.io/dwv/}{link} & \citep{Haak2015} &  \\ \hline
V & Oviyam & 2018 & \href{http://oviyam.raster.in/index.html}{link} & \citep{Haak2015} & Maybe not open source \\ \hline
V & MicroView & 2015 & \href{http://microview.sourceforge.net/}{link} & \citep{Haak2015} &  \\ \hline
V & MIPAV & 2019 & \href{https://mipav.cit.nih.gov/}{link} & \citep{Haak2015} &  \\ \hline
V & Slice:Drop & 2019 & \href{https://github.com/slicedrop/slicedrop.github.com}{link} & \citep{Haak2015} &  \\ \hline
V & Fiji & 2018 & \href{http://fiji.sc/}{link} & \citep{Samala2014} & Distribution of ImageJ \\ \hline
V & GATE & 2019 & \href{http://www.opengatecollaboration.org/}{link} & \citep{Emms2019} &  \\ \hline
V & ITK-SNAP & 2019 & \href{http://www.itksnap.org/pmwiki/pmwiki.php}{link} & \citep{Samala2014} &  \\ \hline
V & ParaView & 2019 & \href{https://github.com/Kitware/ParaView}{link} & \citep{Samala2014} &  \\ \hline
V & MeVisLab & 2018 & \href{https://www.mevislab.de/}{link} & \citep{Samala2014} &  \\ \hline
V & MatrixUser & 2018 & \href{https://sourceforge.net/projects/matrixuser/}{link} & \citep{Hasan2020} &  \\ \hline
V & NextCloud & 2019 & \href{https://github.com/ayselafsar/dicomviewer}{link} & \citep{Hasan2020} &  \\ \hline
C & dcm2niix & 2019 & \href{https://www.nitrc.org/plugins/mwiki/index.php/dcm2nii:MainPage#Download}{link} & \citep{Hasan2020} &  \\ \hline
V & INVESALIUS 3 & 2019 & \href{https://invesalius.github.io/}{link} & \citep{Hasan2020} &  \\ \hline
V & Trimage & 2019 & \href{https://trimage.org/}{link} & \citep{Hasan2020} &  \\ \hline
V & Gwyddion & 2020 & \href{http://gwyddion.net/}{link} & \citep{Hasan2020} &  \\ \hline
 &  &  &  &  &  \\ \hline
V & dicompyler & 2016 & \href{http://www.dicompyler.com/}{link} & \citep{Mu2019} &  \\ \hline
 &  &  &  &  &  \\ \hline
V & \begin{tabular}[c]{@{}l@{}}Sante DICOM\\ Viewer Lite\end{tabular} & 2019 & \href{https://www.santesoft.com/win/sante-dicom-viewer-lite/sante-dicom-viewer-lite.html}{link} & \citep{Mu2019} & Maybe not open source \\ \hline
V & \begin{tabular}[c]{@{}l@{}}Navegatium DICOM\\ Viewer\end{tabular} & 2014? & \href{http://navegatium.com/en-us/dicom-viewer}{link} & \citep{Mu2019} &  \\ \hline
\end{longtable}

\section*{Quality Measurements}

\newpage

\bibliographystyle {plainnat}
\bibliography {../CommonFiles/ResearchProposal}

\end{document}

### 2020-02-02 StateOfPractice/Ao-Notes.tex
Update Ao-Notes

@@
 Jan/30/2020 & Ao Dong & Initial draft\\
+Feb/02/2020 & Ao Dong & Update\\
 \bottomrule
@@
 
-\section*{Software List}
+\section{Software List}
 PACS: Picture Archiving and Communication System\\
@@
 V & Osimis & 2020 & \href{https://bitbucket.org/osimis/osimis-webviewer-plugin/src/master/}{link} & \citep{Bjorn2017} &  \\ \hline
-V & OsiriX Lite & 2013 & \href{https://github.com/pixmeo/osirix}{link} & \begin{tabular}[c]{@{}l@{}}\citep{Bjorn2017}\\ \citep{Bruhschwein2019}\\ \\ \citep{Haak2015}\end{tabular} &  \\ \hline
+V & OsiriX Lite & 2013 & \href{https://github.com/pixmeo/osirix}{link} & \begin{tabular}[c]{@{}l@{}}\citep{Bjorn2017}\\ \citep{Bruhschwein2019}\\ \citep{Haak2015}\end{tabular} &  \\ \hline
 V & jivex & 2019 & \href{https://www.visus.com/en/downloads/jivex-dicom-viewer.html}{link} & \begin{tabular}[c]{@{}l@{}}\citep{Mu2019}\\ \citep{Haak2015}\end{tabular} & Maybe not open source \\ \hline
-V & Open Dicom Viewer & 2015 & \href{https://sourceforge.net/projects/opendicomviewer/}{link} & \begin{tabular}[c]{@{}l@{}}\citep{Mu2019}\\ \citep{Haak2015}\end{tabular} &  \\ \hline
-V & dwv & 2019 & \href{https://ivmartel.github.io/dwv/}{link} & \citep{Bjorn2017} &  \\ \hline
+V & \begin{tabular}[c]{@{}l@{}}Open Dicom\\ Viewer\end{tabular} & 2015 & \href{https://sourceforge.net/projects/opendicomviewer/}{link} & \begin{tabular}[c]{@{}l@{}}\citep{Mu2019}\\ \citep{Haak2015}\end{tabular} &  \\ \hline
+V & dwv & 2019 & \href{https://ivmartel.github.io/dwv/}{link} & \begin{tabular}[c]{@{}l@{}}\citep{Bjorn2017}\\ \citep{Haak2015}\end{tabular} &  \\ \hline
 V & OHIF & 2020 & \href{http://ohif.org/}{link} & \citep{Bjorn2017} &  \\ \hline
@@
 V & cornerstone & 2018 & \href{https://github.com/cornerstonejs/cornerstone/releases}{link} & \citep{Haak2015} &  \\ \hline
-V & dwv & 2019 & \href{https://ivmartel.github.io/dwv/}{link} & \citep{Haak2015} &  \\ \hline
 V & Oviyam & 2018 & \href{http://oviyam.raster.in/index.html}{link} & \citep{Haak2015} & Maybe not open source \\ \hline
@@
 V & Gwyddion & 2020 & \href{http://gwyddion.net/}{link} & \citep{Hasan2020} &  \\ \hline
- &  &  &  &  &  \\ \hline
 V & dicompyler & 2016 & \href{http://www.dicompyler.com/}{link} & \citep{Mu2019} &  \\ \hline
- &  &  &  &  &  \\ \hline
 V & \begin{tabular}[c]{@{}l@{}}Sante DICOM\\ Viewer Lite\end{tabular} & 2019 & \href{https://www.santesoft.com/win/sante-dicom-viewer-lite/sante-dicom-viewer-lite.html}{link} & \citep{Mu2019} & Maybe not open source \\ \hline
-V & \begin{tabular}[c]{@{}l@{}}Navegatium DICOM\\ Viewer\end{tabular} & 2014? & \href{http://navegatium.com/en-us/dicom-viewer}{link} & \citep{Mu2019} &  \\ \hline
+V & \begin{tabular}[c]{@{}l@{}}Navegatium\\ DICOM\\ Viewer\end{tabular} & 2014? & \href{http://navegatium.com/en-us/dicom-viewer}{link} & \citep{Mu2019} &  \\ \hline
 \end{longtable}
 
-\section*{Quality Measurements}
+\newpage
+
+\section{Quality Measurements}
+
+\subsection{Interoperability}
+There are not many measuring methods in papers, and most of them are very
+complicated.
+\begin{itemize}
+\item Can the workstation software (for visualization, analysis, etc.) connect
+with the PACS (server)?
+    \item Does the software use output from or provide input to other software?
+    \item Can the software work with customized plug-ins? 
+\end{itemize}
+
+\noindent Measuring aspects from \citep{SmithEtAl2018}:
+\begin{itemize}
+    \item Does the software interoperate with external systems? ({yes*, no})
+    \item Is there a workflow that uses other softwares? ({yes*, no})
+\item If there are external interactions, is the API clearly defined? ({yes*,
+no, n/a})
+\end{itemize}
+
+\subsection{Visibility/Transparency}
+\begin{itemize}
+\item Does the software use any version and issue tracking system, such as
+Github, Gitlab for development?
+\item Does the software have documents recording the development process and
+status?
+\item Does the software have clear release log with essential information, such
+as release date, bug fixed and new features?
+\end{itemize}
+
+\noindent Measuring aspects from \citep{SmithEtAl2018}:
+\begin{itemize}
+\item Is the development process defined? If yes, what process is used. ({yes*,
+no, n/a})
+\item Ease of external examination relative to other products considered? ({1 ..
+10})
+\end{itemize}
+
+\subsection{Productivity}
+Can be measured by the summation of all output (such as the number of lines of
+new code, the number of pages of new documents and the number of new test cases)
+produced per person-day.
+
+However, it is hard for a third party to carry out the above method without
+knowing the exact number of developers or how much time they spent on the
+project.
+
+\subsection{Completeness}
+
+\subsection{Consistency}
 

### 2020-02-03 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
updated definitions

@@
 \end{defn}
-\noindent \textbf{Proposed Definition}
-
+
+\begin{defn}
+	Verifiability refers to how well software components or the integrated
+	product can be evaluated to demonstrate whether the system functions as
+	expected \citep{wiegers2003softreq}.
+\end{defn}
+\noindent \textbf{Proposed Definition}
 Definition~\ref{Defn_Verifiability1} rephrased:

### 2020-02-03 StateOfPractice/Peter-Notes.tex

@@
 How much time it takes to add a new function (requirement)\\
+How many open and closed issues on git\\
+How quickly are issues closed\\
+Update frequency\\
 
-Metrics for Assessing a Software System's Maintainability (Oman, Hagemeister)\\
+Metrics for Assessing a Software System's Maintainability (Oman, Hagemeister):\\
+Software system metrics:\\
+1. server maturity attributes (age since release, size, stability, maintenance intensity, defect intensity, reliability, reuse, subjective product appraisals)\\
+size = thousands of non commented source statements (KNCSS)\\
+stability = 1 - change factor\\
+defect intensity = defects reported / KNCSS --check full equation\\
+reuse = the percentage of lines of code reused from other systems\\
 
-Software Metrics for Predicting Maintainability (Frappier, Matwin, Mili)
+2. source code (control structure, information structure, typography and naming and commenting) - each of these is broken into system and component subcategories\\
+
+system control structure: modularity, complexity, consistency, nesting, control coupling, encapsulation, module reuse, control flow consistency\\
+
+component control structure: complexity, use of structured constructs, use of unconditional branching, nesting, span of control structures, cohesion\\
+
+system information structure: global data types, global data structures, system coupling, data flow consistency, data type consistency, nesting, I/O complexity, I/O integrity\\
+
+component information structure: local data types, local data structures, data coupling, initialization integrity, span of data \\
+
+system typography, naming and commenting: overall program formatting, overall program commenting, module separation, naming, symbols and case\\
+
+component typography, naming and commenting: statement formatting, vertical spacing, horizontal spacing, intramodule commenting\\
+
+3. supporting documentation (abstraction, physical attributes)\\
+
+supporting documentation abstraction: descriptiveness appraisals, completeness appraisals, correctness appraisals\\
+
+supporting documentation physical attributes: readability appraisals, modifiability appraisals\\
+
+We can then sum up weighted attributes of each dimension, and then further sum up weighted dimensions\\
+
+\noindent Software Metrics for Predicting Maintainability (Frappier, Matwin, Mili): \\
+
+Documentation criterion (metrics): readability, traceability (NR, UR), coupling (M-MC, HK-IF, R-IF, KPL-IF, IF4, CA-DC), cohesion (M-MS, HK-IF, R-IF, KPL-IF, IF4, CA-DC), size\\
+
+Source code criterion (metrics): traceability(NR, UR), control structure(v(g), knots, RLC), independence, readability(Vcd, Vcs, Ls), size (LOC, E, RLC), doc accuracy (DAR), consistency (SCC)\\
+
+UR - Unreferenced Requirements - The  number  of  original  requirements  not  referenced  by  a  lower  document  in  the  documentation hierarchy (Formula: UR = number of R1 not referenced by a R2, where..)\\
+
+NR - Non-referenced requirements - The number of items not referencing an original requirement.(NR = number of R2 not referencing any R1, where..)\\
+
+Module Coupling (M-MC) - A measure of the strength of the relationships between modules.
+(Formula: Myers defines six levels of coupling which are, in order of decreasing strength: coupling, common, external, control, stamp, data)\\
+
+Module Strength (M-MS) - A measure of how strongly related are the elements within a module.(Formula: Module  strength  is  a  subjective  metric.
+Myers  defines  seven  levels  of  strength  which are, in order of increasing strength: coincidental, logical, classical, procedural, communicational, informational, functional)\\
+
+Information Flow (HK-IF) -A measure of the control flow and data flow between modules. (Formula: IFm = (fan-inm*fan-outm)2*int-compm)\\
+
+Integrated Information Flow of Rombach (R-IF) - A  measure  of  intermodule  and  intramodule  complexity  based  on  information  flow. \\
+
+ Information Flow by Kitchenham et al (KPL-IF) - A   measure   of   intermodule   complexity   inspired   from   Henry   and   Kafura's   information flow metric.
+Since Kitchenham et al experienced some difficulties in understanding   the   definition   of   flows   provided   by   Henry   and   Kafura,   they   formulated a new set of definitions.\\
+
+  IF4 -  A measure of intermodule complexity based on information flow.\\
+
+Design Complexity of Card and Agresti (CA-DC) - A  measure  of  intermodule  and  intramodule  complexity  of  a  system  based  on  fan-out, number of modules and input/output variables \\
+
+Cocomo Inspired Metric (COCO) - A selection of appropriate adjustment factors of the intermediate Cocomo metric. \\
+
+Cyclomatic Complexity Number (v(G)) -   The number of independent basic paths in a program. \\
+
+Knots Description -  The  number  of  crossing  lines  (unstructured  goto  statements)  in  a  control  flow  graph\\
+
+Relative Logical Complexity (RLC) -  The number of binary decisions divided by the number of statements \\
+
+Comments Volume of Declarations (Vcd) -  Total number of characters found in the comments of the declaration section of a module.
+The declaration section comprises comments before the module heading up to the first executable statement of the module body.
+\\
+
+ Comments Volume of Structures (Vcs) Description:    Total number of characters in the comments found anywhere in the module except in  the  declaration  section.
+The  declaration  section  comprises  comments  before  the module heading up to the first executable statement of the module body.\\
+
+ Average Length of Variable Names (Ls) Description:     Mean  number  of  characters  of  all  variables  used  in  a  module.    Unused  declared  variables are not included. \\
+
+  Lines of  Code (LOC) Description: The number of lines in the source code excluding blank lines or comment lines.\\
+
+  Software Science Effort (E) Description:    An  estimation  of  programming  effort  based  on  the  number  of  operators  and  operands.  It is a combination of other Software Science metrics.\\
+
+  Documentation Accuracy Ratio (DAR) Description:     A  verification  of  the  accuracy  of  the  CEI Spec, RS and SDD with respect to the source code.\\
+
+  Source Code Consistency (SCC) Description:     The  extent  to  which  the  source  code  contains uniform notation, terminology and symbology within itself. \\
 

### 2020-02-09 StateOfPractice/Ao-Notes.tex
Update notes

@@
 Feb/02/2020 & Ao Dong & Update\\
+Feb/08/2020 & Ao Dong & Changed table to a spreadsheet link\\
 \bottomrule
@@
 \section{Software List}
-PACS: Picture Archiving and Communication System\\
-TK: Tool kit\\
-V: Visualization\\
-A: Analysis\\
-C: Convertion
 
-\begin{longtable}{|l|l|l|l|l|l|}
-\hline
-Functions & Sftwr Name & \begin{tabular}[c]{@{}l@{}}Latest\\ Release\end{tabular} & Link & Source \& reference & Notes \\ \hline
-TK & VTK & 2019 & \href{https://vtk.org/}{link} & \citep{Samala2014} &  \\ \hline
-TK & ITK & 2019 & \href{https://itk.org/}{link} & \citep{Samala2014} &  \\ \hline
-TK & DCMTK & 2019 & \href{https://www.dcmtk.org/dcmtk.php.en}{link} & \begin{tabular}[c]{@{}l@{}}\citep{Hasan2020}\\ \citep{Mu2019}\end{tabular} &  \\ \hline
-TK & dcm4che & 2020 & \href{https://www.dcm4che.org/}{link} & \begin{tabular}[c]{@{}l@{}}\citep{Emms2019}\\ \citep{Bjorn2017}\\ \citep{Hasan2020}\end{tabular} &  \\ \hline
-PACS & orthanc & 2019 & \href{https://www.orthanc-server.com/index.php}{link} & \begin{tabular}[c]{@{}l@{}}\citep{Bjorn2017}\\ \citep{Hasan2020}\end{tabular} &  \\ \hline
-\begin{tabular}[c]{@{}l@{}}PACS\\ V\end{tabular} & ClearCanvas & 2015 & \href{http://clearcanvas.github.io/}{link} & \citep{Bjorn2017} &  \\ \hline
-PACS & Conquest & 2020 & \href{https://ingenium.home.xs4all.nl/dicom.html}{link} & \citep{Bjorn2017} &  \\ \hline
-V & Aliza & 2020 & \href{https://www.aliza-dicom-viewer.com/}{link} & \begin{tabular}[c]{@{}l@{}}\citep{Emms2019}\\ \citep{Mu2019}\end{tabular} & Maybe not open source \\ \hline
-V & 3D Slicer & 2019 & \href{https://www.slicer.org/}{link} & \begin{tabular}[c]{@{}l@{}}\citep{Emms2019}\\ \citep{Hasan2020}\\ \citep{Samala2014}\\ \citep{Haak2015}\end{tabular} &  \\ \hline
-V & Mango & 2019 & \href{http://ric.uthscsa.edu/mango/}{link} & \begin{tabular}[c]{@{}l@{}}\citep{Emms2019}\\ \citep{Hasan2020}\\ \citep{Mu2019}\end{tabular} & \begin{tabular}[c]{@{}l@{}}Only open source for\\ iPad and web app version\end{tabular} \\ \hline
-V & Ginkgo CADx & 2019 & \href{http://ginkgo-cadx.com/en/}{link} & \begin{tabular}[c]{@{}l@{}}\citep{Emms2019}\\ \citep{Bjorn2017}\\ \citep{Hasan2020}\\ \citep{Bruhschwein2019}\\ \citep{Mu2019}\\ \citep{Haak2015}\end{tabular} &  \\ \hline
-V & XMedCon & 2019 & \href{https://xmedcon.sourceforge.io/Main/HomePage}{link} & \begin{tabular}[c]{@{}l@{}}\citep{Emms2019}\\ \citep{Hasan2020}\\ \citep{Mu2019}\\ \citep{Haak2015}\end{tabular} &  \\ \hline
-V & WEASIS & 2019 & \href{https://nroduit.github.io/en/}{link} & \begin{tabular}[c]{@{}l@{}}\citep{Emms2019}\\ \citep{Hasan2020}\\ \citep{Mu2019}\\ \citep{Haak2015}\end{tabular} &  \\ \hline
-V & MRIcroGL & 2019 & \begin{tabular}[c]{@{}l@{}}\href{https://www.nitrc.org/projects/mricrogl/}{link1}\\ \href{https://www.mccauslandcenter.sc.edu/mricrogl/home}{link2}\end{tabular} & \begin{tabular}[c]{@{}l@{}}\citep{Emms2019}\\ \citep{Hasan2020}\\ \citep{Samala2014}\\ \citep{Haak2015}\end{tabular} &  \\ \hline
-V & Amide & 2017 & \href{http://amide.sourceforge.net/}{link} & \begin{tabular}[c]{@{}l@{}}\citep{Emms2019}\\ \citep{Hasan2020}\\ \citep{Mu2019}\end{tabular} &  \\ \hline
-V & SMILI & 2019 & \href{https://smili-project.sourceforge.io/}{link} & \begin{tabular}[c]{@{}l@{}}\citep{Emms2019}\\ \citep{Hasan2020}\\ \citep{Mu2019}\end{tabular} &  \\ \hline
-V & ImageJ & 2019 & \href{https://imagej.nih.gov/ij/index.html}{link} & \begin{tabular}[c]{@{}l@{}}\citep{Emms2019}\\ \citep{Hasan2020}\\ \citep{Samala2014}\\ \citep{Haak2015}\end{tabular} &  \\ \hline
-V & DicomBrowser & 2016 & \href{https://wiki.xnat.org/xnat-tools/dicombrowser}{link} & \begin{tabular}[c]{@{}l@{}}\citep{Emms2019}\\ \citep{Hasan2020}\end{tabular} &  \\ \hline
-V & 3DimViewer & 2019 & \href{https://bitbucket.org/3dimlab/3dimviewer/src/master/}{link} & \begin{tabular}[c]{@{}l@{}}\citep{Bjorn2017}\\ \citep{Hasan2020}\\ \citep{Mu2019}\end{tabular} &  \\ \hline
-V & MicroDicom & 2020 & \href{http://www.microdicom.com/}{link} & \begin{tabular}[c]{@{}l@{}}\citep{Mu2019}\\ \citep{Bjorn2017}\end{tabular} & Maybe not open source \\ \hline
-V & Horos & 2019 & \href{https://github.com/horosproject/horos}{link} & \begin{tabular}[c]{@{}l@{}}\citep{Bjorn2017}\\ \citep{Bruhschwein2019}\end{tabular} &  \\ \hline
-V & Osimis & 2020 & \href{https://bitbucket.org/osimis/osimis-webviewer-plugin/src/master/}{link} & \citep{Bjorn2017} &  \\ \hline
-V & OsiriX Lite & 2013 & \href{https://github.com/pixmeo/osirix}{link} & \begin{tabular}[c]{@{}l@{}}\citep{Bjorn2017}\\ \citep{Bruhschwein2019}\\ \citep{Haak2015}\end{tabular} &  \\ \hline
-V & jivex & 2019 & \href{https://www.visus.com/en/downloads/jivex-dicom-viewer.html}{link} & \begin{tabular}[c]{@{}l@{}}\citep{Mu2019}\\ \citep{Haak2015}\end{tabular} & Maybe not open source \\ \hline
-V & \begin{tabular}[c]{@{}l@{}}Open Dicom\\ Viewer\end{tabular} & 2015 & \href{https://sourceforge.net/projects/opendicomviewer/}{link} & \begin{tabular}[c]{@{}l@{}}\citep{Mu2019}\\ \citep{Haak2015}\end{tabular} &  \\ \hline
-V & dwv & 2019 & \href{https://ivmartel.github.io/dwv/}{link} & \begin{tabular}[c]{@{}l@{}}\citep{Bjorn2017}\\ \citep{Haak2015}\end{tabular} &  \\ \hline
-V & OHIF & 2020 & \href{http://ohif.org/}{link} & \citep{Bjorn2017} &  \\ \hline
-V & vv & 2019 & \href{https://github.com/open-vv/vv}{link} & \citep{Bjorn2017} &  \\ \hline
-V & drishti & 2018 & \href{https://github.com/nci/drishti}{link} & \citep{Bjorn2017} &  \\ \hline
-V & BioImage Suite & 2020 & \href{https://bioimagesuiteweb.github.io/webapp/}{link} & \citep{Haak2015} &  \\ \hline
-V & cornerstone & 2018 & \href{https://github.com/cornerstonejs/cornerstone/releases}{link} & \citep{Haak2015} &  \\ \hline
-V & Oviyam & 2018 & \href{http://oviyam.raster.in/index.html}{link} & \citep{Haak2015} & Maybe not open source \\ \hline
-V & MicroView & 2015 & \href{http://microview.sourceforge.net/}{link} & \citep{Haak2015} &  \\ \hline
-V & MIPAV & 2019 & \href{https://mipav.cit.nih.gov/}{link} & \citep{Haak2015} &  \\ \hline
-V & Slice:Drop & 2019 & \href{https://github.com/slicedrop/slicedrop.github.com}{link} & \citep{Haak2015} &  \\ \hline
-V & Fiji & 2018 & \href{http://fiji.sc/}{link} & \citep{Samala2014} & Distribution of ImageJ \\ \hline
-V & GATE & 2019 & \href{http://www.opengatecollaboration.org/}{link} & \citep{Emms2019} &  \\ \hline
-V & ITK-SNAP & 2019 & \href{http://www.itksnap.org/pmwiki/pmwiki.php}{link} & \citep{Samala2014} &  \\ \hline
-V & ParaView & 2019 & \href{https://github.com/Kitware/ParaView}{link} & \citep{Samala2014} &  \\ \hline
-V & MeVisLab & 2018 & \href{https://www.mevislab.de/}{link} & \citep{Samala2014} &  \\ \hline
-V & MatrixUser & 2018 & \href{https://sourceforge.net/projects/matrixuser/}{link} & \citep{Hasan2020} &  \\ \hline
-V & NextCloud & 2019 & \href{https://github.com/ayselafsar/dicomviewer}{link} & \citep{Hasan2020} &  \\ \hline
-C & dcm2niix & 2019 & \href{https://www.nitrc.org/plugins/mwiki/index.php/dcm2nii:MainPage#Download}{link} & \citep{Hasan2020} &  \\ \hline
-V & INVESALIUS 3 & 2019 & \href{https://invesalius.github.io/}{link} & \citep{Hasan2020} &  \\ \hline
-V & Trimage & 2019 & \href{https://trimage.org/}{link} & \citep{Hasan2020} &  \\ \hline
-V & Gwyddion & 2020 & \href{http://gwyddion.net/}{link} & \citep{Hasan2020} &  \\ \hline
-V & dicompyler & 2016 & \href{http://www.dicompyler.com/}{link} & \citep{Mu2019} &  \\ \hline
-V & \begin{tabular}[c]{@{}l@{}}Sante DICOM\\ Viewer Lite\end{tabular} & 2019 & \href{https://www.santesoft.com/win/sante-dicom-viewer-lite/sante-dicom-viewer-lite.html}{link} & \citep{Mu2019} & Maybe not open source \\ \hline
-V & \begin{tabular}[c]{@{}l@{}}Navegatium\\ DICOM\\ Viewer\end{tabular} & 2014? & \href{http://navegatium.com/en-us/dicom-viewer}{link} & \citep{Mu2019} &  \\ \hline
-\end{longtable}
-
-\newpage
+Link to the software list (Google Sheet): \href{https://docs.google.com/spreadsheets/d/122wU0v3ZtvDcqy8C4zKJ89kU-8fXAbo3Mzn6vcVXOi0/edit?usp=sharing}{LINK}
 

### 2020-02-10 StateOfPractice/Peter-Notes.tex
reusability, portability

@@
 
+Taxonomy of reusability mctrics
+1. Empirical methods
+-
+-
+Module oriented
+- Complexity based
+- Size based
+- Reliability based
+Component oriented
+
+ 2.Qualitative methods
+-
+-Module oriented
+- Style guidelines
+Component oriented
+- Certification guidelines
+- Quality guidelines\\
+
+-----
+
 Chidamber and Kemerer object-oriented metrics:\\
@@
 
-Issues in the Specification and Measurement of Software Portability (Mooney)\\
+Issues in the Specification and Measurement of Software Portability (Mooney):\\
+The term portability
+ refers to the ability of a software
+ unit to be ported (to a given environment).
+ A program is
+ portable if and to the degree that the cost of porting is less than
+ the cost of redevelopment. A software unit would be perfectly
+ portable if it could be ported at zero cost, but this is never
+ possible in practice.
+ Instead, a software unit may be
+ characterized by its degree ofportability,
+ which is a function of
+ the porting and development costs, with respect to a specific
+ target environment.\\
 
-Designing a Measurement Method for the Portability Non-functional Requirement (Talib)\\
+ Some of these issues have been examined by the author
+ [Mooney 931. This work has proposed as a metric the degree of
+ portability
+ of a software unit with respect to a target
+ environment, defined as
+ DPfsu) = 1 - (Cport(su,q) I Crd$\&$req.e2)).\\
+ This metric relates portability to a ratio between the cost
+ of porting (which depends on the properties of the existing
+ software unit and on the target environment), and the cost of
+ redevelopment (which depends on the requirements and the
+ target environment). A series of experiments is underway to
+ refine and validate this metric and to determine how to measure
+ or estimate Cpoyf and Crdev.
+ One study by Sheets [94]
+ suggests that the metric can be badly skewed by secondary
+ elements such as inadequate documentation for the existing
+ software.\\
+
+Designing a Measurement Method for the Portability Non-functional Requirement (Talib):\\
+ EffortNew is the total effort needed in working hours units
+ to develop the software functionalities on a new
+ environment\\
+
+-----------\\
+https://www.softwaretestinghelp.com/what-is-portability-testing/:\\
 
@@
 
-https://www.softwaretestinghelp.com/what-is-portability-testing/ 
+check Installability, Adaptability, Replaceability, Compatibility or Coexistence\\
+
+Installability: validate OS reqs, memory and RAM reqs, clear installation and uninstallation procedures, additional prerequisites\\
+
+Adapatability: hardware and software dependency, language dependency and communication system\\
+
 \end{document}

### 2020-02-10 StateOfPractice/Olu.tex

\documentclass{article}
\usepackage{natbib}
\usepackage{tabularx}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{amsmath,amsthm}

\title{Quality measure notes}

\author{Olu O}

\date{}

\begin{document}

\maketitle

~\newpage
\section{Quality Measurement}
\subsection{Installability}

It depends on a lot of factors:
\begin{itemize}
	\item How experienced is the installer?
	\item How fast the destination computer?
	\item What is the medium from which the software is being installed (Internet download, local network, CD/DVD)?
	\item Are there any manual steps needed during the installation?
\end{itemize}
\subsubitem\citep{SmithEtAl2018}
\begin{itemize}

	\item Are there installation instructions?
	\item Are the installation instructions linear?
	\item Is there something in place to automate the installation?
	\item Is there a means given to validate the installation?
	\item How many steps were involved in the installation?
	\item How many software packages need to be installed?
	\item Run uninstall, if available. Any obvious problems?

\subsubitem
 \url{www.testingstandards.co.uk/installability_guidelines.htm}
\begin{itemize}
	\item Can an untrained user successfully perform an initial installation of the application in an average of 10 minutes?
	\item When installing an upgraded version of the application, are all customizations in the user’s profile retained and converted to the new version’s data format if needed?
	\item Does installation program verify the correctness of the download before beginning the installation process???
	\item Following successful installation, does the installation program delete all temporary, backup, obsolete, and unneeded files associated with the application.
	\item What is the maximum number of installation steps
	\item Time that the installation process takes compared to the expected time it takes
	\item Install instructions available?
	\item Uninstall instructions available?
	\item Does the system handle an incomplete installation, such as one interrupted by a power failure or aborted by the user?
	\item Do applications need to be shut down before performing the installation?
	\item Does the user need the capability to install, uninstall, reinstall, or repair just selected portions of the application if required? (Can this be performed?)
	\item Is the user aware of successful, or unsuccessful, installation?
\end{itemize}

\end{itemize}
Installability addresses the following activities:
\begin{itemize}

\item Initial installation
\item Recovery from an incomplete, incorrect, or user-aborted installation
\item Reinstallation of the same version
\item Installation of a new version
\item Reverting to a previous version
\item Installation of additional components or updates
\item Uninstallation
\end{itemize}

\subsection{Verifiability}

\citep{SmithEtAl2018}
This was for Correctness and Verifiability
\begin{itemize}
	\item Are external libraries used? 
	\item Does the community have confidence in this library? 
	\item Any reference to the requirements specifications of the program?
	\item What tools or techniques are used to build confidence of correctness? (string)
	\item Is there a getting started tutorial, if yes, is the output as expected?
\end{itemize}
\citep{wiegers2003softreq}
\begin{itemize}
	\item What reference reports or other outputs can we use to verify that the system is producing its outputs correctly?
	\item Is the maximum cyclomatic complexity of a module more than 20?
	\item Are there any portions of the system that do not yield deterministic outputs, such that it could be difficult to determine if they were working correctly?
	\item Is it possible to come up with test data sets that have a high probability of revealing any errors in the requirements or in their implementation?
\end{itemize}

\bibliographystyle {plainnat}
\bibliography {../CommonFiles/ResearchProposal}
\end{document}

### 2020-02-13 StateOfPractice/Peter-Notes-QualityMeasurement.tex
peter's notes - cleaned up, continued work on maintainability, robustness; added first draft solution excel list

--- 2020-02-10 StateOfPractice/Peter-Notes.tex
@@
 \usepackage{hyperref}
+\hypersetup{%
+	colorlinks = true,
+	linkcolor  = black
+}
 
-\title{Project Notes}
+\title{Project Notes - Quality Measurement}
 
@@
 
-~\newpage
+\newpage
 
-\section{List of LBM solutions}
-\begin{enumerate}
-	\item aromanro/LatticeBoltzmann\\
-	\href{https://github.com/aromanro/LatticeBoltzmann}{https://github.com/aromanro/LatticeBoltzmann}\\
-	Found from list at \href{https://github.com/topics/lattice-boltzmann}{https://github.com/topics/lattice-boltzmann}\\
-	A 2D Lattice Boltzmann program
-
-	\item Atruszkowska/LBM\_MATLAB\\
-	\href{https://github.com/atruszkowska/LBM_MATLAB}{https://github.com/atruszkowska/LBM\_MATLAB}\\
-		Found from list at \href{https://github.com/topics/lattice-boltzmann}{https://github.com/topics/lattice-boltzmann}\\
-	MPI-style parallelized Shan and Chen LBM with multiscale modeling extension
-
-	\item ch4-project\\
-	\href{https://github.com/ecalzavarini/ch4-project}{https://github.com/ecalzavarini/ch4-project}\\  
-	Found using Google search.\\
-	Eulerian-Lagrangian fluid dynamics platform
-	A general purpose Lattice-Boltzmann code for fluid-dynamics simulations. It includes :
-	fluid dynamics (with several volume forcing terms for Channel flow, Homogeneous Isotropic Turbulence, buoyancy)
-	temperature dynamics (advection, diffusion , sink/source or reaction terms)
-	phase change (enthalpy formulation for solid/liquid systems)
-	scalar transport (same functionalities as temperature)
-	lagrangian dynamics (tracers, heavy/light and active point-like particles; non-shperical Jeffery rotation, gyrotaxis)
-	large eddy simulation (Smagorinsky, Shear Improved Samgorinsky with Kalman Filter)
-
-	\item CUDA-LBM-simulator\\
-	\href{https://github.com/henryfriedlander/CUDA-LBM-simulator}{https://github.com/henryfriedlander/CUDA-LBM-simulator}\\
-		Found from list at \href{https://github.com/topics/lattice-boltzmann}{https://github.com/topics/lattice-boltzmann}\\
-	This is a Lattice-Boltzmann simulation using CUDA GPU graphics optimization.
-
-	\item CudneLB (TCLB)\\
-	\href{https://github.com/CFD-GO/TCLB}{https://github.com/CFD-GO/TCLB}\\
-		Found from list at \href{https://github.com/topics/lattice-boltzmann}{https://github.com/topics/lattice-boltzmann}\\
-	CudneLB is a MPI+CUDA or MPI+CPU high-performance CFD simulation code, based on Lattice Boltzmann Method.
-
-	\item DL\_Meso \\
-	\href{https://www.scd.stfc.ac.uk/Pages/DL_MESO.aspx}{https://www.scd.stfc.ac.uk/Pages/DL\_MESO.aspx} \\
-	Found using Google search.\\
-	DL\_MESO is a general purpose mesoscale simulation package developed by Michael Seaton for CCP5 and UKCOMES​ under a grant provided by EPSRC.
-It is written in Fortran 2003 and C++ and supports both Lattice Boltzmann Equation (LBE) and Dissipative Particle Dynamics (DPD) methods.
-It is supplied with its own Java-based Graphical User Interface (GUI) and is capable of both serial and parallel execution.
-
-	\item ESPResSo
-\\
-	\href{http://espressomd.org/html/doc/index.html}{http://espressomd.org/html/doc/index.html}
-\\
-		Found from list at \href{https://github.com/topics/lattice-boltzmann}{https://github.com/topics/lattice-boltzmann}\\
-	ESPResSo is a simulation package designed to perform Molecular Dynamics (MD) and Monte Carlo (MC) simulations.
-It is meant to be a universal tool for simulations of a variety of soft matter systems.
-It features a broad range of interaction potentials which opens up possibilities for performing simulations using models with different levels of coarse-graining.
-It also includes modern and efficient algorithms for treatment of Electrostatics (P3M, MMM-type algorithms, constant potential simulations, dielectric interfaces, …), hydrodynamic interactions (DPD, Lattice-Boltzmann), and magnetic interactions, only to name a few.
-It is designed to exploit the capabilities of parallel computational environments.
-The program is being continuously extended to keep the pace with current developments both in the algorithms and software.
-
-	\item ESPResSo++
-\\
-	\href{http://www.espresso-pp.de/}{http://www.espresso-pp.de/}
-\\
-			Found from list at \href{https://github.com/topics/lattice-boltzmann}{https://github.com/topics/lattice-boltzmann}\\
-	ESPResSo++ is a software package for the scientific simulation and analysis of coarse-grained atomistic or bead-spring models as they are used in soft matter research.\\
-	ESPResSo++ has a modern C++ core and flexible Python user interface.\\
-	ESPResSo and ESPResSo++ have common roots however their development is independent and they are different software packages.\\
-	ESPResSo++ is free, open-source software published under the GNU General Public License (GPL).
-
-	\item firesim \\ 
-	\href{https://github.com/kynan/firesim}{https://github.com/kynan/firesim}
-\\
-	Found using Google search.\\
-	Lattice Boltzmann Method (LBM) fluid solver driving a particle engine for the simulation and real-time visualization of fire
-
-	\item fvLBM
-\\
-	\href{https://github.com/zhulianhua/fvLBM}{https://github.com/zhulianhua/fvLBM}
-\\
-	Found using Google search.\\
-	finite volume lattice Boltzmann method
-
-	\item HemeLB \\
-	\href{https://github.com/UCL/hemelb}{https://github.com/UCL/hemelb}
-\\
-	Found using Google search.\\
-	A software pipeline that simulates the blood flow through a stent (or other flow diverting device) inserted in a patient’s brain.
-
-	\item JavaCFD
-\\
-	\href{https://github.com/SihaoHuang/JavaCFD}{https://github.com/SihaoHuang/JavaCFD}
-\\
-		Found from list at \href{https://github.com/topics/lattice-boltzmann}{https://github.com/topics/lattice-boltzmann}\\
-	Computational fluid dynamics software written in Java using the Lattice-Boltzmann method. Allows custom-defined, arbitrary geometries in 2D incompressible flow field.
-
-	\item JFlowSim \\
-	\href{https://github.com/ChristianFJanssen/jflowsim}{https://github.com/ChristianFJanssen/jflowsim} \\
-		Found from list at \href{https://github.com/topics/lattice-boltzmann}{https://github.com/topics/lattice-boltzmann}\\
-	jFlowSim is an interactive, thread-parallel Lattice Boltzmann solver in two dimensions.
-
-	\item laboetie
-\\
-	\href{https://github.com/maxlevesque/laboetie}{https://github.com/maxlevesque/laboetie}
-\\
-		Found from list at \href{https://github.com/topics/lattice-boltzmann}{https://github.com/topics/lattice-boltzmann}\\
-	laboetie is a computational fluid dynamics code for chemical applications.
-	It uses the Lattice-Boltzmann algorithm.
-
-	\item LatBo.jl \\
-	\href{https://github.com/UCL/LatBo.jl}{https://github.com/UCL/LatBo.jl}
-\\
-	Found using Google search.\\
-	Lattice-Boltzmann implementation in Julia
-
-	\item lettuce
-\\
-	\href{https://github.com/Olllom/lettuce}{https://github.com/Olllom/lettuce}
-\\
-	Found using Google search.\\
-	GPU-acclerated Lattice Boltzmann in Python
-
-	\item LB2D\_Prime
-\\
-	\href{http://faculty.fiu.edu/~sukopm/LBnD_Prime/LBnD_Prime.html}{http://faculty.fiu.edu/~sukopm/LBnD\_Prime/LBnD\_Prime.html} \\
-	Found using Google search.\\
-	LB2D\_Prime is a lattice Boltzmann (LB) code capable of simulating single and multi-phase flows and solute/heat transport in geometrically complex domains.
-
-	\item LB3D \\ 
-	\href{http://ccs.chem.ucl.ac.uk/lb3d}{http://ccs.chem.ucl.ac.uk/lb3d}
-\\
-	Found using Google search.\\
-	A parallel implementation of the Lattice-Boltzmann method for simulation of interacting amphiphilic fluids.
-LB3D provides functionality to simulate three-dimensional simple, binary oil/water and ternary oil/water/amphiphile fluids using the Shan-Chen model for binary fluid interactions.
-
-	\item LBDEMcoupling-public
-\\
-	\href{https://github.com/ParticulateFlow/LBDEMcoupling-public}{https://github.com/ParticulateFlow/LBDEMcoupling-public}
-\\
-	Found using Google search.\\
-	Coupling between the Lattice-Boltzmann code Palabos and the DEM code LIGGGHTS
-
-	\item LBSim
-\\
-	\href{https://github.com/noirb/lbsim}{https://github.com/noirb/lbsim}
-\\
-	Found using Google search.\\
-	A small and simple Lattice-Boltzmann Method fluid simulator supporting complex boundaries.
-
-	\item Limbes
-\\
-	\href{https://code.google.com/archive/p/limbes/}{https://code.google.com/archive/p/limbes/} \\
-	Found using Google search.\\
-	Open source (GPL) code in 2D based on Gauss-Hermite quadrature, parallel (openmp), fortran 90.
-LIMBES is the recursive acronym for LIMBES Is May be a Boltzmann Equation Solver.
-Version 1.0 solves numerically by a Lattice Boltzmann like method the BGK-Boltzmann equation for gas in two dimensions.
-
-	\item listLBM
-\\
-	\href{https://github.com/sorush-khajepor/listLBM}{https://github.com/sorush-khajepor/listLBM} \\
-		Found from list at \href{https://github.com/topics/lattice-boltzmann}{https://github.com/topics/lattice-boltzmann}\\
-	ListLBM is a sparse lattice Boltzmann solver for multiphase flow in porous media
-
-	\item LUMA
-\\
-	\href{https://github.com/ElsevierSoftwareX/SOFTX-D-18-00007}{https://github.com/ElsevierSoftwareX/SOFTX-D-18-00007}\\
-		Found from list at \href{https://github.com/topics/lattice-boltzmann}{https://github.com/topics/lattice-boltzmann}\\
-	LUMA: A many-core, Fluid–Structure Interaction solver based on the Lattice-Boltzmann Method
-
-	\item MP-LABS
-\\
-	\href{https://github.com/carlosrosales/mplabs}{https://github.com/carlosrosales/mplabs}
-\\
-	Found using Google search.\\
-	MP-LABS is a suite of numerical simulation tools for multiphase flows based on the free energy Lattice Boltzmann Method (LBM).
-The code allows for the simulation of quasi-incompressible two-phase flows, and uses multiphase models that allow for large density ratios.
-MP-LABS provides implementations that use periodic boundary conditions, but it is written in a way that allows for easy inclusion of different boundary conditions.
-The output from MP-LABS is in plain ASCII and VTK format, and can be analyzed using other Open Source tools such as Gnuplot and Paraview.
-
-	The objective of the MP-LABS project is to provide a core set of routines that are well documented, highly portable, and have proven to perform well in a variety of systems.
-The source code is written in Fortran 90 and MPI and uses separate subroutines for most tasks in order to make modifications easier.
-
-	\item Openlb \\  
-	\href{https://www.openlb.net/}{https://www.openlb.net/}
-\\
-		Found from list at \href{https://github.com/topics/lattice-boltzmann}{https://github.com/topics/lattice-boltzmann}\\
-	The OpenLB project provides a C++ package for the implementation of lattice Boltzmann methods that is general enough to address a vast range of tansport problems, e.g.
-in computational fluid dynamics.
-The source code is publicly available and constructed in a well readable, modular way.
-
-	\item Palabos \\
-	\href{https://palabos.unige.ch/}{https://palabos.unige.ch/}
-\\
-	Found using Google search.\\
-	The Palabos library is a framework for general-purpose computational fluid dynamics (CFD), with a kernel based on the lattice Boltzmann (LB) method.
-It is used both as a research and an engineering tool: its programming interface is straightforward and makes it possible to set up fluid flow simulations with relative ease, or, if you are knowledgeable of the lattice Boltzmann method, to extend the library with your own models.
-Palabos stands for Parallel Lattice Boltzmann Solver.
-	The library’s native programming interface in written in C++.
-
-	\item pyLBM
-\\
-	\href{https://github.com/pylbm/pylbm}{https://github.com/pylbm/pylbm}
-\\
-		Found from list at \href{https://github.com/topics/lattice-boltzmann}{https://github.com/topics/lattice-boltzmann}\\
-	pylbm is an all-in-one package for numerical simulations using Lattice Boltzmann solvers.
-	This package gives all the tools to describe your lattice Boltzmann scheme in 1D, 2D and 3D problems.
-
-	\item Sailfish \\
-	\href{https://github.com/sailfish-team/sailfish}{https://github.com/sailfish-team/sailfish}
-\\
-	Found using Google search.\\
-	Lattice Boltzmann (LBM) simulation package for GPUs (CUDA, OpenCL)
-
-	\item siramirsaman/LBM
-\\
-	\href{https://github.com/siramirsaman/LBM}{https://github.com/siramirsaman/LBM}
-\\
-		Found from list at \href{https://github.com/topics/lattice-boltzmann}{https://github.com/topics/lattice-boltzmann}\\
-	Lattice Boltzmann Method Implementation in MATLAB for Curved Boundaries
-
-	\item SunlightLB \\
-	\href{http://sunlightlb.sourceforge.net/}{http://sunlightlb.sourceforge.net/}
-\\
-	Found using Google search.\\
-	SunlightLB is an open-source 3D lattice Boltzmann code which can be used to solve a variety of hydrodynamics problems, including passive scalar transport problems.
-
-	\item Taxila-LBM
-\\
-	\href{https://github.com/ecoon/Taxila-LBM}{https://github.com/ecoon/Taxila-LBM}
-\\
-	Found using Google search.\\
-	Taxila LBM is a parallel implementation of the Lattice Boltzmann Method for simulation of flow in porous and geometrically complex media.
-
-	\item  loliverhennigh /
-Lattice-Boltzmann-fluid-flow-in-Tensorflow 
-\\
-	\href{https://github.com/loliverhennigh/Lattice-Boltzmann-fluid-flow-in-Tensorflow}{https://github.com/loliverhennigh/Lattice-Boltzmann-fluid-flow-in-Tensorflow}
-\\
-	Found using Google search.\\
-	A Lattice Boltzmann fluid flow simulation written in Tensorflow.
-
-	\item turbulent\_lbm\_multigpu
-\\
-	\href{https://github.com/arashb/turbulent_lbm_multigpu}{https://github.com/arashb/turbulent\_lbm\_multigpu}
-\\
-	Found using Google search.\\
-	Lattice Boltzmann simulation of turbulent fluid flow on GPU Cluster
-
-	\item waLBerla
-\\
-	\href{https://www.walberla.net/}{https://www.walberla.net/}
-\\
-	Found using Google search.\\
-	waLBerla uses the lattice Boltzmann method (LBM), which is an alternative to classical Navier-Stokes solvers for computational fluid dynamics simulations.
-All of the common LBM collision models are implemented (SRT, TRT, MRT).
-Additionally, a coupling to the rigid body physics engine pe is available.
-
-	\item wlb
-\\
-	\href{https://github.com/weierstrass/wlb}{https://github.com/weierstrass/wlb}
-\\
-	Found using Google search.\\
-	A Lattice-Boltzmann code for solving coupled equations in electrohydrodynamics. 
-	Three collision operators are implemented for the (incompressible) Navier-Stokes, 
-	Nernst-Planck (advection-diffusion) and Poission's equation for electrostatics 
-	respectively. Various implementations of Dirichlet/Neumann boundary conditions 
-	are also available. The code deals (so far) only with 2D systems.
-	This code is part of a  master thesis project carried out at Chalmers University, 
-	Gothenburg.
-
-	\item Zmhhaha/LBM-Cplusplus-A.A.Mohamad
-\\
-	\href{https://github.com/zmhhaha/LBM-Cplusplus-A.A.Mohamad}{https://github.com/zmhhaha/LBM-Cplusplus-A.A.Mohamad} \\
-		Found from list at \href{https://github.com/topics/lattice-boltzmann}{https://github.com/topics/lattice-boltzmann}\\
-	The C++ version code of "Lattice Boltzmann Method Fundamentals and Engineering Applications with Computer Codes".
-
-\end{enumerate}
+\tableofcontents
+\addtocontents{toc}{\protect\thispagestyle{empty}}
 
@@
 \subsection{Robustness}
-According to our definition, we must place the system in a state not assumed or anticipated in its requirements specification.
-Some ideas are providing invalid input, using the software on different hardware or OS software than required, or not providing (or deleting) additional libraries or directories.
-Such changes should of course be analyzed within reason - a different OS version as opposed to a fundamentally different OS, for example.\\
+\subsubsection{Definition}
+Software possesses the characteristic of robustness if it behaves ``reasonably'' in two situations: i) when it encounters circumstances not anticipated in the requirements specification; and ii) when the assumptions in its requirements specification are violated.
 
-Provide invalid input and observe behaviour: Does the system crash - is it recoverable? Does the system prevent you from providing invalid input? Does the system return an error message?\\
-
-Using software on different hardware/OS software than required (would need to make a reasonable list of variance): Ask similar questions as above. \\
-
-Altering/removing/not providing external libraries and/or directories: Ask similar questions as above. \\
-
-Furthermore, in regard to these questions, is software that crashes easily but provides an error more robust than a program that rarely crash but does not provide an error message?\\
-
-We can additionally measure mean time to recovery from failures\\
-
+\subsubsection{Metrics}
+Using CRASH criteria: (DOI: 10.1007/978-3-642-29032-9\_16)
+\begin{enumerate}
+	\item inject random input at interface (automated testing (selenium?) - generate list of valid input and observe if CRASH)
+	\item use invalid input at interface (testing - generate list of invalid input and observe if CRASH)
+	\item not provide necessary libraries (find if external libraries are needed, not provide, observe if CRASH)
+	\item removing directories (mutation) (for example removal of output directory - find directories for mutation, remove, observe if CRASH)
+	\item using OS and hardware outside of those in the reqs: (find req OS and hardware, make list outside of reqs, attempt to run the software and observe if CRASH)
+\end{enumerate}
+\subsubsection{Notes}
 DOI: 10.1007/978-3-642-29032-9\_16:\\ 
@@
 
+~\newpage
+
+\subsection{Performance}
+\subsubsection{Definition}
+The degree to which a system or component accomplishes its designated functions within given constraints, such as speed (database response times, for instance), throughput (transactions per second), capacity (concurrent usage loads), and timing (hard real-time demands).
+
+\subsubsection{Metrics}
+
 \begin{enumerate}
-	\item Injecting Random input 
-	\item Using invalid input 
-	\item Invalid inputs for each function in interface (valid and invalid inputs are type specific) 
-	\item applying mutation techniques 
-\end{enumerate}
-\subsection{Performance}
-\begin{enumerate}
-	\item How many concurrent loads (this may also require hardware parallelism) - Try to run the multiple tests in one instance of the application (pass/fail)?
-	\item How many application instances can be run? - Try to open and run multiple instances of the application (yes/no)?
 	\item CPU usage (find a tool to monitor the applications use of CPU), calculate average over the run of tests. Compare CPU usage between solutions - keep in mind the solutions vary significantly.
-	\item Memory usage - similar to CPU usage
-	\item Average Response Time (compared to other solutions) - run a pre determined set of test
-	\item Error rate (exceptions) - would need to automate tests
-	\item MTBF - would need to automate tests
+	\item Memory usage
+	\item Concurrent usage loads
+	\item Timing
+	\item Run-time errors
+	\item errors in solution
 \end{enumerate}
 
+\subsubsection{Notes}
+
+~\newpage
+
 \subsection{Maintainability}
-How much time it takes to add a new function (requirement)\\
-How many open and closed issues on git\\
-How quickly are issues closed\\
-Update frequency\\
+\subsubsection{Definition}
+The effort with which a software system or component can be modified to:
+
+\begin{enumerate}
+	\item correct faults
+	\item improve performance or other
+	attributes
+	\item satisfy new requirements
+\end{enumerate}
+
+\subsubsection{Metrics}
+\begin{enumerate}
+	\item Open and closed issues on Git
+	\item Time to close issue avg
+	\item Update frequency
+	\item Number of maintainers
+\end{enumerate}
+
+Metrics for Assessing a Software System's Maintainability (Oman, Hagemeister):
+\begin{enumerate}
+\item Age - since release in months
+\item Size - thousands of non commented source statements (TNCSS)
+\item Stability = 1 - Change Factor -> CF is ( e$\land$number of months * average percentage change of lines of code in number of months)
+\item Defect Intensity =  e$\land$number of months * average percentage of defective lines of code per month
+\item Subjective Product Appraisals = 5 point (very low to very high) for language complexity, application complexity, requirements volatility, product dependencies, complexity of build, installation complexity, intensity of product use, efficiency of the software system
+\item modularity: number of modules, average module size
+\item consistency: std deviation of module size (TNCSS)
+\item cyclomatic complexity
+\item global data types: the number of global data types divided by the total number of defined data types
+\item global data structures: the number of global data structures divided by the total number of defined data structures
+\item Data type consistency = 1 - percentage of data structures that undergo type conversion during assignment operations
+\item I/O complexity = the number of lines of code devoted to I/O divided by TNCSS
+\item Local data types = the number of local data types divided by the total number of defined data types averaged over all modules
+\item Local data types = the number of local data types divided by the total number of defined data types averaged over all modules
+\item Local data structures = the number of local data structures divided by the total number of defined data structures averaged over all modules
+\item Initialization integrity = percentage of variables initilaized prior to use averaged over all modules
+\item Overall Formatting =  percentage of blank lines in the whole program, percentage of modules with blank lines
+\item Commenting =  percentage of comment lines in program, percentage of modules with header comments
+\item Statement formatting =  percentage of uncrowded statements (no more than one statement per line) per module averaged over all modules
+\item intramodule commenting = percentage lines of comments in module, averaged over all modules
+\item subjective evaluation of document descriptiveness = accuracy, consistency, unambiguous
+\item subjective evaluation of document completeness = extent of document set, contents
+\item subjective evaluation of document correctness = traceability, verifiability
+\item subjective evaluation of document readability = organization, accessibility via indices and table of contents, consistency of the writing style, typography, and comprehensibility
+\item subjective evaluation of document modifiability = document set redundancies
+\end{enumerate}
+\subsubsection{Notes}
 
 Metrics for Assessing a Software System's Maintainability (Oman, Hagemeister):\\
-Software system metrics:\\
+Software system metrics divided into 3 categories:\\
+
 1. server maturity attributes (age since release, size, stability, maintenance intensity, defect intensity, reliability, reuse, subjective product appraisals)\\
-size = thousands of non commented source statements (KNCSS)\\
-stability = 1 - change factor\\
-defect intensity = defects reported / KNCSS --check full equation\\
-reuse = the percentage of lines of code reused from other systems\\
 
@@
 supporting documentation physical attributes: readability appraisals, modifiability appraisals\\
-
-We can then sum up weighted attributes of each dimension, and then further sum up weighted dimensions\\
 
@@
 
+~\newpage
+
 \subsection{Reusability}
+\subsubsection{Definition}
+The extent to which a software component can be used with or without adaptation in a problem solution other than the one for which it was originally developed.
+
+\subsubsection{Metrics}
+\subsubsection{Notes}
 Measuring Software Reusability (Poulin)\\
@@
 
+3.1 Prieto-Diaz and Freeman:
+Their process model encourages
+white-box reuse and consists of finding candidate
+reusable modules, evaluating each, deciding which
+module the programmer can modify the easiest, then
+adapting the module. In this model they identify
+four module-oricntcd metrics and a fifth metric used
+to modify the fust four.
+
+Program size. Reuse depends on a small module
+size, as indicated by lines of source code.
+Program structure. Reuse depends o n a simple
+program structure as indicated by fewer links to
+other modules (low coupling) and low
+cyclomatic complexity.
+Program documentation.
+Reuse depends on
+excellent documentation as indicated by a sub-
+jective overall rating o n a scale of 1 to 10.
+Programming language. Reuse depends on pro-
+gamming language to the extent that it helps to
+reuse a module written in the same programming
+language. If a reusable module in the same lan-
+guage does not exist, the degree of similarity
+between the target language and the one used in
+the module affects the difficulty of modifying the
+module to meet the new requirement.
+Reuse experience. The experience of the reuser
+in the programming language and in the appliea-
+tion domain affects the previous metrics because
+every programmer views a module from their
+own perspective. For example, programmers
+wtll have diffcrent views of what makes a
+“small” module, depending on their background.
+?’lis fifth metric serves to modify the values of
+the other mctrics.
+
 -----
@@
 Reusability Index: A Measure for Assessing Software Assets Reusability (Ampatzoglou)
+
+~\newpage
 \subsection{Portability}
+\subsubsection{Definition}
+Effort required to transfer a program between system environments (including hardware and software).\\
+\subsubsection{Metrics}
+\subsubsection{Notes}
 compare effort to port to effort to redevelop\\

### 2020-02-13 StateOfPractice/Peter-Notes-SolutionDescriptions.tex
peter's notes - cleaned up, continued work on maintainability, robustness; added first draft solution excel list

--- 2020-02-01 StateOfPractice/Peter-Notes.tex
@@
 
-\title{Project Notes}
+\title{Project Notes - Solution Descriptions}
 
@@
 
-~\newpage
-\section{Variabilities}
-
-Input Variabilities
-\begin{enumerate}
-	\item boundary parameters
-	\item dimension
-	\item number of velocity directions
-\end{enumerate}
-
-\noindent Calculation Variabilities
-\begin{enumerate}
-	\item computational model
-	\item decomposition technique
-	\item coefficient weights
-	\item input check
-	\item encoding of output
-	\item exception check
-\end{enumerate}
-
-\noindent Other Variabilities
-\begin{enumerate}
-	\item language
-	\item license
-\end{enumerate}
-
-~\newpage
-\section{Quality Measurement}
-\subsection{Robustness}
-According to our definition, we must place the system in a state not assumed or anticipated in its requirements specification.
-Some ideas are providing invalid input, using the software on different hardware or OS software than required, or not providing (or deleting) additional libraries or directories.
-Such changes should of course be analyzed within reason - a different OS version as opposed to a fundamentally different OS, for example.\\
-
-Provide invalid input and observe behaviour: Does the system crash - is it recoverable? Does the system prevent you from providing invalid input? Does the system return an error message?\\
-
-Using software on different hardware/OS software than required (would need to make a reasonable list of variance): Ask similar questions as above. \\
-
-Altering/removing/not providing external libraries and/or directories: Ask similar questions as above. \\
-
-Furthermore, in regard to these questions, is software that crashes easily but provides an error more robust than a program that rarely crash but does not provide an error message?\\
-
-We can additionally measure mean time to recovery from failures\\
-
-DOI: 10.1007/978-3-642-29032-9\_16:\\ 
-The robustness failures are typically classified according to the CRASH criteria [540]: Catastrophic (the whole system crashes or reboots), Restart (the application has to be restarted), Abort (the application terminates abnormally), Silent (invalid operation is performed without error signal), and Hindering (incorrect error code is returned–note that returning a proper error code is considered as robust operation).
-The measure of robustness can be given as the ratio of test cases that exposes robustness faults, or, from the system’s point of view, as the number of robustness faults exposed by a given test suite.\\
-
-\begin{enumerate}
-	\item Injecting Random input 
-	\item Using invalid input 
-	\item Invalid inputs for each function in interface (valid and invalid inputs are type specific) 
-	\item applying mutation techniques 
-\end{enumerate}
-\subsection{Performance}
-\begin{enumerate}
-	\item How many concurrent loads (this may also require hardware parallelism) - Try to run the multiple tests in one instance of the application (pass/fail)?
-	\item How many application instances can be run? - Try to open and run multiple instances of the application (yes/no)?
-	\item CPU usage (find a tool to monitor the applications use of CPU), calculate average over the run of tests. Compare CPU usage between solutions - keep in mind the solutions vary significantly.
-	\item Memory usage - similar to CPU usage
-	\item Average Response Time (compared to other solutions) - run a pre determined set of test
-	\item Error rate (exceptions) - would need to automate tests
-	\item MTBF - would need to automate tests
-\end{enumerate}
-
-\subsection{Maintainability}
-How much time it takes to add a new function (requirement)\\
-
-Metrics for Assessing a Software System's Maintainability (Oman, Hagemeister)\\
-
-Software Metrics for Predicting Maintainability (Frappier, Matwin, Mili)
-
-\subsection{Reusability}
-Measuring Software Reusability (Poulin)\\
-
-Chidamber and Kemerer object-oriented metrics:\\
-
-https://www.aivosto.com/project/help/pm-oo-ck.html:\\
-
-eg weighted methods per class, number of children, coupling per class\\
-
-Software reusability metrics estimation: Algorithms, models and optimization techniques (Padhy)\\
-
-Cyclomatic complexity: independent paths through source code\\
-
-Software Reuse and Reusability Metrics and Models (Frakes, Terry)\\
-
-A new reusability metric for object-oriented software (Barnard)\\
-
-A metrics suite for measuring reusability of software components (Washizaki)\\
-
-Reusability Index: A Measure for Assessing Software Assets Reusability (Ampatzoglou) 
-\subsection{Portability}
-compare effort to port to effort to redevelop\\
-
-determine if the system can be ported: ram, processor, resolution, OS, browser\\
-
-Issues in the Specification and Measurement of Software Portability (Mooney)\\
-
-Designing a Measurement Method for the Portability Non-functional Requirement (Talib)\\
-
-ISO 9126 breaks down portability testing: installability, compatability, adapatability, and replaceability.\\
-
-https://www.softwaretestinghelp.com/what-is-portability-testing/ 
 \end{document}

### 2020-02-14 StateOfPractice/Peter-Notes-QualityMeasurement.tex
peter's notes - further cleaned up maintainability, robustness, performance

@@
 \begin{enumerate}
-	\item inject random input at interface (automated testing (selenium?) - generate list of valid input and observe if CRASH)
-	\item use invalid input at interface (testing - generate list of invalid input and observe if CRASH)
-	\item not provide necessary libraries (find if external libraries are needed, not provide, observe if CRASH)
-	\item removing directories (mutation) (for example removal of output directory - find directories for mutation, remove, observe if CRASH)
-	\item using OS and hardware outside of those in the reqs: (find req OS and hardware, make list outside of reqs, attempt to run the software and observe if CRASH)
+	\item How does the system react when random but valid input is provided?\\
+
+	Inject random input at interface (automated testing (selenium?) - generate list of valid input and observe if CRASH)
+	\item How does the system react when invalid input is provided?\\
+
+	Use invalid input at interface (testing - generate list of invalid input and observe if CRASH)
+	\item How does the system react when required libraries are not provided?\\
+
+	Not provide necessary libraries (find if external libraries are needed, not provide, observe if CRASH)
+	\item How does the system react when directories and/or files are mutated?\\
+
+	Removing directories (mutation) (for example removal of output directory - find directories for mutation, remove, observe if CRASH)
+
+	\item How does the system react when run on OS and/or hardware that is outside of the requirement specification?\\
+
+	Using OS and hardware outside of those in the reqs: (find req OS and hardware, make list outside of reqs, attempt to run the software and observe if CRASH)
 \end{enumerate}
@@
 \begin{enumerate}
-	\item CPU usage (find a tool to monitor the applications use of CPU), calculate average over the run of tests. Compare CPU usage between solutions - keep in mind the solutions vary significantly.
-	\item Memory usage
-	\item Concurrent usage loads
-	\item Timing
-	\item Run-time errors
-	\item errors in solution
+	\item How does the application compare in CPU utilization compared to other solutions running on the same hardware?\\
+
+	Measure average CPU utilization of each application using the same hardware.
+	\item What is the average memory usage of the application?\\
+
+	Measure the average memory usage of the application using the same hardware.
+	\item How many concurrent usage loads can the system handle?\\
+
+	Run multiple instances of the application. 
+	\item What is the average amount of time that the system needs to provide a solution?\\
+
+	Measure the average amount of time that each application takes to provide a solution.
+	\item How many errors does the system experience in X hours of run-time?\\
+
+	Run automated tests and count the number of run-time errors.
+	\item How accurate, on average, are the results of the solution?\\
+
+	Measure the average error of the solution when compared to a pseudo oracle.
 \end{enumerate}
@@
 \begin{enumerate}
-	\item Open and closed issues on Git
-	\item Time to close issue avg
-	\item Update frequency
+	\item Open and closed issues on Git\\
+
+	Count the ratio of open to closed issues on Git
+	\item Average time to close issues\\
+
+	Calculate the average time taken to close an issue on Git
+	\item Update frequency\\
+
+	Count average number of days between major update releases?
 	\item Number of maintainers
+\\
+
+	Count how many maintainers have worked on this project.\\
 \end{enumerate}
@@
 \begin{enumerate}
-\item Age - since release in months
-\item Size - thousands of non commented source statements (TNCSS)
-\item Stability = 1 - Change Factor -> CF is ( e$\land$number of months * average percentage change of lines of code in number of months)
-\item Defect Intensity =  e$\land$number of months * average percentage of defective lines of code per month
-\item Subjective Product Appraisals = 5 point (very low to very high) for language complexity, application complexity, requirements volatility, product dependencies, complexity of build, installation complexity, intensity of product use, efficiency of the software system
-\item modularity: number of modules, average module size
-\item consistency: std deviation of module size (TNCSS)
-\item cyclomatic complexity
-\item global data types: the number of global data types divided by the total number of defined data types
-\item global data structures: the number of global data structures divided by the total number of defined data structures
-\item Data type consistency = 1 - percentage of data structures that undergo type conversion during assignment operations
-\item I/O complexity = the number of lines of code devoted to I/O divided by TNCSS
-\item Local data types = the number of local data types divided by the total number of defined data types averaged over all modules
-\item Local data types = the number of local data types divided by the total number of defined data types averaged over all modules
-\item Local data structures = the number of local data structures divided by the total number of defined data structures averaged over all modules
-\item Initialization integrity = percentage of variables initilaized prior to use averaged over all modules
-\item Overall Formatting =  percentage of blank lines in the whole program, percentage of modules with blank lines
-\item Commenting =  percentage of comment lines in program, percentage of modules with header comments
-\item Statement formatting =  percentage of uncrowded statements (no more than one statement per line) per module averaged over all modules
-\item intramodule commenting = percentage lines of comments in module, averaged over all modules
-\item subjective evaluation of document descriptiveness = accuracy, consistency, unambiguous
-\item subjective evaluation of document completeness = extent of document set, contents
-\item subjective evaluation of document correctness = traceability, verifiability
-\item subjective evaluation of document readability = organization, accessibility via indices and table of contents, consistency of the writing style, typography, and comprehensibility
-\item subjective evaluation of document modifiability = document set redundancies
+\item Age\\
+
+Count number of months since release.
+\item Size\\
+
+Count thousands of non commented source statements (TNCSS)
+\item Stability
+
+Calculate: Stability = 1 - Change Factor when CF is ( e$\land$number of months * average percentage change of lines of code in number of months)
+\item Defect Intensity\\
+
+Calculate: Defect intensity =  e$\land$number of months * average percentage of defective lines of code per month
+\item Subjective Product Appraisals\\
+
+Measure: Subjective Product Appraisal = 5 point (very low to very high) for language complexity, application complexity, requirements volatility, product dependencies, complexity of build, installation complexity, intensity of product use, efficiency of the software system
+\item Modularity\\
+
+Measure: number of modules, average module size
+\item Consistency\\
+
+Measure: std deviation of module size (TNCSS)
+\item Global Data Types\\
+
+Measure: the number of global data types divided by the total number of defined data types
+\item Global Data Structures\\
+
+Measure: the number of global data structures divided by the total number of defined data structures
+\item Data Type Consistency\\
+
+Calculate: 1 - percentage of data structures that undergo type conversion during assignment operations
+\item I/O Complexity\\
+
+Measure: the number of lines of code devoted to I/O divided by TNCSS
+\item Local Data Types\\
+
+Calculate: the number of local data types divided by the total number of defined data types averaged over all modules
+\item Local Data Types\\
+
+Calculate: the number of local data types divided by the total number of defined data types averaged over all modules
+\item Local Data Structures\\
+
+Calculate: the number of local data structures divided by the total number of defined data structures averaged over all modules
+\item Initialization Integrity\\
+
+Calculate: percentage of variables initialized prior to use averaged over all modules
+\item Overall Formatting\\
+
+Calculate: percentage of blank lines in the whole program, percentage of modules with blank lines
+\item Commenting\\
+
+Calculate: percentage of comment lines in program, percentage of modules with header comments
+\item Statement Formatting\\
+
+Calculate: percentage of uncrowded statements (no more than one statement per line) per module averaged over all modules
+\item Intramodule Commenting\\
+
+Calculate: percentage lines of comments in module, averaged over all modules
+\item Subjective Evaluation of Document Descriptiveness\\
+
+Categorize based on accuracy, consistency, unambiguous
+\item Subjective Evaluation of Document Completeness\\
+
+Categorize based on extent of document set, contents
+\item Subjective Evaluation of Document Correctness\\
+
+Categorize based on traceability, verifiability
+\item Subjective Evaluation of Document Readability\\
+
+Categorize based on organization, accessibility via indices and table of contents, consistency of the writing style, typography, and comprehensibility
+\item Subjective Evaluation of Document Modifiability\\
+
+Categorize based on document set redundancies
 \end{enumerate}
@@
 supporting documentation physical attributes: readability appraisals, modifiability appraisals\\
-
-\noindent Software Metrics for Predicting Maintainability (Frappier, Matwin, Mili): \\
-
-Documentation criterion (metrics): readability, traceability (NR, UR), coupling (M-MC, HK-IF, R-IF, KPL-IF, IF4, CA-DC), cohesion (M-MS, HK-IF, R-IF, KPL-IF, IF4, CA-DC), size\\
-
-Source code criterion (metrics): traceability(NR, UR), control structure(v(g), knots, RLC), independence, readability(Vcd, Vcs, Ls), size (LOC, E, RLC), doc accuracy (DAR), consistency (SCC)\\
-
-UR - Unreferenced Requirements - The  number  of  original  requirements  not  referenced  by  a  lower  document  in  the  documentation hierarchy (Formula: UR = number of R1 not referenced by a R2, where..)\\
-
-NR - Non-referenced requirements - The number of items not referencing an original requirement.(NR = number of R2 not referencing any R1, where..)\\
-
-Module Coupling (M-MC) - A measure of the strength of the relationships between modules.
-(Formula: Myers defines six levels of coupling which are, in order of decreasing strength: coupling, common, external, control, stamp, data)\\
-
-Module Strength (M-MS) - A measure of how strongly related are the elements within a module.(Formula: Module  strength  is  a  subjective  metric.
-Myers  defines  seven  levels  of  strength  which are, in order of increasing strength: coincidental, logical, classical, procedural, communicational, informational, functional)\\
-
-Information Flow (HK-IF) -A measure of the control flow and data flow between modules. (Formula: IFm = (fan-inm*fan-outm)2*int-compm)\\
-
-Integrated Information Flow of Rombach (R-IF) - A  measure  of  intermodule  and  intramodule  complexity  based  on  information  flow. \\
-
- Information Flow by Kitchenham et al (KPL-IF) - A   measure   of   intermodule   complexity   inspired   from   Henry   and   Kafura's   information flow metric.
-Since Kitchenham et al experienced some difficulties in understanding   the   definition   of   flows   provided   by   Henry   and   Kafura,   they   formulated a new set of definitions.\\
-
-  IF4 -  A measure of intermodule complexity based on information flow.\\
-
-Design Complexity of Card and Agresti (CA-DC) - A  measure  of  intermodule  and  intramodule  complexity  of  a  system  based  on  fan-out, number of modules and input/output variables \\
-
-Cocomo Inspired Metric (COCO) - A selection of appropriate adjustment factors of the intermediate Cocomo metric. \\
-
-Cyclomatic Complexity Number (v(G)) -   The number of independent basic paths in a program. \\
-
-Knots Description -  The  number  of  crossing  lines  (unstructured  goto  statements)  in  a  control  flow  graph\\
-
-Relative Logical Complexity (RLC) -  The number of binary decisions divided by the number of statements \\
-
-Comments Volume of Declarations (Vcd) -  Total number of characters found in the comments of the declaration section of a module.
-The declaration section comprises comments before the module heading up to the first executable statement of the module body.
-\\
-
- Comments Volume of Structures (Vcs) Description:    Total number of characters in the comments found anywhere in the module except in  the  declaration  section.
-The  declaration  section  comprises  comments  before  the module heading up to the first executable statement of the module body.\\
-
- Average Length of Variable Names (Ls) Description:     Mean  number  of  characters  of  all  variables  used  in  a  module.    Unused  declared  variables are not included. \\
-
-  Lines of  Code (LOC) Description: The number of lines in the source code excluding blank lines or comment lines.\\
-
-  Software Science Effort (E) Description:    An  estimation  of  programming  effort  based  on  the  number  of  operators  and  operands.  It is a combination of other Software Science metrics.\\
-
-  Documentation Accuracy Ratio (DAR) Description:     A  verification  of  the  accuracy  of  the  CEI Spec, RS and SDD with respect to the source code.\\
-
-  Source Code Consistency (SCC) Description:     The  extent  to  which  the  source  code  contains uniform notation, terminology and symbology within itself. \\
 

### 2020-02-15 StateOfPractice/Peter-Notes-QualityMeasurement.tex
peter's notes - cleaned up, added metrics to reusability, portability

@@
 \subsubsection{Metrics}
+\noindent Poulin (non-direct source):
+\begin{enumerate}
+	\item Size\\
+
+	Count lines of source code.
+	\item Structure\\
+
+	Count links to other modules.
+	\item Documentation\\
+
+	Subjective rating on a scale of 1 to 10.
+	\item Language\\
+
+	Subjective rating based on how common the language is.
+	\item Simple interfaces\\
+
+	Count the average number of passed variables to modules.
+	\item Few calls to other modules\\
+
+	Count the average number of calls to other modules.
+	\item Cyclomatic complexity\\
+
+	Measure cyclomatic complexity
+	\item Self descriptiveness\\
+
+	Measure average percentage of comment lines in module
+	\item Number of problem reports\\
+
+	Measure average number of outstanding defects in the module.
+\end{enumerate}
+
+\noindent Chidamber and Kemerer:
+\begin{enumerate}
+	\item Methods per class\\
+
+	Measure the average number of methods defined in each class (want to keep low).
+	\item Number of children\\
+
+	Measure the number of immediate child classes derived from a base class (wan this to be high). 
+	\item Coupling between object classes\\
+
+	Measure the average number of methods or variables per class that are used by other classes (want to keep this low).
+\end{enumerate}
+
+\noindent Software reusability metrics estimation: Algorithms, models and optimization techniques (Padhy):\\ 
+\begin{enumerate}
+	\item Methods per class (WMC)\\
+
+	Measure the number of methods per class.
+	\item Depth of inheritance tree (DIT)\\
+
+	Measure the highest length from the rood to the node.
+	\item Number of children (NOC)\\
+
+	Measure the average total number of total number of sub-classes.
+
+	\item Method complexity (MC)\\
+
+	Measure the average total number of methods present in a class.
+\end{enumerate}
+
+\noindent A new reusability metric for object-oriented software (Barnard):\\ 
+\begin{enumerate}
+	\item Coupling (calls to foreign classes)
+	\item Number of methods
+	\item Number of attributes
+	\item Meaningful name
+	\item Documentation
+	\item Lines of code
+	\item Comment lines
+	\item Depth of inheritance
+	\item Number of children
+	\item Cohesion 
+	\item Class variables
+	\item Complexity of data structure
+	\item Interface parameters
+	\item Calls to foreign classes
+	\item Calls to library classes
+	\item Access (public, protected, private)
+\end{enumerate}
+
+\noindent Reusability Index: A Measure for Assessing Software Assets Reusability (Ampatzoglou):\\ 
+\begin{enumerate}
+	\item Number of classes in a module
+	\item Number of properties
+	\item Interface complexity
+	\item Documentation quality
+	\item Number of open bugs
+\end{enumerate}
+
 \subsubsection{Notes}
-Measuring Software Reusability (Poulin)\\
+\noindent Measuring Software Reusability (Poulin)\\
 
-Taxonomy of reusability mctrics
-1. Empirical methods
--
--
-Module oriented
-- Complexity based
-- Size based
-- Reliability based
-Component oriented
+Taxonomy of reusability metrics:\\
+1. Empirical methods:\\
+Module oriented (complexity based, size based, reliability based), 
+Component oriented\\
 
- 2.Qualitative methods
--
--Module oriented
-- Style guidelines
-Component oriented
-- Certification guidelines
-- Quality guidelines\\
+2.Qualitative methods:\\
+Module oriented (style guidelines), Component oriented (certification guidelines, quality guidelines)\\
 
-3.1 Prieto-Diaz and Freeman:
-Their process model encourages
-white-box reuse and consists of finding candidate
-reusable modules, evaluating each, deciding which
-module the programmer can modify the easiest, then
-adapting the module. In this model they identify
-four module-oricntcd metrics and a fifth metric used
-to modify the fust four.
+\noindent Chidamber and Kemerer object-oriented metrics (https://www.aivosto.com/project/help/pm-oo-ck.html):\\
+The Chidamber and Kemerer metrics suite originally consists of 6 metrics calculated for each class: WMC, DIT, NOC, CBO, RFC and LCOM1.
+The original suite has later been amended by RFC´, LCOM2, LCOM3 and LCOM4 by other authors.\\
 
-Program size. Reuse depends on a small module
-size, as indicated by lines of source code.
-Program structure. Reuse depends o n a simple
-program structure as indicated by fewer links to
-other modules (low coupling) and low
-cyclomatic complexity.
-Program documentation.
-Reuse depends on
-excellent documentation as indicated by a sub-
-jective overall rating o n a scale of 1 to 10.
-Programming language. Reuse depends on pro-
-gamming language to the extent that it helps to
-reuse a module written in the same programming
-language. If a reusable module in the same lan-
-guage does not exist, the degree of similarity
-between the target language and the one used in
-the module affects the difficulty of modifying the
-module to meet the new requirement.
-Reuse experience. The experience of the reuser
-in the programming language and in the appliea-
-tion domain affects the previous metrics because
-every programmer views a module from their
-own perspective. For example, programmers
-wtll have diffcrent views of what makes a
-“small” module, depending on their background.
-?’lis fifth metric serves to modify the values of
-the other mctrics.
+\noindent Software reusability metrics estimation: Algorithms, models and optimization techniques (Padhy)\\
 
------
+\noindent A new reusability metric for object-oriented software (Barnard)\\
 
-Chidamber and Kemerer object-oriented metrics:\\
-
-https://www.aivosto.com/project/help/pm-oo-ck.html:\\
-
-eg weighted methods per class, number of children, coupling per class\\
-
-Software reusability metrics estimation: Algorithms, models and optimization techniques (Padhy)\\
-
-Cyclomatic complexity: independent paths through source code\\
-
-Software Reuse and Reusability Metrics and Models (Frakes, Terry)\\
-
-A new reusability metric for object-oriented software (Barnard)\\
-
-A metrics suite for measuring reusability of software components (Washizaki)\\
-
-Reusability Index: A Measure for Assessing Software Assets Reusability (Ampatzoglou)
+\noindent Reusability Index: A Measure for Assessing Software Assets Reusability (Ampatzoglou)
 
@@
 \subsubsection{Metrics}
+Same as installability?
+\begin{enumerate}
+	\item Time to port
+	\item OS platforms
+	\item OS versions
+	\item Hardware platforms
+	\item Number of external libraries needed
+\end{enumerate}
+
 \subsubsection{Notes}
@@
 
-Issues in the Specification and Measurement of Software Portability (Mooney):\\
+\noindent Issues in the Specification and Measurement of Software Portability (Mooney):\\
 The term portability
@@
 
-Designing a Measurement Method for the Portability Non-functional Requirement (Talib):\\
- EffortNew is the total effort needed in working hours units
- to develop the software functionalities on a new
- environment\\
+  The principal types of portability usually considered are binary portability (porting the executable form) and source portability (porting the source language representation).
+Binary portability offers several advantages, but is possible only across strongly similar environments.Source portability assumes availability of source code, but provides opportunities to adapt a software unit to a wide range of environments.
+In this paper we consider only source portability.The porting process has two principal components which we call transportation and adaptation.
+Transportation is physical movement;  this may not be trivial since compatible media must be used and various types of representation conversion may be required.
+Adaptation is any modification that must be performed on the original version;  we take this to mean both mechanical translation such as by language processors, and manual modification by humans\\
 
------------\\
-https://www.softwaretestinghelp.com/what-is-portability-testing/:\\
+\noindent Designing a Measurement Method for the Portability Non-functional Requirement (Talib):\\
+
+\noindent https://www.softwaretestinghelp.com/what-is-portability-testing/:\\
 

### 2020-02-17 StateOfPractice/Peter-Notes-QualityMeasurement.tex
peter's notes - added portability measures; updated first 6 solutions in excel file

@@
 
-Input Variabilities
 \begin{enumerate}
-	\item boundary parameters
-	\item dimension
-	\item number of velocity directions
-\end{enumerate}
-
-\noindent Calculation Variabilities
-\begin{enumerate}
+	\item category
+	\item dimensions
 	\item computational model
 	\item decomposition technique
-	\item coefficient weights
 	\item input check
@@
 	\item exception check
-\end{enumerate}
-
-\noindent Other Variabilities
-\begin{enumerate}
 	\item language
 	\item license
+	\item parallel
+	\item open source
+	\item output model
+	\item turbulent
+	\item complex geometry
+	\item non-Newtonian fluids
+	\item dependencies
 \end{enumerate}
@@
 \begin{enumerate}
-	\item Time to port
-	\item OS platforms
-	\item OS versions
-	\item Hardware platforms
-	\item Number of external libraries needed
+	\item OS platforms\\
+
+	Measure the average time (or steps taken) to port the program between a list of OS environments.
+	\item OS versions\\
+
+	Measure the average time (or steps taken) to port the program between a list of OS versions.
+	\item Hardware platforms\\
+
+	Measure the average time (or steps taken) to port the program between a list of hardware platforms.
+
+	\item Number of external libraries needed\\
+
+	Count the number of external libraries needed to set up the environment.
+
 \end{enumerate}

### 2020-02-17 StateOfPractice/Peter-Notes-SolutionDescriptions.tex
peter's notes - added portability measures; updated first 6 solutions in excel file

@@
 	\item ch4-project\\
-	\href{https://github.com/ecalzavarini/ch4-project}{https://github.com/ecalzavarini/ch4-project}\\  
+	\href{https://github.com/SoftwareImpacts/SIMPAC-2019-1}{https://github.com/SoftwareImpacts/SIMPAC-2019-1}\\  
 	Found using Google search.\\

### 2020-02-24 StateOfPractice/Peter-Notes/Peter-Notes-SolutionDescriptions.tex
cleaned up State of Practice directory - added a subdir for my notes

--- 2020-02-17 StateOfPractice/Peter-Notes-SolutionDescriptions.tex
@@
 	A software pipeline that simulates the blood flow through a stent (or other flow diverting device) inserted in a patient’s brain.
-
-	\item JavaCFD
-\\
-	\href{https://github.com/SihaoHuang/JavaCFD}{https://github.com/SihaoHuang/JavaCFD}
-\\
-		Found from list at \href{https://github.com/topics/lattice-boltzmann}{https://github.com/topics/lattice-boltzmann}\\
-	Computational fluid dynamics software written in Java using the Lattice-Boltzmann method. Allows custom-defined, arbitrary geometries in 2D incompressible flow field.
 

### 2020-02-24 StateOfPractice/Peter-Notes/Commonality-Analysis-LB-Systems.tex
updated commonality analysis template

\documentclass[12pt]{article}

\usepackage{titling}
\usepackage{blindtext}
\usepackage{fullpage}
\usepackage[round]{natbib}
\usepackage{multirow}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{siunitx}
\usepackage{graphicx}
\usepackage{float}
\usepackage{hyperref}
\hypersetup{
    colorlinks,
    citecolor=blue,
    filecolor=black,
    linkcolor=red,
    urlcolor=blue
}
\newcommand{\colAwidth}{0.24\textwidth}
\newcommand{\colBwidth}{0.76\textwidth}

\newcommand{\colCwidth}{0.28\textwidth}
\newcommand{\colDwidth}{0.72\textwidth}

\newcommand{\colEwidth}{0.33\textwidth}
\newcommand{\colFwidth}{0.67\textwidth}

\newcounter{comnum} %Commonality Number
\newcommand{\cthecomnum}{C\thecomnum}
\newcommand{\cref}[1]{C\ref{#1}}

\newcounter{varnum} %Variability Number
\newcommand{\vthevarnum}{V\thevarnum}
\newcommand{\vref}[1]{V\ref{#1}}

\newcounter{parnum} %Parameter of Variation Number
\newcommand{\ptheparnum}{P\theparnum}
\newcommand{\pref}[1]{P\ref{#1}}

\newif\ifcomments\commentstrue

\ifcomments
\newcommand{\authornote}[3]{\textcolor{#1}{[#3 ---#2]}}
\newcommand{\todo}[1]{\textcolor{red}{[TODO: #1]}}
\else
\newcommand{\authornote}[3]{}
\newcommand{\todo}[1]{}
\fi

\newcommand{\wss}[1]{\authornote{blue}{SS}{#1}} %Spencer Smith
\newcommand{\jc}[1]{\authornote{red}{JC}{#1}} %Jacques Carette
\newcommand{\oo}[1]{\authornote{magenta}{OO}{#1}} %Olu Owojaiye
\newcommand{\pmi}[1]{\authornote{green}{PM}{#1}} %Peter Michalski
\newcommand{\ad}[1]{\authornote{cyan}{AD}{#1}} %Ao Dong

\begin{document}

\title{\bf{Commonality Analysis for Lattice Boltzmann Systems}} 
\author{Peter Michalski, \\Department of Computing and Software\\McMaster University}
\date{\today}

\begin{titlepage}
	\clearpage\maketitle
	\thispagestyle{empty}
	\begin{abstract}
		This report presents a commonality analysis for Lattice Boltzmann systems. The document reviews both the methodology of commonality analysis and the details of Lattice Boltzmann systems. The commonality analysis itself consists of the following:
		i) terminology and definitions; ii) commonalities, or features that are common to all potential family members; iii) variabilities, or features and characteristics that may vary among family members; and, iv) parameters of variation, or the potential values that can be assigned to the variabilities. 
		The documentation of the above items for Lattice Boltzmann systems is clarified by decomposing each item into subsections on Lattice Boltzmann methods, input, output, nonfunctional
		requirements, and, where appropriate, system constraints.\\

		\noindent\textbf{Keywords:} Commonality analysis, Lattice Boltzmann methods, program family.

	\end{abstract}
\end{titlepage}

\pagenumbering{roman}

\newpage

\tableofcontents

\newpage

\pagenumbering{arabic}

\section{Introduction}
Mesh generating systems are well suited to development as a program family because they
fit Parnas’s definition of a program family: “a set of programs whose common properties
are so extensive that it is advantageous to study the common properties of the programs
before analyzing individual members” (Parnas, 1976). Developing mesh generating sys-
tems as a program family is advantageous because mesh generators share many common
features, or commonalities. Furthermore, when there are differences between systems, the
variabilities between them can be systematically considered. The purpose of this document
is to record these commonalities and variabilities and show the relationships between them,
and thus facilitate the specification and development of mesh generator program family
members. This document will be valuable in all future phases of system development and
analysis. For instance, the requirements documentation for any mesh generator will use
the commonality analysis, since the requirements should refine the commonalities, which
are shared requirements of all mesh generators. Moreover, the design of any future system
will use this documentation to facilitate consideration of the variabilities, so that likely
changes can be made to the system with a minimal amount of work.
The scope of the commonality analysis presented here can be considered both from the
point of view of mesh generating systems and from the point of view of software engineering
methodologies. From the mesh generator perspective, the starting point for the current
document is the commonality analysis conducted by Chen (2003). However, the scope of
the current document is broader than that of Chen (2003), which was restricted to two-
dimensional meshes and did not consider non-conformal, hybrid or mixed meshes. The
current commonality analysis is intended to cover all mesh generators that are targeted
at finite element applications. Meshes for cartography or other uses are not explicitly
considered, although much of the information in this document does overlap with other
mesh uses. With respect to software engineering methodologies, the scope of the current
report is restricted to informal methods, with the intention that the informal requirements
will form a starting point for later development and refinement by formal methods.
The first section below provides an overview of the program family of mesh generators,
by reviewing what is involved in a commonality analysis and the basics of mesh generating
systems. After this, the basic terminology and definitions necessary for understanding the
remainder of the document are provided. The definitions include terms used in describing
a commonality analysis and terms that are used in defining the characteristics and prop-
erties of meshes. The next three sections consist of the lists of commonalities, variabilities
and parameters of variation, respectively. These three sections form the heart of the docu-
mentation and include an extensive set of cross-references to demonstrate the relationships
between the different items. The final section provides information on unresolved issues.

\newpage
\section{Overview}
This section provides both an overview of the process of commonality analysis and of mesh
generation. The subsection on commonality analysis briefly introduces this topic, along
with references that can be searched for further information. The subsection on mesh
generators outlines the following: i) the basics of mesh generation, ii) the scope of the
current analysis, iii) the overall philosophy that has been adopted for the commonality
analysis, and iv) the arguments in favour of the development of mesh generators as a
program family.

\subsection{Commonality Analysis}
In some situations it is advantageous to develop a collection of related software products
as a program family. The idea is that if the software products are similar enough, then
it should be possible to predict what the products have in common, what differs between
them and then reuse the common aspects and thus support rapid development of the
family members. The idea of program families was introduced by Dijkstra (1972) and later
investigated by Parnas (Parnas, 1976, 1979). More recently, Weiss (Weiss, 1997, 1998;
Ardis and Weiss, 1997) has considered the concept of a program family in the context of
what he terms Family oriented Abstraction, Specification and Translation (FAST) (Cuka
and Weiss, 1997; Weiss and Lai, 1999).
In the approach advocated by Weiss, the first step is a commonality analysis. This
analysis consists of systematically identifying and documenting the commonalities that all
program family members share, the variabilities between family members and the termi-
nology used in describing the family. A commonality analysis provides a systematic way of
gaining confidence that a family is worth building and of deciding what the family members
will be. A commonality analysis document provides the following benefits (Weiss, 1997,
1998):
1. A starting point for the design of a domain specific languages (DSL): Once a DSL,
or application modelling language, is developed the program family members can be
rapidly generated by specifying a given family member using the language.
2. A basis for a common design for all family members: When the software engineers
come to designing the individual family members, they can take advantage of the
commonalities to reuse code. Moreover, the variabilities can be considered in the
design so that they can be easily accommodated. One approach to the design may
be to decompose the system into components that can each be customized by speci-
fication of values for its various parameters, where the parameters correspond to the
parameters of variation identified in the commonality analysis document.
3. A historical reference: This document records the important issues concerning the
scope and the nature of the family (as well as some unsolved issues) to facilitate the
involvement of the participants in maintaining and evolving the family.
CAS-04-10-SS
4
4. A basis for reengineering a domain: Existing projects may not have been developed
using software engineering methodologies. The projects can be systematically reor-
ganized and redesigned with the aid of a commonality analysis to unify the existing
products.
5. A basic training reference for new software developers: This document provides the
necessary basic information for a new team-member to understand the family.
The next section will show that the above uses of the commonality analysis document
will be beneficial for the development of a program family of mesh generators. The com-
monality analysis will benefit all subsequent stages in the software development process.
For instance, as mentioned in the introduction, the commonalities will act as requirements
that will be the starting point for writing a software requirements specification. The com-
monalities will be refined into specific requirements by fixing the value of their associated
variabilities. The change in the values of the variabilities then corresponds to the change
from one program family member to another.
As commonalities and variabilities are requirements, they should express “What” func-
tionalities and qualities the system should have, and not mention “How” these requirements
are to be accomplished. That is, the commonalities and variabilities should not involve
design decisions. The design decisions will be made after the requirements for a family
member have been specified. The one exception to this is system constraints, which are
requirements that explicitly make design decisions.
Besides the “What” versus “How” test, there are other tests that can be used to review
commonalities and variabilities, as proposed by Weiss (1997). One such test is the “what
is ruled out” test. This test determines if a commonality or variability actually makes a
decision because if no alternatives are ruled out then no decision has really been made.
Another test is the “negation” test. If the negation of a decision represents a position
that someone might argue for, then the original decision is likely to be meaningful. For
instance, the statement that “the software should be reliable” has a negation that no one
would likely argue for and thus the statement does not represent a good characterization
of a goal for the system.
In Weiss (1997) the stages of a commonality analysis are described in a systematic way.
The stages include the following: prepare, plan, analyze, quantify and review. The stages
are completed through the aid of a moderator and a series of meeting and preliminary
documents and documentation reviews. Although the systematic approach advocated by
Weiss has its advantages, for the case of writing a commonality analysis document for mesh
generating systems it was decided that a less structured approach is feasible. Mesh genera-
tors are simpler than other software systems in the sense that they have fewer interactions
with the environment. Also, the theory of mesh generation has a solid mathematical ba-
sis that can be used to remove some of the ambiguity that Weiss’s approach is aimed
at reducing. Therefore, the approach adopted here is to revise the original commonality
analysis document produced by Chen (2003) and then make the new document available
to others for review. The new document will be maintained in a concurrent versioning
CAS-04-10-SS
5
system repository so that multiple authors can work on it, and more importantly, so that
the documentation’s revision history can be tracked and the documentation can be rolled
back to an earlier version if necessary.

\subsection{Lattice Boltzmann Systems}
\newpage
\section{Terminology and Definitions}
This section is divided into two subsections. The first discusses the terminology that comes
from the software engineering field, while the second presents the definitions used in mesh
generation. Common acronyms are also listed in this section. The lists are not intended
to be read sequentially, but rather to be consulted for reference purposes; therefore, the
terms are ordered alphabetically, with the consequence that some terms that appear early
in the list depend on the definitions of later terms.

\subsection{Software Engineering Related Definitions and Acronyms}
Commonality: A requirement or goal common to all family members.
Goal: “Goals capture, at different levels of abstraction, the various objectives the system
under consideration should achieve.” van Lamsweerde (2001)
Program Family: A set of programs that are analyzed and designed together starting
from the initial stages of the software development life-cycle.
Requirements: A software requirement is: i) a condition or capability needed by a user
to solve a problem or achieve an objective; ii) a condition or capability that must be
met or possessed by a system or system component to satisfy a contract, standard,
specification, or other formally imposed document; or, iii) a documented represen-
CAS-04-10-SS
10
tation of a condition or capability as in the above two definitions.
Dorfman, 2000)
Variability: A requirement or goal that varies between family members.
\subsection{Lattice Boltzmann Related Definitions and Acronyms}
\newpage
\section{Commonalities}
\subsection{Lattice Boltzmann Method Solvers}
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
		\hline
		\bf Item Number& C\refstepcounter{comnum}\thecomnum \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
		\hline
		\bf Description & \\
		\hline
		\bf Related Variability & \vref{}\\
		\hline
		\bf History & \\
		\hline
	\end{tabular}
\end{minipage}\\
~\newline\\
~\newline
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
		\hline
		\bf Item Number& C\refstepcounter{comnum}\thecomnum \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
		\hline
		\bf Description & \\
		\hline
		\bf Related Variability & \vref{}\\
		\hline
		\bf History & \\
		\hline
	\end{tabular}
\end{minipage}\\
\subsection{Input}
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
		\hline
		\bf Item Number& C\refstepcounter{comnum}\thecomnum \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
		\hline
		\bf Description & \\
		\hline
		\bf Related Variability & \vref{}\\
		\hline
		\bf History & \\
		\hline
	\end{tabular}
\end{minipage}\\
~\newline\\
~\newline
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
		\hline
		\bf Item Number& C\refstepcounter{comnum}\thecomnum \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
		\hline
		\bf Description & \\
		\hline
		\bf Related Variability & \vref{}\\
		\hline
		\bf History & \\
		\hline
	\end{tabular}
\end{minipage}\\
\subsection{Output}
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
		\hline
		\bf Item Number& C\refstepcounter{comnum}\thecomnum \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
		\hline
		\bf Description & \\
		\hline
		\bf Related Variability & \vref{}\\
		\hline
		\bf History & \\
		\hline
	\end{tabular}
\end{minipage}\\
~\newline\\
~\newline
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
		\hline
		\bf Item Number& C\refstepcounter{comnum}\thecomnum \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
		\hline
		\bf Description & \\
		\hline
		\bf Related Variability & \vref{}\\
		\hline
		\bf History & \\
		\hline
	\end{tabular}
\end{minipage}\\
\subsection{Nonfunctional Requirements}
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
		\hline
		\bf Item Number& C\refstepcounter{comnum}\thecomnum \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
		\hline
		\bf Description & \\
		\hline
		\bf Related Variability & \vref{}\\
		\hline
		\bf History & \\
		\hline
	\end{tabular}
\end{minipage}\\
~\newline\\
~\newline

\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
		\hline
		\bf Item Number& C\refstepcounter{comnum}\thecomnum \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
		\hline
		\bf Description & \\
		\hline
		\bf Related Variability & \vref{}\\
		\hline
		\bf History & \\
		\hline
	\end{tabular}
\end{minipage}\\
\newpage
\section{Variabilities}
\subsection{Lattice Boltzmann Method Solvers}
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
		\hline
		\bf Item Number& V\refstepcounter{varnum}\thevarnum \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
		\hline
		\bf Description & \\
		\hline
		\bf Related Commonality & \cref{}\\
		\hline
		\bf Related Parameter & \pref{}\\
		\hline
		\bf History & \\
		\hline
	\end{tabular}
\end{minipage}\\
~\newline\\
~\newline
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
		\hline
		\bf Item Number& V\refstepcounter{varnum}\thevarnum \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
		\hline
		\bf Description & \\
		\hline
		\bf Related Commonality & \cref{}\\
		\hline
		\bf Related Parameter & \pref{}\\
		\hline
		\bf History & \\
		\hline
	\end{tabular}
\end{minipage}\\
\subsection{Input}
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
		\hline
		\bf Item Number& V\refstepcounter{varnum}\thevarnum \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
		\hline
		\bf Description & \\
		\hline
		\bf Related Commonality & \cref{}\\
		\hline
		\bf Related Parameter & \pref{}\\
		\hline
		\bf History & \\
		\hline
	\end{tabular}
\end{minipage}\\
~\newline\\
~\newline
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
		\hline
		\bf Item Number& V\refstepcounter{varnum}\thevarnum \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
		\hline
		\bf Description & \\
		\hline
		\bf Related Commonality & \cref{}\\
		\hline
		\bf Related Parameter & \pref{}\\
		\hline
		\bf History & \\
		\hline
	\end{tabular}
\end{minipage}\\
\subsection{Output}
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
		\hline
		\bf Item Number& V\refstepcounter{varnum}\thevarnum \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
		\hline
		\bf Description & \\
		\hline
		\bf Related Commonality & \cref{}\\
		\hline
		\bf Related Parameter & \pref{}\\
		\hline
		\bf History & \\
		\hline
	\end{tabular}
\end{minipage}\\
~\newline\\
~\newline
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
		\hline
		\bf Item Number& V\refstepcounter{varnum}\thevarnum \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
		\hline
		\bf Description & \\
		\hline
		\bf Related Commonality & \cref{}\\
		\hline
		\bf Related Parameter & \pref{}\\
		\hline
		\bf History & \\
		\hline
	\end{tabular}
\end{minipage}\\
\subsection{System Constraints}
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
		\hline
		\bf Item Number& V\refstepcounter{varnum}\thevarnum \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
		\hline
		\bf Description & \\
		\hline
		\bf Related Commonality & \cref{}\\
		\hline
		\bf Related Parameter & \pref{}\\
		\hline
		\bf History & \\
		\hline
	\end{tabular}
\end{minipage}\\
~\newline\\
~\newline
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
		\hline
		\bf Item Number& V\refstepcounter{varnum}\thevarnum \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
		\hline
		\bf Description & \\
		\hline
		\bf Related Commonality & \cref{}\\
		\hline
		\bf Related Parameter & \pref{}\\
		\hline
		\bf History & \\
		\hline
	\end{tabular}
\end{minipage}\\
\subsection{Nonfunctional Requirements}
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
		\hline
		\bf Item Number& V\refstepcounter{varnum}\thevarnum \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
		\hline
		\bf Description & \\
		\hline
		\bf Related Commonality & \cref{}\\
		\hline
		\bf Related Parameter & \pref{}\\
		\hline
		\bf History & \\
		\hline
	\end{tabular}
\end{minipage}\\
~\newline\\
~\newline
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
		\hline
		\bf Item Number& V\refstepcounter{varnum}\thevarnum \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
		\hline
		\bf Description & \\
		\hline
		\bf Related Commonality & \cref{}\\
		\hline
		\bf Related Parameter & \pref{}\\
		\hline
		\bf History & \\
		\hline
	\end{tabular}
\end{minipage}\\
\newpage
\section{Parameters of Variation}
\subsection{Lattice Boltzmann Method Solvers}
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
		\hline
		\bf Item Number& P\refstepcounter{parnum}\theparnum \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
		\hline
		\bf Corresponding Variability & \vref{}\\
		\hline
		\bf Range of Parameters & \\
		\hline
		\bf Binding Time & \\
		\hline
	\end{tabular}
\end{minipage}\\
~\newline\\
~\newline
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
		\hline
		\bf Item Number& P\refstepcounter{parnum}\theparnum \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
		\hline
		\bf Corresponding Variability & \vref{}\\
		\hline
		\bf Range of Parameters & \\
		\hline
		\bf Binding Time & \\
		\hline
	\end{tabular}
\end{minipage}\\
\subsection{Input}
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
		\hline
		\bf Item Number& P\refstepcounter{parnum}\theparnum \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
		\hline
		\bf Corresponding Variability & \vref{}\\
		\hline
		\bf Range of Parameters & \\
		\hline
		\bf Binding Time & \\
		\hline
	\end{tabular}
\end{minipage}\\
~\newline\\
~\newline
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
		\hline
		\bf Item Number& P\refstepcounter{parnum}\theparnum \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
		\hline
		\bf Corresponding Variability & \vref{}\\
		\hline
		\bf Range of Parameters & \\
		\hline
		\bf Binding Time & \\
		\hline
	\end{tabular}
\end{minipage}\\
\subsection{Output}
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
		\hline
		\bf Item Number& P\refstepcounter{parnum}\theparnum \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
		\hline
		\bf Corresponding Variability & \vref{}\\
		\hline
		\bf Range of Parameters & \\
		\hline
		\bf Binding Time & \\
		\hline
	\end{tabular}
\end{minipage}\\
~\newline\\
~\newline
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
		\hline
		\bf Item Number& P\refstepcounter{parnum}\theparnum \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
		\hline
		\bf Corresponding Variability & \vref{}\\
		\hline
		\bf Range of Parameters & \\
		\hline
		\bf Binding Time & \\
		\hline
	\end{tabular}
\end{minipage}\\
\subsection{System Constraints}
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
		\hline
		\bf Item Number& P\refstepcounter{parnum}\theparnum \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
		\hline
		\bf Corresponding Variability & \vref{}\\
		\hline
		\bf Range of Parameters & \\
		\hline
		\bf Binding Time & \\
		\hline
	\end{tabular}
\end{minipage}\\
~\newline\\
~\newline
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
		\hline
		\bf Item Number& P\refstepcounter{parnum}\theparnum \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
		\hline
		\bf Corresponding Variability & \vref{}\\
		\hline
		\bf Range of Parameters & \\
		\hline
		\bf Binding Time & \\
		\hline
	\end{tabular}
\end{minipage}\\
\subsection{Nonfunctional Requirements}
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
		\hline
		\bf Item Number& P\refstepcounter{parnum}\theparnum \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
		\hline
		\bf Corresponding Variability & \vref{}\\
		\hline
		\bf Range of Parameters & \\
		\hline
		\bf Binding Time & \\
		\hline
	\end{tabular}
\end{minipage}\\
~\newline\\
~\newline
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
		\hline
		\bf Item Number& P\refstepcounter{parnum}\theparnum \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
		\hline
		\bf Corresponding Variability & \vref{}\\
		\hline
		\bf Range of Parameters & \\
		\hline
		\bf Binding Time & \\
		\hline
	\end{tabular}
\end{minipage}\\
\newpage
\section{Issues}
\newpage
\bibliographystyle {plainnat}
\bibliography {../../CommonFiles/ResearchProposal}
\newpage
\section{Appendix}
\end{document}

### 2020-03-01 StateOfPractice/Ao-Notes/Ao-Notes.tex
Relocate to a new folder

--- 2020-02-09 StateOfPractice/Ao-Notes.tex
@@
 
-Link to the software list (Google Sheet): \href{https://docs.google.com/spreadsheets/d/122wU0v3ZtvDcqy8C4zKJ89kU-8fXAbo3Mzn6vcVXOi0/edit?usp=sharing}{LINK}
+The link to the software list (Google Sheet): \href{https://docs.google.com/spreadsheets/d/122wU0v3ZtvDcqy8C4zKJ89kU-8fXAbo3Mzn6vcVXOi0/edit?usp=sharing}{LINK}
+
+\noindent There are 50 packages in the list.
+
+\subsection{Sources}
+
+The above list is built based on 7 sources - 3 academical papers, 3 blog articles, and 1 online forum discussion. All of the sources are recorded on the second sheet in the list.
+
+\subsection{Orders}
+
+The software packages are put in the list roughly by the order of how many times they are mentioned in the above sources.
+
+\subsection{Selection}
+
+A few software packages mentioned in the sources are excluded, since they are in the following cases:
+\begin{itemize}
+    \item Commercialized. Some software are not free any more. For example, PacsOne Premium, JVS-DiComPlus, RadiAnt, Synedra, TurtleSeg, AlgoM, Athena.
+    \item No recent updates.
+A few software without recent release are excluded, such as PacsOne Basic (2005), K-PACS (developed for Windows XP or Windows VISTA), JVS-DiCom (2012), MITO-DICOM Viewer (2015), Aeskulap (2007), Fusionviewer (2008), Kradview (2013), Mayam (2013), Eviewbox (2015), EzDICOM (2013), Agnosco (2011), ONIS Free Version (2016 and not open-source), DICOMscope (2001).
+\end{itemize}
+
+\subsection{Categories}
+
+The software packages are roughly divided into 3 major categories - Tool Kit, PACS, and Viewer.
+\begin{itemize}
+    \item Tool Kits are usually used by developers to create new software, and there are only a few of them in the list. 
+    \item PACS stands for Picture Archiving and Communications System, which can be regarded as a type of server.
+There are also only several of them in the list, since most PACS solutions are commercial nowadays.
+    \item  Most software in the list are in the Viewer category, and many of these ones also provide other functions such as analysis, conversion, segmentation, etc.
+\end{itemize}
 

### 2020-03-05 StateOfPractice/Peter-Notes/Peter-Notes-SolutionDescriptions.tex
added a few LBM solvers

@@
 It is supplied with its own Java-based Graphical User Interface (GUI) and is capable of both serial and parallel execution.
+
+	\item elbe \\
+	\href{https://www.tuhh.de/alt/elbe/home.html}{https://www.tuhh.de/alt/elbe/home.html}\\
+	Found using www.swmath.org \\
+	elbe is a highly efficient academic flow solver for the simulation of non-linear three-dimensional flow problems, such as violent impact, tank sloshing or Tsunami propagation, and highly turbulent singlephase flows for drag resistance, aerodynamic noise control and fluid-structure interactions.
+elbe, the efficient lattice boltzmann environment, is accelerated with recent graphics hardware (graphics processing units, GPUs).
+Thanks to the very efficient numerical backend, three-dimensional simulations of complex flows are possible in or near real-time.
+elbe can be considered as a prototype for next-generation CFD tools that allow for a real-time analysis of complex flow fields and simulation-based design.
+Being developed since late 2010, the main development efforts currently take place at the Institute for Fluid Dynamics and Ship Theory (FDS) at Hamburg University of Technology (TUHH), with external partners and contributors at University of Rhode Island and TU Braunschweig.
+
+	\item eLBM \\
+	No direct source\\
+	Found using www.swmath.org \\
+	Direct simulation of pore-scale two-phase visco-capillary flow on large digital rock images using a phase-field lattice Boltzmann method on general-purpose graphics processing units.
 
@@
 
-	\item lettuce
-\\
-	\href{https://github.com/Olllom/lettuce}{https://github.com/Olllom/lettuce}
-\\
-	Found using Google search.\\
-	GPU-acclerated Lattice Boltzmann in Python
-
 	\item LB2D\_Prime
@@
 
+	\item LBM-EP \\
+	No direct source\\
+	Found using www.swmath.org \\	
+	LBM-EP: lattice-Boltzmann method for fast cardiac electrophysiology simulation from 3D images.
+Current treatments of heart rhythm troubles require careful planning and guidance for optimal outcomes.
+Computational models of cardiac electrophysiology are being proposed for therapy planning but current approaches are either too simplified or too computationally intensive for patient-specific simulations in clinical practice.
+This paper presents a novel approach, LBM-EP, to solve any type of mono-domain cardiac electrophysiology models at near real-time that is especially tailored for patient-specific simulations.
+The domain is discretized on a Cartesian grid with a level-set representation of patient’s heart geometry, previously estimated from images automatically.
+The cell model is calculated node-wise, while the transmembrane potential is diffused using Lattice-Boltzmann method within the domain defined by the level-set.
+Experiments on synthetic cases, on a data set from CESC’10 and on one patient with myocardium scar showed that LBM-EP provides results comparable to an FEM implementation, while being 10 − 45 times faster.
+Fast, accurate, scalable and requiring no specific meshing, LBM-EP paves the way to efficient and detailed models of cardiac electrophysiology for therapy planning.
+
+	\item lbmpy \\
+	\href{https://pypi.org/project/lbmpy/}{https://pypi.org/project/lbmpy/}\\
+	Found using www.swmath.org \\
+	Run fast fluid simulations based on the lattice Boltzmann method in Python on CPUs and GPUs.
+lbmpy creates highly optimized LB compute kernels in C or CUDA, for a wide variety of different collision operators, including MRT, entropic, and cumulant schemes.
+
 	\item LBSim
@@
 	A small and simple Lattice-Boltzmann Method fluid simulator supporting complex boundaries.
+
+	\item lettuce
+	\\
+	\href{https://github.com/Olllom/lettuce}{https://github.com/Olllom/lettuce}
+	\\
+	Found using Google search.\\
+	GPU-acclerated Lattice Boltzmann in Python
 
@@
 
+	\item  loliverhennigh /
+	Lattice-Boltzmann-fluid-flow-in-Tensorflow 
+	\\
+	\href{https://github.com/loliverhennigh/Lattice-Boltzmann-fluid-flow-in-Tensorflow}{https://github.com/loliverhennigh/Lattice-Boltzmann-fluid-flow-in-Tensorflow}
+	\\
+	Found using Google search.\\
+	A Lattice Boltzmann fluid flow simulation written in Tensorflow.
+
+	\item Ludwig
+	\\
+	\href{https://github.com/ludwig-cf/ludwig}{https://github.com/ludwig-cf/ludwig}\\
+	Found using Google search.\\
+	Ludwig is a parallel code for the simulation of complex fluids, which include mixtures, colloidal suspensions, gels, and liquid crystals.
+It takes its name from Ludwig Boltzmann, as it uses a lattice Boltzmann method as a basis for numerical solution of the Navier Stokes equations for hydrodynamics.
+It typically combines hydrodynamics with a coarse-grained order parameter (or order parameters) to represent the "complex" part in a free energy picture.
+The code is written in standard ANSI C, and uses the Message Passing Interface for distributed memory parallelism.
+Threaded parallelism is also available via a lightweight abstraction layer ("Target Data Parallel" or "TargetDP") which currently supports either OpenMP or CUDA (NVIDIA GPUs) from a single source.
+
 	\item LUMA
@@
 	LUMA: A many-core, Fluid–Structure Interaction solver based on the Lattice-Boltzmann Method
+
+	\item Mechsys \\
+	\href{http://mechsys.nongnu.org/}{http://mechsys.nongnu.org/}\\
+	Found using www.swmath.org \\
+	MechSys is a programming library for the implementation of simulation tools in mechanics. Its source code is mainly written in C++ with easier to use templates for further customization.
 
@@
 
+	\item openLBMflow \\
+	\href{https://github.com/wme7/openLBMflow}{https://github.com/wme7/openLBMflow}\\
+	Found using www.swmath.org \\
+	The openLBMflow is an fast fluid flow solver based on Lattice Boltzmann Method.
+Main future are 2D and 3D code, single and multiphase models, Output data in VTK format can be directly open in Paraview.
+Download Windows or Linux binary version now.
+
 	\item Palabos \\
@@
 	The library’s native programming interface in written in C++.
+
+	\item ParallelLbmCranfield \\
+	\href{https://github.com/mate-szoke/ParallelLbmCranfield}{https://github.com/mate-szoke/ParallelLbmCranfield}\\
+	Found using www.swmath.org \\
+	This software is a two dimensional parallel Lattice Boltzmann Method solver implemented in CUDA and PGAS UPC
+
+	\item PowerFLOW \\
+	\href{https://www.3ds.com/products-services/simulia/products/powerflow/}{https://www.3ds.com/products-services/simulia/products/powerflow/}\\
+	Found using www.swmath.org \\
+	PowerFLOW: Through the usage of our unique, inherently transient Lattice Boltzmann-based physics, Exa’s PowerFLOW® CFD solution performs simulations that accurately predict real world conditions.
+Using the PowerFLOW suite, engineers evaluate product performance early in the design process prior to any prototype being built — when the impact of change is most significant for design and budgets.
+PowerFLOW imports fully complex model geometry and accurately and efficiently performs aerodynamic, aeroacoustic and thermal management simulations.
+
+	\item ProLB \\
+	\href{http://www.prolb-cfd.com/}{http://www.prolb-cfd.com/}
+	\\
+	Found using Google search.\\
+	ProLB is an innovative Computational Fluid Dynamics (CFD) software solution.
+Based on the Lattice-Boltzmann method, its successfully-validated solver performs inherently transient simulations of highly complex flows with a competitive turnaround time.
+ProLB‘s accurate aerodynamic and aeroacoustic modeling allows engineers to make early design decisions that optimize and shorten the product development process.
 
@@
 	Taxila LBM is a parallel implementation of the Lattice Boltzmann Method for simulation of flow in porous and geometrically complex media.
-
-	\item  loliverhennigh /
-Lattice-Boltzmann-fluid-flow-in-Tensorflow 
-\\
-	\href{https://github.com/loliverhennigh/Lattice-Boltzmann-fluid-flow-in-Tensorflow}{https://github.com/loliverhennigh/Lattice-Boltzmann-fluid-flow-in-Tensorflow}
-\\
-	Found using Google search.\\
-	A Lattice Boltzmann fluid flow simulation written in Tensorflow.
 

### 2020-03-05 StateOfPractice/Ao-Notes/Ao-Notes.tex

@@
 \usepackage{longtable}
+\theoremstyle{definition}
+\newtheorem{defn}{Definition}
 
@@
 Feb/08/2020 & Ao Dong & Changed table to a spreadsheet link\\
+Mar/01/2020 & Ao Dong & Updated quality measurements\\
 \bottomrule
@@
 
-The link to the software list (Google Sheet): \href{https://docs.google.com/spreadsheets/d/122wU0v3ZtvDcqy8C4zKJ89kU-8fXAbo3Mzn6vcVXOi0/edit?usp=sharing}{LINK}
+The link to the software list (Google Sheet):
+\href{https://docs.google.com/spreadsheets/d/122wU0v3ZtvDcqy8C4zKJ89kU-8fXAbo3Mzn6vcVXOi0/edit?usp=sharing}{LINK}
 
@@
 
-The above list is built based on 7 sources - 3 academical papers, 3 blog articles, and 1 online forum discussion. All of the sources are recorded on the second sheet in the list.
+The above list is built based on 7 sources - 3 academical papers, 3 blog
+articles, and 1 online forum discussion. All of the sources are recorded on the
+second sheet in the list.
 
@@
 
-The software packages are put in the list roughly by the order of how many times they are mentioned in the above sources.
+The software packages are put in the list roughly by the order of how many times
+they are mentioned in the above sources.
 
@@
 
-A few software packages mentioned in the sources are excluded, since they are in the following cases:
+A few software packages mentioned in the sources are excluded, since they are in
+the following cases:
 \begin{itemize}
-    \item Commercialized. Some software are not free any more. For example, PacsOne Premium, JVS-DiComPlus, RadiAnt, Synedra, TurtleSeg, AlgoM, Athena.
-    \item No recent updates.
-A few software without recent release are excluded, such as PacsOne Basic (2005), K-PACS (developed for Windows XP or Windows VISTA), JVS-DiCom (2012), MITO-DICOM Viewer (2015), Aeskulap (2007), Fusionviewer (2008), Kradview (2013), Mayam (2013), Eviewbox (2015), EzDICOM (2013), Agnosco (2011), ONIS Free Version (2016 and not open-source), DICOMscope (2001).
+\item Commercialized. Some software are not free any more. For example, PacsOne
+Premium, JVS-DiComPlus, RadiAnt, Synedra, TurtleSeg, AlgoM, Athena.
+\item No recent updates. A few software without recent release are excluded,
+such as PacsOne Basic (2005), K-PACS (developed for Windows XP or Windows
+VISTA), JVS-DiCom (2012), MITO-DICOM Viewer (2015), Aeskulap (2007),
+Fusionviewer (2008), Kradview (2013), Mayam (2013), Eviewbox (2015), EzDICOM
+(2013), Agnosco (2011), ONIS Free Version (2016 and not open-source), DICOMscope
+(2001).
 \end{itemize}
@@
 
-The software packages are roughly divided into 3 major categories - Tool Kit, PACS, and Viewer.
+The software packages are roughly divided into 3 major categories - Tool Kit,
+PACS, and Viewer.
 \begin{itemize}
-    \item Tool Kits are usually used by developers to create new software, and there are only a few of them in the list. 
-    \item PACS stands for Picture Archiving and Communications System, which can be regarded as a type of server.
-There are also only several of them in the list, since most PACS solutions are commercial nowadays.
-    \item  Most software in the list are in the Viewer category, and many of these ones also provide other functions such as analysis, conversion, segmentation, etc.
+\item Tool Kits are usually used by developers to create new software, and there
+are only a few of them in the list.
+\item PACS stands for Picture Archiving and Communications System, which can be
+regarded as a type of server. There are also only several of them in the list,
+since most PACS solutions are commercial nowadays.
+\item Most software in the list are in the Viewer category, and many of these
+ones also provide other functions such as analysis, conversion, segmentation,
+etc.
 \end{itemize}
@@
 \subsection{Interoperability}
+\begin{defn}
+The degree to which two or more systems, products or components can exchange
+information and use the information that has been exchanged.
+\end{defn}
+
 There are not many measuring methods in papers, and most of them are very
@@
 \begin{itemize}
-\item Can the workstation software (for visualization, analysis, etc.) connect
-with the PACS (server)?
-    \item Does the software use output from or provide input to other software?
+\item Does the software use any API to transfer information?
+\item Does the software generate output files with uncommon formats or
+extensions that can only be used by itself?
+\item Can the output be used by other software as input?
+\item Can it take output from other software as input?
 \item Can the software work with customized plug-ins? 
@@
 \subsection{Visibility/Transparency}
+\begin{defn}
+The extent to which all of the steps of a software development process and the
+current status of it are conveyed clearly.
+\end{defn}
+The ones with a * mark might be hard to know.
 \begin{itemize}
 \item Does the software use any version and issue tracking system, such as
-Github, Gitlab for development?
+Github, Gitlab, JIRA, Rally, VersionOne, etc.?
+\item The percentage of (open) issues assigned to specific developers.
 \item Does the software have documents recording the development process and
@@
 as release date, bug fixed and new features?
+\item *Are there weekly or monthly status reports for the development process?
+\item *Are there any project management tools used?
+\item *Are there any project indicators used, such as burndown charts,
+cumulative flow diagrams, and other metrics?
 \end{itemize}
@@
 no, n/a})
-\item Ease of external examination relative to other products considered? ({1 ..
-10})
+\item Ease of external examination relative to other products considered? ({1 ..10})
 \end{itemize}
@@
 \subsection{Productivity}
-Can be measured by the summation of all output (such as the number of lines of
-new code, the number of pages of new documents and the number of new test cases)
+\begin{defn}
+Productivity is the amount of output per unit of input used, which can be
+measured by the summation of all output (such as the number of lines of new
+code, the number of pages of new documents and the number of new test cases)
 produced per person-day.
+\end{defn}
 
-However, it is hard for a third party to carry out the above method without
-knowing the exact number of developers or how much time they spent on the
-project.
+It can be hard for a third party to know how many days the developers worked on
+the project.
+
+\begin{itemize}
+\item The lines of new code produced per person-day.
+\item The pages of new documents produced per person-day.
+\item The number of new test cases finished per person-day.
+\item The number of issues closed per person-day.
+\end{itemize}
 
 \subsection{Completeness}
+\begin{defn}
+A specification is complete to the extent that all of its parts are present and
+each part is fully developed.
+\end{defn}
+
+Are there the following documents?
+\begin{itemize}
+\item SRS
+\item VnV plan \& reports
+\item MG \& MIS
+\item Readme
+\item User manual (including tutorials installation instructions) 
+\item Release log
+\end{itemize}
+Inspired by \citep{Boehm1984}:
+\begin{itemize}
+\item Is there anything left to be determined?
+\item Is there any references in the specification to functions, inputs, or
+outputs (including databases) not defined in the specification?
+\item Is there any specification items missing?
+\item Is there any functions missing?
+\item For multi-products software package, is there any products missing?
+\end{itemize}
 
 \subsection{Consistency}
+\begin{defn}
+A specification is consistent to the extent that its provisions do not conflict
+with each other or with governing specifications and objectives.
+\end{defn}
+
+Are the following items of documents consistent?
+\begin{itemize}
+\item Language
+\item File format (pdf, html)
+\item Location (GitHub wiki, offline)
+\item Wording format
+\item Syntax
+\item Abbreviations and Acronyms
+\end{itemize}
+
+Inspired by \citep{Boehm1984}:
+\begin{itemize}
+\item Is there any specification items conflicting with each other in one
+document?
+\item Is there any conflicts between different documents?
+\end{itemize}
 
@@
 \bibliographystyle {plainnat}
-\bibliography {../CommonFiles/ResearchProposal}
+\bibliography {../../CommonFiles/ResearchProposal}
 

### 2020-03-12 StateOfPractice/Methodology/Methodology.tex
Updates to Methodology for assessing quality

@@
 
+\section{Overview of Steps in Assessing Quality of the Domain Software}
+
+\begin{enumerate}
+\item Identify domain.  (Provide criteria on a candidate domain.)
+\item 
+\end{enumerate}
+
+\section{Identify Candidate Software}
+
 \section{Domain Analysis} \label{SecDomainAnalysis}
@@
 
-\section{Quality Measures}
+\section{Empirical Measures}
 
-Go through each quality and describe how it will be measured.
+Measures that can be extracted from on-line repos.
 
-Start with decided upon definition.  (If discussion is on-going, please select
-the ``best'' definition, but include a warning that it may change.)
+\section{User Experiments}
+
+Describe experiments with users to assess usability, performance etc.
+
+\section{Analytic Hierarchy Process}
+
+Describe process.  Domain expert review.
+
+\section{Quality Specific Measures}
 

### 2020-03-16 StateOfPractice/Methodology/Methodology.tex
Update Methodology

@@
 
+\subsection{Raw Data}
 Measures that can be extracted from on-line repos.
+
+\ad{Still at brainstorm stage.}
+\begin{itemize}
+\item number of contributors
+\item number of watches
+\item number of stars
+\item number of forks
+\item number of clones
+\item number of commits
+\item number of total/code/document files
+\item lines of total/logical/comment code
+\item lines/pages of documents (can pdf be extracted?)
+\item number of total/open/closed/merged pull requests
+\item number of total/open/closed issues
+\item number of total/open/closed issues with assignees
+\end{itemize}
+
+\subsection{Processed Data}
+Metrics that can be calculated from the raw data.
+
+\ad{Still at brainstorm stage.}
+\begin{itemize}
+\item percentage of total/open/closed issues with assignees -
+Visibility/Transparency
+\item lines of new code produced per person-day - Productivity
+\item lines/pages of new documents produced per person-day - Productivity
+\item number of issues closed per person-day - Productivity
+\item percentage of comment lines in the code - maintainability \ad{Not Ao's
+qualities}
+\end{itemize}
+
+\subsection{Tool Test - HubListener}
+\ad{This section is currently a note of unorganized contents. Most parts will be
+removed or relocated.}
+
+\hyperlink{https://github.com/pjmc-oliveira/HubListener}{GitHub repo}
+
+Raw:
+\begin{itemize}
+\item Number of Files
+\item Number of Lines
+\item Number of Logical Lines
+\item Number of Comments
+\end{itemize}
+
+Cyclomatic:
+\hyperlink{https://www.geeksforgeeks.org/cyclomatic-complexity/}{Intro}
+\begin{itemize}
+\item Cyclomatic Complexity
+\end{itemize}
+
+Halstead:
+\hyperlink{https://www.geeksforgeeks.org/software-engineering-halsteads-software-metrics/}{Intro}
+\begin{itemize}
+\item Halstead Effort
+\item Halstead Bugs
+\item Halstead Length
+\item Halstead Difficulty
+\item Halstead Time
+\item Halstead Vocabulary
+\item Halstead Volume
+\end{itemize}
+
+Environment 1:
+\begin{itemize}
+\item Hardware: laptop
+\item OS: Windows 10
+\item Node.js version: 12
+\item npm version: 6
+\item Does it work: No
+\item Errors: errors during "npm install" step
+\end{itemize}
+
+Environment 2:
+\begin{itemize}
+\item Hardware: Google Cloud VM
+\item OS: Ubuntu 18.04
+\item Node.js version: 8
+\item npm version: 3
+\item Does it work: No
+\item Errors: errors during "npm install" step
+\end{itemize}
+
+Environment 3:
+\begin{itemize}
+\item Hardware: Google Cloud VM
+\item OS: Ubuntu 18.04
+\item Node.js version: 8
+\item npm version: 6
+\item Does it work: Yes
+\item Errors: see following tests
+\end{itemize}
+
+Test 1 on environment 3: 
+\begin{itemize}
+\item Repo: 3D Slicer \hyperlink{https://github.com/Slicer/Slicer}{Github repo}
+\item Command line or web app: web app
+\item Quick Analyze checked: Yes
+\item Time Consumption: 96s
+\item Errors/problems: data of total/open/closed issues are missing, data of
+Cyclomatic and Halstead metrics are missing.
+\end{itemize}
+
+Test 2 on environment 3: 
+\begin{itemize}
+\item Repo: 3D Slicer \hyperlink{https://github.com/Slicer/Slicer}{Github repo}
+\item Command line or web app: web app
+\item Quick Analyze checked: No
+\item Time Consumption: 4.3hrs
+\item Errors/problems: the web app lost response after the analysis
+\end{itemize}
+
+Test 3/4 on environment 3: 
+\begin{itemize}
+\item Ao's repo \hyperlink{https://github.com/Ao99/e-commerce-django}{Github
+repo}
+\item Command line or web app: web app
+\item Quick Analyze checked: Yes/No
+\item Time Consumption: 121s/117s
+\item Errors/problems: most data are missing
+\end{itemize}
+
+Test 5/6 on environment 3: 
+\begin{itemize}
+\item HubListener
+\hyperlink{https://github.com/pjmc-oliveira/HubListener}{Github repo}
+\item Command line or web app: web app
+\item Quick Analyze checked: Yes/No
+\item Time Consumption: 49s/131s
+\item Errors/problems: none
+\end{itemize}
+
+\ad{This citation needs to be deleted later. It's here because my compiler
+doesn't work with 0 citations}
+\cite{Emms2019}
 

### 2020-03-16 StateOfPractice/Methodology/Methodology.tex
Updated website links

@@
 
-\hyperlink{https://github.com/pjmc-oliveira/HubListener}{GitHub repo}
+\href{https://github.com/pjmc-oliveira/HubListener}{GitHub repo}
 
@@
 Cyclomatic:
-\hyperlink{https://www.geeksforgeeks.org/cyclomatic-complexity/}{Intro}
+\href{https://www.geeksforgeeks.org/cyclomatic-complexity/}{Intro}
 \begin{itemize}
@@
 Halstead:
-\hyperlink{https://www.geeksforgeeks.org/software-engineering-halsteads-software-metrics/}{Intro}
+\href{https://www.geeksforgeeks.org/software-engineering-halsteads-software-metrics/}{Intro}
 \begin{itemize}
@@
 \begin{itemize}
-\item Repo: 3D Slicer \hyperlink{https://github.com/Slicer/Slicer}{Github repo}
+\item Repo: 3D Slicer \href{https://github.com/Slicer/Slicer}{Github repo}
 \item Command line or web app: web app
@@
 \begin{itemize}
-\item Ao's repo \hyperlink{https://github.com/Ao99/e-commerce-django}{Github
+\item Ao's repo \href{https://github.com/Ao99/e-commerce-django}{Github
 repo}
@@
 \item HubListener
-\hyperlink{https://github.com/pjmc-oliveira/HubListener}{Github repo}
+\href{https://github.com/pjmc-oliveira/HubListener}{Github repo}
 \item Command line or web app: web app

### 2020-03-20 StateOfPractice/Methodology/Methodology.tex
Update Methodology

@@
 
+The data that HubListener can extract.
+
 Raw:
@@
 
-Environment 1:
-\begin{itemize}
-\item Hardware: laptop
-\item OS: Windows 10
-\item Node.js version: 12
-\item npm version: 6
-\item Does it work: No
-\item Errors: errors during "npm install" step
-\end{itemize}
-
-Environment 2:
-\begin{itemize}
-\item Hardware: Google Cloud VM
-\item OS: Ubuntu 18.04
-\item Node.js version: 8
-\item npm version: 3
-\item Does it work: No
-\item Errors: errors during "npm install" step
-\end{itemize}
-
-Environment 3:
-\begin{itemize}
-\item Hardware: Google Cloud VM
-\item OS: Ubuntu 18.04
-\item Node.js version: 8
-\item npm version: 6
-\item Does it work: Yes
-\item Errors: see following tests
-\end{itemize}
-
-Test 1 on environment 3: 
-\begin{itemize}
-\item Repo: 3D Slicer \href{https://github.com/Slicer/Slicer}{Github repo}
-\item Command line or web app: web app
-\item Quick Analyze checked: Yes
-\item Time Consumption: 96s
-\item Errors/problems: data of total/open/closed issues are missing, data of
-Cyclomatic and Halstead metrics are missing.
-\end{itemize}
-
-Test 2 on environment 3: 
-\begin{itemize}
-\item Repo: 3D Slicer \href{https://github.com/Slicer/Slicer}{Github repo}
-\item Command line or web app: web app
-\item Quick Analyze checked: No
-\item Time Consumption: 4.3hrs
-\item Errors/problems: the web app lost response after the analysis
-\end{itemize}
-
-Test 3/4 on environment 3: 
-\begin{itemize}
-\item Ao's repo \href{https://github.com/Ao99/e-commerce-django}{Github
-repo}
-\item Command line or web app: web app
-\item Quick Analyze checked: Yes/No
-\item Time Consumption: 121s/117s
-\item Errors/problems: most data are missing
-\end{itemize}
-
-Test 5/6 on environment 3: 
-\begin{itemize}
-\item HubListener
-\href{https://github.com/pjmc-oliveira/HubListener}{Github repo}
-\item Command line or web app: web app
-\item Quick Analyze checked: Yes/No
-\item Time Consumption: 49s/131s
-\item Errors/problems: none
-\end{itemize}
+Test results: HubListener works well on the repo of itself, but it did not work well on some other repos.
 

### 2020-03-31 StateOfPractice/Methodology/Methodology.tex
update tool tests

@@
 
-\subsection{Tool Test - HubListener}
-\ad{This section is currently a note of unorganized contents. Most parts will be
-removed or relocated.}
+\subsection{Tool Tests}
+\ad{This section is currently a note of unorganized contents. Most parts will beremoved or relocated.}
 
+\ad{This citation needs to be deleted later. It's here because my compiler
+doesn't work with 0 citations}
+\cite{Emms2019}
+
+Most tests were done targeting to the repo of 3D Slicer
+\href{https://github.com/tomgi/git_stats}{GitHub repo}
+
+\subsubsection{git-stats}
+\href{https://github.com/tomgi/git_stats}{GitHub repo}
+
+Test results:
+\href{http://git-stats-slicer.ao9.io/}{http://git-stats-slicer.ao9.io/} the
+results are output as webpages, so I hosted for you to check. Data can be
+downloaded as spreadsheets.
+
+\subsubsection{git-of-theseus}
+\href{https://github.com/erikbern/git-of-theseus}{GitHub repo}
+
+Test results: It took about 100 minutes for one repo on a 8 core 16G ram Linux
+machine. It only outputs graphs.
+
+\subsubsection{hercules}
+\href{https://github.com/src-d/hercules}{GitHub repo}
+
+Test results: this one seems to be promising, but the installation is
+complicated with various errors.
+
+\subsubsection{git-repo-analysis}
+\href{https://github.com/larsxschneider/git-repo-analysis}{GitHub repo}
+
+\subsubsection{HubListener}
 \href{https://github.com/pjmc-oliveira/HubListener}{GitHub repo}
@@
 
-Test results: HubListener works well on the repo of itself, but it did not work well on some other repos.
+Test results: HubListener works well on the repo of itself, but it did not work
+well on some other repos.
 
-\ad{This citation needs to be deleted later. It's here because my compiler
-doesn't work with 0 citations}
-\cite{Emms2019}
+\subsubsection{gitinspector}
+\href{https://github.com/ejwa/gitinspector}{GitHub repo}
+
+Test results: it doesn't work well. Instead of creating output results, it
+prints the results directly in the console.
 

### 2020-04-01 StateOfPractice/Peter-Notes/Commonality-Analysis-LB-Systems.tex
commonality analysis for LB

@@
 	\begin{abstract}
-		This report presents a commonality analysis for Lattice Boltzmann systems.
+		This report presents a commonality analysis for Lattice Boltzmann systems using a commonality analysis template based on \cite{SmithAndChen2004}.
 The document reviews both the methodology of commonality analysis and the details of Lattice Boltzmann systems.
 The commonality analysis itself consists of the following:
-		i) terminology and definitions; ii) commonalities, or features that are common to all potential family members; iii) variabilities, or features and characteristics that may vary among family members; and, iv) parameters of variation, or the potential values that can be assigned to the variabilities.
-		The documentation of the above items for Lattice Boltzmann systems is clarified by decomposing each item into subsections on Lattice Boltzmann methods, input, output, nonfunctional
-		requirements, and, where appropriate, system constraints.\\
+		\textit{i}) terminology and definitions; \textit{ii}) commonalities, or features that are common to all potential family members; \textit{iii}) variabilities, or features and characteristics that may vary among family members; and, \textit{iv}) parameters of variation, or the potential values that can be assigned to the variabilities.
+		The documentation of the above items for Lattice Boltzmann systems is clarified by decomposing each item into subsections on Lattice Boltzmann methods, input, output, nonfunctional requirements, and, where appropriate, system constraints.\\
 
@@
 \section{Introduction}
-Mesh generating systems are well suited to development as a program family because they
-fit Parnas’s definition of a program family: “a set of programs whose common properties
-are so extensive that it is advantageous to study the common properties of the programs
-before analyzing individual members” (Parnas, 1976). Developing mesh generating sys-
-tems as a program family is advantageous because mesh generators share many common
-features, or commonalities. Furthermore, when there are differences between systems, the
-variabilities between them can be systematically considered. The purpose of this document
-is to record these commonalities and variabilities and show the relationships between them,
-and thus facilitate the specification and development of mesh generator program family
-members. This document will be valuable in all future phases of system development and
-analysis. For instance, the requirements documentation for any mesh generator will use
-the commonality analysis, since the requirements should refine the commonalities, which
-are shared requirements of all mesh generators. Moreover, the design of any future system
-will use this documentation to facilitate consideration of the variabilities, so that likely
-changes can be made to the system with a minimal amount of work.
-The scope of the commonality analysis presented here can be considered both from the
-point of view of mesh generating systems and from the point of view of software engineering
-methodologies. From the mesh generator perspective, the starting point for the current
-document is the commonality analysis conducted by Chen (2003). However, the scope of
-the current document is broader than that of Chen (2003), which was restricted to two-
-dimensional meshes and did not consider non-conformal, hybrid or mixed meshes. The
-current commonality analysis is intended to cover all mesh generators that are targeted
-at finite element applications. Meshes for cartography or other uses are not explicitly
-considered, although much of the information in this document does overlap with other
-mesh uses. With respect to software engineering methodologies, the scope of the current
-report is restricted to informal methods, with the intention that the informal requirements
-will form a starting point for later development and refinement by formal methods.
-The first section below provides an overview of the program family of mesh generators,
-by reviewing what is involved in a commonality analysis and the basics of mesh generating
-systems. After this, the basic terminology and definitions necessary for understanding the
-remainder of the document are provided. The definitions include terms used in describing
-a commonality analysis and terms that are used in defining the characteristics and prop-
-erties of meshes. The next three sections consist of the lists of commonalities, variabilities
-and parameters of variation, respectively. These three sections form the heart of the docu-
-mentation and include an extensive set of cross-references to demonstrate the relationships
-between the different items. The final section provides information on unresolved issues.
+Lattice Boltzmann systems are well suited to development as a program family because they fit Parnas’s definition of a program family: “a set of programs whose common properties are so extensive that it is advantageous to study the common properties of the programs before analyzing individual members” \citep{parnas1976design}.
+Developing such systems as a program family is advantageous because Lattice Boltzmann systems share many common features, or commonalities.
+Furthermore, when there are differences between systems, the variabilities between them can be systematically considered.
+The purpose of this document is to record these commonalities and variabilities and show the relationships between them, and thus facilitate the specification and development of Lattice Boltzmann program family members.
+This document will be valuable in all future phases of system development and analysis.
+For instance, the requirements documentation for any Lattice Boltzmann systems will use the commonality analysis, since the requirements should refine the commonalities, which are shared requirements of all such systems.
+Moreover, the design of any future system will use this documentation to facilitate consideration of the variabilities, so that likely changes can be made to the system with a minimal amount of work.
+
+The scope of the commonality analysis presented here can be considered both from the point of view of Lattice Boltzmann systems and from the point of view of software engineering methodologies.
+From the Lattice Boltzmann system perspective, the starting point for the current document is a commonality analysis.
+With respect to software engineering methodologies, the scope of the current report is restricted to informal methods, with the intention that the informal requirements will form a starting point for later development and refinement by formal methods.
+
+The first section below provides an overview of the program family of Lattice Boltzmann systems, by reviewing what is involved in a commonality analysis and the basics of Lattice Boltzmann systems.
+After this, the basic terminology and definitions necessary for understanding the remainder of the document are provided.
+The definitions include terms used in describing a commonality analysis and terms that are used in defining the characteristics and properties of Lattice Boltzmann methods.
+The next three sections consist of the lists of commonalities, variabilities and parameters of variation, respectively.
+These three sections form the heart of the documentation and include an extensive set of cross-references to demonstrate the relationships between the different items.
+The final section provides information on unresolved issues.
 
@@
 \section{Overview}
-This section provides both an overview of the process of commonality analysis and of mesh
-generation. The subsection on commonality analysis briefly introduces this topic, along
-with references that can be searched for further information. The subsection on mesh
-generators outlines the following: i) the basics of mesh generation, ii) the scope of the
-current analysis, iii) the overall philosophy that has been adopted for the commonality
-analysis, and iv) the arguments in favour of the development of mesh generators as a
-program family.
+This section provides both an overview of the process of commonality analysis and of Lattice Boltzmann methods.
+The subsection on commonality analysis briefly introduces this topic, along with references that can be searched for further information.
+The subsection on Lattice Boltzmann methods outlines the basics of Lattice Boltzmann methods and of the scope of the commonality analysis.
 
 \subsection{Commonality Analysis}
-In some situations it is advantageous to develop a collection of related software products
-as a program family. The idea is that if the software products are similar enough, then
-it should be possible to predict what the products have in common, what differs between
-them and then reuse the common aspects and thus support rapid development of the
-family members. The idea of program families was introduced by Dijkstra (1972) and later
-investigated by Parnas (Parnas, 1976, 1979). More recently, Weiss (Weiss, 1997, 1998;
-Ardis and Weiss, 1997) has considered the concept of a program family in the context of
-what he terms Family oriented Abstraction, Specification and Translation (FAST) (Cuka
-and Weiss, 1997; Weiss and Lai, 1999).
-In the approach advocated by Weiss, the first step is a commonality analysis. This
-analysis consists of systematically identifying and documenting the commonalities that all
-program family members share, the variabilities between family members and the termi-
-nology used in describing the family. A commonality analysis provides a systematic way of
-gaining confidence that a family is worth building and of deciding what the family members
-will be. A commonality analysis document provides the following benefits (Weiss, 1997,
-1998):
-1. A starting point for the design of a domain specific languages (DSL): Once a DSL,
-or application modelling language, is developed the program family members can be
-rapidly generated by specifying a given family member using the language.
-2. A basis for a common design for all family members: When the software engineers
-come to designing the individual family members, they can take advantage of the
-commonalities to reuse code. Moreover, the variabilities can be considered in the
-design so that they can be easily accommodated. One approach to the design may
-be to decompose the system into components that can each be customized by speci-
-fication of values for its various parameters, where the parameters correspond to the
-parameters of variation identified in the commonality analysis document.
-3. A historical reference: This document records the important issues concerning the
-scope and the nature of the family (as well as some unsolved issues) to facilitate the
-involvement of the participants in maintaining and evolving the family.
-CAS-04-10-SS
-4
-4. A basis for reengineering a domain: Existing projects may not have been developed
-using software engineering methodologies. The projects can be systematically reor-
-ganized and redesigned with the aid of a commonality analysis to unify the existing
-products.
-5. A basic training reference for new software developers: This document provides the
-necessary basic information for a new team-member to understand the family.
-The next section will show that the above uses of the commonality analysis document
-will be beneficial for the development of a program family of mesh generators. The com-
-monality analysis will benefit all subsequent stages in the software development process.
-For instance, as mentioned in the introduction, the commonalities will act as requirements
-that will be the starting point for writing a software requirements specification. The com-
-monalities will be refined into specific requirements by fixing the value of their associated
-variabilities. The change in the values of the variabilities then corresponds to the change
-from one program family member to another.
-As commonalities and variabilities are requirements, they should express “What” func-
-tionalities and qualities the system should have, and not mention “How” these requirements
-are to be accomplished. That is, the commonalities and variabilities should not involve
-design decisions. The design decisions will be made after the requirements for a family
-member have been specified. The one exception to this is system constraints, which are
-requirements that explicitly make design decisions.
-Besides the “What” versus “How” test, there are other tests that can be used to review
-commonalities and variabilities, as proposed by Weiss (1997). One such test is the “what
-is ruled out” test. This test determines if a commonality or variability actually makes a
-decision because if no alternatives are ruled out then no decision has really been made.
-Another test is the “negation” test. If the negation of a decision represents a position
-that someone might argue for, then the original decision is likely to be meaningful. For
-instance, the statement that “the software should be reliable” has a negation that no one
-would likely argue for and thus the statement does not represent a good characterization
-of a goal for the system.
-In Weiss (1997) the stages of a commonality analysis are described in a systematic way.
-The stages include the following: prepare, plan, analyze, quantify and review. The stages
-are completed through the aid of a moderator and a series of meeting and preliminary
-documents and documentation reviews. Although the systematic approach advocated by
-Weiss has its advantages, for the case of writing a commonality analysis document for mesh
-generating systems it was decided that a less structured approach is feasible. Mesh genera-
-tors are simpler than other software systems in the sense that they have fewer interactions
-with the environment. Also, the theory of mesh generation has a solid mathematical ba-
-sis that can be used to remove some of the ambiguity that Weiss’s approach is aimed
-at reducing. Therefore, the approach adopted here is to revise the original commonality
-analysis document produced by Chen (2003) and then make the new document available
-to others for review. The new document will be maintained in a concurrent versioning
-CAS-04-10-SS
-5
-system repository so that multiple authors can work on it, and more importantly, so that
-the documentation’s revision history can be tracked and the documentation can be rolled
-back to an earlier version if necessary.
+In some situations it is advantageous to develop a collection of related software products as a program family.
+The idea is that if the software products are similar enough, then it should be possible to predict what the products have in common, what differs between them and then reuse the common aspects and thus support rapid development of the family members.
+The idea of program families was introduced by \cite{dijkstra1972} and later investigated by \cite{parnas1976design,parnas1979designing}.
+More recently, \cite{weiss1997defining, weiss1998commonality, ardis1997defining} has considered the concept of a program family in the context of what he terms Family oriented Abstraction, Specification and Translation (FAST) \citep{cuka1997specifying,weiss1999software}.
+
+In the approach advocated by Weiss, the first step is a commonality analysis.
+This analysis consists of systematically identifying and documenting the commonalities that all program family members share, the variabilities between family members and the terminology used in describing the family.
+A commonality analysis provides a systematic way of gaining confidence that a family is worth building and of deciding what the family members will be.
+A commonality analysis document provides the following benefits \cite{weiss1997defining, weiss1998commonality}:
+
+\begin{enumerate}
+	\item A starting point for the design of a domain specific languages (DSL): Once a DSL, or application modelling language, is developed the program family members can be rapidly generated by specifying a given family member using the language.
+	\item A basis for a common design for all family members: When the software engineers come to designing the individual family members, they can take advantage of the commonalities to reuse code.
+Moreover, the variabilities can be considered in the design so that they can be easily accommodated.
+One approach to the design may be to decompose the system into components that can each be customized by specification of values for its various parameters, where the parameters correspond to the parameters of variation identified in the commonality analysis document.
+	\item A historical reference: This document records the important issues concerning the scope and the nature of the family (as well as some unsolved issues) to facilitate the involvement of the participants in maintaining and evolving the family.
+	\item A basis for re-engineering a domain: Existing projects may not have been developed using software engineering methodologies.
+The projects can be systematically reorganized and redesigned with the aid of a commonality analysis to unify the existing products.
+	\item A basic training reference for new software developers: This document provides the necessary basic information for a new team-member to understand the family. 
+\end{enumerate}
+
+The next section will show that the above uses of the commonality analysis document will be beneficial for the development of a family of Lattice Boltzmann programs.
+The commonality analysis will benefit all subsequent stages in the software development process.
+For instance, as mentioned in the introduction, the commonalities will act as requirements that will be the starting point for writing a software requirements specification.
+The commonalities will be refined into specific requirements by fixing the value of their associated variabilities.
+The change in the values of the variabilities then corresponds to the change from one program family member to another.
+
+As commonalities and variabilities are requirements, they should express “What” functionalities and qualities the system should have, and not mention “How” these requirements are to be accomplished.
+That is, the commonalities and variabilities should not involve design decisions.
+The design decisions will be made after the requirements for a family member have been specified.
+The one exception to this is system constraints, which are requirements that explicitly make design decisions.
+
+Besides the “What” versus “How” test, there are other tests that can be used to review commonalities and variabilities, as proposed by \cite{weiss1997defining}.
+One such test is the “what is ruled out” test.
+This test determines if a commonality or variability actually makes a decision because if no alternatives are ruled out then no decision has really been made.
+Another test is the “negation” test.
+If the negation of a decision represents a position that someone might argue for, then the original decision is likely to be meaningful.
+For instance, the statement that “the software should be reliable” has a negation that no one would likely argue for and thus the statement does not represent a good characterization of a goal for the system.
+
+In \cite{weiss1997defining} the stages of a commonality analysis are described in a systematic way.
+The stages include the following: prepare, plan, analyze, quantify and review.
+The stages are completed through the aid of a moderator and a series of meeting and preliminary documents and documentation reviews.
+The approach adopted here is to revise the original commonality analysis document produced by \cite{chen2003} and then make the new document available to others for review.
+The new document will be maintained in a concurrent versioning system repository so that multiple authors can work on it, and more importantly, so that the documentation’s revision history can be tracked and the documentation can be rolled back to an earlier version if necessary.
 
 \subsection{Lattice Boltzmann Systems}
+Lattice Boltzmann Methods (LBM) are a family of fluid dynamics algorithms for simulating single-phase and multiphase fluid flows, often incorporating additional physical complexities \citep{chen1998lattice}.
+They consider the behaviours of a collection of particles as a single unit at the mesoscopic scale.
+These methods predict the positional probability of a collection of particles moving through a lattice structure.
+Off the shelf (OTS) Lattice Boltzmann Solvers (LBS) allow for a range of fluid and physical model input parameters, computational parameters, and output parameters.
+
+As LBS model fluid dynamics within a boundary using a predefined lattice structure, the methods rely on a two step calculation process.
+The first process is streaming, where the particles move along the lattice via links.
+The second process is collision, where energy and momentum is transferred among particles that collide \citep{bao2011lattice}.
+There are many standardized lattice models - individual solvers within the family might only use a subset of them.
+The LBM uses the initial parameters of the fluid to find the probability of where along the lattice linkages a group of particles are most likely to travel.
+It then moves the particles into the next node, and transfers the energy and momentum if a collision occurs.
+Then the process repeats for the duration of the modeling instance.
+
 \newpage
 \section{Terminology and Definitions}
-This section is divided into two subsections. The first discusses the terminology that comes
-from the software engineering field, while the second presents the definitions used in mesh
-generation. Common acronyms are also listed in this section. The lists are not intended
-to be read sequentially, but rather to be consulted for reference purposes; therefore, the
-terms are ordered alphabetically, with the consequence that some terms that appear early
-in the list depend on the definitions of later terms.
+This section is divided into two subsections.
+The first discusses the terminology that comes from the software engineering field, while the second presents the definitions used in Lattice Boltzmann solvers.
+Common acronyms are also listed in this section.
+The lists are not intended to be read sequentially, but rather to be consulted for reference purposes; therefore, the terms are ordered alphabetically, with the consequence that some terms that appear early in the list depend on the definitions of later terms.
 
 \subsection{Software Engineering Related Definitions and Acronyms}
-Commonality: A requirement or goal common to all family members.
-Goal: “Goals capture, at different levels of abstraction, the various objectives the system
-under consideration should achieve.” van Lamsweerde (2001)
-Program Family: A set of programs that are analyzed and designed together starting
-from the initial stages of the software development life-cycle.
-Requirements: A software requirement is: i) a condition or capability needed by a user
-to solve a problem or achieve an objective; ii) a condition or capability that must be
-met or possessed by a system or system component to satisfy a contract, standard,
-specification, or other formally imposed document; or, iii) a documented represen-
-CAS-04-10-SS
-10
-tation of a condition or capability as in the above two definitions.
-Dorfman, 2000)
-Variability: A requirement or goal that varies between family members.
+\noindent\textbf{Commonality}: A requirement or goal common to all family members.\\
+
+\noindent\textbf{Goal}: “Goals capture, at different levels of abstraction, the various objectives the system under consideration should achieve.” \cite{van2001goal}\\
+
+\noindent\textbf{Program Family}: A set of programs that are analyzed and designed together starting from the initial stages of the software development life-cycle.\\
+
+\noindent\textbf{Requirements}: A software requirement is: \textit{i}) a condition or capability needed by a user to solve a problem or achieve an objective; \textit{ii}) a condition or capability that must be met or possessed by a system or system component to satisfy a contract, standard, specification, or other formally imposed document; or, \textit{iii}) a documented representation of a condition or capability as in the above two definitions.
+\cite{thayer2000ieee}\\
+
+\noindent\textbf{Variability}: A requirement or goal that varies between family members.
+
 \subsection{Lattice Boltzmann Related Definitions and Acronyms}
+
+\noindent\textbf{1D} 1-Dimensional\\
+
+\noindent\textbf{2D}: 2-Dimensional\\
+
+\noindent\textbf{3D}: 3-Dimensional\\
+
+\noindent\textbf{LBM}: Lattice Boltzmann Methods\\
+
+\noindent\textbf{LBS}: Lattice Boltzmann Solvers\\
+
+\noindent\textbf{Velocity Directions}: The number of links connecting to each lattice node in the chosen model from neighbouring nodes.
+All nodes in a chosen lattice model will have the same number of links.
+A single link will connect between two adjacent nodes.
+
 \newpage
-\section{Commonalities}
+\section{Commonalities}\label{comm_sec}
+
+This section lists all the common features among all the potential family members.
+The commonalities are organized using the following abstraction of the system, which can be used to describe all Lattice Boltzmann systems: input information, then generate the simulation and finally output the results.
+Section \ref{comm_lbm} describes the commonalities for the simulation step.
+Section \ref{comm_in} highlights the input information that is required for all Lattice Boltzmann systems.
+The next section, Section \ref{comm_out}, shows the common features for the output of Lattice Boltzmann systems, such as the requirement that mesh information be written to files.
+(Although the output information could simply be written to the computer’s memory, in all practical applications it is desirable to have a persistent record of the output that was created.) The final section covers qualities of the system that cannot be classified as input, simulation generation or output.
+These commonalities are termed nonfunctional requirements of the system.
+For instance, all systems will have the goal that the response time to a user’s request is small enough to allow the user to focus on his/her problem and to maintain his/her train of thought, without being distracted by excessive waiting time.
+The commonality in this case is refined by a later variability because the specific requirement on the response time will depend on the intended usage of the mesh generating system.
+
+Each commonality below uses the same structure.
+All of the commonalities are assigned a unique item number, which takes the form of a natural number with the prefix ``C''.
+Following this, a description of the commonality is provided along with a list of related variabilities, which are given as hyperlinks that allow navigation of the document to the text describing the variability.
+Finally, each commonality ends with a summary of the history, including the date of creation and any dates of modification, along with a brief description of the modification.
+If necessary, a previous version of the document can be obtained by using the concurrent versioning system where the files are stored.
+
+\subsection{Lattice Boltzmann Method Solvers}\label{comm_lbm}
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
+		\hline
+		\bf Item Number& C\refstepcounter{comnum}\thecomnum{}\label{lattice} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
+		\hline
+		\bf Description & A lattice discretizes a computational domain into a finite number of points. All LBS discretize the computational domain using a regular, evenly spaced grid.\\
+		\hline
+		\bf Related Variability & \vref{numveldir} \vref{varcolop}\\
+		\hline
+		\bf History & Created April 1, 2020 \\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+~\newline\\
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
+		\hline
+		\bf Item Number& C\refstepcounter{comnum}\thecomnum\label{collisionoperator} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
+		\hline
+		\bf Description & All Lattice Boltzmann versions use a collision operator which concerns collisions between particles.\\
+		\hline
+		\bf Related Variability & \vref{numveldir} \vref{varcolop}\\
+		\hline
+		\bf History & Created April 1, 2020\\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+~\newline\\
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
+		\hline
+		\bf Item Number& C\refstepcounter{comnum}\thecomnum\label{probabilitydensityfunc} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
+		\hline
+		\bf Description & All Lattice Boltzmann versions user a probability density function to give the probability that fluid has moved into a specific domain.\\
+		\hline
+		\bf Related Variability & \vref{dimensions}\\
+		\hline
+		\bf History & Created April 1, 2020\\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+~\newline\\
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
+		\hline
+		\bf Item Number& C\refstepcounter{comnum}\thecomnum\label{discretizevelocity} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
+		\hline
+		\bf Description & All Lattice Boltzmann versions discretize velocity into a finite number of directions.\\
+		\hline
+		\bf Related Variability & \vref{numveldir}\\
+		\hline
+		\bf History & Created April 1, 2020\\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+~\newline\\
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
+		\hline
+		\bf Item Number& C\refstepcounter{comnum}\thecomnum\label{relaxationupdate} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
+		\hline
+		\bf Description & All Lattice Boltzmann versions update the velocity distribution during each LBM iteration.\\
+		\hline
+		\bf Related Variability & \vref{varrelaxops}\\
+		\hline
+		\bf History & Created April 1, 2020\\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+~\newline\\
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
+		\hline
+		\bf Item Number& C\refstepcounter{comnum}\thecomnum\label{equildistfunc} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
+		\hline
+		\bf Description & Every Lattice Boltzmann version uses an equilibrium distribution function to capture the probability distribution of particles.\\
+		\hline
+		\bf Related Variability & \vref{varedf} \vref{edfstoreage}\\
+		\hline
+		\bf History & Created April 1, 2020\\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+~\newline\\
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
+		\hline
+		\bf Item Number& C\refstepcounter{comnum}\thecomnum\label{Boltzmanntransequ} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
+		\hline
+		\bf Description & Every Lattice Boltzmann version uses a Boltzmann transport equation to describe the statistical behaviour of a system that does not have collisions.\\
+		\hline
+		\bf Related Variability & \vref{vartraneq}\\
+		\hline
+		\bf History & Created April 1, 2020\\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+\subsection{Input}\label{comm_in}
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
+		\hline
+		\bf Item Number& C\refstepcounter{comnum}\thecomnum\label{fluidandboundaryinfo} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
+		\hline
+		\bf Description & The LBS require fluid, model, and boundary information for the problem.\\
+		\hline
+		\bf Related Variability & \vref{interface} \vref{numberoffluids} \vref{typefluidparm}\\
+		\hline
+		\bf History & Created April 1, 2020\\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+\subsection{Output}\label{comm_out}
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
+		\hline
+		\bf Item Number& C\refstepcounter{comnum}\thecomnum\label{predictionstomemory} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
+		\hline
+		\bf Description & LBS write fluid parameter predictions to memory.\\
+		\hline
+		\bf Related Variability & \vref{varvisual} \vref{formatoutput} \\
+		\hline
+		\bf History & Created April 1, 2020\\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+\subsection{Nonfunctional Requirements}\label{comm_nfr}
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
+		\hline
+		\bf Item Number& C\refstepcounter{comnum}\thecomnum\label{precision} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
+		\hline
+		\bf Description & LBS provides the precision required for the particular problems it is intended to solve.\\
+		\hline
+		\bf Related Variability & \vref{varprecise}\\
+		\hline
+		\bf History & Created April 1, 2020\\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+~\newline\\
+~\newline
+
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
+		\hline
+		\bf Item Number& C\refstepcounter{comnum}\thecomnum\label{accuracy} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
+		\hline
+		\bf Description & LBS provides the accuracy required for the particular problems it is intended to solve.\\
+		\hline
+		\bf Related Variability & \vref{varaccuracy}\\
+		\hline
+		\bf History & Created April 1, 2020\\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+~\newline\\
+~\newline
+
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
+		\hline
+		\bf Item Number& C\refstepcounter{comnum}\thecomnum\label{performance} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
+		\hline
+		\bf Description & The response time is small enough to allow the user to focus on their problem without being distracted by excessive waiting times.\\
+		\hline
+		\bf Related Variability & \vref{varresponsetime}\\
+		\hline
+		\bf History & Created April 1, 2020\\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+~\newline\\
+~\newline
+
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
+		\hline
+		\bf Item Number& C\refstepcounter{comnum}\thecomnum\label{portability} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
+		\hline
+		\bf Description & LBS will be as portable to other operating systems as required by the users.\\
+		\hline
+		\bf Related Variability & \vref{varportability}\\
+		\hline
+		\bf History & Created April 1, 2020\\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+~\newline\\
+~\newline
+
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
+		\hline
+		\bf Item Number& C\refstepcounter{comnum}\thecomnum\label{usability} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
+		\hline
+		\bf Description & LBS shall be efficiently easy to use.\\
+		\hline
+		\bf Related Variability & \vref{varusability}\\
+		\hline
+		\bf History & Created April 1, 2020\\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+\newpage
+\section{Variabilities}\label{var_sec}
+This section provides a list of characteristics that may vary among family members.
+As in Section \ref{comm_sec}, the first three subsections on variabilities are organized into the following sublists: Mesh Generation, Input and Output.
+The final two subsections list variabilities that can be characterized as system constraints and as nonfunctional requirements.
+
+As for the commonalities, each variability is labelled with a unique item number.
+In this case the numbers are prepended with the letter ``V''.
+The other four headings provided for each variability are: Description, Related Commonality, Related Parameter and History.
+The related commonalities and parameters are given as a set of identifiers that respectively refer back to the previous section on commonalities or refer forward to the next section on parameters of variation.
+
 \subsection{Lattice Boltzmann Method Solvers}
@@
 	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
-		\hline
-		\bf Item Number& C\refstepcounter{comnum}\thecomnum \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
-		\hline
-		\bf Description & \\
-		\hline
-		\bf Related Variability & \vref{}\\
-		\hline
-		\bf History & \\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-~\newline\\
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
-		\hline
-		\bf Item Number& C\refstepcounter{comnum}\thecomnum \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
-		\hline
-		\bf Description & \\
-		\hline
-		\bf Related Variability & \vref{}\\
-		\hline
-		\bf History & \\
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
+		\hline
+		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{varrelaxops} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
+		\hline
+		\bf Description & Different relaxation times are used to update the velocity distribution.\\
+		\hline
+		\bf Related Commonality & \cref{relaxationupdate}\\
+		\hline
+		\bf Related Parameter & \pref{parrelax}\\
+		\hline
+		\bf History & Created April 1, 2020\\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+~\newline\\
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
+		\hline
+		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{parallel} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
+		\hline
+		\bf Description & LBS may use a framework for parallel processing of the model.\\
+		\hline
+		\bf Related Commonality & None\\
+		\hline
+		\bf Related Parameter & \pref{parparallel}\\
+		\hline
+		\bf History & Created April 1, 2020\\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+~\newline\\
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
+		\hline
+		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{varedf} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
+		\hline
+		\bf Description & Different versions of an equilibrium distribution function can capture the probability distribution of particles.\\
+		\hline
+		\bf Related Commonality & \cref{equildistfunc}\\
+		\hline
+		\bf Related Parameter & \pref{paredf}\\
+		\hline
+		\bf History & Created April 1, 2020\\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+~\newline\\
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
+		\hline
+		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{edfstorage} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
+		\hline
+		\bf Description & Storage patters for distribution function can vary.\\
+		\hline
+		\bf Related Commonality & \cref{equildistfunc}\\
+		\hline
+		\bf Related Parameter & \pref{parstorage}\\
+		\hline
+		\bf History & Created April 1, 2020\\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+~\newline\\
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
+		\hline
+		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{dimensions} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
+		\hline
+		\bf Description & The number of dimensions in the lattice of the model can vary.\\
+		\hline
+		\bf Related Commonality & \cref{probabilitydensityfunc}\\
+		\hline
+		\bf Related Parameter & \pref{pardimensions}\\
+		\hline
+		\bf History & Created April 1, 2020\\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+~\newline\\
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
+		\hline
+		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{numveldir} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
+		\hline 
+		\bf Description & The number of velocity directions in the lattice of the model can vary.\\
+		\hline
+		\bf Related Commonality & \cref{lattice} \cref{collisionoperator} \cref{discretizevelocity}\\
+		\hline
+		\bf Related Parameter & \pref{parveldir}\\
+		\hline
+		\bf History & Created April 1, 2020\\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+~\newline\\
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
+		\hline
+		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{varcolop} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
+		\hline
+		\bf Description & Various collision operators can be used.\\
+		\hline
+		\bf Related Commonality & \cref{lattice} \cref{collisionoperator}\\
+		\hline
+		\bf Related Parameter & \pref{parcollisop}\\
+		\hline
+		\bf History & Created April 1, 2020\\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+~\newline\\
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
+		\hline
+		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{vartraneq} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
+		\hline
+		\bf Description & Various Transport Equations can be used to describe the statistical behaviour of the system\\
+		\hline
+		\bf Related Commonality & \cref{Boltzmanntransequ}\\
+		\hline
+		\bf Related Parameter & \pref{partransequ}\\
+		\hline
+		\bf History & Created April 1, 2020\\
 		\hline
@@
 	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
-		\hline
-		\bf Item Number& C\refstepcounter{comnum}\thecomnum \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
-		\hline
-		\bf Description & \\
-		\hline
-		\bf Related Variability & \vref{}\\
-		\hline
-		\bf History & \\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-~\newline\\
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
-		\hline
-		\bf Item Number& C\refstepcounter{comnum}\thecomnum \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
-		\hline
-		\bf Description & \\
-		\hline
-		\bf Related Variability & \vref{}\\
-		\hline
-		\bf History & \\
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
+		\hline
+		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{interface} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
+		\hline
+		\bf Description & The input interface can vary between LBS.\\
+		\hline
+		\bf Related Commonality & \cref{fluidandboundaryinfo}\\
+		\hline
+		\bf Related Parameter & \pref{parinterface}\\
+		\hline
+		\bf History & Created April 1, 2020\\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+~\newline\\
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
+		\hline
+		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{numberoffluids} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
+		\hline
+		\bf Description & The number of fluids allowed in the LBS.\\
+		\hline
+		\bf Related Commonality & \cref{fluidandboundaryinfo}\\
+		\hline
+		\bf Related Parameter & \pref{parfluids}\\
+		\hline
+		\bf History & Created April 1, 2020\\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+~\newline\\
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
+		\hline
+		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{typefluidparm} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
+		\hline
+		\bf Description & The type of fluid parameters.\\
+		\hline
+		\bf Related Commonality & \cref{fluidandboundaryinfo}\\
+		\hline
+		\bf Related Parameter & \pref{partypeparm}\\
+		\hline
+		\bf History & Created April 1, 2020\\
 		\hline
@@
 	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
-		\hline
-		\bf Item Number& C\refstepcounter{comnum}\thecomnum \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
-		\hline
-		\bf Description & \\
-		\hline
-		\bf Related Variability & \vref{}\\
-		\hline
-		\bf History & \\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-~\newline\\
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
-		\hline
-		\bf Item Number& C\refstepcounter{comnum}\thecomnum \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
-		\hline
-		\bf Description & \\
-		\hline
-		\bf Related Variability & \vref{}\\
-		\hline
-		\bf History & \\
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
+		\hline
+		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{varvisual} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
+		\hline
+		\bf Description & Visual presentation of the prediction.\\
+		\hline
+		\bf Related Commonality & \cref{predictionstomemory}\\
+		\hline
+		\bf Related Parameter & \pref{parvarvisual}\\
+		\hline
+		\bf History & Created April 1, 2020\\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+~\newline\\
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
+		\hline
+		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{formatoutput} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
+		\hline
+		\bf Description & Format of prediction information.\\
+		\hline
+		\bf Related Commonality & \cref{predictionstomemory}\\
+		\hline
+		\bf Related Parameter & \pref{parformatoutput}\\
+		\hline
+		\bf History & Created April 1, 2020\\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+\subsection{System Constraints}
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
+		\hline
+		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{processing} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
+		\hline
+		\bf Description & Hardware which processes the calculations\\
+		\hline
+		\bf Related Commonality & None \\
+		\hline
+		\bf Related Parameter & \pref{parprocessing}\\
+		\hline
+		\bf History & Created April 1, 2020\\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+~\newline\\
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
+		\hline
+		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{operatingsystem} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
+		\hline
+		\bf Description & Operating systems on which LBS runs.\\
+		\hline
+		\bf Related Commonality & None \\
+		\hline
+		\bf Related Parameter & \pref{paropsystem}\\
+		\hline
+		\bf History & Created April 1, 2020\\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+~\newline\\
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
+		\hline
+		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{storage} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
+		\hline
+		\bf Description & Amount of storage and memory needed for the LBS.\\
+		\hline
+		\bf Related Commonality & None \\
+		\hline
+		\bf Related Parameter & \pref{parstorageconst}\\
+		\hline
+		\bf History & Created April 1, 2020\\
 		\hline
@@
 	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
-		\hline
-		\bf Item Number& C\refstepcounter{comnum}\thecomnum \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
-		\hline
-		\bf Description & \\
-		\hline
-		\bf Related Variability & \vref{}\\
-		\hline
-		\bf History & \\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-~\newline\\
-~\newline
-
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
-		\hline
-		\bf Item Number& C\refstepcounter{comnum}\thecomnum \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
-		\hline
-		\bf Description & \\
-		\hline
-		\bf Related Variability & \vref{}\\
-		\hline
-		\bf History & \\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-\newpage
-\section{Variabilities}
-\subsection{Lattice Boltzmann Method Solvers}
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
-		\hline
-		\bf Item Number& V\refstepcounter{varnum}\thevarnum \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & \\
-		\hline
-		\bf Related Commonality & \cref{}\\
-		\hline
-		\bf Related Parameter & \pref{}\\
-		\hline
-		\bf History & \\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-~\newline\\
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
-		\hline
-		\bf Item Number& V\refstepcounter{varnum}\thevarnum \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & \\
-		\hline
-		\bf Related Commonality & \cref{}\\
-		\hline
-		\bf Related Parameter & \pref{}\\
-		\hline
-		\bf History & \\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-\subsection{Input}
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
-		\hline
-		\bf Item Number& V\refstepcounter{varnum}\thevarnum \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & \\
-		\hline
-		\bf Related Commonality & \cref{}\\
-		\hline
-		\bf Related Parameter & \pref{}\\
-		\hline
-		\bf History & \\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-~\newline\\
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
-		\hline
-		\bf Item Number& V\refstepcounter{varnum}\thevarnum \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & \\
-		\hline
-		\bf Related Commonality & \cref{}\\
-		\hline
-		\bf Related Parameter & \pref{}\\
-		\hline
-		\bf History & \\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-\subsection{Output}
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
-		\hline
-		\bf Item Number& V\refstepcounter{varnum}\thevarnum \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & \\
-		\hline
-		\bf Related Commonality & \cref{}\\
-		\hline
-		\bf Related Parameter & \pref{}\\
-		\hline
-		\bf History & \\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-~\newline\\
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
-		\hline
-		\bf Item Number& V\refstepcounter{varnum}\thevarnum \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & \\
-		\hline
-		\bf Related Commonality & \cref{}\\
-		\hline
-		\bf Related Parameter & \pref{}\\
-		\hline
-		\bf History & \\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-\subsection{System Constraints}
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
-		\hline
-		\bf Item Number& V\refstepcounter{varnum}\thevarnum \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & \\
-		\hline
-		\bf Related Commonality & \cref{}\\
-		\hline
-		\bf Related Parameter & \pref{}\\
-		\hline
-		\bf History & \\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-~\newline\\
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
-		\hline
-		\bf Item Number& V\refstepcounter{varnum}\thevarnum \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & \\
-		\hline
-		\bf Related Commonality & \cref{}\\
-		\hline
-		\bf Related Parameter & \pref{}\\
-		\hline
-		\bf History & \\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-\subsection{Nonfunctional Requirements}
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
-		\hline
-		\bf Item Number& V\refstepcounter{varnum}\thevarnum \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & \\
-		\hline
-		\bf Related Commonality & \cref{}\\
-		\hline
-		\bf Related Parameter & \pref{}\\
-		\hline
-		\bf History & \\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-~\newline\\
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
-		\hline
-		\bf Item Number& V\refstepcounter{varnum}\thevarnum \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & \\
-		\hline
-		\bf Related Commonality & \cref{}\\
-		\hline
-		\bf Related Parameter & \pref{}\\
-		\hline
-		\bf History & \\
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
+		\hline
+		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{varprecise} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
+		\hline
+		\bf Description & The precision needed for each input and output.\\
+		\hline
+		\bf Related Commonality & \cref{precision}\\
+		\hline
+		\bf Related Parameter & \pref{parprecise}\\
+		\hline
+		\bf History & Created April 1, 2020\\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+~\newline\\
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
+		\hline
+		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{varaccuracy} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
+		\hline
+		\bf Description & The required accuracy for the output.\\
+		\hline
+		\bf Related Commonality & \cref{accuracy}\\
+		\hline
+		\bf Related Parameter & \pref{paraccuracy}\\
+		\hline
+		\bf History & Created April 1, 2020\\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+~\newline\\
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
+		\hline
+		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{varresponsetime} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
+		\hline
+		\bf Description & The response time required for user interaction with the system varies.\\
+		\hline
+		\bf Related Commonality & \cref{performance}\\
+		\hline
+		\bf Related Parameter & \pref{parresponse}\\
+		\hline
+		\bf History & Created April 1, 2020\\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+~\newline\\
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
+		\hline
+		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{varportability} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
+		\hline
+		\bf Description & The operating systems on which a LBS can run.\\
+		\hline
+		\bf Related Commonality & \cref{portability}\\
+		\hline
+		\bf Related Parameter & \pref{parportability}\\
+		\hline
+		\bf History & Created April 1, 2020\\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+~\newline\\
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
+		\hline
+		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{varusability} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
+		\hline
+		\bf Description & The ease with which LBS can effectively be run varies.\\
+		\hline
+		\bf Related Commonality & \cref{usability}\\
+		\hline
+		\bf Related Parameter & \pref{parusability}\\
+		\hline
+		\bf History & Created April 1, 2020\\
 		\hline
@@
 \section{Parameters of Variation}
+This section specifies the parameters of variation for the variabilities listed in Section \ref{var_sec}.
+They are organized into the same five subcategories as employed previously: Mesh Generation, Input, Output, System Constraints, Nonfunctional Requirements.
+
+Each parameter of variation is given a unique identifier of the form “P” followed by a natural number.
+The corresponding variability is listed and a hyperlink is provided that allows navigation back to the appropriate item in Section \ref{var_sec}.
+The final entry for each parameter of variation is the binding time, which is the time in the software lifecycle when the variability is fixed.
+The binding time could be during specification, or during building
+of the system (compile time), or during execution of the system (run time).
+It is possible to have a mixture of binding times.
+For instance, a parameter of variation could have a binding time of “specification or building” to represent that the parameter could be set at specification time, or it could be postponed until the given family member is built.
+The choice of postponing the decision until the build would be associated with the presence of a domain specific language that would allow postponing decisions on the values of the parameter of variation.
+
 \subsection{Lattice Boltzmann Method Solvers}
@@
 		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{}\\
-		\hline
-		\bf Range of Parameters & \\
-		\hline
-		\bf Binding Time & \\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-~\newline\\
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
-		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{}\\
-		\hline
-		\bf Range of Parameters & \\
-		\hline
-		\bf Binding Time & \\
+		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parrelax} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
+		\hline
+		\bf Corresponding Variability & \vref{varrelaxops}\\
+		\hline
+		\bf Range of Parameters & Single relaxation rate, two relaxation-time methods, multi-relaxation -time can be used.\\
+		\hline
+		\bf Binding Time & Build Time\\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+~\newline\\
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
+		\hline
+		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parparallel} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
+		\hline
+		\bf Corresponding Variability & \vref{parallel}\\
+		\hline
+		\bf Range of Parameters & If parallelism is used, commonly OpenMP, OpenCL, CUDA, or MPI are used.\\
+		\hline
+		\bf Binding Time & Build Time\\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+~\newline\\
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
+		\hline
+		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{paredf} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
+		\hline
+		\bf Corresponding Variability & \vref{parallel}\\
+		\hline
+		\bf Range of Parameters & Equilibrium can be approximated up to different orders in incompressible or compressible versions.\\
+		\hline
+		\bf Binding Time & Build Time\\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+~\newline\\
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
+		\hline
+		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parstorage} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
+		\hline
+		\bf Corresponding Variability & \vref{edfstorage}\\
+		\hline
+		\bf Range of Parameters & Single and multiple array storage patterns can be used.\\
+		\hline
+		\bf Binding Time & Build Time\\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+~\newline\\
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
+		\hline
+		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{pardimensions} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
+		\hline
+		\bf Corresponding Variability & \vref{dimensions}\\
+		\hline
+		\bf Range of Parameters & LBS models can have up to 3 dimensions.\\
+		\hline
+		\bf Binding Time & Build Time or Run Time\\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+~\newline\\
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
+		\hline
+		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parveldir} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
+		\hline
+		\bf Corresponding Variability & \vref{numveldir}\\
+		\hline
+		\bf Range of Parameters & One dimensional models include options of 2, 3, and 5 velocity directions.
+Two dimensional models include options of 9, 13, and 15 velocity directions.
+Three dimensional models include options of 15, 19, and 27 velocity directions.
+\\
+		\hline
+		\bf Binding Time & Build Time or Run Time\\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+~\newline\\
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
+		\hline
+		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parcollisop} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
+		\hline
+		\bf Corresponding Variability & \vref{varcolop}\\
+		\hline
+		\bf Range of Parameters & SRT, TRT, BGK collision operators.\\
+		\hline
+		\bf Binding Time & Build Time\\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+~\newline\\
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
+		\hline
+		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{partransequ} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
+		\hline
+		\bf Corresponding Variability & \vref{vartraneq}\\
+		\hline
+		\bf Range of Parameters & Collision and collision free transport equations.\\
+		\hline
+		\bf Binding Time & Build Time\\
 		\hline
@@
 		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{}\\
-		\hline
-		\bf Range of Parameters & \\
-		\hline
-		\bf Binding Time & \\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-~\newline\\
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
-		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{}\\
-		\hline
-		\bf Range of Parameters & \\
-		\hline
-		\bf Binding Time & \\
+		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parinterface} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
+		\hline
+		\bf Corresponding Variability & \vref{interface}\\
+		\hline
+		\bf Range of Parameters & Input can be graphical, text or file.\\
+		\hline
+		\bf Binding Time & Build Time\\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+~\newline\\
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
+		\hline
+		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parfluids} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
+		\hline
+		\bf Corresponding Variability & \vref{numberoffluids}\\
+		\hline
+		\bf Range of Parameters & LBS can model one or more fluids.\\
+		\hline
+		\bf Binding Time & Build Time\\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+~\newline\\
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
+		\hline
+		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{partypeparm} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
+		\hline
+		\bf Corresponding Variability & \vref{typefluidparm}\\
+		\hline
+		\bf Range of Parameters & LBS fluid parameters include Reynolds Number, density, viscosity, time, pressure, force, direction, relaxation rate, turbulence.\\
+		\hline
+		\bf Binding Time & Build Time\\
 		\hline
@@
 		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{}\\
-		\hline
-		\bf Range of Parameters & \\
-		\hline
-		\bf Binding Time & \\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-~\newline\\
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
-		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{}\\
-		\hline
-		\bf Range of Parameters & \\
-		\hline
-		\bf Binding Time & \\
+		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parvarvisual} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
+		\hline
+		\bf Corresponding Variability & \vref{varvisual}\\
+		\hline
+		\bf Range of Parameters & LBS can provide 1D, 2D, and 3D rendering of  the model.\\
+		\hline
+		\bf Binding Time & Build Time or Run Time\\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+~\newline\\
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
+		\hline
+		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parformatoutput} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
+		\hline
+		\bf Corresponding Variability & \vref{formatoutput}\\
+		\hline
+		\bf Range of Parameters & LBS prediction information is output in either text or binary format.\\
+		\hline
+		\bf Binding Time & Build Time\\
 		\hline
@@
 		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{}\\
-		\hline
-		\bf Range of Parameters & \\
-		\hline
-		\bf Binding Time & \\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-~\newline\\
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
-		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{}\\
-		\hline
-		\bf Range of Parameters & \\
-		\hline
-		\bf Binding Time & \\
+		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parprocessing} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
+		\hline
+		\bf Corresponding Variability & \vref{processing}\\
+		\hline
+		\bf Range of Parameters & The LBS model can be calculated on the CPU or GPU.\\
+		\hline
+		\bf Binding Time & Build Time\\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+~\newline\\
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
+		\hline
+		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{paropsystem} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
+		\hline
+		\bf Corresponding Variability & \vref{operatingsystem}\\
+		\hline
+		\bf Range of Parameters & LBS can be run on Windows, MacOS, or Linux versions.\\
+		\hline
+		\bf Binding Time & Build Time\\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+~\newline\\
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
+		\hline
+		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parstorageconst} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
+		\hline
+		\bf Corresponding Variability & \vref{storage}\\
+		\hline
+		\bf Range of Parameters & The amount of memory and storage varies between LBS. LBS sizes begin in the low gigabytes. \\
+		\hline
+		\bf Binding Time & Build Time\\
 		\hline
@@
 		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{}\\
-		\hline
-		\bf Range of Parameters & \\
-		\hline
-		\bf Binding Time & \\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-~\newline\\
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
-		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{}\\
-		\hline
-		\bf Range of Parameters & \\
-		\hline
-		\bf Binding Time & \\
+		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parprecise} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
+		\hline
+		\bf Corresponding Variability & \vref{varprecise}\\
+		\hline
+		\bf Range of Parameters & Each input and each output will have a specified the precision with which the real number is stored and/or calculated.\\
+		\hline
+		\bf Binding Time & Build Time or Run Time\\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+~\newline\\
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
+		\hline
+		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{paraccuracy} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
+		\hline
+		\bf Corresponding Variability & \vref{varaccuracy}\\
+		\hline
+		\bf Range of Parameters & The tolerance allowed for each output produced by the system so that the system produces usable results. \\
+		\hline
+		\bf Binding Time & Build Time or Run Time\\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+~\newline\\
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
+		\hline
+		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parresponse} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
+		\hline
+		\bf Corresponding Variability & \vref{varresponsetime}\\
+		\hline
+		\bf Range of Parameters & The maximum amount of time that the user is expected to wait for the system to produce a result ranges between systems and models from near instantaneous to several minutes or longer.\\
+		\hline
+		\bf Binding Time & Specification\\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+~\newline\\
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
+		\hline
+		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parportability} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
+		\hline
+		\bf Corresponding Variability & \vref{varportability}\\
+		\hline
+		\bf Range of Parameters & The ease with which LBS can run on Windows, MacOS, and Linux system varies. Some LBS can run on all three, while some can run on multiple or only one operating system.\\
+		\hline
+		\bf Binding Time & Specification\\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+~\newline\\
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
+		\hline
+		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parusability} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
+		\hline
+		\bf Corresponding Variability & \vref{varusability}\\
+		\hline
+		\bf Range of Parameters & Some LBS require programming skills and domain expert knowledge while some LBS can be run without such skills.\\
+		\hline
+		\bf Binding Time & Specification\\
 		\hline

### 2020-04-03 StateOfPractice/Peter-Notes/Commonality-Analysis-LB-Systems.tex
fixed typo in commonality analysis and generated new methodology pdf
added feature diagram to Peter's directory and fixed errors in commonality analysis

@@
 		\hline
-		\bf Description & A lattice discretizes a computational domain into a finite number of points. All LBS discretize the computational domain using a regular, evenly spaced grid.\\
-		\hline
-		\bf Related Variability & \vref{numveldir} \vref{varcolop}\\
+		\bf Description & A lattice discretizes a computational domain into a finite number of points. All LBS discretize the computational domain using a regular, evenly spaced grid within a boundary.\\
+		\hline
+		\bf Related Variability & \vref{numveldir} \vref{varcolop} \vref{varboundary}\\
 		\hline
@@
 		\hline
-		\bf Related Variability & \vref{numveldir} \vref{varcolop}\\
+		\bf Related Variability & \vref{dimensions} \vref{numveldir}\\
 		\hline
@@
 \end{minipage}\\
-~\newline\\
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
-		\hline
-		\bf Item Number& C\refstepcounter{comnum}\thecomnum\label{discretizevelocity} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
-		\hline
-		\bf Description & All Lattice Boltzmann versions discretize velocity into a finite number of directions.\\
-		\hline
-		\bf Related Variability & \vref{numveldir}\\
-		\hline
-		\bf History & Created April 1, 2020\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-~\newline\\
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
-		\hline
-		\bf Item Number& C\refstepcounter{comnum}\thecomnum\label{relaxationupdate} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
-		\hline
-		\bf Description & All Lattice Boltzmann versions update the velocity distribution during each LBM iteration.\\
-		\hline
-		\bf Related Variability & \vref{varrelaxops}\\
-		\hline
-		\bf History & Created April 1, 2020\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-~\newline\\
-~\newline
-\noindent
 \begin{minipage}{\textwidth}
@@
 		\hline
-		\bf Related Variability & \vref{varedf} \vref{edfstoreage}\\
+		\bf Related Variability & \vref{varedf} \vref{edfstorage} \vref{edfcoeff}\\
 		\hline
@@
 		\hline
-		\bf Related Variability & \vref{interface} \vref{numberoffluids} \vref{typefluidparm}\\
+		\bf Related Variability & \vref{interface}\\
 		\hline
@@
 		\hline
-		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{varrelaxops} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & Different relaxation times are used to update the velocity distribution.\\
-		\hline
-		\bf Related Commonality & \cref{relaxationupdate}\\
-		\hline
-		\bf Related Parameter & \pref{parrelax}\\
-		\hline
-		\bf History & Created April 1, 2020\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-~\newline\\
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
-		\hline
 		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{parallel} \\
@@
 		\hline
+		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{edfcoeff} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
+		\hline
+		\bf Description & Coefficients for distribution function can vary.\\
+		\hline
+		\bf Related Commonality & \cref{equildistfunc}\\
+		\hline
+		\bf Related Parameter & \pref{paredfcoeff}\\
+		\hline
+		\bf History & Created April 1, 2020\\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+~\newline\\
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
+		\hline
 		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{dimensions} \\
@@
 		\hline
-		\bf Related Commonality & \cref{probabilitydensityfunc}\\
+		\bf Related Commonality & \cref{lattice} \cref{probabilitydensityfunc}\\
 		\hline
@@
 		\hline
-		\bf Related Commonality & \cref{lattice} \cref{collisionoperator} \cref{discretizevelocity}\\
+		\bf Related Commonality & \cref{lattice} \cref{collisionoperator}\\
 		\hline
@@
 \end{minipage}\\
+~\newline\\
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
+		\hline
+		\bf Item Number&  V\refstepcounter{varnum}\thevarnum\label{numberoffluids} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
+		\hline
+		\bf Description & The number of fluids allowed in the LBS.\\
+		\hline
+		\bf Related Commonality & \cref{fluidandboundaryinfo}\\
+		\hline
+		\bf Related Parameter & \pref{parfluids}\\
+		\hline
+		\bf History & Created April 1, 2020\\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+~\newline\\
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
+		\hline
+		\bf Item Number&  V\refstepcounter{varnum}\thevarnum\label{typefluidparm} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
+		\hline
+		\bf Description & The type of fluid parameters.\\
+		\hline
+		\bf Related Commonality & \cref{fluidandboundaryinfo}\\
+		\hline
+		\bf Related Parameter & \pref{partypeparm}\\
+		\hline
+		\bf History & Created April 1, 2020\\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+~\newline\\
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
+		\hline
+		\bf Item Number&  V\refstepcounter{varnum}\thevarnum\label{varboundary} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
+		\hline
+		\bf Description & The boundary of the lattice can have varied conditions.\\
+		\hline
+		\bf Related Commonality & \cref{lattice}\\
+		\hline
+		\bf Related Parameter & \pref{parboundary}\\
+		\hline
+		\bf History & Created April 1, 2020\\
+		\hline
+	\end{tabular}
+\end{minipage}\\
 \subsection{Input}
@@
 		\bf Related Parameter & \pref{parinterface}\\
-		\hline
-		\bf History & Created April 1, 2020\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-~\newline\\
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
-		\hline
-		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{numberoffluids} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & The number of fluids allowed in the LBS.\\
-		\hline
-		\bf Related Commonality & \cref{fluidandboundaryinfo}\\
-		\hline
-		\bf Related Parameter & \pref{parfluids}\\
-		\hline
-		\bf History & Created April 1, 2020\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-~\newline\\
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
-		\hline
-		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{typefluidparm} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & The type of fluid parameters.\\
-		\hline
-		\bf Related Commonality & \cref{fluidandboundaryinfo}\\
-		\hline
-		\bf Related Parameter & \pref{partypeparm}\\
 		\hline
@@
 		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parrelax} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{varrelaxops}\\
-		\hline
-		\bf Range of Parameters & Single relaxation rate, two relaxation-time methods, multi-relaxation -time can be used.\\
+		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parparallel} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
+		\hline
+		\bf Corresponding Variability & \vref{parallel}\\
+		\hline
+		\bf Range of Parameters & If parallelism is used, commonly OpenMP, OpenCL, CUDA, or MPI are used.\\
 		\hline
@@
 		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parparallel} \\
+		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{paredf} \\
 		\hline
@@
 		\hline
-		\bf Range of Parameters & If parallelism is used, commonly OpenMP, OpenCL, CUDA, or MPI are used.\\
+		\bf Range of Parameters & Equilibrium can be approximated up to different orders in incompressible or compressible versions.\\
 		\hline
@@
 		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{paredf} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{parallel}\\
-		\hline
-		\bf Range of Parameters & Equilibrium can be approximated up to different orders in incompressible or compressible versions.\\
+		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parstorage} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
+		\hline
+		\bf Corresponding Variability & \vref{edfstorage}\\
+		\hline
+		\bf Range of Parameters & Single and multiple array storage patterns can be used.\\
 		\hline
@@
 		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parstorage} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{edfstorage}\\
-		\hline
-		\bf Range of Parameters & Single and multiple array storage patterns can be used.\\
+		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{paredfcoeff} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
+		\hline
+		\bf Corresponding Variability & \vref{edfcoeff}\\
+		\hline
+		\bf Range of Parameters & Numerous coefficients for equilibrium distribution function based on number of velocity directions. \\
 		\hline
@@
 		\hline
-		\bf Range of Parameters & SRT, TRT, BGK collision operators.\\
+		\bf Range of Parameters & SRT, TRT, MRT, BGK collision operators.\\
 		\hline
@@
 \end{minipage}\\
+~\newline\\
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
+		\hline
+		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parfluids} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
+		\hline
+		\bf Corresponding Variability & \vref{numberoffluids}\\
+		\hline
+		\bf Range of Parameters & LBS can model one or more fluids.\\
+		\hline
+		\bf Binding Time & Build Time\\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+~\newline\\
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
+		\hline
+		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{partypeparm} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
+		\hline
+		\bf Corresponding Variability & \vref{typefluidparm}\\
+		\hline
+		\bf Range of Parameters & LBS fluid parameters include Reynolds Number, density, viscosity, time, pressure, force, direction, relaxation rate, turbulence.\\
+		\hline
+		\bf Binding Time & Build Time\\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+~\newline\\
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
+		\hline
+		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parboundary} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
+		\hline
+		\bf Corresponding Variability & \vref{varboundary}\\
+		\hline
+		\bf Range of Parameters & Lattice boundary can have reflective or non-reflective conditions.\\
+		\hline
+		\bf Binding Time & Build Time\\
+		\hline
+	\end{tabular}
+\end{minipage}\\
 \subsection{Input}
@@
 		\bf Range of Parameters & Input can be graphical, text or file.\\
-		\hline
-		\bf Binding Time & Build Time\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-~\newline\\
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
-		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parfluids} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{numberoffluids}\\
-		\hline
-		\bf Range of Parameters & LBS can model one or more fluids.\\
-		\hline
-		\bf Binding Time & Build Time\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-~\newline\\
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
-		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{partypeparm} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{typefluidparm}\\
-		\hline
-		\bf Range of Parameters & LBS fluid parameters include Reynolds Number, density, viscosity, time, pressure, force, direction, relaxation rate, turbulence.\\
 		\hline

### 2020-04-13 StateOfPractice/Methodology/Methodology.tex
commonality analysis for LB
update Methodology

@@
 
+Instead of only focus on the current status of the above numbers, we may find
+the time history of them to be more valuable. For example, the number of
+contributors over time, the number of lines of code over time, the number of
+open issues over time, etc.
+
 \subsection{Processed Data}
@@
 Visibility/Transparency
-\item lines of new code produced per person-day - Productivity
-\item lines/pages of new documents produced per person-day - Productivity
-\item number of issues closed per person-day - Productivity
+\item lines of new code produced per person-month - Productivity
+\item lines/pages of new documents produced per person-month - Productivity
+\item number of issues closed per person-month - Productivity
 \item percentage of comment lines in the code - maintainability \ad{Not Ao's
@@
 \end{itemize}
+
+In the above calculations, a month can be determined to be 30 days.
 

### 2020-04-13 StateOfPractice/Methodology/Methodology.tex
Revert "update Methodology"
This reverts commit 71dc42cc24a9df57377a1b5d1579554b50ecfb42.

--- 2020-03-31 StateOfPractice/Methodology/Methodology.tex
@@
 \usepackage{amsmath,amsthm}
+\usepackage{hyperref}
 
@@
 
+\begin{enumerate}
+	\item Must be open source.
+	\item Must have GitHub repository.
+\end{enumerate}
+
 \section{Domain Analysis} \label{SecDomainAnalysis}
@@
 Commonality analysis.  Follow as for mesh generator (likely with less detail).
+
+Commonality analysis document Steps:
+\begin{enumerate}
+	\item Introduction
+	\item Overview of Domain
+	\item Add Commonalities - Split into simulation, input, output, and nonfunctional requirements
+	\item Add Variabilites - Split into simulation, input, output, system constraints, and nonfunctional requirements
+	\item Add Parameters of Variation - Split into simulation, input, output, system constraints, and nonfunctional requirements
+	\item Add Terminology, Definitions, Acronyms
+\end{enumerate}
+
+Commonality analysis for Lattice Boltzmann Solvers can be found \href{run:../Peter-Notes/Commonality-Analysis-LB-Systems.pdf}{here}.
 

### 2020-04-14 StateOfPractice/Methodology/GradingTemplate.tex
Copy the TemplateToGradeSCSoft to here

The table below lists the full set of measures that are assessed for each
software product.  The measures are grouped under headings for each quality, and
one for summary information.  Following each measure, the type for a valid
result is given in brackets.  Many of the types are given as enumerated sets.
For instance, the response on many of the questions is one of ``yes,'' ``no,''
or ``unclear.''  The type ``number'' means natural number, a positive integer.  The
types for date and url are not explicitly defined, but they are what one would
expect from their names.  In some cases the response for a given question is not
necessarily limited to one answer, such as the question on what platforms are
supported by the software product.  Case like this are indicated by ``set of''
preceding the type of an individual answer.  The type in these cases are then
the power set of the individual response type.  In some cases a superscript $^*$
is used to indicate that a response of this type should be accompanied by
explanatory text.  For instance, if problems were caused by uninstall, the
reviewer should note what problems were caused.

\begin{longtable}{p{16cm}}
  \caption{Grading Template}   \label{table:TemplateFull}\\
  \toprule
  \textbf{Summary Information}\\
  \midrule
  Software name? (string)\\
  URL? (url)\\
  Educational institution (string)\\
  Software purpose (string)\\
  Number of developers (number)\\
  How is the project funded (string)\\
  Number of downloads for current version (number)\\
  Release date (date)\\
  Last updated (date)\\
  Status (\{alive, dead, unclear\})\\
  License (\{GNU GPL, BSD, MIT, terms of use, trial, none, unclear\})\\
  Platforms (set of \{Windows, Linux, OS X, Android, Other OS\})\\
  Category (\{concept, public, private\})\\
  Development model (\{open source, freeware, commercial\})\\
  Publications using the software (set of url)\\
  Publications about the software (set of url) \\
  Is source code available? (\{yes, no\})\\
  Programming language(s) (set of \{FORTRAN, Matlab, C, \CC, Java, R, Ruby,
  Python, Cython, BASIC, Pascal, IDL, unclear\})\\

  \midrule
  \textbf{Installability} (Measured via installation on a virtual machine.)\\
  \midrule

  Are there installation instructions? (\{yes, no\})\\
  Are the installation instructions linear? (\{yes, no, n/a\})\\
  Is there something in place to automate the installation? (\{yes$^*$, no\})\\
  Is there a specified way to validate the installation, such as a test suite? (\{yes$^*$, no\})\\
  How many steps were involved in the installation? (number)\\
  How many software packages need to be installed before or during installation?
  (number)\\
  Run uninstall, if available. Were any obvious problems caused? (\{unavail, yes$^*$, no\})\\
  Overall impression? (\{1 .. 10\})\\

  \midrule
  \textbf{Correctness and Verifiability}\\
  \midrule

  Are external libraries used? (\{yes$^*$, no, unclear\})\\
  Does the community have confidence in this library? (\{yes, no, unclear\})\\
  Any reference to the requirements specifications of the program?
  (\{yes$^*$, no, unclear\})\\
  What tools or techniques are used to build confidence of correctness? (string)\\
  If there is a getting started tutorial, is the output as expected? (\{yes, no$^*$, n/a\})\\
  Overall impression? (\{1 .. 10\})\\

  \midrule
  \textbf{Surface Reliability}\\
  \midrule

  Did the software ``break'' during installation? (\{yes$^*$, no\})\\
  Did the software ``break'' during the initial tutorial testing? (\{yes$^*$, no, n/a\})\\
  Overall impression? (\{1 .. 10\})\\

  \midrule
  \textbf{Surface Robustness}\\
  \midrule

  Does the software handle garbage input reasonably? (\{yes, no$^*$\})\\
  For any plain text input files, if all new lines are replaced with new lines
  and carriage returns, will the software handle this gracefully? (\{yes,
  no$^*$, n/a\})\\
  Overall impression? (\{1 .. 10\})\\

  \midrule
  \textbf{Surface Performance}\\
  \midrule

  Is there evidence that performance was considered? (\{yes$^*$, no\})\\
  Overall impression? (\{1 .. 10\})\\

  \midrule
  \textbf{Surface Usability}\\
  \midrule

  Is there a getting started tutorial? (\{yes, no\})\\
  Is there a standard example that is explained? (\{yes, no\})\\
  Is there a user manual? (\{yes, no\})\\
  Does the application have the usual ``look and feel'' for the platform it is
  on? (\{yes, no$^*$\})\\
  Are there any features that show a lack of visibility? (\{yes, no$^*$\})\\
  Are expected user characteristics documented? (\{yes, no\})\\
  What is the user support model? (string)\\
  Overall impression? (\{1 .. 10\})\\

  \midrule
  \textbf{Maintainability}\\
  \midrule

  Is there a history of multiple versions of the software?  (\{yes, no, unclear\})\\
  Is there any information on how code is reviewed, or how to contribute?
  (\{yes$^*$, no\})\\
  Is there a changelog?  (\{yes, no\})\\
  What is the maintenance type? (set of \{corrective, adaptive, perfective, unclear\})\\
  What issue tracking tool is employed? (set of \{Trac, JIRA, Redmine, e-mail,
  discussion board, sourceforge, google code, git, none, unclear\})\\
  Are the majority of identified bugs fixed? (\{yes, no$^*$, unclear\})\\
  Which version control system is in use? (\{svn, cvs, git, github, unclear\})\\
  Is there evidence that maintainability was considered in the design?
  (\{yes$^*$, no\})\\
  Are there code clones? (\{yes$^*$, no, unclear\})\\
  Overall impression? (\{1 .. 10\})\\

  \midrule
  \textbf{Reusability}\\
  \midrule

  Are any portions of the software used by another package? (\{yes$^*$, no\})\\
  Is there evidence that reusability was considered in the design? (API
  documented, web service, command line tools, ...) (\{yes$^*$, no, unclear\})\\
  Overall impression? (\{1 .. 10\})\\

  \midrule
  \textbf{Portability}\\
  \midrule

  What platforms is the software advertised to work on?
  (set of \{Windows, Linux, OS X, Android, Other OS\})\\
  Are special steps taken in the source code to handle portability? (\{yes$^*$,
  no, n/a\}) \\
  Is portability explicitly identified as NOT being important? (\{yes, no\})\\
  Convincing evidence that portability has been achieved? (\{yes$^*$, no\})\\
  Overall impression? (\{1 .. 10\})\\

  \midrule
  \textbf{Surface Understandability} (Based on 10 random source files)\\
  \midrule

  Consistent indentation and formatting style? (\{yes, no, n/a\})\\
  Explicit identification of a coding standard? (\{yes$^*$, no, n/a\})\\
  Are the code identifiers consistent, distinctive, and
  meaningful? (\{yes, no$^*$, n/a\})\\
  Are constants (other than 0 and 1) hard coded into the program? (\{yes, no$^*$, n/a\})\\
  Comments are clear, indicate what is being done, not how? (\{yes, no$^*$, n/a\})\\
  Is the name/URL of any algorithms used mentioned?
  (\{yes, no$^*$, n/a\})\\
  Parameters are in the same order for all functions? (\{yes, no$^*$, n/a\})\\
  Is code modularized? (\{yes, no$^*$, n/a\})\\
  Descriptive names of source code files? (\{yes, no$^*$, n/a\})\\
  Is a design document provided? (\{yes$^*$, no, n/a\})\\
  Overall impression? (\{1 .. 10\})\\

  \midrule
  \textbf{Interoperability}\\
  \midrule

  Does the software interoperate with external systems? (\{yes$^*$, no\})\\
  Is there a workflow that uses other softwares? (\{yes$^*$, no\})\\
  If there are external interactions, is the API clearly defined? (\{yes$^*$, no, n/a\})\\
  Overall impression? (\{1 .. 10\})\\

  \midrule
  \textbf{Visibility/Transparency}\\
  \midrule

  Is the development process defined? If yes, what process is used. (\{yes$^*$, no, n/a\})\\
  Ease of external examination relative to other products
  considered?  (\{1 .. 10\})\\
  Overall impression? (\{1 .. 10\})\\

  \midrule
  \textbf{Reproducibility}\\
  \midrule

  Is there a record of the environment used for their development and testing?
  (\{yes$^*$, no\})\\
  Is test data available for verification?  (\{yes, no\})\\
  Are automated tools used to capture experimental context?  (\{yes$^*$, no\})\\
  Overall impression? (\{1 .. 10\})\\

  \bottomrule

\end{longtable}

### 2020-04-14 StateOfPractice/Methodology/GradingTemplate.tex
start comments and suggestions

@@
   Publications about the software (set of url) \\
-  Is source code available? (\{yes, no\})\\
+  Is source code available? (\{yes, no\})\ad{It might be better to directly ask for a set of url, and the answer can be n/a if it's not open source, or unclear if it can't be determined.}\\
+  Source code URL (\{set of url, n/a, unclear\})\ad{example}\\
   Programming language(s) (set of \{FORTRAN, Matlab, C, \CC, Java, R, Ruby,

### 2020-04-15 StateOfPractice/Methodology/GradingTemplate.tex
some comments to the grading template

@@
   Software purpose (string)\\
-  Number of developers (number)\\
+  Number of developers (number)\ad{It might be hard to decide.
+For example 3D Slicer has 171 people with commits to its repo, but on its website, 18 contributors(including people who are not developers) are acknowledged.}\\
   How is the project funded (string)\\
@@
   Is there something in place to automate the installation? (\{yes$^*$, no\})\\
-  Is there a specified way to validate the installation, such as a test suite? (\{yes$^*$, no\})\\
-  How many steps were involved in the installation? (number)\\
+  Is there a specified way to validate the installation, such as a test suite? \ad{What's the difference between this one and the getting started tutorial?} (\{yes$^*$, no\})\\
+  How many steps were involved in the installation? (number) \ad{what if the installation has different steps on different OS?}\\
   How many software packages need to be installed before or during installation?
-  (number)\\
+  (number) \ad{should it be 0 if no packages needed? I saw some answers with "no" or "none"}\\
   Run uninstall, if available. Were any obvious problems caused? (\{unavail, yes$^*$, no\})\\
@@
   Any reference to the requirements specifications of the program?
-  (\{yes$^*$, no, unclear\})\\
-  What tools or techniques are used to build confidence of correctness? (string)\\
+  (\{yes$^*$, no, unclear\})\ad{Seems like a good question to put in the questionnaire}\\
+  What tools or techniques are used to build confidence of correctness? (string)\ad{Seems like a good question to put in the questionnaire}\\
   If there is a getting started tutorial, is the output as expected? (\{yes, no$^*$, n/a\})\\
@@
 
-  Does the software handle garbage input reasonably? (\{yes, no$^*$\})\\
+  Does the software handle garbage input reasonably? (\{yes, no$^*$\}) \ad{How do we prepare garbage input?}\\
   For any plain text input files, if all new lines are replaced with new lines
@@
   Are expected user characteristics documented? (\{yes, no\})\\
-  What is the user support model? (string)\\
+  What is the user support model? (string)\ad{Can GitHub be one? Many users directly ask questions there by open new issues.}\\
   Overall impression? (\{1 .. 10\})\\
@@
   What issue tracking tool is employed? (set of \{Trac, JIRA, Redmine, e-mail,
-  discussion board, sourceforge, google code, git, none, unclear\})\\
-  Are the majority of identified bugs fixed? (\{yes, no$^*$, unclear\})\\
+  discussion board, sourceforge, google code, git, none, unclear\}) \ad{Does git mean GitHub and GitLab? Should we add BitBucket here?}\\
+  Are the majority of identified bugs fixed? (\{yes, no$^*$, unclear\})\ad{It might be better to record a number here}\\
+  What is the percentage of identified issues that are closed? (percentage)\ad{an example}\\
   Which version control system is in use? (\{svn, cvs, git, github, unclear\})\\
@@
   (\{yes$^*$, no\})\\
-  Are there code clones? (\{yes$^*$, no, unclear\})\\
+  Are there code clones? (\{yes$^*$, no, unclear\})\ad{Might need to test some static analyzer.}\\
   Overall impression? (\{1 .. 10\})\\
@@
 
-  Are any portions of the software used by another package? (\{yes$^*$, no\})\\
+  Are any portions of the software used by another package? (\{yes$^*$, no\})\ad{It seems hard to find an answer for some software.}\\
   Is there evidence that reusability was considered in the design? (API

### 2020-04-15 OverallResearchProposal/ObjectivesAndResearchQuestions.tex
First draft of research objectives and research questions

\documentclass[12pt]{article}

\usepackage[round]{natbib}
\usepackage{booktabs}

\usepackage{hyperref}
\hypersetup{
    colorlinks=true,       % false: boxed links; true: colored links
    linkcolor=red,          % color of internal links (change box color with linkbordercolor)
    citecolor=blue,       % color of links to bibliography
    filecolor=magenta,   % color of file links
    urlcolor=cyan           % color of external links
}

%% Comments
\newif\ifcomments\commentstrue

\ifcomments
\newcommand{\authornote}[3]{\textcolor{#1}{[#3 ---#2]}}
\newcommand{\todo}[1]{\textcolor{red}{[TODO: #1]}}
\else
\newcommand{\authornote}[3]{}
\newcommand{\todo}[1]{}
\fi

\newcommand{\wss}[1]{\authornote{blue}{SS}{#1}} %Spencer Smith
\newcommand{\jc}[1]{\authornote{red}{JC}{#1}} %Jacques Carette
\newcommand{\oo}[1]{\authornote{magenta}{OO}{#1}} %Olu Owojaiye
\newcommand{\pmi}[1]{\authornote{green}{PM}{#1}} %Peter Michalski
\newcommand{\ad}[1]{\authornote{cyan}{AD}{#1}} %Ao Dong

%\oddsidemargin 0mm
%\evensidemargin 0mm
%\textwidth 160mm
%\textheight 200mm

\begin{document}

\title{Empirical Study of Scientific Computer Software: Objectives and Research Questions} 
\author{Spencer Smith}
\date{\today}

%\maketitle

\subsection* {Overall Objective}

\textbf{What knowledge needs to be captured and what principles, processes and
  methodologies followed to produce sustainable \{software + artifacts\} with
  a reasonable amount of time and energy?}\\

In the above definition, software is considered to be an application (app)
and/or a library.  The outcome of the development process is not just software.
It is \{software + artifacts\}.  \wss{softwartifacts?} The artifacts consist of
such work products as documentation, build scripts, code, test cases,
contributor's guides, etc.  The quality of sustainability cannot be assessed by
looking at the software alone, or the code alone; it depends on everything,
including the processes and methodologies used.  Further information on the
relevant definitions and context is given in the next section.

The notion of ``reasonable amount of time and energy'' is part of the objective,
but reasonable is not defined in the objective.  What is reasonable is going to
depend on the context.  Problems that are large, complex and/or critical are
going to have a greater ``budget'' of time and energy.  When comparing different
alternatives for processes and methodologies, preference would be given to the
alternative that costs less time and energy, assuming that this alternative
achieves the same level of sustainability as its competition.  The assumptions
listed in the next section outline the typical space of problems where it makes
sense to aim for sustainable \{software + artifacts\}.

In the statement of the objective, principles, processes and methodologies are
listed, but not tools.  Tools are needed to support methodologies, but tools
change rapidly.  Our goal is to be more general.  If the right principles,
processes and methodologies are found, the tools will adapt to match.  However,
understanding the current state of the practice will involve looking at tools,
as shown in the later list of research questions.

The objective is not phrased to target a specific software domain, but one of
the research assumptions is that we will focus on Scientific Computing Software
(SCS) examples and applications.  This focus is because creating sustainable
software+artifacts is more feasible when the software domain is well understood,
which is the case for SCS.  Moreover, SCS is an important, often neglected,
domain of software.

\subsection*{Context, Definitions and Assumptions}

The overall objective is too ambiguous as given.  It requires additional
information/interpretation.  Therefore this section provides definitions of the
relevant terms and a list of the assumptions that define the potential contexts
where sustainable software+artifact development is reasonable.

\subsubsection*{Definitions}

\begin{description}
\item[Knowledge] Facts, definitions, theories, assumptions, etc. \wss{needs a
    proper definition, with citation(s)}
\item[Process] Organization of the software development activities.  \wss{needs a
    proper definition, with citation(s)}
\item[Principles] Principles are ``general and abstract statements describing
  desirable properties of software processes and products.''  \citep[p.\
  41]{GhezziEtAl2003}
\item[Methods] ``Methods are general guidelines that govern the execution of some activity;
  they are rigorous, systematic, and disciplined approaches.''  \citep[p.\
  41]{GhezziEtAl2003}
\item[Techniques] ``Techniques are more technical and mechanical than methods.''
  \cite[p.\ 41]{GhezziEtAl2003}
\item[Methodology] Combination of methods and techniques.  \citep[p.\
  41]{GhezziEtAl2003}
\item[Tools] ``Tools ... are developed to support the application of techniques,
  method and methodologies.'' \cite[p.\ 41]{GhezziEtAl2003}

\item[Scientific Computing] the use of computational tools to analyze or
  simulate (continuous) mathematical models of real world systems of engineering
  or scientific importance so that we can better understand and (potentially)
  predict the system's behaviour. \citep{SmithAndLai2005}
\item[Sustainable] We start with the general definition:

``Sustainable development is development that meets the needs
  of the present without compromising the ability of future generations to meet
  their own needs.'' \citep{Brundtland1987}

  This definition is general; it is not specific to software and it is silent on
  the specific needs of the present and the future.  The definition can be made
  less abstract by introducing the context of software+artifacts and thinking
  separately about the needs of the present and the future.  For the needs of
  the present, the software should be correct, reliable and usable. For the
  needs of the future, given uncertainty, this means supporting change from
  relevant concerns (political, economic, social, technical, legal and
  environmental) should be possible with a reasonable level of resources.
  Supporting the future then means producing software+artifacts that are
  maintainable, portable, reusable and reproducible.

  Summarizing the above, the proposed definition for sustainability is:

  \emph{Sustainable software means software+artifacts that are correct,
    reliable, and usable (for the present) and maintainable, portable, reusable
    and reproducible (for the future).}

This definition has brought in several other qualities.  These qualities are
defined in the ``Quality  Definitions of Qualities'' document.
\wss{Reproducibility above might be changed to replicability, or possibly both
  terms should be included?}
\wss{Should performance be included in the list of qualities?}

\item[Software] Application and/or library.
\item[Application (App)] An executable program. \wss{needs a
    proper definition, with citation(s)}
\item[Library] Services that are available to other programs, but not an
  executable program itself. \wss{needs a
    proper definition, with citation(s)}
\item[Artifacts] Work products generated during the process of creating
  software.  The work products include requirements documentation, design
  documentation, verification and validation plans, verification and validation
  reports, contributor's guides, user guides, build scripts, code, test cases,
  etc.
\end{description}

\subsubsection*{Assumptions}

Producing sustainable software+artifacts is not a trivial undertaking.  Not
every project needs to aim for sustainability.  Our position is that the time
and energy is justified when one or more of the following assumptions applies.

\begin{itemize}
\item The software project is going to be long lived, where long life means at
  least 10 years.
\item The software+artifacts should interest multiple stakeholders (not just the
  original developer), with different interests and backgrounds.
\item The software is safety relevant, such as software for nuclear safety,
  medical imaging or computational medicine.
\item The software explicitly, or implicitly, is part of a program family of
  related software.  When the program family assumption means that there are
  related programs that are less effort to design and build together than to
  design and build as separate projects.  Program family development depends on
  satisfying three hypotheses \citep{Weiss1997}:
\begin{itemize}
\item Redevelopment hypothesis – most software development involved in producing
  an individual family members should be redevelopment because the family
  members have so much in common.
\item Oracle Hypothesis – the changes that are likely to occur during the soft-
  ware’s lifetime are predictable.
\item Organizational Hypothesis – designers can organize the software and the
  development effort so that predicted changes can be made independently.
\end{itemize}
\item The project recognizes that software artifacts other than the code are
  relevant.
\item The domain of the software project is well understood.  This assumption is
  related to the oracle hypothesis.
\end{itemize}

Although this study is not specifically restricted to open-source software, for
practical reasons (especially the lack of access to the code), commercial
software will not be strongly emphasized.

\subsubsection*{Current State of the Practice Research Questions}

A good starting point for understanding how to develop sustainable SCS software+artifacts is
to look at how SCS is currently developed.  By measuring relevant qualities and
relating them to development practices, we can form an idea of what is currently
working well.

Following the key points of the overall objective, the research questions are
based on investigating the following: i) knowledge, ii) principles, processes
and methodologies, iii) software qualities and iv) the necessary investment of
time and energy.  The scope of the research questions are not trying to cover
everything, but rather to focus on the most important aspects for SCS.

\begin{enumerate}
\item What knowledge is currently captured in the artifacts generated by
  existing open source SCS projects?
\item What principles, processes, methodologies and tools are used by existing
  open source SCS projects?
\item What are the ``pain points'' for developers working on SCS projects?  What
  aspects of the existing processes, methodologies and tools do they consider
  could potentially be improved?
\item For a given software product (software+artifacts), how can we produce a
  reasonable measure of the software qualities (correctness, reliability,
  usability etc.) with a few hours of effort?
\item For a given software product (software+artifacts), how can we produce a
  meaningful measure of usability with a few days worth of effort?
\item For a given software product (software+artifacts), how can we produce a
  meaningful measure of maintainability (with respect to likely changes) with a
  few days worth of effort?
\item For a given software product (software+artifacts), how can we produce a
  meaningful measure of reproducibility/replicability with a few days worth of
  effort?
\item With a reasonable effort and time, how can we compare the productivity
  between different processes, methodologies and tools?
\item Can a correlation be identified between software projects that use best
  practices for the development process and methodologies and improvement in the
  qualities of usability, performance and reproducibility/replicability?
\end{enumerate}

\subsubsection*{Ideal Process}

Some empirical research studies of SCS stop at the previous step.  They aim to
identify what is currently working well, and then implicitly assume that the
identified process/methodologies/tools provide the best guidelines for others to
follow.  However, identifying the best current approach to developing SCS does
not imply that the best possible approach has been found.  Our instinct is that
there is plenty of room for improvement with SCS development.  The following
research questions explore this possibility.

\begin{enumerate}
\item Ignoring the time and effort required, what knowledge is necessary to
  capture for sustainable SCS?  How can this knowledge best be documented in
  software artifacts?
\item How can Model Driven Engineering and Code Generation be fit in the
  development process for SCS to remove as many repetitive, tedious and error
  prone tasks as possible?  We will call this the ``ideal process.''
\item How well do existing modelling and code generation tools do at
  implementing the ``ideal process''?  Candidates for comparison include Drasil
  and Epsilon.
\item How well does a model driven approach to compared to a traditional
  approach for usability, productivity and sustainability?
\item What are the attitudes and preferences of typical users towards
  an implementation of the ``ideal process''?
\end{enumerate}

\bibliographystyle{plainnat}
\bibliography {../CommonFiles/ResearchProposal}

\end{document}

### 2020-04-15 StateOfPractice/Methodology/GradingTemplate.tex
more comments to the grading template

@@
   Publications about the software (set of url) \\
-  Is source code available? (\{yes, no\})\ad{It might be better to directly ask for a set of url, and the answer can be n/a if it's not open source, or unclear if it can't be determined.}\\
-  Source code URL (\{set of url, n/a, unclear\})\ad{example}\\
+  Is source code available?
+(\{yes, no\})\ad{It might be better to directly ask for a set of url, and the answer can be n/a if it's not open source, or unclear if it can't be determined.
+An example is as follows.}\\
+  \ad{Example}Source code URL (\{set of url, n/a, unclear\})\\
   Programming language(s) (set of \{FORTRAN, Matlab, C, \CC, Java, R, Ruby,
@@
   discussion board, sourceforge, google code, git, none, unclear\}) \ad{Does git mean GitHub and GitLab? Should we add BitBucket here?}\\
-  Are the majority of identified bugs fixed? (\{yes, no$^*$, unclear\})\ad{It might be better to record a number here}\\
-  What is the percentage of identified issues that are closed? (percentage)\ad{an example}\\
+  Are the majority of identified bugs fixed? (\{yes, no$^*$, unclear\})\ad{It might be better to record a number here. An example is as follows.}\\
+  \ad{Example}What is the percentage of identified issues that are closed? (percentage)\\
   Which version control system is in use? (\{svn, cvs, git, github, unclear\})\\
@@
   Is portability explicitly identified as NOT being important? (\{yes, no\})\\
-  Convincing evidence that portability has been achieved? (\{yes$^*$, no\})\\
+  Convincing evidence that portability has been achieved? (\{yes$^*$, no\})\ad{Seems like a good question to put in the questionnaire}\\
   Overall impression? (\{1 .. 10\})\\
@@
   Descriptive names of source code files? (\{yes, no$^*$, n/a\})\\
-  Is a design document provided? (\{yes$^*$, no, n/a\})\\
+  Is a design document provided? (\{yes$^*$, no, n/a\})\ad{What is a design document? There can be some examples.}\\
   Overall impression? (\{1 .. 10\})\\
@@
   If there are external interactions, is the API clearly defined? (\{yes$^*$, no, n/a\})\\
+  \ad{Some more suggested questions are as follows:}\\
+  \ad{1.}Can the output be used by other software as input?(\{yes$^*$, no\})\\
+  \ad{2.}Can it take output from other software as input?(\{yes$^*$, no\})\\
+  \ad{3.}Does the software generate output files with uncommon formats or extensions? Such as the ones that can only be used by itself.(\{yes$^*$, no, n/a\})\\
+  \ad{4.}Can the software work with customized plug-ins? (\{yes, no\})\\
   Overall impression? (\{1 .. 10\})\\
@@
 
-  Is the development process defined? If yes, what process is used. (\{yes$^*$, no, n/a\})\\
+  Is the development process defined? If yes, what process is used. (\{yes$^*$, no, n/a\})\ad{Seems like a good question to put in the questionnaire}\\
   Ease of external examination relative to other products
   considered?  (\{1 .. 10\})\\
+  \ad{Some more suggested questions are as follows:}\\
+  \ad{1.}Are there any documents recording the development process and status? (\{yes$^*$, no,\})\\
+  \ad{2.}Are there release notes? (\{yes, no\})\\
+  \ad{3.}In the issue tracking system, are there at least some issues assigned to specific developers?
+(\{yes, no\})\ad{It would be better to get a percentage, but the empirical tool can't achieve this.
+It might be worth to manually count the most recent 50 issues.}\\
+  \ad{4.}In the issue tracking system, are there at least some issues with tags of types and priorities?
+(\{yes, no\})\ad{In GitHub, tags are optional.
+In BitBucket, it seems to be mandatory.
+In SourceForge, it seems to be not supported.}\\
+  \ad{5.}Are there any project management tools used during the development?\ad{Seems like a good question to put in the questionnaire}\\
   Overall impression? (\{1 .. 10\})\\

### 2020-04-15 StateOfPractice/Methodology/GradingTemplate.tex
Hard wrap with to 80

@@
 The table below lists the full set of measures that are assessed for each
-software product.  The measures are grouped under headings for each quality, and
-one for summary information.  Following each measure, the type for a valid
+software product. The measures are grouped under headings for each quality,
+an done for summary information. Following each measure, the type for a valid
 result is given in brackets.  Many of the types are given as enumerated sets.
 For instance, the response on many of the questions is one of ``yes,'' ``no,''
-or ``unclear.''  The type ``number'' means natural number, a positive integer.  The
+or ``unclear.'' The type ``number'' means natural number, a positive integer.
+The
 types for date and url are not explicitly defined, but they are what one would
-expect from their names.  In some cases the response for a given question is not
-necessarily limited to one answer, such as the question on what platforms are
+expect from their names. In some cases the response for a given question is
+not necessarily limited to one answer, such as the question on what platforms are
 supported by the software product.  Case like this are indicated by ``set of''
 preceding the type of an individual answer.  The type in these cases are then
-the power set of the individual response type.  In some cases a superscript $^*$
-is used to indicate that a response of this type should be accompanied by
+the power set of the individual response type. In some cases a superscript
+$^*$is used to indicate that a response of this type should be accompanied by
 explanatory text.  For instance, if problems were caused by uninstall, the
@@
   Software purpose (string)\\
-  Number of developers (number)\ad{It might be hard to decide.
-For example 3D Slicer has 171 people with commits to its repo, but on its website, 18 contributors(including people who are not developers) are acknowledged.}\\
+Number of developers (number)\ad{It might be hard to decide. For example 3D
+Slicer has 171 people with commits to its repo, but on its website, 18
+contributors(including people who are not developers) are acknowledged.}\\
   How is the project funded (string)\\
@@
   Publications about the software (set of url) \\
-  Is source code available?
-(\{yes, no\})\ad{It might be better to directly ask for a set of url, and the answer can be n/a if it's not open source, or unclear if it can't be determined.
-An example is as follows.}\\
+Is source code available? (\{yes, no\})\ad{It might be better to directly ask
+for a set of url, and the answer can be n/a if it's not open source, or unclear
+if it can't be determined. An example is as follows.}\\
   \ad{Example}Source code URL (\{set of url, n/a, unclear\})\\
@@
   Is there something in place to automate the installation? (\{yes$^*$, no\})\\
-  Is there a specified way to validate the installation, such as a test suite? \ad{What's the difference between this one and the getting started tutorial?} (\{yes$^*$, no\})\\
-  How many steps were involved in the installation? (number) \ad{what if the installation has different steps on different OS?}\\
-  How many software packages need to be installed before or during installation?
-  (number) \ad{should it be 0 if no packages needed? I saw some answers with "no" or "none"}\\
-  Run uninstall, if available. Were any obvious problems caused? (\{unavail, yes$^*$, no\})\\
+Is there a specified way to validate the installation, such as a test suite?
+\ad{What's the difference between this one and the getting started tutorial?}
+(\{yes$^*$, no\})\\
+How many steps were involved in the installation? (number) \ad{what if the
+installation has different steps on different OS?}\\
+How many software packages need to be installed before or during
+installation?(number) \ad{should it be 0 if no packages needed? I saw some
+answers with "no"
+or "none"}\\
+Run uninstall, if available. Were any obvious problems caused? (\{unavail,
+yes$^*$, no\})\\
   Overall impression? (\{1 .. 10\})\\
@@
   Any reference to the requirements specifications of the program?
-  (\{yes$^*$, no, unclear\})\ad{Seems like a good question to put in the questionnaire}\\
-  What tools or techniques are used to build confidence of correctness? (string)\ad{Seems like a good question to put in the questionnaire}\\
-  If there is a getting started tutorial, is the output as expected? (\{yes, no$^*$, n/a\})\\
+(\{yes$^*$, no, unclear\})\ad{Seems like a good question to put in the
+questionnaire}\\
+What tools or techniques are used to build confidence of correctness?
+(string)\ad{Seems like a good question to put in the questionnaire}\\
+If there is a getting started tutorial, is the output as expected? (\{yes,
+no$^*$, n/a\})\\
   Overall impression? (\{1 .. 10\})\\
@@
   Did the software ``break'' during installation? (\{yes$^*$, no\})\\
-  Did the software ``break'' during the initial tutorial testing? (\{yes$^*$, no, n/a\})\\
+Did the software ``break'' during the initial tutorial testing? (\{yes$^*$, no,
+n/a\})\\
   Overall impression? (\{1 .. 10\})\\
@@
 
-  Does the software handle garbage input reasonably? (\{yes, no$^*$\}) \ad{How do we prepare garbage input?}\\
+Does the software handle garbage input reasonably? (\{yes, no$^*$\}) \ad{How do
+we prepare garbage input?}\\
   For any plain text input files, if all new lines are replaced with new lines
@@
   Are expected user characteristics documented? (\{yes, no\})\\
-  What is the user support model? (string)\ad{Can GitHub be one? Many users directly ask questions there by open new issues.}\\
+What is the user support model? (string)\ad{Can GitHub be one? Many users
+directly ask questions there by open new issues.}\\
   Overall impression? (\{1 .. 10\})\\
@@
 
-  Is there a history of multiple versions of the software?  (\{yes, no, unclear\})\\
+Is there a history of multiple versions of the software? (\{yes, no,
+unclear\})\\
   Is there any information on how code is reviewed, or how to contribute?
@@
   Is there a changelog?  (\{yes, no\})\\
-  What is the maintenance type? (set of \{corrective, adaptive, perfective, unclear\})\\
+What is the maintenance type? (set of \{corrective, adaptive, perfective,
+unclear\})\\
   What issue tracking tool is employed? (set of \{Trac, JIRA, Redmine, e-mail,
-  discussion board, sourceforge, google code, git, none, unclear\}) \ad{Does git mean GitHub and GitLab? Should we add BitBucket here?}\\
-  Are the majority of identified bugs fixed? (\{yes, no$^*$, unclear\})\ad{It might be better to record a number here. An example is as follows.}\\
-  \ad{Example}What is the percentage of identified issues that are closed? (percentage)\\
+discussion board, sourceforge, google code, git, none, unclear\}) \ad{Does git
+mean GitHub and GitLab? Should we add BitBucket here?}\\
+Are the majority of identified bugs fixed? (\{yes, no$^*$, unclear\})\ad{It
+might be better to record a number here. An example is as follows.}\\
+\ad{Example}What is the percentage of identified issues that are closed?
+(percentage)\\
 Which version control system is in use? (\{svn, cvs, git, github, unclear\})\\
@@
   (\{yes$^*$, no\})\\
-  Are there code clones? (\{yes$^*$, no, unclear\})\ad{Might need to test some static analyzer.}\\
+Are there code clones? (\{yes$^*$, no, unclear\})\ad{Might need to test some
+static analyzer.}\\
   Overall impression? (\{1 .. 10\})\\
@@
 
-  Are any portions of the software used by another package? (\{yes$^*$, no\})\ad{It seems hard to find an answer for some software.}\\
+Are any portions of the software used by another package? (\{yes$^*$,
+no\})\ad{It seems hard to find an answer for some software.}\\
   Is there evidence that reusability was considered in the design? (API
@@
   Is portability explicitly identified as NOT being important? (\{yes, no\})\\
-  Convincing evidence that portability has been achieved? (\{yes$^*$, no\})\ad{Seems like a good question to put in the questionnaire}\\
+Convincing evidence that portability has been achieved? (\{yes$^*$,
+no\})\ad{Seems like a good question to put in the questionnaire}\\
   Overall impression? (\{1 .. 10\})\\
@@
   meaningful? (\{yes, no$^*$, n/a\})\\
-  Are constants (other than 0 and 1) hard coded into the program? (\{yes, no$^*$, n/a\})\\
-  Comments are clear, indicate what is being done, not how? (\{yes, no$^*$, n/a\})\\
+Are constants (other than 0 and 1) hard coded into the program? (\{yes, no$^*$,
+n/a\})\\
+Comments are clear, indicate what is being done, not how? (\{yes, no$^*$,
+n/a\})\\
   Is the name/URL of any algorithms used mentioned?
@@
   Descriptive names of source code files? (\{yes, no$^*$, n/a\})\\
-  Is a design document provided? (\{yes$^*$, no, n/a\})\ad{What is a design document? There can be some examples.}\\
+Is a design document provided? (\{yes$^*$, no, n/a\})\ad{What is a design
+document? There can be some examples.}\\
   Overall impression? (\{1 .. 10\})\\
@@
   Is there a workflow that uses other softwares? (\{yes$^*$, no\})\\
-  If there are external interactions, is the API clearly defined? (\{yes$^*$, no, n/a\})\\
+If there are external interactions, is the API clearly defined? (\{yes$^*$, no,
+n/a\})\\
   \ad{Some more suggested questions are as follows:}\\
@@
   \ad{2.}Can it take output from other software as input?(\{yes$^*$, no\})\\
-  \ad{3.}Does the software generate output files with uncommon formats or extensions? Such as the ones that can only be used by itself.(\{yes$^*$, no, n/a\})\\
+\ad{3.}Does the software generate output files with uncommon formats or
+extensions? Such as the ones that can only be used by itself.(\{yes$^*$, no,
+n/a\})\\
   \ad{4.}Can the software work with customized plug-ins? (\{yes, no\})\\
@@
 
-  Is the development process defined? If yes, what process is used. (\{yes$^*$, no, n/a\})\ad{Seems like a good question to put in the questionnaire}\\
+Is the development process defined? If yes, what process is used. (\{yes$^*$,
+no, n/a\})\ad{Seems like a good question to put in the questionnaire}\\
   Ease of external examination relative to other products
@@
   \ad{Some more suggested questions are as follows:}\\
-  \ad{1.}Are there any documents recording the development process and status? (\{yes$^*$, no,\})\\
+\ad{1.}Are there any documents recording the development process and status?
+(\{yes$^*$, no,\})\\
   \ad{2.}Are there release notes? (\{yes, no\})\\
-  \ad{3.}In the issue tracking system, are there at least some issues assigned to specific developers?
-(\{yes, no\})\ad{It would be better to get a percentage, but the empirical tool can't achieve this.
-It might be worth to manually count the most recent 50 issues.}\\
-  \ad{4.}In the issue tracking system, are there at least some issues with tags of types and priorities?
-(\{yes, no\})\ad{In GitHub, tags are optional.
-In BitBucket, it seems to be mandatory.
-In SourceForge, it seems to be not supported.}\\
-  \ad{5.}Are there any project management tools used during the development?\ad{Seems like a good question to put in the questionnaire}\\
+\ad{3.}In the issue tracking system, are there at least some issues assigned to
+specific developers? (\{yes, no\})\ad{It would be better to get a percentage,
+but the empirical tool can't achieve this. It might be worth to manually count
+the most recent 50 issues.}\\
+\ad{4.}In the issue tracking system, are there at least some issues with tags of 
+types and priorities? (\{yes, no\})\ad{In GitHub, tags are optional. In
+BitBucket, it seems to be mandatory. In SourceForge, it seems to be not
+supported.}\\
+\ad{5.}Are there any project management tools used during the
+development?\ad{Seems like a good question to put in the questionnaire}\\
   Overall impression? (\{1 .. 10\})\\

### 2020-04-15 StateOfPractice/Peter-Notes/GradingTemplate.tex
comments on old template

--- 2020-04-15 StateOfPractice/Methodology/GradingTemplate.tex
@@
   Software purpose (string)\\
-  Number of developers (number)\ad{It might be hard to decide.
-For example 3D Slicer has 171 people with commits to its repo, but on its website, 18 contributors(including people who are not developers) are acknowledged.}\\
+Number of developers (number)\\
   How is the project funded (string)\\
@@
   Publications about the software (set of url) \\
-  Is source code available? (\{yes, no\})\ad{It might be better to directly ask for a set of url, and the answer can be n/a if it's not open source, or unclear if it can't be determined.}\\
-  Source code URL (\{set of url, n/a, unclear\})\ad{example}\\
+  Is source code available? (\{yes, no\})\\
+  Source code URL (\{set of url, n/a, unclear\})\\
   Programming language(s) (set of \{FORTRAN, Matlab, C, \CC, Java, R, Ruby,
@@
   \midrule
-
+ \pmi{Are required (known to work) operating system versions listed?}\\
   Are there installation instructions? (\{yes, no\})\\
@@
   Is there something in place to automate the installation? (\{yes$^*$, no\})\\
-  Is there a specified way to validate the installation, such as a test suite? \ad{What's the difference between this one and the getting started tutorial?} (\{yes$^*$, no\})\\
-  How many steps were involved in the installation? (number) \ad{what if the installation has different steps on different OS?}\\
-  How many software packages need to be installed before or during installation?
-  (number) \ad{should it be 0 if no packages needed? I saw some answers with "no" or "none"}\\
-  Run uninstall, if available. Were any obvious problems caused? (\{unavail, yes$^*$, no\})\\
+Is there a specified way to validate the installation, such as a test suite?
+(\{yes$^*$, no\})\\
+How many steps were involved in the installation? (number) \\
+How many software packages need to be installed before or during
+installation?(number) \\
+\pmi{are specific versions of required packages listed?}\\
+\pmi{are there links to or instructions for the installation of required packages?}\\
+Run uninstall, if available. Were any obvious problems caused? (\{unavail,
+yes$^*$, no\})\\
+\pmi{In your opinion could a non-developer of the software effectively install this software by verbatim following the instructions}\\
+\pmi{Was troubleshooting the instructions necessary to install the system?}
   Overall impression? (\{1 .. 10\})\\
@@
   Are external libraries used? (\{yes$^*$, no, unclear\})\\
+  \pmi{are versions of external libraries specified?}\\
+  \pmi{are instructions for finding/installing the external libraries specified?}\\
   Does the community have confidence in this library? (\{yes, no, unclear\})\\
   Any reference to the requirements specifications of the program?
-  (\{yes$^*$, no, unclear\})\ad{Seems like a good question to put in the questionnaire}\\
-  What tools or techniques are used to build confidence of correctness? (string)\ad{Seems like a good question to put in the questionnaire}\\
-  If there is a getting started tutorial, is the output as expected? (\{yes, no$^*$, n/a\})\\
+(\{yes$^*$, no, unclear\})\\
+What tools or techniques are used to build confidence of correctness?
+(string)\\
+If there is a getting started tutorial, is the output as expected? (\{yes,
+no$^*$, n/a\})\\
+\pmi{Does the getting started tutorial provide an answer to verify against?}\\
   Overall impression? (\{1 .. 10\})\\
@@
   Did the software ``break'' during installation? (\{yes$^*$, no\})\\
-  Did the software ``break'' during the initial tutorial testing? (\{yes$^*$, no, n/a\})\\
+  \pmi{was a helpful error message provided?}
+    \pmi{was the installation recoverable?}
+Did the software ``break'' during the initial tutorial testing? (\{yes$^*$, no,
+n/a\})\\
+  \pmi{was a helpful error message provided?}\\
+  \pmi{was the testing recoverable?}\\
   Overall impression? (\{1 .. 10\})\\
@@
 
-  Does the software handle garbage input reasonably? (\{yes, no$^*$\}) \ad{How do we prepare garbage input?}\\
+Does the software handle garbage input reasonably? (\{yes, no$^*$\}) \\
   For any plain text input files, if all new lines are replaced with new lines
@@
   no$^*$, n/a\})\\
+  \pmi{Does the system react gracefully when required libraries or input files are not provided? Descriptive message?}
   Overall impression? (\{1 .. 10\})\\
@@
   Is there evidence that performance was considered? (\{yes$^*$, no\})\\
+  \pmi{Does the software (optionally) use GPU for processing?}\\
   Overall impression? (\{1 .. 10\})\\
@@
   Is there a standard example that is explained? (\{yes, no\})\\
+  \pmi{Is a solution provided?}\\
   Is there a user manual? (\{yes, no\})\\
@@
   on? (\{yes, no$^*$\})\\
+  \pmi{this question is vague to me}\\
   Are there any features that show a lack of visibility? (\{yes, no$^*$\})\\
   Are expected user characteristics documented? (\{yes, no\})\\
-  What is the user support model? (string)\ad{Can GitHub be one? Many users directly ask questions there by open new issues.}\\
+What is the user support model? (string)\\
   Overall impression? (\{1 .. 10\})\\
@@
   Is there a changelog?  (\{yes, no\})\\
-  What is the maintenance type? (set of \{corrective, adaptive, perfective, unclear\})\\
+What is the maintenance type? (set of \{corrective, adaptive, perfective,
+unclear\})\\
   What issue tracking tool is employed? (set of \{Trac, JIRA, Redmine, e-mail,
-  discussion board, sourceforge, google code, git, none, unclear\}) \ad{Does git mean GitHub and GitLab? Should we add BitBucket here?}\\
-  Are the majority of identified bugs fixed? (\{yes, no$^*$, unclear\})\ad{It might be better to record a number here}\\
-  What is the percentage of identified issues that are closed? (percentage)\ad{an example}\\
+discussion board, sourceforge, google code, git, none, unclear\})\\
+Are the majority of identified bugs fixed? (\{yes, no$^*$, unclear\})\\
+What is the percentage of identified issues that are closed?
+(percentage)\\
+\pmi{is there a way to measure average time to close issues?}\\
+\pmi{are files well commented? - percentage?}\\
+\pmi{comment on module / file size}\\
 Which version control system is in use? (\{svn, cvs, git, github, unclear\})\\
@@
   (\{yes$^*$, no\})\\
-  Are there code clones? (\{yes$^*$, no, unclear\})\ad{Might need to test some static analyzer.}\\
+Are there code clones? (\{yes$^*$, no, unclear\})\\
   Overall impression? (\{1 .. 10\})\\
@@
 
-  Are any portions of the software used by another package? (\{yes$^*$, no\})\ad{It seems hard to find an answer for some software.}\\
+Are any portions of the software used by another package? (\{yes$^*$,
+no\})\\
   Is there evidence that reusability was considered in the design? (API
 documented, web service, command line tools, ...) (\{yes$^*$, no, unclear\})\\
+\pmi{Are module interfaces clearly defined?}\\
+\pmi{Are modules described in the user guide?}\\
+\pmi{Is the system modularized?}\\
 Overall impression? (\{1 .. 10\})\\
@@
   (set of \{Windows, Linux, OS X, Android, Other OS\})\\
+  \pmi{What versions of these platforms is the software advertised to work on?}\\
+  \pmi{How many external libraries are needed?}\\
   Are special steps taken in the source code to handle portability? (\{yes$^*$,
@@
   Is portability explicitly identified as NOT being important? (\{yes, no\})\\
-  Convincing evidence that portability has been achieved? (\{yes$^*$, no\})\\
+Convincing evidence that portability has been achieved? (\{yes$^*$,
+no\})\\
   Overall impression? (\{1 .. 10\})\\
@@
   Is there a workflow that uses other softwares? (\{yes$^*$, no\})\\
-  If there are external interactions, is the API clearly defined? (\{yes$^*$, no, n/a\})\\
-  Overall impression? (\{1 .. 10\})\\
+If there are external interactions, is the API clearly defined? (\{yes$^*$, no,
+n/a\})\\
 
@@
 
-  Is the development process defined? If yes, what process is used. (\{yes$^*$, no, n/a\})\\
+Is the development process defined? If yes, what process is used. (\{yes$^*$,
+no, n/a\})\\
   Ease of external examination relative to other products

### 2020-04-15 StateOfPractice/Peter-Notes/TemplateToGradeSCSoft_Comments.tex
comments on old template

--- 2020-04-13 StateOfPractice/Methodology/Methodology.tex
@@
-\documentclass[letterpaper,cleveref]{lipics-v2019}
+\documentclass[12pt]{article}
 
-\usepackage{natbib}
 \usepackage{booktabs}
-\usepackage{amsmath,amsthm}
+\usepackage{longtable}
+\usepackage{xspace}
+\usepackage{fullpage}
+\usepackage{hyperref}
 
-\usepackage{hyperref}
 \hypersetup{
@@
 
-\newcommand{\notdone}[1]{\textcolor{red}{#1}}
-\newcommand{\done}[1]{\textcolor{black}{#1}}
+\def\CC{{C\nolinebreak[4]\hspace{-.05em}\raisebox{.4ex}{\tiny\bf ++}}\xspace}
 
-%\oddsidemargin 0mm
-%\evensidemargin 0mm
-%\textwidth 160mm
-%\textheight 200mm
+\begin{document}
 
-\theoremstyle{definition}
-\newtheorem{defn}{Definition}
-
-\title{Methodology for Assessing the State of the Practice for Domain X} 
-\author{Spencer Smith}{McMaster University, Canada}{smiths@mcmaster.ca}{}{}
-\author{Jacques Carette}{McMaster University, Canada}{carette@mcmaster.ca}{}{}
-\author{Olu Owojaiye}{McMaster University, Canada}{owojaiyo@mcmaster.ca}{}{}
-\author{Peter Michalski}{McMaster University, Canada}{michap@mcmaster.ca}{}{}
-\author{Ao Dong}{McMaster University, Canada}{donga9@mcmaster.ca}{}{}
-
-\authorrunning{Smith et al.}  \Copyright{Spencer Smith and Jacques Carette and
-Olu Owojaiye and Peter Michalski and Ao Dong}
-
+\title{Template for Grading Scientific Computing Software} 
+\author{Spencer Smith, Adam Lazzarato, Yue Sun, Zheng Zeng,\\ Vasudha Kapil and Jacques Carette}
 \date{\today}
 
-\hideLIPIcs
-\nolinenumbers
-
-\begin{document}
 \maketitle
-
-\begin{abstract}
-	...
-\end{abstract}
-
-\tableofcontents
-
-\section{Introduction} \label{SecIntroduction}
-
-Purpose and scope of the document.  \wss{Needs to be filled in.  Should
-	reference the overall research proposal, and the ``state of the practice''
-	exercise in particular.}
-
-\section{Overview of Steps in Assessing Quality of the Domain Software}
-
-\begin{enumerate}
-\item Identify domain.  (Provide criteria on a candidate domain.)
-\item 
-\end{enumerate}
-
-\section{Identify Candidate Software}
-
-\section{Domain Analysis} \label{SecDomainAnalysis}
-
-Commonality analysis.  Follow as for mesh generator (likely with less detail).
-
-\section{Empirical Measures}
-
-\subsection{Raw Data}
-Measures that can be extracted from on-line repos.
-
-\ad{Still at brainstorm stage.}
-\begin{itemize}
-\item number of contributors
-\item number of watches
-\item number of stars
-\item number of forks
-\item number of clones
-\item number of commits
-\item number of total/code/document files
-\item lines of total/logical/comment code
-\item lines/pages of documents (can pdf be extracted?)
-\item number of total/open/closed/merged pull requests
-\item number of total/open/closed issues
-\item number of total/open/closed issues with assignees
-\end{itemize}
-
-Instead of only focus on the current status of the above numbers, we may find
-the time history of them to be more valuable. For example, the number of
-contributors over time, the number of lines of code over time, the number of
-open issues over time, etc.
-
-\subsection{Processed Data}
-Metrics that can be calculated from the raw data.
-
-\ad{Still at brainstorm stage.}
-\begin{itemize}
-\item percentage of total/open/closed issues with assignees -
-Visibility/Transparency
-\item lines of new code produced per person-month - Productivity
-\item lines/pages of new documents produced per person-month - Productivity
-\item number of issues closed per person-month - Productivity
-\item percentage of comment lines in the code - maintainability \ad{Not Ao's
-qualities}
-\end{itemize}
-
-In the above calculations, a month can be determined to be 30 days.
-
-\subsection{Tool Tests}
-\ad{This section is currently a note of unorganized contents. Most parts will beremoved or relocated.}
-
-\ad{This citation needs to be deleted later. It's here because my compiler
-doesn't work with 0 citations}
-\cite{Emms2019}
-
-Most tests were done targeting to the repo of 3D Slicer
-\href{https://github.com/tomgi/git_stats}{GitHub repo}
-
-\subsubsection{git-stats}
-\href{https://github.com/tomgi/git_stats}{GitHub repo}
-
-Test results:
-\href{http://git-stats-slicer.ao9.io/}{http://git-stats-slicer.ao9.io/} the
-results are output as webpages, so I hosted for you to check. Data can be
-downloaded as spreadsheets.
-
-\subsubsection{git-of-theseus}
-\href{https://github.com/erikbern/git-of-theseus}{GitHub repo}
-
-Test results: It took about 100 minutes for one repo on a 8 core 16G ram Linux
-machine. It only outputs graphs.
-
-\subsubsection{hercules}
-\href{https://github.com/src-d/hercules}{GitHub repo}
-
-Test results: this one seems to be promising, but the installation is
-complicated with various errors.
-
-\subsubsection{git-repo-analysis}
-\href{https://github.com/larsxschneider/git-repo-analysis}{GitHub repo}
-
-\subsubsection{HubListener}
-\href{https://github.com/pjmc-oliveira/HubListener}{GitHub repo}
-
-The data that HubListener can extract.
-
-Raw:
-\begin{itemize}
-\item Number of Files
-\item Number of Lines
-\item Number of Logical Lines
-\item Number of Comments
-\end{itemize}
-
-Cyclomatic:
-\href{https://www.geeksforgeeks.org/cyclomatic-complexity/}{Intro}
-\begin{itemize}
-\item Cyclomatic Complexity
-\end{itemize}
-
-Halstead:
-\href{https://www.geeksforgeeks.org/software-engineering-halsteads-software-metrics/}{Intro}
-\begin{itemize}
-\item Halstead Effort
-\item Halstead Bugs
-\item Halstead Length
-\item Halstead Difficulty
-\item Halstead Time
-\item Halstead Vocabulary
-\item Halstead Volume
-\end{itemize}
-
-Test results: HubListener works well on the repo of itself, but it did not work
-well on some other repos.
-
-\subsubsection{gitinspector}
-\href{https://github.com/ejwa/gitinspector}{GitHub repo}
-
-Test results: it doesn't work well. Instead of creating output results, it
-prints the results directly in the console.
-
-\section{User Experiments}
-
-Describe experiments with users to assess usability, performance etc.
-
-\section{Analytic Hierarchy Process}
-
-Describe process.  Domain expert review.
-
-\section{Quality Specific Measures}
-
-\subsection{\notdone{Installability} \oo{owner}}
-
-\subsection{\notdone{Correctness} \oo{owner}}
-
-\subsection{\notdone{Verifiability/Testability} \oo{owner}}
-
-\subsection{\notdone{Validatability} \oo{owner}}
-
-\subsection{\notdone{Reliability} \oo{owner}}
-
-\subsection{\notdone{Robustness} \pmi{owner}}
-
-\subsection{\notdone{Performance} \pmi{owner}}
-
-\subsection{\notdone{Usability} \jc{owner}}
-
-\subsection{\notdone{Maintainability} \pmi{owner}}
-
-\subsection{\notdone{Reusability} \pmi{owner}}
-
-\subsection{\notdone{Portability} \pmi{owner}}
-
-\subsection{\notdone{Understandability} \jc{owner}}
-
-\subsection{\notdone{Interoperability} \ad{owner}}
-
-\subsection{\notdone{Visibility/Transparency} \ad{owner}}
-
-\subsection{\notdone{Reproducibility} \wss{owner}}
-
-\subsection{\notdone{Productivity} \ad{owner}}
-
-\subsection{\notdone{Sustainability} \wss{owner}}
-
-\subsection{\notdone{Completeness} \ad{owner}}
-
-\subsection{\notdone{Consistency} \ad{owner}}
-
-\subsection{\notdone{Modifiability} \jc{owner}}
-
-\subsection{\notdone{Traceability} \jc{owner}}
-
-\subsection{\notdone{Unambiguity} \wss{owner}}
-
-\subsection{\notdone{Verifiability} \wss{owner}}
-
-\subsection{\notdone{Abstract} \wss{owner}}
-
-\section{Using Data to Rank Family Members}
-
-Describe AHP process (or similar).
-
-\newpage
-
-\bibliographystyle {plainnat}
-\bibliography {../../CommonFiles/ResearchProposal}
+\include{GradingTemplate}
 

### 2020-04-16 StateOfPractice/Methodology/Methodology.tex
usability

@@
 \section{User Experiments}
-
+\subsection{Usability Experiment}
+Steps:
+\begin {enumerate}
+\item Write research questions
+\item Prepare survey questions
+\item Select a list of projects
+\item Identify tasks for the study subjects to perform (tasks can be defined based on user category, at least one task to modify the software)
+\item Select participants 
+\item Survey the study subject
+\item Perform tasks
+\item Observe participants (take notes, record sessions)
+\item Perform pairwise comparison analysis
+\item Prepare Usability report
+\end {enumerate}
 Describe experiments with users to assess usability, performance etc.

### 2020-04-16 StateOfPractice/Methodology/Methodology.tex
usability

--- 2020-04-13 StateOfPractice/Methodology/Methodology.tex
@@
 \section{User Experiments}
+\subsection{Usability Experiment}
+Steps:
+\begin {enumerate}
+\item Write research questions
+\item Prepare logistic details(location, time table, recording setup, moderators etc)
+\item Prepare survey questions
+\item Select a list of projects
+\item Design the tasks for the study subjects to perform (tasks can be defined based on user category, at least one task to modify the software)
+\item Select participants
+\item Conduct usability session(s) 
+\item Survey the study subject to collect pre-experiment data
+\item Perform tasks
+\item Observe the study subjects (take notes, record sessions, watch out for body language and verbal cues)
+\item Survey the study subjects to collect final feedback
+\item Prepare experiment report
+\item Perform pairwise comparison analysis
+\item Prepare analysis report
+\end {enumerate}
 
+- Pairwise Comparison Methodology needs to be described
+- Downloaded OBS to record experiment sessions, will test it's capability
 Describe experiments with users to assess usability, performance etc.

### 2020-04-20 StateOfPractice/Ao-Notes/Questionnaire.tex
create questionnaire

--- 2020-04-15 StateOfPractice/Peter-Notes/TemplateToGradeSCSoft_Comments.tex
@@
 
-\def\CC{{C\nolinebreak[4]\hspace{-.05em}\raisebox{.4ex}{\tiny\bf ++}}\xspace}
+\title{Questionnaire to Developers} 
+\author{Ao Dong}
+\date{\today}
 
@@
 
-\title{Template for Grading Scientific Computing Software} 
-\author{Spencer Smith, Adam Lazzarato, Yue Sun, Zheng Zeng,\\ Vasudha Kapil and Jacques Carette}
-\date{\today}
+\maketitle
 
-\maketitle
-\include{GradingTemplate}
+\begin{enumerate}
+\item \ad{Visibility/Transparency} Is there a certain type of development methodologies used during the development? (\{Waterfall, Scrum, Kanban, else\})
+\item \ad{Visibility/Transparency} Is there a clearly defined development process? If yes, what is it. (\{yes$^*$, no\})
+\item \ad{Visibility/Transparency} Are there any project management tools used during the development? If yes, what are they. (\{yes$^*$, no\})
+\item \ad{Correctness and Verifiability} During the process of development, what tools or techniques are used to build confidence of correctness? (string)
+\item \ad{Correctness and Verifiability} Is there any document about the requirements specifications of the program? If yes, what is it. (\{yes$^*$, no\})
+\item \ad{Portability} Do you think that portability has been achieved? If yes, how? (\{yes$^*$, no\})
+\item \ad{Maintainability} How was maintainability considered in the design? (string)
+\item \ad{Maintainability} What is the maintenance type? (set of \{corrective, adaptive, perfective,
+unclear\})
+\item \ad{Reusability} How was reusability considered in the design? (string)
+\item \ad{Reusability} Are any portions of the software used by another package? If yes, how they are used. ({yes$^*$, no})
+\item \ad{Completeness} Is any of the following documents used during the development? ({yes$^*$, no})
+\begin{itemize}
+\item Module Guide
+\item Module Interface Specification
+\item System Verification and Validation Plan
+\item Unit Verification and Validation Plan
+\end{itemize}
+\end{enumerate}
 

### 2020-04-20 StateOfPractice/Ao-Notes/Questionnaire.tex
update questionnaire

@@
 \maketitle
+Several questions are borrowed from \href{https://gitlab.cas.mcmaster.ca/smiths/pub/-/blob/master/Jegatheesan2016.pdf}{Jegatheesan2016}
+
+\section{Information about the developers and users}
+\begin{enumerate}
+\item Interviewees' current position/title? degrees?
+\item Interviewees' contribution to/relationship with the software?
+\item Length of time the interviewee has been involved with this software?
+\item How large is the development group?
+\item What is the typical background of a developer?
+\item How large is the user group?
+\item What is the typical background of a user?
+\end{enumerate}
+
+\section{Information about the software}
 
 \begin{enumerate}
+\item \ad{General} What is the most important software quality(ies) to your work? (set of selected qualities plus "else")
+\item \ad{General} Are there any examples where the documentation helped? If yes, how it helped. ({yes$^*$, no})
+\item \ad{General} Is there any documentation you feel you should produce and do not? If yes, what is it and why? ({yes$^*$, no})
+\item \ad{Completeness} Do you address any of your quality concerns using documentation? If yes, what are the qualities and the documents. ({yes$^*$, no})
 \item \ad{Visibility/Transparency} Is there a certain type of development methodologies used during the development? (\{Waterfall, Scrum, Kanban, else\})
@@
 \item \ad{Visibility/Transparency} Are there any project management tools used during the development? If yes, what are they. (\{yes$^*$, no\})
+\item \ad{Visibility/Transparency} Going forward, will your approach to documentation of requirements and design
+change? If not, why not. (\{yes, no$^*$\})
 \item \ad{Correctness and Verifiability} During the process of development, what tools or techniques are used to build confidence of correctness? (string)
+\item \ad{Correctness and Verifiability} Do you use any tools to support testing? If yes, what are they. (e.g. unit testing tools, regression testing suites) (\{yes$^*$, no\})
 \item \ad{Correctness and Verifiability} Is there any document about the requirements specifications of the program? If yes, what is it. (\{yes$^*$, no\})
@@
 \item \ad{Reusability} Are any portions of the software used by another package? If yes, how they are used. ({yes$^*$, no})
+\item \ad{Reproducibility} Is reproducibility important to you? ({yes$^*$, no})
+\item \ad{Reproducibility} Do you use tools to help reproduce previous software results? If yes, what are they. (e.g. version control, configuration management) ({yes$^*$, no})
 \item \ad{Completeness} Is any of the following documents used during the development? ({yes$^*$, no})
+\item \ad{General} Will this experience influence how you develop software? Do you see yourself maintaining the same level of documentation, tool support as you go forward? (string)
 \begin{itemize}
@@
 \item Module Interface Specification
-\item System Verification and Validation Plan
-\item Unit Verification and Validation Plan
+\item Verification and Validation Plan
+\item Verification and Validation Report
 \end{itemize}

### 2020-04-20 StateOfPractice/Methodology/Methodology.tex
update survey into this Methodology

@@
 
+\appendix
+\section{Appendix}
+\subsection{Survey for the Selected Projects}
+\ad{Several questions are borrowed from \href{https://gitlab.cas.mcmaster.ca/smiths/pub/-/blob/master/Jegatheesan2016.pdf}{Jegatheesan2016}, and needed to be cited later.}
+\subsubsection{Information about the developers and users}
+\begin{enumerate}
+\item Interviewees' current position/title? degrees?
+\item Interviewees' contribution to/relationship with the software?
+\item Length of time the interviewee has been involved with this software?
+\item How large is the development group?
+\item What is the typical background of a developer?
+\item How large is the user group?
+\item What is the typical background of a user?
+\end{enumerate}
+
+\subsubsection{Information about the software}
+
+\begin{enumerate}
+\item \ad{General} What is the most important software quality(ies) to your work? (set of selected qualities plus "else")
+\item \ad{General} Are there any examples where the documentation helped? If yes, how it helped. ({yes$^*$, no})
+\item \ad{General} Is there any documentation you feel you should produce and do not? If yes, what is it and why? ({yes$^*$, no})
+\item \ad{Completeness} Do you address any of your quality concerns using documentation? If yes, what are the qualities and the documents. ({yes$^*$, no})
+\item \ad{Visibility/Transparency} Is there a certain type of development methodologies used during the development? (\{Waterfall, Scrum, Kanban, else\})
+\item \ad{Visibility/Transparency} Is there a clearly defined development process? If yes, what is it. (\{yes$^*$, no\})
+\item \ad{Visibility/Transparency} Are there any project management tools used during the development? If yes, what are they. (\{yes$^*$, no\})
+\item \ad{Visibility/Transparency} Going forward, will your approach to documentation of requirements and design
+change? If not, why not. (\{yes, no$^*$\})
+\item \ad{Correctness and Verifiability} During the process of development, what tools or techniques are used to build confidence of correctness? (string)
+\item \ad{Correctness and Verifiability} Do you use any tools to support testing? If yes, what are they. (e.g. unit testing tools, regression testing suites) (\{yes$^*$, no\})
+\item \ad{Correctness and Verifiability} Is there any document about the requirements specifications of the program? If yes, what is it. (\{yes$^*$, no\})
+\item \ad{Portability} Do you think that portability has been achieved? If yes, how? (\{yes$^*$, no\})
+\item \ad{Maintainability} How was maintainability considered in the design? (string)
+\item \ad{Maintainability} What is the maintenance type? (set of \{corrective, adaptive, perfective,
+unclear\})
+\item \ad{Reusability} How was reusability considered in the design? (string)
+\item \ad{Reusability} Are any portions of the software used by another package? If yes, how they are used. ({yes$^*$, no})
+\item \ad{Reproducibility} Is reproducibility important to you? ({yes$^*$, no})
+\item \ad{Reproducibility} Do you use tools to help reproduce previous software results? If yes, what are they. (e.g. version control, configuration management) ({yes$^*$, no})
+\item \ad{Completeness} Is any of the following documents used during the development? ({yes$^*$, no})
+\item \ad{General} Will this experience influence how you develop software? Do you see yourself maintaining the same level of documentation, tool support as you go forward? (string)
+\begin{itemize}
+\item Module Guide
+\item Module Interface Specification
+\item Verification and Validation Plan
+\item Verification and Validation Report
+\end{itemize}
+\end{enumerate}
+
 \newpage

### 2020-04-20 StateOfPractice/Methodology/Methodology.tex
updated .bib, methodology file

--- 2020-04-16 StateOfPractice/Methodology/Methodology.tex
@@
 
-- Pairwise Comparison Methodology needs to be described
-- Downloaded OBS to record experiment sessions, will test it's capability
-Describe experiments with users to assess usability, performance etc.
+- Pairwise Comparison Methodology needs to be described\\
+- Downloaded OBS to record experiment sessions, will test it's capability\\
+- Describe experiments with users to assess usability, performance etc.
 

### 2020-04-21 StateOfPractice/Peter-Notes/GradingTemplate.tex

@@
   \midrule
- \pmi{Are required (known to work) operating system versions listed?}\\
+ \pmi{Are required (known to work) operating system versions listed? (\{yes, no\})}\\
   Are there installation instructions? (\{yes, no\})\\
@@
 installation?(number) \\
-\pmi{are specific versions of required packages listed?}\\
-\pmi{are there links to or instructions for the installation of required packages?}\\
+\pmi{are specific versions of required packages listed? (\{yes, no, n/a\})}\\
+\pmi{are there links to or instructions for the installation of required packages? (\{yes, no, n/a\})}\\
 Run uninstall, if available. Were any obvious problems caused? (\{unavail,
 yes$^*$, no\})\\
-\pmi{In your opinion could a non-developer of the software effectively install this software by verbatim following the instructions}\\
-\pmi{Was troubleshooting the instructions necessary to install the system?}
+\pmi{In your opinion could a non-developer of the software effectively install this software by verbatim following the instructions (\{yes, no, n/a\})}\\
+\pmi{Was troubleshooting the instructions necessary to install the system? (\{yes, no, n/a\})}
   Overall impression? (\{1 .. 10\})\\
@@
   Are external libraries used? (\{yes$^*$, no, unclear\})\\
-  \pmi{are versions of external libraries specified?}\\
-  \pmi{are instructions for finding/installing the external libraries specified?}\\
+  \pmi{are versions of external libraries specified? (\{yes, no, n/a\})}\\
   Does the community have confidence in this library? (\{yes, no, unclear\})\\
@@
 no$^*$, n/a\})\\
-\pmi{Does the getting started tutorial provide an answer to verify against?}\\
+\pmi{Does the getting started tutorial provide an answer to verify against? (\{yes, no, n/a\})}\\
   Overall impression? (\{1 .. 10\})\\
@@
   Did the software ``break'' during installation? (\{yes$^*$, no\})\\
-  \pmi{was a helpful error message provided?}
-    \pmi{was the installation recoverable?}
+  \pmi{was a helpful error message provided? (\{yes, no, n/a\})}
+    \pmi{was the installation recoverable? (\{yes, no, n/a\})}
 Did the software ``break'' during the initial tutorial testing? (\{yes$^*$, no,
 n/a\})\\
-  \pmi{was a helpful error message provided?}\\
-  \pmi{was the testing recoverable?}\\
+  \pmi{was a helpful error message provided? (\{yes, no, n/a\})}\\
+  \pmi{was the testing recoverable? (\{yes, no, n/a\})}\\
   Overall impression? (\{1 .. 10\})\\
@@
   no$^*$, n/a\})\\
-  \pmi{Does the system react gracefully when required libraries or input files are not provided? Descriptive message?}
   Overall impression? (\{1 .. 10\})\\
@@
   Is there evidence that performance was considered? (\{yes$^*$, no\})\\
-  \pmi{Does the software (optionally) use GPU for processing?}\\
+  \pmi{Does the software (optionally) use GPU for processing? (\{yes, no, unclear\})}\\
   Overall impression? (\{1 .. 10\})\\
@@
   Is there a standard example that is explained? (\{yes, no\})\\
-  \pmi{Is a solution provided?}\\
+  \pmi{Is a solution provided? (\{yes, no, n/a\})}\\
   Is there a user manual? (\{yes, no\})\\
@@
   on? (\{yes, no$^*$\})\\
-  \pmi{this question is vague to me}\\
   Are there any features that show a lack of visibility? (\{yes, no$^*$\})\\
@@
   Is there a changelog?  (\{yes, no\})\\
+ \pmi{Is technical documentation (requirements, architecture, code, algorithms, interfaces, APIs) provided? (\{yes, no, unclear\})}\\ 
 What is the maintenance type? (set of \{corrective, adaptive, perfective,
@@
 (percentage)\\
-\pmi{is there a way to measure average time to close issues?}\\
-\pmi{are files well commented? - percentage?}\\
-\pmi{comment on module / file size}\\
+\pmi{Are files well commented? (\{yes, no, unclear\})}\\
+\pmi{Are files of reasonable length? (\{yes, no, unclear\})}\\
 Which version control system is in use? (\{svn, cvs, git, github, unclear\})\\
@@
 documented, web service, command line tools, ...) (\{yes$^*$, no, unclear\})\\
-\pmi{Are module interfaces clearly defined?}\\
-\pmi{Are modules described in the user guide?}\\
-\pmi{Is the system modularized?}\\
+\pmi{Is the system modularized? (\{yes, no, unclear\})}\\
+\pmi{Are module interfaces clearly defined? (\{yes, no, n/a\})}\\
+\pmi{Are modules described in the user guide? (\{yes, no, n/a\})}\\
 Overall impression? (\{1 .. 10\})\\
@@
   (set of \{Windows, Linux, OS X, Android, Other OS\})\\
-  \pmi{What versions of these platforms is the software advertised to work on?}\\
-  \pmi{How many external libraries are needed?}\\
+  \pmi{Are versions of these platforms specified? (\{yes,
+  	no\})}\\
+  \pmi{How many external libraries are needed to be manually installed? (\{int, unclear\})}\\
   Are special steps taken in the source code to handle portability? (\{yes$^*$,

### 2020-04-21 OverallResearchProposal/ObjectivesAndResearchQuestions.tex
update with comments

@@
 and/or a library.  The outcome of the development process is not just software.
-It is \{software + artifacts\}.  \wss{softwartifacts?} The artifacts consist of
+It is \{software + artifacts\}.  \wss{softwartifacts?} \ad{try softifacts?}The artifacts consist of
 such work products as documentation, build scripts, code, test cases,
@@
 looking at the software alone, or the code alone; it depends on everything,
-including the processes and methodologies used.  Further information on the
+including the processes and methodologies used \ad{during the development?}.  Further information on the
 relevant definitions and context is given in the next section.
@@
 alternative that costs less time and energy, assuming that this alternative
-achieves the same level of sustainability as its competition.  The assumptions
+achieves the same \ad{or higher} level of sustainability as its competition.  The assumptions
 listed in the next section outline the typical space of problems where it makes
@@
     proper definition, with citation(s)}
+
+``Awareness or familiarity gained by experience of a fact or situation.'' \citep{OxfordKnowledge2020}
+\ad{Definition from Lexico.com (Oxford) dictionary}
+
 \item[Process] Organization of the software development activities.  \wss{needs a
     proper definition, with citation(s)}
+
+``A series of actions or steps taken in order to achieve a particular end.''\citep{OxfordProcess2020}
+\ad{Definition from Lexico.com (Oxford) dictionary}
+
 \item[Principles] Principles are ``general and abstract statements describing
@@
 \wss{Should performance be included in the list of qualities?}
+\ad{Almost every SCS needs to be correct, reliable and usable, but many of them
+may not need to provide edge performance. How do we choose which qualities to be
+included here?}
 
@@
     proper definition, with citation(s)}
+
+``Application software is a program or group of programs designed for end users.
+These programs are divided into two classes: system software and application
+software. While system software consists of low-level programs that interact
+with computers at a basic level, application software resides above system
+software and includes applications such as database programs, word processors
+and spreadsheets. Application software may be bundled with system software or
+published alone.''\citep{TechopediaApplication2018}\ad{Definition from
+Techopedia.com}
+
+``A program or piece of software designed and written to fulfill a particular
+purpose of the user.''\citep{OxfordApplication2020}\ad{Definition from
+Lexico.com (Oxford) dictionary}
+
 \item[Library] Services that are available to other programs, but not an
@@
     proper definition, with citation(s)}
+
+``A software library is a suite of data and programming code that is used to
+develop software programs and applications. It is designed to assist both the
+programmer and the programming language compiler in building and executing
+software.''\citep{TechopediaLibrary2016}\ad{Definition from Techopedia.com}
+
+``A collection of programs and software packages made generally available, often
+loaded and stored on disk for immediate
+use.''\citep{OxfordLibrary2020}\ad{Definition from Lexico.com (Oxford)
+dictionary}
+
 \item[Artifacts] Work products generated during the process of creating
@@
 \item Oracle Hypothesis – the changes that are likely to occur during the soft-
-  ware’s lifetime are predictable.
+  ware’s \ad{software's} lifetime are predictable.
 \item Organizational Hypothesis – designers can organize the software and the
@@
   between different processes, methodologies and tools?
-\item Can a correlation be identified between software projects that use best
+\item Can a correlation be identified between software projects that use best \ad{better}
   practices for the development process and methodologies and improvement in the
   qualities of usability, performance and reproducibility/replicability?
+\ad{Change "best" to "better"? The best one isn't likely to be found, and better
+practices may also result in improvements.}
 \end{enumerate}

### 2020-04-28 OverallResearchProposal/ObjectivesAndResearchQuestions.tex
Edits to address comments by Ao

@@
 and/or a library.  The outcome of the development process is not just software.
-It is \{software + artifacts\}.  \wss{softwartifacts?} \ad{try softifacts?}The artifacts consist of
-such work products as documentation, build scripts, code, test cases,
-contributor's guides, etc.  The quality of sustainability cannot be assessed by
-looking at the software alone, or the code alone; it depends on everything,
-including the processes and methodologies used \ad{during the development?}.  Further information on the
-relevant definitions and context is given in the next section.
+It is \{software + artifacts\}.  From this point onward, we will use the term
+softifacts when we are referencing the combination of software and the
+associated artifacts.  The artifacts consist of such work products as
+documentation, build scripts, code, test cases, contributor's guides, etc.  The
+quality of sustainability cannot be assessed by looking at the software alone,
+or the code alone; it depends on everything, including the processes and
+methodologies used during the development.  Further information on the relevant
+definitions and context is given in the next section.
 
@@
 alternative that costs less time and energy, assuming that this alternative
-achieves the same \ad{or higher} level of sustainability as its competition.  The assumptions
-listed in the next section outline the typical space of problems where it makes
-sense to aim for sustainable \{software + artifacts\}.
+achieves the same (or a higher) level of sustainability as its competition.  The
+assumptions listed in the next section outline the typical space of problems
+where it makes sense to aim for sustainable softifacts.
 
@@
 (SCS) examples and applications.  This focus is because creating sustainable
-software+artifacts is more feasible when the software domain is well understood,
+softifacts is more feasible when the software domain is well understood,
 which is the case for SCS.  Moreover, SCS is an important, often neglected,
@@
 relevant terms and a list of the assumptions that define the potential contexts
-where sustainable software+artifact development is reasonable.
+where sustainable softifacts development is reasonable.
 
@@
   the specific needs of the present and the future.  The definition can be made
-  less abstract by introducing the context of software+artifacts and thinking
+  less abstract by introducing the context of softifacts and thinking
   separately about the needs of the present and the future.  For the needs of
@@
   environmental) should be possible with a reasonable level of resources.
-  Supporting the future then means producing software+artifacts that are
+  Supporting the future then means producing softifacts that are
   maintainable, portable, reusable and reproducible.
@@
 
-  \emph{Sustainable software means software+artifacts that are correct,
+  \emph{Sustainable software means softifacts that are correct,
     reliable, and usable (for the present) and maintainable, portable, reusable
@@
 \item[Software] Application and/or library.
-\item[Application (App)] An executable program. \wss{needs a
-    proper definition, with citation(s)}
+\item[Application (App)] ``Software designed to fulfill specific needs of a
+  user.'' \citep{IEEEStdGlossarySET1990}
 
@@
 and spreadsheets. Application software may be bundled with system software or
-published alone.''\citep{TechopediaApplication2018}\ad{Definition from
+published alone.''\citep{TechopediaApplication2018} \ad{Definition from
 Techopedia.com}
@@
 
-Producing sustainable software+artifacts is not a trivial undertaking.  Not
+Producing sustainable softifacts is not a trivial undertaking.  Not
 every project needs to aim for sustainability.  Our position is that the time
@@
   least 10 years.
-\item The software+artifacts should interest multiple stakeholders (not just the
+\item The softifacts should interest multiple stakeholders (not just the
   original developer), with different interests and backgrounds.
@@
 
-A good starting point for understanding how to develop sustainable SCS software+artifacts is
+A good starting point for understanding how to develop sustainable SCS softifacts is
 to look at how SCS is currently developed.  By measuring relevant qualities and
@@
 
+Define what is meant by meaningful (what criteria need to be satisfied), since
+this term is used several times in the research questions.
+
 \begin{enumerate}
@@
   could potentially be improved?
-\item For a given software product (software+artifacts), how can we produce a
+\item For a given software product (softifacts), how can we produce a
   reasonable measure of the software qualities (correctness, reliability,
   usability etc.) with a few hours of effort?
-\item For a given software product (software+artifacts), how can we produce a
+\item For a given software product (softifacts), how can we produce a
   meaningful measure of usability with a few days worth of effort?
-\item For a given software product (software+artifacts), how can we produce a
+\item For a given software product (softifacts), how can we produce a
   meaningful measure of maintainability (with respect to likely changes) with a
   few days worth of effort?
-\item For a given software product (software+artifacts), how can we produce a
+\item For a given software product (softifacts), how can we produce a
   meaningful measure of reproducibility/replicability with a few days worth of
@@
   between different processes, methodologies and tools?
-\item Can a correlation be identified between software projects that use best \ad{better}
+\item Can a correlation be identified between software projects that use best 
   practices for the development process and methodologies and improvement in the
   qualities of usability, performance and reproducibility/replicability?
-\ad{Change "best" to "better"? The best one isn't likely to be found, and better
-practices may also result in improvements.}
 \end{enumerate}
@@
 \end{document}
+
+NOTES
+
+- what is needed to produce softifacts sustainably?
+
+- measuring process aimed at 'producing softifacts sustainably'. part of it could be measuring effort.

### 2020-04-30 StateOfPractice/Ao-Notes/Ao-Notes.tex
update notes

@@
 
+\section{Questions to Domain Experts}
+
+\subsection{Software}
+\begin{itemize}
+\item What do you think of our process of selecting software candidates?
+\item What do you think of the list of the selected software?
+\item Is there any software you think should be removed from / added to the
+list?
+\end{itemize}
+
+\subsection{Schedule}
+\begin{itemize}
+\item We propose to start this project from ... and finish it before ... . What
+do you think?
+\item Our meeting schedule
+\end{itemize}
+
+\subsection{Team}
+\begin{itemize}
+\item How many members do you think will be in the team for analyzing this
+domain?
+\item How many hours can a team member commit?
+\item Is there any other experts you think we should keep them involved?
+\end{itemize}
+
+\subsection{Technical}
+\begin{itemize}
+\item Equipment such as computers, test tools, OS
+\item Technical details such as screen recording, questionnaires.
+\end{itemize}
+
 \newpage

### 2020-04-30 StateOfPractice/Methodology/Methodology.tex
update to usability section

--- 2020-04-20 StateOfPractice/Methodology/Methodology.tex
@@
 \item Select participants
-\item Conduct usability session(s) 
+\item Conduct usability session
 \item Survey the study subject to collect pre-experiment data
@@
 
-- Pairwise Comparison Methodology needs to be described
-- Downloaded OBS to record experiment sessions, will test it's capability
-Describe experiments with users to assess usability, performance etc.
+\subsection{Empirical Measures}
+ \begin {enumerate}
+ \item Qualitative/Quantitative
+ \item Pairwise Comparison methodology needs to be described
+ \item Measure time to complete a task and compare actual with expected time of completion
+ \item From a list of given tasks, measure how many tasks were completed correctly
+\end {enumerate}
+
+\subsection{Tools}
+\begin {enumerate} 
+\item Downloaded OBS to record experiment sessions
+\end {enumerate}
+\subsection{To do}
+\begin {enumerate}
+ \item Detail plan
+ \item Describe Pairwise methodology
+ \item Define tasks
+ \item prepare survey questions
+ \end {enumerate}
 

### 2020-05-01 StateOfPractice/Methodology/Methodology.tex
updated usability measures
added questionnaire and task criteria sections

@@
  \item From a list of given tasks, measure how many tasks were completed correctly
+ \item Learnability
+ \item Efficiency
+ \item Memorability
+ \item Errors
+ \item Satisfaction
+ \item Heuristic evaluation: Experts go through the interface to identify elements that violate usability heuristics.Popular because of its low cost and low time commitment
 \end {enumerate}
 
+\subsection{Nielsen’s Heuristics}
+\begin{enumerate}
+	\item Consistency: Check if standards and conventions in product design are followed
+ \begin{itemize}
+ 	\item Sequences of actions (skill acquisition).
+ 	\item Color (categorization).
+ 	\item Layout and position (spatial consistency).
+ 	\item Font, capitalization (levels of organization).
+ 	\item Terminology (delete, del, remove, rm) and language (words, phrases).
+ 	\item Standards (e.g.,	blue underlined text for unvisited hyper links).
+ \end{itemize}
+\item Visibility: Users should be informed about what is going on with the system through appropriate feedback and display of information e.g. 
+ \begin{itemize}
+ \item What is the current state of the system?
+ \item What can be done at current state?
+ \item Where can users go?
+ \item What change is made after an action?
+ \end{itemize}
+\item Match between system and world. The image of the system perceived by users should match the model the users have about the system. 
+\begin {itemize}
+\item User model matches system image.
+\item Actions provided by the system should match actions performed by users.
+\item Objects on the system should match objects of the task.
+\end {itemize}
+\item Minimalist: Any extraneous information is a distraction and a slow-down. 
+\begin{itemize}
+\item Less is more.
+\item Simple is not equivalent to abstract and general.
+\item Simple is efficient.
+\item  Progressive levels of detail.
+\end{itemize}
+\item Minimize memory load. Users should not be required to memorize a lot of information to carry out tasks. Memory load reduces users’ capacity to carry out the main tasks. 
+\begin{itemize}
+\item Recognition vs. recall (e.g., menu vs. commands).
+\item Externalize information through visualization.
+\item Perceptual procedures.
+\item Hierarchical structure.
+\item Default values.
+\item Concrete examples (DD/MM/YY, e.g., 10/20/99).
+\item Generic rules and actions (e.g., drag objects).
+\item Intuitive procedure
+\end{itemize}
+\item Informative feedback. Users should be given prompt and informative feedback about their actions. 
+\begin{itemize}
+\item Information that can be directly perceived, interpreted, and evaluated.
+\item Levels of feedback (novice and expert).
+\item Concrete and specific, not abstract and general.
+\item Response time:\\
+-0.1 s for instantaneously reacting;\\
+-1.0 s for uninterrupted flow of thought;\\
+-10 s for the limit of attention.
+\end{itemize}
+\item Flexibility and efficiency. Users always learn and users are always different. Give users the flexibility of creating customization and shortcuts to accelerate their performance. 
+\begin{itemize}
+	\item Shortcuts for experienced users.
+	\item Shortcuts or macros for frequently used operations.
+	\item Skill acquisition through chunking.
+	\item Examples:
+	- Abbreviations, function keys, hot keys, command keys, macros, aliases, templates, type-ahead, bookmarks, hot links, history, default values, etc.
+\end{itemize}
+\item Good error messages. 
+\begin {itemize}
+\item The messages should be informative enough such that users can understand the nature of errors, learn from errors, and recover from errors. 
+\item Phrased in clear language, avoid obscure codes. Example of obscure code: “system crashed, error code 147.”
+\item Precise, not vague or general. Example of general comment: “Cannot open document.”
+\item Constructive.
+\item Polite. Examples of impolite message: “illegal user action,” “job aborted,” “system was crashed,” “fatal error,” etc.
+\end {itemize}
+\item Prevent errors. It is always better to design interfaces that prevent errors from happening in the first place.
+\begin {itemize} 
+\item Interfaces that make errors impossible.
+\item Avoid modes (e.g., vi, text wrap). Or use informative feedback, e.g., different sounds.
+\item Execution error vs. evaluation error.
+\item Various types of slips and mistakes.
+\end {itemize}
+\item Clear closure. Every task has a beginning and an end. Users should be clearly notified about the completion of a task. 
+\begin {itemize}
+\item Clear beginning, middle, and end.
+\item Complete 7-stages of actions.
+\item Clear feedback to indicate goals are achieved and current stacks of goals can be released. Examples of good closures include many dialogues.
+\end {itemize}
+\item Reversible actions. 
+\begin {itemize}
+\item Users should be allowed to recover from errors. Reversible actions also encourage exploratory learning. 
+\item At different levels: a single action, a subtask, or a complete task.
+\item Multiple steps.
+\item Encourage exploratory learning.
+\item Prevent serious errors.
+\end {itemize}
+\item Use users’ language. The language should be always presented in a form understandable by the intended users.
+\begin{itemize}
+\item Use standard meanings of words.
+\item Specialized language for specialized group.
+\item User defined aliases.
+\item Users’ perspective. Example: “we have bought four tickets for you” (bad) vs. “you bought four tickets” (good).
+\end {itemize}
+\item Help and documentation. Always provide help when needed. 
+\begin {itemize}
+\item Context-sensitive help.
+\item Four types of help.\\
+-task-oriented;\\
+-alphabetically ordered;\\
+-semantically organized;\\
+-search.
+\item Help embedded in contents.
+\end {itemize}
+\item Users in control. Do not give users that impression that they are controlled by the systems. 
+\begin {itemize}
+\item Users are initiators of actors, not responders to actions.
+\item Avoid surprising actions, unexpected outcomes, tedious sequences of actions, etc.
+\end {itemize}
+\end{enumerate}
 \subsection{Tools}
@@
 
+\subsection{Task definition/selection criteria}
+\begin {enumerate}
+\item Tasks based on difficulty levels (Difficult, Medium and easy)
+\item Commonly performed tasks (applicable to all the selected software)
+\item Task that would/should generate errors messages or error codes
+\item Tasks to check standards and conventions
+\item Tasks that should provide users with the system's current state and what actions can be done or where to go or what change was made
+\item Tasks related to what users will do in the real world
+\item End to End tasks that require users to go through sequential or hierarchical steps.
+\item Tasks that allow users to reuse a previously defined value without having to memorize or jot it.
+\item Tasks that should provide users with feedback
+\item Tasks that allow users to customize and use shortcuts, bookmarks, command keys etc.
+\item Tasks to verify if the system is capable of preventing and /or reversing errors
+\item Tasks that might lead the user to use help or documentation
+
+\end {enumerate}
+
+\subsection{Usability Questionnaire}
+This questions need to be reworded to make them as open ended as possible\\
+source \url{https://www.usabilitest.com/sus-pdf-generator}- 20-29
+\begin {enumerate}
+\item Overall, I am satisfied with how easy it is to use this system.
+\item It was simple to use this system.
+\item I could effectively complete the tasks and scenarios using this system.
+\item I was able to complete the tasks and scenarios quickly using this system.
+\item I was able to efficiently complete the tasks and scenarios using this system.
+\item I felt comfortable using this system.
+\item It was easy to learn to use this system.
+\item I believe I could become productive quickly using this system.
+\item The system gave error messages that clearly told me how to fix problems.
+\item Whenever I made a mistake using the system, I could recover easily and quickly.
+\item The information (such as online help, on-screen messages, and other documentation) provided with this system was clear.
+\item It was easy to find the information I needed.
+\item The information provided for the system was easy to understand.
+\item The information was effective in helping me complete the tasks and scenarios.
+\item The organization of information on the system screens was clear.
+\item The interface of this system was pleasant.
+\item I liked using the interface of this system.
+\item This system has all the functions and capabilities I expect it to have.
+\item Overall, I am satisfied with this system.
+\item I think that I would like to use this system frequently.
+\item I found the system unnecessarily complex.
+\item I thought the system was easy to use.
+\item I think that I would need the support of a technical person to be able to use this system.
+\item I found the various functions in this system were well integrated.
+\item I thought there was too much inconsistency in this system.
+\item I would imagine that most people would learn to use this system very quickly.
+\item I found the system very cumbersome to use.
+\item I felt very confident using the system.
+\item I needed to learn a lot of things before I could get going with this system.
+\end {enumerate}
 \section{Analytic Hierarchy Process}

### 2020-05-04 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
clean up usability

@@
 
+\subsection{Additional Definitions}
+
+Some standards give additional definitions for terms that are used a lot.
+These are reproduced here, to aid in the interpretation of the definitions
+below (and their refinement).
+
+\begin{defn} 
+\textsf{Effectiveness}:
+The accuracy and completeness with which users achieve specified
+goals. 
+\cite{ISO9241:11}
+\end{defn}
+
+\begin{defn} 
+\textsf{Efficiency}: 
+The resources expended in relation to the accuracy and
+completeness with which users achieve goals.
+\cite{ISO9241:11}
+\end{defn}
+
+\begin{defn}
+\textsf{Satisfaction}: 
+Freedom from discomfort, and positive attitude to the use of the product.
+\cite{ISO9241:11}
+\end{defn}
+
+\begin{defn}
+\textsf{Context of use}:
+characteristics of the users, tasks and the organizational and physical
+environments 
+\cite{ISO9241:11}
+\end{defn}
+
+\begin{defn}
+\textsf{Goal}: intended outcome
+\cite{ISO9241:11}
+\end{defn}
+
+\begin{defn}
+\textsf{Task}: activities required to achieve a goal 
+\cite{ISO9241:11}
+\end{defn}
+
 \section{Qualities of Software Products, Artifacts and
@@
   specified context of use.
-\end{defn}
-ISO defines usability as
+  \cite{ISO16982:2002} and \cite{ISO9241:11}
+\end{defn}
 
 Nielsen and (separately) Schneidermann have defined usability as part of
-usefulness and
-is composed of:
+usefulness and is composed of:
+\begin{defn}
 \begin{itemize}
@@
 economics sense of the word).
-
-There are two ISO standards covering this, namely ISO/TR 16982:202 and ISO9241.
-
-The Interaction Design Foundation \url{https://www.interaction-design.org/literature/topics/usability} further
-lists the following desirable outcomes:
-
+\cite{NielsenNorman}
+\end{defn}
+
+\begin{defn}
+Desirable outcomes of usability:
 \begin{enumerate}
@@
 using the user interface during the first contact with the website. For
-  example, if a travel agent’s website is a well-designed one, the user should
+example, if a travel agent's website is a well-designed one, the user should
 be able to move through the sequence of actions to book a ticket quickly.
@@
 \item It should be easy to recall the user interface and how to use it on
-  subsequent visits. So, a good design on the travel agent’s site means the user
+  subsequent visits. So, a good design on the travel agent's site means the user
   should learn from the first time and book a second ticket just as easily.
 \end{enumerate}
+From the Interaction Design Foundation
+\url{https://www.interaction-design.org/literature/topics/usability}.
+\end{defn}
 
@@
   ``The effort required to learn, operate, prepare input, and interpret output
-  of a program'' \citep{McCallEtAl1977}. (As summarized in
-  \citet{VanVliet2000}.)  \wss{Once we have verified McCall said this, we can
-    remove the VanVliet reference.} \pmi{Yes this is almost exactly from McCall, except he did not have the word "The" at the beginning.}
+  of a program'' \citep{McCallEtAl1977}. 
 \end{defn}
@@
 
-\wss{Still needs to be completed}
-
-\noindent \textbf{Reasoning}
-
-\wss{Needs to be completed}
+\begin{defn}
+  The extent to which a product can be used by specified users to achieve
+  specified goals with effectiveness, efficiency, and satisfaction in a
+  specified context of use.
+  \cite{ISO16982:2002} and \cite{ISO9241:11}
+\end{defn}
+
+\noindent \textbf{Reasoning}
+
+There has been a lot of thought put behind the ISO definition, from experts.
+The standard has been revisited, but the definition has not changed, which 
+means that it has stood the test of time.
 
@@
   things can be easily determined.  Documentation is probably how the
-  information will be conveyed, but it doesn't have to be how it is done.}
+  information will be conveyed, but it doesn't have to be how it is done.}  \ad{Adjusted}
 

### 2020-05-08 StateOfPractice/Peter-Notes/Peter-Notes-Installability.tex
Added Advice on Improving Installability into Peter's Notes

\documentclass{article}

\usepackage{tabularx}
\usepackage{booktabs}
\usepackage{hyperref}
\hypersetup{%
	colorlinks = true,
	linkcolor  = black
}

\title{Project Notes - Installability}

\author{Peter Michalski}

\date{}

\begin{document}

\maketitle

~\newpage
\section{Advice on Improving Installability}

\begin{enumerate}
	\item Project needs an installation guide.
	\item The installation steps should be linear.
	\item All installation and testing information should ideally be in one document, or in only a few locations. In the latter case the documents should be clearly mapped.
	\item The development environment should be indicated. Specify operating system and libraries used when developing the system. Specify the exact versions. Also specify programming language versions.
	\item The installation guide should lay out all steps to install the program on a fresh virtual box/operating system. The assumption should be that absolutely nothing has been installed on the box. The guide should specify which operating system version is used to install the system using the installation guide. 
	\item The installation guide should indicate all library versions that are required, including those libraries required to view any visual output.
	\item Test cases should be available.
	\item The document should state the steps required to run a test case. 
	\item Test case solutions should be provided. 
\end{enumerate}

\end{document}

### 2020-05-11 StateOfPractice/Peter-Notes/RevisedGradingTemplate.tex
First revision of Measurement Template

--- 2020-04-21 StateOfPractice/Peter-Notes/GradingTemplate.tex
@@
   Software purpose (string)\\
-Number of developers (number)\\
+  Number of developers (number - use repo commit logs)\\
   How is the project funded (string)\\
@@
   Publications using the software (set of url)\\
-  Publications about the software (set of url) \\
+  Publications about the software (set of url)\\
   Is source code available? (\{yes, no\})\\
@@
   \midrule
- \pmi{Are required (known to work) operating system versions listed? (\{yes, no\})}\\
   Are there installation instructions? (\{yes, no\})\\
+  Are the installation instructions in one document? (\{yes, no, n/a\})\\
   Are the installation instructions linear? (\{yes, no, n/a\})\\
-  Is there something in place to automate the installation? (\{yes$^*$, no\})\\
-Is there a specified way to validate the installation, such as a test suite?
-(\{yes$^*$, no\})\\
-How many steps were involved in the installation? (number) \\
-How many software packages need to be installed before or during
-installation?(number) \\
-\pmi{are specific versions of required packages listed? (\{yes, no, n/a\})}\\
-\pmi{are there links to or instructions for the installation of required packages? (\{yes, no, n/a\})}\\
-Run uninstall, if available. Were any obvious problems caused? (\{unavail,
-yes$^*$, no\})\\
-\pmi{In your opinion could a non-developer of the software effectively install this software by verbatim following the instructions (\{yes, no, n/a\})}\\
-\pmi{Was troubleshooting the instructions necessary to install the system? (\{yes, no, n/a\})}
+  Do the installation instructions begin with a fresh operating system? (\{yes, no, unclear\})\\
+  Are compatible operating system versions listed? (\{yes, no\})\\
+  Is there something in place to automate the installation (makefile, script,
+  installer, etc)? (\{yes$^*$, no\})\\
+  Is there a specified way to validate the installation, such as a test suite? (\{yes$^*$, no\})\\
+  How many steps were involved in the installation? What OS was used? (number, set of \{Windows, Linux, OS X, Android, Other OS\}) \\
+  How many software packages need to be installed before or during installation?(number) \\
+  Are required package versions listed? (\{yes, no, n/a\})\\
+  Are there instructions for the installation of required packages? (\{yes, no, n/a\})\\
+  Run uninstall, if available. Were any obvious problems caused? (\{unavail, yes$^*$, no\})\\
   Overall impression? (\{1 .. 10\})\\
@@
   Are external libraries used? (\{yes$^*$, no, unclear\})\\
-  \pmi{are versions of external libraries specified? (\{yes, no, n/a\})}\\
+  Are versions of external libraries specified? (\{yes, no, n/a\})\\
   Does the community have confidence in this library? (\{yes, no, unclear\})\\
-  Any reference to the requirements specifications of the program?
-(\{yes$^*$, no, unclear\})\\
-What tools or techniques are used to build confidence of correctness?
-(string)\\
-If there is a getting started tutorial, is the output as expected? (\{yes,
-no$^*$, n/a\})\\
-\pmi{Does the getting started tutorial provide an answer to verify against? (\{yes, no, n/a\})}\\
+  Any reference to the requirements specifications of the program? (\{yes$^*$, no, unclear\})\\
+  What tools or techniques are used to build confidence of correctness? (string)\\
+  If there is a getting started tutorial? (\{yes, no$^*$\})\\
+  Are the tutorial instructions linear? (\{yes, no, n/a\})\\
+  Does the getting started tutorial provide an expected output? (\{yes, no, n/a\})\\
+  Does running the tutorial create output that matches the expected output? (\{yes, no$^*$, n/a\})\\
   Overall impression? (\{1 .. 10\})\\
@@
   Did the software ``break'' during installation? (\{yes$^*$, no\})\\
-  \pmi{was a helpful error message provided? (\{yes, no, n/a\})}
-    \pmi{was the installation recoverable? (\{yes, no, n/a\})}
-Did the software ``break'' during the initial tutorial testing? (\{yes$^*$, no,
-n/a\})\\
-  \pmi{was a helpful error message provided? (\{yes, no, n/a\})}\\
-  \pmi{was the testing recoverable? (\{yes, no, n/a\})}\\
+  If the software installation broke, was a descriptive error message displayed? (\{yes, no, n/a\})\\
+  If the software installation broke, was the installation recoverable? (\{yes, no, n/a\})\\
+  Did the software ``break'' during the initial tutorial testing? (\{yes$^*$, no, n/a\})\\
+  If the tutorial testing broke, was a descriptive error message displayed? (\{yes, no, n/a\})\\
+  If the tutorial testing broke, was the tutorial testing recoverable? (\{yes, no, n/a\})\\
   Overall impression? (\{1 .. 10\})\\
@@
 
-Does the software handle garbage input reasonably? (\{yes, no$^*$\}) \\
+  Does the software handle unexpected/unanticipated input (like data of the wrong type, empty
+input, missing files or links) reasonably? (a reasonable response can include
+an appropriate error message.)  (\{yes, no$^*$\}) \\
   For any plain text input files, if all new lines are replaced with new lines
@@
   Is there evidence that performance was considered? (\{yes$^*$, no\})\\
-  \pmi{Does the software (optionally) use GPU for processing? (\{yes, no, unclear\})}\\
+  Does the software allow the use of a GPU for processing? (\{yes, no, unclear\})\\
   Overall impression? (\{1 .. 10\})\\
@@
   Is there a standard example that is explained? (\{yes, no\})\\
-  \pmi{Is a solution provided? (\{yes, no, n/a\})}\\
   Is there a user manual? (\{yes, no\})\\
@@
 
-Is there a history of multiple versions of the software? (\{yes, no,
-unclear\})\\
+  Is there a history of multiple versions of the software? (\{yes, no, unclear\})\\
   Is there any information on how code is reviewed, or how to contribute?
@@
   Is there a changelog?  (\{yes, no\})\\
- \pmi{Is technical documentation (requirements, architecture, code, algorithms, interfaces, APIs) provided? (\{yes, no, unclear\})}\\ 
-What is the maintenance type? (set of \{corrective, adaptive, perfective,
-unclear\})\\
-  What issue tracking tool is employed? (set of \{Trac, JIRA, Redmine, e-mail,
-discussion board, sourceforge, google code, git, none, unclear\})\\
+  Are technical documents available (requirements, architecture, code, algorithms, interfaces, APIs)? (\{yes, no, unclear\})\\ 
+  What is the maintenance type? (set of \{corrective, adaptive, perfective, unclear\})\\
+  What issue tracking tool is employed? (set of \{Trac, JIRA, Redmine, e-mail, discussion board, sourceforge, google code, git, BitBucket, none, unclear\})\\
+  How many known bugs have been fixed? (\{number\})\\
+  How many open known bugs remain? (\{number\})\\
   Are the majority of identified bugs fixed? (\{yes, no$^*$, unclear\})\\
-What is the percentage of identified issues that are closed?
-(percentage)\\
-\pmi{Are files well commented? (\{yes, no, unclear\})}\\
-\pmi{Are files of reasonable length? (\{yes, no, unclear\})}\\
+  What is the percentage of identified issues that are closed? (percentage)\\
+  Are files commented well? (\{yes, no, unclear\})\\
   Which version control system is in use? (\{svn, cvs, git, github, unclear\})\\
@@
 
-Are any portions of the software used by another package? (\{yes$^*$,
-no\})\\
-  Is there evidence that reusability was considered in the design? (API
-documented, web service, command line tools, ...) (\{yes$^*$, no, unclear\})\\
-\pmi{Is the system modularized? (\{yes, no, unclear\})}\\
-\pmi{Are module interfaces clearly defined? (\{yes, no, n/a\})}\\
-\pmi{Are modules described in the user guide? (\{yes, no, n/a\})}\\
+  Are any portions of the software used by another package? (\{yes$^*$, no\})\\
+  Is there evidence that reusability was considered in the design? (API documented, web service, command line tools, ...) (\{yes$^*$, no, unclear\})\\
+  Is the system modularized? (\{yes, no, unclear\})\\
+  Are module interfaces clearly defined? (\{yes, no, n/a\})\\
+  Are modules described in the user guide? (\{yes, no, n/a\})\\
   Overall impression? (\{1 .. 10\})\\
@@
   (set of \{Windows, Linux, OS X, Android, Other OS\})\\
-  \pmi{Are versions of these platforms specified? (\{yes,
-  	no\})}\\
-  \pmi{How many external libraries are needed to be manually installed? (\{int, unclear\})}\\
   Are special steps taken in the source code to handle portability? (\{yes$^*$,
@@
   Is portability explicitly identified as NOT being important? (\{yes, no\})\\
-Convincing evidence that portability has been achieved? (\{yes$^*$,
-no\})\\
+  Convincing evidence that portability has been achieved? (\{yes$^*$, no\})\\
   Overall impression? (\{1 .. 10\})\\
@@
   meaningful? (\{yes, no$^*$, n/a\})\\
-Are constants (other than 0 and 1) hard coded into the program? (\{yes, no$^*$,
-n/a\})\\
-Comments are clear, indicate what is being done, not how? (\{yes, no$^*$,
-n/a\})\\
+  Are constants (other than 0 and 1) hard coded into the program? (\{yes, no$^*$, n/a\})\\
+  Comments are clear, indicate what is being done, not how? (\{yes, no$^*$, n/a\})\\
   Is the name/URL of any algorithms used mentioned?
@@
   Descriptive names of source code files? (\{yes, no$^*$, n/a\})\\
-Is a design document provided? (\{yes$^*$, no, n/a\})\\
+  Is a design document (data structures, architecture, interface, procedures) provided? (\{yes$^*$, no, n/a\})\\
   Overall impression? (\{1 .. 10\})\\
@@
   Is there a workflow that uses other softwares? (\{yes$^*$, no\})\\
-If there are external interactions, is the API clearly defined? (\{yes$^*$, no,
-n/a\})\\
+  If there are external interactions, is the API clearly defined? (\{yes$^*$, no, n/a\})\\
+  Could the software output be used by other software as input? (\{yes$^*$, no, unclear\})\\
+  Could the software take output from other software as input? (\{yes$^*$, no, unclear\})\\
+  Does the software generate files with uncommon formats? (require uncommon software to open) (\{yes$^*$, no, unclear\})\\
+  Could the software work with customized plug-ins? (\{yes$^*$, no, unclear\})\\
+  Overall impression? (\{1 .. 10\})\\
 
@@
 
-Is the development process defined? If yes, what process is used. (\{yes$^*$,
-no, n/a\})\\
+  Is the development process defined? If yes, what process is used. (\{yes$^*$, no, n/a\})\\
   Ease of external examination relative to other products
   considered?  (\{1 .. 10\})\\
+  Are there any documents recording the development process and status? (\{yes$^*$, no\})\\ 
+  Is the development environment documented? (\{yes$^*$, no, unclear\})\\
+  Are there release notes? (\{yes$^*$, no\})\\
+  In the issue tracking system, what percentage of issues are assigned to specific developers? (\{number, n/a\})\\
+  In the issue tracking system, are some issues tagged with types and/or priorities? (\{yes, no, n/a\})\\
+  Were there any project management tools used during development? (\{yes, no\})\\
   Overall impression? (\{1 .. 10\})\\

### 2020-05-13 StateOfPractice/Ao-Notes/Questionnaire.tex
hard wrap to 80 width
Update questionnaire

--- 2020-04-15 StateOfPractice/Peter-Notes/TemplateToGradeSCSoft_Comments.tex
@@
 
+\usepackage{natbib}
 \usepackage{booktabs}
@@
 
-\title{Template for Grading Scientific Computing Software} 
-\author{Spencer Smith, Adam Lazzarato, Yue Sun, Zheng Zeng,\\ Vasudha Kapil and Jacques Carette}
-\date{\today}
+\maketitle
 
-\maketitle
-\include{GradingTemplate}
+\section{Information about the developers and users}\citep{Jegatheesan2016}
+\begin{enumerate}
+\item Interviewees' current position/title? degrees?
+\item Interviewees' contribution to/relationship with the software?
+\item Length of time the interviewee has been involved with this software?
+\item How large is the development group?
+\item What is the typical background of a developer?
+\item How large is the user group?
+\item What is the typical background of a user?
+\end{enumerate}
+
+\section{Information about the software}
+
+\ad{I drafted another version without considering our measurements}
+
+\noindent\ad{3 aspects: dev process, software quality, user feedback}
+
+\begin{enumerate}
+\item Currently, what is the biggest obstacle against your development process?
+\item In order to solve this problem, is there any improvement needed for your
+development process?
+\item In the past, is there any major obstacle to your development process that
+has been solved? How did you solve it?
+\item What is your project management process? Do you think improve this process
+can tackle the current problem?
+\item What is your documentation process? Do you think improve this process can
+tackle the current problem?
+\item Can you name several software qualities that are the most important to
+this project?
+\item Is any of them hard to be achieved? If yes, what methods have considered
+or practiced to improve the situation? If practiced, does it work?
+\item Do you think the users can understand your software well? Has there been
+any problem to use this software?
+\item Do you think the current documentation can clearly convey all necessary
+knowledge to the users? If yes, how did you successfully achieve it? If no, what
+improvements are needed?
+\end{enumerate}
+
+\newpage
+
+\bibliographystyle {plainnat}
+\bibliography {../../CommonFiles/ResearchProposal}
 

### 2020-05-13 StateOfPractice/Ao-Notes/Questions to Domain Experts.tex
Created a separate file for the questions to experts, and modified these questions

--- 2020-04-20 StateOfPractice/Ao-Notes/Questionnaire.tex
@@
 
-\title{Questionnaire to Developers} 
+\title{Questions to Domain Experts} 
 \author{Ao Dong}
@@
 
-\begin{enumerate}
-\item \ad{Visibility/Transparency} Is there a certain type of development methodologies used during the development? (\{Waterfall, Scrum, Kanban, else\})
-\item \ad{Visibility/Transparency} Is there a clearly defined development process? If yes, what is it. (\{yes$^*$, no\})
-\item \ad{Visibility/Transparency} Are there any project management tools used during the development? If yes, what are they. (\{yes$^*$, no\})
-\item \ad{Correctness and Verifiability} During the process of development, what tools or techniques are used to build confidence of correctness? (string)
-\item \ad{Correctness and Verifiability} Is there any document about the requirements specifications of the program? If yes, what is it. (\{yes$^*$, no\})
-\item \ad{Portability} Do you think that portability has been achieved? If yes, how? (\{yes$^*$, no\})
-\item \ad{Maintainability} How was maintainability considered in the design? (string)
-\item \ad{Maintainability} What is the maintenance type? (set of \{corrective, adaptive, perfective,
-unclear\})
-\item \ad{Reusability} How was reusability considered in the design? (string)
-\item \ad{Reusability} Are any portions of the software used by another package? If yes, how they are used. ({yes$^*$, no})
-\item \ad{Completeness} Is any of the following documents used during the development? ({yes$^*$, no})
+\section{Paper}
 \begin{itemize}
-\item Module Guide
-\item Module Interface Specification
-\item System Verification and Validation Plan
-\item Unit Verification and Validation Plan
+\item Where should we publish this paper?
+\item Who are our targeted readers?
 \end{itemize}
-\end{enumerate}
+
+\section{Schedule \& Team}
+\begin{itemize}
+\item We propose to start this project from ... and finish it before ... . What
+do you think?
+\item Our meeting schedule is ... does it fit your plan?
+\item We need ... extra members from your team, each working ... hours per week
+for ... weeks. What do you think?
+\item Is there any other experts you think we should keep them involved?
+\end{itemize}
+
+\section{Software List}
+\begin{itemize}
+\item The top 10 software of your choice?
+\item What do you think of our process of selecting software candidates?
+\item What do you think of the list of the selected software?
+\item Is there any software you think should be removed from / added to the
+list?
+\end{itemize}
+
+\section{Qualities \& Measurements}
+\begin{itemize}
+\item Are there any pain points when you develop/use software?
+\item In order to test the usability, what is a good small task for the test?
+\item Are there any suggestions on surface measure to qualities?
+\item Are there any suggestions on deep measure to qualities?
+\item Are there any suggestions on commonality analysis?
+\end{itemize}
+
+\section{Other}
+\begin{itemize}
+\item We are facing difficulties such as ..., do you have suggestions?
+\item How to track our tasks. e.g. using GitHub issues.
+\item Equipment such as computers, test tools, OS
+\end{itemize}
 

### 2020-05-14 StateOfPractice/Methodology/Methodology.tex
added one question to task criteria
Some additions to the methodology

@@
 	reference the overall research proposal, and the ``state of the practice''
-	exercise in particular.}
+	exercise in particular.  Reference questions we are trying to answer.}
+
+\section{Research Questions}
+
+In general questions:
+
+\begin{enumerate}
+\item Comparison between domains
+\item How to measure qualities
+\item How does the quality compare for projects with the most resources to those
+  with the fewest?
+\item What skills/knowledge are needed by future developers?
+\item How can the development process be improved?
+\item What are the common pain points?
+\end{enumerate}
+
+For each domain questions''
+
+\begin{enumerate}
+\item Best examples within the domain
+\item What software artifacts?
+\item What are the pain points?
+\item Any advice on what can be done about the pain points?
+\end{enumerate}
+
+Measure the effort invested and the reward.  Related to sustainability.
+
+Collect the data and see what conclusions follow.  For an individual domain,
+between domains.  The process isn't so much about ranking the software as it is
+about looking at the software closely and see what conclusions arise.  The
+measurements are intended to force scrutiny, from different perspectives.
 
@@
 \begin{enumerate}
+\item Start with state of practice research questions.
 \item Identify domain.  (Provide criteria on a candidate domain.)
-\item 
+\item Identify list of candidate software in the domain.  (Section~\ref{SecIdentifyCandSoft})
+\item \emph{Domain Expert}: Vet domain software list.  
+\item Domain Analysis. (Section~\ref{SecDomainAnalysis})
+\item \emph{Domain Expert}: Vet domain analysis
+\item Measure using ``shallow'' measurement template.
+\item Empirical measures.
+\item Use AHP process to rank the software.  Tool for estimating AHP results?
+  Tool for ranking and sorting by various criteria?  Conditional formula to
+  assess software within each measure, or just use opinion of measurer?
+\item Identify short list for deeper exploration.
+\item \emph{Domain Expert}: vet AHP ranking and short list.
+\item With short list:
+\begin{enumerate}
+\item Survey developers
+\item Usability experiments
+\item Performance benchmarks
+\item Maintainability experiments
+\end{enumerate}
+\item Compare shallow and deep measures and ranking
+\item Answers for research questions.
 \end{enumerate}
 
-\section{Identify Candidate Software}
+\wss{The domain expert is involved in multiple steps in the process.  How best
+  to get their feedback?  The domain experts are busy and are unlikely to devote
+  significant time to the project.  We need to quickly get to the point.  Maybe
+  something around task based inspection?  Directed interview?}
+
+\section{Identify Candidate Software} \label{SecIdentifyCandSoft}
 
@@
 Commonality analysis.  Follow as for mesh generator (likely with less detail).
+Final result will be tables of commonalities, variabilities and parameters of
+variation.
 
@@
 \item Overview of Domain
-	\item Add Commonalities - Split into simulation, input, output, and nonfunctional requirements
-	\item Add Variabilites - Split into simulation, input, output, system constraints, and nonfunctional requirements
-	\item Add Parameters of Variation - Split into simulation, input, output, system constraints, and nonfunctional requirements
+\item Add Commonalities - Split into simulation, input, output, and
+  nonfunctional requirements
+\item Add Variabilities - Split into simulation, input, output, system
+  constraints, and nonfunctional requirements
+\item Add Parameters of Variation - Split into simulation, input, output, system
+  constraints, and nonfunctional requirements
 \item Add Terminology, Definitions, Acronyms
@@
 
-Commonality analysis for Lattice Boltzmann Solvers can be found \href{run:../Peter-Notes/Commonality-Analysis-LB-Systems.pdf}{here}.
+Commonality analysis for Lattice Boltzmann Solvers can be found
+\href{run:../Peter-Notes/Commonality-Analysis-LB-Systems.pdf}{here}.
 
@@
 \item Tasks that might lead the user to use help or documentation
+\item Task to modify the source code
 

### 2020-05-14 StateOfPractice/Methodology/Methodology.tex
usability experiment section update

@@
 \section{User Experiments}
-\subsection{Usability Experiment}
-Steps:
+\subsection{Usability Experimental Procedure}
+
+\subsection {Procedure}
 \begin {enumerate}
-\item Write research questions
-\item Prepare logistic details(location, time table, recording setup, moderators etc)
-\item Prepare survey questions
-\item Select a list of projects
-\item Design the tasks for the study subjects to perform (tasks can be defined based on user category, at least one task to modify the software)
-\item Select participants
-\item Conduct usability session
-\item Survey the study subject to collect pre-experiment data
+%\item Write research questions
+%\item Prepare logistic details(location, time table, recording setup, moderators etc)
+%\item Prepare survey questions
+%\item Select a list of projects
+%\item Design the tasks for the study subjects to perform (tasks can be defined based on user category, at least one task to modify the software)
+%\item Select participants
+
+\item Survey participants to collect pre-experiment data
 \item Perform tasks
-\item Observe the study subjects (take notes, record sessions, watch out for body language and verbal cues)
-\item Survey the study subjects to collect final feedback
+\item Observe the study subjects (take notes, record sessions, watch out for body languages and verbal cues)
+\item Survey the study subjects to collect feedback
 \item Prepare experiment report
@@
 \begin {enumerate}
- \item Detail plan
- \item Describe Pairwise methodology
  \item Define tasks
@@
  \end {enumerate}
-
-\subsection{Task definition/selection criteria}
+\subsection{Defining usability tasks}
+Below are the requirements to be used in defining usability tasks
+\begin {itemize}
+\item Tasks should take no more than 2 hours for completion - review timing
+\item Tasks should be defined for different three different levels of participant: Expert, intermediate and novice (These 3 groups of users should be defined)
+\item Tasks should be defined based on 3 difficulty levels : Difficult, Medium and easy (Difficulty levels should be defined) 
+\end {itemize}
+\subsection{Task selection criteria}
 \begin {enumerate}
-\item Tasks based on difficulty levels (Difficult, Medium and easy)
 \item Commonly performed tasks (applicable to all the selected software)
@@
 \item Tasks to verify if the system is capable of preventing and /or reversing errors
-\item Tasks that might lead the user to use help or documentation
+\item Tasks that can lead the user to use help or documentation
 \item Task to modify the source code

### 2020-05-15 StateOfPractice/Methodology/Methodology.tex
merged sections 7.5 and 7.5

@@
 \begin {enumerate}
+%\item Select a list of projects
 %\item Write research questions
@@
 %\item Prepare survey questions
-%\item Select a list of projects
 %\item Design the tasks for the study subjects to perform (tasks can be defined based on user category, at least one task to modify the software)
@@
 \item Survey participants to collect pre-experiment data
-\item Perform tasks
+\item Participants perform tasks
 \item Observe the study subjects (take notes, record sessions, watch out for body languages and verbal cues)
-\item Survey the study subjects to collect feedback
+\item Survey the study subjects to collect feedback (post experiment interview)
 \item Prepare experiment report
@@
 
-\subsection{Empirical Measures}
- \begin {enumerate}
- \item Qualitative/Quantitative
- \item Pairwise Comparison methodology needs to be described
- \item Measure time to complete a task and compare actual with expected time of completion
- \item From a list of given tasks, measure how many tasks were completed correctly
- \item Learnability
- \item Efficiency
- \item Memorability
- \item Errors
- \item Satisfaction
- \item Heuristic evaluation: Experts go through the interface to identify elements that violate usability heuristics.Popular because of its low cost and low time commitment
-\end {enumerate}
+%\subsection{Empirical Measures}
+% \begin {enumerate}
+% \item Qualitative/Quantitative
+% \item Pairwise Comparison methodology needs to be described
+% \item Measure time to complete a task and compare actual with expected time of completion
+% \item From a list of given tasks, measure how many tasks were completed correctly
+% \item Learnability
+% \item Efficiency
+% \item Memorability
+% \item Errors
+% \item Satisfaction
+% \item Heuristic evaluation: Experts go through the interface to identify elements that violate usability heuristics.Popular because of its low cost and low time commitment
+%\end {enumerate}
 
@@
  \end {enumerate}
-\subsection{Defining usability tasks}
-Below are the requirements to be used in defining usability tasks
-\begin {itemize}
-\item Tasks should take no more than 2 hours for completion - review timing
+
+\subsection{Task selection criteria}
+Below are the criteria to be used in defining usability experiment tasks
+\begin {enumerate}
+\item Tasks should take no more than 2 hours(in one sitting) for completion.
 \item Tasks should be defined for different three different levels of participant: Expert, intermediate and novice (These 3 groups of users should be defined)
 \item Tasks should be defined based on 3 difficulty levels : Difficult, Medium and easy (Difficulty levels should be defined) 
-\end {itemize}
-\subsection{Task selection criteria}
-\begin {enumerate}
-\item Commonly performed tasks (applicable to all the selected software)
-\item Task that would/should generate errors messages or error codes
+\item What are the most common tasks that users perform (these applies to all selected projects collectively) on the software application.
+\item Task that would generate errors messages or error codes
 \item Tasks to check standards and conventions
-\item Tasks that should provide users with the system's current state and what actions can be done or where to go or what change was made
+\item Tasks that provide users with the system's current state and what actions can be done or where to go or what change was made
 \item Tasks related to what users will do in the real world
-\item End to End tasks that require users to go through sequential or hierarchical steps.
+\item End to end tasks that require users to go through sequential or hierarchical steps.
 \item Tasks that allow users to reuse a previously defined value without having to memorize or jot it.
-\item Tasks that should provide users with feedback
+\item Tasks that provide users with feedback.
 \item Tasks that allow users to customize and use shortcuts, bookmarks, command keys etc.
 \item Tasks to verify if the system is capable of preventing and /or reversing errors
-\item Tasks that can lead the user to use help or documentation
-\item Task to modify the source code
+\item Tasks that would lead the user to use help or documentation
+%\item Task to modify the source code
 

### 2020-05-19 StateOfPractice/Peter-Notes/Peter-Notes-Installability.tex
added bare bones installation check advice to installability notes

@@
 	\item The development environment should be indicated. Specify operating system and libraries used when developing the system. Specify the exact versions. Also specify programming language versions.
-	\item The installation guide should lay out all steps to install the program on a fresh virtual box/operating system.
+	\item The installation guide should lay out all steps to install the program on a bare bones virtual box/operating system.
 The assumption should be that absolutely nothing has been installed on the box.
 The guide should specify which operating system version is used to install the system using the installation guide.
+	\item The installation process should be checked using a bare bones virtual machine
 	\item The installation guide should indicate all library versions that are required, including those libraries required to view any visual output.

### 2020-05-27 StateOfPractice/Methodology/Methodology.tex
update usability experiment section

@@
 \item Participants perform tasks
-\item Observe the study subjects (take notes, record sessions, watch out for body languages and verbal cues)
+\item Observe the study subjects (take notes, record sessions(OBS screen recorder), watch out for body languages and verbal cues)
 \item Survey the study subjects to collect feedback (post experiment interview)
@@
 
-\subsection{Nielsen’s Heuristics}
-\begin{enumerate}
-	\item Consistency: Check if standards and conventions in product design are followed
- \begin{itemize}
- 	\item Sequences of actions (skill acquisition).
- 	\item Color (categorization).
- 	\item Layout and position (spatial consistency).
- 	\item Font, capitalization (levels of organization).
- 	\item Terminology (delete, del, remove, rm) and language (words, phrases).
- 	\item Standards (e.g.,	blue underlined text for unvisited hyper links).
- \end{itemize}
-\item Visibility: Users should be informed about what is going on with the system through appropriate feedback and display of information e.g. 
- \begin{itemize}
- \item What is the current state of the system?
- \item What can be done at current state?
- \item Where can users go?
- \item What change is made after an action?
- \end{itemize}
-\item Match between system and world. The image of the system perceived by users should match the model the users have about the system. 
-\begin {itemize}
-\item User model matches system image.
-\item Actions provided by the system should match actions performed by users.
-\item Objects on the system should match objects of the task.
-\end {itemize}
-\item Minimalist: Any extraneous information is a distraction and a slow-down. 
-\begin{itemize}
-\item Less is more.
-\item Simple is not equivalent to abstract and general.
-\item Simple is efficient.
-\item  Progressive levels of detail.
-\end{itemize}
-\item Minimize memory load. Users should not be required to memorize a lot of information to carry out tasks. Memory load reduces users’ capacity to carry out the main tasks. 
-\begin{itemize}
-\item Recognition vs. recall (e.g., menu vs. commands).
-\item Externalize information through visualization.
-\item Perceptual procedures.
-\item Hierarchical structure.
-\item Default values.
-\item Concrete examples (DD/MM/YY, e.g., 10/20/99).
-\item Generic rules and actions (e.g., drag objects).
-\item Intuitive procedure
-\end{itemize}
-\item Informative feedback. Users should be given prompt and informative feedback about their actions. 
-\begin{itemize}
-\item Information that can be directly perceived, interpreted, and evaluated.
-\item Levels of feedback (novice and expert).
-\item Concrete and specific, not abstract and general.
-\item Response time:\\
--0.1 s for instantaneously reacting;\\
--1.0 s for uninterrupted flow of thought;\\
--10 s for the limit of attention.
-\end{itemize}
-\item Flexibility and efficiency. Users always learn and users are always different. Give users the flexibility of creating customization and shortcuts to accelerate their performance. 
-\begin{itemize}
-	\item Shortcuts for experienced users.
-	\item Shortcuts or macros for frequently used operations.
-	\item Skill acquisition through chunking.
-	\item Examples:
-	- Abbreviations, function keys, hot keys, command keys, macros, aliases, templates, type-ahead, bookmarks, hot links, history, default values, etc.
-\end{itemize}
-\item Good error messages. 
-\begin {itemize}
-\item The messages should be informative enough such that users can understand the nature of errors, learn from errors, and recover from errors. 
-\item Phrased in clear language, avoid obscure codes. Example of obscure code: “system crashed, error code 147.”
-\item Precise, not vague or general. Example of general comment: “Cannot open document.”
-\item Constructive.
-\item Polite. Examples of impolite message: “illegal user action,” “job aborted,” “system was crashed,” “fatal error,” etc.
-\end {itemize}
-\item Prevent errors. It is always better to design interfaces that prevent errors from happening in the first place.
-\begin {itemize} 
-\item Interfaces that make errors impossible.
-\item Avoid modes (e.g., vi, text wrap). Or use informative feedback, e.g., different sounds.
-\item Execution error vs. evaluation error.
-\item Various types of slips and mistakes.
-\end {itemize}
-\item Clear closure. Every task has a beginning and an end. Users should be clearly notified about the completion of a task. 
-\begin {itemize}
-\item Clear beginning, middle, and end.
-\item Complete 7-stages of actions.
-\item Clear feedback to indicate goals are achieved and current stacks of goals can be released. Examples of good closures include many dialogues.
-\end {itemize}
-\item Reversible actions. 
-\begin {itemize}
-\item Users should be allowed to recover from errors. Reversible actions also encourage exploratory learning. 
-\item At different levels: a single action, a subtask, or a complete task.
-\item Multiple steps.
-\item Encourage exploratory learning.
-\item Prevent serious errors.
-\end {itemize}
-\item Use users’ language. The language should be always presented in a form understandable by the intended users.
-\begin{itemize}
-\item Use standard meanings of words.
-\item Specialized language for specialized group.
-\item User defined aliases.
-\item Users’ perspective. Example: “we have bought four tickets for you” (bad) vs. “you bought four tickets” (good).
-\end {itemize}
-\item Help and documentation. Always provide help when needed. 
-\begin {itemize}
-\item Context-sensitive help.
-\item Four types of help.\\
--task-oriented;\\
--alphabetically ordered;\\
--semantically organized;\\
--search.
-\item Help embedded in contents.
-\end {itemize}
-\item Users in control. Do not give users that impression that they are controlled by the systems. 
-\begin {itemize}
-\item Users are initiators of actors, not responders to actions.
-\item Avoid surprising actions, unexpected outcomes, tedious sequences of actions, etc.
-\end {itemize}
-\end{enumerate}
-\subsection{Tools}
+%\subsection{Nielsen’s Heuristics}
+%\begin{enumerate}
+%	\item Consistency: Check if standards and conventions in product design are followed
+% \begin{itemize}
+% 	\item Sequences of actions (skill acquisition).
+% 	\item Color (categorization).
+% 	\item Layout and position (spatial consistency).
+% 	\item Font, capitalization (levels of organization).
+% 	\item Terminology (delete, del, remove, rm) and language (words, phrases).
+% 	\item Standards (e.g.,	blue underlined text for unvisited hyper links).
+% \end{itemize}
+%\item Visibility: Users should be informed about what is going on with the system through appropriate feedback and display of information e.g. 
+% \begin{itemize}
+% \item What is the current state of the system?
+% \item What can be done at current state?
+% \item Where can users go?
+% \item What change is made after an action?
+% \end{itemize}
+%\item Match between system and world. The image of the system perceived by users should match the model the users have about the system. 
+%\begin {itemize}
+%\item User model matches system image.
+%\item Actions provided by the system should match actions performed by users.
+%\item Objects on the system should match objects of the task.
+%\end {itemize}
+%\item Minimalist: Any extraneous information is a distraction and a slow-down. 
+%\begin{itemize}
+%\item Less is more.
+%\item Simple is not equivalent to abstract and general.
+%\item Simple is efficient.
+%\item  Progressive levels of detail.
+%\end{itemize}
+%\item Minimize memory load. Users should not be required to memorize a lot of information to carry out tasks. Memory load reduces users’ capacity to carry out the main tasks. 
+%\begin{itemize}
+%\item Recognition vs. recall (e.g., menu vs. commands).
+%\item Externalize information through visualization.
+%\item Perceptual procedures.
+%\item Hierarchical structure.
+%\item Default values.
+%\item Concrete examples (DD/MM/YY, e.g., 10/20/99).
+%\item Generic rules and actions (e.g., drag objects).
+%\item Intuitive procedure
+%\end{itemize}
+%\item Informative feedback. Users should be given prompt and informative feedback about their actions. 
+%\begin{itemize}
+%\item Information that can be directly perceived, interpreted, and evaluated.
+%\item Levels of feedback (novice and expert).
+%\item Concrete and specific, not abstract and general.
+%\item Response time:\\
+%-0.1 s for instantaneously reacting;\\
+%-1.0 s for uninterrupted flow of thought;\\
+%-10 s for the limit of attention.
+%\end{itemize}
+%%\item Flexibility and efficiency. Users always learn and users are always different. Give users the flexibility of creating customization and shortcuts to accelerate their performance. 
+%%\begin{itemize}
+%%	\item Shortcuts for experienced users.
+%%	\item Shortcuts or macros for frequently used operations.
+%%	\item Skill acquisition through chunking.
+%%	\item Examples:
+%%	- Abbreviations, function keys, hot keys, command keys, macros, aliases, templates, type-ahead, bookmarks, hot links, history, default values, etc.
+%%\end{itemize}
+%\item Good error messages. 
+%\begin {itemize}
+%\item The messages should be informative enough such that users can understand the nature of errors, learn from errors, and recover from errors. 
+%\item Phrased in clear language, avoid obscure codes. Example of obscure code: “system crashed, error code 147.”
+%\item Precise, not vague or general. Example of general comment: “Cannot open document.”
+%\item Constructive.
+%\item Polite. Examples of impolite message: “illegal user action,” “job aborted,” “system was crashed,” “fatal error,” etc.
+%\end {itemize}
+%\item Prevent errors. It is always better to design interfaces that prevent errors from happening in the first place.
+%\begin {itemize} 
+%\item Interfaces that make errors impossible.
+%\item Avoid modes (e.g., vi, text wrap). Or use informative feedback, e.g., different sounds.
+%\item Execution error vs. evaluation error.
+%\item Various types of slips and mistakes.
+%\end {itemize}
+%\item Clear closure. Every task has a beginning and an end. Users should be clearly notified about the completion of a task. 
+%\begin {itemize}
+%\item Clear beginning, middle, and end.
+%\item Complete 7-stages of actions.
+%\item Clear feedback to indicate goals are achieved and current stacks of goals can be released. Examples of good closures include many dialogues.
+%\end {itemize}
+%\item Reversible actions. 
+%\begin {itemize}
+%\item Users should be allowed to recover from errors. Reversible actions also encourage exploratory learning. 
+%\item At different levels: a single action, a subtask, or a complete task.
+%\item Multiple steps.
+%\item Encourage exploratory learning.
+%\item Prevent serious errors.
+%\end {itemize}
+%\item Use users’ language. The language should be always presented in a form understandable by the intended users.
+%\begin{itemize}
+%\item Use standard meanings of words.
+%\item Specialized language for specialized group.
+%\item User defined aliases.
+%\item Users’ perspective. Example: “we have bought four tickets for you” (bad) vs. “you bought four tickets” (good).
+%\end {itemize}
+%\item Help and documentation. Always provide help when needed. 
+%\begin {itemize}
+%\item Context-sensitive help.
+%\item Four types of help.\\
+%-task-oriented;\\
+%-alphabetically ordered;\\
+%-semantically organized;\\
+%-search.
+%\item Help embedded in contents.
+%\end {itemize}
+%\item Users in control. Do not give users that impression that they are controlled by the systems. 
+%\begin {itemize}
+%\item Users are initiators of actors, not responders to actions.
+%\item Avoid surprising actions, unexpected outcomes, tedious sequences of actions, etc.
+%\end {itemize}
+%\end{enumerate}
+
+\subsection{Task selection criteria}
+Below are the criteria to be used in defining usability experiment tasks.
+The task selection will be determined with the aid of the domain expert attached to any of the selected projects.
+The domain expert will be asked to consider the below criteria when helping us to determine a task.
 \begin {enumerate}
-\item Downloaded OBS to record experiment sessions
-\end {enumerate}
-\subsection{To do}
-\begin {enumerate}
- \item Define tasks
- \item prepare survey questions
+\item All tasks should take no more than 2 hours (in one sitting) for completion.
+\item Common and popular tasks that users would perform (these apply to all selected projects collectively) with the software application.
+%\item Task that would generate errors messages or error codes
+%\item Tasks to check standards and conventions
+%\item Tasks that provide users with the system's current state and what actions can be done or where to go or what change was made
+\item Tasks related to what users will do in the real world
+\item End to end tasks that require users to go through sequential or hierarchical steps.
+%\item Tasks that allow users to reuse a previously defined value without having to memorize or jot it.
+%\item Tasks that provide users with feedback.
+%\item Tasks that allow users to customize and use shortcuts, bookmarks, command keys etc.
+%\item Tasks to verify if the system is capable of preventing and /or reversing errors
+%\item Tasks that would lead the user to use help or documentation
 \end {enumerate}
 
-\subsection{Task selection criteria}
-Below are the criteria to be used in defining usability experiment tasks
-\begin {enumerate}
-\item Tasks should take no more than 2 hours(in one sitting) for completion.
-\item Tasks should be defined for different three different levels of participant: Expert, intermediate and novice (These 3 groups of users should be defined)
-\item Tasks should be defined based on 3 difficulty levels : Difficult, Medium and easy (Difficulty levels should be defined) 
-\item What are the most common tasks that users perform (these applies to all selected projects collectively) on the software application.
-\item Task that would generate errors messages or error codes
-\item Tasks to check standards and conventions
-\item Tasks that provide users with the system's current state and what actions can be done or where to go or what change was made
-\item Tasks related to what users will do in the real world
-\item End to end tasks that require users to go through sequential or hierarchical steps.
-\item Tasks that allow users to reuse a previously defined value without having to memorize or jot it.
-\item Tasks that provide users with feedback.
-\item Tasks that allow users to customize and use shortcuts, bookmarks, command keys etc.
-\item Tasks to verify if the system is capable of preventing and /or reversing errors
-\item Tasks that would lead the user to use help or documentation
-%\item Task to modify the source code
-
-\end {enumerate}
+**Domain experts will be asked to identify what background knowledge is necessary for the suggested tasks (Expert, Intermediate, Novice).
 
@@
 \item I needed to learn a lot of things before I could get going with this system.
+\item \textbf{The below were added to this list by Olu based on Nielson's Heuristics}
+\item The software provided me system's current state and what actions can be done or where to go or what change was made.
+\item I was able reuse a previously defined value without having to memorize or jot it.
+\item I was able to reverse an error and didn't have to restart a task due to error.
+\item I was able to refer to help/documentation easily when I needed it.
+\item I was pre-warned of errors based on the actions I performed
 \end {enumerate}

### 2020-05-27 StateOfPractice/Ao-Notes/A Guide to Empirical Measures.tex

\documentclass[12pt]{article}

\usepackage{natbib}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{xspace}
\usepackage{fullpage}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{float}

\hypersetup{
colorlinks=true,       % false: boxed links; true: colored links
linkcolor=red,          % color of internal links (change box color with
%linkbordercolor)
citecolor=blue,       % color of links to bibliography
filecolor=magenta,   % color of file links
urlcolor=cyan           % color of external links
}

%% Global noindent
\setlength\parindent{0pt}

%% Comments
\newif\ifcomments\commentstrue

\ifcomments
\newcommand{\authornote}[3]{\textcolor{#1}{[#3 ---#2]}}
\newcommand{\todo}[1]{\textcolor{red}{[TODO: #1]}}
\else
\newcommand{\authornote}[3]{}
\newcommand{\todo}[1]{}
\fi

\newcommand{\wss}[1]{\authornote{blue}{SS}{#1}} %Spencer Smith
\newcommand{\jc}[1]{\authornote{red}{JC}{#1}} %Jacques Carette
\newcommand{\oo}[1]{\authornote{magenta}{OO}{#1}} %Olu Owojaiye
\newcommand{\pmi}[1]{\authornote{green}{PM}{#1}} %Peter Michalski
\newcommand{\ad}[1]{\authornote{cyan}{AD}{#1}} %Ao Dong

\usepackage{listings}
\usepackage{color}
\usepackage{xcolor}

\definecolor{dkgreen}{rgb}{0,0.6,0}
\definecolor{gray}{rgb}{0.5,0.5,0.5}
\definecolor{mauve}{rgb}{0.58,0,0.82}

%% Code blocks
\lstset{
  language=bash,
  aboveskip=3mm,
  belowskip=3mm,
  showstringspaces=false,
  columns=flexible,
  basicstyle={\small\ttfamily\color{white}},
  numbers=none,
  backgroundcolor = \color{black},
  commentstyle=\color{gray},
  stringstyle=\color{mauve},
  breaklines=true,
  breakatwhitespace=true,
  tabsize=3
}

\title{A Guide to Empirical Measures} 
\author{Ao Dong}
\date{\today}

\begin{document}

\maketitle

This document is a general guide to the implements of empirical measures.

There are several sections in this guide, with one tool to use in each of them,
and usually measures different aspect of a software package. There is no
mandatory order to use which tool first.

\section{git\_stats}
\subsection{Introduction}
Name: git\_stats

Source Code: \href{https://github.com/tomgi/git_stats}{GitHub repo}

\subsection{User Manual}
\label{git_stats_manual}
Official Manual: \href{https://github.com/tomgi/git_stats}{GitHub repo}

\subsection{Demo of Installation and Running the Tool}
This is the showcase of how to install and run this tool.

The installation steps on your machine may be different from this section.
Please refer to Section \ref{git_stats_manual}.

Hardware: a virtual machine with 8 cores and 16 GB RAM

OS: Debian GNU/Linux 9.11

\begin{enumerate}
\item Install ruby/gem environment
\begin{lstlisting}
apt-get install ruby ruby-nokogiri ruby-nokogiri-diff ruby-nokogumbo
\end{lstlisting}

Check the installation:
\begin{lstlisting}
gem --version
\end{lstlisting}

\item Install the tool
\begin{lstlisting}
sudo gem install git_stats
\end{lstlisting}

\item Prepare the target repo

Make sure the target repo (the repo to be analyzed, not the repo of this tool)
is on your machine.
In this demo, the target repo is downloaded from a GitHub repo:
\begin{lstlisting}
# change [project path] to your desired folder
cd [project path]
git clone https://github.com/nroduit/Weasis.git
\end{lstlisting}

\item Generate analytics
\begin{lstlisting}
# make sure [repo path] is the target repo path
# the [output path] can be anywhere you desire
git_stats generate -p [repo path] -o [output path]
# e.g. git_stats generate -p /home/user/git-stats/Weasis -o
/home/user/git-stats/Weasis-analytics
\end{lstlisting}

\item View the analytics

View the analytic results by open [output path]/index.html with any browser or
other software supporting HTML web page format.
\begin{figure}[H]
\includegraphics[scale=0.5]{git-stats-demo.JPG}
\end{figure}

\item Download the data

On most of the taps of this web page, the data can be downloaded for more
analytics by clicking the menu button beside the data-range section.
\end{enumerate}

\section{scc}
\subsection{Introduction}
Name: scc

Source Code: \href{https://github.com/boyter/scc}{GitHub repo}

\subsection{User Manual}
\label{scc_manual}
Official Manual: \href{https://github.com/boyter/scc}{GitHub repo}

\subsection{Demo of Installation and Running the Tool}
This is the showcase of how to install and run this tool.

The installation steps on your machine may be different from this section.
Please refer to Section \ref{scc_manual}.

Hardware: a virtual machine with 8 cores and 16 GB RAM

OS: Debian GNU/Linux 9.11

\begin{enumerate}
\item Install Golang

Follow the \href{https://golang.org/doc/install}{official instructions}, or the
following demo,

download the installation package:
\begin{lstlisting}
wget https://dl.google.com/go/go1.14.3.linux-amd64.tar.gz
\end{lstlisting}

unpack to /usr/local:
\begin{lstlisting}
sudo tar -C /usr/local -xzf go1.14.3.linux-amd64.tar.gz
\end{lstlisting}

use a text editor to open \textasciitilde/.profile, e.g.:
\begin{lstlisting}
nano ~/.profile
\end{lstlisting}

add the following lines to the end of this file:
\begin{lstlisting}
export GOPATH=$HOME/go
export PATH=$PATH:/usr/local/go/bin:$GOPATH/bin
\end{lstlisting}

save the file, and load the commands into the current shell instance:
\begin{lstlisting}
source ~/.profile
\end{lstlisting}

check the installation:
\begin{lstlisting}
go version
\end{lstlisting}

\item Install the tool
\begin{lstlisting}
go get -u github.com/boyter/scc/
\end{lstlisting}

\item Prepare the target repo

Make sure the target repo (the repo to be analyzed, not the repo of this tool)
is on your machine.
In this demo, the target repo is downloaded from a GitHub repo:
\begin{lstlisting}
# change [project path] to your desired folder
cd [project path]
git clone https://github.com/nroduit/Weasis.git
\end{lstlisting}

\item Generate analytics
\begin{lstlisting}
# make sure [repo path] is the target repo path
cd [repo path]
# use scc to generate analytics
scc
\end{lstlisting}

\item View the analytics

The results will be directly shown.
\begin{figure}[H]
\includegraphics[scale=0.7]{scc-demo.JPG}
\end{figure}

\end{enumerate}

\end{document}

### 2020-06-02 OverallResearchProposal/ObjectivesAndResearchQuestions.tex
Minor notes added to quality definition document related to reproducibility

@@
 
-\textbf{What knowledge needs to be captured and what principles, processes and
+OLD \textbf{What is needed knowledge needs to be captured and what principles, processes and
   methodologies followed to produce sustainable \{software + artifacts\} with
@@
 
-In the above definition, software is considered to be an application (app)
+\textbf{What is needed to produce sustainable \{software + artifacts\}?}\\
+
+Define software.
+
+Define sustainable.  We start with the general definition:
+
+``Sustainable development is development that meets the needs
+  of the present without compromising the ability of future generations to meet
+  their own needs.'' \citep{Brundtland1987}
+
+Might be better to use the adopted definition - since it uses qualities, which
+are terms that we need.
+
+Define softifacts.  In the above definition, software is considered to be an application (app)
 and/or a library.  The outcome of the development process is not just software.
@@
 
+What is needed - In the statement of the objective, principles, processes and
+methodologies are listed, but not tools.  Tools are needed to support
+methodologies, but tools change rapidly.  Our goal is to be more general.  If
+the right principles, processes and methodologies are found, the tools will
+adapt to match.  However, understanding the current state of the practice will
+involve looking at tools, as shown in the later list of research questions.
+
 The notion of ``reasonable amount of time and energy'' is part of the objective,
@@
 where it makes sense to aim for sustainable softifacts.
-
-In the statement of the objective, principles, processes and methodologies are
-listed, but not tools.  Tools are needed to support methodologies, but tools
-change rapidly.  Our goal is to be more general.  If the right principles,
-processes and methodologies are found, the tools will adapt to match.  However,
-understanding the current state of the practice will involve looking at tools,
-as shown in the later list of research questions.
 
@@
   separately about the needs of the present and the future.  For the needs of
-  the present, the software should be correct, reliable and usable. For the
+  the present, the software should meet its requirements specification.
+  \wss{highlight correctness? reliability?}  For the
   needs of the future, given uncertainty, this means supporting change from
@@
   Supporting the future then means producing softifacts that are
-  maintainable, portable, reusable and reproducible.
+  maintainable, reusable and reproducible.
 

### 2020-06-03 OverallResearchProposal/ObjectivesAndResearchQuestions.tex
Simplifyin the overall objective to simply be "What is needed to produce sustainable softifacts?"

@@
 
-Define software.
+To understand this research objective, we will decompose each constituent part
+in turn, starting with the end of the sentence.  In the above objective,
+software consists of either programs, which run (execute) on a computer, or
+libraries, which provide services to be used by programs.  Both programs and
+libraries are created using computer code, written in a programming language.
+Many developers place most, if not all, of their emphasis on the computer code
+that makes up programs and libraries.  However, we consider the outcome of the
+development process to be more than just software and the associated computer
+code; we consider the products of the development process to be \{software +
+artifacts\}.  The artifacts include code, but also all the other potential work
+products, such as documentation, build scripts, test cases, contributor's
+guides, etc.  From this point onward, we will use the term \emph{softifacts}
+when we are referencing the combination of software and the associated
+artifacts.  As we will see, assessing the quality of sustainability is not
+possible by only considering the computer code; it depends on all of the
+artifacts, and on the processes and methodologies used during software
+development.  \wss{Can we consider software as both the computer code and the
+  executable (program or library), or should the term software just mean the
+  executable?  That is, when we mean code, should we be careful to not instead
+  say software?}
 
@@
 are terms that we need.
-
-Define softifacts.  In the above definition, software is considered to be an application (app)
-and/or a library.  The outcome of the development process is not just software.
-It is \{software + artifacts\}.  From this point onward, we will use the term
-softifacts when we are referencing the combination of software and the
-associated artifacts.  The artifacts consist of such work products as
-documentation, build scripts, code, test cases, contributor's guides, etc.  The
-quality of sustainability cannot be assessed by looking at the software alone,
-or the code alone; it depends on everything, including the processes and
-methodologies used during the development.  Further information on the relevant
-definitions and context is given in the next section.
 
@@
 domain of software.
+
+Further information on the relevant definitions and context is given in the next
+section.  Roadmap.
 

### 2020-06-10 OverallResearchProposal/ObjectivesAndResearchQuestions.tex
Restoring previous version of file that was removed by a recent merge

@@
 
-OLD \textbf{What is needed knowledge needs to be captured and what principles, processes and
-  methodologies followed to produce sustainable \{software + artifacts\} with
-  a reasonable amount of time and energy?}\\
-
 \textbf{What is needed to produce sustainable \{software + artifacts\}?}\\
 
-To understand this research objective, we will decompose each constituent part
-in turn, starting with the end of the sentence.  In the above objective,
+To fully explain this research objective, we will decompose each constituent
+part in turn, starting with the end of the sentence.  In the above objective,
 software consists of either programs, which run (execute) on a computer, or
@@
 when we are referencing the combination of software and the associated
-artifacts.  As we will see, assessing the quality of sustainability is not
+artifacts.  As we will explain below, assessing the quality of sustainability is not
 possible by only considering the computer code; it depends on all of the
@@
 
-Define sustainable.  We start with the general definition:
+We define sustainable softifacts, as follows: \emph{Sustainable softifacts
+  satisfy, for a reasonable amount of energy, the software requirements for the
+  present, while also being maintainable, reusable and reproducible for the
+  future.}  This definition is a specialization of the general definition of
+sustainable development from \citet{Brundtland1987}, as discussed in
+Section~\ref{Sec_Context}.  The notion of ``reasonable effort'' is part of the
+definition of sustainability, but reasonable cannot be completely defined
+because what is reasonable is going to depend on the context.  Problems that are
+large, complex and/or critical are going to have a greater ``budget'' of time
+and energy.  When comparing different alternatives for processes and
+methodologies, preference would be given to the alternative that costs less
+effort, assuming that this alternative achieves the same (or a higher) level of
+sustainability as its competition.  The assumptions listed in the next section
+(Section~\ref{Sec_Context}) outline the typical space of problems where it makes
+sense to aim for sustainable softifacts.
 
-``Sustainable development is development that meets the needs
-  of the present without compromising the ability of future generations to meet
-  their own needs.'' \citep{Brundtland1987}
-
-Might be better to use the adopted definition - since it uses qualities, which
-are terms that we need.
-
-What is needed - In the statement of the objective, principles, processes and
-methodologies are listed, but not tools.  Tools are needed to support
-methodologies, but tools change rapidly.  Our goal is to be more general.  If
-the right principles, processes and methodologies are found, the tools will
-adapt to match.  However, understanding the current state of the practice will
-involve looking at tools, as shown in the later list of research questions.
-
-The notion of ``reasonable amount of time and energy'' is part of the objective,
-but reasonable is not defined in the objective.  What is reasonable is going to
-depend on the context.  Problems that are large, complex and/or critical are
-going to have a greater ``budget'' of time and energy.  When comparing different
-alternatives for processes and methodologies, preference would be given to the
-alternative that costs less time and energy, assuming that this alternative
-achieves the same (or a higher) level of sustainability as its competition.  The
-assumptions listed in the next section outline the typical space of problems
-where it makes sense to aim for sustainable softifacts.
+The overall objectives asks what is needed for sustainability.  This ``what''
+can be decomposed further.  We consider the ``what'' to be the principles,
+processes and methodologies used for software development.  We exclude tools
+from this list because, although tools are needed to support processes and
+methodologies, tools change rapidly.  Our goal is to be general.  If the right
+principles, processes and methodologies are found, the tools will adapt to
+match.  However, understanding the current state of the practice will involve
+looking at tools, as shown in the later list of research questions
+(Section~\ref{Sec_ResearchQuestions}).
 
 The objective is not phrased to target a specific software domain, but one of
-the research assumptions is that we will focus on Scientific Computing Software
-(SCS) examples and applications.  This focus is because creating sustainable
-softifacts is more feasible when the software domain is well understood,
-which is the case for SCS.  Moreover, SCS is an important, often neglected,
-domain of software.
+the research assumptions (Section~\ref{Sec_Context}) is that we will focus on
+Research Software (RS) examples and applications.  This focus is because
+creating sustainable softifacts is more feasible when the software domain is
+well understood, which is the case for RS.  Moreover, RS is an important,
+often neglected, domain of software.
 
@@
 
-\subsection*{Context, Definitions and Assumptions}
+\subsection*{Context, Definitions and Assumptions} \label{Sec_Context}
 
@@
 
-  \emph{Sustainable software means softifacts that are correct,
-    reliable, and usable (for the present) and maintainable, portable, reusable
-    and reproducible (for the future).}
+  \emph{Sustainable softifacts satisfy, for a reasonable amount of energy, the
+    software requirements for the present, while also being maintainable,
+    reusable and reproducible for the future.}
 
@@
 
-\subsubsection*{Current State of the Practice Research Questions}
+\subsubsection*{Current State of the Practice Research Questions} \label{Sec_ResearchQuestions}
 
@@
 
-- what is needed to produce softifacts sustainably?
-
 - measuring process aimed at 'producing softifacts sustainably'. part of it could be measuring effort.

### 2020-06-11 StateOfPractice/Methodology/Methodology.tex
updated selection criteria

@@
 
-%\subsection{Empirical Measures}
-% \begin {enumerate}
-% \item Qualitative/Quantitative
-% \item Pairwise Comparison methodology needs to be described
-% \item Measure time to complete a task and compare actual with expected time of completion
-% \item From a list of given tasks, measure how many tasks were completed correctly
-% \item Learnability
-% \item Efficiency
-% \item Memorability
-% \item Errors
-% \item Satisfaction
-% \item Heuristic evaluation: Experts go through the interface to identify elements that violate usability heuristics.Popular because of its low cost and low time commitment
-%\end {enumerate}
-
 %\subsection{Nielsen’s Heuristics}
@@
 \subsection{Task selection criteria}
-Below are the criteria to be used in defining usability experiment tasks.
-The task selection will be determined with the aid of the domain expert attached to any of the selected projects.
-The domain expert will be asked to consider the below criteria when helping us to determine a task.
+**The task selection will be determined with the aid of the domain expert attached to any of the selected projects.\\ 
+**The domain expert will be asked to consider the below criteria when defining a task. \\
+**Domain experts will also be asked to identify what background knowledge is necessary for the suggested tasks - Novice, Intermediate, Advanced
 \begin {enumerate}
-\item All tasks should take no more than 2 hours (in one sitting) for completion.
-\item Common and popular tasks that users would perform (these apply to all selected projects collectively) with the software application.
-%\item Task that would generate errors messages or error codes
-%\item Tasks to check standards and conventions
-%\item Tasks that provide users with the system's current state and what actions can be done or where to go or what change was made
-\item Tasks related to what users will do in the real world
-\item End to end tasks that require users to go through sequential or hierarchical steps.
-%\item Tasks that allow users to reuse a previously defined value without having to memorize or jot it.
-%\item Tasks that provide users with feedback.
-%\item Tasks that allow users to customize and use shortcuts, bookmarks, command keys etc.
-%\item Tasks to verify if the system is capable of preventing and /or reversing errors
-%\item Tasks that would lead the user to use help or documentation
+\item Collectively all tasks should not take no more than 2 hours.
+\item Selected tasks should reflect common use cases of  the software
+\item Include tasks that require a set of sequential or hierarchical steps to be completed
+
 \end {enumerate} 
-
-**Domain experts will be asked to identify what background knowledge is necessary for the suggested tasks (Expert, Intermediate, Novice).
+\oo   {Are domain experts expected to provide step by step instructions to complete a tasks}
 

### 2020-06-11 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
Minor notes added to quality definition document related to reproducibility

@@
 
-\noindent \textbf{Reasoning}
+The importance of reproducibility is seen by retractions of influential papers,
+like the review of
+\href{https://www.theguardian.com/commentisfree/2020/jun/05/lancet-had-to-do-one-of-the-biggest-retractions-in-modern-history-how-could-this-happen}
+{hydroxychloroquine trials}
+
+\noindent \textbf{Reasoning}
+
+\wss{Reproducibility literature review is given in \citet{FeinbergEtAl2020}.
+  Includes examples where replicability was found to be lacking.  Has a good
+  litereature review in Section 2.}
 

### 2020-06-11 OverallResearchProposal/ObjectivesAndResearchQuestions.tex
Addition of some ideas for the next set of edits to the research questions

@@
 
-\textbf{What is needed to produce sustainable \{software + artifacts\}?}\\
+\textbf{What is needed to produce \{software + artifacts\} sustainably?}\\
 
@@
 We define sustainable softifacts, as follows: \emph{Sustainable softifacts
-  satisfy, for a reasonable amount of energy, the software requirements for the
+  satisfy, for a reasonable amount of effort, the software requirements for the
   present, while also being maintainable, reusable and reproducible for the
@@
 effort, assuming that this alternative achieves the same (or a higher) level of
-sustainability as its competition.  The assumptions listed in the next section
+sustainability as its competition.  \wss{reasonable effort, without a large
+  cost, means productivity} The assumptions listed in the next section
 (Section~\ref{Sec_Context}) outline the typical space of problems where it makes
@@
 
+what else do we need - knowledge - knowledge captured in softifacts.  - overlap
+plus duplication - required and desired, but still want to be productive.  To be
+productive, want to say once and view as needed.  Suggests generation - maybe
+put this elsewhere?
+
 The objective is not phrased to target a specific software domain, but one of
@@
 section.  Roadmap.
+
+- what is needed to produce softifacts sustainably motivates our ideal process
+(sequence of choices).
 
@@
 
-\begin{itemize}
+- number or name assumptions
+
+\begin{enumerate}
 \item The software project is going to be long lived, where long life means at
@@
   medical imaging or computational medicine.
-\item The software explicitly, or implicitly, is part of a program family of
+\item \wss{this one is about feasibility - separate this out} The software explicitly, or implicitly, is part of a program family of
   related software.  When the program family assumption means that there are
@@
 \item The domain of the software project is well understood.  This assumption is
-  related to the oracle hypothesis.
-\end{itemize}
+  related to the oracle hypothesis. \wss{related to feasibility}
+\end{enumerate}
 
+- program family comes up to make it economically feasible, not necessary, but
+important (JC gave example of space probe - not a family, but still makes sense
+to do it right.)
+- sustainable - part of reasonable effort
 Although this study is not specifically restricted to open-source software, for

### 2020-06-12 StateOfPractice/Methodology/Methodology.tex
updated questionnaire section

@@
 \item Collectively all tasks should not take no more than 2 hours.
-\item Selected tasks should reflect common use cases of  the software
+\item Selected tasks should reflect common use cases of  the software.
 \item Include tasks that require a set of sequential or hierarchical steps to be completed
@@
 \end {enumerate}
-\oo   {Are domain experts expected to provide step by step instructions to complete a tasks}
 
 \subsection{Usability Questionnaire}
-This questions need to be reworded to make them as open ended as possible\\
-source \url{https://www.usabilitest.com/sus-pdf-generator}- 20-29
-\begin {enumerate}
-\item Overall, I am satisfied with how easy it is to use this system.
-\item It was simple to use this system.
-\item I could effectively complete the tasks and scenarios using this system.
-\item I was able to complete the tasks and scenarios quickly using this system.
-\item I was able to efficiently complete the tasks and scenarios using this system.
-\item I felt comfortable using this system.
-\item It was easy to learn to use this system.
-\item I believe I could become productive quickly using this system.
-\item The system gave error messages that clearly told me how to fix problems.
-\item Whenever I made a mistake using the system, I could recover easily and quickly.
-\item The information (such as online help, on-screen messages, and other documentation) provided with this system was clear.
-\item It was easy to find the information I needed.
-\item The information provided for the system was easy to understand.
-\item The information was effective in helping me complete the tasks and scenarios.
-\item The organization of information on the system screens was clear.
-\item The interface of this system was pleasant.
-\item I liked using the interface of this system.
-\item This system has all the functions and capabilities I expect it to have.
-\item Overall, I am satisfied with this system.
-\item I think that I would like to use this system frequently.
-\item I found the system unnecessarily complex.
-\item I thought the system was easy to use.
-\item I think that I would need the support of a technical person to be able to use this system.
-\item I found the various functions in this system were well integrated.
-\item I thought there was too much inconsistency in this system.
-\item I would imagine that most people would learn to use this system very quickly.
-\item I found the system very cumbersome to use.
-\item I felt very confident using the system.
-\item I needed to learn a lot of things before I could get going with this system.
-\item \textbf{The below were added to this list by Olu based on Nielson's Heuristics}
-\item The software provided me system's current state and what actions can be done or where to go or what change was made.
-\item I was able reuse a previously defined value without having to memorize or jot it.
-\item I was able to reverse an error and didn't have to restart a task due to error.
-\item I was able to refer to help/documentation easily when I needed it.
-\item I was pre-warned of errors based on the actions I performed
-\end {enumerate}
+Two sources of standardized usability questionnaire we could use.
+
+- \url{https://www.usabilitest.com/sus-pdf-generator}- 20-29 - SUS.\\
+- \url{https://uiuxtrend.com/pssuq-post-study-system-usability-questionnaire/} - PSSUQ
+
 \section{Analytic Hierarchy Process}

### 2020-06-12 OverallResearchProposal/ObjectivesAndResearchQuestions.tex

@@
 To fully explain this research objective, we will decompose each constituent
-part in turn, starting with the end of the sentence.  In the above objective,
-software consists of either programs, which run (execute) on a computer, or
-libraries, which provide services to be used by programs.  Both programs and
-libraries are created using computer code, written in a programming language.
-Many developers place most, if not all, of their emphasis on the computer code
-that makes up programs and libraries.  However, we consider the outcome of the
-development process to be more than just software and the associated computer
-code; we consider the products of the development process to be \{software +
-artifacts\}.  The artifacts include code, but also all the other potential work
-products, such as documentation, build scripts, test cases, contributor's
-guides, etc.  From this point onward, we will use the term \emph{softifacts}
-when we are referencing the combination of software and the associated
-artifacts.  As we will explain below, assessing the quality of sustainability is not
-possible by only considering the computer code; it depends on all of the
-artifacts, and on the processes and methodologies used during software
-development.  \wss{Can we consider software as both the computer code and the
-  executable (program or library), or should the term software just mean the
-  executable?  That is, when we mean code, should we be careful to not instead
-  say software?}
+part in turn.  In the above objective, \emph{software} consists of either
+programs, which run (execute) on a computer, or libraries, which provide
+services to be used by programs.  Both programs and libraries are created using
+computer code, written in a programming language.  Many developers place most,
+if not all, of their emphasis on the computer code that makes up programs and
+libraries.  However, we consider the outcome of the development process to be
+more than just software and the associated computer code; we consider the
+products of the development process to be \{software + artifacts\}.  The
+\emph{artifacts} include code, but also all the other potential work products,
+such as documentation, build scripts, test cases, contributor's guides, etc.
+From this point onward, we will use the term \emph{softifacts} when we are
+referencing the combination of software and the associated artifacts.  As we
+will explain below, assessing the quality of sustainability is not possible by
+only considering the computer code; it depends on all of the artifacts, and on
+the processes and methodologies used during software development.  \wss{Can we
+  consider software as both the computer code and the executable (program or
+  library), or should the term software just mean the executable?  That is, when
+  we mean code, should we be careful to not instead say software?}
 
 We define sustainable softifacts, as follows: \emph{Sustainable softifacts
-  satisfy, for a reasonable amount of effort, the software requirements for the
+  satisfy, with a reasonable amount of effort, the software requirements for the
   present, while also being maintainable, reusable and reproducible for the
@@
 effort, assuming that this alternative achieves the same (or a higher) level of
-sustainability as its competition.  \wss{reasonable effort, without a large
-  cost, means productivity} The assumptions listed in the next section
-(Section~\ref{Sec_Context}) outline the typical space of problems where it makes
-sense to aim for sustainable softifacts.
+sustainability as its competition.  Achieving a reasonable effort means that
+sustainability depends on the productivity the adopted development and
+maintenance process.  The assumptions listed in the next section
+(Section~\ref{Sec_Context}) outline the typical space of problems where it
+sustainable softifacts are necessary, and what problem constraints suggest the
+possibility of a reasonable level of effort.
 
-The overall objectives asks what is needed for sustainability.  This ``what''
-can be decomposed further.  We consider the ``what'' to be the principles,
-processes and methodologies used for software development.  We exclude tools
-from this list because, although tools are needed to support processes and
-methodologies, tools change rapidly.  Our goal is to be general.  If the right
-principles, processes and methodologies are found, the tools will adapt to
-match.  However, understanding the current state of the practice will involve
-looking at tools, as shown in the later list of research questions
-(Section~\ref{Sec_ResearchQuestions}).
+The overall objectives asks what is needed for produce softifacts sustainably.
+We consider the \emph{what} to be the principles, processes and methodologies
+used for software development.  We exclude tools from this list because,
+although tools are needed to support processes and methodologies, tools change
+rapidly.  Our goal is to be general.  If the right principles, processes and
+methodologies are found, the tools will adapt to match.  However, understanding
+the current state of the practice will involve looking at tools, as shown in the
+later list of research questions (Section~\ref{Sec_ResearchQuestions}).  As
+mentioned above, for sustainability to be achieved, the productivity of the
+adopted process must be considered.
 
@@
 
+Traditional approach struggles.  Try generative.  Does generative supply what is
+needed?  Is there an ideal process?  - what is needed to produce softifacts
+sustainably motivates our ideal process (sequence of choices).
+
 Further information on the relevant definitions and context is given in the next
 section.  Roadmap.
-
-- what is needed to produce softifacts sustainably motivates our ideal process
-(sequence of choices).
 

### 2020-06-16 OverallResearchProposal/ObjectivesAndResearchQuestions.tex
Updates to objectives section, more explicitly splitting softifacts (knowledge) from process

@@
 
-\subsection* {Overall Objective}
+\section {Overall Objective}
 
@@
 The overall objectives asks what is needed for produce softifacts sustainably.
-We consider the \emph{what} to be the principles, processes and methodologies
-used for software development.  We exclude tools from this list because,
-although tools are needed to support processes and methodologies, tools change
-rapidly.  Our goal is to be general.  If the right principles, processes and
-methodologies are found, the tools will adapt to match.  However, understanding
-the current state of the practice will involve looking at tools, as shown in the
-later list of research questions (Section~\ref{Sec_ResearchQuestions}).  As
-mentioned above, for sustainability to be achieved, the productivity of the
-adopted process must be considered.
+The \emph{what} is found by answering two questions:
 
-what else do we need - knowledge - knowledge captured in softifacts.  - overlap
+\begin{enumerate}
+\item What softifacts, in addition to the code, are needed?
+\item What software engineering principles, processes and methodologies should
+  be employed?
+\end{enumerate}
+
+To determine what softifacts are necessary for sustainability, we need to
+determine what knowledge needs to be captured and documented.  The knowledge
+comes from the problem domain, from computing, from documentation rules and from
+the development process.  We also need to determine how to document the
+softifacts.  That is, what information should appear in which artifacts?, how
+should it be formatted and structured?
+
+The adopted software engineering principles, processes and methodologies
+significantly contribute to sustainability.  We exclude tools from this list
+because, although tools are needed to support processes and methodologies, tools
+change rapidly.  Our goal is to be general.  If the right principles, processes
+and methodologies are found, the tools will adapt to match.  However,
+understanding the current state of the practice will involve looking at tools,
+as shown in the later list of research questions
+(Section~\ref{Sec_ResearchQuestions}).  For sustainability to be achieved, the
+productivity of the adopted process must be considered.  The process should lead
+to high quality artifacts with, as mentioned earlier, a reasonable amount of
+effort.
+
+To be successful, the process, methods and tools, need:
+
+to avoid repetition (knowledge should only be entered once)
+reduce/remove drudgery
+provide flexibility for change
+I think there are more things in this list ...
+
+This leads to our hypothesis: We can use a process that uses knowledge capture
+and artifact generation to avoid repetition, reduce drudgery and provide
+flexibility.  We have been calling this process MDE, although I'm not sure
+@carette is happy with that terminology.  For simplicity though, I'll use that
+term in this message.
+
+If MDE satisfies the qualities of a sustainable softifact generating process,
+then following the bullet points backwards supports the argument that we can
+produce softifacts sustainably using MDE.  This argument is just on the process
+though; MDE doesn't really dictate what artifacts we want.  That is a separate
+research question.
+
+- overlap
 plus duplication - required and desired, but still want to be productive.  To be
@@
 Further information on the relevant definitions and context is given in the next
-section.  Roadmap.
+section.  Roadmap.  Research questions for softifacts and knowledge.  Research
+questions for the process.
 
-\subsection*{Context, Definitions and Assumptions} \label{Sec_Context}
+\section{Good Research Questions}
+
+The criteria for a good research question are :
+
+Focused on a single problem or issue
+Researchable using primary and/or secondary sources
+Feasible to answer within the timeframe and practical constraints
+Specific enough to answer thoroughly
+Complex enough to develop the answer over the space of a paper or thesis
+Relevant to your field of study and/or society more broadly
+
+\section {Context, Definitions and Assumptions} \label{Sec_Context}
 
@@
 
-\subsubsection*{Definitions}
+\subsection{Definitions}
 
@@
 
-\subsubsection*{Assumptions}
+\subsection{Assumptions}
 
@@
 
-\subsubsection*{Current State of the Practice Research Questions} \label{Sec_ResearchQuestions}
+\subsection{Current State of the Practice Research Questions} \label{Sec_ResearchQuestions}
 
@@
 
-\subsubsection*{Ideal Process}
+\subsection{Ideal Process}
 

### 2020-06-17 OverallResearchProposal/ObjectivesAndResearchQuestions.tex
Further edits, nothing major

@@
 
-Focused on a single problem or issue
-Researchable using primary and/or secondary sources
-Feasible to answer within the timeframe and practical constraints
-Specific enough to answer thoroughly
-Complex enough to develop the answer over the space of a paper or thesis
-Relevant to your field of study and/or society more broadly
+\begin{itemize}
+\item Focused on a single problem or issue
+\item Researchable using primary and/or secondary sources
+\item Feasible to answer within the timeframe and practical constraints
+\item Specific enough to answer thoroughly
+\item Complex enough to develop the answer over the space of a paper or thesis
+\item Relevant to your field of study and/or society more broadly
+\end{itemize}
 

### 2020-06-21 StateOfPractice/Ao-Notes/Questions to Developers.tex
Renamed documents
update questions to developers

--- 2020-05-13 StateOfPractice/Ao-Notes/Questionnaire.tex
@@
 \usepackage{hyperref}
+\usepackage[normalem]{ulem}
 
@@
 
-\title{Questionnaire to Developers} 
+\title{Questions to Developers} 
 \author{Ao Dong}
@@
 has been solved? How did you solve it?
+\item What is your software development model? For example, waterfall, agile, etc.
 \item What is your project management process? Do you think improve this process
-can tackle the current problem?
+can tackle the current problem? Were any project management tools used?
 \item What is your documentation process? Do you think improve this process can
 tackle the current problem?
-\item Can you name several software qualities that are the most important to
-this project?
-\item Is any of them hard to be achieved? If yes, what methods have considered
-or practiced to improve the situation? If practiced, does it work?
+\item \sout{Can you name several software qualities that are the most important to
+this project?}
+\item Was it hard to ensure the correctness of the software? If there were any
+ obstacles, what methods have been considered or practiced to improve the situation?
+  If practiced, did it work?
+\item When designing the software, did you consider the ease of future changes?
+ For example, will it be hard to change the structure of the system, modules or code
+  blocks? What measures have been taken to ensure the ease of future changes
+   and maintains?
 \item Do you think the users can understand your software well? Has there been

### 2020-07-01 StateOfPractice/Ao-Notes/Meeting Agenda with Domain Experts.tex
updated and renamed the Meeting Agenda with Domain Experts

\documentclass[12pt]{article}

\usepackage{natbib}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{xspace}
\usepackage{xcolor}
\usepackage{fullpage}
\usepackage{hyperref}

\hypersetup{
colorlinks=true,       % false: boxed links; true: colored links
linkcolor=red,          % color of internal links (change box color with
%linkbordercolor)
citecolor=blue,       % color of links to bibliography
filecolor=magenta,   % color of file links
urlcolor=cyan           % color of external links
}

%% Comments
\newif\ifcomments\commentstrue

\ifcomments
\newcommand{\authornote}[3]{\textcolor{#1}{[#3 ---#2]}}
\newcommand{\todo}[1]{\textcolor{red}{[TODO: #1]}}
\else
\newcommand{\authornote}[3]{}
\newcommand{\todo}[1]{}
\fi

\newcommand{\wss}[1]{\authornote{blue}{SS}{#1}} %Spencer Smith
\newcommand{\jc}[1]{\authornote{red}{JC}{#1}} %Jacques Carette
\newcommand{\oo}[1]{\authornote{magenta}{OO}{#1}} %Olu Owojaiye
\newcommand{\pmi}[1]{\authornote{green}{PM}{#1}} %Peter Michalski
\newcommand{\ad}[1]{\authornote{brown}{AD}{#1}} %Ao Dong

%% Global noindent
% \setlength\parindent{0pt}

\title{Meeting Agenda with Domain Experts} 
\author{Ao Dong}
\date{\today}

\begin{document}

\maketitle

\section{Introduction}
Presenter: \ad{TBD}

A brief introduction about the project. \ad{To be written}

\subsection{About this Meeting}
Presenter: \ad{TBD}

The meeting will be roughly one hour. We will follow this document and go
through all the steps.

In the interest of time, some topics will only be discussed briefly during this
meeting, and tasks will be assigned to domain experts to finish later. For
example, Task \ref{task_overall_objective} in Section \ref{project_objectives}.

We will create issues on GitHub to track the progress of the tasks. Details can
be found in Appendix \ref{proj_mgmt_tools}. The tasks can also be found in
Appendix \ref{tasks_domain_experts}.

\subsection{Project Objectives}
\label{project_objectives}
Presenter: Dr.\ Spencer Smith and Dr.\ Jacques Carette \ad{TBD}

The overall objectives are documented in the section Overall Objective of
\href{https://github.com/smiths/AIMSS/blob/master/OverallResearchProposal/ObjectivesAndResearchQuestions.pdf}{Objectives
And Research Questions}.

A brief introduction. \ad{To be written}

Do you have any suggestions? Feedback can be provided later in Task
\ref{task_overall_objective}.

\subsection{Research Questions}
\label{research_questions}
Presenter: Dr.\ Spencer Smith \ad{TBD}

The research questions are documented in the section Current State of the
Practice Research Questions of
\href{https://github.com/smiths/AIMSS/blob/master/OverallResearchProposal/ObjectivesAndResearchQuestions.pdf}{Objectives
And Research Questions}.

A brief introduction. \ad{To be written}

Do you have any thoughts on one of the questions - What are the ``pain points''
for developers working on SCS projects? What aspects of the existing processes,
methodologies and tools do they consider could potentially be improved?

Do you have any suggestions? Feedback can be provided later in Task
\ref{task_research_questions}.

\subsection{Research Proposal}
\label{research_proposal}
Presenter: Dr.\ Spencer Smith \ad{TBD}

The research proposal is documented in
\href{https://github.com/smiths/AIMSS/blob/master/OverallResearchProposal/ResearchProposal.pdf}{Research
Proposal}.

A brief introduction. \ad{To be written}

Do you have any suggestions? Feedback can be provided later in Task
\ref{task_research_proposal}.

\section{Software List}
Presenter: Ao or Peter \ad{TBD}

In this section, we would like to discuss the list of software packages and the
process of creating and shortening it.

\subsection{Top 10 List}
\label{top_10_list}
What will be the top 10 software of your choice? Feedback can be provided later
in Task \ref{task_top_10_list}.

\subsection{Domain Analysis}
The domain analysis can be found in Section 3.2 of
\href{https://github.com/Ao99/MISEG/blob/master/docs/SRS/SRS.pdf}{SRS for
Medical Imaging Software} or
\href{https://github.com/peter-michalski/LatticeBoltzmannSolvers/blob/master/docs/SRS/CA.pdf}{CA
of Lattice Boltzmann Solvers}.

What do you think of our analysis?

\subsection{Selecting Process}
The process of creating and shortening the list of software packages is
documented in
\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Ao-Notes/Ao-Notes.pdf}{Ao's
Notes} or Peter's Notes \ad{TBD}

What do you think of our process of selecting software candidates?

\subsection{Software List}
\label{software_list}
The software list is
\href{https://docs.google.com/spreadsheets/d/122wU0v3ZtvDcqy8C4zKJ89kU-8fXAbo3Mzn6vcVXOi0/edit?usp=sharing}{Medical
Imaging Software List} or Lattice Boltzmann Solver List. \ad{TBD}

What do you think of the list of the selected software?

Is there any software you would like to remove from or add to the list?

Feedback can be provided later in Task \ref{task_software_list}.

\section{Qualities \& Measurements}

\subsection{Measurement Template}
\label{measurement_template}
Presenter: \ad{TBD}

The template for software quality measurements can be found at
\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Combined_MeasurementTemplate_EmpiricalMeasures.xlsx}{Measurement
Template}.

Do you have any suggestions? Feedback can be provided later in Task
\ref{task_measurement_template}.

\subsection{Survey for Short List Projects}
\label{survey_short_list_projects}
Presenter: \ad{TBD}

After the first round of measurement, we would like to select a short list of
projects (about 5 of them for each domain) and directly ask questions to the
developers.

We would like to do the selection based on the results of the first round
measurement, such as selecting the top 5 ones with highest scores. However, we
can also include 1 or 2 most recommended ones from you. What do you think of
this selection process?

The survey questions to the developers can be found at
\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Ao-Notes/Questions%20to%20Developers.pdf}{Questions
to Developers}.

Do you have any suggestions? Feedback can be provided later in Task
\ref{task_survey_short_list_projects}

\subsection{Usability Tests}
\label{usability_tests}
Presenter: Olu \ad{TBD}

A brief introduction. \ad{To be written}

The usability test can be found at ... \ad{TBD}

Do you have any suggestions? Feedback can be provided later in Task
\ref{task_usability_tests}

\section{Publication}
Presenter: Dr.\ Spencer Smith and Dr.\ Jacques Carette \ad{TBD}

\begin{itemize}
\item Where should we publish this paper?
\item Who are our targeted readers?
\end{itemize}

\section{Schedule \& Team}
Presenter: \ad{TBD}

\begin{itemize}
\item We propose to start this project from ... and finish it before ... . What
do you think?
\item Our meeting schedule is ... does it fit your plan?
\item We need ... extra members from your team, each working ... hours per week
for ... weeks. What do you think?
\item Is there any other experts you think we should keep them involved?
\end{itemize}

\appendix
\section{Project Management Tools}
\label{proj_mgmt_tools}
We would like to use GitHub for task tracking and document version control. The
repository for this project is
\href{https://github.com/smiths/AIMSS}{https://github.com/smiths/AIMSS}.

Tasks will be created as issues on the repository and assigned to individual
stakeholders. An open issue usually means a task to be fulfilled, and it can be
closed once it is finished. Here is an example of an issue
\href{https://github.com/smiths/AIMSS/issues/19}{https://github.com/smiths/AIMSS/issues/19}.

\section{Tasks for Domain Experts}
\label{tasks_domain_experts}

\subsection{Review the Overall Objective}
\label{task_overall_objective}
GitHub Issue:
\href{https://github.com/smiths/AIMSS/issues}{https://github.com/smiths/AIMSS/issues}
\ad{To be created}

\noindent Back to Section \ref{project_objectives}

\subsection{Review the Research Questions}
\label{task_research_questions}
GitHub Issue:
\href{https://github.com/smiths/AIMSS/issues}{https://github.com/smiths/AIMSS/issues}
\ad{To be created}

\noindent Back to Section \ref{research_questions}

\subsection{Review the Research Proposal}
\label{task_research_proposal}
GitHub Issue:
\href{https://github.com/smiths/AIMSS/issues}{https://github.com/smiths/AIMSS/issues}
\ad{To be created}

\noindent Back to Section \ref{research_proposal}

\subsection{Create a Top 10 List}
\label{task_top_10_list}
GitHub Issue:
\href{https://github.com/smiths/AIMSS/issues}{https://github.com/smiths/AIMSS/issues}
\ad{To be created}

\noindent Back to Section \ref{top_10_list}

\subsection{Review the Software List}
\label{task_software_list}
GitHub Issue:
\href{https://github.com/smiths/AIMSS/issues}{https://github.com/smiths/AIMSS/issues}
\ad{To be created}

\noindent Back to Section \ref{software_list}

\subsection{Review the Measurement Template}
\label{task_measurement_template}
GitHub Issue:
\href{https://github.com/smiths/AIMSS/issues}{https://github.com/smiths/AIMSS/issues}
\ad{To be created}

\noindent Back to Section \ref{measurement_template}

\subsection{Review the Survey for Short List Projects}
\label{task_survey_short_list_projects}
GitHub Issue:
\href{https://github.com/smiths/AIMSS/issues}{https://github.com/smiths/AIMSS/issues}
\ad{To be created}

\noindent Back to Section \ref{survey_short_list_projects}

\subsection{Review the Usability Tests}
\label{task_usability_tests}
GitHub Issue:
\href{https://github.com/smiths/AIMSS/issues}{https://github.com/smiths/AIMSS/issues}
\ad{To be created}

\noindent Back to Section \ref{usability_tests}

\end{document}

### 2020-07-01 StateOfPractice/Ao-Notes/Questions to Developers.tex
updated questions to developers

@@
 \item What is the typical background of a developer?
-\item How large is the user group?
+\item How large is the user group, and how do you estimate the user group size?
 \item What is the typical background of a user?
@@
 development process?
+\item What is your documentation process? Do you think improve this process can
+solve this problem?
 \item In the past, is there any major obstacle to your development process that
@@
 can tackle the current problem? Were any project management tools used?
-\item What is your documentation process? Do you think improve this process can
-tackle the current problem?
 \item \sout{Can you name several software qualities that are the most important to
@@
 improvements are needed?
+\item Do you have any concern that your computational results won't be
+reproducible in the future? Have you taken any steps to ensure reproducibility?
 \end{enumerate}

### 2020-07-01 StateOfPractice/Ao-Notes/Meeting Agenda with Domain Experts.tex
fixed several grammar mistakes

@@
 
-A brief introduction about the project. \ad{To be written}
+A brief introduction to the project. \ad{To be written}
 
@@
 We would like to do the selection based on the results of the first round
-measurement, such as selecting the top 5 ones with highest scores. However, we
+measurement, such as selecting the top 5 ones with the highest scores. However, we
 can also include 1 or 2 most recommended ones from you. What do you think of
@@
 for ... weeks. What do you think?
-\item Is there any other experts you think we should keep them involved?
+\item Are there any other experts you think we should keep them involved?
 \end{itemize}

### 2020-07-22 StateOfPractice/Ao-Notes/Questions to Developers.tex

@@
 \usepackage{xspace}
+\usepackage{xcolor}
 \usepackage{fullpage}
@@
 \newcommand{\pmi}[1]{\authornote{green}{PM}{#1}} %Peter Michalski
-\newcommand{\ad}[1]{\authornote{cyan}{AD}{#1}} %Ao Dong
+\newcommand{\ad}[1]{\authornote{brown}{AD}{#1}} %Ao Dong
 
@@
 \item How large is the development group?
+\item Do you have a defined process for accepting new contributions into your team?
 \item What is the typical background of a developer?
-\item How large is the user group, and how do you estimate the user group size?
+\item What is your estimated number of users? How did you come up with that estimate?
 \item What is the typical background of a user?
@@
 
-\ad{I drafted another version without considering our measurements}
-
 \noindent\ad{3 aspects: dev process, software quality, user feedback}
@@
 \begin{enumerate}
-\item Currently, what is the biggest obstacle against your development process?
-\item In order to solve this problem, is there any improvement needed for your
-development process?
-\item What is your documentation process? Do you think improve this process can
-solve this problem?
+\item Currently, what are the most significant obstacles in your development
+process?
+\item How might you change your development process to remove or reduce these
+obstacles?
+\item How does documentation fit into your development process? Would improved
+documentation help with the obstacles you typically face?
 \item In the past, is there any major obstacle to your development process that

### 2020-08-03 StateOfPractice/AHP2020/LBM/Source code/DataProcess/src/Main.java
added AHP code for LBM

@@ -1,0 +1,146 @@
+import java.io.File; 
+import java.io.FileInputStream;
+import java.io.FileOutputStream;
+import java.util.Iterator;
+
+import javax.swing.JOptionPane;
+
+import org.apache.poi.ss.usermodel.Cell;
+import org.apache.poi.ss.usermodel.Row;
+import org.apache.poi.xssf.usermodel.XSSFSheet;
+import org.apache.poi.xssf.usermodel.XSSFWorkbook;
+
+public class Main {
+	
+	public static void main(String[] args)
+    {
+		int numberQualities = 9;
+		int numberSolvers = 23; //also need to change rownum below
+		Cell[][] table=new Cell[numberQualities][numberSolvers];
+		Double[][][] ahpTable= new Double[numberQualities][numberSolvers][numberSolvers-1];
+        try
+        {
+            FileInputStream file = new FileInputStream(new File("AHP_Template.xlsx"));
+ 
+            //Create Workbook instance holding reference to .xlsx file
+            XSSFWorkbook workbook = new XSSFWorkbook(file);
+ 
+            //Get second sheet from the workbook
+            XSSFSheet sheet = workbook.getSheetAt(1);
+ 
+            //Iterate through each rows one by one
+            Iterator<Row> rowIterator = sheet.iterator();            
+            int r=0;
+            int col=0;
+            while (rowIterator.hasNext())
+            {
+                Row row = rowIterator.next();
+                //For each row, iterate through all the columns
+                Iterator<Cell> cellIterator = row.cellIterator();
+                 
+                while (cellIterator.hasNext())
+                {
+                    Cell cell = cellIterator.next();
+                    //Check the cell type and format accordingly
+                    switch (cell.getCellType())
+                    {
+                        case Cell.CELL_TYPE_NUMERIC:
+                        	table[r][col]=cell;    
+                        	col++;
+                            break;
+                        case Cell.CELL_TYPE_STRING:
+                            break;
+                        default:
+                           	break;
+                    }                    
+                }
+                r++;
+                col=0;
+            }
+            XSSFSheet sheetnew;
+            if(workbook.getSheet("AHP Example")==null){
+            	sheetnew= workbook.createSheet("AHPDATA");
+            }
+            else{
+            	sheetnew=workbook.getSheet("AHP Example");
+            }
+            double x=0;
+            double y=0;
+            for(int i=0;i<numberSolvers;i++)
+            {
[…]

### 2020-08-05 StateOfPractice/Methodology/Methodology.tex
added some points and questions to 'Overview of Steps in Assessing Quality of the Domain Software'

@@
 \item Identify domain.  (Provide criteria on a candidate domain.)
-\item Identify list of candidate software in the domain.  (Section~\ref{SecIdentifyCandSoft})
-\item \emph{Domain Expert}: Vet domain software list.  
+\item Identify list of candidate software in the domain. \pmi{Do we want to outline how to find such a list?} (Section~\ref{SecIdentifyCandSoft})
+\item Filter domain software list.
+(Judge on professionalism, available documentation, status, availability of source code) \pmi{The original document lists the Domain Expert as responsible for this.
+Should we have had them vet the list prior to taking our measurements?}
 \item Domain Analysis. (Section~\ref{SecDomainAnalysis})
-\item \emph{Domain Expert}: Vet domain analysis
-\item Measure using ``shallow'' measurement template.
+\item \emph{Domain Expert}: Vet domain analysis.
+\item Gather source code and documentation.
 \item Empirical measures.
+\item Measure using ``shallow'' measurement template. \pmi{Do we want to comment on the creation of this template and why certain qualities were chosen?}
 \item Use AHP process to rank the software.  Tool for estimating AHP results?
   Tool for ranking and sorting by various criteria?  Conditional formula to
-  assess software within each measure, or just use opinion of measurer?
-\item Identify short list for deeper exploration.
+  assess software within each measure, or just use opinion of measurer? \pmi{Will we want to rank the qualities before we implement AHP? Will we provide our AHP tool?}
+\item Decide on desired short list size.  
+\item Identify short list for deeper exploration using AHP rankings of shallow measurements.
 \item \emph{Domain Expert}: vet AHP ranking and short list.
@@
 \end{enumerate}
-\item Compare shallow and deep measures and ranking
-\item Answers for research questions.
+\item Rank short list using XXXXXX \pmi{How will we use the above step to rank the short list?}
+\item Compare shallow and deep measures and ranking.
+\item Document answers for research questions.
 \end{enumerate}
@@
 \begin{enumerate}
-\item Must be open source.
-\item Must have GitHub repository.
+\item Must have viewable source code.
+\item Ideally have a git repository. \pmi{Not all of mine had this..}
 \end{enumerate}

### 2020-08-20 StateOfPractice/Methodology/Methodology.tex
further revisions to overview of steps in assessing quality of the domain software; moved Meeting Agenda files to Methodology folder

--- 2020-06-12 StateOfPractice/Methodology/Methodology.tex
@@
 \begin{enumerate}
-\item Start with state of practice research questions.
-\item Identify domain.  (Provide criteria on a candidate domain.)
-\item Identify list of candidate software in the domain.  (Section~\ref{SecIdentifyCandSoft})
-\item \emph{Domain Expert}: Vet domain software list.  
-\item Domain Analysis. (Section~\ref{SecDomainAnalysis})
-\item \emph{Domain Expert}: Vet domain analysis
-\item Measure using ``shallow'' measurement template.
-\item Empirical measures.
-\item Use AHP process to rank the software.  Tool for estimating AHP results?
-  Tool for ranking and sorting by various criteria?  Conditional formula to
-  assess software within each measure, or just use opinion of measurer?
-\item Identify short list for deeper exploration.
-\item \emph{Domain Expert}: vet AHP ranking and short list.
+\item Start with state of practice research questions.\pmi{add link to above section}
+\item Identify the domain.  (Provide criteria on a candidate domain.) \pmi{as an example what criteria? Do we need to add a section below explaining this?}
+\item \emph{Domain Expert}: Create a top ten list of software packages in the domain. \pmi{The details of this will be specified in the Meeting Agenda, so link to that section}
+\item Brief the Domain Expert on the overall objective, research proposal, research questions, measurement template, survey for short list projects, usability tests, performance benchmarks, maintainability experiments.\pmi{Link this to Meeting Agenda, and create a small summary below; Reword or remove anything from this sentence?}
+\item Identify broad list of candidate software packages in the domain. (Section~\ref{SecIdentifyCandSoft}) \pmi{To be completed}
+\item Preliminary filter of software packages list. \pmi{Add section below; Judge on professionalism, available documentation, status, availability of source code}
+\item \emph{Domain Expert}: Review domain software list. \pmi{any comments on this? Does this need a section below? Link to appropriate section of Meeting Agenda?}
+\item Domain Analysis. (Section~\ref{SecDomainAnalysis})\pmi{To be completed}
+\item \emph{Domain Expert}: Vet domain analysis.\pmi{This is currently not listed in Meeting Agenda and was part of the original Steps in Assessing Quality list.
+We need to add it to the Meeting Agenda and link it from here}
+\item Gather source code and documentation for each prospective software package.
+\item Collect Empirical measures. \pmi{To do: Link to below and edit the section}
+\item Measure using ``shallow'' measurement template. \pmi{Add section below and link to actual template}
+\item Use AHP process to rank the software packages. \pmi{Link to below section (which needs to be edited) outlining the tool we used, and link the below section to the AHP README file}
+\item Identify a short list of five software packages for deeper exploration according to the AHP rankings of the shallow measurements.
+\item \emph{Domain Expert}: vet AHP ranking and short list.\pmi{Currently not listed in Meeting Agenda, add it and link to it?
+...or is this redundant considering they have made a top ten list and have already vetted the original prospective list (both of these latter Domain Expert actions are listed in the Meeting Agenda)?}
 \item With short list:
 \begin{enumerate}
-\item Survey developers
-\item Usability experiments
-\item Performance benchmarks
-\item Maintainability experiments
+\item Survey developers \pmi{To do: Link to survey; Add brief below?}
+\item Usability experiments \pmi{Link to section below; link to any other document?}
+\item Performance benchmarks\pmi{note: this is still in consideration}
+\item Maintainability experiments\pmi{note: this is still in consideration}
 \end{enumerate}
-\item Compare shallow and deep measures and ranking
-\item Answers for research questions.
+\item Rank short list using pairwise comparison of shortlist results.\pmi{Add section below? Link to another doc?}
+\item Compare shallow and deep measures and ranking.\pmi{Add section below? Link to another doc? Anything specific to consider when comparing?}
+\item Document answers for research questions.
 \end{enumerate}
@@
 
-\section{Identify Candidate Software} \label{SecIdentifyCandSoft}
-
+\section{How to Identify Candidate Software} \label{SecIdentifyCandSoft}
+\pmi{outline here how to find such a list}
 \begin{enumerate}
-\item Must be open source.
-\item Must have GitHub repository.
+\item Must have viewable source code.
+\item Ideally have a git repository. \pmi{Not all of mine had this..}
 \end{enumerate}

### 2020-08-29 StateOfPractice/Methodology/Methodology.tex
updated 'Steps in Assessing Qualilty'

--- 2020-06-12 StateOfPractice/Methodology/Methodology.tex
@@
 
-\section{Research Questions}
+\section{Research Questions}\label{ResearchQuestions}
 
@@
 \begin{enumerate}
-\item Start with state of practice research questions.
-\item Identify domain.  (Provide criteria on a candidate domain.)
-\item Identify list of candidate software in the domain.  (Section~\ref{SecIdentifyCandSoft})
-\item \emph{Domain Expert}: Vet domain software list.  
-\item Domain Analysis. (Section~\ref{SecDomainAnalysis})
-\item \emph{Domain Expert}: Vet domain analysis
-\item Measure using ``shallow'' measurement template.
-\item Empirical measures.
-\item Use AHP process to rank the software.  Tool for estimating AHP results?
-  Tool for ranking and sorting by various criteria?  Conditional formula to
-  assess software within each measure, or just use opinion of measurer?
-\item Identify short list for deeper exploration.
-\item \emph{Domain Expert}: vet AHP ranking and short list.
+\item Start with state of practice research questions. (Section~\ref{ResearchQuestions}) \pmi{To be completed}
+\item Identify the domain. (Section~\ref{SecIdentifyDomain}) \pmi{To be completed}
+\item \emph{Domain Experts}: Create a top ten list of software packages in the domain. (\href{run:Meeting Agenda with Domain Experts.pdf}{Meeting Agenda with Domain Experts})
+\item Brief the Domain Experts on the overall objective, research proposal, research questions, measurement template, survey for short list projects, usability tests, performance benchmarks, maintainability experiments.
+(\href{run:Meeting Agenda with Domain Experts.pdf}{Meeting Agenda with Domain Experts})
+\item Identify broad list of candidate software packages in the domain. (Section~\ref{SecIdentifyCandSoft}) \pmi{To be completed}
+\item Preliminary filter of software packages list. (Section~\ref{SecInitialFilter}) \pmi{To be completed}
+\item \emph{Domain Experts}: Review domain software list. (\href{run:Meeting Agenda with Domain Experts.pdf}{Meeting Agenda with Domain Experts})
+\item Domain Analysis. (Section~\ref{SecDomainAnalysis})\pmi{To be completed}
+\item \emph{Domain Experts}: Vet domain analysis.
+(\href{run:Meeting Agenda with Domain Experts.pdf}{Meeting Agenda with Domain Experts}) \pmi{This was part of the original Steps in Assessing Quality list.
+We need to add it to the Meeting Agenda}
+\item Gather source code and documentation for each prospective software package.
+\item Collect empirical measures. (Section~\ref{SecEmpiricalMeasures}) \pmi{To be completed}
+\item Measure using ``shallow'' measurement template. (Section~\ref{SecShallowMeasure}) \pmi{To be completed}
+\item Use AHP process to rank the software packages. (Section~\ref{SecAHP}) \pmi{To be completed}
+\item Identify a short list of top software packages, typically four to six, for deeper exploration according to the AHP rankings of the shallow measurements.
+\item \emph{Domain Experts}: Vet AHP ranking and short list.
+(\href{run:Meeting Agenda with Domain Experts.pdf}{Meeting Agenda with Domain Experts}) \pmi{This was part of the original Steps in Assessing Quality list.
+We need to add it to the Meeting Agenda}
 \item With short list:
 \begin{enumerate}
-\item Survey developers
-\item Usability experiments
-\item Performance benchmarks
-\item Maintainability experiments
+\item Survey developers \pmi{To do: Link to survey}
+\item Usability experiments \pmi{Move the below usability section to a new doc, and link to it.}
+\item Performance benchmarks\pmi{note: this is still in consideration}
+\item Maintainability experiments\pmi{note: this is still in consideration}
 \end{enumerate}
-\item Compare shallow and deep measures and ranking
-\item Answers for research questions.
+\item Rank short list. (Section~\ref{SecRankShortList}) \pmi{To be completed}
+\item Document answers for research questions.
 \end{enumerate}
@@
 
-\section{Identify Candidate Software} \label{SecIdentifyCandSoft}
+\section{How to Identify the Domain} \label{SecIdentifyDomain}
+\begin{enumerate}
+	\item Needs to be "a community of"/many people using and studying this domain
+	\item There needs to be "many" software solutions
+	\item There needs to be a user community using these solutions
+	\item Must be in the research software scope
+	\item Specific enough scope
+	\item Must be a well understood domain / mature
+\end{enumerate}
 
+\section{How to Identify Candidate Software} \label{SecIdentifyCandSoft}
+\pmi{outline here how to find such a list}
 \begin{enumerate}
-\item Must be open source.
-\item Must have GitHub repository.
+\item Must have viewable source code.
+\item Ideally have a git repository. \pmi{Not all of mine had this..}
+\end{enumerate}
+
+\section{How to Initially Filter the Software List} \label{SecInitialFilter}
+The person doing this must keep track of both the initial and filtered lists. The reasons for this is traceability bc..
+\begin{enumerate}
+	\item Professionalism..
+	\item Available documentation..
+	\item Status..
+	\item Availability of source code..
 \end{enumerate}
@@
 
-\section{Empirical Measures}
+\section{Empirical Measures} \label{SecEmpiricalMeasures}
 
@@
 prints the results directly in the console.
+
+\section{Measure Using Shallow Measurement Template} \label{SecShallowMeasure}
+This step...
+The measurement template can be found \href{run:Combined_MeasurementTemplate_EmpiricalMeasures.xlsx}{here}.
+\begin{enumerate}
+	\item Step 1..
+\end{enumerate}
 
@@
 \subsection{Task selection criteria}
-**The task selection will be determined with the aid of the domain expert attached to any of the selected projects.\\ 
-**The domain expert will be asked to consider the below criteria when defining a task. \\
+**The task selection will be determined with the aid of the domain experts attached to any of the selected projects.\\ 
+**The domain experts will be asked to consider the below criteria when defining a task. \\
 **Domain experts will also be asked to identify what background knowledge is necessary for the suggested tasks - Novice, Intermediate, Advanced
@@
 
-\section{Analytic Hierarchy Process}
+\section{Analytic Hierarchy Process} \label{SecAHP}
 
-Describe process.  Domain expert review.
+Describe process. Domain experts review. Outline the tool we used. The README file of the tool can be found \href{run:../AHP2020/LBM/README.txt}{here}.
+
+\section{Rank Short List} \label{SecRankShortList}
+Rank using pairwise comparison of short list results.
+\begin{enumerate}
+	\item Compare with respect to usability
+	\item ...
+\end{enumerate}
 

### 2020-08-30 StateOfPractice/Methodology/Methodology.tex
updated 'How to Identify the Domain' in Methodology document

@@
 \begin{enumerate}	
-	\item Needs to be "a community of"/many people using and studying this domain
-	\item There needs to be "many" software solutions
-	\item There needs to be a user community using these solutions
-	\item Must be in the research software scope
-	\item Specific enough scope
-	\item Must be a well understood domain / mature
+	\item The domain must fall within the research software scope.
+	\item The domain must be specific.
+	\item The domain must be well understood.
+	\item There must be a community of people studying the domain.
+	\item There must to be a variety of software solutions for the domain.
+	\item These software solutions must have a user community.
 \end{enumerate}

### 2020-08-31 StateOfPractice/Methodology/Methodology.tex
updated 'Initially Filter the Software List' in Methodology document

@@
 \item Start with state of practice research questions. (Section~\ref{ResearchQuestions}) \pmi{To be completed}
-\item Identify the domain. (Section~\ref{SecIdentifyDomain}) \pmi{To be completed}
+\item Identify the domain. (Section~\ref{SecIdentifyDomain}) \pmi{To be reviewed}
 \item \emph{Domain Experts}: Create a top ten list of software packages in the domain. (\href{run:Meeting Agenda with Domain Experts.pdf}{Meeting Agenda with Domain Experts})
@@
 (\href{run:Meeting Agenda with Domain Experts.pdf}{Meeting Agenda with Domain Experts})
-\item Identify broad list of candidate software packages in the domain. (Section~\ref{SecIdentifyCandSoft}) \pmi{To be completed}
-\item Preliminary filter of software packages list. (Section~\ref{SecInitialFilter}) \pmi{To be completed}
+\item Identify broad list of candidate software packages in the domain. (Section~\ref{SecIdentifyCandSoft}) \pmi{To be reviewed}
+\item Preliminary filter of software packages list. (Section~\ref{SecInitialFilter}) \pmi{To be reviewed}
 \item \emph{Domain Experts}: Review domain software list. (\href{run:Meeting Agenda with Domain Experts.pdf}{Meeting Agenda with Domain Experts})
@@
 \section{How to Identify Candidate Software} \label{SecIdentifyCandSoft}
-\pmi{outline here how to find such a list}
+The candidate software can be found through search engine queries and domain related publications. The candidate software should have the following properties:
 \begin{enumerate}
+	\item Major function(s) must fall within the identified domain.
 	\item Must have viewable source code.
-\item Ideally have a git repository. \pmi{Not all of mine had this..}
+	\item Ideally have a git repository or ability to gather emprical measures found in Section \ref{SecEmpiricalMeasures}.
 \end{enumerate}
@@
 \section{How to Initially Filter the Software List} \label{SecInitialFilter}
-The person doing this must keep track of both the initial and filtered lists. The reasons for this is traceability bc..
+The initial list of candidate software should be filtered using the following properties:
 \begin{enumerate}
-	\item Professionalism..
-	\item Available documentation..
-	\item Status..
-	\item Availability of source code..
+	\item Organization - The software and any related documentation should appear to be easy to gather and understand.
+	\item Available documentation - The purpose of the software and the installation and usage procedures should appear to be moderately clear or easy to find.
+	\item Status - The software cannot be marked as incomplete or in an initial development phase.
 \end{enumerate}
+
+Copies of both the initial and filtered lists should be kept for traceability purposes.
 

### 2020-08-31 StateOfPractice/Methodology/Methodology.tex
moved 'User Experiments' from Methodology doc into its own file

@@
 \item Gather source code and documentation for each prospective software package.
-\item Collect empirical measures. (Section~\ref{SecEmpiricalMeasures}) \pmi{To be completed}
+\item Collect empirical measures. (Section~\ref{SecEmpiricalMeasures}) \pmi{To be completed - go over how to clean this up in meeting, or through an issue}
 \item Measure using ``shallow'' measurement template. (Section~\ref{SecShallowMeasure}) \pmi{To be completed}
@@
 \begin{enumerate}
-\item Survey developers \pmi{To do: Link to survey}
-\item Usability experiments \pmi{Move the below usability section to a new doc, and link to it.}
+\item Survey developers (\href{run:Questions to Developers.pdf}{Questions to Developers})
+\item Usability experiments (\href{run:User Experiments.pdf}{User Experiments})
 \item Performance benchmarks\pmi{note: this is still in consideration}
@@
 
-\section{User Experiments}
-\subsection{Usability Experimental Procedure}
-
-\subsection {Procedure}
-\begin {enumerate}
-%\item Select a list of projects
-%\item Write research questions
-%\item Prepare logistic details(location, time table, recording setup, moderators etc)
-%\item Prepare survey questions
-%\item Design the tasks for the study subjects to perform (tasks can be defined based on user category, at least one task to modify the software)
-%\item Select participants
-
-\item Survey participants to collect pre-experiment data
-\item Participants perform tasks
-\item Observe the study subjects (take notes, record sessions(OBS screen recorder), watch out for body languages and verbal cues)
-\item Survey the study subjects to collect feedback (post experiment interview)
-\item Prepare experiment report
-\item Perform pairwise comparison analysis
-\item Prepare analysis report
-\end {enumerate}
-
-%\subsection{Nielsen’s Heuristics}
-%\begin{enumerate}
-%	\item Consistency: Check if standards and conventions in product design are followed
-% \begin{itemize}
-% 	\item Sequences of actions (skill acquisition).
-% 	\item Color (categorization).
-% 	\item Layout and position (spatial consistency).
-% 	\item Font, capitalization (levels of organization).
-% 	\item Terminology (delete, del, remove, rm) and language (words, phrases).
-% 	\item Standards (e.g.,	blue underlined text for unvisited hyper links).
-% \end{itemize}
-%\item Visibility: Users should be informed about what is going on with the system through appropriate feedback and display of information e.g. 
-% \begin{itemize}
-% \item What is the current state of the system?
-% \item What can be done at current state?
-% \item Where can users go?
-% \item What change is made after an action?
-% \end{itemize}
-%\item Match between system and world. The image of the system perceived by users should match the model the users have about the system. 
-%\begin {itemize}
-%\item User model matches system image.
-%\item Actions provided by the system should match actions performed by users.
-%\item Objects on the system should match objects of the task.
-%\end {itemize}
-%\item Minimalist: Any extraneous information is a distraction and a slow-down. 
-%\begin{itemize}
-%\item Less is more.
-%\item Simple is not equivalent to abstract and general.
-%\item Simple is efficient.
-%\item  Progressive levels of detail.
-%\end{itemize}
-%\item Minimize memory load. Users should not be required to memorize a lot of information to carry out tasks. Memory load reduces users’ capacity to carry out the main tasks. 
-%\begin{itemize}
-%\item Recognition vs. recall (e.g., menu vs. commands).
-%\item Externalize information through visualization.
-%\item Perceptual procedures.
-%\item Hierarchical structure.
-%\item Default values.
-%\item Concrete examples (DD/MM/YY, e.g., 10/20/99).
-%\item Generic rules and actions (e.g., drag objects).
-%\item Intuitive procedure
-%\end{itemize}
-%\item Informative feedback. Users should be given prompt and informative feedback about their actions. 
-%\begin{itemize}
-%\item Information that can be directly perceived, interpreted, and evaluated.
-%\item Levels of feedback (novice and expert).
-%\item Concrete and specific, not abstract and general.
-%\item Response time:\\
-%-0.1 s for instantaneously reacting;\\
-%-1.0 s for uninterrupted flow of thought;\\
-%-10 s for the limit of attention.
-%\end{itemize}
-%%\item Flexibility and efficiency. Users always learn and users are always different. Give users the flexibility of creating customization and shortcuts to accelerate their performance. 
-%%\begin{itemize}
-%%	\item Shortcuts for experienced users.
-%%	\item Shortcuts or macros for frequently used operations.
-%%	\item Skill acquisition through chunking.
-%%	\item Examples:
-%%	- Abbreviations, function keys, hot keys, command keys, macros, aliases, templates, type-ahead, bookmarks, hot links, history, default values, etc.
-%%\end{itemize}
-%\item Good error messages. 
-%\begin {itemize}
-%\item The messages should be informative enough such that users can understand the nature of errors, learn from errors, and recover from errors. 
-%\item Phrased in clear language, avoid obscure codes. Example of obscure code: “system crashed, error code 147.”
-%\item Precise, not vague or general. Example of general comment: “Cannot open document.”
-%\item Constructive.
-%\item Polite. Examples of impolite message: “illegal user action,” “job aborted,” “system was crashed,” “fatal error,” etc.
-%\end {itemize}
-%\item Prevent errors. It is always better to design interfaces that prevent errors from happening in the first place.
-%\begin {itemize} 
-%\item Interfaces that make errors impossible.
-%\item Avoid modes (e.g., vi, text wrap). Or use informative feedback, e.g., different sounds.
-%\item Execution error vs. evaluation error.
-%\item Various types of slips and mistakes.
-%\end {itemize}
-%\item Clear closure. Every task has a beginning and an end. Users should be clearly notified about the completion of a task. 
-%\begin {itemize}
-%\item Clear beginning, middle, and end.
-%\item Complete 7-stages of actions.
-%\item Clear feedback to indicate goals are achieved and current stacks of goals can be released. Examples of good closures include many dialogues.
-%\end {itemize}
-%\item Reversible actions. 
-%\begin {itemize}
-%\item Users should be allowed to recover from errors. Reversible actions also encourage exploratory learning. 
-%\item At different levels: a single action, a subtask, or a complete task.
-%\item Multiple steps.
-%\item Encourage exploratory learning.
-%\item Prevent serious errors.
-%\end {itemize}
-%\item Use users’ language. The language should be always presented in a form understandable by the intended users.
-%\begin{itemize}
-%\item Use standard meanings of words.
-%\item Specialized language for specialized group.
-%\item User defined aliases.
-%\item Users’ perspective. Example: “we have bought four tickets for you” (bad) vs. “you bought four tickets” (good).
-%\end {itemize}
-%\item Help and documentation. Always provide help when needed. 
-%\begin {itemize}
-%\item Context-sensitive help.
-%\item Four types of help.\\
-%-task-oriented;\\
-%-alphabetically ordered;\\
-%-semantically organized;\\
-%-search.
-%\item Help embedded in contents.
-%\end {itemize}
-%\item Users in control. Do not give users that impression that they are controlled by the systems. 
-%\begin {itemize}
-%\item Users are initiators of actors, not responders to actions.
-%\item Avoid surprising actions, unexpected outcomes, tedious sequences of actions, etc.
-%\end {itemize}
-%\end{enumerate}
-
-\subsection{Task selection criteria}
-**The task selection will be determined with the aid of the domain experts attached to any of the selected projects.\\ 
-**The domain experts will be asked to consider the below criteria when defining a task. \\
-**Domain experts will also be asked to identify what background knowledge is necessary for the suggested tasks - Novice, Intermediate, Advanced
-\begin {enumerate}
-\item Collectively all tasks should not take no more than 2 hours.
-\item Selected tasks should reflect common use cases of  the software.
-\item Include tasks that require a set of sequential or hierarchical steps to be completed
-
-\end {enumerate}
-
-\subsection{Usability Questionnaire}
-Two sources of standardized usability questionnaire we could use.
-
-- \url{https://www.usabilitest.com/sus-pdf-generator}- 20-29 - SUS.\\
-- \url{https://uiuxtrend.com/pssuq-post-study-system-usability-questionnaire/} - PSSUQ
-
 \section{Analytic Hierarchy Process} \label{SecAHP}

### 2020-08-31 StateOfPractice/Methodology/Methodology.tex
updated 'Measure Using Shallow Measurement Template' in Methodology document

@@
 \item Collect empirical measures. (Section~\ref{SecEmpiricalMeasures}) \pmi{To be completed - go over how to clean this up in meeting, or through an issue}
-\item Measure using ``shallow'' measurement template. (Section~\ref{SecShallowMeasure}) \pmi{To be completed}
+\item Measure using ``shallow'' measurement template. (Section~\ref{SecShallowMeasure}) \pmi{To be reviewed}
 \item Use AHP process to rank the software packages. (Section~\ref{SecAHP}) \pmi{To be completed}
@@
 
-\subsubsection{git-stats}
+\subsubsection{git-stats}\label{gitstats}
 \href{https://github.com/tomgi/git_stats}{GitHub repo}
@@
 downloaded as spreadsheets.
+
+\subsubsection{git-stats}\label{scc}
+\href{https://github.com/boyter/scc}{GitHub repo}
 
@@
 \section{Measure Using Shallow Measurement Template} \label{SecShallowMeasure}
-This step...
-The measurement template can be found \href{run:Combined_MeasurementTemplate_EmpiricalMeasures.xlsx}{here}.
+For each software package fill out one column in the \href{run:Combined_MeasurementTemplate_EmpiricalMeasures.xlsx}{Measurement Template} following these steps:
 \begin{enumerate}
-	\item Step 1..
+	\item Gather the summary information for rows 3 to 20
+	\item Using the GitStats tool found in Section \ref{gitstats} gather the measurements for rows 105 to 112
+	\item Using the SCC tool found in Section \ref{scc} gather the measurements for rows 114 to 118
+	\item If the software package is found on git, gather the measurements for rows 120 to 124
+	\item Review installation documentation and attempt to install the software package
+	\item Gather the measurements for rows 22 to 37
+	\item Gather the measurements for correctness and verifiability in rows 39 to 48
+	\item Gather the measurements for surface reliability in rows 50 to 56
+	\item Gather the measurements for surface robustness in rows 58 to 61
+	\item Gather the measurements for surface usability in rows 63 to 68
+	\item Gather the measurements for maintainability in rows 70 to 78
+	\item Gather the measurements for reusability in rows 80 to 83
+	\item Gather the measurements for surface understandability in rows 85 to 94
+	\item Gather the measurements for visibility and transparency in rows 96 to 101
 \end{enumerate}

### 2020-09-01 StateOfPractice/Methodology/Methodology.tex
updated 'AHP' in Methodology document

@@
 \item Measure using ``shallow'' measurement template. (Section~\ref{SecShallowMeasure}) \pmi{To be reviewed}
-\item Use AHP process to rank the software packages. (Section~\ref{SecAHP}) \pmi{To be completed}
+\item Use AHP process to rank the software packages. (Section~\ref{SecAHP}) \pmi{To be reviewed}
 \item Identify a short list of top software packages, typically four to six, for deeper exploration according to the AHP rankings of the shallow measurements.
@@
 \section{Analytic Hierarchy Process} \label{SecAHP}
-
-Describe process. Domain experts review. Outline the tool we used. The README file of the tool can be found \href{run:../AHP2020/LBM/README.txt}{here}.
+The analytical hierarchy process (AHP) ranks software packages using pairwise comparisons of quality scores.
+We have developed a tool for conducting this process.
+The tool includes an AHP JAR script and a sensitivity analysis JAR script that is used to ensure that the software package rankings are appropriate with respect to the uncertainty of the quality scores.
+The README file of the tool is found \href{run:../AHP2020/LBM/README.txt}{here}.
+This file outlines the requirements for, and configuration and usage of the JAR scripts.
+The JAR scripts, source code, and required libraries are located in the same folder as the README file.
 

### 2020-09-02 StateOfPractice/Methodology/Methodology.tex
updated 'AHP' in Methodology document

--- 2020-08-31 StateOfPractice/Methodology/Methodology.tex
@@
 \section{Analytic Hierarchy Process} \label{SecAHP}
-
-Describe process. Domain experts review. Outline the tool we used. The README file of the tool can be found \href{run:../AHP2020/LBM/README.txt}{here}.
+The Analytical Hierarchy Process (AHP) is a decision making technique that can be used when comparing multiple options by multiple criteria.
+In our work AHP is used for comparing and ranking the software packages of a domain using the quality scores that are gathered in the \href{run:Combined_MeasurementTemplate_EmpiricalMeasures.xlsx}{Measurement Template}.
+AHP performs a pairwise analysis between each of the quality options using a matrix which is then used to generate an overall score for each software package for the given criteria.
+More information on AHP is available in \cite{SmithEtAl2016} We have developed a tool for conducting this process.
+The tool includes an AHP JAR script and a sensitivity analysis JAR script that is used to ensure that the software package rankings are appropriate with respect to the uncertainty of the quality scores.The README file of the tool is found \href{run:../AHP2020/LBM/README.txt}{here}.
+This file outlines the requirements for, and configuration and usage of the JAR scripts.
+The JAR scripts, source code, and required libraries are located in the same folder as the README file.
 

### 2020-09-07 StateOfPractice/Methodology/Methodology.tex
copied Empirical Measures into Appendix in Methodology document for the purpose of traceability
updated AHP section of Methodology

--- 2020-09-01 StateOfPractice/Methodology/Methodology.tex
@@
 \section{Analytic Hierarchy Process} \label{SecAHP}
-The analytical hierarchy process (AHP) ranks software packages using pairwise comparisons of quality scores.
+The Analytical Hierarchy Process (AHP) is a decision-making technique that can be used when comparing multiple options by multiple criteria.
+In our work AHP is used for comparing and ranking the software packages of a domain using the quality scores that are gathered in the \href{run:Combined_MeasurementTemplate_EmpiricalMeasures.xlsx}{Measurement Template}.
+AHP performs a pairwise analysis between each of the quality options using a matrix which is then used to generate an overall score for each software package for the given criteria.
+\cite{SmithEtAl2016} shows how AHP is applied to ranking software based on quality measures.
 We have developed a tool for conducting this process.
@@
 The README file of the tool is found \href{run:../AHP2020/LBM/README.txt}{here}.
-This file outlines the requirements for, and configuration and usage of the JAR scripts.
+This file outlines the requirements for, and configuration and usage of, the JAR scripts.
 The JAR scripts, source code, and required libraries are located in the same folder as the README file.
@@
 
+\subsection{Empirical Measures Considerations - Raw Data}
+
+Measures that can be extracted from on-line repos.
+
+\ad{Still at brainstorm stage.}
+\begin{itemize}
+	\item number of contributors
+	\item number of watches
+	\item number of stars
+	\item number of forks
+	\item number of clones
+	\item number of commits
+	\item number of total/code/document files
+	\item lines of total/logical/comment code
+	\item lines/pages of documents (can pdf be extracted?)
+	\item number of total/open/closed/merged pull requests
+	\item number of total/open/closed issues
+	\item number of total/open/closed issues with assignees
+\end{itemize}
+
+Instead of only focus on the current status of the above numbers, we may find
+the time history of them to be more valuable. For example, the number of
+contributors over time, the number of lines of code over time, the number of
+open issues over time, etc.
+
+\subsection{Empirical Measures Considerations - Processed Data}
+Metrics that can be calculated from the raw data.
+
+\ad{Still at brainstorm stage.}
+\begin{itemize}
+	\item percentage of total/open/closed issues with assignees -
+	Visibility/Transparency
+	\item lines of new code produced per person-month - Productivity
+	\item lines/pages of new documents produced per person-month - Productivity
+	\item number of issues closed per person-month - Productivity
+	\item percentage of comment lines in the code - maintainability \ad{Not Ao's
+		qualities}
+\end{itemize}
+
+In the above calculations, a month can be determined to be 30 days.
+
+\subsection{Empirical Measures Considerations - Tool Tests}
+\ad{This section is currently a note of unorganized contents. Most parts will beremoved or relocated.}
+
+\ad{This citation needs to be deleted later. It's here because my compiler
+	doesn't work with 0 citations}
+\cite{Emms2019}
+
+Most tests were done targeting to the repo of 3D Slicer
+\href{https://github.com/tomgi/git_stats}{GitHub repo}
+
+\subsubsection{git-stats}\label{gitstats}
+\href{https://github.com/tomgi/git_stats}{GitHub repo}
+
+Test results:
+\href{http://git-stats-slicer.ao9.io/}{http://git-stats-slicer.ao9.io/} the
+results are output as webpages, so I hosted for you to check. Data can be
+downloaded as spreadsheets.
+
+\subsubsection{git-stats}\label{scc}
+\href{https://github.com/boyter/scc}{GitHub repo}
+
+\subsubsection{git-of-theseus}
+\href{https://github.com/erikbern/git-of-theseus}{GitHub repo}
+
+Test results: It took about 100 minutes for one repo on a 8 core 16G ram Linux
+machine. It only outputs graphs.
+
+\subsubsection{hercules}
+\href{https://github.com/src-d/hercules}{GitHub repo}
+
+Test results: this one seems to be promising, but the installation is
+complicated with various errors.
+
+\subsubsection{git-repo-analysis}
+\href{https://github.com/larsxschneider/git-repo-analysis}{GitHub repo}
+
+\subsubsection{HubListener}
+\href{https://github.com/pjmc-oliveira/HubListener}{GitHub repo}
+
+The data that HubListener can extract.
+
+Raw:
+\begin{itemize}
+	\item Number of Files
+	\item Number of Lines
+	\item Number of Logical Lines
+	\item Number of Comments
+\end{itemize}
+
+Cyclomatic:
+\href{https://www.geeksforgeeks.org/cyclomatic-complexity/}{Intro}
+\begin{itemize}
+	\item Cyclomatic Complexity
+\end{itemize}
+
+Halstead:
+\href{https://www.geeksforgeeks.org/software-engineering-halsteads-software-metrics/}{Intro}
+\begin{itemize}
+	\item Halstead Effort
+	\item Halstead Bugs
+	\item Halstead Length
+	\item Halstead Difficulty
+	\item Halstead Time
+	\item Halstead Vocabulary
+	\item Halstead Volume
+\end{itemize}
+
+Test results: HubListener works well on the repo of itself, but it did not work
+well on some other repos.
+
+\subsubsection{gitinspector}
+\href{https://github.com/ejwa/gitinspector}{GitHub repo}
+
+Test results: it doesn't work well. Instead of creating output results, it
+prints the results directly in the console.
+
 \newpage

### 2020-09-08 StateOfPractice/Methodology/Methodology.tex
updated Domain Analysis section of Methodology

@@
 \begin{enumerate}
-\item Start with state of practice research questions. (Section~\ref{ResearchQuestions}) \pmi{To be completed}
+\item Start with state of practice research questions.
+(Section~\ref{ResearchQuestions}) \pmi{To be completed; should I remove section 2 from above and just link to ../OverallResearchProposal/ObjectivesAndResearchQuestions.pdf?, move the doc to this folder?}
 \item Identify the domain. (Section~\ref{SecIdentifyDomain}) \pmi{To be reviewed}
@@
 \item \emph{Domain Experts}: Review domain software list. (\href{run:Meeting Agenda with Domain Experts.pdf}{Meeting Agenda with Domain Experts})
-\item Domain Analysis. (Section~\ref{SecDomainAnalysis})\pmi{To be completed}
+\item Domain Analysis. (Section~\ref{SecDomainAnalysis})\pmi{To be reviewed}
 \item \emph{Domain Experts}: Vet domain analysis.
@@
 \section{Domain Analysis} \label{SecDomainAnalysis}
+The domain analysis consists of a commonality analysis of the software packages.
+Its purpose is to show the relationships between these packages, and to facilitate an understanding of the informal specification and development of them.
+The final result of the analysis will be tables of commonalities, variabilities and parameters of variation.
 
-Commonality analysis.  Follow as for mesh generator (likely with less detail).
-Final result will be tables of commonalities, variabilities and parameters of
-variation.
-
-Commonality analysis document Steps:
+\noindent Steps to produce a commonality analysis:
 \begin{enumerate}
-\item Introduction
-\item Overview of Domain
-\item Add Commonalities - Split into simulation, input, output, and
-  nonfunctional requirements
-\item Add Variabilities - Split into simulation, input, output, system
-  constraints, and nonfunctional requirements
-\item Add Parameters of Variation - Split into simulation, input, output, system
-  constraints, and nonfunctional requirements
+\item Write an Introduction
+\item Write the Overview of Domain
+\item List Commonalities - Split into sections for the simulation step, input information, output information, and nonfunctional requirements
+\item List Variabilities - Split into sections for the simulation step, input information, output information, and nonfunctional requirements
+\item List Parameters of Variation - Split into sections for the simulation step, input information, output information, system constraints, and nonfunctional requirements
 \item Add Terminology, Definitions, Acronyms
@@
 
-Commonality analysis for Lattice Boltzmann Solvers can be found
+A sample commonality analysis for Lattice Boltzmann Solvers can be found
 \href{run:../Peter-Notes/Commonality-Analysis-LB-Systems.pdf}{here}.

### 2020-09-09 StateOfPractice/Methodology/Methodology.tex
updated Emprirical Measures section of Methodology

@@
 \subsection{Raw Data}
-Measures that can be extracted from on-line repos.
+The following raw data measures are extracted from repositories:
 
-\ad{Still at brainstorm stage.}
 \begin{itemize}
-\item number of contributors
-\item number of watches
-\item number of stars
-\item number of forks
-\item number of clones
-\item number of commits
-\item number of total/code/document files
-\item lines of total/logical/comment code
-\item lines/pages of documents (can pdf be extracted?)
-\item number of total/open/closed/merged pull requests
-\item number of total/open/closed issues
-\item number of total/open/closed issues with assignees
+\item Number of stars.
+\item Number of forks.
+\item Number of people watching the repository.
+\item Number of open pull requests.
+\item Number of closed pull requests.	
+\item Number of developers.	
+\item Number of open issues.
+\item Number of closed issues.
+\item Initial release date.
+\item Last commit date.
+\item Programming languages used.
+\item Number of text-based files.
+\item Number of total lines in text-based files.
+\item Number of code lines in text-based files.
+\item Number of comment lines in text-based files.
+\item Number of blank lines in text-based files.
+\item Number of binary files.  
+\item Number of total lines added to text-based files.
+\item Number of total lines deleted from text-based files.
+\item Number of total commits.
+\item Numbers of commits by year in the last 5 years. (Count from as early as possible if the project is younger than 5 years.) 
+\item Numbers of commits by month in the last 12 months.
 \end{itemize}
 
-Instead of only focus on the current status of the above numbers, we may find
-the time history of them to be more valuable. For example, the number of
-contributors over time, the number of lines of code over time, the number of
-open issues over time, etc.
+\subsection{Processed Data}
+The following measures are calculated from the raw data:
 
-\subsection{Processed Data}
-Metrics that can be calculated from the raw data.
-
-\ad{Still at brainstorm stage.}
 \begin{itemize}
-\item percentage of total/open/closed issues with assignees -
-Visibility/Transparency
-\item lines of new code produced per person-month - Productivity
-\item lines/pages of new documents produced per person-month - Productivity
-\item number of issues closed per person-month - Productivity
-\item percentage of comment lines in the code - maintainability \ad{Not Ao's
-qualities}
+\item Status of software package (dead or alive).
+\item Percentage of identified issues that are closed.
+\item Percentage of code that is comments.
 \end{itemize}
 
-In the above calculations, a month can be determined to be 30 days.
-
 \subsection{Tool Tests}
-\ad{This section is currently a note of unorganized contents. Most parts will beremoved or relocated.}
-
-\ad{This citation needs to be deleted later. It's here because my compiler
-doesn't work with 0 citations}
-\cite{Emms2019}
-
-Most tests were done targeting to the repo of 3D Slicer
-\href{https://github.com/tomgi/git_stats}{GitHub repo}
+The following tools are used to gather some of the raw data from the repositories:
 
@@
 
-Test results:
-\href{http://git-stats-slicer.ao9.io/}{http://git-stats-slicer.ao9.io/} the
-results are output as webpages, so I hosted for you to check. Data can be
-downloaded as spreadsheets.
-
-\subsubsection{git-stats}\label{scc}
+\subsubsection{scc}\label{scc}
 \href{https://github.com/boyter/scc}{GitHub repo}
-
-\subsubsection{git-of-theseus}
-\href{https://github.com/erikbern/git-of-theseus}{GitHub repo}
-
-Test results: It took about 100 minutes for one repo on a 8 core 16G ram Linux
-machine. It only outputs graphs.
-
-\subsubsection{hercules}
-\href{https://github.com/src-d/hercules}{GitHub repo}
-
-Test results: this one seems to be promising, but the installation is
-complicated with various errors.
-
-\subsubsection{git-repo-analysis}
-\href{https://github.com/larsxschneider/git-repo-analysis}{GitHub repo}
-
-\subsubsection{HubListener}
-\href{https://github.com/pjmc-oliveira/HubListener}{GitHub repo}
-
-The data that HubListener can extract.
-
-Raw:
-\begin{itemize}
-\item Number of Files
-\item Number of Lines
-\item Number of Logical Lines
-\item Number of Comments
-\end{itemize}
-
-Cyclomatic:
-\href{https://www.geeksforgeeks.org/cyclomatic-complexity/}{Intro}
-\begin{itemize}
-\item Cyclomatic Complexity
-\end{itemize}
-
-Halstead:
-\href{https://www.geeksforgeeks.org/software-engineering-halsteads-software-metrics/}{Intro}
-\begin{itemize}
-\item Halstead Effort
-\item Halstead Bugs
-\item Halstead Length
-\item Halstead Difficulty
-\item Halstead Time
-\item Halstead Vocabulary
-\item Halstead Volume
-\end{itemize}
-
-Test results: HubListener works well on the repo of itself, but it did not work
-well on some other repos.
-
-\subsubsection{gitinspector}
-\href{https://github.com/ejwa/gitinspector}{GitHub repo}
-
-Test results: it doesn't work well. Instead of creating output results, it
-prints the results directly in the console.
 
@@
 
-\subsubsection{git-stats}\label{gitstats}
+\subsubsection{git-stats}
 \href{https://github.com/tomgi/git_stats}{GitHub repo}
@@
 
-\subsubsection{git-stats}\label{scc}
+\subsubsection{scc}
 \href{https://github.com/boyter/scc}{GitHub repo}

### 2020-09-09 StateOfPractice/Methodology/Methodology.tex
commented on Rank Short List section of Methodology

@@
 \end{enumerate}
-\item Rank short list. (Section~\ref{SecRankShortList}) \pmi{To be completed}
+\item Rank short list. (Section~\ref{SecRankShortList}) \pmi{To be completed; how will we do this? status of usability experiments?}
 \item Document answers for research questions.
@@
 \section{Rank Short List} \label{SecRankShortList}
-Rank using pairwise comparison of short list results.
-\begin{enumerate}
-	\item Compare with respect to usability
-	\item ...
-\end{enumerate}
+Rank using pairwise comparison of short list software packages with respect to usability survey results.
 

### 2020-09-17 StateOfPractice/Methodology/Methodology.tex
update Methodology

@@
 \usepackage{hyperref}
+\usepackage{xcolor}
 
-\usepackage{hyperref}
 \hypersetup{
@@
 	\item Ideally have a git repository or ability to gather emprical measures found in Section \ref{SecEmpiricalMeasures}.
+	\item \ad{``Ideally have the latest release or source code commit within the last 5 years.'' How about this one? Do we care about the ``vintage'' software?}
 \end{enumerate}
@@
 	\item Available documentation - The purpose of the software and the installation and usage procedures should appear to be moderately clear or easy to find.
+	\ad{What if installation or usage procedures are unavailable? I think we're measuring that if the software have installation instructions and user manuals,
+	so perhaps they should be kept even if these docs are missing.}
 	\item Status - The software cannot be marked as incomplete or in an initial development phase.
+	\item \ad{What if there are still more candidates than we need? Should we also filter out the ``older'' or ``less popular'' ones if needed? }
 \end{enumerate}
@@
 For each software package fill out one column in the \href{run:Combined_MeasurementTemplate_EmpiricalMeasures.xlsx}{Measurement Template} following these steps:
+\ad{Maybe a one or two sentences introduction to the template?}
 \begin{enumerate}
@@
 	\item Using the GitStats tool found in Section \ref{gitstats} gather the measurements for rows 105 to 112
+	\ad{Will it be easier to use the section names instead of row numbers? Otherwise it can be hard to maintain.}
 	\item Using the SCC tool found in Section \ref{scc} gather the measurements for rows 114 to 118
@@
 	\item Gather the measurements for correctness and verifiability in rows 39 to 48
+	\ad{for 7 to 14, maybe we can put them together, to something like "measurements for the followings sections..."}
 	\item Gather the measurements for surface reliability in rows 50 to 56

### 2020-09-19 StateOfPractice/Methodology/Methodology.tex
copied over Olu's Research Questions into Methodology doc Section 2, made some adjustment, added comments - Section is still WOP

@@
 \section{Research Questions}\label{ResearchQuestions}
-
-In general questions:
-
 	\begin{enumerate} 
-\item Comparison between domains
-\item How to measure qualities
-\item How does the quality compare for projects with the most resources to those
-  with the fewest?
-\item What skills/knowledge are needed by future developers?
-\item How can the development process be improved?
-\item What are the common pain points?
+		\item What artifacts are recommended for research software development?
+\pmi{remove}
+		\item What artifacts do current software projects produce?
+\pmi{keep}
+		\item What artifacts are produced in existing software development processes. 
+\pmi{remove}
+		\item What information is captured by current projects artifacts?
+\pmi{keep}
+		\item What information is necessary to capture for sustainability and productivity?
+\pmi{need to review - subjective, can we capture this from gathering state of practice data?}
+		\item What commonalities exist among software artifacts?
+\pmi{keep}
+		\item What are sustainable and productive methods to produce softifacts?
+\pmi{remove}
+		\item What are the consequences of technology tools towards sustainability and productivity  \pmi{keep, reword: what technology tools are used? what do they capture? benefits?}
+		\item What are the common factors that lead to sustainability and productivity?
+\pmi{need to review, reword. How do we gather sustainability from SOP data?}
+		\item What relationships exist, if any, between 'high ranking' software projects and sustainability/productivity?
+ \pmi{need to review}	
+		\item What skills/knowledge are needed by future developers?	\pmi{copied from original - review}
+		\item What are the common pain points?	\pmi{copied from original - review}
+		\item How can the development process be improved?	\pmi{copied from original - review}
 	\end{enumerate}
 
-For each domain questions''
-
-\begin{enumerate}
-\item Best examples within the domain
-\item What software artifacts?
-\item What are the pain points?
-\item Any advice on what can be done about the pain points?
-\end{enumerate}
-
-Measure the effort invested and the reward.  Related to sustainability.
-
-Collect the data and see what conclusions follow.  For an individual domain,
-between domains.  The process isn't so much about ranking the software as it is
-about looking at the software closely and see what conclusions arise.  The
-measurements are intended to force scrutiny, from different perspectives.
+%In general questions:
+%
+%\begin{enumerate}
+%\item Comparison between domains
+%\item How to measure qualities
+%\item How does the quality compare for projects with the most resources to %those
+%  with the fewest?
+%\item What skills/knowledge are needed by future developers?
+%\item How can the development process be improved?
+%\item What are the common pain points?
+%\end{enumerate}
+%
+%For each domain questions''
+%
+%\begin{enumerate}
+%\item Best examples within the domain
+%\item What software artifacts?
+%\item What are the pain points?
+%\item Any advice on what can be done about the pain points?
+%\end{enumerate}
+%
+%Measure the effort invested and the reward.  Related to sustainability.
+%
+%Collect the data and see what conclusions follow.  For an individual domain,
+%between domains.  The process isn't so much about ranking the software as it is
+%about looking at the software closely and see what conclusions arise.  The
+%measurements are intended to force scrutiny, from different perspectives.
 

### 2020-09-20 StateOfPractice/Methodology/Methodology.tex
updated section 3 of Methodology doc

@@
 
+Note the following formatting conventions of this document.
+Red text denotes a link to information within the document.
+Purple text denotes a link to supplementary information that is external to this document.\pmi{needs to be added to intro to explain convention}
+
 \section{Research Questions}\label{ResearchQuestions}
@@
 
-\section{Overview of Steps in Assessing Quality of the Domain Software}
+\section{Overview of Steps in Assessing Quality of the Domain Software}\label{StepsAQDS}
 
 \begin{enumerate}
-\item Start with state of practice research questions.
-(Section~\ref{ResearchQuestions}) \pmi{To be completed; should I remove section 2 from above and just link to ../OverallResearchProposal/ObjectivesAndResearchQuestions.pdf?, move the doc to this folder?}
+\item Start with state of practice research questions. (Section~\ref{ResearchQuestions}) \pmi{To be completed}
 \item Identify the domain. (Section~\ref{SecIdentifyDomain}) \pmi{To be reviewed}
@@
 \item Collect empirical measures. (Section~\ref{SecEmpiricalMeasures}) \pmi{To be completed - go over how to clean this up in meeting, or through an issue}
-\item Measure using ``shallow'' measurement template. (Section~\ref{SecShallowMeasure}) \pmi{To be reviewed}
+\item Measure using measurement template. (Section~\ref{SecShallowMeasure}) \pmi{To be reviewed}
 \item Use AHP process to rank the software packages. (Section~\ref{SecAHP}) \pmi{To be reviewed}
-\item Identify a short list of top software packages, typically four to six, for deeper exploration according to the AHP rankings of the shallow measurements.
-\item \emph{Domain Experts}: Vet AHP ranking and short list.
-(\href{run:Meeting Agenda with Domain Experts.pdf}{Meeting Agenda with Domain Experts}) \pmi{This was part of the original Steps in Assessing Quality list.
-We need to add it to the Meeting Agenda}
+\item Identify a short list of top software packages, typically four to six, for deeper exploration according to the AHP rankings of the measurements.
+\item \emph{Domain Experts}: Vet AHP ranking and short list. (\href{run:Meeting Agenda with Domain Experts.pdf}{Meeting Agenda with Domain Experts})
 \item With short list:
@@
 \item Performance benchmarks\pmi{note: this is still in consideration}
-\item Maintainability experiments\pmi{note: this is still in consideration}
+\item Maintainability experiments\pmi{note: these experiments need to be completed - to be discussed at next meeting}
 \end{enumerate}
-\item Rank short list. (Section~\ref{SecRankShortList}) \pmi{To be completed; how will we do this? status of usability experiments?}
+\item Rank short list. (Section~\ref{SecRankShortList}) \pmi{To be completed}
 \item Document answers for research questions.
@@
 
-\section{Measure Using Shallow Measurement Template} \label{SecShallowMeasure}
+\section{Measure Using Measurement Template} \label{SecShallowMeasure}
 For each software package fill out one column in the \href{run:Combined_MeasurementTemplate_EmpiricalMeasures.xlsx}{Measurement Template} following these steps:
@@
 Rank using pairwise comparison of short list software packages with respect to usability survey results.
+
+\pmi{SS: let's ask the people that do the usability/performance/maintainability experiments to do an AHP pair-wise comparison between the short-list software packages}
 

### 2020-09-20 StateOfPractice/Methodology/Methodology.tex
updated section 5 of Methodology doc

@@
 \section{How to Identify Candidate Software} \label{SecIdentifyCandSoft}
-The candidate software can be found through search engine queries and domain related publications. The candidate software should have the following properties:
+The candidate software can be found through search engine queries targeting authoritative lists of software from the domain, such as lists on GitHub and swMATH, and through searching domain related publications.
+Domain experts are also asked for their suggestions and are asked to review the list.
+The candidate software should have the following properties:
+
 \begin{enumerate}
@@
 	\item Must have viewable source code.
-	\item Ideally have a git repository or ability to gather emprical measures found in Section \ref{SecEmpiricalMeasures}.
-	\item \ad{``Ideally have the latest release or source code commit within the last 5 years.'' How about this one? Do we care about the ``vintage'' software?}
+	\item Ideally have a git repository or ability to gather empirical measures found in Section \ref{SecEmpiricalMeasures}.
+	\item \ad{``Ideally have the latest release or source code commit within the last 5 years.'' How about this one?
+Do we care about the ``vintage'' software?}\pmi{I tested some software that has not been updated in about 8 years, as this software was mentioned in multiple places and looked professional.
+That being said, age should be noted as something that is considered along with other qualities.
+"Ideally the latest release or source code commit should be relatively recent, such as within the last 5 years, unless the candidate software appears to be well recommended and currently in use."}
 \end{enumerate}

### 2020-09-20 StateOfPractice/Methodology/Methodology.tex
updated section 6 of Methodology doc

@@
 \section{How to Initially Filter the Software List} \label{SecInitialFilter}
+The initial list of candidate software packages should first be filtered to a manageable size.
+Our target is about 30 packages.
+We want the list to include packages that are not marked as incomplete and are at least somewhat organized and understandable.
+We also want the list to be a good representation of the software domain, with packages having enough in common with respect to functionality that it makes sense to compare the software.
 The initial list of candidate software should be filtered using the following properties:
+
 \begin{enumerate}
@@
 	so perhaps they should be kept even if these docs are missing.}
+    \pmi{How are we installing the software if such documents and procedures are missing?
+Should we expect users to be able to install?
+I did not encounter a situation where even basic installation steps were missing.}
 	\item Status - The software cannot be marked as incomplete or in an initial development phase.
-	\item \ad{What if there are still more candidates than we need? Should we also filter out the ``older'' or ``less popular'' ones if needed? }
+	\item \ad{What if there are still more candidates than we need? Should we also filter out the ``older'' or ``less popular'' ones if needed? }\pmi{Good question. Let's discuss during meeting.}
 \end{enumerate}

### 2020-09-20 StateOfPractice/Methodology/Methodology.tex
updated section 4 of Methodology doc

@@
 \item Start with state of practice research questions. (Section~\ref{ResearchQuestions}) \pmi{To be completed}
-\item Identify the domain. (Section~\ref{SecIdentifyDomain}) \pmi{To be reviewed}
+\item Identify the domain. (Section~\ref{SecIdentifyDomain}) \pmi{To be completed based on suggestions}
 \item \emph{Domain Experts}: Create a top ten list of software packages in the domain. (\href{run:Meeting Agenda with Domain Experts.pdf}{Meeting Agenda with Domain Experts})
@@
 \section{How to Identify the Domain} \label{SecIdentifyDomain}
+\pmi{add introduction}
+\pmi{add a definition in the introduction for what we mean by research software}
 \begin{enumerate}	
 	\item The domain must fall within the research software scope.
-	\item The domain must be specific.
 	\item The domain must be well understood.
 	\item There must be a community of people studying the domain.
-	\item There must to be a variety of software solutions for the domain.
-	\item These software solutions must have a user community.
+	\item A preliminary search, or discussion with experts, suggests that there will be numerous, at least about 15, candidate software packages to study.
+	\item The software packages must have open source options. 
 \end{enumerate}
+\pmi{SS:"Add examples of well understood domains. We can include pointers to previous "state of the practice" projects as examples."}
 

### 2020-09-20 StateOfPractice/Methodology/Methodology.tex
updated section 9 of Methodology doc

@@
 \section{Measure Using Measurement Template} \label{SecShallowMeasure}
-For each software package fill out one column in the \href{run:Combined_MeasurementTemplate_EmpiricalMeasures.xlsx}{Measurement Template} following these steps:
-\ad{Maybe a one or two sentences introduction to the template?}
+For each software package fill out one column of the \href{run:Combined_MeasurementTemplate_EmpiricalMeasures.xlsx}{Measurement Template} spreadsheet found in our repository. Follow these steps:
+\ad{Maybe a one or two sentences introduction to the template?} \pmi{SS:"A screenshot of a portion of the template would be very helpful here"}
+
+\pmi{SS:"please add an estimate of how long each measurement typically takes"}
+
+\pmi{SS:"explain any tricky parts with interpreting the parameters of variation for measuring a particular indicator. A few example would help."}
+
 \begin{enumerate} 
-	\item Gather the summary information for rows 3 to 20
-	\item Using the GitStats tool found in Section \ref{gitstats} gather the measurements for rows 105 to 112
-	\ad{Will it be easier to use the section names instead of row numbers? Otherwise it can be hard to maintain.}
-	\item Using the SCC tool found in Section \ref{scc} gather the measurements for rows 114 to 118
-	\item If the software package is found on git, gather the measurements for rows 120 to 124
-	\item Review installation documentation and attempt to install the software package
-	\item Gather the measurements for rows 22 to 37
-	\item Gather the measurements for correctness and verifiability in rows 39 to 48
-	\ad{for 7 to 14, maybe we can put them together, to something like "measurements for the followings sections..."}
-	\item Gather the measurements for surface reliability in rows 50 to 56
-	\item Gather the measurements for surface robustness in rows 58 to 61
-	\item Gather the measurements for surface usability in rows 63 to 68
-	\item Gather the measurements for maintainability in rows 70 to 78
-	\item Gather the measurements for reusability in rows 80 to 83
-	\item Gather the measurements for surface understandability in rows 85 to 94
-	\item Gather the measurements for visibility and transparency in rows 96 to 101
+	\item Gather the summary information into the top section of the document
+	\item Using the GitStats tool found in Section \ref{gitstats} gather the measurements for the Repo Metrics - GitStats section found near the bottom of the document
+	\item Using the SCC tool found in Section \ref{scc} gather the measurements for the Repo Metrics - SCC section found near the bottom of the document
+	\item If the software package is found on git, gather the measurements for the Repo Metrics - GitHub section found near the bottom of the document
+	\item Review installation documentation and attempt to install the software package on a virtual machine
+	\item Gather the measurements for installability
+	\pmi{SS:"we should mention that the project owners will be contacted for help with installation, if necessary, but that there is a cap on the amount of time spent on installation"}
+	\item Gather the measurements for correctness and verifiability
+	\item Gather the measurements for surface reliability
+	\item Gather the measurements for surface robustness
+	\item Gather the measurements for surface usability
+	\item Gather the measurements for maintainability
+	\item Gather the measurements for reusability
+	\item Gather the measurements for surface understandability
+	\item Gather the measurements for visibility and transparency
+	\item Assign a score out of ten for each quality \pmi{SS:"add some text about how the measure out of 10 is assigned for each quality", link to scoring document}
 \end{enumerate}
+\pmi{SS:"add something about the sensitivity analysis tool and its purpose", explain what the tool is and its purpose, indicate that it will be used in the next section to verify rankings, add it to the next section}
 

### 2020-09-21 StateOfPractice/Methodology/Methodology.tex
updated section 7 of Methodology doc

@@
 \section{Domain Analysis} \label{SecDomainAnalysis}
+\pmi{SS:"We should have more of an introduction on what is a commonality analysis, and more generally, what is a program family.
+We should also have more of a literature review, and a pointer to examples.
+Some publications to look at include ArdisAndWeiss1997, Chen2003, SmithAndChen2004, SmithEtAl2008, SmithMcCutchanAndCarette2017, Weiss1998"}
 The domain analysis consists of a commonality analysis of the software packages.
@@
 \item Write the Overview of Domain
-\item List Commonalities - Split into sections for the simulation step, input information, output information, and nonfunctional requirements
-\item List Variabilities - Split into sections for the simulation step, input information, output information, and nonfunctional requirements
-\item List Parameters of Variation - Split into sections for the simulation step, input information, output information, system constraints, and nonfunctional requirements
+\item List Commonalities
+\item List Variabilities
+\item List Parameters of Variation
 \item Add Terminology, Definitions, Acronyms

### 2020-09-21 StateOfPractice/Methodology/Methodology.tex
updated section 7 link of Methodology doc
updated section 8 of Methodology doc

@@
 Red text denotes a link to information within the document.
-Purple text denotes a link to supplementary information that is external to this document.\pmi{needs to be added to intro to explain convention}
+Purple text denotes a link to supplementary information that is external to this document.
+Blue text denotes a URL link.\pmi{needs to be added to intro to explain convention}
 
@@
 A sample commonality analysis for Lattice Boltzmann Solvers can be found
-\href{run:../Peter-Notes/Commonality-Analysis-LB-Systems.pdf}{here}.
+\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Peter-Notes/Commonality-Analysis-LB-Systems.pdf}{here}.
 
 \section{Empirical Measures} \label{SecEmpiricalMeasures}
+\pmi{Add introduction.  We should orient the reader on the purpose of this section, and provide a brief roadmap of the subsections.}
 
@@
 The following raw data measures are extracted from repositories:
-
+\pmi{SS:"we should say why we are collecting this data.
+It is okay to say that we are collecting data that is reasonably easy to collect.
+We will then put the data together and analyze it.
+However, we should also relate the measures we are collecting to questions we want to answer.
+Ideally, we will relate the data to our research questions.
+For instance, we are collecting some of the data to see how large a project is.
+More files, more lines of code, etc, means a larger project.
+The number of developers is another measure of size.
+Other measures are intended to ascertain a project's popularity, like the number of stars and forks.
+Other measures are intended to see how active the project is, like the measures related to the history of commits."}
 \begin{itemize}
@@
 \begin{itemize}
-\item Status of software package (dead or alive).
+\item Status of software package (dead or alive). Alive is defined as the presence of repository commits or software package version releases in the last 18 months.
 \item Percentage of identified issues that are closed.
@@
 
-\subsection{Tool Tests}
-The following tools are used to gather some of the raw data from the repositories:
-
-\subsubsection{git-stats}\label{gitstats}
-\href{https://github.com/tomgi/git_stats}{GitHub repo}
-
-\subsubsection{scc}\label{scc}
-\href{https://github.com/boyter/scc}{GitHub repo}
+\subsection{Tools}
+\href{https://github.com/tomgi/git_stats}{GitStats} is used to measure the number of binary files as well as the number of added and deleted lines in a repository.
+The tool is also used to measuring the number of commits over different intervals of time.
+\href{https://github.com/boyter/scc}{Sloc Cloc and Code (scc)} is used to measure the number of text based files as well as the number of total, code, comment, and blank lines in a repository.
+These tools were selected due to their installability, usability, and ability to gather the above empirical measures.
+A guide for installing and running them can be found \href{run:A Guide to Empirical Measures.pdf}{here}.
 

### 2020-09-27 StateOfPractice/Methodology/Methodology.tex
updated section 2 of Methodology

@@
 \section{Research Questions}\label{ResearchQuestions}
+
+Questions from Olu's Research Proposal:
 	\begin{enumerate} 
@@
 \pmi{keep}
-		\item What artifacts are produced in existing software development processes. 
-\pmi{remove}
-		\item What information is captured by current projects artifacts?
-\pmi{keep}
-		\item What information is necessary to capture for sustainability and productivity?
-\pmi{need to review - subjective, can we capture this from gathering state of practice data?}
-		\item What commonalities exist among software artifacts?
-\pmi{keep}
-		\item What are sustainable and productive methods to produce softifacts?
-\pmi{remove}
-		\item What are the consequences of technology tools towards sustainability and productivity  \pmi{keep, reword: what technology tools are used? what do they capture? benefits?}
-		\item What are the common factors that lead to sustainability and productivity?
-\pmi{need to review, reword. How do we gather sustainability from SOP data?}
-		\item What relationships exist, if any, between 'high ranking' software projects and sustainability/productivity?
- \pmi{need to review}	
-		\item What skills/knowledge are needed by future developers?	\pmi{copied from original - review}
-		\item What are the common pain points?	\pmi{copied from original - review}
-		\item How can the development process be improved?	\pmi{copied from original - review}
+\item What information is captured by current software project artifacts?\pmi{keep}
+\item What information is necessary to capture for sustainability?\pmi{remove}
+\item What information is necessary to capture for productivity?\pmi{remove}
+\item What commonalities exist among software artifacts?\pmi{need to review}
+\item How can we improve sustainability in research softifacts development practices?\pmi{remove}
+\item What are the consequences of technology tools towards sustainability and productivity?\pmi{remove}
+\item What are attitudes of research software developers and engineer towards sustainability and productivity with respect to producing softifacts? \pmi{keep}
+\item What are the factors that lead to sustainability and productivity?\pmi{remove}
+\item What relationships exist between `high rated' software projects and sustainability with respect to research softifacts production? \pmi{remove}
+\end{enumerate}
+
+Questions from Overall Research Project folder
+\begin{enumerate}
+	\item What knowledge is currently captured in the artifacts generated by
+	existing open source SCS projects?\pmi{keep}
+	\item What principles, processes, methodologies and tools are used by existing open source SCS projects?\pmi{keep}
+	\item What are the pain points" for developers working on SCS projects?
+	What aspects of the existing processes, methodologies and tools do
+	they consider could potentially be improved?\pmi{keep}
+	\item For a given software product (softifacts), how can we produce a reasonable measure of the software qualities (correctness, reliability, usability etc.) with a few hours of effort?\pmi{remove}
+	\item For a given software product (softifacts), how can we produce a meaningful measure of usability with a few days worth of effort?\pmi{remove}
+	\item For a given software product (softifacts), how can we produce a meaningful measure of maintainability (with respect to likely changes) with
+	a few days worth of effort?\pmi{remove}
+	\item For a given software product (softifacts), how can we produce a meaningful measure of reproducibility/replicability with a few days worth of
+	effort?\pmi{remove}
+	\item With a reasonable effort and time, how can we compare the productivity between different processes, methodologies and tools?\pmi{remove}
+	\item Can a correlation be identified between software projects that use
+	best practices for the development process and methodologies and improvement in the qualities of usability, performance and reproducibility/replicability?\pmi{remove}
+	\item Ignoring the time and effort required, what knowledge is necessary to
+	capture for sustainable SCS? How can this knowledge best be documented in software artifacts?\pmi{remove}
+	\item How can Model Driven Engineering and Code Generation be t in the
+	development process for SCS to remove as many repetitive, tedious and
+	error prone tasks as possible? We will call this the ideal process."\pmi{remove}
+	\item How well do existing modelling and code generation tools do at implementing the ideal process"? Candidates for comparison include Drasil
+	and Epsilon.\pmi{remove}
+	\item How well does a model driven approach to compared to a traditional
+	approach for usability, productivity and sustainability?\pmi{remove}
+	\item What are the attitudes and preferences of typical users towards an
+	implementation of the ideal process"?\pmi{remove}
 \end{enumerate}

### 2020-09-28 StateOfPractice/Methodology/Methodology.tex
updated section 4 of Methodology

@@
 \section{How to Identify the Domain} \label{SecIdentifyDomain}
-\pmi{add introduction}
-\pmi{add a definition in the introduction for what we mean by research software}
+A domain of research software must be identified.
+Research software is defined in this exercise as ``software that is used to generate, process or analyze results that [are intended] to appear in a publication'' \citep{hettrick2014uk}.
+The chosen domain must have the following properties:
+
 \begin{enumerate}	
@@
 \end{enumerate}
-\pmi{SS:"Add examples of well understood domains. We can include pointers to previous "state of the practice" projects as examples."}
+
+Some examples of domains that fit these criteria are seismology software, as well as mesh generators and processors.
+These are described in \citep{SmithEtAl2018} and \citep{smith2016state}, respectively.
 

### 2020-09-28 StateOfPractice/Methodology/Methodology.tex
updated section 9 of Methodology

@@
 \section{Measure Using Measurement Template} \label{SecShallowMeasure}
-For each software package fill out one column of the \href{run:Combined_MeasurementTemplate_EmpiricalMeasures.xlsx}{Measurement Template} spreadsheet found in our repository. Follow these steps:
-\ad{Maybe a one or two sentences introduction to the template?} \pmi{SS:"A screenshot of a portion of the template would be very helpful here"}
 
-\pmi{SS:"please add an estimate of how long each measurement typically takes"}
+\begin{figure}[h!]
+	\begin{center}
+		\includegraphics[width=1.0\textwidth]{measurement_template}
+		\caption{Top of Measurement Template}
+		\label{measurement_template_image}
 
-\pmi{SS:"explain any tricky parts with interpreting the parameters of variation for measuring a particular indicator. A few example would help."}
+	\end{center}
+\end{figure}
+
+The \href{run:Combined_MeasurementTemplate_EmpiricalMeasures.xlsx}{Measurement Template} is used to track measurements and quality scores for all of the software packages in the domain.
+For each software package, fill out one column of the template.
+This process can take between 2 to 5 hours for each package.
+Project developers can be contacted for help regarding installation, if necessary, but a cap of about 2 hours should be imposed on the entire installation process.
+For completing the template, please follow these steps:
+
+\pmi{SS:"explain any tricky parts with interpreting the parameters of variation for measuring a particular indicator. A few example would help."}\pmi{Ask Ao for input}
 
 \begin{enumerate} 
-	\item Gather the summary information into the top section of the document
-	\item Using the GitStats tool found in Section \ref{gitstats} gather the measurements for the Repo Metrics - GitStats section found near the bottom of the document
-	\item Using the SCC tool found in Section \ref{scc} gather the measurements for the Repo Metrics - SCC section found near the bottom of the document
+	\item Gather the summary information into the top section of the document (Figure \ref{measurement_template_image})
+	\item Using the GitStats tool that is described in Section \ref{SecEmpiricalMeasures} gather the measurements for the Repo Metrics - GitStats section found near the bottom of the document
+	\item Using the SCC tool that is also described in Section \ref{SecEmpiricalMeasures} gather the measurements for the Repo Metrics - SCC section found near the bottom of the document
 	\item If the software package is found on git, gather the measurements for the Repo Metrics - GitHub section found near the bottom of the document
@@
 	\item Gather the measurements for installability
-	\pmi{SS:"we should mention that the project owners will be contacted for help with installation, if necessary, but that there is a cap on the amount of time spent on installation"}
 	\item Gather the measurements for correctness and verifiability
@@
 	\item Gather the measurements for visibility and transparency
-	\item Assign a score out of ten for each quality \pmi{SS:"add some text about how the measure out of 10 is assigned for each quality", link to scoring document}
+	\item Assign a score out of ten for each quality.
+The score can be measured using the \href{run:MeasurementTemplate_ImpressionCalculator.xlsx}{Measurement Template Impression Calculator}.
+For each quality measurement, the file indicates the appropriate score to assign the measurement based on possible measurement values.
 \end{enumerate}
-\pmi{SS:"add something about the sensitivity analysis tool and its purpose", explain what the tool is and its purpose, indicate that it will be used in the next section to verify rankings, add it to the next section}
 

### 2020-09-28 StateOfPractice/Methodology/Methodology.tex
updated section 8 of Methodology

@@
 \section{Empirical Measures} \label{SecEmpiricalMeasures}
-\pmi{Add introduction.  We should orient the reader on the purpose of this section, and provide a brief roadmap of the subsections.}
+Some quality measurements rely on the gathering of raw and processed empirical data.
+We focus on data that is reasonably easy to collect, which we combine and analyze for this assessment.
+The measures that are collected relate to the research questions that we want to answer.
+For instance, we collect some of the data to see how large a project is.
+Other measures intend to ascertain a project’s popularity, or how active the project is.
+Section \ref{rawdata} will orient the reader as to what raw data is collected. 
+Some of this data can be observed from Git repository metrics.
+The rest can be collected using freeware tools.
+\href{https://github.com/tomgi/git_stats}{GitStats} is used to measure the number of binary files as well as the number of added and deleted lines in a repository.
+The tool is also used to measure the number of commits over different intervals of time.
+\href{https://github.com/boyter/scc}{Sloc Cloc and Code (scc)} is used to measure the number of text based files as well as the number of total, code, comment, and blank lines in a repository.
+These tools were selected due to their installability, usability, and ability to gather the empirical measures listed below.
+A guide for installing and running them can be found \href{run:A Guide to Empirical Measures.pdf}{here}.
+Section \ref{processeddata} introduces the required processed data, which is calculated using the raw data.
 
-\subsection{Raw Data}
+\subsection{Raw Data}\label{rawdata}
 The following raw data measures are extracted from repositories:
-\pmi{SS:"we should say why we are collecting this data.
-It is okay to say that we are collecting data that is reasonably easy to collect.
-We will then put the data together and analyze it.
-However, we should also relate the measures we are collecting to questions we want to answer.
-Ideally, we will relate the data to our research questions.
-For instance, we are collecting some of the data to see how large a project is.
-More files, more lines of code, etc, means a larger project.
-The number of developers is another measure of size.
-Other measures are intended to ascertain a project's popularity, like the number of stars and forks.
-Other measures are intended to see how active the project is, like the measures related to the history of commits."}
+
 \begin{itemize}
@@
 
-\subsection{Processed Data}
+\subsection{Processed Data}\label{processeddata}
 The following measures are calculated from the raw data:
@@
 \end{itemize}
-
-\subsection{Tools}
-\href{https://github.com/tomgi/git_stats}{GitStats} is used to measure the number of binary files as well as the number of added and deleted lines in a repository.
-The tool is also used to measuring the number of commits over different intervals of time.
-\href{https://github.com/boyter/scc}{Sloc Cloc and Code (scc)} is used to measure the number of text based files as well as the number of total, code, comment, and blank lines in a repository.
-These tools were selected due to their installability, usability, and ability to gather the above empirical measures.
-A guide for installing and running them can be found \href{run:A Guide to Empirical Measures.pdf}{here}.
 

### 2020-10-01 StateOfPractice/Methodology/Methodology.tex
Some revisions to the research questions section

@@
 
-Note the following formatting conventions of this document.
-Red text denotes a link to information within the document.
-Purple text denotes a link to supplementary information that is external to this document.
-Blue text denotes a URL link.\pmi{needs to be added to intro to explain convention}
+Note the following formatting conventions of this document. Red text denotes a
+link to information within the document. Purple text denotes a link to
+supplementary information that is external to this document. Blue text denotes a
+URL link.\pmi{needs to be added to intro to explain convention}
 
@@
 Questions from Olu's Research Proposal:
+
 \begin{enumerate}
-		\item What artifacts are recommended for research software development?
-\pmi{remove}
-		\item What artifacts do current software projects produce?
-\pmi{keep}
-\item What information is captured by current software project artifacts?\pmi{keep}
-\item What information is necessary to capture for sustainability?\pmi{remove}
-\item What information is necessary to capture for productivity?\pmi{remove}
-\item What commonalities exist among software artifacts?\pmi{need to review}
-\item How can we improve sustainability in research softifacts development practices?\pmi{remove}
-\item What are the consequences of technology tools towards sustainability and productivity?\pmi{remove}
-\item What are attitudes of research software developers and engineer towards sustainability and productivity with respect to producing softifacts? \pmi{keep}
-\item What are the factors that lead to sustainability and productivity?\pmi{remove}
-\item What relationships exist between `high rated' software projects and sustainability with respect to research softifacts production? \pmi{remove}
+\item What artifacts do current software projects produce?  \pmi{keep}
+\item What information is captured by current software project
+  artifacts? \pmi{keep} \wss{This question probably isn't part of the state of
+    the practice exercise, but part of Olu's PhD research.}
+\item What are attitudes of research software developers and engineer toward:
+\begin{enumerate}
+\item sustainability?
+\item productivity?
+\end{enumerate} \pmi{keep} \wss{What are we hoping to learn from this question?
+  This might be a question for Olu, since she currently has the same research question.}
+\item For a given domain of research software, what current projects provide the
+  best examples of high quality software and documentation? \wss{Reworded
+    version of one of Olu's research questions.}
 \end{enumerate}
 
-Questions from Overall Research Project folder
+\wss{Rather than focus on sustainability and productivity, should we focus on
+  usability and modifiability for the SOP exercise?  Our current questions and
+  experiments focus more on these topics.}
+
+Questions from Overall Research Project folder:
+
 \begin{enumerate}
-	\item What knowledge is currently captured in the artifacts generated by
-	existing open source SCS projects?\pmi{keep}
-	\item What principles, processes, methodologies and tools are used by existing open source SCS projects?\pmi{keep}
-	\item What are the pain points" for developers working on SCS projects?
-	What aspects of the existing processes, methodologies and tools do
-	they consider could potentially be improved?\pmi{keep}
-	\item For a given software product (softifacts), how can we produce a reasonable measure of the software qualities (correctness, reliability, usability etc.) with a few hours of effort?\pmi{remove}
-	\item For a given software product (softifacts), how can we produce a meaningful measure of usability with a few days worth of effort?\pmi{remove}
-	\item For a given software product (softifacts), how can we produce a meaningful measure of maintainability (with respect to likely changes) with
-	a few days worth of effort?\pmi{remove}
-	\item For a given software product (softifacts), how can we produce a meaningful measure of reproducibility/replicability with a few days worth of
-	effort?\pmi{remove}
-	\item With a reasonable effort and time, how can we compare the productivity between different processes, methodologies and tools?\pmi{remove}
-	\item Can a correlation be identified between software projects that use
-	best practices for the development process and methodologies and improvement in the qualities of usability, performance and reproducibility/replicability?\pmi{remove}
-	\item Ignoring the time and effort required, what knowledge is necessary to
-	capture for sustainable SCS? How can this knowledge best be documented in software artifacts?\pmi{remove}
-	\item How can Model Driven Engineering and Code Generation be t in the
-	development process for SCS to remove as many repetitive, tedious and
-	error prone tasks as possible? We will call this the ideal process."\pmi{remove}
-	\item How well do existing modelling and code generation tools do at implementing the ideal process"? Candidates for comparison include Drasil
-	and Epsilon.\pmi{remove}
-	\item How well does a model driven approach to compared to a traditional
-	approach for usability, productivity and sustainability?\pmi{remove}
-	\item What are the attitudes and preferences of typical users towards an
-	implementation of the ideal process"?\pmi{remove}
+\item What principles, processes, methodologies and tools are used by existing
+  open source SCS projects? \pmi{keep}
+\item What are the pain points for developers working on SCS projects?  What
+  aspects of the existing processes, methodologies and tools do they consider
+  could potentially be improved? \pmi{keep}
 \end{enumerate}
@@
 %
-%For each domain questions''
+%For each domain questions:
 %

### 2020-10-01 StateOfPractice/Methodology/Methodology.tex
updated and commented on sections 2, 5, 6 of Methodology based on meeting discussion

@@
 \item What artifacts do current software projects produce?  \pmi{keep}
-\item What information is captured by current software project
-  artifacts? \pmi{keep} \wss{This question probably isn't part of the state of
-    the practice exercise, but part of Olu's PhD research.}
-\item What are attitudes of research software developers and engineer toward:
+\item What are the attitudes and actions of research software developers and toward:
 \begin{enumerate}
 \item sustainability?
-\item productivity?
+\item usability?
+\item modifiability? \pmi{think about and possibly remove this one}
 \end{enumerate} \pmi{keep} \wss{What are we hoping to learn from this question?
@@
 	\item Ideally have a git repository or ability to gather empirical measures found in Section \ref{SecEmpiricalMeasures}.
-	\item \ad{``Ideally have the latest release or source code commit within the last 5 years.'' How about this one?
-Do we care about the ``vintage'' software?}\pmi{I tested some software that has not been updated in about 8 years, as this software was mentioned in multiple places and looked professional.
-That being said, age should be noted as something that is considered along with other qualities.
-"Ideally the latest release or source code commit should be relatively recent, such as within the last 5 years, unless the candidate software appears to be well recommended and currently in use."}
+	\item Status - The software cannot be marked as incomplete or in an initial development phase.
 \end{enumerate}
@@
 \begin{enumerate}
-	\item Organization - The software and any related documentation should appear to be easy to gather and understand.
+	\item Restrict scope (functionality?)\pmi{elaborate on this}
+	\item Can we install? \pmi{elaborate on this, copy from point below?}
 	\item Available documentation - The purpose of the software and the installation and usage procedures should appear to be moderately clear or easy to find.
@@
 I did not encounter a situation where even basic installation steps were missing.}
-	\item Status - The software cannot be marked as incomplete or in an initial development phase.
-	\item \ad{What if there are still more candidates than we need? Should we also filter out the ``older'' or ``less popular'' ones if needed? }\pmi{Good question. Let's discuss during meeting.}
+	\item \ad{What if there are still more candidates than we need?
+Should we also filter out the ``older'' or ``less popular'' ones if needed?
+}\pmi{Good question.
+Let's discuss during meeting.}\pmi{meeting: add point about dropping off oldest ones, look at the point below copied from previous section}
+	\item \ad{``Ideally have the latest release or source code commit within the last 5 years.'' How about this one?
+Do we care about the ``vintage'' software?}\pmi{I tested some software that has not been updated in about 8 years, as this software was mentioned in multiple places and looked professional.
+That being said, age should be noted as something that is considered along with other qualities.
+"Ideally the latest release or source code commit should be relatively recent, such as within the last 5 years, unless the candidate software appears to be well recommended and currently in use."}
+
 \end{enumerate}

### 2020-10-11 StateOfPractice/Methodology/Methodology.tex
fixed typos in Methodology document

@@
 \item What artifacts do current software projects produce?  \pmi{keep}
-\item What are the attitudes and actions of research software developers and toward:
+\item What are the attitudes and actions of research software developers toward:
 \begin{enumerate}
@@
 	\item Ideally have a git repository or ability to gather empirical measures found in Section \ref{SecEmpiricalMeasures}.
-	\item Status - The software cannot be marked as incomplete or in an initial development phase.
+	\item The software cannot be marked as incomplete or in an initial development phase.
 \end{enumerate}

### 2020-10-11 StateOfPractice/Methodology/Methodology.tex
fixed typos in Methodology document

@@
 \section{How to Identify Candidate Software} \label{SecIdentifyCandSoft}
-The candidate software can be found through search engine queries targeting authoritative lists of software from the domain, such as lists on GitHub and swMATH, and through searching domain related publications.
+The candidate software can be found through search engine queries targeting authoritative lists of software from the domain, such as lists on GitHub and swMATH, and by searching domain related publications.
 Domain experts are also asked for their suggestions and are asked to review the list.

### 2020-10-11 StateOfPractice/Methodology/Methodology.tex
fixed link in Section 8 of Methodology document
updated html links in Methodology document

@@
 \item Identify the domain. (Section~\ref{SecIdentifyDomain}) \pmi{To be completed based on suggestions}
-\item \emph{Domain Experts}: Create a top ten list of software packages in the domain. (\href{run:Meeting Agenda with Domain Experts.pdf}{Meeting Agenda with Domain Experts})
+\item \emph{Domain Experts}: Create a top ten list of software packages in the domain.
+(\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Meeting Agenda withmDomainmExperts.pdf}{Meeting Agenda with Domain Experts})
 \item Brief the Domain Experts on the overall objective, research proposal, research questions, measurement template, survey for short list projects, usability tests, performance benchmarks, maintainability experiments.
-(\href{run:Meeting Agenda with Domain Experts.pdf}{Meeting Agenda with Domain Experts})
+(\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Meeting Agenda withmDomainmExperts.pdf}{Meeting Agenda with Domain Experts})
 \item Identify broad list of candidate software packages in the domain. (Section~\ref{SecIdentifyCandSoft}) \pmi{To be reviewed}
 \item Preliminary filter of software packages list. (Section~\ref{SecInitialFilter}) \pmi{To be reviewed}
-\item \emph{Domain Experts}: Review domain software list. (\href{run:Meeting Agenda with Domain Experts.pdf}{Meeting Agenda with Domain Experts})
+\item \emph{Domain Experts}: Review domain software list.
+(\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Meeting Agenda withmDomainmExperts.pdf}{Meeting Agenda with Domain Experts})
 \item Domain Analysis. (Section~\ref{SecDomainAnalysis})\pmi{To be reviewed}
 \item \emph{Domain Experts}: Vet domain analysis.
-(\href{run:Meeting Agenda with Domain Experts.pdf}{Meeting Agenda with Domain Experts}) \pmi{This was part of the original Steps in Assessing Quality list.
+(\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Meeting Agenda withmDomainmExperts.pdf}{Meeting Agenda with Domain Experts}) \pmi{This was part of the original Steps in Assessing Quality list.
 We need to add it to the Meeting Agenda}
@@
 \item Identify a short list of top software packages, typically four to six, for deeper exploration according to the AHP rankings of the measurements.
-\item \emph{Domain Experts}: Vet AHP ranking and short list. (\href{run:Meeting Agenda with Domain Experts.pdf}{Meeting Agenda with Domain Experts})
+\item \emph{Domain Experts}: Vet AHP ranking and short list.
+(\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Meeting Agenda withmDomainmExperts.pdf}{Meeting Agenda with Domain Experts})
 \item With short list:
 \begin{enumerate}
-\item Survey developers (\href{run:Questions to Developers.pdf}{Questions to Developers})
-\item Usability experiments (\href{run:User Experiments.pdf}{User Experiments})
+\item Survey developers (\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Questions to Developers.pdf}{Questions to Developers})
+\item Usability experiments (\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/User Experiments.pdf}{User Experiments})
 \item Performance benchmarks\pmi{note: this is still in consideration}
@@
 These tools were selected due to their installability, usability, and ability to gather the empirical measures listed below.
-A guide for installing and running them can be found \href{run:A Guide to Empirical Measures.pdf}{here}.
+A guide for installing and running them can be found \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/A Guide to Empirical Measures.pdf}{here}.
 Section \ref{processeddata} introduces the required processed data, which is calculated using the raw data.
@@
 
-The \href{run:Combined_MeasurementTemplate_EmpiricalMeasures.xlsx}{Measurement Template} is used to track measurements and quality scores for all of the software packages in the domain.
+The \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Combined_MeasurementTemplate_EmpiricalMeasures.xlsx}{Measurement Template} is used to track measurements and quality scores for all of the software packages in the domain.
 For each software package, fill out one column of the template.
@@
 	\item Assign a score out of ten for each quality.
-The score can be measured using the \href{run:MeasurementTemplate_ImpressionCalculator.xlsx}{Measurement Template Impression Calculator}.
+The score can be measured using the \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/MeasurementTemplate_ImpressionCalculator.xlsx}{Measurement Template Impression Calculator}.
 For each quality measurement, the file indicates the appropriate score to assign the measurement based on possible measurement values.
@@
 The Analytical Hierarchy Process (AHP) is a decision-making technique that can be used when comparing multiple options by multiple criteria.
-In our work AHP is used for comparing and ranking the software packages of a domain using the quality scores that are gathered in the \href{run:Combined_MeasurementTemplate_EmpiricalMeasures.xlsx}{Measurement Template}.
+In our work AHP is used for comparing and ranking the software packages of a domain using the quality scores that are gathered in the \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Combined_MeasurementTemplate_EmpiricalMeasures.xlsx}{Measurement Template}.
 AHP performs a pairwise analysis between each of the quality options using a matrix which is then used to generate an overall score for each software package for the given criteria.
@@
 The tool includes an AHP JAR script and a sensitivity analysis JAR script that is used to ensure that the software package rankings are appropriate with respect to the uncertainty of the quality scores.
-The README file of the tool is found \href{run:../AHP2020/LBM/README.txt}{here}.
+The README file of the tool is found \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/AHP2020/LBM/README.txt}{here}.
 This file outlines the requirements for, and configuration and usage of, the JAR scripts.

### 2020-10-11 StateOfPractice/Methodology/Methodology.tex
updated Section 3 of Methodology

@@
 \begin{enumerate}
-\item Start with state of practice research questions. (Section~\ref{ResearchQuestions}) \pmi{To be completed}
-\item Identify the domain. (Section~\ref{SecIdentifyDomain}) \pmi{To be completed based on suggestions}
+\item Start with state of practice research questions. (Section~\ref{ResearchQuestions}) \pmi{To be updated based on feedback}
+\item Identify the domain. (Section~\ref{SecIdentifyDomain}) \pmi{In second review}
 \item \emph{Domain Experts}: Create a top ten list of software packages in the domain.
@@
 (\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Meeting Agenda withmDomainmExperts.pdf}{Meeting Agenda with Domain Experts})
-\item Identify broad list of candidate software packages in the domain. (Section~\ref{SecIdentifyCandSoft}) \pmi{To be reviewed}
-\item Preliminary filter of software packages list. (Section~\ref{SecInitialFilter}) \pmi{To be reviewed}
+\item Identify broad list of candidate software packages in the domain. (Section~\ref{SecIdentifyCandSoft})
+\item Preliminary filter of software packages list. (Section~\ref{SecInitialFilter}) \pmi{To be edited based on feedback}
 \item \emph{Domain Experts}: Review domain software list.
 (\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Meeting Agenda withmDomainmExperts.pdf}{Meeting Agenda with Domain Experts})
-\item Domain Analysis. (Section~\ref{SecDomainAnalysis})\pmi{To be reviewed}
+\item Domain Analysis. (Section~\ref{SecDomainAnalysis})\pmi{To be edited - add intro}
 \item \emph{Domain Experts}: Vet domain analysis.
-(\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Meeting Agenda withmDomainmExperts.pdf}{Meeting Agenda with Domain Experts}) \pmi{This was part of the original Steps in Assessing Quality list.
-We need to add it to the Meeting Agenda}
+(\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Meeting Agenda withmDomainmExperts.pdf}{Meeting Agenda with Domain Experts})
 \item Gather source code and documentation for each prospective software package.
-\item Collect empirical measures. (Section~\ref{SecEmpiricalMeasures}) \pmi{To be completed - go over how to clean this up in meeting, or through an issue}
-\item Measure using measurement template. (Section~\ref{SecShallowMeasure}) \pmi{To be reviewed}
-\item Use AHP process to rank the software packages. (Section~\ref{SecAHP}) \pmi{To be reviewed}
+\item Collect empirical measures. (Section~\ref{SecEmpiricalMeasures})
+\item Measure using measurement template. (Section~\ref{SecShallowMeasure}) \pmi{To be edited based on feedback}
+\item Use AHP process to rank the software packages. (Section~\ref{SecAHP})
 \item Identify a short list of top software packages, typically four to six, for deeper exploration according to the AHP rankings of the measurements.

### 2020-10-12 StateOfPractice/Methodology/Methodology.tex
updated Section 2 of Methodology

@@
 
-Questions from Olu's Research Proposal:
+The following are research questions that we want to answer in order to assess the state of practice of research software in our chosen domain:
 
 \begin{enumerate}
-\item What artifacts do current software projects produce?  \pmi{keep}
+\item What artifacts do current software packages produce? 
+\item What tools are used by current software packages?
+\item What principles, processes, and methodologies are used in the development of current software packages?
+\item What are the pain points for developers working on SCS projects?
+What aspects of the existing processes, methodologies and tools do they consider could potentially be improved?
+What processes, methodologies and tools have they changed in order to reduce obstacles?
 \item What are the attitudes and actions of research software developers toward:
@@
 \item usability?
-\item modifiability? \pmi{think about and possibly remove this one}
-\end{enumerate} \pmi{keep} \wss{What are we hoping to learn from this question?
-  This might be a question for Olu, since she currently has the same research question.}
-\item For a given domain of research software, what current projects provide the
-  best examples of high quality software and documentation? \wss{Reworded
-    version of one of Olu's research questions.}
+\item modifiability?
 \end{enumerate} 
-
-\wss{Rather than focus on sustainability and productivity, should we focus on
-  usability and modifiability for the SOP exercise?  Our current questions and
-  experiments focus more on these topics.}
-
-Questions from Overall Research Project folder:
-
-\begin{enumerate}
-\item What principles, processes, methodologies and tools are used by existing
-  open source SCS projects? \pmi{keep}
-\item What are the pain points for developers working on SCS projects?  What
-  aspects of the existing processes, methodologies and tools do they consider
-  could potentially be improved? \pmi{keep}
+\item For a given domain of research software, what current projects provide the best examples of high quality software and documentation?
 \end{enumerate}
@@
 \begin{enumerate}
-\item Start with state of practice research questions. (Section~\ref{ResearchQuestions}) \pmi{To be updated based on feedback}
+\item Start with the state of practice research questions. (Section~\ref{ResearchQuestions}) \pmi{To be updated based on feedback}
 \item Identify the domain. (Section~\ref{SecIdentifyDomain}) \pmi{In second review}

### 2020-10-12 StateOfPractice/Methodology/Methodology.tex
updated Section 6 of Methodology

@@
 We also want the list to be a good representation of the software domain, with packages having enough in common with respect to functionality that it makes sense to compare the software.
-The initial list of candidate software should be filtered using the following properties:
+The initial list of candidate software should be filtered using the following properties sequentially:
 
 \begin{enumerate}
-	\item Restrict scope (functionality?)\pmi{elaborate on this}
-	\item Can we install? \pmi{elaborate on this, copy from point below?}
-	\item Available documentation - The purpose of the software and the installation and usage procedures should appear to be moderately clear or easy to find.
-	\ad{What if installation or usage procedures are unavailable? I think we're measuring that if the software have installation instructions and user manuals,
-	so perhaps they should be kept even if these docs are missing.}
-    \pmi{How are we installing the software if such documents and procedures are missing?
-Should we expect users to be able to install?
-I did not encounter a situation where even basic installation steps were missing.}
-	\item \ad{What if there are still more candidates than we need?
-Should we also filter out the ``older'' or ``less popular'' ones if needed?
-}\pmi{Good question.
-Let's discuss during meeting.}\pmi{meeting: add point about dropping off oldest ones, look at the point below copied from previous section}
-	\item \ad{``Ideally have the latest release or source code commit within the last 5 years.'' How about this one?
-Do we care about the ``vintage'' software?}\pmi{I tested some software that has not been updated in about 8 years, as this software was mentioned in multiple places and looked professional.
-That being said, age should be noted as something that is considered along with other qualities.
-"Ideally the latest release or source code commit should be relatively recent, such as within the last 5 years, unless the candidate software appears to be well recommended and currently in use."}
-
+	\item Scope: The functionality of the domain software should be increasingly specific.
+	\item Usage: The installation procedure should appear to be clear or easy to figure out.
+    \item Age: Ideally the latest release or source code commit should be relatively recent, such as within the last 5 years, unless the candidate software appears to be well recommended and currently in use.
 \end{enumerate}

### 2020-10-12 StateOfPractice/Methodology/Methodology.tex

@@
 \section{Domain Analysis} \label{SecDomainAnalysis}
-\pmi{SS:"We should have more of an introduction on what is a commonality analysis, and more generally, what is a program family.
-We should also have more of a literature review, and a pointer to examples.
-Some publications to look at include ArdisAndWeiss1997, Chen2003, SmithAndChen2004, SmithEtAl2008, SmithMcCutchanAndCarette2017, Weiss1998"}
-The domain analysis consists of a commonality analysis of the software packages.
+
+The domain analysis consists of a commonality analysis of the family of software packages.
 Its purpose is to show the relationships between these packages, and to facilitate an understanding of the informal specification and development of them.
-The final result of the analysis will be tables of commonalities, variabilities and parameters of variation.
+\cite{weiss1997defining} defines commonality analysis as an ``approach to defining a family by identifying commonalities, i.e., assumptions that are true for all family members, variabilities, i.e., assumptions about what can vary among family members, and common terminology for the family''.
+The final result of the domain analysis will be tables of commonalities, variabilities, and parameters of variation of a program family, which \cite{parnas1976design} defines as ``a set of programs whose common properties are so extensive that it is advantageous to study the common properties of the programs before analyzing individual members''.
+\cite{smith2008commonality} present a template for conducting a commonality analysis, which was referred to when conducting this work.
+\cite{weiss1998commonality} describes another commonality analysis technique for deciding the members of a program family which readers are encouraged to be familiar with.
+\cite{SmithAndChen2004} provide an example of a commonality analysis for mesh generating software.
+The steps to produce a commonality analysis are:
 
-\noindent Steps to produce a commonality analysis:
 \begin{enumerate}

### 2020-10-14 StateOfPractice/Methodology/Methodology.tex
updated commonality and variability in Section 7 of Methodology

@@
 Its purpose is to show the relationships between these packages, and to facilitate an understanding of the informal specification and development of them.
-\cite{weiss1997defining} defines commonality analysis as an ``approach to defining a family by identifying commonalities, i.e., assumptions that are true for all family members, variabilities, i.e., assumptions about what can vary among family members, and common terminology for the family''.
+\cite{weiss1997defining} defines commonality analysis as an approach to defining a family by identifying commonalities, variabilities, and common terminology for the family.
+Commonalities are goals, theories, models, definitions and assumptions that are common between family members.
+Likewise, variabilities are goals, theories, models, definitions and assumptions that are different between family members.
 The final result of the domain analysis will be tables of commonalities, variabilities, and parameters of variation of a program family, which \cite{parnas1976design} defines as ``a set of programs whose common properties are so extensive that it is advantageous to study the common properties of the programs before analyzing individual members''.

### 2020-10-15 StateOfPractice/Methodology/Methodology.tex
added another commonality analysis reference to Section 7 of Methodology
fixed type Section 7 of Methodology
updated methodology with meeting comments

@@
 \begin{enumerate}
-\item What artifacts do current software packages produce? 
+\item What are the artifacts that current software packages consist of? 
 \item What tools are used by current software packages?
@@
 What processes, methodologies and tools have they changed in order to reduce obstacles?
-\item What are the attitudes and actions of research software developers toward:
+\item What are the actions of research software developers toward:\pmi{to do: reverse engineer from questions to developers; might remove or completely alter points below}
 \begin{enumerate}
-\item sustainability?
+\item sustainability?\pmi{meeting discussion did consider removing this}
 \item usability?
@@
 \end{enumerate} 
-\item For a given domain of research software, what current projects provide the best examples of high quality software and documentation?
+\item How does software designated as high quality by this methodology compare with top rated software by the community? \pmi{review}
 \end{enumerate}
@@
 \item Performance benchmarks\pmi{note: this is still in consideration}
-\item Maintainability experiments\pmi{note: these experiments need to be completed - to be discussed at next meeting}
+\item Maintainability experiments\pmi{note: these experiments need to be completed - to be discussed at next meeting}\pmi{meeting: start thinking about maintainability experiments; maybe change to modifiability; at least describe what we would do for such experiments - measure sensible changes, what does a small change mean exactly, document 'likely changes' - see Parnas; modifiability - we want to be explicit about what we want to modify; how to we interpret the results?; Add these experiments to user experiments file, in such case change file name, keep format similar to user experiments (procedure, no questionnaire, add preamble describing comments above), such experiments would get quantitative measures}\pmi{SS:For the maintainability experiment, I think we should write it up in the methodology, even if we don't do it.
+Discussing maintainability is needed to fit with the research questions (since maintainability is part of sustainability).
+We want sustainability to fit into the bigger context of what interests research software developers.
+As far as the maintainability experiment goes, we can just ask the domain experts to make a likely change and ask them to record their experience with making the change.
+It isn't quantifiable, but it would give us qualitative data.}
 \end{enumerate}
@@
 
-\wss{The domain expert is involved in multiple steps in the process.  How best
-  to get their feedback?  The domain experts are busy and are unlikely to devote
-  significant time to the project.  We need to quickly get to the point.  Maybe
-  something around task based inspection?  Directed interview?}
+%\wss{The domain expert is involved in multiple steps in the process.  How best
+%  to get their feedback?  The domain experts are busy and are unlikely to devote
+%  significant time to the project.  We need to quickly get to the point.  Maybe
+%  something around task based inspection?  Directed interview?}
 
@@
 \cite{weiss1998commonality} describes another commonality analysis technique for deciding the members of a program family which readers are encouraged to be familiar with.
-\cite{SmithAndChen2004} provide an example of a commonality analysis for mesh generating software.
+\cite{SmithAndChen2004} and \cite{SmithMcCutchanAndCarette2017} are examples of a commonality analysis for a family of mesh generating software and a family of material models, respectively.
 The steps to produce a commonality analysis are:

### 2020-10-15 StateOfPractice/Methodology/Methodology.tex
updated section 6 of Methodology

@@
 \section{How to Initially Filter the Software List} \label{SecInitialFilter}
-The initial list of candidate software packages should first be filtered to a manageable size.
-Our target is about 30 packages.
-We want the list to include packages that are not marked as incomplete and are at least somewhat organized and understandable.
-We also want the list to be a good representation of the software domain, with packages having enough in common with respect to functionality that it makes sense to compare the software.
-The initial list of candidate software should be filtered using the following properties sequentially:
+If the list of software is too long (over around 30 packages), then steps need to be taken to create a more manageable list.
+To reduce the length of the list, the following filters are applied in the priority order listed, stopping once the desired number of projects is reached:
 
@@
 	\item Usage: The installation procedure should appear to be clear or easy to figure out.
-    \item Age: Ideally the latest release or source code commit should be relatively recent, such as within the last 5 years, unless the candidate software appears to be well recommended and currently in use.
+    \item Age: Ideally the latest release or source code commit should be relatively recent for software in the domain unless the candidate software appears to be well recommended and currently in use.
 \end{enumerate}

### 2020-10-15 StateOfPractice/Methodology/Methodology.tex
updated section 2 of Methodology

@@
 
-The following are research questions that we want to answer in order to assess the state of practice of research software in our chosen domain:
+The following are research questions that we want to answer to assess the state of practice of research software in our chosen domain:
 
@@
 What aspects of the existing processes, methodologies and tools do they consider could potentially be improved?
-What processes, methodologies and tools have they changed in order to reduce obstacles?
-\item What are the actions of research software developers toward:\pmi{to do: reverse engineer from questions to developers; might remove or completely alter points below}
+What processes, methodologies and tools have they changed to reduce obstacles?
+\item For research software developers, what specific actions are taken to address the following:\pmi{to do: reverse engineer from questions to developers; might remove or completely alter points below}\pmi{removed attitudes as per meeting discussion}
 \begin{enumerate}
@@
 \item usability?
-\item modifiability?
+\item modifiability?\pmi{SS comment in issue 57: I think we want to keep modifiability in the list.
+The reasons you have listed are good reasons.
+Also, modifiability is such an important software quality, especially when it comes to having a good software development process.}
 \end{enumerate} 

### 2020-10-23 StateOfPractice/Methodology/Methodology.tex
updated Section 2 Research Questions of Methodology, actions by devs

@@
 What processes, methodologies and tools have they changed to reduce obstacles?
-\item For research software developers, what specific actions are taken to address the following:\pmi{to do: reverse engineer from questions to developers; might remove or completely alter points below}\pmi{removed attitudes as per meeting discussion}
+\item For research software developers, what specific actions are taken to address the following:
 \begin{enumerate}
-\item sustainability?\pmi{meeting discussion did consider removing this}
-\item usability?
-\item modifiability?\pmi{SS comment in issue 57: I think we want to keep modifiability in the list.
-The reasons you have listed are good reasons.
-Also, modifiability is such an important software quality, especially when it comes to having a good software development process.}
+\item sustainability?\pmi{previous meeting discussion did consider removing this; I do not see any current questions to developers that best fit this quality; confirm remove or add new question to developers?}
+\item usability \pmi{question 10 touches on this but will need to be reworded or broken up - ``What if any actions were taken to address usability issues?''}
+\item traceability \pmi{question to developers 3}
+\item modifiability \pmi{question to developers 9}
+\item maintainability \pmi{question to developers 9}
+\item correctness \pmi{question to developers 8}
+\item understandability \pmi{question to developers 10 touches on this; need to reword question in terms of an action}
+\item unambiguity \pmi{question to developers 11}
+\item reproducibility \pmi{question to developers 12}
+\item visibility/transparency \pmi{questions 1,3,4,5,6 build to this but a question would need to be modified to include something along the line of ``has this information been added to the documentation?''}
 \end{enumerate} 

### 2020-10-23 StateOfPractice/Methodology/Experiments.tex
updated Section 2 Modifiability experiment note, renamed experiments file, moved comments into file regarding modifiability experiments

\documentclass[letterpaper,cleveref]{lipics-v2019}

\usepackage{natbib}
\usepackage{booktabs}
\usepackage{amsmath,amsthm}
\usepackage{hyperref}

\usepackage{hyperref}
\hypersetup{
colorlinks=true,       % false: boxed links; true: colored links
linkcolor=red,          % color of internal links (change box color with
%linkbordercolor)
citecolor=blue,       % color of links to bibliography
filecolor=magenta,   % color of file links
urlcolor=cyan           % color of external links
}

%% Comments
\newif\ifcomments\commentstrue

\ifcomments
\newcommand{\authornote}[3]{\textcolor{#1}{[#3 ---#2]}}
\newcommand{\todo}[1]{\textcolor{red}{[TODO: #1]}}
\else
\newcommand{\authornote}[3]{}
\newcommand{\todo}[1]{}
\fi

\newcommand{\wss}[1]{\authornote{blue}{SS}{#1}} %Spencer Smith
\newcommand{\jc}[1]{\authornote{red}{JC}{#1}} %Jacques Carette
\newcommand{\oo}[1]{\authornote{magenta}{OO}{#1}} %Olu Owojaiye
\newcommand{\pmi}[1]{\authornote{green}{PM}{#1}} %Peter Michalski
\newcommand{\ad}[1]{\authornote{cyan}{AD}{#1}} %Ao Dong

\newcommand{\notdone}[1]{\textcolor{red}{#1}}
\newcommand{\done}[1]{\textcolor{black}{#1}}

%\oddsidemargin 0mm
%\evensidemargin 0mm
%\textwidth 160mm
%\textheight 200mm

\theoremstyle{definition}
\newtheorem{defn}{Definition}

\title{Experiments} 
\author{Spencer Smith}{McMaster University, Canada}{smiths@mcmaster.ca}{}{}
\author{Jacques Carette}{McMaster University, Canada}{carette@mcmaster.ca}{}{}
\author{Olu Owojaiye}{McMaster University, Canada}{owojaiyo@mcmaster.ca}{}{}
\author{Peter Michalski}{McMaster University, Canada}{michap@mcmaster.ca}{}{}
\author{Ao Dong}{McMaster University, Canada}{donga9@mcmaster.ca}{}{}

\authorrunning{Smith et al.}  \Copyright{Spencer Smith and Jacques Carette and
Olu Owojaiye and Peter Michalski and Ao Dong}

\date{\today}

\hideLIPIcs
\nolinenumbers

\begin{document}
\maketitle

~\newpage
\section{User Experiments}
\subsection{Usability Experimental Procedure}

\subsection {Procedure}
\begin {enumerate}
%\item Select a list of projects
%\item Write research questions
%\item Prepare logistic details(location, time table, recording setup, moderators etc)
%\item Prepare survey questions
%\item Design the tasks for the study subjects to perform (tasks can be defined based on user category, at least one task to modify the software)
%\item Select participants

\item Survey participants to collect pre-experiment data
\item Participants perform tasks
\item Observe the study subjects (take notes, record sessions(OBS screen recorder), watch out for body languages and verbal cues)
\item Survey the study subjects to collect feedback (post experiment interview)
\item Prepare experiment report
\item Perform pairwise comparison analysis
\item Prepare analysis report
\end {enumerate}

%\subsection{Nielsen’s Heuristics}
%\begin{enumerate}
%	\item Consistency: Check if standards and conventions in product design are followed
% \begin{itemize}
% 	\item Sequences of actions (skill acquisition).
% 	\item Color (categorization).
% 	\item Layout and position (spatial consistency).
% 	\item Font, capitalization (levels of organization).
% 	\item Terminology (delete, del, remove, rm) and language (words, phrases).
% 	\item Standards (e.g.,	blue underlined text for unvisited hyper links).
% \end{itemize}
%\item Visibility: Users should be informed about what is going on with the system through appropriate feedback and display of information e.g. 
% \begin{itemize}
% \item What is the current state of the system?
% \item What can be done at current state?
% \item Where can users go?
% \item What change is made after an action?
% \end{itemize}
%\item Match between system and world. The image of the system perceived by users should match the model the users have about the system. 
%\begin {itemize}
%\item User model matches system image.
%\item Actions provided by the system should match actions performed by users.
%\item Objects on the system should match objects of the task.
%\end {itemize}
%\item Minimalist: Any extraneous information is a distraction and a slow-down. 
%\begin{itemize}
%\item Less is more.
%\item Simple is not equivalent to abstract and general.
%\item Simple is efficient.
%\item  Progressive levels of detail.
%\end{itemize}
%\item Minimize memory load. Users should not be required to memorize a lot of information to carry out tasks. Memory load reduces users’ capacity to carry out the main tasks. 
%\begin{itemize}
%\item Recognition vs. recall (e.g., menu vs. commands).
%\item Externalize information through visualization.
%\item Perceptual procedures.
%\item Hierarchical structure.
%\item Default values.
%\item Concrete examples (DD/MM/YY, e.g., 10/20/99).
%\item Generic rules and actions (e.g., drag objects).
%\item Intuitive procedure
%\end{itemize}
%\item Informative feedback. Users should be given prompt and informative feedback about their actions. 
%\begin{itemize}
%\item Information that can be directly perceived, interpreted, and evaluated.
%\item Levels of feedback (novice and expert).
%\item Concrete and specific, not abstract and general.
%\item Response time:\\
%-0.1 s for instantaneously reacting;\\
%-1.0 s for uninterrupted flow of thought;\\
%-10 s for the limit of attention.
%\end{itemize}
%%\item Flexibility and efficiency. Users always learn and users are always different. Give users the flexibility of creating customization and shortcuts to accelerate their performance. 
%%\begin{itemize}
%%	\item Shortcuts for experienced users.
%%	\item Shortcuts or macros for frequently used operations.
%%	\item Skill acquisition through chunking.
%%	\item Examples:
%%	- Abbreviations, function keys, hot keys, command keys, macros, aliases, templates, type-ahead, bookmarks, hot links, history, default values, etc.
%%\end{itemize}
%\item Good error messages. 
%\begin {itemize}
%\item The messages should be informative enough such that users can understand the nature of errors, learn from errors, and recover from errors. 
%\item Phrased in clear language, avoid obscure codes. Example of obscure code: “system crashed, error code 147.”
%\item Precise, not vague or general. Example of general comment: “Cannot open document.”
%\item Constructive.
%\item Polite. Examples of impolite message: “illegal user action,” “job aborted,” “system was crashed,” “fatal error,” etc.
%\end {itemize}
%\item Prevent errors. It is always better to design interfaces that prevent errors from happening in the first place.
%\begin {itemize} 
%\item Interfaces that make errors impossible.
%\item Avoid modes (e.g., vi, text wrap). Or use informative feedback, e.g., different sounds.
%\item Execution error vs. evaluation error.
%\item Various types of slips and mistakes.
%\end {itemize}
%\item Clear closure. Every task has a beginning and an end. Users should be clearly notified about the completion of a task. 
%\begin {itemize}
%\item Clear beginning, middle, and end.
%\item Complete 7-stages of actions.
%\item Clear feedback to indicate goals are achieved and current stacks of goals can be released. Examples of good closures include many dialogues.
%\end {itemize}
%\item Reversible actions. 
%\begin {itemize}
%\item Users should be allowed to recover from errors. Reversible actions also encourage exploratory learning. 
%\item At different levels: a single action, a subtask, or a complete task.
%\item Multiple steps.
%\item Encourage exploratory learning.
%\item Prevent serious errors.
%\end {itemize}
%\item Use users’ language. The language should be always presented in a form understandable by the intended users.
%\begin{itemize}
%\item Use standard meanings of words.
%\item Specialized language for specialized group.
%\item User defined aliases.
%\item Users’ perspective. Example: “we have bought four tickets for you” (bad) vs. “you bought four tickets” (good).
%\end {itemize}
%\item Help and documentation. Always provide help when needed. 
%\begin {itemize}
%\item Context-sensitive help.
%\item Four types of help.\\
%-task-oriented;\\
%-alphabetically ordered;\\
%-semantically organized;\\
%-search.
%\item Help embedded in contents.
%\end {itemize}
%\item Users in control. Do not give users that impression that they are controlled by the systems. 
%\begin {itemize}
%\item Users are initiators of actors, not responders to actions.
%\item Avoid surprising actions, unexpected outcomes, tedious sequences of actions, etc.
%\end {itemize}
%\end{enumerate}

\subsection{Task selection criteria}
**The task selection will be determined with the aid of the domain experts attached to any of the selected projects.\\ 
**The domain experts will be asked to consider the below criteria when defining a task. \\
**Domain experts will also be asked to identify what background knowledge is necessary for the suggested tasks - Novice, Intermediate, Advanced
\begin {enumerate}
\item Collectively all tasks should not take no more than 2 hours.
\item Selected tasks should reflect common use cases of  the software.
\item Include tasks that require a set of sequential or hierarchical steps to be completed

\end {enumerate}

\subsection{Usability Questionnaire}
Two sources of standardized usability questionnaire we could use.

- \url{https://www.usabilitest.com/sus-pdf-generator}- 20-29 - SUS.\\
- \url{https://uiuxtrend.com/pssuq-post-study-system-usability-questionnaire/} - PSSUQ

~\newpage
\section{Modifiability Experiments}
copied from Methodology:

\pmi{note: these experiments need to be completed - to be discussed at next meeting}\pmi{meeting: start thinking about maintainability experiments; maybe change to modifiability; at least describe what we would do for such experiments - measure sensible changes, what does a small change mean exactly, document 'likely changes' - see Parnas; modifiability - we want to be explicit about what we want to modify; how to we interpret the results?; Add these experiments to user experiments file, in such case change file name, keep format similar to user experiments (procedure, no questionnaire, add preamble describing comments above), such experiments would get quantitative measures}\pmi{SS:For the maintainability experiment, I think we should write it up in the methodology, even if we don't do it. Discussing maintainability is needed to fit with the research questions (since maintainability is part of sustainability). We want sustainability to fit into the bigger context of what interests research software developers. As far as the maintainability experiment goes, we can just ask the domain experts to make a likely change and ask them to record their experience with making the change. It isn't quantifiable, but it would give us qualitative data.}
\subsection {Procedure}

\newpage

\bibliographystyle {plainnat}
\bibliography {../../CommonFiles/ResearchProposal}

\end{document}

### 2020-10-23 StateOfPractice/Methodology/Methodology.tex
updated Section 2 Modifiability experiment note, renamed experiments file, moved comments into file regarding modifiability experiments

@@
 \item Survey developers (\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Questions to Developers.pdf}{Questions to Developers})
-\item Usability experiments (\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/User Experiments.pdf}{User Experiments})
+\item Usability experiments (\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Experiments.pdf}{Experiments})
 \item Performance benchmarks\pmi{note: this is still in consideration}
-\item Maintainability experiments\pmi{note: these experiments need to be completed - to be discussed at next meeting}\pmi{meeting: start thinking about maintainability experiments; maybe change to modifiability; at least describe what we would do for such experiments - measure sensible changes, what does a small change mean exactly, document 'likely changes' - see Parnas; modifiability - we want to be explicit about what we want to modify; how to we interpret the results?; Add these experiments to user experiments file, in such case change file name, keep format similar to user experiments (procedure, no questionnaire, add preamble describing comments above), such experiments would get quantitative measures}\pmi{SS:For the maintainability experiment, I think we should write it up in the methodology, even if we don't do it.
-Discussing maintainability is needed to fit with the research questions (since maintainability is part of sustainability).
-We want sustainability to fit into the bigger context of what interests research software developers.
-As far as the maintainability experiment goes, we can just ask the domain experts to make a likely change and ask them to record their experience with making the change.
-It isn't quantifiable, but it would give us qualitative data.}
+\item Modifiability experiments (\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Experiments.pdf}{Experiments})
 \end{enumerate}

### 2020-10-23 StateOfPractice/Methodology/Experiments.tex
added comments into Experiments file regarding modifiability experiment

@@
 \section{Modifiability Experiments}
-copied from Methodology:
+\pmi{add preamble here describing some of the notes below}
 
-\pmi{note: these experiments need to be completed - to be discussed at next meeting}\pmi{meeting: start thinking about maintainability experiments; maybe change to modifiability; at least describe what we would do for such experiments - measure sensible changes, what does a small change mean exactly, document 'likely changes' - see Parnas; modifiability - we want to be explicit about what we want to modify; how to we interpret the results?; Add these experiments to user experiments file, in such case change file name, keep format similar to user experiments (procedure, no questionnaire, add preamble describing comments above), such experiments would get quantitative measures}\pmi{SS:For the maintainability experiment, I think we should write it up in the methodology, even if we don't do it.
-Discussing maintainability is needed to fit with the research questions (since maintainability is part of sustainability).
-We want sustainability to fit into the bigger context of what interests research software developers.
-As far as the maintainability experiment goes, we can just ask the domain experts to make a likely change and ask them to record their experience with making the change.
-It isn't quantifiable, but it would give us qualitative data.}
+What can be done to measure modifiability?
+
+Idea: measure sensible changes
+
+Can we measure sensible changes? How?
+
+We could gather qualitative data about the process of attempting to make sensible changes.
+Perceived difficulty, available documentation, prerequisite knowledge, time/effort required, tools required, errors and issues encountered and ability to overcome those issues with the available resources, is there dev support?
+How to interpret results?
+
+What does a small change mean? What are likely changes - See Parnas
+
+Parnas identifies likely changes as:
+
+ Likely Changes: if the system is required to be easy to change, the requirements should  contain  a  definition  of  the  areas  that  are  considered  likely  to  change.
+You  cannot design a system so that everything is equally easy to change.
+Programmers should not have to decide which changes are most likely\cite{parnas1986rational}
+
+We want to be explicit about what we want to modify.
+
 \subsection {Procedure}
+\begin{enumerate}
+	\item Identify likely changes using procedure x
+	\item Select from likely change list (using what criteria)
+	\item Gather relevant documentation?
+	\item Ask the domain expert to make the (selected) likely change(s)
+	\item Have the domain expert answer a questionnaire about their attempt to make the likely change
+\end{enumerate}
 

### 2020-10-23 StateOfPractice/Methodology/Methodology.tex
cleaned up intro, section 3 status comments, section 11 rank short list

@@
 \section{Introduction} \label{SecIntroduction}
+The scope of this document is to conduct a current state of practice analysis of software in a scientific computing domain for the purpose of understanding quality and assisting in future development.
+The methodology presented here is part of an overall assessment of the impact of model driven engineering and code generation on the sustainability of scientific computing software.
+The research proposal for this assessment, which contains a more detailed description of this state of practice exercise, can be found \href{https://github.com/smiths/AIMSS/blob/master/OverallResearchProposal/ResearchProposal.pdf}{here}.
+The research questions that this exercise will answer are found in Section \ref{ResearchQuestions}.
+Section \ref{StepsAQDS} outlines the steps to be taken.
 
-Purpose and scope of the document.  \wss{Needs to be filled in.  Should
-  reference the overall research proposal, and the ``state of the practice''
-  exercise in particular.  Reference questions we are trying to answer.}
-
-Note the following formatting conventions of this document. Red text denotes a
-link to information within the document. Purple text denotes a link to
-supplementary information that is external to this document. Blue text denotes a
-URL link.\pmi{needs to be added to intro to explain convention}
+ Note the following formatting conventions of this document. \textcolor{red}{Red} text denotes a link to information within this document. \textcolor{cyan}{Cyan} text denotes a URL link.
 
@@
 \item Identify broad list of candidate software packages in the domain. (Section~\ref{SecIdentifyCandSoft})
-\item Preliminary filter of software packages list. (Section~\ref{SecInitialFilter}) \pmi{To be edited based on feedback}
+\item Preliminary filter of software packages list. (Section~\ref{SecInitialFilter}) \pmi{in final review}
 \item \emph{Domain Experts}: Review domain software list.
 (\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Meeting Agenda withmDomainmExperts.pdf}{Meeting Agenda with Domain Experts})
-\item Domain Analysis. (Section~\ref{SecDomainAnalysis})\pmi{To be edited - add intro}
+\item Domain Analysis. (Section~\ref{SecDomainAnalysis})
 \item \emph{Domain Experts}: Vet domain analysis.
@@
 \section{Rank Short List} \label{SecRankShortList}
-Rank using pairwise comparison of short list software packages with respect to usability survey results.
-
-\pmi{SS: let's ask the people that do the usability/performance/maintainability experiments to do an AHP pair-wise comparison between the short-list software packages}
+An AHP pairwise comparison between the short list packages with respect to usability and modifiability survey results, as well as performance benchmarks, is conducted by the domain expert.
+The tool can be found here.
+The usability and modifiability survey can be found \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Experiments.pdf}{here}.
+Performance benchmarks can be found here.
+\pmi{are we adding a tool for this?}
+\pmi{performance benchmarks will be domain specific, how to proceed with providing a resource/guidance?}
 

### 2020-10-23 StateOfPractice/Methodology/Methodology.tex
added some notes into Section 9 Measure Using Measurment Template regarding ambiguities of measurement tamplate

@@
 
+\pmi{first pass for `tricky parts'; under summary information:
+	``Publications about the software''
+	list of publications can be found on some software websites, for others you may use google scholar or similar indexes,
+	``Is there evidence that performance was considered?''	Look for mention of any items, broad scope to answer yes but must include explanatory text: speed, storage, throughput consideration as well as comments on performance optimization/development, GPU processing, parallelism, multicore processing, or similar considerations}
+
 \begin{enumerate} 

### 2020-10-23 StateOfPractice/Methodology/Methodology.tex
updated Section 4 Methodology based on comments

@@
 \begin{enumerate}	
-	\item The domain must fall within the research software scope.
-	\item The domain must be well understood.
+	\item The domain must have a well-defined, stable theoretical underpinning. Standard textbooks, preferably understandable by an upper undergraduate student, is a good sign of this.
 	\item There must be a community of people studying the domain.
-	\item A preliminary search, or discussion with experts, suggests that there will be numerous, at least about 15, candidate software packages to study.
 	\item The software packages must have open source options. 
+	\item A preliminary search, or discussion with experts, suggests that there will be numerous, at least about 15, candidate software packages in the domain that qualify as `research software'.
 \end{enumerate}
 
-Some examples of domains that fit these criteria are seismology software, as well as mesh generators and processors.
-These are described in \citep{SmithEtAl2018} and \citep{smith2016state}, respectively.
+Some examples of domains that fit these criteria are finite element analysis, quantum chemistry, seismology, as well as mesh generators and processors.
+These are described in \citep{szabo1996finite}, \citep{veryazov20042molcas}, \citep{SmithEtAl2018} and \citep{smith2016state}, respectively.
 

### 2020-10-25 StateOfPractice/Methodology/Methodology.tex
added comments on "Measure Using Measurement Template"

@@
 	list of publications can be found on some software websites, for others you may use google scholar or similar indexes,
-	``Is there evidence that performance was considered?''	Look for mention of any items, broad scope to answer yes but must include explanatory text: speed, storage, throughput consideration as well as comments on performance optimization/development, GPU processing, parallelism, multicore processing, or similar considerations}
+``Is there evidence that performance was considered?'' Look for mention of any
+items, broad scope to answer yes but must include explanatory text: speed,
+storage, throughput consideration as well as comments on performance
+optimization/development, GPU processing, parallelism, multicore processing, or
+similar considerations} \ad{For ``performance'', the powerful search function
+from Github can help. What I did is searching ``performance'' in the repo on
+github.com, and usually I can get some author's comments regarding it. For
+projects not hosting on github (e.g. on SourceForge), I just upload it as my own
+repo, then to the search.}
+
+\ad{Tricky parts: (1). I second ``Publications'' - it's really hard to search by
+the software name sometimes. For example, I have software with names such as
+``DICOM Viewer'' and ``GATE''. (2). ``Initial release date'' - it's very hard to
+find the exact date, so for most of them, I used the release year. (3).
+``continuous integration'' - sometimes it's hard to find out from the repo if
+the authors don't mention it in any document. (4). ``getting started tutorial''
+vs ``user manual'' - many projects don't have a clear distinction between these
+two.}
+
+\ad{Tricky parts: (1). I second ``Publications'' - it's really hard to search by
+the software name sometimes. For example, I have software with names such as
+``DICOM Viewer'' and ``GATE''. (2). ``Initial release date'' - it's very hard to
+find the exact date, so for most of them, I used the release year. (3).
+``continuous integration'' - sometimes it's hard to find out from the repo if
+the authors don't mention it in any document. (4). ``getting started tutorial''
+vs ``user manual'' - many projects don't have a clear distinction between these
+two.}
+
+\ad{Might need improvements: ``Are constants (other than 0 and 1) hard coded
+into the program? (\{yes, no* , unclear\})''. Why we need extra comments if the
+answer is ``no''?}
 

### 2020-10-29 StateOfPractice/Methodology/Questions to Developers.tex
added comment about pilot into Questions to Developers
updated quesiton 10 an added comments into Questions to Developers

--- 2020-07-22 StateOfPractice/Ao-Notes/Questions to Developers.tex
@@
 \maketitle
-
+\pmi{A pilot will be conducted with these questions in order to test them. Test subjects may include PhD students.}
 \section{Information about the developers and users}\citep{Jegatheesan2016}
@@
 
+\pmi{consider adding `What, if any, actions were taken to address x?' to most/all questions, similar to question 10}
+\pmi{questions 1,3,4,5,6 build to visibility/transparency but a question would need to be modified to include something along the line of `has this information been added to the documentation?' or as above asking about actions}
 \begin{enumerate}
 \item Currently, what are the most significant obstacles in your development
-process?
+process? [visibility/transparency]
 \item How might you change your development process to remove or reduce these
@@
 \item How does documentation fit into your development process? Would improved
-documentation help with the obstacles you typically face?
+documentation help with the obstacles you typically face? [traceability, visibility/transparency]
 \item In the past, is there any major obstacle to your development process that
-has been solved? How did you solve it?
-\item What is your software development model? For example, waterfall, agile, etc.
+has been solved? How did you solve it? [visibility/transparency]
+\item What is your software development model? For example, waterfall, agile, etc. [visibility/transparency]
 \item What is your project management process? Do you think improve this process
-can tackle the current problem? Were any project management tools used?
+can tackle the current problem? Were any project management tools used? [visibility/transparency]
 \item \sout{Can you name several software qualities that are the most important to
@@
 \item Was it hard to ensure the correctness of the software? If there were any
- obstacles, what methods have been considered or practiced to improve the situation?
-  If practiced, did it work?
+ obstacles, what methods have been considered or practiced to improve the situation? If practiced, did it work? [correctness]
 \item When designing the software, did you consider the ease of future changes?
- For example, will it be hard to change the structure of the system, modules or code
-  blocks? What measures have been taken to ensure the ease of future changes
-   and maintains?
-\item Do you think the users can understand your software well? Has there been
-any problem to use this software?
+ For example, will it be hard to change the structure of the system, modules or code blocks?
+What measures have been taken to ensure the ease of future changes and maintains?
+[maintainability, modifiability]
+\item Provide instances where users have misunderstood the software.
+What, if any, actions were taken to address usability issues?
+[usability, understandability] \pmi{should this be split into two questions?}
 \item Do you think the current documentation can clearly convey all necessary
 knowledge to the users? If yes, how did you successfully achieve it? If no, what
-improvements are needed?
+improvements are needed? [unambiguity]
 \item Do you have any concern that your computational results won't be
-reproducible in the future? Have you taken any steps to ensure reproducibility?
+reproducible in the future? Have you taken any steps to ensure reproducibility? [reproducibility]
 \end{enumerate}

### 2020-10-29 StateOfPractice/Methodology/Methodology.tex
added comment about performance benchmarks into methodology

@@
 \section{Introduction} \label{SecIntroduction}
+
+\pmi{comment from meeting: remove performance benchmarks from scope of exercise because enough other people consider this.
+Add similar comment into introduction explaining (link to examples of performance considered elsewhere in similar projects??) why performance is not considered}
+
 The scope of this document is to conduct a current state of practice analysis of software in a scientific computing domain for the purpose of understanding quality and assisting in future development.

### 2020-10-29 StateOfPractice/Methodology/Methodology.tex
added comments from meeting regarding Rank Short List section of Methodology

@@
 \section{Rank Short List} \label{SecRankShortList}
+
+\pmi{comments from meeting: this part is to `enable conversation'; the ranking starts the conversation - `why do they like this software'; yes to tool question - spreadsheet; enable conversation regarding what's good about the specific softwares (also important highlights of domain software?); drop performance benchmarks}
+
 An AHP pairwise comparison between the short list packages with respect to usability and modifiability survey results, as well as performance benchmarks, is conducted by the domain expert.

### 2020-11-06 StateOfPractice/Methodology/Methodology.tex
removed sustainability from section 2 of methodology
removed performance benchmarks from scope

--- 2020-10-29 StateOfPractice/Methodology/Methodology.tex
@@
 Add similar comment into introduction explaining (link to examples of performance considered elsewhere in similar projects??) why performance is not considered}
+\pmi{do we have specific examples of other people considering performance benchmarks?}
 
 The scope of this document is to conduct a current state of practice analysis of software in a scientific computing domain for the purpose of understanding quality and assisting in future development.
+We leave the assessment of performance of such software to other similar projects.
 The methodology presented here is part of an overall assessment of the impact of model driven engineering and code generation on the sustainability of scientific computing software.
@@
 \begin{enumerate}
-\item sustainability?\pmi{previous meeting discussion did consider removing this; I do not see any current questions to developers that best fit this quality; confirm remove or add new question to developers?}
 \item usability \pmi{question 10 touches on this but will need to be reworded or broken up - ``What if any actions were taken to address usability issues?''}
@@
 (\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Meeting Agenda withmDomainmExperts.pdf}{Meeting Agenda with Domain Experts})
-\item Brief the Domain Experts on the overall objective, research proposal, research questions, measurement template, survey for short list projects, usability tests, performance benchmarks, maintainability experiments.
+\item Brief the Domain Experts on the overall objective, research proposal, research questions, measurement template, survey for short list projects, usability tests, maintainability experiments.
 (\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Meeting Agenda withmDomainmExperts.pdf}{Meeting Agenda with Domain Experts})
@@
 \item Usability experiments (\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Experiments.pdf}{Experiments})
-\item Performance benchmarks\pmi{note: this is still in consideration}
 \item Modifiability experiments (\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Experiments.pdf}{Experiments})
@@
 \section{Rank Short List} \label{SecRankShortList}
-An AHP pairwise comparison between the short list packages with respect to usability and modifiability survey results, as well as performance benchmarks, is conducted by the domain expert.
+
+\pmi{comments from meeting: this part is to `enable conversation'; the ranking starts the conversation - `why do they like this software'; yes to tool question - spreadsheet; enable conversation regarding what's good about the specific softwares (also important highlights of domain software?)}
+
+An AHP pairwise comparison between the short list packages with respect to usability and modifiability survey results is conducted by the domain expert.
 The tool can be found here.
 The usability and modifiability survey can be found \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Experiments.pdf}{here}.
-Performance benchmarks can be found here.
 \pmi{are we adding a tool for this?}
-\pmi{performance benchmarks will be domain specific, how to proceed with providing a resource/guidance?}
 

### 2020-11-06 StateOfPractice/Methodology/Methodology.tex
updated section 2 research questions, methodology, refs to questions to devs

@@
 Add similar comment into introduction explaining (link to examples of performance considered elsewhere in similar projects??) why performance is not considered}
-\pmi{do we have specific examples of other people considering performance benchmarks?}
+\pmi{do we have specific examples of other works considering performance benchmarks?}
 
@@
 \begin{enumerate}
-\item usability \pmi{question 10 touches on this but will need to be reworded or broken up - ``What if any actions were taken to address usability issues?''}
+\item usability \pmi{question to developers 11}
 \item traceability \pmi{question to developers 3}
@@
 \item correctness \pmi{question to developers 8}
-\item understandability \pmi{question to developers 10 touches on this; need to reword question in terms of an action}
-\item unambiguity \pmi{question to developers 11}
-\item reproducibility \pmi{question to developers 12}
-\item visibility/transparency \pmi{questions 1,3,4,5,6 build to this but a question would need to be modified to include something along the line of ``has this information been added to the documentation?''}
+\item understandability \pmi{question to developers 10}
+\item unambiguity \pmi{question to developers 12}
+\item reproducibility \pmi{question to developers 13}
+\item visibility/transparency \pmi{question to developers 3}
 \end{enumerate} 
-\item How does software designated as high quality by this methodology compare with top rated software by the community? \pmi{review}
+\item How does software designated as high quality by this methodology compare with top rated software by the community?
 \end{enumerate}

### 2020-11-06 StateOfPractice/Methodology/Methodology.tex
updated section 3 and 11, methodology

@@
 \begin{enumerate}
-\item Start with the state of practice research questions. (Section~\ref{ResearchQuestions}) \pmi{To be updated based on feedback}
-\item Identify the domain. (Section~\ref{SecIdentifyDomain}) \pmi{In second review}
+\item Start with the state of practice research questions. (Section~\ref{ResearchQuestions}) \pmi{In review}
+\item Identify the domain. (Section~\ref{SecIdentifyDomain})
 \item \emph{Domain Experts}: Create a top ten list of software packages in the domain.
@@
 \item Identify broad list of candidate software packages in the domain. (Section~\ref{SecIdentifyCandSoft})
-\item Preliminary filter of software packages list. (Section~\ref{SecInitialFilter}) \pmi{in final review}
+\item Preliminary filter of software packages list. (Section~\ref{SecInitialFilter})
 \item \emph{Domain Experts}: Review domain software list.
@@
 
-\pmi{comments from meeting: this part is to `enable conversation'; the ranking starts the conversation - `why do they like this software'; yes to tool question - spreadsheet; enable conversation regarding what's good about the specific softwares (also important highlights of domain software?)}
-
 An AHP pairwise comparison between the short list packages with respect to usability and modifiability survey results is conducted by the domain expert.
@@
 The usability and modifiability survey can be found \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Experiments.pdf}{here}.
-\pmi{are we adding a tool for this?}
+This step is intended to start a conversation about the features of specific software packages.
 

### 2020-11-06 StateOfPractice/Methodology/Methodology.tex
updated section 9, methodology

@@
 \item Collect empirical measures. (Section~\ref{SecEmpiricalMeasures})
-\item Measure using measurement template. (Section~\ref{SecShallowMeasure}) \pmi{To be edited based on feedback}
+\item Measure using measurement template. (Section~\ref{SecShallowMeasure}) \pmi{To be reviewed}
 \item Use AHP process to rank the software packages. (Section~\ref{SecAHP})
@@
 \end{enumerate}
-\item Rank short list. (Section~\ref{SecRankShortList}) \pmi{To be completed}
+\item Rank short list. (Section~\ref{SecRankShortList}) \pmi{To be completed - add pairwise comparison tool}
 \item Document answers for research questions.
@@
 Project developers can be contacted for help regarding installation, if necessary, but a cap of about 2 hours should be imposed on the entire installation process.
-For completing the template, please follow these steps:
 
-\pmi{SS:"explain any tricky parts with interpreting the parameters of variation for measuring a particular indicator. A few example would help."}\pmi{Ask Ao for input}
+Some guidance on applying sections of the template:
 
-\pmi{first pass for `tricky parts'; under summary information:
-	``Publications about the software''
-	list of publications can be found on some software websites, for others you may use google scholar or similar indexes,
-``Is there evidence that performance was considered?'' Look for mention of any
-items, broad scope to answer yes but must include explanatory text: speed,
-storage, throughput consideration as well as comments on performance
-optimization/development, GPU processing, parallelism, multicore processing, or
-similar considerations} \ad{For ``performance'', the powerful search function
-from Github can help. What I did is searching ``performance'' in the repo on
-github.com, and usually I can get some author's comments regarding it. For
-projects not hosting on github (e.g. on SourceForge), I just upload it as my own
-repo, then to the search.}
-
-\ad{Tricky parts: (1). I second ``Publications'' - it's really hard to search by
-the software name sometimes. For example, I have software with names such as
-``DICOM Viewer'' and ``GATE''. (2). ``Initial release date'' - it's very hard to
-find the exact date, so for most of them, I used the release year. (3).
-``continuous integration'' - sometimes it's hard to find out from the repo if
-the authors don't mention it in any document. (4). ``getting started tutorial''
-vs ``user manual'' - many projects don't have a clear distinction between these
-two.}
-
-\ad{Tricky parts: (1). I second ``Publications'' - it's really hard to search by
-the software name sometimes. For example, I have software with names such as
-``DICOM Viewer'' and ``GATE''. (2). ``Initial release date'' - it's very hard to
-find the exact date, so for most of them, I used the release year. (3).
-``continuous integration'' - sometimes it's hard to find out from the repo if
-the authors don't mention it in any document. (4). ``getting started tutorial''
-vs ``user manual'' - many projects don't have a clear distinction between these
-two.}
+\begin{enumerate}
+	\item Publications about the software: A list of publications can be found directly on the website of some software packages. For others use Google Scholar or a similar index.
+	\item Initial release date: Mark the release year if an exact date is not available.
+	\item Continuous integration: Search the software artifacts for any mention of continuous integration. The search function on GitHub can help. 
+	\item Is there evidence that performance was considered?: Search the software artifacts for any mention of speed, storage, throughput, performance optimization, parallelism, multi-core processing, or similar considerations.
+The search function on GitHub can help.
+	\item Getting started tutorial: Sometimes this is found within another artifact, like the user manual.
+\end{enumerate}
 
 \ad{Might need improvements: ``Are constants (other than 0 and 1) hard coded
-into the program? (\{yes, no* , unclear\})''. Why we need extra comments if the
-answer is ``no''?}
+	into the program? (\{yes, no* , unclear\})''. Why we need extra comments if the	answer is ``no''?}\pmi{good question, to be brought up at meeting}
+
+For completing the template, please follow these steps:
 

### 2020-11-10 StateOfPractice/Methodology/Methodology.tex
updated section 9, methodology
Reformatted methodology document to have fixed width for lines of text

@@
 
-\pmi{comment from meeting: remove performance benchmarks from scope of exercise because enough other people consider this.
-Add similar comment into introduction explaining (link to examples of performance considered elsewhere in similar projects??) why performance is not considered}
-\pmi{do we have specific examples of other works considering performance benchmarks?}
+\pmi{comment from meeting: remove performance benchmarks from scope of exercise
+  because enough other people consider this. Add similar comment into
+  introduction explaining (link to examples of performance considered elsewhere
+  in similar projects??) why performance is not considered} \pmi{do we have
+  specific examples of other works considering performance benchmarks?}
 
-The scope of this document is to conduct a current state of practice analysis of software in a scientific computing domain for the purpose of understanding quality and assisting in future development.
-We leave the assessment of performance of such software to other similar projects.
-The methodology presented here is part of an overall assessment of the impact of model driven engineering and code generation on the sustainability of scientific computing software.
-The research proposal for this assessment, which contains a more detailed description of this state of practice exercise, can be found \href{https://github.com/smiths/AIMSS/blob/master/OverallResearchProposal/ResearchProposal.pdf}{here}.
-The research questions that this exercise will answer are found in Section \ref{ResearchQuestions}.
-Section \ref{StepsAQDS} outlines the steps to be taken.
+The scope of this document is to conduct a current state of practice analysis of
+software in a scientific computing domain for the purpose of understanding
+quality and assisting in future development. We leave the assessment of
+performance of such software to other similar projects. The methodology
+presented here is part of an overall assessment of the impact of model driven
+engineering and code generation on the sustainability of scientific computing
+software. The research proposal for this assessment, which contains a more
+detailed description of this state of practice exercise, can be found
+\href{https://github.com/smiths/AIMSS/blob/master/OverallResearchProposal/ResearchProposal.pdf}{here}. The
+research questions that this exercise will answer are found in Section
+\ref{ResearchQuestions}. Section \ref{StepsAQDS} outlines the steps to be taken.
 
- Note the following formatting conventions of this document. \textcolor{red}{Red} text denotes a link to information within this document. \textcolor{cyan}{Cyan} text denotes a URL link.
+Note the following formatting conventions of this document. \textcolor{red}{Red}
+text denotes a link to information within this document. \textcolor{cyan}{Cyan}
+text denotes a URL link.
 
@@
 
-The following are research questions that we want to answer to assess the state of practice of research software in our chosen domain:
+The following are research questions that we want to answer to assess the state
+of practice of research software in our chosen domain:
 
@@
 \item What tools are used by current software packages?
-\item What principles, processes, and methodologies are used in the development of current software packages?
-\item What are the pain points for developers working on SCS projects?
-What aspects of the existing processes, methodologies and tools do they consider could potentially be improved?
-What processes, methodologies and tools have they changed to reduce obstacles?
+\item What principles, processes, and methodologies are used in the development
+  of current software packages?
+\item What are the pain points for developers working on SCS projects? What
+  aspects of the existing processes, methodologies and tools do they consider
+  could potentially be improved? What processes, methodologies and tools have
+  they changed to reduce obstacles?
 \item For research software developers, what specific actions are taken to address the following:
@@
 \end{enumerate} 
-\item How does software designated as high quality by this methodology compare with top rated software by the community?
+\item How does software designated as high quality by this methodology compare
+  with top rated software by the community?
 \end{enumerate}
@@
 \begin{enumerate}
-\item Start with the state of practice research questions. (Section~\ref{ResearchQuestions}) \pmi{In review}
+\item Start with the state of practice research
+  questions. (Section~\ref{ResearchQuestions}) \pmi{In review}
 \item Identify the domain. (Section~\ref{SecIdentifyDomain})
-\item \emph{Domain Experts}: Create a top ten list of software packages in the domain.
-(\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Meeting Agenda withmDomainmExperts.pdf}{Meeting Agenda with Domain Experts})
-\item Brief the Domain Experts on the overall objective, research proposal, research questions, measurement template, survey for short list projects, usability tests, maintainability experiments.
-(\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Meeting Agenda withmDomainmExperts.pdf}{Meeting Agenda with Domain Experts})
-\item Identify broad list of candidate software packages in the domain. (Section~\ref{SecIdentifyCandSoft})
-\item Preliminary filter of software packages list. (Section~\ref{SecInitialFilter})
-\item \emph{Domain Experts}: Review domain software list.
-(\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Meeting Agenda withmDomainmExperts.pdf}{Meeting Agenda with Domain Experts})
+\item \emph{Domain Experts}: Create a top ten list of software packages in the
+  domain. (\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Meeting
+    Agenda withmDomainmExperts.pdf}{Meeting Agenda with Domain Experts})
+\item Brief the Domain Experts on the overall objective, research proposal,
+  research questions, measurement template, survey for short list projects,
+  usability tests, maintainability
+  experiments. (\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Meeting
+    Agenda withmDomainmExperts.pdf}{Meeting Agenda with Domain Experts})
+\item Identify broad list of candidate software packages in the
+  domain. (Section~\ref{SecIdentifyCandSoft})
+\item Preliminary filter of software packages
+  list. (Section~\ref{SecInitialFilter})
+\item \emph{Domain Experts}: Review domain software
+  list. (\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Meeting
+    Agenda withmDomainmExperts.pdf}{Meeting Agenda with Domain Experts})
 \item Domain Analysis. (Section~\ref{SecDomainAnalysis})
-\item \emph{Domain Experts}: Vet domain analysis.
-(\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Meeting Agenda withmDomainmExperts.pdf}{Meeting Agenda with Domain Experts})
-\item Gather source code and documentation for each prospective software package.
+\item \emph{Domain Experts}: Vet domain
+  analysis. (\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Meeting
+    Agenda withmDomainmExperts.pdf}{Meeting Agenda with Domain Experts})
+\item Gather source code and documentation for each prospective software
+  package.
 \item Collect empirical measures. (Section~\ref{SecEmpiricalMeasures})
-\item Measure using measurement template. (Section~\ref{SecShallowMeasure}) \pmi{To be reviewed}
+\item Measure using measurement template. (Section~\ref{SecShallowMeasure})
+  \pmi{To be reviewed}
 \item Use AHP process to rank the software packages. (Section~\ref{SecAHP})
-\item Identify a short list of top software packages, typically four to six, for deeper exploration according to the AHP rankings of the measurements.
-\item \emph{Domain Experts}: Vet AHP ranking and short list.
-(\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Meeting Agenda withmDomainmExperts.pdf}{Meeting Agenda with Domain Experts})
+\item Identify a short list of top software packages, typically four to six, for
+  deeper exploration according to the AHP rankings of the measurements.
+\item \emph{Domain Experts}: Vet AHP ranking and short
+  list. (\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Meeting
+    Agenda withmDomainmExperts.pdf}{Meeting Agenda with Domain Experts})
 \item With short list:
 \begin{enumerate}
-\item Survey developers (\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Questions to Developers.pdf}{Questions to Developers})
-\item Usability experiments (\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Experiments.pdf}{Experiments})
-\item Modifiability experiments (\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Experiments.pdf}{Experiments})
+\item Survey developers
+  (\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Questions
+    to Developers.pdf}{Questions to Developers})
+\item Usability experiments
+  (\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Experiments.pdf}{Experiments})
+\item Modifiability experiments
+  (\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Experiments.pdf}{Experiments})
 \end{enumerate}
-\item Rank short list. (Section~\ref{SecRankShortList}) \pmi{To be completed - add pairwise comparison tool}
+\item Rank short list. (Section~\ref{SecRankShortList}) \pmi{To be completed -
+    add pairwise comparison tool}
 \item Document answers for research questions.
@@
 \section{How to Identify the Domain} \label{SecIdentifyDomain}
-A domain of research software must be identified.
-Research software is defined in this exercise as ``software that is used to generate, process or analyze results that [are intended] to appear in a publication'' \citep{hettrick2014uk}.
-The chosen domain must have the following properties:
+
+A domain of research software must be identified. Research software is defined
+in this exercise as ``software that is used to generate, process or analyze
+results that [are intended] to appear in a publication''
+\citep{hettrick2014uk}. The chosen domain must have the following properties:
 
 \begin{enumerate}
-	\item The domain must have a well-defined, stable theoretical underpinning. Standard textbooks, preferably understandable by an upper undergraduate student, is a good sign of this.
+\item The domain must have a well-defined, stable theoretical
+  underpinning. Standard textbooks, preferably understandable by an upper
+  undergraduate student, is a good sign of this.
 \item There must be a community of people studying the domain.
 \item The software packages must have open source options.
-	\item A preliminary search, or discussion with experts, suggests that there will be numerous, at least about 15, candidate software packages in the domain that qualify as `research software'.
+\item A preliminary search, or discussion with experts, suggests that there will
+  be numerous, at least about 15, candidate software packages in the domain that
+  qualify as `research software'.
 \end{enumerate}
 
-Some examples of domains that fit these criteria are finite element analysis, quantum chemistry, seismology, as well as mesh generators and processors.
-These are described in \citep{szabo1996finite}, \citep{veryazov20042molcas}, \citep{SmithEtAl2018} and \citep{smith2016state}, respectively.
+Some examples of domains that fit these criteria are finite element analysis,
+quantum chemistry, seismology, as well as mesh generators and processors. These
+are described in \citep{szabo1996finite}, \citep{veryazov20042molcas},
+\citep{SmithEtAl2018} and \citep{smith2016state}, respectively.
 
 \section{How to Identify Candidate Software} \label{SecIdentifyCandSoft}
-The candidate software can be found through search engine queries targeting authoritative lists of software from the domain, such as lists on GitHub and swMATH, and by searching domain related publications.
-Domain experts are also asked for their suggestions and are asked to review the list.
-The candidate software should have the following properties:
+
+The candidate software can be found through search engine queries targeting
+authoritative lists of software from the domain, such as lists on GitHub and
+swMATH, and by searching domain related publications. Domain experts are also
+asked for their suggestions and are asked to review the list. The candidate
+software should have the following properties:
 
@@
 \item Must have viewable source code.
-	\item Ideally have a git repository or ability to gather empirical measures found in Section \ref{SecEmpiricalMeasures}.
-	\item The software cannot be marked as incomplete or in an initial development phase.
+\item Ideally have a git repository or ability to gather empirical measures
+  found in Section \ref{SecEmpiricalMeasures}.
+\item The software cannot be marked as incomplete or in an initial development
+  phase.
 \end{enumerate}
@@
 \section{How to Initially Filter the Software List} \label{SecInitialFilter}
-If the list of software is too long (over around 30 packages), then steps need to be taken to create a more manageable list.
-To reduce the length of the list, the following filters are applied in the priority order listed, stopping once the desired number of projects is reached:
+
+If the list of software is too long (over around 30 packages), then steps need
+to be taken to create a more manageable list. To reduce the length of the list,
+the following filters are applied in the priority order listed, stopping once
+the desired number of projects is reached:
 
 \begin{enumerate}
-	\item Scope: The functionality of the domain software should be increasingly specific.
-	\item Usage: The installation procedure should appear to be clear or easy to figure out.
-    \item Age: Ideally the latest release or source code commit should be relatively recent for software in the domain unless the candidate software appears to be well recommended and currently in use.
+\item Scope: The functionality of the domain software should be increasingly
+  specific.
+\item Usage: The installation procedure should appear to be clear or easy to
+  figure out.
+\item Age: Ideally the latest release or source code commit should be relatively
+  recent for software in the domain unless the candidate software appears to be
+  well recommended and currently in use.
 \end{enumerate}
@@
 
-The domain analysis consists of a commonality analysis of the family of software packages.
-Its purpose is to show the relationships between these packages, and to facilitate an understanding of the informal specification and development of them.
-\cite{weiss1997defining} defines commonality analysis as an approach to defining a family by identifying commonalities, variabilities, and common terminology for the family.
-Commonalities are goals, theories, models, definitions and assumptions that are common between family members.
-Likewise, variabilities are goals, theories, models, definitions and assumptions that are different between family members.
-The final result of the domain analysis will be tables of commonalities, variabilities, and parameters of variation of a program family, which \cite{parnas1976design} defines as ``a set of programs whose common properties are so extensive that it is advantageous to study the common properties of the programs before analyzing individual members''.
-\cite{smith2008commonality} present a template for conducting a commonality analysis, which was referred to when conducting this work.
-\cite{weiss1998commonality} describes another commonality analysis technique for deciding the members of a program family which readers are encouraged to be familiar with.
-\cite{SmithAndChen2004} and \cite{SmithMcCutchanAndCarette2017} are examples of a commonality analysis for a family of mesh generating software and a family of material models, respectively.
-The steps to produce a commonality analysis are:
+The domain analysis consists of a commonality analysis of the family of software
+packages. Its purpose is to show the relationships between these packages, and
+to facilitate an understanding of the informal specification and development of
+them. \cite{weiss1997defining} defines commonality analysis as an approach to
+defining a family by identifying commonalities, variabilities, and common
+terminology for the family. Commonalities are goals, theories, models,
+definitions and assumptions that are common between family members. Likewise,
+variabilities are goals, theories, models, definitions and assumptions that are
+different between family members. The final result of the domain analysis will
+be tables of commonalities, variabilities, and parameters of variation of a
+program family, which \cite{parnas1976design} defines as ``a set of programs
+whose common properties are so extensive that it is advantageous to study the
+common properties of the programs before analyzing individual
+members''. \cite{smith2008commonality} present a template for conducting a
+commonality analysis, which was referred to when conducting this
+work. \cite{weiss1998commonality} describes another commonality analysis
+technique for deciding the members of a program family which readers are
+encouraged to be familiar with. \cite{SmithAndChen2004} and
+\cite{SmithMcCutchanAndCarette2017} are examples of a commonality analysis for a
+family of mesh generating software and a family of material models,
+respectively. The steps to produce a commonality analysis are:
 
@@
 \section{Empirical Measures} \label{SecEmpiricalMeasures}
-Some quality measurements rely on the gathering of raw and processed empirical data.
-We focus on data that is reasonably easy to collect, which we combine and analyze for this assessment.
-The measures that are collected relate to the research questions that we want to answer.
-For instance, we collect some of the data to see how large a project is.
-Other measures intend to ascertain a project’s popularity, or how active the project is.
-Section \ref{rawdata} will orient the reader as to what raw data is collected. 
-Some of this data can be observed from Git repository metrics.
-The rest can be collected using freeware tools.
-\href{https://github.com/tomgi/git_stats}{GitStats} is used to measure the number of binary files as well as the number of added and deleted lines in a repository.
-The tool is also used to measure the number of commits over different intervals of time.
-\href{https://github.com/boyter/scc}{Sloc Cloc and Code (scc)} is used to measure the number of text based files as well as the number of total, code, comment, and blank lines in a repository.
-These tools were selected due to their installability, usability, and ability to gather the empirical measures listed below.
-A guide for installing and running them can be found \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/A Guide to Empirical Measures.pdf}{here}.
-Section \ref{processeddata} introduces the required processed data, which is calculated using the raw data.
+
+Some quality measurements rely on the gathering of raw and processed empirical
+data. We focus on data that is reasonably easy to collect, which we combine and
+analyze for this assessment. The measures that are collected relate to the
+research questions that we want to answer. For instance, we collect some of the
+data to see how large a project is. Other measures intend to ascertain a
+project’s popularity, or how active the project is.  Section \ref{rawdata} will
+orient the reader as to what raw data is collected.  Some of this data can be
+observed from Git repository metrics. The rest can be collected using freeware
+tools. \href{https://github.com/tomgi/git_stats}{GitStats} is used to measure
+the number of binary files as well as the number of added and deleted lines in a
+repository. The tool is also used to measure the number of commits over
+different intervals of time. \href{https://github.com/boyter/scc}{Sloc Cloc and
+  Code (scc)} is used to measure the number of text based files as well as the
+number of total, code, comment, and blank lines in a repository. These tools
+were selected due to their installability, usability, and ability to gather the
+empirical measures listed below. A guide for installing and running them can be
+found
+\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/A
+  Guide to Empirical Measures.pdf} {here}. Section \ref{processeddata}
+introduces the required processed data, which is calculated using the raw data.
 
 \subsection{Raw Data}\label{rawdata}
+
 The following raw data measures are extracted from repositories:
@@
 \subsection{Processed Data}\label{processeddata}
+
 The following measures are calculated from the raw data:
@@
 \begin{itemize}
-\item Status of software package (dead or alive). Alive is defined as the presence of repository commits or software package version releases in the last 18 months.
+\item Status of software package (dead or alive). Alive is defined as the
+  presence of repository commits or software package version releases in the
+  last 18 months.
 \item Percentage of identified issues that are closed.
@@
     \label{measurement_template_image}
-
   \end{center}
@@
 
-The \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Combined_MeasurementTemplate_EmpiricalMeasures.xlsx}{Measurement Template} is used to track measurements and quality scores for all of the software packages in the domain.
-For each software package, fill out one column of the template.
-This process can take between 2 to 5 hours for each package.
-Project developers can be contacted for help regarding installation, if necessary, but a cap of about 2 hours should be imposed on the entire installation process.
+The
+\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Combined_MeasurementTemplate_EmpiricalMeasures.xlsx}
+{Measurement Template} is used to track measurements and quality scores for all
+of the software packages in the domain. For each software package, fill out one
+column of the template. This process can take between 2 to 5 hours for each
+package.  Project developers can be contacted for help regarding installation,
+if necessary, but a cap of about 2 hours should be imposed on the entire
+installation process.
 
@@
 \begin{enumerate}
-	\item Publications about the software: A list of publications can be found directly on the website of some software packages. For others use Google Scholar or a similar index.
-	\item Initial release date: Mark the release year if an exact date is not available.
-	\item Continuous integration: Search the software artifacts for any mention of continuous integration. The search function on GitHub can help. 
-	\item Is there evidence that performance was considered?: Search the software artifacts for any mention of speed, storage, throughput, performance optimization, parallelism, multi-core processing, or similar considerations.
-The search function on GitHub can help.
-	\item Getting started tutorial: Sometimes this is found within another artifact, like the user manual.
+\item Initial release date: Mark the release year if an exact date is not
+  available.
+\item Publications about the software: A list of publications can be found
+  directly on the website of some software packages. For others use Google
+  Scholar or a similar index.
+\item Is there evidence that performance was considered?: Search the software
+  artifacts for any mention of speed, storage, throughput, performance
+  optimization, parallelism, multi-core processing, or similar
+  considerations. The search function on GitHub can help.
+\item Getting started tutorial: Sometimes this is found within another artifact,
+  like the user manual.
+\item Continuous integration: Search the software artifacts for any mention of
+  continuous integration. The search function on GitHub can help.
 \end{enumerate}
@@
 \ad{Might need improvements: ``Are constants (other than 0 and 1) hard coded
-	into the program? (\{yes, no* , unclear\})''. Why we need extra comments if the	answer is ``no''?}\pmi{good question, to be brought up at meeting}
+  into the program? (\{yes, no* , unclear\})''. Why we need extra comments if
+  the answer is ``no''?}\pmi{good question, to be brought up at meeting}
 
@@
 \begin{enumerate}
-	\item Gather the summary information into the top section of the document (Figure \ref{measurement_template_image})
-	\item Using the GitStats tool that is described in Section \ref{SecEmpiricalMeasures} gather the measurements for the Repo Metrics - GitStats section found near the bottom of the document
-	\item Using the SCC tool that is also described in Section \ref{SecEmpiricalMeasures} gather the measurements for the Repo Metrics - SCC section found near the bottom of the document
-	\item If the software package is found on git, gather the measurements for the Repo Metrics - GitHub section found near the bottom of the document
-	\item Review installation documentation and attempt to install the software package on a virtual machine
+\item Gather the summary information into the top section of the document
+  (Figure \ref{measurement_template_image})
+\item Using the GitStats tool that is described in Section
+  \ref{SecEmpiricalMeasures} gather the measurements for the Repo Metrics -
+  GitStats section found near the bottom of the document
+\item Using the SCC tool that is also described in Section
+  \ref{SecEmpiricalMeasures} gather the measurements for the Repo Metrics - SCC
+  section found near the bottom of the document
+\item If the software package is found on git, gather the measurements for the
+  Repo Metrics - GitHub section found near the bottom of the document
+\item Review installation documentation and attempt to install the software
+  package on a virtual machine
 \item Gather the measurements for installability
@@
 \item Gather the measurements for visibility and transparency
-	\item Assign a score out of ten for each quality.
-The score can be measured using the \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/MeasurementTemplate_ImpressionCalculator.xlsx}{Measurement Template Impression Calculator}.
-For each quality measurement, the file indicates the appropriate score to assign the measurement based on possible measurement values.
+\item Assign a score out of ten for each quality. The score can be measured
+  using the
+  \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/MeasurementTemplate_ImpressionCalculator.xlsx}
+  {Measurement Template Impression Calculator}. For each quality measurement,
+  the file indicates the appropriate score to assign the measurement based on
+  possible measurement values.
 \end{enumerate}
@@
 \section{Analytic Hierarchy Process} \label{SecAHP}
-The Analytical Hierarchy Process (AHP) is a decision-making technique that can be used when comparing multiple options by multiple criteria.
-In our work AHP is used for comparing and ranking the software packages of a domain using the quality scores that are gathered in the \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Combined_MeasurementTemplate_EmpiricalMeasures.xlsx}{Measurement Template}.
-AHP performs a pairwise analysis between each of the quality options using a matrix which is then used to generate an overall score for each software package for the given criteria.
-\cite{SmithEtAl2016} shows how AHP is applied to ranking software based on quality measures.
-We have developed a tool for conducting this process.
-The tool includes an AHP JAR script and a sensitivity analysis JAR script that is used to ensure that the software package rankings are appropriate with respect to the uncertainty of the quality scores.
-The README file of the tool is found \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/AHP2020/LBM/README.txt}{here}.
-This file outlines the requirements for, and configuration and usage of, the JAR scripts.
-The JAR scripts, source code, and required libraries are located in the same folder as the README file.
+
+The Analytical Hierarchy Process (AHP) is a decision-making technique that can
+be used when comparing multiple options by multiple criteria. In our work AHP is
+used for comparing and ranking the software packages of a domain using the
+quality scores that are gathered in the
+\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Combined_MeasurementTemplate_EmpiricalMeasures.xlsx}
+{Measurement Template}. AHP performs a pairwise analysis between each of the
+quality options using a matrix which is then used to generate an overall score
+for each software package for the given criteria. \cite{SmithEtAl2016} shows how
+AHP is applied to ranking software based on quality measures. We have developed
+a tool for conducting this process. The tool includes an AHP JAR script and a
+sensitivity analysis JAR script that is used to ensure that the software package
+rankings are appropriate with respect to the uncertainty of the quality
+scores. The README file of the tool is found
+\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/AHP2020/LBM/README.txt}{here}. This
+file outlines the requirements for, and configuration and usage of, the JAR
+scripts. The JAR scripts, source code, and required libraries are located in the
+same folder as the README file.
 
@@
 
-An AHP pairwise comparison between the short list packages with respect to usability and modifiability survey results is conducted by the domain expert.
-The tool can be found here.
-The usability and modifiability survey can be found \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Experiments.pdf}{here}.
-This step is intended to start a conversation about the features of specific software packages.
+An AHP pairwise comparison between the short list packages with respect to
+usability and modifiability survey results is conducted by the domain
+expert. The tool can be found here. The usability and modifiability survey can
+be found
+\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Experiments.pdf}
+{here}. This step is intended to start a conversation about the features of
+specific software packages.
 
@@
 \appendix
+
 \section{Appendix}
+
 \subsection{Survey for the Selected Projects}
-\ad{Several questions are borrowed from \href{https://gitlab.cas.mcmaster.ca/smiths/pub/-/blob/master/Jegatheesan2016.pdf}{Jegatheesan2016}, and needed to be cited later.}
+
+\ad{Several questions are borrowed from
+  \href{https://gitlab.cas.mcmaster.ca/smiths/pub/-/blob/master/Jegatheesan2016.pdf}{Jegatheesan2016},
+  and needed to be cited later.}
+
 \subsubsection{Information about the developers and users}
+
 \begin{enumerate}
@@
 \begin{enumerate}
-\item \ad{General} What is the most important software quality(ies) to your work? (set of selected qualities plus "else")
-\item \ad{General} Are there any examples where the documentation helped? If yes, how it helped. ({yes$^*$, no})
-\item \ad{General} Is there any documentation you feel you should produce and do not? If yes, what is it and why? ({yes$^*$, no})
-\item \ad{Completeness} Do you address any of your quality concerns using documentation? If yes, what are the qualities and the documents. ({yes$^*$, no})
-\item \ad{Visibility/Transparency} Is there a certain type of development methodologies used during the development? (\{Waterfall, Scrum, Kanban, else\})
-\item \ad{Visibility/Transparency} Is there a clearly defined development process? If yes, what is it. (\{yes$^*$, no\})
-\item \ad{Visibility/Transparency} Are there any project management tools used during the development? If yes, what are they. (\{yes$^*$, no\})
-\item \ad{Visibility/Transparency} Going forward, will your approach to documentation of requirements and design
-change? If not, why not. (\{yes, no$^*$\})
-\item \ad{Correctness and Verifiability} During the process of development, what tools or techniques are used to build confidence of correctness? (string)
-\item \ad{Correctness and Verifiability} Do you use any tools to support testing? If yes, what are they. (e.g. unit testing tools, regression testing suites) (\{yes$^*$, no\})
-\item \ad{Correctness and Verifiability} Is there any document about the requirements specifications of the program? If yes, what is it. (\{yes$^*$, no\})
-\item \ad{Portability} Do you think that portability has been achieved? If yes, how? (\{yes$^*$, no\})
-\item \ad{Maintainability} How was maintainability considered in the design? (string)
-\item \ad{Maintainability} What is the maintenance type? (set of \{corrective, adaptive, perfective,
-unclear\})
+\item \ad{General} What is the most important software quality(ies) to your
+  work? (set of selected qualities plus "else")
+\item \ad{General} Are there any examples where the documentation helped? If
+  yes, how it helped. ({yes$^*$, no})
+\item \ad{General} Is there any documentation you feel you should produce and do
+  not? If yes, what is it and why? ({yes$^*$, no})
+\item \ad{Completeness} Do you address any of your quality concerns using
+  documentation? If yes, what are the qualities and the documents. ({yes$^*$,
+    no})
+\item \ad{Visibility/Transparency} Is there a certain type of development
+  methodologies used during the development? (\{Waterfall, Scrum, Kanban,
+  else\})
+\item \ad{Visibility/Transparency} Is there a clearly defined development
+  process? If yes, what is it. (\{yes$^*$, no\})
+\item \ad{Visibility/Transparency} Are there any project management tools used
+  during the development? If yes, what are they. (\{yes$^*$, no\})
+\item \ad{Visibility/Transparency} Going forward, will your approach to
+  documentation of requirements and design change? If not, why not. (\{yes,
+  no$^*$\})
+\item \ad{Correctness and Verifiability} During the process of development, what
+  tools or techniques are used to build confidence of correctness? (string)
+\item \ad{Correctness and Verifiability} Do you use any tools to support
+  testing? If yes, what are they. (e.g. unit testing tools, regression testing
+  suites) (\{yes$^*$, no\})
+\item \ad{Correctness and Verifiability} Is there any document about the
+  requirements specifications of the program? If yes, what is it. (\{yes$^*$,
+  no\})
+\item \ad{Portability} Do you think that portability has been achieved? If yes,
+  how? (\{yes$^*$, no\})
+\item \ad{Maintainability} How was maintainability considered in the design?
+  (string)
+\item \ad{Maintainability} What is the maintenance type? (set of \{corrective,
+  adaptive, perfective, unclear\})
 \item \ad{Reusability} How was reusability considered in the design? (string)
-\item \ad{Reusability} Are any portions of the software used by another package? If yes, how they are used. ({yes$^*$, no})
+\item \ad{Reusability} Are any portions of the software used by another package?
+  If yes, how they are used. ({yes$^*$, no})
 \item \ad{Reproducibility} Is reproducibility important to you? ({yes$^*$, no})
-\item \ad{Reproducibility} Do you use tools to help reproduce previous software results? If yes, what are they. (e.g. version control, configuration management) ({yes$^*$, no})
-\item \ad{Completeness} Is any of the following documents used during the development? ({yes$^*$, no})
-\item \ad{General} Will this experience influence how you develop software? Do you see yourself maintaining the same level of documentation, tool support as you go forward? (string)
+\item \ad{Reproducibility} Do you use tools to help reproduce previous software
+  results? If yes, what are they. (e.g. version control, configuration
+  management) ({yes$^*$, no})
+\item \ad{Completeness} Is any of the following documents used during the
+  development? ({yes$^*$, no})
+\item \ad{General} Will this experience influence how you develop software? Do
+  you see yourself maintaining the same level of documentation, tool support as
+  you go forward? (string)
 \begin{itemize}
@@
 \subsection{Empirical Measures Considerations - Processed Data}
+
 Metrics that can be calculated from the raw data.
@@
 \subsection{Empirical Measures Considerations - Tool Tests}
-\ad{This section is currently a note of unorganized contents. Most parts will beremoved or relocated.}
+
+\ad{This section is currently a note of unorganized contents. Most parts will
+  beremoved or relocated.}
 
 \ad{This citation needs to be deleted later. It's here because my compiler
-	doesn't work with 0 citations}
-\cite{Emms2019}
+  doesn't work with 0 citations} \cite{Emms2019}
 
@@
 \subsubsection{git-stats}
+
 \href{https://github.com/tomgi/git_stats}{GitHub repo}
@@
 \subsubsection{scc}
+
 \href{https://github.com/boyter/scc}{GitHub repo}
@@
 \subsubsection{git-of-theseus}
+
 \href{https://github.com/erikbern/git-of-theseus}{GitHub repo}
@@
 \subsubsection{hercules}
+
 \href{https://github.com/src-d/hercules}{GitHub repo}
@@
 \subsubsection{gitinspector}
+
 \href{https://github.com/ejwa/gitinspector}{GitHub repo}

### 2020-11-10 StateOfPractice/Methodology/Methodology.tex
Revisions to Research Questions section

@@
 
-The following are research questions that we want to answer to assess the state
-of practice of research software in our chosen domain:
+The following are the research questions that we wish to answer for each of our
+selected domains.  In addition to answering the questions for each domain, we
+also wish to combine the data from multiple domains to answer these questions
+for research software in general.
 
 \begin{enumerate}
-\item What are the artifacts that current software packages consist of? 
+\item What artifacts are present in current software packages? 
 \item What tools are used by current software packages?
@@
   of current software packages?
-\item What are the pain points for developers working on SCS projects? What
-  aspects of the existing processes, methodologies and tools do they consider
-  could potentially be improved? What processes, methodologies and tools have
-  they changed to reduce obstacles?
-\item For research software developers, what specific actions are taken to address the following:
+\item What are the pain points for developers working on research software
+  projects?  What aspects of the existing processes, methodologies and tools do
+  they consider as potentially needing improvement?  How should processes,
+  methodologies and tools be changed to improve software development and
+  software quality?
+\item For research software developers, what specific actions are taken to
+  address the following:
 \begin{enumerate}

### 2020-11-12 StateOfPractice/Methodology/Methodology.tex
updated wording in section 6 methodology

@@
 the following filters are applied in the priority order listed, stopping once
-the desired number of projects is reached:
+a manageable number of projects is reached:
 

### 2020-11-12 StateOfPractice/Methodology/Methodology.tex
updated introduction section 1 methodology, added performance benchmark note

@@
 
-\pmi{comment from meeting: remove performance benchmarks from scope of exercise
-  because enough other people consider this. Add similar comment into
-  introduction explaining (link to examples of performance considered elsewhere
-  in similar projects??) why performance is not considered} \pmi{do we have
-  specific examples of other works considering performance benchmarks?}
-
 The scope of this document is to conduct a current state of practice analysis of
 software in a scientific computing domain for the purpose of understanding
-quality and assisting in future development. We leave the assessment of
-performance of such software to other similar projects. The methodology
-presented here is part of an overall assessment of the impact of model driven
-engineering and code generation on the sustainability of scientific computing
-software. The research proposal for this assessment, which contains a more
-detailed description of this state of practice exercise, can be found
-\href{https://github.com/smiths/AIMSS/blob/master/OverallResearchProposal/ResearchProposal.pdf}{here}. The
+quality and assisting in future development.
+We observe product, artifact, and process quality in our analysis.
+We leave the assessment of scientific computing software using performance benchmarks to other projects, such as the work of \cite{kaagstrom1998gemm}.
+The methodology presented here is part of an overall assessment of the impact of model driven engineering and code generation on the sustainability of scientific computing software.
+The research proposal for this assessment, which contains a more detailed description of this state of practice exercise, can be found \href{https://github.com/smiths/AIMSS/blob/master/OverallResearchProposal/ResearchProposal.pdf}{here}.
+The
 research questions that this exercise will answer are found in Section
@@
 Note the following formatting conventions of this document. \textcolor{red}{Red}
-text denotes a link to information within this document. \textcolor{cyan}{Cyan}
-text denotes a URL link.
+text denotes a link to sections of this document. \textcolor{cyan}{Cyan}
+text denotes a URL link. \textcolor{blue}{Blue} text denotes a citation.
 

### 2020-11-13 StateOfPractice/Methodology/Methodology.tex
updated measurement template constants question, related methodology section 9 comment, and section2 status
Edits to Overview of Steps in Assessing Quality section, delete section listing qualities

@@
 software in a scientific computing domain for the purpose of understanding
-quality and assisting in future development.
-We observe product, artifact, and process quality in our analysis.
-We leave the assessment of scientific computing software using performance benchmarks to other projects, such as the work of \cite{kaagstrom1998gemm}.
-The methodology presented here is part of an overall assessment of the impact of model driven engineering and code generation on the sustainability of scientific computing software.
-The research proposal for this assessment, which contains a more detailed description of this state of practice exercise, can be found \href{https://github.com/smiths/AIMSS/blob/master/OverallResearchProposal/ResearchProposal.pdf}{here}.
-The
+quality and assisting in future development. We observe product, artifact, and
+process quality in our analysis. We leave the assessment of scientific computing
+software using performance benchmarks to other projects, such as the work of
+\cite{kaagstrom1998gemm}. The methodology presented here is part of an overall
+assessment of the impact of model driven engineering and code generation on the
+sustainability of scientific computing software. The research proposal for this
+assessment, which contains a more detailed description of this state of practice
+exercise, can be found
+\href{https://github.com/smiths/AIMSS/blob/master/OverallResearchProposal/ResearchProposal.pdf}{here}. The
 research questions that this exercise will answer are found in Section
@@
 
-\section{Overview of Steps in Assessing Quality of the Domain Software}\label{StepsAQDS}
+\section{Overview of Steps in Assessing Quality of the Domain Software} \label{StepsAQDS}
+
+To answer the above research questions (Section~\ref{ResearchQuestions}), we
+systematically measure the quality of the software through data collection and a
+series of experiments.  An overview of the measurement process is given in the
+following steps, starting from determining a domain that is suitable for
+measurement:
 
 \begin{enumerate}
-\item Start with the state of practice research
-  questions. (Section~\ref{ResearchQuestions}) \pmi{In review}
 \item Identify the domain. (Section~\ref{SecIdentifyDomain})
@@
 \item Measure using measurement template. (Section~\ref{SecShallowMeasure})
-  \pmi{To be reviewed}
 \item Use AHP process to rank the software packages. (Section~\ref{SecAHP})
@@
 
-\ad{Might need improvements: ``Are constants (other than 0 and 1) hard coded
-  into the program? (\{yes, no* , unclear\})''. Why we need extra comments if
-  the answer is ``no''?}\pmi{good question, to be brought up at meeting}
-
 For completing the template, please follow these steps:
@@
 specific software packages.
-
-\section{Quality Specific Measures}
-
-\subsection{\notdone{Installability} \oo{owner}}
-
-\subsection{\notdone{Correctness} \oo{owner}}
-
-\subsection{\notdone{Verifiability/Testability} \oo{owner}}
-
-\subsection{\notdone{Validatability} \oo{owner}}
-
-\subsection{\notdone{Reliability} \oo{owner}}
-
-\subsection{\notdone{Robustness} \pmi{owner}}
-
-\subsection{\notdone{Performance} \pmi{owner}}
-
-\subsection{\notdone{Usability} \jc{owner}}
-
-\subsection{\notdone{Maintainability} \pmi{owner}}
-
-\subsection{\notdone{Reusability} \pmi{owner}}
-
-\subsection{\notdone{Portability} \pmi{owner}}
-
-\subsection{\notdone{Understandability} \jc{owner}}
-
-\subsection{\notdone{Interoperability} \ad{owner}}
-
-\subsection{\notdone{Visibility/Transparency} \ad{owner}}
-
-\subsection{\notdone{Reproducibility} \wss{owner}}
-
-\subsection{\notdone{Productivity} \ad{owner}}
-
-\subsection{\notdone{Sustainability} \wss{owner}}
-
-\subsection{\notdone{Completeness} \ad{owner}}
-
-\subsection{\notdone{Consistency} \ad{owner}}
-
-\subsection{\notdone{Modifiability} \jc{owner}}
-
-\subsection{\notdone{Traceability} \jc{owner}}
-
-\subsection{\notdone{Unambiguity} \wss{owner}}
-
-\subsection{\notdone{Verifiability} \wss{owner}}
-
-\subsection{\notdone{Abstract} \wss{owner}}
 

### 2020-11-13 StateOfPractice/Methodology/Questions to Developers.tex
added meeting notes into Experiments regarding procedure for modifiablity; edited comment in Questions to Developers
moved traceability from methodology to questions to developers

--- 2020-07-22 StateOfPractice/Ao-Notes/Questions to Developers.tex
@@
 \maketitle
-
+\pmi{A pilot will be conducted with these questions to test them. Test subjects may include PhD students.}
 \section{Information about the developers and users}\citep{Jegatheesan2016}
@@
 
+\pmi{consider adding `What, if any, actions were taken to address x?' to most/all questions, similar to question 10}
+
+Square brackets are used for traceability to qualities and relevant research questions found in the  \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Methodology.pdf}{methodology document}.
+
 \begin{enumerate}
@@
 \item How does documentation fit into your development process? Would improved
-documentation help with the obstacles you typically face?
+documentation help with the obstacles you typically face? [traceability - research question 5b, visibility/transparency - research question 5i]
 \item In the past, is there any major obstacle to your development process that
@@
 \item Was it hard to ensure the correctness of the software? If there were any
- obstacles, what methods have been considered or practiced to improve the situation?
-  If practiced, did it work?
+ obstacles, what methods have been considered or practiced to improve the situation? If practiced, did it work? [correctness - research question 5e]
 \item When designing the software, did you consider the ease of future changes?
- For example, will it be hard to change the structure of the system, modules or code
-  blocks? What measures have been taken to ensure the ease of future changes
-   and maintains?
-\item Do you think the users can understand your software well? Has there been
-any problem to use this software?
+ For example, will it be hard to change the structure of the system, modules or code blocks?
+What measures have been taken to ensure the ease of future changes and maintains?
+[maintainability - research question 5d, modifiability - research question 5c]
+\item Provide instances where users have misunderstood the software. What, if any, actions were taken to address understandability issues? [understandability - research question 5f]
+\item What, if any, actions were taken to address usability issues? [usability - research question 5a]
 \item Do you think the current documentation can clearly convey all necessary
 knowledge to the users? If yes, how did you successfully achieve it? If no, what
-improvements are needed?
+improvements are needed? [unambiguity - research question 5g]
 \item Do you have any concern that your computational results won't be
-reproducible in the future? Have you taken any steps to ensure reproducibility?
+reproducible in the future? Have you taken any steps to ensure reproducibility? [reproducibility - research question 5h]
 \end{enumerate}

### 2020-11-13 StateOfPractice/Methodology/Experiments.tex
added meeting notes into Experiments regarding procedure for modifiablity; edited comment in Questions to Developers
updated preamble and procedure of modifiability experiment

@@
 ~\newpage
-\section{Modifiability Experiments}
-\pmi{add preamble here describing some of the notes below}
+\section{Modifiability Experiment}
 
-What can be done to measure modifiability?
-
-Idea: measure sensible changes
-
-Can we measure sensible changes? How?
-
-We could gather qualitative data about the process of attempting to make sensible changes.
-Perceived difficulty, available documentation, prerequisite knowledge, time/effort required, tools required, errors and issues encountered and ability to overcome those issues with the available resources, is there dev support?
-How to interpret results?
-
-What does a small change mean? What are likely changes - See Parnas
+This experiment is designed to gather qualitative data regarding the modifiability of each software package in a domain.
+The steps outlined below will result in the creation, by domain experts, of a list of likely changes for software in the domain.
+This list will then be used to analyze which software packages have been designed to accommodate these expected modifications.
 
@@
 
-We want to be explicit about what we want to modify.
+\subsection {Procedure}
 
-\subsection {Procedure}
 \begin{enumerate}
-	\item Identify likely changes using procedure x
-	\item Select from likely change list (using what criteria)
-	\item Gather relevant documentation?
-	\item Ask the domain expert to make the (selected) likely change(s)
-	\item Have the domain expert answer a questionnaire about their attempt to make the likely change
+	\item \textit{Domain Expert}: List all likely changes that a domain expert might make in a software package in the domain.
+	\item Using the short list of software packages listed in the  \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Methodology.pdf}{methodology document}:
+	\begin{enumerate}
+		\item Identify which of the above changes each package is likely designed to accommodate.
+		\item Prepare a short report outlining the likely changes for software in the domain, and which changes each software package is likely designed to accommodate.
+	\end{enumerate}
 \end{enumerate}

### 2020-11-14 StateOfPractice/AHP2020/LBM/Source code/DataProcess/src/Main_ShortList.java
added AHP for short list

@@ -1,0 +1,146 @@
+import java.io.File; 
+import java.io.FileInputStream;
+import java.io.FileOutputStream;
+import java.util.Iterator;
+
+import javax.swing.JOptionPane;
+
+import org.apache.poi.ss.usermodel.Cell;
+import org.apache.poi.ss.usermodel.Row;
+import org.apache.poi.xssf.usermodel.XSSFSheet;
+import org.apache.poi.xssf.usermodel.XSSFWorkbook;
+
+public class Main_ShortList {
+	
+	public static void main(String[] args)
+    {
+		int numberQualities = 2;
+		int numberSoftwarePackages = 7; //also need to change rownum below
+		Cell[][] table=new Cell[numberQualities][numberSoftwarePackages];
+		Double[][][] ahpTable= new Double[numberQualities][numberSoftwarePackages][numberSoftwarePackages-1];
+        try
+        {
+            FileInputStream file = new FileInputStream(new File("AHP_ShortList.xlsx"));
+ 
+            //Create Workbook instance holding reference to .xlsx file
+            XSSFWorkbook workbook = new XSSFWorkbook(file);
+ 
+            //Get second sheet from the workbook
+            XSSFSheet sheet = workbook.getSheet("AssignedScores");
+ 
+            //Iterate through each rows one by one
+            Iterator<Row> rowIterator = sheet.iterator();            
+            int r=0;
+            int col=0;
+            while (rowIterator.hasNext())
+            {
+                Row row = rowIterator.next();
+                //For each row, iterate through all the columns
+                Iterator<Cell> cellIterator = row.cellIterator();
+                 
+                while (cellIterator.hasNext())
+                {
+                    Cell cell = cellIterator.next();
+                    //Check the cell type and format accordingly
+                    switch (cell.getCellType())
+                    {
+                        case Cell.CELL_TYPE_NUMERIC:
+                        	table[r][col]=cell;    
+                        	col++;
+                            break;
+                        case Cell.CELL_TYPE_STRING:
+                            break;
+                        default:
+                           	break;
+                    }                    
+                }
+                r++;
+                col=0;
+            }
+            XSSFSheet sheetnew;
+            if(workbook.getSheet("AHP Example")==null){
+            	sheetnew= workbook.createSheet("AHPDATA");
+            }
+            else{
+            	sheetnew=workbook.getSheet("AHP Example");
+            }
+            double x=0;
+            double y=0;
[…]

### 2020-11-20 StateOfPractice/Methodology/Methodology.tex
moved traceability from methodology to questions to developers
Revisions to candidate software indentification section

@@
 \begin{enumerate}
-\item usability \pmi{question to developers 11}
-\item traceability \pmi{question to developers 3}
-\item modifiability \pmi{question to developers 9}
-\item maintainability \pmi{question to developers 9}
-\item correctness \pmi{question to developers 8}
-\item understandability \pmi{question to developers 10}
-\item unambiguity \pmi{question to developers 12}
-\item reproducibility \pmi{question to developers 13}
-\item visibility/transparency \pmi{question to developers 3}
+\item usability
+\item traceability
+\item modifiability
+\item maintainability
+\item correctness
+\item understandability
+\item unambiguity
+\item reproducibility
+\item visibility/transparency
 \end{enumerate} 
@@
 in this exercise as ``software that is used to generate, process or analyze
-results that [are intended] to appear in a publication''
-\citep{hettrick2014uk}. The chosen domain must have the following properties:
+results that [are intended] to appear in a publication'' \citep{hettrick2014uk}
+in a scientific or engineering context.  Research software is a more general
+term that covers what is often called scientific computing.  To be applicable
+for the methodology described in this document, the chosen domain must have the
+following properties:
 
 \begin{enumerate}
-\item The domain must have a well-defined, stable theoretical
-  underpinning. Standard textbooks, preferably understandable by an upper
-  undergraduate student, is a good sign of this.
+\item The domain must have well-defined and stable theoretical underpinning.  A
+  good sign of this is the existance of standard textbooks, preferably
+  understandable by an upper year undergraduate student.
 \item There must be a community of people studying the domain.
@@
 \item A preliminary search, or discussion with experts, suggests that there will
-  be numerous, at least about 15, candidate software packages in the domain that
+  be numerous, at least 15, candidate software packages in the domain that
   qualify as `research software'.
@@
 
-Some examples of domains that fit these criteria are finite element analysis,
-quantum chemistry, seismology, as well as mesh generators and processors. These
-are described in \citep{szabo1996finite}, \citep{veryazov20042molcas},
-\citep{SmithEtAl2018} and \citep{smith2016state}, respectively.
+Some examples of domains that fit these criteria are finite element analysis
+\citep{szabo1996finite}, quantum chemistry \citep{veryazov20042molcas},
+seismology \citep{SmithEtAl2018}, as well as mesh generators
+\citep{smith2016state}.
 
@@
 
-The candidate software can be found through search engine queries targeting
-authoritative lists of software from the domain, such as lists on GitHub and
-swMATH, and by searching domain related publications. Domain experts are also
-asked for their suggestions and are asked to review the list. The candidate
-software should have the following properties:
+Once the domain of interest is identified, the candidate software for measuring
+can be found through search engine queries targeting authoritative lists of
+software.  Potential places to search include \href{https://github.com/}
+{GitHub}, \href{https://swmath.org/} {swMATH} and domain related publications,
+such as review articles. Domain experts are also asked for their suggestions and
+are asked to review the initial draft of the software list.
+
+When forming the list and reviewing the candidate software the following
+properties should be considered:
 
 \begin{enumerate}
-\item Major function(s) must fall within the identified domain.
-\item Must have viewable source code.
-\item Ideally have a git repository or ability to gather empirical measures
-  found in Section \ref{SecEmpiricalMeasures}.
+\item The software functionality must fall within the identified domain.
+\item The source code must be viewable.
+\item The empirical measures listed in Section \ref{SecEmpiricalMeasures} should
+  ideally be available, which implies a preference for GitHub repositories.
 \item The software cannot be marked as incomplete or in an initial development

### 2020-11-24 StateOfPractice/Methodology/Methodology.tex
Updates to "How to Initially Filter the Software List" section

@@
 to be taken to create a more manageable list. To reduce the length of the list,
-the following filters are applied in the priority order listed, stopping once
-a manageable number of projects is reached:
+the following filters are applied.  The filters are applied in the priority
+order listed, with the filtering process stopped once the list size is
+manageable.
 
 \begin{enumerate}
-\item Scope: The functionality of the domain software should be increasingly
-  specific.
-\item Usage: The installation procedure should appear to be clear or easy to
-  figure out.
-\item Age: Ideally the latest release or source code commit should be relatively
-  recent for software in the domain unless the candidate software appears to be
-  well recommended and currently in use.
+\item Scope: Software is removed by narrowing what functionality is considered
+  to be within the scope of the domain.
+\item Usage: Software packages are eliminated if their installation procedure is
+  not clear and easy to follow.
+\item Age: The older software packages (age being measured by the last date when
+  a change was made) are eliminated, except in the cases where an older software
+  package appears to be highly recommended and currently in use.
 \end{enumerate}

### 2020-11-26 StateOfPractice/Methodology/Meeting Agenda with Domain Experts.tex
added notes into Meeting Agenda document to address issues 54 and 56
Complete editing pass through methodology document

--- 2020-07-01 StateOfPractice/Ao-Notes/Meeting Agenda with Domain Experts.tex
@@
 
-What do you think of our analysis?
+What do you think of our analysis?  \wss{This is too big a question for the
+  domain experts to answer in a reasonable amount of time.  Moreover, without
+  more guidance their feedback is likely to be shallow.  I think we are better
+  off with a task based inspection, using issues assigned via a GitHub project
+  page.  We also will likely be more successful asking for reviews from grad
+  students, rather than from faculty members.  We should also consider
+  summarizing the commonalities, variabilities and parameters of variation in
+  tables, or possibly graphs.}
 
@@
 
+\subsection{Review the Domain Analysis}
+\pmi{added this task as requested in AIMSS issue  \href{https://github.com/smiths/AIMSS/issues/56}{56}}
+
+\subsection{Review the AHP Ranking and Short List of Software Packages}
+\pmi{added this task as requested in AIMSS issue  \href{https://github.com/smiths/AIMSS/issues/54}{54}}
+
 \subsection{Review the Measurement Template}

### 2020-11-26 StateOfPractice/Methodology/Methodology.tex
Complete editing pass through methodology document

@@
 
-\usepackage{natbib}
+\usepackage[round]{natbib}
 \usepackage{booktabs}
@@
 
-\section{Research Questions}\label{ResearchQuestions}
+\section{Research Questions} \label{ResearchQuestions}
 
@@
 
-The domain analysis consists of a commonality analysis of the family of software
-packages. Its purpose is to show the relationships between these packages, and
-to facilitate an understanding of the informal specification and development of
-them. \cite{weiss1997defining} defines commonality analysis as an approach to
-defining a family by identifying commonalities, variabilities, and common
-terminology for the family. Commonalities are goals, theories, models,
-definitions and assumptions that are common between family members. Likewise,
-variabilities are goals, theories, models, definitions and assumptions that are
-different between family members. The final result of the domain analysis will
-be tables of commonalities, variabilities, and parameters of variation of a
-program family, which \cite{parnas1976design} defines as ``a set of programs
-whose common properties are so extensive that it is advantageous to study the
-common properties of the programs before analyzing individual
-members''. \cite{smith2008commonality} present a template for conducting a
+Since each domain we will study will have a reasonably small scope, we
+will be able to view the software as constituting a program
+family.  The concept of a program family is defined by \citet{parnas1976design}
+as ``a set of programs whose common properties are so extensive that it is
+advantageous to study the common properties of the programs before analyzing
+individual members''.  Studying the common properties within a family of related
+programs is termed a domain analysis.
+
+The domain analysis consists of a commonality analysis of
+the family of software packages. Its purpose is to show the relationships
+between these packages, and to facilitate an understanding of the informal
+specification and development of them. \cite{weiss1997defining} defines
+commonality analysis as an approach to defining a family by identifying
+commonalities, variabilities, and common terminology for the
+family. Commonalities are goals, theories, models, definitions and assumptions
+that are common between family members.  Variabilities are goals, theories,
+models, definitions and assumptions that differ between family
+members. Associated with each variability are its parameters of variation, which
+summarize the possible values for that variability, along with their potential
+binding time.  The binding time is when the value of the variability is set.  It
+could be set as specification time, build time (when the program is being
+compiled) or run time (when the code is executing).
+
+The final result of the domain analysis will be
+tables of commonalities, variabilities, and parameters of variation of a program
+family.   \cite{smith2008commonality} present a template for conducting a
 commonality analysis, which was referred to when conducting this
@@
 A sample commonality analysis for Lattice Boltzmann Solvers can be found
-\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Peter-Notes/Commonality-Analysis-LB-Systems.pdf}{here}.
+\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Peter-Notes/Commonality-Analysis-LB-Systems.pdf}
+{here}.
 
@@
 analyze for this assessment. The measures that are collected relate to the
-research questions that we want to answer. For instance, we collect some of the
+research questions (Section~\ref{ResearchQuestions}). For instance, we collect some of the
 data to see how large a project is. Other measures intend to ascertain a
-project’s popularity, or how active the project is.  Section \ref{rawdata} will
-orient the reader as to what raw data is collected.  Some of this data can be
+project’s popularity, or how active the project is.
+
+Section \ref{rawdata} orients the reader on what raw data is collected.  Some of this data can be
 observed from Git repository metrics. The rest can be collected using freeware
@@
 the number of binary files as well as the number of added and deleted lines in a
-repository. The tool is also used to measure the number of commits over
+repository. This tool is also used to measure the number of commits over
 different intervals of time. \href{https://github.com/boyter/scc}{Sloc Cloc and
@@
 were selected due to their installability, usability, and ability to gather the
-empirical measures listed below. A guide for installing and running them can be
-found
+empirical measures listed below. Details on installing and running the tools can
+be found in our 
 \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/A
-  Guide to Empirical Measures.pdf} {here}. Section \ref{processeddata}
+  Guide to Empirical Measures.pdf} {Guide to Empirical Measures}. Section \ref{processeddata}
 introduces the required processed data, which is calculated using the raw data.
@@
 \item Number of total commits.
-\item Numbers of commits by year in the last 5 years. (Count from as early as possible if the project is younger than 5 years.) 
+\item Numbers of commits by year in the last 5 years. (Count from as early as
+  possible if the project is younger than 5 years.)
 \item Numbers of commits by month in the last 12 months.
@@
 {Measurement Template} is used to track measurements and quality scores for all
-of the software packages in the domain. For each software package, fill out one
+of the software packages in the domain. For each software package, we fill out one
 column of the template. This process can take between 2 to 5 hours for each
@@
 if necessary, but a cap of about 2 hours should be imposed on the entire
-installation process.
+installation process, to keep the overall measurement time feasible.
 
-Some guidance on applying sections of the template:
+Most of the data to be collected should be straightforward from reviewing the
+measurement template.  However, in some cases some extra guidance is necessary
+to eliminate ambiguity, as follows:
 
@@
 
-For completing the template, please follow these steps:
+For completing the template, the following steps should be followed:
 
@@
 \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Experiments.pdf}
-{here}. This step is intended to start a conversation about the features of
-specific software packages.
+{here}. This step is intended to start a conversation about the relative quality of
+short-listed software packages.
 
@@
 Describe AHP process (or similar).
+
+\newpage
 

### 2020-11-29 StateOfPractice/Methodology/Experiments.tex
updated modifiability experiment introduction

@@
 
-This experiment is designed to gather qualitative data regarding the modifiability of each software package in a domain.
-The steps outlined below will result in the creation, by domain experts, of a list of likely changes for software in the domain.
-This list will then be used to analyze which software packages have been designed to accommodate these expected modifications.
+This experiment is designed to gather qualitative data regarding the modifiability of each software package.
+The initial step outlined below produces a list of likely software changes for members of a scientific computing software domain.
+This list is then used to analyze whether specific software packages have been designed to accommodate these expected modifications.
 
-Parnas identifies likely changes as:
-
- Likely Changes: if the system is required to be easy to change, the requirements should  contain  a  definition  of  the  areas  that  are  considered  likely  to  change.
+Parnas's definition of likely changes has informed this analysis: ``If the system is required to be easy to change, the requirements should contain a definition of the areas that are considered likely to change.
 You cannot design a system so that everything is equally easy to change.
-Programmers should not have to decide which changes are most likely\cite{parnas1986rational}
+Programmers should not have to decide which changes are most likely'' \cite{parnas1986rational}.
 
@@
 	\item \textit{Domain Expert}: List all likely changes that a domain expert might make in a software package in the domain.
-	\item Using the short list of software packages listed in the  \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Methodology.pdf}{methodology document}:
+	\item Using the short list of software packages listed in the  \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Methodology.pdf}{Methodology document}:
 	\begin{enumerate}

### 2020-11-29 StateOfPractice/Methodology/Experiments.tex
updated modifiability experiment introduction, added Clements

@@
 
-Parnas's definition of likely changes has informed this analysis: ``If the system is required to be easy to change, the requirements should contain a definition of the areas that are considered likely to change.
+Parnas and Clements' definition of likely changes has informed this analysis: ``If the system is required to be easy to change, the requirements should contain a definition of the areas that are considered likely to change.
 You cannot design a system so that everything is equally easy to change.

### 2020-11-29 StateOfPractice/Methodology/Methodology.tex
cleaned up Methodology appendix, moved one part to Questions to Devs

@@
 
-\subsection{Survey for the Selected Projects}
-
-\ad{Several questions are borrowed from
-  \href{https://gitlab.cas.mcmaster.ca/smiths/pub/-/blob/master/Jegatheesan2016.pdf}{Jegatheesan2016},
-  and needed to be cited later.}
-
-\subsubsection{Information about the developers and users}
-
-\begin{enumerate}
-\item Interviewees' current position/title? degrees?
-\item Interviewees' contribution to/relationship with the software?
-\item Length of time the interviewee has been involved with this software?
-\item How large is the development group?
-\item What is the typical background of a developer?
-\item How large is the user group?
-\item What is the typical background of a user?
-\end{enumerate}
-
-\subsubsection{Information about the software}
-
-\begin{enumerate}
-\item \ad{General} What is the most important software quality(ies) to your
-  work? (set of selected qualities plus "else")
-\item \ad{General} Are there any examples where the documentation helped? If
-  yes, how it helped. ({yes$^*$, no})
-\item \ad{General} Is there any documentation you feel you should produce and do
-  not? If yes, what is it and why? ({yes$^*$, no})
-\item \ad{Completeness} Do you address any of your quality concerns using
-  documentation? If yes, what are the qualities and the documents. ({yes$^*$,
-    no})
-\item \ad{Visibility/Transparency} Is there a certain type of development
-  methodologies used during the development? (\{Waterfall, Scrum, Kanban,
-  else\})
-\item \ad{Visibility/Transparency} Is there a clearly defined development
-  process? If yes, what is it. (\{yes$^*$, no\})
-\item \ad{Visibility/Transparency} Are there any project management tools used
-  during the development? If yes, what are they. (\{yes$^*$, no\})
-\item \ad{Visibility/Transparency} Going forward, will your approach to
-  documentation of requirements and design change? If not, why not. (\{yes,
-  no$^*$\})
-\item \ad{Correctness and Verifiability} During the process of development, what
-  tools or techniques are used to build confidence of correctness? (string)
-\item \ad{Correctness and Verifiability} Do you use any tools to support
-  testing? If yes, what are they. (e.g. unit testing tools, regression testing
-  suites) (\{yes$^*$, no\})
-\item \ad{Correctness and Verifiability} Is there any document about the
-  requirements specifications of the program? If yes, what is it. (\{yes$^*$,
-  no\})
-\item \ad{Portability} Do you think that portability has been achieved? If yes,
-  how? (\{yes$^*$, no\})
-\item \ad{Maintainability} How was maintainability considered in the design?
-  (string)
-\item \ad{Maintainability} What is the maintenance type? (set of \{corrective,
-  adaptive, perfective, unclear\})
-\item \ad{Reusability} How was reusability considered in the design? (string)
-\item \ad{Reusability} Are any portions of the software used by another package?
-  If yes, how they are used. ({yes$^*$, no})
-\item \ad{Reproducibility} Is reproducibility important to you? ({yes$^*$, no})
-\item \ad{Reproducibility} Do you use tools to help reproduce previous software
-  results? If yes, what are they. (e.g. version control, configuration
-  management) ({yes$^*$, no})
-\item \ad{Completeness} Is any of the following documents used during the
-  development? ({yes$^*$, no})
-\item \ad{General} Will this experience influence how you develop software? Do
-  you see yourself maintaining the same level of documentation, tool support as
-  you go forward? (string)
-\begin{itemize}
-\item Module Guide
-\item Module Interface Specification
-\item Verification and Validation Plan
-\item Verification and Validation Report
-\end{itemize}
-\end{enumerate}
-
-\subsection{Empirical Measures Considerations - Raw Data}
-
-Measures that can be extracted from on-line repos.
-
-\ad{Still at brainstorm stage.}
-\begin{itemize}
-\item number of contributors
-\item number of watches
-\item number of stars
-\item number of forks
-\item number of clones
-\item number of commits
-\item number of total/code/document files
-\item lines of total/logical/comment code
-\item lines/pages of documents (can pdf be extracted?)
-\item number of total/open/closed/merged pull requests
-\item number of total/open/closed issues
-\item number of total/open/closed issues with assignees
-\end{itemize}
-
-Instead of only focus on the current status of the above numbers, we may find
-the time history of them to be more valuable. For example, the number of
-contributors over time, the number of lines of code over time, the number of
-open issues over time, etc.
-
-\subsection{Empirical Measures Considerations - Processed Data}
-
-Metrics that can be calculated from the raw data.
-
-\ad{Still at brainstorm stage.}
-\begin{itemize}
-\item percentage of total/open/closed issues with assignees -
-  Visibility/Transparency
-\item lines of new code produced per person-month - Productivity
-\item lines/pages of new documents produced per person-month - Productivity
-\item number of issues closed per person-month - Productivity
-\item percentage of comment lines in the code - maintainability \ad{Not Ao's
-    qualities}
-\end{itemize}
-
-In the above calculations, a month can be determined to be 30 days.
-
 \subsection{Empirical Measures Considerations - Tool Tests}

### 2020-11-29 StateOfPractice/Methodology/Questions to Developers.tex
cleaned up Methodology appendix, moved one part to Questions to Devs

@@
 
+\subsection{Old notes from methodology: Information about the software}
+Compare to questions above and then delete this subsection
+
+\begin{enumerate}
+	\item \ad{General} What is the most important software quality(ies) to your
+	work? (set of selected qualities plus "else")
+	\item \ad{General} Are there any examples where the documentation helped? If
+	yes, how it helped. ({yes$^*$, no})
+	\item \ad{General} Is there any documentation you feel you should produce and do
+	not? If yes, what is it and why? ({yes$^*$, no})
+	\item \ad{Completeness} Do you address any of your quality concerns using
+	documentation? If yes, what are the qualities and the documents. ({yes$^*$,
+		no})
+	\item \ad{Visibility/Transparency} Is there a certain type of development
+	methodologies used during the development? (\{Waterfall, Scrum, Kanban,
+	else\})
+	\item \ad{Visibility/Transparency} Is there a clearly defined development
+	process? If yes, what is it. (\{yes$^*$, no\})
+	\item \ad{Visibility/Transparency} Are there any project management tools used
+	during the development? If yes, what are they. (\{yes$^*$, no\})
+	\item \ad{Visibility/Transparency} Going forward, will your approach to
+	documentation of requirements and design change? If not, why not. (\{yes,
+	no$^*$\})
+	\item \ad{Correctness and Verifiability} During the process of development, what
+	tools or techniques are used to build confidence of correctness? (string)
+	\item \ad{Correctness and Verifiability} Do you use any tools to support
+	testing? If yes, what are they. (e.g. unit testing tools, regression testing
+	suites) (\{yes$^*$, no\})
+	\item \ad{Correctness and Verifiability} Is there any document about the
+	requirements specifications of the program? If yes, what is it. (\{yes$^*$,
+	no\})
+	\item \ad{Portability} Do you think that portability has been achieved? If yes,
+	how? (\{yes$^*$, no\})
+	\item \ad{Maintainability} How was maintainability considered in the design?
+	(string)
+	\item \ad{Maintainability} What is the maintenance type? (set of \{corrective,
+	adaptive, perfective, unclear\})
+	\item \ad{Reusability} How was reusability considered in the design? (string)
+	\item \ad{Reusability} Are any portions of the software used by another package?
+	If yes, how they are used. ({yes$^*$, no})
+	\item \ad{Reproducibility} Is reproducibility important to you? ({yes$^*$, no})
+	\item \ad{Reproducibility} Do you use tools to help reproduce previous software
+	results? If yes, what are they. (e.g. version control, configuration
+	management) ({yes$^*$, no})
+	\item \ad{Completeness} Is any of the following documents used during the
+	development? ({yes$^*$, no})
+	\item \ad{General} Will this experience influence how you develop software? Do
+	you see yourself maintaining the same level of documentation, tool support as
+	you go forward? (string)
+	\begin{itemize}
+		\item Module Guide
+		\item Module Interface Specification
+		\item Verification and Validation Plan
+		\item Verification and Validation Report
+	\end{itemize}
+\end{enumerate}
+
 \newpage

### 2020-12-01 StateOfPractice/Methodology/Methodology-Traceability.tex
moved the list of considered empirical measurement tools from the appendix into a new traceability document

--- 2020-11-29 StateOfPractice/Methodology/Methodology.tex
@@
 
-\title{Methodology for Assessing the State of the Practice for Domain X} 
+\title{Methodology Document Traceability} 
 \author{Spencer Smith}{McMaster University, Canada}{smiths@mcmaster.ca}{}{}
@@
 \maketitle
-
-\begin{abstract}
-	...
-\end{abstract}
-
-\tableofcontents
-
-\section{Introduction} \label{SecIntroduction}
-
-The scope of this document is to conduct a current state of practice analysis of
-software in a scientific computing domain for the purpose of understanding
-quality and assisting in future development. We observe product, artifact, and
-process quality in our analysis. We leave the assessment of scientific computing
-software using performance benchmarks to other projects, such as the work of
-\cite{kaagstrom1998gemm}. The methodology presented here is part of an overall
-assessment of the impact of model driven engineering and code generation on the
-sustainability of scientific computing software. The research proposal for this
-assessment, which contains a more detailed description of this state of practice
-exercise, can be found
-\href{https://github.com/smiths/AIMSS/blob/master/OverallResearchProposal/ResearchProposal.pdf}{here}. The
-research questions that this exercise will answer are found in Section
-\ref{ResearchQuestions}. Section \ref{StepsAQDS} outlines the steps to be taken.
-
-Note the following formatting conventions of this document. \textcolor{red}{Red}
-text denotes a link to sections of this document. \textcolor{cyan}{Cyan}
-text denotes a URL link. \textcolor{blue}{Blue} text denotes a citation.
-
-\section{Research Questions} \label{ResearchQuestions}
-
-The following are the research questions that we wish to answer for each of our
-selected domains.  In addition to answering the questions for each domain, we
-also wish to combine the data from multiple domains to answer these questions
-for research software in general.
-
-\begin{enumerate}
-\item What artifacts are present in current software packages? 
-\item What tools are used by current software packages?
-\item What principles, processes, and methodologies are used in the development
-  of current software packages?
-\item What are the pain points for developers working on research software
-  projects?  What aspects of the existing processes, methodologies and tools do
-  they consider as potentially needing improvement?  How should processes,
-  methodologies and tools be changed to improve software development and
-  software quality?
-\item For research software developers, what specific actions are taken to
-  address the following:
-\begin{enumerate}
-\item usability
-\item traceability
-\item modifiability
-\item maintainability
-\item correctness
-\item understandability
-\item unambiguity
-\item reproducibility
-\item visibility/transparency
-\end{enumerate} 
-\item How does software designated as high quality by this methodology compare
-  with top rated software by the community?
-\end{enumerate}
-
-%In general questions:
-%
-%\begin{enumerate}
-%\item Comparison between domains
-%\item How to measure qualities
-%\item How does the quality compare for projects with the most resources to %those
-%  with the fewest?
-%\item What skills/knowledge are needed by future developers?
-%\item How can the development process be improved?
-%\item What are the common pain points?
-%\end{enumerate}
-%
-%For each domain questions:
-%
-%\begin{enumerate}
-%\item Best examples within the domain
-%\item What software artifacts?
-%\item What are the pain points?
-%\item Any advice on what can be done about the pain points?
-%\end{enumerate}
-%
-%Measure the effort invested and the reward.  Related to sustainability.
-%
-%Collect the data and see what conclusions follow.  For an individual domain,
-%between domains.  The process isn't so much about ranking the software as it is
-%about looking at the software closely and see what conclusions arise.  The
-%measurements are intended to force scrutiny, from different perspectives.
-
-\section{Overview of Steps in Assessing Quality of the Domain Software} \label{StepsAQDS}
-
-To answer the above research questions (Section~\ref{ResearchQuestions}), we
-systematically measure the quality of the software through data collection and a
-series of experiments.  An overview of the measurement process is given in the
-following steps, starting from determining a domain that is suitable for
-measurement:
-
-\begin{enumerate}
-\item Identify the domain. (Section~\ref{SecIdentifyDomain})
-\item \emph{Domain Experts}: Create a top ten list of software packages in the
-  domain. (\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Meeting
-    Agenda withmDomainmExperts.pdf}{Meeting Agenda with Domain Experts})
-\item Brief the Domain Experts on the overall objective, research proposal,
-  research questions, measurement template, survey for short list projects,
-  usability tests, maintainability
-  experiments. (\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Meeting
-    Agenda withmDomainmExperts.pdf}{Meeting Agenda with Domain Experts})
-\item Identify broad list of candidate software packages in the
-  domain. (Section~\ref{SecIdentifyCandSoft})
-\item Preliminary filter of software packages
-  list. (Section~\ref{SecInitialFilter})
-\item \emph{Domain Experts}: Review domain software
-  list. (\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Meeting
-    Agenda withmDomainmExperts.pdf}{Meeting Agenda with Domain Experts})
-\item Domain Analysis. (Section~\ref{SecDomainAnalysis})
-\item \emph{Domain Experts}: Vet domain
-  analysis. (\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Meeting
-    Agenda withmDomainmExperts.pdf}{Meeting Agenda with Domain Experts})
-\item Gather source code and documentation for each prospective software
-  package.
-\item Collect empirical measures. (Section~\ref{SecEmpiricalMeasures})
-\item Measure using measurement template. (Section~\ref{SecShallowMeasure})
-\item Use AHP process to rank the software packages. (Section~\ref{SecAHP})
-\item Identify a short list of top software packages, typically four to six, for
-  deeper exploration according to the AHP rankings of the measurements.
-\item \emph{Domain Experts}: Vet AHP ranking and short
-  list. (\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Meeting
-    Agenda withmDomainmExperts.pdf}{Meeting Agenda with Domain Experts})
-\item With short list:
-\begin{enumerate}
-\item Survey developers
-  (\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Questions
-    to Developers.pdf}{Questions to Developers})
-\item Usability experiments
-  (\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Experiments.pdf}{Experiments})
-\item Modifiability experiments
-  (\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Experiments.pdf}{Experiments})
-\end{enumerate}
-\item Rank short list. (Section~\ref{SecRankShortList}) \pmi{To be completed -
-    add pairwise comparison tool}
-\item Document answers for research questions.
-\end{enumerate}
-
-%\wss{The domain expert is involved in multiple steps in the process.  How best
-%  to get their feedback?  The domain experts are busy and are unlikely to devote
-%  significant time to the project.  We need to quickly get to the point.  Maybe
-%  something around task based inspection?  Directed interview?}
-
-\section{How to Identify the Domain} \label{SecIdentifyDomain}
-
-A domain of research software must be identified. Research software is defined
-in this exercise as ``software that is used to generate, process or analyze
-results that [are intended] to appear in a publication'' \citep{hettrick2014uk}
-in a scientific or engineering context.  Research software is a more general
-term that covers what is often called scientific computing.  To be applicable
-for the methodology described in this document, the chosen domain must have the
-following properties:
-
-\begin{enumerate}
-\item The domain must have well-defined and stable theoretical underpinning.  A
-  good sign of this is the existance of standard textbooks, preferably
-  understandable by an upper year undergraduate student.
-\item There must be a community of people studying the domain.
-\item The software packages must have open source options.
-\item A preliminary search, or discussion with experts, suggests that there will
-  be numerous, at least 15, candidate software packages in the domain that
-  qualify as `research software'.
-\end{enumerate}
-
-Some examples of domains that fit these criteria are finite element analysis
-\citep{szabo1996finite}, quantum chemistry \citep{veryazov20042molcas},
-seismology \citep{SmithEtAl2018}, as well as mesh generators
-\citep{smith2016state}.
-
-\section{How to Identify Candidate Software} \label{SecIdentifyCandSoft}
-
-Once the domain of interest is identified, the candidate software for measuring
-can be found through search engine queries targeting authoritative lists of
-software.  Potential places to search include \href{https://github.com/}
-{GitHub}, \href{https://swmath.org/} {swMATH} and domain related publications,
-such as review articles. Domain experts are also asked for their suggestions and
-are asked to review the initial draft of the software list.
-
-When forming the list and reviewing the candidate software the following
-properties should be considered:
-
-\begin{enumerate}
-\item The software functionality must fall within the identified domain.
-\item The source code must be viewable.
-\item The empirical measures listed in Section \ref{SecEmpiricalMeasures} should
-  ideally be available, which implies a preference for GitHub repositories.
-\item The software cannot be marked as incomplete or in an initial development
-  phase.
-\end{enumerate}
-
-\section{How to Initially Filter the Software List} \label{SecInitialFilter}
-
-If the list of software is too long (over around 30 packages), then steps need
-to be taken to create a more manageable list. To reduce the length of the list,
-the following filters are applied.  The filters are applied in the priority
-order listed, with the filtering process stopped once the list size is
-manageable.
-
-\begin{enumerate}
-\item Scope: Software is removed by narrowing what functionality is considered
-  to be within the scope of the domain.
-\item Usage: Software packages are eliminated if their installation procedure is
-  not clear and easy to follow.
-\item Age: The older software packages (age being measured by the last date when
-  a change was made) are eliminated, except in the cases where an older software
-  package appears to be highly recommended and currently in use.
-\end{enumerate}
-
-Copies of both the initial and filtered lists should be kept for traceability purposes.
-
-\section{Domain Analysis} \label{SecDomainAnalysis}
-
-Since each domain we will study will have a reasonably small scope, we
-will be able to view the software as constituting a program
-family.  The concept of a program family is defined by \citet{parnas1976design}
-as ``a set of programs whose common properties are so extensive that it is
-advantageous to study the common properties of the programs before analyzing
-individual members''.  Studying the common properties within a family of related
-programs is termed a domain analysis.
-
-The domain analysis consists of a commonality analysis of
-the family of software packages. Its purpose is to show the relationships
-between these packages, and to facilitate an understanding of the informal
-specification and development of them. \cite{weiss1997defining} defines
-commonality analysis as an approach to defining a family by identifying
-commonalities, variabilities, and common terminology for the
-family. Commonalities are goals, theories, models, definitions and assumptions
-that are common between family members.  Variabilities are goals, theories,
-models, definitions and assumptions that differ between family
-members. Associated with each variability are its parameters of variation, which
-summarize the possible values for that variability, along with their potential
-binding time.  The binding time is when the value of the variability is set.  It
-could be set as specification time, build time (when the program is being
-compiled) or run time (when the code is executing).
-
-The final result of the domain analysis will be
-tables of commonalities, variabilities, and parameters of variation of a program
-family.   \cite{smith2008commonality} present a template for conducting a
-commonality analysis, which was referred to when conducting this
-work. \cite{weiss1998commonality} describes another commonality analysis
-technique for deciding the members of a program family which readers are
-encouraged to be familiar with. \cite{SmithAndChen2004} and
-\cite{SmithMcCutchanAndCarette2017} are examples of a commonality analysis for a
-family of mesh generating software and a family of material models,
-respectively. The steps to produce a commonality analysis are:
-
-\begin{enumerate}
-\item Write an Introduction
-\item Write the Overview of Domain
-\item List Commonalities
-\item List Variabilities
-\item List Parameters of Variation
-\item Add Terminology, Definitions, Acronyms
-\end{enumerate}
-
-A sample commonality analysis for Lattice Boltzmann Solvers can be found
-\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Peter-Notes/Commonality-Analysis-LB-Systems.pdf}
-{here}.
-
-\section{Empirical Measures} \label{SecEmpiricalMeasures}
-
-Some quality measurements rely on the gathering of raw and processed empirical
-data. We focus on data that is reasonably easy to collect, which we combine and
-analyze for this assessment. The measures that are collected relate to the
-research questions (Section~\ref{ResearchQuestions}). For instance, we collect some of the
-data to see how large a project is. Other measures intend to ascertain a
-project’s popularity, or how active the project is.
-
-Section \ref{rawdata} orients the reader on what raw data is collected.  Some of this data can be
-observed from Git repository metrics. The rest can be collected using freeware
-tools. \href{https://github.com/tomgi/git_stats}{GitStats} is used to measure
-the number of binary files as well as the number of added and deleted lines in a
-repository. This tool is also used to measure the number of commits over
-different intervals of time. \href{https://github.com/boyter/scc}{Sloc Cloc and
-  Code (scc)} is used to measure the number of text based files as well as the
-number of total, code, comment, and blank lines in a repository. These tools
-were selected due to their installability, usability, and ability to gather the
-empirical measures listed below. Details on installing and running the tools can
-be found in our 
-\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/A
-  Guide to Empirical Measures.pdf} {Guide to Empirical Measures}. Section \ref{processeddata}
-introduces the required processed data, which is calculated using the raw data.
-
-\subsection{Raw Data}\label{rawdata}
-
-The following raw data measures are extracted from repositories:
-
-\begin{itemize}
-\item Number of stars.
-\item Number of forks.
-\item Number of people watching the repository.
-\item Number of open pull requests.
-\item Number of closed pull requests.	
-\item Number of developers.	
-\item Number of open issues.
-\item Number of closed issues.
-\item Initial release date.
-\item Last commit date.
-\item Programming languages used.
-\item Number of text-based files.
-\item Number of total lines in text-based files.
-\item Number of code lines in text-based files.
-\item Number of comment lines in text-based files.
-\item Number of blank lines in text-based files.
-\item Number of binary files.  
-\item Number of total lines added to text-based files.
-\item Number of total lines deleted from text-based files.
-\item Number of total commits.
-\item Numbers of commits by year in the last 5 years. (Count from as early as
-  possible if the project is younger than 5 years.)
-\item Numbers of commits by month in the last 12 months.
-\end{itemize}
-
-\subsection{Processed Data}\label{processeddata}
-
-The following measures are calculated from the raw data:
-
-\begin{itemize}
-\item Status of software package (dead or alive). Alive is defined as the
-  presence of repository commits or software package version releases in the
-  last 18 months.
-\item Percentage of identified issues that are closed.
-\item Percentage of code that is comments.
-\end{itemize}
-
-\section{Measure Using Measurement Template} \label{SecShallowMeasure}
-
-\begin{figure}[h!]
-  \begin{center}
-    \includegraphics[width=1.0\textwidth]{measurement_template}
-    \caption{Top of Measurement Template}
-    \label{measurement_template_image}
-  \end{center}
-\end{figure}
-
-The
-\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Combined_MeasurementTemplate_EmpiricalMeasures.xlsx}
-{Measurement Template} is used to track measurements and quality scores for all
-of the software packages in the domain. For each software package, we fill out one
-column of the template. This process can take between 2 to 5 hours for each
-package.  Project developers can be contacted for help regarding installation,
-if necessary, but a cap of about 2 hours should be imposed on the entire
-installation process, to keep the overall measurement time feasible.
-
-Most of the data to be collected should be straightforward from reviewing the
-measurement template.  However, in some cases some extra guidance is necessary
-to eliminate ambiguity, as follows:
-
-\begin{enumerate}
-\item Initial release date: Mark the release year if an exact date is not
-  available.
-\item Publications about the software: A list of publications can be found
-  directly on the website of some software packages. For others use Google
-  Scholar or a similar index.
-\item Is there evidence that performance was considered?: Search the software
-  artifacts for any mention of speed, storage, throughput, performance
-  optimization, parallelism, multi-core processing, or similar
-  considerations. The search function on GitHub can help.
-\item Getting started tutorial: Sometimes this is found within another artifact,
-  like the user manual.
-\item Continuous integration: Search the software artifacts for any mention of
-  continuous integration. The search function on GitHub can help.
-\end{enumerate}
-
-For completing the template, the following steps should be followed:
-
-\begin{enumerate}
-\item Gather the summary information into the top section of the document
-  (Figure \ref{measurement_template_image})
-\item Using the GitStats tool that is described in Section
-  \ref{SecEmpiricalMeasures} gather the measurements for the Repo Metrics -
-  GitStats section found near the bottom of the document
-\item Using the SCC tool that is also described in Section
-  \ref{SecEmpiricalMeasures} gather the measurements for the Repo Metrics - SCC
-  section found near the bottom of the document
-\item If the software package is found on git, gather the measurements for the
-  Repo Metrics - GitHub section found near the bottom of the document
-\item Review installation documentation and attempt to install the software
-  package on a virtual machine
-\item Gather the measurements for installability
-\item Gather the measurements for correctness and verifiability
-\item Gather the measurements for surface reliability
-\item Gather the measurements for surface robustness
-\item Gather the measurements for surface usability
-\item Gather the measurements for maintainability
-\item Gather the measurements for reusability
-\item Gather the measurements for surface understandability
-\item Gather the measurements for visibility and transparency
-\item Assign a score out of ten for each quality. The score can be measured
-  using the
-  \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/MeasurementTemplate_ImpressionCalculator.xlsx}
-  {Measurement Template Impression Calculator}. For each quality measurement,
-  the file indicates the appropriate score to assign the measurement based on
-  possible measurement values.
-\end{enumerate}
-
-\section{Analytic Hierarchy Process} \label{SecAHP}
-
-The Analytical Hierarchy Process (AHP) is a decision-making technique that can
-be used when comparing multiple options by multiple criteria. In our work AHP is
-used for comparing and ranking the software packages of a domain using the
-quality scores that are gathered in the
-\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Combined_MeasurementTemplate_EmpiricalMeasures.xlsx}
-{Measurement Template}. AHP performs a pairwise analysis between each of the
-quality options using a matrix which is then used to generate an overall score
-for each software package for the given criteria. \cite{SmithEtAl2016} shows how
-AHP is applied to ranking software based on quality measures. We have developed
-a tool for conducting this process. The tool includes an AHP JAR script and a
-sensitivity analysis JAR script that is used to ensure that the software package
-rankings are appropriate with respect to the uncertainty of the quality
-scores. The README file of the tool is found
-\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/AHP2020/LBM/README.txt}{here}. This
-file outlines the requirements for, and configuration and usage of, the JAR
-scripts. The JAR scripts, source code, and required libraries are located in the
-same folder as the README file.
-
-\section{Rank Short List} \label{SecRankShortList}
-
-An AHP pairwise comparison between the short list packages with respect to
-usability and modifiability survey results is conducted by the domain
-expert. The tool can be found here. The usability and modifiability survey can
-be found
-\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Experiments.pdf}
-{here}. This step is intended to start a conversation about the relative quality of
-short-listed software packages.
-
-\section{Using Data to Rank Family Members}
-
-Describe AHP process (or similar).
-
 \newpage
-
-\appendix
-
-\section{Appendix}
+\section{Empirical Measurement}
 
@@
 
-\ad{This section is currently a note of unorganized contents. Most parts will
-  beremoved or relocated.}
-
-\ad{This citation needs to be deleted later. It's here because my compiler
-  doesn't work with 0 citations} \cite{Emms2019}
+The following tools were considered when selecting a tool for taking empirical measurements of repositories.
 

### 2020-12-01 StateOfPractice/Methodology/Methodology.tex
moved the list of considered empirical measurement tools from the appendix into a new traceability document

@@
 
-\section{Appendix}
-
-\subsection{Empirical Measures Considerations - Tool Tests}
-
-\ad{This section is currently a note of unorganized contents. Most parts will
-  beremoved or relocated.}
-
-\ad{This citation needs to be deleted later. It's here because my compiler
-  doesn't work with 0 citations} \cite{Emms2019}
-
-Most tests were done targeting to the repo of 3D Slicer
-\href{https://github.com/tomgi/git_stats}{GitHub repo}
-
-\subsubsection{git-stats}
-
-\href{https://github.com/tomgi/git_stats}{GitHub repo}
-
-Test results:
-\href{http://git-stats-slicer.ao9.io/}{http://git-stats-slicer.ao9.io/} the
-results are output as webpages, so I hosted for you to check. Data can be
-downloaded as spreadsheets.
-
-\subsubsection{scc}
-
-\href{https://github.com/boyter/scc}{GitHub repo}
-
-\subsubsection{git-of-theseus}
-
-\href{https://github.com/erikbern/git-of-theseus}{GitHub repo}
-
-Test results: It took about 100 minutes for one repo on a 8 core 16G ram Linux
-machine. It only outputs graphs.
-
-\subsubsection{hercules}
-
-\href{https://github.com/src-d/hercules}{GitHub repo}
-
-Test results: this one seems to be promising, but the installation is
-complicated with various errors.
-
-\subsubsection{git-repo-analysis}
-\href{https://github.com/larsxschneider/git-repo-analysis}{GitHub repo}
-
-\subsubsection{HubListener}
-\href{https://github.com/pjmc-oliveira/HubListener}{GitHub repo}
-
-The data that HubListener can extract.
-
-Raw:
-\begin{itemize}
-\item Number of Files
-\item Number of Lines
-\item Number of Logical Lines
-\item Number of Comments
-\end{itemize}
-
-Cyclomatic:
-\href{https://www.geeksforgeeks.org/cyclomatic-complexity/}{Intro}
-\begin{itemize}
-\item Cyclomatic Complexity
-\end{itemize}
-
-Halstead:
-\href{https://www.geeksforgeeks.org/software-engineering-halsteads-software-metrics/}{Intro}
-\begin{itemize}
-\item Halstead Effort
-\item Halstead Bugs
-\item Halstead Length
-\item Halstead Difficulty
-\item Halstead Time
-\item Halstead Vocabulary
-\item Halstead Volume
-\end{itemize}
-
-Test results: HubListener works well on the repo of itself, but it did not work
-well on some other repos.
-
-\subsubsection{gitinspector}
-
-\href{https://github.com/ejwa/gitinspector}{GitHub repo}
-
-Test results: it doesn't work well. Instead of creating output results, it
-prints the results directly in the console.
-
 \newpage

### 2020-12-01 StateOfPractice/Methodology/Methodology.tex
Updates to methodology

@@
 
+\ccsdesc{Software and its engineering}
+\ccsdesc{Software and its engineering~Software product lines}
+\ccsdesc{General and reference~Empirical studies}
+
+\keywords{software quality, domain analysis, scientific computing software,
+  research software, empirical measures, analytic hierarchy process}
+
 \date{\today}
@@
 \begin{document}
+
 \maketitle
@@
 
+~\newpage
+
 \tableofcontents
+
+~\newpage
 
@@
 exercise, can be found
-\href{https://github.com/smiths/AIMSS/blob/master/OverallResearchProposal/ResearchProposal.pdf}{here}. The
-research questions that this exercise will answer are found in Section
-\ref{ResearchQuestions}. Section \ref{StepsAQDS} outlines the steps to be taken.
+\href{https://github.com/smiths/AIMSS/blob/master/OverallResearchProposal/ResearchProposal.pdf}
+{here}. The research questions that this exercise will answer are found in
+Section \ref{ResearchQuestions}. Section \ref{StepsAQDS} outlines the steps to
+be taken.
 

### 2020-12-03 StateOfPractice/Methodology/Methodology.tex
removed old note
Beginning of some additional content for the introduction - still need to add refs to bib, need to complete into

@@
 \section{Introduction} \label{SecIntroduction}
+
+Research software involves using scientific computing tools to simulate
+mathematical models of real world systems so that we can better understand and
+predict the system's behaviour.  A small sample of some important examples of
+research software include the following: designing new automotive parts,
+analysing the flow of blood in the body, and determining the concentration of a
+pollutant released into the groundwater.  As these examples illustrate, research
+software can be used for tackling problems that impact such areas as
+manufacturing, financial planning, environmental policy, and the health, welfare
+and safety of communities.
+
+The challenging.  Room for improvement.  SEI, BME.  Problem in that current
+state-of-the-practice for research software does not incorporate
+state-of-the-art SE tools and methods \citep{JohansonAndHasselbring2018}.  The
+lack of SE tools and methods is viewed as contributing to sustainability and
+reliability problems \citep{FaulkEtAl2009}.  Embarrassing failures have
+occurred, like a retraction of derived molecular protein structures
+\citep{Miller2006} and fixing and then reintroducing the same error in a large
+code base three times in 20 year \citep{MilewiczAndRaybourn2018}.  To improve
+this situation, we need to more fully understand the current state of the
+practice for research software.
 
@@
 
-\section{Using Data to Rank Family Members}
-
-Describe AHP process (or similar).
-
 \newpage

### 2020-12-04 StateOfPractice/Methodology/Methodology.tex
Further work on the introduction section of the methodology

--- 2020-12-01 StateOfPractice/Methodology/Methodology.tex
@@
 \section{Introduction} \label{SecIntroduction}
+
+Research software uses computing to simulate mathematical models of real world
+systems so that we can better understand and predict the system's behaviour.  A
+small sample of some important examples of research software include the
+following: designing new automotive parts, analysing the flow of blood in the
+body, and determining the concentration of a pollutant released into the
+groundwater.  As these examples illustrate, research software can be used for
+tackling important problems that impact such areas as manufacturing, financial
+planning, environmental policy, and the health, welfare and safety of
+communities.
+
+Given the importance of research software, scientists and engineers are pushing
+for sustainable development practices.  This is evident from the existence of
+such groups as SEI, BME.  Groups promoting sustainable practices are necessary
+because the current state-of-the-practice for research software does not
+incorporate state-of-the-art SE tools and methods
+\citep{JohansonAndHasselbring2018}.  The lack of SE tools and methods
+contributes to sustainability and reliability problems \citep{FaulkEtAl2009}.
+Problems with the current state of the practice are evident from embarrassing
+failures, like a retraction of derived molecular protein structures
+\citep{Miller2006}, false reproduction of sonoluminescent fusion
+\citep{PostAndVotta2005}, and fixing and then reintroducing the same error in a
+large code base three times in 20 year \citep{MilewiczAndRaybourn2018}.  To
+improve this situation, we need to more fully understand the current state of
+the practice for research software.
+
+To understand the current quality of SCS and to assist in developing methods for
+measuring qualities, we will identify several domains within SCS and endeavour
+to summarize the current state of the practice within those domains.  This
+``state of the practice'' exercise will build off of prior work on
+measuring/assessing the state of software development practice in several SCS
+domains.  We will update the work that was done previously for domains such as
+Geographic Information Systems \citep{SmithEtAl2018_arXivGIS}, Mesh Generators
+\citep{SmithEtAl2016}, Seismology software \citep{SmithEtAl2018}, and
+Statistical software for psychology \citep{SmithEtAl2018_StatSoft}.  We may
+return to these domains and/or introduce new domains.  Potential
+domains/collaborators are listed in the subsections below.
+
+In the previous project, we measured 30 software projects for each domain, but
+the measures were relatively shallow.  With the re-boot we will refine the llist
+into a short list of the best projects for more detailed measurement.
+
+In the previous state of the practice assessing exercise
+\citep{SmithEtAl2018_arXivGIS}, a series of questions and simple metrics was
+devised to measure the quality of the documentation and adherence to best
+software development practices.  The previous set of questions is provided
+\href{https://github.com/adamlazz/DomainX/blob/master/TemplateToGradeSCSoft.pdf}
+{here}. As described below, the new project critically assess the previous set of questions and
+revises as necessary.  In addition, the following data is also collected for each domain:
+
+\begin{itemize}
+\item Characterization of the functionality provided by the software in the
+  domain.  What services does each member of the domain provide?  This
+  information would be summarized in a commonality analysis document.
+\item Usability testing of each software in the domain.
+\item Collect empirical software engineering related data, such as the number of
+  files, number of lines of code, number of open issues,
+  number of closed issues, the number of lines of documentation (?). 
+\end{itemize}
+
+One of the guidelines for successfully on-boarding new developers is to
+  ``Keep knowledge up to date and findable'' \citep{ShollerEtAl2019}.  The
+  selected measurements should be able to assist with judging whether a project
+  succeeds on this front.
+
+The above data will be combined to rank the software in the domain.  The
+specifics are yet to be determined, but in the past the Analytic Hierarchy
+Process (AHP) \citep{Saaty1980} proved helpful in this.  In the past the AHP
+results were calculated by a grade out of 10 from each student evaluator for
+each quality for each measured product.  A better approach may be to have a
+group of people familiar with the software do a pairwise comparison between each
+product to come up with a number.  The number would be based on the data
+collected, but not in a numerical way.  The data collected would serve the same
+role as a decision support system.  It would provided insight, but human beings
+would do the final pairwise comparisons, and thus the final ranking.
+
+Key to the success of this exercise will be to involve and engage a partner for
+each state of practice project.  In the previous effort we did not involve domain
+experts with the rationale of excluding potential bias, but this advantage is
+not worth not being able to evaluate the functionality/usability of the software.
+Moreover, not having an expert makes publication more difficult, since there is
+no one to advice on how best to approach journals and publishers.  The domain
+expert will be asked to help in the following ways:
+
+GitHub will be used to coordinate the work of the large team of people that will
+be involved in this project.  In addition to the project record left on GitHub,
+the final data will be exported to Mendeley (as was done for the previous
+quality measurement exercise, for instance the grades for GIS software are
+summarized at \href{https://data.mendeley.com/datasets/6kprpvv7r7/1}
+{https://data.mendeley.com/datasets/6kprpvv7r7/1}.)
+
+The data for each of the different subdomains within SCS will be standardized to
+facilitate a meta-analysis that combines all of the data.
 

### 2020-12-11 StateOfPractice/Methodology/Methodology.tex
Further work updating methodology document introduction

@@
 Research software uses computing to simulate mathematical models of real world
-systems so that we can better understand and predict the system's behaviour.  A
-small sample of some important examples of research software include the
-following: designing new automotive parts, analysing the flow of blood in the
+systems so that we can better understand and predict those system's behaviour.  A
+small set of examples of important research software includes the
+following: designing new automotive parts, analyzing the flow of blood in the
 body, and determining the concentration of a pollutant released into the
@@
 Given the importance of research software, scientists and engineers are pushing
-for sustainable development practices.  This is evident from the existence of
-such groups as SEI, BME.  Groups promoting sustainable practices are necessary
-because the current state-of-the-practice for research software does not
-incorporate state-of-the-art SE tools and methods
+for methods and tools to sustainably develop hiqh quality software.  This is
+evident from the existence of such groups as the Software Sustainability
+Institute (\href{https://www.software.ac.uk/} {SEI}) and Better Scientific
+Software (\href{https://bssw.io/} {BSS}).  Groups such as these are necessary
+for promoting sustainable practices because unfortunately the current
+state-of-the-practice for research software often does not incorporate
+state-of-the-art Software Engineering (SE) tools and methods
 \citep{JohansonAndHasselbring2018}.  The lack of SE tools and methods
@@
 large code base three times in 20 year \citep{MilewiczAndRaybourn2018}.  To
-improve this situation, we need to more fully understand the current state of
+improve this situation, we need to first more fully understand the current state of
 the practice for research software.
 
-To understand the current quality of SCS and to assist in developing methods for
-measuring qualities, we will identify several domains within SCS and endeavour
-to summarize the current state of the practice within those domains.  This
-``state of the practice'' exercise will build off of prior work on
-measuring/assessing the state of software development practice in several SCS
-domains.  We will update the work that was done previously for domains such as
-Geographic Information Systems \citep{SmithEtAl2018_arXivGIS}, Mesh Generators
-\citep{SmithEtAl2016}, Seismology software \citep{SmithEtAl2018}, and
-Statistical software for psychology \citep{SmithEtAl2018_StatSoft}.  We may
-return to these domains and/or introduce new domains.  Potential
-domains/collaborators are listed in the subsections below.
+To understand the current quality of research software and to assist in
+developing methods for measuring qualities, we will identify several domains
+within research software and endeavour to summarize the current state of the
+practice within those domains.  This ``state of the practice'' exercise will
+build off of prior work on measuring/assessing the state of software development
+practice in several research domains.  We will update the work that was done
+previously for domains such as Geographic Information Systems
+\citep{SmithEtAl2018_arXivGIS}, Mesh Generators \citep{SmithEtAl2016},
+Seismology software \citep{SmithEtAl2018}, and Statistical software for
+psychology \citep{SmithEtAl2018_StatSoft}.
 
 In the previous project, we measured 30 software projects for each domain, but
-the measures were relatively shallow.  With the re-boot we will refine the llist
+the measures were relatively shallow.  With the re-boot we are refining the list
 into a short list of the best projects for more detailed measurement.

### 2020-12-11 StateOfPractice/Methodology/Methodology.tex
fixed error in Section 11 Methodology

@@
 An AHP pairwise comparison between the short list packages with respect to
-usability and modifiability survey results is conducted by the domain
-expert. The tool can be found here. The usability and modifiability survey can
-be found
+usability and modifiability survey results is conducted, and then reviewed by the domain expert. The tool can be found here. The usability and modifiability survey can be found
 \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Experiments.pdf}
-{here}. This step is intended to start a conversation about the relative quality of
-short-listed software packages.
+{here}. This step is intended to start a conversation about the relative quality of short-listed software packages.
 

### 2020-12-11 StateOfPractice/Methodology/Meeting Agenda with Domain Experts.tex
added modifiability experiment to meeting agenda with domain experts
added 2 tasks, and reordered tasks based on order of Methodology steps
cleaned up Meeting Agenda document

@@
 
+\subsection{Research Proposal}
+\label{research_proposal}
+Presenter: Dr.\ Spencer Smith \ad{TBD}
+
+The research proposal is documented in
+\href{https://github.com/smiths/AIMSS/blob/master/OverallResearchProposal/ResearchProposal.pdf}{Research
+	Proposal}.
+
+A brief introduction. \ad{To be written}
+
+Do you have any suggestions? Feedback can be provided later in Task
+\ref{task_research_proposal}.
+
 \subsection{Research Questions}
@@
 \ref{task_research_questions}.
-
-\subsection{Research Proposal}
-\label{research_proposal}
-Presenter: Dr.\ Spencer Smith \ad{TBD}
-
-The research proposal is documented in
-\href{https://github.com/smiths/AIMSS/blob/master/OverallResearchProposal/ResearchProposal.pdf}{Research
-Proposal}.
-
-A brief introduction. \ad{To be written}
-
-Do you have any suggestions? Feedback can be provided later in Task
-\ref{task_research_proposal}.
-
-\section{Software List}
-Presenter: Ao or Peter \ad{TBD}
-
-In this section, we would like to discuss the list of software packages and the
-process of creating and shortening it.
-
-\subsection{Top 10 List}
-\label{top_10_list}
-What will be the top 10 software of your choice? Feedback can be provided later
-in Task \ref{task_top_10_list}.
-
-\subsection{Domain Analysis}
-The domain analysis can be found in Section 3.2 of
-\href{https://github.com/Ao99/MISEG/blob/master/docs/SRS/SRS.pdf}{SRS for
-Medical Imaging Software} or
-\href{https://github.com/peter-michalski/LatticeBoltzmannSolvers/blob/master/docs/SRS/CA.pdf}{CA
-of Lattice Boltzmann Solvers}.
-
-What do you think of our analysis?  \wss{This is too big a question for the
-  domain experts to answer in a reasonable amount of time.  Moreover, without
-  more guidance their feedback is likely to be shallow.  I think we are better
-  off with a task based inspection, using issues assigned via a GitHub project
-  page.  We also will likely be more successful asking for reviews from grad
-  students, rather than from faculty members.  We should also consider
-  summarizing the commonalities, variabilities and parameters of variation in
-  tables, or possibly graphs.}
-
-\subsection{Selecting Process}
-The process of creating and shortening the list of software packages is
-documented in
-\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Ao-Notes/Ao-Notes.pdf}{Ao's
-Notes} or Peter's Notes \ad{TBD}
-
-What do you think of our process of selecting software candidates?
-
-\subsection{Software List}
-\label{software_list}
-The software list is
-\href{https://docs.google.com/spreadsheets/d/122wU0v3ZtvDcqy8C4zKJ89kU-8fXAbo3Mzn6vcVXOi0/edit?usp=sharing}{Medical
-Imaging Software List} or Lattice Boltzmann Solver List. \ad{TBD}
-
-What do you think of the list of the selected software?
-
-Is there any software you would like to remove from or add to the list?
-
-Feedback can be provided later in Task \ref{task_software_list}.
 
@@
 
+\subsection{Modifiability Experiment}
+\label{modifiability_experiment}
+Presenter:
+
+A brief introduction.
+
+The usability test can be found..
+
+Do you have any suggestions? Feedback can be provided later in Task
+\ref{modifiability_experiment}
+
+\section{Software List}
+Presenter: Ao or Peter \ad{TBD}
+
+In this section, we would like to discuss the list of software packages and the
+process of creating and shortening it.
+
+\subsection{Top 10 List}
+\label{top_10_list}
+What will be the top 10 software of your choice? Feedback can be provided later
+in Task \ref{task_top_10_list}.
+
+\subsection{Software List}
+\label{software_list}
+The software list is
+\href{https://docs.google.com/spreadsheets/d/122wU0v3ZtvDcqy8C4zKJ89kU-8fXAbo3Mzn6vcVXOi0/edit?usp=sharing}{Medical
+	Imaging Software List} or Lattice Boltzmann Solver List. \ad{TBD}
+
+What do you think of the list of the selected software?
+
+Is there any software you would like to remove from or add to the list?
+
+Feedback can be provided later in Task \ref{task_software_list}.
+
+\subsection{Domain Analysis}
+\label{domain_analysis}
+The domain analysis can be found in Section 3.2 of
+\href{https://github.com/Ao99/MISEG/blob/master/docs/SRS/SRS.pdf}{SRS for
+Medical Imaging Software} or
+\href{https://github.com/peter-michalski/LatticeBoltzmannSolvers/blob/master/docs/SRS/CA.pdf}{CA
+of Lattice Boltzmann Solvers}.
+
+What do you think of our analysis?  \wss{This is too big a question for the
+  domain experts to answer in a reasonable amount of time.  Moreover, without
+  more guidance their feedback is likely to be shallow.  I think we are better
+  off with a task based inspection, using issues assigned via a GitHub project
+  page.  We also will likely be more successful asking for reviews from grad
+  students, rather than from faculty members.  We should also consider
+  summarizing the commonalities, variabilities and parameters of variation in
+  tables, or possibly graphs.}
+
+\subsection{First AHP Ranking}
+\label{first_AHP}
+Presenter:
+
+A brief introduction.
+
+Do you have any feedback?
+
+\subsection{Final AHP Ranking}
+\label{final_AHP}
+Presenter:
+
+A brief introduction.
+
+Do you have any feedback?
+
 \section{Publication}
@@
 
+\newpage
+
 \section{Tasks for Domain Experts}
@@
 
-\noindent Back to Section \ref{project_objectives}
+\noindent Refer to Section \ref{project_objectives}
+
+\subsection{Review the Research Proposal}
+\label{task_research_proposal}
+GitHub Issue:
+\href{https://github.com/smiths/AIMSS/issues}{https://github.com/smiths/AIMSS/issues}
+\ad{To be created}
+
+\noindent Refer to Section \ref{research_proposal}
 
@@
 
-\noindent Back to Section \ref{research_questions}
+\noindent Refer to Section \ref{research_questions}
 
-\subsection{Review the Research Proposal}
-\label{task_research_proposal}
+\subsection{Review the Measurement Template}
+\label{task_measurement_template}
 GitHub Issue:
@@
 
-\noindent Back to Section \ref{research_proposal}
+\noindent Refer to Section \ref{measurement_template}
+
+\subsection{Review the Survey for Short List Projects}
+\label{task_survey_short_list_projects}
+GitHub Issue:
+\href{https://github.com/smiths/AIMSS/issues}{https://github.com/smiths/AIMSS/issues}
+\ad{To be created}
+
+\noindent Refer to Section \ref{survey_short_list_projects}
+
+\subsection{Review the Usability Tests}
+\label{task_usability_tests}
+GitHub Issue:
+\href{https://github.com/smiths/AIMSS/issues}{https://github.com/smiths/AIMSS/issues}
+\ad{To be created}
+
+\noindent Refer to Section \ref{usability_tests}
+
+\subsection{Review the Modifiability Experiment}
+\label{task_modifiability_experiment}
+GitHub Issue:
+\href{https://github.com/smiths/AIMSS/issues}{https://github.com/smiths/AIMSS/issues}
+\pmi{To be created}
+
+\noindent Refer to Section \ref{modifiability_experiment}
 
@@
 
-\noindent Back to Section \ref{top_10_list}
+\noindent Refer to Section \ref{top_10_list}
 
@@
 
-\noindent Back to Section \ref{software_list}
+\noindent Refer to Section \ref{software_list}
 
 \subsection{Review the Domain Analysis}
-\pmi{added this task as requested in AIMSS issue  \href{https://github.com/smiths/AIMSS/issues/56}{56}}
+GitHub Issue:
+\href{https://github.com/smiths/AIMSS/issues}{https://github.com/smiths/AIMSS/issues}
+\pmi{To be created}
+
+\noindent Refer to Section \ref{domain_analysis}
 
 \subsection{Review the AHP Ranking and Short List of Software Packages}
-\pmi{added this task as requested in AIMSS issue  \href{https://github.com/smiths/AIMSS/issues/54}{54}}
-
-\subsection{Review the Measurement Template}
-\label{task_measurement_template}
 GitHub Issue:
 \href{https://github.com/smiths/AIMSS/issues}{https://github.com/smiths/AIMSS/issues}
-\ad{To be created}
+\pmi{To be created}
 
-\noindent Back to Section \ref{measurement_template}
+\noindent Refer to Section \ref{first_AHP}
 
-\subsection{Review the Survey for Short List Projects}
-\label{task_survey_short_list_projects}
+\subsection{Comment on Final Short List Ranking}
+\label{task_final_ranking}
 GitHub Issue:
 \href{https://github.com/smiths/AIMSS/issues}{https://github.com/smiths/AIMSS/issues}
-\ad{To be created}
+\pmi{To be created}
 
-\noindent Back to Section \ref{survey_short_list_projects}
+\noindent Refer to Section \ref{final_AHP}
 
-\subsection{Review the Usability Tests}
-\label{task_usability_tests}
-GitHub Issue:
-\href{https://github.com/smiths/AIMSS/issues}{https://github.com/smiths/AIMSS/issues}
-\ad{To be created}
-
-\noindent Back to Section \ref{usability_tests}
+%
+%\section{Deleted Notes}
+%\subsection{Selecting Process}
+%The process of creating and shortening the list of software packages is
+%documented in
+%\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Ao-Notes/Ao-N%otes.pdf}{Ao's
+%	Notes} or Peter's Notes \ad{TBD}
+%
+%What do you think of our process of selecting software candidates?
+%
 

### 2020-12-14 StateOfPractice/Methodology/Methodology.tex
Further edits to the introduciton section.

@@
 Institute (\href{https://www.software.ac.uk/} {SEI}) and Better Scientific
-Software (\href{https://bssw.io/} {BSS}).  Groups such as these are necessary
-for promoting sustainable practices because unfortunately the current
-state-of-the-practice for research software often does not incorporate
-state-of-the-art Software Engineering (SE) tools and methods
-\citep{JohansonAndHasselbring2018}.  The lack of SE tools and methods
-contributes to sustainability and reliability problems \citep{FaulkEtAl2009}.
-Problems with the current state of the practice are evident from embarrassing
-failures, like a retraction of derived molecular protein structures
-\citep{Miller2006}, false reproduction of sonoluminescent fusion
-\citep{PostAndVotta2005}, and fixing and then reintroducing the same error in a
-large code base three times in 20 year \citep{MilewiczAndRaybourn2018}.  To
-improve this situation, we need to first more fully understand the current state of
-the practice for research software.
+Software (\href{https://bssw.io/} {BSS}).  Sustainability promoting groups such
+as these are necessary because unfortunately the current state-of-the-practice
+for research software often does not incorporate state-of-the-art Software
+Engineering (SE) tools and methods \citep{JohansonAndHasselbring2018}.  The lack
+of SE tools and methods contributes to sustainability and reliability problems
+\citep{FaulkEtAl2009}.  Problems with the current state of the practice are
+evident from embarrassing failures, like a retraction of derived molecular
+protein structures \citep{Miller2006}, false reproduction of sonoluminescent
+fusion \citep{PostAndVotta2005}, and fixing and then reintroducing the same
+error in a large code base three times in 20 year
+\citep{MilewiczAndRaybourn2018}.  To improve this situation, we need to first
+fully understand the current state of the practice for research software.
 
-To understand the current quality of research software and to assist in
-developing methods for measuring qualities, we will identify several domains
-within research software and endeavour to summarize the current state of the
-practice within those domains.  This ``state of the practice'' exercise will
-build off of prior work on measuring/assessing the state of software development
-practice in several research domains.  We will update the work that was done
-previously for domains such as Geographic Information Systems
-\citep{SmithEtAl2018_arXivGIS}, Mesh Generators \citep{SmithEtAl2016},
-Seismology software \citep{SmithEtAl2018}, and Statistical software for
-psychology \citep{SmithEtAl2018_StatSoft}.
+The purpose of our proposed methodology is to gain insight on how software
+quality is impacted by the software development principles, processes and tools
+currently used within research software communities.  Since research software is
+so broad a category, we will reduce the scope of each application of the
+methodology by applying it to a specific domain within research software.  This
+``state of the practice for domain X'' exercise will build off of prior work on
+measuring/assessing the state of software development practice in several
+research domains.  We will update the work that was done previously for domains
+such as Geographic Information Systems \citep{SmithEtAl2018_arXivGIS}, Mesh
+Generators \citep{SmithEtAl2016}, Seismology software \citep{SmithEtAl2018}, and
+Statistical software for psychology \citep{SmithEtAl2018_StatSoft}.
 
 In the previous project, we measured 30 software projects for each domain, but
-the measures were relatively shallow.  With the re-boot we are refining the list
-into a short list of the best projects for more detailed measurement.
+the measures were relatively shallow.  With this re-boot we are adding more
+measures and experiments.  refining the list
+into a short list of the best projects for more detailed measurement.  More details.
 

### 2020-12-18 StateOfPractice/Methodology/Methodology.tex
Completed revision of introduction of methodology section

@@
 Research software uses computing to simulate mathematical models of real world
-systems so that we can better understand and predict those system's behaviour.  A
+systems so that we can better understand and predict those systems's behaviour.  A
 small set of examples of important research software includes the
@@
 
-The purpose of our proposed methodology is to gain insight on how software
-quality is impacted by the software development principles, processes and tools
-currently used within research software communities.  Since research software is
-so broad a category, we will reduce the scope of each application of the
-methodology by applying it to a specific domain within research software.  This
-``state of the practice for domain X'' exercise will build off of prior work on
-measuring/assessing the state of software development practice in several
-research domains.  We will update the work that was done previously for domains
-such as Geographic Information Systems \citep{SmithEtAl2018_arXivGIS}, Mesh
-Generators \citep{SmithEtAl2016}, Seismology software \citep{SmithEtAl2018}, and
-Statistical software for psychology \citep{SmithEtAl2018_StatSoft}.
+The purpose of our proposed methodology is to understand how software quality is
+impacted by the software development principles, processes and tools currently
+used within research software communities.  Since research software is so broad
+a category, we will reduce the scope of our methodology to focus on one specific
+research software domain at a time.  This ``state of the practice for domain X''
+exercise will build off of prior work on measuring/assessing the state of
+software development practice in several research domains.  We will update the
+work that was done previously for domains such as Geographic Information Systems
+\citep{SmithEtAl2018_arXivGIS}, Mesh Generators \citep{SmithEtAl2016},
+Seismology software \citep{SmithEtAl2018}, and Statistical software for
+psychology \citep{SmithEtAl2018_StatSoft}.
 
-In the previous project, we measured 30 software projects for each domain, but
-the measures were relatively shallow.  With this re-boot we are adding more
-measures and experiments.  refining the list
-into a short list of the best projects for more detailed measurement.  More details.
+In the previous ``state of the practice'' project, we measured 30 software
+projects for each domain, but the measures were relatively shallow.  With this
+re-boot we will still target about 30 software examples from each domain, but we
+will be collecting more data by adding more measures and experiments.  In
+keeping with the previous project, we still have the constraint that the work
+load for applying the methodology to a given domain needs to be feasible for a
+team as small as one individual, and for a time that is short, ideally 3 to 4
+months of part-time work per domain.
 
-In the previous state of the practice assessing exercise
-\citep{SmithEtAl2018_arXivGIS}, a series of questions and simple metrics was
-devised to measure the quality of the documentation and adherence to best
-software development practices.  The previous set of questions is provided
+To start with, the 
 \href{https://github.com/adamlazz/DomainX/blob/master/TemplateToGradeSCSoft.pdf}
-{here}. As described below, the new project critically assess the previous set of questions and
-revises as necessary.  In addition, the following data is also collected for each domain:
+{previous set of questions} were critically assessed and modified.  In addition,
+the following data is part of the new methodology:
 
@@
 \item Characterization of the functionality provided by the software in the
-  domain.  What services does each member of the domain provide?  This
-  information would be summarized in a commonality analysis document.
-\item Usability testing of each software in the domain.
-\item Collect empirical software engineering related data, such as the number of
-  files, number of lines of code, number of open issues,
-  number of closed issues, the number of lines of documentation (?). 
+  domain via a commonality analysis.
+\item Usability testing for a sample of software in the domain.
+\item Empirical software engineering related data, such as the number of
+  files, number of lines of code, percentage of issues that are closed, etc. 
 \end{itemize}
 
-One of the guidelines for successfully on-boarding new developers is to
-  ``Keep knowledge up to date and findable'' \citep{ShollerEtAl2019}.  The
-  selected measurements should be able to assist with judging whether a project
-  succeeds on this front.
+Unlike the previous measurement process, the new methodology will involves and
+engages a domain expert partner throughout.  We did not previously engage the
+domain expert with the rationale that we wished to exclude potential bias.
+However, this advantage is not worth not being able to evaluate the
+functionality/usability of the software.  Moreover, not having an expert makes
+publication more difficult, since there is no one to provide advice on how best
+to approach journals and publishers.
 
-The above data will be combined to rank the software in the domain.  The
-specifics are yet to be determined, but in the past the Analytic Hierarchy
-Process (AHP) \citep{Saaty1980} proved helpful in this.  In the past the AHP
-results were calculated by a grade out of 10 from each student evaluator for
-each quality for each measured product.  A better approach may be to have a
-group of people familiar with the software do a pairwise comparison between each
-product to come up with a number.  The number would be based on the data
-collected, but not in a numerical way.  The data collected would serve the same
-role as a decision support system.  It would provided insight, but human beings
-would do the final pairwise comparisons, and thus the final ranking.
+In the proposed methodology, the collected data will be combined to rank the
+software within each domain using the Analytic Hierarchy Process (AHP)
+\citep{Saaty1980}.  As in the past process, we will use AHP to develop a list of software
+ranked by quality.  However, in the new process we will not stop with this
+list.  The domain expert will be consulted to verify the ordering, and to
+discuss the decisions that led to the ranking.  The AHP process will be used to
+facilitate a conversation with the domain experts as a means to deepen our
+understanding of the software in the domain, and the needs of typical developers.
 
-Key to the success of this exercise will be to involve and engage a partner for
-each state of practice project.  In the previous effort we did not involve domain
-experts with the rationale of excluding potential bias, but this advantage is
-not worth not being able to evaluate the functionality/usability of the software.
-Moreover, not having an expert makes publication more difficult, since there is
-no one to advice on how best to approach journals and publishers.  The domain
-expert will be asked to help in the following ways:
-
-GitHub will be used to coordinate the work of the large team of people that will
+GitHub will be used to coordinate the work of the team of people that will
 be involved in this project.  In addition to the project record left on GitHub,
@@
 
-The data for each of the different subdomains within SCS will be standardized to
-facilitate a meta-analysis that combines all of the data.
-
-The scope of this document is to conduct a current state of practice analysis of
-software in a scientific computing domain for the purpose of understanding
-quality and assisting in future development. We observe product, artifact, and
-process quality in our analysis. We leave the assessment of scientific computing
+The scope of this methodology includes observations on product, artifact and
+process quality for research softwares. We leave the assessment of research
 software using performance benchmarks to other projects, such as the work of
-\cite{kaagstrom1998gemm}. The methodology presented here is part of an overall
-assessment of the impact of model driven engineering and code generation on the
-sustainability of scientific computing software. The research proposal for this
-assessment, which contains a more detailed description of this state of practice
-exercise, can be found
-\href{https://github.com/smiths/AIMSS/blob/master/OverallResearchProposal/ResearchProposal.pdf}
-{here}. The research questions that this exercise will answer are found in
-Section \ref{ResearchQuestions}. Section \ref{StepsAQDS} outlines the steps to
-be taken.
+\cite{kaagstrom1998gemm}.
 
@@
 \item The domain must have well-defined and stable theoretical underpinning.  A
-  good sign of this is the existance of standard textbooks, preferably
+  good sign of this is the existence of standard textbooks, preferably
   understandable by an upper year undergraduate student.
@@
 An AHP pairwise comparison between the short list packages with respect to
-usability and modifiability survey results is conducted, and then reviewed by the domain expert. The tool can be found here. The usability and modifiability survey can be found
+usability and modifiability survey results is conducted, and then reviewed by
+the domain expert. The tool can be found here. The usability and modifiability
+survey can be found
 \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Experiments.pdf}
-{here}. This step is intended to start a conversation about the relative quality of short-listed software packages.
+{here}. This step is intended to start a conversation about the relative quality
+of short-listed software packages.
 

### 2021-01-04 StateOfPractice/Methodology/Meeting Agenda with Domain Experts.tex
Deleted tasks asking for feedbacks in Sections 1.2, 1.3, 1.4, 2.1, 2.2, 2.4

@@
 meeting, and tasks will be assigned to domain experts to finish later. For
-example, Task \ref{task_overall_objective} in Section \ref{project_objectives}.
+example, Task \ref{task_usability_tests} in Section \ref{usability_tests}.
 
@@
 
-Do you have any suggestions? Feedback can be provided later in Task
-\ref{task_overall_objective}.
-
 \subsection{Research Proposal}
@@
 A brief introduction. \ad{To be written}
-
-Do you have any suggestions? Feedback can be provided later in Task
-\ref{task_research_proposal}.
 
@@
 
-Do you have any suggestions? Feedback can be provided later in Task
-\ref{task_research_questions}.
-
 \section{Qualities \& Measurements}
@@
 	Template}.
-
-Do you have any suggestions? Feedback can be provided later in Task
-\ref{task_measurement_template}.
 
@@
 	to Developers}.
-
-Do you have any suggestions? Feedback can be provided later in Task
-\ref{task_survey_short_list_projects}
 
@@
 The usability test can be found..
-
-Do you have any suggestions? Feedback can be provided later in Task
-\ref{modifiability_experiment}
 
@@
 
-\subsection{Review the Overall Objective}
-\label{task_overall_objective}
-GitHub Issue:
-\href{https://github.com/smiths/AIMSS/issues}{https://github.com/smiths/AIMSS/issues}
-\ad{To be created}
-
-\noindent Refer to Section \ref{project_objectives}
-
-\subsection{Review the Research Proposal}
-\label{task_research_proposal}
-GitHub Issue:
-\href{https://github.com/smiths/AIMSS/issues}{https://github.com/smiths/AIMSS/issues}
-\ad{To be created}
-
-\noindent Refer to Section \ref{research_proposal}
-
-\subsection{Review the Research Questions}
-\label{task_research_questions}
-GitHub Issue:
-\href{https://github.com/smiths/AIMSS/issues}{https://github.com/smiths/AIMSS/issues}
-\ad{To be created}
-
-\noindent Refer to Section \ref{research_questions}
-
-\subsection{Review the Measurement Template}
-\label{task_measurement_template}
-GitHub Issue:
-\href{https://github.com/smiths/AIMSS/issues}{https://github.com/smiths/AIMSS/issues}
-\ad{To be created}
-
-\noindent Refer to Section \ref{measurement_template}
-
-\subsection{Review the Survey for Short List Projects}
-\label{task_survey_short_list_projects}
-GitHub Issue:
-\href{https://github.com/smiths/AIMSS/issues}{https://github.com/smiths/AIMSS/issues}
-\ad{To be created}
-
-\noindent Refer to Section \ref{survey_short_list_projects}
-
 \subsection{Review the Usability Tests}
@@
 \noindent Refer to Section \ref{usability_tests}
-
-\subsection{Review the Modifiability Experiment}
-\label{task_modifiability_experiment}
-GitHub Issue:
-\href{https://github.com/smiths/AIMSS/issues}{https://github.com/smiths/AIMSS/issues}
-\pmi{To be created}
-
-\noindent Refer to Section \ref{modifiability_experiment}
 

### 2021-01-04 StateOfPractice/Methodology/Meeting Agenda with Domain Experts.tex
Clarified 3 times of interaction with the domain experts

@@
 
-\subsection{About this Meeting}
+\subsection{3 Steps for Domain Experts}
+The domain experts are expected to participate in the following 3 steps:
+\begin{itemize}
+\item Step 1 - introductory meeting. This is our current meeting, which will
+lasts for about one hour.
+\item Step 2 - measuring. Usability experiments, modifiability experiments, and
+developer interviews are included in this step. More details will be discussed
+later in Section \ref{qualities_Measurements}.
+\item Step 3 - concluding meeting. Our second and the last meeting, aiming to be
+one hour in duration. Results of ranking will be discussed.
+\end{itemize}
+
+\subsection{About this Meeting (Introductory Meeting)}
 Presenter: \ad{TBD}
@@
 \section{Qualities \& Measurements}
+\label{qualities_Measurements}
 

### 2021-01-04 StateOfPractice/Methodology/Meeting Agenda with Domain Experts.tex
modified "Survey for Short List Projects"

@@
 After the first round of measurement, we would like to select a short list of
-projects (about 5 of them for each domain) and directly ask questions to the
+projects (about 5 to 10 of them for each domain) and directly ask questions to the
 developers.
 
-We would like to do the selection based on the results of the first round
-measurement, such as selecting the top 5 ones with the highest scores. However, we
-can also include 1 or 2 most recommended ones from you. What do you think of
-this selection process?
+Considering the fact that some development teams may ignore our requests for
+this interview, We would like to send invitations to all of the teams developing
+the software packages in the list. We estimate that there will be a smaller
+proportion of them accepting our request. What do you think of
+this process?
 
 The survey questions to the developers can be found at
-\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Ao-Notes/Questions%20to%20Developers.pdf}{Questions
+\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Questions%20to%20Developers.pdf}{Questions
 	to Developers}.
+
+We also would like to ask the domain experts a modified version of the above
+questions. The purpose will be to learn how our domain experts and their
+students create software. If we time it right, the interview with the domain
+experts is also a chance to do a dry run with our questions.
 

### 2021-01-05 StateOfPractice/Methodology/Methodology.tex
added feedbacks after reviewing Methodology

@@
 \item What tools are used by current software packages?
+\ad{The ``tools'' here seems ambiguous. Does it mean the tools used by
+developers for developing and maintaining the packages, or the dependencies
+(such as libraries) used by software?}
 \item What principles, processes, and methodologies are used in the development
@@
 \item With short list:
+\ad{From my understanding, we'll have two short lists. One is for the top
+software packages for deeper AHP ranking, the other contains the software
+packages whose developers are willing to take the survey. Should we separate
+them here?}
 \begin{enumerate}
@@
   ideally be available, which implies a preference for GitHub repositories.
+\ad{For empirical measures, even if it's not a GitHub repo, a git repo is better
+than a Mercurial or Subversion one. Should we describe our preference in such
+order - GitHub > other git repos (in GitLab, BitBucket, SourceForge, etc.) >
+other repos?}
 \item The software cannot be marked as incomplete or in an initial development
@@
 Section \ref{rawdata} orients the reader on what raw data is collected.  Some of this data can be
-observed from Git repository metrics. The rest can be collected using freeware
+observed from Git\ad{It's better to be ``GitHub'' here. GitLab, BitBucket and
+SourceForge also mainly have git repos, but not providing all the metrics.}
+repository metrics. The rest can be collected using freeware
 tools. \href{https://github.com/tomgi/git_stats}{GitStats} is used to measure

### 2021-01-09 StateOfPractice/Methodology/Methodology.tex
Ao suggested updates to methodology document

--- 2020-12-18 StateOfPractice/Methodology/Methodology.tex
@@
 \item What artifacts are present in current software packages? 
-\item What tools are used by current software packages?
+\item What tools (development, dependencies, project management) are used by current software packages?
 \item What principles, processes, and methodologies are used in the development
@@
 \item Measure using measurement template. (Section~\ref{SecShallowMeasure})
+\item Survey developers
+(\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Questions
+	to Developers.pdf}{Questions to Developers})
 \item Use AHP process to rank the software packages. (Section~\ref{SecAHP})
@@
 \begin{enumerate}
-\item Survey developers
-  (\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Questions
-    to Developers.pdf}{Questions to Developers})
 \item Usability experiments
@@
 \item The empirical measures listed in Section \ref{SecEmpiricalMeasures} should
-  ideally be available, which implies a preference for GitHub repositories.
+  ideally be available, which implies a preference for GitHub-style repositories.
 \item The software cannot be marked as incomplete or in an initial development
@@
 
-Section \ref{rawdata} orients the reader on what raw data is collected.  Some of this data can be
-observed from Git repository metrics. The rest can be collected using freeware
-tools. \href{https://github.com/tomgi/git_stats}{GitStats} is used to measure
+Section \ref{rawdata} orients the reader on what raw data is collected.
+Some of this data can be observed from GitHub repository metrics.
+The rest can be collected using freeware tools.
+\href{https://github.com/tomgi/git_stats}{GitStats} is used to measure
 the number of binary files as well as the number of added and deleted lines in a
@@
 An AHP pairwise comparison between the short list packages with respect to
-usability and modifiability survey results is conducted, and then reviewed by
-the domain expert. The tool can be found here. The usability and modifiability
+usability and modifiability experiment results is conducted, and then reviewed by the domain expert. The tool can be found here. The usability and modifiability
 survey can be found

### 2021-01-16 StateOfPractice/Methodology/Interactions with Domain Experts.tex
updated interactions with domain experts

--- 2021-01-04 StateOfPractice/Methodology/Meeting Agenda with Domain Experts.tex
@@
 
-\title{Meeting Agenda with Domain Experts} 
+\title{Interactions with Domain Experts} 
 \author{Ao Dong}
@@
 
-\section{Introduction}
-Presenter: \ad{TBD}
-
-A brief introduction to the project. \ad{To be written}
-
-\subsection{3 Steps for Domain Experts}
-The domain experts are expected to participate in the following 3 steps:
+The purpose of the document is to describe the interactions with the domain
+experts, which will be through 3 steps as follows,
 \begin{itemize}
-\item Step 1 - introductory meeting. This is our current meeting, which will
-lasts for about one hour.
-\item Step 2 - measuring. Usability experiments, modifiability experiments, and
-developer interviews are included in this step. More details will be discussed
-later in Section \ref{qualities_Measurements}.
-\item Step 3 - concluding meeting. Our second and the last meeting, aiming to be
+\item Step 1 - Introductory Meeting in Section \ref{introductory_meeting}. In
+this meeting, we will be focusing on introducing the background and objectives
+of the project and vetting some details and plans.
+\item Step 2 - Measuring in Section \ref{qualities_measurements}. Vetting the
+Domain Analysis, usability experiments, modifiability experiments, and
+developer interviews are included in this step.
+\item Step 3 - Concluding Meeting in Section \ref{concluding_meeting}. Our
+second and the last meeting, aiming to be
 one hour in duration. Results of ranking will be discussed.
@@
 
-\subsection{About this Meeting (Introductory Meeting)}
+\section{Introductory Meeting}
+\label{introductory_meeting}
+
+\subsection{Prerequisites}
+This section contains the tasks of preparing for the introductory meeting. These
+tasks need to be finished before the meeting.
+
+\subsubsection{Top 10 Software List}
+\label{top_10_list}
+
+The domain experts need to prepare the top 10 software packages of their choice.
+Feedback can be provided in Task \ref{task_top_10_list}.
+
+\subsection{About this Meeting}
 Presenter: \ad{TBD}
 
-The meeting will be roughly one hour. We will follow this document and go
-through all the steps.
+The meeting will be roughly one hour. We will go
+through all the steps in Section \ref{introductory_meeting}.
 
@@
 
-\subsection{Project Objectives}
-\label{project_objectives}
+\subsection{Overall Objective}
+\label{overall_objective}
 Presenter: Dr.\ Spencer Smith and Dr.\ Jacques Carette \ad{TBD}
 
-The overall objectives are documented in the section Overall Objective of
+As described in the document 
 \href{https://github.com/smiths/AIMSS/blob/master/OverallResearchProposal/ObjectivesAndResearchQuestions.pdf}{Objectives
-And Research Questions}.
+And Research Questions}, the overall objective asks ``What is needed to produce
+\{software + artifacts\} sustainably?'' The definitions of \emph{software},
+\emph{artifacts} and \emph{sustainably} can be found in the same document. The
+term \emph{softifacts} will be used to refer the combination of software and
+artifacts.
 
-A brief introduction. \ad{To be written}
+According to the above document, to answer the above question, we will need to
+find answers for these two questions,
+\begin{enumerate}
+\item ``What softifacts, in addition to the code, are needed?''
+\item ``What software engineering principles, processes and methodologies should
+be employed?''
+\end{enumerate}
 
@@
 \href{https://github.com/smiths/AIMSS/blob/master/OverallResearchProposal/ObjectivesAndResearchQuestions.pdf}{Objectives
-And Research Questions}.
-
-A brief introduction. \ad{To be written}
+And Research Questions}. Guided by overall objective in Section
+\ref{overall_objective}, these questions focus on researching on ``i) knowledge,
+ii) principles, processes
+and methodologies, iii) software qualities and iv) the necessary investment of
+time and energy''.
 
@@
 
-\section{Qualities \& Measurements}
-\label{qualities_Measurements}
+\subsection{Vetting the List of Candidate Software}
+\label{software_list}
 
-\subsection{Measurement Template}
-\label{measurement_template}
-Presenter: \ad{TBD}
-
-The template for software quality measurements can be found at
-\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Combined_MeasurementTemplate_EmpiricalMeasures.xlsx}{Measurement
-	Template}.
-
-\subsection{Survey for Short List Projects}
-\label{survey_short_list_projects}
-Presenter: \ad{TBD}
-
-After the first round of measurement, we would like to select a short list of
-projects (about 5 to 10 of them for each domain) and directly ask questions to the
-developers.
-
-Considering the fact that some development teams may ignore our requests for
-this interview, We would like to send invitations to all of the teams developing
-the software packages in the list. We estimate that there will be a smaller
-proportion of them accepting our request. What do you think of
-this process?
-
-The survey questions to the developers can be found at
-\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Questions%20to%20Developers.pdf}{Questions
-	to Developers}.
-
-We also would like to ask the domain experts a modified version of the above
-questions. The purpose will be to learn how our domain experts and their
-students create software. If we time it right, the interview with the domain
-experts is also a chance to do a dry run with our questions.
-
-\subsection{Usability Tests}
-\label{usability_tests}
-Presenter: Olu \ad{TBD}
-
-A brief introduction. \ad{To be written}
-
-The usability test can be found at ... \ad{TBD}
-
-Do you have any suggestions? Feedback can be provided later in Task
-\ref{task_usability_tests}
-
-\subsection{Modifiability Experiment}
-\label{modifiability_experiment}
-Presenter:
-
-A brief introduction.
-
-The usability test can be found..
-
-\section{Software List}
-Presenter: Ao or Peter \ad{TBD}
-
-In this section, we would like to discuss the list of software packages and the
-process of creating and shortening it.
-
-\subsection{Top 10 List}
-\label{top_10_list}
-What will be the top 10 software of your choice? Feedback can be provided later
-in Task \ref{task_top_10_list}.
-
-\subsection{Software List}
-\label{software_list}
 The software list is
@@
 
-\subsection{Domain Analysis}
-\label{domain_analysis}
-The domain analysis can be found in Section 3.2 of
-\href{https://github.com/Ao99/MISEG/blob/master/docs/SRS/SRS.pdf}{SRS for
-Medical Imaging Software} or
-\href{https://github.com/peter-michalski/LatticeBoltzmannSolvers/blob/master/docs/SRS/CA.pdf}{CA
-of Lattice Boltzmann Solvers}.
-
-What do you think of our analysis?  \wss{This is too big a question for the
-  domain experts to answer in a reasonable amount of time.  Moreover, without
-  more guidance their feedback is likely to be shallow.  I think we are better
-  off with a task based inspection, using issues assigned via a GitHub project
-  page.  We also will likely be more successful asking for reviews from grad
-  students, rather than from faculty members.  We should also consider
-  summarizing the commonalities, variabilities and parameters of variation in
-  tables, or possibly graphs.}
-
-\subsection{First AHP Ranking}
-\label{first_AHP}
-Presenter:
-
-A brief introduction.
-
-Do you have any feedback?
-
-\subsection{Final AHP Ranking}
-\label{final_AHP}
-Presenter:
-
-A brief introduction.
-
-Do you have any feedback?
-
-\section{Publication}
+\subsection{Publication}
 Presenter: Dr.\ Spencer Smith and Dr.\ Jacques Carette \ad{TBD}
@@
 
-\section{Schedule \& Team}
+\subsection{Schedule \& Team}
 Presenter: \ad{TBD}
@@
 \end{itemize}
+
+\subsection{Introducing the Next Steps}
+
+\section{Qualities \& Measurements}
+\label{qualities_measurements}
+
+\subsection{Vetting the Domain Analysis}
+\label{domain_analysis}
+Task leader: Ao and Peter
+
+The domain analysis can be found in Section 3.2 of
+\href{https://github.com/Ao99/MISEG/blob/master/docs/SRS/SRS.pdf}{SRS for
+Medical Imaging Software} or
+\href{https://github.com/peter-michalski/LatticeBoltzmannSolvers/blob/master/docs/SRS/CA.pdf}{CA
+of Lattice Boltzmann Solvers}.
+
+Please review the Figure 1 in \href{https://github.com/Ao99/MISEG/blob/master/docs/SRS/SRS.pdf}{SRS for
+Medical Imaging Software} or Figure/Table ? \ad{TBD. It can be a summary of the
+commonalities, variabilities and parameters of variation in tables, or possibly
+graphs.} in
+\href{https://github.com/peter-michalski/LatticeBoltzmannSolvers/blob/master/docs/SRS/CA.pdf}{CA
+of Lattice Boltzmann Solvers}.
+What do you think of our analysis for the scope of the family, commonalities, and variabilities  for this domain?
+
+Our preference is to have feedback from several graduate students. Feedback can be provided in Task \ref{task_domain_analysis}.
+
+\subsection{Measurement Template}
+\label{measurement_template}
+Task leader: Ao or Peter
+
+The template for software quality measurements can be found at
+\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Combined_MeasurementTemplate_EmpiricalMeasures.xlsx}{Measurement
+	Template}.
+
+As described in Task \ref{task_measurement_template}, domain experts are
+expected to take a look at this template. Feedback are welcome but not required.
+
+\subsection{Survey for Short List Projects}
+\label{survey_short_list_projects}
+Task leader: Ao or Peter
+
+After the first round of measurement, we would like to select a short list of
+projects (about 5 to 10 of them for each domain) and directly ask questions to the
+developers.
+
+Considering the fact that some development teams may ignore our requests for
+this interview, We would like to send invitations to all of the teams developing
+the software packages in the list. We estimate that there will be a smaller
+proportion of them accepting our request. What do you think of
+this process? Feedback can be provided in Task \ref{task_review_short_list}.
+
+The survey questions to the developers can be found at
+\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Questions%20to%20Developers.pdf}{Questions
+	to Developers}.
+
+We also would like to ask the domain experts a modified version of the above
+questions. The purpose will be to learn how our domain experts and their
+students create software. If we time it right, the interview with the domain
+experts is also a chance to do a dry run with our questions. We can use Task
+\ref{task_short_list_survey} to manage the progress of the survey with domain
+experts.
+
+\subsection{Usability Tests}
+\label{usability_tests}
+Task leader: Olu
+
+A brief introduction. \ad{To be written}
+
+The usability test can be found at ... \ad{TBD}
+
+Feedback can be provided later in Task
+\ref{task_usability_tests}. We can also use the same task to manage the progress of the tests.
+
+\subsection{Modifiability Experiment}
+\label{modifiability_experiment}
+Task leader: \ad{TBD}
+
+A brief introduction. \ad{To be written}
+
+The modifiability experiment can be found at.. \ad{TBD}
+
+Feedback can be provided later in Task
+\ref{task_modifiability_experiment}. We can also use the same task to manage the
+progress of the tests.
+
+\section{Concluding Meeting}
+\label{concluding_meeting}
+
+\subsection{First AHP Ranking}
+\label{first_AHP}
+Presenter: \ad{TBD}
+
+A brief introduction about the ranking results.
+
+Do you have any feedback? Feedback can be provided in Task \ref{task_first_ranking}.
+
+\subsection{Final AHP Ranking}
+\label{final_AHP}
+Presenter: \ad{TBD}
+
+A brief introduction about the ranking results for the short list software packages.
+
+Do you have any feedback? Feedback can be provided in Task \ref{task_final_ranking}.
 
@@
 
-\subsection{Review the Usability Tests}
-\label{task_usability_tests}
-GitHub Issue:
-\href{https://github.com/smiths/AIMSS/issues}{https://github.com/smiths/AIMSS/issues}
-\ad{To be created}
-
-\noindent Refer to Section \ref{usability_tests}
-
 \subsection{Create a Top 10 List}
@@
 
-\subsection{Review the Software List}
+\subsection{Vet the Software List}
 \label{task_software_list}
@@
 
-\subsection{Review the Domain Analysis}
+\subsection{Vet the Domain Analysis}
+\label{task_domain_analysis}
 GitHub Issue:
@@
 
-\subsection{Review the AHP Ranking and Short List of Software Packages}
+\subsection{Read the Measurement Template}
+\label{task_measurement_template}
+GitHub Issue:
+\href{https://github.com/smiths/AIMSS/issues}{https://github.com/smiths/AIMSS/issues}
+\ad{To be created}
+
+\noindent Refer to Section \ref{measurement_template}
+
+\subsection{Review the Process of Creating a Short List}
+\label{task_review_short_list}
+GitHub Issue:
+\href{https://github.com/smiths/AIMSS/issues}{https://github.com/smiths/AIMSS/issues}
+\ad{To be created}
+
+\noindent Refer to Section \ref{survey_short_list_projects}
+
+\subsection{Take the Survey for the Short List Projects}
+\label{task_short_list_survey}
+GitHub Issue:
+\href{https://github.com/smiths/AIMSS/issues}{https://github.com/smiths/AIMSS/issues}
+\ad{To be created}
+
+\noindent Refer to Section \ref{survey_short_list_projects}
+
+\subsection{Participate in the Usability Tests}
+\label{task_usability_tests}
+GitHub Issue:
+\href{https://github.com/smiths/AIMSS/issues}{https://github.com/smiths/AIMSS/issues}
+\ad{To be created}
+
+\noindent Refer to Section \ref{usability_tests}
+
+\subsection{Participate in the Modifiability Experiment}
+\label{task_modifiability_experiment}
+GitHub Issue:
+\href{https://github.com/smiths/AIMSS/issues}{https://github.com/smiths/AIMSS/issues}
+\ad{To be created}
+
+\noindent Refer to Section \ref{modifiability_experiment}
+
+\subsection{Review the First AHP Ranking}
+\label{task_first_ranking}
 GitHub Issue:
@@
 
-\subsection{Comment on Final Short List Ranking}
+\subsection{Review the Final Short List Ranking}
 \label{task_final_ranking}

### 2021-01-16 StateOfPractice/Methodology/MeetingPresentations/Presentation for the Introductory Meeting.tex
created presentation for introductory meeting

%\documentclass[handout]{beamer} 
\documentclass[t,12pt,numbers,fleqn]{beamer}
%\documentclass[ignorenonframetext]{beamer}

\newif\ifquestions
%\questionstrue
\questionsfalse

\usepackage{pgfpages} 
\usepackage{hyperref}
\hypersetup{colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    filecolor=blue,
    urlcolor=blue,
    unicode=false}
\urlstyle{same}

\usepackage{booktabs}
\usepackage{bibentry}
\usepackage[round, authoryear]{natbib}
\bibliographystyle{plainnat}

%\usetheme{Iimenau}

\useoutertheme{split} %so the footline can be seen, without needing pgfpages

%\pgfpagesuselayout{resize to}[letterpaper,border shrink=5mm,landscape]  %if this is uncommented, the hyperref links do not work

\mode<presentation>{}

\input{./def-beamer}

\newcommand{\topic}{Introductory Meeting}

\input{./titlepage}

\begin{document}

%\nobibliography{../../../CommonFiles/ResearchProposal}

\input{./footline}

%% Comments
\newif\ifcomments\commentstrue

\ifcomments
\newcommand{\authornote}[3]{\textcolor{#1}{[#3 ---#2]}}
\newcommand{\todo}[1]{\textcolor{red}{[TODO: #1]}}
\else
\newcommand{\authornote}[3]{}
\newcommand{\todo}[1]{}
\fi

\newcommand{\wss}[1]{\authornote{blue}{SS}{#1}} %Spencer Smith
\newcommand{\jc}[1]{\authornote{red}{JC}{#1}} %Jacques Carette
\newcommand{\oo}[1]{\authornote{magenta}{OO}{#1}} %Olu Owojaiye
\newcommand{\pmi}[1]{\authornote{green}{PM}{#1}} %Peter Michalski
\newcommand{\ad}[1]{\authornote{brown}{AD}{#1}} %Ao Dong

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}
\frametitle{Interactions with Domain Experts}

\bi
\item Introductory Meeting
\item Measuring
\item Concluding Meeting
\ei

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}
\frametitle{About this Meeting}

\bi
\item Introductory Meeting
\bi
\item Overall Objective
\item Research Proposal
\item Research Questions
\item Vetting the List of Candidate Software
\item Publication
\item Schedule \& Team
\item Introducing the Next Steps
\ei
\item Measuring
\item Concluding Meeting
\ei

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}
\frametitle{Overall Objective}

``What is needed to produce \{software + artifacts\} sustainably?''

\href{https://github.com/smiths/AIMSS/blob/master/OverallResearchProposal/ObjectivesAndResearchQuestions.pdf}{Objectives
And Research Questions}

\bi
\item Definitions
\bi
\item software
\item artifacts
\item \{software + artifacts\} - softifacts
\item sustainable softifacts
\ei
\item What softifacts, in addition to the code, are needed?
\item What software engineering principles, processes and methodologies should be employed?
\ei

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}
\frametitle{Research Proposal}

\href{https://github.com/smiths/AIMSS/blob/master/OverallResearchProposal/ResearchProposal.pdf}{Research
Proposal}

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}
\frametitle{Research Questions}

\href{https://github.com/smiths/AIMSS/blob/master/OverallResearchProposal/ObjectivesAndResearchQuestions.pdf}{Objectives
And Research Questions}

\bi
\item To investigate
\bi
\item knowledge
\item principles, processes and methodologies
\item software qualities
\item the necessary investment of time and energy
\ei
\item What are the ``pain points'' for
developers working on SCS projects?
\item What aspects of the existing processes, methodologies and tools do they consider could potentially be improved?
\ei

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}
\frametitle{Vetting the List of Candidate Software}

\href{https://docs.google.com/spreadsheets/d/122wU0v3ZtvDcqy8C4zKJ89kU-8fXAbo3Mzn6vcVXOi0/edit?usp=sharing}{Medical Imaging Software List}

Task:
\bi
\item What do you think of the list of the selected software?
\item Is there any software you would like to remove from or add to the list?
\ei

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}
\frametitle{Publication}

\bi
\item Where should we publish this paper?
\item Who are our targeted readers?
\ei

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}
\frametitle{Schedule \& Team}

\bi
\item Project timeline
\item Schedule for concluding meeting
\item ... extra members from your team, each working ... hours per week for ...
weeks.
\item Any other experts?
\ei

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}
\frametitle{Introducing the Next Steps}

\bi
\item Introductory Meeting
\item Measuring
\bi
\item Vetting the Domain Analysis
\item Measurement Template
\item Survey for Short List Projects
\item Usability Tests
\item Modifiability Experiment
\ei
\item Concluding Meeting
\bi
\item First AHP Ranking
\item Final AHP Ranking
\ei
\ei

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{document}

### 2021-01-16 StateOfPractice/Methodology/MeetingPresentations/titlepage.tex
created presentation for introductory meeting

%Title page for slides

\newcommand{\presenters}{Dr.~Spencer Smith, Dr.~Jacques Carette,\\Olu Owojaiye, Peter Michalski, Ao Dong} %use to switch presenters
\newcommand{\projectTitle}{Assessing the State of the Practice for Domain X}
\newcommand{\department}{Department of Computing and Software}
\newcommand{\instituteName}{Faculty of Engineering, McMaster University}

%\setbeamerfont{structure}{series=\bfseries}
%\usefonttheme[stillsansseriftext,stillsansserifmath]{serif}
\setbeamertemplate{navigation symbols}{} 
\setbeamertemplate{itemize item}[ball]

\title{
  {\normalsize \bf 
    \borange{\projectTitle}}\\[2ex]
  {\Large \bf \topic}}

\author{\presenters}

\institute{\instituteName}

\date{
\today
\bc
  \includegraphics[scale = 0.2, keepaspectratio]
  {./mcmaster-logo-full-color.jpg}
\ec
}

\renewcommand{\borange}[1] %orange is too hard to read
{
   \bred{#1}
}

### 2021-01-17 StateOfPractice/Methodology/Interactions with Domain Experts.tex
updated Research Proposal

@@
 
-A brief introduction. \ad{To be written}
+\subsubsection{State of the Practice}
+
+MEng students will perform assessments to the state of the practice in several
+Scientific Computing Software (SCS) domains. Domain analysis will be conducted
+(related to Section \ref{domain_analysis}). About 20 to 30 software projects
+will be chosen for the measurements, and the software list is in Section
+\ref{software_list}. The questions and metrics (including empirical
+measurements) to assess the projects are documented in the Section
+\ref{measurement_template}.
+
+Additionally, we will also perform Usability Tests (Section
+\ref{usability_tests}) and Modifiability Experiment (Section
+\ref{modifiability_experiment})
+
+To compare each software quality and the overall quality, we will use the
+Analytic Hierarchy Process (AHP) to rank software projects in each domain. The
+first ranking will be for all projects, and the final ranking will be for the
+short list projects (more details in Section \ref{survey_short_list_projects}).
+The ranking results will be the topic for our last meeting - the concluding
+meeting in Section \ref{concluding_meeting}.
+
+\subsubsection{Impact of MDE on the Sustainability of SCS}
 

### 2021-01-29 StateOfPractice/Methodology/Experiments.tex
added isability intro

@@
 \subsection{Usability Experimental Procedure}
+The purpose of this experiment is to assess the usability of each software package and collect qualitative data.
+An initial step is required to define a set of tasks for the experiment.
+The guideline below would assist domain experts in defining a list usability tasks for each domain.
+This list will then be used to evaluate the usability of each selected software package.
+It is expected that the list would allow experimenters perform tasks in  different aspects of the software package and provide their evaluation using a standardized usability questionnaire ( \url{https://www.usabilitest.com/sus-pdf-generator}- 20-29 , \url{https://uiuxtrend.com/pssuq-post-study-system-usability-questionnaire/} - PSSUQ) - (we need to decide on the questionnaire to use eventually)
 

### 2021-01-29 StateOfPractice/Methodology/Experiments.tex
small changes

@@
 \section{User Experiments}
-\subsection{Usability Experimental Procedure}
+\subsection{Usability Experiment}
 The purpose of this experiment is to assess the usability of each software package and collect qualitative data.
@@
 
-\subsection {Procedure}
+\subsubsection{Task selection criteria for domain experts}
+The following criteria will be considered by the domain expert when defining the tasks for each software package.
+
 \begin {enumerate}
-%\item Select a list of projects
-%\item Write research questions
-%\item Prepare logistic details(location, time table, recording setup, moderators etc)
-%\item Prepare survey questions
-%\item Design the tasks for the study subjects to perform (tasks can be defined based on user category, at least one task to modify the software)
-%\item Select participants
-
-\item Survey participants to collect pre-experiment data
-\item Participants perform tasks
-\item Observe the study subjects (take notes, record sessions(OBS screen recorder), watch out for body languages and verbal cues)
-\item Survey the study subjects to collect feedback (post experiment interview)
-\item Prepare experiment report
-\item Perform pairwise comparison analysis
-\item Prepare analysis report
-\end {enumerate}
-
-%\subsection{Nielsen’s Heuristics}
-%\begin{enumerate}
-%	\item Consistency: Check if standards and conventions in product design are followed
-% \begin{itemize}
-% 	\item Sequences of actions (skill acquisition).
-% 	\item Color (categorization).
-% 	\item Layout and position (spatial consistency).
-% 	\item Font, capitalization (levels of organization).
-% 	\item Terminology (delete, del, remove, rm) and language (words, phrases).
-% 	\item Standards (e.g.,	blue underlined text for unvisited hyper links).
-% \end{itemize}
-%\item Visibility: Users should be informed about what is going on with the system through appropriate feedback and display of information e.g. 
-% \begin{itemize}
-% \item What is the current state of the system?
-% \item What can be done at current state?
-% \item Where can users go?
-% \item What change is made after an action?
-% \end{itemize}
-%\item Match between system and world. The image of the system perceived by users should match the model the users have about the system. 
-%\begin {itemize}
-%\item User model matches system image.
-%\item Actions provided by the system should match actions performed by users.
-%\item Objects on the system should match objects of the task.
-%\end {itemize}
-%\item Minimalist: Any extraneous information is a distraction and a slow-down. 
-%\begin{itemize}
-%\item Less is more.
-%\item Simple is not equivalent to abstract and general.
-%\item Simple is efficient.
-%\item  Progressive levels of detail.
-%\end{itemize}
-%\item Minimize memory load. Users should not be required to memorize a lot of information to carry out tasks. Memory load reduces users’ capacity to carry out the main tasks. 
-%\begin{itemize}
-%\item Recognition vs. recall (e.g., menu vs. commands).
-%\item Externalize information through visualization.
-%\item Perceptual procedures.
-%\item Hierarchical structure.
-%\item Default values.
-%\item Concrete examples (DD/MM/YY, e.g., 10/20/99).
-%\item Generic rules and actions (e.g., drag objects).
-%\item Intuitive procedure
-%\end{itemize}
-%\item Informative feedback. Users should be given prompt and informative feedback about their actions. 
-%\begin{itemize}
-%\item Information that can be directly perceived, interpreted, and evaluated.
-%\item Levels of feedback (novice and expert).
-%\item Concrete and specific, not abstract and general.
-%\item Response time:\\
-%-0.1 s for instantaneously reacting;\\
-%-1.0 s for uninterrupted flow of thought;\\
-%-10 s for the limit of attention.
-%\end{itemize}
-%%\item Flexibility and efficiency. Users always learn and users are always different. Give users the flexibility of creating customization and shortcuts to accelerate their performance. 
-%%\begin{itemize}
-%%	\item Shortcuts for experienced users.
-%%	\item Shortcuts or macros for frequently used operations.
-%%	\item Skill acquisition through chunking.
-%%	\item Examples:
-%%	- Abbreviations, function keys, hot keys, command keys, macros, aliases, templates, type-ahead, bookmarks, hot links, history, default values, etc.
-%%\end{itemize}
-%\item Good error messages. 
-%\begin {itemize}
-%\item The messages should be informative enough such that users can understand the nature of errors, learn from errors, and recover from errors. 
-%\item Phrased in clear language, avoid obscure codes. Example of obscure code: “system crashed, error code 147.”
-%\item Precise, not vague or general. Example of general comment: “Cannot open document.”
-%\item Constructive.
-%\item Polite. Examples of impolite message: “illegal user action,” “job aborted,” “system was crashed,” “fatal error,” etc.
-%\end {itemize}
-%\item Prevent errors. It is always better to design interfaces that prevent errors from happening in the first place.
-%\begin {itemize} 
-%\item Interfaces that make errors impossible.
-%\item Avoid modes (e.g., vi, text wrap). Or use informative feedback, e.g., different sounds.
-%\item Execution error vs. evaluation error.
-%\item Various types of slips and mistakes.
-%\end {itemize}
-%\item Clear closure. Every task has a beginning and an end. Users should be clearly notified about the completion of a task. 
-%\begin {itemize}
-%\item Clear beginning, middle, and end.
-%\item Complete 7-stages of actions.
-%\item Clear feedback to indicate goals are achieved and current stacks of goals can be released. Examples of good closures include many dialogues.
-%\end {itemize}
-%\item Reversible actions. 
-%\begin {itemize}
-%\item Users should be allowed to recover from errors. Reversible actions also encourage exploratory learning. 
-%\item At different levels: a single action, a subtask, or a complete task.
-%\item Multiple steps.
-%\item Encourage exploratory learning.
-%\item Prevent serious errors.
-%\end {itemize}
-%\item Use users’ language. The language should be always presented in a form understandable by the intended users.
-%\begin{itemize}
-%\item Use standard meanings of words.
-%\item Specialized language for specialized group.
-%\item User defined aliases.
-%\item Users’ perspective. Example: “we have bought four tickets for you” (bad) vs. “you bought four tickets” (good).
-%\end {itemize}
-%\item Help and documentation. Always provide help when needed. 
-%\begin {itemize}
-%\item Context-sensitive help.
-%\item Four types of help.\\
-%-task-oriented;\\
-%-alphabetically ordered;\\
-%-semantically organized;\\
-%-search.
-%\item Help embedded in contents.
-%\end {itemize}
-%\item Users in control. Do not give users that impression that they are controlled by the systems. 
-%\begin {itemize}
-%\item Users are initiators of actors, not responders to actions.
-%\item Avoid surprising actions, unexpected outcomes, tedious sequences of actions, etc.
-%\end {itemize}
-%\end{enumerate}
-
-\subsection{Task selection criteria}
-**The task selection will be determined with the aid of the domain experts attached to any of the selected projects.\\ 
-**The domain experts will be asked to consider the below criteria when defining a task. \\
-**Domain experts will also be asked to identify what background knowledge is necessary for the suggested tasks - Novice, Intermediate, Advanced
-\begin {enumerate}
-\item Collectively all tasks should not take no more than 2 hours.
-\item Selected tasks should reflect common use cases of  the software.
-\item Include tasks that require a set of sequential or hierarchical steps to be completed
+\item Tasks should be executable for subjects with novice to intermediate experience.
+\item All tasks should take no more than one hour.
+\item Tasks should include the basic/common use cases of the software package.
+\item Include tasks that require sequential or hierarchical steps for completion
 
@@
 
-\subsection{Usability Questionnaire}
-Two sources of standardized usability questionnaire we could use.
+\subsubsection {Procedure}
+\begin {enumerate}
 
-- \url{https://www.usabilitest.com/sus-pdf-generator}- 20-29 - SUS.\\
-- \url{https://uiuxtrend.com/pssuq-post-study-system-usability-questionnaire/} - PSSUQ
+\item Survey participants to collect pre-experiment data (background, experience of subjects)
+\item Participants perform tasks based on task defined by domain experts.
+\item Observe the study subjects (take notes, record sessions(OBS screen recorder), watch out for body languages and verbal cues)
+\item Survey the study subjects to collect feedback (post-experiment interview)
+\item Prepare a summary report of experiment
+
+\end {enumerate}
 

### 2021-02-18 StateOfPractice/Methodology/Interactions with Domain Experts.tex
Some edits to the interaction with domain experts document

@@
 The purpose of the document is to describe the interactions with the domain
-experts, which will be through 3 steps as follows,
+experts.  For each domain that is measured, a domain expert is a member of the
+assessment team.  Their role is to provide their expertise on the domain and its
+existing software.   The domain expert will typically be a faculty member.
+Ideally their graduate students will also assist with the measurement process.
+
+The interactions with the domain experts and their graduate students are divided
+into 3 stages:
+
 \begin{itemize}
-\item Step 1 - Introductory Meeting in Section \ref{introductory_meeting}. In
+\item Stage 1 - Introductory Meeting in Section \ref{introductory_meeting}. In
 this meeting, we will be focusing on introducing the background and objectives
-of the project and vetting some details and plans.
-\item Step 2 - Measuring in Section \ref{qualities_measurements}. Vetting the
-Domain Analysis, usability experiments, modifiability experiments, and
-developer interviews are included in this step.
-\item Step 3 - Concluding Meeting in Section \ref{concluding_meeting}. Our
-second and the last meeting, aiming to be
-one hour in duration. Results of ranking will be discussed.
+of the project and vetting the initial software list.  We will also discuss the
+resources that the domain expert can make available for Stage 2.
+\item Stage 2 - This stage isn't isolated to a single meeting, but is rather a
+  series of measurement and assessment tasks, as described in Section
+  \ref{qualities_measurements}.  During this stage the domain expert and their
+  students will help with vetting the Domain Analysis, usability
+  experiments, and modifiability experiments.  We will also interview the domain
+  expert in this step, using similar questions to those that we are asking the
+  developers.  (By developers, we mean the programmers that have created the
+  programs in the domain that are being measured.)
+\item Stage 3 - The domain expert interaction concludes with a final meeting, as
+  detailed in Section \ref{concluding_meeting}.  In this meeting we will discuss
+  the results of Stage 2 and go over the final ranking.
 \end{itemize}

### 2021-05-09 StateOfPractice/Ao-Report/01abstract.tex
created Ao's report folder

\addcontentsline{toc}{chapter}{Abstract}
\begin{center}
\textbf{\large Abstract}
\end{center}

Abstract here

### 2021-05-09 StateOfPractice/Ao-Report/02ack.tex
created Ao's report folder

\addcontentsline{toc}{chapter}{Acknowledgments}
\begin{center}
\textbf{\textup{\Large Acknowledgments}}
\end{center}

acknowledgements here

### 2021-05-09 StateOfPractice/Ao-Report/11Introduction.tex
created Ao's report folder

\chapter{Introduction}
\label{ch_intro}

introduction

### 2021-05-09 StateOfPractice/Ao-Report/12Background.tex
created Ao's report folder

\chapter{Background}
\label{ch_background}

\section{Categories and Status}

Briefly discuss open source, freeware and commercial software.

\section{Software Quality Definitions}

Refer to the Quality Definitions of Qualities created by us

\section{Analytic Hierarchy Process}

### 2021-05-09 StateOfPractice/Ao-Report/13MeasurementMethods.tex
created Ao's report folder

\chapter{Measurement Methods}
\label{ch_methods}

Where to include Domain experts, in Software Product Selection?

\section{Overall Process}

Briefly introduce the following subsections.

\section{Software Product Selection}

My process of collecting and screening the packages step by step.

\section{Grading Template}

Manual measurements and empirical tools.

\section{Measuring Qualities}

e.g. virtual machines, time spent per software, where to look for docs, etc.

### 2021-05-09 StateOfPractice/Ao-Report/14MeasurementResults.tex
created Ao's report folder

\chapter{Measurement Results}
\label{ch_results}

For some qualities, it might be a good idea to cross-compare with the empirical scores.

\section{Selected Software List}

\section{Installability}

\section{Correctness \& Verifiability}

\section{Reliability}

\section{Robustness}

\section{Usability}

\section{Maintainability}

\section{Reusability}

\section{Understandability}

\section{Visibility \& Transparency}

### 2021-05-09 StateOfPractice/Ao-Report/15InterviewsWithDevelopers.tex
created Ao's report folder

\chapter{Interviews with Developers}
\label{ch_interviews}

\section{Interview Methods}

The way to ask questions, the way to transcript answers and the technologies used.

\section{Interviewee Selection}

Generally speaking, ask all the teams which we can find contacts, and continue with the ones who are willing to participate.

\section{Interview Question Selection}

The aspects we focused on.

\section{Summary of Answers}

\begin{itemize}
\item Start with one by one, with commonalities and interesting special cases.

\item Shorten and summarize later.
\end{itemize}

\section{Discussions}

Any conclusions?

### 2021-05-09 StateOfPractice/Ao-Report/31FullGradingTemplate.tex
created Ao's report folder

\chapter{Full Grading Template}
\label{ap_grading}

appendix here

### 2021-05-09 StateOfPractice/Ao-Report/33InterviewAnswers.tex
created Ao's report folder

\chapter{Interview Answers}
\label{ap_interview}

appendix here

### 2021-05-10 StateOfPractice/Ao-Report/13MeasurementMethods.tex
updated the outlines

@@
 
+It might be a good idea to make the methods general and independent to domains.
+
 \section{Overall Process}
@@
 Briefly introduce the following subsections.
+
+\section{Domain Selection}
+How do we choose SCS domains, and point out again that the whole section should be able to apply to most domains.
 
@@
 e.g. virtual machines, time spent per software, where to look for docs, etc.
+
+\section{Interviewee Selection}
+
+Generally speaking, ask all the teams which we can find contacts, and continue with the ones who are willing to participate.
+
+\section{Interview Question Selection}
+
+The aspects we focused on.
+
+\section{Interview Methods}
+
+The way to ask questions, the way to transcript answers and the technologies used.

### 2021-05-12 StateOfPractice/Peter-Notes/Report/Report.tex
partial first draft LBM report

\documentclass[12pt, notitlepage]{article}
\usepackage{amsmath, mathtools}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{colortbl}
\usepackage{xr}
\usepackage{hyperref}
\usepackage{longtable}
\usepackage{xfrac}
\usepackage{tabularx}
\usepackage{float}
\usepackage{siunitx}
\usepackage{booktabs}
\usepackage{caption}
\usepackage{pdflscape}
\usepackage{afterpage}
\usepackage{titling}
\usepackage{blindtext}
\usepackage{indentfirst}

\usepackage[round]{natbib}

%\usepackage{refcheck}

\hypersetup{
    bookmarks=true,         % show bookmarks bar?
      colorlinks=true,       % false: boxed links; true: colored links
    linkcolor=red,          % color of internal links (change box color with linkbordercolor)
    citecolor=green,        % color of links to bibliography
    filecolor=magenta,      % color of file links
    urlcolor=cyan           % color of external links
}

% For easy change of table widths
\newcommand{\colZwidth}{1.0\textwidth}
\newcommand{\colAwidth}{0.24\textwidth}
\newcommand{\colBwidth}{0.76\textwidth}
\newcommand{\colCwidth}{0.28\textwidth}
\newcommand{\colDwidth}{0.72\textwidth}
\newcommand{\colEwidth}{0.33\textwidth}
\newcommand{\colFwidth}{0.67\textwidth}
\newcommand{\colGwidth}{0.5\textwidth}
\newcommand{\colHwidth}{0.28\textwidth}

\newcounter{comnum} %Commonality Number
\newcommand{\cthecomnum}{C\thecomnum}
\newcommand{\cref}[1]{C\ref{#1}}

\newcounter{varnum} %Variability Number
\newcommand{\vthevarnum}{V\thevarnum}
\newcommand{\vref}[1]{V\ref{#1}}

\newcounter{parnum} %Parameter of Variation Number
\newcommand{\ptheparnum}{P\theparnum}
\newcommand{\pref}[1]{P\ref{#1}}

% Used so that cross-references have a meaningful prefix
\newcounter{defnum} %Definition Number
\newcommand{\dthedefnum}{GD\thedefnum}
\newcommand{\dref}[1]{GD\ref{#1}}
\newcounter{datadefnum} %Datadefinition Number
\newcommand{\ddthedatadefnum}{DD\thedatadefnum}
\newcommand{\ddref}[1]{DD\ref{#1}}
\newcounter{theorynum} %Theory Number
\newcommand{\tthetheorynum}{T\thetheorynum}
\newcommand{\tref}[1]{T\ref{#1}}
\newcounter{tablenum} %Table Number
\newcommand{\tbthetablenum}{T\thetablenum}
\newcommand{\tbref}[1]{TB\ref{#1}}
\newcounter{assumpnum} %Assumption Number
\newcommand{\atheassumpnum}{P\theassumpnum}
\newcommand{\aref}[1]{A\ref{#1}}
\newcounter{goalnum} %Goal Number
\newcommand{\gthegoalnum}{P\thegoalnum}
\newcommand{\gsref}[1]{GS\ref{#1}}
\newcounter{instnum} %Instance Number
\newcommand{\itheinstnum}{IM\theinstnum}
\newcommand{\iref}[1]{IM\ref{#1}}
\newcounter{reqnum} %Requirement Number
\newcommand{\rthereqnum}{P\thereqnum}
\newcommand{\rref}[1]{R\ref{#1}}
\newcounter{lcnum} %Likely change number
\newcommand{\lthelcnum}{LC\thelcnum}
\newcommand{\lcref}[1]{LC\ref{#1}}

\usepackage{fullpage}
\usepackage{setspace}
\doublespacing
\title{Title} 
\author{Peter Michalski}
\date{\today}

\begin{document}
	\maketitle
	\begin{singlespace}
	\begin{abstract}
		Abstract here ...
	\end{abstract}

\newpage

\tableofcontents
\addtocontents{toc}{\protect\thispagestyle{empty}}

~\newpage

\listoffigures

\listoftables

~\newpage
\pagenumbering{gobble}
\pagenumbering{arabic}

\clearpage
\section{Reference Material}

\subsection{Software Engineering Related Definitions and Acronyms}

\noindent\textbf{AHP} Analytical Hierarchy Process\\

\noindent\textbf{Commonality}: A requirement or goal common to all family members.\\

\noindent\textbf{Goal}: “Goals capture, at different levels of abstraction, the various objectives the system under consideration should achieve.” \citep{van2001goal}\\

\noindent\textbf{Program Family}: A set of programs that are analyzed and designed together starting from the initial stages of the software development life-cycle.\\

\noindent\textbf{Requirements}: A software requirement is: \textit{i}) a condition or capability needed by a user to solve a problem or achieve an objective; \textit{ii}) a condition or capability that must be met or possessed by a system or system component to satisfy a contract, standard, specification, or other formally imposed document; or, \textit{iii}) a documented representation of a condition or capability as in the above two definitions. \citep{thayer2000ieee}\\

\noindent\textbf{Variability}: A requirement or goal that varies between family members.

\newpage
\subsection{Lattice Boltzmann Related Definitions and Acronyms}

\noindent\textbf{1D} 1-Dimensional\\

\noindent\textbf{2D}: 2-Dimensional\\

\noindent\textbf{3D}: 3-Dimensional\\

\noindent\textbf{BGK}: Bhatnagar-Gross-Krook\\

\noindent\textbf{CPU}: central processing unit\\

\noindent\textbf{GPU}: graphics processing unit\\

\noindent\textbf{MRT}: multi-relaxation-time\\

\noindent\textbf{SRT}: single-relaxation-time\\

\noindent\textbf{TRT}: two-relaxation-time\\

\noindent\textbf{LBM}: Lattice Boltzmann Methods\\

\noindent\textbf{LBS}: Lattice Boltzmann Solvers\\

\noindent\textbf{Velocity Directions}: The number of links connecting to each lattice node in the chosen model from neighbouring nodes. All nodes in a chosen lattice model will have the same number of links. A single link will connect between two adjacent nodes.

\end{singlespace}
\newpage
\section{Introduction}

- comment on scs?
- comment previous state of the practice projects
- what was done then and how things were done now (methodology)
-Red text denotes a linked sections of this document. Cyan text denotes a URL link. Blue text denotes a citation.

\subsection{Purpose of Report}
-constitutes the author’s M. Eng. Report
-It is therefore intended for the examination committee

\subsection{Motivation of Report}

-provide an example that demonstrates the value of se methods and technologies for SCS
-understand how software quality is impacted by the software development principles, processes and tools currently used within research software communities
- list what the purpose of SOP evaluation of LBM is for
-describe LBM in the contect of SCS
-report on the SOP of LBM software development
-evaluate the development process and software qualities
-suggest improvements to LBM SCS in the context of software development and qualities
- scs shows room for improvement in terms of quality, link papers, also why is this, link more papers

\subsection{Scope of Report}
-measured 23 software projects
- other limitations
- addresses what was done
- notes how things can be improved

\subsection{Organization of Report}
-organization of the document

\newpage
\section{Background}

must update images from pylbm to pyLBM

this is theoretical
-define SCS in more detail than introduction
- importance of qualities in SCS
- ensuring the quality can take time
-define software family (defined more/further in domain analysis)
-background of LBM and what they are used for, link to domain analysis section for further detail

\newpage

\section{Domain Analysis - DONE SECTION 1st DRAFT}

\subsection{Lattice Boltzmann Systems}

Lattice Boltzmann Methods (LBM) are a family of fluid dynamics algorithms for simulating single-phase and multiphase fluid flows, often incorporating additional physical complexities \citep{chen1998lattice}. They consider the behaviours of a collection of particles as a single unit at the mesoscopic scale. These methods predict the positional probability of a collection of particles moving through a lattice structure. Off the shelf (OTS) Lattice Boltzmann Solvers (LBS) allow for a range of fluid and physical model input parameters, computational parameters, and output parameters.

As LBS model fluid dynamics within a boundary using a predefined lattice structure, the methods rely on a two step calculation process. The first process is streaming, where the particles move along the lattice via links. The second process is collision, where energy and momentum is transferred among particles that collide \citep{bao2011lattice}.
There are many standardized lattice models - individual solvers within the family might only use a subset of them.
The LBM uses the initial parameters of the fluid to find the probability of where along the lattice linkages a group of particles are most likely to travel. It then moves the particles into the next node, and transfers the energy and momentum if a collision occurs. Then the process repeats for the duration of the modeling instance.

\begin{figure}[h!]
	\begin{center}
		\includegraphics[width=0.5\textwidth]{FeatureDiagramLBS}
		\caption{LBS Model Feature Diagram}
		\label{LBSFeatureDiagram}
	\end{center}
\end{figure}

\subsection{Commonalities}\label{comm_sec}

This section lists all the common features among all the potential family members. The commonalities are organized using the following abstraction of the system, which can be used to describe all Lattice Boltzmann systems: input information, then generate the simulation and finally output the results. Section \ref{comm_lbm} describes the commonalities for the simulation step. Section \ref{comm_in} highlights the input information that is required for all Lattice Boltzmann systems. The next section, Section \ref{comm_out}, shows the common features for the output of Lattice Boltzmann systems, such as the requirement that mesh information be written to files. (Although the output information could simply be written to the computer’s memory, in all practical applications it is desirable to have a persistent record of the output that was created.) The final section covers qualities of the system that cannot be classified as input, simulation generation or output. These commonalities are termed nonfunctional requirements of the system. For instance, all systems will have the goal that the response time to a user’s request is small enough to allow the user to focus on his/her problem and to maintain his/her train of thought, without being distracted by excessive waiting time. The commonality in this case is refined by a later variability because the specific requirement on the response time will depend on the intended usage of the mesh generating system.

Each commonality below uses the same structure. All of the commonalities are assigned a unique item number, which takes the form of a natural number with the prefix ``C''. Following this, a description of the commonality is provided along with a list of related variabilities, which are given as hyperlinks that allow navigation of the document to the text describing the variability. Finally, each commonality ends with a summary of the history, including the date of creation and any dates of modification, along with a brief description of the modification. If necessary, a previous version of the document can be obtained by using the concurrent versioning system where the files are stored.

\subsubsection{Lattice Boltzmann Method Solvers}\label{comm_lbm}
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
		\hline
		\bf Item Number& C\refstepcounter{comnum}\thecomnum{}\label{lattice} \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
		\hline
		\bf Description & A lattice discretizes a computational domain into a finite number of points. All LBS discretize the computational domain using a regular, evenly spaced grid within a boundary.\\
		\hline
		\bf Related Variability & \vref{numveldir} \vref{varcolop} \vref{varboundary}\\
		\hline
	\end{tabular}
\end{minipage}
~\newline
~\newline
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
		\hline
		\bf Item Number& C\refstepcounter{comnum}\thecomnum\label{collisionoperator} \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
		\hline
		\bf Description & All Lattice Boltzmann versions use a collision operator which concerns collisions between particles.\\
		\hline
		\bf Related Variability & \vref{dimensions} \vref{numveldir}\\
		\hline
	\end{tabular}
\end{minipage}
~\newline
~\newline
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
		\hline
		\bf Item Number& C\refstepcounter{comnum}\thecomnum\label{probabilitydensityfunc} \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
		\hline
		\bf Description & All Lattice Boltzmann versions user a probability density function to give the probability that fluid has moved into a specific domain.\\
		\hline
		\bf Related Variability & \vref{dimensions}\\
		\hline
	\end{tabular}
\end{minipage}
~\newline
~\newline
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
		\hline
		\bf Item Number& C\refstepcounter{comnum}\thecomnum\label{equildistfunc} \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
		\hline
		\bf Description & Every Lattice Boltzmann version uses an equilibrium distribution function to capture the probability distribution of particles.\\
		\hline
		\bf Related Variability & \vref{varedf} \vref{edfstorage} \vref{edfcoeff}\\
		\hline
	\end{tabular}
\end{minipage}
~\newline
~\newline
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
		\hline
		\bf Item Number& C\refstepcounter{comnum}\thecomnum\label{Boltzmanntransequ} \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
		\hline
		\bf Description & Every Lattice Boltzmann version uses a Boltzmann transport equation to describe the statistical behaviour of a system that does not have collisions.\\
		\hline
		\bf Related Variability & \vref{vartraneq}\\
		\hline
	\end{tabular}
\end{minipage}\\
\subsubsection{Input}\label{comm_in}
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
		\hline
		\bf Item Number& C\refstepcounter{comnum}\thecomnum\label{fluidandboundaryinfo} \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
		\hline
		\bf Description & The LBS require fluid, model, and boundary information for the problem.\\
		\hline
		\bf Related Variability & \vref{interface}\\
		\hline
	\end{tabular}
\end{minipage}\\
\subsubsection{Output}\label{comm_out}
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
		\hline
		\bf Item Number& C\refstepcounter{comnum}\thecomnum\label{predictionstomemory} \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
		\hline
		\bf Description & LBS write fluid parameter predictions to memory.\\
		\hline
		\bf Related Variability & \vref{varvisual} \vref{formatoutput} \\
		\hline
	\end{tabular}
\end{minipage}\\
\subsubsection{Nonfunctional Requirements}\label{comm_nfr}
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
		\hline
		\bf Item Number& C\refstepcounter{comnum}\thecomnum\label{precision} \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
		\hline
		\bf Description & LBS provides the precision required for the particular problems it is intended to solve.\\
		\hline
		\bf Related Variability & \vref{varprecise}\\
		\hline
	\end{tabular}
\end{minipage}
~\newline
~\newline
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
		\hline
		\bf Item Number& C\refstepcounter{comnum}\thecomnum\label{accuracy} \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
		\hline
		\bf Description & LBS provides the accuracy required for the particular problems it is intended to solve.\\
		\hline
		\bf Related Variability & \vref{varaccuracy}\\
		\hline
	\end{tabular}
\end{minipage}
~\newline
~\newline
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
		\hline
		\bf Item Number& C\refstepcounter{comnum}\thecomnum\label{performance} \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
		\hline
		\bf Description & The response time is small enough to allow the user to focus on their problem without being distracted by excessive waiting times.\\
		\hline
		\bf Related Variability & \vref{varresponsetime}\\
		\hline
	\end{tabular}
\end{minipage}
~\newline
~\newline
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
		\hline
		\bf Item Number& C\refstepcounter{comnum}\thecomnum\label{portability} \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
		\hline
		\bf Description & LBS will be as portable to other operating systems as required by the users.\\
		\hline
		\bf Related Variability & \vref{varportability}\\
		\hline
	\end{tabular}
\end{minipage}
~\newline
~\newline
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
		\hline
		\bf Item Number& C\refstepcounter{comnum}\thecomnum\label{usability} \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
		\hline
		\bf Description & The use of LBS is easy and efficient.\\
		\hline
		\bf Related Variability & \vref{varusability}\\
		\hline
	\end{tabular}
\end{minipage}\\

\subsection{Variabilities}\label{var_sec}
This section provides a list of characteristics that may vary among family members. As in Section \ref{comm_sec}, the first three subsections on variabilities are organized into the following sublists: Mesh Generation, Input and Output. The final two subsections list variabilities that can be characterized as system constraints and as nonfunctional requirements.

As for the commonalities, each variability is labelled with a unique item number. In this case the numbers are prepended with the letter ``V''. The other four headings provided for each variability are: Description, Related Commonality, Related Parameter and History. The related commonalities and parameters are given as a set of identifiers that respectively refer back to the previous section on commonalities or refer forward to the next section on parameters of variation.

\subsubsection{Lattice Boltzmann Method Solvers}
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
		\hline
		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{parallel} \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
		\hline
		\bf Description & LBS may use a framework for parallel processing of the model.\\
		\hline
		\bf Related Commonality & None\\
		\hline
		\bf Related Parameter & \pref{parparallel}\\
		\hline
	\end{tabular}
\end{minipage}\\
~\newline\\
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
		\hline
		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{varedf} \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
		\hline
		\bf Description & Different versions of an equilibrium distribution function can capture the probability distribution of particles.\\
		\hline
		\bf Related Commonality & \cref{equildistfunc}\\
		\hline
		\bf Related Parameter & \pref{paredf}\\
		\hline
	\end{tabular}
\end{minipage}
~\newline
~\newline
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
		\hline
		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{edfstorage} \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
		\hline
		\bf Description & Storage patters for distribution function can vary.\\
		\hline
		\bf Related Commonality & \cref{equildistfunc}\\
		\hline
		\bf Related Parameter & \pref{parstorage}\\
		\hline
	\end{tabular}
\end{minipage}
~\newline
~\newline
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
		\hline
		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{edfcoeff} \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
		\hline
		\bf Description & Coefficients for distribution function can vary.\\
		\hline
		\bf Related Commonality & \cref{equildistfunc}\\
		\hline
		\bf Related Parameter & \pref{paredfcoeff}\\
		\hline
	\end{tabular}
\end{minipage}
~\newline
~\newline
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
		\hline
		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{dimensions} \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
		\hline
		\bf Description & The number of dimensions in the lattice of the model can vary.\\
		\hline
		\bf Related Commonality & \cref{lattice} \cref{probabilitydensityfunc}\\
		\hline
		\bf Related Parameter & \pref{pardimensions}\\
		\hline
	\end{tabular}
\end{minipage}
~\newline
~\newline
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
		\hline
		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{numveldir} \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
		\hline 
		\bf Description & The number of velocity directions in the lattice of the model can vary.\\
		\hline
		\bf Related Commonality & \cref{lattice} \cref{collisionoperator}\\
		\hline
		\bf Related Parameter & \pref{parveldir}\\
		\hline
	\end{tabular}
\end{minipage}
~\newline
~\newline
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
		\hline
		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{varcolop} \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
		\hline
		\bf Description & Various collision operators can be used.\\
		\hline
		\bf Related Commonality & \cref{lattice} \cref{collisionoperator}\\
		\hline
		\bf Related Parameter & \pref{parcollisop}\\
		\hline
	\end{tabular}
\end{minipage}
~\newline
~\newline
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
		\hline
		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{vartraneq} \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
		\hline
		\bf Description & Various transport equations can be used to describe the statistical behaviour of the system\\
		\hline
		\bf Related Commonality & \cref{Boltzmanntransequ}\\
		\hline
		\bf Related Parameter & \pref{partransequ}\\
		\hline
	\end{tabular}
\end{minipage}
~\newline
~\newline
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
		\hline
		\bf Item Number&  V\refstepcounter{varnum}\thevarnum\label{numberoffluids} \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
		\hline
		\bf Description & The number of fluids allowed in the LBS.\\
		\hline
		\bf Related Commonality & \cref{fluidandboundaryinfo}\\
		\hline
		\bf Related Parameter & \pref{parfluids}\\
		\hline
	\end{tabular}
\end{minipage}
~\newline
~\newline
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
		\hline
		\bf Item Number&  V\refstepcounter{varnum}\thevarnum\label{typefluidparm} \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
		\hline
		\bf Description & The type of fluid parameters.\\
		\hline
		\bf Related Commonality & \cref{fluidandboundaryinfo}\\
		\hline
		\bf Related Parameter & \pref{partypeparm}\\
		\hline
	\end{tabular}
\end{minipage}
~\newline
~\newline
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
		\hline
		\bf Item Number&  V\refstepcounter{varnum}\thevarnum\label{varboundary} \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
		\hline
		\bf Description & The boundary of the lattice can have varied conditions.\\
		\hline
		\bf Related Commonality & \cref{lattice}\\
		\hline
		\bf Related Parameter & \pref{parboundary}\\
		\hline
	\end{tabular}
\end{minipage}\\
\subsubsection{Input}
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
		\hline
		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{interface} \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
		\hline
		\bf Description & The input interface can vary between LBS.\\
		\hline
		\bf Related Commonality & \cref{fluidandboundaryinfo}\\
		\hline
		\bf Related Parameter & \pref{parinterface}\\
		\hline
	\end{tabular}
\end{minipage}\\
\subsubsection{Output}
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
		\hline
		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{varvisual} \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
		\hline
		\bf Description & Visual presentation of the prediction.\\
		\hline
		\bf Related Commonality & \cref{predictionstomemory}\\
		\hline
		\bf Related Parameter & \pref{parvarvisual}\\
		\hline
	\end{tabular}
\end{minipage}
~\newline
~\newline
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
		\hline
		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{formatoutput} \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
		\hline
		\bf Description & Format of prediction information.\\
		\hline
		\bf Related Commonality & \cref{predictionstomemory}\\
		\hline
		\bf Related Parameter & \pref{parformatoutput}\\
		\hline
	\end{tabular}
\end{minipage}\\
\subsubsection{System Constraints}
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
		\hline
		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{processing} \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
		\hline
		\bf Description & Hardware which processes the calculations\\
		\hline
		\bf Related Commonality & None \\
		\hline
		\bf Related Parameter & \pref{parprocessing}\\
		\hline
	\end{tabular}
\end{minipage}
~\newline
~\newline
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
		\hline
		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{operatingsystem} \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
		\hline
		\bf Description & Operating systems on which LBS runs.\\
		\hline
		\bf Related Commonality & None \\
		\hline
		\bf Related Parameter & \pref{paropsystem}\\
		\hline
	\end{tabular}
\end{minipage}
~\newline
~\newline
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
		\hline
		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{storage} \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
		\hline
		\bf Description & Amount of storage and memory needed for the LBS.\\
		\hline
		\bf Related Commonality & None \\
		\hline
		\bf Related Parameter & \pref{parstorageconst}\\
		\hline
	\end{tabular}
\end{minipage}\\
\subsubsection{Nonfunctional Requirements}
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
		\hline
		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{varprecise} \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
		\hline
		\bf Description & The precision needed for each input and output.\\
		\hline
		\bf Related Commonality & \cref{precision}\\
		\hline
		\bf Related Parameter & \pref{parprecise}\\
		\hline
	\end{tabular}
\end{minipage}
~\newline
~\newline
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
		\hline
		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{varaccuracy} \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
		\hline
		\bf Description & The required accuracy for the output.\\
		\hline
		\bf Related Commonality & \cref{accuracy}\\
		\hline
		\bf Related Parameter & \pref{paraccuracy}\\
		\hline
	\end{tabular}
\end{minipage}
~\newline
~\newline
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
		\hline
		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{varresponsetime} \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
		\hline
		\bf Description & The response time required for user interaction with the system varies.\\
		\hline
		\bf Related Commonality & \cref{performance}\\
		\hline
		\bf Related Parameter & \pref{parresponse}\\
		\hline
	\end{tabular}
\end{minipage}
~\newline
~\newline
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
		\hline
		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{varportability} \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
		\hline
		\bf Description & The operating systems on which a LBS can run.\\
		\hline
		\bf Related Commonality & \cref{portability}\\
		\hline
		\bf Related Parameter & \pref{parportability}\\
		\hline
	\end{tabular}
\end{minipage}
~\newline
~\newline
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
		\hline
		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{varusability} \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
		\hline
		\bf Description & The ease with which LBS can effectively be run varies.\\
		\hline
		\bf Related Commonality & \cref{usability}\\
		\hline
		\bf Related Parameter & \pref{parusability}\\
		\hline
	\end{tabular}
\end{minipage}\\
\newpage
\subsection{Parameters of Variation}
This section specifies the parameters of variation for the variabilities listed in Section \ref{var_sec}. They are organized into the same five subcategories as employed previously: Mesh Generation, Input, Output, System Constraints, Nonfunctional Requirements.

Each parameter of variation is given a unique identifier of the form “P” followed by a natural number. The corresponding variability is listed and a hyperlink is provided that allows navigation back to the appropriate item in Section \ref{var_sec}. The final entry for each parameter of variation is the binding time, which is the time in the software lifecycle when the variability is fixed. The binding time could be during specification, or during building
of the system (compile time), or during execution of the system (run time). It is possible to have a mixture of binding times. For instance, a parameter of variation could have a binding time of “specification or building” to represent that the parameter could be set at specification time, or it could be postponed until the given family member is built. The choice of postponing the decision until the build would be associated with the presence of a domain specific language that would allow postponing decisions on the values of the parameter of variation.

\subsubsection{Lattice Boltzmann Method Solvers}
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
		\hline
		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parparallel} \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
		\hline
		\bf Corresponding Variability & \vref{parallel}\\
		\hline
		\bf Range of Parameters & OpenMP, OpenCL, CUDA, or MPI are used if LBS is parallel.\\
		\hline
		\bf Binding Time & Build Time\\
		\hline
	\end{tabular}
\end{minipage}
~\newline
~\newline
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
		\hline
		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{paredf} \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
		\hline
		\bf Corresponding Variability & \vref{parallel}\\
		\hline
		\bf Range of Parameters & Equilibrium can be approximated up to different orders in incompressible or compressible versions.\\
		\hline
		\bf Binding Time & Build Time\\
		\hline
	\end{tabular}
\end{minipage}
~\newline
~\newline
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
		\hline
		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parstorage} \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
		\hline
		\bf Corresponding Variability & \vref{edfstorage}\\
		\hline
		\bf Range of Parameters & Single and multiple array storage patterns can be used.\\
		\hline
		\bf Binding Time & Build Time\\
		\hline
	\end{tabular}
\end{minipage}
~\newline
~\newline
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
		\hline
		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{paredfcoeff} \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
		\hline
		\bf Corresponding Variability & \vref{edfcoeff}\\
		\hline
		\bf Range of Parameters & Numerous coefficients for equilibrium distribution function based on number of velocity directions. \\
		\hline
		\bf Binding Time & Build Time\\
		\hline
	\end{tabular}
\end{minipage}
~\newline
~\newline
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
		\hline
		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{pardimensions} \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
		\hline
		\bf Corresponding Variability & \vref{dimensions}\\
		\hline
		\bf Range of Parameters & LBS models can have up to 3 dimensions.\\
		\hline
		\bf Binding Time & Build Time or Run Time\\
		\hline
	\end{tabular}
\end{minipage}
~\newline
~\newline
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
		\hline
		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parveldir} \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
		\hline
		\bf Corresponding Variability & \vref{numveldir}\\
		\hline
		\bf Range of Parameters & One dimensional models include options of 2, 3, and 5 velocity directions. Two dimensional models include options of 9, 13, and 15 velocity directions. Three dimensional models include options of 15, 19, and 27 velocity directions. \\
		\hline
		\bf Binding Time & Build Time or Run Time\\
		\hline
	\end{tabular}
\end{minipage}
~\newline
~\newline
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
		\hline
		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parcollisop} \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
		\hline
		\bf Corresponding Variability & \vref{varcolop}\\
		\hline
		\bf Range of Parameters & SRT, TRT, MRT, BGK collision operators.\\
		\hline
		\bf Binding Time & Build Time\\
		\hline
	\end{tabular}
\end{minipage}
~\newline
~\newline
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
		\hline
		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{partransequ} \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
		\hline
		\bf Corresponding Variability & \vref{vartraneq}\\
		\hline
		\bf Range of Parameters & Collision and collision free transport equations.\\
		\hline
		\bf Binding Time & Build Time\\
		\hline
	\end{tabular}
\end{minipage}
~\newline
~\newline
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
		\hline
		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parfluids} \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
		\hline
		\bf Corresponding Variability & \vref{numberoffluids}\\
		\hline
		\bf Range of Parameters & LBS can model one or more fluids.\\
		\hline
		\bf Binding Time & Build Time\\
		\hline
	\end{tabular}
\end{minipage}
~\newline
~\newline
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
		\hline
		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{partypeparm} \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
		\hline
		\bf Corresponding Variability & \vref{typefluidparm}\\
		\hline
		\bf Range of Parameters & LBS fluid parameters include Reynolds Number, density, viscosity, time, pressure, force, direction, relaxation rate, turbulence.\\
		\hline
		\bf Binding Time & Build Time\\
		\hline
	\end{tabular}
\end{minipage}
~\newline
~\newline
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
		\hline
		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parboundary} \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
		\hline
		\bf Corresponding Variability & \vref{varboundary}\\
		\hline
		\bf Range of Parameters & Lattice boundary can have reflective or non-reflective conditions.\\
		\hline
		\bf Binding Time & Build Time\\
		\hline
	\end{tabular}
\end{minipage}\\
\subsubsection{Input}
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
		\hline
		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parinterface} \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
		\hline
		\bf Corresponding Variability & \vref{interface}\\
		\hline
		\bf Range of Parameters & Input can be graphical, text or file.\\
		\hline
		\bf Binding Time & Build Time\\
		\hline
	\end{tabular}
\end{minipage}\\
\subsubsection{Output}
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
		\hline
		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parvarvisual} \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
		\hline
		\bf Corresponding Variability & \vref{varvisual}\\
		\hline
		\bf Range of Parameters & LBS can provide 1D, 2D, and 3D rendering of  the model.\\
		\hline
		\bf Binding Time & Build Time or Run Time\\
		\hline
	\end{tabular}
\end{minipage}
~\newline
~\newline
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
		\hline
		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parformatoutput} \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
		\hline
		\bf Corresponding Variability & \vref{formatoutput}\\
		\hline
		\bf Range of Parameters & LBS prediction information is output in either text or binary format.\\
		\hline
		\bf Binding Time & Build Time\\
		\hline
	\end{tabular}
\end{minipage}\\
\subsubsection{System Constraints}
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
		\hline
		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parprocessing} \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
		\hline
		\bf Corresponding Variability & \vref{processing}\\
		\hline
		\bf Range of Parameters & The LBS model can be calculated on the CPU or GPU.\\
		\hline
		\bf Binding Time & Build Time\\
		\hline
	\end{tabular}
\end{minipage}
~\newline
~\newline
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
		\hline
		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{paropsystem} \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
		\hline
		\bf Corresponding Variability & \vref{operatingsystem}\\
		\hline
		\bf Range of Parameters & LBS can be run on Windows, MacOS, or Linux versions.\\
		\hline
		\bf Binding Time & Build Time\\
		\hline
	\end{tabular}
\end{minipage}
~\newline
~\newline
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
		\hline
		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parstorageconst} \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
		\hline
		\bf Corresponding Variability & \vref{storage}\\
		\hline
		\bf Range of Parameters & The amount of memory and storage varies between LBS. LBS sizes begin in the low gigabytes. \\
		\hline
		\bf Binding Time & Build Time\\
		\hline
	\end{tabular}
\end{minipage}\\
\subsubsection{Nonfunctional Requirements}
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
		\hline
		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parprecise} \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
		\hline
		\bf Corresponding Variability & \vref{varprecise}\\
		\hline
		\bf Range of Parameters & Each input and each output will have a specified the precision with which the real number is stored and/or calculated.\\
		\hline
		\bf Binding Time & Build Time or Run Time\\
		\hline
	\end{tabular}
\end{minipage}
~\newline
~\newline
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
		\hline
		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{paraccuracy} \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
		\hline
		\bf Corresponding Variability & \vref{varaccuracy}\\
		\hline
		\bf Range of Parameters & The tolerance allowed for each output produced by the system so that the system produces usable results. \\
		\hline
		\bf Binding Time & Build Time or Run Time\\
		\hline
	\end{tabular}
\end{minipage}
~\newline
~\newline
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
		\hline
		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parresponse} \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
		\hline
		\bf Corresponding Variability & \vref{varresponsetime}\\
		\hline
		\bf Range of Parameters & The maximum amount of time that the user is expected to wait for the system to produce a result ranges between systems and models from near instantaneous to several minutes or longer.\\
		\hline
		\bf Binding Time & Specification\\
		\hline
	\end{tabular}
\end{minipage}
~\newline
~\newline
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
		\hline
		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parportability} \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
		\hline
		\bf Corresponding Variability & \vref{varportability}\\
		\hline
		\bf Range of Parameters & The ease with which LBS can run on Windows, MacOS, and Linux system varies. Some LBS can run on all three, while some can run on multiple or only one operating system.\\
		\hline
		\bf Binding Time & Specification\\
		\hline
	\end{tabular}
\end{minipage}
~\newline
~\newline
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
		\hline
		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parusability} \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
		\hline
		\bf Corresponding Variability & \vref{varusability}\\
		\hline
		\bf Range of Parameters & Some LBS require programming skills and domain expert knowledge while some LBS can be run without such skills.\\
		\hline
		\bf Binding Time & Specification\\
		\hline
	\end{tabular}
\end{minipage}

\newpage
\section{Method - DONE SECTION 1st DRAFT}

In this project we set out to answer the research questions found in Section \ref{researchquestions} for the Lattice Boltzmann scientific computing software domain. The process involved systematically measuring and analyzing members of this software family along the quality attributes found in Section \ref{softwarequalities}. This included gathering quantitative and qualitative measurements. A goal of the project was to also produce a hierarchical quality assessment for LBM software.

We collected quantitative data using the measures found in the measurement template of Section \ref{measurementtemplate}. Some of this was empirical software engineering related data, such as the number of files, number of lines of code, percentage of issues that are closed. Most was gathered by manually observing the software, it's source code, and it's artifacts, while some was gathered using the tools listed in Section \ref{tools}.

We also collected qualitative data by interviewing the software package developers and asking them the questions found in Section \ref{interviewquestions}. Furthermore, we solicited the assistance of domain experts to better access each software package by leveraging their experience to access the functional and non-functional domain software requirements.

This section begins with the steps of the overall process we used to select, measure and compare LBM software. This is followed by quality definitions and how they were applied in our project. The rest of the section provides an overview of how candidate software packages were selected and filtered, the empirical measurements and software tools, and the analytic hierarchy process (AHP).

\subsection{Process}

 An overview of the assessment process is given in the following steps:

\begin{enumerate}
	\item Identify a broad list of candidate software packages in the domain. This is discussed in Section \ref{identifysoftware}.
	\item Filter the software package list. This is discussed in Section \ref{filtersoftware}.
	\item Gather the source code and documentation for each software package.
	\item Collect empirical measures. This is discussed in Section \ref{empiricalmeasures}.
	\item Measure using the measurement template. This is discussed in Section \ref{empiricalmeasures}. The measurement template can be found in Section \ref{measurementtemplate}.
	\item Survey the developers. The developers of each software package of the filtered software list were contacted for voluntary interviews. Four interviews were conducted in total. The interview questions can be found in Section \ref{interviewquestions}.
	\item Use AHP to rank the software packages. This is discussed in Section \ref{AHP}.
	\item Analyze the results and answer the research questions. The answers can be found in section \ref{answersquestions}.
\end{enumerate}

\subsection{Software Qualities}\label{softwarequalities}

Software quality attributes facilitate the measurement and comparison of software packages in this state of the practice assessment. In this report we adopt software quality definitions from various researchers and subject matter expert entities. The quality measurement results are found in Section \ref{qualityresults} of this report. Section \ref{qualityrecommentations} lists recommendations to address software qualities in LBM software packages. The following are the software quality definitions used in this state of practice exercise, along with comments regarding their quantitative and qualitative measurement.

\subsubsection{Installability}

Installability is measured by the effort required for the installation, uninstallation or reinstallation of a software product in a specified
environment \citep{ISO/IEC25010} \citep{lenhard2013measuring}. A good measure of installability correlates with scenarios when low or moderate effort is required to gather and prepare software for its general use on a system for which it was designed. In this case effort includes the time spent finding and understanding the installation instructions, the man-time and resources spent performing the installation procedure, and the absence or ease of overcoming system compatibility issues. The ability to reasonably validate the installation procedure can also have a positive effect on the measure of installability. Similarly, the ease of uninstallation has an affect on the measure of installability.

\subsubsection{Correctness}

 A software program is correct if it behaves according to its stated
specifications \citep{GhezziEtAl2003}. This requires that the specification is available. Software is less likely to have a formal specification if it is not developed by seasoned or professional software developers. Since some software does not have a specification available, the correctness of software cannot always be verified. Despite an absent specification, the correctness of the output of scientific computing software can sometimes be manually verified by applying domain knowledge. A good measure of correctness correlates with the availability of a requirements specification and reference to domain theory, as well as the explicit use of tools or techniques for building confidence of correctness, such as documentation generators and software analysis tools.

\subsubsection{Verifiability}

Verifiability is measured by the extent to which a set of tests can be written and executed, to demonstrate that the delivered system meets the specification \citep{sommerville}. Similarly to correctness, verifiability is positively correlated with the availability of a specification and with reference to domain knowledge. A good measure of verifiability is further positively correlated with the availability of well written tutorials that include expected output, with software unit testing documentation, and with evidence of continuous integration during the development process.

\subsubsection{Reliability}

Reliability is measured by the probability of failure-free operation of a computer program in a specified environment for a specified time, i.e. the average time interval between two failures also known as the mean time to failure (MTTF) \citep{GhezziEtAl2003} \citep{musa1987software}. Reliability is thus positively correlated with the absence of errors during installation and use. Recoverability from errors also improves the measure of reliability. In turn the presence of descriptive error messages and logs increases the ability to recover the system.

\subsubsection{Robustness}

Software possesses the characteristic of robustness if it behaves “reasonably” in two situations: i) when it encounters circumstances not anticipated in the requirements specification; and ii) when the assumptions in its requirements specification are violated \citep{boehm2007software} \citep{ghezzi1991fundamentals}. A good measure of robustness correlates with a reasonable reaction to unexpected input like data of the wrong type, empty input, or missing files or links. A reasonable reaction includes an appropriate error message and the ability to recover the system.

\subsubsection{Performance}

Performance is measured by the degree to which a system or component accomplishes its designated functions within given constraints, such as speed (database response times, for instance), throughput (transactions per second), capacity (concurrent usage loads), and timing (hard real-time demands) \citep{IEEEStdGlossarySET1990} \citep{wiegers2003softreq}. In this state of the practice assessment performance was not quantitatively measured. Instead the documentation of each software package was scanned for information that alludes to a consideration of performance, such as parallelization tools.

\subsubsection{Usability}

Usability is measured by the extent to which a software product can be used by specified users to achieve specified goals with effectiveness, efficiency, and satisfaction in a specified context of use \citep{nielsonusability}. A good measure of usability correlates with the presence of documentation, specifically tutorials, manuals, and defined user characteristics, and user support. Preferably the user support model has avenues to contact developers and report issues.

\subsubsection{Maintainability}

A measure of maintainability is the effort with which a software
system or component can be modified to correct faults, improve performance or other attributes, and satisfy new requirements \citep{IEEEStdGlossarySET1990} \citep{boehm2007software}. In this state of practice analysis maintainability is measured by the quality of documentation artifacts, and the presence of version control and issue tracking. These qualities can greatly decrease the effort needed to modify software. There are many documentation artifacts that can improve maintainability, including user and developer manuals, specifications, README files, change logs, release notes, publications, forums, and instructional websites.

\subsubsection{Modifiability}

Modifiability refers to the ease with which stable changes can be made to a system and the flexibility of the system to adopt such changes \citep{8016712}.
This state of practice assessment did not quantitatively measure modifiability. Developers were asked in interviews if they considered the ease of future changes when developing the software packages, specifically changes to the structure of the system, modules and code blocks. A follow up question asked if any measures had been taken.

\subsubsection{Reusability}

Reusability refers to the extent to which components of a software package can be used with or without adaptation in other software packages \citep{kalagiakos2003non}. A good measure of reusability results from a large number of easily reusable components. Increased software modularization, smaller components with well defined interfaces, is important. For this state of practice assessment, a good measure of reusablity correlates with an increased number of code files, and the availability of API documentation.

\subsubsection{Understandability}

Understandability is measured by the capability of the software package to enable the user to understand its suitability and function \citep{ISO9126}.
It is an artifact-dependent quality. Understandability is different for the user-interface, source code, documentation. In this state of practice analysis understandability focuses on the source code. It is measured by the consistency of a formatting style, the extend of modularization, the explicit identification of coding standards, the presence of meaningful identifiers, and clarity of comments.

\subsubsection{Traceability}

Traceability refers to the ability to link the software implementation and the software artifacts, especially the requirement specification \citep{McCallEtAl1977}. Similar to the quality of correctness, this requires some form of specification to be available. The quality refers to keeping track of information as it changes forms or relates between artifacts. This state of practice assessment did not quantitatively measure traceability. Developers were asked in interviews how documentation fits into their development process.

\subsubsection{Visibility and Transparency}

Visibility and transparency refer to the extent to which all of the steps of a software development process, and the current status of it, are conveyed clearly \citep{ghezzi1991fundamentals}. In this state of practice assessment a good measure of visibility and transparency correlates with a well defined development process, the presence of development process and environment documentation, and software package version release notes.

\subsubsection{Reproducibility}

Software achieves reproducibility if another developer can take the requirements documentation and re-obtain the same software artifacts \citep{BenureauAndRougier2017}. In this context, reproducibility refers to the similarity between artifacts. Reproducibility can also refer to the output of the software, irrespective of the artifacts, where the scientific results are compared between software implementations, or between software and manual implementations. This state of practice assessment did not quantitatively measure reproducibility. Developers were asked in interviews if they have any concern that their computational results won’t be reproducible in the future, and if they have taken any steps to ensure reproducibility.

\subsubsection{Unambiguity}

Unambiguity refers to the extent to which two readers have similar interpretations when reading software artifacts. In other words, artifacts are unambiguous if, and only if, they only have one interpretation \citep{IEEE1998}. This state of practice assessment did not quantitatively measure unambiguity. Developers were asked in interviews if they think that the current documentation can clearly convey all necessary knowledge to the users, and how they achieved this or what improvements are needed ot achieve it.

\subsection{Identify Candidate Software}\label{identifysoftware}

The candidate software was found through search engine queries targeting authoritative lists of software. We found LBM software listed on the websites GitHub and swMATH, as well as through articles found in scholarly journals and databases.

When forming the list and reviewing the candidate software the following properties were considered:

\begin{enumerate}
	\item The software functionality must fall within the identified domain.
	\item The source code must be viewable.
	\item The empirical measures listed in Section 8 should ideally be available, which implies a preference for GitHub-style repositories.
	\item The software cannot be marked as incomplete or in an initial development phase.
\end{enumerate}

The initial list had 46 packages, including a few packages that were later found to not have publicly available source code, or to be incomplete.

\subsection{Filter the Software List}\label{filtersoftware}

In order to reduce the number of members of the candidate software list, the following filters were applied. The filters were applied in the priority order listed.

\begin{enumerate}
	\item Scope: Software is removed by narrowing what functionality is considered to be within the scope of the domain.
	\item Usage: Software packages were eliminated if their installation procedure was missing or not clear and easy to follow.
	\item Age: The older software packages (age being measured by the last date when a change was made) are eliminated, except in the cases where an older software package appears to be highly recommended and currently in use. 
\end{enumerate}

For the third item in the above filter, software packages were characterized as `alive' if their related documentation had been updated within
the last 18 month. Packages were categorized as `dead' if the last update of this information was more than 18 month ago.\\

\begin{center}
	\begin{tabular}{ p{4cm}p{3cm}p{3cm} }
		\hline
		Name & Released & Updated\\
		\hline
		DL\_MESO (LBE) & unclear & 2020 Mar\\
		waLBerla & 2008 Aug & 2020 Jul\\
		Sailfish & 2012 Nov & 2019 Jun\\
		Palabos & unclear & 2020 Jul\\
		OpenLB & 2007 Jul & 2019 Oct\\
		MechSys   & 2008 Jun    & 2020 Jul\\
		LUMA&   2016 Nov   & 2020 Feb\\
		Ludwig& 2018 Aug & 2020 Jul\\
		lettuce    & 2019 May & 2020 Jul\\
		pyLBM & 2015 Jun&   2020 Jun\\
		lbmpy& unknown  & 2020 Jun  \\
		TCLB & 2013 Jun  & 2020 Apr\\
		ESPResSo & 2010 Nov & 2020 Jun\\
		ESPResSo++ & 2011 Feb & 2020 Apr\\
		\hline
	\end{tabular}
	\captionof{table}{Alive Software Packages}\label{alivepackages}
\end{center}

While the initial list had 46 packages, filtering by scope, usage, and age, decreased the size of the list to 23 packages. Many of the 23 packages that were removed could not be tested as there was no installation guide, they were incomplete, source code was not publicly available, a license was needed, and the project was out of scope or not up to a standard that would support incorporating them into this study. Of the remaining 23 packages that were studied, some were kept on the list despite being marked as dead due to their prevalence on authoritative lists on LBM software and their surface excellence, specifically the clearly considerable time that was put into these projects.  
The final list of software that was analyzed in this project can be found in the following two tables. Table \ref{alivepackages} lists packages that fell into the `alive' category as of mid 2020, and Table \ref{deadpackages} lists packages that were `dead' at that time. \\

\begin{center}
	\begin{tabular}{ p{4cm}p{3cm}p{3cm} }
		\hline
		Name & Released & Updated\\
		\hline
		SunlightLB & 2005 Sep & 2012 Nov\\
		MP-LABS & 2008 Jun & 2014 Oct\\
		LIMBES & 2010 Nov & 2014 Dec\\
		LB3D & unclear & 2012 Mar\\
		LB3D-Prime & 2005 & 2011 Oct\\
		LB2D-Prime & 2005 & 2012 Apr\\
		LatBo.jl & 2014 Aug & 2017 Feb\\
		HemeLB & 2007 Jun & 2018 Aug\\
		laboetie & 2014 Nov & 2018 Aug\\
		\hline
	\end{tabular}
	\captionof{table}{Dead Software Packages}\label{deadpackages}
\end{center}

There is considerable variation among these software packages, including their intended purpose, size, user interfaces, and their base programming languages. For example, the OpenLB software package is predominantly a C++ package that makes use of hybrid parallelization and was designed to address a range of computational fluid dynamics problems \citep{heuveline2009towards}. The software package pyLBM is an all-in-one Python language package for numerical simulations \citep{graille2017pylbm}. ESPResSo is an extensible simulation package that is specifically for research on soft matter, and is written in C++ and Python \citep{weik2019espresso}. The HemeLB is used for efficient simulation of fluid flow in several medical domains, and is written predominantly in C, C++, and Python \citep{mazzeo2008hemelb}.

\subsection{Empirical Measures}\label{empiricalmeasures}

The quality measurements in this assessment rely on the gathering and analyzing of raw and processed empirical data related to the research questions listed in Section \ref{researchquestions}.

All of the quality measurements that are part of the AHP analysis are in this category, while qualitative data gathered during interviews with developers is not. This part of the assessment focuses on data that is reasonably easy to collect. Much of the data was gathered by manually reviewing the artifacts of each software package, while some was gathered using freeware tools listed in Section \ref{tools}. The data that was gathered is listed in the measurement template found in Section \ref{measurementtemplate}. Some of the data required processing other data within the template, including the status of the software package which relied on the last commit date, the percentage of issues that are closed which relied on the number of open and closed issues, and the percentage of code that is comments which relied on the number of total lines and comment lines in text-based files. The complete measurement template data was then analyzed using the grading template found in Section \ref{gradingtemplate}, the output of which was analyzed using the AHP described in Section \ref{AHP}.

\subsubsection{Software Tools}\label{tools}

Most of the measurement template data was gathered by observing GitHub repository metrics and software package artifacts, or by processing data gathered using freeware tools. Data in the final three sets of the measurement template was collected using these tools. The tool \href{https://github.com/tomgi/git_stats}{GitStats} was used to measure each software package's GitHub repository for the number of binary files, the number of added and deleted lines, and the number of commits over varying time intervals. The tool \href{https://github.com/boyter/scc}{Sloc Cloc and Code (scc)} was used to measure the number of text based files as well as the number of total, code, comment, and blank lines in each GitHub repository. Details on installing and running these tools can be found in the \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/A
	Guide to Empirical Measures.pdf} {Guide to Empirical Measures} file in the AIMSS repository.

\subsection{Analytical Hierarchy Process}\label{AHP}

The Analytical Hierarchy Process (AHP) is a decision-making technique that is used to compare multiple options by multiple criteria. In our work AHP was used for comparing and ranking the LBM software packages using the overall impression quality scores that were gathered in the measurement template found in Section \ref{measurementtemplate} using the grading template found in Section \ref{gradingtemplate}. AHP performs a pairwise analysis using a matrix and generates an overall score as well as quality scores for each software package. \cite{SmithEtAl2016} shows how AHP is applied to ranking software based on quality measures.

This project used a tool for conducting this process. The tool includes a sensitivity analysis that was used to ensure that the software package rankings are appropriate with respect to the uncertainty of the quality scores. The README file of the tool, which includes requirements and usage information, is found \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/AHP2020/LBM/README.txt}{here}.

\newpage
\section{Results - DONE SECTION 1st DRAFT}\label{qualityresults}

This section presents the AHP results for qualities that were quantitatively measured. Qualitiative data from interviews with some developers is also found in this section. Working quality definitions are found in section \ref{softwarequalities}.

\subsection{Installability}

All of the software packages that were tested have installation instructions. As noted previously, many of the 23 software packages removed from the original list of 46 software packages are missing documentation, including installation instructions.

All 23 packages can be installed on some Unix-like systems. Seven packages could be installed on Windows, and 5 on macOS. Operating system compatibility is found in the documentation of 19 software packages. For this state of the practice assessment, all but one of the software packages were tested on Ubuntu. The remaining package was tested on CentOS.

Of the software packages that were tested, most installation instructions are located in one place, often in an instruction manual or on a web page. Sometimes incomplete installation instructions are found on a home page, with more detailed instructions located on another web page or within formal documentation. Maintainability and correctness of these instructions could be improved if they were limited to one location.

All but two of the software packages have an automation procedure for at least part of the installation process. Most of these packages use make to automate the installation. A couple of packages use custom scripts for automation.

Errors encountered during the installation process were sometimes quickly fixed thanks to descriptive error messages. At other times, error messages were vague and difficult to troubleshoot. Fourteen software packages definitively broke during installation. A couple of packages did not provide a definitive message of the success or failure of installation. In these instances validating the installation required performing a tutorial or running a script, as described below, if they were available. The installation of most software packages could be recovered if a descriptive error message was displayed. Only three software packages that displayed a descriptive error message were not recoverable, and most of these instances were due to hardware and operating system incompatibility, such as the requirement of CUDA.

About half of the installation instructions are not written as if the person doing the installation has none of the dependent packages installed, and many software packages do not list all of their dependencies. Many dependencies needed to be manually installed. Sometimes only an error message during the installation process informs the user of the requirement of these packages. A detailed rewrite of the installation instructions from the point of view of installation on a clean operating system is suggested.

Sixteen software packages require less than 10 dependencies to be installed. All but one software package require less than 20 dependencies. Some packages may automatically install additional dependencies in the background. Eighteen of the software packages do not explicitly indicate software dependency versions. Some software package installation issues, specifically those occurring when manual installation of dependencies is required, may be avoided if the dependency versions are specified. Fifteen software packages do not have detailed instructions for installing dependencies.

Sixteen software packages have less than ten manual installation steps. If commands for the installation of dependencies is streamlined then none of the software packages take more than 20 steps to install. The average number of steps is about 8, and the fewest is 3.

All but six of the software packages have a way to verify the installation. Twelve have some sort of tutorial examples that can be run by the user. Most of these generate output that can be compared to the documentation. Some packages do not have a way to verify the output of the tutorials, and the user is left to assume correctness if there is no error prompted during the tutorial. Some other ways of installation validation include validation scripts, automatic validation after the installation, and instructions to manually review the file system.

Instructions for uninstallation is easily found for only one of the software packages.

\begin{figure}[h!]
	\begin{center}
		\includegraphics[width=1.0\textwidth]{installability_chart}
		\caption{AHP Installability Score}
		\label{Fig_Installability}
	\end{center}
\end{figure}

Figure \ref{Fig_Installability} shows the installability ranking of the software packages using AHP. Software packages with a higher score tend to have one set of linear installation instructions that is written as if the person doing the installation has none of the dependencies installed. The instructions often have compatible operating system versions listed and include instructions for the installation of dependencies. The top ranked packages also often incorporate some sort of the automation in the installation process and have fewer manual installation steps. The number of dependencies a package has does not correlate with a higher score. Installation validation, often through test examples with expected output, is correlated with a higher score.

Many software packages would benefit from a rewrite or reorganization of installation instructions. A single location for installation instructions would improve their maintainability and correctness. Listing compatible operating system and dependency versions would decrease installation time and errors, as would adding instructions on installing dependencies. Errors that may be encountered during the installation process need to prompt detailed messages or logs. Once a software package is installed, either a transparent automatic validation needs to be performed or the user needs to be able to perform a manual validation using test examples and output. Finally, uninstallation instructions should be included in the documentation.

\subsection{Surface Correctness and Verifiability}

Sixteen of the software packages have a requirement specification or explicitly reference domain theory. This information is often found in a user manual, on a web page, or is mentioned in publications by the developers. In the latter case the user may need to spend significant time to find this information.

Interviews with developers confirmed that many of the software packages were developed by domain experts with backgrounds in physics, mathematics and mechanical engineering. Many did not have formal software engineering education. Some development teams include computer scientists. Despite a lack of visible domain documentation and a resulting lower surface correctness and verifiability score, many software packages do have a strong domain background.

Document generation tools are explicitly used by 12 software packages. Sphinx is used by 8 of them, and Doxygen is used by 7. Some of the packages use both.

Tutorials are available for all but 5 of the software packages. Generally they are linearly written and easy to follow. However, only 8 tutorials provided an expected output and it is not possible to verify the correctness of the output of the software packages that are missing this key information. In this case the user cannot be sure of correctness and may need to assume it if there are no immediately visible errors.

Unit tests are only explicitly available for one of the software packages. Code modularization of most packages allows for a user to create tests with varying degrees of effort. These tests allow developers and users to verify the correctness of fragments of the source code, and in doing so better assess the correctness of the entire package.

The documentation or presence of continuous integration tools and techniques alludes to a more refined development process where faults are isolated and better recognized. On the surface only 2 of the packages mentioned applying the practice of continuous integration in their development process.

Interviews with the developers of software packages suggest a more frequent use of both unit testing and continuous integration in the development processes than initially surveyed. For example, OpenLB, pyLBM, and TCLB use such methods during development despite it not being explicitly clear from a surface analysis.

\begin{figure}[h!]
	\begin{center}
		\includegraphics[width=1.0\textwidth]{correctnessverifiability_chart}
		\caption{AHP Surface Correctness and Verifiability Score}
		\label{Fig_CorrectnessVerifiability}
	\end{center}
\end{figure}

Figure \ref{Fig_CorrectnessVerifiability} shows the surface correctness and verifiability ranking of the software packages using AHP. Software packages with a higher score tend to have a visible requirements specification or references to theory documentation. They also explicitly use at least one document generation tool that builds confidence of correctness. The top ranked software packages all include an easy to follow getting started tutorial in their documentation, and most provide expected output for these tutorials. Only the top ranked package, Ludwig, provided unit testing. It and the second ranked package, ESPResSo, explicitly incorporated continuous integration in the development process. During developer interviews it was discovered that some other packages also use continuous integration and unit testing, despite not being explicitly stated in documentation. Such packages, including OpenLB, pyLBM, and TCLB, have higher true correctness and verificablity than is noted from surface analysis.

The inclusion of requirements specification and theory documentation greatly benefits the correctness and verifiability of software packages. The use of document generation tools can help build confidence in correctness. The addition of easy to follow tutorials, along with expected output, further helps users verify the software and have confidence in its correctness. Unit testing documentation and capability, as well as the use of continuous integration tools and techniques such as Bamboo, Jenkins, and Travis CI, help verify correctness. Several interviewed developers alluded to difficulty with testing large numbers of features, and some even manually tested output. The use of well defined unit testing tools could decrease the time spent testing some feature.

\subsection{Surface Reliability}

Errors occurred when installing 16 of the software packages, and every instance prompted a message. Error messages included unrecognized commands (even when following the installation guide), missing links, missing dependencies, syntax errors in code files, and vague output. Several automatic installation processes could not find and load dependency packages. The installation files pointed to outdated external repositories. Seven of the installations were clearly recovered and one of the installations was assumed to be recovered due to the absence of installation verification. The installation of eight of the software packages could not be recovered. Most of these installations could not reach a dependency repository, had an incompatible error in their core installation files, or prompted vague error messages.

Of the 13 software packages that installed correctly and had tutorials, 4 broke during tutorial testing. All four error resulted in an error message being displayed. One error was due to a missing tutorial dependency, another was due to an invalid command despite following the tutorial, and the final two errors were execution errors. Of the 4 broken tutorial instances, only the one that was missing a dependency was recoverable.

\begin{figure}[h!]
	\begin{center}
		\includegraphics[width=1.0\textwidth]{reliability_chart}
		\caption{AHP Surface Reliability Score}
		\label{Fig_Reliability}
	\end{center}
\end{figure}

Figure \ref{Fig_Reliability} shows the surface reliability ranking of the software packages using AHP. Software packages with a high score did not break during installation, or the installation was recoverable. All of the top 5 ranked packages had tutorials. One of these packages, pyLBM, broke during tutorial testing, but a descriptive error message helped in recovery.

Overall, lower ranked software packages were lacking clear documentation, testing or tutorial examples, and descriptive error messages, and had broken automatic dependencies. Thus, regarding surface reliability, software packages would benefit from clear up-to-date installation and tutorial documentation, and descriptive error messages.

\subsection{Surface Robustness}

The software packages were tested to handle unexpected input, including incorrect data types, empty input, and missing files or links. Success predicated on a reasonable response from the system, including appropriate error messages and an absence of unrecoverable system failures.

\begin{figure}[h!]
	\begin{center}
		\includegraphics[width=1.0\textwidth]{robustness_chart}
		\caption{AHP Surface Robustness Score}
		\label{Fig_Robustness}
	\end{center}
\end{figure}

Figure \ref{Fig_Robustness} shows the surface robustness ranking of the software packages using AHP. Software packages with a high score behaved reasonably in response to unexpected input as described above. All of the software packages that installed correctly passed this test. They output descriptive error messages or did not crash. Software packages with a lower surface robustness score had not installed correctly, so their score here may not be a true reflection of runtime robustness. Similarly, all software packages that installed correctly and require plain text input files handled an unexpected change, a replacement of new lines with carriage returns, to these input files correctly.

\subsection{Surface Performance}

Each software package was not quantitatively measured for surface performance.
Although the software packages all apply Lattice Boltzmann methods to solve scientific computing problems, the packages focus on varied domains, with varying parameters, and are technically different from each other. Due to this, a comparison of performance is not appropriate. Instead, in this project we looked through each software package's artifacts for evidence that performance was considered. The artifacts of 17 software packages mentioned the use of some sort of parallelization for execution. This included GPU processing and the CUDA parallel computing platform. GPUs provide superior processing power and speed compared to CPUs, and are often used for scientific computing when a large amount of data is involved. The software package TCLB is implemented in a highly efficient multi-GPU code to achieve performance suitable for model optimization \citep{rutkowski2020open}. In the Ludwig package, a so-called mixed mode approach is used where fine-grained parallelism is implemented on the GPU, and MPI is used for even larger scale parallelism \citep{gray2013ludwig}. While some software packages required CUDA and GPU processing, some had the option of using either the GPU or the CPU. The packages that require GPU and CUDA have better performance at the expense of installability and surface reliability.

\subsection{Surface Usability}

Software package artifacts were reviewed for the presence of a tutorial, a user manual, documented user characteristics, and a user support model. In total 18 software packages have a tutorial and 13 have a user manual. The tutorials vary in scope and substance, and 8 include an expected output to compare user results to. Most user manuals are in the form of a file that can be downloaded, while some are rendered on a web-page. Some packages do not have a user manual but do have useful documentation distributed on their web-pages. Expected user characteristics are documented in 4 software packages. Users are typically scientists or engineers trying to test a new application. Their background is often physics, chemistry, biophysics, or mathematics. All but one of the packages have a user support model, and several of them have multiple avenues of user support. The most popular avenue of support is git, followed by email and forums. One software package has a FAQ page.

\begin{figure}[h!]
	\begin{center}
		\includegraphics[width=1.0\textwidth]{usability_chart}
		\caption{AHP Surface Usability Score}
		\label{Fig_Usability}
	\end{center}
\end{figure}

Figure \ref{Fig_Usability} shows the surface usability ranking of the software packages using AHP. Software packages with a high score have a tutorial and user manual, and sometimes have documented user characteristics, and have at least one user support model, with many having several.

Interviews with developers revealed several usability issues. Some users have misunderstood boundaries of LBM and computational fluid dynamics in general, and have combined or applied methods that are not physically sound together. Sometimes users have applied LBM to poorly defined or inappropriate fluid dynamics problems. The users do not realize the limitations of the methods, of the software, and of how much time and effort is required to properly model a problem that can technically be modeled by the software. The software is not as easy as out of the box plug and play. Developers of some of the software packages mitigated this by editing the source code to prevent unsound combination from being applied, and by updating the documentation to better inform users of LBM limitations, and of the requirements to properly model appropriate problems, including what algorithms and parameters to use.

Some additional infrequent software usability issues were commented on. Users have had trouble with installation and understanding how to maneuver the interfaces and how to set up or run models. These issues are addressed by various user support models, including frequently asked questions sections on the software package websites, user guides, and hardware and software requirement specifications.

One software package changed its scripting language to Python to make it more usable. The developer commented that this was the biggest step in addressing usability, commenting that Python was a much easier language for users to learn and understand.

\subsection{Maintainability}

Software packages were reviewed for the presence of artifacts. Every type of artifact or file that is not a code file was recorded. The software packages were also reviewed for version numbers. This information could be used to better troubleshoot software issues and organize documentation. All but 3 software packages have source code release and documentation version numbers.

Information on how code is reviewed, or how to contribute to the project was also noted. In total, eleven software packages have this information, which was found in various artifacts, including in developer guides, contributor guides, user guides, developer web-pages, and README files.

Issue tracking is used in 22 software packages, 15 of which use git, 6 use email, and one uses SourceForge. Most software packages that use git have most of their issues closed, and only 3 have less than 50 percent of their issues closed. Furthermore, most packages that use git for issue tracking use GitHub as a version control system, one package uses CVS for issue tracking, and 8 packages do not appear to use any such system.

Software package code files were further measured for the percentage of code that is comments. Packages with a higher percentage of comments were designated as more maintainable. Comments represent more than 10 percent of code files in 15 packages, and the average percentage of comments is 13.7.

\begin{figure}[h!]
	\begin{center}
		\includegraphics[width=1.0\textwidth]{maintainability_chart}
		\caption{AHP Maintainability Score}
		\label{Fig_Maintainability}
	\end{center}
\end{figure}

Figure \ref{Fig_Maintainability} shows the maintainability ranking of the software packages using AHP. Software packages with a high score provide version numbers on documents and source code releases, have an abundance of high quality artifacts, and use an issue tracking tool and version control system. These packages also appear to reasonably handle tracked issues, having most of their issues closed. Their code files are well commented with more than 10 percent of the code being comments.

Interviews with developers revealed that most projects do have a defined process for accepting contributions from current team members or those that at some time had been linked to the packages or developers. The packages rarely get contributions from outside developers, but the process would be similar as for the aforementioned group.

Contributions are made through GitHub, and are then reviewed and pulled by lead developers, often with consultation with a group of core developers depending on the organizational model. Continuous integration is part of the process for some packages.

Some developers noted that their software package does not have well defined contributing guide in the repository but it might be a good idea to add one in the near future. They would be very happy to see spontaneous contribution from outside of their organization, but now it is currently not the case.

Furthermore, maintainability has been addressed by increasing source code modularity, reducing duplicate information, and improving abstraction. Several software packages have had large sections of their code base redeveloped with languages that the developers felt are more understandable and readable, and that are better supported and have preferred dependencies. Data structures have also been redeveloped and storage has been improved. One software developer mentioned that the mathematical foundations that the geometries and models are designed from have been re-evaluated and optimized.

\subsection{Modifiability}

Software packages were not quantitatively measured for modifiability.
In this project we asked developers to comment on modifiability when we interviewed them. Specifically, we asked if ease of future changes to the system, modules, and code blocks was considered when designing the software. We also asked if any measures had been taken to ensure the ease of future changes. 
All of the developers that were interviewed noted that the ease of future changes was considered and that measures to ensure it had been taken.

A high degree of code modularity and abstraction was noted by one developer as a measure to ensure ease of future changes. They also noted that some of the code base was transitioned from c to c++, which could ease modifiability of that software package. Another developer noted that their software package was designed to easily allow for the addition of some LBM features, but changes to major aspects of the LBM models would be difficult. Furthermore, the package was designed with flexible data structures and storage in mind.

Some software packages, like Palabos, provide validation benchmarks for their core fundamental algorithmic ingredients \citep{latt2021palabos}. The stated intent of these benchmarks is to showcase the validity and usefulness of the package in order to stimulate the development of third-user extensions. The Palabos package identifies as as a development framework for modeling problems in various computational fluid dynamics areas.

\subsection{Reusability}

The total number of source code files was measured for each software package. A larger number of source files was associated with increased reusability since this generally represented either more software package functionality or better modularization. The software packages were also reviewed for the presence of API documentation, which indicates that a software package was developed with interaction between other software applications in mind.

\begin{figure}[h!]
	\begin{center}
		\includegraphics[width=1.0\textwidth]{reuseability_chart}
		\caption{AHP Reusability Score}
		\label{Fig_Reusabilty}
	\end{center}
\end{figure}

Figure \ref{Fig_Reusabilty} shows the reusability ranking of the software packages using AHP. Software packages with a high score have thousands of source code files and API documentation. The highest scoring packages, ESPResSo and waLBerla, have extensive functionality, including graphical visualizations as well as modeling that does not use LBM. For this reason a comparison between other software packages is not on a level field, but regardless of this these packages do have an abundance of reusable components.

There was a strong focus on modularity when designing the waLBerla framework in order to enhance productivity, reusability, and maintainability \citep{bauer2021walberla}. The software design has enabled for waLBerla to be successfully applied in several projects as a basis for various extensions \citep{bauer2021walberla}.

\subsection{Surface Understandability}

Ten random source code files of each software package were reviewed for several measures. This measure of surface understandability may not perfectly reflect each package due to this limitation. All of the packages appear to have a consistent indentation and formatting style. Only LUMA and HemeLB explicitly identify coding standards that are used during development. Generally, the software packages use consistent, distinctive, and meaningful code identifiers. Only 4 packages appear to use vague identifiers. Hard coded constants were observed in the source code of 12 packages. The constants are used for various parameters, mathematical constants, and matrix definitions. All of the packages are well commented. Comments clearly indicate what is being done. Domain algorithms are noted in the source code of 11 packages. All of the packages were well modularized. When observing the source code files, it was found that 13 of the packages had a consistent style and order for function parameters.

\begin{figure}[h!]
	\begin{center}
		\includegraphics[width=1.0\textwidth]{understandability_chart}
		\caption{AHP Understandability Score}
		\label{Fig_Understandability}
	\end{center}
\end{figure}

Figure \ref{Fig_Understandability} shows the surface understandability ranking of the software packages using AHP. Software packages with a high score have a consistent indentation and formatting style, and consistent, distinctive and meaningful code identifiers. They also have hard coded constants, and explicitly identify mathematical and LBM algorithms. Their comments are clear and indicative of what is being done. The source code is well modularized and structured.

Software developers noted that in their opinion users have generally found their packages to be understandable. One developer commented that some users have attempted to run physically incompatible LBM methods, and the solution was to edit the code to prevent such combinations, as well as update the documentation to prevent misunderstanding the methods. Similarly, another developer noted that some users had issues setting up parameters for LBM schemes. The solution was to this was to update the interface where these parameters are set, as well as adding functionality to test the stability of the parameters. A third developer noted that some users lack the background knowledge to easily model flow problems using their software. A frequently asked questions section was added to the source package website to help users find answers to common questions. The package also has detailed documentation, including guides and usage requirements specification.

\subsection{Traceability}

Software packages were not quantitatively measured for traceability.
In this project we asked developers to comment on traceability when we interviewed them. They were asked to comment on their documentation and how it fits into their development process.

One developer noted that all major additions to their package had accompanying changes to artifacts, especially noting user documentation, and how it is updated to note major usability issues. Considerable effort had been put into the documentation. They want to lower the entry barrier for new developers so there is considerable documentation aimed at developers, and it is frequently updated. This documentation informs developers on how to get started, orients them to the artifacts, source code, and system architecture, as well as how the software package build system works, and how the coupling between the simulation engine and the interface works.

Another developer noted the importance of documentation for both user and developers of their software. New features are always well documented in the artifacts. The developers use documentation to stay up to date on the status of the software package, and to expand or use features, like computational models or algorithms. This is necessary so that the coding standard for these models is kept consistent as developers come and go.

The importance of documentation for both users and developers was stressed throughout the interviews. However, it was noted several times that a lack of time and funding had negative affect on the documentation. Most of these developers are scientific researchers evaluated on the scientific papers that they produce. Writing and updating documentation is something that is done in free time, if that time arises, and if writing documentation is a concern. Sometimes it is a last priority for the developers. Finding ways to hasten updating documentation would increase the frequency of such updates and benefit both users and developers.

One developer noted the use of documentation generators like Doxygen. It would be advisable for more projects to use such automatic document generation tools, if they are a good technological fit.

\subsection{Visibility and Transparency}

Software package artifacts were reviewed for the identification of a specific development model, like a waterfall of agile development model, and the presence of documentation recording the development process and standard. They were also reviewed for the identification of the development environment, and the presence of release notes. The packages tended to not explicitly use well-known development models. This was also noted in the interviews with developers detailed below. The development teams of these packages are fairly small and easily organized without the need for such processes. Seven of the software packages did have some artifacts outlining the general development process, how to contribute, and the status of the package or its components. Eight of the packages have artifacts that note the development environment. While this information could help developers, and would improve transparency, the small close-knit nature of the development teams make explicitly publicly specifying this information practically unnecessary. Version release notes were found in 9 of the software packages.

\begin{figure}[h!]
	\begin{center}
		\includegraphics[width=1.0\textwidth]{visibilitytransparency_chart}
		\caption{AHP Visibility and Transparency Score}
		\label{Fig_VisibilityTransparency}
	\end{center}
\end{figure}

Figure \ref{Fig_VisibilityTransparency} shows the visibility and transparency ranking of the software packages using AHP. Software packages with a high score have an explicit development model and defined development process. They also 
had detailed and easy to access notes accompanying software releases.

Developers were asked to comment on the obstacles in their development process. One developer noted that a lot of their source code had been written with a specific application in mind, and that there is too much coupling between components. Addressing this issue would help with code modifiability and reusability. Updating the development process would help resolve this issues and prevent such issues in the future. Improving the visibility of software changes and the peer review process would also help. Improving the software engineering education or experience of developers is also an idea that was brought up by several developers. Specifically, developers should always write code that is decoupled and modular, and should keep in mind the visibility of their contributions by better updating documentation and ensuring that their contributions are transparent to the rest of the developers. This would help catch issues in the code contributions, and improve source code maintenance.

According to one developer, some obstacles to the development processes of their package had been overcome by the introduction of continuous integration practices, and a peer review process for contributions. These practices decrease development and maintenance times. Another developer mentioned that their package had two sets of code, for executing the models on the CPU and GPU, and that maintenance was decreased by introducing macros, the use of which is now part of the development process.

Developers were also asked how documentation fits into their development process. Several developers noted that developer documentation plays an important role in familiarizing potential contributors to the software system architecture. Without the guidance that the documentation provides it would be unlikely that contributions would pass the peer review process.

None of the software packages whose developers were interviewed have a formal software development model. The packages all have fairly small development teams. These teams do accept outside contributors, but generally the teams are tight-knit, often working at the same institution, although one of the packages has an international team. One developer noted that while no formal model is used, their development model is something similar to a combination of agile and waterfall development models.

The developers noted similar project management processes. In teams of only a couple of developers, the addition of new features or major changes were discussed with the entire team. Projects with more than a couple of developers had lead developer roles. These lead developers would review potential additions to the software. The software packages use GitHub for managing the project. Typically there are several development branches as well as the master branch.

\subsection{Reproducibility}

Software packages were not quantitatively measured for reproducibility.
In this project we asked developers to comment on reproducibility when we interviewed them.

Developers were asked if they have any concern that their computational results would not be reproducible in the future, and if they had taken any steps to ensure reproducibility.

One developer noted that they compare the results of their methods against manually calculated results. These tests comparing these results are automatically run for all source code changes that are suggested. The tests are run when a pull request is opened on GitHub. Even once these tests are complete, a peer review process is done before changes are fully committed to the appropriate branch. The results for all of the LBM schemes on the software package development branch are also compared daily for correctness, ensuring that the system output reflects the expected output.

Several developers noted that they currently do not have a system in place to test for reproducibility, but it is of interest and could be implemented in the future. Generally, the mathematical foundations of the models are verified, but the output of the software packages is not compared to other output. Depending on the package and how it outputs LBM solutions, it may not be practical or feasible. A correct output may not be exactly reproducible, as it may be dependent on a probability distribution, so strictly comparing results may not be appropriate.

The source code and artifacts of some software packages may be reproducible. There is considerable variance in the quality of the software specification and other developer documentation. Some packages are well detailed, and translating the specification into source code will produce similar results across developers.

\subsection{Unambiguity}

Software packages were not quantitatively measured for unambiguity.
In this project we asked developers to comment on unambiguity when we interviewed them.

Developers were asked if they thought that the current documentation can clearly convey all necessary knowledge to the users, and if they had taken any steps to ensure clarity.

One developer noted that their documentation was meant for users that are already familiar with underlying physics and computational fluid dynamics methods, and that these concepts are not explained in detail. The documentation focuses solely on how to technically use the software package, and includes a user guide and tutorial walk through of how to set up and run a simulation. With this in mind the developer believes that their documentation is in reasonable shape for a those with a minimum knowledge of the underlying physics to learn to use the package. If new users have technical questions these can be addressed in further revisions of the documentation. New developers should also find that the code documentation is reasonably clear and useful. Information that is missing, like information explaining dependency packages, is reasonably referenced.

Another developer also noted that their documentation was in reasonable shape, but they would really appreciate more user feedback to improve it. They also noted that they believe a lack of knowledge of the underlying physics and computational fluid dynamics concepts can be an issue for some users. This information can be referenced in the documentation, but it is not something that the documentation needs to detail.

\subsection{Overall Quality}

Figure \ref{Fig_OverallScore} shows the overall ranking of the software packages using AHP. Software packages with an overall high score had ranked high in at least several of the individual qualities that were quantitatively measured.

\begin{figure}[h!]
	\begin{center}
		\includegraphics[width=1.0\textwidth]{finalscore_chart}
		\caption{AHP Overall Score}
		\label{Fig_OverallScore}
	\end{center}
\end{figure}

Looking at the top three ranked packages. ESPResSo had achieved a relative high score in installability, surface correctness and verifiability, surface usability, maintainability, reusability, and visibility and transparency. Ludwig scored high in surface correctness and verifiability, surface robustness, surface usability, maintainability, and visibility and transparency. Palabos scored high in installability, surface reliability, surface robustness, maintainability, and understandability.

Section \ref{qualityrecommentations} further analyzes these findings and offers some software quality recommendations for future development of LBM software.

\newpage
\section{Answers To Research Questions}\label{answersquestions}

\subsection{Artifacts Present}

\subsection{Tools Used}

\subsection{Principles, Processes, and Methodologies}

\subsection{Pain Points}

\subsection{Quality Recommendations}\label{qualityrecommentations}

create section for each quality?

consider what worked for some of the packages

\subsection{Designation Comparison}

answers last question of research questions:

How does software designated as high quality by this methodology compare	with top rated software by the community?

\newpage
\section{Conclusion}

- key findings restated
- comment on future work

\newpage
\section{Appendix}
\begin{singlespace}

\subsection{Research Questions}\label{researchquestions}
	The following are the research questions that are considered.

	\begin{enumerate}
		\item What artifacts are present in current software packages? 
		\item What tools (development, dependencies, project management) are used by current software packages?
		\item What principles, processes, and methodologies are used in the development
		of current software packages?
		\item What are the pain points for developers working on research software
		projects?  What aspects of the existing processes, methodologies and tools do
		they consider as potentially needing improvement?  How should processes,
		methodologies and tools be changed to improve software development and
		software quality?
		\item For research software developers, what specific actions are taken to
		address the following:
		\begin{enumerate}
			\item installability
			\item correctness and verifiability
			\item reliability
			\item robustness
			\item performance
			\item usability
			\item maintainability
			\item modifiability
			\item reusability
			\item understandability
			\item traceability
			\item visibility and transparency
			\item reproducibility
			\item unambiguity
		\end{enumerate} 
		\item How does software designated as high quality by this methodology compare	with top rated software by the community?
	\end{enumerate}

\newpage

\subsection{Measurement Template}
The table below lists the set of measures that are used to assess each software product. The first set identifies summary information, followed by 9 sets for software qualities and 3 sets for raw metrics. Each measure is followed by the type for a valid result. A superscript ∗ indicate that a response of this type needs to be accompanied by explanatory text.

\captionof{table}{Measurement Template}\label{measurementtemplate}
\def\arraystretch{1.22}
\begin{tabular}{p{16cm}}
		\hline
		\textbf{Summary Information}\\
		\hline
		Software name? (string)\\
		URL? (URL)\\
		Affiliation (institution(s)) (string or {N/A})\\
		Software purpose (string)\\
		Number of developers (all developers that have contributed at least one commit to the project) (use repo commit logs) (number)\\
		How is the project funded? (unfunded, unclear, funded*) where * requires a string to say the source of funding\\
		Initial release date? (date)\\
		Last commit date? (date)\\
		Status? (alive is defined as presence of commits in the last 18 months) ({alive, dead, unclear})\\
		License? ({GNU GPL, BSD, MIT, terms of use, trial, none, unclear, other*}) * given via a string \\
		Platforms? (set of {Windows, Linux, OS X, Android, other*}) * given via string\\
		Software Category? The concept category includes software that does not have an officially released version. Public software has a released version in the public domain. Private software has a released version available to authorized users only. ({concept, public, private})\\
		Development model? ({open source, freeware, commercial, unclear})\\
		Publications about the software? Refers to publications that have used or mentioned the software. (number or {unknown})\\
		Source code URL? ({set of url, n/a, unclear})\\
		Programming language(s)? (set of {FORTRAN, Matlab, C, C++, Java, R, Ruby, Python, Cython, BASIC, Pascal, IDL, unclear, other*}) * given via string \\
		Is there evidence that performance was considered? Performance refers to either speed, storage, or throughput. ({yes∗, no})\\
		Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have) \\
		\hline
\end{tabular}

\def\arraystretch{1.5}
\begin{tabular}{p{16cm}}
		\hline		
		\textbf{Installability  (Measured via installation on a virtual machine.) }\\
		\hline
		Are there installation instructions? ({yes, no})\\
		Are the installation instructions in one place? Place referring to a single document or web page. ({yes, no, n/a})\\
		Are the installation instructions linear? Linear meaning progressing  in a single series of steps. ({yes, no, n/a})\\
		Are the instructions written as if the person doing the installation has none of the dependent packages installed? ({yes, no, unclear})\\
		Are compatible operating system versions listed? ({yes, no})\\
		Is there something in place to automate the installation (makefile, script, installer, etc)? ({yes*, no})\\
		If the software installation broke, was a descriptive error message displayed? ({yes, no, n/a})\\
		Is there a specified way to validate the installation? ({yes*, no})\\
		How many steps were involved in the installation? (Includes manual steps like unzipping files) Specify OS. (number, OS)\\
		What OS was used for the installation? ({Windows, Linux, OS X, Android, other* }) *given via string\\
		How many extra software packages need to be installed before or during installation? (number)\\
		Are required package versions listed? ({yes, no, n/a})\\
		Are there instructions for the installation of required packages / dependencies? ({yes, no, n/a})\\
		Run uninstall, if available. Were any obvious problems caused? ({yes∗ , no, unavail})\\
		Overall impression? ({1 .. 10})\\
		Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have)\\
		\hline
\end{tabular}

\def\arraystretch{1.33}
\begin{tabular}{p{16cm}}
	\hline		
	\textbf{Correctness and Verifiability}\\
	\hline
	Any reference to the requirements specifications of the program or theory manuals? ({yes∗ , no, unclear})\\
	What tools or techniques are used to build confidence of correctness? ({literate programming, automated testing, symbolic execution, model checking, assertions used in the code, Sphinx, Doxygen, Javadoc, confluence, unclear, other*}) * given via string\\
	If there is a getting started tutorial? ({yes, no})\\
	Are the tutorial instructions linear? ({yes, no, n/a})\\
	Does the getting started tutorial provide an expected output? ({yes, no*, n/a})\\
	Does your tutorial output match the expected output? ({yes, no, n/a})\\
	Are unit tests available?  ({yes, no, unclear})\\
	Is there evidence of continuous integration? (for example mentioned in documentation, Jenkins, Travis CI, Bamboo, other) ({yes*, no, unclear})\\
	Overall impression? ({1 .. 10})\\
	Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have) \\
	\hline	
	\textbf{Surface Reliability}\\
	\hline
	Did the software “break” during installation? ({yes∗ , no})\\
	If the software installation broke, was the installation instance recoverable? ({yes, no, n/a})\\
	Did the software “break” during the initial tutorial testing? ({yes∗, no, n/a})\\
	If the tutorial testing broke, was a descriptive error message displayed? ({yes, no, n/a})\\
	If the tutorial testing broke, was the tutorial testing instance recoverable? ({yes, no, n/a})\\
	Overall impression? ({1 .. 10})\\
	Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have)\\
	\hline		
	\textbf{Surface Robustness}\\
	\hline
	Does the software handle unexpected/unanticipated input (like data of the wrong type, empty input, missing files or links) reasonably? (a reasonable response can include an appropriate error message.) ({yes, no∗ })\\
	For any plain text input files, if all new lines are replaced with new lines and carriage returns, will the software handle this gracefully? ({yes, no∗, n/a})\\
	Overall impression? ({1 .. 10})\\
	Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have)\\
	\hline
\end{tabular}

\def\arraystretch{1.4}
\begin{tabular}{p{16cm}}
	\hline		
	\textbf{Surface Usability}\\
	\hline
	Is there a getting started tutorial? ({yes, no})\\
	Is there a user manual? ({yes, no})\\
	Are expected user characteristics documented? ({yes, no})\\
	What is the user support model? FAQ? User forum? E-mail address to direct questions? Etc. (string)\\
	Overall impression? ({1 .. 10})\\
	Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have)\\
	\hline	
	\textbf{Maintainability}\\
	\hline
	What is the current version number? (number)\\
	Is there any information on how code is reviewed, or how to contribute? ({yes*, no})\\
	Are artifacts available? (List every type of file that is not a code file – for examples please look at the ‘Artifact Name’ column of https://gitlab.cas.mcmaster.ca/SEforSC/se4sc/-/blob/git-svn/GradStudents/Olu/ResearchProposal/Artifacts\_MiningV3.xlsx) ({yes*, no, unclear}) *list via string\\
	What issue tracking tool is employed? (set of {Trac, JIRA, Redmine, e-mail, discussion board, sourceforge, google code, git, BitBucket, none, unclear, other*}) * given via string\\
	What is the percentage of identified issues that are closed? (percentage)\\
	What percentage of code is comments? (percentage)\\
	Which version control system is in use? ({svn, cvs, git, github, unclear, other*}) * given via string\\
	Overall impression? ({1 .. 10})\\
	Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have)\\
	\hline		
	\textbf{Reusability}\\
	\hline
	How many code files are there? (number)\\
	Is API documented? ({yes, no, n/a})\\
	Overall impression? ({1 .. 10})\\
	Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have)\\
	\hline
\end{tabular}

\def\arraystretch{1.4}
\begin{tabular}{p{16cm}}
	\hline		
	\textbf{Surface Understandability (Based on 10 random source files)}\\
	\hline
	Consistent indentation and formatting style? ({yes, no, n/a})\\
	Explicit identification of a coding standard? ({yes∗, no, n/a})\\
	Are the code identifiers consistent, distinctive, and meaningful? ({yes, no∗ , n/a})\\
	Are constants (other than 0 and 1) hard coded into the program? ({yes, no∗ , n/a})\\
	Comments are clear, indicate what is being done, not how? ({yes, no∗ , n/a})\\
	Is the name/URL of any algorithms used mentioned? ({yes, no∗ , n/a})\\
	Parameters are in the same order for all functions? ({yes, no∗ , n/a})\\
	Is code modularized? ({yes, no∗ , n/a})\\
	Overall impression? ({1 .. 10})\\
	Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have)\\
	\hline		
	\textbf{Visibility/Transparency}\\
	\hline
	Is the development process defined? If yes, what process is used. ({yes∗, no, n/a})\\
	Are there any documents recording the development process and status?  ({yes∗, no}))\\
	Is the development environment documented? ({yes∗, no})\\
	Are there release notes? ({yes∗, no})\\
	Overall impression? ({1 .. 10})\\
	Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have)\\
	\hline		
	\textbf{Raw Metrics (Measured via git\_stats)}\\
	\hline
	Number of text-based files. (number)\\
	Number of binary files. (number)\\
	Number of total lines in text-based files. (number)\\
	Number of total lines added to text-based files. (number)\\
	Number of total lines deleted from text-based files. (number)\\
	Number of total commits. (number)\\
	Numbers of commits by year in the last 5 years. (Count from as early as possible if the project is younger than 5 years.) (list of numbers)\\
	Numbers of commits by month in the last 12 months. (list of numbers)\\
	\hline
\end{tabular}

\def\arraystretch{1.4}
\begin{tabular}{p{16cm}}
	\hline		
	\textbf{Raw Metrics (Measured via scc)}\\
	\hline
	Number of text-based files. (number)\\
	Number of total lines in text-based files. (number)\\
	Number of code lines in text-based files. (number)\\
	Number of comment lines in text-based files. (number)\\
	Number of blank lines in text-based files. (number)\\
	\hline
	\textbf{Repo Metrics (Measured via GitHub)}\\
	\hline
	Number of stars. (number)\\
	Number of forks. (number)\\
	Number of people watching this repo. (number)\\
	Number of open pull requests. (number)\\
	Number of closed pull requests. (number)\\
	\hline
\end{tabular}

\newpage

\subsection{Grading Template}
The table below lists how each quality measure of the measurement template is used to calculate an overall impression in each software quality set.\newline

\captionof{table}{Grading Template}\label{gradingtemplate}
\def\arraystretch{1.5}
\begin{tabular}{p{16cm}}
	\hline		
	\textbf{Installability  (Measured via installation on a virtual machine.) }\\
	\hline
	Are there installation instructions? ({yes=1, no=-1})\\
	Are the installation instructions in one place? Place referring to a single document or web page. ({yes=1, no=0, n/a=0})\\
	Are the installation instructions linear? Linear meaning progressing  in a single series of steps. ({yes=1, no=0, n/a=0})\\
	Are the instructions written as if the person doing the installation has none of the dependent packages installed? ({yes=1, no=0, unclear=0})\\
	Are compatible operating system versions listed? ({yes=1, no=0})\\
	Is there something in place to automate the installation (makefile, script, installer, etc)? ({yes*=1, no=-1})\\
	If the software installation broke, was a descriptive error message displayed? ({yes=0, no=-2, n/a=1})\\
	Is there a specified way to validate the installation? ({yes*=1, no=0})\\
	How many steps were involved in the installation? (Includes manual steps like unzipping files) Specify OS. ($<$10 = 1)\\
	What OS was used for the installation? (does not count)\\
	How many extra software packages need to be installed before or during installation? ($<$10 = 1)\\
	Are required package versions listed? ({yes=1, no=0, n/a=1})\\
	Are there instructions for the installation of required packages / dependencies? ({yes=1, no=0, n/a=1})\\
	Run uninstall, if available. Were any obvious problems caused? ({yes∗=0, no=1, unavail=1})\\
	Overall impression? (a sum of $>$10 is rounded down to 10)\\
	\hline
\end{tabular}

\def\arraystretch{1.33}
\begin{tabular}{p{16cm}}
	\hline		
	\textbf{Correctness and Verifiability}\\
	\hline
	Any reference to the requirements specifications of the program or theory manuals? ({yes∗=2, no=0, unclear=0})\\
	What tools or techniques are used to build confidence of correctness? (any=1, unclear=0)\\
	If there is a getting started tutorial? ({yes=2, no=0})\\
	Are the tutorial instructions linear? ({yes=1, no=0, n/a=0})\\
	Does the getting started tutorial provide an expected output? ({yes=1, no*=0, n/a=0})\\
	Does your tutorial output match the expected output? ({yes=1, no=0, n/a=0})\\
	Are unit tests available?  ({yes=1, no=0, unclear=0})\\
	Is there evidence of continuous integration? (for example mentioned in documentation, Jenkins, Travis CI, Bamboo, other) ({yes*=1, no=0, unclear=0})\\
	\hline	
	\textbf{Surface Reliability}\\
	\hline
	Did the software “break” during installation? ({yes∗=0, no=5})\\
	If the software installation broke, was the installation instance recoverable? ({yes=5, no=0, n/a=0})\\
	Did the software “break” during the initial tutorial testing? ({yes∗=0, no=5, n/a=0})\\
	If the tutorial testing broke, was a descriptive error message displayed? ({yes=2, no=0, n/a=0})\\
	If the tutorial testing broke, was the tutorial testing instance recoverable? ({yes=3, no=0, n/a=0})\\
	\hline		
	\textbf{Surface Robustness}\\
	\hline
	Does the software handle unexpected/unanticipated input (like data of the wrong type, empty input, missing files or links) reasonably? (a reasonable response can include an appropriate error message.) ({yes=5, no∗=0})\\
	For any plain text input files, if all new lines are replaced with new lines and carriage returns, will the software handle this gracefully? ({yes=5, no∗=0, n/a=5})\\
	\hline		
	\textbf{Surface Usability}\\
	\hline
	Is there a getting started tutorial? ({yes=3, no=0})\\
	Is there a user manual? ({yes=4, no=0})\\
	Are expected user characteristics documented? ({yes=1, no=0})\\
	What is the user support model? FAQ? User forum? E-mail address to direct questions? Etc. (one=1, two+=2, none=0)\\
	\hline
\end{tabular}

\def\arraystretch{1.4}
\begin{tabular}{p{16cm}}
	\hline
	\textbf{Maintainability}\\
	\hline
	What is the current version number? (provided=1, nothing=0)\\
	Is there any information on how code is reviewed, or how to contribute? ({yes*=1, no=0})\\
	Are artifacts available? (List every type of file that is not a code file – for examples please look at the ‘Artifact Name’ column of https://gitlab.cas.mcmaster.ca/SEforSC/se4sc/-/blob/git-svn/GradStudents/Olu/ResearchProposal/Artifacts\_MiningV3.xlsx) (Rate 0 – 2 depending on how many and perceived quality)\\
	What issue tracking tool is employed? (nothing=0, email of other private=1, anything public or accessible by all devs (eg git) = 2)\\
	What is the percentage of identified issues that are closed? (50$\%$+=1, $<$50$\%$=0)\\
	What percentage of code is comments? (10$\%$+=1, $<$10$\%$=0)\\
	Which version control system is in use? (anything=2, nothing=0)\\
	\hline		
	\textbf{Reusability}\\
	\hline
	How many code files are there? (0-9=0, 10-49=1, 50-99=3, 100-299=4, 300-599=5, 600-999=6, 1000+=8)\\
	Is API documented? ({yes=2, no=0, n/a=0})\\
	\hline	
	\textbf{Surface Understandability (Based on 10 random source files)}\\
	\hline
	Consistent indentation and formatting style? ({yes=1, no=0, n/a=0})\\
	Explicit identification of a coding standard? ({yes∗=1, no=0, n/a=0})\\
	Are the code identifiers consistent, distinctive, and meaningful? ({yes=2, no∗=0, n/a=0})\\
	Are constants (other than 0 and 1) hard coded into the program? ({yes=1, no∗=0, n/a=0})\\
	Comments are clear, indicate what is being done, not how? ({yes=2, no∗=0, n/a=0})\\
	Is the name/URL of any algorithms used mentioned? ({yes=1, no∗=0, n/a=0})\\
	Parameters are in the same order for all functions? ({yes=1, no∗=0, n/a=0})\\
	Is code modularized? ({yes=1, no∗=0, n/a=0})\\
	\hline		
	\textbf{Visibility/Transparency}\\
	\hline
	Is the development process defined? If yes, what process is used. ({yes∗=3, no=0, n/a=0})\\
	Are there any documents recording the development process and status?  ({yes∗=3, no=0}))\\
	Is the development environment documented? ({yes∗=2, no=0})\\
	Are there release notes? ({yes∗=2, no=0})\\
	\hline
\end{tabular}
\newpage

\subsection{Ethics Approval}
\newpage

\subsection{Interview Questions}\label{interviewquestions}

Information about these interview questions:  This gives you an idea what I would like to learn about the development of {domain} software. Interviews will be one-to-one and will be open-ended (not just “yes or no” answers). Because of this, the exact wording may change a little. Sometimes I will use other short questions to make sure I understand what you told me or if I need more information when we are talking such as: “So, you are saying that …?), to get more information (“Please tell me more?”), or to learn what you think or feel about something (“Why do you think that is…?”).

\begin{enumerate}
	\item Interviewees’ current position/title? degrees?
	\item Interviewees’ contribution to/relationship with the software?
	\item Length of time the interviewee has been involved with this software?
	\item How large is the development group?
	\item Do you have a defined process for accepting new contributions into your team?
	\item What is the typical background of a developer?
	\item What is your estimated number of users? How did you come up with that estimate?
	\item What is the typical background of a user?
	\item Currently, what are the most significant obstacles in your development process?
	\item How might you change your development process to remove or reduce these obstacles?
	\item How does documentation fit into your development process? Would improved documentation help with the obstacles you typically face?
	\item In the past, is there any major obstacle to your development process that has been solved? How did you solve it?
	\item What is your software development model? For example, waterfall, agile, etc.
	\item What is your project management process? Do you think improving this process can
tackle the current problem? Were any project management tools used?
	\item Was it hard to ensure the correctness of the software? If there were any obstacles, what methods have been considered or practiced to improve the situation? If practiced, did it work?
	\item When designing the software, did you consider the ease of future changes? For example, will it be hard to change the structure of the system, modules or code blocks? What measures have been taken to ensure the ease of future changes and maintains?
	\item Provide instances where users have misunderstood the software. What, if any, actions were taken to address understandability issues?
	\item What, if any, actions were taken to address usability issues?
	\item Do you think the current documentation can clearly convey all necessary knowledge to the users? If yes, how did you successfully achieve it? If no, what improvements are needed?
	\item Do you have any concern that your computational results won’t be reproducible in the future? Have you taken any steps to ensure reproducibility?

\end{enumerate}

\newpage

\bibliographystyle {plainnat}
\bibliography {References}

\end{singlespace}

\end{document}

### 2021-05-14 StateOfPractice/Peter-Notes/Report/Report.tex
added draft organization of report and purpose of report, and a few other small changes

@@
 \subsection{Purpose of Report}
--constitutes the author’s M. Eng. Report
--It is therefore intended for the examination committee
+
+This document constitutes the author's M.Eng.
+report in partial fulfillment of the the degree of Master of Engineering.
+It is to be submitted to the Department of Computing and Software in McMaster University's Faculty of Engineer and is intended for review by the examination committee.
+In this context the purposes of the report are the following:
+
+\begin{itemize}
+	\item Describe the State of Practice of SCS project origin, motivations, and goals
+	\item Describe LBM software structure and goals
+	\item Report on the measure of a subset of LBM software packages along quality metrics
+	\item Evaluate the state of practice of LBM software development along quality metrics
+	\item Suggest future work for improving LBM software along quality metrics
+	\item Suggest future work for improving State of Practice of SCS assessments
+\end{itemize}
 
 \subsection{Motivation of Report}
+
+simple about SCS ....
 
@@
 \subsection{Scope of Report}
--measured 23 software projects
-- other limitations
+-measured 23 software projects (remove this)
+- other limitations about the study
 - addresses what was done
@@
 \subsection{Organization of Report}
--organization of the document
+
+The first part of this document includes the Table of Contents, List of Figures, and List of Tables.
+This is followed by Reference Material including a subsections outlining Software Engineering Related Definitions and Acronyms and Lattice Boltzmann Related Definitions and Acronyms.
+
+The rest of core of the report is organized as follows:
+
+\begin{itemize}
+	\item \textbf{Introduction} to the report, including its purpose, motivation, scope, and organization.
+	\item \textbf{Background} of Scientific Computing Software, including software qualities and software families.
+	\item LBM software \textbf{Domain Analysis}.
+	\item \textbf{Methodology} of this State of the Practice Assessment, including the steps of the methodology, definitions of software qualities and how they are applied in this assessment, and an overview of candidate software selection, empirical measures, and the analytical hierarchy process.
+	\item Quantitative and qualitative \textbf{Results}. 
+	\item An analysis of the results and \textbf{Answers To Research Questions}.
+	\item \textbf{Conclusion} to the report including comments on future State of Practice assessments.
+	\item \textbf{Appendix}: This section includes our Research Questions, the Measurement Template, Grading Template, Ethics Approval, and Developer Interview Questions.
+\end{itemize}
 
@@
 \section{Background}
+
+this is about scs background
 
@@
 \newpage
-\section{Method - DONE SECTION 1st DRAFT}
+\section{Methodology - DONE SECTION 1st DRAFT}
 
@@
 \newpage
+
 \section{Answers To Research Questions}\label{answersquestions}
@@
 - key findings restated
-- comment on future work
+- comment on future work in State of Practice SCS
 
@@
 
-\subsection{Interview Questions}\label{interviewquestions}
+\subsection{Developer Interview Questions}\label{interviewquestions}
 

### 2021-05-16 StateOfPractice/Peter-Notes/Report/Report.tex
fixed chart typos
done first draft of introduction section

@@
 \doublespacing
-\title{Title} 
+\title{State of the Practice of Lattice Boltzmann Method Software Development} 
 \author{Peter Michalski}
@@
 	\begin{abstract}
-		Abstract here ...
+		This project analyzed the state of the practice for software development in the Lattice Boltzmann Method 
+		domain by quantitatively and qualitatively measuring 23 software packages along quality attributes...
 	\end{abstract}
@@
 \newpage
-\section{Introduction}
-
-- comment on scs?
-- comment previous state of the practice projects
-- what was done then and how things were done now (methodology)
--Red text denotes a linked sections of this document. Cyan text denotes a URL link. Blue text denotes a citation.
+\section{Introduction - DONE SECTION 1st DRAFT}
+
+This report analyzes the development of software packages that use Lattice Boltzmann Methods (LBM), a class of computational fluid dynamics methods, for fluid simulation.
+The analysis involves taking qualitative and qualitative measures along quality attributes.
+The goal is to analyze the current state of development of Scientific Computing Software (SCS) in this domain to provide insight into its best practices and offer guidance for future development.
+
+In this report red text denotes a linked sections of this document. Cyan text denotes a URL link. Blue text denotes a citation.
 
@@
 \begin{itemize}
-	\item Describe the State of Practice of SCS project origin, motivations, and goals
+	\item Describe the state of the practice of SCS project origin, motivations, and goals
 	\item Describe LBM software structure and goals
 	\item Report on the measure of a subset of LBM software packages along quality metrics
-	\item Evaluate the state of practice of LBM software development along quality metrics
+	\item Evaluate the state of the practice of LBM software development along quality metrics
 	\item Suggest future work for improving LBM software along quality metrics
-	\item Suggest future work for improving State of Practice of SCS assessments
+	\item Suggest future work for improving state of the practice of SCS assessments
 \end{itemize}
@@
 
-simple about SCS ....
-
--provide an example that demonstrates the value of se methods and technologies for SCS
--understand how software quality is impacted by the software development principles, processes and tools currently used within research software communities
-- list what the purpose of SOP evaluation of LBM is for
--describe LBM in the contect of SCS
--report on the SOP of LBM software development
--evaluate the development process and software qualities
--suggest improvements to LBM SCS in the context of software development and qualities
-- scs shows room for improvement in terms of quality, link papers, also why is this, link more papers
+The purpose of state of the practice assessments is to understand how software quality is impacted by software development choices, including principles, processes, and tools, within SCS communities.
+This knowledge can be used to guide future development of SCS, specifically along quality attributes, and reduce software quality failures.
+This work reports on the state of the practice of LBM software development and makes suggestions on improving the quality of software in this domain.
+
+This assessment of the state of the practice of LBM software development builds off of prior work on assessing the state of research software development.
+The previous state of the practice assessments include domains such as Geographic Information Systems \citep{SmithEtAl2018_arXivGIS}, Mesh Generators \citep{SmithEtAl2016}, Oceanographic Software \citep{smith2015state}, Seismology software \citep{SmithEtAl2018}, and statistical software for psychology \citep{SmithEtAl2018_StatSoft}.
+
+In the course of this assessment we updated the methodology that was used during previous assessments.
+Details of the new methodology are found \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Methodology.pdf}{here}.
+The previous set of research questions were critically assessed and modified.
+The updated questions are listed in Section \ref{researchquestions}.
+In this re-boot we collected more quantitative and qualitative data, focused on software quality attributes, by adding more measures, including collecting empirical data and interviewing developers.
+We have also added a domain analysis to better characterize the functionality provided by the software, and have leveraged the expertise of a domain expert.
+As in past assessments, the collected data was combined to rank the software using the Analytic Hierarchy Process (AHP).
+The domain expert was consulted to verify the ordering.
 
 \subsection{Scope of Report}
--measured 23 software projects (remove this)
-- other limitations about the study
-- addresses what was done
-- notes how things can be improved
+
+This report analyzes a filtered set of LBM software packages along quantitatively and qualitatively measures.
+Many of these measures are captured using surface measurements, which can be categorized as initial and easy to capture measurements, of the underlying quality and they may not represent the true quality of the software.
+These surface measurements are taken due to the desire to apply the same measurement, with reasonable effort, along all software packages despite technical and functional variabilities in the set of software packages.
+The measures are used to answer the research questions found in Section \ref{researchquestions}.
+The measured software packages are open source, and it is recommended to keep this in mind when considering the recommendations of this report.
+Practices surrounding close source software development may differ.
+
+The report addresses what was done during the development of the software packages to address the quality attributes listed in Section \ref{softwarequalities}.
+It then provides general guidance on how to improve these qualities when developing LBM software.
+It does not make suggestions on what should have been done or should be done for any one specific package.
+Best practices may at times differ among software packages.
 
@@
 	\item LBM software \textbf{Domain Analysis}.
-	\item \textbf{Methodology} of this State of the Practice Assessment, including the steps of the methodology, definitions of software qualities and how they are applied in this assessment, and an overview of candidate software selection, empirical measures, and the analytical hierarchy process.
+	\item \textbf{Methodology} of this state of the practice assessment, including the steps of the methodology, definitions of software qualities and how they are applied in this assessment, and an overview of candidate software selection, empirical measures, and the analytical hierarchy process.
 	\item Quantitative and qualitative \textbf{Results}. 
 	\item An analysis of the results and \textbf{Answers To Research Questions}.
-	\item \textbf{Conclusion} to the report including comments on future State of Practice assessments.
+	\item \textbf{Conclusion} to the report including comments on future state of the practice assessments.
 	\item \textbf{Appendix}: This section includes our Research Questions, the Measurement Template, Grading Template, Ethics Approval, and Developer Interview Questions.
@@
 this is about scs background
-
-must update images from pylbm to pyLBM
 
@@
 -background of LBM and what they are used for, link to domain analysis section for further detail
+
+" Scientific Computing (SC) is the subdomain of
+computer science that uses computers to simulate and analyze continuous mathematical
+models of physical systems. "
+" As SmithEtAl2016 [8] shows, SC developers value documentation, but
+they cannot see themselves meeting the required time constraints. Maintaining the doc-
+umentation when some requirements change during development process makes the time
+commitment even more difficult [2]. "
+"The current approach to documentation
+is largely a manual exercise, thus the quality is influenced by the author’s SC abilities. "
+
+"The problems with writing high-quality maintainable SC software documentation mo-
+tivated the creation of the Drasil Framework [7]. Drasil consists of a collection of Domain
+Specific Language (DSL) for capturing scientific documents, structures, and computing
+knowledge, and then transforming this knowledge into relevant software artifacts, such as
+requirement documentation, design documentation, and the code for test cases."
 
@@
 Section \ref{qualityrecommentations} lists recommendations to address software qualities in LBM software packages.
-The following are the software quality definitions used in this state of practice exercise, along with comments regarding their quantitative and qualitative measurement.
+The following are the software quality definitions used in this state of the practice exercise, along with comments regarding their quantitative and qualitative measurement.
 
@@
 system or component can be modified to correct faults, improve performance or other attributes, and satisfy new requirements \citep{IEEEStdGlossarySET1990} \citep{boehm2007software}.
-In this state of practice analysis maintainability is measured by the quality of documentation artifacts, and the presence of version control and issue tracking.
+In this state of the practice analysis maintainability is measured by the quality of documentation artifacts, and the presence of version control and issue tracking.
 These qualities can greatly decrease the effort needed to modify software.
@@
 Modifiability refers to the ease with which stable changes can be made to a system and the flexibility of the system to adopt such changes \citep{8016712}.
-This state of practice assessment did not quantitatively measure modifiability.
+This state of the practice assessment did not quantitatively measure modifiability.
 Developers were asked in interviews if they considered the ease of future changes when developing the software packages, specifically changes to the structure of the system, modules and code blocks.
@@
 Increased software modularization, smaller components with well defined interfaces, is important.
-For this state of practice assessment, a good measure of reusablity correlates with an increased number of code files, and the availability of API documentation.
+For this state of the practice assessment, a good measure of reusablity correlates with an increased number of code files, and the availability of API documentation.
 
@@
 Understandability is different for the user-interface, source code, documentation.
-In this state of practice analysis understandability focuses on the source code.
+In this state of the practice analysis understandability focuses on the source code.
 It is measured by the consistency of a formatting style, the extend of modularization, the explicit identification of coding standards, the presence of meaningful identifiers, and clarity of comments.
@@
 The quality refers to keeping track of information as it changes forms or relates between artifacts.
-This state of practice assessment did not quantitatively measure traceability.
+This state of the practice assessment did not quantitatively measure traceability.
 Developers were asked in interviews how documentation fits into their development process.
@@
 Visibility and transparency refer to the extent to which all of the steps of a software development process, and the current status of it, are conveyed clearly \citep{ghezzi1991fundamentals}.
-In this state of practice assessment a good measure of visibility and transparency correlates with a well defined development process, the presence of development process and environment documentation, and software package version release notes.
+In this state of the practice assessment a good measure of visibility and transparency correlates with a well defined development process, the presence of development process and environment documentation, and software package version release notes.
 
@@
 Reproducibility can also refer to the output of the software, irrespective of the artifacts, where the scientific results are compared between software implementations, or between software and manual implementations.
-This state of practice assessment did not quantitatively measure reproducibility.
+This state of the practice assessment did not quantitatively measure reproducibility.
 Developers were asked in interviews if they have any concern that their computational results won’t be reproducible in the future, and if they have taken any steps to ensure reproducibility.
@@
 In other words, artifacts are unambiguous if, and only if, they only have one interpretation \citep{IEEE1998}.
-This state of practice assessment did not quantitatively measure unambiguity.
+This state of the practice assessment did not quantitatively measure unambiguity.
 Developers were asked in interviews if they think that the current documentation can clearly convey all necessary knowledge to the users, and how they achieved this or what improvements are needed ot achieve it.
@@
 	\begin{center}
-		\includegraphics[width=1.0\textwidth]{reuseability_chart}
+		\includegraphics[width=1.0\textwidth]{reusability_chart}
 		\caption{AHP Reusability Score}
@@
 - key findings restated
-- comment on future work in State of Practice SCS
+- comment on future work in state of the practice SCS
 

### 2021-05-17 StateOfPractice/Peter-Notes/Report/Report.tex
some notes on additions to Background section

@@
 
-this is about scs background
-
-this is theoretical
+this section is about scs background and software quality background, separate definitions of those and relationship between, also define software family
+
 -define SCS in more detail than introduction
+-uses computers to simulate and analyze continuous mathematical
+models of physical system[add citation from prev paper]
+- look at other related papers and cite
+-look at other sources and cite
+
 - importance of qualities in SCS
-- ensuring the quality can take time
--define software family (defined more/further in domain analysis)
--background of LBM and what they are used for, link to domain analysis section for further detail
-
-" Scientific Computing (SC) is the subdomain of
-computer science that uses computers to simulate and analyze continuous mathematical
-models of physical systems. "
-" As SmithEtAl2016 [8] shows, SC developers value documentation, but
+- add citations
+- ensuring the quality can take time blurb
+-" As SmithEtAl2016 [8] shows, SC developers value documentation, but
 they cannot see themselves meeting the required time constraints. Maintaining the doc-
 umentation when some requirements change during development process makes the time
-commitment even more difficult [2]. "
+commitment even more difficult [2]." paraphrase and cite
 "The current approach to documentation
-is largely a manual exercise, thus the quality is influenced by the author’s SC abilities. "
+is largely a manual exercise, thus the quality is influenced by the author’s SC abilities. " paraphrase /cite
+- add blurb about Drasil, like below
+-define software family, add citation 
+- examples?
 

### 2021-05-19 StateOfPractice/Peter-Notes/Report/Report.tex
added notes on artifacts present

@@
 \usepackage{blindtext}
+\usepackage{multicol}
 \usepackage{indentfirst}
@@
 \subsection{Artifacts Present}
+
+split into common (16-23), less common (8-15), rare artifacts (0-7) lists
+
+in each split section comment on why almost always present are there and do the best ranked have the common/rare artifacts?, if so comment on how they could have contribted to higher rankings in specific quality AHP rankings
+
+Common (16-23):
+\begin{singlespace}
+\begin{multicols}{2}	
+\begin{itemize}
+	\item License 
+	\item Makefile / Build File 
+	\item README File 
+	\item Tutorial 
+	\item List of Related Articles / Publications 
+	\item Dependency Notes 
+	\item Requirements Specification / Theory Notes
+	\item Authors / Developers
+	\item Installation Guide/Instructions 
+	\item Bug Tracker
+\end{itemize}
+\end{multicols}
+\end{singlespace}
+
+Less Common (8-15):
+\begin{singlespace}
+	\begin{multicols}{2}	
+		\begin{itemize}
+			\item User Manual/Guide
+			\item Version Control
+			\item Change Log / Release Notes
+			\item Functional Specification/Notes
+			\item Design Documentation
+			\item Performance Information/Notes
+			\item Test Plan/Report/Script/Data/Cases
+		\end{itemize}
+	\end{multicols}
+\end{singlespace}
+
+Rare (1-7):
+\begin{singlespace}
+	\begin{multicols}{2}	
+		\begin{itemize}
+			\item Developer/Contributing Manual/Guide
+			\item Video Guide (including YouTube)
+			\item API Documentation
+			\item Verification and Validation Plan/Notes
+			\item FAQ / Forum
+		\end{itemize}
+	\end{multicols}
+\end{singlespace}
 

### 2021-05-20 StateOfPractice/Peter-Notes/Report/Report.tex
1st draft LBM report, artifacts present

@@
 
-\subsection{Artifacts Present}
-
-split into common (16-23), less common (8-15), rare artifacts (0-7) lists
-
-in each split section comment on why almost always present are there and do the best ranked have the common/rare artifacts?, if so comment on how they could have contribted to higher rankings in specific quality AHP rankings
-
-Common (16-23):
+\subsection{Artifacts Present - DONE 1st DRAFT}
+
+The software packages were examined for artifacts.
+Below we have grouped these into common, less common, and rare artifacts.
+Common artifacts were found in 16 to 23 of the software packages.
+Less common artifacts were found in 8 to 15 of the software packages.
+Rare artifacts were found in 1 to 27 of the software packages.
+
+\subsubsection{Common Artifacts}
+The following artifacts were commonly found in the 23 software packages that were tested.
+All of the top four AHP ranked packages, ESPREsSo, Ludwig, Palabos, and OpenLB have each of these artifacts, except for a clear requirements specification or theory notes.
+Only three of the packages were found to have that artifact.
+
 \begin{singlespace}
@@
 \begin{itemize}
+	\item Authors/Developers List
+	\item Bug Tracker
+	\item Dependency Notes 
+	\item Installation Guide/Instructions 
 	\item License
+	\item List of Related Articles/Publications 
 	\item Makefile / Build File 
 	\item README File 
+	\item Requirements Specification / Theory Notes
 	\item Tutorial
-	\item List of Related Articles / Publications 
-	\item Dependency Notes 
-	\item Requirements Specification / Theory Notes
-	\item Authors / Developers
-	\item Installation Guide/Instructions 
-	\item Bug Tracker
 \end{itemize}
@@
 
-Less Common (8-15):
+These common artifacts contribute to the quality of the software in various ways.
+A list of authors and developers helps potential users and contributors more easily contact project members to answer questions affecting many software qualities.
+A bug tracker helps with organizing improvements to the software, also improving many software qualities.
+Dependency notes help users install the software, as does an installation guide.
+Makefiles or other automated build files decrease human error in the installation process.
+Visible licenses promote usage of the software.
+Requirements specifications, linked theory notes, and related articles and publications help clarify the software and make it more understandable, decrease ambiguity, and help to verify its correctness.
+README files also help with understandability.
+Meanwhile tutorials help users become familiar with using the software.
+
+\subsubsection{Less Common Artifacts}
+The following artifacts were less commonly found in the 23 software packages that were tested. The top four AHP ranked packages have most of these artifacts.
+Only one of those four packages does not have a user manual or guide, but it does have a detailed and informative website.
+Another one of the top four packages does not have any visible design documentation.
+A third package from that list does not appear to use a version control system.
+It is probably that such a system is in use but is not visible outside of the development team since the package website notes a current version number.
+
 \begin{singlespace}
@@
 		\begin{itemize}
+			\item Change Log / Release Notes
+			\item Design Documentation
+			\item Functional Specification/Notes	
+			\item Performance Information/Notes
+			\item Test Plan/Report/Script/Data/Cases
 			\item User Manual/Guide
 			\item Version Control
-			\item Change Log / Release Notes
-			\item Functional Specification/Notes
-			\item Design Documentation
-			\item Performance Information/Notes
-			\item Test Plan/Report/Script/Data/Cases
 		\end{itemize}
@@
 
-Rare (1-7):
+These artifacts also contribute to software quality.
+A change log or release notes improve traceability of the software.
+Design documentation helps with maintaining, modifying, and reusing the software.
+Version control also helps to improve these qualities, as well as traceability.
+Functional specifications and notes clarify the software, contributing to usability and understandability.
+A user manual also benefits those qualities and with installability, and correctness and verificability.
+Performance notes suggest that performance was considered when developing the software.
+Test plans, scripts, and cases, help verify correctness.
+
+\subsubsection{Rare Artifacts}
+The following artifacts were rarely found in the 23 software packages that were tested.
+It is not common for the top four AHP ranked packages to have many of these artifacts.
+None of the top four packages have any explicit API documentation.
+Three of these packages have information on contributing to the project.
+Two of them have a FAQ section or forums.
+One has verification and validation notes and a video guide explaining how to use the software.
+
 \begin{singlespace}
@@
 		\begin{itemize}
+			\item API Documentation
 			\item Developer/Contributing Manual/Guide
+			\item FAQ / Forum
+			\item Verification and Validation Plan/Notes
 			\item Video Guide (including YouTube)
-			\item API Documentation
-			\item Verification and Validation Plan/Notes
-			\item FAQ / Forum
 		\end{itemize}
@@
 \end{singlespace}
+
+Although these artifacts were rarely found in our set of LBM software, they also contribute to software quality.
+API documentation helps with reusing the software.
+Developer and contributor manuals and guides help with maintainability, visibility and transparency.
+FAQs and forums increase usability of software.
+Verification and validation notes can be used to help check if a system meets specifications.
+Video guides can contribute to many software qualities, depending on the content of the video.
 

### 2021-05-20 StateOfPractice/Peter-Notes/Report/Report.tex
1st draft LBM report, tools used

@@
 
-\subsection{Tools Used}
+\subsection{Tools Used - DONE 1st Draft}
+The purpose of tools is to support the development, verification, maintenance, and evolution of software, software processes, and artifacts \citep{ghezzi1991fundamentals}.
+Many tools are used by current LBM software packages.
+The tools noted here are subdivided into development tools, dependencies, and project management tools.
+
+\subsubsection{Development Tools}
+
+These tools support the development of end products but do not become part of them, unlike dependencies that remain in the application once it is released \citep{ghezzi1991fundamentals}.
+The following type of development tools were explicitly noted in the artifacts or web pages of the 23 LBM packages that were assessed.
+It is likely that other tools, such as debuggers, were used but are not specified here.
+
+\begin{singlespace}
+	\begin{multicols}{2}	
+		\begin{itemize}
+			\item Continuous Integration Tools
+			\item Code Editors
+			\item Development Environment
+			\item Runtime Environments
+			\item Compilers
+			\item Unit Testing Tools
+			\item Correctness Verification Tools
+		\end{itemize}
+	\end{multicols}
+\end{singlespace}
+
+The use of the above tools can verify the correctness of software during its development.
+Only two of the software packages that were assessed mentioned using continuous integration tools like Travis CI.
+Code editors and compilers were explicitly noted to have been used by several packages, and were likely used by all of them.
+One of the packages explicitly noted the use of unit testing, but did not specify the tool that was used.
+Likewise the use of a tool for verifying the correctness of output was noted by one of the developers.
+It is likely that similar tools were used when developing other members of the software package list.
+
+\subsubsection{Dependencies}
+
+The following types of dependencies were explicitly noted in the artifacts or web pages of the 23 LBM packages that were assessed.
+It is possible that other types of dependencies are part of these software packages but are not clearly specified in their artifacts or web sites and because of that they are not listed here.
+
+\begin{singlespace}
+	\begin{multicols}{2}	
+		\begin{itemize}
+			\item Build Automation Tools
+			\item Technical Libraries
+			\item Domain Specific Libraries
+		\end{itemize}
+	\end{multicols}
+\end{singlespace}
+
+Most of the software packages use some sort of build automation tools, most commonly Make.
+They also all use various technical and domain specific libraries.
+Technical libraries include visualization, data analysis, and message passing libraries among others.
+Domain specific libraries are scientific computing libraries.
+Specific libraries are not listed here but they fall into the above categories.
+Libraries that are not explicitly stated in any artifact may fall outside of the categories.
+
+\subsubsection{Project Management Tools}
+
+Many of the software packages that were assessed were developed by teams of two or more people.
+Their work needed to be coordinated and managed.
+The following types of project management tools were explicitly noted in the artifacts, web pages, or interviews with the developers of the 23 LBM packages that were assessed.
+As with development tools and dependencies, it is possible that other types of project management tools were used to coordinate and manage the projects but are not specified and because of that they are not listed here.
+
+\begin{singlespace}
+	\begin{multicols}{2}	
+		\begin{itemize}
+			\item Collaboration Tools
+			\item Email
+			\item Change Tracking Tools
+			\item Version Control Tools
+			\item Document Generation Tools
+		\end{itemize}
+	\end{multicols}
+\end{singlespace}
+
+Collaboration tools are noted to have been used when developing the software projects.
+Most often email and video conferencing is used.
+Project management software was not explicitly mentioned, but it is possible that some of the project use such software.
+Many of the projects are located on GitHub, and it's developers use the platform to help manage their projects, especially bug related issues.
+Most of the projects appear to use change tracking and version control tools.
+They often use GitHub for this, and a few use Git or CVS.
+Document generation tools are mentioned in the artifacts of 12 of the projects.
+Tools such as Sphinx and Doxygen are explicitly used in this capacity.
+Drasil could be considered as an alternative to generate artifacts since it is geared specifically towards research software domains like LBM software.
 
@@
 \subsection{Pain Points}
+
+- bringing on new developers
+
+- communication of software functionality
+
+- time/funding for documentation
+
+-look at template and interviews
+
+- others
 

### 2021-05-23 StateOfPractice/Peter-Notes/Report/Report.tex
1st draft LBM report, background and principles, processes, methodologies; other minor notes

@@
 \newpage
-\section{Background}
-
-this section is about scs background and software quality background, separate definitions of those and relationship between, also define software family
-
--define SCS in more detail than introduction
--uses computers to simulate and analyze continuous mathematical
-models of physical system[add citation from prev paper]
-- look at other related papers and cite
--look at other sources and cite
-
-- importance of qualities in SCS
-- add citations
-- ensuring the quality can take time blurb
--" As SmithEtAl2016 [8] shows, SC developers value documentation, but
-they cannot see themselves meeting the required time constraints. Maintaining the doc-
-umentation when some requirements change during development process makes the time
-commitment even more difficult [2]." paraphrase and cite
-"The current approach to documentation
-is largely a manual exercise, thus the quality is influenced by the author’s SC abilities. " paraphrase /cite
-- add blurb about Drasil, like below
--define software family, add citation 
-- examples?
-
-"The problems with writing high-quality maintainable SC software documentation mo-
-tivated the creation of the Drasil Framework [7]. Drasil consists of a collection of Domain
-Specific Language (DSL) for capturing scientific documents, structures, and computing
-knowledge, and then transforming this knowledge into relevant software artifacts, such as
-requirement documentation, design documentation, and the code for test cases."
+\section{Background - DONE SECTION 1st Draft}
+
+Scientific computing software (SCS) is defined as the use of computer tools to analyze or simulate mathematical models of real world systems \citep{smith2006systematic}.
+Given the importance of such software, scientists and engineers desire methods and tools to sustainably develop high quality software, which has led to the formation of groups like the Software Sustainability Institute (\href{https://www.software.ac.uk/}{SEI}) and Better Scientific Software (\href{https://bssw.io/}{BSS}).
+The purpose of this state of the practice assessment is to understand how software quality in the LBM SCS family is impacted by software development choices, including principles, processes, and tools.
+A software family is defined
+by \cite{parnas1976design} as “a set of programs whose common properties are so extensive that it is advantageous to study the common properties of the programs before analyzing individual members”.
+The common properties of the LBM family of related programs are listed in the a domain analysis in Section \ref{domainanalysis}.
+The software qualities that are assessed in this report are listed in Section \ref{softwarequalities}.
+Ensuring these qualities can take significant time, but as \cite{SmithEtAl2016} shows, SCS developers often cannot see themselves meeting the required time constraints to better develop some of the artifacts that improve software quality, like detailed documentation.
+Inability to develop such documentation due to time constraints could be mitigated by the use of automatic document generation tools like Drasil.
+The Drasil Framework consists of a collection of Domain Specific Languages (DSL) for capturing scientific documents, structures, and computing
+knowledge, and then transforming this knowledge into relevant software artifacts without having to manually duplicate knowledge into multiple artifacts \citep{zhao2018}.
 
@@
 
-\section{Domain Analysis - DONE SECTION 1st DRAFT}
+\section{Domain Analysis - DONE SECTION 1st DRAFT}\label{domainanalysis}
 
@@
 \end{enumerate}
+
+These steps are further detailed in the Methodology for Assessing the State of the Practice for Domain X document found \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Methodology.pdf}{here}.
 
@@
 
-\subsection{Principles, Processes, and Methodologies}
+\subsection{Principles, Processes, and Methodologies - DONE 1st DRAFT}
+
+Most of the assessed software packages do not explicitly state in their artifacts the motivations or design principles that were considered when developing the software.
+One package, Sailfish, indicates that shortening development time was considered in early stages of design, with the developers opting for using Python and CUDA/OpenCL to achieve this without sacrificing any computational performance.
+The goals of that project are explicitly listed as performance, scalability, agility and extendability, maintenance, and ease of use.
+The project scored well in these listed categories during our assessment, for the categories that were measured.
+
+Processes, like methods, are ways of doing things, especially in an orderly way; while methodologies are defined as systems of methods \citep{ghezzi1991fundamentals}.
+It is not explicitly indicated in the artifacts of most of these packages that development involved following any specific model, like a waterfall of agile development model.
+One developer noted that while no formal model is used, their development model is something similar to a combination of agile and waterfall development models.
+The developers teams of the LBM packages are fairly small and easily organized without the need for such methodologies.
+
+Seven of the software packages contain artifacts outlining the general development process, like basic instructions on how to contribute.
+Most of the development teams accept outside contributors, but generally the teams are tight-knit, often working at the same institution.
+
+The developers that were interviewed all noted similar project management processes.
+In teams of only a couple of developers, the addition of new features or major changes are discussed with the entire team.
+Projects with more than a couple of developers have lead developer roles.
+These lead developers review potential additions to the software.
+One of the developers noted in their interview that a peer review process is used to review major changes and additions.
+
+The software packages frequently use GitHub for managing the project. Typically there are several development branches and changes from these branches are merged into the master branch.
+
+Documentation was also noted as playing a significant role in the development process, specifically with on-boarding new developers.
+An intention is to lower the entry barrier for these new contributors.
+The documentation provides information on how to get started, orients the user to artifacts and the source code, and explains how the system works, including the so-called simulation engine and interface.
+The use of document generation tools is mentioned in the artifacts of 12 software packages, and was noted during interviews with developers.
+The mentioned document generation tools are Sphinx and Doxygen.
+
+Two types of software changes were noted during interviews.
+One is feature additions, which arise from a scientific or functional need.
+The development of these changes was noted to involve more formal discussions within the development team and the execution involves lead developers.
+The other change type is code refactoring, which only sometimes involves formal discussions with the development team.
+New developers were noted to play an increased role in these changes compared to the former changes.
+Software bugs are typically addressed in a similar fashion as code refactoring, and issue trackers are commonly used to manage these changes.
+
+Interviews with the developers of software packages also revealed a more frequent use of both unit testing and continuous integration in the development process than was found by only examining the artifacts.
+The use of automatic installation processes is also common.
+Most often this involved a Make script.
 
@@
 
+"Developers were asked to comment on the obstacles in their development process. One
+developer noted that a lot of their source code had been written with a specific application in
+mind, and that there is too much coupling between components. Addressing this issue would
+help with code modifiability and reusability. Updating the development process would help
+resolve this issues and prevent such issues in the future. Improving the visibility of software
+changes and the peer review process would also help. Improving the software engineering
+education or experience of developers is also an idea that was brought up by several devel-
+opers. Specifically, developers should always write code that is decoupled and modular, and
+should keep in mind the visibility of their contributions by better updating documentation
+and ensuring that their contributions are transparent to the rest of the developers. This
+would help catch issues in the code contributions, and improve source code maintenance.
+According to one developer, some obstacles to the development processes of their package
+had been overcome by the introduction of continuous integration practices, and a peer review
+process for contributions. These practices decrease development and maintenance times.
+Another developer mentioned that their package had two sets of code, for executing the
+models on the CPU and GPU, and that maintenance was decreased by introducing macros,
+the use of which is now part of the development process.
+"
+
 - bringing on new developers
@@
 - others
+
+9. Currently, what are the most significant obstacles in your development process?
+
+12. In the past, is there any major obstacle to your development process that has been
+solved? How did you solve it?
+
+15. Was it hard to ensure the correctness of the software? If there were any obstacles, what
+methods have been considered or practiced to improve the situation? If practiced, did
+it work?
+16. When designing the software, did you consider the ease of future changes? For example,
+will it be hard to change the structure of the system, modules or code blocks? What
+measures have been taken to ensure the ease of future changes and maintains?
+
+17. Provide instances where users have misunderstood the software. What, if any, actions
+were taken to address understandability issues?
+18. What, if any, actions were taken to address usability issues?
+19. Do you think the current documentation can clearly convey all necessary knowledge to
+the users? If yes, how did you successfully achieve it? If no, what improvements are
+needed?
+20. Do you have any concern that your computational results wont be reproducible in the
+future? Have you taken any steps to ensure reproducibility?
 
@@
 \subsection{Ethics Approval}
+This project received ethics clearance from the McMaster Research Ethics Board on February 20, 2021.\newline
+
+Project Title: AIMSS - State of the Practice\newline
+
+MREB\#: 5219
+
 \newpage

### 2021-05-24 StateOfPractice/Ao-Report/17Recommendations.tex
updated outlines, chap 2, and chap 3

\chapter{Recommendations}
\label{ch_recommendations}

I think the recommendations can originate from both parts - measurements and interviews.

### 2021-05-24 StateOfPractice/Peter-Notes/Report/Report.tex
LBM Report: added draft pain points

@@
 
-\subsection{Pain Points}
-
-"Developers were asked to comment on the obstacles in their development process. One
-developer noted that a lot of their source code had been written with a specific application in
-mind, and that there is too much coupling between components. Addressing this issue would
-help with code modifiability and reusability. Updating the development process would help
-resolve this issues and prevent such issues in the future. Improving the visibility of software
-changes and the peer review process would also help. Improving the software engineering
-education or experience of developers is also an idea that was brought up by several devel-
-opers. Specifically, developers should always write code that is decoupled and modular, and
-should keep in mind the visibility of their contributions by better updating documentation
-and ensuring that their contributions are transparent to the rest of the developers. This
-would help catch issues in the code contributions, and improve source code maintenance.
-According to one developer, some obstacles to the development processes of their package
-had been overcome by the introduction of continuous integration practices, and a peer review
-process for contributions. These practices decrease development and maintenance times.
-Another developer mentioned that their package had two sets of code, for executing the
-models on the CPU and GPU, and that maintenance was decreased by introducing macros,
-the use of which is now part of the development process.
-"
-
-- bringing on new developers
-
-- communication of software functionality
-
-- time/funding for documentation
-
--look at template and interviews
-
-- others
-
-9. Currently, what are the most significant obstacles in your development process?
-
-12. In the past, is there any major obstacle to your development process that has been
-solved? How did you solve it?
-
-15. Was it hard to ensure the correctness of the software? If there were any obstacles, what
-methods have been considered or practiced to improve the situation? If practiced, did
-it work?
-16. When designing the software, did you consider the ease of future changes? For example,
-will it be hard to change the structure of the system, modules or code blocks? What
-measures have been taken to ensure the ease of future changes and maintains?
-
-17. Provide instances where users have misunderstood the software. What, if any, actions
-were taken to address understandability issues?
-18. What, if any, actions were taken to address usability issues?
-19. Do you think the current documentation can clearly convey all necessary knowledge to
-the users? If yes, how did you successfully achieve it? If no, what improvements are
-needed?
-20. Do you have any concern that your computational results wont be reproducible in the
-future? Have you taken any steps to ensure reproducibility?
+\subsection{Pain Points - DONE 1st Draft}
+
+Developers were asked to comment on obstacles in their development process, obstacles encountered by user, and potential future obstacles with the software packages.
+
+Several development process obstacles were noted.
+One developer noted that their small development team, common in LBM software packages, has a lack of time to implement new features.
+Team members are almost always part of the same institute or already know each other from other projects.
+External contributions are rare despite many of the projects accepting them.
+
+A lack of software development experience was noted by the developer of a third package where many of the team members on the project are domain experts.
+There is a steep learning curve to get these team members to contribute good quality source code.
+This has been somewhat addressed as the code is now generally written in a way that the developer thinks makes it as easy as possible for contributions.
+
+The developer also noted that there is a lack of incentive and funding in academia for developing good scientific software that people end up using.
+There are no journals that publish such source code.
+There is no place for the development of good quality software.
+Scientific software is often developed by the researchers that temporarily use it in their own research and keep the development and use in-house.
+This software often does not meet the standards that would be required by external users.
+Sometimes the software is commercialized but it is not open source.
+
+Documentation is important, but it's quality could be improved.
+There is often no time or funding for maintaining quality documentation for software that is rarely used externally.
+Furthermore, documentation generally does not teach the underlying metaphysics.
+It sometimes provides an introduction to these topics.
+Users would need to already be familiar with these topics or spend time referencing suitable textbooks.
+The documentation generally focuses on explaining how to use the software.
+It is not feasible for package documentation to address the underlying metaphysics topics in detail.
+Sometimes frequently asked questions on the underlying theory are answered in the documentation, or pointers to answers are noted.
+
+Technically, setting up parallelization was also noted as a pain point for one of the developers, and the introduction of continuous integration by another.
+Software development knowledge could mitigate some pain points.
+As already noted, many of the developers are domain experts and not professional software developers.
+An interviewed developer with more extensive software development experience noted that eliminating equivalent statements using macros had helped improve the quality of their source code, specifically helping with extending code to run on both the CPU and GPU.
+
+Difficulties with ensuring correctness were also noted by several developers.
+They indicated that tests are run on new additions,on  both the individual modules and on the entire system to verify continued system correctness.
+These tests compare the package output to known correct output using test cases.
+A developer commented that the amount of testing data that is needed for some cases is a problem, as free services do not offer capabilities to store and frequently process such large amounts of data and in-house testing solutions needed to be created.
+
+A few obstacles related to users were found.
+Several developers noted that users sometimes try to combine LBM methods with cases incorrectly, and that this is not theoretically sound.
+Some users think that the packages will work out of the box to solve their cases, while the packages require a good understanding of computational fluid dynamics and significant effort in setting up each case.
+These packages are not like more commercial software packages, they are generally set up to solve specific research problems.
+While they are modifiable to solve similar problems, these modifications are not trivial.
+So far this has been addressed by updating the documentation to better inform users of the underlying LBM theory and package requirements.
+Similar issues with LBM parameters were noted by another developer.
+Updating the user interface to better explain explain theoretical principles, as well as test user input, was the implemented solution.
+As noted above, sometimes frequently asked questions on the underlying theory and on how to use the software are answered in the documentation.
+
+One developer commented that parts of their source code had been refactored to Python partially to address usability.
+Python was perceived as a much more usable language that would be easy for future users and developers to learn and understand.
+According to this developer this was one of the biggest steps in terms of addressing usability.
+
+A few potential future obstacles were noted.
+One developer said that their source code had been written with a specific application in mind and that there was too much coupling between components.
+This would have an impact on future modifiability and reusability when trying to extend the software.
+The effected code would need to be refactored.
+
+As noted above, difficulties with ensuring future correctness could also arise.
+As new methods and functionality is added into the software, new test cases and test data will need to be developed and provided.
 

### 2021-05-24 StateOfPractice/Ao-Report/12Background.tex

@@
 
-\section{Categories and Status}
+\section{Software Categories}
+In this section, we discuss three common software categories that are mentioned in Section \ref{sec_software_selection}, and also SC software.
 
-Briefly discuss open source, freeware and commercial software.
+\subsection{Open Source Software}
+\label{sec_open_source_software}
+For Open Source software (OSS), its source code is openly accessible, and users have the right to study, change and distribute it under a license granted by the copyright holder.
+For many OSS projects, the development process is based on the collaboration of different contributors worldwide \cite{Corbly2014}.
+
+\subsection{Freeware}
+Freeware is software that can be used free of charge.
+Unlike with OSS, the authors of freeware typically do not allow users to access or modify the source code of the software \cite{LINFO2006}.
+The term \textit{freeware} should not be confused with \textit{free software}, which is similar to OSS but with a few differences.
+
+\subsection{Commercial Software}
+``Commercial software is software developed by a business as part of its business'' \cite{GNU2019}.
+Typically speaking, the users are required to pay to access all of the features of commercial software, excluding access to the source code.
+However, some commercial software is also free of charge \cite{GNU2019}.
+
+\subsection{Scientific Computing Software}
+Software development in Scientific Computing (SC) depends on the knowledge of three areas - the knowledge of a specific engineering or science domain, the ability to mathematically build models and applying algorithms, and the capability to implement theoretical models and algorithms with computational tools.
+SC software is built with mathematical and computational tools to serve the purpose of solving scientific problems in a domain \cite{Mehta2015}.
 
@@
 
-Refer to the Quality Definitions of Qualities created by us
+The definitions of software qualities are from Smith et al. \cite{SmithEtAl2020}. The order of the qualities follows the grading template in Appendix \ref{ap_grading}.
+
+\begin{itemize}
+\item \textbf{Installability} The effort required for the installation, uninstallation, or reinstallation of a software or product in a specified environment.
+\item \textbf{Correctness \& Verifiability} A program is correct if it behaves according to its stated.
+Verifiability is the extent to which a set of tests can be written and executed, to demonstrate that the delivered system meets the specification.
+\item \textbf{Reliability} The probability of failure-free operation of a computer program in a specified environment for a specified time, i.e.
+the average time interval between two failures also known as the mean time to failure (MTTF).
+\item \textbf{Robustness} Software possesses the characteristic of robustness if it behaves ``reasonably'' in two situations: i) when it encounters circumstances not anticipated in the requirements specification, and ii) when the assumptions in its requirements specification are violated.
+\item \textbf{Usability} The extent to which a product can be used by specified users to achieve specified goals with effectiveness, efficiency, and satisfaction in a specified context of use.
+\item \textbf{Maintainability} The effort with which a software system or component can be modified to i) correct faults; ii) improve performance or other attributes; iii) satisfy new requirements.
+\item \textbf{Reusability} The extent to which a software component can be used with or without adaptation in a problem solution other than the one for which it was originally developed.
+\item \textbf{Understandability} (To be completed)
+\item \textbf{Visibility \& Transparency} The extent to which all of the steps of a software development process and the current status of it are conveyed clearly.
+\end{itemize}
 
 \section{Analytic Hierarchy Process}
+The Analytic Hierarchy Process (AHP) was developed by Thomas L.
+Saaty, and it has been widely used to make and analyze multiple criteria decisions \cite{VaidyaEtAl2006}.
+The AHP organizes multiple criteria factors in a hierarchical structure and pairwise compares the alternatives to calculate relative ratios \cite{Saaty1990}.
+
+For a project with $ m $ criteria, we can use a  $m\times m$ matrix $A$ to record the relative importance between factors.
+By pairwise compare criterion $i$ and criterion $j$, the value of $A_{ij}$ is decided as follows, and the value of $A_{ji}$ is $1/A_{ij}$ \cite{Saaty1990},
+\begin{itemize}
+\item $A_{ij} = 1$ if criterion $i$ and criterion $j$ are equally important;
+\item $A_{ij} = 9$ if criterion $i$ is extremely more important than criterion $j$;
+\item $A_{ij}$ equals to an integer value between 1 and 9 according the the relative importance of criterion $i$ and criterion $j$.
+\end{itemize}
+
+The above process assumes that criterion $i$ is not less important than criterion $j$, otherwise, we need to reverse $i$ and $j$ and determine $A_{ji}$ first, then $A_{ij} = 1/A_{ji}$.
+
+The priority vecotr $w$ can be calculated by solving the following equation \cite{Saaty1990}, \begin{equation}
+Aw = \lambda_{max}w,
+\end{equation}
+where $\lambda_{max}$ is the maximal eigenvalue of $A$.
+
+In this project, $w$ is approximated with the approach classic \textit{mean of normalized values}  \cite{AlessioEtAl2006},
+
+\begin{equation}
+w_i = \frac{1}{m}\sum_{j=1}^{m}\frac{A_{ij}}{\sum_{k=1}^{m}A_{kj}}
+\end{equation}
+
+Suppose there are $n$ alternatives, for criterion $i = 1, 2, ...
+, m$, we can create an $n\times n$ matrix $B_i$ to record the relative preferences between these choices.
+The way of generating $B_i$ is similar to the one for $A$.
+However, unlike comparing the importance between criteria, we pairwise decide how much one alternative is more favored than the other.
+The same method is used to calculate the local priority vector for each $B_i$.
+
+In this project, the 9 software qualities mentioned above are the criteria ($m = 9$), while 29 software packages ($n = 29$) are compared.
+The software are evaluated with the grading template in Appendix \ref{ap_grading} and a subjective score is given for each quality.
+For a pair of qualities or software, $i$ and $j$, such that $i$ is not less significant than $j$, the pairwise comparison result of $i$ versus $j$ is converted from $min((score_i - score_j) + 1, 9)$.

### 2021-05-24 StateOfPractice/Ao-Report/13Methods.tex

\chapter{Methods}
\label{ch_methods}

\ad{Where to include the domain experts? I plan to include them in  Domain Selection and Software Product Selection, and also mention that they'll participate in interviews.}

In this project, we aim to study and research the SC software within scientific domains. However, we try to design the whole process of this project in a more general way, with the hope that it should not be limited to only one or several specific software categories.

The way how we choose a domain is listed in Section \ref{sec_domain_selection}. Within a specific domain, we use the process documented in Section \ref{sec_software_selection} to collect and filter the software packages for our study. Then, all the software is measured by using the grading template in Section \ref{sec_grading_template} and the empirical measurement method in Section \label{sec_empirical_measurements}. Section \ref{sec_measuring_qualities} presents more details about how we apply the above process.

\section{Domain Selection}
\label{sec_domain_selection}
Although the methods of this project may not have such limitations, we limit our selection within scientific domains to fulfill the objective of our research.

One major factor in properly choosing a candidate domain for the study is the ease to select software packages within it. As described in Section \ref{sec_software_selection}, the software product selection process is most likely to have a screening step, and the quantity of final-decision packages may not meet our initial expectation if we do not have enough software candidates to choose from. Section \ref{sec_software_selection} also explains why our process prioritize the OSS. Consequently, a scientific domain with a large number of active OSS is proffered. This also often indicates that the domain has an active community developing and using SC software, making it easier to conduct interviews mentioned in Section \ref{sec_interview_methods}.

Even if we can find an adequate number of software packages in a domain, we may still find the software products are developed to solve various problems within the domain and should be categorized into different sub-groups. So one question needs to be asked - do we prefer a group of software all providing similar functions and features, or do we aim to cross-compare several sub-sections within the same domain? With that answered, it should be easier to determine a favored domain.

Another aspect to consider is the team carrying out the research, more specifically, the domain experts - if there is any - in the team. In our team, the projects are often led by researchers in the software engineering field and supported by experts working in other scientific domains. Having domain experts in the team provides significant benefits in selecting software packages and designing interview questions.

In this project, Medical Imaging (MI) domain is selected. Numerous software products can be found in this domain, and a great number of domain experts use - and even develop - such software. We decide to focus on MI software with the viewing function, and more details about this filtering are in Section \ref{sec_software_selection}. Being able to include MI domain experts in our team also makes this domain more preferred.

\section{Software Product Selection}
\label{sec_software_selection}

The process of selecting software packages contains two steps: i) identify software candidates in the chosen domain, ii) filter the list according to needs \cite{SmithEtAl2021}.

\subsection{Identify Software Candidates}
The candidate software can be found from publications in the domain. Another source is to search various websites, such as \hyperlink{https://github.com/}{GitHub}, \hyperlink{https://swmath.org/}{swMATH} and the Google search results for software recommendation articles. Meanwhile, we should also include the suggested ones from domain experts \cite{SmithEtAl2021}.

As for this project, 48 MI software projects are identified as the candidates, and they are found from publications \cite{Bjorn2017} \cite{Bruhschwein2019} \cite{Haak2015}, online articles related to the domain \cite{Emms2019} \cite{Hasan2020} \cite{Mu2019}, forum discussions related to the domain \cite{Samala2014}, etc.

\subsection{Filter the Software List}
The goal is to build a software list with a length of about 30 \cite{SmithEtAl2021}.

The only mandatory requirement is that the software must be open source, as defined in Section \ref{sec_open_source_software}. This is due to the grading process defined int Section \ref{sec_grading_template}. In order to evaluate all aspects of some software qualities, the source code must be accessible.

The other factors to filter the list can be optional and should be considered according to the number of software candidates and the objectives of a research project.

One of the factors is the functions and purposes of the software. For example, we can choose a group of software with similar functions, so that the cross-comparison is between each individual of them. On the other hand, if our objective is comparing sub-categories in the domain, we should select from candidates in each of the categories.

The empirical measurement tools listed in Section \ref{sec_empirical_measurements} are limited to projects using \hyperlink{https://git-scm.com/}{Git} as the version control tool, so software with Git is preferred. Some manual steps in empirical measurement depend on a few metrics of GitHub, which makes projects hold on GitHub more favored \cite{SmithEtAl2021}.

Some of the OSS projects may experience a lack of recent maintenance. So packages that have not been updated for a long time can be eliminated, unless they are still popular and highly recommended by the users in the domain \cite{SmithEtAl2021}.

In this project, we focus on the MI software providing imaging viewing function. \ad{TBC}

\section{Grading Template}
\label{sec_grading_template}

\ad{The grading template for manual measurements}

\section{Empirical Measurements}
\label{sec_empirical_measurements}

\ad{Explain the empirical tools and how they affect the scores.}

\section{Measuring Qualities}
\label{sec_measuring_qualities}

\ad{e.g. virtual machines, time spent per software, where to look for docs, etc.}

\section{Interview Methods}
\label{sec_interview_methods}

\subsection{Interviewee Selection}

\ad{Generally speaking, ask all the teams which we can find contacts, and continue with the ones who are willing to participate.}

\subsection{Interview Question Selection}

\ad{The aspects we focused on.}

\subsection{Interview Methods}

\ad{The ethics approvals, the way to ask questions, the way to transcript answers, and the technologies used.}

### 2021-05-25 StateOfPractice/Peter-Notes/Report/Report.tex
LBM Report: added draft quality recommendations

@@
 
-\subsection{Artifacts Present - DONE 1st DRAFT}
+\subsection{Artifacts Present - DONE 1st DRAFT}\label{artifacts}
 
@@
 
-\subsection{Quality Recommendations}\label{qualityrecommentations}
-
-create section for each quality?
-
-consider what worked for some of the packages
+\subsection{Quality Recommendations - DONE 1st Draft}\label{qualityrecommentations}
+
+The following points should be considered to improve LBM software quality.
+
+\subsubsection{Installability}
+
+\begin{itemize}
+	\item Include OS compatibility, including versions.
+	\item Include complete installation instructions.
+	\item Include all dependencies in the installation instructions, required versions, and how to install them.
+	\item If possible, automate the installation of dependencies.
+	\item Installation instructions should be written as if the user does not have any dependencies installed and is installing on a clean OS.
+	\item Limit installation instruction location to one location.
+	\item Automate the installation process as much as possible.
+	\item Include descriptive error messages for errors encountered during installation.
+	\item Include a way to validate the installation.
+	\item Include instructions for uninstalling the software.
+\end{itemize}
+
+\subsubsection{Surface Correctness and Verifiability}
+
+\begin{itemize}
+	\item Use a requirements specification document.
+	\item Make public the requirements specification document, or explicitly reference the domain theory that the software is designed from.
+	\item Ensure the above information is easy to find. Consider adding it to the user manual. 
+	\item Development teams should include both domain experts and experiences software developers.
+	\item Use and make public detailed documentation. Consider using automatic document generation tools.
+	\item Include detailed tutorials with expected output.
+	\item Use unit testing during development. Make public the facilities for unit testing.
+	\item Modularize the source code.
+	\item Use continuous integration tools and processes during development.
+\end{itemize}
+
+\subsubsection{Surface Reliability}
+
+\begin{itemize}
+	\item Include descriptive error messages where appropriate.
+	\item In case automatic installation of dependencies fails, the system should indicate to the user what dependencies are needed and the installation procedure should be able to skip installing the dependencies after they have been manually installed.
+	\item The packages should include detailed tutorials, including dependencies, expected output, and supplementary documentation.
+\end{itemize}
+
+\subsubsection{Surface Robustness}
+
+\begin{itemize}
+	\item Include descriptive error messages for cases of unexpected input  including incorrect data types, empty input, and missing files or links.
+\end{itemize}
+
+\subsubsection{Surface Performance}
+
+\begin{itemize}
+	\item Implement using parallelization tools. Consider GPU processing, CUDA, and MPI. 
+	\item The user should be able to process the models on either the CPU or GPU.
+\end{itemize}
+
+\subsubsection{Surface Usability}
+
+\begin{itemize}
+	\item Include user hardware and software requirements.
+	\item Include a user tutorial with expected output.
+	\item Include a detailed user manual. Identify elements of user interfaces.
+	\item Explicitly state the boundaries of the software, and intended applications.
+	\item Identify appropriate fluid dynamics problems that can be solved by the software.
+	\item Provide background domain information or a reference to it.
+	\item Explicitly state the requirements and commitments required of the user to properly model their problem.
+	\item Document expected user characteristics.
+	\item Keep documentation in one location.
+	\item Maintain a user support model (Git, email, forum, FAQ)
+	\item If possible, consider using user-friendly scripting languages like Python. Especially in parts of the software that are likely to be modified or reviewed by users.
+\end{itemize}
+
+\subsubsection{Maintainability}
+
+\begin{itemize}
+	\item Include high quality artifacts. Ideally most of the common and less common artifacts listed in Section \ref{artifacts}.
+	\item Include version numbers and release notes for all major releases. 
+	\item Have a defined process for accepting contributions.
+	\item Include documentation for making contributions to the project. 
+	\item Use an issue tracker (Git, email, SourceForge, other) to manage bugs and changes. Issues should be closed regularly. 
+	\item Use a version control system (GitHib, CVS).
+	\item Source code needs to be well commented.
+	\item Ensure source code is modular.
+	\item Eliminate code duplication.
+	\item If possible, develop using languages that are easy to read and quick to learn.
+\end{itemize}
+
+\subsubsection{Modifiability}
+
+\begin{itemize}
+	\item Ensure a high degree of code modularity and abstraction.
+	\item If possible, develop using languages that are easy to read and quick to learn.
+	\item Consider future modifiabilility in the design stage prior to any development.
+	\item Consider flexibility of data structures and data storage in the design stage.
+	\item Make public validation benchmarks.
+\end{itemize}
+
+\subsubsection{Reusability}
+
+\begin{itemize}
+	\item Modularize the source code.
+	\item Document module interfaces.
+	\item Provide API documentation, if applicable.
+\end{itemize}
+
+\subsubsection{Surface Understandability}
+
+\begin{itemize}
+	\item Maintain consistent formatting style across source code files.
+	\item Identify a coding standard, if one is used during development.
+	\item Use consistent, distinctive, and meaningful code identifiers. Identify algorithms that are used.
+	\item Hard code domain constants.
+	\item Add meaningful comments. They should indicate what is being done in each section of source code.
+	\item Modularize the source code.
+	\item Include user guides and usage requirements.
+	\item In the documentation identify the limits of the software. 
+	\item Identify the cases that the software can be applied to.
+	\item Note the requirements of the user in order to set up their case. Identify what parameters are needed. 
+	\item Consider adding a FAQ section to the documentation. 
+\end{itemize}
+
+\subsubsection{Traceability}
+
+\begin{itemize}
+	\item Update all relevant documentation when a change to the software is made.
+	\item Use automatic document generation tools to limit the time spent on updating documentation.
+	\item Include documentation aimed at developers, like a developers guide.
+	\item Provide documentation that orients users and developers to artifacts, the source code, and system architecture.
+	\item Highlight new system features between artifact versions. 
+\end{itemize}
+
+\subsubsection{Visibility and Transparency}
+
+\begin{itemize}
+	\item Identify the development model that was used, if one was used.
+	\item Identify the development environment.
+	\item Identify the development process and how to contribute. 
+	\item Include notes with all releases.
+	\item Write modular code.
+	\item Communicate all contributions and changes to the entire development team.
+	\item Use continuous integration processes and tools.
+	\item Consider peer review processes for contributions. 
+	\item Decrease code duplication.
+	\item Maintain updated developer documentation.
+	\item Make use of project management software.
+\end{itemize}
+
+\subsubsection{Reproducibility}
+
+\begin{itemize}
+	\item Compare results of methods against manually calculated or known correct results.
+	\item Automate the above comparison. Run the comparison after code changes.
+	\item Consider peer reviews for code contributions.
+	\item Maintain detailed and up to date design documentation.
+\end{itemize}
+
+\subsubsection{Unambiguity}
+
+\begin{itemize}
+	\item Documentation must include all technical requirements and dependencies for using the software. 
+	\item Documentation should detail how to properly use the software.
+	\item Documentation should outline correct and incorrect use cases of the software.
+	\item Documentation should either explain underlying CFD theories or reference appropriate resources.
+	\item Consider asking users for feedback on the documentation.
+\end{itemize}
 

### 2021-05-30 StateOfPractice/Ao-Report/12Background.tex
updated ao's report ch 3

@@
 \subsection{Freeware}
+\label{sec_freeware}
 Freeware is software that can be used free of charge.
@@
 \section{Software Quality Definitions}
+\label{sec_software_quality}
 
-The definitions of software qualities are from Smith et al. \cite{SmithEtAl2020}. The order of the qualities follows the grading template in Appendix \ref{ap_grading}.
+The definitions of software qualities are from Smith et al. \cite{SmithEtAl2020}. The order of the qualities follows the grading template in Appendix \ref{ap_grading_template}.
 
@@
 In this project, the 9 software qualities mentioned above are the criteria ($m = 9$), while 29 software packages ($n = 29$) are compared.
-The software are evaluated with the grading template in Appendix \ref{ap_grading} and a subjective score is given for each quality.
+The software are evaluated with the grading template in Appendix \ref{ap_grading_template} and a subjective score is given for each quality.
 For a pair of qualities or software, $i$ and $j$, such that $i$ is not less significant than $j$, the pairwise comparison result of $i$ versus $j$ is converted from $min((score_i - score_j) + 1, 9)$.

### 2021-05-30 StateOfPractice/Ao-Report/13Methods.tex
updated ao's report ch 3

@@
 
-In this project, we aim to study and research the SC software within scientific domains.
-However, we try to design the whole process of this project in a more general way, with the hope that it should not be limited to only one or several specific software categories.
+In this project, we aimed to study and research the SC software within scientific domains.
+However, we tried to design the whole process of this project in a more general way, with the hope that it should not be limited to only one or several specific software categories.
 
-The way how we choose a domain is listed in Section \ref{sec_domain_selection}.
-Within a specific domain, we use the process documented in Section \ref{sec_software_selection} to collect and filter the software packages for our study.
-Then, all the software is measured by using the grading template in Section \ref{sec_grading_template} and the empirical measurement method in Section \label{sec_empirical_measurements}.
-Section \ref{sec_measuring_qualities} presents more details about how we apply the above process.
+The way how we chose a domain is listed in Section \ref{sec_domain_selection}.
+Within a specific domain, we used the process documented in Section \ref{sec_software_selection} to collect and filter the software packages for our study.
+Then, all the software products were measured by using the grading template in Section \ref{sec_grading_template} and the empirical measurement method in Section \label{sec_empirical_measurements}.
+Section \ref{sec_measuring_qualities} presents more details about how we applied the above process.
 
@@
 \label{sec_domain_selection}
-Although the methods of this project may not have such limitations, we limit our selection within scientific domains to fulfill the objective of our research.
+Although the methods designed by us may not have such limitations, we limited our selection of a domain within scientific domains to fulfill the objective of our research.
 
@@
 
-In this project, Medical Imaging (MI) domain is selected.
-Numerous software products can be found in this domain, and a great number of domain experts use - and even develop - such software.
-We decide to focus on MI software with the viewing function, and more details about this filtering are in Section \ref{sec_software_selection}.
-Being able to include MI domain experts in our team also makes this domain more preferred.
+In this project, Medical Imaging (MI) domain was selected.
+Numerous software products can be found in this domain and a great number of professionals working in this domain use and/or develop such software.
+We decided to focus on MI software with the viewing function, and more details about this filtering are in Section \ref{sec_software_selection}.
+Being able to include MI domain experts in our team also made this domain more preferred.
 
@@
 \subsection{Identify Software Candidates}
-The candidate software can be found from publications in the domain.
+\label{sec_identify_software_candidates}
+The candidate software can be found in publications in the domain.
 Another source is to search various websites, such as \hyperlink{https://github.com/}{GitHub}, \hyperlink{https://swmath.org/}{swMATH} and the Google search results for software recommendation articles.
@@
 
-As for this project, 48 MI software projects are identified as the candidates, and they are found from publications \cite{Bjorn2017} \cite{Bruhschwein2019} \cite{Haak2015}, online articles related to the domain \cite{Emms2019} \cite{Hasan2020} \cite{Mu2019}, forum discussions related to the domain \cite{Samala2014}, etc.
+As for this project, 48 MI software projects were identified as the candidates, and they were found from publications \cite{Bjorn2017} \cite{Bruhschwein2019} \cite{Haak2015}, online articles related to the domain \cite{Emms2019} \cite{Hasan2020} \cite{Mu2019}, forum discussions related to the domain \cite{Samala2014}, etc.
 
@@
 
-The only mandatory requirement is that the software must be open source, as defined in Section \ref{sec_open_source_software}.
+The only ``mandatory'' requirement is that the software must be OSS, as defined in Section \ref{sec_open_source_software}.
 This is due to the grading process defined int Section \ref{sec_grading_template}.
-In order to evaluate all aspects of some software qualities, the source code must be accessible.
+To evaluate all aspects of some software qualities, the source code should be accessible.
 
@@
 The empirical measurement tools listed in Section \ref{sec_empirical_measurements} are limited to projects using \hyperlink{https://git-scm.com/}{Git} as the version control tool, so software with Git is preferred.
-Some manual steps in empirical measurement depend on a few metrics of GitHub, which makes projects hold on GitHub more favored \cite{SmithEtAl2021}.
+Some manual steps in empirical measurement depend on a few metrics of GitHub, which makes projects held on GitHub more favored \cite{SmithEtAl2021}.
 
@@
 
-In this project, we focus on the MI software providing imaging viewing function. \ad{TBC}
+No matter what standards are set to filter the packages, with domain experts in the team, their opinion should be valued.
+For example, if a software package is not OSS and has not been updated for a long while, the domain experts may still identify it as a popular and widely used product.
+In this case, perhaps it is valuable to keep this software on the list.
+
+In this project, among the 48 software packages identified in Section \ref{sec_identify_software_candidates}, there were several ones that we could not find their source code, such as MicroDicom, Aliza, and jivex, etc.
+So these packages are likely to be freeware defined in Section \ref{sec_freeware} and were removed from the list since they might not be OSS.
+
+We focused on the MI software that could work as a MI viewer.
+Some of the software on the list were tool kits or libraries for other software to use as dependencies but not for end-users to view medical images, such as VTK, ITK, and dcm4che, etc.
+These were also eliminated from the list.
+
+After the above two steps, there were 29 software products left on the list.
+We still preferred projects using git and GitHub and being updated recently, but no other filtering was needed since the number of packages was already below 30.
+However, 27 out of the 29 software packages on the final list used git, and 24 of which were held on GitHub.
+Furthermore, 27 packages had the latest updates after the year 2018, and 23 after 2020.
 
@@
 \label{sec_grading_template}
+The full grading template can be found in Appendix \ref{ap_grading_template}. The template contains 101 questions that we used for grading software products.
 
-\ad{The grading template for manual measurements}
+In the first section of the template, some general information about the software is collected, such as the name, purpose, platform, programming language, publications about the software, the first release and the most recent change date, website, and source code repository of the product, etc.
+Information in this section helps us to understand the projects better and may be useful for further analysis, but it does not directly affect the grading scores for the packages.
+
+The next 9 sections in the template are designed for the 9 software qualities mentioned in Section \ref{sec_software_quality}.
+For each quality, several questions are asked and the typical choices of answers are ``yes'', ``no'', ``n/a'', ``unclear'', a number, a string, a date, a set of strings, etc.
+The last question of each quality is asking for an overall score between 1 and 10, which is based on all the previous questions of this section.
+This score is also the grading score for this quality.
+For some qualities, this grading score is also affected by the empirical measurement sections on the template.
+
+All the last 3 sections on the template are related to the empirical measurements.
+Two command-line software tools git\_stats and scc are used to extract information about the source code from the project repositories.
+For projects held on GitHub, additional metrics are collected manually, such as the stars of the GitHub repository, and the numbers of open and closed pull requests.
+Details of how these empirical measurements affect software quality grading scores are presented in Section \ref{sec_empirical_measurements}.
 
@@
 
-\ad{Explain the empirical tools and how they affect the scores.}
+Two command-line tools were used for the empirical measurements in this project.
+One is GitStats that generates statistics for git repositories and display outputs in the format of web pages \cite{Gieniusz2019}; the other one is Sloc Cloc and Code (abbreviated as scc by the author) \cite{Boyter2021}, aiming to count the lines of code, comments, etc.
+
+Both tools can measure the number of text-based files in a git repository, as well as lines of text in these files.
+Based on our experience, most text-based files in a software project repository contain programming source code and are used to compile and build software products, and a minority of these files are instructions and other documents.
+So the lines of text in text-based files are roughly regarded as lines of programming code.
+The two tools usually generate similar but not indentical results as for the above measurements.
+From our understanding, this minor difference is because of the different techniques they use to detect if a file is a text-based or binary file.
+
+Additionally, we also manually collected some information for projects held on GitHub, such as the numbers of stars, forks, people watching this repository, open pull request, and closed pull request.
+
+These empirical measurements helped us from two aspects.
+Firstly, with more statistical details, we could get an overview of a project faster and more accurately.
+For example, the number of commits over the last 12 months shows how active this project has been during this period, and the number of stars and forks may be related to the popularity of it.
+Secondly, the results were factors to consider when we determined the grading scores for some software qualities.
+For example, if the percentage of comment lines is low, we might want to double-check the understandability of the code, and if the ratio of open versus closed pull requests is high, perhaps we should pay more attention to the maintainability of the project.
 
@@
 \label{sec_measuring_qualities}
+27 out of the 29 software packages can be installed on two or three different operating systems such as Windows, macOS, and Linux, and 5 of them are browser-based, making them platform-independent.
+However, in the interest of time, we only performed the measurements for each project by installing it on one of the platforms, most likely Windows.
 
-\ad{e.g. virtual machines, time spent per software, where to look for docs, etc.}
+In order to test the software on a ``clean'' system, we created a new virtual machine (VM) for each software and only installed the necessary dependencies before measuring.
+All the 29 VM were created on the same computer.
+We only started the measuring of the next software after finishing a current one, and after grading each software, the VM was destroyed.
+
+Generally speaking, about two hours were spent on grading one software, unless there were technical issues and more time was need to resolve them.
+For most of the situation, all the measurements for one software were finished on the same day.
+
+The results were filled into the grading template alongside measuring a software in a VM. The sequence of measuring also followed the grading template.
 
@@
 
-\ad{Generally speaking, ask all the teams which we can find contacts, and continue with the ones who are willing to participate.}
+For a software list with a length of roughly 30, we aim to interview about 10 development teams of these projects.
+Interviewing multiple individuals from each team should give us more comprehensive information about a project, but due to the difficulties of finding willing participants, a single engineer well knowing the project can also be sufficient.
+
+Ideally, we select projects after the grading measurements are done and prefer the projects with higher overall scores.
+However, if not enough participants are found, we should also contact all teams on the list.
+
+For MI domain, we started with the teams with higher score on our list, and eventually contacted all of them. There are developers/architects from 8 teams having participated our interviews so far.
+
+The contacts of the the teams were found on the websites related to the software.
+For example, the official web pages, repository websites, publication websites, and bio pages of the teams' institutions.
+For each candidate, we sent at most two emails asking for their support and participating before receiving any replies.
 
@@
 
-\ad{The aspects we focused on.}
+We have a list of 20 questions to guide our interviews with the development teams, which can be found in Appendix \ref{ap_interview}.
+
+Some of the questions are about the background of the software, the development teams, the interviewees, and the way they organize the projects.
+We also ask about their understandings to the users.
+Another part of the interview questions focuses on the major difficulties the team experience currently or in the past, and the solutions that they have found or will try in the future.
+We also discuss with interviewees the importance of documents to their projects and the current situations of these documents.
+One more proportion of the questions are about several specific software qualities, such as maintainability, understandability, usability, and reproducibility.
+
+The interviews are supposed to be semi-structured based on the question list and follow-up questions can be asked.
+Based on our experience, the interviewees usually bring up some interesting topics not expected by us and we found it valuable to continue on these topics and ask for a few more details.
 
 \subsection{Interview Methods}
+Before contacting any interviewee candidate, the study had been reviewed by the McMaster University Research Ethics Board and received ethics clearance.
 
-\ad{The ethics approvals, the way to ask questions, the way to transcript answers, and the technologies used.}
+The members of the development teams are based around the world, so we organized these interviews as virtual meetings online.
+\hyperlink{https://zoom.us/}{Zoom} was used for the meetings.
+After receiving consents from the interviewees, we also recorded our discussions to better transcribe them.

### 2021-05-30 StateOfPractice/Ao-Report/13Methods.tex
ao's report - minor updates

@@
 
-\subsection{Interview Methods}
+\subsection{Interview Process}
 Before contacting any interviewee candidate, the study had been reviewed by the McMaster University Research Ethics Board and received ethics clearance.

### 2021-06-01 StateOfPractice/Ao-Report/12Background.tex
Ao's report - updated ch2

@@
 
+This chapter introduces several different categories of software, based on which we designed the processes to select the software domain and software candidates in Chapter \ref{ch_methods}.
+It also covers the software quality definitions used in the grading template in Appendix \ref{ap_grading_template} and an overview of Analytic Hierarchy Process (AHP), a tool we used to compare and grading software products.
+
 \section{Software Categories}
+To narrow down the scopes when selecting software domains and software packages, we usually target specific software categories.
 In this section, we discuss three common software categories that are mentioned in Section \ref{sec_software_selection}, and also SC software.
+Section \ref{sec_domain_selection} and Section \ref{sec_software_selection} present more details about why we prefer some of these categories.
 
@@
 For many OSS projects, the development process is based on the collaboration of different contributors worldwide \cite{Corbly2014}.
+Accessible source code usually expose more ``secrets'' of a software project, such as the underlying logic of software functions, the styles that how the developers achieve their works, and the flaws and potential risks in the final product.
+Thus, it brings much more convenience to the researches analyzing the qualities of the project.
 
@@
 The term \textit{freeware} should not be confused with \textit{free software}, which is similar to OSS but with a few differences.
+To the end-users, the differences between freeware and OSS often do not bother them.
+The fact that these products are free of charge is likely to make them popular to a large base of users.
+However, software developers, end-users who wish to modify the source code, and researchers looking for inner characteristics of it may find the inaccessible source code to be a problem.
 
@@
 However, some commercial software is also free of charge \cite{GNU2019}.
+Based on our experience, most commercial software products are not OSS.
+
+For some specific software, the backgrounds of commercial software developers often differ from the ones of non-commercial OSS.
+In such a case, the former is usually developed by software engineers, and the latter is likely to have developers who work in the domain of the software and are also end-users of the products.
+One example is software in Scientific Computing (SC), since the developers need to utilize their domain-specific during the development process \cite{WilsonEtAl2014}.
 
 \subsection{Scientific Computing Software}
-Software development in Scientific Computing (SC) depends on the knowledge of three areas - the knowledge of a specific engineering or science domain, the ability to mathematically build models and applying algorithms, and the capability to implement theoretical models and algorithms with computational tools.
+Software development in SC depends on the knowledge of three areas - the knowledge of a specific engineering or science domain, the ability to mathematically build models and applying algorithms, and the capability to implement theoretical models and algorithms with computational tools.
 SC software is built with mathematical and computational tools to serve the purpose of solving scientific problems in a domain \cite{Mehta2015}.
+The majority of scientists developing their own software are self-taught programmers \cite{WilsonEtAl2014}, so there may be a bigger Room for software quality improvement in SC domains.
 
@@
 \begin{itemize}
+\label{def_installability}
 \item \textbf{Installability} The effort required for the installation, uninstallation, or reinstallation of a software or product in a specified environment.
+\label{def_correctness_verifiability}
 \item \textbf{Correctness \& Verifiability} A program is correct if it behaves according to its stated.
 Verifiability is the extent to which a set of tests can be written and executed, to demonstrate that the delivered system meets the specification.
+\label{def_reliability}
 \item \textbf{Reliability} The probability of failure-free operation of a computer program in a specified environment for a specified time, i.e.
 the average time interval between two failures also known as the mean time to failure (MTTF).
+\label{def_robustness}
 \item \textbf{Robustness} Software possesses the characteristic of robustness if it behaves ``reasonably'' in two situations: i) when it encounters circumstances not anticipated in the requirements specification, and ii) when the assumptions in its requirements specification are violated.
+\label{def_usability}
 \item \textbf{Usability} The extent to which a product can be used by specified users to achieve specified goals with effectiveness, efficiency, and satisfaction in a specified context of use.
+\label{def_maintainability}
 \item \textbf{Maintainability} The effort with which a software system or component can be modified to i) correct faults; ii) improve performance or other attributes; iii) satisfy new requirements.
+\label{def_reusability}
 \item \textbf{Reusability} The extent to which a software component can be used with or without adaptation in a problem solution other than the one for which it was originally developed.
+\label{def_understandability}
 \item \textbf{Understandability} (To be completed)
+\label{def_visibility_transparency}
 \item \textbf{Visibility \& Transparency} The extent to which all of the steps of a software development process and the current status of it are conveyed clearly.
@@
 \section{Analytic Hierarchy Process}
-The Analytic Hierarchy Process (AHP) was developed by Thomas L.
+To generate grading scores for a group of software packages, we use the AHP to pairwise compare them.
+This tool was developed by Thomas L.
 Saaty, and it has been widely used to make and analyze multiple criteria decisions \cite{VaidyaEtAl2006}.

### 2021-06-02 StateOfPractice/Peter-Notes/Report/Report.tex
edits to introduction section of LBMM Report

@@
 
+\noindent\textbf{SCS}: Scientific Computing Software\\
+
 \noindent\textbf{Variability}: A requirement or goal that varies between family members.
@@
 
-In this report red text denotes a linked sections of this document. Cyan text denotes a URL link. Blue text denotes a citation.
+In this report red text denotes internally linked sections of this document. Cyan text denotes an external URL link. Green text denotes a link to the bibliography.
 
@@
 	\item Evaluate the state of the practice of LBM software development along quality metrics
-	\item Suggest future work for improving LBM software along quality metrics
-	\item Suggest future work for improving state of the practice of SCS assessments
+	\item Make suggestions for improving LBM software along quality metrics
+	\item Make suggestions for improving state of the practice of SCS assessments
 \end{itemize}
@@
 The updated questions are listed in Section \ref{researchquestions}.
-In this re-boot we collected more quantitative and qualitative data, focused on software quality attributes, by adding more measures, including collecting empirical data and interviewing developers.
+In this re-boot we collected more quantitative and qualitative data, focused on software quality attributes, and added more measures, including collecting empirical data and interviewing developers.
 We have also added a domain analysis to better characterize the functionality provided by the software, and have leveraged the expertise of a domain expert.
@@
 
-This report analyzes a filtered set of LBM software packages along quantitatively and qualitatively measures.
+This report analyzes a filtered set of LBM software packages along quantitative and qualitative measures.
 Many of these measures are captured using surface measurements, which can be categorized as initial and easy to capture measurements, of the underlying quality and they may not represent the true quality of the software.
-These surface measurements are taken due to the desire to apply the same measurement, with reasonable effort, along all software packages despite technical and functional variabilities in the set of software packages.
+Surface measurements are taken as they allow us to apply the same measurement, with reasonable effort, along all software packages despite technical and functional variabilities among the set of software packages.
 The measures are used to answer the research questions found in Section \ref{researchquestions}.
 The measured software packages are open source, and it is recommended to keep this in mind when considering the recommendations of this report.
+Some recommendations may not apply to closed-source software.
 Practices surrounding close source software development may differ.
@@
 It does not make suggestions on what should have been done or should be done for any one specific package.
-Best practices may at times differ among software packages.
+Best practices may differ among software packages due to their inherent organizational and technical differences.
 
@@
 The first part of this document includes the Table of Contents, List of Figures, and List of Tables.
-This is followed by Reference Material including a subsections outlining Software Engineering Related Definitions and Acronyms and Lattice Boltzmann Related Definitions and Acronyms.
+This is followed by Reference Material including subsections outlining Software Engineering Related Definitions and Acronyms and Lattice Boltzmann Related Definitions and Acronyms.
 
@@
 	\item LBM software \textbf{Domain Analysis}.
-	\item \textbf{Methodology} of this state of the practice assessment, including the steps of the methodology, definitions of software qualities and how they are applied in this assessment, and an overview of candidate software selection, empirical measures, and the analytical hierarchy process.
+	\item \textbf{Methodology} of this state of the practice assessment, including the steps of the methodology, software quality definitions and how these qualities are assessed in this report, an overview of candidate software selection, empirical measures gathered, and the analytical hierarchy process.
 	\item Quantitative and qualitative \textbf{Results}. 
 	\item An analysis of the results and \textbf{Answers To Research Questions}.
-	\item \textbf{Conclusion} to the report including comments on future state of the practice assessments.
+	\item \textbf{Conclusion} to the report including comments on suggestions for future state of the practice assessments.
 	\item \textbf{Appendix}: This section includes our Research Questions, the Measurement Template, Grading Template, Ethics Approval, and Developer Interview Questions.

### 2021-06-02 StateOfPractice/Peter-Notes/Report/Report.tex
edits to background section of LBM Report

@@
 
-Scientific computing software (SCS) is defined as the use of computer tools to analyze or simulate mathematical models of real world systems \citep{smith2006systematic}.
-Given the importance of such software, scientists and engineers desire methods and tools to sustainably develop high quality software, which has led to the formation of groups like the Software Sustainability Institute (\href{https://www.software.ac.uk/}{SEI}) and Better Scientific Software (\href{https://bssw.io/}{BSS}).
+SCS is defined as the use of computer tools to analyze or simulate mathematical models of real world systems \citep{smith2006systematic}.
+Given the importance of such software, scientists and engineers desire methods and tools to sustainably develop it with high quality.
+This has led to the formation of groups like the Software Sustainability Institute (\href{https://www.software.ac.uk/}{SEI}) and Better Scientific Software (\href{https://bssw.io/}{BSS}).
 The purpose of this state of the practice assessment is to understand how software quality in the LBM SCS family is impacted by software development choices, including principles, processes, and tools.
@@
 by \cite{parnas1976design} as “a set of programs whose common properties are so extensive that it is advantageous to study the common properties of the programs before analyzing individual members”.
-The common properties of the LBM family of related programs are listed in the a domain analysis in Section \ref{domainanalysis}.
+The common properties of the LBM family of related programs are listed as part of the domain analysis in Section \ref{domainanalysis}.
 The software qualities that are assessed in this report are listed in Section \ref{softwarequalities}.
-Ensuring these qualities can take significant time, but as \cite{SmithEtAl2016} shows, SCS developers often cannot see themselves meeting the required time constraints to better develop some of the artifacts that improve software quality, like detailed documentation.
-Inability to develop such documentation due to time constraints could be mitigated by the use of automatic document generation tools like Drasil.
+Ensuring software quality can take significant time, but as \cite{SmithEtAl2016} shows, SCS developers often cannot see themselves meeting the required time constraints to better develop some of the artifacts that improve qualities, like detailed documentation.
+An inability to develop such documentation due to time constraints could be mitigated by the use of automatic document generation tools like Drasil.
 The Drasil Framework consists of a collection of Domain Specific Languages (DSL) for capturing scientific documents, structures, and computing

### 2021-06-02 StateOfPractice/Peter-Notes/Report/Report.tex
edits to domain analysis section of LBM Report

@@
 
-\noindent\textbf{AHP} Analytical Hierarchy Process\\
+\noindent\textbf{AHP}: Analytical Hierarchy Process\\
 
@@
 \noindent\textbf{Goal}: “Goals capture, at different levels of abstraction, the various objectives the system under consideration should achieve.” \citep{van2001goal}\\
+
+\noindent\textbf{OTS}: Off-The-Shelf\\
 
@@
 
-\noindent\textbf{CPU}: central processing unit\\
-
-\noindent\textbf{GPU}: graphics processing unit\\
-
-\noindent\textbf{MRT}: multi-relaxation-time\\
-
-\noindent\textbf{SRT}: single-relaxation-time\\
-
-\noindent\textbf{TRT}: two-relaxation-time\\
+\noindent\textbf{CPU}: Central Processing Unit\\
+
+\noindent\textbf{GPU}: Graphics Processing Unit\\
+
+\noindent\textbf{MRT}: Multi-Relaxation-Time\\
+
+\noindent\textbf{SRT}: Single-Relaxation-Time\\
+
+\noindent\textbf{TRT}: Two-Relaxation-Time\\
 
@@
 
-Lattice Boltzmann Methods (LBM) are a family of fluid dynamics algorithms for simulating single-phase and multiphase fluid flows, often incorporating additional physical complexities \citep{chen1998lattice}.
+LBM are a family of fluid dynamics algorithms for simulating single-phase and multiphase fluid flows, often incorporating additional physical complexities \citep{chen1998lattice}.
 They consider the behaviours of a collection of particles as a single unit at the mesoscopic scale.
 These methods predict the positional probability of a collection of particles moving through a lattice structure.
-Off the shelf (OTS) Lattice Boltzmann Solvers (LBS) allow for a range of fluid and physical model input parameters, computational parameters, and output parameters.
+Off-the-shelf (OTS) Lattice Boltzmann Solvers (LBS) allow for a range of fluid and physical model input parameters, computational parameters, and output parameters.
 
@@
 There are many standardized lattice models - individual solvers within the family might only use a subset of them.
-The LBM uses the initial parameters of the fluid to find the probability of where along the lattice linkages a group of particles are most likely to travel.
+LBM use the initial parameters of the fluid to find the probability of where along the lattice linkages a group of particles are most likely to travel.
 It then moves the particles into the next node, and transfers the energy and momentum if a collision occurs.
@@
 
-This section lists all the common features among all the potential family members.
-The commonalities are organized using the following abstraction of the system, which can be used to describe all Lattice Boltzmann systems: input information, then generate the simulation and finally output the results.
+This section lists common features among potential family members.
+The commonalities are organized using the following abstraction of the system, which can be used to describe all Lattice Boltzmann systems: input information, generate the simulation, output the results.
 Section \ref{comm_lbm} describes the commonalities for the simulation step.
-Section \ref{comm_in} highlights the input information that is required for all Lattice Boltzmann systems.
+Section \ref{comm_in} highlights the input information that is required for Lattice Boltzmann systems.
 The next section, Section \ref{comm_out}, shows the common features for the output of Lattice Boltzmann systems, such as the requirement that mesh information be written to files.
-(Although the output information could simply be written to the computer’s memory, in all practical applications it is desirable to have a persistent record of the output that was created.) The final section covers qualities of the system that cannot be classified as input, simulation generation or output.
+Although the output information could simply be written to the computer’s memory, in all practical applications it is desirable to have a persistent record of the output that was created.
+The final section covers qualities of the system that cannot be classified as input, simulation generation or output.
 These commonalities are termed nonfunctional requirements of the system.
@@
 Following this, a description of the commonality is provided along with a list of related variabilities, which are given as hyperlinks that allow navigation of the document to the text describing the variability.
-Finally, each commonality ends with a summary of the history, including the date of creation and any dates of modification, along with a brief description of the modification.
-If necessary, a previous version of the document can be obtained by using the concurrent versioning system where the files are stored.
 
@@
 		\hline
-		\bf Description & All Lattice Boltzmann versions user a probability density function to give the probability that fluid has moved into a specific domain.\\
+		\bf Description & All Lattice Boltzmann versions use a probability density function to give the probability that fluid has moved into a specific domain.\\
 		\hline
@@
 		\hline
-		\bf Description & LBS provides the precision required for the particular problems it is intended to solve.\\
+		\bf Description & LBS provide the precision required for the particular problems it is intended to solve.\\
 		\hline
@@
 		\hline
-		\bf Description & LBS provides the accuracy required for the particular problems it is intended to solve.\\
+		\bf Description & LBS provide the accuracy required for the particular problems it is intended to solve.\\
 		\hline
@@
 In this case the numbers are prepended with the letter ``V''.
-The other four headings provided for each variability are: Description, Related Commonality, Related Parameter and History.
+The other three headings provided for each variability are: Description, Related Commonality, and Related Parameter.
 The related commonalities and parameters are given as a set of identifiers that respectively refer back to the previous section on commonalities or refer forward to the next section on parameters of variation.
@@
 		\hline
-		\bf Description & The boundary of the lattice can have varied conditions.\\
+		\bf Description & The boundary of the lattice can have various conditions.\\
 		\hline
@@
 		\hline
-		\bf Description & Hardware which processes the calculations\\
+		\bf Description & Hardware that processes the calculations\\
 		\hline
@@
 		\hline
-		\bf Description & Operating systems on which LBS runs.\\
+		\bf Description & Operating systems on which LBS run.\\
 		\hline
@@
 For instance, a parameter of variation could have a binding time of “specification or building” to represent that the parameter could be set at specification time, or it could be postponed until the given family member is built.
-The choice of postponing the decision until the build would be associated with the presence of a domain specific language that would allow postponing decisions on the values of the parameter of variation.
+The choice of postponing the decision until the build could be associated with the presence of a domain specific language that would allow postponing decisions on the values of the parameter of variation.
 
@@
 		\hline
-		\bf Range of Parameters & OpenMP, OpenCL, CUDA, or MPI are used if LBS is parallel.\\
+		\bf Range of Parameters & OpenMP, OpenCL, CUDA, MPI are used if the execution of the LBS is parallelized.\\
 		\hline
@@
 		\hline
-		\bf Range of Parameters & The amount of memory and storage varies between LBS. LBS sizes begin in the low gigabytes. \\
+		\bf Range of Parameters & The amount of memory and storage varies between LBS.\\
 		\hline
@@
 		\hline
-		\bf Range of Parameters & Each input and each output will have a specified the precision with which the real number is stored and/or calculated.\\
+		\bf Range of Parameters & Each input and each output will have specified the precision with which the real number is stored and/or calculated.\\
 		\hline

### 2021-06-02 StateOfPractice/Peter-Notes/Report/Report.tex
1st edits to methodology section of LBM Report

@@
 
-In this project we set out to answer the research questions found in Section \ref{researchquestions} for the Lattice Boltzmann scientific computing software domain.
+In this project we set out to answer the research questions found in Section \ref{researchquestions} for the LBM SCS domain.
 The process involved systematically measuring and analyzing members of this software family along the quality attributes found in Section \ref{softwarequalities}.
 This included gathering quantitative and qualitative measurements.
-A goal of the project was to also produce a hierarchical quality assessment for LBM software.
+A goal of the project was to produce a hierarchical quality assessment for LBM software packages.
 
@@
 Some of this was empirical software engineering related data, such as the number of files, number of lines of code, percentage of issues that are closed.
-Most was gathered by manually observing the software, it's source code, and it's artifacts, while some was gathered using the tools listed in Section \ref{tools}.
+Most of the data was gathered by manually observing the software, its source code, and its artifacts, while some was gathered using the tools listed in Section \ref{tools}.
 
 We also collected qualitative data by interviewing the software package developers and asking them the questions found in Section \ref{interviewquestions}.
-Furthermore, we solicited the assistance of domain experts to better access each software package by leveraging their experience to access the functional and non-functional domain software requirements.
-
-This section begins with the steps of the overall process we used to select, measure and compare LBM software.
-This is followed by quality definitions and how they were applied in our project.
-The rest of the section provides an overview of how candidate software packages were selected and filtered, the empirical measurements and software tools, and the analytic hierarchy process (AHP).
+Furthermore, we solicited the assistance of domain experts to better assess each software package by leveraging their experience to assess the functional and non-functional requirements for the software domain.
+
+This section begins by describing the steps of the overall process we used to select, measure and compare LBM software.
+This is followed by quality definitions and how these qualities were assessed in our project.
+The rest of the section provides an overview of how candidate software packages were selected and filtered, the empirical measurements and software tools that were used, and the analytic hierarchy process (AHP).
 

### 2021-06-02 StateOfPractice/Peter-Notes/Report/Report.tex
2nd edits to methodology section of LBM Report

@@
 
- An overview of the assessment process is given in the following steps:
+ The following steps provide an overview of how the assessment was conducted:
 
 \begin{enumerate}
-	\item Identify a broad list of candidate software packages in the domain. This is discussed in Section \ref{identifysoftware}.
+	\item List candidate software packages for the domain. This is discussed in Section \ref{identifysoftware}.
 	\item Filter the software package list. This is discussed in Section \ref{filtersoftware}.
@@
 	\item Collect empirical measures. This is discussed in Section \ref{empiricalmeasures}.
-	\item Measure using the measurement template. This is discussed in Section \ref{empiricalmeasures}. The measurement template can be found in Section \ref{measurementtemplate}.
+	\item Measure using the measurement template. This is discussed in Section \ref{empiricalmeasures}. The measurement template can be found in Section \ref{measurementtemplate2}.
 	\item Survey the developers.
-The developers of each software package of the filtered software list were contacted for voluntary interviews.
+The developers of each software package in the filtered software list were contacted for voluntary interviews.
 Four interviews were conducted in total.
@@
 In this report we adopt software quality definitions from various researchers and subject matter expert entities.
+Some of the definitions are from \cite{Smithetal2020}.
 The quality measurement results are found in Section \ref{qualityresults} of this report.
@@
 In this case effort includes the time spent finding and understanding the installation instructions, the man-time and resources spent performing the installation procedure, and the absence or ease of overcoming system compatibility issues.
-The ability to reasonably validate the installation procedure can also have a positive effect on the measure of installability.
+The ability to reasonably validate the installation procedure also has a positive effect on the measure of installability.
 Similarly, the ease of uninstallation has an affect on the measure of installability.
@@
 
-Verifiability is measured by the extent to which a set of tests can be written and executed, to demonstrate that the delivered system meets the specification \citep{sommerville}.
-Similarly to correctness, verifiability is positively correlated with the availability of a specification and with reference to domain knowledge.
-A good measure of verifiability is further positively correlated with the availability of well written tutorials that include expected output, with software unit testing documentation, and with evidence of continuous integration during the development process.
+Verifiability is measured by the extent to which a set of tests can be written and executed to demonstrate that the delivered system meets the specification \citep{sommerville}.
+Similarly to correctness, verifiability is correlated with the availability of a specification and with reference to domain knowledge.
+A good measure of verifiability is further correlated with the availability of well written tutorials that include expected output, with software unit testing documentation, and with evidence of continuous integration during the development process.
 
@@
 Reliability is thus positively correlated with the absence of errors during installation and use.
-Recoverability from errors also improves the measure of reliability.
+Recoverability from errors also improves reliability.
 In turn the presence of descriptive error messages and logs increases the ability to recover the system.
@@
 Software possesses the characteristic of robustness if it behaves “reasonably” in two situations: i) when it encounters circumstances not anticipated in the requirements specification; and ii) when the assumptions in its requirements specification are violated \citep{boehm2007software} \citep{ghezzi1991fundamentals}.
-A good measure of robustness correlates with a reasonable reaction to unexpected input like data of the wrong type, empty input, or missing files or links.
+A good measure of robustness correlates with a reasonable reaction to unexpected input, including data of the wrong type, empty input, or missing files or links.
 A reasonable reaction includes an appropriate error message and the ability to recover the system.
@@
 In this state of the practice assessment performance was not quantitatively measured.
-Instead the documentation of each software package was scanned for information that alludes to a consideration of performance, such as parallelization tools.
+Instead the documentation of each software package was observed for information that alludes to a consideration of performance, such as parallelization tools.
 
@@
 Usability is measured by the extent to which a software product can be used by specified users to achieve specified goals with effectiveness, efficiency, and satisfaction in a specified context of use \citep{nielsonusability}.
-A good measure of usability correlates with the presence of documentation, specifically tutorials, manuals, and defined user characteristics, and user support.
+A good measure of usability correlates with the presence of documentation, including tutorials, manuals, and defined user characteristics, and user support.
 Preferably the user support model has avenues to contact developers and report issues.
@@
 In this state of the practice analysis maintainability is measured by the quality of documentation artifacts, and the presence of version control and issue tracking.
-These qualities can greatly decrease the effort needed to modify software.
+These artifacts can greatly decrease the effort needed to modify software.
 There are many documentation artifacts that can improve maintainability, including user and developer manuals, specifications, README files, change logs, release notes, publications, forums, and instructional websites.
@@
 A good measure of reusability results from a large number of easily reusable components.
-Increased software modularization, smaller components with well defined interfaces, is important.
+Increased software modularization, defined as the presence of smaller components with well defined interfaces, is important.
 For this state of the practice assessment, a good measure of reusablity correlates with an increased number of code files, and the availability of API documentation.
@@
 In this context, reproducibility refers to the similarity between artifacts.
-Reproducibility can also refer to the output of the software, irrespective of the artifacts, where the scientific results are compared between software implementations, or between software and manual implementations.
+Reproducibility can also refer to the output of the software, irrespective of the artifacts, where the scientific results are compared between software implementations, or between software implementations and manually calculated results.
 This state of the practice assessment did not quantitatively measure reproducibility.
@@
 
-When forming the list and reviewing the candidate software the following properties were considered:
+The following properties were considered when creating the list and reviewing the candidate software:
 
@@
 	\item The source code must be viewable.
-	\item The empirical measures listed in Section 8 should ideally be available, which implies a preference for GitHub-style repositories.
+	\item The empirical measures should be available, which implies a preference for GitHub-style repositories.
 	\item The software cannot be marked as incomplete or in an initial development phase.
@@
 
-The initial list had 46 packages, including a few packages that were later found to not have publicly available source code, or to be incomplete.
+The initial list had 46 packages, including a few packages that were later found to not have publicly available source code, or to be in an incomplete state of development.
 
@@
 
-In order to reduce the number of members of the candidate software list, the following filters were applied. The filters were applied in the priority order listed.
+In order to reduce the number of members in the candidate software list, the following filters were applied. The filters were applied in the priority order listed.
 
@@
 	\item Usage: Software packages were eliminated if their installation procedure was missing or not clear and easy to follow.
-	\item Age: The older software packages (age being measured by the last date when a change was made) are eliminated, except in the cases where an older software package appears to be highly recommended and currently in use.
+	\item Age: The older software packages (age being measured by the last date when a change was made) were eliminated, except in the cases where an older software package appears to be highly recommended and currently in use.
 \end{enumerate}
@@
 
-While the initial list had 46 packages, filtering by scope, usage, and age, decreased the size of the list to 23 packages.
-Many of the 23 packages that were removed could not be tested as there was no installation guide, they were incomplete, source code was not publicly available, a license was needed, and the project was out of scope or not up to a standard that would support incorporating them into this study.
-Of the remaining 23 packages that were studied, some were kept on the list despite being marked as dead due to their prevalence on authoritative lists on LBM software and their surface excellence, specifically the clearly considerable time that was put into these projects.
-The final list of software that was analyzed in this project can be found in the following two tables.
+While the initial list had 46 packages, filtering by scope, usage, and age decreased the size of the list to 23 packages.
+Many of the 23 packages that were removed could not be tested as there was no installation guide, they were incomplete, source code was not publicly available, a license was needed, or the project was out of scope or not up to a standard that would support incorporating them into this study.
+Of the remaining 23 packages that were studied, some were kept on the list despite being marked as dead due to their prevalence on authoritative lists on LBM software and due to their surface excellence, specifically the considerable time that was put into these projects.
+The final list of software packages that were analyzed in this project can be found in the following two tables.
 Table \ref{alivepackages} lists packages that fell into the `alive' category as of mid 2020, and Table \ref{deadpackages} lists packages that were `dead' at that time.
@@
 
-There is considerable variation among these software packages, including their intended purpose, size, user interfaces, and their base programming languages.
+There is considerable variation among these software packages, including their intended purpose, size, user interfaces, and software languages used.
 For example, the OpenLB software package is predominantly a C++ package that makes use of hybrid parallelization and was designed to address a range of computational fluid dynamics problems \citep{heuveline2009towards}.
@@
 ESPResSo is an extensible simulation package that is specifically for research on soft matter, and is written in C++ and Python \citep{weik2019espresso}.
-The HemeLB is used for efficient simulation of fluid flow in several medical domains, and is written predominantly in C, C++, and Python \citep{mazzeo2008hemelb}.
+The HemeLB package is used for efficient simulation of fluid flow in several medical domains, and is written predominantly in C, C++, and Python \citep{mazzeo2008hemelb}.
 
@@
 Much of the data was gathered by manually reviewing the artifacts of each software package, while some was gathered using freeware tools listed in Section \ref{tools}.
-The data that was gathered is listed in the measurement template found in Section \ref{measurementtemplate}.
+The data that was gathered is listed in the measurement template found in Section \ref{measurementtemplate2}.
 Some of the data required processing other data within the template, including the status of the software package which relied on the last commit date, the percentage of issues that are closed which relied on the number of open and closed issues, and the percentage of code that is comments which relied on the number of total lines and comment lines in text-based files.
-The complete measurement template data was then analyzed using the grading template found in Section \ref{gradingtemplate}, the output of which was analyzed using the AHP described in Section \ref{AHP}.
+The complete measurement template data was then analyzed using the grading template found in Section \ref{gradingtemplate2}, the output of which was analyzed using the AHP described in Section \ref{AHP}.
 
@@
 The Analytical Hierarchy Process (AHP) is a decision-making technique that is used to compare multiple options by multiple criteria.
-In our work AHP was used for comparing and ranking the LBM software packages using the overall impression quality scores that were gathered in the measurement template found in Section \ref{measurementtemplate} using the grading template found in Section \ref{gradingtemplate}.
-AHP performs a pairwise analysis using a matrix and generates an overall score as well as quality scores for each software package.
+In our work AHP was used for comparing and ranking the LBM software packages using the overall impression quality scores that were gathered in the measurement template found in Section \ref{measurementtemplate2} using the grading template found in Section \ref{gradingtemplate2}.
+AHP performs a pairwise analysis using a matrix and generates an overall score as well as individual quality scores for each software package.
 \cite{SmithEtAl2016} shows how AHP is applied to ranking software based on quality measures.
@@
 
-\subsection{Measurement Template}
+\subsection{Measurement Template}\label{measurementtemplate2}
 The table below lists the set of measures that are used to assess each software product.
@@
 
-\subsection{Grading Template}
+\subsection{Grading Template}\label{gradingtemplate2}
 The table below lists how each quality measure of the measurement template is used to calculate an overall impression in each software quality set.\newline

### 2021-06-02 StateOfPractice/Peter-Notes/Report/Report.tex
1st edits to results section of LBM Report

@@
 
-This section presents the AHP results for qualities that were quantitatively measured.
-Qualitiative data from interviews with some developers is also found in this section.
-Working quality definitions are found in section \ref{softwarequalities}.
+This section presents the quantitative and qualitative findings of this study.
+The qualitative data that is presented was gathered through interviews with some developers.
+AHP results for qualities that were quantitatively measured are also found in this section.
+Quality definitions are found in section \ref{softwarequalities}.
 
@@
 All of the software packages that were tested have installation instructions.
-As noted previously, many of the 23 software packages removed from the original list of 46 software packages are missing documentation, including installation instructions.
-
-All 23 packages can be installed on some Unix-like systems.
-Seven packages could be installed on Windows, and 5 on macOS.
-Operating system compatibility is found in the documentation of 19 software packages.
-For this state of the practice assessment, all but one of the software packages were tested on Ubuntu.
+As noted previously, many of the twenty three software packages that were part of the original long list of forty six software packages but were later removed do not include documentation or installation instructions.
+
+All twenty three packages on the short list can be installed on some Unix-like systems.
+Seven packages could be installed on Windows, and five on macOS.
+Operating system compatibility is found in the documentation of nineteen software packages.
+All but one of the software packages were tested on Ubuntu for this state of the practice assessment.
 The remaining package was tested on CentOS.
@@
 
-All but two of the software packages have an automation procedure for at least part of the installation process.
-Most of these packages use make to automate the installation.
-A couple of packages use custom scripts for automation.
-
-Errors encountered during the installation process were sometimes quickly fixed thanks to descriptive error messages.
-At other times, error messages were vague and difficult to troubleshoot.
+All but two of the software packages have automated some of the installation process. Most of these packages use Make to automate the installation, and a few of them use custom scripts.
+
+Errors encountered during the installation process were often quickly fixed thanks to descriptive error messages.
+Systems that provided vague error messages were more difficult to troubleshoot.
+Only three software packages that displayed a descriptive error message were not recoverable, and most of these instances were due to hardware and operating system incompatibility, such as the requirement of CUDA.
 Fourteen software packages definitively broke during installation.
 A couple of packages did not provide a definitive message of the success or failure of installation.
-In these instances validating the installation required performing a tutorial or running a script, as described below, if they were available.
-The installation of most software packages could be recovered if a descriptive error message was displayed.
-Only three software packages that displayed a descriptive error message were not recoverable, and most of these instances were due to hardware and operating system incompatibility, such as the requirement of CUDA.
+In these instances, validating the installation required performing a tutorial or running a script, as described below, if these were available.
 
@@
 
-Sixteen software packages require less than 10 dependencies to be installed.
-All but one software package require less than 20 dependencies.
+Sixteen software packages require less than ten dependencies to be installed.
+All but one software package require less than twenty dependencies.
 Some packages may automatically install additional dependencies in the background.
 Eighteen of the software packages do not explicitly indicate software dependency versions.
-Some software package installation issues, specifically those occurring when manual installation of dependencies is required, may be avoided if the dependency versions are specified.
+Some software package installation issues, specifically those occurring when manual installation of dependencies is required, may be avoided if versions of dependencies are specified.
 Fifteen software packages do not have detailed instructions for installing dependencies.
@@
 Sixteen software packages have less than ten manual installation steps.
-If commands for the installation of dependencies is streamlined then none of the software packages take more than 20 steps to install.
-The average number of steps is about 8, and the fewest is 3.
+If dependencies are installed in one command then none of the software packages take more than twenty steps to install.
+The average number of steps is about eight, and the fewest is three.
 
@@
 Twelve have some sort of tutorial examples that can be run by the user.
-Most of these generate output that can be compared to the documentation.
-Some packages do not have a way to verify the output of the tutorials, and the user is left to assume correctness if there is no error prompted during the tutorial.
+Most of these tutorials provide expected output.
+Some packages do not have a way to verify the output of the tutorials, and the user is left to assume correctness if no error message is encountered.
 Some other ways of installation validation include validation scripts, automatic validation after the installation, and instructions to manually review the file system.
 
-Instructions for uninstallation is easily found for only one of the software packages.
+Uninstallation instructions were found for only one of the software packages.
 
@@
 Figure \ref{Fig_Installability} shows the installability ranking of the software packages using AHP.
-Software packages with a higher score tend to have one set of linear installation instructions that is written as if the person doing the installation has none of the dependencies installed.
-The instructions often have compatible operating system versions listed and include instructions for the installation of dependencies.
-The top ranked packages also often incorporate some sort of the automation in the installation process and have fewer manual installation steps.
+Software packages with a higher score tend to have one set of linear installation instructions that are written as if the person doing the installation has none of the dependencies installed.
+The instructions often list compatible operating system versions and include instructions for the installation of dependencies.
+The top ranked packages often incorporate some sort of automation of the installation process and have fewer manual installation steps.
 The number of dependencies a package has does not correlate with a higher score.
-Installation validation, often through test examples with expected output, is correlated with a higher score.
+The ability to validate the installation process, often through tutorials or test examples that include expected output, is correlated with a higher score.
 
@@
 Listing compatible operating system and dependency versions would decrease installation time and errors, as would adding instructions on installing dependencies.
-Errors that may be encountered during the installation process need to prompt detailed messages or logs.
-Once a software package is installed, either a transparent automatic validation needs to be performed or the user needs to be able to perform a manual validation using test examples and output.
+Installation process errors should prompt the system to display detailed messages.
+Once a software package is installed, either an automatic validation needs to be performed or the user needs to be able to perform a manual validation using test examples that include expected output.
 Finally, uninstallation instructions should be included in the documentation.

### 2021-06-03 StateOfPractice/Peter-Notes/Report/Report.tex
2nd edits to results section of LBM Report

@@
 
-Sixteen of the software packages have a requirement specification or explicitly reference domain theory.
-This information is often found in a user manual, on a web page, or is mentioned in publications by the developers.
+Sixteen of the software packages include a requirements specification artifact or explicitly reference domain theory.
+This artifact is often found within a user manual, on a web page, or is mentioned in related publications.
 In the latter case the user may need to spend significant time to find this information.
 
-Interviews with developers confirmed that many of the software packages were developed by domain experts with backgrounds in physics, mathematics and mechanical engineering.
-Many did not have formal software engineering education.
-Some development teams include computer scientists.
-Despite a lack of visible domain documentation and a resulting lower surface correctness and verifiability score, many software packages do have a strong domain background.
-
-Document generation tools are explicitly used by 12 software packages. Sphinx is used by 8 of them, and Doxygen is used by 7. Some of the packages use both.
-
-Tutorials are available for all but 5 of the software packages.
+Interviews with developers confirmed that many of the software packages are developed by domain experts with backgrounds in physics, mathematics and mechanical engineering.
+Many of these developers do not have formal software engineering education.
+Some of the development teams include computer scientists.
+Despite a lack of visible domain documentation and a resulting lower surface correctness and verifiability score, many software packages do have a strong domain background on account of their development team.
+
+Document generation tools are explicitly used by twelve software packages. Sphinx is used by eight of them, and Doxygen is used by seven. Several of the packages use both.
+
+Tutorials are available for all but five of the software packages.
 Generally they are linearly written and easy to follow.
-However, only 8 tutorials provided an expected output and it is not possible to verify the correctness of the output of the software packages that are missing this key information.
-In this case the user cannot be sure of correctness and may need to assume it if there are no immediately visible errors.
+However, only eight tutorials provide an expected output.
+It is not possible to verify the correctness of the output of the software packages that are missing this key information.
+In these cases the user may need to assume correctness if there are no visible errors.
 
 Unit tests are only explicitly available for one of the software packages.
-Code modularization of most packages allows for a user to create tests with varying degrees of effort.
+Code modularization of most packages allow for users to create tests with varying degrees of effort.
 These tests allow developers and users to verify the correctness of fragments of the source code, and in doing so better assess the correctness of the entire package.
@@
 The documentation or presence of continuous integration tools and techniques alludes to a more refined development process where faults are isolated and better recognized.
-On the surface only 2 of the packages mentioned applying the practice of continuous integration in their development process.
-
-Interviews with the developers of software packages suggest a more frequent use of both unit testing and continuous integration in the development processes than initially surveyed.
+Only two of the packages mentioned applying the practice of continuous integration in their development process.
+
+Interviews with the developers of software packages suggested a more frequent use of both unit testing and continuous integration in the development processes than initially surveyed.
 For example, OpenLB, pyLBM, and TCLB use such methods during development despite it not being explicitly clear from a surface analysis.
@@
 They also explicitly use at least one document generation tool that builds confidence of correctness.
-The top ranked software packages all include an easy to follow getting started tutorial in their documentation, and most provide expected output for these tutorials.
+The top ranked software packages all include an easy to follow getting started tutorial, and most of these include expected output.
 Only the top ranked package, Ludwig, provided unit testing.
@@
 During developer interviews it was discovered that some other packages also use continuous integration and unit testing, despite not being explicitly stated in documentation.
-Such packages, including OpenLB, pyLBM, and TCLB, have higher true correctness and verificablity than is noted from surface analysis.
+The correctness and verificablity of such packages is not measured well using surface analysis.
 
@@
 The use of document generation tools can help build confidence in correctness.
-The addition of easy to follow tutorials, along with expected output, further helps users verify the software and have confidence in its correctness.
+The addition of easy to follow tutorials further helps users verify the software and have confidence in its correctness.
 Unit testing documentation and capability, as well as the use of continuous integration tools and techniques such as Bamboo, Jenkins, and Travis CI, help verify correctness.
-Several interviewed developers alluded to difficulty with testing large numbers of features, and some even manually tested output.
+Several interviewed developers alluded to difficulty with testing large numbers of features, and some even manually tested program output.
 The use of well defined unit testing tools could decrease the time spent testing some feature.
@@
 
-Errors occurred when installing 16 of the software packages, and every instance prompted a message.
-Error messages included unrecognized commands (even when following the installation guide), missing links, missing dependencies, syntax errors in code files, and vague output.
-Several automatic installation processes could not find and load dependency packages.
-The installation files pointed to outdated external repositories.
-Seven of the installations were clearly recovered and one of the installations was assumed to be recovered due to the absence of installation verification.
+Errors occurred when installing sixteen of the software packages.
+Every instance prompted an error message.
+These messages indicated unrecognized commands (even when following the installation guide), missing links, missing dependencies, syntax errors in code files.
+In some instances the error messages were vague.
+Several automatic installation processes could not find and load dependencies.
+In these instances the installation tried to access outdated external repositories.
+Seven of the installations were recovered and verified, and one of the installations was assumed to be recovered due to the absence of any way to verify it.
 The installation of eight of the software packages could not be recovered.
-Most of these installations could not reach a dependency repository, had an incompatible error in their core installation files, or prompted vague error messages.
-
-Of the 13 software packages that installed correctly and had tutorials, 4 broke during tutorial testing.
-All four error resulted in an error message being displayed.
-One error was due to a missing tutorial dependency, another was due to an invalid command despite following the tutorial, and the final two errors were execution errors.
-Of the 4 broken tutorial instances, only the one that was missing a dependency was recoverable.
+Most of these broken installations could not find external dependencies, encountered system incompatibilities, or displayed vague error messages.
+
+Of the thirteen software packages that installed correctly and also have tutorials, four broke during tutorial testing.
+All of these instances resulted in an error message being displayed.
+One error was due to a missing tutorial dependency, another was due to an invalid command despite following the tutorial, and the final two errors were vague execution errors.
+Of the four broken tutorial instances, only the one that was missing a dependency was recoverable.
 
@@
 Figure \ref{Fig_Reliability} shows the surface reliability ranking of the software packages using AHP.
-Software packages with a high score did not break during installation, or the installation was recoverable.
-All of the top 5 ranked packages had tutorials.
+Software packages with a high score either did not break during installation, or the broken installation was recoverable.
+All of the top five ranked packages have tutorials.
 One of these packages, pyLBM, broke during tutorial testing, but a descriptive error message helped in recovery.
 
-Overall, lower ranked software packages were lacking clear documentation, testing or tutorial examples, and descriptive error messages, and had broken automatic dependencies.
+Overall, lower ranked software packages are lacking clear documentation, testing or tutorial examples, and descriptive error messages, and have broken dependencies.
 Thus, regarding surface reliability, software packages would benefit from clear up-to-date installation and tutorial documentation, and descriptive error messages.
@@
 
-The software packages were tested to handle unexpected input, including incorrect data types, empty input, and missing files or links.
+The software packages were tested for handling unexpected input, including incorrect data types, empty input, and missing files or links.
 Success predicated on a reasonable response from the system, including appropriate error messages and an absence of unrecoverable system failures.
@@
 They output descriptive error messages or did not crash.
-Software packages with a lower surface robustness score had not installed correctly, so their score here may not be a true reflection of runtime robustness.
-Similarly, all software packages that installed correctly and require plain text input files handled an unexpected change, a replacement of new lines with carriage returns, to these input files correctly.
+Software packages with a lower surface robustness score had not installed correctly, so their robustness score may not be a true reflection of runtime robustness.
+Similarly, all software packages that installed correctly and require plain text input files correctly handled an unexpected change to these input files, including a replacement of new lines with carriage returns.
 
@@
 
-Each software package was not quantitatively measured for surface performance.
-Although the software packages all apply Lattice Boltzmann methods to solve scientific computing problems, the packages focus on varied domains, with varying parameters, and are technically different from each other.
+The software packages were not quantitatively measured for surface performance.
+Although the software packages all apply LBM to solve scientific computing problems, the packages focus on varied domains, with varying parameters, and are technically different from each other.
 Due to this, a comparison of performance is not appropriate.
-Instead, in this project we looked through each software package's artifacts for evidence that performance was considered.
-The artifacts of 17 software packages mentioned the use of some sort of parallelization for execution.
+In this project we instead looked through each software package's artifacts for evidence that performance was considered.
+The artifacts of seventeen software packages mentioned the use of some sort of parallelization tool or technique.
 This included GPU processing and the CUDA parallel computing platform.
@@
 In the Ludwig package, a so-called mixed mode approach is used where fine-grained parallelism is implemented on the GPU, and MPI is used for even larger scale parallelism \citep{gray2013ludwig}.
-While some software packages required CUDA and GPU processing, some had the option of using either the GPU or the CPU.
+While some software packages require CUDA and GPU processing, some have the option of using either the GPU or the CPU.
 The packages that require GPU and CUDA have better performance at the expense of installability and surface reliability.

### 2021-06-03 StateOfPractice/Peter-Notes/Report/Report.tex
3rd edits to results section of LBM Report

@@
 
-Of the software packages that were tested, most installation instructions are located in one place, often in an instruction manual or on a web page.
-Sometimes incomplete installation instructions are found on a home page, with more detailed instructions located on another web page or within formal documentation.
+Of the software packages that were tested, most installation instructions are located in one place, often in an instruction manual or on a web-page.
+Sometimes incomplete installation instructions are found on a home page, with more detailed instructions located on another web-page or within formal documentation.
 Maintainability and correctness of these instructions could be improved if they were limited to one location.
@@
 Sixteen of the software packages include a requirements specification artifact or explicitly reference domain theory.
-This artifact is often found within a user manual, on a web page, or is mentioned in related publications.
+This artifact is often found within a user manual, on a web-page, or is mentioned in related publications.
 In the latter case the user may need to spend significant time to find this information.
@@
 Software package artifacts were reviewed for the presence of a tutorial, a user manual, documented user characteristics, and a user support model.
-In total 18 software packages have a tutorial and 13 have a user manual.
-The tutorials vary in scope and substance, and 8 include an expected output to compare user results to.
+In total eighteen software packages have a tutorial and thirteen have a user manual.
+The tutorials vary in scope and substance, and eight include an expected output.
 Most user manuals are in the form of a file that can be downloaded, while some are rendered on a web-page.
 Some packages do not have a user manual but do have useful documentation distributed on their web-pages.
-Expected user characteristics are documented in 4 software packages.
+Expected user characteristics are documented in four software packages.
 Users are typically scientists or engineers trying to test a new application.
@@
 All but one of the packages have a user support model, and several of them have multiple avenues of user support.
-The most popular avenue of support is git, followed by email and forums.
+The most popular avenue of support is Git, followed by email and forums.
 One software package has a FAQ page.
@@
 Figure \ref{Fig_Usability} shows the surface usability ranking of the software packages using AHP.
-Software packages with a high score have a tutorial and user manual, and sometimes have documented user characteristics, and have at least one user support model, with many having several.
+Software packages with a high score have a tutorial and user manual, sometimes have documented user characteristics, and have at least one user support model.
+Many packages have several user models.
 
 Interviews with developers revealed several usability issues.
-Some users have misunderstood boundaries of LBM and computational fluid dynamics in general, and have combined or applied methods that are not physically sound together.
+Some users have misunderstood the boundaries of LBM and computational fluid dynamics, and have combined or applied methods that are not physically sound.
 Sometimes users have applied LBM to poorly defined or inappropriate fluid dynamics problems.
-The users do not realize the limitations of the methods, of the software, and of how much time and effort is required to properly model a problem that can technically be modeled by the software.
-The software is not as easy as out of the box plug and play.
+The users do not realize the limitations of the methods, of the software, and do not understand the requirements to properly model a problem with the software.
+The software packages are not designed to be used out of the box in a plug and play fashion.
 Developers of some of the software packages mitigated this by editing the source code to prevent unsound combination from being applied, and by updating the documentation to better inform users of LBM limitations, and of the requirements to properly model appropriate problems, including what algorithms and parameters to use.
 
-Some additional infrequent software usability issues were commented on.
+Some additional but infrequent software usability issues were commented on by the developers.
 Users have had trouble with installation and understanding how to maneuver the interfaces and how to set up or run models.
-These issues are addressed by various user support models, including frequently asked questions sections on the software package websites, user guides, and hardware and software requirement specifications.
+These issues are addressed by various user support models, including frequently asked questions sections on the software package websites, user guides, and hardware and software requirements specifications.
 
@@
 Every type of artifact or file that is not a code file was recorded.
-The software packages were also reviewed for version numbers.
-This information could be used to better troubleshoot software issues and organize documentation.
-All but 3 software packages have source code release and documentation version numbers.
+The software packages were also reviewed for software release and documentation version numbers.
+This information could be used to better troubleshoot issues and organize documentation.
+All but three software packages have source code release and documentation version numbers.
 
@@
 
-Issue tracking is used in 22 software packages, 15 of which use git, 6 use email, and one uses SourceForge.
-Most software packages that use git have most of their issues closed, and only 3 have less than 50 percent of their issues closed.
-Furthermore, most packages that use git for issue tracking use GitHub as a version control system, one package uses CVS for issue tracking, and 8 packages do not appear to use any such system.
+Issue tracking is used in twenty two software packages, fifteen of which use Git, six use email, and one uses SourceForge.
+Most software packages that use Git have most of their issues closed, and only three have less than fifty percent of their issues closed.
+Furthermore, most packages that use Git for issue tracking use GitHub as a version control system, one package uses CVS for issue tracking, and eight packages do not appear to use any such system.
 
@@
 Packages with a higher percentage of comments were designated as more maintainable.
-Comments represent more than 10 percent of code files in 15 packages, and the average percentage of comments is 13.7.
+Comments represent more than ten percent of code files in fifteen packages, and the average percentage of code comments is about fourteen.
 
@@
 These packages also appear to reasonably handle tracked issues, having most of their issues closed.
-Their code files are well commented with more than 10 percent of the code being comments.
-
-Interviews with developers revealed that most projects do have a defined process for accepting contributions from current team members or those that at some time had been linked to the packages or developers.
+Their code files are well commented with more than ten percent of the code being comments.
+
+Interviews with developers revealed that most projects do have a defined process for accepting contributions from team members.
 The packages rarely get contributions from outside developers, but the process would be similar as for the aforementioned group.
@@
 Some developers noted that their software package does not have well defined contributing guide in the repository but it might be a good idea to add one in the near future.
-They would be very happy to see spontaneous contribution from outside of their organization, but now it is currently not the case.
+They would be very happy to see contribution from outside of their organization, but currently this does not happen.
 
 Furthermore, maintainability has been addressed by increasing source code modularity, reducing duplicate information, and improving abstraction.
-Several software packages have had large sections of their code base redeveloped with languages that the developers felt are more understandable and readable, and that are better supported and have preferred dependencies.
+Several software packages have had large sections of their code base redeveloped with languages that the developers felt are more understandable and readable, and that are better supported.
 Data structures have also been redeveloped and storage has been improved.
-One software developer mentioned that the mathematical foundations that the geometries and models are designed from have been re-evaluated and optimized.
+One software developer mentioned that the mathematical foundations that the geometries and models are designed from had been re-evaluated and optimized.
 
@@
 
-A high degree of code modularity and abstraction was noted by one developer as a measure to ensure ease of future changes.
+A high degree of code modularity and abstraction was noted by one developer as a measure to ensure the ease of future changes.
 They also noted that some of the code base was transitioned from c to c++, which could ease modifiability of that software package.
@@
 The stated intent of these benchmarks is to showcase the validity and usefulness of the package in order to stimulate the development of third-user extensions.
-The Palabos package identifies as as a development framework for modeling problems in various computational fluid dynamics areas.
+The Palabos package identifies as a development framework for modeling problems in various computational fluid dynamics areas.
 
@@
 
-The total number of source code files was measured for each software package.
-A larger number of source files was associated with increased reusability since this generally represented either more software package functionality or better modularization.
+Each software package was measured for the total number of source code files.
+A larger number of source files was associated with increased reusability due to increased modularization.
 The software packages were also reviewed for the presence of API documentation, which indicates that a software package was developed with interaction between other software applications in mind.
@@
 The highest scoring packages, ESPResSo and waLBerla, have extensive functionality, including graphical visualizations as well as modeling that does not use LBM.
-For this reason a comparison between other software packages is not on a level field, but regardless of this these packages do have an abundance of reusable components.
+For this reason a comparison with other software packages is not on a level field.
+However, these packages do have an abundance of reusable components.
 
 There was a strong focus on modularity when designing the waLBerla framework in order to enhance productivity, reusability, and maintainability \citep{bauer2021walberla}.
-The software design has enabled for waLBerla to be successfully applied in several projects as a basis for various extensions \citep{bauer2021walberla}.
+Its software design has enabled waLBerla to be successfully applied in several projects as a basis for various extensions \citep{bauer2021walberla}.
 
@@
 Ten random source code files of each software package were reviewed for several measures.
-This measure of surface understandability may not perfectly reflect each package due to this limitation.
+This measure of surface understandability may not perfectly reflect each package due to the limitation of only testing ten files.
+
 All of the packages appear to have a consistent indentation and formatting style.
@@
 Generally, the software packages use consistent, distinctive, and meaningful code identifiers.
-Only 4 packages appear to use vague identifiers.
-Hard coded constants were observed in the source code of 12 packages.
+Only four packages appear to use vague identifiers.
+Hard coded constants were observed in the source code of twelve packages.
 The constants are used for various parameters, mathematical constants, and matrix definitions.
-All of the packages are well commented.
-Comments clearly indicate what is being done.
-Domain algorithms are noted in the source code of 11 packages.
-All of the packages were well modularized.
-When observing the source code files, it was found that 13 of the packages had a consistent style and order for function parameters.
+All of the packages are well commented and the comments clearly indicate what is being done.
+Domain algorithms are noted in the source code of eleven packages.
+All of the packages are well modularized.
+When observing the source code files, it was found that thirteen of the packages have a consistent style and order for function parameters.
 
@@
 Figure \ref{Fig_Understandability} shows the surface understandability ranking of the software packages using AHP.
-Software packages with a high score have a consistent indentation and formatting style, and consistent, distinctive and meaningful code identifiers.
+Software packages with a high score have a consistent indentation and formatting style, and consistent, distinctive, and meaningful code identifiers.
 They also have hard coded constants, and explicitly identify mathematical and LBM algorithms.
-Their comments are clear and indicative of what is being done.
+Their comments are clear and indicate what is being done in the source code.
 The source code is well modularized and structured.
@@
 Software developers noted that in their opinion users have generally found their packages to be understandable.
-One developer commented that some users have attempted to run physically incompatible LBM methods, and the solution was to edit the code to prevent such combinations, as well as update the documentation to prevent misunderstanding the methods.
+One developer commented that some users have attempted to run physically incompatible LBM methods, and the solution was to edit the code to prevent such combinations, as well as to update the documentation to prevent misunderstanding the methods.
 Similarly, another developer noted that some users had issues setting up parameters for LBM schemes.
-The solution was to this was to update the interface where these parameters are set, as well as adding functionality to test the stability of the parameters.
-A third developer noted that some users lack the background knowledge to easily model flow problems using their software.
-A frequently asked questions section was added to the source package website to help users find answers to common questions.
-The package also has detailed documentation, including guides and usage requirements specification.
+The solution to this was to update the interface where these parameters are set, as well as to add functionality to test the stability of the parameters.
+A third developer noted that some users lack the background knowledge to easily model fluid dynamics problems using their software.
+A frequently asked questions section was added to their package website to help users find answers to common questions.
+The package also has detailed documentation, including guides and usage requirements specification, to better help user understand the software.
 
@@
 These tools support the development of end products but do not become part of them, unlike dependencies that remain in the application once it is released \citep{ghezzi1991fundamentals}.
-The following type of development tools were explicitly noted in the artifacts or web pages of the 23 LBM packages that were assessed.
+The following type of development tools were explicitly noted in the artifacts or web-pages of the 23 LBM packages that were assessed.
 It is likely that other tools, such as debuggers, were used but are not specified here.
@@
 
-The following types of dependencies were explicitly noted in the artifacts or web pages of the 23 LBM packages that were assessed.
+The following types of dependencies were explicitly noted in the artifacts or web-pages of the 23 LBM packages that were assessed.
 It is possible that other types of dependencies are part of these software packages but are not clearly specified in their artifacts or web sites and because of that they are not listed here.
@@
 Their work needed to be coordinated and managed.
-The following types of project management tools were explicitly noted in the artifacts, web pages, or interviews with the developers of the 23 LBM packages that were assessed.
+The following types of project management tools were explicitly noted in the artifacts, web-pages, or interviews with the developers of the 23 LBM packages that were assessed.
 As with development tools and dependencies, it is possible that other types of project management tools were used to coordinate and manage the projects but are not specified and because of that they are not listed here.
@@
 		Are there installation instructions? ({yes, no})\\
-		Are the installation instructions in one place? Place referring to a single document or web page. ({yes, no, n/a})\\
+		Are the installation instructions in one place? Place referring to a single document or web-page. ({yes, no, n/a})\\
 		Are the installation instructions linear? Linear meaning progressing  in a single series of steps. ({yes, no, n/a})\\
@@
 	Are there installation instructions? ({yes=1, no=-1})\\
-	Are the installation instructions in one place? Place referring to a single document or web page. ({yes=1, no=0, n/a=0})\\
+	Are the installation instructions in one place? Place referring to a single document or web-page. ({yes=1, no=0, n/a=0})\\
 	Are the installation instructions linear? Linear meaning progressing  in a single series of steps. ({yes=1, no=0, n/a=0})\\

### 2021-06-03 StateOfPractice/Peter-Notes/Report/Report.tex
4th edits to results section of LBM Report

@@
 Software packages were not quantitatively measured for traceability.
-In this project we asked developers to comment on traceability when we interviewed them. They were asked to comment on their documentation and how it fits into their development process.
-
-One developer noted that all major additions to their package had accompanying changes to artifacts, especially noting user documentation, and how it is updated to note major usability issues.
-Considerable effort had been put into the documentation.
-They want to lower the entry barrier for new developers so there is considerable documentation aimed at developers, and it is frequently updated.
-This documentation informs developers on how to get started, orients them to the artifacts, source code, and system architecture, as well as how the software package build system works, and how the coupling between the simulation engine and the interface works.
-
-Another developer noted the importance of documentation for both user and developers of their software.
-New features are always well documented in the artifacts.
-The developers use documentation to stay up to date on the status of the software package, and to expand or use features, like computational models or algorithms.
-This is necessary so that the coding standard for these models is kept consistent as developers come and go.
+In this project we asked developers to comment on traceability, specifically on their software package's documentation and how it fit into their development process.
+
+One developer noted that all major additions to their package had accompanying changes to artifacts and documentation.
+They noted that considerable effort had been put into the documentation.
+They further commented that they want to lower the entry barrier for new developers, and because of that their package has a considerable amount of developer documentation.
+This documentation informs developers on how to get started, and orients them to the artifacts, source code, and system architecture, as well as how the software package build system works, and how the coupling between the simulation engine and the interface works.
+
+Another developer noted the importance of documentation for both the users and developers of their software.
+New features are always added to the documentation.
+The developers use documentation to stay up to date on the status of the software package, and to help expand features, like computational models or algorithms.
+This is necessary so that the coding standard for these models is kept consistent with new developers.
 
 The importance of documentation for both users and developers was stressed throughout the interviews.
-However, it was noted several times that a lack of time and funding had negative affect on the documentation.
-Most of these developers are scientific researchers evaluated on the scientific papers that they produce.
+However, it was noted several times that a lack of time and funding has a negative affect on the documentation.
+Most of the developers are scientific researchers evaluated on the scientific papers that they produce.
 Writing and updating documentation is something that is done in free time, if that time arises, and if writing documentation is a concern.
@@
 
-One developer noted the use of documentation generators like Doxygen. It would be advisable for more projects to use such automatic document generation tools, if they are a good technological fit.
+One developer noted the use of documentation generators like Doxygen. It would be advisable for more projects to use such automatic document generation tools, since some projects do not do this.
 
@@
 The packages tended to not explicitly use well-known development models.
-This was also noted in the interviews with developers detailed below.
+This was also noted in the interviews with developers, as detailed below.
 The development teams of these packages are fairly small and easily organized without the need for such processes.
@@
 While this information could help developers, and would improve transparency, the small close-knit nature of the development teams make explicitly publicly specifying this information practically unnecessary.
-Version release notes were found in 9 of the software packages.
+Version release notes were found in nine of the software packages.
 
@@
 Addressing this issue would help with code modifiability and reusability.
-Updating the development process would help resolve this issues and prevent such issues in the future.
-Improving the visibility of software changes and the peer review process would also help.
+Updating the development process would help resolve this issue and prevent such issues in the future.
+Improving the visibility of software changes and a peer review process would also help.
 Improving the software engineering education or experience of developers is also an idea that was brought up by several developers.
 Specifically, developers should always write code that is decoupled and modular, and should keep in mind the visibility of their contributions by better updating documentation and ensuring that their contributions are transparent to the rest of the developers.
-This would help catch issues in the code contributions, and improve source code maintenance.
+This would help catch issues in the contributions, and improve source code maintenance.
 
@@
 These practices decrease development and maintenance times.
-Another developer mentioned that their package had two sets of code, for executing the models on the CPU and GPU, and that maintenance was decreased by introducing macros, the use of which is now part of the development process.
+Another developer mentioned that their package had two sets of code, for executing the models on the CPU and GPU, and that maintenance was decreased by introducing macros, a practice which became a common part of the development process.
 
@@
 The developers noted similar project management processes.
-In teams of only a couple of developers, the addition of new features or major changes were discussed with the entire team.
-Projects with more than a couple of developers had lead developer roles.
-These lead developers would review potential additions to the software.
+For teams of only a couple of developers, the addition of new features or major changes are discussed with the entire team.
+Projects with more than a couple of developers have lead developer roles.
+These lead developers review potential additions to the software.
 The software packages use GitHub for managing the project.
@@
 One developer noted that they compare the results of their methods against manually calculated results.
-These tests comparing these results are automatically run for all source code changes that are suggested.
+These comparison tests are automatically run for all source code changes.
 The tests are run when a pull request is opened on GitHub.
@@
 Several developers noted that they currently do not have a system in place to test for reproducibility, but it is of interest and could be implemented in the future.
-Generally, the mathematical foundations of the models are verified, but the output of the software packages is not compared to other output.
+Generally, the mathematical foundations of the models are verified, but the output of the software package is not compared to other output.
 Depending on the package and how it outputs LBM solutions, it may not be practical or feasible.
@@
 The source code and artifacts of some software packages may be reproducible.
-There is considerable variance in the quality of the software specification and other developer documentation.
-Some packages are well detailed, and translating the specification into source code will produce similar results across developers.
+There is considerable variance in the quality of software specifications and other developer documentation.
+Some packages are well detailed, and translating the specifications into source code will produce similar results across developers.
 
@@
 
-One developer noted that their documentation was meant for users that are already familiar with underlying physics and computational fluid dynamics methods, and that these concepts are not explained in detail.
+One developer noted that their documentation was meant for users that are already familiar with the underlying physics and computational fluid dynamics methods, and that these concepts are not explained in detail.
 The documentation focuses solely on how to technically use the software package, and includes a user guide and tutorial walk through of how to set up and run a simulation.
-With this in mind the developer believes that their documentation is in reasonable shape for a those with a minimum knowledge of the underlying physics to learn to use the package.
+With this in mind the developer believes that their documentation is in reasonable shape for users with a minimum knowledge of the underlying physics.
 If new users have technical questions these can be addressed in further revisions of the documentation.
-New developers should also find that the code documentation is reasonably clear and useful.
-Information that is missing, like information explaining dependency packages, is reasonably referenced.
+New developers should also find that the documentation is reasonably clear and useful.
+Information that is missing, like detailed explanations of dependencies, is referenced in the documents.
 
@@
 
-Looking at the top three ranked packages.
-ESPResSo had achieved a relative high score in installability, surface correctness and verifiability, surface usability, maintainability, reusability, and visibility and transparency.
+Looking at the top three ranked packages, ESPResSo had achieved a relative high score in installability, surface correctness and verifiability, surface usability, maintainability, reusability, and visibility and transparency.
 Ludwig scored high in surface correctness and verifiability, surface robustness, surface usability, maintainability, and visibility and transparency.

### 2021-06-04 StateOfPractice/Peter-Notes/Report/Report.tex
1st edits to Answers to R Questions in LBM Report

@@
 
-The software packages were examined for artifacts.
-Below we have grouped these into common, less common, and rare artifacts.
+The software packages were examined for the presence of artifacts, which were then categorized by frequency.
+Below we have grouped them into common, less common, and rare artifacts.
 Common artifacts were found in 16 to 23 of the software packages.
 Less common artifacts were found in 8 to 15 of the software packages.
-Rare artifacts were found in 1 to 27 of the software packages.
+Rare artifacts were found in 1 to 7 of the software packages.
 
@@
 The following artifacts were commonly found in the 23 software packages that were tested.
-All of the top four AHP ranked packages, ESPREsSo, Ludwig, Palabos, and OpenLB have each of these artifacts, except for a clear requirements specification or theory notes.
-Only three of the packages were found to have that artifact.
+All of the top four AHP ranked packages, ESPREsSo, Ludwig, Palabos, and OpenLB have each of these artifacts, except for a requirements specification or theory notes.
+Only three of the top four ranked packages were found to have such an artifact.
 
@@
 
-These common artifacts contribute to the quality of the software in various ways.
-A list of authors and developers helps potential users and contributors more easily contact project members to answer questions affecting many software qualities.
-A bug tracker helps with organizing improvements to the software, also improving many software qualities.
+These common artifacts contribute to the quality of the software in the following ways.
+A list of authors and developers helps potential users and contributors contact project members to answer questions affecting many software qualities.
+A bug tracker helps with organizing improvements to the software.
 Dependency notes help users install the software, as does an installation guide.
 Makefiles or other automated build files decrease human error in the installation process.
-Visible licenses promote usage of the software.
-Requirements specifications, linked theory notes, and related articles and publications help clarify the software and make it more understandable, decrease ambiguity, and help to verify its correctness.
+Licenses promote usage of the software.
+Requirements specifications, linked theory notes, and related articles and publications help make the software more understandable, decrease ambiguity, and help with verifying its correctness.
 README files also help with understandability.
-Meanwhile tutorials help users become familiar with using the software.
+The tutorials help users become familiar with using the software.
 
@@
 Only one of those four packages does not have a user manual or guide, but it does have a detailed and informative website.
-Another one of the top four packages does not have any visible design documentation.
-A third package from that list does not appear to use a version control system.
-It is probably that such a system is in use but is not visible outside of the development team since the package website notes a current version number.
+Another one of the top four packages does not appear to have publicly visible design documentation.
+A third package from the list does not appear to use a version control system.
+It is possible that such a system is used by the package since its website notes package version numbers, but the artifacts do not explicitly state the use of such a system.
 
@@
 Design documentation helps with maintaining, modifying, and reusing the software.
-Version control also helps to improve these qualities, as well as traceability.
+Version control also helps to improve these qualities, as well as with traceability.
 Functional specifications and notes clarify the software, contributing to usability and understandability.
-A user manual also benefits those qualities and with installability, and correctness and verificability.
+A user manual also helps with those qualities and with installability, and correctness and verificability.
 Performance notes suggest that performance was considered when developing the software.
@@
 \subsubsection{Rare Artifacts}
-The following artifacts were rarely found in the 23 software packages that were tested.
+The following artifacts were rarely found in the 23 software packages that were assessed.
 It is not common for the top four AHP ranked packages to have many of these artifacts.
@@
 Three of these packages have information on contributing to the project.
-Two of them have a FAQ section or forums.
-One has verification and validation notes and a video guide explaining how to use the software.
+Two of them have a FAQ section or forum.
+One has verification and validation notes, and a video guide explaining how to use the software.
 
@@
 Developer and contributor manuals and guides help with maintainability, visibility and transparency.
-FAQs and forums increase usability of software.
+FAQs and forums improve usability of software.
 Verification and validation notes can be used to help check if a system meets specifications.
@@
 \subsection{Tools Used - DONE 1st Draft}
-The purpose of tools is to support the development, verification, maintenance, and evolution of software, software processes, and artifacts \citep{ghezzi1991fundamentals}.
-Many tools are used by current LBM software packages.
+Software tools are used to support the development, verification, maintenance, and evolution of software, software processes, and artifacts \citep{ghezzi1991fundamentals}.
+Many tools are used by LBM software packages.
 The tools noted here are subdivided into development tools, dependencies, and project management tools.
@@
 
-The use of the above tools can verify the correctness of software during its development.
+The above tools can verify the correctness of software during its development.
 Only two of the software packages that were assessed mentioned using continuous integration tools like Travis CI.
@@
 One of the packages explicitly noted the use of unit testing, but did not specify the tool that was used.
-Likewise the use of a tool for verifying the correctness of output was noted by one of the developers.
-It is likely that similar tools were used when developing other members of the software package list.
+Likewise, the use of a tool for verifying the correctness of output was noted by one of the developers.
+It is likely that similar tools were used when developing other software packages.
 
@@
 They also all use various technical and domain specific libraries.
-Technical libraries include visualization, data analysis, and message passing libraries among others.
+Technical libraries include visualization, data analysis, and message passing libraries.
 Domain specific libraries are scientific computing libraries.
-Specific libraries are not listed here but they fall into the above categories.
-Libraries that are not explicitly stated in any artifact may fall outside of the categories.
+Libraries that are explicitly stated in artifact, and were not noted during our observations, may fall outside of these categories.
 
@@
 
-Collaboration tools are noted to have been used when developing the software projects.
+Collaboration tools are noted as being used when developing the software projects.
 Most often email and video conferencing is used.
-Project management software was not explicitly mentioned, but it is possible that some of the project use such software.
-Many of the projects are located on GitHub, and it's developers use the platform to help manage their projects, especially bug related issues.
+Project management software was not explicitly mentioned, but it is possible that some of the projects use such software.
+Many of the projects are located on GitHub, and its developers use the platform to help manage their projects, especially bug related issues.
 Most of the projects appear to use change tracking and version control tools.

### 2021-06-04 StateOfPractice/Peter-Notes/Report/Report.tex
2nd edits to Answers to R Questions in LBM Report

@@
 
+\noindent\textbf{OS}: Operating System\\
+
 \noindent\textbf{Program Family}: A set of programs that are analyzed and designed together starting from the initial stages of the software development life-cycle.\\
@@
 
-Most of the assessed software packages do not explicitly state in their artifacts the motivations or design principles that were considered when developing the software.
-One package, Sailfish, indicates that shortening development time was considered in early stages of design, with the developers opting for using Python and CUDA/OpenCL to achieve this without sacrificing any computational performance.
+Most of the software packages do not explicitly state in their artifacts the motivations or design principles that were considered when developing the software.
+One package, Sailfish, indicates that shortening the development time was considered in early stages of design, with the developers opting for using Python and CUDA/OpenCL to achieve this without sacrificing any computational performance.
 The goals of that project are explicitly listed as performance, scalability, agility and extendability, maintenance, and ease of use.
-The project scored well in these listed categories during our assessment, for the categories that were measured.
+The project scored well in these categories during our assessment.
 
 Processes, like methods, are ways of doing things, especially in an orderly way; while methodologies are defined as systems of methods \citep{ghezzi1991fundamentals}.
-It is not explicitly indicated in the artifacts of most of these packages that development involved following any specific model, like a waterfall of agile development model.
+It is not explicitly indicated in the artifacts of most of the packages that development involved following any specific model, like a waterfall or agile development model.
 One developer noted that while no formal model is used, their development model is something similar to a combination of agile and waterfall development models.
-The developers teams of the LBM packages are fairly small and easily organized without the need for such methodologies.
+The developer teams of the LBM packages are fairly small and easily organized without the need for such models.
 
 Seven of the software packages contain artifacts outlining the general development process, like basic instructions on how to contribute.
-Most of the development teams accept outside contributors, but generally the teams are tight-knit, often working at the same institution.
+Most of the development teams accept outside contributors, but generally the teams are centralized, often working at the same institution.
 
 The developers that were interviewed all noted similar project management processes.
-In teams of only a couple of developers, the addition of new features or major changes are discussed with the entire team.
-Projects with more than a couple of developers have lead developer roles.
+In teams of only a couple of developers, additions of new features or major changes are discussed with the entire team.
+Projects with more than a couple developers have lead developer roles.
 These lead developers review potential additions to the software.
-One of the developers noted in their interview that a peer review process is used to review major changes and additions.
-
-The software packages frequently use GitHub for managing the project. Typically there are several development branches and changes from these branches are merged into the master branch.
+One of the developers that was interviewed noted that a peer review process is used to assess major changes and additions.
+
+Many software packages use GitHub for managing the project. Typically there are several simultaneous development branches.
 
 Documentation was also noted as playing a significant role in the development process, specifically with on-boarding new developers.
-An intention is to lower the entry barrier for these new contributors.
+A goal of documentation is to lower the entry barrier for these new contributors.
 The documentation provides information on how to get started, orients the user to artifacts and the source code, and explains how the system works, including the so-called simulation engine and interface.
 The use of document generation tools is mentioned in the artifacts of 12 software packages, and was noted during interviews with developers.
-The mentioned document generation tools are Sphinx and Doxygen.
-
-Two types of software changes were noted during interviews.
+Document generation tools that were mentioned include Sphinx and Doxygen.
+
+Two types of software changes were discussed during interviews with developers.
 One is feature additions, which arise from a scientific or functional need.
-The development of these changes was noted to involve more formal discussions within the development team and the execution involves lead developers.
+The development of such changes was noted to involve more formal discussions within the development team and the addition of them into the main software branch involves lead developers.
 The other change type is code refactoring, which only sometimes involves formal discussions with the development team.
 New developers were noted to play an increased role in these changes compared to the former changes.
-Software bugs are typically addressed in a similar fashion as code refactoring, and issue trackers are commonly used to manage these changes.
-
-Interviews with the developers of software packages also revealed a more frequent use of both unit testing and continuous integration in the development process than was found by only examining the artifacts.
+Software bugs are typically addressed in a similar fashion as code refactoring, and issue tracking is commonly used to manage these changes.
+
+Interviews with the developers of software packages also revealed a more frequent use of both unit testing and continuous integration in the development process than was found by only assessing the artifacts.
 The use of automatic installation processes is also common.
@@
 
-The following points should be considered to improve LBM software quality.
+The following points should be considered when developing high quality LBM software.
 
@@
 \begin{itemize}
-	\item Include OS compatibility, including versions.
+	\item Include OS compatibility, including specific OS versions.
 	\item Include complete installation instructions.
-	\item Include all dependencies in the installation instructions, required versions, and how to install them.
+	\item Installation instructions should be written as if the user does not have any dependencies installed and is installing on a clean OS.
+	\item List all dependencies in the installation instructions, dependency versions, and how to install them.
 	\item If possible, automate the installation of dependencies.
-	\item Installation instructions should be written as if the user does not have any dependencies installed and is installing on a clean OS.
-	\item Limit installation instruction location to one location.
 	\item Automate the installation process as much as possible.
+	\item The installation instruction should only be in one location.
 	\item Include descriptive error messages for errors encountered during installation.
@@
 	\item Ensure the above information is easy to find. Consider adding it to the user manual. 
-	\item Development teams should include both domain experts and experiences software developers.
+	\item Development teams should include both domain experts and experienced software developers.
 	\item Use and make public detailed documentation. Consider using automatic document generation tools.
-	\item Include detailed tutorials with expected output.
-	\item Use unit testing during development. Make public the facilities for unit testing.
+	\item Include detailed tutorials that include expected output.
+	\item Use unit testing during development and make them public.
 	\item Modularize the source code.

### 2021-06-06 StateOfPractice/Peter-Notes/Report/Report.tex
3rd edits to Answers to R Questions in LBM Report

@@
 
-Developers were asked to comment on obstacles in their development process, obstacles encountered by user, and potential future obstacles with the software packages.
-
-Several development process obstacles were noted.
-One developer noted that their small development team, common in LBM software packages, has a lack of time to implement new features.
+Developers were asked to comment on obstacles in their development process, obstacles encountered by users, and potential future obstacles.
+
+Several development process obstacles were identified.
+One developer noted that their small development team has a lack of time to implement new features.
+Small development teams are common for LBM software packages.
 Team members are almost always part of the same institute or already know each other from other projects.
@@
 
-A lack of software development experience was noted by the developer of a third package where many of the team members on the project are domain experts.
-There is a steep learning curve to get these team members to contribute good quality source code.
-This has been somewhat addressed as the code is now generally written in a way that the developer thinks makes it as easy as possible for contributions.
-
-The developer also noted that there is a lack of incentive and funding in academia for developing good scientific software that people end up using.
+A lack of software development experience was noted by another developer.
+Many of the team members on their project are domain experts and there is a steep learning curve to get these team members to contribute good quality source code.
+It was further noted that this has been somewhat addressed, as the code has been re-written to best ensure ease of future contributions.
+The same developer also noted that there is a lack of incentive and funding in academia for developing widely used scientific software.
 There are no journals that publish such source code.
-There is no place for the development of good quality software.
-Scientific software is often developed by the researchers that temporarily use it in their own research and keep the development and use in-house.
-This software often does not meet the standards that would be required by external users.
-Sometimes the software is commercialized but it is not open source.
-
-Documentation is important, but it's quality could be improved.
-There is often no time or funding for maintaining quality documentation for software that is rarely used externally.
-Furthermore, documentation generally does not teach the underlying metaphysics.
-It sometimes provides an introduction to these topics.
-Users would need to already be familiar with these topics or spend time referencing suitable textbooks.
-The documentation generally focuses on explaining how to use the software.
-It is not feasible for package documentation to address the underlying metaphysics topics in detail.
-Sometimes frequently asked questions on the underlying theory are answered in the documentation, or pointers to answers are noted.
-
-Technically, setting up parallelization was also noted as a pain point for one of the developers, and the introduction of continuous integration by another.
-Software development knowledge could mitigate some pain points.
+There is no organization helping with the development of good quality software.
+Scientific software is often developed in-house by the very researchers that temporarily use it in their own research.
+This software has a defined user and purpose, and often does not meet the standards that would be required by external users.
+Such software is rarely commercialized, in which case it is not open-source.
+
+Documentation of LBM software is noted as being important, but its quality could be improved.
+There is often no time or funding for maintaining quality documentation for software that is rarely used outside of the development team.
+Furthermore, the documentation generally only provides a shallow overview of the underlying computational fluid dynamics theory.
+Users would be well advised to already be familiar with these topics to or they should spend significant time referencing theory resources.
+The documentation instead generally focuses on explaining how to use the software.
+It is of course not feasible for package documentation to address the underlying metaphysics topics in detail, so it is advised that the package documentation links to resources that better explain the underlying theory.
+Sometimes, frequently asked questions about the underlying theory are answered in the documentation.
+
+Setting up parallelization was also noted as a technical pain point by one of the developers, and the introduction of continuous integration by another.
+Software development knowledge could mitigate such pain points.
 As already noted, many of the developers are domain experts and not professional software developers.
-An interviewed developer with more extensive software development experience noted that eliminating equivalent statements using macros had helped improve the quality of their source code, specifically helping with extending code to run on both the CPU and GPU.
+One of the interviewed developers that has relatively more extensive software development experience noted that eliminating equivalent statements using macros had helped improve the quality of their source code, specifically helping with reusing code to run on both the CPU and GPU.
 
 Difficulties with ensuring correctness were also noted by several developers.
-They indicated that tests are run on new additions,on  both the individual modules and on the entire system to verify continued system correctness.
+They indicated that tests are run on all new source code additions, testing both individual modules and the system to verify correctness.
 These tests compare the package output to known correct output using test cases.
-A developer commented that the amount of testing data that is needed for some cases is a problem, as free services do not offer capabilities to store and frequently process such large amounts of data and in-house testing solutions needed to be created.
+A developer commented that the amount of testing data that is needed for some cases is a problem as free testing services do not offer capabilities to store and process such large amounts of data, and in-house testing solutions needed to be created to address this limitation.
 
 A few obstacles related to users were found.
-Several developers noted that users sometimes try to combine LBM methods with cases incorrectly, and that this is not theoretically sound.
-Some users think that the packages will work out of the box to solve their cases, while the packages require a good understanding of computational fluid dynamics and significant effort in setting up each case.
-These packages are not like more commercial software packages, they are generally set up to solve specific research problems.
+Several developers noted that users sometimes try to use incorrect LBM method combinations to solve their problems.
+Furthermore, some users think that the packages will work out of the box to solve their cases, while the packages require both a good understanding of computational fluid dynamics and an understanding of the requirements for formulating problems in the individual packages, which can be a significant endeavor.
+These software packages are not like commercial software packages.
+They are generally set up to solve specific research problems, and are otfen primarily used by their developers.
 While they are modifiable to solve similar problems, these modifications are not trivial.
-So far this has been addressed by updating the documentation to better inform users of the underlying LBM theory and package requirements.
+So far this problem of on-boarding new users has been addressed by updating the documentation to better inform users of the underlying LBM theory and package requirements.
 Similar issues with LBM parameters were noted by another developer.
-Updating the user interface to better explain explain theoretical principles, as well as test user input, was the implemented solution.
+Updating the user interface to better explain theoretical principles, as well as test user input for compatibility, was the implemented solution.
 As noted above, sometimes frequently asked questions on the underlying theory and on how to use the software are answered in the documentation.
 
-One developer commented that parts of their source code had been refactored to Python partially to address usability.
-Python was perceived as a much more usable language that would be easy for future users and developers to learn and understand.
-According to this developer this was one of the biggest steps in terms of addressing usability.
+One of the interviewed developers commented that parts of their package's source code had been refactored to Python in order to help address usability issues.
+Python was perceived as a much more usable language, and it would be easy for future users and developers to learn and understand the source code.
 
 A few potential future obstacles were noted.
-One developer said that their source code had been written with a specific application in mind and that there was too much coupling between components.
-This would have an impact on future modifiability and reusability when trying to extend the software.
-The effected code would need to be refactored.
-
-As noted above, difficulties with ensuring future correctness could also arise.
-As new methods and functionality is added into the software, new test cases and test data will need to be developed and provided.
+One developer said that their source code had been written with a specific application in mind and that due to this there was too much coupling between components of the source code.
+This would have an impact on future modifiability and reusability when trying to extend the software, and the code would need to be refactored.
+
+As noted above, difficulties with ensuring future correctness could also arise. As new methods and functionality is added into the software, new test cases and test data will need to be developed.
 

### 2021-06-06 StateOfPractice/Peter-Notes/Report/Report.tex
4th edits to Answers to R Questions in LBM Report

@@
 
-The following points should be considered when developing high quality LBM software.
+The following points regarding software quality should be considered when developing LBM software packages.
 
@@
 	\item Include descriptive error messages where appropriate.
-	\item In case automatic installation of dependencies fails, the system should indicate to the user what dependencies are needed and the installation procedure should be able to skip installing the dependencies after they have been manually installed.
-	\item The packages should include detailed tutorials, including dependencies, expected output, and supplementary documentation.
+	\item In case automatic installation of dependencies fails, the system should indicate to the user what dependencies need to be installed manually.
+	\item The packages should include detailed tutorials, including dependencies, expected output, and any additional supplementary documentation that may be required.
 \end{itemize}
@@
 \begin{itemize}
-	\item Include descriptive error messages for cases of unexpected input  including incorrect data types, empty input, and missing files or links.
+	\item The system must provide descriptive error messages when it encounters unexpected input, including incorrect data types, empty input, and missing files or links.
 \end{itemize}
@@
 \begin{itemize}
-	\item Implement using parallelization tools. Consider GPU processing, CUDA, and MPI. 
-	\item The user should be able to process the models on either the CPU or GPU.
+	\item Integrate parallelization tools and techniques in order to reduce processing time. Consider GPU processing, CUDA, and MPI. 
+	\item The user should be able to choose to process their model on either the CPU or GPU.
 \end{itemize}
@@
 \begin{itemize}
-	\item Include user hardware and software requirements.
-	\item Include a user tutorial with expected output.
-	\item Include a detailed user manual. Identify elements of user interfaces.
-	\item Explicitly state the boundaries of the software, and intended applications.
-	\item Identify appropriate fluid dynamics problems that can be solved by the software.
-	\item Provide background domain information or a reference to it.
-	\item Explicitly state the requirements and commitments required of the user to properly model their problem.
-	\item Document expected user characteristics.
-	\item Keep documentation in one location.
+	\item Include user hardware and software requirements documentation.
+	\item Include a user tutorial that indicates the expected output.
+	\item Include a detailed user manual. It should identify elements of user interfaces, and identify all requirements to model a system.	
+	\item State appropriate fluid dynamics problems that the software is designed to model in the documentation, and explicitly state the limits of the software. 
+	\item Provide documentation that details the background theory information, or provide a reference to such information.
+	\item Identify expected user characteristics.
+	\item Keep all documentation in one location.
+	\item Do not duplicate artifact information.
 	\item Maintain a user support model (Git, email, forum, FAQ)
-	\item If possible, consider using user-friendly scripting languages like Python. Especially in parts of the software that are likely to be modified or reviewed by users.
+	\item If possible, consider using popular user-friendly software languages like Python. Especially consider this for parts of the source code that is likely to be modified or reviewed by users.
 \end{itemize}
@@
 \begin{itemize}
-	\item Include high quality artifacts. Ideally most of the common and less common artifacts listed in Section \ref{artifacts}.
-	\item Include version numbers and release notes for all major releases. 
+	\item Keep artifacts updated.
+	\item Ideally include most of the common and less common artifacts listed in Section \ref{artifacts}.
+	\item Include version numbers and release notes for all major source code and artifact releases. 
 	\item Have a defined process for accepting contributions.
 	\item Include documentation for making contributions to the project. 
-	\item Use an issue tracker (Git, email, SourceForge, other) to manage bugs and changes. Issues should be closed regularly. 
+	\item Use an issue tracker (Git, email, SourceForge, other) to manage bugs and changes. Issues should be regularly reviewed and closed. 
 	\item Use a version control system (GitHib, CVS).
-	\item Source code needs to be well commented.
-	\item Ensure source code is modular.
+	\item Source code needs to be well commented. Typically, more than ten percent of LBM package source code is comments. 
+	\item Ensure source code is modular. Generally, this means that source code is split into files along functionality.
 	\item Eliminate code duplication.
-	\item If possible, develop using languages that are easy to read and quick to learn.
+	\item If possible, consider using popular user-friendly software languages like Python. Especially consider this for parts of the source code that is likely to be modified or reviewed by users.
 \end{itemize}
@@
 	\item Ensure a high degree of code modularity and abstraction.
-	\item If possible, develop using languages that are easy to read and quick to learn.
-	\item Consider future modifiabilility in the design stage prior to any development.
+	\item If possible, consider using popular user-friendly software languages like Python. Especially consider this for parts of the source code that is likely to be modified or reviewed by users.
+	\item Consider future source code modifiabilility as early as the design stage of development.
 	\item Consider flexibility of data structures and data storage in the design stage.
-	\item Make public validation benchmarks.
+	\item Provide validation benchmark tools and data to the public.
 \end{itemize}
@@
 \begin{itemize}
-	\item Modularize the source code.
-	\item Document module interfaces.
+	\item Ensure a high degree of code modularity and abstraction.
+	\item Document module interfaces in design and developer documentation.
 	\item Provide API documentation, if applicable.
@@
 	\item Maintain consistent formatting style across source code files.
-	\item Identify a coding standard, if one is used during development.
-	\item Use consistent, distinctive, and meaningful code identifiers. Identify algorithms that are used.
+	\item Identify a coding standard in the documentation, if a coding standard is used during development.
+	\item Use consistent, distinctive, and meaningful code identifiers. 
+	\item Identify algorithms that are used.
 	\item Hard code domain constants.
 	\item Add meaningful comments. They should indicate what is being done in each section of source code.
-	\item Modularize the source code.
-	\item Include user guides and usage requirements.
-	\item In the documentation identify the limits of the software. 
-	\item Identify the cases that the software can be applied to.
-	\item Note the requirements of the user in order to set up their case. Identify what parameters are needed. 
+	\item Ensure a high degree of code modularity and abstraction.
+	\item Include user guides, and state technical and problem modeling requirements.
+	\item State problem cases that the software is designed to model in the documentation, and explicitly state the limits of the software. 
 	\item Consider adding a FAQ section to the documentation. 
@@
 	\item Use automatic document generation tools to limit the time spent on updating documentation.
-	\item Include documentation aimed at developers, like a developers guide.
-	\item Provide documentation that orients users and developers to artifacts, the source code, and system architecture.
-	\item Highlight new system features between artifact versions. 
+	\item Provide a developers guide.
+	\item Provide a resource that helps orient users and developers to the artifacts, source code, and system architecture.
+	\item Highlight new features and changes between artifact versions. 
 \end{itemize}
@@
 	\item Identify the development environment.
-	\item Identify the development process and how to contribute. 
+	\item Identify the development process and how new users can contribute. 
 	\item Include notes with all releases.
-	\item Write modular code.
-	\item Communicate all contributions and changes to the entire development team.
+	\item Ensure a high degree of code modularity and abstraction.
+	\item Communicate all changes to the development team.
 	\item Use continuous integration processes and tools.
-	\item Consider peer review processes for contributions. 
-	\item Decrease code duplication.
-	\item Maintain updated developer documentation.
-	\item Make use of project management software.
+	\item Consider peer review processes to assess contributions. 
+	\item Eliminate code duplication.
+	\item Update all relevant developer documentation when a change to the software is made.
+	\item Use project management software.
 \end{itemize}
@@
 \begin{itemize}
-	\item Compare results of methods against manually calculated or known correct results.
-	\item Automate the above comparison. Run the comparison after code changes.
-	\item Consider peer reviews for code contributions.
-	\item Maintain detailed and up to date design documentation.
+	\item Test output against manually calculated or known correct results.
+	\item Automate the testing of output. Run tests after all code changes.
+	\item Consider peer review processes to assess contributions. 
+	\item Update all relevant design documentation when a change to the software is made.
 \end{itemize}
@@
 \begin{itemize}
-	\item Documentation must include all technical requirements and dependencies for using the software. 
-	\item Documentation should detail how to properly use the software.
-	\item Documentation should outline correct and incorrect use cases of the software.
-	\item Documentation should either explain underlying CFD theories or reference appropriate resources.
-	\item Consider asking users for feedback on the documentation.
+	\item Documentation must state all technical requirements and software dependencies. 
+	\item A user guide must be provided.
+	\item State appropriate fluid dynamics problems that the software is designed to model in the documentation, and explicitly state the limits of the software. 
+	\item The documentation should either explain the underlying CFD theories or provide a reference to appropriate resources.
+	\item Consider asking users for feedback on the documentation, specifically if it clearly conveys all necessary information to the users.
 \end{itemize}

### 2021-06-07 StateOfPractice/AHP2021/MedicalImaging/Source code/DataProcess/src/Main.java
updated ahp scores

@@ -27,7 +27,7 @@ public class Main {
 
     public static void main(String[] args) {
         // To use the original assigned scores, change this variable to false;
-        boolean SHOULD_APPLY_SENSITIVITY_ANALYSIS = true;
+        boolean SHOULD_APPLY_SENSITIVITY_ANALYSIS = false;
 
         try {
             FileInputStream file = new FileInputStream(new File("AHP_Template.xlsx"));

### 2021-06-09 StateOfPractice/AHP2021/MedicalImaging/Source code/DataProcess/src/Main.java
updated AHP

@@ -27,7 +27,7 @@ public class Main {
 
     public static void main(String[] args) {
         // To use the original assigned scores, change this variable to false;
-        boolean SHOULD_APPLY_SENSITIVITY_ANALYSIS = false;
+        boolean SHOULD_APPLY_SENSITIVITY_ANALYSIS = true;
 
         try {
             FileInputStream file = new FileInputStream(new File("AHP_Template.xlsx"));
@@ -145,8 +145,8 @@ public class Main {
 
     private static void applySensitivityAnalysis(XSSFWorkbook workbook) {
         // Set the lower and upper bound of the random numbers according to needs
-        double LOWER_BOUND = -1;
-        double UPPER_BOUND = 1;
+        double LOWER_BOUND = -0.5;
+        double UPPER_BOUND = 0.5;
 
         // Adjust scores with random numbers
         for (int i = 0; i < numberQualities; i++) {

### 2021-06-10 StateOfPractice/Peter-Notes/Report/Report.tex
applied first set of feedback, up to but not including domain analysis section

@@
 
-\usepackage[round]{natbib}
+\usepackage[]{natbib}
 
@@
 \usepackage{setspace}
+\usepackage{thesis}
+
 \doublespacing
-\title{State of the Practice of Lattice Boltzmann Method Software Development} 
-\author{Peter Michalski}
-\date{\today}
+
+\newcommand{\fulltitle}{State of the Practice for Lattice Boltzmann Method Software}
+\newcommand{\authorname}{Peter Michalski}
 
 \begin{document}
-	\maketitle
 \begin{singlespace}
-	\begin{abstract}
-		This project analyzed the state of the practice for software development in the Lattice Boltzmann Method 
-		domain by quantitatively and qualitatively measuring 23 software packages along quality attributes...
-	\end{abstract}
+
+\thesistitle
+	{\MakeUppercase{\fulltitle}}
+	{\MakeUppercase{\authorname}.}
+	{Master of Engineering in Computing and Software}
+	{\copyright \ Copyright by \authorname, July 2021}
 
 \newpage
-
+\thispagestyle{empty}
+\begin{center}
+	\textbf{\large Abstract}
+\end{center}
+
+This project analyzes the state of the practice for software development in the Lattice Boltzmann Method domain by quantitatively and qualitatively measuring 23 software packages along quality attributes...
+
+\newpage
+\pagenumbering{roman}
 \tableofcontents
-\addtocontents{toc}{\protect\thispagestyle{empty}}
+\addtocontents{toc}{}
 
@@
 ~\newpage
-\pagenumbering{gobble}
-\pagenumbering{arabic}
 
 \clearpage
-\section{Reference Material}
-
-\subsection{Software Engineering Related Definitions and Acronyms}
+\section*{Reference Material}
+
+\subsection*{Software Engineering Related Definitions and Acronyms}
 
@@
 
-\noindent\textbf{Goal}: “Goals capture, at different levels of abstraction, the various objectives the system under consideration should achieve.” \citep{van2001goal}\\
+\noindent\textbf{Goal}: Goals capture, at different levels of abstraction, the various objectives the system under consideration should achieve \citep{van2001goal}.\\
 
@@
 
-\noindent\textbf{Program Family}: A set of programs that are analyzed and designed together starting from the initial stages of the software development life-cycle.\\
-
-\noindent\textbf{Requirements}: A software requirement is: \textit{i}) a condition or capability needed by a user to solve a problem or achieve an objective; \textit{ii}) a condition or capability that must be met or possessed by a system or system component to satisfy a contract, standard, specification, or other formally imposed document; or, \textit{iii}) a documented representation of a condition or capability as in the above two definitions.
-\citep{thayer2000ieee}\\
+\noindent\textbf{Requirements}: A software requirement is: \textit{i}) a condition or capability needed by a user to solve a problem or achieve an objective; \textit{ii}) a condition or capability that must be met or possessed by a system or system component to satisfy a contract, standard, specification, or other formally imposed document; or, \textit{iii}) a documented representation of a condition or capability as in the above two definitions \citep{thayer2000ieee}.\\
 
@@
 
+\noindent\textbf{Software Family}: A set of programs with an extensive amount of common properties \cite{parnas1976design}.\\
+
 \noindent\textbf{Variability}: A requirement or goal that varies between family members.
@@
 \newpage
-\subsection{Lattice Boltzmann Related Definitions and Acronyms}
+\subsection*{Lattice Boltzmann Related Definitions and Acronyms}
 
@@
 
-\noindent\textbf{BGK}: Bhatnagar-Gross-Krook\\
+\noindent\textbf{BGK}: Bhatnagar-Gross-Krook \citep{bhatnagar1954model}\\
 
@@
 \newpage
+\pagenumbering{arabic}
 \section{Introduction - DONE SECTION 1st DRAFT}
@@
 This report analyzes the development of software packages that use Lattice Boltzmann Methods (LBM), a class of computational fluid dynamics methods, for fluid simulation.
-The analysis involves taking qualitative and qualitative measures along quality attributes.
-The goal is to analyze the current state of development of Scientific Computing Software (SCS) in this domain to provide insight into its best practices and offer guidance for future development.
+These LBM algorithms consider the behaviours of a collection of particles as a single unit of an intermediate size, and predict their positional probability as they move through a lattice structure.
+LBM are further detailed in Section \ref{domainanalysis}.
+
+Scientific Computing Software (SCS) is defined as the use of computer tools to analyze or simulate mathematical models of real world systems \citep{smith2006systematic}.
+Given the importance of such software, scientists and engineers desire methods and tools to sustainably develop it with high quality.
+This has led to the formation of groups like the Software Sustainability Institute (\href{https://www.software.ac.uk/}{SEI}) and Better Scientific Software (\href{https://bssw.io/}{BSS}).
+
+The goal of this report is to analyze the current state of development of LBM SCS to provide insight into its best practices and to offer guidance for future development.
+We want to understand how software quality in the LBM SCS family is impacted by software development choices, including principles, processes, and tools.
 
@@
 
-\subsection{Purpose of Report}
-
-This document constitutes the author's M.Eng.
-report in partial fulfillment of the the degree of Master of Engineering.
-It is to be submitted to the Department of Computing and Software in McMaster University's Faculty of Engineer and is intended for review by the examination committee.
-In this context the purposes of the report are the following:
+\subsection{Purpose}
+
+The LBM software packages are assessed for a given set of quality attributes, using quantitative and qualitative data, in order to answer the research questions found in Section \ref{researchquestions}.
+The qualities are defined in Section \ref{softwarequalities}.
+Qualitative data is gathered from interviews with software package developers.
+The interview questions are listed in Section \ref{interviewquestions}.
+Quantitative data is measured using the quality metrics listed in Section \ref{measurementtemplate2}.
+In this context the purposes of the report are as follows:
 
 \begin{itemize}
-	\item Describe the state of the practice of SCS project origin, motivations, and goals
-	\item Describe LBM software structure and goals
+	\item Develop and test an updated methodology for assessing the state of the practice of SCS projects
+	\item Describe the specific example of LBM software structure and goals
 	\item Report on the measure of a subset of LBM software packages along quality metrics
-	\item Evaluate the state of the practice of LBM software development along quality metrics
-	\item Make suggestions for improving LBM software along quality metrics
+	\item Evaluate the state of the practice of LBM software development along quality attributes
+	\item Make suggestions for improving LBM software along quality attributes
 	\item Make suggestions for improving state of the practice of SCS assessments
@@
 
-\subsection{Motivation of Report}
+\subsection{Motivation}
 
@@
 In the course of this assessment we updated the methodology that was used during previous assessments.
-Details of the new methodology are found \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Methodology.pdf}{here}.
+Details of the new methodology are in \citep{methodology2020}.
 The previous set of research questions were critically assessed and modified.
@@
 
-\subsection{Scope of Report}
+\subsection{Scope}
 
 This report analyzes a filtered set of LBM software packages along quantitative and qualitative measures.
-Many of these measures are captured using surface measurements, which can be categorized as initial and easy to capture measurements, of the underlying quality and they may not represent the true quality of the software.
+Many of these measures are captured using surface measurements, which can be categorized as initial and easy to capture measurements, of the underlying quality; they may not represent the true quality of the software.
 Surface measurements are taken as they allow us to apply the same measurement, with reasonable effort, along all software packages despite technical and functional variabilities among the set of software packages.
-The measures are used to answer the research questions found in Section \ref{researchquestions}.
 The measured software packages are open source, and it is recommended to keep this in mind when considering the recommendations of this report.
@@
 
-\subsection{Organization of Report}
-
-The first part of this document includes the Table of Contents, List of Figures, and List of Tables.
-This is followed by Reference Material including subsections outlining Software Engineering Related Definitions and Acronyms and Lattice Boltzmann Related Definitions and Acronyms.
-
-The rest of core of the report is organized as follows:
+\subsection{Organization}
+
+The report is organized as follows:
 
@@
 	\item \textbf{Introduction} to the report, including its purpose, motivation, scope, and organization.
-	\item \textbf{Background} of Scientific Computing Software, including software qualities and software families.
 	\item LBM software \textbf{Domain Analysis}.
+A domain analysis consists of systematically identifying and documenting the commonalities, variabilities, and terminology of a software family \citep{Weiss1997}.
 	\item \textbf{Methodology} of this state of the practice assessment, including the steps of the methodology, software quality definitions and how these qualities are assessed in this report, an overview of candidate software selection, empirical measures gathered, and the analytical hierarchy process.
+This follows a general methodology for assessing the state of the practice for SCS domains.
 	\item Quantitative and qualitative \textbf{Results}. 
@@
 \newpage
-\section{Background - DONE SECTION 1st Draft}
-
-SCS is defined as the use of computer tools to analyze or simulate mathematical models of real world systems \citep{smith2006systematic}.
-Given the importance of such software, scientists and engineers desire methods and tools to sustainably develop it with high quality.
-This has led to the formation of groups like the Software Sustainability Institute (\href{https://www.software.ac.uk/}{SEI}) and Better Scientific Software (\href{https://bssw.io/}{BSS}).
-The purpose of this state of the practice assessment is to understand how software quality in the LBM SCS family is impacted by software development choices, including principles, processes, and tools.
-A software family is defined
-by \cite{parnas1976design} as “a set of programs whose common properties are so extensive that it is advantageous to study the common properties of the programs before analyzing individual members”.
+
+\section{Domain Analysis - DONE SECTION 1st DRAFT}\label{domainanalysis}
+
+A software family is defined by \cite{parnas1976design} as “a set of programs whose common properties are so extensive that it is advantageous to study the common properties of the programs before analyzing individual members”.
 The common properties of the LBM family of related programs are listed as part of the domain analysis in Section \ref{domainanalysis}.
-The software qualities that are assessed in this report are listed in Section \ref{softwarequalities}.
-Ensuring software quality can take significant time, but as \cite{SmithEtAl2016} shows, SCS developers often cannot see themselves meeting the required time constraints to better develop some of the artifacts that improve qualities, like detailed documentation.
-An inability to develop such documentation due to time constraints could be mitigated by the use of automatic document generation tools like Drasil.
-The Drasil Framework consists of a collection of Domain Specific Languages (DSL) for capturing scientific documents, structures, and computing
-knowledge, and then transforming this knowledge into relevant software artifacts without having to manually duplicate knowledge into multiple artifacts \citep{zhao2018}.
-
-\newpage
-
-\section{Domain Analysis - DONE SECTION 1st DRAFT}\label{domainanalysis}
 
@@
 Some of this was empirical software engineering related data, such as the number of files, number of lines of code, percentage of issues that are closed.
-Most of the data was gathered by manually observing the software, its source code, and its artifacts, while some was gathered using the tools listed in Section \ref{tools}.
+Most of the data was gathered by manually observing the software, its source code, and its artifacts, while some was gathered using the tools discussed in Section \ref{empiricalmeasures}.
 
@@
 This part of the assessment focuses on data that is reasonably easy to collect.
-Much of the data was gathered by manually reviewing the artifacts of each software package, while some was gathered using freeware tools listed in Section \ref{tools}.
+Much of the data was gathered by manually reviewing the artifacts of each software package, while some was gathered using the freeware tools discussed below.
 The data that was gathered is listed in the measurement template found in Section \ref{measurementtemplate2}.
@@
 The complete measurement template data was then analyzed using the grading template found in Section \ref{gradingtemplate2}, the output of which was analyzed using the AHP described in Section \ref{AHP}.
-
-\subsubsection{Software Tools}\label{tools}
 
@@
 			\item API Documentation
-			\item Developer/Contributing Manual/Guide
+			\item Developer/Contributor Manual/Guide
 			\item FAQ / Forum
@@
 - comment on future work in state of the practice SCS
+
+the following was moved from the former background section, may need edits as part of conclusion:
+
+  The software qualities that are assessed in this report are listed in Section \ref{softwarequalities}.
+Ensuring software quality can take significant time, but as \cite{SmithEtAl2016} shows, SCS developers often cannot see themselves meeting the required time constraints to better develop some of the artifacts that improve qualities, like detailed documentation.
+An inability to develop such documentation due to time constraints could be mitigated by the use of automatic document generation tools like Drasil.
+The Drasil Framework consists of a collection of Domain Specific Languages (DSL) for capturing scientific documents, structures, and computing
+knowledge, and then transforming this knowledge into relevant software artifacts without having to manually duplicate knowledge into multiple artifacts \citep{zhao2018}.
 

### 2021-06-13 StateOfPractice/Ao-Report/13Methods.tex
Ao's report - updated ch 3

--- 2021-05-24 StateOfPractice/Ao-Report/13Methods.tex
@@
 
-\ad{Where to include the domain experts? I plan to include them in  Domain Selection and Software Product Selection, and also mention that they'll participate in interviews.}
+We designed a general process for evaluating the state of the practice of domain-specific software, that we instantiate to SC software for specific scientific domains.
 
-In this project, we aim to study and research the SC software within scientific domains.
-However, we try to design the whole process of this project in a more general way, with the hope that it should not be limited to only one or several specific software categories.
+Our method involves: 1) choosing a software domain (Section \ref{sec_domain_selection}); 2) collecting and filtering software packages (Section \ref{sec_software_selection}); 3) grading the selected software (Section \ref{sec_grading_software}); 4) interviewing development teams for further information (Section \ref{sec_grading_software}).
 
-The way how we choose a domain is listed in Section \ref{sec_domain_selection}.
-Within a specific domain, we use the process documented in Section \ref{sec_software_selection} to collect and filter the software packages for our study.
-Then, all the software is measured by using the grading template in Section \ref{sec_grading_template} and the empirical measurement method in Section \label{sec_empirical_measurements}.
-Section \ref{sec_measuring_qualities} presents more details about how we apply the above process.
+Details of how we applied the method on the MI domain are expanded upon in Section \ref{sec_applying_method}.
 
@@
 \label{sec_domain_selection}
-Although the methods of this project may not have such limitations, we limit our selection within scientific domains to fulfill the objective of our research.
+Our methods are generic, but our scope only included scientific domains due to the objective of our research, and thus we have only applied this method to scientific domains.
 
-One major factor in properly choosing a candidate domain for the study is the ease to select software packages within it.
-As described in Section \ref{sec_software_selection}, the software product selection process is most likely to have a screening step, and the quantity of final-decision packages may not meet our initial expectation if we do not have enough software candidates to choose from.
-Section \ref{sec_software_selection} also explains why our process prioritize the OSS.
-Consequently, a scientific domain with a large number of active OSS is proffered.
-This also often indicates that the domain has an active community developing and using SC software, making it easier to conduct interviews mentioned in Section \ref{sec_interview_methods}.
+When choosing a candidate domain, we prefer a domain with a large number of active OSS.
+This is because we aim to finalize a list of 30 software packages \cite{SmithEtAl2021} in the domain after screening out the ones not meeting our requirements, such as the ones without recent updates or certain functions.
+Thus the quantity of final-decision packages may not meet our initial expectation if we do not have enough software candidates to choose from.
+Another reason is that our grading method works much better for OSS.
+Moreover, we need the domain to have an active community developing and using SC software, making it easier to conduct interviews with developers.
 
-Even if we can find an adequate number of software packages in a domain, we may still find the software products are developed to solve various problems within the domain and should be categorized into different sub-groups.
-So one question needs to be asked - do we prefer a group of software all providing similar functions and features, or do we aim to cross-compare several sub-sections within the same domain?
-With that answered, it should be easier to determine a favored domain.
+With an adequate number of software packages in a selected domain, we may still find the software products with various purposes and fall into different sub-categories.
+So we should ask one question to ourselves - do we prefer a group of software all providing similar functions and features, or do we aim to cross-compare several sub-sections within the same domain?
+With that answered, we can better determine a favored domain.
 
 Another aspect to consider is the team carrying out the research, more specifically, the domain experts - if there is any - in the team.
-In our team, the projects are often led by researchers in the software engineering field and supported by experts working in other scientific domains.
+In our team, researchers in the software engineering field usually lead the projects, and experts working in scientific domains support them.
 Having domain experts in the team provides significant benefits in selecting software packages and designing interview questions.
-
-In this project, Medical Imaging (MI) domain is selected.
-Numerous software products can be found in this domain, and a great number of domain experts use - and even develop - such software.
-We decide to focus on MI software with the viewing function, and more details about this filtering are in Section \ref{sec_software_selection}.
-Being able to include MI domain experts in our team also makes this domain more preferred.
 
@@
 \subsection{Identify Software Candidates}
-The candidate software can be found from publications in the domain.
+\label{sec_identify_software_candidates}
+We can find candidate software in publications of the domain.
 Another source is to search various websites, such as \hyperlink{https://github.com/}{GitHub}, \hyperlink{https://swmath.org/}{swMATH} and the Google search results for software recommendation articles.
@@
 
-As for this project, 48 MI software projects are identified as the candidates, and they are found from publications \cite{Bjorn2017} \cite{Bruhschwein2019} \cite{Haak2015}, online articles related to the domain \cite{Emms2019} \cite{Hasan2020} \cite{Mu2019}, forum discussions related to the domain \cite{Samala2014}, etc.
-
 \subsection{Filter the Software List}
+\label{sec_filter_software_list}
 The goal is to build a software list with a length of about 30 \cite{SmithEtAl2021}.
 
-The only mandatory requirement is that the software must be open source, as defined in Section \ref{sec_open_source_software}.
+The only ``mandatory'' requirement is that the software must be OSS, as defined in Section \ref{sec_open_source_software}.
 This is due to the grading process defined int Section \ref{sec_grading_template}.
-In order to evaluate all aspects of some software qualities, the source code must be accessible.
+To evaluate all aspects of some software qualities, the source code should be accessible.
 
-The other factors to filter the list can be optional and should be considered according to the number of software candidates and the objectives of a research project.
+The other factors to filter the list can be optional and we should consider them according to the number of software candidates and the objectives of the research project.
 
@@
 For example, we can choose a group of software with similar functions, so that the cross-comparison is between each individual of them.
-On the other hand, if our objective is comparing sub-categories in the domain, we should select from candidates in each of the categories.
+On the other hand, if our objective is to compare sub-categories in the domain, we should select candidates from each of the categories.
 
-The empirical measurement tools listed in Section \ref{sec_empirical_measurements} are limited to projects using \hyperlink{https://git-scm.com/}{Git} as the version control tool, so software with Git is preferred.
-Some manual steps in empirical measurement depend on a few metrics of GitHub, which makes projects hold on GitHub more favored \cite{SmithEtAl2021}.
+The empirical measurement tools listed in Section \ref{sec_empirical_measurements} have limitations that we can only apply them on projects using \hyperlink{https://git-scm.com/}{Git} as the version control tool, so we prefer software with Git.
+Some manual steps in empirical measurement depend on a few metrics of GitHub, which makes projects held on GitHub more favored \cite{SmithEtAl2021}.
 
 Some of the OSS projects may experience a lack of recent maintenance.
-So packages that have not been updated for a long time can be eliminated, unless they are still popular and highly recommended by the users in the domain \cite{SmithEtAl2021}.
+So we can eliminate packages that have not been updated for a long time and unless they are still popular and highly recommended by the users in the domain \cite{SmithEtAl2021}.
 
-In this project, we focus on the MI software providing imaging viewing function. \ad{TBC}
+No matter we set what standards to filter the packages, with domain experts in the team, we should value their opinions.
+For example, if a software package is not OSS and has not been updated for a long while, the domain experts may still identify it as a popular and widely used product.
+In this case, perhaps it is valuable to keep this software on the list.
 
-\section{Grading Template}
+\section{Grading Software}
+\label{sec_grading_software}
+
+We grade the selected software using a template (Section \ref{sec_grading_template}) and a specific empirical method (Section \ref{sec_empirical_measurements}).
+Some technical details for the measurements are in Section \ref{sec_technical_details}.
+
+\subsection{Grading Template}
 \label{sec_grading_template}
+The full grading template can be found in Appendix \ref{ap_grading_template}. The template contains 101 questions that we use for grading software products.
 
-\ad{The grading template for manual measurements}
+We use the first section of the template to collect some general information about the software, such as the name, purpose, platform, programming language, publications about the software, the first release and the most recent change date, website, and source code repository of the product, etc.
+Information in this section helps us to understand the projects better and may be useful for further analysis, but it does not directly affect the grading scores for the packages.
 
-\section{Empirical Measurements}
+We designed the next 9 sections in the template for the 9 software qualities mentioned in Section \ref{sec_software_quality}.
+For each quality, we ask several questions and the typical answers are among the choices of ``yes'', ``no'', ``n/a'', ``unclear'', a number, a string, a date, a set of strings, etc.
+The last question of each quality is asking for an overall score between 1 and 10 based on all the previous questions of this section, which is also the grading score for this quality.
+For some qualities, the empirical measurement sections on the template also affect this grading score.
+
+All the last 3 sections on the template are about the empirical measurements.
+We use two command-line software tools git\_stats and scc to extract information about the source code from the project repositories.
+For projects held on GitHub, we manually collect additional metrics, such as the stars of the GitHub repository, and the numbers of open and closed pull requests.
+Section \ref{sec_empirical_measurements} presents more details of how these empirical measurements affect software quality grading scores.
+
+\subsection{Empirical Measurements}
 \label{sec_empirical_measurements}
 
-\ad{Explain the empirical tools and how they affect the scores.}
+We use two command-line tools for the empirical measurements.
+One is GitStats that generates statistics for git repositories and display outputs in the format of web pages \cite{Gieniusz2019}; the other one is Sloc Cloc and Code (as known as scc) \cite{Boyter2021}, aiming to count the lines of code, comments, etc.
 
-\section{Measuring Qualities}
-\label{sec_measuring_qualities}
+Both tools can measure the number of text-based files in a git repository, as well as lines of text in these files.
+Based on our experience, most text-based files in a software project repository contain programming source code and developers use them to compile and build software products, and a minority of these files are instructions and other documents.
+So we roughly regard the lines of text in text-based files as lines of programming code.
+The two tools usually generate similar but not identical results as for the above measurements.
+From our understanding, this minor difference is because of the different techniques that they use to detect if a file is a text-based or binary file.
 
-\ad{e.g. virtual machines, time spent per software, where to look for docs, etc.}
+Additionally, we also manually collect some information for projects held on GitHub, such as the numbers of stars, forks, people watching this repository, open pull request, and closed pull request.
+
+These empirical measurements help us from two aspects.
+Firstly, with more statistical details, we can get an overview of a project faster and more accurately.
+For example, the number of commits over the last 12 months shows how active this project has been during this period, and the number of stars and forks may reveal its popularity.
+Secondly, the results may affect our decisions regarding the grading scores for some software qualities.
+For example, if the percentage of comment lines is low, we might want to double-check the understandability of the code, and if the ratio of open versus closed pull requests is high, perhaps we should pay more attention to the maintainability of the project.
+
+\subsection{Technical Details}
+\label{sec_technical_details}
+To test the software on a ``clean'' system, we created a new virtual machine (VM) for each software and only installed the necessary dependencies before measuring.
+All 29 VM were created on the same computer.
+We only start the measuring of the next software after finishing a current one, and after grading each software, the VM is destroyed.
+
+Generally speaking, we spend about two hours grading one software, unless there are technical issues and more time is needed to resolve them.
+In most of the situation, we finish all the measurements for one software on the same day.
 
@@
 \subsection{Interviewee Selection}
+For a software list with a length of roughly 30, we aim to interview about 10 development teams of these projects.
+Interviewing multiple individuals from each team should give us more comprehensive information about a project, but if there are difficulties in finding multiple willing participants, a single engineer well knowing the project can also be sufficient.
 
-\ad{Generally speaking, ask all the teams which we can find contacts, and continue with the ones who are willing to participate.}
+Ideally, we select projects after the grading measurements and prefer the projects with higher overall scores.
+However, if we do not find enough participants, we should also contact all teams on the list.
+
+We try to find the contacts of the teams on the websites related to the software, such as the official web pages, repository websites, publication websites, and bio pages of the teams' institutions.
+For each candidate, we send at most two emails asking for their support and participation before receiving any replies.
 
@@
 
-\ad{The aspects we focused on.}
+We have a list of 20 questions to guide our interviews with the development teams, which can be found in Appendix \ref{ap_interview}.
 
-\subsection{Interview Methods}
+Some of the questions are about the background of the software, the development teams, the interviewees, and the way they organize the projects.
+We also ask about their understandings of the users.
+Another part of the interview questions focuses on the major difficulties the team experience currently or in the past, and the solutions that they have found or will try in the future.
+We also discuss with interviewees the importance of documents to their projects and the current situations of these documents.
+One more proportion of the questions are about several specific software qualities, such as maintainability, understandability, usability, and reproducibility.
 
-\ad{The ethics approvals, the way to ask questions, the way to transcript answers, and the technologies used.}
+The interviews are supposed to be semi-structured based on the question list and we also ask follow-up questions when necessary.
+Based on our experience, the interviewees usually bring up some interesting ideas that we do not expect and we can expand on these topics for a few more details.
+
+\subsection{Interview Process}
+Since the members of the development teams are likely to be based around the world, so we organize these interviews as virtual meetings online with \hyperlink{https://zoom.us/}{Zoom}.
+After receiving consent from the interviewees, we also record our discussions to better transcribe them.
+
+\section{An Example of Applying the Method}
+\label{sec_applying_method}
+
+This section shows how we applied our method in Section \ref{ch_methods} on the Medical Imaging (MI) domain.
+
+Based on the principles in Section \ref{sec_domain_selection}, we selected the MI domain, because there are numerous software products and a great number of professionals use and/or develop such software in this domain.
+We decided to focus on MI software with the viewing function, so we needed a large number of software candidates in the domain.
+Being able to include MI domain experts in our team also made this domain more preferred.
+
+By using the method in Section \ref{sec_identify_software_candidates}, we identified 48 MI software projects as the candidates from publications \cite{Bjorn2017} \cite{Bruhschwein2019} \cite{Haak2015}, online articles related to the domain \cite{Emms2019} \cite{Hasan2020} \cite{Mu2019}, forum discussions related to the domain \cite{Samala2014}, etc.
+
+Among the 48 software packages, there were several ones that we could not find their source code, such as MicroDicom, Aliza, and jivex, etc.
+These packages are likely to be freeware defined in Section \ref{sec_freeware} and not OSS.
+So following guidelines in Section \ref{sec_filter_software_list} we removed them from the list.
+We focused on the MI software that could work as a MI viewer.
+Some of the software on the list were tool kits or libraries for other software to use as dependencies but not for end-users to view medical images, such as VTK, ITK, and dcm4che, etc.
+We also eliminated these from the list.
+After that, there were 29 software products left on the list.
+We still preferred projects using git and GitHub and being updated recently, but did not apply another filtering since the number of packages was already below 30.
+However, 27 out of the 29 software packages on the final list used git, and 24 of which were held on GitHub.
+Furthermore, 27 packages had the latest updates after the year 2018, and 23 after 2020.
+
+Then we followed the steps in Section \ref{sec_grading_software} to measure and grade the software.
+27 out of the 29 packages can be installed on two or three different operating systems such as Windows, macOS, and Linux, and 5 of them are browser-based, making them platform-independent.
+However, in the interest of time, we only performed the measurements for each project by installing it on one of the platforms, most likely Windows.
+
+Going through the interview process in Section \ref{sec_interview_methods}, we started with the teams with higher scores on our list, and eventually contacted all of them.
+There are developers/architects from 8 teams having participated in our interviews so far.
+Before contacting any interviewee candidate, we received ethics clearance from the McMaster University Research Ethics Board (Appendix \ref{ap_ethics}).

### 2021-06-13 StateOfPractice/Ao-Report/13Methods.tex
Ao's report - updated ch 3

@@
 
-Our method involves: 1) choosing a software domain (Section \ref{sec_domain_selection}); 2) collecting and filtering software packages (Section \ref{sec_software_selection}); 3) grading the selected software (Section \ref{sec_grading_software}); 4) interviewing development teams for further information (Section \ref{sec_grading_software}).
+Our method involves: 1) choosing a software domain (Section \ref{sec_domain_selection}); 2) collecting and filtering software packages (Section \ref{sec_software_selection}); 3) grading the selected software (Section \ref{sec_grading_software}); 4) interviewing development teams for further information (Section \ref{sec_interview_methods}).
 
@@
 The only ``mandatory'' requirement is that the software must be OSS, as defined in Section \ref{sec_open_source_software}.
-This is due to the grading process defined int Section \ref{sec_grading_template}.
-To evaluate all aspects of some software qualities, the source code should be accessible.
+The reason is that to evaluate all aspects of some software qualities, the source code should be accessible.
 

### 2021-06-15 StateOfPractice/Peter-Notes/Report/Report.tex
applied first set of feedback, up to and including domain analysis section

@@
 
+A domain analysis consists of systematically identifying and documenting the commonalities, variabilities, and parameters of variation of a software family \citep{Weiss1997}.
 A software family is defined by \cite{parnas1976design} as “a set of programs whose common properties are so extensive that it is advantageous to study the common properties of the programs before analyzing individual members”.
-The common properties of the LBM family of related programs are listed as part of the domain analysis in Section \ref{domainanalysis}.
+This domain analysis of the LBM SCS family is organized into the four subsection located below.
+The first subsection reviews the basics of Lattice Boltzmann systems.
+The next three subsections consist of lists of commonalities, variabilities, and parameters of variation, respectively.
+These three sections form the heart of the domain analysis and include an extensive set of cross-references to demonstrate the relationships between the different items.
 
@@
 LBM are a family of fluid dynamics algorithms for simulating single-phase and multiphase fluid flows, often incorporating additional physical complexities \citep{chen1998lattice}.
-They consider the behaviours of a collection of particles as a single unit at the mesoscopic scale.
+They consider the behaviours of a collection of particles as a single unit at the mesoscopic scale, between the nanoscopic and microscopic scales.
 These methods predict the positional probability of a collection of particles moving through a lattice structure.
@@
 
-As LBS model fluid dynamics within a boundary using a predefined lattice structure, the methods rely on a two step calculation process.
+LBS model fluid dynamics within a boundary using a predefined lattice structure and a two step calculation process.
 The first process is streaming, where the particles move along the lattice via links.
@@
 
-\begin{figure}[h!]
-	\begin{center}
-		\includegraphics[width=0.5\textwidth]{FeatureDiagramLBS}
-		\caption{LBS Model Feature Diagram}
-		\label{LBSFeatureDiagram}
-	\end{center}
-\end{figure}
-
 \subsection{Commonalities}\label{comm_sec}
@@
 Although the output information could simply be written to the computer’s memory, in all practical applications it is desirable to have a persistent record of the output that was created.
-The final section covers qualities of the system that cannot be classified as input, simulation generation or output.
-These commonalities are termed nonfunctional requirements of the system.
-For instance, all systems will have the goal that the response time to a user’s request is small enough to allow the user to focus on his/her problem and to maintain his/her train of thought, without being distracted by excessive waiting time.
-The commonality in this case is refined by a later variability because the specific requirement on the response time will depend on the intended usage of the mesh generating system.
+The final section covers nonfunctional requirements of the system.
+For instance, all systems will have the goal that the response time to a user’s request is small enough to allow the user to focus on his/her problem and to maintain his/her train of thought, without excessive waiting time.
+The commonality in this case is refined by a later variability because the specific requirement on the response time will depend on the individual software packages, their simulation models, inputs, and outputs.
 
@@
 		\hline
-		\bf Description & All Lattice Boltzmann versions use a collision operator which concerns collisions between particles.\\
+		\bf Description & All Lattice Boltzmann versions use a collision operator which concerns collisions between particles.
+Collision operators map collisions of particles within the lattice space.
+The Bhatnagar-Gross-Krook Collision Operator is a common LBM collision operator that preserves continuity for a discretized model, for each velocity direction $i$.
+Its equation is $\mathrm{\Omega_i} = \frac{1}{\tau}(f_{i}^{eq}-f_{i})$, where $\tau$ is the relaxation rate towards equilibrium, $f^{eq}$ is the equilibrium particle probability distribution function, and $f$ is the particle probability distribution function.\\
 		\hline
@@
 		\hline
-		\bf Description & Every Lattice Boltzmann version uses a Boltzmann transport equation to describe the statistical behaviour of a system that does not have collisions.\\
+		\bf Description & Every Lattice Boltzmann version uses a Boltzmann transport equation to describe the statistical behaviour of a system that does not have collisions.
+The equation is $(\frac{d}{dt} + \mathrm{e} \cdot \nabla_\mathrm{x} + \frac{F}{m} \cdot \nabla_\mathrm{e})f = \Omega(t)$, where $\mathrm{e}$ is the velocity, $\nabla_\mathrm{x}$ is the position vector gradient, $\nabla_\mathrm{e}$ is the velocity gradient, and $\Omega(t)$ is the collision operator as a function of time.\\
 		\hline
@@
 		\hline
-		\bf Description & The LBS require fluid, model, and boundary information for the problem.\\
+		\bf Description & The LBS require fluid, model, and boundary information for the problem.
+This includes, but is not limited to, fluid acceleration rate, velocity, and viscosity, as well as the number of dimensions in the lattice model, and the number of velocity directions in the lattice.\\
 		\hline
@@
 		\hline
-		\bf Description & LBS write fluid parameter predictions to memory.\\
+		\bf Description & LBS write fluid predictions, such as the output of the transport equation, to memory.\\
 		\hline
@@
 		\hline
-		\bf Description & The use of LBS is easy and efficient.\\
+		\bf Description & The LBS is fairly easy to use. The system should have a good degree of usability, as defined in Section \ref{softwarequalities}.\\
 		\hline
@@
 This section provides a list of characteristics that may vary among family members.
-As in Section \ref{comm_sec}, the first three subsections on variabilities are organized into the following sublists: Mesh Generation, Input and Output.
+As in Section \ref{comm_sec}, the first three subsections on variabilities are organized into the following sublists: Simulation Models, Input and Output.
 The final two subsections list variabilities that can be characterized as system constraints and as nonfunctional requirements.
@@
 		\hline
-		\bf Description & Coefficients for distribution function can vary.\\
+		\bf Description & Coefficients used with the distribution function can vary. These are based on the number of velocity directions in the model.\\
 		\hline
@@
 		\hline
-		\bf Description & The response time required for user interaction with the system varies.\\
+		\bf Description & Usability varies between systems and between models within systems.\\
 		\hline
@@
 This section specifies the parameters of variation for the variabilities listed in Section \ref{var_sec}.
-They are organized into the same five subcategories as employed previously: Mesh Generation, Input, Output, System Constraints, Nonfunctional Requirements.
+They are organized into the same five subcategories as employed previously: Simulation Models, Input, Output, System Constraints, Nonfunctional Requirements.
 
@@
 		\hline
-		\bf Corresponding Variability & \vref{parallel}\\
-		\hline
-		\bf Range of Parameters & Equilibrium can be approximated up to different orders in incompressible or compressible versions.\\
+		\bf Corresponding Variability & \vref{varedf}\\
+		\hline
+		\bf Range of Parameters & Equilibrium approximation varies between incompressible or compressible models.\\		
 		\hline
@@
 		\hline
-		\bf Range of Parameters & Single and multiple array storage patterns can be used.\\
+		\bf Range of Parameters & Various data structures can be used to store function output, including single and multi-dimensional arrays, depending on the problem model and developer preferences.\\
 		\hline
@@
 		\hline
-		\bf Range of Parameters & Numerous coefficients for equilibrium distribution function based on number of velocity directions. \\
+		\bf Range of Parameters & Numerous coefficients for equilibrium distribution function based on number of velocity directions.
+The number of velocity directions is typically 2, 3, 5, 9, 13, 15, 19, or 27.\\
 		\hline
@@
 		\hline
-		\bf Range of Parameters & LBS models can have up to 3 dimensions.\\
+		\bf Range of Parameters & LBS model has 1, 2, or 3 dimensions.\\
 		\hline
@@
 Three dimensional models include options of 15, 19, and 27 velocity directions.
-\\
+Figure \ref{dimension_model} from \citep{thurey2009stable} illustrates two commonly used LBM models in two and three dimensions.\\
 		\hline
@@
 ~\newline
+
+\begin{figure}[h!]
+	\begin{center}
+		\includegraphics[width=0.7\textwidth]{dimension_model}
+		\caption{Commonly used LBM models in two and three dimensions}
+		\label{dimension_model}
+	\end{center}
+\end{figure}
+
 ~\newline
@@
 		\hline
-		\bf Range of Parameters & Collision and collision free transport equations.\\
+		\bf Range of Parameters & Collision and collision free transport equations.
+LBS sometimes use one or the other, often the collision transport equation.
+A Boolean parameter could be used to select between these equations in systems that can apply either equation.\\
 		\hline
@@
 		\hline
-		\bf Range of Parameters & LBS can model one or more fluids.\\
+		\bf Range of Parameters & LBS can model a natural number of fluids.\\
 		\hline
@@
 		\hline
-		\bf Range of Parameters & LBS fluid parameters include Reynolds Number, density, viscosity, time, pressure, force, direction, relaxation rate, turbulence.\\
+		\bf Range of Parameters & LBS fluid parameters include Reynolds Number, density, viscosity, time, pressure, force, direction, relaxation rate, turbulence.
+All of these parameters have the type of real number.\\
 		\hline
@@
 		\hline
-		\bf Range of Parameters & Lattice boundary can have reflective or non-reflective conditions.\\
+		\bf Range of Parameters & Lattice boundary can have reflective or non-reflective conditions.
+Some LBS will only model reflective or non-reflective conditions.
+If there is a choice then this can be indicated by a Boolean parameter.\\
 		\hline
@@
 		\hline
-		\bf Binding Time & Build Time\\
+		\bf Binding Time & Run Time\\
 		\hline
@@
 		\hline
-		\bf Range of Parameters & The maximum amount of time that the user is expected to wait for the system to produce a result ranges between systems and models from near instantaneous to several minutes or longer.\\
+		\bf Range of Parameters & The usability of LBS varies.
+For example, the maximum amount of time that the user is expected to wait for the system to produce a result ranges between systems and models from near instantaneous to several minutes or longer.\\
 		\hline
@@
 		\hline
-		\bf Range of Parameters & The ease with which LBS can run on Windows, MacOS, and Linux system varies. Some LBS can run on all three, while some can run on multiple or only one operating system.\\
+		\bf Range of Parameters & LBS can run on Windows, MacOS, and/or Linux systems. Some LBS can run on all three, while some can run on multiple or only one operating system.\\
 		\hline
@@
 		\hline
-		\bf Range of Parameters & Some LBS require programming skills and domain expert knowledge while some LBS can be run without such skills.\\
+		\bf Range of Parameters & Some LBS require significant time to set up models, while some LBS can be run with relatively less user labour.\\
 		\hline

### 2021-06-15 StateOfPractice/Peter-Notes/Report/Report.tex
updated intro and domain analysis with purpose of CA re functionality; added advantags of LBM and a picture to intro

@@
 
-This report analyzes the development of software packages that use Lattice Boltzmann Methods (LBM), a class of computational fluid dynamics methods, for fluid simulation.
+This report analyzes the development of software packages that use Lattice Boltzmann Methods (LBM), a class of computational fluid dynamics (CFD) methods, for fluid simulation.
 These LBM algorithms consider the behaviours of a collection of particles as a single unit of an intermediate size, and predict their positional probability as they move through a lattice structure.
+Figure \ref{circularflow} from \citep{chen2021phase} presents the streamlines of converged flow past a stationary circular cylinder with varied Raynolds number, a common task of LBM.
+
+\begin{figure}[h!]
+	\begin{center}
+		\includegraphics[width=0.7\textwidth]{circularflow}
+		\caption{Streamlines of the flow past a stationary circular cylinder at Re = 10, 20, and 40}
+		\label{circularflow}
+	\end{center}
+\end{figure}
+
+LBM have several advantages over conventional CFD methods, including a simple calculation procedure, improved parallelization, and robust handling of complex geometries \citep{ganji2015application}.
 LBM are further detailed in Section \ref{domainanalysis}.
@@
 A domain analysis consists of systematically identifying and documenting the commonalities, variabilities, and terminology of a software family \citep{Weiss1997}.
+	The purpose of this analysis in this state of the practice assessment is to better classify the different software products based on their functionality.
 	\item \textbf{Methodology} of this state of the practice assessment, including the steps of the methodology, software quality definitions and how these qualities are assessed in this report, an overview of candidate software selection, empirical measures gathered, and the analytical hierarchy process.
@@
 A software family is defined by \cite{parnas1976design} as “a set of programs whose common properties are so extensive that it is advantageous to study the common properties of the programs before analyzing individual members”.
+
+We added this domain commonality and variability analysis to our original methodology because the first time we did the State of the Practice exercise we found we weren't comparing "apples to apples".
+Software packages can have considerable variation, even when the packages appear to be in the same software family.
+Viewing the ranking exercise as a means for selecting the right tool for a job, functionality is important.
+A 2D Lattice Boltzmann solver is not a substitute for a 3D Lattice Boltzmann solver when the latter is required.
+Therefore, this exercise seeks to classify the different software products based on their functionality.
+Nonfunctional qualities are assessed through other means in our process.
+For instance, through usability experiments as noted in the \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Methodology.pdf}{Methodology for Assessing the State of the Practice for Domain X} document.
+
 This domain analysis of the LBM SCS family is organized into the four subsection located below.
@@
 \end{minipage}\\
-\subsubsection{Nonfunctional Requirements}\label{comm_nfr}
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
-		\hline
-		\bf Item Number& C\refstepcounter{comnum}\thecomnum\label{precision} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
-		\hline
-		\bf Description & LBS provide the precision required for the particular problems it is intended to solve.\\
-		\hline
-		\bf Related Variability & \vref{varprecise}\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
-		\hline
-		\bf Item Number& C\refstepcounter{comnum}\thecomnum\label{accuracy} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
-		\hline
-		\bf Description & LBS provide the accuracy required for the particular problems it is intended to solve.\\
-		\hline
-		\bf Related Variability & \vref{varaccuracy}\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
-		\hline
-		\bf Item Number& C\refstepcounter{comnum}\thecomnum\label{performance} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
-		\hline
-		\bf Description & The response time is small enough to allow the user to focus on their problem without being distracted by excessive waiting times.\\
-		\hline
-		\bf Related Variability & \vref{varresponsetime}\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
-		\hline
-		\bf Item Number& C\refstepcounter{comnum}\thecomnum\label{portability} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
-		\hline
-		\bf Description & LBS will be as portable to other operating systems as required by the users.\\
-		\hline
-		\bf Related Variability & \vref{varportability}\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
-		\hline
-		\bf Item Number& C\refstepcounter{comnum}\thecomnum\label{usability} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
-		\hline
-		\bf Description & The LBS is fairly easy to use. The system should have a good degree of usability, as defined in Section \ref{softwarequalities}.\\
-		\hline
-		\bf Related Variability & \vref{varusability}\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-
 \subsection{Variabilities}\label{var_sec}
@@
 		\bf Related Parameter & \pref{parstorageconst}\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-\subsubsection{Nonfunctional Requirements}
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
-		\hline
-		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{varprecise} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & The precision needed for each input and output.\\
-		\hline
-		\bf Related Commonality & \cref{precision}\\
-		\hline
-		\bf Related Parameter & \pref{parprecise}\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
-		\hline
-		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{varaccuracy} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & The required accuracy for the output.\\
-		\hline
-		\bf Related Commonality & \cref{accuracy}\\
-		\hline
-		\bf Related Parameter & \pref{paraccuracy}\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
-		\hline
-		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{varresponsetime} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & Usability varies between systems and between models within systems.\\
-		\hline
-		\bf Related Commonality & \cref{performance}\\
-		\hline
-		\bf Related Parameter & \pref{parresponse}\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
-		\hline
-		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{varportability} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & The operating systems on which a LBS can run.\\
-		\hline
-		\bf Related Commonality & \cref{portability}\\
-		\hline
-		\bf Related Parameter & \pref{parportability}\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
-		\hline
-		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{varusability} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & The ease with which LBS can effectively be run varies.\\
-		\hline
-		\bf Related Commonality & \cref{usability}\\
-		\hline
-		\bf Related Parameter & \pref{parusability}\\
 		\hline
@@
 \end{minipage}\\
-\subsubsection{Nonfunctional Requirements}
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
-		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parprecise} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{varprecise}\\
-		\hline
-		\bf Range of Parameters & Each input and each output will have specified the precision with which the real number is stored and/or calculated.\\
-		\hline
-		\bf Binding Time & Build Time or Run Time\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
-		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{paraccuracy} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{varaccuracy}\\
-		\hline
-		\bf Range of Parameters & The tolerance allowed for each output produced by the system so that the system produces usable results. \\
-		\hline
-		\bf Binding Time & Build Time or Run Time\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
-		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parresponse} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{varresponsetime}\\
-		\hline
-		\bf Range of Parameters & The usability of LBS varies.
-For example, the maximum amount of time that the user is expected to wait for the system to produce a result ranges between systems and models from near instantaneous to several minutes or longer.\\
-		\hline
-		\bf Binding Time & Specification\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
-		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parportability} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{varportability}\\
-		\hline
-		\bf Range of Parameters & LBS can run on Windows, MacOS, and/or Linux systems. Some LBS can run on all three, while some can run on multiple or only one operating system.\\
-		\hline
-		\bf Binding Time & Specification\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
-		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parusability} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{varusability}\\
-		\hline
-		\bf Range of Parameters & Some LBS require significant time to set up models, while some LBS can be run with relatively less user labour.\\
-		\hline
-		\bf Binding Time & Specification\\
-		\hline
-	\end{tabular}
-\end{minipage}
 

### 2021-06-16 StateOfPractice/Peter-Notes/Report/Report.tex
applied first set of feedback, specifically Methodology section

@@
 
-The LBM software packages are assessed for a given set of quality attributes, using quantitative and qualitative data, in order to answer the research questions found in Section \ref{researchquestions}.
+The LBM software packages are assessed in order to answer the research questions found in the Appendix in \ref{researchquestions}.
+They are assessed for a given set of quality attributes using quantitative and qualitative data.
 The qualities are defined in Section \ref{softwarequalities}.
 Qualitative data is gathered from interviews with software package developers.
-The interview questions are listed in Section \ref{interviewquestions}.
-Quantitative data is measured using the quality metrics listed in Section \ref{measurementtemplate2}.
+The interview questions are listed in the Appendix in \ref{interviewquestions}.
+Quantitative data is measured using the quality metrics listed in the Appendix in \ref{measurementtemplate2}.
 In this context the purposes of the report are as follows:
@@
 The previous set of research questions were critically assessed and modified.
-The updated questions are listed in Section \ref{researchquestions}.
+The updated questions are listed in the Appendix in \ref{researchquestions}.
 In this re-boot we collected more quantitative and qualitative data, focused on software quality attributes, and added more measures, including collecting empirical data and interviewing developers.
@@
 Surface measurements are taken as they allow us to apply the same measurement, with reasonable effort, along all software packages despite technical and functional variabilities among the set of software packages.
+Not all qualities are quantitatively measured.
+For example, performance is not measured by running the code.
+A surface investigation of the documentation of each software package for performance information is conducted instead.
 The measured software packages are open source, and it is recommended to keep this in mind when considering the recommendations of this report.
@@
 
-In this project we set out to answer the research questions found in Section \ref{researchquestions} for the LBM SCS domain.
+In this project we set out to answer the research questions found in the Appendix in \ref{researchquestions} for the LBM SCS domain.
 The process involved systematically measuring and analyzing members of this software family along the quality attributes found in Section \ref{softwarequalities}.
 This included gathering quantitative and qualitative measurements.
-A goal of the project was to produce a hierarchical quality assessment for LBM software packages.
+A goal of the project was to produce a quality assessment for LBM software packages.
+
+The methodology used in this LBM software assessment is a general state of the practice methodology that can be applied to any SCS domain.
+It was developed as an update to previous state of the practice exercises.
+This general methodology is linked in the next subsection.
 
 We collected quantitative data using the measures found in the measurement template of Section \ref{measurementtemplate}.
-Some of this was empirical software engineering related data, such as the number of files, number of lines of code, percentage of issues that are closed.
-Most of the data was gathered by manually observing the software, its source code, and its artifacts, while some was gathered using the tools discussed in Section \ref{empiricalmeasures}.
-
-We also collected qualitative data by interviewing the software package developers and asking them the questions found in Section \ref{interviewquestions}.
+Some of this was empirical software engineering related data, such as the number of files, number of lines of code, percentage of issues that are closed, etc.
+Most of the data was gathered by manually investigating the software, its source code, and its artifacts, while some was gathered using the empirical measurement tools discussed in Section \ref{empiricalmeasures}.
+
+We also collected qualitative data by interviewing the software package developers and asking them the questions found in the Appendix in \ref{interviewquestions}.
 Furthermore, we solicited the assistance of domain experts to better assess each software package by leveraging their experience to assess the functional and non-functional requirements for the software domain.
@@
 	\item Collect empirical measures. This is discussed in Section \ref{empiricalmeasures}.
-	\item Measure using the measurement template. This is discussed in Section \ref{empiricalmeasures}. The measurement template can be found in Section \ref{measurementtemplate2}.
+	\item Measure using the measurement template. This is discussed in Section \ref{empiricalmeasures}. The measurement template can be found in the Appendix in \ref{measurementtemplate2}.
 	\item Survey the developers.
 The developers of each software package in the filtered software list were contacted for voluntary interviews.
-Four interviews were conducted in total.
-The interview questions can be found in Section \ref{interviewquestions}.
+The interview questions can be found in the Appendix in \ref{interviewquestions}.
 	\item Use AHP to rank the software packages. This is discussed in Section \ref{AHP}.
-	\item Analyze the results and answer the research questions. The answers can be found in section \ref{answersquestions}.
+	\item Analyze the results and answer the research questions. The answers can be found in Section \ref{answersquestions}.
 \end{enumerate}
 
-These steps are further detailed in the Methodology for Assessing the State of the Practice for Domain X document found \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Methodology.pdf}{here}.
+These steps are further detailed in the \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Methodology.pdf}{Methodology for Assessing the State of the Practice for Domain X} document.
 
@@
 Software quality attributes facilitate the measurement and comparison of software packages in this state of the practice assessment.
-In this report we adopt software quality definitions from various researchers and subject matter expert entities.
+We adopt software quality definitions from various researchers and subject matter expert entities.
 Some of the definitions are from \cite{Smithetal2020}.
@@
 This requires that the specification is available.
-Software is less likely to have a formal specification if it is not developed by seasoned or professional software developers.
+Software is unlikely to have a formal specification if it is not developed by seasoned or professional software developers.
 Since some software does not have a specification available, the correctness of software cannot always be verified.
@@
 Recoverability from errors also improves reliability.
-In turn the presence of descriptive error messages and logs increases the ability to recover the system.
 
@@
 It is an artifact-dependent quality.
-Understandability is different for the user-interface, source code, documentation.
-In this state of the practice analysis understandability focuses on the source code.
+Understandability is different for the user-interface, source code, and documentation.
+In this state of the practice analysis, understandability focuses on the source code.
 It is measured by the consistency of a formatting style, the extend of modularization, the explicit identification of coding standards, the presence of meaningful identifiers, and clarity of comments.
@@
 Software achieves reproducibility if another developer can take the requirements documentation and re-obtain the same software artifacts \citep{BenureauAndRougier2017}.
-In this context, reproducibility refers to the similarity between artifacts.
-Reproducibility can also refer to the output of the software, irrespective of the artifacts, where the scientific results are compared between software implementations, or between software implementations and manually calculated results.
+This includes the output of the software, where the scientific results are compared between software implementations, or between software implementations and manually calculated results.
 This state of the practice assessment did not quantitatively measure reproducibility.
-Developers were asked in interviews if they have any concern that their computational results won’t be reproducible in the future, and if they have taken any steps to ensure reproducibility.
+Developers were asked in interviews if they have any concern that their computational results won't be reproducible in the future, and if they have taken any steps to ensure reproducibility.
 
@@
 This state of the practice assessment did not quantitatively measure unambiguity.
-Developers were asked in interviews if they think that the current documentation can clearly convey all necessary knowledge to the users, and how they achieved this or what improvements are needed ot achieve it.
+Developers were asked in interviews if they think that the current documentation can clearly convey all necessary knowledge to the users, and how they achieved this or what improvements are needed to achieve it.
 
@@
 
-The initial list had 46 packages, including a few packages that were later found to not have publicly available source code, or to be in an incomplete state of development.
+The initial list had 45 packages, including a few packages that were later found to not have publicly available source code, or to be in an incomplete state of development.
 
@@
 
-In order to reduce the number of members in the candidate software list, the following filters were applied. The filters were applied in the priority order listed.
+To reduce the number of members in the candidate software list to a manageable size, the following filters were applied. The filters were applied in the priority order listed.
 
@@
 		DL\_MESO (LBE) & unclear & 2020 Mar\\
-		waLBerla & 2008 Aug & 2020 Jul\\
-		Sailfish & 2012 Nov & 2019 Jun\\
-		Palabos & unclear & 2020 Jul\\
-		OpenLB & 2007 Jul & 2019 Oct\\
-		MechSys   & 2008 Jun    & 2020 Jul\\
-		LUMA&   2016 Nov   & 2020 Feb\\
-		Ludwig& 2018 Aug & 2020 Jul\\
-		lettuce    & 2019 May & 2020 Jul\\
-		pyLBM & 2015 Jun&   2020 Jun\\
-		lbmpy& unknown  & 2020 Jun  \\
-		TCLB & 2013 Jun  & 2020 Apr\\
 		ESPResSo & 2010 Nov & 2020 Jun\\
 		ESPResSo++ & 2011 Feb & 2020 Apr\\
+		lbmpy& unknown  & 2020 Jun  \\
+		lettuce & 2019 May & 2020 Jul\\
+		Ludwig& 2018 Aug & 2020 Jul\\
+		LUMA& 2016 Nov   & 2020 Feb\\
+		MechSys & 2008 Jun    & 2020 Jul\\
+		OpenLB & 2007 Jul & 2019 Oct\\
+		Palabos & unclear & 2020 Jul\\
+		pyLBM & 2015 Jun&   2020 Jun\\
+		Sailfish & 2012 Nov & 2019 Jun\\
+		TCLB & 2013 Jun  & 2020 Apr\\
+		waLBerla & 2008 Aug & 2020 Jul\\
 		\hline
@@
 
-While the initial list had 46 packages, filtering by scope, usage, and age decreased the size of the list to 23 packages.
-Many of the 23 packages that were removed could not be tested as there was no installation guide, they were incomplete, source code was not publicly available, a license was needed, or the project was out of scope or not up to a standard that would support incorporating them into this study.
+While the initial list had 45 packages, filtering by scope, usage, and age decreased the size of the list to 23 packages.
+Many of the 22 packages that were removed could not be tested as there was no installation guide, they were incomplete, source code was not publicly available, a license was needed, or the project was out of scope or not up to a standard that would support incorporating them into this study.
+These eliminated software packages are listed in the Appendix in Section \ref{eliminatedpackagessection}.
 Of the remaining 23 packages that were studied, some were kept on the list despite being marked as dead due to their prevalence on authoritative lists on LBM software and due to their surface excellence, specifically the considerable time that was put into these projects.
+
 The final list of software packages that were analyzed in this project can be found in the following two tables.
@@
 		\hline
-		SunlightLB & 2005 Sep & 2012 Nov\\
-		MP-LABS & 2008 Jun & 2014 Oct\\
+		HemeLB & 2007 Jun & 2018 Aug\\
+		laboetie & 2014 Nov & 2018 Aug\\		
+		LatBo.jl & 2014 Aug & 2017 Feb\\
 		LIMBES & 2010 Nov & 2014 Dec\\
 		LB3D & unclear & 2012 Mar\\
+		LB2D-Prime & 2005 & 2012 Apr\\
 		LB3D-Prime & 2005 & 2011 Oct\\
-		LB2D-Prime & 2005 & 2012 Apr\\
-		LatBo.jl & 2014 Aug & 2017 Feb\\
-		HemeLB & 2007 Jun & 2018 Aug\\
-		laboetie & 2014 Nov & 2018 Aug\\
+		MP-LABS & 2008 Jun & 2014 Oct\\
+		SunlightLB & 2005 Sep & 2012 Nov\\
 		\hline
@@
 
-The quality measurements in this assessment rely on the gathering and analyzing of raw and processed empirical data related to the research questions listed in Section \ref{researchquestions}.
-
-All of the quality measurements that are part of the AHP analysis are in this category, while qualitative data gathered during interviews with developers is not.
+The quality measurements in this assessment rely on the gathering and analyzing of raw and processed empirical data related to the research questions listed in the Appendix in \ref{researchquestions}.
+
+All of the quality measurements that are part of the AHP analysis are empirical measurements.
+Qualitative data gathered during interviews with developers is not part of the AHP analysis.
 This part of the assessment focuses on data that is reasonably easy to collect.
 Much of the data was gathered by manually reviewing the artifacts of each software package, while some was gathered using the freeware tools discussed below.
-The data that was gathered is listed in the measurement template found in Section \ref{measurementtemplate2}.
-Some of the data required processing other data within the template, including the status of the software package which relied on the last commit date, the percentage of issues that are closed which relied on the number of open and closed issues, and the percentage of code that is comments which relied on the number of total lines and comment lines in text-based files.
-The complete measurement template data was then analyzed using the grading template found in Section \ref{gradingtemplate2}, the output of which was analyzed using the AHP described in Section \ref{AHP}.
+The data that was gathered is listed in the measurement template found in the Appendix in \ref{measurementtemplate2}.
+Some of the data required processing other data within the template, including the status of the software package, which relied on the last commit date; the percentage of issues that are closed, which relied on the number of open and closed issues; and the percentage of code that is comments, which relied on the number of total lines and comment lines in text-based files.
+The complete measurement template data was then analyzed using the grading template found in the Appendix in \ref{gradingtemplate2}, the output of which was analyzed using the AHP described in Section \ref{AHP}.
 
@@
 The Analytical Hierarchy Process (AHP) is a decision-making technique that is used to compare multiple options by multiple criteria.
-In our work AHP was used for comparing and ranking the LBM software packages using the overall impression quality scores that were gathered in the measurement template found in Section \ref{measurementtemplate2} using the grading template found in Section \ref{gradingtemplate2}.
+In our work AHP was used for comparing and ranking the LBM software packages using the overall impression quality scores that were gathered in the measurement template found in the Appendix in \ref{measurementtemplate2} using the grading template found in the Appendix in \ref{gradingtemplate2}.
 AHP performs a pairwise analysis using a matrix and generates an overall score as well as individual quality scores for each software package.
@@
 The tool includes a sensitivity analysis that was used to ensure that the software package rankings are appropriate with respect to the uncertainty of the quality scores.
-The README file of the tool, which includes requirements and usage information, is found \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/AHP2020/LBM/README.txt}{here}.
+The \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/AHP2020/LBM/README.txt}{README} file of the tool includes requirements and usage information.
 
@@
 
-\subsection{Measurement Template}\label{measurementtemplate2}
-The table below lists the set of measures that are used to assess each software product.
-The first set identifies summary information, followed by 9 sets for software qualities and 3 sets for raw metrics.
-Each measure is followed by the type for a valid result.
-A superscript ∗ indicate that a response of this type needs to be accompanied by explanatory text.
-
-\captionof{table}{Measurement Template}\label{measurementtemplate}
-\def\arraystretch{1.22}
-\begin{tabular}{p{16cm}}
-		\hline
-		\textbf{Summary Information}\\
-		\hline
-		Software name? (string)\\
-		URL? (URL)\\
-		Affiliation (institution(s)) (string or {N/A})\\
-		Software purpose (string)\\
-		Number of developers (all developers that have contributed at least one commit to the project) (use repo commit logs) (number)\\
-		How is the project funded? (unfunded, unclear, funded*) where * requires a string to say the source of funding\\
-		Initial release date? (date)\\
-		Last commit date? (date)\\
-		Status? (alive is defined as presence of commits in the last 18 months) ({alive, dead, unclear})\\
-		License? ({GNU GPL, BSD, MIT, terms of use, trial, none, unclear, other*}) * given via a string \\
-		Platforms? (set of {Windows, Linux, OS X, Android, other*}) * given via string\\
-		Software Category?
-The concept category includes software that does not have an officially released version.
-Public software has a released version in the public domain.
-Private software has a released version available to authorized users only.
-({concept, public, private})\\
-		Development model? ({open source, freeware, commercial, unclear})\\
-		Publications about the software? Refers to publications that have used or mentioned the software. (number or {unknown})\\
-		Source code URL? ({set of url, n/a, unclear})\\
-		Programming language(s)? (set of {FORTRAN, Matlab, C, C++, Java, R, Ruby, Python, Cython, BASIC, Pascal, IDL, unclear, other*}) * given via string \\
-		Is there evidence that performance was considered? Performance refers to either speed, storage, or throughput. ({yes∗, no})\\
-		Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have) \\
-		\hline
-\end{tabular}
-
-\def\arraystretch{1.5}
-\begin{tabular}{p{16cm}}
-		\hline		
-		\textbf{Installability  (Measured via installation on a virtual machine.) }\\
-		\hline
-		Are there installation instructions? ({yes, no})\\
-		Are the installation instructions in one place? Place referring to a single document or web-page. ({yes, no, n/a})\\
-		Are the installation instructions linear? Linear meaning progressing  in a single series of steps. ({yes, no, n/a})\\
-		Are the instructions written as if the person doing the installation has none of the dependent packages installed? ({yes, no, unclear})\\
-		Are compatible operating system versions listed? ({yes, no})\\
-		Is there something in place to automate the installation (makefile, script, installer, etc)? ({yes*, no})\\
-		If the software installation broke, was a descriptive error message displayed? ({yes, no, n/a})\\
-		Is there a specified way to validate the installation? ({yes*, no})\\
-		How many steps were involved in the installation? (Includes manual steps like unzipping files) Specify OS. (number, OS)\\
-		What OS was used for the installation? ({Windows, Linux, OS X, Android, other* }) *given via string\\
-		How many extra software packages need to be installed before or during installation? (number)\\
-		Are required package versions listed? ({yes, no, n/a})\\
-		Are there instructions for the installation of required packages / dependencies? ({yes, no, n/a})\\
-		Run uninstall, if available. Were any obvious problems caused? ({yes∗ , no, unavail})\\
-		Overall impression? ({1 .. 10})\\
-		Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have)\\
-		\hline
-\end{tabular}
-
-\def\arraystretch{1.33}
-\begin{tabular}{p{16cm}}
-	\hline		
-	\textbf{Correctness and Verifiability}\\
-	\hline
-	Any reference to the requirements specifications of the program or theory manuals? ({yes∗ , no, unclear})\\
-	What tools or techniques are used to build confidence of correctness?
-({literate programming, automated testing, symbolic execution, model checking, assertions used in the code, Sphinx, Doxygen, Javadoc, confluence, unclear, other*}) * given via string\\
-	If there is a getting started tutorial? ({yes, no})\\
-	Are the tutorial instructions linear? ({yes, no, n/a})\\
-	Does the getting started tutorial provide an expected output? ({yes, no*, n/a})\\
-	Does your tutorial output match the expected output? ({yes, no, n/a})\\
-	Are unit tests available?  ({yes, no, unclear})\\
-	Is there evidence of continuous integration? (for example mentioned in documentation, Jenkins, Travis CI, Bamboo, other) ({yes*, no, unclear})\\
-	Overall impression? ({1 .. 10})\\
-	Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have) \\
-	\hline	
-	\textbf{Surface Reliability}\\
-	\hline
-	Did the software “break” during installation? ({yes∗ , no})\\
-	If the software installation broke, was the installation instance recoverable? ({yes, no, n/a})\\
-	Did the software “break” during the initial tutorial testing? ({yes∗, no, n/a})\\
-	If the tutorial testing broke, was a descriptive error message displayed? ({yes, no, n/a})\\
-	If the tutorial testing broke, was the tutorial testing instance recoverable? ({yes, no, n/a})\\
-	Overall impression? ({1 .. 10})\\
-	Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have)\\
-	\hline		
-	\textbf{Surface Robustness}\\
-	\hline
-	Does the software handle unexpected/unanticipated input (like data of the wrong type, empty input, missing files or links) reasonably?
-(a reasonable response can include an appropriate error message.) ({yes, no∗ })\\
-	For any plain text input files, if all new lines are replaced with new lines and carriage returns, will the software handle this gracefully? ({yes, no∗, n/a})\\
-	Overall impression? ({1 .. 10})\\
-	Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have)\\
-	\hline
-\end{tabular}
-
-\def\arraystretch{1.4}
-\begin{tabular}{p{16cm}}
-	\hline		
-	\textbf{Surface Usability}\\
-	\hline
-	Is there a getting started tutorial? ({yes, no})\\
-	Is there a user manual? ({yes, no})\\
-	Are expected user characteristics documented? ({yes, no})\\
-	What is the user support model? FAQ? User forum? E-mail address to direct questions? Etc. (string)\\
-	Overall impression? ({1 .. 10})\\
-	Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have)\\
-	\hline	
-	\textbf{Maintainability}\\
-	\hline
-	What is the current version number? (number)\\
-	Is there any information on how code is reviewed, or how to contribute? ({yes*, no})\\
-	Are artifacts available?
-(List every type of file that is not a code file – for examples please look at the ‘Artifact Name’ column of https://gitlab.cas.mcmaster.ca/SEforSC/se4sc/-/blob/git-svn/GradStudents/Olu/ResearchProposal/Artifacts\_MiningV3.xlsx) ({yes*, no, unclear}) *list via string\\
-	What issue tracking tool is employed? (set of {Trac, JIRA, Redmine, e-mail, discussion board, sourceforge, google code, git, BitBucket, none, unclear, other*}) * given via string\\
-	What is the percentage of identified issues that are closed? (percentage)\\
-	What percentage of code is comments? (percentage)\\
-	Which version control system is in use? ({svn, cvs, git, github, unclear, other*}) * given via string\\
-	Overall impression? ({1 .. 10})\\
-	Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have)\\
-	\hline		
-	\textbf{Reusability}\\
-	\hline
-	How many code files are there? (number)\\
-	Is API documented? ({yes, no, n/a})\\
-	Overall impression? ({1 .. 10})\\
-	Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have)\\
-	\hline
-\end{tabular}
-
-\def\arraystretch{1.4}
-\begin{tabular}{p{16cm}}
-	\hline		
-	\textbf{Surface Understandability (Based on 10 random source files)}\\
-	\hline
-	Consistent indentation and formatting style? ({yes, no, n/a})\\
-	Explicit identification of a coding standard? ({yes∗, no, n/a})\\
-	Are the code identifiers consistent, distinctive, and meaningful? ({yes, no∗ , n/a})\\
-	Are constants (other than 0 and 1) hard coded into the program? ({yes, no∗ , n/a})\\
-	Comments are clear, indicate what is being done, not how? ({yes, no∗ , n/a})\\
-	Is the name/URL of any algorithms used mentioned? ({yes, no∗ , n/a})\\
-	Parameters are in the same order for all functions? ({yes, no∗ , n/a})\\
-	Is code modularized? ({yes, no∗ , n/a})\\
-	Overall impression? ({1 .. 10})\\
-	Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have)\\
-	\hline		
-	\textbf{Visibility/Transparency}\\
-	\hline
-	Is the development process defined? If yes, what process is used. ({yes∗, no, n/a})\\
-	Are there any documents recording the development process and status?  ({yes∗, no}))\\
-	Is the development environment documented? ({yes∗, no})\\
-	Are there release notes? ({yes∗, no})\\
-	Overall impression? ({1 .. 10})\\
-	Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have)\\
-	\hline		
-	\textbf{Raw Metrics (Measured via git\_stats)}\\
-	\hline
-	Number of text-based files. (number)\\
-	Number of binary files. (number)\\
-	Number of total lines in text-based files. (number)\\
-	Number of total lines added to text-based files. (number)\\
-	Number of total lines deleted from text-based files. (number)\\
-	Number of total commits. (number)\\
-	Numbers of commits by year in the last 5 years. (Count from as early as possible if the project is younger than 5 years.) (list of numbers)\\
-	Numbers of commits by month in the last 12 months. (list of numbers)\\
-	\hline
-\end{tabular}
-
-\def\arraystretch{1.4}
-\begin{tabular}{p{16cm}}
-	\hline		
-	\textbf{Raw Metrics (Measured via scc)}\\
-	\hline
-	Number of text-based files. (number)\\
-	Number of total lines in text-based files. (number)\\
-	Number of code lines in text-based files. (number)\\
-	Number of comment lines in text-based files. (number)\\
-	Number of blank lines in text-based files. (number)\\
-	\hline
-	\textbf{Repo Metrics (Measured via GitHub)}\\
-	\hline
-	Number of stars. (number)\\
-	Number of forks. (number)\\
-	Number of people watching this repo. (number)\\
-	Number of open pull requests. (number)\\
-	Number of closed pull requests. (number)\\
-	\hline
-\end{tabular}
-
-\newpage
-
-\subsection{Grading Template}\label{gradingtemplate2}
-The table below lists how each quality measure of the measurement template is used to calculate an overall impression in each software quality set.\newline
-
-\captionof{table}{Grading Template}\label{gradingtemplate}
-\def\arraystretch{1.5}
-\begin{tabular}{p{16cm}}
-	\hline		
-	\textbf{Installability  (Measured via installation on a virtual machine.) }\\
-	\hline
-	Are there installation instructions? ({yes=1, no=-1})\\
-	Are the installation instructions in one place? Place referring to a single document or web-page. ({yes=1, no=0, n/a=0})\\
-	Are the installation instructions linear? Linear meaning progressing  in a single series of steps. ({yes=1, no=0, n/a=0})\\
-	Are the instructions written as if the person doing the installation has none of the dependent packages installed? ({yes=1, no=0, unclear=0})\\
-	Are compatible operating system versions listed? ({yes=1, no=0})\\
-	Is there something in place to automate the installation (makefile, script, installer, etc)? ({yes*=1, no=-1})\\
-	If the software installation broke, was a descriptive error message displayed? ({yes=0, no=-2, n/a=1})\\
-	Is there a specified way to validate the installation? ({yes*=1, no=0})\\
-	How many steps were involved in the installation? (Includes manual steps like unzipping files) Specify OS. ($<$10 = 1)\\
-	What OS was used for the installation? (does not count)\\
-	How many extra software packages need to be installed before or during installation? ($<$10 = 1)\\
-	Are required package versions listed? ({yes=1, no=0, n/a=1})\\
-	Are there instructions for the installation of required packages / dependencies? ({yes=1, no=0, n/a=1})\\
-	Run uninstall, if available. Were any obvious problems caused? ({yes∗=0, no=1, unavail=1})\\
-	Overall impression? (a sum of $>$10 is rounded down to 10)\\
-	\hline
-\end{tabular}
-
-\def\arraystretch{1.33}
-\begin{tabular}{p{16cm}}
-	\hline		
-	\textbf{Correctness and Verifiability}\\
-	\hline
-	Any reference to the requirements specifications of the program or theory manuals? ({yes∗=2, no=0, unclear=0})\\
-	What tools or techniques are used to build confidence of correctness? (any=1, unclear=0)\\
-	If there is a getting started tutorial? ({yes=2, no=0})\\
-	Are the tutorial instructions linear? ({yes=1, no=0, n/a=0})\\
-	Does the getting started tutorial provide an expected output? ({yes=1, no*=0, n/a=0})\\
-	Does your tutorial output match the expected output? ({yes=1, no=0, n/a=0})\\
-	Are unit tests available?  ({yes=1, no=0, unclear=0})\\
-	Is there evidence of continuous integration? (for example mentioned in documentation, Jenkins, Travis CI, Bamboo, other) ({yes*=1, no=0, unclear=0})\\
-	\hline	
-	\textbf{Surface Reliability}\\
-	\hline
-	Did the software “break” during installation? ({yes∗=0, no=5})\\
-	If the software installation broke, was the installation instance recoverable? ({yes=5, no=0, n/a=0})\\
-	Did the software “break” during the initial tutorial testing? ({yes∗=0, no=5, n/a=0})\\
-	If the tutorial testing broke, was a descriptive error message displayed? ({yes=2, no=0, n/a=0})\\
-	If the tutorial testing broke, was the tutorial testing instance recoverable? ({yes=3, no=0, n/a=0})\\
-	\hline		
-	\textbf{Surface Robustness}\\
-	\hline
-	Does the software handle unexpected/unanticipated input (like data of the wrong type, empty input, missing files or links) reasonably?
-(a reasonable response can include an appropriate error message.) ({yes=5, no∗=0})\\
-	For any plain text input files, if all new lines are replaced with new lines and carriage returns, will the software handle this gracefully? ({yes=5, no∗=0, n/a=5})\\
-	\hline		
-	\textbf{Surface Usability}\\
-	\hline
-	Is there a getting started tutorial? ({yes=3, no=0})\\
-	Is there a user manual? ({yes=4, no=0})\\
-	Are expected user characteristics documented? ({yes=1, no=0})\\
-	What is the user support model? FAQ? User forum? E-mail address to direct questions? Etc. (one=1, two+=2, none=0)\\
-	\hline
-\end{tabular}
-
-\def\arraystretch{1.4}
-\begin{tabular}{p{16cm}}
-	\hline
-	\textbf{Maintainability}\\
-	\hline
-	What is the current version number? (provided=1, nothing=0)\\
-	Is there any information on how code is reviewed, or how to contribute? ({yes*=1, no=0})\\
-	Are artifacts available?
-(List every type of file that is not a code file – for examples please look at the ‘Artifact Name’ column of https://gitlab.cas.mcmaster.ca/SEforSC/se4sc/-/blob/git-svn/GradStudents/Olu/ResearchProposal/Artifacts\_MiningV3.xlsx) (Rate 0 – 2 depending on how many and perceived quality)\\
-	What issue tracking tool is employed? (nothing=0, email of other private=1, anything public or accessible by all devs (eg git) = 2)\\
-	What is the percentage of identified issues that are closed? (50$\%$+=1, $<$50$\%$=0)\\
-	What percentage of code is comments? (10$\%$+=1, $<$10$\%$=0)\\
-	Which version control system is in use? (anything=2, nothing=0)\\
-	\hline		
-	\textbf{Reusability}\\
-	\hline
-	How many code files are there? (0-9=0, 10-49=1, 50-99=3, 100-299=4, 300-599=5, 600-999=6, 1000+=8)\\
-	Is API documented? ({yes=2, no=0, n/a=0})\\
-	\hline	
-	\textbf{Surface Understandability (Based on 10 random source files)}\\
-	\hline
-	Consistent indentation and formatting style? ({yes=1, no=0, n/a=0})\\
-	Explicit identification of a coding standard? ({yes∗=1, no=0, n/a=0})\\
-	Are the code identifiers consistent, distinctive, and meaningful? ({yes=2, no∗=0, n/a=0})\\
-	Are constants (other than 0 and 1) hard coded into the program? ({yes=1, no∗=0, n/a=0})\\
-	Comments are clear, indicate what is being done, not how? ({yes=2, no∗=0, n/a=0})\\
-	Is the name/URL of any algorithms used mentioned? ({yes=1, no∗=0, n/a=0})\\
-	Parameters are in the same order for all functions? ({yes=1, no∗=0, n/a=0})\\
-	Is code modularized? ({yes=1, no∗=0, n/a=0})\\
-	\hline		
-	\textbf{Visibility/Transparency}\\
-	\hline
-	Is the development process defined? If yes, what process is used. ({yes∗=3, no=0, n/a=0})\\
-	Are there any documents recording the development process and status?  ({yes∗=3, no=0}))\\
-	Is the development environment documented? ({yes∗=2, no=0})\\
-	Are there release notes? ({yes∗=2, no=0})\\
-	\hline
-\end{tabular}
-\newpage
-
-\subsection{Ethics Approval}
-This project received ethics clearance from the McMaster Research Ethics Board on February 20, 2021.\newline
-
-Project Title: AIMSS - State of the Practice\newline
-
-MREB\#: 5219
-
-\newpage
-
 \subsection{Developer Interview Questions}\label{interviewquestions}
@@
 
+\subsection{Measurement Template}\label{measurementtemplate2}
+The table below lists the set of measures that are used to assess each software product.
+The first set identifies summary information, followed by 9 sets for software qualities and 3 sets for raw metrics.
+Each measure is followed by the type for a valid result.
+A superscript ∗ indicate that a response of this type needs to be accompanied by explanatory text.
+
+\captionof{table}{Measurement Template}\label{measurementtemplate}
+\def\arraystretch{1.22}
+\begin{tabular}{p{16cm}}
+		\hline
+		\textbf{Summary Information}\\
+		\hline
+		Software name? (string)\\
+		URL? (URL)\\
+		Affiliation (institution(s)) (string or {N/A})\\
+		Software purpose (string)\\
+		Number of developers (all developers that have contributed at least one commit to the project) (use repo commit logs) (number)\\
+		How is the project funded? (unfunded, unclear, funded*) where * requires a string to say the source of funding\\
+		Initial release date? (date)\\
+		Last commit date? (date)\\
+		Status? (alive is defined as presence of commits in the last 18 months) ({alive, dead, unclear})\\
+		License? ({GNU GPL, BSD, MIT, terms of use, trial, none, unclear, other*}) * given via a string \\
+		Platforms? (set of {Windows, Linux, OS X, Android, other*}) * given via string\\
+		Software Category?
+The concept category includes software that does not have an officially released version.
+Public software has a released version in the public domain.
+Private software has a released version available to authorized users only.
+({concept, public, private})\\
+		Development model? ({open source, freeware, commercial, unclear})\\
+		Publications about the software? Refers to publications that have used or mentioned the software. (number or {unknown})\\
+		Source code URL? ({set of url, n/a, unclear})\\
+		Programming language(s)? (set of {FORTRAN, Matlab, C, C++, Java, R, Ruby, Python, Cython, BASIC, Pascal, IDL, unclear, other*}) * given via string \\
+		Is there evidence that performance was considered? Performance refers to either speed, storage, or throughput. ({yes∗, no})\\
+		Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have) \\
+		\hline
+\end{tabular}
+
+\def\arraystretch{1.5}
+\begin{tabular}{p{16cm}}
+		\hline		
+		\textbf{Installability  (Measured via installation on a virtual machine.) }\\
+		\hline
+		Are there installation instructions? ({yes, no})\\
+		Are the installation instructions in one place? Place referring to a single document or web-page. ({yes, no, n/a})\\
+		Are the installation instructions linear? Linear meaning progressing  in a single series of steps. ({yes, no, n/a})\\
+		Are the instructions written as if the person doing the installation has none of the dependent packages installed? ({yes, no, unclear})\\
+		Are compatible operating system versions listed? ({yes, no})\\
+		Is there something in place to automate the installation (makefile, script, installer, etc)? ({yes*, no})\\
+		If the software installation broke, was a descriptive error message displayed? ({yes, no, n/a})\\
+		Is there a specified way to validate the installation? ({yes*, no})\\
+		How many steps were involved in the installation? (Includes manual steps like unzipping files) Specify OS. (number, OS)\\
+		What OS was used for the installation? ({Windows, Linux, OS X, Android, other* }) *given via string\\
+		How many extra software packages need to be installed before or during installation? (number)\\
+		Are required package versions listed? ({yes, no, n/a})\\
+		Are there instructions for the installation of required packages / dependencies? ({yes, no, n/a})\\
+		Run uninstall, if available. Were any obvious problems caused? ({yes∗ , no, unavail})\\
+		Overall impression? ({1 .. 10})\\
+		Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have)\\
+		\hline
+\end{tabular}
+
+\def\arraystretch{1.33}
+\begin{tabular}{p{16cm}}
+	\hline		
+	\textbf{Correctness and Verifiability}\\
+	\hline
+	Any reference to the requirements specifications of the program or theory manuals? ({yes∗ , no, unclear})\\
+	What tools or techniques are used to build confidence of correctness?
+({literate programming, automated testing, symbolic execution, model checking, assertions used in the code, Sphinx, Doxygen, Javadoc, confluence, unclear, other*}) * given via string\\
+	If there is a getting started tutorial? ({yes, no})\\
+	Are the tutorial instructions linear? ({yes, no, n/a})\\
+	Does the getting started tutorial provide an expected output? ({yes, no*, n/a})\\
+	Does your tutorial output match the expected output? ({yes, no, n/a})\\
+	Are unit tests available?  ({yes, no, unclear})\\
+	Is there evidence of continuous integration? (for example mentioned in documentation, Jenkins, Travis CI, Bamboo, other) ({yes*, no, unclear})\\
+	Overall impression? ({1 .. 10})\\
+	Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have) \\
+	\hline	
+	\textbf{Surface Reliability}\\
+	\hline
+	Did the software “break” during installation? ({yes∗ , no})\\
+	If the software installation broke, was the installation instance recoverable? ({yes, no, n/a})\\
+	Did the software “break” during the initial tutorial testing? ({yes∗, no, n/a})\\
+	If the tutorial testing broke, was a descriptive error message displayed? ({yes, no, n/a})\\
+	If the tutorial testing broke, was the tutorial testing instance recoverable? ({yes, no, n/a})\\
+	Overall impression? ({1 .. 10})\\
+	Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have)\\
+	\hline		
+	\textbf{Surface Robustness}\\
+	\hline
+	Does the software handle unexpected/unanticipated input (like data of the wrong type, empty input, missing files or links) reasonably?
+(a reasonable response can include an appropriate error message.) ({yes, no∗ })\\
+	For any plain text input files, if all new lines are replaced with new lines and carriage returns, will the software handle this gracefully? ({yes, no∗, n/a})\\
+	Overall impression? ({1 .. 10})\\
+	Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have)\\
+	\hline
+\end{tabular}
+
+\def\arraystretch{1.4}
+\begin{tabular}{p{16cm}}
+	\hline		
+	\textbf{Surface Usability}\\
+	\hline
+	Is there a getting started tutorial? ({yes, no})\\
+	Is there a user manual? ({yes, no})\\
+	Are expected user characteristics documented? ({yes, no})\\
+	What is the user support model? FAQ? User forum? E-mail address to direct questions? Etc. (string)\\
+	Overall impression? ({1 .. 10})\\
+	Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have)\\
+	\hline	
+	\textbf{Maintainability}\\
+	\hline
+	What is the current version number? (number)\\
+	Is there any information on how code is reviewed, or how to contribute? ({yes*, no})\\
+	Are artifacts available?
+(List every type of file that is not a code file – for examples please look at the ‘Artifact Name’ column of https://gitlab.cas.mcmaster.ca/SEforSC/se4sc/-/blob/git-svn/GradStudents/Olu/ResearchProposal/Artifacts\_MiningV3.xlsx) ({yes*, no, unclear}) *list via string\\
+	What issue tracking tool is employed? (set of {Trac, JIRA, Redmine, e-mail, discussion board, sourceforge, google code, git, BitBucket, none, unclear, other*}) * given via string\\
+	What is the percentage of identified issues that are closed? (percentage)\\
+	What percentage of code is comments? (percentage)\\
+	Which version control system is in use? ({svn, cvs, git, github, unclear, other*}) * given via string\\
+	Overall impression? ({1 .. 10})\\
+	Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have)\\
+	\hline		
+	\textbf{Reusability}\\
+	\hline
+	How many code files are there? (number)\\
+	Is API documented? ({yes, no, n/a})\\
+	Overall impression? ({1 .. 10})\\
+	Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have)\\
+	\hline
+\end{tabular}
+
+\def\arraystretch{1.4}
+\begin{tabular}{p{16cm}}
+	\hline		
+	\textbf{Surface Understandability (Based on 10 random source files)}\\
+	\hline
+	Consistent indentation and formatting style? ({yes, no, n/a})\\
+	Explicit identification of a coding standard? ({yes∗, no, n/a})\\
+	Are the code identifiers consistent, distinctive, and meaningful? ({yes, no∗ , n/a})\\
+	Are constants (other than 0 and 1) hard coded into the program? ({yes, no∗ , n/a})\\
+	Comments are clear, indicate what is being done, not how? ({yes, no∗ , n/a})\\
+	Is the name/URL of any algorithms used mentioned? ({yes, no∗ , n/a})\\
+	Parameters are in the same order for all functions? ({yes, no∗ , n/a})\\
+	Is code modularized? ({yes, no∗ , n/a})\\
+	Overall impression? ({1 .. 10})\\
+	Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have)\\
+	\hline		
+	\textbf{Visibility/Transparency}\\
+	\hline
+	Is the development process defined? If yes, what process is used. ({yes∗, no, n/a})\\
+	Are there any documents recording the development process and status?  ({yes∗, no}))\\
+	Is the development environment documented? ({yes∗, no})\\
+	Are there release notes? ({yes∗, no})\\
+	Overall impression? ({1 .. 10})\\
+	Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have)\\
+	\hline		
+	\textbf{Raw Metrics (Measured via git\_stats)}\\
+	\hline
+	Number of text-based files. (number)\\
+	Number of binary files. (number)\\
+	Number of total lines in text-based files. (number)\\
+	Number of total lines added to text-based files. (number)\\
+	Number of total lines deleted from text-based files. (number)\\
+	Number of total commits. (number)\\
+	Numbers of commits by year in the last 5 years. (Count from as early as possible if the project is younger than 5 years.) (list of numbers)\\
+	Numbers of commits by month in the last 12 months. (list of numbers)\\
+	\hline
+\end{tabular}
+
+\def\arraystretch{1.4}
+\begin{tabular}{p{16cm}}
+	\hline		
+	\textbf{Raw Metrics (Measured via scc)}\\
+	\hline
+	Number of text-based files. (number)\\
+	Number of total lines in text-based files. (number)\\
+	Number of code lines in text-based files. (number)\\
+	Number of comment lines in text-based files. (number)\\
+	Number of blank lines in text-based files. (number)\\
+	\hline
+	\textbf{Repo Metrics (Measured via GitHub)}\\
+	\hline
+	Number of stars. (number)\\
+	Number of forks. (number)\\
+	Number of people watching this repo. (number)\\
+	Number of open pull requests. (number)\\
+	Number of closed pull requests. (number)\\
+	\hline
+\end{tabular}
+
+\newpage
+
+\subsection{Eliminated Software Packages}\label{eliminatedpackagessection}
+
+The table below lists the software packages that were filtered out of the candidate software list.
+
+\begin{center}
+	\begin{tabular}{ p{4cm}p{4cm}}
+		\hline
+		Name & Reason(s) For Removal\\
+		\hline
+		ch4-project&Scope \& Usage\\
+		CUDA-LBM-simulator&Scope \& Age\\
+		elbe&Scope \& Usage\\
+		eLBM&Usage\\
+		firesim&Scope \& Usage \& Age\\
+		fvLBM&Scope \& Usage \& Age\\
+		JFlowSim&Scope \& Usage \& Age\\
+		LatticeBoltzmann&Scope \& Usage\\
+		LBM&Scope \& Usage \& Age\\
+		LBM-Cplusplus&Scope \& Usage\\
+		LBM-EP&Usage\\
+		LBM\_MATLAB&Scope \& Usage \& Age\\
+		LBSim&Scope \& Usage \& Age\\
+		listLBM&Scope \& Usage \& Age\\
+		loliverhennigh&Scope \& Usage \& Age\\
+		openLBMflow&Scope \& Usage \& Age\\
+		ParallelLbmCranfield&Usage \& Age\\
+		PowerFLOW&Usage\\
+		ProLB&Usage\\
+		Taxila-LBM&Usage \& Age\\
+		turbulent\_lbm\_multigpu&Usage \& Age\\
+		wlb&Scope \& Usage \& Age\\
+		\hline
+	\end{tabular}
+	\captionof{table}{Eliminated Software Packages}\label{eliminatedpackages}
+\end{center}
+
+\newpage
+
+\subsection{Grading Template}\label{gradingtemplate2}
+The table below lists how each quality measure of the measurement template is used to calculate an overall impression in each software quality set.\newline
+
+\captionof{table}{Grading Template}\label{gradingtemplate}
+\def\arraystretch{1.5}
+\begin{tabular}{p{16cm}}
+	\hline		
+	\textbf{Installability  (Measured via installation on a virtual machine.) }\\
+	\hline
+	Are there installation instructions? ({yes=1, no=-1})\\
+	Are the installation instructions in one place? Place referring to a single document or web-page. ({yes=1, no=0, n/a=0})\\
+	Are the installation instructions linear? Linear meaning progressing  in a single series of steps. ({yes=1, no=0, n/a=0})\\
+	Are the instructions written as if the person doing the installation has none of the dependent packages installed? ({yes=1, no=0, unclear=0})\\
+	Are compatible operating system versions listed? ({yes=1, no=0})\\
+	Is there something in place to automate the installation (makefile, script, installer, etc)? ({yes*=1, no=-1})\\
+	If the software installation broke, was a descriptive error message displayed? ({yes=0, no=-2, n/a=1})\\
+	Is there a specified way to validate the installation? ({yes*=1, no=0})\\
+	How many steps were involved in the installation? (Includes manual steps like unzipping files) Specify OS. ($<$10 = 1)\\
+	What OS was used for the installation? (does not count)\\
+	How many extra software packages need to be installed before or during installation? ($<$10 = 1)\\
+	Are required package versions listed? ({yes=1, no=0, n/a=1})\\
+	Are there instructions for the installation of required packages / dependencies? ({yes=1, no=0, n/a=1})\\
+	Run uninstall, if available. Were any obvious problems caused? ({yes∗=0, no=1, unavail=1})\\
+	Overall impression? (a sum of $>$10 is rounded down to 10)\\
+	\hline
+\end{tabular}
+
+\def\arraystretch{1.33}
+\begin{tabular}{p{16cm}}
+	\hline		
+	\textbf{Correctness and Verifiability}\\
+	\hline
+	Any reference to the requirements specifications of the program or theory manuals? ({yes∗=2, no=0, unclear=0})\\
+	What tools or techniques are used to build confidence of correctness? (any=1, unclear=0)\\
+	If there is a getting started tutorial? ({yes=2, no=0})\\
+	Are the tutorial instructions linear? ({yes=1, no=0, n/a=0})\\
+	Does the getting started tutorial provide an expected output? ({yes=1, no*=0, n/a=0})\\
+	Does your tutorial output match the expected output? ({yes=1, no=0, n/a=0})\\
+	Are unit tests available?  ({yes=1, no=0, unclear=0})\\
+	Is there evidence of continuous integration? (for example mentioned in documentation, Jenkins, Travis CI, Bamboo, other) ({yes*=1, no=0, unclear=0})\\
+	\hline	
+	\textbf{Surface Reliability}\\
+	\hline
+	Did the software “break” during installation? ({yes∗=0, no=5})\\
+	If the software installation broke, was the installation instance recoverable? ({yes=5, no=0, n/a=0})\\
+	Did the software “break” during the initial tutorial testing? ({yes∗=0, no=5, n/a=0})\\
+	If the tutorial testing broke, was a descriptive error message displayed? ({yes=2, no=0, n/a=0})\\
+	If the tutorial testing broke, was the tutorial testing instance recoverable? ({yes=3, no=0, n/a=0})\\
+	\hline		
+	\textbf{Surface Robustness}\\
+	\hline
+	Does the software handle unexpected/unanticipated input (like data of the wrong type, empty input, missing files or links) reasonably?
+(a reasonable response can include an appropriate error message.) ({yes=5, no∗=0})\\
+	For any plain text input files, if all new lines are replaced with new lines and carriage returns, will the software handle this gracefully? ({yes=5, no∗=0, n/a=5})\\
+	\hline		
+	\textbf{Surface Usability}\\
+	\hline
+	Is there a getting started tutorial? ({yes=3, no=0})\\
+	Is there a user manual? ({yes=4, no=0})\\
+	Are expected user characteristics documented? ({yes=1, no=0})\\
+	What is the user support model? FAQ? User forum? E-mail address to direct questions? Etc. (one=1, two+=2, none=0)\\
+	\hline
+\end{tabular}
+
+\def\arraystretch{1.4}
+\begin{tabular}{p{16cm}}
+	\hline
+	\textbf{Maintainability}\\
+	\hline
+	What is the current version number? (provided=1, nothing=0)\\
+	Is there any information on how code is reviewed, or how to contribute? ({yes*=1, no=0})\\
+	Are artifacts available?
+(List every type of file that is not a code file – for examples please look at the ‘Artifact Name’ column of https://gitlab.cas.mcmaster.ca/SEforSC/se4sc/-/blob/git-svn/GradStudents/Olu/ResearchProposal/Artifacts\_MiningV3.xlsx) (Rate 0 – 2 depending on how many and perceived quality)\\
+	What issue tracking tool is employed? (nothing=0, email of other private=1, anything public or accessible by all devs (eg git) = 2)\\
+	What is the percentage of identified issues that are closed? (50$\%$+=1, $<$50$\%$=0)\\
+	What percentage of code is comments? (10$\%$+=1, $<$10$\%$=0)\\
+	Which version control system is in use? (anything=2, nothing=0)\\
+	\hline		
+	\textbf{Reusability}\\
+	\hline
+	How many code files are there? (0-9=0, 10-49=1, 50-99=3, 100-299=4, 300-599=5, 600-999=6, 1000+=8)\\
+	Is API documented? ({yes=2, no=0, n/a=0})\\
+	\hline	
+	\textbf{Surface Understandability (Based on 10 random source files)}\\
+	\hline
+	Consistent indentation and formatting style? ({yes=1, no=0, n/a=0})\\
+	Explicit identification of a coding standard? ({yes∗=1, no=0, n/a=0})\\
+	Are the code identifiers consistent, distinctive, and meaningful? ({yes=2, no∗=0, n/a=0})\\
+	Are constants (other than 0 and 1) hard coded into the program? ({yes=1, no∗=0, n/a=0})\\
+	Comments are clear, indicate what is being done, not how? ({yes=2, no∗=0, n/a=0})\\
+	Is the name/URL of any algorithms used mentioned? ({yes=1, no∗=0, n/a=0})\\
+	Parameters are in the same order for all functions? ({yes=1, no∗=0, n/a=0})\\
+	Is code modularized? ({yes=1, no∗=0, n/a=0})\\
+	\hline		
+	\textbf{Visibility/Transparency}\\
+	\hline
+	Is the development process defined? If yes, what process is used. ({yes∗=3, no=0, n/a=0})\\
+	Are there any documents recording the development process and status?  ({yes∗=3, no=0}))\\
+	Is the development environment documented? ({yes∗=2, no=0})\\
+	Are there release notes? ({yes∗=2, no=0})\\
+	\hline
+\end{tabular}
+\newpage
+
+\subsection{Ethics Approval}
+This project received ethics clearance from the McMaster Research Ethics Board on February 20, 2021.\newline
+
+Project Title: AIMSS - State of the Practice\newline
+
+MREB\#: 5219
+
+\newpage
+
 \bibliographystyle {plainnat}

### 2021-06-16 StateOfPractice/Peter-Notes/Report/Report.tex
moved research questions to Introduction

@@
 
-\subsection{Purpose}
-
-The LBM software packages are assessed in order to answer the research questions found in the Appendix in \ref{researchquestions}.
-They are assessed for a given set of quality attributes using quantitative and qualitative data.
-The qualities are defined in Section \ref{softwarequalities}.
-Qualitative data is gathered from interviews with software package developers.
-The interview questions are listed in the Appendix in \ref{interviewquestions}.
-Quantitative data is measured using the quality metrics listed in the Appendix in \ref{measurementtemplate2}.
-In this context the purposes of the report are as follows:
-
-\begin{itemize}
-	\item Develop and test an updated methodology for assessing the state of the practice of SCS projects
-	\item Describe the specific example of LBM software structure and goals
-	\item Report on the measure of a subset of LBM software packages along quality metrics
-	\item Evaluate the state of the practice of LBM software development along quality attributes
-	\item Make suggestions for improving LBM software along quality attributes
-	\item Make suggestions for improving state of the practice of SCS assessments
-\end{itemize}
-
-\subsection{Motivation}
-
-The purpose of state of the practice assessments is to understand how software quality is impacted by software development choices, including principles, processes, and tools, within SCS communities.
-This knowledge can be used to guide future development of SCS, specifically along quality attributes, and reduce software quality failures.
-This work reports on the state of the practice of LBM software development and makes suggestions on improving the quality of software in this domain.
-
-This assessment of the state of the practice of LBM software development builds off of prior work on assessing the state of research software development.
-The previous state of the practice assessments include domains such as Geographic Information Systems \citep{SmithEtAl2018_arXivGIS}, Mesh Generators \citep{SmithEtAl2016}, Oceanographic Software \citep{smith2015state}, Seismology software \citep{SmithEtAl2018}, and statistical software for psychology \citep{SmithEtAl2018_StatSoft}.
-
-In the course of this assessment we updated the methodology that was used during previous assessments.
-Details of the new methodology are in \citep{methodology2020}.
-The previous set of research questions were critically assessed and modified.
-The updated questions are listed in the Appendix in \ref{researchquestions}.
-In this re-boot we collected more quantitative and qualitative data, focused on software quality attributes, and added more measures, including collecting empirical data and interviewing developers.
-We have also added a domain analysis to better characterize the functionality provided by the software, and have leveraged the expertise of a domain expert.
-As in past assessments, the collected data was combined to rank the software using the Analytic Hierarchy Process (AHP).
-The domain expert was consulted to verify the ordering.
-
-\subsection{Scope}
-
-This report analyzes a filtered set of LBM software packages along quantitative and qualitative measures.
-Many of these measures are captured using surface measurements, which can be categorized as initial and easy to capture measurements, of the underlying quality; they may not represent the true quality of the software.
-Surface measurements are taken as they allow us to apply the same measurement, with reasonable effort, along all software packages despite technical and functional variabilities among the set of software packages.
-Not all qualities are quantitatively measured.
-For example, performance is not measured by running the code.
-A surface investigation of the documentation of each software package for performance information is conducted instead.
-The measured software packages are open source, and it is recommended to keep this in mind when considering the recommendations of this report.
-Some recommendations may not apply to closed-source software.
-Practices surrounding close source software development may differ.
-
-The report addresses what was done during the development of the software packages to address the quality attributes listed in Section \ref{softwarequalities}.
-It then provides general guidance on how to improve these qualities when developing LBM software.
-It does not make suggestions on what should have been done or should be done for any one specific package.
-Best practices may differ among software packages due to their inherent organizational and technical differences.
-
-\subsection{Organization}
-
-The report is organized as follows:
-
-\begin{itemize}
-	\item \textbf{Introduction} to the report, including its purpose, motivation, scope, and organization.
-	\item LBM software \textbf{Domain Analysis}.
-A domain analysis consists of systematically identifying and documenting the commonalities, variabilities, and terminology of a software family \citep{Weiss1997}.
-	The purpose of this analysis in this state of the practice assessment is to better classify the different software products based on their functionality.
-	\item \textbf{Methodology} of this state of the practice assessment, including the steps of the methodology, software quality definitions and how these qualities are assessed in this report, an overview of candidate software selection, empirical measures gathered, and the analytical hierarchy process.
-This follows a general methodology for assessing the state of the practice for SCS domains.
-	\item Quantitative and qualitative \textbf{Results}. 
-	\item An analysis of the results and \textbf{Answers To Research Questions}.
-	\item \textbf{Conclusion} to the report including comments on suggestions for future state of the practice assessments.
-	\item \textbf{Appendix}: This section includes our Research Questions, the Measurement Template, Grading Template, Ethics Approval, and Developer Interview Questions.
-\end{itemize}
-
-\newpage
-
-\section{Domain Analysis - DONE SECTION 1st DRAFT}\label{domainanalysis}
-
-A domain analysis consists of systematically identifying and documenting the commonalities, variabilities, and parameters of variation of a software family \citep{Weiss1997}.
-A software family is defined by \cite{parnas1976design} as “a set of programs whose common properties are so extensive that it is advantageous to study the common properties of the programs before analyzing individual members”.
-
-We added this domain commonality and variability analysis to our original methodology because the first time we did the State of the Practice exercise we found we weren't comparing "apples to apples".
-Software packages can have considerable variation, even when the packages appear to be in the same software family.
-Viewing the ranking exercise as a means for selecting the right tool for a job, functionality is important.
-A 2D Lattice Boltzmann solver is not a substitute for a 3D Lattice Boltzmann solver when the latter is required.
-Therefore, this exercise seeks to classify the different software products based on their functionality.
-Nonfunctional qualities are assessed through other means in our process.
-For instance, through usability experiments as noted in the \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Methodology.pdf}{Methodology for Assessing the State of the Practice for Domain X} document.
-
-This domain analysis of the LBM SCS family is organized into the four subsection located below.
-The first subsection reviews the basics of Lattice Boltzmann systems.
-The next three subsections consist of lists of commonalities, variabilities, and parameters of variation, respectively.
-These three sections form the heart of the domain analysis and include an extensive set of cross-references to demonstrate the relationships between the different items.
-
-\subsection{Lattice Boltzmann Systems}
-
-LBM are a family of fluid dynamics algorithms for simulating single-phase and multiphase fluid flows, often incorporating additional physical complexities \citep{chen1998lattice}.
-They consider the behaviours of a collection of particles as a single unit at the mesoscopic scale, between the nanoscopic and microscopic scales.
-These methods predict the positional probability of a collection of particles moving through a lattice structure.
-Off-the-shelf (OTS) Lattice Boltzmann Solvers (LBS) allow for a range of fluid and physical model input parameters, computational parameters, and output parameters.
-
-LBS model fluid dynamics within a boundary using a predefined lattice structure and a two step calculation process.
-The first process is streaming, where the particles move along the lattice via links.
-The second process is collision, where energy and momentum is transferred among particles that collide \citep{bao2011lattice}.
-There are many standardized lattice models - individual solvers within the family might only use a subset of them.
-LBM use the initial parameters of the fluid to find the probability of where along the lattice linkages a group of particles are most likely to travel.
-It then moves the particles into the next node, and transfers the energy and momentum if a collision occurs.
-Then the process repeats for the duration of the modeling instance.
-
-\subsection{Commonalities}\label{comm_sec}
-
-This section lists common features among potential family members.
-The commonalities are organized using the following abstraction of the system, which can be used to describe all Lattice Boltzmann systems: input information, generate the simulation, output the results.
-Section \ref{comm_lbm} describes the commonalities for the simulation step.
-Section \ref{comm_in} highlights the input information that is required for Lattice Boltzmann systems.
-The next section, Section \ref{comm_out}, shows the common features for the output of Lattice Boltzmann systems, such as the requirement that mesh information be written to files.
-Although the output information could simply be written to the computer’s memory, in all practical applications it is desirable to have a persistent record of the output that was created.
-The final section covers nonfunctional requirements of the system.
-For instance, all systems will have the goal that the response time to a user’s request is small enough to allow the user to focus on his/her problem and to maintain his/her train of thought, without excessive waiting time.
-The commonality in this case is refined by a later variability because the specific requirement on the response time will depend on the individual software packages, their simulation models, inputs, and outputs.
-
-Each commonality below uses the same structure.
-All of the commonalities are assigned a unique item number, which takes the form of a natural number with the prefix ``C''.
-Following this, a description of the commonality is provided along with a list of related variabilities, which are given as hyperlinks that allow navigation of the document to the text describing the variability.
-
-\subsubsection{Lattice Boltzmann Method Solvers}\label{comm_lbm}
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
-		\hline
-		\bf Item Number& C\refstepcounter{comnum}\thecomnum{}\label{lattice} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
-		\hline
-		\bf Description & A lattice discretizes a computational domain into a finite number of points. All LBS discretize the computational domain using a regular, evenly spaced grid within a boundary.\\
-		\hline
-		\bf Related Variability & \vref{numveldir} \vref{varcolop} \vref{varboundary}\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
-		\hline
-		\bf Item Number& C\refstepcounter{comnum}\thecomnum\label{collisionoperator} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
-		\hline
-		\bf Description & All Lattice Boltzmann versions use a collision operator which concerns collisions between particles.
-Collision operators map collisions of particles within the lattice space.
-The Bhatnagar-Gross-Krook Collision Operator is a common LBM collision operator that preserves continuity for a discretized model, for each velocity direction $i$.
-Its equation is $\mathrm{\Omega_i} = \frac{1}{\tau}(f_{i}^{eq}-f_{i})$, where $\tau$ is the relaxation rate towards equilibrium, $f^{eq}$ is the equilibrium particle probability distribution function, and $f$ is the particle probability distribution function.\\
-		\hline
-		\bf Related Variability & \vref{dimensions} \vref{numveldir}\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
-		\hline
-		\bf Item Number& C\refstepcounter{comnum}\thecomnum\label{probabilitydensityfunc} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
-		\hline
-		\bf Description & All Lattice Boltzmann versions use a probability density function to give the probability that fluid has moved into a specific domain.\\
-		\hline
-		\bf Related Variability & \vref{dimensions}\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
-		\hline
-		\bf Item Number& C\refstepcounter{comnum}\thecomnum\label{equildistfunc} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
-		\hline
-		\bf Description & Every Lattice Boltzmann version uses an equilibrium distribution function to capture the probability distribution of particles.\\
-		\hline
-		\bf Related Variability & \vref{varedf} \vref{edfstorage} \vref{edfcoeff}\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
-		\hline
-		\bf Item Number& C\refstepcounter{comnum}\thecomnum\label{Boltzmanntransequ} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
-		\hline
-		\bf Description & Every Lattice Boltzmann version uses a Boltzmann transport equation to describe the statistical behaviour of a system that does not have collisions.
-The equation is $(\frac{d}{dt} + \mathrm{e} \cdot \nabla_\mathrm{x} + \frac{F}{m} \cdot \nabla_\mathrm{e})f = \Omega(t)$, where $\mathrm{e}$ is the velocity, $\nabla_\mathrm{x}$ is the position vector gradient, $\nabla_\mathrm{e}$ is the velocity gradient, and $\Omega(t)$ is the collision operator as a function of time.\\
-		\hline
-		\bf Related Variability & \vref{vartraneq}\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-\subsubsection{Input}\label{comm_in}
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
-		\hline
-		\bf Item Number& C\refstepcounter{comnum}\thecomnum\label{fluidandboundaryinfo} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
-		\hline
-		\bf Description & The LBS require fluid, model, and boundary information for the problem.
-This includes, but is not limited to, fluid acceleration rate, velocity, and viscosity, as well as the number of dimensions in the lattice model, and the number of velocity directions in the lattice.\\
-		\hline
-		\bf Related Variability & \vref{interface}\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-\subsubsection{Output}\label{comm_out}
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
-		\hline
-		\bf Item Number& C\refstepcounter{comnum}\thecomnum\label{predictionstomemory} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
-		\hline
-		\bf Description & LBS write fluid predictions, such as the output of the transport equation, to memory.\\
-		\hline
-		\bf Related Variability & \vref{varvisual} \vref{formatoutput} \\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-\subsection{Variabilities}\label{var_sec}
-This section provides a list of characteristics that may vary among family members.
-As in Section \ref{comm_sec}, the first three subsections on variabilities are organized into the following sublists: Simulation Models, Input and Output.
-The final two subsections list variabilities that can be characterized as system constraints and as nonfunctional requirements.
-
-As for the commonalities, each variability is labelled with a unique item number.
-In this case the numbers are prepended with the letter ``V''.
-The other three headings provided for each variability are: Description, Related Commonality, and Related Parameter.
-The related commonalities and parameters are given as a set of identifiers that respectively refer back to the previous section on commonalities or refer forward to the next section on parameters of variation.
-
-\subsubsection{Lattice Boltzmann Method Solvers}
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
-		\hline
-		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{parallel} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & LBS may use a framework for parallel processing of the model.\\
-		\hline
-		\bf Related Commonality & None\\
-		\hline
-		\bf Related Parameter & \pref{parparallel}\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-~\newline\\
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
-		\hline
-		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{varedf} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & Different versions of an equilibrium distribution function can capture the probability distribution of particles.\\
-		\hline
-		\bf Related Commonality & \cref{equildistfunc}\\
-		\hline
-		\bf Related Parameter & \pref{paredf}\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
-		\hline
-		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{edfstorage} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & Storage patters for distribution function can vary.\\
-		\hline
-		\bf Related Commonality & \cref{equildistfunc}\\
-		\hline
-		\bf Related Parameter & \pref{parstorage}\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
-		\hline
-		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{edfcoeff} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & Coefficients used with the distribution function can vary. These are based on the number of velocity directions in the model.\\
-		\hline
-		\bf Related Commonality & \cref{equildistfunc}\\
-		\hline
-		\bf Related Parameter & \pref{paredfcoeff}\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
-		\hline
-		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{dimensions} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & The number of dimensions in the lattice of the model can vary.\\
-		\hline
-		\bf Related Commonality & \cref{lattice} \cref{probabilitydensityfunc}\\
-		\hline
-		\bf Related Parameter & \pref{pardimensions}\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
-		\hline
-		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{numveldir} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline 
-		\bf Description & The number of velocity directions in the lattice of the model can vary.\\
-		\hline
-		\bf Related Commonality & \cref{lattice} \cref{collisionoperator}\\
-		\hline
-		\bf Related Parameter & \pref{parveldir}\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
-		\hline
-		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{varcolop} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & Various collision operators can be used.\\
-		\hline
-		\bf Related Commonality & \cref{lattice} \cref{collisionoperator}\\
-		\hline
-		\bf Related Parameter & \pref{parcollisop}\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
-		\hline
-		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{vartraneq} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & Various transport equations can be used to describe the statistical behaviour of the system\\
-		\hline
-		\bf Related Commonality & \cref{Boltzmanntransequ}\\
-		\hline
-		\bf Related Parameter & \pref{partransequ}\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
-		\hline
-		\bf Item Number&  V\refstepcounter{varnum}\thevarnum\label{numberoffluids} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & The number of fluids allowed in the LBS.\\
-		\hline
-		\bf Related Commonality & \cref{fluidandboundaryinfo}\\
-		\hline
-		\bf Related Parameter & \pref{parfluids}\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
-		\hline
-		\bf Item Number&  V\refstepcounter{varnum}\thevarnum\label{typefluidparm} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & The type of fluid parameters.\\
-		\hline
-		\bf Related Commonality & \cref{fluidandboundaryinfo}\\
-		\hline
-		\bf Related Parameter & \pref{partypeparm}\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
-		\hline
-		\bf Item Number&  V\refstepcounter{varnum}\thevarnum\label{varboundary} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & The boundary of the lattice can have various conditions.\\
-		\hline
-		\bf Related Commonality & \cref{lattice}\\
-		\hline
-		\bf Related Parameter & \pref{parboundary}\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-\subsubsection{Input}
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
-		\hline
-		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{interface} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & The input interface can vary between LBS.\\
-		\hline
-		\bf Related Commonality & \cref{fluidandboundaryinfo}\\
-		\hline
-		\bf Related Parameter & \pref{parinterface}\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-\subsubsection{Output}
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
-		\hline
-		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{varvisual} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & Visual presentation of the prediction.\\
-		\hline
-		\bf Related Commonality & \cref{predictionstomemory}\\
-		\hline
-		\bf Related Parameter & \pref{parvarvisual}\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
-		\hline
-		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{formatoutput} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & Format of prediction information.\\
-		\hline
-		\bf Related Commonality & \cref{predictionstomemory}\\
-		\hline
-		\bf Related Parameter & \pref{parformatoutput}\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-\subsubsection{System Constraints}
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
-		\hline
-		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{processing} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & Hardware that processes the calculations\\
-		\hline
-		\bf Related Commonality & None \\
-		\hline
-		\bf Related Parameter & \pref{parprocessing}\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
-		\hline
-		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{operatingsystem} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & Operating systems on which LBS run.\\
-		\hline
-		\bf Related Commonality & None \\
-		\hline
-		\bf Related Parameter & \pref{paropsystem}\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
-		\hline
-		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{storage} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & Amount of storage and memory needed for the LBS.\\
-		\hline
-		\bf Related Commonality & None \\
-		\hline
-		\bf Related Parameter & \pref{parstorageconst}\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-\newpage
-\subsection{Parameters of Variation}
-This section specifies the parameters of variation for the variabilities listed in Section \ref{var_sec}.
-They are organized into the same five subcategories as employed previously: Simulation Models, Input, Output, System Constraints, Nonfunctional Requirements.
-
-Each parameter of variation is given a unique identifier of the form “P” followed by a natural number.
-The corresponding variability is listed and a hyperlink is provided that allows navigation back to the appropriate item in Section \ref{var_sec}.
-The final entry for each parameter of variation is the binding time, which is the time in the software lifecycle when the variability is fixed.
-The binding time could be during specification, or during building
-of the system (compile time), or during execution of the system (run time).
-It is possible to have a mixture of binding times.
-For instance, a parameter of variation could have a binding time of “specification or building” to represent that the parameter could be set at specification time, or it could be postponed until the given family member is built.
-The choice of postponing the decision until the build could be associated with the presence of a domain specific language that would allow postponing decisions on the values of the parameter of variation.
-
-\subsubsection{Lattice Boltzmann Method Solvers}
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
-		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parparallel} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{parallel}\\
-		\hline
-		\bf Range of Parameters & OpenMP, OpenCL, CUDA, MPI are used if the execution of the LBS is parallelized.\\
-		\hline
-		\bf Binding Time & Build Time\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
-		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{paredf} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{varedf}\\
-		\hline
-		\bf Range of Parameters & Equilibrium approximation varies between incompressible or compressible models.\\		
-		\hline
-		\bf Binding Time & Build Time\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
-		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parstorage} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{edfstorage}\\
-		\hline
-		\bf Range of Parameters & Various data structures can be used to store function output, including single and multi-dimensional arrays, depending on the problem model and developer preferences.\\
-		\hline
-		\bf Binding Time & Build Time\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
-		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{paredfcoeff} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{edfcoeff}\\
-		\hline
-		\bf Range of Parameters & Numerous coefficients for equilibrium distribution function based on number of velocity directions.
-The number of velocity directions is typically 2, 3, 5, 9, 13, 15, 19, or 27.\\
-		\hline
-		\bf Binding Time & Build Time\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
-		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{pardimensions} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{dimensions}\\
-		\hline
-		\bf Range of Parameters & LBS model has 1, 2, or 3 dimensions.\\
-		\hline
-		\bf Binding Time & Build Time or Run Time\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
-		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parveldir} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{numveldir}\\
-		\hline
-		\bf Range of Parameters & One dimensional models include options of 2, 3, and 5 velocity directions.
-Two dimensional models include options of 9, 13, and 15 velocity directions.
-Three dimensional models include options of 15, 19, and 27 velocity directions.
-Figure \ref{dimension_model} from \citep{thurey2009stable} illustrates two commonly used LBM models in two and three dimensions.\\
-		\hline
-		\bf Binding Time & Build Time or Run Time\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-
-\begin{figure}[h!]
-	\begin{center}
-		\includegraphics[width=0.7\textwidth]{dimension_model}
-		\caption{Commonly used LBM models in two and three dimensions}
-		\label{dimension_model}
-	\end{center}
-\end{figure}
-
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
-		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parcollisop} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{varcolop}\\
-		\hline
-		\bf Range of Parameters & SRT, TRT, MRT, BGK collision operators.\\
-		\hline
-		\bf Binding Time & Build Time\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
-		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{partransequ} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{vartraneq}\\
-		\hline
-		\bf Range of Parameters & Collision and collision free transport equations.
-LBS sometimes use one or the other, often the collision transport equation.
-A Boolean parameter could be used to select between these equations in systems that can apply either equation.\\
-		\hline
-		\bf Binding Time & Build Time\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
-		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parfluids} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{numberoffluids}\\
-		\hline
-		\bf Range of Parameters & LBS can model a natural number of fluids.\\
-		\hline
-		\bf Binding Time & Build Time\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
-		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{partypeparm} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{typefluidparm}\\
-		\hline
-		\bf Range of Parameters & LBS fluid parameters include Reynolds Number, density, viscosity, time, pressure, force, direction, relaxation rate, turbulence.
-All of these parameters have the type of real number.\\
-		\hline
-		\bf Binding Time & Build Time\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
-		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parboundary} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{varboundary}\\
-		\hline
-		\bf Range of Parameters & Lattice boundary can have reflective or non-reflective conditions.
-Some LBS will only model reflective or non-reflective conditions.
-If there is a choice then this can be indicated by a Boolean parameter.\\
-		\hline
-		\bf Binding Time & Build Time\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-\subsubsection{Input}
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
-		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parinterface} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{interface}\\
-		\hline
-		\bf Range of Parameters & Input can be graphical, text or file.\\
-		\hline
-		\bf Binding Time & Build Time\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-\subsubsection{Output}
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
-		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parvarvisual} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{varvisual}\\
-		\hline
-		\bf Range of Parameters & LBS can provide 1D, 2D, and 3D rendering of  the model.\\
-		\hline
-		\bf Binding Time & Build Time or Run Time\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
-		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parformatoutput} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{formatoutput}\\
-		\hline
-		\bf Range of Parameters & LBS prediction information is output in either text or binary format.\\
-		\hline
-		\bf Binding Time & Build Time\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-\subsubsection{System Constraints}
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
-		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parprocessing} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{processing}\\
-		\hline
-		\bf Range of Parameters & The LBS model can be calculated on the CPU or GPU.\\
-		\hline
-		\bf Binding Time & Build Time\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
-		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{paropsystem} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{operatingsystem}\\
-		\hline
-		\bf Range of Parameters & LBS can be run on Windows, MacOS, or Linux versions.\\
-		\hline
-		\bf Binding Time & Build Time\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
-		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parstorageconst} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{storage}\\
-		\hline
-		\bf Range of Parameters & The amount of memory and storage varies between LBS.\\
-		\hline
-		\bf Binding Time & Run Time\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-
-\newpage
-\section{Methodology - DONE SECTION 1st DRAFT}
-
-In this project we set out to answer the research questions found in the Appendix in \ref{researchquestions} for the LBM SCS domain.
-The process involved systematically measuring and analyzing members of this software family along the quality attributes found in Section \ref{softwarequalities}.
-This included gathering quantitative and qualitative measurements.
-A goal of the project was to produce a quality assessment for LBM software packages.
-
-The methodology used in this LBM software assessment is a general state of the practice methodology that can be applied to any SCS domain.
-It was developed as an update to previous state of the practice exercises.
-This general methodology is linked in the next subsection.
-
-We collected quantitative data using the measures found in the measurement template of Section \ref{measurementtemplate}.
-Some of this was empirical software engineering related data, such as the number of files, number of lines of code, percentage of issues that are closed, etc.
-Most of the data was gathered by manually investigating the software, its source code, and its artifacts, while some was gathered using the empirical measurement tools discussed in Section \ref{empiricalmeasures}.
-
-We also collected qualitative data by interviewing the software package developers and asking them the questions found in the Appendix in \ref{interviewquestions}.
-Furthermore, we solicited the assistance of domain experts to better assess each software package by leveraging their experience to assess the functional and non-functional requirements for the software domain.
-
-This section begins by describing the steps of the overall process we used to select, measure and compare LBM software.
-This is followed by quality definitions and how these qualities were assessed in our project.
-The rest of the section provides an overview of how candidate software packages were selected and filtered, the empirical measurements and software tools that were used, and the analytic hierarchy process (AHP).
-
-\subsection{Process}
-
- The following steps provide an overview of how the assessment was conducted:
-
-\begin{enumerate}
-	\item List candidate software packages for the domain. This is discussed in Section \ref{identifysoftware}.
-	\item Filter the software package list. This is discussed in Section \ref{filtersoftware}.
-	\item Gather the source code and documentation for each software package.
-	\item Collect empirical measures. This is discussed in Section \ref{empiricalmeasures}.
-	\item Measure using the measurement template. This is discussed in Section \ref{empiricalmeasures}. The measurement template can be found in the Appendix in \ref{measurementtemplate2}.
-	\item Survey the developers.
-The developers of each software package in the filtered software list were contacted for voluntary interviews.
-The interview questions can be found in the Appendix in \ref{interviewquestions}.
-	\item Use AHP to rank the software packages. This is discussed in Section \ref{AHP}.
-	\item Analyze the results and answer the research questions. The answers can be found in Section \ref{answersquestions}.
-\end{enumerate}
-
-These steps are further detailed in the \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Methodology.pdf}{Methodology for Assessing the State of the Practice for Domain X} document.
-
-\subsection{Software Qualities}\label{softwarequalities}
-
-Software quality attributes facilitate the measurement and comparison of software packages in this state of the practice assessment.
-We adopt software quality definitions from various researchers and subject matter expert entities.
-Some of the definitions are from \cite{Smithetal2020}.
-The quality measurement results are found in Section \ref{qualityresults} of this report.
-Section \ref{qualityrecommentations} lists recommendations to address software qualities in LBM software packages.
-The following are the software quality definitions used in this state of the practice exercise, along with comments regarding their quantitative and qualitative measurement.
-
-\subsubsection{Installability}
-
-Installability is measured by the effort required for the installation, uninstallation or reinstallation of a software product in a specified
-environment \citep{ISO/IEC25010} \citep{lenhard2013measuring}.
-A good measure of installability correlates with scenarios when low or moderate effort is required to gather and prepare software for its general use on a system for which it was designed.
-In this case effort includes the time spent finding and understanding the installation instructions, the man-time and resources spent performing the installation procedure, and the absence or ease of overcoming system compatibility issues.
-The ability to reasonably validate the installation procedure also has a positive effect on the measure of installability.
-Similarly, the ease of uninstallation has an affect on the measure of installability.
-
-\subsubsection{Correctness}
-
- A software program is correct if it behaves according to its stated
-specifications \citep{GhezziEtAl2003}.
-This requires that the specification is available.
-Software is unlikely to have a formal specification if it is not developed by seasoned or professional software developers.
-Since some software does not have a specification available, the correctness of software cannot always be verified.
-Despite an absent specification, the correctness of the output of scientific computing software can sometimes be manually verified by applying domain knowledge.
-A good measure of correctness correlates with the availability of a requirements specification and reference to domain theory, as well as the explicit use of tools or techniques for building confidence of correctness, such as documentation generators and software analysis tools.
-
-\subsubsection{Verifiability}
-
-Verifiability is measured by the extent to which a set of tests can be written and executed to demonstrate that the delivered system meets the specification \citep{sommerville}.
-Similarly to correctness, verifiability is correlated with the availability of a specification and with reference to domain knowledge.
-A good measure of verifiability is further correlated with the availability of well written tutorials that include expected output, with software unit testing documentation, and with evidence of continuous integration during the development process.
-
-\subsubsection{Reliability}
-
-Reliability is measured by the probability of failure-free operation of a computer program in a specified environment for a specified time, i.e.
-the average time interval between two failures also known as the mean time to failure (MTTF) \citep{GhezziEtAl2003} \citep{musa1987software}.
-Reliability is thus positively correlated with the absence of errors during installation and use.
-Recoverability from errors also improves reliability.
-
-\subsubsection{Robustness}
-
-Software possesses the characteristic of robustness if it behaves “reasonably” in two situations: i) when it encounters circumstances not anticipated in the requirements specification; and ii) when the assumptions in its requirements specification are violated \citep{boehm2007software} \citep{ghezzi1991fundamentals}.
-A good measure of robustness correlates with a reasonable reaction to unexpected input, including data of the wrong type, empty input, or missing files or links.
-A reasonable reaction includes an appropriate error message and the ability to recover the system.
-
-\subsubsection{Performance}
-
-Performance is measured by the degree to which a system or component accomplishes its designated functions within given constraints, such as speed (database response times, for instance), throughput (transactions per second), capacity (concurrent usage loads), and timing (hard real-time demands) \citep{IEEEStdGlossarySET1990} \citep{wiegers2003softreq}.
-In this state of the practice assessment performance was not quantitatively measured.
-Instead the documentation of each software package was observed for information that alludes to a consideration of performance, such as parallelization tools.
-
-\subsubsection{Usability}
-
-Usability is measured by the extent to which a software product can be used by specified users to achieve specified goals with effectiveness, efficiency, and satisfaction in a specified context of use \citep{nielsonusability}.
-A good measure of usability correlates with the presence of documentation, including tutorials, manuals, and defined user characteristics, and user support.
-Preferably the user support model has avenues to contact developers and report issues.
-
-\subsubsection{Maintainability}
-
-A measure of maintainability is the effort with which a software
-system or component can be modified to correct faults, improve performance or other attributes, and satisfy new requirements \citep{IEEEStdGlossarySET1990} \citep{boehm2007software}.
-In this state of the practice analysis maintainability is measured by the quality of documentation artifacts, and the presence of version control and issue tracking.
-These artifacts can greatly decrease the effort needed to modify software.
-There are many documentation artifacts that can improve maintainability, including user and developer manuals, specifications, README files, change logs, release notes, publications, forums, and instructional websites.
-
-\subsubsection{Modifiability}
-
-Modifiability refers to the ease with which stable changes can be made to a system and the flexibility of the system to adopt such changes \citep{8016712}.
-This state of the practice assessment did not quantitatively measure modifiability.
-Developers were asked in interviews if they considered the ease of future changes when developing the software packages, specifically changes to the structure of the system, modules and code blocks.
-A follow up question asked if any measures had been taken.
-
-\subsubsection{Reusability}
-
-Reusability refers to the extent to which components of a software package can be used with or without adaptation in other software packages \citep{kalagiakos2003non}.
-A good measure of reusability results from a large number of easily reusable components.
-Increased software modularization, defined as the presence of smaller components with well defined interfaces, is important.
-For this state of the practice assessment, a good measure of reusablity correlates with an increased number of code files, and the availability of API documentation.
-
-\subsubsection{Understandability}
-
-Understandability is measured by the capability of the software package to enable the user to understand its suitability and function \citep{ISO9126}.
-It is an artifact-dependent quality.
-Understandability is different for the user-interface, source code, and documentation.
-In this state of the practice analysis, understandability focuses on the source code.
-It is measured by the consistency of a formatting style, the extend of modularization, the explicit identification of coding standards, the presence of meaningful identifiers, and clarity of comments.
-
-\subsubsection{Traceability}
-
-Traceability refers to the ability to link the software implementation and the software artifacts, especially the requirement specification \citep{McCallEtAl1977}.
-Similar to the quality of correctness, this requires some form of specification to be available.
-The quality refers to keeping track of information as it changes forms or relates between artifacts.
-This state of the practice assessment did not quantitatively measure traceability.
-Developers were asked in interviews how documentation fits into their development process.
-
-\subsubsection{Visibility and Transparency}
-
-Visibility and transparency refer to the extent to which all of the steps of a software development process, and the current status of it, are conveyed clearly \citep{ghezzi1991fundamentals}.
-In this state of the practice assessment a good measure of visibility and transparency correlates with a well defined development process, the presence of development process and environment documentation, and software package version release notes.
-
-\subsubsection{Reproducibility}
-
-Software achieves reproducibility if another developer can take the requirements documentation and re-obtain the same software artifacts \citep{BenureauAndRougier2017}.
-This includes the output of the software, where the scientific results are compared between software implementations, or between software implementations and manually calculated results.
-This state of the practice assessment did not quantitatively measure reproducibility.
-Developers were asked in interviews if they have any concern that their computational results won't be reproducible in the future, and if they have taken any steps to ensure reproducibility.
-
-\subsubsection{Unambiguity}
-
-Unambiguity refers to the extent to which two readers have similar interpretations when reading software artifacts.
-In other words, artifacts are unambiguous if, and only if, they only have one interpretation \citep{IEEE1998}.
-This state of the practice assessment did not quantitatively measure unambiguity.
-Developers were asked in interviews if they think that the current documentation can clearly convey all necessary knowledge to the users, and how they achieved this or what improvements are needed to achieve it.
-
-\subsection{Identify Candidate Software}\label{identifysoftware}
-
-The candidate software was found through search engine queries targeting authoritative lists of software.
-We found LBM software listed on the websites GitHub and swMATH, as well as through articles found in scholarly journals and databases.
-
-The following properties were considered when creating the list and reviewing the candidate software:
-
-\begin{enumerate}
-	\item The software functionality must fall within the identified domain.
-	\item The source code must be viewable.
-	\item The empirical measures should be available, which implies a preference for GitHub-style repositories.
-	\item The software cannot be marked as incomplete or in an initial development phase.
-\end{enumerate}
-
-The initial list had 45 packages, including a few packages that were later found to not have publicly available source code, or to be in an incomplete state of development.
-
-\subsection{Filter the Software List}\label{filtersoftware}
-
-To reduce the number of members in the candidate software list to a manageable size, the following filters were applied. The filters were applied in the priority order listed.
-
-\begin{enumerate}
-	\item Scope: Software is removed by narrowing what functionality is considered to be within the scope of the domain.
-	\item Usage: Software packages were eliminated if their installation procedure was missing or not clear and easy to follow.
-	\item Age: The older software packages (age being measured by the last date when a change was made) were eliminated, except in the cases where an older software package appears to be highly recommended and currently in use.
-\end{enumerate}
-
-For the third item in the above filter, software packages were characterized as `alive' if their related documentation had been updated within
-the last 18 month. Packages were categorized as `dead' if the last update of this information was more than 18 month ago.\\
-
-\begin{center}
-	\begin{tabular}{ p{4cm}p{3cm}p{3cm} }
-		\hline
-		Name & Released & Updated\\
-		\hline
-		DL\_MESO (LBE) & unclear & 2020 Mar\\
-		ESPResSo & 2010 Nov & 2020 Jun\\
-		ESPResSo++ & 2011 Feb & 2020 Apr\\
-		lbmpy& unknown  & 2020 Jun  \\
-		lettuce & 2019 May & 2020 Jul\\
-		Ludwig& 2018 Aug & 2020 Jul\\
-		LUMA& 2016 Nov   & 2020 Feb\\
-		MechSys & 2008 Jun    & 2020 Jul\\
-		OpenLB & 2007 Jul & 2019 Oct\\
-		Palabos & unclear & 2020 Jul\\
-		pyLBM & 2015 Jun&   2020 Jun\\
-		Sailfish & 2012 Nov & 2019 Jun\\
-		TCLB & 2013 Jun  & 2020 Apr\\
-		waLBerla & 2008 Aug & 2020 Jul\\
-		\hline
-	\end{tabular}
-	\captionof{table}{Alive Software Packages}\label{alivepackages}
-\end{center}
-
-While the initial list had 45 packages, filtering by scope, usage, and age decreased the size of the list to 23 packages.
-Many of the 22 packages that were removed could not be tested as there was no installation guide, they were incomplete, source code was not publicly available, a license was needed, or the project was out of scope or not up to a standard that would support incorporating them into this study.
-These eliminated software packages are listed in the Appendix in Section \ref{eliminatedpackagessection}.
-Of the remaining 23 packages that were studied, some were kept on the list despite being marked as dead due to their prevalence on authoritative lists on LBM software and due to their surface excellence, specifically the considerable time that was put into these projects.
-
-The final list of software packages that were analyzed in this project can be found in the following two tables.
-Table \ref{alivepackages} lists packages that fell into the `alive' category as of mid 2020, and Table \ref{deadpackages} lists packages that were `dead' at that time.
-\\
-
-\begin{center}
-	\begin{tabular}{ p{4cm}p{3cm}p{3cm} }
-		\hline
-		Name & Released & Updated\\
-		\hline
-		HemeLB & 2007 Jun & 2018 Aug\\
-		laboetie & 2014 Nov & 2018 Aug\\		
-		LatBo.jl & 2014 Aug & 2017 Feb\\
-		LIMBES & 2010 Nov & 2014 Dec\\
-		LB3D & unclear & 2012 Mar\\
-		LB2D-Prime & 2005 & 2012 Apr\\
-		LB3D-Prime & 2005 & 2011 Oct\\
-		MP-LABS & 2008 Jun & 2014 Oct\\
-		SunlightLB & 2005 Sep & 2012 Nov\\
-		\hline
-	\end{tabular}
-	\captionof{table}{Dead Software Packages}\label{deadpackages}
-\end{center}
-
-There is considerable variation among these software packages, including their intended purpose, size, user interfaces, and software languages used.
-For example, the OpenLB software package is predominantly a C++ package that makes use of hybrid parallelization and was designed to address a range of computational fluid dynamics problems \citep{heuveline2009towards}.
-The software package pyLBM is an all-in-one Python language package for numerical simulations \citep{graille2017pylbm}.
-ESPResSo is an extensible simulation package that is specifically for research on soft matter, and is written in C++ and Python \citep{weik2019espresso}.
-The HemeLB package is used for efficient simulation of fluid flow in several medical domains, and is written predominantly in C, C++, and Python \citep{mazzeo2008hemelb}.
-
-\subsection{Empirical Measures}\label{empiricalmeasures}
-
-The quality measurements in this assessment rely on the gathering and analyzing of raw and processed empirical data related to the research questions listed in the Appendix in \ref{researchquestions}.
-
-All of the quality measurements that are part of the AHP analysis are empirical measurements.
-Qualitative data gathered during interviews with developers is not part of the AHP analysis.
-This part of the assessment focuses on data that is reasonably easy to collect.
-Much of the data was gathered by manually reviewing the artifacts of each software package, while some was gathered using the freeware tools discussed below.
-The data that was gathered is listed in the measurement template found in the Appendix in \ref{measurementtemplate2}.
-Some of the data required processing other data within the template, including the status of the software package, which relied on the last commit date; the percentage of issues that are closed, which relied on the number of open and closed issues; and the percentage of code that is comments, which relied on the number of total lines and comment lines in text-based files.
-The complete measurement template data was then analyzed using the grading template found in the Appendix in \ref{gradingtemplate2}, the output of which was analyzed using the AHP described in Section \ref{AHP}.
-
-Most of the measurement template data was gathered by observing GitHub repository metrics and software package artifacts, or by processing data gathered using freeware tools.
-Data in the final three sets of the measurement template was collected using these tools.
-The tool \href{https://github.com/tomgi/git_stats}{GitStats} was used to measure each software package's GitHub repository for the number of binary files, the number of added and deleted lines, and the number of commits over varying time intervals.
-The tool \href{https://github.com/boyter/scc}{Sloc Cloc and Code (scc)} was used to measure the number of text based files as well as the number of total, code, comment, and blank lines in each GitHub repository.
-Details on installing and running these tools can be found in the \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/A
-	Guide to Empirical Measures.pdf} {Guide to Empirical Measures} file in the AIMSS repository.
-
-\subsection{Analytical Hierarchy Process}\label{AHP}
-
-The Analytical Hierarchy Process (AHP) is a decision-making technique that is used to compare multiple options by multiple criteria.
-In our work AHP was used for comparing and ranking the LBM software packages using the overall impression quality scores that were gathered in the measurement template found in the Appendix in \ref{measurementtemplate2} using the grading template found in the Appendix in \ref{gradingtemplate2}.
-AHP performs a pairwise analysis using a matrix and generates an overall score as well as individual quality scores for each software package.
-\cite{SmithEtAl2016} shows how AHP is applied to ranking software based on quality measures.
-
-This project used a tool for conducting this process.
-The tool includes a sensitivity analysis that was used to ensure that the software package rankings are appropriate with respect to the uncertainty of the quality scores.
-The \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/AHP2020/LBM/README.txt}{README} file of the tool includes requirements and usage information.
-
-\newpage
-\section{Results - DONE SECTION 1st DRAFT}\label{qualityresults}
-
-This section presents the quantitative and qualitative findings of this study.
-The qualitative data that is presented was gathered through interviews with some developers.
-AHP results for qualities that were quantitatively measured are also found in this section.
-Quality definitions are found in section \ref{softwarequalities}.
-
-\subsection{Installability}
-
-All of the software packages that were tested have installation instructions.
-As noted previously, many of the twenty three software packages that were part of the original long list of forty six software packages but were later removed do not include documentation or installation instructions.
-
-All twenty three packages on the short list can be installed on some Unix-like systems.
-Seven packages could be installed on Windows, and five on macOS.
-Operating system compatibility is found in the documentation of nineteen software packages.
-All but one of the software packages were tested on Ubuntu for this state of the practice assessment.
-The remaining package was tested on CentOS.
-
-Of the software packages that were tested, most installation instructions are located in one place, often in an instruction manual or on a web-page.
-Sometimes incomplete installation instructions are found on a home page, with more detailed instructions located on another web-page or within formal documentation.
-Maintainability and correctness of these instructions could be improved if they were limited to one location.
-
-All but two of the software packages have automated some of the installation process. Most of these packages use Make to automate the installation, and a few of them use custom scripts.
-
-Errors encountered during the installation process were often quickly fixed thanks to descriptive error messages.
-Systems that provided vague error messages were more difficult to troubleshoot.
-Only three software packages that displayed a descriptive error message were not recoverable, and most of these instances were due to hardware and operating system incompatibility, such as the requirement of CUDA.
-Fourteen software packages definitively broke during installation.
-A couple of packages did not provide a definitive message of the success or failure of installation.
-In these instances, validating the installation required performing a tutorial or running a script, as described below, if these were available.
-
-About half of the installation instructions are not written as if the person doing the installation has none of the dependent packages installed, and many software packages do not list all of their dependencies.
-Many dependencies needed to be manually installed.
-Sometimes only an error message during the installation process informs the user of the requirement of these packages.
-A detailed rewrite of the installation instructions from the point of view of installation on a clean operating system is suggested.
-
-Sixteen software packages require less than ten dependencies to be installed.
-All but one software package require less than twenty dependencies.
-Some packages may automatically install additional dependencies in the background.
-Eighteen of the software packages do not explicitly indicate software dependency versions.
-Some software package installation issues, specifically those occurring when manual installation of dependencies is required, may be avoided if versions of dependencies are specified.
-Fifteen software packages do not have detailed instructions for installing dependencies.
-
-Sixteen software packages have less than ten manual installation steps.
-If dependencies are installed in one command then none of the software packages take more than twenty steps to install.
-The average number of steps is about eight, and the fewest is three.
-
-All but six of the software packages have a way to verify the installation.
-Twelve have some sort of tutorial examples that can be run by the user.
-Most of these tutorials provide expected output.
-Some packages do not have a way to verify the output of the tutorials, and the user is left to assume correctness if no error message is encountered.
-Some other ways of installation validation include validation scripts, automatic validation after the installation, and instructions to manually review the file system.
-
-Uninstallation instructions were found for only one of the software packages.
-
-\begin{figure}[h!]
-	\begin{center}
-		\includegraphics[width=1.0\textwidth]{installability_chart}
-		\caption{AHP Installability Score}
-		\label{Fig_Installability}
-	\end{center}
-\end{figure}
-
-Figure \ref{Fig_Installability} shows the installability ranking of the software packages using AHP.
-Software packages with a higher score tend to have one set of linear installation instructions that are written as if the person doing the installation has none of the dependencies installed.
-The instructions often list compatible operating system versions and include instructions for the installation of dependencies.
-The top ranked packages often incorporate some sort of automation of the installation process and have fewer manual installation steps.
-The number of dependencies a package has does not correlate with a higher score.
-The ability to validate the installation process, often through tutorials or test examples that include expected output, is correlated with a higher score.
-
-Many software packages would benefit from a rewrite or reorganization of installation instructions.
-A single location for installation instructions would improve their maintainability and correctness.
-Listing compatible operating system and dependency versions would decrease installation time and errors, as would adding instructions on installing dependencies.
-Installation process errors should prompt the system to display detailed messages.
-Once a software package is installed, either an automatic validation needs to be performed or the user needs to be able to perform a manual validation using test examples that include expected output.
-Finally, uninstallation instructions should be included in the documentation.
-
-\subsection{Surface Correctness and Verifiability}
-
-Sixteen of the software packages include a requirements specification artifact or explicitly reference domain theory.
-This artifact is often found within a user manual, on a web-page, or is mentioned in related publications.
-In the latter case the user may need to spend significant time to find this information.
-
-Interviews with developers confirmed that many of the software packages are developed by domain experts with backgrounds in physics, mathematics and mechanical engineering.
-Many of these developers do not have formal software engineering education.
-Some of the development teams include computer scientists.
-Despite a lack of visible domain documentation and a resulting lower surface correctness and verifiability score, many software packages do have a strong domain background on account of their development team.
-
-Document generation tools are explicitly used by twelve software packages. Sphinx is used by eight of them, and Doxygen is used by seven. Several of the packages use both.
-
-Tutorials are available for all but five of the software packages.
-Generally they are linearly written and easy to follow.
-However, only eight tutorials provide an expected output.
-It is not possible to verify the correctness of the output of the software packages that are missing this key information.
-In these cases the user may need to assume correctness if there are no visible errors.
-
-Unit tests are only explicitly available for one of the software packages.
-Code modularization of most packages allow for users to create tests with varying degrees of effort.
-These tests allow developers and users to verify the correctness of fragments of the source code, and in doing so better assess the correctness of the entire package.
-
-The documentation or presence of continuous integration tools and techniques alludes to a more refined development process where faults are isolated and better recognized.
-Only two of the packages mentioned applying the practice of continuous integration in their development process.
-
-Interviews with the developers of software packages suggested a more frequent use of both unit testing and continuous integration in the development processes than initially surveyed.
-For example, OpenLB, pyLBM, and TCLB use such methods during development despite it not being explicitly clear from a surface analysis.
-
-\begin{figure}[h!]
-	\begin{center}
-		\includegraphics[width=1.0\textwidth]{correctnessverifiability_chart}
-		\caption{AHP Surface Correctness and Verifiability Score}
-		\label{Fig_CorrectnessVerifiability}
-	\end{center}
-\end{figure}
-
-Figure \ref{Fig_CorrectnessVerifiability} shows the surface correctness and verifiability ranking of the software packages using AHP.
-Software packages with a higher score tend to have a visible requirements specification or references to theory documentation.
-They also explicitly use at least one document generation tool that builds confidence of correctness.
-The top ranked software packages all include an easy to follow getting started tutorial, and most of these include expected output.
-Only the top ranked package, Ludwig, provided unit testing.
-It and the second ranked package, ESPResSo, explicitly incorporated continuous integration in the development process.
-During developer interviews it was discovered that some other packages also use continuous integration and unit testing, despite not being explicitly stated in documentation.
-The correctness and verificablity of such packages is not measured well using surface analysis.
-
-The inclusion of requirements specification and theory documentation greatly benefits the correctness and verifiability of software packages.
-The use of document generation tools can help build confidence in correctness.
-The addition of easy to follow tutorials further helps users verify the software and have confidence in its correctness.
-Unit testing documentation and capability, as well as the use of continuous integration tools and techniques such as Bamboo, Jenkins, and Travis CI, help verify correctness.
-Several interviewed developers alluded to difficulty with testing large numbers of features, and some even manually tested program output.
-The use of well defined unit testing tools could decrease the time spent testing some feature.
-
-\subsection{Surface Reliability}
-
-Errors occurred when installing sixteen of the software packages.
-Every instance prompted an error message.
-These messages indicated unrecognized commands (even when following the installation guide), missing links, missing dependencies, syntax errors in code files.
-In some instances the error messages were vague.
-Several automatic installation processes could not find and load dependencies.
-In these instances the installation tried to access outdated external repositories.
-Seven of the installations were recovered and verified, and one of the installations was assumed to be recovered due to the absence of any way to verify it.
-The installation of eight of the software packages could not be recovered.
-Most of these broken installations could not find external dependencies, encountered system incompatibilities, or displayed vague error messages.
-
-Of the thirteen software packages that installed correctly and also have tutorials, four broke during tutorial testing.
-All of these instances resulted in an error message being displayed.
-One error was due to a missing tutorial dependency, another was due to an invalid command despite following the tutorial, and the final two errors were vague execution errors.
-Of the four broken tutorial instances, only the one that was missing a dependency was recoverable.
-
-\begin{figure}[h!]
-	\begin{center}
-		\includegraphics[width=1.0\textwidth]{reliability_chart}
-		\caption{AHP Surface Reliability Score}
-		\label{Fig_Reliability}
-	\end{center}
-\end{figure}
-
-Figure \ref{Fig_Reliability} shows the surface reliability ranking of the software packages using AHP.
-Software packages with a high score either did not break during installation, or the broken installation was recoverable.
-All of the top five ranked packages have tutorials.
-One of these packages, pyLBM, broke during tutorial testing, but a descriptive error message helped in recovery.
-
-Overall, lower ranked software packages are lacking clear documentation, testing or tutorial examples, and descriptive error messages, and have broken dependencies.
-Thus, regarding surface reliability, software packages would benefit from clear up-to-date installation and tutorial documentation, and descriptive error messages.
-
-\subsection{Surface Robustness}
-
-The software packages were tested for handling unexpected input, including incorrect data types, empty input, and missing files or links.
-Success predicated on a reasonable response from the system, including appropriate error messages and an absence of unrecoverable system failures.
-
-\begin{figure}[h!]
-	\begin{center}
-		\includegraphics[width=1.0\textwidth]{robustness_chart}
-		\caption{AHP Surface Robustness Score}
-		\label{Fig_Robustness}
-	\end{center}
-\end{figure}
-
-Figure \ref{Fig_Robustness} shows the surface robustness ranking of the software packages using AHP.
-Software packages with a high score behaved reasonably in response to unexpected input as described above.
-All of the software packages that installed correctly passed this test.
-They output descriptive error messages or did not crash.
-Software packages with a lower surface robustness score had not installed correctly, so their robustness score may not be a true reflection of runtime robustness.
-Similarly, all software packages that installed correctly and require plain text input files correctly handled an unexpected change to these input files, including a replacement of new lines with carriage returns.
-
-\subsection{Surface Performance}
-
-The software packages were not quantitatively measured for surface performance.
-Although the software packages all apply LBM to solve scientific computing problems, the packages focus on varied domains, with varying parameters, and are technically different from each other.
-Due to this, a comparison of performance is not appropriate.
-In this project we instead looked through each software package's artifacts for evidence that performance was considered.
-The artifacts of seventeen software packages mentioned the use of some sort of parallelization tool or technique.
-This included GPU processing and the CUDA parallel computing platform.
-GPUs provide superior processing power and speed compared to CPUs, and are often used for scientific computing when a large amount of data is involved.
-The software package TCLB is implemented in a highly efficient multi-GPU code to achieve performance suitable for model optimization \citep{rutkowski2020open}.
-In the Ludwig package, a so-called mixed mode approach is used where fine-grained parallelism is implemented on the GPU, and MPI is used for even larger scale parallelism \citep{gray2013ludwig}.
-While some software packages require CUDA and GPU processing, some have the option of using either the GPU or the CPU.
-The packages that require GPU and CUDA have better performance at the expense of installability and surface reliability.
-
-\subsection{Surface Usability}
-
-Software package artifacts were reviewed for the presence of a tutorial, a user manual, documented user characteristics, and a user support model.
-In total eighteen software packages have a tutorial and thirteen have a user manual.
-The tutorials vary in scope and substance, and eight include an expected output.
-Most user manuals are in the form of a file that can be downloaded, while some are rendered on a web-page.
-Some packages do not have a user manual but do have useful documentation distributed on their web-pages.
-Expected user characteristics are documented in four software packages.
-Users are typically scientists or engineers trying to test a new application.
-Their background is often physics, chemistry, biophysics, or mathematics.
-All but one of the packages have a user support model, and several of them have multiple avenues of user support.
-The most popular avenue of support is Git, followed by email and forums.
-One software package has a FAQ page.
-
-\begin{figure}[h!]
-	\begin{center}
-		\includegraphics[width=1.0\textwidth]{usability_chart}
-		\caption{AHP Surface Usability Score}
-		\label{Fig_Usability}
-	\end{center}
-\end{figure}
-
-Figure \ref{Fig_Usability} shows the surface usability ranking of the software packages using AHP.
-Software packages with a high score have a tutorial and user manual, sometimes have documented user characteristics, and have at least one user support model.
-Many packages have several user models.
-
-Interviews with developers revealed several usability issues.
-Some users have misunderstood the boundaries of LBM and computational fluid dynamics, and have combined or applied methods that are not physically sound.
-Sometimes users have applied LBM to poorly defined or inappropriate fluid dynamics problems.
-The users do not realize the limitations of the methods, of the software, and do not understand the requirements to properly model a problem with the software.
-The software packages are not designed to be used out of the box in a plug and play fashion.
-Developers of some of the software packages mitigated this by editing the source code to prevent unsound combination from being applied, and by updating the documentation to better inform users of LBM limitations, and of the requirements to properly model appropriate problems, including what algorithms and parameters to use.
-
-Some additional but infrequent software usability issues were commented on by the developers.
-Users have had trouble with installation and understanding how to maneuver the interfaces and how to set up or run models.
-These issues are addressed by various user support models, including frequently asked questions sections on the software package websites, user guides, and hardware and software requirements specifications.
-
-One software package changed its scripting language to Python to make it more usable.
-The developer commented that this was the biggest step in addressing usability, commenting that Python was a much easier language for users to learn and understand.
-
-\subsection{Maintainability}
-
-Software packages were reviewed for the presence of artifacts.
-Every type of artifact or file that is not a code file was recorded.
-The software packages were also reviewed for software release and documentation version numbers.
-This information could be used to better troubleshoot issues and organize documentation.
-All but three software packages have source code release and documentation version numbers.
-
-Information on how code is reviewed, or how to contribute to the project was also noted.
-In total, eleven software packages have this information, which was found in various artifacts, including in developer guides, contributor guides, user guides, developer web-pages, and README files.
-
-Issue tracking is used in twenty two software packages, fifteen of which use Git, six use email, and one uses SourceForge.
-Most software packages that use Git have most of their issues closed, and only three have less than fifty percent of their issues closed.
-Furthermore, most packages that use Git for issue tracking use GitHub as a version control system, one package uses CVS for issue tracking, and eight packages do not appear to use any such system.
-
-Software package code files were further measured for the percentage of code that is comments.
-Packages with a higher percentage of comments were designated as more maintainable.
-Comments represent more than ten percent of code files in fifteen packages, and the average percentage of code comments is about fourteen.
-
-\begin{figure}[h!]
-	\begin{center}
-		\includegraphics[width=1.0\textwidth]{maintainability_chart}
-		\caption{AHP Maintainability Score}
-		\label{Fig_Maintainability}
-	\end{center}
-\end{figure}
-
-Figure \ref{Fig_Maintainability} shows the maintainability ranking of the software packages using AHP.
-Software packages with a high score provide version numbers on documents and source code releases, have an abundance of high quality artifacts, and use an issue tracking tool and version control system.
-These packages also appear to reasonably handle tracked issues, having most of their issues closed.
-Their code files are well commented with more than ten percent of the code being comments.
-
-Interviews with developers revealed that most projects do have a defined process for accepting contributions from team members.
-The packages rarely get contributions from outside developers, but the process would be similar as for the aforementioned group.
-
-Contributions are made through GitHub, and are then reviewed and pulled by lead developers, often with consultation with a group of core developers depending on the organizational model.
-Continuous integration is part of the process for some packages.
-
-Some developers noted that their software package does not have well defined contributing guide in the repository but it might be a good idea to add one in the near future.
-They would be very happy to see contribution from outside of their organization, but currently this does not happen.
-
-Furthermore, maintainability has been addressed by increasing source code modularity, reducing duplicate information, and improving abstraction.
-Several software packages have had large sections of their code base redeveloped with languages that the developers felt are more understandable and readable, and that are better supported.
-Data structures have also been redeveloped and storage has been improved.
-One software developer mentioned that the mathematical foundations that the geometries and models are designed from had been re-evaluated and optimized.
-
-\subsection{Modifiability}
-
-Software packages were not quantitatively measured for modifiability.
-In this project we asked developers to comment on modifiability when we interviewed them.
-Specifically, we asked if ease of future changes to the system, modules, and code blocks was considered when designing the software.
-We also asked if any measures had been taken to ensure the ease of future changes.
-All of the developers that were interviewed noted that the ease of future changes was considered and that measures to ensure it had been taken.
-
-A high degree of code modularity and abstraction was noted by one developer as a measure to ensure the ease of future changes.
-They also noted that some of the code base was transitioned from c to c++, which could ease modifiability of that software package.
-Another developer noted that their software package was designed to easily allow for the addition of some LBM features, but changes to major aspects of the LBM models would be difficult.
-Furthermore, the package was designed with flexible data structures and storage in mind.
-
-Some software packages, like Palabos, provide validation benchmarks for their core fundamental algorithmic ingredients \citep{latt2021palabos}.
-The stated intent of these benchmarks is to showcase the validity and usefulness of the package in order to stimulate the development of third-user extensions.
-The Palabos package identifies as a development framework for modeling problems in various computational fluid dynamics areas.
-
-\subsection{Reusability}
-
-Each software package was measured for the total number of source code files.
-A larger number of source files was associated with increased reusability due to increased modularization.
-The software packages were also reviewed for the presence of API documentation, which indicates that a software package was developed with interaction between other software applications in mind.
-
-\begin{figure}[h!]
-	\begin{center}
-		\includegraphics[width=1.0\textwidth]{reusability_chart}
-		\caption{AHP Reusability Score}
-		\label{Fig_Reusabilty}
-	\end{center}
-\end{figure}
-
-Figure \ref{Fig_Reusabilty} shows the reusability ranking of the software packages using AHP.
-Software packages with a high score have thousands of source code files and API documentation.
-The highest scoring packages, ESPResSo and waLBerla, have extensive functionality, including graphical visualizations as well as modeling that does not use LBM.
-For this reason a comparison with other software packages is not on a level field.
-However, these packages do have an abundance of reusable components.
-
-There was a strong focus on modularity when designing the waLBerla framework in order to enhance productivity, reusability, and maintainability \citep{bauer2021walberla}.
-Its software design has enabled waLBerla to be successfully applied in several projects as a basis for various extensions \citep{bauer2021walberla}.
-
-\subsection{Surface Understandability}
-
-Ten random source code files of each software package were reviewed for several measures.
-This measure of surface understandability may not perfectly reflect each package due to the limitation of only testing ten files.
-
-All of the packages appear to have a consistent indentation and formatting style.
-Only LUMA and HemeLB explicitly identify coding standards that are used during development.
-Generally, the software packages use consistent, distinctive, and meaningful code identifiers.
-Only four packages appear to use vague identifiers.
-Hard coded constants were observed in the source code of twelve packages.
-The constants are used for various parameters, mathematical constants, and matrix definitions.
-All of the packages are well commented and the comments clearly indicate what is being done.
-Domain algorithms are noted in the source code of eleven packages.
-All of the packages are well modularized.
-When observing the source code files, it was found that thirteen of the packages have a consistent style and order for function parameters.
-
-\begin{figure}[h!]
-	\begin{center}
-		\includegraphics[width=1.0\textwidth]{understandability_chart}
-		\caption{AHP Understandability Score}
-		\label{Fig_Understandability}
-	\end{center}
-\end{figure}
-
-Figure \ref{Fig_Understandability} shows the surface understandability ranking of the software packages using AHP.
-Software packages with a high score have a consistent indentation and formatting style, and consistent, distinctive, and meaningful code identifiers.
-They also have hard coded constants, and explicitly identify mathematical and LBM algorithms.
-Their comments are clear and indicate what is being done in the source code.
-The source code is well modularized and structured.
-
-Software developers noted that in their opinion users have generally found their packages to be understandable.
-One developer commented that some users have attempted to run physically incompatible LBM methods, and the solution was to edit the code to prevent such combinations, as well as to update the documentation to prevent misunderstanding the methods.
-Similarly, another developer noted that some users had issues setting up parameters for LBM schemes.
-The solution to this was to update the interface where these parameters are set, as well as to add functionality to test the stability of the parameters.
-A third developer noted that some users lack the background knowledge to easily model fluid dynamics problems using their software.
-A frequently asked questions section was added to their package website to help users find answers to common questions.
-The package also has detailed documentation, including guides and usage requirements specification, to better help user understand the software.
-
-\subsection{Traceability}
-
-Software packages were not quantitatively measured for traceability.
-In this project we asked developers to comment on traceability, specifically on their software package's documentation and how it fit into their development process.
-
-One developer noted that all major additions to their package had accompanying changes to artifacts and documentation.
-They noted that considerable effort had been put into the documentation.
-They further commented that they want to lower the entry barrier for new developers, and because of that their package has a considerable amount of developer documentation.
-This documentation informs developers on how to get started, and orients them to the artifacts, source code, and system architecture, as well as how the software package build system works, and how the coupling between the simulation engine and the interface works.
-
-Another developer noted the importance of documentation for both the users and developers of their software.
-New features are always added to the documentation.
-The developers use documentation to stay up to date on the status of the software package, and to help expand features, like computational models or algorithms.
-This is necessary so that the coding standard for these models is kept consistent with new developers.
-
-The importance of documentation for both users and developers was stressed throughout the interviews.
-However, it was noted several times that a lack of time and funding has a negative affect on the documentation.
-Most of the developers are scientific researchers evaluated on the scientific papers that they produce.
-Writing and updating documentation is something that is done in free time, if that time arises, and if writing documentation is a concern.
-Sometimes it is a last priority for the developers.
-Finding ways to hasten updating documentation would increase the frequency of such updates and benefit both users and developers.
-
-One developer noted the use of documentation generators like Doxygen. It would be advisable for more projects to use such automatic document generation tools, since some projects do not do this.
-
-\subsection{Visibility and Transparency}
-
-Software package artifacts were reviewed for the identification of a specific development model, like a waterfall of agile development model, and the presence of documentation recording the development process and standard.
-They were also reviewed for the identification of the development environment, and the presence of release notes.
-The packages tended to not explicitly use well-known development models.
-This was also noted in the interviews with developers, as detailed below.
-The development teams of these packages are fairly small and easily organized without the need for such processes.
-Seven of the software packages did have some artifacts outlining the general development process, how to contribute, and the status of the package or its components.
-Eight of the packages have artifacts that note the development environment.
-While this information could help developers, and would improve transparency, the small close-knit nature of the development teams make explicitly publicly specifying this information practically unnecessary.
-Version release notes were found in nine of the software packages.
-
-\begin{figure}[h!]
-	\begin{center}
-		\includegraphics[width=1.0\textwidth]{visibilitytransparency_chart}
-		\caption{AHP Visibility and Transparency Score}
-		\label{Fig_VisibilityTransparency}
-	\end{center}
-\end{figure}
-
-Figure \ref{Fig_VisibilityTransparency} shows the visibility and transparency ranking of the software packages using AHP.
-Software packages with a high score have an explicit development model and defined development process.
-They also
-had detailed and easy to access notes accompanying software releases.
-
-Developers were asked to comment on the obstacles in their development process.
-One developer noted that a lot of their source code had been written with a specific application in mind, and that there is too much coupling between components.
-Addressing this issue would help with code modifiability and reusability.
-Updating the development process would help resolve this issue and prevent such issues in the future.
-Improving the visibility of software changes and a peer review process would also help.
-Improving the software engineering education or experience of developers is also an idea that was brought up by several developers.
-Specifically, developers should always write code that is decoupled and modular, and should keep in mind the visibility of their contributions by better updating documentation and ensuring that their contributions are transparent to the rest of the developers.
-This would help catch issues in the contributions, and improve source code maintenance.
-
-According to one developer, some obstacles to the development processes of their package had been overcome by the introduction of continuous integration practices, and a peer review process for contributions.
-These practices decrease development and maintenance times.
-Another developer mentioned that their package had two sets of code, for executing the models on the CPU and GPU, and that maintenance was decreased by introducing macros, a practice which became a common part of the development process.
-
-Developers were also asked how documentation fits into their development process.
-Several developers noted that developer documentation plays an important role in familiarizing potential contributors to the software system architecture.
-Without the guidance that the documentation provides it would be unlikely that contributions would pass the peer review process.
-
-None of the software packages whose developers were interviewed have a formal software development model.
-The packages all have fairly small development teams.
-These teams do accept outside contributors, but generally the teams are tight-knit, often working at the same institution, although one of the packages has an international team.
-One developer noted that while no formal model is used, their development model is something similar to a combination of agile and waterfall development models.
-
-The developers noted similar project management processes.
-For teams of only a couple of developers, the addition of new features or major changes are discussed with the entire team.
-Projects with more than a couple of developers have lead developer roles.
-These lead developers review potential additions to the software.
-The software packages use GitHub for managing the project.
-Typically there are several development branches as well as the master branch.
-
-\subsection{Reproducibility}
-
-Software packages were not quantitatively measured for reproducibility.
-In this project we asked developers to comment on reproducibility when we interviewed them.
-
-Developers were asked if they have any concern that their computational results would not be reproducible in the future, and if they had taken any steps to ensure reproducibility.
-
-One developer noted that they compare the results of their methods against manually calculated results.
-These comparison tests are automatically run for all source code changes.
-The tests are run when a pull request is opened on GitHub.
-Even once these tests are complete, a peer review process is done before changes are fully committed to the appropriate branch.
-The results for all of the LBM schemes on the software package development branch are also compared daily for correctness, ensuring that the system output reflects the expected output.
-
-Several developers noted that they currently do not have a system in place to test for reproducibility, but it is of interest and could be implemented in the future.
-Generally, the mathematical foundations of the models are verified, but the output of the software package is not compared to other output.
-Depending on the package and how it outputs LBM solutions, it may not be practical or feasible.
-A correct output may not be exactly reproducible, as it may be dependent on a probability distribution, so strictly comparing results may not be appropriate.
-
-The source code and artifacts of some software packages may be reproducible.
-There is considerable variance in the quality of software specifications and other developer documentation.
-Some packages are well detailed, and translating the specifications into source code will produce similar results across developers.
-
-\subsection{Unambiguity}
-
-Software packages were not quantitatively measured for unambiguity.
-In this project we asked developers to comment on unambiguity when we interviewed them.
-
-Developers were asked if they thought that the current documentation can clearly convey all necessary knowledge to the users, and if they had taken any steps to ensure clarity.
-
-One developer noted that their documentation was meant for users that are already familiar with the underlying physics and computational fluid dynamics methods, and that these concepts are not explained in detail.
-The documentation focuses solely on how to technically use the software package, and includes a user guide and tutorial walk through of how to set up and run a simulation.
-With this in mind the developer believes that their documentation is in reasonable shape for users with a minimum knowledge of the underlying physics.
-If new users have technical questions these can be addressed in further revisions of the documentation.
-New developers should also find that the documentation is reasonably clear and useful.
-Information that is missing, like detailed explanations of dependencies, is referenced in the documents.
-
-Another developer also noted that their documentation was in reasonable shape, but they would really appreciate more user feedback to improve it.
-They also noted that they believe a lack of knowledge of the underlying physics and computational fluid dynamics concepts can be an issue for some users.
-This information can be referenced in the documentation, but it is not something that the documentation needs to detail.
-
-\subsection{Overall Quality}
-
-Figure \ref{Fig_OverallScore} shows the overall ranking of the software packages using AHP.
-Software packages with an overall high score had ranked high in at least several of the individual qualities that were quantitatively measured.
-
-\begin{figure}[h!]
-	\begin{center}
-		\includegraphics[width=1.0\textwidth]{finalscore_chart}
-		\caption{AHP Overall Score}
-		\label{Fig_OverallScore}
-	\end{center}
-\end{figure}
-
-Looking at the top three ranked packages, ESPResSo had achieved a relative high score in installability, surface correctness and verifiability, surface usability, maintainability, reusability, and visibility and transparency.
-Ludwig scored high in surface correctness and verifiability, surface robustness, surface usability, maintainability, and visibility and transparency.
-Palabos scored high in installability, surface reliability, surface robustness, maintainability, and understandability.
-
-Section \ref{qualityrecommentations} further analyzes these findings and offers some software quality recommendations for future development of LBM software.
-
-\newpage
-
-\section{Answers To Research Questions}\label{answersquestions}
-
-\subsection{Artifacts Present - DONE 1st DRAFT}\label{artifacts}
-
-The software packages were examined for the presence of artifacts, which were then categorized by frequency.
-Below we have grouped them into common, less common, and rare artifacts.
-Common artifacts were found in 16 to 23 of the software packages.
-Less common artifacts were found in 8 to 15 of the software packages.
-Rare artifacts were found in 1 to 7 of the software packages.
-
-\subsubsection{Common Artifacts}
-The following artifacts were commonly found in the 23 software packages that were tested.
-All of the top four AHP ranked packages, ESPREsSo, Ludwig, Palabos, and OpenLB have each of these artifacts, except for a requirements specification or theory notes.
-Only three of the top four ranked packages were found to have such an artifact.
-
-\begin{singlespace}
-\begin{multicols}{2}	
-\begin{itemize}
-	\item Authors/Developers List
-	\item Bug Tracker
-	\item Dependency Notes 
-	\item Installation Guide/Instructions 
-	\item License
-	\item List of Related Articles/Publications 
-	\item Makefile / Build File 
-	\item README File 
-	\item Requirements Specification / Theory Notes
-	\item Tutorial
-\end{itemize}
-\end{multicols}
-\end{singlespace}
-
-These common artifacts contribute to the quality of the software in the following ways.
-A list of authors and developers helps potential users and contributors contact project members to answer questions affecting many software qualities.
-A bug tracker helps with organizing improvements to the software.
-Dependency notes help users install the software, as does an installation guide.
-Makefiles or other automated build files decrease human error in the installation process.
-Licenses promote usage of the software.
-Requirements specifications, linked theory notes, and related articles and publications help make the software more understandable, decrease ambiguity, and help with verifying its correctness.
-README files also help with understandability.
-The tutorials help users become familiar with using the software.
-
-\subsubsection{Less Common Artifacts}
-The following artifacts were less commonly found in the 23 software packages that were tested. The top four AHP ranked packages have most of these artifacts.
-Only one of those four packages does not have a user manual or guide, but it does have a detailed and informative website.
-Another one of the top four packages does not appear to have publicly visible design documentation.
-A third package from the list does not appear to use a version control system.
-It is possible that such a system is used by the package since its website notes package version numbers, but the artifacts do not explicitly state the use of such a system.
-
-\begin{singlespace}
-	\begin{multicols}{2}	
-		\begin{itemize}
-			\item Change Log / Release Notes
-			\item Design Documentation
-			\item Functional Specification/Notes	
-			\item Performance Information/Notes
-			\item Test Plan/Report/Script/Data/Cases
-			\item User Manual/Guide
-			\item Version Control
-		\end{itemize}
-	\end{multicols}
-\end{singlespace}
-
-These artifacts also contribute to software quality.
-A change log or release notes improve traceability of the software.
-Design documentation helps with maintaining, modifying, and reusing the software.
-Version control also helps to improve these qualities, as well as with traceability.
-Functional specifications and notes clarify the software, contributing to usability and understandability.
-A user manual also helps with those qualities and with installability, and correctness and verificability.
-Performance notes suggest that performance was considered when developing the software.
-Test plans, scripts, and cases, help verify correctness.
-
-\subsubsection{Rare Artifacts}
-The following artifacts were rarely found in the 23 software packages that were assessed.
-It is not common for the top four AHP ranked packages to have many of these artifacts.
-None of the top four packages have any explicit API documentation.
-Three of these packages have information on contributing to the project.
-Two of them have a FAQ section or forum.
-One has verification and validation notes, and a video guide explaining how to use the software.
-
-\begin{singlespace}
-	\begin{multicols}{2}	
-		\begin{itemize}
-			\item API Documentation
-			\item Developer/Contributor Manual/Guide
-			\item FAQ / Forum
-			\item Verification and Validation Plan/Notes
-			\item Video Guide (including YouTube)
-		\end{itemize}
-	\end{multicols}
-\end{singlespace}
-
-Although these artifacts were rarely found in our set of LBM software, they also contribute to software quality.
-API documentation helps with reusing the software.
-Developer and contributor manuals and guides help with maintainability, visibility and transparency.
-FAQs and forums improve usability of software.
-Verification and validation notes can be used to help check if a system meets specifications.
-Video guides can contribute to many software qualities, depending on the content of the video.
-
-\subsection{Tools Used - DONE 1st Draft}
-Software tools are used to support the development, verification, maintenance, and evolution of software, software processes, and artifacts \citep{ghezzi1991fundamentals}.
-Many tools are used by LBM software packages.
-The tools noted here are subdivided into development tools, dependencies, and project management tools.
-
-\subsubsection{Development Tools}
-
-These tools support the development of end products but do not become part of them, unlike dependencies that remain in the application once it is released \citep{ghezzi1991fundamentals}.
-The following type of development tools were explicitly noted in the artifacts or web-pages of the 23 LBM packages that were assessed.
-It is likely that other tools, such as debuggers, were used but are not specified here.
-
-\begin{singlespace}
-	\begin{multicols}{2}	
-		\begin{itemize}
-			\item Continuous Integration Tools
-			\item Code Editors
-			\item Development Environment
-			\item Runtime Environments
-			\item Compilers
-			\item Unit Testing Tools
-			\item Correctness Verification Tools
-		\end{itemize}
-	\end{multicols}
-\end{singlespace}
-
-The above tools can verify the correctness of software during its development.
-Only two of the software packages that were assessed mentioned using continuous integration tools like Travis CI.
-Code editors and compilers were explicitly noted to have been used by several packages, and were likely used by all of them.
-One of the packages explicitly noted the use of unit testing, but did not specify the tool that was used.
-Likewise, the use of a tool for verifying the correctness of output was noted by one of the developers.
-It is likely that similar tools were used when developing other software packages.
-
-\subsubsection{Dependencies}
-
-The following types of dependencies were explicitly noted in the artifacts or web-pages of the 23 LBM packages that were assessed.
-It is possible that other types of dependencies are part of these software packages but are not clearly specified in their artifacts or web sites and because of that they are not listed here.
-
-\begin{singlespace}
-	\begin{multicols}{2}	
-		\begin{itemize}
-			\item Build Automation Tools
-			\item Technical Libraries
-			\item Domain Specific Libraries
-		\end{itemize}
-	\end{multicols}
-\end{singlespace}
-
-Most of the software packages use some sort of build automation tools, most commonly Make.
-They also all use various technical and domain specific libraries.
-Technical libraries include visualization, data analysis, and message passing libraries.
-Domain specific libraries are scientific computing libraries.
-Libraries that are explicitly stated in artifact, and were not noted during our observations, may fall outside of these categories.
-
-\subsubsection{Project Management Tools}
-
-Many of the software packages that were assessed were developed by teams of two or more people.
-Their work needed to be coordinated and managed.
-The following types of project management tools were explicitly noted in the artifacts, web-pages, or interviews with the developers of the 23 LBM packages that were assessed.
-As with development tools and dependencies, it is possible that other types of project management tools were used to coordinate and manage the projects but are not specified and because of that they are not listed here.
-
-\begin{singlespace}
-	\begin{multicols}{2}	
-		\begin{itemize}
-			\item Collaboration Tools
-			\item Email
-			\item Change Tracking Tools
-			\item Version Control Tools
-			\item Document Generation Tools
-		\end{itemize}
-	\end{multicols}
-\end{singlespace}
-
-Collaboration tools are noted as being used when developing the software projects.
-Most often email and video conferencing is used.
-Project management software was not explicitly mentioned, but it is possible that some of the projects use such software.
-Many of the projects are located on GitHub, and its developers use the platform to help manage their projects, especially bug related issues.
-Most of the projects appear to use change tracking and version control tools.
-They often use GitHub for this, and a few use Git or CVS.
-Document generation tools are mentioned in the artifacts of 12 of the projects.
-Tools such as Sphinx and Doxygen are explicitly used in this capacity.
-Drasil could be considered as an alternative to generate artifacts since it is geared specifically towards research software domains like LBM software.
-
-\subsection{Principles, Processes, and Methodologies - DONE 1st DRAFT}
-
-Most of the software packages do not explicitly state in their artifacts the motivations or design principles that were considered when developing the software.
-One package, Sailfish, indicates that shortening the development time was considered in early stages of design, with the developers opting for using Python and CUDA/OpenCL to achieve this without sacrificing any computational performance.
-The goals of that project are explicitly listed as performance, scalability, agility and extendability, maintenance, and ease of use.
-The project scored well in these categories during our assessment.
-
-Processes, like methods, are ways of doing things, especially in an orderly way; while methodologies are defined as systems of methods \citep{ghezzi1991fundamentals}.
-It is not explicitly indicated in the artifacts of most of the packages that development involved following any specific model, like a waterfall or agile development model.
-One developer noted that while no formal model is used, their development model is something similar to a combination of agile and waterfall development models.
-The developer teams of the LBM packages are fairly small and easily organized without the need for such models.
-
-Seven of the software packages contain artifacts outlining the general development process, like basic instructions on how to contribute.
-Most of the development teams accept outside contributors, but generally the teams are centralized, often working at the same institution.
-
-The developers that were interviewed all noted similar project management processes.
-In teams of only a couple of developers, additions of new features or major changes are discussed with the entire team.
-Projects with more than a couple developers have lead developer roles.
-These lead developers review potential additions to the software.
-One of the developers that was interviewed noted that a peer review process is used to assess major changes and additions.
-
-Many software packages use GitHub for managing the project. Typically there are several simultaneous development branches.
-
-Documentation was also noted as playing a significant role in the development process, specifically with on-boarding new developers.
-A goal of documentation is to lower the entry barrier for these new contributors.
-The documentation provides information on how to get started, orients the user to artifacts and the source code, and explains how the system works, including the so-called simulation engine and interface.
-The use of document generation tools is mentioned in the artifacts of 12 software packages, and was noted during interviews with developers.
-Document generation tools that were mentioned include Sphinx and Doxygen.
-
-Two types of software changes were discussed during interviews with developers.
-One is feature additions, which arise from a scientific or functional need.
-The development of such changes was noted to involve more formal discussions within the development team and the addition of them into the main software branch involves lead developers.
-The other change type is code refactoring, which only sometimes involves formal discussions with the development team.
-New developers were noted to play an increased role in these changes compared to the former changes.
-Software bugs are typically addressed in a similar fashion as code refactoring, and issue tracking is commonly used to manage these changes.
-
-Interviews with the developers of software packages also revealed a more frequent use of both unit testing and continuous integration in the development process than was found by only assessing the artifacts.
-The use of automatic installation processes is also common.
-Most often this involved a Make script.
-
-\subsection{Pain Points - DONE 1st Draft}
-
-Developers were asked to comment on obstacles in their development process, obstacles encountered by users, and potential future obstacles.
-
-Several development process obstacles were identified.
-One developer noted that their small development team has a lack of time to implement new features.
-Small development teams are common for LBM software packages.
-Team members are almost always part of the same institute or already know each other from other projects.
-External contributions are rare despite many of the projects accepting them.
-
-A lack of software development experience was noted by another developer.
-Many of the team members on their project are domain experts and there is a steep learning curve to get these team members to contribute good quality source code.
-It was further noted that this has been somewhat addressed, as the code has been re-written to best ensure ease of future contributions.
-The same developer also noted that there is a lack of incentive and funding in academia for developing widely used scientific software.
-There are no journals that publish such source code.
-There is no organization helping with the development of good quality software.
-Scientific software is often developed in-house by the very researchers that temporarily use it in their own research.
-This software has a defined user and purpose, and often does not meet the standards that would be required by external users.
-Such software is rarely commercialized, in which case it is not open-source.
-
-Documentation of LBM software is noted as being important, but its quality could be improved.
-There is often no time or funding for maintaining quality documentation for software that is rarely used outside of the development team.
-Furthermore, the documentation generally only provides a shallow overview of the underlying computational fluid dynamics theory.
-Users would be well advised to already be familiar with these topics to or they should spend significant time referencing theory resources.
-The documentation instead generally focuses on explaining how to use the software.
-It is of course not feasible for package documentation to address the underlying metaphysics topics in detail, so it is advised that the package documentation links to resources that better explain the underlying theory.
-Sometimes, frequently asked questions about the underlying theory are answered in the documentation.
-
-Setting up parallelization was also noted as a technical pain point by one of the developers, and the introduction of continuous integration by another.
-Software development knowledge could mitigate such pain points.
-As already noted, many of the developers are domain experts and not professional software developers.
-One of the interviewed developers that has relatively more extensive software development experience noted that eliminating equivalent statements using macros had helped improve the quality of their source code, specifically helping with reusing code to run on both the CPU and GPU.
-
-Difficulties with ensuring correctness were also noted by several developers.
-They indicated that tests are run on all new source code additions, testing both individual modules and the system to verify correctness.
-These tests compare the package output to known correct output using test cases.
-A developer commented that the amount of testing data that is needed for some cases is a problem as free testing services do not offer capabilities to store and process such large amounts of data, and in-house testing solutions needed to be created to address this limitation.
-
-A few obstacles related to users were found.
-Several developers noted that users sometimes try to use incorrect LBM method combinations to solve their problems.
-Furthermore, some users think that the packages will work out of the box to solve their cases, while the packages require both a good understanding of computational fluid dynamics and an understanding of the requirements for formulating problems in the individual packages, which can be a significant endeavor.
-These software packages are not like commercial software packages.
-They are generally set up to solve specific research problems, and are otfen primarily used by their developers.
-While they are modifiable to solve similar problems, these modifications are not trivial.
-So far this problem of on-boarding new users has been addressed by updating the documentation to better inform users of the underlying LBM theory and package requirements.
-Similar issues with LBM parameters were noted by another developer.
-Updating the user interface to better explain theoretical principles, as well as test user input for compatibility, was the implemented solution.
-As noted above, sometimes frequently asked questions on the underlying theory and on how to use the software are answered in the documentation.
-
-One of the interviewed developers commented that parts of their package's source code had been refactored to Python in order to help address usability issues.
-Python was perceived as a much more usable language, and it would be easy for future users and developers to learn and understand the source code.
-
-A few potential future obstacles were noted.
-One developer said that their source code had been written with a specific application in mind and that due to this there was too much coupling between components of the source code.
-This would have an impact on future modifiability and reusability when trying to extend the software, and the code would need to be refactored.
-
-As noted above, difficulties with ensuring future correctness could also arise. As new methods and functionality is added into the software, new test cases and test data will need to be developed.
-
-\subsection{Quality Recommendations - DONE 1st Draft}\label{qualityrecommentations}
-
-The following points regarding software quality should be considered when developing LBM software packages.
-
-\subsubsection{Installability}
-
-\begin{itemize}
-	\item Include OS compatibility, including specific OS versions.
-	\item Include complete installation instructions.
-	\item Installation instructions should be written as if the user does not have any dependencies installed and is installing on a clean OS.
-	\item List all dependencies in the installation instructions, dependency versions, and how to install them.
-	\item If possible, automate the installation of dependencies.
-	\item Automate the installation process as much as possible.
-	\item The installation instruction should only be in one location.
-	\item Include descriptive error messages for errors encountered during installation.
-	\item Include a way to validate the installation.
-	\item Include instructions for uninstalling the software.
-\end{itemize}
-
-\subsubsection{Surface Correctness and Verifiability}
-
-\begin{itemize}
-	\item Use a requirements specification document.
-	\item Make public the requirements specification document, or explicitly reference the domain theory that the software is designed from.
-	\item Ensure the above information is easy to find. Consider adding it to the user manual. 
-	\item Development teams should include both domain experts and experienced software developers.
-	\item Use and make public detailed documentation. Consider using automatic document generation tools.
-	\item Include detailed tutorials that include expected output.
-	\item Use unit testing during development and make them public.
-	\item Modularize the source code.
-	\item Use continuous integration tools and processes during development.
-\end{itemize}
-
-\subsubsection{Surface Reliability}
-
-\begin{itemize}
-	\item Include descriptive error messages where appropriate.
-	\item In case automatic installation of dependencies fails, the system should indicate to the user what dependencies need to be installed manually.
-	\item The packages should include detailed tutorials, including dependencies, expected output, and any additional supplementary documentation that may be required.
-\end{itemize}
-
-\subsubsection{Surface Robustness}
-
-\begin{itemize}
-	\item The system must provide descriptive error messages when it encounters unexpected input, including incorrect data types, empty input, and missing files or links.
-\end{itemize}
-
-\subsubsection{Surface Performance}
-
-\begin{itemize}
-	\item Integrate parallelization tools and techniques in order to reduce processing time. Consider GPU processing, CUDA, and MPI. 
-	\item The user should be able to choose to process their model on either the CPU or GPU.
-\end{itemize}
-
-\subsubsection{Surface Usability}
-
-\begin{itemize}
-	\item Include user hardware and software requirements documentation.
-	\item Include a user tutorial that indicates the expected output.
-	\item Include a detailed user manual. It should identify elements of user interfaces, and identify all requirements to model a system.	
-	\item State appropriate fluid dynamics problems that the software is designed to model in the documentation, and explicitly state the limits of the software. 
-	\item Provide documentation that details the background theory information, or provide a reference to such information.
-	\item Identify expected user characteristics.
-	\item Keep all documentation in one location.
-	\item Do not duplicate artifact information.
-	\item Maintain a user support model (Git, email, forum, FAQ)
-	\item If possible, consider using popular user-friendly software languages like Python. Especially consider this for parts of the source code that is likely to be modified or reviewed by users.
-\end{itemize}
-
-\subsubsection{Maintainability}
-
-\begin{itemize}
-	\item Keep artifacts updated.
-	\item Ideally include most of the common and less common artifacts listed in Section \ref{artifacts}.
-	\item Include version numbers and release notes for all major source code and artifact releases. 
-	\item Have a defined process for accepting contributions.
-	\item Include documentation for making contributions to the project. 
-	\item Use an issue tracker (Git, email, SourceForge, other) to manage bugs and changes. Issues should be regularly reviewed and closed. 
-	\item Use a version control system (GitHib, CVS).
-	\item Source code needs to be well commented. Typically, more than ten percent of LBM package source code is comments. 
-	\item Ensure source code is modular. Generally, this means that source code is split into files along functionality.
-	\item Eliminate code duplication.
-	\item If possible, consider using popular user-friendly software languages like Python. Especially consider this for parts of the source code that is likely to be modified or reviewed by users.
-\end{itemize}
-
-\subsubsection{Modifiability}
-
-\begin{itemize}
-	\item Ensure a high degree of code modularity and abstraction.
-	\item If possible, consider using popular user-friendly software languages like Python. Especially consider this for parts of the source code that is likely to be modified or reviewed by users.
-	\item Consider future source code modifiabilility as early as the design stage of development.
-	\item Consider flexibility of data structures and data storage in the design stage.
-	\item Provide validation benchmark tools and data to the public.
-\end{itemize}
-
-\subsubsection{Reusability}
-
-\begin{itemize}
-	\item Ensure a high degree of code modularity and abstraction.
-	\item Document module interfaces in design and developer documentation.
-	\item Provide API documentation, if applicable.
-\end{itemize}
-
-\subsubsection{Surface Understandability}
-
-\begin{itemize}
-	\item Maintain consistent formatting style across source code files.
-	\item Identify a coding standard in the documentation, if a coding standard is used during development.
-	\item Use consistent, distinctive, and meaningful code identifiers. 
-	\item Identify algorithms that are used.
-	\item Hard code domain constants.
-	\item Add meaningful comments. They should indicate what is being done in each section of source code.
-	\item Ensure a high degree of code modularity and abstraction.
-	\item Include user guides, and state technical and problem modeling requirements.
-	\item State problem cases that the software is designed to model in the documentation, and explicitly state the limits of the software. 
-	\item Consider adding a FAQ section to the documentation. 
-\end{itemize}
-
-\subsubsection{Traceability}
-
-\begin{itemize}
-	\item Update all relevant documentation when a change to the software is made.
-	\item Use automatic document generation tools to limit the time spent on updating documentation.
-	\item Provide a developers guide.
-	\item Provide a resource that helps orient users and developers to the artifacts, source code, and system architecture.
-	\item Highlight new features and changes between artifact versions. 
-\end{itemize}
-
-\subsubsection{Visibility and Transparency}
-
-\begin{itemize}
-	\item Identify the development model that was used, if one was used.
-	\item Identify the development environment.
-	\item Identify the development process and how new users can contribute. 
-	\item Include notes with all releases.
-	\item Ensure a high degree of code modularity and abstraction.
-	\item Communicate all changes to the development team.
-	\item Use continuous integration processes and tools.
-	\item Consider peer review processes to assess contributions. 
-	\item Eliminate code duplication.
-	\item Update all relevant developer documentation when a change to the software is made.
-	\item Use project management software.
-\end{itemize}
-
-\subsubsection{Reproducibility}
-
-\begin{itemize}
-	\item Test output against manually calculated or known correct results.
-	\item Automate the testing of output. Run tests after all code changes.
-	\item Consider peer review processes to assess contributions. 
-	\item Update all relevant design documentation when a change to the software is made.
-\end{itemize}
-
-\subsubsection{Unambiguity}
-
-\begin{itemize}
-	\item Documentation must state all technical requirements and software dependencies. 
-	\item A user guide must be provided.
-	\item State appropriate fluid dynamics problems that the software is designed to model in the documentation, and explicitly state the limits of the software. 
-	\item The documentation should either explain the underlying CFD theories or provide a reference to appropriate resources.
-	\item Consider asking users for feedback on the documentation, specifically if it clearly conveys all necessary information to the users.
-\end{itemize}
-
-\subsection{Designation Comparison}
-
-answers last question of research questions:
-
-How does software designated as high quality by this methodology compare	with top rated software by the community?
-
-\newpage
-\section{Conclusion}
-
-- key findings restated
-- comment on future work in state of the practice SCS
-
-the following was moved from the former background section, may need edits as part of conclusion:
-
-  The software qualities that are assessed in this report are listed in Section \ref{softwarequalities}.
-Ensuring software quality can take significant time, but as \cite{SmithEtAl2016} shows, SCS developers often cannot see themselves meeting the required time constraints to better develop some of the artifacts that improve qualities, like detailed documentation.
-An inability to develop such documentation due to time constraints could be mitigated by the use of automatic document generation tools like Drasil.
-The Drasil Framework consists of a collection of Domain Specific Languages (DSL) for capturing scientific documents, structures, and computing
-knowledge, and then transforming this knowledge into relevant software artifacts without having to manually duplicate knowledge into multiple artifacts \citep{zhao2018}.
-
-\newpage
-\section{Appendix}
-\begin{singlespace}
-
-\subsection{Research Questions}\label{researchquestions}
-	The following are the research questions that are considered.
+\subsection{Purpose}\label{purpose}
+
+The LBM software packages are assessed in order to answer the following research questions:
 
@@
 
+The packages are assessed for a given set of quality attributes using quantitative and qualitative data.
+The qualities are defined in Section \ref{softwarequalities}.
+Qualitative data is gathered from interviews with software package developers.
+The interview questions are listed in the Appendix in \ref{interviewquestions}.
+Quantitative data is measured using the quality metrics listed in the Appendix in \ref{measurementtemplate2}.
+In this context the purposes of the report are as follows:
+
+\begin{itemize}
+	\item Develop and test an updated methodology for assessing the state of the practice of SCS projects
+	\item Describe the specific example of LBM software structure and goals
+	\item Report on the measure of a subset of LBM software packages along quality metrics
+	\item Evaluate the state of the practice of LBM software development along quality attributes
+	\item Make suggestions for improving LBM software along quality attributes
+	\item Make suggestions for improving state of the practice of SCS assessments
+\end{itemize}
+
+\subsection{Motivation}
+
+The purpose of state of the practice assessments is to understand how software quality is impacted by software development choices, including principles, processes, and tools, within SCS communities.
+This knowledge can be used to guide future development of SCS, specifically along quality attributes, and reduce software quality failures.
+This work reports on the state of the practice of LBM software development and makes suggestions on improving the quality of software in this domain.
+
+This assessment of the state of the practice of LBM software development builds off of prior work on assessing the state of research software development.
+The previous state of the practice assessments include domains such as Geographic Information Systems \citep{SmithEtAl2018_arXivGIS}, Mesh Generators \citep{SmithEtAl2016}, Oceanographic Software \citep{smith2015state}, Seismology software \citep{SmithEtAl2018}, and statistical software for psychology \citep{SmithEtAl2018_StatSoft}.
+
+In the course of this assessment we updated the methodology that was used during previous assessments.
+Details of the new methodology are in \citep{methodology2020}.
+The previous set of research questions were critically assessed and modified.
+The updated questions are listed in Section \ref{purpose}.
+In this re-boot we collected more quantitative and qualitative data, focused on software quality attributes, and added more measures, including collecting empirical data and interviewing developers.
+We have also added a domain analysis to better characterize the functionality provided by the software, and have leveraged the expertise of a domain expert.
+As in past assessments, the collected data was combined to rank the software using the Analytic Hierarchy Process (AHP).
+The domain expert was consulted to verify the ordering.
+
+\subsection{Scope}
+
+This report analyzes a filtered set of LBM software packages along quantitative and qualitative measures.
+Many of these measures are captured using surface measurements, which can be categorized as initial and easy to capture measurements, of the underlying quality; they may not represent the true quality of the software.
+Surface measurements are taken as they allow us to apply the same measurement, with reasonable effort, along all software packages despite technical and functional variabilities among the set of software packages.
+Not all qualities are quantitatively measured.
+For example, performance is not measured by running the code.
+A surface investigation of the documentation of each software package for performance information is conducted instead.
+The measured software packages are open source, and it is recommended to keep this in mind when considering the recommendations of this report.
+Some recommendations may not apply to closed-source software.
+Practices surrounding close source software development may differ.
+
+The report addresses what was done during the development of the software packages to address the quality attributes listed in Section \ref{softwarequalities}.
+It then provides general guidance on how to improve these qualities when developing LBM software.
+It does not make suggestions on what should have been done or should be done for any one specific package.
+Best practices may differ among software packages due to their inherent organizational and technical differences.
+
+\subsection{Organization}
+
+The report is organized as follows:
+
+\begin{itemize}
+	\item \textbf{Introduction} to the report, including its purpose, motivation, scope, and organization.
+	\item LBM software \textbf{Domain Analysis}.
+A domain analysis consists of systematically identifying and documenting the commonalities, variabilities, and terminology of a software family \citep{Weiss1997}.
+	The purpose of this analysis in this state of the practice assessment is to better classify the different software products based on their functionality.
+	\item \textbf{Methodology} of this state of the practice assessment, including the steps of the methodology, software quality definitions and how these qualities are assessed in this report, an overview of candidate software selection, empirical measures gathered, and the analytical hierarchy process.
+This follows a general methodology for assessing the state of the practice for SCS domains.
+	\item Quantitative and qualitative \textbf{Results}. 
+	\item An analysis of the results and \textbf{Answers To Research Questions}.
+	\item \textbf{Conclusion} to the report including comments on suggestions for future state of the practice assessments.
+	\item \textbf{Appendix}: This section includes our Research Questions, the Measurement Template, Grading Template, Ethics Approval, and Developer Interview Questions.
+\end{itemize}
+
 \newpage
+
+\section{Domain Analysis - DONE SECTION 1st DRAFT}\label{domainanalysis}
+
+A domain analysis consists of systematically identifying and documenting the commonalities, variabilities, and parameters of variation of a software family \citep{Weiss1997}.
+A software family is defined by \cite{parnas1976design} as “a set of programs whose common properties are so extensive that it is advantageous to study the common properties of the programs before analyzing individual members”.
+
+We added this domain commonality and variability analysis to our original methodology because the first time we did the State of the Practice exercise we found we weren't comparing "apples to apples".
+Software packages can have considerable variation, even when the packages appear to be in the same software family.
+Viewing the ranking exercise as a means for selecting the right tool for a job, functionality is important.
+A 2D Lattice Boltzmann solver is not a substitute for a 3D Lattice Boltzmann solver when the latter is required.
+Therefore, this exercise seeks to classify the different software products based on their functionality.
+Nonfunctional qualities are assessed through other means in our process.
+For instance, through usability experiments as noted in the \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Methodology.pdf}{Methodology for Assessing the State of the Practice for Domain X} document.
+
+This domain analysis of the LBM SCS family is organized into the four subsection located below.
+The first subsection reviews the basics of Lattice Boltzmann systems.
+The next three subsections consist of lists of commonalities, variabilities, and parameters of variation, respectively.
+These three sections form the heart of the domain analysis and include an extensive set of cross-references to demonstrate the relationships between the different items.
+
+\subsection{Lattice Boltzmann Systems}
+
+LBM are a family of fluid dynamics algorithms for simulating single-phase and multiphase fluid flows, often incorporating additional physical complexities \citep{chen1998lattice}.
+They consider the behaviours of a collection of particles as a single unit at the mesoscopic scale, between the nanoscopic and microscopic scales.
+These methods predict the positional probability of a collection of particles moving through a lattice structure.
+Off-the-shelf (OTS) Lattice Boltzmann Solvers (LBS) allow for a range of fluid and physical model input parameters, computational parameters, and output parameters.
+
+LBS model fluid dynamics within a boundary using a predefined lattice structure and a two step calculation process.
+The first process is streaming, where the particles move along the lattice via links.
+The second process is collision, where energy and momentum is transferred among particles that collide \citep{bao2011lattice}.
+There are many standardized lattice models - individual solvers within the family might only use a subset of them.
+LBM use the initial parameters of the fluid to find the probability of where along the lattice linkages a group of particles are most likely to travel.
+It then moves the particles into the next node, and transfers the energy and momentum if a collision occurs.
+Then the process repeats for the duration of the modeling instance.
+
+\subsection{Commonalities}\label{comm_sec}
+
+This section lists common features among potential family members.
+The commonalities are organized using the following abstraction of the system, which can be used to describe all Lattice Boltzmann systems: input information, generate the simulation, output the results.
+Section \ref{comm_lbm} describes the commonalities for the simulation step.
+Section \ref{comm_in} highlights the input information that is required for Lattice Boltzmann systems.
+The next section, Section \ref{comm_out}, shows the common features for the output of Lattice Boltzmann systems, such as the requirement that mesh information be written to files.
+Although the output information could simply be written to the computer’s memory, in all practical applications it is desirable to have a persistent record of the output that was created.
+The final section covers nonfunctional requirements of the system.
+For instance, all systems will have the goal that the response time to a user’s request is small enough to allow the user to focus on his/her problem and to maintain his/her train of thought, without excessive waiting time.
+The commonality in this case is refined by a later variability because the specific requirement on the response time will depend on the individual software packages, their simulation models, inputs, and outputs.
+
+Each commonality below uses the same structure.
+All of the commonalities are assigned a unique item number, which takes the form of a natural number with the prefix ``C''.
+Following this, a description of the commonality is provided along with a list of related variabilities, which are given as hyperlinks that allow navigation of the document to the text describing the variability.
+
+\subsubsection{Lattice Boltzmann Method Solvers}\label{comm_lbm}
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
+		\hline
+		\bf Item Number& C\refstepcounter{comnum}\thecomnum{}\label{lattice} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
+		\hline
+		\bf Description & A lattice discretizes a computational domain into a finite number of points. All LBS discretize the computational domain using a regular, evenly spaced grid within a boundary.\\
+		\hline
+		\bf Related Variability & \vref{numveldir} \vref{varcolop} \vref{varboundary}\\
+		\hline
+	\end{tabular}
+\end{minipage}
+~\newline
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
+		\hline
+		\bf Item Number& C\refstepcounter{comnum}\thecomnum\label{collisionoperator} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
+		\hline
+		\bf Description & All Lattice Boltzmann versions use a collision operator which concerns collisions between particles.
+Collision operators map collisions of particles within the lattice space.
+The Bhatnagar-Gross-Krook Collision Operator is a common LBM collision operator that preserves continuity for a discretized model, for each velocity direction $i$.
+Its equation is $\mathrm{\Omega_i} = \frac{1}{\tau}(f_{i}^{eq}-f_{i})$, where $\tau$ is the relaxation rate towards equilibrium, $f^{eq}$ is the equilibrium particle probability distribution function, and $f$ is the particle probability distribution function.\\
+		\hline
+		\bf Related Variability & \vref{dimensions} \vref{numveldir}\\
+		\hline
+	\end{tabular}
+\end{minipage}
+~\newline
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
+		\hline
+		\bf Item Number& C\refstepcounter{comnum}\thecomnum\label{probabilitydensityfunc} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
+		\hline
+		\bf Description & All Lattice Boltzmann versions use a probability density function to give the probability that fluid has moved into a specific domain.\\
+		\hline
+		\bf Related Variability & \vref{dimensions}\\
+		\hline
+	\end{tabular}
+\end{minipage}
+~\newline
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
+		\hline
+		\bf Item Number& C\refstepcounter{comnum}\thecomnum\label{equildistfunc} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
+		\hline
+		\bf Description & Every Lattice Boltzmann version uses an equilibrium distribution function to capture the probability distribution of particles.\\
+		\hline
+		\bf Related Variability & \vref{varedf} \vref{edfstorage} \vref{edfcoeff}\\
+		\hline
+	\end{tabular}
+\end{minipage}
+~\newline
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
+		\hline
+		\bf Item Number& C\refstepcounter{comnum}\thecomnum\label{Boltzmanntransequ} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
+		\hline
+		\bf Description & Every Lattice Boltzmann version uses a Boltzmann transport equation to describe the statistical behaviour of a system that does not have collisions.
+The equation is $(\frac{d}{dt} + \mathrm{e} \cdot \nabla_\mathrm{x} + \frac{F}{m} \cdot \nabla_\mathrm{e})f = \Omega(t)$, where $\mathrm{e}$ is the velocity, $\nabla_\mathrm{x}$ is the position vector gradient, $\nabla_\mathrm{e}$ is the velocity gradient, and $\Omega(t)$ is the collision operator as a function of time.\\
+		\hline
+		\bf Related Variability & \vref{vartraneq}\\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+\subsubsection{Input}\label{comm_in}
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
+		\hline
+		\bf Item Number& C\refstepcounter{comnum}\thecomnum\label{fluidandboundaryinfo} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
+		\hline
+		\bf Description & The LBS require fluid, model, and boundary information for the problem.
+This includes, but is not limited to, fluid acceleration rate, velocity, and viscosity, as well as the number of dimensions in the lattice model, and the number of velocity directions in the lattice.\\
+		\hline
+		\bf Related Variability & \vref{interface}\\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+\subsubsection{Output}\label{comm_out}
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
+		\hline
+		\bf Item Number& C\refstepcounter{comnum}\thecomnum\label{predictionstomemory} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
+		\hline
+		\bf Description & LBS write fluid predictions, such as the output of the transport equation, to memory.\\
+		\hline
+		\bf Related Variability & \vref{varvisual} \vref{formatoutput} \\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+\subsection{Variabilities}\label{var_sec}
+This section provides a list of characteristics that may vary among family members.
+As in Section \ref{comm_sec}, the first three subsections on variabilities are organized into the following sublists: Simulation Models, Input and Output.
+The final two subsections list variabilities that can be characterized as system constraints and as nonfunctional requirements.
+
+As for the commonalities, each variability is labelled with a unique item number.
+In this case the numbers are prepended with the letter ``V''.
+The other three headings provided for each variability are: Description, Related Commonality, and Related Parameter.
+The related commonalities and parameters are given as a set of identifiers that respectively refer back to the previous section on commonalities or refer forward to the next section on parameters of variation.
+
+\subsubsection{Lattice Boltzmann Method Solvers}
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
+		\hline
+		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{parallel} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
+		\hline
+		\bf Description & LBS may use a framework for parallel processing of the model.\\
+		\hline
+		\bf Related Commonality & None\\
+		\hline
+		\bf Related Parameter & \pref{parparallel}\\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+~\newline\\
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
+		\hline
+		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{varedf} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
+		\hline
+		\bf Description & Different versions of an equilibrium distribution function can capture the probability distribution of particles.\\
+		\hline
+		\bf Related Commonality & \cref{equildistfunc}\\
+		\hline
+		\bf Related Parameter & \pref{paredf}\\
+		\hline
+	\end{tabular}
+\end{minipage}
+~\newline
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
+		\hline
+		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{edfstorage} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
+		\hline
+		\bf Description & Storage patters for distribution function can vary.\\
+		\hline
+		\bf Related Commonality & \cref{equildistfunc}\\
+		\hline
+		\bf Related Parameter & \pref{parstorage}\\
+		\hline
+	\end{tabular}
+\end{minipage}
+~\newline
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
+		\hline
+		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{edfcoeff} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
+		\hline
+		\bf Description & Coefficients used with the distribution function can vary. These are based on the number of velocity directions in the model.\\
+		\hline
+		\bf Related Commonality & \cref{equildistfunc}\\
+		\hline
+		\bf Related Parameter & \pref{paredfcoeff}\\
+		\hline
+	\end{tabular}
+\end{minipage}
+~\newline
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
+		\hline
+		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{dimensions} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
+		\hline
+		\bf Description & The number of dimensions in the lattice of the model can vary.\\
+		\hline
+		\bf Related Commonality & \cref{lattice} \cref{probabilitydensityfunc}\\
+		\hline
+		\bf Related Parameter & \pref{pardimensions}\\
+		\hline
+	\end{tabular}
+\end{minipage}
+~\newline
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
+		\hline
+		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{numveldir} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
+		\hline 
+		\bf Description & The number of velocity directions in the lattice of the model can vary.\\
+		\hline
+		\bf Related Commonality & \cref{lattice} \cref{collisionoperator}\\
+		\hline
+		\bf Related Parameter & \pref{parveldir}\\
+		\hline
+	\end{tabular}
+\end{minipage}
+~\newline
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
+		\hline
+		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{varcolop} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
+		\hline
+		\bf Description & Various collision operators can be used.\\
+		\hline
+		\bf Related Commonality & \cref{lattice} \cref{collisionoperator}\\
+		\hline
+		\bf Related Parameter & \pref{parcollisop}\\
+		\hline
+	\end{tabular}
+\end{minipage}
+~\newline
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
+		\hline
+		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{vartraneq} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
+		\hline
+		\bf Description & Various transport equations can be used to describe the statistical behaviour of the system\\
+		\hline
+		\bf Related Commonality & \cref{Boltzmanntransequ}\\
+		\hline
+		\bf Related Parameter & \pref{partransequ}\\
+		\hline
+	\end{tabular}
+\end{minipage}
+~\newline
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
+		\hline
+		\bf Item Number&  V\refstepcounter{varnum}\thevarnum\label{numberoffluids} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
+		\hline
+		\bf Description & The number of fluids allowed in the LBS.\\
+		\hline
+		\bf Related Commonality & \cref{fluidandboundaryinfo}\\
+		\hline
+		\bf Related Parameter & \pref{parfluids}\\
+		\hline
+	\end{tabular}
+\end{minipage}
+~\newline
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
+		\hline
+		\bf Item Number&  V\refstepcounter{varnum}\thevarnum\label{typefluidparm} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
+		\hline
+		\bf Description & The type of fluid parameters.\\
+		\hline
+		\bf Related Commonality & \cref{fluidandboundaryinfo}\\
+		\hline
+		\bf Related Parameter & \pref{partypeparm}\\
+		\hline
+	\end{tabular}
+\end{minipage}
+~\newline
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
+		\hline
+		\bf Item Number&  V\refstepcounter{varnum}\thevarnum\label{varboundary} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
+		\hline
+		\bf Description & The boundary of the lattice can have various conditions.\\
+		\hline
+		\bf Related Commonality & \cref{lattice}\\
+		\hline
+		\bf Related Parameter & \pref{parboundary}\\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+\subsubsection{Input}
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
+		\hline
+		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{interface} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
+		\hline
+		\bf Description & The input interface can vary between LBS.\\
+		\hline
+		\bf Related Commonality & \cref{fluidandboundaryinfo}\\
+		\hline
+		\bf Related Parameter & \pref{parinterface}\\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+\subsubsection{Output}
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
+		\hline
+		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{varvisual} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
+		\hline
+		\bf Description & Visual presentation of the prediction.\\
+		\hline
+		\bf Related Commonality & \cref{predictionstomemory}\\
+		\hline
+		\bf Related Parameter & \pref{parvarvisual}\\
+		\hline
+	\end{tabular}
+\end{minipage}
+~\newline
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
+		\hline
+		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{formatoutput} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
+		\hline
+		\bf Description & Format of prediction information.\\
+		\hline
+		\bf Related Commonality & \cref{predictionstomemory}\\
+		\hline
+		\bf Related Parameter & \pref{parformatoutput}\\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+\subsubsection{System Constraints}
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
+		\hline
+		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{processing} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
+		\hline
+		\bf Description & Hardware that processes the calculations\\
+		\hline
+		\bf Related Commonality & None \\
+		\hline
+		\bf Related Parameter & \pref{parprocessing}\\
+		\hline
+	\end{tabular}
+\end{minipage}
+~\newline
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
+		\hline
+		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{operatingsystem} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
+		\hline
+		\bf Description & Operating systems on which LBS run.\\
+		\hline
+		\bf Related Commonality & None \\
+		\hline
+		\bf Related Parameter & \pref{paropsystem}\\
+		\hline
+	\end{tabular}
+\end{minipage}
+~\newline
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
+		\hline
+		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{storage} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
+		\hline
+		\bf Description & Amount of storage and memory needed for the LBS.\\
+		\hline
+		\bf Related Commonality & None \\
+		\hline
+		\bf Related Parameter & \pref{parstorageconst}\\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+\newpage
+\subsection{Parameters of Variation}
+This section specifies the parameters of variation for the variabilities listed in Section \ref{var_sec}.
+They are organized into the same five subcategories as employed previously: Simulation Models, Input, Output, System Constraints, Nonfunctional Requirements.
+
+Each parameter of variation is given a unique identifier of the form “P” followed by a natural number.
+The corresponding variability is listed and a hyperlink is provided that allows navigation back to the appropriate item in Section \ref{var_sec}.
+The final entry for each parameter of variation is the binding time, which is the time in the software lifecycle when the variability is fixed.
+The binding time could be during specification, or during building
+of the system (compile time), or during execution of the system (run time).
+It is possible to have a mixture of binding times.
+For instance, a parameter of variation could have a binding time of “specification or building” to represent that the parameter could be set at specification time, or it could be postponed until the given family member is built.
+The choice of postponing the decision until the build could be associated with the presence of a domain specific language that would allow postponing decisions on the values of the parameter of variation.
+
+\subsubsection{Lattice Boltzmann Method Solvers}
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
+		\hline
+		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parparallel} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
+		\hline
+		\bf Corresponding Variability & \vref{parallel}\\
+		\hline
+		\bf Range of Parameters & OpenMP, OpenCL, CUDA, MPI are used if the execution of the LBS is parallelized.\\
+		\hline
+		\bf Binding Time & Build Time\\
+		\hline
+	\end{tabular}
+\end{minipage}
+~\newline
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
+		\hline
+		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{paredf} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
+		\hline
+		\bf Corresponding Variability & \vref{varedf}\\
+		\hline
+		\bf Range of Parameters & Equilibrium approximation varies between incompressible or compressible models.\\		
+		\hline
+		\bf Binding Time & Build Time\\
+		\hline
+	\end{tabular}
+\end{minipage}
+~\newline
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
+		\hline
+		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parstorage} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
+		\hline
+		\bf Corresponding Variability & \vref{edfstorage}\\
+		\hline
+		\bf Range of Parameters & Various data structures can be used to store function output, including single and multi-dimensional arrays, depending on the problem model and developer preferences.\\
+		\hline
+		\bf Binding Time & Build Time\\
+		\hline
+	\end{tabular}
+\end{minipage}
+~\newline
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
+		\hline
+		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{paredfcoeff} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
+		\hline
+		\bf Corresponding Variability & \vref{edfcoeff}\\
+		\hline
+		\bf Range of Parameters & Numerous coefficients for equilibrium distribution function based on number of velocity directions.
+The number of velocity directions is typically 2, 3, 5, 9, 13, 15, 19, or 27.\\
+		\hline
+		\bf Binding Time & Build Time\\
+		\hline
+	\end{tabular}
+\end{minipage}
+~\newline
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
+		\hline
+		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{pardimensions} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
+		\hline
+		\bf Corresponding Variability & \vref{dimensions}\\
+		\hline
+		\bf Range of Parameters & LBS model has 1, 2, or 3 dimensions.\\
+		\hline
+		\bf Binding Time & Build Time or Run Time\\
+		\hline
+	\end{tabular}
+\end{minipage}
+~\newline
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
+		\hline
+		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parveldir} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
+		\hline
+		\bf Corresponding Variability & \vref{numveldir}\\
+		\hline
+		\bf Range of Parameters & One dimensional models include options of 2, 3, and 5 velocity directions.
+Two dimensional models include options of 9, 13, and 15 velocity directions.
+Three dimensional models include options of 15, 19, and 27 velocity directions.
+Figure \ref{dimension_model} from \citep{thurey2009stable} illustrates two commonly used LBM models in two and three dimensions.\\
+		\hline
+		\bf Binding Time & Build Time or Run Time\\
+		\hline
+	\end{tabular}
+\end{minipage}
+~\newline
+
+\begin{figure}[h!]
+	\begin{center}
+		\includegraphics[width=0.7\textwidth]{dimension_model}
+		\caption{Commonly used LBM models in two and three dimensions}
+		\label{dimension_model}
+	\end{center}
+\end{figure}
+
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
+		\hline
+		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parcollisop} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
+		\hline
+		\bf Corresponding Variability & \vref{varcolop}\\
+		\hline
+		\bf Range of Parameters & SRT, TRT, MRT, BGK collision operators.\\
+		\hline
+		\bf Binding Time & Build Time\\
+		\hline
+	\end{tabular}
+\end{minipage}
+~\newline
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
+		\hline
+		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{partransequ} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
+		\hline
+		\bf Corresponding Variability & \vref{vartraneq}\\
+		\hline
+		\bf Range of Parameters & Collision and collision free transport equations.
+LBS sometimes use one or the other, often the collision transport equation.
+A Boolean parameter could be used to select between these equations in systems that can apply either equation.\\
+		\hline
+		\bf Binding Time & Build Time\\
+		\hline
+	\end{tabular}
+\end{minipage}
+~\newline
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
+		\hline
+		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parfluids} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
+		\hline
+		\bf Corresponding Variability & \vref{numberoffluids}\\
+		\hline
+		\bf Range of Parameters & LBS can model a natural number of fluids.\\
+		\hline
+		\bf Binding Time & Build Time\\
+		\hline
+	\end{tabular}
+\end{minipage}
+~\newline
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
+		\hline
+		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{partypeparm} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
+		\hline
+		\bf Corresponding Variability & \vref{typefluidparm}\\
+		\hline
+		\bf Range of Parameters & LBS fluid parameters include Reynolds Number, density, viscosity, time, pressure, force, direction, relaxation rate, turbulence.
+All of these parameters have the type of real number.\\
+		\hline
+		\bf Binding Time & Build Time\\
+		\hline
+	\end{tabular}
+\end{minipage}
+~\newline
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
+		\hline
+		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parboundary} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
+		\hline
+		\bf Corresponding Variability & \vref{varboundary}\\
+		\hline
+		\bf Range of Parameters & Lattice boundary can have reflective or non-reflective conditions.
+Some LBS will only model reflective or non-reflective conditions.
+If there is a choice then this can be indicated by a Boolean parameter.\\
+		\hline
+		\bf Binding Time & Build Time\\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+\subsubsection{Input}
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
+		\hline
+		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parinterface} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
+		\hline
+		\bf Corresponding Variability & \vref{interface}\\
+		\hline
+		\bf Range of Parameters & Input can be graphical, text or file.\\
+		\hline
+		\bf Binding Time & Build Time\\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+\subsubsection{Output}
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
+		\hline
+		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parvarvisual} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
+		\hline
+		\bf Corresponding Variability & \vref{varvisual}\\
+		\hline
+		\bf Range of Parameters & LBS can provide 1D, 2D, and 3D rendering of  the model.\\
+		\hline
+		\bf Binding Time & Build Time or Run Time\\
+		\hline
+	\end{tabular}
+\end{minipage}
+~\newline
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
+		\hline
+		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parformatoutput} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
+		\hline
+		\bf Corresponding Variability & \vref{formatoutput}\\
+		\hline
+		\bf Range of Parameters & LBS prediction information is output in either text or binary format.\\
+		\hline
+		\bf Binding Time & Build Time\\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+\subsubsection{System Constraints}
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
+		\hline
+		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parprocessing} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
+		\hline
+		\bf Corresponding Variability & \vref{processing}\\
+		\hline
+		\bf Range of Parameters & The LBS model can be calculated on the CPU or GPU.\\
+		\hline
+		\bf Binding Time & Build Time\\
+		\hline
+	\end{tabular}
+\end{minipage}
+~\newline
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
+		\hline
+		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{paropsystem} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
+		\hline
+		\bf Corresponding Variability & \vref{operatingsystem}\\
+		\hline
+		\bf Range of Parameters & LBS can be run on Windows, MacOS, or Linux versions.\\
+		\hline
+		\bf Binding Time & Build Time\\
+		\hline
+	\end{tabular}
+\end{minipage}
+~\newline
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
+		\hline
+		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parstorageconst} \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
+		\hline
+		\bf Corresponding Variability & \vref{storage}\\
+		\hline
+		\bf Range of Parameters & The amount of memory and storage varies between LBS.\\
+		\hline
+		\bf Binding Time & Run Time\\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+
+\newpage
+\section{Methodology - DONE SECTION 1st DRAFT}
+
+In this project we set out to answer the research questions listed in Section \ref{purpose} for the LBM SCS domain.
+The process involved systematically measuring and analyzing members of this software family along the quality attributes found in Section \ref{softwarequalities}.
+This included gathering quantitative and qualitative measurements.
+A goal of the project was to produce a quality assessment for LBM software packages.
+
+The methodology used in this LBM software assessment is a general state of the practice methodology that can be applied to any SCS domain.
+It was developed as an update to previous state of the practice exercises.
+This general methodology is linked in the next subsection.
+
+We collected quantitative data using the measures found in the measurement template of Section \ref{measurementtemplate}.
+Some of this was empirical software engineering related data, such as the number of files, number of lines of code, percentage of issues that are closed, etc.
+Most of the data was gathered by manually investigating the software, its source code, and its artifacts, while some was gathered using the empirical measurement tools discussed in Section \ref{empiricalmeasures}.
+
+We also collected qualitative data by interviewing the software package developers and asking them the questions found in the Appendix in \ref{interviewquestions}.
+Furthermore, we solicited the assistance of domain experts to better assess each software package by leveraging their experience to assess the functional and non-functional requirements for the software domain.
+
+This section begins by describing the steps of the overall process we used to select, measure and compare LBM software.
+This is followed by quality definitions and how these qualities were assessed in our project.
+The rest of the section provides an overview of how candidate software packages were selected and filtered, the empirical measurements and software tools that were used, and the analytic hierarchy process (AHP).
+
+\subsection{Process}
+
+ The following steps provide an overview of how the assessment was conducted:
+
+\begin{enumerate}
+	\item List candidate software packages for the domain. This is discussed in Section \ref{identifysoftware}.
+	\item Filter the software package list. This is discussed in Section \ref{filtersoftware}.
+	\item Gather the source code and documentation for each software package.
+	\item Collect empirical measures. This is discussed in Section \ref{empiricalmeasures}.
+	\item Measure using the measurement template. This is discussed in Section \ref{empiricalmeasures}. The measurement template can be found in the Appendix in \ref{measurementtemplate2}.
+	\item Survey the developers.
+The developers of each software package in the filtered software list were contacted for voluntary interviews.
+The interview questions can be found in the Appendix in \ref{interviewquestions}.
+	\item Use AHP to rank the software packages. This is discussed in Section \ref{AHP}.
+	\item Analyze the results and answer the research questions. The answers can be found in Section \ref{answersquestions}.
+\end{enumerate}
+
+These steps are further detailed in the \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Methodology.pdf}{Methodology for Assessing the State of the Practice for Domain X} document.
+
+\subsection{Software Qualities}\label{softwarequalities}
+
+Software quality attributes facilitate the measurement and comparison of software packages in this state of the practice assessment.
+We adopt software quality definitions from various researchers and subject matter expert entities.
+Some of the definitions are from \cite{Smithetal2020}.
+The quality measurement results are found in Section \ref{qualityresults} of this report.
+Section \ref{qualityrecommentations} lists recommendations to address software qualities in LBM software packages.
+The following are the software quality definitions used in this state of the practice exercise, along with comments regarding their quantitative and qualitative measurement.
+
+\subsubsection{Installability}
+
+Installability is measured by the effort required for the installation, uninstallation or reinstallation of a software product in a specified
+environment \citep{ISO/IEC25010} \citep{lenhard2013measuring}.
+A good measure of installability correlates with scenarios when low or moderate effort is required to gather and prepare software for its general use on a system for which it was designed.
+In this case effort includes the time spent finding and understanding the installation instructions, the man-time and resources spent performing the installation procedure, and the absence or ease of overcoming system compatibility issues.
+The ability to reasonably validate the installation procedure also has a positive effect on the measure of installability.
+Similarly, the ease of uninstallation has an affect on the measure of installability.
+
+\subsubsection{Correctness}
+
+ A software program is correct if it behaves according to its stated
+specifications \citep{GhezziEtAl2003}.
+This requires that the specification is available.
+Software is unlikely to have a formal specification if it is not developed by seasoned or professional software developers.
+Since some software does not have a specification available, the correctness of software cannot always be verified.
+Despite an absent specification, the correctness of the output of scientific computing software can sometimes be manually verified by applying domain knowledge.
+A good measure of correctness correlates with the availability of a requirements specification and reference to domain theory, as well as the explicit use of tools or techniques for building confidence of correctness, such as documentation generators and software analysis tools.
+
+\subsubsection{Verifiability}
+
+Verifiability is measured by the extent to which a set of tests can be written and executed to demonstrate that the delivered system meets the specification \citep{sommerville}.
+Similarly to correctness, verifiability is correlated with the availability of a specification and with reference to domain knowledge.
+A good measure of verifiability is further correlated with the availability of well written tutorials that include expected output, with software unit testing documentation, and with evidence of continuous integration during the development process.
+
+\subsubsection{Reliability}
+
+Reliability is measured by the probability of failure-free operation of a computer program in a specified environment for a specified time, i.e.
+the average time interval between two failures also known as the mean time to failure (MTTF) \citep{GhezziEtAl2003} \citep{musa1987software}.
+Reliability is thus positively correlated with the absence of errors during installation and use.
+Recoverability from errors also improves reliability.
+
+\subsubsection{Robustness}
+
+Software possesses the characteristic of robustness if it behaves “reasonably” in two situations: i) when it encounters circumstances not anticipated in the requirements specification; and ii) when the assumptions in its requirements specification are violated \citep{boehm2007software} \citep{ghezzi1991fundamentals}.
+A good measure of robustness correlates with a reasonable reaction to unexpected input, including data of the wrong type, empty input, or missing files or links.
+A reasonable reaction includes an appropriate error message and the ability to recover the system.
+
+\subsubsection{Performance}
+
+Performance is measured by the degree to which a system or component accomplishes its designated functions within given constraints, such as speed (database response times, for instance), throughput (transactions per second), capacity (concurrent usage loads), and timing (hard real-time demands) \citep{IEEEStdGlossarySET1990} \citep{wiegers2003softreq}.
+In this state of the practice assessment performance was not quantitatively measured.
+Instead the documentation of each software package was observed for information that alludes to a consideration of performance, such as parallelization tools.
+
+\subsubsection{Usability}
+
+Usability is measured by the extent to which a software product can be used by specified users to achieve specified goals with effectiveness, efficiency, and satisfaction in a specified context of use \citep{nielsonusability}.
+A good measure of usability correlates with the presence of documentation, including tutorials, manuals, and defined user characteristics, and user support.
+Preferably the user support model has avenues to contact developers and report issues.
+
+\subsubsection{Maintainability}
+
+A measure of maintainability is the effort with which a software
+system or component can be modified to correct faults, improve performance or other attributes, and satisfy new requirements \citep{IEEEStdGlossarySET1990} \citep{boehm2007software}.
+In this state of the practice analysis maintainability is measured by the quality of documentation artifacts, and the presence of version control and issue tracking.
+These artifacts can greatly decrease the effort needed to modify software.
+There are many documentation artifacts that can improve maintainability, including user and developer manuals, specifications, README files, change logs, release notes, publications, forums, and instructional websites.
+
+\subsubsection{Modifiability}
+
+Modifiability refers to the ease with which stable changes can be made to a system and the flexibility of the system to adopt such changes \citep{8016712}.
+This state of the practice assessment did not quantitatively measure modifiability.
+Developers were asked in interviews if they considered the ease of future changes when developing the software packages, specifically changes to the structure of the system, modules and code blocks.
+A follow up question asked if any measures had been taken.
+
+\subsubsection{Reusability}
+
+Reusability refers to the extent to which components of a software package can be used with or without adaptation in other software packages \citep{kalagiakos2003non}.
+A good measure of reusability results from a large number of easily reusable components.
+Increased software modularization, defined as the presence of smaller components with well defined interfaces, is important.
+For this state of the practice assessment, a good measure of reusablity correlates with an increased number of code files, and the availability of API documentation.
+
+\subsubsection{Understandability}
+
+Understandability is measured by the capability of the software package to enable the user to understand its suitability and function \citep{ISO9126}.
+It is an artifact-dependent quality.
+Understandability is different for the user-interface, source code, and documentation.
+In this state of the practice analysis, understandability focuses on the source code.
+It is measured by the consistency of a formatting style, the extend of modularization, the explicit identification of coding standards, the presence of meaningful identifiers, and clarity of comments.
+
+\subsubsection{Traceability}
+
+Traceability refers to the ability to link the software implementation and the software artifacts, especially the requirement specification \citep{McCallEtAl1977}.
+Similar to the quality of correctness, this requires some form of specification to be available.
+The quality refers to keeping track of information as it changes forms or relates between artifacts.
+This state of the practice assessment did not quantitatively measure traceability.
+Developers were asked in interviews how documentation fits into their development process.
+
+\subsubsection{Visibility and Transparency}
+
+Visibility and transparency refer to the extent to which all of the steps of a software development process, and the current status of it, are conveyed clearly \citep{ghezzi1991fundamentals}.
+In this state of the practice assessment a good measure of visibility and transparency correlates with a well defined development process, the presence of development process and environment documentation, and software package version release notes.
+
+\subsubsection{Reproducibility}
+
+Software achieves reproducibility if another developer can take the requirements documentation and re-obtain the same software artifacts \citep{BenureauAndRougier2017}.
+This includes the output of the software, where the scientific results are compared between software implementations, or between software implementations and manually calculated results.
+This state of the practice assessment did not quantitatively measure reproducibility.
+Developers were asked in interviews if they have any concern that their computational results won't be reproducible in the future, and if they have taken any steps to ensure reproducibility.
+
+\subsubsection{Unambiguity}
+
+Unambiguity refers to the extent to which two readers have similar interpretations when reading software artifacts.
+In other words, artifacts are unambiguous if, and only if, they only have one interpretation \citep{IEEE1998}.
+This state of the practice assessment did not quantitatively measure unambiguity.
+Developers were asked in interviews if they think that the current documentation can clearly convey all necessary knowledge to the users, and how they achieved this or what improvements are needed to achieve it.
+
+\subsection{Identify Candidate Software}\label{identifysoftware}
+
+The candidate software was found through search engine queries targeting authoritative lists of software.
+We found LBM software listed on the websites GitHub and swMATH, as well as through articles found in scholarly journals and databases.
+
+The following properties were considered when creating the list and reviewing the candidate software:
+
+\begin{enumerate}
+	\item The software functionality must fall within the identified domain.
+	\item The source code must be viewable.
+	\item The empirical measures should be available, which implies a preference for GitHub-style repositories.
+	\item The software cannot be marked as incomplete or in an initial development phase.
+\end{enumerate}
+
+The initial list had 45 packages, including a few packages that were later found to not have publicly available source code, or to be in an incomplete state of development.
+
+\subsection{Filter the Software List}\label{filtersoftware}
+
+To reduce the number of members in the candidate software list to a manageable size, the following filters were applied. The filters were applied in the priority order listed.
+
+\begin{enumerate}
+	\item Scope: Software is removed by narrowing what functionality is considered to be within the scope of the domain.
+	\item Usage: Software packages were eliminated if their installation procedure was missing or not clear and easy to follow.
+	\item Age: The older software packages (age being measured by the last date when a change was made) were eliminated, except in the cases where an older software package appears to be highly recommended and currently in use.
+\end{enumerate}
+
+For the third item in the above filter, software packages were characterized as `alive' if their related documentation had been updated within
+the last 18 month. Packages were categorized as `dead' if the last update of this information was more than 18 month ago.\\
+
+\begin{center}
+	\begin{tabular}{ p{4cm}p{3cm}p{3cm} }
+		\hline
+		Name & Released & Updated\\
+		\hline
+		DL\_MESO (LBE) & unclear & 2020 Mar\\
+		ESPResSo & 2010 Nov & 2020 Jun\\
+		ESPResSo++ & 2011 Feb & 2020 Apr\\
+		lbmpy& unknown  & 2020 Jun  \\
+		lettuce & 2019 May & 2020 Jul\\
+		Ludwig& 2018 Aug & 2020 Jul\\
+		LUMA& 2016 Nov   & 2020 Feb\\
+		MechSys & 2008 Jun    & 2020 Jul\\
+		OpenLB & 2007 Jul & 2019 Oct\\
+		Palabos & unclear & 2020 Jul\\
+		pyLBM & 2015 Jun&   2020 Jun\\
+		Sailfish & 2012 Nov & 2019 Jun\\
+		TCLB & 2013 Jun  & 2020 Apr\\
+		waLBerla & 2008 Aug & 2020 Jul\\
+		\hline
+	\end{tabular}
+	\captionof{table}{Alive Software Packages}\label{alivepackages}
+\end{center}
+
+While the initial list had 45 packages, filtering by scope, usage, and age decreased the size of the list to 23 packages.
+Many of the 22 packages that were removed could not be tested as there was no installation guide, they were incomplete, source code was not publicly available, a license was needed, or the project was out of scope or not up to a standard that would support incorporating them into this study.
+These eliminated software packages are listed in the Appendix in Section \ref{eliminatedpackagessection}.
+Of the remaining 23 packages that were studied, some were kept on the list despite being marked as dead due to their prevalence on authoritative lists on LBM software and due to their surface excellence, specifically the considerable time that was put into these projects.
+
+The final list of software packages that were analyzed in this project can be found in the following two tables.
+Table \ref{alivepackages} lists packages that fell into the `alive' category as of mid 2020, and Table \ref{deadpackages} lists packages that were `dead' at that time.
+\\
+
+\begin{center}
+	\begin{tabular}{ p{4cm}p{3cm}p{3cm} }
+		\hline
+		Name & Released & Updated\\
+		\hline
+		HemeLB & 2007 Jun & 2018 Aug\\
+		laboetie & 2014 Nov & 2018 Aug\\		
+		LatBo.jl & 2014 Aug & 2017 Feb\\
+		LIMBES & 2010 Nov & 2014 Dec\\
+		LB3D & unclear & 2012 Mar\\
+		LB2D-Prime & 2005 & 2012 Apr\\
+		LB3D-Prime & 2005 & 2011 Oct\\
+		MP-LABS & 2008 Jun & 2014 Oct\\
+		SunlightLB & 2005 Sep & 2012 Nov\\
+		\hline
+	\end{tabular}
+	\captionof{table}{Dead Software Packages}\label{deadpackages}
+\end{center}
+
+There is considerable variation among these software packages, including their intended purpose, size, user interfaces, and software languages used.
+For example, the OpenLB software package is predominantly a C++ package that makes use of hybrid parallelization and was designed to address a range of computational fluid dynamics problems \citep{heuveline2009towards}.
+The software package pyLBM is an all-in-one Python language package for numerical simulations \citep{graille2017pylbm}.
+ESPResSo is an extensible simulation package that is specifically for research on soft matter, and is written in C++ and Python \citep{weik2019espresso}.
+The HemeLB package is used for efficient simulation of fluid flow in several medical domains, and is written predominantly in C, C++, and Python \citep{mazzeo2008hemelb}.
+
+\subsection{Empirical Measures}\label{empiricalmeasures}
+
+The quality measurements in this assessment rely on the gathering and analyzing of raw and processed empirical data related to the research questions listed in Section \ref{purpose}.
+
+All of the quality measurements that are part of the AHP analysis are empirical measurements.
+Qualitative data gathered during interviews with developers is not part of the AHP analysis.
+This part of the assessment focuses on data that is reasonably easy to collect.
+Much of the data was gathered by manually reviewing the artifacts of each software package, while some was gathered using the freeware tools discussed below.
+The data that was gathered is listed in the measurement template found in the Appendix in \ref{measurementtemplate2}.
+Some of the data required processing other data within the template, including the status of the software package, which relied on the last commit date; the percentage of issues that are closed, which relied on the number of open and closed issues; and the percentage of code that is comments, which relied on the number of total lines and comment lines in text-based files.
+The complete measurement template data was then analyzed using the grading template found in the Appendix in \ref{gradingtemplate2}, the output of which was analyzed using the AHP described in Section \ref{AHP}.
+
+Most of the measurement template data was gathered by observing GitHub repository metrics and software package artifacts, or by processing data gathered using freeware tools.
+Data in the final three sets of the measurement template was collected using these tools.
+The tool \href{https://github.com/tomgi/git_stats}{GitStats} was used to measure each software package's GitHub repository for the number of binary files, the number of added and deleted lines, and the number of commits over varying time intervals.
+The tool \href{https://github.com/boyter/scc}{Sloc Cloc and Code (scc)} was used to measure the number of text based files as well as the number of total, code, comment, and blank lines in each GitHub repository.
+Details on installing and running these tools can be found in the \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/A
+	Guide to Empirical Measures.pdf} {Guide to Empirical Measures} file in the AIMSS repository.
+
+\subsection{Analytical Hierarchy Process}\label{AHP}
+
+The Analytical Hierarchy Process (AHP) is a decision-making technique that is used to compare multiple options by multiple criteria.
+In our work AHP was used for comparing and ranking the LBM software packages using the overall impression quality scores that were gathered in the measurement template found in the Appendix in \ref{measurementtemplate2} using the grading template found in the Appendix in \ref{gradingtemplate2}.
+AHP performs a pairwise analysis using a matrix and generates an overall score as well as individual quality scores for each software package.
+\cite{SmithEtAl2016} shows how AHP is applied to ranking software based on quality measures.
+
+This project used a tool for conducting this process.
+The tool includes a sensitivity analysis that was used to ensure that the software package rankings are appropriate with respect to the uncertainty of the quality scores.
+The \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/AHP2020/LBM/README.txt}{README} file of the tool includes requirements and usage information.
+
+\newpage
+\section{Results - DONE SECTION 1st DRAFT}\label{qualityresults}
+
+This section presents the quantitative and qualitative findings of this study.
+The qualitative data that is presented was gathered through interviews with some developers.
+AHP results for qualities that were quantitatively measured are also found in this section.
+Quality definitions are found in section \ref{softwarequalities}.
+
+\subsection{Installability}
+
+All of the software packages that were tested have installation instructions.
+As noted previously, many of the twenty three software packages that were part of the original long list of forty six software packages but were later removed do not include documentation or installation instructions.
+
+All twenty three packages on the short list can be installed on some Unix-like systems.
+Seven packages could be installed on Windows, and five on macOS.
+Operating system compatibility is found in the documentation of nineteen software packages.
+All but one of the software packages were tested on Ubuntu for this state of the practice assessment.
+The remaining package was tested on CentOS.
+
+Of the software packages that were tested, most installation instructions are located in one place, often in an instruction manual or on a web-page.
+Sometimes incomplete installation instructions are found on a home page, with more detailed instructions located on another web-page or within formal documentation.
+Maintainability and correctness of these instructions could be improved if they were limited to one location.
+
+All but two of the software packages have automated some of the installation process. Most of these packages use Make to automate the installation, and a few of them use custom scripts.
+
+Errors encountered during the installation process were often quickly fixed thanks to descriptive error messages.
+Systems that provided vague error messages were more difficult to troubleshoot.
+Only three software packages that displayed a descriptive error message were not recoverable, and most of these instances were due to hardware and operating system incompatibility, such as the requirement of CUDA.
+Fourteen software packages definitively broke during installation.
+A couple of packages did not provide a definitive message of the success or failure of installation.
+In these instances, validating the installation required performing a tutorial or running a script, as described below, if these were available.
+
+About half of the installation instructions are not written as if the person doing the installation has none of the dependent packages installed, and many software packages do not list all of their dependencies.
+Many dependencies needed to be manually installed.
+Sometimes only an error message during the installation process informs the user of the requirement of these packages.
+A detailed rewrite of the installation instructions from the point of view of installation on a clean operating system is suggested.
+
+Sixteen software packages require less than ten dependencies to be installed.
+All but one software package require less than twenty dependencies.
+Some packages may automatically install additional dependencies in the background.
+Eighteen of the software packages do not explicitly indicate software dependency versions.
+Some software package installation issues, specifically those occurring when manual installation of dependencies is required, may be avoided if versions of dependencies are specified.
+Fifteen software packages do not have detailed instructions for installing dependencies.
+
+Sixteen software packages have less than ten manual installation steps.
+If dependencies are installed in one command then none of the software packages take more than twenty steps to install.
+The average number of steps is about eight, and the fewest is three.
+
+All but six of the software packages have a way to verify the installation.
+Twelve have some sort of tutorial examples that can be run by the user.
+Most of these tutorials provide expected output.
+Some packages do not have a way to verify the output of the tutorials, and the user is left to assume correctness if no error message is encountered.
+Some other ways of installation validation include validation scripts, automatic validation after the installation, and instructions to manually review the file system.
+
+Uninstallation instructions were found for only one of the software packages.
+
+\begin{figure}[h!]
+	\begin{center}
+		\includegraphics[width=1.0\textwidth]{installability_chart}
+		\caption{AHP Installability Score}
+		\label{Fig_Installability}
+	\end{center}
+\end{figure}
+
+Figure \ref{Fig_Installability} shows the installability ranking of the software packages using AHP.
+Software packages with a higher score tend to have one set of linear installation instructions that are written as if the person doing the installation has none of the dependencies installed.
+The instructions often list compatible operating system versions and include instructions for the installation of dependencies.
+The top ranked packages often incorporate some sort of automation of the installation process and have fewer manual installation steps.
+The number of dependencies a package has does not correlate with a higher score.
+The ability to validate the installation process, often through tutorials or test examples that include expected output, is correlated with a higher score.
+
+Many software packages would benefit from a rewrite or reorganization of installation instructions.
+A single location for installation instructions would improve their maintainability and correctness.
+Listing compatible operating system and dependency versions would decrease installation time and errors, as would adding instructions on installing dependencies.
+Installation process errors should prompt the system to display detailed messages.
+Once a software package is installed, either an automatic validation needs to be performed or the user needs to be able to perform a manual validation using test examples that include expected output.
+Finally, uninstallation instructions should be included in the documentation.
+
+\subsection{Surface Correctness and Verifiability}
+
+Sixteen of the software packages include a requirements specification artifact or explicitly reference domain theory.
+This artifact is often found within a user manual, on a web-page, or is mentioned in related publications.
+In the latter case the user may need to spend significant time to find this information.
+
+Interviews with developers confirmed that many of the software packages are developed by domain experts with backgrounds in physics, mathematics and mechanical engineering.
+Many of these developers do not have formal software engineering education.
+Some of the development teams include computer scientists.
+Despite a lack of visible domain documentation and a resulting lower surface correctness and verifiability score, many software packages do have a strong domain background on account of their development team.
+
+Document generation tools are explicitly used by twelve software packages. Sphinx is used by eight of them, and Doxygen is used by seven. Several of the packages use both.
+
+Tutorials are available for all but five of the software packages.
+Generally they are linearly written and easy to follow.
+However, only eight tutorials provide an expected output.
+It is not possible to verify the correctness of the output of the software packages that are missing this key information.
+In these cases the user may need to assume correctness if there are no visible errors.
+
+Unit tests are only explicitly available for one of the software packages.
+Code modularization of most packages allow for users to create tests with varying degrees of effort.
+These tests allow developers and users to verify the correctness of fragments of the source code, and in doing so better assess the correctness of the entire package.
+
+The documentation or presence of continuous integration tools and techniques alludes to a more refined development process where faults are isolated and better recognized.
+Only two of the packages mentioned applying the practice of continuous integration in their development process.
+
+Interviews with the developers of software packages suggested a more frequent use of both unit testing and continuous integration in the development processes than initially surveyed.
+For example, OpenLB, pyLBM, and TCLB use such methods during development despite it not being explicitly clear from a surface analysis.
+
+\begin{figure}[h!]
+	\begin{center}
+		\includegraphics[width=1.0\textwidth]{correctnessverifiability_chart}
+		\caption{AHP Surface Correctness and Verifiability Score}
+		\label{Fig_CorrectnessVerifiability}
+	\end{center}
+\end{figure}
+
+Figure \ref{Fig_CorrectnessVerifiability} shows the surface correctness and verifiability ranking of the software packages using AHP.
+Software packages with a higher score tend to have a visible requirements specification or references to theory documentation.
+They also explicitly use at least one document generation tool that builds confidence of correctness.
+The top ranked software packages all include an easy to follow getting started tutorial, and most of these include expected output.
+Only the top ranked package, Ludwig, provided unit testing.
+It and the second ranked package, ESPResSo, explicitly incorporated continuous integration in the development process.
+During developer interviews it was discovered that some other packages also use continuous integration and unit testing, despite not being explicitly stated in documentation.
+The correctness and verificablity of such packages is not measured well using surface analysis.
+
+The inclusion of requirements specification and theory documentation greatly benefits the correctness and verifiability of software packages.
+The use of document generation tools can help build confidence in correctness.
+The addition of easy to follow tutorials further helps users verify the software and have confidence in its correctness.
+Unit testing documentation and capability, as well as the use of continuous integration tools and techniques such as Bamboo, Jenkins, and Travis CI, help verify correctness.
+Several interviewed developers alluded to difficulty with testing large numbers of features, and some even manually tested program output.
+The use of well defined unit testing tools could decrease the time spent testing some feature.
+
+\subsection{Surface Reliability}
+
+Errors occurred when installing sixteen of the software packages.
+Every instance prompted an error message.
+These messages indicated unrecognized commands (even when following the installation guide), missing links, missing dependencies, syntax errors in code files.
+In some instances the error messages were vague.
+Several automatic installation processes could not find and load dependencies.
+In these instances the installation tried to access outdated external repositories.
+Seven of the installations were recovered and verified, and one of the installations was assumed to be recovered due to the absence of any way to verify it.
+The installation of eight of the software packages could not be recovered.
+Most of these broken installations could not find external dependencies, encountered system incompatibilities, or displayed vague error messages.
+
+Of the thirteen software packages that installed correctly and also have tutorials, four broke during tutorial testing.
+All of these instances resulted in an error message being displayed.
+One error was due to a missing tutorial dependency, another was due to an invalid command despite following the tutorial, and the final two errors were vague execution errors.
+Of the four broken tutorial instances, only the one that was missing a dependency was recoverable.
+
+\begin{figure}[h!]
+	\begin{center}
+		\includegraphics[width=1.0\textwidth]{reliability_chart}
+		\caption{AHP Surface Reliability Score}
+		\label{Fig_Reliability}
+	\end{center}
+\end{figure}
+
+Figure \ref{Fig_Reliability} shows the surface reliability ranking of the software packages using AHP.
+Software packages with a high score either did not break during installation, or the broken installation was recoverable.
+All of the top five ranked packages have tutorials.
+One of these packages, pyLBM, broke during tutorial testing, but a descriptive error message helped in recovery.
+
+Overall, lower ranked software packages are lacking clear documentation, testing or tutorial examples, and descriptive error messages, and have broken dependencies.
+Thus, regarding surface reliability, software packages would benefit from clear up-to-date installation and tutorial documentation, and descriptive error messages.
+
+\subsection{Surface Robustness}
+
+The software packages were tested for handling unexpected input, including incorrect data types, empty input, and missing files or links.
+Success predicated on a reasonable response from the system, including appropriate error messages and an absence of unrecoverable system failures.
+
+\begin{figure}[h!]
+	\begin{center}
+		\includegraphics[width=1.0\textwidth]{robustness_chart}
+		\caption{AHP Surface Robustness Score}
+		\label{Fig_Robustness}
+	\end{center}
+\end{figure}
+
+Figure \ref{Fig_Robustness} shows the surface robustness ranking of the software packages using AHP.
+Software packages with a high score behaved reasonably in response to unexpected input as described above.
+All of the software packages that installed correctly passed this test.
+They output descriptive error messages or did not crash.
+Software packages with a lower surface robustness score had not installed correctly, so their robustness score may not be a true reflection of runtime robustness.
+Similarly, all software packages that installed correctly and require plain text input files correctly handled an unexpected change to these input files, including a replacement of new lines with carriage returns.
+
+\subsection{Surface Performance}
+
+The software packages were not quantitatively measured for surface performance.
+Although the software packages all apply LBM to solve scientific computing problems, the packages focus on varied domains, with varying parameters, and are technically different from each other.
+Due to this, a comparison of performance is not appropriate.
+In this project we instead looked through each software package's artifacts for evidence that performance was considered.
+The artifacts of seventeen software packages mentioned the use of some sort of parallelization tool or technique.
+This included GPU processing and the CUDA parallel computing platform.
+GPUs provide superior processing power and speed compared to CPUs, and are often used for scientific computing when a large amount of data is involved.
+The software package TCLB is implemented in a highly efficient multi-GPU code to achieve performance suitable for model optimization \citep{rutkowski2020open}.
+In the Ludwig package, a so-called mixed mode approach is used where fine-grained parallelism is implemented on the GPU, and MPI is used for even larger scale parallelism \citep{gray2013ludwig}.
+While some software packages require CUDA and GPU processing, some have the option of using either the GPU or the CPU.
+The packages that require GPU and CUDA have better performance at the expense of installability and surface reliability.
+
+\subsection{Surface Usability}
+
+Software package artifacts were reviewed for the presence of a tutorial, a user manual, documented user characteristics, and a user support model.
+In total eighteen software packages have a tutorial and thirteen have a user manual.
+The tutorials vary in scope and substance, and eight include an expected output.
+Most user manuals are in the form of a file that can be downloaded, while some are rendered on a web-page.
+Some packages do not have a user manual but do have useful documentation distributed on their web-pages.
+Expected user characteristics are documented in four software packages.
+Users are typically scientists or engineers trying to test a new application.
+Their background is often physics, chemistry, biophysics, or mathematics.
+All but one of the packages have a user support model, and several of them have multiple avenues of user support.
+The most popular avenue of support is Git, followed by email and forums.
+One software package has a FAQ page.
+
+\begin{figure}[h!]
+	\begin{center}
+		\includegraphics[width=1.0\textwidth]{usability_chart}
+		\caption{AHP Surface Usability Score}
+		\label{Fig_Usability}
+	\end{center}
+\end{figure}
+
+Figure \ref{Fig_Usability} shows the surface usability ranking of the software packages using AHP.
+Software packages with a high score have a tutorial and user manual, sometimes have documented user characteristics, and have at least one user support model.
+Many packages have several user models.
+
+Interviews with developers revealed several usability issues.
+Some users have misunderstood the boundaries of LBM and computational fluid dynamics, and have combined or applied methods that are not physically sound.
+Sometimes users have applied LBM to poorly defined or inappropriate fluid dynamics problems.
+The users do not realize the limitations of the methods, of the software, and do not understand the requirements to properly model a problem with the software.
+The software packages are not designed to be used out of the box in a plug and play fashion.
+Developers of some of the software packages mitigated this by editing the source code to prevent unsound combination from being applied, and by updating the documentation to better inform users of LBM limitations, and of the requirements to properly model appropriate problems, including what algorithms and parameters to use.
+
+Some additional but infrequent software usability issues were commented on by the developers.
+Users have had trouble with installation and understanding how to maneuver the interfaces and how to set up or run models.
+These issues are addressed by various user support models, including frequently asked questions sections on the software package websites, user guides, and hardware and software requirements specifications.
+
+One software package changed its scripting language to Python to make it more usable.
+The developer commented that this was the biggest step in addressing usability, commenting that Python was a much easier language for users to learn and understand.
+
+\subsection{Maintainability}
+
+Software packages were reviewed for the presence of artifacts.
+Every type of artifact or file that is not a code file was recorded.
+The software packages were also reviewed for software release and documentation version numbers.
+This information could be used to better troubleshoot issues and organize documentation.
+All but three software packages have source code release and documentation version numbers.
+
+Information on how code is reviewed, or how to contribute to the project was also noted.
+In total, eleven software packages have this information, which was found in various artifacts, including in developer guides, contributor guides, user guides, developer web-pages, and README files.
+
+Issue tracking is used in twenty two software packages, fifteen of which use Git, six use email, and one uses SourceForge.
+Most software packages that use Git have most of their issues closed, and only three have less than fifty percent of their issues closed.
+Furthermore, most packages that use Git for issue tracking use GitHub as a version control system, one package uses CVS for issue tracking, and eight packages do not appear to use any such system.
+
+Software package code files were further measured for the percentage of code that is comments.
+Packages with a higher percentage of comments were designated as more maintainable.
+Comments represent more than ten percent of code files in fifteen packages, and the average percentage of code comments is about fourteen.
+
+\begin{figure}[h!]
+	\begin{center}
+		\includegraphics[width=1.0\textwidth]{maintainability_chart}
+		\caption{AHP Maintainability Score}
+		\label{Fig_Maintainability}
+	\end{center}
+\end{figure}
+
+Figure \ref{Fig_Maintainability} shows the maintainability ranking of the software packages using AHP.
+Software packages with a high score provide version numbers on documents and source code releases, have an abundance of high quality artifacts, and use an issue tracking tool and version control system.
+These packages also appear to reasonably handle tracked issues, having most of their issues closed.
+Their code files are well commented with more than ten percent of the code being comments.
+
+Interviews with developers revealed that most projects do have a defined process for accepting contributions from team members.
+The packages rarely get contributions from outside developers, but the process would be similar as for the aforementioned group.
+
+Contributions are made through GitHub, and are then reviewed and pulled by lead developers, often with consultation with a group of core developers depending on the organizational model.
+Continuous integration is part of the process for some packages.
+
+Some developers noted that their software package does not have well defined contributing guide in the repository but it might be a good idea to add one in the near future.
+They would be very happy to see contribution from outside of their organization, but currently this does not happen.
+
+Furthermore, maintainability has been addressed by increasing source code modularity, reducing duplicate information, and improving abstraction.
+Several software packages have had large sections of their code base redeveloped with languages that the developers felt are more understandable and readable, and that are better supported.
+Data structures have also been redeveloped and storage has been improved.
+One software developer mentioned that the mathematical foundations that the geometries and models are designed from had been re-evaluated and optimized.
+
+\subsection{Modifiability}
+
+Software packages were not quantitatively measured for modifiability.
+In this project we asked developers to comment on modifiability when we interviewed them.
+Specifically, we asked if ease of future changes to the system, modules, and code blocks was considered when designing the software.
+We also asked if any measures had been taken to ensure the ease of future changes.
+All of the developers that were interviewed noted that the ease of future changes was considered and that measures to ensure it had been taken.
+
+A high degree of code modularity and abstraction was noted by one developer as a measure to ensure the ease of future changes.
+They also noted that some of the code base was transitioned from c to c++, which could ease modifiability of that software package.
+Another developer noted that their software package was designed to easily allow for the addition of some LBM features, but changes to major aspects of the LBM models would be difficult.
+Furthermore, the package was designed with flexible data structures and storage in mind.
+
+Some software packages, like Palabos, provide validation benchmarks for their core fundamental algorithmic ingredients \citep{latt2021palabos}.
+The stated intent of these benchmarks is to showcase the validity and usefulness of the package in order to stimulate the development of third-user extensions.
+The Palabos package identifies as a development framework for modeling problems in various computational fluid dynamics areas.
+
+\subsection{Reusability}
+
+Each software package was measured for the total number of source code files.
+A larger number of source files was associated with increased reusability due to increased modularization.
+The software packages were also reviewed for the presence of API documentation, which indicates that a software package was developed with interaction between other software applications in mind.
+
+\begin{figure}[h!]
+	\begin{center}
+		\includegraphics[width=1.0\textwidth]{reusability_chart}
+		\caption{AHP Reusability Score}
+		\label{Fig_Reusabilty}
+	\end{center}
+\end{figure}
+
+Figure \ref{Fig_Reusabilty} shows the reusability ranking of the software packages using AHP.
+Software packages with a high score have thousands of source code files and API documentation.
+The highest scoring packages, ESPResSo and waLBerla, have extensive functionality, including graphical visualizations as well as modeling that does not use LBM.
+For this reason a comparison with other software packages is not on a level field.
+However, these packages do have an abundance of reusable components.
+
+There was a strong focus on modularity when designing the waLBerla framework in order to enhance productivity, reusability, and maintainability \citep{bauer2021walberla}.
+Its software design has enabled waLBerla to be successfully applied in several projects as a basis for various extensions \citep{bauer2021walberla}.
+
+\subsection{Surface Understandability}
+
+Ten random source code files of each software package were reviewed for several measures.
+This measure of surface understandability may not perfectly reflect each package due to the limitation of only testing ten files.
+
+All of the packages appear to have a consistent indentation and formatting style.
+Only LUMA and HemeLB explicitly identify coding standards that are used during development.
+Generally, the software packages use consistent, distinctive, and meaningful code identifiers.
+Only four packages appear to use vague identifiers.
+Hard coded constants were observed in the source code of twelve packages.
+The constants are used for various parameters, mathematical constants, and matrix definitions.
+All of the packages are well commented and the comments clearly indicate what is being done.
+Domain algorithms are noted in the source code of eleven packages.
+All of the packages are well modularized.
+When observing the source code files, it was found that thirteen of the packages have a consistent style and order for function parameters.
+
+\begin{figure}[h!]
+	\begin{center}
+		\includegraphics[width=1.0\textwidth]{understandability_chart}
+		\caption{AHP Understandability Score}
+		\label{Fig_Understandability}
+	\end{center}
+\end{figure}
+
+Figure \ref{Fig_Understandability} shows the surface understandability ranking of the software packages using AHP.
+Software packages with a high score have a consistent indentation and formatting style, and consistent, distinctive, and meaningful code identifiers.
+They also have hard coded constants, and explicitly identify mathematical and LBM algorithms.
+Their comments are clear and indicate what is being done in the source code.
+The source code is well modularized and structured.
+
+Software developers noted that in their opinion users have generally found their packages to be understandable.
+One developer commented that some users have attempted to run physically incompatible LBM methods, and the solution was to edit the code to prevent such combinations, as well as to update the documentation to prevent misunderstanding the methods.
+Similarly, another developer noted that some users had issues setting up parameters for LBM schemes.
+The solution to this was to update the interface where these parameters are set, as well as to add functionality to test the stability of the parameters.
+A third developer noted that some users lack the background knowledge to easily model fluid dynamics problems using their software.
+A frequently asked questions section was added to their package website to help users find answers to common questions.
+The package also has detailed documentation, including guides and usage requirements specification, to better help user understand the software.
+
+\subsection{Traceability}
+
+Software packages were not quantitatively measured for traceability.
+In this project we asked developers to comment on traceability, specifically on their software package's documentation and how it fit into their development process.
+
+One developer noted that all major additions to their package had accompanying changes to artifacts and documentation.
+They noted that considerable effort had been put into the documentation.
+They further commented that they want to lower the entry barrier for new developers, and because of that their package has a considerable amount of developer documentation.
+This documentation informs developers on how to get started, and orients them to the artifacts, source code, and system architecture, as well as how the software package build system works, and how the coupling between the simulation engine and the interface works.
+
+Another developer noted the importance of documentation for both the users and developers of their software.
+New features are always added to the documentation.
+The developers use documentation to stay up to date on the status of the software package, and to help expand features, like computational models or algorithms.
+This is necessary so that the coding standard for these models is kept consistent with new developers.
+
+The importance of documentation for both users and developers was stressed throughout the interviews.
+However, it was noted several times that a lack of time and funding has a negative affect on the documentation.
+Most of the developers are scientific researchers evaluated on the scientific papers that they produce.
+Writing and updating documentation is something that is done in free time, if that time arises, and if writing documentation is a concern.
+Sometimes it is a last priority for the developers.
+Finding ways to hasten updating documentation would increase the frequency of such updates and benefit both users and developers.
+
+One developer noted the use of documentation generators like Doxygen. It would be advisable for more projects to use such automatic document generation tools, since some projects do not do this.
+
+\subsection{Visibility and Transparency}
+
+Software package artifacts were reviewed for the identification of a specific development model, like a waterfall of agile development model, and the presence of documentation recording the development process and standard.
+They were also reviewed for the identification of the development environment, and the presence of release notes.
+The packages tended to not explicitly use well-known development models.
+This was also noted in the interviews with developers, as detailed below.
+The development teams of these packages are fairly small and easily organized without the need for such processes.
+Seven of the software packages did have some artifacts outlining the general development process, how to contribute, and the status of the package or its components.
+Eight of the packages have artifacts that note the development environment.
+While this information could help developers, and would improve transparency, the small close-knit nature of the development teams make explicitly publicly specifying this information practically unnecessary.
+Version release notes were found in nine of the software packages.
+
+\begin{figure}[h!]
+	\begin{center}
+		\includegraphics[width=1.0\textwidth]{visibilitytransparency_chart}
+		\caption{AHP Visibility and Transparency Score}
+		\label{Fig_VisibilityTransparency}
+	\end{center}
+\end{figure}
+
+Figure \ref{Fig_VisibilityTransparency} shows the visibility and transparency ranking of the software packages using AHP.
+Software packages with a high score have an explicit development model and defined development process.
+They also
+had detailed and easy to access notes accompanying software releases.
+
+Developers were asked to comment on the obstacles in their development process.
+One developer noted that a lot of their source code had been written with a specific application in mind, and that there is too much coupling between components.
+Addressing this issue would help with code modifiability and reusability.
+Updating the development process would help resolve this issue and prevent such issues in the future.
+Improving the visibility of software changes and a peer review process would also help.
+Improving the software engineering education or experience of developers is also an idea that was brought up by several developers.
+Specifically, developers should always write code that is decoupled and modular, and should keep in mind the visibility of their contributions by better updating documentation and ensuring that their contributions are transparent to the rest of the developers.
+This would help catch issues in the contributions, and improve source code maintenance.
+
+According to one developer, some obstacles to the development processes of their package had been overcome by the introduction of continuous integration practices, and a peer review process for contributions.
+These practices decrease development and maintenance times.
+Another developer mentioned that their package had two sets of code, for executing the models on the CPU and GPU, and that maintenance was decreased by introducing macros, a practice which became a common part of the development process.
+
+Developers were also asked how documentation fits into their development process.
+Several developers noted that developer documentation plays an important role in familiarizing potential contributors to the software system architecture.
+Without the guidance that the documentation provides it would be unlikely that contributions would pass the peer review process.
+
+None of the software packages whose developers were interviewed have a formal software development model.
+The packages all have fairly small development teams.
+These teams do accept outside contributors, but generally the teams are tight-knit, often working at the same institution, although one of the packages has an international team.
+One developer noted that while no formal model is used, their development model is something similar to a combination of agile and waterfall development models.
+
+The developers noted similar project management processes.
+For teams of only a couple of developers, the addition of new features or major changes are discussed with the entire team.
+Projects with more than a couple of developers have lead developer roles.
+These lead developers review potential additions to the software.
+The software packages use GitHub for managing the project.
+Typically there are several development branches as well as the master branch.
+
+\subsection{Reproducibility}
+
+Software packages were not quantitatively measured for reproducibility.
+In this project we asked developers to comment on reproducibility when we interviewed them.
+
+Developers were asked if they have any concern that their computational results would not be reproducible in the future, and if they had taken any steps to ensure reproducibility.
+
+One developer noted that they compare the results of their methods against manually calculated results.
+These comparison tests are automatically run for all source code changes.
+The tests are run when a pull request is opened on GitHub.
+Even once these tests are complete, a peer review process is done before changes are fully committed to the appropriate branch.
+The results for all of the LBM schemes on the software package development branch are also compared daily for correctness, ensuring that the system output reflects the expected output.
+
+Several developers noted that they currently do not have a system in place to test for reproducibility, but it is of interest and could be implemented in the future.
+Generally, the mathematical foundations of the models are verified, but the output of the software package is not compared to other output.
+Depending on the package and how it outputs LBM solutions, it may not be practical or feasible.
+A correct output may not be exactly reproducible, as it may be dependent on a probability distribution, so strictly comparing results may not be appropriate.
+
+The source code and artifacts of some software packages may be reproducible.
+There is considerable variance in the quality of software specifications and other developer documentation.
+Some packages are well detailed, and translating the specifications into source code will produce similar results across developers.
+
+\subsection{Unambiguity}
+
+Software packages were not quantitatively measured for unambiguity.
+In this project we asked developers to comment on unambiguity when we interviewed them.
+
+Developers were asked if they thought that the current documentation can clearly convey all necessary knowledge to the users, and if they had taken any steps to ensure clarity.
+
+One developer noted that their documentation was meant for users that are already familiar with the underlying physics and computational fluid dynamics methods, and that these concepts are not explained in detail.
+The documentation focuses solely on how to technically use the software package, and includes a user guide and tutorial walk through of how to set up and run a simulation.
+With this in mind the developer believes that their documentation is in reasonable shape for users with a minimum knowledge of the underlying physics.
+If new users have technical questions these can be addressed in further revisions of the documentation.
+New developers should also find that the documentation is reasonably clear and useful.
+Information that is missing, like detailed explanations of dependencies, is referenced in the documents.
+
+Another developer also noted that their documentation was in reasonable shape, but they would really appreciate more user feedback to improve it.
+They also noted that they believe a lack of knowledge of the underlying physics and computational fluid dynamics concepts can be an issue for some users.
+This information can be referenced in the documentation, but it is not something that the documentation needs to detail.
+
+\subsection{Overall Quality}
+
+Figure \ref{Fig_OverallScore} shows the overall ranking of the software packages using AHP.
+Software packages with an overall high score had ranked high in at least several of the individual qualities that were quantitatively measured.
+
+\begin{figure}[h!]
+	\begin{center}
+		\includegraphics[width=1.0\textwidth]{finalscore_chart}
+		\caption{AHP Overall Score}
+		\label{Fig_OverallScore}
+	\end{center}
+\end{figure}
+
+Looking at the top three ranked packages, ESPResSo had achieved a relative high score in installability, surface correctness and verifiability, surface usability, maintainability, reusability, and visibility and transparency.
+Ludwig scored high in surface correctness and verifiability, surface robustness, surface usability, maintainability, and visibility and transparency.
+Palabos scored high in installability, surface reliability, surface robustness, maintainability, and understandability.
+
+Section \ref{qualityrecommentations} further analyzes these findings and offers some software quality recommendations for future development of LBM software.
+
+\newpage
+
+\section{Answers To Research Questions}\label{answersquestions}
+
+\subsection{Artifacts Present - DONE 1st DRAFT}\label{artifacts}
+
+The software packages were examined for the presence of artifacts, which were then categorized by frequency.
+Below we have grouped them into common, less common, and rare artifacts.
+Common artifacts were found in 16 to 23 of the software packages.
+Less common artifacts were found in 8 to 15 of the software packages.
+Rare artifacts were found in 1 to 7 of the software packages.
+
+\subsubsection{Common Artifacts}
+The following artifacts were commonly found in the 23 software packages that were tested.
+All of the top four AHP ranked packages, ESPREsSo, Ludwig, Palabos, and OpenLB have each of these artifacts, except for a requirements specification or theory notes.
+Only three of the top four ranked packages were found to have such an artifact.
+
+\begin{singlespace}
+\begin{multicols}{2}	
+\begin{itemize}
+	\item Authors/Developers List
+	\item Bug Tracker
+	\item Dependency Notes 
+	\item Installation Guide/Instructions 
+	\item License
+	\item List of Related Articles/Publications 
+	\item Makefile / Build File 
+	\item README File 
+	\item Requirements Specification / Theory Notes
+	\item Tutorial
+\end{itemize}
+\end{multicols}
+\end{singlespace}
+
+These common artifacts contribute to the quality of the software in the following ways.
+A list of authors and developers helps potential users and contributors contact project members to answer questions affecting many software qualities.
+A bug tracker helps with organizing improvements to the software.
+Dependency notes help users install the software, as does an installation guide.
+Makefiles or other automated build files decrease human error in the installation process.
+Licenses promote usage of the software.
+Requirements specifications, linked theory notes, and related articles and publications help make the software more understandable, decrease ambiguity, and help with verifying its correctness.
+README files also help with understandability.
+The tutorials help users become familiar with using the software.
+
+\subsubsection{Less Common Artifacts}
+The following artifacts were less commonly found in the 23 software packages that were tested. The top four AHP ranked packages have most of these artifacts.
+Only one of those four packages does not have a user manual or guide, but it does have a detailed and informative website.
+Another one of the top four packages does not appear to have publicly visible design documentation.
+A third package from the list does not appear to use a version control system.
+It is possible that such a system is used by the package since its website notes package version numbers, but the artifacts do not explicitly state the use of such a system.
+
+\begin{singlespace}
+	\begin{multicols}{2}	
+		\begin{itemize}
+			\item Change Log / Release Notes
+			\item Design Documentation
+			\item Functional Specification/Notes	
+			\item Performance Information/Notes
+			\item Test Plan/Report/Script/Data/Cases
+			\item User Manual/Guide
+			\item Version Control
+		\end{itemize}
+	\end{multicols}
+\end{singlespace}
+
+These artifacts also contribute to software quality.
+A change log or release notes improve traceability of the software.
+Design documentation helps with maintaining, modifying, and reusing the software.
+Version control also helps to improve these qualities, as well as with traceability.
+Functional specifications and notes clarify the software, contributing to usability and understandability.
+A user manual also helps with those qualities and with installability, and correctness and verificability.
+Performance notes suggest that performance was considered when developing the software.
+Test plans, scripts, and cases, help verify correctness.
+
+\subsubsection{Rare Artifacts}
+The following artifacts were rarely found in the 23 software packages that were assessed.
+It is not common for the top four AHP ranked packages to have many of these artifacts.
+None of the top four packages have any explicit API documentation.
+Three of these packages have information on contributing to the project.
+Two of them have a FAQ section or forum.
+One has verification and validation notes, and a video guide explaining how to use the software.
+
+\begin{singlespace}
+	\begin{multicols}{2}	
+		\begin{itemize}
+			\item API Documentation
+			\item Developer/Contributor Manual/Guide
+			\item FAQ / Forum
+			\item Verification and Validation Plan/Notes
+			\item Video Guide (including YouTube)
+		\end{itemize}
+	\end{multicols}
+\end{singlespace}
+
+Although these artifacts were rarely found in our set of LBM software, they also contribute to software quality.
+API documentation helps with reusing the software.
+Developer and contributor manuals and guides help with maintainability, visibility and transparency.
+FAQs and forums improve usability of software.
+Verification and validation notes can be used to help check if a system meets specifications.
+Video guides can contribute to many software qualities, depending on the content of the video.
+
+\subsection{Tools Used - DONE 1st Draft}
+Software tools are used to support the development, verification, maintenance, and evolution of software, software processes, and artifacts \citep{ghezzi1991fundamentals}.
+Many tools are used by LBM software packages.
+The tools noted here are subdivided into development tools, dependencies, and project management tools.
+
+\subsubsection{Development Tools}
+
+These tools support the development of end products but do not become part of them, unlike dependencies that remain in the application once it is released \citep{ghezzi1991fundamentals}.
+The following type of development tools were explicitly noted in the artifacts or web-pages of the 23 LBM packages that were assessed.
+It is likely that other tools, such as debuggers, were used but are not specified here.
+
+\begin{singlespace}
+	\begin{multicols}{2}	
+		\begin{itemize}
+			\item Continuous Integration Tools
+			\item Code Editors
+			\item Development Environment
+			\item Runtime Environments
+			\item Compilers
+			\item Unit Testing Tools
+			\item Correctness Verification Tools
+		\end{itemize}
+	\end{multicols}
+\end{singlespace}
+
+The above tools can verify the correctness of software during its development.
+Only two of the software packages that were assessed mentioned using continuous integration tools like Travis CI.
+Code editors and compilers were explicitly noted to have been used by several packages, and were likely used by all of them.
+One of the packages explicitly noted the use of unit testing, but did not specify the tool that was used.
+Likewise, the use of a tool for verifying the correctness of output was noted by one of the developers.
+It is likely that similar tools were used when developing other software packages.
+
+\subsubsection{Dependencies}
+
+The following types of dependencies were explicitly noted in the artifacts or web-pages of the 23 LBM packages that were assessed.
+It is possible that other types of dependencies are part of these software packages but are not clearly specified in their artifacts or web sites and because of that they are not listed here.
+
+\begin{singlespace}
+	\begin{multicols}{2}	
+		\begin{itemize}
+			\item Build Automation Tools
+			\item Technical Libraries
+			\item Domain Specific Libraries
+		\end{itemize}
+	\end{multicols}
+\end{singlespace}
+
+Most of the software packages use some sort of build automation tools, most commonly Make.
+They also all use various technical and domain specific libraries.
+Technical libraries include visualization, data analysis, and message passing libraries.
+Domain specific libraries are scientific computing libraries.
+Libraries that are explicitly stated in artifact, and were not noted during our observations, may fall outside of these categories.
+
+\subsubsection{Project Management Tools}
+
+Many of the software packages that were assessed were developed by teams of two or more people.
+Their work needed to be coordinated and managed.
+The following types of project management tools were explicitly noted in the artifacts, web-pages, or interviews with the developers of the 23 LBM packages that were assessed.
+As with development tools and dependencies, it is possible that other types of project management tools were used to coordinate and manage the projects but are not specified and because of that they are not listed here.
+
+\begin{singlespace}
+	\begin{multicols}{2}	
+		\begin{itemize}
+			\item Collaboration Tools
+			\item Email
+			\item Change Tracking Tools
+			\item Version Control Tools
+			\item Document Generation Tools
+		\end{itemize}
+	\end{multicols}
+\end{singlespace}
+
+Collaboration tools are noted as being used when developing the software projects.
+Most often email and video conferencing is used.
+Project management software was not explicitly mentioned, but it is possible that some of the projects use such software.
+Many of the projects are located on GitHub, and its developers use the platform to help manage their projects, especially bug related issues.
+Most of the projects appear to use change tracking and version control tools.
+They often use GitHub for this, and a few use Git or CVS.
+Document generation tools are mentioned in the artifacts of 12 of the projects.
+Tools such as Sphinx and Doxygen are explicitly used in this capacity.
+Drasil could be considered as an alternative to generate artifacts since it is geared specifically towards research software domains like LBM software.
+
+\subsection{Principles, Processes, and Methodologies - DONE 1st DRAFT}
+
+Most of the software packages do not explicitly state in their artifacts the motivations or design principles that were considered when developing the software.
+One package, Sailfish, indicates that shortening the development time was considered in early stages of design, with the developers opting for using Python and CUDA/OpenCL to achieve this without sacrificing any computational performance.
+The goals of that project are explicitly listed as performance, scalability, agility and extendability, maintenance, and ease of use.
+The project scored well in these categories during our assessment.
+
+Processes, like methods, are ways of doing things, especially in an orderly way; while methodologies are defined as systems of methods \citep{ghezzi1991fundamentals}.
+It is not explicitly indicated in the artifacts of most of the packages that development involved following any specific model, like a waterfall or agile development model.
+One developer noted that while no formal model is used, their development model is something similar to a combination of agile and waterfall development models.
+The developer teams of the LBM packages are fairly small and easily organized without the need for such models.
+
+Seven of the software packages contain artifacts outlining the general development process, like basic instructions on how to contribute.
+Most of the development teams accept outside contributors, but generally the teams are centralized, often working at the same institution.
+
+The developers that were interviewed all noted similar project management processes.
+In teams of only a couple of developers, additions of new features or major changes are discussed with the entire team.
+Projects with more than a couple developers have lead developer roles.
+These lead developers review potential additions to the software.
+One of the developers that was interviewed noted that a peer review process is used to assess major changes and additions.
+
+Many software packages use GitHub for managing the project. Typically there are several simultaneous development branches.
+
+Documentation was also noted as playing a significant role in the development process, specifically with on-boarding new developers.
+A goal of documentation is to lower the entry barrier for these new contributors.
+The documentation provides information on how to get started, orients the user to artifacts and the source code, and explains how the system works, including the so-called simulation engine and interface.
+The use of document generation tools is mentioned in the artifacts of 12 software packages, and was noted during interviews with developers.
+Document generation tools that were mentioned include Sphinx and Doxygen.
+
+Two types of software changes were discussed during interviews with developers.
+One is feature additions, which arise from a scientific or functional need.
+The development of such changes was noted to involve more formal discussions within the development team and the addition of them into the main software branch involves lead developers.
+The other change type is code refactoring, which only sometimes involves formal discussions with the development team.
+New developers were noted to play an increased role in these changes compared to the former changes.
+Software bugs are typically addressed in a similar fashion as code refactoring, and issue tracking is commonly used to manage these changes.
+
+Interviews with the developers of software packages also revealed a more frequent use of both unit testing and continuous integration in the development process than was found by only assessing the artifacts.
+The use of automatic installation processes is also common.
+Most often this involved a Make script.
+
+\subsection{Pain Points - DONE 1st Draft}
+
+Developers were asked to comment on obstacles in their development process, obstacles encountered by users, and potential future obstacles.
+
+Several development process obstacles were identified.
+One developer noted that their small development team has a lack of time to implement new features.
+Small development teams are common for LBM software packages.
+Team members are almost always part of the same institute or already know each other from other projects.
+External contributions are rare despite many of the projects accepting them.
+
+A lack of software development experience was noted by another developer.
+Many of the team members on their project are domain experts and there is a steep learning curve to get these team members to contribute good quality source code.
+It was further noted that this has been somewhat addressed, as the code has been re-written to best ensure ease of future contributions.
+The same developer also noted that there is a lack of incentive and funding in academia for developing widely used scientific software.
+There are no journals that publish such source code.
+There is no organization helping with the development of good quality software.
+Scientific software is often developed in-house by the very researchers that temporarily use it in their own research.
+This software has a defined user and purpose, and often does not meet the standards that would be required by external users.
+Such software is rarely commercialized, in which case it is not open-source.
+
+Documentation of LBM software is noted as being important, but its quality could be improved.
+There is often no time or funding for maintaining quality documentation for software that is rarely used outside of the development team.
+Furthermore, the documentation generally only provides a shallow overview of the underlying computational fluid dynamics theory.
+Users would be well advised to already be familiar with these topics to or they should spend significant time referencing theory resources.
+The documentation instead generally focuses on explaining how to use the software.
+It is of course not feasible for package documentation to address the underlying metaphysics topics in detail, so it is advised that the package documentation links to resources that better explain the underlying theory.
+Sometimes, frequently asked questions about the underlying theory are answered in the documentation.
+
+Setting up parallelization was also noted as a technical pain point by one of the developers, and the introduction of continuous integration by another.
+Software development knowledge could mitigate such pain points.
+As already noted, many of the developers are domain experts and not professional software developers.
+One of the interviewed developers that has relatively more extensive software development experience noted that eliminating equivalent statements using macros had helped improve the quality of their source code, specifically helping with reusing code to run on both the CPU and GPU.
+
+Difficulties with ensuring correctness were also noted by several developers.
+They indicated that tests are run on all new source code additions, testing both individual modules and the system to verify correctness.
+These tests compare the package output to known correct output using test cases.
+A developer commented that the amount of testing data that is needed for some cases is a problem as free testing services do not offer capabilities to store and process such large amounts of data, and in-house testing solutions needed to be created to address this limitation.
+
+A few obstacles related to users were found.
+Several developers noted that users sometimes try to use incorrect LBM method combinations to solve their problems.
+Furthermore, some users think that the packages will work out of the box to solve their cases, while the packages require both a good understanding of computational fluid dynamics and an understanding of the requirements for formulating problems in the individual packages, which can be a significant endeavor.
+These software packages are not like commercial software packages.
+They are generally set up to solve specific research problems, and are otfen primarily used by their developers.
+While they are modifiable to solve similar problems, these modifications are not trivial.
+So far this problem of on-boarding new users has been addressed by updating the documentation to better inform users of the underlying LBM theory and package requirements.
+Similar issues with LBM parameters were noted by another developer.
+Updating the user interface to better explain theoretical principles, as well as test user input for compatibility, was the implemented solution.
+As noted above, sometimes frequently asked questions on the underlying theory and on how to use the software are answered in the documentation.
+
+One of the interviewed developers commented that parts of their package's source code had been refactored to Python in order to help address usability issues.
+Python was perceived as a much more usable language, and it would be easy for future users and developers to learn and understand the source code.
+
+A few potential future obstacles were noted.
+One developer said that their source code had been written with a specific application in mind and that due to this there was too much coupling between components of the source code.
+This would have an impact on future modifiability and reusability when trying to extend the software, and the code would need to be refactored.
+
+As noted above, difficulties with ensuring future correctness could also arise. As new methods and functionality is added into the software, new test cases and test data will need to be developed.
+
+\subsection{Quality Recommendations - DONE 1st Draft}\label{qualityrecommentations}
+
+The following points regarding software quality should be considered when developing LBM software packages.
+
+\subsubsection{Installability}
+
+\begin{itemize}
+	\item Include OS compatibility, including specific OS versions.
+	\item Include complete installation instructions.
+	\item Installation instructions should be written as if the user does not have any dependencies installed and is installing on a clean OS.
+	\item List all dependencies in the installation instructions, dependency versions, and how to install them.
+	\item If possible, automate the installation of dependencies.
+	\item Automate the installation process as much as possible.
+	\item The installation instruction should only be in one location.
+	\item Include descriptive error messages for errors encountered during installation.
+	\item Include a way to validate the installation.
+	\item Include instructions for uninstalling the software.
+\end{itemize}
+
+\subsubsection{Surface Correctness and Verifiability}
+
+\begin{itemize}
+	\item Use a requirements specification document.
+	\item Make public the requirements specification document, or explicitly reference the domain theory that the software is designed from.
+	\item Ensure the above information is easy to find. Consider adding it to the user manual. 
+	\item Development teams should include both domain experts and experienced software developers.
+	\item Use and make public detailed documentation. Consider using automatic document generation tools.
+	\item Include detailed tutorials that include expected output.
+	\item Use unit testing during development and make them public.
+	\item Modularize the source code.
+	\item Use continuous integration tools and processes during development.
+\end{itemize}
+
+\subsubsection{Surface Reliability}
+
+\begin{itemize}
+	\item Include descriptive error messages where appropriate.
+	\item In case automatic installation of dependencies fails, the system should indicate to the user what dependencies need to be installed manually.
+	\item The packages should include detailed tutorials, including dependencies, expected output, and any additional supplementary documentation that may be required.
+\end{itemize}
+
+\subsubsection{Surface Robustness}
+
+\begin{itemize}
+	\item The system must provide descriptive error messages when it encounters unexpected input, including incorrect data types, empty input, and missing files or links.
+\end{itemize}
+
+\subsubsection{Surface Performance}
+
+\begin{itemize}
+	\item Integrate parallelization tools and techniques in order to reduce processing time. Consider GPU processing, CUDA, and MPI. 
+	\item The user should be able to choose to process their model on either the CPU or GPU.
+\end{itemize}
+
+\subsubsection{Surface Usability}
+
+\begin{itemize}
+	\item Include user hardware and software requirements documentation.
+	\item Include a user tutorial that indicates the expected output.
+	\item Include a detailed user manual. It should identify elements of user interfaces, and identify all requirements to model a system.	
+	\item State appropriate fluid dynamics problems that the software is designed to model in the documentation, and explicitly state the limits of the software. 
+	\item Provide documentation that details the background theory information, or provide a reference to such information.
+	\item Identify expected user characteristics.
+	\item Keep all documentation in one location.
+	\item Do not duplicate artifact information.
+	\item Maintain a user support model (Git, email, forum, FAQ)
+	\item If possible, consider using popular user-friendly software languages like Python. Especially consider this for parts of the source code that is likely to be modified or reviewed by users.
+\end{itemize}
+
+\subsubsection{Maintainability}
+
+\begin{itemize}
+	\item Keep artifacts updated.
+	\item Ideally include most of the common and less common artifacts listed in Section \ref{artifacts}.
+	\item Include version numbers and release notes for all major source code and artifact releases. 
+	\item Have a defined process for accepting contributions.
+	\item Include documentation for making contributions to the project. 
+	\item Use an issue tracker (Git, email, SourceForge, other) to manage bugs and changes. Issues should be regularly reviewed and closed. 
+	\item Use a version control system (GitHib, CVS).
+	\item Source code needs to be well commented. Typically, more than ten percent of LBM package source code is comments. 
+	\item Ensure source code is modular. Generally, this means that source code is split into files along functionality.
+	\item Eliminate code duplication.
+	\item If possible, consider using popular user-friendly software languages like Python. Especially consider this for parts of the source code that is likely to be modified or reviewed by users.
+\end{itemize}
+
+\subsubsection{Modifiability}
+
+\begin{itemize}
+	\item Ensure a high degree of code modularity and abstraction.
+	\item If possible, consider using popular user-friendly software languages like Python. Especially consider this for parts of the source code that is likely to be modified or reviewed by users.
+	\item Consider future source code modifiabilility as early as the design stage of development.
+	\item Consider flexibility of data structures and data storage in the design stage.
+	\item Provide validation benchmark tools and data to the public.
+\end{itemize}
+
+\subsubsection{Reusability}
+
+\begin{itemize}
+	\item Ensure a high degree of code modularity and abstraction.
+	\item Document module interfaces in design and developer documentation.
+	\item Provide API documentation, if applicable.
+\end{itemize}
+
+\subsubsection{Surface Understandability}
+
+\begin{itemize}
+	\item Maintain consistent formatting style across source code files.
+	\item Identify a coding standard in the documentation, if a coding standard is used during development.
+	\item Use consistent, distinctive, and meaningful code identifiers. 
+	\item Identify algorithms that are used.
+	\item Hard code domain constants.
+	\item Add meaningful comments. They should indicate what is being done in each section of source code.
+	\item Ensure a high degree of code modularity and abstraction.
+	\item Include user guides, and state technical and problem modeling requirements.
+	\item State problem cases that the software is designed to model in the documentation, and explicitly state the limits of the software. 
+	\item Consider adding a FAQ section to the documentation. 
+\end{itemize}
+
+\subsubsection{Traceability}
+
+\begin{itemize}
+	\item Update all relevant documentation when a change to the software is made.
+	\item Use automatic document generation tools to limit the time spent on updating documentation.
+	\item Provide a developers guide.
+	\item Provide a resource that helps orient users and developers to the artifacts, source code, and system architecture.
+	\item Highlight new features and changes between artifact versions. 
+\end{itemize}
+
+\subsubsection{Visibility and Transparency}
+
+\begin{itemize}
+	\item Identify the development model that was used, if one was used.
+	\item Identify the development environment.
+	\item Identify the development process and how new users can contribute. 
+	\item Include notes with all releases.
+	\item Ensure a high degree of code modularity and abstraction.
+	\item Communicate all changes to the development team.
+	\item Use continuous integration processes and tools.
+	\item Consider peer review processes to assess contributions. 
+	\item Eliminate code duplication.
+	\item Update all relevant developer documentation when a change to the software is made.
+	\item Use project management software.
+\end{itemize}
+
+\subsubsection{Reproducibility}
+
+\begin{itemize}
+	\item Test output against manually calculated or known correct results.
+	\item Automate the testing of output. Run tests after all code changes.
+	\item Consider peer review processes to assess contributions. 
+	\item Update all relevant design documentation when a change to the software is made.
+\end{itemize}
+
+\subsubsection{Unambiguity}
+
+\begin{itemize}
+	\item Documentation must state all technical requirements and software dependencies. 
+	\item A user guide must be provided.
+	\item State appropriate fluid dynamics problems that the software is designed to model in the documentation, and explicitly state the limits of the software. 
+	\item The documentation should either explain the underlying CFD theories or provide a reference to appropriate resources.
+	\item Consider asking users for feedback on the documentation, specifically if it clearly conveys all necessary information to the users.
+\end{itemize}
+
+\subsection{Designation Comparison}
+
+answers last question of research questions:
+
+How does software designated as high quality by this methodology compare	with top rated software by the community?
+
+\newpage
+\section{Conclusion}
+
+- key findings restated
+- comment on future work in state of the practice SCS
+
+the following was moved from the former background section, may need edits as part of conclusion:
+
+  The software qualities that are assessed in this report are listed in Section \ref{softwarequalities}.
+Ensuring software quality can take significant time, but as \cite{SmithEtAl2016} shows, SCS developers often cannot see themselves meeting the required time constraints to better develop some of the artifacts that improve qualities, like detailed documentation.
+An inability to develop such documentation due to time constraints could be mitigated by the use of automatic document generation tools like Drasil.
+The Drasil Framework consists of a collection of Domain Specific Languages (DSL) for capturing scientific documents, structures, and computing
+knowledge, and then transforming this knowledge into relevant software artifacts without having to manually duplicate knowledge into multiple artifacts \citep{zhao2018}.
+
+\newpage
+\section{Appendix}
+\begin{singlespace}
 

### 2021-06-17 StateOfPractice/Peter-Notes/Report/Report.tex
fixed error in LBM measurement template

@@
 AHP results for qualities that were quantitatively measured are also found in this section.
-Quality definitions are found in section \ref{softwarequalities}.
+Quality definitions are found in Section \ref{softwarequalities}.
 
@@
 
-All of the software packages that were tested have installation instructions.
-As noted previously, many of the twenty three software packages that were part of the original long list of forty six software packages but were later removed do not include documentation or installation instructions.
-
-All twenty three packages on the short list can be installed on some Unix-like systems.
+All of the 23 software packages that were tested have installation instructions.
+As noted previously, many of the 23 software packages that were part of the original long list of 45 packages were removed due to not including documentation or installation instructions.
+
+All 23 packages on the short list can be installed on some Unix-like systems.
 Seven packages could be installed on Windows, and five on macOS.
-Operating system compatibility is found in the documentation of nineteen software packages.
-All but one of the software packages were tested on Ubuntu for this state of the practice assessment.
-The remaining package was tested on CentOS.
+Operating system compatibility is found in the documentation of 19 software packages.
+All but one of the software packages, TCLB, were tested on Ubuntu for this state of the practice assessment.
+TCLB was tested on CentOS, which is mentioned in its installation instructions.
 
 Of the software packages that were tested, most installation instructions are located in one place, often in an instruction manual or on a web-page.
-Sometimes incomplete installation instructions are found on a home page, with more detailed instructions located on another web-page or within formal documentation.
-Maintainability and correctness of these instructions could be improved if they were limited to one location.
-
-All but two of the software packages have automated some of the installation process. Most of these packages use Make to automate the installation, and a few of them use custom scripts.
+Sometimes, like with Ludwig, incomplete installation instructions are found on a home page, with more detailed instructions located on another web-page, or within the documentation.
+Maintainability and correctness of these instructions could be improved if all the instructions were in one location.
+
+All but one of the software packages (LatBo.jl) have automated at least some of the installation process.
+Most of these packages, such as waLBerla and SunlightLB, use Make to automate the installation, and a few of them, like lbmpy, use custom scripts.
 
@@
 
-Sixteen software packages require less than ten dependencies to be installed.
-All but one software package require less than twenty dependencies.
+Sixteen software packages require less than 10 dependencies to be installed.
+All but one software package require less than 20 dependencies.
 Some packages may automatically install additional dependencies in the background.
@@
 
-Sixteen software packages have less than ten manual installation steps.
-If dependencies are installed in one command then none of the software packages take more than twenty steps to install.
+Sixteen software packages have less than 10 manual installation steps.
+If dependencies are installed in one command then none of the software packages take more than 20 steps to install.
 The average number of steps is about eight, and the fewest is three.
@@
 
-Document generation tools are explicitly used by twelve software packages. Sphinx is used by eight of them, and Doxygen is used by seven. Several of the packages use both.
+Document generation tools are explicitly used by 12 software packages. Sphinx is used by eight of them, and Doxygen is used by seven. Several of the packages use both.
 
@@
 
-Errors occurred when installing sixteen of the software packages.
+Errors occurred when installing 16 of the software packages.
 Every instance prompted an error message.
@@
 
-Of the thirteen software packages that installed correctly and also have tutorials, four broke during tutorial testing.
+Of the 13 software packages that installed correctly and also have tutorials, four broke during tutorial testing.
 All of these instances resulted in an error message being displayed.
@@
 In this project we instead looked through each software package's artifacts for evidence that performance was considered.
-The artifacts of seventeen software packages mentioned the use of some sort of parallelization tool or technique.
+The artifacts of 17 software packages mentioned the use of some sort of parallelization tool or technique.
 This included GPU processing and the CUDA parallel computing platform.
@@
 Software package artifacts were reviewed for the presence of a tutorial, a user manual, documented user characteristics, and a user support model.
-In total eighteen software packages have a tutorial and thirteen have a user manual.
+In total 18 software packages have a tutorial and 13 have a user manual.
 The tutorials vary in scope and substance, and eight include an expected output.
@@
 Information on how code is reviewed, or how to contribute to the project was also noted.
-In total, eleven software packages have this information, which was found in various artifacts, including in developer guides, contributor guides, user guides, developer web-pages, and README files.
-
-Issue tracking is used in twenty two software packages, fifteen of which use Git, six use email, and one uses SourceForge.
-Most software packages that use Git have most of their issues closed, and only three have less than fifty percent of their issues closed.
+In total, 11 software packages have this information, which was found in various artifacts, including in developer guides, contributor guides, user guides, developer web-pages, and README files.
+
+Issue tracking is used in 22 software packages, 15 of which use Git, six use email, and one uses SourceForge.
+Most software packages that use Git have most of their issues closed, and only three have less than 50 percent of their issues closed.
 Furthermore, most packages that use Git for issue tracking use GitHub as a version control system, one package uses CVS for issue tracking, and eight packages do not appear to use any such system.
@@
 Packages with a higher percentage of comments were designated as more maintainable.
-Comments represent more than ten percent of code files in fifteen packages, and the average percentage of code comments is about fourteen.
+Comments represent more than 10 percent of code files in 15 packages, and the average percentage of code comments is about 14.
 
@@
 These packages also appear to reasonably handle tracked issues, having most of their issues closed.
-Their code files are well commented with more than ten percent of the code being comments.
+Their code files are well commented with more than 10 percent of the code being comments.
 
@@
 Ten random source code files of each software package were reviewed for several measures.
-This measure of surface understandability may not perfectly reflect each package due to the limitation of only testing ten files.
+This measure of surface understandability may not perfectly reflect each package due to the limitation of only testing 10 files.
 
@@
 Only four packages appear to use vague identifiers.
-Hard coded constants were observed in the source code of twelve packages.
+Hard coded constants were observed in the source code of 12 packages.
 The constants are used for various parameters, mathematical constants, and matrix definitions.
 All of the packages are well commented and the comments clearly indicate what is being done.
-Domain algorithms are noted in the source code of eleven packages.
+Domain algorithms are noted in the source code of 11 packages.
 All of the packages are well modularized.
-When observing the source code files, it was found that thirteen of the packages have a consistent style and order for function parameters.
+When observing the source code files, it was found that 13 of the packages have a consistent style and order for function parameters.
 
@@
 	\item Use a version control system (GitHib, CVS).
-	\item Source code needs to be well commented. Typically, more than ten percent of LBM package source code is comments. 
+	\item Source code needs to be well commented. Typically, more than 10 percent of LBM package source code is comments. 
 	\item Ensure source code is modular. Generally, this means that source code is split into files along functionality.

### 2021-06-17 StateOfPractice/Peter-Notes/Report/Report.tex
fixed typo in charts, updated Results Installability sebsection

@@
 Errors encountered during the installation process were often quickly fixed thanks to descriptive error messages.
-Systems that provided vague error messages were more difficult to troubleshoot.
-Only three software packages that displayed a descriptive error message were not recoverable, and most of these instances were due to hardware and operating system incompatibility, such as the requirement of CUDA.
+Systems that provided vague error messages, such as those that did not specify which action or file was at fault, were more difficult to troubleshoot.
+Only three software packages (HemeLB, LB3D, lbmpy) that displayed a descriptive error message were not recoverable, and most of these instances were due to hardware and operating system incompatibility, such as the requirement of CUDA.
 Fourteen software packages definitively broke during installation.
-A couple of packages did not provide a definitive message of the success or failure of installation.
+Some packages, such as LB2D\_Prime and LB3D\_Prime, did not provide a definitive message of the success or failure of installation.
 In these instances, validating the installation required performing a tutorial or running a script, as described below, if these were available.
 
-About half of the installation instructions are not written as if the person doing the installation has none of the dependent packages installed, and many software packages do not list all of their dependencies.
-Many dependencies needed to be manually installed.
-Sometimes only an error message during the installation process informs the user of the requirement of these packages.
+About half of the installation instructions are written as if the person doing the installation has none of the dependent packages installed.
+It is common for software packages to not list all of their dependencies despite listing some.
+This was the case for many packages, including ESPResSo++, Ludwig, and LUMA.
+Sometimes only an error message during the installation process informs the user of the requirement of these additional packages.
 A detailed rewrite of the installation instructions from the point of view of installation on a clean operating system is suggested.
+A clean environment can be achieved for testing purposes by using a virtual machine.
 
 Sixteen software packages require less than 10 dependencies to be installed.
-All but one software package require less than 20 dependencies.
+All but one software package (LatBo.jl) require less than 20 dependencies.
 Some packages may automatically install additional dependencies in the background.
@@
 
-All but six of the software packages have a way to verify the installation.
-Twelve have some sort of tutorial examples that can be run by the user.
-Most of these tutorials provide expected output.
-Some packages do not have a way to verify the output of the tutorials, and the user is left to assume correctness if no error message is encountered.
-Some other ways of installation validation include validation scripts, automatic validation after the installation, and instructions to manually review the file system.
-
-Uninstallation instructions were found for only one of the software packages.
+All but six (ESPResSo, HemeLB, laboetie, LB3D-Prime, lbmpy, waLBerla) of the software packages have a way to specifically verify the installation.
+Most have some sort of tutorial examples that can be run by the user.
+Some other ways of installation validation include validation scripts (LB2D-Prime, lettuce, Ludwig, LUMA), automatic validation after the installation (LatBo.jl), and instructions to manually review the file system (LIMBES).
+
+Uninstallation instructions were found for only one of the software packages, pyLBM.
 
@@
 Figure \ref{Fig_Installability} shows the installability ranking of the software packages using AHP.
-Software packages with a higher score tend to have one set of linear installation instructions that are written as if the person doing the installation has none of the dependencies installed.
+Software packages with a higher score (MechSys, Palabos, ESPResSo, OpenLB) tend to have one set of linear installation instructions that are written as if the person doing the installation has none of the dependencies installed.
 The instructions often list compatible operating system versions and include instructions for the installation of dependencies.
@@
 
-Tutorials are available for all but five of the software packages.
+Tutorials are available for 18 of the software packages.
 Generally they are linearly written and easy to follow.
@@
 
-Looking at the top three ranked packages, ESPResSo had achieved a relative high score in installability, surface correctness and verifiability, surface usability, maintainability, reusability, and visibility and transparency.
-Ludwig scored high in surface correctness and verifiability, surface robustness, surface usability, maintainability, and visibility and transparency.
+Looking at the top three ranked packages, Ludwig scored high in surface correctness and verifiability, surface robustness, surface usability, maintainability, and visibility and transparency.
+ESPResSo had achieved a relative high score in installability, surface correctness and verifiability, surface usability, maintainability, reusability, and visibility and transparency.
 Palabos scored high in installability, surface reliability, surface robustness, maintainability, and understandability.

### 2021-06-19 StateOfPractice/Ao-Report/12Background.tex
Update 12Background.tex

--- 2021-05-30 StateOfPractice/Ao-Report/12Background.tex
@@
 
+This chapter introduces several different categories of software, based on which we designed the processes to select the software domain and software candidates in Chapter \ref{ch_methods}.
+It also covers the software quality definitions used in the grading template in Appendix \ref{ap_grading_template} and an overview of the Analytic Hierarchy Process (AHP), a tool we used to compare and grade software products.
+
 \section{Software Categories}
+We usually target specific software categories to narrow down the scopes when selecting software domains and software packages.
 In this section, we discuss three common software categories that are mentioned in Section \ref{sec_software_selection}, and also SC software.
+Section \ref{sec_domain_selection} and Section \ref{sec_software_selection} present more details about why we prefer some of these categories.
 
@@
 For Open Source software (OSS), its source code is openly accessible, and users have the right to study, change and distribute it under a license granted by the copyright holder.
-For many OSS projects, the development process is based on the collaboration of different contributors worldwide \cite{Corbly2014}.
+For many OSS projects, the development process relies on the collaboration of different contributors worldwide \cite{Corbly2014}.
+Accessible source code usually exposes more ``secrets'' of a software project, such as the underlying logic of software functions, how developers achieve their works, and the flaws and potential risks in the final product.
+Thus, it brings much more convenience to the researchers analyzing the qualities of the project.
 
@@
 The term \textit{freeware} should not be confused with \textit{free software}, which is similar to OSS but with a few differences.
+To the end-users, the differences between freeware and OSS often do not bother them.
+The fact that these products are free of charge is likely to make them popular with many users.
+However, software developers, end-users who wish to modify the source code, and researchers looking for inner characteristics may find the inaccessible source code to be a problem.
 
@@
 However, some commercial software is also free of charge \cite{GNU2019}.
+Based on our experience, most commercial software products are not OSS.
+
+For some specific software, the backgrounds of commercial software developers often differ from the ones of non-commercial OSS.
+In such a case, the former is usually the product of software engineers, and the latter is likely to have developers who work in the domain and are also end-users of the products.
+One example is software in Scientific Computing (SC), since the developers need to utilize their domain-specific during the development process \cite{WilsonEtAl2014}.
 
 \subsection{Scientific Computing Software}
-Software development in Scientific Computing (SC) depends on the knowledge of three areas - the knowledge of a specific engineering or science domain, the ability to mathematically build models and applying algorithms, and the capability to implement theoretical models and algorithms with computational tools.
+Software development in SC depends on the knowledge of three areas - the inside of a specific engineering or science domain, the ability to mathematically build models and apply algorithms, and to implement theoretical models and algorithms with computational tools.
 SC software is built with mathematical and computational tools to serve the purpose of solving scientific problems in a domain \cite{Mehta2015}.
+However, the majority of scientists developing their software are self-taught programmers \cite{WilsonEtAl2014}, so there may be a bigger Room for software quality improvement in SC domains.
 
@@
 \begin{itemize}
+\label{def_installability}
 \item \textbf{Installability} The effort required for the installation, uninstallation, or reinstallation of a software or product in a specified environment.
+\label{def_correctness_verifiability}
 \item \textbf{Correctness \& Verifiability} A program is correct if it behaves according to its stated.
 Verifiability is the extent to which a set of tests can be written and executed, to demonstrate that the delivered system meets the specification.
+\label{def_reliability}
 \item \textbf{Reliability} The probability of failure-free operation of a computer program in a specified environment for a specified time, i.e.
 the average time interval between two failures also known as the mean time to failure (MTTF).
+\label{def_robustness}
 \item \textbf{Robustness} Software possesses the characteristic of robustness if it behaves ``reasonably'' in two situations: i) when it encounters circumstances not anticipated in the requirements specification, and ii) when the assumptions in its requirements specification are violated.
+\label{def_usability}
 \item \textbf{Usability} The extent to which a product can be used by specified users to achieve specified goals with effectiveness, efficiency, and satisfaction in a specified context of use.
+\label{def_maintainability}
 \item \textbf{Maintainability} The effort with which a software system or component can be modified to i) correct faults; ii) improve performance or other attributes; iii) satisfy new requirements.
+\label{def_reusability}
 \item \textbf{Reusability} The extent to which a software component can be used with or without adaptation in a problem solution other than the one for which it was originally developed.
+\label{def_understandability}
 \item \textbf{Understandability} (To be completed)
-\item \textbf{Visibility \& Transparency} The extent to which all of the steps of a software development process and the current status of it are conveyed clearly.
+\label{def_visibility_transparency}
+\item \textbf{Visibility/Transparency} The extent to which all of the steps of a software development process and the current status of it are conveyed clearly.
 \end{itemize}
@@
 \section{Analytic Hierarchy Process}
-The Analytic Hierarchy Process (AHP) was developed by Thomas L.
-Saaty, and it has been widely used to make and analyze multiple criteria decisions \cite{VaidyaEtAl2006}.
+\label{sec_AHP}
+To generate grading scores for a group of software packages, we use the AHP to pairwise compare them.
+Thomas L.
+Saaty developed this tool, and people widely used it to make and analyze multiple criteria decisions \cite{VaidyaEtAl2006}.
 The AHP organizes multiple criteria factors in a hierarchical structure and pairwise compares the alternatives to calculate relative ratios \cite{Saaty1990}.
@@
 The way of generating $B_i$ is similar to the one for $A$.
-However, unlike comparing the importance between criteria, we pairwise decide how much one alternative is more favored than the other.
-The same method is used to calculate the local priority vector for each $B_i$.
+However, unlike comparing the importance between criteria, we pairwise decide how much we favor one alternative over the other.
+We use the same method to calculate the local priority vector for each $B_i$.
 

### 2021-06-19 StateOfPractice/Ao-Report/13Methods.tex
Update 13Methods.tex

@@
 When choosing a candidate domain, we prefer a domain with a large number of active OSS.
-This is because we aim to finalize a list of 30 software packages \cite{SmithEtAl2021} in the domain after screening out the ones not meeting our requirements, such as the ones without recent updates or certain functions.
-Thus the quantity of final-decision packages may not meet our initial expectation if we do not have enough software candidates to choose from.
+The reason is that we aim to finalize a list of 30 software packages \cite{SmithEtAl2021} in the domain after screening out the ones not meeting our requirements, such as the ones without recent updates or specific functions.
+Thus the quantity of final-decision packages may not meet our initial expectation if we do not have enough software candidates.
 Another reason is that our grading method works much better for OSS.
@@
 
-Another aspect to consider is the team carrying out the research, more specifically, the domain experts - if there is any - in the team.
+Another aspect to consider is the team carrying out the research, specifically the domain experts - if there is any - in the team.
 In our team, researchers in the software engineering field usually lead the projects, and experts working in scientific domains support them.
-Having domain experts in the team provides significant benefits in selecting software packages and designing interview questions.
+Having domain experts in the group provides significant benefits in selecting software packages and designing interview questions.
 
@@
 The only ``mandatory'' requirement is that the software must be OSS, as defined in Section \ref{sec_open_source_software}.
-The reason is that to evaluate all aspects of some software qualities, the source code should be accessible.
+Because to evaluate all aspects of some software qualities, the source code should be accessible.
 
-The other factors to filter the list can be optional and we should consider them according to the number of software candidates and the objectives of the research project.
+The other factors to filter the list can be optional, and we should consider them according to the number of software candidates and the objectives of the research project.
 
@@
 Some of the OSS projects may experience a lack of recent maintenance.
-So we can eliminate packages that have not been updated for a long time and unless they are still popular and highly recommended by the users in the domain \cite{SmithEtAl2021}.
+So we can eliminate packages without recent updates, unless they are still popular and highly recommended by the domain users \cite{SmithEtAl2021}.
 
-No matter we set what standards to filter the packages, with domain experts in the team, we should value their opinions.
-For example, if a software package is not OSS and has not been updated for a long while, the domain experts may still identify it as a popular and widely used product.
+Whether we set what standards to filter the packages, with domain experts in the team, we should value their opinions.
+For example, if a software package is not OSS and has no updates for a long while, the domain experts may still identify it as a popular and widely used product.
 In this case, perhaps it is valuable to keep this software on the list.
@@
 We use the first section of the template to collect some general information about the software, such as the name, purpose, platform, programming language, publications about the software, the first release and the most recent change date, website, and source code repository of the product, etc.
-Information in this section helps us to understand the projects better and may be useful for further analysis, but it does not directly affect the grading scores for the packages.
+Information in this section helps us understand the projects better and may be helpful for further analysis, but it does not directly affect the grading scores for the packages.
 
-We designed the next 9 sections in the template for the 9 software qualities mentioned in Section \ref{sec_software_quality}.
+We designed the following 9 sections in the template for the 9 software qualities mentioned in Section \ref{sec_software_quality}.
 For each quality, we ask several questions and the typical answers are among the choices of ``yes'', ``no'', ``n/a'', ``unclear'', a number, a string, a date, a set of strings, etc.
-The last question of each quality is asking for an overall score between 1 and 10 based on all the previous questions of this section, which is also the grading score for this quality.
-For some qualities, the empirical measurement sections on the template also affect this grading score.
+Each quality needs an overall score between 1 and 10 based on all the previous questions.
+For some qualities, the empirical measurement sections also affect this grading score.
 
-All the last 3 sections on the template are about the empirical measurements.
-We use two command-line software tools git\_stats and scc to extract information about the source code from the project repositories.
+All the last three sections on the template are about the empirical measurements.
+We use two command-line software tools \textit{GitStats} and \textit{scc} to extract information about the source code from the project repositories.
 For projects held on GitHub, we manually collect additional metrics, such as the stars of the GitHub repository, and the numbers of open and closed pull requests.
@@
 We use two command-line tools for the empirical measurements.
-One is GitStats that generates statistics for git repositories and display outputs in the format of web pages \cite{Gieniusz2019}; the other one is Sloc Cloc and Code (as known as scc) \cite{Boyter2021}, aiming to count the lines of code, comments, etc.
+One is \textit{GitStats} that generates statistics for git repositories and display outputs in the format of web pages \cite{Gieniusz2019}; the other one is Sloc Cloc and Code (as known as \textit{scc}) \cite{Boyter2021}, aiming to count the lines of code, comments, etc.
 
-Both tools can measure the number of text-based files in a git repository, as well as lines of text in these files.
-Based on our experience, most text-based files in a software project repository contain programming source code and developers use them to compile and build software products, and a minority of these files are instructions and other documents.
+Both tools can measure the number of text-based files in a git repository and lines of text in these files.
+Based on our experience, most text-based files in a software project repository contain programming source code, and developers use them to compile and build software products.
+A minority of these files are instructions and other documents.
 So we roughly regard the lines of text in text-based files as lines of programming code.
@@
 Secondly, the results may affect our decisions regarding the grading scores for some software qualities.
-For example, if the percentage of comment lines is low, we might want to double-check the understandability of the code, and if the ratio of open versus closed pull requests is high, perhaps we should pay more attention to the maintainability of the project.
+For example, if the percentage of comment lines is low, we might want to double-check the understandability of the code.
+On the other hand, if the ratio of open versus closed pull requests is high, perhaps we should pay more attention to the project’s maintainability.
 
@@
 To test the software on a ``clean'' system, we created a new virtual machine (VM) for each software and only installed the necessary dependencies before measuring.
-All 29 VM were created on the same computer.
-We only start the measuring of the next software after finishing a current one, and after grading each software, the VM is destroyed.
+We created All 29 VM on the same computer.
+We only start the measuring of the following software after finishing a current one, and after grading each software, we destroy the VM.
 
@@
 For a software list with a length of roughly 30, we aim to interview about 10 development teams of these projects.
-Interviewing multiple individuals from each team should give us more comprehensive information about a project, but if there are difficulties in finding multiple willing participants, a single engineer well knowing the project can also be sufficient.
+Interviewing multiple individuals from each team should give us more comprehensive information.
+Still, if it is challenging to find various willing participants, a single engineer well-knowing the project can also be sufficient.
 
@@
 We try to find the contacts of the teams on the websites related to the software, such as the official web pages, repository websites, publication websites, and bio pages of the teams' institutions.
-For each candidate, we send at most two emails asking for their support and participation before receiving any replies.
+Then, we send at most two emails asking for their support and participation before receiving any replies for each candidate.
 
@@
 
-Some of the questions are about the background of the software, the development teams, the interviewees, and the way they organize the projects.
+Some questions are for the background of the software, the development teams, the interviewees, and how they organize the projects.
 We also ask about their understandings of the users.
-Another part of the interview questions focuses on the major difficulties the team experience currently or in the past, and the solutions that they have found or will try in the future.
-We also discuss with interviewees the importance of documents to their projects and the current situations of these documents.
-One more proportion of the questions are about several specific software qualities, such as maintainability, understandability, usability, and reproducibility.
+Another part of the interview questions focuses on the significant difficulties the team experience currently or in the past, and the solutions they have found or will try.
+We also discuss the importance of documents to their projects and the current situations of these documents with interviewees.
+A few questions are about specific software qualities, such as \textit{maintainability}, \textit{understandability}, \textit{usability}, and \textit{reproducibility}.
 
-The interviews are supposed to be semi-structured based on the question list and we also ask follow-up questions when necessary.
-Based on our experience, the interviewees usually bring up some interesting ideas that we do not expect and we can expand on these topics for a few more details.
+The interviews are supposed to be semi-structured based on the question list, and we also ask follow-up questions when necessary.
+Based on our experience, the interviewees usually bring up some exciting ideas that we do not expect, and we can expand on these topics for a few more details.
 
@@
 Since the members of the development teams are likely to be based around the world, so we organize these interviews as virtual meetings online with \hyperlink{https://zoom.us/}{Zoom}.
-After receiving consent from the interviewees, we also record our discussions to better transcribe them.
+After receiving consent from the interviewees, we also record our discussions to transcribe them better.
 
@@
 
-Based on the principles in Section \ref{sec_domain_selection}, we selected the MI domain, because there are numerous software products and a great number of professionals use and/or develop such software in this domain.
-We decided to focus on MI software with the viewing function, so we needed a large number of software candidates in the domain.
+Based on the principles in Section \ref{sec_domain_selection}, we selected the MI domain, because there are numerous software products and a significant number of professionals use and develop such software in this domain.
+We decided to focus on MI software with the viewing function, so we needed many software candidates in the domain.
 Being able to include MI domain experts in our team also made this domain more preferred.
@@
 So following guidelines in Section \ref{sec_filter_software_list} we removed them from the list.
-We focused on the MI software that could work as a MI viewer.
-Some of the software on the list were tool kits or libraries for other software to use as dependencies but not for end-users to view medical images, such as VTK, ITK, and dcm4che, etc.
+Next, we focused on the MI software that could work as a MI viewer.
+Some of the software on the list were tool kits or libraries for other software to use as dependencies but not for end-users to view medical images, such as VTK, ITK, dcm4che, etc.
 We also eliminated these from the list.
@@
 We still preferred projects using git and GitHub and being updated recently, but did not apply another filtering since the number of packages was already below 30.
-However, 27 out of the 29 software packages on the final list used git, and 24 of which were held on GitHub.
+However, 27 out of the 29 software packages on the final list used git, and 24 chose GitHub.
 Furthermore, 27 packages had the latest updates after the year 2018, and 23 after 2020.
@@
 Then we followed the steps in Section \ref{sec_grading_software} to measure and grade the software.
-27 out of the 29 packages can be installed on two or three different operating systems such as Windows, macOS, and Linux, and 5 of them are browser-based, making them platform-independent.
+27 out of the 29 packages are compatible with two or three different operating systems such as Windows, macOS, and Linux, and 5 of them are browser-based, making them platform-independent.
 However, in the interest of time, we only performed the measurements for each project by installing it on one of the platforms, most likely Windows.
@@
 Going through the interview process in Section \ref{sec_interview_methods}, we started with the teams with higher scores on our list, and eventually contacted all of them.
-There are developers/architects from 8 teams having participated in our interviews so far.
+As a result, developers/architects from 8 teams have participated in our interviews so far.
 Before contacting any interviewee candidate, we received ethics clearance from the McMaster University Research Ethics Board (Appendix \ref{ap_ethics}).

### 2021-06-19 StateOfPractice/Ao-Report/14MeasurementResults.tex
ao's report - ch 4

@@
 
-For some qualities, it might be a good idea to cross-compare with the empirical scores.
+A summary of the measurement results can be found in Appendix \ref{ap_measurements}, and the detailed data is in the repository \hyperlink{https://github.com/smiths/AIMSS}{https://github.com/smiths/AIMSS}.
 
-\section{Selected Software List}
+All of the 29 software packages that we measured are OSS with functions of displaying medical images.
+The initial release dates of these packages are between the years 1998 and 2018.
+We considered 2 of them as ``dead'' projects since they had the latest updates more than 18 months earlier than the time of measuring, but both had updates after 2017.
+
+We found out that at least 8 of the projects received funding for at least a certain period, but we were not sure about the rest of them due to limited information.
+
+After analyzing the repositories and official websites of the projects, we found out that the number of contributors for a project varied between 1 to roughly 100.
+We considered anyone who made at least one accepted commit to the source code as a contributor, so it does not mean that any development teams had long-term members as many as 100.
+Many of these projects received change requests and code from the community, such as pull requests and git commits on GitHub.
+At least 27 of the software projects had code commits from more than a single developer, and as far as we could find out, at least 13 of them had no less than 10 contributors.
+
+According to our measurements, 27 out of the 29 packages supported multiple operating systems, and 25 of them could work on all three Windows, macOS, and Linux systems.
+However, there was a significant difference in the philosophy to achieve cross-platform compatibility.
+The majority of the 27 were desktop software products built for each operating system separately.
+On the other hand, 5 were web applications, which made them naturally platform-independent and compatible with almost all operating systems with compatible browsers.
+There are certain advantages and disadvantages to developing either a native or a browser-based application, and some of the details can be found from the interview answers in Section \ref{ch_interviews}.
+
+Most of the projects used more than one programming language, including a primary language that the developers used the most.
+The most popular primary language was C++, chosen by 11 projects, and another 3 teams selected C.
+The second most used one was JavaScript, with 6 projects using it as the primary language.
+4 of the projects chose Java, and 3 selected Python.
+
+\section{Overall Scores}
+
+As described in Section \ref{sec_AHP}, for our AHP measurements, there are 9 criteria which are the 9 software qualities and 29 software packages as the alternatives.
+We decided to make all 9 qualities equally important, so the score of each quality affects the overall scores on the same scale.
+
+Figure \ref{fg_overall_scores} shows the overall scores of all 29 software packages in descending order.
+Since we produced the scores from the AHP process, the total sum of the 29 scores is precisely 1.
+
+\begin{figure}[H]
+\includegraphics[scale=0.35]{figures/overall_scores.png}
+\caption{Overall AHP scores for all 9 software qualities}
+\label{fg_overall_scores}
+\end{figure}
+
+The top two software products had higher scores in each criterion.
+For example, \textit{3D Slicer} \cite{Kikinis2014} ranked at the first or second place for 8 of the 9 qualities; \textit{ImageJ} \cite{Rueden2017} was within the top three products for 7 qualities.
+There were 3 software packages that we
+could not install or build correctly.
+Among them, \textit{DICOM Viewer} \cite{Afsar2021} was the only one that did not have an online test version, so that we could not finish all measurements for it.
+Consequently, we might underestimate its score.
 
 \section{Installability}
+\label{sec_result_installability}
+When measuring the \textit{installability}, we checked the existence and quality of installation instructions.
+The user experience was also an important factor, such as the ease to follow the instructions, number of steps, automation tools, and the prerequisite steps for the installation.
+If any problem interrupted the process of installation or uninstallation, we gave a lower score to this quality.
+We also recorded the operating system for the installation test and whether we could verify the installation.
+
+Figure \ref{fg_installability_scores} lists the scores of \textit{installability}, with a total sum of 1.
+
+\begin{figure}[H]
+\includegraphics[scale=0.38]{figures/installability_scores.png}
+\caption{AHP installability scores}
+\label{fg_installability_scores}
+\end{figure}
+
+The projects with higher scores had easy-to-follow installation instructions, and the installation processes were automated, fast, and frustration-free, with all dependencies automatically added.
+There were also no errors during the installation and uninstallation steps.
+
+The ones with the lowest scores often showed severe problems.
+We could not install some of them due to some issues that we could not solve.
+We spent a reasonable amount of time on these problems, then considered them major obstacles for normal users if we still did not figure out any solutions.
+For example, we needed to build some software from the source code, and we failed the building for two of them following the instructions.
+Furthermore, \textit{DICOM Viewer} depended on a cloud platform, and we could not successfully install the dependency.
+We suspect that only a part of the users faced the same problems, and given a lot of time, we might be able to find solutions.
+However, the difficulties greatly impacted the installation experiences, and we graded these software packages with lower scores.
+
+Although we could not locally build two software packages, there were deployed online versions for them. With that, we finished all the measurements for them.
 
@@
 
-\section{Reliability}
+For \textit{correctness}, we checked the projects to identify the techniques to ensure this quality, such as literate programming, automated testing, symbolic execution, model checking, unit tests, etc.
+We also examined that whether the projects used continuous integration.
+As for \textit{verifiability}, we went through the documents of the project to check the requirements specifications, theory manuals, and getting started tutorial.
+If a getting started tutorial existed and provided expected results, we followed it and checked if the outputs match.
 
-\section{Robustness}
+The scores of \textit{correctness \& verifiability} are shown in Figure \ref{fg_correctness_erifiability_scores}.
+Generally speaking, the packages with higher scores adopted more techniques to improve \textit{correctness}, and had better documents for us to verify it.
 
-\section{Usability}
+\begin{figure}[H]
+\includegraphics[scale=0.38]{figures/correctness_verifiability_scores.png}
+\caption{AHP correctness \& verifiability scores}
+\label{fg_correctness_erifiability_scores}
+\end{figure}
+
+After examining the source code, we could not find any evidence of unit testing in more than half of the projects.
+Unit testing benefits most parts of the software's life cycle, such as designing, coding, debugging, and optimization \cite{Hamill2004}.
+It can reveal the bugs at an earlier stage of the development process, and the absence of unit testing may cause worse \textit{correctness \& verifiability}.
+We also could not identify the requirements specifications or theory manuals for most of the projects.
+It seems that even for some projects with well-organized documents, requirements specifications and theory manuals were still missing.
+
+\section{Surface Reliability}
+For \textit{Surface Reliability}, we checked that whether the software broke during the installation and tutorial, whether there were descriptive error messages, and if we could recover the process after the errors.
+
+As described in Section \ref{sec_result_installability}, we could not build two software packages, however, since there were online versions of them, which might be proof that successful deployments were possible.
+So the failure of installation did not affect their scores in \textit{Surface Reliability}.
+Figure \ref{fg_reliability_scores} shows the AHP results.
+
+\begin{figure}[H]
+\includegraphics[scale=0.38]{figures/reliability_scores.png}
+\caption{AHP surface reliability scores}
+\label{fg_reliability_scores}
+\end{figure}
+
+For the software with the lowest scores, \textit{Drishti} \cite{Limaye2012} and \textit{GATE} \cite{Jan2004} crashed during our operations, and we might underestimated the score of \textit{DICOM Viewer} since we could not install or test it.
+
+\section{Surface Robustness}
+
+To test \textit{Surface Robustness}, we checked that how the software handle unexpected/unanticipated input.
+Since all 29 MI software packages had functions to load image files, we prepared broken image files for the software to open.
+For example, a text file (.txt) with a modified extension name (.dcm) can be used as an unexpected/unanticipated input.
+We first tried to load a few correct input files with the software to ensure the function was working correctly, then with the unexpected/unanticipated ones.
+Figure \ref{fg_robustness_scores} presents the scores.
+
+\begin{figure}[H]
+\includegraphics[scale=0.38]{figures/robustness_scores.png}
+\caption{AHP surface robustness scores}
+\label{fg_robustness_scores}
+\end{figure}
+
+The packages with higher scores elegantly handled the unexpected/unanticipated inputs, normally by showing a clear error message.
+Again, we could not test \textit{DICOM Viewer}.
+We also might underestimate the score of \textit{OHIF Viewer} \cite{Ziegler2020} since we needed further customization to load data and the test was not complete.
+\textit{MatrixUser} \cite{Liu2016}, \textit{dwv} \cite{Martelli2021}, \textit{Slice:Drop} \cite{Haehn2013}, and \textit{MRIcroGL} \cite{Rorden2021} ignored the incorrect format of the input files, and displayed blank or meaningless images.
+\textit{Drishti} successfully detected the unexpected/unanticipated inputs, but the software crashed as a result.
+For unknown reasons, \textit{GATE} failed to load both correct and incorrect inputs.
+
+\section{Surface Usability}
+We measured \textit{Surface Usability} by examining the documents of the projects and considered software with a getting started tutorial and a user manual easier to use.
+Meanwhile, we checked if users had any channels to get supports.
+Our impressions and user experiences when testing the software also affected the scores a lot.
+For example, easy-to-use graphical user interfaces gave us better experiences.
+
+Figure \ref{fg_usability_scores} lists the AHP scores for all packages.
+The ones with higher scores usually provided both comprehensive document guidance and a good user experience.
+\textit{INVESALIUS 3} \cite{Amorim2015} set an good example of detailed and clear user manual.
+\textit{GATE} also provided a large number of documents, but we think that they conveyed the ideas poorly, so we had trouble understand and use them.
+
+\begin{figure}[H]
+\includegraphics[scale=0.38]{figures/usability_scores.png}
+\caption{AHP surface usability scores}
+\label{fg_usability_scores}
+\end{figure}
 
@@
 
+Regarding \textit{maintainability}, we tried to search the projects' documents and identify the process of contributing and reviewing code.
+We believe that the artifacts of a project - including source code, documents, building scripts, etc.
+- can significantly affect its  \textit{maintainability}.
+Thus we checked each project for its artifacts, such as API documentation, bug tracker, release notes, test cases, build files, version control, etc.
+We also checked which tools supported issue tracking and version control and the percentages of closed issues and comment lines in code.
+Figure \ref{fg_maintainability_scores} shows the AHP results for \textit{maintainability}.
+
+\begin{figure}[H]
+\includegraphics[scale=0.38]{figures/maintainability_scores.png}
+\caption{AHP maintainability scores}
+\label{fg_maintainability_scores}
+\end{figure}
+
+We marked \textit{3D Slicer} with a much higher score than others because it did very well at closing the identified issues, and more importantly, we found it to have the most comprehensive artifacts.
+For example, as far as we could find out, only 5 out of the 29 projects had a project plan, 6 of them had a developer's manual, and 7 with API documentation, and \textit{3D Slicer} included all these documents.
+
 \section{Reusability}
 
-\section{Understandability}
+To measure \textit{reusability}, we counted the total number of code files for each project.
+We considered the projects with more code files to be more reusable.
+As mentioned in Section \ref{sec_empirical_measurements}, the tools \textit{GitStats} and \textit{scc} count this number with minor differences, and we used the result from \textit{GitStats} for all projects.
+We also decided that the projects with API documentation can deliver better \textit{Reusability}.
+The AHP results are in Figure \ref{fg_reusability_scores}.
 
-\section{Visibility \& Transparency}
+\begin{figure}[H]
+\includegraphics[scale=0.38]{figures/reusability_scores.png}
+\caption{AHP reusability scores}
+\label{fg_reusability_scores}
+\end{figure}
+
+\section{Surface Understandability}
+
+To determine the \textit{surface understandability} for each project, we randomly examined 10 code files.
+We checked the code’s style within each code file, such as whether the identifiers, parameters, indentation, and formatting were consistent, whether the constants (other than 0 and 1) were hardcoded into the code if the developers modularized the code.
+We also checked the descriptive information for the code, such as documents mentioning the coding standard, the comments in the code, and the descriptions or links for algorithms used in the code.
+Figure \ref{fg_surface_understandability_scores} shows the scores.
+
+\begin{figure}[H]
+\includegraphics[scale=0.38]{figures/understandability_scores.png}
+\caption{AHP surface understandability scores}
+\label{fg_surface_understandability_scores}
+\end{figure}
+
+Most projects had a consistent style for the code, but we only found explicit identification of a coding standard for only 3 out of the 29, which are \textit{3D Slicer}, \textit{Weasis} \cite{Roduit2021}, and \textit{ImageJ}.
+
+\section{Visibility/Transparency}
+
+To understand the \textit{visibility/transparency}, such as all of the steps of a software development process and the current status described in Section \ref{sec_software_quality}, we tried to access this information from the existing documents.
+We tried to examine the development process, current status, development environment, and release notes for each project.
+If any information was missing or poorly conveyed, the \textit{visibility/transparency} was not ideal.
+
+Figure \ref{fg_visibility_transparency_scores} shows the AHP scores.
+Generally speaking, the teams that actively documented their development process and plans scored higher because they delivered better communication to people outside the team.
+
+\begin{figure}[H]
+\includegraphics[scale=0.38]{figures/visibility_transparency_scores.png}
+\caption{AHP visibility/transparency scores}
+\label{fg_visibility_transparency_scores}
+\end{figure}

### 2021-06-19 StateOfPractice/Ao-Report/16ThreatToValidity.tex
ao's report - ch 4

\chapter{Threat to Validity}
\label{ch_validity}

\begin{itemize}
\item limited time to each software, surface quality
\item impressions, subjective
\item searched all information by ourselves, so maybe missed some info
\item mainly measured by one person, so maybe biased
\end{itemize}

### 2021-06-21 StateOfPractice/Peter-Notes/Report/Report.tex
updated LBM Report Results section with feedback

@@
 
+\noindent\textbf{CPU}: Central Processing Unit\\
+
 \noindent\textbf{Goal}: Goals capture, at different levels of abstraction, the various objectives the system under consideration should achieve \citep{van2001goal}.\\
 
+\noindent\textbf{GPU}: Graphics Processing Unit\\
+
+\noindent\textbf{LOC}: Lines of Code\\
+
+\noindent\textbf{OS}: Operating System\\
+
 \noindent\textbf{OTS}: Off-The-Shelf\\
-
-\noindent\textbf{OS}: Operating System\\
 
@@
 
-\noindent\textbf{CPU}: Central Processing Unit\\
-
-\noindent\textbf{GPU}: Graphics Processing Unit\\
+\noindent\textbf{CFD}: Computational Fluid Dynamics\\
 
@@
 
-The LBM software packages are assessed in order to answer the following research questions:
+The LBM software packages are assessed to answer the following research questions:
 
@@
 We collected quantitative data using the measures found in the measurement template of Section \ref{measurementtemplate}.
-Some of this was empirical software engineering related data, such as the number of files, number of lines of code, percentage of issues that are closed, etc.
+Some of this was empirical software engineering related data, such as the number of files, number of lines of code (LOC), percentage of issues that are closed, etc.
 Most of the data was gathered by manually investigating the software, its source code, and its artifacts, while some was gathered using the empirical measurement tools discussed in Section \ref{empiricalmeasures}.
@@
 
+\begin{onehalfspacing}
 \begin{center}
@@
 \end{center}
+\end{onehalfspacing}
 
@@
 \end{center}
+\end{onehalfspacing}
 
 There is considerable variation among these software packages, including their intended purpose, size, user interfaces, and software languages used.
-For example, the OpenLB software package is predominantly a C++ package that makes use of hybrid parallelization and was designed to address a range of computational fluid dynamics problems \citep{heuveline2009towards}.
+For example, the OpenLB software package is predominantly a C++ package that makes use of hybrid parallelization and was designed to address a range of CFD problems \citep{heuveline2009towards}.
 The software package pyLBM is an all-in-one Python language package for numerical simulations \citep{graille2017pylbm}.
@@
 
-This section presents the quantitative and qualitative findings of this study.
-The qualitative data that is presented was gathered through interviews with some developers.
-AHP results for qualities that were quantitatively measured are also found in this section.
-Quality definitions are found in Section \ref{softwarequalities}.
-
-\subsection{Installability}
+This section presents the quantitative and qualitative findings from data that was gathered for each quality listed in Section \ref{softwarequalities}.
+Quantitative results are presented in Section \ref{AHPresults}, and qualitative results are presented in Section \ref{interviewresults}.
+
+\subsection{Quantitative Findings and AHP Results}\label{AHPresults}
+
+This subsection presents the quantitative findings from data that was gathered using the measurement template found in Section \ref{measurementtemplate2}.
+The results of the AHP analysis based on that data are also presented in this subsection.
+
+\subsubsection{Installability}
 
@@
 All but one of the software packages, TCLB, were tested on Ubuntu for this state of the practice assessment.
-TCLB was tested on CentOS, which is mentioned in its installation instructions.
+TCLB was tested on CentOS, since this operating system is mentioned in its installation instructions.
 
@@
 Errors encountered during the installation process were often quickly fixed thanks to descriptive error messages.
-Systems that provided vague error messages, such as those that did not specify which action or file was at fault, were more difficult to troubleshoot.
+Systems that provided vague error messages, such as messages that did not specify which action or file was at fault, were more difficult to troubleshoot.
 Only three software packages (HemeLB, LB3D, lbmpy) that displayed a descriptive error message were not recoverable, and most of these instances were due to hardware and operating system incompatibility, such as the requirement of CUDA.
 Fourteen software packages definitively broke during installation.
-Some packages, such as LB2D\_Prime and LB3D\_Prime, did not provide a definitive message of the success or failure of installation.
+Some packages, such as LB2D-Prime and LB3D-Prime, did not provide a definitive message of the success or failure of installation.
 In these instances, validating the installation required performing a tutorial or running a script, as described below, if these were available.
@@
 If dependencies are installed in one command then none of the software packages take more than 20 steps to install.
-The average number of steps is about eight, and the fewest is three.
+The average number of steps is about eight, and the fewest is two (LB3D-Prime).
 
@@
 
-\subsection{Surface Correctness and Verifiability}
-
-Sixteen of the software packages include a requirements specification artifact or explicitly reference domain theory.
+\subsubsection{Surface Correctness and Verifiability}
+
+Sixteen of the software packages include a requirements specification artifact or explicitly reference domain theory, often only the latter.
+Software packages that distribute requirements specification information, such as DL\_MESO (LBE), generally keep it brief and include it within other documentation.
 This artifact is often found within a user manual, on a web-page, or is mentioned in related publications.
 In the latter case the user may need to spend significant time to find this information.
-
-Interviews with developers confirmed that many of the software packages are developed by domain experts with backgrounds in physics, mathematics and mechanical engineering.
-Many of these developers do not have formal software engineering education.
-Some of the development teams include computer scientists.
-Despite a lack of visible domain documentation and a resulting lower surface correctness and verifiability score, many software packages do have a strong domain background on account of their development team.
 
@@
 
-Unit tests are only explicitly available for one of the software packages.
+Unit tests are only explicitly available for one of the software packages, Ludwig.
 Code modularization of most packages allow for users to create tests with varying degrees of effort.
@@
 
-The documentation or presence of continuous integration tools and techniques alludes to a more refined development process where faults are isolated and better recognized.
-Only two of the packages mentioned applying the practice of continuous integration in their development process.
-
-Interviews with the developers of software packages suggested a more frequent use of both unit testing and continuous integration in the development processes than initially surveyed.
-For example, OpenLB, pyLBM, and TCLB use such methods during development despite it not being explicitly clear from a surface analysis.
+The use of continuous integration tools and techniques alludes to a more refined development process where faults are isolated and better recognized.
+Only two of the packages (ESPResSo, Ludwig) mentioned applying the practice of continuous integration in their development process.
 
@@
 It and the second ranked package, ESPResSo, explicitly incorporated continuous integration in the development process.
-During developer interviews it was discovered that some other packages also use continuous integration and unit testing, despite not being explicitly stated in documentation.
-The correctness and verificablity of such packages is not measured well using surface analysis.
 
@@
 Unit testing documentation and capability, as well as the use of continuous integration tools and techniques such as Bamboo, Jenkins, and Travis CI, help verify correctness.
-Several interviewed developers alluded to difficulty with testing large numbers of features, and some even manually tested program output.
-The use of well defined unit testing tools could decrease the time spent testing some feature.
-
-\subsection{Surface Reliability}
-
+
+\subsubsection{Surface Reliability}
+
+The analysis of surface reliability focused on package installation and tutorials.
 Errors occurred when installing 16 of the software packages.
@@
 In these instances the installation tried to access outdated external repositories.
-Seven of the installations were recovered and verified, and one of the installations was assumed to be recovered due to the absence of any way to verify it.
+Seven of the installations were recovered and verified, and one of the installations (LB3D-Prime) was assumed to be recovered due to the absence of any way to verify it.
 The installation of eight of the software packages could not be recovered.
@@
 
-Of the 13 software packages that installed correctly and also have tutorials, four broke during tutorial testing.
+Of the 13 software packages that installed correctly and also have tutorials, four (pyLBM, ESPResSo++, LIMBES, Ludwig) broke during tutorial testing.
 All of these instances resulted in an error message being displayed.
-One error was due to a missing tutorial dependency, another was due to an invalid command despite following the tutorial, and the final two errors were vague execution errors.
+One error (pyLBM) was due to a missing tutorial dependency, another (Ludwig) was due to an invalid command despite following the tutorial, and the final two errors were vague execution errors.
 Of the four broken tutorial instances, only the one that was missing a dependency was recoverable.
@@
 Overall, lower ranked software packages are lacking clear documentation, testing or tutorial examples, and descriptive error messages, and have broken dependencies.
-Thus, regarding surface reliability, software packages would benefit from clear up-to-date installation and tutorial documentation, and descriptive error messages.
-
-\subsection{Surface Robustness}
+Thus, regarding surface reliability, software packages would benefit from clear up-to-date documentation that specifies all dependencies, the inclusion of testing and tutorial examples, and the assurance of descriptive error messages during fault conditions.
+
+\subsubsection{Surface Robustness}
 
@@
 
-\subsection{Surface Performance}
-
-The software packages were not quantitatively measured for surface performance.
-Although the software packages all apply LBM to solve scientific computing problems, the packages focus on varied domains, with varying parameters, and are technically different from each other.
+\subsubsection{Surface Performance}
+
+Although the software packages all apply LBM to solve scientific computing problems, the packages focus on varied CFD problems, with varying parameters, and are technically different from each other.
 Due to this, a comparison of performance is not appropriate.
 In this project we instead looked through each software package's artifacts for evidence that performance was considered.
-The artifacts of 17 software packages mentioned the use of some sort of parallelization tool or technique.
-This included GPU processing and the CUDA parallel computing platform.
+The artifacts of 17 software packages mentioned parallelization.
+This included GPU processing and the CUDA parallel computing platform, which were mentioned in the artifacts of 6 packages (ESPResSo, lbmpy, lettuce, pyLBM, Sailfish, TCLB).
 GPUs provide superior processing power and speed compared to CPUs, and are often used for scientific computing when a large amount of data is involved.
@@
 In the Ludwig package, a so-called mixed mode approach is used where fine-grained parallelism is implemented on the GPU, and MPI is used for even larger scale parallelism \citep{gray2013ludwig}.
-While some software packages require CUDA and GPU processing, some have the option of using either the GPU or the CPU.
+While one software package (Sailfish) required CUDA and GPU processing, some (ESPResSo, lbmpy, lettuce, pyLBM, TCLB) have the option of using either the GPU or the CPU.
 The packages that require GPU and CUDA have better performance at the expense of installability and surface reliability.
 
-\subsection{Surface Usability}
+\subsubsection{Surface Usability}
 
@@
 Most user manuals are in the form of a file that can be downloaded, while some are rendered on a web-page.
-Some packages do not have a user manual but do have useful documentation distributed on their web-pages.
-Expected user characteristics are documented in four software packages.
-Users are typically scientists or engineers trying to test a new application.
+Some packages (waLBerla) do not have a user manual, but do have useful documentation distributed on their web-pages.
+Expected user characteristics are documented in four software packages (laboetie, LIMBES, Ludwig, Palabos).
+Users are typically scientists or engineers.
 Their background is often physics, chemistry, biophysics, or mathematics.
-All but one of the packages have a user support model, and several of them have multiple avenues of user support.
+All but one of the packages (LIMBES) have a user support model, and many of them have multiple avenues of user support.
 The most popular avenue of support is Git, followed by email and forums.
-One software package has a FAQ page.
+One software package (OpenLB) has an FAQ page.
 
@@
 Software packages with a high score have a tutorial and user manual, sometimes have documented user characteristics, and have at least one user support model.
-Many packages have several user models.
-
-Interviews with developers revealed several usability issues.
-Some users have misunderstood the boundaries of LBM and computational fluid dynamics, and have combined or applied methods that are not physically sound.
-Sometimes users have applied LBM to poorly defined or inappropriate fluid dynamics problems.
-The users do not realize the limitations of the methods, of the software, and do not understand the requirements to properly model a problem with the software.
-The software packages are not designed to be used out of the box in a plug and play fashion.
-Developers of some of the software packages mitigated this by editing the source code to prevent unsound combination from being applied, and by updating the documentation to better inform users of LBM limitations, and of the requirements to properly model appropriate problems, including what algorithms and parameters to use.
-
-Some additional but infrequent software usability issues were commented on by the developers.
-Users have had trouble with installation and understanding how to maneuver the interfaces and how to set up or run models.
-These issues are addressed by various user support models, including frequently asked questions sections on the software package websites, user guides, and hardware and software requirements specifications.
-
-One software package changed its scripting language to Python to make it more usable.
-The developer commented that this was the biggest step in addressing usability, commenting that Python was a much easier language for users to learn and understand.
-
-\subsection{Maintainability}
+Many packages have several user support models.
+
+\subsubsection{Maintainability}
 
@@
 This information could be used to better troubleshoot issues and organize documentation.
-All but three software packages have source code release and documentation version numbers.
+All but three software packages (LatBo.jl, LB3D-Prime, MechSys) have source code release and documentation version numbers.\\
+
+\begin{onehalfspacing}
+\begin{center}
+	\begin{tabular}{ p{4cm}p{4cm}p{4cm} }
+		\hline
+		Name & $\%$ Issues Closed & $\%$ Code Comments\\
+		\hline
+		DL\_MESO (LBE) & Not Git & 8.06\\
+		ESPResSo & 89.26 & 21.78\\
+		ESPResSo++ & 66.28 & 17.10\\
+		HemeLB & No Issues & 16.68\\
+		laboetie & 18.75 & 2.47\\		
+		LatBo.jl & 93.33 & 0.40\\
+		LB2D-Prime & Not Git & 13.61\\
+		LB3D & Not Git & 13.76\\
+		LB3D-Prime & Not Git & 14.34\\
+		lbmpy& 58.33  & 2.03  \\
+		lettuce & 33.33 & 8.19\\
+		LIMBES & Not Git & 17.39\\
+		Ludwig& 60.00 & 20.70\\
+		LUMA& 85.71   & 0.20\\
+		MechSys & Not Git & 15.11\\
+		MP-LABS & 100.00 & 26.67\\
+		OpenLB & Not Git & 22.43\\
+		Palabos & 89.47 & 17.76\\
+		pyLBM & 66.67& 16.12\\
+		Sailfish & 22.22 & 9.26\\
+		SunlightLB & Not Git & 17.67\\
+		TCLB & 60.32 & 6.02\\
+		waLBerla & 72.90 & 22.62\\
+		\hline
+	\end{tabular}
+	\captionof{table}{Git Repository Data}\label{gitrepodata}
+\end{center}
+\end{onehalfspacing}
 
@@
 
-Issue tracking is used in 22 software packages, 15 of which use Git, six use email, and one uses SourceForge.
-Most software packages that use Git have most of their issues closed, and only three have less than 50 percent of their issues closed.
-Furthermore, most packages that use Git for issue tracking use GitHub as a version control system, one package uses CVS for issue tracking, and eight packages do not appear to use any such system.
+Issue tracking is used in 22 software packages, 15 of which use Git, six use email, and one (SunlightLB) uses SourceForge.
+Most software packages that use Git have most of their issues closed, and only three (laboetie, lettuce, Sailfish) have less than 50 percent of their issues closed.
+This information is presented in Table \ref{gitrepodata}.
+Furthermore, 13 packages that use Git for issue tracking use GitHub as a version control system, while two (Palabos, waLBerla) use GitLab.
+Of the other packages, one package (SunlightLB) uses CVS for issue tracking, and seven packages do not appear to use any issue tracking system.
 
 Software package code files were further measured for the percentage of code that is comments.
+The findings are presented in Table \ref{gitrepodata}.
 Packages with a higher percentage of comments were designated as more maintainable.
@@
 
-Interviews with developers revealed that most projects do have a defined process for accepting contributions from team members.
-The packages rarely get contributions from outside developers, but the process would be similar as for the aforementioned group.
-
-Contributions are made through GitHub, and are then reviewed and pulled by lead developers, often with consultation with a group of core developers depending on the organizational model.
-Continuous integration is part of the process for some packages.
-
-Some developers noted that their software package does not have well defined contributing guide in the repository but it might be a good idea to add one in the near future.
-They would be very happy to see contribution from outside of their organization, but currently this does not happen.
-
-Furthermore, maintainability has been addressed by increasing source code modularity, reducing duplicate information, and improving abstraction.
-Several software packages have had large sections of their code base redeveloped with languages that the developers felt are more understandable and readable, and that are better supported.
-Data structures have also been redeveloped and storage has been improved.
-One software developer mentioned that the mathematical foundations that the geometries and models are designed from had been re-evaluated and optimized.
-
-\subsection{Modifiability}
-
-Software packages were not quantitatively measured for modifiability.
-In this project we asked developers to comment on modifiability when we interviewed them.
-Specifically, we asked if ease of future changes to the system, modules, and code blocks was considered when designing the software.
-We also asked if any measures had been taken to ensure the ease of future changes.
-All of the developers that were interviewed noted that the ease of future changes was considered and that measures to ensure it had been taken.
-
-A high degree of code modularity and abstraction was noted by one developer as a measure to ensure the ease of future changes.
-They also noted that some of the code base was transitioned from c to c++, which could ease modifiability of that software package.
-Another developer noted that their software package was designed to easily allow for the addition of some LBM features, but changes to major aspects of the LBM models would be difficult.
-Furthermore, the package was designed with flexible data structures and storage in mind.
-
-Some software packages, like Palabos, provide validation benchmarks for their core fundamental algorithmic ingredients \citep{latt2021palabos}.
-The stated intent of these benchmarks is to showcase the validity and usefulness of the package in order to stimulate the development of third-user extensions.
-The Palabos package identifies as a development framework for modeling problems in various computational fluid dynamics areas.
-
-\subsection{Reusability}
+\subsubsection{Reusability}\label{reusabilityresults}
 
@@
 A larger number of source files was associated with increased reusability due to increased modularization.
+Some packages have more features than others, consequently contributing to reusablility since they have more source code that can be reused.
 The software packages were also reviewed for the presence of API documentation, which indicates that a software package was developed with interaction between other software applications in mind.
@@
 
-There was a strong focus on modularity when designing the waLBerla framework in order to enhance productivity, reusability, and maintainability \citep{bauer2021walberla}.
-Its software design has enabled waLBerla to be successfully applied in several projects as a basis for various extensions \citep{bauer2021walberla}.
-
-\subsection{Surface Understandability}
+Table \ref{moduledata} shows file and line of code data of the software packages.
+Packages with a high reusability score do not have many LOC per text file, generally having a few hundred lines or less.
+This suggests that the source code of these packages is likely functionally modularized, and modules could be reused in other projects.
+
+There was a strong focus on modularity when designing the waLBerla framework to enhance productivity, reusability, and maintainability \citep{bauer2021walberla}.
+Its software design has enabled waLBerla to be successfully applied in several projects as a basis for various extensions \citep{bauer2021walberla}.\\
+
+\begin{onehalfspacing}
+	\begin{center}
+		\begin{tabular}{ p{3.5cm}p{2cm}p{2.5cm}p{2cm}p{2.5cm} }
+			\hline
+			Name & Text Files & Binary Files & LOC & Avg. LOC / Text File\\
+			\hline
+			DL\_MESO (LBE) & 310 & 51 & 170223& 549\\
+			ESPResSo &1309& 86 & 186700&143\\
+			ESPResSo++ &5328& 66 & 969196&182\\
+			HemeLB &1065& 48 & 95104&89\\
+			laboetie &133& 1 & 48403&364\\		
+			LatBo.jl &41& 0 & 42172&1029\\
+			LB2D-Prime &82& 19 & 54755&668\\
+			LB3D &99& 76 & 39766&402\\
+			LB3D-Prime &23& 6 & 12944&563\\
+			lbmpy&201&  28 & 46489  &231\\
+			lettuce &62& 0 & 5529&89\\
+			LIMBES &26& 1 & 4872&187\\
+			Ludwig&859& 32 & 109811&128\\
+			LUMA&312& 19 & 4370670&14000\\
+			MechSys &324& 3 & 85543&264\\
+			MP-LABS &307& 3 & 43124&140\\
+			OpenLB &1104& 5 & 209034&189\\
+			Palabos &1829& 67 & 547623&299\\
+			pyLBM &258& 85& 32314&125\\
+			Sailfish &632& 11 & 69398&110\\
+			SunlightLB & 36& 1 & 7646& 212\\
+			TCLB &535& 7 & 43226&81\\
+			waLBerla & 2395 & 67 & 848146&353\\
+			\hline
+		\end{tabular}
+		\captionof{table}{Module Data}\label{moduledata}
+	\end{center}
+\end{onehalfspacing}
+
+\subsubsection{Surface Understandability}
 
 Ten random source code files of each software package were reviewed for several measures.
-This measure of surface understandability may not perfectly reflect each package due to the limitation of only testing 10 files.
+This assessment of surface understandability may not perfectly reflect each package due to the limitation of only testing 10 files.
 
@@
 Generally, the software packages use consistent, distinctive, and meaningful code identifiers.
-Only four packages appear to use vague identifiers.
+Only four packages (LB2D-Prime, LB3D-Prime, LIMBES, MP-LABS) appear to use vague identifiers, such as single letters for variables.
 Hard coded constants were observed in the source code of 12 packages.
@@
 Domain algorithms are noted in the source code of 11 packages.
-All of the packages are well modularized.
-When observing the source code files, it was found that 13 of the packages have a consistent style and order for function parameters.
+Table \ref{moduledata} suggests that the software packages are modularized to various degrees.
+When observing the source code files, it was found that 13 of the packages have a consistent style and order of function parameters.
 
@@
 
-Software developers noted that in their opinion users have generally found their packages to be understandable.
-One developer commented that some users have attempted to run physically incompatible LBM methods, and the solution was to edit the code to prevent such combinations, as well as to update the documentation to prevent misunderstanding the methods.
-Similarly, another developer noted that some users had issues setting up parameters for LBM schemes.
-The solution to this was to update the interface where these parameters are set, as well as to add functionality to test the stability of the parameters.
-A third developer noted that some users lack the background knowledge to easily model fluid dynamics problems using their software.
-A frequently asked questions section was added to their package website to help users find answers to common questions.
-The package also has detailed documentation, including guides and usage requirements specification, to better help user understand the software.
-
-\subsection{Traceability}
-
-Software packages were not quantitatively measured for traceability.
-In this project we asked developers to comment on traceability, specifically on their software package's documentation and how it fit into their development process.
-
-One developer noted that all major additions to their package had accompanying changes to artifacts and documentation.
-They noted that considerable effort had been put into the documentation.
-They further commented that they want to lower the entry barrier for new developers, and because of that their package has a considerable amount of developer documentation.
-This documentation informs developers on how to get started, and orients them to the artifacts, source code, and system architecture, as well as how the software package build system works, and how the coupling between the simulation engine and the interface works.
-
-Another developer noted the importance of documentation for both the users and developers of their software.
-New features are always added to the documentation.
-The developers use documentation to stay up to date on the status of the software package, and to help expand features, like computational models or algorithms.
-This is necessary so that the coding standard for these models is kept consistent with new developers.
-
-The importance of documentation for both users and developers was stressed throughout the interviews.
-However, it was noted several times that a lack of time and funding has a negative affect on the documentation.
-Most of the developers are scientific researchers evaluated on the scientific papers that they produce.
-Writing and updating documentation is something that is done in free time, if that time arises, and if writing documentation is a concern.
-Sometimes it is a last priority for the developers.
-Finding ways to hasten updating documentation would increase the frequency of such updates and benefit both users and developers.
-
-One developer noted the use of documentation generators like Doxygen. It would be advisable for more projects to use such automatic document generation tools, since some projects do not do this.
-
-\subsection{Visibility and Transparency}
+\subsubsection{Visibility and Transparency}
 
@@
 
-Developers were asked to comment on the obstacles in their development process.
-One developer noted that a lot of their source code had been written with a specific application in mind, and that there is too much coupling between components.
-Addressing this issue would help with code modifiability and reusability.
-Updating the development process would help resolve this issue and prevent such issues in the future.
-Improving the visibility of software changes and a peer review process would also help.
-Improving the software engineering education or experience of developers is also an idea that was brought up by several developers.
-Specifically, developers should always write code that is decoupled and modular, and should keep in mind the visibility of their contributions by better updating documentation and ensuring that their contributions are transparent to the rest of the developers.
-This would help catch issues in the contributions, and improve source code maintenance.
-
-According to one developer, some obstacles to the development processes of their package had been overcome by the introduction of continuous integration practices, and a peer review process for contributions.
-These practices decrease development and maintenance times.
-Another developer mentioned that their package had two sets of code, for executing the models on the CPU and GPU, and that maintenance was decreased by introducing macros, a practice which became a common part of the development process.
-
-Developers were also asked how documentation fits into their development process.
-Several developers noted that developer documentation plays an important role in familiarizing potential contributors to the software system architecture.
-Without the guidance that the documentation provides it would be unlikely that contributions would pass the peer review process.
-
-None of the software packages whose developers were interviewed have a formal software development model.
-The packages all have fairly small development teams.
-These teams do accept outside contributors, but generally the teams are tight-knit, often working at the same institution, although one of the packages has an international team.
-One developer noted that while no formal model is used, their development model is something similar to a combination of agile and waterfall development models.
-
-The developers noted similar project management processes.
-For teams of only a couple of developers, the addition of new features or major changes are discussed with the entire team.
-Projects with more than a couple of developers have lead developer roles.
-These lead developers review potential additions to the software.
-The software packages use GitHub for managing the project.
-Typically there are several development branches as well as the master branch.
-
-\subsection{Reproducibility}
-
-Software packages were not quantitatively measured for reproducibility.
-In this project we asked developers to comment on reproducibility when we interviewed them.
-
-Developers were asked if they have any concern that their computational results would not be reproducible in the future, and if they had taken any steps to ensure reproducibility.
-
-One developer noted that they compare the results of their methods against manually calculated results.
-These comparison tests are automatically run for all source code changes.
-The tests are run when a pull request is opened on GitHub.
-Even once these tests are complete, a peer review process is done before changes are fully committed to the appropriate branch.
-The results for all of the LBM schemes on the software package development branch are also compared daily for correctness, ensuring that the system output reflects the expected output.
-
-Several developers noted that they currently do not have a system in place to test for reproducibility, but it is of interest and could be implemented in the future.
-Generally, the mathematical foundations of the models are verified, but the output of the software package is not compared to other output.
-Depending on the package and how it outputs LBM solutions, it may not be practical or feasible.
-A correct output may not be exactly reproducible, as it may be dependent on a probability distribution, so strictly comparing results may not be appropriate.
-
-The source code and artifacts of some software packages may be reproducible.
-There is considerable variance in the quality of software specifications and other developer documentation.
-Some packages are well detailed, and translating the specifications into source code will produce similar results across developers.
-
-\subsection{Unambiguity}
-
-Software packages were not quantitatively measured for unambiguity.
-In this project we asked developers to comment on unambiguity when we interviewed them.
-
-Developers were asked if they thought that the current documentation can clearly convey all necessary knowledge to the users, and if they had taken any steps to ensure clarity.
-
-One developer noted that their documentation was meant for users that are already familiar with the underlying physics and computational fluid dynamics methods, and that these concepts are not explained in detail.
-The documentation focuses solely on how to technically use the software package, and includes a user guide and tutorial walk through of how to set up and run a simulation.
-With this in mind the developer believes that their documentation is in reasonable shape for users with a minimum knowledge of the underlying physics.
-If new users have technical questions these can be addressed in further revisions of the documentation.
-New developers should also find that the documentation is reasonably clear and useful.
-Information that is missing, like detailed explanations of dependencies, is referenced in the documents.
-
-Another developer also noted that their documentation was in reasonable shape, but they would really appreciate more user feedback to improve it.
-They also noted that they believe a lack of knowledge of the underlying physics and computational fluid dynamics concepts can be an issue for some users.
-This information can be referenced in the documentation, but it is not something that the documentation needs to detail.
-
-\subsection{Overall Quality}
-
-Figure \ref{Fig_OverallScore} shows the overall ranking of the software packages using AHP.
-Software packages with an overall high score had ranked high in at least several of the individual qualities that were quantitatively measured.
+\subsubsection{Overall Quality}
 
@@
 
+Figure \ref{Fig_OverallScore} shows the overall ranking of the software packages using AHP.
+Software packages with an overall high score had ranked high in at least several of the individual qualities that were quantitatively measured.
+
 Looking at the top three ranked packages, Ludwig scored high in surface correctness and verifiability, surface robustness, surface usability, maintainability, and visibility and transparency.
@@
 Section \ref{qualityrecommentations} further analyzes these findings and offers some software quality recommendations for future development of LBM software.
+
+\subsection{Qualitative Findings From Developer Interviews}\label{interviewresults}
+
+This subsection presents the qualitative findings from data that was gathered during interviews with developers. The interview questions are found in Section \ref{interviewquestions}.
+
+\subsubsection{Surface Correctness and Verifiability}
+
+Interviews with developers confirmed that these software packages are developed by domain experts with backgrounds in physics, mathematics and mechanical engineering.
+It was noted in these interviews that some of the developers do not have formal software engineering education.
+Some of the development teams include computer scientists.
+Despite a lack of visible domain documentation and a resulting lower surface correctness and verifiability score, it is clear that some of the software packages were developed by teams with significant domain knowledge on account of the academic backgrounds of their developers.
+
+Interviews suggested a more frequent use of both unit testing and continuous integration in the development processes than what was observed from the artifact survey.
+For example, OpenLB, pyLBM, and TCLB use such methods during development despite this not being explicitly clear from an analysis of the material available online.
+The correctness and verificablity of such packages is not measured well using surface analysis.
+
+Several interviewed developers alluded to difficulty with testing the correctness of large numbers of features, and some even manually tested program output.
+The use of well defined unit testing tools could decrease the time spent testing some feature.
+
+\subsubsection{Surface Usability}
+
+Interviews with developers revealed several usability issues.
+Some users have misunderstood the boundaries of LBM and CFD, and have combined or applied methods that are not physically sound.
+Sometimes users have applied LBM to poorly defined or inappropriate fluid dynamics problems.
+For example, they may wish to model flow through or around a structure despite having limited information about the structure or its environment, and having little previous knowledge of CFD.
+The users do not realize the limitations of the methods, of the software, and do not understand the requirements to properly model a problem with the software.
+As the developer of TCLB noted, such software packages are not designed to be used ``out of the box'' in a plug and play fashion, and it could take months so set up the CFD problems correctly.
+Developers of some software packages, including ESPResSo, mitigated this by editing the source code to prevent users from ``combining methods that are not physically sound together'', and by updating the documentation to better inform users of LBM limitations, and of the requirements to properly model appropriate problems, including what algorithms and parameters to use.
+
+Some additional but infrequent software usability issues were commented on by the developers.
+Users have had trouble with installation and understanding how to maneuver the interfaces and how to set up or run models.
+These issues are addressed by various user support models, including frequently asked questions sections on the software package websites, user guides, and hardware and software requirements specifications.
+
+One software package (ESPResSo) changed some of its scripting language to Python to make it more usable.
+The developer commented that this was ``the biggest step in terms of usability over the years'', further commenting that ``most people in the field know [Python]'' and that ``it's easy to learn''.
+
+\subsubsection{Maintainability}
+
+Interviews with developers revealed that most projects do have a defined process for accepting contributions from team members.
+The packages rarely get contributions from outside developers, but the process would be similar as for the aforementioned group.
+
+Contributions are made through GitHub, and are then reviewed and pulled by lead developers, often with consultation with a group of core developers depending on the organizational model.
+Continuous integration is part of the process for some packages.
+
+Some developers noted that their software package does not have well defined contributing guide in the repository, but it might be a good idea to add one in the near future.
+They would be happy to see contribution from outside of their organization, but currently this does not happen.
+
+Furthermore, maintainability has been addressed by increasing source code modularity, reducing duplicate information, and improving abstraction by developing well-defined interfaces.
+This was noted by the developers of ESPResSo and pyLBM.
+Several software packages have had sections of their code base redeveloped with languages that the developers felt are more understandable and readable, and that are better supported, such as Python.
+Data structures have also been redeveloped and storage has been improved.
+A developer of pyLBM mentioned that the geometries and models of their system had been ``decoupled'', using abstraction and modularization of the source code, to make it ``very easy to add [new] features''.
+
+\subsubsection{Modifiability}
+
+Software packages were not quantitatively measured for modifiability.
+In this project we asked developers to comment on modifiability when we interviewed them.
+Specifically, we asked if ease of future changes to the system, modules, and code blocks was considered when designing the software.
+We also asked if any measures had been taken to ensure the ease of future changes.
+All of the developers that were interviewed noted that the ease of future changes was considered and that measures to ensure it had been taken, including
+requiring the separation of software components in the source code architecture.
+
+A high degree of code modularity and abstraction, ensured by separating components and hiding information behind well-define interfaces, was noted by developers as a measure to ensure the ease of future changes.
+The developer of ESPResSo also noted that some of the code base was transitioned from C to C++, which could ease modifiability of that software package.
+The developer of TCLB noted that their software package was designed to allow for the addition of some LBM features, but changes to major aspects of the system would be difficult.
+For example, ``implementing a new model will be an easy contribution'', but changes to the ``Cartesian mesh...will be a nightmare''.
+Furthermore, the package was designed with flexible data structures and storage in mind.
+
+Some software packages, like Palabos, provide validation benchmarks for their core fundamental algorithmic ingredients \citep{latt2021palabos}.
+The stated intent of these benchmarks is to showcase the validity and usefulness of the package to stimulate the development of third-user extensions.
+The Palabos package identifies as a development framework for modeling problems in various CFD areas.
+
+\subsubsection{Surface Understandability}
+
+Software developers noted that they believe users have generally found their packages to be understandable.
+The interviewed developer of ESPResSo commented that some users have attempted to run physically incompatible LBM methods, and the solution was to edit the code to prevent such combinations, as well as to update the documentation to prevent misunderstanding the methods.
+Similarly, a developer of pyLBM noted that some users had issues setting up parameters for LBM schemes.
+The solution to this was to update the interface where these parameters are set, as well as to add functionality to test the stability of the parameters.
+A developer of OpenLB noted that some users lack the background knowledge to easily model fluid dynamics problems using their software.
+A frequently asked questions section was added to their package website to help users find answers to common questions.
+The package also has detailed documentation, including guides and usage requirements specification, to better help user understand the software.
+
+\subsubsection{Traceability}
+
+Software packages were not quantitatively measured for traceability.
+In this project we asked developers to comment on traceability, specifically on their software package's documentation and how it fit into their development process.
+
+The interviewed developer of ESPResSo noted that all major additions to their package had accompanying changes to artifacts and documentation.
+They noted that considerable effort had been put into the documentation.
+They further commented that they want to lower the entry barrier for new developers, and because of that their package has a considerable amount of developer documentation.
+This documentation informs developers on how to get started, and orients them to the artifacts, source code, and system architecture, as well as how the software package build system works, and how the coupling between the simulation engine and the interface works.
+
+Developers noted the importance of documentation for both the users and developers of their software.
+New features are always added to the documentation.
+The developers use documentation to stay up to date on the status of the software package, and to help expand features, like computational models or algorithms.
+This is necessary so that the coding standard for these models is kept consistent with new developers.
+
+The importance of documentation for both users and developers was stressed throughout the interviews.
+However, it was noted several times that a lack of time and funding has a negative affect on the documentation.
+Most of the developers are scientific researchers evaluated on the scientific papers that they produce.
+Writing and updating documentation is something that is done in free time, if that time arises.
+Sometimes it is a last priority for the developers.
+Finding ways to hasten updating documentation would increase the frequency of such updates and benefit both users and developers.
+
+The developer of OpenLB noted the use of documentation generators like Doxygen.
+It would be advisable for more projects to use such automatic document generation tools, since some projects do not do this.
+
+\subsubsection{Visibility and Transparency}
+
+Developers were asked to comment on the obstacles in their development process.
+The developer of ESPResSo noted that a lot of their source code had been written with a specific application in mind, and that there is too much coupling between components.
+Addressing this issue would help with code modifiability and reusability.
+Updating the development process would help resolve this issue and prevent such issues in the future.
+Improving the visibility of software changes and a peer review process would also help.
+Improving the software engineering education or experience of developers is also an idea that was brought up by several developers.
+Specifically, developers should always write code that is decoupled and modular, and should keep in mind the visibility of their contributions by better updating documentation and ensuring that their contributions are transparent to the rest of the developers.
+This would help catch issues in the contributions, and improve source code maintenance.
+
+According to the interviewed developer of ESPResSo, some obstacles to the development processes of their package had been overcome by the introduction of continuous integration practices, and a peer review process for contributions.
+These practices decrease development and maintenance times.
+The developer of TCLB mentioned that their package had two sets of code, for executing the models on the CPU and GPU, and that maintenance was decreased by introducing macros, a practice which became a common part of the development process.
+
+Developers were also asked how documentation fits into their development process.
+Several developers noted that developer documentation plays an important role in familiarizing potential contributors to the software system architecture.
+Without the guidance that the documentation provides it would be unlikely that contributions would pass the peer review process.
+
+None of the software packages whose developers were interviewed have a formal software development model.
+The packages all have fairly small development teams.
+These teams do accept outside contributors, but generally the teams are tight-knit, often working at the same institution, although one of the packages has an international team.
+The developer of ESPResSo noted that while no formal model is used, their development model is something similar to a combination of agile and waterfall development models.
+
+The developers noted similar project management processes.
+For teams of only a couple of developers, the addition of new features or major changes are discussed with the entire team.
+Projects with more than a couple of developers have lead developer roles.
+These lead developers review potential additions to the software.
+The software packages use GitHub for managing the project.
+Typically there are several development branches as well as the master branch.
+
+\subsubsection{Reproducibility}
+
+Software packages were not quantitatively measured for reproducibility.
+In this project we asked developers to comment on reproducibility when we interviewed them.
+
+Developers were asked if they have any concern that their computational results would not be reproducible in the future, and if they had taken any steps to ensure reproducibility.
+
+The developer of ESPResSo noted a comparison of the results of their methods against manually calculated results.
+These comparison tests are automatically run for all source code changes.
+The tests are run when a pull request is opened on GitHub.
+Even once these tests are complete, a peer review process is done before changes are fully committed to the appropriate branch.
+The results for all of the LBM schemes on the software package development branch are also frequently compared for correctness, ensuring that the system output reflects the expected output.
+
+Several developers noted that they currently do not have a system in place to test for reproducibility, but it is of interest and could be implemented in the future.
+Generally, the mathematical foundations of the models are verified, but the output of the software package is not compared to other output.
+Depending on the package and how it outputs solutions, it may not be practical or feasible.
+A correct output may not be exactly reproducible, as it may be dependent on a probability distribution, so strictly comparing results may not be appropriate.
+
+The source code and artifacts of some software packages may be reproducible.
+There is considerable variance in the quality of software specifications and other developer documentation.
+Some packages are well detailed, and translating the specifications into source code will produce similar results across developers.
+Developers were not asked to comment on the reproducibility of their source code from their requirements specifications and design documentation.
+This question should be considered in the next iteration of state of the practice assessments.
+
+\subsubsection{Unambiguity}
+
+Software packages were not quantitatively measured for unambiguity.
+In this project we asked developers to comment on unambiguity when we interviewed them.
+
+Developers were asked if they thought that the current documentation can clearly convey all necessary knowledge to the users, and if they had taken any steps to ensure clarity.
+
+The developer of ESPResSo noted that their documentation was meant for users that are already familiar with the underlying physics and CFD methods.
+These concepts are not explained in detail within the documentation.
+Users should acquire this knowledge from suitable external sources.
+The documentation focuses on how to technically use the software package, and includes a user guide and tutorial walk through of how to set up and run a simulation.
+With this in mind, the developer believes that their documentation is in reasonable shape for users with a minimum knowledge of the underlying physics.
+If new users have technical questions these can be addressed in further revisions of the documentation.
+New developers should find that the documentation is reasonably clear and useful.
+Information that is missing, like detailed explanations of dependencies, is referenced in the documents.
+
+The developer of pyLBM also noted that their documentation was in reasonable shape, but that they ``need more [user] feedback to improve [it]''.
+They also noted that they believe a lack of knowledge of the underlying physics and CFD concepts can be an issue for some users.
+This information can be referenced in the documentation, but it is not something that the documentation needs to detail.
 
@@
 There is often no time or funding for maintaining quality documentation for software that is rarely used outside of the development team.
-Furthermore, the documentation generally only provides a shallow overview of the underlying computational fluid dynamics theory.
+Furthermore, the documentation generally only provides a shallow overview of the underlying CFD theory.
 Users would be well advised to already be familiar with these topics to or they should spend significant time referencing theory resources.
@@
 Several developers noted that users sometimes try to use incorrect LBM method combinations to solve their problems.
-Furthermore, some users think that the packages will work out of the box to solve their cases, while the packages require both a good understanding of computational fluid dynamics and an understanding of the requirements for formulating problems in the individual packages, which can be a significant endeavor.
+Furthermore, some users think that the packages will work out of the box to solve their cases, while the packages require both a good understanding of CFD and an understanding of the requirements for formulating problems in the individual packages, which can be a significant endeavor.
 These software packages are not like commercial software packages.
@@
 
-One of the interviewed developers commented that parts of their package's source code had been refactored to Python in order to help address usability issues.
+One of the interviewed developers commented that parts of their package's source code had been refactored to Python to help address usability issues.
 Python was perceived as a much more usable language, and it would be easy for future users and developers to learn and understand the source code.
@@
 \begin{itemize}
-	\item Integrate parallelization tools and techniques in order to reduce processing time. Consider GPU processing, CUDA, and MPI. 
+	\item Integrate parallelization tools and techniques to reduce processing time. Consider GPU processing, CUDA, and MPI. 
 	\item The user should be able to choose to process their model on either the CPU or GPU.

### 2021-06-21 StateOfPractice/Peter-Notes/Report/Report.tex
fixed some typos LBM Report

@@
 
-This report analyzes the development of software packages that use Lattice Boltzmann Methods (LBM), a class of computational fluid dynamics (CFD) methods, for fluid simulation.
+We analyze the development of software packages that use Lattice Boltzmann Methods (LBM), a class of computational fluid dynamics (CFD) methods, for fluid simulation.
 These LBM algorithms consider the behaviours of a collection of particles as a single unit of an intermediate size, and predict their positional probability as they move through a lattice structure.
-Figure \ref{circularflow} from \citep{chen2021phase} presents the streamlines of converged flow past a stationary circular cylinder with varied Raynolds number, a common task of LBM.
+Figure \ref{circularflow} from \citep{chen2021phase} presents the streamlines of converged flow past a stationary circular cylinder with varied Reynolds number, a common task of LBM.
 
@@
 
-This report analyzes a filtered set of LBM software packages along quantitative and qualitative measures.
+We analyze a filtered set of LBM software packages along quantitative and qualitative measures.
 Many of these measures are captured using surface measurements, which can be categorized as initial and easy to capture measurements, of the underlying quality; they may not represent the true quality of the software.
@@
 
-Interviews suggested a more frequent use of both unit testing and continuous integration in the development processes than what was observed from the artifact survey.
+Interviews suggested a more frequent use of both unit testing and continuous integration in the development processes than what was observed from the initial survey.
 For example, OpenLB, pyLBM, and TCLB use such methods during development despite this not being explicitly clear from an analysis of the material available online.
@@
 
+\subsection{Grading Template}\label{gradingtemplate2}
+The table below lists how each quality measure of the measurement template is used to calculate an overall impression in each software quality set.\newline
+
+\captionof{table}{Grading Template}\label{gradingtemplate}
+\def\arraystretch{1.5}
+\begin{tabular}{p{16cm}}
+	\hline		
+	\textbf{Installability  (Measured via installation on a virtual machine.) }\\
+	\hline
+	Are there installation instructions? ({yes=1, no=-1})\\
+	Are the installation instructions in one place? Place referring to a single document or web-page. ({yes=1, no=0, n/a=0})\\
+	Are the installation instructions linear? Linear meaning progressing  in a single series of steps. ({yes=1, no=0, n/a=0})\\
+	Are the instructions written as if the person doing the installation has none of the dependent packages installed? ({yes=1, no=0, unclear=0})\\
+	Are compatible operating system versions listed? ({yes=1, no=0})\\
+	Is there something in place to automate the installation (makefile, script, installer, etc)? ({yes*=1, no=-1})\\
+	If the software installation broke, was a descriptive error message displayed? ({yes=0, no=-2, n/a=1})\\
+	Is there a specified way to validate the installation? ({yes*=1, no=0})\\
+	How many steps were involved in the installation? (Includes manual steps like unzipping files) Specify OS. ($<$10 = 1)\\
+	What OS was used for the installation? (does not count)\\
+	How many extra software packages need to be installed before or during installation? ($<$10 = 1)\\
+	Are required package versions listed? ({yes=1, no=0, n/a=1})\\
+	Are there instructions for the installation of required packages / dependencies? ({yes=1, no=0, n/a=1})\\
+	Run uninstall, if available. Were any obvious problems caused? ({yes∗=0, no=1, unavail=1})\\
+	Overall impression? (a sum of $>$10 is rounded down to 10)\\
+	\hline
+\end{tabular}
+
+\def\arraystretch{1.33}
+\begin{tabular}{p{16cm}}
+	\hline		
+	\textbf{Correctness and Verifiability}\\
+	\hline
+	Any reference to the requirements specifications of the program or theory manuals? ({yes∗=2, no=0, unclear=0})\\
+	What tools or techniques are used to build confidence of correctness? (any=1, unclear=0)\\
+	If there is a getting started tutorial? ({yes=2, no=0})\\
+	Are the tutorial instructions linear? ({yes=1, no=0, n/a=0})\\
+	Does the getting started tutorial provide an expected output? ({yes=1, no*=0, n/a=0})\\
+	Does your tutorial output match the expected output? ({yes=1, no=0, n/a=0})\\
+	Are unit tests available?  ({yes=1, no=0, unclear=0})\\
+	Is there evidence of continuous integration? (for example mentioned in documentation, Jenkins, Travis CI, Bamboo, other) ({yes*=1, no=0, unclear=0})\\
+	\hline	
+	\textbf{Surface Reliability}\\
+	\hline
+	Did the software “break” during installation? ({yes∗=0, no=5})\\
+	If the software installation broke, was the installation instance recoverable? ({yes=5, no=0, n/a=0})\\
+	Did the software “break” during the initial tutorial testing? ({yes∗=0, no=5, n/a=0})\\
+	If the tutorial testing broke, was a descriptive error message displayed? ({yes=2, no=0, n/a=0})\\
+	If the tutorial testing broke, was the tutorial testing instance recoverable? ({yes=3, no=0, n/a=0})\\
+	\hline		
+	\textbf{Surface Robustness}\\
+	\hline
+	Does the software handle unexpected/unanticipated input (like data of the wrong type, empty input, missing files or links) reasonably?
+(a reasonable response can include an appropriate error message.) ({yes=5, no∗=0})\\
+	For any plain text input files, if all new lines are replaced with new lines and carriage returns, will the software handle this gracefully? ({yes=5, no∗=0, n/a=5})\\
+	\hline		
+	\textbf{Surface Usability}\\
+	\hline
+	Is there a getting started tutorial? ({yes=3, no=0})\\
+	Is there a user manual? ({yes=4, no=0})\\
+	Are expected user characteristics documented? ({yes=1, no=0})\\
+	What is the user support model? FAQ? User forum? E-mail address to direct questions? Etc. (one=1, two+=2, none=0)\\
+	\hline
+\end{tabular}
+
+\def\arraystretch{1.4}
+\begin{tabular}{p{16cm}}
+	\hline
+	\textbf{Maintainability}\\
+	\hline
+	What is the current version number? (provided=1, nothing=0)\\
+	Is there any information on how code is reviewed, or how to contribute? ({yes*=1, no=0})\\
+	Are artifacts available?
+(List every type of file that is not a code file – for examples please look at the ‘Artifact Name’ column of https://gitlab.cas.mcmaster.ca/SEforSC/se4sc/-/blob/git-svn/GradStudents/Olu/ResearchProposal/Artifacts\_MiningV3.xlsx) (Rate 0 – 2 depending on how many and perceived quality)\\
+	What issue tracking tool is employed? (nothing=0, email of other private=1, anything public or accessible by all devs (eg git) = 2)\\
+	What is the percentage of identified issues that are closed? (50$\%$+=1, $<$50$\%$=0)\\
+	What percentage of code is comments? (10$\%$+=1, $<$10$\%$=0)\\
+	Which version control system is in use? (anything=2, nothing=0)\\
+	\hline		
+	\textbf{Reusability}\\
+	\hline
+	How many code files are there? (0-9=0, 10-49=1, 50-99=3, 100-299=4, 300-599=5, 600-999=6, 1000+=8)\\
+	Is API documented? ({yes=2, no=0, n/a=0})\\
+	\hline	
+	\textbf{Surface Understandability (Based on 10 random source files)}\\
+	\hline
+	Consistent indentation and formatting style? ({yes=1, no=0, n/a=0})\\
+	Explicit identification of a coding standard? ({yes∗=1, no=0, n/a=0})\\
+	Are the code identifiers consistent, distinctive, and meaningful? ({yes=2, no∗=0, n/a=0})\\
+	Are constants (other than 0 and 1) hard coded into the program? ({yes=1, no∗=0, n/a=0})\\
+	Comments are clear, indicate what is being done, not how? ({yes=2, no∗=0, n/a=0})\\
+	Is the name/URL of any algorithms used mentioned? ({yes=1, no∗=0, n/a=0})\\
+	Parameters are in the same order for all functions? ({yes=1, no∗=0, n/a=0})\\
+	Is code modularized? ({yes=1, no∗=0, n/a=0})\\
+	\hline		
+	\textbf{Visibility/Transparency}\\
+	\hline
+	Is the development process defined? If yes, what process is used. ({yes∗=3, no=0, n/a=0})\\
+	Are there any documents recording the development process and status?  ({yes∗=3, no=0}))\\
+	Is the development environment documented? ({yes∗=2, no=0})\\
+	Are there release notes? ({yes∗=2, no=0})\\
+	\hline
+\end{tabular}
+\newpage
+
 \subsection{Eliminated Software Packages}\label{eliminatedpackagessection}
@@
 
-\subsection{Grading Template}\label{gradingtemplate2}
-The table below lists how each quality measure of the measurement template is used to calculate an overall impression in each software quality set.\newline
-
-\captionof{table}{Grading Template}\label{gradingtemplate}
-\def\arraystretch{1.5}
-\begin{tabular}{p{16cm}}
-	\hline		
-	\textbf{Installability  (Measured via installation on a virtual machine.) }\\
-	\hline
-	Are there installation instructions? ({yes=1, no=-1})\\
-	Are the installation instructions in one place? Place referring to a single document or web-page. ({yes=1, no=0, n/a=0})\\
-	Are the installation instructions linear? Linear meaning progressing  in a single series of steps. ({yes=1, no=0, n/a=0})\\
-	Are the instructions written as if the person doing the installation has none of the dependent packages installed? ({yes=1, no=0, unclear=0})\\
-	Are compatible operating system versions listed? ({yes=1, no=0})\\
-	Is there something in place to automate the installation (makefile, script, installer, etc)? ({yes*=1, no=-1})\\
-	If the software installation broke, was a descriptive error message displayed? ({yes=0, no=-2, n/a=1})\\
-	Is there a specified way to validate the installation? ({yes*=1, no=0})\\
-	How many steps were involved in the installation? (Includes manual steps like unzipping files) Specify OS. ($<$10 = 1)\\
-	What OS was used for the installation? (does not count)\\
-	How many extra software packages need to be installed before or during installation? ($<$10 = 1)\\
-	Are required package versions listed? ({yes=1, no=0, n/a=1})\\
-	Are there instructions for the installation of required packages / dependencies? ({yes=1, no=0, n/a=1})\\
-	Run uninstall, if available. Were any obvious problems caused? ({yes∗=0, no=1, unavail=1})\\
-	Overall impression? (a sum of $>$10 is rounded down to 10)\\
-	\hline
-\end{tabular}
-
-\def\arraystretch{1.33}
-\begin{tabular}{p{16cm}}
-	\hline		
-	\textbf{Correctness and Verifiability}\\
-	\hline
-	Any reference to the requirements specifications of the program or theory manuals? ({yes∗=2, no=0, unclear=0})\\
-	What tools or techniques are used to build confidence of correctness? (any=1, unclear=0)\\
-	If there is a getting started tutorial? ({yes=2, no=0})\\
-	Are the tutorial instructions linear? ({yes=1, no=0, n/a=0})\\
-	Does the getting started tutorial provide an expected output? ({yes=1, no*=0, n/a=0})\\
-	Does your tutorial output match the expected output? ({yes=1, no=0, n/a=0})\\
-	Are unit tests available?  ({yes=1, no=0, unclear=0})\\
-	Is there evidence of continuous integration? (for example mentioned in documentation, Jenkins, Travis CI, Bamboo, other) ({yes*=1, no=0, unclear=0})\\
-	\hline	
-	\textbf{Surface Reliability}\\
-	\hline
-	Did the software “break” during installation? ({yes∗=0, no=5})\\
-	If the software installation broke, was the installation instance recoverable? ({yes=5, no=0, n/a=0})\\
-	Did the software “break” during the initial tutorial testing? ({yes∗=0, no=5, n/a=0})\\
-	If the tutorial testing broke, was a descriptive error message displayed? ({yes=2, no=0, n/a=0})\\
-	If the tutorial testing broke, was the tutorial testing instance recoverable? ({yes=3, no=0, n/a=0})\\
-	\hline		
-	\textbf{Surface Robustness}\\
-	\hline
-	Does the software handle unexpected/unanticipated input (like data of the wrong type, empty input, missing files or links) reasonably?
-(a reasonable response can include an appropriate error message.) ({yes=5, no∗=0})\\
-	For any plain text input files, if all new lines are replaced with new lines and carriage returns, will the software handle this gracefully? ({yes=5, no∗=0, n/a=5})\\
-	\hline		
-	\textbf{Surface Usability}\\
-	\hline
-	Is there a getting started tutorial? ({yes=3, no=0})\\
-	Is there a user manual? ({yes=4, no=0})\\
-	Are expected user characteristics documented? ({yes=1, no=0})\\
-	What is the user support model? FAQ? User forum? E-mail address to direct questions? Etc. (one=1, two+=2, none=0)\\
-	\hline
-\end{tabular}
-
-\def\arraystretch{1.4}
-\begin{tabular}{p{16cm}}
-	\hline
-	\textbf{Maintainability}\\
-	\hline
-	What is the current version number? (provided=1, nothing=0)\\
-	Is there any information on how code is reviewed, or how to contribute? ({yes*=1, no=0})\\
-	Are artifacts available?
-(List every type of file that is not a code file – for examples please look at the ‘Artifact Name’ column of https://gitlab.cas.mcmaster.ca/SEforSC/se4sc/-/blob/git-svn/GradStudents/Olu/ResearchProposal/Artifacts\_MiningV3.xlsx) (Rate 0 – 2 depending on how many and perceived quality)\\
-	What issue tracking tool is employed? (nothing=0, email of other private=1, anything public or accessible by all devs (eg git) = 2)\\
-	What is the percentage of identified issues that are closed? (50$\%$+=1, $<$50$\%$=0)\\
-	What percentage of code is comments? (10$\%$+=1, $<$10$\%$=0)\\
-	Which version control system is in use? (anything=2, nothing=0)\\
-	\hline		
-	\textbf{Reusability}\\
-	\hline
-	How many code files are there? (0-9=0, 10-49=1, 50-99=3, 100-299=4, 300-599=5, 600-999=6, 1000+=8)\\
-	Is API documented? ({yes=2, no=0, n/a=0})\\
-	\hline	
-	\textbf{Surface Understandability (Based on 10 random source files)}\\
-	\hline
-	Consistent indentation and formatting style? ({yes=1, no=0, n/a=0})\\
-	Explicit identification of a coding standard? ({yes∗=1, no=0, n/a=0})\\
-	Are the code identifiers consistent, distinctive, and meaningful? ({yes=2, no∗=0, n/a=0})\\
-	Are constants (other than 0 and 1) hard coded into the program? ({yes=1, no∗=0, n/a=0})\\
-	Comments are clear, indicate what is being done, not how? ({yes=2, no∗=0, n/a=0})\\
-	Is the name/URL of any algorithms used mentioned? ({yes=1, no∗=0, n/a=0})\\
-	Parameters are in the same order for all functions? ({yes=1, no∗=0, n/a=0})\\
-	Is code modularized? ({yes=1, no∗=0, n/a=0})\\
-	\hline		
-	\textbf{Visibility/Transparency}\\
-	\hline
-	Is the development process defined? If yes, what process is used. ({yes∗=3, no=0, n/a=0})\\
-	Are there any documents recording the development process and status?  ({yes∗=3, no=0}))\\
-	Is the development environment documented? ({yes∗=2, no=0})\\
-	Are there release notes? ({yes∗=2, no=0})\\
-	\hline
-\end{tabular}
-\newpage
-
 \subsection{Ethics Approval}

### 2021-06-22 StateOfPractice/Peter-Notes/Report/Report.tex
updated some of Section 5 of LBM report using feedback; fixed typos

@@
 
-\subsection{Purpose}\label{purpose}
+\subsection{Research Questions}\label{purpose}
 
@@
 		LatBo.jl & 2014 Aug & 2017 Feb\\
+		LB2D-Prime & 2005 & 2012 Apr\\
+		LB3D & unclear & 2012 Mar\\
+		LB3D-Prime & 2005 & 2011 Oct\\
 		LIMBES & 2010 Nov & 2014 Dec\\
-		LB3D & unclear & 2012 Mar\\
-		LB2D-Prime & 2005 & 2012 Apr\\
-		LB3D-Prime & 2005 & 2011 Oct\\
 		MP-LABS & 2008 Jun & 2014 Oct\\
@@
 The ability to validate the installation process, often through tutorials or test examples that include expected output, is correlated with a higher score.
+Furthermore, the top six ranked packages are noted as being alive.
 
@@
 It and the second ranked package, ESPResSo, explicitly incorporated continuous integration in the development process.
+Furthermore, eight of the top 10 ranked packages are noted as being alive.
 
@@
 One of these packages, pyLBM, broke during tutorial testing, but a descriptive error message helped in recovery.
+Furthermore, nine of the top 10 ranked packages are noted as being alive.
 
@@
 Similarly, all software packages that installed correctly and require plain text input files correctly handled an unexpected change to these input files, including a replacement of new lines with carriage returns.
+Furthermore, nine of the top 10 ranked packages are noted as being alive.
 
@@
 Many packages have several user support models.
+Furthermore, four of the top five ranked packages are noted as being alive.
 
@@
 This information could be used to better troubleshoot issues and organize documentation.
-All but three software packages (LatBo.jl, LB3D-Prime, MechSys) have source code release and documentation version numbers.\\
+All but three software packages (LatBo.jl, LB3D-Prime, MechSys) have source code release and documentation version numbers.
+
+Information on how code is reviewed, or how to contribute to the project was also noted.
+In total, 11 software packages have this information, which was found in various artifacts, including in developer guides, contributor guides, user guides, developer web-pages, and README files.
+
+Issue tracking is used in 22 software packages, 15 of which use Git, six use email, and one (SunlightLB) uses SourceForge.
+Most software packages that use Git have most of their issues closed, and only three (laboetie, lettuce, Sailfish) have less than 50 percent of their issues closed.
+Alive packages (11 use Git issue tracking) have 64\% of their issues closed, while dead packages (3 use Git issue tracking) have 71\% of their issues closed.
+This information is presented in Table \ref{gitrepodata}.
+Furthermore, 13 packages that use Git for issue tracking use GitHub as a version control system, while two (Palabos, waLBerla) use GitLab.
+Of the other packages, one package (SunlightLB) uses CVS for issue tracking, and seven packages do not appear to use any issue tracking system.
+
+Software package code files were further measured for the percentage of code that is comments.
+The findings are presented in Table \ref{gitrepodata}.
+Packages with a higher percentage of comments were designated as more maintainable.
+Comments represent more than 10 percent of code files in 15 packages, and the average percentage of code comments is about 14.\\
 
@@
 \begin{center}
-	\begin{tabular}{ p{4cm}p{4cm}p{4cm} }
-		\hline
-		Name & $\%$ Issues Closed & $\%$ Code Comments\\
-		\hline
-		DL\_MESO (LBE) & Not Git & 8.06\\
-		ESPResSo & 89.26 & 21.78\\
-		ESPResSo++ & 66.28 & 17.10\\
-		HemeLB & No Issues & 16.68\\
-		laboetie & 18.75 & 2.47\\		
-		LatBo.jl & 93.33 & 0.40\\
-		LB2D-Prime & Not Git & 13.61\\
-		LB3D & Not Git & 13.76\\
-		LB3D-Prime & Not Git & 14.34\\
-		lbmpy& 58.33  & 2.03  \\
-		lettuce & 33.33 & 8.19\\
-		LIMBES & Not Git & 17.39\\
-		Ludwig& 60.00 & 20.70\\
-		LUMA& 85.71   & 0.20\\
-		MechSys & Not Git & 15.11\\
-		MP-LABS & 100.00 & 26.67\\
-		OpenLB & Not Git & 22.43\\
-		Palabos & 89.47 & 17.76\\
-		pyLBM & 66.67& 16.12\\
-		Sailfish & 22.22 & 9.26\\
-		SunlightLB & Not Git & 17.67\\
-		TCLB & 60.32 & 6.02\\
-		waLBerla & 72.90 & 22.62\\
+	\begin{tabular}{ p{3.5cm}p{3.5cm}p{3.5cm}p{2.5cm} }
+		\hline
+		Name & $\%$ Issues Closed & $\%$ Code Comments& Status\\
+		\hline
+		DL\_MESO (LBE) & Not Git & 8.06&Alive\\
+		ESPResSo & 89.26 & 21.78&Alive\\
+		ESPResSo++ & 66.28 & 17.10&Alive\\
+		HemeLB & No Issues & 16.68&Dead\\
+		laboetie & 18.75 & 2.47&Dead\\		
+		LatBo.jl & 93.33 & 0.40&Dead\\
+		LB2D-Prime & Not Git & 13.61&Dead\\
+		LB3D & Not Git & 13.76&Dead\\
+		LB3D-Prime & Not Git & 14.34&Dead\\
+		lbmpy& 58.33  & 2.03  &Alive\\
+		lettuce & 33.33 & 8.19&Alive\\
+		LIMBES & Not Git & 17.39&Dead\\
+		Ludwig& 60.00 & 20.70&Alive\\
+		LUMA& 85.71   & 0.20&Alive\\
+		MechSys & Not Git & 15.11&Alive\\
+		MP-LABS & 100.00 & 26.67&Dead\\
+		OpenLB & Not Git & 22.43&Alive\\
+		Palabos & 89.47 & 17.76&Alive\\
+		pyLBM & 66.67& 16.12&Alive\\
+		Sailfish & 22.22 & 9.26&Alive\\
+		SunlightLB & Not Git & 17.67&Dead\\
+		TCLB & 60.32 & 6.02&Alive\\
+		waLBerla & 72.90 & 22.62&Alive\\
 		\hline
@@
 \end{onehalfspacing}
-
-Information on how code is reviewed, or how to contribute to the project was also noted.
-In total, 11 software packages have this information, which was found in various artifacts, including in developer guides, contributor guides, user guides, developer web-pages, and README files.
-
-Issue tracking is used in 22 software packages, 15 of which use Git, six use email, and one (SunlightLB) uses SourceForge.
-Most software packages that use Git have most of their issues closed, and only three (laboetie, lettuce, Sailfish) have less than 50 percent of their issues closed.
-This information is presented in Table \ref{gitrepodata}.
-Furthermore, 13 packages that use Git for issue tracking use GitHub as a version control system, while two (Palabos, waLBerla) use GitLab.
-Of the other packages, one package (SunlightLB) uses CVS for issue tracking, and seven packages do not appear to use any issue tracking system.
-
-Software package code files were further measured for the percentage of code that is comments.
-The findings are presented in Table \ref{gitrepodata}.
-Packages with a higher percentage of comments were designated as more maintainable.
-Comments represent more than 10 percent of code files in 15 packages, and the average percentage of code comments is about 14.
 
@@
 Their code files are well commented with more than 10 percent of the code being comments.
+Furthermore, nine of the top 10 ranked packages are noted as being alive.
 
@@
 However, these packages do have an abundance of reusable components.
+Furthermore, nine of the top 10 ranked packages are noted as being alive.
 
@@
 The source code is well modularized and structured.
+Furthermore, four of the top five ranked packages are noted as being alive.
 
@@
 They also
-had detailed and easy to access notes accompanying software releases.
+had detailed and easy to access notes accompanying software releases. Furthermore, four of the top five ranked packages are noted as being alive.
 
@@
 
+This section answers the research questions listed in Section \ref{purpose} using the quantitative and qualitative data presented in Section \ref{qualityresults}, additional data from software package repositories and artifacts, and domain expert feedback.
+Each subsection answers one of the research questions.
+Software package artifacts are examined in Section \ref{artifacts}, tools are listed in Section \ref{tools}, development principles, processes, and methodologies are discussed in Section \ref{prinprocmeth}, development pain points are explored in Section \ref{painpoints}, and our recommendations for improving software qualities are presented in Section \ref{qualityrecommentations}.
+A comparison between our designations of the software packages to community rankings is made in Section \ref{comparison}.
+Finally, threats to the validity of this assessment are noted in Section \ref{threats}.
+
 \subsection{Artifacts Present - DONE 1st DRAFT}\label{artifacts}
+
+This subsection answers the research question: What artifacts are present in current software packages?
 
@@
 Below we have grouped them into common, less common, and rare artifacts.
-Common artifacts were found in 16 to 23 of the software packages.
-Less common artifacts were found in 8 to 15 of the software packages.
-Rare artifacts were found in 1 to 7 of the software packages.
+Common artifacts were found in 16 to 23 ($>$70\%) of the software packages.
+Less common artifacts were found in 8 to 15 (35-70\%) of the software packages.
+Rare artifacts were found in 1 to 7 ($<$35\%) of the software packages.
 
@@
 The following artifacts were commonly found in the 23 software packages that were tested.
-All of the top four AHP ranked packages, ESPREsSo, Ludwig, Palabos, and OpenLB have each of these artifacts, except for a requirements specification or theory notes.
-Only three of the top four ranked packages were found to have such an artifact.
-
-\begin{singlespace}
+All of the top four AHP ranked packages, ESPREsSo, Ludwig, OpenLB, and Palabos, have each of these artifacts, except only three of them have a requirements specification or theory notes.
+Palabos is the only one of these four packages that does not have an artifact from this category.
+
 \begin{multicols}{2}	
 \begin{itemize}
+	\begin{singlespace}
 		\item Authors/Developers List
@@
 		\item Tutorial
+	\end{singlespace}
 \end{itemize}
 \end{multicols}
-\end{singlespace}
 
@@
 The following artifacts were less commonly found in the 23 software packages that were tested. The top four AHP ranked packages have most of these artifacts.
-Only one of those four packages does not have a user manual or guide, but it does have a detailed and informative website.
-Another one of the top four packages does not appear to have publicly visible design documentation.
-A third package from the list does not appear to use a version control system.
+At the time of data collection, only one (Palabos) of the four packages did not have a user manual or guide, but there was a broken link on the package website indicating that such an artifact might exist.
+This broken link was later fixed, but this is not reflected in our data.
+Despite the broken link, Palabos does have a detailed and informative website.
+Another one (Ludwig) of the top four packages does not appear to have publicly visible design documentation.
+A third package (OpenLB) from the list does not appear to use a version control system.
 It is possible that such a system is used by the package since its website notes package version numbers, but the artifacts do not explicitly state the use of such a system.
 
-\begin{singlespace}
 	\begin{multicols}{2}	
 		\begin{itemize}
+			\begin{singlespace}
 			\item Change Log / Release Notes
@@
 			\item Version Control
+		\end{singlespace}
 		\end{itemize}
 	\end{multicols}
-\end{singlespace}
 
@@
 None of the top four packages have any explicit API documentation.
-Three of these packages have information on contributing to the project.
-Two of them have a FAQ section or forum.
-One has verification and validation notes, and a video guide explaining how to use the software.
-
-\begin{singlespace}
+Three of these packages (ESPREsSo, Ludwig, Palabos) have information on contributing to the project.
+Two of them (OpenLB, Palabos) have a FAQ section or forum.
+One (OpenLB) has verification and validation notes, and a video guide of the software.
+
 	\begin{multicols}{2}	
 		\begin{itemize}
+			\begin{singlespace}
 				\item API Documentation
@@
 				\item Video Guide (including YouTube)
+			\end{singlespace}
 		\end{itemize}
 	\end{multicols}
-\end{singlespace}
 
@@
 
-\subsection{Tools Used - DONE 1st Draft}
+\subsection{Tools Used - DONE 1st Draft}\label{tools}
+
+This subsection answers the research question: What tools (development, dependencies, project management) are used by current software packages?
+
 Software tools are used to support the development, verification, maintenance, and evolution of software, software processes, and artifacts \citep{ghezzi1991fundamentals}.
@@
 
-These tools support the development of end products but do not become part of them, unlike dependencies that remain in the application once it is released \citep{ghezzi1991fundamentals}.
+Development tools support the development of end products but do not become part of them, unlike dependencies that remain in the application once it is released \citep{ghezzi1991fundamentals}.
 The following type of development tools were explicitly noted in the artifacts or web-pages of the 23 LBM packages that were assessed.
-It is likely that other tools, such as debuggers, were used but are not specified here.
-
-\begin{singlespace}
+It is likely that other tools, such as debuggers, were used but are not specified in our sources.
+
 	\begin{multicols}{2}	
 		\begin{itemize}
+			\begin{singlespace}
 			\item Continuous Integration Tools
@@
 			\item Correctness Verification Tools
+			\end{singlespace}
 		\end{itemize}
 	\end{multicols}
-\end{singlespace}
 
 The above tools can verify the correctness of software during its development.
-Only two of the software packages that were assessed mentioned using continuous integration tools like Travis CI.
+Only two (ESPResSo, Ludwig) of the software packages that were assessed mentioned using continuous integration tools, like Travis CI.
 Code editors and compilers were explicitly noted to have been used by several packages, and were likely used by all of them.
-One of the packages explicitly noted the use of unit testing, but did not specify the tool that was used.
-Likewise, the use of a tool for verifying the correctness of output was noted by one of the developers.
+One of the packages (Ludwig) explicitly noted the use of proprietary unit testing code written in C.
+Likewise, the use of proprietary code for verifying the correctness of output was noted by one (pyLBM) of the developers.
 It is likely that similar tools were used when developing other software packages.
@@
 
-\begin{singlespace}
 	\begin{multicols}{2}	
 		\begin{itemize}
+			\begin{singlespace}
 			\item Build Automation Tools
@@
 			\item Domain Specific Libraries
+		\end{singlespace}
 		\end{itemize}
 	\end{multicols}
-\end{singlespace}
 
@@
 They also all use various technical and domain specific libraries.
-Technical libraries include visualization, data analysis, and message passing libraries.
-Domain specific libraries are scientific computing libraries.
-Libraries that are explicitly stated in artifact, and were not noted during our observations, may fall outside of these categories.
+Technical libraries include visualization (e.g.
+Matplotlib, ParaView, Pygame, VTK), data analysis (e.g.
+Anaconda, Torch), and message passing libraries (e.g.
+MPICH, Open MPI, PyZMQ).
+Domain specific libraries are scientific computing libraries (e.g.
+SciPy).
+Libraries that are explicitly stated in artifacts, and were not noted during our observations, may fall outside of these categories.
 
@@
 
-\begin{singlespace}
 	\begin{multicols}{2}	
 		\begin{itemize}
+			\begin{singlespace}
 			\item Collaboration Tools
@@
 			\item Document Generation Tools
+		\end{singlespace}
 		\end{itemize}
 	\end{multicols}
-\end{singlespace}
 
@@
 Most of the projects appear to use change tracking and version control tools.
-They often use GitHub for this, and a few use Git or CVS.
+They often use GitHub for this.
+One package (lbmpy) uses Git, and another (SunlightLB) uses CVS.
 Document generation tools are mentioned in the artifacts of 12 of the projects.
-Tools such as Sphinx and Doxygen are explicitly used in this capacity.
-Drasil could be considered as an alternative to generate artifacts since it is geared specifically towards research software domains like LBM software.
-
-\subsection{Principles, Processes, and Methodologies - DONE 1st DRAFT}
+The tools Sphinx and Doxygen are explicitly used in this capacity.
+
+\subsection{Principles, Processes, and Methodologies - DONE 1st DRAFT}\label{prinprocmeth}
+
+This subsection answers the research question: What principles, processes, and methodologies are used in the development of current software packages?
+
+The points and conclusions in this subsection come from developer interviews and reviews of software package artifacts.
 
 Most of the software packages do not explicitly state in their artifacts the motivations or design principles that were considered when developing the software.
-One package, Sailfish, indicates that shortening the development time was considered in early stages of design, with the developers opting for using Python and CUDA/OpenCL to achieve this without sacrificing any computational performance.
+One package, Sailfish, indicates in its artifacts that shortening the development time was considered in early stages of design, with the developers opting for using Python and CUDA/OpenCL to achieve this without sacrificing any computational performance.
 The goals of that project are explicitly listed as performance, scalability, agility and extendability, maintenance, and ease of use.
@@
 It is not explicitly indicated in the artifacts of most of the packages that development involved following any specific model, like a waterfall or agile development model.
-One developer noted that while no formal model is used, their development model is something similar to a combination of agile and waterfall development models.
-The developer teams of the LBM packages are fairly small and easily organized without the need for such models.
+One developer (ESPResSo) noted that while no formal model is used, their development model is something similar to a combination of agile and waterfall development models.
+The developer teams of the LBM packages are fairly small, so it is feasible for them to be organized without the need for such models.
 
 Seven of the software packages contain artifacts outlining the general development process, like basic instructions on how to contribute.
-Most of the development teams accept outside contributors, but generally the teams are centralized, often working at the same institution.
+Eleven of the packages explicitly convey that they would accept outside contributors, but generally the teams are centralized, often working at the same institution.
 
@@
 These lead developers review potential additions to the software.
-One of the developers that was interviewed noted that a peer review process is used to assess major changes and additions.
-
-Many software packages use GitHub for managing the project. Typically there are several simultaneous development branches.
+One of the developers (ESPResSo) that was interviewed noted that an ad hoc peer review process is used to assess major changes and additions.
+
+Thirteen (57\%) of the 23 software packages use GitHub for managing the project, including nine (64\%) of the 14 alive packages, and four of the nine (44\%) dead packages.
+This could be indicative of a transition to such software development and version control tools for SCS.
+Typically there are several simultaneous development branches in these projects.
 
@@
 The use of document generation tools is mentioned in the artifacts of 12 software packages, and was noted during interviews with developers.
-Document generation tools that were mentioned include Sphinx and Doxygen.
+Sphinx and Doxygen are the tools that were mentioned.
 
@@
 
-\subsection{Pain Points - DONE 1st Draft}
+\subsection{Pain Points - DONE 1st Draft}\label{painpoints}
+
+This subsection answers the research question: What are the pain points for developers working on research software projects?
+What aspects of the existing processes, methodologies and tools do they consider as potentially needing improvement?
+How should processes, methodologies and tools be changed to improve software development and software quality?
 
@@
 \subsection{Quality Recommendations - DONE 1st Draft}\label{qualityrecommentations}
+
+This subsection answers the research question: For research software developers, what specific actions are taken to address software qualities?
 
@@
 
-\subsection{Designation Comparison}
+\subsection{Designation Comparison}\label{comparison}
+
+This subsection answers the research question: How does software designated as high quality by this methodology compare with top rated software by the community?
 
@@
 How does software designated as high quality by this methodology compare	with top rated software by the community?
+
+\subsection{Threats To Validity}\label{threats}
+
+This section will point to potential threats to the validity in the measurement section of data collection, for example from above:
+
+Interviews with the developers of software packages also revealed a more frequent use of both unit testing and continuous integration in the development process than was found by only assessing the artifacts.
+The use of automatic installation processes is also common.
+Most often this involved a Make script.
 
@@
 knowledge, and then transforming this knowledge into relevant software artifacts without having to manually duplicate knowledge into multiple artifacts \citep{zhao2018}.
+
+The following was moved from Tools Used subsection (consider adding future work subsection within conclusion, also consider subsection for ideas for future SOP such as asking the developers):
+
+Drasil could be considered as an alternative to generate artifacts since it is geared specifically towards research software domains like LBM software.
+
+future sop idea (from Results section):
+
+Developers were not asked to com-
+ment on the reproducibility of their source code from their requirements specifications and
+design documentation. This question should be considered in the next iteration of state of
+the practice assessments.
 

### 2021-06-23 StateOfPractice/Peter-Notes/Report/Report.tex
LBM 5.4 feedback changes

@@
 Several development process obstacles were identified.
-One developer noted that their small development team has a lack of time to implement new features.
+A developer of pyLBM noted that their small development team has a lack of time to implement new features.
 Small development teams are common for LBM software packages.
@@
 
-A lack of software development experience was noted by another developer.
-Many of the team members on their project are domain experts and there is a steep learning curve to get these team members to contribute good quality source code.
+A lack of software development experience was noted by the developer of TCLB.
+Many of the team members on their project are domain experts and there can be a steep learning curve before these team members contribute good quality source code.
 It was further noted that this has been somewhat addressed, as the code has been re-written to best ensure ease of future contributions.
 The same developer also noted that there is a lack of incentive and funding in academia for developing widely used scientific software.
-There are no journals that publish such source code.
-There is no organization helping with the development of good quality software.
+The importance of funding for scientific software has been discussed in \cite{gewaltig2012quality}, which notes that ``software tools are developed and maintained only for as long as there is explicit or implicit funding''.
+The develop also commented that there are no journals that publish such scientific software source code.
+However, some work has been done to address such concerns in \citep{smith2016software}, which presents a set of software citation principles and discusses ``how they could be used to implement software citation in the scholarly community'' \citep{katz2019software}.
+Another raised concern was that there are no organizations helping with the development of good quality software; but some do exist, including \href{https://bssw.io/}{Better Scientific Software (BSSw)}, \href{https://www.software.ac.uk/}{Software Sustainability Institute}, and \href{https://software-carpentry.org/}{Software Carpentry}.
+Some SCS developers may not be familiar with these organizations.
+
 Scientific software is often developed in-house by the very researchers that temporarily use it in their own research.
-This software has a defined user and purpose, and often does not meet the standards that would be required by external users.
-Such software is rarely commercialized, in which case it is not open-source.
+Empirical studies of such ``professional end-user development'' of SCS is noted in \citep{segal2007end}.
+This kind of software has a defined user and purpose, and often does not meet the standards that would be required by external users.
+It has been categorized as a ``private tool'' by \citep{gewaltig2012quality}, which notes that despite often being made freely available, ``it is not always clear that it is sufficiently mature in terms of domain coverage, validity, documentation or usability, to be useful to other researchers''.
+It is less common for such ad hoc scientific software to become ``user-ready software'', which ``is not only research-ready, but should have most of the attributes commonly expected of commercial software products including broadness of scope, robustness, demonstrable correctness and adequate documentation'' \citep{gewaltig2012quality}.
+As software becomes user-ready, it can become commercialized and closed-source.
 
@@
 Furthermore, the documentation generally only provides a shallow overview of the underlying CFD theory.
-Users would be well advised to already be familiar with these topics to or they should spend significant time referencing theory resources.
+Users would be well advised to already be familiar with these topics, or they should spend significant time referencing theory resources.
 The documentation instead generally focuses on explaining how to use the software.
-It is of course not feasible for package documentation to address the underlying metaphysics topics in detail, so it is advised that the package documentation links to resources that better explain the underlying theory.
+It is of course not feasible for package documentation to address the underlying physics topics in detail, so it is advised that the package documentation links to resources that better explain the underlying theory.
 Sometimes, frequently asked questions about the underlying theory are answered in the documentation.
+OpenLB has an artifact for such questions.
 
 Setting up parallelization was also noted as a technical pain point by one of the developers, and the introduction of continuous integration by another.
-Software development knowledge could mitigate such pain points.
+Software development knowledge, and automatic code generation, could mitigate such pain points.
 As already noted, many of the developers are domain experts and not professional software developers.
-One of the interviewed developers that has relatively more extensive software development experience noted that eliminating equivalent statements using macros had helped improve the quality of their source code, specifically helping with reusing code to run on both the CPU and GPU.
+The developer of TCLB noted that eliminating equivalent statements using macros had helped improve the quality of their source code, specifically helping with reusing code to run on both the CPU and GPU.
 
@@
 These tests compare the package output to known correct output using test cases.
-A developer commented that the amount of testing data that is needed for some cases is a problem as free testing services do not offer capabilities to store and process such large amounts of data, and in-house testing solutions needed to be created to address this limitation.
+The developer of TCLB commented that the amount of testing data that is needed for some cases is a problem as free testing services do not offer capabilities to store and process such large amounts of data, and in-house testing solutions needed to be created to address this limitation.
+The solution for this has been to limit the size of the testing problems, and running tests in small batches of few iterations.
 
@@
 These software packages are not like commercial software packages.
-They are generally set up to solve specific research problems, and are otfen primarily used by their developers.
+They are generally set up to solve specific research problems, and are often primarily used by their developers.
 While they are modifiable to solve similar problems, these modifications are not trivial.
+Better documentation, with attention on traceability, and automatic code generation are suggested when designing software for change, and would help with these modifications.
 So far this problem of on-boarding new users has been addressed by updating the documentation to better inform users of the underlying LBM theory and package requirements.
@@
 
-One of the interviewed developers commented that parts of their package's source code had been refactored to Python to help address usability issues.
+The interviewed developer of ESPResSo commented that parts of their package's source code had been refactored to Python to help address usability issues.
 Python was perceived as a much more usable language, and it would be easy for future users and developers to learn and understand the source code.
@@
 A few potential future obstacles were noted.
-One developer said that their source code had been written with a specific application in mind and that due to this there was too much coupling between components of the source code.
-This would have an impact on future modifiability and reusability when trying to extend the software, and the code would need to be refactored.
+The developer of ESPResSo noted that their source code had been written with a specific application in mind and that due to this there was too much coupling between components of the source code.
+This results in technical debt, and would have an impact on future modifiability and reusability when trying to extend the software, and the code would need to be refactored.
 
 As noted above, difficulties with ensuring future correctness could also arise. As new methods and functionality is added into the software, new test cases and test data will need to be developed.
+
+The developer comments have emphasized an importance on source code, while documentation seems to be of secondary importance.
+It must be stressed that improved management of documentation could improve package source code and help eliminate some of the developer concerns that were raised.
 

### 2021-06-23 StateOfPractice/Peter-Notes/Report/Report.tex
LBM 5.4 small edits

@@
 
-Developers were asked to comment on obstacles in their development process, obstacles encountered by users, and potential future obstacles.
-
-Several development process obstacles were identified.
+Developers were asked to comment on obstacles in their development process, obstacles encountered by users, and potential future obstacles. Several development process obstacles were identified.
+
 A developer of pyLBM noted that their small development team has a lack of time to implement new features.
@@
 External contributions are rare despite many of the projects accepting them.
+Aside from on-boarding new developers, time constraints could be mitigated by increasing developer efficiency, which could be addressed in several ways, including by improving the quality of documentation, or incorporating automatic code generation.
 
@@
 It was further noted that this has been somewhat addressed, as the code has been re-written to best ensure ease of future contributions.
+
 The same developer also noted that there is a lack of incentive and funding in academia for developing widely used scientific software.
 The importance of funding for scientific software has been discussed in \cite{gewaltig2012quality}, which notes that ``software tools are developed and maintained only for as long as there is explicit or implicit funding''.
-The develop also commented that there are no journals that publish such scientific software source code.
-However, some work has been done to address such concerns in \citep{smith2016software}, which presents a set of software citation principles and discusses ``how they could be used to implement software citation in the scholarly community'' \citep{katz2019software}.
+The developer further commented that there are no journals that publish such scientific software source code.
+However, there are ways to get such source code cited.
+Work has been done to address this in \citep{smith2016software}, which presents a set of software citation principles and discusses ``how they could be used to implement software citation in the scholarly community'' \citep{katz2019software}.
+
 Another raised concern was that there are no organizations helping with the development of good quality software; but some do exist, including \href{https://bssw.io/}{Better Scientific Software (BSSw)}, \href{https://www.software.ac.uk/}{Software Sustainability Institute}, and \href{https://software-carpentry.org/}{Software Carpentry}.
@@
 As software becomes user-ready, it can become commercialized and closed-source.
-
 Documentation of LBM software is noted as being important, but its quality could be improved.
-There is often no time or funding for maintaining quality documentation for software that is rarely used outside of the development team.
+As already noted, there is often no time or funding for maintaining quality documentation for software that is rarely used outside of the development team.
 Furthermore, the documentation generally only provides a shallow overview of the underlying CFD theory.

### 2021-06-23 StateOfPractice/Peter-Notes/Report/Report.tex
more updates to chapter 5 LBM report based on feedback

@@
 Only four packages (LB2D-Prime, LB3D-Prime, LIMBES, MP-LABS) appear to use vague identifiers, such as single letters for variables.
-Hard coded constants were observed in the source code of 12 packages.
+Hard coded symbolic constants were observed in the source code of 12 packages.
 The constants are used for various parameters, mathematical constants, and matrix definitions.
@@
 
-Furthermore, maintainability has been addressed by increasing source code modularity, reducing duplicate information, and improving abstraction by developing well-defined interfaces.
+Furthermore, maintainability has been addressed by increasing source code modularity, reducing duplicate information, and improving abstraction by developing well defined interfaces.
 This was noted by the developers of ESPResSo and pyLBM.
@@
 
-A high degree of code modularity and abstraction, ensured by separating components and hiding information behind well-define interfaces, was noted by developers as a measure to ensure the ease of future changes.
+A high degree of code modularity and abstraction, ensured by separating components and hiding information behind well defined interfaces, was noted by developers as a measure to ensure the ease of future changes.
 The developer of ESPResSo also noted that some of the code base was transitioned from C to C++, which could ease modifiability of that software package.
@@
 The developer comments have emphasized an importance on source code, while documentation seems to be of secondary importance.
-It must be stressed that improved management of documentation could improve package source code and help eliminate some of the developer concerns that were raised.
+It must be stressed that improved management of documentation could benefit development and help eliminate some of the developer concerns that were raised.
+The use of automatic document generation tools that capture scientific and computing knowledge, and transform it into software artifacts, is advised.
+Drasil is an automatic document generation tool that is further discussed in Section \ref{conclusion}.
 
@@
 The following points regarding software quality should be considered when developing LBM software packages.
+These points are based on developer interviews, SCS literature, what was found to have worked for packages that were designated as high quality in this assessment.
 
@@
 	\item Installation instructions should be written as if the user does not have any dependencies installed and is installing on a clean OS.
+This can be tested on a clean environment using a virtual machine.
 	\item List all dependencies in the installation instructions, dependency versions, and how to install them.
-	\item If possible, automate the installation of dependencies.
-	\item Automate the installation process as much as possible.
+	\item If possible, automate the installation of dependencies. Use tools such as Make. 
+	\item Automate the installation process as much as possible. Use tools such as Make.
 	\item The installation instruction should only be in one location.
-	\item Include descriptive error messages for errors encountered during installation.
+	\item Include descriptive error messages for errors encountered during installation. This was done by several top ranked packages, including Ludwig and LUMA
 	\item Include a way to validate the installation.
-	\item Include instructions for uninstalling the software.
+This can be done using a custom script, or a test case that specifies expected output.
+This was done by several top ranked packages, including OpenLB, Ludwig, LUMA, and Palabos.
+	\item Include instructions for uninstalling the software. These were included with pyLBM.
 \end{itemize}
@@
 \begin{itemize}
-	\item Use a requirements specification document.
-	\item Make public the requirements specification document, or explicitly reference the domain theory that the software is designed from.
+	\item Use a requirements specification document. A template 
+	\item Make public (on package website or GitHub) the requirements specification document, or explicitly reference the domain theory that the software is designed from.
 	\item Ensure the above information is easy to find. Consider adding it to the user manual. 
 	\item Development teams should include both domain experts and experienced software developers.
-	\item Use and make public detailed documentation. Consider using automatic document generation tools.
-	\item Include detailed tutorials that include expected output.
-	\item Use unit testing during development and make them public.
-	\item Modularize the source code.
-	\item Use continuous integration tools and processes during development.
+	\item Use and make public (on package website or GitHub) detailed documentation. Consider using automatic document generation tools like Doxygen, Drasil, or Sphinx.
+	\item Include detailed tutorials that include expected output, like the \href{https://www.walberla.net/doxygen/index.html}{waLBerla tutorials}.
+	\item Use unit tests during development and make them public (on package website or GitHub). This was available for Ludwig.
+	\item Modularize the source code, separate components, hide information  behind well defined interfaces.
+	\item Use continuous integration tools and processes (Bamboo, Jenkins, and Travis CI) during development. This was done by top ranked packages ESPResSo and Ludwig.
 \end{itemize}
@@
 \begin{itemize}
-	\item Include descriptive error messages where appropriate.
+	\item Include descriptive error messages where appropriate. This was done by most packages that encountered a fault.
 	\item In case automatic installation of dependencies fails, the system should indicate to the user what dependencies need to be installed manually.
+This was done by many packages, including ESPResSo++, Ludwig, LUMA, pyLBM, TCLB.
 	\item The packages should include detailed tutorials, including dependencies, expected output, and any additional supplementary documentation that may be required.
+This was done by many packages, including waLBerla, Palabos, MechSys, LUMA, pyLBM.
 \end{itemize}
@@
 	\item The system must provide descriptive error messages when it encounters unexpected input, including incorrect data types, empty input, and missing files or links.
+Eighteen of the software packages behaved in such reasonable fashion when tested with unexpected input.
 \end{itemize}
@@
 \begin{itemize}
-	\item Integrate parallelization tools and techniques to reduce processing time. Consider GPU processing, CUDA, and MPI. 
-	\item The user should be able to choose to process their model on either the CPU or GPU.
+	\item Integrate parallelization tools and techniques to reduce processing time. Consider GPU processing, CUDA, and MPI. Seventeen software packages mentioned parallelization in their artifacts.
+	\item The user should be able to choose to process their model on either the CPU or GPU. ESPResSo, lbmpy, lettuce, pyLBM, and TCLB allow for either CPU or GPU processing.
 \end{itemize}
@@
 	\item Include user hardware and software requirements documentation.
-	\item Include a user tutorial that indicates the expected output.
-	\item Include a detailed user manual. It should identify elements of user interfaces, and identify all requirements to model a system.	
-	\item State appropriate fluid dynamics problems that the software is designed to model in the documentation, and explicitly state the limits of the software. 
+Hardware requirements are rarely listed in the software packages that were assessed.
+Only those offering GPU processing (ESPResSo, lbmpy, lettuce, pyLBM, Sailfish, TCLB) mentioned any hardware requirements.
+On the other hand, some sort of software requirements were available for all packages, even if only some dependencies or a compatible operating system.
+	\item Include a user tutorial that indicates the expected output. This was done by many packages, including waLBerla, Palabos, MechSys, LUMA, pyLBM.
+	\item Provide a detailed user manual.
+It should identify elements of user interfaces, and identify all requirements to model a system.
+Thirteen packages have a user manual.
+ESPResSo has a well detailed manual.
+	\item State appropriate fluid dynamics problems that the software is designed to model in the documentation, and explicitly state the limits of the software. This is done in the ESPResSo user guide.
 	\item Provide documentation that details the background theory information, or provide a reference to such information.
-	\item Identify expected user characteristics.
+	\item Identify expected user characteristics. LIMBES, Ludwig, laboetie, and Palabos did this. The importance of specifying user characteristics is discussed in \citep{smith2007requirements}.
 	\item Keep all documentation in one location.
@@
 	\item Maintain a user support model (Git, email, forum, FAQ)
-	\item If possible, consider using popular user-friendly software languages like Python. Especially consider this for parts of the source code that is likely to be modified or reviewed by users.
+	\item If possible, consider using popular user-friendly software languages like Python.
+Especially consider this for parts of the source code that is likely to be modified or reviewed by users.
+ESPResSo and Sailfish use Python to shorted development time and improve usability.
 \end{itemize}
@@
 	\item Ideally include most of the common and less common artifacts listed in Section \ref{artifacts}.
-	\item Include version numbers and release notes for all major source code and artifact releases. 
-	\item Have a defined process for accepting contributions.
+	\item Include version numbers and release notes for all major source code and artifact releases. The top five ranked software packages include release notes.
+	\item Have a defined process for accepting contributions. The top five ranked software packages include information on how to contribute.
 	\item Include documentation for making contributions to the project. 
-	\item Use an issue tracker (Git, email, SourceForge, other) to manage bugs and changes. Issues should be regularly reviewed and closed. 
-	\item Use a version control system (GitHib, CVS).
-	\item Source code needs to be well commented. Typically, more than 10 percent of LBM package source code is comments. 
-	\item Ensure source code is modular. Generally, this means that source code is split into files along functionality.
+	\item Use an issue tracker (Git, email, SourceForge, other) to manage bugs and changes.
+Issues should be regularly reviewed and closed.
+All but three (laboetie, lettuce, Sailfish) of the software packages that use Git have most of their issues closed.
+	\item Use a version control system (GitHib, CVS). Four (ESPREsSo, Ludwig, LUMA, Palabos) of the top five ranked packages use a version control system.
+	\item Source code needs to be well commented. Typically, more than 10 percent of LBM package source code is comments, as presented in Table \ref{gitrepodata}.
+	\item Modularize the source code, separate components, hide information behind well defined interfaces.
 	\item Eliminate code duplication.
-	\item If possible, consider using popular user-friendly software languages like Python. Especially consider this for parts of the source code that is likely to be modified or reviewed by users.
+	\item If possible, consider using popular user-friendly software languages like Python.
+Especially consider this for parts of the source code that is likely to be modified or reviewed by users.
+ESPResSo and Sailfish use Python to address several software qualities, including maintainability.
+	\item Consider the recommended points for addressing traceability.
 \end{itemize}
@@
 \begin{itemize}
-	\item Ensure a high degree of code modularity and abstraction.
-	\item If possible, consider using popular user-friendly software languages like Python. Especially consider this for parts of the source code that is likely to be modified or reviewed by users.
-	\item Consider future source code modifiabilility as early as the design stage of development.
-	\item Consider flexibility of data structures and data storage in the design stage.
-	\item Provide validation benchmark tools and data to the public.
+	\item Modularize the source code, separate components, hide information behind well defined interfaces.
+	\item If possible, consider using popular user-friendly software languages like Python.
+Especially consider this for parts of the source code that is likely to be modified or reviewed by users.
+ESPResSo and Sailfish use Python to address several software qualities, including modifiability.
+	\item Consider future source code modifiability as early as the design stage of development.
+	\item Consider flexibility of data structures and data storage in the design stage. The package pyLBM redeveloped data structures to ease future changes.
 \end{itemize}
@@
 \begin{itemize}
-	\item Ensure a high degree of code modularity and abstraction.
+	\item Modularize the source code, separate components, hide information behind well defined interfaces.
 	\item Document module interfaces in design and developer documentation. 
-	\item Provide API documentation, if applicable.
+	\item Provide API documentation, if applicable. Only one (ESPResSo) of the top five ranked packages provided API documentation.
 \end{itemize}
@@
 \begin{itemize}
-	\item Maintain consistent formatting style across source code files.
-	\item Identify a coding standard in the documentation, if a coding standard is used during development.
-	\item Use consistent, distinctive, and meaningful code identifiers. 
-	\item Identify algorithms that are used.
-	\item Hard code domain constants.
-	\item Add meaningful comments. They should indicate what is being done in each section of source code.
-	\item Ensure a high degree of code modularity and abstraction.
-	\item Include user guides, and state technical and problem modeling requirements.
-	\item State problem cases that the software is designed to model in the documentation, and explicitly state the limits of the software. 
-	\item Consider adding a FAQ section to the documentation. 
+	\item Adopt a coding standard and document it in artifacts, including some examples. A coding standard needs to be part of continuous integration. Only
+	LUMA and HemeLB explicitly identify coding standards. 
+	\item Use consistent, distinctive, and meaningful code identifiers. Nineteen of the 23 software packages use such code identifiers.
+	\item Use symbolic constants. Twelve of the 23 software packages use such constants.
+	\item Identify algorithms that are used in source code comments.
+Document them in the artifacts.
+Cite relevant external sources if needed.
+Domain algorithms are noted in the source code of 11 packages.
+	\item Add meaningful comments. Indicate what is being done in each section of source code.
+	\item Modularize the source code, separate components, hide information behind well defined interfaces.
+	\item Include user manuals that identify all technical and problem modeling requirements. Thirteen software packages have a user manual.
+	\item State appropriate fluid dynamics problems that the software is designed to model in the documentation, and explicitly state the limits of the software. This is done in the ESPResSo user guide. 
+	\item Consider adding a FAQ section to the documentation. This helped resolve some usability issues for OpenLB.
 \end{itemize}
@@
 	\item Update all relevant documentation when a change to the software is made.
-	\item Use automatic document generation tools to limit the time spent on updating documentation.
-	\item Provide a developers guide.
-	\item Provide a resource that helps orient users and developers to the artifacts, source code, and system architecture.
-	\item Highlight new features and changes between artifact versions. 
+	\item Use automatic document generation tools (Doxygen, Drasil, Sphinx) to limit the time spent on updating documentation.
+	\item Provide a resource, like a developer's guide, that helps orient developers to the artifacts, source code, and system architecture so that they can better document changes.  
 \end{itemize}
@@
 \begin{itemize}
-	\item Identify the development model that was used, if one was used.
-	\item Identify the development environment.
-	\item Identify the development process and how new users can contribute. 
-	\item Include notes with all releases.
-	\item Ensure a high degree of code modularity and abstraction.
-	\item Communicate all changes to the development team.
-	\item Use continuous integration processes and tools.
-	\item Consider peer review processes to assess contributions. 
-	\item Eliminate code duplication.
-	\item Update all relevant developer documentation when a change to the software is made.
-	\item Use project management software.
+	\item Summarize the development process that is used. Provide information on how new users can contribute. Identify the development model by name (waterfall, agile, etc.), if appropriate.
+	\item Identify the development environment. Eight of the 23 software packages identify the development environment.
+	\item Update all relevant documentation when a change to the software is made.
+	\item Include notes with all releases. Nine of the packages include release notes.
+	\item Communicate all changes within the development team.
+	\item Use continuous integration processes and tools (Bamboo, Jenkins, and Travis CI). This was done by top ranked packages ESPResSo and Ludwig.
+	\item Consider peer review processes to assess contributions and ensure the tracking of information. High ranked package ESPResSo uses a peer review process for contributions. 
+	\item Use project management tools, including change and version control tools (GitHub, GitLab, CVS), collaboration tools (GitHub, GitLab), and document generation tools (Doxygen, Drasil, Sphinx).
 \end{itemize}
@@
 \begin{itemize}
-	\item Test output against manually calculated or known correct results.
-	\item Automate the testing of output. Run tests after all code changes.
-	\item Consider peer review processes to assess contributions. 
+	\item Test output against automatically calculated or known correct results. High ranked package ESPResSo does this.
+	\item Automate the testing of output. Run tests after all code changes. High ranked package ESPResSo does this.
+	\item Consider peer review processes and task based inspection to assess contributions. High ranked package ESPResSo does this.
 	\item Update all relevant design documentation when a change to the software is made.
@@
 \begin{itemize}
-	\item Documentation must state all technical requirements and software dependencies. 
-	\item A user guide must be provided.
-	\item State appropriate fluid dynamics problems that the software is designed to model in the documentation, and explicitly state the limits of the software. 
+	\item Documentation must state all technical requirements and software dependencies. A missing dependency was a frequent cause of fault conditions during this assessment.
+	\item Provide a detailed user manual. Thirteen packages have a user manual. ESPResSo has a well detailed manual.
+	\item Provide a table of symbols in the developer guide that map to names used in the course code.
+	\item State appropriate fluid dynamics problems that the software is designed to model in the documentation, and explicitly state the limits of the software. This is done in the ESPResSo user guide.
 	\item The documentation should either explain the underlying CFD theories or provide a reference to appropriate resources.
-	\item Consider asking users for feedback on the documentation, specifically if it clearly conveys all necessary information to the users.
+	\item Consider asking users for feedback on the documentation. The developer of pyLBM noted that such feedback would be appreciated.
 \end{itemize}
@@
 \newpage
-\section{Conclusion}
+\section{Conclusion}\label{conclusion}
 
@@
 the practice assessments.
+
+On page 86 of the ReportChapt5Feedback pdf SS commented that some of the adivuce from the quality recommendations section would be difficult to do for less mature projects.
+"We should add some getting started advice for the biggest ROI changes".
+In this section I should add a subsection for recommendations, I should highlight the quality recommendations that are pervasive for multiple qualities, like modularization and using software languages that are considered user friendly (Python), etc.
 

### 2021-06-24 StateOfPractice/Peter-Notes/DomainExptMeeting/Presentation for the Introductory Meeting.tex
Add domain specific domain expert meeting slides

%\documentclass[handout]{beamer} 
\documentclass[t,12pt,numbers,fleqn]{beamer}
%\documentclass[ignorenonframetext]{beamer}

\newif\ifquestions
%\questionstrue
\questionsfalse

\usepackage{pgfpages} 
\usepackage{hyperref}
\hypersetup{colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    filecolor=blue,
    urlcolor=blue,
    unicode=false}
\urlstyle{same}

\usepackage{booktabs}
\usepackage{bibentry}
\usepackage[round, authoryear]{natbib}
\bibliographystyle{plainnat}

%\usetheme{Iimenau}

\useoutertheme{split} %so the footline can be seen, without needing pgfpages

%\pgfpagesuselayout{resize to}[letterpaper,border shrink=5mm,landscape]  %if this is uncommented, the hyperref links do not work

\mode<presentation>{}

\input{./def-beamer}

\newcommand{\topic}{State of the Practice for Lattice Boltzmann Solvers}

\input{./titlepage}

\begin{document}

%\nobibliography{../../../CommonFiles/ResearchProposal}

\input{./footline}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}
\frametitle{Overview}

\bi
\item Goals
  \bi
  \item Understand the state of the software development practices for
    LBM Software
  \item Make recommendations for improvements
  \item A publication that is useful to the community
  \ei  
\item We have developed a standard methdology for assessing SOP for Domain X
\item The methodology requires a \textbf{domain expert} to:
  \bi
  \item Vet the preliminary results
  \item Assess the feasibility of the recommendations
  \item Navigate the publication process
  \item Answer \textbf{developer} interview questions on pain points  
  \ei
\item Today's meeting
  \bi
\item Informal
\item Questions do not have to be answered in real time, or by domain expert
  \ei
\ei

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}
\frametitle{Overall Process}

\begin{enumerate}
%\item Identify the domain
\item \emph{Domain Expert}: Create a top ten list
\item Brief \emph{Domain Expert}
\item Initial list of candidate software packages
\item \emph{Domain Expert}: Vet domain software list
\item Domain Analysis
\item \emph{Domain Expert}: Vet domain analysis
%\item Gather source code and documentation for software
\item Collect empirical measures
\item Measure using measurement template
\item Interview developers
\item Use AHP process to rank the software packages
\item \emph{Domain Expert}: Vet AHP ranking
\item \emph{Domain Expert}: Review recommendations
\end{enumerate}

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}
\frametitle{Scope}

[Ao put your scope figure here.  Peter, you can delete this slide.]

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}
\frametitle{Vet Software List}

\bi
\item How does our list compare to the domain expert's list?
\item Is any software missing?
\item Is there software that should be included?
\item Any other questions/comments or concerns?  
\ei

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}
\frametitle{Software List}

[Peter and Ao, please put your list here, with hyperlinks if possible.  You can use
columns to make it all fit on one slide.]

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}
\frametitle{Domain Expert Software List}

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}
\frametitle{Domain Analysis}

\bi
\item So users can pick the right software for their task
\item Think of the software as a family of related software products, like:
\bi
\item Family of automobiles
\item Computer hardware
\item ...
\ei
\item Commonalities
\item Variabilities
\item Parameters of variation
\item Do the following slides capture the commonalities and variabilities?
\item We can follow up over e-mail, if that is easier
\ei

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}
\frametitle{Commonalities}

[Peter, please list your commonalities here; Ao, we'll use a scope slide
instead]

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}
\frametitle{Variabilities}

[Peter, please list your variabilities, with their associated parameters of
variation here; Ao, you can delete this slide]

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}
\frametitle{Summary of Measures}

\bi
\item Empirical measures
  \bi
  \item Examples
  \ei
\item Measurement template
  \bi
  \item qualities
  \ei
\item Interview questions
  \bi
  \item Examples
  \ei
  \item AHP Process with pairwise comparisons
\ei
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}
\frametitle{Thoughts on Overall Ranking?}

[Peter and Ao, give the overall ranking graphic for your domain]

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}
\frametitle{Ranking Follow-Up}

\bi
\item We would like feedback on the ranking for each of the qualities
\item Is it feasible to review all 10 graphs, and the associated write-up?
\item Maybe there is a grad student that can review the rankings?
 \ei

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}
\frametitle{Discuss Recommendations}

\bi
\item 
\ei

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}
\frametitle{Publication}

\bi
\item Where should we publish this paper?
\item Who do you see as the targeted readers?
\ei

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}
\frametitle{Developer Questions (for next time)}

\bi
\item 
\ei

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{document}

### 2021-06-24 StateOfPractice/Ao-Notes/DomainExptMeeting/Presentation for the Introductory Meeting.tex
Update to Ao presentation slides to make them more Ao specific

--- 2021-06-24 StateOfPractice/Peter-Notes/DomainExptMeeting/Presentation for the Introductory Meeting.tex
@@
 \begin{frame}
-\frametitle{Domain Analysis}
-
-\bi
-\item So users can pick the right software for their task
-\item Think of the software as a family of related software products, like:
-\bi
-\item Family of automobiles
-\item Computer hardware
-\item ...
-\ei
-\item Commonalities
-\item Variabilities
-\item Parameters of variation
-\item Do the following slides capture the commonalities and variabilities?
-\item We can follow up over e-mail, if that is easier
-\ei
-
-\end{frame}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{frame}
-\frametitle{Commonalities}
-
-[Peter, please list your commonalities here; Ao, we'll use a scope slide
-instead]
-
-\end{frame}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{frame}
-\frametitle{Variabilities}
-
-[Peter, please list your variabilities, with their associated parameters of
-variation here; Ao, you can delete this slide]
-
-\end{frame}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{frame}
 \frametitle{Summary of Measures}

### 2021-06-24 StateOfPractice/Peter-Notes/DomainExptMeeting/Presentation for the Introductory Meeting.tex
Edits to slides to remove from Peter's slides parts that are Ao specific
updated LBM domain expert presentation

@@
 
+\usepackage{multicol}
 \usepackage{booktabs}
@@
 \begin{frame}
-\frametitle{Scope}
-
-[Ao put your scope figure here.  Peter, you can delete this slide.]
-
-\end{frame}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{frame}
 \frametitle{Vet Software List}
@@
 
-[Peter and Ao, please put your list here, with hyperlinks if possible.  You can use
-columns to make it all fit on one slide.]
+\begin{multicols}{2}	
+	\bi
+			\item \href{https://ludwig.epcc.ed.ac.uk/}{Ludwig}
+			\item \href{http://espressomd.org/wordpress/}{ESPResSo}
+			\item \href{https://palabos.unige.ch/}{Palabos}
+			\item \href{https://www.openlb.net/}{OpenLB}
+			\item \href{https://github.com/aharwood2/LUMA/}{LUMA}
+			\item \href{https://pypi.org/project/pylbm/}{pyLBM}
+			\item \href{https://www.scd.stfc.ac.uk/Pages/DL_MESO.aspx}{DL\_MESO (LBE)}
+			\item \href{https://www.walberla.net}{waLBerla}
+			\item \href{http://sailfish.us.edu.pl/}{Sailfish}
+			\item \href{https://github.com/maxlevesque/laboetie}{laboetie}
+			\item \href{https://docs.tclb.io/}{TCLB}
+			\item \href{http://mechsys.nongnu.org/}{MechSys}
+			\item \href{https://github.com/Olllom/lettuce}{lettuce}
+			\item \href{https://github.com/espressopp/espressopp}{ESPResSo++} 
+			\item \href{https://github.com/carlosrosales/mplabs}{MP-LABS}		
+			\item \href{http://sunlightlb.sourceforge.net/}{SunlightLB}
+			\item \href{http://ccs.chem.ucl.ac.uk/lb3d}{LB3D} - 			
+			\href{https://www.sciencedirect.com/science/article/pii/S0010465517301017}{paper}
+			\item \href{https://code.google.com/archive/p/limbes/}{LIMBES}
+			\item \href{http://faculty.fiu.edu/~sukopm/LBnD_Prime/LBnD_Prime.html}{LB2D-Prime}	
+			\item \href{https://github.com/UCL/hemelb}{HemeLB} 
+			\item \href{https://pypi.org/project/lbmpy/}{lbmpy}	
+			\item \href{http://faculty.fiu.edu/~sukopm/LBnD_Prime/LBnD_Prime.html}{LB3D-Prime}	
+			\item \href{https://github.com/UCL/LatBo.jl}{LatBo.jl}
+	\ei
+\end{multicols}
 
@@
 
-[Peter, please list your commonalities here; Ao, we'll use a scope slide
-instead]
+\bi
+	\item Lattice - discretized domain within a boundary
+	\item Collision operator (The Bhatnagar-Gross-Krook Collision Operator is common)
+	\item Probability density function
+	\item Equilibrium distribution function
+	\item Boltzmann transport equation
+\ei
 
@@
 
-[Peter, please list your variabilities, with their associated parameters of
-variation here; Ao, you can delete this slide]
+\bi
+\item May use parallel processing
+\item Different equilibrium distribution functions
+\item Coefficients for equilibrium distribution function (based on velocity directions)
+\item Lattice dimensions (1D, 2D, 3D)
+\item Varied velocity directions, partially based on dimensions 
+\item Collision operators (SRT, TRT, MRT, BGK)
+\item Collision vs collision-free transport equations
+\item Number of fluid that can be modeled simultaneously
+\item Fluid parameters (Reynolds Num., density, viscosity, etc.)
+\item Lattice boundary (reflective or non-reflective)
+\ei
 

### 2021-06-24 StateOfPractice/Peter-Notes/DomainExptMeeting/Presentation for the Introductory Meeting.tex
updated LBM domain expert presentation 2

@@
 \bi
-\item May use parallel processing
-\item Different equilibrium distribution functions
-\item Coefficients for equilibrium distribution function (based on velocity directions)
+\item May use parallel processing (May use varied tools for this - MPI, CUDA..)
+\item Different equilibrium distribution functions (compressible and incompressible)
+\item Coefficients for equilibrium distribution function (based on velocity directions - 2,3,5,9,13,15,19,27)
 \item Lattice dimensions (1D, 2D, 3D)
-\item Varied velocity directions, partially based on dimensions 
+\item Varied velocity directions, partially based on dimensions (2,3,5,9,13,15,19,27)
 \item Collision operators (SRT, TRT, MRT, BGK)
 \item Collision vs collision-free transport equations
-\item Number of fluid that can be modeled simultaneously
+\item Number of fluids that can be modeled simultaneously
 \item Fluid parameters (Reynolds Num., density, viscosity, etc.)

### 2021-06-25 StateOfPractice/Peter-Notes/DomainExptMeeting/Presentation for the Introductory Meeting.tex
Addition of slides comparing the domain expert and SOP lists
Complete draft of slides for Zahra meeting

@@
 \bi
-\item How does our list compare to the domain expert's list?
-\item Is any software missing?
-\item Is there software that should be included?
-\item Any other questions/comments or concerns?  
+\item \structure{How does our list compare to the domain expert's list?}
+\item \structure{Is any software missing?}
+\item \structure{Is there software that should be included?}
+\item \structure{Any other questions/comments or concerns?}
 \ei
@@
 \begin{frame}
-\frametitle{Software List}
+\frametitle{SOP Software List}
 
@@
 \begin{frame}
-\frametitle{Domain Expert Software List}
+\frametitle{Common With Domain Expert List}
+
+\begin{multicols}{2}	
+  \bi
+\item \href{https://palabos.unige.ch/}{Palabos}
+\item \href{https://www.openlb.net/}{OpenLB}
+\item \href{https://github.com/aharwood2/LUMA/}{LUMA}
+\item \href{https://www.walberla.net}{waLBerla}
+\item \href{http://sailfish.us.edu.pl/}{Sailfish}
+\item \href{https://github.com/UCL/hemelb}{HemeLB}
+\item \href{http://ccs.chem.ucl.ac.uk/lb3d}{LB3D} - \href{https://www.sciencedirect.com/science/article/pii/S0010465517301017}{paper}
+\item \href{https://github.com/carlosrosales/mplabs}{MP-LABS}
+\item \href{http://sunlightlb.sourceforge.net/}{SunlightLB}
+\item \href{https://pypi.org/project/pylbm/}{pyLBM}
+\item  \href{https://www.scd.stfc.ac.uk/Pages/DL_MESO.aspx}{DL\_MESO (LBE)}
+\item \href{https://code.google.com/archive/p/limbes/}{LIMBES}
+\item \href{https://docs.tclb.io/}{TCLB} (also called CudneLB)
+\ei
+\end{multicols}
+
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{frame}
+\frametitle{Only on Domain Expert List}
+
+%\begin{multicols}{2}	
+  \bi
+\item ASL (general purpose tool for PDEs, LBM a small part)
+\item ch4-project (no docs, no installation guide)
+\item Open FSI (uses Palabos for LBM, new project)
+\item LIFE (small and new project)
+\item LBSIM (dead project)
+\item Taxila LBM (dead project, could not install, broken links in
+  documentation)                    
+	\ei
+%\end{multicols}
+
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{frame}
+\frametitle{Only on SOP List}
+
+\begin{multicols}{2}	
+  \bi
+			\item \href{https://ludwig.epcc.ed.ac.uk/}{Ludwig}
+			\item \href{http://espressomd.org/wordpress/}{ESPResSo}
+
+			\item \href{https://github.com/maxlevesque/laboetie}{laboetie}
+			\item \href{http://mechsys.nongnu.org/}{MechSys}
+			\item \href{https://github.com/Olllom/lettuce}{lettuce}
+			\item \href{https://github.com/espressopp/espressopp}{ESPResSo++}
+
+			\item \href{http://faculty.fiu.edu/~sukopm/LBnD_Prime/LBnD_Prime.html}{LB2D-Prime}
+
+			\item \href{https://pypi.org/project/lbmpy/}{lbmpy}	
+			\item \href{http://faculty.fiu.edu/~sukopm/LBnD_Prime/LBnD_Prime.html}{LB3D-Prime}	
+			\item \href{https://github.com/UCL/LatBo.jl}{LatBo.jl}
+	\ei
+\end{multicols}
+
+\structure{Why is pylbm on the domain expert list twice, should one be lbmpy?}
 
@@
 \item Computer hardware
-\item ...
+\item etc.
 \ei
@@
 \item Parameters of variation
-\item Do the following slides capture the commonalities and variabilities?
+\item \structure{Do the following slides capture the commonalities and variabilities?}
 \item We can follow up over e-mail, if that is easier
@@
 \begin{frame}
-\frametitle{Variabilities}
+\frametitle{SOP Variabilities}
 
@@
 \item Different equilibrium distribution functions (compressible and incompressible)
-\item Coefficients for equilibrium distribution function (based on velocity directions - 2,3,5,9,13,15,19,27)
+\item Coefficients for equilibrium distribution function (based on velocity
+  directions - 2, 3, 5, 9, 13, 15, 19, 27)
 \item Lattice dimensions (1D, 2D, 3D)
-\item Varied velocity directions, partially based on dimensions (2,3,5,9,13,15,19,27)
+\item Varied velocity directions, partially based on dimensions (2, 3, 5, 9, 13,
+  15, 19, 27)
 \item Collision operators (SRT, TRT, MRT, BGK)
@@
 \item Number of fluids that can be modeled simultaneously
-\item Fluid parameters (Reynolds Num., density, viscosity, etc.)
+\item Fluid parameters (Reynolds num., density, viscosity, etc.)
 \item Lattice boundary (reflective or non-reflective)
+\ei
+
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{frame}
+\frametitle{Domain Expert Variabilities}
+
+\bi
+\item May use parallel processing (MPI, CUDA, OpenMP, ...)
+\item Lattice dimensions (1D, 2D, 3D)
+\item Turbulent (Yes, No)
+\item FSI (Fluid Structure Interaction)
+\item Complex Geometry
+\item Non-Newtonian Fluids
+\item Programming Language (C++, Python, etc)
 \ei
@@
 \bi
-\item Empirical measures
-  \bi
-  \item Examples
-  \ei
+\item Empirical measures (number of stars, forks, pull requests, lines of code, etc)
 \item Measurement template
   \bi
-  \item qualities
+\item Installability
+\item Correctness and verifiability
+\item Surface reliability
+\item Surface robustness
+\item Surface usability
+\item Maintainability
+\item Surface Understandability
+  \item Visibility and transparency
   \ei
-\item Interview questions
-  \bi
-  \item Examples
-  \ei
+\item Developer interviews (looking for pain points)
 \item AHP Process with pairwise comparisons
@@
 
-[Peter and Ao, give the overall ranking graphic for your domain]
+\bigskip
+
+    \includegraphics[width=1.0\textwidth]{finalscore_chart.png}
 
@@
 \bi
-\item We would like feedback on the ranking for each of the qualities
-\item Is it feasible to review all 10 graphs, and the associated write-up?
-\item Maybe there is a grad student that can review the rankings?
+\item \structure{We would like feedback on the ranking for each of the qualities}
+\item \structure{Is it feasible to review all 10 graphs, and the associated write-up?}
+\item \structure{Maybe there is a grad student that can review the rankings?}
 \ei
@@
 \begin{frame}
-\frametitle{Discuss Recommendations}
+\frametitle{Pain Points from Developer Interviews}
 
+\structure{Do these fit with your experience?}
 \bi
-\item 
+\item Lack of time to implement new features
+\item Team members lack software development experience
+\item Lack of funding for software development
+\item No organizations to help with developing high quality software
+\item Documentation could be improved, only has shallow overview of the theory
+\item Setting up parallelization is challenging
+\item Setting up continuous integration is challenging
+\item Difficulties in ensuring correctness because of size of test cases
+\item Users sometimes use tools inappropriately
+\item Technical debt due to coupling between components
+  \ei
+
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{frame}
+\frametitle{Recommendations}
+
+\structure{Do these seem feasible?  What other ideas do we have?}
+\bi
+\item Consult with software development organization
+  \bi
+\item Better Scientific Software (BSSw)
+\item Software Sustainability Institute
+\item Software Carpentry
+  \ei
+\item Citations for software (Katz project)
+  \item Redefine productivity to include time working on tasks like testing, continuous
+    integration and documentation
 \ei
@@
 \bi
-\item Where should we publish this paper?
-\item Who do you see as the targeted readers?
-\ei
-
-\end{frame}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{frame}
-\frametitle{Developer Questions (for next time)}
-
-\bi
-\item 
+\item \structure{Who do you see as the targeted readers?}
+\item \structure{Where should we publish this paper?}
 \ei

### 2021-06-25 StateOfPractice/Peter-Notes/Report/Report.tex
edits to Chapter 5 of LBM Report

@@
 \pagenumbering{arabic}
-\section{Introduction - DONE SECTION 1st DRAFT}
+\section{Introduction}
 
@@
 
-\section{Domain Analysis - DONE SECTION 1st DRAFT}\label{domainanalysis}
+\section{Domain Analysis}\label{domainanalysis}
 
@@
 \newpage
-\section{Methodology - DONE SECTION 1st DRAFT}
+\section{Methodology}
 
@@
 \newpage
-\section{Results - DONE SECTION 1st DRAFT}\label{qualityresults}
+\section{Results}\label{qualityresults}
 
@@
 Packages with a high reusability score do not have many LOC per text file, generally having a few hundred lines or less.
-This suggests that the source code of these packages is likely functionally modularized, and modules could be reused in other projects.
-
-There was a strong focus on modularity when designing the waLBerla framework to enhance productivity, reusability, and maintainability \citep{bauer2021walberla}.
-Its software design has enabled waLBerla to be successfully applied in several projects as a basis for various extensions \citep{bauer2021walberla}.\\
-
+This suggests that the source code of these packages is likely functionally modularized, and modules could be reused in other projects.\\
+
+	\begin{center}
 		\begin{onehalfspacing}
-	\begin{center}
 		\begin{tabular}{ p{3.5cm}p{2cm}p{2.5cm}p{2cm}p{2.5cm} }
@@
 		\captionof{table}{Module Data}\label{moduledata}
+		\end{onehalfspacing}
 	\end{center}
-\end{onehalfspacing}
+
+There was a strong focus on modularity when designing the waLBerla framework to enhance productivity, reusability, and maintainability \citep{bauer2021walberla}.
+Its software design has enabled waLBerla to be successfully applied in several projects as a basis for various extensions \citep{bauer2021walberla}.
 
@@
 
-\subsection{Artifacts Present - DONE 1st DRAFT}\label{artifacts}
+\subsection{Artifacts Present}\label{artifacts}
 
@@
 
-\subsubsection{Less Common Artifacts}
+\subsubsection{Less Common Artifacts}\label{lesscommon}
 The following artifacts were less commonly found in the 23 software packages that were tested. The top four AHP ranked packages have most of these artifacts.
@@
 
-\subsection{Tools Used - DONE 1st Draft}\label{tools}
+\subsection{Tools Used}\label{tools}
 
@@
 SciPy).
-Libraries that are explicitly stated in artifacts, and were not noted during our observations, may fall outside of these categories.
+Libraries that are not explicitly stated in artifacts, or were not noted during our observations, may fall outside of these categories.
 
@@
 Thirteen (57\%) of the 23 software packages use GitHub for managing the project, including nine (64\%) of the 14 alive packages, and four of the nine (44\%) dead packages.
+Two projects (Palabos, WaLBerla) use GitLab.
 This could be indicative of a transition to such software development and version control tools for SCS.
@@
 
-\subsection{Pain Points - DONE 1st Draft}\label{painpoints}
+\subsection{Pain Points}\label{painpoints}
 
@@
 
-Developers were asked to comment on obstacles in their development process, obstacles encountered by users, and potential future obstacles. Several development process obstacles were identified.
+Developers were asked to comment on obstacles in their development process, obstacles encountered by users, and potential future obstacles.
 
@@
 The developer of TCLB commented that the amount of testing data that is needed for some cases is a problem as free testing services do not offer capabilities to store and process such large amounts of data, and in-house testing solutions needed to be created to address this limitation.
-The solution for this has been to limit the size of the testing problems, and running tests in small batches of few iterations.
+The solution for this has been to limit the size of the testing problems, and to run tests in small batches with few iterations.
 
@@
 The developer of ESPResSo noted that their source code had been written with a specific application in mind and that due to this there was too much coupling between components of the source code.
-This results in technical debt, and would have an impact on future modifiability and reusability when trying to extend the software, and the code would need to be refactored.
+This results in technical debt, having an impact on future modifiability and reusability when trying to extend the software, and the code would need to be refactored.
 
@@
 The developer comments have emphasized an importance on source code, while documentation seems to be of secondary importance.
-It must be stressed that improved management of documentation could benefit development and help eliminate some of the developer concerns that were raised.
+It must be stressed that improving documentation could benefit development and help eliminate some of the developer concerns that were raised.
 The use of automatic document generation tools that capture scientific and computing knowledge, and transform it into software artifacts, is advised.
@@
 
-\subsection{Quality Recommendations - DONE 1st Draft}\label{qualityrecommentations}
+\subsection{Quality Recommendations}\label{qualityrecommentations}
 
@@
 	\item Include OS compatibility, including specific OS versions.
-	\item Include complete installation instructions.
+	\item Provide complete installation instructions.
 	\item Installation instructions should be written as if the user does not have any dependencies installed and is installing on a clean OS.
@@
 	\item Include descriptive error messages for errors encountered during installation. This was done by several top ranked packages, including Ludwig and LUMA
-	\item Include a way to validate the installation.
+	\item Provide a way to validate the installation.
 This can be done using a custom script, or a test case that specifies expected output.
 This was done by several top ranked packages, including OpenLB, Ludwig, LUMA, and Palabos.
-	\item Include instructions for uninstalling the software. These were included with pyLBM.
+	\item Provide instructions for uninstalling the software. These were included with pyLBM.
 \end{itemize}
@@
 \begin{itemize}
-	\item Use a requirements specification document. A template 
+	\item Use a requirements specification document. A template is presented in \citep{smith2005new}. 
 	\item Make public (on package website or GitHub) the requirements specification document, or explicitly reference the domain theory that the software is designed from.
@@
 	\item Use and make public (on package website or GitHub) detailed documentation. Consider using automatic document generation tools like Doxygen, Drasil, or Sphinx.
-	\item Include detailed tutorials that include expected output, like the \href{https://www.walberla.net/doxygen/index.html}{waLBerla tutorials}.
+	\item Provide detailed tutorials that include expected output, like the \href{https://www.walberla.net/doxygen/index.html}{waLBerla tutorials}.
 	\item Use unit tests during development and make them public (on package website or GitHub). This was available for Ludwig.
 	\item Modularize the source code, separate components, hide information  behind well defined interfaces.
-	\item Use continuous integration tools and processes (Bamboo, Jenkins, and Travis CI) during development. This was done by top ranked packages ESPResSo and Ludwig.
+	\item Use continuous integration tools (Bamboo, Jenkins, and Travis CI) and processes during development. This was done by top ranked packages ESPResSo and Ludwig.
 \end{itemize}
@@
 	\item The system must provide descriptive error messages when it encounters unexpected input, including incorrect data types, empty input, and missing files or links.
-Eighteen of the software packages behaved in such reasonable fashion when tested with unexpected input.
+Eighteen of the software packages behaved reasonably when tested with unexpected input.
 \end{itemize}
@@
 Only those offering GPU processing (ESPResSo, lbmpy, lettuce, pyLBM, Sailfish, TCLB) mentioned any hardware requirements.
-On the other hand, some sort of software requirements were available for all packages, even if only some dependencies or a compatible operating system.
-	\item Include a user tutorial that indicates the expected output. This was done by many packages, including waLBerla, Palabos, MechSys, LUMA, pyLBM.
+On the other hand, some sort of software requirements were available for all packages, even if only some dependencies or a compatible operating system were mentioned.
+	\item Provide a user tutorial that indicates the expected output. This was done by many packages, including waLBerla, Palabos, MechSys, LUMA, pyLBM.
 	\item Provide a detailed user manual.
@@
 Especially consider this for parts of the source code that is likely to be modified or reviewed by users.
-ESPResSo and Sailfish use Python to shorted development time and improve usability.
+ESPResSo and Sailfish use Python to shorten development time and improve usability.
 \end{itemize}
@@
 	\item Include version numbers and release notes for all major source code and artifact releases. The top five ranked software packages include release notes.
-	\item Have a defined process for accepting contributions. The top five ranked software packages include information on how to contribute.
-	\item Include documentation for making contributions to the project. 
+	\item Have a defined process for accepting contributions, and make public documentation for making contributions to the project.
+The top five ranked software packages include information on how to contribute.
 	\item Use an issue tracker (Git, email, SourceForge, other) to manage bugs and changes.
@@
 	\item Modularize the source code, separate components, hide information behind well defined interfaces.
-	\item Include user manuals that identify all technical and problem modeling requirements. Thirteen software packages have a user manual.
+	\item Provide a user manual that identifies all technical and problem modeling requirements. Thirteen software packages have a user manual.
 	\item State appropriate fluid dynamics problems that the software is designed to model in the documentation, and explicitly state the limits of the software. This is done in the ESPResSo user guide. 
@@
 	\item Use automatic document generation tools (Doxygen, Drasil, Sphinx) to limit the time spent on updating documentation.
-	\item Provide a resource, like a developer's guide, that helps orient developers to the artifacts, source code, and system architecture so that they can better document changes.  
+	\item Provide a developer's guide to help orient developers to the artifacts, source code, and system architecture so that they can better document changes.  
 \end{itemize}
@@
 	\item Provide a detailed user manual. Thirteen packages have a user manual. ESPResSo has a well detailed manual.
-	\item Provide a table of symbols in the developer guide that map to names used in the course code.
+	\item Provide a table of symbols in the developer's guide that maps to names used in the course code.
 	\item State appropriate fluid dynamics problems that the software is designed to model in the documentation, and explicitly state the limits of the software. This is done in the ESPResSo user guide.
@@
 
-answers last question of research questions:
-
-How does software designated as high quality by this methodology compare	with top rated software by the community?
+This comparison helps evaluate the methodology that was used, and assess the validity of the findings.
+Threats to the validity of this exercise are noted in Section \ref{threats}.
+The software package designations of this report are compared to package repository ranking metrics in Section \ref{repmetrics}, and to software recommended by a domain expert in Section \ref{domainexpertrecommend}.
+
+\subsubsection{Repository Ranking Metrics}\label{repmetrics}
+
+\begin{center}
+	\begin{onehalfspacing}
+		\begin{tabular}{ p{3.5cm}p{2.5cm}p{2.5cm}p{2.5cm} }
+			\hline
+			Name & Our Ranking & Repository Stars & Repository Watches \\
+			\hline
+			Ludwig&1& 19 & 6\\
+			ESPResSo &2& 115 & 23\\
+			Palabos &3& 14 & GitLab\\
+			OpenLB &4& No Git & No Git\\
+			LUMA&5& 26 & 10\\
+			pyLBM &6& 57& 8\\
+			DL\_MESO (LBE) & 7 & No Git & No Git \\
+			waLBerla & 8 & 17 & GitLab\\
+			Sailfish &9& 154 & 39\\
+			laboetie &10& 4 & 5\\
+			TCLB &11& 61 & 16\\
+			MechSys &12& No Git & No Git\\
+			lettuce &13& 8 & 2\\
+			ESPResSo++ &14& 31 & 13\\
+			MP-LABS &15& 7 & 2\\			
+			SunlightLB & 16& No Git & No Git\\
+			LB3D &17& No Git & No Git\\			
+			LIMBES &18& No Git & No Git\\
+			LB2D-Prime &19& No Git & No Git\\		
+			HemeLB &20& 22 & 15\\
+			lbmpy&21&  4 & 2  \\	
+			LB3D-Prime &22& No Git & No Git\\	
+			LatBo.jl &23& 4 & 5\\			
+			\hline
+		\end{tabular}
+		\captionof{table}{Repository Ranking Metrics}\label{repometrics}
+	\end{onehalfspacing}
+\end{center}
+
+Table \ref{repometrics} presents our LBM software package rankings along with the repository ranking metrics of each software package.
+
+Eight packages do not use GitHub or GitLab, so they do not have a measure of repository stars.
+Looking at the repository stars of the other 15 packages, we can observe a slight pattern where packages that have been highly ranked by our assessment do have more stars than lower ranked packages.
+The second ranked package (ESPResSo) has the second most number of stars.
+The ninth ranked package (Sailfish) has the most number of stars.
+Packages designated as lower quality often do not use GitHub or GitLab, or have few stars, except for HemeLB which has 22 stars.
+Our assessment of this package might not be accurate.
+
+The repository watches column contains even less data to compare since two of the packages (Palabos, waLBerla) use GitLab, which does not include this metric.
+The pattern that can be observed with this metric is very similar to that of the stars metric.
+The ninth ranked package (Sailfish) has the most number of watches.
+The second ranked package (ESPResSo) has the second most number of watches.
+
+Besides missing data, another threat to the validity of this comparison is the varying ages of the repositories.
+Older packages have been able to accumulate stars and watches for longer than newer packages.
+The true quality of new packages may not be reflected in their stars and watches.
+
+\subsubsection{Domain Expert Recommended Software}\label{domainexpertrecommend}
 
@@
 
-This section will point to potential threats to the validity in the measurement section of data collection, for example from above:
-
-Interviews with the developers of software packages also revealed a more frequent use of both unit testing and continuous integration in the development process than was found by only assessing the artifacts.
-The use of automatic installation processes is also common.
-Most often this involved a Make script.
+This section examines potential threats to the validity of this state of the practice assessment. These can be categorized into methodology and data collection issues.
+
+The measures listed in our measurement template may not be broad enough to accurately capture some qualities.
+For example, there are only two measures of surface robustness.
+The measurement of robustness could be expanded, as it currently only measures unexpected input.
+Other faults could be introduced, but could require a large investment of time to develop, and might not be a fair measure for all packages.
+Similarly, reusability is assessed along the number of code files and LOC per file.
+While this measure is indicative of modularity, it is possible that some packages have many files, with few LOC, but the files do not contain source code that is easily reusable.
+The files may be poorly formatted, or the source code may be vague and have ambiguous identifiers.
+Furthermore, the measurement of understandability relies on 10 random source code files.
+It is possible that the 10 files that were chosen to represent a software package may not be a good representation of the understandability of that package.
+
+Regarding data collection, a risk to the validity of this assessment is missing or incorrect data.
+Some software package data may not have been measured due to technology issues like broken links.
+This issue arose with the measurement of Palabos, which had a broken link to its user manual, as noted in Section \ref{lesscommon}.
+
+Some pertinent data may not have been specified in public artifacts, or may be obscure within an artifact or web-page.
+The use of unit testing and continuous integration was mentioned in the artifacts of only two (ESPResSo, Ludwig) packages.
+However, interviews suggested a more frequent use of both unit testing and continuous integration in the development processes than what was observed from the initial survey of the artifacts.
+For example, OpenLB, pyLBM, and TCLB use such methods during development despite this not being explicitly clear from an analysis of the material available online.
+
+Furthermore, design documentation was measured to be a ``less common'' artifact in this assessment, but it is probable that such documentation is part of all LBM packages.
+After all, developing SCS is not a trivial endeavor.
+It is likely that many packages have such documentation but did not make it public, and due to this the measured data is not a true reflection of software package quality.
 

### 2021-06-25 StateOfPractice/Ao-Notes/DomainExptMeeting/Presentation for the Introductory Meeting.tex
Updates to Medical Imaging Presentation

@@
 \bibliographystyle{plainnat}
+
+\usepackage{tikz}
+\usetikzlibrary{mindmap}
+%tikstyle commands
+% Define block styles
+\tikzstyle{context} = [rectangle, rounded corners, text width=2.5cm, minimum
+size=2.5cm,text centered, draw=black, fill=white!30]
+\tikzstyle{strategy} = [trapezium, trapezium left angle=70, trapezium right
+angle=110, text width=3cm, minimum height=1cm, text centered, draw=black,
+fill=white!30]
+\tikzstyle{subclaim} = [rectangle, text width=3cm, minimum height=1cm, text
+centered, draw=black, fill=white!30]
+\tikzstyle{goal} = [rectangle, text width=5cm, minimum height=1cm, text
+centered, draw=black, fill=white!30]
+\tikzstyle{assumption} = [ellipse, text width=3cm, minimum height=1cm, text
+centered, draw=black, fill=white!30]
+\tikzstyle{evidence} = [circle, text width=2cm, minimum size=2cm, text
+centered,draw=black, fill=white!30]
+\tikzstyle{arrow} = [thick,->,>=stealth]
+
+\newcommand{\famname}{MIA}
+\newcommand{\progname}{MISEG}
 
@@
 
-\newcommand{\topic}{State of the Practice for Lattice Boltzmann Solvers}
+\newcommand{\topic}{State of the Practice for Medical Imaging Software}
 
@@
   \item Understand the state of the software development practices for
-    LBM Software
+    medical imaging Software
   \item Make recommendations for improvements
@@
 \ei
+
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{frame}[plain, fragile]
+\frametitle{Scope}
+
+\scalebox{0.64}
+{
+\begin{tikzpicture}[mindmap, grow cyclic, every node/.style=concept, concept
+color=orange!40,
+	level 1/.append style={level distance=5cm,sibling angle=90},
+	level 2/.append style={level distance=3cm,sibling angle=90},
+	level 3/.style={level distance=3cm,sibling angle=60}]
+\node{\famname}
+	child[concept color=teal!40] { node {Enhancement}}
+	child[concept color=teal!40] { node {Analysis}
+    	child[concept color=blue!30] { node {Registration}}
+	    child[concept color=purple!50,] { node {Segmentation}}
+	    child[concept color=blue!30] { node {Statistical Analysis}
+	        child { node {Feature Extraction}}
+	        child { node {Classification}}
+	        child { node {Interpretation}}
+	}}
+	child[concept color=teal!40] { node {Simulation/\\Modeling}}
+    child[concept color=teal!40] { node {Visualization}
+	    child[concept color=blue!30] { node {2D Display}}
+	    child[concept color=blue!30] { node {3D Rendering}}
+	    child[concept color=blue!30] { node {Reformatted Views}}
+    	}
+;
+\end{tikzpicture}
+}
 
@@
 \begin{frame}
-\frametitle{Scope}
+\frametitle{Vet Software List}
 
-[Ao put your scope figure here.  Peter, you can delete this slide.]
+\bi
+\item \structure{How does our list compare to the domain expert's list?}
+\item \structure{Is any software missing?}
+\item \structure{Is there software that should be included?}
+\item \structure{Any other questions/comments or concerns?}
+\ei
+
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{frame}[plain]
+%\frametitle{SOP Software List}
+
+\begin{multicols}{2}	
+	\bi
+      \item 3D Slicer
+      \item  Ginkgo CADx
+      \item XMedCon
+      \item Weasis
+      \item MRIcroGL
+      \item SMILI
+      \item ImageJ
+      \item Fiji
+      \item DicomBrowser
+        \item 3DimViewer
+        \item Horos
+        \item OsiriX Lite
+        \item dwv
+        \item Drishti
+        \item BioImage Suite Web
+        \item OHIF Viewer
+          \item Slice:Drop
+          \item GATE
+          \item ITK-SNAP
+          \item ParaView
+          \item MatrixUser
+          \item DICOM Viewer
+          \item INVESALIUS 3
+            \item medInria
+            \item dicompyler
+            \item MicroView
+            \item Papaya
+            \item AMIDE
+            \item Gwyddion \ei
+\end{multicols}
 
@@
 \begin{frame}
-\frametitle{Vet Software List}
+\frametitle{Common With Domain Expert List}
 
+\begin{multicols}{2}	
   \bi
-\item How does our list compare to the domain expert's list?
-\item Is any software missing?
-\item Is there software that should be included?
-\item Any other questions/comments or concerns?  
+\item 
 \ei
+\end{multicols}
 
@@
 \begin{frame}
-\frametitle{Software List}
+\frametitle{Only on Domain Expert List}
 
-[Peter and Ao, please put your list here, with hyperlinks if possible.  You can use
-columns to make it all fit on one slide.]
+%\begin{multicols}{2}	
+  \bi
+\item 
+	\ei
+%\end{multicols}
 
@@
 \begin{frame}
-\frametitle{Domain Expert Software List}
+\frametitle{Only on SOP List}
+
+\begin{multicols}{2}	
+  \bi
+			\item 
+	\ei
+\end{multicols}
 
@@
 
-[Peter and Ao, give the overall ranking graphic for your domain]
+\bigskip
+
+    \includegraphics[width=1.0\textwidth]{overall_scores.png}
 
@@
 \begin{frame}
-\frametitle{Discuss Recommendations}
+\frametitle{Pain Points from Developer Interviews}
 
+\structure{Do these fit with your experience?}
 \bi
-\item 
+\item Lack of time to implement new features
+\item Team members lack software development experience
+\item Lack of funding for software development
+\item No organizations to help with developing high quality software
+\item Difficulty to get test data
+  \ei
+
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{frame}
+\frametitle{Recommendations}
+
+\structure{Do these seem feasible?  What other ideas do we have?}
+\bi
+\item Consult with software development organization
+  \bi
+\item Better Scientific Software (BSSw)
+\item Software Sustainability Institute
+\item Software Carpentry
+  \ei
+\item Citations for software (Katz project)
+  \item Redefine productivity to include time working on tasks like testing, continuous
+    integration and documentation
 \ei
@@
 \bi
-\item Where should we publish this paper?
-\item Who do you see as the targeted readers?
+\item \structure{Who do you see as the targeted readers?}
+\item \structure{Where should we publish this paper?}
 \ei
@@
 \begin{frame}
-\frametitle{Developer Questions (for next time)}
+\frametitle{Developer Questions}
 
 \bi
-\item 
-\ei
+\item \structure{What experience do you, or your students, have with developing
+    software?}
+\item \structure{How big would the development group be?}
+\item \structure{What is the typical background of a developer?}
+\item \structure{How many users would you software typically have?}
+\item \structure{What is the typical background of a user}
+\item \structure{Currently, what are the most significant obstacles in your
+    development process?}
+\item \structure{How might you change your development process to remove or
+    reduce these obstacles?}
+\item \structure{How does documentation fit into your development process? Would improved
+  docu- mentation help with the obstacles you typically face?}      
+\item \structure{In the past, is there any major obstacle to your development process that
+  has been solved? How did you solve it?}
+\item \structure{What is your software development model? For example, waterfall, agile,
+  etc.}
+\item \structure{What is your project management process? Do you think improving this
+  process can tackle the current problem? Were any project management tools
+  used?}  \ei
 

### 2021-06-25 StateOfPractice/Ao-Notes/DomainExptMeeting/Presentation for the Introductory Meeting.tex
updated software lists

@@
 
-\begin{multicols}{2}	
 \bi
-\item 
+\item 3DSlicer
+\item Horos
+\item ImageJ
+\item Fiji (a ``batteries-included" distribution of ImageJ)
+\item Mango (on the initial list, but not the final list)
+\item Papaya (the web version of Mango)
+\item MRIcroGL (MRIcron development has moved to MRIcroGL)
 \ei
-\end{multicols}
 
@@
 
-%\begin{multicols}{2}	
 \bi
-\item 
+\item AFNI \hyperlink{https://afni.nimh.nih.gov}{https://afni.nimh.nih.gov}
+\item FSL \hyperlink{https://fsl.fmrib.ox.ac.uk/fsl/fslwiki}{https://fsl.fmrib.ox.ac.uk/fsl/fslwiki} 
+\item Freesurfer \hyperlink{https://surfer.nmr.mgh.harvard.edu}{https://surfer.nmr.mgh.harvard.edu} 
+\item Tarquin \hyperlink{http://tarquin.sourceforge.net}{http://tarquin.sourceforge.net}  (for MRSpectroscopy data)
+\item Diffusion Toolkit  \hyperlink{http://trackvis.org/dtk/}{http://trackvis.org/dtk/} (diffusion toolkit and trakVis)
+\item MRItrix \hyperlink{https://www.mrtrix.org}{https://www.mrtrix.org}
 \ei
-%\end{multicols}
 
@@
 \bi
-			\item 
+\item Ginkgo CADx
+\item XMedCon
+\item Weasis
+\item SMILI
+\item DicomBrowser
+\item 3DimViewer
+\item OsiriX Lite
+\item dwv
+\item Drishti
+\item BioImage Suite Web
+\item OHIF Viewer
+\item Slice:Drop
+\item GATE
+\item ITK-SNAP
+\item ParaView
+\item MatrixUser
+\item DICOM Viewer
+\item INVESALIUS 3
+\item medInria
+\item dicompyler
+\item MicroView
+\item AMIDE
+\item Gwyddion
 \ei

### 2021-06-25 StateOfPractice/Ao-Notes/DomainExptMeeting/Presentation for the Introductory Meeting.tex
updated color for segmentation

@@
     	child[concept color=blue!30] { node {Registration}}
-	    child[concept color=purple!50,] { node {Segmentation}}
+	    child[concept color=blue!30,] { node {Segmentation}}
 	    child[concept color=blue!30] { node {Statistical Analysis}

### 2021-06-25 StateOfPractice/Ao-Notes/DomainExptMeeting/Presentation for the Introductory Meeting.tex
updated painpoints

@@
 \bi
-\item Lack of time to implement new features
-\item Team members lack software development experience
-\item Lack of funding for software development
+\item Lack of time to 1) implement all requested new features 2) write and maintain good documents 3) review all contributions from the community
+\item It is difficult to find or keep the members with experience in both software development and medical imaging
+\item Lack of funding for software development (especially for maintenance)
 \item No organizations to help with developing high quality software
 \item Difficulty to get test data
+\item Difficulty to achieve cross-platform compatibility: 1) native app - more time and work 2) web app w/o server - worse performance 3) web app w/ servers - more cost
 \ei

### 2021-06-25 StateOfPractice/Ao-Notes/DomainExptMeeting/Presentation for the Introductory Meeting.tex
updated qualities
added all interview questions to the slides

@@
 \item Maintainability
-\item Surface Understandability
-  \item Visibility and transparency
+\item Reusability
+\item Surface understandability
+\item Visibility/transparency
 \ei
@@
     reduce these obstacles?}      
+\ei
+
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{frame}
+\frametitle{Developer Questions}
+
+\bi
 \item \structure{How does documentation fit into your development process? Would improved
@@
 process can tackle the current problem? Were any project management tools
-  used?}  \ei
+used?}
+\ei
+
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{frame}
+\frametitle{Developer Questions}
+
+\bi
+\item \structure{Was it hard to ensure the correctness of the software? If there were any obstacles, what methods
+have been considered or practiced to improve the situation? If practiced, did it work?}
+\item \structure{When designing the software, did you consider the ease of future changes? For example, will it be
+hard to change the structure of the system, modules or code blocks? What measures have been taken
+to ensure the ease of future changes and maintains?}
+\item \structure{Provide instances where users have misunderstood the software. What, if any, actions were taken
+to address understandability issues?}
+\item \structure{What, if any, actions were taken to address usability issues?}
+\ei
+
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{frame}
+\frametitle{Developer Questions}
+
+\bi
+\item \structure{Do you think the current documentation can clearly convey all necessary knowledge to the users?
+If yes, how did you successfully achieve it? If no, what improvements are needed?}
+\item \structure{Do you have any concern that your computational results won’t be reproducible in the future?
+Have you taken any steps to ensure reproducibility?}
+\ei
 

### 2021-06-25 StateOfPractice/Peter-Notes/DomainExptMeeting/Presentation for the Introductory Meeting.tex
Updates to LBM Domain Expert Meeting Slides

@@
   \bi
-  \item Understand the state of the software development practices for
-    LBM Software
+  \item Understand state of software development practice
   \item Make recommendations for improvements
@@
   \ei  
-\item We have developed a standard methdology for assessing SOP for Domain X
+\item We have developed a standard methdology% for assessing SOP for Domain X
 \item The methodology requires a \textbf{domain expert} to:
@@
 \item Informal
-\item Questions do not have to be answered in real time, or by domain expert
+\item Questions do not have to be answered in real time%, or by domain expert
   \ei
@@
 \item \emph{Domain Expert}: Create a top ten list
-\item Brief \emph{Domain Expert}
+%\item Brief \emph{Domain Expert}
 \item Initial list of candidate software packages
 \item \emph{Domain Expert}: Vet domain software list
-\item Domain Analysis
-\item \emph{Domain Expert}: Vet domain analysis
+\item Domain Analysis, Software Features
+\item \emph{Domain Expert}: Vet domain analysis / features
 %\item Gather source code and documentation for software
-\item Collect empirical measures
+\item Collect empirical measures (stars, forks, lines of code, etc.)
 \item Measure using measurement template
-\item Interview developers
+\item Interview developers (pain points)
 \item Use AHP process to rank the software packages
 \item \emph{Domain Expert}: Vet AHP ranking
-\item \emph{Domain Expert}: Review recommendations
+\item \emph{Domain Expert}: Review recommendations %$\Leftarrow$
 \end{enumerate}
@@
 \begin{frame}
-\frametitle{Common With Domain Expert List}
+\frametitle{Domain Expert List (Common in Blue)}
 
@@
 \item \href{https://docs.tclb.io/}{TCLB} (also called CudneLB)
+  \item ASL 
+\item ch4-project
+\item Open FSI
+\item LIFE
+\item LBSIM
+\item Taxila LBM
+
 \ei
@@
 
-\begin{frame}
-\frametitle{Domain Analysis}
+% \begin{frame}
+% \frametitle{Domain Analysis}
 
-\bi
-\item So users can pick the right software for their task
-\item Think of the software as a family of related software products, like:
-\bi
-\item Family of automobiles
-\item Computer hardware
-\item etc.
-\ei
-\item Commonalities
-\item Variabilities
-\item Parameters of variation
-\item \structure{Do the following slides capture the commonalities and variabilities?}
-\item We can follow up over e-mail, if that is easier
-\ei
+% \bi
+% \item So users can pick the right software for their task
+% \item Think of the software as a family of related software products, like:
+% \bi
+% \item Family of automobiles
+% \item Computer hardware
+% \item etc.
+% \ei
+% \item Commonalities
+% \item Variabilities
+% \item Parameters of variation
+% \item \structure{Do the following slides capture the commonalities and variabilities?}
+% \item We can follow up over e-mail, if that is easier
+% \ei
 
-\end{frame}
+% \end{frame}
 
@@
 \begin{frame}
-\frametitle{Commonalities}
+\frametitle{SOP Common Features}
 
@@
 \begin{frame}
-\frametitle{SOP Variabilities}
+\frametitle{SOP Features}
 
@@
 \begin{frame}
-\frametitle{Domain Expert Variabilities}
+\frametitle{Domain Expert Features}
 
@@
 \begin{frame}
-\frametitle{Summary of Measures}
+\frametitle{Potential Next Steps}
 
 \bi
-\item Empirical measures (number of stars, forks, pull requests, lines of code, etc)
-\item Measurement template
+\item \structure{Can we work together on completing the following for the
+    paper?}
   \bi
+\item Domain analysis
+\item Combined list of features
+\item Matrix of software related to features \ei \ei
+
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{frame}
+\frametitle{Measurements Made}
+
+\be
 \item Installability
@@
 \item Maintainability
-\item Surface Understandability
+\item Reusability
+\item Surface understandability
 \item Visibility and transparency
-  \ei
-\item Developer interviews (looking for pain points)
-\item AHP Process with pairwise comparisons
-  \ei
-\end{frame}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{frame}
-\frametitle{Thoughts on Overall Ranking?}
-
-\bigskip
-
-    \includegraphics[width=1.0\textwidth]{finalscore_chart.png}
+ \ee
 
@@
 \begin{frame}
-\frametitle{Ranking Follow-Up}
+\frametitle{Maintainability Measurement}
 
-\bi
-\item \structure{We would like feedback on the ranking for each of the qualities}
-\item \structure{Is it feasible to review all 10 graphs, and the associated write-up?}
-\item \structure{Maybe there is a grad student that can review the rankings?}
-\ei
+\bigskip
+
+    \includegraphics[width=1.0\textwidth]{maintainability_chart.png}
+
+    %\bigskip
+
+    \structure{Does anything stand out?}
 
@@
 \begin{frame}
-\frametitle{Pain Points from Developer Interviews}
+\frametitle{Domain Expert: Ranking Follow-Up}
 
-\structure{Do these fit with your experience?}
 \bi
-\item Lack of time to implement new features
+\item \structure{Can you review the full ranking measurements?}
+\bi
+\item There are 9 measurements
+\item The write up is about 10 pages long \ei \ei
+
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{frame}
+\frametitle{From Developer Interviews: Pain Points}
+
+\begin{enumerate}
+\item Lack of time to implement new features \label{LackOfTime}
 \item Team members lack software development experience
 \item Lack of funding for software development
-\item No organizations to help with developing high quality software
+\item Lack of credit for developing software \label{LackOfCredit}
+\item No organizations to help with developing high quality software \label{NoOrg}
 \item Documentation could be improved, only has shallow overview of the theory
@@
 \item Technical debt due to coupling between components
-  \ei
+\end{enumerate}
 
@@
 \bi
-\item Consult with software development organization
+\item Consult with software development organization (P\ref{NoOrg})
   \bi
@@
   \ei
-\item Citations for software (Katz project)
+\item Citations for software (Katz project) (P\ref{LackOfCredit})
   \item Redefine productivity to include time working on tasks like testing, continuous
@@
 
+\begin{frame}
+\frametitle{Developer Questions}
+
+\bi
+\item \structure{What experience do you, or your students, have with developing
+    software?}
+\item \structure{How big would the development group be?}
+\item \structure{What is the typical background of a developer?}
+\item \structure{How many users would you software typically have?}
+\item \structure{What is the typical background of a user?}
+\item \structure{Currently, what are the most significant obstacles in your
+    development process?}
+\item \structure{How might you change your development process to remove or
+    reduce these obstacles?}
+\item \structure{How does documentation fit into your development process? Would improved
+  documentation help with the obstacles you typically face?}      
+\ei
+
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{frame}
+\frametitle{Developer Questions Continued}
+
+\bi
+\item \structure{In the past, is there any major obstacle to your development process that
+  has been solved? How did you solve it?}
+\item \structure{What is your software development model? For example, waterfall, agile,
+  etc.}
+\item \structure{What is your project management process? Do you think improving this
+  process can tackle the current problem? Were any project management tools
+  used?}
+\item \structure{Was it hard to ensure the correctness of the software? If there
+    were any obstacles, what methods have been considered or practiced to
+    improve the situation? If practiced, did it work?}
+\ei
+
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{frame}
+\frametitle{Developer Questions Continued}
+
+\bi
+\item \structure{When designing the software, did you consider the ease of
+    future changes? For example, will it be hard to change the structure of
+    the system, modules or code blocks? What measures have been taken to ensure
+    the ease of future changes and maintainance?}
+\item \structure{Provide instances where users have misunderstood the
+    software. What, if any, actions were taken to address understandability
+    issues?}
+\item \structure{What, if any, actions were taken to address usability issues?}
+\item \structure{Do you think the current documentation can clearly convey all
+    necessary knowledge to the users? If yes, how did you successfully achieve
+    it? If no, what improvements are needed?}
+\ei
+
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{frame}
+\frametitle{Developer Questions Continued}
+
+\bi
+\item \structure{Do you have any concern that your computational results won't
+    be reproducible in the future? Have you taken any steps to ensure
+    reproducibility?}  \ei
+
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \end{document}

### 2021-06-25 StateOfPractice/Ao-Notes/DomainExptMeeting/Presentation for the Introductory Meeting.tex
Update to medical image domain expert meeting

@@
 \item \structure{How many users would you software typically have?}
-\item \structure{What is the typical background of a user}
+\item \structure{What is the typical background of a user?}
 \item \structure{Currently, what are the most significant obstacles in your
@@
     reduce these obstacles?}
+\item \structure{How does documentation fit into your development process? Would improved
+  documentation help with the obstacles you typically face?}      
 \ei
@@
 \begin{frame}
-\frametitle{Developer Questions}
+\frametitle{Developer Questions Continued}
 
 \bi
-\item \structure{How does documentation fit into your development process? Would improved
-docu- mentation help with the obstacles you typically face?}
 \item \structure{In the past, is there any major obstacle to your development process that
@@
   used?}
+\item \structure{Was it hard to ensure the correctness of the software? If there
+    were any obstacles, what methods have been considered or practiced to
+    improve the situation? If practiced, did it work?}
 \ei
@@
 \begin{frame}
-\frametitle{Developer Questions}
+\frametitle{Developer Questions Continued}
 
 \bi
-\item \structure{Was it hard to ensure the correctness of the software? If there were any obstacles, what methods
-have been considered or practiced to improve the situation? If practiced, did it work?}
-\item \structure{When designing the software, did you consider the ease of future changes? For example, will it be
-hard to change the structure of the system, modules or code blocks? What measures have been taken
-to ensure the ease of future changes and maintains?}
-\item \structure{Provide instances where users have misunderstood the software. What, if any, actions were taken
-to address understandability issues?}
+\item \structure{When designing the software, did you consider the ease of
+    future changes? For example, will it be hard to change the structure of
+    the system, modules or code blocks? What measures have been taken to ensure
+    the ease of future changes and maintainance?}
+\item \structure{Provide instances where users have misunderstood the
+    software. What, if any, actions were taken to address understandability
+    issues?}
 \item \structure{What, if any, actions were taken to address usability issues?}
+\item \structure{Do you think the current documentation can clearly convey all
+    necessary knowledge to the users? If yes, how did you successfully achieve
+    it? If no, what improvements are needed?}
 \ei
@@
 \begin{frame}
-\frametitle{Developer Questions}
+\frametitle{Developer Questions Continued}
 
 \bi
-\item \structure{Do you think the current documentation can clearly convey all necessary knowledge to the users?
-If yes, how did you successfully achieve it? If no, what improvements are needed?}
-\item \structure{Do you have any concern that your computational results won’t be reproducible in the future?
-Have you taken any steps to ensure reproducibility?}
-\ei
+\item \structure{Do you have any concern that your computational results won't
+    be reproducible in the future? Have you taken any steps to ensure
+    reproducibility?}  \ei
 

### 2021-06-25 StateOfPractice/Ao-Notes/DomainExptMeeting/Presentation for the Introductory Meeting.tex
Update to Overview slide

@@
   \bi
-  \item Understand the state of the software development practices for
-    medical imaging Software
+  \item Understand state of software development practice
   \item Make recommendations for improvements

### 2021-06-25 StateOfPractice/Ao-Notes/DomainExptMeeting/Presentation for the Introductory Meeting.tex
Update Overall Process slide
Modify Measurements Made slide
Domain analysis and features, change the maintainability, rather than full ranking

@@
 \begin{frame}
-\frametitle{Summary of Measures}
+\frametitle{Domain Analysis and Features}
 
 \bi
-\item Empirical measures (number of stars, forks, pull requests, lines of code, etc)
-\item Measurement template
+\item \structure{Can we work together on doing the following for the
+    paper?}
   \bi
+\item Domain analysis
+\item List of features
+\item Matrix of software related to features \ei \ei
+
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{frame}
+\frametitle{Measurements Made}
+
+\be
 \item Installability
@@
 \item Surface understandability
-\item Visibility/transparency
-\ei
-\item Developer interviews (looking for pain points)
-\item AHP Process with pairwise comparisons
-  \ei
-\end{frame}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{frame}
-\frametitle{Thoughts on Overall Ranking?}
-
-\bigskip
-
-    \includegraphics[width=1.0\textwidth]{overall_scores.png}
+\item Visibility and transparency
+ \ee
 
@@
 \begin{frame}
-\frametitle{Ranking Follow-Up}
+\frametitle{Maintainability Measurement}
+
+\bigskip
+
+\includegraphics[width=1.0\textwidth]{maintainability_scores.png}
+
+    \structure{Does anything stand out?}
+
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{frame}
+\frametitle{Domain Expert: Ranking Follow-Up}
 
 \bi
-\item \structure{We would like feedback on the ranking for each of the qualities}
-\item \structure{Is it feasible to review all 10 graphs, and the associated write-up?}
-\item \structure{Maybe there is a grad student that can review the rankings?}
-\ei
+\item \structure{Can you review the full ranking measurements?}
+\bi
+\item There are 9 measurements
+\item The write up is about 10 pages long \ei \ei
 

### 2021-06-25 StateOfPractice/Peter-Notes/Report/Report.tex
updated chapter 5 LBM report

@@
 
+Table \ref{rankingcomparison} compares the top 10 ranked LBM software packages with a LBM package ranking made by a domain expert.
+Five (Palabos, OpenLB, LUMA, waLBerla, Sailfish) of the top 10 ranked packages in this assessment are also found in the domain expert's list.
+Interestingly, these five packages are listed in the same order on both lists.
+Looking at the remaining packages, the top two ranked packages in this state of the practice assessment (Ludwig, ESPResSo) are not on the domain expert's list.
+Perhaps the intended applications of these packages, complex fluids for Ludwig and soft matter research for ESPResSo, do not align with the research interests of the domain expert, and this is why they did not make it onto their list.
+Moving down the list, pyLBM and DL\_MESO(LBE) did not make the domain expert's top 10 list but were mentioned by the domain expert.
+The 10th ranked package, laboetie, was not mentioned by the domain expert.
+Looking at the remainder of the domain expert's list, ASL is a general purpose tool for solving partial differential equations and may have fallen out of scope of the authoritative lists that were used to identify the initial LBM software list.
+The domain expert's fifth ranked package (ch4-project) was on the initial software list of this assessment, but was removed due to a lack of documentation.
+Open FSI is a new project that uses Palabos.
+It was made public the year before data was collected for this state of the practice exercise.
+It may not have been on authoritative lists due to its age.
+Similarly, LIFE was also made public recently.
+Finally, the domain expert's 10th ranked package (HemeLB) was ranked 20th in this state of the practice assessment.
+
+\begin{center}
+	\begin{onehalfspacing}
+		\begin{tabular}{ p{2cm}p{3.5cm}p{3.5cm}}
+			\hline
+			Rank & This Assessment & Domain Expert\\
+			\hline
+			1&Ludwig&Palabos\\
+			2&ESPResSo &OpenLB\\
+			3&Palabos &LUMA\\
+			4&OpenLB &ASL\\
+			5&LUMA&ch4-project\\
+			6&pyLBM &Open FSI\\
+			7&DL\_MESO (LBE) & WaLBerla\\
+			8&waLBerla & LIFE\\
+			9&Sailfish &Sailfish\\
+			10&laboetie &HemeLB\\		
+			\hline
+		\end{tabular}
+		\captionof{table}{Ranking Comparison}\label{rankingcomparison}
+	\end{onehalfspacing}
+\end{center}
+
 \subsection{Threats To Validity}\label{threats}

### 2021-06-26 StateOfPractice/Peter-Notes/Report/Report.tex
added Conclusion and Appendix to LBM report

@@
 
-This project analyzes the state of the practice for software development in the Lattice Boltzmann Method domain by quantitatively and qualitatively measuring 23 software packages along quality attributes...
+We analyze the state of the practice of software development of Lattice Boltzmann Methods software domain by quantitatively and qualitatively measuring, and comparing, 23 software packages along quality attributes.
+A methodology for assessing the state of the practice of software development in scientific computing software domains is presented.
+A domain analysis of the Lattice Boltzmann Methods software family is made, and a candidate software package list is identified and curated.
+The packages are assessed to answer software development related research questions to understand how software quality is impacted by software development choices, including principles, processes, and tools.
+Software developers are interviewed to identify development pain points, and to identify how software quality is ensured.
+Quantitative data is used to rank the software packages using an analytical hierarchy process.
+The ranking designations are compared with rankings from the software development community.
+Recommendations for improving software along quality metrics are made.
+This knowledge can be used to guide future development of scientific computing software, specifically along quality attributes, and to reduce software quality failures.\\
+
+\noindent\textbf{Keywords}: Lattice Boltzmann Methods, Scientific Computing, Software Engineering, Software Family, Software Quality
 
@@
 The purpose of state of the practice assessments is to understand how software quality is impacted by software development choices, including principles, processes, and tools, within SCS communities.
-This knowledge can be used to guide future development of SCS, specifically along quality attributes, and reduce software quality failures.
+This knowledge can be used to guide future development of SCS, specifically along quality attributes, and to reduce software quality failures.
 This work reports on the state of the practice of LBM software development and makes suggestions on improving the quality of software in this domain.
@@
 The use of automatic document generation tools that capture scientific and computing knowledge, and transform it into software artifacts, is advised.
-Drasil is an automatic document generation tool that is further discussed in Section \ref{conclusion}.
+Drasil is an automatic document generation tool that is further discussed in Section \ref{highlightedrecommendations}.
 
@@
 
-- key findings restated
-- comment on future work in state of the practice SCS
-
-the following was moved from the former background section, may need edits as part of conclusion:
-
-  The software qualities that are assessed in this report are listed in Section \ref{softwarequalities}.
-Ensuring software quality can take significant time, but as \cite{SmithEtAl2016} shows, SCS developers often cannot see themselves meeting the required time constraints to better develop some of the artifacts that improve qualities, like detailed documentation.
-An inability to develop such documentation due to time constraints could be mitigated by the use of automatic document generation tools like Drasil.
-The Drasil Framework consists of a collection of Domain Specific Languages (DSL) for capturing scientific documents, structures, and computing
-knowledge, and then transforming this knowledge into relevant software artifacts without having to manually duplicate knowledge into multiple artifacts \citep{zhao2018}.
-
-The following was moved from Tools Used subsection (consider adding future work subsection within conclusion, also consider subsection for ideas for future SOP such as asking the developers):
-
-Drasil could be considered as an alternative to generate artifacts since it is geared specifically towards research software domains like LBM software.
-
-future sop idea (from Results section):
-
-Developers were not asked to com-
-ment on the reproducibility of their source code from their requirements specifications and
-design documentation. This question should be considered in the next iteration of state of
-the practice assessments.
-
-On page 86 of the ReportChapt5Feedback pdf SS commented that some of the adivuce from the quality recommendations section would be difficult to do for less mature projects.
-"We should add some getting started advice for the biggest ROI changes".
-In this section I should add a subsection for recommendations, I should highlight the quality recommendations that are pervasive for multiple qualities, like modularization and using software languages that are considered user friendly (Python), etc.
+We analyzed the state of the practice of software development of Lattice Boltzmann Methods software domain by quantitatively and qualitatively measuring, and comparing, 23 software packages along quality attributes.
+The software qualities that were assessed in this report are listed in Section \ref{softwarequalities}.
+A methodology for assessing the state of the practice of software development in scientific computing software domains was presented.
+Domain packages were assessed to answer the software development related research questions listed in Section \ref{purpose} to understand how software quality is impacted by software development choices, including principles, processes, and tools.
+Software developers were interviewed to identify development pain points, and to identify how software quality is ensured.
+Quantitative data was used to rank the software packages using an analytical hierarchy process.
+The ranking designations were compared with rankings from the software development community, and we found that many of our top 10 ranked packages are ranked highly by a domain expert.
+Recommendations for improving software along quality metrics were made, and highlights are presented in Section \ref{highlightedrecommendations} of this conclusion.
+The findings of this report can be used to guide future development of scientific computing software, specifically along quality attributes, and to reduce software quality failures.
+
+We understand that software packages vary in goals, developers, funding, and other aspects.
+Our goal was to highlight software quality successes through an evaluation of the entire domain software family.
+Furthermore, threats to the validity of the findings are highlighted in Section \ref{threats}.
+
+Recommendations for future state of the practice assessments are made in Section \ref{futuresop}.
+
+\subsection{Highlighted Recommendations}\label{highlightedrecommendations}
+
+The following recommendations improve software quality along multiple attributes, and provide the greatest return on investment:
+
+\begin{itemize}
+	\item Provide a detailed user manual. It should identify elements of user interfaces, and identify all requirements to model a system.
+	\item State appropriate fluid dynamics problems that the software is designed to model in the documentation, and explicitly state the limits of the software. 
+	\item Include detailed tutorials, including dependencies, expected
+	output, and any additional supplementary documentation that may be required.
+	\item Keep all documentation in one location.
+	\item If possible, consider using popular user-friendly software languages like Python. Especially consider this for parts of the source code that is likely to be modified or reviewed by users. 
+	\item Modularize the source code, separate components, hide information behind well defined interfaces.
+	\item Include descriptive error messages where appropriate.
+	\item Summarize the development process that is used. Provide information on how new users can contribute. 
+	\item Consider peer review processes and task based inspection to assess contributions.	
+	\item Use continuous integration tools (Bamboo, Jenkins, and Travis CI) and processes during development. 
+	\item Use project management tools, including change and version control tools (GitHub, GitLab, CVS), collaboration tools (GitHub, GitLab), and document generation tools (Doxygen, Drasil, Sphinx).
+\end{itemize}
+
+Ensuring software quality can take significant time.
+An inability to develop high quality documentation due to time constraints could be mitigated by the use of automatic document generation tools like Drasil.
+The Drasil Framework consists of a collection of Domain Specific Languages (DSL) for capturing scientific documents, structures, and computing knowledge, and then transforming this knowledge into relevant software artifacts without having to manually duplicate knowledge into multiple artifacts \citep{zhao2018}.
+
+\subsection{Future State Of The Practice Assessments}\label{futuresop}
+
+This section notes recommendations for future state of the practice assessments.
+
+As mentioned in Section \ref{threats}, the measures listed in our measurement template may not be broad enough to accurately capture some qualities. Adding or extending measures is worth considering.
+
+Section \ref{interviewresults} noted that developers were not asked to comment on the reproducibility of their source code from their requirements specifications and design documentation.
+Adding this question to the interview guide should be considered in the next iteration of state of the practice assessments.
+Furthermore, as mentioned in Section \ref{threats}, it was found that some pertinent information was not specified in public artifacts.
+The use of unit testing and continuous integration by several packages (OpenLB, pyLBM, TCLB) was only discovered during interviews with developers.
+Adding further questions to the interview guide regarding the measures that are on the measurement template could reduce instances of incorrect data being collected.
+This additional interview data could be analyzed and incorporated into the AHP ranking, ensuring quality designations more accurate represent the true quality of the software packages.
 

### 2021-07-05 StateOfPractice/Peter-Notes/Report/Report.tex
incorporated full report feedback from SS into LBM report

@@
 \usepackage{indentfirst}
+\usepackage[toc,page]{appendix}
 
@@
 
-We analyze the state of the practice of software development of Lattice Boltzmann Methods software domain by quantitatively and qualitatively measuring, and comparing, 23 software packages along quality attributes.
-A methodology for assessing the state of the practice of software development in scientific computing software domains is presented.
-A domain analysis of the Lattice Boltzmann Methods software family is made, and a candidate software package list is identified and curated.
+We analyze the state of the practice of software development of Lattice Boltzmann Methods software domain by quantitatively and qualitatively measuring, and comparing, 23 software packages.
+A general methodology for assessing the state of the practice of software development in scientific computing software domains is presented.
+A domain analysis of the Lattice Boltzmann Methods software family is made, and candidate software packages are identified and curated.
 The packages are assessed to answer software development related research questions to understand how software quality is impacted by software development choices, including principles, processes, and tools.
 Software developers are interviewed to identify development pain points, and to identify how software quality is ensured.
-Quantitative data is used to rank the software packages using an analytical hierarchy process.
+Quantitative data is used to rank the software packages using the Analytical Hierarchy Process.
+The packages Ludwig, ESPResSo, and Palabos are ranked as the top three packages.
+Each of them scored high in at least several of the individual qualities that were quantitatively measured.
 The ranking designations are compared with rankings from the software development community.
 Recommendations for improving software along quality metrics are made.
-This knowledge can be used to guide future development of scientific computing software, specifically along quality attributes, and to reduce software quality failures.\\
+These include providing a detailed user manual and tutorials, explicitly stating the limits of the software, using user-friendly software languages, considering a peer review process, communicating development and contribution information, and using continuous integration and project management tools.
+This knowledge can be used to guide future development of scientific computing software along quality attributes to reduce software quality failures.\\
 
@@
 
-We analyze the development of software packages that use Lattice Boltzmann Methods (LBM), a class of computational fluid dynamics (CFD) methods, for fluid simulation.
+We analyze the development of Computational Fluid Dynamics (CFD) software packages that use Lattice Boltzmann Methods (LBM).
 These LBM algorithms consider the behaviours of a collection of particles as a single unit of an intermediate size, and predict their positional probability as they move through a lattice structure.
-Figure \ref{circularflow} from \citep{chen2021phase} presents the streamlines of converged flow past a stationary circular cylinder with varied Reynolds number, a common task of LBM.
+Figure \ref{circularflow} presents the streamlines of converged flow past a stationary circular cylinder with varied Reynolds number, a common task of LBM.
 
@@
 		\includegraphics[width=0.7\textwidth]{circularflow}
-		\caption{Streamlines of the flow past a stationary circular cylinder at Re = 10, 20, and 40}
+		\caption{Streamlines of flow past a stationary circular cylinder at Re = 10, 20, and 40 \citep{chen2021phase}}
 		\label{circularflow}
@@
 The qualities are defined in Section \ref{softwarequalities}.
+Quantitative data is measured using quality metrics.
+The findings are presented in Section \ref{AHPresults}.
 Qualitative data is gathered from interviews with software package developers.
-The interview questions are listed in the Appendix in \ref{interviewquestions}.
-Quantitative data is measured using the quality metrics listed in the Appendix in \ref{measurementtemplate2}.
+The interview findings are presented in Section \ref{interviewresults}.
 In this context the purposes of the report are as follows:
@@
 We have also added a domain analysis to better characterize the functionality provided by the software, and have leveraged the expertise of a domain expert.
-As in past assessments, the collected data was combined to rank the software using the Analytic Hierarchy Process (AHP).
+As in past assessments, the collected data was combined to rank the software using the Analytical Hierarchy Process (AHP).
 The domain expert was consulted to verify the ordering.
@@
 A surface investigation of the documentation of each software package for performance information is conducted instead.
-The measured software packages are open source, and it is recommended to keep this in mind when considering the recommendations of this report.
+The measured software packages are open source.
 Some recommendations may not apply to closed-source software.
@@
 	The purpose of this analysis in this state of the practice assessment is to better classify the different software products based on their functionality.
-	\item \textbf{Methodology} of this state of the practice assessment, including the steps of the methodology, software quality definitions and how these qualities are assessed in this report, an overview of candidate software selection, empirical measures gathered, and the analytical hierarchy process.
+	\item \textbf{Methodology} of this state of the practice assessment, including the steps of the methodology, software quality definitions and how these qualities are assessed in this report, an overview of candidate software selection, empirical measures gathered, and the AHP.
 This follows a general methodology for assessing the state of the practice for SCS domains.
@@
 
-We added this domain commonality and variability analysis to our original methodology because the first time we did the State of the Practice exercise we found we weren't comparing "apples to apples".
+We added this domain commonality and variability analysis to our original methodology because the first time we did the State of the Practice exercise we found we weren't comparing ``apples to apples".
 Software packages can have considerable variation, even when the packages appear to be in the same software family.
@@
 
-LBM are a family of fluid dynamics algorithms for simulating single-phase and multiphase fluid flows, often incorporating additional physical complexities \citep{chen1998lattice}.
+LBM are a family of fluid dynamics algorithms for simulating single-phase and multiphase fluid flows, often incorporating additional physical complexities \citep{chen1998lattice} such as reflective and non-reflective boundaries.
 They consider the behaviours of a collection of particles as a single unit at the mesoscopic scale, between the nanoscopic and microscopic scales.
@@
 Although the output information could simply be written to the computer’s memory, in all practical applications it is desirable to have a persistent record of the output that was created.
-The final section covers nonfunctional requirements of the system.
-For instance, all systems will have the goal that the response time to a user’s request is small enough to allow the user to focus on his/her problem and to maintain his/her train of thought, without excessive waiting time.
-The commonality in this case is refined by a later variability because the specific requirement on the response time will depend on the individual software packages, their simulation models, inputs, and outputs.
 
@@
 		\bf Description & Every Lattice Boltzmann version uses a Boltzmann transport equation to describe the statistical behaviour of a system that does not have collisions.
-The equation is $(\frac{d}{dt} + \mathrm{e} \cdot \nabla_\mathrm{x} + \frac{F}{m} \cdot \nabla_\mathrm{e})f = \Omega(t)$, where $\mathrm{e}$ is the velocity, $\nabla_\mathrm{x}$ is the position vector gradient, $\nabla_\mathrm{e}$ is the velocity gradient, and $\Omega(t)$ is the collision operator as a function of time.\\
+The equation is $(\frac{d}{dt} + \mathrm{e} \cdot \nabla_\mathrm{x} + \frac{F}{m} \cdot \nabla_\mathrm{e})f = \Omega(t)$, where $\mathrm{e}$ is the velocity, $\nabla_\mathrm{x}$ is the position vector gradient, $F$ is the force of the fluid, $m$ is the mass, $\nabla_\mathrm{e}$ is the velocity gradient, $f$ indicates the probability density function, and $\Omega(t)$ is the collision operator as a function of time.\\
 		\hline
@@
 As in Section \ref{comm_sec}, the first three subsections on variabilities are organized into the following sublists: Simulation Models, Input and Output.
-The final two subsections list variabilities that can be characterized as system constraints and as nonfunctional requirements.
+The final subsection lists variabilities that can be characterized as system constraints.
 
@@
 		\hline
-		\bf Description & The number of fluids allowed in the LBS.\\
+		\bf Description & The number of fluids allowed in a multifluid simulation.\\
 		\hline
@@
 This section specifies the parameters of variation for the variabilities listed in Section \ref{var_sec}.
-They are organized into the same five subcategories as employed previously: Simulation Models, Input, Output, System Constraints, Nonfunctional Requirements.
+They are organized into the same five subcategories as employed previously: Simulation Models, Input, Output, and System Constraints.
 
@@
 The binding time could be during specification, or during building
-of the system (compile time), or during execution of the system (run time).
+of the system (build time), or during execution of the system (run time).
 It is possible to have a mixture of binding times.
@@
 In this project we set out to answer the research questions listed in Section \ref{purpose} for the LBM SCS domain.
-The process involved systematically measuring and analyzing members of this software family along the quality attributes found in Section \ref{softwarequalities}.
-This included gathering quantitative and qualitative measurements.
-A goal of the project was to produce a quality assessment for LBM software packages.
+The process involved systematically measuring and analyzing members of this software family along the quality attributes listed in the research questions and described in Section \ref{softwarequalities}.
+The methodology includes gathering quantitative and qualitative measurements.
+A goal of the project is to produce a quality assessment for LBM software packages.
 
@@
 It was developed as an update to previous state of the practice exercises.
-This general methodology is linked in the next subsection.
-
-We collected quantitative data using the measures found in the measurement template of Section \ref{measurementtemplate}.
+
+We collected quantitative data using the measures found in the measurement template in Appendix \ref{measurementtemplate2}.
 Some of this was empirical software engineering related data, such as the number of files, number of lines of code (LOC), percentage of issues that are closed, etc.
@@
 
-We also collected qualitative data by interviewing the software package developers and asking them the questions found in the Appendix in \ref{interviewquestions}.
+We also collected qualitative data by interviewing the software package developers and asking them the questions found in Appendix \ref{interviewquestions}.
+Ethics clearance information can be found in Appendix \ref{ethicsapproval}.
 Furthermore, we solicited the assistance of domain experts to better assess each software package by leveraging their experience to assess the functional and non-functional requirements for the software domain.
@@
 This is followed by quality definitions and how these qualities were assessed in our project.
-The rest of the section provides an overview of how candidate software packages were selected and filtered, the empirical measurements and software tools that were used, and the analytic hierarchy process (AHP).
+The rest of the section provides an overview of how candidate software packages were selected and filtered, the empirical measurements and software tools that were used, and the AHP.
 
@@
 	\item Collect empirical measures. This is discussed in Section \ref{empiricalmeasures}.
-	\item Measure using the measurement template. This is discussed in Section \ref{empiricalmeasures}. The measurement template can be found in the Appendix in \ref{measurementtemplate2}.
+	\item Measure using the measurement template. This is discussed in Section \ref{empiricalmeasures}. The measurement template can be found in Appendix \ref{measurementtemplate2}.
 	\item Survey the developers.
 The developers of each software package in the filtered software list were contacted for voluntary interviews.
-The interview questions can be found in the Appendix in \ref{interviewquestions}.
+The interview questions can be found in Appendix \ref{interviewquestions}.
 	\item Use AHP to rank the software packages. This is discussed in Section \ref{AHP}.
@@
 Much of the data was gathered by manually reviewing the artifacts of each software package, while some was gathered using the freeware tools discussed below.
-The data that was gathered is listed in the measurement template found in the Appendix in \ref{measurementtemplate2}.
+The data that was gathered is listed in the measurement template found in Appendix \ref{measurementtemplate2}.
 Some of the data required processing other data within the template, including the status of the software package, which relied on the last commit date; the percentage of issues that are closed, which relied on the number of open and closed issues; and the percentage of code that is comments, which relied on the number of total lines and comment lines in text-based files.
-The complete measurement template data was then analyzed using the grading template found in the Appendix in \ref{gradingtemplate2}, the output of which was analyzed using the AHP described in Section \ref{AHP}.
+The complete measurement template data was then analyzed using the grading template found in Appendix \ref{gradingtemplate2}, the output of which was analyzed using the AHP described in Section \ref{AHP}.
 
@@
 The Analytical Hierarchy Process (AHP) is a decision-making technique that is used to compare multiple options by multiple criteria.
-In our work AHP was used for comparing and ranking the LBM software packages using the overall impression quality scores that were gathered in the measurement template found in the Appendix in \ref{measurementtemplate2} using the grading template found in the Appendix in \ref{gradingtemplate2}.
+In our work AHP was used for comparing and ranking the LBM software packages using the overall impression quality scores that were gathered in the measurement template found in Appendix \ref{measurementtemplate2} using the grading template found in Appendix \ref{gradingtemplate2}.
 AHP performs a pairwise analysis using a matrix and generates an overall score as well as individual quality scores for each software package.
-\cite{SmithEtAl2016} shows how AHP is applied to ranking software based on quality measures.
+Smith et al.
+(2016) shows how the AHP is applied to ranking software based on quality measures \cite{SmithEtAl2016}.
 
@@
 Furthermore, nine of the top 10 ranked packages are noted as being alive.
+LIMBES is noted as being dead.
 
@@
 Software package artifacts were reviewed for the presence of a tutorial, a user manual, documented user characteristics, and a user support model.
-In total 18 software packages have a tutorial and 13 have a user manual.
+In total 18 software packages have a tutorial, 13 have a user manual, and 11 have both.
 The tutorials vary in scope and substance, and eight include an expected output.
@@
 Most software packages that use Git have most of their issues closed, and only three (laboetie, lettuce, Sailfish) have less than 50 percent of their issues closed.
+Four of the top five overall ranked packages (Ludwig, ESPResSo, Palabos, LUMA) have most of their issues closed.
+Fourth ranked LUMA does not use Git.
 Alive packages (11 use Git issue tracking) have 64\% of their issues closed, while dead packages (3 use Git issue tracking) have 71\% of their issues closed.
@@
 Packages with a higher percentage of comments were designated as more maintainable.
-Comments represent more than 10 percent of code files in 15 packages, and the average percentage of code comments is about 14.\\
+Comments represent more than 10 percent of code files in 15 packages, and the average percentage of code comments is about 14 percent.
+Four of the top five overall ranked packages (Ludwig, ESPResSo, Palabos, OpenLB) have more than the average.
+Fifth ranked LUMA has only 0.2 percent comments, the fewest of any package.
+The package has the most lines of source code, with over four million.
+The next largest package is ESPResSo++ with one million.
+
+\begin{figure}[h!]
+	\begin{center}
+		\includegraphics[width=1.0\textwidth]{maintainability_chart}
+		\caption{AHP Maintainability Score}
+		\label{Fig_Maintainability}
+	\end{center}
+\end{figure}
+
+Figure \ref{Fig_Maintainability} shows the maintainability ranking of the software packages using AHP.
+Software packages with a high score provide version numbers on documents and source code releases, have an abundance of high quality artifacts, and use an issue tracking tool and version control system.
+These packages also appear to reasonably handle tracked issues, having most of their issues closed.
+Their code files are well commented with more than 10 percent of the code being comments.
+Furthermore, nine of the top 10 ranked packages are noted as being alive.
+MP-LABS is noted as being dead.\\
 
@@
 
-\begin{figure}[h!]
-	\begin{center}
-		\includegraphics[width=1.0\textwidth]{maintainability_chart}
-		\caption{AHP Maintainability Score}
-		\label{Fig_Maintainability}
-	\end{center}
-\end{figure}
-
-Figure \ref{Fig_Maintainability} shows the maintainability ranking of the software packages using AHP.
-Software packages with a high score provide version numbers on documents and source code releases, have an abundance of high quality artifacts, and use an issue tracking tool and version control system.
-These packages also appear to reasonably handle tracked issues, having most of their issues closed.
-Their code files are well commented with more than 10 percent of the code being comments.
-Furthermore, nine of the top 10 ranked packages are noted as being alive.
-
 \subsubsection{Reusability}\label{reusabilityresults}
@@
 Furthermore, nine of the top 10 ranked packages are noted as being alive.
-
-Table \ref{moduledata} shows file and line of code data of the software packages.
-Packages with a high reusability score do not have many LOC per text file, generally having a few hundred lines or less.
-This suggests that the source code of these packages is likely functionally modularized, and modules could be reused in other projects.\\
+HemeLB is noted as being dead.
 
@@
 
+Table \ref{moduledata} shows file and line of code data of the software packages.
+Packages with a high reusability score do not have many LOC per text file, generally having a few hundred lines or less.
+This suggests that the source code of these packages is likely functionally modularized, and modules could be reused in other projects.
+
 There was a strong focus on modularity when designing the waLBerla framework to enhance productivity, reusability, and maintainability \citep{bauer2021walberla}.
@@
 Ten random source code files of each software package were reviewed for several measures.
-This assessment of surface understandability may not perfectly reflect each package due to the limitation of only testing 10 files.
+This assessment of surface understandability may not perfectly reflect each package due to the practical limitation of only being able to test 10 files.
 
@@
 Only four packages (LB2D-Prime, LB3D-Prime, LIMBES, MP-LABS) appear to use vague identifiers, such as single letters for variables.
-Hard coded symbolic constants were observed in the source code of 12 packages.
+Symbolic constants were observed in the source code of 12 packages.
 The constants are used for various parameters, mathematical constants, and matrix definitions.
@@
 Software packages with a high score have a consistent indentation and formatting style, and consistent, distinctive, and meaningful code identifiers.
-They also have hard coded constants, and explicitly identify mathematical and LBM algorithms.
+They also have symbolic constants, and explicitly identify mathematical and LBM algorithms.
 Their comments are clear and indicate what is being done in the source code.
@@
 They also
-had detailed and easy to access notes accompanying software releases. Furthermore, four of the top five ranked packages are noted as being alive.
+had detailed and easy to access notes accompanying software releases. Furthermore, four of the top five ranked packages are noted as being alive. MP-LABS is noted as being dead.
 
@@
 Software packages with an overall high score had ranked high in at least several of the individual qualities that were quantitatively measured.
+The overall ranking is found with an assumption of an equal weighting of all qualities.
+If the qualities were weighed differently, the overall software package ranking would change.
 
@@
 
-This subsection presents the qualitative findings from data that was gathered during interviews with developers. The interview questions are found in Section \ref{interviewquestions}.
+This subsection presents the qualitative findings from data that was gathered during interviews with developers. The interview questions are found in Appendix \ref{interviewquestions}.
 
@@
 The users do not realize the limitations of the methods, of the software, and do not understand the requirements to properly model a problem with the software.
-As the developer of TCLB noted, such software packages are not designed to be used ``out of the box'' in a plug and play fashion, and it could take months so set up the CFD problems correctly.
+As the developer of TCLB noted, such software packages are not designed to be used ``out of the box'' in a plug and play fashion, and it could take months or more to set up the CFD problems correctly.
 Developers of some software packages, including ESPResSo, mitigated this by editing the source code to prevent users from ``combining methods that are not physically sound together'', and by updating the documentation to better inform users of LBM limitations, and of the requirements to properly model appropriate problems, including what algorithms and parameters to use.
@@
 
-A high degree of code modularity and abstraction, ensured by separating components and hiding information behind well defined interfaces, was noted by developers as a measure to ensure the ease of future changes.
+A high degree of code modularity and abstraction was noted by developers as a measure to ensure the ease of future changes.
+This can be ensured by separating components and hiding information behind well defined interfaces.
 The developer of ESPResSo also noted that some of the code base was transitioned from C to C++, which could ease modifiability of that software package.
@@
 A frequently asked questions section was added to their package website to help users find answers to common questions.
-The package also has detailed documentation, including guides and usage requirements specification, to better help user understand the software.
+The package also has detailed documentation, including guides and usage requirements specification, to better help users understand the software.
 
@@
 Most of the developers are scientific researchers evaluated on the scientific papers that they produce.
-Writing and updating documentation is something that is done in free time, if that time arises.
+Writing and updating documentation is something that is done in their free time, if that time arises.
 Sometimes it is a last priority for the developers.
@@
 At the time of data collection, only one (Palabos) of the four packages did not have a user manual or guide, but there was a broken link on the package website indicating that such an artifact might exist.
-This broken link was later fixed, but this is not reflected in our data.
+This broken link was later fixed, but this is not reflected in our data because it was not present at the time of data collection.
 Despite the broken link, Palabos does have a detailed and informative website.
@@
 
-Development tools support the development of end products but do not become part of them, unlike dependencies that remain in the application once it is released \citep{ghezzi1991fundamentals}.
+Development tools support the development of end products, but do not become part of them, unlike dependencies that remain in the application once it is released \citep{ghezzi1991fundamentals}.
 The following type of development tools were explicitly noted in the artifacts or web-pages of the 23 LBM packages that were assessed.
@@
 The following types of dependencies were explicitly noted in the artifacts or web-pages of the 23 LBM packages that were assessed.
-It is possible that other types of dependencies are part of these software packages but are not clearly specified in their artifacts or web sites and because of that they are not listed here.
+It is possible that other types of dependencies are part of these software packages, but are not clearly specified in their artifacts or web sites and because of that they are not listed here.
 
@@
 Most of the projects appear to use change tracking and version control tools.
-They often use GitHub for this.
-One package (lbmpy) uses Git, and another (SunlightLB) uses CVS.
+They often use GitHub or GitLab for this.
+One package (SunlightLB) uses CVS.
 Document generation tools are mentioned in the artifacts of 12 of the projects.
@@
 
-\subsection{Principles, Processes, and Methodologies - DONE 1st DRAFT}\label{prinprocmeth}
+\subsection{Principles, Processes, and Methodologies}\label{prinprocmeth}
 
@@
 One is feature additions, which arise from a scientific or functional need.
-The development of such changes was noted to involve more formal discussions within the development team and the addition of them into the main software branch involves lead developers.
+These changes involve formal discussions within the development team, and lead developer participation is mandatory.
 The other change type is code refactoring, which only sometimes involves formal discussions with the development team.
@@
 As software becomes user-ready, it can become commercialized and closed-source.
+
 Documentation of LBM software is noted as being important, but its quality could be improved.
@@
 \begin{itemize}
-	\item Use a requirements specification document. A template is presented in \citep{smith2005new}. 
+	\item Use a requirements specification document. A potential template is presented in \citep{smith2005new}. 
 	\item Make public (on package website or GitHub) the requirements specification document, or explicitly reference the domain theory that the software is designed from.
@@
 	\item Use a version control system (GitHib, CVS). Four (ESPREsSo, Ludwig, LUMA, Palabos) of the top five ranked packages use a version control system.
-	\item Source code needs to be well commented. Typically, more than 10 percent of LBM package source code is comments, as presented in Table \ref{gitrepodata}.
+	\item Source code needs to be well commented.
+Typically, more than 10 percent of LBM package source code is comments, as presented in Table \ref{gitrepodata}.
+Four of the top five overall ranked packages (Ludwig, ESPResSo, Palabos, OpenLB) have about 20 percent of their source code as comments.
 	\item Modularize the source code, separate components, hide information behind well defined interfaces.
@@
 
-\begin{center}
-	\begin{onehalfspacing}
-		\begin{tabular}{ p{3.5cm}p{2.5cm}p{2.5cm}p{2.5cm} }
-			\hline
-			Name & Our Ranking & Repository Stars & Repository Watches \\
-			\hline
-			Ludwig&1& 19 & 6\\
-			ESPResSo &2& 115 & 23\\
-			Palabos &3& 14 & GitLab\\
-			OpenLB &4& No Git & No Git\\
-			LUMA&5& 26 & 10\\
-			pyLBM &6& 57& 8\\
-			DL\_MESO (LBE) & 7 & No Git & No Git \\
-			waLBerla & 8 & 17 & GitLab\\
-			Sailfish &9& 154 & 39\\
-			laboetie &10& 4 & 5\\
-			TCLB &11& 61 & 16\\
-			MechSys &12& No Git & No Git\\
-			lettuce &13& 8 & 2\\
-			ESPResSo++ &14& 31 & 13\\
-			MP-LABS &15& 7 & 2\\			
-			SunlightLB & 16& No Git & No Git\\
-			LB3D &17& No Git & No Git\\			
-			LIMBES &18& No Git & No Git\\
-			LB2D-Prime &19& No Git & No Git\\		
-			HemeLB &20& 22 & 15\\
-			lbmpy&21&  4 & 2  \\	
-			LB3D-Prime &22& No Git & No Git\\	
-			LatBo.jl &23& 4 & 5\\			
-			\hline
-		\end{tabular}
-		\captionof{table}{Repository Ranking Metrics}\label{repometrics}
-	\end{onehalfspacing}
-\end{center}
-
 Table \ref{repometrics} presents our LBM software package rankings along with the repository ranking metrics of each software package.
@@
 Our assessment of this package might not be accurate.
+The number of stars does not necessarily represent the perceptions of the community, but for lack of an alternative measure we will use it this way.
 
@@
 The true quality of new packages may not be reflected in their stars and watches.
+\\
+
+\begin{center}
+	\begin{onehalfspacing}
+		\begin{tabular}{ p{3.5cm}p{2.5cm}p{2cm}p{2cm}p{2cm} }
+			\hline
+			Name & Our Ranking & Repository Stars & Repository Star Rank & Repository Watches \\
+			\hline
+			Ludwig&1& 19 &8& 6\\
+			ESPResSo &2& 115 &2& 23\\
+			Palabos &3& 14 &10& GitLab\\
+			OpenLB &4& No Git &No Git & No Git\\
+			LUMA&5& 26 &6& 10\\
+			pyLBM &6& 57&4& 8\\
+			DL\_MESO (LBE) & 7 & No Git &No Git & No Git \\
+			waLBerla & 8 & 17 &9& GitLab\\
+			Sailfish &9& 154 &1& 39\\
+			laboetie &10& 4 &13& 5\\
+			TCLB &11& 61 &3& 16\\
+			MechSys &12& No Git &No Git & No Git\\
+			lettuce &13& 8 &11& 2\\
+			ESPResSo++ &14& 31 &5& 13\\
+			MP-LABS &15& 7 &12& 2\\			
+			SunlightLB & 16& No Git &No Git & No Git\\
+			LB3D &17& No Git &No Git & No Git\\			
+			LIMBES &18& No Git &No Git & No Git\\
+			LB2D-Prime &19& No Git &No Git & No Git\\		
+			HemeLB &20& 22 &7& 15\\
+			lbmpy&21&  4 &13& 2  \\	
+			LB3D-Prime &22& No Git &No Git & No Git\\	
+			LatBo.jl &23& 4 &13& 5\\			
+			\hline
+		\end{tabular}
+		\captionof{table}{Repository Ranking Metrics}\label{repometrics}
+	\end{onehalfspacing}
+\end{center}
 
@@
 It was made public the year before data was collected for this state of the practice exercise.
-It may not have been on authoritative lists due to its age.
+It may not have been on authoritative lists due to its young age.
 Similarly, LIFE was also made public recently.
@@
 	\begin{onehalfspacing}
-		\begin{tabular}{ p{2cm}p{3.5cm}p{3.5cm}}
+		\begin{tabular}{ p{2cm}p{4.5cm}p{3.5cm}}
 			\hline
@@
 			\hline
-			1&Ludwig&Palabos\\
-			2&ESPResSo &OpenLB\\
-			3&Palabos &LUMA\\
-			4&OpenLB &ASL\\
-			5&LUMA&ch4-project\\
-			6&pyLBM &Open FSI\\
-			7&DL\_MESO (LBE) & WaLBerla\\
-			8&waLBerla & LIFE\\
-			9&Sailfish &Sailfish\\
-			10&laboetie &HemeLB\\		
+			1&Ludwig (N/A) &Palabos\\
+			2&ESPResSo (N/A) &OpenLB\\
+			3&Palabos (1)&LUMA\\
+			4&OpenLB (2)&ASL\\
+			5&LUMA (3)&ch4-project\\
+			6&pyLBM (17)&Open FSI\\
+			7&DL\_MESO (LBE) (12)& WaLBerla\\
+			8&waLBerla (7)& LIFE\\
+			9&Sailfish (9)&Sailfish\\
+			10&laboetie (N/A)&HemeLB\\		
 			\hline
@@
 This section examines potential threats to the validity of this state of the practice assessment. These can be categorized into methodology and data collection issues.
+
+The goal of this assessment isn't to rank the software, but to use the ranking exercise as a means to understand the state of the practice of LBM software development.
 
@@
 The software qualities that were assessed in this report are listed in Section \ref{softwarequalities}.
-A methodology for assessing the state of the practice of software development in scientific computing software domains was presented.
+A methodology for assessing the state of the practice of software development in SCS domains was presented.
 Domain packages were assessed to answer the software development related research questions listed in Section \ref{purpose} to understand how software quality is impacted by software development choices, including principles, processes, and tools.
 Software developers were interviewed to identify development pain points, and to identify how software quality is ensured.
-Quantitative data was used to rank the software packages using an analytical hierarchy process.
+Quantitative data was used to rank the software packages using the AHP.
 The ranking designations were compared with rankings from the software development community, and we found that many of our top 10 ranked packages are ranked highly by a domain expert.
 Recommendations for improving software along quality metrics were made, and highlights are presented in Section \ref{highlightedrecommendations} of this conclusion.
-The findings of this report can be used to guide future development of scientific computing software, specifically along quality attributes, and to reduce software quality failures.
+The findings of this report can be used to guide future development of SCS, specifically along quality attributes, and to reduce software quality failures.
 
@@
 Ensuring software quality can take significant time.
-An inability to develop high quality documentation due to time constraints could be mitigated by the use of automatic document generation tools like Drasil.
+An inability to develop high quality documentation due to time constraints could be mitigated in the future by the use of automatic document generation tools like Drasil.
 The Drasil Framework consists of a collection of Domain Specific Languages (DSL) for capturing scientific documents, structures, and computing knowledge, and then transforming this knowledge into relevant software artifacts without having to manually duplicate knowledge into multiple artifacts \citep{zhao2018}.
@@
 
-As mentioned in Section \ref{threats}, the measures listed in our measurement template may not be broad enough to accurately capture some qualities. Adding or extending measures is worth considering.
+As mentioned in Section \ref{threats}, the measures listed in our measurement template may not be broad enough to accurately capture some qualities.
+Adding or extending measures is worth considering.
+For example, usability experiments and performance benchmarks could be incorporated into the assessment.
 
@@
 \newpage
-\section{Appendix}
+\begin{appendices}
+
 \begin{singlespace}
 
-\subsection{Developer Interview Questions}\label{interviewquestions}
+\section{Measurement Template}\label{measurementtemplate2}
+The table below lists the set of measures that are used to assess each software product.
+The first set identifies summary information, followed by 9 sets for software qualities and 3 sets for raw metrics.
+Each measure is followed by the type for a valid result.
+A superscript ∗ indicate that a response of this type needs to be accompanied by explanatory text.\\
+\newpage
+\captionof{table}{Measurement Template}\label{measurementtemplate}
+\def\arraystretch{1.22}
+\begin{tabular}{p{14cm}}
+	\hline
+	\textbf{Summary Information}\\
+	\hline
+	Software name? (string)\\
+	URL? (URL)\\
+	Affiliation (institution(s)) (string or {N/A})\\
+	Software purpose (string)\\
+	Number of developers (all developers that have contributed at least one commit to the project) (use repo commit logs) (number)\\
+	How is the project funded? (unfunded, unclear, funded*) where * requires a string to say the source of funding\\
+	Initial release date? (date)\\
+	Last commit date? (date)\\
+	Status? (alive is defined as presence of commits in the last 18 months) ({alive, dead, unclear})\\
+	License? ({GNU GPL, BSD, MIT, terms of use, trial, none, unclear, other*}) * given via a string \\
+	Platforms? (set of {Windows, Linux, OS X, Android, other*}) * given via string\\
+	Software Category?
+The concept category includes software that does not have an officially released version.
+Public software has a released version in the public domain.
+Private software has a released version available to authorized users only.
+({concept, public, private})\\
+	Development model? ({open source, freeware, commercial, unclear})\\
+	Publications about the software? Refers to publications that have used or mentioned the software. (number or {unknown})\\
+	Source code URL? ({set of url, n/a, unclear})\\
+	Programming language(s)? (set of {FORTRAN, Matlab, C, C++, Java, R, Ruby, Python, Cython, BASIC, Pascal, IDL, unclear, other*}) * given via string \\
+	Is there evidence that performance was considered? Performance refers to either speed, storage, or throughput. ({yes∗, no})\\
+	Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have) \\
+	\hline
+\end{tabular}
+
+\def\arraystretch{1.5}
+\begin{tabular}{p{14cm}}
+	\hline		
+	\textbf{Installability  (Measured via installation on a virtual machine.) }\\
+	\hline
+	Are there installation instructions? ({yes, no})\\
+	Are the installation instructions in one place? Place referring to a single document or web-page. ({yes, no, n/a})\\
+	Are the installation instructions linear? Linear meaning progressing  in a single series of steps. ({yes, no, n/a})\\
+	Are the instructions written as if the person doing the installation has none of the dependent packages installed? ({yes, no, unclear})\\
+	Are compatible operating system versions listed? ({yes, no})\\
+	Is there something in place to automate the installation (makefile, script, installer, etc)? ({yes*, no})\\
+	If the software installation broke, was a descriptive error message displayed? ({yes, no, n/a})\\
+	Is there a specified way to validate the installation? ({yes*, no})\\
+	How many steps were involved in the installation? (Includes manual steps like unzipping files) Specify OS. (number, OS)\\
+	What OS was used for the installation? ({Windows, Linux, OS X, Android, other* }) *given via string\\
+	How many extra software packages need to be installed before or during installation? (number)\\
+	Are required package versions listed? ({yes, no, n/a})\\
+	Are there instructions for the installation of required packages / dependencies? ({yes, no, n/a})\\
+	Run uninstall, if available. Were any obvious problems caused? ({yes∗ , no, unavail})\\
+	Overall impression? ({1 .. 10})\\
+	Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have)\\
+	\hline
+\end{tabular}
+
+\def\arraystretch{1.33}
+\begin{tabular}{p{14cm}}
+	\hline		
+	\textbf{Correctness and Verifiability}\\
+	\hline
+	Any reference to the requirements specifications of the program or theory manuals? ({yes∗ , no, unclear})\\
+	What tools or techniques are used to build confidence of correctness?
+({literate programming, automated testing, symbolic execution, model checking, assertions used in the code, Sphinx, Doxygen, Javadoc, confluence, unclear, other*}) * given via string\\
+	If there is a getting started tutorial? ({yes, no})\\
+	Are the tutorial instructions linear? ({yes, no, n/a})\\
+	Does the getting started tutorial provide an expected output? ({yes, no*, n/a})\\
+	Does your tutorial output match the expected output? ({yes, no, n/a})\\
+	Are unit tests available?  ({yes, no, unclear})\\
+	Is there evidence of continuous integration? (for example mentioned in documentation, Jenkins, Travis CI, Bamboo, other) ({yes*, no, unclear})\\
+	Overall impression? ({1 .. 10})\\
+	Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have) \\
+	\hline	
+	\textbf{Surface Reliability}\\
+	\hline
+	Did the software “break” during installation? ({yes∗ , no})\\
+	If the software installation broke, was the installation instance recoverable? ({yes, no, n/a})\\
+	Did the software “break” during the initial tutorial testing? ({yes∗, no, n/a})\\
+	If the tutorial testing broke, was a descriptive error message displayed? ({yes, no, n/a})\\
+	If the tutorial testing broke, was the tutorial testing instance recoverable? ({yes, no, n/a})\\
+	Overall impression? ({1 .. 10})\\
+	Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have)\\
+	\hline		
+\end{tabular}
+
+\def\arraystretch{1.4}
+\begin{tabular}{p{14cm}}
+	\hline		
+	\textbf{Surface Robustness}\\
+	\hline
+	Does the software handle unexpected/unanticipated input (like data of the wrong type, empty input, missing files or links) reasonably?
+(a reasonable response can include an appropriate error message.) ({yes, no∗ })\\
+	For any plain text input files, if all new lines are replaced with new lines and carriage returns, will the software handle this gracefully? ({yes, no∗, n/a})\\
+	Overall impression? ({1 .. 10})\\
+	Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have)\\
+	\hline		
+	\textbf{Surface Usability}\\
+	\hline
+	Is there a getting started tutorial? ({yes, no})\\
+	Is there a user manual? ({yes, no})\\
+	Are expected user characteristics documented? ({yes, no})\\
+	What is the user support model? FAQ? User forum? E-mail address to direct questions? Etc. (string)\\
+	Overall impression? ({1 .. 10})\\
+	Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have)\\
+	\hline
+\end{tabular}
+
+\def\arraystretch{1.4}
+\begin{tabular}{p{14cm}}
+		\hline	
+	\textbf{Maintainability}\\
+	\hline
+	What is the current version number? (number)\\
+	Is there any information on how code is reviewed, or how to contribute? ({yes*, no})\\
+	Are artifacts available?
+(List every type of file that is not a code file – for examples please look at the ‘Artifact Name’ column of https://gitlab.cas.mcmaster.ca/SEforSC/se4sc/-/blob/git-svn/GradStudents/Olu/ResearchProposal/Artifacts\_MiningV3.xlsx) ({yes*, no, unclear}) *list via string\\
+	What issue tracking tool is employed? (set of {Trac, JIRA, Redmine, e-mail, discussion board, sourceforge, google code, git, BitBucket, none, unclear, other*}) * given via string\\
+	What is the percentage of identified issues that are closed? (percentage)\\
+	What percentage of code is comments? (percentage)\\
+	Which version control system is in use? ({svn, cvs, git, github, unclear, other*}) * given via string\\
+	Overall impression? ({1 .. 10})\\
+	Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have)\\
+	\hline		
+	\textbf{Reusability}\\
+	\hline
+	How many code files are there? (number)\\
+	Is API documented? ({yes, no, n/a})\\
+	Overall impression? ({1 .. 10})\\
+	Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have)\\
+	\hline		
+\end{tabular}
+
+\def\arraystretch{1.4}
+\begin{tabular}{p{14cm}}
+		\hline		
+	\textbf{Surface Understandability (Based on 10 random source files)}\\
+	\hline
+	Consistent indentation and formatting style? ({yes, no, n/a})\\
+	Explicit identification of a coding standard? ({yes∗, no, n/a})\\
+	Are the code identifiers consistent, distinctive, and meaningful? ({yes, no∗ , n/a})\\
+	Are constants (other than 0 and 1) hard coded into the program? ({yes, no∗ , n/a})\\
+	Comments are clear, indicate what is being done, not how? ({yes, no∗ , n/a})\\
+	Is the name/URL of any algorithms used mentioned? ({yes, no∗ , n/a})\\
+	Parameters are in the same order for all functions? ({yes, no∗ , n/a})\\
+	Is code modularized? ({yes, no∗ , n/a})\\
+	Overall impression? ({1 .. 10})\\
+	Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have)\\
+	\hline		
+	\textbf{Visibility/Transparency}\\
+	\hline
+	Is the development process defined? If yes, what process is used. ({yes∗, no, n/a})\\
+	Are there any documents recording the development process and status?  ({yes∗, no}))\\
+	Is the development environment documented? ({yes∗, no})\\
+	Are there release notes? ({yes∗, no})\\
+	Overall impression? ({1 .. 10})\\
+	Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have)\\
+		\hline		
+	\textbf{Raw Metrics (Measured via git\_stats)}\\
+	\hline
+	Number of text-based files. (number)\\
+	Number of binary files. (number)\\
+	Number of total lines in text-based files. (number)\\
+	Number of total lines added to text-based files. (number)\\
+	Number of total lines deleted from text-based files. (number)\\
+	Number of total commits. (number)\\
+	Numbers of commits by year in the last 5 years. (Count from as early as possible if the project is younger than 5 years.) (list of numbers)\\
+	Numbers of commits by month in the last 12 months. (list of numbers)\\
+	\hline
+\end{tabular}
+
+\def\arraystretch{1.4}
+\begin{tabular}{p{14cm}}
+		\hline		
+	\textbf{Raw Metrics (Measured via scc)}\\
+	\hline
+	Number of text-based files. (number)\\
+	Number of total lines in text-based files. (number)\\
+	Number of code lines in text-based files. (number)\\
+	Number of comment lines in text-based files. (number)\\
+	Number of blank lines in text-based files. (number)\\
+	\hline
+	\textbf{Repo Metrics (Measured via GitHub)}\\
+	\hline
+	Number of stars. (number)\\
+	Number of forks. (number)\\
+	Number of people watching this repo. (number)\\
+	Number of open pull requests. (number)\\
+	Number of closed pull requests. (number)\\
+	\hline
+\end{tabular}
+
+\newpage
+
+\section{Developer Interview Questions}\label{interviewquestions}
 
@@
 
-\subsection{Measurement Template}\label{measurementtemplate2}
-The table below lists the set of measures that are used to assess each software product.
-The first set identifies summary information, followed by 9 sets for software qualities and 3 sets for raw metrics.
-Each measure is followed by the type for a valid result.
-A superscript ∗ indicate that a response of this type needs to be accompanied by explanatory text.
-
-\captionof{table}{Measurement Template}\label{measurementtemplate}
-\def\arraystretch{1.22}
-\begin{tabular}{p{16cm}}
-		\hline
-		\textbf{Summary Information}\\
-		\hline
-		Software name? (string)\\
-		URL? (URL)\\
-		Affiliation (institution(s)) (string or {N/A})\\
-		Software purpose (string)\\
-		Number of developers (all developers that have contributed at least one commit to the project) (use repo commit logs) (number)\\
-		How is the project funded? (unfunded, unclear, funded*) where * requires a string to say the source of funding\\
-		Initial release date? (date)\\
-		Last commit date? (date)\\
-		Status? (alive is defined as presence of commits in the last 18 months) ({alive, dead, unclear})\\
-		License? ({GNU GPL, BSD, MIT, terms of use, trial, none, unclear, other*}) * given via a string \\
-		Platforms? (set of {Windows, Linux, OS X, Android, other*}) * given via string\\
-		Software Category?
-The concept category includes software that does not have an officially released version.
-Public software has a released version in the public domain.
-Private software has a released version available to authorized users only.
-({concept, public, private})\\
-		Development model? ({open source, freeware, commercial, unclear})\\
-		Publications about the software? Refers to publications that have used or mentioned the software. (number or {unknown})\\
-		Source code URL? ({set of url, n/a, unclear})\\
-		Programming language(s)? (set of {FORTRAN, Matlab, C, C++, Java, R, Ruby, Python, Cython, BASIC, Pascal, IDL, unclear, other*}) * given via string \\
-		Is there evidence that performance was considered? Performance refers to either speed, storage, or throughput. ({yes∗, no})\\
-		Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have) \\
-		\hline
-\end{tabular}
-
-\def\arraystretch{1.5}
-\begin{tabular}{p{16cm}}
-		\hline		
-		\textbf{Installability  (Measured via installation on a virtual machine.) }\\
-		\hline
-		Are there installation instructions? ({yes, no})\\
-		Are the installation instructions in one place? Place referring to a single document or web-page. ({yes, no, n/a})\\
-		Are the installation instructions linear? Linear meaning progressing  in a single series of steps. ({yes, no, n/a})\\
-		Are the instructions written as if the person doing the installation has none of the dependent packages installed? ({yes, no, unclear})\\
-		Are compatible operating system versions listed? ({yes, no})\\
-		Is there something in place to automate the installation (makefile, script, installer, etc)? ({yes*, no})\\
-		If the software installation broke, was a descriptive error message displayed? ({yes, no, n/a})\\
-		Is there a specified way to validate the installation? ({yes*, no})\\
-		How many steps were involved in the installation? (Includes manual steps like unzipping files) Specify OS. (number, OS)\\
-		What OS was used for the installation? ({Windows, Linux, OS X, Android, other* }) *given via string\\
-		How many extra software packages need to be installed before or during installation? (number)\\
-		Are required package versions listed? ({yes, no, n/a})\\
-		Are there instructions for the installation of required packages / dependencies? ({yes, no, n/a})\\
-		Run uninstall, if available. Were any obvious problems caused? ({yes∗ , no, unavail})\\
-		Overall impression? ({1 .. 10})\\
-		Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have)\\
-		\hline
-\end{tabular}
-
-\def\arraystretch{1.33}
-\begin{tabular}{p{16cm}}
-	\hline		
-	\textbf{Correctness and Verifiability}\\
-	\hline
-	Any reference to the requirements specifications of the program or theory manuals? ({yes∗ , no, unclear})\\
-	What tools or techniques are used to build confidence of correctness?
-({literate programming, automated testing, symbolic execution, model checking, assertions used in the code, Sphinx, Doxygen, Javadoc, confluence, unclear, other*}) * given via string\\
-	If there is a getting started tutorial? ({yes, no})\\
-	Are the tutorial instructions linear? ({yes, no, n/a})\\
-	Does the getting started tutorial provide an expected output? ({yes, no*, n/a})\\
-	Does your tutorial output match the expected output? ({yes, no, n/a})\\
-	Are unit tests available?  ({yes, no, unclear})\\
-	Is there evidence of continuous integration? (for example mentioned in documentation, Jenkins, Travis CI, Bamboo, other) ({yes*, no, unclear})\\
-	Overall impression? ({1 .. 10})\\
-	Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have) \\
-	\hline	
-	\textbf{Surface Reliability}\\
-	\hline
-	Did the software “break” during installation? ({yes∗ , no})\\
-	If the software installation broke, was the installation instance recoverable? ({yes, no, n/a})\\
-	Did the software “break” during the initial tutorial testing? ({yes∗, no, n/a})\\
-	If the tutorial testing broke, was a descriptive error message displayed? ({yes, no, n/a})\\
-	If the tutorial testing broke, was the tutorial testing instance recoverable? ({yes, no, n/a})\\
-	Overall impression? ({1 .. 10})\\
-	Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have)\\
-	\hline		
-	\textbf{Surface Robustness}\\
-	\hline
-	Does the software handle unexpected/unanticipated input (like data of the wrong type, empty input, missing files or links) reasonably?
-(a reasonable response can include an appropriate error message.) ({yes, no∗ })\\
-	For any plain text input files, if all new lines are replaced with new lines and carriage returns, will the software handle this gracefully? ({yes, no∗, n/a})\\
-	Overall impression? ({1 .. 10})\\
-	Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have)\\
-	\hline
-\end{tabular}
-
-\def\arraystretch{1.4}
-\begin{tabular}{p{16cm}}
-	\hline		
-	\textbf{Surface Usability}\\
-	\hline
-	Is there a getting started tutorial? ({yes, no})\\
-	Is there a user manual? ({yes, no})\\
-	Are expected user characteristics documented? ({yes, no})\\
-	What is the user support model? FAQ? User forum? E-mail address to direct questions? Etc. (string)\\
-	Overall impression? ({1 .. 10})\\
-	Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have)\\
-	\hline	
-	\textbf{Maintainability}\\
-	\hline
-	What is the current version number? (number)\\
-	Is there any information on how code is reviewed, or how to contribute? ({yes*, no})\\
-	Are artifacts available?
-(List every type of file that is not a code file – for examples please look at the ‘Artifact Name’ column of https://gitlab.cas.mcmaster.ca/SEforSC/se4sc/-/blob/git-svn/GradStudents/Olu/ResearchProposal/Artifacts\_MiningV3.xlsx) ({yes*, no, unclear}) *list via string\\
-	What issue tracking tool is employed? (set of {Trac, JIRA, Redmine, e-mail, discussion board, sourceforge, google code, git, BitBucket, none, unclear, other*}) * given via string\\
-	What is the percentage of identified issues that are closed? (percentage)\\
-	What percentage of code is comments? (percentage)\\
-	Which version control system is in use? ({svn, cvs, git, github, unclear, other*}) * given via string\\
-	Overall impression? ({1 .. 10})\\
-	Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have)\\
-	\hline		
-	\textbf{Reusability}\\
-	\hline
-	How many code files are there? (number)\\
-	Is API documented? ({yes, no, n/a})\\
-	Overall impression? ({1 .. 10})\\
-	Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have)\\
-	\hline
-\end{tabular}
-
-\def\arraystretch{1.4}
-\begin{tabular}{p{16cm}}
-	\hline		
-	\textbf{Surface Understandability (Based on 10 random source files)}\\
-	\hline
-	Consistent indentation and formatting style? ({yes, no, n/a})\\
-	Explicit identification of a coding standard? ({yes∗, no, n/a})\\
-	Are the code identifiers consistent, distinctive, and meaningful? ({yes, no∗ , n/a})\\
-	Are constants (other than 0 and 1) hard coded into the program? ({yes, no∗ , n/a})\\
-	Comments are clear, indicate what is being done, not how? ({yes, no∗ , n/a})\\
-	Is the name/URL of any algorithms used mentioned? ({yes, no∗ , n/a})\\
-	Parameters are in the same order for all functions? ({yes, no∗ , n/a})\\
-	Is code modularized? ({yes, no∗ , n/a})\\
-	Overall impression? ({1 .. 10})\\
-	Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have)\\
-	\hline		
-	\textbf{Visibility/Transparency}\\
-	\hline
-	Is the development process defined? If yes, what process is used. ({yes∗, no, n/a})\\
-	Are there any documents recording the development process and status?  ({yes∗, no}))\\
-	Is the development environment documented? ({yes∗, no})\\
-	Are there release notes? ({yes∗, no})\\
-	Overall impression? ({1 .. 10})\\
-	Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have)\\
-	\hline		
-	\textbf{Raw Metrics (Measured via git\_stats)}\\
-	\hline
-	Number of text-based files. (number)\\
-	Number of binary files. (number)\\
-	Number of total lines in text-based files. (number)\\
-	Number of total lines added to text-based files. (number)\\
-	Number of total lines deleted from text-based files. (number)\\
-	Number of total commits. (number)\\
-	Numbers of commits by year in the last 5 years. (Count from as early as possible if the project is younger than 5 years.) (list of numbers)\\
-	Numbers of commits by month in the last 12 months. (list of numbers)\\
-	\hline
-\end{tabular}
-
-\def\arraystretch{1.4}
-\begin{tabular}{p{16cm}}
-	\hline		
-	\textbf{Raw Metrics (Measured via scc)}\\
-	\hline
-	Number of text-based files. (number)\\
-	Number of total lines in text-based files. (number)\\
-	Number of code lines in text-based files. (number)\\
-	Number of comment lines in text-based files. (number)\\
-	Number of blank lines in text-based files. (number)\\
-	\hline
-	\textbf{Repo Metrics (Measured via GitHub)}\\
-	\hline
-	Number of stars. (number)\\
-	Number of forks. (number)\\
-	Number of people watching this repo. (number)\\
-	Number of open pull requests. (number)\\
-	Number of closed pull requests. (number)\\
-	\hline
-\end{tabular}
-
-\newpage
-
-\subsection{Grading Template}\label{gradingtemplate2}
+\section{Grading Template}\label{gradingtemplate2}
 The table below lists how each quality measure of the measurement template is used to calculate an overall impression in each software quality set.\newline
@@
 \def\arraystretch{1.5}
-\begin{tabular}{p{16cm}}
+\begin{tabular}{p{14cm}}
 	\hline		
@@
 \def\arraystretch{1.33}
-\begin{tabular}{p{16cm}}
+\begin{tabular}{p{14cm}}
 	\hline		
@@
 	\hline		
+\end{tabular}
+
+\def\arraystretch{1.4}
+\begin{tabular}{p{14cm}}
+		\hline		
 	\textbf{Surface Usability}\\
@@
 	What is the user support model? FAQ? User forum? E-mail address to direct questions? Etc. (one=1, two+=2, none=0)\\
-	\hline
-\end{tabular}
-
-\def\arraystretch{1.4}
-\begin{tabular}{p{16cm}}
 	\hline
@@
 	\hline
+\end{tabular}
+
+\def\arraystretch{1.4}
+\begin{tabular}{p{14cm}}
+\hline	
 \textbf{Surface Understandability (Based on 10 random source files)}\\
@@
 \end{tabular}
+
 \newpage
 
-\subsection{Eliminated Software Packages}\label{eliminatedpackagessection}
+\section{Eliminated Software Packages}\label{eliminatedpackagessection}
 
@@
 
-\subsection{Ethics Approval}
+\section{Ethics Approval}\label{ethicsapproval}
 This project received ethics clearance from the McMaster Research Ethics Board on February 20, 2021.\newline
@@
 
+\end{singlespace}
+\end{appendices}
 \newpage
-
+\begin{singlespace}
 \bibliographystyle {plainnat}

### 2021-07-13 StateOfPractice/Ao-Report/15InterviewsWithDevelopers.tex
updated the outlines
ao report - Ch 5 & Appd C

@@
 \chapter{Interviews with Developers}
-\label{ch_interviews}
+\label{ch_interview}
 
-\section{Interview Methods}
+This section summarizes some answers from the interviews with developers.
+We interviewed nine developers from eight of the 29 MI software projects.
+The eight projects are \textit{3D Slicer}, \textit{INVESALIUS 3}, \textit{dwv}, \textit{BioImage Suite Web}, \textit{ITK-SNAP}, \textit{MRIcroGL}, \textit{Weasis}, and \textit{OHIF}.
+We spent about 90 minutes for each interview and asked 20 prepared questions.
+We also asked following-up questions when we thought it was worth diving deeper.
+One participant was too busy to have an interview, so he wrote down the answers and sent them to us.
 
-The way to ask questions, the way to transcript answers and the technologies used.
+In this section, we only discuss the answers which we think are the most exciting and essential.
+We summarize the rest part of the interviews in Appendix \ref{ap_interview}.
+The interviewees may have provided multiple answers to each question.
+When counting the number of interviewees for each response, we count all of them with this answer.
 
-\section{Interviewee Selection}
+\section{Current and Past Pain Points}
 
-Generally speaking, ask all the teams which we can find contacts, and continue with the ones who are willing to participate.
+By asking questions 9, 10, and 12, we tried to identify the pain points during the development process in the eight projects.
+The pain points include current and past obstacles.
+We also asked the interviewees how they would solve the problems.
 
-\section{Interview Question Selection}
+Questions 9, 10, and 12 are as follows,
 
-The aspects we focused on.
+\textit{Q9. Currently, what are the most significant obstacles in your development process?}
 
-\section{Summary of Answers}
+\textit{Q10. How might you change your development process to remove or reduce these obstacles?}
+
+\textit{Q12. In the past, is there any major obstacle to your development process that has been solved? How did you solve it?}
+
+Table \ref{tab_obstacles} shows the number of times the interviewees mentioned the current and past obstacles in their projects.
+
+\begin{table}[H]
+\centering
+\begin{tabular}{llll}
+\hline
+\multirow{2}{*}{Group} & \multirow{2}{*}{Obstacle} & \multicolumn{2}{l}{Number of interviewees with the answer} \\ \cline{3-4} 
+ &  & as a current obstacle & as a past obstacle \\ \hline
+\multirow{2}{*}{1} & Lack of fundings & 3 &  \\
+ & Lack of time to devote to the project & 2 & 1 \\ \hdashline
+\multirow{3}{*}{2} & \begin{tabular}[c]{@{}l@{}}Hard to keep up with changes in operating \\ systems and libraries\end{tabular} & 1 &  \\
+ & Hard to support multiple operating systems & 2 &  \\
+ & Hard to support lower-end computers & 1 & 2 \\ \hdashline
+3 & Lack of access to real-world datasets for testing & 3 & 2 \\ \hdashline
+\multirow{7}{*}{Others}
+ & Hard to have a high level roadmap from the start & 1 &  \\
+ & Not enough participants for usability tests & 1 &  \\
+ & \begin{tabular}[c]{@{}l@{}}Codebase is so big and complicated and only\\ a few people fully understand it\end{tabular} & 1 &  \\
+ & Hard to transfer to new technologies & & 2 \\
+ & Hard to understand users' needs & & 1 \\
+ & Hard to maintain good documentations & & 1 \\ \hline
+\end{tabular}
+\caption{\label{tab_obstacles}Current and past obstacles by the numbers of interviewees with the answers}
+\end{table}
+
+The interviewees provided some potential and proven solutions for the problems in Table \ref{tab_obstacles}.
+We group these pain points into three major groups, and put the less mentioned ones into the group \textit{Others}.
+Sections \ref{sec_pain_points_1}, \ref{sec_pain_points_2}, and \ref{sec_pain_points_3} include further discussion about the three major groups of pain points and their solutions.
+
+\subsection{Pain Points in Group 1}
+\label{sec_pain_points_1}
+We summarize the pain points in Group 1 as
+\textbf{the lack of fundings and time.}
+
+We also summarize the potential and proven solutions as follows.
+
+\textit{Potential solutions from interviewees:}
+\begin{itemize}
+\item shifting from development mode toward maintenance mode when developers are not enough;
+\item licensing the software to commercial companies that integrate it into their products;
+\item better documentation to save time for answering users' and developers' questions;
+\item supporting third-party plugins and extensions.
+\end{itemize}
+
+\textit{Proven solutions from interviewees:}
+\begin{itemize}
+\item GitHub Actions, which is a good continuous integration and continuous delivery (CI/CD) tool to save time.
+\end{itemize}
+
+Many interviewees thought lack of fundings and lack of time were the most significant obstacles.
+The interviewees from \textit{3D Slicer} team and \textit{OHIF} team pointed out that it was more challenging to get fundings for software maintenance as opposed to researches.
+The interviewee from the \textit{ITK-SNAP} team thought enough fundings was a way to solve to problem of lacking time, because they could hire more dedicated developers.
+On the other hand, the interviewee from the \textit{ITK-SNAP} team did not think that fundings could solve the same problem, since he still would need a lot of time to supervise the project.
+
+No interviewee suggested any solution to solve the funding problem fundamentally. However, they provided ideas to save time, such as better documentation, third-party plugins, and good CI/CD tools.
+
+\subsection{Pain Points in Group 2}
+\label{sec_pain_points_2}
+\label{sec_pain_points_1}
+We summarize the pain points in Group 2 as
+\textbf{the difficulty to balance between four factors: cross-platform compatibility, convenience to development \& maintenance, performance, and security.} They are also related to the choice between native application and web application.
+
+We also summarize the potential and proven solutions as follows.
+
+\textit{Potential solutions from interviewees:}
+\begin{itemize}
+\item web applications that use computing power from computers GPU;
+\item to better support lower-end computers, adopting a web-based approach with backend servers;
+\item to better support lower-end computers, using memory-mapped files to consume less computer memory;
+\item more fundings;
+\item maintaining better documentations to ease the development \& maintenance processes;
+\end{itemize}
+
+\textit{Proven solutions from interviewees:}
+\begin{itemize}
+\item one interviewee saw the performance problem disappeared over the years when computers became more and more powerful. 
+\end{itemize}
+
+Table \ref{tab_native_vs_web} shows the teams' choices between native application and web application.
+In all the 29 teams on our list, most of them chose to develop native applications.
+For the eight teams we interviewed, three of them were building web applications, and the \textit{MRIcroGL} team was considering web-based solutions.
+So we had a good chance to discuss the differences between the two choices with the interviewees.
+
+\begin{table}[H]
+\centering
+\begin{tabular}{lll}
+\hline
+Software team & Native application & Web application \\ \hline
+3D Slicer & X & \\
+INVESALIUS 3 & X & \\
+dwv & & X \\
+BioImage Suite Web & & X \\
+ITK-SNAP & X & \\
+MRIcroGL & X & \\
+Weasis & X & \\
+OHIF & & X \\ \hdashline
+Total number among the eight teams & 5 & 3 \\
+Total number among the 29 teams & 24 & 5 \\ \hline
+\end{tabular}
+\caption{\label{tab_native_vs_web}Teams' choices between native application and web application}
+\end{table}
+
+The interviewees talked about the advantages and disadvantages of the two choices. We summarize the opinions from the interviewees as follows.
+
+\textit{Native application}:
+\begin{itemize}
+\item Advantages
+    \begin{itemize}
+    \item higher performance
+    \end{itemize}
+\item Disadvantages
+    \begin{itemize}
+    \item hard to achieve cross-platform compatibility
+    \item more complicated build process
+    \end{itemize}
+\end{itemize}
+\textit{Web application}:
+\begin{itemize}
+\item Advantages
+    \begin{itemize}
+    \item easy to achieve cross-platform compatibility
+    \item simpler build process
+    \end{itemize}
+\item Disadvantages
+    \begin{itemize}
+    \item lower performance (without a backend)    \item may not meet the requirements of protecting patients' privacy (with a backend)
+    \item extra cost for backend servers (with a backend)
+    \end{itemize}
+\end{itemize}
+
+The interviewees did not conclude any perfect solutions.
+
+\subsection{Pain Points in Group 3}
+\label{sec_pain_points_3}
+The pain point in Group 3 is
+\textbf{the lack of access to real-world datasets for testing.}
+
+We also summarize the potential and proven solutions as follows.
+
+\textit{Potential solutions from interviewees:}
+\begin{itemize}
+\item using open datasets
+\end{itemize}
+
+\textit{Proven solutions from interviewees:}
+\begin{itemize}
+\item asking the users to provide deidentified copies of medical images if they have problems loading the images;
+\item sending the beta versions of software to medical workers who can access the data and complete the tests;
+\item if (part of) the team belongs to a medical school or a hospital, using the datasets they can access to;
+\item if the team has access to MRI scanners, self-building MI datasets;
+\item if the team has connections with MI equipment manufacturers, asking for their help on data format problems.
+\end{itemize}
+
+No interviewee provided a perfect way to solve this problem. However, connections between the development team and medical professionals/institutions could ease the pain.
+
+\section{Documents in the Projects}
+By asking questions 11 and 19, we tried to understand the interviewees' opinions on documentation. We also aimed to find out the quality of documentations in their projects.
+
+Questions 11 and 19 are as follows,
+
+\textit{Q11. How does documentation fit into your development process? Would improved documentation help with the obstacles you typically face?}
+
+\textit{Q19. Do you think the current documentation can clearly convey all necessary knowledge to the users? If yes, how did you successfully achieve it? If no, what improvements are needed?}
+
+Table \ref{tab_opinion_doc} summarizes interviewees' opinions on documentation.
+Interviewees from each of the eight projects thought that documentation was important to their projects, and most of them said that it could save their time to answer questions from users and developers.
+However, most of them saw the need to improve their documentation, and only three of them thought that their documentations conveyed information clearly enough.
+
+\begin{table}[H]
+\centering
+\begin{tabular}{ll}
+\hline
+Opinion on documentation & Number of interviewees with the answer \\ \hline
+Documentation is vital to the project& 8 \\
+Documentation of the project needs improvements & 7 \\
+Referring to documentation saves time to answer questions & 6 \\
+Lack of time to maintain good documentation & 4 \\
+Documentation of the project conveys information clearly & 3 \\
+Coding is more preferable than documentation & 2 \\
+Users help each other by referring to documentation & 1 \\ \hline
+\end{tabular}
+\caption{\label{tab_opinion_doc}Opinions on documentation by the numbers of interviewees with the answers}
+\end{table}
+
+Table \ref{tab_doc_tools} lists some of the tools and methods mentioned by the interviewees, which they used for documentation.
+
+\begin{table}[H]
+\centering
+\begin{tabular}{ll}
+\hline
+Tool or method for documentation & Number of interviewees with the answer \\ \hline
+Forum discussions & 3 \\
+Videos & 3 \\
+GitHub & 2 \\
+Mediawiki / wiki pages & 2 \\
+Workshops & 2 \\
+Social media & 2 \\
+Writing books & 1 \\
+Google Form & 1 \\
+State management & 1 \\ \hline
+\end{tabular}
+\caption{\label{tab_doc_tools}Tools and methods for documentation by the numbers of interviewees with the answers}
+\end{table}
+
+\section{Contribution Management and Project Management}
+By asking questions 5, 13, and 14, we tried to understand how the teams managed the contributions and their projects.
+
+Questions 5, 13, and 14 are as follows,
+
+\textit{Q5. Do you have a defined process for accepting new contributions into your team?}
+
+\textit{Q13. What is your software development model? For example, waterfall, agile, etc.}
+
+\textit{Q14. What is your project management process? Do you think improving this process can tackle the current problem? Were any project management tools used?}
+
+Maybe some team had a documented process for accepting new contributions, but none talked about it during the interview.
+However, most of them mentioned using GitHub and pull requests to manage contributions from the community.
+The interviewees generally gave very positive feedback on using GitHub.
+Some also said they had handled the project repository with some other tools, and eventually transferred to git and GitHub.
+Table \ref{tab_method_receive_contributions} shows the number of times the interviewees mentioned the methods of receiving contributions.
+
+\begin{table}[H]
+\centering
+\begin{tabular}{lll}
+\hline
+\multirow{2}{*}{Method of receiving contributions} & \multicolumn{2}{l}{Number of interviewees with the answer} \\ \cline{2-3} 
+ & as a current method & as a past method \\ \hline
+GitHub with pull requests & 8 & \\
+Code contributions from emails & & 3 \\
+Code contributions from forums & & 1 \\
+Sharing the git repository by email & & 1 \\ \hline
+\end{tabular}
+\caption{\label{tab_method_receive_contributions}Methods of receiving contributions by the numbers of interviewees with the answers}
+\end{table}
+
+Additionally, the \textit{3D Slicer} team encouraged users to develop their extensions for specific use cases, and the \textit{OHIF} team was trying to enable the use of plug-ins; the interviewee from the \textit{ITK-SNAP} team said one way of accepting new team members was through funded academic projects.
+
+Table \ref{tab_developmen_models} shows the software development models by the numbers of interviewees with the answers.
+Only two interviewees confirmed their development models.
+The others did not think they used a specific model, but three of them suggested that their processes were similar to Waterfall or
+Agile.
+
+\begin{table}[H]
+\centering
+\begin{tabular}{ll}
+\hline
+Software development model & Number of interviewees with the answer \\ \hline
+Undefined/self-directed & 3 \\
+Similar to Agile & 2 \\
+Similar to Waterfall & 1 \\
+Agile & 1 \\
+Waterfall & 1 \\ \hline
+\end{tabular}
+\caption{\label{tab_developmen_models}Software development models by the numbers of interviewees with the answers}
+\end{table}
+
+Some interviewees mentioned the project management tools they used, which are in Table \ref{tab_pm_tools}. Generally speaking, they talked about two types:
 
 \begin{itemize}
-\item Start with one by one, with commonalities and interesting special cases.
-
-\item Shorten and summarize later.
+\item Trackers, including GitHub, issue trackers, bug trackers and Jira;
+\item Documents, including GitHub, Wiki page, Google Doc, and Confluence.
 \end{itemize}
 
-\section{Discussions}
+\begin{table}[H]
+\centering
+\begin{tabular}{ll}
+\hline
+Project management tools & Number of interviewees with the answer \\ \hline
+GitHub & 3 \\
+Issue trackers & 1 \\
+Bug trackers & 1 \\
+Jira & 1 \\
+Wiki page & 1 \\
+Google Doc & 1 \\
+Confluence & 1 \\ \hline
+\end{tabular}
+\caption{\label{tab_pm_tools}Project management tools by the numbers of interviewees with the answers}
+\end{table}
 
-Any conclusions?
+No interviewee introduced any strictly defined project management process.
+The most common way was following the issues, such as bugs and feature requests.
+Additionally, the \textit{3D Slicer} team had weekly meetings to discuss the goals for the project; the \textit{INVESALIUS 3} team relied on the GitHub process for their project management; the \textit{ITK-SNAP} team had a fixed six-month release pace; only the interviewee from the \textit{OHIF} team mentioned that the team has a project manager; the \textit{3D Slicer} team and \textit{BioImage Suite Web} team were doing nightly builds and tests.
+
+Most interviewees skipped the question ``Do you think improving this process can tackle the current problem?", but the interviewee from the \textit{OHIF} team gave a positive answer.

### 2021-07-13 StateOfPractice/Ao-Report/33InterviewAnswers.tex
ao report - Ch 5 & Appd C

@@
-\chapter{Interview Answers}
+\chapter{Other Interview Answers}
 \label{ap_interview}
 
-appendix here
+We asked 20 interview questions to the nine interviewees from eight software projects.
+We discuss the answers to interview questions 5, 9, 10, 11, 12, 13, 14, and 19 in Section \ref{ch_interview}, and summarize the answers to the other questions in this section.
+
+\noindent\textbf{Q1. Interviewees’ current position/title? degrees?}
+
+Six of the nine interviewees revealed their position/title, such as CEO of a company, endowed chair and professor in universities, software engineers in a commercial company and a hospital.
+
+Most of them answered their backgrounds and degrees.
+Table \ref{tab_q1_degrees} shows the highest academic degrees the participants have, and Table \ref{tab_q1_majors} shows what majors they studied.
+Many of the interviewees studied in multiple majors.
+
+\begin{table}[H]
+\centering
+\begin{tabular}{ll}
+\hline
+Highest degree & Number of interviewees \\ \hline
+PHD & 4 \\
+Master & 3 \\
+Bachelor & 0 \\
+Unspecified academic degree & 2 \\ \hline
+\end{tabular}
+\caption{\label{tab_q1_degrees}Interviewees' highest academic degrees}
+\end{table}
+
+\begin{table}[H]
+\centering
+\begin{tabular}{ll}
+\hline
+Major & Number of interviewees \\ \hline
+Computer Science & 4 \\
+Physics & 2 \\
+Biomedical Engineering & 1 \\
+Neuroimaging & 1 \\
+Geology (image analysis) & 1 \\
+Media Arts and Sciences & 1 \\
+Mechanical Engineering & 1 \\
+Materials Engineering & 1 \\
+Psychology & 1 \\ \hline
+\end{tabular}
+\caption{\label{tab_q1_majors}Interviewees' majors at university}
+\end{table}
+
+\noindent\textbf{Q2. Interviewees’ contribution to/relationship with the software?}
+
+Table \ref{tab_q2_roles} shows the interviewees’ roles and responsibilities in the projects.
+One of the participants did not explicitly mention his role, but implicitly revealed that he was a primary contributor to the project.
+
+\begin{table}[H]
+\centering
+\begin{tabular}{ll}
+\hline
+Role in the projects & Number of interviewees \\ \hline
+Chief Architect & 2 \\
+Lead Developer & 1 \\
+Core Developer & 5 \\
+Unspecified & 1 \\ \hline
+\end{tabular}
+\caption{\label{tab_q2_roles}Interviewees' roles in the projects}
+\end{table}
+
+\noindent\textbf{Q3. Length of time the interviewee has been involved with this software?}
+
+Table \ref{tab_q3_years} shows the distribution of the lengths of time the interviewees had worked on the projects.
+
+\begin{table}[H]
+\centering
+\begin{tabular}{ll}
+\hline
+Length of time in the projects & Number of interviewees \\ \hline
+0-1 & 1 \\
+2-5 & 0 \\
+6-10 & 2 \\
+11-15 & 3 \\
+16-20 & 2 \\
+21-25 & 1 \\ \hline
+\end{tabular}
+\caption{\label{tab_q3_years}Lengths of time that the interviewees worked in the projects}
+\end{table}
+
+\noindent\textbf{Q4. How large is the development group?}
+
+The size of each group grows and shrinks over the years.
+Most teams mentioned that the team members join and leave.
+Some teams said that when there was sufficient funding, they could afford more developers.
+
+Table \ref{tab_q4_members} shows the numbers of active members at the time of interviews. The members include people working on development and project management.
+
+\begin{table}[H]
+\centering
+\begin{tabular}{ll}
+\hline
+Number of current members & Number of projects \\ \hline
+1-3 & 5 \\
+4-6 & 3 \\ \hline
+\end{tabular}
+\caption{\label{tab_q4_members}Numbers of current members in the projects}
+\end{table}
+
+As shown in the table, no team had a vast number of members.
+Some projects had more developers, such as \textit{3D Slicer}; on the other hand, some teams such as \textit{dwv} had only one primary developer, plus a maximum of two or three developers occasionally.
+
+\textit{3D Slicer} is a special case, because it supports third-party extensions.
+So there have been community members developing and maintaining these extensions.
+Table \ref{tab_q4_members} does not include these members.
+
+\noindent\textbf{Q6. What is the typical background of a developer?}
+
+Not all interviewees could clearly answer this question.
+Many of them talked about the backgrounds of members with who they were familiar.
+Table \ref{tab_q6_dev_backgrounds} shows the number of times all interviewees mentioned a background.
+
+\begin{table}[H]
+\centering
+\begin{tabular}{ll}
+\hline
+Background of a developer & Number of interviewees with the answer \\ \hline
+\begin{tabular}[c]{@{}l@{}}Computer Science, Information\\ Technology, and Software Development \end{tabular} & 6 \\
+Imaging & 2 \\
+Medical Imaging & 1 \\
+Mathematics & 1 \\
+Biomedical Engineering & 1 \\
+Computer Aided Medical Procedures & 1 \\
+Physician & 1 \\ \hline
+\end{tabular}
+\caption{\label{tab_q6_dev_backgrounds}Backgrounds of developers by the numbers of interviewees with the answers}
+\end{table}
+
+\noindent\textbf{Q7. What is your estimated number of users? How did you come up with that estimate?}
+
+None of the interviewees knew the exact number of users. Some of them provided estimations based on different facts. However, we do not think these numbers are comparable to each other.
+
+\begin{table}[H]
+\centering
+\begin{tabular}{lll}
+\hline
+Software & Rough estimation & Considered facts \\ \hline
+3D Slicer & 100,000 & \begin{tabular}[c]{@{}l@{}} Search results on Google Scholar;\\ number of new posts per year on slicer.org; \\ number of downloads. \end{tabular} \\
+INVESALIUS 3 & 75,000 & Number of random IDs created by new installation.\\
+dwv & No estimation & About 20 companies integrated \textit{dwv} in their products. \\
+BioImage Suite Web & 100 active users & \begin{tabular}[c]{@{}l@{}}  The  interviewee only counted the users from \\ several Universities who were active users. \end{tabular} \\
+ITK-SNAP & 10,000 plus & Number of downloads. \\
+MRIcroGL & No estimation & \begin{tabular}[c]{@{}l@{}}  It is the top 1 downloaded software on this NITRC list \\ \hyperlink{https://www.nitrc.org/top/toplist.php?type=downloads}{https://www.nitrc.org/top/toplist.php?type=downloads} \end{tabular} \\
+Weasis & \begin{tabular}[c]{@{}l@{}}  10,000 user used \\ it as least once \end{tabular} & Number of profiles. \\
+OHIF & About 5000 & \begin{tabular}[c]{@{}l@{}}   Some platforms integrated \textit{OHIF}, and it was hard \\ to know the number of end users. \end{tabular} \\ \hline
+\end{tabular}
+\caption{\label{tab_q7_num_users}Rough Estimations for the Number of Users}
+\end{table}
+
+Table \ref{tab_q7_num_users} shows the estimations and how the interviewees made them.
+It is clear that some estimated only the active users, and some counted users who had used only once.
+So we do not compare these numbers.
+
+\noindent\textbf{Q8. What is the typical background of a user?}
+
+All interviewees provided several different user backgrounds, and all of them mentioned medical researchers or medical professionals.
+Table \ref{tab_q8_user_backgrounds} shows the number of times all interviewees mentioned a background.
+
+\begin{table}[H]
+\centering
+\begin{tabular}{ll}
+\hline
+Background of a user & Number of interviewees with the answer \\ \hline
+Medical Researchers & 6 \\
+Doctors/Health care professionals/Surgeons & 5 \\
+Student Researchers & 4 \\
+Patients & 3 \\
+Paleontologist & 1 \\
+Biomechanical Engineers & 1 \\
+Imaging Researchers & 1 \\
+Mechanical Engineers & 1 \\ \hline
+\end{tabular}
+\caption{\label{tab_q8_user_backgrounds}Backgrounds of users by the numbers of interviewees with the answers}
+\end{table}
+
+\noindent\textbf{Q15.
+Was it hard to ensure the correctness of the software?
+If there were any obstacles, what methods have been considered or practiced to improve the situation?
+If practiced, did it work?}
+
+Table \ref{tab_q15_threats_correctness} shows the threats to \textit{correctness} by the numbers of interviewees with the answers.
+
+\begin{table}[H]
+\centering
+\begin{tabular}{ll}
+\hline
+Threat to correctness & Number of interviewees with the answer \\ \hline
+\begin{tabular}[c]{@{}l@{}}Clinical systems produce data in various formats \\ (e.g. DICOM), which have complicated standards.\\ The software needs to handle the complexity.\end{tabular} & 2 \\
+\begin{tabular}[c]{@{}l@{}}Different types of MI machines can create data \\ in slightly different ways, adding more complexity\\ for the software to handle.\end{tabular} & 2 \\
+\begin{tabular}[c]{@{}l@{}}Besides viewing, the software has additional \\ functions, which lead to extra complexity.\end{tabular} & 1 \\
+There is lack of real word image data for testing. & 1 \\
+\begin{tabular}[c]{@{}l@{}}The team cannot use private data for debugging,\\ even when the data cause problems.\end{tabular} & 1 \\
+\begin{tabular}[c]{@{}l@{}}The software uses huge datasets for testing, so \\ the tests are expensive and time-consuming.\end{tabular} & 1 \\
+It is hard to well manage releases. & 1 \\
+The project has no unit tests. & 1 \\
+The project has no dedicated quality assurance team. & 1 \\ \hline
+\end{tabular}
+\caption{\label{tab_q15_threats_correctness}Threats to correctness by the numbers of interviewees with the answers}
+\end{table}
+
+Table \ref{tab_q15_strategies_correctness} shows the strategies to ensure \textit{correctness} by the numbers of interviewees with the answers.
+The interviewees from the \textit{3D Slicer} and \textit{ITK-SNAP} teams thought that the self-tests and automated tests were beneficial and could significantly save time.
+The interviewee from the \textit{Weasis} team kept collecting medical images for more than ten years.
+These images have caused problems with the software.
+So he had samples to test specific problems.
+
+\begin{table}[H]
+\centering
+\begin{tabular}{ll}
+\hline
+Strategy to ensure correctness & Number of interviewees with the answer \\ \hline
+\begin{tabular}[c]{@{}l@{}}Test-driven development / component tests /\\ integration tests / smoke tests / regression tests.\end{tabular} & 4 \\
+Self tests / automated tests. & 3 \\
+\begin{tabular}[c]{@{}l@{}}Two stage development process / stable release \&\\ nightly builds.\end{tabular} & 3 \\
+CI/CD. & 1 \\
+\begin{tabular}[c]{@{}l@{}}Using deidentified copies of medical images for\\ debugging.\end{tabular} & 1 \\
+\begin{tabular}[c]{@{}l@{}}Sending the beta versions of software to medical\\ workers who can access the data and do the tests.\end{tabular} & 1 \\
+\begin{tabular}[c]{@{}l@{}}Collecting and maintaining a dataset of\\ problematic images.\end{tabular} & 1 \\ \hline
+\end{tabular}
+\caption{\label{tab_q15_strategies_correctness}Strategies to ensure correctness by the numbers of interviewees with the answers}
+\end{table}
+
+\noindent\textbf{Q16.
+When designing the software, did you consider the ease of future changes?
+For example, will it be hard to change the system’s structure, modules, or code blocks?
+What measures have been taken to ensure the ease of future changes and maintains?}
+
+Table \ref{tab_q16_strategies_maintainability} shows the strategies to ensure \textit{maintainability} by the numbers of interviewees with the answers.
+The modular approach is the most talked-about solution to improve \textit{maintainability}.
+The \textit{3D Slicer} team used a well-defined structure for the software, which they named as ``event-driven MVC pattern''.
+Moreover, \textit{3D Slicer} discovers and loads necessary modules at runtime, according to the configuration and installed extensions.
+The \textit{BioImage Suite Web} team had designed and re-designed their software multiple times in the last 10+
+years. They found that their modular approach effectively supported the maintainability \cite{Joshi2011}.
+
+\begin{table}[H]
+\centering
+\begin{tabular}{ll}
+\hline
+Strategy to ensure maintainability & Number of interviewees with the answer \\ \hline
+\begin{tabular}[c]{@{}l@{}}Modular approach / thinking the software\\ as reusable Lego bricks / maintain \\ repetitive functions as libraries.\end{tabular} & 5 \\
+Supporting third-party extensions. & 1 \\
+Easy-to-understand architecture. & 1 \\
+Dedicated architect. & 1 \\
+Starting from simple solutions. & 1 \\
+Documentation. & 1 \\ \hline
+\end{tabular}
+\caption{\label{tab_q16_strategies_maintainability}Strategies to ensure maintainability by the numbers of interviewees with the answers}
+\end{table}
+
+\noindent\textbf{Q17. Provide instances where users have misunderstood the software. What, if any, actions were taken to address understandability issues?}
+
+Table \ref{tab_q17_threats_understandability} shows the threats to \textit{understandability} by the numbers of interviewees with the answers.
+It separates \textit{understandability} issues to users and developers by the horizontal dash line.
+
+\begin{table}[H]
+\centering
+\begin{tabular}{ll}
+\hline
+Threat to understandability & Number of interviewees with the answer \\ \hline
+\begin{tabular}[c]{@{}l@{}}Not all users understand how to use some\\ features of the software.\end{tabular} & 2 \\
+\begin{tabular}[c]{@{}l@{}}The team has no dedicated user experience\\ (UX) designer.\end{tabular} & 1 \\
+\begin{tabular}[c]{@{}l@{}}The software does not make some important\\ indicators noticeable (e.g. a progress bar).\end{tabular} & 1 \\
+\begin{tabular}[c]{@{}l@{}}Not all users understand the purpose of the\\ software.\end{tabular} & 1 \\
+\begin{tabular}[c]{@{}l@{}}Not all users know if the software includes\\ certain features.\end{tabular} & 1 \\
+\begin{tabular}[c]{@{}l@{}}Not all users understand how to use the\\ command line tool.\end{tabular} & 1 \\
+\begin{tabular}[c]{@{}l@{}}Not all users understand that the software is \\a web application.\end{tabular} & 1 \\\hdashline
+\begin{tabular}[c]{@{}l@{}}Not all developers understand how to deploy\\ the software.\end{tabular} & 1 \\
+\begin{tabular}[c]{@{}l@{}}The architecture is difficult for new\\ developers to understand.\end{tabular} & 1 \\ \hline
+\end{tabular}
+\caption{\label{tab_q17_threats_understandability}Threats to understandability by the numbers of interviewees with the answers}
+\end{table}
+
+Table \ref{tab_q17_strategies_understandability} shows the strategies to ensure \textit{understandability} by the numbers of interviewees with the answers.
+
+\begin{table}[H]
+\centering
+\begin{tabular}{ll}
+\hline
+Strategy to ensure understandability & Number of interviewees with the answer \\ \hline
+\begin{tabular}[c]{@{}l@{}}Documentation / user manual /\\ user mailing list / forum.\end{tabular} & 4 \\
+Graphical user interface. & 2 \\
+Testing every release with active users. & 1 \\
+\begin{tabular}[c]{@{}l@{}}Making simple things simple and complicated\\ things possible.\end{tabular} & 1 \\
+Icons with more clear visual expressions. & 1 \\
+Designing the software to be intuitive. & 1 \\
+Having a UX designer with the right experience. & 1 \\
+Dialog windows for important notifications. & 1 \\
+\begin{tabular}[c]{@{}l@{}}Providing an example if the users need to build\\ the software by themselves.\end{tabular} & 1 \\ \hline
+\end{tabular}
+\caption{\label{tab_q17_strategies_understandability}Strategies to ensure understandability by the numbers of interviewees with the answers}
+\end{table}
+
+\noindent\textbf{Q18. What, if any, actions were taken to address usability issues?}
+
+Table \ref{tab_q18_strategies_usability } shows the strategies to ensure \textit{usability} by the numbers of interviewees with the answers.
+
+\begin{table}[H]
+\centering
+\begin{tabular}{ll}
+\hline
+Strategy to ensure usability & Number of interviewees with the answer \\ \hline
+Usability tests and interviews with end users. & 3 \\
+Adjusting according to users' feedbacks. & 3 \\
+\begin{tabular}[c]{@{}l@{}}Straightforward and intuitively designed interface /\\ professional UX designer.\end{tabular} & 2 \\
+\begin{tabular}[c]{@{}l@{}}Providing step-by-step processes, and showing\\ the step numbers.\end{tabular} & 1 \\
+\begin{tabular}[c]{@{}l@{}}Making the basic functions easy to use without\\ reading the documentation.\end{tabular} & 1 \\
+Focusing on limited number of functions. & 1 \\
+Making the software more streamlined. & 1 \\
+Downsampling images to consume less memory. & 1 \\
+\begin{tabular}[c]{@{}l@{}}An option to load only part of the data to boost\\ performance.\end{tabular} & 1 \\ \hline
+\end{tabular}
+\caption{\label{tab_q18_strategies_usability }Strategies to ensure usability by the numbers of interviewees with the answers}
+\end{table}
+
+\noindent\textbf{Q20. Do you have any concern that your computational results won’t be reproducible in the future? Have you taken any steps to ensure reproducibility?}
+
+Table \ref{tab_q20_threats_reproducibility} shows the threats to \textit{reproducibility} by the numbers of interviewees with the answers.
+
+\begin{table}[H]
+\centering
+\begin{tabular}{ll}
+\hline
+Threat to reproducibility & Number of interviewees with the answer \\ \hline
+\begin{tabular}[c]{@{}l@{}}If the software is closed-source, the reproducibility\\is hard to achieve.\end{tabular} & 1 \\
+The project has no user interaction tests. & 1 \\
+The project has no unit tests. & 1 \\
+\begin{tabular}[c]{@{}l@{}}Using different versions of some common libraries\\may cause problems.\end{tabular} & 1 \\
+CPU variability can leads to non-reproducibility. & 1 \\
+\begin{tabular}[c]{@{}l@{}}When reverse-engineering how manufacturers create\\medical images, the team may misinterpret it.\end{tabular} & 1 \\ \hline
+\end{tabular}
+\caption{\label{tab_q20_threats_reproducibility}Threats to reproducibility by the numbers of interviewees with the answers}
+\end{table}
+
+Table \ref{tab_q20_strategies_ reproducibility } shows the strategies to ensure \textit{ reproducibility} by the numbers of interviewees with the answers.
+The interviewee from the \textit{3D Slicer} team provided various suggestions.
+One interviewee from another team suggested that they used \textit{3D Slicer} as the benchmark to test their \textit{reproducibility}.
+
+\begin{table}[H]
+\centering
+\begin{tabular}{ll}
+\hline
+Strategy to ensure reproducibility & Number of interviewees with the answer \\ \hline
+Regression tests / unit tests / having good tests. & 6 \\
+\begin{tabular}[c]{@{}l@{}}Making code, data, and documentation available / making\\ the software open-source / using open-source libraries.\end{tabular} & 5 \\
+Running same tests on all platforms. & 1 \\
+\begin{tabular}[c]{@{}l@{}}A dockerized version of the software,  insulating it from\\ the operating system environment.\end{tabular} & 1 \\
+Using standard libraries. & 1 \\
+Monitoring the upgrades of the libraries. & 1 \\
+Clearly documenting the versions. & 1 \\
+\begin{tabular}[c]{@{}l@{}}Bringing along the exact versions of all the dependencies\\ with the software.\end{tabular} & 1 \\
+Providing checksums of the data. & 1 \\
+\begin{tabular}[c]{@{}l@{}}Benchmark the software against other software with\\ similar purposes.\end{tabular} & 1 \\ \hline
+\end{tabular}
+\caption{\label{tab_q20_strategies_ reproducibility }Strategies to ensure  reproducibility by the numbers of interviewees with the answers}
+\end{table}

### 2021-07-14 StateOfPractice/Ao-Notes/DomainExptMeeting/Presentation for the Introductory Meeting.tex
Add goal to ranking follo-up slide
Updates to domain expert meeting slides

@@
 \item There are 9 measurements
-\item The write up is about 10 pages long \ei \ei
+\item The write up is about 10 pages long
+\item Our goal is to find anything that seems out of place
+  \ei \ei
 
@@
 \begin{frame}
-\frametitle{Pain Points from Developer Interviews}
+\frametitle{Developer Interviews: Pain Points}
 
-\structure{Do these fit with your experience?}
-\bi
-\item Lack of time to 1) implement all requested new features 2) write and maintain good documents 3) review all contributions from the community
-\item It is difficult to find or keep the members with experience in both software development and medical imaging
+\be
+\item Lack of time to 1) implement all requested new features 2) write and
+  maintain good documents 3) review all contributions from the community
+\item It is difficult to find or keep the members with experience in both
+  software development and medical imaging
 \item Lack of funding for software development (especially for maintenance)
-\item No organizations to help with developing high quality software
+\item No organizations to help with developing high quality software \label{NoOrg}
 \item Difficulty to get test data
-\item Difficulty to achieve cross-platform compatibility: 1) native app - more time and work 2) web app w/o server - worse performance 3) web app w/ servers - more cost
-\ei
+\item Difficulty to achieve cross-platform compatibility: 1) native app - more
+  time and work 2) web app w/o server - worse performance 3) web app w/ servers
+  - more cost \ee
 
@@
 \item \structure{What is the typical background of a developer?}
-\item \structure{How many users would you software typically have?}
+\item \structure{How many users would your software typically have?}
 \item \structure{What is the typical background of a user?}

### 2021-07-14 StateOfPractice/Ao-Report/13Methods.tex
ao report - update ch 3

@@
 \label{sec_domain_selection}
-Our methods are generic, but our scope only included scientific domains due to the objective of our research, and thus we have only applied this method to scientific domains.
+Our methods are generic, but we have only applied them to scientific domains due to the objective of our research.
 
-When choosing a candidate domain, we prefer a domain with a large number of active OSS.
-The reason is that we aim to finalize a list of 30 software packages \cite{SmithEtAl2021} in the domain after screening out the ones not meeting our requirements, such as the ones without recent updates or specific functions.
-Thus the quantity of final-decision packages may not meet our initial expectation if we do not have enough software candidates.
-Another reason is that our grading method works much better for OSS.
-Moreover, we need the domain to have an active community developing and using SC software, making it easier to conduct interviews with developers.
+When choosing a candidate domain, we prefer one with a large number of active OSS.
+The reason is that we aim to finalize a list of 30 software packages \cite{SmithEtAl2021} after the screening step.
+For example, we remove the ones without recent updates or specific functions.
+Thus, we need enough software candidates from the beginning.
+Besides, we prefer OSS projects because our grading method works better on them.
+Moreover, we prefer domains with active communities developing and using SC software, so it is easier to invite enough developers for interviews.
 
-With an adequate number of software packages in a selected domain, we may still find the software products with various purposes and fall into different sub-categories.
-So we should ask one question to ourselves - do we prefer a group of software all providing similar functions and features, or do we aim to cross-compare several sub-sections within the same domain?
-With that answered, we can better determine a favored domain.
+We prefer 30 software packages providing similar functions or falling into different sub-categories depending on our research purpose.
+So the domain needs to have enough candidates in one sub-category or enough sub-categories to cross-compare.
 
-Another aspect to consider is the team carrying out the research, specifically the domain experts - if there is any - in the team.
-In our team, researchers in the software engineering field usually lead the projects, and experts working in scientific domains support them.
-Having domain experts in the group provides significant benefits in selecting software packages and designing interview questions.
+We also prefer domains in which our team has expertise.
+We invite domain experts to join and support our projects.
+They help us in many aspects, such as vetting the software list and interview questions.
 
@@
 \label{sec_identify_software_candidates}
-We can find candidate software in publications of the domain.
-Another source is to search various websites, such as \hyperlink{https://github.com/}{GitHub}, \hyperlink{https://swmath.org/}{swMATH} and the Google search results for software recommendation articles.
-Meanwhile, we should also include the suggested ones from domain experts \cite{SmithEtAl2021}.
+We start with finding candidate software in publications of the domain.
+Then, we search various websites, such as \hyperlink{https://github.com/}{GitHub}, \hyperlink{https://swmath.org/}{swMATH} and the Google search results for software recommendation articles.
+Meanwhile, we include the suggested ones from domain experts \cite{SmithEtAl2021}.
 
@@
 The only ``mandatory'' requirement is that the software must be OSS, as defined in Section \ref{sec_open_source_software}.
-Because to evaluate all aspects of some software qualities, the source code should be accessible.
+Because to evaluate all software qualities, our methods need to access the source code.
 
-The other factors to filter the list can be optional, and we should consider them according to the number of software candidates and the objectives of the research project.
+The other factors to filter the list are optional, and we consider them according to the number of software candidates and the objectives of the research project.
 
-One of the factors is the functions and purposes of the software.
-For example, we can choose a group of software with similar functions, so that the cross-comparison is between each individual of them.
-On the other hand, if our objective is to compare sub-categories in the domain, we should select candidates from each of the categories.
+One factor is the functions and purposes of the software.
+For example, we can choose a group of software with similar functions, so that the comparison is between the software; or we can cross-compare sub-categories in the domain, then we need to select candidates from each sub-category.
 
-The empirical measurement tools listed in Section \ref{sec_empirical_measurements} have limitations that we can only apply them on projects using \hyperlink{https://git-scm.com/}{Git} as the version control tool, so we prefer software with Git.
+The empirical measurement tools listed in Section \ref{sec_empirical_measurements} only work on projects using \hyperlink{https://git-scm.com/}{Git}, so we prefer software with Git.
 Some manual steps in empirical measurement depend on a few metrics of GitHub, which makes projects held on GitHub more favored \cite{SmithEtAl2021}.
@@
 Some of the OSS projects may experience a lack of recent maintenance.
-So we can eliminate packages without recent updates, unless they are still popular and highly recommended by the domain users \cite{SmithEtAl2021}.
+So we eliminate packages without recent updates, unless they are still popular and highly recommended by the domain users \cite{SmithEtAl2021}.
 
-Whether we set what standards to filter the packages, with domain experts in the team, we should value their opinions.
-For example, if a software package is not OSS and has no updates for a long while, the domain experts may still identify it as a popular and widely used product.
-In this case, perhaps it is valuable to keep this software on the list.
+With domain experts in the team, we value their opinions on the filtering process.
+For example, if a software package is not OSS and has no updates for a long while, but the domain experts identify it as a valuable product, we still consider it to keep.
 
@@
 
-We use the first section of the template to collect some general information about the software, such as the name, purpose, platform, programming language, publications about the software, the first release and the most recent change date, website, and source code repository of the product, etc.
-Information in this section helps us understand the projects better and may be helpful for further analysis, but it does not directly affect the grading scores for the packages.
+We use the first section of the template to collect general information, such as the name, purpose, platform, programming language, publications about the software, the first release and the most recent change date, website, and source code repository of the product, etc.
+Information in this section helps us understand the projects better and may be helpful for further analysis, but it does not directly affect the grading scores.
 
-We designed the following 9 sections in the template for the 9 software qualities mentioned in Section \ref{sec_software_quality}.
-For each quality, we ask several questions and the typical answers are among the choices of ``yes'', ``no'', ``n/a'', ``unclear'', a number, a string, a date, a set of strings, etc.
+We designed the following nine sections in the template for the nine software qualities mentioned in Section \ref{sec_software_quality}.
+For each quality, we ask several questions and the typical answers are among the collection of ``yes'', ``no'', ``n/a'', ``unclear'', a number, a string, a date, a set of strings, etc.
 Each quality needs an overall score between 1 and 10 based on all the previous questions.
-For some qualities, the empirical measurement sections also affect this grading score.
+For some qualities, the empirical measurements also affect the score.
 
-All the last three sections on the template are about the empirical measurements.
-We use two command-line software tools \textit{GitStats} and \textit{scc} to extract information about the source code from the project repositories.
+All the last three sections are about the empirical measurements.
+We use two command-line software tools \textit{GitStats} and \textit{scc} to extract information from the source code repositories.
 For projects held on GitHub, we manually collect additional metrics, such as the stars of the GitHub repository, and the numbers of open and closed pull requests.
-Section \ref{sec_empirical_measurements} presents more details of how these empirical measurements affect software quality grading scores.
+Section \ref{sec_empirical_measurements} presents more details about the empirical measurements.
 
@@
 
-Both tools can measure the number of text-based files in a git repository and lines of text in these files.
-Based on our experience, most text-based files in a software project repository contain programming source code, and developers use them to compile and build software products.
+Both tools measure the number of text-based files in a git repository and lines of text in these files.
+Based on our experience, most text-based files in a repository contain programming source code, and developers use them to compile and build software products.
 A minority of these files are instructions and other documents.
 So we roughly regard the lines of text in text-based files as lines of programming code.
-The two tools usually generate similar but not identical results as for the above measurements.
-From our understanding, this minor difference is because of the different techniques that they use to detect if a file is a text-based or binary file.
+The two tools usually generate similar but not identical results.
+From our understanding, this minor difference is due to the different techniques to detect if a file is text-based or binary.
 
-Additionally, we also manually collect some information for projects held on GitHub, such as the numbers of stars, forks, people watching this repository, open pull request, and closed pull request.
+Additionally, we also manually collect information for projects held on GitHub, such as the numbers of stars, forks, people watching this repository, open pull requests, and closed pull requests.
 
 These empirical measurements help us from two aspects.
-Firstly, with more statistical details, we can get an overview of a project faster and more accurately.
-For example, the number of commits over the last 12 months shows how active this project has been during this period, and the number of stars and forks may reveal its popularity.
+Firstly, they help us with getting a project overview faster and more accurately.
+For example, the number of commits over the last 12 months shows how active this project has been, and the number of stars and forks may reveal its popularity.
 Secondly, the results may affect our decisions regarding the grading scores for some software qualities.
-For example, if the percentage of comment lines is low, we might want to double-check the understandability of the code.
-On the other hand, if the ratio of open versus closed pull requests is high, perhaps we should pay more attention to the project’s maintainability.
+For example, if the percentage of comment lines is low, we double-check the \textit{understandability} of the code; if the ratio of open versus closed pull requests is high, we pay more attention to the \textit{maintainability}.
 
@@
 \label{sec_technical_details}
-To test the software on a ``clean'' system, we created a new virtual machine (VM) for each software and only installed the necessary dependencies before measuring.
-We created All 29 VM on the same computer.
-We only start the measuring of the following software after finishing a current one, and after grading each software, we destroy the VM.
+To test the software on a ``clean'' system, we create a new virtual machine (VM) for each software and only install the necessary dependencies before measuring.
+We make all 30 VM on the same computer, one at a time, and destroy them after measuring.
 
-Generally speaking, we spend about two hours grading one software, unless there are technical issues and more time is needed to resolve them.
+We spend about two hours grading one package, unless we find technical issues and need more time to resolve them.
 In most of the situation, we finish all the measurements for one software on the same day.
@@
 \subsection{Interviewee Selection}
-For a software list with a length of roughly 30, we aim to interview about 10 development teams of these projects.
-Interviewing multiple individuals from each team should give us more comprehensive information.
-Still, if it is challenging to find various willing participants, a single engineer well-knowing the project can also be sufficient.
+For a software list with a length of roughly 30, we aim to interview about ten development teams.
+Interviewing multiple individuals from each team gives us more comprehensive information, but a single engineer well-knowing the project is also sufficient.
 
-Ideally, we select projects after the grading measurements and prefer the projects with higher overall scores.
-However, if we do not find enough participants, we should also contact all teams on the list.
+Ideally, we select projects after the grading measurements and prefer the ones with higher overall scores. However, if we do not find enough participants, we contact all teams on the list.
 
-We try to find the contacts of the teams on the websites related to the software, such as the official web pages, repository websites, publication websites, and bio pages of the teams' institutions.
-Then, we send at most two emails asking for their support and participation before receiving any replies for each candidate.
+We try to find the contacts of the teams on the projects' websites, such as the official web pages, repositories, publications, and bio pages of the teams' institutions.
+Then, we send at most two emails to one contact asking for its participation before receiving any replies.
 
@@
 
-We have a list of 20 questions to guide our interviews with the development teams, which can be found in Appendix \ref{ap_interview}.
+We have a list of 20 questions to guide our interviews, which can be found in Section \ref{ch_interview} and Appendix \ref{ap_interview}.
 
-Some questions are for the background of the software, the development teams, the interviewees, and how they organize the projects.
+Some questions are about the background of the software, the development teams, the interviewees, and how they organize the projects.
 We also ask about their understandings of the users.
-Another part of the interview questions focuses on the significant difficulties the team experience currently or in the past, and the solutions they have found or will try.
-We also discuss the importance of documents to their projects and the current situations of these documents with interviewees.
+Some questions focus on the current and past difficulties, and the solutions the team has found or will try.
+We also discuss the importance and current situations of documentation.
 A few questions are about specific software qualities, such as \textit{maintainability}, \textit{understandability}, \textit{usability}, and \textit{reproducibility}.
 
-The interviews are supposed to be semi-structured based on the question list, and we also ask follow-up questions when necessary.
-Based on our experience, the interviewees usually bring up some exciting ideas that we do not expect, and we can expand on these topics for a few more details.
+The interviews are semi-structured based on the question list, and we also ask follow-up questions when necessary.
+Based on our experience, the interviewees usually bring up some exciting ideas that we do not expect, and it is worth expanding on these topics.
 
 \subsection{Interview Process}
-Since the members of the development teams are likely to be based around the world, so we organize these interviews as virtual meetings online with \hyperlink{https://zoom.us/}{Zoom}.
+Since the members of the development teams are usually around the world, we organize these interviews as virtual meetings online with \hyperlink{https://zoom.us/}{Zoom}.
 After receiving consent from the interviewees, we also record our discussions to transcribe them better.
 
-\section{An Example of Applying the Method}
+\section{Applying the Method to Medical Imaging (MI)}
 \label{sec_applying_method}
 
-This section shows how we applied our method in Section \ref{ch_methods} on the Medical Imaging (MI) domain.
-
-Based on the principles in Section \ref{sec_domain_selection}, we selected the MI domain, because there are numerous software products and a significant number of professionals use and develop such software in this domain.
-We decided to focus on MI software with the viewing function, so we needed many software candidates in the domain.
-Being able to include MI domain experts in our team also made this domain more preferred.
+Based on the principles in Section \ref{sec_domain_selection}, we selected the MI domain and the viewer sub-category. We also included MI domain experts in our team.
 
@@
 
-Among the 48 software packages, there were several ones that we could not find their source code, such as MicroDicom, Aliza, and jivex, etc.
+Among the 48 software packages, there were several ones that we could not find in their source code, such as \textit{MicroDicom}, \textit{Aliza}, and \textit{jivex}, etc.
 These packages are likely to be freeware defined in Section \ref{sec_freeware} and not OSS.
@@
 Next, we focused on the MI software that could work as a MI viewer.
-Some of the software on the list were tool kits or libraries for other software to use as dependencies but not for end-users to view medical images, such as VTK, ITK, dcm4che, etc.
+Some of the software on the list were tool kits or libraries for other software to use as dependencies but not for end-users to view medical images, such as \textit{VTK}, \textit{ITK}, \textit{dcm4che}, etc.
 We also eliminated these from the list.

### 2021-07-14 StateOfPractice/Ao-Report/13Methods.tex
ao report - minor update

@@
 
-We prefer 30 software packages providing similar functions or falling into different sub-categories depending on our research purpose.
-So the domain needs to have enough candidates in one sub-category or enough sub-categories to cross-compare.
+We prefer 30 software packages providing similar functions or falling into different sub-groups depending on our research purpose.
+So the domain needs to have enough candidates in one sub-group or enough sub-groups to cross-compare.
 
@@
 One factor is the functions and purposes of the software.
-For example, we can choose a group of software with similar functions, so that the comparison is between the software; or we can cross-compare sub-categories in the domain, then we need to select candidates from each sub-category.
+For example, we can choose a group of software with similar functions, so that the comparison is between the software; or we can cross-compare sub-groups in the domain, then we need to select candidates from each sub-group.
 
@@
 
-Based on the principles in Section \ref{sec_domain_selection}, we selected the MI domain and the viewer sub-category. We also included MI domain experts in our team.
+Based on the principles in Section \ref{sec_domain_selection}, we selected the MI domain and the viewer sub-group. We also included MI domain experts in our team.
 

### 2021-07-14 StateOfPractice/Ao-Report/12Background.tex
ao report - update ch 2

@@
 
-This chapter introduces several different categories of software, based on which we designed the processes to select the software domain and software candidates in Chapter \ref{ch_methods}.
-It also covers the software quality definitions used in the grading template in Appendix \ref{ap_grading_template} and an overview of the Analytic Hierarchy Process (AHP), a tool we used to compare and grade software products.
+When designing a method for evaluating the state of the practice of domain-specific software, we included a step to select domain and software.
+Knowledge of different software categories is essential for the selections.
+To compare and grade the software qualities with the grading template in Appendix \ref{ap_grading_template}, we need the definitions of qualities and the Analytic Hierarchy Process (AHP).
 
@@
 We usually target specific software categories to narrow down the scopes when selecting software domains and software packages.
-In this section, we discuss three common software categories that are mentioned in Section \ref{sec_software_selection}, and also SC software.
-Section \ref{sec_domain_selection} and Section \ref{sec_software_selection} present more details about why we prefer some of these categories.
+In this section, we discuss three common software categories and also SC software.
 
@@
 
-The definitions of software qualities are from Smith et al. \cite{SmithEtAl2020}. The order of the qualities follows the grading template in Appendix \ref{ap_grading_template}.
+Our grading template in Appendix \ref{ap_grading_template} contain nine software qualities.
+This section lists the definitions of them, which are from Smith et al.
+\cite{SmithEtAl2020}.
+The order of the qualities follows the grading template.
 

### 2021-07-14 StateOfPractice/Ao-Report/12Background.tex
ao report - minor updates

@@
 When designing a method for evaluating the state of the practice of domain-specific software, we included a step to select domain and software.
-Knowledge of different software categories is essential for the selections.
+Knowledge of different software categories is essential for the selection.
 To compare and grade the software qualities with the grading template in Appendix \ref{ap_grading_template}, we need the definitions of qualities and the Analytic Hierarchy Process (AHP).
@@
 The fact that these products are free of charge is likely to make them popular with many users.
-However, software developers, end-users who wish to modify the source code, and researchers looking for inner characteristics may find the inaccessible source code to be a problem.
+However, software developers, end-users who wish to modify the source code, and researchers looking for inner characteristics may find the inaccessible source code a problem.
 

### 2021-07-14 StateOfPractice/Ao-Report/32FullSoftwareListBeforeFiltering.tex
ao report - update Appd B

\chapter{Full Software List Before Filtering}
\label{ap_list_before_filtering}

Table \ref{tab_list_before_filtering} lists the 48 software before filtering. We selected 29 of them to the final list, which are all open-source and in the same sub-group. We divided them into four sub-groups: Viewer (V), Tool Kit (TK), Picture Archiving and Communication System (PACS), and Convertion (C). The table also shows the sources of identifying them.

\begin{table}[]
\begin{tabular}{lllllllllll}
\hline
 &  &  &  & \multicolumn{7}{c}{Source} \\ \cline{5-11} 
\multirow{-2}{*}{Software} & \multirow{-2}{*}{\begin{tabular}[c]{@{}l@{}}Final\\ list\end{tabular}} & \multirow{-2}{*}{\begin{tabular}[c]{@{}l@{}}Open-\\ source\end{tabular}} & \multirow{-2}{*}{\begin{tabular}[c]{@{}l@{}}Sub-\\ group\end{tabular}} & \cite{Bjorn2017} & \cite{Bruhschwein2019} & \cite{Haak2015} & \cite{Emms2019} & \cite{Hasan2020} & \cite{Mu2019} & \cite{Samala2014} \\ \hline
3D Slicer & X & X & V &  &  & X & X & X &  & X \\
Ginkgo CADx & X & X & V & X & X & X & X & X & X &  \\
XMedCon & X & X & V &  &  & X & X & X & X &  \\
Weasis & X & X & V &  &  & X & X & X & X &  \\
MRIcroGL & X & X & V &  &  & X & X & X &  & X \\
SMILI & X & X & V &  &  &  & X & X & X &  \\
ImageJ & X & X & V &  &  & X & X & X &  & X \\
Fiji & X & X & V &  &  &  &  &  &  & X \\
DicomBrowser & X & X & V &  &  &  & X & X &  &  \\
3DimViewer & X & X & V & X &  &  & X & X &  &  \\
Horos & X & X & V & X & X &  &  &  &  &  \\
OsiriX Lite & X & X & V & X & X & X &  &  &  &  \\
dwv & X & X & V & X &  & X &  &  &  &  \\
Drishti & X & X & V & X &  &  &  &  &  &  \\
\begin{tabular}[c]{@{}l@{}}BioImage Suite\\ Web\end{tabular} & X & X & V &  &  & X &  &  &  &  \\
OHIF Viewer & X & X & V & \multicolumn{7}{l}{Other source: \cite{Ziegler2020}} \\
Slice:Drop & X & X & V &  &  & X &  &  &  &  \\
\end{tabular}
\end{table}

\begin{table}[]
\begin{tabular}{lllllllllll}
\hline
 &  &  &  & \multicolumn{7}{c}{Source} \\ \cline{5-11} 
\multirow{-2}{*}{Software} & \multirow{-2}{*}{\begin{tabular}[c]{@{}l@{}}Final\\ list\end{tabular}} & \multirow{-2}{*}{\begin{tabular}[c]{@{}l@{}}Open-\\ source\end{tabular}} & \multirow{-2}{*}{\begin{tabular}[c]{@{}l@{}}Sub-\\ group\end{tabular}} & \cite{Bjorn2017} & \cite{Bruhschwein2019} & \cite{Haak2015} & \cite{Emms2019} & \cite{Hasan2020} & \cite{Mu2019} & \cite{Samala2014} \\ \hline
GATE & X & X & V &  &  &  & X &  &  &  \\
ITK-SNAP & X & X & V &  &  &  &  &  &  & X \\
ParaView & X & X & V &  &  &  &  &  &  & X \\
MatrixUser & X & X & V &  &  &  &  & X &  &  \\
DICOM Viewer & X & X & V &  &  &  &  & X &  &  \\
INVESALIUS 3 & X & X & V &  &  &  &  & X &  &  \\
medInria & X & X & V &  &  & X &  &  &  &  \\
dicompyler & X & X & V &  &  &  &  &  & X &  \\
MicroView & X & X & V &  &  & X &  &  &  &  \\
Papaya & X & X & V &  &  &  & X & X & X &  \\
AMIDE & X & X & V &  &  &  & X & X & X &  \\
Gwyddion & X & X & V &  &  &  &  & X &  &  \\
VTK &  & X & TK &  &  &  &  &  &  & X \\
ITK &  & X & TK &  &  &  &  &  &  & X \\
DCMTK &  & X & TK &  &  &  &  & X & X &  \\
XTK &  & X & TK & \multicolumn{7}{l}{Other source: it is used by slice:drop} \\
dcm4che &  & X & TK & X &  &  & X & X &  &  \\
cornerstone &  & X & TK &  &  & X &  &  &  &  \\
\end{tabular}
\end{table}

\begin{table}[]
\begin{tabular}{lllllllllll}
\hline
 &  &  &  & \multicolumn{7}{c}{Source} \\ \cline{5-11} 
\multirow{-2}{*}{Software} & \multirow{-2}{*}{\begin{tabular}[c]{@{}l@{}}Final\\ list\end{tabular}} & \multirow{-2}{*}{\begin{tabular}[c]{@{}l@{}}Open-\\ source\end{tabular}} & \multirow{-2}{*}{\begin{tabular}[c]{@{}l@{}}Sub-\\ group\end{tabular}} & \cite{Bjorn2017} & \cite{Bruhschwein2019} & \cite{Haak2015} & \cite{Emms2019} & \cite{Hasan2020} & \cite{Mu2019} & \cite{Samala2014} \\ \hline
orthanc &  & X & PACS & X &  &  &  & X &  &  \\
Conquest &  & X & PACS & X &  &  &  &  &  &  \\
ClearCanvas &  & X & PACS, V & X &  &  &  &  &  &  \\
dcm2niix &  & X & C &  &  &  &  & X &  &  \\
\begin{tabular}[c]{@{}l@{}}Open Dicom\\ Viewer\end{tabular} &  & X & V &  &  & X &  &  & X &  \\
MicroDicom &  &  & V & X &  &  &  &  & X &  \\
Aliza &  &  & V &  &  &  & X &  & X &  \\
jivex &  &  & V &  &  & X &  &  & X &  \\
MIPAV &  &  & V &  &  & X &  &  &  &  \\
Oviyam &  &  & V &  &  & X &  &  &  &  \\
MeVisLab &  &  & V &  &  &  &  &  &  & X \\
\begin{tabular}[c]{@{}l@{}}Sante DICOM\\ Viewer Lite\end{tabular} &  &  & V &  &  &  &  &  & X &  \\
\begin{tabular}[c]{@{}l@{}}Navegatium\\ DICOM Viewer\end{tabular} &  &  & V &  &  &  &  &  & X &  \\ \hline
\end{tabular}
\caption{\label{tab_list_before_filtering}Full software list before filtering}
\end{table}

### 2021-07-14 StateOfPractice/Ao-Report/13Methods.tex
ao report - ch 3 minor update

@@
 
-Among the 48 software packages, there were several ones that we could not find in their source code, such as \textit{MicroDicom}, \textit{Aliza}, and \textit{jivex}, etc.
+Appendix \ref{ap_list_before_filtering} shows all the 48 software packages.
+Among them, there were several ones that we could not find in their source code, such as \textit{MicroDicom}, \textit{Aliza}, and \textit{jivex}, etc.
 These packages are likely to be freeware defined in Section \ref{sec_freeware} and not OSS.

### 2021-07-15 StateOfPractice/Ao-Report/32FullSoftwareListBeforeFiltering.tex
ao report - update Appd B

@@
 		\multirow{-2}{*}{Software} & \multirow{-2}{*}{\begin{tabular}[c]{@{}l@{}}Final\\ list\end{tabular}} & \multirow{-2}{*}{\begin{tabular}[c]{@{}l@{}}Open-\\ source\end{tabular}} & \multirow{-2}{*}{\begin{tabular}[c]{@{}l@{}}Sub-\\ group\end{tabular}} & \cite{Bjorn2017} & \cite{Bruhschwein2019} & \cite{Haak2015} & \cite{Emms2019} & \cite{Hasan2020} & \cite{Mu2019} & \cite{Samala2014} \\ \hline
-3D Slicer & X & X & V &  &  & X & X & X &  & X \\
-Ginkgo CADx & X & X & V & X & X & X & X & X & X &  \\
-XMedCon & X & X & V &  &  & X & X & X & X &  \\
-Weasis & X & X & V &  &  & X & X & X & X &  \\
-MRIcroGL & X & X & V &  &  & X & X & X &  & X \\
-SMILI & X & X & V &  &  &  & X & X & X &  \\
-ImageJ & X & X & V &  &  & X & X & X &  & X \\
-Fiji & X & X & V &  &  &  &  &  &  & X \\
-DicomBrowser & X & X & V &  &  &  & X & X &  &  \\
-3DimViewer & X & X & V & X &  &  & X & X &  &  \\
-Horos & X & X & V & X & X &  &  &  &  &  \\
-OsiriX Lite & X & X & V & X & X & X &  &  &  &  \\
-dwv & X & X & V & X &  & X &  &  &  &  \\
-Drishti & X & X & V & X &  &  &  &  &  &  \\
-\begin{tabular}[c]{@{}l@{}}BioImage Suite\\ Web\end{tabular} & X & X & V &  &  & X &  &  &  &  \\
-OHIF Viewer & X & X & V & \multicolumn{7}{l}{Other source: \cite{Ziegler2020}} \\
-Slice:Drop & X & X & V &  &  & X &  &  &  &  \\
+		3D Slicer \cite{Kikinis2014} & X & X & V &  &  & X & X & X &  & X \\
+		Ginkgo CADx \cite{Wollny2020} & X & X & V & X & X & X & X & X & X &  \\
+		XMedCon \cite{Nolf2003} & X & X & V &  &  & X & X & X & X &  \\
+		Weasis \cite{Roduit2021} & X & X & V &  &  & X & X & X & X &  \\
+		MRIcroGL \cite{Rorden2021} & X & X & V &  &  & X & X & X &  & X \\
+		SMILI \cite{Chandra2018} & X & X & V &  &  &  & X & X & X &  \\
+		ImageJ \cite{Rueden2017} & X & X & V &  &  & X & X & X &  & X \\
+		Fiji \cite{Schindelin2012} & X & X & V &  &  &  &  &  &  & X \\
+		DicomBrowser \cite{Archie2012} & X & X & V &  &  &  & X & X &  &  \\
+		3DimViewer \cite{TESCAN2020} & X & X & V & X &  &  & X & X &  &  \\
+		Horos \cite{horosproject2020} & X & X & V & X & X &  &  &  &  &  \\
+		OsiriX Lite \cite{PixmeoSARL2019} & X & X & V & X & X & X &  &  &  &  \\
+		dwv \cite{Martelli2021} & X & X & V & X &  & X &  &  &  &  \\
+		Drishti \cite{Limaye2012} & X & X & V & X &  &  &  &  &  &  \\
+		\begin{tabular}[c]{@{}l@{}}BioImage Suite\\ Web \cite{Papademetris2005} \end{tabular} & X & X & V &  &  & X &  &  &  &  \\
+		OHIF Viewer \cite{Ziegler2020} & X & X & V & \multicolumn{7}{l}{Other source: \cite{Ziegler2020}} \\
+		Slice:Drop \cite{Haehn2013} & X & X & V &  &  & X &  &  &  &  \\
 	\end{tabular}
@@
 		\multirow{-2}{*}{Software} & \multirow{-2}{*}{\begin{tabular}[c]{@{}l@{}}Final\\ list\end{tabular}} & \multirow{-2}{*}{\begin{tabular}[c]{@{}l@{}}Open-\\ source\end{tabular}} & \multirow{-2}{*}{\begin{tabular}[c]{@{}l@{}}Sub-\\ group\end{tabular}} & \cite{Bjorn2017} & \cite{Bruhschwein2019} & \cite{Haak2015} & \cite{Emms2019} & \cite{Hasan2020} & \cite{Mu2019} & \cite{Samala2014} \\ \hline
-GATE & X & X & V &  &  &  & X &  &  &  \\
-ITK-SNAP & X & X & V &  &  &  &  &  &  & X \\
-ParaView & X & X & V &  &  &  &  &  &  & X \\
-MatrixUser & X & X & V &  &  &  &  & X &  &  \\
-DICOM Viewer & X & X & V &  &  &  &  & X &  &  \\
-INVESALIUS 3 & X & X & V &  &  &  &  & X &  &  \\
-medInria & X & X & V &  &  & X &  &  &  &  \\
-dicompyler & X & X & V &  &  &  &  &  & X &  \\
-MicroView & X & X & V &  &  & X &  &  &  &  \\
-Papaya & X & X & V &  &  &  & X & X & X &  \\
-AMIDE & X & X & V &  &  &  & X & X & X &  \\
-Gwyddion & X & X & V &  &  &  &  & X &  &  \\
-VTK &  & X & TK &  &  &  &  &  &  & X \\
-ITK &  & X & TK &  &  &  &  &  &  & X \\
-DCMTK &  & X & TK &  &  &  &  & X & X &  \\
-XTK &  & X & TK & \multicolumn{7}{l}{Other source: it is used by slice:drop} \\
-dcm4che &  & X & TK & X &  &  & X & X &  &  \\
-cornerstone &  & X & TK &  &  & X &  &  &  &  \\
+		GATE \cite{Jan2004} & X & X & V &  &  &  & X &  &  &  \\
+		ITK-SNAP \cite{Yushkevich2006} & X & X & V &  &  &  &  &  &  & X \\
+		ParaView \cite{Ahrens2005} & X & X & V &  &  &  &  &  &  & X \\
+		MatrixUser \cite{Liu2016} & X & X & V &  &  &  &  & X &  &  \\
+		DICOM Viewer \cite{Afsar2021} & X & X & V &  &  &  &  & X &  &  \\
+		INVESALIUS 3 \cite{Amorim2015} & X & X & V &  &  &  &  & X &  &  \\
+		medInria \cite{Fillard2012} & X & X & V &  &  & X &  &  &  &  \\
+		dicompyler \cite{Panchal2010} & X & X & V &  &  &  &  &  & X &  \\
+		MicroView \cite{ParallaxInnovations2020} & X & X & V &  &  & X &  &  &  &  \\
+		Papaya \cite{UTHSCSA2019} & X & X & V &  &  &  & X & X & X &  \\
+		AMIDE \cite{Loening2017} & X & X & V &  &  &  & X & X & X &  \\
+		Gwyddion \cite{Nevcas2012} & X & X & V &  &  &  &  & X &  &  \\
+		VTK \cite{SchroederEtAl2006} &  & X & TK &  &  &  &  &  &  & X \\
+		ITK \cite{McCormick2014} &  & X & TK &  &  &  &  &  &  & X \\
+		DCMTK \cite{DCMTK2021} &  & X & TK &  &  &  &  & X & X &  \\
+		XTK \cite{Haehn2012} &  & X & TK & \multicolumn{7}{l}{Other source: it is used by slice:drop} \\
+		dcm4che \cite{dcm4che2021} &  & X & TK & X &  &  & X & X &  &  \\
+		cornerstone \cite{cornerstone2020} &  & X & TK &  &  & X &  &  &  &  \\
 	\end{tabular}
@@
 		\multirow{-2}{*}{Software} & \multirow{-2}{*}{\begin{tabular}[c]{@{}l@{}}Final\\ list\end{tabular}} & \multirow{-2}{*}{\begin{tabular}[c]{@{}l@{}}Open-\\ source\end{tabular}} & \multirow{-2}{*}{\begin{tabular}[c]{@{}l@{}}Sub-\\ group\end{tabular}} & \cite{Bjorn2017} & \cite{Bruhschwein2019} & \cite{Haak2015} & \cite{Emms2019} & \cite{Hasan2020} & \cite{Mu2019} & \cite{Samala2014} \\ \hline
-orthanc &  & X & PACS & X &  &  &  & X &  &  \\
-Conquest &  & X & PACS & X &  &  &  &  &  &  \\
-ClearCanvas &  & X & PACS, V & X &  &  &  &  &  &  \\
-dcm2niix &  & X & C &  &  &  &  & X &  &  \\
-\begin{tabular}[c]{@{}l@{}}Open Dicom\\ Viewer\end{tabular} &  & X & V &  &  & X &  &  & X &  \\
-MicroDicom &  &  & V & X &  &  &  &  & X &  \\
-Aliza &  &  & V &  &  &  & X &  & X &  \\
-jivex &  &  & V &  &  & X &  &  & X &  \\
-MIPAV &  &  & V &  &  & X &  &  &  &  \\
-Oviyam &  &  & V &  &  & X &  &  &  &  \\
-MeVisLab &  &  & V &  &  &  &  &  &  & X \\
-\begin{tabular}[c]{@{}l@{}}Sante DICOM\\ Viewer Lite\end{tabular} &  &  & V &  &  &  &  &  & X &  \\
+		orthanc \cite{Jodogne2018} &  & X & PACS & X &  &  &  & X &  &  \\
+		Conquest \cite{Herk2021} &  & X & PACS & X &  &  &  &  &  &  \\
+		ClearCanvas \cite{ClearCanvas2015} &  & X & PACS, V & X &  &  &  &  &  &  \\
+		dcm2niix \cite{Li2016} &  & X & C &  &  &  &  & X &  &  \\
+		\begin{tabular}[c]{@{}l@{}}Open Dicom\\ Viewer \cite{OpenDicomViewer2011} \end{tabular} &  & X & V &  &  & X &  &  & X &  \\
+		MicroDicom \cite{MicroDicom2021} &  &  & V & X &  &  &  &  & X &  \\
+		Aliza \cite{Aliza2021} &  &  & V &  &  &  & X &  & X &  \\
+		JiveX \cite{Visus2021} &  &  & V &  &  & X &  &  & X &  \\
+		MIPAV \cite{CIT2020} &  &  & V &  &  & X &  &  &  &  \\
+		Oviyam \cite{RasterImages2021} &  &  & V &  &  & X &  &  &  &  \\
+		MeVisLab \cite{Heckel2009} &  &  & V &  &  &  &  &  &  & X \\
+		\begin{tabular}[c]{@{}l@{}}Sante DICOM\\ Viewer Lite \cite{Santesoft2021} \end{tabular} &  &  & V &  &  &  &  &  & X &  \\
 		\begin{tabular}[c]{@{}l@{}}Navegatium\\ DICOM Viewer\end{tabular} &  &  & V &  &  &  &  &  & X &  \\ \hline

### 2021-07-19 StateOfPractice/Peter-Notes/Report/Report.tex
added note on Musubi LBM package to LBM Report
fixed note on Musubi LBM package to LBM Report
added second cover page LBM report

@@
 	{\copyright \ Copyright by \authorname, July 2021}
+
+\newpage
+\thispagestyle{empty}
+
+\noindent\textbf{TITLE:} State of The Practice for Lattice Boltzmann Method Software
+
+\noindent\textbf{DEGREE:} Master of Engineering (Computing and Software)
+
+\noindent\textbf{INSTITUTION:} McMaster University
+
+\noindent\textbf{AUTHOR:} Peter Michalski
+
+\noindent\textbf{SUPERVISOR:} Dr. Spencer Smith and Dr. Jacques Carette
+
+\noindent\textbf{NUMBER OF PAGES:} 119
 
@@
 We found LBM software listed on the websites GitHub and swMATH, as well as through articles found in scholarly journals and databases.
+Software packages that are not on the authoritative lists that we used were not assessed.
+Packages that fit our criteria but were found after data collection and analysis was conducted, such as Musubi, will need to be added to future SOP assessments.
 

### 2021-07-24 StateOfPractice/Ao-Report/13Methods.tex
MI report - updated ch 4

@@
 Each quality needs an overall score between 1 and 10 based on all the previous questions.
-For some qualities, the empirical measurements also affect the score.
+For some qualities, we perform surface measurements, which allow us to carry out on all packages with reasonable efforts.
+The surface measurements reveal some traits of a underlying quality, but may not fully represent it.
+
+\begin{itemize}
+\item \textbf{Installability} We check the existence and quality of installation instructions.
+The user experience is also an important factor, such as the ease to follow the instructions, number of steps, automation tools, and the prerequisite steps for the installation.
+If any problem interrupts the process of installation or uninstallation, we give a lower score to this quality.
+We also record the operating system for the installation test and whether we can verify the installation.
+
+\item \textbf{Correctness \& Verifiability} For \textit{correctness}, we check the projects to identify the techniques to ensure this quality, such as literate programming, automated testing, symbolic execution, model checking, unit tests, etc.
+We also examine whether the projects use continuous integration.
+For \textit{verifiability}, we go through the documents of the projects to check the requirements specifications, theory manuals, and getting started tutorials.
+If a getting started tutorial exists and provides expected results, we follow it and check if the outputs match.
+
+\item \textbf{Surface Reliability} We check that whether the software break during the installations and tutorials, whether there are descriptive error messages, and if we can recover the process after the errors.
+
+\item \textbf{Surface Robustness} We check that how the software handle unexpected/unanticipated input.
+For example, for software packages with the function to load image files, we prepare broken image files for them to open.
+We use a text file (.txt) with a modified extension name (.dcm) as an unexpected/unanticipated input.
+We load a few correct input files to ensure the function is working correctly before testing with the unexpected/unanticipated ones.
+
+\item \textbf{Surface Usability} We examine the documents of the projects and consider software with a getting started tutorial and a user manual easier to use.
+Meanwhile, we check if users have any channels to get supports.
+Our impressions and user experiences when testing the software also affect the scores.
+For example, easy-to-use graphical user interfaces give us a better experience, thus lead to better scores.
+
+\item \textbf{Maintainability} We search the projects' documents and identify the process of contributing and reviewing code.
+We believe that the artifacts of a project - including source code, documents, building scripts, etc.
+- can significantly affect its  \textit{maintainability}.
+Thus we check each project for its artifacts, such as API documentation, bug tracker, release notes, test cases, build files, version control, etc.
+We also check the tools supporting issue tracking and version control and the percentages of closed issues and comment lines in code.
+
+\item \textbf{reusability} We count the total number of code files for each project.
+Projects with a large number of components provide more choices to reuse.
+Furthermore, well modularized code, which tend to have smaller parts in separated files, are easier to reuse.
+Thus, we consider the projects with more code files to be more reusable.
+We use \textit{GitStats} as a tool to count the number of text-based files for all projects, and consider the projects with more text-based files to also have more code files.
+We also decide that the projects with API documentation can deliver better \textit{Reusability}.
+
+\item \textbf{Surface Understandability} We randomly examine 10 code files.
+We check the code’s style within each code file, such as whether the identifiers, parameters, indentation, and formatting are consistent, whether the constants (other than 0 and 1) are hardcoded into the code, and whether the developers modularized the code.
+We also check the descriptive information for the code, such as documents mentioning the coding standard, the comments in the code, and the descriptions or links for algorithms in the code.
+
+\item \textbf{Visibility/Transparency} To measure this quality, such as all of the steps of a software development process and the current status of a project, we check the existing documents.
+We examine the development process, current status, development environment, and release notes for each project.
+If any information is missing or poorly conveyed, the \textit{visibility/transparency} is not ideal.
+\end{itemize}
 
 All the last three sections are about the empirical measurements.
+For some qualities, the empirical measurements also affect the score.
 We use two command-line software tools \textit{GitStats} and \textit{scc} to extract information from the source code repositories.

### 2021-07-24 StateOfPractice/Ao-Report/14MeasurementResults.tex
MI report - updated ch 4

--- 2021-05-09 StateOfPractice/Ao-Report/14MeasurementResults.tex
@@
 
-For some qualities, it might be a good idea to cross-compare with the empirical scores.
+This section shows the summary of the measurement results. The detailed data can be found in the repository \hyperlink{https://github.com/smiths/AIMSS}{https://github.com/smiths/AIMSS}.
 
-\section{Selected Software List}
+Table \ref{tab_final_list} shows the 29 software packages that we measured.
+We tried to find the initial release dates in the documents, such as the introduction and release notes.
+We used the date of the latest change to each code repository to decide the latest update.
+We found out funding information for eight projects, and we are not sure about the rest.
+
+We considered anyone who made at least one accepted commit to the source code as a contributor.
+Thus, it does not mean that any development teams had long-term members as many as 100.
+Many of these projects received change requests and code from the community, such as pull requests and git commits on GitHub.
+
+Table \ref{tab_final_list} also shows the supported operating systems for each software package, and 25 of them could work on all three Windows (W), macOS (M), and Linux (L) systems.
+However, there was a significant difference in the philosophy to achieve cross-platform compatibility.
+Most of them were native software products, but five were naturally platform-independent web applications.
+
+\begin{table}[H]
+\begin{tabular}{lllllllll}
+\hline
+\multirow{2}{*}{Software} & \multirow{2}{*}{Released} & \multirow{2}{*}{Updated} & \multirow{2}{*}{Contributors} & \multirow{2}{*}{Funded} & \multicolumn{3}{c}{OS} & \multirow{2}{*}{Web} \\ \cline{6-8}
+ &  &  &  &  & W & M & L &  \\ \hline
+3D Slicer \cite{Kikinis2014} & 1998 & 2020-08 & 100 & X & X & X & X &  \\
+Ginkgo CADx \cite{Wollny2020} & 2010 & 2019-05 & 3 &  & X & X & X &  \\
+XMedCon \cite{Nolf2003} & 2000 & 2020-08 & 2 &  & X & X & X &  \\
+Weasis \cite{Roduit2021} & 2010 & 2020-08 & 8 &  & X & X & X &  \\
+MRIcroGL \cite{Rorden2021} & 2015 & 2020-08 & 2 &  & X & X & X &  \\
+SMILI \cite{Chandra2018} & 2014 & 2020-06 & 9 &  & X & X & X &  \\
+ImageJ \cite{Rueden2017} & 1997 & 2020-08 & 18 & X & X & X & X &  \\
+Fiji \cite{Schindelin2012} & 2011 & 2020-08 & 55 & X & X & X & X &  \\
+DicomBrowser \cite{Archie2012} & 2012 & 2020-08 & 3 &  & X & X & X &  \\
+3DimViewer \cite{TESCAN2020} & unclear & 2020-03 & 3 & X & X & X &  &  \\
+Horos \cite{horosproject2020} & unclear & 2020-04 & 21 &  &  & X &  &  \\
+OsiriX Lite \cite{PixmeoSARL2019} & 2004 & 2019-11 & 9 &  &  & X &  &  \\
+dwv \cite{Martelli2021} & 2012 & 2020-09 & 22 &  & X & X & X & X \\
+Drishti \cite{Limaye2012} & 2012 & 2020-08 & 1 &  & X & X & X &  \\
+\begin{tabular}[c]{@{}l@{}}BioImage Suite\\ Web \cite{Papademetris2005}\end{tabular} & 2018 & 2020-10 & 13 & X & X & X & X & X \\
+OHIF Viewer \cite{Ziegler2020} & 2015 & 2020-10 & 76 &  & X & X & X & X \\
+Slice:Drop \cite{Haehn2013} & 2012 & 2020-04 & 3 &  & X & X & X & X \\
+GATE \cite{Jan2004} & 2011 & 2020-10 & 45 &  &  & X & X &  \\
+ITK-SNAP \cite{Yushkevich2006} & 2006 & 2020-06 & 13 & X & X & X & X &  \\
+ParaView \cite{Ahrens2005} & 2002 & 2020-10 & 100 & X & X & X & X & X \\
+MatrixUser \cite{Liu2016} & 2013 & 2018-07 & 1 &  & X & X & X &  \\
+DICOM Viewer \cite{Afsar2021} & 2018 & 2020-04 & 5 & X & X & X & X &  \\
+INVESALIUS 3 \cite{Amorim2015} & 2009 & 2020-09 & 10 &  & X & X & X &  \\
+\end{tabular}
+\end{table}
+
+\begin{table}[H]
+\begin{tabular}{lllllllll}
+\hline
+\multirow{2}{*}{Software} & \multirow{2}{*}{Released} & \multirow{2}{*}{Updated} & \multirow{2}{*}{Contributors} & \multirow{2}{*}{Funded} & \multicolumn{3}{c}{OS} & \multirow{2}{*}{Web} \\ \cline{6-8}
+ &  &  &  &  & W & M & L &  \\ \hline
+medInria \cite{Fillard2012} & 2009 & 2020-11 & 21 &  & X & X & X &  \\
+dicompyler \cite{Panchal2010} & 2009 & 2020-01 & 2 &  & X & X &  &  \\
+MicroView \cite{ParallaxInnovations2020} & 2015 & 2020-08 & 2 &  & X & X & X &  \\
+Papaya \cite{UTHSCSA2019} & 2012 & 2019-05 & 9 &  & X & X & X &  \\
+AMIDE \cite{Loening2017} & 2006 & 2017-01 & 4 &  & X & X & X &  \\
+Gwyddion \cite{Nevcas2012} & 2004 & 2020-11 & 38 &  & X & X & X &  \\ \hline
+\end{tabular}
+\caption{\label{tab_final_list}Final software list}
+\end{table}
+
+Most of the projects used more than one programming language, including a primary language that the developers used the most.
+Figure \ref{fig_language} shows the primary languages versus the number of projects using them.
+
+\begin{figure}[H]
+\centering
+\begin{tikzpicture}
+\centering
+\begin{axis}
+[
+ybar,
+height=8cm,
+width=12cm,
+ylabel={Number of projects},
+xlabel={\ Primary language},
+symbolic x coords={C++, JavaScript, Java, C, Python, Pascal, Matlab},
+xtick=data,
+nodes near coords,
+nodes near coords align={vertical},
+]
+\addplot[black,fill=blue!55!black] coordinates {(C++,11) (JavaScript,6) (Java,4) (C,3) (Python,3) (Pascal,1) (Matlab,1) };
+\end{axis}  
+\end{tikzpicture}
+\caption{\label{fig_language}Primary languages versus number of projects using them}
+\end{figure}
+
+We failed installing \textit{DICOM Viewer}, so we could not test \textit{surface reliability} and \textit{surface robustness} for it.
+We might underestimate the scores of the two qualities for this software.
+We kept this software on our list because the other seven qualities do not rely on a successful installation.
+Besides, it used a unique dependency, and we wanted to keep the diversity.
 
 \section{Installability}
+\label{sec_result_installability}
+
+Figure \ref{fg_installability_scores} lists the scores of \textit{installability}.
+
+\begin{figure}[H]
+\includegraphics[scale=0.38]{figures/installability_scores.png}
+\caption{AHP installability scores}
+\label{fg_installability_scores}
+\end{figure}
+
+We found installation instructions for 16 projects.
+Among the ones without such an instruction, \textit{BioImage Suite Web} and \textit{Slice:Drop} were web applications and provided online versions to use, thus they needed no installation.
+Installing 10 of the projects needed extra dependencies.
+Five of them are the web applications in Table \ref{tab_final_list}, and depended on a browser; \textit{dwv}, \textit{OHIF Viewer}, and \textit{GATE} needed extra dependencies to build; \textit{ImageJ} and	\textit{Fiji} needed an unzip tool; \textit{MatrixUser} was based on Matlab; \textit{DICOM Viewer} needed to work on Nextcloud platform.
+
+\textit{3D Slicer} has the highest score because it had easy-to-follow installation instructions, and the installation processes were automated, fast, and frustration-free, with all dependencies automatically added.
+There were also no errors during the installation and uninstallation steps.
+Many other software packages also had installation instructions and automated installers, and we had no trouble installing them, such as \textit{INVESALIUS 3}, \textit{Gwyddion}, \textit{XMedCon}, and \textit{MicroView}.
+We gave them various scores based on the understandability of the instructions, installation steps, and user experience.
+Since \textit{BioImage Suite Web} and \textit{Slice:Drop} needed no installation, we gave them higher scores due to the significant convenience.
+\textit{BioImage Suite Web} also provided an option to download cache to local for offline usage, which was easy to apply.
+
+\textit{dwv}, \textit{GATE}, and \textit{DICOM Viewer} showed severe problems.
+We could not install them due to some issues that we could not solve.
+We spent a reasonable amount of time on these problems, then considered them major obstacles for normal users if we still did not figure out any solutions.
+We suspect that only a part of the users faced the same problems, and given a lot of time, we might be able to find solutions.
+However, the difficulties greatly impacted the installation experiences, and we graded these software packages with lower scores.
+For example, \textit{dwv} and \textit{GATE} had the option to build from the source code, and we failed the building processes following the instructions.
+Although we could not locally build them, we could use a deployed online version for \textit{dwv}, and a VM version for \textit{GATE}.
+With those, we finished all the measurements for them.
+Furthermore, \textit{DICOM Viewer} depended on a cloud platform, and we could not successfully install the dependency.
+
+\textit{MatrixUser} has a lower score because it depended on Matlab. We considered installing Matlab takes many more steps and time, and some users may not have a license to use Matlab.
 
 \section{Correctness \& Verifiability}
+The scores of \textit{correctness \& verifiability} are shown in Figure \ref{fg_correctness_erifiability_scores}.
+Generally speaking, the packages with higher scores adopted more techniques to improve \textit{correctness}, and had better documents for us to verify it.
 
-\section{Reliability}
+\begin{figure}[H]
+\includegraphics[scale=0.38]{figures/correctness_verifiability_scores.png}
+\caption{AHP correctness \& verifiability scores}
+\label{fg_correctness_erifiability_scores}
+\end{figure}
 
-\section{Robustness}
+After examining the source code, we could not find any evidence of unit testing in more than half of the projects.
+Unit testing benefits most parts of the software's life cycle, such as designing, coding, debugging, and optimization \cite{Hamill2004}.
+It can reveal the bugs at an earlier stage of the development process, and the absence of unit testing may cause worse \textit{correctness \& verifiability}.
 
-\section{Usability}
+We could not find requirements specifications for most projects.
+The only document we found is a road map of \textit{3D Slicer}, which contained design requirements for the following changes.
+However, it did not record the conditions for previous versions.
+We also could not identify the theory manuals for all of the projects.
+It seems that even for some projects with well-organized documents, requirements specifications and theory manuals were still missing.
+
+We identified five projects using continuous integration and continuous delivery (CI/CD) tools, which are \textit{3D Slicer}, \textit{ImageJ
+}, \textit{Fiji}, \textit{dwv}, and \textit{OHIF Viewer}.
+
+\section{Surface Reliability}
+As described in Section \ref{sec_result_installability}, we could not build \textit{dwv} and \textit{GATE}.
+However, since there was an online or VM version of them, successful deployment is possible.
+So the failure of installation did not affect their scores in \textit{surface reliability}.
+Figure \ref{fg_reliability_scores} shows the AHP results.
+
+\begin{figure}[H]
+\includegraphics[scale=0.38]{figures/reliability_scores.png}
+\caption{AHP surface reliability scores}
+\label{fg_reliability_scores}
+\end{figure}
+
+When applying basic operations with the software packages, we found out that \textit{Drishti} crashed during loading damaged image files, and \textit{GATE} could not open macro files and lost response several times.
+
+\section{Surface Robustness}
+Figure \ref{fg_robustness_scores} presents the scores for \textit{surface robustness}.
+
+\begin{figure}[H]
+\includegraphics[scale=0.38]{figures/robustness_scores.png}
+\caption{AHP surface robustness scores}
+\label{fg_robustness_scores}
+\end{figure}
+
+The packages with higher scores elegantly handled the unexpected/unanticipated inputs, normally showing a clear error message.
+We might underestimate the score of \textit{OHIF Viewer} since we needed further customization to load data, and the test was not complete.
+\textit{MatrixUser}, \textit{dwv}, \textit{Slice:Drop}, and \textit{MRIcroGL} ignored the incorrect format of the input files, and displayed blank or meaningless images.
+\textit{Drishti} successfully detected the unexpected/unanticipated inputs, but the software crashed as a result.
+For unknown reasons, \textit{GATE} failed to load both correct and incorrect inputs.
+
+\section{Surface Usability}
+Figure \ref{fg_usability_scores} lists the AHP scores for \textit{surface usability}.
+
+\begin{figure}[H]
+\includegraphics[scale=0.38]{figures/usability_scores.png}
+\caption{AHP surface usability scores}
+\label{fg_usability_scores}
+\end{figure}
+
+We found a getting started tutorial for only 11 projects but a user manual for 22 projects. \textit{MRIcroGL} was the only one with expected user characteristics documented.
+
+The ones with higher scores usually provided both comprehensive document guidance and a good user experience.
+\textit{INVESALIUS 3} set an excellent example of a detailed and precise user manual.
+\textit{GATE} also provided a large number of documents, but we think that they conveyed the ideas poorly, as we had trouble understanding and using them.
+
+Table \ref{tab_user_support_model} shows the user support models by the number of projects using them.
+Maybe not every team intended to use GitHub issues to answer users' questions, but many users use them to seek help.
+
+\begin{table}[H]
+\centering
+\begin{tabular}{ll}
+\hline
+\multicolumn{1}{c}{User support model} & Number of projects \\ \hline
+GitHub issue & 24 \\
+GitLab issue, SourceForge discussions & 2 \\
+FAQ & 12 \\
+Forum & 10 \\
+E-mail address & 9 \\
+Troubleshooting & 2 \\
+Contact form & 1 \\ \hline
+\end{tabular}
+\caption{\label{tab_user_support_model}User support models by number of projects}
+\end{table}
 
 \section{Maintainability}
+Figure \ref{fg_maintainability_scores} shows the AHP results for \textit{maintainability}.
+
+\begin{figure}[H]
+\includegraphics[scale=0.38]{figures/maintainability_scores.png}
+\caption{AHP maintainability scores}
+\label{fg_maintainability_scores}
+\end{figure}
+
+We marked \textit{3D Slicer} with a much higher score than others because it did very well at closing the identified issues, and more importantly, we found it to have the most comprehensive artifacts.
+For example, as far as we could find out, only a few of the 29 projects had a project plan, developer's manual, or API documentation, and only \textit{3D Slicer}, \textit{ImageJ}, \textit{Fiji} included all three documents.
+Meanwhile, \textit{3D Slicer} has a much higher percentage of closed issues (91.65\%) than \textit{ImageJ} (52.49\%) and \textit{Fiji} (63.79\%).
+Table \ref{tab_maintainability_docs} shows which projects had these documents.
+
+\begin{table}[H]
+\centering
+\begin{tabular}{llll}
+\hline
+\multicolumn{1}{c}{Software} & Proj plan & Dev manual & API doc \\ \hline
+3D Slicer & X & X & X \\
+Weasis &  & X &  \\
+SMILI &  &  & X \\
+ImageJ & X & X & X \\
+Fiji & X & X & X \\
+dwv &  &  & X \\
+BioImage Suite Web &  & X &  \\
+OHIF Viewer &  & X & X \\
+ParaView & X &  &  \\
+INVESALIUS 3 & X &  &  \\
+medInria &  & X &  \\
+Gwyddion &  & X & X \\ \hline
+\end{tabular}
+\caption{\label{tab_maintainability_docs}Software with the maintainability documents}
+\end{table}
+
+27 of the 29 projects used git as the version control tool; \textit{AMIDE} used Mercurial; \textit{Gwyddion} used Subversion. 24 projects used GitHub for their repositories; \textit{XMedCon
+}, \textit{AMIDE}, and \textit{Gwyddion} used SourceForge; \textit{DicomBrowser} and \textit{3DimViewer} used BitBucket.
 
 \section{Reusability}
+Figure 4.7 shows the AHP results for \textit{reusability}.
 
-\section{Understandability}
+\begin{figure}[H]
+\includegraphics[scale=0.38]{figures/reusability_scores.png}
+\caption{AHP reusability scores}
+\label{fg_reusability_scores}
+\end{figure}
 
-\section{Visibility \& Transparency}
+As described in Section \ref{sec_grading_template}, we gave higher scores to the projects with an API document and more code files.
+As shown in Table \ref{tab_maintainability_docs}, seven projects had API documents.
+Figure \ref{fg_num_text_files} shows the number of text-based files by projects.
+
+\begin{figure}[H]
+\includegraphics[scale=0.38]{figures/num_text_files.png}
+\caption{Number of text-based files files by projects}
+\label{fg_num_text_files}
+\end{figure}
+
+\section{Surface Understandability}
+Figure \ref{fg_surface_understandability_scores} shows the scores for \textit{surface understandability}.
+
+\begin{figure}[H]
+\includegraphics[scale=0.38]{figures/understandability_scores.png}
+\caption{AHP surface understandability scores}
+\label{fg_surface_understandability_scores}
+\end{figure}
+
+All projects had a consistent code style with parameters in the same order for all functions; the code was modularized; the comments were clear, indicating what is being done, not how.
+However, we only found explicit identification of a coding standard for only 3 out of the 29, which are \textit{3D Slicer}, \textit{Weasis}, and \textit{ImageJ}.
+We also found hard-coded constants in \textit{medInria}, \textit{dicompyler}, \textit{MicroView}, and \textit{Papaya}.
+We did not find any reference to the used algorithms in projects \textit{XMedCon}, \textit{DicomBrowser}, \textit{3DimViewer}, \textit{BioImage Suite Web}, \textit{Slice:Drop}, \textit{MatrixUser}, \textit{DICOM Viewer}, \textit{dicompyler}, and \textit{Papaya}.
+
+\section{Visibility/Transparency}
+Figure \ref{fg_visibility_transparency_scores} shows the AHP scores for \textit{visibility/transparency}.
+Generally speaking, the teams that actively documented their development process and plans scored higher because they delivered better communication to people outside the team.
+
+\begin{figure}[H]
+\includegraphics[scale=0.38]{figures/visibility_transparency_scores.png}
+\caption{AHP visibility/transparency scores}
+\label{fg_visibility_transparency_scores}
+\end{figure}
+
+Table \ref{tab_Visibility/Transparency_docs} shows the projects which had documents for the development process, project status, development environment, and release notes.
+
+\begin{table}[]
+\centering
+\begin{tabular}{lllll}
+\hline
+Software & Dev process & Proj status & Dev env & Rls notes \\ \hline
+3D Slicer & X & X & X & X \\
+Weasis \cite{Roduit2021} &  &  & X & X \\
+MRIcroGL \cite{Rorden2021} &  &  &  & X \\
+SMILI \cite{Chandra2018} &  &  &  & X \\
+ImageJ \cite{Rueden2017} & X & X & X & X \\
+Fiji \cite{Schindelin2012} & X & X & X &  \\
+Horos \cite{horosproject2020} &  &  &  & X \\
+OsiriX Lite \cite{PixmeoSARL2019} &  &  &  & X \\
+dwv \cite{Martelli2021} &  &  &  & X \\
+Drishti \cite{Limaye2012} &  &  &  & X \\
+\begin{tabular}[c]{@{}l@{}}BioImage Suite\\ Web\end{tabular} &  &  & X &  \\
+OHIF Viewer \cite{Ziegler2020} &  &  & X & X \\
+GATE \cite{Jan2004} &  &  &  & X \\
+ITK-SNAP \cite{Yushkevich2006} &  &  &  & X \\
+ParaView \cite{Ahrens2005} &  & X &  &  \\
+MatrixUser \cite{Liu2016} &  &  &  & X \\
+DICOM Viewer \cite{Afsar2021} &  &  & X & X \\
+INVESALIUS 3 \cite{Amorim2015} &  &  &  & X \\
+medInria \cite{Fillard2012} &  &  & X & X \\
+MicroView \cite{ParallaxInnovations2020} &  &  &  & X \\
+Gwyddion \cite{Nevcas2012} &  &  &  & X \\ \hline
+\end{tabular}
+\caption{\label{tab_Visibility/Transparency_docs}Software with the visibility/transparency documents}
+\end{table}
+
+\section{Overall Scores}
+
+As described in Section \ref{sec_AHP}, for our AHP measurements, there are nine criteria which are the nine software qualities and 29 software packages as the alternatives.
+We decided to make all nine qualities equally important, so the score of each quality affects the overall scores on the same scale.
+
+Figure \ref{fg_overall_scores} shows the overall scores of all 29 software packages in descending order.
+Since we produced the scores from the AHP process, the total sum of the 29 scores is precisely 1.
+
+\begin{figure}[H]
+\includegraphics[scale=0.38]{figures/overall_scores.png}
+\caption{Overall AHP scores for all 9 software qualities}
+
+\label{fg_overall_scores}
+\end{figure}
+
+The top software products had higher scores in most criteria.
+For example, \textit{3D Slicer} ranked in the top two software products for all qualities except \textit{surface robustness}; \textit{ImageJ} ranked in the top three products for \textit{correctness \& verifiability}, \textit{surface reliability}, \textit{surface usability}, \textit{maintainability}, \textit{ surface understandability}, and \textit{visibility/transparency}; \textit{OHIF Viewer} ranked in the top five products for \textit{correctness \& verifiability}, \textit{surface reliability}, \textit{surface usability}, \textit{maintainability}, and \textit{reusability}.
+We might underestimate its scores of two qualities for \textit{DICOM Viewer}, but equally compared it with the other software for the rest seven qualities.

### 2021-07-24 StateOfPractice/Ao-Report/15InterviewsWithDevelopers.tex
minor update
MI report - updated ch 4

@@
 The interviewees may have provided multiple answers to each question.
-When counting the number of interviewees for each response, we count all of them with this answer.
+When counting the number of interviewees for each response, we count every interviewee with this answer.
 
@@
 \begin{itemize}
-\item GitHub Actions, which is a good continuous integration and continuous delivery (CI/CD) tool to save time.
+\item GitHub Actions, which is a good CI/CD tool to save time.
 \end{itemize}

### 2021-07-24 StateOfPractice/Ao-Report/14MeasurementResults.tex
MI report - minor updates

@@
 We considered anyone who made at least one accepted commit to the source code as a contributor.
-Thus, it does not mean that any development teams had long-term members as many as 100.
+Thus, it does not mean that any development team has are 100 long-term members.
 Many of these projects received change requests and code from the community, such as pull requests and git commits on GitHub.
@@
 We failed installing \textit{DICOM Viewer}, so we could not test \textit{surface reliability} and \textit{surface robustness} for it.
-We might underestimate the scores of the two qualities for this software.
 We kept this software on our list because the other seven qualities do not rely on a successful installation.
@@
 
-We identified five projects using continuous integration and continuous delivery (CI/CD) tools, which are \textit{3D Slicer}, \textit{ImageJ
+We identified five projects using CI/CD tools, which are \textit{3D Slicer}, \textit{ImageJ
 }, \textit{Fiji}, \textit{dwv}, and \textit{OHIF Viewer}.
@@
 
-The top software products had higher scores in most criteria.
-For example, \textit{3D Slicer} ranked in the top two software products for all qualities except \textit{surface robustness}; \textit{ImageJ} ranked in the top three products for \textit{correctness \& verifiability}, \textit{surface reliability}, \textit{surface usability}, \textit{maintainability}, \textit{ surface understandability}, and \textit{visibility/transparency}; \textit{OHIF Viewer} ranked in the top five products for \textit{correctness \& verifiability}, \textit{surface reliability}, \textit{surface usability}, \textit{maintainability}, and \textit{reusability}.
-We might underestimate its scores of two qualities for \textit{DICOM Viewer}, but equally compared it with the other software for the rest seven qualities.
+The top three software products \textit{3D Slicer}, \textit{ImageJ}, and \textit{OHIF Viewer} had higher scores in most criteria.
+\textit{3D Slicer} ranked in the top two software products for all qualities except \textit{surface robustness}; \textit{ImageJ} ranked in the top three products for \textit{correctness \& verifiability}, \textit{surface reliability}, \textit{surface usability}, \textit{maintainability}, \textit{ surface understandability}, and \textit{visibility/transparency}; \textit{OHIF Viewer} ranked in the top five products for \textit{correctness \& verifiability}, \textit{surface reliability}, \textit{surface usability}, \textit{maintainability}, and \textit{reusability}.
+We might underestimate its scores of qualities \textit{surface reliability} and \textit{surface robustness} for \textit{DICOM Viewer}, but equally compared it with the other software for the rest seven qualities.

### 2021-07-24 StateOfPractice/Ao-Report/13Methods.tex
MI report - minor updates
MI report - updated software table in ch 4

@@
 If any problem interrupts the process of installation or uninstallation, we give a lower score to this quality.
-We also record the operating system for the installation test and whether we can verify the installation.
+We also record the operating system (OS) for the installation test and whether we can verify the installation.
 
 \item \textbf{Correctness \& Verifiability} For \textit{correctness}, we check the projects to identify the techniques to ensure this quality, such as literate programming, automated testing, symbolic execution, model checking, unit tests, etc.
-We also examine whether the projects use continuous integration.
+We also examine whether the projects use continuous integration and continuous delivery (CI/CD).
 For \textit{verifiability}, we go through the documents of the projects to check the requirements specifications, theory manuals, and getting started tutorials.
@@
 Then we followed the steps in Section \ref{sec_grading_software} to measure and grade the software.
-27 out of the 29 packages are compatible with two or three different operating systems such as Windows, macOS, and Linux, and 5 of them are browser-based, making them platform-independent.
+27 out of the 29 packages are compatible with two or three different OS such as Windows, macOS, and Linux, and 5 of them are browser-based, making them platform-independent.
 However, in the interest of time, we only performed the measurements for each project by installing it on one of the platforms, most likely Windows.

### 2021-07-24 StateOfPractice/Ao-Report/14MeasurementResults.tex
MI report - updated software table in ch 4

@@
 Table \ref{tab_final_list} shows the 29 software packages that we measured.
-We tried to find the initial release dates in the documents, such as the introduction and release notes.
+We tried to find the initial release dates (Rlsd) in the documents, such as the introduction and release notes.
+We did not find this date for two projects and marked them with ``?".
 We used the date of the latest change to each code repository to decide the latest update.
-We found out funding information for eight projects, and we are not sure about the rest.
+We found out funding information (Fnd) for eight projects, and we are not sure about the rest.
 
-We considered anyone who made at least one accepted commit to the source code as a contributor.
+We counted the number of contributors (NOC) and lines of code (LOC).
+we considered anyone who made at least one accepted commit to the source code as a contributor.
 Thus, it does not mean that any development team has are 100 long-term members.
@@
 
-Table \ref{tab_final_list} also shows the supported operating systems for each software package, and 25 of them could work on all three Windows (W), macOS (M), and Linux (L) systems.
+Table \ref{tab_final_list} also shows the supported OS for each software package, and 25 of them could work on all three Windows (W), macOS (M), and Linux (L) systems.
 However, there was a significant difference in the philosophy to achieve cross-platform compatibility.
@@
 \begin{table}[H]
-\begin{tabular}{lllllllll}
+\begin{tabular}{llllllllll}
 \hline
-\multirow{2}{*}{Software} & \multirow{2}{*}{Released} & \multirow{2}{*}{Updated} & \multirow{2}{*}{Contributors} & \multirow{2}{*}{Funded} & \multicolumn{3}{c}{OS} & \multirow{2}{*}{Web} \\ \cline{6-8}
- &  &  &  &  & W & M & L &  \\ \hline
-3D Slicer \cite{Kikinis2014} & 1998 & 2020-08 & 100 & X & X & X & X &  \\
-Ginkgo CADx \cite{Wollny2020} & 2010 & 2019-05 & 3 &  & X & X & X &  \\
-XMedCon \cite{Nolf2003} & 2000 & 2020-08 & 2 &  & X & X & X &  \\
-Weasis \cite{Roduit2021} & 2010 & 2020-08 & 8 &  & X & X & X &  \\
-MRIcroGL \cite{Rorden2021} & 2015 & 2020-08 & 2 &  & X & X & X &  \\
-SMILI \cite{Chandra2018} & 2014 & 2020-06 & 9 &  & X & X & X &  \\
-ImageJ \cite{Rueden2017} & 1997 & 2020-08 & 18 & X & X & X & X &  \\
-Fiji \cite{Schindelin2012} & 2011 & 2020-08 & 55 & X & X & X & X &  \\
-DicomBrowser \cite{Archie2012} & 2012 & 2020-08 & 3 &  & X & X & X &  \\
-3DimViewer \cite{TESCAN2020} & unclear & 2020-03 & 3 & X & X & X &  &  \\
-Horos \cite{horosproject2020} & unclear & 2020-04 & 21 &  &  & X &  &  \\
-OsiriX Lite \cite{PixmeoSARL2019} & 2004 & 2019-11 & 9 &  &  & X &  &  \\
-dwv \cite{Martelli2021} & 2012 & 2020-09 & 22 &  & X & X & X & X \\
-Drishti \cite{Limaye2012} & 2012 & 2020-08 & 1 &  & X & X & X &  \\
-\begin{tabular}[c]{@{}l@{}}BioImage Suite\\ Web \cite{Papademetris2005}\end{tabular} & 2018 & 2020-10 & 13 & X & X & X & X & X \\
-OHIF Viewer \cite{Ziegler2020} & 2015 & 2020-10 & 76 &  & X & X & X & X \\
-Slice:Drop \cite{Haehn2013} & 2012 & 2020-04 & 3 &  & X & X & X & X \\
-GATE \cite{Jan2004} & 2011 & 2020-10 & 45 &  &  & X & X &  \\
-ITK-SNAP \cite{Yushkevich2006} & 2006 & 2020-06 & 13 & X & X & X & X &  \\
-ParaView \cite{Ahrens2005} & 2002 & 2020-10 & 100 & X & X & X & X & X \\
-MatrixUser \cite{Liu2016} & 2013 & 2018-07 & 1 &  & X & X & X &  \\
-DICOM Viewer \cite{Afsar2021} & 2018 & 2020-04 & 5 & X & X & X & X &  \\
-INVESALIUS 3 \cite{Amorim2015} & 2009 & 2020-09 & 10 &  & X & X & X &  \\
+\multirow{2}{*}{Software} & \multirow{2}{*}{Rlsd} & \multirow{2}{*}{Updated} & \multirow{2}{*}{Fnd} & \multirow{2}{*}{NOC} & \multirow{2}{*}{LOC} & \multicolumn{3}{c}{OS} & \multirow{2}{*}{W} \\ \cline{7-9}
+ &  &  &  &  &  & W & M & L &  \\ \hline
+3D Slicer \cite{Kikinis2014} & 1998 & 2020-08 & X & 100 & 501451 & X & X & X &  \\
+Ginkgo CADx \cite{Wollny2020} & 2010 & 2019-05 &  & 3 & 257144 & X & X & X &  \\
+XMedCon \cite{Nolf2003} & 2000 & 2020-08 &  & 2 & 96767 & X & X & X &  \\
+Weasis \cite{Roduit2021} & 2010 & 2020-08 &  & 8 & 123272 & X & X & X &  \\
+MRIcroGL \cite{Rorden2021} & 2015 & 2020-08 &  & 2 & 8493 & X & X & X &  \\
+SMILI \cite{Chandra2018} & 2014 & 2020-06 &  & 9 & 62626 & X & X & X &  \\
+ImageJ \cite{Rueden2017} & 1997 & 2020-08 & X & 18 & 9681 & X & X & X &  \\
+Fiji \cite{Schindelin2012} & 2011 & 2020-08 & X & 55 & 10833 & X & X & X &  \\
+DicomBrowser \cite{Archie2012} & 2012 & 2020-08 &  & 3 & 5505 & X & X & X &  \\
+3DimViewer \cite{TESCAN2020} & ? & 2020-03 & X & 3 & 178065 & X & X &  &  \\
+Horos \cite{horosproject2020} & ? & 2020-04 &  & 21 & 561617 &  & X &  &  \\
+OsiriX Lite \cite{PixmeoSARL2019} & 2004 & 2019-11 &  & 9 & 544304 &  & X &  &  \\
+dwv \cite{Martelli2021} & 2012 & 2020-09 &  & 22 & 47815 & X & X & X & X \\
+Drishti \cite{Limaye2012} & 2012 & 2020-08 &  & 1 & 268168 & X & X & X &  \\
+\begin{tabular}[c]{@{}l@{}}BioImage Suite\\ Web \cite{Papademetris2005}\end{tabular} & 2018 & 2020-10 & X & 13 & 139699 & X & X & X & X \\
+OHIF Viewer \cite{Ziegler2020} & 2015 & 2020-10 &  & 76 & 63951 & X & X & X & X \\
+Slice:Drop \cite{Haehn2013} & 2012 & 2020-04 &  & 3 & 19020 & X & X & X & X \\
+GATE \cite{Jan2004} & 2011 & 2020-10 &  & 45 & 207122 &  & X & X &  \\
+ITK-SNAP \cite{Yushkevich2006} & 2006 & 2020-06 & X & 13 & 88530 & X & X & X &  \\
+ParaView \cite{Ahrens2005} & 2002 & 2020-10 & X & 100 & 886326 & X & X & X & X \\
+MatrixUser \cite{Liu2016} & 2013 & 2018-07 &  & 1 & 23121 & X & X & X &  \\
 \end{tabular}
@@
 \begin{table}[H]
-\begin{tabular}{lllllllll}
+\begin{tabular}{llllllllll}
 \hline
-\multirow{2}{*}{Software} & \multirow{2}{*}{Released} & \multirow{2}{*}{Updated} & \multirow{2}{*}{Contributors} & \multirow{2}{*}{Funded} & \multicolumn{3}{c}{OS} & \multirow{2}{*}{Web} \\ \cline{6-8}
- &  &  &  &  & W & M & L &  \\ \hline
-medInria \cite{Fillard2012} & 2009 & 2020-11 & 21 &  & X & X & X &  \\
-dicompyler \cite{Panchal2010} & 2009 & 2020-01 & 2 &  & X & X &  &  \\
-MicroView \cite{ParallaxInnovations2020} & 2015 & 2020-08 & 2 &  & X & X & X &  \\
-Papaya \cite{UTHSCSA2019} & 2012 & 2019-05 & 9 &  & X & X & X &  \\
-AMIDE \cite{Loening2017} & 2006 & 2017-01 & 4 &  & X & X & X &  \\
-Gwyddion \cite{Nevcas2012} & 2004 & 2020-11 & 38 &  & X & X & X &  \\ \hline
+\multirow{2}{*}{Software} & \multirow{2}{*}{Rlsd} & \multirow{2}{*}{Updated} & \multirow{2}{*}{Fnd} & \multirow{2}{*}{NOC} & \multirow{2}{*}{LOC} & \multicolumn{3}{c}{OS} & \multirow{2}{*}{W} \\ \cline{7-9}
+ &  &  &  &  &  & W & M & L &  \\ \hline
+DICOM Viewer \cite{Afsar2021} & 2018 & 2020-04 & X & 5 & 30761 & X & X & X &  \\
+INVESALIUS 3 \cite{Amorim2015} & 2009 & 2020-09 &  & 10 & 48605 & X & X & X &  \\
+medInria \cite{Fillard2012} & 2009 & 2020-11 &  & 21 & 148924 & X & X & X &  \\
+dicompyler \cite{Panchal2010} & 2009 & 2020-01 &  & 2 & 15941 & X & X &  &  \\
+MicroView \cite{ParallaxInnovations2020} & 2015 & 2020-08 &  & 2 & 27470 & X & X & X &  \\
+Papaya \cite{UTHSCSA2019} & 2012 & 2019-05 &  & 9 & 71831 & X & X & X &  \\
+AMIDE \cite{Loening2017} & 2006 & 2017-01 &  & 4 & 102827 & X & X & X &  \\
+Gwyddion \cite{Nevcas2012} & 2004 & 2020-11 &  & 38 & 643427 & X & X & X &  \\ \hline
 \end{tabular}

### 2021-07-24 StateOfPractice/Ao-Report/15InterviewsWithDevelopers.tex
MI report - updated software table in ch 4

@@
  & Lack of time to devote to the project & 2 & 1 \\ \hdashline
-\multirow{3}{*}{2} & \begin{tabular}[c]{@{}l@{}}Hard to keep up with changes in operating \\ systems and libraries\end{tabular} & 1 &  \\
- & Hard to support multiple operating systems & 2 &  \\
+\multirow{3}{*}{2} & \begin{tabular}[c]{@{}l@{}}Hard to keep up with changes in OS \\ and libraries\end{tabular} & 1 &  \\
+ & Hard to support multiple OS & 2 &  \\
  & Hard to support lower-end computers & 1 & 2 \\ \hdashline

### 2021-07-26 StateOfPractice/Ao-Report/11Introduction.tex
MI report - added ch 1 and ch 6

@@
 \label{ch_intro}
+This paper analyzes the state of the practice for Medical Imaging (MI) software.
+MI is the clinical tool to image the interior of a body, providing information for diagnostic, analytic, and medical applications \cite{FDA2021} \cite{enwiki:1034887445}.
+MI is an essential part of collecting accurate information during clinical diagnosis \cite{Zhang2008}.
+MI computing and software aim to visualize and process medical images and produce clinically meaningful information \cite{enwiki:1034877594}.
 
-introduction
+Scientiﬁc computing (SC) is the technique and process of mathematically modeling real-world engineering or scientific systems with computing tools, as well as analysis and prediction with the models \cite{Smith2006}.
+MI software belongs to a specific domain of SC.
+
+We aim to study the current status of SC software development in the MI domain; understand the current merits, drawbacks, and pain points during the development process, as well as the software qualities in the domain; provide guidelines and recommendations for future development.
+
+Section \ref{sec_motivation} presents our motivation to start the research set the above goals, Section \ref{sec_research_questions} lists our research questions, and Section \ref{sec_scope} explains the domain analysis of MI software scope of our research.
+
+\section{Motivation}
+\label{sec_motivation}
+Most scientists think developing and using SC software play significant roles in their research \cite{Hannay2009}.
+They spend a substantial proportion of their working hours on SC software development \cite{Hannay2009} \cite{Prabhu2011}, and this proportion of time has increased over the years \cite{Hannay2009}.
+
+Developing SC software requires solid knowledge in specific domains \cite{Wilson2014}.
+Many of them learn software engineering skills by themselves or from their peers, instead of proper training \cite{Hannay2009}.
+Hannay et al.
+\cite{Hannay2009} also pointed out that many scientists showed ignorance and indifference to standard software engineering concepts.
+According to a survey by Prabhu et al.
+\cite{Prabhu2011}, more than half of the 114 subjects did not use any proper debugger for their software.
+
+Due to its nature, SC software born from one project can be part of many other projects in the future, with the potential to disproportionately causing damages to scientific researches \cite{Wilson2014}.
+
+As a result, the development process and quality of SC software concern us. We want to understand their status in SC domains and improve them.
+
+\section{Research Questions}
+\label{sec_research_questions}
+To achieve our objectives, we designed a few research questions and tried to answer them by our research methods. The questions are as follows,
+
+\begin{enumerate}
+\item What are the pain points for developers working on MI software projects? What are the solutions for these pain points?
+\item What artifacts the projects generated?
+\item What role does documentation play in the projects? What are the developers' attitudes toward it?
+\item What principles, processes, methodologies, and tools the projects used?
+\item What is the current status of the following software qualities for the projects? What actions have the developers taken to address them?
+\begin{itemize}
+	\item Installability
+	\item Correctness \& Verifiability
+	\item Reliability
+	\item Robustness
+	\item Usability
+	\item Maintainability
+	\item Reusability
+	\item Understandability
+	\item Visibility/Transparency
+\end{itemize}
+\item How does software quality ranking generated by our methods compare with the ratings from the community?
+\end{enumerate}
+
+\section{Scope}
+\label{sec_scope}
+According to Bankman \cite{Bankman2000}, MI software deals with six different basic problems, while Angenentet et al.
+\cite{Angenent2006} pointed out that four fundamental problems are solved by MI software.
+While both mentioned Segmentation, Registration, and Visualization of medical images, Bankman also included Enhancement, Quantification, and a section covering some other functions \cite{Bankman2000}.
+On the other hand, Angenent's team included Simulation \cite{Angenent2006}.
+According to Wikipedia contributors \cite{enwiki:1034877594}, MI software has primary functions in categories such as Segmentation, Registration, Visualization (including the basic display, reformatted views, and 3D volume rendering), Statistical Analysis, Image-based Physiological Modelling, etc.
+As Kim et al.
+\cite{Kim2011} describe, the general steps of medical image analysis after obtaining digital data include Enhancement, Segmentation, Feature Extraction, Classification, and Interpretation.
+Besides the above major functions, some MI software provides supportive functions.
+For example, with Tool Kit libraries VTK \cite{SchroederEtAl2006} and ITK \cite{McCormick2014}, developers build software with Visualization and Analysis functions; Picture Archiving and Communication System (PACS) helps users to economically store and conveniently access images \cite{Choplin1992}.
+
+We divided MI software into five sub-groups and several sub-sub-groups by their major functions shown in Figure \ref{fig_mi_functions}.
+
+\begin{figure}
+\centering
+\begin{tikzpicture}[mindmap, grow cyclic, every node/.style=concept, concept
+color=orange!40,
+level 1/.append style={level distance=5cm,sibling angle=72},
+level 2/.append style={level distance=3cm,sibling angle=90},
+level 3/.style={level distance=2.3cm,sibling angle=90}]
+\node{MI software}
+child[concept color=teal!40] { node {Enhancement}}
+child[concept color=teal!40] { node {Analysis}
+    child[concept color=blue!30] { node {Registration}}
+    child[concept color=blue!30,] { node {Segmentation}}
+    child[concept color=blue!30] { node {Statistical Analysis}
+    child[concept color=blue!20] { node {Feature Extraction}}
+    child[concept color=blue!20] { node {Classification}}
+    child[concept color=blue!20] { node {Interpretation}}
+    }}
+child[concept color=teal!40] { node {Simulation/\\Modeling}}
+child[concept color=teal!40] { node {Supporting}
+    child[concept color=blue!30] { node {Tool Kit}}
+    child[concept color=blue!30] { node {PACS}}
+}
+child[concept color=teal!40] { node {Visualization}
+    child[concept color=blue!30] { node {2D Display}}
+    child[concept color=blue!30] { node {3D Rendering}}
+    child[concept color=blue!30] { node {Reformatted Views}}
+};
+\end{tikzpicture}
+\caption{Major functions of MI software}
+\label{fig_mi_functions}
+\end{figure}
+
+In this project, the scope of the software is limited to the software library providing the Visualization tools and functions.

### 2021-07-26 StateOfPractice/Ao-Report/13Methods.tex
MI report - added ch 1 and ch 6

@@
 
-\section{Applying the Method to Medical Imaging (MI)}
+\section{Applying the Method to MI}
 \label{sec_applying_method}
 
-Based on the principles in Section \ref{sec_domain_selection}, we selected the MI domain and the viewer sub-group. We also included MI domain experts in our team.
+Based on the principles in Section \ref{sec_domain_selection}, we selected the MI domain and the sub-group of software with Visualization function shown in Figure \ref{fig_mi_functions}.
+We also included MI domain experts in our team.
 
@@
 So following guidelines in Section \ref{sec_filter_software_list} we removed them from the list.
-Next, we focused on the MI software that could work as a MI viewer.
+Next, we focused on the MI software providing visualization functions.
 Some of the software on the list were tool kits or libraries for other software to use as dependencies but not for end-users to view medical images, such as \textit{VTK}, \textit{ITK}, \textit{dcm4che}, etc.

### 2021-07-26 StateOfPractice/Ao-Report/16ThreatToValidity.tex
MI report - added ch 1 and ch 6

@@
 
-\begin{itemize}
-\item limited time to each software, surface quality
-\item impressions, subjective
-\item searched all information by ourselves, so maybe missed some info
-\item mainly measured by one person, so maybe biased
-\end{itemize}
+This section lists all the potential threats to the validity of our research.
+
+\begin{enumerate}
+\item We compared nine software qualities for 29 software packages, so we could only spend a limited time on each of them.
+As a result, our assessments may not be thorough in revealing their status fully.
+\item We used the grading template in Appendix \ref{ap_grading_template} to guide our measurements.
+Our impressions of the software - such as user experience - were factors in deciding some scores.
+Thus, there is a risk that some scores may be subjective and biased.
+\item It was not practical to ask each development team for every piece of information.
+We collected much information - such as artifacts and funding situations of software - by ourselves.
+There may be cases that we missed some information.
+\item We interviewed eight teams, which is a good proportion of the 29. However, there is still a risk that they might not well represent the whole MI software community.
+\item As mentioned in Section \ref{ch_interview}, one interviewee was too busy to participate in a full interview, so he provided a version of written answers to us.
+Since we did not have the chance to explain our questions or ask him follow-up questions, there is a possibility of misinterpretation of the questions or answers.
+\end{enumerate}

### 2021-07-26 StateOfPractice/Ao-Report/32FullSoftwareListBeforeFiltering.tex
MI report - update styles
MI report - added ch 1 and ch 6

@@
 Table \ref{tab_list_before_filtering} lists the 48 software before filtering.
-We selected 29 of them to the final list, which are all open-source and in the same sub-group.
-We divided them into four sub-groups: Viewer (V), Tool Kit (TK), Picture Archiving and Communication System (PACS), and Convertion (C).
+We selected 29 of them to the final list, which are all open-source and in Visualization (V) sub-group.
+We found software packages in sub-groups Tool Kit (TK) and PACS but removed them from the final list.
 The table also shows the sources of identifying them.
 
-\begin{table}[]
+\begin{table}[H]
 \begin{tabular}{lllllllllll}
@@
 DicomBrowser \cite{Archie2012} & X & X & V &  &  &  & X & X &  &  \\
+\end{tabular}
+\end{table}
+
+\begin{table}[H]
+\begin{tabular}{lllllllllll}
+\hline
+ &  &  &  & \multicolumn{7}{c}{Source} \\ \cline{5-11} 
+\multirow{-2}{*}{Software} & \multirow{-2}{*}{\begin{tabular}[c]{@{}l@{}}Final\\ list\end{tabular}} & \multirow{-2}{*}{\begin{tabular}[c]{@{}l@{}}Open-\\ source\end{tabular}} & \multirow{-2}{*}{\begin{tabular}[c]{@{}l@{}}Sub-\\ group\end{tabular}} & \cite{Bjorn2017} & \cite{Bruhschwein2019} & \cite{Haak2015} & \cite{Emms2019} & \cite{Hasan2020} & \cite{Mu2019} & \cite{Samala2014} \\ \hline
 3DimViewer \cite{TESCAN2020} & X & X & V & X &  &  & X & X &  &  \\
@@
 Slice:Drop \cite{Haehn2013} & X & X & V &  &  & X &  &  &  &  \\
-	\end{tabular}
-\end{table}
-
-\begin{table}[]
-	\begin{tabular}{lllllllllll}
-		\hline
-		&  &  &  & \multicolumn{7}{c}{Source} \\ \cline{5-11} 
-		\multirow{-2}{*}{Software} & \multirow{-2}{*}{\begin{tabular}[c]{@{}l@{}}Final\\ list\end{tabular}} & \multirow{-2}{*}{\begin{tabular}[c]{@{}l@{}}Open-\\ source\end{tabular}} & \multirow{-2}{*}{\begin{tabular}[c]{@{}l@{}}Sub-\\ group\end{tabular}} & \cite{Bjorn2017} & \cite{Bruhschwein2019} & \cite{Haak2015} & \cite{Emms2019} & \cite{Hasan2020} & \cite{Mu2019} & \cite{Samala2014} \\ \hline
 GATE \cite{Jan2004} & X & X & V &  &  &  & X &  &  &  \\
@@
 ITK \cite{McCormick2014} &  & X & TK &  &  &  &  &  &  & X \\
+\end{tabular}
+\end{table}
+
+\begin{table}[H]
+\begin{tabular}{lllllllllll}
+\hline
+ &  &  &  & \multicolumn{7}{c}{Source} \\ \cline{5-11} 
+\multirow{-2}{*}{Software} & \multirow{-2}{*}{\begin{tabular}[c]{@{}l@{}}Final\\ list\end{tabular}} & \multirow{-2}{*}{\begin{tabular}[c]{@{}l@{}}Open-\\ source\end{tabular}} & \multirow{-2}{*}{\begin{tabular}[c]{@{}l@{}}Sub-\\ group\end{tabular}} & \cite{Bjorn2017} & \cite{Bruhschwein2019} & \cite{Haak2015} & \cite{Emms2019} & \cite{Hasan2020} & \cite{Mu2019} & \cite{Samala2014} \\ \hline
 DCMTK \cite{DCMTK2021} &  & X & TK &  &  &  &  & X & X &  \\
@@
 cornerstone \cite{cornerstone2020} &  & X & TK &  &  & X &  &  &  &  \\
-	\end{tabular}
-\end{table}
-
-\begin{table}[]
-	\begin{tabular}{lllllllllll}
-		\hline
-		&  &  &  & \multicolumn{7}{c}{Source} \\ \cline{5-11} 
-		\multirow{-2}{*}{Software} & \multirow{-2}{*}{\begin{tabular}[c]{@{}l@{}}Final\\ list\end{tabular}} & \multirow{-2}{*}{\begin{tabular}[c]{@{}l@{}}Open-\\ source\end{tabular}} & \multirow{-2}{*}{\begin{tabular}[c]{@{}l@{}}Sub-\\ group\end{tabular}} & \cite{Bjorn2017} & \cite{Bruhschwein2019} & \cite{Haak2015} & \cite{Emms2019} & \cite{Hasan2020} & \cite{Mu2019} & \cite{Samala2014} \\ \hline
+dcm2niix \cite{Li2016} &  & X & TK &  &  &  &  & X &  &  \\
 orthanc \cite{Jodogne2018} &  & X & PACS & X &  &  &  & X &  &  \\
@@
 ClearCanvas \cite{ClearCanvas2015} &  & X & PACS, V & X &  &  &  &  &  &  \\
-		dcm2niix \cite{Li2016} &  & X & C &  &  &  &  & X &  &  \\
 \begin{tabular}[c]{@{}l@{}}Open Dicom\\ Viewer \cite{OpenDicomViewer2011} \end{tabular} &  & X & V &  &  & X &  &  & X &  \\

### 2021-07-27 StateOfPractice/Ao-Report/31FullGradingTemplate.tex
updated ao's report ch 3
MI report - added append A

@@
 \chapter{Full Grading Template}
-\label{ap_grading}
+\label{ap_grading_template}
 
-appendix here
+\newpage
+\begingroup
+\renewcommand{\arraystretch}{0.8}
+
+\begin{table}[H]
+\centering
+\caption{Measurement Template}\label{measurementtemplate}
+\begin{tabular}{p{14cm}}
+\hline
+\textbf{Summary Information}\\
+\hline
+Software name? (string)\\
+URL? (URL)\\
+Affiliation (institution(s)) (string or {N/A})\\
+Software purpose (string)\\
+Number of developers (all developers that have contributed at least one commit to the project) (use repo commit logs) (number)\\
+How is the project funded? (unfunded, unclear, funded$\ast$) where $\ast$ requires a string to say the source of funding\\
+Initial release date? (date)\\
+Last commit date? (date)\\
+Status? (alive is defined as presence of commits in the last 18 months) ({alive, dead, unclear})\\
+License? ({GNU GPL, BSD, MIT, terms of use, trial, none, unclear, other$\ast$}) $\ast$ given via a string \\
+Platforms? (set of {Windows, Linux, OS X, Android, other$\ast$}) $\ast$ given via string\\
+Software Category?
+The concept category includes software that does not have an officially released version.
+Public software has a released version in the public domain.
+Private software has a released version available to authorized users only.
+({concept, public, private})\\
+Development model? ({open source, freeware, commercial, unclear})\\
+Publications about the software? Refers to publications that have used or mentioned the software. (number or {unknown})\\
+Source code URL? ({set of url, n/a, unclear})\\
+Programming language(s)? (set of {FORTRAN, Matlab, C, C++, Java, R, Ruby, Python, Cython, BASIC, Pascal, IDL, unclear, other$\ast$}) $\ast$ given via string \\
+Is there evidence that performance was considered? Performance refers to either speed, storage, or throughput. ({yes$\ast$, no})\\
+Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have) \\
+\hline
+\end{tabular}
+\end{table}
+
+\begin{table}[H]
+\centering
+\begin{tabular}{p{14cm}}
+\hline		
+\textbf{Installability  (Measured via installation on a virtual machine.) }\\
+\hline
+Are there installation instructions? ({yes, no})\\
+Are the installation instructions in one place? Place referring to a single document or web-page. ({yes, no, n/a})\\
+Are the installation instructions linear? Linear meaning progressing  in a single series of steps. ({yes, no, n/a})\\
+Are the instructions written as if the person doing the installation has none of the dependent packages installed? ({yes, no, unclear})\\
+Are compatible operating system versions listed? ({yes, no})\\
+Is there something in place to automate the installation (makefile, script, installer, etc)? ({yes$\ast$, no})\\
+If the software installation broke, was a descriptive error message displayed? ({yes, no, n/a})\\
+Is there a specified way to validate the installation? ({yes$\ast$, no})\\
+How many steps were involved in the installation? (Includes manual steps like unzipping files) Specify OS. (number, OS)\\
+What OS was used for the installation? ({Windows, Linux, OS X, Android, other$\ast$ }) $\ast$given via string\\
+How many extra software packages need to be installed before or during installation? (number)\\
+Are required package versions listed? ({yes, no, n/a})\\
+Are there instructions for the installation of required packages / dependencies? ({yes, no, n/a})\\
+Run uninstall, if available. Were any obvious problems caused? ({yes$\ast$ , no, unavail})\\
+Overall impression? ({1 .. 10})\\
+Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have)\\
+\hline
+\end{tabular}
+\end{table}
+
+\begin{table}[H]
+\centering
+\begin{tabular}{p{14cm}}
+\hline		
+\textbf{Correctness and Verifiability}\\
+\hline
+Any reference to the requirements specifications of the program or theory manuals? ({yes$\ast$ , no, unclear})\\
+What tools or techniques are used to build confidence of correctness?
+({literate programming, automated testing, symbolic execution, model checking, assertions used in the code, Sphinx, Doxygen, Javadoc, confluence, unclear, other$\ast$}) $\ast$ given via string\\
+If there is a getting started tutorial? ({yes, no})\\
+Are the tutorial instructions linear? ({yes, no, n/a})\\
+Does the getting started tutorial provide an expected output? ({yes, no$\ast$, n/a})\\
+Does your tutorial output match the expected output? ({yes, no, n/a})\\
+Are unit tests available?  ({yes, no, unclear})\\
+Is there evidence of continuous integration? (for example mentioned in documentation, Jenkins, Travis CI, Bamboo, other) ({yes$\ast$, no, unclear})\\
+Overall impression? ({1 .. 10})\\
+Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have) \\
+\hline	
+\textbf{Surface Reliability}\\
+\hline
+Did the software “break” during installation? ({yes$\ast$ , no})\\
+If the software installation broke, was the installation instance recoverable? ({yes, no, n/a})\\
+Did the software “break” during the initial tutorial testing? ({yes$\ast$, no, n/a})\\
+If the tutorial testing broke, was a descriptive error message displayed? ({yes, no, n/a})\\
+If the tutorial testing broke, was the tutorial testing instance recoverable? ({yes, no, n/a})\\
+Overall impression? ({1 .. 10})\\
+Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have)\\
+\hline		
+\end{tabular}
+\end{table}
+
+\begin{table}[H]
+\centering
+\begin{tabular}{p{14cm}}
+\hline		
+\textbf{Surface Robustness}\\
+\hline
+Does the software handle unexpected/unanticipated input (like data of the wrong type, empty input, missing files or links) reasonably?
+(a reasonable response can include an appropriate error message.) ({yes, no$\ast$ })\\
+For any plain text input files, if all new lines are replaced with new lines and carriage returns, will the software handle this gracefully? ({yes, no$\ast$, n/a})\\
+Overall impression? ({1 .. 10})\\
+Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have)\\
+\hline		
+\textbf{Surface Usability}\\
+\hline
+Is there a getting started tutorial? ({yes, no})\\
+Is there a user manual? ({yes, no})\\
+Are expected user characteristics documented? ({yes, no})\\
+What is the user support model? FAQ? User forum? E-mail address to direct questions? Etc. (string)\\
+Overall impression? ({1 .. 10})\\
+Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have)\\
+\hline
+\end{tabular}
+\end{table}
+
+\begin{table}[H]
+\begin{tabular}{p{14cm}}
+\hline	
+\textbf{Maintainability}\\
+\hline
+What is the current version number? (number)\\
+Is there any information on how code is reviewed, or how to contribute? ({yes$\ast$, no})\\
+Are artifacts available?
+(List every type of file that is not a code file – for examples please look at the ‘Artifact Name’ column of https://gitlab.cas.mcmaster.ca/SEforSC/se4sc/-/blob/git-svn/GradStudents/Olu/ResearchProposal/Artifacts\_MiningV3.xlsx) ({yes$\ast$, no, unclear}) $\ast$list via string\\
+What issue tracking tool is employed? (set of {Trac, JIRA, Redmine, e-mail, discussion board, sourceforge, google code, git, BitBucket, none, unclear, other$\ast$}) $\ast$ given via string\\
+What is the percentage of identified issues that are closed? (percentage)\\
+What percentage of code is comments? (percentage)\\
+Which version control system is in use? ({svn, cvs, git, github, unclear, other$\ast$}) $\ast$ given via string\\
+Overall impression? ({1 .. 10})\\
+Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have)\\
+\hline		
+\textbf{Reusability}\\
+\hline
+How many code files are there? (number)\\
+Is API documented? ({yes, no, n/a})\\
+Overall impression? ({1 .. 10})\\
+Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have)\\
+\hline		
+\end{tabular}
+\end{table}
+
+\begin{table}[H]
+\begin{tabular}{p{14cm}}
+\hline		
+\textbf{Surface Understandability (Based on 10 random source files)}\\
+\hline
+Consistent indentation and formatting style? ({yes, no, n/a})\\
+Explicit identification of a coding standard? ({yes$\ast$, no, n/a})\\
+Are the code identifiers consistent, distinctive, and meaningful? ({yes, no$\ast$ , n/a})\\
+Are constants (other than 0 and 1) hard coded into the program? ({yes, no$\ast$ , n/a})\\
+Comments are clear, indicate what is being done, not how? ({yes, no$\ast$ , n/a})\\
+Is the name/URL of any algorithms used mentioned? ({yes, no$\ast$ , n/a})\\
+Parameters are in the same order for all functions? ({yes, no$\ast$ , n/a})\\
+Is code modularized? ({yes, no$\ast$ , n/a})\\
+Overall impression? ({1 .. 10})\\
+Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have)\\
+\hline		
+\textbf{Visibility/Transparency}\\
+\hline
+Is the development process defined? If yes, what process is used. ({yes$\ast$, no, n/a})\\
+Are there any documents recording the development process and status?  ({yes$\ast$, no}))\\
+Is the development environment documented? ({yes$\ast$, no})\\
+Are there release notes? ({yes$\ast$, no})\\
+Overall impression? ({1 .. 10})\\
+Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have)\\
+\hline
+\end{tabular}
+\end{table}
+
+\begin{table}[H]
+\begin{tabular}{p{14cm}}
+\hline		
+\textbf{Raw Metrics (Measured via git\_stats)}\\
+\hline
+Number of text-based files. (number)\\
+Number of binary files. (number)\\
+Number of total lines in text-based files. (number)\\
+Number of total lines added to text-based files. (number)\\
+Number of total lines deleted from text-based files. (number)\\
+Number of total commits. (number)\\
+Numbers of commits by year in the last 5 years. (Count from as early as possible if the project is younger than 5 years.) (list of numbers)\\
+Numbers of commits by month in the last 12 months. (list of numbers)\\
+\hline		
+\textbf{Raw Metrics (Measured via scc)}\\
+\hline
+Number of text-based files. (number)\\
+Number of total lines in text-based files. (number)\\
+Number of code lines in text-based files. (number)\\
+Number of comment lines in text-based files. (number)\\
+Number of blank lines in text-based files. (number)\\
+\hline
+\textbf{Repo Metrics (Measured via GitHub)}\\
+\hline
+Number of people watching this repo. (number)\\
+Number of stars. (number)\\
+Number of forks. (number)\\
+Number of open pull requests. (number)\\
+Number of closed pull requests. (number)\\
+Number of months on GitHub. (number)\\
+Accessed date. (date)\\
+\hline
+\end{tabular}
+\end{table}
+
+\endgroup

### 2021-07-27 StateOfPractice/Ao-Report/11Introduction.tex
MI report - updated ch 1 and ch 6
MI report - updated ch 6 and some other parts

@@
 
-As a result, the development process and quality of SC software concern us. We want to understand their status in SC domains and improve them.
+As a result, the development process and quality of SC software concern us.
+We want to understand their status in SC domains and improve them.
+We build and refine our methodology, based on our previous work in scientific domains such as oceanography \cite{Smith2015}, mesh generation \cite{smith2016state}, geographic information systems \cite{smith2018state}, psychology \cite{smith2018statistical} and seismology \cite{Smith2018Seismology}.
 
@@
 \begin{enumerate}
-\item What are the pain points for developers working on MI software projects? What are the solutions for these pain points?
+\item What are the pain points for developers working on MI software projects? What solutions the developers have tried to address the pain points?
 \item What artifacts the projects generated?
@@
 	\item Visibility/Transparency
+	\item Reproducibility
 \end{itemize}
-\item How does software quality ranking generated by our methods compare with the ratings from the community?
+\item How does the software quality ranking generated by our methods compare with the ratings from the community?
 \end{enumerate}

### 2021-07-27 StateOfPractice/Ao-Report/12Background.tex
MI report - updated ch 6 and some other parts

@@
 \cite{SmithEtAl2020}.
-The order of the qualities follows the grading template.
+The order of the first nine qualities follows the grading template.
+We do not measure \textit{reproducibility} with the grading template, but we discuss it with the developers by interviews.
 
@@
 	\label{def_understandability}
-	\item \textbf{Understandability} (To be completed)
+	\item \textbf{Understandability} The capability of the software product to enable the user to understand whether the software is suitable, and how it can be used for particular tasks and conditions of use.
 	\label{def_visibility_transparency}
 	\item \textbf{Visibility/Transparency} The extent to which all of the steps of a software development process and the current status of it are conveyed clearly.
+	\label{def_reproducibility}
+	\item \textbf{Reproducibility} A result is said to be reproducible if another researcher can take the original code and input data, execute it, and re-obtain the same result.
 \end{itemize}

### 2021-07-27 StateOfPractice/Ao-Report/13Methods.tex
MI report - updated ch 6 and some other parts

@@
 
-Additionally, we also manually collect information for projects held on GitHub, such as the numbers of stars, forks, people watching this repository, open pull requests, and closed pull requests.
+Additionally, we also manually collect information for projects held on GitHub, such as the numbers of stars, forks, people watching this repository, open pull requests, closed pull requests, number of months a repository has been on GitHub.
+A git repository can have a creation date much earlier than the first day on GitHub.
+For example, the developers created the git repository of \textit{3D Slicer} in 2002, but did not upload a copy of it to GitHub until 2020.
+We get the creation date of the GitHub copy by using API \textit{https://api.github.com/repos/{:owner}/{:repository}} (e.g.
+\hyperlink{https://api.github.com/repos/slicer/slicer}{https://api.github.com/repos/slicer/slicer}).
+In the response, the value of ``created\_at" is what we want.
+The number of months a repository has been on GitHub helps us to understand the average change of metrics over time, e.g.
+the average new stars per month.
 

### 2021-07-27 StateOfPractice/Ao-Report/14MeasurementResults.tex
MI report - updated ch 6 and some other parts

@@
 \section{Surface Robustness}
+\label{sec_result_robustness}
 Figure \ref{fg_robustness_scores} presents the scores for \textit{surface robustness}.
@@
 \section{Maintainability}
+\label{sec_score_maintainability}
 Figure \ref{fg_maintainability_scores} shows the AHP results for \textit{maintainability}.

### 2021-07-27 StateOfPractice/Ao-Report/15InterviewsWithDevelopers.tex
MI report - updated ch 6 and some other parts

@@
 \section{Current and Past Pain Points}
+\label{sec_interview_pain_points}
 
@@
 \section{Documents in the Projects}
+\label{sec_interview_documents}
 By asking questions 11 and 19, we tried to understand the interviewees' opinions on documentation. We also aimed to find out the quality of documentations in their projects.
@@
 \section{Contribution Management and Project Management}
+\label{sec_contribution_pm}
 By asking questions 5, 13, and 14, we tried to understand how the teams managed the contributions and their projects.
@@
 Most interviewees skipped the question ``Do you think improving this process can tackle the current problem?", but the interviewee from the \textit{OHIF} team gave a positive answer.
+
+\section{Discussions on Software Qualities}
+\label{sec_interview_software_qualities}
+Questions 15, 16, 17, 18, and 20 are about software qualities \textit{correctness}, \textit{maintainability}, \textit{understandability}, \textit{usability}, and \textit{reproducibility} respectively.
+We asked these questions to understand the threats to these qualities and the developers' strategies to improve them.
+
+\subsection{Correctness}
+\textit{Q15. Was it hard to ensure the correctness of the software? If there were any obstacles, what methods have been considered or practiced to improve the situation? If practiced, did it work?}
+
+Table \ref{tab_q15_threats_correctness} shows the threats to \textit{correctness} by the numbers of interviewees with the answers.
+
+\begin{table}[H]
+\centering
+\begin{tabular}{ll}
+\hline
+Threat to correctness & Number of interviewees with the answer \\ \hline
+\begin{tabular}[c]{@{}l@{}}Clinical systems produce data in various formats \\ (e.g. DICOM), which have complicated standards.\\ The software needs to handle the complexity.\end{tabular} & 2 \\
+\begin{tabular}[c]{@{}l@{}}Different types of MI machines can create data \\ in slightly different ways, adding more complexity\\ for the software to handle.\end{tabular} & 2 \\
+\begin{tabular}[c]{@{}l@{}}Besides viewing, the software has additional \\ functions, which lead to extra complexity.\end{tabular} & 1 \\
+There is lack of real word image data for testing. & 1 \\
+\begin{tabular}[c]{@{}l@{}}The team cannot use private data for debugging,\\ even when the data cause problems.\end{tabular} & 1 \\
+\begin{tabular}[c]{@{}l@{}}The software uses huge datasets for testing, so \\ the tests are expensive and time-consuming.\end{tabular} & 1 \\
+It is hard to well manage releases. & 1 \\
+The project has no unit tests. & 1 \\
+The project has no dedicated quality assurance team. & 1 \\ \hline
+\end{tabular}
+\caption{\label{tab_q15_threats_correctness}Threats to correctness by the numbers of interviewees with the answers}
+\end{table}
+
+Table \ref{tab_q15_strategies_correctness} shows the strategies to ensure \textit{correctness} by the numbers of interviewees with the answers.
+The interviewees from the \textit{3D Slicer} and \textit{ITK-SNAP} teams thought that the self-tests and automated tests were beneficial and could significantly save time.
+The interviewee from the \textit{Weasis} team kept collecting medical images for more than ten years.
+These images have caused problems with the software.
+So he had samples to test specific problems.
+
+\begin{table}[H]
+\centering
+\begin{tabular}{ll}
+\hline
+Strategy to ensure correctness & Number of interviewees with the answer \\ \hline
+\begin{tabular}[c]{@{}l@{}}Test-driven development / component tests /\\ integration tests / smoke tests / regression tests.\end{tabular} & 4 \\
+Self tests / automated tests. & 3 \\
+\begin{tabular}[c]{@{}l@{}}Two stage development process / stable release \&\\ nightly builds.\end{tabular} & 3 \\
+CI/CD. & 1 \\
+\begin{tabular}[c]{@{}l@{}}Using deidentified copies of medical images for\\ debugging.\end{tabular} & 1 \\
+\begin{tabular}[c]{@{}l@{}}Sending the beta versions of software to medical\\ workers who can access the data and do the tests.\end{tabular} & 1 \\
+\begin{tabular}[c]{@{}l@{}}Collecting and maintaining a dataset of\\ problematic images.\end{tabular} & 1 \\ \hline
+\end{tabular}
+\caption{\label{tab_q15_strategies_correctness}Strategies to ensure correctness by the numbers of interviewees with the answers}
+\end{table}
+
+\subsection{Maintainability}
+\textit{Q16.
+When designing the software, did you consider the ease of future changes?
+For example, will it be hard to change the system’s structure, modules, or code blocks?
+What measures have been taken to ensure the ease of future changes and maintains?}
+
+Table \ref{tab_q16_strategies_maintainability} shows the strategies to ensure \textit{maintainability} by the numbers of interviewees with the answers.
+The modular approach is the most talked-about solution to improve \textit{maintainability}.
+The \textit{3D Slicer} team used a well-defined structure for the software, which they named as ``event-driven MVC pattern''.
+Moreover, \textit{3D Slicer} discovers and loads necessary modules at runtime, according to the configuration and installed extensions.
+The \textit{BioImage Suite Web} team had designed and re-designed their software multiple times in the last 10+
+years. They found that their modular approach effectively supported the maintainability \cite{Joshi2011}.
+
+\begin{table}[H]
+\centering
+\begin{tabular}{ll}
+\hline
+Strategy to ensure maintainability & Number of interviewees with the answer \\ \hline
+\begin{tabular}[c]{@{}l@{}}Modular approach / thinking the software\\ as reusable Lego bricks / maintain \\ repetitive functions as libraries.\end{tabular} & 5 \\
+Supporting third-party extensions. & 1 \\
+Easy-to-understand architecture. & 1 \\
+Dedicated architect. & 1 \\
+Starting from simple solutions. & 1 \\
+Documentation. & 1 \\ \hline
+\end{tabular}
+\caption{\label{tab_q16_strategies_maintainability}Strategies to ensure maintainability by the numbers of interviewees with the answers}
+\end{table}
+
+\subsection{Understandability}
+\textit{Q17. Provide instances where users have misunderstood the software. What, if any, actions were taken to address understandability issues?}
+
+Table \ref{tab_q17_threats_understandability} shows the threats to \textit{understandability} by the numbers of interviewees with the answers.
+It separates \textit{understandability} issues to users and developers by the horizontal dash line.
+
+\begin{table}[H]
+\centering
+\begin{tabular}{ll}
+\hline
+Threat to understandability & Number of interviewees with the answer \\ \hline
+\begin{tabular}[c]{@{}l@{}}Not all users understand how to use some\\ features of the software.\end{tabular} & 2 \\
+\begin{tabular}[c]{@{}l@{}}The team has no dedicated user experience\\ (UX) designer.\end{tabular} & 1 \\
+\begin{tabular}[c]{@{}l@{}}The software does not make some important\\ indicators noticeable (e.g. a progress bar).\end{tabular} & 1 \\
+\begin{tabular}[c]{@{}l@{}}Not all users understand the purpose of the\\ software.\end{tabular} & 1 \\
+\begin{tabular}[c]{@{}l@{}}Not all users know if the software includes\\ certain features.\end{tabular} & 1 \\
+\begin{tabular}[c]{@{}l@{}}Not all users understand how to use the\\ command line tool.\end{tabular} & 1 \\
+\begin{tabular}[c]{@{}l@{}}Not all users understand that the software is \\a web application.\end{tabular} & 1 \\\hdashline
+\begin{tabular}[c]{@{}l@{}}Not all developers understand how to deploy\\ the software.\end{tabular} & 1 \\
+\begin{tabular}[c]{@{}l@{}}The architecture is difficult for new\\ developers to understand.\end{tabular} & 1 \\ \hline
+\end{tabular}
+\caption{\label{tab_q17_threats_understandability}Threats to understandability by the numbers of interviewees with the answers}
+\end{table}
+
+Table \ref{tab_q17_strategies_understandability} shows the strategies to ensure \textit{understandability} by the numbers of interviewees with the answers.
+
+\begin{table}[H]
+\centering
+\begin{tabular}{ll}
+\hline
+Strategy to ensure understandability & Number of interviewees with the answer \\ \hline
+\begin{tabular}[c]{@{}l@{}}Documentation / user manual /\\ user mailing list / forum.\end{tabular} & 4 \\
+Graphical user interface. & 2 \\
+Testing every release with active users. & 1 \\
+\begin{tabular}[c]{@{}l@{}}Making simple things simple and complicated\\ things possible.\end{tabular} & 1 \\
+Icons with more clear visual expressions. & 1 \\
+Designing the software to be intuitive. & 1 \\
+Having a UX designer with the right experience. & 1 \\
+Dialog windows for important notifications. & 1 \\
+\begin{tabular}[c]{@{}l@{}}Providing an example if the users need to build\\ the software by themselves.\end{tabular} & 1 \\ \hline
+\end{tabular}
+\caption{\label{tab_q17_strategies_understandability}Strategies to ensure understandability by the numbers of interviewees with the answers}
+\end{table}
+
+\subsection{Usability}
+\textit{Q18. What, if any, actions were taken to address usability issues?}
+
+Table \ref{tab_q18_strategies_usability} shows the strategies to ensure \textit{usability} by the numbers of interviewees with the answers.
+
+\begin{table}[H]
+\centering
+\begin{tabular}{ll}
+\hline
+Strategy to ensure usability & Number of interviewees with the answer \\ \hline
+Usability tests and interviews with end users. & 3 \\
+Adjusting according to users' feedbacks. & 3 \\
+\begin{tabular}[c]{@{}l@{}}Straightforward and intuitively designed interface /\\ professional UX designer.\end{tabular} & 2 \\
+\begin{tabular}[c]{@{}l@{}}Providing step-by-step processes, and showing\\ the step numbers.\end{tabular} & 1 \\
+\begin{tabular}[c]{@{}l@{}}Making the basic functions easy to use without\\ reading the documentation.\end{tabular} & 1 \\
+Focusing on limited number of functions. & 1 \\
+Making the software more streamlined. & 1 \\
+Downsampling images to consume less memory. & 1 \\
+\begin{tabular}[c]{@{}l@{}}An option to load only part of the data to boost\\ performance.\end{tabular} & 1 \\ \hline
+\end{tabular}
+\caption{\label{tab_q18_strategies_usability}Strategies to ensure usability by the numbers of interviewees with the answers}
+\end{table}
+
+\subsection{Reproducibility}
+\textit{Q20. Do you have any concern that your computational results won’t be reproducible in the future? Have you taken any steps to ensure reproducibility?}
+
+Table \ref{tab_q20_threats_reproducibility} shows the threats to \textit{reproducibility} by the numbers of interviewees with the answers.
+
+\begin{table}[H]
+\centering
+\begin{tabular}{ll}
+\hline
+Threat to reproducibility & Number of interviewees with the answer \\ \hline
+\begin{tabular}[c]{@{}l@{}}If the software is closed-source, the reproducibility\\is hard to achieve.\end{tabular} & 1 \\
+The project has no user interaction tests. & 1 \\
+The project has no unit tests. & 1 \\
+\begin{tabular}[c]{@{}l@{}}Using different versions of some common libraries\\may cause problems.\end{tabular} & 1 \\
+CPU variability can leads to non-reproducibility. & 1 \\
+\begin{tabular}[c]{@{}l@{}}When reverse-engineering how manufacturers create\\medical images, the team may misinterpret it.\end{tabular} & 1 \\ \hline
+\end{tabular}
+\caption{\label{tab_q20_threats_reproducibility}Threats to reproducibility by the numbers of interviewees with the answers}
+\end{table}
+
+Table \ref{tab_q20_strategies_reproducibility} shows the strategies to ensure \textit{ reproducibility} by the numbers of interviewees with the answers.
+The interviewee from the \textit{3D Slicer} team provided various suggestions.
+One interviewee from another team suggested that they used \textit{3D Slicer} as the benchmark to test their \textit{reproducibility}.
+
+\begin{table}[H]
+\centering
+\begin{tabular}{ll}
+\hline
+Strategy to ensure reproducibility & Number of interviewees with the answer \\ \hline
+Regression tests / unit tests / having good tests. & 6 \\
+\begin{tabular}[c]{@{}l@{}}Making code, data, and documentation available / making\\ the software open-source / using open-source libraries.\end{tabular} & 5 \\
+Running same tests on all platforms. & 1 \\
+\begin{tabular}[c]{@{}l@{}}A dockerized version of the software,  insulating it from\\ the operating system environment.\end{tabular} & 1 \\
+Using standard libraries. & 1 \\
+Monitoring the upgrades of the libraries. & 1 \\
+Clearly documenting the versions. & 1 \\
+\begin{tabular}[c]{@{}l@{}}Bringing along the exact versions of all the dependencies\\ with the software.\end{tabular} & 1 \\
+Providing checksums of the data. & 1 \\
+\begin{tabular}[c]{@{}l@{}}Benchmark the software against other software with\\ similar purposes.\end{tabular} & 1 \\ \hline
+\end{tabular}
+\caption{\label{tab_q20_strategies_reproducibility}Strategies to ensure  reproducibility by the numbers of interviewees with the answers}
+\end{table}

### 2021-07-27 StateOfPractice/Ao-Report/16AnswersToResearchQuestions.tex
MI report - updated ch 1 and ch 6
MI report - updated ch 6 and some other parts

--- 2021-07-26 StateOfPractice/Ao-Report/16ThreatToValidity.tex
@@
-\chapter{Threat to Validity}
-\label{ch_validity}
+\chapter{Answers to Research Questions}
+\label{ch_answers}
 
+This section answers our research questions in Section \ref{sec_research_questions}.
+The answers are based on our quality measurements and developer interviews.
+Sections \ref{sec_rq1_pain_points}, \ref{sec_rq2_artifacts}, \ref{sec_rq3_documentation}, \ref{sec_rq4_PPMT}, \ref{sec_rq5_qualities}, and \ref{sec_rq6_comparison} summarize the answers to the six questions respectively.
+Section \ref{sec_threats_to_validity} lists the threats to the validity of our research.
+
+\section{Pain Points and Solutions}
+\label{sec_rq1_pain_points}
+\textit{RQ1. What are the pain points for developers working on MI software projects? What solutions have the developers tried to address the pain points?}
+
+We answer this question primarily by interviewing the developers. As described in Section \ref{sec_interview_pain_points}, the top three pain points are as follows,
+\begin{itemize}
+\item the lack of fundings and time;
+\item the difficulty to balance between four factors: cross-platform compatibility, convenience to development \& maintenance, performance, and security;
+\item the lack of access to real-world datasets for testing.
+\end{itemize}
+
+Section \ref{sec_interview_pain_points} also lists potential and proven solutions from the developers.
+
+\section{Artifacts in the Projects}
+\label{sec_rq2_artifacts}
+\textit{RQ2. What artifacts the projects generated?}
+
+We answer this question by searching and scanning the source code and documentation of the projects.
+The complete list of artifacts is in the repository \hyperlink{https://github.com/smiths/AIMSS}{https://github.com/smiths/AIMSS}.
+
+Part of the answer is in Section \ref{ch_results}.
+Table \ref{tab_maintainability_docs} shows which projects contained a project plan, development manual, or API documentation, and Table \ref{tab_Visibility/Transparency_docs} lists the projects with a documented development process, project status, development environment, or release notes.
+
+Table \ref{tab_artifacts_frequency} shows the frequency of some artifacts in the 29 MI projects.
+
+\begin{table}[H]
+\centering
+\begin{tabular}{ll}
+\hline
+Artifact & Number of projects \\ \hline
+Source code & 29 \\
+Version Control & 29 \\
+README & 29 \\
+License & 28 \\
+Bug tracker & 28 \\
+Change request & 28 \\
+User Manual & 22 \\
+Release notes & 22 \\
+Build file & 18 \\
+Tutorials & 18 \\
+Makefile & 18 \\
+Installation Guide & 16 \\
+Test cases & 15 \\
+Authors & 14 \\
+FAQ & 14 \\
+Acknowledgements & 12 \\
+Executable files & 10 \\
+Developer's Manual & 8 \\
+API documentation & 7 \\
+Troubleshooting guide & 6 \\
+Project Plan & 5 \\ \hline
+\end{tabular}
+\caption{\label{tab_artifacts_frequency}Artifacts by their frequency in the 29 MI projects}
+\end{table}
+
+\section{Documentation in the Projects}
+\label{sec_rq3_documentation}
+\textit{RQ3. What role does documentation play in the projects? What are the developers' attitudes toward it?}
+
+We answer this question by interviewing the developers.
+As shown in Section \ref{sec_interview_documents}, most of the nine interviewees thought that documentation was essential to their projects.
+However, they hold the common opinion that their documentation needed improvements.
+Nearly half of them also believed that the lack of time prevented them from improving the documentation.
+
+\section{Principles, Processes, Methodologies, and Tools in the Projects}
+\label{sec_rq4_PPMT}
+\textit{RQ4. What principles, processes, methodologies, and tools the projects used?}
+
+We answer this question by measuring the qualities and interviewing the developers.
+
+Table \ref{tab_maintainability_docs}, \ref{tab_Visibility/Transparency_docs}, and \ref{tab_artifacts_frequency} reveal part of the answer, such as which and how many teams used a project plan, API documentation, release notes, version control, etc.
+
+We summarize the other part of the answer from the interviews.
+Sections \ref{sec_interview_documents} and \ref{sec_contribution_pm} list some principles, processes, methodologies, and tools for documentation, contribution management, and project management.
+As shown in Section \ref{sec_interview_software_qualities}, the developers also shared some of their principles, processes, methodologies, and tools to improve \textit{correctness}, \textit{maintainability}, \textit{understandability}, \textit{usability}, and \textit{reproducibility}.
+
+\section{Software Qualities}
+\label{sec_rq5_qualities}
+\textit{RQ5. What is the current status of the following software qualities for the projects? What actions have the developers taken to address them?}
+
+We answer this question by measuring the qualities and interviewing the developers.
+Section \ref{ch_results} contains our measurement results for \textit{Installability}, \textit{Correctness \& Verifiability}, \textit{Reliability}, \textit{Robustness}, \textit{Usability}, \textit{Maintainability}, \textit{Reusability}, \textit{Understandability}, and \textit{Visibility/Transparency}; Section \ref{sec_interview_software_qualities} shows results from the interviews for \textit{correctness}, \textit{maintainability}, \textit{understandability}, \textit{usability}, and \textit{reproducibility}.
+
+We list some of the key points that may improve each quality as follows,
+\begin{itemize}
+\item \textbf{Installability}
+\begin{itemize}
+    \item clear instructions;
+    \item automated installer;
+    \item including all dependencies in the installer;
+    \item avoiding heavily depending on other commercial products (e.g. Matlab);
+    \item considering building a web application that needs no installation.
+\end{itemize}
+\item \textbf{Correctness \& Verifiability}
+\begin{itemize}
+    \item test-driven development with unit tests, integration tests, nightly tests, etc.
+    \item two stage development process with stable release \& nightly builds;
+    \item CI/CD;
+    \item documents of requirements specifications and theory manuals.
+\end{itemize}
+\item \textbf{Reliability}
+\begin{itemize}
+    \item test-driven development with unit tests, integration tests, nightly tests, etc.
+    \item two stage development process with stable release \& nightly builds;
+    \item descriptive error messages.
+\end{itemize}
+\item \textbf{Robustness}
+\begin{itemize}
+    \item designing with exceptions and make the software failures elegant;
+    \item descriptive error messages.
+\end{itemize}
+\item \textbf{Usability}
+\begin{itemize}
+    \item usability tests and interviews with end users;
+    \item adjusting according to users’ feedbacks;
+    \item getting started tutorials;
+    \item user manuals;
+    \item professional UX designs;
+    \item active supports to users.
+\end{itemize}
+\item \textbf{Maintainability}
+\begin{itemize}
+    \item using GitHub;
+    \item modular approach;
+    \item documentation for developers, such as project plan, developer’s manual, and API documentation.
+\end{itemize}
+\item \textbf{Reusability}
+\begin{itemize}
+    \item modular approach;
+    \item API documentation.
+\end{itemize}
+\item \textbf{Understandability}
+\begin{itemize}
+    \item modular approach;
+    \item consistent codding style;
+    \item clear comments;
+    \item description of used algorithms;
+    \item documentation;
+    \item communication between developers and users via GitHub issues, mailing lists, forums etc.
+    \item graphical user interface.
+\end{itemize}
+\item \textbf{Visibility/Transparency}
+\begin{itemize}
+    \item documents for the development process, project status, development environment, and release notes.
+\end{itemize}
+\item \textbf{Reproducibility}
+\begin{itemize}
+    \item test-driven development with unit tests, integration tests, nightly tests, etc.
+    \item open-source;
+    \item making data and documentation available;
+    \item using open-source libraries.
+\end{itemize}
+\end{itemize}
+
+\section{Our Ranking versus the Community Ratings}
+\label{sec_rq6_comparison}
+\textit{RQ6. How does the software quality ranking generated by our methods compare with the ratings from the community?}
+
+We answer this question by grading the qualities of the software, then comparing the ranking with the community ratings on GitHub, such as GitHub stars, number of forks, and number of people watching the projects.
+
+Table \ref{tab_ranking_vs_GitHub} shows our ranking to the 29 MI projects, and their GitHub metrics if applicable.
+As mentioned in Section \ref{sec_score_maintainability}, 24 projects used GitHub.
+Since GitHub repositories have different creation dates, we collect the number of months each stayed on GitHub, and calculate the average number of new stars, people watching, and forks per 12 months.
+The method of getting the creation date is described in Section \ref{sec_empirical_measurements}, and we obtained these metrics in July, 2021.
+
+Generally speaking, most of the top-ranking MI software projects also received greater attention and popularity on GitHub.
+Project \textit{dwv} was popular on GitHub, but we ranked it low.
+As mentioned in Section \ref{sec_result_installability}, we failed to build it locally, and used the test version on its websites for the measurements.
+We followed the instructions and tried to run the command ``yarn run test" locally, which did not work.
+In addition, the test version did not detect a broken DICOM file and displayed a blank image as described in Section \ref{sec_result_robustness}.
+We might underestimate the scores for \textit{dwv} due to uncommon technical issues.
+
+\begingroup
+\renewcommand{\arraystretch}{0.85}
+\begin{table}[H]
+\centering
+\begin{tabular}{lllll}
+\hline
+Software & Our ranking & Stars/yr & Watch/yr & Forks/yr \\ \hline
+3D Slicer & 1 & 284.25 & 18.75 & 128.25 \\
+ImageJ & 2 & 83.58 & 9.37 & 30.00 \\
+OHIF Viewer & 3 & 277.04 & 19.30 & 223.83 \\
+Fiji & 4 & 44.20 & 4.98 & 20.68 \\
+ParaView & 5 & 66.76 & 7.11 & 28.36 \\
+Weasis & 6 & 36.00 & 5.10 & 18.90 \\
+medInria & 7 & 7.04 & 3.35 & 6.35 \\
+BioImage Suite Web & 8 & 17.85 & 4.62 & 6.77 \\
+OsiriX Lite & 9 & 34.14 & 8.62 & 23.77 \\
+INVESALIUS 3 & 10 & 39.66 & 4.11 & 17.37 \\
+Gwyddion & 11 & n/a & n/a & n/a \\
+Horos & 12 & 48.62 & 9.23 & 18.00 \\
+SMILI & 13 & 3.04 & 0.91 & 1.52 \\
+Ginkgo CADx & 14 & 18.83 & 4.43 & 5.72 \\
+ITK-SNAP & 15 & 9.10 & 0.97 & 3.59 \\
+MicroView & 16 & 1.48 & 0.82 & 0.82 \\
+Papaya & 17 & 44.70 & 4.84 & 19.71 \\
+MRIcroGL & 18 & 23.63 & 3.38 & 3.38 \\
+XMedCon & 19 & n/a & n/a & n/a \\
+Slice:Drop & 20 & 9.73 & 2.16 & 4.97 \\
+DicomBrowser & 21 & n/a & n/a & n/a \\
+AMIDE & 22 & n/a & n/a & n/a \\
+dwv & 23 & 123.56 & 11.59 & 51.36 \\
+3DimViewer & 24 & n/a & n/a & n/a \\
+GATE & 25 & 18.90 & 5.79 & 25.79 \\
+dicompyler & 26 & 35.18 & 4.77 & 13.64 \\
+Drishti & 27 & 16.04 & 3.78 & 4.43 \\
+MatrixUser & 28 & 2.00 & 0.33 & 0.33 \\
+DICOM Viewer & 29 & 42.86 & 5.71 & 8.57 \\ \hline
+\end{tabular}
+\caption{\label{tab_ranking_vs_GitHub}Software ranking versus GitHub metrics}
+\end{table}
+\endgroup
+
+\section{Threats to Validity}
+\label{sec_threats_to_validity}
 This section lists all the potential threats to the validity of our research.
@@
 Since we did not have the chance to explain our questions or ask him follow-up questions, there is a possibility of misinterpretation of the questions or answers.
+\item As mentioned in Section \ref{sec_rq6_comparison}, we gave \textit{dwv} much lower scores than its GitHub popularity. We might underestimate its rank due to uncommon technical issues.
 \end{enumerate}

### 2021-07-27 StateOfPractice/Ao-Report/33InterviewAnswers.tex
MI report - updated ch 6 and some other parts

--- 2021-05-09 StateOfPractice/Ao-Report/33InterviewAnswers.tex
@@
-\chapter{Interview Answers}
+\chapter{Other Interview Answers}
 \label{ap_interview}
 
-appendix here
+We asked 20 interview questions to the nine interviewees from eight software projects.
+We discuss the answers to interview questions 5, 9, 10, 11, 12, 13, 14, and 19 in Section \ref{ch_interview}, and summarize the answers to the other questions in this section.
+
+\noindent\textbf{Q1. Interviewees’ current position/title? degrees?}
+
+Six of the nine interviewees revealed their position/title, such as CEO of a company, endowed chair and professor in universities, software engineers in a commercial company and a hospital.
+
+Most of them answered their backgrounds and degrees.
+Table \ref{tab_q1_degrees} shows the highest academic degrees the participants have, and Table \ref{tab_q1_majors} shows what majors they studied.
+Many of the interviewees studied in multiple majors.
+
+\begin{table}[H]
+\centering
+\begin{tabular}{ll}
+\hline
+Highest degree & Number of interviewees \\ \hline
+PHD & 4 \\
+Master & 3 \\
+Bachelor & 0 \\
+Unspecified academic degree & 2 \\ \hline
+\end{tabular}
+\caption{\label{tab_q1_degrees}Interviewees' highest academic degrees}
+\end{table}
+
+\begin{table}[H]
+\centering
+\begin{tabular}{ll}
+\hline
+Major & Number of interviewees \\ \hline
+Computer Science & 4 \\
+Physics & 2 \\
+Biomedical Engineering & 1 \\
+Neuroimaging & 1 \\
+Geology (image analysis) & 1 \\
+Media Arts and Sciences & 1 \\
+Mechanical Engineering & 1 \\
+Materials Engineering & 1 \\
+Psychology & 1 \\ \hline
+\end{tabular}
+\caption{\label{tab_q1_majors}Interviewees' majors at university}
+\end{table}
+
+\noindent\textbf{Q2. Interviewees’ contribution to/relationship with the software?}
+
+Table \ref{tab_q2_roles} shows the interviewees’ roles and responsibilities in the projects.
+One of the participants did not explicitly mention his role, but implicitly revealed that he was a primary contributor to the project.
+
+\begin{table}[H]
+\centering
+\begin{tabular}{ll}
+\hline
+Role in the projects & Number of interviewees \\ \hline
+Chief Architect & 2 \\
+Lead Developer & 1 \\
+Core Developer & 5 \\
+Unspecified & 1 \\ \hline
+\end{tabular}
+\caption{\label{tab_q2_roles}Interviewees' roles in the projects}
+\end{table}
+
+\noindent\textbf{Q3. Length of time the interviewee has been involved with this software?}
+
+Table \ref{tab_q3_years} shows the distribution of the lengths of time the interviewees had worked on the projects.
+
+\begin{table}[H]
+\centering
+\begin{tabular}{ll}
+\hline
+Length of time in the projects & Number of interviewees \\ \hline
+0-1 & 1 \\
+2-5 & 0 \\
+6-10 & 2 \\
+11-15 & 3 \\
+16-20 & 2 \\
+21-25 & 1 \\ \hline
+\end{tabular}
+\caption{\label{tab_q3_years}Lengths of time that the interviewees worked in the projects}
+\end{table}
+
+\noindent\textbf{Q4. How large is the development group?}
+
+The size of each group grows and shrinks over the years.
+Most teams mentioned that the team members join and leave.
+Some teams said that when there was sufficient funding, they could afford more developers.
+
+Table \ref{tab_q4_members} shows the numbers of active members at the time of interviews. The members include people working on development and project management.
+
+\begin{table}[H]
+\centering
+\begin{tabular}{ll}
+\hline
+Number of current members & Number of projects \\ \hline
+1-3 & 5 \\
+4-6 & 3 \\ \hline
+\end{tabular}
+\caption{\label{tab_q4_members}Numbers of current members in the projects}
+\end{table}
+
+As shown in the table, no team had a vast number of members.
+Some projects had more developers, such as \textit{3D Slicer}; on the other hand, some teams such as \textit{dwv} had only one primary developer, plus a maximum of two or three developers occasionally.
+
+\textit{3D Slicer} is a special case, because it supports third-party extensions.
+So there have been community members developing and maintaining these extensions.
+Table \ref{tab_q4_members} does not include these members.
+
+\noindent\textbf{Q6. What is the typical background of a developer?}
+
+Not all interviewees could clearly answer this question.
+Many of them talked about the backgrounds of members with who they were familiar.
+Table \ref{tab_q6_dev_backgrounds} shows the number of times all interviewees mentioned a background.
+
+\begin{table}[H]
+\centering
+\begin{tabular}{ll}
+\hline
+Background of a developer & Number of interviewees with the answer \\ \hline
+\begin{tabular}[c]{@{}l@{}}Computer Science, Information\\ Technology, and Software Development \end{tabular} & 6 \\
+Imaging & 2 \\
+Medical Imaging & 1 \\
+Mathematics & 1 \\
+Biomedical Engineering & 1 \\
+Computer Aided Medical Procedures & 1 \\
+Physician & 1 \\ \hline
+\end{tabular}
+\caption{\label{tab_q6_dev_backgrounds}Backgrounds of developers by the numbers of interviewees with the answers}
+\end{table}
+
+\noindent\textbf{Q7. What is your estimated number of users? How did you come up with that estimate?}
+
+None of the interviewees knew the exact number of users. Some of them provided estimations based on different facts. However, we do not think these numbers are comparable to each other.
+
+\begin{table}[H]
+\centering
+\begin{tabular}{lll}
+\hline
+Software & Rough estimation & Considered facts \\ \hline
+3D Slicer & 100,000 & \begin{tabular}[c]{@{}l@{}} Search results on Google Scholar;\\ number of new posts per year on slicer.org; \\ number of downloads. \end{tabular} \\
+INVESALIUS 3 & 75,000 & Number of random IDs created by new installation.\\
+dwv & No estimation & About 20 companies integrated \textit{dwv} in their products. \\
+BioImage Suite Web & 100 active users & \begin{tabular}[c]{@{}l@{}}  The  interviewee only counted the users from \\ several Universities who were active users. \end{tabular} \\
+ITK-SNAP & 10,000 plus & Number of downloads. \\
+MRIcroGL & No estimation & \begin{tabular}[c]{@{}l@{}}  It is the top 1 downloaded software on this NITRC list \\ \hyperlink{https://www.nitrc.org/top/toplist.php?type=downloads}{https://www.nitrc.org/top/toplist.php?type=downloads} \end{tabular} \\
+Weasis & \begin{tabular}[c]{@{}l@{}}  10,000 user used \\ it as least once \end{tabular} & Number of profiles. \\
+OHIF & About 5000 & \begin{tabular}[c]{@{}l@{}}   Some platforms integrated \textit{OHIF}, and it was hard \\ to know the number of end users. \end{tabular} \\ \hline
+\end{tabular}
+\caption{\label{tab_q7_num_users}Rough Estimations for the Number of Users}
+\end{table}
+
+Table \ref{tab_q7_num_users} shows the estimations and how the interviewees made them.
+It is clear that some estimated only the active users, and some counted users who had used only once.
+So we do not compare these numbers.
+
+\noindent\textbf{Q8. What is the typical background of a user?}
+
+All interviewees provided several different user backgrounds, and all of them mentioned medical researchers or medical professionals.
+Table \ref{tab_q8_user_backgrounds} shows the number of times all interviewees mentioned a background.
+
+\begin{table}[H]
+\centering
+\begin{tabular}{ll}
+\hline
+Background of a user & Number of interviewees with the answer \\ \hline
+Medical Researchers & 6 \\
+Doctors/Health care professionals/Surgeons & 5 \\
+Student Researchers & 4 \\
+Patients & 3 \\
+Paleontologist & 1 \\
+Biomechanical Engineers & 1 \\
+Imaging Researchers & 1 \\
+Mechanical Engineers & 1 \\ \hline
+\end{tabular}
+\caption{\label{tab_q8_user_backgrounds}Backgrounds of users by the numbers of interviewees with the answers}
+\end{table}

### 2021-07-27 StateOfPractice/Ao-Report/16AnswersToResearchQuestions.tex
MI report - moved some words from ch 6 to ch 7

@@
 We answer this question by measuring the qualities and interviewing the developers.
-Section \ref{ch_results} contains our measurement results for \textit{Installability}, \textit{Correctness \& Verifiability}, \textit{Reliability}, \textit{Robustness}, \textit{Usability}, \textit{Maintainability}, \textit{Reusability}, \textit{Understandability}, and \textit{Visibility/Transparency}; Section \ref{sec_interview_software_qualities} shows results from the interviews for \textit{correctness}, \textit{maintainability}, \textit{understandability}, \textit{usability}, and \textit{reproducibility}.
 
-We list some of the key points that may improve each quality as follows,
-\begin{itemize}
-\item \textbf{Installability}
-\begin{itemize}
-    \item clear instructions;
-    \item automated installer;
-    \item including all dependencies in the installer;
-    \item avoiding heavily depending on other commercial products (e.g. Matlab);
-    \item considering building a web application that needs no installation.
-\end{itemize}
-\item \textbf{Correctness \& Verifiability}
-\begin{itemize}
-    \item test-driven development with unit tests, integration tests, nightly tests, etc.
-    \item two stage development process with stable release \& nightly builds;
-    \item CI/CD;
-    \item documents of requirements specifications and theory manuals.
-\end{itemize}
-\item \textbf{Reliability}
-\begin{itemize}
-    \item test-driven development with unit tests, integration tests, nightly tests, etc.
-    \item two stage development process with stable release \& nightly builds;
-    \item descriptive error messages.
-\end{itemize}
-\item \textbf{Robustness}
-\begin{itemize}
-    \item designing with exceptions and make the software failures elegant;
-    \item descriptive error messages.
-\end{itemize}
-\item \textbf{Usability}
-\begin{itemize}
-    \item usability tests and interviews with end users;
-    \item adjusting according to users’ feedbacks;
-    \item getting started tutorials;
-    \item user manuals;
-    \item professional UX designs;
-    \item active supports to users.
-\end{itemize}
-\item \textbf{Maintainability}
-\begin{itemize}
-    \item using GitHub;
-    \item modular approach;
-    \item documentation for developers, such as project plan, developer’s manual, and API documentation.
-\end{itemize}
-\item \textbf{Reusability}
-\begin{itemize}
-    \item modular approach;
-    \item API documentation.
-\end{itemize}
-\item \textbf{Understandability}
-\begin{itemize}
-    \item modular approach;
-    \item consistent codding style;
-    \item clear comments;
-    \item description of used algorithms;
-    \item documentation;
-    \item communication between developers and users via GitHub issues, mailing lists, forums etc.
-    \item graphical user interface.
-\end{itemize}
-\item \textbf{Visibility/Transparency}
-\begin{itemize}
-    \item documents for the development process, project status, development environment, and release notes.
-\end{itemize}
-\item \textbf{Reproducibility}
-\begin{itemize}
-    \item test-driven development with unit tests, integration tests, nightly tests, etc.
-    \item open-source;
-    \item making data and documentation available;
-    \item using open-source libraries.
-\end{itemize}
-\end{itemize}
+Section \ref{ch_results} contains our measurement results for \textit{Installability}, \textit{Correctness \& Verifiability}, \textit{Reliability}, \textit{Robustness}, \textit{Usability}, \textit{Maintainability}, \textit{Reusability}, \textit{Understandability}, and \textit{Visibility/Transparency}.
+We applied (surface) measurements to the nine qualities, and hope that the grading scores represent their current status.
+We believe that our assessments also revealed some of the development teams' practices addressing the qualities.
+
+Section \ref{sec_interview_software_qualities} shows findings from the interviews for \textit{correctness}, \textit{maintainability}, \textit{understandability}, \textit{usability}, and \textit{reproducibility}.
+The interviewees expressed their thoughts on these five qualities.
+We discussed the current or past threats to them, and found out what actions they believed would improve the qualities.
 

### 2021-07-27 StateOfPractice/Ao-Report/17Recommendations.tex
MI report - moved some words from ch 6 to ch 7

@@
 
-I think the recommendations can originate from both parts - measurements and interviews.
+\section{Recommendations to Software Qualities}
+
+We list some of the key points that may improve each quality as follows,
+\begin{itemize}
+\item \textbf{Installability}
+\begin{itemize}
+    \item clear instructions;
+    \item automated installer;
+    \item including all dependencies in the installer;
+    \item avoiding heavily depending on other commercial products (e.g. Matlab);
+    \item considering building a web application that needs no installation.
+\end{itemize}
+\item \textbf{Correctness \& Verifiability}
+\begin{itemize}
+    \item test-driven development with unit tests, integration tests, nightly tests, etc.
+    \item two stage development process with stable release \& nightly builds;
+    \item CI/CD;
+    \item documents of requirements specifications and theory manuals.
+\end{itemize}
+\item \textbf{Reliability}
+\begin{itemize}
+    \item test-driven development with unit tests, integration tests, nightly tests, etc.
+    \item two stage development process with stable release \& nightly builds;
+    \item descriptive error messages.
+\end{itemize}
+\item \textbf{Robustness}
+\begin{itemize}
+    \item designing with exceptions and make the software failures elegant;
+    \item descriptive error messages.
+\end{itemize}
+\item \textbf{Usability}
+\begin{itemize}
+    \item usability tests and interviews with end users;
+    \item adjusting according to users’ feedbacks;
+    \item getting started tutorials;
+    \item user manuals;
+    \item professional UX designs;
+    \item active supports to users.
+\end{itemize}
+\item \textbf{Maintainability}
+\begin{itemize}
+    \item using GitHub;
+    \item modular approach;
+    \item documentation for developers, such as project plan, developer’s manual, and API documentation.
+\end{itemize}
+\item \textbf{Reusability}
+\begin{itemize}
+    \item modular approach;
+    \item API documentation.
+\end{itemize}
+\item \textbf{Understandability}
+\begin{itemize}
+    \item modular approach;
+    \item consistent codding style;
+    \item clear comments;
+    \item description of used algorithms;
+    \item documentation;
+    \item communication between developers and users via GitHub issues, mailing lists, forums etc.
+    \item graphical user interface.
+\end{itemize}
+\item \textbf{Visibility/Transparency}
+\begin{itemize}
+    \item documents for the development process, project status, development environment, and release notes.
+\end{itemize}
+\item \textbf{Reproducibility}
+\begin{itemize}
+    \item test-driven development with unit tests, integration tests, nightly tests, etc.
+    \item open-source;
+    \item making data and documentation available;
+    \item using open-source libraries.
+\end{itemize}
+\end{itemize}

### 2021-07-27 StateOfPractice/Peter-Notes/Report/Report.tex
updated LBM report Abstract

@@
 
-We analyze the state of the practice of software development of Lattice Boltzmann Methods software domain by quantitatively and qualitatively measuring, and comparing, 23 software packages.
-A general methodology for assessing the state of the practice of software development in scientific computing software domains is presented.
-A domain analysis of the Lattice Boltzmann Methods software family is made, and candidate software packages are identified and curated.
-The packages are assessed to answer software development related research questions to understand how software quality is impacted by software development choices, including principles, processes, and tools.
-Software developers are interviewed to identify development pain points, and to identify how software quality is ensured.
-Quantitative data is used to rank the software packages using the Analytical Hierarchy Process.
-The packages Ludwig, ESPResSo, and Palabos are ranked as the top three packages.
-Each of them scored high in at least several of the individual qualities that were quantitatively measured.
-The ranking designations are compared with rankings from the software development community.
-Recommendations for improving software along quality metrics are made.
-These include providing a detailed user manual and tutorials, explicitly stating the limits of the software, using user-friendly software languages, considering a peer review process, communicating development and contribution information, and using continuous integration and project management tools.
-This knowledge can be used to guide future development of scientific computing software along quality attributes to reduce software quality failures.\\
+We analyze the state of the practice of software development in the Lattice Boltzmann Methods software domain by quantitatively and qualitatively measuring and comparing 23 software packages.
+We present a general methodology for assessing the state of the practice of software development in scientific computing software domains.
+We present a domain analysis of the Lattice Boltzmann Methods software family, and identify candidate software packages.
+We assess these packages to answer software development related research questions to understand how software quality is impacted by software development choices, including principles, processes, and tools.
+We interview software developers to identify development pain points, and to identify how software quality is ensured.
+We use quantitative data to rank the software packages using the Analytical Hierarchy Process, and rank Ludwig, ESPResSo, and Palabos as the top three packages.
+Each of them score high in at least several of the individual qualities that we quantitatively measure.
+We compare these rankings with rankings from the software development community.
+We make recommendations, such as providing a detailed user manual and tutorials, explicitly stating the limits of the software, using user-friendly software languages, considering a peer review process, communicating development and contribution information, and using continuous integration and project management tools.\\
 
@@
 
-We analyzed the state of the practice of software development of Lattice Boltzmann Methods software domain by quantitatively and qualitatively measuring, and comparing, 23 software packages along quality attributes.
+We analyzed the state of the practice of software development in the Lattice Boltzmann Methods software domain by quantitatively and qualitatively measuring, and comparing, 23 software packages along quality attributes.
 The software qualities that were assessed in this report are listed in Section \ref{softwarequalities}.

### 2021-07-27 StateOfPractice/Peter-Notes/Report/Report.tex
updated LBM report, split section 4

@@
 
-\noindent\textbf{NUMBER OF PAGES:} 119
+\noindent\textbf{NUMBER OF PAGES:} 120
 
@@
 This follows a general methodology for assessing the state of the practice for SCS domains.
-	\item Quantitative and qualitative \textbf{Results}. 
+	\item \textbf{Quantitative Results}. 
+	\item \textbf{Qualitative Results}. 
 	\item An analysis of the results and \textbf{Answers To Research Questions}.
@@
 Some of the definitions are from \cite{Smithetal2020}.
-The quality measurement results are found in Section \ref{qualityresults} of this report.
+The quality measurement results are found in Section \ref{AHPresults} of this report.
 Section \ref{qualityrecommentations} lists recommendations to address software qualities in LBM software packages.
@@
 \newpage
-\section{Results}\label{qualityresults}
-
-This section presents the quantitative and qualitative findings from data that was gathered for each quality listed in Section \ref{softwarequalities}.
-Quantitative results are presented in Section \ref{AHPresults}, and qualitative results are presented in Section \ref{interviewresults}.
-
-\subsection{Quantitative Findings and AHP Results}\label{AHPresults}
-
-This subsection presents the quantitative findings from data that was gathered using the measurement template found in Section \ref{measurementtemplate2}.
-The results of the AHP analysis based on that data are also presented in this subsection.
-
-\subsubsection{Installability}
+
+\section{Quantitative Findings and AHP Results}\label{AHPresults}
+
+This section presents the quantitative findings from data that was gathered using the measurement template found in Section \ref{measurementtemplate2} for qualities listed in Section \ref{softwarequalities}.
+The results of the AHP analysis based on that data are also presented in this section.
+
+\subsection{Installability}
 
@@
 
-\subsubsection{Surface Correctness and Verifiability}
+\subsection{Surface Correctness and Verifiability}
 
@@
 
-\subsubsection{Surface Reliability}
+\subsection{Surface Reliability}
 
@@
 
-\subsubsection{Surface Robustness}
+\subsection{Surface Robustness}
 
@@
 
-\subsubsection{Surface Performance}
+\subsection{Surface Performance}
 
@@
 
-\subsubsection{Surface Usability}
+\subsection{Surface Usability}
 
@@
 
-\subsubsection{Maintainability}
+\subsection{Maintainability}
 
@@
 
-\subsubsection{Reusability}\label{reusabilityresults}
+\subsection{Reusability}\label{reusabilityresults}
 
@@
 
-\subsubsection{Surface Understandability}
+\subsection{Surface Understandability}
 
@@
 
-\subsubsection{Visibility and Transparency}
+\subsection{Visibility and Transparency}
 
@@
 
-\subsubsection{Overall Quality}
+\subsection{Overall Quality}
 
@@
 
-\subsection{Qualitative Findings From Developer Interviews}\label{interviewresults}
-
-This subsection presents the qualitative findings from data that was gathered during interviews with developers. The interview questions are found in Appendix \ref{interviewquestions}.
-
-\subsubsection{Surface Correctness and Verifiability}
+\newpage
+
+\section{Qualitative Findings From Developer Interviews}\label{interviewresults}
+
+This section presents the qualitative findings from data that was gathered during interviews with developers for qualities listed in Section \ref{softwarequalities}.
+The interview questions are found in Appendix \ref{interviewquestions}.
+
+\subsection{Surface Correctness and Verifiability}
 
@@
 
-\subsubsection{Modifiability}
+\subsection{Modifiability}
 
@@
 
-\subsubsection{Traceability}
+\subsection{Traceability}
 
@@
 
-\subsubsection{Reproducibility}
+\subsection{Reproducibility}
 
@@
 
-\subsubsection{Unambiguity}
+\subsection{Unambiguity}
 
@@
 
-This section answers the research questions listed in Section \ref{purpose} using the quantitative and qualitative data presented in Section \ref{qualityresults}, additional data from software package repositories and artifacts, and domain expert feedback.
+This section answers the research questions listed in Section \ref{purpose} using the quantitative data presented in Section \ref{AHPresults}, qualitative data presented in Section \ref{interviewresults}, additional data from software package repositories and artifacts, and domain expert feedback.
 Each subsection answers one of the research questions.

### 2021-07-27 StateOfPractice/Peter-Notes/Report/Report.tex
updated LBM report section 6.1

@@
 
-\noindent\textbf{NUMBER OF PAGES:} 120
+\noindent\textbf{NUMBER OF PAGES:} 121
 
@@
 The software packages were examined for the presence of artifacts, which were then categorized by frequency.
-Below we have grouped them into common, less common, and rare artifacts.
+We have grouped them into common, less common, and rare artifacts in Table \ref{artifactspresent}.
 Common artifacts were found in 16 to 23 ($>$70\%) of the software packages.
@@
 
+\begin{onehalfspace}
+	\begin{center}
+		\begin{tabular}{ p{8 cm} }
+			\hline
+			Common\\
+			\hline
+			Authors / Developers List\\
+			Bug Tracker\\
+			Dependency Notes\\
+			Installation Guide / Instructions\\
+			License\\
+			List of Related Articles / Publications\\
+			Makefile / Build File\\
+			README File\\
+			Requirements Specification / Theory Notes\\
+			Tutorial\\
+			\hline
+			Less Common\\
+			\hline
+			Change Log / Release Notes\\
+			Design Documentation\\
+			Functional Specification / Notes\\
+			Performance Information / Notes\\
+			Test Plan / Report / Script / Data / Cases\\
+			User Manual/Guide\\
+			Version Control\\
+			\hline
+			Rare\\
+			\hline
+			API Documentation\\
+			Developer / Contributor Manual / Guide\\
+			FAQ / Forum\\
+			Verification and Validation Plan / Notes\\
+			Video Guide (including YouTube)\\
+			\hline
+		\end{tabular}
+		\captionof{table}{Artifacts Present}\label{artifactspresent}
+	\end{center}
+\end{onehalfspace}
+
 \subsubsection{Common Artifacts}
-The following artifacts were commonly found in the 23 software packages that were tested.
-All of the top four AHP ranked packages, ESPREsSo, Ludwig, OpenLB, and Palabos, have each of these artifacts, except only three of them have a requirements specification or theory notes.
+All of the top four AHP ranked packages, ESPREsSo, Ludwig, OpenLB, and Palabos, have each of the commonly found artifacts, except only three of them have a requirements specification or theory notes.
 Palabos is the only one of these four packages that does not have an artifact from this category.
 
-\begin{multicols}{2}	
-\begin{itemize}
-	\begin{singlespace}
-		\item Authors/Developers List
-		\item Bug Tracker
-		\item Dependency Notes 
-		\item Installation Guide/Instructions 
-		\item License
-		\item List of Related Articles/Publications 
-		\item Makefile / Build File 
-		\item README File 
-		\item Requirements Specification / Theory Notes
-		\item Tutorial
-	\end{singlespace}
-\end{itemize}
-\end{multicols}
-
-These common artifacts contribute to the quality of the software in the following ways.
+The common artifacts contribute to the quality of the software in the following ways.
 A list of authors and developers helps potential users and contributors contact project members to answer questions affecting many software qualities.
@@
 \subsubsection{Less Common Artifacts}\label{lesscommon}
-The following artifacts were less commonly found in the 23 software packages that were tested. The top four AHP ranked packages have most of these artifacts.
+The top four AHP ranked packages have most of the less common artifacts.
 At the time of data collection, only one (Palabos) of the four packages did not have a user manual or guide, but there was a broken link on the package website indicating that such an artifact might exist.
@@
 
-	\begin{multicols}{2}	
-		\begin{itemize}
-			\begin{singlespace}
-			\item Change Log / Release Notes
-			\item Design Documentation
-			\item Functional Specification/Notes	
-			\item Performance Information/Notes
-			\item Test Plan/Report/Script/Data/Cases
-			\item User Manual/Guide
-			\item Version Control
-		\end{singlespace}
-		\end{itemize}
-	\end{multicols}
-
-These artifacts also contribute to software quality.
+The less common artifacts also contribute to software quality.
 A change log or release notes improve traceability of the software.
@@
 \subsubsection{Rare Artifacts}
-The following artifacts were rarely found in the 23 software packages that were assessed.
-It is not common for the top four AHP ranked packages to have many of these artifacts.
+It is not common for the top four AHP ranked packages to have many of the rare artifacts.
 None of the top four packages have any explicit API documentation.
@@
 
-	\begin{multicols}{2}	
-		\begin{itemize}
-			\begin{singlespace}
-				\item API Documentation
-				\item Developer/Contributor Manual/Guide
-				\item FAQ / Forum
-				\item Verification and Validation Plan/Notes
-				\item Video Guide (including YouTube)
-			\end{singlespace}
-		\end{itemize}
-	\end{multicols}
-
-Although these artifacts were rarely found in our set of LBM software, they also contribute to software quality.
+Although the artifacts were rarely found in our set of LBM software, they also contribute to software quality.
 API documentation helps with reusing the software.

### 2021-07-28 StateOfPractice/Ao-Report/15InterviewsWithDevelopers.tex
MI report - added ch 7

@@
 \label{sec_pain_points_2}
-\label{sec_pain_points_1}
 We summarize the pain points in Group 2 as
@@
 \item if the team has access to MRI scanners, self-building MI datasets;
-\item if the team has connections with MI equipment manufacturers, asking for their help on data format problems.
+\item if the team has connections with MI equipment manufacturers, asking for their help on data format problems;
+\item storing all images that cause special problems, and maintaining this special dataset over time.
 \end{itemize}

### 2021-07-28 StateOfPractice/Ao-Report/17Recommendations.tex
MI report - added ch 7

@@
 
-\section{Recommendations to Software Qualities}
+This section presents our recommendations on SC software development.
+We give these suggestions to the development in all SC domains, unless specifically mentioning that some guidelines are only for MI software.
 
-We list some of the key points that may improve each quality as follows,
+Section \ref{sec_recommendations_qualities} discusses the actions that can potentially improve the ten software qualities.
+Sections \ref{sec_recommendations_limited_resources}, \ref{sec_recommendations_technology_stack}, and \ref{sec_recommendations_testing_dataset} are based on the primary pain points collected from the developers in the MI domain, but we believe scientists and developers are likely to face them in most SC domains.
+These sections contain our general suggestions to tackle them.
+
+\section{Recommendations on Improving Software Qualities}
+\label{sec_recommendations_qualities}
+
+Based on our quality measurements in Sections \ref{ch_results} and discussions with the developers in Sections \ref{sec_interview_software_qualities}, we collected many key points that may improve the software qualities.
+We list the primary ones by each quality as follows,
 \begin{itemize}
@@
 \end{itemize}
+
+\section{Recommendations on Dealing With Limited Resources}
+\label{sec_recommendations_limited_resources}
+
+The limitation of resources has many faces. We regard the lack of fundings, time, and developers as representations of this problem.
+
+We summarize our discussion with the MI software developers in Section \ref{sec_pain_points_1} to the following recommendations,
+\begin{itemize}
+\item \textbf{Identify the root cause.} More fundings or developers may not solve the problem of lacking time. It is beneficial to identify the underlying obstacles to the team.
+
+\item \textbf{Maintain a good documentation.} Creating and updating documentation consumes time, but can save much more time in the long term.
+If the users and developers can find answers to their questions by themselves, they are less likely to abuse the team's issue tracker.
+
+\item \textbf{Adopt time-saving tools.} A good CI/CD tool (e.g., GitHub Actions) saves time for building and deploying the product, and automated tests can work in the background while developers are focusing on other tasks.
+
+\item \textbf{Use test-driven development process.} Many people think writing test cases is less fun than building the functional code, but this is only true before the bugs show.
+Identifying and fixing bugs can cause substantial resources.
+Setting up the test cases costs time, but generates more benefits in the long run.
+
+\item \textbf{Consider supporting third-party plugins or extensions.} Why not let users share the burden?
+No software product can deliver every user's needs, and the large quantity of features leads to more bugs and difficulty to maintain.
+So it may be a good idea to shift some development and maintenance responsibilities to the users.
+The users may also be happy about the extra flexibility.
+
+\item \textbf{Consider ``hibernating" for a while.} When developers are not enough, the team can shift from development mode toward maintenance mode for some time.
+Stop building new features, and fix bugs and design problems in the past.
+The software qualities may also improve as a result.
+
+\item \textbf{Commercialization is not always toxic.} Licensing the software to commercial companies to use as internal modules of their products may bring financial supports to the team.
+Meanwhile, the project can stay open-source to the community.
+\end{itemize}
+
+\section{Recommendations on Choosing A Tech Stack}
+\label{sec_recommendations_tech_stack}
+
+A tech stack refers to a set of technologies used by a team to build software and manage the project.
+Section \ref{sec_pain_points_2} lists the advantages and disadvantages between native and web applications.
+In this section, we give further suggestions on the choice of a tech stack to address the \textit{compatibility}, \textit{maintainability}, \textit{performance}, and \textit{security} of software.
+
+\begin{itemize}
+\item \textbf{Identify the priorities of the qualities.} It is hard to cover all aspects.
+Some teams achieve all four above qualities for their software, but it is not an easy task.
+Sections \ref{sec_pain_points_2} contains more details about the difficulty of balancing between the four qualities.
+A team needs to prioritize its objectives according to its resource and experience.
+
+\item \textbf{Be open-minded about new technologies.} Web applications with only a frontend are known for worse \textit{performance} than native applications.
+However, new technologies may ease this difference.
+For example, some JavaScript libraries can help the frontend with harnessing the power of computer GPU and accelerate graphical computing.
+In addition, there are new frameworks helping developers with cross-platform \textit{compatibility}.
+For example, the \hyperlink{https://flutter.dev/}{Flutter} project enables support for web, mobile, and desktop OS with one codebase.
+
+\item \textbf{Use git and GitHub.} 
+As mentioned in Sections \ref{sec_score_maintainability}, almost all of the 29 MI software projects used git, and the majority of them used GitHub.
+We found from the projects' websites and our interviews with developers that, some projects moved from other version control tools to git and GitHub.
+GitHub provides convenient repository and project management, and OSS projects receive more attention and contribution on
+GitHub.
+
+\item \textbf{Web applications can also deliver high performance.} Web applications with backend servers may perform even better than native applications.
+If a team needs to support lower-end computers, it is good to use back-end servers for heavy computing tasks.
+
+\item \textbf{Backend servers can have low costs.} It is worth exploring the serverless solutions from major cloud service providers.
+Serverless still uses a server, but the team is only charged when they use it.
+The solution is event-driven, and costs the team by the number of requests it processes.
+Thus, serverless can be very cost-effective for the less intensively used functions.
+
+\item \textbf{Web transmission may diminish security.} Transferring sensitive data online can be a problem for projects requiring high security.
+Regulations in some SC domains may forbid doing so.
+In this case, a web application with a backend may not be a good choice.
+
+\item \textbf{Maintain a good documentation.} No matter what tech stack a team uses, good documents such as well-maintained project plan, developer's manual, and API documentation always help team members to contribute more and make fewer mistakes.
+\end{itemize}
+
+\section{Recommendations on Enriching the Testing Datasets}
+\label{sec_recommendations_testing_dataset}
+
+As described in Section \ref{sec_interview_pain_points}, it was difficult for some software development teams in the MI domain to access real-world medical imaging datasets.
+This problem restricted their capability and flexibility to test their software.
+We believe software developers in other SC domains may also face similar issues.
+
+Based on Section \ref{sec_pain_points_3}, we provide some suggestions as follows,
+\begin{itemize}
+\item \textbf{Build and maintain good connections to datasets.} A team can build connections with professionals working in the SC domain, who may have access to private datasets and perform tests for the team.
+Moreover, if a team has such professionals as internal members, the process can be even simpler.
+
+\item \textbf{Collect and maintain datasets over time.} A team may face all kinds of strange problems caused by various unique inputs over the years of development.
+It is worth collecting and maintaining these data, which can form a good dataset for testing.
+
+\item \textbf{Establish community collaboration in the domain.} During our interviews with developers in the MI domain, we heard many stories of asking for supports from other professionals or equipment manufacturers.
+However, we believe that broader collaboration between development teams can address this problem better.
+Some datasets are too sensitive to share, but if the community has some kind of ``group discussion", teams can better express their needs, and professionals can better offer their voluntary supports for testing.
+Ultimately, perhaps the community can establish a nonprofit organization as a third-party, which maintains large datasets, tests OSS in the domain, and protects privacy.
+\end{itemize}

### 2021-07-28 StateOfPractice/Peter-Notes/Report/Report.tex
updated LBM report, organized 6.4

@@
 
+\subsubsection{Lack of Development Time}
+
 A developer of pyLBM noted that their small development team has a lack of time to implement new features.
@@
 
+\subsubsection{Lack of Software Development Experience}
+
 A lack of software development experience was noted by the developer of TCLB.
@@
 It was further noted that this has been somewhat addressed, as the code has been re-written to best ensure ease of future contributions.
+
+\subsubsection{Lack of Incentive and Funding}
 
@@
 Work has been done to address this in \citep{smith2016software}, which presents a set of software citation principles and discusses ``how they could be used to implement software citation in the scholarly community'' \citep{katz2019software}.
+
+\subsubsection{Lack of External Support}
 
@@
 
-Documentation of LBM software is noted as being important, but its quality could be improved.
-As already noted, there is often no time or funding for maintaining quality documentation for software that is rarely used outside of the development team.
-Furthermore, the documentation generally only provides a shallow overview of the underlying CFD theory.
-Users would be well advised to already be familiar with these topics, or they should spend significant time referencing theory resources.
-The documentation instead generally focuses on explaining how to use the software.
-It is of course not feasible for package documentation to address the underlying physics topics in detail, so it is advised that the package documentation links to resources that better explain the underlying theory.
-Sometimes, frequently asked questions about the underlying theory are answered in the documentation.
-OpenLB has an artifact for such questions.
+\subsubsection{Parallelization and Continuous Integration}
 
@@
 The developer of TCLB noted that eliminating equivalent statements using macros had helped improve the quality of their source code, specifically helping with reusing code to run on both the CPU and GPU.
+
+\subsubsection{Ensuring Correctness}
 
@@
 The solution for this has been to limit the size of the testing problems, and to run tests in small batches with few iterations.
+
+\subsubsection{Usability}
 
@@
 
+\subsubsection{Technical Debt}
+
 A few potential future obstacles were noted.
@@
 
-The developer comments have emphasized an importance on source code, while documentation seems to be of secondary importance.
+\subsubsection{Quality of Documentation}
+
+Interviewees commented that documentation is important and that its quality could be improved.
+As already noted, there is often no time or funding for maintaining quality documentation for software that is rarely used outside of the development team.
+Furthermore, the documentation generally only provides a shallow overview of the underlying CFD theory.
+Users would be well advised to already be familiar with these topics, or they should spend significant time referencing theory resources.
+The documentation instead generally focuses on explaining how to use the software.
+It is of course not feasible for package documentation to address the underlying physics topics in detail, so it is advised that the package documentation links to resources that better explain the underlying theory.
+Sometimes, frequently asked questions about the underlying theory are answered in the documentation.
+OpenLB has an artifact for such questions.
+
+The developer comments emphasized an importance on source code, while documentation seems to be of secondary importance.
 It must be stressed that improving documentation could benefit development and help eliminate some of the developer concerns that were raised.

### 2021-07-28 StateOfPractice/Ao-Report/18Conclusion.tex
MI report - added ch 8

\chapter{Conclusions}
\label{ch_conclusions}

We analyzed the state of the practice for SC software in the MI domain. To better achieve our goals in Section \ref{ch_intro}, we proposed six research questions, that we list in Section \ref{sec_research_questions}.

Our methods in Section \ref{ch_methods} form a general process to evaluate domain-specific software, that we apply on specific SC domains. As mentioned in Section \ref{sec_applying_method}, following this process, we chose the MI domain, identified 48 SC software candidates in it, then selected 29 of them to our final list. Section \ref{ch_results} lists our measurements to nine software qualities for the 29 projects, and Section \ref{ch_interview} contains our interviews with eight of the 29 teams, discussing their development process and five software qualities.

With the measurement and interview results in Section \ref{ch_results} and \ref{ch_interview}, we reveal some current status of SC software development and qualities in the MI domain. The interview results also show some merits, drawbacks, and pain points within the development process. With these findings, we answer our six research questions in Section \ref{ch_answers}. In addition, Section \ref{ch_recommendations} presents our recommendations on SC software development, such as the ones on improving software qualities and easing pain points during development.

With learnings from this project, we summarized recommendations for the future state of the practice assessments,
\begin{itemize}
	\item we can further automate the measurements on the grading temple in Appendix \ref{ap_grading_template}. For example, with automation scripts and a GitHub API, we may save significant time on retrieving the GitHub metrics;
	\item the grading standard can be more explicit. For example, we can explicitly define scores for each item in the grading temple.
	\item we can better organize the interview questions. We think they are good questions. However, since we use audio conversion tools to transcribe the answers, we should aim to make the transcription easier to read. For example, we can order them together for questions about the five software qualities and compose a similar structure for each.
	\item we can mark the follow-up interview questions with keywords. For example, say ``this is a follow-up question" every time asking one. Thus, we record this sentence in the transcription, and it will be much easier to distinguish the follow-up questions from the 20 designed questions.
\end{itemize}

### 2021-07-28 StateOfPractice/Peter-Notes/Report/Report.tex
updated LBM report, added rationale for recommendations

@@
 
-\noindent\textbf{NUMBER OF PAGES:} 121
+\noindent\textbf{NUMBER OF PAGES:} 123
 
@@
 \begin{itemize}
-	\item Include OS compatibility, including specific OS versions.
-	\item Provide complete installation instructions.
+	\item Include OS compatibility, including specific OS versions. This was done by several top ranked packages, including ESPResSo and LUMA.
+	\item Provide complete installation instructions. This was done by all of the top five ranked packages.
 	\item Installation instructions should be written as if the user does not have any dependencies installed and is installing on a clean OS.
+This was done by several top ranked packages, including ESPResSo, OpenLB, and Palabos.
 This can be tested on a clean environment using a virtual machine.
 	\item List all dependencies in the installation instructions, dependency versions, and how to install them.
-	\item If possible, automate the installation of dependencies. Use tools such as Make. 
-	\item Automate the installation process as much as possible. Use tools such as Make.
-	\item The installation instruction should only be in one location.
+This was done by several top ranked packages, including ESPResSo, LUMA, OpenLB, and Palabos.
+	\item If possible, automate the installation of dependencies. Use tools such as Make. This was done by all of the top five ranked packages.
+	\item Automate the installation process as much as possible. Use tools such as Make. This was done by all of the top five ranked packages.
+	\item The installation instruction should only be in one location. This was done by several top ranked packages, including ESPResSo, OpenLB, and Palabos.
 	\item Include descriptive error messages for errors encountered during installation. This was done by several top ranked packages, including Ludwig and LUMA
@@
 \begin{itemize}
-	\item Use a requirements specification document. A potential template is presented in \citep{smith2005new}. 
+	\item Use a requirements specification document.
+This is suggested in SCS literature, and several of the top ranked packages had such a document or reference to theory manuals.
+A potential template is presented in \citep{smith2005new}.
 	\item Make public (on package website or GitHub) the requirements specification document, or explicitly reference the domain theory that the software is designed from.
-	\item Ensure the above information is easy to find. Consider adding it to the user manual. 
-	\item Development teams should include both domain experts and experienced software developers.
-	\item Use and make public (on package website or GitHub) detailed documentation. Consider using automatic document generation tools like Doxygen, Drasil, or Sphinx.
+This was done by several top ranked packages, including ESPResSo, Ludwig, LUMA, and OpenLB.
+	\item Ensure the above information is easy to find. Consider adding it to the user manual. The information is easy to find in most top ranked packages.
+	\item Development teams should include both domain experts and experienced software developers. This suggestion is based on developer comments.
+	\item Use and make public (on package website or GitHub) detailed documentation.
+Consider using automatic document generation tools like Doxygen, Drasil, or Sphinx.
+This was done by all of the top five ranked packages.
 	\item Provide detailed tutorials that include expected output, like the \href{https://www.walberla.net/doxygen/index.html}{waLBerla tutorials}.
 	\item Use unit tests during development and make them public (on package website or GitHub). This was available for Ludwig.
-	\item Modularize the source code, separate components, hide information  behind well defined interfaces.
+	\item Modularize the source code, separate components, hide information  behind well defined interfaces. This is suggested in SCS literature, and in developer comments.
 	\item Use continuous integration tools (Bamboo, Jenkins, and Travis CI) and processes during development. This was done by top ranked packages ESPResSo and Ludwig.
@@
 	\item The packages should include detailed tutorials, including dependencies, expected output, and any additional supplementary documentation that may be required.
-This was done by many packages, including waLBerla, Palabos, MechSys, LUMA, pyLBM.
+This was done by many packages, including waLBerla, Palabos, MechSys, LUMA, and pyLBM.
 \end{itemize}
@@
 	\item Provide documentation that details the background theory information, or provide a reference to such information.
+This was done by several top ranked packages, including ESPResSo, Ludwig, LUMA, and OpenLB.
 	\item Identify expected user characteristics. LIMBES, Ludwig, laboetie, and Palabos did this. The importance of specifying user characteristics is discussed in \citep{smith2007requirements}.
-	\item Keep all documentation in one location.
-	\item Do not duplicate artifact information.
-	\item Maintain a user support model (Git, email, forum, FAQ)
+	\item Keep all documentation in one location. This was done by top ranked packages.
+	\item Do not duplicate artifact information. This is suggested as good software development practice. Duplicate information is difficult to maintain. 
+	\item Maintain a user support model (Git, email, forum, FAQ). This was done by all of the top five ranked packages.
 	\item If possible, consider using popular user-friendly software languages like Python.
@@
 \begin{itemize}
-	\item Keep artifacts updated.
+	\item Keep artifacts updated. This is suggested as good software maintenance practice.
 	\item Ideally include most of the common and less common artifacts listed in Section \ref{artifacts}.
@@
 Four of the top five overall ranked packages (Ludwig, ESPResSo, Palabos, OpenLB) have about 20 percent of their source code as comments.
-	\item Modularize the source code, separate components, hide information behind well defined interfaces.
-	\item Eliminate code duplication.
+	\item Modularize the source code, separate components, hide information behind well defined interfaces. This is suggested in SCS literature, and in developer comments.
+	\item Eliminate code duplication. This is suggested as good software development practice. Duplicate code is difficult to maintain.
 	\item If possible, consider using popular user-friendly software languages like Python.
@@
 \begin{itemize}
-	\item Modularize the source code, separate components, hide information behind well defined interfaces.
+	\item Modularize the source code, separate components, hide information behind well defined interfaces. This is suggested in SCS literature, and in developer comments.
 	\item If possible, consider using popular user-friendly software languages like Python.
@@
 ESPResSo and Sailfish use Python to address several software qualities, including modifiability.
-	\item Consider future source code modifiability as early as the design stage of development.
+	\item Consider future source code modifiability as early as the design stage of development. This is suggested as good software development practice.
 	\item Consider flexibility of data structures and data storage in the design stage. The package pyLBM redeveloped data structures to ease future changes.
@@
 \begin{itemize}
-	\item Modularize the source code, separate components, hide information behind well defined interfaces.
-	\item Document module interfaces in design and developer documentation. 
+	\item Modularize the source code, separate components, hide information behind well defined interfaces. This is suggested in SCS literature, and in developer comments.
+	\item Document module interfaces in design and developer documentation. This is suggested as good software development practice.
 	\item Provide API documentation, if applicable. Only one (ESPResSo) of the top five ranked packages provided API documentation.
@@
 Domain algorithms are noted in the source code of 11 packages.
-	\item Add meaningful comments. Indicate what is being done in each section of source code.
-	\item Modularize the source code, separate components, hide information behind well defined interfaces.
+	\item Add meaningful comments. Indicate what is being done in each section of source code. All of the top ranked packages have meaningful comments in their source code.
+	\item Modularize the source code, separate components, hide information behind well defined interfaces. This is suggested in SCS literature, and in developer comments.
 	\item Provide a user manual that identifies all technical and problem modeling requirements. Thirteen software packages have a user manual.
@@
 \begin{itemize}
-	\item Update all relevant documentation when a change to the software is made.
-	\item Use automatic document generation tools (Doxygen, Drasil, Sphinx) to limit the time spent on updating documentation.
+	\item Update all relevant documentation when a change to the software is made. This is suggested as good software development practice.
+	\item Use automatic document generation tools (Doxygen, Drasil, Sphinx) to limit the time spent on updating documentation. This was done by all of the top five ranked packages.
 	\item Provide a developer's guide to help orient developers to the artifacts, source code, and system architecture so that they can better document changes.
+Top ranked package OpenLB has a developer guide.
 \end{itemize}
@@
 \begin{itemize}
-	\item Summarize the development process that is used. Provide information on how new users can contribute. Identify the development model by name (waterfall, agile, etc.), if appropriate.
+	\item Summarize the development process that is used.
+Provide information on how new users can contribute.
+Identify the development model by name (waterfall, agile, etc.), if appropriate.
+Seven of the software packages have some artifacts outlining the general development process.
+Eleven packages have information on how to contribute.
 	\item Identify the development environment. Eight of the 23 software packages identify the development environment.
-	\item Update all relevant documentation when a change to the software is made.
+	\item Update all relevant documentation when a change to the software is made. This is suggested as good software development practice.
 	\item Include notes with all releases. Nine of the packages include release notes.
-	\item Communicate all changes within the development team.
+	\item Communicate all changes within the development team. This suggestion is based on developer comments.
 	\item Use continuous integration processes and tools (Bamboo, Jenkins, and Travis CI). This was done by top ranked packages ESPResSo and Ludwig.
@@
 	\item Use project management tools, including change and version control tools (GitHub, GitLab, CVS), collaboration tools (GitHub, GitLab), and document generation tools (Doxygen, Drasil, Sphinx).
+This was done by all of the top five ranked packages.
 \end{itemize}
@@
 	\item Consider peer review processes and task based inspection to assess contributions. High ranked package ESPResSo does this.
-	\item Update all relevant design documentation when a change to the software is made.
+	\item Update all relevant design documentation when a change to the software is made. This is suggested as good software development practice.
 \end{itemize}
@@
 	\item Provide a detailed user manual. Thirteen packages have a user manual. ESPResSo has a well detailed manual.
-	\item Provide a table of symbols in the developer's guide that maps to names used in the course code.
+	\item Provide a table of symbols in the developer's guide that maps to names used in the source code. This is suggested as good software development practice.
 	\item State appropriate fluid dynamics problems that the software is designed to model in the documentation, and explicitly state the limits of the software. This is done in the ESPResSo user guide.
 	\item The documentation should either explain the underlying CFD theories or provide a reference to appropriate resources.
+This was done by several top ranked packages, including ESPResSo, Ludwig, LUMA, and OpenLB.
 	\item Consider asking users for feedback on the documentation. The developer of pyLBM noted that such feedback would be appreciated.
@@
 \begin{itemize}
-	\item Provide a detailed user manual. It should identify elements of user interfaces, and identify all requirements to model a system.
-	\item State appropriate fluid dynamics problems that the software is designed to model in the documentation, and explicitly state the limits of the software. 
+	\item Provide a detailed user manual.
+It should identify elements of user interfaces, and identify all requirements to model a system.
+Thirteen packages have a user manual.
+ESPResSo has a well detailed manual.
+	\item State appropriate fluid dynamics problems that the software is designed to model in the documentation, and explicitly state the limits of the software. This is done in the ESPResSo user guide.
 	\item Include detailed tutorials, including dependencies, expected
-	output, and any additional supplementary documentation that may be required.
-	\item Keep all documentation in one location.
-	\item If possible, consider using popular user-friendly software languages like Python. Especially consider this for parts of the source code that is likely to be modified or reviewed by users. 
-	\item Modularize the source code, separate components, hide information behind well defined interfaces.
-	\item Include descriptive error messages where appropriate.
-	\item Summarize the development process that is used. Provide information on how new users can contribute. 
-	\item Consider peer review processes and task based inspection to assess contributions.	
-	\item Use continuous integration tools (Bamboo, Jenkins, and Travis CI) and processes during development. 
+	output, and any additional supplementary documentation that may be required. This was done by many packages, including waLBerla, Palabos, MechSys, LUMA, and pyLBM.
+	\item Keep all documentation in one location. This was done by top ranked packages.
+	\item If possible, consider using popular user-friendly software languages like Python.
+Especially consider this for parts of the source code that is likely to be modified or reviewed by users.
+ESPResSo and Sailfish use Python to address several software qualities, including maintainability, modifiability and usability.
+	\item Modularize the source code, separate components, hide information behind well defined interfaces. This is suggested in SCS literature, and in developer comments.
+	\item Include descriptive error messages where appropriate. This was done by most packages that encountered a fault.
+	\item Summarize the development process that is used.
+Provide information on how new users can contribute.
+Seven of the software packages have some artifacts outlining the general development process.
+Eleven packages have information on how to contribute.
+	\item Consider peer review processes and task based inspection to assess contributions. High ranked package ESPResSo does this.
+	\item Use continuous integration tools (Bamboo, Jenkins, and Travis CI) and processes during development. This was done by top ranked packages ESPResSo and Ludwig.
 	\item Use project management tools, including change and version control tools (GitHub, GitLab, CVS), collaboration tools (GitHub, GitLab), and document generation tools (Doxygen, Drasil, Sphinx).
+This was done by all of the top five ranked packages.
 \end{itemize}

### 2021-07-28 StateOfPractice/Ao-Report/01abstract.tex
MI report - added abstract

@@
 
-Abstract here
+We present a method to assess the state of the practice for software in Scientific Computing (SC) domains and apply the method to the Medical Imaging (MI) domain.
+This method guided us to select 29 MI software projects from 48 candidates, assess nine software qualities by answering 103 questions for each software, and interview eight of the 29 development teams.
+The results helped us with revealing the current status of MI software development, such as the merits, drawbacks, and pain points during the development process and the software qualities.
+Based on our findings, we suggested guidelines and recommendations for future SC software development.
+\newline
+
+\noindent\textbf{Keywords:} Medical Imaging, Scientific Computing, software engineering, software quality, Analytic Hierarchy Process, developer interview

### 2021-07-28 StateOfPractice/Ao-Report/11Introduction.tex
MI report - added ch 8
MI report - added Abbreviations and Acronyms

@@
 \label{ch_intro}
-This paper analyzes the state of the practice for Medical Imaging (MI) software.
+This report analyzes the state of the practice for Medical Imaging (MI) software.
 MI is the clinical tool to image the interior of a body, providing information for diagnostic, analytic, and medical applications \cite{FDA2021} \cite{enwiki:1034887445}.
@@
 
-Scientiﬁc computing (SC) is the technique and process of mathematically modeling real-world engineering or scientific systems with computing tools, as well as analysis and prediction with the models \cite{Smith2006}.
+Scientiﬁc Computing (SC) is the technique and process of mathematically modeling real-world engineering or scientific systems with computing tools, as well as analysis and prediction with the models \cite{Smith2006}.
 MI software belongs to a specific domain of SC.
@@
 Besides the above major functions, some MI software provides supportive functions.
-For example, with Tool Kit libraries VTK \cite{SchroederEtAl2006} and ITK \cite{McCormick2014}, developers build software with Visualization and Analysis functions; Picture Archiving and Communication System (PACS) helps users to economically store and conveniently access images \cite{Choplin1992}.
+For example, with Tool Kit (TK) libraries VTK \cite{SchroederEtAl2006} and ITK \cite{McCormick2014}, developers build software with Visualization and Analysis functions; Picture Archiving and Communication System (PACS) helps users to economically store and conveniently access images \cite{Choplin1992}.
 
@@
 child[concept color=teal!40] { node {Supporting}
-    child[concept color=blue!30] { node {Tool Kit}}
+    child[concept color=blue!30] { node {TK}}
     child[concept color=blue!30] { node {PACS}}

### 2021-07-28 StateOfPractice/Ao-Report/12Background.tex
MI report - added Abbreviations and Acronyms

@@
 In such a case, the former is usually the product of software engineers, and the latter is likely to have developers who work in the domain and are also end-users of the products.
-One example is software in Scientific Computing (SC), since the developers need to utilize their domain-specific during the development process \cite{WilsonEtAl2014}.
+One example is software in SC, since the developers need to utilize their domain-specific during the development process \cite{WilsonEtAl2014}.
 

### 2021-07-30 StateOfPractice/Ao-Report/11Introduction.tex
MI report - updated ch 2

@@
 
-Scientiﬁc Computing (SC) is the technique and process of mathematically modeling real-world engineering or scientific systems with computing tools, as well as analysis and prediction with the models \cite{Smith2006}.
+We define SC as ``the use of computer tools to analyze or simulate mathematical models of real world systems of engineering or scientiﬁc importance so that we can better understand and predict the system’s behaviour" \cite{Smith2006}.
+Software development in SC depends on the knowledge of three areas: engineering or science domain knowledge, mathematical algorithm knowledge, and computational algorithm knowledge.
 MI software belongs to a specific domain of SC.

### 2021-07-30 StateOfPractice/Ao-Report/12Background.tex
MI report - updated ch 2

--- 2021-07-14 StateOfPractice/Ao-Report/12Background.tex
@@
 
+In this section, we introduce the relevant software categories (Section \ref{sec_software_categories}).
+We also cover the software quality definitions (Section \ref{sec_software_quality}) and an overview of the Analytic Hierarchy Process (AHP) (Section \ref{sec_AHP}).
+
 When designing a method for evaluating the state of the practice of domain-specific software, we included a step to select domain and software.
 Knowledge of different software categories is essential for the selection.
-To compare and grade the software qualities with the grading template in Appendix \ref{ap_grading_template}, we need the definitions of qualities and the Analytic Hierarchy Process (AHP).
+To compare and rank the software qualities with the grading template in Appendix \ref{ap_grading_template}, we need the definitions of qualities and the AHP.
 
 \section{Software Categories}
-We usually target specific software categories to narrow down the scopes when selecting software domains and software packages.
-In this section, we discuss three common software categories and also SC software.
+\label{sec_software_categories}
+We target specific software categories to narrow down the scope when selecting software packages for measuring.
+In this section, we discuss three common software categories: Open source software (OSS), freeware, and commercial software.
 
@@
 \label{sec_open_source_software}
-For Open Source software (OSS), its source code is openly accessible, and users have the right to study, change and distribute it under a license granted by the copyright holder.
+For OSS, the source code is openly accessible.
+Users have the right to study, change and distribute it under a license granted by the copyright holder.
 For many OSS projects, the development process relies on the collaboration of different contributors worldwide \cite{Corbly2014}.
 Accessible source code usually exposes more ``secrets'' of a software project, such as the underlying logic of software functions, how developers achieve their works, and the flaws and potential risks in the final product.
-Thus, it brings much more convenience to the researchers analyzing the qualities of the project.
+Thus, it brings much more convenience to researchers analyzing the qualities of the project.
 
@@
 Freeware is software that can be used free of charge.
-Unlike with OSS, the authors of freeware typically do not allow users to access or modify the source code of the software \cite{LINFO2006}.
-The term \textit{freeware} should not be confused with \textit{free software}, which is similar to OSS but with a few differences.
+Unlike with OSS, the authors of freeware do not allow users to access or modify the source code of the software \cite{LINFO2006}.
+The term \textit{freeware} should not be confused with \textit{free software}, which is similar to OSS.
 To the end-users, the differences between freeware and OSS often do not bother them.
 The fact that these products are free of charge is likely to make them popular with many users.
-However, software developers, end-users who wish to modify the source code, and researchers looking for inner characteristics may find the inaccessible source code a problem.
+However, software developers, end-users who wish to modify the source code, and researchers looking for insight into software development process will find the inaccessible source code a problem.
 
@@
 In such a case, the former is usually the product of software engineers, and the latter is likely to have developers who work in the domain and are also end-users of the products.
-One example is software in Scientific Computing (SC), since the developers need to utilize their domain-specific during the development process \cite{WilsonEtAl2014}.
-
-\subsection{Scientific Computing Software}
-Software development in SC depends on the knowledge of three areas - the inside of a specific engineering or science domain, the ability to mathematically build models and apply algorithms, and to implement theoretical models and algorithms with computational tools.
-SC software is built with mathematical and computational tools to serve the purpose of solving scientific problems in a domain \cite{Mehta2015}.
-However, the majority of scientists developing their software are self-taught programmers \cite{WilsonEtAl2014}, so there may be a bigger Room for software quality improvement in SC domains.
+One example of such software is SC software, since the developers need to utilize their domain-specific during the development process \cite{Wilson2014}.
 
@@
 
-Our grading template in Appendix \ref{ap_grading_template} contain nine software qualities.
-This section lists the definitions of them, which are from Smith et al.
+This section lists the definitions of 10 software qualities, which are from Smith et al.
 \cite{SmithEtAl2020}.
-The order of the qualities follows the grading template.
+We aim to measure each of them for selected SC software packages.
+The order of the first nine qualities follows our grading template in Appendix \ref{ap_grading_template}.
+We do not measure \textit{reproducibility} with the grading template, but discuss it with the developers by interviews.
 
 \begin{itemize}
-	\label{def_installability}
-	\item \textbf{Installability} The effort required for the installation, uninstallation, or reinstallation of a software or product in a specified environment.
-	\label{def_correctness_verifiability}
+	\item \textbf{Installability} The effort required for the installation, uninstallation, or reinstallation of a software or product in a specified environment \cite{ISO/IEC25010} \cite{lenhard2013measuring}.
+
 	\item \textbf{Correctness \& Verifiability} A program is correct if it behaves according to its stated.
-Verifiability is the extent to which a set of tests can be written and executed, to demonstrate that the delivered system meets the specification.
-	\label{def_reliability}
+Verifiability is the extent to which a set of tests can be written and executed, to demonstrate that the delivered system meets the specification \cite{GhezziEtAl2003}.
+
 	\item \textbf{Reliability} The probability of failure-free operation of a computer program in a specified environment for a specified time, i.e.
-the average time interval between two failures also known as the mean time to failure (MTTF).
-	\label{def_robustness}
-	\item \textbf{Robustness} Software possesses the characteristic of robustness if it behaves ``reasonably'' in two situations: i) when it encounters circumstances not anticipated in the requirements specification, and ii) when the assumptions in its requirements specification are violated.
-	\label{def_usability}
-	\item \textbf{Usability} The extent to which a product can be used by specified users to achieve specified goals with effectiveness, efficiency, and satisfaction in a specified context of use.
-	\label{def_maintainability}
-	\item \textbf{Maintainability} The effort with which a software system or component can be modified to i) correct faults; ii) improve performance or other attributes; iii) satisfy new requirements.
-	\label{def_reusability}
-	\item \textbf{Reusability} The extent to which a software component can be used with or without adaptation in a problem solution other than the one for which it was originally developed.
-	\label{def_understandability}
-	\item \textbf{Understandability} (To be completed)
-	\label{def_visibility_transparency}
-	\item \textbf{Visibility/Transparency} The extent to which all of the steps of a software development process and the current status of it are conveyed clearly.
+the average time interval between two failures also known as the mean time to failure (MTTF) \cite{musa1987software}
+	\cite{GhezziEtAl2003}.
+
+	\item \textbf{Robustness} Software possesses the characteristic of robustness if it behaves ``reasonably'' in two situations: i) when it encounters circumstances not anticipated in the requirements specification, and ii) when the assumptions in its requirements specification are violated \cite{ghezzi1991fundamentals} \cite{boehm2007software}.
+
+	\item \textbf{Usability} ``The extent to which a product can be used by specified users to achieve specified goals with effectiveness, efficiency, and satisfaction in a specified context of use" \cite{ISO/TR16982:2002} \cite{ISO9241-11:2018}.
+
+	\item \textbf{Maintainability} The effort with which a software system or component can be modified to i) correct faults; ii) improve performance or other attributes; iii) satisfy new requirements \cite{IEEEStdGlossarySET1990} \cite{boehm2007software}.
+
+	\item \textbf{Reusability} ``The extent to which a software component can be used with or without adaptation in a problem solution other than the one for which it was originally developed" \cite{kalagiakos2003non}.
+
+	\item \textbf{Understandability} ``The capability of the software product to enable the user to understand whether the software is suitable, and how it can be used for particular tasks and conditions of use" \cite{iso2001iec}.
+
+	\item \textbf{Visibility/Transparency} The extent to which all of the steps of a software development process and the current status of it are conveyed clearly \cite{ghezzi1991fundamentals}.
+
+	\item \textbf{Reproducibility} ``A result is said to be reproducible if another researcher can take the original code and input data, execute it, and re-obtain the same result" \cite{BenureauAndRougier2017}.
 \end{itemize}
@@
 \label{sec_AHP}
-To generate grading scores for a group of software packages, we use the AHP to pairwise compare them.
+To generate ranking scores for a set of software packages, we use AHP, which utilizes pairwise comparisons between all of the packages.
 Thomas L.
 Saaty developed this tool, and people widely used it to make and analyze multiple criteria decisions \cite{VaidyaEtAl2006}.
-The AHP organizes multiple criteria factors in a hierarchical structure and pairwise compares the alternatives to calculate relative ratios \cite{Saaty1990}.
+AHP organizes multiple criteria factors in a hierarchical structure and uses pairwise comparisons between alternatives to calculate relative ratios \cite{Saaty1990}.
 
-For a project with $ m $ criteria, we can use a  $m\times m$ matrix $A$ to record the relative importance between factors.
+For a project with $ m $ criteria, we can use an $m\times m$ matrix $A$ to record the relative importance between factors.
 By pairwise compare criterion $i$ and criterion $j$, the value of $A_{ij}$ is decided as follows, and the value of $A_{ji}$ is $1/A_{ij}$ \cite{Saaty1990},
@@
 
-The priority vecotr $w$ can be calculated by solving the following equation \cite{Saaty1990}, \begin{equation}
+The priority vector $w$ can be calculated by solving the following equation \cite{Saaty1990}, \begin{equation}
 Aw = \lambda_{max}w,
@@
 
-In this project, $w$ is approximated with the approach classic \textit{mean of normalized values}  \cite{AlessioEtAl2006},
+In this project, $w$ is approximated with the classic \textit{mean of normalized values} approach \cite{AlessioEtAl2006},
 
@@
 
-In this project, the 9 software qualities mentioned above are the criteria ($m = 9$), while 29 software packages ($n = 29$) are compared.
-The software are evaluated with the grading template in Appendix \ref{ap_grading_template} and a subjective score is given for each quality.
-For a pair of qualities or software, $i$ and $j$, such that $i$ is not less significant than $j$, the pairwise comparison result of $i$ versus $j$ is converted from $min((score_i - score_j) + 1, 9)$.
+In this project, the 9 software qualities mentioned above are the criteria ($m = 9$), while 29 software packages ($n = 29$) are compared for each of the $m$ criteria.
+The software are evaluated with the grading template in Appendix \ref{ap_grading_template} and a subjective score from one to ten is given for each quality for each package.
+To convert the results into pair-wise scores, we apply the following steps: for each quality, for a pair of packages $i$ and $j$, such that $score_i >= score_j$, the pairwise comparison result of $i$ versus $j$ is converted from $min((score_i - score_j) + 1, 9)$.

### 2021-08-08 StateOfPractice/Ao-Report/13Methods.tex
MI report - added Abbreviations and Acronyms
MI report - update ch 3

--- 2021-07-26 StateOfPractice/Ao-Report/13Methods.tex
@@
-\chapter{Methods}
+\chapter{Research Methodology}
 \label{ch_methods}
 
-We designed a general process for evaluating the state of the practice of domain-specific software, that we instantiate to SC software for specific scientific domains.
+We designed a general process for evaluating the state of the practice of domain-specific software, that we instantiate to SC software for a specific scientific domain.
 
-Our method involves: 1) choosing a software domain (Section \ref{sec_domain_selection}); 2) collecting and filtering software packages (Section \ref{sec_software_selection}); 3) grading the selected software (Section \ref{sec_grading_software}); 4) interviewing development teams for further information (Section \ref{sec_interview_methods}).
+Our methodology involves several steps, shown as follows:
+\begin{enumerate}
+\item choosing a software domain (Section \ref{sec_domain_selection});
+\item collecting and filtering software packages (Section \ref{sec_software_selection});
+\item grading the selected software (Section \ref{sec_grading_software});
+\item interviewing development teams for further information (Section \ref{sec_interview_methods}).
+\end{enumerate}
 
-Details of how we applied the method on the MI domain are expanded upon in Section \ref{sec_applying_method}.
+Section \ref{sec_applying_method} presents an example of how we applied the methodology on the MI domain.
 
@@
 Thus, we need enough software candidates from the beginning.
-Besides, we prefer OSS projects because our grading method works better on them.
-Moreover, we prefer domains with active communities developing and using SC software, so it is easier to invite enough developers for interviews.
+Besides, we prefer OSS projects because our grading method requires access to the code.
+In addition, we prefer a domain with an active community developing and using the software.
+As a result, it is easier to invite enough developers for interviews.
 
@@
 Then, we search various websites, such as \hyperlink{https://github.com/}{GitHub}, \hyperlink{https://swmath.org/}{swMATH} and the Google search results for software recommendation articles.
-Meanwhile, we include the suggested ones from domain experts \cite{SmithEtAl2021}.
+We should also include the ones suggested by the domain experts \cite{SmithEtAl2021}.
 
@@
 The only ``mandatory'' requirement is that the software must be OSS, as defined in Section \ref{sec_open_source_software}.
-Because to evaluate all software qualities, our methods need to access the source code.
+We need this because evaluating some software qualities requires the source code.
 
-The other factors to filter the list are optional, and we consider them according to the number of software candidates and the objectives of the research project.
+The other filters are optional, and we consider them according to the number of software candidates and the objectives of the research project. We try to apply them in the following priority order:
 
-One factor is the functions and purposes of the software.
+\begin{enumerate}
+\item The functions and purpose of the software.
 For example, we can choose a group of software with similar functions, so that the comparison is between the software; or we can cross-compare sub-groups in the domain, then we need to select candidates from each sub-group.
 
+\item The version control tool.
 The empirical measurement tools listed in Section \ref{sec_empirical_measurements} only work on projects using \hyperlink{https://git-scm.com/}{Git}, so we prefer software with Git.
@@
 
+\item The age of software.
 Some of the OSS projects may experience a lack of recent maintenance.
 So we eliminate packages without recent updates, unless they are still popular and highly recommended by the domain users \cite{SmithEtAl2021}.
+We consider a software project as ``alive" if it has any update within the last 18 months; otherwise, we mark it as ``dead".
+\end{enumerate}
 
-With domain experts in the team, we value their opinions on the filtering process.
+The order of filters 2 and 3 is flexible. We adjust it according to the number of software packages affected by the filters, and the number of ones remaining on the list.
+
+Before showing our filtered list to the domain experts, we ask them to list the top 10 software in the domain.
+Then, we cross-compare the two lists and discuss the commons and differences.
+We value their opinions on the filtering process.
 For example, if a software package is not OSS and has no updates for a long while, but the domain experts identify it as a valuable product, we still consider it to keep.
@@
 \label{sec_grading_template}
-The full grading template can be found in Appendix \ref{ap_grading_template}. The template contains 101 questions that we use for grading software products.
+The full grading template can be found in Appendix \ref{ap_grading_template}.
+The template contains 103 questions that we use for grading software products.
+Figure \ref{fg_grading_template_example} shows an example of this grading template.
+
+\begin{figure}[h]
+\includegraphics[scale=0.42]{figures/template.png}
+\caption{Grading template example}
+\label{fg_grading_template_example}
+\end{figure}
 
@@
 Thus we check each project for its artifacts, such as API documentation, bug tracker, release notes, test cases, build files, version control, etc.
-We also check the tools supporting issue tracking and version control and the percentages of closed issues and comment lines in code.
+We also check the tools supporting issue tracking and version control, the percentages of closed issues, and the proportion of comment lines in code.
 
@@
 Projects with a large number of components provide more choices to reuse.
-Furthermore, well modularized code, which tend to have smaller parts in separated files, are easier to reuse.
+Furthermore, well-modularized code, which tends to have smaller parts in separated files, is easier to reuse.
 Thus, we consider the projects with more code files to be more reusable.
@@
 \item \textbf{Surface Understandability} We randomly examine 10 code files.
-We check the code’s style within each code file, such as whether the identifiers, parameters, indentation, and formatting are consistent, whether the constants (other than 0 and 1) are hardcoded into the code, and whether the developers modularized the code.
+We check the code’s style within each file, such as whether the identifiers, parameters, indentation, and formatting are consistent, whether the constants (other than 0 and 1) are hardcoded, and whether the developers modularized the code.
 We also check the descriptive information for the code, such as documents mentioning the coding standard, the comments in the code, and the descriptions or links for algorithms in the code.
@@
 For some qualities, the empirical measurements also affect the score.
-We use two command-line software tools \textit{GitStats} and \textit{scc} to extract information from the source code repositories.
+We use tools to extract information from the source code repositories.
 For projects held on GitHub, we manually collect additional metrics, such as the stars of the GitHub repository, and the numbers of open and closed pull requests.
@@
 
-Additionally, we also manually collect information for projects held on GitHub, such as the numbers of stars, forks, people watching this repository, open pull requests, and closed pull requests.
+Additionally, we manually collect information for projects held on GitHub, such as the numbers of stars, forks, people watching this repository, open pull requests, closed pull requests, number of months a repository has been on GitHub.
+A git repository can have a creation date much earlier than the first day on GitHub.
+For example, the developers created the git repository of \textit{3D Slicer} in 2002, but did not upload a copy of it to GitHub until 2020.
+We get the creation date of the GitHub copy by using API \textit{https://api.github.com/repos/{:owner}/{:repository}} (e.g., \hyperlink{https://api.github.com/repos/slicer/slicer}{https://api.github.com/repos/slicer/slicer}).
+In the response, the value of ``created\_at" is what we want.
+The number of months a repository has been on GitHub helps us understand the average change of metrics over time, e.g., the average new stars per month.
 
@@
 
-We spend about two hours grading one package, unless we find technical issues and need more time to resolve them.
+We spend about two hours grading each package, unless we find technical issues and need more time to resolve them.
 In most of the situation, we finish all the measurements for one software on the same day.
@@
 
-\subsection{Interviewee Selection}
-For a software list with a length of roughly 30, we aim to interview about ten development teams.
-Interviewing multiple individuals from each team gives us more comprehensive information, but a single engineer well-knowing the project is also sufficient.
+This section introduces our interview questions (Section \ref{sec_interview_questions}), method of selecting interviewees (Section \ref{sec_interviewee_selection}), and interview process (Section \ref{sec_interview_process}).
 
-Ideally, we select projects after the grading measurements and prefer the ones with higher overall scores. However, if we do not find enough participants, we contact all teams on the list.
-
-We try to find the contacts of the teams on the projects' websites, such as the official web pages, repositories, publications, and bio pages of the teams' institutions.
-Then, we send at most two emails to one contact asking for its participation before receiving any replies.
-
-\subsection{Interview Question Selection}
-
-We have a list of 20 questions to guide our interviews, which can be found in Section \ref{ch_interview} and Appendix \ref{ap_interview}.
+\subsection{Interview Questions}
+\label{sec_interview_questions}
+We designed a list of 20 questions to guide our interviews, which can be found in Section \ref{ch_interview} and Appendix \ref{ap_interview}.
 
@@
 
+\subsection{Interviewee Selection}
+\label{sec_interviewee_selection}
+For a software list with a length of roughly 30, we aim to interview about ten development teams.
+Interviewing multiple individuals from each team gives us more comprehensive information, but a single engineer well-knowing the project is also sufficient.
+
+Ideally, we select projects after the grading measurements and prefer the ones with higher overall scores. However, if we do not find enough participants, we contact all teams on the list.
+
+We try to find the contacts of the teams on the projects' websites, such as the official web pages, repositories, publications, and bio pages of the teams' institutions.
+Then, we send at most two emails to one contact asking for its participation before receiving any replies.
+We operate the invitation according to our ethics approval, such as the one in Appendix \ref{ap_ethics}.
+For example, we ask for participants' consent before interviewing them, recording the conversation, or including it in our report.
+
 \subsection{Interview Process}
+\label{sec_interview_process}
 Since the members of the development teams are usually around the world, we organize these interviews as virtual meetings online with \hyperlink{https://zoom.us/}{Zoom}.
-After receiving consent from the interviewees, we also record our discussions to transcribe them better.
+After receiving consent from the interviewees, we also record and transcribe our discussions.
 
@@
 
-Based on the principles in Section \ref{sec_domain_selection}, we selected the MI domain and the sub-group of software with Visualization function shown in Figure \ref{fig_mi_functions}.
-We also included MI domain experts in our team.
+Based on the principles in Section \ref{sec_domain_selection}, we selected the MI domain and the sub-group of software with the Visualization function shown in Figure \ref{fig_mi_functions}.
+We also included Dr.
+Michael Noseworthy and some of his students as the MI domain experts in our team.
 
@@
 
-Appendix \ref{ap_list_before_filtering} shows all the 48 software packages.
-Among them, there were several ones that we could not find in their source code, such as \textit{MicroDicom}, \textit{Aliza}, and \textit{jivex}, etc.
+Appendix \ref{ap_list_before_filtering} shows all the 48 software packages. Our filtering process is as follows:
+
+\begin{enumerate}
+\item Among them, there were eight ones that we could not find their source code, such as \textit{MicroDicom}, \textit{Aliza}, and \textit{jivex}.
 These packages are likely to be freeware defined in Section \ref{sec_freeware} and not OSS.
 So following guidelines in Section \ref{sec_filter_software_list} we removed them from the list.
-Next, we focused on the MI software providing visualization functions.
-Some of the software on the list were tool kits or libraries for other software to use as dependencies but not for end-users to view medical images, such as \textit{VTK}, \textit{ITK}, \textit{dcm4che}, etc.
+
+\item Next, we focused on the MI software providing visualization functions, as described in Section \ref{sec_scope}.
+Seven of the software on the list were tool kits or libraries for other software to use as dependencies, but not for end-users to view medical images, such as \textit{VTK}, \textit{ITK}, and \textit{dcm4che}; another three were PACS.
 We also eliminated these from the list.
+
+\item Finally, we removed \textit{Open Dicom Viewer} from the list because it had not received any updates for a long time.
+\end{enumerate}
+
 After that, there were 29 software products left on the list.
-We still preferred projects using git and GitHub and being updated recently, but did not apply another filtering since the number of packages was already below 30.
+We still preferred projects using git and GitHub and being updated recently, but did not apply this filter since packages were already below 30.
 However, 27 out of the 29 software packages on the final list used git, and 24 chose GitHub.
@@
 
-Going through the interview process in Section \ref{sec_interview_methods}, we started with the teams with higher scores on our list, and eventually contacted all of them.
+Before contacting any interviewee candidate, we received ethics clearance from the McMaster University Research Ethics Board (Appendix \ref{ap_ethics}).
+Going through the interview process in Section \ref{sec_interview_methods}, we contacted all of the 29 teams.
 As a result, developers/architects from 8 teams have participated in our interviews so far.
-Before contacting any interviewee candidate, we received ethics clearance from the McMaster University Research Ethics Board (Appendix \ref{ap_ethics}).

### 2021-08-08 StateOfPractice/Ao-Report/13Methods.tex
MI report - update ch 3

@@
 
-\item Finally, we removed \textit{Open Dicom Viewer} from the list because it had not received any updates for a long time.
+\item Finally, we removed \textit{Open Dicom Viewer} from the list because it had not received any updates for a long time (since 2011).
+After that, only \textit{MatrixUser} and \textit{AMIDE} were considered as ``dead".
+However, both of them had much more recent updates (after 2017) than \textit{Open Dicom Viewer}.
 \end{enumerate}
 
-After that, there were 29 software products left on the list.
+There were 29 software products left on the list.
 We still preferred projects using git and GitHub and being updated recently, but did not apply this filter since packages were already below 30.
-However, 27 out of the 29 software packages on the final list used git, and 24 chose GitHub.
-Furthermore, 27 packages had the latest updates after the year 2018, and 23 after 2020.
+Even without this filter, 27 out of the 29 software packages on the final list used git, and 24 chose GitHub.
 

### 2021-08-10 StateOfPractice/Ao-Report/15InterviewsWithDevelopers.tex
MI report - update ch 5 and other parts

@@
 We also asked following-up questions when we thought it was worth diving deeper.
-One participant was too busy to have an interview, so he wrote down the answers and sent them to us.
+One participant was too busy to have an interview, so they wrote down their answers.
 
-In this section, we only discuss the answers which we think are the most exciting and essential.
-We summarize the rest part of the interviews in Appendix \ref{ap_interview}.
+In this section, we only discuss the answers that we think are the most exciting and essential.
+We summarize the rest of the interviews in Appendix \ref{ap_interview}.
 The interviewees may have provided multiple answers to each question.
@@
 We also asked the interviewees how they would solve the problems.
+Questions 9, 10, and 12 are as follows:
 
-Questions 9, 10, and 12 are as follows,
-
-\textit{Q9. Currently, what are the most significant obstacles in your development process?}
-
-\textit{Q10. How might you change your development process to remove or reduce these obstacles?}
-
-\textit{Q12. In the past, is there any major obstacle to your development process that has been solved? How did you solve it?}
+\begin{description}
+\item[Q9.] Currently, what are the most significant obstacles in your development process?
+\item[Q10.] How might you change your development process to remove or reduce these obstacles?
+\item[Q12.] In the past, is there any major obstacle to your development process that has been solved? How did you solve it?
+\end{description}
 
@@
 \hline
-\multirow{2}{*}{Group} & \multirow{2}{*}{Obstacle} & \multicolumn{2}{l}{Number of interviewees with the answer} \\ \cline{3-4} 
- &  & as a current obstacle & as a past obstacle \\ \hline
+\multirow{2}{*}{Grp} & \multirow{2}{*}{Obstacle} & \multicolumn{2}{l}{Num ans.} \\ \cline{3-4} 
+ &  & current & past \\ \hline
 \multirow{2}{*}{1} & Lack of fundings & 3 &  \\
  & Lack of time to devote to the project & 2 & 1 \\ \hdashline
-\multirow{3}{*}{2} & \begin{tabular}[c]{@{}l@{}}Hard to keep up with changes in OS \\ and libraries\end{tabular} & 1 &  \\
+\multirow{3}{*}{2} & Hard to keep up with changes in OS and libraries & 1 &  \\
  & Hard to support multiple OS & 2 &  \\
@@
 3 & Lack of access to real-world datasets for testing & 3 & 2 \\ \hdashline
-\multirow{7}{*}{Others}
+\multirow{7}{*}{Oths}
  & Hard to have a high level roadmap from the start & 1 &  \\
  & Not enough participants for usability tests & 1 &  \\
- & \begin{tabular}[c]{@{}l@{}}Codebase is so big and complicated and only\\ a few people fully understand it\end{tabular} & 1 &  \\
+ & Only a few people fully understand the large codebase & 1 &  \\
  & Hard to transfer to new technologies & & 2 \\
@@
 
-\textit{Potential solutions from interviewees:}
+\noindent\textit{Potential solutions from interviewees:}
 \begin{itemize}
-\item shifting from development mode toward maintenance mode when developers are not enough;
+\item when the team does not have enough developers for building new features and fixing bugs at the same time, shifting from development mode toward maintenance mode;
 \item licensing the software to commercial companies that integrate it into their products;
@@
 
-\textit{Proven solutions from interviewees:}
+\noindent\textit{Proven solutions from interviewees:}
 \begin{itemize}
@@
 Many interviewees thought lack of fundings and lack of time were the most significant obstacles.
-The interviewees from \textit{3D Slicer} team and \textit{OHIF} team pointed out that it was more challenging to get fundings for software maintenance as opposed to researches.
-The interviewee from the \textit{ITK-SNAP} team thought enough fundings was a way to solve to problem of lacking time, because they could hire more dedicated developers.
-On the other hand, the interviewee from the \textit{ITK-SNAP} team did not think that fundings could solve the same problem, since he still would need a lot of time to supervise the project.
+The interviewees from \textit{3D Slicer} team and \textit{OHIF} team pointed out that it was more challenging to get fundings for software maintenance as opposed to research.
+The interviewee from the \textit{ITK-SNAP} team thought more fundings was a way to solve the lack of time problem, because they could hire more dedicated developers.
+On the other hand, the interviewee from the \textit{Weasis} team did not think that fundings could solve the same problem, since he still would need a lot of time to supervise the project.
 
-No interviewee suggested any solution to solve the funding problem fundamentally. However, they provided ideas to save time, such as better documentation, third-party plugins, and good CI/CD tools.
+No interviewee suggested any solution to bring extra funding to the project. However, they provided ideas to save time, such as better documentation, third-party plugins, and good CI/CD tools.
 
@@
 
-\textit{Potential solutions from interviewees:}
+\noindent\textit{Potential solutions from interviewees:}
 \begin{itemize}
@@
 \item to better support lower-end computers, using memory-mapped files to consume less computer memory;
-\item more fundings;
+\item more funding;
 \item maintaining better documentations to ease the development \& maintenance processes;
@@
 
-The interviewees talked about the advantages and disadvantages of the two choices. We summarize the opinions from the interviewees as follows.
+The interviewees talked about the advantages and disadvantages of the two choices. We summarize the opinions from the interviewees in Table \ref{tab_pros_cons_native_vs_web}.
 
-\textit{Native application}:
-\begin{itemize}
-\item Advantages
-    \begin{itemize}
-    \item higher performance
-    \end{itemize}
-\item Disadvantages
-    \begin{itemize}
-    \item hard to achieve cross-platform compatibility
-    \item more complicated build process
-    \end{itemize}
-\end{itemize}
-\textit{Web application}:
-\begin{itemize}
-\item Advantages
-    \begin{itemize}
-    \item easy to achieve cross-platform compatibility
-    \item simpler build process
-    \end{itemize}
-\item Disadvantages
-    \begin{itemize}
-    \item lower performance (without a backend)    \item may not meet the requirements of protecting patients' privacy (with a backend)
-    \item extra cost for backend servers (with a backend)
-    \end{itemize}
-\end{itemize}
+\begin{table}[H]
+\centering
+\hspace*{-2cm}\begin{tabular}{lll}
+\hline
+ & Native application & Web application \\ \hline
+Ad & - higher performance & \begin{tabular}[c]{@{}l@{}}- easy to achieve cross-platform compatibility\\ - simpler build process\end{tabular} \\ \hdashline
+Disad & \begin{tabular}[c]{@{}l@{}}- hard to achieve cross-platform compatibility\\ - more complicated build process\end{tabular} & \begin{tabular}[c]{@{}l@{}}\textit{Without a backend:} \\ - lower performance\\ \textit{With a backend:} \\ - harder for privacy protection \\ - extra cost for backend servers \end{tabular} \\ \hline
+\end{tabular}
+\caption{\label{tab_pros_cons_native_vs_web}Advantages and disadvantages of native application and web application}
+\end{table}
 
@@
 \label{sec_interview_documents}
-By asking questions 11 and 19, we tried to understand the interviewees' opinions on documentation. We also aimed to find out the quality of documentations in their projects.
+We tried to understand the interviewees' opinions on documentation and the quality of documentations with questions 11 and 19:
 
-Questions 11 and 19 are as follows,
-
-\textit{Q11. How does documentation fit into your development process? Would improved documentation help with the obstacles you typically face?}
-
-\textit{Q19. Do you think the current documentation can clearly convey all necessary knowledge to the users? If yes, how did you successfully achieve it? If no, what improvements are needed?}
+\begin{description}
+\item[Q11.] How does documentation fit into your development process? Would improved documentation help with the obstacles you typically face?
+\item[Q19.] Do you think the current documentation can clearly convey all necessary knowledge to the users? If yes, how did you successfully achieve it? If no, what improvements are needed?
+\end{description}
 
@@
 \hline
-Opinion on documentation & Number of interviewees with the answer \\ \hline
+Opinion on documentation & Num ans. \\ \hline
 Documentation is vital to the project& 8 \\
@@
 \hline
-Tool or method for documentation & Number of interviewees with the answer \\ \hline
+Tool or method for documentation & Num ans. \\ \hline
 Forum discussions & 3 \\
@@
 \label{sec_contribution_pm}
-By asking questions 5, 13, and 14, we tried to understand how the teams managed the contributions and their projects.
+We tried to understand how the teams managed the contributions and their projects by asking the questions:
 
-Questions 5, 13, and 14 are as follows,
+\begin{description}
+\item[Q5.] Do you have a defined process for accepting new contributions into your team?
+\item[Q13.] What is your software development model? For example, waterfall, agile, etc.
+\item[Q14.] What is your project management process? Do you think improving this process can tackle the current problem? Were any project management tools used?
+\end{description}
 
-\textit{Q5. Do you have a defined process for accepting new contributions into your team?}
+\textit{}
 
-\textit{Q13. What is your software development model? For example, waterfall, agile, etc.}
+\textit{}
 
-\textit{Q14. What is your project management process? Do you think improving this process can tackle the current problem? Were any project management tools used?}
+\textit{}
 
@@
 \hline
-\multirow{2}{*}{Method of receiving contributions} & \multicolumn{2}{l}{Number of interviewees with the answer} \\ \cline{2-3} 
- & as a current method & as a past method \\ \hline
+\multirow{2}{*}{Method of receiving contributions} & \multicolumn{2}{l}{Num ans.} \\ \cline{2-3} 
+ & current & past \\ \hline
 GitHub with pull requests & 8 & \\
@@
 \hline
-Software development model & Number of interviewees with the answer \\ \hline
+Software development model & Num ans. \\ \hline
 Undefined/self-directed & 3 \\
@@
 \hline
-Project management tools & Number of interviewees with the answer \\ \hline
+Project management tools & Num ans. \\ \hline
 GitHub & 3 \\
@@
 
-Most interviewees skipped the question ``Do you think improving this process can tackle the current problem?", but the interviewee from the \textit{OHIF} team gave a positive answer.
+Most interviewees skipped the question ``Do you think improving this process can tackle the current problem?".
+We concluded that a yes-or-no question is not very informative here.
+The interviewee from the \textit{OHIF} team gave a positive answer to this question.
+They believed that a better project management process can improve the efficiency of junior developers.
+They also improved the project management tools (from public Jira to public GitHub repository plus private Jira), so they could better communicate externally and internally.
 
@@
 \subsection{Correctness}
-\textit{Q15. Was it hard to ensure the correctness of the software? If there were any obstacles, what methods have been considered or practiced to improve the situation? If practiced, did it work?}
+
+\begin{description}
+\item[Q15.] Was it hard to ensure the correctness of the software? If there were any obstacles, what methods have been considered or practiced to improve the situation? If practiced, did it work?
+\end{description}
 
@@
 \hline
-Threat to correctness & Number of interviewees with the answer \\ \hline
+Threat to correctness & Num ans. \\ \hline
 \begin{tabular}[c]{@{}l@{}}Clinical systems produce data in various formats \\ (e.g. DICOM), which have complicated standards.\\ The software needs to handle the complexity.\end{tabular} & 2 \\
@@
 \hline
-Strategy to ensure correctness & Number of interviewees with the answer \\ \hline
+Strategy to ensure correctness & Num ans. \\ \hline
 \begin{tabular}[c]{@{}l@{}}Test-driven development / component tests /\\ integration tests / smoke tests / regression tests.\end{tabular} & 4 \\
@@
 \subsection{Maintainability}
-\textit{Q16.
-When designing the software, did you consider the ease of future changes?
+
+\begin{description}
+\item[Q16.] When designing the software, did you consider the ease of future changes?
 For example, will it be hard to change the system’s structure, modules, or code blocks?
-What measures have been taken to ensure the ease of future changes and maintains?}
+What measures have been taken to ensure the ease of future changes and maintains?
+\end{description}
 
@@
 \hline
-Strategy to ensure maintainability & Number of interviewees with the answer \\ \hline
+Strategy to ensure maintainability & Num ans. \\ \hline
 \begin{tabular}[c]{@{}l@{}}Modular approach / thinking the software\\ as reusable Lego bricks / maintain \\ repetitive functions as libraries.\end{tabular} & 5 \\
@@
 \subsection{Understandability}
-\textit{Q17. Provide instances where users have misunderstood the software. What, if any, actions were taken to address understandability issues?}
+
+\begin{description}
+\item[Q17.] Provide instances where users have misunderstood the software. What, if any, actions were taken to address understandability issues?
+\end{description}
 
@@
 \hline
-Threat to understandability & Number of interviewees with the answer \\ \hline
+Threat to understandability & Num ans. \\ \hline
 \begin{tabular}[c]{@{}l@{}}Not all users understand how to use some\\ features of the software.\end{tabular} & 2 \\
@@
 \hline
-Strategy to ensure understandability & Number of interviewees with the answer \\ \hline
+Strategy to ensure understandability & Num ans. \\ \hline
 \begin{tabular}[c]{@{}l@{}}Documentation / user manual /\\ user mailing list / forum.\end{tabular} & 4 \\
@@
 \subsection{Usability}
-\textit{Q18. What, if any, actions were taken to address usability issues?}
+
+\begin{description}
+\item[Q17.] What, if any, actions were taken to address usability issues?
+\end{description}
 
@@
 \hline
-Strategy to ensure usability & Number of interviewees with the answer \\ \hline
+Strategy to ensure usability & Num ans. \\ \hline
 Usability tests and interviews with end users. & 3 \\
@@
 \subsection{Reproducibility}
-\textit{Q20. Do you have any concern that your computational results won’t be reproducible in the future? Have you taken any steps to ensure reproducibility?}
+
+\begin{description}
+\item[Q20.] Do you have any concern that your computational results won’t be reproducible in the future? Have you taken any steps to ensure reproducibility?
+\end{description}
 
@@
 \hline
-Threat to reproducibility & Number of interviewees with the answer \\ \hline
+Threat to reproducibility & Num ans. \\ \hline
 \begin{tabular}[c]{@{}l@{}}If the software is closed-source, the reproducibility\\is hard to achieve.\end{tabular} & 1 \\
@@
 \hline
-Strategy to ensure reproducibility & Number of interviewees with the answer \\ \hline
+Strategy to ensure reproducibility & Num ans. \\ \hline
 Regression tests / unit tests / having good tests. & 6 \\

### 2021-08-10 StateOfPractice/Ao-Report/18Conclusion.tex
MI report - update ch 5 and other parts

@@
 
-With learnings from this project, we summarized recommendations for the future state of the practice assessments,
+With learnings from this project, we summarized recommendations for the future state of the practice assessments:
 \begin{itemize}
@@
 	\item the grading standard can be more explicit. For example, we can explicitly define scores for each item in the grading temple.
+	\item we can improve some interview questions.
+For example, in question 14, ``Do you think improving this process can tackle the current problem?" is a yes-or-no question, which is not informative enough.
+As mentioned in Section \ref{sec_contribution_pm}, most interviewees ignored it.
+We can change it to ``By improving this process, what current problems can be tackled?"
 	\item we can better organize the interview questions.
-We think they are good questions.
-However, since we use audio conversion tools to transcribe the answers, we should aim to make the transcription easier to read.
+Since we use audio conversion tools to transcribe the answers, we should aim to make the transcription easier to read.
 For example, we can order them together for questions about the five software qualities and compose a similar structure for each.

### 2021-08-10 StateOfPractice/Ao-Report/33InterviewAnswers.tex
MI report - update ch 5 and other parts

@@
 
-\noindent\textbf{Q1. Interviewees’ current position/title? degrees?}
+\begin{description}
+\item[Q1.] Interviewees’ current position/title? degrees?
+\end{description}
 
@@
 \hline
-Highest degree & Number of interviewees \\ \hline
+Highest degree & Num ans. \\ \hline
 PHD & 4 \\
@@
 \hline
-Major & Number of interviewees \\ \hline
+Major & Num ans. \\ \hline
 Computer Science & 4 \\
@@
 
-\noindent\textbf{Q2. Interviewees’ contribution to/relationship with the software?}
+\begin{description}
+\item[Q2.] Interviewees’ contribution to/relationship with the software?
+\end{description}
 
@@
 \hline
-Role in the projects & Number of interviewees \\ \hline
+Role in the projects & Num ans. \\ \hline
 Chief Architect & 2 \\
@@
 
-\noindent\textbf{Q3. Length of time the interviewee has been involved with this software?}
+\begin{description}
+\item[Q3.] Length of time the interviewee has been involved with this software?
+\end{description}
 
@@
 \hline
-Length of time in the projects & Number of interviewees \\ \hline
+Years & Num ans. \\ \hline
 0-1 & 1 \\
@@
 
-\noindent\textbf{Q4. How large is the development group?}
+\begin{description}
+\item[Q4.] How large is the development group?
+\end{description}
 
@@
 \hline
-Number of current members & Number of projects \\ \hline
+Num mbrs & Num ans. \\ \hline
 1-3 & 5 \\
@@
 
-\noindent\textbf{Q6. What is the typical background of a developer?}
+\begin{description}
+\item[Q6.] What is the typical background of a developer?
+\end{description}
 
@@
 \hline
-Background of a developer & Number of interviewees with the answer \\ \hline
+Background of a developer & Num ans. \\ \hline
 \begin{tabular}[c]{@{}l@{}}Computer Science, Information\\ Technology, and Software Development \end{tabular} & 6 \\
@@
 
-\noindent\textbf{Q7. What is your estimated number of users? How did you come up with that estimate?}
+\begin{description}
+\item[Q7.] What is your estimated number of users? How did you come up with that estimate?
+\end{description}
 
@@
 \centering
-\begin{tabular}{lll}
+\hspace*{-2cm}\begin{tabular}{lll}
 \hline
 Software & Rough estimation & Considered facts \\ \hline
-3D Slicer & 100,000 & \begin{tabular}[c]{@{}l@{}} Search results on Google Scholar;\\ number of new posts per year on slicer.org; \\ number of downloads. \end{tabular} \\
+3D Slicer & 100,000 & \begin{tabular}[c]{@{}l@{}} 1) search results on Google Scholar;\\ 2) number of new posts per year on slicer.org; \\ 3) number of downloads. \end{tabular} \\
 INVESALIUS 3 & 75,000 & Number of random IDs created by new installation.\\
 dwv & No estimation & About 20 companies integrated \textit{dwv} in their products. \\
-BioImage Suite Web & 100 active users & \begin{tabular}[c]{@{}l@{}}  The  interviewee only counted the users from \\ several Universities who were active users. \end{tabular} \\
+BioImage Suite Web & 100 active users &  Only counted the active users from several Universities. \\
 ITK-SNAP & 10,000 plus & Number of downloads. \\
-MRIcroGL & No estimation & \begin{tabular}[c]{@{}l@{}}  It is the top 1 downloaded software on this NITRC list \\ \hyperlink{https://www.nitrc.org/top/toplist.php?type=downloads}{https://www.nitrc.org/top/toplist.php?type=downloads} \end{tabular} \\
-Weasis & \begin{tabular}[c]{@{}l@{}}  10,000 user used \\ it as least once \end{tabular} & Number of profiles. \\
-OHIF & About 5000 & \begin{tabular}[c]{@{}l@{}}   Some platforms integrated \textit{OHIF}, and it was hard \\ to know the number of end users. \end{tabular} \\ \hline
+MRIcroGL & No estimation & The top 1 downloaded software on the NITRC list. \\
+Weasis & 10,000 incl. one-time users & Number of profiles. \\
+OHIF & About 5000 & Some platforms integrated \textit{OHIF}; hard to estimate. \\ \hline
 \end{tabular}
@@
 
-\noindent\textbf{Q8. What is the typical background of a user?}
+\begin{description}
+\item[Q8.] What is the typical background of a user?
+\end{description}
 
@@
 \hline
-Background of a user & Number of interviewees with the answer \\ \hline
+Background of a user & Num ans. \\ \hline
 Medical Researchers & 6 \\

### 2021-08-10 StateOfPractice/Ao-Report/11Introduction.tex
MI report - updated research question format

@@
 
-\begin{enumerate}
-\item What are the pain points for developers working on MI software projects? What solutions the developers have tried to address the pain points?
-\item What artifacts the projects generated?
-\item What role does documentation play in the projects? What are the developers' attitudes toward it?
-\item What principles, processes, methodologies, and tools the projects used?
-\item What is the current status of the following software qualities for the projects? What actions have the developers taken to address them?
+\begin{description}
+\item[RQ1.] What are the pain points for developers working on MI software projects? What solutions the developers have tried to address the pain points?
+\item[RQ2.] What artifacts the projects generated?
+\item[RQ3.] What role does documentation play in the projects? What are the developers' attitudes toward it?
+\item[RQ4.] What principles, processes, methodologies, and tools the projects used?
+\item[RQ5.] What is the current status of the following software qualities for the projects? What actions have the developers taken to address them?
 \begin{itemize}
@@
 \end{itemize}
-\item How does the software quality ranking generated by our methods compare with the ratings from the community?
-\end{enumerate}
+\item[RQ6.] How does the software quality ranking generated by our methods compare with the ratings from the community?
+\end{description}
 

### 2021-08-10 StateOfPractice/Ao-Report/16AnswersToResearchQuestions.tex
MI report - updated research question format

@@
 \label{sec_rq1_pain_points}
-\textit{RQ1. What are the pain points for developers working on MI software projects? What solutions have the developers tried to address the pain points?}
+\begin{description}\item[RQ1.] What are the pain points for developers working on MI software projects? What solutions have the developers tried to address the pain points?\end{description}
 
@@
 \label{sec_rq2_artifacts}
-\textit{RQ2. What artifacts the projects generated?}
+\begin{description}\item[RQ2.] What artifacts the projects generated?\end{description}
 
@@
 \label{sec_rq3_documentation}
-\textit{RQ3. What role does documentation play in the projects? What are the developers' attitudes toward it?}
+\begin{description}\item[RQ3.] What role does documentation play in the projects? What are the developers' attitudes toward it?\end{description}
 
@@
 \label{sec_rq4_PPMT}
-\textit{RQ4. What principles, processes, methodologies, and tools the projects used?}
+\begin{description}\item[RQ4.] What principles, processes, methodologies, and tools the projects used?\end{description}
 
@@
 \label{sec_rq5_qualities}
-\textit{RQ5. What is the current status of the following software qualities for the projects? What actions have the developers taken to address them?}
+\begin{description}\item[RQ5.] What is the current status of the following software qualities for the projects? What actions have the developers taken to address them?\end{description}
 
@@
 \label{sec_rq6_comparison}
-\textit{RQ6. How does the software quality ranking generated by our methods compare with the ratings from the community?}
+\begin{description}\item[RQ6.] How does the software quality ranking generated by our methods compare with the ratings from the community?\end{description}
 

### 2021-08-10 StateOfPractice/Ao-Report/17Recommendations.tex
MI report - minor updates

@@
 Section \ref{sec_recommendations_qualities} discusses the actions that can potentially improve the ten software qualities.
-Sections \ref{sec_recommendations_limited_resources}, \ref{sec_recommendations_technology_stack}, and \ref{sec_recommendations_testing_dataset} are based on the primary pain points collected from the developers in the MI domain, but we believe scientists and developers are likely to face them in most SC domains.
+Sections \ref{sec_recommendations_limited_resources}, \ref{sec_recommendations_tech_stack}, and \ref{sec_recommendations_testing_dataset} are based on the primary pain points collected from the developers in the MI domain, but we believe scientists and developers are likely to face them in most SC domains.
 These sections contain our general suggestions to tackle them.

### 2021-08-11 StateOfPractice/Ao-Report/13Methods.tex
MI report - updates

@@
-\chapter{Research Methodology}
+\chapter{Methodology}
 \label{ch_methods}
@@
 
-Our methodology involves several steps, shown as follows:
+Our method involves several steps, shown as follows:
 \begin{enumerate}
@@
 
-Section \ref{sec_applying_method} presents an example of how we applied the methodology on the MI domain.
+Section \ref{sec_applying_method} presents an example of how we applied the method on the MI domain.
 
@@
 
+\subsection{Vet the Software List}
+\label{sec_vet_software_list}
 Before showing our filtered list to the domain experts, we ask them to list the top 10 software in the domain.
 Then, we cross-compare the two lists and discuss the commons and differences.
-We value their opinions on the filtering process.
+In addition, we ask them to vet our filtered list.
+They provide opinions on whether the list is reasonable to them.
+We also value their opinions on the filtering process.
 For example, if a software package is not OSS and has no updates for a long while, but the domain experts identify it as a valuable product, we still consider it to keep.
@@
 \label{sec_interview_process}
+Before contacting any interviewee candidate, we need to receive ethics clearance from the McMaster University Research Ethics Board.
 Since the members of the development teams are usually around the world, we organize these interviews as virtual meetings online with \hyperlink{https://zoom.us/}{Zoom}.
@@
 \label{sec_applying_method}
+This section shows an overview of applying our method to the MI domain.
 
+\subsection{Domain Selection}
 Based on the principles in Section \ref{sec_domain_selection}, we selected the MI domain and the sub-group of software with the Visualization function shown in Figure \ref{fig_mi_functions}.
@@
 
+\subsection{Software Product Selection}
 By using the method in Section \ref{sec_identify_software_candidates}, we identified 48 MI software projects as the candidates from publications \cite{Bjorn2017} \cite{Bruhschwein2019} \cite{Haak2015}, online articles related to the domain \cite{Emms2019} \cite{Hasan2020} \cite{Mu2019}, forum discussions related to the domain \cite{Samala2014}, etc.
+Appendix \ref{ap_list_before_filtering} shows all the 48 software packages.
 
-Appendix \ref{ap_list_before_filtering} shows all the 48 software packages. Our filtering process is as follows:
+Guided by the method in Section \ref{sec_filter_software_list}, we filtered the list with a process as follows:
 
@@
 
-There were 29 software products left on the list.
 We still preferred projects using git and GitHub and being updated recently, but did not apply this filter since packages were already below 30.
-Even without this filter, 27 out of the 29 software packages on the final list used git, and 24 chose GitHub.
+Even without this filter, 27 out of the 29 software packages on the filtered list used git, and 24 chose GitHub.
 
+Following the process in Section \ref{sec_vet_software_list}, our domain experts provided a list of top software that contains 12 software products (Table \ref{tab_top_software_experts}).
+We compared two lists and found six common ones.
+We did not include the rest in our final list, because their major function was not Visualization.
+
+\begin{table}[H]
+\centering
+\begin{tabular}{ll}
+\hline
+Software & On both lists \\ \hline
+3D Slicer & X \\
+Horos & X \\
+ImageJ & X \\
+Fiji & X \\
+AFNI &  \\
+FSL &  \\
+Freesurfer &  \\
+Mricron & X \\
+Mango & X \\
+Tarquin &  \\
+Diffusion Toolkit &  \\
+MRItrix &  \\ \hline
+\end{tabular}
+\caption{\label{tab_top_software_experts}Top software by the MI domain experts}
+\end{table}
+
+We included \textit{Mango} in the initial list, but removed it because it was not OSS.
+However, we kept \textit{Papaya}, a the web version of \textit{Mango}.
+Instead of \textit{MRIcron}, we chose \textit{MRIcroGL}, because \textit{MRIcron} development had moved to
+\textit{MRIcroGL} \cite{Rorden2021b}.
+
+After vetting our filtered list, the domain experts believed it was reasonable and did not identify any problem.
+Thus, as shown in Appendix \ref{ap_list_before_filtering}, eventually we had 29 software products on the final list.
+
+\subsection{Grading Software}
 Then we followed the steps in Section \ref{sec_grading_software} to measure and grade the software.
@@
 
-Before contacting any interviewee candidate, we received ethics clearance from the McMaster University Research Ethics Board (Appendix \ref{ap_ethics}).
+\subsection{Interviews}
+We received ethics clearance from the McMaster University Research Ethics Board (Appendix \ref{ap_ethics}).
 Going through the interview process in Section \ref{sec_interview_methods}, we contacted all of the 29 teams.

### 2021-08-11 StateOfPractice/Ao-Report/14MeasurementResults.tex
MI report - minor updates
MI report - updates

@@
 
-This section shows the summary of the measurement results. The detailed data can be found in the repository \hyperlink{https://github.com/smiths/AIMSS}{https://github.com/smiths/AIMSS}.
+As discussed in Section \ref{sec_applying_method}, we applied our method to the MI domain.
+This section shows the summary of the measurement results.
+The detailed data can be found in the repository \hyperlink{https://data.mendeley.com/datasets/k3pcdvdzj2/1}{https://data.mendeley.com/datasets/k3pcdvdzj2/1}.
 
 Table \ref{tab_final_list} shows the 29 software packages that we measured.
-We tried to find the initial release dates (Rlsd) in the documents, such as the introduction and release notes.
-We did not find this date for two projects and marked them with ``?".
+We found the initial release dates (Rlsd) in their documents for most projects and marked the rest two with ``?".
 We used the date of the latest change to each code repository to decide the latest update.
-We found out funding information (Fnd) for eight projects, and we are not sure about the rest.
+We found out funding information (Fnd) for only eight projects.
 

### 2021-08-11 StateOfPractice/Ao-Report/15InterviewsWithDevelopers.tex
MI report - updates

@@
 
-This section summarizes some answers from the interviews with developers.
+We applied our method to the MI domain (Section \ref{sec_applying_method}) and summarize some answers from the interviews with developers in this section.
+We only discuss the answers that we think are the most exciting and essential in this section, and summarize the rest in Appendix \ref{ap_interview}.
+
 We interviewed nine developers from eight of the 29 MI software projects.
@@
 One participant was too busy to have an interview, so they wrote down their answers.
-
-In this section, we only discuss the answers that we think are the most exciting and essential.
-We summarize the rest of the interviews in Appendix \ref{ap_interview}.
 The interviewees may have provided multiple answers to each question.
-When counting the number of interviewees for each response, we count every interviewee with this answer.
+Thus, when counting the number of answers, the total result can be larger than nine.
 

### 2021-08-12 StateOfPractice/GitHubMetricsCollector/src/main/java/io/ao9/GitHubMetricsCollector/App.java
Created GitHub metrics collector

@@ -1,0 +1,217 @@
+package io.ao9.GitHubMetricsCollector;
+
+import com.fasterxml.jackson.databind.ObjectMapper;
+import org.apache.poi.ss.usermodel.Cell;
+import org.apache.poi.ss.usermodel.Row;
+import org.apache.poi.xssf.usermodel.XSSFSheet;
+import org.apache.poi.xssf.usermodel.XSSFWorkbook;
+
+import java.io.FileInputStream;
+import java.io.FileOutputStream;
+import java.io.IOException;
+import java.net.URI;
+import java.net.http.HttpClient;
+import java.net.http.HttpRequest;
+import java.net.http.HttpResponse;
+import java.time.LocalDate;
+import java.time.Period;
+import java.time.ZoneId;
+import java.util.*;
+
+public class App {
+    private final static String fileName = "GitHub_Metrics_Template.xlsx";
+    private final static String sheetName = "Sheet1";
+    private final static String gitHubUrl = "https://github.com";
+    private final static String gitHubApiUri = "https://api.github.com/graphql";
+    private final static String gitHubUsername = "your_GitHub_username";
+    // Do NOT commit your token to this git repo
+    private final static String gitHubToken = "your_GitHub_token";
+
+    public static void main(String[] args) {
+
+        try {
+            FileInputStream fileInput = new FileInputStream(fileName);
+            XSSFWorkbook workbook = new XSSFWorkbook(fileInput);
+            XSSFSheet sheet = workbook.getSheet(sheetName);
+            fileInput.close();
+
+            System.out.printf("Reading input data from file '%s'...\n\n", fileName);
+            List<String> repoUrls = readInputData(sheet);
+
+            System.out.println("Fetching repo data by GitHub API...\n");
+            List<RepoData> repoDataList = fetchRepoData(sheet, repoUrls);
+
+            System.out.println("Preparing output data...\n");
+            prepareOutputData(sheet, repoDataList);
+
+            System.out.printf("Writing to file '%s'...\n\n", fileName);
+            FileOutputStream fileOutput = new FileOutputStream(fileName);
+            workbook.write(fileOutput);
+            fileOutput.close();
+
+            System.out.println("Process Done!");
+        } catch (Exception e) {
+            e.printStackTrace();
+        }
+    }
+
+    private static List<String> readInputData(XSSFSheet sheet) {
+        /*
+         * Read repo urls
[…]

### 2021-08-12 StateOfPractice/GitHubMetricsCollector/src/main/java/io/ao9/GitHubMetricsCollector/RepoData.java

@@ -1,0 +1,25 @@
+package io.ao9.GitHubMetricsCollector;
+
+import lombok.*;
+
+import java.util.Date;
+import java.util.Map;
+
+/*
+ * lombok simply saves time to write
+ * getter, setter, constructor and toString
+ */
+@NoArgsConstructor
+@Data
+@ToString
+public class RepoData {
+    private Map<String, Integer> stargazers;
+    private Map<String, Integer> forks;
+    private Map<String, Integer> watchers;
+    private Map<String, Integer> openIssues;
+    private Map<String, Integer> totalIssues;
+    private Map<String, Integer> openPullRequests;
+    private Map<String, Integer> totalPullRequests;
+    private Date createdAt;
+    private Date updatedAt;
+}

### 2021-08-12 StateOfPractice/GitHubMetricsCollector/src/test/java/io/ao9/GitHubMetricsCollector/AppTest.java

@@ -1,0 +1,20 @@
+package io.ao9.GitHubMetricsCollector;
+
+import static org.junit.Assert.assertTrue;
+
+import org.junit.Test;
+
+/**
+ * Unit test for simple App.
+ */
+public class AppTest 
+{
+    /**
+     * Rigorous Test :-)
+     */
+    @Test
+    public void shouldAnswerWithTrue()
+    {
+        assertTrue( true );
+    }
+}

### 2021-08-16 StateOfPractice/Ao-Report/01abstract.tex
MI report - updated Abstract and table formats

--- 2021-05-09 StateOfPractice/Ao-Report/01abstract.tex
@@
 
-Abstract here
+\begingroup
+\setstretch{1.25}
+We present a general method to assess the state of the practice for Scientific Computing (SC) software and apply the method to the Medical Imaging (MI) software.
+This method guided us to select 29 MI software projects from 48 candidates, assess 10 software qualities (\textit{Installability}, \textit{Correctness \& Verifiability}, \textit{Reliability}, \textit{Robustness}, \textit{Usability}, \textit{Maintainability}, \textit{Reusability}, \textit{Understandability}, \textit{Visibility/Transparency}, and \textit{Reproducibility}) by answering 103 questions for each software, and interview eight of the 29 development teams.
+The results helped us with revealing the current status of MI software development.
+Based on the quantitative data for the first nine qualities, we ranked the MI software with the Analytic Hierarchy Process (AHP).
+The top three software products were \textit{3D Slicer}, \textit{ImageJ}, and \textit{OHIF Viewer}, which received high scores for most qualities.
+\textit{3D Slicer} was among the top two for all nine qualities except \textit{robustness}.
+By interviewing the developers, we identified three major types of pain points during their development process: i) the lack of resources; ii) the difficulty to balance between four factors: \textit{compatibility}, \textit{maintainability}, \textit{performance}, and \textit{security};
+iii) the lack of access to real-world datasets for testing.
+We collected proven and potential solutions for these problems.
+The interviews also helped us to understand the status of documentation, project management, and five qualities (\textit{correctness}, \textit{maintainability}, \textit{understandability}, \textit{usability}, and \textit{reproducibility}) in the projects.
+We summarized the threats and strategies to these qualities.
+For future SC software development, we proposed recommendations on improving software qualities, dealing with limited resources, choosing a tech stack, and enriching the testing datasets.
+The recommendations include adopting test-driven development, using continuous integration and continuous delivery (CI/CD), using git and GitHub, maintaining good documents, supporting third-party plugins or extensions, considering web application solutions, and establishing community collaboration in a SC domain.
+\newline
+
+\noindent\textbf{Keywords:} Medical Imaging, Scientific Computing, software engineering, software quality, Analytic Hierarchy Process, developer interview
+\endgroup

### 2021-08-16 StateOfPractice/Ao-Report/14MeasurementResults.tex
MI report - updated Abstract and table formats

@@
 Drishti \cite{Limaye2012} & 2012 & 2020-08 &  & 1 & 268168 & X & X & X &  \\
-\begin{tabular}[c]{@{}l@{}}BioImage Suite\\ Web \cite{Papademetris2005}\end{tabular} & 2018 & 2020-10 & X & 13 & 139699 & X & X & X & X \\
+BioImage Suite Web \cite{Papademetris2005} & 2018 & 2020-10 & X & 13 & 139699 & X & X & X & X \\
 OHIF Viewer \cite{Ziegler2020} & 2015 & 2020-10 &  & 76 & 63951 & X & X & X & X \\
@@
 MatrixUser \cite{Liu2016} & 2013 & 2018-07 &  & 1 & 23121 & X & X & X &  \\
-\end{tabular}
-\end{table}
-
-\begin{table}[H]
-\begin{tabular}{llllllllll}
-\hline
-\multirow{2}{*}{Software} & \multirow{2}{*}{Rlsd} & \multirow{2}{*}{Updated} & \multirow{2}{*}{Fnd} & \multirow{2}{*}{NOC} & \multirow{2}{*}{LOC} & \multicolumn{3}{c}{OS} & \multirow{2}{*}{W} \\ \cline{7-9}
- &  &  &  &  &  & W & M & L &  \\ \hline
 DICOM Viewer \cite{Afsar2021} & 2018 & 2020-04 & X & 5 & 30761 & X & X & X &  \\
@@
 AMIDE \cite{Loening2017} & 2006 & 2017-01 &  & 4 & 102827 & X & X & X &  \\
-Gwyddion \cite{Nevcas2012} & 2004 & 2020-11 &  & 38 & 643427 & X & X & X &  \\ \hline
+Gwyddion \cite{Nevcas2012} & 2004 & 2020-11 &  & 38 & 643427 & X & X & X &  \\
+\hline
 \end{tabular}

### 2021-08-16 StateOfPractice/Ao-Report/15InterviewsWithDevelopers.tex
MI report - updated Abstract and table formats

@@
 \centering
-\begin{tabular}{ll}
+\hspace*{-1cm}\begin{tabular}{ll}
 \hline
 Threat to correctness & Num ans. \\ \hline
-\begin{tabular}[c]{@{}l@{}}Clinical systems produce data in various formats \\ (e.g. DICOM), which have complicated standards.\\ The software needs to handle the complexity.\end{tabular} & 2 \\
-\begin{tabular}[c]{@{}l@{}}Different types of MI machines can create data \\ in slightly different ways, adding more complexity\\ for the software to handle.\end{tabular} & 2 \\
-\begin{tabular}[c]{@{}l@{}}Besides viewing, the software has additional \\ functions, which lead to extra complexity.\end{tabular} & 1 \\
-There is lack of real word image data for testing. & 1 \\
-\begin{tabular}[c]{@{}l@{}}The team cannot use private data for debugging,\\ even when the data cause problems.\end{tabular} & 1 \\
-\begin{tabular}[c]{@{}l@{}}The software uses huge datasets for testing, so \\ the tests are expensive and time-consuming.\end{tabular} & 1 \\
+Complexity - data in various formats and complicated standards. & 2 \\
+Complexity - different MI machines create data in different ways. & 2 \\
+Complexity - additional functions beside of viewing. & 1 \\
+The lack of real word image data for testing. & 1 \\
+The team cannot use private data for debugging even when the data cause problems. & 1 \\
+With huge datasets for testing, the tests are expensive and time-consuming. & 1 \\
 It is hard to well manage releases. & 1 \\
@@
 \centering
-\begin{tabular}{ll}
+\hspace*{-1cm}\begin{tabular}{ll}
 \hline
 Strategy to ensure correctness & Num ans. \\ \hline
-\begin{tabular}[c]{@{}l@{}}Test-driven development / component tests /\\ integration tests / smoke tests / regression tests.\end{tabular} & 4 \\
+Test-driven development, component / integration / smoke / regression tests. & 4 \\
 Self tests / automated tests. & 3 \\
-\begin{tabular}[c]{@{}l@{}}Two stage development process / stable release \&\\ nightly builds.\end{tabular} & 3 \\
+Two stage development process / stable release \& nightly builds. & 3 \\
 CI/CD. & 1 \\
-\begin{tabular}[c]{@{}l@{}}Using deidentified copies of medical images for\\ debugging.\end{tabular} & 1 \\
-\begin{tabular}[c]{@{}l@{}}Sending the beta versions of software to medical\\ workers who can access the data and do the tests.\end{tabular} & 1 \\
-\begin{tabular}[c]{@{}l@{}}Collecting and maintaining a dataset of\\ problematic images.\end{tabular} & 1 \\ \hline
+Using deidentified copies of medical images for debugging. & 1 \\
+Sending beta versions to medical workers who can access the data to do the tests. & 1 \\
+Collecting and maintaining a dataset of problematic images. & 1 \\ \hline
 \end{tabular}
@@
 Strategy to ensure maintainability & Num ans. \\ \hline
-\begin{tabular}[c]{@{}l@{}}Modular approach / thinking the software\\ as reusable Lego bricks / maintain \\ repetitive functions as libraries.\end{tabular} & 5 \\
+Modular approach / maintain repetitive functions as libraries. & 5 \\
 Supporting third-party extensions. & 1 \\
@@
 Threat to understandability & Num ans. \\ \hline
-\begin{tabular}[c]{@{}l@{}}Not all users understand how to use some\\ features of the software.\end{tabular} & 2 \\
-\begin{tabular}[c]{@{}l@{}}The team has no dedicated user experience\\ (UX) designer.\end{tabular} & 1 \\
-\begin{tabular}[c]{@{}l@{}}The software does not make some important\\ indicators noticeable (e.g. a progress bar).\end{tabular} & 1 \\
-\begin{tabular}[c]{@{}l@{}}Not all users understand the purpose of the\\ software.\end{tabular} & 1 \\
-\begin{tabular}[c]{@{}l@{}}Not all users know if the software includes\\ certain features.\end{tabular} & 1 \\
-\begin{tabular}[c]{@{}l@{}}Not all users understand how to use the\\ command line tool.\end{tabular} & 1 \\
-\begin{tabular}[c]{@{}l@{}}Not all users understand that the software is \\a web application.\end{tabular} & 1 \\\hdashline
-\begin{tabular}[c]{@{}l@{}}Not all developers understand how to deploy\\ the software.\end{tabular} & 1 \\
-\begin{tabular}[c]{@{}l@{}}The architecture is difficult for new\\ developers to understand.\end{tabular} & 1 \\ \hline
+Not all users understand how to use some features. & 2 \\
+The team has no dedicated user experience (UX) designer. & 1 \\
+Some important indicators are not noticeable (e.g. a progress bar). & 1 \\
+Not all users understand the purpose of the software. & 1 \\
+Not all users know if the software includes certain features. & 1 \\
+Not all users understand how to use the command line tool. & 1 \\
+Not all users understand that the software is a web application. & 1 \\\hdashline
+Not all developers understand how to deploy the software. & 1 \\
+The architecture is difficult for new developers to understand. & 1 \\ \hline
 \end{tabular}
@@
 Strategy to ensure understandability & Num ans. \\ \hline
-\begin{tabular}[c]{@{}l@{}}Documentation / user manual /\\ user mailing list / forum.\end{tabular} & 4 \\
+Documentation / user manual / user mailing list / forum. & 4 \\
 Graphical user interface. & 2 \\
 Testing every release with active users. & 1 \\
-\begin{tabular}[c]{@{}l@{}}Making simple things simple and complicated\\ things possible.\end{tabular} & 1 \\
+Making simple things simple and complicated things possible. & 1 \\
 Icons with more clear visual expressions. & 1 \\
@@
 Dialog windows for important notifications. & 1 \\
-\begin{tabular}[c]{@{}l@{}}Providing an example if the users need to build\\ the software by themselves.\end{tabular} & 1 \\ \hline
+Providing an example if the users need to build the software by themselves. & 1 \\ \hline
 \end{tabular}
@@
 Adjusting according to users' feedbacks. & 3 \\
-\begin{tabular}[c]{@{}l@{}}Straightforward and intuitively designed interface /\\ professional UX designer.\end{tabular} & 2 \\
-\begin{tabular}[c]{@{}l@{}}Providing step-by-step processes, and showing\\ the step numbers.\end{tabular} & 1 \\
-\begin{tabular}[c]{@{}l@{}}Making the basic functions easy to use without\\ reading the documentation.\end{tabular} & 1 \\
+Straightforward and intuitively designed interface / professional UX designer. & 2 \\
+Providing step-by-step processes, and showing the step numbers. & 1 \\
+Making the basic functions easy to use without reading the documentation. & 1 \\
 Focusing on limited number of functions. & 1 \\
@@
 Downsampling images to consume less memory. & 1 \\
-\begin{tabular}[c]{@{}l@{}}An option to load only part of the data to boost\\ performance.\end{tabular} & 1 \\ \hline
+An option to load only part of the data to boost performance. & 1 \\ \hline
 \end{tabular}
@@
 Threat to reproducibility & Num ans. \\ \hline
-\begin{tabular}[c]{@{}l@{}}If the software is closed-source, the reproducibility\\is hard to achieve.\end{tabular} & 1 \\
+If the software is closed-source, the reproducibility is hard to achieve. & 1 \\
 The project has no user interaction tests. & 1 \\
 The project has no unit tests. & 1 \\
-\begin{tabular}[c]{@{}l@{}}Using different versions of some common libraries\\may cause problems.\end{tabular} & 1 \\
+Using different versions of some common libraries may cause problems. & 1 \\
 CPU variability can leads to non-reproducibility. & 1 \\
-\begin{tabular}[c]{@{}l@{}}When reverse-engineering how manufacturers create\\medical images, the team may misinterpret it.\end{tabular} & 1 \\ \hline
+The team may misinterpret how manufacturers create medical images. & 1 \\ \hline
 \end{tabular}
@@
 Regression tests / unit tests / having good tests. & 6 \\
-\begin{tabular}[c]{@{}l@{}}Making code, data, and documentation available / making\\ the software open-source / using open-source libraries.\end{tabular} & 5 \\
+Making code, data, and documentation available / OSS / open-source libraries. & 5 \\
 Running same tests on all platforms. & 1 \\
-\begin{tabular}[c]{@{}l@{}}A dockerized version of the software,  insulating it from\\ the operating system environment.\end{tabular} & 1 \\
+A dockerized version of the software,  insulating it from the OS environment. & 1 \\
 Using standard libraries. & 1 \\
@@
 Clearly documenting the versions. & 1 \\
-\begin{tabular}[c]{@{}l@{}}Bringing along the exact versions of all the dependencies\\ with the software.\end{tabular} & 1 \\
+Bringing along the exact versions of all the dependencies with the software. & 1 \\
 Providing checksums of the data. & 1 \\
-\begin{tabular}[c]{@{}l@{}}Benchmark the software against other software with\\ similar purposes.\end{tabular} & 1 \\ \hline
+Benchmarking the software against other software with similar purposes. & 1 \\ \hline
 \end{tabular}

### 2021-08-18 StateOfPractice/Ao-Report/11Introduction.tex
MI report - updated ch 1 and more

@@
 \label{ch_intro}
-This report analyzes the state of the practice for Medical Imaging (MI) software.
+We define Scientific Computing (SC) as ``the use of computer tools to analyze or simulate mathematical models of real world systems of engineering or scientiﬁc importance so that we can better understand and predict the system’s behaviour" \cite{Smith2006}.
+Many researchers consider SC as the third pillar of science and engineering, along with theory and experiment \cite{Landau2005}.
+Almost all areas in science and engineering use computers for modeling \cite{Golub2014}, and software plays an essential role in modern scientific research \cite{Hannay2009} \cite{Wilson2014}.
+Software development in SC depends on three fields of knowledge: engineering or scientific domain knowledge, mathematical algorithm knowledge, and computational algorithm knowledge \cite{Landau2005} \cite{Mehta2015}.
+Thus, most SC software developers are scientists in SC domains \cite{Wilson2014}.
+However, they do not always use the modern software development techniques, tools, and methods \cite{Wilson2014}.
+Therefore, we developed a methodology for assessing the state of the practice for SC software.
+We apply this process to Medical Imaging (MI) software that belongs to a specific domain of SC.
+
+This report analyzes the state of the practice for MI software.
 MI is the clinical tool to image the interior of a body, providing information for diagnostic, analytic, and medical applications \cite{FDA2021} \cite{enwiki:1034887445}.
 MI is an essential part of collecting accurate information during clinical diagnosis \cite{Zhang2008}.
-MI computing and software aim to visualize and process medical images and produce clinically meaningful information \cite{enwiki:1034877594}.
-
-We define SC as ``the use of computer tools to analyze or simulate mathematical models of real world systems of engineering or scientiﬁc importance so that we can better understand and predict the system’s behaviour" \cite{Smith2006}.
-Software development in SC depends on the knowledge of three areas: engineering or science domain knowledge, mathematical algorithm knowledge, and computational algorithm knowledge.
-MI software belongs to a specific domain of SC.
+MI software aims to visualize and process medical images and produce clinically meaningful information \cite{enwiki:1034877594}.
 
@@
 
-Section \ref{sec_motivation} presents our motivation to start the research set the above goals, Section \ref{sec_research_questions} lists our research questions, and Section \ref{sec_scope} explains the domain analysis of MI software scope of our research.
+Section \ref{sec_motivation} presents our motivation to start the research set the above goals, Section \ref{sec_research_questions} lists our research questions, and Section \ref{sec_scope} presents the scope of MI software in our research.
 
@@
 \label{sec_motivation}
-Most scientists think developing and using SC software play significant roles in their research \cite{Hannay2009}.
+Most scientists think developing and using SC software plays a significant role in their research \cite{Hannay2009}.
 They spend a substantial proportion of their working hours on SC software development \cite{Hannay2009} \cite{Prabhu2011}, and this proportion of time has increased over the years \cite{Hannay2009}.
@@
 Developing SC software requires solid knowledge in specific domains \cite{Wilson2014}.
-Many of them learn software engineering skills by themselves or from their peers, instead of proper training \cite{Hannay2009}.
+Many developers learn software engineering skills by themselves or from their peers, instead of from proper training \cite{Hannay2009}.
 Hannay et al.
-\cite{Hannay2009} also pointed out that many scientists showed ignorance and indifference to standard software engineering concepts.
+\cite{Hannay2009} observe that many scientists showed ignorance and indifference to standard software engineering concepts.
 According to a survey by Prabhu et al.
@@
 
-Due to its nature, SC software born from one project can be part of many other projects in the future, with the potential to disproportionately causing damages to scientific researches \cite{Wilson2014}.
+Due to its nature, SC software born from one project can be part of many other projects in the future, with the potential to disproportionately causing damages to scientific research \cite{Wilson2014}.
 
@@
 We want to understand their status in SC domains and improve them.
+In addition, we want to understand whether problems like these mentioned above occur in all SC domains, or whether the state of the practice varies between domains.
 We build and refine our methodology, based on our previous work in scientific domains such as oceanography \cite{Smith2015}, mesh generation \cite{smith2016state}, geographic information systems \cite{smith2018state}, psychology \cite{smith2018statistical} and seismology \cite{Smith2018Seismology}.
@@
 \label{sec_research_questions}
-To achieve our objectives, we designed a few research questions and tried to answer them by our research methods. The questions are as follows,
+To achieve our objectives, we devised a few research questions as follows:
 
 \begin{description}
-\item[RQ1.] What are the pain points for developers working on MI software projects? What solutions the developers have tried to address the pain points?
-\item[RQ2.] What artifacts the projects generated?
-\item[RQ3.] What role does documentation play in the projects? What are the developers' attitudes toward it?
-\item[RQ4.] What principles, processes, methodologies, and tools the projects used?
+\item[RQ1.] What artifacts are present in current software projects? What role does documentation play in the projects? What are the developers' attitudes toward it?
+\item[RQ2.] What tools are used in the development of current software packages?
+\item[RQ3.] What principles, processes, and methodologies are used in the development of current software packages?
+\item[RQ4.] What are the pain points for developers working on research software projects?
+What aspects of the existing processes, methodologies, and tools do they consider as potentially needing improvement?
+What changes to processes, methodologies, and tools can improve software development and software quality?
 \item[RQ5.] What is the current status of the following software qualities for the projects? What actions have the developers taken to address them?
@@
 \cite{Angenent2006} pointed out that four fundamental problems are solved by MI software.
-While both mentioned Segmentation, Registration, and Visualization of medical images, Bankman also included Enhancement, Quantification, and a section covering some other functions \cite{Bankman2000}.
+While both mentioned Segmentation, Registration, and Visualization of medical images, Bankman also included Enhancement, Quantification, and three functions for MI archiving and telemedicine systems (Compression, Storage, and Communication) \cite{Bankman2000}.
 On the other hand, Angenent's team included Simulation \cite{Angenent2006}.
-According to Wikipedia contributors \cite{enwiki:1034877594}, MI software has primary functions in categories such as Segmentation, Registration, Visualization (including the basic display, reformatted views, and 3D volume rendering), Statistical Analysis, Image-based Physiological Modelling, etc.
+According to Wikipedia contributors \cite{enwiki:1034877594}, MI software has primary functions in categories Segmentation, Registration, Visualization (including the basic display, reformatted views, and 3D volume rendering), Statistical Analysis, and Image-based Physiological Modelling.
 As Kim et al.
@@
 
-We divided MI software into five sub-groups and several sub-sub-groups by their major functions shown in Figure \ref{fig_mi_functions}.
+Based on our literature survey, we divided MI software into five sub-groups and several sub-sub-groups by their major functions, as shown in Figure \ref{fig_mi_functions}.
 
-\begin{figure}
+\begin{figure}[h]
 \centering
@@
 
-In this project, the scope of the software is limited to the software library providing the Visualization tools and functions.
+To keep the data collection and analysis feasible, we limited the scope of the software to the software packages providing the Visualization tools and functions in this project.
+
+\section{Overview of the Methodology}
+We designed a general method to assess the state of the practice for SC software.
+With this method, we choose an SC domain and identify software candidates in it.
+Then, we filter the candidates and produce a final list of about 30 software packages.
+We measure the qualities of each software by answering questions on a grading template, as shown in Appendix \ref{ap_grading_template}.
+With the quantitative data generated by the template, we rank the software with the Analytic Hierarchy Process (AHP).
+After that, we interview some of the development teams to further understand the status of their development process.
+Finally, we summarize the results and propose recommendations for future SC software development.
+
+\section{Organization}
+We organize this report as follows:
+\begin{itemize}
+\item \textbf{Introduction} to our research and this report.
+\item \textbf{Background} of our research and methodology.
+\item \textbf{Methodology} of our state of the practice assessment, including an overview of applying it to the MI software.
+\item \textbf{Measurement Results} for a list of selected MI software, including measurement data generated by our grading template, and our ranking to the software on this list.
+\item \textbf{Interviews with Developers}, including the pain points and other status of their development process.
+\item \textbf{Answers to Research Questions.} 
+\item \textbf{Recommendations} to future SC software development.
+\item \textbf{Conclusions} to this report, including recommendations to the future state of the practice study.
+\item \textbf{Appendix}, including our Full Grading Template, Full Software List Before Filtering, Other Interview Answers, and Ethics Approval.
+\end{itemize}

### 2021-08-18 StateOfPractice/Ao-Report/12Background.tex
MI report - updated ch 1 and more
MI report - update ch 2

@@
 \label{ch_background}
-
-In this section, we introduce the relevant software categories (Section \ref{sec_software_categories}).
-We also cover the software quality definitions (Section \ref{sec_software_quality}) and an overview of the Analytic Hierarchy Process (AHP) (Section \ref{sec_AHP}).
 
@@
 
+In this section, we introduce the relevant software categories (Section \ref{sec_software_categories}).
+We also cover the software quality definitions (Section \ref{sec_software_quality}) and an overview of the AHP (Section \ref{sec_AHP}).
+
 \section{Software Categories}
@@
 We target specific software categories to narrow down the scope when selecting software packages for measuring.
-In this section, we discuss three common software categories: Open source software (OSS), freeware, and commercial software.
+One way to categorize them is grouping by functions, such as analyzing the scope for the MI software (Section \ref{sec_scope}), which requires specific domain knowledge.
+In this section, we discuss three common software categories that we can apply to all SC domains: Open source software (OSS), freeware, and commercial software.
 
@@
 
-	\item \textbf{Correctness \& Verifiability} A program is correct if it behaves according to its stated.
+	\item \textbf{Correctness \& Verifiability} A program is correct if it behaves according to its stated specifications.
 Verifiability is the extent to which a set of tests can be written and executed, to demonstrate that the delivered system meets the specification \cite{GhezziEtAl2003}.
@@
 For a project with $ m $ criteria, we can use an $m\times m$ matrix $A$ to record the relative importance between factors.
-By pairwise compare criterion $i$ and criterion $j$, the value of $A_{ij}$ is decided as follows, and the value of $A_{ji}$ is $1/A_{ij}$ \cite{Saaty1990},
+When comparing criterion $i$ and criterion $j$, the value of $A_{ij}$ is decided as follows, and the value of $A_{ji}$ is $1/A_{ij}$ \cite{Saaty1990},
 \begin{itemize}
@@
 The software are evaluated with the grading template in Appendix \ref{ap_grading_template} and a subjective score from one to ten is given for each quality for each package.
-To convert the results into pair-wise scores, we apply the following steps: for each quality, for a pair of packages $i$ and $j$, such that $score_i >= score_j$, the pairwise comparison result of $i$ versus $j$ is converted from $min((score_i - score_j) + 1, 9)$.
+For each quality, for a pair of packages $i$ and $j$, such that $score_i >= score_j$, the difference between two scores is $diff_{ij} = score_i - score_j$.
+The relationship between $A_{ij} = 1$ and $diff_{ij}$ is as follows:
+\begin{itemize}
+\item $A_{ij} = 1$ and $diff_{ij} = 0$ when criterion $i$ and criterion $j$ are equally important;
+\item $A_{ij}$ increases when $diff_{ij}$ increases;
+\item $A_{ij} = 9$ and $diff_{ij} = 9$ when criterion $i$ is extremely more important than criterion $j$.
+\end{itemize}
+Thus, we approximate the pairwise comparison result of $i$ versus $j$ by the following equation:
+\begin{equation}
+A_{ij} = min(score_i - score_j + 1, 9)
+\end{equation}

### 2021-08-18 StateOfPractice/Ao-Report/11Introduction.tex
MI report - updated ch 6 and more

@@
 \begin{description}
-\item[RQ1.] What artifacts are present in current software projects? What role does documentation play in the projects? What are the developers' attitudes toward it?
+\item[RQ1.] What artifacts are present in current software projects? What role does documentation play in the projects? What are the developers' attitude toward it?
 \item[RQ2.] What tools are used in the development of current software packages?
@@
 \end{itemize}
-\item[RQ6.] How does the software quality ranking generated by our methods compare with the ratings from the community?
+\item[RQ6.] How does software designated as high quality by this methodology compare with top-rated software by the community?
 \end{description}

### 2021-08-18 StateOfPractice/Ao-Report/13Methods.tex
MI report - updated ch 6 and more

@@
 \subsection{Software Product Selection}
+\label{sec_mi_software_selection}
 By using the method in Section \ref{sec_identify_software_candidates}, we identified 48 MI software projects as the candidates from publications \cite{Bjorn2017} \cite{Bruhschwein2019} \cite{Haak2015}, online articles related to the domain \cite{Emms2019} \cite{Hasan2020} \cite{Mu2019}, forum discussions related to the domain \cite{Samala2014}, etc.

### 2021-08-18 StateOfPractice/Ao-Report/16AnswersToResearchQuestions.tex
MI report - minor updates
MI report - updated ch 6 and more

@@
 The answers are based on our quality measurements and developer interviews.
-Sections \ref{sec_rq1_pain_points}, \ref{sec_rq2_artifacts}, \ref{sec_rq3_documentation}, \ref{sec_rq4_PPMT}, \ref{sec_rq5_qualities}, and \ref{sec_rq6_comparison} summarize the answers to the six questions respectively.
+Sections \ref{sec_rq_artifacts}, \ref{sec_rq_tools}, \ref{sec_rq_PPM}, \ref{sec_rq_pain_points}, \ref{sec_rq_qualities}, and \ref{sec_rq_comparison} summarize the answers to the six questions respectively.
 Section \ref{sec_threats_to_validity} lists the threats to the validity of our research.
 
-\section{Pain Points and Solutions}
-\label{sec_rq1_pain_points}
-\begin{description}\item[RQ1.] What are the pain points for developers working on MI software projects? What solutions have the developers tried to address the pain points?\end{description}
+\section{Artifacts in the Projects}
+\label{sec_rq_artifacts}
+\begin{description}
+\item[RQ1.] What artifacts are present in current software projects?
+\end{description}
 
-We answer this question primarily by interviewing the developers. As described in Section \ref{sec_interview_pain_points}, the top three pain points are as follows,
-\begin{itemize}
-\item the lack of fundings and time;
-\item the difficulty to balance between four factors: cross-platform compatibility, convenience to development \& maintenance, performance, and security;
-\item the lack of access to real-world datasets for testing.
-\end{itemize}
-
-Section \ref{sec_interview_pain_points} also lists potential and proven solutions from the developers.
-
-\section{Artifacts in the Projects}
-\label{sec_rq2_artifacts}
-\begin{description}\item[RQ2.] What artifacts the projects generated?\end{description}
-
-We answer this question by searching and scanning the source code and documentation of the projects.
-The complete list of artifacts is in the repository \hyperlink{https://github.com/smiths/AIMSS}{https://github.com/smiths/AIMSS}.
+We answer this question by examining the documentation, scanning the source code, and interviewing the developers of the projects.
 
@@
 
-\section{Documentation in the Projects}
-\label{sec_rq3_documentation}
-\begin{description}\item[RQ3.] What role does documentation play in the projects? What are the developers' attitudes toward it?\end{description}
+\section{Tools in the Projects}
+\label{sec_rq_tools}
+\begin{description}
+\item[RQ2.] What tools are used in the development of current software packages?
+\end{description}
 
-We answer this question by interviewing the developers.
-As shown in Section \ref{sec_interview_documents}, most of the nine interviewees thought that documentation was essential to their projects.
+We answer this question by measuring the qualities and interviewing the developers.
+This section summarizes the tools used for CI/CD, user support, version control, documentation, contribution management, and project management.
+
+\subsection{CI/CD Tools}
+As mentioned in Section \ref{sec_result_correctness_verifiability}, we identified five projects using CI/CD tools.
+\textit{3D Slicer} and \textit{OHIF Viewer} used CircleCI; \textit{ImageJ}, \textit{Fiji}, and \textit{dwv} used Travis.
+According to the interviews with developers, \textit{dwv} and \textit{Weasis} used GitHub Actions.
+
+\subsection{User Support Tools}
+Table \ref{tab_user_support_model} shows the user support tools by the number of projects using them.
+
+\subsection{Version Control Tools}
+Briefly speaking, 27 of the 29 projects chose git as the version control tool, and 24 used GitHub to host the repositories. Section \ref{sec_score_maintainability} lists more details.
+
+\subsection{Documentation Tools}
+Table \ref{tab_doc_tools} lists the documentation tools by the number of answers from the interviewees.
+
+\subsection{Contribution Management Tools}
+Table \ref{tab_method_receive_contributions} shows the ways of receiving contributions by the number of answers from the interviewees.
+
+\subsection{Project Management Tools}
+Table \ref{tab_pm_tools} shows the tools for project management by the number of answers from the interviewees.
+
+\section{Principles, Processes, and Methodologies in the Projects}
+\label{sec_rq_PPM}
+\begin{description}
+\item[RQ3.] What principles, processes, and methodologies are used in the development of current software packages?
+\end{description}
+
+We answer this question by measuring the qualities and interviewing the developers.
+This section shows the principles, processes, and methodologies in software testing, documentation, contribution management, project management, and improving software qualities.
+
+\subsection{Software Testing}
+As shown in Section \ref{fg_correctness_erifiability_scores}, we only identified the use of unit testing in less than half of the 29 projects.
+Most projects did not have requirements specifications and theory manuals.
+
+On the other hand, the interviewees believed that testing (including usability tests with users) was the top solution to improve \textit{correctness}, \textit{usability}, and \textit{reproducibility}, as shown in Section \ref{sec_interview_correctness}, \ref{sec_interview_usability}, and \ref{sec_interview_reproducibility}.
+
+Table \ref{tab_q15_threats_correctness} lists a threat to \textit{correctness}: with huge datasets for testing, the tests are expensive and time-consuming.
+Table \ref{tab_q15_strategies_correctness} shows that three interviewees endorsed self tests / automated tests, which may save the time for testing.
+
+Section \ref{sec_pain_points_3} introduces a pain point related to testing, and developers' strategies to address it.
+
+\subsection{Documentation}
+Section \ref{fg_correctness_erifiability_scores} shows that most projects did not have requirements specifications and theory manuals.
+Table \ref{tab_maintainability_docs} presents which projects had a project plan, developer’s manual, or API documentation, which implies better \textit{maintainability}.
+Table \ref{tab_Visibility/Transparency_docs} lists the existing documents that can improve \textit{visibility/transparency} of the projects.
+Table \ref{tab_artifacts_frequency} reveals the number of all the artifacts in the 29 projects.
+
+Most of the nine interviewees thought that documentation was essential to their projects.
 However, they hold the common opinion that their documentation needed improvements.
 Nearly half of them also believed that the lack of time prevented them from improving the documentation.
+See more details in Section \ref{sec_interview_documents}.
 
-\section{Principles, Processes, Methodologies, and Tools in the Projects}
-\label{sec_rq4_PPMT}
-\begin{description}\item[RQ4.] What principles, processes, methodologies, and tools the projects used?\end{description}
+\subsection{Contribution Management}
+According to the interviews, all the eight teams were using GitHub and pull requests to accept new contributions from the community.
+Some of them had relied on emails and forums for the purpose.
+More details are in Section \ref{sec_contribution_pm}.
 
-We answer this question by measuring the qualities and interviewing the developers.
+\subsection{Project Management}
+As presented in Section \ref{sec_contribution_pm}, most teams in the interviews did not have a clearly defined software development model.
+Many of them relied on GitHub, issue/bug trackers to manage their projects.
 
-Table \ref{tab_maintainability_docs}, \ref{tab_Visibility/Transparency_docs}, and \ref{tab_artifacts_frequency} reveal part of the answer, such as which and how many teams used a project plan, API documentation, release notes, version control, etc.
+\section{Pain Points and Solutions}
+\label{sec_rq_pain_points}
+\begin{description}
+\item[RQ4.] What are the pain points for developers working on research software projects?
+What aspects of the existing processes, methodologies, and tools do they consider as potentially needing improvement?
+What changes to processes, methodologies, and tools can improve software development and software quality?
+\end{description}
 
-We summarize the other part of the answer from the interviews.
-Sections \ref{sec_interview_documents} and \ref{sec_contribution_pm} list some principles, processes, methodologies, and tools for documentation, contribution management, and project management.
-As shown in Section \ref{sec_interview_software_qualities}, the developers also shared some of their principles, processes, methodologies, and tools to improve \textit{correctness}, \textit{maintainability}, \textit{understandability}, \textit{usability}, and \textit{reproducibility}.
+We answer this question primarily by interviewing the developers. This section presents the pain points and the solutions from the developers and our team.
+
+\subsection{Pain Points}
+As described in Section \ref{sec_interview_pain_points}, the top three pain points are as follows,
+\begin{itemize}
+\item the lack of fundings and time;
+\item the difficulty to balance between four factors: cross-platform compatibility, convenience to development \& maintenance, performance, and security;
+\item the lack of access to real-world datasets for testing.
+\end{itemize}
+
+\subsection{Solutions}
+Section \ref{sec_interview_pain_points} lists the details about the potential and proven solutions from the developers in the MI domain.
+We also summarize our general recommendations on software development in all SC domains in Section \ref{ch_recommendations}.
 
 \section{Software Qualities}
-\label{sec_rq5_qualities}
+\label{sec_rq_qualities}
 \begin{description}\item[RQ5.] What is the current status of the following software qualities for the projects? What actions have the developers taken to address them?\end{description}
 
-We answer this question by measuring the qualities and interviewing the developers.
+This section includes our answers from the qualities measurements and interviews with the developers.
 
+\subsection{Results by Measuring Software Qualities}
 Section \ref{ch_results} contains our measurement results for \textit{Installability}, \textit{Correctness \& Verifiability}, \textit{Reliability}, \textit{Robustness}, \textit{Usability}, \textit{Maintainability}, \textit{Reusability}, \textit{Understandability}, and \textit{Visibility/Transparency}.
@@
 
+\subsection{Results by Interviewing Developers}
 Section \ref{sec_interview_software_qualities} shows findings from the interviews for \textit{correctness}, \textit{maintainability}, \textit{understandability}, \textit{usability}, and \textit{reproducibility}.
 The interviewees expressed their thoughts on these five qualities.
-We discussed the current or past threats to them, and found out what actions they believed would improve the qualities.
+We discussed the current or past threats, and found out what actions they believed would improve the qualities.
 
 \section{Our Ranking versus the Community Ratings}
-\label{sec_rq6_comparison}
-\begin{description}\item[RQ6.] How does the software quality ranking generated by our methods compare with the ratings from the community?\end{description}
+\label{sec_rq_comparison}
+\begin{description}
+\item[RQ6.] How does software designated as high quality by this methodology compare with top-rated software by the community?
+\end{description}
 
-We answer this question by grading the qualities of the software, then comparing the ranking with the community ratings on GitHub, such as GitHub stars, number of forks, and number of people watching the projects.
+We answer this question by grading the qualities of the software, collecting ratings from the MI community, then conducting two comparisons:
+\begin{itemize}
+\item comparing our ranking with the community ratings on GitHub, such as GitHub stars, number of forks, and number of people watching the projects (Section \ref{sec_ranking_vs_github});
+\item comparing top-rated software designated by our methodology with the ones recommended by our domain experts (Section \ref{sec_designation_vs_experts}).
+\end{itemize}
+
+\subsection{Our Ranking versus the GitHub Popularity}
+\label{sec_ranking_vs_github}
 
@@
 
+\subsection{Designated Top Software versus the Domain Experts' Recommendation}
+\label{sec_designation_vs_experts}
+
+As shown in Section \ref{sec_mi_software_selection}, our domain experts recommended a list of top software with 12 software products.
+Table \ref{tab_top_software_vs_experts} compares the top 12 software projects ranked by our methodology with the ones from the domain experts.
+
+An example of interpreting Table \ref{tab_top_software_vs_experts} is as follows: in row 2, \textit{ImageJ} has rank 2 by our assessment, and the number \textit{(3)} means that it has rank 3 by the experts; likewise, \textit{Horos} has rank 2 by domain expers, and rank 12 by our assessment.
+
+All of the top 4 software from the domain experts are among the top 12 ones ranked by our methodology.
+3 of the top 4 on both lists are the same ones: \textit{3D Slicer}, \textit{ImageJ}, and \textit{Fiji}.
+\textit{3D Slicer} is also the top 1 of both rankings.
+
+As mentioned in Section \ref{sec_mi_software_selection}, six of the recommended software packages did not have Visualization as the primary function, so we did not include them on our final list.
+
+\begin{table}[H]
+\centering
+\begin{tabular}{lll}
+\hline
+Rank & This assessment & Domain expert \\ \hline
+1 & 3D Slicer (1) & 3D Slicer (1) \\
+2 & ImageJ (3) & Horos (12) \\
+3 & OHIF Viewer (n/a) & ImageJ (2) \\
+4 & Fiji (4) & Fiji (4) \\
+5 & ParaView (n/a) & AFNI (n/a) \\
+6 & Weasis (n/a) & FSL (n/a) \\
+7 & medInria (n/a) & Freesurfer (n/a) \\
+8 & BioImage Suite Web (n/a) & Mricron (18*) \\
+9 & OsiriX Lite (n/a) & Mango (17**) \\
+10 & INVESALIUS 3 (n/a) & Tarquin (n/a) \\
+11 & Gwyddion (n/a) & Diffusion Toolkit (n/a) \\
+12 & Horos (2) & MRItrix (n/a) \\ \hline
+\end{tabular}
+\caption{\label{tab_top_software_vs_experts}Top software versus domain experts' recommendation}
+\end{table}
+
+* \textit{MRIcron} development had moved to \textit{MRIcroGL}, as mentioned in Section \ref{sec_mi_software_selection}. Thus, we measured and ranked \textit{MRIcroGL} at 18.
+
+** We included \textit{Mango} in the initial list, but removed it because it was not OSS. \textit{Papaya} is a the web version OSS of \textit{Mango}. We measured and ranked \textit{Papaya} at 17.
+
 \section{Threats to Validity}
@@
 Since we did not have the chance to explain our questions or ask him follow-up questions, there is a possibility of misinterpretation of the questions or answers.
-\item As mentioned in Section \ref{sec_rq6_comparison}, we gave \textit{dwv} much lower scores than its GitHub popularity. We might underestimate its rank due to uncommon technical issues.
+\item As mentioned in Section \ref{sec_rq_comparison}, we gave \textit{dwv} much lower scores than its GitHub popularity. We might underestimate its rank due to uncommon technical issues.
 \end{enumerate}

### 2021-08-19 StateOfPractice/Ao-Report/11Introduction.tex
MI report - update ch 3 and more

@@
 Besides the above major functions, some MI software provides supportive functions.
-For example, with Tool Kit (TK) libraries VTK \cite{SchroederEtAl2006} and ITK \cite{McCormick2014}, developers build software with Visualization and Analysis functions; Picture Archiving and Communication System (PACS) helps users to economically store and conveniently access images \cite{Choplin1992}.
+For example, with Tool Kit libraries VTK \cite{SchroederEtAl2006} and ITK \cite{McCormick2014}, developers build software with Visualization and Analysis functions; Picture Archiving and Communication System (PACS) helps users to economically store and conveniently access images \cite{Choplin1992}.
 
@@
 child[concept color=teal!40] { node {Supporting}
-    child[concept color=blue!30] { node {TK}}
+    child[concept color=blue!30] { node {Tool Kit}}
     child[concept color=blue!30] { node {PACS}}

### 2021-08-19 StateOfPractice/Ao-Report/12Background.tex
MI report - update ch 3 and more

@@
 
-	\item \textbf{Reliability} The probability of failure-free operation of a computer program in a specified environment for a specified time, i.e.
-the average time interval between two failures also known as the mean time to failure (MTTF) \cite{musa1987software}
+	\item \textbf{Reliability} The probability of failure-free operation of a computer program in a specified environment for a specified time, i.e., the average time interval between two failures also known as the mean time to failure (MTTF) \cite{musa1987software}
 	\cite{GhezziEtAl2003}.

### 2021-08-19 StateOfPractice/Ao-Report/13Methods.tex
MI report - update ch 3 and more

@@
 
-We designed a general process for evaluating the state of the practice of domain-specific software, that we instantiate to SC software for a specific scientific domain.
+We designed a general process for evaluating the state of the practice of domain-specific SC software, that we instantiate for a specific scientific domain.
 
-Our method involves several steps, shown as follows:
+Our method involves four steps:
 \begin{enumerate}
@@
 
-Section \ref{sec_applying_method} presents an example of how we applied the method on the MI domain.
+\noindent Section \ref{sec_applying_method} presents an example of how we applied the method on the MI domain.
 
@@
 \label{sec_domain_selection}
-Our methods are generic, but we have only applied them to scientific domains due to the objective of our research.
-
+Our methods are generic for any SC software, but they need to be applied to a specific domain.
 When choosing a candidate domain, we prefer one with a large number of active OSS.
@@
 For example, we remove the ones without recent updates or specific functions.
-Thus, we need enough software candidates from the beginning.
 Besides, we prefer OSS projects because our grading method requires access to the code.
@@
 
-The process of selecting software packages contains two steps: i) identify software candidates in the chosen domain, ii) filter the list according to needs \cite{SmithEtAl2021}.
+The process of selecting software packages contains two steps: i) identify software candidates in the chosen domain, ii) filter the list according to our needs \cite{SmithEtAl2021}.
 
@@
 
-The other filters are optional, and we consider them according to the number of software candidates and the objectives of the research project. We try to apply them in the following priority order:
+The other filters are optional, and we consider them according to the number of software candidates and the objectives of the research project. We apply them in the following priority order:
 
@@
 \item The functions and purpose of the software.
-For example, we can choose a group of software with similar functions, so that the comparison is between the software; or we can cross-compare sub-groups in the domain, then we need to select candidates from each sub-group.
+An SC domain often contains software with various functions and purposes.
+For example, some MI software packages are Tool Kit for developers to use, and some others offer Visualization function to end-users.
+We have two options:
+\begin{itemize}
+\item selecting a set of software with the same major function.
+In this case, we can use the identical process to assess all packages, e.g., same input to measure \textit{Robustness}.
+Also, when we give impression scores to qualities such as \textit{Installability} and \textit{Usability}, the results are more comparable.
+Thus, it is more feasible to collect the results and rank the software.
+\item selecting software from a set of different sub-groups.
+For example, we can choose 10 MI software from each of the three sub-groups: Visualization, Tool Kit, and PACS.
+The downside: we may need different processes to measure each sub-group; it may be less accurate to mix all three sub-groups and rank them together.
+The benefit: we can cross-compare the development processes between the sub-groups.
+\end{itemize}
 
@@
 
-\item The age of software.
+\item The age of the software.
 Some of the OSS projects may experience a lack of recent maintenance.
@@
 \label{sec_vet_software_list}
-Before showing our filtered list to the domain experts, we ask them to list the top 10 software in the domain.
-Then, we cross-compare the two lists and discuss the commons and differences.
+Before showing our filtered list to the domain experts, we ask them to list their top 10 software in the domain.
+Then, we cross-compare the two lists and discuss the commonality and variability.
 In addition, we ask them to vet our filtered list.
-They provide opinions on whether the list is reasonable to them.
-We also value their opinions on the filtering process.
-For example, if a software package is not OSS and has no updates for a long while, but the domain experts identify it as a valuable product, we still consider it to keep.
+They provide views on whether the list is reasonable.
+We also use their opinions for a filtering process.
+For example, if a software package is not OSS and has had no updates for a long time, but the domain experts identify it as a valuable product, we still consider it in our final list.
 
@@
 
-We use the first section of the template to collect general information, such as the name, purpose, platform, programming language, publications about the software, the first release and the most recent change date, website, and source code repository of the product, etc.
+We use the first section of the template to collect general information, such as the name, purpose, platform, programming language, publications about the software, the first release and the most recent change date, website, source code repository of the product, etc.
 Information in this section helps us understand the projects better and may be helpful for further analysis, but it does not directly affect the grading scores.
@@
 Each quality needs an overall score between 1 and 10 based on all the previous questions.
-For some qualities, we perform surface measurements, which allow us to carry out on all packages with reasonable efforts.
+For some qualities, we perform surface measurements, which allow us to measure all packages with reasonable efforts.
 The surface measurements reveal some traits of a underlying quality, but may not fully represent it.
@@
 If any problem interrupts the process of installation or uninstallation, we give a lower score to this quality.
-We also record the operating system (OS) for the installation test and whether we can verify the installation.
+We also record the Operating System (OS) for the installation test and whether we can verify the installation.
 
@@
 
-\item \textbf{Surface Reliability} We check that whether the software break during the installations and tutorials, whether there are descriptive error messages, and if we can recover the process after the errors.
+\item \textbf{Surface Reliability} We check whether the software breaks during the installations and tutorials, whether there are descriptive error messages, and if we can recover the process after an error.
 
-\item \textbf{Surface Robustness} We check that how the software handle unexpected/unanticipated input.
-For example, for software packages with the function to load image files, we prepare broken image files for them to open.
+\item \textbf{Surface Robustness} We check how the software handles unexpected/unanticipated input.
+For example, we prepare broken image files for software packages with the function to load image files.
 We use a text file (.txt) with a modified extension name (.dcm) as an unexpected/unanticipated input.
@@
 
-\item \textbf{Surface Usability} We examine the documents of the projects and consider software with a getting started tutorial and a user manual easier to use.
-Meanwhile, we check if users have any channels to get supports.
-Our impressions and user experiences when testing the software also affect the scores.
-For example, easy-to-use graphical user interfaces give us a better experience, thus lead to better scores.
+\item \textbf{Surface Usability} We examine the documentation of the projects, and we consider software with a getting started tutorial and a user manual easier to use.
+Meanwhile, we check if users have any channels to get support.
+We also record our impressions and user experiences when testing the software.
+Easy-to-use graphical user interfaces give us a better experience, which leads to better scores.
 
@@
 
-\item \textbf{reusability} We count the total number of code files for each project.
+\item \textbf{Reusability} We count the total number of code files for each project.
 Projects with a large number of components provide more choices to reuse.
-Furthermore, well-modularized code, which tends to have smaller parts in separated files, is easier to reuse.
+Furthermore, well-modularized code, which tends to have smaller parts in separated files, is typically easier to reuse.
 Thus, we consider the projects with more code files to be more reusable.
@@
 
-\item \textbf{Visibility/Transparency} To measure this quality, such as all of the steps of a software development process and the current status of a project, we check the existing documents.
+\item \textbf{Visibility/Transparency} To measure this quality, we check the existing documents to find out whether the software development process and current status of a project are visible and transparent.
 We examine the development process, current status, development environment, and release notes for each project.
-If any information is missing or poorly conveyed, the \textit{visibility/transparency} is not ideal.
+If any information is missing or poorly conveyed, the \textit{visibility/transparency} will be lower.
 \end{itemize}
 
-All the last three sections are about the empirical measurements.
 For some qualities, the empirical measurements also affect the score.
@@
 We use two command-line tools for the empirical measurements.
-One is \textit{GitStats} that generates statistics for git repositories and display outputs in the format of web pages \cite{Gieniusz2019}; the other one is Sloc Cloc and Code (as known as \textit{scc}) \cite{Boyter2021}, aiming to count the lines of code, comments, etc.
+One is \textit{GitStats} that generates statistics for git repositories and displays outputs in the form of web pages \cite{Gieniusz2019}; the other one is Sloc Cloc and Code (as known as \textit{scc}) \cite{Boyter2021}, aiming to count the lines of code, comments, etc.
 
@@
 
-Additionally, we manually collect information for projects held on GitHub, such as the numbers of stars, forks, people watching this repository, open pull requests, closed pull requests, number of months a repository has been on GitHub.
+Additionally, we manually collect information for projects held on GitHub, such as the numbers of stars, forks, people watching this repository, open pull requests, closed pull requests, and the number of months a repository has been on GitHub.
 A git repository can have a creation date much earlier than the first day on GitHub.
@@
 To test the software on a ``clean'' system, we create a new virtual machine (VM) for each software and only install the necessary dependencies before measuring.
-We make all 30 VM on the same computer, one at a time, and destroy them after measuring.
+We make all 30 VMs on the same computer, one at a time, and destroy them after measuring.
 
@@
 
-The interviews are semi-structured based on the question list, and we also ask follow-up questions when necessary.
-Based on our experience, the interviewees usually bring up some exciting ideas that we do not expect, and it is worth expanding on these topics.
+The interviews are semi-structured based on the question list; we ask follow-up questions when necessary.
+Based on our experience, the interviewees usually bring up some exciting ideas that we did not expect, and it is worth expanding on these topics.
 
@@
 
-Ideally, we select projects after the grading measurements and prefer the ones with higher overall scores. However, if we do not find enough participants, we contact all teams on the list.
+Ideally, we select projects after the grading measurements and prefer the ones with higher overall scores.
+However, if we do not find enough participants, we reach all teams on the list.
+As mentioned in Section  \ref{sec_apply_to_mi_interviews}, when we applied this process to the MI domain, we eventually contacted all teams.
 
@@
 We also included Dr.
-Michael Noseworthy and some of his students as the MI domain experts in our team.
+Michael Noseworthy, a professor of Electrical and Computer Engineering at McMaster University, Co-Director of the McMaster School of Biomedical Engineering, and Director of Medical Imaging Physics and Engineering at St.
+Joseph’s Healthcare, and some of his students as the MI domain experts in our team.
 
@@
 By using the method in Section \ref{sec_identify_software_candidates}, we identified 48 MI software projects as the candidates from publications \cite{Bjorn2017} \cite{Bruhschwein2019} \cite{Haak2015}, online articles related to the domain \cite{Emms2019} \cite{Hasan2020} \cite{Mu2019}, forum discussions related to the domain \cite{Samala2014}, etc.
-Appendix \ref{ap_list_before_filtering} shows all the 48 software packages.
+Appendix \ref{ap_list_before_filtering} shows all 48 software packages.
 
@@
 \begin{enumerate}
-\item Among them, there were eight ones that we could not find their source code, such as \textit{MicroDicom}, \textit{Aliza}, and \textit{jivex}.
+\item Among them, there were eight that we could not find their source code, such as \textit{MicroDicom}, \textit{Aliza}, and \textit{jivex}.
 These packages are likely to be freeware defined in Section \ref{sec_freeware} and not OSS.
@@
 
-\item Next, we focused on the MI software providing visualization functions, as described in Section \ref{sec_scope}.
-Seven of the software on the list were tool kits or libraries for other software to use as dependencies, but not for end-users to view medical images, such as \textit{VTK}, \textit{ITK}, and \textit{dcm4che}; another three were PACS.
+\item Next, we focused on the MI software providing Visualization functions, as described in Section \ref{sec_scope}.
+Seven of the software on the list were Tool Kits or libraries for other software to use as dependencies, but not for end-users to view medical images, such as \textit{VTK}, \textit{ITK}, and \textit{dcm4che}; another three were PACS.
 We also eliminated these from the list.
@@
 We compared two lists and found six common ones.
-We did not include the rest in our final list, because their major function was not Visualization.
 
@@
 
+Six software packages on the domain experts' list were not on our filtered list.
+We believed their primary function was Analysis mentioned in Section \ref{sec_scope}.
+Thus, we did not include them in our final list.
+
 After vetting our filtered list, the domain experts believed it was reasonable and did not identify any problem.
-Thus, as shown in Appendix \ref{ap_list_before_filtering}, eventually we had 29 software products on the final list.
+Thus, as shown in Appendix \ref{ap_list_before_filtering}, eventually, we had 29 software products on the final list.
 
@@
 \subsection{Interviews}
+\label{sec_apply_to_mi_interviews}
 We received ethics clearance from the McMaster University Research Ethics Board (Appendix \ref{ap_ethics}).

### 2021-08-20 StateOfPractice/Ao-Report/13Methods.tex
MI report - updated ch 4 and more

@@
 
-\item \textbf{Surface Reliability} We check whether the software breaks during the installations and tutorials, whether there are descriptive error messages, and if we can recover the process after an error.
+\item \textbf{Surface Reliability} We check whether the software breaks during the installation and the operations following a getting started tutorials, whether there are descriptive error messages, and if we can recover the process after an error.
+We use the software to load damaged images during the assessment to \textit{surface robustness}, and lower its score for \textit{surface reliability} if a software product ``break" in this process.
 
 \item \textbf{Surface Robustness} We check how the software handles unexpected/unanticipated input.
-For example, we prepare broken image files for software packages with the function to load image files.
+For example, we prepare broken image files for MI software packages with the function to load image files.
 We use a text file (.txt) with a modified extension name (.dcm) as an unexpected/unanticipated input.
@@
 Furthermore, well-modularized code, which tends to have smaller parts in separated files, is typically easier to reuse.
-Thus, we consider the projects with more code files to be more reusable.
-We use \textit{GitStats} as a tool to count the number of text-based files for all projects, and consider the projects with more text-based files to also have more code files.
+Thus, we consider the projects with more code files and less code lines per file to be more reusable.
+We use a command-line tool \textit{scc} to count the number of  text-based files and LOC for all projects.
 We also decide that the projects with API documentation can deliver better \textit{Reusability}.

### 2021-08-20 StateOfPractice/Ao-Report/14MeasurementResults.tex
MI report - updated ch 6 and more
MI report - updated ch 4 and more

@@
 
-Table \ref{tab_final_list} shows the 29 software packages that we measured.
-We found the initial release dates (Rlsd) in their documents for most projects and marked the rest two with ``?".
+Table \ref{tab_final_list} shows the 29 software packages that we measured, along with some summary data collected in the year 2020.
+As mentioned in Section \ref{sec_grading_template}, we used \textit{scc} (Section \ref{sec_empirical_measurements}) to count the Lines of Code (LOC), excluding the comment and blank lines.
+We arrange the items in the descending order of the LOC.
+We found the initial release dates (Rlsd) in their documents for most projects and marked the two unknown dates with ``?".
 We used the date of the latest change to each code repository to decide the latest update.
@@
 
-We counted the number of contributors (NOC) and lines of code (LOC).
-we considered anyone who made at least one accepted commit to the source code as a contributor.
-Thus, it does not mean that any development team has are 100 long-term members.
+We counted the number of contributors (NOC).
+We considered anyone who made at least one accepted commit to the source code as a contributor.
+Thus, the NOC is not usually the same as the number of long-term members.
 Many of these projects received change requests and code from the community, such as pull requests and git commits on GitHub.
 
-Table \ref{tab_final_list} also shows the supported OS for each software package, and 25 of them could work on all three Windows (W), macOS (M), and Linux (L) systems.
+Table \ref{tab_final_list} also shows the supported OS for each software package.
+Twenty-five of them could work on all three OSs: Windows (W), macOS (M), and Linux (L).
 However, there was a significant difference in the philosophy to achieve cross-platform compatibility.
-Most of them were native software products, but five were naturally platform-independent web applications.
+Most of them were native software products, but five were naturally platform-independent web applications, as shown in column ``Web".
 
 \begin{table}[H]
+\centering
 \begin{tabular}{llllllllll}
 \hline
-\multirow{2}{*}{Software} & \multirow{2}{*}{Rlsd} & \multirow{2}{*}{Updated} & \multirow{2}{*}{Fnd} & \multirow{2}{*}{NOC} & \multirow{2}{*}{LOC} & \multicolumn{3}{c}{OS} & \multirow{2}{*}{W} \\ \cline{7-9}
+\multirow{2}{*}{Software} & \multirow{2}{*}{Rlsd} & \multirow{2}{*}{Updated} & \multirow{2}{*}{Fnd} & \multirow{2}{*}{NOC} & \multirow{2}{*}{LOC} & \multicolumn{3}{c}{OS} & \multirow{2}{*}{Web} \\ \cline{7-9}
  &  &  &  &  &  & W & M & L &  \\ \hline
-3D Slicer \cite{Kikinis2014} & 1998 & 2020-08 & X & 100 & 501451 & X & X & X &  \\
-Ginkgo CADx \cite{Wollny2020} & 2010 & 2019-05 &  & 3 & 257144 & X & X & X &  \\
-XMedCon \cite{Nolf2003} & 2000 & 2020-08 &  & 2 & 96767 & X & X & X &  \\
-Weasis \cite{Roduit2021} & 2010 & 2020-08 &  & 8 & 123272 & X & X & X &  \\
-MRIcroGL \cite{Rorden2021} & 2015 & 2020-08 &  & 2 & 8493 & X & X & X &  \\
-SMILI \cite{Chandra2018} & 2014 & 2020-06 &  & 9 & 62626 & X & X & X &  \\
-ImageJ \cite{Rueden2017} & 1997 & 2020-08 & X & 18 & 9681 & X & X & X &  \\
-Fiji \cite{Schindelin2012} & 2011 & 2020-08 & X & 55 & 10833 & X & X & X &  \\
-DicomBrowser \cite{Archie2012} & 2012 & 2020-08 &  & 3 & 5505 & X & X & X &  \\
-3DimViewer \cite{TESCAN2020} & ? & 2020-03 & X & 3 & 178065 & X & X &  &  \\
+ParaView \cite{Ahrens2005} & 2002 & 2020-10 & X & 100 & 886326 & X & X & X & X \\
+Gwyddion \cite{Nevcas2012} & 2004 & 2020-11 &  & 38 & 643427 & X & X & X &  \\
 Horos \cite{horosproject2020} & ? & 2020-04 &  & 21 & 561617 &  & X &  &  \\
 OsiriX Lite \cite{PixmeoSARL2019} & 2004 & 2019-11 &  & 9 & 544304 &  & X &  &  \\
+3D Slicer \cite{Kikinis2014} & 1998 & 2020-08 & X & 100 & 501451 & X & X & X &  \\
+Drishti \cite{Limaye2012} & 2012 & 2020-08 &  & 1 & 268168 & X & X & X &  \\
+Ginkgo CADx \cite{Wollny2020} & 2010 & 2019-05 &  & 3 & 257144 & X & X & X &  \\
+GATE \cite{Jan2004} & 2011 & 2020-10 &  & 45 & 207122 &  & X & X &  \\
+3DimViewer \cite{TESCAN2020} & ? & 2020-03 & X & 3 & 178065 & X & X &  &  \\
+medInria \cite{Fillard2012} & 2009 & 2020-11 &  & 21 & 148924 & X & X & X &  \\
+BioImage Suite Web \cite{Papademetris2005} & 2018 & 2020-10 & X & 13 & 139699 & X & X & X & X \\
+Weasis \cite{Roduit2021} & 2010 & 2020-08 &  & 8 & 123272 & X & X & X &  \\
+AMIDE \cite{Loening2017} & 2006 & 2017-01 &  & 4 & 102827 & X & X & X &  \\
+XMedCon \cite{Nolf2003} & 2000 & 2020-08 &  & 2 & 96767 & X & X & X &  \\
+ITK-SNAP \cite{Yushkevich2006} & 2006 & 2020-06 & X & 13 & 88530 & X & X & X &  \\
+Papaya \cite{UTHSCSA2019} & 2012 & 2019-05 &  & 9 & 71831 & X & X & X &  \\
+OHIF Viewer \cite{Ziegler2020} & 2015 & 2020-10 &  & 76 & 63951 & X & X & X & X \\
+SMILI \cite{Chandra2018} & 2014 & 2020-06 &  & 9 & 62626 & X & X & X &  \\
+INVESALIUS 3 \cite{Amorim2015} & 2009 & 2020-09 &  & 10 & 48605 & X & X & X &  \\
 dwv \cite{Martelli2021} & 2012 & 2020-09 &  & 22 & 47815 & X & X & X & X \\
-Drishti \cite{Limaye2012} & 2012 & 2020-08 &  & 1 & 268168 & X & X & X &  \\
-BioImage Suite Web \cite{Papademetris2005} & 2018 & 2020-10 & X & 13 & 139699 & X & X & X & X \\
-OHIF Viewer \cite{Ziegler2020} & 2015 & 2020-10 &  & 76 & 63951 & X & X & X & X \\
+DICOM Viewer \cite{Afsar2021} & 2018 & 2020-04 & X & 5 & 30761 & X & X & X &  \\
+MicroView \cite{ParallaxInnovations2020} & 2015 & 2020-08 &  & 2 & 27470 & X & X & X &  \\
+MatrixUser \cite{Liu2016} & 2013 & 2018-07 &  & 1 & 23121 & X & X & X &  \\
 Slice:Drop \cite{Haehn2013} & 2012 & 2020-04 &  & 3 & 19020 & X & X & X & X \\
-GATE \cite{Jan2004} & 2011 & 2020-10 &  & 45 & 207122 &  & X & X &  \\
-ITK-SNAP \cite{Yushkevich2006} & 2006 & 2020-06 & X & 13 & 88530 & X & X & X &  \\
-ParaView \cite{Ahrens2005} & 2002 & 2020-10 & X & 100 & 886326 & X & X & X & X \\
-MatrixUser \cite{Liu2016} & 2013 & 2018-07 &  & 1 & 23121 & X & X & X &  \\
-DICOM Viewer \cite{Afsar2021} & 2018 & 2020-04 & X & 5 & 30761 & X & X & X &  \\
-INVESALIUS 3 \cite{Amorim2015} & 2009 & 2020-09 &  & 10 & 48605 & X & X & X &  \\
-medInria \cite{Fillard2012} & 2009 & 2020-11 &  & 21 & 148924 & X & X & X &  \\
 dicompyler \cite{Panchal2010} & 2009 & 2020-01 &  & 2 & 15941 & X & X &  &  \\
-MicroView \cite{ParallaxInnovations2020} & 2015 & 2020-08 &  & 2 & 27470 & X & X & X &  \\
-Papaya \cite{UTHSCSA2019} & 2012 & 2019-05 &  & 9 & 71831 & X & X & X &  \\
-AMIDE \cite{Loening2017} & 2006 & 2017-01 &  & 4 & 102827 & X & X & X &  \\
-Gwyddion \cite{Nevcas2012} & 2004 & 2020-11 &  & 38 & 643427 & X & X & X &  \\
-\hline
+Fiji \cite{Schindelin2012} & 2011 & 2020-08 & X & 55 & 10833 & X & X & X &  \\
+ImageJ \cite{Rueden2017} & 1997 & 2020-08 & X & 18 & 9681 & X & X & X &  \\
+MRIcroGL \cite{Rorden2021} & 2015 & 2020-08 &  & 2 & 8493 & X & X & X &  \\
+DicomBrowser \cite{Archie2012} & 2012 & 2020-08 &  & 3 & 5505 & X & X & X &  \\ \hline
 \end{tabular}
@@
 
-We failed installing \textit{DICOM Viewer}, so we could not test \textit{surface reliability} and \textit{surface robustness} for it.
+We failed to install \textit{DICOM Viewer}, so we could not test its \textit{surface reliability} and \textit{surface robustness}.
 We kept this software on our list because the other seven qualities do not rely on a successful installation.
-Besides, it used a unique dependency, and we wanted to keep the diversity.
+Besides, the \textit{DICOM Viewer} team built it as a plugin software for NextCloud (\hyperlink{https://apps.nextcloud.com/}{https://apps.nextcloud.com/}) platform, which was a unique choice we had not seen before.
+We wanted to keep it to enrich the diversity.
 
@@
 We found installation instructions for 16 projects.
-Among the ones without such an instruction, \textit{BioImage Suite Web} and \textit{Slice:Drop} were web applications and provided online versions to use, thus they needed no installation.
-Installing 10 of the projects needed extra dependencies.
-Five of them are the web applications in Table \ref{tab_final_list}, and depended on a browser; \textit{dwv}, \textit{OHIF Viewer}, and \textit{GATE} needed extra dependencies to build; \textit{ImageJ} and	\textit{Fiji} needed an unzip tool; \textit{MatrixUser} was based on Matlab; \textit{DICOM Viewer} needed to work on Nextcloud platform.
+Among the ones without instructions, \textit{BioImage Suite Web} and \textit{Slice:Drop} are web applications with online versions to use, thus they do not need installation.
+Installing 10 of the projects required extra dependencies.
+Five of them are the web applications in Table \ref{tab_final_list}, and depended on a browser; \textit{dwv}, \textit{OHIF Viewer}, and \textit{GATE} needed extra dependencies to build; \textit{ImageJ} and	\textit{Fiji} needed an unzip tool; \textit{MatrixUser} was based on Matlab; \textit{DICOM Viewer} needed to work on a Nextcloud platform.
 
-\textit{3D Slicer} has the highest score because it had easy-to-follow installation instructions, and the installation processes were automated, fast, and frustration-free, with all dependencies automatically added.
+\textit{3D Slicer} has the highest score because it had easy to follow installation instructions, and the installation processes were automated, fast, and frustration-free, with all dependencies automatically added.
 There were also no errors during the installation and uninstallation steps.
@@
 We gave them various scores based on the understandability of the instructions, installation steps, and user experience.
-Since \textit{BioImage Suite Web} and \textit{Slice:Drop} needed no installation, we gave them higher scores due to the significant convenience.
+Since \textit{BioImage Suite Web} and \textit{Slice:Drop} needed no installation, we gave them higher scores.
 \textit{BioImage Suite Web} also provided an option to download cache to local for offline usage, which was easy to apply.
@@
 \textit{dwv}, \textit{GATE}, and \textit{DICOM Viewer} showed severe problems.
-We could not install them due to some issues that we could not solve.
-We spent a reasonable amount of time on these problems, then considered them major obstacles for normal users if we still did not figure out any solutions.
+We could not install them.
+We spent a reasonable amount of time on these problems, then considered them as major obstacles for normal users if we could not figure out any solutions.
 We suspect that only a part of the users faced the same problems, and given a lot of time, we might be able to find solutions.
@@
 With those, we finished all the measurements for them.
-Furthermore, \textit{DICOM Viewer} depended on a cloud platform, and we could not successfully install the dependency.
+Furthermore, \textit{DICOM Viewer} depended on the NextCloud platform, and we could not successfully install the dependency.
 
@@
 \section{Correctness \& Verifiability}
+\label{sec_result_correctness_verifiability}
+
 The scores of \textit{correctness \& verifiability} are shown in Figure \ref{fg_correctness_erifiability_scores}.
@@
 Unit testing benefits most parts of the software's life cycle, such as designing, coding, debugging, and optimization \cite{Hamill2004}.
-It can reveal the bugs at an earlier stage of the development process, and the absence of unit testing may cause worse \textit{correctness \& verifiability}.
+It can reveal the bugs at an earlier stage of the development process, and the absence of unit testing may cause problems for \textit{correctness \& verifiability}.
 
 We could not find requirements specifications for most projects.
-The only document we found is a road map of \textit{3D Slicer}, which contained design requirements for the following changes.
+The only document we found is a road map of \textit{3D Slicer}, which contained design requirements for the upcoming changes.
 However, it did not record the conditions for previous versions.
 We also could not identify the theory manuals for all of the projects.
-It seems that even for some projects with well-organized documents, requirements specifications and theory manuals were still missing.
+Even for some projects with well-organized documents, requirements specifications and theory manuals were still missing.
 
@@
 
-When applying basic operations with the software packages, we found out that \textit{Drishti} crashed during loading damaged image files, and \textit{GATE} could not open macro files and lost response several times.
+As shown in Section \ref{sec_result_installability}, most of the software products did not ``break" during installation or did not need installation; \textit{dwv} and \textit{GATE} broke in the building stage, and the processes were not recoverable; we could not install the dependency for \textit{DICOM Viewer}.
+
+Of the seven software packages with a getting started tutorial and operation steps in the tutorial, most showed no error when we followed the steps.
+However, \textit{GATE} could not open macro files and became unresponsive several times, without any descriptive error message.
+When assessing \textit{surface robustness} (Section \ref{sec_result_robustness}), we found out that \textit{Drishti} crashed during loading damaged image files and did not show any descriptive error message.
+On the other hand, we did not see any problems with the online version of \textit{dwv}.
 
@@
 
-The packages with higher scores elegantly handled the unexpected/unanticipated inputs, normally showing a clear error message.
+The packages with higher scores elegantly handled the unexpected/unanticipated inputs, typically showing a clear error message.
 We might underestimate the score of \textit{OHIF Viewer} since we needed further customization to load data, and the test was not complete.
-\textit{MatrixUser}, \textit{dwv}, \textit{Slice:Drop}, and \textit{MRIcroGL} ignored the incorrect format of the input files, and displayed blank or meaningless images.
-\textit{Drishti} successfully detected the unexpected/unanticipated inputs, but the software crashed as a result.
-For unknown reasons, \textit{GATE} failed to load both correct and incorrect inputs.
+
+Digital Imaging and Communications in Medicine (DICOM) is a widely used MI standard, and ``it defines the formats for medical images that can be exchanged with the data and quality necessary for clinical use" \cite{MITA2021}.
+According to their documentation, all 29 software packages should support the DICOM standard.
+We prepared two types of image files: the ones in correct formats and the broken ones.
+We used two MI sample files in the DICOM format as the image files with valid formats; we created a standard text file, changed its extension name from ``.txt" to ``.dcm", and used it as the unexpected/unanticipated input.
+
+Being tested with the input files with correct formats, all software packages except \textit{GATE} loaded the images correctly. \textit{GATE} failed this test for unknown reasons.
+
+With the unexpected/unanticipated input, \textit{MatrixUser}, \textit{dwv}, and \textit{Slice:Drop} ignored the incorrect format of the file and loaded it regardless.
+They did not show any error message and displayed a blank image.
+\textit{MRIcroGL} behaved similarly except that it showed a meaningless image with noise pixels.
+\textit{Drishti} successfully detected the broken format of the file, but the software crashed as a result.
+We recorded \textit{Drishti}'s issue to the measurement of its \textit{reliability} in Section \ref{sec_result_installability}.
 
@@
 
-We found a getting started tutorial for only 11 projects but a user manual for 22 projects. \textit{MRIcroGL} was the only one with expected user characteristics documented.
+We found a getting started tutorial for only 11 projects but a user manual for 22 projects. \textit{MRIcroGL} was the only one with documentation for the expected user characteristics.
 
-The ones with higher scores usually provided both comprehensive document guidance and a good user experience.
+The software with higher scores usually provided both comprehensive document guidance and a good user experience.
 \textit{INVESALIUS 3} set an excellent example of a detailed and precise user manual.
@@
 Meanwhile, \textit{3D Slicer} has a much higher percentage of closed issues (91.65\%) than \textit{ImageJ} (52.49\%) and \textit{Fiji} (63.79\%).
-Table \ref{tab_maintainability_docs} shows which projects had these documents.
+Table \ref{tab_maintainability_docs} shows which projects had these documents, in the descending order of their \textit{maintainability} scores.
 
@@
 3D Slicer & X & X & X \\
+ImageJ & X & X & X \\
 Weasis &  & X &  \\
+OHIF Viewer &  & X & X \\
+Fiji & X & X & X \\
+ParaView & X &  &  \\
 SMILI &  &  & X \\
-ImageJ & X & X & X \\
-Fiji & X & X & X \\
+medInria &  & X &  \\
+INVESALIUS 3 & X &  &  \\
 dwv &  &  & X \\
 BioImage Suite Web &  & X &  \\
-OHIF Viewer &  & X & X \\
-ParaView & X &  &  \\
-INVESALIUS 3 & X &  &  \\
-medInria &  & X &  \\
 Gwyddion &  & X & X \\ \hline
@@
 
-As described in Section \ref{sec_grading_template}, we gave higher scores to the projects with an API document and more code files.
+As described in Section \ref{sec_grading_template}, we gave higher scores to the projects with an API document.
 As shown in Table \ref{tab_maintainability_docs}, seven projects had API documents.
-Figure \ref{fg_num_text_files} shows the number of text-based files by projects.
+We also considered projects with more code files and less LOC per code file as more reusable.
+Table \ref{tab_loc_per_file} shows the number of text-based files by projects, which we used to approximate the number of code files.
+The table also lists the total number of lines (including comments and blanks), LOC, and average LOC per file.
+We arranged the items in the descending order of their \textit{reusability} scores.
 
-\begin{figure}[H]
-\includegraphics[scale=0.38]{figures/num_text_files.png}
-\caption{Number of text-based files files by projects}
-\label{fg_num_text_files}
-\end{figure}
+\begin{table}[H]
+\centering
+\begin{tabular}{lllll}
+\hline
+\multirow{2}{*}{Software} & \multirow{2}{*}{Text files} & \multirow{2}{*}{Total lines} & \multirow{2}{*}{LOC} & \multirow{2}{*}{LOC/file} \\
+ &  &  &  &  \\ \hline
+OHIF Viewer & 1162 & 86306 & 63951 & 55 \\
+3D Slicer & 3386 & 709143 & 501451 & 148 \\
+Gwyddion & 2060 & 787966 & 643427 & 312 \\
+ParaView & 5556 & 1276863 & 886326 & 160 \\
+OsiriX Lite & 2270 & 873025 & 544304 & 240 \\
+Horos & 2346 & 912496 & 561617 & 239 \\
+medInria & 1678 & 214607 & 148924 & 89 \\
+Weasis & 1027 & 156551 & 123272 & 120 \\
+BioImage Suite Web & 931 & 203810 & 139699 & 150 \\
+GATE & 1720 & 311703 & 207122 & 120 \\
+Ginkgo CADx & 974 & 361207 & 257144 & 264 \\
+SMILI & 275 & 90146 & 62626 & 228 \\
+Fiji & 136 & 13764 & 10833 & 80 \\
+Drishti & 757 & 345225 & 268168 & 354 \\
+ITK-SNAP & 677 & 139880 & 88530 & 131 \\
+3DimViewer & 730 & 240627 & 178065 & 244 \\
+DICOM Viewer & 302 & 34701 & 30761 & 102 \\
+ImageJ & 40 & 10740 & 9681 & 242 \\
+dwv & 188 & 71099 & 47815 & 254 \\
+MatrixUser & 216 & 31336 & 23121 & 107 \\
+INVESALIUS 3 & 156 & 59328 & 48605 & 312 \\
+AMIDE & 183 & 139658 & 102827 & 562 \\
+Papaya & 110 & 95594 & 71831 & 653 \\
+MicroView & 137 & 36173 & 27470 & 201 \\
+XMedCon & 202 & 129991 & 96767 & 479 \\
+MRIcroGL & 97 & 50445 & 8493 & 88 \\
+Slice:Drop & 77 & 25720 & 19020 & 247 \\
+DicomBrowser & 54 & 7375 & 5505 & 102 \\
+dicompyler & 48 & 19201 & 15941 & 332 \\ \hline
+\end{tabular}
+\caption{\label{tab_loc_per_file}Number of files and lines}
+\end{table}
 
@@
 All projects had a consistent code style with parameters in the same order for all functions; the code was modularized; the comments were clear, indicating what is being done, not how.
-However, we only found explicit identification of a coding standard for only 3 out of the 29, which are \textit{3D Slicer}, \textit{Weasis}, and \textit{ImageJ}.
+However, we only found explicit identification of a coding standard for 3 out of the 29: \textit{3D Slicer}, \textit{Weasis}, and \textit{ImageJ}.
 We also found hard-coded constants in \textit{medInria}, \textit{dicompyler}, \textit{MicroView}, and \textit{Papaya}.
@@
 
-Table \ref{tab_Visibility/Transparency_docs} shows the projects which had documents for the development process, project status, development environment, and release notes.
+Table \ref{tab_Visibility/Transparency_docs} shows the projects which had documents for the development process, project status, development environment, and release notes, in the descending order of their \textit{visibility/transparency} scores.
 
-\begin{table}[]
+\begin{table}[H]
 \centering
@@
 3D Slicer & X & X & X & X \\
-Weasis \cite{Roduit2021} &  &  & X & X \\
-MRIcroGL \cite{Rorden2021} &  &  &  & X \\
-SMILI \cite{Chandra2018} &  &  &  & X \\
-ImageJ \cite{Rueden2017} & X & X & X & X \\
-Fiji \cite{Schindelin2012} & X & X & X &  \\
-Horos \cite{horosproject2020} &  &  &  & X \\
-OsiriX Lite \cite{PixmeoSARL2019} &  &  &  & X \\
-dwv \cite{Martelli2021} &  &  &  & X \\
-Drishti \cite{Limaye2012} &  &  &  & X \\
-\begin{tabular}[c]{@{}l@{}}BioImage Suite\\ Web\end{tabular} &  &  & X &  \\
-OHIF Viewer \cite{Ziegler2020} &  &  & X & X \\
-GATE \cite{Jan2004} &  &  &  & X \\
-ITK-SNAP \cite{Yushkevich2006} &  &  &  & X \\
-ParaView \cite{Ahrens2005} &  & X &  &  \\
-MatrixUser \cite{Liu2016} &  &  &  & X \\
-DICOM Viewer \cite{Afsar2021} &  &  & X & X \\
-INVESALIUS 3 \cite{Amorim2015} &  &  &  & X \\
-medInria \cite{Fillard2012} &  &  & X & X \\
-MicroView \cite{ParallaxInnovations2020} &  &  &  & X \\
-Gwyddion \cite{Nevcas2012} &  &  &  & X \\ \hline
+ImageJ & X & X & X & X \\
+Fiji & X & X & X &  \\
+MRIcroGL &  &  &  & X \\
+Weasis &  &  & X & X \\
+ParaView &  & X &  &  \\
+OHIF Viewer &  &  & X & X \\
+DICOM Viewer &  &  & X & X \\
+medInria &  &  & X & X \\
+SMILI &  &  &  & X \\
+Drishti &  &  &  & X \\
+INVESALIUS 3 &  &  &  & X \\
+OsiriX Lite &  &  &  & X \\
+GATE &  &  &  & X \\
+MicroView &  &  &  & X \\
+MatrixUser &  &  &  & X \\
+BioImage Suite Web &  &  & X &  \\
+ITK-SNAP &  &  &  & X \\
+Horos &  &  &  & X \\
+dwv &  &  &  & X \\
+Gwyddion &  &  &  & X \\ \hline
 \end{tabular}
@@
 As described in Section \ref{sec_AHP}, for our AHP measurements, there are nine criteria which are the nine software qualities and 29 software packages as the alternatives.
-We decided to make all nine qualities equally important, so the score of each quality affects the overall scores on the same scale.
+In the absence of requirements for an actual project, we made all nine qualities equally important, so the score of each quality affects the overall scores on the same scale.
 

### 2021-08-27 StateOfPractice/Ao-Report/15InterviewsWithDevelopers.tex
MI report - updated ch 6 and more
MI report - updated ch 5

@@
 
-We applied our method to the MI domain (Section \ref{sec_applying_method}) and summarize some answers from the interviews with developers in this section.
-We only discuss the answers that we think are the most exciting and essential in this section, and summarize the rest in Appendix \ref{ap_interview}.
+Our method involves interviews with developers in the MI domain (Section \ref{sec_applying_method}).
+We summarize some answers from the interviews in this section.
+We highlight the answers that are the most informative and interesting in this section, and summarize the rest in Appendix \ref{ap_interview}.
 
@@
 The interviewees may have provided multiple answers to each question.
-Thus, when counting the number of answers, the total result can be larger than nine.
+Thus, when counting the number of answers, the total result is sometimes larger than nine.
 
@@
 
-By asking questions 9, 10, and 12, we tried to identify the pain points during the development process in the eight projects.
+By asking questions \hyperlink{q9}{9}, \hyperlink{q10}{10}, and \hyperlink{q12}{12}, we tried to identify the pain points during the development process in the eight projects.
 The pain points include current and past obstacles.
@@
 \begin{description}
-\item[Q9.] Currently, what are the most significant obstacles in your development process?
-\item[Q10.] How might you change your development process to remove or reduce these obstacles?
-\item[Q12.] In the past, is there any major obstacle to your development process that has been solved? How did you solve it?
+\item[Q9.] \hypertarget{q9}Currently, what are the most significant obstacles in your development process?
+\item[Q10.] \hypertarget{q10}How might you change your development process to remove or reduce these obstacles?
+\item[Q12.] \hypertarget{q12}In the past, is there any major obstacle to your development process that has been solved? How did you solve it?
 \end{description}
@@
 \hline
-\multirow{2}{*}{Grp} & \multirow{2}{*}{Obstacle} & \multicolumn{2}{l}{Num ans.} \\ \cline{3-4} 
+\multirow{2}{*}{Category} & \multirow{2}{*}{Obstacle} & \multicolumn{2}{c}{Num ans.} \\ \cline{3-4} 
  &  & current & past \\ \hline
-\multirow{2}{*}{1} & Lack of fundings & 3 &  \\
+\multirow{2}{*}{Resource} & Lack of fundings & 3 &  \\
  & Lack of time to devote to the project & 2 & 1 \\ \hdashline
-\multirow{3}{*}{2} & Hard to keep up with changes in OS and libraries & 1 &  \\
+\multirow{3}{*}{Balance} & Hard to keep up with changes in OS and libraries & 1 &  \\
  & Hard to support multiple OS & 2 &  \\
  & Hard to support lower-end computers & 1 & 2 \\ \hdashline
-3 & Lack of access to real-world datasets for testing & 3 & 2 \\ \hdashline
-\multirow{7}{*}{Oths}
+Testing & Lack of access to real-world datasets for testing & 3 & 2 \\ \hdashline
+\multirow{7}{*}{Others}
  & Hard to have a high level roadmap from the start & 1 &  \\
@@
 The interviewees provided some potential and proven solutions for the problems in Table \ref{tab_obstacles}.
-We group these pain points into three major groups, and put the less mentioned ones into the group \textit{Others}.
+We group these pain points into three major categories of obstacles: \textit{resource}, \textit{balance}, and \textit{testing}.
+We put the less mentioned ones into the category \textit{Others}.
 Sections \ref{sec_pain_points_1}, \ref{sec_pain_points_2}, and \ref{sec_pain_points_3} include further discussion about the three major groups of pain points and their solutions.
 
-\subsection{Pain Points in Group 1}
+\subsection{Resource Pain Points}
 \label{sec_pain_points_1}
-We summarize the pain points in Group 1 as
+We summarize the pain points in the \textit{resource} category as
 \textbf{the lack of fundings and time.}
 
-We also summarize the potential and proven solutions as follows.
+The potential and proven solutions are:
 
@@
 
-\subsection{Pain Points in Group 2}
+\subsection{Balance Pain Points}
 \label{sec_pain_points_2}
-We summarize the pain points in Group 2 as
+We summarize the pain points in the \textit{balance} category as
 \textbf{the difficulty to balance between four factors: cross-platform compatibility, convenience to development \& maintenance, performance, and security.} They are also related to the choice between native application and web application.
 
-We also summarize the potential and proven solutions as follows.
+The potential and proven solutions are:
 
@@
 
-The interviewees did not conclude any perfect solutions.
+\subsection{Testing Pain Point}
+\label{sec_pain_points_3}
+The pain point in the \textit{testing} category is \textbf{the lack of access to real-world datasets for testing.}
 
-\subsection{Pain Points in Group 3}
-\label{sec_pain_points_3}
-The pain point in Group 3 is
-\textbf{the lack of access to real-world datasets for testing.}
-
-We also summarize the potential and proven solutions as follows.
+The potential and proven solutions are:
 
@@
 \item sending the beta versions of software to medical workers who can access the data and complete the tests;
-\item if (part of) the team belongs to a medical school or a hospital, using the datasets they can access to;
-\item if the team has access to MRI scanners, self-building MI datasets;
+\item if (part of) the team belongs to a medical school or a hospital, using the datasets they can access;
+\item if the team has access to MRI scanners, self-building sample images for testing;
 \item if the team has connections with MI equipment manufacturers, asking for their help on data format problems;
@@
 Interviewees from each of the eight projects thought that documentation was important to their projects, and most of them said that it could save their time to answer questions from users and developers.
-However, most of them saw the need to improve their documentation, and only three of them thought that their documentations conveyed information clearly enough.
+Most of them saw the need to improve their documentation, and only three of them thought that their documentations conveyed information clearly enough.
 
@@
 Documentation of the project conveys information clearly & 3 \\
-Coding is more preferable than documentation & 2 \\
+Coding is more fun than documentation & 2 \\
 Users help each other by referring to documentation & 1 \\ \hline
@@
 
-Table \ref{tab_doc_tools} lists some of the tools and methods mentioned by the interviewees, which they used for documentation.
+Table \ref{tab_doc_tools} lists some of the documentation tools and methods mentioned by the interviewees.
 
@@
 \label{sec_contribution_pm}
-We tried to understand how the teams managed the contributions and their projects by asking the questions:
+We tried to understand how the teams managed the contributions and their projects by asking the following questions:
 
@@
 \item[Q13.] What is your software development model? For example, waterfall, agile, etc.
-\item[Q14.] What is your project management process? Do you think improving this process can tackle the current problem? Were any project management tools used?
+\item[Q14.] \hypertarget{q14}What is your project management process? Do you think improving this process can tackle the current problem? Were any project management tools used?
 \end{description}
 
-\textit{}
-
-\textit{}
-
-\textit{}
-
-Maybe some team had a documented process for accepting new contributions, but none talked about it during the interview.
+Although some team may have a documented process for accepting new contributions, no one talked about it during the interview.
 However, most of them mentioned using GitHub and pull requests to manage contributions from the community.
@@
 
-Additionally, the \textit{3D Slicer} team encouraged users to develop their extensions for specific use cases, and the \textit{OHIF} team was trying to enable the use of plug-ins; the interviewee from the \textit{ITK-SNAP} team said one way of accepting new team members was through funded academic projects.
+For managing contributions, the \textit{3D Slicer} team encouraged users to develop their extensions for specific use cases, and the \textit{OHIF} team was trying to enable the use of plug-ins; the interviewee from the \textit{ITK-SNAP} team said one way of accepting new team members was through funded academic projects.
 
@@
 
-Most interviewees skipped the question ``Do you think improving this process can tackle the current problem?".
-We concluded that a yes-or-no question is not very informative here.
+Most interviewees skipped the second part of \hyperlink{q14}{Q14} ``Do you think improving this process can tackle the current problem?".
+In retrospect, we should not have asked a yes-or-no question, since it is not very informative.
 The interviewee from the \textit{OHIF} team gave a positive answer to this question.
@@
 \label{sec_interview_software_qualities}
-Questions 15, 16, 17, 18, and 20 are about software qualities \textit{correctness}, \textit{maintainability}, \textit{understandability}, \textit{usability}, and \textit{reproducibility} respectively.
+Questions 15--18, and 20 are about the software qualities of \textit{correctness}, \textit{maintainability}, \textit{understandability}, \textit{usability}, and \textit{reproducibility}, respectively.
 We asked these questions to understand the threats to these qualities and the developers' strategies to improve them.
+We discuss each quality in a separate section below.
 
 \subsection{Correctness}
-
+\label{sec_interview_correctness}
 \begin{description}
@@
 Complexity - different MI machines create data in different ways. & 2 \\
-Complexity - additional functions beside of viewing. & 1 \\
+Complexity - additional functions besides viewing. & 1 \\
 The lack of real word image data for testing. & 1 \\
@@
 With huge datasets for testing, the tests are expensive and time-consuming. & 1 \\
-It is hard to well manage releases. & 1 \\
+It is hard to manage releases. & 1 \\
 The project has no unit tests. & 1 \\
@@
 \centering
-\hspace*{-1cm}\begin{tabular}{ll}
+\hspace*{-1.5cm}\begin{tabular}{ll}
 \hline
 Strategy to ensure correctness & Num ans. \\ \hline
-Test-driven development, component / integration / smoke / regression tests. & 4 \\
+Test-driven development, component tests, integration tests, smoke tests, regression tests. & 4 \\
 Self tests / automated tests. & 3 \\
@@
 \subsection{Maintainability}
-
+\label{sec_interview_maintainability}
 \begin{description}
-\item[Q16.] When designing the software, did you consider the ease of future changes?
+\item[Q16.] \hypertarget{q16}When designing the software, did you consider the ease of future changes?
 For example, will it be hard to change the system’s structure, modules, or code blocks?
@@
 \begin{description}
-\item[Q17.] Provide instances where users have misunderstood the software. What, if any, actions were taken to address understandability issues?
+\item[Q17.] \hypertarget{q17}Provide instances where users have misunderstood the software. What, if any, actions were taken to address understandability issues?
 \end{description}
@@
 \subsection{Usability}
-
+\label{sec_interview_usability}
 \begin{description}
-\item[Q17.] What, if any, actions were taken to address usability issues?
+\item[Q18.] \hypertarget{q18}What, if any, actions were taken to address usability issues?
 \end{description}
@@
 \subsection{Reproducibility}
-
+\label{sec_interview_reproducibility}
 \begin{description}

### 2021-08-27 StateOfPractice/Ao-Report/16AnswersToResearchQuestions.tex
MI report - updated ch 6

@@
 
-This section answers our research questions in Section \ref{sec_research_questions}.
+This section answers our research questions from Section \ref{sec_research_questions}.
 The answers are based on our quality measurements and developer interviews.
-Sections \ref{sec_rq_artifacts}, \ref{sec_rq_tools}, \ref{sec_rq_PPM}, \ref{sec_rq_pain_points}, \ref{sec_rq_qualities}, and \ref{sec_rq_comparison} summarize the answers to the six questions respectively.
+Sections \ref{sec_rq_artifacts}-- \ref{sec_rq_comparison} summarize the answers to the six questions, respectively.
 Section \ref{sec_threats_to_validity} lists the threats to the validity of our research.
@@
 
-Table \ref{tab_artifacts_frequency} shows the frequency of some artifacts in the 29 MI projects.
+As mentioned in Section \ref{sec_grading_template}, we search for all the artifacts in a project when measuring \textit{maintainability}.
+The detailed records of the existing artifacts in the 29 MI projects are at \hyperlink{https://data.mendeley.com/datasets/k3pcdvdzj2/1}{https://data.mendeley.com/datasets/k3pcdvdzj2/1}.
+Table \ref{tab_artifacts_frequency} summarizes the frequency of the artifacts .
 
@@
 \textit{3D Slicer} and \textit{OHIF Viewer} used CircleCI; \textit{ImageJ}, \textit{Fiji}, and \textit{dwv} used Travis.
+We identified the above projects and tools by examining the documentation and source code of all projects.
+Thus, we may missed some projects or tools.
 According to the interviews with developers, \textit{dwv} and \textit{Weasis} used GitHub Actions.
@@
 \subsection{User Support Tools}
-Table \ref{tab_user_support_model} shows the user support tools by the number of projects using them.
+Table \ref{tab_user_support_model} shows the user support tools by the number of projects using them. The two most popular tools are GitHub issue and FAQ.
 
 \subsection{Version Control Tools}
-Briefly speaking, 27 of the 29 projects chose git as the version control tool, and 24 used GitHub to host the repositories. Section \ref{sec_score_maintainability} lists more details.
+Briefly speaking, 27 of the 29 projects chose git as the version control tool, and 24 used GitHub to host the repositories. Section \ref{sec_score_maintainability} provides more details.
 
 \subsection{Documentation Tools}
-Table \ref{tab_doc_tools} lists the documentation tools by the number of answers from the interviewees.
+Table \ref{tab_doc_tools} lists the documentation tools by the number of answers from the interviewees. The two most popular tools are forum discussions and videos.
 
@@
 Table \ref{tab_method_receive_contributions} shows the ways of receiving contributions by the number of answers from the interviewees.
+All eight development teams were using GitHub with pull requests as the tool.
+The most popular tool in the past was email.
 
 \subsection{Project Management Tools}
-Table \ref{tab_pm_tools} shows the tools for project management by the number of answers from the interviewees.
+Table \ref{tab_pm_tools} shows the tools for project management by the number of answers from the interviewees. GitHub was the top-chosen tool.
 
@@
 \subsection{Software Testing}
-As shown in Section \ref{fg_correctness_erifiability_scores}, we only identified the use of unit testing in less than half of the 29 projects.
-Most projects did not have requirements specifications and theory manuals.
+As shown in Section \ref{fg_correctness_erifiability_scores}, we identified the use of unit testing in less than half of the 29 projects.
 
@@
 Table \ref{tab_q15_threats_correctness} lists a threat to \textit{correctness}: with huge datasets for testing, the tests are expensive and time-consuming.
-Table \ref{tab_q15_strategies_correctness} shows that three interviewees endorsed self tests / automated tests, which may save the time for testing.
+Table \ref{tab_q15_strategies_correctness} shows that three interviewees endorsed self tests / automated tests, which may save time for testing.
 
@@
 \subsection{Documentation}
-Section \ref{fg_correctness_erifiability_scores} shows that most projects did not have requirements specifications and theory manuals.
+Section \ref{fg_correctness_erifiability_scores} shows that all projects did not have theory manuals.
+We identified a road map in the \textit{3D Slicer} project, and no requirements specifications for the rest.
 Table \ref{tab_maintainability_docs} presents which projects had a project plan, developer’s manual, or API documentation, which implies better \textit{maintainability}.
@@
 
-Most of the nine interviewees thought that documentation was essential to their projects.
+Eight of the nine interviewees thought that documentation was essential to their projects.
 However, they hold the common opinion that their documentation needed improvements.
@@
 
-We answer this question primarily by interviewing the developers. This section presents the pain points and the solutions from the developers and our team.
+We answer this question primarily by interviewing the developers. This section presents the pain points and the solutions proposed by the developers.
 
@@
 Section \ref{sec_interview_pain_points} lists the details about the potential and proven solutions from the developers in the MI domain.
-We also summarize our general recommendations on software development in all SC domains in Section \ref{ch_recommendations}.
 
@@
 We applied (surface) measurements to the nine qualities, and hope that the grading scores represent their current status.
-We believe that our assessments also revealed some of the development teams' practices addressing the qualities.
 
@@
 We might underestimate the scores for \textit{dwv} due to uncommon technical issues.
+We also ranked \textit{DICOM Viewer} much lower than its popularity.
+As mentioned in Section \ref{sec_result_installability} it depended on the NextCloud platform that we could not successfully install.
+Thus, we might underestimate the scores of its \textit{surface reliability} and \textit{surface robustness}.
+We weighted all qualities equally, which is not likely to be the case with all end-users.
+As a result, some projects may have received higher popularity because of their better performance in part of the nine qualities.
 
@@
 \hline
-Software & Our ranking & Stars/yr & Watch/yr & Forks/yr \\ \hline
+Software & Our ranking & Stars/yr & Watches/yr & Forks/yr \\ \hline
 3D Slicer & 1 & 284.25 & 18.75 & 128.25 \\
-ImageJ & 2 & 83.58 & 9.37 & 30.00 \\
-OHIF Viewer & 3 & 277.04 & 19.30 & 223.83 \\
-Fiji & 4 & 44.20 & 4.98 & 20.68 \\
+OHIF Viewer & 3 & 277.04 & 19.3 & 223.83 \\
+dwv & 23 & 123.56 & 11.59 & 51.36 \\
+ImageJ & 2 & 83.58 & 9.37 & 30 \\
 ParaView & 5 & 66.76 & 7.11 & 28.36 \\
-Weasis & 6 & 36.00 & 5.10 & 18.90 \\
+Horos & 12 & 48.62 & 9.23 & 18 \\
+Papaya & 17 & 44.7 & 4.84 & 19.71 \\
+Fiji & 4 & 44.2 & 4.98 & 20.68 \\
+DICOM Viewer & 29 & 42.86 & 5.71 & 8.57 \\
+INVESALIUS 3 & 10 & 39.66 & 4.11 & 17.37 \\
+Weasis & 6 & 36 & 5.1 & 18.9 \\
+dicompyler & 26 & 35.18 & 4.77 & 13.64 \\
+OsiriX Lite & 9 & 34.14 & 8.62 & 23.77 \\
+MRIcroGL & 18 & 23.63 & 3.38 & 3.38 \\
+GATE & 25 & 18.9 & 5.79 & 25.79 \\
+Ginkgo CADx & 14 & 18.83 & 4.43 & 5.72 \\
+BioImage Suite Web & 8 & 17.85 & 4.62 & 6.77 \\
+Drishti & 27 & 16.04 & 3.78 & 4.43 \\
+Slice:Drop & 20 & 9.73 & 2.16 & 4.97 \\
+ITK-SNAP & 15 & 9.1 & 0.97 & 3.59 \\
 medInria & 7 & 7.04 & 3.35 & 6.35 \\
-BioImage Suite Web & 8 & 17.85 & 4.62 & 6.77 \\
-OsiriX Lite & 9 & 34.14 & 8.62 & 23.77 \\
-INVESALIUS 3 & 10 & 39.66 & 4.11 & 17.37 \\
+SMILI & 13 & 3.04 & 0.91 & 1.52 \\
+MatrixUser & 28 & 2 & 0.33 & 0.33 \\
+MicroView & 16 & 1.48 & 0.82 & 0.82 \\
 Gwyddion & 11 & n/a & n/a & n/a \\
-Horos & 12 & 48.62 & 9.23 & 18.00 \\
-SMILI & 13 & 3.04 & 0.91 & 1.52 \\
-Ginkgo CADx & 14 & 18.83 & 4.43 & 5.72 \\
-ITK-SNAP & 15 & 9.10 & 0.97 & 3.59 \\
-MicroView & 16 & 1.48 & 0.82 & 0.82 \\
-Papaya & 17 & 44.70 & 4.84 & 19.71 \\
-MRIcroGL & 18 & 23.63 & 3.38 & 3.38 \\
 XMedCon & 19 & n/a & n/a & n/a \\
-Slice:Drop & 20 & 9.73 & 2.16 & 4.97 \\
-DicomBrowser & 21 & n/a & n/a & n/a \\
+DicomBrower & 21 & n/a & n/a & n/a \\
 AMIDE & 22 & n/a & n/a & n/a \\
-dwv & 23 & 123.56 & 11.59 & 51.36 \\
-3DimViewer & 24 & n/a & n/a & n/a \\
-GATE & 25 & 18.90 & 5.79 & 25.79 \\
-dicompyler & 26 & 35.18 & 4.77 & 13.64 \\
-Drishti & 27 & 16.04 & 3.78 & 4.43 \\
-MatrixUser & 28 & 2.00 & 0.33 & 0.33 \\
-DICOM Viewer & 29 & 42.86 & 5.71 & 8.57 \\ \hline
+3DimViewer & 24 & n/a & n/a & n/a \\ \hline
 \end{tabular}
@@
 
-An example of interpreting Table \ref{tab_top_software_vs_experts} is as follows: in row 2, \textit{ImageJ} has rank 2 by our assessment, and the number \textit{(3)} means that it has rank 3 by the experts; likewise, \textit{Horos} has rank 2 by domain expers, and rank 12 by our assessment.
+An example of interpreting Table \ref{tab_top_software_vs_experts} is as follows: in row 2, \textit{ImageJ} has rank 2 by our assessment, and the number \textit{(3)} means that it has rank 3 by the experts; likewise, \textit{Horos} has rank 2 by the domain experts, and rank 12 by our assessment.
 
@@
 
-As mentioned in Section \ref{sec_mi_software_selection}, six of the recommended software packages did not have Visualization as the primary function, so we did not include them on our final list.
+As mentioned in Section \ref{sec_mi_software_selection}, six of the recommended software packages did not have visualization as the primary function, so we did not include them on our final list.
 
@@
 \begin{enumerate}
-\item We compared nine software qualities for 29 software packages, so we could only spend a limited time on each of them.
-As a result, our assessments may not be thorough in revealing their status fully.
+\item We compared nine software qualities for 29 software packages, so we could only spend a limited time on each of them. As a result, our assessments may have missed something relevant.
 \item We used the grading template in Appendix \ref{ap_grading_template} to guide our measurements.
@@
 There may be cases that we missed some information.
+\item Our ranking is partly based on surface (shallow) measurement, which may not fully reveal the underlying qualities.
 \item We interviewed eight teams, which is a good proportion of the 29. However, there is still a risk that they might not well represent the whole MI software community.
@@
 Since we did not have the chance to explain our questions or ask him follow-up questions, there is a possibility of misinterpretation of the questions or answers.
-\item As mentioned in Section \ref{sec_rq_comparison}, we gave \textit{dwv} much lower scores than its GitHub popularity. We might underestimate its rank due to uncommon technical issues.
+\item As mentioned in Section \ref{sec_result_installability}, we could not install or build \textit{dwv}, \textit{GATE}, and \textit{DICOM Viewer}.
+We used a deployed online version for \textit{dwv}, a VM version for \textit{GATE}, but no alternative for \textit{DICOM Viewer}.
+We might underestimate their rank due to uncommon technical issues.
+\item Our ranking gave all qualities equal weight, which may not be the case with all end-users. Thus, it may not represent the popularity of software among end-users.
+\item The number of GitHub stars, watches, and forks are not perfect measures of popularity, but they are what we had available.
 \end{enumerate}

### 2021-08-28 StateOfPractice/Ao-Report/14MeasurementResults.tex
MI report - updated ch 6 and more

@@
 \section{Surface Reliability}
+\label{sec_result_reliability}
 As described in Section \ref{sec_result_installability}, we could not build \textit{dwv} and \textit{GATE}.
@@
 \section{Surface Usability}
+\label{sec_result_usability}
 Figure \ref{fg_usability_scores} lists the AHP scores for \textit{surface usability}.
@@
 GitHub issue & 24 \\
-GitLab issue, SourceForge discussions & 2 \\
-FAQ & 12 \\
+Frequently Asked Questions (FAQ) & 12 \\
 Forum & 10 \\
 E-mail address & 9 \\
+GitLab issue, SourceForge discussions & 2 \\
 Troubleshooting & 2 \\
@@
 \section{Reusability}
+\label{sec_result_reusability}
 Figure 4.7 shows the AHP results for \textit{reusability}.
@@
 \section{Surface Understandability}
+\label{sec_result_understandability}
 Figure \ref{fg_surface_understandability_scores} shows the scores for \textit{surface understandability}.
@@
 \section{Visibility/Transparency}
+\label{sec_result_visibility_transparency}
 Figure \ref{fg_visibility_transparency_scores} shows the AHP scores for \textit{visibility/transparency}.

### 2021-08-28 StateOfPractice/Ao-Report/16AnswersToResearchQuestions.tex
MI report - updated ch 6 and more

@@
 This section answers our research questions from Section \ref{sec_research_questions}.
-The answers are based on our quality measurements and developer interviews.
 Sections \ref{sec_rq_artifacts}-- \ref{sec_rq_comparison} summarize the answers to the six questions, respectively.
 Section \ref{sec_threats_to_validity} lists the threats to the validity of our research.
+The answers are based on our quality measurements in Section \ref{ch_results} and developer interviews in Section \ref{ch_interview}.
+We refer to these sections to avoid repetition, and organize the references in tables (e.g.
+Table \ref{tab_reference_rq1}).
 
@@
 We answer this question by examining the documentation, scanning the source code, and interviewing the developers of the projects.
+Table \ref{tab_reference_rq1} shows the sections and tables containing answers to this research question.
 
-Part of the answer is in Section \ref{ch_results}.
-Table \ref{tab_maintainability_docs} shows which projects contained a project plan, development manual, or API documentation, and Table \ref{tab_Visibility/Transparency_docs} lists the projects with a documented development process, project status, development environment, or release notes.
+\begin{table}[H]
+\centering
+\begin{tabular}{ll}
+\hline
+Section or table & Description \\ \hline
+Table \ref{tab_maintainability_docs} & Maintainability documents \\
+Table \ref{tab_Visibility/Transparency_docs} & Visibility/transparency documents \\ \hline
+\end{tabular}
+\caption{\label{tab_reference_rq1}Sections and tables with answers to RQ1}
+\end{table}
 
@@
 The detailed records of the existing artifacts in the 29 MI projects are at \hyperlink{https://data.mendeley.com/datasets/k3pcdvdzj2/1}{https://data.mendeley.com/datasets/k3pcdvdzj2/1}.
-Table \ref{tab_artifacts_frequency} summarizes the frequency of the artifacts .
+Table \ref{tab_artifacts_frequency} summarizes the frequency of the artifacts.
 
@@
 This section summarizes the tools used for CI/CD, user support, version control, documentation, contribution management, and project management.
+Table \ref{tab_reference_rq2} shows the sections and tables containing answers to this research question.
 
-\subsection{CI/CD Tools}
+\begin{table}[H]
+\centering
+\begin{tabular}{ll}
+\hline
+Section or table & Description \\ \hline
+Section \ref{sec_result_correctness_verifiability} & CI/CD tools \\
+Table \ref{tab_user_support_model} & User support tools \\
+Section \ref{sec_score_maintainability} & Version control tools \\
+Table \ref{tab_doc_tools} & Documentation tools \\
+Table \ref{tab_method_receive_contributions} & Contribution management tools \\
+Table \ref{tab_pm_tools} & Project management tools\\ \hline
+\end{tabular}
+\caption{\label{tab_reference_rq2}Sections and tables with answers to RQ2}
+\end{table}
+
 As mentioned in Section \ref{sec_result_correctness_verifiability}, we identified five projects using CI/CD tools.
@@
 According to the interviews with developers, \textit{dwv} and \textit{Weasis} used GitHub Actions.
-
-\subsection{User Support Tools}
-Table \ref{tab_user_support_model} shows the user support tools by the number of projects using them. The two most popular tools are GitHub issue and FAQ.
-
-\subsection{Version Control Tools}
-Briefly speaking, 27 of the 29 projects chose git as the version control tool, and 24 used GitHub to host the repositories. Section \ref{sec_score_maintainability} provides more details.
-
-\subsection{Documentation Tools}
-Table \ref{tab_doc_tools} lists the documentation tools by the number of answers from the interviewees. The two most popular tools are forum discussions and videos.
-
-\subsection{Contribution Management Tools}
-Table \ref{tab_method_receive_contributions} shows the ways of receiving contributions by the number of answers from the interviewees.
-All eight development teams were using GitHub with pull requests as the tool.
-The most popular tool in the past was email.
-
-\subsection{Project Management Tools}
-Table \ref{tab_pm_tools} shows the tools for project management by the number of answers from the interviewees. GitHub was the top-chosen tool.
 
@@
 This section shows the principles, processes, and methodologies in software testing, documentation, contribution management, project management, and improving software qualities.
+Table \ref{tab_reference_rq3} shows the sections and tables containing answers to this research question.
 
-\subsection{Software Testing}
-As shown in Section \ref{fg_correctness_erifiability_scores}, we identified the use of unit testing in less than half of the 29 projects.
+\begin{table}[H]
+\centering
+\begin{tabular}{ll}
+\hline
+Section or table & Description \\ \hline
+Section \ref{fg_correctness_erifiability_scores}, \ref{sec_pain_points_3}, \ref{sec_interview_correctness}, \ref{sec_interview_usability}, and \ref{sec_interview_reproducibility}; Table \ref{tab_q15_threats_correctness} and \ref{tab_q15_strategies_correctness}.
+& Software testing \\
+Section \ref{fg_correctness_erifiability_scores} and \ref{sec_interview_documents}; Table \ref{tab_maintainability_docs}, \ref{tab_Visibility/Transparency_docs}, and \ref{tab_artifacts_frequency} & Documentation \\
+Section \ref{sec_contribution_pm} & Contribution management \\
+Section \ref{sec_contribution_pm} & Project management \\ \hline
+\end{tabular}
+\caption{\label{tab_reference_rq3}Sections and tables with answers to RQ3}
+\end{table}
 
-On the other hand, the interviewees believed that testing (including usability tests with users) was the top solution to improve \textit{correctness}, \textit{usability}, and \textit{reproducibility}, as shown in Section \ref{sec_interview_correctness}, \ref{sec_interview_usability}, and \ref{sec_interview_reproducibility}.
+We identified the use of unit testing in less than half of the 29 projects.
+On the other hand, the interviewees believed that testing (including usability tests with users) was the top solution to improve \textit{correctness}, \textit{usability}, and \textit{reproducibility}.
+One pain point in the development process is the lack of access to real-world datasets for testing.
+The developers' strategies to address it is in Section \ref{sec_pain_points_3}.
+One threat to \textit{correctness} is: with huge datasets for testing, the tests are expensive and time-consuming.
+Three interviewees endorsed self tests / automated tests, which may save time for testing.
 
-Table \ref{tab_q15_threats_correctness} lists a threat to \textit{correctness}: with huge datasets for testing, the tests are expensive and time-consuming.
-Table \ref{tab_q15_strategies_correctness} shows that three interviewees endorsed self tests / automated tests, which may save time for testing.
-
-Section \ref{sec_pain_points_3} introduces a pain point related to testing, and developers' strategies to address it.
-
-\subsection{Documentation}
-Section \ref{fg_correctness_erifiability_scores} shows that all projects did not have theory manuals.
+All 29 projects did not have theory manuals.
 We identified a road map in the \textit{3D Slicer} project, and no requirements specifications for the rest.
-Table \ref{tab_maintainability_docs} presents which projects had a project plan, developer’s manual, or API documentation, which implies better \textit{maintainability}.
-Table \ref{tab_Visibility/Transparency_docs} lists the existing documents that can improve \textit{visibility/transparency} of the projects.
-Table \ref{tab_artifacts_frequency} reveals the number of all the artifacts in the 29 projects.
-
 Eight of the nine interviewees thought that documentation was essential to their projects.
@@
 Nearly half of them also believed that the lack of time prevented them from improving the documentation.
-See more details in Section \ref{sec_interview_documents}.
-
-\subsection{Contribution Management}
-According to the interviews, all the eight teams were using GitHub and pull requests to accept new contributions from the community.
-Some of them had relied on emails and forums for the purpose.
-More details are in Section \ref{sec_contribution_pm}.
-
-\subsection{Project Management}
-As presented in Section \ref{sec_contribution_pm}, most teams in the interviews did not have a clearly defined software development model.
-Many of them relied on GitHub, issue/bug trackers to manage their projects.
 
@@
 
-We answer this question primarily by interviewing the developers. This section presents the pain points and the solutions proposed by the developers.
-
-\subsection{Pain Points}
-As described in Section \ref{sec_interview_pain_points}, the top three pain points are as follows,
-\begin{itemize}
-\item the lack of fundings and time;
-\item the difficulty to balance between four factors: cross-platform compatibility, convenience to development \& maintenance, performance, and security;
-\item the lack of access to real-world datasets for testing.
-\end{itemize}
-
-\subsection{Solutions}
-Section \ref{sec_interview_pain_points} lists the details about the potential and proven solutions from the developers in the MI domain.
+We answer this question by interviewing the developers.
+The answers to this question, including the pain points and the solutions proposed by the developers, are in Section \ref{sec_interview_pain_points}.
 
@@
 This section includes our answers from the qualities measurements and interviews with the developers.
+Table \ref{tab_reference_rq4} shows the sections and tables containing answers to this research question.
 
-\subsection{Results by Measuring Software Qualities}
-Section \ref{ch_results} contains our measurement results for \textit{Installability}, \textit{Correctness \& Verifiability}, \textit{Reliability}, \textit{Robustness}, \textit{Usability}, \textit{Maintainability}, \textit{Reusability}, \textit{Understandability}, and \textit{Visibility/Transparency}.
-We applied (surface) measurements to the nine qualities, and hope that the grading scores represent their current status.
-
-\subsection{Results by Interviewing Developers}
-Section \ref{sec_interview_software_qualities} shows findings from the interviews for \textit{correctness}, \textit{maintainability}, \textit{understandability}, \textit{usability}, and \textit{reproducibility}.
-The interviewees expressed their thoughts on these five qualities.
-We discussed the current or past threats, and found out what actions they believed would improve the qualities.
+\begin{table}[H]
+\centering
+\begin{tabular}{ll}
+\hline
+Section or table & Description \\ \hline
+Section \ref{sec_result_installability} & Installability \\
+Section \ref{sec_result_correctness_verifiability} and \ref{sec_interview_correctness} & Correctness \& verifiability \\
+Section \ref{sec_result_reliability} & Reliability \\
+Section \ref{sec_result_robustness} & Robustness \\
+Section \ref{sec_result_usability} and \ref{sec_interview_usability}& Usability \\
+Section \ref{sec_score_maintainability} and \ref{sec_interview_maintainability} & Maintainability \\
+Section \ref{sec_result_reusability} & Reusability \\
+Section \ref{sec_result_understandability} and \ref{sec_interview_understandability} & Understandability \\
+Section \ref{sec_result_visibility_transparency} & Visibility/transparency \\
+Section \ref{sec_interview_reproducibility} & Reproducibility \\ \hline
+\end{tabular}
+\caption{\label{tab_reference_rq4}Sections and tables with answers to RQ4}
+\end{table}
 
@@
 
-Table \ref{tab_ranking_vs_GitHub} shows our ranking to the 29 MI projects, and their GitHub metrics if applicable.
+Table \ref{tab_ranking_vs_GitHub} shows our ranking to the 29 MI projects, and their GitHub metrics if applicable, in the descending order of the average number of new stars.
 As mentioned in Section \ref{sec_score_maintainability}, 24 projects used GitHub.
@@
 Thus, we might underestimate the scores of its \textit{surface reliability} and \textit{surface robustness}.
-We weighted all qualities equally, which is not likely to be the case with all end-users.
-As a result, some projects may have received higher popularity because of their better performance in part of the nine qualities.
+We weighted all qualities equally, which is not likely to be the case with all users.
+As a result, some projects with high popularity may not perform well in all qualities.
 
@@
 We might underestimate their rank due to uncommon technical issues.
-\item Our ranking gave all qualities equal weight, which may not be the case with all end-users. Thus, it may not represent the popularity of software among end-users.
+\item Our ranking gave all qualities equal weight, which may not be the case with all users. Thus, it may not represent the popularity of software among users.
 \item The number of GitHub stars, watches, and forks are not perfect measures of popularity, but they are what we had available.

### 2021-08-28 StateOfPractice/Ao-Report/18Conclusion.tex
MI report - updated ch 6 and more

@@
 	\item we can improve some interview questions.
-For example, in question 14, ``Do you think improving this process can tackle the current problem?" is a yes-or-no question, which is not informative enough.
+For example, in \hyperlink{q14}{Q14}, ``Do you think improving this process can tackle the current problem?" is a yes-or-no question, which is not informative enough.
 As mentioned in Section \ref{sec_contribution_pm}, most interviewees ignored it.
-We can change it to ``By improving this process, what current problems can be tackled?"
+We can change it to ``By improving this process, what current problems can be tackled?" In \hyperlink{q16}{Q16}, we can ask more details about the modular approach, such as "What principles did you use to divide code into modules?
+Can you give an example of using the principles?".
+\hyperlink{q17}{Q17} and  \hyperlink{q18}{Q18} should respectively ask \textit{understandability} to developers and \textit{usability} to end-users.
 	\item we can better organize the interview questions.

### 2021-08-28 StateOfPractice/Ao-Report/17Recommendations.tex
MI report - update ch 7
MI report - updated ch 7

@@
 This section presents our recommendations on SC software development.
-We give these suggestions to the development in all SC domains, unless specifically mentioning that some guidelines are only for MI software.
+In general, our suggestions apply to all SC domains, unless we specifically mention that a particular guideline is only for MI software.
 
@@
 Sections \ref{sec_recommendations_limited_resources}, \ref{sec_recommendations_tech_stack}, and \ref{sec_recommendations_testing_dataset} are based on the primary pain points collected from the developers in the MI domain, but we believe scientists and developers are likely to face them in most SC domains.
-These sections contain our general suggestions to tackle them.
+These sections contain our general suggestions tackling them.
 
@@
 \begin{itemize}
-    \item test-driven development with unit tests, integration tests, nightly tests, etc.
+    \item test-driven development with unit tests, integration tests, and nightly tests;
     \item two stage development process with stable release \& nightly builds;
     \item CI/CD;
-    \item documents of requirements specifications and theory manuals.
+    \item requirements specifications and theory manuals \cite{Smith2016} \cite{SmithAndLai2005}.
+    \item static code analysis tools (e.g. Lint and SonarQube)
 \end{itemize}
@@
 \begin{itemize}
-    \item test-driven development with unit tests, integration tests, nightly tests, etc.
+    \item test-driven development with unit tests, integration tests, and nightly tests.
     \item two stage development process with stable release \& nightly builds;
@@
 \begin{itemize}
-    \item designing with exceptions and make the software failures elegant;
+    \item designing with exceptions and make the software failures graceful;
     \item descriptive error messages.
@@
     \item using GitHub;
-    \item modular approach;
-    \item documentation for developers, such as project plan, developer’s manual, and API documentation.
+    \item modular approach with the design principle proposed by Parnas: ``system details that are likely to change independently should be the secrets of separate modules; the only assumptions that should appear in the interfaces between modules are those that are considered unlikely to change." \cite{ParnasEtAl2000}
+    \item documentation for developers: project plan, developer’s manual, and API documentation.
 \end{itemize}
@@
     \item modular approach;
-    \item API documentation.
+    \item API documentation;
+    \item tools that generate software documentation for developers (e.g. Doxygen, Javadoc, and Sphinx).
 \end{itemize}
@@
     \item modular approach;
-    \item consistent codding style;
-    \item clear comments;
+    \item good coding style: consistent indentation and formatting style; consistent, distinctive, and meaningful code identifiers; keeping parameters in the same order for all functions; avoiding hard-coded constants (other than 0 and 1);
+    \item clear comments, indicating what is being done, not how;
     \item description of used algorithms;
-    \item documentation;
-    \item communication between developers and users via GitHub issues, mailing lists, forums etc.
+    \item documentation of explicit requirements on coding standard;
+    \item communication between developers and users via GitHub issues, mailing lists, and forums.
     \item graphical user interface.
@@
 \begin{itemize}
-    \item test-driven development with unit tests, integration tests, nightly tests, etc.
+    \item test-driven development with unit tests, integration tests, and nightly tests.
     \item open-source;
@@
 
-We summarize our discussion with the MI software developers in Section \ref{sec_pain_points_1} to the following recommendations,
+We summarize our discussion with the MI software developers in Section \ref{sec_pain_points_1} with the following recommendations,
 \begin{itemize}
@@
 \item \textbf{Maintain a good documentation.} Creating and updating documentation consumes time, but can save much more time in the long term.
-If the users and developers can find answers to their questions by themselves, they are less likely to abuse the team's issue tracker.
+If the users and developers can find answers to their questions themselves, they are less likely to abuse the team's issue tracker.
 
@@
 
-\item \textbf{Use test-driven development process.} Many people think writing test cases is less fun than building the functional code, but this is only true before the bugs show.
-Identifying and fixing bugs can cause substantial resources.
+\item \textbf{Use test-driven development process.} Many people think writing test cases is less fun than building the functional code, but this is only true before we encounter the bugs.
+Identifying and fixing bugs can consume substantial resources.
 Setting up the test cases costs time, but generates more benefits in the long run.
@@
 \item \textbf{Consider supporting third-party plugins or extensions.} Why not let users share the burden?
-No software product can deliver every user's needs, and the large quantity of features leads to more bugs and difficulty to maintain.
+No software product can deliver every user's needs, and the large quantity of features leads to more bugs and maintenance problems.
 So it may be a good idea to shift some development and maintenance responsibilities to the users.
@@
 \item \textbf{Consider ``hibernating" for a while.} When developers are not enough, the team can shift from development mode toward maintenance mode for some time.
-Stop building new features, and fix bugs and design problems in the past.
-The software qualities may also improve as a result.
+Stop building new features, and instead fix bugs and design problems from the past.
+If the development team can repay some of its technical debt, the software qualities may improve as a result.
 
 \item \textbf{Commercialization is not always toxic.} Licensing the software to commercial companies to use as internal modules of their products may bring financial supports to the team.
-Meanwhile, the project can stay open-source to the community.
+Meanwhile, the project can stay open-source for the community.
 \end{itemize}
@@
 However, new technologies may ease this difference.
-For example, some JavaScript libraries can help the frontend with harnessing the power of computer GPU and accelerate graphical computing.
+For example, some JavaScript libraries can help the frontend harness the power of computer GPU and accelerate graphical computing.
 In addition, there are new frameworks helping developers with cross-platform \textit{compatibility}.
@@
 
-\item \textbf{Maintain a good documentation.} No matter what tech stack a team uses, good documents such as well-maintained project plan, developer's manual, and API documentation always help team members to contribute more and make fewer mistakes.
+\item \textbf{Maintain a good documentation.} No matter what tech stack a team uses, a well-maintained project plan, developer's manual, and API documentation always help team members to contribute more and make fewer mistakes.
 \end{itemize}
@@
 \item \textbf{Collect and maintain datasets over time.} A team may face all kinds of strange problems caused by various unique inputs over the years of development.
-It is worth collecting and maintaining these data, which can form a good dataset for testing.
+It is worth collecting and maintaining this data, which can form a good dataset for testing.
+
+\item \textbf{Search for open data sources.} In general, there are many open datasets in different SC domains.
+Take MI as an example, there are Chest X-ray Datasets by National Institute of Health (\hyperlink{https://nihcc.app.box.com/v/ChestXray-NIHCC}{https://nihcc.app.box.com/v/ChestXray-NIHCC}) \cite{WangEtAl2017}, Cancer Imaging Archive (\hyperlink{https://www.cancerimagingarchive.net/}{https://www.cancerimagingarchive.net/}) \cite{PriorEtAl2017}, and MedPix by National Library of Medicine (\hyperlink{https://medpix.nlm.nih.gov/home}{https://medpix.nlm.nih.gov/home}) \cite{Smirniotopoulos2014}.
+A team developing MI software should be able to find more open datasets according to their needs.
+
+\item \textbf{Create sample data for testing.} If a team can access tools creating sample data, they may also self-build datasets for testing.
+For example, an MI software development team can use an MRI scanner to create images of objects, animals, and volunteers.
+The team can build the images based on specific testing requirements.
+
+\item \textbf{Remove privacy from sensitive data.} For data with sensitive information, a team can ask the data owner to remove such information or add noise to protect privacy.
+One example is using deidentified copies of medical images for testing.
 
@@
 However, we believe that broader collaboration between development teams can address this problem better.
-Some datasets are too sensitive to share, but if the community has some kind of ``group discussion", teams can better express their needs, and professionals can better offer their voluntary supports for testing.
-Ultimately, perhaps the community can establish a nonprofit organization as a third-party, which maintains large datasets, tests OSS in the domain, and protects privacy.
+Some datasets are too sensitive to share, but if the community has some kind of ``group discussion", teams can better express their needs, and professionals can better offer voluntary support for testing.
+Ultimately, the community can establish a nonprofit organization as a third-party, which maintains large datasets, tests OSS in the domain, and protects privacy.
 \end{itemize}

### 2021-08-28 StateOfPractice/Ao-Report/11Introduction.tex
MI report - pointed answers to research questions

@@
 \begin{description}
-\item[RQ1.] What artifacts are present in current software projects? What role does documentation play in the projects? What are the developers' attitude toward it?
-\item[RQ2.] What tools are used in the development of current software packages?
-\item[RQ3.] What principles, processes, and methodologies are used in the development of current software packages?
-\item[RQ4.] What are the pain points for developers working on research software projects?
+\item[RQ1.] \hypertarget{rq1}What artifacts are present in current software projects? What role does documentation play in the projects? What are the developers' attitude toward it?
+\item[RQ2.] \hypertarget{rq2}What tools are used in the development of current software packages?
+\item[RQ3.] \hypertarget{rq3}What principles, processes, and methodologies are used in the development of current software packages?
+\item[RQ4.] \hypertarget{rq4}What are the pain points for developers working on research software projects?
 What aspects of the existing processes, methodologies, and tools do they consider as potentially needing improvement?
 What changes to processes, methodologies, and tools can improve software development and software quality?
-\item[RQ5.] What is the current status of the following software qualities for the projects? What actions have the developers taken to address them?
+\item[RQ5.] \hypertarget{rq5}What is the current status of the following software qualities for the projects? What actions have the developers taken to address them?
 \begin{itemize}
@@
 \end{itemize}
-\item[RQ6.] How does software designated as high quality by this methodology compare with top-rated software by the community?
+\item[RQ6.] \hypertarget{rq6}How does software designated as high quality by this methodology compare with top-rated software by the community?
 \end{description}

### 2021-08-28 StateOfPractice/Ao-Report/14MeasurementResults.tex
MI report - pointed answers to research questions

@@
 The detailed data can be found in the repository \hyperlink{https://data.mendeley.com/datasets/k3pcdvdzj2/1}{https://data.mendeley.com/datasets/k3pcdvdzj2/1}.
+This section contains part of the answers to \hyperlink{rq5}{RQ5}.
 
@@
 
+In this section, the information about CI/CD tools is part of the answers to \hyperlink{rq2}{RQ2}, and the information about software testing and documentation is part of the answers to \hyperlink{rq3}{RQ3}.
+
 \section{Surface Reliability}
@@
 Table \ref{tab_user_support_model} shows the user support models by the number of projects using them.
+This table contains part of the answers to \hyperlink{rq2}{RQ2}.
 Maybe not every team intended to use GitHub issues to answer users' questions, but many users use them to seek help.
@@
 Table \ref{tab_maintainability_docs} shows which projects had these documents, in the descending order of their \textit{maintainability} scores.
+This table contains part of the answers to \hyperlink{rq1}{RQ1} and \hyperlink{rq3}{RQ3}.
 
@@
 }, \textit{AMIDE}, and \textit{Gwyddion} used SourceForge; \textit{DicomBrowser} and \textit{3DimViewer} used BitBucket.
+The information about version control tools is part of the answers to \hyperlink{rq2}{RQ2}.
 
@@
 Table \ref{tab_Visibility/Transparency_docs} shows the projects which had documents for the development process, project status, development environment, and release notes, in the descending order of their \textit{visibility/transparency} scores.
+This table contains part of the answers to \hyperlink{rq1}{RQ1} and \hyperlink{rq3}{RQ3}.
 

### 2021-08-28 StateOfPractice/Ao-Report/15InterviewsWithDevelopers.tex
MI report - updated ch 6 and more
MI report - pointed answers to research questions

@@
 We also asked the interviewees how they would solve the problems.
+This section contains the answers to \hyperlink{rq4}{RQ4}.
 Questions 9, 10, and 12 are as follows:
@@
 \label{sec_pain_points_3}
-The pain point in the \textit{testing} category is \textbf{the lack of access to real-world datasets for testing.}
+The pain point in the \textit{testing} category is \textbf{the lack of access to real-world datasets for testing.} The information about software testing in this section is part of the answers to \hyperlink{rq3}{RQ3}.
 
@@
 \label{sec_interview_documents}
-We tried to understand the interviewees' opinions on documentation and the quality of documentations with questions 11 and 19:
+We tried to understand the interviewees' opinions on documentation and the quality of documentations with questions 11 and 19.
+The information about documentation is part of the answers to \hyperlink{rq3}{RQ3}.
 
@@
 
-Table \ref{tab_doc_tools} lists some of the documentation tools and methods mentioned by the interviewees.
+Table \ref{tab_doc_tools} lists some of the documentation tools and methods mentioned by the interviewees. This table contains part of the answers to \hyperlink{rq2}{RQ2}.
 
@@
 Table \ref{tab_method_receive_contributions} shows the number of times the interviewees mentioned the methods of receiving contributions.
+This table contains part of the answers to \hyperlink{rq2}{RQ2}.
 
@@
 
-Some interviewees mentioned the project management tools they used, which are in Table \ref{tab_pm_tools}. Generally speaking, they talked about two types:
+Some interviewees mentioned the project management tools they used, which are in Table \ref{tab_pm_tools}.
+This table contains part of the answers to \hyperlink{rq2}{RQ2}.
+Generally speaking, the interviewees talked about two types:
 
@@
 
+The information about contribution management and project management in this section is part of the answers to \hyperlink{rq3}{RQ3}.
+
 \section{Discussions on Software Qualities}
@@
 We discuss each quality in a separate section below.
+This section contains part of the answers to \hyperlink{rq5}{RQ5}.
 
@@
 
+The information about software testing in this section is part of the answers to \hyperlink{rq3}{RQ3}.
+
 \subsection{Maintainability}
@@
 \subsection{Understandability}
-
+\label{sec_interview_understandability}
 \begin{description}
@@
 Table \ref{tab_q18_strategies_usability} shows the strategies to ensure \textit{usability} by the numbers of interviewees with the answers.
+The information about software testing in this table is part of the answers to \hyperlink{rq3}{RQ3}.
 
@@
 \end{table}
+
+The information about software testing in this section is part of the answers to \hyperlink{rq3}{RQ3}.

### 2021-08-28 StateOfPractice/Ao-Report/16AnswersToResearchQuestions.tex
MI report - pointed answers to research questions

@@
 Table \ref{tab_artifacts_frequency} summarizes the frequency of the artifacts.
+This table also contains part of the answers to \hyperlink{rq3}{RQ3}.
 
@@
 Section or table & Description \\ \hline
-Section \ref{fg_correctness_erifiability_scores}, \ref{sec_pain_points_3}, \ref{sec_interview_correctness}, \ref{sec_interview_usability}, and \ref{sec_interview_reproducibility}; Table \ref{tab_q15_threats_correctness} and \ref{tab_q15_strategies_correctness}.
-& Software testing \\
+Section \ref{fg_correctness_erifiability_scores}, \ref{sec_pain_points_3}, \ref{sec_interview_correctness}, and \ref{sec_interview_reproducibility}; Table \ref{tab_q18_strategies_usability} & Software testing \\
 Section \ref{fg_correctness_erifiability_scores} and \ref{sec_interview_documents}; Table \ref{tab_maintainability_docs}, \ref{tab_Visibility/Transparency_docs}, and \ref{tab_artifacts_frequency} & Documentation \\

### 2021-08-29 StateOfPractice/Ao-Report/18Conclusion.tex
MI report - updated ch 8 and more

--- 2021-08-10 StateOfPractice/Ao-Report/18Conclusion.tex
@@
 We analyzed the state of the practice for SC software in the MI domain.
-To better achieve our goals in Section \ref{ch_intro}, we proposed six research questions, that we list in Section \ref{sec_research_questions}.
+To better achieve our goals in Section \ref{ch_intro}, we proposed six research questions in Section \ref{sec_research_questions}.
 
@@
 
-With the measurement and interview results in Section \ref{ch_results} and \ref{ch_interview}, we reveal some current status of SC software development and qualities in the MI domain.
-The interview results also show some merits, drawbacks, and pain points within the development process.
-With these findings, we answer our six research questions in Section \ref{ch_answers}.
-In addition, Section \ref{ch_recommendations} presents our recommendations on SC software development, such as the ones on improving software qualities and easing pain points during development.
+We answered our six research questions in Section \ref{ch_answers}. In addition, Section \ref{ch_recommendations} presents our recommendations on SC software development.
+
+\section{Key Findings}
+
+With the measurement results in Section \ref{ch_results}, we revealed some current status of SC software development and qualities in the MI domain.
+We ranked the 29 software projects in nine qualities based on the grading scores.
+\textit{3D Slicer}, \textit{ImageJ}, and \textit{OHIF Viewer} are the top three software by their overall scores.
+
+The interview results in Section \ref{ch_interview} show some merits, drawbacks, and pain points within the development process. The three primary categories of pain points are:
+\begin{itemize}
+\item the lack of fundings and time;
+\item the difficulty to balance between four factors: cross-platform compatibility, convenience to development \& maintenance, performance, and security;
+\item the lack of access to real-world datasets for testing.
+\end{itemize}
+We summarized the solutions from the developers to address these problems.
+We also collected the status of software testing, documentation, contribution management, and project management in the eight projects.
+
+Our answers to the research questions (Section \ref{ch_answers}) are based on the above findings.
+We identified the existing artifacts, tools, principles, processes, and methodologies in the 29 projects.
+By comparisons in Section \ref{sec_rq_comparison}, we found out: 1) four of the top five software projects in our ranking were also among the top five ones receiving the most GitHub stars per year (Table \ref{tab_ranking_vs_GitHub}); 2) three of the top four in our ranking were among the top four provided by the domain experts (Table \ref{tab_top_software_vs_experts}).
+
+Section \ref{ch_recommendations} presents our recommendations on improving software qualities and easing pain points during development. Some highlighted ones are:
+\begin{itemize}
+\item adopting test-driven development with unit tests, integration tests, and nightly tests;
+\item maintaining good documentation (e.g., installation instructions, requirements specifications, theory manuals, getting started tutorials, user manuals, project plan, developer’s manual, API documentation, requirements on coding standards, development process, project status, development environment, and release notes);
+\item using CI/CD;
+\item using git and GitHub;
+\item modular approach with the design principle proposed by Parnas \cite{ParnasEtAl2000};
+\item considering newer technologies (e.g., web application and serverless solution);
+\item various ways of enriching the testing datasets in Section \ref{sec_recommendations_testing_dataset}.
+\end{itemize}
+
+\section{Future Works}
 
@@
 \begin{itemize}
+    \item we can make the surface measurements less shallow. For example:
+    \begin{itemize}
+        \item \textit{surface reliability}: our current measurement relies on the processes of installation and getting started tutorials.
+However, not all software needs installation or has a getting started tutorial.
+We can design a list of operation steps, perform the same operations with each software, and record any errors.
+        \item \textit{surface robustness}: we used damaged images as inputs for this measuring MI software.
+This process is similar to fuzz testing \cite{enwiki:1039424308}, which is one type of fault injection \cite{enwiki:1039005082}.
+We may adopt more fault injection methods, and identify tools and libraries to automate this process.
+        \item \textit{surface usability}: we can design usability tests and test all software projects with end-users. The end-users can be volunteers and domain experts.
+        \item \textit{surface understandability}: our current method does not require understanding the source code.
+As software engineers, perhaps we can select a small module of each project, read the source code and documentation, try to understand the logic, and score the ease of the process.
+    \end{itemize}
 	\item we can further automate the measurements on the grading temple in Appendix \ref{ap_grading_template}.
@@
 	\item the grading standard can be more explicit. For example, we can explicitly define scores for each item in the grading temple.
-	\item we can improve some interview questions.
-For example, in question 14, ``Do you think improving this process can tackle the current problem?" is a yes-or-no question, which is not informative enough.
+	\item we can improve some interview questions. Some examples are:
+	\begin{itemize}
+	    \item in \hyperlink{q14}{Q14}, ``Do you think improving this process can tackle the current problem?" is a yes-or-no question, which is not informative enough.
 As mentioned in Section \ref{sec_contribution_pm}, most interviewees ignored it.
-We can change it to ``By improving this process, what current problems can be tackled?"
+We can change it to ``By improving this process, what current problems can be tackled?";
+	    \item in \hyperlink{q16}{Q16}, we can ask more details about the modular approach, such as "What principles did you use to divide code into modules?
+Can you give an example of using the principles?";
+	    \item \hyperlink{q17}{Q17} and  \hyperlink{q18}{Q18} should respectively ask \textit{understandability} to developers and \textit{usability} to end-users.
+	\end{itemize}
 	\item we can better organize the interview questions.
-Since we use audio conversion tools to transcribe the answers, we should aim to make the transcription easier to read.
+Since we use audio conversion tools to transcribe the answers, we should make the transcription easier to read.
 For example, we can order them together for questions about the five software qualities and compose a similar structure for each.
@@
 \end{itemize}
+
+\noindent In addition, we propose a few SC domains that are potentially suitable for future works:
+\begin{itemize}
+    \item Metallurgy
+    \item Quantitative Finance
+    \item Computational Fluid Dynamics
+    \item Basic Linear Algebra
+    \item Finite Elements
+    \item Sparse Linear Solvers
+\end{itemize}
+
+\noindent After applying our method on various domains, we can start a meta-study to compare the state of the practice for software in different domains.

### 2021-08-30 StateOfPractice/Ao-Report/12Background.tex
MI report - updated ch 2

@@
 The software are evaluated with the grading template in Appendix \ref{ap_grading_template} and a subjective score from one to ten is given for each quality for each package.
-For each quality, for a pair of packages $i$ and $j$, such that $score_i >= score_j$, the difference between two scores is $diff_{ij} = score_i - score_j$.
-The relationship between $A_{ij} = 1$ and $diff_{ij}$ is as follows:
+For each quality, for a pair of packages $i$ and $j$, such that $score_i >= score_j$, the difference between two scores is $\mathit{diff_{ij}} = score_i - score_j$.
+The relationship between $A_{ij} = 1$ and $\mathit{diff_{ij}}$ is as follows:
 \begin{itemize}
-\item $A_{ij} = 1$ and $diff_{ij} = 0$ when criterion $i$ and criterion $j$ are equally important;
-\item $A_{ij}$ increases when $diff_{ij}$ increases;
-\item $A_{ij} = 9$ and $diff_{ij} = 9$ when criterion $i$ is extremely more important than criterion $j$.
+\item $A_{ij} = 1$ and $\mathit{diff_{ij}} = 0$ when criterion $i$ and criterion $j$ are equally important;
+\item $A_{ij}$ increases when $\mathit{diff_{ij}}$ increases;
+\item $A_{ij} = 9$ and $\mathit{diff_{ij}} = 9$ when criterion $i$ is extremely more important than criterion $j$.
 \end{itemize}

### 2021-09-02 StateOfPractice/Ao-Report/13Methods.tex
MI report - updates

@@
 \subsection{Grading Software}
+\label{sec_applying_method_grading}
 Then we followed the steps in Section \ref{sec_grading_software} to measure and grade the software.
@@
 Going through the interview process in Section \ref{sec_interview_methods}, we contacted all of the 29 teams.
-As a result, developers/architects from 8 teams have participated in our interviews so far.
+Members from eight teams responded and agreed to participate.
+As a result, we interviewed nine developers and architects from the eight teams.

### 2021-09-02 StateOfPractice/Ao-Report/17Recommendations.tex
MI report - updates

@@
 \begin{itemize}
-\item \textbf{Installability}
+\item \textbf{Installability} (Section \ref{sec_result_installability})
 \begin{itemize}
@@
 \end{itemize}
-\item \textbf{Correctness \& Verifiability}
+\item \textbf{Correctness \& Verifiability} (Section \ref{sec_result_correctness_verifiability} and \ref{sec_interview_correctness})
 \begin{itemize}
@@
 \end{itemize}
-\item \textbf{Reliability}
+\item \textbf{Reliability} (Section \ref{sec_result_reliability})
 \begin{itemize}
@@
 \end{itemize}
-\item \textbf{Robustness}
+\item \textbf{Robustness} (Section \ref{sec_result_robustness})
 \begin{itemize}
@@
 \end{itemize}
-\item \textbf{Usability}
+\item \textbf{Usability} (Section \ref{sec_result_usability} and \ref{sec_interview_usability})
 \begin{itemize}
@@
 \end{itemize}
-\item \textbf{Maintainability}
+\item \textbf{Maintainability} (Section \ref{sec_score_maintainability} and \ref{sec_interview_maintainability})
 \begin{itemize}
@@
 \end{itemize}
-\item \textbf{Reusability}
+\item \textbf{Reusability} (Section \ref{sec_result_reusability})
 \begin{itemize}
@@
 \end{itemize}
-\item \textbf{Understandability}
+\item \textbf{Understandability} (Section \ref{sec_result_understandability} and \ref{sec_interview_understandability})
 \begin{itemize}
@@
 \end{itemize}
-\item \textbf{Visibility/Transparency}
+\item \textbf{Visibility/Transparency} (Section \ref{sec_result_visibility_transparency})
 \begin{itemize}
@@
 \end{itemize}
-\item \textbf{Reproducibility}
+\item \textbf{Reproducibility} (Section \ref{sec_interview_reproducibility})
 \begin{itemize}

### 2021-09-07 StateOfPractice/Ao-Report/16AnswersToResearchQuestions.tex
MI report - updated ch 8 and more
MI report - updated ch 6

@@
 Artifact & Number of projects \\ \hline
-Source code & 29 \\
+README & 29 \\
 Version Control & 29 \\
-README & 29 \\
 License & 28 \\
@@
 Tutorials & 18 \\
-Makefile & 18 \\
 Installation Guide & 16 \\
@@
 \end{table}
+
+We summarized the definitions of the artifacts in \hyperlink{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Artifacts_MiningV3.xlsx}{https://github.com/.../Artifacts\_MiningV3.xlsx}.
+Source code is a type of artifact.
+Since we only included OSS on the final list (Section \ref{sec_mi_software_selection}), every project's source code was available.
+Thus, we excluded it in the above table.
 
@@
 
-Table \ref{tab_ranking_vs_GitHub} shows our ranking to the 29 MI projects, and their GitHub metrics if applicable, in the descending order of the average number of new stars.
+Table \ref{tab_ranking_vs_GitHub} shows our ranking to the 29 MI projects, and their GitHub metrics if applicable.
 As mentioned in Section \ref{sec_score_maintainability}, 24 projects used GitHub.
@@
 
+In Table \ref{tab_ranking_vs_GitHub}, we used the average number of new stars per 12 months as an indicator of the GitHub popularity, and listed the items in the descending order of this number.
+We ordered the non-GitHub items by our ranking.
 Generally speaking, most of the top-ranking MI software projects also received greater attention and popularity on GitHub.
+Between our ranking and the GitHub stars-per-year ranking, four of the top five software projects are the same.
+
 Project \textit{dwv} was popular on GitHub, but we ranked it low.
@@
 Software & Our ranking & Stars/yr & Watches/yr & Forks/yr \\ \hline
-3D Slicer & 1 & 284.25 & 18.75 & 128.25 \\
-OHIF Viewer & 3 & 277.04 & 19.3 & 223.83 \\
-dwv & 23 & 123.56 & 11.59 & 51.36 \\
-ImageJ & 2 & 83.58 & 9.37 & 30 \\
-ParaView & 5 & 66.76 & 7.11 & 28.36 \\
-Horos & 12 & 48.62 & 9.23 & 18 \\
-Papaya & 17 & 44.7 & 4.84 & 19.71 \\
-Fiji & 4 & 44.2 & 4.98 & 20.68 \\
-DICOM Viewer & 29 & 42.86 & 5.71 & 8.57 \\
-INVESALIUS 3 & 10 & 39.66 & 4.11 & 17.37 \\
-Weasis & 6 & 36 & 5.1 & 18.9 \\
-dicompyler & 26 & 35.18 & 4.77 & 13.64 \\
-OsiriX Lite & 9 & 34.14 & 8.62 & 23.77 \\
-MRIcroGL & 18 & 23.63 & 3.38 & 3.38 \\
-GATE & 25 & 18.9 & 5.79 & 25.79 \\
-Ginkgo CADx & 14 & 18.83 & 4.43 & 5.72 \\
-BioImage Suite Web & 8 & 17.85 & 4.62 & 6.77 \\
-Drishti & 27 & 16.04 & 3.78 & 4.43 \\
-Slice:Drop & 20 & 9.73 & 2.16 & 4.97 \\
-ITK-SNAP & 15 & 9.1 & 0.97 & 3.59 \\
-medInria & 7 & 7.04 & 3.35 & 6.35 \\
-SMILI & 13 & 3.04 & 0.91 & 1.52 \\
-MatrixUser & 28 & 2 & 0.33 & 0.33 \\
-MicroView & 16 & 1.48 & 0.82 & 0.82 \\
+3D Slicer & 1 & 284 & 19 & 128 \\
+OHIF Viewer & 3 & 277 & 19 & 224 \\
+dwv & 23 & 124 & 12 & 51 \\
+ImageJ & 2 & 84 & 9 & 30 \\
+ParaView & 5 & 67 & 7 & 28 \\
+Horos & 12 & 49 & 9 & 18 \\
+Papaya & 17 & 45 & 5 & 20 \\
+Fiji & 4 & 44 & 5 & 21 \\
+DICOM Viewer & 29 & 43 & 6 & 9 \\
+INVESALIUS 3 & 10 & 40 & 4 & 17 \\
+Weasis & 6 & 36 & 5 & 19 \\
+dicompyler & 26 & 35 & 5 & 14 \\
+OsiriX Lite & 9 & 34 & 9 & 24 \\
+MRIcroGL & 18 & 24 & 3 & 3 \\
+GATE & 25 & 19 & 6 & 26 \\
+Ginkgo CADx & 14 & 19 & 4 & 6 \\
+BioImage Suite Web & 8 & 18 & 5 & 7 \\
+Drishti & 27 & 16 & 4 & 4 \\
+Slice:Drop & 20 & 10 & 2 & 5 \\
+ITK-SNAP & 15 & 9 & 1 & 4 \\
+medInria & 7 & 7 & 3 & 6 \\
+SMILI & 13 & 3 & 1 & 2 \\
+MatrixUser & 28 & 2 & 0 & 0 \\
+MicroView & 16 & 1 & 1 & 1 \\
 Gwyddion & 11 & n/a & n/a & n/a \\
@@
 As shown in Section \ref{sec_mi_software_selection}, our domain experts recommended a list of top software with 12 software products.
-Table \ref{tab_top_software_vs_experts} compares the top 12 software projects ranked by our methodology with the ones from the domain experts.
-
-An example of interpreting Table \ref{tab_top_software_vs_experts} is as follows: in row 2, \textit{ImageJ} has rank 2 by our assessment, and the number \textit{(3)} means that it has rank 3 by the experts; likewise, \textit{Horos} has rank 2 by the domain experts, and rank 12 by our assessment.
+Table \ref{tab_top_software_vs_experts} and \ref{tab_experts_vs_top_software} compare the top 12 software projects ranked by our methodology with the ones from the domain experts.
 
@@
 \hline
-Rank & This assessment & Domain expert \\ \hline
-1 & 3D Slicer (1) & 3D Slicer (1) \\
-2 & ImageJ (3) & Horos (12) \\
-3 & OHIF Viewer (n/a) & ImageJ (2) \\
-4 & Fiji (4) & Fiji (4) \\
-5 & ParaView (n/a) & AFNI (n/a) \\
-6 & Weasis (n/a) & FSL (n/a) \\
-7 & medInria (n/a) & Freesurfer (n/a) \\
-8 & BioImage Suite Web (n/a) & Mricron (18*) \\
-9 & OsiriX Lite (n/a) & Mango (17**) \\
-10 & INVESALIUS 3 (n/a) & Tarquin (n/a) \\
-11 & Gwyddion (n/a) & Diffusion Toolkit (n/a) \\
-12 & Horos (2) & MRItrix (n/a) \\ \hline
+Our ranking & Assessed software & Domain experts \\ \hline
+1 & 3D Slicer & 1 \\
+2 & ImageJ & 3 \\
+3 & OHIF Viewer & n/a \\
+4 & Fiji & 4 \\
+5 & ParaView & n/a \\
+6 & Weasis & n/a \\
+7 & medInria & n/a \\
+8 & BioImage Suite Web & n/a \\
+9 & OsiriX Lite & n/a \\
+10 & INVESALIUS 3 & n/a \\
+11 & Gwyddion & n/a \\
+12 & Horos & 2 \\ \hline
 \end{tabular}
-\caption{\label{tab_top_software_vs_experts}Top software versus domain experts' recommendation}
+\caption{\label{tab_top_software_vs_experts}Top software by our ranking versus domain experts' recommendation}
+\end{table}
+
+\begin{table}[H]
+\centering
+\begin{tabular}{lll}
+\hline
+Our ranking & Recommended software & Domain experts \\ \hline
+1 & 3D Slicer & 1 \\
+12 & Horos & 2 \\
+2 & ImageJ & 3 \\
+4 & Fiji & 4 \\
+n/a & AFNI & 5 \\
+n/a & FSL & 6 \\
+n/a & Freesurfer & 7 \\
+18* & Mricron & 8 \\
+17** & Mango & 9 \\
+n/a & Tarquin & 10 \\
+n/a & Diffusion Toolkit & 11 \\
+n/a & MRItrix & 12 \\ \hline
+\end{tabular}
+\caption{\label{tab_experts_vs_top_software}Domain experts' recommendation versus our ranking}
 \end{table}

### 2021-09-13 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
Some edits to the introduction of the quality definitions paper

@@
 colorlinks=true,        % false: boxed links; true: colored links
-linkcolor=red,          % color of internal links (change box color with
-%linkbordercolor)
+linkcolor=red,          % color of internal links (change box color with linkbordercolor)
 citecolor=blue,         % color of links to bibliography
@@
 
-Purpose and scope of the document.  \wss{Needs to be filled in.  Should
-	reference the overall research proposal, and the ``state of the practice''
-	exercise in particular.}
-
-The presentation is divided into two main sections: i) qualities that apply to
+Purpose and scope of the document.
+
+\wss{ Needs to be filled in.  Should reference the overall research proposal,
+	and the ``state of the practice'' exercise in particular.}
+
+Given that our research focuses on improving the quality of software, we often
+find it necessary to have clear definitions of the different software qualities,
+like maintainability, usability, sustainability, etc. Unfortunately, the
+definitions that we find in the literature are not always satisfying.  They are
+not always adequate for unambiguous communication, or they simply are	not useful
+for providing guidance on what software decisions will lead to quality
+improvement. Moreover, methods for measuring qualities based on the definitions
+are not always clear.  Therefore, we have endeavored to build a lis of
+definitions that can aid and guide our research. We are publishing our
+collection, since others may likewise find them valuable.
+
+Methodology:
+\begin{itemize}
+  \item Collect the list of qualities to define, based on the qualities that are
+  listed in software engineering texts, such as Ghezzi Et al and VanHliet.
+  \item For each quality find as many definitions as possible
+  \item Based on the existing definitions, propose a high quality definition of
+  each quality.  The proposed definition may use one of the existing, or combine
+  several of the options.  We may also make other modifications.  The goal is
+  for the proposed definition to address our acceptance criteria.
+\end{itemize}
+
 software products, software artifacts and software development processes, and
@@
 \begin{description}
+  \item[Unambiguous]
   \item[Consistency] The final list of definitions should be consistent in the
@@
 \end{description}
+
+Not necessary to have every quality measurable.  Analogy to the definition of
+true error in scientific computing.
+
+\subsection{Background}
+
+Quality definitions history: McCall, Ghezzi et al, VanHleit, Boehm, ISO standards.
 

### 2021-09-20 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
Updates to quality of qualities paper, reformat, start introduction

@@
 \usepackage{amsmath,amsthm}
-
 \usepackage{hyperref}
@@
 
+\begin{CCSXML}
+<ccs2012>
+<concept>
+<concept_id>10011007</concept_id>
+<concept_desc>Software and its engineering</concept_desc>
+<concept_significance>500</concept_significance>
+</concept>
+</ccs2012>
+\end{CCSXML}
+
+\ccsdesc[500]{Software and its engineering}
+
+\keywords{software qualities, installability, correctness, verfiability,
+  validatability, testability, reliability, robustness, performance,
+  usability, maintainability, reusability, portability, understandability,
+  interoperability, visibility/transparency, reproducibility, productivity,
+  sustainability, completness, consistency, modifiability, traceability,
+  unambiguity, verifiability, abstract}
+
 \begin{document}
@@
 \begin{abstract}
-	...
+	Existing definitions of software qualities (also called quality
+  characteristics) are collected and analyzed.  For each quality a
+  ``best'\rq{} definition is recommended based on the following criteria:
+  unambiguity, simplicity, consistency, ...
 \end{abstract}
@@
 
-Purpose and scope of the document.
-
-\wss{ Needs to be filled in.  Should reference the overall research proposal,
-	and the ``state of the practice'' exercise in particular.}
-
-Given that our research focuses on improving the quality of software, we often
-find it necessary to have clear definitions of the different software qualities,
-like maintainability, usability, sustainability, etc. Unfortunately, the
-definitions that we find in the literature are not always satisfying.  They are
-not always adequate for unambiguous communication, or they simply are	not useful
-for providing guidance on what software decisions will lead to quality
-improvement. Moreover, methods for measuring qualities based on the definitions
-are not always clear.  Therefore, we have endeavored to build a lis of
-definitions that can aid and guide our research. We are publishing our
+% Purpose
+
+Given that our research focuses on improving the quality of software, we
+often find it necessary to have clear definitions of the different software
+qualities, like maintainability, usability, sustainability, etc.
+Unfortunately, the definitions that we find in the literature are not always
+satisfying.  This became clear when we recently undertook an exercise for
+measuring the \lq\lq{}state of the practice\rq\rq{} in different software
+domains, including the domains of seismology software \citep{SmithEtAl2018},
+Geographic Information Systems \citep{SmithEtAl2018_arXivGIS} and mesh
+generators \citep{smith2016state}. In trying to determine methods to quickly
+assess quality of large samples of existing software, we found communication
+between ourselves and other challenging.  We craved unambigous and
+standardized definitions to enable methods for measuring qualities.
+Therefore, we have endeavored to collect an exhaustive list of definitions
+and from this build a set of recommended definitions, based on our quality
+criteria for high-quality quality definitions.  We are publishing our
 collection, since others may likewise find them valuable.
 
+% Scope
+
+Not specifically emphasizing scientific computing, but when a specific
+context is needed for a definition, the scientific context will be used.  For
+instance, the ISO definition of accuracy is defined in terms of precision, but
+in scientific computing accuracy and precision are two different concepts.
+
+\wss{List of definitions don't include textbooks that simply repeat
+definitions from the literature.}
+
+% Methodology
 Methodology:
+
 \begin{itemize}
+
   \item Collect the list of qualities to define, based on the qualities that are
-  listed in software engineering texts, such as Ghezzi Et al and VanHliet.
+  listed in software engineering texts, such as Ghezzi Et al and Van Hliet.
+
   \item For each quality find as many definitions as possible
+
   \item Based on the existing definitions, propose a high quality definition of
@@
   for the proposed definition to address our acceptance criteria.
+
 \end{itemize}
@@
 
+\wss{introduce the terminology of quality characteristics, as in ISO9126?}
+
+\wss{tell the reader that the definitions will be sorted in the order of their
+publication.}
+
 In determining our recommended definition for each quality, we used the
@@
 \begin{description}
+
   \item[Unambiguous]
-  \item[Consistency] The final list of definitions should be consistent in the
-  terminology used.  For instance, we do not use synonyms for variety; we use
-  the same word to mean the same concept throughout the final list of
-  definitions.  This is why measurement is given by ``the effort required..''
-  and not the various synonyms for this that are seen in the literature. \wss{We
-  may change this phrase we use for measuring something.  I just wanted to put
-  this in here as a placeholder.}
+
+  \item[Consistency] The final list of definitions should be consistent in
+  the terminology used.  For instance, we do not use synonyms for variety;
+  we use the same word to mean the same concept throughout the final list of
+  definitions.  This is why measurement is given by ``the effort required.''
+  and not the various synonyms for this that are seen in the literature.
+  \wss{We may change this phrase we use for measuring something.  I just
+  wanted to put this in here as a placeholder.}
+
   \item[Generality]
+
   \item[Measurability]
+
   \item[Etc.]
+
 \end{description}
@@
 
+\wss{add a definition related to capability?  The word seems to come up in  many definitions.}
+
 To assess the current state of software development, and to understand how
@@
 we mean by quality.  The concept of quality is decomposed into a set of separate
-components that together make up ``quality''. Unfortunately, these are
-called \emph{qualities}. These are associated to the software product,
-the software artifacts (documentation, test cases, etc) and to the software
-development process itself, and combinations thereof.
+components that together make up ``quality''. Unfortunately, these are called
+\emph{qualities}. These are associated to the software product, the software
+artifacts (documentation, test cases, etc) and to the software development
+process itself, and combinations thereof.
 
@@
 
-\subsection{{Installability} \oo{owner}}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{Installability \oo{owner}}
 
@@
   The capability of the software product to be installed in a specified
-  environment \citep{ISO9126}.
+  environment \citep{ISO9126}. %2001
 \end{defn}
@@
 	be successfully installed and/or uninstalled in a specified environment
-	\citep{ISO/IEC25010}.
+	\citep{ISO/IEC25010}. %2011
 \end{defn}
@@
 Combined definitions \ref{Installability_Selected2} and
-\ref{Installability_Selected1} rephrased:
- Installability refers to the
-effort required for the installation, uninstallation or reinstallation of a
-software or product in a specified environment.
-
-\noindent \textbf{Reasoning}
-
- The definition captures a vital metric of measure for this quality - 
- effort, and considers not just the installation but uninstallation
- and re-installation which are all related to installation activities.
-
-\subsection{{Correctness} \oo{owner}}
+\ref{Installability_Selected1} rephrased: Installability refers to the effort
+required for the installation, uninstallation or reinstallation of a software or
+product in a specified environment.
+
+\wss{Remove the repetition of the word being defined in the definition?  Remove
+  reinstallation?  This quality characteristic is a product quality.  It is an
+  external quality.  Installability should be something that can be measured
+  directly.}
+
+\noindent \textbf{Reasoning}
+
+ The definition captures a vital metric of measure for this quality - effort,
+ and considers not just the installation but uninstallation and re-installation
+ which are all related to installation activities.
+
+ \wss{remove re-installation - it doesn't seem different from installation.
+ Connection between installability and portability.  Use the terminology effort
+ (without cost) to be consistent}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+ \subsection{Correctness \oo{owner}}
+
+\begin{defn} 
+  The extent to which a program satisfies its specifications and fulfills the
+  user's mission objectives \citep{McCallEtAl1977}. 
+\end{defn}
+
+\begin{defn}
+  The ability of software products to perform their exact tasks, as defined by
+  their specification \citep{meyer1988object}.
+\end{defn}
+
+\begin{defn}  
+  The degree to which a system or component is free from faults in its
+  specification, design and implementation \citep{IEEEComputerDictionary1991}.
+\end{defn}
 
@@
 
-\begin{defn}
-  The ability of software products to perform their exact tasks, as defined by
-  their specification \citep{meyer1988object}.
-\end{defn}
-
-\begin{defn} 
-  The extent to which a program satisfies its specifications and fulfills the
-  user's mission objectives \citep{McCallEtAl1977}. 
-\end{defn}
-
-\begin{defn}  
-  The degree to which a system or component is free from faults in its
-  specification, design and implementation \citep{IEEEComputerDictionary1991}.
-\end{defn}
-
 \noindent \textbf{Proposed Definition}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \subsection{\notdone{Verifiability} \oo{owner}}
+
+\begin{defn}
+	 Verification is the process of evaluating a system or component to
+   determine whether the products of a given development phase satisfy the
+   conditions imposed at the start of that phase
+   \citep{IEEEStdGlossarySET1990}.
+\end{defn}
 
@@
 
-\begin{defn} \label{Defn_Verifiability1}
-Verifiability ``means that you should be able to write a set of tests that can
-demonstrate that the delivered system meets each specified requirement'' \citep{sommerville}.	
-\end{defn}
-
-\begin{defn}
-	The evaluation of whether or not a product, service, or result complies with
-	a regulation, requirement, specification, or imposed condition \citep{project2017guide}
-	\oo {verification definition}
-\end{defn}
-
-\begin{defn}
-	By verification, we mean all activities that are undertaken to ascertain that the
-	software meets its objectives \citep{GhezziEtAl2003}.
-\end{defn}
-
-\begin{defn}
-	 Verification is the process of evaluating a system or component to determine whether the products
-	 of a given development phase satisfy the conditions imposed at the start of that phase \citep{IEEEStdGlossarySET1990}.
+\begin{defn}
+	By verification, we mean all activities that are undertaken to ascertain
+  that the software meets its objectives \citep{GhezziEtAl2003}.
 \end{defn}
@@
 \end{defn}
-\noindent \textbf{Proposed Definition}
-Definition~\ref{Defn_Verifiability1} rephrased:
-Verifiability is the extent to which a set of tests can be written and executed, to demonstrate
-that the delivered system meets the specification.
+
+\begin{defn} \label{Defn_Verifiability1}
+  Verifiability ``means that you should be able to write a set of tests that
+  can demonstrate that the delivered system meets each specified
+  requirement'' \citep{sommerville}. %2011
+\end{defn}
+
+\begin{defn}
+	The evaluation of whether or not a product, service, or result complies
+  with a regulation, requirement, specification, or imposed condition
+  \citep{project2017guide} \oo {verification definition}
+\end{defn}
+
+\noindent \textbf{Proposed Definition} Definition~\ref{Defn_Verifiability1}
+rephrased: Verifiability is the extent to which a set of tests can be
+written and executed, to demonstrate that the delivered system meets the
+specification.
 
@@
 Definition is concise and measurable.  Verifiability involves solving the
-equations right \citep[p.~23]{Roache1998} and the definition states that system needs to be verifiable based on the specifications..
+equations right \citep[p.~23]{Roache1998} and the definition states that
+system needs to be verifiable based on the specifications.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
 \subsection{\notdone{Validatability} \oo{owner}}
-
-\begin{defn}
-  Validation means solving the right equations \citep[p.~23]{Roache1998}.
-\end{defn}
-
-\begin{defn}
-  The assurance that a product, service, or result meets the needs of the customer and other identified
-  stakeholders \citep{project2017guide}.
-\end{defn}
-\begin{defn}
-	Software validation is the process of checking that the system conforms to its specification and
-	that it meets the real needs of the users of the system \citep{sommerville}.
-
-\end{defn}
-\begin{defn}
-	Software validation, where the software is checked to ensure that it is what the
-	customer requires \citep{sommerville}.
-\end{defn}
-\begin{defn}
-  Software validation is achieved through a series of tests that demonstrate conformity
-  with requirements \citep{pressman2005software}.
-\end{defn}
-
-\begin{defn}
-	Validation is concerned with establishing that the product fulfills its
-	intended use \citep{van2008software}.
-\end{defn}
 
@@
 
-\noindent \textbf{Proposed Definition}
-
-Definition \ref{Defn_Validatability} rephrased:  The effort required in evaluating
-a system or component during or at the end of the development process to determine
-whether it satisfies specified requirements
-
-\noindent \textbf{Reasoning}
-
-Definition describes that system should be measurable through the development life cycle,
-It is concise and can be measured.
+\begin{defn}
+  Validation means solving the right equations \citep[p.~23]{Roache1998}.
+\end{defn}
+
+\begin{defn}
+  Software validation is achieved through a series of tests that demonstrate
+  conformity with requirements \citep{pressman2005software}.
+\end{defn}
+
+\begin{defn}
+	Validation is concerned with establishing that the product fulfills its
+	intended use \citep{van2008software}.
+\end{defn}
+
+\begin{defn}
+  Software validation is the process of checking that the system conforms to
+  its specification and that it meets the real needs of the users of the
+  system \citep{sommerville}. %2011
+\end{defn}
+
+\begin{defn}
+	Software validation, where the software is checked to ensure that it is
+  what the customer requires \citep{sommerville}. %2011
+\end{defn}
+
+\begin{defn}
+  The assurance that a product, service, or result meets the needs of the
+  customer and other identified stakeholders \citep{project2017guide}.
+\end{defn}
+
+\noindent \textbf{Proposed Definition}
+
+Definition \ref{Defn_Validatability} rephrased:  The effort required in
+evaluating a system or component during or at the end of the development
+process to determine whether it satisfies specified requirements
+
+\noindent \textbf{Reasoning}
+
+Definition describes that system should be measurable through the
+development life cycle, It is concise and can be measured.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \subsection{\notdone{Testability} \oo{owner}}
-
-\begin{defn}
-	The degree to which you can unit-test and system-test a system;
-	the degree to which you can verify that the system meets its requirements \citep{mcconnell2004code}.
-\end{defn}
 
@@
 
+\begin{defn}
+	The degree to which you can unit-test and system-test a system; the degree
+  to which you can verify that the system meets its requirements
+  \citep{mcconnell2004code}.
+\end{defn}
+
 \noindent \textbf{Proposed Definition}
@@
 
-Definition is concise and measurable
+Definition is concise and measurable.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 \begin{defn}
-  Informally, software is reliable if the user can depend on it \citep{GhezziEtAl2003}.
-\end{defn}
-
-\begin{defn}
- The capability of the software product to maintain a specified level of performance when used under specified conditions \cite{ISO9126}.
-\end{defn}
+  Code possesses the characteristic reliability to the extent that it can be
+  expected to perform its intended functions satisfactorily
+  \citep{boehm1976quantitative}.
+\end{defn}
+
+\begin{defn}
+  Extent to which a program can be expected to perform its intended function
+  with required precision \citep{McCallEtAl1977}.
+\end{defn}
+
 \begin{defn} \label{reliabilitySelected1} It is the probability of failure-free
@@
 \begin{defn}
-  Code possesses the characteristic reliability to the extent that it can be
-  expected to perform its intended functions satisfactorily
-  \citep{boehm1976quantitative}.
+  The ability of a system or component to perform its required function under
+  stated conditions for a specified period of time
+  \citep{IEEEComputerDictionary1991}.
+\end{defn}
+
+\begin{defn}
+ The capability of the software product to maintain a specified level of
+ performance when used under specified conditions \cite{ISO9126}. %2001
+\end{defn}
+
+\begin{defn}
+  Informally, software is reliable if the user can depend on it
+  \citep{GhezziEtAl2003}.
+\end{defn}
+
+\begin{defn} \label{reliabilitySelected2} Reliability is concerned with measuring
+	the probability of occurence of faliure - that is, the probability of the
+	observable effects of errors \citep{GhezziEtAl2003}.
+\end{defn}
+
+\begin{defn}
+  The ability of a system to perform its required functions under stated
+  conditions whenever required—having a long mean time between failures
+  \citep{mcconnell2004code}.
+\end{defn}
+
+\begin{defn}
+  Informally, the reliability of a system is the probability, over a given
+  period of time, that the system will correctly deliver services as expected by
+  the user \citep{sommerville}. %2011
 \end{defn}
@@
 \end{defn}
-
-\begin{defn}
-  Extent to which a program can be expected to perform its intended function
-  with required precision \citep{McCallEtAl1977}.
-\end{defn}
-
-\begin{defn}
-  Informally, the reliability of a system is the probability, over a given
-  period of time, that the system will correctly deliver services as expected by
-  the user \citep{sommerville}.
-\end{defn}
-
-\begin{defn}
-  The ability of a system or component to perform its required function under
-  stated conditions for a specified period of time
-  \citep{IEEEComputerDictionary1991}.
-\end{defn}
-
-\begin{defn}
-  The ability of a system to perform its required functions under stated
-  conditions whenever required—having a long mean time between failures
-  \citep{mcconnell2004code}.
-\end{defn}
-\begin{defn} \label{reliabilitySelected2} Reliability is concerned with measuring
-	the probability of occurence of faliure - that is, the probability of the
-	observable effects of errors \citep{GhezziEtAl2003}.
-\end{defn}
-\oo{Statistical definition from  \citep{GhezziEtAl2003} added}
 
@@
 
-\subsection{{Robustness} \pmi{owner}}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{Robustness \pmi{owner}}
+
 \begin{defn}
@@
 \end{defn}
-\begin{defn}
-	The quality can be further informally refined as the ability of a software to
-	keep an acceptable behaviour, expressed in terms of robustness requirements,
-	in
-	spite of exceptional or unforeseen execution conditions (such as the
-	unavailability of system resources, communication failures, invalid or
-	stressful inputs, etc.) \citep{fernandez2005model}.
-\end{defn}
+
+\begin{defn} \label{RobustnessDefnSelected}
+	A program is robust if it behaves ``reasonably'', even in circumstances that
+	were not anticipated in the requirements specification - for example, when it
+	encounters incorrect input data or some hardware malfunction
+	\citep{ghezzi1991fundamentals}. \wss{Use Ghezzi 2003 instead?}
+\end{defn}
+
+\begin{defn}
+	The quality can be further informally refined as the ability of a software
+  to keep an acceptable behaviour, expressed in terms of robustness
+  requirements, in spite of exceptional or unforeseen execution conditions
+  (such as the unavailability of system resources, communication failures,
+  invalid or stressful inputs, etc.) \citep{fernandez2005model}.
+\end{defn}
+
 \begin{defn} \label{RobustnessPartialDefn}
@@
 	specification \citep{boehm2007software}.
-\end{defn}
-\begin{defn} \label{RobustnessDefnSelected}
-	A program is robust if it behaves ``reasonably'', even in circumstances that
-	were not anticipated in the requirements specification - for example, when it
-	encounters incorrect input data or some hardware malfunction
-	\citep{ghezzi1991fundamentals}.
 \end{defn}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \subsection{\notdone{Performance} \pmi{owner}}
@@
 \begin{defn} \label{PerformanceDefnSelected}
-	The degree to which a system or component accomplishes its designated functions
-	within given constraints, such as speed, accuracy, or memory usage
-	\citep{IEEEStdGlossarySET1990}.
-\end{defn}
+  The degree to which a system or component accomplishes its designated
+  functions within given constraints, such as speed, accuracy, or memory
+  usage \citep{IEEEStdGlossarySET1990}.
+\end{defn}
+
+\begin{defn}
+	In software engineering we often equate performance with efficiency. A
+  software system is efficient if it uses computing resources economically
+  \citep{ghezzi1991fundamentals}.  \wss{Change to 2003 reference?}
+\end{defn}
+
 \begin{defn} \label{PerformanceDefnSelected2}
@@
 	real-time demands) \citep{wiegers2003softreq}.
-\end{defn}
-\begin{defn}
-	In software engineering we often equate performance with efficiency. A software system is efficient if it uses computing resources economically
-	\citep{ghezzi1991fundamentals}.
 \end{defn}
@@
 
-\pmi{I will check if Wiegers discusses the difference between speed and timing.}
+\pmi{I will check if Wiegers discusses the difference between speed and
+timing.}
+
 \pmi{Wiegers did not discuss this difference.}
-\pmi{Speed - change over time; timing - related to specification: when something
-	should happen in relation to something else - state transitions}
+
+\pmi{Speed - change over time; timing - related to specification: when
+something should happen in relation to something else - state transitions}
+
 \wss{If we need both speed and timing, we should define them.  We aren't
   including accuracy from the IEEE definition.  Is that on purpose?}
-\pmi{We have decided not to include accuracy. SS will find difference b/w speed and timing from a colleague.}
+
+\pmi{We have decided not to include accuracy. SS will find difference b/w
+speed and timing from a colleague.}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
 \subsection{\notdone{Usability} \jc{owner}}
+
+\begin{defn}
+  ``The effort required to learn, operate, prepare input, and interpret output
+  of a program'' \citep{McCallEtAl1977}.
+\end{defn}
+
+\begin{defn}
+  ``A software is usable - or user friendly - if its human users find it easy
+  to use'' \citep{ghezzi1991fundamentals}. \wss{switch to 2003 ref?}
+\end{defn}
 
@@
   specified goals with effectiveness, efficiency, and satisfaction in a
-  specified context of use.
-  \cite{ISO16982:2002} and \cite{ISO9241:11}
+  specified context of use. \cite{ISO16982:2002} and \cite{ISO9241:11}
 \end{defn}
@@
 economics sense of the word).
-\cite{NielsenNorman}
+\cite{NielsenNorman} %2012
 \end{defn}
@@
 \begin{defn}
-  ``The effort required to learn, operate, prepare input, and interpret output
-  of a program'' \citep{McCallEtAl1977}. 
-\end{defn}
-
-\begin{defn}
-  ``A software is usable - or user friendly - if its human users find it easy
-  to use'' \citep{ghezzi1991fundamentals}.
-\end{defn}
-
-\begin{defn}
   The capability of the software to be understood, learned, used and liked by
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \subsection{\notdone{Maintainability} \pmi{owner}}
+
+\begin{defn}
+  Effort required to locate and fix an error in an operational program
+  \citep{McCallEtAl1977}.
+
+  \wss{It looks like \citet{pressman2005software} was using
+  \citet{McCallEtAl1977} for their definition.} \wss{PM - can you please
+  look into whether \citet{pressman2005software} cite
+  \citet{McCallEtAl1977}?  Do they give a reason for dropping the word
+  operational?}  \pmi{Yes, \citet{pressman2005software} does cite the
+  relevant section of the textbook with \citet{McCallEtAl1977}. While they
+  do not give a reason for specifically dropping the word operational, they
+  do indicate that their given definition ``is a very limited definition''.
+  Perhaps their intent was breadth.} \wss{Peter, please verify the McCall
+  reference definition, and then just cite that.  Olu has found the McCall
+  reference, so you shouldn't have to search for it.  We can remove both the
+  VanVliet and Pressman reference in preference of the McCall definition.}
+
+  \pmi{I have verified the McCall definition and removed the Pressman and
+  van Vliet references above. I will leave these comments until all similar
+  definitions in this document have been edited.}
+\end{defn}
 
@@
 \begin{defn}
-  ISO/IEC 25010 refers to maintainability as the degree of effectiveness and
-  efficiency with which a product or system can be modified by the intended
-  maintainers \citep{ISO/IEC25010}.
+  We will view maintainability as two separate qualities: repairability and
+  evolvability. Software is repairable if it allows the fixing of defects;
+  it is evolvable if it allows changes that enable it to satisfy new
+  requirements \citep{ghezzi1991fundamentals}. \wss{update to 2003 ref?}
+\end{defn}
+
+\begin{defn}
+  The capability of the software to be modified.  Modifications may include
+  corrections, improvements or adaptation of the software to changes in
+  environment, and in requirements and functional specifications.
+  \cite{ISO9126} %2001
 \end{defn}
@@
 
-\begin{defn}
-  We will view maintainability as two separate qualities: repairability and
-  evolvability. Software is repairable if it allows the fixing of defects; it is
-  evolvable if it allows changes that enable it to satisfy new requirements
-  \citep{ghezzi1991fundamentals}.
-\end{defn}
-
 \begin{defn} \label{MaintainabilityDefnSelected2}
@@
 \begin{defn}
-  Effort required to locate and fix an error in an operational program
-  \citep{McCallEtAl1977}. \wss{It looks
-    like \citet{pressman2005software} was using \citet{McCallEtAl1977} for their
-    definition.} \wss{PM - can you please look into whether
-    \citet{pressman2005software} cite \citet{McCallEtAl1977}?  Do they give a
-    reason for dropping the word operational?}  \pmi{Yes,
-    \citet{pressman2005software} does cite the relevant section of the textbook
-    with \citet{McCallEtAl1977}. While they do not give a reason for
-    specifically dropping the word operational, they do indicate that their
-    given definition ``is a very limited definition''. Perhaps their intent was
-    breadth.} \wss{Peter, please verify the McCall reference definition, and
-    then just cite that.  Olu has found the McCall reference, so you shouldn't
-    have to search for it.  We can remove both the VanVliet and Pressman
-    reference in  preference of the McCall definition.}
-\pmi{I have verified the McCall definition and removed the Pressman and van Vliet references above. I will leave these comments until all similar definitions in this document have been edited.}
-\end{defn}
-
-\begin{defn}
-  The capability of the software to be modified.  Modifications may include
-  corrections, improvements or adaptation of the software to changes in
-  environment, and in requirements and functional specifications. \cite{ISO9126}
+  ISO/IEC 25010 refers to maintainability as the degree of effectiveness and
+  efficiency with which a product or system can be modified by the intended
+  maintainers \citep{ISO/IEC25010}. %2011
 \end{defn}
@@
 
-\wss{What are the other attributes?  What is an attribute?  Is it the
-	same as a quality?  We are calling performance a quality in this document.}
-\pmi{The term quality attribute has come up often in the literature - qualities
-	appear to be a subset of attributes. Quality attributes seem to be associated
-	with non-functional requirements. The term resource attribute was encountered in
-	a journal and implied an association with a functional requirement. The two
-	might differ along those lines of NFR and FR.}
+\wss{What are the other attributes?  What is an attribute?  Is it the same
+as a quality?  We are calling performance a quality in this document.}
+
+\pmi{The term quality attribute has come up often in the literature -
+qualities appear to be a subset of attributes. Quality attributes seem to be
+associated with non-functional requirements. The term resource attribute was
+encountered in a journal and implied an association with a functional
+requirement. The two might differ along those lines of NFR and FR.}
 
@@
 
-\subsection{{Reusability} \pmi{owner}}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{Reusability \pmi{owner}}
+
+\begin{defn}
+  Extent to which a program can be used in other applications - related to
+  the packaging and scope of the functions that programs perform
+  \citep{McCallEtAl1977}.
+\end{defn}
 
@@
 \end{defn}
-\begin{defn}
-  Extent to which a program can be used in other applications - related to the packaging and scope of the functions that programs perform \citep{McCallEtAl1977}. 
-\end{defn}
+
+\begin{defn}
+  A product is reusable if we can ``use the product - perhaps with minor changes
+  - to build another product'' \citep{ghezzi1991fundamentals}.  \wss{Update
+  to 2003?}
+\end{defn}
+
 \begin{defn} \label{ReusabilityDefnSelected}
@@
 \end{defn}
-\begin{defn}
-  Reusability is the likelihood a segment of source code that can be used again
-  to add new functionalities with slight or no modification
+
+\begin{defn}
+  Reusability is the likelihood a segment of source code that can be used
+  again to add new functionalities with slight or no modification
   \citep{sandhu2010survey}.
-\end{defn}
-
-\begin{defn}
-  A product is reusable if we can ``use the product - perhaps with minor changes
-  - to build another product'' \citep{ghezzi1991fundamentals}.
 \end{defn}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \subsection{{Portability} \pmi{owner}}
 
-\begin{defn}
-  The ease with which a system or component can be transferred from one hardware
-  or software environment to another \citep{IEEEStdGlossarySET1990}.
-\end{defn}
-
-\begin{defn}
-  An application is portable across a class of environments to the degree that
-  the effort required to transport and adapt it to a new environment in the
-  class is less than the effort of redevelopment \citep{mooney1990strategies}.
-\end{defn}
-
-\begin{defn} \label{PortabilityDefnSelected} Effort required to transfer a program from one hardware configuration and/or software system environment to another \citep{McCallEtAl1977}.
+\begin{defn} \label{PortabilityDefnSelected}
+  Effort required to transfer a program from one hardware configuration
+  and/or software system environment to another \citep{McCallEtAl1977}.
+\end{defn}
+
+\begin{defn}
+  The ease with which a system or component can be transferred from one
+  hardware or software environment to another \citep{IEEEStdGlossarySET1990}.
+\end{defn}
+
+\begin{defn}
+  An application is portable across a class of environments to the degree
+  that the effort required to transport and adapt it to a new environment in
+  the class is less than the effort of redevelopment
+  \citep{mooney1990strategies}.
+\end{defn}
+
+\begin{defn}
+  Portability refers to the ability to run a system on different hardware
+  platforms \citep{ghezzi1991fundamentals}.  \wss{update to 2003?}
+\end{defn}
+
+\begin{defn}
+  The capability of software to be transferred from one environment to
+  another \cite{ISO9126}. %2001
 \end{defn}
@@
 \begin{defn}
-  Code possesses the characteristic of portability to the extent that it can be
-  operated easily and well on computer configurations other than its current
-  one. This implies that special function features, not easily available at
-  other facilities, are not used, that standard library functions and
-  subroutines are selected for universal applicability, and so on
-  \citep{boehm2007software}.
-\end{defn}
-
-\begin{defn}
-  Portability refers to the ability to run a system on different hardware
-  platforms \citep{ghezzi1991fundamentals}.
-\end{defn}
-
-\begin{defn}
-  The capability of software to be transferred from one environment to another
-  \cite{ISO9126}.
-\end{defn}
+  Code possesses the characteristic of portability to the extent that it can
+  be operated easily and well on computer configurations other than its
+  current one. This implies that special function features, not easily
+  available at other facilities, are not used, that standard library
+  functions and subroutines are selected for universal applicability, and so
+  on \citep{boehm2007software}.
+\end{defn}
+
 \noindent \textbf{Proposed Definition}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \subsection{\notdone{Understandability} \jc{owner}}
 
+\begin{defn}
+  The effort ``to uncover the logic of the application''
+  \citep{ghezzi1991fundamentals}.
+\end{defn}
+
+\begin{defn}
+  The capability of the software product to enable the user to understand
+  whether the software is suitable, and how it can be used for particular tasks
+  and conditions of use \cite{ISO9126}. %2001
+\end{defn}
+
 Understandability is artifact-dependent.  What it means for a user-interface
-(graphical or otherwise) to be understandable is wildly different than what it
-means for the code, and even the user documentation.
-
-The literature here is thin and scattered.  More work will need to be done to
-find something useful.
-
-Interestingly, the business literature seems to have taken more care to define
-this.  Here we encounter
+(graphical or otherwise) to be understandable is wildly different than what
+it means for the code, and even the user documentation.
+
+The literature here is thin and scattered.  More work will need to be done
+to find something useful.
+
+Interestingly, the business literature seems to have taken more care to
+define this.  Here we encounter:
+
 \begin{quote}
@@
 \end{quote}
+
 At least this brings in the idea that the \emph{reader} is actively involved,
@@
 
-\begin{defn}
-  The effort ``to uncover the logic of the application''
-  \citep{ghezzi1991fundamentals}.
-\end{defn}
-
-\begin{defn}
-  The capability of the software product to enable the user to understand
-  whether the software is suitable, and how it can be used for particular tasks
-  and conditions of use \cite{ISO9126}.
-\end{defn}
-
 \noindent \textbf{Proposed Definition}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \subsection{\notdone{Interoperability} \ad{owner}}
@@
 \begin{defn}
-  Effort required to couple one system with another
-  \citep{McCallEtAl1977}. \wss{Ao, if
-    this is from McCall, we should just cite the primary source and remove the
-    Van Vliet reference.} \pmi{Yes this is almost exactly from McCall, except he did not have the word "The" at the beginning.}
+  Effort required to couple one system with another \citep{McCallEtAl1977}.
+  \wss{Ao, if this is from McCall, we should just cite the primary source
+  and remove the Van Vliet reference.} \pmi{Yes this is almost exactly from
+  McCall, except he did not have the word "The" at the beginning.}
   \ad{Someone with access to the book added it. I found the book in the pub
@@
 \begin{defn}
-  \label{InteroperabilitySelected}
-  The degree to which two or more systems, products or components can exchange
-  information and use the information that has been exchanged
-  \citep{ISO/IEC25010}.
-\end{defn}
-
-\begin{defn}
   The ability of a system to coexist and cooperate with other systems.
-  \citep{ghezzi1991fundamentals}.
+  \citep{ghezzi1991fundamentals}.  \wss{2003 ref?}
+\end{defn}
+
+\begin{defn}
+  The capability of the software to interact with one or more specified
+  systems \cite{ISO9126}. %2001
 \end{defn}
@@
   The capability to communicate, execute programs, and transfer data among
-  various functional units in a manner that requires the user to have little or
-  no knowledge of the unique characteristics of those units
-  \citep{ISO/IEC/IEEE24765}.
-\end{defn}
-
-\begin{defn}
-  Interoperability is a characteristic of a product or system, whose interfaces
-  are completely understood, to work with other products or systems, present or
-  future, in either implementation or access, without any restrictions
-  \citep{AFUL2019}.
+  various functional units in a manner that requires the user to have little
+  or no knowledge of the unique characteristics of those units
+  \citep{ISO/IEC/IEEE24765}. %2010
+\end{defn}
+
+\begin{defn}
+  \label{InteroperabilitySelected} The degree to which two or more systems,
+  products or components can exchange information and use the information
+  that has been exchanged \citep{ISO/IEC25010}. %2011
+\end{defn}
+
+\begin{defn}
+  Interoperability is a characteristic of a product or system, whose
+  interfaces are completely understood, to work with other products or
+  systems, present or future, in either implementation or access, without
+  any restrictions \citep{AFUL2019}.
 \end{defn}
@@
 
-\begin{defn}
-  The capability of the software to interact with one or more specified systems
-  \cite{ISO9126}.
-\end{defn}
 \noindent \textbf{Proposed Definition}
@@
 
-This definition is concise and also detailed enough to show the concept not only
-on system, but also on products and components. It also covered the concept with
-more than 2 systems.
+This definition is concise and also detailed enough to show the concept not
+only on system, but also on products and components. It also covered the
+concept with more than 2 systems.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 \begin{defn}
-  \label{VisibilitySelected}
-  A software development process is visible if all of its steps and its current
-  status are documented clearly. Another term used to characterize this property
-  is transparency \citep{ghezzi1991fundamentals}.
+  \label{VisibilitySelected} A software development process is visible if
+  all of its steps and its current status are documented clearly. Another
+  term used to characterize this property is transparency
+  \citep{ghezzi1991fundamentals}. \wss{change to 2003 ref?}
+\end{defn}
+
+\begin{defn}
+  Business process visibility, also called process visibility, is the ability to
+  accurately and completely view the processes, transactions and other
+  activities operating within an enterprise \citep{Rouse2013}.
 \end{defn}
@@
 \begin{defn}
-  Business process visibility, also called process visibility, is the ability to
-  accurately and completely view the processes, transactions and other
-  activities operating within an enterprise \citep{Rouse2013}.
-\end{defn}
-
-\begin{defn}
   Process transparency refers to the ability to look inside. The “look inside”
@@
 
-Definition \ref{VisibilitySelected} rephrased: The extent to which all of the
-steps of a software development process and the current status of it are 
-conveyed clearly. \wss{I wonder if rather than
-  ``documented clearly'', we should be more abstract and talk about how these
-  things can be easily determined.  Documentation is probably how the
-  information will be conveyed, but it doesn't have to be how it is done.}  \ad{Adjusted}
+Definition \ref{VisibilitySelected} rephrased: The extent to which all of
+the steps of a software development process and the current status of it are
+conveyed clearly. \wss{I wonder if rather than ``documented clearly'', we
+should be more abstract and talk about how these things can be easily
+determined.  Documentation is probably how the information will be conveyed,
+  but it doesn't have to be how it is done.}  \ad{Adjusted}
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \subsection{\notdone{Reproducibility} \wss{owner}}
@@
 Reproducibility is a required component of the scientific method
-\citep{Davison2012}.  Although QA has, ``a bad name among creative scientists
-and engineers'' \citep[p.~352]{Roache1998}, the community need to recognize that
-participating in QA management also improves reproducibility.  Reproducibility,
-like QA, benefits from a consistent and repeatable computing environment,
-version control and separating code from configuration/parameters
-\citep{Davison2012}.
+\citep{Davison2012}.  Although QA has, ``a bad name among creative
+scientists and engineers'' \citep[p.~352]{Roache1998}, the community need to
+recognize that participating in QA management also improves reproducibility.
+Reproducibility, like QA, benefits from a consistent and repeatable
+computing environment, version control and separating code from
+configuration/parameters \citep{Davison2012}.
 
@@
   A result is said to be reproducible if another researcher can take the
-  original code and input data, execute it, and re-obtain the same result (Peng,
-  Dominici, and Zeger, 2006), as cited in \citet{BenureauAndRougier2017}.
+  original code and input data, execute it, and re-obtain the same result
+  (Peng, Dominici, and Zeger, 2006), as cited in
+  \citet{BenureauAndRougier2017}.
 \end{defn}
@@
 
-The importance of reproducibility is seen by retractions of influential papers,
-like the review of
-\href{https://www.theguardian.com/commentisfree/2020/jun/05/lancet-had-to-do-one-of-the-biggest-retractions-in-modern-history-how-could-this-happen}
-{hydroxychloroquine trials}
+The importance of reproducibility is seen by retractions of influential
+papers, like the review of \href{https://www.theguardian
+.com/commentisfree/2020/jun/05/lancet-had-to-do-one-of-the-biggest
+-retractions-in-modern-history-how-could-this-happen} {hydroxychloroquine
+trials}
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \subsection{\notdone{Productivity} \ad{owner}}
 
-\begin{defn}
-  A quality of the software production process, referring to its efficiency and
-  performance \citep{ghezzi1991fundamentals}.
-\end{defn}
 \begin{defn}
@@
 \end{defn}
-\begin{defn}
-	\label{ProductivitySelected1}
-Productivity is the amount of output (what is produced) per unit of input used.
-In general, productivity is difficult to measure because outputs and inputs are
-typically quite diverse and are often themselves difficult to measure. In the
-context of software, productivity measurement is usually based on a simple ratio
-of product size to project effort. Thus,If we can measure the size of the
-software product and the effort required to develop the product, we have:
+
+\begin{defn}
+  A quality of the software production process, referring to its efficiency
+  and performance \citep{ghezzi1991fundamentals}.  \wss{2003 ref?}
+\end{defn}
+
+\begin{defn}
+	\label{ProductivitySelected2} The number of lines of new code developed per
+  person-day (an imperfect measure of productivity but one that could be
+  measured consistently) \citep{MacCormack2003}.
+\end{defn}
+
+\begin{defn}
+	\label{ProductivitySelected1} Productivity is the amount of output (what is
+  produced) per unit of input used. In general, productivity is difficult to
+  measure because outputs and inputs are typically quite diverse and are
+  often themselves difficult to measure. In the context of software,
+  productivity measurement is usually based on a simple ratio of product
+  size to project effort. Thus,If we can measure the size of the software
+  product and the effort required to develop the product, we have:
 	\begin{align}
@@
 	\end{align}
-Equation (1) assumes that size is the output of the software production process
-and effort is the input to the process. This can be contrasted with the
-	viewpoint of software cost models where we use size as an independent variable
-	(i.e., an input) to predict effort which is treated as an output. Equation (1)
-	is simple to operationalize if we have a single dominant size measure, for
-	example, product size measured in lines of code \citep{Kitchenham2004}.
-\end{defn}
-\begin{defn}
-	\label{ProductivitySelected2}
-The number of lines of new code developed per person-day (an imperfect measure
-of productivity but one that could be measured consistently)
-\citep{MacCormack2003}.
-\end{defn}
-
-\noindent \textbf{Proposed Definition}
-
-The revision of the combination of Definition \ref{ProductivitySelected1} and
-Definition \ref{ProductivitySelected2}: Productivity is the
-amount of output per unit of input used, which can be measured by the summation
-of all output (such as the number of lines of new code, the number of pages of
-new documents and the number of new test cases) produced per person-day.
-
-\noindent \textbf{Reasoning}
-
-It is concise and measurable. \wss{What is the output?  What is the input?  I
-	think the definition needs to give more information on these.  In particular,
-	the above definitions focus on code as the output, but documentation, test
-	cases etc should also be part of the output.  If we are going to measure this,
-	we need a better idea of what we are measuring for outputs and inputs.}
-	\ad{I added another def and made it more measurable}
+  Equation (1) assumes that size is the output of the software production
+  process and effort is the input to the process. This can be contrasted
+  with the viewpoint of software cost models where we use size as an
+  independent variable (i.e., an input) to predict effort which is treated
+  as an output. Equation (1) is simple to operationalize if we have a single
+  dominant size measure, for example, product size measured in lines of code
+  \citep{Kitchenham2004}.
+\end{defn}
+
+\noindent \textbf{Proposed Definition}
+
+The revision of the combination of Definition \ref{ProductivitySelected1}
+and Definition \ref{ProductivitySelected2}: Productivity is the amount of
+output per unit of input used, which can be measured by the summation of all
+output (such as the number of lines of new code, the number of pages of new
+documents and the number of new test cases) produced per person-day.
+
+\noindent \textbf{Reasoning}
+
+It is concise and measurable. \wss{What is the output?  What is the input?
+I think the definition needs to give more information on these.  In
+particular, the above definitions focus on code as the output, but
+documentation, test cases etc should also be part of the output.  If we are
+going to measure this, we need a better idea of what we are measuring for
+outputs and inputs.} \ad{I added another def and made it more measurable}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 \begin{defn}
-  \noindent The ability to meet the needs of the present without compromising
-  the ability of future generations to meet their own needs
+  \noindent The ability to meet the needs of the present without
+  compromising the ability of future generations to meet their own needs
   \citep{Brundtland1987}.
@@
 The fourth dimension of technical sustainability is also added by
-\citep{WolframEtAl2017}.  Technical sustainability is the focus on the thesis by
-\citet{Hygerth2016}.
+\citep{WolframEtAl2017}.  Technical sustainability is the focus on the
+thesis by \citet{Hygerth2016}.
 
@@
   \noindent Sustainable development is a mindset (principles) and an
-  accompanying set of practices that enable a team to achieve and maintain an
-  optimal development pace indefinitely \citep{Tate2005}.
+  accompanying set of practices that enable a team to achieve and maintain
+  an optimal development pace indefinitely \citep{Tate2005}.
 \end{defn}
@@
   \noindent The capacity of the software to endure. In other words,
-  sustainability means that the software will continue to be available in the
-  future, on new platforms, meeting new needs \citep{Katz2016}.
+  sustainability means that the software will continue to be available in
+  the future, on new platforms, meeting new needs \citep{Katz2016}.
 \end{defn}
@@
 \begin{defn}
-Sustainability encompasses cost efficient maintainability and evolvability \citep{SehestedtEtAl2014}.
-\end{defn}
-
-\citet{SehestedtEtAl2014} goes on to say that sustainability can be observed by
-evaluating the four criteria of the architectural model: completeness,
+Sustainability encompasses cost efficient maintainability and evolvability
+\citep{SehestedtEtAl2014}.
+\end{defn}
+
+\citet{SehestedtEtAl2014} goes on to say that sustainability can be observed
+by evaluating the four criteria of the architectural model: completeness,
 consistency, correctness and clarity.
@@
 {Willenbring.pdf}
-lists sustainability factors: extensible, interoperable, maintainable, portable,
-reusable, scalable and usable.  \wss{Sounds like they are listing almost all
-  software qualities.  It seems that sustainability is at least in part achieved
-  by having high quality software.}
+lists sustainability factors: extensible, interoperable, maintainable,
+portable, reusable, scalable and usable.  \wss{Sounds like they are listing
+almost all software qualities.  It seems that sustainability is at least in
+part achieved by having high quality software.}
 
@@
 \end{description}
+
 Find paper that combines nonfunctional qualities into sustainability.
@@
 
-To achieve the qualities listed in Section~\ref{SecQualities}, the documentation
-should achieve the qualities listed in this section.  All but the final quality
-listed (abstraction), are adapted from the IEEE recommended practise for
-producing good software requirements \citep{IEEE1998}.  Abstraction means only
-revealing relevant details, which in a requirements document means stating what
-is to be achieved, but remaining silent on how it is to be achieved.
-Abstraction is an important software development principle for dealing with
-complexity \citep[p.~40]{GhezziEtAl2003}.  Correctness was in the above list, so
-it is not repeated here.  \citet{SmithAndKoothoor2016} present further details
-on the qualities of documentation for SCS.
+To achieve the qualities listed in Section~\ref{SecQualities}, the
+documentation should achieve the qualities listed in this section.  All but
+the final quality listed (abstraction), are adapted from the IEEE
+recommended practise for producing good software requirements
+\citep{IEEE1998}.  Abstraction means only revealing relevant details, which
+in a requirements document means stating what is to be achieved, but
+remaining silent on how it is to be achieved. Abstraction is an important
+software development principle for dealing with complexity \citep[p
+.~40]{GhezziEtAl2003}.  Correctness was in the above list, so it is not
+repeated here.  \citet{SmithAndKoothoor2016} present further details on the
+qualities of documentation for SCS.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 \begin{defn}
-  \label{CompletenessSelected}
-  A specification is complete to the extent that all of its parts are present
-  and each part is fully developed. A software specification must exhibit
-  several properties to assure its completeness \citep{Boehm1984}:
+  Those attributes of the software that provide full implementation of the
+  functions required. \citep{McCallEtAl1977}. \wss{Ao, if this is from McCall,
+  we should just cite the primary source and remove the Van Vliet reference.}
+  \pmi{McCall's definition verbatim: Those attributes of the software that
+  provide full implementation of the functions required.} \ad{Originally added
+  by someone else so I didn't have access to the book. But now it's available
+  and I double checked, Peter got it accurately.}
+\end{defn}
+
+\begin{defn}
+  \label{CompletenessSelected} A specification is complete to the extent
+  that all of its parts are present and each part is fully developed. A
+  software specification must exhibit several properties to assure its
+  completeness \citep{Boehm1984}:
   \begin{itemize}
@@
 \begin{defn}
-  Those attributes of the software that provide full implementation of the 
-  functions required. \citep{McCallEtAl1977}. \wss{Ao, if this is from McCall, 
-  we should just cite the primary source and remove the Van Vliet reference.} 
-  \pmi{McCall's definition verbatim: Those attributes of the software that 
-  provide full implementation of the functions required.} \ad{Originally added 
-  by someone else so I didn't have access to the book. But now it's available 
-  and I double checked, Peter got it accurately.}
-\end{defn}
-
-\begin{defn}
 	An SRS is complete if, and only if, it includes the following elements:
@@
 definition of completeness from \citet[p.\ 5--6]{IEEE1998}.  This definition
-	is for
-	requirements, but maybe there is something we can generalize from the
+is for requirements, but maybe there is something we can generalize from the
 definition?  We could also look for definitions outside of software
-	development.}
-\ad{added IEEE and dictionary}
-\ad{\url{https://annals-csis.org/proceedings/2016/pliks/468.pdf} has a method of
-measuring completeness and consistency, but I haven't summarized any def from it
-yet}
+development.} \ad{added IEEE and dictionary} \ad{\url{https://annals-csis
+.org/proceedings/2016/pliks/468.pdf} has a method of measuring completeness
+and consistency, but I haven't summarized any def from it yet}
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \subsection{\notdone{Consistency} \ad{owner}}
@@
 \begin{defn}
-  \label{ConsistencySelected}
-  A specification is consistent to the extent that its provisions do not
-  conflict with each other or with governing specifications and
-  objectives. Specifications require consistency in several ways
-  \citep{Boehm1984}.
+  Those attributes of the software that provide uniform design and
+  implementation techniques and notation \citep{McCallEtAl1977}.  \wss{Ao,
+    if this is from McCall, we should just cite the primary source and
+    remove the Van Vliet reference.} \pmi{McCall's definition verbatim:
+  Those attributes of the software that provide uniform design and
+  implementation techniques and notation.}\ad{Originally added by someone
+  else so I didn't have access to the book. But now it's available and I
+  double checked, Peter got it accurately.}
+\end{defn}
+
+\begin{defn}
+  \label{ConsistencySelected} A specification is consistent to the extent
+  that its provisions do not conflict with each other or with governing
+  specifications and objectives. Specifications require consistency in
+  several ways \citep{Boehm1984}.
   \begin{itemize}
@@
 \begin{defn}
-  Consistency requires that no two or more requirements in a specification
-  contradict each other. It is also often regarded as the case where words and
-  terms have the same meaning throughout the requirements specifications
-  (consistent use of terminology). These two views of consistency imply that
-  mutually exclusive statements and clashes in terminology should be avoided
-  \citep{ZOWGHI2003}.
-\end{defn}
-\begin{defn}
-  Consistency: 1. the degree of uniformity, standardization, and freedom from
-  contradiction among the documents or parts of a system or component
-  2. software attributes that provide uniform design and implementation
-  techniques and notations \citep{ISO/IEC/IEEE24765}.
-\end{defn}
-
-\begin{defn}
-  Those attributes of the software that provide uniform design and 
-  implementation techniques and notation \citep{McCallEtAl1977}.  \wss{Ao, if 
-  this is from McCall, we should just cite
-    the primary source and remove the Van Vliet reference.} \pmi{McCall's 
-    definition verbatim: Those attributes of the software that provide uniform 
-    design and implementation techniques and notation.}\ad{Originally added 
-  by someone else so I didn't have access to the book. But now it's available 
-  and I double checked, Peter got it accurately.}
-\end{defn}
-
-\begin{defn}
 Consistency refers to internal consistency. If an SRS does not agree with some
@@
 \end{defn}
+
 \begin{defn}
  An SRS is internally consistent if, and only if, no subset of individual
-requirements described in it conflict. The three types of likely conflicts in an
-SRS are as follows \citep{IEEE1998}:
+ requirements described in it conflict. The three types of likely conflicts
+ in an SRS are as follows \citep{IEEE1998}:
  \begin{enumerate}[a)]
-\item The specified characteristics of real-world objects may conflict. For
-example,
+ \item The specified characteristics of real-world
+ objects may conflict. For example,
  \begin{enumerate}[1)]
-\item The format of an output report may be described in one requirement as
-tabular but in another as textual.
-\item One requirement may state that all lights shall be green while another may
+ \item The format of an output report may be described
+ in one requirement as tabular but in another as textual.
+ \item One
+ requirement may state that all lights shall be green while another may
  state that all lights shall be blue.
@@
 \end{defn}
+
+\begin{defn}
+  Consistency requires that no two or more requirements in a specification
+  contradict each other. It is also often regarded as the case where words
+  and terms have the same meaning throughout the requirements specifications
+  (consistent use of terminology). These two views of consistency imply that
+  mutually exclusive statements and clashes in terminology should be avoided
+  \citep{ZOWGHI2003}.
+\end{defn}
+
+\begin{defn}
+  Consistency: 1. the degree of uniformity, standardization, and freedom
+  from contradiction among the documents or parts of a system or component 2
+  . software attributes that provide uniform design and implementation
+  techniques and notations \citep{ISO/IEC/IEEE24765}.
+\end{defn}
+
 \begin{defn}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \subsection{\notdone{Modifiability} \jc{owner}}
@@
 \begin{defn}
-  Modifiability is the degree of ease at which changes can be made to a system,
-  and the flexibility with which the system adapts to such changes.
+  Modifiability is the degree of ease at which changes can be made to a
+  system, and the flexibility with which the system adapts to such changes.
 \end{defn}
@@
 used in other domains.  A generic definition that is still quite useful is
-\begin{defn}
-  The capability (and implementation) of keeping track of a given set or type of
-  information to a given degree, or the ability to chronologically interrelate
-  uniquely identifiable entities in a way that is verifiable.
-\end{defn}
+
+\begin{defn}
+  The capability (and implementation) of keeping track of a given set or
+  type of information to a given degree, or the ability to chronologically
+  interrelate uniquely identifiable entities in a way that is verifiable.
+\end{defn}
+
 By specializing the above to software artifacts, ``interrelate'' to ``why is
-this here'' (for forward tracing from requirements), this does indeed give what
-is meant in SE.
+this here'' (for forward tracing from requirements), this does indeed give
+what is meant in SE.
 
@@
 24765-2017 - ISO/IEC/IEEE International Standard - Systems and software
-engineering--Vocabulary
-has a full definition, namely
+engineering--Vocabulary has a full definition, namely
 \begin{enumerate}
@@
 \end{defn} \wss{Once we have verified McCall, we can remove the Van Vliet
-  reference.}  \pmi{McCall's definition verbatim: Those attributes of the software that provide a thread from the requirements to the implementation with respect to the specific development and operational environment.}
+reference.}  \pmi{McCall's definition verbatim: Those attributes of the
+software that provide a thread from the requirements to the implementation
+with respect to the specific development and operational environment.}
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \subsection{\notdone{Unambiguity} \wss{owner}}
 
-A specification is unambiguous when it has a unique interpretation.  If there is
-a possibility that two readers will have two different interpretations, than the
-specification is ambiguous.  \wss{When I get the Ghezzi text back from Olu, I'll
-  check to see if they have anything to add to this definition.}
+A specification is unambiguous when it has a unique interpretation.  If
+there is a possibility that two readers will have two different
+interpretations, than the specification is ambiguous.  \wss{When I get the
+Ghezzi text back from Olu, I'll check to see if they have anything to add to
+this definition.}
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \subsection{\notdone{Verifiability} \wss{owner}}
@@
 
-\item Verification - Are we building the product right?  Are we implementing the
-  requirements correctly (internal)
-\item Validation - Are we building the right product? Are we getting the right
-  requirements (external)
-\item According to
-
-  \href{https://en.wikipedia.org/wiki/Software_verification_and_validation}{Capability
-    Maturity Model (CMM)}
+\item Verification - Are we building the product right?  Are we implementing
+the requirements correctly (internal) \item Validation - Are we building the
+right product? Are we getting the right requirements (external) \item
+According to  \href{https://en.wikipedia.org/wiki/Software_verification_and_validation}
+{Capability Maturity Model (CMM)}
   \begin{itemize}
@@
 
-Verifiability is related to testability, which is defined by McCall et al.\ as
-``The effort required to test a program to ensure that it performs its intended
-function'' \citep{VanVliet2000}. \wss{Need to verify McCall reference, then
-  delete VanVliet.} \pmi{McCall's definition verbatim: Effort required to test a program to insure it performs its intended function.}
+Verifiability is related to testability, which is defined by McCall et al.\
+as ``The effort required to test a program to ensure that it performs its
+intended function'' \citep{VanVliet2000}. \wss{Need to verify McCall
+reference, then delete VanVliet.} \pmi{McCall's definition verbatim: Effort
+required to test a program to insure it performs its intended function.}
 
@@
 \wss{Needs to be completed}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 

### 2021-09-20 StateOfPractice/Ao-Report/11Introduction.tex
MI report - updated according to the external review

@@
 We measure the qualities of each software by answering questions on a grading template, as shown in Appendix \ref{ap_grading_template}.
-With the quantitative data generated by the template, we rank the software with the Analytic Hierarchy Process (AHP).
+Comparing with our previous research mentioned in Section \ref{sec_motivation}, this project includes a new type of measurement, the empirical measurements, in the grading template.
+We use the empirical measurements to collect metrics from project repositories and source code.
+With the quantitative data generated by the template, we rank the software with the Analytic Hierarchy Process (AHP), which is a technique for group decision-making in complex situations.
+Section \ref{sec_AHP} includes description about the AHP.
 After that, we interview some of the development teams to further understand the status of their development process.

### 2021-09-21 StateOfPractice/Ao-Report/02ack.tex
MI report - updated acknowledgements

@@
 
-acknowledgements here
+I would like to express my appreciation to my instructors, Dr.
+Spencer Smith and Dr.
+Jacques Carette, for guiding me through the research and report writing.
+I also would like to thank other members of our team, Peter Michalski, Oluwaseun Owojaiye, and Dr.
+Michael Noseworthy.
+
+I have my special thanks to my wife, who has always been considerate and supportive during the whole of my graduate study.

### 2021-09-22 StateOfPractice/Papers/MedImageSoft_SOP/MedImageSoft_SOP_AcademicRadiology.tex
Edits to quality paper - mostly changes and additions to the introduction

%% 
%% Copyright 2007-2020 Elsevier Ltd
%% 
%% This file is part of the 'Elsarticle Bundle'.
%% ---------------------------------------------
%% 
%% It may be distributed under the conditions of the LaTeX Project Public
%% License, either version 1.2 of this license or (at your option) any
%% later version.  The latest version of this license is in
%%    http://www.latex-project.org/lppl.txt
%% and version 1.2 or later is part of all distributions of LaTeX
%% version 1999/12/01 or later.
%% 
%% The list of all files belonging to the 'Elsarticle Bundle' is
%% given in the file `manifest.txt'.
%% 
%% Template article for Elsevier's document class `elsarticle'
%% with harvard style bibliographic references

\documentclass[preprint,12pt,authoryear]{elsarticle}

%% Use the option review to obtain double line spacing
%% \documentclass[authoryear,preprint,review,12pt]{elsarticle}

%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
%% for a journal layout:
%% \documentclass[final,1p,times,authoryear]{elsarticle}
%% \documentclass[final,1p,times,twocolumn,authoryear]{elsarticle}
%% \documentclass[final,3p,times,authoryear]{elsarticle}
%% \documentclass[final,3p,times,twocolumn,authoryear]{elsarticle}
%% \documentclass[final,5p,times,authoryear]{elsarticle}
%% \documentclass[final,5p,times,twocolumn,authoryear]{elsarticle}

%% For including figures, graphicx.sty has been loaded in
%% elsarticle.cls. If you prefer to use the old commands
%% please give \usepackage{epsfig}

%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
%% The amsthm package provides extended theorem environments
%% \usepackage{amsthm}

%% The lineno packages adds line numbers. Start line numbering with
%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
%% for the whole article with \linenumbers.
%% \usepackage{lineno}

\journal{Academic Radiology}

\begin{document}

\begin{frontmatter}

%% Title, authors and addresses

%% use the tnoteref command within \title for footnotes;
%% use the tnotetext command for theassociated footnote;
%% use the fnref command within \author or \affiliation for footnotes;
%% use the fntext command for theassociated footnote;
%% use the corref command within \author for corresponding author footnotes;
%% use the cortext command for theassociated footnote;
%% use the ead command for the email address,
%% and the form \ead[url] for the home page:
%% \title{Title\tnoteref{label1}}
%% \tnotetext[label1]{}
%% \author{Name\corref{cor1}\fnref{label2}}
%% \ead{email address}
%% \ead[url]{home page}
%% \fntext[label2]{}
%% \cortext[cor1]{}
%% \affiliation{organization={},
%%            addressline={}, 
%%            city={},
%%            postcode={}, 
%%            state={},
%%            country={}}
%% \fntext[label3]{}

\title{Assessing the State of the Practice for Medical Imaging Software}

%% use optional labels to link authors explicitly to addresses:
%% \author[label1,label2]{}
%% \affiliation[label1]{organization={},
%%             addressline={},
%%             city={},
%%             postcode={},
%%             state={},
%%             country={}}
%%
%% \affiliation[label2]{organization={},
%%             addressline={},
%%             city={},
%%             postcode={},
%%             state={},
%%             country={}}

\author[CAS]{Spencer Smith}
\author[CAS]{Ao Dong}
\author[CAS]{Jacques Carette}
\author[ECE]{Mike Noseworthy}

\affiliation[CAS]{organization={McMaster University, Computing and Software Department},%Department and Organization
            addressline={1280 Main Street West}, 
            city={Hamilton},
            postcode={L8S 4K1}, 
            state={Ontario},
            country={Canada}}

\affiliation[ECE]{organization={McMaster University, Electrical Engineering},%Department and Organization
            addressline={1280 Main Street West}, 
            city={Hamilton},
            postcode={L8S 4K1}, 
            state={Ontario},
            country={Canada}}

\begin{abstract}
%% Text of abstract

\end{abstract}

%%Graphical abstract
%\begin{graphicalabstract}
%\includegraphics{grabs}
%\end{graphicalabstract}

%%Research highlights
%\begin{highlights}
%\item Research highlight 1
%\item Research highlight 2
%\end{highlights}

\begin{keyword}
%% keywords here, in the form: keyword \sep keyword

%% PACS codes here, in the form: \PACS code \sep code

%% MSC codes here, in the form: \MSC code \sep code
%% or \MSC[2008] code \sep code (2000 is the default)

\end{keyword}

\end{frontmatter}

%% \linenumbers

%% main text
\section{}
\label{}

%% The Appendices part is started with the command \appendix;
%% appendix sections are then done as normal sections
%% \appendix

%% \section{}
%% \label{}

%% If you have bibdatabase file and want bibtex to generate the
%% bibitems, please use
%%
%%  \bibliographystyle{elsarticle-harv} 
%%  \bibliography{<your bibdatabase>}

%% else use the following coding to input the bibitems directly in the
%% TeX file.

\begin{thebibliography}{00}

%% \bibitem[Author(year)]{label}
%% Text of bibliographic item

\bibitem[ ()]{}

\end{thebibliography}
\end{document}

\endinput
%%
%% End of file `elsarticle-template-harv.tex'.

### 2021-09-22 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
Edits to quality paper - mostly changes and additions to the introduction

@@
-\documentclass[letterpaper,cleveref]{lipics-v2019}
+\documentclass[letterpaper, cleveref]{lipics-v2019}
 
@@
 }
+\usepackage{bbding}
 
@@
 
-Given that our research focuses on improving the quality of software, we
-often find it necessary to have clear definitions of the different software
-qualities, like maintainability, usability, sustainability, etc.
-Unfortunately, the definitions that we find in the literature are not always
-satisfying.  This became clear when we recently undertook an exercise for
-measuring the \lq\lq{}state of the practice\rq\rq{} in different software
-domains, including the domains of seismology software \citep{SmithEtAl2018},
-Geographic Information Systems \citep{SmithEtAl2018_arXivGIS} and mesh
-generators \citep{smith2016state}. In trying to determine methods to quickly
-assess quality of large samples of existing software, we found communication
-between ourselves and other challenging.  We craved unambigous and
-standardized definitions to enable methods for measuring qualities.
-Therefore, we have endeavored to collect an exhaustive list of definitions
-and from this build a set of recommended definitions, based on our quality
-criteria for high-quality quality definitions.  We are publishing our
-collection, since others may likewise find them valuable.
+In the course of conducting and presenting our software engineering related
+research we have often desired clear definitions for relevant software
+qualities, like maintainability, usability, productivity, sustainability, etc.
+Unfortunately, the definitions of qualities that we found in the literature were
+not always satisfying.  The inadequacy of existing definitions became
+particularly clear when we recently undertook an exercise measuring the
+\lq\lq{}state of the practice\rq\rq{} in different software domains, including
+the domains of seismology software \citep{SmithEtAl2018}, Geographic Information
+Systems \citep{SmithEtAl2018_arXivGIS} and mesh generators
+\citep{smith2016state}. When determining methods to quickly assess quality
+across large samples of existing software, we found communication amongst
+ourselves and with others challenging. We craved unambiguous and standardized
+definitions to enable and inspire methods for measuring software qualities.  To
+help ground our future research, and possibly help others in the same situation,
+we have collected an exhaustive list of definitions.  From this list we have
+produced our set of recommended software quality definitions.
+
+In the current work, quality is defined as a measure of the excellence or worth
+of an entity.  As is standard practice, we do not think of quality as a single
+measure, but rather as a set of measures.  In the overview of managing software
+quality provided by \citet[p 110]{VanVliet2000} the different measures are
+called quality attributes, quality factors, and quality criteria.  In
+\citet{ISO9126} they are called quality characteristics, subcharacteristics and
+attributes.  To keep our presentation simple, and to not imply a classification
+hierarchy that we don't intend, we will just use the term qualities to refer to
+the different measures of quality.
+
+Many taxonomies have been proposed for software qualities, as in
+\citet{McCallEtAl1977, boehm1976quantitative, ISO9126}. Rather than aiming for a
+complete coverage of the history of qualities, the wide variety of potential
+qualities, or the relationships between them, we will instead take a practical
+approach; we will focus on the qualities that we have found most relevant in our
+research.  This means we will include the usual software engineering suspects,
+like usability and maintainability.  Furthermore, since we also do research on
+scientific computing software, we will also include scientifically relevant
+qualities, like validatability and reproducibility relevant.
+
+Given our respect for \citet[p.\ 15--33]{GhezziEtAl2003}, we used their list of
+qualities as a starting point. Our list, shown in
+Table~\ref{Tab_SoftwareQualities}, includes all of their qualities, except for
+timeliness, which we excluded because the already included quality of
+productivity implies the importance of timeliness.  To the list from
+\citet{GhezziEtAl2003} we have added the practical quality of installability,
+since our aforementioned studies on the state of the practice in different
+software domains (for instance \citet{SmithEtAl2018_arXivGIS}) have shown us
+that none of the other qualities matter if the software cannot be properly
+installed.  Given our interest in scientific computing, we also added the
+qualities of validatability and reproducibility. Since we were not able to find
+much information in the literature on the quality of verifiability, we also
+added the closely related quality of testability to our List.  The quality of
+sustainability was added since this term has emerged as a popular goal for
+software projects. The order of the qualities in
+Table~\ref{Tab_SoftwareQualities} follow the order used in
+\cite{GhezziEtAl2003}, which puts related qualities (like correctness and
+reliability) together.  The order is roughly the order in which qualities are
+considered.  We followed these same guidelines when inserting our additional
+qualities into the list.
+
+\begin{table}[h!]
+  \begin{center}
+  \begin{tabular}{lllll}
+  \toprule
+  & External & Internal & Product & Process \\  
+  \midrule
+  Installability & \Checkmark & & \Checkmark & \\
+  Correctness & \Checkmark & \Checkmark & \Checkmark & \Checkmark \\
+  Reliability & \Checkmark & & \Checkmark & \Checkmark \\
+  Robustness & \Checkmark & & \Checkmark & \Checkmark \\     
+  Performance & \Checkmark & & \Checkmark & \\
+  Usability & \Checkmark & & \Checkmark & \Checkmark \\
+  Verifiability & & \Checkmark & \Checkmark & \Checkmark \\
+  Validatability & \Checkmark & & \Checkmark & \\     
+  Testability & & \Checkmark & \Checkmark & \\
+  Maintainability &  & \Checkmark & \Checkmark & \Checkmark\\
+  Reusability & & \Checkmark & \Checkmark & \Checkmark\\
+  Portability & & \Checkmark & \Checkmark & \\
+  Understandability & & \Checkmark &  & \\     
+  Interoperability & \Checkmark & & \Checkmark & \\
+  Productivity &  &  &  & \Checkmark \\
+  Visibility &  & &  & \Checkmark \\
+  Reproducibility & \Checkmark & & \Checkmark & \\
+  Sustainability & & \Checkmark & \Checkmark & \\      
+  \bottomrule
+\end{tabular}
+\end{center}
+\caption{Software Qualities}
+\label{Tab_SoftwareQualities}
+\end{table}
+
+Table~\ref{Tab_SoftwareQualities} makes a distinction between external and
+internal qualities.  External qualities are the qualities that are relevant to
+the user of the software while internal qualities are those that are relevant to
+developers \citep[p.\ 16]{GhezziEtAl2003}.  Although the users are not typically
+interested in internal qualities, it is the internal qualities that influence
+external quality.  External qualities are assessed using the produced software
+and any user relevant documentation.  Internal qualities reference such things
+as the code, documentation, test cases, and the software development process.
+Correctness is shown as both an internal and external quality because it is
+possible to judge the correctness of software and of the various internal
+artifacts produced while developing software.  The external quality of
+reliability is influenced by the internal quality of verifiability.  Since
+validatability is related to getting the right requirements, it is shown as a
+external quality.  Although one could judge the usability of internal
+documentation, the quality of usability is considered as external, with the
+quality of understandability introduced to cover the internal equivalent.
+Filling in the table the way we want to use it.
+
+Some qualities focus on the product, others on the process, and yet others on
+both.  This is shown in Table~\ref{Tab_SoftwareQualities}.  Define process.
+Define product. Process quality improves product quality.  The check marks are
+for qualities that can be applied to a process, not the qualities that a process
+improves.  Examples.  Performance isn't for process.  For process use
+productivity.
 
@@
 
-Not specifically emphasizing scientific computing, but when a specific
-context is needed for a definition, the scientific context will be used.  For
-instance, the ISO definition of accuracy is defined in terms of precision, but
-in scientific computing accuracy and precision are two different concepts.
+The topic of quality has been visited and revisited many times through the
+course of software engineering.  Starting with McCall, Boehm.  Iso standards.
+Taxonomies, relationships between qualities.  Rather than introduce distinctions
+that aren't relevant, we'll just focus on qualities of interest. We are not
+endeavoring to review all of the available literature, nor are we trying to
+provide a complete framework for all qualities and the relationships between
+them.  We are taking a more practical approach.  We have identified the
+qualities that come up most in our research.  Not specifically emphasizing
+scientific computing, but when a specific context is needed for a definition,
+the scientific context will be used.  For instance, the ISO definition of
+accuracy is defined in terms of precision, but in scientific computing accuracy
+and precision are two different concepts.  Some are product and some are process
+and some are both.  Some are internal and some are external.
 
@@
 
+Roadmap.
 \subsection{Background}
@@
 The accuracy and completeness with which users achieve specified
-goals. 
-\cite{ISO9241:11}
+goals. \cite{ISO9241:11}
 \end{defn}
@@
 The resources expended in relation to the accuracy and
-completeness with which users achieve goals.
-\cite{ISO9241:11}
+completeness with which users achieve goals. \cite{ISO9241:11}
 \end{defn}
@@
 characteristics of the users, tasks and the organizational and physical
-environments 
-\cite{ISO9241:11}
-\end{defn}
-
-\begin{defn}
-\textsf{Goal}: intended outcome
-\cite{ISO9241:11}
-\end{defn}
-
-\begin{defn}
-\textsf{Task}: activities required to achieve a goal 
-\cite{ISO9241:11}
-\end{defn}
-
-\section{Qualities of Software Products, Artifacts and
-	Processes} \label{SecQualities}
+environments. \cite{ISO9241:11}
+\end{defn}
+
+\begin{defn}
+\textsf{Goal}: intended outcome. \cite{ISO9241:11}
+\end{defn}
+
+\begin{defn}
+\textsf{Task}: activities required to achieve a goal. \cite{ISO9241:11}
+\end{defn}
+
+\section{Qualities of Software Products, Artifacts and Processes}
+\label{SecQualities}
 
@@
 this definition captures it all.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{\notdone{Verifiability} \oo{owner}}
-
-\begin{defn}
-	 Verification is the process of evaluating a system or component to
-   determine whether the products of a given development phase satisfy the
-   conditions imposed at the start of that phase
-   \citep{IEEEStdGlossarySET1990}.
-\end{defn}
-
-\begin{defn}
-	Verification involves solving the equations right \citep[p.~23]{Roache1998}.
-\end{defn}
-
-\begin{defn}
-  A software is verifiable if its properties can be verified easily
-  \citep{GhezziEtAl2003}.  \wss{Is this a direct quote?} \oo{yes}
-\end{defn}
-
-\begin{defn}
-	By verification, we mean all activities that are undertaken to ascertain
-  that the software meets its objectives \citep{GhezziEtAl2003}.
-\end{defn}
-
-\begin{defn}
-	Verifiability refers to how well software components or the integrated
-  product can be evaluated to demonstrate whether the system functions as
-  expected \citep{wiegers2003softreq}.
-\end{defn}
-
-\begin{defn} \label{Defn_Verifiability1}
-  Verifiability ``means that you should be able to write a set of tests that
-  can demonstrate that the delivered system meets each specified
-  requirement'' \citep{sommerville}. %2011
-\end{defn}
-
-\begin{defn}
-	The evaluation of whether or not a product, service, or result complies
-  with a regulation, requirement, specification, or imposed condition
-  \citep{project2017guide} \oo {verification definition}
-\end{defn}
-
-\noindent \textbf{Proposed Definition} Definition~\ref{Defn_Verifiability1}
-rephrased: Verifiability is the extent to which a set of tests can be
-written and executed, to demonstrate that the delivered system meets the
-specification.
-
-\noindent \textbf{Reasoning}
-
-Definition is concise and measurable.  Verifiability involves solving the
-equations right \citep[p.~23]{Roache1998} and the definition states that
-system needs to be verifiable based on the specifications.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{\notdone{Validatability} \oo{owner}}
-
-\begin{defn} \label{Defn_Validatability}
-  The process of evaluating a system or component during or at the end of the
-  development process to determine whether it satisfies specified requirements
-  \citep{IEEEStdGlossarySET1990}.
-\end{defn}
-
-\begin{defn}
-  Validation means solving the right equations \citep[p.~23]{Roache1998}.
-\end{defn}
-
-\begin{defn}
-  Software validation is achieved through a series of tests that demonstrate
-  conformity with requirements \citep{pressman2005software}.
-\end{defn}
-
-\begin{defn}
-	Validation is concerned with establishing that the product fulfills its
-	intended use \citep{van2008software}.
-\end{defn}
-
-\begin{defn}
-  Software validation is the process of checking that the system conforms to
-  its specification and that it meets the real needs of the users of the
-  system \citep{sommerville}. %2011
-\end{defn}
-
-\begin{defn}
-	Software validation, where the software is checked to ensure that it is
-  what the customer requires \citep{sommerville}. %2011
-\end{defn}
-
-\begin{defn}
-  The assurance that a product, service, or result meets the needs of the
-  customer and other identified stakeholders \citep{project2017guide}.
-\end{defn}
-
-\noindent \textbf{Proposed Definition}
-
-Definition \ref{Defn_Validatability} rephrased:  The effort required in
-evaluating a system or component during or at the end of the development
-process to determine whether it satisfies specified requirements
-
-\noindent \textbf{Reasoning}
-
-Definition describes that system should be measurable through the
-development life cycle, It is concise and can be measured.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{\notdone{Testability} \oo{owner}}
-
-\begin{defn}
-	Code possesses the characteristic testability to the extent that it
-  facilitates the establishment of verification criteria and supports
-  evaluation of its performance \citep{boehm1976quantitative}
-\end{defn}
-
-\begin{defn}
-	Effort required to test a program to ensure it performs
-	its intended function \citep{McCallEtAl1977}.
-\end{defn}
-
-\begin{defn} \label{Defn_Testability}
-	 The degree to which a system or component facilitates the establishment of
-	 test criteria and the performance of tests to determine whether those
-	 criteria have been met \citep{IEEEStdGlossarySET1990}.
-\end{defn}
-
-\begin{defn}
-	The degree to which you can unit-test and system-test a system; the degree
-  to which you can verify that the system meets its requirements
-  \citep{mcconnell2004code}.
-\end{defn}
-
-\noindent \textbf{Proposed Definition}
-
-Definition \ref{Defn_Testability}: The degree to which a system or component
- facilitates the establishment of test criteria and the performance of tests
- to determine whether those criteria have been met.
-
-\noindent \textbf{Reasoning}
-
-Definition is concise and measurable.
 
@@
 
+\subsection{\notdone{Verifiability} \oo{owner}}
+
+\begin{defn}
+	 Verification is the process of evaluating a system or component to
+   determine whether the products of a given development phase satisfy the
+   conditions imposed at the start of that phase
+   \citep{IEEEStdGlossarySET1990}.
+\end{defn}
+
+\begin{defn}
+	Verification involves solving the equations right \citep[p.~23]{Roache1998}.
+\end{defn}
+
+\begin{defn}
+  A software is verifiable if its properties can be verified easily
+  \citep{GhezziEtAl2003}.  \wss{Is this a direct quote?} \oo{yes}
+\end{defn}
+
+\begin{defn}
+	By verification, we mean all activities that are undertaken to ascertain
+  that the software meets its objectives \citep{GhezziEtAl2003}.
+\end{defn}
+
+\begin{defn}
+	Verifiability refers to how well software components or the integrated
+  product can be evaluated to demonstrate whether the system functions as
+  expected \citep{wiegers2003softreq}.
+\end{defn}
+
+\begin{defn} \label{Defn_Verifiability1}
+  Verifiability ``means that you should be able to write a set of tests that
+  can demonstrate that the delivered system meets each specified
+  requirement'' \citep{sommerville}. %2011
+\end{defn}
+
+\begin{defn}
+	The evaluation of whether or not a product, service, or result complies
+  with a regulation, requirement, specification, or imposed condition
+  \citep{project2017guide} \oo {verification definition}
+\end{defn}
+
+\noindent \textbf{Proposed Definition} Definition~\ref{Defn_Verifiability1}
+rephrased: Verifiability is the extent to which a set of tests can be
+written and executed, to demonstrate that the delivered system meets the
+specification.
+
+\noindent \textbf{Reasoning}
+
+Definition is concise and measurable.  Verifiability involves solving the
+equations right \citep[p.~23]{Roache1998} and the definition states that
+system needs to be verifiable based on the specifications.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{\notdone{Validatability} \oo{owner}}
+
+\begin{defn} \label{Defn_Validatability}
+  The process of evaluating a system or component during or at the end of the
+  development process to determine whether it satisfies specified requirements
+  \citep{IEEEStdGlossarySET1990}.
+\end{defn}
+
+\begin{defn}
+  Validation means solving the right equations \citep[p.~23]{Roache1998}.
+\end{defn}
+
+\begin{defn}
+  Software validation is achieved through a series of tests that demonstrate
+  conformity with requirements \citep{pressman2005software}.
+\end{defn}
+
+\begin{defn}
+	Validation is concerned with establishing that the product fulfills its
+	intended use \citep{van2008software}.
+\end{defn}
+
+\begin{defn}
+  Software validation is the process of checking that the system conforms to
+  its specification and that it meets the real needs of the users of the
+  system \citep{sommerville}. %2011
+\end{defn}
+
+\begin{defn}
+	Software validation, where the software is checked to ensure that it is
+  what the customer requires \citep{sommerville}. %2011
+\end{defn}
+
+\begin{defn}
+  The assurance that a product, service, or result meets the needs of the
+  customer and other identified stakeholders \citep{project2017guide}.
+\end{defn}
+
+\noindent \textbf{Proposed Definition}
+
+Definition \ref{Defn_Validatability} rephrased:  The effort required in
+evaluating a system or component during or at the end of the development
+process to determine whether it satisfies specified requirements
+
+\noindent \textbf{Reasoning}
+
+Definition describes that system should be measurable through the
+development life cycle, It is concise and can be measured.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{\notdone{Testability} \oo{owner}}
+
+\begin{defn}
+	Code possesses the characteristic testability to the extent that it
+  facilitates the establishment of verification criteria and supports
+  evaluation of its performance \citep{boehm1976quantitative}
+\end{defn}
+
+\begin{defn}
+	Effort required to test a program to ensure it performs
+	its intended function \citep{McCallEtAl1977}.
+\end{defn}
+
+\begin{defn} \label{Defn_Testability}
+	 The degree to which a system or component facilitates the establishment of
+	 test criteria and the performance of tests to determine whether those
+	 criteria have been met \citep{IEEEStdGlossarySET1990}.
+\end{defn}
+
+\begin{defn}
+	The degree to which you can unit-test and system-test a system; the degree
+  to which you can verify that the system meets its requirements
+  \citep{mcconnell2004code}.
+\end{defn}
+
+\noindent \textbf{Proposed Definition}
+
+Definition \ref{Defn_Testability}: The degree to which a system or component
+ facilitates the establishment of test criteria and the performance of tests
+ to determine whether those criteria have been met.
+
+\noindent \textbf{Reasoning}
+
+Definition is concise and measurable.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \subsection{\notdone{Maintainability} \pmi{owner}}
@@
 concept with more than 2 systems.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{\notdone{Visibility/Transparency} \ad{owner}}
-
-\begin{defn}
-  \label{VisibilitySelected} A software development process is visible if
-  all of its steps and its current status are documented clearly. Another
-  term used to characterize this property is transparency
-  \citep{ghezzi1991fundamentals}. \wss{change to 2003 ref?}
-\end{defn}
-
-\begin{defn}
-  Business process visibility, also called process visibility, is the ability to
-  accurately and completely view the processes, transactions and other
-  activities operating within an enterprise \citep{Rouse2013}.
-\end{defn}
-
-\begin{defn}
-  Visibility provides transparency into the development process. It is the
-  ability to see progress at any point and determine the distance to completion
-  of a goal.  Visibility provides status of not only the progress of the
-  project, but the product itself \citep{GSA2019}.
-\end{defn}
-
-\begin{defn}
-  Process transparency refers to the ability to look inside. The “look inside”
-  provides an in-depth and clear visibility into the business processes and how
-  these operate \citep{PRIME2019}.
-\end{defn}
-
-\begin{defn}
-The degree to which something is seen by the public
-\citep{CambridgeVisibility2019}.
-
-The degree to which something is seen or known about
-\citep{CambridgeVisibility2019}.
-\end{defn}
-
-\noindent \textbf{Proposed Definition}
-
-Definition \ref{VisibilitySelected} rephrased: The extent to which all of
-the steps of a software development process and the current status of it are
-conveyed clearly. \wss{I wonder if rather than ``documented clearly'', we
-should be more abstract and talk about how these things can be easily
-determined.  Documentation is probably how the information will be conveyed,
-  but it doesn't have to be how it is done.}  \ad{Adjusted}
-
-\noindent \textbf{Reasoning}
-
-Definition \ref{VisibilitySelected} points out that documentation is the way to
-improve visibility. It is rephrased because the original one might refer to
-binary status - ``visible'' or ``invisible''.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{\notdone{Reproducibility} \wss{owner}}
-
-Reproducibility is a required component of the scientific method
-\citep{Davison2012}.  Although QA has, ``a bad name among creative
-scientists and engineers'' \citep[p.~352]{Roache1998}, the community need to
-recognize that participating in QA management also improves reproducibility.
-Reproducibility, like QA, benefits from a consistent and repeatable
-computing environment, version control and separating code from
-configuration/parameters \citep{Davison2012}.
-
-Reproducibility is defined as:
-
-\begin{defn}
-  A result is said to be reproducible if another researcher can take the
-  original code and input data, execute it, and re-obtain the same result
-  (Peng, Dominici, and Zeger, 2006), as cited in
-  \citet{BenureauAndRougier2017}.
-\end{defn}
-
-The related concept of replicable is defined as:
-
-\begin{defn}
-  Documentation achieves replicability if the description it provides of the
-  algorithms is sufficiently precise and complete for an independent researcher
-  to re-obtain the results it presents.  \citep{BenureauAndRougier2017}
-\end{defn}
-
-It would be worthwhile to look for some additional definitions.
-
-\noindent \textbf{Proposed Definition}
-
-\wss{Needs to be completed}
-
-The importance of reproducibility is seen by retractions of influential
-papers, like the review of \href{https://www.theguardian
-.com/commentisfree/2020/jun/05/lancet-had-to-do-one-of-the-biggest
--retractions-in-modern-history-how-could-this-happen} {hydroxychloroquine
-trials}
-
-\noindent \textbf{Reasoning}
-
-\wss{Reproducibility literature review is given in \citet{FeinbergEtAl2020}.
-Includes examples where replicability was found to be lacking.  Has a good
-litereature review in Section 2.}
-
-\wss{Needs to be completed}
 
@@
 outputs and inputs.} \ad{I added another def and made it more measurable}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{\notdone{Visibility/Transparency} \ad{owner}}
+
+\begin{defn}
+  \label{VisibilitySelected} A software development process is visible if
+  all of its steps and its current status are documented clearly. Another
+  term used to characterize this property is transparency
+  \citep{ghezzi1991fundamentals}. \wss{change to 2003 ref?}
+\end{defn}
+
+\begin{defn}
+  Business process visibility, also called process visibility, is the ability to
+  accurately and completely view the processes, transactions and other
+  activities operating within an enterprise \citep{Rouse2013}.
+\end{defn}
+
+\begin{defn}
+  Visibility provides transparency into the development process. It is the
+  ability to see progress at any point and determine the distance to completion
+  of a goal.  Visibility provides status of not only the progress of the
+  project, but the product itself \citep{GSA2019}.
+\end{defn}
+
+\begin{defn}
+  Process transparency refers to the ability to look inside. The “look inside”
+  provides an in-depth and clear visibility into the business processes and how
+  these operate \citep{PRIME2019}.
+\end{defn}
+
+\begin{defn}
+The degree to which something is seen by the public
+\citep{CambridgeVisibility2019}.
+
+The degree to which something is seen or known about
+\citep{CambridgeVisibility2019}.
+\end{defn}
+
+\noindent \textbf{Proposed Definition}
+
+Definition \ref{VisibilitySelected} rephrased: The extent to which all of
+the steps of a software development process and the current status of it are
+conveyed clearly. \wss{I wonder if rather than ``documented clearly'', we
+should be more abstract and talk about how these things can be easily
+determined.  Documentation is probably how the information will be conveyed,
+  but it doesn't have to be how it is done.}  \ad{Adjusted}
+
+\noindent \textbf{Reasoning}
+
+Definition \ref{VisibilitySelected} points out that documentation is the way to
+improve visibility. It is rephrased because the original one might refer to
+binary status - ``visible'' or ``invisible''.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{\notdone{Reproducibility} \wss{owner}}
+
+Reproducibility is a required component of the scientific method
+\citep{Davison2012}.  Although QA has, ``a bad name among creative
+scientists and engineers'' \citep[p.~352]{Roache1998}, the community need to
+recognize that participating in QA management also improves reproducibility.
+Reproducibility, like QA, benefits from a consistent and repeatable
+computing environment, version control and separating code from
+configuration/parameters \citep{Davison2012}.
+
+Reproducibility is defined as:
+
+\begin{defn}
+  A result is said to be reproducible if another researcher can take the
+  original code and input data, execute it, and re-obtain the same result
+  (Peng, Dominici, and Zeger, 2006), as cited in
+  \citet{BenureauAndRougier2017}.
+\end{defn}
+
+The related concept of replicable is defined as:
+
+\begin{defn}
+  Documentation achieves replicability if the description it provides of the
+  algorithms is sufficiently precise and complete for an independent researcher
+  to re-obtain the results it presents.  \citep{BenureauAndRougier2017}
+\end{defn}
+
+It would be worthwhile to look for some additional definitions.
+
+\noindent \textbf{Proposed Definition}
+
+\wss{Needs to be completed}
+
+The importance of reproducibility is seen by retractions of influential
+papers, like the review of \href{https://www.theguardian
+.com/commentisfree/2020/jun/05/lancet-had-to-do-one-of-the-biggest
+-retractions-in-modern-history-how-could-this-happen} {hydroxychloroquine
+trials}
+
+\noindent \textbf{Reasoning}
+
+\wss{Reproducibility literature review is given in \citet{FeinbergEtAl2020}.
+Includes examples where replicability was found to be lacking.  Has a good
+litereature review in Section 2.}
+
+\wss{Needs to be completed}
 

### 2021-09-30 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
Completed installability and almost completed correctness

@@
 colorlinks=true,        % false: boxed links; true: colored links
-linkcolor=red,          % color of internal links (change box color with linkbordercolor)
+linkcolor=blue,         % color of internal links (change box color with linkbordercolor)
 citecolor=blue,         % color of links to bibliography
@@
 \usepackage{bbding}
+\usepackage{tcolorbox}
+
+% new tcolorbox environment
+% #1: tcolorbox options
+\newtcolorbox{mybox}[1][]
+{
+  colframe = yellow!50!red,
+  colback  = white,
+  width=\dimexpr\textwidth+12mm\relax,
+  enlarge left by=-6mm,
+  #1,
+}
 
@@
 \newcommand{\done}[1]{\textcolor{black}{#1}}
-
-%\oddsidemargin 0mm
-%\evensidemargin 0mm
-%\textwidth 160mm
-%\textheight 200mm
 
@@
 
+\newpage
+
 \tableofcontents
 
+\newpage
+
 \section{Introduction} \label{SecIntroduction}
-
-% Purpose
 
@@
 qualities, like maintainability, usability, productivity, sustainability, etc.
-Unfortunately, the definitions of qualities that we found in the literature were
-not always satisfying.  The inadequacy of existing definitions became
+Unfortunately, the definitions of qualities that we have found in the literature
+are not always satisfying.  This inadequacy of existing definitions became
 particularly clear when we recently undertook an exercise measuring the
@@
 
-In the current work, quality is defined as a measure of the excellence or worth
-of an entity.  As is standard practice, we do not think of quality as a single
-measure, but rather as a set of measures.  In the overview of managing software
-quality provided by \citet[p 110]{VanVliet2000} the different measures are
-called quality attributes, quality factors, and quality criteria.  In
-\citet{ISO9126} they are called quality characteristics, subcharacteristics and
-attributes.  To keep our presentation simple, and to not imply a classification
-hierarchy that we don't intend, we will just use the term qualities to refer to
-the different measures of quality.
+In the current work, software is defined as either programs, which run (execute)
+on a computer, or libraries, which provide services to be used by programs. Both
+programs and libraries are created using computer code, written in a programming
+language.
+
+Quality is defined as a measure of the excellence or worth of an entity.  As is
+standard practice, we do not think of quality as a single measure, but rather as
+a set of measures.  That is, quality is a collection of different qualities,
+often called ``ilities.''  In the overview of managing software quality provided
+by \citet[p 110]{VanVliet2000} the different measures are called quality
+attributes, quality factors, and quality criteria.  In \citet{ISO9126} they are
+called quality characteristics, sub-characteristics and attributes.  To keep our
+presentation simple, and to not imply a classification hierarchy that we don't
+intend, we will just use the term qualities to refer to the different measures
+of quality.
+
+The qualities of most interest to project managers and users are summarized in
+the next subsection (\ref{Sec_IndirectQualities}).  Since these qualities cannot
+generally be measured directly, they are characterized as indirect qualities.
+The relevant indirect qualities are listed in this section, while being
+characterized as external, internal, product or process qualities.  The next
+subsection lists the relevant direct qualities that assist in achieving the
+indirect qualities.  The final subsection presents the methodology that was
+employed to summarize, analyze and define each quality.
+
+\subsection{Indirect Qualities} \label{Sec_IndirectQualities}
 
 Many taxonomies have been proposed for software qualities, as in
-\citet{McCallEtAl1977, boehm1976quantitative, ISO9126}. Rather than aiming for a
+\citet{McCallEtAl1977, boehm1976quantitative, ISO9126}. Rather than aiming for
 complete coverage of the history of qualities, the wide variety of potential
@@
 research.  This means we will include the usual software engineering suspects,
-like usability and maintainability.  Furthermore, since we also do research on
+like usability and maintainability.  In addition, since we also do research on
 scientific computing software, we will also include scientifically relevant
-qualities, like validatability and reproducibility relevant.
+qualities, like validatability and reproducibility.
 
@@
 Table~\ref{Tab_SoftwareQualities}, includes all of their qualities, except for
-timeliness, which we excluded because the already included quality of
-productivity implies the importance of timeliness.  To the list from
+timeliness, which was excluded because the already included quality of
+productivity overlaps with timeliness.  To the list from
 \citet{GhezziEtAl2003} we have added the practical quality of installability,
@@
 much information in the literature on the quality of verifiability, we also
-added the closely related quality of testability to our List.  The quality of
-sustainability was added since this term has emerged as a popular goal for
-software projects. The order of the qualities in
-Table~\ref{Tab_SoftwareQualities} follow the order used in
-\cite{GhezziEtAl2003}, which puts related qualities (like correctness and
-reliability) together.  The order is roughly the order in which qualities are
-considered.  We followed these same guidelines when inserting our additional
-qualities into the list.
+added the closely related quality of testability.  The quality of sustainability
+was added since this term has emerged as a popular goal for software projects.
+
+The order of the qualities in Table~\ref{Tab_SoftwareQualities} follow the order
+used in \cite{GhezziEtAl2003}, which puts related qualities (like correctness
+and reliability) together.  Moreover, the order is roughly the order in which
+qualities are considered in practice.  We followed these same guidelines when
+inserting our additional qualities into the list.
 
@@
   Verifiability & & \Checkmark & \Checkmark & \Checkmark \\
-  Validatability & \Checkmark & & \Checkmark & \\     
+  Validatability & \Checkmark & & \Checkmark & \Checkmark\\     
   Testability & & \Checkmark & \Checkmark & \\
@@
   Portability & & \Checkmark & \Checkmark & \\
-  Understandability & & \Checkmark &  & \\     
+  Understandability & & \Checkmark & \Checkmark & \\     
   Interoperability & \Checkmark & & \Checkmark & \\
@@
 Correctness is shown as both an internal and external quality because it is
-possible to judge the correctness of software and of the various internal
-artifacts produced while developing software.  The external quality of
-reliability is influenced by the internal quality of verifiability.  Since
-validatability is related to getting the right requirements, it is shown as a
-external quality.  Although one could judge the usability of internal
-documentation, the quality of usability is considered as external, with the
-quality of understandability introduced to cover the internal equivalent.
-Filling in the table the way we want to use it.
+possible to judge the correctness of software and to judge the correctness of
+the various internal artifacts produced while developing the software.  The
+external quality of reliability is influenced by the internal quality of
+verifiability. Since validatability is related to getting the right
+requirements, it is shown as an external quality.  Although one could judge the
+usability of internal documentation, the quality of usability is considered as
+external, with the quality of understandability designated to cover the internal
+equivalent of usability.
 
 Some qualities focus on the product, others on the process, and yet others on
-both.  This is shown in Table~\ref{Tab_SoftwareQualities}.  Define process.
-Define product. Process quality improves product quality.  The check marks are
-for qualities that can be applied to a process, not the qualities that a process
-improves.  Examples.  Performance isn't for process.  For process use
-productivity.
-
-% Scope
-
-The topic of quality has been visited and revisited many times through the
-course of software engineering.  Starting with McCall, Boehm.  Iso standards.
-Taxonomies, relationships between qualities.  Rather than introduce distinctions
-that aren't relevant, we'll just focus on qualities of interest. We are not
-endeavoring to review all of the available literature, nor are we trying to
-provide a complete framework for all qualities and the relationships between
-them.  We are taking a more practical approach.  We have identified the
-qualities that come up most in our research.  Not specifically emphasizing
-scientific computing, but when a specific context is needed for a definition,
-the scientific context will be used.  For instance, the ISO definition of
-accuracy is defined in terms of precision, but in scientific computing accuracy
-and precision are two different concepts.  Some are product and some are process
-and some are both.  Some are internal and some are external.
-
-\wss{List of definitions don't include textbooks that simply repeat
-definitions from the literature.}
-
-% Methodology
-Methodology:
-
-\begin{itemize}
-
-  \item Collect the list of qualities to define, based on the qualities that are
-  listed in software engineering texts, such as Ghezzi Et al and Van Hliet.
-
-  \item For each quality find as many definitions as possible
-
-  \item Based on the existing definitions, propose a high quality definition of
-  each quality.  The proposed definition may use one of the existing, or combine
-  several of the options.  We may also make other modifications.  The goal is
-  for the proposed definition to address our acceptance criteria.
-
-\end{itemize}
-
-software products, software artifacts and software development processes, and
-ii) qualities that are considered important for good specifications.  The
-specification could be a specification of requirements, design or a test plan.
-
-The pattern in the presentation for each of the qualities is the same.  First we
-summarize all of the definitions that we could find in the literature.  To keep
-the clutter of quotation marks down, we have adopted the convention that each
-definition is given verbatim from the cited source, but without showing
-quotation marks.  In cases where the definition has to be rephrased, quotation
-marks are used to show those portions that are taken verbatim for the original
-source.  After the summary of the existing definitions, we propose the
-definition that we would like to work with going forward.  This definition can
-either be our preference from the existing definitions, or a new definition,
-which is often found by combining existing definitions.  Following the proposed
-definition is an explanation for the reasoning that led to this choice.
-
-\wss{introduce the terminology of quality characteristics, as in ISO9126?}
-
-\wss{tell the reader that the definitions will be sorted in the order of their
-publication.}
-
-In determining our recommended definition for each quality, we used the
-following criteria:
+both, as shown in Table~\ref{Tab_SoftwareQualities}.  The term product refers to
+both the external and internal products.  The external products are the
+user-visible products, like the software itself and any user documentation.  The
+internal products are the products (sometimes called work products or artifacts)
+relevant to the developers, like requirements documentation, design
+documentation, test cases, makefiles, etc.  The process refers to the sequence
+of activities that produce the software.  The process generates both the
+internal and external products.  The process is based on principles, methods and
+techniques.  The process is supported by tools.  All product qualities (internal
+and external) are effected by the process, so this information is not shown in
+Table~\ref{Tab_SoftwareQualities}.  Instead, the checkmarks in
+Table~\ref{Tab_SoftwareQualities} are for qualities that can be applied to a
+process.  For instance, we do discuss the installability or correctness of a
+process, but it does make sense to think about the portability of a process.
+Rather than discuss the performance of a process, we introduce the process
+specific quality of productivity.  Similarly, although we can discuss the
+usability of a process, we instead introduce the term understandability to
+exclusively apply to internal products (like the code).
+
+The qualities shown in Table~\ref{Tab_SoftwareQualities} are qualities that can
+only be measured indirectly \citep[p.\ 109]{VanVliet2000}.  For instance, a
+direct measure of maintainability doesn't exist.  Since it cannot be measured
+directly, we base our estimation of maintainability on direct measures that are
+likely to influence maintainability, like consistency, traceability, etc.
+Qualities that can be measured directly are discussed in the next section.
+
+\subsection{Direct Qualities} \label{SecDirectQsIntro}
+
+To achieve the qualities listed in Section~\ref{Sec_IndirectQualities},
+conventional wisdom holds that the internal products should aim for the
+qualities listed below.  All but the final quality listed (abstraction), are
+adapted from the IEEE recommended practice for producing good software
+requirements \citep{IEEE1998}. Abstraction is an important software development
+principle for dealing with complexity \citep[p .~40]{GhezziEtAl2003}.
+\citet{IEEE1998} also includes correctness and verifiability, but we have
+excluded these from the list of direct qualities, since they belong in the list
+of indirect qualities in the previous section.
+
+\begin{itemize}
+  \item Completeness
+  \item Consistency
+  \item Modifiability
+  \item Traceability
+  \item Unambiguity
+  \item Abstract
+\end{itemize}
+
+All of the direct qualities listed above are internal qualities.  Furthermore,
+they apply to both the (internal) product and the process.
+
+\subsection{Methodology}
+
+For each quality we collect as many distinct instances of the definition that we
+can find.  Some of the definitions occur in multiple places (for instance the
+definitions from \citet{McCallEtAl1977} appear in many software engineering
+textbooks).  Since these instances do not add any new information, we do not
+make an effort to identify how frequently the definitions reappear in different
+sources.  Once all the definitions are collected, we determine the definition
+that we would like to work with going forward.  This definition can either be
+our preference from the existing definitions, or a new definition, which is
+often found by combining existing definitions.
+
+To ensure that our recommended definitions are quality definitions, they are
+based on direct quality measures (\ref{SecDirectQsIntro}).  To the list of
+direct qualities, we add the quality criteria of measurability.  How the direct
+qualities are applied to develop definitions of qualities is described below.
 
@@
 
-  \item[Unambiguous]
-
-  \item[Consistency] The final list of definitions should be consistent in
-  the terminology used.  For instance, we do not use synonyms for variety;
-  we use the same word to mean the same concept throughout the final list of
-  definitions.  This is why measurement is given by ``the effort required.''
-  and not the various synonyms for this that are seen in the literature.
-  \wss{We may change this phrase we use for measuring something.  I just
-  wanted to put this in here as a placeholder.}
-
-  \item[Generality]
-
-  \item[Measurability]
-
-  \item[Etc.]
+  \item[Completeness] To judge the completeness of our definitions, we verify
+  that all of the relevant points from the collected definitions have been
+  included.  We also verify that all facets of the definition are covered when a
+  quality applies to multiple categories of internal, external, product and/or
+  process.  Finally, we check that all of the qualities listed here have been
+  considered.
+
+  \item[Consistency] The final list of definitions should be consistent in the
+  terminology used.  For instance, we do not use synonyms for variety; we use
+  the same word to mean the same concept throughout the final list of
+  definitions.  Specific decisions on terminology are as follows:
+  \begin{itemize}
+    \item To represent the idea of cost, effort, time, resources, and
+    efficiency, we say \emph{effort}.
+    \item To represent ``the degree to which'' or the ``the extent to which'',
+    we will say \emph{the extent to which}.
+    \item Fill in as go through definitions.
+  \end{itemize}
+
+  \item[Modifiability] The modifiability of the definitions is aided by the fact
+  that the definitions are short.  Changes are not difficult to make.  The
+  modifiability is further aided by the related qualities of traceability and
+  abstraction.
+
+  \item[Traceability] Each definition includes a section for the rationale.  The
+  rationale explicitly links the recommended definition to the previously
+  published definitions.  The rationale also explains why any aspects of the
+  previous definitions were ignored.  The rationale also traces the selected
+  definition to these qualities.
+
+  \item[Unambiguity] With natural language it is difficult to remove all
+  ambiguity, but every effort is made to keep the definitions simple and to use
+  standard, consistent, terminology.
+
+  \item[Abstract] Quality definitions should say what the quality achieves, but
+  not how to achieve it.  We aim for abstract definitions that do not assume any
+  specific methodology or tools.  For those qualities that fit into multiple
+  categories of internal, external, product and/or process, we aim to have one
+  definition that will apply to all.  When that is not possible, we will clarify
+  how the definition is modified between categories.
+
+  \item[Measurability] Although many qualities can only be indirectly measurable,
+  the definition of the quality should imply a measure that is conceptually
+  possible, even if the data needed to complete the measurement is often
+  unavailable.  The scale type of the units of measure are as follows, in order
+  of preference: ratio, interval, ordinal and nominal.  These scale types are
+  defined in \citet[p.\ 107]{VanVliet2000}.  The ratio and interval scale types
+  are preferred to the ordinal and nominal types because more they can convey
+  more information.  Specific consequences of aiming for measurability are as follows:
+  \begin{itemize}
+    \item We will avoid the common phrase in definitions of ``the capability
+    of'' because this implies a binary measure.  By this definition a product
+    (or process) is either capable or not; it is a binary measure.  Instead of
+    measuring capability, we will prefer to use the phrase ``the effort
+    required''.
+    \item etc.
+  \end{itemize}
 
@@
 
-Not necessary to have every quality measurable.  Analogy to the definition of
-true error in scientific computing.
-
-Roadmap.
-\subsection{Background}
-
-Quality definitions history: McCall, Ghezzi et al, VanHleit, Boehm, ISO standards.
-
-\subsection{Additional Definitions}
-
-Some standards give additional definitions for terms that are used a lot.
-These are reproduced here, to aid in the interpretation of the definitions
-below (and their refinement).
-
-\begin{defn} 
-\textsf{Effectiveness}:
-The accuracy and completeness with which users achieve specified
-goals. \cite{ISO9241:11}
-\end{defn}
-
-\begin{defn} 
-\textsf{Efficiency}: 
-The resources expended in relation to the accuracy and
-completeness with which users achieve goals. \cite{ISO9241:11}
-\end{defn}
-
-\begin{defn}
-\textsf{Satisfaction}: 
-Freedom from discomfort, and positive attitude to the use of the product.
-\cite{ISO9241:11}
-\end{defn}
-
-\begin{defn}
-\textsf{Context of use}:
-characteristics of the users, tasks and the organizational and physical
-environments. \cite{ISO9241:11}
-\end{defn}
-
-\begin{defn}
-\textsf{Goal}: intended outcome. \cite{ISO9241:11}
-\end{defn}
-
-\begin{defn}
-\textsf{Task}: activities required to achieve a goal. \cite{ISO9241:11}
-\end{defn}
-
-\section{Qualities of Software Products, Artifacts and Processes}
+Our aim is to capture the essence of the quality in an unambiguous way, even if
+it cannot be measured. This is analogous to the definition of true error in
+scientific computing.  The true error, which is the difference between the
+calculated and the true solution, can rarely be calculated since the true
+solution is rarely known. Even though it can usually only be estimated, the
+concept of true error is still a concept that is relevant for analyzing and
+understanding the behaviour of numerical algorithms.
+
+After completing each definition we verify that the definition works for all of
+the quality categories (internal, external, product and process) that apply.  If
+not, the definition is modified.  The rationale section is written to explicitly
+addresses each of these points.  Each definition is also checked to verify that
+it is complete, consistent, traceable, unambiguous, abstract and measurable.
+The rational section is written to explain how these qualities, for each quality
+definition, are achieved.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Indirect Qualities of Software, Artifacts and Processes}
 \label{SecQualities}
 
-\wss{add a definition related to capability?  The word seems to come up in  many definitions.}
-
-To assess the current state of software development, and to understand how
-future changes impact software development, we need a clear definition of what
-we mean by quality.  The concept of quality is decomposed into a set of separate
-components that together make up ``quality''. Unfortunately, these are called
-\emph{qualities}. These are associated to the software product, the software
-artifacts (documentation, test cases, etc) and to the software development
-process itself, and combinations thereof.
-
-Our analysis is centred around a set of software qualities.  Quality is not
-considered as a single measure, but a collection of different qualities, often
-called ``ilities.''  These qualities highlight the desirable nonfunctional
-properties for software artifacts, which include both documentation and
-code. Some qualities, such as visibility and productivity, apply to the process
-used for developing the software. The following list of qualities is based on
-\cite{GhezziEtAl2003}. To the list from \cite{GhezziEtAl2003}, we have added
-three qualities important for SC: installability, reproducibility and
-sustainability.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{Installability \oo{owner}}
-
-\begin{defn}
+The presentation below for each of the qualities is the same.  First, we
+summarize, in the order of their publication, all of the definitions that we
+could find in the literature.  To keep the clutter of quotation marks down, we
+have adopted the convention that each definition is given verbatim from the
+cited source, but without showing quotation marks.  In cases where the
+definition has to be rephrased, quotation marks are used to show those portions
+that are taken verbatim for the original source.  Second, after the summary of
+the existing definitions, we propose the definition that we would like to work
+with going forward. Third, following the proposed definition is an explanation
+for the reasoning that led to this choice.  The definitions apply to internal,
+external, product and process following the designations in
+Table~\ref{Tab_SoftwareQualities}. Where necessary, this is clarified in the
+definition and its rationale.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{Installability (external, product)} %\oo{owner}
+
+\begin{defn} \label{Installability_ISO9126}
   The capability of the software product to be installed in a specified
@@
 
-\noindent \textbf{Proposed Definition}
-
-Combined definitions \ref{Installability_Selected2} and
-\ref{Installability_Selected1} rephrased: Installability refers to the effort
-required for the installation, uninstallation or reinstallation of a software or
-product in a specified environment.
-
-\wss{Remove the repetition of the word being defined in the definition?  Remove
-  reinstallation?  This quality characteristic is a product quality.  It is an
-  external quality.  Installability should be something that can be measured
-  directly.}
-
-\noindent \textbf{Reasoning}
-
- The definition captures a vital metric of measure for this quality - effort,
- and considers not just the installation but uninstallation and re-installation
- which are all related to installation activities.
-
- \wss{remove re-installation - it doesn't seem different from installation.
- Connection between installability and portability.  Use the terminology effort
- (without cost) to be consistent}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
- \subsection{Correctness \oo{owner}}
-
-\begin{defn} 
+\begin{mybox}
+\subsubsection*{Installability}
+
+The effort required for the installation and/or uninstallation of software in a
+specified environment.
+
+\end{mybox}
+
+\subsubsection*{Reasoning}
+
+The recommended definition comes from combining Definitions
+\ref{Installability_Selected2} and \ref{Installability_Selected1}.  The
+different reasons for the proposed definition are as follows:
+
+\begin{itemize}
+  \item For consistency uses the term \textbf{effort} in place of efficiency.
+  \item For measurability avoids the term ``the capability of'' from
+  Definition~\ref{Installability_ISO9126}.
+  \item The ``effectiveness'' from Definition~\ref{Installability_Selected2} is
+  dropped because installation typically is judged as a binary result.  The
+  installation is either completed, or not.
+  \item From Definition~\ref{Installability_Selected1} the phrase ``given an
+  installation is possible to begin with'' is removed because this notion is
+  covered by the measure of effort.  The effort will be infinite if an
+  installation is impossible.
+\end{itemize}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+ \subsection{Correctness (external, internal, product, process)} %\oo{owner}
+
+\begin{defn} \label{Correctness_McCallEtAl1977}
   The extent to which a program satisfies its specifications and fulfills the
@@
 
-\begin{defn}
+\begin{defn} \label{Correctness_meyer1988object}
   The ability of software products to perform their exact tasks, as defined by
@@
 
-\begin{defn}  
+\begin{defn} \label{Correctness_IEEEComputerDictionary1991}
   The degree to which a system or component is free from faults in its
@@
 
-\begin{defn} \label{Correctness_Selected}
+\begin{defn} \label{Correctness_GhezziEtAl2003}
   A program is functionally correct if it behaves according to its stated
@@
 
-\begin{defn}
-  The degree to which a system is free from faults in its specification, design, and 
-  implementation \citep{mcconnell2004code}.
-\end{defn}
-
-\noindent \textbf{Proposed Definition}
-
-Definition~\ref{Correctness_Selected} rephrased: A program is correct if it
-behaves according to its stated specifications.
-
-\noindent \textbf{Reasoning}
-
-To determine whether a program is correct, we must be able to refer to both its
-specification, design and implementation to understand the intended purpose and
-this definition captures it all.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{\notdone{Reliability} \oo{owner}}
+\begin{defn} \label{Correctness_mcconnell2004code}
+  The degree to which a system is free from faults in its specification, design,
+  and implementation \citep{mcconnell2004code}.
+\end{defn}
+
+\begin{mybox}
+\subsubsection*{Correctness}
+An entity is correct if it matches its specification.  The entity can be either
+software, an internal product (work product or artifact) or process.  The
+specification can either be explicitly stated or implicit.
+\end{mybox}
+
+\subsubsection*{Reasoning}
+
+The selection definition is a rephrasing of
+Definition~\ref{Correctness_GhezziEtAl2003}. The different reasons for the
+proposed definition are as follows:
+
+\begin{itemize}
+  \item Definition~\ref{Correctness_McCallEtAl1977} mentions ``mission
+  objectives''.  We feel that the mission objectives are incorporated into the
+  specification, so this element is not need to be mentioned separately in the
+  recommended definition.
+  \item Definition~\ref{Correctness_GhezziEtAl2003} refers to the functional
+  specification, but we believe that software is incorrect if it does not
+  follow the nonfunctional specification as well.  For instance, for safety
+  critical software the performance requirements need to be satisfied, or the
+  software will not be trustworthy.
+  \item Definition~\ref{Correctness_IEEEComputerDictionary1991} and
+  Definition~\ref{Correctness_mcconnell2004code} are almost the same, except
+  Definition~\ref{Correctness_IEEEComputerDictionary1991} mentions ``system or
+  component'' rather than just ``system''.  To keep our recommended definition
+  simple, we do not introduce the distinction of components; we just focus on
+  the entities of the software, internal products and processes.
+  \item Several of the previous definitions imply measurement of different
+  levels of correctness by phrases such as ``The extent to which''
+  (Definition~\ref{Correctness_McCallEtAl1977}) and ``The degree to which''
+  (Definitions~\ref{Correctness_IEEEComputerDictionary1991}
+  and~\ref{Correctness_mcconnell2004code}). However, the proposed definition has
+  a binary range of possible measurement - either correct or incorrect.  This
+  doesn't follow our stated goal of having the quality of measurement cover the
+  largest possible scale type.  We chose to recommend this simpler definition of
+  correctness to satisfy the competing quality of unambiguity.
+  Definition~\ref{Correctness_IEEEComputerDictionary1991} also implies a binary
+  measure with the phrase ``the ability of''.  Making correctness a binary
+  quality makes the definition unambiguous and it frees up room for the quality
+  of reliability to cover the more nuanced aspects of the extent to which an
+  entity is ``correct''.
+  \item Internal Product,  Implied by
+  Definition~\ref{Correctness_IEEEComputerDictionary1991}.  Changed to mention
+  stated and unstated specification (explicit or implicit), since internal
+  products do not have a specification.  May have a template.
+  \item Process.
+\end{itemize}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{Reliability (external, product, process)} %\oo{owner} %NOTDONE
 
@@
 
-\begin{defn} \label{reliabilitySelected2} Reliability is concerned with measuring
-	the probability of occurence of faliure - that is, the probability of the
-	observable effects of errors \citep{GhezziEtAl2003}.
+\begin{defn} \label{reliabilitySelected2} Reliability is concerned with
+	measuring the probability of occurrence of failure - that is, the probability
+	of the observable effects of errors \citep{GhezziEtAl2003}.
 \end{defn}
@@
 
-\noindent \textbf{Proposed Definition}
-
-Definition \ref{reliabilitySelected1} and \ref{reliabilitySelected2} rephrased:
+\begin{mybox}
+\subsubsection*{Reliability}
 Probability of failure-free operation of a computer program in a specified
@@
 failures also known as the mean time to failure (MTTF).
-
-\wss{Needs to be completed.}\oo{Done}
-
-\noindent \textbf{Reasoning}
-
+\end{mybox}
+
+\subsubsection*{Reasoning}
+
+Definition \ref{reliabilitySelected1} and \ref{reliabilitySelected2} rephrased:.
 Reliability is defined such that the measurement metrics are visible - the
-specified environment and time.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{Robustness \pmi{owner}}
+specified environment and time.   The different reasons for the proposed
+definition are as follows:
+
+\begin{itemize}
+  \item reason 1
+  \item reason 2
+  \item ...
+\end{itemize}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{Robustness (external, product, process)} %\pmi{owner}
 
@@
 
-\noindent \textbf{Proposed Definition}
-
+\begin{mybox}
+\subsubsection*{Robustness}
 Definition \ref{RobustnessPartialDefn} and Definition
@@
 the assumptions in its requirements specification are violated.
-
-\noindent \textbf{Reasoning}
+\end{mybox}
+
+\subsubsection*{Reasoning}
 
 This definition indicates that robustness is related to the quality of
-correctness (both within and outside of it).
-
-\pmi{Still may want to refine what we mean by reasonable}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{\notdone{Performance} \pmi{owner}}
+correctness (both within and outside of it). \pmi{Still may want to refine what
+we mean by reasonable}  The different reasons for the proposed definition are as
+follows:
+
+\begin{itemize}
+  \item reason 1
+  \item reason 2
+  \item ...
+\end{itemize}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{Performance (external, product)} %NOTDONE \pmi{owner}
 
@@
 
-\begin{defn} \label{PerformanceDefnSelected2}
-	How well or how rapidly the system must perform specific functions. Performance
-	requirements encompass speed (database response times, for instance), throughput
-	(transactions per second), capacity (concurrent usage loads), and timing (hard
-	real-time demands) \citep{wiegers2003softreq}.
-\end{defn}
-
-\noindent \textbf{Proposed Definition}
-
+\begin{defn} \label{PerformanceDefnSelected2} How well or how rapidly the system
+	must perform specific functions. Performance requirements encompass speed
+	(database response times, for instance), throughput (transactions per second),
+	capacity (concurrent usage loads), and timing (hard real-time demands)
+	\citep{wiegers2003softreq}.
+\end{defn}
+
+\begin{mybox}
+\subsubsection*{Performance}
 Combined definition \ref{PerformanceDefnSelected} and
@@
 (database response times, for instance), throughput (transactions per second),
-capacity (concurrent usage loads), and timing (hard real-time
-demands). \wss{What is the difference between speed and timing?}
-
-\noindent \textbf{Reasoning}
+capacity (concurrent usage loads), and timing (hard real-time demands).
+\wss{What is the difference between speed and timing?}
+\end{mybox}
+
+\subsubsection*{Reasoning}
 
@@
 associated with software performance, such as speed, throughput, capacity, and
-timing.
-
-\pmi{I will check if Wiegers discusses the difference between speed and
-timing.}
-
-\pmi{Wiegers did not discuss this difference.}
-
-\pmi{Speed - change over time; timing - related to specification: when
-something should happen in relation to something else - state transitions}
-
-\wss{If we need both speed and timing, we should define them.  We aren't
-  including accuracy from the IEEE definition.  Is that on purpose?}
-
-\pmi{We have decided not to include accuracy. SS will find difference b/w
-speed and timing from a colleague.}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{\notdone{Usability} \jc{owner}}
+timing. \pmi{I will check if Wiegers discusses the difference between speed and
+timing.} \pmi{Wiegers did not discuss this difference.}  \pmi{Speed - change
+over time; timing - related to specification: when something should happen in
+relation to something else - state transitions}  \wss{If we need both speed and
+timing, we should define them.  We aren't including accuracy from the IEEE
+definition.  Is that on purpose?}  \pmi{We have decided not to include accuracy.
+SS will find difference b/w speed and timing from a colleague.}   The
+different reasons for the proposed definition are as follows:
+
+\begin{itemize}
+  \item reason 1
+  \item reason 2
+  \item ...
+\end{itemize}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{Usability (external, product, process) } %\notdone \jc{owner}
 
@@
 \end{defn}
-\noindent \textbf{Proposed Definition}
-
-\begin{defn}
+
+\begin{mybox}
+\subsubsection*{Usability}
 The extent to which a product can be used by specified users to achieve
-  specified goals with effectiveness, efficiency, and satisfaction in a
-  specified context of use. \cite{ISO16982:2002} and \cite{ISO9241:11}
-\end{defn}
-
-\noindent \textbf{Reasoning}
-
-There has been a lot of thought put behind the ISO definition, from experts.
-The standard has been revisited, but the definition has not changed, which 
-means that it has stood the test of time.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{\notdone{Verifiability} \oo{owner}}
+specified goals with effectiveness, efficiency, and satisfaction in a specified
+context of use. \cite{ISO16982:2002} and \cite{ISO9241:11}
+\end{mybox}
+
+\subsubsection*{Reasoning}
+
+There has been a lot of thought put behind the ISO definition, from experts. The
+standard has been revisited, but the definition has not changed, which means
+that it has stood the test of time.  The different reasons for the proposed
+definition are as follows:
+
+\begin{itemize}
+  \item reason 1
+  \item reason 2
+  \item ...
+\end{itemize}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{Verifiability (internal, product, process) } %\notdone \oo{owner}
 
@@
 
-\noindent \textbf{Proposed Definition} Definition~\ref{Defn_Verifiability1}
+\begin{itemize}
+
+\item Verification - Are we building the product right?  Are we implementing
+the requirements correctly (internal) \item Validation - Are we building the
+right product? Are we getting the right requirements (external) \item
+According to  \href{https://en.wikipedia.org/wiki/Software_verification_and_validation}
+{Capability Maturity Model (CMM)}
+
+\begin{itemize}
+\item Software Verification: The process of evaluating software to determine
+      whether the products of a given development phase satisfy the conditions
+      imposed at the start of that phase. [IEEE-STD-610] \wss{Need a proper
+      citation}
+\item Software Validation: The process of evaluating software during or at the
+      end of the development process to determine whether it satisfies specified
+      requirements. [IEEE-STD-610] \wss{Need a proper citation}
+\end{itemize}
+
+\end{itemize}
+
+``An SRS is verifiable if, and only if, every requirement stated therein is
+verifiable. A requirement is verifiable if, and only if, there exists some
+finite cost-effective process with which a person or machine can check that the
+software product meets the requirement. In general any ambiguous requirement is
+not verifiable.'' \citep{IEEE1998}
+
+Verifiability is related to testability, which is defined by McCall et al.\
+as ``The effort required to test a program to ensure that it performs its
+intended function'' \citep{VanVliet2000}. \wss{Need to verify McCall
+reference, then delete VanVliet.} \pmi{McCall's definition verbatim: Effort
+required to test a program to insure it performs its intended function.}
+
+\wss{When I get the Ghezzi text back from Olu, I'll check to see if they have anything to add to this definition.}
+
+\begin{defn}
+  A software system is verifiable if its properties can be verified easily
+  \citep{ghezzi1991fundamentals}.
+\end{defn}
+
+\begin{mybox}
+\subsubsection*{Verifiability} Definition~\ref{Defn_Verifiability1}
 rephrased: Verifiability is the extent to which a set of tests can be
@@
 specification.
-
-\noindent \textbf{Reasoning}
+\end{mybox}
+
+\subsubsection*{Reasoning}
 
 Definition is concise and measurable.  Verifiability involves solving the
-equations right \citep[p.~23]{Roache1998} and the definition states that
-system needs to be verifiable based on the specifications.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{\notdone{Validatability} \oo{owner}}
+equations right \citep[p.~23]{Roache1998} and the definition states that system
+needs to be verifiable based on the specifications.  When applied to process
+means ``did we apply the process correctly'', as opposed to validatability, we
+refers to ``did we apply the correct process''.   The different reasons for the
+proposed definition are as follows:
+
+\begin{itemize}
+  \item reason 1
+  \item reason 2
+  \item ...
+\end{itemize}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{Validatability (external, product, process)} %\notdone \oo{owner}
 
@@
 
-\noindent \textbf{Proposed Definition}
-
+\begin{mybox}
+\subsubsection*{Validatability}
 Definition \ref{Defn_Validatability} rephrased:  The effort required in
@@
 process to determine whether it satisfies specified requirements
-
-\noindent \textbf{Reasoning}
+\end{mybox}
+
+\subsubsection*{Reasoning}
 
 Definition describes that system should be measurable through the
-development life cycle, It is concise and can be measured.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{\notdone{Testability} \oo{owner}}
+development life cycle, It is concise and can be measured.  The
+different reasons for the proposed definition are as follows:
+
+\begin{itemize}
+  \item reason 1
+  \item reason 2
+  \item ...
+\end{itemize}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{Testability (internal, product)} %\notdone \oo{owner}
 
@@
 
-\noindent \textbf{Proposed Definition}
-
+\begin{mybox}
+\subsubsection*{Testability}
 Definition \ref{Defn_Testability}: The degree to which a system or component
@@
  to determine whether those criteria have been met.
-
-\noindent \textbf{Reasoning}
-
-Definition is concise and measurable.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{\notdone{Maintainability} \pmi{owner}}
+\end{mybox}
+
+\subsubsection*{Reasoning}
+
+Definition is concise and measurable.   The different reasons for the proposed
+definition are as follows:
+
+\begin{itemize}
+  \item reason 1
+  \item reason 2
+  \item ...
+\end{itemize}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{Maintainability (internal, product, process)} %\notdone \pmi{owner}
 
@@
 
-\noindent \textbf{Proposed Definition}
-
+\begin{mybox}
+\subsubsection*{Maintainability}
 Combined altered Definition \ref{MaintainabilityDefnSelected1} and Definition
@@
 component can be modified to:
-
 \begin{enumerate}
@@
 \end{enumerate}
-
-\noindent \textbf{Reasoning}
+\end{mybox}
+
+\subsubsection*{Reasoning}
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{Reusability \pmi{owner}}
+The different reasons for the proposed definition are as follows:
+
+\begin{itemize}
+  \item reason 1
+  \item reason 2
+  \item ...
+\end{itemize}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{Reusability (internal, product, process)} %\pmi{owner}
 
@@
 
-\noindent \textbf{Proposed Definition}
-
+\begin{mybox}
+\subsubsection*{Reusability}
 Definition \ref{ReusabilityDefnSelected}: The extent to which a software
@@
 than the one for which it was originally developed.
-
-\noindent \textbf{Reasoning}
+\end{mybox}
+
+\subsubsection*{Reasoning}
 
@@
 
-\subsection{{Portability} \pmi{owner}}
+\subsection{Portability (internal, product)} %\pmi{owner}
 
@@
 
-\noindent \textbf{Proposed Definition}
-
+\begin{mybox}
+\subsubsection*{Portability}
 Definition \ref{PortabilityDefnSelected} rephrased: Effort required to transfer
 a program between system environments (including hardware and software).
-
-\noindent \textbf{Reasoning}
-
-This is measurable and succinct.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{\notdone{Understandability} \jc{owner}}
+\end{mybox}
+
+\subsubsection*{Reasoning}
+
+This is measurable and succinct.   The different reasons for the proposed
+definition are as follows:
+
+\begin{itemize}
+  \item reason 1
+  \item reason 2
+  \item ...
+\end{itemize}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{Understandability (internal, product)} %\notdone \jc{owner}
 
@@
 
-\noindent \textbf{Proposed Definition}
-
+\begin{mybox}
+\subsubsection*{Understandability}
 \wss{Still needs to be completed}
-
-\noindent \textbf{Reasoning}
+\end{mybox}
+
+\subsubsection*{Reasoning}
 
@@
 
-\subsection{\notdone{Interoperability} \ad{owner}}
+\subsection{Interoperability (external, product)} %\notdone \ad{owner}
 
@@
 
-\noindent \textbf{Proposed Definition}
-
+\begin{mybox}
+\subsubsection*{Interoperability}
 Definition \ref{InteroperabilitySelected}.
-
-\noindent \textbf{Reasoning}
-
-This definition is concise and also detailed enough to show the concept not
-only on system, but also on products and components. It also covered the
-concept with more than 2 systems.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{\notdone{Productivity} \ad{owner}}
+\end{mybox}
+
+\subsubsection*{Reasoning}
+
+This definition is concise and also detailed enough to show the concept not only
+on system, but also on products and components. It also covered the concept with
+more than 2 systems.  The different reasons for the proposed definition are as
+follows:
+
+\begin{itemize}
+  \item reason 1
+  \item reason 2
+  \item ...
+\end{itemize}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{Productivity (process)} %\notdone \ad{owner}}
 
@@
 
-\noindent \textbf{Proposed Definition}
-
+\begin{mybox}
+\subsubsection*{Productivity}
 The revision of the combination of Definition \ref{ProductivitySelected1}
@@
 documents and the number of new test cases) produced per person-day.
-
-\noindent \textbf{Reasoning}
-
-It is concise and measurable. \wss{What is the output?  What is the input?
-I think the definition needs to give more information on these.  In
-particular, the above definitions focus on code as the output, but
-documentation, test cases etc should also be part of the output.  If we are
-going to measure this, we need a better idea of what we are measuring for
-outputs and inputs.} \ad{I added another def and made it more measurable}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{\notdone{Visibility/Transparency} \ad{owner}}
+\wss{Use Long Term Productivity paper}
+\end{mybox}
+
+\subsubsection*{Reasoning}
+
+It is concise and measurable. \wss{What is the output?  What is the input? I
+think the definition needs to give more information on these.  In particular,
+the above definitions focus on code as the output, but documentation, test cases
+etc should also be part of the output.  If we are going to measure this, we need
+a better idea of what we are measuring for outputs and inputs.} \ad{I added
+another def and made it more measurable.}   The different reasons for the
+proposed definition are as follows:
+
+\begin{itemize}
+  \item reason 1
+  \item reason 2
+  \item ...
+\end{itemize}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{Visibility/Transparency (process)} %\notdone \ad{owner}
 
@@
 
-\noindent \textbf{Proposed Definition}
-
-Definition \ref{VisibilitySelected} rephrased: The extent to which all of
-the steps of a software development process and the current status of it are
-conveyed clearly. \wss{I wonder if rather than ``documented clearly'', we
-should be more abstract and talk about how these things can be easily
-determined.  Documentation is probably how the information will be conveyed,
-  but it doesn't have to be how it is done.}  \ad{Adjusted}
-
-\noindent \textbf{Reasoning}
+\begin{mybox}
+\subsubsection*{Visibility/Transparency} 
+Definition \ref{VisibilitySelected} rephrased: The extent to which all of the
+steps of a software development process and the current status of it are
+conveyed clearly. \wss{I wonder if rather than ``documented clearly'', we should
+be more abstract and talk about how these things can be easily determined.
+Documentation is probably how the information will be conveyed, but it doesn't
+have to be how it is done.}  \ad{Adjusted}
+\end{mybox}
+
+\subsubsection*{Reasoning}
 
@@
 improve visibility. It is rephrased because the original one might refer to
-binary status - ``visible'' or ``invisible''.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{\notdone{Reproducibility} \wss{owner}}
+binary status - ``visible'' or ``invisible''.  The different reasons for the
+proposed definition are as follows:
+
+\begin{itemize}
+  \item reason 1
+  \item reason 2
+  \item ...
+\end{itemize}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{Reproducibility (external, product)} %\notdone \wss{owner}
 
@@
 
-\noindent \textbf{Proposed Definition}
-
+\begin{mybox}
+\subsubsection*{Reproducibility} 
 \wss{Needs to be completed}
-
 The importance of reproducibility is seen by retractions of influential
@@
 trials}
-
-\noindent \textbf{Reasoning}
+\end{mybox}
+
+\subsubsection*{Reasoning}
 
@@
 
-\subsection{\notdone{Sustainability} \wss{owner}}
+\subsection{Sustainability (internal, product)} %\notdone \wss{owner}
 
@@
 
-\noindent \textbf{Proposed Definition}
-
+\begin{mybox}
+\subsubsection*{Sustainability} 
 \wss{Needs to be completed}
-
-\noindent \textbf{Reasoning}
-
-\wss{Needs to be completed}
-
-\section{Desirable Qualities of Good Specifications} \label{SecDesirableQs}
-
-To achieve the qualities listed in Section~\ref{SecQualities}, the
-documentation should achieve the qualities listed in this section.  All but
-the final quality listed (abstraction), are adapted from the IEEE
-recommended practise for producing good software requirements
-\citep{IEEE1998}.  Abstraction means only revealing relevant details, which
-in a requirements document means stating what is to be achieved, but
-remaining silent on how it is to be achieved. Abstraction is an important
-software development principle for dealing with complexity \citep[p
-.~40]{GhezziEtAl2003}.  Correctness was in the above list, so it is not
-repeated here.  \citet{SmithAndKoothoor2016} present further details on the
-qualities of documentation for SCS.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{\notdone{Completeness} \ad{owner}}
+\end{mybox}
+
+\subsubsection*{Reasoning}
+
+\wss{Needs to be completed}.   The different reasons for the proposed definition
+are as follows:
+
+\begin{itemize}
+  \item reason 1
+  \item reason 2
+  \item ...
+\end{itemize}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Direct Qualities} \label{SecDirectQs}
+
+To achieve the  qualities listed in Section~\ref{SecQualities}, the internal
+products and process should achieve the qualities listed in this section.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{Completeness} %\ontdone \ad{owner}
 
@@
 
-\noindent \textbf{Proposed Definition}
-
+\begin{mybox}
+\subsubsection*{Completeness} 
 The first sentence of Definition \ref{CompletenessSelected}: A specification is
@@
 developed.
-
-\noindent \textbf{Reasoning}
-
-It is concise and measurable.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{\notdone{Consistency} \ad{owner}}
+\end{mybox}
+
+\subsubsection*{Reasoning}
+
+It is concise and measurable.  The different reasons for the proposed definition
+are as follows:
+
+\begin{itemize}
+  \item reason 1
+  \item reason 2
+  \item ...
+\end{itemize}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{Consistency} %\notdone \ad{owner}
 
@@
 
-\noindent \textbf{Proposed Definition}
-
+\begin{mybox}
+\subsubsection*{Consistency} 
 The first sentence of Definition \ref{ConsistencySelected}: A specification is
@@
 with governing specifications and objectives.
-
-\noindent \textbf{Reasoning}
-
-It is concise and measurable.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{\notdone{Modifiability} \jc{owner}}
+\end{mybox}
+
+\subsubsection*{Reasoning}
+
+It is concise and measurable.  The different reasons for the proposed definition
+are as follows:
+
+\begin{itemize}
+  \item reason 1
+  \item reason 2
+  \item ...
+\end{itemize}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{Modifiability} %\notdone \jc{owner}
 
@@
 
-\noindent \textbf{Reasoning}
-
-\wss{Needs to be completed}
-
-\subsection{\notdone{Traceability} \jc{owner}}
+\begin{mybox}
+\subsubsection*{Modifiability} 
+The first sentence of Definition \ref{ConsistencySelected}: A specification is
+consistent to the extent that its provisions do not conflict with each other or
+with governing specifications and objectives.
+\end{mybox}
+
+\subsubsection*{Reasoning}
+
+\wss{Needs to be completed}   The
+different reasons for the proposed definition are as follows:
+
+\begin{itemize}
+  \item reason 1
+  \item reason 2
+  \item ...
+\end{itemize}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{Traceability} %\notdone \jc{owner}
 
@@
 
-\noindent \textbf{Proposed Definition}
-
+\begin{mybox}
+\subsubsection*{Traceability} 
 \wss{Needs to be completed.}
-
-\noindent \textbf{Reasoning}
-
-\wss{Needs to be completed}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{\notdone{Unambiguity} \wss{owner}}
+\end{mybox}
+
+\subsubsection*{Reasoning}
+
+\wss{Needs to be completed}  The different reasons for the proposed definition
+are as follows:
+
+\begin{itemize}
+  \item reason 1
+  \item reason 2
+  \item ...
+\end{itemize}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{Unambiguity} %\notdone \wss{owner}
 
@@
 
-\noindent \textbf{Proposed Definition}
-
+\begin{mybox}
+\subsubsection*{Unambiguity} 
 \wss{Needs to be completed.}
-
-\noindent \textbf{Reasoning}
+\end{mybox}
+
+\subsubsection*{Reasoning}
 
@@
 
-\subsection{\notdone{Verifiability} \wss{owner}}
-
-\begin{itemize}
-
-\item Verification - Are we building the product right?  Are we implementing
-the requirements correctly (internal) \item Validation - Are we building the
-right product? Are we getting the right requirements (external) \item
-According to  \href{https://en.wikipedia.org/wiki/Software_verification_and_validation}
-{Capability Maturity Model (CMM)}
-  \begin{itemize}
-  \item Software Verification: The process of evaluating software to determine
-    whether the products of a given development phase satisfy the conditions
-    imposed at the start of that phase. [IEEE-STD-610] \wss{Need a proper
-      citation}
-  \item Software Validation: The process of evaluating software during or at the
-    end of the development process to determine whether it satisfies specified
-    requirements. [IEEE-STD-610] \wss{Need a proper citation}
-  \end{itemize}
-\end{itemize}
-
-``An SRS is verifiable if, and only if, every requirement stated therein is
-verifiable. A requirement is verifiable if, and only if, there exists some
-finite cost-effective process with which a person or machine can check that the
-software product meets the requirement. In general any ambiguous requirement is
-not verifiable.'' \citep{IEEE1998}
-
-Verifiability is related to testability, which is defined by McCall et al.\
-as ``The effort required to test a program to ensure that it performs its
-intended function'' \citep{VanVliet2000}. \wss{Need to verify McCall
-reference, then delete VanVliet.} \pmi{McCall's definition verbatim: Effort
-required to test a program to insure it performs its intended function.}
-
-\wss{When I get the Ghezzi text back from Olu, I'll check to see if they have
-  anything to add to this definition.}
-
-\begin{defn}
-  A software system is verifiable if its properties can be verified easily
-  \citep{ghezzi1991fundamentals}.
-\end{defn}
-
-\noindent \textbf{Proposed Definition}
-
-\wss{Needs to be completed}
-
-\noindent \textbf{Reasoning}
-
-\wss{Needs to be completed}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{\notdone{Abstract} \wss{owner}}
+\subsection{Abstract} %\notdone \wss{owner}
 
@@
 
-\noindent \textbf{Proposed Definition}
-
+\begin{mybox}
+\subsubsection*{Abstract} 
 \wss{Needs to be completed}
-
-\noindent \textbf{Reasoning}
+\end{mybox}
+
+\subsubsection*{Reasoning}
 
 \wss{Needs to be completed}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 \end{document}
+
+% possibly useful definitions
+
+\subsection{Additional Definitions}
+
+Some standards give additional definitions for terms that are used a lot.
+These are reproduced here, to aid in the interpretation of the definitions
+below (and their refinement).
+
+\begin{defn} 
+\textsf{Effectiveness}:
+The accuracy and completeness with which users achieve specified
+goals. \cite{ISO9241:11}
+\end{defn}
+
+\begin{defn} 
+\textsf{Efficiency}: 
+The resources expended in relation to the accuracy and
+completeness with which users achieve goals. \cite{ISO9241:11}
+\end{defn}
+
+\begin{defn}
+\textsf{Satisfaction}: 
+Freedom from discomfort, and positive attitude to the use of the product.
+\cite{ISO9241:11}
+\end{defn}
+
+\begin{defn}
+\textsf{Context of use}:
+characteristics of the users, tasks and the organizational and physical
+environments. \cite{ISO9241:11}
+\end{defn}
+
+\begin{defn}
+\textsf{Goal}: intended outcome. \cite{ISO9241:11}
+\end{defn}
+
+\begin{defn}
+\textsf{Task}: activities required to achieve a goal. \cite{ISO9241:11}
+\end{defn}

### 2021-10-01 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
Updates to installability rationale

@@
 \begin{abstract}
-	Existing definitions of software qualities (also called quality
-  characteristics) are collected and analyzed.  For each quality a
-  ``best'\rq{} definition is recommended based on the following criteria:
-  unambiguity, simplicity, consistency, ...
+	We have found some of the existing definitions of software qualities
+	inadequate as a basis for conducting and presenting our software engineering
+	research.  Therefore, we have endeavored to create a list of definitions that
+	are consistent (same words/phrases are always used for the same concepts),
+	measurable (all qualities are conceptually measurable, even if the data needed
+	for measurement is often unavailable), abstract (quality definitions state
+	what the quality achieves, but not how to achieve it) and traceable (explicit
+	links are given to all previous definitions and the ideas within those
+	definitions).  Our methodology starts with an exhaustive literature survey to
+	find existing definitions.  We then determine the recommended definition,
+	which is either one of the existing definitions, or a new definition, found by
+	combining existing definitions.  Each recommended definition is accompanied by
+	a rationale that explicitly justifies the definition.  By presenting the
+	rationales, we hope our software engineering colleagues will scrutinize our
+	work and provide recommendations for future modifications. The qualities that
+	are defined include installability, correctness, reliability, robustness,
+	performance, usability, verifiability, validatability, testability,
+	maintainability, reusability, portability, understandability,
+	interoperability, productivity, visibility/transparency, reproducibility,
+	sustainability, completeness, consistency, modifiability, traceability,
+	unambiguity and abstract.  Each quality is classified into one or more of the
+	following categories: direct, indirect, internal, external, product or
+	process.
 \end{abstract}
@@
 often called ``ilities.''  In the overview of managing software quality provided
-by \citet[p 110]{VanVliet2000} the different measures are called quality
+by \citet[p.~110]{VanVliet2000} the different measures are called quality
 attributes, quality factors, and quality criteria.  In \citet{ISO9126} they are
@@
 The qualities of most interest to project managers and users are summarized in
-the next subsection (\ref{Sec_IndirectQualities}).  Since these qualities cannot
-generally be measured directly, they are characterized as indirect qualities.
-The relevant indirect qualities are listed in this section, while being
-characterized as external, internal, product or process qualities.  The next
-subsection lists the relevant direct qualities that assist in achieving the
-indirect qualities.  The final subsection presents the methodology that was
-employed to summarize, analyze and define each quality.
+the next subsection (Section~\ref{Sec_IndirectQualities}).  Since these
+qualities cannot generally be measured directly, they are characterized as
+indirect qualities. The relevant indirect qualities are listed in this section
+and each quality is grouped into one or more of the following categories:
+external, internal, product or process. The next subsection lists the relevant
+direct qualities that assist in achieving the indirect qualities.  The final
+subsection presents the methodology that was employed to summarize, analyze and
+define each quality.
 
@@
 research.  This means we will include the usual software engineering suspects,
-like usability and maintainability.  In addition, since we also do research on
-scientific computing software, we will also include scientifically relevant
-qualities, like validatability and reproducibility.
+like usability and maintainability.  Since we also do research on scientific
+computing software, we will also include scientifically relevant qualities, like
+validatability and reproducibility.
 
@@
 timeliness, which was excluded because the already included quality of
-productivity overlaps with timeliness.  To the list from
+productivity sufficiently overlaps with timeliness.  To the list from
 \citet{GhezziEtAl2003} we have added the practical quality of installability,
@@
 installed.  Given our interest in scientific computing, we also added the
-qualities of validatability and reproducibility. Since we were not able to find
-much information in the literature on the quality of verifiability, we also
-added the closely related quality of testability.  The quality of sustainability
-was added since this term has emerged as a popular goal for software projects.
+qualities of validatability and reproducibility. Since we found few definitions
+in the literature on the quality of verifiability, we also added the closely
+related quality of testability.  The quality of sustainability was added since
+this term has emerged as a popular goal for software projects.
 
@@
 usability of internal documentation, the quality of usability is considered as
-external, with the quality of understandability designated to cover the internal
+external, since the quality of understandability designated to cover the internal
 equivalent of usability.
@@
 internal and external products.  The process is based on principles, methods and
-techniques.  The process is supported by tools.  All product qualities (internal
-and external) are effected by the process, so this information is not shown in
-Table~\ref{Tab_SoftwareQualities}.  Instead, the checkmarks in
-Table~\ref{Tab_SoftwareQualities} are for qualities that can be applied to a
-process.  For instance, we do discuss the installability or correctness of a
-process, but it does make sense to think about the portability of a process.
-Rather than discuss the performance of a process, we introduce the process
-specific quality of productivity.  Similarly, although we can discuss the
-usability of a process, we instead introduce the term understandability to
-exclusively apply to internal products (like the code).
+techniques while being supported by tools.  All product qualities (internal and
+external) are effected by the process, so the checkmarks in
+Table~\ref{Tab_SoftwareQualities} are used for a different purpose.  The
+checkmarks show which qualities are applied to the process itself.  For
+instance, we can consider the correctness and maintainability of a process, but
+discussing the portability of a process makes little sense. Rather than apply
+the quality of performance to a process, we instead introduce the process
+specific quality of productivity.  Although we can imagine discussing the
+understandability of a process, to avoid confusion and overlap, we reserve
+understandability for the internal quality applied to products (like code). For
+the related process quality, we designate the term usability.
 
@@
 conventional wisdom holds that the internal products should aim for the
-qualities listed below.  All but the final quality listed (abstraction), are
-adapted from the IEEE recommended practice for producing good software
-requirements \citep{IEEE1998}. Abstraction is an important software development
-principle for dealing with complexity \citep[p .~40]{GhezziEtAl2003}.
-\citet{IEEE1998} also includes correctness and verifiability, but we have
-excluded these from the list of direct qualities, since they belong in the list
-of indirect qualities in the previous section.
+qualities listed in this section.  All but the final quality listed
+(abstraction), are adapted from the IEEE recommended practice for producing good
+software requirements \citep{IEEE1998}. Abstraction is an important addition
+because it is the software development principle that allows us to deal with
+complexity \citep[p.~40]{GhezziEtAl2003}. \citet{IEEE1998} also includes
+correctness and verifiability, but we have excluded these from our list, since
+they belong in the list of indirect qualities in the previous section.
 
@@
 
-\subsection{Methodology}
+\subsection{Methodology} \label{SecMethodology}
 
@@
 definitions from \citet{McCallEtAl1977} appear in many software engineering
-textbooks).  Since these instances do not add any new information, we do not
-make an effort to identify how frequently the definitions reappear in different
-sources.  Once all the definitions are collected, we determine the definition
-that we would like to work with going forward.  This definition can either be
-our preference from the existing definitions, or a new definition, which is
-often found by combining existing definitions.
+textbooks).  Since these instances do not add any new information, they are not
+included; we do not make an effort to identify how frequently the definitions
+reappear in different sources.  Once all the definitions are collected, we
+determine the definition that we would like to recommend.  The recommended
+definition can either be our preference from the existing definitions, or a new
+definition, which is often found by combining existing definitions.
 
 To ensure that our recommended definitions are quality definitions, they are
-based on direct quality measures (\ref{SecDirectQsIntro}).  To the list of
-direct qualities, we add the quality criteria of measurability.  How the direct
-qualities are applied to develop definitions of qualities is described below.
+based on the direct quality measures (Section~\ref{SecDirectQsIntro}).  To the
+list of direct qualities, we also add the quality criteria of measurability.
+How the direct qualities are applied to develop quality definitions of qualities
+is described below.
 
@@
   that all of the relevant points from the collected definitions have been
-  included.  We also verify that all facets of the definition are covered when a
-  quality applies to multiple categories of internal, external, product and/or
-  process.  Finally, we check that all of the qualities listed here have been
+  included, or explicitly rejected in the definition's rationale.  We also
+  verify that all facets of the definition are covered when a quality applies to
+  multiple categories of internal, external, product and/or process.  Finally,
+  we check that all of the qualities listed in this section have been
   considered.
@@
   \item[Consistency] The final list of definitions should be consistent in the
-  terminology used.  For instance, we do not use synonyms for variety; we use
-  the same word to mean the same concept throughout the final list of
+  terminology used.  For instance, we do not use synonyms; we use
+  the same word or phrase to mean the same concept throughout the final list of
   definitions.  Specific decisions on terminology are as follows:
@@
     we will say \emph{the extent to which}.
-    \item Fill in as go through definitions.
+    \item \wss{Fill in additional consistency decisions, as we go through
+    definitions.}
   \end{itemize}
@@
   published definitions.  The rationale also explains why any aspects of the
-  previous definitions were ignored.  The rationale also traces the selected
-  definition to these qualities.
+  previous definitions were ignored.  The rationale also explicitly traces the
+  selected definition to the qualities in this section.
 
@@
   ambiguity, but every effort is made to keep the definitions simple and to use
-  standard, consistent, terminology.
+  standard and consistent terminology.  The rationale given for each recommended
+  definition also serves the purpose of making the definitions unambiguous.
 
@@
 
-  \item[Measurability] Although many qualities can only be indirectly measurable,
+  \item[Measurability] Although many qualities can only be indirectly measured,
   the definition of the quality should imply a measure that is conceptually
@@
   defined in \citet[p.\ 107]{VanVliet2000}.  The ratio and interval scale types
-  are preferred to the ordinal and nominal types because more they can convey
-  more information.  Specific consequences of aiming for measurability are as follows:
+  are preferred to the ordinal and nominal types because they can convey more
+  information.  Specific consequences of aiming for measurability are as
+  follows:
   \begin{itemize}
@@
     required''.
-    \item etc.
+    \item \wss{Add to this list as measurability specific decisions are made.}
   \end{itemize}
@@
 calculated and the true solution, can rarely be calculated since the true
-solution is rarely known. Even though it can usually only be estimated, the
-concept of true error is still a concept that is relevant for analyzing and
-understanding the behaviour of numerical algorithms.
+solution is generally unknown. Even though true error can only be estimated, the
+concept of true error is integral to analyzing and understanding the behaviour
+of numerical algorithms.
 
@@
 
-The recommended definition comes from combining Definitions
-\ref{Installability_Selected2} and \ref{Installability_Selected1}.  The
-different reasons for the proposed definition are as follows:
-
-\begin{itemize}
-  \item For consistency uses the term \textbf{effort} in place of efficiency.
-  \item For measurability avoids the term ``the capability of'' from
-  Definition~\ref{Installability_ISO9126}.
+The recommended definition comes from combining
+Definitions~\ref{Installability_Selected2} and \ref{Installability_Selected1}.
+The rationales for the proposed definition are as follows:
+
+\begin{itemize}
+  \item For consistency we use the term \textbf{effort} in place of efficiency.
   \item The ``effectiveness'' from Definition~\ref{Installability_Selected2} is
-  dropped because installation typically is judged as a binary result.  The
-  installation is either completed, or not.
+  dropped because the success or failure of an installation is typically judged
+  as a binary result.  The installation is generally either completed, or not.
+  \item Although we are okay with a binary result for success or failure of an
+  installation, the measure of effort required to complete an installation is
+  not a binary measure.  For this reason, we do not use the term ``the
+  capability of'' (from Definition~\ref{Installability_ISO9126}), so that we can
+  be consistent with our other recommended definitions where we avoid that term
+  for measurability reasons.
   \item From Definition~\ref{Installability_Selected1} the phrase ``given an
@@
 The selection definition is a rephrasing of
-Definition~\ref{Correctness_GhezziEtAl2003}. The different reasons for the
-proposed definition are as follows:
+Definition~\ref{Correctness_GhezziEtAl2003}.  All of the other definitions
+(Definition~\ref{Correctness_McCallEtAl1977},
+Definition~\ref{Correctness_meyer1988object},
+Definition~\ref{Correctness_IEEEComputerDictionary1991} and
+Definition~\ref{Correctness_mcconnell2004code}) also mention specification.  The
+rationale for the proposed definition follows:
 
@@
   \item Definition~\ref{Correctness_GhezziEtAl2003} refers to the functional
-  specification, but we believe that software is incorrect if it does not
-  follow the nonfunctional specification as well.  For instance, for safety
-  critical software the performance requirements need to be satisfied, or the
-  software will not be trustworthy.
+  specification, but we believe that correct software also follows the
+  nonfunctional specification.  For instance, for safety critical software the
+  performance requirements need to be satisfied, or the software will not be
+  trustworthy and therefore not correct.
   \item Definition~\ref{Correctness_IEEEComputerDictionary1991} and
@@
   and~\ref{Correctness_mcconnell2004code}). However, the proposed definition has
-  a binary range of possible measurement - either correct or incorrect.  This
-  doesn't follow our stated goal of having the quality of measurement cover the
-  largest possible scale type.  We chose to recommend this simpler definition of
-  correctness to satisfy the competing quality of unambiguity.
+  a binary range of possible measurement -- either correct or incorrect.  This
+  doesn't follow our stated goal (in Section~\ref{SecMethodology}) of having the
+  quality of measurement cover the largest possible scale type.  Like the
+  recommended definition,
   Definition~\ref{Correctness_IEEEComputerDictionary1991} also implies a binary
@@
   quality makes the definition unambiguous and it frees up room for the quality
-  of reliability to cover the more nuanced aspects of the extent to which an
-  entity is ``correct''.
-  \item Internal Product,  Implied by
-  Definition~\ref{Correctness_IEEEComputerDictionary1991}.  Changed to mention
-  stated and unstated specification (explicit or implicit), since internal
-  products do not have a specification.  May have a template.
-  \item Process.
+  of reliability to cover the more nuanced aspects of how an entity is
+  ``correct''.
+  \item Like the recommended definition,
+  Definition~\ref{Correctness_IEEEComputerDictionary1991} implies that
+  correctness can be judged for products.
+  Definition~\ref{Correctness_IEEEComputerDictionary1991} mentions the
+  ``specification, design and implementation''.
+  \item So that our definition of correctness can be abstract and thus also
+  apply to internal products, the definition mentions internal products, and
+  states that the specification can be implicit.  To be correct an internal
+  product needs to follow any specified template or documentation conventions.
+  Moreover, although they may not be stated explicitly, the requirements for a
+  document will be implicitly specified in the minds of the products
+  authors/creators.
+  \item So that our definition of correctness can be abstract and thus also
+  apply to processes, the definition mentions processes and states the
+  specification can be implicit.  If the rules of the process are explicitly
+  specified, then developers can check whether the rules have been followed.
+  Even when the rules are not stated in writing, the developers will have
+  implicit (possibly changing and inconsistent) rules in their minds.
+  \item The definition for correctness allows the specification to be implicit,
+  even for the external product of the software itself.  Not every project
+  writes down the requirements, but the requirements still exist inside the
+  heads of the developers.  This is why testing can be done for software, even
+  when the developers do not specify their requirements.
 \end{itemize}
@@
 \subsubsection*{Robustness}
-Definition \ref{RobustnessPartialDefn} and Definition
-\ref{RobustnessDefnSelected} rephrased: Software possesses the characteristic of
-robustness if it behaves ``reasonably'' in two situations: i) when it encounters
-circumstances not anticipated in the requirements specification; and ii) when
-the assumptions in its requirements specification are violated.
+Definition~\ref{RobustnessPartialDefn} and
+Definition~\ref{RobustnessDefnSelected} rephrased: Software possesses the
+characteristic of robustness if it behaves ``reasonably'' in two situations: i)
+when it encounters circumstances not anticipated in the requirements
+specification; and ii) when the assumptions in its requirements specification
+are violated.
 \end{mybox}
@@
 \subsubsection*{Performance}
-Combined definition \ref{PerformanceDefnSelected} and
-\ref{PerformanceDefnSelected2}: The degree to which a system or component
-accomplishes its designated functions within given constraints, such as speed
-(database response times, for instance), throughput (transactions per second),
-capacity (concurrent usage loads), and timing (hard real-time demands).
+The degree to which a system or component accomplishes its designated functions
+within given constraints, such as speed (database response times, for instance),
+throughput (transactions per second), capacity (concurrent usage loads), and
+timing (hard real-time demands).
 \wss{What is the difference between speed and timing?}
@@
 
-This definition offers a comprehensive list of constraints that are commonly
-associated with software performance, such as speed, throughput, capacity, and
-timing. \pmi{I will check if Wiegers discusses the difference between speed and
-timing.} \pmi{Wiegers did not discuss this difference.}  \pmi{Speed - change
-over time; timing - related to specification: when something should happen in
-relation to something else - state transitions}  \wss{If we need both speed and
-timing, we should define them.  We aren't including accuracy from the IEEE
-definition.  Is that on purpose?}  \pmi{We have decided not to include accuracy.
-SS will find difference b/w speed and timing from a colleague.}   The
-different reasons for the proposed definition are as follows:
+The definition is found by combining Definition~\ref{PerformanceDefnSelected}
+and Definition~\ref{PerformanceDefnSelected2}.  This definition offers a
+comprehensive list of constraints that are commonly associated with software
+performance, such as speed, throughput, capacity, and timing. \pmi{I will check
+if Wiegers discusses the difference between speed and timing.} \pmi{Wiegers did
+not discuss this difference.}  \pmi{Speed - change over time; timing - related
+to specification: when something should happen in relation to something else -
+state transitions}  \wss{If we need both speed and timing, we should define
+them.  We aren't including accuracy from the IEEE definition.  Is that on
+purpose?}  \pmi{We have decided not to include accuracy. SS will find difference
+b/w speed and timing from a colleague.}   The different reasons for the proposed
+definition are as follows:
 
@@
 \subsubsection*{Completeness} 
-The first sentence of Definition \ref{CompletenessSelected}: A specification is
+The first sentence of Definition~\ref{CompletenessSelected}: A specification is
 complete to the extent that all of its parts are present and each part is fully
@@
 \subsubsection*{Consistency} 
-The first sentence of Definition \ref{ConsistencySelected}: A specification is
+The first sentence of Definition~\ref{ConsistencySelected}: A specification is
 consistent to the extent that its provisions do not conflict with each other or
@@
 \subsubsection*{Modifiability} 
-The first sentence of Definition \ref{ConsistencySelected}: A specification is
+The first sentence of Definition~\ref{ConsistencySelected}: A specification is
 consistent to the extent that its provisions do not conflict with each other or

### 2021-10-04 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
Minor update to unambiguous quality section

@@
 
-A specification is unambiguous when it has a unique interpretation.  If
-there is a possibility that two readers will have two different
-interpretations, than the specification is ambiguous.  \wss{When I get the
-Ghezzi text back from Olu, I'll check to see if they have anything to add to
-this definition.}
-
+\begin{defn}
  A Software Requirements Specification (SRS) is unambiguous if, and only if,
  every requirement stated therein has only one interpretation \citep{IEEE1998}.
+\end{defn}
 
@@
 
-\wss{Needs to be completed}
+\wss{Needs to be completed}  The different reasons for the proposed definition
+are as follows:
+
+\begin{itemize}
+  \item reason 1
+  \item reason 2
+  \item ...
+\end{itemize}
 

### 2021-10-06 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
Completed the quality of performance, explicitly indicated \notdone{}

@@
 
-\subsection{Reliability (external, product, process)} %\oo{owner} %NOTDONE
+\subsection{\notdone{Reliability (external, product, process)}} %\oo{owner} %NOTDONE
 
@@
 Reliability is defined such that the measurement metrics are visible - the
-specified environment and time.   The different reasons for the proposed
-definition are as follows:
+specified environment and time.   \wss{Accuracy should be mentioned here, since
+it is related to how close the solution is to the correct solution.  Accuracy is
+not part of performance.  Maybe make accuracy part of correctness?}  The
+different reasons for the proposed definition are as follows:
 
@@
 
-\subsection{Robustness (external, product, process)} %\pmi{owner}
+\subsection{\notdone{Robustness (external, product, process)}} %\pmi{owner}
 
@@
 
-\begin{defn}
-	In software engineering we often equate performance with efficiency. A
-  software system is efficient if it uses computing resources economically
-  \citep{ghezzi1991fundamentals}.  \wss{Change to 2003 reference?}
+\begin{defn} \label{PerformanceGhezzieEtAl2003}
+	Performance ... is an external quality based on user requirements.
+  \citep[p.\ 20]{GhezziEtAl2003}.
 \end{defn}
@@
 \subsubsection*{Performance}
-The degree to which a system or component accomplishes its designated functions
-within given constraints, such as speed (database response times, for instance),
-throughput (transactions per second), capacity (concurrent usage loads), and
-timing (hard real-time demands).
-\wss{What is the difference between speed and timing?}
+The extent to which a system or component accomplishes its functional
+requirements within the given constraints of speed (database response times,
+for instance), throughput (transactions per second), capacity (concurrent usage
+loads), and timing (hard real-time demands).
 \end{mybox}
@@
 The definition is found by combining Definition~\ref{PerformanceDefnSelected}
-and Definition~\ref{PerformanceDefnSelected2}.  This definition offers a
-comprehensive list of constraints that are commonly associated with software
-performance, such as speed, throughput, capacity, and timing. \pmi{I will check
-if Wiegers discusses the difference between speed and timing.} \pmi{Wiegers did
-not discuss this difference.}  \pmi{Speed - change over time; timing - related
-to specification: when something should happen in relation to something else -
-state transitions}  \wss{If we need both speed and timing, we should define
-them.  We aren't including accuracy from the IEEE definition.  Is that on
-purpose?}  \pmi{We have decided not to include accuracy. SS will find difference
-b/w speed and timing from a colleague.}   The different reasons for the proposed
-definition are as follows:
-
-\begin{itemize}
-  \item reason 1
-  \item reason 2
-  \item ...
-\end{itemize}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{Usability (external, product, process) } %\notdone \jc{owner}
+and Definition~\ref{PerformanceDefnSelected2}.  The two definitions are fairly
+similar.  We liked the structure of Definition~\ref{PerformanceDefnSelected} and
+the more detailed list of performance aspects in
+Definition~\ref{PerformanceDefnSelected2}.  The specific reasons for the
+proposed definition are as follows:
+
+\begin{itemize}
+  \item We say ``the extent to which'' to be consistent with our other
+  definitions.
+  \item We changed ``designated functions'' to ``functional requirements'' for
+  the goal of being unambiguous.  We already introduced the distinction between
+  functional and nonfunctional requirements in the discussion of correctness.
+  This also connects the proposed definition to the user requirements, as is
+  done by Definition~\ref{PerformanceGhezzieEtAl2003}.
+  \item To remove ambiguity, we explicitly list all aspects of performance
+  (speed, throughput, capacity and timing), instead of introducing a list that
+  is implied to be incomplete via the phrase ``such as''.
+  \item We do not include accuracy (from
+  Definition~\ref{PerformanceDefnSelected}) because we consider accuracy to be
+  related to correctness and reliability.
+\end{itemize}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{\notdone{Usability (external, product, process)}} %\notdone \jc{owner}
 
@@
 
-\subsection{Verifiability (internal, product, process) } %\notdone \oo{owner}
+\subsection{\notdone{Verifiability (internal, product, process)}} %\notdone \oo{owner}
 
@@
 
-\subsection{Validatability (external, product, process)} %\notdone \oo{owner}
+\subsection{\notdone{Validatability (external, product, process)}} %\notdone \oo{owner}
 
@@
 
-\subsection{Testability (internal, product)} %\notdone \oo{owner}
+\subsection{\notdone{Testability (internal, product)}} %\notdone \oo{owner}
 
@@
 
-\subsection{Maintainability (internal, product, process)} %\notdone \pmi{owner}
+\subsection{\notdone{Maintainability (internal, product, process)}} %\notdone \pmi{owner}
 
@@
 
-\subsection{Reusability (internal, product, process)} %\pmi{owner}
+\subsection{\notdone{Reusability (internal, product, process)}} %\pmi{owner}
 
@@
 
-\subsection{Portability (internal, product)} %\pmi{owner}
+\subsection{\notdone{Portability (internal, product)}} %\pmi{owner}
 
@@
 
-\subsection{Understandability (internal, product)} %\notdone \jc{owner}
+\subsection{\notdone{Understandability (internal, product)}} %\notdone \jc{owner}
 
@@
 
-\subsection{Interoperability (external, product)} %\notdone \ad{owner}
+\subsection{\notdone{Interoperability (external, product)}} %\notdone \ad{owner}
 
@@
 
-\subsection{Productivity (process)} %\notdone \ad{owner}}
+\subsection{\notdone{Productivity (process)}} %\notdone \ad{owner}}
 
@@
 
-\subsection{Visibility/Transparency (process)} %\notdone \ad{owner}
+\subsection{\notdone{Visibility/Transparency (process)}} %\notdone \ad{owner}
 
@@
 
-\subsection{Reproducibility (external, product)} %\notdone \wss{owner}
+\subsection{\notdone{Reproducibility (external, product)}} %\notdone \wss{owner}
 
@@
 
-\subsection{Sustainability (internal, product)} %\notdone \wss{owner}
+\subsection{\notdone{Sustainability (internal, product)}} %\notdone \wss{owner}
 
@@
 
-\subsection{Completeness} %\ontdone \ad{owner}
+\subsection{\notdone{Completeness}} %\ontdone \ad{owner}
 
@@
 
-\subsection{Consistency} %\notdone \ad{owner}
+\subsection{\notdone{Consistency}} %\notdone \ad{owner}
 
@@
 
-\subsection{Modifiability} %\notdone \jc{owner}
+\subsection{\notdone{Modifiability}} %\notdone \jc{owner}
 
@@
 
-\subsection{Traceability} %\notdone \jc{owner}
+\subsection{\notdone{Traceability}} %\notdone \jc{owner}
 
@@
 
-\subsection{Unambiguity} %\notdone \wss{owner}
+\subsection{\notdone{Unambiguity}} %\notdone \wss{owner}
 
@@
 
-\subsection{Abstract} %\notdone \wss{owner}
+\subsection{\notdone{Abstract}} %\notdone \wss{owner}
 

### 2021-10-06 StateOfPractice/DomainMeasurements/LBM-Peter/Commonality-Analysis-LB-Systems.tex
Rearrange domain measurement reports to have more consistent naming and locations

--- 2020-04-03 StateOfPractice/Peter-Notes/Commonality-Analysis-LB-Systems.tex
@@
 It then moves the particles into the next node, and transfers the energy and momentum if a collision occurs.
-Then the process repeats for the duration of the modeling instance.
+Then the process repeats for the duration of the modeling instance.\\
+
+\begin{figure}[h!]
+	\begin{center}
+		\includegraphics[width=0.5\textwidth]{FeatureDiagramLBS}
+		\caption{LBS Model Feature Diagram}
+		\label{LBSFeatureDiagram}
+	\end{center}
+\end{figure}
 
@@
 \noindent\textbf{3D}: 3-Dimensional\\
+
+\noindent\textbf{BGK}: Bhatnagar-Gross-Krook\\
+
+\noindent\textbf{CPU}: central processing unit\\
+
+\noindent\textbf{GPU}: graphics processing unit\\
+
+\noindent\textbf{MRT}: multi-relaxation-time\\
+
+\noindent\textbf{SRT}: single-relaxation-time\\
+
+\noindent\textbf{TRT}: two-relaxation-time\\
 
@@
 		\hline
-		\bf Description & LBS shall be efficiently easy to use.\\
+		\bf Description & The use of LBS is easy and efficient.\\
 		\hline
@@
 		\hline
-		\bf Description & Various Transport Equations can be used to describe the statistical behaviour of the system\\
+		\bf Description & Various transport equations can be used to describe the statistical behaviour of the system\\
 		\hline
@@
 		\hline
-		\bf Range of Parameters & If parallelism is used, commonly OpenMP, OpenCL, CUDA, or MPI are used.\\
+		\bf Range of Parameters & OpenMP, OpenCL, CUDA, or MPI are used if LBS is parallel.\\
 		\hline

### 2021-10-06 StateOfPractice/DomainMeasurements/LBM-Peter/Commonality-Analysis-Template.tex
Rearrange domain measurement reports to have more consistent naming and locations

--- 2020-04-03 StateOfPractice/Peter-Notes/Commonality-Analysis-LB-Systems.tex
@@
 \section{Introduction}
-Lattice Boltzmann systems are well suited to development as a program family because they fit Parnas’s definition of a program family: “a set of programs whose common properties are so extensive that it is advantageous to study the common properties of the programs before analyzing individual members” \citep{parnas1976design}.
-Developing such systems as a program family is advantageous because Lattice Boltzmann systems share many common features, or commonalities.
+Mesh generating systems are well suited to development as a program family because they fit Parnas’s definition of a program family: “a set of programs whose common properties are so extensive that it is advantageous to study the common properties of the programs before analyzing individual members” \citep{parnas1976design}.
+Developing mesh generating systems as a program family is advantageous because mesh generators share many common features, or commonalities.
 Furthermore, when there are differences between systems, the variabilities between them can be systematically considered.
-The purpose of this document is to record these commonalities and variabilities and show the relationships between them, and thus facilitate the specification and development of Lattice Boltzmann program family members.
+The purpose of this document
+is to record these commonalities and variabilities and show the relationships between them, and thus facilitate the specification and development of mesh generator program family members.
 This document will be valuable in all future phases of system development and analysis.
-For instance, the requirements documentation for any Lattice Boltzmann systems will use the commonality analysis, since the requirements should refine the commonalities, which are shared requirements of all such systems.
-Moreover, the design of any future system will use this documentation to facilitate consideration of the variabilities, so that likely changes can be made to the system with a minimal amount of work.
-
-The scope of the commonality analysis presented here can be considered both from the point of view of Lattice Boltzmann systems and from the point of view of software engineering methodologies.
-From the Lattice Boltzmann system perspective, the starting point for the current document is a commonality analysis.
+For instance, the requirements documentation for any mesh generator will use the commonality analysis, since the requirements should refine the commonalities, which are shared requirements of all mesh generators.
+Moreover, the design of any future system will use this documentation to facilitate consideration of the variabilities, so that likely
+changes can be made to the system with a minimal amount of work.
+
+The scope of the commonality analysis presented here can be considered both from the point of view of mesh generating systems and from the point of view of software engineering methodologies.
+From the mesh generator perspective, the starting point for the current document is the commonality analysis conducted by \cite{chen2003}.
+However, the scope of the current document is broader than that of \cite{chen2003}, which was restricted to two-dimensional meshes and did not consider non-conformal, hybrid or mixed meshes.
+The current commonality analysis is intended to cover all mesh generators that are targeted at finite element applications.
+Meshes for cartography or other uses are not explicitly
+considered, although much of the information in this document does overlap with other mesh uses.
 With respect to software engineering methodologies, the scope of the current report is restricted to informal methods, with the intention that the informal requirements will form a starting point for later development and refinement by formal methods.
 
-The first section below provides an overview of the program family of Lattice Boltzmann systems, by reviewing what is involved in a commonality analysis and the basics of Lattice Boltzmann systems.
+The first section below provides an overview of the program family of mesh generators, by reviewing what is involved in a commonality analysis and the basics of mesh generating systems.
 After this, the basic terminology and definitions necessary for understanding the remainder of the document are provided.
-The definitions include terms used in describing a commonality analysis and terms that are used in defining the characteristics and properties of Lattice Boltzmann methods.
+The definitions include terms used in describing a commonality analysis and terms that are used in defining the characteristics and properties of meshes.
 The next three sections consist of the lists of commonalities, variabilities and parameters of variation, respectively.
@@
 \section{Overview}
-This section provides both an overview of the process of commonality analysis and of Lattice Boltzmann methods.
+This section provides both an overview of the process of commonality analysis and of mesh generation.
 The subsection on commonality analysis briefly introduces this topic, along with references that can be searched for further information.
-The subsection on Lattice Boltzmann methods outlines the basics of Lattice Boltzmann methods and of the scope of the commonality analysis.
+The subsection on mesh generators outlines the following: \textit{i}) the basics of mesh generation, \textit{ii}) the scope of the current analysis, \textit{iii}) the overall philosophy that has been adopted for the commonality analysis, and \textit{iv}) the arguments in favour of the development of mesh generators as a program family.
 
@@
 
-The next section will show that the above uses of the commonality analysis document will be beneficial for the development of a family of Lattice Boltzmann programs.
+The next section will show that the above uses of the commonality analysis document will be beneficial for the development of a program family of mesh generators.
 The commonality analysis will benefit all subsequent stages in the software development process.
@@
 The stages are completed through the aid of a moderator and a series of meeting and preliminary documents and documentation reviews.
-The approach adopted here is to revise the original commonality analysis document produced by \cite{chen2003} and then make the new document available to others for review.
+Although the systematic approach advocated by Weiss has its advantages, for the case of writing a commonality analysis document for mesh generating systems it was decided that a less structured approach is feasible.
+Mesh generators are simpler than other software systems in the sense that they have fewer interactions with the environment.
+Also, the theory of mesh generation has a solid mathematical basis that can be used to remove some of the ambiguity that Weiss’s approach is aimed at reducing.
+Therefore, the approach adopted here is to revise the original commonality analysis document produced by \cite{chen2003} and then make the new document available to others for review.
 The new document will be maintained in a concurrent versioning system repository so that multiple authors can work on it, and more importantly, so that the documentation’s revision history can be tracked and the documentation can be rolled back to an earlier version if necessary.
@@
 \subsection{Lattice Boltzmann Systems}
-Lattice Boltzmann Methods (LBM) are a family of fluid dynamics algorithms for simulating single-phase and multiphase fluid flows, often incorporating additional physical complexities \citep{chen1998lattice}.
-They consider the behaviours of a collection of particles as a single unit at the mesoscopic scale.
-These methods predict the positional probability of a collection of particles moving through a lattice structure.
-Off the shelf (OTS) Lattice Boltzmann Solvers (LBS) allow for a range of fluid and physical model input parameters, computational parameters, and output parameters.
-
-As LBS model fluid dynamics within a boundary using a predefined lattice structure, the methods rely on a two step calculation process.
-The first process is streaming, where the particles move along the lattice via links.
-The second process is collision, where energy and momentum is transferred among particles that collide \citep{bao2011lattice}.
-There are many standardized lattice models - individual solvers within the family might only use a subset of them.
-The LBM uses the initial parameters of the fluid to find the probability of where along the lattice linkages a group of particles are most likely to travel.
-It then moves the particles into the next node, and transfers the energy and momentum if a collision occurs.
-Then the process repeats for the duration of the modeling instance.
-
 \newpage
@@
 This section is divided into two subsections.
-The first discusses the terminology that comes from the software engineering field, while the second presents the definitions used in Lattice Boltzmann solvers.
+The first discusses the terminology that comes from the software engineering field, while the second presents the definitions used in mesh generation.
 Common acronyms are also listed in this section.
@@
 \subsection{Lattice Boltzmann Related Definitions and Acronyms}
-
-\noindent\textbf{1D} 1-Dimensional\\
-
-\noindent\textbf{2D}: 2-Dimensional\\
-
-\noindent\textbf{3D}: 3-Dimensional\\
-
-\noindent\textbf{LBM}: Lattice Boltzmann Methods\\
-
-\noindent\textbf{LBS}: Lattice Boltzmann Solvers\\
-
-\noindent\textbf{Velocity Directions}: The number of links connecting to each lattice node in the chosen model from neighbouring nodes.
-All nodes in a chosen lattice model will have the same number of links.
-A single link will connect between two adjacent nodes.
-
 \newpage
@@
 This section lists all the common features among all the potential family members.
-The commonalities are organized using the following abstraction of the system, which can be used to describe all Lattice Boltzmann systems: input information, then generate the simulation and finally output the results.
-Section \ref{comm_lbm} describes the commonalities for the simulation step.
-Section \ref{comm_in} highlights the input information that is required for all Lattice Boltzmann systems.
-The next section, Section \ref{comm_out}, shows the common features for the output of Lattice Boltzmann systems, such as the requirement that mesh information be written to files.
-(Although the output information could simply be written to the computer’s memory, in all practical applications it is desirable to have a persistent record of the output that was created.) The final section covers qualities of the system that cannot be classified as input, simulation generation or output.
+The commonalities are organized using the following abstraction of the system, which can be used to describe all mesh generators: input information, then generate the mesh and finally output the results.
+Section \ref{comm_lbm} describes the commonalities for the mesh generation step, which includes the discretization of the domain, as well as other information on the problem such as the boundary conditions, material properties, \textit{etc}.
+Section \ref{comm_in} highlights the input information that is required for all mesh generators, such as the geometry of the domain that is going to be discretized.
+The next section, Section \ref{comm_out}, shows the common features for the output of mesh generators, such as the requirement that mesh information be written to files.
+(Although the mesh information could simply be written to the computer’s memory, in all practical applications it is desirable to have a persistent record of the mesh that was created.) The final section covers qualities of the system that cannot be classified as input, mesh generation or output.
 These commonalities are termed nonfunctional requirements of the system.
@@
 
+The commonalities listed in this section are for a mesh generator that is acting as a preprocessor for a finite element analysis program, which means that the mesh generator will produce data files with information on the discretization of the domain, the boundary conditions, the material properties, the system properties, etc.
+In some cases the mesh generating systems only focus on the discretization of the domain and leave it to another system to provide the other information necessary for the finite element analysis.
+For systems that only focus on the discretization of the domain the following subset of the commonalities applies: .
+
 \subsection{Lattice Boltzmann Method Solvers}\label{comm_lbm}
@@
 		\hline
-		\bf Item Number& C\refstepcounter{comnum}\thecomnum{}\label{lattice} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
-		\hline
-		\bf Description & A lattice discretizes a computational domain into a finite number of points. All LBS discretize the computational domain using a regular, evenly spaced grid within a boundary.\\
-		\hline
-		\bf Related Variability & \vref{numveldir} \vref{varcolop} \vref{varboundary}\\
-		\hline
-		\bf History & Created April 1, 2020 \\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-~\newline\\
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
-		\hline
-		\bf Item Number& C\refstepcounter{comnum}\thecomnum\label{collisionoperator} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
-		\hline
-		\bf Description & All Lattice Boltzmann versions use a collision operator which concerns collisions between particles.\\
-		\hline
-		\bf Related Variability & \vref{dimensions} \vref{numveldir}\\
-		\hline
-		\bf History & Created April 1, 2020\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-~\newline\\
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
-		\hline
-		\bf Item Number& C\refstepcounter{comnum}\thecomnum\label{probabilitydensityfunc} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
-		\hline
-		\bf Description & All Lattice Boltzmann versions user a probability density function to give the probability that fluid has moved into a specific domain.\\
-		\hline
-		\bf Related Variability & \vref{dimensions}\\
-		\hline
-		\bf History & Created April 1, 2020\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
-		\hline
-		\bf Item Number& C\refstepcounter{comnum}\thecomnum\label{equildistfunc} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
-		\hline
-		\bf Description & Every Lattice Boltzmann version uses an equilibrium distribution function to capture the probability distribution of particles.\\
-		\hline
-		\bf Related Variability & \vref{varedf} \vref{edfstorage} \vref{edfcoeff}\\
-		\hline
-		\bf History & Created April 1, 2020\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-~\newline\\
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
-		\hline
-		\bf Item Number& C\refstepcounter{comnum}\thecomnum\label{Boltzmanntransequ} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
-		\hline
-		\bf Description & Every Lattice Boltzmann version uses a Boltzmann transport equation to describe the statistical behaviour of a system that does not have collisions.\\
-		\hline
-		\bf Related Variability & \vref{vartraneq}\\
-		\hline
-		\bf History & Created April 1, 2020\\
+		\bf Item Number& C\refstepcounter{comnum}\thecomnum \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
+		\hline
+		\bf Description & \\
+		\hline
+		\bf Related Variability & \vref{}\\
+		\hline
+		\bf History & \\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+~\newline\\
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
+		\hline
+		\bf Item Number& C\refstepcounter{comnum}\thecomnum \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
+		\hline
+		\bf Description & \\
+		\hline
+		\bf Related Variability & \vref{}\\
+		\hline
+		\bf History & \\
 		\hline
@@
 		\hline
-		\bf Item Number& C\refstepcounter{comnum}\thecomnum\label{fluidandboundaryinfo} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
-		\hline
-		\bf Description & The LBS require fluid, model, and boundary information for the problem.\\
-		\hline
-		\bf Related Variability & \vref{interface}\\
-		\hline
-		\bf History & Created April 1, 2020\\
+		\bf Item Number& C\refstepcounter{comnum}\thecomnum \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
+		\hline
+		\bf Description & \\
+		\hline
+		\bf Related Variability & \vref{}\\
+		\hline
+		\bf History & \\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+~\newline\\
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
+		\hline
+		\bf Item Number& C\refstepcounter{comnum}\thecomnum \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
+		\hline
+		\bf Description & \\
+		\hline
+		\bf Related Variability & \vref{}\\
+		\hline
+		\bf History & \\
 		\hline
@@
 		\hline
-		\bf Item Number& C\refstepcounter{comnum}\thecomnum\label{predictionstomemory} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
-		\hline
-		\bf Description & LBS write fluid parameter predictions to memory.\\
-		\hline
-		\bf Related Variability & \vref{varvisual} \vref{formatoutput} \\
-		\hline
-		\bf History & Created April 1, 2020\\
+		\bf Item Number& C\refstepcounter{comnum}\thecomnum \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
+		\hline
+		\bf Description & \\
+		\hline
+		\bf Related Variability & \vref{}\\
+		\hline
+		\bf History & \\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+~\newline\\
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
+		\hline
+		\bf Item Number& C\refstepcounter{comnum}\thecomnum \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
+		\hline
+		\bf Description & \\
+		\hline
+		\bf Related Variability & \vref{}\\
+		\hline
+		\bf History & \\
 		\hline
@@
 		\hline
-		\bf Item Number& C\refstepcounter{comnum}\thecomnum\label{precision} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
-		\hline
-		\bf Description & LBS provides the precision required for the particular problems it is intended to solve.\\
-		\hline
-		\bf Related Variability & \vref{varprecise}\\
-		\hline
-		\bf History & Created April 1, 2020\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-~\newline\\
-~\newline
-
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
-		\hline
-		\bf Item Number& C\refstepcounter{comnum}\thecomnum\label{accuracy} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
-		\hline
-		\bf Description & LBS provides the accuracy required for the particular problems it is intended to solve.\\
-		\hline
-		\bf Related Variability & \vref{varaccuracy}\\
-		\hline
-		\bf History & Created April 1, 2020\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-~\newline\\
-~\newline
-
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
-		\hline
-		\bf Item Number& C\refstepcounter{comnum}\thecomnum\label{performance} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
-		\hline
-		\bf Description & The response time is small enough to allow the user to focus on their problem without being distracted by excessive waiting times.\\
-		\hline
-		\bf Related Variability & \vref{varresponsetime}\\
-		\hline
-		\bf History & Created April 1, 2020\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-~\newline\\
-~\newline
-
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
-		\hline
-		\bf Item Number& C\refstepcounter{comnum}\thecomnum\label{portability} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
-		\hline
-		\bf Description & LBS will be as portable to other operating systems as required by the users.\\
-		\hline
-		\bf Related Variability & \vref{varportability}\\
-		\hline
-		\bf History & Created April 1, 2020\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-~\newline\\
-~\newline
-
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
-		\hline
-		\bf Item Number& C\refstepcounter{comnum}\thecomnum\label{usability} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
-		\hline
-		\bf Description & LBS shall be efficiently easy to use.\\
-		\hline
-		\bf Related Variability & \vref{varusability}\\
-		\hline
-		\bf History & Created April 1, 2020\\
+		\bf Item Number& C\refstepcounter{comnum}\thecomnum \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
+		\hline
+		\bf Description & \\
+		\hline
+		\bf Related Variability & \vref{}\\
+		\hline
+		\bf History & \\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+~\newline\\
+~\newline
+
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
+		\hline
+		\bf Item Number& C\refstepcounter{comnum}\thecomnum \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
+		\hline
+		\bf Description & \\
+		\hline
+		\bf Related Variability & \vref{}\\
+		\hline
+		\bf History & \\
 		\hline
@@
 		\hline
-		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{parallel} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & LBS may use a framework for parallel processing of the model.\\
-		\hline
-		\bf Related Commonality & None\\
-		\hline
-		\bf Related Parameter & \pref{parparallel}\\
-		\hline
-		\bf History & Created April 1, 2020\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-~\newline\\
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
-		\hline
-		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{varedf} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & Different versions of an equilibrium distribution function can capture the probability distribution of particles.\\
-		\hline
-		\bf Related Commonality & \cref{equildistfunc}\\
-		\hline
-		\bf Related Parameter & \pref{paredf}\\
-		\hline
-		\bf History & Created April 1, 2020\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-~\newline\\
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
-		\hline
-		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{edfstorage} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & Storage patters for distribution function can vary.\\
-		\hline
-		\bf Related Commonality & \cref{equildistfunc}\\
-		\hline
-		\bf Related Parameter & \pref{parstorage}\\
-		\hline
-		\bf History & Created April 1, 2020\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-~\newline\\
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
-		\hline
-		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{edfcoeff} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & Coefficients for distribution function can vary.\\
-		\hline
-		\bf Related Commonality & \cref{equildistfunc}\\
-		\hline
-		\bf Related Parameter & \pref{paredfcoeff}\\
-		\hline
-		\bf History & Created April 1, 2020\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-~\newline\\
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
-		\hline
-		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{dimensions} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & The number of dimensions in the lattice of the model can vary.\\
-		\hline
-		\bf Related Commonality & \cref{lattice} \cref{probabilitydensityfunc}\\
-		\hline
-		\bf Related Parameter & \pref{pardimensions}\\
-		\hline
-		\bf History & Created April 1, 2020\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-~\newline\\
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
-		\hline
-		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{numveldir} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline 
-		\bf Description & The number of velocity directions in the lattice of the model can vary.\\
-		\hline
-		\bf Related Commonality & \cref{lattice} \cref{collisionoperator}\\
-		\hline
-		\bf Related Parameter & \pref{parveldir}\\
-		\hline
-		\bf History & Created April 1, 2020\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-~\newline\\
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
-		\hline
-		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{varcolop} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & Various collision operators can be used.\\
-		\hline
-		\bf Related Commonality & \cref{lattice} \cref{collisionoperator}\\
-		\hline
-		\bf Related Parameter & \pref{parcollisop}\\
-		\hline
-		\bf History & Created April 1, 2020\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-~\newline\\
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
-		\hline
-		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{vartraneq} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & Various Transport Equations can be used to describe the statistical behaviour of the system\\
-		\hline
-		\bf Related Commonality & \cref{Boltzmanntransequ}\\
-		\hline
-		\bf Related Parameter & \pref{partransequ}\\
-		\hline
-		\bf History & Created April 1, 2020\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-~\newline\\
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
-		\hline
-		\bf Item Number&  V\refstepcounter{varnum}\thevarnum\label{numberoffluids} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & The number of fluids allowed in the LBS.\\
-		\hline
-		\bf Related Commonality & \cref{fluidandboundaryinfo}\\
-		\hline
-		\bf Related Parameter & \pref{parfluids}\\
-		\hline
-		\bf History & Created April 1, 2020\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-~\newline\\
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
-		\hline
-		\bf Item Number&  V\refstepcounter{varnum}\thevarnum\label{typefluidparm} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & The type of fluid parameters.\\
-		\hline
-		\bf Related Commonality & \cref{fluidandboundaryinfo}\\
-		\hline
-		\bf Related Parameter & \pref{partypeparm}\\
-		\hline
-		\bf History & Created April 1, 2020\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-~\newline\\
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
-		\hline
-		\bf Item Number&  V\refstepcounter{varnum}\thevarnum\label{varboundary} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & The boundary of the lattice can have varied conditions.\\
-		\hline
-		\bf Related Commonality & \cref{lattice}\\
-		\hline
-		\bf Related Parameter & \pref{parboundary}\\
-		\hline
-		\bf History & Created April 1, 2020\\
+		\bf Item Number& V\refstepcounter{varnum}\thevarnum \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
+		\hline
+		\bf Description & \\
+		\hline
+		\bf Related Commonality & \cref{}\\
+		\hline
+		\bf Related Parameter & \pref{}\\
+		\hline
+		\bf History & \\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+~\newline\\
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
+		\hline
+		\bf Item Number& V\refstepcounter{varnum}\thevarnum \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
+		\hline
+		\bf Description & \\
+		\hline
+		\bf Related Commonality & \cref{}\\
+		\hline
+		\bf Related Parameter & \pref{}\\
+		\hline
+		\bf History & \\
 		\hline
@@
 		\hline
-		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{interface} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & The input interface can vary between LBS.\\
-		\hline
-		\bf Related Commonality & \cref{fluidandboundaryinfo}\\
-		\hline
-		\bf Related Parameter & \pref{parinterface}\\
-		\hline
-		\bf History & Created April 1, 2020\\
+		\bf Item Number& V\refstepcounter{varnum}\thevarnum \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
+		\hline
+		\bf Description & \\
+		\hline
+		\bf Related Commonality & \cref{}\\
+		\hline
+		\bf Related Parameter & \pref{}\\
+		\hline
+		\bf History & \\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+~\newline\\
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
+		\hline
+		\bf Item Number& V\refstepcounter{varnum}\thevarnum \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
+		\hline
+		\bf Description & \\
+		\hline
+		\bf Related Commonality & \cref{}\\
+		\hline
+		\bf Related Parameter & \pref{}\\
+		\hline
+		\bf History & \\
 		\hline
@@
 		\hline
-		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{varvisual} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & Visual presentation of the prediction.\\
-		\hline
-		\bf Related Commonality & \cref{predictionstomemory}\\
-		\hline
-		\bf Related Parameter & \pref{parvarvisual}\\
-		\hline
-		\bf History & Created April 1, 2020\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-~\newline\\
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
-		\hline
-		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{formatoutput} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & Format of prediction information.\\
-		\hline
-		\bf Related Commonality & \cref{predictionstomemory}\\
-		\hline
-		\bf Related Parameter & \pref{parformatoutput}\\
-		\hline
-		\bf History & Created April 1, 2020\\
+		\bf Item Number& V\refstepcounter{varnum}\thevarnum \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
+		\hline
+		\bf Description & \\
+		\hline
+		\bf Related Commonality & \cref{}\\
+		\hline
+		\bf Related Parameter & \pref{}\\
+		\hline
+		\bf History & \\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+~\newline\\
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
+		\hline
+		\bf Item Number& V\refstepcounter{varnum}\thevarnum \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
+		\hline
+		\bf Description & \\
+		\hline
+		\bf Related Commonality & \cref{}\\
+		\hline
+		\bf Related Parameter & \pref{}\\
+		\hline
+		\bf History & \\
 		\hline
@@
 		\hline
-		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{processing} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & Hardware which processes the calculations\\
-		\hline
-		\bf Related Commonality & None \\
-		\hline
-		\bf Related Parameter & \pref{parprocessing}\\
-		\hline
-		\bf History & Created April 1, 2020\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-~\newline\\
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
-		\hline
-		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{operatingsystem} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & Operating systems on which LBS runs.\\
-		\hline
-		\bf Related Commonality & None \\
-		\hline
-		\bf Related Parameter & \pref{paropsystem}\\
-		\hline
-		\bf History & Created April 1, 2020\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-~\newline\\
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
-		\hline
-		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{storage} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & Amount of storage and memory needed for the LBS.\\
-		\hline
-		\bf Related Commonality & None \\
-		\hline
-		\bf Related Parameter & \pref{parstorageconst}\\
-		\hline
-		\bf History & Created April 1, 2020\\
+		\bf Item Number& V\refstepcounter{varnum}\thevarnum \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
+		\hline
+		\bf Description & \\
+		\hline
+		\bf Related Commonality & \cref{}\\
+		\hline
+		\bf Related Parameter & \pref{}\\
+		\hline
+		\bf History & \\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+~\newline\\
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
+		\hline
+		\bf Item Number& V\refstepcounter{varnum}\thevarnum \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
+		\hline
+		\bf Description & \\
+		\hline
+		\bf Related Commonality & \cref{}\\
+		\hline
+		\bf Related Parameter & \pref{}\\
+		\hline
+		\bf History & \\
 		\hline
@@
 		\hline
-		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{varprecise} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & The precision needed for each input and output.\\
-		\hline
-		\bf Related Commonality & \cref{precision}\\
-		\hline
-		\bf Related Parameter & \pref{parprecise}\\
-		\hline
-		\bf History & Created April 1, 2020\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-~\newline\\
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
-		\hline
-		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{varaccuracy} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & The required accuracy for the output.\\
-		\hline
-		\bf Related Commonality & \cref{accuracy}\\
-		\hline
-		\bf Related Parameter & \pref{paraccuracy}\\
-		\hline
-		\bf History & Created April 1, 2020\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-~\newline\\
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
-		\hline
-		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{varresponsetime} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & The response time required for user interaction with the system varies.\\
-		\hline
-		\bf Related Commonality & \cref{performance}\\
-		\hline
-		\bf Related Parameter & \pref{parresponse}\\
-		\hline
-		\bf History & Created April 1, 2020\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-~\newline\\
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
-		\hline
-		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{varportability} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & The operating systems on which a LBS can run.\\
-		\hline
-		\bf Related Commonality & \cref{portability}\\
-		\hline
-		\bf Related Parameter & \pref{parportability}\\
-		\hline
-		\bf History & Created April 1, 2020\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-~\newline\\
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
-		\hline
-		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{varusability} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & The ease with which LBS can effectively be run varies.\\
-		\hline
-		\bf Related Commonality & \cref{usability}\\
-		\hline
-		\bf Related Parameter & \pref{parusability}\\
-		\hline
-		\bf History & Created April 1, 2020\\
+		\bf Item Number& V\refstepcounter{varnum}\thevarnum \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
+		\hline
+		\bf Description & \\
+		\hline
+		\bf Related Commonality & \cref{}\\
+		\hline
+		\bf Related Parameter & \pref{}\\
+		\hline
+		\bf History & \\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+~\newline\\
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
+		\hline
+		\bf Item Number& V\refstepcounter{varnum}\thevarnum \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
+		\hline
+		\bf Description & \\
+		\hline
+		\bf Related Commonality & \cref{}\\
+		\hline
+		\bf Related Parameter & \pref{}\\
+		\hline
+		\bf History & \\
 		\hline
@@
 		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parparallel} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{parallel}\\
-		\hline
-		\bf Range of Parameters & If parallelism is used, commonly OpenMP, OpenCL, CUDA, or MPI are used.\\
-		\hline
-		\bf Binding Time & Build Time\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-~\newline\\
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
-		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{paredf} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{parallel}\\
-		\hline
-		\bf Range of Parameters & Equilibrium can be approximated up to different orders in incompressible or compressible versions.\\
-		\hline
-		\bf Binding Time & Build Time\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-~\newline\\
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
-		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parstorage} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{edfstorage}\\
-		\hline
-		\bf Range of Parameters & Single and multiple array storage patterns can be used.\\
-		\hline
-		\bf Binding Time & Build Time\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-~\newline\\
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
-		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{paredfcoeff} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{edfcoeff}\\
-		\hline
-		\bf Range of Parameters & Numerous coefficients for equilibrium distribution function based on number of velocity directions. \\
-		\hline
-		\bf Binding Time & Build Time\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-~\newline\\
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
-		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{pardimensions} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{dimensions}\\
-		\hline
-		\bf Range of Parameters & LBS models can have up to 3 dimensions.\\
-		\hline
-		\bf Binding Time & Build Time or Run Time\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-~\newline\\
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
-		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parveldir} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{numveldir}\\
-		\hline
-		\bf Range of Parameters & One dimensional models include options of 2, 3, and 5 velocity directions.
-Two dimensional models include options of 9, 13, and 15 velocity directions.
-Three dimensional models include options of 15, 19, and 27 velocity directions.
-\\
-		\hline
-		\bf Binding Time & Build Time or Run Time\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-~\newline\\
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
-		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parcollisop} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{varcolop}\\
-		\hline
-		\bf Range of Parameters & SRT, TRT, MRT, BGK collision operators.\\
-		\hline
-		\bf Binding Time & Build Time\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-~\newline\\
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
-		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{partransequ} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{vartraneq}\\
-		\hline
-		\bf Range of Parameters & Collision and collision free transport equations.\\
-		\hline
-		\bf Binding Time & Build Time\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-~\newline\\
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
-		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parfluids} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{numberoffluids}\\
-		\hline
-		\bf Range of Parameters & LBS can model one or more fluids.\\
-		\hline
-		\bf Binding Time & Build Time\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-~\newline\\
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
-		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{partypeparm} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{typefluidparm}\\
-		\hline
-		\bf Range of Parameters & LBS fluid parameters include Reynolds Number, density, viscosity, time, pressure, force, direction, relaxation rate, turbulence.\\
-		\hline
-		\bf Binding Time & Build Time\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-~\newline\\
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
-		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parboundary} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{varboundary}\\
-		\hline
-		\bf Range of Parameters & Lattice boundary can have reflective or non-reflective conditions.\\
-		\hline
-		\bf Binding Time & Build Time\\
+		\bf Item Number& P\refstepcounter{parnum}\theparnum \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
+		\hline
+		\bf Corresponding Variability & \vref{}\\
+		\hline
+		\bf Range of Parameters & \\
+		\hline
+		\bf Binding Time & \\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+~\newline\\
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
+		\hline
+		\bf Item Number& P\refstepcounter{parnum}\theparnum \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
+		\hline
+		\bf Corresponding Variability & \vref{}\\
+		\hline
+		\bf Range of Parameters & \\
+		\hline
+		\bf Binding Time & \\
 		\hline
@@
 		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parinterface} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{interface}\\
-		\hline
-		\bf Range of Parameters & Input can be graphical, text or file.\\
-		\hline
-		\bf Binding Time & Build Time\\
+		\bf Item Number& P\refstepcounter{parnum}\theparnum \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
+		\hline
+		\bf Corresponding Variability & \vref{}\\
+		\hline
+		\bf Range of Parameters & \\
+		\hline
+		\bf Binding Time & \\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+~\newline\\
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
+		\hline
+		\bf Item Number& P\refstepcounter{parnum}\theparnum \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
+		\hline
+		\bf Corresponding Variability & \vref{}\\
+		\hline
+		\bf Range of Parameters & \\
+		\hline
+		\bf Binding Time & \\
 		\hline
@@
 		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parvarvisual} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{varvisual}\\
-		\hline
-		\bf Range of Parameters & LBS can provide 1D, 2D, and 3D rendering of  the model.\\
-		\hline
-		\bf Binding Time & Build Time or Run Time\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-~\newline\\
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
-		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parformatoutput} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{formatoutput}\\
-		\hline
-		\bf Range of Parameters & LBS prediction information is output in either text or binary format.\\
-		\hline
-		\bf Binding Time & Build Time\\
+		\bf Item Number& P\refstepcounter{parnum}\theparnum \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
+		\hline
+		\bf Corresponding Variability & \vref{}\\
+		\hline
+		\bf Range of Parameters & \\
+		\hline
+		\bf Binding Time & \\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+~\newline\\
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
+		\hline
+		\bf Item Number& P\refstepcounter{parnum}\theparnum \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
+		\hline
+		\bf Corresponding Variability & \vref{}\\
+		\hline
+		\bf Range of Parameters & \\
+		\hline
+		\bf Binding Time & \\
 		\hline
@@
 		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parprocessing} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{processing}\\
-		\hline
-		\bf Range of Parameters & The LBS model can be calculated on the CPU or GPU.\\
-		\hline
-		\bf Binding Time & Build Time\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-~\newline\\
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
-		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{paropsystem} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{operatingsystem}\\
-		\hline
-		\bf Range of Parameters & LBS can be run on Windows, MacOS, or Linux versions.\\
-		\hline
-		\bf Binding Time & Build Time\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-~\newline\\
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
-		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parstorageconst} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{storage}\\
-		\hline
-		\bf Range of Parameters & The amount of memory and storage varies between LBS. LBS sizes begin in the low gigabytes. \\
-		\hline
-		\bf Binding Time & Build Time\\
+		\bf Item Number& P\refstepcounter{parnum}\theparnum \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
+		\hline
+		\bf Corresponding Variability & \vref{}\\
+		\hline
+		\bf Range of Parameters & \\
+		\hline
+		\bf Binding Time & \\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+~\newline\\
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
+		\hline
+		\bf Item Number& P\refstepcounter{parnum}\theparnum \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
+		\hline
+		\bf Corresponding Variability & \vref{}\\
+		\hline
+		\bf Range of Parameters & \\
+		\hline
+		\bf Binding Time & \\
 		\hline
@@
 		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parprecise} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{varprecise}\\
-		\hline
-		\bf Range of Parameters & Each input and each output will have a specified the precision with which the real number is stored and/or calculated.\\
-		\hline
-		\bf Binding Time & Build Time or Run Time\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-~\newline\\
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
-		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{paraccuracy} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{varaccuracy}\\
-		\hline
-		\bf Range of Parameters & The tolerance allowed for each output produced by the system so that the system produces usable results. \\
-		\hline
-		\bf Binding Time & Build Time or Run Time\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-~\newline\\
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
-		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parresponse} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{varresponsetime}\\
-		\hline
-		\bf Range of Parameters & The maximum amount of time that the user is expected to wait for the system to produce a result ranges between systems and models from near instantaneous to several minutes or longer.\\
-		\hline
-		\bf Binding Time & Specification\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-~\newline\\
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
-		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parportability} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{varportability}\\
-		\hline
-		\bf Range of Parameters & The ease with which LBS can run on Windows, MacOS, and Linux system varies. Some LBS can run on all three, while some can run on multiple or only one operating system.\\
-		\hline
-		\bf Binding Time & Specification\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-~\newline\\
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
-		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parusability} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{varusability}\\
-		\hline
-		\bf Range of Parameters & Some LBS require programming skills and domain expert knowledge while some LBS can be run without such skills.\\
-		\hline
-		\bf Binding Time & Specification\\
+		\bf Item Number& P\refstepcounter{parnum}\theparnum \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
+		\hline
+		\bf Corresponding Variability & \vref{}\\
+		\hline
+		\bf Range of Parameters & \\
+		\hline
+		\bf Binding Time & \\
+		\hline
+	\end{tabular}
+\end{minipage}\\
+~\newline\\
+~\newline
+\noindent
+\begin{minipage}{\textwidth}
+	\renewcommand*{\arraystretch}{1.5}
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
+		\hline
+		\bf Item Number& P\refstepcounter{parnum}\theparnum \\
+		\hline
+	\end{tabular}\\
+
+	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
+		\hline
+		\bf Corresponding Variability & \vref{}\\
+		\hline
+		\bf Range of Parameters & \\
+		\hline
+		\bf Binding Time & \\
 		\hline

### 2021-10-06 StateOfPractice/DomainMeasurements/LBM-Peter/Peter-Notes-SolutionDescriptions.tex
Rearrange domain measurement reports to have more consistent naming and locations

--- 2020-03-05 StateOfPractice/Peter-Notes/Peter-Notes-SolutionDescriptions.tex
@@
 \item ch4-project\\
-	\href{https://github.com/SoftwareImpacts/SIMPAC-2019-1}{https://github.com/SoftwareImpacts/SIMPAC-2019-1}\\  
+\href{https://github.com/ecalzavarini/ch4-project}{https://github.com/ecalzavarini/ch4-project}\\  
 Found using Google search.\\
@@
 Found using Google search.\\
-	finite volume lattice Boltzmann method
+A finite volume lattice Boltzmann method solver implementated on OpenFOAM.
 
@@
 LB3D provides functionality to simulate three-dimensional simple, binary oil/water and ternary oil/water/amphiphile fluids using the Shan-Chen model for binary fluid interactions.
-
-	\item LBDEMcoupling-public
-	\\
-	\href{https://github.com/ParticulateFlow/LBDEMcoupling-public}{https://github.com/ParticulateFlow/LBDEMcoupling-public}
-	\\
-	Found using Google search.\\
-	Coupling between the Lattice-Boltzmann code Palabos and the DEM code LIGGGHTS
 

### 2021-10-06 StateOfPractice/DomainMeasurements/LBM-Peter/Report/Report.tex
Rearrange domain measurement reports to have more consistent naming and locations

--- 2021-07-28 StateOfPractice/Peter-Notes/Report/Report.tex
@@
 	{Master of Engineering in Computing and Software}
-	{\copyright \ Copyright by \authorname, July 2021}
+	{\copyright \ Copyright by \authorname, August 2021}
 
@@
 
-\noindent\textbf{NUMBER OF PAGES:} 123
+\noindent\textbf{NUMBER OF PAGES:} 125
 
@@
 		\includegraphics[width=0.7\textwidth]{circularflow}
-		\caption{Streamlines of flow past a stationary circular cylinder at Re = 10, 20, and 40 \citep{chen2021phase}}
+		\caption{Streamlines of flow past a stationary circular cylinder at Reynolds number = 10, 20, and 40 \citep{chen2021phase}}
 		\label{circularflow}
@@
 This assessment of the state of the practice of LBM software development builds off of prior work on assessing the state of research software development.
-The previous state of the practice assessments include domains such as Geographic Information Systems \citep{SmithEtAl2018_arXivGIS}, Mesh Generators \citep{SmithEtAl2016}, Oceanographic Software \citep{smith2015state}, Seismology software \citep{SmithEtAl2018}, and statistical software for psychology \citep{SmithEtAl2018_StatSoft}.
+Updates to the methodology used in prior assessments are detailed below.
+Those assessments include domains such as Geographic Information Systems \citep{SmithEtAl2018_arXivGIS}, Mesh Generators \citep{SmithEtAl2016}, Oceanographic Software \citep{smith2015state}, Seismology software \citep{SmithEtAl2018}, and statistical software for psychology \citep{SmithEtAl2018_StatSoft}.
 
@@
 A domain analysis consists of systematically identifying and documenting the commonalities, variabilities, and parameters of variation of a software family \citep{Weiss1997}.
-A software family is defined by \cite{parnas1976design} as “a set of programs whose common properties are so extensive that it is advantageous to study the common properties of the programs before analyzing individual members”.
+A software family is defined by \cite{parnas1976design} as ``a set of programs whose common properties are so extensive that it is advantageous to study the common properties of the programs before analyzing individual members''.
 
@@
 Software packages can have considerable variation, even when the packages appear to be in the same software family.
-Viewing the ranking exercise as a means for selecting the right tool for a job, functionality is important.
+Viewing our final ranking of the software packages as a means for selecting the right tool for a job, functionality is important.
 A 2D Lattice Boltzmann solver is not a substitute for a 3D Lattice Boltzmann solver when the latter is required.
-Therefore, this exercise seeks to classify the different software products based on their functionality.
+For example, an analysis of a vascular system may require a 3D model to convey all required information to a researcher.
+This domain analysis helps ensure that the packages that we compare are appropriately similar, and highlights potential differences.
+This information further helps when analyzing the appropriateness of the final rank of the packages.
+This exercise seeks to classify the different software products based on their functionality.
 Nonfunctional qualities are assessed through other means in our process.
 For instance, through usability experiments as noted in the \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Methodology.pdf}{Methodology for Assessing the State of the Practice for Domain X} document.
+This exercise was not part of the methodology of the previous State of the Practice exercise.
 
@@
 This section lists common features among potential family members.
+These features were chosen after a review of LBM literature and software package specifications.
 The commonalities are organized using the following abstraction of the system, which can be used to describe all Lattice Boltzmann systems: input information, generate the simulation, output the results.
@@
 The next section, Section \ref{comm_out}, shows the common features for the output of Lattice Boltzmann systems, such as the requirement that mesh information be written to files.
-Although the output information could simply be written to the computer’s memory, in all practical applications it is desirable to have a persistent record of the output that was created.
+Although the output information could simply be written to the memory, in all practical applications it is desirable to have a persistent record of the output that was created.
 
@@
 \end{minipage}\\
+
 \subsection{Variabilities}\label{var_sec}
 This section provides a list of characteristics that may vary among family members.
+These features were chosen after a review of LBM literature and software package specifications.
 As in Section \ref{comm_sec}, the first three subsections on variabilities are organized into the following sublists: Simulation Models, Input and Output.

### 2021-10-06 StateOfPractice/DomainMeasurements/MedImaging-Ao/Empirical Measures Template.tex
Rearrange domain measurement reports to have more consistent naming and locations

--- 2020-04-15 StateOfPractice/Peter-Notes/TemplateToGradeSCSoft_Comments.tex
@@
 
-\title{Template for Grading Scientific Computing Software} 
-\author{Spencer Smith, Adam Lazzarato, Yue Sun, Zheng Zeng,\\ Vasudha Kapil and Jacques Carette}
+\title{Empirical Measures Template} 
+\author{Ao Dong}
 \date{\today}
@@
 \maketitle
-\include{GradingTemplate}
+
+The tools for Empirical Measures:
+\begin{itemize}
+\item \href{https://github.com/tomgi/git_stats}{git\_stats}
+\item \href{https://github.com/boyter/scc}{scc}
+\end{itemize}
+
+\ad{For now it only contains raw metrics. We'll have to consider how to use them to calculate more meaningful metrics.}
+
+\begin{longtable}{p{16cm}}
+  \caption{Grading Template}   \label{table:TemplateFull}\\
+  \toprule
+  \textbf{Raw Metrics} (Measured via git\_stats)\\
+  \midrule
+  Number of text-based files. (number)\\
+  Number of binary files. (number)\\
+  Number of total lines in text-based files. (number)\\
+  Number of total lines added to text-based files. (number)\\
+  Number of total lines deleted from text-based files. (number)\\
+  Number of total commits. (number)\\
+Numbers of commits by year in the last 5 years. (Count from as early as possible
+if the project is younger than 5 years.) (list of numbers)\\
+  Numbers of commits by month in the last 12 months. (list of numbers)\\
+
+  \midrule
+  \textbf{Raw Metrics} (Measured via scc)\\
+  \midrule
+  Number of text-based files. (number)\\
+  Number of total lines in text-based files. (number)\\
+  Number of code lines in text-based files. (number)\\
+  Number of comment lines in text-based files. (number)\\
+  Number of blank lines in text-based files. (number)\\
+
+  \midrule
+  \textbf{Raw Metrics} (Measured via GitHub API)\\
+  \midrule
+Numbers of commits, added lines, and deleted lines by author by week.
+(2-dimensional list of numbers) \ad{I know we're not likely to measure this.
+Just a note that we can get this data. It's only available for GitHub repos, and
+I have to test tools for parsing JSON data (e.g. to Excel table).}\\
+  \bottomrule
+
+\end{longtable}
 

### 2021-10-06 StateOfPractice/DomainMeasurements/MedImaging-Ao/Report/01abstract.tex
Rearrange domain measurement reports to have more consistent naming and locations

--- 2021-08-16 StateOfPractice/Ao-Report/01abstract.tex
@@
 For future SC software development, we proposed recommendations on improving software qualities, dealing with limited resources, choosing a tech stack, and enriching the testing datasets.
-The recommendations include adopting test-driven development, using continuous integration and continuous delivery (CI/CD), using git and GitHub, maintaining good documents, supporting third-party plugins or extensions, considering web application solutions, and establishing community collaboration in a SC domain.
+The recommendations include adopting test-driven development, using continuous integration and continuous delivery (CI/CD), using git and GitHub, maintaining good documentation, supporting third-party plugins or extensions, considering web application solutions, and establishing community collaboration in an SC domain.
 \newline

### 2021-10-06 StateOfPractice/DomainMeasurements/MedImaging-Ao/Report/12Background.tex
Rearrange domain measurement reports to have more consistent naming and locations

--- 2021-08-30 StateOfPractice/Ao-Report/12Background.tex
@@
 
-In this project, the 9 software qualities mentioned above are the criteria ($m = 9$), while 29 software packages ($n = 29$) are compared for each of the $m$ criteria.
+In this project, the first nine software qualities mentioned in Section \ref{sec_software_quality} are the criteria ($m = 9$), while 29 software packages ($n = 29$) are compared for each of the $m$ criteria.
 The software are evaluated with the grading template in Appendix \ref{ap_grading_template} and a subjective score from one to ten is given for each quality for each package.

### 2021-10-06 StateOfPractice/DomainMeasurements/MedImaging-Ao/Report/14MeasurementResults.tex
Rearrange domain measurement reports to have more consistent naming and locations

--- 2021-08-28 StateOfPractice/Ao-Report/14MeasurementResults.tex
@@
 
-As discussed in Section \ref{sec_applying_method}, we applied our method to the MI domain.
+As discussed in Section \ref{sec_grading_software}, we use a grading template and a empirical method to measure the selected software.
+We applied this step to the MI domain (Section \ref{sec_applying_method_grading}).
 This section shows the summary of the measurement results.

### 2021-10-06 StateOfPractice/DomainMeasurements/MedImaging-Ao/Report/15InterviewsWithDevelopers.tex
Rearrange domain measurement reports to have more consistent naming and locations

--- 2021-08-28 StateOfPractice/Ao-Report/15InterviewsWithDevelopers.tex
@@
 
-Our method involves interviews with developers in the MI domain (Section \ref{sec_applying_method}).
-We summarize some answers from the interviews in this section.
+The measurement results in Section \ref{ch_results} are based on the information collected by ourselves.
+Such information is sufficient to measure the projects with reasonable efforts, but incomplete for us to understand the development process more deeply.
+For example, we usually cannot identify the following information in a project's documents: the pain points during the development, the threats to certain software qualities, and the developers' strategies to address them.
+We believe interviews with developers can collect the additional information we need.
+As a result, our method involves interviews with developers in a domain (Section \ref{sec_interview_methods}).
+We applied this step to the MI domain (Section \ref{sec_apply_to_mi_interviews}).
+
+In this section, we summarize some answers from the interviews.
 We highlight the answers that are the most informative and interesting in this section, and summarize the rest in Appendix \ref{ap_interview}.
 
-We interviewed nine developers from eight of the 29 MI software projects.
+As mentioned in Section \ref{sec_apply_to_mi_interviews}, we contacted all 29 teams.
+Some of them responded and participated in the interviews.
+Eventually, we interviewed nine developers from eight of the 29 MI software projects.
 The eight projects are \textit{3D Slicer}, \textit{INVESALIUS 3}, \textit{dwv}, \textit{BioImage Suite Web}, \textit{ITK-SNAP}, \textit{MRIcroGL}, \textit{Weasis}, and \textit{OHIF}.

### 2021-10-06 StateOfPractice/DomainMeasurements/MedImaging-Ao/Report/16AnswersToResearchQuestions.tex
Rearrange domain measurement reports to have more consistent naming and locations

--- 2021-09-07 StateOfPractice/Ao-Report/16AnswersToResearchQuestions.tex
@@
 \label{sec_threats_to_validity}
-This section lists all the potential threats to the validity of our research.
 
-\begin{enumerate}
+This section lists all the potential threats to validity in our research. The definitions of common validity are \cite{AmpatzoglouEtAl2019}\cite{ZhouEtAl2016}:
+\begin{description}
+\item[Construct Validity:] ``Deﬁnes how eﬀectively a test or experiment measures up to its claims.
+This aspect deals with whether or not the researcher measures what is intended to be measured" \cite{AmpatzoglouEtAl2019}.
+\item[Internal Validity:] ``This aspect relates to the examination of causal relations.
+Internal validity examines whether an experimental treatment/condition makes a diﬀerence or not, and whether there is evidence to support the claim" \cite{AmpatzoglouEtAl2019}.
+\item[External Validity:] ``Define the domain to which a study's findings can be
+generalized" \cite{ZhouEtAl2016}.
+\item[Conclusion Validity:] ``Demonstrate that the operations of a study such as the data collection procedure can be repeated, with the same results" \cite{ZhouEtAl2016}.
+\end{description}
+
+\noindent We categorize and present the threats to validity in the following subsections.
+
+\subsection{Threats to Construct Validity}
+\begin{itemize}
 \item We compared nine software qualities for 29 software packages, so we could only spend a limited time on each of them. As a result, our assessments may have missed something relevant.
-\item We used the grading template in Appendix \ref{ap_grading_template} to guide our measurements.
-Our impressions of the software - such as user experience - were factors in deciding some scores.
-Thus, there is a risk that some scores may be subjective and biased.
+\item Our ranking is partly based on surface (shallow) measurement, which may not fully reveal the underlying qualities.
+\end{itemize}
+
+\subsection{Threats to Internal Validity}
+\begin{itemize}
 \item It was not practical to ask each development team for every piece of information.
@@
 There may be cases that we missed some information.
-\item Our ranking is partly based on surface (shallow) measurement, which may not fully reveal the underlying qualities.
-\item We interviewed eight teams, which is a good proportion of the 29. However, there is still a risk that they might not well represent the whole MI software community.
 \item As mentioned in Section \ref{ch_interview}, one interviewee was too busy to participate in a full interview, so he provided a version of written answers to us.
@@
 We might underestimate their rank due to uncommon technical issues.
+\end{itemize}
+
+\subsection{Threats to External Validity}
+\begin{itemize}
+\item We interviewed eight teams, which is a good proportion of the 29. However, there is still a risk that they might not well represent the whole MI software community.
 \item Our ranking gave all qualities equal weight, which may not be the case with all users. Thus, it may not represent the popularity of software among users.
 \item The number of GitHub stars, watches, and forks are not perfect measures of popularity, but they are what we had available.
-\end{enumerate}
+\end{itemize}
+
+\subsection{Threats to Conclusion Validity}
+\begin{itemize}
+\item We used the grading template in Appendix \ref{ap_grading_template} to guide our measurements.
+Our impressions of the software - such as user experience - were factors in deciding some scores.
+Thus, there is a risk that some scores may be subjective and biased.
+\end{itemize}

### 2021-10-06 StateOfPractice/DomainMeasurements/MedImaging-Ao/Report/18Conclusion.tex
Rearrange domain measurement reports to have more consistent naming and locations

--- 2021-08-29 StateOfPractice/Ao-Report/18Conclusion.tex
@@
 	\item we can further automate the measurements on the grading temple in Appendix \ref{ap_grading_template}.
-For example, with automation scripts and a GitHub API, we may save significant time on retrieving the GitHub metrics;
+For example, with automation scripts and the GitHub API, we may save significant time on retrieving the GitHub metrics.
+We have started to build a tool for this purpose, with its source code at this repository \hyperlink{https://github.com/smiths/AIMSS/tree/master/StateOfPractice/GitHubMetricsCollector}{https://github.com/smiths/AIMSS/.../GitHubMetricsCollector}.
+This GitHub Metrics Collector can take GitHub repository links as input, automatically collect metrics from the GitHub API, and record the results.
+We can improve and use this tool in our future projects;
 	\item the grading standard can be more explicit. For example, we can explicitly define scores for each item in the grading temple.

### 2021-10-06 StateOfPractice/DomainMeasurements/MedImaging-Ao/Report/32FullSoftwareListBeforeFiltering.tex
Rearrange domain measurement reports to have more consistent naming and locations

\chapter{Full Software List Before Filtering}
\label{ap_list_before_filtering}

Table \ref{tab_list_before_filtering} lists the 48 software before filtering. We selected 29 of them to the final list, which are all open-source and in Visualization (V) sub-group. We found software packages in sub-groups Tool Kit (TK) and Picture Archiving and Communication System (PACS) but removed them from the final list. The table also shows the sources of identifying them.

\begin{table}[H]
\hspace*{-1cm}\begin{tabular}{lllllllllll}
\hline
 &  &  &  & \multicolumn{7}{c}{Source} \\ \cline{5-11} 
\multirow{-2}{*}{Software} & \multirow{-2}{*}{\begin{tabular}[c]{@{}l@{}}Final\\ list\end{tabular}} & \multirow{-2}{*}{\begin{tabular}[c]{@{}l@{}}Open-\\ source\end{tabular}} & \multirow{-2}{*}{\begin{tabular}[c]{@{}l@{}}Sub-\\ group\end{tabular}} & \cite{Bjorn2017} & \cite{Bruhschwein2019} & \cite{Haak2015} & \cite{Emms2019} & \cite{Hasan2020} & \cite{Mu2019} & \cite{Samala2014} \\ \hline
3D Slicer \cite{Kikinis2014} & X & X & V &  &  & X & X & X &  & X \\
Ginkgo CADx \cite{Wollny2020} & X & X & V & X & X & X & X & X & X &  \\
XMedCon \cite{Nolf2003} & X & X & V &  &  & X & X & X & X &  \\
Weasis \cite{Roduit2021} & X & X & V &  &  & X & X & X & X &  \\
MRIcroGL \cite{Rorden2021} & X & X & V &  &  & X & X & X &  & X \\
SMILI \cite{Chandra2018} & X & X & V &  &  &  & X & X & X &  \\
ImageJ \cite{Rueden2017} & X & X & V &  &  & X & X & X &  & X \\
Fiji \cite{Schindelin2012} & X & X & V &  &  &  &  &  &  & X \\
DicomBrowser \cite{Archie2012} & X & X & V &  &  &  & X & X &  &  \\
3DimViewer \cite{TESCAN2020} & X & X & V & X &  &  & X & X &  &  \\
Horos \cite{horosproject2020} & X & X & V & X & X &  &  &  &  &  \\
OsiriX Lite \cite{PixmeoSARL2019} & X & X & V & X & X & X &  &  &  &  \\
dwv \cite{Martelli2021} & X & X & V & X &  & X &  &  &  &  \\
Drishti \cite{Limaye2012} & X & X & V & X &  &  &  &  &  &  \\
BioImage Suite Web \cite{Papademetris2005} & X & X & V &  &  & X &  &  &  &  \\
\end{tabular}
\end{table}

\begin{table}[H]
\hspace*{-1cm}\begin{tabular}{lllllllllll}
\hline
 &  &  &  & \multicolumn{7}{c}{Source} \\ \cline{5-11} 
\multirow{-2}{*}{Software} & \multirow{-2}{*}{\begin{tabular}[c]{@{}l@{}}Final\\ list\end{tabular}} & \multirow{-2}{*}{\begin{tabular}[c]{@{}l@{}}Open-\\ source\end{tabular}} & \multirow{-2}{*}{\begin{tabular}[c]{@{}l@{}}Sub-\\ group\end{tabular}} & \cite{Bjorn2017} & \cite{Bruhschwein2019} & \cite{Haak2015} & \cite{Emms2019} & \cite{Hasan2020} & \cite{Mu2019} & \cite{Samala2014} \\ \hline
OHIF Viewer \cite{Ziegler2020} & X & X & V & \multicolumn{7}{l}{Other source: \cite{Ziegler2020}} \\
Slice:Drop \cite{Haehn2013} & X & X & V &  &  & X &  &  &  &  \\
GATE \cite{Jan2004} & X & X & V &  &  &  & X &  &  &  \\
ITK-SNAP \cite{Yushkevich2006} & X & X & V &  &  &  &  &  &  & X \\
ParaView \cite{Ahrens2005} & X & X & V &  &  &  &  &  &  & X \\
MatrixUser \cite{Liu2016} & X & X & V &  &  &  &  & X &  &  \\
DICOM Viewer \cite{Afsar2021} & X & X & V &  &  &  &  & X &  &  \\
INVESALIUS 3 \cite{Amorim2015} & X & X & V &  &  &  &  & X &  &  \\
medInria \cite{Fillard2012} & X & X & V &  &  & X &  &  &  &  \\
dicompyler \cite{Panchal2010} & X & X & V &  &  &  &  &  & X &  \\
MicroView \cite{ParallaxInnovations2020} & X & X & V &  &  & X &  &  &  &  \\
Papaya \cite{UTHSCSA2019} & X & X & V &  &  &  & X & X & X &  \\
AMIDE \cite{Loening2017} & X & X & V &  &  &  & X & X & X &  \\
Gwyddion \cite{Nevcas2012} & X & X & V &  &  &  &  & X &  &  \\
VTK \cite{SchroederEtAl2006} &  & X & TK &  &  &  &  &  &  & X \\
ITK \cite{McCormick2014} &  & X & TK &  &  &  &  &  &  & X \\
DCMTK \cite{DCMTK2021} &  & X & TK &  &  &  &  & X & X &  \\
XTK \cite{Haehn2012} &  & X & TK & \multicolumn{7}{l}{Other source: it is used by slice:drop} \\
dcm4che \cite{dcm4che2021} &  & X & TK & X &  &  & X & X &  &  \\
cornerstone \cite{cornerstone2020} &  & X & TK &  &  & X &  &  &  &  \\
dcm2niix \cite{Li2016} &  & X & TK &  &  &  &  & X &  &  \\
orthanc \cite{Jodogne2018} &  & X & PACS & X &  &  &  & X &  &  \\
Conquest \cite{Herk2021} &  & X & PACS & X &  &  &  &  &  &  \\
ClearCanvas \cite{ClearCanvas2015} &  & X & PACS, V & X &  &  &  &  &  &  \\
\begin{tabular}[c]{@{}l@{}}Open Dicom\\ Viewer \cite{OpenDicomViewer2011} \end{tabular} &  & X & V &  &  & X &  &  & X &  \\
MicroDicom \cite{MicroDicom2021} &  &  & V & X &  &  &  &  & X &  \\
Aliza \cite{Aliza2021} &  &  & V &  &  &  & X &  & X &  \\
JiveX \cite{Visus2021} &  &  & V &  &  & X &  &  & X &  \\
MIPAV \cite{CIT2020} &  &  & V &  &  & X &  &  &  &  \\
Oviyam \cite{RasterImages2021} &  &  & V &  &  & X &  &  &  &  \\
MeVisLab \cite{Heckel2009} &  &  & V &  &  &  &  &  &  & X \\
\begin{tabular}[c]{@{}l@{}}Sante DICOM\\ Viewer Lite \cite{Santesoft2021} \end{tabular} &  &  & V &  &  &  &  &  & X &  \\
\begin{tabular}[c]{@{}l@{}}Navegatium\\ DICOM Viewer\end{tabular} &  &  & V &  &  &  &  &  & X &  \\ \hline
\end{tabular}
\caption{\label{tab_list_before_filtering}Full software list before filtering}
\end{table}

### 2021-10-08 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
added definition of unambiguity, chose definition, added reasoning

@@
 
+\begin{defn}
+	\label{UnambiguitySelected} (1) Not having two or more possible meanings. (2) Not susceptible to different interpretations. (3) Not obscure, not vague. (4) Clear, definite, certain. \citep{FDA2014}.
+\end{defn}
+
 \begin{mybox}
 \subsubsection*{Unambiguity} 
-\wss{Needs to be completed.}
-\end{mybox}
-
-\subsubsection*{Reasoning}
-
-\wss{Needs to be completed}  The different reasons for the proposed definition
-are as follows:
-
-\begin{itemize}
-  \item reason 1
-  \item reason 2
-  \item ...
+Definition~\ref{UnambiguitySelected} rephrased: Unambiguity is the quality of having one clear interpretation, and not having any other possible interpretations.
+\end{mybox}
+
+\subsubsection*{Reasoning}
+
+\begin{itemize}
+  \item This definition succinctly encompasses parts of the chosen definition, and includes the basis of the alternate definition.
 \end{itemize}

### 2021-10-09 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
added traceability definitions

@@
 
+\begin{defn}
+This is the property of a system description technique which allows changes in one of the three system descriptions -- requirements, specification, implementation -- to be traced to the corresponding portions of the other descriptions.
+The correspondence should be maintained throughout the lifetime of the system \citep{greenspan1978structuring}.
+\end{defn}
+
+\begin{defn}
+The ability to relate requirements specifications with other artefacts created
+in the development life-cycle of a software system \citep{spanoudakis2002plausible}.
+\end{defn}
+
+\begin{defn}
+Symbol identification that permits symbols to be traced throughout a software system. Also, continuity between program versions \citep{graf1999modern}.
+\end{defn}
+
 \begin{mybox}

### 2021-10-09 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
added modifiability definitions

@@
 
+\begin{defn}
+\begin{enumerate}
+	\item Ease with which a system can be changed without introducing defects.
+	\item Degree to which a product or system can be effectively and efficiently modified without introducing defects or degrading existing product quality \citep{iso2017iec}.
+\end{enumerate}
+\end{defn}
+
+\begin{defn}
+Modifiability is about the cost of change and refers to the ease with which a software system can accommodate changes \citep{northrop2004achieving}.
+\end{defn}
+
 \begin{mybox}

### 2021-10-09 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
added reproducibility definitions

@@
 
+\begin{defn}
+Closeness of the agreement between the results of measurements of the same measurand carried out under changed conditions of measurement \citep{iso2017iec}.
+\end{defn}
+
+\begin{defn}
+Obtaining consistent results using the same input data, computational steps,  methods, and code, and conditions of analysis \citep{national2019reproducibility}.
+\end{defn}
+
+\begin{defn}
+Running the same software on the same input data and obtaining the same results \citep{rougier2017sustainable}.
+\end{defn}
+
 It would be worthwhile to look for some additional definitions.

### 2021-10-09 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
added sustainability definitions

@@
 
+\begin{defn}
+Sustainable Software is software whose direct and indirect negative impacts on economy, society, human beings, and the environment resulting from development, deployment, and usage of the software is minimal and/or has a positive effect on sustainable development \citep{dick2010model}.
+\end{defn}
+
+\begin{defn}
+The art of developing sustainable software with a sustainable software engineering process so that negative and positive impacts result in and/or are expected to result from the software product over its whole life cycle are continuously assessed, documented, and used for further optimization of the software product \citep{venters2014software}.
+\end{defn}
+
 Find paper that combines nonfunctional qualities into sustainability.

### 2021-10-10 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
added abstraction definitions

@@
 
+\begin{defn}
+Abstraction refers to the process of removing irrelevant details in order to focus on the essential parts of the system \citep{filipovikj2017pattern}.
+\end{defn}
+
+\begin{defn}
+Abstraction is a special case of separation of concerns wherein we separate the concerns of the important aspects from the concern of the less important details \citep{GhezziEtAl2003}.
+\end{defn}
+
+\begin{defn}
+An abstraction is a simplification of some complex element of a system used to communicate meaning in a single phrase \citep{roger2015software}.
+\end{defn}
+
 Abstraction is related to reusability (and other qualities).

### 2021-10-11 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
added reasoning for unambiguity definition

@@
 \begin{defn}
- A Software Requirements Specification (SRS) is unambiguous if, and only if,
+ \label{IEEEunambiguity}A Software Requirements Specification (SRS) is unambiguous if, and only if,
  every requirement stated therein has only one interpretation \citep{IEEE1998}.
@@
 \subsubsection*{Unambiguity} 
-Definition~\ref{UnambiguitySelected} rephrased: Unambiguity is the quality of having one clear interpretation, and not having any other possible interpretations.
-\end{mybox}
-
-\subsubsection*{Reasoning}
-
-\begin{itemize}
-  \item This definition succinctly encompasses parts of the chosen definition, and includes the basis of the alternate definition.
+Unambiguity is the quality of having one clear interpretation, and not having any other possible interpretations.
+\end{mybox}
+
+\subsubsection*{Reasoning}
+
+The recommended definition comes from rephrasing Definition \ref{UnambiguitySelected} while keeping in mind its similarities with Definition \ref{IEEEunambiguity}.
+The rationale for the proposed definition follows:
+
+\begin{itemize}
+  \item Definition \ref{IEEEunambiguity} mentions the ``Software Requirements Specification'' document and, the specifically related term, ``requirement''.
+We feel that for the sake of abstraction we should not have the definition focus solely on this specific document and its related term, so we do not include these in the chosen definition.
+  \item The term ``interpretation'' is found in both Definition \ref{IEEEunambiguity} and Definition \ref{UnambiguitySelected} and is kept in the chosen definition.
+  \item The first two part of Definition \ref{UnambiguitySelected} can be succinctly restated as ``having one interpretation''.
+  \item The fourth part of Definition \ref{UnambiguitySelected} can be added to the above phrase, resulting in the new phrase ``having one clear interpretation''.
+This removes the possibility of one vague interpretation, also addressing the third part of Definition \ref{UnambiguitySelected}.
+  \item The new phrase ``having one clear interpretation'' is still ambiguous as there could be one clear interpretation, and one or more vague interpretations.
+This is addressed by adding a phrase stating that there are no other possible interpretations besides the one clear interpretation.
 \end{itemize}

### 2021-10-13 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
fixed typo
added mathematical definition of abstraction comment

@@
   \item The term ``interpretation'' is found in both Definition \ref{IEEEunambiguity} and Definition \ref{UnambiguitySelected} and is kept in the chosen definition.
-  \item The first two part of Definition \ref{UnambiguitySelected} can be succinctly restated as ``having one interpretation''.
+  \item The first two parts of Definition \ref{UnambiguitySelected} can be succinctly restated as ``having one interpretation''.
   \item The fourth part of Definition \ref{UnambiguitySelected} can be added to the above phrase, resulting in the new phrase ``having one clear interpretation''.
@@
 
+We are not exploring a mathematical definition of abstraction at this time. \cite{ma1991types} discuss type-based abstraction captured in mathematical theory.
+
 Abstraction is related to reusability (and other qualities).

### 2021-10-15 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
added mathematical definition of abstraction comment - fixed error
Updates to the definition of the quality of usability

@@
     we will say \emph{the extent to which}.
+    \item Definitions use phrases like ``goals'' and ``requirements''.  We will
+    use the term ``requirements'', since it implies something more concrete than
+    a goal.
     \item \wss{Fill in additional consistency decisions, as we go through
@@
 
-\begin{defn}
-  ``The effort required to learn, operate, prepare input, and interpret output
-  of a program'' \citep{McCallEtAl1977}.
-\end{defn}
-
-\begin{defn}
-  ``A software is usable - or user friendly - if its human users find it easy
-  to use'' \citep{ghezzi1991fundamentals}. \wss{switch to 2003 ref?}
-\end{defn}
-
-\begin{defn}
+\begin{defn} \label{UsabilityMcCallEtAl1977}
+  The effort required to learn, operate, prepare input, and interpret output
+  of a program \citep{McCallEtAl1977}.
+\end{defn}
+
+\begin{defn} \label{UsabilityISO9126}
+  The capability of the software to be understood, learned, used and liked by
+  the user, when used under specified conditions. \cite{ISO9126}
+\end{defn}
+
+\begin{defn} \label{UsabilityGhezziEtAl2003}
+  A software system is usable - or \emph{user friendly} - if its human users
+  find it easy to use. \citep[p.\ 22]{GhezziEtAl2003}.
+\end{defn}
+
+\begin{defn} \label{UsabilityISO16982:2002}
   The extent to which a product can be used by specified users to achieve
   specified goals with effectiveness, efficiency, and satisfaction in a
-  specified context of use. \cite{ISO16982:2002} and \cite{ISO9241:11}
-\end{defn}
-
-Nielsen and (separately) Schneidermann have defined usability as part of
-usefulness and is composed of:
-\begin{defn}
+  specified context of use. \citep{ISO16982:2002, ISO9241:11}
+\end{defn}
+
+\begin{defn} \label{UsabilityNielsenNorman}
+Usability as part of usefulness and is composed of:
 \begin{itemize}
@@
 
-\begin{defn}
+\begin{defn} \label{UsabilityDesFoundation}
 Desirable outcomes of usability:
@@
 
-One core reference, for definitions and metrics, is
-\citet{bevan1995measuring}.
-
-\begin{defn}
-  The capability of the software to be understood, learned, used and liked by
-  the user, when used under specified conditions \cite{ISO9126}.
-\end{defn}
+% One core reference, for definitions and metrics, is
+% \citet{bevan1995measuring}.
 
@@
 \subsubsection*{Usability}
-The extent to which a product can be used by specified users to achieve
-specified goals with effectiveness, efficiency, and satisfaction in a specified
-context of use. \cite{ISO16982:2002} and \cite{ISO9241:11}
-\end{mybox}
-
-\subsubsection*{Reasoning}
-
-There has been a lot of thought put behind the ISO definition, from experts. The
-standard has been revisited, but the definition has not changed, which means
-that it has stood the test of time.  The different reasons for the proposed
-definition are as follows:
-
-\begin{itemize}
-  \item reason 1
-  \item reason 2
-  \item ...
+The extent to which software can be used by specified users to achieve specified
+requirements with effectiveness, efficiency, and satisfaction in a specified
+context of use. 
+\end{mybox}
+
+\subsubsection*{Reasoning}
+
+The selected definition is a modified version of
+Definition~\ref{UsabilityISO16982:2002}.  This definition comes from an ISO
+standard that has been revisited multiple times, without the definition for
+usability changing; therefore, the definition has stood the test of time.
+Additional justifications for selecting the proposed definition are as follows:
+
+\begin{itemize}
+  \item The selected definition uses our selected phrase of ``the extent to
+  which'' for consistency with the other definitions.
+  \item The selected definition is less ambiguous compared
+  to~\ref{UsabilityMcCallEtAl1977},~\ref{UsabilityISO9126},~\ref{UsabilityGhezziEtAl2003}
+  and ~\ref{UsabilityDesFoundation} because the selected definition talks about
+  specified users, goals and context of use.
+  \item The proposed definition covers the ``learn, operate, prepare input, and
+  interpret output'' from Definition~\ref{UsabilityMcCallEtAl1977} and ``Learnability'', Memorability'' and ``Errors'' from Definition~\ref{UsabilityNielsenNorman} by the more
+  abstract phrase ``achieve specified requirements''.
+  \item We changed ``the product'' from Definition~\ref{UsabilityISO16982:2002}
+  to ``software'', since this document covers software qualities.
+  \item We changed ``goals'' from Definition~\ref{UsabilityISO16982:2002} to
+  ``requirements'' for the goal of being unambiguous and consistent.
 \end{itemize}
@@
 
-We are not exploring a mathematical definition of abstraction at this time. \cite{ma1991types} discuss type-based abstraction captured in mathematical theory.
+We are not exploring a mathematical definition of abstraction at this time. \cite{reynolds1983types} discusses type-based abstraction captured in mathematical theory.
 

### 2021-10-15 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
Completion of usability quality

@@
 
-\subsection{\notdone{Usability (external, product, process)}} %\notdone \jc{owner}
+\subsection{Usability (external, product, process)} %\jc{owner}
 
@@
 \subsubsection*{Usability}
-The extent to which software can be used by specified users to achieve specified
-requirements with effectiveness, efficiency, and satisfaction in a specified
-context of use. 
+The extent to which a software product, or development process, can be used by
+specified users to achieve specified requirements with effectiveness,
+efficiency, and satisfaction in a specified context of use. 
 \end{mybox}
@@
   \item The proposed definition covers the ``learn, operate, prepare input, and
-  interpret output'' from Definition~\ref{UsabilityMcCallEtAl1977} and ``Learnability'', Memorability'' and ``Errors'' from Definition~\ref{UsabilityNielsenNorman} by the more
-  abstract phrase ``achieve specified requirements''.
+  interpret output'' from Definition~\ref{UsabilityMcCallEtAl1977} and
+  ``Learnability'', Memorability'' and ``Errors'' from
+  Definition~\ref{UsabilityNielsenNorman} by the more abstract phrase ``achieve
+  specified requirements''.
   \item We changed ``the product'' from Definition~\ref{UsabilityISO16982:2002}
-  to ``software'', since this document covers software qualities.
+  to ``software product or development process'', since this document covers
+  software qualities and the quality definition applies to both the product and
+  the process.
   \item We changed ``goals'' from Definition~\ref{UsabilityISO16982:2002} to
   ``requirements'' for the goal of being unambiguous and consistent.
+  \item Everything from Definition~\ref{UsabilityISO9126} is included in the
+  proposed definition because ``understood and learned and used'' are covered by
+  effectively and efficiently achieving specified requirements.  The term
+  ``liked'' in Definition~\ref{UsabilityISO9126} is covered by user satisfaction
+  in the proposed definition.
+  \item The user friendliness synonym for usability, as given by
+  Definition~\ref{UsabilityGhezziEtAl2003}, is excluded for consistency; we only
+  want to introduce one term: usability.
+  \item Definition~\ref{UsabilityDesFoundation} is consistent with the selected
+  definition, but Definition~\ref{UsabilityDesFoundation} itself is not
+  abstract, since it incorporates examples.
 \end{itemize}

### 2021-10-20 StateOfPractice/Methodology/Methodology.tex
Updates to Methodology document

@@
 \usepackage{xcolor}
+\usepackage{longtable}
+\usepackage{paralist}
+\usepackage{totcount}
+\usepackage{booktabs}
 
@@
 
-\theoremstyle{definition}
-\newtheorem{defn}{Definition}
+\newcommand{\CC}{C\nolinebreak\hspace{-.05em}\raisebox{.4ex}{\small\bf
++}\nolinebreak\hspace{-.10em}\raisebox{.4ex}{\small\bf +}}
+
+\newcounter{totHours} %to track the estimate of total hours to assess SOP
+\setcounter{totHours}{0}
+\regtotcounter{totHours}
 
@@
 \begin{abstract}
-	...
+	To improve software development methods and tools for research software, we
+	first need to understand the current state of the practice.  Therefore, we
+	have developed a methodology for assessing the state of the software
+	development practices for a given research software domain.  The methodology
+	is applied to one domain at a time in recognition that software development in
+	different domains is likely to have adopted different best practices.
+	Moreover, providing a means to measure different domains facilitates
+	comparison of development practices between domains.  For each domain we wish
+	to answer questions such as: 
+  \begin{inparaenum}[i)]
+    \item What artifacts (documents, code, test cases, etc.) are present?
+    \item What tools are used?
+    \item What principles, process and methodologies are used?
+    \item What are the pain points for developers?
+    \item What actions are used to improve qualities like maintainability and
+    reproducibility?
+  \end{inparaenum} 
+  To answer these questions, our methodology prescribes the following steps: 
+  \begin{inparaenum}[i)] 
+    \item Identify the domain;
+    \item Identify a list of candidate software packages;
+    \item Filter the list to a length of about 30 packages;
+    \item Gather source code and documentation for each package;
+    \item Collect raw data on each software repo, like number of stars, number
+    of open issues, number of lines of code, and so on;
+    \item Fill in the measurement template (the template consists of 108 questions
+    to assess 9 qualities (like installability, usability and visibility));
+    \item Interview developers (the interview consists of 20 questions and takes
+    about an hour);
+    \item Rank the software using the Analytic Hierarchy Process (AHP); and,
+    \item Analyze the data to answer the questions posed above.
+  \end{inparaenum}
+  A domain expert should be engaged throughout the process, to ensure that
+  implicit information about the domain is properly represented and to assist
+  with conducing an analysis of the commonalities and variabilities between the
+  30 selected packages.  Using our methodology, spreadsheet templates and AHP
+  tool, we estimate (based on our experience with using the process) the time to
+  complete an assessment for a given domain at \total{totHours} person hours.
 \end{abstract}
@@
 Given the importance of research software, scientists and engineers are pushing
-for methods and tools to sustainably develop hiqh quality software.  This is
+for methods and tools to sustainably develop high quality software.  This is
 evident from the existence of such groups as the Software Sustainability
@@
 \citep{MilewiczAndRaybourn2018}.  To improve this situation, we need to first
-fully understand the current state of the practice for research software.
+fully understand the current state of the practice for research software
+development.
 
@@
 a category, we will reduce the scope of our methodology to focus on one specific
-research software domain at a time.  This ``state of the practice for domain X''
-exercise will build off of prior work on measuring/assessing the state of
-software development practice in several research domains.  We will update the
-work that was done previously for domains such as Geographic Information Systems
-\citep{SmithEtAl2018_arXivGIS}, Mesh Generators \citep{SmithEtAl2016},
-Seismology software \citep{SmithEtAl2018}, and Statistical software for
-psychology \citep{SmithEtAl2018_StatSoft}.
+research software domain at a time.  To emphasize that the method is generic, we
+will label the specific domain as X within this document.
+
+This ``state of the practice for domain X'' exercise builds off of prior work on
+measuring/assessing the state of software development practice in several
+research domains.  We have updated the work that was done previously for domains
+such as Geographic Information Systems \citep{SmithEtAl2018_arXivGIS}, Mesh
+Generators \citep{SmithEtAl2016}, Seismology software \citep{SmithEtAl2018}, and
+Statistical software for psychology \citep{SmithEtAl2018_StatSoft}.  Initial
+tests of the new methodology have been done for medical image analysis software
+\citep{Dong2021} and for Lattice Boltzmann Method (LBM) software
+\citep{Michalski2021}.
 
@@
 projects for each domain, but the measures were relatively shallow.  With this
-re-boot we will still target about 30 software examples from each domain, but we
-will be collecting more data by adding more measures and experiments.  In
-keeping with the previous project, we still have the constraint that the work
-load for applying the methodology to a given domain needs to be feasible for a
-team as small as one individual, and for a time that is short, ideally 3 to 4
-months of part-time work per domain.
+re-boot we still target about 30 software examples from each domain, but we are
+now collecting more data.  In keeping with the previous project, we still have
+the constraint that the work load for applying the methodology to a given domain
+needs to be feasible for a team as small as one individual, and for a time that
+is short, ideally under a person month per domain.\footnote{A person month is
+considered to be $20$ working days ($4$ weeks in a month, with $5$ days of work
+per week) at $8$ person hours per day, or $20 \cdot 8 = 160$ person hours.}
 
-To start with, the 
+To begin our re-boot of the previous methodology we critically assessed, and
+subsequently modified, our
 \href{https://github.com/adamlazz/DomainX/blob/master/TemplateToGradeSCSoft.pdf}
-{previous set of questions} were critically assessed and modified.  In addition,
-the following data is part of the new methodology:
+{previous set of measures}.  In addition, the following data has been added to
+the new methodology:
 
@@
   domain via a commonality analysis.
-\item Usability testing for a sample of software in the domain.
-\item Empirical software engineering related data, such as the number of
-  files, number of lines of code, percentage of issues that are closed, etc. 
+\item Project repository related data, such as the number of files, number of
+  lines of code, percentage of issues that are closed, etc. 
+\item Interviews with software developers in domain X.
+%\item Usability testing for a sample of software in the domain.  
 \end{itemize}
 
-Unlike the previous measurement process, the new methodology will involves and
+Unlike the previous measurement process, the new methodology involves and
 engages a domain expert partner throughout.  We did not previously engage the
-domain expert with the rationale that we wished to exclude potential bias.
-However, this advantage is not worth not being able to evaluate the
-functionality/usability of the software.  Moreover, not having an expert makes
-publication more difficult, since there is no one to provide advice on how best
-to approach journals and publishers.
+domain expert with the rationale that we wished to eliminate potential bias.
+However, this advantage is not worth the inability to evaluate the functionality
+of the software.  Moreover, not having an expert makes navigating the on-line
+resources difficult, since some information will be implicit to domain experts,
+and not every statement found on-line will be accurate   The importance of the
+domain expert is particularly noteworthy when it comes time to publication and
+dissemination of the state of the practice assessment.  Throughout this document
+the person (or persons) that provide domaine expertise will be designated as the
+\textit{Domain Expert}.
 
-In the proposed methodology, the collected data will be combined to rank the
-software within each domain using the Analytic Hierarchy Process (AHP)
-\citep{Saaty1980}.  As in the past process, we will use AHP to develop a list of software
-ranked by quality.  However, in the new process we will not stop with this
-list.  The domain expert will be consulted to verify the ordering, and to
-discuss the decisions that led to the ranking.  The AHP process will be used to
-facilitate a conversation with the domain experts as a means to deepen our
-understanding of the software in the domain, and the needs of typical developers.
+In the proposed methodology, the collected data is combined to rank the software
+within each domain using the Analytic Hierarchy Process (AHP) \citep{Saaty1980}.
+As in the previous measurement exercise, we use AHP to develop a list of
+software ranked by quality.  However, in the new process we do not stop with
+this list.  The Domain Expert is consulted to verify the ordering, and to
+discuss the decisions that led to the ranking.  The AHP process is used to
+facilitate a conversation with the Domain Expert as a means to deepen our
+understanding of the software in the domain, and the needs of typical
+developers.
 
-GitHub will be used to coordinate the work of the team of people that will
-be involved in this project.  In addition to the project record left on GitHub,
-the final data will be exported to Mendeley (as was done for the previous
-quality measurement exercise, for instance the grades for GIS software are
-summarized at \href{https://data.mendeley.com/datasets/6kprpvv7r7/1}
-{https://data.mendeley.com/datasets/6kprpvv7r7/1}.)
+So that the collected data for a given domain can benefit the scientific
+community, our recommendation is that all collected data be made public.  For
+instance, the data collection for each domain could be put on a GitHub
+repository.  In addition to the project record left on GitHub, the final data
+can be exported to \href{https://data.mendeley.com/}{Mendeley Data}.  As an
+example, the measurements for the
+\href{https://data.mendeley.com/datasets/6kprpvv7r7/1} {state of practice on GIS
+software} are available on Mendeley.  Ideally the full analysis of the state of
+the practice for domain X will be published in a suitable journal.
 
@@
 software using performance benchmarks to other projects, such as the work of
-\cite{kaagstrom1998gemm}.
+\cite{kaagstrom1998gemm}.  Currently we are also leaving experiments to measure
+usability and modifiability as future work, as discussed in
+Section~\ref{SecFutureWork}.
 
-Note the following formatting conventions of this document. \textcolor{red}{Red}
-text denotes a link to sections of this document. \textcolor{cyan}{Cyan}
-text denotes a URL link. \textcolor{blue}{Blue} text denotes a citation.
+Within this document the following typographic conventions have been adopted:
+\begin{inparaenum}[i)] 
+  \item \textcolor{red}{Red} text denotes a link to sections of this document. 
+  \item \textcolor{cyan}{Cyan} text denotes a URL link.
+  \item \textcolor{blue}{Blue} text denotes a citation.
+\end{inparaenum}
 
-\section{Research Questions} \label{ResearchQuestions}
+The full methodology is presented in the sections that follow.
+Section~\ref{SecResearchQuestions} highlights the research questions that are to
+be answered for each measured domain. These questions are answered by the data
+collected using the process outlined in Section~\ref{StepsAQDS}.  The major
+steps in the process are outlined in
+Sections~\ref{SecIdentifyDomain}--\ref{SecAHP}.  Following this, the time
+required for assessing a single domain is given in
+Section~\ref{SecEstTimeRequired}.
+
+\section{Research Questions} \label{SecResearchQuestions}
 
@@
 \item What artifacts are present in current software packages? 
-\item What tools (development, dependencies, project management) are used by current software packages?
+\item What tools (development, dependencies, project management) are used by
+current software packages?
 \item What principles, processes, and methodologies are used in the development
@@
 
-%In general questions:
-%
-%\begin{enumerate}
-%\item Comparison between domains
-%\item How to measure qualities
-%\item How does the quality compare for projects with the most resources to %those
-%  with the fewest?
-%\item What skills/knowledge are needed by future developers?
-%\item How can the development process be improved?
-%\item What are the common pain points?
-%\end{enumerate}
-%
-%For each domain questions:
-%
-%\begin{enumerate}
-%\item Best examples within the domain
-%\item What software artifacts?
-%\item What are the pain points?
-%\item Any advice on what can be done about the pain points?
-%\end{enumerate}
-%
-%Measure the effort invested and the reward.  Related to sustainability.
-%
-%Collect the data and see what conclusions follow.  For an individual domain,
-%between domains.  The process isn't so much about ranking the software as it is
-%about looking at the software closely and see what conclusions arise.  The
-%measurements are intended to force scrutiny, from different perspectives.
+\section{Overview of Steps in Assessing Quality of the Domain Software}
+\label{StepsAQDS}
 
-\section{Overview of Steps in Assessing Quality of the Domain Software} \label{StepsAQDS}
-
-To answer the above research questions (Section~\ref{ResearchQuestions}), we
+To answer the above research questions (Section~\ref{SecResearchQuestions}), we
 systematically measure the quality of the software through data collection and a
@@
 \begin{enumerate}
-\item Identify the domain. (Section~\ref{SecIdentifyDomain})
-\item \emph{Domain Experts}: Create a top ten list of software packages in the
-  domain. (\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Meeting
-    Agenda withmDomainmExperts.pdf}{Meeting Agenda with Domain Experts})
-\item Brief the Domain Experts on the overall objective, research proposal,
-  research questions, measurement template, survey for short list projects,
-  usability tests, maintainability
-  experiments. (\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Meeting
-    Agenda withmDomainmExperts.pdf}{Meeting Agenda with Domain Experts})
-\item Identify broad list of candidate software packages in the
-  domain. (Section~\ref{SecIdentifyCandSoft})
-\item Preliminary filter of software packages
-  list. (Section~\ref{SecInitialFilter})
-\item \emph{Domain Experts}: Review domain software
-  list. (\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Meeting
-    Agenda withmDomainmExperts.pdf}{Meeting Agenda with Domain Experts})
+\item Identify the domain (X). (Section~\ref{SecIdentifyDomain})
+\item Ask the Domain Expert to create a top ten list of software packages in the
+  domain. (Section~\ref{SecDomainExpert})
+\item Meet with the Domain Expert to brief them on the overall objective,
+  research proposal, research questions, measurement template, and developer
+  interviews. (Section~\ref{SecDomainExpert}) \label{StepMeeting}
+\item Identify broad list of candidate software packages in the domain.
+  (Section~\ref{SecIdentifyCandSoft})
+\item Preliminary filter of software packages list.
+  (Section~\ref{SecInitialFilter})
+\item Review software list with Domain Expert. (Section~\ref{SecDomainExpert})
 \item Domain Analysis. (Section~\ref{SecDomainAnalysis})
-\item \emph{Domain Experts}: Vet domain
-  analysis. (\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Meeting
-    Agenda withmDomainmExperts.pdf}{Meeting Agenda with Domain Experts})
+\item Ask Domain Expert to vet domain analysis.
+  (Section~\ref{SecDomainExpert})
 \item Gather source code and documentation for each prospective software
   package.
-\item Collect empirical measures. (Section~\ref{SecEmpiricalMeasures})
-\item Measure using measurement template. (Section~\ref{SecShallowMeasure})
-\item Survey developers
-(\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Questions
-	to Developers.pdf}{Questions to Developers})
+\item Collect repository based information. (Section~\ref{SecEmpiricalMeasures})
+\item Measure using measurement template. (Section~\ref{SecMeasureTemplate})
+\item Survey developers (Section~\ref{SecSurveyQuestions})
 \item Use AHP process to rank the software packages. (Section~\ref{SecAHP})
-\item Identify a short list of top software packages, typically four to six, for
-  deeper exploration according to the AHP rankings of the measurements.
-\item \emph{Domain Experts}: Vet AHP ranking and short
-  list. (\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Meeting
-    Agenda withmDomainmExperts.pdf}{Meeting Agenda with Domain Experts})
-\item With short list:
-\begin{enumerate}
-\item Usability experiments
-  (\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Experiments.pdf}{Experiments})
-\item Modifiability experiments
-  (\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Experiments.pdf}{Experiments})
-\end{enumerate}
-\item Rank short list. (Section~\ref{SecRankShortList}) \pmi{To be completed -
-    add pairwise comparison tool}
-\item Document answers for research questions.
+% \item Identify a short list of top software packages, typically four to six, for
+%   deeper exploration according to the AHP rankings of the measurements.
+\item Ask Domain Expert to vet AHP ranking. (Section~\ref{SecDomainExpert})
+%\item Rank short list. (Section~\ref{SecRankShortList})
+\item Answer research questions (from Section~\ref{SecResearchQuestions}) and
+document answers.
 \end{enumerate}
@@
 
-\section{How to Identify the Domain} \label{SecIdentifyDomain}
+\section{How to Identify the Domain?} \label{SecIdentifyDomain}
 
@@
 in a scientific or engineering context.  Research software is a more general
-term that covers what is often called scientific computing.  To be applicable
-for the methodology described in this document, the chosen domain must have the
+term for what is often called scientific computing.  To be applicable for the
+methodology described in this document, the chosen domain must have the
 following properties:
@@
 
-\section{How to Identify Candidate Software} \label{SecIdentifyCandSoft}
+\section{Interaction With Domain Expert} \label{SecDomainExpert}
+
+As mentioned in the introduction (Section~\ref{SecIntroduction}), the Domain
+Expert is an important member of the state of the practice assessment team.
+Pitfalls exist if non-domain experts attempt to acquire an authoritative list of
+software, or perform a commonality analysis or try to rank the software. The
+main source of problems for non-domain experts is that they can only rely on
+information that is available on-line, but on-line data has two potential
+problems:
+\begin{inparaenum}[i)]
+  \item the on-line resources could have false or inaccurate information; and,
+  \item the on-line resources could leave out relevant information that is so
+in-grained with experts that nobody thinks to explicitly record the information.
+\end{inparaenum}
+
+Domain experts may be recruited from academia or industry.  The only
+requirements are knowledge of the domain and a willingness to be engaged in the
+assessment process.  The domain expert does not have to be a software developer,
+but they should be a user of domain software.  Given that the domain experts are
+likely to be busy people, the measurement process needs to benefit from their
+expertise, without putting too much of a burden on their time.
+
+In advance of the first meeting with the Domain Expert (Step~\ref{StepsAQDS} in
+Section~\ref{StepsAQDS}) the expert is asked to create a top ten list of
+software packages in the domain.  This is done to help the expert get in the
+right mind set in advance of the first meeting.  Moreover, by doing the exercise
+in advance, we avoid the potential pitfall of the expert approving the
+discovered list of software without giving it adequate thought.  The emphasis
+during the first meeting is for the Domain Expert to learn what is expected of
+them.  The discussion should also cover avenues for publication and
+dissemination.
+
+The Domain Experts provide are asked to vet the collected data and analysis.  In
+particular, they are asked to vet the proposed list of software packages, the
+domain analysis and the AHP ranking.  These interactions can be done either
+electronically or with in-person (or virtual) meetings.
+
+\section{How to Identify Candidate Software?} \label{SecIdentifyCandSoft}
 
@@
 {GitHub}, \href{https://swmath.org/} {swMATH} and domain related publications,
-such as review articles. Domain experts are also asked for their suggestions and
+such as review articles. Domain Experts are also asked for their suggestions and
 are asked to review the initial draft of the software list.
@@
 \item The empirical measures listed in Section \ref{SecEmpiricalMeasures} should
-  ideally be available, which implies a preference for GitHub-style repositories.
-\item The software cannot be marked as incomplete or in an initial development
+  ideally be available, which implies a preference for GitHub-style
+  repositories.
+\item The software cannot be marked as incomplete, or in an initial development
   phase.
@@
 
-\section{How to Initially Filter the Software List} \label{SecInitialFilter}
+\section{How to Initially Filter the Software List?} \label{SecInitialFilter}
 
@@
 
-Copies of both the initial and filtered lists should be kept for traceability purposes.
+Copies of both the initial and filtered lists, along with the rationale for
+shortening the list, should be kept for traceability purposes.
 
@@
 
-Since each domain we will study will have a reasonably small scope, we
-will be able to view the software as constituting a program
-family.  The concept of a program family is defined by \citet{parnas1976design}
-as ``a set of programs whose common properties are so extensive that it is
-advantageous to study the common properties of the programs before analyzing
-individual members''.  Studying the common properties within a family of related
-programs is termed a domain analysis.
+Since each domain we will study will have a reasonably small scope, we will be
+able to view the software as constituting a program family.  The concept of a
+program family is defined by \citet{parnas1976design} as ``a set of programs
+whose common properties are so extensive that it is advantageous to study the
+common properties of the programs before analyzing individual members''.
+Studying the common properties within a family of related programs is termed a
+domain analysis.
 
-The domain analysis consists of a commonality analysis of
-the family of software packages. Its purpose is to show the relationships
-between these packages, and to facilitate an understanding of the informal
-specification and development of them. \cite{weiss1997defining} defines
-commonality analysis as an approach to defining a family by identifying
-commonalities, variabilities, and common terminology for the
-family. Commonalities are goals, theories, models, definitions and assumptions
-that are common between family members.  Variabilities are goals, theories,
-models, definitions and assumptions that differ between family
-members. Associated with each variability are its parameters of variation, which
-summarize the possible values for that variability, along with their potential
-binding time.  The binding time is when the value of the variability is set.  It
-could be set as specification time, build time (when the program is being
-compiled) or run time (when the code is executing).
+The domain analysis consists of a commonality analysis of the family of software
+packages. Its purpose is to show the relationships between these packages, and
+to facilitate an understanding of the informal specification and development of
+them. \cite{weiss1997defining} defines commonality analysis as an approach to
+defining a family by identifying commonalities, variabilities, and common
+terminology for the family. Commonalities are goals, theories, models,
+definitions and assumptions that are common between family members.
+Variabilities are goals, theories, models, definitions and assumptions that
+differ between family members. Associated with each variability are its
+parameters of variation, which summarize the possible values for that
+variability, along with their potential binding time.  The binding time is when
+the value of the variability is set.  It could be set as specification time,
+build time (when the program is being compiled) or run time (when the code is
+executing).
 
-The final result of the domain analysis will be
-tables of commonalities, variabilities, and parameters of variation of a program
-family.   \cite{smith2008commonality} present a template for conducting a
-commonality analysis, which was referred to when conducting this
-work. \cite{weiss1998commonality} describes another commonality analysis
-technique for deciding the members of a program family which readers are
-encouraged to be familiar with. \cite{SmithAndChen2004} and
+The final result of the domain analysis will be tables of commonalities,
+variabilities, and parameters of variation of a program family.
+\cite{smith2008commonality} present a template for conducting a commonality
+analysis, which was referred to when conducting this work.
+\cite{weiss1998commonality} describes another commonality analysis technique for
+deciding the members of a program family. \cite{SmithAndChen2004} and
 \cite{SmithMcCutchanAndCarette2017} are examples of a commonality analysis for a
@@
 \item Write an Introduction
-\item Write the Overview of Domain
+\item Write an Overview of the Domain Knowledge
 \item List Commonalities
@@
 
-A sample commonality analysis for Lattice Boltzmann Solvers can be found
-\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Peter-Notes/Commonality-Analysis-LB-Systems.pdf}
-{here}.
+% A sample commonality analysis for Lattice Boltzmann Solvers can be found
+% \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Peter-Notes/Commonality-Analysis-LB-Systems.pdf}
+% {here}.
 
-\section{Empirical Measures} \label{SecEmpiricalMeasures}
+\section{Repository Based Measures} \label{SecEmpiricalMeasures}
 
-Some quality measurements rely on the gathering of raw and processed empirical
-data. We focus on data that is reasonably easy to collect, which we combine and
-analyze for this assessment. The measures that are collected relate to the
-research questions (Section~\ref{ResearchQuestions}). For instance, we collect some of the
-data to see how large a project is. Other measures intend to ascertain a
-project’s popularity, or how active the project is.
+Some quality measurements rely on gathering raw and processed data from software
+repositories. We focus on data that is reasonably easy to collect, which we
+combine and analyze. The measures that are collected relate to the research
+questions (Section~\ref{SecResearchQuestions}). For instance, we collect data to
+see how large a project is, to ascertain a project’s popularity, and to
+determine whether to project is being actively developed.
 
-Section \ref{rawdata} orients the reader on what raw data is collected.
-Some of this data can be observed from GitHub repository metrics.
-The rest can be collected using freeware tools.
-\href{https://github.com/tomgi/git_stats}{GitStats} is used to measure
-the number of binary files as well as the number of added and deleted lines in a
-repository. This tool is also used to measure the number of commits over
-different intervals of time. \href{https://github.com/boyter/scc}{Sloc Cloc and
-  Code (scc)} is used to measure the number of text based files as well as the
+Section~\ref{rawdata} lists the raw data that is collected.  Some of this data
+can be observed from GitHub repository metrics. The rest can be collected using
+freeware tools. \href{https://github.com/tomgi/git_stats}{GitStats} is used to
+measure the number of binary files as well as the number of added and deleted
+lines in a repository. This tool is also used to measure the number of commits
+over different intervals of time. \href{https://github.com/boyter/scc}{Sloc Cloc
+and Code (scc)} is used to measure the number of text based files as well as the
 number of total, code, comment, and blank lines in a repository. These tools
@@
 empirical measures listed below. Details on installing and running the tools can
-be found in our 
-\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/A
-  Guide to Empirical Measures.pdf} {Guide to Empirical Measures}. Section \ref{processeddata}
-introduces the required processed data, which is calculated using the raw data.
+be found in Appendix~\ref{SecRepoTools}.  Section \ref{processeddata} introduces
+the required processed data, which is calculated using the raw data.
 
@@
 \begin{itemize}
-\item Status of software package (dead or alive). Alive is defined as the
-  presence of repository commits or software package version releases in the
-  last 18 months.
+\item Status of software package as either dead or alive, where alive is defined
+  as the presence of repository commits or software package version releases in
+  the last 18 months.
 \item Percentage of identified issues that are closed.
@@
 
-\section{Measure Using Measurement Template} \label{SecShallowMeasure}
+\section{Measure Using Measurement Template} \label{SecMeasureTemplate}
 
@@
 
-The
-\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Combined_MeasurementTemplate_EmpiricalMeasures.xlsx}
-{Measurement Template} is used to track measurements and quality scores for all
-of the software packages in the domain. For each software package, we fill out one
-column of the template. This process can take between 2 to 5 hours for each
-package.  Project developers can be contacted for help regarding installation,
-if necessary, but a cap of about 2 hours should be imposed on the entire
-installation process, to keep the overall measurement time feasible.
+The Measurement Template is found in
+Appendix~\ref{SecGradingTemplate}.  This template is used to track measurements
+and quality scores for all of the software packages in the domain. For each
+software package, we fill out one column of the template. This process can take
+between 2 to 5 hours for each package.  Project developers can be contacted for
+help regarding installation, if necessary, but a cap of about 2 hours should be
+imposed on the entire installation process, to keep the overall measurement time
+feasible.
 
@@
 \item Assign a score out of ten for each quality. The score can be measured
-  using the
-  \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/MeasurementTemplate_ImpressionCalculator.xlsx}
-  {Measurement Template Impression Calculator}. For each quality measurement,
-  the file indicates the appropriate score to assign the measurement based on
-  possible measurement values.
+  using the Measurement Template Impression Calculator, found in
+  Appendix~\ref{SecImpressionCalculator}. For each quality measurement, the file
+  indicates the appropriate score to assign the measurement based on possible
+  measurement values.
 \end{enumerate}
@@
 used for comparing and ranking the software packages of a domain using the
-quality scores that are gathered in the
-\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Combined_MeasurementTemplate_EmpiricalMeasures.xlsx}
-{Measurement Template}. AHP performs a pairwise analysis between each of the
-quality options using a matrix which is then used to generate an overall score
-for each software package for the given criteria. \cite{SmithEtAl2016} shows how
-AHP is applied to ranking software based on quality measures. We have developed
-a tool for conducting this process. The tool includes an AHP JAR script and a
+quality scores that are gathered in the Measurement Template
+(Appendix~\ref{SecGradingTemplate}). AHP performs a pairwise analysis between
+each of the quality options using a matrix which is used to generate an overall
+score for each software package for the given criteria. \cite{SmithEtAl2016}
+shows how AHP is applied to ranking software based on quality measures. We have
+developed a
+\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/AHP2020/LBM/README.txt}{tool}
+for conducting this process. The tool includes an AHP JAR script and a
 sensitivity analysis JAR script that is used to ensure that the software package
-rankings are appropriate with respect to the uncertainty of the quality
-scores. The README file of the tool is found
-\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/AHP2020/LBM/README.txt}{here}. This
-file outlines the requirements for, and configuration and usage of, the JAR
+rankings are appropriate with respect to the uncertainty of the quality scores.
+The
+\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/AHP2020/LBM/README.txt}{README
+file} outlines the requirements for, and configuration and usage of, the JAR
 scripts. The JAR scripts, source code, and required libraries are located in the
@@
 
-\section{Rank Short List} \label{SecRankShortList}
+% \section{Rank Short List} \label{SecRankShortList}
 
-An AHP pairwise comparison between the short list packages with respect to
-usability and modifiability experiment results is conducted, and then reviewed by the domain expert. The tool can be found here. The usability and modifiability
-survey can be found
-\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Experiments.pdf}
-{here}. This step is intended to start a conversation about the relative quality
-of short-listed software packages.
+% An AHP pairwise comparison between the short list packages with respect to
+% usability and modifiability experiment results is conducted, and then reviewed
+% by the domain expert. The tool can be found here. The usability and
+% modifiability survey can be found
+% \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Experiments.pdf}
+% {here}. This step is intended to start a conversation about the relative quality
+% of short-listed software packages.
+
+\section{Estimate of Time Required} \label{SecEstTimeRequired}
+
+estimate the time required in person hours.  Add a table.  Assume reviewed
+material, using provided spreadsheets and AHP tool.  Assume domain has been
+identified and domain expert engaged.  Following steps given in
+Section~\ref{StepsAQDS}.
+
+\begin{table}[h]
+  \caption{Estimated Person Hours for Assessing the State of Practice for Domain
+  X} \label{TabPersonHours}
+  \centering
+  \begin{tabular}{p{10cm} l}
+    \toprule
+    \textbf{Task} & \textbf{Hours} \\
+    \midrule
+
+    Initial 1 hour meeting with domain expert plus meeting prep & 10
+    \addtocounter{totHours}{10} \\
+
+    Identify candidate software & 10 \addtocounter{totHours}{10} \\
+
+    Filter software list & 5 \addtocounter{totHours}{5} \\
+
+    Domain analysis & 20 \addtocounter{totHours}{20} \\
+
+    Interaction with domain expert(s) (initial meeting, review software list,
+    vet domain analysis, vet AHP ranking) & 10 \addtocounter{totHours}{10} \\
+
+    Collect raw data and fill-in measurement template & 30
+    \addtocounter{totHours}{30} \\
+
+    Solicit developers for interviews & 2 \addtocounter{totHours}{2} \\
+
+    Conduct interviews & 15 \addtocounter{totHours}{15} \\
+
+    AHP ranking & 2 \addtocounter{totHours}{2} \\
+
+    Analyze data and answer research questions & 30 \addtocounter{totHours}{30}
+    \\
+
+    \midrule
+    \textbf{Total} & \textbf{\thetotHours} \\
+    \bottomrule
+  \end{tabular}
+
+\end{table}
+
+\section{Future Work} \label{SecFutureWork}
+
+- usability and modifiability experiments
+
+Usability experiments (\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Experiments.pdf}{Experiments})
+
+Modifiability experiments (\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Experiments.pdf}{Experiments})
+
+- not just for research software
+
+- use it to inform research on Drasil - what artifacts, what process dev
+currently use, how could the new process promoted by Drasil fit into the current
+development processes?
+
+- meta analysis of collected data.  Already have GIS, mesh generator.  Currently
+working on medical image analysis and LBM.
+
+\section{Concluding Remarks} \label{SecConcludingRemarks}
+
+- summary
 
@@
 \appendix
+
+\section{A Guide to Repository-Based Measurement Tools} \label{SecRepoTools}
+
+This appendix covers the tools used for collecting repository based data
+(Section~\ref{SecEmpiricalMeasures}).  The two tools covered are git\_stats and
+scc.  The tools do not have to be used in any particular order.
+
+\subsection{git\_stats}
+
+\subsubsection{Introduction}
+
+Source Code: \href{https://github.com/tomgi/git_stats}{GitHub repo}
+
+\subsubsection{User Manual} \label{git_stats_manual}
+
+Official Manual: \href{https://github.com/tomgi/git_stats}{GitHub repo}
+
+\subsubsection{Demo of Installation and Running the Tool}
+
+The installation steps on your machine may be different from this section.
+Please refer to the user manual mentioned in Section \ref{git_stats_manual}, if
+necessary.  The steps shown here were executed on a virtual machine with 8 cores
+and 16 GB RAM running Debian GNU/Linux 9.11
+
+\begin{enumerate}
+\item Install ruby/gem environment
+\begin{lstlisting}
+apt-get install ruby ruby-nokogiri ruby-nokogiri-diff ruby-nokogumbo
+\end{lstlisting}
+
+Check the installation:
+\begin{lstlisting}
+gem --version
+\end{lstlisting}
+
+\item Install the tool
+\begin{lstlisting}
+sudo gem install git_stats
+\end{lstlisting}
+
+\item Prepare the target repo
+
+Make sure the target repo (the repo to be analyzed, not the repo of this tool)
+is on your machine.
+In this demo, the target repo is downloaded from a GitHub repo:
+\begin{lstlisting}
+# change [git path] to the url of your target repo
+git clone [git path]
+# e.g. git clone https://github.com/nroduit/Weasis.git
+\end{lstlisting}
+
+\item Generate analytics
+\begin{lstlisting}
+# make sure [repo path] is the target repo path
+# the [output path] can be anywhere you desire
+git_stats generate -p [repo path] -o [output path]
+# e.g. git_stats generate -p /home/user/git-stats/Weasis -o 
+# /home/user/git-stats/Weasis-analytics
+\end{lstlisting}
+
+\item View the analytics
+
+View the analytic results by open [output path]/index.html with any browser or
+other software supporting HTML web page format.
+\begin{figure}[h!]
+\includegraphics[scale=0.4]{git-stats-demo.JPG}
+\end{figure}
+
+\item Download the data
+
+On most of the taps of this web page, the data can be downloaded for more
+analytics by clicking the menu button beside the data-range section.
+
+\end{enumerate}
+
+\subsection{scc}
+
+\subsubsection{Introduction}
+
+Source Code: \href{https://github.com/boyter/scc}{GitHub repo}
+
+\subsubsection{User Manual} \label{scc_manual}
+
+Official Manual: \href{https://github.com/boyter/scc}{GitHub repo}
+
+\subsubsection{Demo of Installation and Running the Tool}
+
+The installation steps on your machine may be different from this section.
+Please refer to the user manual mentioned in Section \ref{scc_manual}, if
+necessary.  These steps were executed on a virtual machine with 8 cores and 16
+GB RAM running Debian GNU/Linux 9.11
+
+\begin{enumerate}
+\item Install Golang
+
+Follow the \href{https://golang.org/doc/install}{official instructions}, or the
+following demo,
+
+download the installation package:
+\begin{lstlisting}
+wget https://dl.google.com/go/go1.14.3.linux-amd64.tar.gz
+\end{lstlisting}
+
+unpack to /usr/local:
+\begin{lstlisting}
+sudo tar -C /usr/local -xzf go1.14.3.linux-amd64.tar.gz
+\end{lstlisting}
+
+use a text editor to open \textasciitilde/.profile, e.g.:
+\begin{lstlisting}
+nano ~/.profile
+\end{lstlisting}
+
+add the following lines to the end of this file:
+\begin{lstlisting}
+export GOPATH=$HOME/go
+export PATH=$PATH:/usr/local/go/bin:$GOPATH/bin
+\end{lstlisting}
+
+save the file, and load the commands into the current shell instance:
+\begin{lstlisting}
+source ~/.profile
+\end{lstlisting}
+
+check the installation:
+\begin{lstlisting}
+go version
+\end{lstlisting}
+
+\item Install the tool
+\begin{lstlisting}
+go get -u github.com/boyter/scc/
+\end{lstlisting}
+
+\item Prepare the target repo
+
+Make sure the target repo (the repo to be analyzed, not the repo of this tool)
+is on your machine.
+In this demo, the target repo is downloaded from a GitHub repo:
+\begin{lstlisting}
+# change [project path] to your desired folder
+cd [project path]
+git clone https://github.com/nroduit/Weasis.git
+\end{lstlisting}
+
+\item Generate analytics
+\begin{lstlisting}
+# make sure [repo path] is the target repo path
+cd [repo path]
+# use scc to generate analytics
+scc
+\end{lstlisting}
+
+\item View the analytics
+
+The results will be appear something like the following.
+
+\begin{figure}[h!]
+\includegraphics[scale=0.5]{scc-demo.JPG}
+\end{figure}
+
+\end{enumerate}
+
+\newpage
+
+\section{Measurement Template} \label{SecGradingTemplate}
+
+The table below lists the full set of measures that are assessed for each
+software product.  The measures are grouped under headings for each quality, and
+one for summary information.  Following each measure, the type for a valid
+result is given in brackets.  Many of the types are given as enumerated sets.
+For instance, the response on many of the questions is one of ``yes,'' ``no,''
+or ``unclear.''  The type ``number'' means natural number, a positive integer.  The
+types for date and url are not explicitly defined, but they are what one would
+expect from their names.  In some cases the response for a given question is not
+necessarily limited to one answer, such as the question on what platforms are
+supported by the software product.  Case like this are indicated by ``set of''
+preceding the type of an individual answer.  The type in these cases are then
+the power set of the individual response type.  In some cases a superscript $^*$
+is used to indicate that a response of this type should be accompanied by
+explanatory text.  For instance, if problems were caused by uninstall, the
+reviewer should note what problems were caused.  An (I) precedes the question
+description when its measurement requires a successful installation.
+
+The table below lists the set of measures that are used to assess each software
+product. The first set identifies summary information, followed by 9 sets for
+software qualities and 3 sets for raw metrics. Each measure is followed by the
+type for a valid result. A superscript $^*$ indicate that a response of this type
+needs to be accompanied by explanatory text.
+
+\newpage
+
+\captionof{table}{Measurement Template}\label{measurementtemplate}
+\def\arraystretch{1.22}
+\begin{tabular}{p{14cm}}
+	\hline
+	\textbf{Summary Information}\\
+	\hline
+	Software name? (string)\\
+	URL? (URL)\\
+	Affiliation (institution(s)) (string or {N/A})\\
+	Software purpose (string)\\
+	Number of developers (all developers that have contributed at least one commit
+	to the project) (use repo commit logs) (number)\\
+	How is the project funded? (unfunded, unclear, funded$^*$) where $^*$ requires a
+	string to say the source of funding\\
+	Initial release date? (date)\\
+	Last commit date? (date)\\
+	Status? (alive is defined as presence of commits in the last 18 months)
+	({alive, dead, unclear})\\
+	License? ({GNU GPL, BSD, MIT, terms of use, trial, none, unclear, other$^*$}) $^*$
+	given via a string \\
+	Platforms? (set of {Windows, Linux, OS X, Android, other$^*$}) $^*$ given via
+	string\\
+	Software Category? The concept category includes software that does not have
+	an officially released version. Public software has a released version in the
+	public domain. Private software has a released version available to authorized
+	users only. ({concept, public, private})\\
+	Development model? ({open source, freeware, commercial, unclear})\\
+	Publications about the software? Refers to publications that have used or
+	mentioned the software. (number or {unknown})\\
+	Source code URL? ({set of url, n/a, unclear})\\
+	Programming language(s)? (set of {FORTRAN, Matlab, C, C++, Java, R, Ruby,
+	Python, Cython, BASIC, Pascal, IDL, unclear, other$^*$}) $^*$ given via string \\
+	Is there evidence that performance was considered? Performance refers to
+	either speed, storage, or throughput. ({yes$^*$, no})\\
+	Additional comments? (can cover any metrics you feel are missing, or any other
+	thoughts you have) \\
+	\hline
+\end{tabular}
+
+\def\arraystretch{1.5}
+\begin{tabular}{p{14cm}}
+	\hline		
+	\textbf{Installability (Measured via installation on a virtual machine.) }\\
+	\hline
+	Are there installation instructions? ({yes, no})\\
+	Are the installation instructions in one place? Place referring to a single
+	document or web-page. ({yes, no, n/a})\\
+	Are the installation instructions linear? Linear meaning progressing  in a
+	single series of steps. ({yes, no, n/a})\\
+	Are the instructions written as if the person doing the installation has none
+	of the dependent packages installed? ({yes, no, unclear})\\
+	Are compatible operating system versions listed? ({yes, no})\\
+	Is there something in place to automate the installation (makefile, script,
+	installer, etc)? ({yes$^*$, no})\\
+	If the software installation broke, was a descriptive error message displayed?
+	({yes, no, n/a})\\
+	Is there a specified way to validate the installation? ({yes$^*$, no})\\
+	How many steps were involved in the installation? (Includes manual steps like
+	unzipping files) Specify OS. (number, OS)\\
+	What OS was used for the installation? ({Windows, Linux, OS X, Android, other$^*$
+	}) $^*$given via string\\
+	How many extra software packages need to be installed before or during
+	installation? (number)\\
+	Are required package versions listed? ({yes, no, n/a})\\
+	Are there instructions for the installation of required packages /
+	dependencies? ({yes, no, n/a})\\
+	Run uninstall, if available. Were any obvious problems caused? ({yes$^*$ , no,
+	unavail})\\
+	Overall impression? ({1 .. 10})\\
+	Additional comments? (can cover any metrics you feel are missing, or any other
+	thoughts you have)\\
+	\hline
+\end{tabular}
+
+\def\arraystretch{1.33}
+\begin{tabular}{p{14cm}}
+	\hline		
+	\textbf{Correctness and Verifiability}\\
+	\hline
+	Any reference to the requirements specifications of the program or theory
+	manuals? ({yes$^*$ , no, unclear})\\
+	What tools or techniques are used to build confidence of correctness?
+	({literate programming, automated testing, symbolic execution, model checking,
+	assertions used in the code, Sphinx, Doxygen, Javadoc, confluence, unclear,
+	other$^*$}) $^*$ given via string\\
+	If there is a getting started tutorial? ({yes, no})\\
+	Are the tutorial instructions linear? ({yes, no, n/a})\\
+	Does the getting started tutorial provide an expected output? ({yes, no$^*$,
+	n/a})\\
+	Does your tutorial output match the expected output? ({yes, no, n/a})\\
+	Are unit tests available?  ({yes, no, unclear})\\
+	Is there evidence of continuous integration? (for example mentioned in
+	documentation, Jenkins, Travis CI, Bamboo, other) ({yes$^*$, no, unclear})\\
+	Overall impression? ({1 .. 10})\\
+	Additional comments? (can cover any metrics you feel are missing, or any other
+	thoughts you have) \\
+	\hline	
+	\textbf{Surface Reliability}\\
+	\hline
+	Did the software “break” during installation? ({yes$^*$ , no})\\
+	If the software installation broke, was the installation instance recoverable?
+	({yes, no, n/a})\\
+	Did the software “break” during the initial tutorial testing? ({yes$^*$, no,
+	n/a})\\
+	If the tutorial testing broke, was a descriptive error message displayed?
+	({yes, no, n/a})\\
+	If the tutorial testing broke, was the tutorial testing instance recoverable?
+	({yes, no, n/a})\\
+	Overall impression? ({1 .. 10})\\
+	Additional comments? (can cover any metrics you feel are missing, or any other
+	thoughts you have)\\
+	\hline		
+\end{tabular}
+
+\def\arraystretch{1.4}
+\begin{tabular}{p{14cm}}
+	\hline		
+	\textbf{Surface Robustness}\\
+	\hline
+	Does the software handle unexpected/unanticipated input (like data of the
+	wrong type, empty input, missing files or links) reasonably? (a reasonable
+	response can include an appropriate error message.) ({yes, no$^*$ })\\
+	For any plain text input files, if all new lines are replaced with new lines
+	and carriage returns, will the software handle this gracefully? ({yes, no$^*$,
+	n/a})\\
+	Overall impression? ({1 .. 10})\\
+	Additional comments? (can cover any metrics you feel are missing, or any other
+	thoughts you have)\\
+	\hline		
+	\textbf{Surface Usability}\\
+	\hline
+	Is there a getting started tutorial? ({yes, no})\\
+	Is there a user manual? ({yes, no})\\
+	Are expected user characteristics documented? ({yes, no})\\
+	What is the user support model? FAQ? User forum? E-mail address to direct
+	questions? Etc. (string)\\
+	Overall impression? ({1 .. 10})\\
+	Additional comments? (can cover any metrics you feel are missing, or any other
+	thoughts you have)\\
+	\hline
+\end{tabular}
+
+\def\arraystretch{1.4}
+\begin{tabular}{p{14cm}}
+		\hline	
+	\textbf{Maintainability}\\
+	\hline
+	What is the current version number? (number)\\
+	Is there any information on how code is reviewed, or how to contribute?
+	({yes$^*$, no})\\
+	Are artifacts available? (List every type of file that is not a code file –
+	for examples please look at the ‘Artifact Name’ column of
+	https://gitlab.cas.mcmaster.ca/SEforSC/se4sc/-/blob/git-svn/GradStudents/Olu/ResearchProposal/Artifacts\_MiningV3.xlsx)
+	({yes$^*$, no, unclear}) $^*$list via string\\
+	What issue tracking tool is employed? (set of {Trac, JIRA, Redmine, e-mail,
+	discussion board, sourceforge, google code, git, BitBucket, none, unclear,
+	other$^*$}) $^*$ given via string\\
+	What is the percentage of identified issues that are closed? (percentage)\\
+	What percentage of code is comments? (percentage)\\
+	Which version control system is in use? ({svn, cvs, git, github, unclear,
+	other$^*$}) $^*$ given via string\\
+	Overall impression? ({1 .. 10})\\
+	Additional comments? (can cover any metrics you feel are missing, or any other
+	thoughts you have)\\
+	\hline		
+	\textbf{Reusability}\\
+	\hline
+	How many code files are there? (number)\\
+	Is API documented? ({yes, no, n/a})\\
+	Overall impression? ({1 .. 10})\\
+	Additional comments? (can cover any metrics you feel are missing, or any other
+	thoughts you have)\\
+	\hline		
+\end{tabular}
+
+\def\arraystretch{1.4}
+\begin{tabular}{p{14cm}}
+		\hline		
+	\textbf{Surface Understandability (Based on 10 random source files)}\\
+	\hline
+	Consistent indentation and formatting style? ({yes, no, n/a})\\
+	Explicit identification of a coding standard? ({yes$^*$, no, n/a})\\
+	Are the code identifiers consistent, distinctive, and meaningful? ({yes, no$^*$ ,
+	n/a})\\
+	Are constants (other than 0 and 1) hard coded into the program? ({yes, no$^*$ ,
+	n/a})\\
+	Comments are clear, indicate what is being done, not how? ({yes, no$^*$ , n/a})\\
+	Parameters are in the same order for all functions? ({yes, no$^*$ , n/a})\\
+	Is the name/URL of any algorithms used mentioned? ({yes, no$^*$ , n/a})\\
+	Is code modularized? ({yes, no$^*$ , n/a})\\
+	Overall impression? ({1 .. 10})\\
+	Additional comments? (can cover any metrics you feel are missing, or any other
+	thoughts you have)\\
+	\hline		
+	\textbf{Visibility/Transparency}\\
+	\hline
+	Is the development process defined? If yes, what process is used. ({yes$^*$, no,
+	n/a})\\
+	Are there any documents recording the development process and status?  ({yes$^*$,
+	no}))\\
+	Is the development environment documented? ({yes$^*$, no})\\
+	Are there release notes? ({yes$^*$, no})\\
+	Overall impression? ({1 .. 10})\\
+	Additional comments? (can cover any metrics you feel are missing, or any other
+	thoughts you have)\\
+		\hline		
+	\textbf{Raw Metrics (Measured via git\_stats)}\\
+	\hline
+	Number of text-based files. (number)\\
+	Number of binary files. (number)\\
+	Number of total lines in text-based files. (number)\\
+	Number of total lines added to text-based files. (number)\\
+	Number of total lines deleted from text-based files. (number)\\
+	Number of total commits. (number)\\
+	Numbers of commits by year in the last 5 years. (Count from as early as
+	possible if the project is younger than 5 years.) (list of numbers)\\
+	Numbers of commits by month in the last 12 months. (list of numbers)\\
+	\hline
+\end{tabular}
+
+\def\arraystretch{1.4}
+\begin{tabular}{p{14cm}}
+		\hline		
+	\textbf{Raw Metrics (Measured via scc)}\\
+	\hline
+	Number of text-based files. (number)\\
+	Number of total lines in text-based files. (number)\\
+	Number of code lines in text-based files. (number)\\
+	Number of comment lines in text-based files. (number)\\
+	Number of blank lines in text-based files. (number)\\
+	\hline
+	\textbf{Repo Metrics (Measured via GitHub)}\\
+	\hline
+	Number of stars. (number)\\
+	Number of forks. (number)\\
+	Number of people watching this repo. (number)\\
+	Number of open pull requests. (number)\\
+	Number of closed pull requests. (number)\\
+	\hline
+\end{tabular}
+
+\newpage
+
+\section{Measurement Template Impression Calculator} \label{SecImpressionCalculator}
+
+The table below lists how each quality measure of the measurement template is
+used to calculate an overall impression in each software quality set.
+
+\newpage
+
+\captionof{table}{Measurement Template}
+\def\arraystretch{1.5}
+\begin{tabular}{p{14cm}}
+	\hline		
+	\textbf{Installability (Measured via installation on a virtual machine.) }\\
+	\hline
+	Are there installation instructions? ({yes=1, no=-1})\\
+	Are the installation instructions in one place? Place referring to a single
+	document or web-page. ({yes=1, no=0, n/a=0})\\
+	Are the installation instructions linear? Linear meaning progressing  in a
+	single series of steps. ({yes=1, no=0, n/a=0})\\
+	Are the instructions written as if the person doing the installation has none
+	of the dependent packages installed? ({yes=1, no=0, unclear=0})\\
+	Are compatible operating system versions listed? ({yes=1, no=0})\\
+	Is there something in place to automate the installation (makefile, script,
+	installer, etc)? ({yes$^*$=1, no=-1})\\
+	If the software installation broke, was a descriptive error message displayed?
+	({yes=0, no=-2, n/a=1})\\
+	Is there a specified way to validate the installation? ({yes$^*$=1, no=0})\\
+	How many steps were involved in the installation? (Includes manual steps like
+	unzipping files) Specify OS. ($<$10 = 1)\\
+	What OS was used for the installation? (does not count)\\
+	How many extra software packages need to be installed before or during
+	installation? ($<$10 = 1)\\
+	Are required package versions listed? ({yes=1, no=0, n/a=1})\\
+	Are there instructions for the installation of required packages /
+	dependencies? ({yes=1, no=0, n/a=1})\\
+	Run uninstall, if available. Were any obvious problems caused? ({yes$^*$=0, no=1,
+	unavail=1})\\
+	Overall impression? (a sum of $>$10 is rounded down to 10)\\
+	\hline
+\end{tabular}
+
+$\mathit{ab} \mathit{val}$
+
+\def\arraystretch{1.33}
+\begin{tabular}{p{14cm}}
+	\hline		
+	\textbf{Correctness and Verifiability}\\
+	\hline
+	Any reference to the requirements specifications of the program or theory
+	manuals? ({yes$^*$=2, no=0, unclear=0})\\
+	What tools or techniques are used to build confidence of correctness? (any=1,
+	unclear=0)\\
+	If there is a getting started tutorial? ({yes=2, no=0})\\
+	Are the tutorial instructions linear? ({yes=1, no=0, n/a=0})\\
+	Does the getting started tutorial provide an expected output? ({yes=1, no$^*$=0,
+	n/a=0})\\
+	Does your tutorial output match the expected output? ({yes=1, no=0, n/a=0})\\
+	Are unit tests available?  ({yes=1, no=0, unclear=0})\\
+	Is there evidence of continuous integration? (for example mentioned in
+	documentation, Jenkins, Travis CI, Bamboo, other) ({yes$^*$=1, no=0,
+	unclear=0})\\
+	\hline	
+	\textbf{Surface Reliability}\\
+	\hline
+	Did the software “break” during installation? ({yes$^*$=0, no=5})\\
+	If the software installation broke, was the installation instance recoverable?
+	({yes=5, no=0, n/a=0})\\
+	Did the software “break” during the initial tutorial testing? ({yes$^*$=0, no=5,
+	n/a=0})\\
+	If the tutorial testing broke, was a descriptive error message displayed?
+	({yes=2, no=0, n/a=0})\\
+	If the tutorial testing broke, was the tutorial testing instance recoverable?
+	({yes=3, no=0, n/a=0})\\
+	\hline		
+	\textbf{Surface Robustness}\\
+	\hline
+	Does the software handle unexpected/unanticipated input (like data of the
+	wrong type, empty input, missing files or links) reasonably? (a reasonable
+	response can include an appropriate error message.) ({yes=5, no$^*$=0})\\
+	For any plain text input files, if all new lines are replaced with new lines
+	and carriage returns, will the software handle this gracefully? ({yes=5,
+	no$^*$=0, n/a=5})\\
+	\hline		
+\end{tabular}
+
+\def\arraystretch{1.4}
+\begin{tabular}{p{14cm}}
+		\hline		
+	\textbf{Surface Usability}\\
+	\hline
+	Is there a getting started tutorial? ({yes=3, no=0})\\
+	Is there a user manual? ({yes=4, no=0})\\
+	Are expected user characteristics documented? ({yes=1, no=0})\\
+	What is the user support model? FAQ? User forum? E-mail address to direct
+	questions? Etc. (one=1, two+=2, none=0)\\
+	\hline
+	\textbf{Maintainability}\\
+	\hline
+	What is the current version number? (provided=1, nothing=0)\\
+	Is there any information on how code is reviewed, or how to contribute?
+	({yes$^*$=1, no=0})\\
+	Are artifacts available? (List every type of file that is not a code file –
+	for examples please look at the ‘Artifact Name’ column of
+	https://gitlab.cas.mcmaster.ca/SEforSC/se4sc/-/blob/git-svn/GradStudents/Olu/ResearchProposal/Artifacts\_MiningV3.xlsx)
+	(Rate 0 – 2 depending on how many and perceived quality)\\
+	What issue tracking tool is employed? (nothing=0, email of other private=1,
+	anything public or accessible by all devs (eg git) = 2)\\
+	What is the percentage of identified issues that are closed? (50$\%$+=1,
+	$<$50$\%$=0)\\
+	What percentage of code is comments? (10$\%$+=1, $<$10$\%$=0)\\
+	Which version control system is in use? (anything=2, nothing=0)\\
+	\hline		
+	\textbf{Reusability}\\
+	\hline
+	How many code files are there? (0-9=0, 10-49=1, 50-99=3, 100-299=4, 300-599=5,
+	600-999=6, 1000+=8)\\
+	Is API documented? ({yes=2, no=0, n/a=0})\\
+	\hline
+\end{tabular}
+
+\def\arraystretch{1.4}
+\begin{tabular}{p{14cm}}
+\hline	
+\textbf{Surface Understandability (Based on 10 random source files)}\\
+\hline
+Consistent indentation and formatting style? ({yes=1, no=0, n/a=0})\\
+Explicit identification of a coding standard? ({yes$^*$=1, no=0, n/a=0})\\
+Are the code identifiers consistent, distinctive, and meaningful? ({yes=2,
+no$^*$=0, n/a=0})\\
+Are constants (other than 0 and 1) hard coded into the program? ({yes=1, no$^*$=0,
+n/a=0})\\
+Comments are clear, indicate what is being done, not how? ({yes=2, no$^*$=0,
+n/a=0})\\
+Is the name/URL of any algorithms used mentioned? ({yes=1, no$^*$=0, n/a=0})\\
+Parameters are in the same order for all functions? ({yes=1, no$^*$=0, n/a=0})\\
+Is code modularized? ({yes=1, no$^*$=0, n/a=0})\\
+\hline		
+\textbf{Visibility/Transparency}\\
+\hline
+Is the development process defined? If yes, what process is used. ({yes$^*$=3,
+no=0, n/a=0})\\
+Are there any documents recording the development process and status?  ({yes$^*$=3,
+no=0}))\\
+Is the development environment documented? ({yes$^*$=2, no=0})\\
+Are there release notes? ({yes$^*$=2, no=0})\\
+\hline
+\end{tabular}
+
+\newpage
+
+\section{Survey Questions for Developers} \label{SecSurveyQuestions}
+
+Introduction.  Roadmap.  Anything about traceability to research questions.
+
+Information about these interview questions:  This gives you an idea what I
+would like to learn about the development of {domain} software. Interviews will
+be one-to-one and will be open-ended (not just “yes or no” answers). Because of
+this, the exact wording may change a little. Sometimes I will use other short
+questions to make sure I understand what you told me or if I need more
+information when we are talking such as: “So, you are saying that ...?), to get
+more information (“Please tell me more?”), or to learn what you think or feel
+about something (“Why do you think that is...?”).
+
+\subsection{Information about the developers and users} \citep{Jegatheesan2016}
+\begin{enumerate}
+\item Interviewees' current position/title? degrees?
+\item Interviewees' contribution to/relationship with the software?
+\item Length of time the interviewee has been involved with this software?
+\item How large is the development group?
+\item Do you have a defined process for accepting new contributions into your team?
+\item What is the typical background of a developer?
+\item What is your estimated number of users? How did you come up with that estimate?
+\item What is the typical background of a user?
+\end{enumerate}
+
+\subsection{Information about the software}
+
+Square brackets are used for traceability to qualities and relevant research
+questions found in Section ...
+
+\begin{enumerate}
+\item Currently, what are the most significant obstacles in your development
+process?
+\item How might you change your development process to remove or reduce these
+obstacles?
+\item How does documentation fit into your development process? Would improved
+documentation help with the obstacles you typically face? [traceability -
+research question 5b, visibility/transparency - research question 5i]
+\item In the past, is there any major obstacle to your development process that
+has been solved? How did you solve it?
+\item What is your software development model? For example, waterfall, agile,
+etc.
+\item What is your project management process? Do you think improve this process
+can tackle the current problem? Were any project management tools used?
+\item Was it hard to ensure the correctness of the software? If there were any
+ obstacles, what methods have been considered or practiced to improve the
+ situation? If practiced, did it work? [correctness - research question 5e]
+\item When designing the software, did you consider the ease of future changes?
+ For example, will it be hard to change the structure of the system, modules or
+ code blocks? What measures have been taken to ensure the ease of future changes
+ and maintains? [maintainability - research question 5d, modifiability -
+ research question 5c]
+\item Provide instances where users have misunderstood the software. What, if
+any, actions were taken to address understandability issues? [understandability
+- research question 5f]
+\item What, if any, actions were taken to address usability issues? [usability -
+research question 5a]
+\item Do you think the current documentation can clearly convey all necessary
+knowledge to the users? If yes, how did you successfully achieve it? If no, what
+improvements are needed? [unambiguity - research question 5g]
+\item Do you have any concern that your computational results won't be
+reproducible in the future? Have you taken any steps to ensure reproducibility?
+[reproducibility - research question 5h]
+\end{enumerate}
 

### 2021-10-20 StateOfPractice/Methodology/Methodology.tex
fixed typo

@@
 Research software uses computing to simulate mathematical models of real world
-systems so that we can better understand and predict those systems's behaviour.  A
+systems so that we can better understand and predict those systems' behaviour.  A
 small set of examples of important research software includes the

### 2021-10-20 StateOfPractice/Methodology/Methodology.tex
Revised intro paragraph to Section 12 (Estimate of Time Required)

@@
 
-estimate the time required in person hours.  Add a table.  Assume reviewed
-material, using provided spreadsheets and AHP tool.  Assume domain has been
-identified and domain expert engaged.  Following steps given in
-Section~\ref{StepsAQDS}.
+Table~\ref{TabPersonHours} estimates the time required (in person hours) to
+complete a state of the practice assessment for domain X.  The table assumes
+that the domain has already been decided and the domain expert has been
+recruited.  The table follows the steps outlined in Section~\ref{StepsAQDS}.
+Time is not included for reviewing the methodology. Moreover, it is assumed that
+the template spreadsheets linked in this document, and the developed AHP tool,
+will be employed, rather than developing new tools.  The person hours given are
+a rough estimate, based on our experience completing assessments for medical
+image analysis software \citep{Dong2021} and for Lattice Boltzmann Method (LBM)
+software \citep{Michalski2021}.  These two domains were assessed at the same
+time as designing the methodology presented in this document.  We did our best
+to estimate the time spent on measurement and separate it from the time spend on
+design and development.  The estimate assumes 30 software packages will be
+measured; the numbers will need to be adjusted if the total packages changes.
 

### 2021-10-20 StateOfPractice/Methodology/Methodology.tex
Updates to time estimates in Section 12

@@
 \item Review software list with Domain Expert. (Section~\ref{SecDomainExpert})
-\item Domain Analysis. (Section~\ref{SecDomainAnalysis})
+\item Domain Analysis (with the help of the Domain Expert).
+(Section~\ref{SecDomainAnalysis})
 \item Ask Domain Expert to vet domain analysis.
@@
 
-The Measurement Template is found in
-Appendix~\ref{SecGradingTemplate}.  This template is used to track measurements
-and quality scores for all of the software packages in the domain. For each
-software package, we fill out one column of the template. This process can take
-between 2 to 5 hours for each package.  Project developers can be contacted for
-help regarding installation, if necessary, but a cap of about 2 hours should be
-imposed on the entire installation process, to keep the overall measurement time
-feasible.
+The Measurement Template is found in Appendix~\ref{SecGradingTemplate}.  This
+template is used to track measurements and quality scores for all of the
+software packages in the domain. For each software package, we fill out one
+column of the template. This process can take between 1 to 4 hours for each
+package.  Project developers can be contacted for help regarding installation,
+if necessary, but a cap of about 2 hours should be imposed on the entire
+installation process, to keep the overall measurement time feasible.
 
@@
 
-    Initial 1 hour meeting with domain expert plus meeting prep & 10
-    \addtocounter{totHours}{10} \\
-
-    Identify candidate software & 10 \addtocounter{totHours}{10} \\
-
-    Filter software list & 5 \addtocounter{totHours}{5} \\
-
-    Domain analysis & 20 \addtocounter{totHours}{20} \\
-
-    Interaction with domain expert(s) (initial meeting, review software list,
-    vet domain analysis, vet AHP ranking) & 10 \addtocounter{totHours}{10} \\
-
-    Collect raw data and fill-in measurement template & 30
-    \addtocounter{totHours}{30} \\
+    Initial 1 hour meeting with domain expert plus meeting prep & 5
+    \addtocounter{totHours}{5} \\
+
+    Identify broad list of candidate software
+    (Section~\ref{SecIdentifyCandSoft}) & 12 \addtocounter{totHours}{12} \\
+
+    Filter software list (Section~\ref{SecInitialFilter}) (10 minutes per
+    package for 30 packages) & 5 \addtocounter{totHours}{5} \\
+
+	Review software list with Domain Expert (Section~\ref{SecDomainExpert}) & 2
+	\addtocounter{totHours}{2} \\
+
+    Domain analysis (with help of Domain Expert)
+    (Section~\ref{SecDomainAnalysis}) & 25 \addtocounter{totHours}{20} \\
+
+	Vet domain analysis with Domain Expert (Section~\ref{SecDomainExpert}) & 3
+	\addtocounter{totHours}{3} \\
+
+	Gather source code and documentation for each package (20 minutes per
+	package for 30 packages)) & 10 \addtocounter{totHours}{10} \\
+
+    Collect repository based data (Section~\ref{SecEmpiricalMeasures}) (10
+    minutes per package for 30 packages) & 5 \addtocounter{totHours}{5} \\
+
+	Measure using measurement template (Section~\ref{SecMeasureTemplate}) (2.5
+    hours per repo for 30 repos) & 75 \addtocounter{totHours}{75} \\
 
@@
 
-    Conduct interviews & 15 \addtocounter{totHours}{15} \\
+    Conduct interviews (1.5 hour interviews with 10 developers (assuming 1 in 3
+    developers agree to an interview)) & 15 \addtocounter{totHours}{15} \\
 
@@
 
-    Analyze data and answer research questions & 30 \addtocounter{totHours}{30}
-    \\
+	Work with Domain Expert to vet AHP ranking & 2 \addtocounter{totHours}{2} \\
+
+    Analyze data and answer research questions & 30 \addtocounter{totHours}{30}\\
 

### 2021-10-20 StateOfPractice/Methodology/Methodology.tex
Revisions to Introduction section

@@
     \item Gather source code and documentation for each package;
-    \item Collect raw data on each software repo, like number of stars, number
-    of open issues, number of lines of code, and so on;
-    \item Fill in the measurement template (the template consists of 108 questions
-    to assess 9 qualities (like installability, usability and visibility));
+    \item Collect repository related data on each software package, like number
+    of stars, number of open issues, number of lines of code;
+    \item Fill in the measurement template (the template consists of 108
+    questions to assess 9 qualities (including the qualities of installability,
+    usability and visibility));
     \item Interview developers (the interview consists of 20 questions and takes
@@
 Software (\href{https://bssw.io/} {BSS}).  Sustainability promoting groups such
-as these are necessary because unfortunately the current state-of-the-practice
-for research software often does not incorporate state-of-the-art Software
-Engineering (SE) tools and methods \citep{JohansonAndHasselbring2018}.  The lack
-of SE tools and methods contributes to sustainability and reliability problems
-\citep{FaulkEtAl2009}.  Problems with the current state of the practice are
-evident from embarrassing failures, like a retraction of derived molecular
+as these are necessary because unfortunately the current ``state of the
+practice'' for research software often does not incorporate ``state of the art''
+Software Engineering (SE) tools and methods \citep{JohansonAndHasselbring2018}.
+The lack of SE tools and methods contributes to sustainability and reliability
+problems \citep{FaulkEtAl2009}.  Problems with the current state of the practice
+are evident from embarrassing failures, like a retraction of derived molecular
 protein structures \citep{Miller2006}, false reproduction of sonoluminescent
@@
 needs to be feasible for a team as small as one individual, and for a time that
-is short, ideally under a person month per domain.\footnote{A person month is
+is short, ideally around a person month per domain.\footnote{A person month is
 considered to be $20$ working days ($4$ weeks in a month, with $5$ days of work
@@
 of the software.  Moreover, not having an expert makes navigating the on-line
-resources difficult, since some information will be implicit to domain experts,
-and not every statement found on-line will be accurate   The importance of the
-domain expert is particularly noteworthy when it comes time to publication and
-dissemination of the state of the practice assessment.  Throughout this document
-the person (or persons) that provide domaine expertise will be designated as the
-\textit{Domain Expert}.
+resources difficult, since on-line resources are often silent on information
+that is implicit to domain experts.  Furthermore, not every statement found
+on-line will necessarily be accurate.  The importance of the domain expert is
+particularly noteworthy when it comes time for publication and dissemination of
+the state of the practice assessment.  Throughout this document the person (or
+persons) that provide domain expertise will be designated as the \textit{Domain
+Expert}.
 
@@
 community, our recommendation is that all collected data be made public.  For
-instance, the data collection for each domain could be put on a GitHub
-repository.  In addition to the project record left on GitHub, the final data
-can be exported to \href{https://data.mendeley.com/}{Mendeley Data}.  As an
-example, the measurements for the
-\href{https://data.mendeley.com/datasets/6kprpvv7r7/1} {state of practice on GIS
-software} are available on Mendeley.  Ideally the full analysis of the state of
-the practice for domain X will be published in a suitable journal.
-
-The scope of this methodology includes observations on product, artifact and
-process quality for research softwares. We leave the assessment of research
-software using performance benchmarks to other projects, such as the work of
-\cite{kaagstrom1998gemm}.  Currently we are also leaving experiments to measure
+instance, the data collection for each domain can be put on a GitHub repository.
+In addition to the project record left on GitHub, the final data can be exported
+to \href{https://data.mendeley.com/}{Mendeley Data}.  As an example, the
+measurements for the \href{https://data.mendeley.com/datasets/6kprpvv7r7/1}
+{state of the practice for GIS software} are available on Mendeley.  Ideally the
+full analysis of the state of the practice for domain X will also be published
+in a suitable journal, allowing for dissemination/feedback and communication.
+
+The scope of this methodology includes observations on product, artifact
+(documentation, test scripts, etc.) and process quality for research software.
+We leave the assessment of the performance of research software, for instance
+using benchmarks, to other projects, such as the work of
+\citet{kaagstrom1998gemm}.  Currently we are also leaving experiments to measure
 usability and modifiability as future work, as discussed in
@@
 collected using the process outlined in Section~\ref{StepsAQDS}.  The major
-steps in the process are outlined in
-Sections~\ref{SecIdentifyDomain}--\ref{SecAHP}.  Following this, the time
-required for assessing a single domain is given in
-Section~\ref{SecEstTimeRequired}.
+steps in the process are outlined in Sections~\ref{SecIdentifyDomain} --
+\ref{SecAHP}.  Following this, the time required for assessing a single domain
+is estimated in Section~\ref{SecEstTimeRequired}.
 

### 2021-10-20 StateOfPractice/Methodology/Methodology.tex
Went through for consistent use of capitalization for Domain Expert

@@
 To answer the above research questions (Section~\ref{SecResearchQuestions}), we
-systematically measure the quality of the software through data collection and a
-series of experiments.  An overview of the measurement process is given in the
-following steps, starting from determining a domain that is suitable for
-measurement:
+systematically measure the quality of the software through data collection.  An
+overview of the measurement process is given in the following steps, starting
+from determining a domain that is suitable for measurement:
 
@@
 requirements are knowledge of the domain and a willingness to be engaged in the
-assessment process.  The domain expert does not have to be a software developer,
+assessment process.  The Domain Expert does not have to be a software developer,
 but they should be a user of domain software.  Given that the domain experts are
@@
 
-The Domain Experts provide are asked to vet the collected data and analysis.  In
+The Domain Experts are asked to vet the collected data and analysis.  In
 particular, they are asked to vet the proposed list of software packages, the
@@
 complete a state of the practice assessment for domain X.  The table assumes
-that the domain has already been decided and the domain expert has been
+that the domain has already been decided and the Domain Expert has been
 recruited.  The table follows the steps outlined in Section~\ref{StepsAQDS}.
@@
 
-    Initial 1 hour meeting with domain expert plus meeting prep & 5
+    Initial 1 hour meeting with the Domain Expert plus meeting prep & 5
     \addtocounter{totHours}{5} \\

### 2021-10-20 StateOfPractice/Methodology/Methodology.tex
Revisions to "How to Identify the Domain?" section

@@
   interviews. (Section~\ref{SecDomainExpert}) \label{StepMeeting}
-\item Identify broad list of candidate software packages in the domain.
+\item Identify the broad list of candidate software packages in the domain.
   (Section~\ref{SecIdentifyCandSoft})
@@
 \item Measure using measurement template. (Section~\ref{SecMeasureTemplate})
-\item Survey developers (Section~\ref{SecSurveyQuestions})
+\item Survey developers. (Section~\ref{SecSurveyQuestions})
 \item Use AHP process to rank the software packages. (Section~\ref{SecAHP})
@@
 
-A domain of research software must be identified. Research software is defined
-in this exercise as ``software that is used to generate, process or analyze
-results that [are intended] to appear in a publication'' \citep{hettrick2014uk}
-in a scientific or engineering context.  Research software is a more general
-term for what is often called scientific computing.  To be applicable for the
-methodology described in this document, the chosen domain must have the
-following properties:
+A domain of research software must be identified to begin the assessment.
+Research software is defined in this exercise as ``software that is used to
+generate, process or analyze results that [are intended] to appear in a
+publication'' \citep{hettrick2014uk} in a scientific or engineering context.
+Research software is a more general term for what is often called scientific
+computing.  To be applicable for the methodology described in this document, the
+chosen domain must have the following properties:
 
@@
 \item A preliminary search, or discussion with experts, suggests that there will
-  be numerous, at least 15, candidate software packages in the domain that
+  be numerous, close to 30, candidate software packages in the domain that
   qualify as `research software'.

### 2021-10-20 StateOfPractice/Methodology/Methodology.tex
Revisions to "Interaction with Domain Experts" section

@@
 Pitfalls exist if non-domain experts attempt to acquire an authoritative list of
-software, or perform a commonality analysis or try to rank the software. The
-main source of problems for non-domain experts is that they can only rely on
-information that is available on-line, but on-line data has two potential
-problems:
+software, or perform a commonality analysis, or try to definitively rank the
+software. The main source of problems for non-domain experts is that they can
+only rely on information that is available on-line, but on-line data has two
+potential problems:
 \begin{inparaenum}[i)]
@@
 but they should be a user of domain software.  Given that the domain experts are
-likely to be busy people, the measurement process needs to benefit from their
-expertise, without putting too much of a burden on their time.
+likely to be busy people, the measurement process cannot put to much of a burden
+on their time.
 

### 2021-10-20 StateOfPractice/Methodology/Methodology.tex
Revisions up to, but not including, Section 10 (Measure using measurement template)
Edits to measurement template section

@@
   a change was made) are eliminated, except in the cases where an older software
-  package appears to be highly recommended and currently in use.
+  package appears to be highly recommended and currently in use.  (The Domain
+  Expert can be consulted on this question, if necessary.)
 \end{enumerate}
@@
 
+\noindent The time frame of 18 months was selected as the separating point
+between alive and dead projects because this is the usual timeframe for
+operating system updates.
+
 \section{Measure Using Measurement Template} \label{SecMeasureTemplate}
-
-\begin{figure}[h!]
-  \begin{center}
-    \includegraphics[width=1.0\textwidth]{measurement_template}
-    \caption{Top of Measurement Template}
-    \label{measurement_template_image}
-  \end{center}
-\end{figure}
 
@@
 if necessary, but a cap of about 2 hours should be imposed on the entire
-installation process, to keep the overall measurement time feasible.
+installation process, to keep the overall measurement time feasible.  To save
+time, a
+\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Combined_MeasurementTemplate_EmpiricalMeasures.xlsx}{blank
+measurement template spreadsheet} has been prepared.  An excerpt of the
+spreadsheet is shown in Figure~\ref{measurement_template_image}.
+
+\begin{figure}[h!]
+	\begin{center}
+	  \includegraphics[width=1.0\textwidth]{measurement_template}
+	  \caption{Excerpt of the Top of the Measurement Template}
+	  \label{measurement_template_image}
+	\end{center}
+  \end{figure}
+
+The full template consists of 108 questions categorized under 9 qualities.  The
+questions were designed to be unambiguous, quantifiable and measurable with
+limited time and domain knowledge. The measures are grouped under headings for
+each quality, and one for summary information. The summary information (shown in
+Figure~\ref{measurement_template_image}) is the first section of the template.
+This section summarizes general information, such as the software name, number
+of developers, etc.  We follow the definitions given by
+\citet{GewaltigAndCannon2012} for the software categories.  Public means software
+intended for public use.  Private means software aimed only at a specific
+group, while the concept category is used for software written simply to demonstrate
+algorithms or concepts. The three categories of development models are: open
+source, where source code is freely available under an open source license;
+free-ware, where a binary or executable is provided for free; and, commercial,
+where the user must pay for the software product.
+
+Following the summary section to measure 9 qualities: 
+\begin{inparaenum}
+	\item installability;
+	\item correctness and verifiability;
+	\item surface reliability;
+	\item surface robustness;
+	\item surface usability;
+	\item maintainability;
+	\item reusability;
+	\item surface understandability; and,
+	\item visibility/transparency. \end}inparaenum} Definitions of these
+qualities are available in a working document on software quality.  Explain the
+use of the word surface.
 
 Most of the data to be collected should be straightforward from reviewing the
-measurement template.  However, in some cases some extra guidance is necessary
+measurement template.  However, in a few cases extra guidance is necessary
 to eliminate ambiguity, as follows:
@@
 \item Continuous integration: Search the software artifacts for any mention of
-  continuous integration. The search function on GitHub can help.
+  continuous integration. The search function on GitHub can help.  In some
+  cases, \texttt{yaml} files will provide a hint that continuous integration is
+  employed.
 \end{enumerate}
 
-For completing the template, the following steps should be followed:
+\noindent To fill-in the spreadsheet template, the following steps should be followed:
 
@@
 \item Gather the summary information into the top section of the document
-  (Figure \ref{measurement_template_image})
-\item Using the GitStats tool that is described in Section
-  \ref{SecEmpiricalMeasures} gather the measurements for the Repo Metrics -
-  GitStats section found near the bottom of the document
-\item Using the SCC tool that is also described in Section
-  \ref{SecEmpiricalMeasures} gather the measurements for the Repo Metrics - SCC
-  section found near the bottom of the document
+  (Figure \ref{measurement_template_image}).
+\item Using the GitStats tool that is described in
+  Section~\ref{SecEmpiricalMeasures} gather the measurements for the Repo
+  Metrics - GitStats section found near the bottom of the spreadsheet.
+\item Using the SCC tool that is also described in
+  Section~\ref{SecEmpiricalMeasures} gather the measurements for the Repo
+  Metrics -- SCC section found near the bottom of the spreadsheet.
 \item If the software package is found on git, gather the measurements for the
-  Repo Metrics - GitHub section found near the bottom of the document
+  Repo Metrics - the GitHub section found near the bottom of the spreadsheet.
 \item Review installation documentation and attempt to install the software
-  package on a virtual machine
+  package on a virtual machine.
 \item Gather the measurements for installability
@@
 \end{enumerate}
+
+As in \citet{SmithEtAl2016}, Virtual machines (VMs) are used to provide an
+optimal testing environments for each MGMP software product. VMs were used
+because it is easier to start with a fresh environment without hav- ing to worry
+about existing libraries and conflicts. Moreover, when the tests are complete
+the VM can be deleted, without any impact on the host operating system. The most
+significant advantage of using VMs is to level the playing field. Every software
+install starts from a clean slate, which removes ``works-on-my-computer``
+errors. In the grading data the details for each VM are noted, including
+hypervisor and operating system version.
 

### 2021-10-20 StateOfPractice/Methodology/Methodology.tex
Updates to methodology section

@@
 template is used to track measurements and quality scores for all of the
-software packages in the domain. For each software package, we fill out one
-column of the template. This process can take between 1 to 4 hours for each
-package.  Project developers can be contacted for help regarding installation,
-if necessary, but a cap of about 2 hours should be imposed on the entire
+software packages in the domain. For each software package, we fill-in the
+template questions. This process can take between 1 to 4 hours for each package.
+Project developers can be contacted for help regarding installation, if
+necessary, but a cap of about 2 hours should be imposed on the entire
 installation process, to keep the overall measurement time feasible.  To save
@@
 \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Combined_MeasurementTemplate_EmpiricalMeasures.xlsx}{blank
-measurement template spreadsheet} has been prepared.  An excerpt of the
-spreadsheet is shown in Figure~\ref{measurement_template_image}.
+measurement template spreadsheet} has been prepared, with the measures as rows.
+An excerpt of the spreadsheet is shown in
+Figure~\ref{measurement_template_image}.  A column should be added to this
+template for each software package to be measured.
 
@@
 	  \includegraphics[width=1.0\textwidth]{measurement_template}
-	  \caption{Excerpt of the Top of the Measurement Template}
+	  \caption{Excerpt of the Top Section of the Measurement Template (Summary Information)}
 	  \label{measurement_template_image}
@@
 
-Following the summary section to measure 9 qualities: 
+Following the summary section are sections to measure 9 qualities: 
 \begin{inparaenum}
@@
 	\item surface understandability; and,
-	\item visibility/transparency. \end}inparaenum} Definitions of these
-qualities are available in a working document on software quality.  Explain the
-use of the word surface.
+	\item visibility/transparency. 
+\end{inparaenum} 
+Definitions of these qualities are available in a
+\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/QDefOfQualities/QDefOfQualities.pdf}{working
+document on software quality}.  Several of the qualities use the word
+``surface''.  This is to highlight that, for these qualities in particular, the
+best that we can do is a shallow measure of the quality.  For instance, we are
+not currently doing any experiments to measure usability.  Instead, we are
+looking for an indication that usability was considered by the developers.  We
+do this by looking for cues in the documentation, like a getting started manual,
+a user manual and documentation of expected user characteristics.
 
@@
 As in \citet{SmithEtAl2016}, Virtual machines (VMs) are used to provide an
-optimal testing environments for each MGMP software product. VMs were used
-because it is easier to start with a fresh environment without hav- ing to worry
-about existing libraries and conflicts. Moreover, when the tests are complete
-the VM can be deleted, without any impact on the host operating system. The most
-significant advantage of using VMs is to level the playing field. Every software
-install starts from a clean slate, which removes ``works-on-my-computer``
-errors. In the grading data the details for each VM are noted, including
-hypervisor and operating system version.
+optimal testing environments for each package. VMs were used because it is
+easier to start with a fresh environment without having to worry about existing
+libraries and conflicts. Moreover, when the tests are complete the VM can be
+deleted, without any impact on the host operating system. The most significant
+advantage of using VMs is to level the playing field. Every software install
+starts from a clean slate, which removes ``works-on-my-computer`` errors. When
+filling in the measurement template spreadsheet, the the details for each VM
+should be noted, including hypervisor and operating system version.
 

### 2021-10-20 StateOfPractice/Methodology/Methodology.tex
Clarification that we aren't estimating how much time the domain expert will spend on the project

@@
 that the domain has already been decided and the Domain Expert has been
-recruited.  The table follows the steps outlined in Section~\ref{StepsAQDS}.
+recruited.  The time spent by the Domain Expert is not included in the numbers
+shown in the table, since the amount of time that the domain expert will work
+independently of the rest of the assessment team will be small.  Moreover, this
+amount of time will vary greatly depending on the preferred work habits of the
+Domain Expert.  The table follows the steps outlined in Section~\ref{StepsAQDS}.
 Time is not included for reviewing the methodology. Moreover, it is assumed that

### 2021-10-20 StateOfPractice/Methodology/Methodology.tex
Further edits to the estimate of time section

@@
     Domain analysis (with help of Domain Expert)
-    (Section~\ref{SecDomainAnalysis}) & 25 \addtocounter{totHours}{20} \\
+    (Section~\ref{SecDomainAnalysis}) & 20 \addtocounter{totHours}{20} \\
 
@@
 
-	Gather source code and documentation for each package (20 minutes per
-	package for 30 packages)) & 10 \addtocounter{totHours}{10} \\
+	Gather source code and documentation for each package (10 minutes per
+	package for 30 packages)) & 5 \addtocounter{totHours}{5} \\
 
@@
 
-    Analyze data and answer research questions & 30 \addtocounter{totHours}{30}\\
+    Analyze data and answer research questions & 20 \addtocounter{totHours}{20}\\
 
@@
 \end{table}
+
+The total number of person hours is \total{totHours} hours.  This is close to
+our goal of 1 month of person hours (160 hours).  The amount of time spent by
+the Domain Expert can be estimated by summing the Domain Expert items in
+Table~\ref{TabPersonHours} and adding an estimate of the time that they will
+independently spend on their assigned tasks.  If we assume that the Domain
+Expert will spend 2 hours on the domain analysis and another 2 hours with
+answering questions, the Domain Expert time will be about 12 person hours.
 

### 2021-10-20 StateOfPractice/Methodology/Methodology.tex
Updates to sections other than future work and conclusions are mostly done

@@
 (Appendix~\ref{SecGradingTemplate}). AHP performs a pairwise analysis between
-each of the quality options using a matrix which is used to generate an overall
-score for each software package for the given criteria. \cite{SmithEtAl2016}
-shows how AHP is applied to ranking software based on quality measures. We have
-developed a
+each of the 9 quality options for each of the 30 software packages.  This
+results in a matrix, which is used to generate an overall score for each
+software package for the given criteria. \cite{SmithEtAl2016} shows how AHP is
+applied to ranking software based on quality measures. We have developed a
 \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/AHP2020/LBM/README.txt}{tool}
@@
 necessary.  The steps shown here were executed on a virtual machine with 8 cores
-and 16 GB RAM running Debian GNU/Linux 9.11
+and 16 GB RAM running Debian GNU/Linux 9.11.
 
@@
 explanatory text.  For instance, if problems were caused by uninstall, the
-reviewer should note what problems were caused.  An (I) precedes the question
-description when its measurement requires a successful installation.
-
-The table below lists the set of measures that are used to assess each software
-product. The first set identifies summary information, followed by 9 sets for
-software qualities and 3 sets for raw metrics. Each measure is followed by the
-type for a valid result. A superscript $^*$ indicate that a response of this type
-needs to be accompanied by explanatory text.
+reviewer should note what problems were caused.  % An (I) precedes the question
+% description when its measurement requires a successful installation.
+The template also include 3 sections at the end for summarizing the repository
+based metrics.  A
+\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Combined_MeasurementTemplate_EmpiricalMeasures.xlsx}
+{blank measurement template spreadsheet} is available to save time with data
+entry.
 
@@
 
-$\mathit{ab} \mathit{val}$
-
 \def\arraystretch{1.33}
@@
 \end{tabular}
+
+% \def\arraystretch{1.22}
+% \begin{tabular}{p{14cm}}
+% \hline
+% \textbf{Repo Metrics (Measured via git_stats)}\\
+% \hline
+% Number of text-based files. (number)\\
+% Number of binary files. (number)\\
+% Number of total lines in text-based files. (number)\\
+% Number of total lines added to text-based files. (number or {not git repo})\\
+% Number of total lines deleted from text-based files. (number or {not git repo})\\
+% Number of total commits. (number or {not git repo})\\
+% Numbers of commits by year in the last 5 years. (Count from as early as possible if the project is younger than 5 years.) (list of numbers or {not git repo}\\
+% Numbers of commits by month in the last 12 months. (list of numbers or {not git repo})\\
+% \hline
+% \end{tabular}
+
+% \def\arraystretch{1.22}
+% \begin{tabular}{p{14cm}}
+% \hline
+% \textbf{Repo Metrics (Measured via scc)}\\
+% \hline
+% Number of text-based files.	(number)\\
+% Number of total lines in text-based files.	(number)\\
+% Number of code lines in text-based files.	(number)\\
+% Number of comment lines in text-based files.	(number)\\
+% Number of blank lines in text-based files.	(number)
+% \hline
+% \end{tabular}
+
+% \def\arraystretch{1.22}
+% \begin{tabular}{p{14cm}}
+% \hline
+% \textbf{Repo Metrics (Measured via GitHub)}\\
+% \hline
+% Number of stars.	(number or {n/a})
+% Number of forks.	(number or {n/a})
+% Number of people watching this repo.	(number or {n/a})
+% Number of open pull requests.	(number or {n/a})
+% Number of closed pull requests.	(number or {n/a})
+% \hline
+% \end{tabular}
 

### 2021-10-22 StateOfPractice/Methodology/Methodology.tex
Final version of Methodology paper, next step to prepare the arXiv version

@@
 colorlinks=true,     % false: boxed links; true: colored links
-linkcolor=red,       % color of internal links (change box color with
-%linkbordercolor)
+linkcolor=blue,      % color of internal links (change box color with linkbordercolor)
 citecolor=blue,      % color of links to bibliography
@@
 \author{Jacques Carette}{McMaster University, Canada}{carette@mcmaster.ca}{}{}
-\author{Olu Owojaiye}{McMaster University, Canada}{owojaiyo@mcmaster.ca}{}{}
 \author{Peter Michalski}{McMaster University, Canada}{michap@mcmaster.ca}{}{}
 \author{Ao Dong}{McMaster University, Canada}{donga9@mcmaster.ca}{}{}
+\author{Olu Owojaiye}{McMaster University, Canada}{owojaiyo@mcmaster.ca}{}{}
 
 \authorrunning{Smith et al.}  \Copyright{Spencer Smith and Jacques Carette and
-Olu Owojaiye and Peter Michalski and Ao Dong}
+ Peter Michalski and Ao Dong and Olu Owojaiye}
 
@@
   implicit information about the domain is properly represented and to assist
-  with conducing an analysis of the commonalities and variabilities between the
+  with conducting an analysis of the commonalities and variabilities between the
   30 selected packages.  Using our methodology, spreadsheet templates and AHP
@@
 \begin{inparaenum}[i)] 
-  \item \textcolor{red}{Red} text denotes a link to sections of this document. 
-  \item \textcolor{cyan}{Cyan} text denotes a URL link.
-  \item \textcolor{blue}{Blue} text denotes a citation.
+  \item \textcolor{red}{Blue} text denotes a link to sections of this document,
+  including citations 
+  \item \textcolor{cyan}{Cyan} text denotes an external URL link.
 \end{inparaenum}
@@
 \item Measure using measurement template. (Section~\ref{SecMeasureTemplate})
-\item Survey developers. (Section~\ref{SecSurveyQuestions})
+\item Survey developers. (Section~\ref{SecSurvey})
 \item Use AHP process to rank the software packages. (Section~\ref{SecAHP})
@@
 
-\begin{figure}[h!]
+\begin{figure}[!ht]
 	\begin{center}
@@
 advantage of using VMs is to level the playing field. Every software install
-starts from a clean slate, which removes ``works-on-my-computer`` errors. When
+starts from a clean slate, which removes ``works-on-my-computer'' errors. When
 filling in the measurement template spreadsheet, the the details for each VM
@@
 
+\section{Survey Developers} \label{SecSurvey}
+
+In the previous state of the practice measurement process
+\citep{SmithEtAl2018_arXivGIS, SmithEtAl2016, SmithEtAl2018}, we only based our
+assessment on information available in on-line software repos.  However, this
+approach meant we weren't able to learn about the development process, the
+attitudes of the developers, the pain points for developers and how the
+developers handle modifiability, reproducibility and usability.  Therefore, in
+the reboot of the measurement process, we have explicitly added a stage for
+interviewing research software developers.
+
+We designed a list of 20 questions to guide our interviews, which can be found
+in Appendix~\ref{SecSurveyQuestions}. Some questions are about the background of
+the software, the development teams, the interviewees, and how they organize the
+projects. We also ask about the developer's understandings of the users. Some
+questions focus on the current and past difficulties, and the solutions the team
+has found, or will try. We also discuss the importance and current situations of
+documentation. A few questions are about specific software qualities, such as
+maintainability, understandability, usability, and reproducibility. The
+interviews are semi-structured based on the question list; we ask follow-up
+questions when necessary. Based on our experience, the interviewees usually
+bring up some exciting ideas that we did not expect, and it is worth expanding
+on these topics.
+
+Our methodology suggests requesting interviews with a developer from each of the
+30 software package.  Requests for interviews are sent to all packages so as to
+not cause a potential bias by singling out any subset of the list. Moreover,
+since not every developer will agree to the interview request, asking 30 times
+will typically yield a reasonable number of responses. In our experience, the
+response rate is between 15\% and 30\%.  In some cases multiple developers from
+the same project will agree to be interviewed. When sending out interview
+requests, we recommend finding the contacts on the projects’ website, or code
+repository, or publications, or the biographic pages of the teams’ institutions.
+We send at most two interview request emails to a contact for each software
+package.  Meeting will typically be held using on-line meeting software, like
+Zoom or Teams.  This facilitates recording and automatic transcription of the
+meetings.
+
+The interviewees should follow a process where they can make informed consent.
+The interviews should follow standard ethics guideline of asking for consent
+before interviewing, recording, and including participant details in the report.
+The interview process presented here was approved by the McMaster University
+Research Ethics Board under the application number 
+\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/MACREM/Application.pdf}
+{MREB\#: 5219}.
+
 \section{Analytic Hierarchy Process} \label{SecAHP}
@@
 The Analytical Hierarchy Process (AHP) is a decision-making technique that can
-be used when comparing multiple options by multiple criteria. In our work AHP is
-used for comparing and ranking the software packages of a domain using the
-quality scores that are gathered in the Measurement Template
-(Appendix~\ref{SecGradingTemplate}). AHP performs a pairwise analysis between
-each of the 9 quality options for each of the 30 software packages.  This
-results in a matrix, which is used to generate an overall score for each
-software package for the given criteria. \cite{SmithEtAl2016} shows how AHP is
-applied to ranking software based on quality measures. We have developed a
+be used when comparing multiple options by multiple criteria. AHP focuses on
+pair-wise comparisons between all options for all criteria.  The advantage of
+pair-wise comparisons is that they facilitates a separation of concerns.  Rather
+than worry about the entire problem, the decision maker can focus on one
+comparison at a time.  In our work AHP is used for comparing and ranking the
+software packages of a domain using the quality scores that are gathered in the
+Measurement Template (Appendix~\ref{SecGradingTemplate}). AHP performs a
+pairwise analysis between each of the 9 quality options for each of the 30
+software packages.  This results in a matrix, which is used to generate an
+overall score for each software package for the given criteria.
+\cite{SmithEtAl2016} shows how AHP is applied to ranking software based on
+quality measures. We have developed a
 \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/AHP2020/LBM/README.txt}{tool}
@@
 
-- usability and modifiability experiments
-
-Usability experiments (\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Experiments.pdf}{Experiments})
-
-Modifiability experiments (\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Experiments.pdf}{Experiments})
-
-- not just for research software
-
-- use it to inform research on Drasil - what artifacts, what process dev
-currently use, how could the new process promoted by Drasil fit into the current
-development processes?
-
-- meta analysis of collected data.  Already have GIS, mesh generator.  Currently
-working on medical image analysis and LBM.
+As explained in the introduction (Section~\ref{SecIntroduction}), our eventual
+goal is to improve the state of practice of software development, which requires
+a baseline means for measuring the current state of the practice.  Now that we
+have a methodology for assessing the state of practice for a given domain, the
+next task is to complete the measurements for multiple domains.  Using the
+previous and updated methodologies, we have measured the state of
+practice for the following domains: Geographic Information Systems
+\citep{SmithEtAl2018_arXivGIS}, Mesh Generators \citep{SmithEtAl2016},
+Seismology software \citep{SmithEtAl2018}, Statistical software for psychology
+\citep{SmithEtAl2018_StatSoft}, medical image analysis software \citep{Dong2021}
+and for Lattice Boltzmann Method software \citep{Michalski2021}.  Future domains
+for measurement could include finite element software, computational medicine,
+machine learning, ordinary differential equation solvers, computer graphics,
+stoichiometry, etc.
+
+With the wealth of data from assessing the state of practice for multiple
+domains, the next step is a meta-analysis.  We would look at how the different
+domains compare. What lessons from one domain could be applied in other domains?
+What (if any) differences exist in the pain points between domains?  Are there
+differences in the tools, processes, and documentation between domains?
+
+The current methodology is constrained by limited resources.  A 4 hour cap on
+the measurement time for each software package limits what can be assessed.
+Within this limit, we can't measure some important qualities, like usability and
+modifiability.  In the future, we propose a more time-consuming process that
+would capture these other quality measures.  To improve the feasibility, the
+more time consuming measurements would not have to be completed for all 30
+packages. Instead, a short list could be identified using the output of the AHP
+ranking to select the top projects, or to select a sample of interesting
+projects across the quality spectrum.
+
+If we can add measures for modifiability and usability, we can start to measure
+the quality impact of software development processes, tools and techniques.  For
+instance, we have a project (called
+\href{https://github.com/JacquesCarette/Drasil} {Drasil}
+\citep{SzymczakEtAl2016}), which facilitates a development process that focuses
+on knowledge capture, followed by generation of code and other software
+artifacts from the captured knowledge.  To understand the advantages and
+disadvantages of Drasil, we could measure its quality.  In particular, we would
+like to see the impact of a generative process on the quality of usability and
+modifiability.
+
+\subsection{Measuring Usability in the Future}
+
+In the future, we propose an experiment for assessing the usability of a given
+software package.  Some initial thoughts on how this might be done are recorded
+in this section.  To do the experiment we need an experimental subject, who will
+be required to complete tasks with the software being studied.  The interaction
+with the software will allow the study subject to experience the software's
+usability.  The tasks for the subject to complete would vary by domain;
+therefore, the tasks would be selected with the help of the Domain Expert.
+Criteria for selecting candidate tasks are as follows:
+
+\begin {enumerate}
+\item Tasks should be executable for subjects with novice to intermediate
+experience.
+\item All tasks should take no more than one hour.
+\item Tasks should include the basic/common use cases of the software package.
+\item Include tasks that require sequential or hierarchical steps for
+completion.
+\end {enumerate}
+
+Once the study subject has experience with the software, they will be in a
+position to judge its usability.  We will measure the usability using a
+standardized usability questionnaire, like the
+\href{https://www.usabilitest.com/sus-pdf-generator} {System Usability Scale
+questionnaire} or the
+\href{https://uiuxtrend.com/pssuq-post-study-system-usability-questionnaire/}
+{Post-Study System Usability Questionnaire}.
+
+As a starting point for the experiment design, the procedure could be something
+like the following:
+
+\begin {enumerate}
+\item Survey participants to collect pre-experiment data (background, experience
+of subjects (especially with domain software)).
+\item Participants perform tasks based on task defined by Domain Experts.
+\item Observe the study subjects (take notes, record sessions (screen
+recorder), watch for body languages and verbal cues).
+\item Survey the study subjects to collect feedback (post-experiment interview),
+complete usability questionnaire.
+\item Prepare a summary report of the experimental results.
+\end {enumerate}
+
+\subsection{Measuring Modifiability in the Future}
+
+The next experiment is designed to gather qualitative data regarding the
+modifiability of each software package. This proposed experiment also involves
+experimental subjects/participants, who in this case are asked to modify a set
+of software packages.  The specific modifications requested will again depend on
+the software domain.  In advance of the experiment the Domain Expert will be
+asked the likely changes for software in this domain.  We emphasize likely
+changes, instead of any changes because software cannot be designed so that
+everything is equally easy to change \citep{parnas1986rational}.  The procedure
+could be along the following lines:
+
+\begin{enumerate}
+	\item Domain Expert lists all likely changes that a developer
+	might be asked to make in a software package in the domain.
+	\item Survey study participants to collect pre-experiment data (background,
+	experience of subjects (especially with domain software)).
+	\item Participants perform modification tasks for likely changes on each
+	software package being studied.
+	\item Observe the study subjects (take notes, record sessions (screen
+	recorder), watch for body languages and verbal cues).
+	\item Record time needed to make the changes.	
+	\item Confirm through testing that the modified software has the correct
+	behaviour.
+	\item Survey the study subjects to collect feedback (post-experiment interview).
+	\item Prepare a summary report of the experimental results.
+\end{enumerate}
+
+The study subjects should make the same changes in multiple pieces of software.
+The reporting of the results will focus more on the relative time differences
+between the set of software packages, rather than the absolute time to make any
+given change. To remove biases caused by the participants experience, the
+different study subjects should use a different order as they go through the
+list of software packages.  Details for this experiment still need to be
+resolved, such as how to take the participants prior knowledge into account,
+especially with respect to their programming experience.
 
@@
 
-- summary
+We have outlined a methodology for assessing the state of the practice for any
+given research software domain.  (Although the scope of the current work has been
+on research software, there is little in the methodology that is specific to
+research software, except for the interview question related to the quality of
+reproducibility.)  When applying the methodology to a given domain, we provide a
+means to answer the following questions:
+\begin{inparaenum}[i)]
+\item What artifacts (documents, code, test cases, etc.) are present?
+\item What tools are used?
+\item What principles, process and methodologies are used?
+\item What are the pain points for developers?
+\item What actions are used to improve qualities like maintainability and
+reproducibility?
+\item What specific actions are taken to achieve the qualities of usability,
+traceability, modifiability, maintainability, correctness, understandability,
+unambiguity, reproducibility and visibility/transparency?
+\item How does software designated as high quality by this methodology compare
+  with top rated software by the community?
+\end{inparaenum}
+
+The methodology depends on the engagement of a Domain Expert.  The Domain
+Expert's role is to ensure that the assessment is consistent with the culture of
+the community of practitioners in the domain.  The Domain Expert also has an
+important role to play with the domain analysis.  For each domain we 
+conduct a domain analysis to look at the commonalities, variabilities and
+parameters of variation, for the family of software in the domain.  The domain
+analysis means that software can be compared not just based on its quality, but
+also based on its functionality.
+
+The methodology follows a systematic procedure that begins with identifying the
+domain and ends with answering the research questions posed above.  In between
+we collect an authoritative list of about 30 software packages.  For each
+package in the list we fill in our measurement template.  The template consists
+of repository related data (like number of open issues, number of lines of code,
+etc.) and 108 measures/questions related to 9 qualities: installability,
+correctness/verifiability, reliability, robustness, usability, maintainability,
+reusability, understandability and visibility/transparency. Filling in the
+template requires installing the software, running simple tests (like completing
+the getting started instructions (if present)), and searching the code,
+documentation and test files.
+
+The data for each domain is used to rank the software package according to each
+quality dimension using AHP.  The ranking is not intended to identify a single
+best software package.  Instead the ranking is intended to provide insights on
+the top set of software for each quality.  The top performers can be contrasted
+with the lesser performers to gain insight into what practices in the domain are
+working.  Deeper insight can be obtained by combining this data with the
+interview data from asking each recruited developer 20 questions.
+
+Combining the quantitative data from the measurement template with the interview
+results, along with the domain experts knowledge, we can determine the current
+state of the practice for domain X.  Using our methodology, spreadsheet
+templates and AHP tool, we estimate (based on our experience with using the
+process) the time to complete an assessment for a given domain at
+\total{totHours} person hours.
 
@@
 other software supporting HTML web page format.
-\begin{figure}[h!]
+
+\begin{figure}[!ht]
 \includegraphics[scale=0.4]{git-stats-demo.JPG}
@@
 
-\begin{figure}[h!]
+\begin{figure}[!ht]
 \includegraphics[scale=0.5]{scc-demo.JPG}
@@
 	What percentage of code is comments? (percentage)\\
-	Which version control system is in use? ({svn, cvs, git, github, unclear,
+	Which version control system is in use? ({svn, cvs, git, Github, unclear,
 	other$^*$}) $^*$ given via string\\
@@
 
-% \def\arraystretch{1.22}
-% \begin{tabular}{p{14cm}}
-% \hline
-% \textbf{Repo Metrics (Measured via git_stats)}\\
-% \hline
-% Number of text-based files. (number)\\
-% Number of binary files. (number)\\
-% Number of total lines in text-based files. (number)\\
-% Number of total lines added to text-based files. (number or {not git repo})\\
-% Number of total lines deleted from text-based files. (number or {not git repo})\\
-% Number of total commits. (number or {not git repo})\\
-% Numbers of commits by year in the last 5 years. (Count from as early as possible if the project is younger than 5 years.) (list of numbers or {not git repo}\\
-% Numbers of commits by month in the last 12 months. (list of numbers or {not git repo})\\
-% \hline
-% \end{tabular}
-
-% \def\arraystretch{1.22}
-% \begin{tabular}{p{14cm}}
-% \hline
-% \textbf{Repo Metrics (Measured via scc)}\\
-% \hline
-% Number of text-based files.	(number)\\
-% Number of total lines in text-based files.	(number)\\
-% Number of code lines in text-based files.	(number)\\
-% Number of comment lines in text-based files.	(number)\\
-% Number of blank lines in text-based files.	(number)
-% \hline
-% \end{tabular}
-
-% \def\arraystretch{1.22}
-% \begin{tabular}{p{14cm}}
-% \hline
-% \textbf{Repo Metrics (Measured via GitHub)}\\
-% \hline
-% Number of stars.	(number or {n/a})
-% Number of forks.	(number or {n/a})
-% Number of people watching this repo.	(number or {n/a})
-% Number of open pull requests.	(number or {n/a})
-% Number of closed pull requests.	(number or {n/a})
-% \hline
-% \end{tabular}
-
 \newpage
@@
 
-Introduction.  Roadmap.  Anything about traceability to research questions.
-
-Information about these interview questions:  This gives you an idea what I
-would like to learn about the development of {domain} software. Interviews will
-be one-to-one and will be open-ended (not just “yes or no” answers). Because of
-this, the exact wording may change a little. Sometimes I will use other short
-questions to make sure I understand what you told me or if I need more
-information when we are talking such as: “So, you are saying that ...?), to get
-more information (“Please tell me more?”), or to learn what you think or feel
-about something (“Why do you think that is...?”).
-
-\subsection{Information about the developers and users} \citep{Jegatheesan2016}
+Interviews will be one-to-one and will be open-ended (not just “yes or no”
+answers). Because of this, the exact wording of the questions may change a
+little. If more information, or clarification, is needed during the
+conversation, follow up questions will be asked, such as: ``So, you are saying
+that ...?'', ``Please tell me more?'', or ``Why do you think that is?''
+
+\subsection{Information about the developers and users}
+
+These questions are based on questions used in \citet{Jegatheesan2016}
+
 \begin{enumerate}
@@
 
-Square brackets are used for traceability to qualities and relevant research
-questions found in Section ...
+Square brackets are used for traceability to the  relevant research
+questions found in Section~\ref{SecResearchQuestions}.
 
@@
 \item How does documentation fit into your development process? Would improved
-documentation help with the obstacles you typically face? [traceability -
-research question 5b, visibility/transparency - research question 5i]
+documentation help with the obstacles you typically face? [research question 5b (traceability),
+research question 5i (visibility/transparency)]
 \item In the past, is there any major obstacle to your development process that
@@
  obstacles, what methods have been considered or practiced to improve the
- situation? If practiced, did it work? [correctness - research question 5e]
+ situation? If practiced, did it work? [research question 5e (correctness)]
 \item When designing the software, did you consider the ease of future changes?
@@
  code blocks? What measures have been taken to ensure the ease of future changes
- and maintains? [maintainability - research question 5d, modifiability -
- research question 5c]
+ and maintains? [research question 5d (maintainability), 
+ research question 5c (modifiability)]
 \item Provide instances where users have misunderstood the software. What, if
-any, actions were taken to address understandability issues? [understandability
-- research question 5f]
-\item What, if any, actions were taken to address usability issues? [usability -
-research question 5a]
+any, actions were taken to address understandability issues? [research question
+5f (understandability)]
+\item What, if any, actions were taken to address usability issues? [
+research question 5a (usability)]
 \item Do you think the current documentation can clearly convey all necessary
 knowledge to the users? If yes, how did you successfully achieve it? If no, what
-improvements are needed? [unambiguity - research question 5g]
+improvements are needed? [research question 5g (unambiguity)]
 \item Do you have any concern that your computational results won't be
 reproducible in the future? Have you taken any steps to ensure reproducibility?
-[reproducibility - research question 5h]
+[research question 5h (reproducibility)]
 \end{enumerate}

### 2021-11-17 StateOfPractice/Papers/MedImageSoft_SOP/MedImageSoft_SOP_AcademicRadiology.tex
Latest edits to the MI State of the Practice paper

@@
 
+\usepackage{hyperref}
+\usepackage{tikz}
+\usetikzlibrary{mindmap}
+\usepackage{multirow}
+\usepackage{pgfplots}
+\usepackage{arydshln}
+
 \journal{Academic Radiology}
+
+\pgfplotsset{compat=1.13}
 
@@
 
-\title{Assessing the State of the Practice for Medical Imaging Software}
+\title{State of the Practice for Medical Imaging Software}
 
@@
 \begin{abstract}
-%% Text of abstract
+
+We present a general method to assess the state of the practice for Scientific
+Computing (SC) software and apply this method to Medical Imaging (MI) software.
+We selected 29 MI software projects from 48 candidates, assessed 10 software
+qualities (installability, correctness/ verifiability, reliability, robustness,
+usability, maintainability, reusability, understandability,
+visibility/transparency and reproducibility) by answering 103 questions for each
+software project, and interviewed 8 of the 29 development teams. Based on the
+quantitative data for the first 9 qualities, we ranked the MI software with the
+Analytic Hierarchy Process (AHP). The top three software products were
+\textit{3D Slicer}, \textit{ImageJ}, and \textit{OHIF Viewer}. By interviewing
+the developers, we identified three major pain points: i) lack of resources; ii)
+difficulty balancing between compatibility, maintainability, performance, and
+security; and, iii) lack of access to real-world datasets for testing. For
+future MI software projects, we propose adopting test-driven development, using
+continuous integration and continuous delivery (CI/CD), using git and GitHub,
+maintaining good documentation, supporting third-party plugins or extensions,
+considering web application solutions, and improving collaboration between
+different MI software projects.
 
@@
 \begin{keyword}
-%% keywords here, in the form: keyword \sep keyword
-
-%% PACS codes here, in the form: \PACS code \sep code
-
-%% MSC codes here, in the form: \MSC code \sep code
-%% or \MSC[2008] code \sep code (2000 is the default)
-
+	medical imaging, scientific computing, software engineering, software
+	quality, Analytic Hierarchy Process, developer interviews
 \end{keyword}
@@
 
-%% main text
-\section{}
-\label{}
+\section{Introduction} \label{ch_intro}
+
+We aim to study the state of software development practice for Medical Imaging
+(MI) software.  MI tools use images of the interior of the body (from sources
+such as Magnetic Resonance Imaging (MRI), Computed Tomography (CT), Positron
+Emission Tomography (PET) and Ultrasound) to provide information for diagnostic,
+analytic, and medical applications \citep{FDA2021, enwiki:1034887445,
+Zhang2008}.  Given the importance of MI software and the high number of
+competing software projects, we wish to understand the merits and drawbacks of
+the current development processes, tools and methodologies.  We aim to assess
+through a software engineering lens the quality of the existing software with
+the hope of highlighting standout examples, understanding current pain points
+and providing guidelines and recommendations for future development.
+
+\subsection{Purpose} \label{sec_motivation}
+
+Not only do we wish to gain insight into the state of the practice for MI
+software, we also wish to understand the development of research software in
+general. We wish to understand the impact of the often cited gap, or chasm,
+between software engineering and scientific programming \citep{Storer2017}.
+Although scientists spend a substantial proportion of their working hours on
+software development \citep{Hannay2009, Prabhu2011}, many developers learn
+software engineering skills by themselves or from their peers, instead of from
+proper training \citep{Hannay2009}. \citet{Hannay2009} observe that many
+scientists showed ignorance and indifference to standard software engineering
+concepts. For instance, according to a survey by \citet{Prabhu2011}, more than
+half of their 114 subjects did not use any proper debugger for their software.
+
+To gain insights, we devised five research questions, which can be applied
+to MI, and to other domains, of research software \citep{SmithEtAl2021}:
+
+\begin{description}
+
+\item[RQ1.] \hypertarget{rq1}What artifacts are present in current software
+projects? What role does documentation play in the projects? What are the
+developers' attitude toward it?
+\item[RQ2.] \hypertarget{rq2}What tools are used in the development of current
+software packages?
+\item[RQ3.] \hypertarget{rq3}What principles, processes, and methodologies are
+used in the development of current software packages?
+\item[RQ4.] \hypertarget{rq4}What are the pain points for developers working on
+research software projects? What aspects of the existing processes,
+methodologies, and tools do they consider as potentially needing improvement?
+What changes to processes, methodologies, and tools can improve software
+development and software quality?
+\item[RQ5.] \hypertarget{rq5}What is the current status of the following
+software qualities for the projects? What actions have the developers taken to
+address them?
+\begin{itemize}
+	\item Installability
+	\item Correctness \& Verifiability
+	\item Reliability
+	\item Robustness
+	\item Usability
+	\item Maintainability
+	\item Reusability
+	\item Understandability
+	\item Visibility/Transparency
+	\item Reproducibility
+\end{itemize}
+\item[RQ6.] \hypertarget{rq6}How does software designated as high quality by
+this methodology compare with top-rated software by the community?
+\end{description}
+
+\subsection{Overview of the Methodology}
+
+We designed a general methodology to assess the state of the practice for SC
+software \citep{SmithEtAl2021}. Details can be found in Section~\ref{}.  This
+methodology builds off prior work to assess the state of the practice for such
+domains as Geographic Information Systems \citep{smith2018state}, Mesh
+Generators \citep{smith2016state}, Seismology software
+\citep{Smith2018Seismology}, and Statistical software for psychology
+\citep{smith2018statistical}.  In keeping with the previous methodology, we have
+maintained the constraint that the work load for measuring a given domain should
+be feasible for a team as small as one person, and for a short time,
+ideally around a person month of effort. A person month is considered to be $20$
+working days ($4$ weeks in a month, with $5$ days of work per week) at $8$
+person hours per day, or $20 \cdot 8 = 160$ person hours.
+
+With our methodology, we first choose an SC domain (in the current case MI) and
+identify a list of about 30 software packages. We approximately measure the
+qualities of each package by filling in a grading template (forward ref?).
+Compared with our previous methodology, the new methodology also includes
+repository based metrics, such as the number of files, number of lines of code,
+percentage of issues that are closed, etc. (forward ref?).  With the
+quantitative data in the grading template, we rank the software with the
+Analytic Hierarchy Process (AHP) (forward ref?). After this, as another addition
+to our previous methodology, we interview some of the development teams to
+further understand the status of their development process (forward ref?).
+Finally, we summarize the results and propose recommendations for future SC
+software development (forward ref?).
+
+\subsection{Scope} \label{sec_scope}
+
+We need to restrict the scope of MI software that we consider so that the
+measurement exercise is feasible within a person month of effort. MI considers a
+broad range of different problems.  According to \citet{Bankman2000}, MI
+software deals with six different basic problems, while \citet{Angenent2006}
+pointed out that four fundamental problems are solved by MI software. While both
+mentioned Segmentation, Registration, and Visualization of medical images,
+Bankman also included Enhancement, Quantification, and three functions for MI
+archiving and telemedicine systems (Compression, Storage, and Communication)
+\citep{Bankman2000}. On the other hand, \citet{Angenent2006} included
+Simulation. According to Wikipedia contributors \citep{enwiki:1034877594}, MI
+software has primary functions in Segmentation, Registration, Visualization
+(including the basic display, reformatted views, and 3D volume rendering),
+Statistical Analysis, and Image-based Physiological Modelling. As
+\citet{Kim2011} describes, the general steps of medical image analysis after
+obtaining digital data include Enhancement, Segmentation, Feature Extraction,
+Classification, and Interpretation. Besides the above major functions, some MI
+software provides supportive functions. For example, with Tool Kit libraries VTK
+\citep{SchroederEtAl2006} and ITK \citep{McCormick2014}, developers build
+software with Visualization and Analysis functions; Picture Archiving and
+Communication System (PACS) helps users to economically store and conveniently
+access images \citep{Choplin1992}.
+
+Based on our literature survey, we divided MI software into five sub-groups and
+several sub-sub-groups by their major functions, as shown in Figure
+\ref{fig_mi_functions}.
+
+\begin{figure}[ht]
+\centering
+\begin{tikzpicture}[mindmap, grow cyclic, every node/.style=concept, concept
+color=orange!40,
+level 1/.append style={level distance=5cm,sibling angle=72},
+level 2/.append style={level distance=3cm,sibling angle=90},
+level 3/.style={level distance=2.3cm,sibling angle=90}]
+\node{MI software}
+child[concept color=teal!40] { node {Enhancement}}
+child[concept color=teal!40] { node {Analysis}
+    child[concept color=blue!30] { node {Registration}}
+    child[concept color=blue!30,] { node {Segmentation}}
+    child[concept color=blue!30] { node {Statistical Analysis}
+    child[concept color=blue!20] { node {Feature Extraction}}
+    child[concept color=blue!20] { node {Classification}}
+    child[concept color=blue!20] { node {Interpretation}}
+    }}
+child[concept color=teal!40] { node {Simulation/\\Modeling}}
+child[concept color=teal!40] { node {Supporting}
+    child[concept color=blue!30] { node {Tool Kit}}
+    child[concept color=blue!30] { node {PACS}}
+}
+child[concept color=teal!40] { node {Visualization}
+    child[concept color=blue!30] { node {2D Display}}
+    child[concept color=blue!30] { node {3D Rendering}}
+    child[concept color=blue!30] { node {Reformatted Views}}
+};
+\end{tikzpicture}
+\caption{Major functions of MI software}
+\label{fig_mi_functions}
+\end{figure}
+
+To keep the data collection and analysis feasible, we limited the scope of the
+software to the software packages providing the Visualization tools and
+functions in this project.
+
+[Roadmap - instead sprinkle forward references throughout the introduction?]
+
+\section{Background} \label{ch_background}
+
+When designing a method for evaluating the state of the practice of
+domain-specific software, we included a step to select domain and software.
+Knowledge of different software categories is essential for the selection. To
+compare and rank the software qualities with the grading template in Appendix
+\ref{ap_grading_template}, we need the definitions of qualities and the AHP.
+
+In this section, we introduce the relevant software categories (Section
+\ref{sec_software_categories}). We also cover the software quality definitions
+(Section \ref{sec_software_quality}) and an overview of the AHP (Section
+\ref{sec_AHP}).
+
+\subsection{Software Categories} \label{sec_software_categories}
+
+We target specific software categories to narrow down the scope when selecting
+software packages for measuring. One way to categorize them is grouping by
+functions, such as analyzing the scope for the MI software (Section
+\ref{sec_scope}), which requires specific domain knowledge. In this section, we
+discuss three common software categories that we can apply to all SC domains:
+Open source software (OSS), freeware, and commercial software.
+
+\subsubsection{Open Source Software} \label{sec_open_source_software}
+
+For OSS, the source code is openly accessible. Users have the right to study,
+change and distribute it under a license granted by the copyright holder. For
+many OSS projects, the development process relies on the collaboration of
+different contributors worldwide \citep{Corbly2014}. Accessible source code
+usually exposes more ``secrets'' of a software project, such as the underlying
+logic of software functions, how developers achieve their works, and the flaws
+and potential risks in the final product. Thus, it brings much more convenience
+to researchers analyzing the qualities of the project.
+
+\subsubsection{Freeware} \label{sec_freeware}
+
+Freeware is software that can be used free of charge. Unlike with OSS, the
+authors of freeware do not allow users to access or modify the source code of
+the software \citep{LINFO2006}. The term \textit{freeware} should not be confused
+with \textit{free software}, which is similar to OSS. To the end-users, the
+differences between freeware and OSS often do not bother them. The fact that
+these products are free of charge is likely to make them popular with many
+users. However, software developers, end-users who wish to modify the source
+code, and researchers looking for insight into software development process will
+find the inaccessible source code a problem.
+
+\subsubsection{Commercial Software}
+
+``Commercial software is software developed by a business as part of its
+business'' \citep{GNU2019}. Typically speaking, the users are required to pay to
+access all of the features of commercial software, excluding access to the
+source code. However, some commercial software is also free of charge
+\citep{GNU2019}. Based on our experience, most commercial software products are
+not OSS.
+
+For some specific software, the backgrounds of commercial software developers
+often differ from the ones of non-commercial OSS. In such a case, the former is
+usually the product of software engineers, and the latter is likely to have
+developers who work in the domain and are also end-users of the products. One
+example of such software is SC software, since the developers need to utilize
+their domain-specific during the development process \citep{Wilson2014}.
+
+\subsection{Software Quality Definitions} \label{sec_software_quality}
+
+This section lists the definitions of 10 software qualities, which are from
+Smith et al. \citep{SmithEtAl2020}. We aim to measure each of them for selected
+SC software packages. The order of the first nine qualities follows our grading
+template in Appendix \ref{ap_grading_template}. We do not measure
+\textit{reproducibility} with the grading template, but discuss it with the
+developers by interviews.
+
+\begin{itemize}
+	\item \textbf{Installability} The effort required for the installation,
+	uninstallation, or reinstallation of a software or product in a specified
+	environment \citep{ISO/IEC25010} \citep{lenhard2013measuring}.
+
+	\item \textbf{Correctness \& Verifiability} A program is correct if it
+	behaves according to its stated specifications. Verifiability is the extent
+	to which a set of tests can be written and executed, to demonstrate that the
+	delivered system meets the specification \citep{GhezziEtAl2003}.
+
+	\item \textbf{Reliability} The probability of failure-free operation of a
+	computer program in a specified environment for a specified time, i.e., the
+	average time interval between two failures also known as the mean time to
+	failure (MTTF) \citep{musa1987software} \citep{GhezziEtAl2003}.
+
+	\item \textbf{Robustness} Software possesses the characteristic of
+	robustness if it behaves ``reasonably'' in two situations: i) when it
+	encounters circumstances not anticipated in the requirements specification,
+	and ii) when the assumptions in its requirements specification are violated
+	\citep{ghezzi1991fundamentals} \citep{boehm2007software}.
+
+	\item \textbf{Usability} ``The extent to which a product can be used by
+	specified users to achieve specified goals with effectiveness, efficiency,
+	and satisfaction in a specified context of use" \citep{ISO/TR16982:2002}
+	\citep{ISO9241-11:2018}.
+
+	\item \textbf{Maintainability} The effort with which a software system or
+	component can be modified to i) correct faults; ii) improve performance or
+	other attributes; iii) satisfy new requirements
+	\citep{IEEEStdGlossarySET1990} \citep{boehm2007software}.
+
+	\item \textbf{Reusability} ``The extent to which a software component can be
+	used with or without adaptation in a problem solution other than the one for
+	which it was originally developed" \citep{kalagiakos2003non}.
+
+	\item \textbf{Understandability} ``The capability of the software product to
+	enable the user to understand whether the software is suitable, and how it
+	can be used for particular tasks and conditions of use" \citep{iso2001iec}.
+
+	\item \textbf{Visibility/Transparency} The extent to which all of the steps
+	of a software development process and the current status of it are conveyed
+	clearly \citep{ghezzi1991fundamentals}.
+
+	\item \textbf{Reproducibility} ``A result is said to be reproducible if
+	another researcher can take the original code and input data, execute it,
+	and re-obtain the same result" \citep{BenureauAndRougier2017}.
+\end{itemize}
+
+\subsection{Analytic Hierarchy Process} \label{sec_AHP}
+
+To generate ranking scores for a set of software packages, we use AHP, which
+utilizes pairwise comparisons between all of the packages. Thomas L. Saaty
+developed this tool, and people widely used it to make and analyze multiple
+criteria decisions \citep{VaidyaEtAl2006}. AHP organizes multiple criteria
+factors in a hierarchical structure and uses pairwise comparisons between
+alternatives to calculate relative ratios \citep{Saaty1990}.
+
+For a project with $ m $ criteria, we can use an $m\times m$ matrix $A$ to
+record the relative importance between factors. When comparing criterion $i$ and
+criterion $j$, the value of $A_{ij}$ is decided as follows, and the value of
+$A_{ji}$ is $1/A_{ij}$ \citep{Saaty1990},
+
+\begin{itemize}
+	\item $A_{ij} = 1$ if criterion $i$ and criterion $j$ are equally important;
+	\item $A_{ij} = 9$ if criterion $i$ is extremely more important than
+	criterion $j$;
+	\item $A_{ij}$ equals to an integer value between 1 and 9 according the the
+	relative importance of criterion $i$ and criterion $j$.
+\end{itemize}
+
+The above process assumes that criterion $i$ is not less important than
+criterion $j$, otherwise, we need to reverse $i$ and $j$ and determine $A_{ji}$
+first, then $A_{ij} = 1/A_{ji}$.
+
+The priority vector $w$ can be calculated by solving the following equation
+\citep{Saaty1990}, 
+\begin{equation} 
+    Aw = \lambda_{max}w,
+\end{equation}
+where $\lambda_{max}$ is the maximal eigenvalue of $A$.
+
+In this project, $w$ is approximated with the classic \textit{mean of normalized
+values} approach \citep{AlessioEtAl2006},
+
+\begin{equation}
+w_i = \frac{1}{m}\sum_{j=1}^{m}\frac{A_{ij}}{\sum_{k=1}^{m}A_{kj}}
+\end{equation}
+
+Suppose there are $n$ alternatives, for criterion $i = 1, 2, ... , m$, we can
+create an $n\times n$ matrix $B_i$ to record the relative preferences between
+these choices. The way of generating $B_i$ is similar to the one for $A$.
+However, unlike comparing the importance between criteria, we pairwise decide
+how much we favor one alternative over the other. We use the same method to
+calculate the local priority vector for each $B_i$.
+
+In this project, the first nine software qualities mentioned in Section
+\ref{sec_software_quality} are the criteria ($m = 9$), while 29 software
+packages ($n = 29$) are compared for each of the $m$ criteria. The software are
+evaluated with the grading template in Appendix \ref{ap_grading_template} and a
+subjective score from one to ten is given for each quality for each package. For
+each quality, for a pair of packages $i$ and $j$, such that $score_i >=
+score_j$, the difference between two scores is $\mathit{diff_{ij}} = score_i -
+score_j$. The relationship between $A_{ij} = 1$ and $\mathit{diff_{ij}}$ is as
+follows:
+
+\begin{itemize}
+\item $A_{ij} = 1$ and $\mathit{diff_{ij}} = 0$ when criterion $i$ and criterion
+$j$ are equally important;
+\item $A_{ij}$ increases when $\mathit{diff_{ij}}$ increases;
+\item $A_{ij} = 9$ and $\mathit{diff_{ij}} = 9$ when criterion $i$ is extremely
+more important than criterion $j$.
+\end{itemize}
+
+Thus, we approximate the pairwise comparison result of $i$ versus $j$ by the
+following equation:
+
+\begin{equation}
+A_{ij} = min(score_i - score_j + 1, 9)
+\end{equation}
+
+\section{Methodology} \label{ch_methods}
+
+We designed a general process for evaluating the state of the practice of
+domain-specific SC software, that we instantiate for a specific scientific
+domain.
+
+Our method involves four steps:
+\begin{enumerate}
+\item choosing a software domain (Section \ref{sec_domain_selection});
+\item collecting and filtering software packages (Section
+\ref{sec_software_selection});
+\item grading the selected software (Section \ref{sec_grading_software});
+\item interviewing development teams for further information (Section
+\ref{sec_interview_methods}).
+\end{enumerate}
+
+\noindent Section \ref{sec_applying_method} presents an example of how we
+applied the method on the MI domain.
+
+\subsection{Domain Selection} \label{sec_domain_selection}
+
+Our methods are generic for any SC software, but they need to be applied to a
+specific domain. When choosing a candidate domain, we prefer one with a large
+number of active OSS. The reason is that we aim to finalize a list of 30
+software packages \citep{SmithEtAl2021} after the screening step. For example, we
+remove the ones without recent updates or specific functions. Besides, we prefer
+OSS projects because our grading method requires access to the code. In
+addition, we prefer a domain with an active community developing and using the
+software. As a result, it is easier to invite enough developers for interviews.
+
+We prefer 30 software packages providing similar functions or falling into
+different sub-groups depending on our research purpose. So the domain needs to
+have enough candidates in one sub-group or enough sub-groups to cross-compare.
+
+We also prefer domains in which our team has expertise. We invite domain experts
+to join and support our projects. They help us in many aspects, such as vetting
+the software list and interview questions.
+
+\subsection{Software Product Selection} \label{sec_software_selection}
+
+The process of selecting software packages contains two steps: i) identify
+software candidates in the chosen domain, ii) filter the list according to our
+needs \citep{SmithEtAl2021}.
+
+\subsubsection{Identify Software Candidates} \label{sec_identify_software_candidates}
+
+We start with finding candidate software in publications of the domain. Then, we
+search various websites, such as \hyperlink{https://github.com/}{GitHub},
+\hyperlink{https://swmath.org/}{swMATH} and the Google search results for
+software recommendation articles. We should also include the ones suggested by
+the domain experts \citep{SmithEtAl2021}.
+
+\subsubsection{Filter the Software List} \label{sec_filter_software_list}
+
+The goal is to build a software list with a length of about 30
+\citep{SmithEtAl2021}.
+
+The only ``mandatory'' requirement is that the software must be OSS, as defined
+in Section \ref{sec_open_source_software}. We need this because evaluating some
+software qualities requires the source code.
+
+The other filters are optional, and we consider them according to the number of
+software candidates and the objectives of the research project. We apply them in
+the following priority order:
+
+\begin{enumerate}
+\item The functions and purpose of the software. An SC domain often contains
+software with various functions and purposes. For example, some MI software
+packages are Tool Kit for developers to use, and some others offer Visualization
+function to end-users. We have two options:
+\begin{itemize}
+\item selecting a set of software with the same major function. In this case, we
+can use the identical process to assess all packages, e.g., same input to
+measure \textit{Robustness}. Also, when we give impression scores to qualities
+such as \textit{Installability} and \textit{Usability}, the results are more
+comparable. Thus, it is more feasible to collect the results and rank the
+software.
+\item selecting software from a set of different sub-groups. For example, we can
+choose 10 MI software from each of the three sub-groups: Visualization, Tool
+Kit, and PACS. The downside: we may need different processes to measure each
+sub-group; it may be less accurate to mix all three sub-groups and rank them
+together. The benefit: we can cross-compare the development processes between
+the sub-groups.
+\end{itemize}
+
+\item The version control tool. The empirical measurement tools listed in
+Section \ref{sec_empirical_measurements} only work on projects using
+\hyperlink{https://git-scm.com/}{Git}, so we prefer software with Git. Some
+manual steps in empirical measurement depend on a few metrics of GitHub, which
+makes projects held on GitHub more favored \citep{SmithEtAl2021}.
+
+\item The age of the software. Some of the OSS projects may experience a lack of
+recent maintenance. So we eliminate packages without recent updates, unless they
+are still popular and highly recommended by the domain users
+\citep{SmithEtAl2021}. We consider a software project as ``alive" if it has any
+update within the last 18 months; otherwise, we mark it as ``dead".
+\end{enumerate}
+
+The order of filters 2 and 3 is flexible. We adjust it according to the number
+of software packages affected by the filters, and the number of ones remaining
+on the list.
+
+\subsubsection{Vet the Software List} \label{sec_vet_software_list}
+
+Before showing our filtered list to the domain experts, we ask them to list
+their top 10 software in the domain. Then, we cross-compare the two lists and
+discuss the commonality and variability.  In addition, we ask them to vet our
+filtered list. They provide views on whether the list is reasonable. We also use
+their opinions for a filtering process. For example, if a software package is
+not OSS and has had no updates for a long time, but the domain experts identify
+it as a valuable product, we still consider it in our final list.
+
+\subsection{Grading Software} \label{sec_grading_software}
+
+We grade the selected software using a template (Section
+\ref{sec_grading_template}) and a specific empirical method (Section
+\ref{sec_empirical_measurements}). Some technical details for the measurements
+are in Section \ref{sec_technical_details}.
+
+\subsubsection{Grading Template} \label{sec_grading_template}
+
+The full grading template can be found in Appendix \ref{ap_grading_template}.
+The template contains 103 questions that we use for grading software products.
+Figure \ref{fg_grading_template_example} shows an example of this grading
+template.
+
+\begin{figure}[ht]
+\includegraphics[scale=0.42]{figures/template.png}
+\caption{Grading template example}
+\label{fg_grading_template_example}
+\end{figure}
+
+We use the first section of the template to collect general information, such as
+the name, purpose, platform, programming language, publications about the
+software, the first release and the most recent change date, website, source
+code repository of the product, etc. Information in this section helps us
+understand the projects better and may be helpful for further analysis, but it
+does not directly affect the grading scores.
+
+We designed the following nine sections in the template for the nine software
+qualities mentioned in Section~\ref{sec_software_quality}. For each quality, we
+ask several questions and the typical answers are among the collection of
+``yes'', ``no'', ``n/a'', ``unclear'', a number, a string, a date, a set of
+strings, etc. Each quality needs an overall score between 1 and 10 based on all
+the previous questions. For some qualities, we perform surface measurements,
+which allow us to measure all packages with reasonable efforts. The surface
+measurements reveal some traits of a underlying quality, but may not fully
+represent it.
+
+\begin{itemize}
+
+\item \textbf{Installability} We check the existence and quality of installation
+instructions. The user experience is also an important factor, such as the ease
+to follow the instructions, number of steps, automation tools, and the
+prerequisite steps for the installation. If any problem interrupts the process
+of installation or uninstallation, we give a lower score to this quality. We
+also record the Operating System (OS) for the installation test and whether we
+can verify the installation.
+
+\item \textbf{Correctness \& Verifiability} For \textit{correctness}, we check
+the projects to identify the techniques to ensure this quality, such as literate
+programming, automated testing, symbolic execution, model checking, unit tests,
+etc. We also examine whether the projects use continuous integration and
+continuous delivery (CI/CD). For \textit{verifiability}, we go through the
+documents of the projects to check the requirements specifications, theory
+manuals, and getting started tutorials. If a getting started tutorial exists and
+provides expected results, we follow it and check if the outputs match.
+
+\item \textbf{Surface Reliability} We check whether the software breaks during
+the installation and the operations following a getting started tutorials,
+whether there are descriptive error messages, and if we can recover the process
+after an error. We use the software to load damaged images during the assessment
+to \textit{surface robustness}, and lower its score for \textit{surface
+reliability} if a software product ``break" in this process.
+
+\item \textbf{Surface Robustness} We check how the software handles
+unexpected/unanticipated input. For example, we prepare broken image files for
+MI software packages with the function to load image files. We use a text file
+(.txt) with a modified extension name (.dcm) as an unexpected/unanticipated
+input. We load a few correct input files to ensure the function is working
+correctly before testing with the unexpected/unanticipated ones.
+
+\item \textbf{Surface Usability} We examine the documentation of the projects,
+and we consider software with a getting started tutorial and a user manual
+easier to use. Meanwhile, we check if users have any channels to get support. We
+also record our impressions and user experiences when testing the software.
+Easy-to-use graphical user interfaces give us a better experience, which leads
+to better scores.
+
+\item \textbf{Maintainability} We search the projects' documents and identify
+the process of contributing and reviewing code. We believe that the artifacts of
+a project - including source code, documents, building scripts, etc. - can
+significantly affect its  \textit{maintainability}. Thus we check each project
+for its artifacts, such as API documentation, bug tracker, release notes, test
+cases, build files, version control, etc. We also check the tools supporting
+issue tracking and version control, the percentages of closed issues, and the
+proportion of comment lines in code.
+
+\item \textbf{Reusability} We count the total number of code files for each
+project. Projects with a large number of components provide more choices to
+reuse. Furthermore, well-modularized code, which tends to have smaller parts in
+separated files, is typically easier to reuse. Thus, we consider the projects
+with more code files and less code lines per file to be more reusable. We use a
+command-line tool \textit{scc} to count the number of  text-based files and LOC
+for all projects. We also decide that the projects with API documentation can
+deliver better \textit{Reusability}.
+
+\item \textbf{Surface Understandability} We randomly examine 10 code files. We
+check the code’s style within each file, such as whether the identifiers,
+parameters, indentation, and formatting are consistent, whether the constants
+(other than 0 and 1) are hardcoded, and whether the developers modularized the
+code. We also check the descriptive information for the code, such as documents
+mentioning the coding standard, the comments in the code, and the descriptions
+or links for algorithms in the code.
+
+\item \textbf{Visibility/Transparency} To measure this quality, we check the
+existing documents to find out whether the software development process and
+current status of a project are visible and transparent. We examine the
+development process, current status, development environment, and release notes
+for each project. If any information is missing or poorly conveyed, the
+\textit{visibility/transparency} will be lower.
+\end{itemize}
+
+For some qualities, the empirical measurements also affect the score. We use
+tools to extract information from the source code repositories. For projects
+held on GitHub, we manually collect additional metrics, such as the stars of the
+GitHub repository, and the numbers of open and closed pull requests. Section
+\ref{sec_empirical_measurements} presents more details about the empirical
+measurements.
+
+\subsubsection{Empirical Measurements} \label{sec_empirical_measurements}
+
+We use two command-line tools for the empirical measurements. One is
+\textit{GitStats} that generates statistics for git repositories and displays
+outputs in the form of web pages \citep{Gieniusz2019}; the other one is Sloc Cloc
+and Code (as known as \textit{scc}) \citep{Boyter2021}, aiming to count the lines
+of code, comments, etc.
+
+Both tools measure the number of text-based files in a git repository and lines
+of text in these files. Based on our experience, most text-based files in a
+repository contain programming source code, and developers use them to compile
+and build software products. A minority of these files are instructions and
+other documents. So we roughly regard the lines of text in text-based files as
+lines of programming code. The two tools usually generate similar but not
+identical results. From our understanding, this minor difference is due to the
+different techniques to detect if a file is text-based or binary.
+
+Additionally, we manually collect information for projects held on GitHub, such
+as the numbers of stars, forks, people watching this repository, open pull
+requests, closed pull requests, and the number of months a repository has been
+on GitHub. A git repository can have a creation date much earlier than the first
+day on GitHub. For example, the developers created the git repository of
+\textit{3D Slicer} in 2002, but did not upload a copy of it to GitHub until
+2020. We get the creation date of the GitHub copy by using API
+\textit{https://api.github.com/repos/{:owner}/{:repository}} (e.g.,
+\hyperlink{https://api.github.com/repos/slicer/slicer}{https://api.github.com/repos/slicer/slicer}).
+In the response, the value of ``created\_at" is what we want. The number of
+months a repository has been on GitHub helps us understand the average change of
+metrics over time, e.g., the average new stars per month.
+
+These empirical measurements help us from two aspects. Firstly, they help us
+with getting a project overview faster and more accurately. For example, the
+number of commits over the last 12 months shows how active this project has
+been, and the number of stars and forks may reveal its popularity. Secondly, the
+results may affect our decisions regarding the grading scores for some software
+qualities. For example, if the percentage of comment lines is low, we
+double-check the \textit{understandability} of the code; if the ratio of open
+versus closed pull requests is high, we pay more attention to the
+\textit{maintainability}.
+
+\subsubsection{Technical Details} \label{sec_technical_details}
+
+To test the software on a ``clean'' system, we create a new virtual machine (VM)
+for each software and only install the necessary dependencies before measuring.
+We make all 30 VMs on the same computer, one at a time, and destroy them after
+measuring.
+
+We spend about two hours grading each package, unless we find technical issues
+and need more time to resolve them. In most of the situation, we finish all the
+measurements for one software on the same day.
+
+\subsection{Interview Methods} \label{sec_interview_methods}
+
+This section introduces our interview questions (Section
+\ref{sec_interview_questions}), method of selecting interviewees (Section
+\ref{sec_interviewee_selection}), and interview process (Section
+\ref{sec_interview_process}).
+
+\subsubsection{Interview Questions} \label{sec_interview_questions}
+
+We designed a list of 20 questions to guide our interviews, which can be found
+in Section \ref{ch_interview} and Appendix \ref{ap_interview}.
+
+Some questions are about the background of the software, the development teams,
+the interviewees, and how they organize the projects. We also ask about their
+understandings of the users. Some questions focus on the current and past
+difficulties, and the solutions the team has found or will try. We also discuss
+the importance and current situations of documentation. A few questions are
+about specific software qualities, such as \textit{maintainability},
+\textit{understandability}, \textit{usability}, and \textit{reproducibility}.
+
+The interviews are semi-structured based on the question list; we ask follow-up
+questions when necessary. Based on our experience, the interviewees usually
+bring up some exciting ideas that we did not expect, and it is worth expanding
+on these topics.
+
+\subsubsection{Interviewee Selection} \label{sec_interviewee_selection}
+
+For a software list with a length of roughly 30, we aim to interview about ten
+development teams. Interviewing multiple individuals from each team gives us
+more comprehensive information, but a single engineer well-knowing the project
+is also sufficient.
+
+Ideally, we select projects after the grading measurements and prefer the ones
+with higher overall scores. However, if we do not find enough participants, we
+reach all teams on the list. As mentioned in Section
+\ref{sec_apply_to_mi_interviews}, when we applied this process to the MI domain,
+we eventually contacted all teams.
+
+We try to find the contacts of the teams on the projects' websites, such as the
+official web pages, repositories, publications, and bio pages of the teams'
+institutions. Then, we send at most two emails to one contact asking for its
+participation before receiving any replies. We operate the invitation according
+to our ethics approval, such as the one in Appendix \ref{ap_ethics}. For
+example, we ask for participants' consent before interviewing them, recording
+the conversation, or including it in our report.
+
+\subsubsection{Interview Process} \label{sec_interview_process}
+
+Before contacting any interviewee candidate, we need to receive ethics clearance
+from the McMaster University Research Ethics Board. Since the members of the
+development teams are usually around the world, we organize these interviews as
+virtual meetings online with \hyperlink{https://zoom.us/}{Zoom}. After receiving
+consent from the interviewees, we also record and transcribe our discussions.
+
+\subsection{Applying the Method to MI} \label{sec_applying_method}
+
+This section shows an overview of applying our method to the MI domain.
+
+\subsubsection{Domain Selection}
+
+Based on the principles in Section \ref{sec_domain_selection}, we selected the
+MI domain and the sub-group of software with the Visualization function shown in
+Figure \ref{fig_mi_functions}. We also included Dr. Michael Noseworthy, a
+professor of Electrical and Computer Engineering at McMaster University,
+Co-Director of the McMaster School of Biomedical Engineering, and Director of
+Medical Imaging Physics and Engineering at St. Joseph’s Healthcare, and some of
+his students as the MI domain experts in our team.
+
+\subsubsection{Software Product Selection} \label{sec_mi_software_selection}
+
+By using the method in Section \ref{sec_identify_software_candidates}, we
+identified 48 MI software projects as the candidates from publications
+\citep{Bjorn2017} \citep{Bruhschwein2019} \citep{Haak2015}, online articles related
+to the domain \citep{Emms2019} \citep{Hasan2020} \citep{Mu2019}, forum discussions
+related to the domain \citep{Samala2014}, etc. Appendix
+\ref{ap_list_before_filtering} shows all 48 software packages.
+
+Guided by the method in Section \ref{sec_filter_software_list}, we filtered the
+list with a process as follows:
+
+\begin{enumerate}
+
+\item Among them, there were eight that we could not find their source code,
+such as \textit{MicroDicom}, \textit{Aliza}, and \textit{jivex}. These packages
+are likely to be freeware defined in Section \ref{sec_freeware} and not OSS. So
+following guidelines in Section \ref{sec_filter_software_list} we removed them
+from the list.
+
+\item Next, we focused on the MI software providing Visualization functions, as
+described in Section \ref{sec_scope}. Seven of the software on the list were
+Tool Kits or libraries for other software to use as dependencies, but not for
+end-users to view medical images, such as \textit{VTK}, \textit{ITK}, and
+\textit{dcm4che}; another three were PACS. We also eliminated these from the
+list.
+
+\item Finally, we removed \textit{Open Dicom Viewer} from the list because it
+had not received any updates for a long time (since 2011). After that, only
+\textit{MatrixUser} and \textit{AMIDE} were considered as ``dead". However, both
+of them had much more recent updates (after 2017) than \textit{Open Dicom
+Viewer}.
+
+\end{enumerate}
+
+We still preferred projects using git and GitHub and being updated recently, but
+did not apply this filter since packages were already below 30. Even without
+this filter, 27 out of the 29 software packages on the filtered list used git,
+and 24 chose GitHub.
+
+Following the process in Section \ref{sec_vet_software_list}, our domain experts
+provided a list of top software that contains 12 software products (Table
+\ref{tab_top_software_experts}). We compared two lists and found six common
+ones.
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{ll}
+\hline
+Software & On both lists \\ \hline
+3D Slicer & X \\
+Horos & X \\
+ImageJ & X \\
+Fiji & X \\
+AFNI &  \\
+FSL &  \\
+Freesurfer &  \\
+Mricron & X \\
+Mango & X \\
+Tarquin &  \\
+Diffusion Toolkit &  \\
+MRItrix &  \\ \hline
+\end{tabular}
+\caption{\label{tab_top_software_experts}Top software by the MI domain experts}
+\end{table}
+
+We included \textit{Mango} in the initial list, but removed it because it was
+not OSS. However, we kept \textit{Papaya}, a the web version of \textit{Mango}.
+Instead of \textit{MRIcron}, we chose \textit{MRIcroGL}, because
+\textit{MRIcron} development had moved to \textit{MRIcroGL} \citep{Rorden2021b}.
+
+Six software packages on the domain experts' list were not on our filtered list.
+We believed their primary function was Analysis mentioned in Section
+\ref{sec_scope}. Thus, we did not include them in our final list.
+
+After vetting our filtered list, the domain experts believed it was reasonable
+and did not identify any problem. Thus, as shown in Appendix
+\ref{ap_list_before_filtering}, eventually, we had 29 software products on the
+final list.
+
+\subsubsection{Grading Software}
+\label{sec_applying_method_grading}
+Then we followed the steps in Section \ref{sec_grading_software} to measure and grade the software.
+27 out of the 29 packages are compatible with two or three different OS such as Windows, macOS, and Linux, and 5 of them are browser-based, making them platform-independent.
+However, in the interest of time, we only performed the measurements for each project by installing it on one of the platforms, most likely Windows.
+
+\subsubsection{Interviews}
+\label{sec_apply_to_mi_interviews}
+We received ethics clearance from the McMaster University Research Ethics Board
+(Appendix \ref{ap_ethics}). Going through the interview process in Section
+\ref{sec_interview_methods}, we contacted all of the 29 teams. Members from
+eight teams responded and agreed to participate. As a result, we interviewed
+nine developers and architects from the eight teams.
+
+\section{Measurement Results} \label{ch_results}
+
+As discussed in Section~\ref{sec_grading_software}, we use a grading template
+and a empirical method to measure the selected software. We applied this step to
+the MI domain (Section~\ref{sec_applying_method_grading}). This section shows
+the summary of the measurement results. The detailed data can be found in the
+repository
+\hyperlink{https://data.mendeley.com/datasets/k3pcdvdzj2/1}{https://data.mendeley.com/datasets/k3pcdvdzj2/1}.
+This section contains part of the answers to \hyperlink{rq5}{RQ5}.
+
+Table \ref{tab_final_list} shows the 29 software packages that we measured,
+along with some summary data collected in the year 2020. As mentioned in Section
+\ref{sec_grading_template}, we used \textit{scc} (Section
+\ref{sec_empirical_measurements}) to count the Lines of Code (LOC), excluding
+the comment and blank lines. We arrange the items in the descending order of the
+LOC. We found the initial release dates (Rlsd) in their documents for most
+projects and marked the two unknown dates with ``?". We used the date of the
+latest change to each code repository to decide the latest update. We found out
+funding information (Fnd) for only eight projects.
+
+We counted the number of contributors (NOC). We considered anyone who made at
+least one accepted commit to the source code as a contributor. Thus, the NOC is
+not usually the same as the number of long-term members. Many of these projects
+received change requests and code from the community, such as pull requests and
+git commits on GitHub.
+
+Table \ref{tab_final_list} also shows the supported OS for each software
+package. Twenty-five of them could work on all three OSs: Windows (W), macOS
+(M), and Linux (L). However, there was a significant difference in the
+philosophy to achieve cross-platform compatibility. Most of them were native
+software products, but five were naturally platform-independent web
+applications, as shown in column ``Web".
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{llllllllll}
+\hline
+\multirow{2}{*}{Software} & \multirow{2}{*}{Rlsd} & \multirow{2}{*}{Updated} & \multirow{2}{*}{Fnd} & \multirow{2}{*}{NOC} & \multirow{2}{*}{LOC} & \multicolumn{3}{c}{OS} & \multirow{2}{*}{Web} \\ \cline{7-9}
+ &  &  &  &  &  & W & M & L &  \\ \hline
+ParaView \citep{Ahrens2005} & 2002 & 2020-10 & X & 100 & 886326 & X & X & X & X \\
+Gwyddion \citep{Nevcas2012} & 2004 & 2020-11 &  & 38 & 643427 & X & X & X &  \\
+Horos \citep{horosproject2020} & ? & 2020-04 &  & 21 & 561617 &  & X &  &  \\
+OsiriX Lite \citep{PixmeoSARL2019} & 2004 & 2019-11 &  & 9 & 544304 &  & X &  &  \\
+3D Slicer \citep{Kikinis2014} & 1998 & 2020-08 & X & 100 & 501451 & X & X & X &  \\
+Drishti \citep{Limaye2012} & 2012 & 2020-08 &  & 1 & 268168 & X & X & X &  \\
+Ginkgo CADx \citep{Wollny2020} & 2010 & 2019-05 &  & 3 & 257144 & X & X & X &  \\
+GATE \citep{Jan2004} & 2011 & 2020-10 &  & 45 & 207122 &  & X & X &  \\
+3DimViewer \citep{TESCAN2020} & ? & 2020-03 & X & 3 & 178065 & X & X &  &  \\
+medInria \citep{Fillard2012} & 2009 & 2020-11 &  & 21 & 148924 & X & X & X &  \\
+BioImage Suite Web \citep{Papademetris2005} & 2018 & 2020-10 & X & 13 & 139699 & X & X & X & X \\
+Weasis \citep{Roduit2021} & 2010 & 2020-08 &  & 8 & 123272 & X & X & X &  \\
+AMIDE \citep{Loening2017} & 2006 & 2017-01 &  & 4 & 102827 & X & X & X &  \\
+XMedCon \citep{Nolf2003} & 2000 & 2020-08 &  & 2 & 96767 & X & X & X &  \\
+ITK-SNAP \citep{Yushkevich2006} & 2006 & 2020-06 & X & 13 & 88530 & X & X & X &  \\
+Papaya \citep{UTHSCSA2019} & 2012 & 2019-05 &  & 9 & 71831 & X & X & X &  \\
+OHIF Viewer \citep{Ziegler2020} & 2015 & 2020-10 &  & 76 & 63951 & X & X & X & X \\
+SMILI \citep{Chandra2018} & 2014 & 2020-06 &  & 9 & 62626 & X & X & X &  \\
+INVESALIUS 3 \citep{Amorim2015} & 2009 & 2020-09 &  & 10 & 48605 & X & X & X &  \\
+dwv \citep{Martelli2021} & 2012 & 2020-09 &  & 22 & 47815 & X & X & X & X \\
+DICOM Viewer \citep{Afsar2021} & 2018 & 2020-04 & X & 5 & 30761 & X & X & X &  \\
+MicroView \citep{ParallaxInnovations2020} & 2015 & 2020-08 &  & 2 & 27470 & X & X & X &  \\
+MatrixUser \citep{Liu2016} & 2013 & 2018-07 &  & 1 & 23121 & X & X & X &  \\
+Slice:Drop \citep{Haehn2013} & 2012 & 2020-04 &  & 3 & 19020 & X & X & X & X \\
+dicompyler \citep{Panchal2010} & 2009 & 2020-01 &  & 2 & 15941 & X & X &  &  \\
+Fiji \citep{Schindelin2012} & 2011 & 2020-08 & X & 55 & 10833 & X & X & X &  \\
+ImageJ \citep{Rueden2017} & 1997 & 2020-08 & X & 18 & 9681 & X & X & X &  \\
+MRIcroGL \citep{Rorden2021} & 2015 & 2020-08 &  & 2 & 8493 & X & X & X &  \\
+DicomBrowser \citep{Archie2012} & 2012 & 2020-08 &  & 3 & 5505 & X & X & X &  \\ \hline
+\end{tabular}
+\caption{\label{tab_final_list}Final software list}
+\end{table}
+
+Most of the projects used more than one programming language, including a
+primary language that the developers used the most. Figure \ref{fig_language}
+shows the primary languages versus the number of projects using them.
+
+% temporarily comment out
+
+% \begin{figure}[ht]
+% \centering
+% \begin{tikzpicture}
+% \centering
+% \begin{axis}
+% [
+% ybar,
+% height=8cm,
+% width=12cm,
+% ylabel={Number of projects},
+% xlabel={\ Primary language},
+% symbolic x coords={C++, JavaScript, Java, C, Python, Pascal, Matlab},
+% xtick=data,
+% nodes near coords,
+% nodes near coords align={vertical},
+% ]
+% \addplot[black,fill=blue!55!black] coordinates {(C++,11) (JavaScript,6) (Java,4) (C,3) (Python,3) (Pascal,1) (Matlab,1) };
+% \end{axis}  
+% \end{tikzpicture}
+% \caption{\label{fig_language}Primary languages versus number of projects using them}
+% \end{figure}
+
+We failed to install \textit{DICOM Viewer}, so we could not test its
+\textit{surface reliability} and \textit{surface robustness}. We kept this
+software on our list because the other seven qualities do not rely on a
+successful installation. Besides, the \textit{DICOM Viewer} team built it as a
+plugin software for NextCloud
+(\hyperlink{https://apps.nextcloud.com/}{https://apps.nextcloud.com/}) platform,
+which was a unique choice we had not seen before. We wanted to keep it to enrich
+the diversity.
+
+\subsection{Installability} \label{sec_result_installability}
+
+Figure \ref{fg_installability_scores} lists the scores of
+\textit{installability}.
+
+\begin{figure}[ht]
+\includegraphics[scale=0.38]{figures/installability_scores.png}
+\caption{AHP installability scores}
+\label{fg_installability_scores}
+\end{figure}
+
+We found installation instructions for 16 projects. Among the ones without
+instructions, \textit{BioImage Suite Web} and \textit{Slice:Drop} are web
+applications with online versions to use, thus they do not need installation.
+Installing 10 of the projects required extra dependencies. Five of them are the
+web applications in Table \ref{tab_final_list}, and depended on a browser;
+\textit{dwv}, \textit{OHIF Viewer}, and \textit{GATE} needed extra dependencies
+to build; \textit{ImageJ} and	\textit{Fiji} needed an unzip tool;
+\textit{MatrixUser} was based on Matlab; \textit{DICOM Viewer} needed to work on
+a Nextcloud platform.
+
+\textit{3D Slicer} has the highest score because it had easy to follow
+installation instructions, and the installation processes were automated, fast,
+and frustration-free, with all dependencies automatically added. There were also
+no errors during the installation and uninstallation steps. Many other software
+packages also had installation instructions and automated installers, and we had
+no trouble installing them, such as \textit{INVESALIUS 3}, \textit{Gwyddion},
+\textit{XMedCon}, and \textit{MicroView}. We gave them various scores based on
+the understandability of the instructions, installation steps, and user
+experience. Since \textit{BioImage Suite Web} and \textit{Slice:Drop} needed no
+installation, we gave them higher scores. \textit{BioImage Suite Web} also
+provided an option to download cache to local for offline usage, which was easy
+to apply.
+
+\textit{dwv}, \textit{GATE}, and \textit{DICOM Viewer} showed severe problems.
+We could not install them. We spent a reasonable amount of time on these
+problems, then considered them as major obstacles for normal users if we could
+not figure out any solutions. We suspect that only a part of the users faced the
+same problems, and given a lot of time, we might be able to find solutions.
+However, the difficulties greatly impacted the installation experiences, and we
+graded these software packages with lower scores. For example, \textit{dwv} and
+\textit{GATE} had the option to build from the source code, and we failed the
+building processes following the instructions. Although we could not locally
+build them, we could use a deployed online version for \textit{dwv}, and a VM
+version for \textit{GATE}. With those, we finished all the measurements for
+them. Furthermore, \textit{DICOM Viewer} depended on the NextCloud platform, and
+we could not successfully install the dependency.
+
+\textit{MatrixUser} has a lower score because it depended on Matlab. We
+considered installing Matlab takes many more steps and time, and some users may
+not have a license to use Matlab.
+
+\subsection{Correctness \& Verifiability} \label{sec_result_correctness_verifiability}
+
+The scores of \textit{correctness \& verifiability} are shown in Figure
+\ref{fg_correctness_erifiability_scores}. Generally speaking, the packages with
+higher scores adopted more techniques to improve \textit{correctness}, and had
+better documents for us to verify it.
+
+\begin{figure}[ht]
+\includegraphics[scale=0.38]{figures/correctness_verifiability_scores.png}
+\caption{AHP correctness \& verifiability scores}
+\label{fg_correctness_erifiability_scores}
+\end{figure}
+
+After examining the source code, we could not find any evidence of unit testing
+in more than half of the projects. Unit testing benefits most parts of the
+software's life cycle, such as designing, coding, debugging, and optimization
+\citep{Hamill2004}. It can reveal the bugs at an earlier stage of the development
+process, and the absence of unit testing may cause problems for
+\textit{correctness \& verifiability}.
+
+We could not find requirements specifications for most projects. The only
+document we found is a road map of \textit{3D Slicer}, which contained design
+requirements for the upcoming changes. However, it did not record the conditions
+for previous versions. We also could not identify the theory manuals for all of
+the projects. Even for some projects with well-organized documents, requirements
+specifications and theory manuals were still missing.
+
+We identified five projects using CI/CD tools, which are \textit{3D Slicer},
+\textit{ImageJ}, \textit{Fiji}, \textit{dwv}, and \textit{OHIF Viewer}.
+
+In this section, the information about CI/CD tools is part of the answers to
+\hyperlink{rq2}{RQ2}, and the information about software testing and
+documentation is part of the answers to \hyperlink{rq3}{RQ3}.
+
+\subsection{Surface Reliability} \label{sec_result_reliability}
+
+As described in Section \ref{sec_result_installability}, we could not build
+\textit{dwv} and \textit{GATE}. However, since there was an online or VM version
+of them, successful deployment is possible. So the failure of installation did
+not affect their scores in \textit{surface reliability}. Figure
+\ref{fg_reliability_scores} shows the AHP results.
+
+\begin{figure}[ht]
+\includegraphics[scale=0.38]{figures/reliability_scores.png}
+\caption{AHP surface reliability scores}
+\label{fg_reliability_scores}
+\end{figure}
+
+As shown in Section \ref{sec_result_installability}, most of the software
+products did not ``break" during installation or did not need installation;
+\textit{dwv} and \textit{GATE} broke in the building stage, and the processes
+were not recoverable; we could not install the dependency for \textit{DICOM
+Viewer}.
+
+Of the seven software packages with a getting started tutorial and operation
+steps in the tutorial, most showed no error when we followed the steps. However,
+\textit{GATE} could not open macro files and became unresponsive several times,
+without any descriptive error message. When assessing \textit{surface
+robustness} (Section \ref{sec_result_robustness}), we found out that
+\textit{Drishti} crashed during loading damaged image files and did not show any
+descriptive error message. On the other hand, we did not see any problems with
+the online version of \textit{dwv}.
+
+\subsection{Surface Robustness} \label{sec_result_robustness}
+
+Figure \ref{fg_robustness_scores} presents the scores for \textit{surface
+robustness}.
+
+\begin{figure}[ht]
+\includegraphics[scale=0.38]{figures/robustness_scores.png}
+\caption{AHP surface robustness scores}
+\label{fg_robustness_scores}
+\end{figure}
+
+The packages with higher scores elegantly handled the unexpected/unanticipated
+inputs, typically showing a clear error message. We might underestimate the
+score of \textit{OHIF Viewer} since we needed further customization to load
+data, and the test was not complete.
+
+Digital Imaging and Communications in Medicine (DICOM) is a widely used MI
+standard, and ``it defines the formats for medical images that can be exchanged
+with the data and quality necessary for clinical use" \citep{MITA2021}. According
+to their documentation, all 29 software packages should support the DICOM
+standard. We prepared two types of image files: the ones in correct formats and
+the broken ones. We used two MI sample files in the DICOM format as the image
+files with valid formats; we created a standard text file, changed its extension
+name from ``.txt" to ``.dcm", and used it as the unexpected/unanticipated input.
+
+Being tested with the input files with correct formats, all software packages
+except \textit{GATE} loaded the images correctly. \textit{GATE} failed this test
+for unknown reasons.
+
+With the unexpected/unanticipated input, \textit{MatrixUser}, \textit{dwv}, and
+\textit{Slice:Drop} ignored the incorrect format of the file and loaded it
+regardless. They did not show any error message and displayed a blank image.
+\textit{MRIcroGL} behaved similarly except that it showed a meaningless image
+with noise pixels. \textit{Drishti} successfully detected the broken format of
+the file, but the software crashed as a result. We recorded \textit{Drishti}'s
+issue to the measurement of its \textit{reliability} in Section
+\ref{sec_result_installability}.
+
+\subsection{Surface Usability} \label{sec_result_usability}
+
+Figure \ref{fg_usability_scores} lists the AHP scores for \textit{surface
+usability}.
+
+\begin{figure}[ht]
+\includegraphics[scale=0.38]{figures/usability_scores.png}
+\caption{AHP surface usability scores}
+\label{fg_usability_scores}
+\end{figure}
+
+We found a getting started tutorial for only 11 projects but a user manual for
+22 projects. \textit{MRIcroGL} was the only one with documentation for the
+expected user characteristics.
+
+The software with higher scores usually provided both comprehensive document
+guidance and a good user experience. \textit{INVESALIUS 3} set an excellent
+example of a detailed and precise user manual. \textit{GATE} also provided a
+large number of documents, but we think that they conveyed the ideas poorly, as
+we had trouble understanding and using them.
+
+Table \ref{tab_user_support_model} shows the user support models by the number
+of projects using them. This table contains part of the answers to
+\hyperlink{rq2}{RQ2}. Maybe not every team intended to use GitHub issues to
+answer users' questions, but many users use them to seek help.
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{ll}
+\hline
+\multicolumn{1}{c}{User support model} & Number of projects \\ \hline
+GitHub issue & 24 \\
+Frequently Asked Questions (FAQ) & 12 \\
+Forum & 10 \\
+E-mail address & 9 \\
+GitLab issue, SourceForge discussions & 2 \\
+Troubleshooting & 2 \\
+Contact form & 1 \\ \hline
+\end{tabular}
+\caption{\label{tab_user_support_model}User support models by number of projects}
+\end{table}
+
+\subsection{Maintainability} \label{sec_score_maintainability}
+
+Figure \ref{fg_maintainability_scores} shows the AHP results for
+\textit{maintainability}.
+
+\begin{figure}[ht]
+\includegraphics[scale=0.38]{figures/maintainability_scores.png}
+\caption{AHP maintainability scores}
+\label{fg_maintainability_scores}
+\end{figure}
+
+We marked \textit{3D Slicer} with a much higher score than others because it did
+very well at closing the identified issues, and more importantly, we found it to
+have the most comprehensive artifacts. For example, as far as we could find out,
+only a few of the 29 projects had a project plan, developer's manual, or API
+documentation, and only \textit{3D Slicer}, \textit{ImageJ}, \textit{Fiji}
+included all three documents. Meanwhile, \textit{3D Slicer} has a much higher
+percentage of closed issues (91.65\%) than \textit{ImageJ} (52.49\%) and
+\textit{Fiji} (63.79\%). Table \ref{tab_maintainability_docs} shows which
+projects had these documents, in the descending order of their
+\textit{maintainability} scores. This table contains part of the answers to
+\hyperlink{rq1}{RQ1} and \hyperlink{rq3}{RQ3}.
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{llll}
+\hline
+\multicolumn{1}{c}{Software} & Proj plan & Dev manual & API doc \\ \hline
+3D Slicer & X & X & X \\
+ImageJ & X & X & X \\
+Weasis &  & X &  \\
+OHIF Viewer &  & X & X \\
+Fiji & X & X & X \\
+ParaView & X &  &  \\
+SMILI &  &  & X \\
+medInria &  & X &  \\
+INVESALIUS 3 & X &  &  \\
+dwv &  &  & X \\
+BioImage Suite Web &  & X &  \\
+Gwyddion &  & X & X \\ \hline
+\end{tabular}
+\caption{\label{tab_maintainability_docs}Software with the maintainability documents}
+\end{table}
+
+27 of the 29 projects used git as the version control tool; \textit{AMIDE} used
+Mercurial; \textit{Gwyddion} used Subversion. 24 projects used GitHub for their
+repositories; \textit{XMedCon}, \textit{AMIDE}, and \textit{Gwyddion} used
+SourceForge; \textit{DicomBrowser} and \textit{3DimViewer} used BitBucket. The
+information about version control tools is part of the answers to
+\hyperlink{rq2}{RQ2}.
+
+\subsection{Reusability} \label{sec_result_reusability}
+
+Figure 4.7 shows the AHP results for \textit{reusability}.
+
+\begin{figure}[ht]
+\includegraphics[scale=0.38]{figures/reusability_scores.png}
+\caption{AHP reusability scores}
+\label{fg_reusability_scores}
+\end{figure}
+
+As described in Section \ref{sec_grading_template}, we gave higher scores to the
+projects with an API document. As shown in Table \ref{tab_maintainability_docs},
+seven projects had API documents. We also considered projects with more code
+files and less LOC per code file as more reusable. Table \ref{tab_loc_per_file}
+shows the number of text-based files by projects, which we used to approximate
+the number of code files. The table also lists the total number of lines
+(including comments and blanks), LOC, and average LOC per file. We arranged the
+items in the descending order of their \textit{reusability} scores.
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{lllll}
+\hline
+\multirow{2}{*}{Software} & \multirow{2}{*}{Text files} & \multirow{2}{*}{Total lines} & \multirow{2}{*}{LOC} & \multirow{2}{*}{LOC/file} \\
+ &  &  &  &  \\ \hline
+OHIF Viewer & 1162 & 86306 & 63951 & 55 \\
+3D Slicer & 3386 & 709143 & 501451 & 148 \\
+Gwyddion & 2060 & 787966 & 643427 & 312 \\
+ParaView & 5556 & 1276863 & 886326 & 160 \\
+OsiriX Lite & 2270 & 873025 & 544304 & 240 \\
+Horos & 2346 & 912496 & 561617 & 239 \\
+medInria & 1678 & 214607 & 148924 & 89 \\
+Weasis & 1027 & 156551 & 123272 & 120 \\
+BioImage Suite Web & 931 & 203810 & 139699 & 150 \\
+GATE & 1720 & 311703 & 207122 & 120 \\
+Ginkgo CADx & 974 & 361207 & 257144 & 264 \\
+SMILI & 275 & 90146 & 62626 & 228 \\
+Fiji & 136 & 13764 & 10833 & 80 \\
+Drishti & 757 & 345225 & 268168 & 354 \\
+ITK-SNAP & 677 & 139880 & 88530 & 131 \\
+3DimViewer & 730 & 240627 & 178065 & 244 \\
+DICOM Viewer & 302 & 34701 & 30761 & 102 \\
+ImageJ & 40 & 10740 & 9681 & 242 \\
+dwv & 188 & 71099 & 47815 & 254 \\
+MatrixUser & 216 & 31336 & 23121 & 107 \\
+INVESALIUS 3 & 156 & 59328 & 48605 & 312 \\
+AMIDE & 183 & 139658 & 102827 & 562 \\
+Papaya & 110 & 95594 & 71831 & 653 \\
+MicroView & 137 & 36173 & 27470 & 201 \\
+XMedCon & 202 & 129991 & 96767 & 479 \\
+MRIcroGL & 97 & 50445 & 8493 & 88 \\
+Slice:Drop & 77 & 25720 & 19020 & 247 \\
+DicomBrowser & 54 & 7375 & 5505 & 102 \\
+dicompyler & 48 & 19201 & 15941 & 332 \\ \hline
+\end{tabular}
+\caption{\label{tab_loc_per_file}Number of files and lines}
+\end{table}
+
+\subsection{Surface Understandability} \label{sec_result_understandability}
+
+Figure \ref{fg_surface_understandability_scores} shows the scores for
+\textit{surface understandability}.
+
+\begin{figure}[ht]
+\includegraphics[scale=0.38]{figures/understandability_scores.png}
+\caption{AHP surface understandability scores}
+\label{fg_surface_understandability_scores}
+\end{figure}
+
+All projects had a consistent code style with parameters in the same order for
+all functions; the code was modularized; the comments were clear, indicating
+what is being done, not how. However, we only found explicit identification of a
+coding standard for 3 out of the 29: \textit{3D Slicer}, \textit{Weasis}, and
+\textit{ImageJ}. We also found hard-coded constants in \textit{medInria},
+\textit{dicompyler}, \textit{MicroView}, and \textit{Papaya}. We did not find
+any reference to the used algorithms in projects \textit{XMedCon},
+\textit{DicomBrowser}, \textit{3DimViewer}, \textit{BioImage Suite Web},
+\textit{Slice:Drop}, \textit{MatrixUser}, \textit{DICOM Viewer},
+\textit{dicompyler}, and \textit{Papaya}.
+
+\subsection{Visibility/Transparency} \label{sec_result_visibility_transparency}
+
+Figure \ref{fg_visibility_transparency_scores} shows the AHP scores for
+\textit{visibility/transparency}. Generally speaking, the teams that actively
+documented their development process and plans scored higher because they
+delivered better communication to people outside the team.
+
+\begin{figure}[ht]
+\includegraphics[scale=0.38]{figures/visibility_transparency_scores.png}
+\caption{AHP visibility/transparency scores}
+\label{fg_visibility_transparency_scores}
+\end{figure}
+
+Table \ref{tab_Visibility/Transparency_docs} shows the projects which had
+documents for the development process, project status, development environment,
+and release notes, in the descending order of their
+\textit{visibility/transparency} scores. This table contains part of the answers
+to \hyperlink{rq1}{RQ1} and \hyperlink{rq3}{RQ3}.
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{lllll}
+\hline
+Software & Dev process & Proj status & Dev env & Rls notes \\ \hline
+3D Slicer & X & X & X & X \\
+ImageJ & X & X & X & X \\
+Fiji & X & X & X &  \\
+MRIcroGL &  &  &  & X \\
+Weasis &  &  & X & X \\
+ParaView &  & X &  &  \\
+OHIF Viewer &  &  & X & X \\
+DICOM Viewer &  &  & X & X \\
+medInria &  &  & X & X \\
+SMILI &  &  &  & X \\
+Drishti &  &  &  & X \\
+INVESALIUS 3 &  &  &  & X \\
+OsiriX Lite &  &  &  & X \\
+GATE &  &  &  & X \\
+MicroView &  &  &  & X \\
+MatrixUser &  &  &  & X \\
+BioImage Suite Web &  &  & X &  \\
+ITK-SNAP &  &  &  & X \\
+Horos &  &  &  & X \\
+dwv &  &  &  & X \\
+Gwyddion &  &  &  & X \\ \hline
+\end{tabular}
+\caption{\label{tab_Visibility/Transparency_docs}Software with the visibility/transparency documents}
+\end{table}
+
+\subsection{Overall Scores}
+
+As described in Section \ref{sec_AHP}, for our AHP measurements, there are nine
+criteria which are the nine software qualities and 29 software packages as the
+alternatives. In the absence of requirements for an actual project, we made all
+nine qualities equally important, so the score of each quality affects the
+overall scores on the same scale.
+
+Figure \ref{fg_overall_scores} shows the overall scores of all 29 software
+packages in descending order. Since we produced the scores from the AHP process,
+the total sum of the 29 scores is precisely 1.
+
+\begin{figure}[ht]
+\includegraphics[scale=0.38]{figures/overall_scores.png}
+\caption{Overall AHP scores for all 9 software qualities}
+
+\label{fg_overall_scores}
+\end{figure}
+
+The top three software products \textit{3D Slicer}, \textit{ImageJ}, and
+\textit{OHIF Viewer} had higher scores in most criteria. \textit{3D Slicer}
+ranked in the top two software products for all qualities except \textit{surface
+robustness}; \textit{ImageJ} ranked in the top three products for
+\textit{correctness \& verifiability}, \textit{surface reliability},
+\textit{surface usability}, \textit{maintainability}, \textit{ surface
+understandability}, and \textit{visibility/transparency}; \textit{OHIF Viewer}
+ranked in the top five products for \textit{correctness \& verifiability},
+\textit{surface reliability}, \textit{surface usability},
+\textit{maintainability}, and \textit{reusability}. We might underestimate its
+scores of qualities \textit{surface reliability} and \textit{surface robustness}
+for \textit{DICOM Viewer}, but equally compared it with the other software for
+the rest seven qualities.
+
+\section{Interviews with Developers} \label{ch_interview}
+
+The measurement results in Section \ref{ch_results} are based on the information
+collected by ourselves. Such information is sufficient to measure the projects
+with reasonable efforts, but incomplete for us to understand the development
+process more deeply. For example, we usually cannot identify the following
+information in a project's documents: the pain points during the development,
+the threats to certain software qualities, and the developers' strategies to
+address them. We believe interviews with developers can collect the additional
+information we need. As a result, our method involves interviews with developers
+in a domain (Section \ref{sec_interview_methods}). We applied this step to the
+MI domain (Section \ref{sec_apply_to_mi_interviews}).
+
+In this section, we summarize some answers from the interviews. We highlight the
+answers that are the most informative and interesting in this section, and
+summarize the rest in Appendix \ref{ap_interview}.
+
+As mentioned in Section \ref{sec_apply_to_mi_interviews}, we contacted all 29
+teams. Some of them responded and participated in the interviews. Eventually, we
+interviewed nine developers from eight of the 29 MI software projects. The eight
+projects are \textit{3D Slicer}, \textit{INVESALIUS 3}, \textit{dwv},
+\textit{BioImage Suite Web}, \textit{ITK-SNAP}, \textit{MRIcroGL},
+\textit{Weasis}, and \textit{OHIF}. We spent about 90 minutes for each interview
+and asked 20 prepared questions. We also asked following-up questions when we
+thought it was worth diving deeper. One participant was too busy to have an
+interview, so they wrote down their answers. The interviewees may have provided
+multiple answers to each question. Thus, when counting the number of answers,
+the total result is sometimes larger than nine.
+
+\subsection{Current and Past Pain Points} \label{sec_interview_pain_points}
+
+By asking questions \hyperlink{q9}{9}, \hyperlink{q10}{10}, and
+\hyperlink{q12}{12}, we tried to identify the pain points during the development
+process in the eight projects. The pain points include current and past
+obstacles. We also asked the interviewees how they would solve the problems.
+This section contains the answers to \hyperlink{rq4}{RQ4}. Questions 9, 10, and
+12 are as follows:
+
+\begin{description}
+\item[Q9.] \hypertarget{q9}Currently, what are the most significant obstacles in
+your development process?
+\item[Q10.] \hypertarget{q10}How might you change your development process to
+remove or reduce these obstacles?
+\item[Q12.] \hypertarget{q12}In the past, is there any major obstacle to your
+development process that has been solved? How did you solve it?
+\end{description}
+
+Table \ref{tab_obstacles} shows the number of times the interviewees mentioned
+the current and past obstacles in their projects.
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{llll}
+\hline
+\multirow{2}{*}{Category} & \multirow{2}{*}{Obstacle} & \multicolumn{2}{c}{Num ans.} \\ \cline{3-4} 
+ &  & current & past \\ \hline
+\multirow{2}{*}{Resource} & Lack of fundings & 3 &  \\
+ & Lack of time to devote to the project & 2 & 1 \\ \hdashline
+\multirow{3}{*}{Balance} & Hard to keep up with changes in OS and libraries & 1 &  \\
+ & Hard to support multiple OS & 2 &  \\
+ & Hard to support lower-end computers & 1 & 2 \\ \hdashline
+Testing & Lack of access to real-world datasets for testing & 3 & 2 \\ \hdashline
+\multirow{7}{*}{Others}
+ & Hard to have a high level roadmap from the start & 1 &  \\
+ & Not enough participants for usability tests & 1 &  \\
+ & Only a few people fully understand the large codebase & 1 &  \\
+ & Hard to transfer to new technologies & & 2 \\
+ & Hard to understand users' needs & & 1 \\
+ & Hard to maintain good documentations & & 1 \\ \hline
+\end{tabular}
+\caption{\label{tab_obstacles}Current and past obstacles by the numbers of interviewees with the answers}
+\end{table}
+
+The interviewees provided some potential and proven solutions for the problems
+in Table \ref{tab_obstacles}. We group these pain points into three major
+categories of obstacles: \textit{resource}, \textit{balance}, and
+\textit{testing}. We put the less mentioned ones into the category
+\textit{Others}. Sections \ref{sec_pain_points_1}, \ref{sec_pain_points_2}, and
+\ref{sec_pain_points_3} include further discussion about the three major groups
+of pain points and their solutions.
+
+\subsubsection{Resource Pain Points} \label{sec_pain_points_1}
+
+We summarize the pain points in the \textit{resource} category as
+\textbf{the lack of fundings and time.}
+
+The potential and proven solutions are:
+
+\noindent\textit{Potential solutions from interviewees:}
+\begin{itemize}
+\item when the team does not have enough developers for building new features
+and fixing bugs at the same time, shifting from development mode toward
+maintenance mode;
+\item licensing the software to commercial companies that integrate it into
+their products;
+\item better documentation to save time for answering users' and developers'
+questions;
+\item supporting third-party plugins and extensions.
+\end{itemize}
+
+\noindent\textit{Proven solutions from interviewees:}
+
+\begin{itemize}
+\item GitHub Actions, which is a good CI/CD tool to save time.
+\end{itemize}
+
+Many interviewees thought lack of fundings and lack of time were the most
+significant obstacles. The interviewees from \textit{3D Slicer} team and
+\textit{OHIF} team pointed out that it was more challenging to get fundings for
+software maintenance as opposed to research. The interviewee from the
+\textit{ITK-SNAP} team thought more fundings was a way to solve the lack of time
+problem, because they could hire more dedicated developers. On the other hand,
+the interviewee from the \textit{Weasis} team did not think that fundings could
+solve the same problem, since he still would need a lot of time to supervise the
+project.
+
+No interviewee suggested any solution to bring extra funding to the project.
+However, they provided ideas to save time, such as better documentation,
+third-party plugins, and good CI/CD tools.
+
+\subsubsection{Balance Pain Points} \label{sec_pain_points_2}
+
+We summarize the pain points in the \textit{balance} category as \textbf{the
+difficulty to balance between four factors: cross-platform compatibility,
+convenience to development \& maintenance, performance, and security.} They are
+also related to the choice between native application and web application.
+
+The potential and proven solutions are:
+
+\noindent\textit{Potential solutions from interviewees:}
+
+\begin{itemize}
+\item web applications that use computing power from computers GPU;
+\item to better support lower-end computers, adopting a web-based approach with backend servers;
+\item to better support lower-end computers, using memory-mapped files to
+consume less computer memory;
+\item more funding;
+\item maintaining better documentations to ease the development \& maintenance
+processes;
+\end{itemize}
+
+\noindent\textit{Proven solutions from interviewees:}
+
+\begin{itemize}
+\item one interviewee saw the performance problem disappeared over the years
+when computers became more and more powerful. 
+\end{itemize}
+
+Table \ref{tab_native_vs_web} shows the teams' choices between native
+application and web application. In all the 29 teams on our list, most of them
+chose to develop native applications. For the eight teams we interviewed, three
+of them were building web applications, and the \textit{MRIcroGL} team was
+considering web-based solutions. So we had a good chance to discuss the
+differences between the two choices with the interviewees.
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{lll}
+\hline
+Software team & Native application & Web application \\ \hline
+3D Slicer & X & \\
+INVESALIUS 3 & X & \\
+dwv & & X \\
+BioImage Suite Web & & X \\
+ITK-SNAP & X & \\
+MRIcroGL & X & \\
+Weasis & X & \\
+OHIF & & X \\ \hdashline
+Total number among the eight teams & 5 & 3 \\
+Total number among the 29 teams & 24 & 5 \\ \hline
+\end{tabular}
+\caption{\label{tab_native_vs_web}Teams' choices between native application and
+web application}
+\end{table}
+
+The interviewees talked about the advantages and disadvantages of the two
+choices. We summarize the opinions from the interviewees in Table
+\ref{tab_pros_cons_native_vs_web}.
+
+\begin{table}[ht]
+\centering
+\hspace*{-2cm}\begin{tabular}{lll}
+\hline
+ & Native application & Web application \\ \hline
+Ad & - higher performance & \begin{tabular}[c]{@{}l@{}}- easy to achieve cross-platform compatibility\\ - simpler build process\end{tabular} \\ \hdashline
+Disad & \begin{tabular}[c]{@{}l@{}}- hard to achieve cross-platform compatibility\\ - more complicated build process\end{tabular} & \begin{tabular}[c]{@{}l@{}}\textit{Without a backend:} \\ - lower performance\\ \textit{With a backend:} \\ - harder for privacy protection \\ - extra cost for backend servers \end{tabular} \\ \hline
+\end{tabular}
+\caption{\label{tab_pros_cons_native_vs_web}Advantages and disadvantages of native application and web application}
+\end{table}
+
+\subsubsection{Testing Pain Point} \label{sec_pain_points_3}
+
+The pain point in the \textit{testing} category is \textbf{the lack of access to
+real-world datasets for testing.} The information about software testing in this
+section is part of the answers to \hyperlink{rq3}{RQ3}.
+
+The potential and proven solutions are:
+
+\noindent\textit{Potential solutions from interviewees:}
+
+\begin{itemize}
+\item using open datasets
+\end{itemize}
+
+\noindent\textit{Proven solutions from interviewees:}
+
+\begin{itemize}
+\item asking the users to provide deidentified copies of medical images if they
+have problems loading the images;
+\item sending the beta versions of software to medical workers who can access
+the data and complete the tests;
+\item if (part of) the team belongs to a medical school or a hospital, using the
+datasets they can access;
+\item if the team has access to MRI scanners, self-building sample images for
+testing;
+\item if the team has connections with MI equipment manufacturers, asking for
+their help on data format problems;
+\item storing all images that cause special problems, and maintaining this
+special dataset over time.
+\end{itemize}
+
+No interviewee provided a perfect way to solve this problem. However,
+connections between the development team and medical professionals/institutions
+could ease the pain.
+
+\subsection{Documents in the Projects} \label{sec_interview_documents}
+
+We tried to understand the interviewees' opinions on documentation and the
+quality of documentations with questions 11 and 19. The information about
+documentation is part of the answers to \hyperlink{rq3}{RQ3}.
+
+\begin{description}
+\item[Q11.] How does documentation fit into your development process? Would
+improved documentation help with the obstacles you typically face?
+\item[Q19.] Do you think the current documentation can clearly convey all
+necessary knowledge to the users? If yes, how did you successfully achieve it?
+If no, what improvements are needed?
+\end{description}
+
+Table \ref{tab_opinion_doc} summarizes interviewees' opinions on documentation.
+Interviewees from each of the eight projects thought that documentation was
+important to their projects, and most of them said that it could save their time
+to answer questions from users and developers. Most of them saw the need to
+improve their documentation, and only three of them thought that their
+documentations conveyed information clearly enough.
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{ll}
+\hline
+Opinion on documentation & Num ans. \\ \hline
+Documentation is vital to the project& 8 \\
+Documentation of the project needs improvements & 7 \\
+Referring to documentation saves time to answer questions & 6 \\
+Lack of time to maintain good documentation & 4 \\
+Documentation of the project conveys information clearly & 3 \\
+Coding is more fun than documentation & 2 \\
+Users help each other by referring to documentation & 1 \\ \hline
+\end{tabular}
+\caption{\label{tab_opinion_doc}Opinions on documentation by the numbers of
+interviewees with the answers}
+\end{table}
+
+Table \ref{tab_doc_tools} lists some of the documentation tools and methods
+mentioned by the interviewees. This table contains part of the answers to
+\hyperlink{rq2}{RQ2}.
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{ll}
+\hline
+Tool or method for documentation & Num ans. \\ \hline
+Forum discussions & 3 \\
+Videos & 3 \\
+GitHub & 2 \\
+Mediawiki / wiki pages & 2 \\
+Workshops & 2 \\
+Social media & 2 \\
+Writing books & 1 \\
+Google Form & 1 \\
+State management & 1 \\ \hline
+\end{tabular}
+\caption{\label{tab_doc_tools}Tools and methods for documentation by the numbers
+of interviewees with the answers}
+\end{table}
+
+\subsection{Contribution Management and Project Management}
+\label{sec_contribution_pm}
+We tried to understand how the teams managed the contributions and their
+projects by asking the following questions:
+
+\begin{description}
+\item[Q5.] Do you have a defined process for accepting new contributions into
+your team?
+\item[Q13.] What is your software development model? For example, waterfall,
+agile, etc.
+\item[Q14.] \hypertarget{q14}What is your project management process? Do you
+think improving this process can tackle the current problem? Were any project
+management tools used?
+\end{description}
+
+Although some team may have a documented process for accepting new
+contributions, no one talked about it during the interview. However, most of
+them mentioned using GitHub and pull requests to manage contributions from the
+community. The interviewees generally gave very positive feedback on using
+GitHub. Some also said they had handled the project repository with some other
+tools, and eventually transferred to git and GitHub. Table
+\ref{tab_method_receive_contributions} shows the number of times the
+interviewees mentioned the methods of receiving contributions. This table
+contains part of the answers to \hyperlink{rq2}{RQ2}.
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{lll}
+\hline
+\multirow{2}{*}{Method of receiving contributions} & \multicolumn{2}{l}{Num ans.} \\ \cline{2-3} 
+ & current & past \\ \hline
+GitHub with pull requests & 8 & \\
+Code contributions from emails & & 3 \\
+Code contributions from forums & & 1 \\
+Sharing the git repository by email & & 1 \\ \hline
+\end{tabular}
+\caption{\label{tab_method_receive_contributions}Methods of receiving
+contributions by the numbers of interviewees with the answers}
+\end{table}
+
+For managing contributions, the \textit{3D Slicer} team encouraged users to
+develop their extensions for specific use cases, and the \textit{OHIF} team was
+trying to enable the use of plug-ins; the interviewee from the \textit{ITK-SNAP}
+team said one way of accepting new team members was through funded academic
+projects.
+
+Table \ref{tab_developmen_models} shows the software development models by the
+numbers of interviewees with the answers. Only two interviewees confirmed their
+development models. The others did not think they used a specific model, but
+three of them suggested that their processes were similar to Waterfall or Agile.
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{ll}
+\hline
+Software development model & Num ans. \\ \hline
+Undefined/self-directed & 3 \\
+Similar to Agile & 2 \\
+Similar to Waterfall & 1 \\
+Agile & 1 \\
+Waterfall & 1 \\ \hline
+\end{tabular}
+\caption{\label{tab_developmen_models}Software development models by the numbers
+of interviewees with the answers}
+\end{table}
+
+Some interviewees mentioned the project management tools they used, which are in
+Table \ref{tab_pm_tools}. This table contains part of the answers to
+\hyperlink{rq2}{RQ2}. Generally speaking, the interviewees talked about two
+types:
+
+\begin{itemize}
+\item Trackers, including GitHub, issue trackers, bug trackers and Jira;
+\item Documents, including GitHub, Wiki page, Google Doc, and Confluence.
+\end{itemize}
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{ll}
+\hline
+Project management tools & Num ans. \\ \hline
+GitHub & 3 \\
+Issue trackers & 1 \\
+Bug trackers & 1 \\
+Jira & 1 \\
+Wiki page & 1 \\
+Google Doc & 1 \\
+Confluence & 1 \\ \hline
+\end{tabular}
+\caption{\label{tab_pm_tools}Project management tools by the numbers of
+interviewees with the answers}
+\end{table}
+
+No interviewee introduced any strictly defined project management process. The
+most common way was following the issues, such as bugs and feature requests.
+Additionally, the \textit{3D Slicer} team had weekly meetings to discuss the
+goals for the project; the \textit{INVESALIUS 3} team relied on the GitHub
+process for their project management; the \textit{ITK-SNAP} team had a fixed
+six-month release pace; only the interviewee from the \textit{OHIF} team
+mentioned that the team has a project manager; the \textit{3D Slicer} team and
+\textit{BioImage Suite Web} team were doing nightly builds and tests.
+
+Most interviewees skipped the second part of \hyperlink{q14}{Q14} ``Do you think
+improving this process can tackle the current problem?". In retrospect, we
+should not have asked a yes-or-no question, since it is not very informative.
+The interviewee from the \textit{OHIF} team gave a positive answer to this
+question. They believed that a better project management process can improve the
+efficiency of junior developers. They also improved the project management tools
+(from public Jira to public GitHub repository plus private Jira), so they could
+better communicate externally and internally.
+
+The information about contribution management and project management in this
+section is part of the answers to \hyperlink{rq3}{RQ3}.
+
+\subsection{Discussions on Software Qualities} \label{sec_interview_software_qualities}
+
+Questions 15--18, and 20 are about the software qualities of
+\textit{correctness}, \textit{maintainability}, \textit{understandability},
+\textit{usability}, and \textit{reproducibility}, respectively. We asked these
+questions to understand the threats to these qualities and the developers'
+strategies to improve them. We discuss each quality in a separate section below.
+This section contains part of the answers to \hyperlink{rq5}{RQ5}.
+
+\subsubsection{Correctness} \label{sec_interview_correctness}
+
+\begin{description}
+\item[Q15.] Was it hard to ensure the correctness of the software? If there were
+any obstacles, what methods have been considered or practiced to improve the
+situation? If practiced, did it work?
+\end{description}
+
+Table \ref{tab_q15_threats_correctness} shows the threats to
+\textit{correctness} by the numbers of interviewees with the answers.
+
+\begin{table}[ht]
+\centering
+\hspace*{-1cm}\begin{tabular}{ll}
+\hline
+Threat to correctness & Num ans. \\ \hline
+Complexity - data in various formats and complicated standards. & 2 \\
+Complexity - different MI machines create data in different ways. & 2 \\
+Complexity - additional functions besides viewing. & 1 \\
+The lack of real word image data for testing. & 1 \\
+The team cannot use private data for debugging even when the data cause
+problems. & 1 \\
+With huge datasets for testing, the tests are expensive and time-consuming. & 1 \\
+It is hard to manage releases. & 1 \\
+The project has no unit tests. & 1 \\
+The project has no dedicated quality assurance team. & 1 \\ \hline
+\end{tabular}
+\caption{\label{tab_q15_threats_correctness}Threats to correctness by the
+numbers of interviewees with the answers}
+\end{table}
+
+Table \ref{tab_q15_strategies_correctness} shows the strategies to ensure
+\textit{correctness} by the numbers of interviewees with the answers. The
+interviewees from the \textit{3D Slicer} and \textit{ITK-SNAP} teams thought
+that the self-tests and automated tests were beneficial and could significantly
+save time. The interviewee from the \textit{Weasis} team kept collecting medical
+images for more than ten years. These images have caused problems with the
+software. So he had samples to test specific problems.
+
+\begin{table}[ht]
+\centering
+\hspace*{-1.5cm}\begin{tabular}{ll}
+\hline
+Strategy to ensure correctness & Num ans. \\ \hline
+Test-driven development, component tests, integration tests, smoke tests, regression tests. & 4 \\
+Self tests / automated tests. & 3 \\
+Two stage development process / stable release \& nightly builds. & 3 \\
+CI/CD. & 1 \\
+Using deidentified copies of medical images for debugging. & 1 \\
+Sending beta versions to medical workers who can access the data to do the tests. & 1 \\
+Collecting and maintaining a dataset of problematic images. & 1 \\ \hline
+\end{tabular}
+\caption{\label{tab_q15_strategies_correctness}Strategies to ensure correctness
+by the numbers of interviewees with the answers}
+\end{table}
+
+The information about software testing in this section is part of the answers to
+\hyperlink{rq3}{RQ3}.
+
+\subsubsection{Maintainability} \label{sec_interview_maintainability}
+
+\begin{description}
+\item[Q16.] \hypertarget{q16}When designing the software, did you consider the
+ease of future changes? For example, will it be hard to change the system’s
+structure, modules, or code blocks? What measures have been taken to ensure the
+ease of future changes and maintains?
+\end{description}
+
+Table \ref{tab_q16_strategies_maintainability} shows the strategies to ensure
+\textit{maintainability} by the numbers of interviewees with the answers. The
+modular approach is the most talked-about solution to improve
+\textit{maintainability}. The \textit{3D Slicer} team used a well-defined
+structure for the software, which they named as ``event-driven MVC pattern''.
+Moreover, \textit{3D Slicer} discovers and loads necessary modules at runtime,
+according to the configuration and installed extensions. The \textit{BioImage
+Suite Web} team had designed and re-designed their software multiple times in
+the last 10+ years. They found that their modular approach effectively supported
+the maintainability \citep{Joshi2011}.
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{ll}
+\hline
+Strategy to ensure maintainability & Num ans. \\ \hline
+Modular approach / maintain repetitive functions as libraries. & 5 \\
+Supporting third-party extensions. & 1 \\
+Easy-to-understand architecture. & 1 \\
+Dedicated architect. & 1 \\
+Starting from simple solutions. & 1 \\
+Documentation. & 1 \\ \hline
+\end{tabular}
+\caption{\label{tab_q16_strategies_maintainability}Strategies to ensure
+maintainability by the numbers of interviewees with the answers}
+\end{table}
+
+\subsubsection{Understandability} \label{sec_interview_understandability}
+
+\begin{description}
+\item[Q17.] \hypertarget{q17}Provide instances where users have misunderstood
+the software. What, if any, actions were taken to address understandability
+issues?
+\end{description}
+
+Table \ref{tab_q17_threats_understandability} shows the threats to
+\textit{understandability} by the numbers of interviewees with the answers. It
+separates \textit{understandability} issues to users and developers by the
+horizontal dash line.
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{ll}
+\hline
+Threat to understandability & Num ans. \\ \hline
+Not all users understand how to use some features. & 2 \\
+The team has no dedicated user experience (UX) designer. & 1 \\
+Some important indicators are not noticeable (e.g. a progress bar). & 1 \\
+Not all users understand the purpose of the software. & 1 \\
+Not all users know if the software includes certain features. & 1 \\
+Not all users understand how to use the command line tool. & 1 \\
+Not all users understand that the software is a web application. & 1 \\\hdashline
+Not all developers understand how to deploy the software. & 1 \\
+The architecture is difficult for new developers to understand. & 1 \\ \hline
+\end{tabular}
+\caption{\label{tab_q17_threats_understandability}Threats to understandability by the numbers of interviewees with the answers}
+\end{table}
+
+Table \ref{tab_q17_strategies_understandability} shows the strategies to ensure \textit{understandability} by the numbers of interviewees with the answers.
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{ll}
+\hline
+Strategy to ensure understandability & Num ans. \\ \hline
+Documentation / user manual / user mailing list / forum. & 4 \\
+Graphical user interface. & 2 \\
+Testing every release with active users. & 1 \\
+Making simple things simple and complicated things possible. & 1 \\
+Icons with more clear visual expressions. & 1 \\
+Designing the software to be intuitive. & 1 \\
+Having a UX designer with the right experience. & 1 \\
+Dialog windows for important notifications. & 1 \\
+Providing an example if the users need to build the software by themselves. & 1
+\\ \hline
+\end{tabular}
+\caption{\label{tab_q17_strategies_understandability}Strategies to ensure understandability by the numbers of interviewees with the answers}
+\end{table}
+
+\subsubsection{Usability} \label{sec_interview_usability}
+
+\begin{description}
+\item[Q18.] \hypertarget{q18}What, if any, actions were taken to address
+usability issues?
+\end{description}
+
+Table \ref{tab_q18_strategies_usability} shows the strategies to ensure
+\textit{usability} by the numbers of interviewees with the answers. The
+information about software testing in this table is part of the answers to
+\hyperlink{rq3}{RQ3}.
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{ll}
+\hline
+Strategy to ensure usability & Num ans. \\ \hline
+Usability tests and interviews with end users. & 3 \\
+Adjusting according to users' feedbacks. & 3 \\
+Straightforward and intuitively designed interface / professional UX designer. & 2 \\
+Providing step-by-step processes, and showing the step numbers. & 1 \\
+Making the basic functions easy to use without reading the documentation. & 1 \\
+Focusing on limited number of functions. & 1 \\
+Making the software more streamlined. & 1 \\
+Downsampling images to consume less memory. & 1 \\
+An option to load only part of the data to boost performance. & 1 \\ \hline
+\end{tabular}
+\caption{\label{tab_q18_strategies_usability}Strategies to ensure usability by
+the numbers of interviewees with the answers}
+\end{table}
+
+\subsubsection{Reproducibility} \label{sec_interview_reproducibility}
+
+\begin{description}
+\item[Q20.] Do you have any concern that your computational results won’t be
+reproducible in the future? Have you taken any steps to ensure reproducibility?
+\end{description}
+
+Table \ref{tab_q20_threats_reproducibility} shows the threats to
+\textit{reproducibility} by the numbers of interviewees with the answers.
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{ll}
+\hline
+Threat to reproducibility & Num ans. \\ \hline
+If the software is closed-source, the reproducibility is hard to achieve. & 1 \\
+The project has no user interaction tests. & 1 \\
+The project has no unit tests. & 1 \\
+Using different versions of some common libraries may cause problems. & 1 \\
+CPU variability can leads to non-reproducibility. & 1 \\
+The team may misinterpret how manufacturers create medical images. & 1 \\ \hline
+\end{tabular}
+\caption{\label{tab_q20_threats_reproducibility}Threats to reproducibility by
+the numbers of interviewees with the answers}
+\end{table}
+
+Table \ref{tab_q20_strategies_reproducibility} shows the strategies to ensure
+\textit{ reproducibility} by the numbers of interviewees with the answers. The
+interviewee from the \textit{3D Slicer} team provided various suggestions. One
+interviewee from another team suggested that they used \textit{3D Slicer} as the
+benchmark to test their \textit{reproducibility}.
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{ll}
+\hline
+Strategy to ensure reproducibility & Num ans. \\ \hline
+Regression tests / unit tests / having good tests. & 6 \\
+Making code, data, and documentation available / OSS / open-source libraries. & 5 \\
+Running same tests on all platforms. & 1 \\
+A dockerized version of the software,  insulating it from the OS environment. & 1 \\
+Using standard libraries. & 1 \\
+Monitoring the upgrades of the libraries. & 1 \\
+Clearly documenting the versions. & 1 \\
+Bringing along the exact versions of all the dependencies with the software. & 1 \\
+Providing checksums of the data. & 1 \\
+Benchmarking the software against other software with similar purposes. & 1 \\ \hline
+\end{tabular}
+\caption{\label{tab_q20_strategies_reproducibility}Strategies to ensure
+reproducibility by the numbers of interviewees with the answers}
+\end{table}
+
+The information about software testing in this section is part of the answers to
+\hyperlink{rq3}{RQ3}.
+
+\section{Answers to Research Questions} \label{ch_answers}
+
+This section answers our research questions from Section
+\ref{sec_research_questions}.  Sections \ref{sec_rq_artifacts}--
+\ref{sec_rq_comparison} summarize the answers to the six questions,
+respectively. Section \ref{sec_threats_to_validity} lists the threats to the
+validity of our research. The answers are based on our quality measurements in
+Section \ref{ch_results} and developer interviews in Section \ref{ch_interview}.
+We refer to these sections to avoid repetition, and organize the references in
+tables (e.g. Table \ref{tab_reference_rq1}).
+
+\subsection{Artifacts in the Projects} \label{sec_rq_artifacts}
+\begin{description}
+\item[RQ1.] What artifacts are present in current software projects?
+\end{description}
+
+We answer this question by examining the documentation, scanning the source
+code, and interviewing the developers of the projects. Table
+\ref{tab_reference_rq1} shows the sections and tables containing answers to this
+research question.
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{ll}
+\hline
+Section or table & Description \\ \hline
+Table \ref{tab_maintainability_docs} & Maintainability documents \\
+Table \ref{tab_Visibility/Transparency_docs} & Visibility/transparency documents \\ \hline
+\end{tabular}
+\caption{\label{tab_reference_rq1}Sections and tables with answers to RQ1}
+\end{table}
+
+As mentioned in Section \ref{sec_grading_template}, we search for all the
+artifacts in a project when measuring \textit{maintainability}. The detailed
+records of the existing artifacts in the 29 MI projects are at
+\hyperlink{https://data.mendeley.com/datasets/k3pcdvdzj2/1}{https://data.mendeley.com/datasets/k3pcdvdzj2/1}.
+Table \ref{tab_artifacts_frequency} summarizes the frequency of the artifacts.
+This table also contains part of the answers to \hyperlink{rq3}{RQ3}.
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{ll}
+\hline
+Artifact & Number of projects \\ \hline
+README & 29 \\
+Version Control & 29 \\
+License & 28 \\
+Bug tracker & 28 \\
+Change request & 28 \\
+User Manual & 22 \\
+Release notes & 22 \\
+Build file & 18 \\
+Tutorials & 18 \\
+Installation Guide & 16 \\
+Test cases & 15 \\
+Authors & 14 \\
+FAQ & 14 \\
+Acknowledgements & 12 \\
+Executable files & 10 \\
+Developer's Manual & 8 \\
+API documentation & 7 \\
+Troubleshooting guide & 6 \\
+Project Plan & 5 \\ \hline
+\end{tabular}
+\caption{\label{tab_artifacts_frequency}Artifacts by their frequency in the 29 MI projects}
+\end{table}
+
+We summarized the definitions of the artifacts in
+\hyperlink{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Artifacts_MiningV3.xlsx}{https://github.com/.../Artifacts\_MiningV3.xlsx}.
+Source code is a type of artifact. Since we only included OSS on the final list
+(Section \ref{sec_mi_software_selection}), every project's source code was
+available. Thus, we excluded it in the above table.
+
+\subsection{Tools in the Projects}
+\label{sec_rq_tools}
+\begin{description}
+\item[RQ2.] What tools are used in the development of current software packages?
+\end{description}
+
+We answer this question by measuring the qualities and interviewing the
+developers. This section summarizes the tools used for CI/CD, user support,
+version control, documentation, contribution management, and project management.
+Table \ref{tab_reference_rq2} shows the sections and tables containing answers
+to this research question.
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{ll}
+\hline
+Section or table & Description \\ \hline
+Section \ref{sec_result_correctness_verifiability} & CI/CD tools \\
+Table \ref{tab_user_support_model} & User support tools \\
+Section \ref{sec_score_maintainability} & Version control tools \\
+Table \ref{tab_doc_tools} & Documentation tools \\
+Table \ref{tab_method_receive_contributions} & Contribution management tools \\
+Table \ref{tab_pm_tools} & Project management tools\\ \hline
+\end{tabular}
+\caption{\label{tab_reference_rq2}Sections and tables with answers to RQ2}
+\end{table}
+
+As mentioned in Section \ref{sec_result_correctness_verifiability}, we
+identified five projects using CI/CD tools. \textit{3D Slicer} and \textit{OHIF
+Viewer} used CircleCI; \textit{ImageJ}, \textit{Fiji}, and \textit{dwv} used
+Travis. We identified the above projects and tools by examining the
+documentation and source code of all projects. Thus, we may missed some projects
+or tools. According to the interviews with developers, \textit{dwv} and
+\textit{Weasis} used GitHub Actions.
+
+\subsection{Principles, Processes, and Methodologies in the Projects}
+\label{sec_rq_PPM}
+\begin{description}
+\item[RQ3.] What principles, processes, and methodologies are used in the
+development of current software packages?
+\end{description}
+
+We answer this question by measuring the qualities and interviewing the
+developers. This section shows the principles, processes, and methodologies in
+software testing, documentation, contribution management, project management,
+and improving software qualities. Table \ref{tab_reference_rq3} shows the
+sections and tables containing answers to this research question.
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{ll}
+\hline
+Section or table & Description \\ \hline
+Section \ref{fg_correctness_erifiability_scores}, \ref{sec_pain_points_3},
+\ref{sec_interview_correctness}, and \ref{sec_interview_reproducibility}; Table
+\ref{tab_q18_strategies_usability} & Software testing \\
+Section \ref{fg_correctness_erifiability_scores} and
+\ref{sec_interview_documents}; Table \ref{tab_maintainability_docs},
+\ref{tab_Visibility/Transparency_docs}, and \ref{tab_artifacts_frequency} &
+Documentation \\
+Section \ref{sec_contribution_pm} & Contribution management \\
+Section \ref{sec_contribution_pm} & Project management \\ \hline
+\end{tabular}
+\caption{\label{tab_reference_rq3}Sections and tables with answers to RQ3}
+\end{table}
+
+We identified the use of unit testing in less than half of the 29 projects. On
+the other hand, the interviewees believed that testing (including usability
+tests with users) was the top solution to improve \textit{correctness},
+\textit{usability}, and \textit{reproducibility}. One pain point in the
+development process is the lack of access to real-world datasets for testing.
+The developers' strategies to address it is in Section \ref{sec_pain_points_3}.
+One threat to \textit{correctness} is: with huge datasets for testing, the tests
+are expensive and time-consuming. Three interviewees endorsed self tests /
+automated tests, which may save time for testing.
+
+All 29 projects did not have theory manuals. We identified a road map in the
+\textit{3D Slicer} project, and no requirements specifications for the rest.
+Eight of the nine interviewees thought that documentation was essential to their
+projects. However, they hold the common opinion that their documentation needed
+improvements. Nearly half of them also believed that the lack of time prevented
+them from improving the documentation.
+
+\subsection{Pain Points and Solutions}
+\label{sec_rq_pain_points}
+\begin{description}
+\item[RQ4.] What are the pain points for developers working on research software
+projects? What aspects of the existing processes, methodologies, and tools do
+they consider as potentially needing improvement? What changes to processes,
+methodologies, and tools can improve software development and software quality?
+\end{description}
+
+We answer this question by interviewing the developers. The answers to this
+question, including the pain points and the solutions proposed by the
+developers, are in Section \ref{sec_interview_pain_points}.
+
+\subsection{Software Qualities}
+\label{sec_rq_qualities}
+\begin{description}\item[RQ5.] What is the current status of the following
+software qualities for the projects? What actions have the developers taken to
+address them?\end{description}
+
+This section includes our answers from the qualities measurements and interviews
+with the developers. Table \ref{tab_reference_rq4} shows the sections and tables
+containing answers to this research question.
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{ll}
+\hline
+Section or table & Description \\ \hline
+Section \ref{sec_result_installability} & Installability \\
+Section \ref{sec_result_correctness_verifiability} and \ref{sec_interview_correctness} & Correctness \& verifiability \\
+Section \ref{sec_result_reliability} & Reliability \\
+Section \ref{sec_result_robustness} & Robustness \\
+Section \ref{sec_result_usability} and \ref{sec_interview_usability}& Usability \\
+Section \ref{sec_score_maintainability} and \ref{sec_interview_maintainability} & Maintainability \\
+Section \ref{sec_result_reusability} & Reusability \\
+Section \ref{sec_result_understandability} and \ref{sec_interview_understandability} & Understandability \\
+Section \ref{sec_result_visibility_transparency} & Visibility/transparency \\
+Section \ref{sec_interview_reproducibility} & Reproducibility \\ \hline
+\end{tabular}
+\caption{\label{tab_reference_rq4}Sections and tables with answers to RQ4}
+\end{table}
+
+\subsection{Our Ranking versus the Community Ratings}
+\label{sec_rq_comparison}
+\begin{description}
+\item[RQ6.] How does software designated as high quality by this methodology
+compare with top-rated software by the community?
+\end{description}
+
+We answer this question by grading the qualities of the software, collecting
+ratings from the MI community, then conducting two comparisons:
+\begin{itemize}
+\item comparing our ranking with the community ratings on GitHub, such as GitHub
+stars, number of forks, and number of people watching the projects (Section
+\ref{sec_ranking_vs_github});
+\item comparing top-rated software designated by our methodology with the ones
+recommended by our domain experts (Section \ref{sec_designation_vs_experts}).
+\end{itemize}
+
+\subsubsection{Our Ranking versus the GitHub Popularity} \label{sec_ranking_vs_github}
+
+Table \ref{tab_ranking_vs_GitHub} shows our ranking to the 29 MI projects, and
+their GitHub metrics if applicable. As mentioned in Section
+\ref{sec_score_maintainability}, 24 projects used GitHub. Since GitHub
+repositories have different creation dates, we collect the number of months each
+stayed on GitHub, and calculate the average number of new stars, people
+watching, and forks per 12 months. The method of getting the creation date is
+described in Section \ref{sec_empirical_measurements}, and we obtained these
+metrics in July, 2021.
+
+In Table \ref{tab_ranking_vs_GitHub}, we used the average number of new stars
+per 12 months as an indicator of the GitHub popularity, and listed the items in
+the descending order of this number. We ordered the non-GitHub items by our
+ranking. Generally speaking, most of the top-ranking MI software projects also
+received greater attention and popularity on GitHub. Between our ranking and the
+GitHub stars-per-year ranking, four of the top five software projects are the
+same.
+
+Project \textit{dwv} was popular on GitHub, but we ranked it low. As mentioned
+in Section \ref{sec_result_installability}, we failed to build it locally, and
+used the test version on its websites for the measurements. We followed the
+instructions and tried to run the command ``yarn run test" locally, which did
+not work. In addition, the test version did not detect a broken DICOM file and
+displayed a blank image as described in Section \ref{sec_result_robustness}. We
+might underestimate the scores for \textit{dwv} due to uncommon technical
+issues. We also ranked \textit{DICOM Viewer} much lower than its popularity. As
+mentioned in Section \ref{sec_result_installability} it depended on the
+NextCloud platform that we could not successfully install. Thus, we might
+underestimate the scores of its \textit{surface reliability} and \textit{surface
+robustness}. We weighted all qualities equally, which is not likely to be the
+case with all users. As a result, some projects with high popularity may not
+perform well in all qualities.
+
+\begingroup
+\renewcommand{\arraystretch}{0.85}
+\begin{table}[ht]
+\centering
+\begin{tabular}{lllll}
+\hline
+Software & Our ranking & Stars/yr & Watches/yr & Forks/yr \\ \hline
+3D Slicer & 1 & 284 & 19 & 128 \\
+OHIF Viewer & 3 & 277 & 19 & 224 \\
+dwv & 23 & 124 & 12 & 51 \\
+ImageJ & 2 & 84 & 9 & 30 \\
+ParaView & 5 & 67 & 7 & 28 \\
+Horos & 12 & 49 & 9 & 18 \\
+Papaya & 17 & 45 & 5 & 20 \\
+Fiji & 4 & 44 & 5 & 21 \\
+DICOM Viewer & 29 & 43 & 6 & 9 \\
+INVESALIUS 3 & 10 & 40 & 4 & 17 \\
+Weasis & 6 & 36 & 5 & 19 \\
+dicompyler & 26 & 35 & 5 & 14 \\
+OsiriX Lite & 9 & 34 & 9 & 24 \\
+MRIcroGL & 18 & 24 & 3 & 3 \\
+GATE & 25 & 19 & 6 & 26 \\
+Ginkgo CADx & 14 & 19 & 4 & 6 \\
+BioImage Suite Web & 8 & 18 & 5 & 7 \\
+Drishti & 27 & 16 & 4 & 4 \\
+Slice:Drop & 20 & 10 & 2 & 5 \\
+ITK-SNAP & 15 & 9 & 1 & 4 \\
+medInria & 7 & 7 & 3 & 6 \\
+SMILI & 13 & 3 & 1 & 2 \\
+MatrixUser & 28 & 2 & 0 & 0 \\
+MicroView & 16 & 1 & 1 & 1 \\
+Gwyddion & 11 & n/a & n/a & n/a \\
+XMedCon & 19 & n/a & n/a & n/a \\
+DicomBrower & 21 & n/a & n/a & n/a \\
+AMIDE & 22 & n/a & n/a & n/a \\
+3DimViewer & 24 & n/a & n/a & n/a \\ \hline
+\end{tabular}
+\caption{\label{tab_ranking_vs_GitHub}Software ranking versus GitHub metrics}
+\end{table}
+\endgroup
+
+\subsubsection{Designated Top Software versus the Domain Experts' Recommendation}
+\label{sec_designation_vs_experts}
+
+As shown in Section \ref{sec_mi_software_selection}, our domain experts
+recommended a list of top software with 12 software products. Table
+\ref{tab_top_software_vs_experts} and \ref{tab_experts_vs_top_software} compare
+the top 12 software projects ranked by our methodology with the ones from the
+domain experts.
+
+All of the top 4 software from the domain experts are among the top 12 ones
+ranked by our methodology. 3 of the top 4 on both lists are the same ones:
+\textit{3D Slicer}, \textit{ImageJ}, and \textit{Fiji}. \textit{3D Slicer} is
+also the top 1 of both rankings.
+
+As mentioned in Section \ref{sec_mi_software_selection}, six of the recommended
+software packages did not have visualization as the primary function, so we did
+not include them on our final list.
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{lll}
+\hline
+Our ranking & Assessed software & Domain experts \\ \hline
+1 & 3D Slicer & 1 \\
+2 & ImageJ & 3 \\
+3 & OHIF Viewer & n/a \\
+4 & Fiji & 4 \\
+5 & ParaView & n/a \\
+6 & Weasis & n/a \\
+7 & medInria & n/a \\
+8 & BioImage Suite Web & n/a \\
+9 & OsiriX Lite & n/a \\
+10 & INVESALIUS 3 & n/a \\
+11 & Gwyddion & n/a \\
+12 & Horos & 2 \\ \hline
+\end{tabular}
+\caption{\label{tab_top_software_vs_experts}Top software by our ranking versus domain experts' recommendation}
+\end{table}
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{lll}
+\hline
+Our ranking & Recommended software & Domain experts \\ \hline
+1 & 3D Slicer & 1 \\
+12 & Horos & 2 \\
+2 & ImageJ & 3 \\
+4 & Fiji & 4 \\
+n/a & AFNI & 5 \\
+n/a & FSL & 6 \\
+n/a & Freesurfer & 7 \\
+18* & Mricron & 8 \\
+17** & Mango & 9 \\
+n/a & Tarquin & 10 \\
+n/a & Diffusion Toolkit & 11 \\
+n/a & MRItrix & 12 \\ \hline
+\end{tabular}
+\caption{\label{tab_experts_vs_top_software}Domain experts' recommendation versus our ranking}
+\end{table}
+
+* \textit{MRIcron} development had moved to \textit{MRIcroGL}, as mentioned in
+Section \ref{sec_mi_software_selection}. Thus, we measured and ranked
+\textit{MRIcroGL} at 18.
+
+** We included \textit{Mango} in the initial list, but removed it because it was
+not OSS. \textit{Papaya} is a the web version OSS of \textit{Mango}. We measured
+and ranked \textit{Papaya} at 17.
+
+\subsection{Threats to Validity} \label{sec_threats_to_validity}
+
+This section lists all the potential threats to validity in our research. The
+definitions of common validity are
+\citep{AmpatzoglouEtAl2019}\citep{ZhouEtAl2016}:
+\begin{description}
+\item[Construct Validity:] ``Defines how effectively a test or experiment measures
+up to its claims. This aspect deals with whether or not the researcher measures
+what is intended to be measured" \citep{AmpatzoglouEtAl2019}.
+\item[Internal Validity:] ``This aspect relates to the examination of causal
+relations. Internal validity examines whether an experimental
+treatment/condition makes a difference or not, and whether there is evidence to
+support the claim" \citep{AmpatzoglouEtAl2019}.
+\item[External Validity:] ``Define the domain to which a study's findings can be
+generalized" \citep{ZhouEtAl2016}.
+\item[Conclusion Validity:] ``Demonstrate that the operations of a study such as
+the data collection procedure can be repeated, with the same results"
+\citep{ZhouEtAl2016}.
+\end{description}
+
+\noindent We categorize and present the threats to validity in the following subsections.
+
+\subsubsection{Threats to Construct Validity}
+\begin{itemize}
+\item We compared nine software qualities for 29 software packages, so we could
+only spend a limited time on each of them. As a result, our assessments may have
+missed something relevant.
+\item Our ranking is partly based on surface (shallow) measurement, which may
+not fully reveal the underlying qualities.
+\end{itemize}
+
+\subsubsection{Threats to Internal Validity}
+\begin{itemize}
+\item It was not practical to ask each development team for every piece of
+information. We collected much information - such as artifacts and funding
+situations of software - by ourselves. There may be cases that we missed some
+information.
+\item As mentioned in Section \ref{ch_interview}, one interviewee was too busy
+to participate in a full interview, so he provided a version of written answers
+to us. Since we did not have the chance to explain our questions or ask him
+follow-up questions, there is a possibility of misinterpretation of the
+questions or answers.
+\item As mentioned in Section \ref{sec_result_installability}, we could not
+install or build \textit{dwv}, \textit{GATE}, and \textit{DICOM Viewer}. We used
+a deployed online version for \textit{dwv}, a VM version for \textit{GATE}, but
+no alternative for \textit{DICOM Viewer}. We might underestimate their rank due
+to uncommon technical issues.
+\end{itemize}
+
+\subsubsection{Threats to External Validity}
+\begin{itemize}
+\item We interviewed eight teams, which is a good proportion of the 29. However,
+there is still a risk that they might not well represent the whole MI software
+community.
+\item Our ranking gave all qualities equal weight, which may not be the case
+with all users. Thus, it may not represent the popularity of software among
+users.
+\item The number of GitHub stars, watches, and forks are not perfect measures of
+popularity, but they are what we had available.
+\end{itemize}
+
+\subsubsection{Threats to Conclusion Validity}
+\begin{itemize}
+\item We used the grading template in Appendix \ref{ap_grading_template} to
+guide our measurements. Our impressions of the software - such as user
+experience - were factors in deciding some scores. Thus, there is a risk that
+some scores may be subjective and biased.
+\end{itemize}
+
+\section{Recommendations} \label{ch_recommendations}
+
+This section presents our recommendations on SC software development. In
+general, our suggestions apply to all SC domains, unless we specifically mention
+that a particular guideline is only for MI software.
+
+Section \ref{sec_recommendations_qualities} discusses the actions that can
+potentially improve the ten software qualities. Sections
+\ref{sec_recommendations_limited_resources},
+\ref{sec_recommendations_tech_stack}, and
+\ref{sec_recommendations_testing_dataset} are based on the primary pain points
+collected from the developers in the MI domain, but we believe scientists and
+developers are likely to face them in most SC domains. These sections contain
+our general suggestions tackling them.
+
+\subsection{Recommendations on Improving Software Qualities} \label{sec_recommendations_qualities}
+
+Based on our quality measurements in Sections \ref{ch_results} and discussions
+with the developers in Sections \ref{sec_interview_software_qualities}, we
+collected many key points that may improve the software qualities. We list the
+primary ones by each quality as follows,
+\begin{itemize}
+\item \textbf{Installability} (Section \ref{sec_result_installability})
+\begin{itemize}
+    \item clear instructions;
+    \item automated installer;
+    \item including all dependencies in the installer;
+    \item avoiding heavily depending on other commercial products (e.g. Matlab);
+    \item considering building a web application that needs no installation.
+\end{itemize}
+\item \textbf{Correctness \& Verifiability} (Section \ref{sec_result_correctness_verifiability} and \ref{sec_interview_correctness})
+\begin{itemize}
+    \item test-driven development with unit tests, integration tests, and nightly tests;
+    \item two stage development process with stable release \& nightly builds;
+    \item CI/CD;
+    \item requirements specifications and theory manuals \citep{Smith2016} \citep{SmithAndLai2005}.
+    \item static code analysis tools (e.g. Lint and SonarQube)
+\end{itemize}
+\item \textbf{Reliability} (Section \ref{sec_result_reliability})
+\begin{itemize}
+    \item test-driven development with unit tests, integration tests, and nightly tests.
+    \item two stage development process with stable release \& nightly builds;
+    \item descriptive error messages.
+\end{itemize}
+\item \textbf{Robustness} (Section \ref{sec_result_robustness})
+\begin{itemize}
+    \item designing with exceptions and make the software failures graceful;
+    \item descriptive error messages.
+\end{itemize}
+\item \textbf{Usability} (Section \ref{sec_result_usability} and \ref{sec_interview_usability})
+\begin{itemize}
+    \item usability tests and interviews with end users;
+    \item adjusting according to users’ feedbacks;
+    \item getting started tutorials;
+    \item user manuals;
+    \item professional UX designs;
+    \item active supports to users.
+\end{itemize}
+\item \textbf{Maintainability} (Section \ref{sec_score_maintainability} and
+\ref{sec_interview_maintainability})
+\begin{itemize}
+    \item using GitHub;
+    \item modular approach with the design principle proposed by Parnas:
+    ``system details that are likely to change independently should be the
+    secrets of separate modules; the only assumptions that should appear in the
+    interfaces between modules are those that are considered unlikely to
+    change." \citep{ParnasEtAl2000}
+    \item documentation for developers: project plan, developer’s manual, and API documentation.
+\end{itemize}
+\item \textbf{Reusability} (Section \ref{sec_result_reusability})
+\begin{itemize}
+    \item modular approach;
+    \item API documentation;
+    \item tools that generate software documentation for developers (e.g.
+    Doxygen, Javadoc, and Sphinx).
+\end{itemize}
+\item \textbf{Understandability} (Section \ref{sec_result_understandability} and
+\ref{sec_interview_understandability})
+\begin{itemize}
+    \item modular approach;
+    \item good coding style: consistent indentation and formatting style;
+    consistent, distinctive, and meaningful code identifiers; keeping parameters
+    in the same order for all functions; avoiding hard-coded constants (other
+    than 0 and 1);
+    \item clear comments, indicating what is being done, not how;
+    \item description of used algorithms;
+    \item documentation of explicit requirements on coding standard;
+    \item communication between developers and users via GitHub issues, mailing lists, and forums.
+    \item graphical user interface.
+\end{itemize}
+\item \textbf{Visibility/Transparency} (Section \ref{sec_result_visibility_transparency})
+\begin{itemize}
+    \item documents for the development process, project status, development
+    environment, and release notes.
+\end{itemize}
+\item \textbf{Reproducibility} (Section \ref{sec_interview_reproducibility})
+\begin{itemize}
+    \item test-driven development with unit tests, integration tests, and nightly tests.
+    \item open-source;
+    \item making data and documentation available;
+    \item using open-source libraries.
+\end{itemize}
+\end{itemize}
+
+\subsection{Recommendations on Dealing With Limited Resources}
+\label{sec_recommendations_limited_resources}
+
+The limitation of resources has many faces. We regard the lack of fundings,
+time, and developers as representations of this problem.
+
+We summarize our discussion with the MI software developers in Section
+\ref{sec_pain_points_1} with the following recommendations,
+\begin{itemize}
+\item \textbf{Identify the root cause.} More fundings or developers may not
+solve the problem of lacking time. It is beneficial to identify the underlying
+obstacles to the team.
+
+\item \textbf{Maintain a good documentation.} Creating and updating
+documentation consumes time, but can save much more time in the long term. If
+the users and developers can find answers to their questions themselves, they
+are less likely to abuse the team's issue tracker.
+
+\item \textbf{Adopt time-saving tools.} A good CI/CD tool (e.g., GitHub Actions)
+saves time for building and deploying the product, and automated tests can work
+in the background while developers are focusing on other tasks.
+
+\item \textbf{Use test-driven development process.} Many people think writing
+test cases is less fun than building the functional code, but this is only true
+before we encounter the bugs. Identifying and fixing bugs can consume
+substantial resources. Setting up the test cases costs time, but generates more
+benefits in the long run.
+
+\item \textbf{Consider supporting third-party plugins or extensions.} Why not
+let users share the burden? No software product can deliver every user's needs,
+and the large quantity of features leads to more bugs and maintenance problems.
+So it may be a good idea to shift some development and maintenance
+responsibilities to the users. The users may also be happy about the extra
+flexibility.
+
+\item \textbf{Consider ``hibernating" for a while.} When developers are not
+enough, the team can shift from development mode toward maintenance mode for
+some time. Stop building new features, and instead fix bugs and design problems
+from the past. If the development team can repay some of its technical debt, the
+software qualities may improve as a result.
+
+\item \textbf{Commercialization is not always toxic.} Licensing the software to
+commercial companies to use as internal modules of their products may bring
+financial supports to the team. Meanwhile, the project can stay open-source for
+the community.
+\end{itemize}
+
+\subsection{Recommendations on Choosing A Tech Stack} \label{sec_recommendations_tech_stack}
+
+A tech stack refers to a set of technologies used by a team to build software
+and manage the project. Section \ref{sec_pain_points_2} lists the advantages and
+disadvantages between native and web applications. In this section, we give
+further suggestions on the choice of a tech stack to address the
+\textit{compatibility}, \textit{maintainability}, \textit{performance}, and
+\textit{security} of software.
+
+\begin{itemize}
+\item \textbf{Identify the priorities of the qualities.} It is hard to cover all
+aspects. Some teams achieve all four above qualities for their software, but it
+is not an easy task. Sections \ref{sec_pain_points_2} contains more details
+about the difficulty of balancing between the four qualities. A team needs to
+prioritize its objectives according to its resource and experience.
+
+\item \textbf{Be open-minded about new technologies.} Web applications with only
+a frontend are known for worse \textit{performance} than native applications.
+However, new technologies may ease this difference. For example, some JavaScript
+libraries can help the frontend harness the power of computer GPU and accelerate
+graphical computing. In addition, there are new frameworks helping developers
+with cross-platform \textit{compatibility}. For example, the
+\hyperlink{https://flutter.dev/}{Flutter} project enables support for web,
+mobile, and desktop OS with one codebase.
+
+\item \textbf{Use git and GitHub.} As mentioned in Sections
+\ref{sec_score_maintainability}, almost all of the 29 MI software projects used
+git, and the majority of them used GitHub. We found from the projects' websites
+and our interviews with developers that, some projects moved from other version
+control tools to git and GitHub. GitHub provides convenient repository and
+project management, and OSS projects receive more attention and contribution on
+GitHub.
+
+\item \textbf{Web applications can also deliver high performance.} Web
+applications with backend servers may perform even better than native
+applications. If a team needs to support lower-end computers, it is good to use
+back-end servers for heavy computing tasks.
+
+\item \textbf{Backend servers can have low costs.} It is worth exploring the
+serverless solutions from major cloud service providers. Serverless still uses a
+server, but the team is only charged when they use it. The solution is
+event-driven, and costs the team by the number of requests it processes. Thus,
+serverless can be very cost-effective for the less intensively used functions.
+
+\item \textbf{Web transmission may diminish security.} Transferring sensitive
+data online can be a problem for projects requiring high security. Regulations
+in some SC domains may forbid doing so. In this case, a web application with a
+backend may not be a good choice.
+
+\item \textbf{Maintain a good documentation.} No matter what tech stack a team
+uses, a well-maintained project plan, developer's manual, and API documentation
+always help team members to contribute more and make fewer mistakes.
+\end{itemize}
+
+\subsection{Recommendations on Enriching the Testing Datasets} \label{sec_recommendations_testing_dataset}
+
+As described in Section \ref{sec_interview_pain_points}, it was difficult for
+some software development teams in the MI domain to access real-world medical
+imaging datasets. This problem restricted their capability and flexibility to
+test their software. We believe software developers in other SC domains may also
+face similar issues.
+
+Based on Section \ref{sec_pain_points_3}, we provide some suggestions as follows,
+\begin{itemize}
+\item \textbf{Build and maintain good connections to datasets.} A team can build
+connections with professionals working in the SC domain, who may have access to
+private datasets and perform tests for the team. Moreover, if a team has such
+professionals as internal members, the process can be even simpler.
+
+\item \textbf{Collect and maintain datasets over time.} A team may face all
+kinds of strange problems caused by various unique inputs over the years of
+development. It is worth collecting and maintaining this data, which can form a
+good dataset for testing.
+
+\item \textbf{Search for open data sources.} In general, there are many open
+datasets in different SC domains. Take MI as an example, there are Chest X-ray
+Datasets by National Institute of Health
+(\hyperlink{https://nihcc.app.box.com/v/ChestXray-NIHCC}{https://nihcc.app.box.com/v/ChestXray-NIHCC})
+\citep{WangEtAl2017}, Cancer Imaging Archive
+(\hyperlink{https://www.cancerimagingarchive.net/}{https://www.cancerimagingarchive.net/})
+\citep{PriorEtAl2017}, and MedPix by National Library of Medicine
+(\hyperlink{https://medpix.nlm.nih.gov/home}{https://medpix.nlm.nih.gov/home})
+\citep{Smirniotopoulos2014}. A team developing MI software should be able to find
+more open datasets according to their needs.
+
+\item \textbf{Create sample data for testing.} If a team can access tools
+creating sample data, they may also self-build datasets for testing. For
+example, an MI software development team can use an MRI scanner to create images
+of objects, animals, and volunteers. The team can build the images based on
+specific testing requirements.
+
+\item \textbf{Remove privacy from sensitive data.} For data with sensitive
+information, a team can ask the data owner to remove such information or add
+noise to protect privacy. One example is using deidentified copies of medical
+images for testing.
+
+\item \textbf{Establish community collaboration in the domain.} During our
+interviews with developers in the MI domain, we heard many stories of asking for
+supports from other professionals or equipment manufacturers. However, we
+believe that broader collaboration between development teams can address this
+problem better. Some datasets are too sensitive to share, but if the community
+has some kind of ``group discussion", teams can better express their needs, and
+professionals can better offer voluntary support for testing. Ultimately, the
+community can establish a nonprofit organization as a third-party, which
+maintains large datasets, tests OSS in the domain, and protects privacy. 
+% look into the database that the Eng Phys guest speaker mentioned?
+\end{itemize}
+
+\section{Conclusions} \label{ch_conclusions}
+
+We analyzed the state of the practice for SC software in the MI domain. To
+better achieve our goals in Section \ref{ch_intro}, we proposed six research
+questions in Section \ref{sec_research_questions}.
+
+Our methods in Section \ref{ch_methods} form a general process to evaluate
+domain-specific software, that we apply on specific SC domains. As mentioned in
+Section \ref{sec_applying_method}, following this process, we chose the MI
+domain, identified 48 SC software candidates in it, then selected 29 of them to
+our final list. Section \ref{ch_results} lists our measurements to nine software
+qualities for the 29 projects, and Section \ref{ch_interview} contains our
+interviews with eight of the 29 teams, discussing their development process and
+five software qualities.
+
+We answered our six research questions in Section \ref{ch_answers}. In addition,
+Section \ref{ch_recommendations} presents our recommendations on SC software
+development.
+
+\subsection{Key Findings}
+
+With the measurement results in Section \ref{ch_results}, we revealed some
+current status of SC software development and qualities in the MI domain. We
+ranked the 29 software projects in nine qualities based on the grading scores.
+\textit{3D Slicer}, \textit{ImageJ}, and \textit{OHIF Viewer} are the top three
+software by their overall scores.
+
+The interview results in Section \ref{ch_interview} show some merits, drawbacks,
+and pain points within the development process. The three primary categories of
+pain points are:
+\begin{itemize}
+\item the lack of fundings and time;
+\item the difficulty to balance between four factors: cross-platform
+compatibility, convenience to development \& maintenance, performance, and
+security;
+\item the lack of access to real-world datasets for testing.
+\end{itemize}
+We summarized the solutions from the developers to address these problems. We
+also collected the status of software testing, documentation, contribution
+management, and project management in the eight projects.
+
+Our answers to the research questions (Section \ref{ch_answers}) are based on
+the above findings. We identified the existing artifacts, tools, principles,
+processes, and methodologies in the 29 projects. By comparisons in Section
+\ref{sec_rq_comparison}, we found out: 1) four of the top five software projects
+in our ranking were also among the top five ones receiving the most GitHub stars
+per year (Table \ref{tab_ranking_vs_GitHub}); 2) three of the top four in our
+ranking were among the top four provided by the domain experts (Table
+\ref{tab_top_software_vs_experts}).
+
+Section \ref{ch_recommendations} presents our recommendations on improving
+software qualities and easing pain points during development. Some highlighted
+ones are:
+\begin{itemize}
+\item adopting test-driven development with unit tests, integration tests, and
+nightly tests;
+\item maintaining good documentation (e.g., installation instructions,
+requirements specifications, theory manuals, getting started tutorials, user
+manuals, project plan, developer’s manual, API documentation, requirements on
+coding standards, development process, project status, development environment,
+and release notes);
+\item using CI/CD;
+\item using git and GitHub;
+\item modular approach with the design principle proposed by Parnas
+\citep{ParnasEtAl2000};
+\item considering newer technologies (e.g., web application and serverless
+solution);
+\item various ways of enriching the testing datasets in Section
+\ref{sec_recommendations_testing_dataset}.
+\end{itemize}
+
+\subsection{Future Works}
+
+With learnings from this project, we summarized recommendations for the future
+state of the practice assessments:
+\begin{itemize}
+    \item we can make the surface measurements less shallow. For example:
+    \begin{itemize}
+        \item \textit{surface reliability}: our current measurement relies on
+        the processes of installation and getting started tutorials. However,
+        not all software needs installation or has a getting started tutorial.
+        We can design a list of operation steps, perform the same operations
+        with each software, and record any errors.
+        \item \textit{surface robustness}: we used damaged images as inputs for
+        this measuring MI software. This process is similar to fuzz testing
+        \citep{enwiki:1039424308}, which is one type of fault injection
+        \citep{enwiki:1039005082}. We may adopt more fault injection methods, and
+        identify tools and libraries to automate this process.
+        \item \textit{surface usability}: we can design usability tests and test
+        all software projects with end-users. The end-users can be volunteers
+        and domain experts.
+        \item \textit{surface understandability}: our current method does not
+        require understanding the source code. As software engineers, perhaps we
+        can select a small module of each project, read the source code and
+        documentation, try to understand the logic, and score the ease of the
+        process.
+    \end{itemize}
+	\item we can further automate the measurements on the grading temple in
+	Appendix \ref{ap_grading_template}. For example, with automation scripts and
+	the GitHub API, we may save significant time on retrieving the GitHub
+	metrics. We have started to build a tool for this purpose, with its source
+	code at this repository
+	\hyperlink{https://github.com/smiths/AIMSS/tree/master/StateOfPractice/GitHubMetricsCollector}{https://github.com/smiths/AIMSS/.../GitHubMetricsCollector}.
+	This GitHub Metrics Collector can take GitHub repository links as input,
+	automatically collect metrics from the GitHub API, and record the results.
+	We can improve and use this tool in our future projects;
+	\item the grading standard can be more explicit. For example, we can
+	explicitly define scores for each item in the grading temple.
+	\item we can improve some interview questions. Some examples are:
+	\begin{itemize}
+	    \item in \hyperlink{q14}{Q14}, ``Do you think improving this process can
+	    tackle the current problem?" is a yes-or-no question, which is not
+	    informative enough. As mentioned in Section \ref{sec_contribution_pm},
+	    most interviewees ignored it. We can change it to ``By improving this
+	    process, what current problems can be tackled?";
+	    \item in \hyperlink{q16}{Q16}, we can ask more details about the modular
+	    approach, such as "What principles did you use to divide code into
+	    modules? Can you give an example of using the principles?";
+	    \item \hyperlink{q17}{Q17} and  \hyperlink{q18}{Q18} should respectively
+	    ask \textit{understandability} to developers and \textit{usability} to
+	    end-users.
+	\end{itemize}
+	\item we can better organize the interview questions. Since we use audio
+	conversion tools to transcribe the answers, we should make the transcription
+	easier to read. For example, we can order them together for questions about
+	the five software qualities and compose a similar structure for each.
+	\item we can mark the follow-up interview questions with keywords. For
+	example, say ``this is a follow-up question" every time asking one. Thus, we
+	record this sentence in the transcription, and it will be much easier to
+	distinguish the follow-up questions from the 20 designed questions.
+\end{itemize}
+
+\noindent In addition, we propose a few SC domains that are potentially suitable
+for future works:
+\begin{itemize}
+    \item Metallurgy
+    \item Quantitative Finance
+    \item Computational Fluid Dynamics
+    \item Basic Linear Algebra
+    \item Finite Elements
+    \item Sparse Linear Solvers
+\end{itemize}
+
+\noindent After applying our method on various domains, we can start a
+meta-study to compare the state of the practice for software in different
+domains.
 
@@
 
-\begin{thebibliography}{00}
+%% Bibliography
+\bibliographystyle{ACM-Reference-Format}
+\bibliography{MedImageSoft_SOP}
 
-%% \bibitem[Author(year)]{label}
-%% Text of bibliographic item
+% \begin{thebibliography}{00}
 
-\bibitem[ ()]{}
+% %% \bibitem[Author(year)]{label}
+% %% Text of bibliographic item
 
-\end{thebibliography}
+% \bibitem[ ()]{}
+
+% \end{thebibliography}
+
 \end{document}
@@
 %% End of file `elsarticle-template-harv.tex'.
+
+%%%%%%%%%%%
+
+% Overview of scientific computing
+We define Scientific Computing (SC) as ``the use of computer tools to analyze or
+simulate mathematical models of real world systems of engineering or scientific
+importance so that we can better understand and predict the system’s behaviour"
+\citep{Smith2006}. Many researchers consider SC as the third pillar of science
+and engineering, along with theory and experiment \citep{Landau2005}. Almost all
+areas in science and engineering use computers for modeling \citep{Golub2014},
+and software plays an essential role in modern scientific research
+\citep{Hannay2009, Wilson2014}.  Software development in SC depends on three
+fields of knowledge: engineering or scientific domain knowledge, mathematical
+algorithm knowledge, and computational algorithm knowledge \citep{Landau2005,
+Mehta2015}. Thus, most SC software developers are scientists in SC domains
+\citep{Wilson2014}. However, they do not always use modern software development
+techniques, tools, and methods \citep{Wilson2014}. Therefore, we developed a
+methodology for assessing the state of the practice for SC software
+\citep{SmithEtAl2021}. We apply this process to Medical Imaging (MI) software
+that belongs to a specific domain of SC.

### 2021-11-17 StateOfPractice/Papers/MedImageSoft_SOP/MedImageSoft_SOP_AcademicRadiology.tex
Updates for Med Image SOP up to section 2 (background)

@@
 
-\documentclass[preprint,12pt,authoryear]{elsarticle}
+%\documentclass[preprint,12pt,authoryear]{elsarticle}
+\documentclass[3p, 12pt,authoryear]{elsarticle}
 
@@
 %% The amsthm package provides extended theorem environments
-%% \usepackage{amsthm}
+\usepackage{amsmath,amsthm}
 
@@
 \usepackage{hyperref}
-\usepackage{tikz}
-\usetikzlibrary{mindmap}
+%\usepackage{tikz}
+%\usetikzlibrary{mindmap}
 \usepackage{multirow}
@@
 
-\pgfplotsset{compat=1.13}
+%\pgfplotsset{compat=1.13}
 
@@
 To gain insights, we devised five research questions, which can be applied
-to MI, and to other domains, of research software \citep{SmithEtAl2021}:
+to MI, as well as to other domains, of research software \citep{SmithEtAl2021}:
 
@@
 software qualities for the projects? What actions have the developers taken to
-address them?
+address them?  (Section~\ref{ch_background} provides definitions of the software
+qualities.)
 \begin{itemize}
@@
 
-\subsection{Overview of the Methodology}
+\subsection{Scope} \label{sec_scope}
+
+We need to restrict the scope of MI software that we consider so that the
+measurement exercise is feasible within the targeted person month of effort. We
+can restrict the scope by selecting a sub-set of the broad range of problems
+considered by MI.  Summing the sets of MI tasks from \citet{Bankman2000} and
+\citet{Angenent2006}, we have the following problems are solved by MI software:
+Segmentation, Registration, Visualization, Enhancement, Quantification,
+Simulation, plus three functions for MI archiving and telemedicine systems
+(Compression, Storage, and Communication). To this list Wikipedia contributors
+\citep{enwiki:1034877594} add Statistical Analysis and Image-based Physiological
+Modelling. \citet{Kim2011} adds Feature Extraction, Classification, and
+Interpretation. Besides the above major functions, some MI software provides
+supportive functions. For example, using Tool Kit libraries VTK
+\citep{SchroederEtAl2006} and ITK \citep{McCormick2014}, developers build
+software with Visualization and Analysis functions. Finally, Picture Archiving
+and Communication System (PACS) helps users to economically store and
+conveniently access images \citep{Choplin1992}.
+
+Based on our literature survey, we divided MI software into five sub-groups and
+several sub-sub-groups by their major functions, as shown in
+Figure~\ref{fig_mi_functions}. We selected the sub-group of Visualization for
+the current state of the practice exercise.
+
+\begin{figure}[ht]
+\centering
+\includegraphics[scale=0.6]{figures/FunctionsOfMiSoftware.pdf}
+\caption{Major functions of MI Software} \label{fig_mi_functions}
+\end{figure}
+
+% \begin{figure}[ht]
+% \centering
+% \begin{tikzpicture}[mindmap, grow cyclic, every node/.style=concept, concept
+% color=orange!40,
+% level 1/.append style={level distance=5cm,sibling angle=72},
+% level 2/.append style={level distance=3cm,sibling angle=90},
+% level 3/.style={level distance=2.3cm,sibling angle=90}]
+% \node{MI software}
+% child[concept color=teal!40] { node {Enhancement}}
+% child[concept color=teal!40] { node {Analysis}
+%     child[concept color=blue!30] { node {Registration}}
+%     child[concept color=blue!30,] { node {Segmentation}}
+%     child[concept color=blue!30] { node {Statistical Analysis}
+%     child[concept color=blue!20] { node {Feature Extraction}}
+%     child[concept color=blue!20] { node {Classification}}
+%     child[concept color=blue!20] { node {Interpretation}}
+%     }}
+% child[concept color=teal!40] { node {Simulation/\\Modeling}}
+% child[concept color=teal!40] { node {Supporting}
+%     child[concept color=blue!30] { node {Tool Kit}}
+%     child[concept color=blue!30] { node {PACS}}
+% }
+% child[concept color=teal!40] { node {Visualization}
+%     child[concept color=blue!30] { node {2D Display}}
+%     child[concept color=blue!30] { node {3D Rendering}}
+%     child[concept color=blue!30] { node {Reformatted Views}}
+% };
+% \end{tikzpicture}
+% \caption{Major functions of MI software}
+% \label{fig_mi_functions}
+% \end{figure}
+
+\subsection{Methodology Overview}
 
 We designed a general methodology to assess the state of the practice for SC
-software \citep{SmithEtAl2021}. Details can be found in Section~\ref{}.  This
-methodology builds off prior work to assess the state of the practice for such
-domains as Geographic Information Systems \citep{smith2018state}, Mesh
-Generators \citep{smith2016state}, Seismology software
-\citep{Smith2018Seismology}, and Statistical software for psychology
+software \citep{SmithEtAl2021}. Details can be found in
+Section~\ref{ch_methods}.  This methodology builds off prior work to assess the
+state of the practice for such domains as Geographic Information Systems
+\citep{smith2018state}, Mesh Generators \citep{smith2016state}, Seismology
+software \citep{Smith2018Seismology}, and Statistical software for psychology
 \citep{smith2018statistical}.  In keeping with the previous methodology, we have
 maintained the constraint that the work load for measuring a given domain should
-be feasible for a team as small as one person, and for a short time,
-ideally around a person month of effort. A person month is considered to be $20$
-working days ($4$ weeks in a month, with $5$ days of work per week) at $8$
-person hours per day, or $20 \cdot 8 = 160$ person hours.
+be feasible for a team as small as one person, and for a short time, ideally
+around a person month of effort. A person month is considered to be $20$ working
+days ($4$ weeks in a month, with $5$ days of work per week) at $8$ person hours
+per day, or $20 \times 8 = 160$ person hours.
 
 With our methodology, we first choose an SC domain (in the current case MI) and
-identify a list of about 30 software packages. We approximately measure the
-qualities of each package by filling in a grading template (forward ref?).
-Compared with our previous methodology, the new methodology also includes
-repository based metrics, such as the number of files, number of lines of code,
-percentage of issues that are closed, etc. (forward ref?).  With the
+identify a list of about 30 software packages. (For measuring MI we used 29
+software packages.)  We approximately measure the qualities of each package by
+filling in a grading template. Compared with our previous methodology, the new
+methodology also includes repository based metrics, such as the number of files,
+number of lines of code, percentage of issues that are closed, etc.  With the
 quantitative data in the grading template, we rank the software with the
-Analytic Hierarchy Process (AHP) (forward ref?). After this, as another addition
-to our previous methodology, we interview some of the development teams to
-further understand the status of their development process (forward ref?).
-Finally, we summarize the results and propose recommendations for future SC
-software development (forward ref?).
-
-\subsection{Scope} \label{sec_scope}
-
-We need to restrict the scope of MI software that we consider so that the
-measurement exercise is feasible within a person month of effort. MI considers a
-broad range of different problems.  According to \citet{Bankman2000}, MI
-software deals with six different basic problems, while \citet{Angenent2006}
-pointed out that four fundamental problems are solved by MI software. While both
-mentioned Segmentation, Registration, and Visualization of medical images,
-Bankman also included Enhancement, Quantification, and three functions for MI
-archiving and telemedicine systems (Compression, Storage, and Communication)
-\citep{Bankman2000}. On the other hand, \citet{Angenent2006} included
-Simulation. According to Wikipedia contributors \citep{enwiki:1034877594}, MI
-software has primary functions in Segmentation, Registration, Visualization
-(including the basic display, reformatted views, and 3D volume rendering),
-Statistical Analysis, and Image-based Physiological Modelling. As
-\citet{Kim2011} describes, the general steps of medical image analysis after
-obtaining digital data include Enhancement, Segmentation, Feature Extraction,
-Classification, and Interpretation. Besides the above major functions, some MI
-software provides supportive functions. For example, with Tool Kit libraries VTK
-\citep{SchroederEtAl2006} and ITK \citep{McCormick2014}, developers build
-software with Visualization and Analysis functions; Picture Archiving and
-Communication System (PACS) helps users to economically store and conveniently
-access images \citep{Choplin1992}.
-
-Based on our literature survey, we divided MI software into five sub-groups and
-several sub-sub-groups by their major functions, as shown in Figure
-\ref{fig_mi_functions}.
-
-\begin{figure}[ht]
-\centering
-\begin{tikzpicture}[mindmap, grow cyclic, every node/.style=concept, concept
-color=orange!40,
-level 1/.append style={level distance=5cm,sibling angle=72},
-level 2/.append style={level distance=3cm,sibling angle=90},
-level 3/.style={level distance=2.3cm,sibling angle=90}]
-\node{MI software}
-child[concept color=teal!40] { node {Enhancement}}
-child[concept color=teal!40] { node {Analysis}
-    child[concept color=blue!30] { node {Registration}}
-    child[concept color=blue!30,] { node {Segmentation}}
-    child[concept color=blue!30] { node {Statistical Analysis}
-    child[concept color=blue!20] { node {Feature Extraction}}
-    child[concept color=blue!20] { node {Classification}}
-    child[concept color=blue!20] { node {Interpretation}}
-    }}
-child[concept color=teal!40] { node {Simulation/\\Modeling}}
-child[concept color=teal!40] { node {Supporting}
-    child[concept color=blue!30] { node {Tool Kit}}
-    child[concept color=blue!30] { node {PACS}}
-}
-child[concept color=teal!40] { node {Visualization}
-    child[concept color=blue!30] { node {2D Display}}
-    child[concept color=blue!30] { node {3D Rendering}}
-    child[concept color=blue!30] { node {Reformatted Views}}
-};
-\end{tikzpicture}
-\caption{Major functions of MI software}
-\label{fig_mi_functions}
-\end{figure}
-
-To keep the data collection and analysis feasible, we limited the scope of the
-software to the software packages providing the Visualization tools and
-functions in this project.
-
-[Roadmap - instead sprinkle forward references throughout the introduction?]
+Analytic Hierarchy Process (AHP) (Details are found in
+Section~\ref{ch_background}). After this, as another addition to our previous
+methodology, we interview some of the development teams to further understand
+the status of their development process. Finally, we summarize the measurements
+(Section~\ref{ch_results}), summarize the interview answers
+(Section~\ref{ch_interview}), answer the research questions
+(Section~\ref{ch_answers}) and propose recommendations
+(Section~\ref{ch_recommendations}) for future SC software development.
 
@@
 
-When designing a method for evaluating the state of the practice of
-domain-specific software, we included a step to select domain and software.
-Knowledge of different software categories is essential for the selection. To
-compare and rank the software qualities with the grading template in Appendix
-\ref{ap_grading_template}, we need the definitions of qualities and the AHP.
-
 In this section, we introduce the relevant software categories (Section
-\ref{sec_software_categories}). We also cover the software quality definitions
-(Section \ref{sec_software_quality}) and an overview of the AHP (Section
-\ref{sec_AHP}).
+\ref{sec_software_categories}). The categories will be used when collecting and
+assessing the software packages.  We also review the definitions of software
+qualities (Section \ref{sec_software_quality}) and provide an overview of the
+AHP (Section \ref{sec_AHP}).
 
@@
 
-We target specific software categories to narrow down the scope when selecting
-software packages for measuring. One way to categorize them is grouping by
-functions, such as analyzing the scope for the MI software (Section
-\ref{sec_scope}), which requires specific domain knowledge. In this section, we
-discuss three common software categories that we can apply to all SC domains:
-Open source software (OSS), freeware, and commercial software.
-
-\subsubsection{Open Source Software} \label{sec_open_source_software}
-
-For OSS, the source code is openly accessible. Users have the right to study,
-change and distribute it under a license granted by the copyright holder. For
-many OSS projects, the development process relies on the collaboration of
-different contributors worldwide \citep{Corbly2014}. Accessible source code
-usually exposes more ``secrets'' of a software project, such as the underlying
-logic of software functions, how developers achieve their works, and the flaws
-and potential risks in the final product. Thus, it brings much more convenience
-to researchers analyzing the qualities of the project.
-
-\subsubsection{Freeware} \label{sec_freeware}
-
-Freeware is software that can be used free of charge. Unlike with OSS, the
-authors of freeware do not allow users to access or modify the source code of
-the software \citep{LINFO2006}. The term \textit{freeware} should not be confused
-with \textit{free software}, which is similar to OSS. To the end-users, the
-differences between freeware and OSS often do not bother them. The fact that
-these products are free of charge is likely to make them popular with many
-users. However, software developers, end-users who wish to modify the source
-code, and researchers looking for insight into software development process will
-find the inaccessible source code a problem.
-
-\subsubsection{Commercial Software}
-
-``Commercial software is software developed by a business as part of its
-business'' \citep{GNU2019}. Typically speaking, the users are required to pay to
-access all of the features of commercial software, excluding access to the
-source code. However, some commercial software is also free of charge
-\citep{GNU2019}. Based on our experience, most commercial software products are
-not OSS.
-
-For some specific software, the backgrounds of commercial software developers
-often differ from the ones of non-commercial OSS. In such a case, the former is
-usually the product of software engineers, and the latter is likely to have
-developers who work in the domain and are also end-users of the products. One
-example of such software is SC software, since the developers need to utilize
-their domain-specific during the development process \citep{Wilson2014}.
+When assessing software packages, we need to know what license the software is
+distributed under.  In particular, we need to know whether the source code will
+be available to us or not.  In this section, we present three common software
+categories:
+
+\begin{itemize}
+
+\item \textbf{Open Source Software (OSS)} For OSS, the source code is openly
+accessible. Users have the right to study, change and distribute it under a
+license granted by the copyright holder. For many OSS projects, the development
+process relies on the collaboration of different contributors worldwide
+\citep{Corbly2014}. Accessible source code usually exposes more ``secrets'' of a
+software project, such as the underlying logic of software functions, how
+developers achieve their works, and the flaws and potential risks in the final
+product. Thus, it brings much more convenience to researchers analyzing the
+qualities of the project.
+
+\item \textbf{Freeware} Freeware is software that can be used free of charge.
+Unlike with OSS, the authors of do not allow users to access or modify the
+source code \citep{LINFO2006}. To many end-users, the differences between
+freeware and OSS may not be relevant. However, software developers, end-users
+who wish to modify the source code, and researchers looking for insight into
+software development process may find the inaccessible source code a problem.
+
+\item \textbf{Commercial Software} ``Commercial software is software developed
+by a business as part of its business'' \citep{GNU2019}. Typically speaking, the
+users are required to pay to access all of the features of commercial software,
+excluding access to the source code. However, some commercial software is also
+free of charge \citep{GNU2019}. Based on our experience, most commercial
+software products are not OSS.
+
+\end{itemize}
 
@@
 
-This section lists the definitions of 10 software qualities, which are from
-Smith et al. \citep{SmithEtAl2020}. We aim to measure each of them for selected
-SC software packages. The order of the first nine qualities follows our grading
-template in Appendix \ref{ap_grading_template}. We do not measure
-\textit{reproducibility} with the grading template, but discuss it with the
-developers by interviews.
-
-\begin{itemize}
-	\item \textbf{Installability} The effort required for the installation,
-	uninstallation, or reinstallation of a software or product in a specified
-	environment \citep{ISO/IEC25010} \citep{lenhard2013measuring}.
+Quality is defined as a measure of the excellence or worth of an entity.  As is
+common practice, we do not think of quality as a single measure, but rather as a
+set of measures.  That is, quality is a collection of different qualities, often
+called ``ilities.''  Below we list the 10 qualities of interest for this study,
+as summarized in \citet{SmithEtAl2021-QDef}.  The order of the qualities follows
+the order used in \cite{GhezziEtAl2003}, which puts related qualities (like
+correctness and reliability) together.  Moreover, the order is roughly the same
+as the order qualities are considered in practice.  [*** To be completed - not
+currently in sync with QDef paper, QDef paper not completely done ***]
+
+\begin{itemize}
+	\item \textbf{Installability} The effort required for the installation
+    and/or uninstallation of software in a specified environment.
 
 	\item \textbf{Correctness \& Verifiability} A program is correct if it
-	behaves according to its stated specifications. Verifiability is the extent
-	to which a set of tests can be written and executed, to demonstrate that the
-	delivered system meets the specification \citep{GhezziEtAl2003}.
+    matches its specification.  The specification can either be explicitly or
+    implicitly stated.  Verifiability is the extent to which the software
+    components or the integrated product can be evaluated to demonstrate whether
+    the system functions as expected.
 
@@
 
-To generate ranking scores for a set of software packages, we use AHP, which
-utilizes pairwise comparisons between all of the packages. Thomas L. Saaty
-developed this tool, and people widely used it to make and analyze multiple
-criteria decisions \citep{VaidyaEtAl2006}. AHP organizes multiple criteria
-factors in a hierarchical structure and uses pairwise comparisons between
-alternatives to calculate relative ratios \citep{Saaty1990}.
-
-For a project with $ m $ criteria, we can use an $m\times m$ matrix $A$ to
-record the relative importance between factors. When comparing criterion $i$ and
-criterion $j$, the value of $A_{ij}$ is decided as follows, and the value of
-$A_{ji}$ is $1/A_{ij}$ \citep{Saaty1990},
-
-\begin{itemize}
-	\item $A_{ij} = 1$ if criterion $i$ and criterion $j$ are equally important;
-	\item $A_{ij} = 9$ if criterion $i$ is extremely more important than
-	criterion $j$;
-	\item $A_{ij}$ equals to an integer value between 1 and 9 according the the
-	relative importance of criterion $i$ and criterion $j$.
-\end{itemize}
-
-The above process assumes that criterion $i$ is not less important than
-criterion $j$, otherwise, we need to reverse $i$ and $j$ and determine $A_{ji}$
-first, then $A_{ij} = 1/A_{ji}$.
-
-The priority vector $w$ can be calculated by solving the following equation
-\citep{Saaty1990}, 
+AHP is used to generate a ranking for a set of software packages. Thomas L.\
+Saaty developed AHP in the 1970s, and people have widely used it since to make
+and analyze multiple criteria decisions \citep{VaidyaEtAl2006}. AHP organizes
+multiple criteria factors in a hierarchical structure and uses pairwise
+comparisons between alternatives to calculate relative ratios \citep{Saaty1990}.
+
+For a project with $m$ criteria, we can use an $m\times m$ matrix $A$ to record
+the relative importance between factors. When comparing criterion $i$ and
+criterion $j$, the value of $A_{ij}$ is decided as follows, with the value of
+$A_{ji}$ generally as $1/A_{ij}$ \citep{Saaty1990}.  $A_{ij} = 1$ if criterion
+$i$ and criterion $j$ are equally important, while $A_{ij} = 9$ if criterion $i$
+is extremely more important than criterion $j$.  The natural numbers between 1
+and 9 are used to show the different levels of relative importance between these
+two extremes. The above assumes that criterion $i$ is not less important than
+criterion $j$.  If that is not the case, we reverse $i$ and $j$ and determine
+$A_{ji}$ first, then $A_{ij} = 1/A_{ji}$.
+
+The priority vector $w$, which ranks the criteria by their importance, can be
+calculated by solving the equation \citep{Saaty1990}:
 \begin{equation} 
-    Aw = \lambda_{max}w,
+    A w = \lambda_{max} w,
 \end{equation}
-where $\lambda_{max}$ is the maximal eigenvalue of $A$.
-
-In this project, $w$ is approximated with the classic \textit{mean of normalized
-values} approach \citep{AlessioEtAl2006},
+where $\lambda_{max}$ is the maximal eigenvalue of $A$.  In this project, $w$ is
+approximated with the classic \textit{mean of normalized values} approach
+\citep{AlessioEtAl2006}:
 
@@
 
-Suppose there are $n$ alternatives, for criterion $i = 1, 2, ... , m$, we can
-create an $n\times n$ matrix $B_i$ to record the relative preferences between
-these choices. The way of generating $B_i$ is similar to the one for $A$.
-However, unlike comparing the importance between criteria, we pairwise decide
-how much we favor one alternative over the other. We use the same method to
-calculate the local priority vector for each $B_i$.
+If there are $n$ alternatives, for criterion $i = 1, 2, ... , m$, we can create
+an $n\times n$ matrix $B_i$ to record the relative preferences between these
+choices for each criterion. The way of generating $B_i$ is similar to the one
+for $A$. However, rather than comparing the importance between criteria, we
+pairwise decide how much we favor one alternative over the other. We use the
+same method to calculate the local priority vector for each $B_i$.  The local
+priority vector in this case ranks the $n$ alternatives for criterion $i$.
 
@@
 packages ($n = 29$) are compared for each of the $m$ criteria. The software are
-evaluated with the grading template in Appendix \ref{ap_grading_template} and a
-subjective score from one to ten is given for each quality for each package. For
-each quality, for a pair of packages $i$ and $j$, such that $score_i >=
-score_j$, the difference between two scores is $\mathit{diff_{ij}} = score_i -
-score_j$. The relationship between $A_{ij} = 1$ and $\mathit{diff_{ij}}$ is as
-follows:
+evaluated with the grading template in Section~\ref{sec_grading_template}, which
+includes a subjective score from $1$ to $10$ for each quality for each package.
+For each quality, for a pair of packages $i$ and $j$, with $\mathit{score}_i >=
+\mathit{score}_j$, the difference between the two scores is $\mathit{diff_{ij}}
+= \mathit{score}_i - \mathit{score}_j$. The mapping between $\mathit{diff_{ij}}$
+(which can vary between 0 and 9) and the values in $A_{ij}$ (which can vary
+between 1 and 9) is as follows:
 
@@
 
-Thus, we approximate the pairwise comparison result of $i$ versus $j$ by the
-following equation:
+\noindent Thus, we approximate the pairwise comparison result of $i$ versus $j$
+by the following equation:
 
 \begin{equation}
-A_{ij} = min(score_i - score_j + 1, 9)
+A_{ij} = \text{min}(\mathit{score}_i - \mathit{score}_j + 1, 9)
 \end{equation}
@@
 \ref{sec_empirical_measurements}). Some technical details for the measurements
-are in Section \ref{sec_technical_details}.
+are in Section \ref{sec_technical_details}.  We do not measure
+\textit{reproducibility} with the grading template, but discuss it with the
+developers by interviews.
 
@@
 \label{sec_applying_method_grading}
-Then we followed the steps in Section \ref{sec_grading_software} to measure and grade the software.
-27 out of the 29 packages are compatible with two or three different OS such as Windows, macOS, and Linux, and 5 of them are browser-based, making them platform-independent.
-However, in the interest of time, we only performed the measurements for each project by installing it on one of the platforms, most likely Windows.
+Then we followed the steps in Section \ref{sec_grading_software} to measure and
+grade the software. 27 out of the 29 packages are compatible with two or three
+different OS such as Windows, macOS, and Linux, and 5 of them are browser-based,
+making them platform-independent. However, in the interest of time, we only
+performed the measurements for each project by installing it on one of the
+platforms, most likely Windows.
 

### 2021-11-17 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
Updates for Med Image SOP up to section 2 (background)

@@
 
-\wss{When I get the Ghezzi text back from Olu, I'll check to see if they have anything to add to this definition.}
+\wss{When I get the Ghezzi text back from Olu, I'll check to see if they have 
+anything to add to this definition.}
 
@@
 \begin{mybox}
-\subsubsection*{Verifiability} Definition~\ref{Defn_Verifiability1}
+\subsubsection*{Verifiability} The extent to which software components or the
+integrated product can be evaluated to demonstrate whether the system functions
+as expected \citep{wiegers2003softreq}.
+
+Definition~\ref{Defn_Verifiability1}
 rephrased: Verifiability is the extent to which a set of tests can be

### 2021-11-18 StateOfPractice/Papers/MedImageSoft_SOP/MedImageSoft_SOP_AcademicRadiology.tex
Udpates to SOP for MI up to Section 3.2 (List of Representative Software)

@@
 \usepackage{arydshln}
+\usepackage{paralist}
+
+%% Comments
+\newif\ifcomments\commentstrue
+
+\ifcomments
+\newcommand{\authornote}[3]{\textcolor{#1}{[#3 ---#2]}}
+\newcommand{\todo}[1]{\textcolor{red}{[TODO: #1]}}
+\else
+\newcommand{\authornote}[3]{}
+\newcommand{\todo}[1]{}
+\fi
+
+\newcommand{\wss}[1]{\authornote{blue}{SS}{#1}} %Spencer Smith
+\newcommand{\jc}[1]{\authornote{red}{JC}{#1}} %Jacques Carette
+\newcommand{\mn}[1]{\authornote{magenta}{MN}{#1}} %Mike Noseworthy
+\newcommand{\ad}[1]{\authornote{cyan}{AD}{#1}} %Ao Dong
+
+\newcommand{\notdone}[1]{\textcolor{red}{#1}}
+\newcommand{\done}[1]{\textcolor{black}{#1}}
 
@@
 We designed a general methodology to assess the state of the practice for SC
-software \citep{SmithEtAl2021}. Details can be found in
-Section~\ref{ch_methods}.  This methodology builds off prior work to assess the
+software \citep{SmithEtAl2021}. Details can be found in Section~\ref{ch_methods}
+and in \citep{Dong2021}.  This methodology builds off prior work to assess the
 state of the practice for such domains as Geographic Information Systems
@@
 
-We designed a general process for evaluating the state of the practice of
-domain-specific SC software, that we instantiate for a specific scientific
-domain.
-
-Our method involves four steps:
+We developed a methodology for evaluating the state of the practice of SC
+software \citep{SmithEtAl2021}.  The  methodology can be instantiated for a
+specific domain of scientific software, which in the current case is medical
+imaging software for visualization.  Our methodology involves and engages a
+domain expert partner throughout, as discussed in
+Section~\ref{sec_vet_software_list}.  The four main steps of the methodology are:
+
 \begin{enumerate}
-\item choosing a software domain (Section \ref{sec_domain_selection});
-\item collecting and filtering software packages (Section
-\ref{sec_software_selection});
-\item grading the selected software (Section \ref{sec_grading_software});
-\item interviewing development teams for further information (Section
-\ref{sec_interview_methods}).
+\item Identify list of representative software packages
+(Section~\ref{sec_software_selection});
+\item Measure (or grade) the selected software (Section~\ref{sec_grading_software});
+\item Interview developers (Section~\ref{sec_interview_methods}).
+\item Answer the research questions (as given in Section~\ref{sec_motivation})
 \end{enumerate}
 
-\noindent Section \ref{sec_applying_method} presents an example of how we
-applied the method on the MI domain.
-
-\subsection{Domain Selection} \label{sec_domain_selection}
-
-Our methods are generic for any SC software, but they need to be applied to a
-specific domain. When choosing a candidate domain, we prefer one with a large
-number of active OSS. The reason is that we aim to finalize a list of 30
-software packages \citep{SmithEtAl2021} after the screening step. For example, we
-remove the ones without recent updates or specific functions. Besides, we prefer
-OSS projects because our grading method requires access to the code. In
-addition, we prefer a domain with an active community developing and using the
-software. As a result, it is easier to invite enough developers for interviews.
-
-We prefer 30 software packages providing similar functions or falling into
-different sub-groups depending on our research purpose. So the domain needs to
-have enough candidates in one sub-group or enough sub-groups to cross-compare.
-
-We also prefer domains in which our team has expertise. We invite domain experts
-to join and support our projects. They help us in many aspects, such as vetting
-the software list and interview questions.
-
-\subsection{Software Product Selection} \label{sec_software_selection}
-
-The process of selecting software packages contains two steps: i) identify
-software candidates in the chosen domain, ii) filter the list according to our
-needs \citep{SmithEtAl2021}.
-
-\subsubsection{Identify Software Candidates} \label{sec_identify_software_candidates}
-
-We start with finding candidate software in publications of the domain. Then, we
-search various websites, such as \hyperlink{https://github.com/}{GitHub},
-\hyperlink{https://swmath.org/}{swMATH} and the Google search results for
-software recommendation articles. We should also include the ones suggested by
-the domain experts \citep{SmithEtAl2021}.
-
-\subsubsection{Filter the Software List} \label{sec_filter_software_list}
-
-The goal is to build a software list with a length of about 30
-\citep{SmithEtAl2021}.
-
-The only ``mandatory'' requirement is that the software must be OSS, as defined
-in Section \ref{sec_open_source_software}. We need this because evaluating some
-software qualities requires the source code.
-
-The other filters are optional, and we consider them according to the number of
-software candidates and the objectives of the research project. We apply them in
-the following priority order:
+In the sections below we provide additional detail on the above steps, while
+concurrently giving examples of how we applied the methodology to the MI domain.
+
+\subsection{Interaction With Domain Expert} \label{sec_vet_software_list}
+
+The Domain Expert is an important member of the state of the practice assessment
+team. Pitfalls exist if non-experts attempt to acquire an authoritative list of
+software, or try to definitively rank the software. Non-experts have the problem
+that they can only rely on information available on-line, which has the
+following drawbacks:
+\begin{inparaenum}[i)]
+  \item the on-line resources could have false or inaccurate information; and,
+  \item the on-line resources could leave out relevant information that is so
+in-grained with experts that nobody thinks to explicitly record it.
+\end{inparaenum}
+
+Domain experts may be recruited from academia or industry.  The only
+requirements are knowledge of the domain and a willingness to be engaged in the
+assessment process.  The Domain Expert does not have to be a software developer,
+but they should be a user of domain software.  Given that the domain experts are
+likely to be busy people, the measurement process cannot put to much of a burden
+on their time.  For the current assessment, our Domain Expert (and paper
+co-author) is Dr.\ Michael Noseworthy, Professor of Electrical and Computer
+Engineering at McMaster University, Co-Director of the McMaster School of
+Biomedical Engineering, and Director of Medical Imaging Physics and Engineering
+at St. Joseph’s Healthcare, Hamilton, Ontario, Canada.
+
+In advance of the first meeting with the Domain Expert, the expert is asked to
+create a list of top software packages in the domain.  This is done to help
+the expert get in the right mind set in advance of the meeting.  Moreover,
+by doing the exercise in advance, we avoid the potential pitfall of the expert
+approving the discovered list of software without giving it adequate thought.
+
+The Domain Experts are asked to vet the collected data and analysis.  In
+particular, they are asked to vet the proposed list of software packages and the
+AHP ranking.  These interactions can be done either electronically or with
+in-person (or virtual) meetings.
+
+\subsection{List of Representative Software} \label{sec_software_selection}
+
+We have a two step process for selecting software packages: i) identify
+software candidates in the chosen domain; and, ii) filter the list to remove less relevant members \citep{SmithEtAl2021}.
+
+We initially identified 48 MI candidate software projects from the literature
+\citep{Bjorn2017, Bruhschwein2019, Haak2015}, online articles \citep{Emms2019,
+Hasan2020, Mu2019}, and forum discussions \citep{Samala2014}.  The full list of
+48 packages is available in \citet{Dong2021}.  To reduce the length of the list
+to a manageable number (29 in this case), we filtered the original list as
+follows:
 
 \begin{enumerate}
-\item The functions and purpose of the software. An SC domain often contains
-software with various functions and purposes. For example, some MI software
-packages are Tool Kit for developers to use, and some others offer Visualization
-function to end-users. We have two options:
-\begin{itemize}
-\item selecting a set of software with the same major function. In this case, we
-can use the identical process to assess all packages, e.g., same input to
-measure \textit{Robustness}. Also, when we give impression scores to qualities
-such as \textit{Installability} and \textit{Usability}, the results are more
-comparable. Thus, it is more feasible to collect the results and rank the
-software.
-\item selecting software from a set of different sub-groups. For example, we can
-choose 10 MI software from each of the three sub-groups: Visualization, Tool
-Kit, and PACS. The downside: we may need different processes to measure each
-sub-group; it may be less accurate to mix all three sub-groups and rank them
-together. The benefit: we can cross-compare the development processes between
-the sub-groups.
-\end{itemize}
-
-\item The version control tool. The empirical measurement tools listed in
-Section \ref{sec_empirical_measurements} only work on projects using
-\hyperlink{https://git-scm.com/}{Git}, so we prefer software with Git. Some
-manual steps in empirical measurement depend on a few metrics of GitHub, which
-makes projects held on GitHub more favored \citep{SmithEtAl2021}.
-
-\item The age of the software. Some of the OSS projects may experience a lack of
-recent maintenance. So we eliminate packages without recent updates, unless they
-are still popular and highly recommended by the domain users
-\citep{SmithEtAl2021}. We consider a software project as ``alive" if it has any
-update within the last 18 months; otherwise, we mark it as ``dead".
+
+\item We removed the packages that did not have source code available, such as
+\textit{MicroDicom}, \textit{Aliza}, and \textit{jivex}.
+
+\item We focused on the MI software that provides visualization functions, as
+described in Section~\ref{sec_scope}. We removed seven packages that were
+toolkits or libraries, such as \textit{VTK}, \textit{ITK}, and \textit{dcm4che}.
+We removed another three that were for PACS.
+
+\item We removed \textit{Open Dicom Viewer}, since it has not received any
+updates for a long time (since 2011).
+
 \end{enumerate}
 
-The order of filters 2 and 3 is flexible. We adjust it according to the number
-of software packages affected by the filters, and the number of ones remaining
-on the list.
-
-\subsubsection{Vet the Software List} \label{sec_vet_software_list}
-
-Before showing our filtered list to the domain experts, we ask them to list
-their top 10 software in the domain. Then, we cross-compare the two lists and
-discuss the commonality and variability.  In addition, we ask them to vet our
-filtered list. They provide views on whether the list is reasonable. We also use
-their opinions for a filtering process. For example, if a software package is
-not OSS and has had no updates for a long time, but the domain experts identify
-it as a valuable product, we still consider it in our final list.
+The Domain Expert provided a list of his top 12 software packages.  We compared
+his list to our list of 29.  We found 6 packages were on both lists: \textit{3D
+Slicer}, \textit{Horos}, \textit{ImageJ}, \textit{Fiji}, \textit{MRIcron} (we
+actually use the update version \textit{MRIcroGL}) and \textit{Mango} (we
+actually use the web version \textit{Papaya}).  Six software packages
+(\textit{AFNI}, \textit{FSL}, \textit{Freesurfer}, \textit{Tarquin},
+\textit{Diffusion Toolkit}, and \textit{MRItrix}) were on the Domain Expert
+list, but not on our filtered list.  However, when we examined those packages,
+we found they were out of scope, since their primary function was not
+visualization.  The Domain Expert agreed with our final choice of 29 packages.
 
@@
 developers by interviews.
-
-\subsubsection{Grading Template} \label{sec_grading_template}
 
@@
 measurements.
-
-\subsubsection{Empirical Measurements} \label{sec_empirical_measurements}
 
@@
 
-\subsubsection{Technical Details} \label{sec_technical_details}
-
-To test the software on a ``clean'' system, we create a new virtual machine (VM)
-for each software and only install the necessary dependencies before measuring.
-We make all 30 VMs on the same computer, one at a time, and destroy them after
-measuring.
+As in \citet{SmithEtAl2016}, Virtual machines (VMs) are used to provide an
+optimal testing environments for each package. VMs were used because it is
+easier to start with a fresh environment without having to worry about existing
+libraries and conflicts. Moreover, when the tests are complete the VM can be
+deleted, without any impact on the host operating system. The most significant
+advantage of using VMs is to level the playing field. Every software install
+starts from a clean slate, which removes ``works-on-my-computer'' errors. When
+filling in the measurement template spreadsheet, the the details for each VM
+should be noted, including hypervisor and operating system version.
 
@@
 measurements for one software on the same day.
+
+Then we followed the steps in Section \ref{sec_grading_software} to measure and
+grade the software. 27 out of the 29 packages are compatible with two or three
+different OS such as Windows, macOS, and Linux, and 5 of them are browser-based,
+making them platform-independent. However, in the interest of time, we only
+performed the measurements for each project by installing it on one of the
+platforms, most likely Windows.
 
@@
 \ref{sec_interview_process}).
-
-\subsubsection{Interview Questions} \label{sec_interview_questions}
 
@@
 
-\subsubsection{Interviewee Selection} \label{sec_interviewee_selection}
-
 For a software list with a length of roughly 30, we aim to interview about ten
@@
 
-\subsubsection{Interview Process} \label{sec_interview_process}
-
 Before contacting any interviewee candidate, we need to receive ethics clearance
@@
 
-\subsection{Applying the Method to MI} \label{sec_applying_method}
-
-This section shows an overview of applying our method to the MI domain.
-
-\subsubsection{Domain Selection}
-
-Based on the principles in Section \ref{sec_domain_selection}, we selected the
-MI domain and the sub-group of software with the Visualization function shown in
-Figure \ref{fig_mi_functions}. We also included Dr. Michael Noseworthy, a
-professor of Electrical and Computer Engineering at McMaster University,
-Co-Director of the McMaster School of Biomedical Engineering, and Director of
-Medical Imaging Physics and Engineering at St. Joseph’s Healthcare, and some of
-his students as the MI domain experts in our team.
-
-\subsubsection{Software Product Selection} \label{sec_mi_software_selection}
-
-By using the method in Section \ref{sec_identify_software_candidates}, we
-identified 48 MI software projects as the candidates from publications
-\citep{Bjorn2017} \citep{Bruhschwein2019} \citep{Haak2015}, online articles related
-to the domain \citep{Emms2019} \citep{Hasan2020} \citep{Mu2019}, forum discussions
-related to the domain \citep{Samala2014}, etc. Appendix
-\ref{ap_list_before_filtering} shows all 48 software packages.
-
-Guided by the method in Section \ref{sec_filter_software_list}, we filtered the
-list with a process as follows:
-
-\begin{enumerate}
-
-\item Among them, there were eight that we could not find their source code,
-such as \textit{MicroDicom}, \textit{Aliza}, and \textit{jivex}. These packages
-are likely to be freeware defined in Section \ref{sec_freeware} and not OSS. So
-following guidelines in Section \ref{sec_filter_software_list} we removed them
-from the list.
-
-\item Next, we focused on the MI software providing Visualization functions, as
-described in Section \ref{sec_scope}. Seven of the software on the list were
-Tool Kits or libraries for other software to use as dependencies, but not for
-end-users to view medical images, such as \textit{VTK}, \textit{ITK}, and
-\textit{dcm4che}; another three were PACS. We also eliminated these from the
-list.
-
-\item Finally, we removed \textit{Open Dicom Viewer} from the list because it
-had not received any updates for a long time (since 2011). After that, only
-\textit{MatrixUser} and \textit{AMIDE} were considered as ``dead". However, both
-of them had much more recent updates (after 2017) than \textit{Open Dicom
-Viewer}.
-
-\end{enumerate}
-
-We still preferred projects using git and GitHub and being updated recently, but
-did not apply this filter since packages were already below 30. Even without
-this filter, 27 out of the 29 software packages on the filtered list used git,
-and 24 chose GitHub.
-
-Following the process in Section \ref{sec_vet_software_list}, our domain experts
-provided a list of top software that contains 12 software products (Table
-\ref{tab_top_software_experts}). We compared two lists and found six common
-ones.
-
-\begin{table}[ht]
-\centering
-\begin{tabular}{ll}
-\hline
-Software & On both lists \\ \hline
-3D Slicer & X \\
-Horos & X \\
-ImageJ & X \\
-Fiji & X \\
-AFNI &  \\
-FSL &  \\
-Freesurfer &  \\
-Mricron & X \\
-Mango & X \\
-Tarquin &  \\
-Diffusion Toolkit &  \\
-MRItrix &  \\ \hline
-\end{tabular}
-\caption{\label{tab_top_software_experts}Top software by the MI domain experts}
-\end{table}
-
-We included \textit{Mango} in the initial list, but removed it because it was
-not OSS. However, we kept \textit{Papaya}, a the web version of \textit{Mango}.
-Instead of \textit{MRIcron}, we chose \textit{MRIcroGL}, because
-\textit{MRIcron} development had moved to \textit{MRIcroGL} \citep{Rorden2021b}.
-
-Six software packages on the domain experts' list were not on our filtered list.
-We believed their primary function was Analysis mentioned in Section
-\ref{sec_scope}. Thus, we did not include them in our final list.
-
-After vetting our filtered list, the domain experts believed it was reasonable
-and did not identify any problem. Thus, as shown in Appendix
-\ref{ap_list_before_filtering}, eventually, we had 29 software products on the
-final list.
-
-\subsubsection{Grading Software}
-\label{sec_applying_method_grading}
-Then we followed the steps in Section \ref{sec_grading_software} to measure and
-grade the software. 27 out of the 29 packages are compatible with two or three
-different OS such as Windows, macOS, and Linux, and 5 of them are browser-based,
-making them platform-independent. However, in the interest of time, we only
-performed the measurements for each project by installing it on one of the
-platforms, most likely Windows.
-
-\subsubsection{Interviews}
-\label{sec_apply_to_mi_interviews}
 We received ethics clearance from the McMaster University Research Ethics Board

### 2021-11-24 StateOfPractice/Papers/MedImageSoft_SOP/MedImageSoft_SOP_AcademicRadiology.tex
Edits to MI SOP up to (but not including) Section 5 (Interviews with Developers)

@@
 %\documentclass[preprint,12pt,authoryear]{elsarticle}
-\documentclass[3p, 12pt,authoryear]{elsarticle}
+%\documentclass[3p, 12pt,authoryear]{elsarticle}
+\documentclass[final, 3p, times, authoryear]{elsarticle}
 
@@
 correctness and reliability) together.  Moreover, the order is roughly the same
-as the order qualities are considered in practice.  [*** To be completed - not
-currently in sync with QDef paper, QDef paper not completely done ***]
+as the order qualities are considered in practice.  \wss{To be completed - not
+currently in sync with QDef paper, QDef paper not completely done.}
 
@@
 packages ($n = 29$) are compared for each of the $m$ criteria. The software are
-evaluated with the grading template in Section~\ref{sec_grading_template}, which
+evaluated with the grading template in Section~\ref{sec_grading_software}, which
 includes a subjective score from $1$ to $10$ for each quality for each package.
@@
 
-We have a two step process for selecting software packages: i) identify
-software candidates in the chosen domain; and, ii) filter the list to remove less relevant members \citep{SmithEtAl2021}.
+We have a two step process for selecting software packages: i) identify software
+candidates in the chosen domain; and, ii) filter the list to remove less
+relevant members \citep{SmithEtAl2021}.
 
@@
 
-We grade the selected software using a template (Section
-\ref{sec_grading_template}) and a specific empirical method (Section
-\ref{sec_empirical_measurements}). Some technical details for the measurements
-are in Section \ref{sec_technical_details}.  We do not measure
-\textit{reproducibility} with the grading template, but discuss it with the
-developers by interviews.
-
-The full grading template can be found in Appendix \ref{ap_grading_template}.
-The template contains 103 questions that we use for grading software products.
-Figure \ref{fg_grading_template_example} shows an example of this grading
-template.
+We grade the selected software using the measurement template summarized
+in~\citet{SmithEtAl2021}.  The template summarizes measures of the qualities
+listed in Section~\ref{sec_software_quality}, except for reproducibility, which
+is assessed through the developer interviews
+(Section~\ref{sec_interview_methods}). For each software package, we fill-in the
+template questions. This process can take between 1 to 4 hours for each package.
+Project developers can be contacted for help regarding installation, if
+necessary, but a cap of about 2 hours is imposed on the
+installation process, to keep the overall measurement time feasible. An excerpt
+of the spreadsheet is shown in Figure~\ref{fg_grading_template_example}.
+\wss{The Figure should be updated to be easier to read.}  A column is included
+to for each measured software package.
 
@@
 
-We use the first section of the template to collect general information, such as
-the name, purpose, platform, programming language, publications about the
-software, the first release and the most recent change date, website, source
-code repository of the product, etc. Information in this section helps us
-understand the projects better and may be helpful for further analysis, but it
-does not directly affect the grading scores.
-
-We designed the following nine sections in the template for the nine software
-qualities mentioned in Section~\ref{sec_software_quality}. For each quality, we
-ask several questions and the typical answers are among the collection of
-``yes'', ``no'', ``n/a'', ``unclear'', a number, a string, a date, a set of
-strings, etc. Each quality needs an overall score between 1 and 10 based on all
-the previous questions. For some qualities, we perform surface measurements,
-which allow us to measure all packages with reasonable efforts. The surface
-measurements reveal some traits of a underlying quality, but may not fully
-represent it.
-
-\begin{itemize}
-
-\item \textbf{Installability} We check the existence and quality of installation
-instructions. The user experience is also an important factor, such as the ease
-to follow the instructions, number of steps, automation tools, and the
-prerequisite steps for the installation. If any problem interrupts the process
-of installation or uninstallation, we give a lower score to this quality. We
-also record the Operating System (OS) for the installation test and whether we
-can verify the installation.
-
-\item \textbf{Correctness \& Verifiability} For \textit{correctness}, we check
-the projects to identify the techniques to ensure this quality, such as literate
-programming, automated testing, symbolic execution, model checking, unit tests,
-etc. We also examine whether the projects use continuous integration and
-continuous delivery (CI/CD). For \textit{verifiability}, we go through the
-documents of the projects to check the requirements specifications, theory
-manuals, and getting started tutorials. If a getting started tutorial exists and
-provides expected results, we follow it and check if the outputs match.
-
-\item \textbf{Surface Reliability} We check whether the software breaks during
-the installation and the operations following a getting started tutorials,
-whether there are descriptive error messages, and if we can recover the process
-after an error. We use the software to load damaged images during the assessment
-to \textit{surface robustness}, and lower its score for \textit{surface
-reliability} if a software product ``break" in this process.
+The full template consists of 108 questions categorized under 9 qualities.  The
+questions were designed to be unambiguous, quantifiable and measurable with
+limited time and domain knowledge. The measures are grouped under headings for
+each quality, and one for summary information. The summary information (shown in
+Figure~\ref{fg_grading_template_example}) is the first section of the template.
+This section summarizes general information, such as the software name, purpose,
+platform, programming language, publications about the software, the first
+release and the most recent change date, website, source code repository of the
+product, number of developers, etc.  We follow the definitions given by
+\citet{GewaltigAndCannon2012} for the software categories.  Public means
+software intended for public use.  Private means software aimed only at a
+specific group, while the concept category is used for software written simply
+to demonstrate algorithms or concepts. The three categories of development
+models are (open source, free-ware and commercial) are discussed in
+Section~\ref{sec_software_categories}.  Information in the summary section sets
+the context for the project, but it does not directly affect the grading scores.
+
+For measuring each quality, we ask several questions and the typical answers are
+among the collection of ``yes'', ``no'', ``n/a'', ``unclear'', a number, a
+string, a date, a set of strings, etc. Each quality is assigned an overall
+score, between 1 and 10, based on all the previous questions.  Several of the
+qualities use the word ``surface''.  This is to highlight that, for these
+qualities in particular, the best that we can do is a shallow measure.  For
+instance, we are not currently doing any experiments to measure usability.
+Instead, we are looking for an indication that usability was considered by the
+developers.  We do this by looking for cues in the documentation, like a getting
+started manual, a user manual and documentation of expected user
+characteristics.  Below is a summary of how each quality is measured.
+
+\begin{itemize}
+
+\item \textbf{Installability} We assess the following: 
+\begin{inparaenum}[i)]
+    \item existence and quality of installation instructions;
+    \item the quality of the user experience via the ease of following
+    instructions, number of steps, automation tools; and,
+    \item whether there is a means to verify the installation.
+\end{inparaenum}
+If any problem interrupts the process of installation or uninstallation, we give
+a lower score. We also record the Operating System (OS) used for the
+installation test.
+
+\item \textbf{Correctness \& Verifiability} We check each project to identify
+any techniques used to ensure this quality, such as literate programming,
+automated testing, symbolic execution, model checking, unit tests, etc. We also
+examine whether the projects use continuous integration and continuous delivery
+(CI/CD). For verifiability, we go through the documents of the projects to check
+for the presence of requirements specifications, theory manuals, and getting
+started tutorials. If a getting started tutorial exists and provides expected
+results, we follow it to check the correctness of the output.
+
+\item \textbf{Surface Reliability} We check the following: 
+\begin{inparaenum}
+    \item whether the software breaks during installation;
+    \item the operation of the software following the getting started tutorial (if present);
+    \item whether the error messages are descriptive; and,
+    \item whether we can recover the process after an error.
+\end{inparaenum}
 
@@
 unexpected/unanticipated input. For example, we prepare broken image files for
-MI software packages with the function to load image files. We use a text file
-(.txt) with a modified extension name (.dcm) as an unexpected/unanticipated
-input. We load a few correct input files to ensure the function is working
-correctly before testing with the unexpected/unanticipated ones.
-
-\item \textbf{Surface Usability} We examine the documentation of the projects,
-and we consider software with a getting started tutorial and a user manual
-easier to use. Meanwhile, we check if users have any channels to get support. We
-also record our impressions and user experiences when testing the software.
-Easy-to-use graphical user interfaces give us a better experience, which leads
-to better scores.
-
-\item \textbf{Maintainability} We search the projects' documents and identify
-the process of contributing and reviewing code. We believe that the artifacts of
-a project - including source code, documents, building scripts, etc. - can
-significantly affect its  \textit{maintainability}. Thus we check each project
-for its artifacts, such as API documentation, bug tracker, release notes, test
-cases, build files, version control, etc. We also check the tools supporting
-issue tracking and version control, the percentages of closed issues, and the
-proportion of comment lines in code.
+MI software packages that load image files. We use a text file (.txt) with a
+modified extension name (.dcm) as an unexpected/unanticipated input. We load a
+few correct input files to ensure the function is working correctly before
+testing the unexpected/unanticipated ones.
+
+\item \textbf{Surface Usability} We examine the project's documentation,
+checking for the presence of getting started tutorials and/or a user manual. We
+also check whether users have channels to request support, such as an e-mail
+address, or issue tracker. Our impressions of usability are based on our
+interaction with the software during testing.  In general, an easy-to-use
+graphical user interface scores higher.
+
+\item \textbf{Maintainability} We believe that the artifacts of a project,
+including source code, documents, and building scripts, significantly influence
+its maintainability. Thus we check each project for the presence of such
+artifacts as API documentation, bug tracker information, release notes, test
+cases, and build files. We also check for the use of tools supporting issue
+tracking and version control, the percentages of closed issues, and the
+proportion of comment lines in the code.
 
 \item \textbf{Reusability} We count the total number of code files for each
-project. Projects with a large number of components provide more choices to
-reuse. Furthermore, well-modularized code, which tends to have smaller parts in
-separated files, is typically easier to reuse. Thus, we consider the projects
-with more code files and less code lines per file to be more reusable. We use a
-command-line tool \textit{scc} to count the number of  text-based files and LOC
-for all projects. We also decide that the projects with API documentation can
-deliver better \textit{Reusability}.
-
-\item \textbf{Surface Understandability} We randomly examine 10 code files. We
-check the code’s style within each file, such as whether the identifiers,
-parameters, indentation, and formatting are consistent, whether the constants
-(other than 0 and 1) are hardcoded, and whether the developers modularized the
-code. We also check the descriptive information for the code, such as documents
-mentioning the coding standard, the comments in the code, and the descriptions
-or links for algorithms in the code.
+project. Projects with a large number of components potentially provide more
+choices for reuse. Furthermore, well-modularized code, which tends to have
+smaller parts in separate files, is typically easier to reuse. Thus, we consider
+the projects with more code files and less Lines of Code (LOC) per file to be
+more reusable. We also consider projects with API documentation as delivering
+better reusability.
+
+\item \textbf{Surface Understandability} Given that time is a constraint, we
+cannot look at all code files for each project; therefore, we randomly examine
+10 code files for their understandability. We check the code’s style within each
+file, such as whether the identifiers, parameters, indentation, and formatting
+are consistent, whether the constants (other than 0 and 1) are hardcoded, and
+whether the code is modularized. We also check the descriptive information for
+the code, such as documents mentioning the coding standard, the comments in the
+code, and the descriptions or links for details on algorithms in the code.
 
 \item \textbf{Visibility/Transparency} To measure this quality, we check the
-existing documents to find out whether the software development process and
+existing documents to find whether the software development process and
 current status of a project are visible and transparent. We examine the
 development process, current status, development environment, and release notes
-for each project. If any information is missing or poorly conveyed, the
-\textit{visibility/transparency} will be lower.
-\end{itemize}
-
-For some qualities, the empirical measurements also affect the score. We use
-tools to extract information from the source code repositories. For projects
-held on GitHub, we manually collect additional metrics, such as the stars of the
-GitHub repository, and the numbers of open and closed pull requests. Section
-\ref{sec_empirical_measurements} presents more details about the empirical
-measurements.
-
-We use two command-line tools for the empirical measurements. One is
-\textit{GitStats} that generates statistics for git repositories and displays
-outputs in the form of web pages \citep{Gieniusz2019}; the other one is Sloc Cloc
-and Code (as known as \textit{scc}) \citep{Boyter2021}, aiming to count the lines
-of code, comments, etc.
+for each project.
+\end{itemize}
+
+As part of filling in the measurement template, we use freeware tools to collect
+repository related data. \href{https://github.com/tomgi/git_stats}{GitStats}
+\citep{Gieniusz2019} is used to measure the number of binary files as well as
+the number of added and deleted lines in a repository. This tool is also used to
+measure the number of commits over different intervals of time.
+\href{https://github.com/boyter/scc}{Sloc Cloc and Code (scc)}
+\citep{Boyter2021} is used to measure the number of text based files as well as
+the number of total, code, comment, and blank lines in a repository.
 
@@
 
-Additionally, we manually collect information for projects held on GitHub, such
-as the numbers of stars, forks, people watching this repository, open pull
-requests, closed pull requests, and the number of months a repository has been
+For projects on GitHub we manually collect additional information, such as the
+numbers of stars, forks, people watching this repository, and open pull requests
 on GitHub. A git repository can have a creation date much earlier than the first
-day on GitHub. For example, the developers created the git repository of
-\textit{3D Slicer} in 2002, but did not upload a copy of it to GitHub until
-2020. We get the creation date of the GitHub copy by using API
-\textit{https://api.github.com/repos/{:owner}/{:repository}} (e.g.,
-\hyperlink{https://api.github.com/repos/slicer/slicer}{https://api.github.com/repos/slicer/slicer}).
-In the response, the value of ``created\_at" is what we want. The number of
-months a repository has been on GitHub helps us understand the average change of
-metrics over time, e.g., the average new stars per month.
-
-These empirical measurements help us from two aspects. Firstly, they help us
-with getting a project overview faster and more accurately. For example, the
-number of commits over the last 12 months shows how active this project has
-been, and the number of stars and forks may reveal its popularity. Secondly, the
-results may affect our decisions regarding the grading scores for some software
-qualities. For example, if the percentage of comment lines is low, we
-double-check the \textit{understandability} of the code; if the ratio of open
-versus closed pull requests is high, we pay more attention to the
-\textit{maintainability}.
-
-As in \citet{SmithEtAl2016}, Virtual machines (VMs) are used to provide an
+pull, closed pull requests, and the number of months a repository has been on
+GitHub. For example, the developers created the git repository for \textit{3D
+Slicer} in 2002, but did not upload a copy of it to GitHub until 2020. Some
+GitHub data can be found using its the GitHub Application Program Interface
+(API) via the following url:
+\textit{https://api.github.com/repos/[owner]/[repository]} where [owner] and
+[repository] are replaced by the repo specific values. The number of months a
+repository has been on GitHub helps us understand the average change of metrics
+over time, like the average new stars per month.
+
+The repository measures help us in wo ways. Firstly, they help us with get a
+fast and accurate project overview. For example, the number of commits over the
+last 12 months shows how active a project has been, and the number of stars and
+forks may reveal its popularity. Secondly, the results may affect our decisions
+regarding the grading scores for some software qualities. For example, if the
+percentage of comment lines is low, we double-check the
+\textit{understandability} of the code; if the ratio of open versus closed pull
+requests is high, we pay more attention to \textit{maintainability}.
+
+As in \citet{SmithEtAl2016}, Virtual machines (VMs) were used to provide an
 optimal testing environments for each package. VMs were used because it is
@@
 
-We spend about two hours grading each package, unless we find technical issues
-and need more time to resolve them. In most of the situation, we finish all the
-measurements for one software on the same day.
-
-Then we followed the steps in Section \ref{sec_grading_software} to measure and
-grade the software. 27 out of the 29 packages are compatible with two or three
-different OS such as Windows, macOS, and Linux, and 5 of them are browser-based,
-making them platform-independent. However, in the interest of time, we only
-performed the measurements for each project by installing it on one of the
-platforms, most likely Windows.
+When grading the software, we found 27 out of the 29 packages are compatible
+with two or three different OS such as Windows, macOS, and Linux, and 5 of them
+are browser-based, making them platform-independent. However, in the interest of
+time, we only performed the measurements for each project by installing it on
+one of the platforms.  When it was an option, we selected Windows as the host
+OS.
 
@@
 
-This section introduces our interview questions (Section
-\ref{sec_interview_questions}), method of selecting interviewees (Section
-\ref{sec_interviewee_selection}), and interview process (Section
-\ref{sec_interview_process}).
-
 We designed a list of 20 questions to guide our interviews, which can be found
-in Section \ref{ch_interview} and Appendix \ref{ap_interview}.
-
-Some questions are about the background of the software, the development teams,
-the interviewees, and how they organize the projects. We also ask about their
-understandings of the users. Some questions focus on the current and past
-difficulties, and the solutions the team has found or will try. We also discuss
-the importance and current situations of documentation. A few questions are
-about specific software qualities, such as \textit{maintainability},
-\textit{understandability}, \textit{usability}, and \textit{reproducibility}.
-
-The interviews are semi-structured based on the question list; we ask follow-up
+in \citet{SmithEtAl2021}. Some questions are about the background of
+the software, the development teams, the interviewees, and how they organize the
+projects. We also ask about the developer's understandings of the users. Some
+questions focus on the current and past difficulties, and the solutions the team
+has found, or will try. We also discuss the importance and current situations of
+documentation. A few questions are about specific software qualities, such as
+maintainability, understandability, usability, and reproducibility. The
+interviews are semi-structured based on the question list; we ask follow-up
 questions when necessary. Based on our experience, the interviewees usually
@@
 
-For a software list with a length of roughly 30, we aim to interview about ten
-development teams. Interviewing multiple individuals from each team gives us
-more comprehensive information, but a single engineer well-knowing the project
-is also sufficient.
-
-Ideally, we select projects after the grading measurements and prefer the ones
-with higher overall scores. However, if we do not find enough participants, we
-reach all teams on the list. As mentioned in Section
-\ref{sec_apply_to_mi_interviews}, when we applied this process to the MI domain,
-we eventually contacted all teams.
-
-We try to find the contacts of the teams on the projects' websites, such as the
-official web pages, repositories, publications, and bio pages of the teams'
-institutions. Then, we send at most two emails to one contact asking for its
-participation before receiving any replies. We operate the invitation according
-to our ethics approval, such as the one in Appendix \ref{ap_ethics}. For
-example, we ask for participants' consent before interviewing them, recording
-the conversation, or including it in our report.
-
-Before contacting any interviewee candidate, we need to receive ethics clearance
-from the McMaster University Research Ethics Board. Since the members of the
-development teams are usually around the world, we organize these interviews as
-virtual meetings online with \hyperlink{https://zoom.us/}{Zoom}. After receiving
-consent from the interviewees, we also record and transcribe our discussions.
-
-We received ethics clearance from the McMaster University Research Ethics Board
-(Appendix \ref{ap_ethics}). Going through the interview process in Section
-\ref{sec_interview_methods}, we contacted all of the 29 teams. Members from
-eight teams responded and agreed to participate. As a result, we interviewed
-nine developers and architects from the eight teams.
+We sent out interview requests to all 29 projects, with 9 projects responding.
+In some cases multiple developers from the same project agreed to be
+interviewed. We found contact information from projects websites, code
+repository, publications, and from biographic pages at the teams’ institutions.
+Meetings were held on-line using either Zoom or Teams, which facilitates
+recording and automatic transcription of the meetings.  The interview process
+presented here was approved by the McMaster University Research Ethics Board
+under the application number 
+\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/MACREM/Application.pdf}
+{MREB\#: 5219}.
 
@@
 
-As discussed in Section~\ref{sec_grading_software}, we use a grading template
-and a empirical method to measure the selected software. We applied this step to
-the MI domain (Section~\ref{sec_applying_method_grading}). This section shows
-the summary of the measurement results. The detailed data can be found in the
-repository
-\hyperlink{https://data.mendeley.com/datasets/k3pcdvdzj2/1}{https://data.mendeley.com/datasets/k3pcdvdzj2/1}.
-This section contains part of the answers to \hyperlink{rq5}{RQ5}.
-
 Table \ref{tab_final_list} shows the 29 software packages that we measured,
-along with some summary data collected in the year 2020. As mentioned in Section
-\ref{sec_grading_template}, we used \textit{scc} (Section
-\ref{sec_empirical_measurements}) to count the Lines of Code (LOC), excluding
-the comment and blank lines. We arrange the items in the descending order of the
-LOC. We found the initial release dates (Rlsd) in their documents for most
+along with summary data collected in the year 2020. We arrange the items in
+descending order of LOC. We found the initial release dates (Rlsd) for most
 projects and marked the two unknown dates with ``?". We used the date of the
-latest change to each code repository to decide the latest update. We found out
-funding information (Fnd) for only eight projects.
-
-We counted the number of contributors (NOC). We considered anyone who made at
-least one accepted commit to the source code as a contributor. Thus, the NOC is
-not usually the same as the number of long-term members. Many of these projects
-received change requests and code from the community, such as pull requests and
-git commits on GitHub.
-
-Table \ref{tab_final_list} also shows the supported OS for each software
-package. Twenty-five of them could work on all three OSs: Windows (W), macOS
-(M), and Linux (L). However, there was a significant difference in the
-philosophy to achieve cross-platform compatibility. Most of them were native
-software products, but five were naturally platform-independent web
-applications, as shown in column ``Web".
-
-\begin{table}[ht]
-\centering
-\begin{tabular}{llllllllll}
+latest change to each code repository to decide the latest update. We found
+funding information (Fnd) for only eight projects.  For the number of
+contributors (NOC) we considered anyone who made at least one accepted commit as
+a contributor. The NOC is not usually the same as the number of long-term
+project members, since many projects received change requests and code from the
+community.  For the supported OS, 25 packages work on all three OSs: Windows
+(W), macOS (M), and Linux (L). Although the usual approach to cross-platform
+compatibility was to work natively on multiple OSes, five projects achieved
+platform-independence via web applications. The full measurement data for all
+packages is available in \citet{Dong2021-Data}.
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{p{6cm}lllllllll}
 \hline
@@
 
-Most of the projects used more than one programming language, including a
-primary language that the developers used the most. Figure \ref{fig_language}
-shows the primary languages versus the number of projects using them.
-
-% temporarily comment out
-
-% \begin{figure}[ht]
-% \centering
-% \begin{tikzpicture}
-% \centering
-% \begin{axis}
-% [
-% ybar,
-% height=8cm,
-% width=12cm,
-% ylabel={Number of projects},
-% xlabel={\ Primary language},
-% symbolic x coords={C++, JavaScript, Java, C, Python, Pascal, Matlab},
-% xtick=data,
-% nodes near coords,
-% nodes near coords align={vertical},
-% ]
-% \addplot[black,fill=blue!55!black] coordinates {(C++,11) (JavaScript,6) (Java,4) (C,3) (Python,3) (Pascal,1) (Matlab,1) };
-% \end{axis}  
-% \end{tikzpicture}
-% \caption{\label{fig_language}Primary languages versus number of projects using them}
-% \end{figure}
-
-We failed to install \textit{DICOM Viewer}, so we could not test its
-\textit{surface reliability} and \textit{surface robustness}. We kept this
-software on our list because the other seven qualities do not rely on a
-successful installation. Besides, the \textit{DICOM Viewer} team built it as a
-plugin software for NextCloud
-(\hyperlink{https://apps.nextcloud.com/}{https://apps.nextcloud.com/}) platform,
-which was a unique choice we had not seen before. We wanted to keep it to enrich
-the diversity.
+Figure \ref{fig_language} shows the primary languages versus the number of
+projects using them.  The primary language is the language used for the majority
+of the project's code; in most cases projects also use other languages.  The
+most popular language is C++, with almost 40\% of projects (11 of 29).  The two
+least popular choices are Pascal and  Matlab, with around 3\% of projects each
+(1 of 29).
+
+\begin{figure}[ht]
+\centering
+\includegraphics{figures/PrimaryLanguages.pdf}
+\centering
+\caption{\label{fig_language}Primary languages versus number of projects}
+\end{figure}
 
@@
 
-Figure \ref{fg_installability_scores} lists the scores of
-\textit{installability}.
+Figure \ref{fg_installability_scores} lists the installability scores.  We found
+installation instructions for 16 projects. Among the ones without instructions,
+\textit{BioImage Suite Web} and \textit{Slice:Drop} do not need installation,
+since they are web applications. Installing 10 of the projects required extra
+dependencies. Five of them are the web applications (as shown in
+Table~\ref{tab_final_list}) and depended on a browser; \textit{dwv},
+\textit{OHIF Viewer}, and \textit{GATE} needed extra dependencies to build;
+\textit{ImageJ} and	\textit{Fiji} needed an unzip tool; \textit{MatrixUser} was
+based on Matlab; \textit{DICOM Viewer} needed to work on a Nextcloud platform.
 
@@
 \end{figure}
-
-We found installation instructions for 16 projects. Among the ones without
-instructions, \textit{BioImage Suite Web} and \textit{Slice:Drop} are web
-applications with online versions to use, thus they do not need installation.
-Installing 10 of the projects required extra dependencies. Five of them are the
-web applications in Table \ref{tab_final_list}, and depended on a browser;
-\textit{dwv}, \textit{OHIF Viewer}, and \textit{GATE} needed extra dependencies
-to build; \textit{ImageJ} and	\textit{Fiji} needed an unzip tool;
-\textit{MatrixUser} was based on Matlab; \textit{DICOM Viewer} needed to work on
-a Nextcloud platform.
 
@@
 installation, we gave them higher scores. \textit{BioImage Suite Web} also
-provided an option to download cache to local for offline usage, which was easy
+provided an option to download cache for offline usage, which was easy
 to apply.
 
-\textit{dwv}, \textit{GATE}, and \textit{DICOM Viewer} showed severe problems.
-We could not install them. We spent a reasonable amount of time on these
-problems, then considered them as major obstacles for normal users if we could
-not figure out any solutions. We suspect that only a part of the users faced the
-same problems, and given a lot of time, we might be able to find solutions.
-However, the difficulties greatly impacted the installation experiences, and we
-graded these software packages with lower scores. For example, \textit{dwv} and
-\textit{GATE} had the option to build from the source code, and we failed the
-building processes following the instructions. Although we could not locally
-build them, we could use a deployed online version for \textit{dwv}, and a VM
-version for \textit{GATE}. With those, we finished all the measurements for
-them. Furthermore, \textit{DICOM Viewer} depended on the NextCloud platform, and
-we could not successfully install the dependency.
-
-\textit{MatrixUser} has a lower score because it depended on Matlab. We
-considered installing Matlab takes many more steps and time, and some users may
-not have a license to use Matlab.
+\textit{dwv}, \textit{GATE}, and \textit{DICOM Viewer} showed severe
+installation problems. We were not able to install them, even after a reasonable
+amount of time (2 hours).  For \textit{dwv} and \textit{GATE} we failed to build
+from the source code, but we were able to proceed with measuring other qualities
+using a deployed online version for \textit{dwv}, and a VM version for
+\textit{GATE}. For \textit{DICOM Viewer} we could not install the NextCloud
+dependency, and we did not have another option for running the software.
+Therefore, for \textit{DICOM Viewer} we could not measure reliability or
+robustness.  The other seven qualities could be measured, since they do not require installation.
+
+\textit{MatrixUser} has a lower score because it depended on Matlab. We assessed
+the score from the point of view of a user that would have to install Matlab and
+acquire a license.  Of course, for users that already work within Matlab, the
+installability score would be higher.
 
@@
 
-The scores of \textit{correctness \& verifiability} are shown in Figure
-\ref{fg_correctness_erifiability_scores}. Generally speaking, the packages with
-higher scores adopted more techniques to improve \textit{correctness}, and had
-better documents for us to verify it.
+The scores of correctness \& verifiability are shown in
+Figure~\ref{fg_correctness_verifiability_scores}. Generally speaking, the
+packages with higher scores adopted more techniques to improve correctness, and
+had better documentation for us to verify this.  For instance, we looked for
+evidence of unit testing, since it benefits most parts of the software's life
+cycle, such as designing, coding, debugging, and optimization
+\citep{Hamill2004}.  We only found evidence of unit testing for about half of
+the projects.
 
@@
 \caption{AHP correctness \& verifiability scores}
-\label{fg_correctness_erifiability_scores}
+\label{fg_correctness_verifiability_scores}
 \end{figure}
 
-After examining the source code, we could not find any evidence of unit testing
-in more than half of the projects. Unit testing benefits most parts of the
-software's life cycle, such as designing, coding, debugging, and optimization
-\citep{Hamill2004}. It can reveal the bugs at an earlier stage of the development
-process, and the absence of unit testing may cause problems for
-\textit{correctness \& verifiability}.
-
-We could not find requirements specifications for most projects. The only
-document we found is a road map of \textit{3D Slicer}, which contained design
-requirements for the upcoming changes. However, it did not record the conditions
-for previous versions. We also could not identify the theory manuals for all of
-the projects. Even for some projects with well-organized documents, requirements
-specifications and theory manuals were still missing.
-
-We identified five projects using CI/CD tools, which are \textit{3D Slicer},
+Even for some projects with well-organized documents, requirements
+specifications and theory manuals were still missing.  We could not identify
+theory manuals for all projects and we did not find requirements specifications
+for most projects. The only document we found was a road map of \textit{3D
+Slicer}, which contained design requirements for upcoming changes.
+
+We identified five projects using CI/CD tools: \textit{3D Slicer},
 \textit{ImageJ}, \textit{Fiji}, \textit{dwv}, and \textit{OHIF Viewer}.
 
-In this section, the information about CI/CD tools is part of the answers to
-\hyperlink{rq2}{RQ2}, and the information about software testing and
-documentation is part of the answers to \hyperlink{rq3}{RQ3}.
+% In this section, the information about CI/CD tools is part of the answers to
+% \hyperlink{rq2}{RQ2}, and the information about software testing and
+% documentation is part of the answers to \hyperlink{rq3}{RQ3}.
 
@@
 
-As described in Section \ref{sec_result_installability}, we could not build
-\textit{dwv} and \textit{GATE}. However, since there was an online or VM version
-of them, successful deployment is possible. So the failure of installation did
-not affect their scores in \textit{surface reliability}. Figure
-\ref{fg_reliability_scores} shows the AHP results.
+Figure~\ref{fg_reliability_scores} shows the AHP results.  As shown in
+Section~\ref{sec_result_installability}, most of the software products did not
+``break" during installation, or did not need installation; \textit{dwv} and
+\textit{GATE} broke in the building stage, and the processes were not
+recoverable; we could not install the dependency for \textit{DICOM Viewer}. Of
+the seven software packages with a getting started tutorial and operation steps
+in the tutorial, most showed no error when we followed the steps. However,
+\textit{GATE} could not open macro files and became unresponsive several times,
+without any descriptive error message. When assessing robustness
+(Section~\ref{sec_result_robustness}), we found that \textit{Drishti} crashed
+when loading damaged image files, without showing any descriptive error message.
+On the other hand, we did not find any problems with the online version of
+\textit{dwv}.
 
@@
 
-As shown in Section \ref{sec_result_installability}, most of the software
-products did not ``break" during installation or did not need installation;
-\textit{dwv} and \textit{GATE} broke in the building stage, and the processes
-were not recoverable; we could not install the dependency for \textit{DICOM
-Viewer}.
-
-Of the seven software packages with a getting started tutorial and operation
-steps in the tutorial, most showed no error when we followed the steps. However,
-\textit{GATE} could not open macro files and became unresponsive several times,
-without any descriptive error message. When assessing \textit{surface
-robustness} (Section \ref{sec_result_robustness}), we found out that
-\textit{Drishti} crashed during loading damaged image files and did not show any
-descriptive error message. On the other hand, we did not see any problems with
-the online version of \textit{dwv}.
-
 \subsection{Surface Robustness} \label{sec_result_robustness}
 
-Figure \ref{fg_robustness_scores} presents the scores for \textit{surface
-robustness}.
+Figure \ref{fg_robustness_scores} presents the scores for surface robustness.
+The packages with higher scores elegantly handled unexpected/unanticipated
+inputs, typically showing a clear error message. We may have underestimated the
+score of \textit{OHIF Viewer}, since we needed further customization to load
+data.
 
@@
 
-The packages with higher scores elegantly handled the unexpected/unanticipated
-inputs, typically showing a clear error message. We might underestimate the
-score of \textit{OHIF Viewer} since we needed further customization to load
-data, and the test was not complete.
-
-Digital Imaging and Communications in Medicine (DICOM) is a widely used MI
-standard, and ``it defines the formats for medical images that can be exchanged
-with the data and quality necessary for clinical use" \citep{MITA2021}. According
-to their documentation, all 29 software packages should support the DICOM
-standard. We prepared two types of image files: the ones in correct formats and
-the broken ones. We used two MI sample files in the DICOM format as the image
-files with valid formats; we created a standard text file, changed its extension
-name from ``.txt" to ``.dcm", and used it as the unexpected/unanticipated input.
-
-Being tested with the input files with correct formats, all software packages
-except \textit{GATE} loaded the images correctly. \textit{GATE} failed this test
-for unknown reasons.
-
-With the unexpected/unanticipated input, \textit{MatrixUser}, \textit{dwv}, and
+Digital Imaging and Communications in Medicine (DICOM) ``defines the formats for
+medical images that can be exchanged with the data and quality necessary for
+clinical use" \citep{MITA2021}. According to their documentation, all 29
+software packages should support the DICOM standard. To test robustness, we
+prepared two types of image files: correct format and broken format (created by
+relabelled a text file to have the ``.dcm'' extension).  All software packages
+loaded the correct format image, except for \textit{GATE}, which failed for
+unknown reasons.  For the broken format, \textit{MatrixUser}, \textit{dwv}, and
 \textit{Slice:Drop} ignored the incorrect format of the file and loaded it
@@
 with noise pixels. \textit{Drishti} successfully detected the broken format of
-the file, but the software crashed as a result. We recorded \textit{Drishti}'s
-issue to the measurement of its \textit{reliability} in Section
-\ref{sec_result_installability}.
+the file, but the software crashed as a result.
 
@@
 
-Figure \ref{fg_usability_scores} lists the AHP scores for \textit{surface
-usability}.
+Figure \ref{fg_usability_scores} shows the AHP scores for surface usability. The
+software with higher scores usually provided both comprehensive documented
+guidance and a good user experience. \textit{INVESALIUS 3} provided an excellent
+example of a detailed and precise user manual. \textit{GATE} also provided a
+large number of documents, but unfortunately we had difficulty understanding and
+using them. We found getting started tutorials for only 11 projects, but a user
+manual for 22 projects. \textit{MRIcroGL} was the only project that explicitly
+documented expected user characteristics.
 
@@
 
-We found a getting started tutorial for only 11 projects but a user manual for
-22 projects. \textit{MRIcroGL} was the only one with documentation for the
-expected user characteristics.
-
-The software with higher scores usually provided both comprehensive document
-guidance and a good user experience. \textit{INVESALIUS 3} set an excellent
-example of a detailed and precise user manual. \textit{GATE} also provided a
-large number of documents, but we think that they conveyed the ideas poorly, as
-we had trouble understanding and using them.
-
-Table \ref{tab_user_support_model} shows the user support models by the number
-of projects using them. This table contains part of the answers to
-\hyperlink{rq2}{RQ2}. Maybe not every team intended to use GitHub issues to
-answer users' questions, but many users use them to seek help.
+Table~\ref{tab_user_support_model} summarizes the user support models by the number
+of projects using them. We do not know whether the prevalent use of GitHub issues for user support is by design, or whether this just naturally happened as users sought help.
+%  This table contains part of the answers to \hyperlink{rq2}{RQ2}.
 
@@
 
-Figure \ref{fg_maintainability_scores} shows the AHP results for
-\textit{maintainability}.
+Figure~\ref{fg_maintainability_scores} shows the ranking results for
+maintainability. We marked \textit{3D Slicer} with a higher score because
+we found it to have the most comprehensive artifacts. For example, as far as we
+could find, only a few of the 29 projects had a project plan, developer's
+manual, or API documentation, and only \textit{3D Slicer}, \textit{ImageJ},
+\textit{Fiji} included all three documents. Moreover, \textit{3D Slicer} has a
+much higher percentage of closed issues (91.65\%) compared to \textit{ImageJ}
+(52.49\%) and \textit{Fiji} (63.79\%). Table~\ref{tab_maintainability_docs}
+shows which projects had these documents, in the descending order of their
+maintainability scores. 
+% This table contains part of the answers to \hyperlink{rq1}{RQ1} and \hyperlink{rq3}{RQ3}.
 
@@
 \end{figure}
-
-We marked \textit{3D Slicer} with a much higher score than others because it did
-very well at closing the identified issues, and more importantly, we found it to
-have the most comprehensive artifacts. For example, as far as we could find out,
-only a few of the 29 projects had a project plan, developer's manual, or API
-documentation, and only \textit{3D Slicer}, \textit{ImageJ}, \textit{Fiji}
-included all three documents. Meanwhile, \textit{3D Slicer} has a much higher
-percentage of closed issues (91.65\%) than \textit{ImageJ} (52.49\%) and
-\textit{Fiji} (63.79\%). Table \ref{tab_maintainability_docs} shows which
-projects had these documents, in the descending order of their
-\textit{maintainability} scores. This table contains part of the answers to
-\hyperlink{rq1}{RQ1} and \hyperlink{rq3}{RQ3}.
 
@@
 
-27 of the 29 projects used git as the version control tool; \textit{AMIDE} used
-Mercurial; \textit{Gwyddion} used Subversion. 24 projects used GitHub for their
-repositories; \textit{XMedCon}, \textit{AMIDE}, and \textit{Gwyddion} used
-SourceForge; \textit{DicomBrowser} and \textit{3DimViewer} used BitBucket. The
-information about version control tools is part of the answers to
-\hyperlink{rq2}{RQ2}.
+27 of the 29 projects used git as the version control tool, with 24 of these
+using GitHub. \textit{AMIDE} used Mercurial and \textit{Gwyddion} used
+Subversion. \textit{XMedCon}, \textit{AMIDE}, and \textit{Gwyddion} used
+SourceForge. \textit{DicomBrowser} and \textit{3DimViewer} used BitBucket. 
+% The information about version control tools is part of the answers to
+% \hyperlink{rq2}{RQ2}.
 
@@
 
-Figure 4.7 shows the AHP results for \textit{reusability}.
+Figure 4.7 shows the AHP results for reusability. As described in
+Section~\ref{sec_grading_software}, we gave higher scores to the projects with
+API documentation. As shown in Table \ref{tab_maintainability_docs}, seven
+projects had API documents. We also considered projects with more code files and
+less LOC per code file as more reusable. Table \ref{tab_loc_per_file} shows the
+number of text-based files by projects, which we used to approximate the number
+of code files. The table also lists the total number of lines (including
+comments and blanks), LOC, and average LOC per file. We arranged the items in
+descending order of their reusability scores.
 
@@
 \end{figure}
-
-As described in Section \ref{sec_grading_template}, we gave higher scores to the
-projects with an API document. As shown in Table \ref{tab_maintainability_docs},
-seven projects had API documents. We also considered projects with more code
-files and less LOC per code file as more reusable. Table \ref{tab_loc_per_file}
-shows the number of text-based files by projects, which we used to approximate
-the number of code files. The table also lists the total number of lines
-(including comments and blanks), LOC, and average LOC per file. We arranged the
-items in the descending order of their \textit{reusability} scores.
 
@@
 
-Figure \ref{fg_surface_understandability_scores} shows the scores for
-\textit{surface understandability}.
+Figure~\ref{fg_surface_understandability_scores} shows the scores for surface
+understandability. All projects had a consistent code style with parameters in
+the same order for all functions; the code was modularized; the comments were
+clear, indicating what is being done, not how. However, we only found explicit
+identification of a coding standard for 3 out of the 29: \textit{3D Slicer},
+\textit{Weasis}, and \textit{ImageJ}. We also found hard-coded constants in
+\textit{medInria}, \textit{dicompyler}, \textit{MicroView}, and \textit{Papaya}.
+We did not find any reference to the used algorithms in projects
+\textit{XMedCon}, \textit{DicomBrowser}, \textit{3DimViewer}, \textit{BioImage
+Suite Web}, \textit{Slice:Drop}, \textit{MatrixUser}, \textit{DICOM Viewer},
+\textit{dicompyler}, and \textit{Papaya}.
 
@@
 
-All projects had a consistent code style with parameters in the same order for
-all functions; the code was modularized; the comments were clear, indicating
-what is being done, not how. However, we only found explicit identification of a
-coding standard for 3 out of the 29: \textit{3D Slicer}, \textit{Weasis}, and
-\textit{ImageJ}. We also found hard-coded constants in \textit{medInria},
-\textit{dicompyler}, \textit{MicroView}, and \textit{Papaya}. We did not find
-any reference to the used algorithms in projects \textit{XMedCon},
-\textit{DicomBrowser}, \textit{3DimViewer}, \textit{BioImage Suite Web},
-\textit{Slice:Drop}, \textit{MatrixUser}, \textit{DICOM Viewer},
-\textit{dicompyler}, and \textit{Papaya}.
-
 \subsection{Visibility/Transparency} \label{sec_result_visibility_transparency}
@@
 \textit{visibility/transparency}. Generally speaking, the teams that actively
-documented their development process and plans scored higher because they
-delivered better communication to people outside the team.
+documented their development process and plans scored higher. Table
+\ref{tab_Visibility/Transparency_docs} shows the projects that had documents for
+the development process, project status, development environment, and release
+notes, in the descending order of their visibility/transparency scores.
+%This table contains part of the answers to \hyperlink{rq1}{RQ1} and \hyperlink{rq3}{RQ3}.
 
@@
 \end{figure}
-
-Table \ref{tab_Visibility/Transparency_docs} shows the projects which had
-documents for the development process, project status, development environment,
-and release notes, in the descending order of their
-\textit{visibility/transparency} scores. This table contains part of the answers
-to \hyperlink{rq1}{RQ1} and \hyperlink{rq3}{RQ3}.
 
@@
 
-As described in Section \ref{sec_AHP}, for our AHP measurements, there are nine
-criteria which are the nine software qualities and 29 software packages as the
-alternatives. In the absence of requirements for an actual project, we made all
-nine qualities equally important, so the score of each quality affects the
-overall scores on the same scale.
-
-Figure \ref{fg_overall_scores} shows the overall scores of all 29 software
-packages in descending order. Since we produced the scores from the AHP process,
-the total sum of the 29 scores is precisely 1.
+As described in Section~\ref{sec_AHP}, for our AHP measurements, we have nine
+criteria (qualities) and 29 alternatives (software packages). In the absence of
+a specific real world context, we assumed all nine qualities are equally
+important. Figure \ref{fg_overall_scores} shows the overall scores in descending
+order. Since we produced the scores from the AHP process, the total sum of the
+29 scores is precisely 1.
 
@@
 ranked in the top two software products for all qualities except \textit{surface
-robustness}; \textit{ImageJ} ranked in the top three products for
-\textit{correctness \& verifiability}, \textit{surface reliability},
-\textit{surface usability}, \textit{maintainability}, \textit{ surface
-understandability}, and \textit{visibility/transparency}; \textit{OHIF Viewer}
-ranked in the top five products for \textit{correctness \& verifiability},
-\textit{surface reliability}, \textit{surface usability},
-\textit{maintainability}, and \textit{reusability}. We might underestimate its
-scores of qualities \textit{surface reliability} and \textit{surface robustness}
-for \textit{DICOM Viewer}, but equally compared it with the other software for
-the rest seven qualities.
+robustness}; \textit{ImageJ} ranked in the top three for correctness \&
+verifiability, surface reliability, surface usability, maintainability, surface
+understandability, and visibility/transparency. \textit{OHIF Viewer} ranked in
+the top five products for correctness \& verifiability, surface reliability,
+surface usability, maintainability, and reusability. Given the installation
+problems, we may have underestimated the scores on reliability and robustness
+for \textit{DICOM Viewer}, but we compared it equally for the other seven
+qualities.
 
@@
 We refer to these sections to avoid repetition, and organize the references in
-tables (e.g. Table \ref{tab_reference_rq1}).
+tables (e.g. Table \ref{tab_reference_rq1}).  RQ5 answered in
+Section~\ref{ch_results}.
 
@@
 situations of software - by ourselves. There may be cases that we missed some
-information.
+information.  There also may be cases where we did not find evidence of
+something, like unit testing, not because the project didn't do it, but because
+no artifacts of this activity remained in the publicly available repository.
 \item As mentioned in Section \ref{ch_interview}, one interviewee was too busy

### 2021-11-30 StateOfPractice/Papers/MedImageSoft_SOP/MedImageSoft_SOP_AcademicRadiology.tex
First complete draft of MI SOP arXiv version

@@
 
-To gain insights, we devised five research questions, which can be applied
-to MI, as well as to other domains, of research software \citep{SmithEtAl2021}:
+To gain insights, we devised five research questions, which can be applied to
+MI, as well as to other domains, of research software \citep{SmithEtAl2021}. The
+first question below refers to artifacts.  We use this term to represent the
+documents, scripts and code that constitutes a software development project.
+Example artifacts include requirements, specifications, user manuals, unit
+tests, system tests, usability tests, build scripts, API (Application
+Programming Interface) documentation, READMEs, license documents, process
+documents, and code.
 
@@
 
-\begin{figure}[ht]
+\begin{figure}[!ht]
 \centering
@@
 
-% \begin{figure}[ht]
+% \begin{figure}[!ht]
 % \centering
@@
 	specified users to achieve specified goals with effectiveness, efficiency,
-	and satisfaction in a specified context of use" \citep{ISO/TR16982:2002}
+	and satisfaction in a specified context of use'' \citep{ISO/TR16982:2002}
 	\citep{ISO9241-11:2018}.
@@
 	used with or without adaptation in a problem solution other than the one for
-	which it was originally developed" \citep{kalagiakos2003non}.
+	which it was originally developed'' \citep{kalagiakos2003non}.
 
@@
 	enable the user to understand whether the software is suitable, and how it
-	can be used for particular tasks and conditions of use" \citep{iso2001iec}.
+	can be used for particular tasks and conditions of use'' \citep{iso2001iec}.
 
@@
 	another researcher can take the original code and input data, execute it,
-	and re-obtain the same result" \citep{BenureauAndRougier2017}.
+	and re-obtain the same result'' \citep{BenureauAndRougier2017}.
 \end{itemize}
@@
 
-\begin{figure}[ht]
+\begin{figure}[!ht]
 \includegraphics[scale=0.42]{figures/template.png}
@@
 automated testing, symbolic execution, model checking, unit tests, etc. We also
-examine whether the projects use continuous integration and continuous delivery
+examine whether the projects use Continuous Integration and Continuous Delivery
 (CI/CD). For verifiability, we go through the documents of the projects to check
@@
 \item \textbf{Surface Reliability} We check the following: 
-\begin{inparaenum}
+\begin{inparaenum}[i)]
     \item whether the software breaks during installation;
@@
 descending order of LOC. We found the initial release dates (Rlsd) for most
-projects and marked the two unknown dates with ``?". We used the date of the
+projects and marked the two unknown dates with ``?''. We used the date of the
 latest change to each code repository to decide the latest update. We found
@@
 
-\begin{table}[ht]
+\begin{table}[!ht]
 \centering
@@
 
-\begin{figure}[ht]
+\begin{figure}[!ht]
 \includegraphics[scale=0.38]{figures/installability_scores.png}
@@
 \citep{Hamill2004}.  We only found evidence of unit testing for about half of
-the projects.
-
-\begin{figure}[ht]
+the projects. We identified five projects using CI/CD tools: \textit{3D Slicer},
+\textit{ImageJ}, \textit{Fiji}, \textit{dwv}, and \textit{OHIF Viewer}.
+
+\begin{figure}[!ht]
 \includegraphics[scale=0.38]{figures/correctness_verifiability_scores.png}
@@
 
-We identified five projects using CI/CD tools: \textit{3D Slicer},
-\textit{ImageJ}, \textit{Fiji}, \textit{dwv}, and \textit{OHIF Viewer}.
-
 % In this section, the information about CI/CD tools is part of the answers to
@@
 Section~\ref{sec_result_installability}, most of the software products did not
-``break" during installation, or did not need installation; \textit{dwv} and
+``break'' during installation, or did not need installation; \textit{dwv} and
 \textit{GATE} broke in the building stage, and the processes were not
@@
 
-\begin{figure}[ht]
+\begin{figure}[!ht]
 \includegraphics[scale=0.38]{figures/reliability_scores.png}
@@
 
-\begin{figure}[ht]
+\begin{figure}[!ht]
 \includegraphics[scale=0.38]{figures/robustness_scores.png}
@@
 medical images that can be exchanged with the data and quality necessary for
-clinical use" \citep{MITA2021}. According to their documentation, all 29
+clinical use'' \citep{MITA2021}. According to their documentation, all 29
 software packages should support the DICOM standard. To test robustness, we
@@
 
-\begin{figure}[ht]
+\begin{figure}[!ht]
 \includegraphics[scale=0.38]{figures/usability_scores.png}
@@
 
-Table~\ref{tab_user_support_model} summarizes the user support models by the number
-of projects using them. We do not know whether the prevalent use of GitHub issues for user support is by design, or whether this just naturally happened as users sought help.
+Table~\ref{tab_user_support_model} summarizes the user support models by the
+number of projects using them. We do not know whether the prevalent use of
+GitHub issues for user support is by design, or whether this just naturally
+happened as users sought help.
 %  This table contains part of the answers to \hyperlink{rq2}{RQ2}.
 
-\begin{table}[ht]
+\begin{table}[!ht]
 \centering
@@
 
-\begin{figure}[ht]
+\begin{figure}[!ht]
 \includegraphics[scale=0.38]{figures/maintainability_scores.png}
@@
 
-\begin{figure}[ht]
+\begin{figure}[!ht]
 \includegraphics[scale=0.38]{figures/reusability_scores.png}
@@
 
-\begin{figure}[ht]
+\begin{figure}[!ht]
 \includegraphics[scale=0.38]{figures/understandability_scores.png}
@@
 
-\begin{figure}[ht]
+\begin{figure}[!ht]
 \includegraphics[scale=0.38]{figures/visibility_transparency_scores.png}
@@
 
-\begin{figure}[ht]
+\begin{figure}[!ht]
 \includegraphics[scale=0.38]{figures/overall_scores.png}
@@
 
-The measurement results in Section \ref{ch_results} are based on the information
-collected by ourselves. Such information is sufficient to measure the projects
-with reasonable efforts, but incomplete for us to understand the development
-process more deeply. For example, we usually cannot identify the following
-information in a project's documents: the pain points during the development,
-the threats to certain software qualities, and the developers' strategies to
-address them. We believe interviews with developers can collect the additional
-information we need. As a result, our method involves interviews with developers
-in a domain (Section \ref{sec_interview_methods}). We applied this step to the
-MI domain (Section \ref{sec_apply_to_mi_interviews}).
-
-In this section, we summarize some answers from the interviews. We highlight the
-answers that are the most informative and interesting in this section, and
-summarize the rest in Appendix \ref{ap_interview}.
-
-As mentioned in Section \ref{sec_apply_to_mi_interviews}, we contacted all 29
-teams. Some of them responded and participated in the interviews. Eventually, we
-interviewed nine developers from eight of the 29 MI software projects. The eight
-projects are \textit{3D Slicer}, \textit{INVESALIUS 3}, \textit{dwv},
-\textit{BioImage Suite Web}, \textit{ITK-SNAP}, \textit{MRIcroGL},
-\textit{Weasis}, and \textit{OHIF}. We spent about 90 minutes for each interview
-and asked 20 prepared questions. We also asked following-up questions when we
-thought it was worth diving deeper. One participant was too busy to have an
-interview, so they wrote down their answers. The interviewees may have provided
-multiple answers to each question. Thus, when counting the number of answers,
-the total result is sometimes larger than nine.
+The measurement results in Section \ref{ch_results} are based on information we
+collected from on-line resources. This information is incomplete because it
+doesn't generally capture the development process, the developer pain points,
+the perceived threats to software quality, and the developers' strategies to
+address these threats. To collect this information we interviewed nine
+developers from eight of the 29 MI software projects. The eight projects are
+\textit{3D Slicer}, \textit{INVESALIUS 3}, \textit{dwv}, \textit{BioImage Suite
+Web}, \textit{ITK-SNAP}, \textit{MRIcroGL}, \textit{Weasis}, and \textit{OHIF}.
+We spent about 90 minutes for each interview and asked 20 prepared questions
+(summarized in~\citet{SmithEtAl2021}). We also asked following-up questions as
+the conversation dictated. One participant was too busy to have an interview, so
+they wrote down their answers. The interviewees may have provided multiple
+answers to each question. Thus, when we summarize the frequency of different
+responses, the total is sometimes larger than nine.
+
+The full interview answers can be found in \citet{Dong2021}.  Selected questions
+and answers are highlighted and analyzed below.  The question numbers refer to
+the numbering in \citet{SmithEtAl2021}.
 
@@
 
-By asking questions \hyperlink{q9}{9}, \hyperlink{q10}{10}, and
-\hyperlink{q12}{12}, we tried to identify the pain points during the development
-process in the eight projects. The pain points include current and past
-obstacles. We also asked the interviewees how they would solve the problems.
-This section contains the answers to \hyperlink{rq4}{RQ4}. Questions 9, 10, and
-12 are as follows:
+The pain points include current and past obstacles, as explored via the
+following questions:
 
@@
 Table \ref{tab_obstacles} shows the number of times the interviewees mentioned
-the current and past obstacles in their projects.
-
-\begin{table}[ht]
+the current and past obstacles in their projects.   We group these pain points
+into three major categories of obstacles: \textit{resource}, \textit{balance},
+and \textit{testing}. We put the less mentioned ones into the category
+\textit{Others}.
+
+\begin{table}[!ht]
 \centering
@@
 
-The interviewees provided some potential and proven solutions for the problems
-in Table \ref{tab_obstacles}. We group these pain points into three major
-categories of obstacles: \textit{resource}, \textit{balance}, and
-\textit{testing}. We put the less mentioned ones into the category
-\textit{Others}. Sections \ref{sec_pain_points_1}, \ref{sec_pain_points_2}, and
-\ref{sec_pain_points_3} include further discussion about the three major groups
-of pain points and their solutions.
-
-\subsubsection{Resource Pain Points} \label{sec_pain_points_1}
-
-We summarize the pain points in the \textit{resource} category as
-\textbf{the lack of fundings and time.}
-
-The potential and proven solutions are:
-
-\noindent\textit{Potential solutions from interviewees:}
-\begin{itemize}
-\item when the team does not have enough developers for building new features
-and fixing bugs at the same time, shifting from development mode toward
-maintenance mode;
-\item licensing the software to commercial companies that integrate it into
+Developers identified a resource related pain point caused by a lack of funding
+and time.  Potential and proven solutions suggested by the interviewees include:
+
+\begin{itemize}
+\item Shifting from development to maintenance when the team does not have
+enough developers for building new features and fixing bugs at the same time;
+\item Licensing the software to commercial companies that integrate it into
 their products;
-\item better documentation to save time for answering users' and developers'
+\item Improving documentation to save time answering users' and developers'
 questions;
-\item supporting third-party plugins and extensions.
-\end{itemize}
-
-\noindent\textit{Proven solutions from interviewees:}
-
-\begin{itemize}
-\item GitHub Actions, which is a good CI/CD tool to save time.
-\end{itemize}
-
-Many interviewees thought lack of fundings and lack of time were the most
+\item Supporting third-party plugins and extensions; and,
+\item Using GitHub Actions for CI/CD to save time.
+\end{itemize}
+
+Many interviewees thought lack of fundings and lack of time were their most
 significant obstacles. The interviewees from \textit{3D Slicer} team and
@@
 problem, because they could hire more dedicated developers. On the other hand,
-the interviewee from the \textit{Weasis} team did not think that fundings could
+the interviewee from the \textit{Weasis} team did not feel that fundings could
 solve the same problem, since he still would need a lot of time to supervise the
-project.
-
-No interviewee suggested any solution to bring extra funding to the project.
-However, they provided ideas to save time, such as better documentation,
-third-party plugins, and good CI/CD tools.
-
-\subsubsection{Balance Pain Points} \label{sec_pain_points_2}
-
-We summarize the pain points in the \textit{balance} category as \textbf{the
-difficulty to balance between four factors: cross-platform compatibility,
-convenience to development \& maintenance, performance, and security.} They are
-also related to the choice between native application and web application.
-
-The potential and proven solutions are:
-
-\noindent\textit{Potential solutions from interviewees:}
-
-\begin{itemize}
-\item web applications that use computing power from computers GPU;
-\item to better support lower-end computers, adopting a web-based approach with backend servers;
-\item to better support lower-end computers, using memory-mapped files to
-consume less computer memory;
-\item more funding;
-\item maintaining better documentations to ease the development \& maintenance
+project.  No interviewee suggested any solution to bring extra funding to the
+project. However, they provided ideas to save time, such as better
+documentation, third-party plugins, and good CI/CD tools.
+
+With respect to balance, developers expressed difficulty balancing between four
+factors: cross-platform compatibility, convenience to development \&
+maintenance, performance, and security.  The potential and proven solutions are:
+
+\begin{itemize}
+\item Adopting a web-based approach with backend servers, to better support
+lower-end computers;
+\item Using memory-mapped files to consume less computer memory, to better
+support lower-end computers, 
+\item Using computing power from the computers GPU for web applications;
+\item Increasing funding;
+\item Maintaining better documentations to ease the development \& maintenance
 processes;
-\end{itemize}
-
-\noindent\textit{Proven solutions from interviewees:}
-
-\begin{itemize}
-\item one interviewee saw the performance problem disappeared over the years
-when computers became more and more powerful. 
-\end{itemize}
-
-Table \ref{tab_native_vs_web} shows the teams' choices between native
-application and web application. In all the 29 teams on our list, most of them
-chose to develop native applications. For the eight teams we interviewed, three
-of them were building web applications, and the \textit{MRIcroGL} team was
+\item Improving performance via more powerful computers, which one interviewee
+pointed out has already happened to reduce the balance problem.
+\end{itemize}
+
+As the above list shows, web-based applications are perceived to help the
+balance problem.  Table~\ref{tab_native_vs_web} shows the teams' choices between
+native application and web application. In all the 29 teams on our list, most of
+them chose to develop native applications. For the eight teams we interviewed,
+three of them were building web applications, and the \textit{MRIcroGL} team was
 considering web-based solutions. So we had a good chance to discuss the
@@
 
-The interviewees talked about the advantages and disadvantages of the two
-choices. We summarize the opinions from the interviewees in Table
-\ref{tab_pros_cons_native_vs_web}.
-
-\begin{table}[ht]
-\centering
-\hspace*{-2cm}\begin{tabular}{lll}
-\hline
- & Native application & Web application \\ \hline
-Ad & - higher performance & \begin{tabular}[c]{@{}l@{}}- easy to achieve cross-platform compatibility\\ - simpler build process\end{tabular} \\ \hdashline
-Disad & \begin{tabular}[c]{@{}l@{}}- hard to achieve cross-platform compatibility\\ - more complicated build process\end{tabular} & \begin{tabular}[c]{@{}l@{}}\textit{Without a backend:} \\ - lower performance\\ \textit{With a backend:} \\ - harder for privacy protection \\ - extra cost for backend servers \end{tabular} \\ \hline
-\end{tabular}
-\caption{\label{tab_pros_cons_native_vs_web}Advantages and disadvantages of native application and web application}
-\end{table}
-
-\subsubsection{Testing Pain Point} \label{sec_pain_points_3}
-
-The pain point in the \textit{testing} category is \textbf{the lack of access to
-real-world datasets for testing.} The information about software testing in this
-section is part of the answers to \hyperlink{rq3}{RQ3}.
-
-The potential and proven solutions are:
-
-\noindent\textit{Potential solutions from interviewees:}
-
-\begin{itemize}
-\item using open datasets
-\end{itemize}
-
-\noindent\textit{Proven solutions from interviewees:}
-
-\begin{itemize}
-\item asking the users to provide deidentified copies of medical images if they
+The advantage for native applications is higher performance, while web
+applications have the advantage of cross-platform compatibility and a simpler
+build process.  These web advantages are mirrored by the native application
+disadvantages of difficulty with cross-platform compatibility and a complex
+build process.  The lower performance disadvantage of web-applications can be
+improved with a server backend, but in this case there are disadvantages for
+privacy protection and the cost of the servers.
+
+For the pain point category of testing developers identified the the lack of access to
+real-world datasets.  Some proposed and proven solutions are as follows:
+
+\begin{itemize}
+\item Using open datasets
+\item Asking the users to provide de-identified copies of medical images if they
 have problems loading the images;
-\item sending the beta versions of software to medical workers who can access
+\item Sending the beta versions of software to medical workers who can access
 the data and complete the tests;
-\item if (part of) the team belongs to a medical school or a hospital, using the
+\item If (part of) the team belongs to a medical school or a hospital, using the
 datasets they can access;
-\item if the team has access to MRI scanners, self-building sample images for
+\item If the team has access to MRI scanners, self-building sample images for
 testing;
-\item if the team has connections with MI equipment manufacturers, asking for
+\item If the team has connections with MI equipment manufacturers, asking for
 their help on data format problems;
-\item storing all images that cause special problems, and maintaining this
+\item Storing all images that cause special problems, and maintaining this
 special dataset over time.
@@
 
-No interviewee provided a perfect way to solve this problem. However,
-connections between the development team and medical professionals/institutions
-could ease the pain.
+No interviewee provided a perfect solution to the testing problem. However,
+increased connections between the development team and medical
+professionals/institutions could ease the pain.
 
@@
 We tried to understand the interviewees' opinions on documentation and the
-quality of documentations with questions 11 and 19. The information about
-documentation is part of the answers to \hyperlink{rq3}{RQ3}.
+quality of documentations with two questions:
 
@@
 
-Table \ref{tab_opinion_doc} summarizes interviewees' opinions on documentation.
+Table~\ref{tab_opinion_doc} summarizes interviewees' opinions on documentation.
 Interviewees from each of the eight projects thought that documentation was
@@
 
-Table \ref{tab_doc_tools} lists some of the documentation tools and methods
-mentioned by the interviewees. This table contains part of the answers to
-\hyperlink{rq2}{RQ2}.
-
-\begin{table}[ht]
-\centering
-\begin{tabular}{ll}
-\hline
-Tool or method for documentation & Num ans. \\ \hline
-Forum discussions & 3 \\
-Videos & 3 \\
-GitHub & 2 \\
-Mediawiki / wiki pages & 2 \\
-Workshops & 2 \\
-Social media & 2 \\
-Writing books & 1 \\
-Google Form & 1 \\
-State management & 1 \\ \hline
-\end{tabular}
-\caption{\label{tab_doc_tools}Tools and methods for documentation by the numbers
-of interviewees with the answers}
-\end{table}
-
-\subsection{Contribution Management and Project Management}
-\label{sec_contribution_pm}
-We tried to understand how the teams managed the contributions and their
-projects by asking the following questions:
+For documentation tools and methods mentioned by the interviewees, the most
+popular (mentioned by about 30\% of developers) were forum discussions and
+videos.  The second most popular options (mentioned by about 20\% of developers)
+were GitHub, wiki pages, workshops, and social media.  The least frequently
+mentioned options (about 10\% of developers) included writing books, google
+forms and state management.
+
+\subsection{Project Management} \label{sec_contribution_pm}
+
+To investigate project management we asked three questions:
 
@@
 agile, etc.
-\item[Q14.] \hypertarget{q14}What is your project management process? Do you
+\item[Q14.] What is your project management process? Do you
 think improving this process can tackle the current problem? Were any project
@@
 
-Although some team may have a documented process for accepting new
-contributions, no one talked about it during the interview. However, most of
-them mentioned using GitHub and pull requests to manage contributions from the
-community. The interviewees generally gave very positive feedback on using
-GitHub. Some also said they had handled the project repository with some other
-tools, and eventually transferred to git and GitHub. Table
-\ref{tab_method_receive_contributions} shows the number of times the
-interviewees mentioned the methods of receiving contributions. This table
-contains part of the answers to \hyperlink{rq2}{RQ2}.
-
-\begin{table}[ht]
-\centering
-\begin{tabular}{lll}
-\hline
-\multirow{2}{*}{Method of receiving contributions} & \multicolumn{2}{l}{Num ans.} \\ \cline{2-3} 
- & current & past \\ \hline
-GitHub with pull requests & 8 & \\
-Code contributions from emails & & 3 \\
-Code contributions from forums & & 1 \\
-Sharing the git repository by email & & 1 \\ \hline
-\end{tabular}
-\caption{\label{tab_method_receive_contributions}Methods of receiving
-contributions by the numbers of interviewees with the answers}
-\end{table}
+Although teams may have a process for accepting new contributions, no one
+discussed this during their interview. However, most teams (8 of 9) mentioned
+using GitHub and pull requests to manage contributions from the community. The
+interviewees generally gave very positive feedback on using GitHub. Some teams
+previously used a different approach to version control and eventually
+transferred to git and GitHub.  The past approaches included contributions
+from e-mail (3 teams), contributions from forums (1 team) and sharing the
+git repository via e-mail (1 team).  \wss{I do not know what this last one means.
+Should ask Ao.}
 
 For managing contributions, the \textit{3D Slicer} team encouraged users to
-develop their extensions for specific use cases, and the \textit{OHIF} team was
-trying to enable the use of plug-ins; the interviewee from the \textit{ITK-SNAP}
-team said one way of accepting new team members was through funded academic
-projects.
-
-Table \ref{tab_developmen_models} shows the software development models by the
-numbers of interviewees with the answers. Only two interviewees confirmed their
-development models. The others did not think they used a specific model, but
-three of them suggested that their processes were similar to Waterfall or Agile.
-
-\begin{table}[ht]
-\centering
-\begin{tabular}{ll}
-\hline
-Software development model & Num ans. \\ \hline
-Undefined/self-directed & 3 \\
-Similar to Agile & 2 \\
-Similar to Waterfall & 1 \\
-Agile & 1 \\
-Waterfall & 1 \\ \hline
-\end{tabular}
-\caption{\label{tab_developmen_models}Software development models by the numbers
-of interviewees with the answers}
-\end{table}
-
-Some interviewees mentioned the project management tools they used, which are in
-Table \ref{tab_pm_tools}. This table contains part of the answers to
-\hyperlink{rq2}{RQ2}. Generally speaking, the interviewees talked about two
-types:
-
-\begin{itemize}
-\item Trackers, including GitHub, issue trackers, bug trackers and Jira;
-\item Documents, including GitHub, Wiki page, Google Doc, and Confluence.
-\end{itemize}
-
-\begin{table}[ht]
-\centering
-\begin{tabular}{ll}
-\hline
-Project management tools & Num ans. \\ \hline
-GitHub & 3 \\
-Issue trackers & 1 \\
-Bug trackers & 1 \\
-Jira & 1 \\
-Wiki page & 1 \\
-Google Doc & 1 \\
-Confluence & 1 \\ \hline
-\end{tabular}
-\caption{\label{tab_pm_tools}Project management tools by the numbers of
-interviewees with the answers}
-\end{table}
+develop their extensions for specific use cases, while the \textit{OHIF} team
+enabled the use of plug-ins.  The interviewee from the \textit{ITK-SNAP} team
+said one way of accepting new team members was through funded academic projects.
+
+For the most part, the responses about development model were vague, with only
+two interviewees following a definite development model. In some cases the
+interviewees felt their process was similar to an existing development model.
+Three teams (about 38\%) either followed agile, or something similar to agile.
+Two teams (25\%) either followed a waterfall process, or something similar.
+Three teams (about 38\%) explicitly stated that their process was undefined or
+self-directed.
+
+Some interviewees mentioned the project management tools they used. Generally
+speaking, the interviewees talked about two types of tools:
+\begin{inparaenum}[i)]
+\item trackers, including GitHub, issue trackers, bug trackers and Jira; and,
+\item documentation tools, including GitHub, Wiki page, Google Doc, and Confluence.
+\end{inparaenum}
+Of the specifically named tools in the above lists, GitHub was mentioned 3
+times, while each of the other tools was only mentioned once.
 
 No interviewee introduced any strictly defined project management process. The
-most common way was following the issues, such as bugs and feature requests.
+most common approach was following the issues, such as bugs and feature requests.
 Additionally, the \textit{3D Slicer} team had weekly meetings to discuss the
@@
 mentioned that the team has a project manager; the \textit{3D Slicer} team and
-\textit{BioImage Suite Web} team were doing nightly builds and tests.
-
-Most interviewees skipped the second part of \hyperlink{q14}{Q14} ``Do you think
-improving this process can tackle the current problem?". In retrospect, we
-should not have asked a yes-or-no question, since it is not very informative.
-The interviewee from the \textit{OHIF} team gave a positive answer to this
-question. They believed that a better project management process can improve the
-efficiency of junior developers. They also improved the project management tools
-(from public Jira to public GitHub repository plus private Jira), so they could
-better communicate externally and internally.
-
-The information about contribution management and project management in this
-section is part of the answers to \hyperlink{rq3}{RQ3}.
+\textit{BioImage Suite Web} team do nightly builds and tests.
+
+Most interviewees skipped the part of the last question asking ``Do you think
+improving this process can tackle the current problem?''. In retrospect, we
+should not have asked a yes-or-no question, since a binary answer is not very
+informative. The interviewee from the \textit{OHIF} team gave a positive answer
+to this question. They believed that a better project management process can
+improve the efficiency of junior developers. They also improved the project
+management tools (from public Jira to public GitHub repository plus private
+Jira), so they could better communicate externally and internally.
 
@@
 
-Questions 15--18, and 20 are about the software qualities of
-\textit{correctness}, \textit{maintainability}, \textit{understandability},
-\textit{usability}, and \textit{reproducibility}, respectively. We asked these
-questions to understand the threats to these qualities and the developers'
-strategies to improve them. We discuss each quality in a separate section below.
-This section contains part of the answers to \hyperlink{rq5}{RQ5}.
-
-\subsubsection{Correctness} \label{sec_interview_correctness}
+Questions 15--18, and 20 cover the software qualities of correctness,
+maintainability, understandability, usability, and reproducibility,
+respectively. We discuss each quality below.
 
@@
 
-Table \ref{tab_q15_threats_correctness} shows the threats to
-\textit{correctness} by the numbers of interviewees with the answers.
-
-\begin{table}[ht]
-\centering
-\hspace*{-1cm}\begin{tabular}{ll}
-\hline
-Threat to correctness & Num ans. \\ \hline
-Complexity - data in various formats and complicated standards. & 2 \\
-Complexity - different MI machines create data in different ways. & 2 \\
-Complexity - additional functions besides viewing. & 1 \\
-The lack of real word image data for testing. & 1 \\
-The team cannot use private data for debugging even when the data cause
-problems. & 1 \\
-With huge datasets for testing, the tests are expensive and time-consuming. & 1 \\
-It is hard to manage releases. & 1 \\
-The project has no unit tests. & 1 \\
-The project has no dedicated quality assurance team. & 1 \\ \hline
-\end{tabular}
-\caption{\label{tab_q15_threats_correctness}Threats to correctness by the
-numbers of interviewees with the answers}
-\end{table}
-
-Table \ref{tab_q15_strategies_correctness} shows the strategies to ensure
-\textit{correctness} by the numbers of interviewees with the answers. The
-interviewees from the \textit{3D Slicer} and \textit{ITK-SNAP} teams thought
-that the self-tests and automated tests were beneficial and could significantly
-save time. The interviewee from the \textit{Weasis} team kept collecting medical
-images for more than ten years. These images have caused problems with the
-software. So he had samples to test specific problems.
-
-\begin{table}[ht]
-\centering
-\hspace*{-1.5cm}\begin{tabular}{ll}
-\hline
-Strategy to ensure correctness & Num ans. \\ \hline
-Test-driven development, component tests, integration tests, smoke tests, regression tests. & 4 \\
-Self tests / automated tests. & 3 \\
-Two stage development process / stable release \& nightly builds. & 3 \\
-CI/CD. & 1 \\
-Using deidentified copies of medical images for debugging. & 1 \\
-Sending beta versions to medical workers who can access the data to do the tests. & 1 \\
-Collecting and maintaining a dataset of problematic images. & 1 \\ \hline
-\end{tabular}
-\caption{\label{tab_q15_strategies_correctness}Strategies to ensure correctness
-by the numbers of interviewees with the answers}
-\end{table}
-
-The information about software testing in this section is part of the answers to
-\hyperlink{rq3}{RQ3}.
-
-\subsubsection{Maintainability} \label{sec_interview_maintainability}
+The interviewees identified multiple threats to correctness.  The most
+frequently mentioned threat was complexity.  Complexity enters the software by
+various means, including a variety of data formats, complicated data standards,
+differing outputs between medical imaging machines, and the addition of
+(non-viewing related) functionality.  Other threats to correctness identified by at least one time
+include the following:
+\begin{itemize}
+\item Lack of real world image data for testing;
+\item The team cannot use private data for debugging even when the data causes
+problems; \wss{Why not?  Should ask Ao}
+\item Expense and time consumed by tests because of the huge datasets;
+\item Difficulty managing releases;
+\item No unit tests; and,
+\item No dedicated quality assurance team.
+\end{itemize}
+
+Testing was the most often mentioned strategy for ensuring correctness.  Seven
+teams mentioned test related activities, including test-driven development,
+component tests, integration tests, smoke tests, regression tests, self tests
+and automated tests.  Another approach frequently adopted (mentioned by 3
+interviewees) is a two state development process with stable releases and
+nightly builds.  Other strategies for ensuring correctness that came up during
+the interviews include CI/CD, using de-identified copies of medical images for
+debugging, sending beta versions to medical workers who can access the data to
+do the tests, and collecting/maintaining a dataset of problematic images.
 
 \begin{description}
-\item[Q16.] \hypertarget{q16}When designing the software, did you consider the
-ease of future changes? For example, will it be hard to change the system’s
-structure, modules, or code blocks? What measures have been taken to ensure the
-ease of future changes and maintains?
+\item[Q16.] When designing the software, did you consider the ease of future
+changes? For example, will it be hard to change the system’s structure, modules,
+or code blocks? What measures have been taken to ensure the ease of future
+changes and maintains?
 \end{description}
 
-Table \ref{tab_q16_strategies_maintainability} shows the strategies to ensure
-\textit{maintainability} by the numbers of interviewees with the answers. The
-modular approach is the most talked-about solution to improve
-\textit{maintainability}. The \textit{3D Slicer} team used a well-defined
-structure for the software, which they named as ``event-driven MVC pattern''.
-Moreover, \textit{3D Slicer} discovers and loads necessary modules at runtime,
-according to the configuration and installed extensions. The \textit{BioImage
-Suite Web} team had designed and re-designed their software multiple times in
-the last 10+ years. They found that their modular approach effectively supported
-the maintainability \citep{Joshi2011}.
-
-\begin{table}[ht]
-\centering
-\begin{tabular}{ll}
-\hline
-Strategy to ensure maintainability & Num ans. \\ \hline
-Modular approach / maintain repetitive functions as libraries. & 5 \\
-Supporting third-party extensions. & 1 \\
-Easy-to-understand architecture. & 1 \\
-Dedicated architect. & 1 \\
-Starting from simple solutions. & 1 \\
-Documentation. & 1 \\ \hline
-\end{tabular}
-\caption{\label{tab_q16_strategies_maintainability}Strategies to ensure
-maintainability by the numbers of interviewees with the answers}
-\end{table}
-
-\subsubsection{Understandability} \label{sec_interview_understandability}
+The most popular, with five out of nine interviewees mentioning it, strategy for
+ensuring maintainability was to use a modular approach, with often repeated
+functions in a library.  Other strategies that were mentioned for improving
+maintainability include supporting third-party extensions, an easy-to-understand
+architecture, a dedicated architect, starting from simple solutions, and
+documentation.  The \textit{3D Slicer} team used a well-defined structure for
+the software, which they named as an ``event-driven MVC pattern''. Moreover,
+\textit{3D Slicer} discovers and loads necessary modules at runtime, according
+to the configuration and installed extensions. The \textit{BioImage Suite Web}
+team had designed and re-designed their software multiple times in the last 10+
+years. They found that their modular approach effectively supports
+maintainability \citep{Joshi2011}.
 
 \begin{description}
-\item[Q17.] \hypertarget{q17}Provide instances where users have misunderstood
-the software. What, if any, actions were taken to address understandability
-issues?
+\item[Q17.] Provide instances where users have misunderstood the software. What,
+if any, actions were taken to address understandability issues?
 \end{description}
 
-Table \ref{tab_q17_threats_understandability} shows the threats to
-\textit{understandability} by the numbers of interviewees with the answers. It
-separates \textit{understandability} issues to users and developers by the
-horizontal dash line.
-
-\begin{table}[ht]
-\centering
-\begin{tabular}{ll}
-\hline
-Threat to understandability & Num ans. \\ \hline
-Not all users understand how to use some features. & 2 \\
-The team has no dedicated user experience (UX) designer. & 1 \\
-Some important indicators are not noticeable (e.g. a progress bar). & 1 \\
-Not all users understand the purpose of the software. & 1 \\
-Not all users know if the software includes certain features. & 1 \\
-Not all users understand how to use the command line tool. & 1 \\
-Not all users understand that the software is a web application. & 1 \\\hdashline
-Not all developers understand how to deploy the software. & 1 \\
-The architecture is difficult for new developers to understand. & 1 \\ \hline
-\end{tabular}
-\caption{\label{tab_q17_threats_understandability}Threats to understandability by the numbers of interviewees with the answers}
-\end{table}
-
-Table \ref{tab_q17_strategies_understandability} shows the strategies to ensure \textit{understandability} by the numbers of interviewees with the answers.
-
-\begin{table}[ht]
-\centering
-\begin{tabular}{ll}
-\hline
-Strategy to ensure understandability & Num ans. \\ \hline
-Documentation / user manual / user mailing list / forum. & 4 \\
-Graphical user interface. & 2 \\
-Testing every release with active users. & 1 \\
-Making simple things simple and complicated things possible. & 1 \\
-Icons with more clear visual expressions. & 1 \\
-Designing the software to be intuitive. & 1 \\
-Having a UX designer with the right experience. & 1 \\
-Dialog windows for important notifications. & 1 \\
-Providing an example if the users need to build the software by themselves. & 1
-\\ \hline
-\end{tabular}
-\caption{\label{tab_q17_strategies_understandability}Strategies to ensure understandability by the numbers of interviewees with the answers}
-\end{table}
-
-\subsubsection{Usability} \label{sec_interview_usability}
+The discussion with the developers focused on understandability issues for two
+classes of users: the end users and other developers.  The threats to
+understandability for end users include users not understanding how to use
+features (\wss{Because the users don't have sufficient background?  Ask Ao.}),
+the team not having a dedicated user experience (UX) designer, some important
+indicators are not noticeable (e.g. a progress bar), not all users understand
+the purpose of the software, not all users know if the software includes certain
+features, not all users understand how to use the command line tool, not all
+users understand that the software is a web application. For developers the
+threats to understandability include developers not understanding how to deploy
+the software (\wss{This sounds like a consequence of a lack of
+understandability, rather than threat to it.  Or is the point that developers
+lack sufficient background?  Ask Ao}), and the architecture is difficult for new
+developers to understand.
+
+The most common strategy, cited by 4 developers, for ensuring understandability
+was to use documentation (user manuals, mailing lists, forums).  Other suggested
+and practiced strategies include a graphical user interface, testing every
+release with active users, making simple things simple and complicated things
+possible, icons with clear visual expressions, designing the software to be
+intuitive, having a UX designer with the right experience, dialog windows for
+important notifications, and providing an example for users to follow.
 
 \begin{description}
-\item[Q18.] \hypertarget{q18}What, if any, actions were taken to address
-usability issues?
+\item[Q18.] What, if any, actions were taken to address usability issues?
 \end{description}
 
-Table \ref{tab_q18_strategies_usability} shows the strategies to ensure
-\textit{usability} by the numbers of interviewees with the answers. The
-information about software testing in this table is part of the answers to
-\hyperlink{rq3}{RQ3}.
-
-\begin{table}[ht]
-\centering
-\begin{tabular}{ll}
-\hline
-Strategy to ensure usability & Num ans. \\ \hline
-Usability tests and interviews with end users. & 3 \\
-Adjusting according to users' feedbacks. & 3 \\
-Straightforward and intuitively designed interface / professional UX designer. & 2 \\
-Providing step-by-step processes, and showing the step numbers. & 1 \\
-Making the basic functions easy to use without reading the documentation. & 1 \\
-Focusing on limited number of functions. & 1 \\
-Making the software more streamlined. & 1 \\
-Downsampling images to consume less memory. & 1 \\
-An option to load only part of the data to boost performance. & 1 \\ \hline
-\end{tabular}
-\caption{\label{tab_q18_strategies_usability}Strategies to ensure usability by
-the numbers of interviewees with the answers}
-\end{table}
-
-\subsubsection{Reproducibility} \label{sec_interview_reproducibility}
+The two most frequently mentioned strategies (mentioned 3 times each) for
+improving usability are:
+\begin{inparaenum}[i)]
+    \item usability tests and interviews with end users; and,
+    \item adjusting the software according to user feedback.
+\end{inparaenum}
+Other ideas that came up include a straightforward and intuitively designed
+interface / professional UX designer, providing step-by-step processes, making
+the basic functions easy to use without reading the documentation, focusing on
+limited number of functions, making the software more streamlined, downsampling
+images to consume less memory, and an option to load only part of the data to
+boost performance.
 
@@
 
-Table \ref{tab_q20_threats_reproducibility} shows the threats to
-\textit{reproducibility} by the numbers of interviewees with the answers.
-
-\begin{table}[ht]
-\centering
-\begin{tabular}{ll}
-\hline
-Threat to reproducibility & Num ans. \\ \hline
-If the software is closed-source, the reproducibility is hard to achieve. & 1 \\
-The project has no user interaction tests. & 1 \\
-The project has no unit tests. & 1 \\
-Using different versions of some common libraries may cause problems. & 1 \\
-CPU variability can leads to non-reproducibility. & 1 \\
-The team may misinterpret how manufacturers create medical images. & 1 \\ \hline
-\end{tabular}
-\caption{\label{tab_q20_threats_reproducibility}Threats to reproducibility by
-the numbers of interviewees with the answers}
-\end{table}
-
-Table \ref{tab_q20_strategies_reproducibility} shows the strategies to ensure
-\textit{ reproducibility} by the numbers of interviewees with the answers. The
-interviewee from the \textit{3D Slicer} team provided various suggestions. One
-interviewee from another team suggested that they used \textit{3D Slicer} as the
-benchmark to test their \textit{reproducibility}.
-
-\begin{table}[ht]
-\centering
-\begin{tabular}{ll}
-\hline
-Strategy to ensure reproducibility & Num ans. \\ \hline
-Regression tests / unit tests / having good tests. & 6 \\
-Making code, data, and documentation available / OSS / open-source libraries. & 5 \\
-Running same tests on all platforms. & 1 \\
-A dockerized version of the software,  insulating it from the OS environment. & 1 \\
-Using standard libraries. & 1 \\
-Monitoring the upgrades of the libraries. & 1 \\
-Clearly documenting the versions. & 1 \\
-Bringing along the exact versions of all the dependencies with the software. & 1 \\
-Providing checksums of the data. & 1 \\
-Benchmarking the software against other software with similar purposes. & 1 \\ \hline
-\end{tabular}
-\caption{\label{tab_q20_strategies_reproducibility}Strategies to ensure
-reproducibility by the numbers of interviewees with the answers}
-\end{table}
-
-The information about software testing in this section is part of the answers to
-\hyperlink{rq3}{RQ3}.
+We discussed threats to reproducibility and strategies for improving it.  The
+threats that were mentioned include closed-source software, no user interaction
+tests, no unit tests, using different versions of some common libraries
+\wss{because the interface and/or behaviour of the library could have changed?
+Ask Ao}, variability between CPUs, and misinterpretation of how manufacturers
+create medical images. The most commonly cited (by 6 teams) strategy to improve
+reproducibility was testing (regression tests, unit tests, having good tests).
+The second most common strategy (mentioned by 5 teams) is making code, data, and
+documentation available, possibly by creating open-source libraries.  Other
+ideas that were mentioned include running the same tests on all platforms, a
+dockerized version of the software to insulate it from the OS environment, using
+standard libraries, monitoring the upgrades of the library dependencies, clearly
+documenting the version information, bringing along the exact versions of all
+the dependencies with the software, providing checksums of the data, and
+benchmarking the software against other software that overlaps in functionality.
+Specifically one interviewee suggested using \textit{3D Slicer} as the benchmark
+to test their reproducibility.
 
@@
 
-This section answers our research questions from Section
-\ref{sec_research_questions}.  Sections \ref{sec_rq_artifacts}--
-\ref{sec_rq_comparison} summarize the answers to the six questions,
-respectively. Section \ref{sec_threats_to_validity} lists the threats to the
-validity of our research. The answers are based on our quality measurements in
-Section \ref{ch_results} and developer interviews in Section \ref{ch_interview}.
-We refer to these sections to avoid repetition, and organize the references in
-tables (e.g. Table \ref{tab_reference_rq1}).  RQ5 answered in
-Section~\ref{ch_results}.
-
-\subsection{Artifacts in the Projects} \label{sec_rq_artifacts}
+This section answers the research questions from
+Section~\ref{sec_motivation}.  We will go through each of the questions
+in turn.
+
 \begin{description}
@@
 
-We answer this question by examining the documentation, scanning the source
-code, and interviewing the developers of the projects. Table
-\ref{tab_reference_rq1} shows the sections and tables containing answers to this
-research question.
-
-\begin{table}[ht]
-\centering
-\begin{tabular}{ll}
-\hline
-Section or table & Description \\ \hline
-Table \ref{tab_maintainability_docs} & Maintainability documents \\
-Table \ref{tab_Visibility/Transparency_docs} & Visibility/transparency documents \\ \hline
-\end{tabular}
-\caption{\label{tab_reference_rq1}Sections and tables with answers to RQ1}
-\end{table}
-
-As mentioned in Section \ref{sec_grading_template}, we search for all the
-artifacts in a project when measuring \textit{maintainability}. The detailed
-records of the existing artifacts in the 29 MI projects are at
-\hyperlink{https://data.mendeley.com/datasets/k3pcdvdzj2/1}{https://data.mendeley.com/datasets/k3pcdvdzj2/1}.
-Table \ref{tab_artifacts_frequency} summarizes the frequency of the artifacts.
-This table also contains part of the answers to \hyperlink{rq3}{RQ3}.
-
-\begin{table}[ht]
-\centering
-\begin{tabular}{ll}
-\hline
-Artifact & Number of projects \\ \hline
-README & 29 \\
-Version Control & 29 \\
-License & 28 \\
-Bug tracker & 28 \\
-Change request & 28 \\
-User Manual & 22 \\
-Release notes & 22 \\
-Build file & 18 \\
-Tutorials & 18 \\
-Installation Guide & 16 \\
-Test cases & 15 \\
-Authors & 14 \\
-FAQ & 14 \\
-Acknowledgements & 12 \\
-Executable files & 10 \\
-Developer's Manual & 8 \\
-API documentation & 7 \\
-Troubleshooting guide & 6 \\
-Project Plan & 5 \\ \hline
-\end{tabular}
-\caption{\label{tab_artifacts_frequency}Artifacts by their frequency in the 29 MI projects}
-\end{table}
-
-We summarized the definitions of the artifacts in
-\hyperlink{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Artifacts_MiningV3.xlsx}{https://github.com/.../Artifacts\_MiningV3.xlsx}.
-Source code is a type of artifact. Since we only included OSS on the final list
-(Section \ref{sec_mi_software_selection}), every project's source code was
-available. Thus, we excluded it in the above table.
-
-\subsection{Tools in the Projects}
-\label{sec_rq_tools}
+We answered this question by examining and analyzing the repositories for each
+of the 29 projects.  In particular, as part of assessing maintainability, we
+listed the artifacts (documents, test scripts etc.) that appear in each project.
+The full measurement results are summarized in \citet{Dong2021-Data}.  The
+artifacts found were as follows (with the number of occurrences in brackets):
+Source code (29), README (29), Version Control (29), License (28), Bug tracker
+(28), Change request (28), User Manual (22), Release notes (22), Build file
+(18), Tutorials (18), Installation Guide (16), Test cases (15), Authors (14),
+FAQ (14), Acknowledgements (12), Executable files (10), Developer's Manual (8),
+API documentation (7), Troubleshooting guide (6), Project Plan (5).  The
+traceability of some document types to the projects in which they occur was
+previously presented in Table~\ref{tab_maintainability_docs} for maintainability
+related documents, and Table~\ref{tab_Visibility/Transparency_docs} for
+visibility/transparency related documents.
+
+The list in the previous paragraph shows that some types of software engineering
+recommended artifacts are missing in MI projects.  For instance, we did not find
+evidence of user stories (an Agile software development tool for capturing
+software features from an end-user perspective), user requirements, test plans,
+test code coverage, etc.  Our preliminary work suggests that this is not unique
+to MI software.  A gap exists in scientific computing development practices and
+software engineering recommendations \citep{OwojaiyeEtAl2021_CSE}.
+
 \begin{description}
@@
 We answer this question by measuring the qualities and interviewing the
-developers. This section summarizes the tools used for CI/CD, user support,
-version control, documentation, contribution management, and project management.
-Table \ref{tab_reference_rq2} shows the sections and tables containing answers
-to this research question.
-
-\begin{table}[ht]
-\centering
-\begin{tabular}{ll}
-\hline
-Section or table & Description \\ \hline
-Section \ref{sec_result_correctness_verifiability} & CI/CD tools \\
-Table \ref{tab_user_support_model} & User support tools \\
-Section \ref{sec_score_maintainability} & Version control tools \\
-Table \ref{tab_doc_tools} & Documentation tools \\
-Table \ref{tab_method_receive_contributions} & Contribution management tools \\
-Table \ref{tab_pm_tools} & Project management tools\\ \hline
-\end{tabular}
-\caption{\label{tab_reference_rq2}Sections and tables with answers to RQ2}
-\end{table}
+developers. We learned that developers use tools for CI/CD
+(Section~\ref{sec_result_correctness_verifiability}), user support
+(Table~\ref{tab_user_support_model}), version control
+(Section~\ref{sec_score_maintainability}), documentation
+(Section~\ref{sec_contribution_pm}), contribution management
+(Section~\ref{sec_contribution_pm}), and project management
+(Section~\ref{sec_contribution_pm}).
 
 As mentioned in Section \ref{sec_result_correctness_verifiability}, we
-identified five projects using CI/CD tools. \textit{3D Slicer} and \textit{OHIF
-Viewer} used CircleCI; \textit{ImageJ}, \textit{Fiji}, and \textit{dwv} used
+identified five projects using CI/CD tools: \textit{3D Slicer} and \textit{OHIF
+Viewer} use CircleCI; \textit{ImageJ}, \textit{Fiji}, and \textit{dwv} use
 Travis. We identified the above projects and tools by examining the
-documentation and source code of all projects. Thus, we may missed some projects
-or tools. According to the interviews with developers, \textit{dwv} and
+documentation and source code of all projects. Not all tools leave identifiable
+artifacts in the repositories; therefore, we may have missed the use of some
+tools. According to the interviews with developers, \textit{dwv} and
 \textit{Weasis} used GitHub Actions.
 
-\subsection{Principles, Processes, and Methodologies in the Projects}
-\label{sec_rq_PPM}
 \begin{description}
@@
 We answer this question by measuring the qualities and interviewing the
-developers. This section shows the principles, processes, and methodologies in
-software testing, documentation, contribution management, project management,
-and improving software qualities. Table \ref{tab_reference_rq3} shows the
-sections and tables containing answers to this research question.
-
-\begin{table}[ht]
-\centering
-\begin{tabular}{ll}
-\hline
-Section or table & Description \\ \hline
-Section \ref{fg_correctness_erifiability_scores}, \ref{sec_pain_points_3},
-\ref{sec_interview_correctness}, and \ref{sec_interview_reproducibility}; Table
-\ref{tab_q18_strategies_usability} & Software testing \\
-Section \ref{fg_correctness_erifiability_scores} and
-\ref{sec_interview_documents}; Table \ref{tab_maintainability_docs},
-\ref{tab_Visibility/Transparency_docs}, and \ref{tab_artifacts_frequency} &
-Documentation \\
-Section \ref{sec_contribution_pm} & Contribution management \\
-Section \ref{sec_contribution_pm} & Project management \\ \hline
-\end{tabular}
-\caption{\label{tab_reference_rq3}Sections and tables with answers to RQ3}
-\end{table}
+developers. We previously summarized principles, processes, and methodologies in
+software testing (Section~\ref{sec_interview_pain_points} and
+Section~\ref{sec_interview_software_qualities}), documentation
+(Section~\ref{sec_interview_documents}, Table~\ref{tab_maintainability_docs},
+Table~\ref{tab_Visibility/Transparency_docs}, and Section~\ref{ch_answers}),
+contribution management (Section~\ref{sec_contribution_pm}), and project
+management (Section~\ref{sec_contribution_pm}).
 
@@
 the other hand, the interviewees believed that testing (including usability
-tests with users) was the top solution to improve \textit{correctness},
-\textit{usability}, and \textit{reproducibility}. One pain point in the
-development process is the lack of access to real-world datasets for testing.
-The developers' strategies to address it is in Section \ref{sec_pain_points_3}.
-One threat to \textit{correctness} is: with huge datasets for testing, the tests
-are expensive and time-consuming. Three interviewees endorsed self tests /
-automated tests, which may save time for testing.
-
-All 29 projects did not have theory manuals. We identified a road map in the
-\textit{3D Slicer} project, and no requirements specifications for the rest.
-Eight of the nine interviewees thought that documentation was essential to their
-projects. However, they hold the common opinion that their documentation needed
+tests with users) was the top solution to improve correctness, usability, and
+reproducibility. One pain point in the development process is the lack of access
+to real-world datasets for testing. The developers' strategies to address this
+are summarized in Section~\ref{sec_pain_points_3}. One threat to correctness is:
+with huge datasets for testing, the tests are expensive and time-consuming.
+Three interviewees endorsed self tests / automated tests, which may save time
+for testing.
+
+Documentation is apparently not emphasized in any of the 29 projects.  None of
+them had theory manuals, although we did identify a road map in the \textit{3D
+Slicer} project.  No requirements specifications were found. Eight of the nine
+interviewees thought that documentation was essential to their projects.
+However, they hold the common opinion that their documentation needed
 improvements. Nearly half of them also believed that the lack of time prevented
@@
 
-\subsection{Pain Points and Solutions}
-\label{sec_rq_pain_points}
 \begin{description}
@@
 question, including the pain points and the solutions proposed by the
-developers, are in Section \ref{sec_interview_pain_points}.
-
-\subsection{Software Qualities}
-\label{sec_rq_qualities}
-\begin{description}\item[RQ5.] What is the current status of the following
+developers, are in Section~\ref{sec_interview_pain_points}.
+
+\begin{description}
+\item[RQ5.] What is the current status of the following
 software qualities for the projects? What actions have the developers taken to
-address them?\end{description}
-
-This section includes our answers from the qualities measurements and interviews
-with the developers. Table \ref{tab_reference_rq4} shows the sections and tables
-containing answers to this research question.
-
-\begin{table}[ht]
-\centering
-\begin{tabular}{ll}
-\hline
-Section or table & Description \\ \hline
-Section \ref{sec_result_installability} & Installability \\
-Section \ref{sec_result_correctness_verifiability} and \ref{sec_interview_correctness} & Correctness \& verifiability \\
-Section \ref{sec_result_reliability} & Reliability \\
-Section \ref{sec_result_robustness} & Robustness \\
-Section \ref{sec_result_usability} and \ref{sec_interview_usability}& Usability \\
-Section \ref{sec_score_maintainability} and \ref{sec_interview_maintainability} & Maintainability \\
-Section \ref{sec_result_reusability} & Reusability \\
-Section \ref{sec_result_understandability} and \ref{sec_interview_understandability} & Understandability \\
-Section \ref{sec_result_visibility_transparency} & Visibility/transparency \\
-Section \ref{sec_interview_reproducibility} & Reproducibility \\ \hline
-\end{tabular}
-\caption{\label{tab_reference_rq4}Sections and tables with answers to RQ4}
-\end{table}
-
-\subsection{Our Ranking versus the Community Ratings}
-\label{sec_rq_comparison}
+address them?
+\end{description}
+
+The current status of the software qualities was assessed through filling in the
+measurement template \citep{Dong2021-Data}.  A summary of the comparison between
+all projects can be found in Section~\ref{ch_results}.  We explored some of the
+qualities more deeply, especially with respect to techniques addressing quality
+concerns, through the interview questions.  Specifically, the qualities of
+correctness \& verifiability, usability, maintainability, understandability and
+reproducibility are explored in Section~\ref{ch_interview}.
+
 \begin{description}
@@
 \item comparing our ranking with the community ratings on GitHub, such as GitHub
-stars, number of forks, and number of people watching the projects (Section
-\ref{sec_ranking_vs_github});
+stars, number of forks, and number of people watching the projects (Section~\ref{sec_grading_software});
 \item comparing top-rated software designated by our methodology with the ones
-recommended by our domain experts (Section \ref{sec_designation_vs_experts}).
-\end{itemize}
-
-\subsubsection{Our Ranking versus the GitHub Popularity} \label{sec_ranking_vs_github}
-
-Table \ref{tab_ranking_vs_GitHub} shows our ranking to the 29 MI projects, and
-their GitHub metrics if applicable. As mentioned in Section
-\ref{sec_score_maintainability}, 24 projects used GitHub. Since GitHub
+recommended by our domain experts (Section \ref{sec_software_selection}).
+\end{itemize}
+
+Table~\ref{tab_ranking_vs_GitHub} shows our ranking of the 29 MI projects, and
+their GitHub metrics, if applicable. As mentioned in
+Section~\ref{sec_score_maintainability}, 24 projects used GitHub. Since GitHub
 repositories have different creation dates, we collect the number of months each
@@
 watching, and forks per 12 months. The method of getting the creation date is
-described in Section \ref{sec_empirical_measurements}, and we obtained these
-metrics in July, 2021.
-
-In Table \ref{tab_ranking_vs_GitHub}, we used the average number of new stars
+described in Section~\ref{sec_grading_software}.  We obtained these
+measures in July, 2021.
+
+In Table~\ref{tab_ranking_vs_GitHub}, we used the average number of new stars
 per 12 months as an indicator of the GitHub popularity, and listed the items in
@@
 Project \textit{dwv} was popular on GitHub, but we ranked it low. As mentioned
-in Section \ref{sec_result_installability}, we failed to build it locally, and
+in Section~\ref{sec_result_installability}, we failed to build it locally, and
 used the test version on its websites for the measurements. We followed the
-instructions and tried to run the command ``yarn run test" locally, which did
+instructions and tried to run the command ``yarn run test'' locally, which did
 not work. In addition, the test version did not detect a broken DICOM file and
@@
 \renewcommand{\arraystretch}{0.85}
-\begin{table}[ht]
+\begin{table}[!ht]
 \centering
@@
 
-\subsubsection{Designated Top Software versus the Domain Experts' Recommendation}
-\label{sec_designation_vs_experts}
-
-As shown in Section \ref{sec_mi_software_selection}, our domain experts
-recommended a list of top software with 12 software products. Table
-\ref{tab_top_software_vs_experts} and \ref{tab_experts_vs_top_software} compare
-the top 12 software projects ranked by our methodology with the ones from the
-domain experts.
-
-All of the top 4 software from the domain experts are among the top 12 ones
-ranked by our methodology. 3 of the top 4 on both lists are the same ones:
-\textit{3D Slicer}, \textit{ImageJ}, and \textit{Fiji}. \textit{3D Slicer} is
-also the top 1 of both rankings.
-
-As mentioned in Section \ref{sec_mi_software_selection}, six of the recommended
-software packages did not have visualization as the primary function, so we did
-not include them on our final list.
-
-\begin{table}[ht]
-\centering
-\begin{tabular}{lll}
-\hline
-Our ranking & Assessed software & Domain experts \\ \hline
-1 & 3D Slicer & 1 \\
-2 & ImageJ & 3 \\
-3 & OHIF Viewer & n/a \\
-4 & Fiji & 4 \\
-5 & ParaView & n/a \\
-6 & Weasis & n/a \\
-7 & medInria & n/a \\
-8 & BioImage Suite Web & n/a \\
-9 & OsiriX Lite & n/a \\
-10 & INVESALIUS 3 & n/a \\
-11 & Gwyddion & n/a \\
-12 & Horos & 2 \\ \hline
-\end{tabular}
-\caption{\label{tab_top_software_vs_experts}Top software by our ranking versus domain experts' recommendation}
-\end{table}
-
-\begin{table}[ht]
-\centering
-\begin{tabular}{lll}
-\hline
-Our ranking & Recommended software & Domain experts \\ \hline
-1 & 3D Slicer & 1 \\
-12 & Horos & 2 \\
-2 & ImageJ & 3 \\
-4 & Fiji & 4 \\
-n/a & AFNI & 5 \\
-n/a & FSL & 6 \\
-n/a & Freesurfer & 7 \\
-18* & Mricron & 8 \\
-17** & Mango & 9 \\
-n/a & Tarquin & 10 \\
-n/a & Diffusion Toolkit & 11 \\
-n/a & MRItrix & 12 \\ \hline
-\end{tabular}
-\caption{\label{tab_experts_vs_top_software}Domain experts' recommendation versus our ranking}
-\end{table}
-
-* \textit{MRIcron} development had moved to \textit{MRIcroGL}, as mentioned in
-Section \ref{sec_mi_software_selection}. Thus, we measured and ranked
-\textit{MRIcroGL} at 18.
-
-** We included \textit{Mango} in the initial list, but removed it because it was
-not OSS. \textit{Papaya} is a the web version OSS of \textit{Mango}. We measured
-and ranked \textit{Papaya} at 17.
-
-\subsection{Threats to Validity} \label{sec_threats_to_validity}
-
-This section lists all the potential threats to validity in our research. The
-definitions of common validity are
-\citep{AmpatzoglouEtAl2019}\citep{ZhouEtAl2016}:
+As shown in Section~\ref{sec_software_selection}, our domain experts recommended
+a list of top software with 12 software products.  All of the top 4 software
+from the Domain Expert's list are among the top 12 ranked by our methodology. 3
+of the top 4 on both lists are the same: \textit{3D Slicer}, \textit{ImageJ},
+and \textit{Fiji}. \textit{3D Slicer} is top project by both rankings (and by
+the GithHub stars measure as well).  The Domain Expert ranked \textit{Horos} as
+their second choice, while we ranked it twelth.  Our third ranked project,
+\textit{OHIF Viewer} was not listed by the Domain Expert.  Neither were the
+software packages that we ranked from fifth to eleventh (\textit{ParaView},
+\textit{Weasis}, \textit{medInria}, \textit{BioImage Suite Web}, \textit{OsiriX
+Lite}, \textit{INVESALIUS}, and \textit{Gwyddion}).  The software mentioned by
+the Domain Expert that we did not rank were the six recommended packages that
+did not have visualization as the primary function (as discussed in
+Section~\ref{sec_software_selection}).  The differences between the list
+recommended by our methodology and the Domain Expert are not surprising.  As
+mentioned above, the methodology weights all qualities equally, but that may not
+be the case for the Domain Expert's impressions.  Moreover, although the Domain
+Expert has significant experience with MI software, he has not used every one of
+the 29 packages.
+
+\wss{In the shorter journal paper version, the focus could be just on answering the 
+research questions and deleting the previous two sections.}
+
+\section{Threats to Validity} \label{sec_threats_to_validity}
+
+Below we categorize and list the threats to validity that we have identified.
+The categories come from \citet{AmpatzoglouEtAl2019} and \citet{ZhouEtAl2016}.
+
 \begin{description}
@@
     up to its claims. This aspect deals with whether or not the researcher measures
-what is intended to be measured" \citep{AmpatzoglouEtAl2019}.
-\item[Internal Validity:] ``This aspect relates to the examination of causal
-relations. Internal validity examines whether an experimental
-treatment/condition makes a difference or not, and whether there is evidence to
-support the claim" \citep{AmpatzoglouEtAl2019}.
-\item[External Validity:] ``Define the domain to which a study's findings can be
-generalized" \citep{ZhouEtAl2016}.
-\item[Conclusion Validity:] ``Demonstrate that the operations of a study such as
-the data collection procedure can be repeated, with the same results"
-\citep{ZhouEtAl2016}.
+    what is intended to be measured'' \citep{AmpatzoglouEtAl2019}.
 \end{description}
 
-\noindent We categorize and present the threats to validity in the following subsections.
-
-\subsubsection{Threats to Construct Validity}
-\begin{itemize}
-\item We compared nine software qualities for 29 software packages, so we could
-only spend a limited time on each of them. As a result, our assessments may have
-missed something relevant.
+\begin{itemize}
+\item For practical considerations the time spent measuring each package had to
+be limited.  The time limit may have caused an assessment to have missed
+something something relevant.
 \item Our ranking is partly based on surface (shallow) measurement, which may
 not fully reveal the underlying qualities.
-\end{itemize}
-
-\subsubsection{Threats to Internal Validity}
-\begin{itemize}
-\item It was not practical to ask each development team for every piece of
-information. We collected much information - such as artifacts and funding
-situations of software - by ourselves. There may be cases that we missed some
-information.  There also may be cases where we did not find evidence of
+\item The questions in the measurement template may not actually measure the
+qualities they are associated with.  For instance, we have assumed that
+maintainability is improved if a high percentage of identified issues are
+closed, but it is possible for a project with a wealth of ideas to have many
+open issues, and still be maintainable.
+\item With the exception of the interview data for 8 projects, we collected all
+of the information for each project from the artifacts available on the
+Internet. In some cases this source of data may mean we did not find evidence of
 something, like unit testing, not because the project didn't do it, but because
 no artifacts of this activity remained in the publicly available repository.
-\item As mentioned in Section \ref{ch_interview}, one interviewee was too busy
-to participate in a full interview, so he provided a version of written answers
-to us. Since we did not have the chance to explain our questions or ask him
+\item As mentioned in Section~\ref{ch_interview}, one interviewee was too busy
+to participate in a full interview, so they provided a version of written answers
+to us. Since we did not have the chance to explain our questions or ask them
 follow-up questions, there is a possibility of misinterpretation of the
 questions or answers.
-\item As mentioned in Section \ref{sec_result_installability}, we could not
+\item As mentioned in Section~\ref{sec_result_installability}, we could not
 install or build \textit{dwv}, \textit{GATE}, and \textit{DICOM Viewer}. We used
@@
 no alternative for \textit{DICOM Viewer}. We might underestimate their rank due
-to uncommon technical issues.
-\end{itemize}
-
-\subsubsection{Threats to External Validity}
+to an uncommon technical issue.
+\end{itemize}
+
+\begin{description}
+    \item[Internal Validity:] ``This aspect relates to the examination of causal
+    relations. Internal validity examines whether an experimental
+    treatment/condition makes a difference or not, and whether there is evidence
+    to support the claim'' \citep{AmpatzoglouEtAl2019}.
+\end{description}
+
+\begin{itemize}
+    \item Many of the measurement template questions look for the presence of
+    certain artifacts, like a user manual, or a getting started tutorial.  We
+    have implicitly assumed that the presence of these artifacts is an
+    indication of a given quality, but this is an indirect measure.  It may be
+    possible to achieve qualities without the artifacts we sought.  A direct
+    measure of quality, like an usability experiment, would not have this
+    problem.
+    \item We compared our rankings to the rankings by the community and the
+    Domain Expert, but we have assumed all qualities are equally weighted.  The
+    community and the Domain Expert likely have a more complex weighting between
+    qualities.
+    \item We assumed there was a casual relationship between the number of files
+    and reusability, since we assumed multiple files implies modularity.  Of
+    course the code can be modular, and not divided into separate files.
+    Moreover, having many files does not immediately in itself imply a
+    decomposition based on sound design principles, like information hiding.
+\end{itemize}
+
+\begin{description}
+    \item[External Validity:] ``Define the domain to which a study's findings
+    can be generalized'' \citep{ZhouEtAl2016}.
+\end{description}
+
 \begin{itemize}
 \item We interviewed eight teams, which is a good proportion of the 29. However,
-there is still a risk that they might not well represent the whole MI software
-community.
-\item Our ranking gave all qualities equal weight, which may not be the case
-with all users. Thus, it may not represent the popularity of software among
-users.
-\item The number of GitHub stars, watches, and forks are not perfect measures of
-popularity, but they are what we had available.
-\end{itemize}
-
-\subsubsection{Threats to Conclusion Validity}
-\begin{itemize}
-\item We used the grading template in Appendix \ref{ap_grading_template} to
-guide our measurements. Our impressions of the software - such as user
-experience - were factors in deciding some scores. Thus, there is a risk that
-some scores may be subjective and biased.
+there is still a risk that the subset of the entire group does not represent the
+whole MI software community.
+\item We identified qualities that we believe the community will be interested
+in, and we weighted these qualities equally, but the qualities we chose, and
+their weighting, may not match the external reality.
+\item The number of GitHub stars, watches, and forks are not ideal measures of
+popularity.  \wss{Can we find a reference for this?}
+\end{itemize}
+
+\begin{description}
+    \item[Conclusion Validity:] ``Demonstrate that the operations of a study
+    such as the data collection procedure can be repeated, with the same
+    results'' \citep{ZhouEtAl2016}.
+\end{description}
+
+\begin{itemize}
+\item The grading template included an entry for the reviewer's impression.  We
+aimed for objectivity, but there is a risk that some scores may be subjective
+and biased.
+\item The measurements are made at an instant in time, and the instant differs
+by a few weeks between projects, due to the time needed to measure them. For the
+most part, the projects are living and continually changing, which means if we
+measured at a different time, our results may change.
 \end{itemize}
@@
 
-This section presents our recommendations on SC software development. In
-general, our suggestions apply to all SC domains, unless we specifically mention
-that a particular guideline is only for MI software.
-
-Section \ref{sec_recommendations_qualities} discusses the actions that can
-potentially improve the ten software qualities. Sections
-\ref{sec_recommendations_limited_resources},
+This section presents our recommendations on MI software development. Although
+our focus is on MI software, unless noted otherwise, our recommendations apply
+to any scientific computing software.
+Section~\ref{sec_recommendations_qualities} discusses the actions that can
+potentially improve the ten software qualities.
+Sections~\ref{sec_recommendations_limited_resources},
 \ref{sec_recommendations_tech_stack}, and
 \ref{sec_recommendations_testing_dataset} are based on the primary pain points
-collected from the developers in the MI domain, but we believe scientists and
-developers are likely to face them in most SC domains. These sections contain
-our general suggestions tackling them.
-
-\subsection{Recommendations on Improving Software Qualities} \label{sec_recommendations_qualities}
-
-Based on our quality measurements in Sections \ref{ch_results} and discussions
-with the developers in Sections \ref{sec_interview_software_qualities}, we
-collected many key points that may improve the software qualities. We list the
-primary ones by each quality as follows,
-\begin{itemize}
-\item \textbf{Installability} (Section \ref{sec_result_installability})
-\begin{itemize}
-    \item clear instructions;
-    \item automated installer;
+(from Section~\ref{sec_interview_pain_points}) collected from developers in the
+MI domain.
+
+\subsection{Recommendations on Improving Software Qualities}
+\label{sec_recommendations_qualities}
+
+Based on our quality measurements in Section~\ref{ch_results} and discussions
+with the developers in Section~\ref{ch_interview}, we collected key points for
+improving software qualities. These points should be considered for new and
+existing projects, but we are not saying that every project should do everything
+on our list.  All projects will have finite resources, so strategies will have
+to be selected that provide the greatest return on investment.  Our specific
+recommendations for each quality are as follows:
+
+\begin{itemize}
+\item \textbf{Installability} (Section~\ref{sec_result_installability})
+\begin{itemize}
+    \item providing clear instructions;
+    \item automating installation;
     \item including all dependencies in the installer;
-    \item avoiding heavily depending on other commercial products (e.g. Matlab);
-    \item considering building a web application that needs no installation.
-\end{itemize}
-\item \textbf{Correctness \& Verifiability} (Section \ref{sec_result_correctness_verifiability} and \ref{sec_interview_correctness})
-\begin{itemize}
-    \item test-driven development with unit tests, integration tests, and nightly tests;
+    \item avoiding heavily depending on other commercial products (e.g.\ Matlab);
+    \item potentially building a web application that needs no installation.
+\end{itemize}
+\item \textbf{Correctness \& Verifiability}
+(Section~\ref{sec_result_correctness_verifiability} and
+Section~\ref{sec_interview_software_qualities})
+\begin{itemize}
+    \item test-driven development with unit tests, integration tests, and
+    nightly tests;
     \item two stage development process with stable release \& nightly builds;
     \item CI/CD;
-    \item requirements specifications and theory manuals \citep{Smith2016} \citep{SmithAndLai2005}.
+    \item requirements specifications and theory manuals \citep{Smith2016, SmithAndLai2005}.
     \item static code analysis tools (e.g. Lint and SonarQube)
 \end{itemize}
-\item \textbf{Reliability} (Section \ref{sec_result_reliability})
+\item \textbf{Reliability} (Section~\ref{sec_result_reliability})
 \begin{itemize}
@@
 \end{itemize}
-\item \textbf{Robustness} (Section \ref{sec_result_robustness})
-\begin{itemize}
-    \item designing with exceptions and make the software failures graceful;
+\item \textbf{Robustness} (Section~\ref{sec_result_robustness})
+\begin{itemize}
+    \item designing with exception handling so the software can fail gracefully;
     \item descriptive error messages.
 \end{itemize}
-\item \textbf{Usability} (Section \ref{sec_result_usability} and \ref{sec_interview_usability})
+\item \textbf{Usability} (Section~\ref{sec_result_usability} and Section~\ref{sec_interview_software_qualities})
 \begin{itemize}
@@
     \item user manuals;
-    \item professional UX designs;
+    \item professional UX (User eXperience) designs;
+    \item graphical user interface;
     \item active supports to users.
 \end{itemize}
-\item \textbf{Maintainability} (Section \ref{sec_score_maintainability} and
-\ref{sec_interview_maintainability})
-\begin{itemize}
-    \item using GitHub;
-    \item modular approach with the design principle proposed by Parnas:
-    ``system details that are likely to change independently should be the
-    secrets of separate modules; the only assumptions that should appear in the
-    interfaces between modules are those that are considered unlikely to
-    change." \citep{ParnasEtAl2000}
+\item \textbf{Maintainability} (Section~\ref{sec_score_maintainability} and Section~\ref{sec_interview_software_qualities})
+\begin{itemize}
+    \item use GitHub (or equivalent version control and project management environment);
+    \item use a modular approach to design (we advocate basing the design on
+    Parnas's principle of information hiding: ``system details that are likely
+    to change independently should be the secrets of separate modules; the only
+    assumptions that should appear in the interfaces between modules are those
+    that are considered unlikely to change.'' \citep{ParnasEtAl2000})
     \item documentation for developers: project plan, developer’s manual, and API documentation.
 \end{itemize}
-\item \textbf{Reusability} (Section \ref{sec_result_reusability})
+\item \textbf{Reusability} (Section~\ref{sec_result_reusability})
 \begin{itemize}
@@
     \item API documentation;
-    \item tools that generate software documentation for developers (e.g.
+    \item tools that generate software documentation for developers (e.g.\
     Doxygen, Javadoc, and Sphinx).
 \end{itemize}
-\item \textbf{Understandability} (Section \ref{sec_result_understandability} and
-\ref{sec_interview_understandability})
+\item \textbf{Understandability} (Section~\ref{sec_result_understandability} and
+Section~\ref{sec_interview_software_qualities})
 \begin{itemize}
@@
     \item clear comments, indicating what is being done, not how;
-    \item description of used algorithms;
-    \item documentation of explicit requirements on coding standard;
-    \item communication between developers and users via GitHub issues, mailing lists, and forums.
-    \item graphical user interface.
-\end{itemize}
-\item \textbf{Visibility/Transparency} (Section \ref{sec_result_visibility_transparency})
-\begin{itemize}
-    \item documents for the development process, project status, development
-    environment, and release notes.
-\end{itemize}
-\item \textbf{Reproducibility} (Section \ref{sec_interview_reproducibility})
+    \item description of algorithms used;
+    \item documentation of explicit requirements on coding standards;
+    \item communication between developers and users via GitHub issues, mailing
+    lists, and forums.
+\end{itemize}
+\item \textbf{Visibility/Transparency} (Section~\ref{sec_result_visibility_transparency})
+\begin{itemize}
+    \item documents for the development process, contributor's guide, project
+    status, development environment, and release notes.
+\end{itemize}
+\item \textbf{Reproducibility} (Section~\ref{sec_interview_software_qualities})
 \begin{itemize}
@@
 
+\wss{Could ``reverse'' the list and list the recommendations and beside each the qualities 
+that that recommendation helps with - this would reduce repetition.}
+
 \subsection{Recommendations on Dealing With Limited Resources}
@@
 
-The limitation of resources has many faces. We regard the lack of fundings,
-time, and developers as representations of this problem.
-
-We summarize our discussion with the MI software developers in Section
-\ref{sec_pain_points_1} with the following recommendations,
+The limitation of resources has many faces, with the key manifestations being
+lack of fundings, time, and developers. We summarized our discussion with the MI
+software developers in Section~\ref{sec_interview_pain_points}.  Many of our
+recommendations involve investing more effort into processes, tools and methods,
+so that the limited resources can be used more effectively.  Our recommendations
+are as follows:
+
 \begin{itemize}
 \item \textbf{Identify the root cause.} More fundings or developers may not
-solve the problem of lacking time. It is beneficial to identify the underlying
-obstacles to the team.
-
-\item \textbf{Maintain a good documentation.} Creating and updating
-documentation consumes time, but can save much more time in the long term. If
-the users and developers can find answers to their questions themselves, they
-are less likely to abuse the team's issue tracker.
-
-\item \textbf{Adopt time-saving tools.} A good CI/CD tool (e.g., GitHub Actions)
+solve problems called by a lack of time. It is beneficial to identify the
+underlying obstacles to the team.  \wss{Is there something from the mythical man
+month that should be cited here?}
+
+\item \textbf{Maintain good documentation.} Creating and updating documentation
+consumes time, but can save much more time in the long term. If the users and
+developers can find answers to their questions themselves, they are less likely
+to abuse the team's issue tracker.  \wss{If it is an option, we could cite one
+of our productivity related papers.}
+
+\item \textbf{Adopt time-saving tools.} A good CI/CD tool (e.g.\ GitHub Actions)
 saves time for building and deploying the product, and automated tests can work
@@
 \item \textbf{Use test-driven development process.} Many people think writing
-test cases is less fun than building the functional code, but this is only true
-before we encounter the bugs. Identifying and fixing bugs can consume
-substantial resources. Setting up the test cases costs time, but generates more
-benefits in the long run.
+test cases is less rewarding than writing code, but without testing identifying
+and fixing bugs can consume substantial resources. Setting up the test cases
+costs time, but generates more benefits in the long run.
 
@@
 So it may be a good idea to shift some development and maintenance
-responsibilities to the users. The users may also be happy about the extra
+responsibilities to the users. The users may also be happy with the extra
 flexibility.
 
-\item \textbf{Consider ``hibernating" for a while.} When developers are not
-enough, the team can shift from development mode toward maintenance mode for
-some time. Stop building new features, and instead fix bugs and design problems
-from the past. If the development team can repay some of its technical debt, the
-software qualities may improve as a result.
+\item \textbf{Consider ``hibernating'' for a while.} When there aren't enough
+developers, the team can shift from development mode toward maintenance mode.
+The team can stop building new features for a while and instead fix bugs and
+design problems from the past. If the development team can repay some of its
+technical debt \citep{KruchtenEtAl2012}, the software qualities will likely
+improve as a result.
 
@@
 commercial companies to use as internal modules of their products may bring
-financial supports to the team. Meanwhile, the project can stay open-source for
+financial support to the team. Meanwhile, the project can stay open-source for
 the community.
@@
 
-\subsection{Recommendations on Choosing A Tech Stack} \label{sec_recommendations_tech_stack}
+\subsection{Recommendations on Choosing A Tech Stack}
+\label{sec_recommendations_tech_stack}
 
 A tech stack refers to a set of technologies used by a team to build software
-and manage the project. Section \ref{sec_pain_points_2} lists the advantages and
-disadvantages between native and web applications. In this section, we give
-further suggestions on the choice of a tech stack to address the
-\textit{compatibility}, \textit{maintainability}, \textit{performance}, and
-\textit{security} of software.
-
-\begin{itemize}
-\item \textbf{Identify the priorities of the qualities.} It is hard to cover all
-aspects. Some teams achieve all four above qualities for their software, but it
-is not an easy task. Sections \ref{sec_pain_points_2} contains more details
-about the difficulty of balancing between the four qualities. A team needs to
-prioritize its objectives according to its resource and experience.
+and manage the project. Section~\ref{sec_interview_pain_points} lists the
+advantages and disadvantages between native and web applications. In this
+section, we give further suggestions on the choice of a tech stack to improve
+the four priority factors identified by developers: compatibility,
+maintainability, performance and security.  The suggestions are intended to
+provide ideas and avenues for exploration; not all of the suggestions will be
+the right fit for all projects and all teams.
+
+\begin{itemize}
+\item \textbf{Identify the priorities between the factors.} Simultaneously
+achieving high levels for all four factors (compatibility, maintainability,
+performance and security) is difficult. A team needs to prioritize its
+objectives according to its resource and experience.
 
 \item \textbf{Be open-minded about new technologies.} Web applications with only
-a frontend are known for worse \textit{performance} than native applications.
-However, new technologies may ease this difference. For example, some JavaScript
-libraries can help the frontend harness the power of computer GPU and accelerate
-graphical computing. In addition, there are new frameworks helping developers
-with cross-platform \textit{compatibility}. For example, the
-\hyperlink{https://flutter.dev/}{Flutter} project enables support for web,
-mobile, and desktop OS with one codebase.
-
-\item \textbf{Use git and GitHub.} As mentioned in Sections
-\ref{sec_score_maintainability}, almost all of the 29 MI software projects used
-git, and the majority of them used GitHub. We found from the projects' websites
-and our interviews with developers that, some projects moved from other version
-control tools to git and GitHub. GitHub provides convenient repository and
-project management, and OSS projects receive more attention and contribution on
-GitHub.
+a frontend are known for worse performance than native applications. However,
+new technologies may ease this difference. For example, some JavaScript
+libraries can help the frontend harness the power of the computer's GPU and
+accelerate graphical computing. In addition, there are new frameworks helping
+developers with cross-platform compatibility. For example, the
+\href{https://flutter.dev/}{Flutter} project enables support for web, mobile,
+and desktop OS with one codebase.
+
+\item \textbf{Use git and GitHub.} As mentioned in
+Section~\ref{sec_score_maintainability}, almost all of the 29 MI software
+projects used git, and the majority of them used GitHub. We found from the
+projects' websites and our interviews with developers that, some projects moved
+from other version control tools to git and GitHub. GitHub provides convenient
+repository and project management, and OSS projects tend to receive more
+attention and contributions on GitHub.
 
@@
 
-\item \textbf{Backend servers can have low costs.} It is worth exploring the
-serverless solutions from major cloud service providers. Serverless still uses a
+\item \textbf{Backend servers can have low costs.} Serverless solutions from
+major cloud service providers may be worth exploring. Serverless still uses a
 server, but the team is only charged when they use it. The solution is
 event-driven, and costs the team by the number of requests it processes. Thus,
-serverless can be very cost-effective for the less intensively used functions.
+serverless can be very cost-effective for less intensively used functions.
 
@@
 data online can be a problem for projects requiring high security. Regulations
-in some SC domains may forbid doing so. In this case, a web application with a
-backend may not be a good choice.
-
-\item \textbf{Maintain a good documentation.} No matter what tech stack a team
+for some MI applications may forbid doing web transmissions. In this case, a web
+application with a backend may not be an option.
+
+\item \textbf{Maintain good documentation.} No matter what tech stack a team
 uses, a well-maintained project plan, developer's manual, and API documentation
-always help team members to contribute more and make fewer mistakes.
-\end{itemize}
-
-\subsection{Recommendations on Enriching the Testing Datasets} \label{sec_recommendations_testing_dataset}
-
-As described in Section \ref{sec_interview_pain_points}, it was difficult for
-some software development teams in the MI domain to access real-world medical
-imaging datasets. This problem restricted their capability and flexibility to
-test their software. We believe software developers in other SC domains may also
-face similar issues.
-
-Based on Section \ref{sec_pain_points_3}, we provide some suggestions as follows,
+generally help team members to contribute more and make fewer mistakes.
+\end{itemize}
+
+\subsection{Recommendations on Enriching the Testing Datasets} 
+\label{sec_recommendations_testing_dataset}
+
+As described in Section~\ref{sec_interview_pain_points}, it is difficult for
+software development teams to access real-world medical imaging datasets. This
+problem restricts their capability and flexibility for testing. We provide some
+suggestions as follows:
+
 \begin{itemize}
 \item \textbf{Build and maintain good connections to datasets.} A team can build
-connections with professionals working in the SC domain, who may have access to
-private datasets and perform tests for the team. Moreover, if a team has such
-professionals as internal members, the process can be even simpler.
-
-\item \textbf{Collect and maintain datasets over time.} A team may face all
-kinds of strange problems caused by various unique inputs over the years of
-development. It is worth collecting and maintaining this data, which can form a
-good dataset for testing.
-
-\item \textbf{Search for open data sources.} In general, there are many open
-datasets in different SC domains. Take MI as an example, there are Chest X-ray
-Datasets by National Institute of Health
-(\hyperlink{https://nihcc.app.box.com/v/ChestXray-NIHCC}{https://nihcc.app.box.com/v/ChestXray-NIHCC})
-\citep{WangEtAl2017}, Cancer Imaging Archive
-(\hyperlink{https://www.cancerimagingarchive.net/}{https://www.cancerimagingarchive.net/})
-\citep{PriorEtAl2017}, and MedPix by National Library of Medicine
-(\hyperlink{https://medpix.nlm.nih.gov/home}{https://medpix.nlm.nih.gov/home})
-\citep{Smirniotopoulos2014}. A team developing MI software should be able to find
-more open datasets according to their needs.
+connections with professionals working in the medical domain, who may have
+access to private datasets and can perform tests for the team. If a team has
+such professionals as internal members, the process can be simplified.
+
+\item \textbf{Collect and maintain datasets over time.} A team may face problems
+caused by various unique inputs over the years of software development. This
+data should be collected and maintained over time to form a good, comprehensive,
+dataset for testing.
+
+\item \textbf{Search for open data sources.} In general, there are many open MI
+datasets.  For instance, there are
+\href{https://nihcc.app.box.com/v/ChestXray-NIHCC}{Chest X-ray Datasets} by
+National Institute of Health \citep{WangEtAl2017},
+\href{https://www.cancerimagingarchive.net/}{Cancer Imaging Archive}
+\citep{PriorEtAl2017}, and \href{https://medpix.nlm.nih.gov/home}{MedPix} by
+National Library of Medicine \citep{Smirniotopoulos2014}. A team developing MI
+software should be able to find more open datasets according to their needs.
 
@@
 problem better. Some datasets are too sensitive to share, but if the community
-has some kind of ``group discussion", teams can better express their needs, and
+has some kind of ``group discussion'', teams can better express their needs, and
 professionals can better offer voluntary support for testing. Ultimately, the
@@
 maintains large datasets, tests OSS in the domain, and protects privacy.
-% look into the database that the Eng Phys guest speaker mentioned?
+
 \end{itemize}
@@
 
-We analyzed the state of the practice for SC software in the MI domain. To
-better achieve our goals in Section \ref{ch_intro}, we proposed six research
-questions in Section \ref{sec_research_questions}.
-
-Our methods in Section \ref{ch_methods} form a general process to evaluate
-domain-specific software, that we apply on specific SC domains. As mentioned in
-Section \ref{sec_applying_method}, following this process, we chose the MI
-domain, identified 48 SC software candidates in it, then selected 29 of them to
-our final list. Section \ref{ch_results} lists our measurements to nine software
-qualities for the 29 projects, and Section \ref{ch_interview} contains our
-interviews with eight of the 29 teams, discussing their development process and
-five software qualities.
-
-We answered our six research questions in Section \ref{ch_answers}. In addition,
-Section \ref{ch_recommendations} presents our recommendations on SC software
-development.
+We analyzed the state of the practice for the MI domain with the goal of
+understanding current practice, answering our six research questions
+(Section~\ref{sec_motivation}) and providing recommendations for current
+and future projects.  Our methods in Section~\ref{ch_methods} form a general
+process to evaluate domain-specific software, that we apply to the specific
+domain of MI software. We identified 48 MI software candidates, then, with the
+help of the Domain Expert selected 29 of them to our final list.
+Section~\ref{ch_results} lists our measurements to nine software qualities for
+the 29 projects, and Section~\ref{ch_interview} contains our interviews with
+eight of the 29 teams, discussing their development process and five software
+qualities.  We answered our six research questions in Section~\ref{ch_answers}.
+In addition, Section~\ref{ch_recommendations} presents our recommendations on SC
+software development.
 
@@
 
-With the measurement results in Section \ref{ch_results}, we revealed some
-current status of SC software development and qualities in the MI domain. We
-ranked the 29 software projects in nine qualities based on the grading scores.
-\textit{3D Slicer}, \textit{ImageJ}, and \textit{OHIF Viewer} are the top three
-software by their overall scores.
-
-The interview results in Section \ref{ch_interview} show some merits, drawbacks,
+With the measurement results in Section~\ref{ch_results}, we summarized the
+current status of MI software development. We ranked the 29 software projects in
+nine qualities.  Based on on the grading scores \textit{3D Slicer},
+\textit{ImageJ}, and \textit{OHIF Viewer} are the top three software packages.
+
+The interview results in Section~\ref{ch_interview} show some merits, drawbacks,
 and pain points within the development process. The three primary categories of
@@
 \end{itemize}
-We summarized the solutions from the developers to address these problems. We
-also collected the status of software testing, documentation, contribution
-management, and project management in the eight projects.
-
-Our answers to the research questions (Section \ref{ch_answers}) are based on
+We summarized the solutions from the developers to address these problems,
+including developing a web-based approach with backend servers and maintaining
+better documentation. We also collected the status of documentation.  We found
+that for all 8 interviewed teams that documentation is felt to be vital to a
+project, with the most popular form of documentation being forum discussions and
+videos.  With respect to project management almost all teams used GitHub and
+pull requests to manage contributions.  Very few teams used a specific
+development model.  It appears that the development process is more ad hoc than
+planned for the majority of projects.
+
+Our answers to the research questions (Section~\ref{ch_answers}) are based on
 the above findings. We identified the existing artifacts, tools, principles,
-processes, and methodologies in the 29 projects. By comparisons in Section
-\ref{sec_rq_comparison}, we found out: 1) four of the top five software projects
-in our ranking were also among the top five ones receiving the most GitHub stars
-per year (Table \ref{tab_ranking_vs_GitHub}); 2) three of the top four in our
-ranking were among the top four provided by the domain experts (Table
-\ref{tab_top_software_vs_experts}).
-
-Section \ref{ch_recommendations} presents our recommendations on improving
+processes, and methodologies in the 29 projects. By comparisons with the implied
+popularity of existing projects we found: 1) four of the top five software
+projects in our ranking were also among the top five ones receiving the most
+GitHub stars per year (Table~\ref{tab_ranking_vs_GitHub}); 2) three of the top
+four in our ranking were among the top four provided by the domain experts.
+
+Section~\ref{ch_recommendations} presents our recommendations on improving
 software qualities and easing pain points during development. Some highlighted
-ones are:
+recommendations are as follows:
 \begin{itemize}
@@
 \item using git and GitHub;
-\item modular approach with the design principle proposed by Parnas
-\citep{ParnasEtAl2000};
-\item considering newer technologies (e.g., web application and serverless
+\item modular approach with the design principle proposed by \citet{ParnasEtAl2000};
+\item considering newer technologies (e.g.\ web application and serverless
 solution);
-\item various ways of enriching the testing datasets in Section
-\ref{sec_recommendations_testing_dataset}.
+\item various ways of enriching the testing datasets, such as using existing
+open data sources and establishing greater community collaboration in the MI
+domain (Section~\ref{sec_recommendations_testing_dataset}).
 \end{itemize}
@@
     \begin{itemize}
-        \item \textit{surface reliability}: our current measurement relies on
+        \item surface reliability: our current measurement relies on
         the processes of installation and getting started tutorials. However,
@@
         with each software, and record any errors.
-        \item \textit{surface robustness}: we used damaged images as inputs for
+        \item surface robustness: we used damaged images as inputs for
         this measuring MI software. This process is similar to fuzz testing
@@
         identify tools and libraries to automate this process.
-        \item \textit{surface usability}: we can design usability tests and test
+        \item surface usability: we can design usability tests and test
         all software projects with end-users. The end-users can be volunteers
         and domain experts.
-        \item \textit{surface understandability}: our current method does not
+        \item surface understandability: our current method does not
         require understanding the source code. As software engineers, perhaps we
@@
         documentation, try to understand the logic, and score the ease of the
-        process.
-    \end{itemize}
-	\item we can further automate the measurements on the grading temple in
-	Appendix \ref{ap_grading_template}. For example, with automation scripts and
-	the GitHub API, we may save significant time on retrieving the GitHub
-	metrics. We have started to build a tool for this purpose, with its source
-	code at this repository
-	\hyperlink{https://github.com/smiths/AIMSS/tree/master/StateOfPractice/GitHubMetricsCollector}{https://github.com/smiths/AIMSS/.../GitHubMetricsCollector}.
-	This GitHub Metrics Collector can take GitHub repository links as input,
-	automatically collect metrics from the GitHub API, and record the results.
-	We can improve and use this tool in our future projects;
-	\item the grading standard can be more explicit. For example, we can
-	explicitly define scores for each item in the grading temple.
+        process.  Ideas for getting started are available in \citet{SmithEtAl2021}.
+        \item measure modifiability as part of the measurement of
+        maintainability.  An experiment could be conducted asking participants
+        to make modifications, observing the study subjects during the
+        modifications, testing the resulting software and surveying the
+        participants \citep{SmithEtAl2021}.
+    \end{itemize}
+	\item we can further automate the measurements on the grading template. For
+	example, with automation scripts and the GitHub API, we may save significant
+	time on retrieving the GitHub metrics through a GitHub Metric Collector.
+	This Collector can take GitHub repository links as input, automatically
+	collect metrics from the GitHub API, and record the results.
+	\item the rubric for the grading standard can be made more explicit.
 	\item we can improve some interview questions. Some examples are:
 	\begin{itemize}
-	    \item in \hyperlink{q14}{Q14}, ``Do you think improving this process can
-	    tackle the current problem?" is a yes-or-no question, which is not
-	    informative enough. As mentioned in Section \ref{sec_contribution_pm},
+	    \item in {Q14}, ``Do you think improving this process can
+	    tackle the current problem?'' is a yes-or-no question, which is not
+	    informative enough. As mentioned in Section~\ref{sec_contribution_pm},
 	    most interviewees ignored it. We can change it to ``By improving this
-	    process, what current problems can be tackled?";
-	    \item in \hyperlink{q16}{Q16}, we can ask more details about the modular
-	    approach, such as "What principles did you use to divide code into
-	    modules? Can you give an example of using the principles?";
-	    \item \hyperlink{q17}{Q17} and  \hyperlink{q18}{Q18} should respectively
-	    ask \textit{understandability} to developers and \textit{usability} to
-	    end-users.
+	    process, what current problems can be tackled?'';
+	    \item in {Q16}, we can ask for more details about the modular
+	    approach, such as ``What principles did you use to divide code into
+	    modules? Can you describe an example of using your principles?'';
+	    \item {Q17} and {Q18} should respectively ask understandability to
+	    developers and usability to end-users, since there was confusion during
+	    the interviews as to which group was being discussed.
 	\end{itemize}
@@
 	\item we can mark the follow-up interview questions with keywords. For
-	example, say ``this is a follow-up question" every time asking one. Thus, we
+	example, say ``this is a follow-up question'' every time asking one. Thus, we
 	record this sentence in the transcription, and it will be much easier to
@@
 \end{itemize}
-
-\noindent In addition, we propose a few SC domains that are potentially suitable
-for future works:
-\begin{itemize}
-    \item Metallurgy
-    \item Quantitative Finance
-    \item Computational Fluid Dynamics
-    \item Basic Linear Algebra
-    \item Finite Elements
-    \item Sparse Linear Solvers
-\end{itemize}
-
-\noindent After applying our method on various domains, we can start a
-meta-study to compare the state of the practice for software in different
-domains.
 
@@
 simulate mathematical models of real world systems of engineering or scientific
-importance so that we can better understand and predict the system’s behaviour"
+importance so that we can better understand and predict the system’s behaviour''
 \citep{Smith2006}. Many researchers consider SC as the third pillar of science

### 2021-12-03 StateOfPractice/Methodology/Methodology_arXiv_Version/Methodology.tex
arXiv version of the methodology paper
Added some thoughts at the end of some potential additions to the MI SOP paper

--- 2021-10-22 StateOfPractice/Methodology/Methodology.tex
@@
 \begin{inparaenum}[i)] 
-  \item \textcolor{red}{Blue} text denotes a link to sections of this document,
+  \item \textcolor{blue}{Blue} text denotes a link to sections of this document,
   including citations 
@@
 \bibliographystyle {plainnat}
-\bibliography {../../CommonFiles/ResearchProposal}
+\bibliography {Methodology_arXiv_version}
 
 \end{document}
+
+Thoughts - add the recent research on pain points (Pinto et al, 2018 and Wiese
+et al 2020), contrast the current study to others - the others try to cover all
+research software, we are focusing on specific domains

### 2021-12-07 StateOfPractice/Papers/MedImageSoft_SOP-arXiv/MedImageSoft_SOP_AcademicRadiology.tex
Changed use of italics to be more consistent, fixed spelling mistake

--- 2021-11-30 StateOfPractice/Papers/MedImageSoft_SOP/MedImageSoft_SOP_AcademicRadiology.tex
@@
 percentage of comment lines is low, we double-check the
-\textit{understandability} of the code; if the ratio of open versus closed pull
-requests is high, we pay more attention to \textit{maintainability}.
+understandability of the code; if the ratio of open versus closed pull
+requests is high, we pay more attention to maintainability.
 
@@
 Figure \ref{fg_visibility_transparency_scores} shows the AHP scores for
-\textit{visibility/transparency}. Generally speaking, the teams that actively
+visibility/transparency. Generally speaking, the teams that actively
 documented their development process and plans scored higher. Table
@@
 NextCloud platform that we could not successfully install. Thus, we might
-underestimate the scores of its \textit{surface reliability} and \textit{surface
-robustness}. We weighted all qualities equally, which is not likely to be the
+underestimate the scores of its surface reliability and surface
+robustness. We weighted all qualities equally, which is not likely to be the
 case with all users. As a result, some projects with high popularity may not
@@
 the GithHub stars measure as well).  The Domain Expert ranked \textit{Horos} as
-their second choice, while we ranked it twelth.  Our third ranked project,
+their second choice, while we ranked it twelfth.  Our third ranked project,
 \textit{OHIF Viewer} was not listed by the Domain Expert.  Neither were the
@@
 that belongs to a specific domain of SC.
+
+TO DO
+
+- add more on comparison to existing literature surveys
+- ask Mike:
+    - for journal version ask for more figures?

### 2021-12-10 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Work in progress on LBM SOP Paper

--- 2021-07-28 StateOfPractice/Peter-Notes/Report/Report.tex
@@
-\documentclass[12pt, notitlepage]{article}
-\usepackage{amsmath, mathtools}
-\usepackage{amsfonts}
+%% 
+%% Copyright 2007-2020 Elsevier Ltd
+%% 
+%% This file is part of the 'Elsarticle Bundle'.
+%% ---------------------------------------------
+%% 
+%% It may be distributed under the conditions of the LaTeX Project Public
+%% License, either version 1.2 of this license or (at your option) any
+%% later version.  The latest version of this license is in
+%%    http://www.latex-project.org/lppl.txt
+%% and version 1.2 or later is part of all distributions of LaTeX
+%% version 1999/12/01 or later.
+%% 
+%% The list of all files belonging to the 'Elsarticle Bundle' is
+%% given in the file `manifest.txt'.
+%% 
+%% Template article for Elsevier's document class `elsarticle'
+%% with harvard style bibliographic references
+
+%\documentclass[preprint,12pt,authoryear]{elsarticle}
+%\documentclass[3p, 12pt,authoryear]{elsarticle}
+\documentclass[final, 3p, times, authoryear]{elsarticle}
+
+%% Use the option review to obtain double line spacing
+%% \documentclass[authoryear,preprint,review,12pt]{elsarticle}
+
+%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
+%% for a journal layout:
+%% \documentclass[final,1p,times,authoryear]{elsarticle}
+%% \documentclass[final,1p,times,twocolumn,authoryear]{elsarticle}
+%% \documentclass[final,3p,times,authoryear]{elsarticle}
+%% \documentclass[final,3p,times,twocolumn,authoryear]{elsarticle}
+%% \documentclass[final,5p,times,authoryear]{elsarticle}
+%% \documentclass[final,5p,times,twocolumn,authoryear]{elsarticle}
+
+%% For including figures, graphicx.sty has been loaded in
+%% elsarticle.cls. If you prefer to use the old commands
+%% please give \usepackage{epsfig}
+
+%% The amssymb package provides various useful mathematical symbols
 \usepackage{amssymb}
-\usepackage{graphicx}
-\usepackage{colortbl}
-\usepackage{xr}
+%% The amsthm package provides extended theorem environments
+\usepackage{amsmath,amsthm}
+
+%% The lineno packages adds line numbers. Start line numbering with
+%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
+%% for the whole article with \linenumbers.
+%% \usepackage{lineno}
+
 \usepackage{hyperref}
-\usepackage{longtable}
-\usepackage{xfrac}
-\usepackage{tabularx}
-\usepackage{float}
-\usepackage{siunitx}
-\usepackage{booktabs}
-\usepackage{caption}
-\usepackage{pdflscape}
-\usepackage{afterpage}
-\usepackage{titling}
-\usepackage{blindtext}
+\usepackage{multirow}
+\usepackage{pgfplots}
+\usepackage{arydshln}
+\usepackage{paralist}
 \usepackage{multicol}
-\usepackage{indentfirst}
-\usepackage[toc,page]{appendix}
-
-\usepackage[]{natbib}
-
-%\usepackage{refcheck}
-
-\hypersetup{
-    bookmarks=true,         % show bookmarks bar?
-      colorlinks=true,       % false: boxed links; true: colored links
-    linkcolor=red,          % color of internal links (change box color with linkbordercolor)
-    citecolor=green,        % color of links to bibliography
-    filecolor=magenta,      % color of file links
-    urlcolor=cyan           % color of external links
-}
+
+%% Comments
+\newif\ifcomments\commentstrue
+
+\ifcomments
+\newcommand{\authornote}[3]{\textcolor{#1}{[#3 ---#2]}}
+\newcommand{\todo}[1]{\textcolor{red}{[TODO: #1]}}
+\else
+\newcommand{\authornote}[3]{}
+\newcommand{\todo}[1]{}
+\fi
+
+\newcommand{\wss}[1]{\authornote{blue}{SS}{#1}} %Spencer Smith
+\newcommand{\jc}[1]{\authornote{red}{JC}{#1}} %Jacques Carette
+\newcommand{\zm}[1]{\authornote{magenta}{MN}{#1}} %Zahra Motamed
+\newcommand{\pmm}[1]{\authornote{cyan}{AD}{#1}} %Peter Michalski
+
+\newcommand{\notdone}[1]{\textcolor{red}{#1}}
+\newcommand{\done}[1]{\textcolor{black}{#1}}
 
@@
 
-% Used so that cross-references have a meaningful prefix
-\newcounter{defnum} %Definition Number
-\newcommand{\dthedefnum}{GD\thedefnum}
-\newcommand{\dref}[1]{GD\ref{#1}}
-\newcounter{datadefnum} %Datadefinition Number
-\newcommand{\ddthedatadefnum}{DD\thedatadefnum}
-\newcommand{\ddref}[1]{DD\ref{#1}}
-\newcounter{theorynum} %Theory Number
-\newcommand{\tthetheorynum}{T\thetheorynum}
-\newcommand{\tref}[1]{T\ref{#1}}
-\newcounter{tablenum} %Table Number
-\newcommand{\tbthetablenum}{T\thetablenum}
-\newcommand{\tbref}[1]{TB\ref{#1}}
-\newcounter{assumpnum} %Assumption Number
-\newcommand{\atheassumpnum}{P\theassumpnum}
-\newcommand{\aref}[1]{A\ref{#1}}
-\newcounter{goalnum} %Goal Number
-\newcommand{\gthegoalnum}{P\thegoalnum}
-\newcommand{\gsref}[1]{GS\ref{#1}}
-\newcounter{instnum} %Instance Number
-\newcommand{\itheinstnum}{IM\theinstnum}
-\newcommand{\iref}[1]{IM\ref{#1}}
-\newcounter{reqnum} %Requirement Number
-\newcommand{\rthereqnum}{P\thereqnum}
-\newcommand{\rref}[1]{R\ref{#1}}
-\newcounter{lcnum} %Likely change number
-\newcommand{\lthelcnum}{LC\thelcnum}
-\newcommand{\lcref}[1]{LC\ref{#1}}
-
-\usepackage{fullpage}
-\usepackage{setspace}
-\usepackage{thesis}
-
-\doublespacing
-
-\newcommand{\fulltitle}{State of the Practice for Lattice Boltzmann Method Software}
-\newcommand{\authorname}{Peter Michalski}
+\newcommand{\CC}{C\nolinebreak\hspace{-.05em}\raisebox{.4ex}{\small\bf
++}\nolinebreak\hspace{-.10em}\raisebox{.4ex}{\small\bf +}}
+
+\journal{???}
 
 \begin{document}
-\begin{singlespace}
-
-\thesistitle
-	{\MakeUppercase{\fulltitle}}
-	{\MakeUppercase{\authorname}.}
-	{Master of Engineering in Computing and Software}
-	{\copyright \ Copyright by \authorname, July 2021}
-
-\newpage
-\thispagestyle{empty}
-
-\noindent\textbf{TITLE:} State of The Practice for Lattice Boltzmann Method Software
-
-\noindent\textbf{DEGREE:} Master of Engineering (Computing and Software)
-
-\noindent\textbf{INSTITUTION:} McMaster University
-
-\noindent\textbf{AUTHOR:} Peter Michalski
-
-\noindent\textbf{SUPERVISOR:} Dr. Spencer Smith and Dr. Jacques Carette
-
-\noindent\textbf{NUMBER OF PAGES:} 123
-
-\newpage
-\thispagestyle{empty}
-\begin{center}
-	\textbf{\large Abstract}
-\end{center}
-
-We analyze the state of the practice of software development in the Lattice Boltzmann Methods software domain by quantitatively and qualitatively measuring and comparing 23 software packages.
-We present a general methodology for assessing the state of the practice of software development in scientific computing software domains.
-We present a domain analysis of the Lattice Boltzmann Methods software family, and identify candidate software packages.
-We assess these packages to answer software development related research questions to understand how software quality is impacted by software development choices, including principles, processes, and tools.
-We interview software developers to identify development pain points, and to identify how software quality is ensured.
-We use quantitative data to rank the software packages using the Analytical Hierarchy Process, and rank Ludwig, ESPResSo, and Palabos as the top three packages.
-Each of them score high in at least several of the individual qualities that we quantitatively measure.
-We compare these rankings with rankings from the software development community.
-We make recommendations, such as providing a detailed user manual and tutorials, explicitly stating the limits of the software, using user-friendly software languages, considering a peer review process, communicating development and contribution information, and using continuous integration and project management tools.\\
-
-\noindent\textbf{Keywords}: Lattice Boltzmann Methods, Scientific Computing, Software Engineering, Software Family, Software Quality
-
-\newpage
-\pagenumbering{roman}
-\tableofcontents
-\addtocontents{toc}{}
-
-~\newpage
-
-\listoffigures
-
-\listoftables
-
-~\newpage
-
-\clearpage
-\section*{Reference Material}
-
-\subsection*{Software Engineering Related Definitions and Acronyms}
-
-\noindent\textbf{AHP}: Analytical Hierarchy Process\\
-
-\noindent\textbf{Commonality}: A requirement or goal common to all family members.\\
-
-\noindent\textbf{CPU}: Central Processing Unit\\
-
-\noindent\textbf{Goal}: Goals capture, at different levels of abstraction, the various objectives the system under consideration should achieve \citep{van2001goal}.\\
-
-\noindent\textbf{GPU}: Graphics Processing Unit\\
-
-\noindent\textbf{LOC}: Lines of Code\\
-
-\noindent\textbf{OS}: Operating System\\
-
-\noindent\textbf{OTS}: Off-The-Shelf\\
-
-\noindent\textbf{Requirements}: A software requirement is: \textit{i}) a condition or capability needed by a user to solve a problem or achieve an objective; \textit{ii}) a condition or capability that must be met or possessed by a system or system component to satisfy a contract, standard, specification, or other formally imposed document; or, \textit{iii}) a documented representation of a condition or capability as in the above two definitions \citep{thayer2000ieee}.\\
-
-\noindent\textbf{SCS}: Scientific Computing Software\\
-
-\noindent\textbf{Software Family}: A set of programs with an extensive amount of common properties \cite{parnas1976design}.\\
-
-\noindent\textbf{Variability}: A requirement or goal that varies between family members.
-
-\newpage
-\subsection*{Lattice Boltzmann Related Definitions and Acronyms}
-
-\noindent\textbf{1D} 1-Dimensional\\
-
-\noindent\textbf{2D}: 2-Dimensional\\
-
-\noindent\textbf{3D}: 3-Dimensional\\
-
-\noindent\textbf{BGK}: Bhatnagar-Gross-Krook \citep{bhatnagar1954model}\\
-
-\noindent\textbf{CFD}: Computational Fluid Dynamics\\
-
-\noindent\textbf{MRT}: Multi-Relaxation-Time\\
-
-\noindent\textbf{SRT}: Single-Relaxation-Time\\
-
-\noindent\textbf{TRT}: Two-Relaxation-Time\\
-
-\noindent\textbf{LBM}: Lattice Boltzmann Methods\\
-
-\noindent\textbf{LBS}: Lattice Boltzmann Solvers\\
-
-\noindent\textbf{Velocity Directions}: The number of links connecting to each lattice node in the chosen model from neighbouring nodes.
-All nodes in a chosen lattice model will have the same number of links.
-A single link will connect between two adjacent nodes.
-
-\end{singlespace}
-\newpage
-\pagenumbering{arabic}
+
+\begin{frontmatter}
+
+%% Title, authors and addresses
+
+%% use the tnoteref command within \title for footnotes;
+%% use the tnotetext command for theassociated footnote;
+%% use the fnref command within \author or \affiliation for footnotes;
+%% use the fntext command for theassociated footnote;
+%% use the corref command within \author for corresponding author footnotes;
+%% use the cortext command for theassociated footnote;
+%% use the ead command for the email address,
+%% and the form \ead[url] for the home page:
+%% \title{Title\tnoteref{label1}}
+%% \tnotetext[label1]{}
+%% \author{Name\corref{cor1}\fnref{label2}}
+%% \ead{email address}
+%% \ead[url]{home page}
+%% \fntext[label2]{}
+%% \cortext[cor1]{}
+%% \affiliation{organization={},
+%%            addressline={}, 
+%%            city={},
+%%            postcode={}, 
+%%            state={},
+%%            country={}}
+%% \fntext[label3]{}
+
+\title{State of the Practice for Lattice Boltzmann Method Software}
+
+%% use optional labels to link authors explicitly to addresses:
+%% \author[label1,label2]{}
+%% \affiliation[label1]{organization={},
+%%             addressline={},
+%%             city={},
+%%             postcode={},
+%%             state={},
+%%             country={}}
+%%
+%% \affiliation[label2]{organization={},
+%%             addressline={},
+%%             city={},
+%%             postcode={},
+%%             state={},
+%%             country={}}
+
+\author[CAS]{Spencer Smith}
+\author[CAS]{Peter Michalski}
+\author[CAS]{Jacques Carette}
+\author[ME]{Zahra Motamed}
+
+\affiliation[CAS]{organization={McMaster University, Computing and Software
+Department},%Department and Organization
+            addressline={1280 Main Street West}, 
+            city={Hamilton},
+            postcode={L8S 4K1}, 
+            state={Ontario},
+            country={Canada}}
+
+\affiliation[ME]{organization={McMaster University, Mechanical Engineering},%Department and Organization
+            addressline={1280 Main Street West}, 
+            city={Hamilton},
+            postcode={L8S 4K1}, 
+            state={Ontario},
+            country={Canada}}
+
+\begin{abstract}
+
+    We analyze the state of software development practice for Lattice Boltzmann
+    solvers by quantitatively and qualitatively measuring and comparing 24
+    software packages for 10 software qualities (installability, correctness/
+    verifiability, reliability, robustness, usability, maintainability,
+    reusability, understandability, visibility/transparency and
+    reproducibility). Our reproducible analysis method employs a measurement
+    template (containing 103 measures that are manually and automatically
+    extracted from the project repository) and developer interviews (consisting
+    of a set of 20 questions). From the measurement results, we ranked the
+    software using the Analytic Hierarchy Process (AHP), a multi-criteria
+    decision making method appropriate for cases with a mix of qualitative and
+    quantitative factors. Our ranking was roughly consistent with ranking the
+    repositories by the number of GitHub stars, although the number one project
+    by star count (Sailfish) is ranked 9th by our methodology. Our top three
+    packages were Ludwig, ESPResSo, and Palabos (all in the top 10 by star
+    count). We interviewed 4 developers to gain insight into their current pain
+    points. Identified challenges include lack of development time, lack of
+    funding, and difficulty with ensuring correctness.  Based on our assessment,
+    we make the following recommendations for current and future LBM projects:
+    provide a detailed user manual and tutorials, explicitly state the limits of
+    the software, use user-friendly programming languages, consider a peer
+    review process, communicate development and contribution information, and
+    use continuous integration and project management tools.
+
+\end{abstract}
+
+%%Graphical abstract
+%\begin{graphicalabstract}
+%\includegraphics{grabs}
+%\end{graphicalabstract}
+
+%%Research highlights
+%\begin{highlights}
+%\item Research highlight 1
+%\item Research highlight 2
+%\end{highlights}
+
+\begin{keyword}
+	Lattice Boltzmann Method (LBM), scientific computing, software engineering,
+	software quality, Analytic Hierarchy Process
+\end{keyword}
+
+\end{frontmatter}
+
+%% \linenumbers
+
 \section{Introduction}
 
-We analyze the development of Computational Fluid Dynamics (CFD) software packages that use Lattice Boltzmann Methods (LBM).
-These LBM algorithms consider the behaviours of a collection of particles as a single unit of an intermediate size, and predict their positional probability as they move through a lattice structure.
-Figure \ref{circularflow} presents the streamlines of converged flow past a stationary circular cylinder with varied Reynolds number, a common task of LBM.
+We analyze the development of Computational Fluid Dynamics (CFD) software
+packages that use the Lattice Boltzmann Method (LBM). LBM packages form a family
+of algorithms for simulating single-phase and multiphase fluid flows, often
+incorporating additional physical complexities \citep{chen1998lattice} such as
+reflective and non-reflective boundaries. They consider the behaviour of a
+collection of particles as a single unit at the mesoscopic scale, between the
+nanoscopic and microscopic scales. LBM solvers predict the positional
+probability of a collection of particles moving through a lattice structure
+following a two step process: i) streaming, where the particles
+move along the lattice via links; and, ii) colliding, where energy and
+momentum is transferred among particles that collide \citep{bao2011lattice}. As
+an example of the output of LBM, Figure~\ref{circularflow} presents the streamlines of
+converged flow past a stationary circular cylinder with varying Reynolds numbers.
+\wss{Zahra might have a better example?  A journal publication isn't likely
+going to accept a figure produced for another paper.}  LBM has several
+advantages over conventional CFD methods, including a simple calculation
+procedure, improved parallelization, and robust handling of complex geometries
+\citep{ganji2015application}.
 
@@
 	\begin{center}
-		\includegraphics[width=0.7\textwidth]{circularflow}
-		\caption{Streamlines of flow past a stationary circular cylinder at Re = 10, 20, and 40 \citep{chen2021phase}}
+		\includegraphics[width=0.7\textwidth]{./figures/circularflow}
+		\caption{Streamlines of flow past a stationary circular cylinder at
+		Reynolds number = 10, 20, and 40 \citep{chen2021phase}}
 		\label{circularflow}
@@
 
-LBM have several advantages over conventional CFD methods, including a simple calculation procedure, improved parallelization, and robust handling of complex geometries \citep{ganji2015application}.
-LBM are further detailed in Section \ref{domainanalysis}.
-
-Scientific Computing Software (SCS) is defined as the use of computer tools to analyze or simulate mathematical models of real world systems \citep{smith2006systematic}.
-Given the importance of such software, scientists and engineers desire methods and tools to sustainably develop it with high quality.
-This has led to the formation of groups like the Software Sustainability Institute (\href{https://www.software.ac.uk/}{SEI}) and Better Scientific Software (\href{https://bssw.io/}{BSS}).
-
-The goal of this report is to analyze the current state of development of LBM SCS to provide insight into its best practices and to offer guidance for future development.
-We want to understand how software quality in the LBM SCS family is impacted by software development choices, including principles, processes, and tools.
-
-In this report red text denotes internally linked sections of this document. Cyan text denotes an external URL link. Green text denotes a link to the bibliography.
-
-\subsection{Research Questions}\label{purpose}
-
-The LBM software packages are assessed to answer the following research questions:
+A small sample of important applications of LBM include the following: designing
+fuel cells \citep{ZhangEtAl2018}, modelling groundwater flow
+\citep{AnwarAndSukop2009}, and analyzing the flow of blood in the cardiovascular
+system \citep{SadeghiEtAl2020}.  As these examples illustrate, LBM can be used
+for tackling problems that impact such areas as environmental policy,
+manufacturing, and health and safety.  Given the important applications of LBM,
+users of the libraries will be concerned with software qualities like
+reliability, robustness, reproducibility, performance, correctness and
+verifiability.  Given the value of their time, developers that create or modify
+LBM libraries will have additional quality concerns, including maintainability,
+reusability, and understandability.  With the quality considerations in mind,
+and the potentially overwhelming number of choices for LBM libraries, the time
+is right to assess the state of the practice.  The goal of this report is to
+analyze the current state of LBM software development to provide insight into
+its best practices and to offer guidance for future development.
+
+To focus our efforts, we devised five research questions, which can be applied
+to LBM, as well as to other domains, of research software \citep{SmithEtAl2021}.
+The first question below refers to artifacts, which is the term we use to
+represent the documents, scripts and code that constitutes a software
+development project. Example artifacts include requirements, specifications,
+user manuals, unit tests, system tests, usability tests, build scripts, API
+(Application Programming Interface) documentation, READMEs, license documents,
+process documents, and code.
 
@@
 	\item What artifacts are present in current software packages? 
-	\item What tools (development, dependencies, project management) are used by current software packages?
+	\item What tools (development, dependencies, project management) are used by
+	current software packages?
 	\item What principles, processes, and methodologies are used in the development
@@
 	\end{enumerate} 
-	\item How does software designated as high quality by this methodology compare	with top rated software by the community?
+	\item How does software designated as high quality by this methodology
+	compare	with top rated software by the community?
 \end{enumerate}
 
-The packages are assessed for a given set of quality attributes using quantitative and qualitative data.
-The qualities are defined in Section \ref{softwarequalities}.
-Quantitative data is measured using quality metrics.
-The findings are presented in Section \ref{AHPresults}.
-Qualitative data is gathered from interviews with software package developers.
-The interview findings are presented in Section \ref{interviewresults}.
-In this context the purposes of the report are as follows:
-
-\begin{itemize}
-	\item Develop and test an updated methodology for assessing the state of the practice of SCS projects
-	\item Describe the specific example of LBM software structure and goals
-	\item Report on the measure of a subset of LBM software packages along quality metrics
-	\item Evaluate the state of the practice of LBM software development along quality attributes
-	\item Make suggestions for improving LBM software along quality attributes
-	\item Make suggestions for improving state of the practice of SCS assessments
-\end{itemize}
-
-\subsection{Motivation}
-
-The purpose of state of the practice assessments is to understand how software quality is impacted by software development choices, including principles, processes, and tools, within SCS communities.
-This knowledge can be used to guide future development of SCS, specifically along quality attributes, and to reduce software quality failures.
-This work reports on the state of the practice of LBM software development and makes suggestions on improving the quality of software in this domain.
-
-This assessment of the state of the practice of LBM software development builds off of prior work on assessing the state of research software development.
-The previous state of the practice assessments include domains such as Geographic Information Systems \citep{SmithEtAl2018_arXivGIS}, Mesh Generators \citep{SmithEtAl2016}, Oceanographic Software \citep{smith2015state}, Seismology software \citep{SmithEtAl2018}, and statistical software for psychology \citep{SmithEtAl2018_StatSoft}.
-
-In the course of this assessment we updated the methodology that was used during previous assessments.
-Details of the new methodology are in \citep{methodology2020}.
-The previous set of research questions were critically assessed and modified.
-The updated questions are listed in Section \ref{purpose}.
-In this re-boot we collected more quantitative and qualitative data, focused on software quality attributes, and added more measures, including collecting empirical data and interviewing developers.
-We have also added a domain analysis to better characterize the functionality provided by the software, and have leveraged the expertise of a domain expert.
-As in past assessments, the collected data was combined to rank the software using the Analytical Hierarchy Process (AHP).
-The domain expert was consulted to verify the ordering.
-
-\subsection{Scope}
-
-We analyze a filtered set of LBM software packages along quantitative and qualitative measures.
-Many of these measures are captured using surface measurements, which can be categorized as initial and easy to capture measurements, of the underlying quality; they may not represent the true quality of the software.
-Surface measurements are taken as they allow us to apply the same measurement, with reasonable effort, along all software packages despite technical and functional variabilities among the set of software packages.
-Not all qualities are quantitatively measured.
-For example, performance is not measured by running the code.
-A surface investigation of the documentation of each software package for performance information is conducted instead.
-The measured software packages are open source.
-Some recommendations may not apply to closed-source software.
-Practices surrounding close source software development may differ.
-
-The report addresses what was done during the development of the software packages to address the quality attributes listed in Section \ref{softwarequalities}.
-It then provides general guidance on how to improve these qualities when developing LBM software.
-It does not make suggestions on what should have been done or should be done for any one specific package.
-Best practices may differ among software packages due to their inherent organizational and technical differences.
-
-\subsection{Organization}
-
-The report is organized as follows:
-
-\begin{itemize}
-	\item \textbf{Introduction} to the report, including its purpose, motivation, scope, and organization.
-	\item LBM software \textbf{Domain Analysis}.
-A domain analysis consists of systematically identifying and documenting the commonalities, variabilities, and terminology of a software family \citep{Weiss1997}.
-	The purpose of this analysis in this state of the practice assessment is to better classify the different software products based on their functionality.
-	\item \textbf{Methodology} of this state of the practice assessment, including the steps of the methodology, software quality definitions and how these qualities are assessed in this report, an overview of candidate software selection, empirical measures gathered, and the AHP.
-This follows a general methodology for assessing the state of the practice for SCS domains.
-	\item \textbf{Quantitative Results}. 
-	\item \textbf{Qualitative Results}. 
-	\item An analysis of the results and \textbf{Answers To Research Questions}.
-	\item \textbf{Conclusion} to the report including comments on suggestions for future state of the practice assessments.
-	\item \textbf{Appendix}: This section includes our Research Questions, the Measurement Template, Grading Template, Ethics Approval, and Developer Interview Questions.
-\end{itemize}
-
-\newpage
-
-\section{Domain Analysis}\label{domainanalysis}
-
-A domain analysis consists of systematically identifying and documenting the commonalities, variabilities, and parameters of variation of a software family \citep{Weiss1997}.
-A software family is defined by \cite{parnas1976design} as “a set of programs whose common properties are so extensive that it is advantageous to study the common properties of the programs before analyzing individual members”.
-
-We added this domain commonality and variability analysis to our original methodology because the first time we did the State of the Practice exercise we found we weren't comparing ``apples to apples".
-Software packages can have considerable variation, even when the packages appear to be in the same software family.
-Viewing the ranking exercise as a means for selecting the right tool for a job, functionality is important.
-A 2D Lattice Boltzmann solver is not a substitute for a 3D Lattice Boltzmann solver when the latter is required.
-Therefore, this exercise seeks to classify the different software products based on their functionality.
-Nonfunctional qualities are assessed through other means in our process.
-For instance, through usability experiments as noted in the \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Methodology.pdf}{Methodology for Assessing the State of the Practice for Domain X} document.
-
-This domain analysis of the LBM SCS family is organized into the four subsection located below.
-The first subsection reviews the basics of Lattice Boltzmann systems.
-The next three subsections consist of lists of commonalities, variabilities, and parameters of variation, respectively.
-These three sections form the heart of the domain analysis and include an extensive set of cross-references to demonstrate the relationships between the different items.
-
-\subsection{Lattice Boltzmann Systems}
-
-LBM are a family of fluid dynamics algorithms for simulating single-phase and multiphase fluid flows, often incorporating additional physical complexities \citep{chen1998lattice} such as reflective and non-reflective boundaries.
-They consider the behaviours of a collection of particles as a single unit at the mesoscopic scale, between the nanoscopic and microscopic scales.
-These methods predict the positional probability of a collection of particles moving through a lattice structure.
-Off-the-shelf (OTS) Lattice Boltzmann Solvers (LBS) allow for a range of fluid and physical model input parameters, computational parameters, and output parameters.
-
-LBS model fluid dynamics within a boundary using a predefined lattice structure and a two step calculation process.
-The first process is streaming, where the particles move along the lattice via links.
-The second process is collision, where energy and momentum is transferred among particles that collide \citep{bao2011lattice}.
-There are many standardized lattice models - individual solvers within the family might only use a subset of them.
-LBM use the initial parameters of the fluid to find the probability of where along the lattice linkages a group of particles are most likely to travel.
-It then moves the particles into the next node, and transfers the energy and momentum if a collision occurs.
-Then the process repeats for the duration of the modeling instance.
-
-\subsection{Commonalities}\label{comm_sec}
-
-This section lists common features among potential family members.
-The commonalities are organized using the following abstraction of the system, which can be used to describe all Lattice Boltzmann systems: input information, generate the simulation, output the results.
-Section \ref{comm_lbm} describes the commonalities for the simulation step.
-Section \ref{comm_in} highlights the input information that is required for Lattice Boltzmann systems.
-The next section, Section \ref{comm_out}, shows the common features for the output of Lattice Boltzmann systems, such as the requirement that mesh information be written to files.
-Although the output information could simply be written to the computer’s memory, in all practical applications it is desirable to have a persistent record of the output that was created.
-
-Each commonality below uses the same structure.
-All of the commonalities are assigned a unique item number, which takes the form of a natural number with the prefix ``C''.
-Following this, a description of the commonality is provided along with a list of related variabilities, which are given as hyperlinks that allow navigation of the document to the text describing the variability.
-
-\subsubsection{Lattice Boltzmann Method Solvers}\label{comm_lbm}
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
-		\hline
-		\bf Item Number& C\refstepcounter{comnum}\thecomnum{}\label{lattice} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
-		\hline
-		\bf Description & A lattice discretizes a computational domain into a finite number of points. All LBS discretize the computational domain using a regular, evenly spaced grid within a boundary.\\
-		\hline
-		\bf Related Variability & \vref{numveldir} \vref{varcolop} \vref{varboundary}\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
-		\hline
-		\bf Item Number& C\refstepcounter{comnum}\thecomnum\label{collisionoperator} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
-		\hline
-		\bf Description & All Lattice Boltzmann versions use a collision operator which concerns collisions between particles.
-Collision operators map collisions of particles within the lattice space.
-The Bhatnagar-Gross-Krook Collision Operator is a common LBM collision operator that preserves continuity for a discretized model, for each velocity direction $i$.
-Its equation is $\mathrm{\Omega_i} = \frac{1}{\tau}(f_{i}^{eq}-f_{i})$, where $\tau$ is the relaxation rate towards equilibrium, $f^{eq}$ is the equilibrium particle probability distribution function, and $f$ is the particle probability distribution function.\\
-		\hline
-		\bf Related Variability & \vref{dimensions} \vref{numveldir}\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
-		\hline
-		\bf Item Number& C\refstepcounter{comnum}\thecomnum\label{probabilitydensityfunc} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
-		\hline
-		\bf Description & All Lattice Boltzmann versions use a probability density function to give the probability that fluid has moved into a specific domain.\\
-		\hline
-		\bf Related Variability & \vref{dimensions}\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
-		\hline
-		\bf Item Number& C\refstepcounter{comnum}\thecomnum\label{equildistfunc} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
-		\hline
-		\bf Description & Every Lattice Boltzmann version uses an equilibrium distribution function to capture the probability distribution of particles.\\
-		\hline
-		\bf Related Variability & \vref{varedf} \vref{edfstorage} \vref{edfcoeff}\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
-		\hline
-		\bf Item Number& C\refstepcounter{comnum}\thecomnum\label{Boltzmanntransequ} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
-		\hline
-		\bf Description & Every Lattice Boltzmann version uses a Boltzmann transport equation to describe the statistical behaviour of a system that does not have collisions.
-The equation is $(\frac{d}{dt} + \mathrm{e} \cdot \nabla_\mathrm{x} + \frac{F}{m} \cdot \nabla_\mathrm{e})f = \Omega(t)$, where $\mathrm{e}$ is the velocity, $\nabla_\mathrm{x}$ is the position vector gradient, $F$ is the force of the fluid, $m$ is the mass, $\nabla_\mathrm{e}$ is the velocity gradient, $f$ indicates the probability density function, and $\Omega(t)$ is the collision operator as a function of time.\\
-		\hline
-		\bf Related Variability & \vref{vartraneq}\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-\subsubsection{Input}\label{comm_in}
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
-		\hline
-		\bf Item Number& C\refstepcounter{comnum}\thecomnum\label{fluidandboundaryinfo} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
-		\hline
-		\bf Description & The LBS require fluid, model, and boundary information for the problem.
-This includes, but is not limited to, fluid acceleration rate, velocity, and viscosity, as well as the number of dimensions in the lattice model, and the number of velocity directions in the lattice.\\
-		\hline
-		\bf Related Variability & \vref{interface}\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-\subsubsection{Output}\label{comm_out}
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
-		\hline
-		\bf Item Number& C\refstepcounter{comnum}\thecomnum\label{predictionstomemory} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
-		\hline
-		\bf Description & LBS write fluid predictions, such as the output of the transport equation, to memory.\\
-		\hline
-		\bf Related Variability & \vref{varvisual} \vref{formatoutput} \\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-\subsection{Variabilities}\label{var_sec}
-This section provides a list of characteristics that may vary among family members.
-As in Section \ref{comm_sec}, the first three subsections on variabilities are organized into the following sublists: Simulation Models, Input and Output.
-The final subsection lists variabilities that can be characterized as system constraints.
-
-As for the commonalities, each variability is labelled with a unique item number.
-In this case the numbers are prepended with the letter ``V''.
-The other three headings provided for each variability are: Description, Related Commonality, and Related Parameter.
-The related commonalities and parameters are given as a set of identifiers that respectively refer back to the previous section on commonalities or refer forward to the next section on parameters of variation.
-
-\subsubsection{Lattice Boltzmann Method Solvers}
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
-		\hline
-		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{parallel} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & LBS may use a framework for parallel processing of the model.\\
-		\hline
-		\bf Related Commonality & None\\
-		\hline
-		\bf Related Parameter & \pref{parparallel}\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-~\newline\\
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
-		\hline
-		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{varedf} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & Different versions of an equilibrium distribution function can capture the probability distribution of particles.\\
-		\hline
-		\bf Related Commonality & \cref{equildistfunc}\\
-		\hline
-		\bf Related Parameter & \pref{paredf}\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
-		\hline
-		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{edfstorage} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & Storage patters for distribution function can vary.\\
-		\hline
-		\bf Related Commonality & \cref{equildistfunc}\\
-		\hline
-		\bf Related Parameter & \pref{parstorage}\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
-		\hline
-		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{edfcoeff} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & Coefficients used with the distribution function can vary. These are based on the number of velocity directions in the model.\\
-		\hline
-		\bf Related Commonality & \cref{equildistfunc}\\
-		\hline
-		\bf Related Parameter & \pref{paredfcoeff}\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
-		\hline
-		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{dimensions} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & The number of dimensions in the lattice of the model can vary.\\
-		\hline
-		\bf Related Commonality & \cref{lattice} \cref{probabilitydensityfunc}\\
-		\hline
-		\bf Related Parameter & \pref{pardimensions}\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
-		\hline
-		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{numveldir} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline 
-		\bf Description & The number of velocity directions in the lattice of the model can vary.\\
-		\hline
-		\bf Related Commonality & \cref{lattice} \cref{collisionoperator}\\
-		\hline
-		\bf Related Parameter & \pref{parveldir}\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
-		\hline
-		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{varcolop} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & Various collision operators can be used.\\
-		\hline
-		\bf Related Commonality & \cref{lattice} \cref{collisionoperator}\\
-		\hline
-		\bf Related Parameter & \pref{parcollisop}\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
-		\hline
-		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{vartraneq} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & Various transport equations can be used to describe the statistical behaviour of the system\\
-		\hline
-		\bf Related Commonality & \cref{Boltzmanntransequ}\\
-		\hline
-		\bf Related Parameter & \pref{partransequ}\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
-		\hline
-		\bf Item Number&  V\refstepcounter{varnum}\thevarnum\label{numberoffluids} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & The number of fluids allowed in a multifluid simulation.\\
-		\hline
-		\bf Related Commonality & \cref{fluidandboundaryinfo}\\
-		\hline
-		\bf Related Parameter & \pref{parfluids}\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
-		\hline
-		\bf Item Number&  V\refstepcounter{varnum}\thevarnum\label{typefluidparm} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & The type of fluid parameters.\\
-		\hline
-		\bf Related Commonality & \cref{fluidandboundaryinfo}\\
-		\hline
-		\bf Related Parameter & \pref{partypeparm}\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
-		\hline
-		\bf Item Number&  V\refstepcounter{varnum}\thevarnum\label{varboundary} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & The boundary of the lattice can have various conditions.\\
-		\hline
-		\bf Related Commonality & \cref{lattice}\\
-		\hline
-		\bf Related Parameter & \pref{parboundary}\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-\subsubsection{Input}
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
-		\hline
-		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{interface} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & The input interface can vary between LBS.\\
-		\hline
-		\bf Related Commonality & \cref{fluidandboundaryinfo}\\
-		\hline
-		\bf Related Parameter & \pref{parinterface}\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-\subsubsection{Output}
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
-		\hline
-		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{varvisual} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & Visual presentation of the prediction.\\
-		\hline
-		\bf Related Commonality & \cref{predictionstomemory}\\
-		\hline
-		\bf Related Parameter & \pref{parvarvisual}\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
-		\hline
-		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{formatoutput} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & Format of prediction information.\\
-		\hline
-		\bf Related Commonality & \cref{predictionstomemory}\\
-		\hline
-		\bf Related Parameter & \pref{parformatoutput}\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-\subsubsection{System Constraints}
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
-		\hline
-		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{processing} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & Hardware that processes the calculations\\
-		\hline
-		\bf Related Commonality & None \\
-		\hline
-		\bf Related Parameter & \pref{parprocessing}\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
-		\hline
-		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{operatingsystem} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & Operating systems on which LBS run.\\
-		\hline
-		\bf Related Commonality & None \\
-		\hline
-		\bf Related Parameter & \pref{paropsystem}\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
-		\hline
-		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{storage} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
-		\hline
-		\bf Description & Amount of storage and memory needed for the LBS.\\
-		\hline
-		\bf Related Commonality & None \\
-		\hline
-		\bf Related Parameter & \pref{parstorageconst}\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-\newpage
-\subsection{Parameters of Variation}
-This section specifies the parameters of variation for the variabilities listed in Section \ref{var_sec}.
-They are organized into the same five subcategories as employed previously: Simulation Models, Input, Output, and System Constraints.
-
-Each parameter of variation is given a unique identifier of the form “P” followed by a natural number.
-The corresponding variability is listed and a hyperlink is provided that allows navigation back to the appropriate item in Section \ref{var_sec}.
-The final entry for each parameter of variation is the binding time, which is the time in the software lifecycle when the variability is fixed.
-The binding time could be during specification, or during building
-of the system (build time), or during execution of the system (run time).
-It is possible to have a mixture of binding times.
-For instance, a parameter of variation could have a binding time of “specification or building” to represent that the parameter could be set at specification time, or it could be postponed until the given family member is built.
-The choice of postponing the decision until the build could be associated with the presence of a domain specific language that would allow postponing decisions on the values of the parameter of variation.
-
-\subsubsection{Lattice Boltzmann Method Solvers}
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
-		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parparallel} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{parallel}\\
-		\hline
-		\bf Range of Parameters & OpenMP, OpenCL, CUDA, MPI are used if the execution of the LBS is parallelized.\\
-		\hline
-		\bf Binding Time & Build Time\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
-		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{paredf} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{varedf}\\
-		\hline
-		\bf Range of Parameters & Equilibrium approximation varies between incompressible or compressible models.\\		
-		\hline
-		\bf Binding Time & Build Time\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
-		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parstorage} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{edfstorage}\\
-		\hline
-		\bf Range of Parameters & Various data structures can be used to store function output, including single and multi-dimensional arrays, depending on the problem model and developer preferences.\\
-		\hline
-		\bf Binding Time & Build Time\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
-		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{paredfcoeff} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{edfcoeff}\\
-		\hline
-		\bf Range of Parameters & Numerous coefficients for equilibrium distribution function based on number of velocity directions.
-The number of velocity directions is typically 2, 3, 5, 9, 13, 15, 19, or 27.\\
-		\hline
-		\bf Binding Time & Build Time\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
-		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{pardimensions} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{dimensions}\\
-		\hline
-		\bf Range of Parameters & LBS model has 1, 2, or 3 dimensions.\\
-		\hline
-		\bf Binding Time & Build Time or Run Time\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
-		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parveldir} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{numveldir}\\
-		\hline
-		\bf Range of Parameters & One dimensional models include options of 2, 3, and 5 velocity directions.
-Two dimensional models include options of 9, 13, and 15 velocity directions.
-Three dimensional models include options of 15, 19, and 27 velocity directions.
-Figure \ref{dimension_model} from \citep{thurey2009stable} illustrates two commonly used LBM models in two and three dimensions.\\
-		\hline
-		\bf Binding Time & Build Time or Run Time\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-
-\begin{figure}[h!]
-	\begin{center}
-		\includegraphics[width=0.7\textwidth]{dimension_model}
-		\caption{Commonly used LBM models in two and three dimensions}
-		\label{dimension_model}
-	\end{center}
-\end{figure}
-
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
-		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parcollisop} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{varcolop}\\
-		\hline
-		\bf Range of Parameters & SRT, TRT, MRT, BGK collision operators.\\
-		\hline
-		\bf Binding Time & Build Time\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
-		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{partransequ} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{vartraneq}\\
-		\hline
-		\bf Range of Parameters & Collision and collision free transport equations.
-LBS sometimes use one or the other, often the collision transport equation.
-A Boolean parameter could be used to select between these equations in systems that can apply either equation.\\
-		\hline
-		\bf Binding Time & Build Time\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
-		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parfluids} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{numberoffluids}\\
-		\hline
-		\bf Range of Parameters & LBS can model a natural number of fluids.\\
-		\hline
-		\bf Binding Time & Build Time\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
-		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{partypeparm} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{typefluidparm}\\
-		\hline
-		\bf Range of Parameters & LBS fluid parameters include Reynolds Number, density, viscosity, time, pressure, force, direction, relaxation rate, turbulence.
-All of these parameters have the type of real number.\\
-		\hline
-		\bf Binding Time & Build Time\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
-		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parboundary} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{varboundary}\\
-		\hline
-		\bf Range of Parameters & Lattice boundary can have reflective or non-reflective conditions.
-Some LBS will only model reflective or non-reflective conditions.
-If there is a choice then this can be indicated by a Boolean parameter.\\
-		\hline
-		\bf Binding Time & Build Time\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-\subsubsection{Input}
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
-		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parinterface} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{interface}\\
-		\hline
-		\bf Range of Parameters & Input can be graphical, text or file.\\
-		\hline
-		\bf Binding Time & Build Time\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-\subsubsection{Output}
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
-		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parvarvisual} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{varvisual}\\
-		\hline
-		\bf Range of Parameters & LBS can provide 1D, 2D, and 3D rendering of  the model.\\
-		\hline
-		\bf Binding Time & Build Time or Run Time\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
-		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parformatoutput} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{formatoutput}\\
-		\hline
-		\bf Range of Parameters & LBS prediction information is output in either text or binary format.\\
-		\hline
-		\bf Binding Time & Build Time\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-\subsubsection{System Constraints}
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
-		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parprocessing} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{processing}\\
-		\hline
-		\bf Range of Parameters & The LBS model can be calculated on the CPU or GPU.\\
-		\hline
-		\bf Binding Time & Build Time\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
-		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{paropsystem} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{operatingsystem}\\
-		\hline
-		\bf Range of Parameters & LBS can be run on Windows, MacOS, or Linux versions.\\
-		\hline
-		\bf Binding Time & Build Time\\
-		\hline
-	\end{tabular}
-\end{minipage}
-~\newline
-~\newline
-\noindent
-\begin{minipage}{\textwidth}
-	\renewcommand*{\arraystretch}{1.5}
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}
-		\hline
-		\bf Item Number& P\refstepcounter{parnum}\theparnum\label{parstorageconst} \\
-		\hline
-	\end{tabular}\\
-
-	\begin{tabular}{| p{\colEwidth} | p{\colFwidth}|}		
-		\hline
-		\bf Corresponding Variability & \vref{storage}\\
-		\hline
-		\bf Range of Parameters & The amount of memory and storage varies between LBS.\\
-		\hline
-		\bf Binding Time & Run Time\\
-		\hline
-	\end{tabular}
-\end{minipage}\\
-
-\newpage
-\section{Methodology}
-
-In this project we set out to answer the research questions listed in Section \ref{purpose} for the LBM SCS domain.
-The process involved systematically measuring and analyzing members of this software family along the quality attributes listed in the research questions and described in Section \ref{softwarequalities}.
-The methodology includes gathering quantitative and qualitative measurements.
-A goal of the project is to produce a quality assessment for LBM software packages.
-
-The methodology used in this LBM software assessment is a general state of the practice methodology that can be applied to any SCS domain.
-It was developed as an update to previous state of the practice exercises.
-
-We collected quantitative data using the measures found in the measurement template in Appendix \ref{measurementtemplate2}.
-Some of this was empirical software engineering related data, such as the number of files, number of lines of code (LOC), percentage of issues that are closed, etc.
-Most of the data was gathered by manually investigating the software, its source code, and its artifacts, while some was gathered using the empirical measurement tools discussed in Section \ref{empiricalmeasures}.
-
-We also collected qualitative data by interviewing the software package developers and asking them the questions found in Appendix \ref{interviewquestions}.
-Ethics clearance information can be found in Appendix \ref{ethicsapproval}.
-Furthermore, we solicited the assistance of domain experts to better assess each software package by leveraging their experience to assess the functional and non-functional requirements for the software domain.
-
-This section begins by describing the steps of the overall process we used to select, measure and compare LBM software.
-This is followed by quality definitions and how these qualities were assessed in our project.
-The rest of the section provides an overview of how candidate software packages were selected and filtered, the empirical measurements and software tools that were used, and the AHP.
+We investigated the research questions by applying the general methodology
+summarized in \citet{SmithEtAl2021}.  The specific application of the
+methodology to LBM is reviewed in Section~\ref{methodology}, with the full
+details in \citep{Michalski2021}.  The current methodology updates the approach
+used in prior assessments of domains like Geographic Information Systems
+\citep{SmithEtAl2018_arXivGIS}, Mesh Generators \citep{SmithEtAl2016},
+Oceanographic Software \citep{smith2015state}, Seismology software
+\citep{SmithEtAl2018}, and statistical software for psychology
+\citep{SmithEtAl2018_StatSoft}.  \wss{add citation to medical imaging software,
+if it is an option}
+
+With our methodology, we first choose a software domain (in the current case
+LBM) and identify a list of 24 software packages.  We approximately measure the
+qualities of each package by filling in a grading template. Compared with our
+previous methodology, the new methodology also includes repository based
+metrics, such as the number of files, number of lines of code, percentage of
+issues that are closed, etc.  With the quantitative data in the grading
+template, we rank the software with the Analytic Hierarchy Process (AHP). After
+this, as another addition to our previous methodology, we interview some of the
+development teams to further understand the status of their development process.
+Finally, we summarize the measurements (Section~\ref{AHPresults}), summarize the
+interview answers (Section~\ref{interviewresults}), answer the research
+questions and propose recommendations for LBM software development
+(Section~\ref{answersquestions}).
+
+\section{Methodology} \label{methodology}
+
+We developed a methodology for evaluating the state of the practice of
+scientific software \citep{SmithEtAl2021}.  This methodology can be instantiated
+for a specific domain of scientific software, which in the current case is LBM
+software.  The first section below describes the steps of the process we used to
+select, measure and compare LBM software. This is followed by quality
+definitions and how these qualities were assessed in our project. The rest of
+the section provides the details of the methodology.
+% an overview of the specifics of candidate software selection, the quantitative
+% measurements, developer interviews, the ranking process via AHP and
+% interaction with the Domain Expert.
 
@@
 
- The following steps provide an overview of how the assessment was conducted:
+The assessment was conducted via the following steps:
 
 \begin{enumerate}
-	\item List candidate software packages for the domain. This is discussed in Section \ref{identifysoftware}.
-	\item Filter the software package list. This is discussed in Section \ref{filtersoftware}.
+	\item List candidate software packages for the domain. This is discussed in Section~\ref{identifysoftware}.
+	\item Filter the software package list. This is discussed in Section~\ref{filtersoftware}.
 	\item Gather the source code and documentation for each software package.
-	\item Collect empirical measures. This is discussed in Section \ref{empiricalmeasures}.
-	\item Measure using the measurement template. This is discussed in Section \ref{empiricalmeasures}. The measurement template can be found in Appendix \ref{measurementtemplate2}.
-	\item Survey the developers.
-The developers of each software package in the filtered software list were contacted for voluntary interviews.
-The interview questions can be found in Appendix \ref{interviewquestions}.
-	\item Use AHP to rank the software packages. This is discussed in Section \ref{AHP}.
-	\item Analyze the results and answer the research questions. The answers can be found in Section \ref{answersquestions}.
+	\item Collect quantitative measures from the project repositories. This is discussed in Section~\ref{empiricalmeasures}.
+	\item Measure using the measurement template. This is discussed in Section~\ref{empiricalmeasures}. The full measurement template can be found in \citet{SmithEtAl2021}.
+	\item Survey the developers. The developers of each software package in the
+	filtered software list were contacted for voluntary interviews. The details are given in Section~\ref{SecSurvey}.
+	\item Use AHP to rank the software packages. This is discussed in Section~\ref{AHP}.
+	\item Analyze the results and answer the research questions. The answers can be found in Section~\ref{answersquestions}.
 \end{enumerate}
 
-These steps are further detailed in the \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Methodology.pdf}{Methodology for Assessing the State of the Practice for Domain X} document.
-
-\subsection{Software Qualities}\label{softwarequalities}
-
-Software quality attributes facilitate the measurement and comparison of software packages in this state of the practice assessment.
-We adopt software quality definitions from various researchers and subject matter expert entities.
-Some of the definitions are from \cite{Smithetal2020}.
-The quality measurement results are found in Section \ref{AHPresults} of this report.
-Section \ref{qualityrecommentations} lists recommendations to address software qualities in LBM software packages.
-The following are the software quality definitions used in this state of the practice exercise, along with comments regarding their quantitative and qualitative measurement.
+The above steps depend on interaction with a domain expert partner, as discussed
+in Section~\ref{sec_vet_software_list}.
+
+\subsection{Software Qualities} \label{softwarequalities}
+
+We adopt software quality definitions from various researchers and subject
+matter experts. Some of the definitions are from \cite{Smithetal2020}. The
+following are the software quality definitions used in this exercise, along with
+comments regarding their quantitative and qualitative measurement.  \wss{Update
+when the Quality Definition of Qualities document is completed.}
 
@@
 
-Installability is measured by the effort required for the installation, uninstallation or reinstallation of a software product in a specified
-environment \citep{ISO/IEC25010} \citep{lenhard2013measuring}.
-A good measure of installability correlates with scenarios when low or moderate effort is required to gather and prepare software for its general use on a system for which it was designed.
-In this case effort includes the time spent finding and understanding the installation instructions, the man-time and resources spent performing the installation procedure, and the absence or ease of overcoming system compatibility issues.
-The ability to reasonably validate the installation procedure also has a positive effect on the measure of installability.
-Similarly, the ease of uninstallation has an affect on the measure of installability.
+Installability is measured by the effort required for the installation,
+uninstallation or reinstallation of a software product in a specified
+environment \citep{ISO/IEC25010, lenhard2013measuring}. A good measure of
+installability correlates with scenarios when low or moderate effort is required
+to gather and prepare software for its general use on a system for which it was
+designed. In our case effort includes the time spent finding and understanding
+the installation instructions, the person-time and resources spent performing
+the installation procedure, and the absence or ease of overcoming system
+compatibility issues. The ability to reasonably validate the installation
+procedure and the ease of installation also have a positive effect on the
+measure of installability.
 
@@
 A software program is correct if it behaves according to its stated
-specifications \citep{GhezziEtAl2003}.
-This requires that the specification is available.
-Software is unlikely to have a formal specification if it is not developed by seasoned or professional software developers.
-Since some software does not have a specification available, the correctness of software cannot always be verified.
-Despite an absent specification, the correctness of the output of scientific computing software can sometimes be manually verified by applying domain knowledge.
-A good measure of correctness correlates with the availability of a requirements specification and reference to domain theory, as well as the explicit use of tools or techniques for building confidence of correctness, such as documentation generators and software analysis tools.
+specifications \citep{GhezziEtAl2003}. This requires that the specification is
+available. Scientific software is unlikely to have a formal specification, since
+this is not common practice in the field. As a consequence, the correctness of
+software often cannot be measured directly. Despite an absent specification, the
+correctness can indirectly be assessed by looking at the output and applying
+domain knowledge. For our measurement of correctness we have assumed that the
+following factors suggest correctness was considered: availability of a
+requirements specification, reference to domain theory, and explicit use of
+tools and/or techniques for building confidence of correctness, such as
+documentation generators and software analysis tools.
 
@@
 
-Verifiability is measured by the extent to which a set of tests can be written and executed to demonstrate that the delivered system meets the specification \citep{sommerville}.
-Similarly to correctness, verifiability is correlated with the availability of a specification and with reference to domain knowledge.
-A good measure of verifiability is further correlated with the availability of well written tutorials that include expected output, with software unit testing documentation, and with evidence of continuous integration during the development process.
+Verifiability is measured by the extent to which a set of tests can be written
+and executed to demonstrate that the delivered system meets the specification
+\citep{sommerville}. Similarly to correctness, verifiability is correlated with
+the availability of a specification and with reference to domain knowledge. A
+good measure of verifiability is further correlated with the availability of
+well written tutorials that include expected output, with software unit testing
+documentation, and with evidence of continuous integration during the
+development process. In our process we measure correctness and verifiability
+together.
 
@@
 
-Reliability is measured by the probability of failure-free operation of a computer program in a specified environment for a specified time, i.e.
-the average time interval between two failures also known as the mean time to failure (MTTF) \citep{GhezziEtAl2003} \citep{musa1987software}.
-Reliability is thus positively correlated with the absence of errors during installation and use.
-Recoverability from errors also improves reliability.
+Reliability is measured by the probability of failure-free operation of a
+computer program in a specified environment for a specified time; that is,
+reliability can be measured by the average time interval between two failures,
+also known as the mean time to failure (MTTF) \citep{GhezziEtAl2003}
+\citep{musa1987software}. Reliability is thus positively correlated with the
+absence of errors during installation and use. Recoverability from errors also
+improves reliability.
 
@@
 
-Software possesses the characteristic of robustness if it behaves “reasonably” in two situations: i) when it encounters circumstances not anticipated in the requirements specification; and ii) when the assumptions in its requirements specification are violated \citep{boehm2007software} \citep{ghezzi1991fundamentals}.
-A good measure of robustness correlates with a reasonable reaction to unexpected input, including data of the wrong type, empty input, or missing files or links.
-A reasonable reaction includes an appropriate error message and the ability to recover the system.
+Software possesses the characteristic of robustness if it behaves ``reasonably''
+in two situations: i) when it encounters circumstances not anticipated in the
+requirements specification; and ii) when the assumptions in its requirements
+specification are violated \citep{boehm2007software, ghezzi1991fundamentals}. A
+good measure of robustness correlates with a reasonable reaction to unexpected
+input, including data of the wrong type, empty input, or missing files or links.
+A reasonable reaction includes an appropriate error message and the ability to
+recover the system.
 
@@
 
-Performance is measured by the degree to which a system or component accomplishes its designated functions within given constraints, such as speed (database response times, for instance), throughput (transactions per second), capacity (concurrent usage loads), and timing (hard real-time demands) \citep{IEEEStdGlossarySET1990} \citep{wiegers2003softreq}.
-In this state of the practice assessment performance was not quantitatively measured.
-Instead the documentation of each software package was observed for information that alludes to a consideration of performance, such as parallelization tools.
+Performance is measured by the degree to which a system or component
+accomplishes its designated functions within given constraints, such as speed
+(database response times, for instance), throughput (transactions per second),
+capacity (concurrent usage loads), and timing (hard real-time demands)
+\citep{IEEEStdGlossarySET1990, wiegers2003softreq}. In this state of the
+practice assessment performance was not directly quantitatively measured.
+Instead the documentation of each software package was observed for information
+that alludes to a consideration of performance, such as parallelization tools.
 
@@
 
-Usability is measured by the extent to which a software product can be used by specified users to achieve specified goals with effectiveness, efficiency, and satisfaction in a specified context of use \citep{nielsonusability}.
-A good measure of usability correlates with the presence of documentation, including tutorials, manuals, and defined user characteristics, and user support.
-Preferably the user support model has avenues to contact developers and report issues.
+Usability is measured by the extent to which a software product can be used by
+specified users to achieve specified goals with effectiveness, efficiency, and
+satisfaction in a specified context of use \citep{nielsonusability}. We assumed
+that a high usability correlates with the presence of documentation,
+including tutorials, manuals, and defined user characteristics, and user
+support. Preferably the user support model has avenues to contact developers and
+report issues.
 
@@
 
-A measure of maintainability is the effort with which a software
-system or component can be modified to correct faults, improve performance or other attributes, and satisfy new requirements \citep{IEEEStdGlossarySET1990} \citep{boehm2007software}.
-In this state of the practice analysis maintainability is measured by the quality of documentation artifacts, and the presence of version control and issue tracking.
-These artifacts can greatly decrease the effort needed to modify software.
-There are many documentation artifacts that can improve maintainability, including user and developer manuals, specifications, README files, change logs, release notes, publications, forums, and instructional websites.
+A measure of maintainability is the effort with which a software system or
+component can be modified to correct faults, improve performance or other
+attributes, and satisfy new requirements \citep{IEEEStdGlossarySET1990,
+boehm2007software}. In the current work maintainability is measured by the
+quality of documentation artifacts, and the presence of version control and
+issue tracking. These artifacts can greatly decrease the effort needed to modify
+software. There are many documentation artifacts that can improve
+maintainability, including user and developer manuals, specifications, README
+files, change logs, release notes, publications, forums, and instructional
+websites.
 
@@
 
-Modifiability refers to the ease with which stable changes can be made to a system and the flexibility of the system to adopt such changes \citep{8016712}.
-This state of the practice assessment did not quantitatively measure modifiability.
-Developers were asked in interviews if they considered the ease of future changes when developing the software packages, specifically changes to the structure of the system, modules and code blocks.
-A follow up question asked if any measures had been taken.
+Modifiability refers to the ease with which stable changes can be made to a
+system and the flexibility of the system to adopt such changes \citep{8016712}.
+We did not directly measure modifiability. Instead, developers were asked in
+interviews if they considered the ease of future changes when developing the
+software packages, specifically changes to the structure of the system, modules
+and code blocks. A follow up question is asked if any measures had been taken.
 
@@
 
-Reusability refers to the extent to which components of a software package can be used with or without adaptation in other software packages \citep{kalagiakos2003non}.
-A good measure of reusability results from a large number of easily reusable components.
-Increased software modularization, defined as the presence of smaller components with well defined interfaces, is important.
-For this state of the practice assessment, a good measure of reusablity correlates with an increased number of code files, and the availability of API documentation.
+Reusability refers to the extent to which components of a software package can
+be used with or without adaptation in other software packages
+\citep{kalagiakos2003non}. A good measure of reusability results from a large
+number of easily reusable components. Increased software modularization, defined
+as the presence of smaller components with well defined interfaces, is
+important. For this state of the practice assessment, a good measure of
+reusability correlates with an increased number of code files, and the
+availability of API documentation.
 
@@
 
-Understandability is measured by the capability of the software package to enable the user to understand its suitability and function \citep{ISO9126}.
-It is an artifact-dependent quality.
-Understandability is different for the user-interface, source code, and documentation.
-In this state of the practice analysis, understandability focuses on the source code.
-It is measured by the consistency of a formatting style, the extend of modularization, the explicit identification of coding standards, the presence of meaningful identifiers, and clarity of comments.
+Understandability is measured by the capability of the software package to
+enable the user to understand its suitability and function \citep{ISO9126}. It
+is an artifact-dependent quality. Understandability is different for the
+user-interface, source code, and documentation. In this state of the practice
+analysis, understandability focuses on the source code. It is measured by the
+consistency of a formatting style, the extend of modularization, the explicit
+identification of coding standards, the presence of meaningful identifiers, and
+clarity of comments.
 
@@
 
-Traceability refers to the ability to link the software implementation and the software artifacts, especially the requirement specification \citep{McCallEtAl1977}.
-Similar to the quality of correctness, this requires some form of specification to be available.
-The quality refers to keeping track of information as it changes forms or relates between artifacts.
-This state of the practice assessment did not quantitatively measure traceability.
-Developers were asked in interviews how documentation fits into their development process.
+Traceability refers to the ability to link the software implementation and the
+software artifacts, especially the requirement specification
+\citep{McCallEtAl1977}. Similar to the quality of correctness, this requires the
+availability of some form of specification. This quality refers to keeping track
+of information as it changes forms or relates between artifacts. We did not
+quantitatively measure traceability. Instead, developers were asked in
+interviews how documentation fits into their development process.
 
@@
 
-Visibility and transparency refer to the extent to which all of the steps of a software development process, and the current status of it, are conveyed clearly \citep{ghezzi1991fundamentals}.
-In this state of the practice assessment a good measure of visibility and transparency correlates with a well defined development process, the presence of development process and environment documentation, and software package version release notes.
+Visibility and transparency refer to the extent to which all of the steps of a
+software development process, and the current status of it, are conveyed clearly
+\citep{ghezzi1991fundamentals}. In this state of the practice assessment a good
+measure of visibility and transparency correlates with a well defined
+development process, documentation of the development process and environment
+and software package version release notes.
 
@@
 
-Software achieves reproducibility if another developer can take the requirements documentation and re-obtain the same software artifacts \citep{BenureauAndRougier2017}.
-This includes the output of the software, where the scientific results are compared between software implementations, or between software implementations and manually calculated results.
-This state of the practice assessment did not quantitatively measure reproducibility.
-Developers were asked in interviews if they have any concern that their computational results won't be reproducible in the future, and if they have taken any steps to ensure reproducibility.
+Software achieves reproducibility if another developer can take the requirements
+documentation and re-obtain the same software artifacts
+\citep{BenureauAndRougier2017}. This includes the output of the software, where
+the scientific results are compared between software implementations, or between
+software implementations and manually calculated results. We measured
+reproducibility qualitatively by asking developers if they have any concern that
+their computational results won't be reproducible, and if they have taken any
+steps to ensure reproducibility.
 
@@
 
-Unambiguity refers to the extent to which two readers have similar interpretations when reading software artifacts.
-In other words, artifacts are unambiguous if, and only if, they only have one interpretation \citep{IEEE1998}.
-This state of the practice assessment did not quantitatively measure unambiguity.
-Developers were asked in interviews if they think that the current documentation can clearly convey all necessary knowledge to the users, and how they achieved this or what improvements are needed to achieve it.
-
-\subsection{Identify Candidate Software}\label{identifysoftware}
-
-The candidate software was found through search engine queries targeting authoritative lists of software.
-We found LBM software listed on the websites GitHub and swMATH, as well as through articles found in scholarly journals and databases.
-Software packages that are not on the authoritative lists that we used were not assessed.
-Packages that fit our criteria but were found after data collection and analysis was conducted, such as Musubi, will need to be added to future SOP assessments.
-
-The following properties were considered when creating the list and reviewing the candidate software:
+Unambiguity refers to the extent to which two readers have similar
+interpretations when reading software artifacts. In other words, artifacts are
+unambiguous if, and only if, they only have one interpretation \citep{IEEE1998}.
+This state of the practice assessment did not quantitatively measure
+unambiguity, but we did ask developers if they think that the current
+documentation can clearly convey all necessary knowledge to the users, and how
+they achieved this or what improvements are needed to achieve it.
+
+\subsection{Identify Candidate Software} \label{identifysoftware}
+
+The candidate software was found through search engine queries targeting
+authoritative lists of software. We found LBM software listed on the websites
+GitHub and swMATH, as well as through articles found in scholarly journals and
+databases.  The Domain Expert (Section~\ref{domainexpertrecommend}) was also
+engaged in selecting the candidate software.  \wss{Packages that fit our
+criteria but were found after data collection and analysis was conducted, such
+as Musubi \citep{HasertEtAl2014}, will need to be added to future SOP
+assessments.}
+
+The following properties were considered when creating the list and reviewing
+the candidate software:
 
@@
 	\item The source code must be viewable.
-	\item The empirical measures should be available, which implies a preference for GitHub-style repositories.
-	\item The software cannot be marked as incomplete or in an initial development phase.
+	\item The empirical measures should be available, which implies a preference
+	for GitHub-style repositories.
+	\item The software cannot be marked as incomplete or in an initial
+	development phase.
 \end{enumerate}
 
-The initial list had 45 packages, including a few packages that were later found to not have publicly available source code, or to be in an incomplete state of development.
-
-\subsection{Filter the Software List}\label{filtersoftware}
-
-To reduce the number of members in the candidate software list to a manageable size, the following filters were applied. The filters were applied in the priority order listed.
+The initial list had 45 packages, including a few packages that were later found
+to not have publicly available source code, or to be in an incomplete state of
+development.
+
+\subsection{Filter the Software List} \label{filtersoftware}
+
+To reduce the number of members in the candidate software list to a manageable
+size, the following filters were applied. The filters were applied in the
+priority order listed.
 
 \begin{enumerate}
-	\item Scope: Software is removed by narrowing what functionality is considered to be within the scope of the domain.
-	\item Usage: Software packages were eliminated if their installation procedure was missing or not clear and easy to follow.
-	\item Age: The older software packages (age being measured by the last date when a change was made) were eliminated, except in the cases where an older software package appears to be highly recommended and currently in use.
+	\item Scope: Software is removed by narrowing what functionality is
+	considered to be within the scope of the domain.
+	\item Usage: Software packages were eliminated if their installation
+	procedure was missing or not clear and easy to follow.
+	\item Age: The older software packages (age being measured by the last date
+	when a change was made) were eliminated, except in the cases where an older
+	software package appears to be highly recommended and currently in use. 
 \end{enumerate}
 
-For the third item in the above filter, software packages were characterized as `alive' if their related documentation had been updated within
-the last 18 month. Packages were categorized as `dead' if the last update of this information was more than 18 month ago.\\
-
-\begin{onehalfspacing}
+For the third item in the above filter, software packages were characterized as
+`alive' if their related documentation had been updated within the last 18
+month. Packages were categorized as `dead' if the last update of this
+information was more than 18 month ago.
+
+\begin{table}
 \begin{center}
@@
 	\end{tabular}
-	\captionof{table}{Alive Software Packages}\label{alivepackages}
+	\caption{Alive Software Packages} \label{alivepackages}
 \end{center}
-\end{onehalfspacing}
-
-While the initial list had 45 packages, filtering by scope, usage, and age decreased the size of the list to 23 packages.
-Many of the 22 packages that were removed could not be tested as there was no installation guide, they were incomplete, source code was not publicly available, a license was needed, or the project was out of scope or not up to a standard that would support incorporating them into this study.
-These eliminated software packages are listed in the Appendix in Section \ref{eliminatedpackagessection}.
-Of the remaining 23 packages that were studied, some were kept on the list despite being marked as dead due to their prevalence on authoritative lists on LBM software and due to their surface excellence, specifically the considerable time that was put into these projects.
-
-The final list of software packages that were analyzed in this project can be found in the following two tables.
-Table \ref{alivepackages} lists packages that fell into the `alive' category as of mid 2020, and Table \ref{deadpackages} lists packages that were `dead' at that time.
-\\
-
-\begin{onehalfspacing}
+\end{table}
+
+While the initial list had 46 packages, filtering by scope, usage, and age
+decreased the size of the list to 24 packages. Many of the 22 packages that were
+removed could not be tested as there was no installation guide, they were
+incomplete, source code was not publicly available, a license was needed, or the
+project was out of scope or not up to a standard that would support
+incorporating them into this study. These eliminated software packages are
+listed in the Appendix of \citet{Michalski2021}. Of the remaining 24 packages
+that were studied, some were kept on the list despite being marked as dead due
+to their prevalence on authoritative lists on LBM software and due to their
+surface excellence.  The final list of software packages that were analyzed in
+this project can be found in the following two tables. Table~\ref{alivepackages}
+lists packages that fell into the `alive' category as of mid 2020, and
+Table~\ref{deadpackages} lists packages that were `dead' at that time.
+
+\begin{table}
 \begin{center}
@@
 	\end{tabular}
-	\captionof{table}{Dead Software Packages}\label{deadpackages}
+	\caption{Dead Software Packages} \label{deadpackages}
 \end{center}
-\end{onehalfspacing}
-
-There is considerable variation among these software packages, including their intended purpose, size, user interfaces, and software languages used.
-For example, the OpenLB software package is predominantly a C++ package that makes use of hybrid parallelization and was designed to address a range of CFD problems \citep{heuveline2009towards}.
-The software package pyLBM is an all-in-one Python language package for numerical simulations \citep{graille2017pylbm}.
-ESPResSo is an extensible simulation package that is specifically for research on soft matter, and is written in C++ and Python \citep{weik2019espresso}.
-The HemeLB package is used for efficient simulation of fluid flow in several medical domains, and is written predominantly in C, C++, and Python \citep{mazzeo2008hemelb}.
-
-\subsection{Empirical Measures}\label{empiricalmeasures}
-
-The quality measurements in this assessment rely on the gathering and analyzing of raw and processed empirical data related to the research questions listed in Section \ref{purpose}.
-
-All of the quality measurements that are part of the AHP analysis are empirical measurements.
-Qualitative data gathered during interviews with developers is not part of the AHP analysis.
-This part of the assessment focuses on data that is reasonably easy to collect.
-Much of the data was gathered by manually reviewing the artifacts of each software package, while some was gathered using the freeware tools discussed below.
-The data that was gathered is listed in the measurement template found in Appendix \ref{measurementtemplate2}.
-Some of the data required processing other data within the template, including the status of the software package, which relied on the last commit date; the percentage of issues that are closed, which relied on the number of open and closed issues; and the percentage of code that is comments, which relied on the number of total lines and comment lines in text-based files.
-The complete measurement template data was then analyzed using the grading template found in Appendix \ref{gradingtemplate2}, the output of which was analyzed using the AHP described in Section \ref{AHP}.
-
-Most of the measurement template data was gathered by observing GitHub repository metrics and software package artifacts, or by processing data gathered using freeware tools.
-Data in the final three sets of the measurement template was collected using these tools.
-The tool \href{https://github.com/tomgi/git_stats}{GitStats} was used to measure each software package's GitHub repository for the number of binary files, the number of added and deleted lines, and the number of commits over varying time intervals.
-The tool \href{https://github.com/boyter/scc}{Sloc Cloc and Code (scc)} was used to measure the number of text based files as well as the number of total, code, comment, and blank lines in each GitHub repository.
-Details on installing and running these tools can be found in the \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/A
-	Guide to Empirical Measures.pdf} {Guide to Empirical Measures} file in the AIMSS repository.
-
-\subsection{Analytical Hierarchy Process}\label{AHP}
-
-The Analytical Hierarchy Process (AHP) is a decision-making technique that is used to compare multiple options by multiple criteria.
-In our work AHP was used for comparing and ranking the LBM software packages using the overall impression quality scores that were gathered in the measurement template found in Appendix \ref{measurementtemplate2} using the grading template found in Appendix \ref{gradingtemplate2}.
-AHP performs a pairwise analysis using a matrix and generates an overall score as well as individual quality scores for each software package.
-Smith et al.
-(2016) shows how the AHP is applied to ranking software based on quality measures \cite{SmithEtAl2016}.
-
-This project used a tool for conducting this process.
-The tool includes a sensitivity analysis that was used to ensure that the software package rankings are appropriate with respect to the uncertainty of the quality scores.
-The \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/AHP2020/LBM/README.txt}{README} file of the tool includes requirements and usage information.
-
-\newpage
-
-\section{Quantitative Findings and AHP Results}\label{AHPresults}
-
-This section presents the quantitative findings from data that was gathered using the measurement template found in Section \ref{measurementtemplate2} for qualities listed in Section \ref{softwarequalities}.
-The results of the AHP analysis based on that data are also presented in this section.
+\end{table}
+
+There is considerable variation among these software packages, including their
+intended purpose, size, user interfaces, and software languages used. For
+example, the OpenLB software package is predominantly a \CC{} package that makes
+use of hybrid parallelization and was designed to address a range of CFD
+problems \citep{heuveline2009towards}. The software package pyLBM is an
+all-in-one Python language package for numerical simulations
+\citep{graille2017pylbm}. ESPResSo is an extensible simulation package that is
+specifically for research on soft matter, and is written in \CC{} and Python
+\citep{weik2019espresso}. The HemeLB package is used for efficient simulation of
+fluid flow in several medical domains, and is written predominantly in C, \CC,
+and Python \citep{mazzeo2008hemelb}.
+
+\subsection{Quantitative Measures} \label{empiricalmeasures}
+
+The measurement template is found in \citet{SmithEtAl2021}.  This template is
+used to track measurements and quality scores for all of the software packages
+in the domain. For each software package, we fill-in the rows of the template.
+This process takes about 2 hours per package, with a cap of 4 hours.  The time
+constraint is necessary so that the work load for measuring a given domain is
+feasible for a team as small as one, since we aim for a cap of 160 person hours
+for the measurement phase \citet{SmithEtAl2021}.  An excerpt of the spreadsheet
+is shown in Figure~\ref{measurement_template_image}.  A column is added to this
+template for each of the 24 software package measured.
+
+\begin{figure}[!ht]
+	\begin{center}
+	  \includegraphics[width=1.0\textwidth]{./figures/measurement_template}
+	  \caption{Excerpt of the Top Section of the Measurement Template (Summary
+	  Information) \wss{Should update this figure to something easier to read}}
+	  \label{measurement_template_image}
+	\end{center}
+\end{figure}
+
+The full template consists of 108 questions categorized under 9 qualities:
+\begin{inparaenum}
+	\item installability;
+	\item correctness and verifiability;
+	\item surface reliability;
+	\item surface robustness;
+	\item surface usability;
+	\item maintainability;
+	\item reusability;
+	\item surface understandability; and,
+	\item visibility/transparency. 
+\end{inparaenum} 
+The questions were designed to be unambiguous, quantifiable and measurable with
+limited time and domain knowledge. The measures are grouped under headings for
+each quality, and one for summary information
+(Figure~\ref{measurement_template_image}).   The summary section provides
+general information, such as the software name, number of developers, etc.  We
+follow the definitions given by \citet{GewaltigAndCannon2012} for the software
+categories.  Public means software intended for public use.  Private means
+software aimed only at a specific group, while the concept category is used for
+software written simply to demonstrate algorithms or concepts. The three
+categories of development models are: open source, where source code is freely
+available under an open source license; free-ware, where a binary or executable
+is provided for free; and, commercial, where the user must pay for the software
+product.
+
+Several of the qualities use the word ``surface''.  This is to highlight that,
+for these qualities in particular, the best that we can do is a shallow measure
+of the quality.  For instance, we are not currently doing any experiments to
+measure usability.  Instead, we are looking for an indication that usability was
+considered by the developers.  We do this by looking for cues in the
+documentation, like a getting started manual, a user manual and documentation of
+expected user characteristics.
+
+Most of the data was gathered by manually investigating the software, its source
+code, and its artifacts, while some was gathered using the automatic repository
+measurement tools.  Tools were use to find data such as the number of files,
+number of lines of code (LOC), percentage of issues that are closed, etc. The
+tool \href{https://github.com/tomgi/git_stats}{GitStats} was used to measure
+each software package's GitHub repository for the number of binary files, the
+number of added and deleted lines, and the number of commits over varying time
+intervals. The tool \href{https://github.com/boyter/scc}{Sloc Cloc and Code
+(scc)} was used to measure the number of text based files as well as the number
+of total, code, comment, and blank lines in each GitHub repository.
+
+\section{Survey Developers} \label{SecSurvey}
+
+We designed a list of 20 questions to guide our interviews, which can be found
+in \citet{SmithEtAl2021}. Some questions are about the background of the
+software, the development teams, the interviewees, and how they organize the
+projects. We also ask about the developer's understandings of the users. Some
+questions focus on the current and past difficulties, and the solutions the team
+has found, or will try. We also discuss the importance and current situations of
+documentation. A few questions are about specific software qualities, such as
+maintainability, understandability, usability, and reproducibility. The
+interviews are semi-structured based on the question list; we ask follow-up
+questions when necessary. Based on our experience, the interviewees usually
+bring up some exciting ideas that we did not expect, and it is worth expanding
+on these topics.
+
+We requested interviews from all 24 packages, with 4 positive responses.  To
+send out interview requests we found we found contacts on the projects’ website,
+or code repository, or publications, or the biographic pages of the teams’
+institutions. We send at most two interview request emails to a contact for each
+software package.  Meeting will typically be held using on-line meeting
+software, like Zoom or Teams.  This facilitates recording and automatic
+transcription of the meetings.
+
+The interviews followed the standard ethics guideline of asking for consent
+before interviewing, recording, and including participant details in the report.
+The interview process presented here was approved by the McMaster University
+Research Ethics Board under the application number 
+\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/MACREM/Application.pdf}
+{MREB\#: 5219}.
+
+\subsection{Analytical Hierarchy Process} \label{AHP}
+
+The Analytical Hierarchy Process (AHP) is a decision-making technique that is
+used to compare multiple options by multiple criteria. In our work AHP was used
+for comparing and ranking the LBM software packages using the overall impression
+quality scores that were gathered in the measurement template found in Appendix
+~\ref{measurementtemplate2} using the grading template found in Appendix
+~\ref{gradingtemplate2}. AHP performs a pairwise analysis using a matrix and
+generates an overall score as well as individual quality scores for each
+software package. Smith et al. (2016) shows how the AHP is applied to ranking
+software based on quality measures \cite{SmithEtAl2016}.
+
+This project used a tool for conducting this process. The tool includes a
+sensitivity analysis that was used to ensure that the software package rankings
+are appropriate with respect to the uncertainty of the quality scores. The
+\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/AHP2020/LBM/README.txt}{README}
+file of the tool includes requirements and usage information.
+
+\subsection{Interaction With Domain Expert} \label{sec_vet_software_list}
+
+The Domain Expert is an important member of the state of the practice assessment
+team. Pitfalls exist if non-experts attempt to acquire an authoritative list of
+software, or try to definitively rank the software. Non-experts have the problem
+that they can only rely on information available on-line, which has the
+following drawbacks:
+\begin{inparaenum}[i)]
+  \item the on-line resources could have false or inaccurate information; and,
+  \item the on-line resources could leave out relevant information that is so
+in-grained with experts that nobody thinks to explicitly record it.
+\end{inparaenum}
+
+Domain experts may be recruited from academia or industry.  The only
+requirements are knowledge of the domain and a willingness to be engaged in the
+assessment process.  The Domain Expert does not have to be a software developer,
+but they should be a user of domain software.  Given that the domain experts are
+likely to be busy people, the measurement process cannot put to much of a burden
+on their time.  For the current assessment, our Domain Expert (and paper
+co-author) is Dr.\ Michael Noseworthy, Professor of Electrical and Computer
+Engineering at McMaster University, Co-Director of the McMaster School of
+Biomedical Engineering, and Director of Medical Imaging Physics and Engineering
+at St. Joseph’s Healthcare, Hamilton, Ontario, Canada.
+
+In advance of the first meeting with the Domain Expert, the expert is asked to
+create a list of top software packages in the domain.  This is done to help
+the expert get in the right mind set in advance of the meeting.  Moreover,
+by doing the exercise in advance, we avoid the potential pitfall of the expert
+approving the discovered list of software without giving it adequate thought.
+
+The Domain Experts are asked to vet the collected data and analysis.  In
+particular, they are asked to vet the proposed list of software packages and the
+AHP ranking.  These interactions can be done either electronically or with
+in-person (or virtual) meetings.
+
+\section{Quantitative Findings and AHP Results} \label{AHPresults}
+
+This section presents the quantitative findings from data that was gathered
+using the measurement template found in Section~\ref{measurementtemplate2} for
+qualities listed in Section~\ref{softwarequalities}. The results of the AHP
+analysis based on that data are also presented in this section.
+
+Table that summarizes the software by variabilities.
+
+\begin{table}
+    \begin{center}
+        \begin{tabular}{ p{3cm}lllllll}
+            \hline
+            Name & Dim & Pll & Com & Rflx & Mfl & NN & OS\\
+            \hline
+            DL\_MESO (LBE) & ? & ? & ? & ? & ? & ? & ?\\
+            ESPResSo & ? & ? & ? & ? & ? & ? & ?\\
+            ESPResSo++ & ? & ? & ? & ? & ? & ? & ?\\
+            lbmpy & ? & ? & ? & ? & ? & ? & ?\\
+            lettuce & ? & ? & ? & ? & ? & ? & ?\\
+            Ludwig & ? & ? & ? & ? & ? & ? & ?\\
+            LUMA & ? & ? & ? & ? & ? & ? & ?\\
+            MechSys & ? & ? & ? & ? & ? & ? & ?\\
+            OpenLB & ? & ? & ? & ? & ? & ? & ?\\
+            Palabos & ? & ? & ? & ? & ? & ? & ?\\
+            pyLBM & ? & ? & ? & ? & ? & ? & ?\\
+            Sailfish & ? & ? & ? & ? & ? & ? & ?\\
+            TCLB & ? & ? & ? & ? & ? & ? & ?\\
+            waLBerla & ? & ? & ? & ? & ? & ? & ?\\
+            HemeLB & ? & ? & ? & ? & ? & ? & ?\\
+            laboetie & ? & ? & ? & ? & ? & ? & ?\\
+            LatBo.jl & ? & ? & ? & ? & ? & ? & ?\\
+            LB2D-Prime & ? & ? & ? & ? & ? & ? & ?\\
+            LB3D & ? & ? & ? & ? & ? & ? & ?\\
+            LB3D-Prime & ? & ? & ? & ? & ? & ? & ?\\
+            LIMBES & ? & ? & ? & ? & ? & ? & ?\\
+            MP-LABS & ? & ? & ? & ? & ? & ? & ?\\
+            SunlightLB & ? & ? & ? & ? & ? & ? & ?\\
+
+            \hline
+        \end{tabular}
+        \caption{Features of Software Packages (Dim for Dimension (1, 2 or 3),
+        Pll for Parallel (Yes or No), Com for Compressible (Yes or No), Rflx for
+        Reflexive boundary condition (Yes or No), MFl for Multifluid (Yes or
+        No), NN for Non-Newtonian, OS (Win, Mac, Lin, )} \label{tbl_features}
+    \end{center}
+    \end{table}
 
@@
 All of the 23 software packages that were tested have installation instructions.
-As noted previously, many of the 23 software packages that were part of the original long list of 45 packages were removed due to not including documentation or installation instructions.
+As noted previously, many of the 23 software packages that were part of the
+original long list of 45 packages were removed due to not including
+documentation or installation instructions.
 
 All 23 packages on the short list can be installed on some Unix-like systems.
-Seven packages could be installed on Windows, and five on macOS.
-Operating system compatibility is found in the documentation of 19 software packages.
-All but one of the software packages, TCLB, were tested on Ubuntu for this state of the practice assessment.
-TCLB was tested on CentOS, since this operating system is mentioned in its installation instructions.
-
-Of the software packages that were tested, most installation instructions are located in one place, often in an instruction manual or on a web-page.
-Sometimes, like with Ludwig, incomplete installation instructions are found on a home page, with more detailed instructions located on another web-page, or within the documentation.
-Maintainability and correctness of these instructions could be improved if all the instructions were in one location.
-
-All but one of the software packages (LatBo.jl) have automated at least some of the installation process.
-Most of these packages, such as waLBerla and SunlightLB, use Make to automate the installation, and a few of them, like lbmpy, use custom scripts.
-
-Errors encountered during the installation process were often quickly fixed thanks to descriptive error messages.
-Systems that provided vague error messages, such as messages that did not specify which action or file was at fault, were more difficult to troubleshoot.
-Only three software packages (HemeLB, LB3D, lbmpy) that displayed a descriptive error message were not recoverable, and most of these instances were due to hardware and operating system incompatibility, such as the requirement of CUDA.
-Fourteen software packages definitively broke during installation.
-Some packages, such as LB2D-Prime and LB3D-Prime, did not provide a definitive message of the success or failure of installation.
-In these instances, validating the installation required performing a tutorial or running a script, as described below, if these were available.
-
-About half of the installation instructions are written as if the person doing the installation has none of the dependent packages installed.
-It is common for software packages to not list all of their dependencies despite listing some.
+Seven packages could be installed on Windows, and five on macOS. Operating
+system compatibility is found in the documentation of 19 software packages. All
+but one of the software packages, TCLB, were tested on Ubuntu for this state of
+the practice assessment. TCLB was tested on CentOS, since this operating system
+is mentioned in its installation instructions.
+
+Of the software packages that were tested, most installation instructions are
+located in one place, often in an instruction manual or on a web-page.
+Sometimes, like with Ludwig, incomplete installation instructions are found on a
+home page, with more detailed instructions located on another web-page, or
+within the documentation. Maintainability and correctness of these instructions
+could be improved if all the instructions were in one location.
+
+All but one of the software packages (LatBo.jl) have automated at least some of
+the installation process. Most of these packages, such as waLBerla and
+SunlightLB, use Make to automate the installation, and a few of them, like
+lbmpy, use custom scripts.
+
+Errors encountered during the installation process were often quickly fixed
+thanks to descriptive error messages. Systems that provided vague error
+messages, such as messages that did not specify which action or file was at
+fault, were more difficult to troubleshoot. Only three software packages
+(HemeLB, LB3D, lbmpy) that displayed a descriptive error message were not
+recoverable, and most of these instances were due to hardware and operating
+system incompatibility, such as the requirement of CUDA. Fourteen software
+packages definitively broke during installation. Some packages, such as
+LB2D-Prime and LB3D-Prime, did not provide a definitive message of the success
+or failure of installation. In these instances, validating the installation
+required performing a tutorial or running a script, as described below, if these
+were available.
+
+About half of the installation instructions are written as if the person doing
+the installation has none of the dependent packages installed. It is common for
+software packages to not list all of their dependencies despite listing some.
 This was the case for many packages, including ESPResSo++, Ludwig, and LUMA.
-Sometimes only an error message during the installation process informs the user of the requirement of these additional packages.
-A detailed rewrite of the installation instructions from the point of view of installation on a clean operating system is suggested.
-A clean environment can be achieved for testing purposes by using a virtual machine.
-
-Sixteen software packages require less than 10 dependencies to be installed.
-All but one software package (LatBo.jl) require less than 20 dependencies.
-Some packages may automatically install additional dependencies in the background.
-Eighteen of the software packages do not explicitly indicate software dependency versions.
-Some software package installation issues, specifically those occurring when manual installation of dependencies is required, may be avoided if versions of dependencies are specified.
-Fifteen software packages do not have detailed instructions for installing dependencies.
-
-Sixteen software packages have less than 10 manual installation steps.
-If dependencies are installed in one command then none of the software packages take more than 20 steps to install.
-The average number of steps is about eight, and the fewest is two (LB3D-Prime).
-
-All but six (ESPResSo, HemeLB, laboetie, LB3D-Prime, lbmpy, waLBerla) of the software packages have a way to specifically verify the installation.
-Most have some sort of tutorial examples that can be run by the user.
-Some other ways of installation validation include validation scripts (LB2D-Prime, lettuce, Ludwig, LUMA), automatic validation after the installation (LatBo.jl), and instructions to manually review the file system (LIMBES).
-
-Uninstallation instructions were found for only one of the software packages, pyLBM.
+Sometimes only an error message during the installation process informs the user
+of the requirement of these additional packages. A detailed rewrite of the
+installation instructions from the point of view of installation on a clean
+operating system is suggested. A clean environment can be achieved for testing
+purposes by using a virtual machine.
+
+Sixteen software packages require less than 10 dependencies to be installed. All
+but one software package (LatBo.jl) require less than 20 dependencies. Some
+packages may automatically install additional dependencies in the background.
+Eighteen of the software packages do not explicitly indicate software dependency
+versions. Some software package installation issues, specifically those
+occurring when manual installation of dependencies is required, may be avoided
+if versions of dependencies are specified. Fifteen software packages do not have
+detailed instructions for installing dependencies.
+
+Sixteen software packages have less than 10 manual installation steps. If
+dependencies are installed in one command then none of the software packages
+take more than 20 steps to install. The average number of steps is about eight,
+and the fewest is two (LB3D-Prime).
+
+All but six (ESPResSo, HemeLB, laboetie, LB3D-Prime, lbmpy, waLBerla) of the
+software packages have a way to specifically verify the installation. Most have
+some sort of tutorial examples that can be run by the user. Some other ways of
+installation validation include validation scripts (LB2D-Prime, lettuce, Ludwig,
+LUMA), automatic validation after the installation (LatBo.jl), and instructions
+to manually review the file system (LIMBES).
+
+Uninstallation instructions were found for only one of the software packages,
+pyLBM.
 
@@
 	\begin{center}
-		\includegraphics[width=1.0\textwidth]{installability_chart}
+		\includegraphics[width=1.0\textwidth]{./figures/installability_chart}
 		\caption{AHP Installability Score}
@@
 
-Figure \ref{Fig_Installability} shows the installability ranking of the software packages using AHP.
-Software packages with a higher score (MechSys, Palabos, ESPResSo, OpenLB) tend to have one set of linear installation instructions that are written as if the person doing the installation has none of the dependencies installed.
-The instructions often list compatible operating system versions and include instructions for the installation of dependencies.
-The top ranked packages often incorporate some sort of automation of the installation process and have fewer manual installation steps.
-The number of dependencies a package has does not correlate with a higher score.
-The ability to validate the installation process, often through tutorials or test examples that include expected output, is correlated with a higher score.
-Furthermore, the top six ranked packages are noted as being alive.
-
-Many software packages would benefit from a rewrite or reorganization of installation instructions.
-A single location for installation instructions would improve their maintainability and correctness.
-Listing compatible operating system and dependency versions would decrease installation time and errors, as would adding instructions on installing dependencies.
-Installation process errors should prompt the system to display detailed messages.
-Once a software package is installed, either an automatic validation needs to be performed or the user needs to be able to perform a manual validation using test examples that include expected output.
-Finally, uninstallation instructions should be included in the documentation.
+Figure~\ref{Fig_Installability} shows the installability ranking of the software
+packages using AHP. Software packages with a higher score (MechSys, Palabos,
+ESPResSo, OpenLB) tend to have one set of linear installation instructions that
+are written as if the person doing the installation has none of the dependencies
+installed. The instructions often list compatible operating system versions and
+include instructions for the installation of dependencies. The top ranked
+packages often incorporate some sort of automation of the installation process
+and have fewer manual installation steps. The number of dependencies a package
+has does not correlate with a higher score. The ability to validate the
+installation process, often through tutorials or test examples that include
+expected output, is correlated with a higher score. Furthermore, the top six
+ranked packages are noted as being alive.
+
+Many software packages would benefit from a rewrite or reorganization of
+installation instructions. A single location for installation instructions would
+improve their maintainability and correctness. Listing compatible operating
+system and dependency versions would decrease installation time and errors, as
+would adding instructions on installing dependencies. Installation process
+errors should prompt the system to display detailed messages. Once a software
+package is installed, either an automatic validation needs to be performed or
+the user needs to be able to perform a manual validation using test examples
+that include expected output. Finally, uninstallation instructions should be
+included in the documentation.
 
@@
 
-Sixteen of the software packages include a requirements specification artifact or explicitly reference domain theory, often only the latter.
-Software packages that distribute requirements specification information, such as DL\_MESO (LBE), generally keep it brief and include it within other documentation.
-This artifact is often found within a user manual, on a web-page, or is mentioned in related publications.
-In the latter case the user may need to spend significant time to find this information.
-
-Document generation tools are explicitly used by 12 software packages. Sphinx is used by eight of them, and Doxygen is used by seven. Several of the packages use both.
-
-Tutorials are available for 18 of the software packages.
-Generally they are linearly written and easy to follow.
-However, only eight tutorials provide an expected output.
-It is not possible to verify the correctness of the output of the software packages that are missing this key information.
-In these cases the user may need to assume correctness if there are no visible errors.
-
-Unit tests are only explicitly available for one of the software packages, Ludwig.
-Code modularization of most packages allow for users to create tests with varying degrees of effort.
-These tests allow developers and users to verify the correctness of fragments of the source code, and in doing so better assess the correctness of the entire package.
-
-The use of continuous integration tools and techniques alludes to a more refined development process where faults are isolated and better recognized.
-Only two of the packages (ESPResSo, Ludwig) mentioned applying the practice of continuous integration in their development process.
+Sixteen of the software packages include a requirements specification artifact
+or explicitly reference domain theory, often only the latter. Software packages
+that distribute requirements specification information, such as DL\_MESO (LBE),
+generally keep it brief and include it within other documentation. This artifact
+is often found within a user manual, on a web-page, or is mentioned in related
+publications. In the latter case the user may need to spend significant time to
+find this information.
+
+Document generation tools are explicitly used by 12 software packages. Sphinx is
+used by eight of them, and Doxygen is used by seven. Several of the packages use
+both.
+
+Tutorials are available for 18 of the software packages. Generally they are
+linearly written and easy to follow. However, only eight tutorials provide an
+expected output. It is not possible to verify the correctness of the output of
+the software packages that are missing this key information. In these cases the
+user may need to assume correctness if there are no visible errors.
+
+Unit tests are only explicitly available for one of the software packages,
+Ludwig. Code modularization of most packages allow for users to create tests
+with varying degrees of effort. These tests allow developers and users to verify
+the correctness of fragments of the source code, and in doing so better assess
+the correctness of the entire package.
+
+The use of continuous integration tools and techniques alludes to a more refined
+development process where faults are isolated and better recognized. Only two of
+the packages (ESPResSo, Ludwig) mentioned applying the practice of continuous
+integration in their development process.
 
@@
 	\begin{center}
-		\includegraphics[width=1.0\textwidth]{correctnessverifiability_chart}
+		\includegraphics[width=1.0\textwidth]{./figures/correctnessverifiability_chart}
 		\caption{AHP Surface Correctness and Verifiability Score}
@@
 
-Figure \ref{Fig_CorrectnessVerifiability} shows the surface correctness and verifiability ranking of the software packages using AHP.
-Software packages with a higher score tend to have a visible requirements specification or references to theory documentation.
-They also explicitly use at least one document generation tool that builds confidence of correctness.
-The top ranked software packages all include an easy to follow getting started tutorial, and most of these include expected output.
-Only the top ranked package, Ludwig, provided unit testing.
-It and the second ranked package, ESPResSo, explicitly incorporated continuous integration in the development process.
-Furthermore, eight of the top 10 ranked packages are noted as being alive.
-
-The inclusion of requirements specification and theory documentation greatly benefits the correctness and verifiability of software packages.
-The use of document generation tools can help build confidence in correctness.
-The addition of easy to follow tutorials further helps users verify the software and have confidence in its correctness.
-Unit testing documentation and capability, as well as the use of continuous integration tools and techniques such as Bamboo, Jenkins, and Travis CI, help verify correctness.
+Figure~\ref{Fig_CorrectnessVerifiability} shows the surface correctness and
+verifiability ranking of the software packages using AHP. Software packages with
+a higher score tend to have a visible requirements specification or references
+to theory documentation. They also explicitly use at least one document
+generation tool that builds confidence of correctness. The top ranked software
+packages all include an easy to follow getting started tutorial, and most of
+these include expected output. Only the top ranked package, Ludwig, provided
+unit testing. It and the second ranked package, ESPResSo, explicitly
+incorporated continuous integration in the development process. Furthermore,
+eight of the top 10 ranked packages are noted as being alive.
+
+The inclusion of requirements specification and theory documentation greatly
+benefits the correctness and verifiability of software packages. The use of
+document generation tools can help build confidence in correctness. The addition
+of easy to follow tutorials further helps users verify the software and have
+confidence in its correctness. Unit testing documentation and capability, as
+well as the use of continuous integration tools and techniques such as Bamboo,
+Jenkins, and Travis CI, help verify correctness.
 
@@
 
-The analysis of surface reliability focused on package installation and tutorials.
-Errors occurred when installing 16 of the software packages.
-Every instance prompted an error message.
-These messages indicated unrecognized commands (even when following the installation guide), missing links, missing dependencies, syntax errors in code files.
-In some instances the error messages were vague.
-Several automatic installation processes could not find and load dependencies.
-In these instances the installation tried to access outdated external repositories.
-Seven of the installations were recovered and verified, and one of the installations (LB3D-Prime) was assumed to be recovered due to the absence of any way to verify it.
-The installation of eight of the software packages could not be recovered.
-Most of these broken installations could not find external dependencies, encountered system incompatibilities, or displayed vague error messages.
-
-Of the 13 software packages that installed correctly and also have tutorials, four (pyLBM, ESPResSo++, LIMBES, Ludwig) broke during tutorial testing.
-All of these instances resulted in an error message being displayed.
-One error (pyLBM) was due to a missing tutorial dependency, another (Ludwig) was due to an invalid command despite following the tutorial, and the final two errors were vague execution errors.
-Of the four broken tutorial instances, only the one that was missing a dependency was recoverable.
+The analysis of surface reliability focused on package installation and
+tutorials. Errors occurred when installing 16 of the software packages. Every
+instance prompted an error message. These messages indicated unrecognized
+commands (even when following the installation guide), missing links, missing
+dependencies, syntax errors in code files. In some instances the error messages
+were vague. Several automatic installation processes could not find and load
+dependencies. In these instances the installation tried to access outdated
+external repositories. Seven of the installations were recovered and verified,
+and one of the installations (LB3D-Prime) was assumed to be recovered due to the
+absence of any way to verify it. The installation of eight of the software
+packages could not be recovered. Most of these broken installations could not
+find external dependencies, encountered system incompatibilities, or displayed
+vague error messages.
+
+Of the 13 software packages that installed correctly and also have tutorials,
+four (pyLBM, ESPResSo++, LIMBES, Ludwig) broke during tutorial testing. All of
+these instances resulted in an error message being displayed. One error (pyLBM)
+was due to a missing tutorial dependency, another (Ludwig) was due to an invalid
+command despite following the tutorial, and the final two errors were vague
+execution errors. Of the four broken tutorial instances, only the one that was
+missing a dependency was recoverable.
 
@@
 	\begin{center}
-		\includegraphics[width=1.0\textwidth]{reliability_chart}
+		\includegraphics[width=1.0\textwidth]{./figures/reliability_chart}
 		\caption{AHP Surface Reliability Score}
@@
 
-Figure \ref{Fig_Reliability} shows the surface reliability ranking of the software packages using AHP.
-Software packages with a high score either did not break during installation, or the broken installation was recoverable.
-All of the top five ranked packages have tutorials.
-One of these packages, pyLBM, broke during tutorial testing, but a descriptive error message helped in recovery.
+Figure~\ref{Fig_Reliability} shows the surface reliability ranking of the
+software packages using AHP. Software packages with a high score either did not
+break during installation, or the broken installation was recoverable. All of
+the top five ranked packages have tutorials. One of these packages, pyLBM, broke
+during tutorial testing, but a descriptive error message helped in recovery.
 Furthermore, nine of the top 10 ranked packages are noted as being alive.
 
-Overall, lower ranked software packages are lacking clear documentation, testing or tutorial examples, and descriptive error messages, and have broken dependencies.
-Thus, regarding surface reliability, software packages would benefit from clear up-to-date documentation that specifies all dependencies, the inclusion of testing and tutorial examples, and the assurance of descriptive error messages during fault conditions.
+Overall, lower ranked software packages are lacking clear documentation, testing
+or tutorial examples, and descriptive error messages, and have broken
+dependencies. Thus, regarding surface reliability, software packages would
+benefit from clear up-to-date documentation that specifies all dependencies, the
+inclusion of testing and tutorial examples, and the assurance of descriptive
+error messages during fault conditions.
 
@@
 
-The software packages were tested for handling unexpected input, including incorrect data types, empty input, and missing files or links.
-Success predicated on a reasonable response from the system, including appropriate error messages and an absence of unrecoverable system failures.
+The software packages were tested for handling unexpected input, including
+incorrect data types, empty input, and missing files or links. Success
+predicated on a reasonable response from the system, including appropriate error
+messages and an absence of unrecoverable system failures.
 
@@
 	\begin{center}
-		\includegraphics[width=1.0\textwidth]{robustness_chart}
+		\includegraphics[width=1.0\textwidth]{./figures/robustness_chart}
 		\caption{AHP Surface Robustness Score}
@@
 
-Figure \ref{Fig_Robustness} shows the surface robustness ranking of the software packages using AHP.
-Software packages with a high score behaved reasonably in response to unexpected input as described above.
-All of the software packages that installed correctly passed this test.
-They output descriptive error messages or did not crash.
-Software packages with a lower surface robustness score had not installed correctly, so their robustness score may not be a true reflection of runtime robustness.
-Similarly, all software packages that installed correctly and require plain text input files correctly handled an unexpected change to these input files, including a replacement of new lines with carriage returns.
-Furthermore, nine of the top 10 ranked packages are noted as being alive.
-LIMBES is noted as being dead.
+Figure~\ref{Fig_Robustness} shows the surface robustness ranking of the software
+packages using AHP. Software packages with a high score behaved reasonably in
+response to unexpected input as described above. All of the software packages
+that installed correctly passed this test. They output descriptive error
+messages or did not crash. Software packages with a lower surface robustness
+score had not installed correctly, so their robustness score may not be a true
+reflection of runtime robustness. Similarly, all software packages that
+installed correctly and require plain text input files correctly handled an
+unexpected change to these input files, including a replacement of new lines
+with carriage returns. Furthermore, nine of the top 10 ranked packages are noted
+as being alive. LIMBES is noted as being dead.
 
@@
 
-Although the software packages all apply LBM to solve scientific computing problems, the packages focus on varied CFD problems, with varying parameters, and are technically different from each other.
-Due to this, a comparison of performance is not appropriate.
-In this project we instead looked through each software package's artifacts for evidence that performance was considered.
-The artifacts of 17 software packages mentioned parallelization.
-This included GPU processing and the CUDA parallel computing platform, which were mentioned in the artifacts of 6 packages (ESPResSo, lbmpy, lettuce, pyLBM, Sailfish, TCLB).
-GPUs provide superior processing power and speed compared to CPUs, and are often used for scientific computing when a large amount of data is involved.
-The software package TCLB is implemented in a highly efficient multi-GPU code to achieve performance suitable for model optimization \citep{rutkowski2020open}.
-In the Ludwig package, a so-called mixed mode approach is used where fine-grained parallelism is implemented on the GPU, and MPI is used for even larger scale parallelism \citep{gray2013ludwig}.
-While one software package (Sailfish) required CUDA and GPU processing, some (ESPResSo, lbmpy, lettuce, pyLBM, TCLB) have the option of using either the GPU or the CPU.
-The packages that require GPU and CUDA have better performance at the expense of installability and surface reliability.
+Although the software packages all apply LBM to solve scientific computing
+problems, the packages focus on varied CFD problems, with varying parameters,
+and are technically different from each other. Due to this, a comparison of
+performance is not appropriate. In this project we instead looked through each
+software package's artifacts for evidence that performance was considered. The
+artifacts of 17 software packages mentioned parallelization. This included GPU
+processing and the CUDA parallel computing platform, which were mentioned in the
+artifacts of 6 packages (ESPResSo, lbmpy, lettuce, pyLBM, Sailfish, TCLB). GPUs
+provide superior processing power and speed compared to CPUs, and are often used
+for scientific computing when a large amount of data is involved. The software
+package TCLB is implemented in a highly efficient multi-GPU code to achieve
+performance suitable for model optimization \citep{rutkowski2020open}. In the
+Ludwig package, a so-called mixed mode approach is used where fine-grained
+parallelism is implemented on the GPU, and MPI is used for even larger scale
+parallelism \citep{gray2013ludwig}. While one software package (Sailfish)
+required CUDA and GPU processing, some (ESPResSo, lbmpy, lettuce, pyLBM, TCLB)
+have the option of using either the GPU or the CPU. The packages that require
+GPU and CUDA have better performance at the expense of installability and
+surface reliability.
 
@@
 
-Software package artifacts were reviewed for the presence of a tutorial, a user manual, documented user characteristics, and a user support model.
-In total 18 software packages have a tutorial, 13 have a user manual, and 11 have both.
-The tutorials vary in scope and substance, and eight include an expected output.
-Most user manuals are in the form of a file that can be downloaded, while some are rendered on a web-page.
-Some packages (waLBerla) do not have a user manual, but do have useful documentation distributed on their web-pages.
-Expected user characteristics are documented in four software packages (laboetie, LIMBES, Ludwig, Palabos).
-Users are typically scientists or engineers.
-Their background is often physics, chemistry, biophysics, or mathematics.
-All but one of the packages (LIMBES) have a user support model, and many of them have multiple avenues of user support.
-The most popular avenue of support is Git, followed by email and forums.
-One software package (OpenLB) has an FAQ page.
+Software package artifacts were reviewed for the presence of a tutorial, a user
+manual, documented user characteristics, and a user support model. In total 18
+software packages have a tutorial, 13 have a user manual, and 11 have both. The
+tutorials vary in scope and substance, and eight include an expected output.
+Most user manuals are in the form of a file that can be downloaded, while some
+are rendered on a web-page. Some packages (waLBerla) do not have a user manual,
+but do have useful documentation distributed on their web-pages. Expected user
+characteristics are documented in four software packages (laboetie, LIMBES,
+Ludwig, Palabos). Users are typically scientists or engineers. Their background
+is often physics, chemistry, biophysics, or mathematics. All but one of the
+packages (LIMBES) have a user support model, and many of them have multiple
+avenues of user support. The most popular avenue of support is Git, followed by
+email and forums. One software package (OpenLB) has an FAQ page.
 
@@
 	\begin{center}
-		\includegraphics[width=1.0\textwidth]{usability_chart}
+		\includegraphics[width=1.0\textwidth]{./figures/usability_chart}
 		\caption{AHP Surface Usability Score}
@@
 
-Figure \ref{Fig_Usability} shows the surface usability ranking of the software packages using AHP.
-Software packages with a high score have a tutorial and user manual, sometimes have documented user characteristics, and have at least one user support model.
-Many packages have several user support models.
-Furthermore, four of the top five ranked packages are noted as being alive.
+Figure~\ref{Fig_Usability} shows the surface usability ranking of the software
+packages using AHP. Software packages with a high score have a tutorial and user
+manual, sometimes have documented user characteristics, and have at least one
+user support model. Many packages have several user support models. Furthermore,
+four of the top five ranked packages are noted as being alive.
 
@@
 
-Software packages were reviewed for the presence of artifacts.
-Every type of artifact or file that is not a code file was recorded.
-The software packages were also reviewed for software release and documentation version numbers.
-This information could be used to better troubleshoot issues and organize documentation.
-All but three software packages (LatBo.jl, LB3D-Prime, MechSys) have source code release and documentation version numbers.
-
-Information on how code is reviewed, or how to contribute to the project was also noted.
-In total, 11 software packages have this information, which was found in various artifacts, including in developer guides, contributor guides, user guides, developer web-pages, and README files.
-
-Issue tracking is used in 22 software packages, 15 of which use Git, six use email, and one (SunlightLB) uses SourceForge.
-Most software packages that use Git have most of their issues closed, and only three (laboetie, lettuce, Sailfish) have less than 50 percent of their issues closed.
-Four of the top five overall ranked packages (Ludwig, ESPResSo, Palabos, LUMA) have most of their issues closed.
-Fourth ranked LUMA does not use Git.
-Alive packages (11 use Git issue tracking) have 64\% of their issues closed, while dead packages (3 use Git issue tracking) have 71\% of their issues closed.
-This information is presented in Table \ref{gitrepodata}.
-Furthermore, 13 packages that use Git for issue tracking use GitHub as a version control system, while two (Palabos, waLBerla) use GitLab.
-Of the other packages, one package (SunlightLB) uses CVS for issue tracking, and seven packages do not appear to use any issue tracking system.
-
-Software package code files were further measured for the percentage of code that is comments.
-The findings are presented in Table \ref{gitrepodata}.
-Packages with a higher percentage of comments were designated as more maintainable.
-Comments represent more than 10 percent of code files in 15 packages, and the average percentage of code comments is about 14 percent.
-Four of the top five overall ranked packages (Ludwig, ESPResSo, Palabos, OpenLB) have more than the average.
-Fifth ranked LUMA has only 0.2 percent comments, the fewest of any package.
-The package has the most lines of source code, with over four million.
-The next largest package is ESPResSo++ with one million.
+Software packages were reviewed for the presence of artifacts. Every type of
+artifact or file that is not a code file was recorded. The software packages
+were also reviewed for software release and documentation version numbers. This
+information could be used to better troubleshoot issues and organize
+documentation. All but three software packages (LatBo.jl, LB3D-Prime, MechSys)
+have source code release and documentation version numbers.
+
+Information on how code is reviewed, or how to contribute to the project was
+also noted. In total, 11 software packages have this information, which was
+found in various artifacts, including in developer guides, contributor guides,
+user guides, developer web-pages, and README files.
+
+Issue tracking is used in 22 software packages, 15 of which use Git, six use
+email, and one (SunlightLB) uses SourceForge. Most software packages that use
+Git have most of their issues closed, and only three (laboetie, lettuce,
+Sailfish) have less than 50 percent of their issues closed. Four of the top five
+overall ranked packages (Ludwig, ESPResSo, Palabos, LUMA) have most of their
+issues closed. Fourth ranked LUMA does not use Git. Alive packages (11 use Git
+issue tracking) have 64\% of their issues closed, while dead packages (3 use Git
+issue tracking) have 71\% of their issues closed. This information is presented
+in Table~\ref{gitrepodata}. Furthermore, 13 packages that use Git for issue
+tracking use GitHub as a version control system, while two (Palabos, waLBerla)
+use GitLab. Of the other packages, one package (SunlightLB) uses CVS for issue
+tracking, and seven packages do not appear to use any issue tracking system.
+
+Software package code files were further measured for the percentage of code
+that is comments. The findings are presented in Table~\ref{gitrepodata}.
+Packages with a higher percentage of comments were designated as more
+maintainable. Comments represent more than 10 percent of code files in 15
+packages, and the average percentage of code comments is about 14 percent. Four
+of the top five overall ranked packages (Ludwig, ESPResSo, Palabos, OpenLB) have
+more than the average. Fifth ranked LUMA has only 0.2 percent comments, the
+fewest of any package. The package has the most lines of source code, with over
+four million. The next largest package is ESPResSo++ with one million.
 
@@
 	\begin{center}
-		\includegraphics[width=1.0\textwidth]{maintainability_chart}
+		\includegraphics[width=1.0\textwidth]{./figures/maintainability_chart}
 		\caption{AHP Maintainability Score}
@@
 
-Figure \ref{Fig_Maintainability} shows the maintainability ranking of the software packages using AHP.
-Software packages with a high score provide version numbers on documents and source code releases, have an abundance of high quality artifacts, and use an issue tracking tool and version control system.
-These packages also appear to reasonably handle tracked issues, having most of their issues closed.
-Their code files are well commented with more than 10 percent of the code being comments.
-Furthermore, nine of the top 10 ranked packages are noted as being alive.
-MP-LABS is noted as being dead.\\
-
-\begin{onehalfspacing}
+Figure~\ref{Fig_Maintainability} shows the maintainability ranking of the
+software packages using AHP. Software packages with a high score provide version
+numbers on documents and source code releases, have an abundance of high quality
+artifacts, and use an issue tracking tool and version control system. These
+packages also appear to reasonably handle tracked issues, having most of their
+issues closed. Their code files are well commented with more than 10 percent of
+the code being comments. Furthermore, nine of the top 10 ranked packages are
+noted as being alive. MP-LABS is noted as being dead.\\
+
+\begin{table}
 \begin{center}
@@
 	\end{tabular}
-	\captionof{table}{Git Repository Data}\label{gitrepodata}
+	\caption{Git Repository Data} \label{gitrepodata}
 \end{center}
-\end{onehalfspacing}
-
-\subsection{Reusability}\label{reusabilityresults}
-
-Each software package was measured for the total number of source code files.
-A larger number of source files was associated with increased reusability due to increased modularization.
-Some packages have more features than others, consequently contributing to reusablility since they have more source code that can be reused.
-The software packages were also reviewed for the presence of API documentation, which indicates that a software package was developed with interaction between other software applications in mind.
+\end{table}
+
+\subsection{Reusability} \label{reusabilityresults}
+
+Each software package was measured for the total number of source code files. A
+larger number of source files was associated with increased reusability due to
+increased modularization. Some packages have more features than others,
+consequently contributing to reusablility since they have more source code that
+can be reused. The software packages were also reviewed for the presence of API
+documentation, which indicates that a software package was developed with
+interaction between other software applications in mind.
 
@@
 	\begin{center}
-		\includegraphics[width=1.0\textwidth]{reusability_chart}
+		\includegraphics[width=1.0\textwidth]{./figures/reusability_chart}
 		\caption{AHP Reusability Score}
@@
 
-Figure \ref{Fig_Reusabilty} shows the reusability ranking of the software packages using AHP.
-Software packages with a high score have thousands of source code files and API documentation.
-The highest scoring packages, ESPResSo and waLBerla, have extensive functionality, including graphical visualizations as well as modeling that does not use LBM.
-For this reason a comparison with other software packages is not on a level field.
-However, these packages do have an abundance of reusable components.
-Furthermore, nine of the top 10 ranked packages are noted as being alive.
-HemeLB is noted as being dead.
-
+Figure~\ref{Fig_Reusabilty} shows the reusability ranking of the software
+packages using AHP. Software packages with a high score have thousands of source
+code files and API documentation. The highest scoring packages, ESPResSo and
+waLBerla, have extensive functionality, including graphical visualizations as
+well as modeling that does not use LBM. For this reason a comparison with other
+software packages is not on a level field. However, these packages do have an
+abundance of reusable components. Furthermore, nine of the top 10 ranked
+packages are noted as being alive. HemeLB is noted as being dead.
+
+\begin{table}
 	\begin{center}
-		\begin{onehalfspacing}
 		\begin{tabular}{ p{3.5cm}p{2cm}p{2.5cm}p{2cm}p{2.5cm} }
@@
 		\end{tabular}
-		\captionof{table}{Module Data}\label{moduledata}
-		\end{onehalfspacing}
+		\caption{Module Data} \label{moduledata}
 	\end{center}
-
-Table \ref{moduledata} shows file and line of code data of the software packages.
-Packages with a high reusability score do not have many LOC per text file, generally having a few hundred lines or less.
-This suggests that the source code of these packages is likely functionally modularized, and modules could be reused in other projects.
-
-There was a strong focus on modularity when designing the waLBerla framework to enhance productivity, reusability, and maintainability \citep{bauer2021walberla}.
-Its software design has enabled waLBerla to be successfully applied in several projects as a basis for various extensions \citep{bauer2021walberla}.
+\end{table}
+
+Table~\ref{moduledata} shows file and line of code data of the software
+packages. Packages with a high reusability score do not have many LOC per text
+file, generally having a few hundred lines or less. This suggests that the
+source code of these packages is likely functionally modularized, and modules
+could be reused in other projects.
+
+There was a strong focus on modularity when designing the waLBerla framework to
+enhance productivity, reusability, and maintainability
+\citep{bauer2021walberla}. Its software design has enabled waLBerla to be
+successfully applied in several projects as a basis for various extensions
+\citep{bauer2021walberla}.
 
@@
 
-Ten random source code files of each software package were reviewed for several measures.
-This assessment of surface understandability may not perfectly reflect each package due to the practical limitation of only being able to test 10 files.
-
-All of the packages appear to have a consistent indentation and formatting style.
-Only LUMA and HemeLB explicitly identify coding standards that are used during development.
-Generally, the software packages use consistent, distinctive, and meaningful code identifiers.
-Only four packages (LB2D-Prime, LB3D-Prime, LIMBES, MP-LABS) appear to use vague identifiers, such as single letters for variables.
-Symbolic constants were observed in the source code of 12 packages.
-The constants are used for various parameters, mathematical constants, and matrix definitions.
-All of the packages are well commented and the comments clearly indicate what is being done.
-Domain algorithms are noted in the source code of 11 packages.
-Table \ref{moduledata} suggests that the software packages are modularized to various degrees.
-When observing the source code files, it was found that 13 of the packages have a consistent style and order of function parameters.
+Ten random source code files of each software package were reviewed for several
+measures. This assessment of surface understandability may not perfectly reflect
+each package due to the practical limitation of only being able to test 10
+files.
+
+All of the packages appear to have a consistent indentation and formatting
+style. Only LUMA and HemeLB explicitly identify coding standards that are used
+during development. Generally, the software packages use consistent,
+distinctive, and meaningful code identifiers. Only four packages (LB2D-Prime,
+LB3D-Prime, LIMBES, MP-LABS) appear to use vague identifiers, such as single
+letters for variables. Symbolic constants were observed in the source code of 12
+packages. The constants are used for various parameters, mathematical constants,
+and matrix definitions. All of the packages are well commented and the comments
+clearly indicate what is being done. Domain algorithms are noted in the source
+code of 11 packages. Table~\ref{moduledata} suggests that the software packages
+are modularized to various degrees. When observing the source code files, it was
+found that 13 of the packages have a consistent style and order of function
+parameters.
 
@@
 	\begin{center}
-		\includegraphics[width=1.0\textwidth]{understandability_chart}
+		\includegraphics[width=1.0\textwidth]{./figures/understandability_chart}
 		\caption{AHP Understandability Score}
@@
 
-Figure \ref{Fig_Understandability} shows the surface understandability ranking of the software packages using AHP.
-Software packages with a high score have a consistent indentation and formatting style, and consistent, distinctive, and meaningful code identifiers.
-They also have symbolic constants, and explicitly identify mathematical and LBM algorithms.
-Their comments are clear and indicate what is being done in the source code.
-The source code is well modularized and structured.
-Furthermore, four of the top five ranked packages are noted as being alive.
+Figure~\ref{Fig_Understandability} shows the surface understandability ranking
+of the software packages using AHP. Software packages with a high score have a
+consistent indentation and formatting style, and consistent, distinctive, and
+meaningful code identifiers. They also have symbolic constants, and explicitly
+identify mathematical and LBM algorithms. Their comments are clear and indicate
+what is being done in the source code. The source code is well modularized and
+structured. Furthermore, four of the top five ranked packages are noted as being
+alive.
 
@@
 
-Software package artifacts were reviewed for the identification of a specific development model, like a waterfall of agile development model, and the presence of documentation recording the development process and standard.
-They were also reviewed for the identification of the development environment, and the presence of release notes.
-The packages tended to not explicitly use well-known development models.
-This was also noted in the interviews with developers, as detailed below.
-The development teams of these packages are fairly small and easily organized without the need for such processes.
-Seven of the software packages did have some artifacts outlining the general development process, how to contribute, and the status of the package or its components.
-Eight of the packages have artifacts that note the development environment.
-While this information could help developers, and would improve transparency, the small close-knit nature of the development teams make explicitly publicly specifying this information practically unnecessary.
-Version release notes were found in nine of the software packages.
+Software package artifacts were reviewed for the identification of a specific
+development model, like a waterfall of agile development model, and the presence
+of documentation recording the development process and standard. They were also
+reviewed for the identification of the development environment, and the presence
+of release notes. The packages tended to not explicitly use well-known
+development models. This was also noted in the interviews with developers, as
+detailed below. The development teams of these packages are fairly small and
+easily organized without the need for such processes. Seven of the software
+packages did have some artifacts outlining the general development process, how
+to contribute, and the status of the package or its components. Eight of the
+packages have artifacts that note the development environment. While this
+information could help developers, and would improve transparency, the small
+close-knit nature of the development teams make explicitly publicly specifying
+this information practically unnecessary. Version release notes were found in
+nine of the software packages.
 
@@
 	\begin{center}
-		\includegraphics[width=1.0\textwidth]{visibilitytransparency_chart}
+		\includegraphics[width=1.0\textwidth]{./figures/visibilitytransparency_chart}
 		\caption{AHP Visibility and Transparency Score}
@@
 
-Figure \ref{Fig_VisibilityTransparency} shows the visibility and transparency ranking of the software packages using AHP.
-Software packages with a high score have an explicit development model and defined development process.
-They also
-had detailed and easy to access notes accompanying software releases. Furthermore, four of the top five ranked packages are noted as being alive. MP-LABS is noted as being dead.
+Figure~\ref{Fig_VisibilityTransparency} shows the visibility and transparency
+ranking of the software packages using AHP. Software packages with a high score
+have an explicit development model and defined development process. They also
+had detailed and easy to access notes accompanying software releases.
+Furthermore, four of the top five ranked packages are noted as being alive.
+MP-LABS is noted as being dead.
 
@@
 \begin{figure}[h!]
-	\begin{center}
-		\includegraphics[width=1.0\textwidth]{finalscore_chart}
+	\centering
+		\includegraphics[width=1.0\textwidth]{./figures/finalscore_chart}
 		\caption{AHP Overall Score}
 		\label{Fig_OverallScore}
-	\end{center}
 \end{figure}
 
-Figure \ref{Fig_OverallScore} shows the overall ranking of the software packages using AHP.
-Software packages with an overall high score had ranked high in at least several of the individual qualities that were quantitatively measured.
-The overall ranking is found with an assumption of an equal weighting of all qualities.
-If the qualities were weighed differently, the overall software package ranking would change.
-
-Looking at the top three ranked packages, Ludwig scored high in surface correctness and verifiability, surface robustness, surface usability, maintainability, and visibility and transparency.
-ESPResSo had achieved a relative high score in installability, surface correctness and verifiability, surface usability, maintainability, reusability, and visibility and transparency.
-Palabos scored high in installability, surface reliability, surface robustness, maintainability, and understandability.
-
-Section \ref{qualityrecommentations} further analyzes these findings and offers some software quality recommendations for future development of LBM software.
-
-\newpage
-
-\section{Qualitative Findings From Developer Interviews}\label{interviewresults}
-
-This section presents the qualitative findings from data that was gathered during interviews with developers for qualities listed in Section \ref{softwarequalities}.
-The interview questions are found in Appendix \ref{interviewquestions}.
+Figure~\ref{Fig_OverallScore} shows the overall ranking of the software packages
+using AHP. Software packages with an overall high score had ranked high in at
+least several of the individual qualities that were quantitatively measured. The
+overall ranking is found with an assumption of an equal weighting of all
+qualities. If the qualities were weighed differently, the overall software
+package ranking would change.
+
+Looking at the top three ranked packages, Ludwig scored high in surface
+correctness and verifiability, surface robustness, surface usability,
+maintainability, and visibility and transparency. ESPResSo had achieved a
+relative high score in installability, surface correctness and verifiability,
+surface usability, maintainability, reusability, and visibility and
+transparency. Palabos scored high in installability, surface reliability,
+surface robustness, maintainability, and understandability.
+
+Section~\ref{qualityrecommentations} further analyzes these findings and offers
+some software quality recommendations for future development of LBM software.
+
+\section{Qualitative Findings From Developer Interviews} \label{interviewresults}
+
+The interview process
+presented here was approved by the McMaster University Research Ethics Board
+under the application number 
+\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/MACREM/Application.pdf}
+{MREB\#: 5219}.
+
+This section presents the qualitative findings from data that was gathered
+during interviews with developers for qualities listed in Section
+~\ref{softwarequalities}. The interview questions are found in Appendix
+~\ref{interviewquestions}.
 
@@
 
-Interviews with developers confirmed that these software packages are developed by domain experts with backgrounds in physics, mathematics and mechanical engineering.
-It was noted in these interviews that some of the developers do not have formal software engineering education.
-Some of the development teams include computer scientists.
-Despite a lack of visible domain documentation and a resulting lower surface correctness and verifiability score, it is clear that some of the software packages were developed by teams with significant domain knowledge on account of the academic backgrounds of their developers.
-
-Interviews suggested a more frequent use of both unit testing and continuous integration in the development processes than what was observed from the initial survey.
-For example, OpenLB, pyLBM, and TCLB use such methods during development despite this not being explicitly clear from an analysis of the material available online.
-The correctness and verificablity of such packages is not measured well using surface analysis.
-
-Several interviewed developers alluded to difficulty with testing the correctness of large numbers of features, and some even manually tested program output.
-The use of well defined unit testing tools could decrease the time spent testing some feature.
+Interviews with developers confirmed that these software packages are developed
+by domain experts with backgrounds in physics, mathematics and mechanical
+engineering. It was noted in these interviews that some of the developers do not
+have formal software engineering education. Some of the development teams
+include computer scientists. Despite a lack of visible domain documentation and
+a resulting lower surface correctness and verifiability score, it is clear that
+some of the software packages were developed by teams with significant domain
+knowledge on account of the academic backgrounds of their developers.
+
+Interviews suggested a more frequent use of both unit testing and continuous
+integration in the development processes than what was observed from the initial
+survey. For example, OpenLB, pyLBM, and TCLB use such methods during development
+despite this not being explicitly clear from an analysis of the material
+available online. The correctness and verifiability of such packages is not
+measured well using surface analysis.
+
+Several interviewed developers alluded to difficulty with testing the
+correctness of large numbers of features, and some even manually tested program
+output. The use of well defined unit testing tools could decrease the time spent
+testing some feature.
 
@@
 
-Interviews with developers revealed several usability issues.
-Some users have misunderstood the boundaries of LBM and CFD, and have combined or applied methods that are not physically sound.
-Sometimes users have applied LBM to poorly defined or inappropriate fluid dynamics problems.
-For example, they may wish to model flow through or around a structure despite having limited information about the structure or its environment, and having little previous knowledge of CFD.
-The users do not realize the limitations of the methods, of the software, and do not understand the requirements to properly model a problem with the software.
-As the developer of TCLB noted, such software packages are not designed to be used ``out of the box'' in a plug and play fashion, and it could take months or more to set up the CFD problems correctly.
-Developers of some software packages, including ESPResSo, mitigated this by editing the source code to prevent users from ``combining methods that are not physically sound together'', and by updating the documentation to better inform users of LBM limitations, and of the requirements to properly model appropriate problems, including what algorithms and parameters to use.
-
-Some additional but infrequent software usability issues were commented on by the developers.
-Users have had trouble with installation and understanding how to maneuver the interfaces and how to set up or run models.
-These issues are addressed by various user support models, including frequently asked questions sections on the software package websites, user guides, and hardware and software requirements specifications.
-
-One software package (ESPResSo) changed some of its scripting language to Python to make it more usable.
-The developer commented that this was ``the biggest step in terms of usability over the years'', further commenting that ``most people in the field know [Python]'' and that ``it's easy to learn''.
+Interviews with developers revealed several usability issues. Some users have
+misunderstood the boundaries of LBM and CFD, and have combined or applied
+methods that are not physically sound. Sometimes users have applied LBM to
+poorly defined or inappropriate fluid dynamics problems. For example, they may
+wish to model flow through or around a structure despite having limited
+information about the structure or its environment, and having little previous
+knowledge of CFD. The users do not realize the limitations of the methods, of
+the software, and do not understand the requirements to properly model a problem
+with the software. As the developer of TCLB noted, such software packages are
+not designed to be used ``out of the box'' in a plug and play fashion, and it
+could take months or more to set up the CFD problems correctly. Developers of
+some software packages, including ESPResSo, mitigated this by editing the source
+code to prevent users from ``combining methods that are not physically sound
+together'', and by updating the documentation to better inform users of LBM
+limitations, and of the requirements to properly model appropriate problems,
+including what algorithms and parameters to use.
+
+Some additional but infrequent software usability issues were commented on by
+the developers. Users have had trouble with installation and understanding how
+to maneuver the interfaces and how to set up or run models. These issues are
+addressed by various user support models, including frequently asked questions
+sections on the software package websites, user guides, and hardware and
+software requirements specifications.
+
+One software package (ESPResSo) changed some of its scripting language to Python
+to make it more usable. The developer commented that this was ``the biggest step
+in terms of usability over the years'', further commenting that ``most people in
+the field know [Python]'' and that ``it's easy to learn''.
 
@@
 
-Interviews with developers revealed that most projects do have a defined process for accepting contributions from team members.
-The packages rarely get contributions from outside developers, but the process would be similar as for the aforementioned group.
-
-Contributions are made through GitHub, and are then reviewed and pulled by lead developers, often with consultation with a group of core developers depending on the organizational model.
-Continuous integration is part of the process for some packages.
-
-Some developers noted that their software package does not have well defined contributing guide in the repository, but it might be a good idea to add one in the near future.
-They would be happy to see contribution from outside of their organization, but currently this does not happen.
-
-Furthermore, maintainability has been addressed by increasing source code modularity, reducing duplicate information, and improving abstraction by developing well defined interfaces.
-This was noted by the developers of ESPResSo and pyLBM.
-Several software packages have had sections of their code base redeveloped with languages that the developers felt are more understandable and readable, and that are better supported, such as Python.
-Data structures have also been redeveloped and storage has been improved.
-A developer of pyLBM mentioned that the geometries and models of their system had been ``decoupled'', using abstraction and modularization of the source code, to make it ``very easy to add [new] features''.
+Interviews with developers revealed that most projects do have a defined process
+for accepting contributions from team members. The packages rarely get
+contributions from outside developers, but the process would be similar as for
+the aforementioned group.
+
+Contributions are made through GitHub, and are then reviewed and pulled by lead
+developers, often with consultation with a group of core developers depending on
+the organizational model. Continuous integration is part of the process for some
+packages.
+
+Some developers noted that their software package does not have well defined
+contributing guide in the repository, but it might be a good idea to add one in
+the near future. They would be happy to see contribution from outside of their
+organization, but currently this does not happen.
+
+Furthermore, maintainability has been addressed by increasing source code
+modularity, reducing duplicate information, and improving abstraction by
+developing well defined interfaces. This was noted by the developers of ESPResSo
+and pyLBM. Several software packages have had sections of their code base
+redeveloped with languages that the developers felt are more understandable and
+readable, and that are better supported, such as Python. Data structures have
+also been redeveloped and storage has been improved. A developer of pyLBM
+mentioned that the geometries and models of their system had been ``decoupled'',
+using abstraction and modularization of the source code, to make it ``very easy
+to add [new] features''.
 
@@
 
-Software packages were not quantitatively measured for modifiability.
-In this project we asked developers to comment on modifiability when we interviewed them.
-Specifically, we asked if ease of future changes to the system, modules, and code blocks was considered when designing the software.
-We also asked if any measures had been taken to ensure the ease of future changes.
-All of the developers that were interviewed noted that the ease of future changes was considered and that measures to ensure it had been taken, including
-requiring the separation of software components in the source code architecture.
-
-A high degree of code modularity and abstraction was noted by developers as a measure to ensure the ease of future changes.
-This can be ensured by separating components and hiding information behind well defined interfaces.
-The developer of ESPResSo also noted that some of the code base was transitioned from C to C++, which could ease modifiability of that software package.
-The developer of TCLB noted that their software package was designed to allow for the addition of some LBM features, but changes to major aspects of the system would be difficult.
-For example, ``implementing a new model will be an easy contribution'', but changes to the ``Cartesian mesh...will be a nightmare''.
-Furthermore, the package was designed with flexible data structures and storage in mind.
-
-Some software packages, like Palabos, provide validation benchmarks for their core fundamental algorithmic ingredients \citep{latt2021palabos}.
-The stated intent of these benchmarks is to showcase the validity and usefulness of the package to stimulate the development of third-user extensions.
-The Palabos package identifies as a development framework for modeling problems in various CFD areas.
+Software packages were not quantitatively measured for modifiability. In this
+project we asked developers to comment on modifiability when we interviewed
+them. Specifically, we asked if ease of future changes to the system, modules,
+and code blocks was considered when designing the software. We also asked if any
+measures had been taken to ensure the ease of future changes. All of the
+developers that were interviewed noted that the ease of future changes was
+considered and that measures to ensure it had been taken, including requiring
+the separation of software components in the source code architecture.
+
+A high degree of code modularity and abstraction was noted by developers as a
+measure to ensure the ease of future changes. This can be ensured by separating
+components and hiding information behind well defined interfaces. The developer
+of ESPResSo also noted that some of the code base was transitioned from C to
+C++, which could ease modifiability of that software package. The developer of
+TCLB noted that their software package was designed to allow for the addition of
+some LBM features, but changes to major aspects of the system would be
+difficult. For example, ``implementing a new model will be an easy
+contribution'', but changes to the ``Cartesian mesh...will be a nightmare''.
+Furthermore, the package was designed with flexible data structures and storage
+in mind.
+
+Some software packages, like Palabos, provide validation benchmarks for their
+core fundamental algorithmic ingredients \citep{latt2021palabos}. The stated
+intent of these benchmarks is to showcase the validity and usefulness of the
+package to stimulate the development of third-user extensions. The Palabos
+package identifies as a development framework for modeling problems in various
+CFD areas.
 
@@
 
-Software developers noted that they believe users have generally found their packages to be understandable.
-The interviewed developer of ESPResSo commented that some users have attempted to run physically incompatible LBM methods, and the solution was to edit the code to prevent such combinations, as well as to update the documentation to prevent misunderstanding the methods.
-Similarly, a developer of pyLBM noted that some users had issues setting up parameters for LBM schemes.
-The solution to this was to update the interface where these parameters are set, as well as to add functionality to test the stability of the parameters.
-A developer of OpenLB noted that some users lack the background knowledge to easily model fluid dynamics problems using their software.
-A frequently asked questions section was added to their package website to help users find answers to common questions.
-The package also has detailed documentation, including guides and usage requirements specification, to better help users understand the software.
+Software developers noted that they believe users have generally found their
+packages to be understandable. The interviewed developer of ESPResSo commented
+that some users have attempted to run physically incompatible LBM methods, and
+the solution was to edit the code to prevent such combinations, as well as to
+update the documentation to prevent misunderstanding the methods. Similarly, a
+developer of pyLBM noted that some users had issues setting up parameters for
+LBM schemes. The solution to this was to update the interface where these
+parameters are set, as well as to add functionality to test the stability of the
+parameters. A developer of OpenLB noted that some users lack the background
+knowledge to easily model fluid dynamics problems using their software. A
+frequently asked questions section was added to their package website to help
+users find answers to common questions. The package also has detailed
+documentation, including guides and usage requirements specification, to better
+help users understand the software.
 
@@
 
-Software packages were not quantitatively measured for traceability.
-In this project we asked developers to comment on traceability, specifically on their software package's documentation and how it fit into their development process.
-
-The interviewed developer of ESPResSo noted that all major additions to their package had accompanying changes to artifacts and documentation.
-They noted that considerable effort had been put into the documentation.
-They further commented that they want to lower the entry barrier for new developers, and because of that their package has a considerable amount of developer documentation.
-This documentation informs developers on how to get started, and orients them to the artifacts, source code, and system architecture, as well as how the software package build system works, and how the coupling between the simulation engine and the interface works.
-
-Developers noted the importance of documentation for both the users and developers of their software.
-New features are always added to the documentation.
-The developers use documentation to stay up to date on the status of the software package, and to help expand features, like computational models or algorithms.
-This is necessary so that the coding standard for these models is kept consistent with new developers.
-
-The importance of documentation for both users and developers was stressed throughout the interviews.
-However, it was noted several times that a lack of time and funding has a negative affect on the documentation.
-Most of the developers are scientific researchers evaluated on the scientific papers that they produce.
-Writing and updating documentation is something that is done in their free time, if that time arises.
-Sometimes it is a last priority for the developers.
-Finding ways to hasten updating documentation would increase the frequency of such updates and benefit both users and developers.
+Software packages were not quantitatively measured for traceability. In this
+project we asked developers to comment on traceability, specifically on their
+software package's documentation and how it fit into their development process.
+
+The interviewed developer of ESPResSo noted that all major additions to their
+package had accompanying changes to artifacts and documentation. They noted that
+considerable effort had been put into the documentation. They further commented
+that they want to lower the entry barrier for new developers, and because of
+that their package has a considerable amount of developer documentation. This
+documentation informs developers on how to get started, and orients them to the
+artifacts, source code, and system architecture, as well as how the software
+package build system works, and how the coupling between the simulation engine
+and the interface works.
+
+Developers noted the importance of documentation for both the users and
+developers of their software. New features are always added to the
+documentation. The developers use documentation to stay up to date on the status
+of the software package, and to help expand features, like computational models
+or algorithms. This is necessary so that the coding standard for these models is
+kept consistent with new developers.
+
+The importance of documentation for both users and developers was stressed
+throughout the interviews. However, it was noted several times that a lack of
+time and funding has a negative affect on the documentation. Most of the
+developers are scientific researchers evaluated on the scientific papers that
+they produce. Writing and updating documentation is something that is done in
+their free time, if that time arises. Sometimes it is a last priority for the
+developers. Finding ways to hasten updating documentation would increase the
+frequency of such updates and benefit both users and developers.
 
 The developer of OpenLB noted the use of documentation generators like Doxygen.
-It would be advisable for more projects to use such automatic document generation tools, since some projects do not do this.
+It would be advisable for more projects to use such automatic document
+generation tools, since some projects do not do this.
 
@@
 Developers were asked to comment on the obstacles in their development process.
-The developer of ESPResSo noted that a lot of their source code had been written with a specific application in mind, and that there is too much coupling between components.
-Addressing this issue would help with code modifiability and reusability.
-Updating the development process would help resolve this issue and prevent such issues in the future.
-Improving the visibility of software changes and a peer review process would also help.
-Improving the software engineering education or experience of developers is also an idea that was brought up by several developers.
-Specifically, developers should always write code that is decoupled and modular, and should keep in mind the visibility of their contributions by better updating documentation and ensuring that their contributions are transparent to the rest of the developers.
-This would help catch issues in the contributions, and improve source code maintenance.
-
-According to the interviewed developer of ESPResSo, some obstacles to the development processes of their package had been overcome by the introduction of continuous integration practices, and a peer review process for contributions.
-These practices decrease development and maintenance times.
-The developer of TCLB mentioned that their package had two sets of code, for executing the models on the CPU and GPU, and that maintenance was decreased by introducing macros, a practice which became a common part of the development process.
-
-Developers were also asked how documentation fits into their development process.
-Several developers noted that developer documentation plays an important role in familiarizing potential contributors to the software system architecture.
-Without the guidance that the documentation provides it would be unlikely that contributions would pass the peer review process.
-
-None of the software packages whose developers were interviewed have a formal software development model.
-The packages all have fairly small development teams.
-These teams do accept outside contributors, but generally the teams are tight-knit, often working at the same institution, although one of the packages has an international team.
-The developer of ESPResSo noted that while no formal model is used, their development model is something similar to a combination of agile and waterfall development models.
-
-The developers noted similar project management processes.
-For teams of only a couple of developers, the addition of new features or major changes are discussed with the entire team.
-Projects with more than a couple of developers have lead developer roles.
-These lead developers review potential additions to the software.
-The software packages use GitHub for managing the project.
+The developer of ESPResSo noted that a lot of their source code had been written
+with a specific application in mind, and that there is too much coupling between
+components. Addressing this issue would help with code modifiability and
+reusability. Updating the development process would help resolve this issue and
+prevent such issues in the future. Improving the visibility of software changes
+and a peer review process would also help. Improving the software engineering
+education or experience of developers is also an idea that was brought up by
+several developers. Specifically, developers should always write code that is
+decoupled and modular, and should keep in mind the visibility of their
+contributions by better updating documentation and ensuring that their
+contributions are transparent to the rest of the developers. This would help
+catch issues in the contributions, and improve source code maintenance.
+
+According to the interviewed developer of ESPResSo, some obstacles to the
+development processes of their package had been overcome by the introduction of
+continuous integration practices, and a peer review process for contributions.
+These practices decrease development and maintenance times. The developer of
+TCLB mentioned that their package had two sets of code, for executing the models
+on the CPU and GPU, and that maintenance was decreased by introducing macros, a
+practice which became a common part of the development process.
+
+Developers were also asked how documentation fits into their development
+process. Several developers noted that developer documentation plays an
+important role in familiarizing potential contributors to the software system
+architecture. Without the guidance that the documentation provides it would be
+unlikely that contributions would pass the peer review process.
+
+None of the software packages whose developers were interviewed have a formal
+software development model. The packages all have fairly small development
+teams. These teams do accept outside contributors, but generally the teams are
+tight-knit, often working at the same institution, although one of the packages
+has an international team. The developer of ESPResSo noted that while no formal
+model is used, their development model is something similar to a combination of
+agile and waterfall development models.
+
+The developers noted similar project management processes. For teams of only a
+couple of developers, the addition of new features or major changes are
+discussed with the entire team. Projects with more than a couple of developers
+have lead developer roles. These lead developers review potential additions to
+the software. The software packages use GitHub for managing the project.
 Typically there are several development branches as well as the master branch.
@@
 
-Software packages were not quantitatively measured for reproducibility.
-In this project we asked developers to comment on reproducibility when we interviewed them.
-
-Developers were asked if they have any concern that their computational results would not be reproducible in the future, and if they had taken any steps to ensure reproducibility.
-
-The developer of ESPResSo noted a comparison of the results of their methods against manually calculated results.
-These comparison tests are automatically run for all source code changes.
-The tests are run when a pull request is opened on GitHub.
-Even once these tests are complete, a peer review process is done before changes are fully committed to the appropriate branch.
-The results for all of the LBM schemes on the software package development branch are also frequently compared for correctness, ensuring that the system output reflects the expected output.
-
-Several developers noted that they currently do not have a system in place to test for reproducibility, but it is of interest and could be implemented in the future.
-Generally, the mathematical foundations of the models are verified, but the output of the software package is not compared to other output.
-Depending on the package and how it outputs solutions, it may not be practical or feasible.
-A correct output may not be exactly reproducible, as it may be dependent on a probability distribution, so strictly comparing results may not be appropriate.
+Software packages were not quantitatively measured for reproducibility. In this
+project we asked developers to comment on reproducibility when we interviewed
+them.
+
+Developers were asked if they have any concern that their computational results
+would not be reproducible in the future, and if they had taken any steps to
+ensure reproducibility.
+
+The developer of ESPResSo noted a comparison of the results of their methods
+against manually calculated results. These comparison tests are automatically
+run for all source code changes. The tests are run when a pull request is opened
+on GitHub. Even once these tests are complete, a peer review process is done
+before changes are fully committed to the appropriate branch. The results for
+all of the LBM schemes on the software package development branch are also
+frequently compared for correctness, ensuring that the system output reflects
+the expected output.
+
+Several developers noted that they currently do not have a system in place to
+test for reproducibility, but it is of interest and could be implemented in the
+future. Generally, the mathematical foundations of the models are verified, but
+the output of the software package is not compared to other output. Depending on
+the package and how it outputs solutions, it may not be practical or feasible. A
+correct output may not be exactly reproducible, as it may be dependent on a
+probability distribution, so strictly comparing results may not be appropriate.
 
 The source code and artifacts of some software packages may be reproducible.
-There is considerable variance in the quality of software specifications and other developer documentation.
-Some packages are well detailed, and translating the specifications into source code will produce similar results across developers.
-Developers were not asked to comment on the reproducibility of their source code from their requirements specifications and design documentation.
-This question should be considered in the next iteration of state of the practice assessments.
+There is considerable variance in the quality of software specifications and
+other developer documentation. Some packages are well detailed, and translating
+the specifications into source code will produce similar results across
+developers. Developers were not asked to comment on the reproducibility of their
+source code from their requirements specifications and design documentation.
+This question should be considered in the next iteration of state of the
+practice assessments.
 
@@
 
-Software packages were not quantitatively measured for unambiguity.
-In this project we asked developers to comment on unambiguity when we interviewed them.
-
-Developers were asked if they thought that the current documentation can clearly convey all necessary knowledge to the users, and if they had taken any steps to ensure clarity.
-
-The developer of ESPResSo noted that their documentation was meant for users that are already familiar with the underlying physics and CFD methods.
-These concepts are not explained in detail within the documentation.
-Users should acquire this knowledge from suitable external sources.
-The documentation focuses on how to technically use the software package, and includes a user guide and tutorial walk through of how to set up and run a simulation.
-With this in mind, the developer believes that their documentation is in reasonable shape for users with a minimum knowledge of the underlying physics.
-If new users have technical questions these can be addressed in further revisions of the documentation.
-New developers should find that the documentation is reasonably clear and useful.
-Information that is missing, like detailed explanations of dependencies, is referenced in the documents.
-
-The developer of pyLBM also noted that their documentation was in reasonable shape, but that they ``need more [user] feedback to improve [it]''.
-They also noted that they believe a lack of knowledge of the underlying physics and CFD concepts can be an issue for some users.
-This information can be referenced in the documentation, but it is not something that the documentation needs to detail.
-
-\newpage
-
-\section{Answers To Research Questions}\label{answersquestions}
-
-This section answers the research questions listed in Section \ref{purpose} using the quantitative data presented in Section \ref{AHPresults}, qualitative data presented in Section \ref{interviewresults}, additional data from software package repositories and artifacts, and domain expert feedback.
-Each subsection answers one of the research questions.
-Software package artifacts are examined in Section \ref{artifacts}, tools are listed in Section \ref{tools}, development principles, processes, and methodologies are discussed in Section \ref{prinprocmeth}, development pain points are explored in Section \ref{painpoints}, and our recommendations for improving software qualities are presented in Section \ref{qualityrecommentations}.
-A comparison between our designations of the software packages to community rankings is made in Section \ref{comparison}.
-Finally, threats to the validity of this assessment are noted in Section \ref{threats}.
-
-\subsection{Artifacts Present}\label{artifacts}
-
-This subsection answers the research question: What artifacts are present in current software packages?
-
-The software packages were examined for the presence of artifacts, which were then categorized by frequency.
-We have grouped them into common, less common, and rare artifacts in Table \ref{artifactspresent}.
-Common artifacts were found in 16 to 23 ($>$70\%) of the software packages.
-Less common artifacts were found in 8 to 15 (35-70\%) of the software packages.
-Rare artifacts were found in 1 to 7 ($<$35\%) of the software packages.
-
-\begin{onehalfspace}
+Software packages were not quantitatively measured for unambiguity. In this
+project we asked developers to comment on unambiguity when we interviewed them.
+
+Developers were asked if they thought that the current documentation can clearly
+convey all necessary knowledge to the users, and if they had taken any steps to
+ensure clarity.
+
+The developer of ESPResSo noted that their documentation was meant for users
+that are already familiar with the underlying physics and CFD methods. These
+concepts are not explained in detail within the documentation. Users should
+acquire this knowledge from suitable external sources. The documentation focuses
+on how to technically use the software package, and includes a user guide and
+tutorial walk through of how to set up and run a simulation. With this in mind,
+the developer believes that their documentation is in reasonable shape for users
+with a minimum knowledge of the underlying physics. If new users have technical
+questions these can be addressed in further revisions of the documentation. New
+developers should find that the documentation is reasonably clear and useful.
+Information that is missing, like detailed explanations of dependencies, is
+referenced in the documents.
+
+The developer of pyLBM also noted that their documentation was in reasonable
+shape, but that they ``need more [user] feedback to improve [it]''. They also
+noted that they believe a lack of knowledge of the underlying physics and CFD
+concepts can be an issue for some users. This information can be referenced in
+the documentation, but it is not something that the documentation needs to
+detail.
+
+\section{Answers To Research Questions} \label{answersquestions}
+
+This section answers the research questions listed in Section~\ref{purpose}
+using the quantitative data presented in Section~\ref{AHPresults}, qualitative
+data presented in Section~\ref{interviewresults}, additional data from software
+package repositories and artifacts, and domain expert feedback. Each subsection
+answers one of the research questions. Software package artifacts are examined
+in Section~\ref{artifacts}, tools are listed in Section~\ref{tools}, development
+principles, processes, and methodologies are discussed in Section
+~\ref{prinprocmeth}, development pain points are explored in Section
+~\ref{painpoints}, and our recommendations for improving software qualities are
+presented in Section~\ref{qualityrecommentations}. A comparison between our
+designations of the software packages to community rankings is made in Section
+~\ref{comparison}. Finally, threats to the validity of this assessment are noted
+in Section~\ref{threats}.
+
+\subsection{Artifacts Present} \label{artifacts}
+
+This subsection answers the research question: What artifacts are present in
+current software packages?
+
+The software packages were examined for the presence of artifacts, which were
+then categorized by frequency. We have grouped them into common, less common,
+and rare artifacts in Table~\ref{artifactspresent}. Common artifacts were found
+in 16 to 23 ($>$70\%) of the software packages. Less common artifacts were found
+in 8 to 15 (35-70\%) of the software packages. Rare artifacts were found in 1 to
+7 ($<$35\%) of the software packages.
+
+\begin{table}
 	\begin{center}
@@
 		\end{tabular}
-		\captionof{table}{Artifacts Present}\label{artifactspresent}
+		\caption{Artifacts Present} \label{artifactspresent}
 	\end{center}
-\end{onehalfspace}
+\end{table}
 
 \subsubsection{Common Artifacts}
-All of the top four AHP ranked packages, ESPREsSo, Ludwig, OpenLB, and Palabos, have each of the commonly found artifacts, except only three of them have a requirements specification or theory notes.
-Palabos is the only one of these four packages that does not have an artifact from this category.
-
-The common artifacts contribute to the quality of the software in the following ways.
-A list of authors and developers helps potential users and contributors contact project members to answer questions affecting many software qualities.
-A bug tracker helps with organizing improvements to the software.
-Dependency notes help users install the software, as does an installation guide.
-Makefiles or other automated build files decrease human error in the installation process.
-Licenses promote usage of the software.
-Requirements specifications, linked theory notes, and related articles and publications help make the software more understandable, decrease ambiguity, and help with verifying its correctness.
-README files also help with understandability.
-The tutorials help users become familiar with using the software.
-
-\subsubsection{Less Common Artifacts}\label{lesscommon}
-The top four AHP ranked packages have most of the less common artifacts.
-At the time of data collection, only one (Palabos) of the four packages did not have a user manual or guide, but there was a broken link on the package website indicating that such an artifact might exist.
-This broken link was later fixed, but this is not reflected in our data because it was not present at the time of data collection.
-Despite the broken link, Palabos does have a detailed and informative website.
-Another one (Ludwig) of the top four packages does not appear to have publicly visible design documentation.
-A third package (OpenLB) from the list does not appear to use a version control system.
-It is possible that such a system is used by the package since its website notes package version numbers, but the artifacts do not explicitly state the use of such a system.
-
-The less common artifacts also contribute to software quality.
-A change log or release notes improve traceability of the software.
-Design documentation helps with maintaining, modifying, and reusing the software.
-Version control also helps to improve these qualities, as well as with traceability.
-Functional specifications and notes clarify the software, contributing to usability and understandability.
-A user manual also helps with those qualities and with installability, and correctness and verificability.
-Performance notes suggest that performance was considered when developing the software.
-Test plans, scripts, and cases, help verify correctness.
+
+All of the top four AHP ranked packages, ESPREsSo, Ludwig, OpenLB, and Palabos,
+have each of the commonly found artifacts, except only three of them have a
+requirements specification or theory notes. Palabos is the only one of these
+four packages that does not have an artifact from this category.
+
+The common artifacts contribute to the quality of the software in the following
+ways. A list of authors and developers helps potential users and contributors
+contact project members to answer questions affecting many software qualities. A
+bug tracker helps with organizing improvements to the software. Dependency notes
+help users install the software, as does an installation guide. Makefiles or
+other automated build files decrease human error in the installation process.
+Licenses promote usage of the software. Requirements specifications, linked
+theory notes, and related articles and publications help make the software more
+understandable, decrease ambiguity, and help with verifying its correctness.
+README files also help with understandability. The tutorials help users become
+familiar with using the software.
+
+\subsubsection{Less Common Artifacts} \label{lesscommon}
+
+The top four AHP ranked packages have most of the less common artifacts. At the
+time of data collection, only one (Palabos) of the four packages did not have a
+user manual or guide, but there was a broken link on the package website
+indicating that such an artifact might exist. This broken link was later fixed,
+but this is not reflected in our data because it was not present at the time of
+data collection. Despite the broken link, Palabos does have a detailed and
+informative website. Another one (Ludwig) of the top four packages does not
+appear to have publicly visible design documentation. A third package (OpenLB)
+from the list does not appear to use a version control system. It is possible
+that such a system is used by the package since its website notes package
+version numbers, but the artifacts do not explicitly state the use of such a
+system.
+
+The less common artifacts also contribute to software quality. A change log or
+release notes improve traceability of the software. Design documentation helps
+with maintaining, modifying, and reusing the software. Version control also
+helps to improve these qualities, as well as with traceability. Functional
+specifications and notes clarify the software, contributing to usability and
+understandability. A user manual also helps with those qualities and with
+installability, and correctness and verificability. Performance notes suggest
+that performance was considered when developing the software. Test plans,
+scripts, and cases, help verify correctness.
 
 \subsubsection{Rare Artifacts}
-It is not common for the top four AHP ranked packages to have many of the rare artifacts.
-None of the top four packages have any explicit API documentation.
-Three of these packages (ESPREsSo, Ludwig, Palabos) have information on contributing to the project.
-Two of them (OpenLB, Palabos) have a FAQ section or forum.
-One (OpenLB) has verification and validation notes, and a video guide of the software.
-
-Although the artifacts were rarely found in our set of LBM software, they also contribute to software quality.
-API documentation helps with reusing the software.
-Developer and contributor manuals and guides help with maintainability, visibility and transparency.
-FAQs and forums improve usability of software.
-Verification and validation notes can be used to help check if a system meets specifications.
-Video guides can contribute to many software qualities, depending on the content of the video.
-
-\subsection{Tools Used}\label{tools}
-
-This subsection answers the research question: What tools (development, dependencies, project management) are used by current software packages?
-
-Software tools are used to support the development, verification, maintenance, and evolution of software, software processes, and artifacts \citep{ghezzi1991fundamentals}.
-Many tools are used by LBM software packages.
-The tools noted here are subdivided into development tools, dependencies, and project management tools.
+
+It is not common for the top four AHP ranked packages to have many of the rare
+artifacts. None of the top four packages have any explicit API documentation.
+Three of these packages (ESPREsSo, Ludwig, Palabos) have information on
+contributing to the project. Two of them (OpenLB, Palabos) have a FAQ section or
+forum. One (OpenLB) has verification and validation notes, and a video guide of
+the software.
+
+Although the artifacts were rarely found in our set of LBM software, they also
+contribute to software quality. API documentation helps with reusing the
+software. Developer and contributor manuals and guides help with
+maintainability, visibility and transparency. FAQs and forums improve usability
+of software. Verification and validation notes can be used to help check if a
+system meets specifications. Video guides can contribute to many software
+qualities, depending on the content of the video.
+
+\subsection{Tools Used} \label{tools}
+
+This subsection answers the research question: What tools (development,
+dependencies, project management) are used by current software packages?
+
+Software tools are used to support the development, verification, maintenance,
+and evolution of software, software processes, and artifacts
+\citep{ghezzi1991fundamentals}. Many tools are used by LBM software packages.
+The tools noted here are subdivided into development tools, dependencies, and
+project management tools.
 
@@
 
-Development tools support the development of end products, but do not become part of them, unlike dependencies that remain in the application once it is released \citep{ghezzi1991fundamentals}.
-The following type of development tools were explicitly noted in the artifacts or web-pages of the 23 LBM packages that were assessed.
-It is likely that other tools, such as debuggers, were used but are not specified in our sources.
+Development tools support the development of end products, but do not become
+part of them, unlike dependencies that remain in the application once it is
+released \citep{ghezzi1991fundamentals}. The following type of development tools
+were explicitly noted in the artifacts or web-pages of the 23 LBM packages that
+were assessed. It is likely that other tools, such as debuggers, were used but
+are not specified in our sources.
 
@@
 		\begin{itemize}
-			\begin{singlespace}
+
 			\item Continuous Integration Tools
@@
 			\item Correctness Verification Tools
-			\end{singlespace}
+
 		\end{itemize}
@@
 The above tools can verify the correctness of software during its development.
-Only two (ESPResSo, Ludwig) of the software packages that were assessed mentioned using continuous integration tools, like Travis CI.
-Code editors and compilers were explicitly noted to have been used by several packages, and were likely used by all of them.
-One of the packages (Ludwig) explicitly noted the use of proprietary unit testing code written in C.
-Likewise, the use of proprietary code for verifying the correctness of output was noted by one (pyLBM) of the developers.
-It is likely that similar tools were used when developing other software packages.
+Only two (ESPResSo, Ludwig) of the software packages that were assessed
+mentioned using continuous integration tools, like Travis CI. Code editors and
+compilers were explicitly noted to have been used by several packages, and were
+likely used by all of them. One of the packages (Ludwig) explicitly noted the
+use of proprietary unit testing code written in C. Likewise, the use of
+proprietary code for verifying the correctness of output was noted by one
+(pyLBM) of the developers. It is likely that similar tools were used when
+developing other software packages.
 
@@
 
-The following types of dependencies were explicitly noted in the artifacts or web-pages of the 23 LBM packages that were assessed.
-It is possible that other types of dependencies are part of these software packages, but are not clearly specified in their artifacts or web sites and because of that they are not listed here.
+The following types of dependencies were explicitly noted in the artifacts or
+web-pages of the 23 LBM packages that were assessed. It is possible that other
+types of dependencies are part of these software packages, but are not clearly
+specified in their artifacts or web sites and because of that they are not
+listed here.
 
@@
 		\begin{itemize}
-			\begin{singlespace}
+
 			\item Build Automation Tools
@@
 			\item Domain Specific Libraries
-		\end{singlespace}
+
 		\end{itemize}
@@
 
-Most of the software packages use some sort of build automation tools, most commonly Make.
-They also all use various technical and domain specific libraries.
-Technical libraries include visualization (e.g.
-Matplotlib, ParaView, Pygame, VTK), data analysis (e.g.
-Anaconda, Torch), and message passing libraries (e.g.
-MPICH, Open MPI, PyZMQ).
-Domain specific libraries are scientific computing libraries (e.g.
-SciPy).
-Libraries that are not explicitly stated in artifacts, or were not noted during our observations, may fall outside of these categories.
+Most of the software packages use some sort of build automation tools, most
+commonly Make. They also all use various technical and domain specific
+libraries. Technical libraries include visualization (e.g. Matplotlib, ParaView,
+Pygame, VTK), data analysis (e.g. Anaconda, Torch), and message passing
+libraries (e.g. MPICH, Open MPI, PyZMQ). Domain specific libraries are
+scientific computing libraries (e.g. SciPy). Libraries that are not explicitly
+stated in artifacts, or were not noted during our observations, may fall outside
+of these categories.
 
@@
 
-Many of the software packages that were assessed were developed by teams of two or more people.
-Their work needed to be coordinated and managed.
-The following types of project management tools were explicitly noted in the artifacts, web-pages, or interviews with the developers of the 23 LBM packages that were assessed.
-As with development tools and dependencies, it is possible that other types of project management tools were used to coordinate and manage the projects but are not specified and because of that they are not listed here.
+Many of the software packages that were assessed were developed by teams of two
+or more people. Their work needed to be coordinated and managed. The following
+types of project management tools were explicitly noted in the artifacts,
+web-pages, or interviews with the developers of the 23 LBM packages that were
+assessed. As with development tools and dependencies, it is possible that other
+types of project management tools were used to coordinate and manage the
+projects but are not specified and because of that they are not listed here.
 
@@
 		\begin{itemize}
-			\begin{singlespace}
+
 			\item Collaboration Tools
@@
 			\item Document Generation Tools
-		\end{singlespace}
+
 		\end{itemize}
@@
 
-Collaboration tools are noted as being used when developing the software projects.
-Most often email and video conferencing is used.
-Project management software was not explicitly mentioned, but it is possible that some of the projects use such software.
-Many of the projects are located on GitHub, and its developers use the platform to help manage their projects, especially bug related issues.
-Most of the projects appear to use change tracking and version control tools.
-They often use GitHub or GitLab for this.
-One package (SunlightLB) uses CVS.
-Document generation tools are mentioned in the artifacts of 12 of the projects.
-The tools Sphinx and Doxygen are explicitly used in this capacity.
-
-\subsection{Principles, Processes, and Methodologies}\label{prinprocmeth}
-
-This subsection answers the research question: What principles, processes, and methodologies are used in the development of current software packages?
-
-The points and conclusions in this subsection come from developer interviews and reviews of software package artifacts.
-
-Most of the software packages do not explicitly state in their artifacts the motivations or design principles that were considered when developing the software.
-One package, Sailfish, indicates in its artifacts that shortening the development time was considered in early stages of design, with the developers opting for using Python and CUDA/OpenCL to achieve this without sacrificing any computational performance.
-The goals of that project are explicitly listed as performance, scalability, agility and extendability, maintenance, and ease of use.
-The project scored well in these categories during our assessment.
-
-Processes, like methods, are ways of doing things, especially in an orderly way; while methodologies are defined as systems of methods \citep{ghezzi1991fundamentals}.
-It is not explicitly indicated in the artifacts of most of the packages that development involved following any specific model, like a waterfall or agile development model.
-One developer (ESPResSo) noted that while no formal model is used, their development model is something similar to a combination of agile and waterfall development models.
-The developer teams of the LBM packages are fairly small, so it is feasible for them to be organized without the need for such models.
-
-Seven of the software packages contain artifacts outlining the general development process, like basic instructions on how to contribute.
-Eleven of the packages explicitly convey that they would accept outside contributors, but generally the teams are centralized, often working at the same institution.
-
-The developers that were interviewed all noted similar project management processes.
-In teams of only a couple of developers, additions of new features or major changes are discussed with the entire team.
-Projects with more than a couple developers have lead developer roles.
-These lead developers review potential additions to the software.
-One of the developers (ESPResSo) that was interviewed noted that an ad hoc peer review process is used to assess major changes and additions.
-
-Thirteen (57\%) of the 23 software packages use GitHub for managing the project, including nine (64\%) of the 14 alive packages, and four of the nine (44\%) dead packages.
-Two projects (Palabos, WaLBerla) use GitLab.
-This could be indicative of a transition to such software development and version control tools for SCS.
+Collaboration tools are noted as being used when developing the software
+projects. Most often email and video conferencing is used. Project management
+software was not explicitly mentioned, but it is possible that some of the
+projects use such software. Many of the projects are located on GitHub, and its
+developers use the platform to help manage their projects, especially bug
+related issues. Most of the projects appear to use change tracking and version
+control tools. They often use GitHub or GitLab for this. One package
+(SunlightLB) uses CVS. Document generation tools are mentioned in the artifacts
+of 12 of the projects. The tools Sphinx and Doxygen are explicitly used in this
+capacity.
+
+\subsection{Principles, Processes, and Methodologies} \label{prinprocmeth}
+
+This subsection answers the research question: What principles, processes, and
+methodologies are used in the development of current software packages?
+
+The points and conclusions in this subsection come from developer interviews and
+reviews of software package artifacts.
+
+Most of the software packages do not explicitly state in their artifacts the
+motivations or design principles that were considered when developing the
+software. One package, Sailfish, indicates in its artifacts that shortening the
+development time was considered in early stages of design, with the developers
+opting for using Python and CUDA/OpenCL to achieve this without sacrificing any
+computational performance. The goals of that project are explicitly listed as
+performance, scalability, agility and extendability, maintenance, and ease of
+use. The project scored well in these categories during our assessment.
+
+Processes, like methods, are ways of doing things, especially in an orderly way;
+while methodologies are defined as systems of methods
+\citep{ghezzi1991fundamentals}. It is not explicitly indicated in the artifacts
+of most of the packages that development involved following any specific model,
+like a waterfall or agile development model. One developer (ESPResSo) noted that
+while no formal model is used, their development model is something similar to a
+combination of agile and waterfall development models. The developer teams of
+the LBM packages are fairly small, so it is feasible for them to be organized
+without the need for such models.
+
+Seven of the software packages contain artifacts outlining the general
+development process, like basic instructions on how to contribute. Eleven of the
+packages explicitly convey that they would accept outside contributors, but
+generally the teams are centralized, often working at the same institution.
+
+The developers that were interviewed all noted similar project management
+processes. In teams of only a couple of developers, additions of new features or
+major changes are discussed with the entire team. Projects with more than a
+couple developers have lead developer roles. These lead developers review
+potential additions to the software. One of the developers (ESPResSo) that was
+interviewed noted that an ad hoc peer review process is used to assess major
+changes and additions.
+
+Thirteen (57\%) of the 23 software packages use GitHub for managing the project,
+including nine (64\%) of the 14 alive packages, and four of the nine (44\%) dead
+packages. Two projects (Palabos, WaLBerla) use GitLab. This could be indicative
+of a transition to such software development and version control tools for SCS.
 Typically there are several simultaneous development branches in these projects.
 
-Documentation was also noted as playing a significant role in the development process, specifically with on-boarding new developers.
-A goal of documentation is to lower the entry barrier for these new contributors.
-The documentation provides information on how to get started, orients the user to artifacts and the source code, and explains how the system works, including the so-called simulation engine and interface.
-The use of document generation tools is mentioned in the artifacts of 12 software packages, and was noted during interviews with developers.
-Sphinx and Doxygen are the tools that were mentioned.
+Documentation was also noted as playing a significant role in the development
+process, specifically with on-boarding new developers. A goal of documentation
+is to lower the entry barrier for these new contributors. The documentation
+provides information on how to get started, orients the user to artifacts and
+the source code, and explains how the system works, including the so-called
+simulation engine and interface. The use of document generation tools is
+mentioned in the artifacts of 12 software packages, and was noted during
+interviews with developers. Sphinx and Doxygen are the tools that were
+mentioned.
 
@@
 One is feature additions, which arise from a scientific or functional need.
-These changes involve formal discussions within the development team, and lead developer participation is mandatory.
-The other change type is code refactoring, which only sometimes involves formal discussions with the development team.
-New developers were noted to play an increased role in these changes compared to the former changes.
-Software bugs are typically addressed in a similar fashion as code refactoring, and issue tracking is commonly used to manage these changes.
-
-Interviews with the developers of software packages also revealed a more frequent use of both unit testing and continuous integration in the development process than was found by only assessing the artifacts.
-The use of automatic installation processes is also common.
-Most often this involved a Make script.
-
-\subsection{Pain Points}\label{painpoints}
-
-This subsection answers the research question: What are the pain points for developers working on research software projects?
-What aspects of the existing processes, methodologies and tools do they consider as potentially needing improvement?
-How should processes, methodologies and tools be changed to improve software development and software quality?
-
-Developers were asked to comment on obstacles in their development process, obstacles encountered by users, and potential future obstacles.
+These changes involve formal discussions within the development team, and lead
+developer participation is mandatory. The other change type is code refactoring,
+which only sometimes involves formal discussions with the development team. New
+developers were noted to play an increased role in these changes compared to the
+former changes. Software bugs are typically addressed in a similar fashion as
+code refactoring, and issue tracking is commonly used to manage these changes.
+
+Interviews with the developers of software packages also revealed a more
+frequent use of both unit testing and continuous integration in the development
+process than was found by only assessing the artifacts. The use of automatic
+installation processes is also common. Most often this involved a Make script.
+
+\subsection{Pain Points} \label{painpoints}
+
+This subsection answers the research question: What are the pain points for
+developers working on research software projects? What aspects of the existing
+processes, methodologies and tools do they consider as potentially needing
+improvement? How should processes, methodologies and tools be changed to improve
+software development and software quality?
+
+Developers were asked to comment on obstacles in their development process,
+obstacles encountered by users, and potential future obstacles.
 
@@
 
-A developer of pyLBM noted that their small development team has a lack of time to implement new features.
-Small development teams are common for LBM software packages.
-Team members are almost always part of the same institute or already know each other from other projects.
-External contributions are rare despite many of the projects accepting them.
-Aside from on-boarding new developers, time constraints could be mitigated by increasing developer efficiency, which could be addressed in several ways, including by improving the quality of documentation, or incorporating automatic code generation.
+A developer of pyLBM noted that their small development team has a lack of time
+to implement new features. Small development teams are common for LBM software
+packages. Team members are almost always part of the same institute or already
+know each other from other projects. External contributions are rare despite
+many of the projects accepting them. Aside from on-boarding new developers, time
+constraints could be mitigated by increasing developer efficiency, which could
+be addressed in several ways, including by improving the quality of
+documentation, or incorporating automatic code generation.
 
@@
 A lack of software development experience was noted by the developer of TCLB.
-Many of the team members on their project are domain experts and there can be a steep learning curve before these team members contribute good quality source code.
-It was further noted that this has been somewhat addressed, as the code has been re-written to best ensure ease of future contributions.
+Many of the team members on their project are domain experts and there can be a
+steep learning curve before these team members contribute good quality source
+code. It was further noted that this has been somewhat addressed, as the code
+has been re-written to best ensure ease of future contributions.
 
@@
 
-The same developer also noted that there is a lack of incentive and funding in academia for developing widely used scientific software.
-The importance of funding for scientific software has been discussed in \cite{gewaltig2012quality}, which notes that ``software tools are developed and maintained only for as long as there is explicit or implicit funding''.
-The developer further commented that there are no journals that publish such scientific software source code.
-However, there are ways to get such source code cited.
-Work has been done to address this in \citep{smith2016software}, which presents a set of software citation principles and discusses ``how they could be used to implement software citation in the scholarly community'' \citep{katz2019software}.
+The same developer also noted that there is a lack of incentive and funding in
+academia for developing widely used scientific software. The importance of
+funding for scientific software has been discussed in
+\cite{gewaltig2012quality}, which notes that ``software tools are developed and
+maintained only for as long as there is explicit or implicit funding''. The
+developer further commented that there are no journals that publish such
+scientific software source code. However, there are ways to get such source code
+cited. Work has been done to address this in \citep{smith2016software}, which
+presents a set of software citation principles and discusses ``how they could be
+used to implement software citation in the scholarly community''
+\citep{katz2019software}.
 
@@
 
-Another raised concern was that there are no organizations helping with the development of good quality software; but some do exist, including \href{https://bssw.io/}{Better Scientific Software (BSSw)}, \href{https://www.software.ac.uk/}{Software Sustainability Institute}, and \href{https://software-carpentry.org/}{Software Carpentry}.
-Some SCS developers may not be familiar with these organizations.
-
-Scientific software is often developed in-house by the very researchers that temporarily use it in their own research.
-Empirical studies of such ``professional end-user development'' of SCS is noted in \citep{segal2007end}.
-This kind of software has a defined user and purpose, and often does not meet the standards that would be required by external users.
-It has been categorized as a ``private tool'' by \citep{gewaltig2012quality}, which notes that despite often being made freely available, ``it is not always clear that it is sufficiently mature in terms of domain coverage, validity, documentation or usability, to be useful to other researchers''.
-It is less common for such ad hoc scientific software to become ``user-ready software'', which ``is not only research-ready, but should have most of the attributes commonly expected of commercial software products including broadness of scope, robustness, demonstrable correctness and adequate documentation'' \citep{gewaltig2012quality}.
-As software becomes user-ready, it can become commercialized and closed-source.
+Another raised concern was that there are no organizations helping with the
+development of good quality software; but some do exist, including
+\href{https://bssw.io/}{Better Scientific Software (BSSw)},
+\href{https://www.software.ac.uk/}{Software Sustainability Institute}, and
+\href{https://software-carpentry.org/}{Software Carpentry}. Some SCS developers
+may not be familiar with these organizations.
+
+Scientific software is often developed in-house by the very researchers that
+temporarily use it in their own research. Empirical studies of such
+``professional end-user development'' of SCS is noted in \citep{segal2007end}.
+This kind of software has a defined user and purpose, and often does not meet
+the standards that would be required by external users. It has been categorized
+as a ``private tool'' by \citep{gewaltig2012quality}, which notes that despite
+often being made freely available, ``it is not always clear that it is
+sufficiently mature in terms of domain coverage, validity, documentation or
+usability, to be useful to other researchers''. It is less common for such ad
+hoc scientific software to become ``user-ready software'', which ``is not only
+research-ready, but should have most of the attributes commonly expected of
+commercial software products including broadness of scope, robustness,
+demonstrable correctness and adequate documentation''
+\citep{gewaltig2012quality}. As software becomes user-ready, it can become
+commercialized and closed-source.
 
@@
 
-Setting up parallelization was also noted as a technical pain point by one of the developers, and the introduction of continuous integration by another.
-Software development knowledge, and automatic code generation, could mitigate such pain points.
-As already noted, many of the developers are domain experts and not professional software developers.
-The developer of TCLB noted that eliminating equivalent statements using macros had helped improve the quality of their source code, specifically helping with reusing code to run on both the CPU and GPU.
+Setting up parallelization was also noted as a technical pain point by one of
+the developers, and the introduction of continuous integration by another.
+Software development knowledge, and automatic code generation, could mitigate
+such pain points. As already noted, many of the developers are domain experts
+and not professional software developers. The developer of TCLB noted that
+eliminating equivalent statements using macros had helped improve the quality of
+their source code, specifically helping with reusing code to run on both the CPU
+and GPU.
 
@@
 Difficulties with ensuring correctness were also noted by several developers.
-They indicated that tests are run on all new source code additions, testing both individual modules and the system to verify correctness.
-These tests compare the package output to known correct output using test cases.
-The developer of TCLB commented that the amount of testing data that is needed for some cases is a problem as free testing services do not offer capabilities to store and process such large amounts of data, and in-house testing solutions needed to be created to address this limitation.
-The solution for this has been to limit the size of the testing problems, and to run tests in small batches with few iterations.
+They indicated that tests are run on all new source code additions, testing both
+individual modules and the system to verify correctness. These tests compare the
+package output to known correct output using test cases. The developer of TCLB
+commented that the amount of testing data that is needed for some cases is a
+problem as free testing services do not offer capabilities to store and process
+such large amounts of data, and in-house testing solutions needed to be created
+to address this limitation. The solution for this has been to limit the size of
+the testing problems, and to run tests in small batches with few iterations.
 
@@
 
-A few obstacles related to users were found.
-Several developers noted that users sometimes try to use incorrect LBM method combinations to solve their problems.
-Furthermore, some users think that the packages will work out of the box to solve their cases, while the packages require both a good understanding of CFD and an understanding of the requirements for formulating problems in the individual packages, which can be a significant endeavor.
-These software packages are not like commercial software packages.
-They are generally set up to solve specific research problems, and are often primarily used by their developers.
-While they are modifiable to solve similar problems, these modifications are not trivial.
-Better documentation, with attention on traceability, and automatic code generation are suggested when designing software for change, and would help with these modifications.
-So far this problem of on-boarding new users has been addressed by updating the documentation to better inform users of the underlying LBM theory and package requirements.
-Similar issues with LBM parameters were noted by another developer.
-Updating the user interface to better explain theoretical principles, as well as test user input for compatibility, was the implemented solution.
-As noted above, sometimes frequently asked questions on the underlying theory and on how to use the software are answered in the documentation.
-
-The interviewed developer of ESPResSo commented that parts of their package's source code had been refactored to Python to help address usability issues.
-Python was perceived as a much more usable language, and it would be easy for future users and developers to learn and understand the source code.
+A few obstacles related to users were found. Several developers noted that users
+sometimes try to use incorrect LBM method combinations to solve their problems.
+Furthermore, some users think that the packages will work out of the box to
+solve their cases, while the packages require both a good understanding of CFD
+and an understanding of the requirements for formulating problems in the
+individual packages, which can be a significant endeavor. These software
+packages are not like commercial software packages. They are generally set up to
+solve specific research problems, and are often primarily used by their
+developers. While they are modifiable to solve similar problems, these
+modifications are not trivial. Better documentation, with attention on
+traceability, and automatic code generation are suggested when designing
+software for change, and would help with these modifications. So far this
+problem of on-boarding new users has been addressed by updating the
+documentation to better inform users of the underlying LBM theory and package
+requirements. Similar issues with LBM parameters were noted by another
+developer. Updating the user interface to better explain theoretical principles,
+as well as test user input for compatibility, was the implemented solution. As
+noted above, sometimes frequently asked questions on the underlying theory and
+on how to use the software are answered in the documentation.
+
+The interviewed developer of ESPResSo commented that parts of their package's
+source code had been refactored to Python to help address usability issues.
+Python was perceived as a much more usable language, and it would be easy for
+future users and developers to learn and understand the source code.
 
@@
 
-A few potential future obstacles were noted.
-The developer of ESPResSo noted that their source code had been written with a specific application in mind and that due to this there was too much coupling between components of the source code.
-This results in technical debt, having an impact on future modifiability and reusability when trying to extend the software, and the code would need to be refactored.
-
-As noted above, difficulties with ensuring future correctness could also arise. As new methods and functionality is added into the software, new test cases and test data will need to be developed.
+A few potential future obstacles were noted. The developer of ESPResSo noted
+that their source code had been written with a specific application in mind and
+that due to this there was too much coupling between components of the source
+code. This results in technical debt, having an impact on future modifiability
+and reusability when trying to extend the software, and the code would need to
+be refactored.
+
+As noted above, difficulties with ensuring future correctness could also arise.
+As new methods and functionality is added into the software, new test cases and
+test data will need to be developed.
 
@@
 
-Interviewees commented that documentation is important and that its quality could be improved.
-As already noted, there is often no time or funding for maintaining quality documentation for software that is rarely used outside of the development team.
-Furthermore, the documentation generally only provides a shallow overview of the underlying CFD theory.
-Users would be well advised to already be familiar with these topics, or they should spend significant time referencing theory resources.
-The documentation instead generally focuses on explaining how to use the software.
-It is of course not feasible for package documentation to address the underlying physics topics in detail, so it is advised that the package documentation links to resources that better explain the underlying theory.
-Sometimes, frequently asked questions about the underlying theory are answered in the documentation.
-OpenLB has an artifact for such questions.
-
-The developer comments emphasized an importance on source code, while documentation seems to be of secondary importance.
-It must be stressed that improving documentation could benefit development and help eliminate some of the developer concerns that were raised.
-The use of automatic document generation tools that capture scientific and computing knowledge, and transform it into software artifacts, is advised.
-Drasil is an automatic document generation tool that is further discussed in Section \ref{highlightedrecommendations}.
-
-\subsection{Quality Recommendations}\label{qualityrecommentations}
-
-This subsection answers the research question: For research software developers, what specific actions are taken to address software qualities?
-
-The following points regarding software quality should be considered when developing LBM software packages.
-These points are based on developer interviews, SCS literature, what was found to have worked for packages that were designated as high quality in this assessment.
+Interviewees commented that documentation is important and that its quality
+could be improved. As already noted, there is often no time or funding for
+maintaining quality documentation for software that is rarely used outside of
+the development team. Furthermore, the documentation generally only provides a
+shallow overview of the underlying CFD theory. Users would be well advised to
+already be familiar with these topics, or they should spend significant time
+referencing theory resources. The documentation instead generally focuses on
+explaining how to use the software. It is of course not feasible for package
+documentation to address the underlying physics topics in detail, so it is
+advised that the package documentation links to resources that better explain
+the underlying theory. Sometimes, frequently asked questions about the
+underlying theory are answered in the documentation. OpenLB has an artifact for
+such questions.
+
+The developer comments emphasized an importance on source code, while
+documentation seems to be of secondary importance. It must be stressed that
+improving documentation could benefit development and help eliminate some of the
+developer concerns that were raised. The use of automatic document generation
+tools that capture scientific and computing knowledge, and transform it into
+software artifacts, is advised. Drasil is an automatic document generation tool
+that is further discussed in Section~\ref{highlightedrecommendations}.
+
+\subsection{Quality Recommendations} \label{qualityrecommentations}
+
+This subsection answers the research question: For research software developers,
+what specific actions are taken to address software qualities?
+
+The following points regarding software quality should be considered when
+developing LBM software packages. These points are based on developer
+interviews, SCS literature, what was found to have worked for packages that were
+designated as high quality in this assessment.  Our recommendations are not
+lists of what should have been done in the past, or what should be done now;
+they are just suggestions for consideration in the future.
 
@@
 \begin{itemize}
-	\item Include OS compatibility, including specific OS versions. This was done by several top ranked packages, including ESPResSo and LUMA.
+	\item Include OS compatibility, including specific OS versions. This was
+	done by several top ranked packages, including ESPResSo and LUMA.
 	\item Provide complete installation instructions. This was done by all of the top five ranked packages.
-	\item Installation instructions should be written as if the user does not have any dependencies installed and is installing on a clean OS.
-This was done by several top ranked packages, including ESPResSo, OpenLB, and Palabos.
-This can be tested on a clean environment using a virtual machine.
-	\item List all dependencies in the installation instructions, dependency versions, and how to install them.
-This was done by several top ranked packages, including ESPResSo, LUMA, OpenLB, and Palabos.
-	\item If possible, automate the installation of dependencies. Use tools such as Make. This was done by all of the top five ranked packages.
-	\item Automate the installation process as much as possible. Use tools such as Make. This was done by all of the top five ranked packages.
-	\item The installation instruction should only be in one location. This was done by several top ranked packages, including ESPResSo, OpenLB, and Palabos.
-	\item Include descriptive error messages for errors encountered during installation. This was done by several top ranked packages, including Ludwig and LUMA
-	\item Provide a way to validate the installation.
-This can be done using a custom script, or a test case that specifies expected output.
-This was done by several top ranked packages, including OpenLB, Ludwig, LUMA, and Palabos.
+	\item Installation instructions should be written as if the user does not
+	have any dependencies installed and is installing on a clean OS. This was
+	done by several top ranked packages, including ESPResSo, OpenLB, and
+	Palabos. This can be tested on a clean environment using a virtual machine.
+	\item List all dependencies in the installation instructions, dependency
+	versions, and how to install them. This was done by several top ranked
+	packages, including ESPResSo, LUMA, OpenLB, and Palabos.
+	\item If possible, automate the installation of dependencies. Use tools such
+	as Make. This was done by all of the top five ranked packages.
+	\item Automate the installation process as much as possible. Use tools such
+	as Make. This was done by all of the top five ranked packages.
+	\item The installation instruction should only be in one location. This was
+	done by several top ranked packages, including ESPResSo, OpenLB, and
+	Palabos.
+	\item Include descriptive error messages for errors encountered during
+	installation. This was done by several top ranked packages, including Ludwig
+	and LUMA
+	\item Provide a way to validate the installation. This can be done using a
+	custom script, or a test case that specifies expected output. This was done
+	by several top ranked packages, including OpenLB, Ludwig, LUMA, and Palabos.
 	\item Provide instructions for uninstalling the software. These were included with pyLBM.
@@
 \begin{itemize}
-	\item Use a requirements specification document.
-This is suggested in SCS literature, and several of the top ranked packages had such a document or reference to theory manuals.
-A potential template is presented in \citep{smith2005new}.
-	\item Make public (on package website or GitHub) the requirements specification document, or explicitly reference the domain theory that the software is designed from.
-This was done by several top ranked packages, including ESPResSo, Ludwig, LUMA, and OpenLB.
+	\item Use a requirements specification document. This is suggested in SCS
+	literature, and several of the top ranked packages had such a document or
+	reference to theory manuals. A potential template is presented in
+	\citep{smith2005new}. 
+	\item Make public (on package website or GitHub) the requirements
+	specification document, or explicitly reference the domain theory that the
+	software is designed from.  This was done by several top ranked packages,
+	including ESPResSo, Ludwig, LUMA, and OpenLB.
 	\item Ensure the above information is easy to find. Consider adding it to the user manual. The information is easy to find in most top ranked packages.
@@
 	\item Keep artifacts updated. This is suggested as good software maintenance practice.
-	\item Ideally include most of the common and less common artifacts listed in Section \ref{artifacts}.
+	\item Ideally include most of the common and less common artifacts listed in Section~\ref{artifacts}.
 	\item Include version numbers and release notes for all major source code and artifact releases. The top five ranked software packages include release notes.
@@
 	\item Source code needs to be well commented.
-Typically, more than 10 percent of LBM package source code is comments, as presented in Table \ref{gitrepodata}.
+Typically, more than 10 percent of LBM package source code is comments, as presented in Table~\ref{gitrepodata}.
 Four of the top five overall ranked packages (Ludwig, ESPResSo, Palabos, OpenLB) have about 20 percent of their source code as comments.
@@
 
-\subsection{Designation Comparison}\label{comparison}
-
-This subsection answers the research question: How does software designated as high quality by this methodology compare with top rated software by the community?
-
-This comparison helps evaluate the methodology that was used, and assess the validity of the findings.
-Threats to the validity of this exercise are noted in Section \ref{threats}.
-The software package designations of this report are compared to package repository ranking metrics in Section \ref{repmetrics}, and to software recommended by a domain expert in Section \ref{domainexpertrecommend}.
-
-\subsubsection{Repository Ranking Metrics}\label{repmetrics}
-
-Table \ref{repometrics} presents our LBM software package rankings along with the repository ranking metrics of each software package.
-
-Eight packages do not use GitHub or GitLab, so they do not have a measure of repository stars.
-Looking at the repository stars of the other 15 packages, we can observe a slight pattern where packages that have been highly ranked by our assessment do have more stars than lower ranked packages.
-The second ranked package (ESPResSo) has the second most number of stars.
-The ninth ranked package (Sailfish) has the most number of stars.
-Packages designated as lower quality often do not use GitHub or GitLab, or have few stars, except for HemeLB which has 22 stars.
-Our assessment of this package might not be accurate.
-The number of stars does not necessarily represent the perceptions of the community, but for lack of an alternative measure we will use it this way.
-
-The repository watches column contains even less data to compare since two of the packages (Palabos, waLBerla) use GitLab, which does not include this metric.
-The pattern that can be observed with this metric is very similar to that of the stars metric.
-The ninth ranked package (Sailfish) has the most number of watches.
-The second ranked package (ESPResSo) has the second most number of watches.
-
-Besides missing data, another threat to the validity of this comparison is the varying ages of the repositories.
-Older packages have been able to accumulate stars and watches for longer than newer packages.
-The true quality of new packages may not be reflected in their stars and watches.
-\\
-
+\subsection{Designation Comparison} \label{comparison}
+
+This subsection answers the research question: How does software designated as
+high quality by this methodology compare with top rated software by the
+community?
+
+This comparison helps evaluate the methodology that was used, and assess the
+validity of the findings. Threats to the validity of this exercise are noted in
+Section~\ref{threats}. The software package designations of this report are
+compared to package repository ranking metrics in Section~\ref{repmetrics}, and
+to software recommended by a domain expert in Section
+~\ref{domainexpertrecommend}.
+
+\subsubsection{Repository Ranking Metrics} \label{repmetrics}
+
+Table~\ref{repometrics} presents our LBM software package rankings along with
+the repository ranking metrics of each software package.
+
+Eight packages do not use GitHub or GitLab, so they do not have a measure of
+repository stars. Looking at the repository stars of the other 15 packages, we
+can observe a slight pattern where packages that have been highly ranked by our
+assessment do have more stars than lower ranked packages. The second ranked
+package (ESPResSo) has the second most number of stars. The ninth ranked package
+(Sailfish) has the most number of stars. Packages designated as lower quality
+often do not use GitHub or GitLab, or have few stars, except for HemeLB which
+has 22 stars. Our assessment of this package might not be accurate. The number
+of stars does not necessarily represent the perceptions of the community, but
+for lack of an alternative measure we will use it this way.
+
+The repository watches column contains even less data to compare since two of
+the packages (Palabos, waLBerla) use GitLab, which does not include this metric.
+The pattern that can be observed with this metric is very similar to that of the
+stars metric. The ninth ranked package (Sailfish) has the most number of
+watches. The second ranked package (ESPResSo) has the second most number of
+watches.
+
+Besides missing data, another threat to the validity of this comparison is the
+varying ages of the repositories. Older packages have been able to accumulate
+stars and watches for longer than newer packages. The true quality of new
+packages may not be reflected in their stars and watches. \\
+
+\begin{table}
 \begin{center}
-	\begin{onehalfspacing}
 		\begin{tabular}{ p{3.5cm}p{2.5cm}p{2cm}p{2cm}p{2cm} }
@@
 		\end{tabular}
-		\captionof{table}{Repository Ranking Metrics}\label{repometrics}
-	\end{onehalfspacing}
+		\caption{Repository Ranking Metrics} \label{repometrics}
 \end{center}
-
-\subsubsection{Domain Expert Recommended Software}\label{domainexpertrecommend}
-
-Table \ref{rankingcomparison} compares the top 10 ranked LBM software packages with a LBM package ranking made by a domain expert.
-Five (Palabos, OpenLB, LUMA, waLBerla, Sailfish) of the top 10 ranked packages in this assessment are also found in the domain expert's list.
-Interestingly, these five packages are listed in the same order on both lists.
-Looking at the remaining packages, the top two ranked packages in this state of the practice assessment (Ludwig, ESPResSo) are not on the domain expert's list.
-Perhaps the intended applications of these packages, complex fluids for Ludwig and soft matter research for ESPResSo, do not align with the research interests of the domain expert, and this is why they did not make it onto their list.
-Moving down the list, pyLBM and DL\_MESO(LBE) did not make the domain expert's top 10 list but were mentioned by the domain expert.
-The 10th ranked package, laboetie, was not mentioned by the domain expert.
-Looking at the remainder of the domain expert's list, ASL is a general purpose tool for solving partial differential equations and may have fallen out of scope of the authoritative lists that were used to identify the initial LBM software list.
-The domain expert's fifth ranked package (ch4-project) was on the initial software list of this assessment, but was removed due to a lack of documentation.
-Open FSI is a new project that uses Palabos.
-It was made public the year before data was collected for this state of the practice exercise.
-It may not have been on authoritative lists due to its young age.
-Similarly, LIFE was also made public recently.
-Finally, the domain expert's 10th ranked package (HemeLB) was ranked 20th in this state of the practice assessment.
-
+\end{table}
+
+\subsubsection{Domain Expert Recommended Software} \label{domainexpertrecommend}
+
+Table~\ref{rankingcomparison} compares the top 10 ranked LBM software packages
+with a LBM package ranking made by a domain expert. Five (Palabos, OpenLB, LUMA,
+waLBerla, Sailfish) of the top 10 ranked packages in this assessment are also
+found in the domain expert's list. Interestingly, these five packages are listed
+in the same order on both lists. Looking at the remaining packages, the top two
+ranked packages in this state of the practice assessment (Ludwig, ESPResSo) are
+not on the domain expert's list. Perhaps the intended applications of these
+packages, complex fluids for Ludwig and soft matter research for ESPResSo, do
+not align with the research interests of the domain expert, and this is why they
+did not make it onto their list. Moving down the list, pyLBM and DL\_MESO(LBE)
+did not make the domain expert's top 10 list but were mentioned by the domain
+expert. The 10th ranked package, laboetie, was not mentioned by the domain
+expert. Looking at the remainder of the domain expert's list, ASL is a general
+purpose tool for solving partial differential equations and may have fallen out
+of scope of the authoritative lists that were used to identify the initial LBM
+software list. The domain expert's fifth ranked package (ch4-project) was on the
+initial software list of this assessment, but was removed due to a lack of
+documentation. Open FSI is a new project that uses Palabos. It was made public
+the year before data was collected for this state of the practice exercise. It
+may not have been on authoritative lists due to its young age. Similarly, LIFE
+was also made public recently. Finally, the domain expert's 10th ranked package
+(HemeLB) was ranked 20th in this state of the practice assessment.
+
+\begin{table}
 \begin{center}
-	\begin{onehalfspacing}
 		\begin{tabular}{ p{2cm}p{4.5cm}p{3.5cm}}
@@
 		\end{tabular}
-		\captionof{table}{Ranking Comparison}\label{rankingcomparison}
-	\end{onehalfspacing}
+		\caption{Ranking Comparison} \label{rankingcomparison}
 \end{center}
-
-\subsection{Threats To Validity}\label{threats}
-
-This section examines potential threats to the validity of this state of the practice assessment. These can be categorized into methodology and data collection issues.
-
-The goal of this assessment isn't to rank the software, but to use the ranking exercise as a means to understand the state of the practice of LBM software development.
-
-The measures listed in our measurement template may not be broad enough to accurately capture some qualities.
-For example, there are only two measures of surface robustness.
-The measurement of robustness could be expanded, as it currently only measures unexpected input.
-Other faults could be introduced, but could require a large investment of time to develop, and might not be a fair measure for all packages.
-Similarly, reusability is assessed along the number of code files and LOC per file.
-While this measure is indicative of modularity, it is possible that some packages have many files, with few LOC, but the files do not contain source code that is easily reusable.
-The files may be poorly formatted, or the source code may be vague and have ambiguous identifiers.
-Furthermore, the measurement of understandability relies on 10 random source code files.
-It is possible that the 10 files that were chosen to represent a software package may not be a good representation of the understandability of that package.
-
-Regarding data collection, a risk to the validity of this assessment is missing or incorrect data.
-Some software package data may not have been measured due to technology issues like broken links.
-This issue arose with the measurement of Palabos, which had a broken link to its user manual, as noted in Section \ref{lesscommon}.
-
-Some pertinent data may not have been specified in public artifacts, or may be obscure within an artifact or web-page.
-The use of unit testing and continuous integration was mentioned in the artifacts of only two (ESPResSo, Ludwig) packages.
-However, interviews suggested a more frequent use of both unit testing and continuous integration in the development processes than what was observed from the initial survey of the artifacts.
-For example, OpenLB, pyLBM, and TCLB use such methods during development despite this not being explicitly clear from an analysis of the material available online.
-
-Furthermore, design documentation was measured to be a ``less common'' artifact in this assessment, but it is probable that such documentation is part of all LBM packages.
-After all, developing SCS is not a trivial endeavor.
-It is likely that many packages have such documentation but did not make it public, and due to this the measured data is not a true reflection of software package quality.
-
-\newpage
-\section{Conclusion}\label{conclusion}
-
-We analyzed the state of the practice of software development in the Lattice Boltzmann Methods software domain by quantitatively and qualitatively measuring, and comparing, 23 software packages along quality attributes.
-The software qualities that were assessed in this report are listed in Section \ref{softwarequalities}.
-A methodology for assessing the state of the practice of software development in SCS domains was presented.
-Domain packages were assessed to answer the software development related research questions listed in Section \ref{purpose} to understand how software quality is impacted by software development choices, including principles, processes, and tools.
-Software developers were interviewed to identify development pain points, and to identify how software quality is ensured.
-Quantitative data was used to rank the software packages using the AHP.
-The ranking designations were compared with rankings from the software development community, and we found that many of our top 10 ranked packages are ranked highly by a domain expert.
-Recommendations for improving software along quality metrics were made, and highlights are presented in Section \ref{highlightedrecommendations} of this conclusion.
-The findings of this report can be used to guide future development of SCS, specifically along quality attributes, and to reduce software quality failures.
-
-We understand that software packages vary in goals, developers, funding, and other aspects.
-Our goal was to highlight software quality successes through an evaluation of the entire domain software family.
-Furthermore, threats to the validity of the findings are highlighted in Section \ref{threats}.
-
-Recommendations for future state of the practice assessments are made in Section \ref{futuresop}.
-
-\subsection{Highlighted Recommendations}\label{highlightedrecommendations}
-
-The following recommendations improve software quality along multiple attributes, and provide the greatest return on investment:
+\end{table}
+
+\subsection{Threats To Validity} \label{threats}
+
+This section examines potential threats to the validity of this state of the
+practice assessment. These can be categorized into methodology and data
+collection issues.
+
+The goal of this assessment isn't to rank the software, but to use the ranking
+exercise as a means to understand the state of the practice of LBM software
+development.
+
+The measures listed in our measurement template may not be broad enough to
+accurately capture some qualities. For example, there are only two measures of
+surface robustness. The measurement of robustness could be expanded, as it
+currently only measures unexpected input. Other faults could be introduced, but
+could require a large investment of time to develop, and might not be a fair
+measure for all packages. Similarly, reusability is assessed along the number of
+code files and LOC per file. While this measure is indicative of modularity, it
+is possible that some packages have many files, with few LOC, but the files do
+not contain source code that is easily reusable. The files may be poorly
+formatted, or the source code may be vague and have ambiguous identifiers.
+Furthermore, the measurement of understandability relies on 10 random source
+code files. It is possible that the 10 files that were chosen to represent a
+software package may not be a good representation of the understandability of
+that package.
+
+Regarding data collection, a risk to the validity of this assessment is missing
+or incorrect data. Some software package data may not have been measured due to
+technology issues like broken links. This issue arose with the measurement of
+Palabos, which had a broken link to its user manual, as noted in Section
+~\ref{lesscommon}.
+
+Some pertinent data may not have been specified in public artifacts, or may be
+obscure within an artifact or web-page. The use of unit testing and continuous
+integration was mentioned in the artifacts of only two (ESPResSo, Ludwig)
+packages. However, interviews suggested a more frequent use of both unit testing
+and continuous integration in the development processes than what was observed
+from the initial survey of the artifacts. For example, OpenLB, pyLBM, and TCLB
+use such methods during development despite this not being explicitly clear from
+an analysis of the material available online.
+
+Furthermore, design documentation was measured to be a ``less common'' artifact
+in this assessment, but it is probable that such documentation is part of all
+LBM packages. After all, developing SCS is not a trivial endeavor. It is likely
+that many packages have such documentation but did not make it public, and due
+to this the measured data is not a true reflection of software package quality.
+
+\section{Conclusion} \label{conclusion}
+
+We analyzed the state of the practice of software development in the Lattice
+Boltzmann Methods software domain by quantitatively and qualitatively measuring,
+and comparing, 23 software packages along quality attributes. The software
+qualities that were assessed in this report are listed in Section
+~\ref{softwarequalities}. A methodology for assessing the state of the practice
+of software development in SCS domains was presented. Domain packages were
+assessed to answer the software development related research questions listed in
+Section~\ref{purpose} to understand how software quality is impacted by software
+development choices, including principles, processes, and tools. Software
+developers were interviewed to identify development pain points, and to identify
+how software quality is ensured. Quantitative data was used to rank the software
+packages using the AHP. The ranking designations were compared with rankings
+from the software development community, and we found that many of our top 10
+ranked packages are ranked highly by a domain expert. Recommendations for
+improving software along quality metrics were made, and highlights are presented
+in Section~\ref{highlightedrecommendations} of this conclusion. The findings of
+this report can be used to guide future development of SCS, specifically along
+quality attributes, and to reduce software quality failures.
+
+We understand that software packages vary in goals, developers, funding, and
+other aspects. Our goal was to highlight software quality successes through an
+evaluation of the entire domain software family. Furthermore, threats to the
+validity of the findings are highlighted in Section~\ref{threats}.
+
+Recommendations for future state of the practice assessments are made in Section
+~\ref{futuresop}.
+
+\subsection{Highlighted Recommendations} \label{highlightedrecommendations}
+
+The following recommendations improve software quality along multiple
+attributes, and provide the greatest return on investment:
 
@@
 
-Ensuring software quality can take significant time.
-An inability to develop high quality documentation due to time constraints could be mitigated in the future by the use of automatic document generation tools like Drasil.
-The Drasil Framework consists of a collection of Domain Specific Languages (DSL) for capturing scientific documents, structures, and computing knowledge, and then transforming this knowledge into relevant software artifacts without having to manually duplicate knowledge into multiple artifacts \citep{zhao2018}.
-
-\subsection{Future State Of The Practice Assessments}\label{futuresop}
+Ensuring software quality can take significant time. An inability to develop
+high quality documentation due to time constraints could be mitigated in the
+future by the use of automatic document generation tools like Drasil. The Drasil
+Framework consists of a collection of Domain Specific Languages (DSL) for
+capturing scientific documents, structures, and computing knowledge, and then
+transforming this knowledge into relevant software artifacts without having to
+manually duplicate knowledge into multiple artifacts \citep{zhao2018}.
+
+\subsection{Future State Of The Practice Assessments} \label{futuresop}
 
@@
 
-As mentioned in Section \ref{threats}, the measures listed in our measurement template may not be broad enough to accurately capture some qualities.
-Adding or extending measures is worth considering.
-For example, usability experiments and performance benchmarks could be incorporated into the assessment.
-
-Section \ref{interviewresults} noted that developers were not asked to comment on the reproducibility of their source code from their requirements specifications and design documentation.
-Adding this question to the interview guide should be considered in the next iteration of state of the practice assessments.
-Furthermore, as mentioned in Section \ref{threats}, it was found that some pertinent information was not specified in public artifacts.
-The use of unit testing and continuous integration by several packages (OpenLB, pyLBM, TCLB) was only discovered during interviews with developers.
-Adding further questions to the interview guide regarding the measures that are on the measurement template could reduce instances of incorrect data being collected.
-This additional interview data could be analyzed and incorporated into the AHP ranking, ensuring quality designations more accurate represent the true quality of the software packages.
-
-\newpage
-\begin{appendices}
-
-\begin{singlespace}
-
-\section{Measurement Template}\label{measurementtemplate2}
-The table below lists the set of measures that are used to assess each software product.
-The first set identifies summary information, followed by 9 sets for software qualities and 3 sets for raw metrics.
-Each measure is followed by the type for a valid result.
-A superscript ∗ indicate that a response of this type needs to be accompanied by explanatory text.\\
-\newpage
-\captionof{table}{Measurement Template}\label{measurementtemplate}
-\def\arraystretch{1.22}
-\begin{tabular}{p{14cm}}
-	\hline
-	\textbf{Summary Information}\\
-	\hline
-	Software name? (string)\\
-	URL? (URL)\\
-	Affiliation (institution(s)) (string or {N/A})\\
-	Software purpose (string)\\
-	Number of developers (all developers that have contributed at least one commit to the project) (use repo commit logs) (number)\\
-	How is the project funded? (unfunded, unclear, funded*) where * requires a string to say the source of funding\\
-	Initial release date? (date)\\
-	Last commit date? (date)\\
-	Status? (alive is defined as presence of commits in the last 18 months) ({alive, dead, unclear})\\
-	License? ({GNU GPL, BSD, MIT, terms of use, trial, none, unclear, other*}) * given via a string \\
-	Platforms? (set of {Windows, Linux, OS X, Android, other*}) * given via string\\
-	Software Category?
-The concept category includes software that does not have an officially released version.
-Public software has a released version in the public domain.
-Private software has a released version available to authorized users only.
-({concept, public, private})\\
-	Development model? ({open source, freeware, commercial, unclear})\\
-	Publications about the software? Refers to publications that have used or mentioned the software. (number or {unknown})\\
-	Source code URL? ({set of url, n/a, unclear})\\
-	Programming language(s)? (set of {FORTRAN, Matlab, C, C++, Java, R, Ruby, Python, Cython, BASIC, Pascal, IDL, unclear, other*}) * given via string \\
-	Is there evidence that performance was considered? Performance refers to either speed, storage, or throughput. ({yes∗, no})\\
-	Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have) \\
-	\hline
-\end{tabular}
-
-\def\arraystretch{1.5}
-\begin{tabular}{p{14cm}}
-	\hline		
-	\textbf{Installability  (Measured via installation on a virtual machine.) }\\
-	\hline
-	Are there installation instructions? ({yes, no})\\
-	Are the installation instructions in one place? Place referring to a single document or web-page. ({yes, no, n/a})\\
-	Are the installation instructions linear? Linear meaning progressing  in a single series of steps. ({yes, no, n/a})\\
-	Are the instructions written as if the person doing the installation has none of the dependent packages installed? ({yes, no, unclear})\\
-	Are compatible operating system versions listed? ({yes, no})\\
-	Is there something in place to automate the installation (makefile, script, installer, etc)? ({yes*, no})\\
-	If the software installation broke, was a descriptive error message displayed? ({yes, no, n/a})\\
-	Is there a specified way to validate the installation? ({yes*, no})\\
-	How many steps were involved in the installation? (Includes manual steps like unzipping files) Specify OS. (number, OS)\\
-	What OS was used for the installation? ({Windows, Linux, OS X, Android, other* }) *given via string\\
-	How many extra software packages need to be installed before or during installation? (number)\\
-	Are required package versions listed? ({yes, no, n/a})\\
-	Are there instructions for the installation of required packages / dependencies? ({yes, no, n/a})\\
-	Run uninstall, if available. Were any obvious problems caused? ({yes∗ , no, unavail})\\
-	Overall impression? ({1 .. 10})\\
-	Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have)\\
-	\hline
-\end{tabular}
-
-\def\arraystretch{1.33}
-\begin{tabular}{p{14cm}}
-	\hline		
-	\textbf{Correctness and Verifiability}\\
-	\hline
-	Any reference to the requirements specifications of the program or theory manuals? ({yes∗ , no, unclear})\\
-	What tools or techniques are used to build confidence of correctness?
-({literate programming, automated testing, symbolic execution, model checking, assertions used in the code, Sphinx, Doxygen, Javadoc, confluence, unclear, other*}) * given via string\\
-	If there is a getting started tutorial? ({yes, no})\\
-	Are the tutorial instructions linear? ({yes, no, n/a})\\
-	Does the getting started tutorial provide an expected output? ({yes, no*, n/a})\\
-	Does your tutorial output match the expected output? ({yes, no, n/a})\\
-	Are unit tests available?  ({yes, no, unclear})\\
-	Is there evidence of continuous integration? (for example mentioned in documentation, Jenkins, Travis CI, Bamboo, other) ({yes*, no, unclear})\\
-	Overall impression? ({1 .. 10})\\
-	Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have) \\
-	\hline	
-	\textbf{Surface Reliability}\\
-	\hline
-	Did the software “break” during installation? ({yes∗ , no})\\
-	If the software installation broke, was the installation instance recoverable? ({yes, no, n/a})\\
-	Did the software “break” during the initial tutorial testing? ({yes∗, no, n/a})\\
-	If the tutorial testing broke, was a descriptive error message displayed? ({yes, no, n/a})\\
-	If the tutorial testing broke, was the tutorial testing instance recoverable? ({yes, no, n/a})\\
-	Overall impression? ({1 .. 10})\\
-	Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have)\\
-	\hline		
-\end{tabular}
-
-\def\arraystretch{1.4}
-\begin{tabular}{p{14cm}}
-	\hline		
-	\textbf{Surface Robustness}\\
-	\hline
-	Does the software handle unexpected/unanticipated input (like data of the wrong type, empty input, missing files or links) reasonably?
-(a reasonable response can include an appropriate error message.) ({yes, no∗ })\\
-	For any plain text input files, if all new lines are replaced with new lines and carriage returns, will the software handle this gracefully? ({yes, no∗, n/a})\\
-	Overall impression? ({1 .. 10})\\
-	Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have)\\
-	\hline		
-	\textbf{Surface Usability}\\
-	\hline
-	Is there a getting started tutorial? ({yes, no})\\
-	Is there a user manual? ({yes, no})\\
-	Are expected user characteristics documented? ({yes, no})\\
-	What is the user support model? FAQ? User forum? E-mail address to direct questions? Etc. (string)\\
-	Overall impression? ({1 .. 10})\\
-	Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have)\\
-	\hline
-\end{tabular}
-
-\def\arraystretch{1.4}
-\begin{tabular}{p{14cm}}
-		\hline	
-	\textbf{Maintainability}\\
-	\hline
-	What is the current version number? (number)\\
-	Is there any information on how code is reviewed, or how to contribute? ({yes*, no})\\
-	Are artifacts available?
-(List every type of file that is not a code file – for examples please look at the ‘Artifact Name’ column of https://gitlab.cas.mcmaster.ca/SEforSC/se4sc/-/blob/git-svn/GradStudents/Olu/ResearchProposal/Artifacts\_MiningV3.xlsx) ({yes*, no, unclear}) *list via string\\
-	What issue tracking tool is employed? (set of {Trac, JIRA, Redmine, e-mail, discussion board, sourceforge, google code, git, BitBucket, none, unclear, other*}) * given via string\\
-	What is the percentage of identified issues that are closed? (percentage)\\
-	What percentage of code is comments? (percentage)\\
-	Which version control system is in use? ({svn, cvs, git, github, unclear, other*}) * given via string\\
-	Overall impression? ({1 .. 10})\\
-	Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have)\\
-	\hline		
-	\textbf{Reusability}\\
-	\hline
-	How many code files are there? (number)\\
-	Is API documented? ({yes, no, n/a})\\
-	Overall impression? ({1 .. 10})\\
-	Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have)\\
-	\hline		
-\end{tabular}
-
-\def\arraystretch{1.4}
-\begin{tabular}{p{14cm}}
-		\hline		
-	\textbf{Surface Understandability (Based on 10 random source files)}\\
-	\hline
-	Consistent indentation and formatting style? ({yes, no, n/a})\\
-	Explicit identification of a coding standard? ({yes∗, no, n/a})\\
-	Are the code identifiers consistent, distinctive, and meaningful? ({yes, no∗ , n/a})\\
-	Are constants (other than 0 and 1) hard coded into the program? ({yes, no∗ , n/a})\\
-	Comments are clear, indicate what is being done, not how? ({yes, no∗ , n/a})\\
-	Is the name/URL of any algorithms used mentioned? ({yes, no∗ , n/a})\\
-	Parameters are in the same order for all functions? ({yes, no∗ , n/a})\\
-	Is code modularized? ({yes, no∗ , n/a})\\
-	Overall impression? ({1 .. 10})\\
-	Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have)\\
-	\hline		
-	\textbf{Visibility/Transparency}\\
-	\hline
-	Is the development process defined? If yes, what process is used. ({yes∗, no, n/a})\\
-	Are there any documents recording the development process and status?  ({yes∗, no}))\\
-	Is the development environment documented? ({yes∗, no})\\
-	Are there release notes? ({yes∗, no})\\
-	Overall impression? ({1 .. 10})\\
-	Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have)\\
-		\hline		
-	\textbf{Raw Metrics (Measured via git\_stats)}\\
-	\hline
-	Number of text-based files. (number)\\
-	Number of binary files. (number)\\
-	Number of total lines in text-based files. (number)\\
-	Number of total lines added to text-based files. (number)\\
-	Number of total lines deleted from text-based files. (number)\\
-	Number of total commits. (number)\\
-	Numbers of commits by year in the last 5 years. (Count from as early as possible if the project is younger than 5 years.) (list of numbers)\\
-	Numbers of commits by month in the last 12 months. (list of numbers)\\
-	\hline
-\end{tabular}
-
-\def\arraystretch{1.4}
-\begin{tabular}{p{14cm}}
-		\hline		
-	\textbf{Raw Metrics (Measured via scc)}\\
-	\hline
-	Number of text-based files. (number)\\
-	Number of total lines in text-based files. (number)\\
-	Number of code lines in text-based files. (number)\\
-	Number of comment lines in text-based files. (number)\\
-	Number of blank lines in text-based files. (number)\\
-	\hline
-	\textbf{Repo Metrics (Measured via GitHub)}\\
-	\hline
-	Number of stars. (number)\\
-	Number of forks. (number)\\
-	Number of people watching this repo. (number)\\
-	Number of open pull requests. (number)\\
-	Number of closed pull requests. (number)\\
-	\hline
-\end{tabular}
-
-\newpage
-
-\section{Developer Interview Questions}\label{interviewquestions}
-
-Information about these interview questions:  This gives you an idea what I would like to learn about the development of {domain} software.
-Interviews will be one-to-one and will be open-ended (not just “yes or no” answers).
-Because of this, the exact wording may change a little.
-Sometimes I will use other short questions to make sure I understand what you told me or if I need more information when we are talking such as: “So, you are saying that …?), to get more information (“Please tell me more?”), or to learn what you think or feel about something (“Why do you think that is…?”).
-
-\begin{enumerate}
-	\item Interviewees’ current position/title? degrees?
-	\item Interviewees’ contribution to/relationship with the software?
-	\item Length of time the interviewee has been involved with this software?
-	\item How large is the development group?
-	\item Do you have a defined process for accepting new contributions into your team?
-	\item What is the typical background of a developer?
-	\item What is your estimated number of users? How did you come up with that estimate?
-	\item What is the typical background of a user?
-	\item Currently, what are the most significant obstacles in your development process?
-	\item How might you change your development process to remove or reduce these obstacles?
-	\item How does documentation fit into your development process? Would improved documentation help with the obstacles you typically face?
-	\item In the past, is there any major obstacle to your development process that has been solved? How did you solve it?
-	\item What is your software development model? For example, waterfall, agile, etc.
-	\item What is your project management process? Do you think improving this process can
-	tackle the current problem? Were any project management tools used?
-	\item Was it hard to ensure the correctness of the software? If there were any obstacles, what methods have been considered or practiced to improve the situation? If practiced, did it work?
-	\item When designing the software, did you consider the ease of future changes?
-For example, will it be hard to change the structure of the system, modules or code blocks?
-What measures have been taken to ensure the ease of future changes and maintains?
-	\item Provide instances where users have misunderstood the software. What, if any, actions were taken to address understandability issues?
-	\item What, if any, actions were taken to address usability issues?
-	\item Do you think the current documentation can clearly convey all necessary knowledge to the users? If yes, how did you successfully achieve it? If no, what improvements are needed?
-	\item Do you have any concern that your computational results won’t be reproducible in the future? Have you taken any steps to ensure reproducibility?
-
-\end{enumerate}
-
-\newpage
-
-\section{Grading Template}\label{gradingtemplate2}
-The table below lists how each quality measure of the measurement template is used to calculate an overall impression in each software quality set.\newline
-
-\captionof{table}{Grading Template}\label{gradingtemplate}
-\def\arraystretch{1.5}
-\begin{tabular}{p{14cm}}
-	\hline		
-	\textbf{Installability  (Measured via installation on a virtual machine.) }\\
-	\hline
-	Are there installation instructions? ({yes=1, no=-1})\\
-	Are the installation instructions in one place? Place referring to a single document or web-page. ({yes=1, no=0, n/a=0})\\
-	Are the installation instructions linear? Linear meaning progressing  in a single series of steps. ({yes=1, no=0, n/a=0})\\
-	Are the instructions written as if the person doing the installation has none of the dependent packages installed? ({yes=1, no=0, unclear=0})\\
-	Are compatible operating system versions listed? ({yes=1, no=0})\\
-	Is there something in place to automate the installation (makefile, script, installer, etc)? ({yes*=1, no=-1})\\
-	If the software installation broke, was a descriptive error message displayed? ({yes=0, no=-2, n/a=1})\\
-	Is there a specified way to validate the installation? ({yes*=1, no=0})\\
-	How many steps were involved in the installation? (Includes manual steps like unzipping files) Specify OS. ($<$10 = 1)\\
-	What OS was used for the installation? (does not count)\\
-	How many extra software packages need to be installed before or during installation? ($<$10 = 1)\\
-	Are required package versions listed? ({yes=1, no=0, n/a=1})\\
-	Are there instructions for the installation of required packages / dependencies? ({yes=1, no=0, n/a=1})\\
-	Run uninstall, if available. Were any obvious problems caused? ({yes∗=0, no=1, unavail=1})\\
-	Overall impression? (a sum of $>$10 is rounded down to 10)\\
-	\hline
-\end{tabular}
-
-\def\arraystretch{1.33}
-\begin{tabular}{p{14cm}}
-	\hline		
-	\textbf{Correctness and Verifiability}\\
-	\hline
-	Any reference to the requirements specifications of the program or theory manuals? ({yes∗=2, no=0, unclear=0})\\
-	What tools or techniques are used to build confidence of correctness? (any=1, unclear=0)\\
-	If there is a getting started tutorial? ({yes=2, no=0})\\
-	Are the tutorial instructions linear? ({yes=1, no=0, n/a=0})\\
-	Does the getting started tutorial provide an expected output? ({yes=1, no*=0, n/a=0})\\
-	Does your tutorial output match the expected output? ({yes=1, no=0, n/a=0})\\
-	Are unit tests available?  ({yes=1, no=0, unclear=0})\\
-	Is there evidence of continuous integration? (for example mentioned in documentation, Jenkins, Travis CI, Bamboo, other) ({yes*=1, no=0, unclear=0})\\
-	\hline	
-	\textbf{Surface Reliability}\\
-	\hline
-	Did the software “break” during installation? ({yes∗=0, no=5})\\
-	If the software installation broke, was the installation instance recoverable? ({yes=5, no=0, n/a=0})\\
-	Did the software “break” during the initial tutorial testing? ({yes∗=0, no=5, n/a=0})\\
-	If the tutorial testing broke, was a descriptive error message displayed? ({yes=2, no=0, n/a=0})\\
-	If the tutorial testing broke, was the tutorial testing instance recoverable? ({yes=3, no=0, n/a=0})\\
-	\hline		
-	\textbf{Surface Robustness}\\
-	\hline
-	Does the software handle unexpected/unanticipated input (like data of the wrong type, empty input, missing files or links) reasonably?
-(a reasonable response can include an appropriate error message.) ({yes=5, no∗=0})\\
-	For any plain text input files, if all new lines are replaced with new lines and carriage returns, will the software handle this gracefully? ({yes=5, no∗=0, n/a=5})\\
-	\hline		
-\end{tabular}
-
-\def\arraystretch{1.4}
-\begin{tabular}{p{14cm}}
-		\hline		
-	\textbf{Surface Usability}\\
-	\hline
-	Is there a getting started tutorial? ({yes=3, no=0})\\
-	Is there a user manual? ({yes=4, no=0})\\
-	Are expected user characteristics documented? ({yes=1, no=0})\\
-	What is the user support model? FAQ? User forum? E-mail address to direct questions? Etc. (one=1, two+=2, none=0)\\
-	\hline
-	\textbf{Maintainability}\\
-	\hline
-	What is the current version number? (provided=1, nothing=0)\\
-	Is there any information on how code is reviewed, or how to contribute? ({yes*=1, no=0})\\
-	Are artifacts available?
-(List every type of file that is not a code file – for examples please look at the ‘Artifact Name’ column of https://gitlab.cas.mcmaster.ca/SEforSC/se4sc/-/blob/git-svn/GradStudents/Olu/ResearchProposal/Artifacts\_MiningV3.xlsx) (Rate 0 – 2 depending on how many and perceived quality)\\
-	What issue tracking tool is employed? (nothing=0, email of other private=1, anything public or accessible by all devs (eg git) = 2)\\
-	What is the percentage of identified issues that are closed? (50$\%$+=1, $<$50$\%$=0)\\
-	What percentage of code is comments? (10$\%$+=1, $<$10$\%$=0)\\
-	Which version control system is in use? (anything=2, nothing=0)\\
-	\hline		
-	\textbf{Reusability}\\
-	\hline
-	How many code files are there? (0-9=0, 10-49=1, 50-99=3, 100-299=4, 300-599=5, 600-999=6, 1000+=8)\\
-	Is API documented? ({yes=2, no=0, n/a=0})\\
-	\hline
-\end{tabular}
-
-\def\arraystretch{1.4}
-\begin{tabular}{p{14cm}}
-\hline	
-\textbf{Surface Understandability (Based on 10 random source files)}\\
-\hline
-Consistent indentation and formatting style? ({yes=1, no=0, n/a=0})\\
-Explicit identification of a coding standard? ({yes∗=1, no=0, n/a=0})\\
-Are the code identifiers consistent, distinctive, and meaningful? ({yes=2, no∗=0, n/a=0})\\
-Are constants (other than 0 and 1) hard coded into the program? ({yes=1, no∗=0, n/a=0})\\
-Comments are clear, indicate what is being done, not how? ({yes=2, no∗=0, n/a=0})\\
-Is the name/URL of any algorithms used mentioned? ({yes=1, no∗=0, n/a=0})\\
-Parameters are in the same order for all functions? ({yes=1, no∗=0, n/a=0})\\
-Is code modularized? ({yes=1, no∗=0, n/a=0})\\
-\hline		
-\textbf{Visibility/Transparency}\\
-\hline
-Is the development process defined? If yes, what process is used. ({yes∗=3, no=0, n/a=0})\\
-Are there any documents recording the development process and status?  ({yes∗=3, no=0}))\\
-Is the development environment documented? ({yes∗=2, no=0})\\
-Are there release notes? ({yes∗=2, no=0})\\
-\hline
-\end{tabular}
-
-\newpage
-
-\section{Eliminated Software Packages}\label{eliminatedpackagessection}
-
-The table below lists the software packages that were filtered out of the candidate software list.
-
-\begin{center}
-	\begin{tabular}{ p{4cm}p{4cm}}
-		\hline
-		Name & Reason(s) For Removal\\
-		\hline
-		ch4-project&Scope \& Usage\\
-		CUDA-LBM-simulator&Scope \& Age\\
-		elbe&Scope \& Usage\\
-		eLBM&Usage\\
-		firesim&Scope \& Usage \& Age\\
-		fvLBM&Scope \& Usage \& Age\\
-		JFlowSim&Scope \& Usage \& Age\\
-		LatticeBoltzmann&Scope \& Usage\\
-		LBM&Scope \& Usage \& Age\\
-		LBM-Cplusplus&Scope \& Usage\\
-		LBM-EP&Usage\\
-		LBM\_MATLAB&Scope \& Usage \& Age\\
-		LBSim&Scope \& Usage \& Age\\
-		listLBM&Scope \& Usage \& Age\\
-		loliverhennigh&Scope \& Usage \& Age\\
-		openLBMflow&Scope \& Usage \& Age\\
-		ParallelLbmCranfield&Usage \& Age\\
-		PowerFLOW&Usage\\
-		ProLB&Usage\\
-		Taxila-LBM&Usage \& Age\\
-		turbulent\_lbm\_multigpu&Usage \& Age\\
-		wlb&Scope \& Usage \& Age\\
-		\hline
-	\end{tabular}
-	\captionof{table}{Eliminated Software Packages}\label{eliminatedpackages}
-\end{center}
-
-\newpage
-
-\section{Ethics Approval}\label{ethicsapproval}
-This project received ethics clearance from the McMaster Research Ethics Board on February 20, 2021.\newline
-
-Project Title: AIMSS - State of the Practice\newline
-
-MREB\#: 5219
-
-\end{singlespace}
-\end{appendices}
-\newpage
-\begin{singlespace}
-\bibliographystyle {plainnat}
-\bibliography {References}
-
-\end{singlespace}
+As mentioned in Section~\ref{threats}, the measures listed in our measurement
+template may not be broad enough to accurately capture some qualities. Adding or
+extending measures is worth considering. For example, usability experiments and
+performance benchmarks could be incorporated into the assessment.
+
+Section~\ref{interviewresults} noted that developers were not asked to comment
+on the reproducibility of their source code from their requirements
+specifications and design documentation. Adding this question to the interview
+guide should be considered in the next iteration of state of the practice
+assessments. Furthermore, as mentioned in Section~\ref{threats}, it was found
+that some pertinent information was not specified in public artifacts. The use
+of unit testing and continuous integration by several packages (OpenLB, pyLBM,
+TCLB) was only discovered during interviews with developers. Adding further
+questions to the interview guide regarding the measures that are on the
+measurement template could reduce instances of incorrect data being collected.
+This additional interview data could be analyzed and incorporated into the AHP
+ranking, ensuring quality designations more accurate represent the true quality
+of the software packages.
+
+\bibliographystyle{ACM-Reference-Format}
+\bibliography{LatticeBoltz_SOP}
 
 \end{document}
+
+TO DO
+
+- add table of features - include Mohammed columns
+- add new program that wasn't measured - means measuring 24, not 23
+- update simple stats for other projects
+- compare to existing survey results (do this in Ao's too)
+- read more of the existing survey studies, as highlighted by Carver paper
+- create issues for Peter
+- data on Mendeley

### 2021-12-11 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Fixed interview subsection heading
Updates to sections on AHP and Interaction with the Domain Expert

@@
 
-\section{Survey Developers} \label{SecSurvey}
+\subsection{Interview Developers} \label{SecSurvey}
 
@@
 for comparing and ranking the LBM software packages using the overall impression
-quality scores that were gathered in the measurement template found in Appendix
-~\ref{measurementtemplate2} using the grading template found in Appendix
-~\ref{gradingtemplate2}. AHP performs a pairwise analysis using a matrix and
-generates an overall score as well as individual quality scores for each
-software package. Smith et al. (2016) shows how the AHP is applied to ranking
-software based on quality measures \cite{SmithEtAl2016}.
+quality scores that were gathered in the measurement template.  AHP performs a
+pairwise analysis using a matrix and generates an overall score as well as
+individual quality scores for each software package. \citep{SmithEtAl2016} shows
+how AHP is applied to ranking software based on quality measures.
 
@@
 sensitivity analysis that was used to ensure that the software package rankings
-are appropriate with respect to the uncertainty of the quality scores. The
+are appropriate with respect to the uncertainty of the quality scores. For the
+sensitivity analysis we modified the score by 10\% for each package and verified
+that the overall ranking was stable.  The
 \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/AHP2020/LBM/README.txt}{README}
@@
 \end{inparaenum}
-
-Domain experts may be recruited from academia or industry.  The only
-requirements are knowledge of the domain and a willingness to be engaged in the
-assessment process.  The Domain Expert does not have to be a software developer,
-but they should be a user of domain software.  Given that the domain experts are
-likely to be busy people, the measurement process cannot put to much of a burden
-on their time.  For the current assessment, our Domain Expert (and paper
-co-author) is Dr.\ Michael Noseworthy, Professor of Electrical and Computer
-Engineering at McMaster University, Co-Director of the McMaster School of
-Biomedical Engineering, and Director of Medical Imaging Physics and Engineering
-at St. Joseph’s Healthcare, Hamilton, Ontario, Canada.
-
-In advance of the first meeting with the Domain Expert, the expert is asked to
-create a list of top software packages in the domain.  This is done to help
-the expert get in the right mind set in advance of the meeting.  Moreover,
-by doing the exercise in advance, we avoid the potential pitfall of the expert
-approving the discovered list of software without giving it adequate thought.
-
-The Domain Experts are asked to vet the collected data and analysis.  In
-particular, they are asked to vet the proposed list of software packages and the
-AHP ranking.  These interactions can be done either electronically or with
-in-person (or virtual) meetings.
+For the current assessment, our Domain Expert (and paper co-author) is Dr.\
+Zahra Motamed, Assistant Professor of Mechanical Engineering at McMaster
+University, Hamilton, Ontario, Canada.
+
+The Domain Expert has an important role with verifying the list of LBM packages.
+In advance of the first meeting with the Domain Expert, they were asked to
+create a list of top software packages in the domain.  This is done to help the
+expert get in the right mind set in advance of the meeting.  Moreover, by doing
+the exercise in advance, we avoid the potential pitfall of the expert approving
+the discovered list of software without giving it adequate thought.  The Domain
+Expert was also asked to vet the collected data and analysis.  In particular,
+they were asked to vet the proposed list of software packages and the AHP
+ranking.  These interactions were done via virtual meetings.
 

### 2021-12-11 StateOfPractice/Papers/MedImageSoft_SOP-arXiv/MedImageSoft_SOP_AcademicRadiology.tex
Minor updates to MI SOP paper
Answer questions to MI paper

@@
 git repository via e-mail (1 team).  \wss{I do not know what this last one means.
-Should ask Ao.}
+Should ask Ao.} \ad{The last one happened when cloud-based repositories (Github,
+Gitlab, Bitbucket, etc.) were not widely used. The project git repository was
+stored locally on the project owner's local computer. The owner shared the whole
+git repo to team members and other individuals who wanted to contribute. The
+contributors made changes to the repo, and sent it back. They transferred the
+repo back and forth by email.}
 
@@
 \item The team cannot use private data for debugging even when the data causes
-problems; \wss{Why not?  Should ask Ao}
+problems; \wss{Why not?  Should ask Ao} \ad{For example, when a physician faces
+some problems to use the software to display a medical image, he/she might reach
+to the team for help. However, the image is private, so the physician can only
+describe the problems, but not share the image with the team to reproduce the
+problems for debugging.}
 \item Expense and time consumed by tests because of the huge datasets;
@@
 understandability for end users include users not understanding how to use
-features (\wss{Because the users don't have sufficient background?  Ask Ao.}),
+features (\wss{Because the users don't have sufficient background?  Ask Ao.})
+\ad{Did you mean medical or medical imaging background? No, it was majorly
+caused by the software usability. Based on the context of our interview
+conversations, it means that users don't understand how to complete certain
+operations or what functions a feature provides. It's due to unclear description,
+lack of documentation, poorly designed UI or nonintuitive process},
 the team not having a dedicated user experience (UX) designer, some important
@@
 understandability, rather than threat to it.  Or is the point that developers
-lack sufficient background?  Ask Ao}), and the architecture is difficult for new
+lack sufficient background?  Ask Ao}) \ad{Yes, this should be a consequence of a
+lack of understandability. I think the direct cause is that the project doesn't
+clearly explain how to deploy. The background of the developers wasn't discussed
+at that point.}, and the architecture is difficult for new
 developers to understand.
@@
 \wss{because the interface and/or behaviour of the library could have changed?
-Ask Ao}, variability between CPUs, and misinterpretation of how manufacturers
+Ask Ao} \ad{Yes. Behaviour change could be caused by a lib changing to use
+different algorithms or different dependencies}, variability
+between CPUs, and misinterpretation of how manufacturers
 create medical images. The most commonly cited (by 6 teams) strategy to improve
@@
 to real-world datasets for testing. The developers' strategies to address this
-are summarized in Section~\ref{sec_pain_points_3}. One threat to correctness is:
+are summarized in Section~\ref{sec_pain_points_3} \ad{to be fixed}. One threat to correctness is:
 with huge datasets for testing, the tests are expensive and time-consuming.
@@
     - for journal version ask for more figures?
+    - for journal double blind solution
+    - can he submit
+    - potential reviewers
+    - feedback
+- Radiology - should not exceed 6000 words, 50 refs, 8 figures and 4 tables

### 2021-12-11 StateOfPractice/Papers/MedImageSoft_SOP-arXiv/MedImageSoft_SOP_AcademicRadiology.tex
Fixed a typo in the section about virtual machines

--- 2021-12-07 StateOfPractice/Papers/MedImageSoft_SOP-arXiv/MedImageSoft_SOP_AcademicRadiology.tex
@@
 starts from a clean slate, which removes ``works-on-my-computer'' errors. When
-filling in the measurement template spreadsheet, the the details for each VM
+filling in the measurement template spreadsheet, the details for each VM
 should be noted, including hypervisor and operating system version.

### 2021-12-11 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Updates to the installation section

@@
 
+As in \citet{SmithEtAl2016}, Virtual machines (VMs) were used to provide an
+optimal testing environments for each package. VMs were used because it is
+easier to start with a fresh environment without having to worry about existing
+libraries and conflicts. Moreover, when the tests are complete the VM can be
+deleted, without any impact on the host operating system. The most significant
+advantage of using VMs is to level the playing field. Every software install
+starts from a clean slate, which removes ``works-on-my-computer'' errors. When
+filling in the measurement template spreadsheet, the details for each VM
+should be noted, including hypervisor and operating system version.
+
 \subsection{Interview Developers} \label{SecSurvey}
@@
 
-This section presents the quantitative findings from data that was gathered
-using the measurement template found in Section~\ref{measurementtemplate2} for
-qualities listed in Section~\ref{softwarequalities}. The results of the AHP
-analysis based on that data are also presented in this section.
-
-Table that summarizes the software by variabilities.
+Table~\ref{} presents a summary of the software that was measured.  The table
+shows the differences between the packages based on features.  \wss{Add some
+discussion of some example features.}  The sections below summarize how the
+software packages compare based on qualities, as measured by the measurement
+template.   The results of the AHP ranking are also discussed.
 
@@
 original long list of 45 packages were removed due to not including
-documentation or installation instructions.
+documentation or installation instructions. Of the 23 software packages that
+were tested, most had installation instructions located in one place, often in
+an instruction manual or on a web-page. Sometimes, like with Ludwig, incomplete
+installation instructions are found on a home page, with more detailed
+instructions located on another web-page, or within the documentation.
+Maintainability and correctness of these instructions could be improved if all
+the instructions were in one location.
 
@@
 system compatibility is found in the documentation of 19 software packages. All
-but one of the software packages, TCLB, were tested on Ubuntu for this state of
+but one of the software packages (TCLB) were tested on Ubuntu for this state of
 the practice assessment. TCLB was tested on CentOS, since this operating system
 is mentioned in its installation instructions.
-
-Of the software packages that were tested, most installation instructions are
-located in one place, often in an instruction manual or on a web-page.
-Sometimes, like with Ludwig, incomplete installation instructions are found on a
-home page, with more detailed instructions located on another web-page, or
-within the documentation. Maintainability and correctness of these instructions
-could be improved if all the instructions were in one location.
 
@@
 About half of the installation instructions are written as if the person doing
-the installation has none of the dependent packages installed. It is common for
-software packages to not list all of their dependencies despite listing some.
-This was the case for many packages, including ESPResSo++, Ludwig, and LUMA.
-Sometimes only an error message during the installation process informs the user
-of the requirement of these additional packages. A detailed rewrite of the
-installation instructions from the point of view of installation on a clean
-operating system is suggested. A clean environment can be achieved for testing
-purposes by using a virtual machine.
+the installation has none of the dependent packages installed. Unfortunately,
+software packages, like ESPResSo++, Ludwig, and LUMA, frequently don't list all
+of their dependencies, or provide only a partial list. Sometimes only an error
+message during the installation process informs the user of the requirement of
+these additional packages. A detailed rewrite of the installation instructions
+from the point of view of installation on a clean operating system is suggested.
+A clean environment can be achieved for testing purposes by using a virtual
+machine.
 
@@
 
-Uninstallation instructions were found for only one of the software packages,
-pyLBM.
+Uninstallation instructions were found for only one of the software packages: pyLBM.
 

### 2021-12-12 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Finished revising up to and including Section 3 (Quantitative Findings)

@@
 if versions of dependencies are specified. Fifteen software packages do not have
-detailed instructions for installing dependencies.
-
-Sixteen software packages have less than 10 manual installation steps. If
-dependencies are installed in one command then none of the software packages
-take more than 20 steps to install. The average number of steps is about eight,
-and the fewest is two (LB3D-Prime).
+detailed instructions for installing dependencies. Sixteen software packages
+have less than 10 manual installation steps. If dependencies are installed in
+one command then none of the software packages take more than 20 steps to
+install. The average number of steps is about eight, and the fewest is two
+(LB3D-Prime).
 
 All but six (ESPResSo, HemeLB, laboetie, LB3D-Prime, lbmpy, waLBerla) of the
-software packages have a way to specifically verify the installation. Most have
-some sort of tutorial examples that can be run by the user. Some other ways of
-installation validation include validation scripts (LB2D-Prime, lettuce, Ludwig,
-LUMA), automatic validation after the installation (LatBo.jl), and instructions
-to manually review the file system (LIMBES).
-
-Uninstallation instructions were found for only one of the software packages: pyLBM.
+software packages have a way to verify the installation. Most have some sort of
+tutorial examples that can be run by the user. Some other ways of installation
+validation include validation scripts (LB2D-Prime, lettuce, Ludwig, LUMA),
+automatic validation after the installation (LatBo.jl), and instructions to
+manually review the file system (LIMBES).  Uninstallation instructions were
+found for only one of the software packages: pyLBM.
 
@@
 of easy to follow tutorials further helps users verify the software and have
-confidence in its correctness. Unit testing documentation and capability, as
-well as the use of continuous integration tools and techniques such as Bamboo,
-Jenkins, and Travis CI, help verify correctness.
+confidence in its correctness. Unit testing, as well as the use of continuous
+integration tools and techniques such as Bamboo, Jenkins, and Travis CI, help
+verify correctness.
 
@@
 commands (even when following the installation guide), missing links, missing
-dependencies, syntax errors in code files. In some instances the error messages
+dependencies, and syntax errors in code files. In some instances the error messages
 were vague. Several automatic installation processes could not find and load
@@
 inclusion of testing and tutorial examples, and the assurance of descriptive
-error messages during fault conditions.
+error messages after faults.
 
@@
 messages or did not crash. Software packages with a lower surface robustness
-score had not installed correctly, so their robustness score may not be a true
-reflection of runtime robustness. Similarly, all software packages that
-installed correctly and require plain text input files correctly handled an
-unexpected change to these input files, including a replacement of new lines
-with carriage returns. Furthermore, nine of the top 10 ranked packages are noted
-as being alive. LIMBES is noted as being dead.
+score did not install correctly, so their robustness score may not be a true
+reflection of runtime robustness. All successfully installed software packages
+that require a plain text input file correctly handled an unexpected change to
+these input files, including a replacement of new lines with carriage returns.
+Furthermore, nine of the top 10 ranked packages are noted as being alive. LIMBES
+is noted as being classified as dead.
 
@@
 problems, the packages focus on varied CFD problems, with varying parameters,
-and are technically different from each other. Due to this, a comparison of
-performance is not appropriate. In this project we instead looked through each
-software package's artifacts for evidence that performance was considered. The
-artifacts of 17 software packages mentioned parallelization. This included GPU
-processing and the CUDA parallel computing platform, which were mentioned in the
-artifacts of 6 packages (ESPResSo, lbmpy, lettuce, pyLBM, Sailfish, TCLB). GPUs
-provide superior processing power and speed compared to CPUs, and are often used
-for scientific computing when a large amount of data is involved. The software
+and are technically different from each other (as shown in
+Table~\ref{tbl_features}). Due to this, a comparison of performance is not
+appropriate. In this project we instead looked through each software package's
+artifacts for evidence that performance was considered. The artifacts of 17
+software packages mentioned parallelization. This included GPU processing and
+the CUDA parallel computing platform, which were mentioned in the artifacts of 6
+packages (ESPResSo, lbmpy, lettuce, pyLBM, Sailfish, TCLB). GPUs provide
+superior processing power and speed compared to CPUs, and are often used for
+scientific computing when a large amount of data is involved. The software
 package TCLB is implemented in a highly efficient multi-GPU code to achieve
@@
 software packages have a tutorial, 13 have a user manual, and 11 have both. The
-tutorials vary in scope and substance, and eight include an expected output.
+tutorials vary in scope and substance, and eight include expected output.
 Most user manuals are in the form of a file that can be downloaded, while some
@@
 were also reviewed for software release and documentation version numbers. This
-information could be used to better troubleshoot issues and organize
-documentation. All but three software packages (LatBo.jl, LB3D-Prime, MechSys)
-have source code release and documentation version numbers.
-
-Information on how code is reviewed, or how to contribute to the project was
-also noted. In total, 11 software packages have this information, which was
-found in various artifacts, including in developer guides, contributor guides,
-user guides, developer web-pages, and README files.
+information can be used to troubleshoot issues and organize documentation. All
+but three software packages (LatBo.jl, LB3D-Prime, MechSys) have source code
+release and documentation version numbers.
+
+When present, information on how code is reviewed, or how to contribute to the
+project was also noted. In total, 11 software packages have this information,
+which was found in various artifacts, including in developer guides, contributor
+guides, user guides, developer web-pages, and README files.
 
@@
 
+\begin{table}
+	\begin{center}
+		\begin{tabular}{ p{3.5cm}p{3.5cm}p{3.5cm}p{2.5cm} }
+			\hline
+			Name & $\%$ Issues Closed & $\%$ Code Comments & Status\\
+			\hline
+			DL\_MESO (LBE) & Not Git & 8.06& Alive\\
+			ESPResSo & 89.26 & 21.78 & Alive\\
+			ESPResSo++ & 66.28 & 17.10 & Alive\\
+			HemeLB & No Issues & 16.68 & Dead\\
+			laboetie & 18.75 & 2.47 & Dead\\		
+			LatBo.jl & 93.33 & 0.40 & Dead\\
+			LB2D-Prime & Not Git & 13.61 & Dead\\
+			LB3D & Not Git & 13.76 & Dead\\
+			LB3D-Prime & Not Git & 14.34 & Dead\\
+			lbmpy& 58.33  & 2.03 & Alive\\
+			lettuce & 33.33 & 8.19 & Alive\\
+			LIMBES & Not Git & 17.39 & Dead\\
+			Ludwig& 60.00 & 20.70 & Alive\\
+			LUMA& 85.71   & 0.20 & Alive\\
+			MechSys & Not Git & 15.11 & Alive\\
+			MP-LABS & 100.00 & 26.67 & Dead\\
+			OpenLB & Not Git & 22.43 & Alive\\
+			Palabos & 89.47 & 17.76 & Alive\\
+			pyLBM & 66.67& 16.12 & Alive\\
+			Sailfish & 22.22 & 9.26 & Alive\\
+			SunlightLB & Not Git & 17.67 & Dead\\
+			TCLB & 60.32 & 6.02 & Alive\\
+			waLBerla & 72.90 & 22.62 & Alive\\
+			\hline
+		\end{tabular}
+		\caption{Git Repository Data} \label{gitrepodata}
+	\end{center}
+	\end{table}
+
 Software package code files were further measured for the percentage of code
@@
 more than the average. Fifth ranked LUMA has only 0.2 percent comments, the
-fewest of any package. The package has the most lines of source code, with over
+fewest of any package. This package has the most lines of source code, with over
 four million. The next largest package is ESPResSo++ with one million.
@@
 artifacts, and use an issue tracking tool and version control system. These
-packages also appear to reasonably handle tracked issues, having most of their
+packages also appear to reasonably handle issue tracking, having most of their
 issues closed. Their code files are well commented with more than 10 percent of
@@
 
-\begin{table}
-\begin{center}
-	\begin{tabular}{ p{3.5cm}p{3.5cm}p{3.5cm}p{2.5cm} }
-		\hline
-		Name & $\%$ Issues Closed & $\%$ Code Comments& Status\\
-		\hline
-		DL\_MESO (LBE) & Not Git & 8.06&Alive\\
-		ESPResSo & 89.26 & 21.78&Alive\\
-		ESPResSo++ & 66.28 & 17.10&Alive\\
-		HemeLB & No Issues & 16.68&Dead\\
-		laboetie & 18.75 & 2.47&Dead\\		
-		LatBo.jl & 93.33 & 0.40&Dead\\
-		LB2D-Prime & Not Git & 13.61&Dead\\
-		LB3D & Not Git & 13.76&Dead\\
-		LB3D-Prime & Not Git & 14.34&Dead\\
-		lbmpy& 58.33  & 2.03  &Alive\\
-		lettuce & 33.33 & 8.19&Alive\\
-		LIMBES & Not Git & 17.39&Dead\\
-		Ludwig& 60.00 & 20.70&Alive\\
-		LUMA& 85.71   & 0.20&Alive\\
-		MechSys & Not Git & 15.11&Alive\\
-		MP-LABS & 100.00 & 26.67&Dead\\
-		OpenLB & Not Git & 22.43&Alive\\
-		Palabos & 89.47 & 17.76&Alive\\
-		pyLBM & 66.67& 16.12&Alive\\
-		Sailfish & 22.22 & 9.26&Alive\\
-		SunlightLB & Not Git & 17.67&Dead\\
-		TCLB & 60.32 & 6.02&Alive\\
-		waLBerla & 72.90 & 22.62&Alive\\
-		\hline
-	\end{tabular}
-	\caption{Git Repository Data} \label{gitrepodata}
-\end{center}
-\end{table}
+\wss{Sort ``Git Repository Data'' table in descending order of percentage of
+code comments.}
 
@@
 
-Each software package was measured for the total number of source code files. A
-larger number of source files was associated with increased reusability due to
-increased modularization. Some packages have more features than others,
-consequently contributing to reusablility since they have more source code that
-can be reused. The software packages were also reviewed for the presence of API
-documentation, which indicates that a software package was developed with
-interaction between other software applications in mind.
+We measured the total number of source code files for each project. A larger
+number of source files is associated with increased reusability, due to our
+assumption that this indicates increased modularization. Some packages have more
+features than others.  This is assumed to contribute to reusablility, since they have
+more source code for potential reuse. The software packages were also reviewed
+for the presence of API documentation, which indicates that a software package
+was developed with interaction between other software applications in mind.
 
@@
 code files and API documentation. The highest scoring packages, ESPResSo and
-waLBerla, have extensive functionality, including graphical visualizations as
-well as modeling that does not use LBM. For this reason a comparison with other
-software packages is not on a level field. However, these packages do have an
-abundance of reusable components. Furthermore, nine of the top 10 ranked
-packages are noted as being alive. HemeLB is noted as being dead.
+waLBerla, have extensive functionality, including graphical visualizations and
+non-LBM modeling. For this reason, a comparison with other software packages is
+not on a level field. However, these packages do have an abundance of reusable
+components. Furthermore, nine of the top 10 ranked packages are noted as being
+alive. HemeLB is considered to be dead.
 
@@
 			\hline
-			DL\_MESO (LBE) & 310 & 51 & 170223& 549\\
-			ESPResSo &1309& 86 & 186700&143\\
-			ESPResSo++ &5328& 66 & 969196&182\\
-			HemeLB &1065& 48 & 95104&89\\
-			laboetie &133& 1 & 48403&364\\		
-			LatBo.jl &41& 0 & 42172&1029\\
-			LB2D-Prime &82& 19 & 54755&668\\
-			LB3D &99& 76 & 39766&402\\
-			LB3D-Prime &23& 6 & 12944&563\\
-			lbmpy&201&  28 & 46489  &231\\
-			lettuce &62& 0 & 5529&89\\
-			LIMBES &26& 1 & 4872&187\\
-			Ludwig&859& 32 & 109811&128\\
-			LUMA&312& 19 & 4370670&14000\\
-			MechSys &324& 3 & 85543&264\\
-			MP-LABS &307& 3 & 43124&140\\
-			OpenLB &1104& 5 & 209034&189\\
-			Palabos &1829& 67 & 547623&299\\
-			pyLBM &258& 85& 32314&125\\
-			Sailfish &632& 11 & 69398&110\\
-			SunlightLB & 36& 1 & 7646& 212\\
-			TCLB &535& 7 & 43226&81\\
-			waLBerla & 2395 & 67 & 848146&353\\
+			DL\_MESO (LBE) & 310 & 51 & 170223 & 549\\
+			ESPResSo & 1309& 86 & 186700 & 143\\
+			ESPResSo++ & 5328& 66 & 969196 & 182\\
+			HemeLB & 1065& 48 & 95104 & 89\\
+			laboetie & 133& 1 & 48403 & 364\\		
+			LatBo.jl & 41& 0 & 42172 & 1029\\
+			LB2D-Prime & 82& 19 & 54755 & 668\\
+			LB3D & 99 & 76 & 39766 & 402\\
+			LB3D-Prime & 2 3& 6 & 12944 & 563\\
+			lbmpy& 201 & 28 & 46489 & 231\\
+			lettuce & 62 & 0 & 5529 & 89\\
+			LIMBES & 26 & 1 & 4872 & 187\\
+			Ludwig & 859 & 32 & 109811 & 128\\
+			LUMA & 312 & 19 & 4370670 & 14000\\
+			MechSys & 324 & 3 & 85543 & 264\\
+			MP-LABS & 307 & 3 & 43124 & 140\\
+			OpenLB & 1104 & 5 & 209034 & 189\\
+			Palabos & 1829 & 67 & 547623 & 299\\
+			pyLBM & 258 & 85 & 32314 & 125\\
+			Sailfish & 632 & 11 & 69398 & 110\\
+			SunlightLB & 36 & 1 & 7646 & 212\\
+			TCLB & 535 & 7 & 43226 & 81\\
+			waLBerla & 2395 & 67 & 848146 & 353\\
 			\hline
@@
 
-Table~\ref{moduledata} shows file and line of code data of the software
-packages. Packages with a high reusability score do not have many LOC per text
-file, generally having a few hundred lines or less. This suggests that the
+Table~\ref{moduledata} shows file and Lines Of Code (LOC) data for the software
+packages. Packages with a high reusability score do not have as many LOC per
+text file, generally having a few hundred lines or less. This suggests that the
 source code of these packages is likely functionally modularized, and modules
@@
 enhance productivity, reusability, and maintainability
-\citep{bauer2021walberla}. Its software design has enabled waLBerla to be
+\citep{bauer2021walberla}. The design of waLBerla has enabled it to be
 successfully applied in several projects as a basis for various extensions
@@
 measures. This assessment of surface understandability may not perfectly reflect
-each package due to the practical limitation of only being able to test 10
-files.
-
-All of the packages appear to have a consistent indentation and formatting
-style. Only LUMA and HemeLB explicitly identify coding standards that are used
-during development. Generally, the software packages use consistent,
-distinctive, and meaningful code identifiers. Only four packages (LB2D-Prime,
-LB3D-Prime, LIMBES, MP-LABS) appear to use vague identifiers, such as single
-letters for variables. Symbolic constants were observed in the source code of 12
-packages. The constants are used for various parameters, mathematical constants,
-and matrix definitions. All of the packages are well commented and the comments
-clearly indicate what is being done. Domain algorithms are noted in the source
-code of 11 packages. Table~\ref{moduledata} suggests that the software packages
-are modularized to various degrees. When observing the source code files, it was
-found that 13 of the packages have a consistent style and order of function
-parameters.
+each package, due to the practical limitation of only examining 10 files.
+
+All of the packages appear to have consistent indentation and formatting. Only
+LUMA and HemeLB explicitly identify coding standards that are used during
+development. Generally, the software packages use consistent, distinctive, and
+meaningful code identifiers. Only four packages (LB2D-Prime, LB3D-Prime, LIMBES,
+MP-LABS) appear to use vague identifiers, such as single letters for variables.
+Symbolic constants were observed in the source code of 12 packages. The
+constants are used for various parameters, mathematical constants, and matrix
+definitions. All of the packages are reasonably well commented, with the
+comments clearly indicating what is being done (as opposed to how it is being
+done). Domain algorithms are noted in the source code of 11 packages.
+Table~\ref{moduledata} suggests that the software packages are modularized to
+various degrees. When observing the source code files, it was found that 13 of
+the packages have a consistent style and order of function parameters.
 
@@
 Figure~\ref{Fig_OverallScore} shows the overall ranking of the software packages
-using AHP. Software packages with an overall high score had ranked high in at
-least several of the individual qualities that were quantitatively measured. The
-overall ranking is found with an assumption of an equal weighting of all
-qualities. If the qualities were weighed differently, the overall software
-package ranking would change.
-
-Looking at the top three ranked packages, Ludwig scored high in surface
+using AHP. In the absence of other information on priorities, the overall
+ranking was calculated by assuming an equal weighting between all qualities. If the
+qualities were weighed differently, the overall software package ranking would
+change. Software packages with an overall high score ranked high in at least
+several of the individual qualities that were quantitatively measured.
+
+Looking at the top three ranked packages: Ludwig scored high in surface
 correctness and verifiability, surface robustness, surface usability,
@@
 surface robustness, maintainability, and understandability.
-
-Section~\ref{qualityrecommentations} further analyzes these findings and offers
-some software quality recommendations for future development of LBM software.
 

### 2021-12-13 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
added variabilities into table

@@
 	\begin{center}
-        \begin{tabular}{ p{3cm}lllllll}
-            \hline
-            Name & Dim & Pll & Com & Rflx & Mfl & NN & OS\\
-            \hline
-            DL\_MESO (LBE) & ? & ? & ? & ? & ? & ? & ?\\
-            ESPResSo & ? & ? & ? & ? & ? & ? & ?\\
-            ESPResSo++ & ? & ? & ? & ? & ? & ? & ?\\
-            lbmpy & ? & ? & ? & ? & ? & ? & ?\\
-            lettuce & ? & ? & ? & ? & ? & ? & ?\\
-            Ludwig & ? & ? & ? & ? & ? & ? & ?\\
-            LUMA & ? & ? & ? & ? & ? & ? & ?\\
-            MechSys & ? & ? & ? & ? & ? & ? & ?\\
-            OpenLB & ? & ? & ? & ? & ? & ? & ?\\
-            Palabos & ? & ? & ? & ? & ? & ? & ?\\
-            pyLBM & ? & ? & ? & ? & ? & ? & ?\\
-            Sailfish & ? & ? & ? & ? & ? & ? & ?\\
-            TCLB & ? & ? & ? & ? & ? & ? & ?\\
-            waLBerla & ? & ? & ? & ? & ? & ? & ?\\
-            HemeLB & ? & ? & ? & ? & ? & ? & ?\\
-            laboetie & ? & ? & ? & ? & ? & ? & ?\\
-            LatBo.jl & ? & ? & ? & ? & ? & ? & ?\\
-            LB2D-Prime & ? & ? & ? & ? & ? & ? & ?\\
-            LB3D & ? & ? & ? & ? & ? & ? & ?\\
-            LB3D-Prime & ? & ? & ? & ? & ? & ? & ?\\
-            LIMBES & ? & ? & ? & ? & ? & ? & ?\\
-            MP-LABS & ? & ? & ? & ? & ? & ? & ?\\
-            SunlightLB & ? & ? & ? & ? & ? & ? & ?\\
-
+		\begin{tabular}{ p{3cm}llllllll}
+			\hline
+			Name & Dim & Pll & Com & Rflx & MFl & Turb & CGE & OS\\
+			\hline
+			DL\_MESO (LBE) & 2, 3 & Y & Y & Y & Y & Y & Y & Win, Mac, Lin\\
+			ESPResSo & 1, 2, 3 & Y & Y & Y & Y & Y & Y & Mac, Lin\\
+			ESPResSo++ & 1, 2, 3 & Y & Y & Y & Y & Y & Y & Lin\\
+			HemeLB & 3 & Y & Y & Y & Y & Y & Y & Lin\\
+			laboetie & 2, 3 & Y & Y & Y & Y & Y & Y & Lin\\
+			LatBo.jl & 2, 3 & N & Y & Y & Y & N & Y & Lin\\
+			LB2D-Prime & 2 & Y & Y & Y & Y & Y & Y & Win, Lin\\
+			LB3D & 3 & Y & N & Y & Y & Y & Y & Lin\\
+			LB3D-Prime & 3 & Y & Y & Y & Y & Y & Y & Win, Lin\\
+			lbmpy & 2, 3 & Y & Y & Y & Y & Y & Y & Lin\\
+			lettuce & 2, 3 & N & Y & Y & Y & Y & Y & Win, Mac, Lin\\
+			LIMBES & 2 & Y & Y & Y & N & N & Y & Lin\\
+			Ludwig & 2, 3 & Y & Y & Y & Y & Y & Y & Lin\\
+			LUMA & 2, 3 & Y & Y & Y & Y & Y & Y & Win, Mac, Lin\\
+			MechSys & 2, 3 & N & Y & Y & Y & Y & Y & Lin\\
+			MP-LABS & 2, 3 & Y & N & Y & Y & N & N & Lin\\
+			OpenLB & 1, 2, 3 & Y & Y & Y & Y & Y & Y & Win, Mac, Lin\\
+			Palabos & 2, 3 & Y & Y & Y & Y & Y & Y & Win, Lin\\
+			pyLBM & 1, 2, 3 & Y & Y & Y & N & Y & Y & Win, Mac, Lin\\
+			Sailfish & 2, 3 & Y & Y & Y & Y & Y & Y & Mac, Lin\\
+			SunlightLB & 3 & N & Y & Y & N & N & Y & Lin\\
+			TCLB & 2, 3 & Y & Y & Y & Y & Y & Y & Lin\\
+			waLBerla & 2, 3 & Y & Y & Y & Y & Y & Y & Lin\\
 			\hline
 		\end{tabular}
-        \caption{Features of Software Packages (Dim for Dimension (1, 2 or 3),
+		\caption{Features of Software Packages (Dim for Dimension (1, 2, 3),
 			Pll for Parallel (Yes or No), Com for Compressible (Yes or No), Rflx for
-        Reflexive boundary condition (Yes or No), MFl for Multifluid (Yes or
-        No), NN for Non-Newtonian, OS (Win, Mac, Lin, )} \label{tbl_features}
+			Reflexive Boundary Condition (Yes or No), MFl for Multifluid (Yes or
+			No), Turb for Turbulent (Yes or No), CGE for Complex Geometries (Yes or No), OS for Operating System (Win, Mac, Lin))} \label{tbl_features}
 	\end{center}

### 2021-12-13 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
added code for operating systems

--- 2021-12-12 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
@@
 	\begin{center}
-        \begin{tabular}{ p{3cm}lllllll}
-            \hline
-            Name & Dim & Pll & Com & Rflx & Mfl & NN & OS\\
-            \hline
-            DL\_MESO (LBE) & ? & ? & ? & ? & ? & ? & ?\\
-            ESPResSo & ? & ? & ? & ? & ? & ? & ?\\
-            ESPResSo++ & ? & ? & ? & ? & ? & ? & ?\\
-            lbmpy & ? & ? & ? & ? & ? & ? & ?\\
-            lettuce & ? & ? & ? & ? & ? & ? & ?\\
-            Ludwig & ? & ? & ? & ? & ? & ? & ?\\
-            LUMA & ? & ? & ? & ? & ? & ? & ?\\
-            MechSys & ? & ? & ? & ? & ? & ? & ?\\
-            OpenLB & ? & ? & ? & ? & ? & ? & ?\\
-            Palabos & ? & ? & ? & ? & ? & ? & ?\\
-            pyLBM & ? & ? & ? & ? & ? & ? & ?\\
-            Sailfish & ? & ? & ? & ? & ? & ? & ?\\
-            TCLB & ? & ? & ? & ? & ? & ? & ?\\
-            waLBerla & ? & ? & ? & ? & ? & ? & ?\\
-            HemeLB & ? & ? & ? & ? & ? & ? & ?\\
-            laboetie & ? & ? & ? & ? & ? & ? & ?\\
-            LatBo.jl & ? & ? & ? & ? & ? & ? & ?\\
-            LB2D-Prime & ? & ? & ? & ? & ? & ? & ?\\
-            LB3D & ? & ? & ? & ? & ? & ? & ?\\
-            LB3D-Prime & ? & ? & ? & ? & ? & ? & ?\\
-            LIMBES & ? & ? & ? & ? & ? & ? & ?\\
-            MP-LABS & ? & ? & ? & ? & ? & ? & ?\\
-            SunlightLB & ? & ? & ? & ? & ? & ? & ?\\
-
+		\begin{tabular}{ p{3cm}llllllll}
+			\hline
+			Name & Dim & Pll & Com & Rflx & MFl & Turb & CGE & OS\\
+			\hline
+			DL\_MESO (LBE) & 2, 3 & Y & Y & Y & Y & Y & Y & W, M, L\\
+			ESPResSo & 1, 2, 3 & Y & Y & Y & Y & Y & Y & M, L\\
+			ESPResSo++ & 1, 2, 3 & Y & Y & Y & Y & Y & Y & L\\
+			HemeLB & 3 & Y & Y & Y & Y & Y & Y & L\\
+			laboetie & 2, 3 & Y & Y & Y & Y & Y & Y & L\\
+			LatBo.jl & 2, 3 & N & Y & Y & Y & N & Y & L\\
+			LB2D-Prime & 2 & Y & Y & Y & Y & Y & Y & W, L\\
+			LB3D & 3 & Y & N & Y & Y & Y & Y & L\\
+			LB3D-Prime & 3 & Y & Y & Y & Y & Y & Y & W, L\\
+			lbmpy & 2, 3 & Y & Y & Y & Y & Y & Y & L\\
+			lettuce & 2, 3 & N & Y & Y & Y & Y & Y & W, M, L\\
+			LIMBES & 2 & Y & Y & Y & N & N & Y & L\\
+			Ludwig & 2, 3 & Y & Y & Y & Y & Y & Y & L\\
+			LUMA & 2, 3 & Y & Y & Y & Y & Y & Y & W, M, L\\
+			MechSys & 2, 3 & N & Y & Y & Y & Y & Y & L\\
+			MP-LABS & 2, 3 & Y & N & Y & Y & N & N & L\\
+			OpenLB & 1, 2, 3 & Y & Y & Y & Y & Y & Y & W, M, L\\
+			Palabos & 2, 3 & Y & Y & Y & Y & Y & Y & W, L\\
+			pyLBM & 1, 2, 3 & Y & Y & Y & N & Y & Y & W, M, L\\
+			Sailfish & 2, 3 & Y & Y & Y & Y & Y & Y & M, L\\
+			SunlightLB & 3 & N & Y & Y & N & N & Y & L\\
+			TCLB & 2, 3 & Y & Y & Y & Y & Y & Y & L\\
+			waLBerla & 2, 3 & Y & Y & Y & Y & Y & Y & L\\
 			\hline
 		\end{tabular}
-        \caption{Features of Software Packages (Dim for Dimension (1, 2 or 3),
+		\caption{Features of Software Packages (Dim for Dimension (1, 2, 3),
 			Pll for Parallel (Yes or No), Com for Compressible (Yes or No), Rflx for
-        Reflexive boundary condition (Yes or No), MFl for Multifluid (Yes or
-        No), NN for Non-Newtonian, OS (Win, Mac, Lin, )} \label{tbl_features}
+			Reflexive Boundary Condition (Yes or No), MFl for Multifluid (Yes or
+			No), Turb for Turbulent (Yes or No), CGE for Complex Geometries (Yes or No), OS for Operating System (Windows (W), macOS (M), Linux (L)))} \label{tbl_features}
 	\end{center}

### 2021-12-13 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
specified parallel options

--- 2021-12-12 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
@@
 	\begin{center}
-        \begin{tabular}{ p{3cm}lllllll}
-            \hline
-            Name & Dim & Pll & Com & Rflx & Mfl & NN & OS\\
-            \hline
-            DL\_MESO (LBE) & ? & ? & ? & ? & ? & ? & ?\\
-            ESPResSo & ? & ? & ? & ? & ? & ? & ?\\
-            ESPResSo++ & ? & ? & ? & ? & ? & ? & ?\\
-            lbmpy & ? & ? & ? & ? & ? & ? & ?\\
-            lettuce & ? & ? & ? & ? & ? & ? & ?\\
-            Ludwig & ? & ? & ? & ? & ? & ? & ?\\
-            LUMA & ? & ? & ? & ? & ? & ? & ?\\
-            MechSys & ? & ? & ? & ? & ? & ? & ?\\
-            OpenLB & ? & ? & ? & ? & ? & ? & ?\\
-            Palabos & ? & ? & ? & ? & ? & ? & ?\\
-            pyLBM & ? & ? & ? & ? & ? & ? & ?\\
-            Sailfish & ? & ? & ? & ? & ? & ? & ?\\
-            TCLB & ? & ? & ? & ? & ? & ? & ?\\
-            waLBerla & ? & ? & ? & ? & ? & ? & ?\\
-            HemeLB & ? & ? & ? & ? & ? & ? & ?\\
-            laboetie & ? & ? & ? & ? & ? & ? & ?\\
-            LatBo.jl & ? & ? & ? & ? & ? & ? & ?\\
-            LB2D-Prime & ? & ? & ? & ? & ? & ? & ?\\
-            LB3D & ? & ? & ? & ? & ? & ? & ?\\
-            LB3D-Prime & ? & ? & ? & ? & ? & ? & ?\\
-            LIMBES & ? & ? & ? & ? & ? & ? & ?\\
-            MP-LABS & ? & ? & ? & ? & ? & ? & ?\\
-            SunlightLB & ? & ? & ? & ? & ? & ? & ?\\
-
+		\begin{tabular}{ p{3cm}llllllll}
+			\hline
+			Name & Dim & Pll & Com & Rflx & MFl & Turb & CGE & OS\\
+			\hline
+			DL\_MESO (LBE) & 2, 3 & MPI/OMP & Y & Y & Y & Y & Y & W, M, L\\
+			ESPResSo & 1, 2, 3 & CUDA/MPI & Y & Y & Y & Y & Y & M, L\\
+			ESPResSo++ & 1, 2, 3 & MPI & Y & Y & Y & Y & Y & L\\
+			HemeLB & 3 & MPI & Y & Y & Y & Y & Y & L\\
+			laboetie & 2, 3 & MPI & Y & Y & Y & Y & Y & L\\
+			LatBo.jl & 2, 3 & - & Y & Y & Y & N & Y & L\\
+			LB2D-Prime & 2 & MPI & Y & Y & Y & Y & Y & W, L\\
+			LB3D & 3 & MPI & N & Y & Y & Y & Y & L\\
+			LB3D-Prime & 3 & MPI & Y & Y & Y & Y & Y & W, L\\
+			lbmpy & 2, 3 & CUDA & Y & Y & Y & Y & Y & L\\
+			lettuce & 2, 3 & CUDA & Y & Y & Y & Y & Y & W, M, L\\
+			LIMBES & 2 & OMP & Y & Y & N & N & Y & L\\
+			Ludwig & 2, 3 & MPI & Y & Y & Y & Y & Y & L\\
+			LUMA & 2, 3 & MPI & Y & Y & Y & Y & Y & W, M, L\\
+			MechSys & 2, 3 & - & Y & Y & Y & Y & Y & L\\
+			MP-LABS & 2, 3 & MPI/OMP & N & Y & Y & N & N & L\\
+			OpenLB & 1, 2, 3 & MPI & Y & Y & Y & Y & Y & W, M, L\\
+			Palabos & 2, 3 & MPI & Y & Y & Y & Y & Y & W, L\\
+			pyLBM & 1, 2, 3 & MPI & Y & Y & N & Y & Y & W, M, L\\
+			Sailfish & 2, 3 & CUDA & Y & Y & Y & Y & Y & M, L\\
+			SunlightLB & 3 & - & Y & Y & N & N & Y & L\\
+			TCLB & 2, 3 & CUDA/MPI & Y & Y & Y & Y & Y & L\\
+			waLBerla & 2, 3 & MPI & Y & Y & Y & Y & Y & L\\
 			\hline
 		\end{tabular}
-        \caption{Features of Software Packages (Dim for Dimension (1, 2 or 3),
-        Pll for Parallel (Yes or No), Com for Compressible (Yes or No), Rflx for
-        Reflexive boundary condition (Yes or No), MFl for Multifluid (Yes or
-        No), NN for Non-Newtonian, OS (Win, Mac, Lin, )} \label{tbl_features}
+		\caption{Features of Software Packages (Dim for Dimension (1, 2, 3),
+			Pll for Parallel (CUDA, MPI, OpenMP (OMP)), Com for Compressible (Yes or No), Rflx for
+			Reflexive Boundary Condition (Yes or No), MFl for Multifluid (Yes or
+			No), Turb for Turbulent (Yes or No), CGE for Complex Geometries (Yes or No), OS for Operating System (Windows (W), macOS (M), Linux (L)))} \label{tbl_features}
 	\end{center}

### 2021-12-14 StateOfPractice/Papers/MedImageSoft_SOP-arXiv/MedImageSoft_SOP_AcademicRadiology.tex

@@
 previously used a different approach to version control and eventually
-transferred to git and GitHub.  The past approaches included contributions
-from e-mail (3 teams), contributions from forums (1 team) and sharing the
-git repository via e-mail (1 team).  \wss{I do not know what this last one means.
-Should ask Ao.}
+transferred to git and GitHub.  The past approaches included contributions from
+e-mail (3 teams), contributions from forums (1 team) and the unconventional
+approach of e-mailing a git repository between team members because at the time
+no free host git servers were available (1 team).
 
@@
 \item Lack of real world image data for testing;
-\item The team cannot use private data for debugging even when the data causes
-problems; \wss{Why not?  Should ask Ao}
+\item The team is told (say by a physician) of data that causes a problem, but
+because of data privacy concerns, the developers are not allowed access to the
+data for debugging.
 \item Expense and time consumed by tests because of the huge datasets;
@@
 classes of users: the end users and other developers.  The threats to
-understandability for end users include users not understanding how to use
-features (\wss{Because the users don't have sufficient background?  Ask Ao.}),
-the team not having a dedicated user experience (UX) designer, some important
-indicators are not noticeable (e.g. a progress bar), not all users understand
-the purpose of the software, not all users know if the software includes certain
-features, not all users understand how to use the command line tool, not all
-users understand that the software is a web application. For developers the
-threats to understandability include developers not understanding how to deploy
-the software (\wss{This sounds like a consequence of a lack of
-understandability, rather than threat to it.  Or is the point that developers
-lack sufficient background?  Ask Ao}), and the architecture is difficult for new
-developers to understand.
+understandability for end users include lack of documentation, an nonintuitive
+process, the team not having a dedicated user experience (UX) designer,
+important indicators not being noticeable (e.g. a progress bar), not all users
+understanding the purpose of the software, not all users knowing if the software
+includes certain features, not all users understanding how to use the command
+line tool, not all users understanding that the software is a web application.
+For developers the threats to understandability include unclear instructions on
+software deployment, and the architecture being difficult for new developers to
+understand.
 
@@
 threats that were mentioned include closed-source software, no user interaction
-tests, no unit tests, using different versions of some common libraries
-\wss{because the interface and/or behaviour of the library could have changed?
-Ask Ao}, variability between CPUs, and misinterpretation of how manufacturers
-create medical images. The most commonly cited (by 6 teams) strategy to improve
-reproducibility was testing (regression tests, unit tests, having good tests).
-The second most common strategy (mentioned by 5 teams) is making code, data, and
-documentation available, possibly by creating open-source libraries.  Other
-ideas that were mentioned include running the same tests on all platforms, a
-dockerized version of the software to insulate it from the OS environment, using
-standard libraries, monitoring the upgrades of the library dependencies, clearly
-documenting the version information, bringing along the exact versions of all
-the dependencies with the software, providing checksums of the data, and
-benchmarking the software against other software that overlaps in functionality.
-Specifically one interviewee suggested using \textit{3D Slicer} as the benchmark
-to test their reproducibility.
+tests, no unit tests, using different versions of some common libraries (causing
+a changes to the library's interface), variability between CPUs, and
+misinterpretation of how manufacturers create medical images. The most commonly
+cited (by 6 teams) strategy to improve reproducibility was testing (regression
+tests, unit tests, having good tests). The second most common strategy
+(mentioned by 5 teams) is making code, data, and documentation available,
+possibly by creating open-source libraries.  Other ideas that were mentioned
+include running the same tests on all platforms, a dockerized version of the
+software to insulate it from the OS environment, using standard libraries,
+monitoring the upgrades of the library dependencies, clearly documenting the
+version information, bringing along the exact versions of all the dependencies
+with the software, providing checksums of the data, and benchmarking the
+software against other software that overlaps in functionality. Specifically one
+interviewee suggested using \textit{3D Slicer} as the benchmark to test their
+reproducibility.
 
@@
 to real-world datasets for testing. The developers' strategies to address this
-are summarized in Section~\ref{sec_pain_points_3}. One threat to correctness is:
-with huge datasets for testing, the tests are expensive and time-consuming.
-Three interviewees endorsed self tests / automated tests, which may save time
-for testing.
+are summarized in Section~\ref{sec_interview_pain_points}. One threat to
+correctness is: with huge datasets for testing, the tests are expensive and
+time-consuming. Three interviewees endorsed self tests / automated tests, which
+may save time for testing.
 

### 2021-12-14 StateOfPractice/Papers/MedImageSoft_SOP-arXiv/MedImageSoft_SOP_AcademicRadiology.tex
Clarified RQs apply to MI software

--- 2021-12-11 StateOfPractice/Papers/MedImageSoft_SOP-arXiv/MedImageSoft_SOP_AcademicRadiology.tex
@@
 
-\item[RQ1.] \hypertarget{rq1}What artifacts are present in current software
+\item[RQ1.] \hypertarget{rq1}What artifacts are present in current MI software
 projects? What role does documentation play in the projects? What are the
 developers' attitude toward it?
-\item[RQ2.] \hypertarget{rq2}What tools are used in the development of current
-software packages?
+\item[RQ2.] \hypertarget{rq2}What tools are used in the development of MI packages?
 \item[RQ3.] \hypertarget{rq3}What principles, processes, and methodologies are
-used in the development of current software packages?
+used in the development of current MI packages?
 \item[RQ4.] \hypertarget{rq4}What are the pain points for developers working on
-research software projects? What aspects of the existing processes,
-methodologies, and tools do they consider as potentially needing improvement?
-What changes to processes, methodologies, and tools can improve software
-development and software quality?
+MI software projects? What aspects of the existing processes, methodologies, and
+tools do they consider as potentially needing improvement? What changes to
+processes, methodologies, and tools can improve software development and
+software quality?
 \item[RQ5.] \hypertarget{rq5}What is the current status of the following
-software qualities for the projects? What actions have the developers taken to
+software qualities for LBM projects? What actions have the developers taken to
 address them?  (Section~\ref{ch_background} provides definitions of the software
@@
 \centering
-\includegraphics[scale=0.6]{figures/FunctionsOfMiSoftware.pdf}
+\includegraphics[scale=0.5]{figures/FunctionsOfMiSoftware.pdf}
 \caption{Major functions of MI Software} \label{fig_mi_functions}
@@
 \centering
-\includegraphics{figures/PrimaryLanguages.pdf}
+\includegraphics[scale=0.7]{figures/PrimaryLanguages.pdf}
 \centering
@@
 Therefore, for \textit{DICOM Viewer} we could not measure reliability or
-robustness.  The other seven qualities could be measured, since they do not require installation.
+robustness.  The other seven qualities could be measured, since they do not
+require installation.
 

### 2021-12-14 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
update RQs to be specific to LBM

@@
 \begin{enumerate}
-	\item What artifacts are present in current software packages? 
+	\item What artifacts are present in current LBM packages? 
 	\item What tools (development, dependencies, project management) are used by
-	current software packages?
+	current LBM solvers?
 	\item What principles, processes, and methodologies are used in the development
-	of current software packages?
-	\item What are the pain points for developers working on research software
+	of current LBM packages?
+	\item What are the pain points for developers working on LBM software
 	projects?  What aspects of the existing processes, methodologies and tools do
@@
 	software quality?
-	\item For research software developers, what specific actions are taken to
+	\item For LBM software developers, what specific actions are taken to
 	address the following:
@@
 
-Table~\ref{} presents a summary of the software that was measured.  The table
-shows the differences between the packages based on features.  \wss{Add some
-discussion of some example features.}  The sections below summarize how the
-software packages compare based on qualities, as measured by the measurement
-template.   The results of the AHP ranking are also discussed.
+Before discussing quality differences between the packages and the results of
+the AHP ranking, we will contrast the different packages based on features, as
+shown in Table~\ref{tbl_features}. \wss{Add some discussion of some example
+features.  Add Non-Newtonian and Fluid Structure interaction.  Listed in
+alphabetical order.}  The sections below summarize how the software packages
+compare based on qualities, as measured by the measurement template.   The
+results of the AHP ranking are also discussed.
 

### 2021-12-14 StateOfPractice/Papers/MedImageSoft_SOP-arXiv/MedImageSoft_SOP_AcademicRadiology.tex
Added a bit more of an introduction to the RQs

@@
 MI, as well as to other domains, of research software \citep{SmithEtAl2021}. The
-first question below refers to artifacts.  We use this term to represent the
-documents, scripts and code that constitutes a software development project.
-Example artifacts include requirements, specifications, user manuals, unit
-tests, system tests, usability tests, build scripts, API (Application
-Programming Interface) documentation, READMEs, license documents, process
-documents, and code.
+questions are designed to learn about the community's interest in, and
+experience with, artifacts, tools, principles, processes, methodologies and
+qualities.  When we mention artifacts we mean documents, scripts and code that
+constitutes a software development project. Example artifacts include
+requirements, specifications, user manuals, unit tests, system tests, usability
+tests, build scripts, API (Application Programming Interface) documentation,
+READMEs, license documents, process documents, and code.  We aimed to collect
+enough information to understand the current pain points experienced by the MI
+software development community so that we can make some preliminary
+recommendations for future improvements.
 

### 2021-12-14 StateOfPractice/Papers/MedImageSoft_SOP-arXiv/MedImageSoft_SOP_AcademicRadiology.tex
Simplified the scope section to explicitly state visualization is our scope, and then list what is excluded.  Removed figure.

@@
 
-We need to restrict the scope of MI software that we consider so that the
-measurement exercise is feasible within the targeted person month of effort. We
-can restrict the scope by selecting a sub-set of the broad range of problems
-considered by MI.  Summing the sets of MI tasks from \citet{Bankman2000} and
-\citet{Angenent2006}, we have the following problems are solved by MI software:
+To make the project feasible, we only cover MI visualization software.  As a
+consequence we are excluding many other categories of MI software, including
 Segmentation, Registration, Visualization, Enhancement, Quantification,
-Simulation, plus three functions for MI archiving and telemedicine systems
-(Compression, Storage, and Communication). To this list Wikipedia contributors
-\citep{enwiki:1034877594} add Statistical Analysis and Image-based Physiological
-Modelling. \citet{Kim2011} adds Feature Extraction, Classification, and
-Interpretation. Besides the above major functions, some MI software provides
-supportive functions. For example, using Tool Kit libraries VTK
-\citep{SchroederEtAl2006} and ITK \citep{McCormick2014}, developers build
-software with Visualization and Analysis functions. Finally, Picture Archiving
-and Communication System (PACS) helps users to economically store and
-conveniently access images \citep{Choplin1992}.
-
-Based on our literature survey, we divided MI software into five sub-groups and
-several sub-sub-groups by their major functions, as shown in
-Figure~\ref{fig_mi_functions}. We selected the sub-group of Visualization for
-the current state of the practice exercise.
-
-\begin{figure}[!ht]
-\centering
-\includegraphics[scale=0.5]{figures/FunctionsOfMiSoftware.pdf}
-\caption{Major functions of MI Software} \label{fig_mi_functions}
-\end{figure}
+Simulation, plus MI archiving and telemedicine systems (Compression, Storage,
+and Communication) (as summarized by \citet{Bankman2000} and
+\citet{Angenent2006}).  We also exclude Statistical Analysis and Image-based
+Physiological Modelling \citep{enwiki:1034877594} and Feature Extraction,
+Classification, and Interpretation \citep{Kim2011}. Software that provides MI
+support functions is also out of scope; therefore, we have not assessed the
+toolkit libraries VTK \citep{SchroederEtAl2006} and ITK \citep{McCormick2014}.
+Finally, Picture Archiving and Communication System (PACS), which helps users to
+economically store and conveniently access images \citep{Choplin1992}, are
+considered out of scope.
+
+% Based on our literature survey, we divided MI software into five sub-groups and
+% several sub-sub-groups by their major functions, as shown in
+% Figure~\ref{fig_mi_functions}.
+
+% \begin{figure}[!ht]
+% \centering
+% \includegraphics[scale=0.5]{figures/FunctionsOfMiSoftware.pdf}
+% \caption{Major functions of MI Software} \label{fig_mi_functions}
+% \end{figure}
 

### 2021-12-14 StateOfPractice/AHP2020/LBM/Source code/DataProcess/src/Main.java
updated AHP results to include Musubi

@@ -15,7 +15,7 @@ public class Main {
 	public static void main(String[] args)
     {
 		int numberQualities = 9;
-		int numberSoftwarePackages = 23; //also need to change rownum below
+		int numberSoftwarePackages = 24; //also need to change rownum below
 		Cell[][] table=new Cell[numberQualities][numberSoftwarePackages];
 		Double[][][] ahpTable= new Double[numberQualities][numberSoftwarePackages][numberSoftwarePackages-1];
         try

### 2021-12-17 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Update to LBM arXiv paper to use wider and shorter charts for the AHP results

@@
 
-We requested interviews from all 24 packages, with 4 positive responses.  To
-send out interview requests we found we found contacts on the projects’ website,
-or code repository, or publications, or the biographic pages of the teams’
-institutions. We send at most two interview request emails to a contact for each
-software package.  Meeting will typically be held using on-line meeting
-software, like Zoom or Teams.  This facilitates recording and automatic
-transcription of the meetings.
+We requested interviews from all 24 packages, except Munubi, since Munubi was
+added after our ethics approved study window had expired.  We were able to
+recruit 4 developers to participate in our study.  To send out interview
+requests we found we found contacts on the projects’ website, or code
+repository, or publications, or the biographic pages of the teams’ institutions.
+We send at most two interview request emails to a contact for each software
+package.  Meeting will typically be held using on-line meeting software, like
+Zoom or Teams.  This facilitates recording and automatic transcription of the
+meetings.
 
@@
 	\begin{center}
-		\includegraphics[width=1.0\textwidth]{./figures/installability_chart}
+		\includegraphics[width=1.0\textwidth]{./figures/installability_chart.pdf}
 		\caption{AHP Installability Score}
@@
 
-The interview process
-presented here was approved by the McMaster University Research Ethics Board
-under the application number 
-\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/MACREM/Application.pdf}
-{MREB\#: 5219}.
-
-This section presents the qualitative findings from data that was gathered
-during interviews with developers for qualities listed in Section
-~\ref{softwarequalities}. The interview questions are found in Appendix
-~\ref{interviewquestions}.
-
-\subsection{Surface Correctness and Verifiability}
+This section presents qualitative findings on software qualities as gleaned from
+interviewing 4 developers. The full interview questions are found in
+\citet{SmithEtAl2021}.
+
+\subsection{Correctness and Verifiability}
 
@@
 by domain experts with backgrounds in physics, mathematics and mechanical
-engineering. It was noted in these interviews that some of the developers do not
-have formal software engineering education. Some of the development teams
-include computer scientists. Despite a lack of visible domain documentation and
-a resulting lower surface correctness and verifiability score, it is clear that
-some of the software packages were developed by teams with significant domain
-knowledge on account of the academic backgrounds of their developers.
+engineering. \wss{Could point to surveys that have shown the same thing.}  It
+was noted in these interviews that some of the developers do not have formal
+software engineering education. Some of the development teams include computer
+scientists. Despite a lack of visible domain documentation and a resulting lower
+surface correctness and verifiability score, it is clear that some of the
+software packages were developed by teams with significant domain knowledge on
+account of the academic backgrounds of their developers.
 
@@
 integration in the development processes than what was observed from the initial
-survey. For example, OpenLB, pyLBM, and TCLB use such methods during development
+survey. For example, OpenLB, pyLBM, and TCLB use such methods during development,
 despite this not being explicitly clear from an analysis of the material
-available online. The correctness and verifiability of such packages is not
-measured well using surface analysis.
-
-Several interviewed developers alluded to difficulty with testing the
-correctness of large numbers of features, and some even manually tested program
-output. The use of well defined unit testing tools could decrease the time spent
-testing some feature.
-
-\subsection{Surface Usability}
+available online. \wss{Add this to threats to validity.}  The correctness and
+verifiability of such packages is not measured well using surface analysis.
+
+Several developers alluded to difficulty with testing the correctness of large
+numbers of features, and some even manually tested program output. The use of
+well defined unit testing tools could decrease the time spent testing some
+feature.  \wss{link this to recommendations}
+
+\subsection{Usability}
 
@@
 
-\subsection{Surface Understandability}
+\subsection{Understandability}
 
@@
 - data on Mendeley
+
+- add a comparison to other studies
+- add conclusions related to how LBM compares to other fields, and the "average" of all SC
+
+- make recommendations in order of conservative to ``bleeding edge''
+- add pointer to project checklists (as sent to Jacques, as promoted during workshop)
+
+- value of tools to enforce good behaviour of developers, like R, like MOOSE infrastructure

### 2021-12-22 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Updated paper to use the new better aspect ratio figures
Updates up to and including section 4

@@
 
-\section{Introduction}
+\section{Introduction} \label{secIntro}
 
@@
 incorporating additional physical complexities \citep{chen1998lattice} such as
-reflective and non-reflective boundaries. They consider the behaviour of a
+reflective and non-reflective boundaries. They consider the behavior of a
 collection of particles as a single unit at the mesoscopic scale, between the
@@
 A small sample of important applications of LBM include the following: designing
-fuel cells \citep{ZhangEtAl2018}, modelling groundwater flow
+fuel cells \citep{ZhangEtAl2018}, modeling groundwater flow
 \citep{AnwarAndSukop2009}, and analyzing the flow of blood in the cardiovascular
@@
 \begin{enumerate}
-	\item List candidate software packages for the domain. This is discussed in Section~\ref{identifysoftware}.
-	\item Filter the software package list. This is discussed in Section~\ref{filtersoftware}.
+	\item List candidate software packages for the domain. This is discussed in
+	Section~\ref{identifysoftware}.
+	\item Filter the software package list. This is discussed in
+	Section~\ref{filtersoftware}.
 	\item Gather the source code and documentation for each software package.
-	\item Collect quantitative measures from the project repositories. This is discussed in Section~\ref{empiricalmeasures}.
-	\item Measure using the measurement template. This is discussed in Section~\ref{empiricalmeasures}. The full measurement template can be found in \citet{SmithEtAl2021}.
+	\item Collect quantitative measures from the project repositories. This is
+	discussed in Section~\ref{empiricalmeasures}.
+	\item Measure using the measurement template. This is discussed in
+	Section~\ref{empiricalmeasures}. The full measurement template can be found
+	in \citet{SmithEtAl2021}.
 	\item Survey the developers. The developers of each software package in the
-	filtered software list were contacted for voluntary interviews. The details are given in Section~\ref{SecSurvey}.
-	\item Use AHP to rank the software packages. This is discussed in Section~\ref{AHP}.
-	\item Analyze the results and answer the research questions. The answers can be found in Section~\ref{answersquestions}.
+	filtered software list were contacted for voluntary interviews. The details
+	are given in Section~\ref{SecSurvey}.
+	\item Use AHP to rank the software packages. This is discussed in
+	Section~\ref{AHP}.
+	\item Analyze the results and answer the research questions. The answers can
+	be found in Section~\ref{answersquestions}.
 \end{enumerate}
@@
 general information, such as the software name, number of developers, etc.  We
-follow the definitions given by \citet{GewaltigAndCannon2012} for the software
+follow the definitions given by \citet{gewaltig2012quality} for the software
 categories.  Public means software intended for public use.  Private means
@@
 recruit 4 developers to participate in our study.  To send out interview
-requests we found we found contacts on the projects’ website, or code
-repository, or publications, or the biographic pages of the teams’ institutions.
-We send at most two interview request emails to a contact for each software
-package.  Meeting will typically be held using on-line meeting software, like
-Zoom or Teams.  This facilitates recording and automatic transcription of the
-meetings.
+requests we found we found contacts on the projects' website, or code
+repository, or publications, or the biographic pages of the teams'
+institutions. We send at most two interview request emails to a contact for each
+software package.  Meeting will typically be held using on-line meeting
+software, like Zoom or Teams.  This facilitates recording and automatic
+transcription of the meetings.
 
@@
 			Pll for Parallel (CUDA, MPI, OpenMP (OMP)), Com for Compressible (Yes or No), Rflx for
-			Reflexive Boundary Condition (Yes or No), MFl for Multifluid (Yes or
+			Reflexive Boundary Condition (Yes or No), MFl for Multi-fluid (Yes or
 			No), Turb for Turbulent (Yes or No), CGE for Complex Geometries (Yes or No), OS for Operating System (Windows (W), macOS (M), Linux (L)))} \label{tbl_features}
@@
 	\begin{center}
-		\includegraphics[width=1.0\textwidth]{./figures/correctnessverifiability_chart}
+		\includegraphics[width=1.0\textwidth]{./figures/correctnessverifiability.pdf}
 		\caption{AHP Surface Correctness and Verifiability Score}
@@
 	\begin{center}
-		\includegraphics[width=1.0\textwidth]{./figures/reliability_chart}
+		\includegraphics[width=1.0\textwidth]{./figures/reliability_chart.pdf}
 		\caption{AHP Surface Reliability Score}
@@
 	\begin{center}
-		\includegraphics[width=1.0\textwidth]{./figures/robustness_chart}
+		\includegraphics[width=1.0\textwidth]{./figures/robustness_chart.pdf}
 		\caption{AHP Surface Robustness Score}
@@
 	\begin{center}
-		\includegraphics[width=1.0\textwidth]{./figures/usability_chart}
+		\includegraphics[width=1.0\textwidth]{./figures/usability_chart.pdf}
 		\caption{AHP Surface Usability Score}
@@
 	\begin{center}
-		\includegraphics[width=1.0\textwidth]{./figures/maintainability_chart}
+		\includegraphics[width=1.0\textwidth]{./figures/maintainability_chart.pdf}
 		\caption{AHP Maintainability Score}
@@
 assumption that this indicates increased modularization. Some packages have more
-features than others.  This is assumed to contribute to reusablility, since they have
+features than others.  This is assumed to contribute to reusability, since they have
 more source code for potential reuse. The software packages were also reviewed
@@
 	\begin{center}
-		\includegraphics[width=1.0\textwidth]{./figures/reusability_chart}
+		\includegraphics[width=1.0\textwidth]{./figures/reusability_chart.pdf}
 		\caption{AHP Reusability Score}
@@
 	\begin{center}
-		\includegraphics[width=1.0\textwidth]{./figures/understandability_chart}
+		\includegraphics[width=1.0\textwidth]{./figures/understandability_chart.pdf}
 		\caption{AHP Understandability Score}
@@
 	\begin{center}
-		\includegraphics[width=1.0\textwidth]{./figures/visibilitytransparency_chart}
+		\includegraphics[width=1.0\textwidth]{./figures/visibilitytransparency_chart.pdf}
 		\caption{AHP Visibility and Transparency Score}
@@
 	\centering
-		\includegraphics[width=1.0\textwidth]{./figures/finalscore_chart}
+		\includegraphics[width=1.0\textwidth]{./figures/finalscore_chart.pdf}
 		\caption{AHP Overall Score}
@@
 
+, without a need for a rigorous software development model. The developer of
+ESPResSo noted that while no formal model is used, their development model is
+something similar to a combination of agile and waterfall development models.
+
+Improving the software engineering
+education or experience of developers is also an idea that was brought up by
+several developers.
+
 This section presents qualitative findings on software qualities as gleaned from
@@
 was noted in these interviews that some of the developers do not have formal
-software engineering education. Some of the development teams include computer
-scientists. Despite a lack of visible domain documentation and a resulting lower
-surface correctness and verifiability score, it is clear that some of the
-software packages were developed by teams with significant domain knowledge on
-account of the academic backgrounds of their developers.
+software engineering education. \wss{Can point to other studies to support
+this.} Some of the development teams include computer scientists. Despite a lack
+of visible domain documentation and a resulting lower surface correctness and
+verifiability score, it is clear that some of the software packages were
+developed by teams with significant domain knowledge.
 
@@
 Several developers alluded to difficulty with testing the correctness of large
-numbers of features, and some even manually tested program output. The use of
-well defined unit testing tools could decrease the time spent testing some
-feature.  \wss{link this to recommendations}
+numbers of features, and some even manually tested program output, as opposed to
+using automated testing. The use of well defined unit testing tools could
+decrease the time spent testing some feature.  \wss{link this to
+recommendations}
 
@@
 limitations, and of the requirements to properly model appropriate problems,
-including what algorithms and parameters to use.
-
-Some additional but infrequent software usability issues were commented on by
-the developers. Users have had trouble with installation and understanding how
-to maneuver the interfaces and how to set up or run models. These issues are
-addressed by various user support models, including frequently asked questions
-sections on the software package websites, user guides, and hardware and
-software requirements specifications.
+including what algorithms and parameters to use.  \wss{Could point to this for
+recommendations related to documentation.  Could point to this for a generative
+approach that won't allow inadmissible combinations, or at least includes an
+explicit warning.}
+
+Usability concerns were also raised for installation, understanding how to
+maneuver the interfaces and understanding how to set up and run models. These
+issues are addressed by various user support models, including frequently asked
+questions sections on the software package websites, user guides, and hardware
+and software requirements specifications.  \wss{Could point to there being room
+for improvement here, especially for requirements specification.}
 
@@
 in terms of usability over the years'', further commenting that ``most people in
-the field know [Python]'' and that ``it's easy to learn''.
+the field know [Python]'' and that ``it's easy to learn''.  \wss{A generative
+approach facilitates switching to whatever language is currently popular.}
 
@@
 
-Interviews with developers revealed that most projects do have a defined process
-for accepting contributions from team members. The packages rarely get
-contributions from outside developers, but the process would be similar as for
-the aforementioned group.
-
-Contributions are made through GitHub, and are then reviewed and pulled by lead
-developers, often with consultation with a group of core developers depending on
-the organizational model. Continuous integration is part of the process for some
-packages.
+Although not always visible from the publicly available artifacts, interviews
+with developers revealed that most projects have a defined process for accepting
+contributions from team members. The packages rarely get contributions from
+outside developers, but the process would be similar as for the aforementioned
+group.  In general, contributions are made through GitHub, and are then reviewed
+and pulled by lead developers, often with consultation with a group of core
+developers depending on the organizational model. Continuous integration is part
+of the process for some packages. \wss{Recommendation to increase.  Successful
+projects use it, a reasonable next step for projects that have not yet
+incorporated CI.}
 
@@
 the near future. They would be happy to see contribution from outside of their
-organization, but currently this does not happen.
+organization, but currently this does not happen.  \wss{A comment was made at
+the recent DOE workshop that you get new users through documentation and
+tutorials.  Maybe highlight the example of Musubi?  Should trace back to this
+for the recommendations part.}
 
@@
 using abstraction and modularization of the source code, to make it ``very easy
-to add [new] features''.
+to add [new] features''.  \wss{Highlights the advantage of design.  Other
+projects can look to these examples.  Maybe cite the Anshu paper about Flash?
+Dune?  and point to the advantages of design.}
+
+The developer of TCLB mentioned that their package had two sets of code, for
+executing the models on the CPU and GPU, and that maintenance was decreased by
+introducing macros, a practice which became a common part of the development
+process.
 
@@
 considered and that measures to ensure it had been taken, including requiring
-the separation of software components in the source code architecture.
-
-A high degree of code modularity and abstraction was noted by developers as a
-measure to ensure the ease of future changes. This can be ensured by separating
-components and hiding information behind well defined interfaces. The developer
-of ESPResSo also noted that some of the code base was transitioned from C to
-C++, which could ease modifiability of that software package. The developer of
-TCLB noted that their software package was designed to allow for the addition of
-some LBM features, but changes to major aspects of the system would be
-difficult. For example, ``implementing a new model will be an easy
-contribution'', but changes to the ``Cartesian mesh...will be a nightmare''.
-Furthermore, the package was designed with flexible data structures and storage
-in mind.
+the separation of software components in the source code architecture.  \wss{A
+good approach for other developers to adopt - part of recommendations.  Trace
+back to this comment.}  A high degree of code modularity and abstraction was
+noted by developers as a measure to ensure the ease of future changes. This can
+be ensured by separating components and hiding information behind well defined
+interfaces. The developer of ESPResSo also noted that some of the code base was
+transitioned from C to C++, which could ease modifiability of that software
+package. The developer of TCLB noted that their software package was designed to
+allow for the addition of some LBM features, but changes to major aspects of the
+system would be difficult. For example, ``implementing a new model will be an
+easy contribution'', but changes to the ``Cartesian mesh...will be a
+nightmare''. Furthermore, the package was designed with flexible data structures
+and storage in mind.
 
@@
 package identifies as a development framework for modeling problems in various
-CFD areas.
+CFD areas.  \wss{Add validation benchmarks as a ``best practice'' used by
+Palabos - others could emmulate.  Goal of this exercise is for people in the
+community to learn from each other, and potentially from our advice too.}
+
+Developers were asked to comment on the obstacles in their development process.
+The developer of ESPResSo noted that a lot of their source code had been written
+with a specific application in mind, and that there is too much coupling between
+components. Addressing this issue, possibly by making the design review process
+more transparent (Section~\ref{secVisib}), would help with code modifiability
+and reusability.  \wss{Traceback for recommendations.}
 
@@
 Software developers noted that they believe users have generally found their
-packages to be understandable. The interviewed developer of ESPResSo commented
-that some users have attempted to run physically incompatible LBM methods, and
-the solution was to edit the code to prevent such combinations, as well as to
-update the documentation to prevent misunderstanding the methods. Similarly, a
-developer of pyLBM noted that some users had issues setting up parameters for
-LBM schemes. The solution to this was to update the interface where these
-parameters are set, as well as to add functionality to test the stability of the
-parameters. A developer of OpenLB noted that some users lack the background
-knowledge to easily model fluid dynamics problems using their software. A
-frequently asked questions section was added to their package website to help
-users find answers to common questions. The package also has detailed
-documentation, including guides and usage requirements specification, to better
-help users understand the software.
+packages to be understandable. \wss{Confusion between understandability and
+usability - threat to validity example.}  The interviewed developer of ESPResSo
+commented that some users have attempted to run physically incompatible LBM
+methods, and the solution was to edit the code to prevent such combinations, as
+well as to update the documentation to prevent misunderstanding the methods.
+Similarly, a developer of pyLBM noted that some users had issues setting up
+parameters for LBM schemes. The solution to this was to update the interface
+where these parameters are set, as well as to add functionality to test the
+stability of the parameters. A developer of OpenLB noted that some users lack
+the background knowledge to easily model fluid dynamics problems using their
+software. A frequently asked questions section was added to their package
+website to help users find answers to common questions. The package also has
+detailed documentation, including guides and usage requirements specification,
+to better help users understand the software.  \wss{Peter - the discussion here
+is more about user usability than the understandability of the code for
+developers.  Did any of the developers discuss developer understandability?  The
+confusion between understandability and usability was also a confusion during
+the interviews Ao conducted.}
+
+\wss{I moved this from the visibility section, since it is more about
+understandability.}  Developers were also asked how documentation fits into their
+development process. Several developers noted that developer documentation plays
+an important role in familiarizing potential contributors to the software system
+architecture. Without the guidance that the documentation provides it would be
+unlikely that contributions would pass the peer review process.
 
@@
 package build system works, and how the coupling between the simulation engine
-and the interface works.
+and the interface works.  \wss{Peter - these points about developer
+documentation seem like they would fit better under understandability, don't you
+think?}  \wss{If ESPresSo scored high (I think it did), we can use that to
+further promote the value of documentation.  Also a threat to validity is that
+we have implicitly assumed documentation is valuable when measuring quality.}
 
@@
 or algorithms. This is necessary so that the coding standard for these models is
-kept consistent with new developers.
+kept consistent for new developers. \wss{Peter - this also sounds like
+discussion that is relevant for understandability.}
 
@@
 developers. Finding ways to hasten updating documentation would increase the
-frequency of such updates and benefit both users and developers.
-
-The developer of OpenLB noted the use of documentation generators like Doxygen.
-It would be advisable for more projects to use such automatic document
-generation tools, since some projects do not do this.
-
-\subsection{Visibility and Transparency}
-
-Developers were asked to comment on the obstacles in their development process.
-The developer of ESPResSo noted that a lot of their source code had been written
-with a specific application in mind, and that there is too much coupling between
-components. Addressing this issue would help with code modifiability and
-reusability. Updating the development process would help resolve this issue and
-prevent such issues in the future. Improving the visibility of software changes
-and a peer review process would also help. Improving the software engineering
-education or experience of developers is also an idea that was brought up by
-several developers. Specifically, developers should always write code that is
-decoupled and modular, and should keep in mind the visibility of their
-contributions by better updating documentation and ensuring that their
-contributions are transparent to the rest of the developers. This would help
-catch issues in the contributions, and improve source code maintenance.
-
-According to the interviewed developer of ESPResSo, some obstacles to the
-development processes of their package had been overcome by the introduction of
-continuous integration practices, and a peer review process for contributions.
-These practices decrease development and maintenance times. The developer of
-TCLB mentioned that their package had two sets of code, for executing the models
-on the CPU and GPU, and that maintenance was decreased by introducing macros, a
-practice which became a common part of the development process.
-
-Developers were also asked how documentation fits into their development
-process. Several developers noted that developer documentation plays an
-important role in familiarizing potential contributors to the software system
-architecture. Without the guidance that the documentation provides it would be
-unlikely that contributions would pass the peer review process.
-
-None of the software packages whose developers were interviewed have a formal
-software development model. The packages all have fairly small development
-teams. These teams do accept outside contributors, but generally the teams are
-tight-knit, often working at the same institution, although one of the packages
-has an international team. The developer of ESPResSo noted that while no formal
-model is used, their development model is something similar to a combination of
-agile and waterfall development models.
-
-The developers noted similar project management processes. For teams of only a
-couple of developers, the addition of new features or major changes are
-discussed with the entire team. Projects with more than a couple of developers
-have lead developer roles. These lead developers review potential additions to
-the software. The software packages use GitHub for managing the project.
-Typically there are several development branches as well as the master branch.
+frequency of such updates and benefit both users and developers.  \wss{Same
+thing has come up in papers.  I'll try to find a reference for this.  This is
+also something we can point to for recommendations related to code generation.}
+\wss{Peter - this also sounds like understandability related discussion?  I
+wonder if we should drop Traceability as a quality that we assessed?  There
+doesn't seem to be much content specific to traceability.}  The developer of
+OpenLB noted the use of documentation generators like Doxygen. It would be
+advisable for more projects to use such automatic document generation tools,
+since some projects do not do this.  \wss{We can refer back to this for the
+recommendations.}
+
+\subsection{Visibility and Transparency} \label{secVisib}
+
+For the developers interviewed, visibility is not currently a priority.
+Visibility is not particularly necessary because the teams are fairly small.
+Although they accept outside contributors, these are rare and the teams remain
+tight-knit.  Most of the teams interviewed have all members at the same
+institution, although one team is spread out internationally. \wss{Peter, can
+you say which team is international, so that we can add this detail to the
+paper?}  For teams of only a few developers, new features or major changes are
+discussed with the entire team. Projects with more than a few developers have
+lead developer roles. These lead developers review potential additions to the
+software.
+
+Although visibility is not a priority, positive steps are being undertaken. For
+instance, all interviewed teams used GitHub, with its visibility enhancing
+features of branching, issue tracking, insights on contributors, the blame
+feature, wikis, pull requests, etc. Continuous integration, as used by ESPResSo,
+makes projects even more visible, since contributors can explicitly see the bar
+that must be crossed for code to be incorporated into the master branch.
+ESPResSo also uses peer review, which improves visibility by explicitly asking
+for feedback from other developers.
+
+Previously mentioned concerns with maintainability, modifiability, and
+understandability  could be improved by further improving visibility.  For
+future sustainability, projects will likely have to expand their development
+teams by recruiting more external developers. One component of recruiting
+developers is transparency in the development process, including the process for
+making a contribution.  \wss{in recommendations could point out what teams could
+do to get more external contributors, and potentially a longer life for their
+project.}
 
@@
 
-Software packages were not quantitatively measured for reproducibility. In this
-project we asked developers to comment on reproducibility when we interviewed
-them.
-
-Developers were asked if they have any concern that their computational results
+Software packages were not quantitatively measured for reproducibility. Instead,
+developers were asked if they had concerns that their computational results
 would not be reproducible in the future, and if they had taken any steps to
-ensure reproducibility.
-
-The developer of ESPResSo noted a comparison of the results of their methods
-against manually calculated results. These comparison tests are automatically
-run for all source code changes. The tests are run when a pull request is opened
-on GitHub. Even once these tests are complete, a peer review process is done
-before changes are fully committed to the appropriate branch. The results for
-all of the LBM schemes on the software package development branch are also
-frequently compared for correctness, ensuring that the system output reflects
-the expected output.
-
-Several developers noted that they currently do not have a system in place to
-test for reproducibility, but it is of interest and could be implemented in the
-future. Generally, the mathematical foundations of the models are verified, but
-the output of the software package is not compared to other output. Depending on
-the package and how it outputs solutions, it may not be practical or feasible. A
-correct output may not be exactly reproducible, as it may be dependent on a
-probability distribution, so strictly comparing results may not be appropriate.
-
-The source code and artifacts of some software packages may be reproducible.
-There is considerable variance in the quality of software specifications and
-other developer documentation. Some packages are well detailed, and translating
-the specifications into source code will produce similar results across
-developers. Developers were not asked to comment on the reproducibility of their
-source code from their requirements specifications and design documentation.
-This question should be considered in the next iteration of state of the
-practice assessments.
+ensure reproducibility.  The developers all agreed that reproducibility is
+important, but they differ in how far they go to support reproducibility.
+ESPResSo does the most to support reproducibility.  For instance, they compare
+the results of their methods against manually calculated results. These
+comparison tests are automatically run for all source code changes. The tests
+are run when a pull request is opened on GitHub. Even once these tests are
+complete, a peer review process is undertaken before changes are fully committed
+to the appropriate branch. The results for all of the LBM schemes on the
+software package development branch are also frequently compared for
+correctness, ensuring that the system output reflects the expected output.
+
+For other teams, the priority is verifying the mathematical foundations of the
+models, but the output of the software package is not compared to other output.
+Depending on the package and how it outputs solutions, this may not be practical
+or feasible. A correct output may not be exactly reproducible, as it may be
+dependent on a probability distribution, so strictly comparing results may not
+be appropriate. \wss{Ideas for dealing with reproducibility in these situations
+will be explored in the recommendations section (like properties of the solution,
+such as conservation of mass).}
+
+For the current project, we did not assess the stronger form of
+``reproducibility'' known as replicability.  A software project is considered
+replicable if the documentation alone (without the code) is sufficiently precise
+and complete to allow a third-part to re-obtain the results of the original code
+\citep{BenureauAndRougier2017}.  Replicability is difficult to achieve with
+current documentation practices.  This question should be considered in the next
+iteration of state of the practice assessments.
 
@@
 
-Software packages were not quantitatively measured for unambiguity. In this
-project we asked developers to comment on unambiguity when we interviewed them.
-
-Developers were asked if they thought that the current documentation can clearly
-convey all necessary knowledge to the users, and if they had taken any steps to
-ensure clarity.
-
-The developer of ESPResSo noted that their documentation was meant for users
-that are already familiar with the underlying physics and CFD methods. These
-concepts are not explained in detail within the documentation. Users should
-acquire this knowledge from suitable external sources. The documentation focuses
-on how to technically use the software package, and includes a user guide and
-tutorial walk through of how to set up and run a simulation. With this in mind,
-the developer believes that their documentation is in reasonable shape for users
-with a minimum knowledge of the underlying physics. If new users have technical
-questions these can be addressed in further revisions of the documentation. New
-developers should find that the documentation is reasonably clear and useful.
-Information that is missing, like detailed explanations of dependencies, is
-referenced in the documents.
+Software packages were not quantitatively measured for unambiguity. We asked
+developers if they thought that the current documentation can clearly convey all
+necessary knowledge to the users, and if they had taken any steps to ensure
+clarity. The developer of ESPResSo noted that their documentation was meant for
+users that are already familiar with the underlying physics and CFD methods.
+These concepts are not explained in detail within the documentation. Users
+should acquire this knowledge from suitable external sources. The documentation
+focuses on how to technically use the software package, and includes a user
+guide and tutorial walk through of how to set up and run a simulation. With this
+in mind, the developer believes that their documentation is in reasonable shape
+for users with a minimum knowledge of the underlying physics. If new users have
+technical questions these can be addressed in further revisions of the
+documentation. New developers should find that the documentation is reasonably
+clear and useful. Information that is missing, like detailed explanations of
+dependencies, is referenced in the documents.
 
@@
 
-This section answers the research questions listed in Section~\ref{purpose}
+This section answers the research questions listed in Section~\ref{secIntro}
 using the quantitative data presented in Section~\ref{AHPresults}, qualitative
@@
 understandability. A user manual also helps with those qualities and with
-installability, and correctness and verificability. Performance notes suggest
+installability, and correctness and verifiability. Performance notes suggest
 that performance was considered when developing the software. Test plans,
@@
 All but three (laboetie, lettuce, Sailfish) of the software packages that use Git have most of their issues closed.
-	\item Use a version control system (GitHib, CVS). Four (ESPREsSo, Ludwig, LUMA, Palabos) of the top five ranked packages use a version control system.
+	\item Use a version control system (GitHub, CVS). Four (ESPREsSo, Ludwig, LUMA, Palabos) of the top five ranked packages use a version control system.
 	\item Source code needs to be well commented.
@@
 \begin{itemize}
-	\item Documentation must state all technical requirements and software dependencies. A missing dependency was a frequent cause of fault conditions during this assessment.
-	\item Provide a detailed user manual. Thirteen packages have a user manual. ESPResSo has a well detailed manual.
-	\item Provide a table of symbols in the developer's guide that maps to names used in the source code. This is suggested as good software development practice.
-	\item State appropriate fluid dynamics problems that the software is designed to model in the documentation, and explicitly state the limits of the software. This is done in the ESPResSo user guide.
-	\item The documentation should either explain the underlying CFD theories or provide a reference to appropriate resources.
-This was done by several top ranked packages, including ESPResSo, Ludwig, LUMA, and OpenLB.
-	\item Consider asking users for feedback on the documentation. The developer of pyLBM noted that such feedback would be appreciated.
+	\item Documentation must state all technical requirements and software
+	dependencies. A missing dependency was a frequent cause of fault conditions
+	during this assessment.
+	\item Provide a detailed user manual. Thirteen packages have a user manual.
+	ESPResSo has a well detailed manual.
+	\item Provide a table of symbols in the developer's guide that maps to names
+	used in the source code. This is suggested as good software development
+	practice.
+	\item State appropriate fluid dynamics problems that the software is
+	designed to model in the documentation, and explicitly state the limits of
+	the software. This is done in the ESPResSo user guide.
+	\item The documentation should either explain the underlying CFD theories or
+	provide a reference to appropriate resources. This was done by several top
+	ranked packages, including ESPResSo, Ludwig, LUMA, and OpenLB.
+	\item Consider asking users for feedback on the documentation. The developer
+	of pyLBM noted that such feedback would be appreciated.
 \end{itemize}
@@
 assessed to answer the software development related research questions listed in
-Section~\ref{purpose} to understand how software quality is impacted by software
+Section~\ref{secIntro} to understand how software quality is impacted by software
 development choices, including principles, processes, and tools. Software
@@
 
-- value of tools to enforce good behaviour of developers, like R, like MOOSE infrastructure
+- value of tools to enforce good behavior of developers, like R, like MOOSE infrastructure
+
+- change RQs - make them more about understanding the range of development
+practices, and identifying projects that are successful.  Also identify room for
+improvement, as identified by the developers themselves, and as identified by
+us.  Compare the state of the practice of LBM to the state of practice of other
+domains.  Where does LBM stand relative to other domains.
+
+- read \citet{WieseEtAl2019} (Naming the Pain in Developing Scientific Software)

### 2021-12-29 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
updated LBM paper to include Musubi measurements, up to the end of Section 3

@@
     by star count (Sailfish) is ranked 9th by our methodology. Our top three
-    packages were Ludwig, ESPResSo, and Palabos (all in the top 10 by star
+    packages were ESPResSo, Ludwig, and Palabos (all in the top 10 by star
     count). We interviewed 4 developers to gain insight into their current pain
@@
 		ESPResSo++ & 2011 Feb & 2020 Apr\\
-		lbmpy& unknown  & 2020 Jun  \\
+		lbmpy& unclear  & 2020 Jun  \\
 		lettuce & 2019 May & 2020 Jul\\
@@
 		LUMA& 2016 Nov   & 2020 Feb\\
-		MechSys & 2008 Jun    & 2020 Jul\\
+		MechSys & 2008 Jun    & 2021 Oct\\
+		Musubi & 2013 Sep & 2019 Aug\\
 		OpenLB & 2007 Jul & 2019 Oct\\
@@
 			MP-LABS & 2, 3 & MPI/OMP & N & Y & Y & N & N & L\\
+			Musubi & 2, 3 & MPI & Y & Y & Y & Y & Y & W, L\\
 			OpenLB & 1, 2, 3 & MPI & Y & Y & Y & Y & Y & W, M, L\\
@@
 
-All of the 23 software packages that were tested have installation instructions.
-As noted previously, many of the 23 software packages that were part of the
-original long list of 45 packages were removed due to not including
-documentation or installation instructions. Of the 23 software packages that
+All of the 24 software packages that were tested have installation instructions.
+As noted previously, many of the 22 software packages that were part of the
+original long list of 46 packages were removed due to not including
+documentation or installation instructions. Of the 24 software packages that
 were tested, most had installation instructions located in one place, often in
@@
 
-All 23 packages on the short list can be installed on some Unix-like systems.
-Seven packages could be installed on Windows, and five on macOS. Operating
-system compatibility is found in the documentation of 19 software packages. All
-but one of the software packages (TCLB) were tested on Ubuntu for this state of
-the practice assessment. TCLB was tested on CentOS, since this operating system
-is mentioned in its installation instructions.
+All 24 packages on the short list can be installed on some Unix-like systems.
+Eight packages could be installed on Windows, and five on macOS.
+Operating system requirements are listed in the documentation of 20 software packages, but compatible operating system versions are only specified in four packages (ESPResSo, LUMA, Mechsys, Sailfish).
+All but one of the software packages (TCLB) were tested on Ubuntu for this state of the practice assessment.
+TCLB was tested on CentOS, since this operating system is mentioned in its installation instructions.
 
@@
 
-Sixteen software packages require less than 10 dependencies to be installed. All
-but one software package (LatBo.jl) require less than 20 dependencies. Some
+Seventeen software packages require less than 10 dependencies to be installed. All but one software package (LatBo.jl) require less than 20 dependencies. Some
 packages may automatically install additional dependencies in the background.
-Eighteen of the software packages do not explicitly indicate software dependency
+Twenty of the software packages do not explicitly indicate software dependency
 versions. Some software package installation issues, specifically those
 occurring when manual installation of dependencies is required, may be avoided
-if versions of dependencies are specified. Fifteen software packages do not have
+if versions of dependencies are specified. Sixteen software packages do not have
 detailed instructions for installing dependencies. Sixteen software packages
@@
 Figure~\ref{Fig_Installability} shows the installability ranking of the software
-packages using AHP. Software packages with a higher score (MechSys, Palabos,
-ESPResSo, OpenLB) tend to have one set of linear installation instructions that
-are written as if the person doing the installation has none of the dependencies
-installed. The instructions often list compatible operating system versions and
-include instructions for the installation of dependencies. The top ranked
-packages often incorporate some sort of automation of the installation process
-and have fewer manual installation steps. The number of dependencies a package
-has does not correlate with a higher score. The ability to validate the
-installation process, often through tutorials or test examples that include
-expected output, is correlated with a higher score. Furthermore, the top six
-ranked packages are noted as being alive.
+packages using AHP.
+Software packages with a higher score (ESPResSo, MechSys, OpenLB, DL\_MESO, Palabos) tend to have one set of linear installation instructions that are written as if the person doing the installation has none of the dependencies installed.
+The instructions often list compatible operating system versions and include instructions for the installation of dependencies.
+The top ranked packages often incorporate some sort of automation of the installation process and have fewer manual installation steps.
+The number of dependencies a package has does not correlate with a higher score.
+The ability to validate the installation process, often through tutorials or test examples that include expected output, is correlated with a higher score.
+Furthermore, the top seven ranked packages are noted as being alive.
 
@@
 
-Sixteen of the software packages include a requirements specification artifact
+Seventeen of the software packages include a requirements specification artifact
 or explicitly reference domain theory, often only the latter. Software packages
@@
 
-Document generation tools are explicitly used by 12 software packages. Sphinx is
-used by eight of them, and Doxygen is used by seven. Several of the packages use
+Document generation tools are explicitly used by 13 software packages. Sphinx is
+used by eight of them, and Doxygen is used by eight. Several of the packages use
 both.
 
-Tutorials are available for 18 of the software packages. Generally they are
+Tutorials are available for 19 of the software packages. Generally they are
 linearly written and easy to follow. However, only eight tutorials provide an
@@
 
-Unit tests are only explicitly available for one of the software packages,
-Ludwig. Code modularization of most packages allow for users to create tests
-with varying degrees of effort. These tests allow developers and users to verify
-the correctness of fragments of the source code, and in doing so better assess
-the correctness of the entire package.
+Unit tests are only explicitly available for two of the software packages,
+Ludwig and Musubi.
+Code modularization of most packages allow for users to create tests with varying degrees of effort.
+These tests allow developers and users to verify the correctness of fragments of the source code, and in doing so better assess the correctness of the entire package.
 
 The use of continuous integration tools and techniques alludes to a more refined
-development process where faults are isolated and better recognized. Only two of
-the packages (ESPResSo, Ludwig) mentioned applying the practice of continuous
-integration in their development process.
+development process where faults are isolated and better recognized.
+Only three of the packages (ESPResSo, Ludwig, Musubi) mentioned applying the practice of continuous integration in their development process.
 
@@
 packages all include an easy to follow getting started tutorial, and most of
-these include expected output. Only the top ranked package, Ludwig, provided
-unit testing. It and the second ranked package, ESPResSo, explicitly
-incorporated continuous integration in the development process. Furthermore,
-eight of the top 10 ranked packages are noted as being alive.
+these include expected output.
+The two top ranked packages, Ludwig and Musubi, indicate the use of unit testing in their documentation.
+They also incorporated continuous integration in the development process.
+Furthermore, eight of the top 10 ranked packages are noted as being alive.
 
@@
 commands (even when following the installation guide), missing links, missing
-dependencies, and syntax errors in code files. In some instances the error messages
-were vague. Several automatic installation processes could not find and load 
-dependencies. In these instances the installation tried to access outdated
-external repositories. Seven of the installations were recovered and verified,
-and one of the installations (LB3D-Prime) was assumed to be recovered due to the
-absence of any way to verify it. The installation of eight of the software
-packages could not be recovered. Most of these broken installations could not
-find external dependencies, encountered system incompatibilities, or displayed
-vague error messages.
-
-Of the 13 software packages that installed correctly and also have tutorials,
+dependencies, and syntax errors in code files.
+In some instances the error messages were vague.
+Several automatic installation processes could not find and load dependencies.
+In these instances the installation tried to access outdated external repositories.
+Seven of the installations were recovered and verified, and one of the installations (LB3D-Prime) was assumed to be recovered due to the absence of any way to verify it.
+The installation of eight of the software packages could not be recovered.
+Most of these broken installations could not find external dependencies, encountered system incompatibilities, or displayed vague error messages.
+
+Of the 14 software packages that installed correctly and also have tutorials,
 four (pyLBM, ESPResSo++, LIMBES, Ludwig) broke during tutorial testing. All of
@@
 break during installation, or the broken installation was recoverable. All of
-the top five ranked packages have tutorials. One of these packages, pyLBM, broke
+the top five ranked packages have tutorials. The package pyLBM broke
 during tutorial testing, but a descriptive error message helped in recovery.
-Furthermore, nine of the top 10 ranked packages are noted as being alive.
+Nine of the top 10 ranked packages are noted as being alive.
 
@@
 appropriate. In this project we instead looked through each software package's
-artifacts for evidence that performance was considered. The artifacts of 17
+artifacts for evidence that performance was considered. The artifacts of 18
 software packages mentioned parallelization. This included GPU processing and
@@
 Software package artifacts were reviewed for the presence of a tutorial, a user
-manual, documented user characteristics, and a user support model. In total 18
+manual, documented user characteristics, and a user support model. In total 19
 software packages have a tutorial, 13 have a user manual, and 11 have both. The
@@
 but do have useful documentation distributed on their web-pages. Expected user
-characteristics are documented in four software packages (laboetie, LIMBES,
-Ludwig, Palabos). Users are typically scientists or engineers. Their background
-is often physics, chemistry, biophysics, or mathematics. All but one of the
-packages (LIMBES) have a user support model, and many of them have multiple
-avenues of user support. The most popular avenue of support is Git, followed by
-email and forums. One software package (OpenLB) has an FAQ page.
+characteristics are documented in five software packages (laboetie, LIMBES,
+Ludwig, Musubi, Palabos).
+Users are typically scientists or engineers.
+Their background is often physics, chemistry, biophysics, or mathematics.
+All but one of the packages (LIMBES) have a user support model, and many of them have multiple avenues of user support.
+The most popular avenue of support is Git, followed by email and forums.
+One software package (OpenLB) has an FAQ page.
 
@@
 
-Issue tracking is used in 22 software packages, 15 of which use Git, six use
+Issue tracking is used in 23 software packages, 15 of which use Git, seven use
 email, and one (SunlightLB) uses SourceForge. Most software packages that use
 Git have most of their issues closed, and only three (laboetie, lettuce,
-Sailfish) have less than 50 percent of their issues closed. Four of the top five
-overall ranked packages (Ludwig, ESPResSo, Palabos, LUMA) have most of their
-issues closed. Fourth ranked LUMA does not use Git. Alive packages (11 use Git
-issue tracking) have 64\% of their issues closed, while dead packages (3 use Git
-issue tracking) have 71\% of their issues closed. This information is presented
-in Table~\ref{gitrepodata}. Furthermore, 13 packages that use Git for issue
-tracking use GitHub as a version control system, while two (Palabos, waLBerla)
-use GitLab. Of the other packages, one package (SunlightLB) uses CVS for issue
-tracking, and seven packages do not appear to use any issue tracking system.
+Sailfish) have less than 50 percent of their issues closed. All of the top five
+overall ranked packages have most of their issues closed.
+Alive packages (11 use Git issue tracking) have 64\% of their issues closed, while dead packages (3 use Git issue tracking) have 71\% of their issues closed.
+This information is presented in Table~\ref{gitrepodata}.
+Furthermore, 13 packages that use Git for issue tracking use GitHub as a version control system, while two (Palabos, waLBerla) use GitLab.
+Of the other packages, one package (SunlightLB) uses CVS for issue tracking, and seven packages do not appear to use any issue tracking system.
 
@@
 			MP-LABS & 100.00 & 26.67 & Dead\\
+			Musubi & Not Git & 24.19 & Alive\\
 			OpenLB & Not Git & 22.43 & Alive\\
@@
 Packages with a higher percentage of comments were designated as more
-maintainable. Comments represent more than 10 percent of code files in 15
-packages, and the average percentage of code comments is about 14 percent. Four
-of the top five overall ranked packages (Ludwig, ESPResSo, Palabos, OpenLB) have
-more than the average. Fifth ranked LUMA has only 0.2 percent comments, the
-fewest of any package. This package has the most lines of source code, with over
-four million. The next largest package is ESPResSo++ with one million.
+maintainable. Comments represent more than 10 percent of code files in 16
+packages, and the average percentage of code comments is about 14 percent. All
+of the top five overall ranked packages have more than the average.
+LUMA has only 0.2 percent comments, the fewest of any package.
+This package has the most lines of source code, with over four million.
+The next largest package is ESPResSo++ with one million.
 
@@
 issues closed. Their code files are well commented with more than 10 percent of
-the code being comments. Furthermore, nine of the top 10 ranked packages are
+the code being comments. Furthermore, four of the top 5 ranked packages are
 noted as being alive. MP-LABS is noted as being dead.\\
@@
 			DL\_MESO (LBE) & 310 & 51 & 170223 & 549\\
-			ESPResSo & 1309& 86 & 186700 & 143\\
-			ESPResSo++ & 5328& 66 & 969196 & 182\\
-			HemeLB & 1065& 48 & 95104 & 89\\
+			ESPResSo & 1390& 83 & 195083 & 140\\
+			ESPResSo++ & 1406& 30 & 165194 & 118\\
+			HemeLB & 1102& 44 & 123806 & 112\\
 			laboetie & 133& 1 & 48403 & 364\\		
@@
 			LB3D & 99 & 76 & 39766 & 402\\
-			LB3D-Prime & 2 3& 6 & 12944 & 563\\
-			lbmpy& 201 & 28 & 46489 & 231\\
-			lettuce & 62 & 0 & 5529 & 89\\
+			LB3D-Prime & 23& 6 & 12944 & 563\\
+			lbmpy& 250 & 29 & 61632 & 247\\
+			lettuce & 73 & 1 & 7660 & 105\\
 			LIMBES & 26 & 1 & 4872 & 187\\
-			Ludwig & 859 & 32 & 109811 & 128\\
-			LUMA & 312 & 19 & 4370670 & 14000\\
-			MechSys & 324 & 3 & 85543 & 264\\
+			Ludwig & 954 & 32 & 162518 & 170\\
+			LUMA & 314 & 19 & 4399723 & 14011\\
+			MechSys & 333 & 3 & 95707 & 287\\
 			MP-LABS & 307 & 3 & 43124 & 140\\
-			OpenLB & 1104 & 5 & 209034 & 189\\
-			Palabos & 1829 & 67 & 547623 & 299\\
-			pyLBM & 258 & 85 & 32314 & 125\\
+			Musubi & 1347 & 1839 & 281879 & 209\\
+			OpenLB & 1438 & 7 & 218406 & 152\\
+			Palabos & 1974 & 71 & 563841 & 286\\
+			pyLBM & 272 & 108 & 37234 & 137\\
 			Sailfish & 632 & 11 & 69398 & 110\\
 			SunlightLB & 36 & 1 & 7646 & 212\\
-			TCLB & 535 & 7 & 43226 & 81\\
-			waLBerla & 2395 & 67 & 848146 & 353\\
+			TCLB & 594 & 7 & 49156 & 83\\
+			waLBerla & 2643 & 69 & 873988 & 331\\
 			\hline
@@
 
-All of the packages appear to have consistent indentation and formatting. Only
-LUMA and HemeLB explicitly identify coding standards that are used during
-development. Generally, the software packages use consistent, distinctive, and
-meaningful code identifiers. Only four packages (LB2D-Prime, LB3D-Prime, LIMBES,
-MP-LABS) appear to use vague identifiers, such as single letters for variables.
-Symbolic constants were observed in the source code of 12 packages. The
-constants are used for various parameters, mathematical constants, and matrix
-definitions. All of the packages are reasonably well commented, with the
-comments clearly indicating what is being done (as opposed to how it is being
-done). Domain algorithms are noted in the source code of 11 packages.
-Table~\ref{moduledata} suggests that the software packages are modularized to
-various degrees. When observing the source code files, it was found that 13 of
-the packages have a consistent style and order of function parameters.
+All of the packages appear to have consistent indentation and formatting.
+Only HemeLB, LUMA, and Musubi explicitly identify coding standards that are used during development.
+Generally, the software packages use consistent, distinctive, and meaningful code identifiers.
+Only four packages (LB2D-Prime, LB3D-Prime, LIMBES, MP-LABS) appear to use vague identifiers, such as single letters for variables.
+Symbolic constants were observed in the source code of 13 packages.
+The constants are used for various parameters, mathematical constants, and matrix definitions.
+All of the packages are reasonably well commented, with the comments clearly indicating what is being done (as opposed to how it is being done).
+Domain algorithms are noted in the source code of 11 packages.
+Table~\ref{moduledata} suggests that the software packages are modularized to various degrees.
+When observing the source code files, it was found that 14 of the packages have a consistent style and order of function parameters.
 
@@
 what is being done in the source code. The source code is well modularized and
-structured. Furthermore, four of the top five ranked packages are noted as being
+structured. Furthermore, all of the top five ranked packages are noted as being
 alive.
@@
 
-Looking at the top three ranked packages: Ludwig scored high in surface
-correctness and verifiability, surface robustness, surface usability,
-maintainability, and visibility and transparency. ESPResSo had achieved a
-relative high score in installability, surface correctness and verifiability,
-surface usability, maintainability, reusability, and visibility and
-transparency. Palabos scored high in installability, surface reliability,
-surface robustness, maintainability, and understandability.
+Looking at the top three ranked packages: ESPResSo had achieved a
+relative high score in installability, surface usability, maintainability, reusability, and visibility and transparency.
+Ludwig scored high in surface correctness and verifiability, surface robustness, surface usability, maintainability, and visibility and transparency.
+Palabos scored high in installability, surface reliability, surface robustness, maintainability, understandability, and visibility and transparency.
 

### 2021-12-29 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
updated LBM paper to include Musubi measurements, up to but not including Section 5.6

@@
 and rare artifacts in Table~\ref{artifactspresent}. Common artifacts were found
-in 16 to 23 ($>$70\%) of the software packages. Less common artifacts were found
-in 8 to 15 (35-70\%) of the software packages. Rare artifacts were found in 1 to
-7 ($<$35\%) of the software packages.
+in 16 to 24 ($>$63\%) of the software packages. Less common artifacts were found
+in 8 to 15 (30-63\%) of the software packages. Rare artifacts were found in 1 to
+7 ($<$30\%) of the software packages.
 
@@
 
-All of the top four AHP ranked packages, ESPREsSo, Ludwig, OpenLB, and Palabos,
-have each of the commonly found artifacts, except only three of them have a
+All of the top four AHP ranked packages, ESPREsSo, Ludwig, Palabos, and OpenLB, have each of the commonly found artifacts, except only three of them have a
 requirements specification or theory notes. Palabos is the only one of these
@@
 data collection. Despite the broken link, Palabos does have a detailed and
-informative website. Another one (Ludwig) of the top four packages does not
-appear to have publicly visible design documentation. A third package (OpenLB)
-from the list does not appear to use a version control system. It is possible
+informative website. Another one of the top four packages, Ludwig, does not
+appear to have publicly visible design documentation. A third package from the list, OpenLB, does not appear to use a version control system. It is possible
 that such a system is used by the package since its website notes package
@@
 released \citep{ghezzi1991fundamentals}. The following type of development tools
-were explicitly noted in the artifacts or web-pages of the 23 LBM packages that
+were explicitly noted in the artifacts or web-pages of the 24 LBM packages that
 were assessed. It is likely that other tools, such as debuggers, were used but
@@
 The above tools can verify the correctness of software during its development.
-Only two (ESPResSo, Ludwig) of the software packages that were assessed
-mentioned using continuous integration tools, like Travis CI. Code editors and
-compilers were explicitly noted to have been used by several packages, and were
-likely used by all of them. One of the packages (Ludwig) explicitly noted the
-use of proprietary unit testing code written in C. Likewise, the use of
-proprietary code for verifying the correctness of output was noted by one
-(pyLBM) of the developers. It is likely that similar tools were used when
-developing other software packages.
+Only three (ESPResSo, Ludwig, Musubi) of the software packages that were assessed mentioned using continuous integration tools, like Travis CI.
+Code editors and compilers were explicitly noted to have been used by several packages, and were likely used by all of them.
+One of the packages (Ludwig) explicitly noted the use of proprietary unit testing code written in C.
+Likewise, the use of proprietary code for verifying the correctness of output was noted by one (pyLBM) of the developers.
+It is likely that similar tools were used when developing other software packages.
 
@@
 The following types of dependencies were explicitly noted in the artifacts or
-web-pages of the 23 LBM packages that were assessed. It is possible that other
+web-pages of the 24 LBM packages that were assessed. It is possible that other
 types of dependencies are part of these software packages, but are not clearly
@@
 types of project management tools were explicitly noted in the artifacts,
-web-pages, or interviews with the developers of the 23 LBM packages that were
+web-pages, or interviews with the developers of the 24 LBM packages that were
 assessed. As with development tools and dependencies, it is possible that other
@@
 (SunlightLB) uses CVS. Document generation tools are mentioned in the artifacts
-of 12 of the projects. The tools Sphinx and Doxygen are explicitly used in this
+of 13 of the projects. The tools Sphinx and Doxygen are explicitly used in this
 capacity.
@@
 
-Seven of the software packages contain artifacts outlining the general
-development process, like basic instructions on how to contribute. Eleven of the
-packages explicitly convey that they would accept outside contributors, but
-generally the teams are centralized, often working at the same institution.
+Seven of the software packages contain artifacts recording the general
+development process and status. Eleven of the packages explicitly convey that they would accept outside contributors, but generally the teams are centralized, often working at the same institution.
 
@@
 
-Thirteen (57\%) of the 23 software packages use GitHub for managing the project,
-including nine (64\%) of the 14 alive packages, and four of the nine (44\%) dead
+Thirteen (54\%) of the 24 software packages use GitHub for managing the project,
+including nine (60\%) of the 15 alive packages, and four of the nine (44\%) dead
 packages. Two projects (Palabos, WaLBerla) use GitLab. This could be indicative
@@
 simulation engine and interface. The use of document generation tools is
-mentioned in the artifacts of 12 software packages, and was noted during
+mentioned in the artifacts of 13 software packages, and was noted during
 interviews with developers. Sphinx and Doxygen are the tools that were
@@
 	\item The system must provide descriptive error messages when it encounters unexpected input, including incorrect data types, empty input, and missing files or links.
-Eighteen of the software packages behaved reasonably when tested with unexpected input.
+Nineteen of the software packages behaved reasonably when tested with unexpected input.
 \end{itemize}
@@
 \begin{itemize}
-	\item Integrate parallelization tools and techniques to reduce processing time. Consider GPU processing, CUDA, and MPI. Seventeen software packages mentioned parallelization in their artifacts.
+	\item Integrate parallelization tools and techniques to reduce processing time. Consider GPU processing, CUDA, and MPI. Eighteen software packages mentioned parallelization in their artifacts.
 	\item The user should be able to choose to process their model on either the CPU or GPU. ESPResSo, lbmpy, lettuce, pyLBM, and TCLB allow for either CPU or GPU processing.
@@
 	\item Adopt a coding standard and document it in artifacts, including some examples. A coding standard needs to be part of continuous integration. Only
-	LUMA and HemeLB explicitly identify coding standards. 
-	\item Use consistent, distinctive, and meaningful code identifiers. Nineteen of the 23 software packages use such code identifiers.
-	\item Use symbolic constants. Twelve of the 23 software packages use such constants.
+	LUMA, HemeLB, and Musubi explicitly identify coding standards. 
+	\item Use consistent, distinctive, and meaningful code identifiers. Twenty of the 24 software packages use such code identifiers.
+	\item Use symbolic constants. Thirteen of the 24 software packages use such constants.
 	\item Identify algorithms that are used in source code comments.
@@
 Eleven packages have information on how to contribute.
-	\item Identify the development environment. Eight of the 23 software packages identify the development environment.
+	\item Identify the development environment. Eight of the 24 software packages identify the development environment.
 	\item Update all relevant documentation when a change to the software is made. This is suggested as good software development practice.

### 2021-12-29 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
updated LBM paper to include Musubi measurements, up to end of paper

@@
 
-Eight packages do not use GitHub or GitLab, so they do not have a measure of
+Nine packages do not use GitHub, so they do not have a measure of
 repository stars. Looking at the repository stars of the other 15 packages, we
 can observe a slight pattern where packages that have been highly ranked by our
-assessment do have more stars than lower ranked packages. The second ranked
+assessment do have more stars than lower ranked packages. The best ranked
 package (ESPResSo) has the second most number of stars. The ninth ranked package
 (Sailfish) has the most number of stars. Packages designated as lower quality
-often do not use GitHub or GitLab, or have few stars, except for HemeLB which
-has 22 stars. Our assessment of this package might not be accurate. The number
-of stars does not necessarily represent the perceptions of the community, but
-for lack of an alternative measure we will use it this way.
+often do not use GitHub or GitLab, or have few stars.
+HemeLB only has 12 stars, but its location recently moved from a different GitHub repository.
+Our assessment of this package might not be accurate.
+The number of stars does not necessarily represent the perceptions of the community, but for lack of an alternative measure we will use it this way.
 
@@
 stars metric. The ninth ranked package (Sailfish) has the most number of
-watches. The second ranked package (ESPResSo) has the second most number of
+watches. The best ranked package (ESPResSo) has the second most number of
 watches.
@@
 			\hline
-			Ludwig&1& 19 &8& 6\\
-			ESPResSo &2& 115 &2& 23\\
-			Palabos &3& 14 &10& GitLab\\
+			ESPResSo &1& 145 &2& 19\\
+			Ludwig&2& 27 &8& 6\\
+			Palabos &3& 34 &6& GitLab\\
 			OpenLB &4& No Git &No Git & No Git\\
-			LUMA&5& 26 &6& 10\\
-			pyLBM &6& 57&4& 8\\
+			LUMA&5& 33 &7& 12\\
+			pyLBM &6& 95&3& 10\\
 			DL\_MESO (LBE) & 7 & No Git &No Git & No Git \\
-			waLBerla & 8 & 17 &9& GitLab\\
-			Sailfish &9& 154 &1& 39\\
-			laboetie &10& 4 &13& 5\\
-			TCLB &11& 61 &3& 16\\
-			MechSys &12& No Git &No Git & No Git\\
-			lettuce &13& 8 &11& 2\\
-			ESPResSo++ &14& 31 &5& 13\\
-			MP-LABS &15& 7 &12& 2\\			
-			SunlightLB & 16& No Git &No Git & No Git\\
-			LB3D &17& No Git &No Git & No Git\\			
-			LIMBES &18& No Git &No Git & No Git\\
-			LB2D-Prime &19& No Git &No Git & No Git\\		
-			HemeLB &20& 22 &7& 15\\
-			lbmpy&21&  4 &13& 2  \\	
-			LB3D-Prime &22& No Git &No Git & No Git\\	
-			LatBo.jl &23& 4 &13& 5\\			
+			Musubi & 8 & No Git &No Git & No Git \\
+			Sailfish &9& 186 &1& 41\\
+			waLBerla & 10 & 20 &9& GitLab\\
+			laboetie &11& 4 &13& 5\\
+			TCLB &12& 95 &3& 16\\
+			MechSys &13& No Git &No Git & No Git\\
+			lettuce &14& 48 &4& 5\\
+			ESPResSo++ &15& 35 &5& 12\\
+			MP-LABS &16& 12 &11& 2\\			
+			SunlightLB & 17& No Git &No Git & No Git\\
+			LB3D &18& No Git &No Git & No Git\\			
+			LIMBES &19& No Git &No Git & No Git\\
+			LB2D-Prime &20& No Git &No Git & No Git\\		
+			HemeLB &21& 12 &11& 12\\
+			lbmpy&22&  11 &12&  2\\	
+			LB3D-Prime &23& No Git &No Git & No Git\\	
+			LatBo.jl &24& 17 &10& 8\\			
 			\hline
@@
 Table~\ref{rankingcomparison} compares the top 10 ranked LBM software packages
-with a LBM package ranking made by a domain expert. Five (Palabos, OpenLB, LUMA,
-waLBerla, Sailfish) of the top 10 ranked packages in this assessment are also
-found in the domain expert's list. Interestingly, these five packages are listed
-in the same order on both lists. Looking at the remaining packages, the top two
-ranked packages in this state of the practice assessment (Ludwig, ESPResSo) are
-not on the domain expert's list. Perhaps the intended applications of these
-packages, complex fluids for Ludwig and soft matter research for ESPResSo, do
-not align with the research interests of the domain expert, and this is why they
-did not make it onto their list. Moving down the list, pyLBM and DL\_MESO(LBE)
-did not make the domain expert's top 10 list but were mentioned by the domain
-expert. The 10th ranked package, laboetie, was not mentioned by the domain
-expert. Looking at the remainder of the domain expert's list, ASL is a general
-purpose tool for solving partial differential equations and may have fallen out
-of scope of the authoritative lists that were used to identify the initial LBM
-software list. The domain expert's fifth ranked package (ch4-project) was on the
-initial software list of this assessment, but was removed due to a lack of
-documentation. Open FSI is a new project that uses Palabos. It was made public
-the year before data was collected for this state of the practice exercise. It
-may not have been on authoritative lists due to its young age. Similarly, LIFE
-was also made public recently. Finally, the domain expert's 10th ranked package
-(HemeLB) was ranked 20th in this state of the practice assessment.
+with a LBM package ranking made by a domain expert.
+Five (Palabos, OpenLB, LUMA, Sailfish, waLBerla) of the top 10 ranked packages in this assessment are also found in the domain expert's top 10 ranking.
+Interestingly, the first three of these five packages are listed in the same order on both lists.
+Looking at the remaining packages, the top two ranked packages in this state of the practice assessment (ESPResSo, Ludwig) are not on the domain expert's list.
+Perhaps the intended applications of these packages, complex fluids for Ludwig and soft matter research for ESPResSo, do not align with the research interests of the domain expert, and this is why they did not make it onto their list.
+This reasoning may also apply to eighth ranked Musubi.
+Moving down the list, pyLBM and DL\_MESO(LBE) did not make the domain expert's top 10 list but were mentioned by the domain expert.
+Looking at the remainder of the domain expert's list, ASL is a general purpose tool for solving partial differential equations and may have fallen out of scope of the authoritative lists that were used to identify the initial LBM software list.
+The domain expert's fifth ranked package (ch4-project) was on the initial software list of this assessment, but was removed due to a lack of documentation.
+Open FSI is a new project that uses Palabos.
+It was made public the year before data was collected for this state of the practice exercise.
+It may not have been on authoritative lists due to its young age.
+Similarly, LIFE was also made public recently.
+Finally, the domain expert's 10th ranked package (HemeLB) was ranked 21st in this state of the practice assessment.
 
@@
 			\hline
-			1&Ludwig (N/A) &Palabos\\
-			2&ESPResSo (N/A) &OpenLB\\
+			1&ESPResSo (N/A) &Palabos\\
+			2&Ludwig (N/A) &OpenLB\\
 			3&Palabos (1)&LUMA\\
@@
 			5&LUMA (3)&ch4-project\\
-			6&pyLBM (17)&Open FSI\\
-			7&DL\_MESO (LBE) (12)& WaLBerla\\
-			8&waLBerla (7)& LIFE\\
+			6&pyLBM (16)&Open FSI\\
+			7&DL\_MESO (LBE) (11)& WaLBerla\\
+			8&Musubi (N/A)& LIFE\\
 			9&Sailfish (9)&Sailfish\\
-			10&laboetie (N/A)&HemeLB\\		
+			10&WaLBerla (7)&HemeLB\\		
 			\hline
@@
 obscure within an artifact or web-page. The use of unit testing and continuous
-integration was mentioned in the artifacts of only two (ESPResSo, Ludwig)
-packages. However, interviews suggested a more frequent use of both unit testing
-and continuous integration in the development processes than what was observed
-from the initial survey of the artifacts. For example, OpenLB, pyLBM, and TCLB
-use such methods during development despite this not being explicitly clear from
-an analysis of the material available online.
+integration was mentioned in the artifacts of only three (ESPResSo, Ludwig, Musubi) packages.
+However, interviews suggested a more frequent use of both unit testing and continuous integration in the development processes than what was observed from the initial survey of the artifacts.
+For example, OpenLB, pyLBM, and TCLB use such methods during development despite this not being explicitly clear from an analysis of the material available online.
 

### 2021-12-29 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Replied to LBM paper comments re traceability and understandability

@@
 \newcommand{\zm}[1]{\authornote{magenta}{MN}{#1}} %Zahra Motamed
-\newcommand{\pmm}[1]{\authornote{cyan}{AD}{#1}} %Peter Michalski
+\newcommand{\pmm}[1]{\authornote{cyan}{PM}{#1}} %Peter Michalski
 
@@
 confusion between understandability and usability was also a confusion during
-the interviews Ao conducted.}
+the interviews Ao conducted.} \pmm{The developer of ESPResSo did comment on developer understandability, specifically for developers that are/were part of their peer review process.
+He commented that code architecture is documented, and that code is made to be readable, and that these help peer reviews of the code.}
 
@@
 documentation seem like they would fit better under understandability, don't you
-think?}  \wss{If ESPresSo scored high (I think it did), we can use that to
+think?}  \pmm{After reviewing the definitions of the two qualities, I think you are right.
+I think that the point presented in the first sentence of this paragraph belongs here in traceability (but could be reworded to note that ESPResSo also has detailed change logs in their docs - change logs are also available for Palabos and waLBerla), but the rest of the paragraph can be moved to understandability.} \wss{If ESPresSo scored high (I think it did), we can use that to
 further promote the value of documentation.  Also a threat to validity is that
@@
 kept consistent for new developers. \wss{Peter - this also sounds like
-discussion that is relevant for understandability.}
+discussion that is relevant for understandability.} \pmm{The points here sound like they could work for both.
+The definition of traceability notes that `This quality refers to keeping track of information as it changes forms or relates between artifacts'.
+So the point `New features are always added to the documentation' could be reworded to fit that part of the definition of traceability.
+As noted above, this is similar to the point about some packages having including detailed change logs.
+Same with `The developers use documentation to stay up to date on the status of the software package, and to help expand features'.
+I agree that at least some of this paragraph could be in understandability, especially the first and second sentences.}
 
@@
 wonder if we should drop Traceability as a quality that we assessed?  There
-doesn't seem to be much content specific to traceability.}  The developer of
-OpenLB noted the use of documentation generators like Doxygen. It would be
-advisable for more projects to use such automatic document generation tools,
-since some projects do not do this.  \wss{We can refer back to this for the
-recommendations.}
+doesn't seem to be much content specific to traceability.}  \pmm{Based on the comments I just added above, I think we can keep at least a small part of what is here in traceability, rewording parts to highlight the importance of change logs.
+Some of the content in this section should indeed be moved to understandability.} The developer of OpenLB noted the use of documentation generators like Doxygen.
+It would be advisable for more projects to use such automatic document generation tools, since some projects do not do this.
+\wss{We can refer back to this for the recommendations.}
 
@@
 you say which team is international, so that we can add this detail to the
-paper?}  For teams of only a few developers, new features or major changes are
+paper?} \pmm{The interviewed developer of OpenLB noted that there were tasks for local and international developers when discussing their software development model.
+He also noted that they have frequent local meetings and annual international meetings.}  For teams of only a few developers, new features or major changes are
 discussed with the entire team. Projects with more than a few developers have
@@
 Boltzmann Methods software domain by quantitatively and qualitatively measuring,
-and comparing, 23 software packages along quality attributes. The software
+and comparing, 24 software packages along quality attributes. The software
 qualities that were assessed in this report are listed in Section

### 2022-01-03 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Edits to LBM SOP

--- 2021-12-29 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
@@
 \usepackage{multicol}
+\usepackage{booktabs}
 
@@
     by star count (Sailfish) is ranked 9th by our methodology. Our top three
-    packages were ESPResSo, Ludwig, and Palabos (all in the top 10 by star
+    packages were Ludwig, ESPResSo, and Palabos (all in the top 10 by star
     count). We interviewed 4 developers to gain insight into their current pain
@@
 		ESPResSo++ & 2011 Feb & 2020 Apr\\
-		lbmpy& unclear  & 2020 Jun  \\
+		lbmpy& unknown  & 2020 Jun  \\
 		lettuce & 2019 May & 2020 Jul\\
@@
 		LUMA& 2016 Nov   & 2020 Feb\\
-		MechSys & 2008 Jun    & 2021 Oct\\
-		Musubi & 2013 Sep & 2019 Aug\\
+		MechSys & 2008 Jun    & 2020 Jul\\
 		OpenLB & 2007 Jul & 2019 Oct\\
@@
 			MP-LABS & 2, 3 & MPI/OMP & N & Y & Y & N & N & L\\
-			Musubi & 2, 3 & MPI & Y & Y & Y & Y & Y & W, L\\
 			OpenLB & 1, 2, 3 & MPI & Y & Y & Y & Y & Y & W, M, L\\
@@
 
-All of the 24 software packages that were tested have installation instructions.
-As noted previously, many of the 22 software packages that were part of the
-original long list of 46 packages were removed due to not including
-documentation or installation instructions. Of the 24 software packages that
+All of the 23 software packages that were tested have installation instructions.
+As noted previously, many of the 23 software packages that were part of the
+original long list of 45 packages were removed due to not including
+documentation or installation instructions. Of the 23 software packages that
 were tested, most had installation instructions located in one place, often in
@@
 
-All 24 packages on the short list can be installed on some Unix-like systems.
-Eight packages could be installed on Windows, and five on macOS.
-Operating system requirements are listed in the documentation of 20 software packages, but compatible operating system versions are only specified in four packages (ESPResSo, LUMA, Mechsys, Sailfish).
-All but one of the software packages (TCLB) were tested on Ubuntu for this state of the practice assessment.
-TCLB was tested on CentOS, since this operating system is mentioned in its installation instructions.
+All 23 packages on the short list can be installed on some Unix-like systems.
+Seven packages could be installed on Windows, and five on macOS. Operating
+system compatibility is found in the documentation of 19 software packages. All
+but one of the software packages (TCLB) were tested on Ubuntu for this state of
+the practice assessment. TCLB was tested on CentOS, since this operating system
+is mentioned in its installation instructions.
 
@@
 
-Seventeen software packages require less than 10 dependencies to be installed. All but one software package (LatBo.jl) require less than 20 dependencies. Some
+Sixteen software packages require less than 10 dependencies to be installed. All
+but one software package (LatBo.jl) require less than 20 dependencies. Some
 packages may automatically install additional dependencies in the background.
-Twenty of the software packages do not explicitly indicate software dependency
+Eighteen of the software packages do not explicitly indicate software dependency
 versions. Some software package installation issues, specifically those
 occurring when manual installation of dependencies is required, may be avoided
-if versions of dependencies are specified. Sixteen software packages do not have
+if versions of dependencies are specified. Fifteen software packages do not have
 detailed instructions for installing dependencies. Sixteen software packages
@@
 Figure~\ref{Fig_Installability} shows the installability ranking of the software
-packages using AHP.
-Software packages with a higher score (ESPResSo, MechSys, OpenLB, DL\_MESO, Palabos) tend to have one set of linear installation instructions that are written as if the person doing the installation has none of the dependencies installed.
-The instructions often list compatible operating system versions and include instructions for the installation of dependencies.
-The top ranked packages often incorporate some sort of automation of the installation process and have fewer manual installation steps.
-The number of dependencies a package has does not correlate with a higher score.
-The ability to validate the installation process, often through tutorials or test examples that include expected output, is correlated with a higher score.
-Furthermore, the top seven ranked packages are noted as being alive.
+packages using AHP. Software packages with a higher score (MechSys, Palabos,
+ESPResSo, OpenLB) tend to have one set of linear installation instructions that
+are written as if the person doing the installation has none of the dependencies
+installed. The instructions often list compatible operating system versions and
+include instructions for the installation of dependencies. The top ranked
+packages often incorporate some sort of automation of the installation process
+and have fewer manual installation steps. The number of dependencies a package
+has does not correlate with a higher score. The ability to validate the
+installation process, often through tutorials or test examples that include
+expected output, is correlated with a higher score. Furthermore, the top six
+ranked packages are noted as being alive.
 
@@
 
-Seventeen of the software packages include a requirements specification artifact
+Sixteen of the software packages include a requirements specification artifact
 or explicitly reference domain theory, often only the latter. Software packages
@@
 
-Document generation tools are explicitly used by 13 software packages. Sphinx is
-used by eight of them, and Doxygen is used by eight. Several of the packages use
+Document generation tools are explicitly used by 12 software packages. Sphinx is
+used by eight of them, and Doxygen is used by seven. Several of the packages use
 both.
 
-Tutorials are available for 19 of the software packages. Generally they are
+Tutorials are available for 18 of the software packages. Generally they are
 linearly written and easy to follow. However, only eight tutorials provide an
@@
 
-Unit tests are only explicitly available for two of the software packages,
-Ludwig and Musubi.
-Code modularization of most packages allow for users to create tests with varying degrees of effort.
-These tests allow developers and users to verify the correctness of fragments of the source code, and in doing so better assess the correctness of the entire package.
+Unit tests are only explicitly available for one of the software packages,
+Ludwig. Code modularization of most packages allow for users to create tests
+with varying degrees of effort. These tests allow developers and users to verify
+the correctness of fragments of the source code, and in doing so better assess
+the correctness of the entire package.
 
 The use of continuous integration tools and techniques alludes to a more refined
-development process where faults are isolated and better recognized.
-Only three of the packages (ESPResSo, Ludwig, Musubi) mentioned applying the practice of continuous integration in their development process.
+development process where faults are isolated and better recognized. Only two of
+the packages (ESPResSo, Ludwig) mentioned applying the practice of continuous
+integration in their development process.
 
@@
 packages all include an easy to follow getting started tutorial, and most of
-these include expected output.
-The two top ranked packages, Ludwig and Musubi, indicate the use of unit testing in their documentation.
-They also incorporated continuous integration in the development process.
-Furthermore, eight of the top 10 ranked packages are noted as being alive.
+these include expected output. Only the top ranked package, Ludwig, provided
+unit testing. It and the second ranked package, ESPResSo, explicitly
+incorporated continuous integration in the development process. Furthermore,
+eight of the top 10 ranked packages are noted as being alive.
 
@@
 commands (even when following the installation guide), missing links, missing
-dependencies, and syntax errors in code files.
-In some instances the error messages were vague.
-Several automatic installation processes could not find and load dependencies.
-In these instances the installation tried to access outdated external repositories.
-Seven of the installations were recovered and verified, and one of the installations (LB3D-Prime) was assumed to be recovered due to the absence of any way to verify it.
-The installation of eight of the software packages could not be recovered.
-Most of these broken installations could not find external dependencies, encountered system incompatibilities, or displayed vague error messages.
-
-Of the 14 software packages that installed correctly and also have tutorials,
+dependencies, and syntax errors in code files. In some instances the error messages
+were vague. Several automatic installation processes could not find and load 
+dependencies. In these instances the installation tried to access outdated
+external repositories. Seven of the installations were recovered and verified,
+and one of the installations (LB3D-Prime) was assumed to be recovered due to the
+absence of any way to verify it. The installation of eight of the software
+packages could not be recovered. Most of these broken installations could not
+find external dependencies, encountered system incompatibilities, or displayed
+vague error messages.
+
+Of the 13 software packages that installed correctly and also have tutorials,
 four (pyLBM, ESPResSo++, LIMBES, Ludwig) broke during tutorial testing. All of
@@
 break during installation, or the broken installation was recoverable. All of
-the top five ranked packages have tutorials. The package pyLBM broke
+the top five ranked packages have tutorials. One of these packages, pyLBM, broke
 during tutorial testing, but a descriptive error message helped in recovery.
-Nine of the top 10 ranked packages are noted as being alive.
+Furthermore, nine of the top 10 ranked packages are noted as being alive.
 
@@
 appropriate. In this project we instead looked through each software package's
-artifacts for evidence that performance was considered. The artifacts of 18
+artifacts for evidence that performance was considered. The artifacts of 17
 software packages mentioned parallelization. This included GPU processing and
@@
 Software package artifacts were reviewed for the presence of a tutorial, a user
-manual, documented user characteristics, and a user support model. In total 19
+manual, documented user characteristics, and a user support model. In total 18
 software packages have a tutorial, 13 have a user manual, and 11 have both. The
@@
 but do have useful documentation distributed on their web-pages. Expected user
-characteristics are documented in five software packages (laboetie, LIMBES,
-Ludwig, Musubi, Palabos).
-Users are typically scientists or engineers.
-Their background is often physics, chemistry, biophysics, or mathematics.
-All but one of the packages (LIMBES) have a user support model, and many of them have multiple avenues of user support.
-The most popular avenue of support is Git, followed by email and forums.
-One software package (OpenLB) has an FAQ page.
+characteristics are documented in four software packages (laboetie, LIMBES,
+Ludwig, Palabos). Users are typically scientists or engineers. Their background
+is often physics, chemistry, biophysics, or mathematics. All but one of the
+packages (LIMBES) have a user support model, and many of them have multiple
+avenues of user support. The most popular avenue of support is Git, followed by
+email and forums. One software package (OpenLB) has an FAQ page.
 
@@
 
-Issue tracking is used in 23 software packages, 15 of which use Git, seven use
+Issue tracking is used in 22 software packages, 15 of which use Git, six use
 email, and one (SunlightLB) uses SourceForge. Most software packages that use
 Git have most of their issues closed, and only three (laboetie, lettuce,
-Sailfish) have less than 50 percent of their issues closed. All of the top five
-overall ranked packages have most of their issues closed.
-Alive packages (11 use Git issue tracking) have 64\% of their issues closed, while dead packages (3 use Git issue tracking) have 71\% of their issues closed.
-This information is presented in Table~\ref{gitrepodata}.
-Furthermore, 13 packages that use Git for issue tracking use GitHub as a version control system, while two (Palabos, waLBerla) use GitLab.
-Of the other packages, one package (SunlightLB) uses CVS for issue tracking, and seven packages do not appear to use any issue tracking system.
+Sailfish) have less than 50 percent of their issues closed. Four of the top five
+overall ranked packages (Ludwig, ESPResSo, Palabos, LUMA) have most of their
+issues closed. Fourth ranked LUMA does not use Git. Alive packages (11 use Git
+issue tracking) have 64\% of their issues closed, while dead packages (3 use Git
+issue tracking) have 71\% of their issues closed. This information is presented
+in Table~\ref{gitrepodata}. Furthermore, 13 packages that use Git for issue
+tracking use GitHub as a version control system, while two (Palabos, waLBerla)
+use GitLab. Of the other packages, one package (SunlightLB) uses CVS for issue
+tracking, and seven packages do not appear to use any issue tracking system.
 
@@
 			MP-LABS & 100.00 & 26.67 & Dead\\
-			Musubi & Not Git & 24.19 & Alive\\
 			OpenLB & Not Git & 22.43 & Alive\\
@@
 Packages with a higher percentage of comments were designated as more
-maintainable. Comments represent more than 10 percent of code files in 16
-packages, and the average percentage of code comments is about 14 percent. All
-of the top five overall ranked packages have more than the average.
-LUMA has only 0.2 percent comments, the fewest of any package.
-This package has the most lines of source code, with over four million.
-The next largest package is ESPResSo++ with one million.
+maintainable. Comments represent more than 10 percent of code files in 15
+packages, and the average percentage of code comments is about 14 percent. Four
+of the top five overall ranked packages (Ludwig, ESPResSo, Palabos, OpenLB) have
+more than the average. Fifth ranked LUMA has only 0.2 percent comments, the
+fewest of any package. This package has the most lines of source code, with over
+four million. The next largest package is ESPResSo++ with one million.
 
@@
 issues closed. Their code files are well commented with more than 10 percent of
-the code being comments. Furthermore, four of the top 5 ranked packages are
+the code being comments. Furthermore, nine of the top 10 ranked packages are
 noted as being alive. MP-LABS is noted as being dead.\\
@@
 			DL\_MESO (LBE) & 310 & 51 & 170223 & 549\\
-			ESPResSo & 1390& 83 & 195083 & 140\\
-			ESPResSo++ & 1406& 30 & 165194 & 118\\
-			HemeLB & 1102& 44 & 123806 & 112\\
+			ESPResSo & 1309& 86 & 186700 & 143\\
+			ESPResSo++ & 5328& 66 & 969196 & 182\\
+			HemeLB & 1065& 48 & 95104 & 89\\
 			laboetie & 133& 1 & 48403 & 364\\		
@@
 			LB3D & 99 & 76 & 39766 & 402\\
-			LB3D-Prime & 23& 6 & 12944 & 563\\
-			lbmpy& 250 & 29 & 61632 & 247\\
-			lettuce & 73 & 1 & 7660 & 105\\
+			LB3D-Prime & 2 3& 6 & 12944 & 563\\
+			lbmpy& 201 & 28 & 46489 & 231\\
+			lettuce & 62 & 0 & 5529 & 89\\
 			LIMBES & 26 & 1 & 4872 & 187\\
-			Ludwig & 954 & 32 & 162518 & 170\\
-			LUMA & 314 & 19 & 4399723 & 14011\\
-			MechSys & 333 & 3 & 95707 & 287\\
+			Ludwig & 859 & 32 & 109811 & 128\\
+			LUMA & 312 & 19 & 4370670 & 14000\\
+			MechSys & 324 & 3 & 85543 & 264\\
 			MP-LABS & 307 & 3 & 43124 & 140\\
-			Musubi & 1347 & 1839 & 281879 & 209\\
-			OpenLB & 1438 & 7 & 218406 & 152\\
-			Palabos & 1974 & 71 & 563841 & 286\\
-			pyLBM & 272 & 108 & 37234 & 137\\
+			OpenLB & 1104 & 5 & 209034 & 189\\
+			Palabos & 1829 & 67 & 547623 & 299\\
+			pyLBM & 258 & 85 & 32314 & 125\\
 			Sailfish & 632 & 11 & 69398 & 110\\
 			SunlightLB & 36 & 1 & 7646 & 212\\
-			TCLB & 594 & 7 & 49156 & 83\\
-			waLBerla & 2643 & 69 & 873988 & 331\\
+			TCLB & 535 & 7 & 43226 & 81\\
+			waLBerla & 2395 & 67 & 848146 & 353\\
 			\hline
@@
 
-All of the packages appear to have consistent indentation and formatting.
-Only HemeLB, LUMA, and Musubi explicitly identify coding standards that are used during development.
-Generally, the software packages use consistent, distinctive, and meaningful code identifiers.
-Only four packages (LB2D-Prime, LB3D-Prime, LIMBES, MP-LABS) appear to use vague identifiers, such as single letters for variables.
-Symbolic constants were observed in the source code of 13 packages.
-The constants are used for various parameters, mathematical constants, and matrix definitions.
-All of the packages are reasonably well commented, with the comments clearly indicating what is being done (as opposed to how it is being done).
-Domain algorithms are noted in the source code of 11 packages.
-Table~\ref{moduledata} suggests that the software packages are modularized to various degrees.
-When observing the source code files, it was found that 14 of the packages have a consistent style and order of function parameters.
+All of the packages appear to have consistent indentation and formatting. Only
+LUMA and HemeLB explicitly identify coding standards that are used during
+development. Generally, the software packages use consistent, distinctive, and
+meaningful code identifiers. Only four packages (LB2D-Prime, LB3D-Prime, LIMBES,
+MP-LABS) appear to use vague identifiers, such as single letters for variables.
+Symbolic constants were observed in the source code of 12 packages. The
+constants are used for various parameters, mathematical constants, and matrix
+definitions. All of the packages are reasonably well commented, with the
+comments clearly indicating what is being done (as opposed to how it is being
+done). Domain algorithms are noted in the source code of 11 packages.
+Table~\ref{moduledata} suggests that the software packages are modularized to
+various degrees. When observing the source code files, it was found that 13 of
+the packages have a consistent style and order of function parameters.
 
@@
 what is being done in the source code. The source code is well modularized and
-structured. Furthermore, all of the top five ranked packages are noted as being
+structured. Furthermore, four of the top five ranked packages are noted as being
 alive.
@@
 
-Looking at the top three ranked packages: ESPResSo had achieved a
-relative high score in installability, surface usability, maintainability, reusability, and visibility and transparency.
-Ludwig scored high in surface correctness and verifiability, surface robustness, surface usability, maintainability, and visibility and transparency.
-Palabos scored high in installability, surface reliability, surface robustness, maintainability, understandability, and visibility and transparency.
+Looking at the top three ranked packages: Ludwig scored high in surface
+correctness and verifiability, surface robustness, surface usability,
+maintainability, and visibility and transparency. ESPResSo had achieved a
+relative high score in installability, surface correctness and verifiability,
+surface usability, maintainability, reusability, and visibility and
+transparency. Palabos scored high in installability, surface reliability,
+surface robustness, maintainability, and understandability.
 
@@
 
-, without a need for a rigorous software development model. The developer of
-ESPResSo noted that while no formal model is used, their development model is
-something similar to a combination of agile and waterfall development models.
-
-Improving the software engineering
-education or experience of developers is also an idea that was brought up by
-several developers.
+Improving the software engineering education or experience of developers is also
+an idea that was brought up by several developers.
 
@@
 
-\subsection{Artifacts Present} \label{artifacts}
-
-This subsection answers the research question: What artifacts are present in
-current software packages?
+\subsection{What artifacts are present in current software projects?}
+\label{artifacts}
 
@@
 and rare artifacts in Table~\ref{artifactspresent}. Common artifacts were found
-in 16 to 24 ($>$63\%) of the software packages. Less common artifacts were found
-in 8 to 15 (30-63\%) of the software packages. Rare artifacts were found in 1 to
-7 ($<$30\%) of the software packages.
+in 16 to 23 ($>$70\%) of the software packages. Less common artifacts were found
+in 8 to 15 (35-70\%) of the software packages. Rare artifacts were found in 1 to
+7 ($<$35\%) of the software packages.
 
@@
 \begin{center}
-		\begin{tabular}{ p{8 cm} }
-			\hline
-			Common\\
-			\hline
-			Authors / Developers List\\
-			Bug Tracker\\
-			Dependency Notes\\
-			Installation Guide / Instructions\\
-			License\\
-			List of Related Articles / Publications\\
-			Makefile / Build File\\
-			README File\\
-			Requirements Specification / Theory Notes\\
-			Tutorial\\
-			\hline
-			Less Common\\
-			\hline
-			Change Log / Release Notes\\
-			Design Documentation\\
-			Functional Specification / Notes\\
-			Performance Information / Notes\\
-			Test Plan / Report / Script / Data / Cases\\
-			User Manual/Guide\\
-			Version Control\\
-			\hline
-			Rare\\
-			\hline
-			API Documentation\\
-			Developer / Contributor Manual / Guide\\
-			FAQ / Forum\\
-			Verification and Validation Plan / Notes\\
-			Video Guide (including YouTube)\\
-			\hline
+\begin{tabular}{ p{5.3 cm} p{4.9 cm} p{5 cm}}
+\toprule
+\textbf{Common} & \textbf{Less Common} & \textbf{Rare} \\
+\midrule
+Authors / Developers List & Change Log / Release Notes & API Documentation\\
+Bug Tracker & Design Documentation & Developer / Contributor Manual\\
+Dependency Notes & Functional Spec. / Notes & FAQ / Forum\\
+Installation Guide / Instructions & Performance Information & Verification and Validation Plan\\
+Requirements Spec. / Theory Notes & Test Plan / Script / Cases & Video Guide (including YouTube)\\
+List of Related Publications & User Manual/Guide & \\
+Makefile / Build File & Version Control & \\
+README File & & \\
+License & & \\
+Tutorial & & \\
+\bottomrule
 \end{tabular}
@@
 
-\subsubsection{Common Artifacts}
-
-All of the top four AHP ranked packages, ESPREsSo, Ludwig, Palabos, and OpenLB, have each of the commonly found artifacts, except only three of them have a
+All of the top four AHP ranked packages, ESPREsSo, Ludwig, OpenLB, and Palabos,
+have each of the commonly found artifacts, except only three of them have a
 requirements specification or theory notes. Palabos is the only one of these
 four packages that does not have an artifact from this category.
-
-The common artifacts contribute to the quality of the software in the following
-ways. A list of authors and developers helps potential users and contributors
-contact project members to answer questions affecting many software qualities. A
-bug tracker helps with organizing improvements to the software. Dependency notes
-help users install the software, as does an installation guide. Makefiles or
-other automated build files decrease human error in the installation process.
-Licenses promote usage of the software. Requirements specifications, linked
-theory notes, and related articles and publications help make the software more
-understandable, decrease ambiguity, and help with verifying its correctness.
-README files also help with understandability. The tutorials help users become
-familiar with using the software.
-
-\subsubsection{Less Common Artifacts} \label{lesscommon}
 
@@
 data collection. Despite the broken link, Palabos does have a detailed and
-informative website. Another one of the top four packages, Ludwig, does not
-appear to have publicly visible design documentation. A third package from the list, OpenLB, does not appear to use a version control system. It is possible
-that such a system is used by the package since its website notes package
+informative website. Another one (Ludwig) of the top four packages does not
+appear to have publicly visible design documentation. A third package (OpenLB)
+from the list does not appear to use a version control system. It is possible
+that such a system is used by the package, since its website notes package
 version numbers, but the artifacts do not explicitly state the use of such a
@@
 
-The less common artifacts also contribute to software quality. A change log or
-release notes improve traceability of the software. Design documentation helps
-with maintaining, modifying, and reusing the software. Version control also
-helps to improve these qualities, as well as with traceability. Functional
-specifications and notes clarify the software, contributing to usability and
-understandability. A user manual also helps with those qualities and with
-installability, and correctness and verifiability. Performance notes suggest
-that performance was considered when developing the software. Test plans,
-scripts, and cases, help verify correctness.
-
-\subsubsection{Rare Artifacts}
-
-It is not common for the top four AHP ranked packages to have many of the rare
-artifacts. None of the top four packages have any explicit API documentation.
-Three of these packages (ESPREsSo, Ludwig, Palabos) have information on
-contributing to the project. Two of them (OpenLB, Palabos) have a FAQ section or
-forum. One (OpenLB) has verification and validation notes, and a video guide of
-the software.
-
-Although the artifacts were rarely found in our set of LBM software, they also
-contribute to software quality. API documentation helps with reusing the
-software. Developer and contributor manuals and guides help with
-maintainability, visibility and transparency. FAQs and forums improve usability
-of software. Verification and validation notes can be used to help check if a
-system meets specifications. Video guides can contribute to many software
-qualities, depending on the content of the video.
-
-\subsection{Tools Used} \label{tools}
-
-This subsection answers the research question: What tools (development,
-dependencies, project management) are used by current software packages?
+The top four ranked packages do not have many of the rare artifacts. None of the
+top four packages have any explicit API documentation. Three of these packages
+(ESPREsSo, Ludwig, Palabos) have information on contributing to the project. Two
+of them (OpenLB, Palabos) have a FAQ section or forum. One (OpenLB) has
+verification and validation notes, and a video guide of the software.
+
+\subsection{What tools (development, dependencies, project management) are used by current software packages?} \label{tools}
 
@@
 and evolution of software, software processes, and artifacts
-\citep{ghezzi1991fundamentals}. Many tools are used by LBM software packages.
-The tools noted here are subdivided into development tools, dependencies, and
-project management tools.
-
-\subsubsection{Development Tools}
+\citep{ghezzi1991fundamentals}. Many tools are used by LBM software packages, as
+summarized in Table~\ref{tbl_tools}.  The tools are subdivided into development
+tools, dependencies, and project management tools.
+
+\begin{table}
+	\begin{center}
+	\begin{tabular}{ p{5.3 cm} p{4.9 cm} p{5 cm}}
+	\toprule
+	\textbf{Development Tools} & \textbf{Dependencies} & \textbf{Project
+	Management Tools} \\
+	\midrule
+	Continuous Integration & Build Automation Tools & Collaboration Tools\\
+	Code Editors & Technical Libraries & Email\\
+	Development Environment & Domain Specific Libraries & Change Tracking Tools\\
+	Runtime Environments &  & Version Control Tools\\
+	Compilers &  & Document Generation Tools\\
+	Unit Testing Tools &  & \\
+	Correctness Verification Tools &  & \\
+	\bottomrule
+	\end{tabular}
+	\caption{Observed development tools, dependencies and project management tools} \label{tbl_tools}
+	\end{center}
+\end{table}
 
@@
 part of them, unlike dependencies that remain in the application once it is
-released \citep{ghezzi1991fundamentals}. The following type of development tools
-were explicitly noted in the artifacts or web-pages of the 24 LBM packages that
-were assessed. It is likely that other tools, such as debuggers, were used but
-are not specified in our sources.
-
-	\begin{multicols}{2}	
-		\begin{itemize}
-
-			\item Continuous Integration Tools
-			\item Code Editors
-			\item Development Environment
-			\item Runtime Environments
-			\item Compilers
-			\item Unit Testing Tools
-			\item Correctness Verification Tools
-
-		\end{itemize}
-	\end{multicols}
-
-The above tools can verify the correctness of software during its development.
-Only three (ESPResSo, Ludwig, Musubi) of the software packages that were assessed mentioned using continuous integration tools, like Travis CI.
-Code editors and compilers were explicitly noted to have been used by several packages, and were likely used by all of them.
-One of the packages (Ludwig) explicitly noted the use of proprietary unit testing code written in C.
-Likewise, the use of proprietary code for verifying the correctness of output was noted by one (pyLBM) of the developers.
-It is likely that similar tools were used when developing other software packages.
-
-\subsubsection{Dependencies}
-
-The following types of dependencies were explicitly noted in the artifacts or
-web-pages of the 24 LBM packages that were assessed. It is possible that other
-types of dependencies are part of these software packages, but are not clearly
-specified in their artifacts or web sites and because of that they are not
-listed here.
-
-	\begin{multicols}{2}	
-		\begin{itemize}
-
-			\item Build Automation Tools
-			\item Technical Libraries
-			\item Domain Specific Libraries
-
-		\end{itemize}
-	\end{multicols}
-
-Most of the software packages use some sort of build automation tools, most
-commonly Make. They also all use various technical and domain specific
-libraries. Technical libraries include visualization (e.g. Matplotlib, ParaView,
-Pygame, VTK), data analysis (e.g. Anaconda, Torch), and message passing
-libraries (e.g. MPICH, Open MPI, PyZMQ). Domain specific libraries are
-scientific computing libraries (e.g. SciPy). Libraries that are not explicitly
-stated in artifacts, or were not noted during our observations, may fall outside
-of these categories.
-
-\subsubsection{Project Management Tools}
+released \citep{ghezzi1991fundamentals}. Although not shown in
+Table~\ref{tbl_tools}, debuggers were also likely used.  Only two (ESPResSo and
+Ludwig) packages mentioned continuous integration tools, like Travis CI. Code
+editors and compilers were explicitly noted to have been used by several
+packages, and were likely used by all of them. One of the packages (Ludwig)
+explicitly noted the use of proprietary unit testing code written in C.
+Likewise, the use of proprietary code for verifying the correctness of output
+was noted by one of the developers (pyLBM). Similar tools were likely used when
+developing the other software packages.
+
+For the dependency tools (Table~\ref{tbl_tools}), we observed that most of the
+software packages use some sort of build automation tools, most commonly Make.
+They all use various technical and domain specific libraries. Technical
+libraries include visualization (e.g.\ Matplotlib, ParaView, Pygame, VTK), data
+analysis (e.g.\ Anaconda, Torch), and message passing libraries (e.g.\ MPICH,
+Open MPI, PyZMQ). Domain specific libraries include scientific computing
+libraries (e.g.\ SciPy). Some used libraries will be missing, if they are not
+explicitly stated in the artifacts we observed.
 
 Many of the software packages that were assessed were developed by teams of two
-or more people. Their work needed to be coordinated and managed. The following
-types of project management tools were explicitly noted in the artifacts,
-web-pages, or interviews with the developers of the 24 LBM packages that were
-assessed. As with development tools and dependencies, it is possible that other
-types of project management tools were used to coordinate and manage the
-projects but are not specified and because of that they are not listed here.
-
-	\begin{multicols}{2}	
-		\begin{itemize}
-
-			\item Collaboration Tools
-			\item Email
-			\item Change Tracking Tools
-			\item Version Control Tools
-			\item Document Generation Tools
-
-		\end{itemize}
-	\end{multicols}
-
-Collaboration tools are noted as being used when developing the software
-projects. Most often email and video conferencing is used. Project management
-software was not explicitly mentioned, but it is possible that some of the
-projects use such software. Many of the projects are located on GitHub, and its
-developers use the platform to help manage their projects, especially bug
-related issues. Most of the projects appear to use change tracking and version
-control tools. They often use GitHub or GitLab for this. One package
-(SunlightLB) uses CVS. Document generation tools are mentioned in the artifacts
-of 13 of the projects. The tools Sphinx and Doxygen are explicitly used in this
-capacity.
-
-\subsection{Principles, Processes, and Methodologies} \label{prinprocmeth}
-
-This subsection answers the research question: What principles, processes, and
-methodologies are used in the development of current software packages?
-
-The points and conclusions in this subsection come from developer interviews and
-reviews of software package artifacts.
+or more people. Their work needed to be coordinated and managed.
+Table~\ref{tbl_tools} shows the types of project management tools that were
+explicitly noted in the artifacts, web-pages, or interviews with the developers.
+As with development tools and dependencies, it is possible that other types of
+project management tools are used, but they were not visible in the artifacts to
+which we had access.  Collaboration tools are noted as being used when
+developing the software projects. Most often email and video conferencing is
+used. Project management software was not explicitly mentioned, but it is
+possible that some of the projects use such software. Many of the projects are
+located on GitHub, and its developers use the platform to help manage their
+projects, especially bug related issues. Most of the projects appear to use
+change tracking and version control tools. They often use GitHub or GitLab for
+this. One package (SunlightLB) uses CVS. Document generation tools are mentioned
+in the artifacts of 12 of the projects. The tools Sphinx and Doxygen are
+explicitly used in this capacity. \wss{look at what the other papers say are
+commonly used tools and bring that information into this section.}
+
+\subsection{What principles, processes, and methodologies are used?}
+\label{prinprocmeth}
 
@@
 the LBM packages are fairly small, so it is feasible for them to be organized
-without the need for such models.
-
-Seven of the software packages contain artifacts recording the general
-development process and status. Eleven of the packages explicitly convey that they would accept outside contributors, but generally the teams are centralized, often working at the same institution.
+without the need for such models.  \wss{some repetition from the previous
+section - should try to figure out how to remove the repetition.}
+
+Seven of the software packages contain artifacts outlining the general
+development process, like basic instructions on how to contribute. Eleven of the
+packages explicitly convey that they would accept outside contributors, but
+generally the teams are centralized, often working at the same institution.
 
@@
 
-Thirteen (54\%) of the 24 software packages use GitHub for managing the project,
-including nine (60\%) of the 15 alive packages, and four of the nine (44\%) dead
+Thirteen (57\%) of the 23 software packages use GitHub for managing the project,
+including nine (64\%) of the 14 alive packages, and four of the nine (44\%) dead
 packages. Two projects (Palabos, WaLBerla) use GitLab. This could be indicative
@@
 simulation engine and interface. The use of document generation tools is
-mentioned in the artifacts of 13 software packages, and was noted during
+mentioned in the artifacts of 12 software packages, and was noted during
 interviews with developers. Sphinx and Doxygen are the tools that were
@@
 
-\subsection{Pain Points} \label{painpoints}
+\subsection{What are the pain points?} \label{painpoints}
 
@@
 improvement? How should processes, methodologies and tools be changed to improve
-software development and software quality?
-
-Developers were asked to comment on obstacles in their development process,
-obstacles encountered by users, and potential future obstacles.
+software development and software quality?  Developers were asked to comment on
+obstacles in their development process, obstacles encountered by users, and
+potential future obstacles.
 
@@
 
-\subsection{Quality Recommendations} \label{qualityrecommentations}
+\subsection{What is the current software quality for the projects?}
+\label{qualityrecommentations}
 
@@
 	by several top ranked packages, including OpenLB, Ludwig, LUMA, and Palabos.
-	\item Provide instructions for uninstalling the software. These were included with pyLBM.
+	\item Provide instructions for uninstalling the software. These were
+	included with pyLBM.
 \end{itemize}
@@
 	including ESPResSo, Ludwig, LUMA, and OpenLB.
-	\item Ensure the above information is easy to find. Consider adding it to the user manual. The information is easy to find in most top ranked packages.
-	\item Development teams should include both domain experts and experienced software developers. This suggestion is based on developer comments.
-	\item Use and make public (on package website or GitHub) detailed documentation.
-Consider using automatic document generation tools like Doxygen, Drasil, or Sphinx.
-This was done by all of the top five ranked packages.
-	\item Provide detailed tutorials that include expected output, like the \href{https://www.walberla.net/doxygen/index.html}{waLBerla tutorials}.
-	\item Use unit tests during development and make them public (on package website or GitHub). This was available for Ludwig.
-	\item Modularize the source code, separate components, hide information  behind well defined interfaces. This is suggested in SCS literature, and in developer comments.
-	\item Use continuous integration tools (Bamboo, Jenkins, and Travis CI) and processes during development. This was done by top ranked packages ESPResSo and Ludwig.
+	\item Ensure the above information is easy to find. Consider adding it to
+	the user manual. The information is easy to find in most top ranked
+	packages.
+	\item Development teams should include both domain experts and experienced
+	software developers. This suggestion is based on developer comments.
+	\item Use and make public (on package website or GitHub) detailed
+	documentation. Consider using automatic document generation tools like
+	Doxygen, Drasil, or Sphinx. This was done by all of the top five ranked
+	packages.
+	\item Provide detailed tutorials that include expected output, like the
+	\href{https://www.walberla.net/doxygen/index.html}{waLBerla tutorials}.
+	\item Use unit tests during development and make them public (on package
+	website or GitHub). This was available for Ludwig.
+	\item Modularize the source code, separate components, hide information
+	behind well defined interfaces. This is suggested in SCS literature, and in
+	developer comments.
+	\item Use continuous integration tools (Bamboo, Jenkins, and Travis CI) and
+	processes during development. This was done by top ranked packages ESPResSo
+	and Ludwig.
 \end{itemize}
@@
 \begin{itemize}
-	\item Include descriptive error messages where appropriate. This was done by most packages that encountered a fault.
-	\item In case automatic installation of dependencies fails, the system should indicate to the user what dependencies need to be installed manually.
-This was done by many packages, including ESPResSo++, Ludwig, LUMA, pyLBM, TCLB.
-	\item The packages should include detailed tutorials, including dependencies, expected output, and any additional supplementary documentation that may be required.
-This was done by many packages, including waLBerla, Palabos, MechSys, LUMA, and pyLBM.
+	\item Include descriptive error messages where appropriate. This was done by
+	most packages that encountered a fault.
+	\item In case automatic installation of dependencies fails, the system
+	should indicate to the user what dependencies need to be installed manually.
+	This was done by many packages, including ESPResSo++, Ludwig, LUMA, pyLBM,
+	TCLB.
+	\item The packages should include detailed tutorials, including
+	dependencies, expected output, and any additional supplementary
+	documentation that may be required. This was done by many packages,
+	including waLBerla, Palabos, MechSys, LUMA, and pyLBM.
 \end{itemize}
@@
 \begin{itemize}
-	\item The system must provide descriptive error messages when it encounters unexpected input, including incorrect data types, empty input, and missing files or links.
-Nineteen of the software packages behaved reasonably when tested with unexpected input.
+	\item The system must provide descriptive error messages when it encounters
+	unexpected input, including incorrect data types, empty input, and missing
+	files or links. Eighteen of the software packages behaved reasonably when
+	tested with unexpected input.
 \end{itemize}
@@
 \begin{itemize}
-	\item Integrate parallelization tools and techniques to reduce processing time. Consider GPU processing, CUDA, and MPI. Eighteen software packages mentioned parallelization in their artifacts.
-	\item The user should be able to choose to process their model on either the CPU or GPU. ESPResSo, lbmpy, lettuce, pyLBM, and TCLB allow for either CPU or GPU processing.
+	\item Integrate parallelization tools and techniques to reduce processing
+	time. Consider GPU processing, CUDA, and MPI. Seventeen software packages
+	mentioned parallelization in their artifacts.
+	\item The user should be able to choose to process their model on either the
+	CPU or GPU. ESPResSo, lbmpy, lettuce, pyLBM, and TCLB allow for either CPU
+	or GPU processing.
 \end{itemize}
@@
 	\item Include user hardware and software requirements documentation.
-Hardware requirements are rarely listed in the software packages that were assessed.
-Only those offering GPU processing (ESPResSo, lbmpy, lettuce, pyLBM, Sailfish, TCLB) mentioned any hardware requirements.
-On the other hand, some sort of software requirements were available for all packages, even if only some dependencies or a compatible operating system were mentioned.
-	\item Provide a user tutorial that indicates the expected output. This was done by many packages, including waLBerla, Palabos, MechSys, LUMA, pyLBM.
-	\item Provide a detailed user manual.
-It should identify elements of user interfaces, and identify all requirements to model a system.
-Thirteen packages have a user manual.
-ESPResSo has a well detailed manual.
-	\item State appropriate fluid dynamics problems that the software is designed to model in the documentation, and explicitly state the limits of the software. This is done in the ESPResSo user guide.
-	\item Provide documentation that details the background theory information, or provide a reference to such information.
-This was done by several top ranked packages, including ESPResSo, Ludwig, LUMA, and OpenLB.
-	\item Identify expected user characteristics. LIMBES, Ludwig, laboetie, and Palabos did this. The importance of specifying user characteristics is discussed in \citep{smith2007requirements}.
-	\item Keep all documentation in one location. This was done by top ranked packages.
-	\item Do not duplicate artifact information. This is suggested as good software development practice. Duplicate information is difficult to maintain. 
-	\item Maintain a user support model (Git, email, forum, FAQ). This was done by all of the top five ranked packages.
-	\item If possible, consider using popular user-friendly software languages like Python.
-Especially consider this for parts of the source code that is likely to be modified or reviewed by users.
-ESPResSo and Sailfish use Python to shorten development time and improve usability.
+	Hardware requirements are rarely listed in the software packages that were
+	assessed. Only those offering GPU processing (ESPResSo, lbmpy, lettuce,
+	pyLBM, Sailfish, TCLB) mentioned any hardware requirements. On the other
+	hand, some sort of software requirements were available for all packages,
+	even if only some dependencies or a compatible operating system were
+	mentioned.
+	\item Provide a user tutorial that indicates the expected output. This was
+	done by many packages, including waLBerla, Palabos, MechSys, LUMA, pyLBM.
+	\item Provide a detailed user manual. It should identify elements of user
+	interfaces, and identify all requirements to model a system. Thirteen
+	packages have a user manual. ESPResSo has a well detailed manual.
+	\item State appropriate fluid dynamics problems that the software is
+	designed to model in the documentation, and explicitly state the limits of
+	the software. This is done in the ESPResSo user guide.
+	\item Provide documentation that details the background theory information,
+	or provide a reference to such information. This was done by several top
+	ranked packages, including ESPResSo, Ludwig, LUMA, and OpenLB.
+	\item Identify expected user characteristics. LIMBES, Ludwig, laboetie, and
+	Palabos did this. The importance of specifying user characteristics is
+	discussed in \citep{smith2007requirements}.
+	\item Keep all documentation in one location. This was done by top ranked
+	packages.
+	\item Do not duplicate artifact information. This is suggested as good
+	software development practice. Duplicate information is difficult to
+	maintain. 
+	\item Maintain a user support model (Git, email, forum, FAQ). This was done
+	by all of the top five ranked packages.
+	\item If possible, consider using popular user-friendly software languages
+	like Python. Especially consider this for parts of the source code that is
+	likely to be modified or reviewed by users. ESPResSo and Sailfish use Python
+	to shorten development time and improve usability.
 \end{itemize}
@@
 \begin{itemize}
-	\item Keep artifacts updated. This is suggested as good software maintenance practice.
-	\item Ideally include most of the common and less common artifacts listed in Section~\ref{artifacts}.
-	\item Include version numbers and release notes for all major source code and artifact releases. The top five ranked software packages include release notes.
-	\item Have a defined process for accepting contributions, and make public documentation for making contributions to the project.
-The top five ranked software packages include information on how to contribute.
-	\item Use an issue tracker (Git, email, SourceForge, other) to manage bugs and changes.
-Issues should be regularly reviewed and closed.
-All but three (laboetie, lettuce, Sailfish) of the software packages that use Git have most of their issues closed.
-	\item Use a version control system (GitHub, CVS). Four (ESPREsSo, Ludwig, LUMA, Palabos) of the top five ranked packages use a version control system.
-	\item Source code needs to be well commented.
-Typically, more than 10 percent of LBM package source code is comments, as presented in Table~\ref{gitrepodata}.
-Four of the top five overall ranked packages (Ludwig, ESPResSo, Palabos, OpenLB) have about 20 percent of their source code as comments.
-	\item Modularize the source code, separate components, hide information behind well defined interfaces. This is suggested in SCS literature, and in developer comments.
-	\item Eliminate code duplication. This is suggested as good software development practice. Duplicate code is difficult to maintain.
-	\item If possible, consider using popular user-friendly software languages like Python.
-Especially consider this for parts of the source code that is likely to be modified or reviewed by users.
-ESPResSo and Sailfish use Python to address several software qualities, including maintainability.
+	\item Keep artifacts updated. This is suggested as good software maintenance
+	practice.
+	\item Ideally include most of the common and less common artifacts listed in
+	Section~\ref{artifacts}.
+	\item Include version numbers and release notes for all major source code
+	and artifact releases. The top five ranked software packages include release
+	notes.
+	\item Have a defined process for accepting contributions, and make public
+	documentation for making contributions to the project. The top five ranked
+	software packages include information on how to contribute. 
+	\item Use an issue tracker (Git, email, SourceForge, other) to manage bugs
+	and changes. Issues should be regularly reviewed and closed. All but three
+	(laboetie, lettuce, Sailfish) of the software packages that use Git have
+	most of their issues closed.
+	\item Use a version control system (GitHub, CVS). Four (ESPREsSo, Ludwig,
+	LUMA, Palabos) of the top five ranked packages use a version control system.
+	\item Source code needs to be well commented. Typically, more than 10
+	percent of LBM package source code is comments, as presented in
+	Table~\ref{gitrepodata}. Four of the top five overall ranked packages
+	(Ludwig, ESPResSo, Palabos, OpenLB) have about 20 percent of their source
+	code as comments.
+	\item Modularize the source code, separate components, hide information
+	behind well defined interfaces. This is suggested in SCS literature, and in
+	developer comments.
+	\item Eliminate code duplication. This is suggested as good software
+	development practice. Duplicate code is difficult to maintain.
+	\item If possible, consider using popular user-friendly software languages
+	like Python. Especially consider this for parts of the source code that is
+	likely to be modified or reviewed by users. ESPResSo and Sailfish use Python
+	to address several software qualities, including maintainability.
 	\item Consider the recommended points for addressing traceability.
@@
 \begin{itemize}
-	\item Modularize the source code, separate components, hide information behind well defined interfaces. This is suggested in SCS literature, and in developer comments.
-	\item If possible, consider using popular user-friendly software languages like Python.
-Especially consider this for parts of the source code that is likely to be modified or reviewed by users.
-ESPResSo and Sailfish use Python to address several software qualities, including modifiability.
-	\item Consider future source code modifiability as early as the design stage of development. This is suggested as good software development practice.
-	\item Consider flexibility of data structures and data storage in the design stage. The package pyLBM redeveloped data structures to ease future changes.
+	\item Modularize the source code, separate components, hide information
+	behind well defined interfaces. This is suggested in SCS literature, and in
+	developer comments.
+	\item If possible, consider using popular user-friendly software languages
+	like Python. Especially consider this for parts of the source code that is
+	likely to be modified or reviewed by users. ESPResSo and Sailfish use Python
+	to address several software qualities, including modifiability.
+	\item Consider future source code modifiability as early as the design stage
+	of development. This is suggested as good software development practice.
+	\item Consider flexibility of data structures and data storage in the design
+	stage. The package pyLBM redeveloped data structures to ease future changes.
 \end{itemize}
@@
 \begin{itemize}
-	\item Modularize the source code, separate components, hide information behind well defined interfaces. This is suggested in SCS literature, and in developer comments.
-	\item Document module interfaces in design and developer documentation. This is suggested as good software development practice.
-	\item Provide API documentation, if applicable. Only one (ESPResSo) of the top five ranked packages provided API documentation.
+	\item Modularize the source code, separate components, hide information
+	behind well defined interfaces. This is suggested in SCS literature, and in
+	developer comments.
+	\item Document module interfaces in design and developer documentation. This
+	is suggested as good software development practice.
+	\item Provide API documentation, if applicable. Only one (ESPResSo) of the
+	top five ranked packages provided API documentation.
 \end{itemize}
@@
 \begin{itemize}
-	\item Adopt a coding standard and document it in artifacts, including some examples. A coding standard needs to be part of continuous integration. Only
-	LUMA, HemeLB, and Musubi explicitly identify coding standards. 
-	\item Use consistent, distinctive, and meaningful code identifiers. Twenty of the 24 software packages use such code identifiers.
-	\item Use symbolic constants. Thirteen of the 24 software packages use such constants.
-	\item Identify algorithms that are used in source code comments.
-Document them in the artifacts.
-Cite relevant external sources if needed.
-Domain algorithms are noted in the source code of 11 packages.
-	\item Add meaningful comments. Indicate what is being done in each section of source code. All of the top ranked packages have meaningful comments in their source code.
-	\item Modularize the source code, separate components, hide information behind well defined interfaces. This is suggested in SCS literature, and in developer comments.
-	\item Provide a user manual that identifies all technical and problem modeling requirements. Thirteen software packages have a user manual.
-	\item State appropriate fluid dynamics problems that the software is designed to model in the documentation, and explicitly state the limits of the software. This is done in the ESPResSo user guide. 
-	\item Consider adding a FAQ section to the documentation. This helped resolve some usability issues for OpenLB.
+	\item Adopt a coding standard and document it in artifacts, including some
+	examples. A coding standard needs to be part of continuous integration. Only
+	LUMA and HemeLB explicitly identify coding standards. 
+	\item Use consistent, distinctive, and meaningful code identifiers. Nineteen
+	of the 23 software packages use such code identifiers.
+	\item Use symbolic constants. Twelve of the 23 software packages use such
+	constants.
+	\item Identify algorithms that are used in source code comments. Document
+	them in the artifacts. Cite relevant external sources if needed. Domain
+	algorithms are noted in the source code of 11 packages.
+	\item Add meaningful comments. Indicate what is being done in each section
+	of source code. All of the top ranked packages have meaningful comments in
+	their source code.
+	\item Modularize the source code, separate components, hide information
+	behind well defined interfaces. This is suggested in SCS literature, and in
+	developer comments.
+	\item Provide a user manual that identifies all technical and problem
+	modeling requirements. Thirteen software packages have a user manual.
+	\item State appropriate fluid dynamics problems that the software is
+	designed to model in the documentation, and explicitly state the limits of
+	the software. This is done in the ESPResSo user guide. 
+	\item Consider adding a FAQ section to the documentation. This helped
+	resolve some usability issues for OpenLB.
 \end{itemize}
@@
 \begin{itemize}
-	\item Update all relevant documentation when a change to the software is made. This is suggested as good software development practice.
-	\item Use automatic document generation tools (Doxygen, Drasil, Sphinx) to limit the time spent on updating documentation. This was done by all of the top five ranked packages.
-	\item Provide a developer's guide to help orient developers to the artifacts, source code, and system architecture so that they can better document changes.
-Top ranked package OpenLB has a developer guide.
+	\item Update all relevant documentation when a change to the software is
+	made. This is suggested as good software development practice.
+	\item Use automatic document generation tools (Doxygen, Drasil, Sphinx) to
+	limit the time spent on updating documentation. This was done by all of the
+	top five ranked packages.
+	\item Provide a developer's guide to help orient developers to the
+	artifacts, source code, and system architecture so that they can better
+	document changes. Top ranked package OpenLB has a developer guide. 
 \end{itemize}
@@
 \begin{itemize}
-	\item Summarize the development process that is used.
-Provide information on how new users can contribute.
-Identify the development model by name (waterfall, agile, etc.), if appropriate.
-Seven of the software packages have some artifacts outlining the general development process.
-Eleven packages have information on how to contribute.
-	\item Identify the development environment. Eight of the 24 software packages identify the development environment.
-	\item Update all relevant documentation when a change to the software is made. This is suggested as good software development practice.
+	\item Summarize the development process that is used. Provide information on
+	how new users can contribute. Identify the development model by name
+	(waterfall, agile, etc.), if appropriate.  Seven of the software packages
+	have some artifacts outlining the general development process. Eleven
+	packages have information on how to contribute. 
+	\item Identify the development environment. Eight of the 23 software
+	packages identify the development environment.
+	\item Update all relevant documentation when a change to the software is
+	made. This is suggested as good software development practice.
 	\item Include notes with all releases. Nine of the packages include release notes.
-	\item Communicate all changes within the development team. This suggestion is based on developer comments.
-	\item Use continuous integration processes and tools (Bamboo, Jenkins, and Travis CI). This was done by top ranked packages ESPResSo and Ludwig.
-	\item Consider peer review processes to assess contributions and ensure the tracking of information. High ranked package ESPResSo uses a peer review process for contributions. 
-	\item Use project management tools, including change and version control tools (GitHub, GitLab, CVS), collaboration tools (GitHub, GitLab), and document generation tools (Doxygen, Drasil, Sphinx).
-This was done by all of the top five ranked packages.
+	\item Communicate all changes within the development team. This suggestion
+	is based on developer comments.
+	\item Use continuous integration processes and tools (Bamboo, Jenkins, and
+	Travis CI). This was done by top ranked packages ESPResSo and Ludwig.
+	\item Consider peer review processes to assess contributions and ensure the
+	tracking of information. High ranked package ESPResSo uses a peer review
+	process for contributions. 
+	\item Use project management tools, including change and version control
+	tools (GitHub, GitLab, CVS), collaboration tools (GitHub, GitLab), and
+	document generation tools (Doxygen, Drasil, Sphinx). This was done by all of
+	the top five ranked packages. 
 \end{itemize}
@@
 \begin{itemize}
-	\item Test output against automatically calculated or known correct results. High ranked package ESPResSo does this.
-	\item Automate the testing of output. Run tests after all code changes. High ranked package ESPResSo does this.
-	\item Consider peer review processes and task based inspection to assess contributions. High ranked package ESPResSo does this.
-	\item Update all relevant design documentation when a change to the software is made. This is suggested as good software development practice.
+	\item Test output against automatically calculated or known correct results.
+	High ranked package ESPResSo does this.
+	\item Automate the testing of output. Run tests after all code changes. High
+	ranked package ESPResSo does this.
+	\item Consider peer review processes and task based inspection to assess
+	contributions. High ranked package ESPResSo does this.
+	\item Update all relevant design documentation when a change to the software
+	is made. This is suggested as good software development practice.
 \end{itemize}
@@
 
-\subsection{Designation Comparison} \label{comparison}
+\subsection{How does software designated as high quality by this methodology
+compare with top-rated software by the community?} \label{comparison}
 
@@
 	\begin{tabular}{ p{2cm}p{4.5cm}p{3.5cm}}
-			\hline
+		\toprule
 		Rank & This Assessment & Domain Expert\\
-			\hline
-			1&Ludwig (N/A) &Palabos\\
-			2&ESPResSo (N/A) &OpenLB\\
-			3&Palabos (1)&LUMA\\
-			4&OpenLB (2)&ASL\\
-			5&LUMA (3)&ch4-project\\
-			6&pyLBM (17)&Open FSI\\
-			7&DL\_MESO (LBE) (12)& WaLBerla\\
-			8&waLBerla (7)& LIFE\\
-			9&Sailfish (9)&Sailfish\\
-			10&laboetie (N/A)&HemeLB\\		
-			\hline
+		\midrule
+		1 & Ludwig (N/A) & Palabos\\
+		2 & ESPResSo (N/A) & OpenLB\\
+		3 & Palabos (1)& LUMA\\
+		4 & OpenLB (2)& ASL\\
+		5 & LUMA (3)& ch4-project\\
+		6 & pyLBM (17)& Open FSI\\
+		7 & DL\_MESO (LBE) (12)& WaLBerla\\
+		8 & waLBerla (7)& LIFE\\
+		9 & Sailfish (9)& Sailfish\\
+		10 & laboetie (N/A)& HemeLB\\		
+		\bottomrule
 	\end{tabular}
@@
 technology issues like broken links. This issue arose with the measurement of
-Palabos, which had a broken link to its user manual, as noted in Section
-~\ref{lesscommon}.
+Palabos, which had a broken link to its user manual, as noted in
+Section~\ref{artifacts}.
 
@@
 - read \citet{WieseEtAl2019} (Naming the Pain in Developing Scientific Software)
+
+- change \citet{ghezzi1991fundamentals} to \citet{GhezziEtAl2003} (need to check
+each instance, add page numbers).

### 2022-01-04 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex

--- 2021-12-29 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
@@
 	\begin{tabular}{ p{4cm}p{3cm}p{3cm} }
-		\hline
+		\toprule
 		Name & Released & Updated\\
-		\hline
+		\midrule
 		DL\_MESO (LBE) & unclear & 2020 Mar\\
@@
 		waLBerla & 2008 Aug & 2020 Jul\\
-		\hline
+		\bottomrule
 	\end{tabular}
@@
 	\begin{tabular}{ p{4cm}p{3cm}p{3cm} }
-		\hline
+		\toprule
 		Name & Released & Updated\\
-		\hline
+		\midrule
 		HemeLB & 2007 Jun & 2018 Aug\\
@@
 		SunlightLB & 2005 Sep & 2012 Nov\\
-		\hline
+		\bottomrule
 	\end{tabular}
@@
 		\begin{tabular}{ p{3cm}llllllll}
-			\hline
+			\toprule
 			Name & Dim & Pll & Com & Rflx & MFl & Turb & CGE & OS\\
-			\hline
+			\midrule
 			DL\_MESO (LBE) & 2, 3 & MPI/OMP & Y & Y & Y & Y & Y & W, M, L\\
@@
 			waLBerla & 2, 3 & MPI & Y & Y & Y & Y & Y & L\\
-			\hline
+			\bottomrule
 		\end{tabular}
@@
 		\begin{tabular}{ p{3.5cm}p{3.5cm}p{3.5cm}p{2.5cm} }
-			\hline
+			\toprule
 			Name & $\%$ Issues Closed & $\%$ Code Comments & Status\\
-			\hline
+			\midrule
 			DL\_MESO (LBE) & Not Git & 8.06& Alive\\
@@
 			waLBerla & 72.90 & 22.62 & Alive\\
-			\hline
+			\bottomrule
 		\end{tabular}
@@
 		\begin{tabular}{ p{3.5cm}p{2cm}p{2.5cm}p{2cm}p{2.5cm} }
-			\hline
+			\toprule
 			Name & Text Files & Binary Files & LOC & Avg. LOC / Text File\\
-			\hline
+			\midrule
 			DL\_MESO (LBE) & 310 & 51 & 170223 & 549\\
@@
 			waLBerla & 2643 & 69 & 873988 & 331\\
-			\hline
+			\bottomrule
 		\end{tabular}
@@
 
-\subsubsection{Common Artifacts}
-
-All of the top four AHP ranked packages, ESPREsSo, Ludwig, Palabos, and OpenLB, have each of the commonly found artifacts, except only three of them have a
+All of the top four AHP ranked packages, ESPREsSo, Ludwig, Palabos, and OpenLB,
+have each of the commonly found artifacts, except only three of them have a
 requirements specification or theory notes. Palabos is the only one of these
 four packages that does not have an artifact from this category.
-
-The common artifacts contribute to the quality of the software in the following
-ways. A list of authors and developers helps potential users and contributors
-contact project members to answer questions affecting many software qualities. A
-bug tracker helps with organizing improvements to the software. Dependency notes
-help users install the software, as does an installation guide. Makefiles or
-other automated build files decrease human error in the installation process.
-Licenses promote usage of the software. Requirements specifications, linked
-theory notes, and related articles and publications help make the software more
-understandable, decrease ambiguity, and help with verifying its correctness.
-README files also help with understandability. The tutorials help users become
-familiar with using the software.
-
-\subsubsection{Less Common Artifacts} \label{lesscommon}
 
@@
 
-The less common artifacts also contribute to software quality. A change log or
-release notes improve traceability of the software. Design documentation helps
-with maintaining, modifying, and reusing the software. Version control also
-helps to improve these qualities, as well as with traceability. Functional
-specifications and notes clarify the software, contributing to usability and
-understandability. A user manual also helps with those qualities and with
-installability, and correctness and verifiability. Performance notes suggest
-that performance was considered when developing the software. Test plans,
-scripts, and cases, help verify correctness.
-
-\subsubsection{Rare Artifacts}
-
-It is not common for the top four AHP ranked packages to have many of the rare
-artifacts. None of the top four packages have any explicit API documentation.
-Three of these packages (ESPREsSo, Ludwig, Palabos) have information on
-contributing to the project. Two of them (OpenLB, Palabos) have a FAQ section or
-forum. One (OpenLB) has verification and validation notes, and a video guide of
-the software.
-
-Although the artifacts were rarely found in our set of LBM software, they also
-contribute to software quality. API documentation helps with reusing the
-software. Developer and contributor manuals and guides help with
-maintainability, visibility and transparency. FAQs and forums improve usability
-of software. Verification and validation notes can be used to help check if a
-system meets specifications. Video guides can contribute to many software
-qualities, depending on the content of the video.
-
-\subsection{Tools Used} \label{tools}
-
-This subsection answers the research question: What tools (development,
-dependencies, project management) are used by current software packages?
+The top four ranked packages do not have many of the rare artifacts. None of the
+top four packages have any explicit API documentation. Three of these packages
+(ESPREsSo, Ludwig, Palabos) have information on contributing to the project. Two
+of them (OpenLB, Palabos) have a FAQ section or forum. One (OpenLB) has
+verification and validation notes, and a video guide of the software.
+
+\subsection{What tools (development, dependencies, project management) are used
+by current software packages?} \label{tools}
 
@@
 part of them, unlike dependencies that remain in the application once it is
-released \citep{ghezzi1991fundamentals}. The following type of development tools
-were explicitly noted in the artifacts or web-pages of the 24 LBM packages that
-were assessed. It is likely that other tools, such as debuggers, were used but
-are not specified in our sources.
-
-	\begin{multicols}{2}	
-		\begin{itemize}
-
-			\item Continuous Integration Tools
-			\item Code Editors
-			\item Development Environment
-			\item Runtime Environments
-			\item Compilers
-			\item Unit Testing Tools
-			\item Correctness Verification Tools
-
-		\end{itemize}
-	\end{multicols}
-
-The above tools can verify the correctness of software during its development.
-Only three (ESPResSo, Ludwig, Musubi) of the software packages that were assessed mentioned using continuous integration tools, like Travis CI.
-Code editors and compilers were explicitly noted to have been used by several packages, and were likely used by all of them.
-One of the packages (Ludwig) explicitly noted the use of proprietary unit testing code written in C.
-Likewise, the use of proprietary code for verifying the correctness of output was noted by one (pyLBM) of the developers.
-It is likely that similar tools were used when developing other software packages.
-
-\subsubsection{Dependencies}
-
-The following types of dependencies were explicitly noted in the artifacts or
-web-pages of the 24 LBM packages that were assessed. It is possible that other
-types of dependencies are part of these software packages, but are not clearly
-specified in their artifacts or web sites and because of that they are not
-listed here.
-
-	\begin{multicols}{2}	
-		\begin{itemize}
-
-			\item Build Automation Tools
-			\item Technical Libraries
-			\item Domain Specific Libraries
-
-		\end{itemize}
-	\end{multicols}
-
-Most of the software packages use some sort of build automation tools, most
-commonly Make. They also all use various technical and domain specific
-libraries. Technical libraries include visualization (e.g. Matplotlib, ParaView,
-Pygame, VTK), data analysis (e.g. Anaconda, Torch), and message passing
-libraries (e.g. MPICH, Open MPI, PyZMQ). Domain specific libraries are
-scientific computing libraries (e.g. SciPy). Libraries that are not explicitly
-stated in artifacts, or were not noted during our observations, may fall outside
-of these categories.
-
-\subsubsection{Project Management Tools}
+released \citep{ghezzi1991fundamentals}. Although not shown in
+Table~\ref{tbl_tools}, debuggers were also likely used.  Only three (ESPResSo, 
+Ludwig and Musubi) packages mentioned continuous integration tools, like Travis CI. Code
+editors and compilers were explicitly noted to have been used by several
+packages, and were likely used by all of them. One of the packages (Ludwig)
+explicitly noted the use of proprietary unit testing code written in C.
+Likewise, the use of proprietary code for verifying the correctness of output
+was noted by one of the developers (pyLBM). Similar tools were likely used when
+developing the other software packages.
+
+For the dependency tools (Table~\ref{tbl_tools}), we observed that most of the
+software packages use some sort of build automation tools, most commonly Make.
+They all use various technical and domain specific libraries. Technical
+libraries include visualization (e.g.\ Matplotlib, ParaView, Pygame, VTK), data
+analysis (e.g.\ Anaconda, Torch), and message passing libraries (e.g.\ MPICH,
+Open MPI, PyZMQ). Domain specific libraries include scientific computing
+libraries (e.g.\ SciPy). Some used libraries will be missing, if they are not
+explicitly stated in the artifacts we observed.
 
 Many of the software packages that were assessed were developed by teams of two
-or more people. Their work needed to be coordinated and managed. The following
-types of project management tools were explicitly noted in the artifacts,
-web-pages, or interviews with the developers of the 24 LBM packages that were
-assessed. As with development tools and dependencies, it is possible that other
-types of project management tools were used to coordinate and manage the
-projects but are not specified and because of that they are not listed here.
-
-	\begin{multicols}{2}	
-		\begin{itemize}
-
-			\item Collaboration Tools
-			\item Email
-			\item Change Tracking Tools
-			\item Version Control Tools
-			\item Document Generation Tools
-
-		\end{itemize}
-	\end{multicols}
-
-Collaboration tools are noted as being used when developing the software
-projects. Most often email and video conferencing is used. Project management
-software was not explicitly mentioned, but it is possible that some of the
-projects use such software. Many of the projects are located on GitHub, and its
-developers use the platform to help manage their projects, especially bug
-related issues. Most of the projects appear to use change tracking and version
-control tools. They often use GitHub or GitLab for this. One package
-(SunlightLB) uses CVS. Document generation tools are mentioned in the artifacts
-of 13 of the projects. The tools Sphinx and Doxygen are explicitly used in this
-capacity.
-
-\subsection{Principles, Processes, and Methodologies} \label{prinprocmeth}
-
-This subsection answers the research question: What principles, processes, and
-methodologies are used in the development of current software packages?
-
-The points and conclusions in this subsection come from developer interviews and
-reviews of software package artifacts.
+or more people. Their work needed to be coordinated and managed.
+Table~\ref{tbl_tools} shows the types of project management tools that were
+explicitly noted in the artifacts, web-pages, or interviews with the developers.
+As with development tools and dependencies, it is possible that other types of
+project management tools are used, but they were not visible in the artifacts to
+which we had access.  Collaboration tools are noted as being used when
+developing the software projects. Most often email and video conferencing is
+used. Project management software was not explicitly mentioned, but it is
+possible that some of the projects use such software. Many of the projects are
+located on GitHub, and its developers use the platform to help manage their
+projects, especially bug related issues. Most of the projects appear to use
+change tracking and version control tools. They often use GitHub or GitLab for
+this. One package (SunlightLB) uses CVS. Document generation tools are mentioned
+in the artifacts of 12 of the projects. The tools Sphinx and Doxygen are
+explicitly used in this capacity. \wss{look at what the other papers say are
+commonly used tools and bring that information into this section.}
+
+\subsection{What principles, processes, and methodologies are used?}
+\label{prinprocmeth}
 
@@
 the LBM packages are fairly small, so it is feasible for them to be organized
-without the need for such models.
+without the need for such models.  \wss{some repetition from the previous
+section - should try to figure out how to remove the repetition.}
 
 Seven of the software packages contain artifacts recording the general
-development process and status. Eleven of the packages explicitly convey that they would accept outside contributors, but generally the teams are centralized, often working at the same institution.
+development process and status. Eleven of the packages explicitly convey that
+they would accept outside contributors, but generally the teams are centralized,
+often working at the same institution.
 
@@
 	done by several top ranked packages, including ESPResSo and LUMA.
-	\item Provide complete installation instructions. This was done by all of the top five ranked packages.
+	\item Provide complete installation instructions. This was done by all of
+	the top five ranked packages.
 	\item Installation instructions should be written as if the user does not
@@
 \begin{itemize}
-	\item The system must provide descriptive error messages when it encounters unexpected input, including incorrect data types, empty input, and missing files or links.
-Nineteen of the software packages behaved reasonably when tested with unexpected input.
+	\item The system must provide descriptive error messages when it encounters
+	unexpected input, including incorrect data types, empty input, and missing
+	files or links. Nineteen of the software packages behaved reasonably when
+	tested with unexpected input.
 \end{itemize}
@@
 \begin{itemize}
-	\item Integrate parallelization tools and techniques to reduce processing time. Consider GPU processing, CUDA, and MPI. Eighteen software packages mentioned parallelization in their artifacts.
-	\item The user should be able to choose to process their model on either the CPU or GPU. ESPResSo, lbmpy, lettuce, pyLBM, and TCLB allow for either CPU or GPU processing.
+	\item Integrate parallelization tools and techniques to reduce processing
+	time. Consider GPU processing, CUDA, and MPI. Eighteen software packages
+	mentioned parallelization in their artifacts.
+	\item The user should be able to choose to process their model on either the
+	CPU or GPU. ESPResSo, lbmpy, lettuce, pyLBM, and TCLB allow for either CPU
+	or GPU processing.
 \end{itemize}
@@
 \begin{itemize}
-	\item Adopt a coding standard and document it in artifacts, including some examples. A coding standard needs to be part of continuous integration. Only
+	\item Adopt a coding standard and document it in artifacts, including some
+	examples. A coding standard needs to be part of continuous integration. Only
 	LUMA, HemeLB, and Musubi explicitly identify coding standards. 
-	\item Use consistent, distinctive, and meaningful code identifiers. Twenty of the 24 software packages use such code identifiers.
-	\item Use symbolic constants. Thirteen of the 24 software packages use such constants.
-	\item Identify algorithms that are used in source code comments.
-Document them in the artifacts.
-Cite relevant external sources if needed.
-Domain algorithms are noted in the source code of 11 packages.
-	\item Add meaningful comments. Indicate what is being done in each section of source code. All of the top ranked packages have meaningful comments in their source code.
-	\item Modularize the source code, separate components, hide information behind well defined interfaces. This is suggested in SCS literature, and in developer comments.
-	\item Provide a user manual that identifies all technical and problem modeling requirements. Thirteen software packages have a user manual.
-	\item State appropriate fluid dynamics problems that the software is designed to model in the documentation, and explicitly state the limits of the software. This is done in the ESPResSo user guide. 
-	\item Consider adding a FAQ section to the documentation. This helped resolve some usability issues for OpenLB.
+	\item Use consistent, distinctive, and meaningful code identifiers. Twenty
+	of the 24 software packages use such code identifiers.
+	\item Use symbolic constants. Thirteen of the 24 software packages use such
+	constants.
+	\item Identify algorithms that are used in source code comments. Document
+	them in the artifacts. Cite relevant external sources if needed. Domain
+	algorithms are noted in the source code of 11 packages.
+	\item Add meaningful comments. Indicate what is being done in each section
+	of source code. All of the top ranked packages have meaningful comments in
+	their source code.
+	\item Modularize the source code, separate components, hide information
+	behind well defined interfaces. This is suggested in SCS literature, and in
+	developer comments.
+	\item Provide a user manual that identifies all technical and problem
+	modeling requirements. Thirteen software packages have a user manual.
+	\item State appropriate fluid dynamics problems that the software is
+	designed to model in the documentation, and explicitly state the limits of
+	the software. This is done in the ESPResSo user guide. 
+	\item Consider adding a FAQ section to the documentation. This helped
+	resolve some usability issues for OpenLB.
 \end{itemize}
@@
 \begin{itemize}
-	\item Summarize the development process that is used.
-Provide information on how new users can contribute.
-Identify the development model by name (waterfall, agile, etc.), if appropriate.
-Seven of the software packages have some artifacts outlining the general development process.
-Eleven packages have information on how to contribute.
-	\item Identify the development environment. Eight of the 24 software packages identify the development environment.
-	\item Update all relevant documentation when a change to the software is made. This is suggested as good software development practice.
+	\item Summarize the development process that is used. Provide information on
+	how new users can contribute. Identify the development model by name
+	(waterfall, agile, etc.), if appropriate.  Seven of the software packages
+	have some artifacts outlining the general development process. Eleven
+	packages have information on how to contribute. 
+	\item Identify the development environment. Eight of the 24 software
+	packages identify the development environment.
+	\item Update all relevant documentation when a change to the software is
+	made. This is suggested as good software development practice.
 	\item Include notes with all releases. Nine of the packages include release notes.
-	\item Communicate all changes within the development team. This suggestion is based on developer comments.
-	\item Use continuous integration processes and tools (Bamboo, Jenkins, and Travis CI). This was done by top ranked packages ESPResSo and Ludwig.
-	\item Consider peer review processes to assess contributions and ensure the tracking of information. High ranked package ESPResSo uses a peer review process for contributions. 
-	\item Use project management tools, including change and version control tools (GitHub, GitLab, CVS), collaboration tools (GitHub, GitLab), and document generation tools (Doxygen, Drasil, Sphinx).
-This was done by all of the top five ranked packages.
+	\item Communicate all changes within the development team. This suggestion
+	is based on developer comments.
+	\item Use continuous integration processes and tools (Bamboo, Jenkins, and
+	Travis CI). This was done by top ranked packages ESPResSo and Ludwig.
+	\item Consider peer review processes to assess contributions and ensure the
+	tracking of information. High ranked package ESPResSo uses a peer review
+	process for contributions. 
+	\item Use project management tools, including change and version control
+	tools (GitHub, GitLab, CVS), collaboration tools (GitHub, GitLab), and
+	document generation tools (Doxygen, Drasil, Sphinx). This was done by all of
+	the top five ranked packages. 
 \end{itemize}
@@
 
-Nine packages do not use GitHub, so they do not have a measure of
-repository stars. Looking at the repository stars of the other 15 packages, we
-can observe a slight pattern where packages that have been highly ranked by our
-assessment do have more stars than lower ranked packages. The best ranked
-package (ESPResSo) has the second most number of stars. The ninth ranked package
+Nine packages do not use GitHub, so they do not have a measure of repository
+stars. Looking at the repository stars of the other 15 packages, we can observe
+a slight pattern where packages that have been highly ranked by our assessment
+do have more stars than lower ranked packages. The best ranked package
+(ESPResSo) has the second most number of stars. The ninth ranked package
 (Sailfish) has the most number of stars. Packages designated as lower quality
-often do not use GitHub or GitLab, or have few stars.
-HemeLB only has 12 stars, but its location recently moved from a different GitHub repository.
-Our assessment of this package might not be accurate.
-The number of stars does not necessarily represent the perceptions of the community, but for lack of an alternative measure we will use it this way.
+often do not use GitHub or GitLab, or have few stars. HemeLB only has 12 stars,
+but its location recently moved from a different GitHub repository. Our
+assessment of this package might not be accurate. The number of stars does not
+necessarily represent the perceptions of the community, but for lack of an
+alternative measure we will use it this way.
 
@@
 		\begin{tabular}{ p{3.5cm}p{2.5cm}p{2cm}p{2cm}p{2cm} }
-			\hline
+			\toprule
 			Name & Our Ranking & Repository Stars & Repository Star Rank & Repository Watches \\
-			\hline
+			\midrule
 			ESPResSo &1& 145 &2& 19\\
@@
 			LatBo.jl &24& 17 &10& 8\\			
-			\hline
+			\bottomrule
 		\end{tabular}
@@
 Table~\ref{rankingcomparison} compares the top 10 ranked LBM software packages
-with a LBM package ranking made by a domain expert.
-Five (Palabos, OpenLB, LUMA, Sailfish, waLBerla) of the top 10 ranked packages in this assessment are also found in the domain expert's top 10 ranking.
-Interestingly, the first three of these five packages are listed in the same order on both lists.
-Looking at the remaining packages, the top two ranked packages in this state of the practice assessment (ESPResSo, Ludwig) are not on the domain expert's list.
-Perhaps the intended applications of these packages, complex fluids for Ludwig and soft matter research for ESPResSo, do not align with the research interests of the domain expert, and this is why they did not make it onto their list.
-This reasoning may also apply to eighth ranked Musubi.
-Moving down the list, pyLBM and DL\_MESO(LBE) did not make the domain expert's top 10 list but were mentioned by the domain expert.
-Looking at the remainder of the domain expert's list, ASL is a general purpose tool for solving partial differential equations and may have fallen out of scope of the authoritative lists that were used to identify the initial LBM software list.
-The domain expert's fifth ranked package (ch4-project) was on the initial software list of this assessment, but was removed due to a lack of documentation.
-Open FSI is a new project that uses Palabos.
-It was made public the year before data was collected for this state of the practice exercise.
-It may not have been on authoritative lists due to its young age.
-Similarly, LIFE was also made public recently.
-Finally, the domain expert's 10th ranked package (HemeLB) was ranked 21st in this state of the practice assessment.
+with a LBM package ranking made by a domain expert. Five (Palabos, OpenLB, LUMA,
+Sailfish, waLBerla) of the top 10 ranked packages in this assessment are also
+found in the domain expert's top 10 ranking. Interestingly, the first three of
+these five packages are listed in the same order on both lists. Looking at the
+remaining packages, the top two ranked packages in this state of the practice
+assessment (ESPResSo, Ludwig) are not on the domain expert's list. Perhaps the
+intended applications of these packages, complex fluids for Ludwig and soft
+matter research for ESPResSo, do not align with the research interests of the
+domain expert, and this is why they did not make it onto their list. This
+reasoning may also apply to eighth ranked Musubi. Moving down the list, pyLBM
+and DL\_MESO(LBE) did not make the domain expert's top 10 list but were
+mentioned by the domain expert. Looking at the remainder of the domain expert's
+list, ASL is a general purpose tool for solving partial differential equations
+and may have fallen out of scope of the authoritative lists that were used to
+identify the initial LBM software list. The domain expert's fifth ranked package
+(ch4-project) was on the initial software list of this assessment, but was
+removed due to a lack of documentation. Open FSI is a new project that uses
+Palabos. It was made public the year before data was collected for this state of
+the practice exercise. It may not have been on authoritative lists due to its
+young age. Similarly, LIFE was also made public recently. Finally, the domain
+expert's 10th ranked package (HemeLB) was ranked 21st in this state of the
+practice assessment.
 
@@
 	\begin{tabular}{ p{2cm}p{4.5cm}p{3.5cm}}
-			\hline
+		\toprule
 		Rank & This Assessment & Domain Expert\\
-			\hline
-			1&ESPResSo (N/A) &Palabos\\
-			2&Ludwig (N/A) &OpenLB\\
-			3&Palabos (1)&LUMA\\
-			4&OpenLB (2)&ASL\\
-			5&LUMA (3)&ch4-project\\
-			6&pyLBM (16)&Open FSI\\
-			7&DL\_MESO (LBE) (11)& WaLBerla\\
-			8&Musubi (N/A)& LIFE\\
-			9&Sailfish (9)&Sailfish\\
-			10&WaLBerla (7)&HemeLB\\		
-			\hline
+		\midrule
+		1 & ESPResSo (N/A) & Palabos\\
+		2 & Ludwig (N/A) & OpenLB\\
+		3 & Palabos (1) & LUMA\\
+		4 & OpenLB (2) & ASL\\
+		5 & LUMA (3) & ch4-project\\
+		6 & pyLBM (16) & Open FSI\\
+		7 & DL\_MESO (LBE) (11) & WaLBerla\\
+		8 & Musubi (N/A) & LIFE\\
+		9 & Sailfish (9) & Sailfish\\
+		10 & WaLBerla (7) & HemeLB\\		
+		\bottomrule
 	\end{tabular}
@@
 obscure within an artifact or web-page. The use of unit testing and continuous
-integration was mentioned in the artifacts of only three (ESPResSo, Ludwig, Musubi) packages.
-However, interviews suggested a more frequent use of both unit testing and continuous integration in the development processes than what was observed from the initial survey of the artifacts.
-For example, OpenLB, pyLBM, and TCLB use such methods during development despite this not being explicitly clear from an analysis of the material available online.
+integration was mentioned in the artifacts of only three (ESPResSo, Ludwig,
+Musubi) packages. However, interviews suggested a more frequent use of both unit
+testing and continuous integration in the development processes than what was
+observed from the initial survey of the artifacts. For example, OpenLB, pyLBM,
+and TCLB use such methods during development despite this not being explicitly
+clear from an analysis of the material available online.
 
@@
 \begin{itemize}
-	\item Provide a detailed user manual.
-It should identify elements of user interfaces, and identify all requirements to model a system.
-Thirteen packages have a user manual.
-ESPResSo has a well detailed manual.
-	\item State appropriate fluid dynamics problems that the software is designed to model in the documentation, and explicitly state the limits of the software. This is done in the ESPResSo user guide.
-	\item Include detailed tutorials, including dependencies, expected
-	output, and any additional supplementary documentation that may be required. This was done by many packages, including waLBerla, Palabos, MechSys, LUMA, and pyLBM.
+	\item Provide a detailed user manual. It should identify elements of user
+	interfaces, and identify all requirements to model a system. Thirteen
+	packages have a user manual. ESPResSo has a well detailed manual.
+	\item State appropriate fluid dynamics problems that the software is
+	designed to model in the documentation, and explicitly state the limits of
+	the software. This is done in the ESPResSo user guide.
+	\item Include detailed tutorials, including dependencies, expected output,
+	and any additional supplementary documentation that may be required. This
+	was done by many packages, including waLBerla, Palabos, MechSys, LUMA, and
+	pyLBM.
 	\item Keep all documentation in one location. This was done by top ranked packages.
-	\item If possible, consider using popular user-friendly software languages like Python.
-Especially consider this for parts of the source code that is likely to be modified or reviewed by users.
-ESPResSo and Sailfish use Python to address several software qualities, including maintainability, modifiability and usability.
-	\item Modularize the source code, separate components, hide information behind well defined interfaces. This is suggested in SCS literature, and in developer comments.
-	\item Include descriptive error messages where appropriate. This was done by most packages that encountered a fault.
-	\item Summarize the development process that is used.
-Provide information on how new users can contribute.
-Seven of the software packages have some artifacts outlining the general development process.
-Eleven packages have information on how to contribute.
-	\item Consider peer review processes and task based inspection to assess contributions. High ranked package ESPResSo does this.
-	\item Use continuous integration tools (Bamboo, Jenkins, and Travis CI) and processes during development. This was done by top ranked packages ESPResSo and Ludwig.
-	\item Use project management tools, including change and version control tools (GitHub, GitLab, CVS), collaboration tools (GitHub, GitLab), and document generation tools (Doxygen, Drasil, Sphinx).
-This was done by all of the top five ranked packages.
+	\item If possible, consider using popular user-friendly software languages
+	like Python. Especially consider this for parts of the source code that is
+	likely to be modified or reviewed by users. ESPResSo and Sailfish use Python
+	to address several software qualities, including maintainability,
+	modifiability and usability.
+	\item Modularize the source code, separate components, hide information
+	behind well defined interfaces. This is suggested in SCS literature, and in
+	developer comments.
+	\item Include descriptive error messages where appropriate. This was done by
+	most packages that encountered a fault.
+	\item Summarize the development process that is used. Provide information on
+	how new users can contribute. Seven of the software packages have some
+	artifacts outlining the general development process. Eleven packages have
+	information on how to contribute.
+	\item Consider peer review processes and task based inspection to assess
+	contributions. High ranked package ESPResSo does this.
+	\item Use continuous integration tools (Bamboo, Jenkins, and Travis CI) and
+	processes during development. This was done by top ranked packages ESPResSo
+	and Ludwig.
+	\item Use project management tools, including change and version control
+	tools (GitHub, GitLab, CVS), collaboration tools (GitHub, GitLab), and
+	document generation tools (Doxygen, Drasil, Sphinx). This was done by all of
+	the top five ranked packages.
 \end{itemize}

### 2022-01-04 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Complete pass through LBM SOP - start of reorg of current, best and future practices

@@
 %% Template article for Elsevier's document class `elsarticle'
-%% with harvard style bibliographic references
+%% with Harvard style bibliographic references
 
@@
 All 24 packages on the short list can be installed on some Unix-like systems.
-Eight packages could be installed on Windows, and five on macOS.
-Operating system requirements are listed in the documentation of 20 software packages, but compatible operating system versions are only specified in four packages (ESPResSo, LUMA, Mechsys, Sailfish).
-All but one of the software packages (TCLB) were tested on Ubuntu for this state of the practice assessment.
-TCLB was tested on CentOS, since this operating system is mentioned in its installation instructions.
+Eight packages could be installed on Windows, and five on macOS. Operating
+system requirements are listed in the documentation of 20 software packages, but
+compatible operating system versions are only specified in four packages
+(ESPResSo, LUMA, Mechsys, Sailfish). All but one of the software packages (TCLB)
+were tested on Ubuntu for this state of the practice assessment. TCLB was tested
+on CentOS, since this operating system is mentioned in its installation
+instructions.
 
@@
 publications. In the latter case the user may need to spend significant time to
-find this information.
+find this information.  \wss{More on requirements as a pain point is given in
+Weiss, and other papers cited by Carver (2021).  Isn't generally requirements in
+the SE sense of following a template or similar.}
 
@@
 
+\section{Comparison to Community Ranking} \label{repmetrics}
+
+Table~\ref{repometrics} presents our LBM software package rankings along with
+the repository ranking metrics of each software package. Nine packages do not
+use GitHub, so they do not have a measure of repository stars. Looking at the
+repository stars of the other 15 packages, we can observe a slight pattern where
+packages that have been highly ranked by our assessment do have more stars than
+lower ranked packages. The best ranked package (ESPResSo) has the second most
+number of stars. The ninth ranked package (Sailfish) has the most number of
+stars. Packages designated as lower quality often do not use GitHub or GitLab,
+or have few stars. HemeLB only has 12 stars, but its location recently moved
+from a different GitHub repository. Our assessment of this package might not be
+accurate. The number of stars does not necessarily represent the perceptions of
+the community, but for lack of an alternative measure we will use it this way.
+Besides missing data, another threat to the validity of this comparison is the
+varying ages of the repositories. Older packages have been able to accumulate
+stars and watches for longer than newer packages. The true quality of new
+packages may not be reflected in their stars and watches.
+
+The repository watches column contains even less data to compare since two of
+the packages (Palabos, waLBerla) use GitLab, which does not include this metric.
+The pattern that can be observed with this metric is very similar to that of the
+stars metric. The ninth ranked package (Sailfish) has the most number of
+watches. The best ranked package (ESPResSo) has the second most number of
+watches.
+
+\begin{table}
+\begin{center}
+	\begin{tabular}{ p{3.5cm}p{2.5cm}p{2cm}p{2cm}p{2cm} }
+		\toprule
+		Name & Our Ranking & Repository Stars & Repository Star Rank & Repository Watches \\
+		\midrule
+		ESPResSo & 1 & 145 & 2 & 19\\
+		Ludwig & 2 & 27 & 8 & 6\\
+		Palabos & 3 & 34 & 6 & GitLab\\
+		OpenLB & 4 & No Git & No Git & No Git\\
+		LUMA & 5 & 33 & 7 & 12\\
+		pyLBM & 6 & 95 & 3 & 10\\
+		DL\_MESO (LBE) & 7 & No Git & No Git & No Git \\
+		Musubi & 8 & No Git & No Git & No Git \\
+		Sailfish & 9 & 186 & 1 & 41\\
+		waLBerla & 10 & 20 & 9 & GitLab\\
+		laboetie & 11 & 4 & 13 & 5\\
+		TCLB & 12 & 95 & 3 & 16\\
+		MechSys & 13 & No Git & No Git & No Git\\
+		lettuce & 14 & 48 & 4 & 5\\
+		ESPResSo++ & 15 & 35 & 5 & 12\\
+		MP-LABS & 16 & 12 & 11 & 2\\			
+		SunlightLB & 17 & No Git & No Git & No Git\\
+		LB3D & 18 & No Git & No Git & No Git\\			
+		LIMBES & 19 & No Git & No Git & No Git\\
+		LB2D-Prime & 20 & No Git & No Git & No Git\\		
+		HemeLB & 21 & 12 & 11 & 12\\
+		lbmpy & 22 &  11 & 12 & 2\\	
+		LB3D-Prime & 23 & No Git & No Git & No Git\\	
+		LatBo.jl & 24 & 17 & 10 & 8\\			
+		\bottomrule
+	\end{tabular}
+	\caption{Repository Ranking Metrics} \label{repometrics}
+\end{center}
+\end{table}
+
+Table~\ref{rankingcomparison} compares the top 10 ranked LBM software packages
+with a LBM package ranking made by a domain expert. Five (Palabos, OpenLB, LUMA,
+Sailfish, waLBerla) of the top 10 ranked packages in this assessment are also
+found in the domain expert's top 10 ranking. Interestingly, the first three of
+these five packages are listed in the same order on both lists. Looking at the
+remaining packages, the top two ranked packages in this state of the practice
+assessment (ESPResSo, Ludwig) are not on the domain expert's list. Perhaps the
+intended applications of these packages, complex fluids for Ludwig and soft
+matter research for ESPResSo, do not align with the research interests of the
+domain expert. This reasoning may also apply to eighth ranked Musubi. Moving
+down the list, pyLBM and DL\_MESO(LBE) did not make the domain expert's top 10
+list but were mentioned by the domain expert. Looking at the remainder of the
+domain expert's list, ASL is a general purpose tool for solving partial
+differential equations and may have fallen out of scope of the authoritative
+lists that were used to identify the initial LBM software list. The domain
+expert's fifth ranked package (ch4-project) was on the initial software list of
+this assessment, but was removed due to a lack of documentation. Open FSI is a
+new project that uses Palabos. It was made public the year before data was
+collected for this state of the practice exercise. It may not have been on
+authoritative lists due to its young age. Similarly, LIFE was also made public
+recently. Finally, the domain expert's 10th ranked package (HemeLB) was ranked
+21st in this state of the practice assessment.
+
+\begin{table}
+\begin{center}
+	\begin{tabular}{ p{2cm}p{4.5cm}p{3.5cm}}
+		\toprule
+		Rank & This Assessment & Domain Expert\\
+		\midrule
+		1 & ESPResSo (N/A) & Palabos\\
+		2 & Ludwig (N/A) & OpenLB\\
+		3 & Palabos (1) & LUMA\\
+		4 & OpenLB (2) & ASL\\
+		5 & LUMA (3) & ch4-project\\
+		6 & pyLBM (16) & Open FSI\\
+		7 & DL\_MESO (LBE) (11) & WaLBerla\\
+		8 & Musubi (N/A) & LIFE\\
+		9 & Sailfish (9) & Sailfish\\
+		10 & WaLBerla (7) & HemeLB\\		
+		\bottomrule
+	\end{tabular}
+	\caption{Ranking Comparison} \label{rankingcomparison}
+\end{center}
+\end{table}
+
 \section{Qualitative Findings From Developer Interviews} \label{interviewresults}
-
-Improving the software engineering education or experience of developers is also
-an idea that was brought up by several developers.
 
@@
 documentation seem like they would fit better under understandability, don't you
-think?}  \pmm{After reviewing the definitions of the two qualities, I think you are right.
-I think that the point presented in the first sentence of this paragraph belongs here in traceability (but could be reworded to note that ESPResSo also has detailed change logs in their docs - change logs are also available for Palabos and waLBerla), but the rest of the paragraph can be moved to understandability.} \wss{If ESPresSo scored high (I think it did), we can use that to
-further promote the value of documentation.  Also a threat to validity is that
-we have implicitly assumed documentation is valuable when measuring quality.}
+think?}  \pmm{After reviewing the definitions of the two qualities, I think you
+are right. I think that the point presented in the first sentence of this
+paragraph belongs here in traceability (but could be reworded to note that
+ESPResSo also has detailed change logs in their docs - change logs are also
+available for Palabos and waLBerla), but the rest of the paragraph can be moved
+to understandability.} \wss{If ESPresSo scored high (I think it did), we can use
+that to further promote the value of documentation.  Also a threat to validity
+is that we have implicitly assumed documentation is valuable when measuring
+quality.}
 
@@
 kept consistent for new developers. \wss{Peter - this also sounds like
-discussion that is relevant for understandability.} \pmm{The points here sound like they could work for both.
-The definition of traceability notes that `This quality refers to keeping track of information as it changes forms or relates between artifacts'.
-So the point `New features are always added to the documentation' could be reworded to fit that part of the definition of traceability.
-As noted above, this is similar to the point about some packages having including detailed change logs.
-Same with `The developers use documentation to stay up to date on the status of the software package, and to help expand features'.
-I agree that at least some of this paragraph could be in understandability, especially the first and second sentences.}
+discussion that is relevant for understandability.} \pmm{The points here sound
+like they could work for both. The definition of traceability notes that `This
+quality refers to keeping track of information as it changes forms or relates
+between artifacts'. So the point `New features are always added to the
+documentation' could be reworded to fit that part of the definition of
+traceability. As noted above, this is similar to the point about some packages
+having including detailed change logs. Same with `The developers use
+documentation to stay up to date on the status of the software package, and to
+help expand features'. I agree that at least some of this paragraph could be in
+understandability, especially the first and second sentences.}
 
@@
 wonder if we should drop Traceability as a quality that we assessed?  There
-doesn't seem to be much content specific to traceability.}  \pmm{Based on the comments I just added above, I think we can keep at least a small part of what is here in traceability, rewording parts to highlight the importance of change logs.
-Some of the content in this section should indeed be moved to understandability.} The developer of OpenLB noted the use of documentation generators like Doxygen.
-It would be advisable for more projects to use such automatic document generation tools, since some projects do not do this.
+doesn't seem to be much content specific to traceability.}  \pmm{Based on the
+comments I just added above, I think we can keep at least a small part of what
+is here in traceability, rewording parts to highlight the importance of change
+logs. Some of the content in this section should indeed be moved to
+understandability.} The developer of OpenLB noted the use of documentation
+generators like Doxygen. It would be advisable for more projects to use such
+automatic document generation tools, since some projects do not do this.
 \wss{We can refer back to this for the recommendations.}
@@
 ~\ref{prinprocmeth}, development pain points are explored in Section
-~\ref{painpoints}, and our recommendations for improving software qualities are
-presented in Section~\ref{qualityrecommentations}. A comparison between our
-designations of the software packages to community rankings is made in Section
-~\ref{comparison}. Finally, threats to the validity of this assessment are noted
-in Section~\ref{threats}.
+~\ref{painpoints}, and summary of current (Section~\ref{currentPractices}), best
+(Section~\ref{bestPractices}) and future practices
+(Section~\ref{futurePractices}). Finally, threats to the validity of this
+assessment are noted in Section~\ref{threats}.
 
@@
 \end{tabular}
-\caption{Artifacts Present} \label{artifactspresent}
+\caption{Artifacts Present in Surveyed Packages, Classified by Frequency}
+\label{artifactspresent}
 \end{center}
@@
 The top four AHP ranked packages have most of the less common artifacts. At the
-time of data collection, only one (Palabos) of the four packages did not have a
+time of data collection, only one of the four packages (Palabos) did not have a
 user manual or guide, but there was a broken link on the package website
@@
 verification and validation notes, and a video guide of the software.
+
+\wss{Weiss mentions pain points related to design documentation.  Carver (2021) 
+also mentions design documentation in the literature review.}
 
@@
 	\end{tabular}
-	\caption{Observed development tools, dependencies and project management tools} \label{tbl_tools}
+	\caption{Observed development tools, dependencies and project management
+	tools} \label{tbl_tools}
 	\end{center}
@@
 Table~\ref{tbl_tools}, debuggers were also likely used.  Only three (ESPResSo,
-Ludwig and Musubi) packages mentioned continuous integration tools, like Travis CI. Code
-editors and compilers were explicitly noted to have been used by several
-packages, and were likely used by all of them. One of the packages (Ludwig)
-explicitly noted the use of proprietary unit testing code written in C.
+Ludwig and Musubi) packages mentioned continuous integration tools, like Travis
+CI. Code editors and compilers were explicitly noted to have been used by
+several packages, and were likely used by all of them. One of the packages
+(Ludwig) explicitly noted the use of proprietary unit testing code written in C.
 Likewise, the use of proprietary code for verifying the correctness of output
@@
 know each other from other projects. External contributions are rare despite
-many of the projects accepting them. Aside from on-boarding new developers, time
-constraints could be mitigated by increasing developer efficiency, which could
-be addressed in several ways, including by improving the quality of
-documentation, or incorporating automatic code generation.
+many of the projects accepting them. \wss{Other projects cite the same point
+point - Wiese paper}  Aside from on-boarding new developers, time constraints
+could be mitigated by increasing developer efficiency, which could be addressed
+in several ways, including by improving the quality of documentation, or
+incorporating automatic code generation. \wss{Potential ways to address the pain
+points might go into a separate section?  Maybe propose the strategy, like code
+generation, and then list the pain points that are addressed?}
 
@@
 Many of the team members on their project are domain experts and there can be a
-steep learning curve before these team members contribute good quality source
-code. It was further noted that this has been somewhat addressed, as the code
-has been re-written to best ensure ease of future contributions.
+steep learning curve before these members contribute good quality source code.
+This problem has been mitigated by re-writing the code to more easily enable
+future contributions.  On a related point, improving the software engineering
+education or experience of developers is also an idea that was brought up by
+several developers.  \wss{The Wiese paper and others talk about the background
+of developers.}
 
@@
 
-The same developer also noted that there is a lack of incentive and funding in
-academia for developing widely used scientific software. The importance of
-funding for scientific software has been discussed in
-\cite{gewaltig2012quality}, which notes that ``software tools are developed and
-maintained only for as long as there is explicit or implicit funding''. The
-developer further commented that there are no journals that publish such
-scientific software source code. However, there are ways to get such source code
-cited. Work has been done to address this in \citep{smith2016software}, which
-presents a set of software citation principles and discusses ``how they could be
-used to implement software citation in the scholarly community''
-\citep{katz2019software}.
+The TCLB developer noted a lack of incentive and funding in academia for
+developing widely used scientific software. The importance of funding for
+scientific software has been discussed in \cite{gewaltig2012quality}, which
+notes that ``software tools are developed and maintained only for as long as
+there is explicit or implicit funding''. The developer further commented that
+there are no journals that publish such scientific software source code.
+However, there are ways to get such source code cited. Work has been done to
+address this in \citep{smith2016software}, which presents a set of software
+citation principles and discusses ``how they could be used to implement software
+citation in the scholarly community'' \citep{katz2019software}.  \wss{other
+papers also mention this pain point.  Solutions proposed by Katz, journal of
+open source etc.}
 
@@
 \citep{gewaltig2012quality}. As software becomes user-ready, it can become
-commercialized and closed-source.
+commercialized and closed-source.  \wss{remove this paragraph?}
+
+\wss{Remove this section?  The external support organizations will likely come
+up as the mitigation for some of the other pain points.}
 
@@
 their source code, specifically helping with reusing code to run on both the CPU
-and GPU.
+and GPU. \wss{The point about macros isn't really related to parallelization or
+continuous integration.}
 
@@
 the testing problems, and to run tests in small batches with few iterations.
+\wss{The Weiss paper alludes this this as well.  I think the Kelly and Sanders
+paper likely talks about testing problems and correctness.}
 
@@
 traceability, and automatic code generation are suggested when designing
-software for change, and would help with these modifications. So far this
-problem of on-boarding new users has been addressed by updating the
-documentation to better inform users of the underlying LBM theory and package
-requirements. Similar issues with LBM parameters were noted by another
-developer. Updating the user interface to better explain theoretical principles,
-as well as test user input for compatibility, was the implemented solution. As
-noted above, sometimes frequently asked questions on the underlying theory and
-on how to use the software are answered in the documentation.
-
-The interviewed developer of ESPResSo commented that parts of their package's
-source code had been refactored to Python to help address usability issues.
-Python was perceived as a much more usable language, and it would be easy for
-future users and developers to learn and understand the source code.
+software for change, and would help with these modifications. \wss{Should decide
+if want to introduce ideas for improving within the presentation of the pain
+point, or separately?}  So far this problem of on-boarding new users has been
+addressed by updating the documentation to better inform users of the underlying
+LBM theory and package requirements. Similar issues with LBM parameters were
+noted by another developer. Updating the user interface to better explain
+theoretical principles, as well as test user input for compatibility, was the
+implemented solution. As noted above, sometimes frequently asked questions on
+the underlying theory and on how to use the software are answered in the
+documentation.  The interviewed developer of ESPResSo commented that parts of
+their package's source code had been refactored to Python to help address
+usability issues. Python was perceived as a much more usable language, and it
+would be easy for future users and developers to learn and understand the source
+code.
 
@@
 and reusability when trying to extend the software, and the code would need to
-be refactored.
-
-As noted above, difficulties with ensuring future correctness could also arise.
-As new methods and functionality is added into the software, new test cases and
-test data will need to be developed.
+be refactored.  As noted above, difficulties with ensuring future correctness
+could also arise. As new methods and functionality is added into the software,
+new test cases and test data will need to be developed.  \wss{Weiss mentions
+mainainability related issues under pain points.}
 
@@
 underlying theory are answered in the documentation. OpenLB has an artifact for
-such questions.
+such questions.  \wss{Weise mentions documentation.}
 
@@
 
-\subsection{What is the current software quality for the projects?}
-\label{qualityrecommentations}
-
-This subsection answers the research question: For research software developers,
-what specific actions are taken to address software qualities?
+\subsection{Current Practices} \label{currentPractices}
 
@@
 lists of what should have been done in the past, or what should be done now;
-they are just suggestions for consideration in the future.
+they are just suggestions for consideration in the future.  \wss{Rework this
+section to highlight what the top projects are doing.  Point to checklists for
+other advice?}
 
@@
 	installation. This was done by several top ranked packages, including Ludwig
-	and LUMA
+	and LUMA.
 	\item Provide a way to validate the installation. This can be done using a
@@
 	behind well defined interfaces. This is suggested in SCS literature, and in
-	developer comments.
+	developer comments.  \wss{one of Wilson's best practices.}
 	\item Use continuous integration tools (Bamboo, Jenkins, and Travis CI) and
 	processes during development. This was done by top ranked packages ESPResSo
-	and Ludwig.
+	and Ludwig.  \wss{Common in LBM - more common than in other domains}
 \end{itemize}
@@
 	documentation that may be required. This was done by many packages,
-	including waLBerla, Palabos, MechSys, LUMA, and pyLBM.
+	including waLBerla, Palabos, MechSys, LUMA, and pyLBM.  \wss{Comment from
+	ASCR attendee that success with open source requires good tutorials and
+	documentation - is this published anywhere?}
 \end{itemize}
@@
 	\item Eliminate code duplication. This is suggested as good software
-	development practice. Duplicate code is difficult to maintain.
+	development practice. Duplicate code is difficult to maintain. \wss{No
+	support for this one or citation, either find or remove it.}
 	\item If possible, consider using popular user-friendly software languages
@@
 	behind well defined interfaces. This is suggested in SCS literature, and in
-	developer comments.
-	\item If possible, consider using popular user-friendly software languages
-	like Python. Especially consider this for parts of the source code that is
-	likely to be modified or reviewed by users. ESPResSo and Sailfish use Python
-	to address several software qualities, including modifiability.
+	developer comments.  \wss{add citation - Wilson mentions - this points seems
+	repetitive}
 	\item Consider future source code modifiability as early as the design stage
 	of development. This is suggested as good software development practice.
+	\wss{citation, support, or remove.}
 	\item Consider flexibility of data structures and data storage in the design
@@
 	behind well defined interfaces. This is suggested in SCS literature, and in
-	developer comments.
+	developer comments.  \wss{repeated from elsewhere - there should be an
+	easier way to do this - give the advice and then link it to the qualities,
+	instead of this way?}
 	\item Document module interfaces in design and developer documentation. This
-	is suggested as good software development practice.
+	is suggested as good software development practice. \wss{citation or support}
 	\item Provide API documentation, if applicable. Only one (ESPResSo) of the
@@
 	behind well defined interfaces. This is suggested in SCS literature, and in
-	developer comments.
+	developer comments. \wss{repetitive}
 	\item Provide a user manual that identifies all technical and problem
@@
 	resolve some usability issues for OpenLB.
+	\item \wss{This quality is basically going through the list of things that
+	were measured by the template.}
 \end{itemize}
@@
 	made. This is suggested as good software development practice.
+	\wss{support, cite or remove}
 	\item Use automatic document generation tools (Doxygen, Drasil, Sphinx) to
 	limit the time spent on updating documentation. This was done by all of the
-	top five ranked packages.
+	top five ranked packages. \wss{repetitive}
 	\item Provide a developer's guide to help orient developers to the
 	artifacts, source code, and system architecture so that they can better
-	document changes. Top ranked package OpenLB has a developer guide. 
+	document changes. Top ranked package OpenLB has a developer guide. \wss{How
+	is this related to traceability?}
 \end{itemize}
@@
 	document generation tools (Doxygen, Drasil, Sphinx). This was done by all of
-	the top five ranked packages. 
+	the top five ranked packages. \wss{repetitive}
 \end{itemize}
@@
 	is made. This is suggested as good software development practice.
+	\wss{cite, support, or remove}
 \end{itemize}
@@
 
-\subsection{How does software designated as high quality by this methodology
-compare with top-rated software by the community?} \label{comparison}
-
-This subsection answers the research question: How does software designated as
-high quality by this methodology compare with top rated software by the
-community?
-
-This comparison helps evaluate the methodology that was used, and assess the
-validity of the findings. Threats to the validity of this exercise are noted in
-Section~\ref{threats}. The software package designations of this report are
-compared to package repository ranking metrics in Section~\ref{repmetrics}, and
-to software recommended by a domain expert in Section
-~\ref{domainexpertrecommend}.
-
-\subsubsection{Repository Ranking Metrics} \label{repmetrics}
-
-Table~\ref{repometrics} presents our LBM software package rankings along with
-the repository ranking metrics of each software package.
-
-Nine packages do not use GitHub, so they do not have a measure of repository
-stars. Looking at the repository stars of the other 15 packages, we can observe
-a slight pattern where packages that have been highly ranked by our assessment
-do have more stars than lower ranked packages. The best ranked package
-(ESPResSo) has the second most number of stars. The ninth ranked package
-(Sailfish) has the most number of stars. Packages designated as lower quality
-often do not use GitHub or GitLab, or have few stars. HemeLB only has 12 stars,
-but its location recently moved from a different GitHub repository. Our
-assessment of this package might not be accurate. The number of stars does not
-necessarily represent the perceptions of the community, but for lack of an
-alternative measure we will use it this way.
-
-The repository watches column contains even less data to compare since two of
-the packages (Palabos, waLBerla) use GitLab, which does not include this metric.
-The pattern that can be observed with this metric is very similar to that of the
-stars metric. The ninth ranked package (Sailfish) has the most number of
-watches. The best ranked package (ESPResSo) has the second most number of
-watches.
-
-Besides missing data, another threat to the validity of this comparison is the
-varying ages of the repositories. Older packages have been able to accumulate
-stars and watches for longer than newer packages. The true quality of new
-packages may not be reflected in their stars and watches. \\
-
-\begin{table}
-\begin{center}
-		\begin{tabular}{ p{3.5cm}p{2.5cm}p{2cm}p{2cm}p{2cm} }
-			\toprule
-			Name & Our Ranking & Repository Stars & Repository Star Rank & Repository Watches \\
-			\midrule
-			ESPResSo &1& 145 &2& 19\\
-			Ludwig&2& 27 &8& 6\\
-			Palabos &3& 34 &6& GitLab\\
-			OpenLB &4& No Git &No Git & No Git\\
-			LUMA&5& 33 &7& 12\\
-			pyLBM &6& 95&3& 10\\
-			DL\_MESO (LBE) & 7 & No Git &No Git & No Git \\
-			Musubi & 8 & No Git &No Git & No Git \\
-			Sailfish &9& 186 &1& 41\\
-			waLBerla & 10 & 20 &9& GitLab\\
-			laboetie &11& 4 &13& 5\\
-			TCLB &12& 95 &3& 16\\
-			MechSys &13& No Git &No Git & No Git\\
-			lettuce &14& 48 &4& 5\\
-			ESPResSo++ &15& 35 &5& 12\\
-			MP-LABS &16& 12 &11& 2\\			
-			SunlightLB & 17& No Git &No Git & No Git\\
-			LB3D &18& No Git &No Git & No Git\\			
-			LIMBES &19& No Git &No Git & No Git\\
-			LB2D-Prime &20& No Git &No Git & No Git\\		
-			HemeLB &21& 12 &11& 12\\
-			lbmpy&22&  11 &12&  2\\	
-			LB3D-Prime &23& No Git &No Git & No Git\\	
-			LatBo.jl &24& 17 &10& 8\\			
-			\bottomrule
-		\end{tabular}
-		\caption{Repository Ranking Metrics} \label{repometrics}
-\end{center}
-\end{table}
-
-\subsubsection{Domain Expert Recommended Software} \label{domainexpertrecommend}
-
-Table~\ref{rankingcomparison} compares the top 10 ranked LBM software packages
-with a LBM package ranking made by a domain expert. Five (Palabos, OpenLB, LUMA,
-Sailfish, waLBerla) of the top 10 ranked packages in this assessment are also
-found in the domain expert's top 10 ranking. Interestingly, the first three of
-these five packages are listed in the same order on both lists. Looking at the
-remaining packages, the top two ranked packages in this state of the practice
-assessment (ESPResSo, Ludwig) are not on the domain expert's list. Perhaps the
-intended applications of these packages, complex fluids for Ludwig and soft
-matter research for ESPResSo, do not align with the research interests of the
-domain expert, and this is why they did not make it onto their list. This
-reasoning may also apply to eighth ranked Musubi. Moving down the list, pyLBM
-and DL\_MESO(LBE) did not make the domain expert's top 10 list but were
-mentioned by the domain expert. Looking at the remainder of the domain expert's
-list, ASL is a general purpose tool for solving partial differential equations
-and may have fallen out of scope of the authoritative lists that were used to
-identify the initial LBM software list. The domain expert's fifth ranked package
-(ch4-project) was on the initial software list of this assessment, but was
-removed due to a lack of documentation. Open FSI is a new project that uses
-Palabos. It was made public the year before data was collected for this state of
-the practice exercise. It may not have been on authoritative lists due to its
-young age. Similarly, LIFE was also made public recently. Finally, the domain
-expert's 10th ranked package (HemeLB) was ranked 21st in this state of the
-practice assessment.
-
-\begin{table}
-\begin{center}
-	\begin{tabular}{ p{2cm}p{4.5cm}p{3.5cm}}
-		\toprule
-		Rank & This Assessment & Domain Expert\\
-		\midrule
-		1 & ESPResSo (N/A) & Palabos\\
-		2 & Ludwig (N/A) & OpenLB\\
-		3 & Palabos (1) & LUMA\\
-		4 & OpenLB (2) & ASL\\
-		5 & LUMA (3) & ch4-project\\
-		6 & pyLBM (16) & Open FSI\\
-		7 & DL\_MESO (LBE) (11) & WaLBerla\\
-		8 & Musubi (N/A) & LIFE\\
-		9 & Sailfish (9) & Sailfish\\
-		10 & WaLBerla (7) & HemeLB\\		
-		\bottomrule
-	\end{tabular}
-	\caption{Ranking Comparison} \label{rankingcomparison}
-\end{center}
-\end{table}
-
-\subsection{Threats To Validity} \label{threats}
+\subsection{Best Practices} \label{bestPractices}
+
+practices by the best measured.
+
+practices mentioned in literature, but not practiced (if any)
+
+\subsection{Future Practices} \label{futurePractices}
+
+practices for the future.
+
+\section{Threats To Validity} \label{threats}
 
@@
 practice assessment. These can be categorized into methodology and data
-collection issues.
-
-The goal of this assessment isn't to rank the software, but to use the ranking
-exercise as a means to understand the state of the practice of LBM software
-development.
+collection issues. The goal of this assessment isn't to rank the software, but
+to use the ranking exercise as a means to understand the state of the practice
+of LBM software development.
 
@@
 to this the measured data is not a true reflection of software package quality.
+
+\wss{There are more threats to validity.  Brainstorm.  Watch for them while 
+editing.  Look at Ao paper.}
 
@@
 quality attributes, and to reduce software quality failures.
+
+\wss{More on the lessons to learn from the top packages.  The comparison to
+common practice - is LBM ahead or behind?  More on best practices comparison.
+Future practices.}
 
@@
 manually duplicate knowledge into multiple artifacts \citep{zhao2018}.
+\wss{Yuzhi is not the best reference to use for the Drasil work.}
 
 \subsection{Future State Of The Practice Assessments} \label{futuresop}
-
-This section notes recommendations for future state of the practice assessments.
 
@@
 ranking, ensuring quality designations more accurate represent the true quality
-of the software packages.
+of the software packages.  \wss{clarify the distinction between usability and
+understandability.}
 
@@
 - read more of the existing survey studies, as highlighted by Carver paper
-- create issues for Peter
 - data on Mendeley
@@
 each instance, add page numbers).
+
+- add links or citations to all of the packages (could add them to the summary table)

### 2022-01-05 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Further divided practices to have a separate list of those that are recommended, but not mentioned by LBM developers or present in their projects

@@
 
-\subsubsection{Installability}
-
 \begin{itemize}
-	\item Include OS compatibility, including specific OS versions. This was
-	done by several top ranked packages, including ESPResSo and LUMA.
-	\item Provide complete installation instructions. This was done by all of
-	the top five ranked packages.
-	\item Installation instructions should be written as if the user does not
-	have any dependencies installed and is installing on a clean OS. This was
-	done by several top ranked packages, including ESPResSo, OpenLB, and
-	Palabos. This can be tested on a clean environment using a virtual machine.
-	\item List all dependencies in the installation instructions, dependency
-	versions, and how to install them. This was done by several top ranked
-	packages, including ESPResSo, LUMA, OpenLB, and Palabos.
-	\item If possible, automate the installation of dependencies. Use tools such
-	as Make. This was done by all of the top five ranked packages.
-	\item Automate the installation process as much as possible. Use tools such
-	as Make. This was done by all of the top five ranked packages.
-	\item The installation instruction should only be in one location. This was
-	done by several top ranked packages, including ESPResSo, OpenLB, and
-	Palabos.
-	\item Include descriptive error messages for errors encountered during
-	installation. This was done by several top ranked packages, including Ludwig
-	and LUMA.
-	\item Provide a way to validate the installation. This can be done using a
-	custom script, or a test case that specifies expected output. This was done
-	by several top ranked packages, including OpenLB, Ludwig, LUMA, and Palabos.
-	\item Provide instructions for uninstalling the software. These were
-	included with pyLBM.
-\end{itemize}
-
-\subsubsection{Surface Correctness and Verifiability}
-
-\begin{itemize}
-	\item Use a requirements specification document. This is suggested in SCS
-	literature, and several of the top ranked packages had such a document or
-	reference to theory manuals. A potential template is presented in
-	\citep{smith2005new}. 
-	\item Make public (on package website or GitHub) the requirements
-	specification document, or explicitly reference the domain theory that the
-	software is designed from.  This was done by several top ranked packages,
-	including ESPResSo, Ludwig, LUMA, and OpenLB.
-	\item Ensure the above information is easy to find. Consider adding it to
-	the user manual. The information is easy to find in most top ranked
-	packages.
-	\item Development teams should include both domain experts and experienced
-	software developers. This suggestion is based on developer comments.
-	\item Use and make public (on package website or GitHub) detailed
-	documentation. Consider using automatic document generation tools like
-	Doxygen, Drasil, or Sphinx. This was done by all of the top five ranked
-	packages.
-	\item Provide detailed tutorials that include expected output, like the
-	\href{https://www.walberla.net/doxygen/index.html}{waLBerla tutorials}.
-	\item Use unit tests during development and make them public (on package
-	website or GitHub). This was available for Ludwig.
-	\item Modularize the source code, separate components, hide information
-	behind well defined interfaces. This is suggested in SCS literature, and in
-	developer comments.  \wss{one of Wilson's best practices.}
-	\item Use continuous integration tools (Bamboo, Jenkins, and Travis CI) and
-	processes during development. This was done by top ranked packages ESPResSo
-	and Ludwig.  \wss{Common in LBM - more common than in other domains}
-\end{itemize}
-
-\subsubsection{Surface Reliability}
-
-\begin{itemize}
-	\item Include descriptive error messages where appropriate. This was done by
-	most packages that encountered a fault.
-	\item In case automatic installation of dependencies fails, the system
-	should indicate to the user what dependencies need to be installed manually.
-	This was done by many packages, including ESPResSo++, Ludwig, LUMA, pyLBM,
-	TCLB.
-	\item The packages should include detailed tutorials, including
-	dependencies, expected output, and any additional supplementary
+	\item Surface Reliability. Include descriptive error messages where
+	appropriate. This was done by most packages that encountered a fault.
+	\item Surface Reliability. In case automatic installation of dependencies
+	fails, the system should indicate to the user what dependencies need to be
+	installed manually. This was done by many packages, including ESPResSo++,
+	Ludwig, LUMA, pyLBM, TCLB.
+	\item Surface Reliability. The packages should include detailed tutorials,
+	including dependencies, expected output, and any additional supplementary
 	documentation that may be required. This was done by many packages,
@@
 	documentation - is this published anywhere?}
+	\item Surface Robustness. The system must provide descriptive error messages
+	when it encounters unexpected input, including incorrect data types, empty
+	input, and missing files or links. Nineteen of the software packages behaved
+	reasonably when tested with unexpected input.
+	\item Surface Performance. Integrate parallelization tools and techniques to
+	reduce processing time. Consider GPU processing, CUDA, and MPI. Eighteen
+	software packages mentioned parallelization in their artifacts.
+	\item Surface Performance. The user should be able to choose to process
+	their model on either the CPU or GPU. ESPResSo, lbmpy, lettuce, pyLBM, and
+	TCLB allow for either CPU or GPU processing.
+	\item Surface Usability. Include user hardware and software requirements
+	documentation. Hardware requirements are rarely listed in the software
+	packages that were assessed. Only those offering GPU processing (ESPResSo,
+	lbmpy, lettuce, pyLBM, Sailfish, TCLB) mentioned any hardware requirements.
+	On the other hand, some sort of software requirements were available for all
+	packages, even if only some dependencies or a compatible operating system
+	were mentioned.
+	\item Surface Usability. Provide a user tutorial that indicates the expected
+	output. This was done by many packages, including waLBerla, Palabos,
+	MechSys, LUMA, pyLBM.
+	\item Surface Usability. Provide a detailed user manual. It should identify
+	elements of user interfaces, and identify all requirements to model a
+	system. Thirteen packages have a user manual. ESPResSo has a well detailed
+	manual.
+	\item Maintainability. Ideally include most of the common and less common
+	artifacts listed in Section~\ref{artifacts}.
+	\item Surface Understandability. Use consistent, distinctive, and meaningful
+	code identifiers. Twenty of the 24 software packages use such code
+	identifiers.
+	\item Surface Understandability. Use symbolic constants. Thirteen of the 24
+	software packages use such constants.
+	\item Surface Understandability. Identify algorithms that are used in source
+	code comments. Document them in the artifacts. Cite relevant external
+	sources if needed. Domain algorithms are noted in the source code of 11
+	packages.
+	\item Surface Understandability. Provide a user manual that identifies all
+	technical and problem modeling requirements. Thirteen software packages have
+	a user manual.
+	\item Unambiguity. Provide a detailed user manual. Thirteen packages have a
+	user manual. ESPResSo has a well detailed manual.
+
 \end{itemize}
 
-\subsubsection{Surface Robustness}
+\subsection{Best Practices} \label{bestPractices}
+
+practices by the best measured.
 
 \begin{itemize}
-	\item The system must provide descriptive error messages when it encounters
-	unexpected input, including incorrect data types, empty input, and missing
-	files or links. Nineteen of the software packages behaved reasonably when
-	tested with unexpected input.
-\end{itemize}
-
-\subsubsection{Surface Performance}
-
-\begin{itemize}
-	\item Integrate parallelization tools and techniques to reduce processing
-	time. Consider GPU processing, CUDA, and MPI. Eighteen software packages
-	mentioned parallelization in their artifacts.
-	\item The user should be able to choose to process their model on either the
-	CPU or GPU. ESPResSo, lbmpy, lettuce, pyLBM, and TCLB allow for either CPU
-	or GPU processing.
-\end{itemize}
-
-\subsubsection{Surface Usability}
-
-\begin{itemize}
-	\item Include user hardware and software requirements documentation.
-	Hardware requirements are rarely listed in the software packages that were
-	assessed. Only those offering GPU processing (ESPResSo, lbmpy, lettuce,
-	pyLBM, Sailfish, TCLB) mentioned any hardware requirements. On the other
-	hand, some sort of software requirements were available for all packages,
-	even if only some dependencies or a compatible operating system were
-	mentioned.
-	\item Provide a user tutorial that indicates the expected output. This was
-	done by many packages, including waLBerla, Palabos, MechSys, LUMA, pyLBM.
-	\item Provide a detailed user manual. It should identify elements of user
-	interfaces, and identify all requirements to model a system. Thirteen
-	packages have a user manual. ESPResSo has a well detailed manual.
-	\item State appropriate fluid dynamics problems that the software is
-	designed to model in the documentation, and explicitly state the limits of
-	the software. This is done in the ESPResSo user guide.
-	\item Provide documentation that details the background theory information,
-	or provide a reference to such information. This was done by several top
-	ranked packages, including ESPResSo, Ludwig, LUMA, and OpenLB.
-	\item Identify expected user characteristics. LIMBES, Ludwig, laboetie, and
-	Palabos did this. The importance of specifying user characteristics is
-	discussed in \citep{smith2007requirements}.
-	\item Keep all documentation in one location. This was done by top ranked
+	\item Installability. Include OS compatibility, including specific OS
+	versions. This was done by several top ranked packages, including ESPResSo
+	and LUMA.
+	\item Installability. Provide complete installation instructions. This was
+	done by all of the top five ranked packages.
+	\item Installability. Installation instructions should be written as if the
+	user does not have any dependencies installed and is installing on a clean
+	OS. This was done by several top ranked packages, including ESPResSo,
+	OpenLB, and Palabos. This can be tested on a clean environment using a
+	virtual machine.
+	\item Installability. List all dependencies in the installation
+	instructions, dependency versions, and how to install them. This was done by
+	several top ranked packages, including ESPResSo, LUMA, OpenLB, and Palabos.
+	\item Installability. If possible, automate the installation of
+	dependencies. Use tools such as Make. This was done by all of the top five
+	ranked packages.
+	\item Installability. Automate the installation process as much as possible.
+	Use tools such as Make. This was done by all of the top five ranked
 	packages.
-	\item Do not duplicate artifact information. This is suggested as good
-	software development practice. Duplicate information is difficult to
-	maintain. 
-	\item Maintain a user support model (Git, email, forum, FAQ). This was done
-	by all of the top five ranked packages.
-	\item If possible, consider using popular user-friendly software languages
-	like Python. Especially consider this for parts of the source code that is
-	likely to be modified or reviewed by users. ESPResSo and Sailfish use Python
-	to shorten development time and improve usability.
-\end{itemize}
-
-\subsubsection{Maintainability}
-
-\begin{itemize}
-	\item Keep artifacts updated. This is suggested as good software maintenance
-	practice.
-	\item Ideally include most of the common and less common artifacts listed in
-	Section~\ref{artifacts}.
-	\item Include version numbers and release notes for all major source code
-	and artifact releases. The top five ranked software packages include release
-	notes.
-	\item Have a defined process for accepting contributions, and make public
-	documentation for making contributions to the project. The top five ranked
-	software packages include information on how to contribute. 
-	\item Use an issue tracker (Git, email, SourceForge, other) to manage bugs
-	and changes. Issues should be regularly reviewed and closed. All but three
-	(laboetie, lettuce, Sailfish) of the software packages that use Git have
-	most of their issues closed.
-	\item Use a version control system (GitHub, CVS). Four (ESPREsSo, Ludwig,
-	LUMA, Palabos) of the top five ranked packages use a version control system.
-	\item Source code needs to be well commented. Typically, more than 10
-	percent of LBM package source code is comments, as presented in
+	\item Installability. The installation instruction should only be in one
+	location. This was done by several top ranked packages, including ESPResSo,
+	OpenLB, and Palabos.
+	\item Installability. Include descriptive error messages for errors
+	encountered during installation. This was done by several top ranked
+	packages, including Ludwig and LUMA.
+	\item Installability. Provide a way to validate the installation. This can
+	be done using a custom script, or a test case that specifies expected
+	output. This was done by several top ranked packages, including OpenLB,
+	Ludwig, LUMA, and Palabos.
+	\item Installability. Provide instructions for uninstalling the software.
+	These were included with pyLBM.
+	\item Surface Correctness and Verifiability. Make public (on package website
+	or GitHub) the requirements specification document, or explicitly reference
+	the domain theory that the software is designed from.  This was done by
+	several top ranked packages, including ESPResSo, Ludwig, LUMA, and OpenLB.
+	\item Surface Correctness and Verifiability. Ensure the above information is
+	easy to find. Consider adding it to the user manual. The information is easy
+	to find in most top ranked packages.
+	\item Surface Correctness and Verifiability. Development teams should
+	include both domain experts and experienced software developers. This
+	suggestion is based on developer comments.
+	\item Surface Correctness and Verifiability. Traceability. Visibility and Transparency. Use and make public (on package
+	website or GitHub) detailed documentation. Consider using automatic document
+	generation tools like Doxygen, Drasil, or Sphinx. This was done by all of
+	the top five ranked packages.
+	\item Surface Correctness and Verifiability. Provide detailed tutorials that
+	include expected output, like the
+	\href{https://www.walberla.net/doxygen/index.html}{waLBerla tutorials}.
+	\item Surface Correctness and Verifiability. Use unit tests during
+	development and make them public (on package website or GitHub). This was
+	available for Ludwig.
+	\item Surface Correctness and Verifiability. Modularize the source code,
+	separate components, hide information behind well defined interfaces. This
+	is suggested in SCS literature, and in developer comments.  \wss{one of
+	Wilson's best practices.}
+	\item Surface Understandability. Maintainability.  Modifiability.
+	Reusability. Modularize the source code, separate components, hide
+	information behind well defined interfaces. This is suggested in SCS
+	literature, and in developer comments. \wss{add citation - Wilson mentions -
+	this points seems repetitive}
+	\item Surface Correctness and Verifiability. Use continuous integration
+	tools (Bamboo, Jenkins, and Travis CI) and processes during development.
+	This was done by top ranked packages ESPResSo and Ludwig.  \wss{Common in
+	LBM - more common than in other domains}
+	\item Surface Usability. State appropriate fluid dynamics problems that the
+	software is designed to model in the documentation, and explicitly state the
+	limits of the software. This is done in the ESPResSo user guide.
+	\item Surface Usability. Provide documentation that details the background
+	theory information, or provide a reference to such information. This was
+	done by several top ranked packages, including ESPResSo, Ludwig, LUMA, and
+	OpenLB.
+	\item Surface Usability. Identify expected user characteristics. LIMBES,
+	Ludwig, laboetie, and Palabos did this. The importance of specifying user
+	characteristics is discussed in \citep{smith2007requirements}.
+	\item Surface Usability. Keep all documentation in one location. This was
+	done by top ranked packages.
+	\item Surface Usability. Maintain a user support model (Git, email, forum,
+	FAQ). This was done by all of the top five ranked packages.
+	\item Surface Usability. If possible, consider using popular user-friendly
+	software languages like Python. Especially consider this for parts of the
+	source code that is likely to be modified or reviewed by users. ESPResSo and
+	Sailfish use Python to shorten development time and improve usability.
+	\item Maintainability. Include version numbers and release notes for all
+	major source code and artifact releases. The top five ranked software
+	packages include release notes.
+	\item Maintainability. Have a defined process for accepting contributions,
+	and make public documentation for making contributions to the project. The
+	top five ranked software packages include information on how to contribute. 
+	\item Maintainability. Use an issue tracker (Git, email, SourceForge, other)
+	to manage bugs and changes. Issues should be regularly reviewed and closed.
+	All but three (laboetie, lettuce, Sailfish) of the software packages that
+	use Git have most of their issues closed.
+	\item Maintainability. Use a version control system (GitHub, CVS). Four
+	(ESPREsSo, Ludwig, LUMA, Palabos) of the top five ranked packages use a
+	version control system.
+	\item Maintainability. Source code needs to be well commented. Typically,
+	more than 10 percent of LBM package source code is comments, as presented in
 	Table~\ref{gitrepodata}. Four of the top five overall ranked packages
@@
 	code as comments.
-	\item Modularize the source code, separate components, hide information
-	behind well defined interfaces. This is suggested in SCS literature, and in
-	developer comments.
-	\item Eliminate code duplication. This is suggested as good software
-	development practice. Duplicate code is difficult to maintain. \wss{No
-	support for this one or citation, either find or remove it.}
-	\item If possible, consider using popular user-friendly software languages
-	like Python. Especially consider this for parts of the source code that is
-	likely to be modified or reviewed by users. ESPResSo and Sailfish use Python
-	to address several software qualities, including maintainability.
-	\item Consider the recommended points for addressing traceability.
-\end{itemize}
-
-\subsubsection{Modifiability}
-
-\begin{itemize}
-	\item Modularize the source code, separate components, hide information
-	behind well defined interfaces. This is suggested in SCS literature, and in
-	developer comments.  \wss{add citation - Wilson mentions - this points seems
-	repetitive}
-	\item Consider future source code modifiability as early as the design stage
-	of development. This is suggested as good software development practice.
-	\wss{citation, support, or remove.}
-	\item Consider flexibility of data structures and data storage in the design
-	stage. The package pyLBM redeveloped data structures to ease future changes.
-\end{itemize}
-
-\subsubsection{Reusability}
-
-\begin{itemize}
-	\item Modularize the source code, separate components, hide information
-	behind well defined interfaces. This is suggested in SCS literature, and in
-	developer comments.  \wss{repeated from elsewhere - there should be an
-	easier way to do this - give the advice and then link it to the qualities,
-	instead of this way?}
-	\item Document module interfaces in design and developer documentation. This
-	is suggested as good software development practice. \wss{citation or support}
-	\item Provide API documentation, if applicable. Only one (ESPResSo) of the
-	top five ranked packages provided API documentation.
-\end{itemize}
-
-\subsubsection{Surface Understandability}
-
-\begin{itemize}
-	\item Adopt a coding standard and document it in artifacts, including some
-	examples. A coding standard needs to be part of continuous integration. Only
-	LUMA, HemeLB, and Musubi explicitly identify coding standards. 
-	\item Use consistent, distinctive, and meaningful code identifiers. Twenty
-	of the 24 software packages use such code identifiers.
-	\item Use symbolic constants. Thirteen of the 24 software packages use such
-	constants.
-	\item Identify algorithms that are used in source code comments. Document
-	them in the artifacts. Cite relevant external sources if needed. Domain
-	algorithms are noted in the source code of 11 packages.
-	\item Add meaningful comments. Indicate what is being done in each section
-	of source code. All of the top ranked packages have meaningful comments in
-	their source code.
-	\item Modularize the source code, separate components, hide information
-	behind well defined interfaces. This is suggested in SCS literature, and in
-	developer comments. \wss{repetitive}
-	\item Provide a user manual that identifies all technical and problem
-	modeling requirements. Thirteen software packages have a user manual.
-	\item State appropriate fluid dynamics problems that the software is
-	designed to model in the documentation, and explicitly state the limits of
-	the software. This is done in the ESPResSo user guide. 
-	\item Consider adding a FAQ section to the documentation. This helped
-	resolve some usability issues for OpenLB.
-	\item \wss{This quality is basically going through the list of things that
-	were measured by the template.}
-\end{itemize}
-
-\subsubsection{Traceability}
-
-\begin{itemize}
-	\item Update all relevant documentation when a change to the software is
-	made. This is suggested as good software development practice.
-	\wss{support, cite or remove}
-	\item Use automatic document generation tools (Doxygen, Drasil, Sphinx) to
-	limit the time spent on updating documentation. This was done by all of the
-	top five ranked packages. \wss{repetitive}
-	\item Provide a developer's guide to help orient developers to the
-	artifacts, source code, and system architecture so that they can better
+	\item Maintainability. If possible, consider using popular user-friendly
+	software languages like Python. Especially consider this for parts of the
+	source code that is likely to be modified or reviewed by users. ESPResSo and
+	Sailfish use Python to address several software qualities, including
+	maintainability.
+	\item Modifiability. Consider flexibility of data structures and data
+	storage in the design stage. The package pyLBM redeveloped data structures
+	to ease future changes.
+	\item Reusability. Provide API documentation, if applicable. Only one
+	(ESPResSo) of the top five ranked packages provided API documentation.
+	\item Surface Understandability. Adopt a coding standard and document it in
+	artifacts, including some examples. A coding standard needs to be part of
+	continuous integration. Only LUMA, HemeLB, and Musubi explicitly identify
+	coding standards. 
+	\item Surface Understandability. Add meaningful comments. Indicate what is
+	being done in each section of source code. All of the top ranked packages
+	have meaningful comments in their source code.
+	\item Surface Understandability. State appropriate fluid dynamics problems
+	that the software is designed to model in the documentation, and explicitly
+	state the limits of the software. This is done in the ESPResSo user guide. 
+	\item Surface Understandability. Consider adding a FAQ section to the
+	documentation. This helped resolve some usability issues for OpenLB.
+	\item Traceability. Provide a developer's guide to help orient developers to
+	the artifacts, source code, and system architecture so that they can better
 	document changes. Top ranked package OpenLB has a developer guide. \wss{How
 	is this related to traceability?}
+	\item Visibility and Transparency. Summarize the development process that is
+	used. Provide information on how new users can contribute. Identify the
+	development model by name (waterfall, agile, etc.), if appropriate.  Seven
+	of the software packages have some artifacts outlining the general
+	development process. Eleven packages have information on how to contribute. 
+	\item Visibility and Transparency. Identify the development environment.
+	Eight of the 24 software packages identify the development environment.
+	\item Visibility and Transparency. Include notes with all releases. Nine of
+	the packages include release notes.
+	\item Visibility and Transparency. Communicate all changes within the
+	development team. This suggestion is based on developer comments.
+	\item Visibility and Transparency. Use continuous integration processes and
+	tools (Bamboo, Jenkins, and Travis CI). This was done by top ranked packages
+	ESPResSo and Ludwig.
+	\item Visibility and Transparency. Consider peer review processes to assess
+	contributions and ensure the tracking of information. High ranked package
+	ESPResSo uses a peer review process for contributions. 
+	\item Reproducibility. Test output against automatically calculated or known
+	correct results. High ranked package ESPResSo does this.
+	\item Reproducibility. Automate the testing of output. Run tests after all
+	code changes. High ranked package ESPResSo does this.
+	\item Reproducibility. Consider peer review processes and task based
+	inspection to assess contributions. High ranked package ESPResSo does this.
+	\item Unambiguity. State appropriate fluid dynamics problems that the
+	software is designed to model in the documentation, and explicitly state the
+	limits of the software. This is done in the ESPResSo user guide.
+	\item Unambiguity. The documentation should either explain the underlying
+	CFD theories or provide a reference to appropriate resources. This was done
+	by several top ranked packages, including ESPResSo, Ludwig, LUMA, and
+	OpenLB.
+	\item Unambiguity. Consider asking users for feedback on the documentation.
+	The developer of pyLBM noted that such feedback would be appreciated.
+
 \end{itemize}
 
-\subsubsection{Visibility and Transparency}
+practices mentioned in literature, but not practiced (if any).  Start with the
+practices that aren't derived from developer interviews.
 
 \begin{itemize}
-	\item Summarize the development process that is used. Provide information on
-	how new users can contribute. Identify the development model by name
-	(waterfall, agile, etc.), if appropriate.  Seven of the software packages
-	have some artifacts outlining the general development process. Eleven
-	packages have information on how to contribute. 
-	\item Identify the development environment. Eight of the 24 software
-	packages identify the development environment.
-	\item Update all relevant documentation when a change to the software is
-	made. This is suggested as good software development practice.
-	\item Include notes with all releases. Nine of the packages include release notes.
-	\item Communicate all changes within the development team. This suggestion
-	is based on developer comments.
-	\item Use continuous integration processes and tools (Bamboo, Jenkins, and
-	Travis CI). This was done by top ranked packages ESPResSo and Ludwig.
-	\item Consider peer review processes to assess contributions and ensure the
-	tracking of information. High ranked package ESPResSo uses a peer review
-	process for contributions. 
-	\item Use project management tools, including change and version control
-	tools (GitHub, GitLab, CVS), collaboration tools (GitHub, GitLab), and
-	document generation tools (Doxygen, Drasil, Sphinx). This was done by all of
-	the top five ranked packages. \wss{repetitive}
+	\item Surface Correctness and Verifiability. Use a requirements
+	specification document. This is suggested in SCS literature, and several of
+	the top ranked packages had such a document or reference to theory manuals.
+	A potential template is presented in \citep{smith2005new}. 
+	\item Surface Usability. Do not duplicate artifact information. This is
+	suggested as good software development practice. Duplicate information is
+	difficult to maintain. 
+	\item Maintainability. Visibility and Transparency. Reproducibility. Keep artifacts updated.
+	This is suggested as good software maintenance practice.
+	\item Maintainability. Eliminate code duplication. This is suggested as good
+	software development practice. Duplicate code is difficult to maintain.
+	\wss{No support for this one or citation, either find or remove it.}
+	\item Modifiability. Consider future source code modifiability as early as
+	the design stage of development. This is suggested as good software
+	development practice. \wss{citation, support, or remove.}
+	\item Reusability. Document module interfaces in design and developer
+	documentation. This is suggested as good software development practice.
+	\wss{citation or support}
+	\item Traceability. Update all relevant documentation when a change to the
+	software is made. This is suggested as good software development practice.
+	\wss{support, cite or remove}
+	\item Unambiguity. Documentation must state all technical requirements and
+	software dependencies. A missing dependency was a frequent cause of fault
+	conditions during this assessment.
+	\item Unambiguity. Provide a table of symbols in the developer's guide that
+	maps to names used in the source code. This is suggested as good software
+	development practice.
+
 \end{itemize}
-
-\subsubsection{Reproducibility}
-
-\begin{itemize}
-	\item Test output against automatically calculated or known correct results.
-	High ranked package ESPResSo does this.
-	\item Automate the testing of output. Run tests after all code changes. High
-	ranked package ESPResSo does this.
-	\item Consider peer review processes and task based inspection to assess
-	contributions. High ranked package ESPResSo does this.
-	\item Update all relevant design documentation when a change to the software
-	is made. This is suggested as good software development practice.
-	\wss{cite, support, or remove}
-\end{itemize}
-
-\subsubsection{Unambiguity}
-
-\begin{itemize}
-	\item Documentation must state all technical requirements and software
-	dependencies. A missing dependency was a frequent cause of fault conditions
-	during this assessment.
-	\item Provide a detailed user manual. Thirteen packages have a user manual.
-	ESPResSo has a well detailed manual.
-	\item Provide a table of symbols in the developer's guide that maps to names
-	used in the source code. This is suggested as good software development
-	practice.
-	\item State appropriate fluid dynamics problems that the software is
-	designed to model in the documentation, and explicitly state the limits of
-	the software. This is done in the ESPResSo user guide.
-	\item The documentation should either explain the underlying CFD theories or
-	provide a reference to appropriate resources. This was done by several top
-	ranked packages, including ESPResSo, Ludwig, LUMA, and OpenLB.
-	\item Consider asking users for feedback on the documentation. The developer
-	of pyLBM noted that such feedback would be appreciated.
-\end{itemize}
-
-\subsection{Best Practices} \label{bestPractices}
-
-practices by the best measured.
-
-practices mentioned in literature, but not practiced (if any)
 

### 2022-01-07 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
sorted LBM paper Table 4 by percent code comments

@@
 			\midrule
-			DL\_MESO (LBE) & Not Git & 8.06& Alive\\
+			MP-LABS & 100.00 & 26.67 & Dead\\
+			Musubi & Not Git & 24.19 & Alive\\
+			waLBerla & 72.90 & 22.62 & Alive\\
+			OpenLB & Not Git & 22.43 & Alive\\
 			ESPResSo & 89.26 & 21.78 & Alive\\
+			Ludwig& 60.00 & 20.70 & Alive\\
+			Palabos & 89.47 & 17.76 & Alive\\
+			SunlightLB & Not Git & 17.67 & Dead\\
+			LIMBES & Not Git & 17.39 & Dead\\
 			ESPResSo++ & 66.28 & 17.10 & Alive\\
 			HemeLB & No Issues & 16.68 & Dead\\
+			pyLBM & 66.67& 16.12 & Alive\\
+			MechSys & Not Git & 15.11 & Alive\\
+			LB3D-Prime & Not Git & 14.34 & Dead\\
+			LB3D & Not Git & 13.76 & Dead\\
+			LB2D-Prime & Not Git & 13.61 & Dead\\
+			Sailfish & 22.22 & 9.26 & Alive\\
+			lettuce & 33.33 & 8.19 & Alive\\
+			DL\_MESO (LBE) & Not Git & 8.06& Alive\\
+			TCLB & 60.32 & 6.02 & Alive\\
 			laboetie & 18.75 & 2.47 & Dead\\	
+			lbmpy& 58.33  & 2.03 & Alive\\	
 			LatBo.jl & 93.33 & 0.40 & Dead\\
-			LB2D-Prime & Not Git & 13.61 & Dead\\
-			LB3D & Not Git & 13.76 & Dead\\
-			LB3D-Prime & Not Git & 14.34 & Dead\\
-			lbmpy& 58.33  & 2.03 & Alive\\
-			lettuce & 33.33 & 8.19 & Alive\\
-			LIMBES & Not Git & 17.39 & Dead\\
-			Ludwig& 60.00 & 20.70 & Alive\\
 			LUMA& 85.71   & 0.20 & Alive\\
-			MechSys & Not Git & 15.11 & Alive\\
-			MP-LABS & 100.00 & 26.67 & Dead\\
-			Musubi & Not Git & 24.19 & Alive\\
-			OpenLB & Not Git & 22.43 & Alive\\
-			Palabos & 89.47 & 17.76 & Alive\\
-			pyLBM & 66.67& 16.12 & Alive\\
-			Sailfish & 22.22 & 9.26 & Alive\\
-			SunlightLB & Not Git & 17.67 & Dead\\
-			TCLB & 60.32 & 6.02 & Alive\\
-			waLBerla & 72.90 & 22.62 & Alive\\
 			\bottomrule

### 2022-01-07 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Further edits related to common practices

--- 2022-01-05 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
@@
     review process, communicate development and contribution information, and
-    use continuous integration and project management tools.
+    use continuous integration and project management tools.  \wss{Change the
+    abstract to discuss how LBM compares to common practices and best practices.
+    Highlight some future practices.}
 
@@
 GitHub and swMATH, as well as through articles found in scholarly journals and
-databases.  The Domain Expert (Section~\ref{domainexpertrecommend}) was also
+databases.  The Domain Expert (Section~\ref{sec_vet_software_list}) was also
 engaged in selecting the candidate software.  \wss{Packages that fit our
@@
 \href{https://bssw.io/}{Better Scientific Software (BSSw)},
-\href{https://www.software.ac.uk/}{Software Sustainability Institute}, and
+\href{https://www.software.ac.uk/}{Software Sustainability Institute} \citep{CrouchEtAl2013}, and
 \href{https://software-carpentry.org/}{Software Carpentry}. Some SCS developers
-may not be familiar with these organizations.
+may not be familiar with these organizations. \citep{WilsonAndLumsdaine2006}
 
@@
 
-\subsection{Current Practices} \label{currentPractices}
-
-The following points regarding software quality should be considered when
-developing LBM software packages. These points are based on developer
-interviews, SCS literature, what was found to have worked for packages that were
-designated as high quality in this assessment.  Our recommendations are not
-lists of what should have been done in the past, or what should be done now;
-they are just suggestions for consideration in the future.  \wss{Rework this
-section to highlight what the top projects are doing.  Point to checklists for
-other advice?}
+\subsection{Common Practices} \label{currentPractices}
+
+The practices that arose frequently.  Compare and contrast to what is done with
+research software in general.  The following points regarding software quality
+should be considered when developing LBM software packages. These points are
+based on developer interviews, SCS literature, what was found to have worked for
+packages that were designated as high quality in this assessment.  Our
+recommendations are not lists of what should have been done in the past, or what
+should be done now; they are just suggestions for consideration in the future.
+\wss{Rework this section to highlight what the top projects are doing.  Point to
+checklists for other advice?}  Highlights of common practices.  Put in context.
+
+Conclusion that LBM software common practice is similar to the common practice
+elsewhere.  Maybe point out where LBM is ahead.  Use results from common
+artifacts, common process etc. to support conclusion.  (Continuous integration, version control, README, Req etc. - point to Table 8.)
 
 \begin{itemize}
-	\item Surface Reliability. Include descriptive error messages where
-	appropriate. This was done by most packages that encountered a fault.
-	\item Surface Reliability. In case automatic installation of dependencies
-	fails, the system should indicate to the user what dependencies need to be
-	installed manually. This was done by many packages, including ESPResSo++,
-	Ludwig, LUMA, pyLBM, TCLB.
-	\item Surface Reliability. The packages should include detailed tutorials,
-	including dependencies, expected output, and any additional supplementary
-	documentation that may be required. This was done by many packages,
-	including waLBerla, Palabos, MechSys, LUMA, and pyLBM.  \wss{Comment from
-	ASCR attendee that success with open source requires good tutorials and
-	documentation - is this published anywhere?}
-	\item Surface Robustness. The system must provide descriptive error messages
-	when it encounters unexpected input, including incorrect data types, empty
-	input, and missing files or links. Nineteen of the software packages behaved
-	reasonably when tested with unexpected input.
-	\item Surface Performance. Integrate parallelization tools and techniques to
-	reduce processing time. Consider GPU processing, CUDA, and MPI. Eighteen
-	software packages mentioned parallelization in their artifacts.
-	\item Surface Performance. The user should be able to choose to process
-	their model on either the CPU or GPU. ESPResSo, lbmpy, lettuce, pyLBM, and
-	TCLB allow for either CPU or GPU processing.
+	\item Surface Reliability. Surface Usability. The packages should include
+	detailed tutorials, including dependencies, expected output, and any
+	additional supplementary documentation that may be required. This was done
+	by many packages, including waLBerla, Palabos, MechSys, LUMA, and pyLBM.
+	\wss{Comment from ASCR attendee that success with open source requires good
+	tutorials and documentation - is this published anywhere?  I think I can
+	find this recommendation in other papers.}
+	\item Surface Usability. Unambiguity. Surface Understandability. Provide a detailed user manual. It
+	should identify elements of user interfaces, and identify all requirements
+	to model a system. Thirteen packages have a user manual. ESPResSo has a well
+	detailed manual. \wss{Include this with tutorial?  Find papers that
+	recommend this kind of documentation?}
 	\item Surface Usability. Include user hardware and software requirements
@@
 	packages, even if only some dependencies or a compatible operating system
-	were mentioned.
-	\item Surface Usability. Provide a user tutorial that indicates the expected
-	output. This was done by many packages, including waLBerla, Palabos,
-	MechSys, LUMA, pyLBM.
-	\item Surface Usability. Provide a detailed user manual. It should identify
-	elements of user interfaces, and identify all requirements to model a
-	system. Thirteen packages have a user manual. ESPResSo has a well detailed
-	manual.
-	\item Maintainability. Ideally include most of the common and less common
-	artifacts listed in Section~\ref{artifacts}.
-	\item Surface Understandability. Use consistent, distinctive, and meaningful
-	code identifiers. Twenty of the 24 software packages use such code
-	identifiers.
-	\item Surface Understandability. Use symbolic constants. Thirteen of the 24
-	software packages use such constants.
-	\item Surface Understandability. Identify algorithms that are used in source
-	code comments. Document them in the artifacts. Cite relevant external
-	sources if needed. Domain algorithms are noted in the source code of 11
-	packages.
-	\item Surface Understandability. Provide a user manual that identifies all
-	technical and problem modeling requirements. Thirteen software packages have
-	a user manual.
-	\item Unambiguity. Provide a detailed user manual. Thirteen packages have a
-	user manual. ESPResSo has a well detailed manual.
-
+	were mentioned.  \wss{This is a key part of reproducibility.}
+    \item Not much of a process - find papers that support this - papers recommend "Define clear and transparent contribution, governance and communication processes" JimenezEtAl2017
 \end{itemize}
@@
 - add links or citations to all of the packages (could add them to the summary table)
+
+- introduce the idea of using our measurement template as a checklist, instead of, or in addition to other checklists - cite the other checklists \citep{SchlauchEtAl2018, BrettEtAl2021, vanGompelEtAl2016}, \href{Software Sustainability Institute Online Form}{https://docs.google.com/forms/d/e/1FAIpQLSf0ccsVdN-nXJCHLluJ-hANZlp8rDKgprJa0oTYiLZSDxh3DA/viewform}
+
+- discuss Wilson papers ``Best Practices'' \citep{WilsonEtAl2014} and ``Good Enough Practices'' \citep{WilsonEtAl2016}
+
+- mention flake8 for Python?
+
+- mention \href{https://joss.theoj.org/}{The Journal of Open Source Software}
+
+- reproducibility - can point to the PiccoloAndFrampton2016 paper
+
+- AlNoamanyAndBorghi2018 lists the most popular languages, Python is number 1
+
+- JimenezEtAl2017 talks about Define clear and transparent contribution, governance and communication processes

### 2022-01-07 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
added links to repos in Table 1 and 2

--- 2022-01-07 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
@@
 		\midrule
-		DL\_MESO (LBE) & unclear & 2020 Mar\\
-		ESPResSo & 2010 Nov & 2020 Jun\\
-		ESPResSo++ & 2011 Feb & 2020 Apr\\
-		lbmpy& unclear  & 2020 Jun  \\
-		lettuce & 2019 May & 2020 Jul\\
-		Ludwig& 2018 Aug & 2020 Jul\\
-		LUMA& 2016 Nov   & 2020 Feb\\
-		MechSys & 2008 Jun    & 2021 Oct\\
-		Musubi & 2013 Sep & 2019 Aug\\
-		OpenLB & 2007 Jul & 2019 Oct\\
-		Palabos & unclear & 2020 Jul\\
-		pyLBM & 2015 Jun&   2020 Jun\\
-		Sailfish & 2012 Nov & 2019 Jun\\
-		TCLB & 2013 Jun  & 2020 Apr\\
-		waLBerla & 2008 Aug & 2020 Jul\\
+		\href{https://www.ccp5.ac.uk/sites/www.ccp5.ac.uk/files/dl_meso/data/dl_meso_2.7.zip}{DL\_MESO (LBE)} & unclear & 2020 Mar\\
+		\href{https://github.com/espressomd/espresso}{ESPResSo} & 2010 Nov & 2020 Jun\\
+		\href{https://github.com/espressopp/espressopp}{ESPResSo++} & 2011 Feb & 2020 Apr\\
+		\href{https://github.com/mabau/lbmpy}{lbmpy}& unclear  & 2020 Jun  \\
+		\href{https://github.com/Olllom/lettuce}{lettuce} & 2019 May & 2020 Jul\\
+		\href{https://github.com/ludwig-cf/ludwig}{Ludwig} & 2018 Aug & 2020 Jul\\
+		\href{https://github.com/aharwood2/LUMA}{LUMA} & 2016 Nov   & 2020 Feb\\
+		\href{http://hg.savannah.gnu.org/hgweb/mechsys/file/tip/}{MechSys} & 2008 Jun    & 2021 Oct\\
+		\href{https://osdn.net/projects/apes/scm/hg/musubi/}{Musubi} & 2013 Sep & 2019 Aug\\
+		\href{https://www.openlb.net/download/}{OpenLB} & 2007 Jul & 2019 Oct\\
+		\href{https://gitlab.com/unigespc/palabos}{Palabos} & unclear & 2020 Jul\\
+		\href{https://github.com/pylbm/pylbm}{pyLBM} & 2015 Jun&   2020 Jun\\
+		\href{https://github.com/sailfish-team/sailfish}{Sailfish} & 2012 Nov & 2019 Jun\\
+		\href{https://github.com/CFD-GO/TCLB}{TCLB} & 2013 Jun  & 2020 Apr\\
+		\href{https://i10git.cs.fau.de/walberla/walberla}{waLBerla} & 2008 Aug & 2020 Jul\\
 		\bottomrule
@@
 		\midrule
-		HemeLB & 2007 Jun & 2018 Aug\\
-		laboetie & 2014 Nov & 2018 Aug\\		
-		LatBo.jl & 2014 Aug & 2017 Feb\\
-		LB2D-Prime & 2005 & 2012 Apr\\
-		LB3D & unclear & 2012 Mar\\
-		LB3D-Prime & 2005 & 2011 Oct\\
-		LIMBES & 2010 Nov & 2014 Dec\\
-		MP-LABS & 2008 Jun & 2014 Oct\\
-		SunlightLB & 2005 Sep & 2012 Nov\\
+		\href{https://github.com/UCL/hemelb}{HemeLB} & 2007 Jun & 2018 Aug\\
+		\href{https://github.com/maxlevesque/laboetie}{laboetie} & 2014 Nov & 2018 Aug\\		
+		\href{https://github.com/UCL/LatBo.jl}{LatBo.jl} & 2014 Aug & 2017 Feb\\
+		\href{https://code.google.com/p/lb2d-prime-dev/source}{LB2D-Prime} & 2005 & 2012 Apr\\
+		\href{http://ccs.chem.ucl.ac.uk/sites/ccs.chem.ucl.ac.uk/themes/ccs2/files/lb3d-2012-03-12.tgz}{LB3D} & unclear & 2012 Mar\\
+		\href{https://code.google.com/archive/p/lb3d-prime-dev/source}{LB3D-Prime} & 2005 & 2011 Oct\\
+		\href{https://code.google.com/archive/p/limbes/source}{LIMBES} & 2010 Nov & 2014 Dec\\
+		\href{https://github.com/carlosrosales/mplabs}{MP-LABS} & 2008 Jun & 2014 Oct\\
+		\href{https://sourceforge.net/projects/sunlightlb/files/latest/download}{SunlightLB} & 2005 Sep & 2012 Nov\\
 		\bottomrule

### 2022-01-07 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
added some publications into tables

--- 2022-01-07 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
@@
 \begin{center}
-	\begin{tabular}{ p{4cm}p{3cm}p{3cm} }
+	\begin{tabular}{ p{3cm}p{2cm}p{2cm}p{7cm} }
 		\toprule
-		Name & Released & Updated\\
+		Name & Released & Updated & Relevant Publication\\
 		\midrule
-		DL\_MESO (LBE) & unclear & 2020 Mar\\
-		ESPResSo & 2010 Nov & 2020 Jun\\
-		ESPResSo++ & 2011 Feb & 2020 Apr\\
-		lbmpy& unclear  & 2020 Jun  \\
-		lettuce & 2019 May & 2020 Jul\\
-		Ludwig& 2018 Aug & 2020 Jul\\
-		LUMA& 2016 Nov   & 2020 Feb\\
-		MechSys & 2008 Jun    & 2021 Oct\\
-		Musubi & 2013 Sep & 2019 Aug\\
-		OpenLB & 2007 Jul & 2019 Oct\\
-		Palabos & unclear & 2020 Jul\\
-		pyLBM & 2015 Jun&   2020 Jun\\
-		Sailfish & 2012 Nov & 2019 Jun\\
-		TCLB & 2013 Jun  & 2020 Apr\\
-		waLBerla & 2008 Aug & 2020 Jul\\
+		\href{https://www.ccp5.ac.uk/sites/www.ccp5.ac.uk/files/dl_meso/data/dl_meso_2.7.zip}{DL\_MESO (LBE)} & unclear & 2020 Mar&\\
+		\href{https://github.com/espressomd/espresso}{ESPResSo} & 2010 Nov & 2020 Jun&\\
+		\href{https://github.com/espressopp/espressopp}{ESPResSo++} & 2011 Feb & 2020 Apr&\\
+		\href{https://github.com/mabau/lbmpy}{lbmpy}& unclear  & 2020 Jun & \\
+		\href{https://github.com/Olllom/lettuce}{lettuce} & 2019 May & 2020 Jul&\\
+		\href{https://github.com/ludwig-cf/ludwig}{Ludwig} & 2018 Aug & 2020 Jul&\\
+		\href{https://github.com/aharwood2/LUMA}{LUMA} & 2016 Nov   & 2020 Feb &\\
+		\href{http://hg.savannah.gnu.org/hgweb/mechsys/file/tip/}{MechSys} & 2008 Jun    & 2021 Oct &\\
+		\href{https://osdn.net/projects/apes/scm/hg/musubi/}{Musubi} & 2013 Sep & 2019 Aug &\\
+		\href{https://www.openlb.net/download/}{OpenLB} & 2007 Jul & 2019 Oct &\\
+		\href{https://gitlab.com/unigespc/palabos}{Palabos} & unclear & 2020 Jul &\\
+		\href{https://github.com/pylbm/pylbm}{pyLBM} & 2015 Jun&   2020 Jun &\\
+		\href{https://github.com/sailfish-team/sailfish}{Sailfish} & 2012 Nov & 2019 Jun &\\
+		\href{https://github.com/CFD-GO/TCLB}{TCLB} & 2013 Jun  & 2020 Apr & Adjoint Lattice Boltzmann for topology optimization on multi-GPU architecture \citep{rokicki2016adjoint}\\
+		\href{https://i10git.cs.fau.de/walberla/walberla}{waLBerla} & 2008 Aug & 2020 Jul & WALBERLA: A block-structured high-performance framework for multiphysics simulations \citep{bauer2021walberla}\\
 		\bottomrule
@@
 \begin{center}
-	\begin{tabular}{ p{4cm}p{3cm}p{3cm} }
+	\begin{tabular}{ p{3cm}p{2cm}p{2cm}p{7cm} }
 		\toprule
-		Name & Released & Updated\\
+		Name & Released & Updated & Relevant Publication\\
 		\midrule
-		HemeLB & 2007 Jun & 2018 Aug\\
-		laboetie & 2014 Nov & 2018 Aug\\		
-		LatBo.jl & 2014 Aug & 2017 Feb\\
-		LB2D-Prime & 2005 & 2012 Apr\\
-		LB3D & unclear & 2012 Mar\\
-		LB3D-Prime & 2005 & 2011 Oct\\
-		LIMBES & 2010 Nov & 2014 Dec\\
-		MP-LABS & 2008 Jun & 2014 Oct\\
-		SunlightLB & 2005 Sep & 2012 Nov\\
+		\href{https://github.com/UCL/hemelb}{HemeLB} & 2007 Jun & 2018 Aug&\\
+		\href{https://github.com/maxlevesque/laboetie}{laboetie} & 2014 Nov & 2018 Aug&\\		
+		\href{https://github.com/UCL/LatBo.jl}{LatBo.jl} & 2014 Aug & 2017 Feb&\\
+		\href{https://code.google.com/p/lb2d-prime-dev/source}{LB2D-Prime} & 2005 & 2012 Apr&\\
+		\href{http://ccs.chem.ucl.ac.uk/sites/ccs.chem.ucl.ac.uk/themes/ccs2/files/lb3d-2012-03-12.tgz}{LB3D} & unclear & 2012 Mar&\\
+		\href{https://code.google.com/archive/p/lb3d-prime-dev/source}{LB3D-Prime} & 2005 & 2011 Oct&\\
+		\href{https://code.google.com/archive/p/limbes/source}{LIMBES} & 2010 Nov & 2014 Dec&\\
+		\href{https://github.com/carlosrosales/mplabs}{MP-LABS} & 2008 Jun & 2014 Oct&\\
+		\href{https://sourceforge.net/projects/sunlightlb/files/latest/download}{SunlightLB} & 2005 Sep & 2012 Nov&\\
 		\bottomrule

### 2022-01-07 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Updates to LBM SOP

@@
 
+\subsection{Analytical Hierarchy Process} \label{AHP}
+
+The Analytical Hierarchy Process (AHP) is a decision-making technique that is
+used to compare multiple options by multiple criteria. In our work AHP was used
+for comparing and ranking the LBM software packages using the overall impression
+quality scores that were gathered in the measurement template.  AHP performs a
+pairwise analysis using a matrix and generates an overall score as well as
+individual quality scores for each software package. \citep{SmithEtAl2016} shows
+how AHP is applied to ranking software based on quality measures.
+
+This project used a tool for conducting this process. The tool includes a
+sensitivity analysis that was used to ensure that the software package rankings
+are appropriate with respect to the uncertainty of the quality scores. For the
+sensitivity analysis we modified the score by 10\% for each package and verified
+that the overall ranking was stable.  The
+\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/AHP2020/LBM/README.txt}{README}
+file of the tool includes requirements and usage information.
+
 \subsection{Interview Developers} \label{SecSurvey}
@@
 {MREB\#: 5219}.
-
-\subsection{Analytical Hierarchy Process} \label{AHP}
-
-The Analytical Hierarchy Process (AHP) is a decision-making technique that is
-used to compare multiple options by multiple criteria. In our work AHP was used
-for comparing and ranking the LBM software packages using the overall impression
-quality scores that were gathered in the measurement template.  AHP performs a
-pairwise analysis using a matrix and generates an overall score as well as
-individual quality scores for each software package. \citep{SmithEtAl2016} shows
-how AHP is applied to ranking software based on quality measures.
-
-This project used a tool for conducting this process. The tool includes a
-sensitivity analysis that was used to ensure that the software package rankings
-are appropriate with respect to the uncertainty of the quality scores. For the
-sensitivity analysis we modified the score by 10\% for each package and verified
-that the overall ranking was stable.  The
-\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/AHP2020/LBM/README.txt}{README}
-file of the tool includes requirements and usage information.
 
@@
 
-\section{Quantitative Findings and AHP Results} \label{AHPresults}
+\section{Ranking Projects Based on Quality Measures} \label{AHPresults}
 
@@
 	Code Editors & Technical Libraries & Email\\
-	Development Environment & Domain Specific Libraries & Change Tracking Tools\\
+	Development Environments & Domain Specific Libraries & Change Tracking Tools\\
 	Runtime Environments &  & Version Control Tools\\

### 2022-01-09 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
added publications to Tables 1 and 2

@@
 	\begin{center}
-	\begin{tabular}{ p{3cm}p{2cm}p{2cm}p{7cm} }
+		\begin{tabular}{ p{3cm}p{2cm}p{2cm}p{5.5cm} }
 			\toprule
@@
 			\midrule
-		\href{https://www.ccp5.ac.uk/sites/www.ccp5.ac.uk/files/dl_meso/data/dl_meso_2.7.zip}{DL\_MESO (LBE)} & unclear & 2020 Mar&\\
-		\href{https://github.com/espressomd/espresso}{ESPResSo} & 2010 Nov & 2020 Jun&\\
-		\href{https://github.com/espressopp/espressopp}{ESPResSo++} & 2011 Feb & 2020 Apr&\\
-		\href{https://github.com/mabau/lbmpy}{lbmpy}& unclear  & 2020 Jun & \\
-		\href{https://github.com/Olllom/lettuce}{lettuce} & 2019 May & 2020 Jul&\\
-		\href{https://github.com/ludwig-cf/ludwig}{Ludwig} & 2018 Aug & 2020 Jul&\\
-		\href{https://github.com/aharwood2/LUMA}{LUMA} & 2016 Nov   & 2020 Feb &\\
-		\href{http://hg.savannah.gnu.org/hgweb/mechsys/file/tip/}{MechSys} & 2008 Jun    & 2021 Oct &\\
-		\href{https://osdn.net/projects/apes/scm/hg/musubi/}{Musubi} & 2013 Sep & 2019 Aug &\\
-		\href{https://www.openlb.net/download/}{OpenLB} & 2007 Jul & 2019 Oct &\\
-		\href{https://gitlab.com/unigespc/palabos}{Palabos} & unclear & 2020 Jul &\\
+			\href{https://www.ccp5.ac.uk/sites/www.ccp5.ac.uk/files/dl_meso/data/dl_meso_2.7.zip}{DL\_MESO (LBE)} & unclear & 2020 Mar&\citep{seaton2013dlmeso}\\
+			\href{https://github.com/espressomd/espresso}{ESPResSo} & 2010 Nov & 2020 Jun&\citep{weik2019espresso}\\
+			\href{https://github.com/espressopp/espressopp}{ESPResSo++} & 2011 Feb & 2020 Apr&\citep{halverson2013espresso++}\\
+			\href{https://github.com/mabau/lbmpy}{lbmpy}& unclear  & 2020 Jun & \citep{bauer2021lbmpy}\\
+			\href{https://github.com/Olllom/lettuce}{lettuce} & 2019 May & 2020 Jul&\citep{bedrunka2021lettuce}\\
+			\href{https://github.com/ludwig-cf/ludwig}{Ludwig} & 2018 Aug & 2020 Jul&\citep{desplat2001ludwig}\\
+			\href{https://github.com/aharwood2/LUMA}{LUMA} & 2016 Nov   & 2020 Feb &\citep{harwood2018luma}\\
+			\href{http://hg.savannah.gnu.org/hgweb/mechsys/file/tip/}{MechSys} & 2008 Jun    & 2021 Oct &\citep{galindo2013coupled}\\
+			\href{https://osdn.net/projects/apes/scm/hg/musubi/}{Musubi} & 2013 Sep & 2019 Aug &\citep{hasert2014complex}\\
+			\href{https://www.openlb.net/download/}{OpenLB} & 2007 Jul & 2019 Oct &\citep{heuveline2010openlb}\\
+			\href{https://gitlab.com/unigespc/palabos}{Palabos} & unclear & 2020 Jul &\citep{latt2021palabos}\\
 			\href{https://github.com/pylbm/pylbm}{pyLBM} & 2015 Jun&   2020 Jun &\\
-		\href{https://github.com/sailfish-team/sailfish}{Sailfish} & 2012 Nov & 2019 Jun &\\
-		\href{https://github.com/CFD-GO/TCLB}{TCLB} & 2013 Jun  & 2020 Apr & Adjoint Lattice Boltzmann for topology optimization on multi-GPU architecture \citep{rokicki2016adjoint}\\
-		\href{https://i10git.cs.fau.de/walberla/walberla}{waLBerla} & 2008 Aug & 2020 Jul & WALBERLA: A block-structured high-performance framework for multiphysics simulations \citep{bauer2021walberla}\\
+			\href{https://github.com/sailfish-team/sailfish}{Sailfish} & 2012 Nov & 2019 Jun & \citep{januszewski2014sailfish}\\
+			\href{https://github.com/CFD-GO/TCLB}{TCLB} & 2013 Jun  & 2020 Apr & \citep{rokicki2016adjoint}\\
+			\href{https://i10git.cs.fau.de/walberla/walberla}{waLBerla} & 2008 Aug & 2020 Jul & \citep{bauer2021walberla}\\
 			\bottomrule
@@
 			\midrule
-		\href{https://github.com/UCL/hemelb}{HemeLB} & 2007 Jun & 2018 Aug&\\
-		\href{https://github.com/maxlevesque/laboetie}{laboetie} & 2014 Nov & 2018 Aug&\\		
+			\href{https://github.com/UCL/hemelb}{HemeLB} & 2007 Jun & 2018 Aug&\citep{mazzeo2008hemelb}\\
+			\href{https://github.com/maxlevesque/laboetie}{laboetie} & 2014 Nov & 2018 Aug&\citep{levesque2013accounting}\\		
 			\href{https://github.com/UCL/LatBo.jl}{LatBo.jl} & 2014 Aug & 2017 Feb&\\
 			\href{https://code.google.com/p/lb2d-prime-dev/source}{LB2D-Prime} & 2005 & 2012 Apr&\\
-		\href{http://ccs.chem.ucl.ac.uk/sites/ccs.chem.ucl.ac.uk/themes/ccs2/files/lb3d-2012-03-12.tgz}{LB3D} & unclear & 2012 Mar&\\
+			\href{http://ccs.chem.ucl.ac.uk/sites/ccs.chem.ucl.ac.uk/themes/ccs2/files/lb3d-2012-03-12.tgz}{LB3D} & unclear & 2012 Mar&\citep{schmieschek2017lb3d}\\
 			\href{https://code.google.com/archive/p/lb3d-prime-dev/source}{LB3D-Prime} & 2005 & 2011 Oct&\\

### 2022-01-09 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Updates to LBM SOP

@@
 confusion between understandability and usability was also a confusion during
-the interviews Ao conducted.} \pmm{The developer of ESPResSo did comment on developer understandability, specifically for developers that are/were part of their peer review process.
-He commented that code architecture is documented, and that code is made to be readable, and that these help peer reviews of the code.}
+the interviews Ao conducted.} \pmm{The developer of ESPResSo did comment on
+developer understandability, specifically for developers that are/were part of
+their peer review process. He commented that code architecture is documented,
+and that code is made to be readable, and that these help peer reviews of the
+code.}
 
@@
 you say which team is international, so that we can add this detail to the
-paper?} \pmm{The interviewed developer of OpenLB noted that there were tasks for local and international developers when discussing their software development model.
-He also noted that they have frequent local meetings and annual international meetings.}  For teams of only a few developers, new features or major changes are
-discussed with the entire team. Projects with more than a few developers have
-lead developer roles. These lead developers review potential additions to the
-software.
+paper?} \pmm{The interviewed developer of OpenLB noted that there were tasks for
+local and international developers when discussing their software development
+model. He also noted that they have frequent local meetings and annual
+international meetings.}  For teams of only a few developers, new features or
+major changes are discussed with the entire team. Projects with more than a few
+developers have lead developer roles. These lead developers review potential
+additions to the software.
 
@@
 Dependency Notes & Functional Spec. / Notes & FAQ / Forum\\
-Installation Guide / Instructions & Performance Information & Verification and Validation Plan\\
-Requirements Spec. / Theory Notes & Test Plan / Script / Cases & Video Guide (including YouTube)\\
+Installation Guide / Instructions & Performance Information & Verification and
+Validation Plan\\
+Requirements Spec. / Theory Notes & Test Plan / Script / Cases & Video Guide
+(including YouTube)\\
 List of Related Publications & User Manual/Guide & \\
@@
 \href{https://bssw.io/}{Better Scientific Software (BSSw)},
-\href{https://www.software.ac.uk/}{Software Sustainability Institute} \citep{CrouchEtAl2013}, and
-\href{https://software-carpentry.org/}{Software Carpentry}. Some SCS developers
-may not be familiar with these organizations. \citep{WilsonAndLumsdaine2006}
+\href{https://www.software.ac.uk/}{Software Sustainability Institute}
+\citep{CrouchEtAl2013}, and \href{https://software-carpentry.org/}{Software
+Carpentry}. Some SCS developers may not be familiar with these organizations.
+\citep{WilsonAndLumsdaine2006}
 
@@
 elsewhere.  Maybe point out where LBM is ahead.  Use results from common
-artifacts, common process etc. to support conclusion.  (Continuous integration, version control, README, Req etc. - point to Table 8.)
+artifacts, common process etc. to support conclusion.  (Continuous integration,
+version control, README, Req etc. - point to Table 8.)
 
@@
 	find this recommendation in other papers.}
-	\item Surface Usability. Unambiguity. Surface Understandability. Provide a detailed user manual. It
-	should identify elements of user interfaces, and identify all requirements
-	to model a system. Thirteen packages have a user manual. ESPResSo has a well
-	detailed manual. \wss{Include this with tutorial?  Find papers that
-	recommend this kind of documentation?}
+	\item Surface Usability. Unambiguity. Surface Understandability. Provide a
+	detailed user manual. It should identify elements of user interfaces, and
+	identify all requirements to model a system. Thirteen packages have a user
+	manual. ESPResSo has a well detailed manual. \wss{Include this with
+	tutorial?  Find papers that recommend this kind of documentation?}
 	\item Surface Usability. Include user hardware and software requirements
@@
 	were mentioned.  \wss{This is a key part of reproducibility.}
-    \item Not much of a process - find papers that support this - papers recommend "Define clear and transparent contribution, governance and communication processes" JimenezEtAl2017
+    \item Not much of a process - find papers that support this - papers
+    recommend "Define clear and transparent contribution, governance and
+    communication processes" JimenezEtAl2017
 \end{itemize}
@@
 	suggestion is based on developer comments.
-	\item Surface Correctness and Verifiability. Traceability. Visibility and Transparency. Use and make public (on package
-	website or GitHub) detailed documentation. Consider using automatic document
-	generation tools like Doxygen, Drasil, or Sphinx. This was done by all of
-	the top five ranked packages.
+	\item Surface Correctness and Verifiability. Traceability. Visibility and
+	Transparency. Use and make public (on package website or GitHub) detailed
+	documentation. Consider using automatic document generation tools like
+	Doxygen, Drasil, or Sphinx. This was done by all of the top five ranked
+	packages.
 	\item Surface Correctness and Verifiability. Provide detailed tutorials that
@@
 	difficult to maintain. 
-	\item Maintainability. Visibility and Transparency. Reproducibility. Keep artifacts updated.
-	This is suggested as good software maintenance practice.
+	\item Maintainability. Visibility and Transparency. Reproducibility. Keep
+	artifacts updated. This is suggested as good software maintenance practice.
 	\item Maintainability. Eliminate code duplication. This is suggested as good
@@
 
-- introduce the idea of using our measurement template as a checklist, instead of, or in addition to other checklists - cite the other checklists \citep{SchlauchEtAl2018, BrettEtAl2021, vanGompelEtAl2016}, \href{Software Sustainability Institute Online Form}{https://docs.google.com/forms/d/e/1FAIpQLSf0ccsVdN-nXJCHLluJ-hANZlp8rDKgprJa0oTYiLZSDxh3DA/viewform}
-
-- discuss Wilson papers ``Best Practices'' \citep{WilsonEtAl2014} and ``Good Enough Practices'' \citep{WilsonEtAl2016}
+- introduce the idea of using our measurement template as a checklist, instead
+of, or in addition to other checklists - cite the other checklists
+\citep{SchlauchEtAl2018, BrettEtAl2021, vanGompelEtAl2016}, \href{Software
+Sustainability Institute Online
+Form}{https://docs.google.com/forms/d/e/1FAIpQLSf0ccsVdN-nXJCHLluJ-hANZlp8rDKgprJa0oTYiLZSDxh3DA/viewform}
+
+- discuss Wilson papers ``Best Practices'' \citep{WilsonEtAl2014} and ``Good
+Enough Practices'' \citep{WilsonEtAl2016}
 
@@
 
-- JimenezEtAl2017 talks about Define clear and transparent contribution, governance and communication processes
+- JimenezEtAl2017 talks about Define clear and transparent contribution,
+governance and communication processes

### 2022-01-10 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Revised RQs

@@
 +}\nolinebreak\hspace{-.10em}\raisebox{.4ex}{\small\bf +}}
+
+\newcounter{rqnum} %research question number
+\newcommand{\rqtherqnum}{RQ`'\therqnum}
+\newcommand{\rqref}[1]{RQ\ref{#1}}
 
@@
 \begin{enumerate}
-	\item What artifacts are present in current LBM packages? 
-	\item What tools (development, dependencies, project management) are used by
-	current LBM solvers?
-	\item What principles, processes, and methodologies are used in the development
-	of current LBM packages?
-	\item What are the pain points for developers working on LBM software
-	projects?  What aspects of the existing processes, methodologies and tools do
-	they consider as potentially needing improvement?  How should processes,
-	methodologies and tools be changed to improve software development and
-	software quality?
-	\item For LBM software developers, what specific actions are taken to
-	address the following:
+	\item[RQ\refstepcounter{rqnum}\therqnum \label{RQ_WhatProjects}:] What LBM
+	software projects exist, with the constraint that the source code must be
+	available for all identified projects?
+	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_HighestQuality}:] Which
+	of the projects identified in \rqref{RQ_WhatProjects} follow current best
+	practices, based on evidence found by experimenting with the software and
+	searching the artifacts available in each project's repository?
+	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareHQ2Popular}:] How
+	similar is the list of top projects identified in \rqref{RQ_HighestQuality}
+	to the most popular projects, as viewed by the scientific community?
+    \item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareArtifacts}:] How
+	do LBM projects compare to research software in general with respect to
+	the artifacts present in their repositories?
+	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareToolsProjMngmnt}:]
+	How do LBM projects compare to research software in general with
+	respect to the use of tools for:
 	\begin{enumerate} 
-		\item installability
-		\item correctness and verifiability
-		\item reliability
-		\item robustness
-		\item performance
-		\item usability
-		\item maintainability
-		\item modifiability
-		\item reusability
-		\item understandability
-		\item traceability
-		\item visibility and transparency
-		\item reproducibility
-		\item unambiguity
+		\item [\rqref{RQ_CompareToolsProjMngmnt}.a] development; and,
+		\item [\rqref{RQ_CompareToolsProjMngmnt}.b] project management?
 	\end{enumerate}
-	\item How does software designated as high quality by this methodology
-	compare	with top rated software by the community?
+	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_PainPoints}:] What are
+	the pain points for developers working on LBM software projects? 
+	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_ComparePainPoints}:] How
+	do the pain points of developers from LBM compare to the pain points
+	for research software in general?
+	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Concerns}:] For LBM
+	developers:
+	\begin{enumerate} 
+		\item [\rqref{RQ_Concerns}.a] what software development concerns exist; and,
+		\item [\rqref{RQ_Concerns}.b] what specific actions are taken to address
+		these concerns?
+	\end{enumerate}
+	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareConcerns}:]
+	Compared to research software in general, how is LBM software (as answered
+	in \rqref{RQ_Concerns}) different with respect to:
+	\begin{enumerate} 
+		\item [\rqref{RQ_CompareConcerns}.a] software development concerns?; and,
+		\item [\rqref{RQ_CompareConcerns}.b] specific actions taken to address
+		concerns?
+	\end{enumerate}
+	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Future}:] What actions
+	could be employed by future LBM developers to address the concerns
+	identified in \rqref{RQ_Concerns} and the pain points found in
+	\rqref{RQ_PainPoints}?
+
 \end{enumerate}

### 2022-01-10 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Revised order of sections, added more explicit references to the RQs

@@
 	\end{enumerate}
+	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareMethodologies}:]
+	How do LBM projects compare to research software in general with respect to
+	principles, processes and methodologies used?
 	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_PainPoints}:] What are
@@
 		\item [\rqref{RQ_Concerns}.a] what software development concerns exist; and,
-		\item [\rqref{RQ_Concerns}.b] what specific actions are taken to address
-		these concerns?
+		\item [\rqref{RQ_Concerns}.b] what specific practices are taken to
+		address these concerns?
 	\end{enumerate}
-	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareConcerns}:]
-	Compared to research software in general, how is LBM software (as answered
-	in \rqref{RQ_Concerns}) different with respect to:
-	\begin{enumerate} 
-		\item [\rqref{RQ_CompareConcerns}.a] software development concerns?; and,
-		\item [\rqref{RQ_CompareConcerns}.b] specific actions taken to address
-		concerns?
-	\end{enumerate}
-	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Future}:] What actions
-	could be employed by future LBM developers to address the concerns
-	identified in \rqref{RQ_Concerns} and the pain points found in
-	\rqref{RQ_PainPoints}?
+	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Recommend}:]
+	What, if any, existing research software best practice, not currently used
+	by LBM software, could address the concerns identified in
+	\rqref{RQ_Concerns}).
+	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Future}:] What practices
+	could be employed by future LBM developers to address the pain points found
+	in \rqref{RQ_PainPoints} and the concerns identified in \rqref{RQ_Concerns}.a?
 
@@
 We developed a methodology for evaluating the state of the practice of
-scientific software \citep{SmithEtAl2021}.  This methodology can be instantiated
-for a specific domain of scientific software, which in the current case is LBM
-software.  The first section below describes the steps of the process we used to
-select, measure and compare LBM software. This is followed by quality
-definitions and how these qualities were assessed in our project. The rest of
-the section provides the details of the methodology.
+scientific software \citep{SmithEtAl2021}.  The methodology aims to answer the
+research questions posed in the previous section. This methodology can be
+instantiated for a specific domain of scientific software, which in the current
+case is LBM software.  The first section below describes the steps of the
+process we used to select, measure and compare LBM software. This is followed by
+quality definitions and how these qualities were assessed in our project. The
+rest of the section provides the details of the methodology.
 % an overview of the specifics of candidate software selection, the quantitative
@@
 
-\subsection{Process}
+\subsection{Overall Process}
 
@@
 
-The candidate software was found through search engine queries targeting
-authoritative lists of software. We found LBM software listed on the websites
-GitHub and swMATH, as well as through articles found in scholarly journals and
-databases.  The Domain Expert (Section~\ref{sec_vet_software_list}) was also
-engaged in selecting the candidate software.  \wss{Packages that fit our
-criteria but were found after data collection and analysis was conducted, such
-as Musubi \citep{HasertEtAl2014}, will need to be added to future SOP
-assessments.}
+Part of how we answer \rqref{RQ_WhatProjects}.  The candidate software was found
+through search engine queries targeting authoritative lists of software. We
+found LBM software listed on the websites GitHub and swMATH, as well as through
+articles found in scholarly journals and databases.  The Domain Expert
+(Section~\ref{sec_vet_software_list}) was also engaged in selecting the
+candidate software.  \wss{Packages that fit our criteria but were found after
+data collection and analysis was conducted, such as Musubi
+\citep{HasertEtAl2014}, will need to be added to future SOP assessments.}
 
@@
 
-To reduce the number of members in the candidate software list to a manageable
-size, the following filters were applied. The filters were applied in the
-priority order listed.
+With previous section, how we answer \rqref{RQ_WhatProjects}.  Answer to
+\rqref{RQ_WhatProjects} provided in this section.  To reduce the number of
+members in the candidate software list to a manageable size, the following
+filters were applied. The filters were applied in the priority order listed.
 
@@
 
-The measurement template is found in \citet{SmithEtAl2021}.  This template is
-used to track measurements and quality scores for all of the software packages
-in the domain. For each software package, we fill-in the rows of the template.
-This process takes about 2 hours per package, with a cap of 4 hours.  The time
+Part of how we address \rqref{RQ_HighestQuality} and
+\rqref{RQ_CompareArtifacts}.  Once have \rqref{RQ_HighestQuality} can combine
+with estimate of popularity to answer \rqref{RQ_CompareHQ2Popular}.  The
+measurement template is found in \citet{SmithEtAl2021}.  This template is used
+to track measurements and quality scores for all of the software packages in the
+domain. For each software package, we fill-in the rows of the template. This
+process takes about 2 hours per package, with a cap of 4 hours.  The time
 constraint is necessary so that the work load for measuring a given domain is
@@
 
-The Analytical Hierarchy Process (AHP) is a decision-making technique that is
-used to compare multiple options by multiple criteria. In our work AHP was used
-for comparing and ranking the LBM software packages using the overall impression
-quality scores that were gathered in the measurement template.  AHP performs a
-pairwise analysis using a matrix and generates an overall score as well as
-individual quality scores for each software package. \citep{SmithEtAl2016} shows
-how AHP is applied to ranking software based on quality measures.
+Part of how we address \rqref{RQ_HighestQuality}.  The Analytical Hierarchy
+Process (AHP) is a decision-making technique that is used to compare multiple
+options by multiple criteria. In our work AHP was used for comparing and ranking
+the LBM software packages using the overall impression quality scores that were
+gathered in the measurement template.  AHP performs a pairwise analysis using a
+matrix and generates an overall score as well as individual quality scores for
+each software package. \citep{SmithEtAl2016} shows how AHP is applied to ranking
+software based on quality measures.
 
@@
 
-We designed a list of 20 questions to guide our interviews, which can be found
-in \citet{SmithEtAl2021}. Some questions are about the background of the
-software, the development teams, the interviewees, and how they organize the
-projects. We also ask about the developer's understandings of the users. Some
-questions focus on the current and past difficulties, and the solutions the team
-has found, or will try. We also discuss the importance and current situations of
-documentation. A few questions are about specific software qualities, such as
-maintainability, understandability, usability, and reproducibility. The
-interviews are semi-structured based on the question list; we ask follow-up
-questions when necessary. Based on our experience, the interviewees usually
-bring up some exciting ideas that we did not expect, and it is worth expanding
-on these topics.
+Part of how we address \rqref{RQ_CompareToolsProjMngmnt},
+\rqref{RQ_CompareMethodologies}, \rqref{RQ_PainPoints} (which enables
+\rqref{RQ_ComparePainPoints}, and \rqref{RQ_Concerns} (which enables
+\rqref{RQ_CompareConcerns}). We designed a list of 20 questions to guide our
+interviews, which can be found in \citet{SmithEtAl2021}. Some questions are
+about the background of the software, the development teams, the interviewees,
+and how they organize the projects. We also ask about the developer's
+understandings of the users. Some questions focus on the current and past
+difficulties, and the solutions the team has found, or will try. We also discuss
+the importance and current situations of documentation. A few questions are
+about specific software qualities, such as maintainability, understandability,
+usability, and reproducibility. The interviews are semi-structured based on the
+question list; we ask follow-up questions when necessary. Based on our
+experience, the interviewees usually bring up some exciting ideas that we did
+not expect, and it is worth expanding on these topics.
 
@@
 
-The Domain Expert is an important member of the state of the practice assessment
-team. Pitfalls exist if non-experts attempt to acquire an authoritative list of
-software, or try to definitively rank the software. Non-experts have the problem
-that they can only rely on information available on-line, which has the
-following drawbacks:
+Part of how we address \rqref{RQ_WhatProjects} and \rqref{RQ_HighestQuality}.
+\wss{Should ask Zahra about \rqref{RQ_CompareHQ2Popular}, see if he has anything
+to add about \rqref{RQ_PainPoints}.} The Domain Expert is an important member of
+the state of the practice assessment team. Pitfalls exist if non-experts attempt
+to acquire an authoritative list of software, or try to definitively rank the
+software. Non-experts have the problem that they can only rely on information
+available on-line, which has the following drawbacks:
 \begin{inparaenum}[i)]
@@
 
-Before discussing quality differences between the packages and the results of
+Answers \rqref{RQ_HighestQuality}.  Before discussing quality differences between the packages and the results of
 the AHP ranking, we will contrast the different packages based on features, as
@@
 
-Table~\ref{repometrics} presents our LBM software package rankings along with
-the repository ranking metrics of each software package. Nine packages do not
-use GitHub, so they do not have a measure of repository stars. Looking at the
-repository stars of the other 15 packages, we can observe a slight pattern where
-packages that have been highly ranked by our assessment do have more stars than
-lower ranked packages. The best ranked package (ESPResSo) has the second most
-number of stars. The ninth ranked package (Sailfish) has the most number of
-stars. Packages designated as lower quality often do not use GitHub or GitLab,
-or have few stars. HemeLB only has 12 stars, but its location recently moved
-from a different GitHub repository. Our assessment of this package might not be
-accurate. The number of stars does not necessarily represent the perceptions of
-the community, but for lack of an alternative measure we will use it this way.
-Besides missing data, another threat to the validity of this comparison is the
-varying ages of the repositories. Older packages have been able to accumulate
-stars and watches for longer than newer packages. The true quality of new
-packages may not be reflected in their stars and watches.
+Answers \rqref{RQ_CompareHQ2Popular}. Table~\ref{repometrics} presents our LBM
+software package rankings along with the repository ranking metrics of each
+software package. Nine packages do not use GitHub, so they do not have a measure
+of repository stars. Looking at the repository stars of the other 15 packages,
+we can observe a slight pattern where packages that have been highly ranked by
+our assessment do have more stars than lower ranked packages. The best ranked
+package (ESPResSo) has the second most number of stars. The ninth ranked package
+(Sailfish) has the most number of stars. Packages designated as lower quality
+often do not use GitHub or GitLab, or have few stars. HemeLB only has 12 stars,
+but its location recently moved from a different GitHub repository. Our
+assessment of this package might not be accurate. The number of stars does not
+necessarily represent the perceptions of the community, but for lack of an
+alternative measure we will use it this way. Besides missing data, another
+threat to the validity of this comparison is the varying ages of the
+repositories. Older packages have been able to accumulate stars and watches for
+longer than newer packages. The true quality of new packages may not be
+reflected in their stars and watches.
 
@@
 
-\section{Qualitative Findings From Developer Interviews} \label{interviewresults}
-
-This section presents qualitative findings on software qualities as gleaned from
-interviewing 4 developers. The full interview questions are found in
-\citet{SmithEtAl2021}.
-
-\subsection{Correctness and Verifiability}
-
-Interviews with developers confirmed that these software packages are developed
-by domain experts with backgrounds in physics, mathematics and mechanical
-engineering. \wss{Could point to surveys that have shown the same thing.}  It
-was noted in these interviews that some of the developers do not have formal
-software engineering education. \wss{Can point to other studies to support
-this.} Some of the development teams include computer scientists. Despite a lack
-of visible domain documentation and a resulting lower surface correctness and
-verifiability score, it is clear that some of the software packages were
-developed by teams with significant domain knowledge.
-
-Interviews suggested a more frequent use of both unit testing and continuous
-integration in the development processes than what was observed from the initial
-survey. For example, OpenLB, pyLBM, and TCLB use such methods during development,
-despite this not being explicitly clear from an analysis of the material
-available online. \wss{Add this to threats to validity.}  The correctness and
-verifiability of such packages is not measured well using surface analysis.
-
-Several developers alluded to difficulty with testing the correctness of large
-numbers of features, and some even manually tested program output, as opposed to
-using automated testing. The use of well defined unit testing tools could
-decrease the time spent testing some feature.  \wss{link this to
-recommendations}
-
-\subsection{Usability}
-
-Interviews with developers revealed several usability issues. Some users have
-misunderstood the boundaries of LBM and CFD, and have combined or applied
-methods that are not physically sound. Sometimes users have applied LBM to
-poorly defined or inappropriate fluid dynamics problems. For example, they may
-wish to model flow through or around a structure despite having limited
-information about the structure or its environment, and having little previous
-knowledge of CFD. The users do not realize the limitations of the methods, of
-the software, and do not understand the requirements to properly model a problem
-with the software. As the developer of TCLB noted, such software packages are
-not designed to be used ``out of the box'' in a plug and play fashion, and it
-could take months or more to set up the CFD problems correctly. Developers of
-some software packages, including ESPResSo, mitigated this by editing the source
-code to prevent users from ``combining methods that are not physically sound
-together'', and by updating the documentation to better inform users of LBM
-limitations, and of the requirements to properly model appropriate problems,
-including what algorithms and parameters to use.  \wss{Could point to this for
-recommendations related to documentation.  Could point to this for a generative
-approach that won't allow inadmissible combinations, or at least includes an
-explicit warning.}
-
-Usability concerns were also raised for installation, understanding how to
-maneuver the interfaces and understanding how to set up and run models. These
-issues are addressed by various user support models, including frequently asked
-questions sections on the software package websites, user guides, and hardware
-and software requirements specifications.  \wss{Could point to there being room
-for improvement here, especially for requirements specification.}
-
-One software package (ESPResSo) changed some of its scripting language to Python
-to make it more usable. The developer commented that this was ``the biggest step
-in terms of usability over the years'', further commenting that ``most people in
-the field know [Python]'' and that ``it's easy to learn''.  \wss{A generative
-approach facilitates switching to whatever language is currently popular.}
-
-\subsection{Maintainability}
-
-Although not always visible from the publicly available artifacts, interviews
-with developers revealed that most projects have a defined process for accepting
-contributions from team members. The packages rarely get contributions from
-outside developers, but the process would be similar as for the aforementioned
-group.  In general, contributions are made through GitHub, and are then reviewed
-and pulled by lead developers, often with consultation with a group of core
-developers depending on the organizational model. Continuous integration is part
-of the process for some packages. \wss{Recommendation to increase.  Successful
-projects use it, a reasonable next step for projects that have not yet
-incorporated CI.}
-
-Some developers noted that their software package does not have well defined
-contributing guide in the repository, but it might be a good idea to add one in
-the near future. They would be happy to see contribution from outside of their
-organization, but currently this does not happen.  \wss{A comment was made at
-the recent DOE workshop that you get new users through documentation and
-tutorials.  Maybe highlight the example of Musubi?  Should trace back to this
-for the recommendations part.}
-
-Furthermore, maintainability has been addressed by increasing source code
-modularity, reducing duplicate information, and improving abstraction by
-developing well defined interfaces. This was noted by the developers of ESPResSo
-and pyLBM. Several software packages have had sections of their code base
-redeveloped with languages that the developers felt are more understandable and
-readable, and that are better supported, such as Python. Data structures have
-also been redeveloped and storage has been improved. A developer of pyLBM
-mentioned that the geometries and models of their system had been ``decoupled'',
-using abstraction and modularization of the source code, to make it ``very easy
-to add [new] features''.  \wss{Highlights the advantage of design.  Other
-projects can look to these examples.  Maybe cite the Anshu paper about Flash?
-Dune?  and point to the advantages of design.}
-
-The developer of TCLB mentioned that their package had two sets of code, for
-executing the models on the CPU and GPU, and that maintenance was decreased by
-introducing macros, a practice which became a common part of the development
-process.
-
-\subsection{Modifiability}
-
-Software packages were not quantitatively measured for modifiability. In this
-project we asked developers to comment on modifiability when we interviewed
-them. Specifically, we asked if ease of future changes to the system, modules,
-and code blocks was considered when designing the software. We also asked if any
-measures had been taken to ensure the ease of future changes. All of the
-developers that were interviewed noted that the ease of future changes was
-considered and that measures to ensure it had been taken, including requiring
-the separation of software components in the source code architecture.  \wss{A
-good approach for other developers to adopt - part of recommendations.  Trace
-back to this comment.}  A high degree of code modularity and abstraction was
-noted by developers as a measure to ensure the ease of future changes. This can
-be ensured by separating components and hiding information behind well defined
-interfaces. The developer of ESPResSo also noted that some of the code base was
-transitioned from C to C++, which could ease modifiability of that software
-package. The developer of TCLB noted that their software package was designed to
-allow for the addition of some LBM features, but changes to major aspects of the
-system would be difficult. For example, ``implementing a new model will be an
-easy contribution'', but changes to the ``Cartesian mesh...will be a
-nightmare''. Furthermore, the package was designed with flexible data structures
-and storage in mind.
-
-Some software packages, like Palabos, provide validation benchmarks for their
-core fundamental algorithmic ingredients \citep{latt2021palabos}. The stated
-intent of these benchmarks is to showcase the validity and usefulness of the
-package to stimulate the development of third-user extensions. The Palabos
-package identifies as a development framework for modeling problems in various
-CFD areas.  \wss{Add validation benchmarks as a ``best practice'' used by
-Palabos - others could emmulate.  Goal of this exercise is for people in the
-community to learn from each other, and potentially from our advice too.}
-
-Developers were asked to comment on the obstacles in their development process.
-The developer of ESPResSo noted that a lot of their source code had been written
-with a specific application in mind, and that there is too much coupling between
-components. Addressing this issue, possibly by making the design review process
-more transparent (Section~\ref{secVisib}), would help with code modifiability
-and reusability.  \wss{Traceback for recommendations.}
-
-\subsection{Understandability}
-
-Software developers noted that they believe users have generally found their
-packages to be understandable. \wss{Confusion between understandability and
-usability - threat to validity example.}  The interviewed developer of ESPResSo
-commented that some users have attempted to run physically incompatible LBM
-methods, and the solution was to edit the code to prevent such combinations, as
-well as to update the documentation to prevent misunderstanding the methods.
-Similarly, a developer of pyLBM noted that some users had issues setting up
-parameters for LBM schemes. The solution to this was to update the interface
-where these parameters are set, as well as to add functionality to test the
-stability of the parameters. A developer of OpenLB noted that some users lack
-the background knowledge to easily model fluid dynamics problems using their
-software. A frequently asked questions section was added to their package
-website to help users find answers to common questions. The package also has
-detailed documentation, including guides and usage requirements specification,
-to better help users understand the software. \wss{Peter - the discussion here
-is more about user usability than the understandability of the code for
-developers. Did any of the developers discuss developer understandability?  The
-confusion between understandability and usability was also a confusion during
-the interviews Ao conducted.} \pmm{The developer of ESPResSo did comment on
-developer understandability, specifically for developers that are/were part of
-their peer review process. He commented that code architecture is documented,
-and that code is made to be readable, and that these help peer reviews of the
-code.}
-
-\wss{I moved this from the visibility section, since it is more about
-understandability.}  Developers were also asked how documentation fits into their
-development process. Several developers noted that developer documentation plays
-an important role in familiarizing potential contributors to the software system
-architecture. Without the guidance that the documentation provides it would be
-unlikely that contributions would pass the peer review process.
-
-\subsection{Traceability}
-
-Software packages were not quantitatively measured for traceability. In this
-project we asked developers to comment on traceability, specifically on their
-software package's documentation and how it fit into their development process.
-
-The interviewed developer of ESPResSo noted that all major additions to their
-package had accompanying changes to artifacts and documentation. They noted that
-considerable effort had been put into the documentation. They further commented
-that they want to lower the entry barrier for new developers, and because of
-that their package has a considerable amount of developer documentation. This
-documentation informs developers on how to get started, and orients them to the
-artifacts, source code, and system architecture, as well as how the software
-package build system works, and how the coupling between the simulation engine
-and the interface works.  \wss{Peter - these points about developer
-documentation seem like they would fit better under understandability, don't you
-think?}  \pmm{After reviewing the definitions of the two qualities, I think you
-are right. I think that the point presented in the first sentence of this
-paragraph belongs here in traceability (but could be reworded to note that
-ESPResSo also has detailed change logs in their docs - change logs are also
-available for Palabos and waLBerla), but the rest of the paragraph can be moved
-to understandability.} \wss{If ESPresSo scored high (I think it did), we can use
-that to further promote the value of documentation.  Also a threat to validity
-is that we have implicitly assumed documentation is valuable when measuring
-quality.}
-
-Developers noted the importance of documentation for both the users and
-developers of their software. New features are always added to the
-documentation. The developers use documentation to stay up to date on the status
-of the software package, and to help expand features, like computational models
-or algorithms. This is necessary so that the coding standard for these models is
-kept consistent for new developers. \wss{Peter - this also sounds like
-discussion that is relevant for understandability.} \pmm{The points here sound
-like they could work for both. The definition of traceability notes that `This
-quality refers to keeping track of information as it changes forms or relates
-between artifacts'. So the point `New features are always added to the
-documentation' could be reworded to fit that part of the definition of
-traceability. As noted above, this is similar to the point about some packages
-having including detailed change logs. Same with `The developers use
-documentation to stay up to date on the status of the software package, and to
-help expand features'. I agree that at least some of this paragraph could be in
-understandability, especially the first and second sentences.}
-
-The importance of documentation for both users and developers was stressed
-throughout the interviews. However, it was noted several times that a lack of
-time and funding has a negative affect on the documentation. Most of the
-developers are scientific researchers evaluated on the scientific papers that
-they produce. Writing and updating documentation is something that is done in
-their free time, if that time arises. Sometimes it is a last priority for the
-developers. Finding ways to hasten updating documentation would increase the
-frequency of such updates and benefit both users and developers.  \wss{Same
-thing has come up in papers.  I'll try to find a reference for this.  This is
-also something we can point to for recommendations related to code generation.}
-\wss{Peter - this also sounds like understandability related discussion?  I
-wonder if we should drop Traceability as a quality that we assessed?  There
-doesn't seem to be much content specific to traceability.}  \pmm{Based on the
-comments I just added above, I think we can keep at least a small part of what
-is here in traceability, rewording parts to highlight the importance of change
-logs. Some of the content in this section should indeed be moved to
-understandability.} The developer of OpenLB noted the use of documentation
-generators like Doxygen. It would be advisable for more projects to use such
-automatic document generation tools, since some projects do not do this.
-\wss{We can refer back to this for the recommendations.}
-
-\subsection{Visibility and Transparency} \label{secVisib}
-
-For the developers interviewed, visibility is not currently a priority.
-Visibility is not particularly necessary because the teams are fairly small.
-Although they accept outside contributors, these are rare and the teams remain
-tight-knit.  Most of the teams interviewed have all members at the same
-institution, although one team is spread out internationally. \wss{Peter, can
-you say which team is international, so that we can add this detail to the
-paper?} \pmm{The interviewed developer of OpenLB noted that there were tasks for
-local and international developers when discussing their software development
-model. He also noted that they have frequent local meetings and annual
-international meetings.}  For teams of only a few developers, new features or
-major changes are discussed with the entire team. Projects with more than a few
-developers have lead developer roles. These lead developers review potential
-additions to the software.
-
-Although visibility is not a priority, positive steps are being undertaken. For
-instance, all interviewed teams used GitHub, with its visibility enhancing
-features of branching, issue tracking, insights on contributors, the blame
-feature, wikis, pull requests, etc. Continuous integration, as used by ESPResSo,
-makes projects even more visible, since contributors can explicitly see the bar
-that must be crossed for code to be incorporated into the master branch.
-ESPResSo also uses peer review, which improves visibility by explicitly asking
-for feedback from other developers.
-
-Previously mentioned concerns with maintainability, modifiability, and
-understandability  could be improved by further improving visibility.  For
-future sustainability, projects will likely have to expand their development
-teams by recruiting more external developers. One component of recruiting
-developers is transparency in the development process, including the process for
-making a contribution.  \wss{in recommendations could point out what teams could
-do to get more external contributors, and potentially a longer life for their
-project.}
-
-\subsection{Reproducibility}
-
-Software packages were not quantitatively measured for reproducibility. Instead,
-developers were asked if they had concerns that their computational results
-would not be reproducible in the future, and if they had taken any steps to
-ensure reproducibility.  The developers all agreed that reproducibility is
-important, but they differ in how far they go to support reproducibility.
-ESPResSo does the most to support reproducibility.  For instance, they compare
-the results of their methods against manually calculated results. These
-comparison tests are automatically run for all source code changes. The tests
-are run when a pull request is opened on GitHub. Even once these tests are
-complete, a peer review process is undertaken before changes are fully committed
-to the appropriate branch. The results for all of the LBM schemes on the
-software package development branch are also frequently compared for
-correctness, ensuring that the system output reflects the expected output.
-
-For other teams, the priority is verifying the mathematical foundations of the
-models, but the output of the software package is not compared to other output.
-Depending on the package and how it outputs solutions, this may not be practical
-or feasible. A correct output may not be exactly reproducible, as it may be
-dependent on a probability distribution, so strictly comparing results may not
-be appropriate. \wss{Ideas for dealing with reproducibility in these situations
-will be explored in the recommendations section (like properties of the solution,
-such as conservation of mass).}
-
-For the current project, we did not assess the stronger form of
-``reproducibility'' known as replicability.  A software project is considered
-replicable if the documentation alone (without the code) is sufficiently precise
-and complete to allow a third-part to re-obtain the results of the original code
-\citep{BenureauAndRougier2017}.  Replicability is difficult to achieve with
-current documentation practices.  This question should be considered in the next
-iteration of state of the practice assessments.
-
-\subsection{Unambiguity}
-
-Software packages were not quantitatively measured for unambiguity. We asked
-developers if they thought that the current documentation can clearly convey all
-necessary knowledge to the users, and if they had taken any steps to ensure
-clarity. The developer of ESPResSo noted that their documentation was meant for
-users that are already familiar with the underlying physics and CFD methods.
-These concepts are not explained in detail within the documentation. Users
-should acquire this knowledge from suitable external sources. The documentation
-focuses on how to technically use the software package, and includes a user
-guide and tutorial walk through of how to set up and run a simulation. With this
-in mind, the developer believes that their documentation is in reasonable shape
-for users with a minimum knowledge of the underlying physics. If new users have
-technical questions these can be addressed in further revisions of the
-documentation. New developers should find that the documentation is reasonably
-clear and useful. Information that is missing, like detailed explanations of
-dependencies, is referenced in the documents.
-
-The developer of pyLBM also noted that their documentation was in reasonable
-shape, but that they ``need more [user] feedback to improve [it]''. They also
-noted that they believe a lack of knowledge of the underlying physics and CFD
-concepts can be an issue for some users. This information can be referenced in
-the documentation, but it is not something that the documentation needs to
-detail.
-
-\section{Answers To Research Questions} \label{answersquestions}
-
-This section answers the research questions listed in Section~\ref{secIntro}
-using the quantitative data presented in Section~\ref{AHPresults}, qualitative
-data presented in Section~\ref{interviewresults}, additional data from software
-package repositories and artifacts, and domain expert feedback. Each subsection
-answers one of the research questions. Software package artifacts are examined
-in Section~\ref{artifacts}, tools are listed in Section~\ref{tools}, development
-principles, processes, and methodologies are discussed in Section
-~\ref{prinprocmeth}, development pain points are explored in Section
-~\ref{painpoints}, and summary of current (Section~\ref{currentPractices}), best
-(Section~\ref{bestPractices}) and future practices
-(Section~\ref{futurePractices}). Finally, threats to the validity of this
-assessment are noted in Section~\ref{threats}.
-
-\subsection{What artifacts are present in current software projects?}
-\label{artifacts}
-
-The software packages were examined for the presence of artifacts, which were
-then categorized by frequency. We have grouped them into common, less common,
-and rare artifacts in Table~\ref{artifactspresent}. Common artifacts were found
-in 16 to 24 ($>$63\%) of the software packages. Less common artifacts were found
-in 8 to 15 (30-63\%) of the software packages. Rare artifacts were found in 1 to
-7 ($<$30\%) of the software packages.
+\section{Comparison of Artifacts to Research Software in General}
+\label{Sec_CompareArtifacts}
+
+Answers \rqref{RQ_CompareArtifacts}.  The software packages were examined for
+the presence of artifacts, which were then categorized by frequency. We have
+grouped them into common, less common, and rare artifacts in
+Table~\ref{artifactspresent}. Common artifacts were found in 16 to 24 ($>$63\%)
+of the software packages. Less common artifacts were found in 8 to 15 (30-63\%)
+of the software packages. Rare artifacts were found in 1 to 7 ($<$30\%) of the
+software packages.
 
@@
 
-\subsection{What tools (development, dependencies, project management) are used
-by current software packages?} \label{tools}
-
-Software tools are used to support the development, verification, maintenance,
-and evolution of software, software processes, and artifacts
-\citep{ghezzi1991fundamentals}. Many tools are used by LBM software packages, as
-summarized in Table~\ref{tbl_tools}.  The tools are subdivided into development
-tools, dependencies, and project management tools.
+\section{Comparison of Tool Usage to Research Software in General}
+\label{Sec_CompareTools}
+
+Answers \rqref{RQ_CompareToolsProjMngmnt}.  Software tools are used to support
+the development, verification, maintenance, and evolution of software, software
+processes, and artifacts \citep{ghezzi1991fundamentals}. Many tools are used by
+LBM software packages, as summarized in Table~\ref{tbl_tools}.  The tools are
+subdivided into development tools, dependencies, and project management tools.
 
@@
 
-\subsection{What principles, processes, and methodologies are used?}
-\label{prinprocmeth}
-
-Most of the software packages do not explicitly state in their artifacts the
-motivations or design principles that were considered when developing the
-software. One package, Sailfish, indicates in its artifacts that shortening the
-development time was considered in early stages of design, with the developers
-opting for using Python and CUDA/OpenCL to achieve this without sacrificing any
-computational performance. The goals of that project are explicitly listed as
-performance, scalability, agility and extendability, maintenance, and ease of
-use. The project scored well in these categories during our assessment.
+\section{Comparison of Principles, Process and Methodologies to Research Software in General} \label{Sec_CompareMethodologies}
+
+Answers \rqref{RQ_CompareMethodologies}.  Most of the software packages do not
+explicitly state in their artifacts the motivations or design principles that
+were considered when developing the software. One package, Sailfish, indicates
+in its artifacts that shortening the development time was considered in early
+stages of design, with the developers opting for using Python and CUDA/OpenCL to
+achieve this without sacrificing any computational performance. The goals of
+that project are explicitly listed as performance, scalability, agility and
+extendability, maintenance, and ease of use. The project scored well in these
+categories during our assessment.
 
@@
 
-\subsection{What are the pain points?} \label{painpoints}
-
-This subsection answers the research question: What are the pain points for
-developers working on research software projects? What aspects of the existing
-processes, methodologies and tools do they consider as potentially needing
-improvement? How should processes, methodologies and tools be changed to improve
-software development and software quality?  Developers were asked to comment on
-obstacles in their development process, obstacles encountered by users, and
-potential future obstacles.
+\section{Developer Pain Points} \label{painpoints}
+
+Answers \rqref{RQ_PainPoints} and \rqref{RQ_ComparePainPoints}.  This subsection
+answers the research question: What are the pain points for developers working
+on research software projects? What aspects of the existing processes,
+methodologies and tools do they consider as potentially needing improvement? How
+should processes, methodologies and tools be changed to improve software
+development and software quality? Developers were asked to comment on obstacles
+in their development process, obstacles encountered by users, and potential
+future obstacles.
 
@@
 that is further discussed in Section~\ref{highlightedrecommendations}.
+
+\section{LBM Developer Identified Concerns} \label{Sec_Concerns}
+
+Answers \rqref{RQ_Concerns}.  This section presents qualitative findings on
+software qualities as gleaned from interviewing 4 developers. The full interview
+questions are found in \citet{SmithEtAl2021}.
+
+\subsection{Correctness and Verifiability}
+
+Interviews with developers confirmed that these software packages are developed
+by domain experts with backgrounds in physics, mathematics and mechanical
+engineering. \wss{Could point to surveys that have shown the same thing.}  It
+was noted in these interviews that some of the developers do not have formal
+software engineering education. \wss{Can point to other studies to support
+this.} Some of the development teams include computer scientists. Despite a lack
+of visible domain documentation and a resulting lower surface correctness and
+verifiability score, it is clear that some of the software packages were
+developed by teams with significant domain knowledge.
+
+Interviews suggested a more frequent use of both unit testing and continuous
+integration in the development processes than what was observed from the initial
+survey. For example, OpenLB, pyLBM, and TCLB use such methods during development,
+despite this not being explicitly clear from an analysis of the material
+available online. \wss{Add this to threats to validity.}  The correctness and
+verifiability of such packages is not measured well using surface analysis.
+
+Several developers alluded to difficulty with testing the correctness of large
+numbers of features, and some even manually tested program output, as opposed to
+using automated testing. The use of well defined unit testing tools could
+decrease the time spent testing some feature.  \wss{link this to
+recommendations}
+
+\subsection{Usability}
+
+Interviews with developers revealed several usability issues. Some users have
+misunderstood the boundaries of LBM and CFD, and have combined or applied
+methods that are not physically sound. Sometimes users have applied LBM to
+poorly defined or inappropriate fluid dynamics problems. For example, they may
+wish to model flow through or around a structure despite having limited
+information about the structure or its environment, and having little previous
+knowledge of CFD. The users do not realize the limitations of the methods, of
+the software, and do not understand the requirements to properly model a problem
+with the software. As the developer of TCLB noted, such software packages are
+not designed to be used ``out of the box'' in a plug and play fashion, and it
+could take months or more to set up the CFD problems correctly. Developers of
+some software packages, including ESPResSo, mitigated this by editing the source
+code to prevent users from ``combining methods that are not physically sound
+together'', and by updating the documentation to better inform users of LBM
+limitations, and of the requirements to properly model appropriate problems,
+including what algorithms and parameters to use.  \wss{Could point to this for
+recommendations related to documentation.  Could point to this for a generative
+approach that won't allow inadmissible combinations, or at least includes an
+explicit warning.}
+
+Usability concerns were also raised for installation, understanding how to
+maneuver the interfaces and understanding how to set up and run models. These
+issues are addressed by various user support models, including frequently asked
+questions sections on the software package websites, user guides, and hardware
+and software requirements specifications.  \wss{Could point to there being room
+for improvement here, especially for requirements specification.}
+
+One software package (ESPResSo) changed some of its scripting language to Python
+to make it more usable. The developer commented that this was ``the biggest step
+in terms of usability over the years'', further commenting that ``most people in
+the field know [Python]'' and that ``it's easy to learn''.  \wss{A generative
+approach facilitates switching to whatever language is currently popular.}
+
+\subsection{Maintainability}
+
+Although not always visible from the publicly available artifacts, interviews
+with developers revealed that most projects have a defined process for accepting
+contributions from team members. The packages rarely get contributions from
+outside developers, but the process would be similar as for the aforementioned
+group.  In general, contributions are made through GitHub, and are then reviewed
+and pulled by lead developers, often with consultation with a group of core
+developers depending on the organizational model. Continuous integration is part
+of the process for some packages. \wss{Recommendation to increase.  Successful
+projects use it, a reasonable next step for projects that have not yet
+incorporated CI.}
+
+Some developers noted that their software package does not have well defined
+contributing guide in the repository, but it might be a good idea to add one in
+the near future. They would be happy to see contribution from outside of their
+organization, but currently this does not happen.  \wss{A comment was made at
+the recent DOE workshop that you get new users through documentation and
+tutorials.  Maybe highlight the example of Musubi?  Should trace back to this
+for the recommendations part.}
+
+Furthermore, maintainability has been addressed by increasing source code
+modularity, reducing duplicate information, and improving abstraction by
+developing well defined interfaces. This was noted by the developers of ESPResSo
+and pyLBM. Several software packages have had sections of their code base
+redeveloped with languages that the developers felt are more understandable and
+readable, and that are better supported, such as Python. Data structures have
+also been redeveloped and storage has been improved. A developer of pyLBM
+mentioned that the geometries and models of their system had been ``decoupled'',
+using abstraction and modularization of the source code, to make it ``very easy
+to add [new] features''.  \wss{Highlights the advantage of design.  Other
+projects can look to these examples.  Maybe cite the Anshu paper about Flash?
+Dune?  and point to the advantages of design.}
+
+The developer of TCLB mentioned that their package had two sets of code, for
+executing the models on the CPU and GPU, and that maintenance was decreased by
+introducing macros, a practice which became a common part of the development
+process.
+
+\subsection{Modifiability}
+
+Software packages were not quantitatively measured for modifiability. In this
+project we asked developers to comment on modifiability when we interviewed
+them. Specifically, we asked if ease of future changes to the system, modules,
+and code blocks was considered when designing the software. We also asked if any
+measures had been taken to ensure the ease of future changes. All of the
+developers that were interviewed noted that the ease of future changes was
+considered and that measures to ensure it had been taken, including requiring
+the separation of software components in the source code architecture.  \wss{A
+good approach for other developers to adopt - part of recommendations.  Trace
+back to this comment.}  A high degree of code modularity and abstraction was
+noted by developers as a measure to ensure the ease of future changes. This can
+be ensured by separating components and hiding information behind well defined
+interfaces. The developer of ESPResSo also noted that some of the code base was
+transitioned from C to C++, which could ease modifiability of that software
+package. The developer of TCLB noted that their software package was designed to
+allow for the addition of some LBM features, but changes to major aspects of the
+system would be difficult. For example, ``implementing a new model will be an
+easy contribution'', but changes to the ``Cartesian mesh...will be a
+nightmare''. Furthermore, the package was designed with flexible data structures
+and storage in mind.
+
+Some software packages, like Palabos, provide validation benchmarks for their
+core fundamental algorithmic ingredients \citep{latt2021palabos}. The stated
+intent of these benchmarks is to showcase the validity and usefulness of the
+package to stimulate the development of third-user extensions. The Palabos
+package identifies as a development framework for modeling problems in various
+CFD areas.  \wss{Add validation benchmarks as a ``best practice'' used by
+Palabos - others could emmulate.  Goal of this exercise is for people in the
+community to learn from each other, and potentially from our advice too.}
+
+Developers were asked to comment on the obstacles in their development process.
+The developer of ESPResSo noted that a lot of their source code had been written
+with a specific application in mind, and that there is too much coupling between
+components. Addressing this issue, possibly by making the design review process
+more transparent (Section~\ref{secVisib}), would help with code modifiability
+and reusability.  \wss{Traceback for recommendations.}
+
+\subsection{Understandability}
+
+Software developers noted that they believe users have generally found their
+packages to be understandable. \wss{Confusion between understandability and
+usability - threat to validity example.}  The interviewed developer of ESPResSo
+commented that some users have attempted to run physically incompatible LBM
+methods, and the solution was to edit the code to prevent such combinations, as
+well as to update the documentation to prevent misunderstanding the methods.
+Similarly, a developer of pyLBM noted that some users had issues setting up
+parameters for LBM schemes. The solution to this was to update the interface
+where these parameters are set, as well as to add functionality to test the
+stability of the parameters. A developer of OpenLB noted that some users lack
+the background knowledge to easily model fluid dynamics problems using their
+software. A frequently asked questions section was added to their package
+website to help users find answers to common questions. The package also has
+detailed documentation, including guides and usage requirements specification,
+to better help users understand the software. \wss{Peter - the discussion here
+is more about user usability than the understandability of the code for
+developers. Did any of the developers discuss developer understandability?  The
+confusion between understandability and usability was also a confusion during
+the interviews Ao conducted.} \pmm{The developer of ESPResSo did comment on
+developer understandability, specifically for developers that are/were part of
+their peer review process. He commented that code architecture is documented,
+and that code is made to be readable, and that these help peer reviews of the
+code.}
+
+\wss{I moved this from the visibility section, since it is more about
+understandability.}  Developers were also asked how documentation fits into their
+development process. Several developers noted that developer documentation plays
+an important role in familiarizing potential contributors to the software system
+architecture. Without the guidance that the documentation provides it would be
+unlikely that contributions would pass the peer review process.
+
+\subsection{Traceability}
+
+Software packages were not quantitatively measured for traceability. In this
+project we asked developers to comment on traceability, specifically on their
+software package's documentation and how it fit into their development process.
+
+The interviewed developer of ESPResSo noted that all major additions to their
+package had accompanying changes to artifacts and documentation. They noted that
+considerable effort had been put into the documentation. They further commented
+that they want to lower the entry barrier for new developers, and because of
+that their package has a considerable amount of developer documentation. This
+documentation informs developers on how to get started, and orients them to the
+artifacts, source code, and system architecture, as well as how the software
+package build system works, and how the coupling between the simulation engine
+and the interface works.  \wss{Peter - these points about developer
+documentation seem like they would fit better under understandability, don't you
+think?}  \pmm{After reviewing the definitions of the two qualities, I think you
+are right. I think that the point presented in the first sentence of this
+paragraph belongs here in traceability (but could be reworded to note that
+ESPResSo also has detailed change logs in their docs - change logs are also
+available for Palabos and waLBerla), but the rest of the paragraph can be moved
+to understandability.} \wss{If ESPresSo scored high (I think it did), we can use
+that to further promote the value of documentation.  Also a threat to validity
+is that we have implicitly assumed documentation is valuable when measuring
+quality.}
+
+Developers noted the importance of documentation for both the users and
+developers of their software. New features are always added to the
+documentation. The developers use documentation to stay up to date on the status
+of the software package, and to help expand features, like computational models
+or algorithms. This is necessary so that the coding standard for these models is
+kept consistent for new developers. \wss{Peter - this also sounds like
+discussion that is relevant for understandability.} \pmm{The points here sound
+like they could work for both. The definition of traceability notes that `This
+quality refers to keeping track of information as it changes forms or relates
+between artifacts'. So the point `New features are always added to the
+documentation' could be reworded to fit that part of the definition of
+traceability. As noted above, this is similar to the point about some packages
+having including detailed change logs. Same with `The developers use
+documentation to stay up to date on the status of the software package, and to
+help expand features'. I agree that at least some of this paragraph could be in
+understandability, especially the first and second sentences.}
+
+The importance of documentation for both users and developers was stressed
+throughout the interviews. However, it was noted several times that a lack of
+time and funding has a negative affect on the documentation. Most of the
+developers are scientific researchers evaluated on the scientific papers that
+they produce. Writing and updating documentation is something that is done in
+their free time, if that time arises. Sometimes it is a last priority for the
+developers. Finding ways to hasten updating documentation would increase the
+frequency of such updates and benefit both users and developers.  \wss{Same
+thing has come up in papers.  I'll try to find a reference for this.  This is
+also something we can point to for recommendations related to code generation.}
+\wss{Peter - this also sounds like understandability related discussion?  I
+wonder if we should drop Traceability as a quality that we assessed?  There
+doesn't seem to be much content specific to traceability.}  \pmm{Based on the
+comments I just added above, I think we can keep at least a small part of what
+is here in traceability, rewording parts to highlight the importance of change
+logs. Some of the content in this section should indeed be moved to
+understandability.} The developer of OpenLB noted the use of documentation
+generators like Doxygen. It would be advisable for more projects to use such
+automatic document generation tools, since some projects do not do this.
+\wss{We can refer back to this for the recommendations.}
+
+\subsection{Visibility and Transparency} \label{secVisib}
+
+For the developers interviewed, visibility is not currently a priority.
+Visibility is not particularly necessary because the teams are fairly small.
+Although they accept outside contributors, these are rare and the teams remain
+tight-knit.  Most of the teams interviewed have all members at the same
+institution, although one team is spread out internationally. \wss{Peter, can
+you say which team is international, so that we can add this detail to the
+paper?} \pmm{The interviewed developer of OpenLB noted that there were tasks for
+local and international developers when discussing their software development
+model. He also noted that they have frequent local meetings and annual
+international meetings.}  For teams of only a few developers, new features or
+major changes are discussed with the entire team. Projects with more than a few
+developers have lead developer roles. These lead developers review potential
+additions to the software.
+
+Although visibility is not a priority, positive steps are being undertaken. For
+instance, all interviewed teams used GitHub, with its visibility enhancing
+features of branching, issue tracking, insights on contributors, the blame
+feature, wikis, pull requests, etc. Continuous integration, as used by ESPResSo,
+makes projects even more visible, since contributors can explicitly see the bar
+that must be crossed for code to be incorporated into the master branch.
+ESPResSo also uses peer review, which improves visibility by explicitly asking
+for feedback from other developers.
+
+Previously mentioned concerns with maintainability, modifiability, and
+understandability  could be improved by further improving visibility.  For
+future sustainability, projects will likely have to expand their development
+teams by recruiting more external developers. One component of recruiting
+developers is transparency in the development process, including the process for
+making a contribution.  \wss{in recommendations could point out what teams could
+do to get more external contributors, and potentially a longer life for their
+project.}
+
+\subsection{Reproducibility}
+
+Software packages were not quantitatively measured for reproducibility. Instead,
+developers were asked if they had concerns that their computational results
+would not be reproducible in the future, and if they had taken any steps to
+ensure reproducibility.  The developers all agreed that reproducibility is
+important, but they differ in how far they go to support reproducibility.
+ESPResSo does the most to support reproducibility.  For instance, they compare
+the results of their methods against manually calculated results. These
+comparison tests are automatically run for all source code changes. The tests
+are run when a pull request is opened on GitHub. Even once these tests are
+complete, a peer review process is undertaken before changes are fully committed
+to the appropriate branch. The results for all of the LBM schemes on the
+software package development branch are also frequently compared for
+correctness, ensuring that the system output reflects the expected output.
+
+For other teams, the priority is verifying the mathematical foundations of the
+models, but the output of the software package is not compared to other output.
+Depending on the package and how it outputs solutions, this may not be practical
+or feasible. A correct output may not be exactly reproducible, as it may be
+dependent on a probability distribution, so strictly comparing results may not
+be appropriate. \wss{Ideas for dealing with reproducibility in these situations
+will be explored in the recommendations section (like properties of the solution,
+such as conservation of mass).}
+
+For the current project, we did not assess the stronger form of
+``reproducibility'' known as replicability.  A software project is considered
+replicable if the documentation alone (without the code) is sufficiently precise
+and complete to allow a third-part to re-obtain the results of the original code
+\citep{BenureauAndRougier2017}.  Replicability is difficult to achieve with
+current documentation practices.  This question should be considered in the next
+iteration of state of the practice assessments.
+
+\subsection{Unambiguity}
+
+Software packages were not quantitatively measured for unambiguity. We asked
+developers if they thought that the current documentation can clearly convey all
+necessary knowledge to the users, and if they had taken any steps to ensure
+clarity. The developer of ESPResSo noted that their documentation was meant for
+users that are already familiar with the underlying physics and CFD methods.
+These concepts are not explained in detail within the documentation. Users
+should acquire this knowledge from suitable external sources. The documentation
+focuses on how to technically use the software package, and includes a user
+guide and tutorial walk through of how to set up and run a simulation. With this
+in mind, the developer believes that their documentation is in reasonable shape
+for users with a minimum knowledge of the underlying physics. If new users have
+technical questions these can be addressed in further revisions of the
+documentation. New developers should find that the documentation is reasonably
+clear and useful. Information that is missing, like detailed explanations of
+dependencies, is referenced in the documents.
+
+The developer of pyLBM also noted that their documentation was in reasonable
+shape, but that they ``need more [user] feedback to improve [it]''. They also
+noted that they believe a lack of knowledge of the underlying physics and CFD
+concepts can be an issue for some users. This information can be referenced in
+the documentation, but it is not something that the documentation needs to
+detail.
+
+\section{Recommendations for Addressing Developer Concerns} \label{Sec_AddressConcerns}
+
+Answers \rqref{RQ_Recommend}.  This section answers the research questions
+listed in Section~\ref{secIntro} using the quantitative data presented in
+Section~\ref{AHPresults}, qualitative data presented in
+Section~\ref{interviewresults}, additional data from software package
+repositories and artifacts, and domain expert feedback. Each subsection answers
+one of the research questions. Software package artifacts are examined in
+Section~\ref{artifacts}, tools are listed in Section~\ref{tools}, development
+principles, processes, and methodologies are discussed in Section
+~\ref{prinprocmeth}, development pain points are explored in Section
+~\ref{painpoints}, and summary of current (Section~\ref{currentPractices}), best
+(Section~\ref{bestPractices}) and future practices
+(Section~\ref{futurePractices}). Finally, threats to the validity of this
+assessment are noted in Section~\ref{threats}.
 
@@
 
-practices for the future.
+Answers \rqref{RQ_Future}.  practices for the future.
 

### 2022-01-13 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Edits to add information on comparison to other research software

@@
 
+Pain points that others, but not LBM developers, mentioned in surveys include
+\citep{PintoEtAl2018}: Cross-platform compatibility, interruptions while coding,
+scope bloat, lack of user feedback, mismatch between coding skills and
+subject-matter skills (for instance, ``I am a mathematician, not an expert on
+software''), lack of formal reward system, hard to collaborate on software
+projects, aloneness.
+
 \subsubsection{Lack of Development Time}
@@
 
+Lack of time was also mentioned by \citet{PintoEtAl2018} and
+\citet{PintoEtAl2016}.
+
 \subsubsection{Lack of Software Development Experience}
@@
 several developers.  \wss{The Wiese paper and others talk about the background
-of developers.}
+of developers. The knowledge for developing research software primarily comes
+from peers and through self-study, not from formal training
+\citep{HannayEtAl2009}.
+The thinking of the developer on the role of formal education matches the trend \citet{PintoEtAl2018} observes, where in their replication of a previous study \citept{HannayEtAl2009}, they show a growing interest in formal training (From 13\% of respondants in 2009 to 22\% in 2018).}
+
+Poor software documentation is also mentioned by \citet{PintoEtAl2018} and
+\citet{LethbridgeEtAl2003}.  The survey from \citet{LethbridgeEtAl2003} is for
+software in general, not just research software.
 
@@
 \wss{The Weiss paper alludes this this as well.  I think the Kelly and Sanders
-paper likely talks about testing problems and correctness.}
+paper likely talks about testing problems and correctness.  ``while many
+scientists believe that software testing is important, a smaller number believe
+they have sufficient understanding about testing concepts''
+\citep{HannayEtAl2009}.}
 
@@
 
+``unique challenges of scientific research (e.g., the frequent and unforeseen
+changes in requirements ...)'' \citep{PintoEtAl2018}. We can point out that in
+the future this viewpoint could change and why.  Not really a unique challenge.
+Cite Smith 2016?
+
 \section{Threats To Validity} \label{threats}

### 2022-01-18 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Updates to SOP for LBM, focus on comparing to research software in general

@@
     of a set of 20 questions). From the measurement results, we ranked the
-    software using the Analytic Hierarchy Process (AHP), a multi-criteria
-    decision making method appropriate for cases with a mix of qualitative and
-    quantitative factors. Our ranking was roughly consistent with ranking the
-    repositories by the number of GitHub stars, although the number one project
-    by star count (Sailfish) is ranked 9th by our methodology. Our top three
-    packages were ESPResSo, Ludwig, and Palabos (all in the top 10 by star
-    count). We interviewed 4 developers to gain insight into their current pain
-    points. Identified challenges include lack of development time, lack of
-    funding, and difficulty with ensuring correctness.  Based on our assessment,
-    we make the following recommendations for current and future LBM projects:
-    provide a detailed user manual and tutorials, explicitly state the limits of
-    the software, use user-friendly programming languages, consider a peer
-    review process, communicate development and contribution information, and
-    use continuous integration and project management tools.  \wss{Change the
-    abstract to discuss how LBM compares to common practices and best practices.
-    Highlight some future practices.}
+    software using the Analytic Hierarchy Process (AHP). Our ranking was roughly
+    consistent with ranking the repositories by the number of GitHub stars,
+    although the number one project by star count (Sailfish) is ranked 9th by
+    our methodology. Our top three packages were ESPResSo, Ludwig, and Palabos
+    (all in the top 10 by star count). We interviewed 4 developers to gain
+    insight into their current pain points. Identified challenges include lack
+    of development time, lack of funding, and difficulty with ensuring
+    correctness.  Based on our assessment, we make the following recommendations
+    for current and future LBM projects: provide a detailed user manual and
+    tutorials, explicitly state the limits of the software, use user-friendly
+    programming languages, consider a peer review process, communicate
+    development and contribution information, and use continuous integration and
+    project management tools.  \wss{Change the abstract to discuss how LBM
+    compares to common practices and best practices. Highlight some future
+    practices.  \wss{Follow the general research software trend of increasing
+    use of version control systems and project management on GitHub or
+    similar.}}
 
@@
 The use of continuous integration tools and techniques alludes to a more refined
-development process where faults are isolated and better recognized.
-Only three of the packages (ESPResSo, Ludwig, Musubi) mentioned applying the practice of continuous integration in their development process.
+development process where faults are isolated and better recognized. Only three
+of the packages (ESPResSo, Ludwig, Musubi) mentioned applying the practice of
+continuous integration in their development process. \wss{Forward reference that
+suggests more tools use CI (Section 7 Comparison of Principles, Process and
+Methodologies to Research Software in General).}
 
@@
 have each of the commonly found artifacts, except only three of them have a
-requirements specification or theory notes. Palabos is the only one of these
-four packages that does not have an artifact from this category.
+requirements specification or theory notes. \wss{Change to clarify that it is
+theory notes, not a requirements specification.  Remove requirements spec from
+table and put it as rare.}  Palabos is the only one of these four packages that
+does not have an artifact from this category.  \wss{** Requirements **}
+\wss{Discuss requirements here. Extra information from Peter: I interpreted it
+to mean what the software should do. As noted in my report, this was not common,
+especially finding any formal document like an SRS. I do not recall any package
+having an SRS. Some packages did note in their manuals or artifacts some of the
+points that would be found in an SRS document. For example, DL\_MESO notes
+functional (``The following is a list of the features that DLMESO currently
+supports....'') and non-functional (should work in both ``serial and parallel
+running; it can be run on standalone machines, clusters and supercomputers..'')
+requirements in its documentation (more on this below). Yes, there was often
+some sort of a theory resource. Often it was simply a link or reference to the
+theory used in the package. There was significant discrepancy in how much theory
+was documented or referenced. Meanwhile, requirements were rare. A few packages
+did note some of this information within other artifacts but it was not
+comprehensive, and it was not presented in a SRS document. DL\_MESO has some of
+the information that belongs in an SRS in its user manual. It is not
+comprehensive or presented as it would be in an SRS, but it was probably the
+best example of requirements. It has a nice subsection on Functionality.
+\citet{Nguyen-HoanEtAl2010} showed requirements are the least commonly produced
+type of documentation for scientific software in general.  LBM matches this
+trend.}
 
@@
 projects, especially bug related issues. Most of the projects appear to use
-change tracking and version control tools. They often use GitHub or GitLab for
-this. One package (SunlightLB) uses CVS. Document generation tools are mentioned
-in the artifacts of 12 of the projects. The tools Sphinx and Doxygen are
-explicitly used in this capacity. \wss{look at what the other papers say are
-commonly used tools and bring that information into this section.}
+change tracking and version control tools. \wss{**Version Control**} \wss{This
+contrasts with work from a little over 10 years ago that shows version control
+was only used in 50\% of research software projects \citep{Nguyen-HoanEtAl2010}.
+At that time \citet{Nguyen-HoanEtAl2010} noted an increase from previous levels
+of version control software use.  Anecdotally this confirms our observations.}
+They often use GitHub or GitLab for this. One package (SunlightLB) uses CVS.
+Document generation tools are mentioned in the artifacts of 12 of the projects.
+The tools Sphinx and Doxygen are explicitly used in this capacity. \wss{look at
+what the other papers say are commonly used tools and bring that information
+into this section.}
 
@@
 \citep{HannayEtAl2009}.
-The thinking of the developer on the role of formal education matches the trend \citet{PintoEtAl2018} observes, where in their replication of a previous study \citept{HannayEtAl2009}, they show a growing interest in formal training (From 13\% of respondants in 2009 to 22\% in 2018).}
+The thinking of the developer on the role of formal education matches the trend \citet{PintoEtAl2018} observes, where in their replication of a previous study \citep{HannayEtAl2009}, they show a growing interest in formal training (From 13\% of respondants in 2009 to 22\% in 2018).}
 
@@
 individual modules and the system to verify correctness. These tests compare the
-package output to known correct output using test cases. The developer of TCLB
-commented that the amount of testing data that is needed for some cases is a
-problem as free testing services do not offer capabilities to store and process
-such large amounts of data, and in-house testing solutions needed to be created
-to address this limitation. The solution for this has been to limit the size of
-the testing problems, and to run tests in small batches with few iterations.
-\wss{The Weiss paper alludes this this as well.  I think the Kelly and Sanders
-paper likely talks about testing problems and correctness.  ``while many
-scientists believe that software testing is important, a smaller number believe
-they have sufficient understanding about testing concepts''
+package output to known correct output using test cases. \wss{oracle}  The
+developer of TCLB commented that the amount of testing data that is needed for
+some cases is a problem as free testing services do not offer capabilities to
+store and process such large amounts of data, and in-house testing solutions
+needed to be created to address this limitation. The solution for this has been
+to limit the size of the testing problems, and to run tests in small batches
+with few iterations. \wss{The Weiss paper alludes this this as well.  I think
+the Kelly and Sanders paper likely talks about testing problems and correctness.
+``while many scientists believe that software testing is important, a smaller
+number believe they have sufficient understanding about testing concepts''
 \citep{HannayEtAl2009}.}
+
+\wss{** BEGIN Testing and Verification **}
+\citet{Nguyen-HoanEtAl2010} says the following:
+\begin{itemize}
+	\item 38 of 60 respondents performed unit tests
+	\item 23 of 60 respondents performed integration tests
+	\item ``In our results, the four most common activities involve comparisons
+	with known, trusted, or prior data. Verification and peer review were three
+	of the four least commonly performed activities.'' \wss{testing is not the
+	only way to verify software.  This addresses the oracle problem.}
+\end{itemize}
+\wss{** END Testing and Verification **}
+
+``The questions of what constitutes an adequate testing oracle for scientific software and what constitutes adequate test data are left open.
+Some of the validation testing we researched was running up against the practical limits; they were using whatever oracle they could scrape together from industry data, benchmarks, and expert judgment.'' \citep{KellyAndSanders2008}
+
+\wss{Seems that LBM not concerned with the oracle problem?}
 

### 2022-01-18 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Revision of text connecting RQs

@@
 of algorithms for simulating single-phase and multiphase fluid flows, often
-incorporating additional physical complexities \citep{chen1998lattice} such as
-reflective and non-reflective boundaries. They consider the behavior of a
-collection of particles as a single unit at the mesoscopic scale, between the
-nanoscopic and microscopic scales. LBM solvers predict the positional
-probability of a collection of particles moving through a lattice structure
-following a two step process: i) streaming, where the particles
-move along the lattice via links; and, ii) colliding, where energy and
-momentum is transferred among particles that collide \citep{bao2011lattice}. As
-an example of the output of LBM, Figure~\ref{circularflow} presents the streamlines of
-converged flow past a stationary circular cylinder with varying Reynolds numbers.
-\wss{Zahra might have a better example?  A journal publication isn't likely
-going to accept a figure produced for another paper.}  LBM has several
-advantages over conventional CFD methods, including a simple calculation
-procedure, improved parallelization, and robust handling of complex geometries
-\citep{ganji2015application}.
+incorporating additional physical complexities \citep{chen1998lattice}, such as
+reflective and non-reflective boundaries. LBM considers the behavior of a
+collection of particles as a single unit at the mesoscopic scale, which lies
+between the nanoscopic and microscopic scales. LBM solvers predict the
+positional probability of a collection of particles moving through a lattice
+structure following a two step process: i) streaming, where the particles move
+along the lattice via links; and, ii) colliding, where energy and momentum is
+transferred among particles that collide \citep{bao2011lattice}. As an example
+of the output of LBM, Figure~\ref{circularflow} presents the streamlines of
+converged flow past a stationary circular cylinder with varying Reynolds
+numbers. \wss{We should ask Zahra and her students for a better example.}  LBM
+has several advantages over conventional CFD methods, including a simple
+calculation procedure, improved parallelization, and robust handling of complex
+geometries \citep{ganji2015application}.
 
@@
 is right to assess the state of the practice.  The goal of this report is to
-analyze the current state of LBM software development to provide insight into
-its best practices and to offer guidance for future development.
-
-To focus our efforts, we devised five research questions, which can be applied
-to LBM, as well as to other domains, of research software \citep{SmithEtAl2021}.
-The first question below refers to artifacts, which is the term we use to
-represent the documents, scripts and code that constitutes a software
-development project. Example artifacts include requirements, specifications,
-user manuals, unit tests, system tests, usability tests, build scripts, API
-(Application Programming Interface) documentation, READMEs, license documents,
-process documents, and code.
+analyze the current state of LBM software development and compare it to the
+development practices employed for research software in general.  We want to
+highlight the success stories that LBM developers can share amongst their
+community and with the broader research software community.  We also want to
+identify areas for future improvement.
+
+To focus our efforts, we devised 11 research questions.  The questions are
+inspired by the questions that underpin our data collection methodology
+\citep{SmithEtAl2021}. To start with, we need to identify the relevant examples
+of LBM software.  Knowing that our data collection methodology \citep{SmithEtAl2021} requires access to source code, we build that constraint into our research question:
 
@@
 	available for all identified projects?
+\end{enumerate}
+
+We next wish to assess the representative software to determine how well they
+apply current software development best practices.  At this point we will base
+our assessment on publicly available artifacts, where artifacts are the
+documents, scripts and code that we find in a project's repository. Example
+artifacts include requirements, specifications, user manuals, unit test cases,
+system tests, usability tests, build scripts, API (Application Programming
+Interface) documentation, READMEs, license documents, process documents, and
+code.  Following best practices does not guarantee popularity, so we will also
+estimate which projects are most often used by the scientists.
+
+\begin{enumerate}
 	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_HighestQuality}:] Which
@@
 	to the most popular projects, as viewed by the scientific community?
+\end{enumerate}
+
+To understand the state of the practice we wish to learn the frequency with
+which different artifacts appear, the types of development tools used and the
+methodologies used for software development.  With this data, we can ask
+questions about how the LBM community compares to other research software.  That
+is, in what ways does LBM follow the trends from other communities, and in which
+ways is it different?
+
+\begin{enumerate}
     \item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareArtifacts}:] How
@@
 	principles, processes and methodologies used?
+\end{enumerate}
+
+Only so much information can be gleaned by looking at software repositories.  To
+gain additional insight, we need to talk to developers.  We need to learn their
+concerns, how they deal with these concerns, and what pain points exist for
+them.  We wish to know what practices are used by the top LBM projects, so that
+others can potentially emulate these practices. We also wish to identify new
+practices by borrowing successful ideas from other domains. For LBM developers
+that may be interested in cutting edge software practices, ideas that are
+proposed, but not yet in practice, should also be identified.  All of this is
+covered by the questions outlined below:
+
+\begin{enumerate}
 	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_PainPoints}:] What are
@@
 
-\subsubsection{Lack of Development Time}
+\subsection{Lack of Development Time}
 
@@
 
-\subsubsection{Lack of Software Development Experience}
+\subsection{Lack of Software Development Experience}
 
@@
 
-\subsubsection{Lack of Incentive and Funding}
+\subsection{Lack of Incentive and Funding}
 
@@
 
-\subsubsection{Lack of External Support}
+\subsection{Lack of External Support}
 
@@
 
-\subsubsection{Parallelization and Continuous Integration}
+\subsection{Parallelization and Continuous Integration}
 
@@
 
-\subsubsection{Ensuring Correctness}
+\subsection{Ensuring Correctness}
 
@@
 \wss{** BEGIN Testing and Verification **}
+
 \citet{Nguyen-HoanEtAl2010} says the following:
@@
 \end{itemize}
+
+May do more testing than artifacts indicate, but that is still a problem.
+Artifacts should be available so that others can verify the results, verify
+their installation.  Part of an assurance case means that a third party can
+verify the work.
+
 \wss{** END Testing and Verification **}
@@
 
-\subsubsection{Usability}
+\subsection{Usability}
 
@@
 
-\subsubsection{Technical Debt}
+\subsection{Technical Debt}
 
@@
 
-\subsubsection{Quality of Documentation}
+\subsection{Quality of Documentation}
 
@@
 Section~\ref{interviewresults}, additional data from software package
-repositories and artifacts, and domain expert feedback. Each subsection answers
-one of the research questions. Software package artifacts are examined in
-Section~\ref{artifacts}, tools are listed in Section~\ref{tools}, development
-principles, processes, and methodologies are discussed in Section
-~\ref{prinprocmeth}, development pain points are explored in Section
-~\ref{painpoints}, and summary of current (Section~\ref{currentPractices}), best
-(Section~\ref{bestPractices}) and future practices
-(Section~\ref{futurePractices}). Finally, threats to the validity of this
-assessment are noted in Section~\ref{threats}.
+repositories and artifacts, and domain expert feedback.
 
@@
 \begin{itemize}
-	\item Installability. Include OS compatibility, including specific OS
-	versions. This was done by several top ranked packages, including ESPResSo
-	and LUMA.
-	\item Installability. Provide complete installation instructions. This was
-	done by all of the top five ranked packages.
-	\item Installability. Installation instructions should be written as if the
-	user does not have any dependencies installed and is installing on a clean
-	OS. This was done by several top ranked packages, including ESPResSo,
-	OpenLB, and Palabos. This can be tested on a clean environment using a
-	virtual machine.
-	\item Installability. List all dependencies in the installation
-	instructions, dependency versions, and how to install them. This was done by
-	several top ranked packages, including ESPResSo, LUMA, OpenLB, and Palabos.
-	\item Installability. If possible, automate the installation of
-	dependencies. Use tools such as Make. This was done by all of the top five
-	ranked packages.
-	\item Installability. Automate the installation process as much as possible.
-	Use tools such as Make. This was done by all of the top five ranked
-	packages.
-	\item Installability. The installation instruction should only be in one
-	location. This was done by several top ranked packages, including ESPResSo,
-	OpenLB, and Palabos.
-	\item Installability. Include descriptive error messages for errors
-	encountered during installation. This was done by several top ranked
-	packages, including Ludwig and LUMA.
-	\item Installability. Provide a way to validate the installation. This can
-	be done using a custom script, or a test case that specifies expected
-	output. This was done by several top ranked packages, including OpenLB,
-	Ludwig, LUMA, and Palabos.
-	\item Installability. Provide instructions for uninstalling the software.
-	These were included with pyLBM.
+	\item The top projects provide great examples of achieving installability,
+	by taking such steps as: providing complete installation instructions,
+	writing instructions as if the user does not have any of the dependencies
+	installed, automation of the installation process (for instance using make),
+	including descriptive error message, and including steps for uninstallation.
 	\item Surface Correctness and Verifiability. Make public (on package website

### 2022-01-20 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Revisions completed for Section 1 (Introduction)

@@
 converged flow past a stationary circular cylinder with varying Reynolds
-numbers. \wss{We should ask Zahra and her students for a better example.}  LBM
-has several advantages over conventional CFD methods, including a simple
-calculation procedure, improved parallelization, and robust handling of complex
-geometries \citep{ganji2015application}.
+numbers. \wss{Zahra can likely provide a better example.}  LBM has several
+advantages over conventional CFD methods, including a simple calculation
+procedure, improved parallelization, and robust handling of complex geometries
+\citep{ganji2015application}.
 
@@
 reliability, robustness, reproducibility, performance, correctness and
-verifiability.  Given the value of their time, developers that create or modify
+verifiability.  Since their time is valuable, developers that create or modify
 LBM libraries will have additional quality concerns, including maintainability,
 reusability, and understandability.  With the quality considerations in mind,
-and the potentially overwhelming number of choices for LBM libraries, the time
-is right to assess the state of the practice.  The goal of this report is to
-analyze the current state of LBM software development and compare it to the
-development practices employed for research software in general.  We want to
-highlight the success stories that LBM developers can share amongst their
-community and with the broader research software community.  We also want to
-identify areas for future improvement.
-
-To focus our efforts, we devised 11 research questions.  The questions are
-inspired by the questions that underpin our data collection methodology
-\citep{SmithEtAl2021}. To start with, we need to identify the relevant examples
-of LBM software.  Knowing that our data collection methodology \citep{SmithEtAl2021} requires access to source code, we build that constraint into our research question:
+and the potentially overwhelming number of choices for existing LBM libraries,
+the time is right to assess the state of the practice.  The goal of this report
+is to analyze the current state of LBM software development and compare it to
+the development practices employed for research software in general.  We want to
+highlight success stories that LBM developers can share amongst their
+community, and with the broader research software community, while at the same time watching for areas for potential future improvement.
+
+To focus our efforts, we devised 11 research questions.  The questions are used
+to structure the discussion in the paper, so for each research question below we
+point to the section that contains our answer.  The questions are inspired by
+the research questions that underpin our data collection methodology
+\citep{SmithEtAl2021}. We start with identifying the examples of LBM software
+where we have access to the source code:
 
@@
 	software projects exist, with the constraint that the source code must be
-	available for all identified projects?
+	available for all identified projects? (Section~\ref{methodology})
 \end{enumerate}
@@
 We next wish to assess the representative software to determine how well they
-apply current software development best practices.  At this point we will base
-our assessment on publicly available artifacts, where artifacts are the
-documents, scripts and code that we find in a project's repository. Example
-artifacts include requirements, specifications, user manuals, unit test cases,
-system tests, usability tests, build scripts, API (Application Programming
-Interface) documentation, READMEs, license documents, process documents, and
-code.  Following best practices does not guarantee popularity, so we will also
-estimate which projects are most often used by the scientists.
+apply current software development best practices.  At this point in the
+process, to remove potential user/developer bias, we will base our assessment
+only on publicly available artifacts, where artifacts are the documents, scripts
+and code that we find in a project's public repository. Example artifacts
+include requirements, specifications, user manuals, unit test cases, system
+tests, usability tests, build scripts, API (Application Programming Interface)
+documentation, READMEs, license documents, process documents, and code.
+Following best practices does not guarantee popularity, so we will also compare
+our ranking to how the user community itself ranks the identified projects.
 
@@
 	practices, based on evidence found by experimenting with the software and
-	searching the artifacts available in each project's repository?
+	searching the artifacts available in each project's repository? (Section~\ref{AHPresults})
 	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareHQ2Popular}:] How
 	similar is the list of top projects identified in \rqref{RQ_HighestQuality}
-	to the most popular projects, as viewed by the scientific community?
+	to the most popular projects, as viewed by the scientific community? (Section~\ref{repmetrics})
 \end{enumerate}
@@
 methodologies used for software development.  With this data, we can ask
-questions about how the LBM community compares to other research software.  That
-is, in what ways does LBM follow the trends from other communities, and in which
-ways is it different?
+questions about how LBM software compares to other research software.  That is,
+in what ways does LBM follow the trends from other developer communities, and in
+which ways is it different?
 
@@
     \item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareArtifacts}:] How
-	do LBM projects compare to research software in general with respect to
-	the artifacts present in their repositories?
+	do LBM projects compare to research software in general with respect to the
+	artifacts present in their repositories?
+	(Section~\ref{Sec_CompareArtifacts})
 	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareToolsProjMngmnt}:]
-	How do LBM projects compare to research software in general with
-	respect to the use of tools for:
+	How do LBM projects compare to research software in general with respect to
+	the use of tools (Section~\ref{Sec_CompareTools}) for:
 	\begin{enumerate} 
@@
 	How do LBM projects compare to research software in general with respect to
-	principles, processes and methodologies used?
+	principles, processes and methodologies used? (Section~\ref{Sec_CompareMethodologies})
 \end{enumerate}
@@
 Only so much information can be gleaned by looking at software repositories.  To
-gain additional insight, we need to talk to developers.  We need to learn their
-concerns, how they deal with these concerns, and what pain points exist for
-them.  We wish to know what practices are used by the top LBM projects, so that
-others can potentially emulate these practices. We also wish to identify new
-practices by borrowing successful ideas from other domains. For LBM developers
-that may be interested in cutting edge software practices, ideas that are
-proposed, but not yet in practice, should also be identified.  All of this is
+gain additional insight, we need to interview developers.  We need to learn
+their concerns, how they deal with these concerns, and what pain points exist
+for them.  We wish to know what practices are used by the top LBM projects, so
+that others can potentially emulate these practices. We also wish to identify
+new practices by borrowing successful ideas from other domains. For LBM
+developers interested in future cutting edge practices, ideas that are proposed,
+but not yet in practice, should also be identified.  The above points are
 covered by the questions outlined below:
@@
 	the pain points for developers working on LBM software projects?
+	(Section~\ref{painpoints})
 	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_ComparePainPoints}:] How
 	do the pain points of developers from LBM compare to the pain points
-	for research software in general?
+	for research software in general? (Section~\ref{painpoints})
 	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Concerns}:] For LBM
@@
 	\begin{enumerate} 
-		\item [\rqref{RQ_Concerns}.a] what software development concerns exist; and,
+		\item [\rqref{RQ_Concerns}.a] what software development concerns exist; and, (Section~\ref{Sec_Concerns})
 		\item [\rqref{RQ_Concerns}.b] what specific practices are taken to
-		address these concerns?
+		address these concerns? (Section~\ref{Sec_AddressConcerns})
 	\end{enumerate}
@@
 	by LBM software, could address the concerns identified in
-	\rqref{RQ_Concerns}).
+	\rqref{RQ_Concerns}). (Section~\ref{Sec_AddressConcerns})
 	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Future}:] What practices
 	could be employed by future LBM developers to address the pain points found
-	in \rqref{RQ_PainPoints} and the concerns identified in \rqref{RQ_Concerns}.a?
+	in~\rqref{RQ_PainPoints} and the concerns identified
+	in~\rqref{RQ_Concerns}.a? (Section~\ref{Sec_AddressConcerns})
 
@@
 Oceanographic Software \citep{smith2015state}, Seismology software
-\citep{SmithEtAl2018}, and statistical software for psychology
-\citep{SmithEtAl2018_StatSoft}.  \wss{add citation to medical imaging software,
-if it is an option}
-
-With our methodology, we first choose a software domain (in the current case
-LBM) and identify a list of 24 software packages.  We approximately measure the
-qualities of each package by filling in a grading template. Compared with our
-previous methodology, the new methodology also includes repository based
-metrics, such as the number of files, number of lines of code, percentage of
-issues that are closed, etc.  With the quantitative data in the grading
-template, we rank the software with the Analytic Hierarchy Process (AHP). After
-this, as another addition to our previous methodology, we interview some of the
-development teams to further understand the status of their development process.
-Finally, we summarize the measurements (Section~\ref{AHPresults}), summarize the
-interview answers (Section~\ref{interviewresults}), answer the research
-questions and propose recommendations for LBM software development
-(Section~\ref{answersquestions}).
+\citep{SmithEtAl2018}, statistical software for psychology
+\citep{SmithEtAl2018_StatSoft} and medical image analysis software
+\citep{Dong2021}.  \wss{add citation to medical imaging software, if it is an
+option}
+
+With our methodology we start by identifying 24 LBM software packages.  We then
+approximately measure the application of best practices for each package by
+filling in a grading template. Compared with our previous methodology (as used
+in \citep{SmithEtAl2016} for instance), the new methodology also includes
+repository based metrics, such as the number of files, number of lines of code,
+percentage of issues that are closed, etc.  With the quantitative data in the
+grading template, we rank the software with the Analytic Hierarchy Process
+(AHP). After this, as another addition to our previous methodology, we interview
+some of the development teams to further understand the status of their
+development process. The quantitative and qualitative data is then used to
+answer the research questions (Sections~\ref{methodology}
+to~\ref{Sec_AddressConcerns}).  Finally, we summarize the threats to validity
+(Section~\ref{threats}) and the final conclusions and recommendations
+(Section~\ref{conclusion}).
 
@@
 
-\section{Recommendations for Addressing Developer Concerns} \label{Sec_AddressConcerns}
-
-Answers \rqref{RQ_Recommend}.  This section answers the research questions
-listed in Section~\ref{secIntro} using the quantitative data presented in
+\section{Lessons from LBM Developers} \label{Sec_AddressConcerns}
+
+Answers \rqref{RQ_Recommend}.  Recommendations for Addressing Developer
+Concerns.  This section answers the research questions listed in
+Section~\ref{secIntro} using the quantitative data presented in
 Section~\ref{AHPresults}, qualitative data presented in

### 2022-01-22 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Rearrangement of Section 2 of LBM SOP paper

@@
 We developed a methodology for evaluating the state of the practice of
-scientific software \citep{SmithEtAl2021}.  The methodology aims to answer the
-research questions posed in the previous section. This methodology can be
-instantiated for a specific domain of scientific software, which in the current
-case is LBM software.  The first section below describes the steps of the
-process we used to select, measure and compare LBM software. This is followed by
-quality definitions and how these qualities were assessed in our project. The
-rest of the section provides the details of the methodology.
-% an overview of the specifics of candidate software selection, the quantitative
-% measurements, developer interviews, the ranking process via AHP and
-% interaction with the Domain Expert.
-
-\subsection{Overall Process}
-
-The assessment was conducted via the following steps:
-
-\begin{enumerate}
-	\item List candidate software packages for the domain. This is discussed in
-	Section~\ref{identifysoftware}.
-	\item Filter the software package list. This is discussed in
-	Section~\ref{filtersoftware}.
-	\item Gather the source code and documentation for each software package.
-	\item Collect quantitative measures from the project repositories. This is
-	discussed in Section~\ref{empiricalmeasures}.
-	\item Measure using the measurement template. This is discussed in
-	Section~\ref{empiricalmeasures}. The full measurement template can be found
-	in \citet{SmithEtAl2021}.
-	\item Survey the developers. The developers of each software package in the
-	filtered software list were contacted for voluntary interviews. The details
-	are given in Section~\ref{SecSurvey}.
-	\item Use AHP to rank the software packages. This is discussed in
-	Section~\ref{AHP}.
-	\item Analyze the results and answer the research questions. The answers can
-	be found in Section~\ref{answersquestions}.
-\end{enumerate}
-
-The above steps depend on interaction with a domain expert partner, as discussed
-in Section~\ref{sec_vet_software_list}.
+scientific software \citep{SmithEtAl2021}. This methodology can be instantiated
+for a specific domain of scientific software, which in the current case is LBM
+software.  Since we group best practices around conventional software qualities,
+the first section below provides definitions for the qualities of interest
+(Section~\ref{softwarequalities}), along with information on how we assessed
+these qualities. The following section (Section~\ref{Sec_OverallProcess})
+provides an overview of the steps we used to select, measure and compare LBM
+software.  The remaining sections provide details on each step of the steps in
+the methodology. 
+% of the next sections provide an overview of the specifics of identifying
+% (Section~\ref{identifysoftware}) and filtering (Section~\ref{filtersoftware})
+% the candidate software, our quantitative measurements
+% (Section~\ref{empiricalmeasures}), the AHP ranking process
+% (Section~\ref{AHP}), developer interviews (Section~\ref{SecSurvey}), and
+% interactions with one of the key members of the assessment team -- the Domain
+% Expert (Section~\ref{sec_vet_software_list}).
 
@@
 they achieved this or what improvements are needed to achieve it.
+
+\subsection{Overall Process} \label{Sec_OverallProcess}
+
+The assessment was conducted via the following steps:
+
+\begin{enumerate}
+	\item List candidate software packages for the domain.
+	(Section~\ref{identifysoftware})
+	\item Filter the software package list. (Section~\ref{filtersoftware})
+	\item Gather the source code and documentation for each software package.
+	\item Collect quantitative measures from the project repositories.
+	(Section~\ref{empiricalmeasures})
+	\item Measure using the measurement template.  The full measurement template
+	can be found in \citet{SmithEtAl2021}. (Section~\ref{empiricalmeasures})
+	\item Survey the developers. (Section~\ref{SecSurvey})
+	\item Use AHP to rank the software packages. (Section~\ref{AHP})
+	\item Analyze the results and answer the research questions.
+\end{enumerate}
+
+The above steps depend on interaction with a domain expert partner, as discussed
+in Section~\ref{sec_vet_software_list}.
 
@@
 \rqref{RQ_ComparePainPoints}, and \rqref{RQ_Concerns} (which enables
-\rqref{RQ_CompareConcerns}). We designed a list of 20 questions to guide our
+\rqref{RQ_Recommend}). We designed a list of 20 questions to guide our
 interviews, which can be found in \citet{SmithEtAl2021}. Some questions are

### 2022-01-24 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Updates to Section 2.1 on Qualities

@@
 We adopt software quality definitions from various researchers and subject
-matter experts. Some of the definitions are from \cite{Smithetal2020}. The
-following are the software quality definitions used in this exercise, along with
-comments regarding their quantitative and qualitative measurement.  \wss{Update
-when the Quality Definition of Qualities document is completed.}
-
-\subsubsection{Installability}
-
-Installability is measured by the effort required for the installation,
-uninstallation or reinstallation of a software product in a specified
-environment \citep{ISO/IEC25010, lenhard2013measuring}. A good measure of
-installability correlates with scenarios when low or moderate effort is required
-to gather and prepare software for its general use on a system for which it was
-designed. In our case effort includes the time spent finding and understanding
-the installation instructions, the person-time and resources spent performing
-the installation procedure, and the absence or ease of overcoming system
-compatibility issues. The ability to reasonably validate the installation
-procedure and the ease of installation also have a positive effect on the
-measure of installability.
-
-\subsubsection{Correctness}
-
-A software program is correct if it behaves according to its stated
-specifications \citep{GhezziEtAl2003}. This requires that the specification is
-available. Scientific software is unlikely to have a formal specification, since
-this is not common practice in the field. As a consequence, the correctness of
-software often cannot be measured directly. Despite an absent specification, the
-correctness can indirectly be assessed by looking at the output and applying
-domain knowledge. For our measurement of correctness we have assumed that the
-following factors suggest correctness was considered: availability of a
-requirements specification, reference to domain theory, and explicit use of
-tools and/or techniques for building confidence of correctness, such as
+matter experts. The following are the software quality definitions used in this
+exercise, along with comments regarding their measurement.
+
+\begin{description}
+
+	\item[Installability] Installability is measured by the effort required for
+	the installation, uninstallation or reinstallation of a software product in
+	a specified environment \citep{ISO/IEC25010, lenhard2013measuring}. A good
+	measure of installability correlates with scenarios when low or moderate
+	effort is required to gather and prepare software for its general use on a
+	system for which it was designed. In our case effort includes the time spent
+	finding and understanding the installation instructions, the person-time and
+	resources spent performing the installation procedure, and the absence or
+	ease of overcoming system compatibility issues. An installability score
+	improves if the software has a means to validate the installation and
+	instructions for uninstallation.
+
+	\item[Correctness] A software program is correct if it behaves according to
+	its stated specifications \citep{GhezziEtAl2003}. This requires that the
+	specification is available. Scientific software is unlikely to have an
+	explicit specification, since this is not common practice in the field. As a
+	consequence, the correctness of software cannot often be measured directly.
+	We assess correctness indirectly by looking for the following: availability
+	of a requirements specification, reference to domain theory, and explicit
+	use of tools and/or techniques for building confidence, such as
 	documentation generators and software analysis tools.
 
-\subsubsection{Verifiability}
-
-Verifiability is measured by the extent to which a set of tests can be written
-and executed to demonstrate that the delivered system meets the specification
-\citep{sommerville}. Similarly to correctness, verifiability is correlated with
-the availability of a specification and with reference to domain knowledge. A
-good measure of verifiability is further correlated with the availability of
-well written tutorials that include expected output, with software unit testing
-documentation, and with evidence of continuous integration during the
-development process. In our process we measure correctness and verifiability
-together.
-
-\subsubsection{Reliability}
-
-Reliability is measured by the probability of failure-free operation of a
-computer program in a specified environment for a specified time; that is,
-reliability can be measured by the average time interval between two failures,
-also known as the mean time to failure (MTTF) \citep{GhezziEtAl2003}
-\citep{musa1987software}. Reliability is thus positively correlated with the
-absence of errors during installation and use. Recoverability from errors also
-improves reliability.
-
-\subsubsection{Robustness}
-
-Software possesses the characteristic of robustness if it behaves ``reasonably''
-in two situations: i) when it encounters circumstances not anticipated in the
-requirements specification; and ii) when the assumptions in its requirements
-specification are violated \citep{boehm2007software, ghezzi1991fundamentals}. A
-good measure of robustness correlates with a reasonable reaction to unexpected
-input, including data of the wrong type, empty input, or missing files or links.
-A reasonable reaction includes an appropriate error message and the ability to
-recover the system.
-
-\subsubsection{Performance}
-
-Performance is measured by the degree to which a system or component
-accomplishes its designated functions within given constraints, such as speed
-(database response times, for instance), throughput (transactions per second),
-capacity (concurrent usage loads), and timing (hard real-time demands)
-\citep{IEEEStdGlossarySET1990, wiegers2003softreq}. In this state of the
-practice assessment performance was not directly quantitatively measured.
-Instead the documentation of each software package was observed for information
-that alludes to a consideration of performance, such as parallelization tools.
-
-\subsubsection{Usability}
-
-Usability is measured by the extent to which a software product can be used by
-specified users to achieve specified goals with effectiveness, efficiency, and
-satisfaction in a specified context of use \citep{nielsonusability}. We assumed
-that a high usability correlates with the presence of documentation,
-including tutorials, manuals, and defined user characteristics, and user
-support. Preferably the user support model has avenues to contact developers and
-report issues.
-
-\subsubsection{Maintainability}
-
-A measure of maintainability is the effort with which a software system or
-component can be modified to correct faults, improve performance or other
-attributes, and satisfy new requirements \citep{IEEEStdGlossarySET1990,
-boehm2007software}. In the current work maintainability is measured by the
-quality of documentation artifacts, and the presence of version control and
-issue tracking. These artifacts can greatly decrease the effort needed to modify
-software. There are many documentation artifacts that can improve
-maintainability, including user and developer manuals, specifications, README
-files, change logs, release notes, publications, forums, and instructional
-websites.
-
-\subsubsection{Modifiability}
-
-Modifiability refers to the ease with which stable changes can be made to a
-system and the flexibility of the system to adopt such changes \citep{8016712}.
-We did not directly measure modifiability. Instead, developers were asked in
-interviews if they considered the ease of future changes when developing the
-software packages, specifically changes to the structure of the system, modules
-and code blocks. A follow up question is asked if any measures had been taken.
-
-\subsubsection{Reusability}
-
-Reusability refers to the extent to which components of a software package can
-be used with or without adaptation in other software packages
-\citep{kalagiakos2003non}. A good measure of reusability results from a large
-number of easily reusable components. Increased software modularization, defined
-as the presence of smaller components with well defined interfaces, is
-important. For this state of the practice assessment, a good measure of
-reusability correlates with an increased number of code files, and the
-availability of API documentation.
-
-\subsubsection{Understandability}
-
-Understandability is measured by the capability of the software package to
-enable the user to understand its suitability and function \citep{ISO9126}. It
-is an artifact-dependent quality. Understandability is different for the
-user-interface, source code, and documentation. In this state of the practice
-analysis, understandability focuses on the source code. It is measured by the
-consistency of a formatting style, the extend of modularization, the explicit
-identification of coding standards, the presence of meaningful identifiers, and
-clarity of comments.
-
-\subsubsection{Traceability}
-
-Traceability refers to the ability to link the software implementation and the
-software artifacts, especially the requirement specification
-\citep{McCallEtAl1977}. Similar to the quality of correctness, this requires the
-availability of some form of specification. This quality refers to keeping track
-of information as it changes forms or relates between artifacts. We did not
-quantitatively measure traceability. Instead, developers were asked in
-interviews how documentation fits into their development process.
-
-\subsubsection{Visibility and Transparency}
-
-Visibility and transparency refer to the extent to which all of the steps of a
-software development process, and the current status of it, are conveyed clearly
-\citep{ghezzi1991fundamentals}. In this state of the practice assessment a good
-measure of visibility and transparency correlates with a well defined
-development process, documentation of the development process and environment
-and software package version release notes.
-
-\subsubsection{Reproducibility}
-
-Software achieves reproducibility if another developer can take the requirements
-documentation and re-obtain the same software artifacts
-\citep{BenureauAndRougier2017}. This includes the output of the software, where
-the scientific results are compared between software implementations, or between
-software implementations and manually calculated results. We measured
-reproducibility qualitatively by asking developers if they have any concern that
-their computational results won't be reproducible, and if they have taken any
-steps to ensure reproducibility.
-
-\subsubsection{Unambiguity}
-
-Unambiguity refers to the extent to which two readers have similar
-interpretations when reading software artifacts. In other words, artifacts are
-unambiguous if, and only if, they only have one interpretation \citep{IEEE1998}.
-This state of the practice assessment did not quantitatively measure
-unambiguity, but we did ask developers if they think that the current
-documentation can clearly convey all necessary knowledge to the users, and how
-they achieved this or what improvements are needed to achieve it.
+	\item[Verifiability] An artifact (like code, or a theory manual) is
+	verifiable if, and only if, every statement therein is verifiable. A
+	statement is verifiable if, and only if, there exists some finite
+	cost-effective process with which a person or machine can check that each
+	statement is correct (adapted from recommended practice for requirements
+	specifications \citep{IEEE1998}). Similarly to correctness, verifiability is
+	correlated with the availability of a specification and with reference to
+	domain knowledge. A good measure of verifiability is further correlated with
+	the availability of well written tutorials that include expected output,
+	with software unit testing, and with evidence of continuous integration. In
+	our process we measure correctness and verifiability together.
+
+	\item[Reliability] Reliability is measured by the probability of
+	failure-free operation of a computer program in a specified environment for
+	a specified time \citep{GhezziEtAl2003, musa1987software}. Reliability is
+	thus positively correlated with the absence of errors during installation
+	and use. Recoverability from errors also improves reliability.
+
+	\item[Robustness] Software possesses the characteristic of robustness if it
+	behaves ``reasonably'' in two situations: i) when it encounters
+	circumstances not anticipated in the requirements specification; and ii)
+	when the assumptions in its requirements specification are violated
+	\citep{boehm2007software, ghezzi1991fundamentals}. A good measure of
+	robustness correlates with a reasonable reaction to unexpected input,
+	including data of the wrong type, empty input, or missing files or links.
+	Examples of reasonable reaction include an appropriate error message and the
+	ability to recover the system.
+
+	\item[Performance] Performance is measured by the degree to which a system
+	or component accomplishes its designated functions within given constraints,
+	such as speed (database response times, for instance), throughput
+	(transactions per second), capacity (concurrent usage loads), and timing
+	(hard real-time demands) \citep{IEEEStdGlossarySET1990, wiegers2003softreq}.
+	In this state of the practice assessment performance was not directly
+	quantitatively measured. Instead the documentation of each software package
+	was observed for information alluding to consideration of performance, such
+	testing results, or parallelization instructions.
+
+	\item[Usability] Usability is measured by the extent to which a software
+	product can be used by specified users to achieve specified goals with
+	effectiveness, efficiency, and satisfaction in a specified context of use
+	\citep{nielsonusability}. We assumed that high usability correlates with
+	the presence of documentation, including tutorials, manuals, defined
+	user characteristics, and user support. Preferably the user support model
+	will have avenues to contact developers and report issues.
+
+	\item[Maintainability] A measure of maintainability is the effort with which
+	a software system or component can be modified to correct faults, improve
+	performance or other attributes, and satisfy new requirements
+	\citep{IEEEStdGlossarySET1990, boehm2007software}. In the current work
+	maintainability is measured by the quality of documentation artifacts, and
+	the presence of version control and issue tracking. These artifacts can
+	greatly decrease the effort needed to modify software. There are many
+	documentation artifacts that can improve maintainability, including user and
+	developer manuals, specifications, README files, change logs, release notes,
+	publications, forums, and instructional websites.
+
+	\item[Modifiability] Modifiability refers to the ease with which stable
+	changes can be made to a system and the flexibility of the system to adopt
+	such changes \citep{8016712}. We did not directly measure modifiability.
+	Instead, developers were asked in interviews if they considered the ease of
+	future changes when developing the software packages, specifically changes
+	to the structure of the system, modules and code blocks. A follow up
+	question is asked if any measures had been taken.
+
+	\item[Reusability] Reusability refers to the extent to which components of a
+	software package can be used with or without adaptation in other software
+	packages \citep{kalagiakos2003non}. A good measure of reusability results
+	from a large number of easily reusable components. Increased software
+	modularization, defined as the presence of smaller components with well
+	defined interfaces, is important. For this state of the practice assessment,
+	we assumed that a good measure of reusability correlates with an increased
+	number of code files, and the availability of API documentation.
+
+	\item[Understandability] Understandability is measured by the capability of
+	the software package to enable the user to understand its suitability and
+	function \citep{ISO9126}. It is an artifact-dependent quality.
+	Understandability is different for the user-interface, source code, and
+	documentation. In this state of the practice analysis, understandability
+	focuses on the source code. It is measured by the consistency of a
+	formatting style, the extend of modularization, the explicit identification
+	of coding standards, the presence of meaningful identifiers, and the clarity
+	of comments.
+
+	\item[Traceability] Traceability refers to the ability to link the software
+	implementation and the software artifacts, especially the requirement
+	specification \citep{McCallEtAl1977}. Similar to the quality of correctness,
+	this requires the availability of some form of specification. This quality
+	refers to keeping track of information as it changes forms or relates
+	between artifacts. We did not quantitatively measure traceability. Instead,
+	developers were asked in interviews how documentation fits into their
+	development process.
+
+	\item[Visibility and Transparency] Visibility and transparency refer to the
+	extent to which all of the steps of a software development process, and the
+	current status of it, are conveyed clearly \citep{ghezzi1991fundamentals}.
+	In this state of the practice assessment a good measure of visibility and
+	transparency correlates with a well defined development process,
+	documentation of the development process and environment and software
+	package version release notes.
+
+	\item[Reproducibility] Software achieves reproducibility if another
+	developer can take the requirements documentation and re-obtain the same
+	software artifacts \citep{BenureauAndRougier2017}. This includes the output
+	of the software, where the scientific results are compared between software
+	implementations, or between software implementations and manually calculated
+	results. We measured reproducibility qualitatively by asking developers if
+	they have any concern that their computational results won't be
+	reproducible, and if they have taken any steps to ensure reproducibility.
+
+	\item[Unambiguity] Unambiguity refers to the extent to which two readers
+	have similar interpretations when reading software artifacts. In other
+	words, artifacts are unambiguous if, and only if, they only have one
+	interpretation \citep{IEEE1998}. This state of the practice assessment did
+	not quantitatively measure unambiguity, but we did ask developers if they
+	think that the current documentation can clearly convey all necessary
+	knowledge to the users, and how they achieved this or what improvements are
+	needed to achieve it.
+
+\end{description}
 

### 2022-01-24 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Edits to Sections on identifying and filtering the software - answers RQ1

@@
 
-The above steps depend on interaction with a domain expert partner, as discussed
+The above steps depend on interaction with a Domain Expert partner, as discussed
 in Section~\ref{sec_vet_software_list}.
@@
 
-Part of how we answer \rqref{RQ_WhatProjects}.  The candidate software was found
-through search engine queries targeting authoritative lists of software. We
-found LBM software listed on the websites GitHub and swMATH, as well as through
-articles found in scholarly journals and databases.  The Domain Expert
+To answer~\rqref{RQ_WhatProjects} we needed to identify existing LBM projects.
+The candidate software was found through search engine queries targeting
+authoritative lists of software. We found LBM software listed on
+\href{https://github.com/} {GitHub} and \href{https://swmath.org/} {swMATH}, as
+well as through scholarly articles. The Domain Expert
 (Section~\ref{sec_vet_software_list}) was also engaged in selecting the
-candidate software.  \wss{Packages that fit our criteria but were found after
-data collection and analysis was conducted, such as Musubi
-\citep{HasertEtAl2014}, will need to be added to future SOP assessments.}
+candidate software.
 
@@
 	for GitHub-style repositories.
-	\item The software cannot be marked as incomplete or in an initial
-	development phase.
+	\item The software cannot be marked as incomplete or in an initial state of
+	development.
 \end{enumerate}
 
-The initial list had 45 packages, including a few packages that were later found
-to not have publicly available source code, or to be in an incomplete state of
-development.
+The initial list had 45 packages, including a few that were later found to not
+have publicly available source code, or to be incomplete projects.
 
@@
 
-With previous section, how we answer \rqref{RQ_WhatProjects}.  Answer to
-\rqref{RQ_WhatProjects} provided in this section.  To reduce the number of
-members in the candidate software list to a manageable size, the following
-filters were applied. The filters were applied in the priority order listed.
+To reduce the length of the list of software packages and to
+answer~\rqref{RQ_WhatProjects}, we filter the initial list of 45.  The following
+filters were applied in the priority order listed.
 
@@
 
-While the initial list had 46 packages, filtering by scope, usage, and age
-decreased the size of the list to 24 packages. Many of the 22 packages that were
-removed could not be tested as there was no installation guide, they were
-incomplete, source code was not publicly available, a license was needed, or the
-project was out of scope or not up to a standard that would support
-incorporating them into this study. These eliminated software packages are
-listed in the Appendix of \citet{Michalski2021}. Of the remaining 24 packages
-that were studied, some were kept on the list despite being marked as dead due
-to their prevalence on authoritative lists on LBM software and due to their
-surface excellence.  The final list of software packages that were analyzed in
-this project can be found in the following two tables. Table~\ref{alivepackages}
-lists packages that fell into the `alive' category as of mid 2020, and
-Table~\ref{deadpackages} lists packages that were `dead' at that time.
+Filtering by scope, usage, and age decreased the size of the list to 23
+packages. Many of the 22 packages that were removed could not be tested as there
+was no installation guide, they were incomplete, source code was not publicly
+available, a license was needed, or the project was out of scope or not up to a
+standard that would support incorporating them into this study. These eliminated
+software packages are listed in the Appendix of \citet{Michalski2021}. Of the
+remaining 23 packages that were studied, some were kept on the list despite
+being marked as dead due to their prevalence on authoritative lists on LBM
+software and due to their surface excellence.  After the initial measurement
+exercise, which took place in mid 2020, we added one more package in January
+2022: Musubi.  Musubi did not come up in our initial search, but once we learned
+of it, we felt that this report is incomplete without including it.
+
+The final list of 24 software packages that were analyzed in this project can be
+found in two tables. Table~\ref{alivepackages} lists packages that fell into the
+`alive' category as of mid 2020 (January 2022 for Musubi), and
+Table~\ref{deadpackages} lists packages that were `dead' at that time. The
+tables include a hyperlink to each project and, when available, citations for
+relevant publications.  The final list of software packages shows considerable
+variation in purpose, size, user interfaces, and programming languages. For
+example, the OpenLB software package is predominantly a \CC{} package that makes
+use of hybrid parallelization and was designed to address a range of CFD
+problems \citep{heuveline2009towards}. The software package pyLBM is an
+all-in-one Python language package for numerical simulations
+\citep{graille2017pylbm}. ESPResSo is an extensible simulation package that is
+specifically for research on soft matter, and is written in \CC{} and Python
+\citep{weik2019espresso}. The HemeLB package is used for efficient simulation of
+fluid flow in several medical domains, and is written predominantly in C, \CC,
+and Python \citep{mazzeo2008hemelb}.
 
@@
 \end{table}
-
-There is considerable variation among these software packages, including their
-intended purpose, size, user interfaces, and software languages used. For
-example, the OpenLB software package is predominantly a \CC{} package that makes
-use of hybrid parallelization and was designed to address a range of CFD
-problems \citep{heuveline2009towards}. The software package pyLBM is an
-all-in-one Python language package for numerical simulations
-\citep{graille2017pylbm}. ESPResSo is an extensible simulation package that is
-specifically for research on soft matter, and is written in \CC{} and Python
-\citep{weik2019espresso}. The HemeLB package is used for efficient simulation of
-fluid flow in several medical domains, and is written predominantly in C, \CC,
-and Python \citep{mazzeo2008hemelb}.
 

### 2022-01-25 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Revisions complete for Section 2, revised the sample of the measurement template

@@
 
-Part of how we address \rqref{RQ_HighestQuality} and
-\rqref{RQ_CompareArtifacts}.  Once have \rqref{RQ_HighestQuality} can combine
-with estimate of popularity to answer \rqref{RQ_CompareHQ2Popular}.  The
-measurement template is found in \citet{SmithEtAl2021}.  This template is used
-to track measurements and quality scores for all of the software packages in the
-domain. For each software package, we fill-in the rows of the template. This
-process takes about 2 hours per package, with a cap of 4 hours.  The time
-constraint is necessary so that the work load for measuring a given domain is
-feasible for a team as small as one, since we aim for a cap of 160 person hours
-for the measurement phase \citet{SmithEtAl2021}.  An excerpt of the spreadsheet
-is shown in Figure~\ref{measurement_template_image}.  A column is added to this
-template for each of the 24 software package measured.
+We rank the projects by how well they follow best practices
+(\rqref{RQ_HighestQuality}) via a measurement template, as described in
+\citet{SmithEtAl2021}.  For each software package (each column in the template),
+we fill-in the rows of the template. This process takes about 2 hours per
+package, with a cap of 4 hours. The time constraint is necessary so that the
+work load is feasible for a team as small as one, given our aim to cap the
+measurement phase at 160 person hours \citep{SmithEtAl2021}.  An excerpt of the
+template, in spreadsheet form, is shown in
+Figure~\ref{measurement_template_image}.
 
@@
 	\begin{center}
-	  \includegraphics[width=1.0\textwidth]{./figures/measurement_template}
-	  \caption{Excerpt of the Top Section of the Measurement Template (Summary
-	  Information) \wss{Should update this figure to something easier to read}}
+	  \includegraphics[width=1.0\textwidth]{./figures/measurement_template.pdf}
+	  \caption{Excerpt of the Top Sections of the Measurement Template}
 	  \label{measurement_template_image}
@@
 The full template consists of 108 questions categorized under 9 qualities:
-\begin{inparaenum}
+\begin{inparaenum}[(i)]
 	\item installability;
@@
 \end{inparaenum}
+
 The questions were designed to be unambiguous, quantifiable and measurable with
@@
 Several of the qualities use the word ``surface''.  This is to highlight that,
-for these qualities in particular, the best that we can do is a shallow measure
-of the quality.  For instance, we are not currently doing any experiments to
-measure usability.  Instead, we are looking for an indication that usability was
-considered by the developers.  We do this by looking for cues in the
-documentation, like a getting started manual, a user manual and documentation of
-expected user characteristics.
-
-Most of the data was gathered by manually investigating the software, its source
-code, and its artifacts, while some was gathered using the automatic repository
-measurement tools.  Tools were use to find data such as the number of files,
+for these qualities in particular, the best that we can do is a shallow measure.
+For instance, we do not conduct experiments to measure usability. Instead, we
+are looking for an indication that usability was considered by the developers by
+looking for cues in the documentation, such as getting started instructions, a
+user manual or documentation of expected user characteristics.
+
+Tools were use to find some of the measurements, such as the number of files,
 number of lines of code (LOC), percentage of issues that are closed, etc. The
@@
 advantage of using VMs is to level the playing field. Every software install
-starts from a clean slate, which removes ``works-on-my-computer'' errors. When
-filling in the measurement template spreadsheet, the details for each VM
-should be noted, including hypervisor and operating system version.
+starts from a clean slate, which removes ``works-on-my-computer'' errors.
 
@@
 
-Part of how we address \rqref{RQ_HighestQuality}.  The Analytical Hierarchy
-Process (AHP) is a decision-making technique that is used to compare multiple
-options by multiple criteria. In our work AHP was used for comparing and ranking
-the LBM software packages using the overall impression quality scores that were
-gathered in the measurement template.  AHP performs a pairwise analysis using a
-matrix and generates an overall score as well as individual quality scores for
-each software package. \citep{SmithEtAl2016} shows how AHP is applied to ranking
+Once we have measured each package, we still need to rank them to answer
+\rqref{RQ_HighestQuality}.  To do this, we used the Analytical Hierarchy Process
+(AHP), a decision-making technique that is used to compare multiple options by
+multiple criteria. AHP was used for comparing and ranking the LBM software
+packages using the overall impression quality scores that were gathered in the
+measurement template.  AHP performs a pairwise analysis using a matrix and
+generates an overall score as well as individual quality scores for each
+software package. \citep{SmithEtAl2016} shows how AHP is applied to ranking
 software based on quality measures.
@@
 
-Part of how we address \rqref{RQ_CompareToolsProjMngmnt},
-\rqref{RQ_CompareMethodologies}, \rqref{RQ_PainPoints} (which enables
-\rqref{RQ_ComparePainPoints}, and \rqref{RQ_Concerns} (which enables
-\rqref{RQ_Recommend}). We designed a list of 20 questions to guide our
-interviews, which can be found in \citet{SmithEtAl2021}. Some questions are
-about the background of the software, the development teams, the interviewees,
-and how they organize the projects. We also ask about the developer's
-understandings of the users. Some questions focus on the current and past
+Several of the research question (\rqref{RQ_CompareToolsProjMngmnt},
+\rqref{RQ_CompareMethodologies}, \rqref{RQ_PainPoints}, \rqref{RQ_Concerns} and
+\rqref{RQ_Recommend}) require going beyond the quantitative data from the
+measurement template. To gain the required insight, we interviewed developers
+using a list of 20 questions from \citet{SmithEtAl2021}. The questions cover the
+background of the development teams, the interviewees, and the software itself.
+We ask the developers how they organize their projects. We also ask them about
+their understanding of the users. Some questions focus on the current and past
 difficulties, and the solutions the team has found, or will try. We also discuss
-the importance and current situations of documentation. A few questions are
-about specific software qualities, such as maintainability, understandability,
-usability, and reproducibility. The interviews are semi-structured based on the
-question list; we ask follow-up questions when necessary. Based on our
-experience, the interviewees usually bring up some exciting ideas that we did
-not expect, and it is worth expanding on these topics.
-
-We requested interviews from all 24 packages, except Munubi, since Munubi was
-added after our ethics approved study window had expired.  We were able to
-recruit 4 developers to participate in our study.  To send out interview
-requests we found we found contacts on the projects' website, or code
-repository, or publications, or the biographic pages of the teams'
-institutions. We send at most two interview request emails to a contact for each
-software package.  Meeting will typically be held using on-line meeting
-software, like Zoom or Teams.  This facilitates recording and automatic
-transcription of the meetings.
-
-The interviews followed the standard ethics guideline of asking for consent
-before interviewing, recording, and including participant details in the report.
-The interview process presented here was approved by the McMaster University
-Research Ethics Board under the application number 
+the importance of, and the current situation for, documentation. A few questions
+are about specific software qualities, such as maintainability,
+understandability, usability, and reproducibility. The interviews are
+semi-structured based on the question list; we ask follow-up questions when
+necessary. Based on our experience, the interviewees often bring up exciting and
+unexpected ideas.
+
+We requested interviews with developers for all 24 packages, except Munubi,
+since Munubi was added after our ethics approved had expired.  We were able to
+recruit 4 developers to participate in our study.  Our potential interview
+subjects were found from project websites, code repositories, publications, and
+biographic information on institution web-pages. Meeting were held using on-line
+meeting software (either Zoom or Teams), which facilitated recording and
+automatic transcription of the meetings.  The interview process was approved by
+the McMaster University Research Ethics Board under the application number 
 \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/MACREM/Application.pdf}
@@
 
-Part of how we address \rqref{RQ_WhatProjects} and \rqref{RQ_HighestQuality}.
-\wss{Should ask Zahra about \rqref{RQ_CompareHQ2Popular}, see if he has anything
-to add about \rqref{RQ_PainPoints}.} The Domain Expert is an important member of
-the state of the practice assessment team. Pitfalls exist if non-experts attempt
-to acquire an authoritative list of software, or try to definitively rank the
-software. Non-experts have the problem that they can only rely on information
-available on-line, which has the following drawbacks:
+We partnered with a Domain Expert to vet our list of projects
+(\rqref{RQ_WhatProjects}) and our ranking (\rqref{RQ_HighestQuality}).  The
+Domain Expert is an important member of the state of the practice assessment
+team. Pitfalls exist if non-experts attempt to acquire an authoritative list of
+software, or try to definitively rank the software. Non-experts have the problem
+that they can only rely on information available on-line, which has the
+following drawbacks:
 \begin{inparaenum}[i)]
@@
 
-Looking at the top three ranked packages: ESPResSo had achieved a
-relative high score in installability, surface usability, maintainability, reusability, and visibility and transparency.
-Ludwig scored high in surface correctness and verifiability, surface robustness, surface usability, maintainability, and visibility and transparency.
-Palabos scored high in installability, surface reliability, surface robustness, maintainability, understandability, and visibility and transparency.
+Looking at the top three ranked packages: ESPResSo had achieved a relative high
+score in installability, surface usability, maintainability, reusability, and
+visibility and transparency. Ludwig scored high in surface correctness and
+verifiability, surface robustness, surface usability, maintainability, and
+visibility and transparency. Palabos scored high in installability, surface
+reliability, surface robustness, maintainability, understandability, and
+visibility and transparency.
 

### 2022-01-25 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Edits up to and including Section 3.1 on Installability.

@@
 
-Answers \rqref{RQ_HighestQuality}.  Before discussing quality differences between the packages and the results of
-the AHP ranking, we will contrast the different packages based on features, as
-shown in Table~\ref{tbl_features}. \wss{Add some discussion of some example
-features.  Add Non-Newtonian and Fluid Structure interaction.  Listed in
-alphabetical order.}  The sections below summarize how the software packages
-compare based on qualities, as measured by the measurement template.   The
-results of the AHP ranking are also discussed.
+In this section we answer \rqref{RQ_HighestQuality} by using AHP to rank the
+identified LBM software against best practices.  Before discussing quality
+differences between the packages and the results of the AHP ranking, we will
+contrast the different packages based on their features, as shown in
+Table~\ref{tbl_features}. The set of LBM programs form a program family, or a
+software product line.  The packages all have the commonality of using LBM
+techniques to simulate the motion of fluids, but there are also variabilities
+between them.  LBM packages can be distinguished by the dimension of problems
+they simulate, whether or not (and how) they support parallel computation,
+whether or not compressibility can be considered, whether reflexive boundaries
+can be employed, whether the simulation handles multi-fluids, whether the
+simulations handles turbulence, and whether complex geometries can be simulated.
+All of these variabilities are listed as columns in Table~\ref{tbl_features}
+with the software listed in alphabetical order. A more complete analysis of the
+commonalities and variabilities between LBM packages is provided in
+\citet{Michalski2021}.
 
@@
 		\end{tabular}
-		\caption{Features of Software Packages (Dim for Dimension (1, 2, 3),
-			Pll for Parallel (CUDA, MPI, OpenMP (OMP)), Com for Compressible (Yes or No), Rflx for
-			Reflexive Boundary Condition (Yes or No), MFl for Multi-fluid (Yes or
-			No), Turb for Turbulent (Yes or No), CGE for Complex Geometries (Yes or No), OS for Operating System (Windows (W), macOS (M), Linux (L)))} \label{tbl_features}
+		\caption{Features of Software Packages (Dim for Dimension (1, 2, 3), Pll
+			for Parallel (CUDA, MPI, OpenMP (OMP)), Com for Compressible (Yes or
+			No), Rflx for Reflexive Boundary Condition (Yes or No), MFl for
+			Multi-fluid (Yes or No), Turb for Turbulent (Yes or No), CGE for
+			Complex Geometries (Yes or No), OS for Operating System (Windows
+			(W), macOS (M), Linux (L)))} \label{tbl_features}
 	\end{center}
@@
 
+The sections below summarize how the software packages compare based on the
+qualities discussed in Section~\ref{softwarequalities}, as measured via the
+measurement template (Section~\ref{empiricalmeasures}).  For the purpose of
+discussion, the last subsection combines all the qualities into one ``global''
+ranking.
+
 \subsection{Installability}
@@
 All of the 24 software packages that were tested have installation instructions.
-As noted previously, many of the 22 software packages that were part of the
-original long list of 46 packages were removed due to not including
-documentation or installation instructions. Of the 24 software packages that
-were tested, most had installation instructions located in one place, often in
-an instruction manual or on a web-page. Sometimes, like with Ludwig, incomplete
+Most have their installation instructions located in one place, often in an
+instruction manual or on a web-page. Sometimes, like with Ludwig, incomplete
 installation instructions are found on a home page, with more detailed
 instructions located on another web-page, or within the documentation.
-Maintainability and correctness of these instructions could be improved if all
-the instructions were in one location.
-
-All 24 packages on the short list can be installed on some Unix-like systems.
-Eight packages could be installed on Windows, and five on macOS. Operating
-system requirements are listed in the documentation of 20 software packages, but
+Installability of the software, and maintainability and correctness of the
+instructions themselves, are improved if all the instructions are in one
+location.
+
+All 24 packages on the short list can be installed on a Unix-like system. Eight
+packages could be installed on Windows, and five on macOS. Operating system
+requirements are listed in the documentation of 20 software packages, but
 compatible operating system versions are only specified in four packages
@@
 were tested on Ubuntu for this state of the practice assessment. TCLB was tested
-on CentOS, since this operating system is mentioned in its installation
-instructions.
+on CentOS, since CentOS is mentioned in its installation instructions.
 
@@
 system incompatibility, such as the requirement of CUDA. Fourteen software
-packages definitively broke during installation. Some packages, such as
-LB2D-Prime and LB3D-Prime, did not provide a definitive message of the success
-or failure of installation. In these instances, validating the installation
-required performing a tutorial or running a script, as described below, if these
-were available.
+packages broke during installation. Some packages, such as LB2D-Prime and
+LB3D-Prime, did not provide a definitive message of the success or failure of
+installation. In these instances, validating the installation required
+performing a tutorial or running a script, as described below, if these were
+available.
 
@@
 
-Seventeen software packages require less than 10 dependencies to be installed. All but one software package (LatBo.jl) require less than 20 dependencies. Some
+Seventeen software packages require less than 10 dependencies to be installed.
+All but one software package (LatBo.jl) require less than 20 dependencies. Some
 packages may automatically install additional dependencies in the background.
-Twenty of the software packages do not explicitly indicate software dependency
-versions. Some software package installation issues, specifically those
-occurring when manual installation of dependencies is required, may be avoided
-if versions of dependencies are specified. Sixteen software packages do not have
-detailed instructions for installing dependencies. Sixteen software packages
-have less than 10 manual installation steps. If dependencies are installed in
-one command then none of the software packages take more than 20 steps to
-install. The average number of steps is about eight, and the fewest is two
-(LB3D-Prime).
+Twenty of the software packages do not explicitly indicate the versions of
+software dependencies. Some software package installation issues, specifically
+those occurring when manual installation of dependencies is required, may be
+avoided if versions of dependencies are specified. Sixteen software packages do
+not have detailed instructions for installing dependencies. Sixteen software
+packages have less than 10 manual installation steps. If dependencies are
+installed in one command then none of the software packages take more than 20
+steps to install. The average number of steps is about eight, and the fewest is
+two (LB3D-Prime).
 
@@
 Figure~\ref{Fig_Installability} shows the installability ranking of the software
-packages using AHP.
-Software packages with a higher score (ESPResSo, MechSys, OpenLB, DL\_MESO, Palabos) tend to have one set of linear installation instructions that are written as if the person doing the installation has none of the dependencies installed.
-The instructions often list compatible operating system versions and include instructions for the installation of dependencies.
-The top ranked packages often incorporate some sort of automation of the installation process and have fewer manual installation steps.
-The number of dependencies a package has does not correlate with a higher score.
-The ability to validate the installation process, often through tutorials or test examples that include expected output, is correlated with a higher score.
-Furthermore, the top seven ranked packages are noted as being alive.
-
-Many software packages would benefit from a rewrite or reorganization of
-installation instructions. A single location for installation instructions would
-improve their maintainability and correctness. Listing compatible operating
-system and dependency versions would decrease installation time and errors, as
-would adding instructions on installing dependencies. Installation process
-errors should prompt the system to display detailed messages. Once a software
-package is installed, either an automatic validation needs to be performed or
-the user needs to be able to perform a manual validation using test examples
-that include expected output. Finally, uninstallation instructions should be
-included in the documentation.
+packages using AHP. Software packages with a higher score (ESPResSo, MechSys,
+OpenLB, DL\_MESO, Palabos) tend to have one set of linear installation
+instructions written as if the person doing the installation has none of the
+dependencies installed. The instructions often list compatible operating system
+versions and include instructions for the installation of dependencies. The top
+ranked packages often incorporate some sort of automation of the installation
+process and have fewer manual installation steps. The number of dependencies a
+package has does not correlate with a higher score. The ability to validate the
+installation process, often through tutorials or test examples that include
+expected output, is correlated with a higher score. Active work on a project
+seems to improve installability, since the top seven ranked packages are all
+classified as alive.
 
@@
 software packages have a tutorial, 13 have a user manual, and 11 have both. The
-tutorials vary in scope and substance, and eight include expected output.
-Most user manuals are in the form of a file that can be downloaded, while some
-are rendered on a web-page. Some packages (waLBerla) do not have a user manual,
-but do have useful documentation distributed on their web-pages. Expected user
+tutorials vary in scope and substance, and eight include expected output. Most
+user manuals are in the form of a file that can be downloaded, while some are
+rendered on a web-page. Some packages (waLBerla) do not have a user manual, but
+do have useful documentation distributed on their web-pages. Expected user
 characteristics are documented in five software packages (laboetie, LIMBES,
-Ludwig, Musubi, Palabos).
-Users are typically scientists or engineers.
-Their background is often physics, chemistry, biophysics, or mathematics.
-All but one of the packages (LIMBES) have a user support model, and many of them have multiple avenues of user support.
-The most popular avenue of support is Git, followed by email and forums.
-One software package (OpenLB) has an FAQ page.
+Ludwig, Musubi, Palabos). Users are typically scientists or engineers. Their
+background is often physics, chemistry, biophysics, or mathematics. All but one
+of the packages (LIMBES) have a user support model, and many of them have
+multiple avenues of user support. The most popular avenue of support is Git,
+followed by email and forums. One software package (OpenLB) has an FAQ page.
 
@@
 Sailfish) have less than 50 percent of their issues closed. All of the top five
-overall ranked packages have most of their issues closed.
-Alive packages (11 use Git issue tracking) have 64\% of their issues closed, while dead packages (3 use Git issue tracking) have 71\% of their issues closed.
-This information is presented in Table~\ref{gitrepodata}.
-Furthermore, 13 packages that use Git for issue tracking use GitHub as a version control system, while two (Palabos, waLBerla) use GitLab.
-Of the other packages, one package (SunlightLB) uses CVS for issue tracking, and seven packages do not appear to use any issue tracking system.
+overall ranked packages have most of their issues closed. Alive packages (11 use
+Git issue tracking) have 64\% of their issues closed, while dead packages (3 use
+Git issue tracking) have 71\% of their issues closed. This information is
+presented in Table~\ref{gitrepodata}. Furthermore, 13 packages that use Git for
+issue tracking use GitHub as a version control system, while two (Palabos,
+waLBerla) use GitLab. Of the other packages, one package (SunlightLB) uses CVS
+for issue tracking, and seven packages do not appear to use any issue tracking
+system.
 
@@
 
-\subsection{Reproducibility}
+\subsection{Reproducibility} \label{SecReproducibility}
 
@@
 Section~\ref{secIntro} using the quantitative data presented in
-Section~\ref{AHPresults}, qualitative data presented in
-Section~\ref{interviewresults}, additional data from software package
-repositories and artifacts, and domain expert feedback.
+Section~\ref{AHPresults}, qualitative data from developer interviews
+(Section~\ref{SecSurvey}), additional data from software package repositories
+and artifacts, and domain expert feedback.
 
@@
 Palabos, which had a broken link to its user manual, as noted in
-Section~\ref{artifacts}.
+Section~\ref{Sec_CompareArtifacts}.
 
@@
 \subsection{Highlighted Recommendations} \label{highlightedrecommendations}
+
+Many software packages would benefit from a rewrite or reorganization of
+installation instructions. A single location for installation instructions would
+improve their maintainability and correctness. Listing compatible operating
+system and dependency versions would decrease installation time and errors, as
+would adding instructions on installing dependencies. Installation process
+errors should prompt the system to display detailed messages. Once a software
+package is installed, either an automatic validation needs to be performed or
+the user needs to be able to perform a manual validation using test examples
+that include expected output. Finally, uninstallation instructions should be
+included in the documentation. \wss{Have to decide how much to emphasizse
+installability.}  \wss{Maybe at some point should mention more modern
+installation support, since most of the projects use make, and there are other
+options.}
 
@@
 
-Section~\ref{interviewresults} noted that developers were not asked to comment
+Section~\ref{SecReproducibility} noted that developers were not asked to comment
 on the reproducibility of their source code from their requirements

### 2022-01-26 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Revisions to section 3.2 Surface Correctness and Verifiability

@@
 
-\subsection{Surface Correctness and Verifiability}
-
-Seventeen of the software packages include a requirements specification artifact
-or explicitly reference domain theory, often only the latter. Software packages
-that distribute requirements specification information, such as DL\_MESO (LBE),
-generally keep it brief and include it within other documentation. This artifact
-is often found within a user manual, on a web-page, or is mentioned in related
-publications. In the latter case the user may need to spend significant time to
-find this information.  \wss{More on requirements as a pain point is given in
-Weiss, and other papers cited by Carver (2021).  Isn't generally requirements in
-the SE sense of following a template or similar.}
+\subsection{Surface Correctness and Verifiability} \label{SecSurfCorrectAndVerifiab}
+
+Seventeen of the software packages explicitly reference domain theory, but a
+full requirements specification, in the software engineering sense
+\citep{IEEE1998, RobertsonAndRobertson1999Vol, ESA1991}, is absent from all of
+the projects. Software packages that include a subset of a requirements
+specification, such as DL\_MESO (LBE), keep the information brief and include it
+within other documents. When present, theory related information is generally
+found within a user manual, on a web-page, or in a cited publication. In the
+latter case significant time may be needed to track down the information.
 
@@
 Unit tests are only explicitly available for two of the software packages,
-Ludwig and Musubi.
-Code modularization of most packages allow for users to create tests with varying degrees of effort.
-These tests allow developers and users to verify the correctness of fragments of the source code, and in doing so better assess the correctness of the entire package.
+Ludwig and Musubi. Code modularization of most packages allow for users to
+create tests with varying degrees of effort. These tests allow developers and
+users to verify the correctness of fragments of the source code, and in doing so
+better assess the correctness of the entire package.
 
@@
 of the packages (ESPResSo, Ludwig, Musubi) mentioned applying the practice of
-continuous integration in their development process. \wss{Forward reference that
-suggests more tools use CI (Section 7 Comparison of Principles, Process and
-Methodologies to Research Software in General).}
+continuous integration in their development process.
 
@@
 
-Figure~\ref{Fig_CorrectnessVerifiability} shows the surface correctness and
-verifiability ranking of the software packages using AHP. Software packages with
-a higher score tend to have a visible requirements specification or references
-to theory documentation. They also explicitly use at least one document
-generation tool that builds confidence of correctness. The top ranked software
-packages all include an easy to follow getting started tutorial, and most of
-these include expected output.
-The two top ranked packages, Ludwig and Musubi, indicate the use of unit testing in their documentation.
-They also incorporated continuous integration in the development process.
-Furthermore, eight of the top 10 ranked packages are noted as being alive.
-
-The inclusion of requirements specification and theory documentation greatly
-benefits the correctness and verifiability of software packages. The use of
-document generation tools can help build confidence in correctness. The addition
-of easy to follow tutorials further helps users verify the software and have
-confidence in its correctness. Unit testing, as well as the use of continuous
-integration tools and techniques such as Bamboo, Jenkins, and Travis CI, help
-verify correctness.
+Figure~\ref{Fig_CorrectnessVerifiability} shows the ranking for surface
+correctness and verifiability. Software packages with a higher score tend to
+have theory documentation. They also explicitly use at least one document
+generation tool that helps with verification, thus building confidence in
+correctness. The top ranked software packages all include an easy to follow
+getting started tutorial, and most of these include expected output. The two top
+ranked packages, Ludwig and Musubi, indicate the use of unit testing in their
+documentation. They also incorporated continuous integration in the development
+process. Furthermore, eight of the top 10 ranked packages are noted as being
+alive.
 
@@
 
+\wss{Document theory (17 of the packages), but not requirements.  Might think
+ this is good enough, but a theory manual doesn't include the traceability to
+ the code.  For instance, what are the consequences of changing an assumption?}
+
+ \wss{More on requirements as a pain point is given in Weiss, and other papers
+cited by Carver (2021).  Isn't generally requirements in the SE sense of
+following a template or similar.}
+
 The top four AHP ranked packages have most of the less common artifacts. At the
@@
 
-\section{Comparison of Principles, Process and Methodologies to Research Software in General} \label{Sec_CompareMethodologies}
+Section \ref{SecSurfCorrectAndVerifiab} shows that continuous integration is
+rare in LBM.
+
+\section{Comparison of Principles, Process and Methodologies to Research
+Software in General} \label{Sec_CompareMethodologies}
 

### 2022-01-26 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Edits to Section 3.3 Surface Reliability

@@
 commands (even when following the installation guide), missing links, missing
-dependencies, and syntax errors in code files.
-In some instances the error messages were vague.
-Several automatic installation processes could not find and load dependencies.
-In these instances the installation tried to access outdated external repositories.
-Seven of the installations were recovered and verified, and one of the installations (LB3D-Prime) was assumed to be recovered due to the absence of any way to verify it.
-The installation of eight of the software packages could not be recovered.
-Most of these broken installations could not find external dependencies, encountered system incompatibilities, or displayed vague error messages.
+dependencies, and syntax errors in code files. In some instances the error
+messages were vague. Several automatic installation processes could not find and
+load dependencies. In these instances the installation tried to access outdated
+external repositories. Seven of the installations were recovered and verified,
+and one of the installations (LB3D-Prime) was assumed to be recovered due to the
+absence of any way to verify it. The installation of eight of the software
+packages could not be recovered. Most of these broken installations could not
+find external dependencies, encountered system incompatibilities, or displayed
+vague error messages.
 
@@
 Nine of the top 10 ranked packages are noted as being alive.
-
-Overall, lower ranked software packages are lacking clear documentation, testing
-or tutorial examples, and descriptive error messages, and have broken
-dependencies. Thus, regarding surface reliability, software packages would
-benefit from clear up-to-date documentation that specifies all dependencies, the
-inclusion of testing and tutorial examples, and the assurance of descriptive
-error messages after faults.
 

### 2022-01-27 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Updated section 3.7 on Maintainability, incorporated information from Peter #121

@@
 The software packages were tested for handling unexpected input, including
-incorrect data types, empty input, and missing files or links. Success
+incorrect data types, empty input, and missing files or links. Success is
 predicated on a reasonable response from the system, including appropriate error
@@
 that installed correctly passed this test. They output descriptive error
-messages or did not crash. Software packages with a lower surface robustness
-score did not install correctly, so their robustness score may not be a true
-reflection of runtime robustness. All successfully installed software packages
-that require a plain text input file correctly handled an unexpected change to
-these input files, including a replacement of new lines with carriage returns.
-Furthermore, nine of the top 10 ranked packages are noted as being alive. LIMBES
-is noted as being classified as dead.
+messages or at least do not crash. Software packages with a lower surface
+robustness score did not install correctly, so their robustness score may not be
+a true reflection of runtime robustness. All successfully installed software
+packages that require a plain text input file correctly handled an unexpected
+change to these input files, including a replacement of new lines with carriage
+returns. Furthermore, nine of the top 10 ranked packages are noted as being
+alive.
 
@@
 and are technically different from each other (as shown in
-Table~\ref{tbl_features}). Due to this, a comparison of performance is not
-appropriate. In this project we instead looked through each software package's
-artifacts for evidence that performance was considered. The artifacts of 18
-software packages mentioned parallelization. This included GPU processing and
-the CUDA parallel computing platform, which were mentioned in the artifacts of 6
-packages (ESPResSo, lbmpy, lettuce, pyLBM, Sailfish, TCLB). GPUs provide
-superior processing power and speed compared to CPUs, and are often used for
-scientific computing when a large amount of data is involved. The software
-package TCLB is implemented in a highly efficient multi-GPU code to achieve
-performance suitable for model optimization \citep{rutkowski2020open}. In the
-Ludwig package, a so-called mixed mode approach is used where fine-grained
-parallelism is implemented on the GPU, and MPI is used for even larger scale
-parallelism \citep{gray2013ludwig}. While one software package (Sailfish)
-required CUDA and GPU processing, some (ESPResSo, lbmpy, lettuce, pyLBM, TCLB)
-have the option of using either the GPU or the CPU. The packages that require
-GPU and CUDA have better performance at the expense of installability and
-surface reliability.
+Table~\ref{tbl_features}). Due to this, a comparison of run-time performance is
+not appropriate. Instead we looked through each software package's artifacts for
+evidence that performance was considered. The artifacts of 18 software packages
+mentioned parallelization. This included GPU processing and the CUDA parallel
+computing platform, which were mentioned in the artifacts of 6 packages
+(ESPResSo, lbmpy, lettuce, pyLBM, Sailfish, TCLB). When a high degree of
+parallelization is possible GPUs provide superior speed compared to CPUs. The
+software package TCLB is implemented in a highly efficient multi-GPU code to
+achieve performance suitable for model optimization \citep{rutkowski2020open}.
+In the Ludwig package, a so-called mixed mode approach is used where
+fine-grained parallelism is implemented on the GPU, and MPI is used for even
+larger scale parallelism \citep{gray2013ludwig}. While one software package
+(Sailfish) required CUDA and GPU processing, some (ESPResSo, lbmpy, lettuce,
+pyLBM, TCLB) have the option of using either the GPU or the CPU. In general, the
+packages that require GPU and CUDA have better performance at the expense of
+installability and surface reliability.
 
@@
 rendered on a web-page. Some packages (waLBerla) do not have a user manual, but
-do have useful documentation distributed on their web-pages. Expected user
-characteristics are documented in five software packages (laboetie, LIMBES,
+do have useful documentation distributed throughout their web-pages. Expected
+user characteristics are documented in five software packages (laboetie, LIMBES,
 Ludwig, Musubi, Palabos). Users are typically scientists or engineers. Their
@@
 user support model. Many packages have several user support models. Furthermore,
-four of the top five ranked packages are noted as being alive.
+four of the top five packages are noted as being alive.
 
@@
 
-Issue tracking is used in 23 software packages, 15 of which use Git, seven use
-email, and one (SunlightLB) uses SourceForge. Most software packages that use
-Git have most of their issues closed, and only three (laboetie, lettuce,
-Sailfish) have less than 50 percent of their issues closed. All of the top five
-overall ranked packages have most of their issues closed. Alive packages (11 use
-Git issue tracking) have 64\% of their issues closed, while dead packages (3 use
-Git issue tracking) have 71\% of their issues closed. This information is
-presented in Table~\ref{gitrepodata}. Furthermore, 13 packages that use Git for
-issue tracking use GitHub as a version control system, while two (Palabos,
-waLBerla) use GitLab. Of the other packages, one package (SunlightLB) uses CVS
-for issue tracking, and seven packages do not appear to use any issue tracking
-system.
+Issue tracking is used in 23 software packages, 15 of which use GitHub or GitLab
+to host their git repository, seven use email, and one (SunlightLB) uses
+SourceForge. Most software packages that use GitHub/GitLab have most of their
+issues closed, and only three (laboetie, lettuce, Sailfish) have less than 50
+percent of their issues closed. All of the top five overall ranked packages have
+most of their issues closed. Alive packages (11 use issue tracking) have 64\% of
+their issues closed, while dead packages (3 use issue tracking) have 71\% of
+their issues closed. This information is presented in Table~\ref{gitrepodata}.
+Furthermore, 13 packages that use git for version control use GitHub, while two
+(Palabos, waLBerla) use GitLab. Of the other packages, one package (SunlightLB)
+uses CVS for issue tracking, and seven packages do not appear to use any issue
+tracking system.
 
@@
 		\end{tabular}
-		\caption{Git Repository Data} \label{gitrepodata}
+		\caption{Git Repository Data, Sorted by Percentage of Code that is
+		Comments} \label{gitrepodata}
 	\end{center}
@@
 Software package code files were further measured for the percentage of code
-that is comments. The findings are presented in Table~\ref{gitrepodata}.
-Packages with a higher percentage of comments were designated as more
-maintainable. Comments represent more than 10 percent of code files in 16
-packages, and the average percentage of code comments is about 14 percent. All
-of the top five overall ranked packages have more than the average.
-LUMA has only 0.2 percent comments, the fewest of any package.
-This package has the most lines of source code, with over four million.
-The next largest package is ESPResSo++ with one million.
+that is comments. The findings are presented in Table~\ref{gitrepodata}, sorted
+by the percentage. Packages with a higher percentage of comments were assumed to be more maintainable. Comments represent more than 10 percent of code files in
+16 packages, and the average percentage of code comments is about 14 percent.
+All of the top five overall ranked packages have more than the average. LUMA has
+only 0.2 percent comments, the fewest of any package. This package has the most
+lines of source code, with over four million. The next largest package is
+ESPResSo++ with one million.
 
@@
 Figure~\ref{Fig_Maintainability} shows the maintainability ranking of the
-software packages using AHP. Software packages with a high score provide version
-numbers on documents and source code releases, have an abundance of high quality
+software packages. Software packages with a high score provide version numbers
+on documents and source code releases, have an abundance of high quality
 artifacts, and use an issue tracking tool and version control system. These
@@
 the code being comments. Furthermore, four of the top 5 ranked packages are
-noted as being alive. MP-LABS is noted as being dead.\\
-
-\wss{Sort ``Git Repository Data'' table in descending order of percentage of
-code comments.}
+noted as being alive.
 

### 2022-01-27 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Revisions to Maintainability section (Section 3.7)

@@
 their issues closed. This information is presented in Table~\ref{gitrepodata}.
-Furthermore, 13 packages that use git for version control use GitHub, while two
+With respect to the host for the git repository, 13 packages use GitHub and two
 (Palabos, waLBerla) use GitLab. Of the other packages, one package (SunlightLB)
@@
 artifacts, and use an issue tracking tool and version control system. These
-packages also appear to reasonably handle issue tracking, having most of their
-issues closed. Their code files are well commented with more than 10 percent of
-the code being comments. Furthermore, four of the top 5 ranked packages are
-noted as being alive.
+packages also appear to manage their issue tracking, having most of their issues
+closed. Their code files are well commented with more than 10 percent of the
+code being comments. Four of the top 5 ranked packages are noted as being alive.
 

### 2022-01-27 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Revisions to Section 3.8 on Reusability

@@
 
-We measured the total number of source code files for each project. A larger
-number of source files is associated with increased reusability, due to our
-assumption that this indicates increased modularization. Some packages have more
-features than others.  This is assumed to contribute to reusability, since they have
-more source code for potential reuse. The software packages were also reviewed
-for the presence of API documentation, which indicates that a software package
-was developed with interaction between other software applications in mind.
+We measured the total number of source code files for each project. We assume
+that a large number of source files is associated with increased reusability,
+due to our assumption that this indicates increased modularization. Some
+packages have more features than others.  This is assumed to contribute to
+reusability, since they have more source code for potential reuse. The software
+packages were also reviewed for the presence of API documentation, which
+indicates that a software package was developed with interaction between other
+software applications in mind.
 
@@
 Figure~\ref{Fig_Reusabilty} shows the reusability ranking of the software
-packages using AHP. Software packages with a high score have thousands of source
-code files and API documentation. The highest scoring packages, ESPResSo and
+packages. Software packages with a high score have thousands of source code
+files and API documentation. The highest scoring packages, ESPResSo and
 waLBerla, have extensive functionality, including graphical visualizations and
@@
 not on a level field. However, these packages do have an abundance of reusable
-components. Furthermore, nine of the top 10 ranked packages are noted as being
-alive. HemeLB is considered to be dead.
+components. Nine of the top 10 ranked packages are noted as being alive.
 
@@
 
-There was a strong focus on modularity when designing the waLBerla framework to
-enhance productivity, reusability, and maintainability
-\citep{bauer2021walberla}. The design of waLBerla has enabled it to be
-successfully applied in several projects as a basis for various extensions
-\citep{bauer2021walberla}.
+The package waLBerla scored high on reusability because of its focus on
+modularity.  The modularity in the waLBerla framework was included to enhance
+productivity, reusability, and maintainability \citep{bauer2021walberla}. The
+design of waLBerla has enabled it to be successfully applied in several projects
+as a basis for various extensions \citep{bauer2021walberla}.
 
@@
 
+The assumption that more code files is an indicator of reusability
+(Section~\ref{reusabilityresults}) may not be a valid assumption.
+
 \section{Conclusion} \label{conclusion}

### 2022-01-27 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Removed extra space in header for Table 5
Finished revision of Section 3 (Ranking Projects)

@@
 Figure~\ref{Fig_Installability} shows the installability ranking of the software
-packages using AHP. Software packages with a higher score (ESPResSo, MechSys,
-OpenLB, DL\_MESO, Palabos) tend to have one set of linear installation
-instructions written as if the person doing the installation has none of the
-dependencies installed. The instructions often list compatible operating system
-versions and include instructions for the installation of dependencies. The top
-ranked packages often incorporate some sort of automation of the installation
-process and have fewer manual installation steps. The number of dependencies a
-package has does not correlate with a higher score. The ability to validate the
+packages. Software packages with a higher score (ESPResSo, MechSys, OpenLB,
+DL\_MESO, Palabos) tend to have one set of linear installation instructions
+written as if the person doing the installation has none of the dependencies
+installed. The instructions often list compatible operating system versions and
+include instructions for the installation of dependencies. The top ranked
+packages often incorporate some sort of automation of the installation process
+and have fewer manual installation steps. The number of dependencies a package
+has does not correlate with a higher score. The ability to validate the
 installation process, often through tutorials or test examples that include
@@
 Figure~\ref{Fig_Reliability} shows the surface reliability ranking of the
-software packages using AHP. Software packages with a high score either did not
-break during installation, or the broken installation was recoverable. All of
-the top five ranked packages have tutorials. The package pyLBM broke
-during tutorial testing, but a descriptive error message helped in recovery.
-Nine of the top 10 ranked packages are noted as being alive.
+software packages. Software packages with a high score either did not break
+during installation, or the broken installation was recoverable. All of the top
+five ranked packages have tutorials. The package pyLBM broke during tutorial
+testing, but a descriptive error message helped in recovery. Nine of the top 10
+ranked packages are noted as being alive.
 
@@
 Figure~\ref{Fig_Robustness} shows the surface robustness ranking of the software
-packages using AHP. Software packages with a high score behaved reasonably in
-response to unexpected input as described above. All of the software packages
-that installed correctly passed this test. They output descriptive error
-messages or at least do not crash. Software packages with a lower surface
-robustness score did not install correctly, so their robustness score may not be
-a true reflection of runtime robustness. All successfully installed software
-packages that require a plain text input file correctly handled an unexpected
-change to these input files, including a replacement of new lines with carriage
-returns. Furthermore, nine of the top 10 ranked packages are noted as being
-alive.
+packages. Software packages with a high score behaved reasonably in response to
+unexpected input as described above. All of the software packages that installed
+correctly passed this test. They output descriptive error messages or at least
+do not crash. Software packages with a lower surface robustness score did not
+install correctly, so their robustness score may not be a true reflection of
+runtime robustness. All successfully installed software packages that require a
+plain text input file correctly handled an unexpected change to these input
+files, including a replacement of new lines with carriage returns. Furthermore,
+nine of the top 10 ranked packages are noted as being alive.
 
@@
 Figure~\ref{Fig_Usability} shows the surface usability ranking of the software
-packages using AHP. Software packages with a high score have a tutorial and user
-manual, sometimes have documented user characteristics, and have at least one
-user support model. Many packages have several user support models. Furthermore,
-four of the top five packages are noted as being alive.
+packages. Software packages with a high score have a tutorial and user manual,
+sometimes have documented user characteristics, and have at least one user
+support model. Many packages have several user support models. Furthermore, four
+of the top five packages are noted as being alive.
 
@@
 			\toprule
-			Name & Text Files & Binary Files & LOC & Avg. LOC / Text File\\
+			Name & Text Files & Binary Files & LOC & Avg.\ LOC / Text File\\
 			\midrule
@@
 
-\subsection{Surface Understandability}
+\subsection{Surface Understandability} \label{Sec_SurfUnderstandability}
 
@@
 
-All of the packages appear to have consistent indentation and formatting.
-Only HemeLB, LUMA, and Musubi explicitly identify coding standards that are used during development.
-Generally, the software packages use consistent, distinctive, and meaningful code identifiers.
-Only four packages (LB2D-Prime, LB3D-Prime, LIMBES, MP-LABS) appear to use vague identifiers, such as single letters for variables.
-Symbolic constants were observed in the source code of 13 packages.
-The constants are used for various parameters, mathematical constants, and matrix definitions.
-All of the packages are reasonably well commented, with the comments clearly indicating what is being done (as opposed to how it is being done).
-Domain algorithms are noted in the source code of 11 packages.
-Table~\ref{moduledata} suggests that the software packages are modularized to various degrees.
-When observing the source code files, it was found that 14 of the packages have a consistent style and order of function parameters.
+All of the packages appear to have consistent indentation and formatting. Only
+HemeLB, LUMA, and Musubi explicitly identify coding standards that are used
+during development. Generally, the software packages use consistent,
+distinctive, and meaningful code identifiers. Only four packages (LB2D-Prime,
+LB3D-Prime, LIMBES, MP-LABS) appear to use vague identifiers, such as single
+letters for variables. Symbolic constants were observed in the source code of 13
+packages. The constants are used for various parameters, mathematical constants,
+and matrix definitions. All of the packages are reasonably well commented, with
+the comments clearly indicating what is being done (as opposed to how it is
+being done). Domain algorithms are noted in the source code of 11 packages.
+Table~\ref{moduledata} suggests that the software packages are modularized to
+various degrees. When observing the source code files, it was found that 14 of
+the packages have a consistent style and order of function parameters.
 
@@
 Figure~\ref{Fig_Understandability} shows the surface understandability ranking
-of the software packages using AHP. Software packages with a high score have a
-consistent indentation and formatting style, and consistent, distinctive, and
-meaningful code identifiers. They also have symbolic constants, and explicitly
-identify mathematical and LBM algorithms. Their comments are clear and indicate
-what is being done in the source code. The source code is well modularized and
-structured. Furthermore, all of the top five ranked packages are noted as being
-alive.
+of the software packages. Software packages with a high score have a consistent
+indentation and formatting style, and consistent, distinctive, and meaningful
+code identifiers. They also have symbolic constants, and explicitly identify
+mathematical and LBM algorithms. Their comments are clear and indicate what is
+being done in the source code. The source code is well modularized and
+structured. All of the top five ranked packages are noted as being alive.
 
@@
 of release notes. The packages tended to not explicitly use well-known
-development models. This was also noted in the interviews with developers, as
-detailed below. The development teams of these packages are fairly small and
-easily organized without the need for such processes. Seven of the software
-packages did have some artifacts outlining the general development process, how
-to contribute, and the status of the package or its components. Eight of the
-packages have artifacts that note the development environment. While this
-information could help developers, and would improve transparency, the small
-close-knit nature of the development teams make explicitly publicly specifying
-this information practically unnecessary. Version release notes were found in
-nine of the software packages.
+development models. This was also noted in the interviews with developers
+(Section~\ref{Sec_CompareMethodologies}). The development teams of these
+packages are fairly small and easily organized without the need for such
+processes. Seven of the software packages did have some artifacts outlining the
+general development process, how to contribute, and the status of the package or
+its components. Eight of the packages have artifacts that note the development
+environment. While this information could help developers, and would improve
+transparency, the small close-knit nature of the development teams make
+explicitly publicly specifying this information practically unnecessary. Version
+release notes were found in nine of the software packages.
 
@@
 Figure~\ref{Fig_VisibilityTransparency} shows the visibility and transparency
-ranking of the software packages using AHP. Software packages with a high score
-have an explicit development model and defined development process. They also
-had detailed and easy to access notes accompanying software releases.
-Furthermore, four of the top five ranked packages are noted as being alive.
-MP-LABS is noted as being dead.
-
-\subsection{Overall Quality}
+ranking of the software packages. Software packages with a high score have an
+explicit development model and defined development process. They also had
+detailed and easy to access notes accompanying software releases. Four of the
+top five ranked packages are noted as being alive.
+
+\subsection{Overall Quality} \label{Sec_OverallQuality}
 
@@
 
-Figure~\ref{Fig_OverallScore} shows the overall ranking of the software packages
-using AHP. In the absence of other information on priorities, the overall
-ranking was calculated by assuming an equal weighting between all qualities. If the
+Figure~\ref{Fig_OverallScore} shows the overall ranking of the software
+packages. In the absence of other information on priorities, the overall ranking
+was calculated by assuming an equal weighting between all qualities. If the
 qualities were weighed differently, the overall software package ranking would
@@
 
-Looking at the top three ranked packages: ESPResSo had achieved a relative high
+Looking at the top three ranked packages: ESPResSo achieved a relatively high
 score in installability, surface usability, maintainability, reusability, and
@@
 
-The assumption that more code files is an indicator of reusability
+Problems with the measurement of qualities:
+
+\begin{itemize}
+\item The assumption that more code files is an indicator of reusability
 (Section~\ref{reusabilityresults}) may not be a valid assumption.
+\item Understandability was measured using only 10 random files
+(Section~\ref{Sec_SurfUnderstandability}).  This could be improved.
+\end{itemize}
+
+Overall ranking with equally weighted qualities isn't realistic (Section~\ref{Sec_OverallQuality}).
 

### 2022-01-27 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
sorted table 5 LBM SOP paper

@@
 			\midrule
+			LUMA & 314 & 19 & 4399723 & 14011\\
+			LatBo.jl & 41& 0 & 42172 & 1029\\
+			LB2D-Prime & 82& 19 & 54755 & 668\\
+			LB3D-Prime & 23& 6 & 12944 & 563\\
 			DL\_MESO (LBE) & 310 & 51 & 170223 & 549\\
+			LB3D & 99 & 76 & 39766 & 402\\
+			laboetie & 133& 1 & 48403 & 364\\
+			waLBerla & 2643 & 69 & 873988 & 331\\
+			MechSys & 333 & 3 & 95707 & 287\\
+			Palabos & 1974 & 71 & 563841 & 286\\
+			lbmpy& 250 & 29 & 61632 & 247\\
+			SunlightLB & 36 & 1 & 7646 & 212\\
+			Musubi & 1347 & 1839 & 281879 & 209\\
+			LIMBES & 26 & 1 & 4872 & 187\\
+			Ludwig & 954 & 32 & 162518 & 170\\
+			OpenLB & 1438 & 7 & 218406 & 152\\
 			ESPResSo & 1390& 83 & 195083 & 140\\
+			MP-LABS & 307 & 3 & 43124 & 140\\
+			pyLBM & 272 & 108 & 37234 & 137\\
 			ESPResSo++ & 1406& 30 & 165194 & 118\\
 			HemeLB & 1102& 44 & 123806 & 112\\
-			laboetie & 133& 1 & 48403 & 364\\		
-			LatBo.jl & 41& 0 & 42172 & 1029\\
-			LB2D-Prime & 82& 19 & 54755 & 668\\
-			LB3D & 99 & 76 & 39766 & 402\\
-			LB3D-Prime & 23& 6 & 12944 & 563\\
-			lbmpy& 250 & 29 & 61632 & 247\\
+			Sailfish & 632 & 11 & 69398 & 110\\
 			lettuce & 73 & 1 & 7660 & 105\\
-			LIMBES & 26 & 1 & 4872 & 187\\
-			Ludwig & 954 & 32 & 162518 & 170\\
-			LUMA & 314 & 19 & 4399723 & 14011\\
-			MechSys & 333 & 3 & 95707 & 287\\
-			MP-LABS & 307 & 3 & 43124 & 140\\
-			Musubi & 1347 & 1839 & 281879 & 209\\
-			OpenLB & 1438 & 7 & 218406 & 152\\
-			Palabos & 1974 & 71 & 563841 & 286\\
-			pyLBM & 272 & 108 & 37234 & 137\\
-			Sailfish & 632 & 11 & 69398 & 110\\
-			SunlightLB & 36 & 1 & 7646 & 212\\
 			TCLB & 594 & 7 & 49156 & 83\\
-			waLBerla & 2643 & 69 & 873988 & 331\\
 			\bottomrule

### 2022-01-29 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Updates to section comparing our ranking to an estimate of the community's ranking

@@
 
-Answers \rqref{RQ_CompareHQ2Popular}. Table~\ref{repometrics} presents our LBM
-software package rankings along with the repository ranking metrics of each
-software package. Nine packages do not use GitHub, so they do not have a measure
-of repository stars. Looking at the repository stars of the other 15 packages,
-we can observe a slight pattern where packages that have been highly ranked by
-our assessment do have more stars than lower ranked packages. The best ranked
-package (ESPResSo) has the second most number of stars. The ninth ranked package
-(Sailfish) has the most number of stars. Packages designated as lower quality
-often do not use GitHub or GitLab, or have few stars. HemeLB only has 12 stars,
-but its location recently moved from a different GitHub repository. Our
-assessment of this package might not be accurate. The number of stars does not
-necessarily represent the perceptions of the community, but for lack of an
-alternative measure we will use it this way. Besides missing data, another
-threat to the validity of this comparison is the varying ages of the
-repositories. Older packages have been able to accumulate stars and watches for
-longer than newer packages. The true quality of new packages may not be
-reflected in their stars and watches.
-
-The repository watches column contains even less data to compare since two of
-the packages (Palabos, waLBerla) use GitLab, which does not include this metric.
-The pattern that can be observed with this metric is very similar to that of the
-stars metric. The ninth ranked package (Sailfish) has the most number of
-watches. The best ranked package (ESPResSo) has the second most number of
-watches.
+To address \rqref{RQ_CompareHQ2Popular}, Table~\ref{repometrics} compares our
+LBM software package rankings against their popularity in the research
+community, as estimated by repository stars and watches. Nine packages do not
+use GitHub, so they do not have a measure of repository stars. Looking at the
+repository stars of the other 15 packages, we can observe a pattern where
+packages that have been highly ranked by our assessment tend to have more stars
+than lower ranked packages. Our best ranked package (ESPResSo) has the second
+most stars, while our ninth ranked package (Sailfish) has the highest number of
+stars. The same correlation is observed in the repository watch column, although
+this column contains less data, since two of the packages (Palabos, waLBerla)
+use GitLab, which does not track watches. Consistent with the star measure, our
+ninth ranked package (Sailfish) has the highest number of watches and our best
+ranked package (ESPResSo) has the second most watches. Packages designated as
+lower quality often do not use GitHub or GitLab, or have only a few stars/watches.
+Although the details of the rankings differ, our measures show that following
+best practices correlates with increased popularity in the scientific community.
 
@@
 			\toprule
-		Name & Our Ranking & Repository Stars & Repository Star Rank & Repository Watches \\
+			Name & Our Ranking & Repository Stars & Repository Star Rank &
+			Repository Watches \\
 			\midrule
@@
 
-Table~\ref{rankingcomparison} compares the top 10 ranked LBM software packages
-with a LBM package ranking made by a domain expert. Five (Palabos, OpenLB, LUMA,
-Sailfish, waLBerla) of the top 10 ranked packages in this assessment are also
-found in the domain expert's top 10 ranking. Interestingly, the first three of
-these five packages are listed in the same order on both lists. Looking at the
-remaining packages, the top two ranked packages in this state of the practice
-assessment (ESPResSo, Ludwig) are not on the domain expert's list. Perhaps the
-intended applications of these packages, complex fluids for Ludwig and soft
-matter research for ESPResSo, do not align with the research interests of the
-domain expert. This reasoning may also apply to eighth ranked Musubi. Moving
-down the list, pyLBM and DL\_MESO(LBE) did not make the domain expert's top 10
-list but were mentioned by the domain expert. Looking at the remainder of the
-domain expert's list, ASL is a general purpose tool for solving partial
-differential equations and may have fallen out of scope of the authoritative
-lists that were used to identify the initial LBM software list. The domain
-expert's fifth ranked package (ch4-project) was on the initial software list of
-this assessment, but was removed due to a lack of documentation. Open FSI is a
-new project that uses Palabos. It was made public the year before data was
-collected for this state of the practice exercise. It may not have been on
-authoritative lists due to its young age. Similarly, LIFE was also made public
-recently. Finally, the domain expert's 10th ranked package (HemeLB) was ranked
-21st in this state of the practice assessment.
-
-\begin{table}
-\begin{center}
-	\begin{tabular}{ p{2cm}p{4.5cm}p{3.5cm}}
-		\toprule
-		Rank & This Assessment & Domain Expert\\
-		\midrule
-		1 & ESPResSo (N/A) & Palabos\\
-		2 & Ludwig (N/A) & OpenLB\\
-		3 & Palabos (1) & LUMA\\
-		4 & OpenLB (2) & ASL\\
-		5 & LUMA (3) & ch4-project\\
-		6 & pyLBM (16) & Open FSI\\
-		7 & DL\_MESO (LBE) (11) & WaLBerla\\
-		8 & Musubi (N/A) & LIFE\\
-		9 & Sailfish (9) & Sailfish\\
-		10 & WaLBerla (7) & HemeLB\\		
-		\bottomrule
-	\end{tabular}
-	\caption{Ranking Comparison} \label{rankingcomparison}
-\end{center}
-\end{table}
+Although our ranking and the estimate of popularity are consistent, trying to
+make more specific conclusions from this data would be problematic.  To start
+with, using stars and watches may not accurately measure popularity or
+necessarily represent the perceptions of the community. Even if the stars and
+watches are a good proxy for popularity, they are an unfair measure since
+repositories vary in age. Older packages have had more time to accumulate stars
+and watches. For instance, HemeLB only has 12 stars, but it is young because (at
+the time of measurement) HemeLB had recently moved from a different GitHub
+repository. From the other side, our estimate of ranking by best practices is
+also an approximation.  Besides our use of shallow measures, the assumption of
+equal weighting between all qualities to create our overall ranking
+(Section~\ref{Sec_OverallQuality}) does not represent reality.  For real
+projects some qualities are going to be more important than others.
+
+Although our ranking and the estimate of the community's ranking are not perfect
+measures, they do suggest a correlation between best practices and popularity.
+We do not know which comes first, the use of best practices or popularity, but
+we do know that the top ranked packages share common practices.  The next
+sections will explore how the practices of the LBM community compare to the
+broader research software community.  We will also investigate the practices
+from the top projects that others within the LBM community, and within the
+broader research software community, can potentially adopt.
 
@@
 rare in LBM.
+
+\wss{Also compare to recommended practices for research software via the
+checklists and other articles.}
 

### 2022-01-30 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
added new column into table 6

@@
 	\begin{center}
-		\begin{tabular}{ p{3.5cm}p{2.5cm}p{2cm}p{2cm}p{2cm} }
+		\begin{tabular}{ p{3cm}p{1.25cm}p{1.75cm}p{1.75cm}p{1.75cm}p{1.75cm} }
 			\toprule
 			Name & Our Ranking & Repository Stars & Repository Star Rank &
-			Repository Watches \\
+			Repository Watches & Repository Watch Rank\\
 			\midrule
-			ESPResSo & 1 & 145 & 2 & 19\\
-			Ludwig & 2 & 27 & 8 & 6\\
-			Palabos & 3 & 34 & 6 & GitLab\\
-			OpenLB & 4 & No Git & No Git & No Git\\
-			LUMA & 5 & 33 & 7 & 12\\
-			pyLBM & 6 & 95 & 3 & 10\\
-			DL\_MESO (LBE) & 7 & No Git & No Git & No Git \\
-			Musubi & 8 & No Git & No Git & No Git \\
-			Sailfish & 9 & 186 & 1 & 41\\
-			waLBerla & 10 & 20 & 9 & GitLab\\
-			laboetie & 11 & 4 & 13 & 5\\
-			TCLB & 12 & 95 & 3 & 16\\
-			MechSys & 13 & No Git & No Git & No Git\\
-			lettuce & 14 & 48 & 4 & 5\\
-			ESPResSo++ & 15 & 35 & 5 & 12\\
-			MP-LABS & 16 & 12 & 11 & 2\\			
-			SunlightLB & 17 & No Git & No Git & No Git\\
-			LB3D & 18 & No Git & No Git & No Git\\			
-			LIMBES & 19 & No Git & No Git & No Git\\
-			LB2D-Prime & 20 & No Git & No Git & No Git\\		
-			HemeLB & 21 & 12 & 11 & 12\\
-			lbmpy & 22 &  11 & 12 & 2\\	
-			LB3D-Prime & 23 & No Git & No Git & No Git\\	
-			LatBo.jl & 24 & 17 & 10 & 8\\			
+			ESPResSo & 1 & 145 & 2 & 19& 2\\
+			Ludwig & 2 & 27 & 8 & 6& 7\\
+			Palabos & 3 & 34 & 6 & GitLab& GitLab\\
+			OpenLB & 4 & No Git & No Git & No Git& No Git\\
+			LUMA & 5 & 33 & 7 & 12& 4\\
+			pyLBM & 6 & 95 & 3 & 10& 5\\
+			DL\_MESO (LBE) & 7 & No Git & No Git & No Git & No Git\\
+			Musubi & 8 & No Git & No Git & No Git & No Git\\
+			Sailfish & 9 & 186 & 1 & 41& 1\\
+			waLBerla & 10 & 20 & 9 & GitLab& GitLab\\
+			laboetie & 11 & 4 & 13 & 5& 8\\
+			TCLB & 12 & 95 & 3 & 16& 3\\
+			MechSys & 13 & No Git & No Git & No Git& No Git\\
+			lettuce & 14 & 48 & 4 & 5& 8\\
+			ESPResSo++ & 15 & 35 & 5 & 12& 4\\
+			MP-LABS & 16 & 12 & 11 & 2& 9\\			
+			SunlightLB & 17 & No Git & No Git & No Git& No Git\\
+			LB3D & 18 & No Git & No Git & No Git& No Git\\			
+			LIMBES & 19 & No Git & No Git & No Git& No Git\\
+			LB2D-Prime & 20 & No Git & No Git & No Git& No Git\\		
+			HemeLB & 21 & 12 & 11 & 12& 4\\
+			lbmpy & 22 &  11 & 12 & 2& 9\\	
+			LB3D-Prime & 23 & No Git & No Git & No Git& No Git\\	
+			LatBo.jl & 24 & 17 & 10 & 8& 6\\			
 			\bottomrule

### 2022-01-31 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Incorporating additional information into the comparison b/w our ranking and the community's ranking

@@
 incorporating additional physical complexities \citep{chen1998lattice}, such as
-reflective and non-reflective boundaries. LBM considers the behavior of a
+reflective and non-reflective boundaries. LBM considers the behaviour of a
 collection of particles as a single unit at the mesoscopic scale, which lies
@@
 A small sample of important applications of LBM include the following: designing
-fuel cells \citep{ZhangEtAl2018}, modeling groundwater flow
+fuel cells \citep{ZhangEtAl2018}, modelling groundwater flow
 \citep{AnwarAndSukop2009}, and analyzing the flow of blood in the cardiovascular
@@
 2022: Musubi.  Musubi did not come up in our initial search, but once we learned
-of it, we felt that this report is incomplete without including it.
+of it, we felt that this report is incomplete without including it.  With the
+addition of Musubi we did not update the manual collection of data for the
+previous packages, due to time constraints, but we did redo all data collection
+that was automated by tools.
 
@@
 waLBerla, have extensive functionality, including graphical visualizations and
-non-LBM modeling. For this reason, a comparison with other software packages is
+non-LBM modelling. For this reason, a comparison with other software packages is
 not on a level field. However, these packages do have an abundance of reusable
@@
 ranked package (ESPResSo) has the second most watches. Packages designated as
-lower quality often do not use GitHub or GitLab, or have only a few stars/watches.
-Although the details of the rankings differ, our measures show that following
-best practices correlates with increased popularity in the scientific community.
+lower quality often do not use GitHub or GitLab, or have only a few
+stars/watches. Although the details of the rankings differ, our measures show
+that following best practices tends to correlate with increased popularity in
+the scientific community.
+
+Notable exceptions to this trend occur for three projects that rank high by star
+count, but relatively low in the ``follow best practices'' ranking: TCLB (star
+rank 3, our rank 12), lettuce (star rank 4, our rank 14) and ESPResSo++ (star
+rank 5, our rank 15).  Part of the discrepancy is because all projects are
+ranked by our process, but not all projects are measured by stars.  Also, our
+process equally weights all qualities (Section~\ref{Sec_OverallQuality}), but
+for real users some qualities will be more important than others. Discrepancies
+between our measures and the star measures may also be due to inaccuracy with
+using stars to approximate popularity, since stars and watches tend to increase
+over time. Older packages (like ESPResSo++ and TCLB) have had more time to
+accumulate stars and watches. Young projects have less time to accumulate stars,
+like HemeLB, which only 12 stars (at the time of measurement) having recently
+moved from a different GitHub repository.  Lettuce is a young project (released
+in 2019) with a relatively high number of stars.  The discrepancy in rankings
+here may be due to a mismatch between the time when lettuce was manually
+measured by our template and when the stars were counted.  The late addition of
+Musubi meant a large gap between when projects were manually measured and the
+time when our automated data (like stars) was collected.
 
@@
 
-Although our ranking and the estimate of popularity are consistent, trying to
-make more specific conclusions from this data would be problematic.  To start
-with, using stars and watches may not accurately measure popularity or
-necessarily represent the perceptions of the community. Even if the stars and
-watches are a good proxy for popularity, they are an unfair measure since
-repositories vary in age. Older packages have had more time to accumulate stars
-and watches. For instance, HemeLB only has 12 stars, but it is young because (at
-the time of measurement) HemeLB had recently moved from a different GitHub
-repository. From the other side, our estimate of ranking by best practices is
-also an approximation.  Besides our use of shallow measures, the assumption of
-equal weighting between all qualities to create our overall ranking
-(Section~\ref{Sec_OverallQuality}) does not represent reality.  For real
-projects some qualities are going to be more important than others.
-
 Although our ranking and the estimate of the community's ranking are not perfect
@@
 We do not know which comes first, the use of best practices or popularity, but
-we do know that the top ranked packages share common practices.  The next
-sections will explore how the practices of the LBM community compare to the
+we do know that the top ranked packages tend to incorporate best practices.  The
+next sections will explore how the practices of the LBM community compare to the
 broader research software community.  We will also investigate the practices
@@
 and an understanding of the requirements for formulating problems in the
-individual packages, which can be a significant endeavor. These software
+individual packages, which can be a significant endeavour. These software
 packages are not like commercial software packages. They are generally set up to
@@
 integration in the development processes than what was observed from the initial
-survey. For example, OpenLB, pyLBM, and TCLB use such methods during development,
-despite this not being explicitly clear from an analysis of the material
-available online. \wss{Add this to threats to validity.}  The correctness and
-verifiability of such packages is not measured well using surface analysis.
+survey. For example, OpenLB, pyLBM, and TCLB use such methods during
+development, despite this not being explicitly clear from an analysis of the
+material available online. \wss{Add this to threats to validity.}  The
+correctness and verifiability of such packages is not measured well using
+surface analysis.
 
@@
 package to stimulate the development of third-user extensions. The Palabos
-package identifies as a development framework for modeling problems in various
+package identifies as a development framework for modelling problems in various
 CFD areas.  \wss{Add validation benchmarks as a ``best practice'' used by
@@
 in this assessment, but it is probable that such documentation is part of all
-LBM packages. After all, developing SCS is not a trivial endeavor. It is likely
+LBM packages. After all, developing SCS is not a trivial endeavour. It is likely
 that many packages have such documentation but did not make it public, and due
@@
 Overall ranking with equally weighted qualities isn't realistic (Section~\ref{Sec_OverallQuality}).
+
+Threat to validity - the gap between measures of everything else and Musubi
+(Section~\ref{identifysoftware}), and then the gap between the manual measures
+and the automated measures (Section~\ref{repmetrics}).
 

### 2022-02-02 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Initial edits to Section 5 (Comparison of Artifacts to Research Software in General)

@@
 
-Answers \rqref{RQ_CompareArtifacts}.  The software packages were examined for
-the presence of artifacts, which were then categorized by frequency. We have
+In this section we answers \rqref{RQ_CompareArtifacts} by comparing the
+artifacts that we observed in LBM repositories to those observed and recommended
+for research software in general.  roadmap?
+
+As part of filling in the measurement template the LBM packages were examined
+for the presence of artifacts, which were then categorized by frequency. We have
 grouped them into common, less common, and rare artifacts in
@@
 Authors / Developers List & Change Log / Release Notes & API Documentation\\
-Bug Tracker & Design Documentation & Developer / Contributor Manual\\
+Issue Tracker & Design Documentation & Developer / Contributor Manual\\
 Dependency Notes & Functional Spec. / Notes & FAQ / Forum\\
@@
 Validation Plan\\
-Requirements Spec. / Theory Notes & Test Plan / Script / Cases & Video Guide
+Theory Notes & Test Plan / Script / Cases & Video Guide
 (including YouTube)\\
-List of Related Publications & User Manual/Guide & \\
+List of Related Publications & User Manual/Guide & Requirements Spec.\\
 Makefile / Build File & Version Control & \\
@@
 \end{tabular}
-\caption{Artifacts Present in Surveyed Packages, Classified by Frequency}
+\caption{Artifacts Present in LBM Packages, Classified by Frequency}
 \label{artifactspresent}
@@
 All of the top four AHP ranked packages, ESPREsSo, Ludwig, Palabos, and OpenLB,
-have each of the commonly found artifacts, except only three of them have a
-requirements specification or theory notes. \wss{Change to clarify that it is
-theory notes, not a requirements specification.  Remove requirements spec from
-table and put it as rare.}  Palabos is the only one of these four packages that
-does not have an artifact from this category.  \wss{** Requirements **}
+have each of the commonly found artifacts, except Palabos does not have any
+theory notes.  The top four AHP ranked packages have most of the less common
+artifacts. At the time of data collection, only one of the four packages
+(Palabos) did not have a user manual or guide, but there was a broken link on
+the package website indicating that such an artifact might exist. This broken
+link was later fixed, but this is not reflected in our data because it was not
+present at the time of initial data collection (mid 2020). Despite the broken
+link, Palabos does have a detailed and informative website. Another one (Ludwig)
+of the top four packages does not appear to have publicly visible design
+documentation. A third package (OpenLB) from the list does not appear to use a
+version control system. It is possible that such a system is used by the
+package, since its website notes package version numbers, but the artifacts do
+not explicitly state the use of such a system.
+
+The top four ranked packages do not have many of the rare artifacts. None of the
+top four packages have any explicit API documentation. Three of these packages
+(ESPREsSo, Ludwig, Palabos) have information on contributing to the project. Two
+of them (OpenLB, Palabos) have a FAQ section or forum. One (OpenLB) has
+verification and validation notes, and a video guide of the software.
+
+\wss{Change to clarify that it is theory notes, not a requirements
+specification.  Remove requirements spec from table and put it as rare.}
+
+\wss{** Requirements **}
 \wss{Discuss requirements here. Extra information from Peter: I interpreted it
@@
 
+ \wss{Kelly paper about how researchers document their theory.  Close to being enough - just need traceability and other things.}
+
  \wss{More on requirements as a pain point is given in Weiss, and other papers
@@
 
-The top four AHP ranked packages have most of the less common artifacts. At the
-time of data collection, only one of the four packages (Palabos) did not have a
-user manual or guide, but there was a broken link on the package website
-indicating that such an artifact might exist. This broken link was later fixed,
-but this is not reflected in our data because it was not present at the time of
-data collection. Despite the broken link, Palabos does have a detailed and
-informative website. Another one (Ludwig) of the top four packages does not
-appear to have publicly visible design documentation. A third package (OpenLB)
-from the list does not appear to use a version control system. It is possible
-that such a system is used by the package, since its website notes package
-version numbers, but the artifacts do not explicitly state the use of such a
-system.
-
-The top four ranked packages do not have many of the rare artifacts. None of the
-top four packages have any explicit API documentation. Three of these packages
-(ESPREsSo, Ludwig, Palabos) have information on contributing to the project. Two
-of them (OpenLB, Palabos) have a FAQ section or forum. One (OpenLB) has
-verification and validation notes, and a video guide of the software.
-
 \wss{Weiss mentions pain points related to design documentation.  Carver (2021) 
 also mentions design documentation in the literature review.}
+
+\wss{README files discussed in PranaEtAl2018.}
+
+\wss{At some point should get into the intended use of the software.  The
+artifacts will change depending on the intention for the software.  Maybe just
+say at the start that writing as if all software is intended for research
+readiness?}
 
@@
 and comparing, 24 software packages along quality attributes. The software
-qualities that were assessed in this report are listed in Section
-~\ref{softwarequalities}. A methodology for assessing the state of the practice
-of software development in SCS domains was presented. Domain packages were
-assessed to answer the software development related research questions listed in
-Section~\ref{secIntro} to understand how software quality is impacted by software
-development choices, including principles, processes, and tools. Software
-developers were interviewed to identify development pain points, and to identify
-how software quality is ensured. Quantitative data was used to rank the software
-packages using the AHP. The ranking designations were compared with rankings
-from the software development community, and we found that many of our top 10
-ranked packages are ranked highly by a domain expert. Recommendations for
+qualities that were assessed in this report are listed in
+Section~\ref{softwarequalities}. A methodology for assessing the state of the
+practice of software development in SCS domains was presented. Domain packages
+were assessed to answer the software development related research questions
+listed in Section~\ref{secIntro} to understand how software quality is impacted
+by software development choices, including principles, processes, and tools.
+Software developers were interviewed to identify development pain points, and to
+identify how software quality is ensured. Quantitative data was used to rank the
+software packages using the AHP. The ranking designations were compared with
+rankings from the software development community, and we found that many of our
+top 10 ranked packages are ranked highly by a domain expert. Recommendations for
 improving software along quality metrics were made, and highlights are presented

### 2022-02-03 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Additional notes added to Section 5.

@@
 
-In this section we answers \rqref{RQ_CompareArtifacts} by comparing the
-artifacts that we observed in LBM repositories to those observed and recommended
-for research software in general.  roadmap?
-
 As part of filling in the measurement template the LBM packages were examined
@@
 of the software packages. Rare artifacts were found in 1 to 7 ($<$30\%) of the
-software packages.
+software packages. In this section we answers \rqref{RQ_CompareArtifacts} by
+comparing the artifacts that we observed in LBM repositories to those observed
+and recommended for research software in general.
 
@@
 (including YouTube)\\
-List of Related Publications & User Manual/Guide & Requirements Spec.\\
+List of Related Publications & User Manual / Guide & Requirements Spec.\\
 Makefile / Build File & Version Control & \\
@@
 readiness?}
+
+\wss{From Peter: Roadmap - I did see this mentioned, but no specific artifacts
+for it. (``rare'' mentions) Code of conduct, policies for team members - do not
+recall ever seeing this Citation - definitely saw this several times (``less
+common'') Programming style guide - I did see this as part of developers guide
+(``rare'') description of team processes - I do recall seeing ``workflow'' detailed
+(``rare''). Not sure if this is similar. peer review - This was mentioned a few
+times in the interviews, but I do not recall seeing artifacts for this.
+checklists - I did see this a few times (``rare'') linters - I did see this
+(``rare'') uninstall - I do not recall seeing this.}
 

### 2022-02-07 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Additional edits to Section 5 on comparing LBM to research software for artifacts

@@
 Seventeen of the software packages explicitly reference domain theory, but a
-full requirements specification, in the software engineering sense
+full and rigorous requirements specification, in the software engineering sense
 \citep{IEEE1998, RobertsonAndRobertson1999Vol, ESA1991}, is absent from all of
@@
 the software packages that are missing this key information. In these cases the
-user may need to assume correctness if there are no visible errors.
+user may need to assume correctness if there are no visible errors.  A
+particularly  detailed tutorial that include expected output is provided by
+\href{https://www.walberla.net/doxygen/index.html}{waLBerla}.
 
@@
 
-\section{Comparison of Artifacts to Research Software in General}
+\section{Comparison Between LBM and Research Software for Artifacts}
 \label{Sec_CompareArtifacts}
@@
 comparing the artifacts that we observed in LBM repositories to those observed
-and recommended for research software in general.
+and recommended for research software in general. Our comparison may point out
+areas where some LBM software packages fall short of current best practices.
+This is not intended to be a criticism of any existing packages, especially
+since in practice not every project needs to achieve the highest possible
+quality.  However, rather than delve into the nuances of which software can
+justify compromising which practices we will write our comparison under the
+ideal assumption that every project has sufficient resources to match best
+practices.
 
@@
 Issue Tracker & Design Documentation & Developer / Contributor Manual\\
-Dependency Notes & Functional Spec. / Notes & FAQ / Forum\\
+Library Dependency List & Functional Spec. / Notes & FAQ / Forum\\
 Installation Guide / Instructions & Performance Information & Verification and
@@
 of them (OpenLB, Palabos) have a FAQ section or forum. One (OpenLB) has
-verification and validation notes, and a video guide of the software.
-
-\wss{Change to clarify that it is theory notes, not a requirements
-specification.  Remove requirements spec from table and put it as rare.}
-
-\wss{** Requirements **}
-\wss{Discuss requirements here. Extra information from Peter: I interpreted it
-to mean what the software should do. As noted in my report, this was not common,
-especially finding any formal document like an SRS. I do not recall any package
-having an SRS. Some packages did note in their manuals or artifacts some of the
-points that would be found in an SRS document. For example, DL\_MESO notes
-functional (``The following is a list of the features that DLMESO currently
-supports....'') and non-functional (should work in both ``serial and parallel
-running; it can be run on standalone machines, clusters and supercomputers..'')
-requirements in its documentation (more on this below). Yes, there was often
-some sort of a theory resource. Often it was simply a link or reference to the
-theory used in the package. There was significant discrepancy in how much theory
-was documented or referenced. Meanwhile, requirements were rare. A few packages
-did note some of this information within other artifacts but it was not
-comprehensive, and it was not presented in a SRS document. DL\_MESO has some of
-the information that belongs in an SRS in its user manual. It is not
-comprehensive or presented as it would be in an SRS, but it was probably the
-best example of requirements. It has a nice subsection on Functionality.
-\citet{Nguyen-HoanEtAl2010} showed requirements are the least commonly produced
-type of documentation for scientific software in general.  LBM matches this
-trend.}
-
-\wss{Document theory (17 of the packages), but not requirements.  Might think
- this is good enough, but a theory manual doesn't include the traceability to
- the code.  For instance, what are the consequences of changing an assumption?}
-
- \wss{Kelly paper about how researchers document their theory.  Close to being enough - just need traceability and other things.}
+verification and validation notes, and a video guide for the software.
+
+The majority of LBM generated artifacts correspond to general recommendations
+from research software developers.  A union of the three columns in
+Table~\ref{artifactspresent} mostly corresponds to recommendations made by the
+research software community.  Most of the items in Table~\ref{artifactspresent}
+are included in guidelines for writing research software, such as the DLR
+(German Aerospace Centre) Software Engineering Guidelines
+\citep{TobiasEtAl2018}, xSDK (Extreme-scale Scientific Software Development Kit)
+Community Package Policies \citep{SmithAndRoscoe2018}, the EURISE (European
+Research Infrastructure Software Engineers') Network Technical Reference
+\cite{ThielEtAl2020} and CLARIAH (Common Lab Research Infrastructure for the
+Arts and Humanities) Guidelines for Software Quality \citep{vanGompelEtAl2016}.
+As an example, LBM developers list the external libraries on which the software
+depends as recommended by \citet{BrettEtAl2021}, \citet{SmithAndRoscoe2018} and
+\citet{vanGompelEtAl2016}. Although in the rare column, some LBM developers
+include an FAQ (Frequently Asked Questions) document as recommended by
+\citep{OrvizEtAl2017}, \citep{ThielEtAl2020} and \citep{vanGompelEtAl2016}. LBM
+developers also follow the ubiquitous recommendation to have a README file,
+although, as for other software, the content of README files is not standardized
+\citep{PranaEtAl2018}.
+
+Three of the items that appear in Table~\ref{artifactspresent} do not explicitly
+appear in the software development guidelines: i) list of related publications,
+ii) performance information and iii) video guides.  The occurrence of the first
+two items in this list are likely more a consequence of how we collected our
+list of LBM artifacts, rather than being a unique practice of LBM developers. We
+highlighted publications and performance information because we were explicitly
+looking for both as part of filling in our measurement template. Related
+publications and performance information certainly appear in other research
+software projects; the practices just aren't highlighted in guidelines because
+they are implicit in other documentation recommendations, such as theory
+documentation and user guides. The observation of a video guide artifact is more
+novel.  Other projects likely use videos, but the practice has apparently not
+yet made it into community guidelines.  Putting aside the debate of the
+effectiveness of learning from print versus video, video documentation is
+growing in popularity, with a majority of Generation Z learners preferring video
+to print \citep{Genota2018}.
+
+Although the LBM community participates in most of the practices we found listed
+in general research software guidelines, some recommended practices were not observed or rarely observed.
+In addition to the items in the last column of Table~\ref{artifactspresent}, we can add the following community recommended items that we rarely, if ever, observed:
+
+\begin{itemize}
+\item A roadmap showing what is planned for the future \citep{Yo2021,
+ThielEtAl2020, vanGompelEtAl2016}.  We did see this information mentioned in
+some repos, but we did not observe a specific artifact devoted to this purpose.
+\item A code of conduct to explicitly say how developers should treat one
+another \citep{Yo2021, ThielEtAl2020}.  We did not observe a policy of this
+sort, possibly because the number of developers for each LBM project is small.
+\item Programming style guidelines so that programming labels and formatting are
+consistent \citep{TobiasEtAl2018, ThielEtAl2020, OrvizEtAl2017, Zadka2018,
+vanGompelEtAl2016, WilsonEtAl2014}.  We saw this information as part of
+developer guides, but only rarely.
+\item Checklists can be used in projects to ensure that best practices are
+followed by all developers.  Some examples include checklists merging branches
+into master \citep{Brown2015}, checklists for saving and sharing changes to the
+project \citep{WilsonEtAl2016}, checklists for new and departing team members
+\citep{HerouxAndBernholdt2018}, checklists for processes related to commits and
+releases \citep{HerouxEtAl2008} and checklists for overall software quality
+\citep{ThielEtAl2020, SSI2022}.  We did see evidence of checklists in the LBM
+projects, but they were used only rarely.
+\item Although not observed for LBM projects, having instructions to uninstall
+the software is part of best practices \citep{vanGompelEtAl2016}.
+\end{itemize}
+
+The above discussion shows that, taken together, LBM projects fall short of
+recommended practices for research software.  However, LBM software is not alone
+in this.  Info on req, design and testing.  \citet{Nguyen-HoanEtAl2010} showed
+requirements are the least commonly produced type of documentation for
+scientific software in general.  Carver points out do not do requirements.  Not
+everyone recommends requirements - can find citations.  Feasible to do
+requirements.  Theory documents are already common (17 of 24? packages).  Need
+to add just a little more information.
+
+\wss{Document theory (17 of the packages), but not requirements.  Opportunity to
+ go a little further and have a more complete requirements specificaiton. Add
+ traceability to the code.  For instance, what are the consequences of changing
+ an assumption?  adding an abstract transformation from theorty to code - when
+ and how.  Example have a PDE, sate use finite difference. Consequence of change
+ - likely and unlikely changes - give example of 2D vs 3D.  Basic idea is to
+ give information on some simple things to add that would make a theory document
+ like a requirements document. }
 
@@
 
-\wss{README files discussed in PranaEtAl2018.}
-
-\wss{At some point should get into the intended use of the software.  The
-artifacts will change depending on the intention for the software.  Maybe just
-say at the start that writing as if all software is intended for research
-readiness?}
-
-\wss{From Peter: Roadmap - I did see this mentioned, but no specific artifacts
-for it. (``rare'' mentions) Code of conduct, policies for team members - do not
-recall ever seeing this Citation - definitely saw this several times (``less
-common'') Programming style guide - I did see this as part of developers guide
-(``rare'') description of team processes - I do recall seeing ``workflow'' detailed
-(``rare''). Not sure if this is similar. peer review - This was mentioned a few
-times in the interviews, but I do not recall seeing artifacts for this.
-checklists - I did see this a few times (``rare'') linters - I did see this
-(``rare'') uninstall - I do not recall seeing this.}
+In Table~\ref{artifactspresent} a rigorous requirements specification is rare,
+but parts of a full specification do occur, with theory notes being common and
+portions of the functional specification (statements about what the software
+does) being less common.  Discuss requirements stuff.
 
@@
 
-Answers \rqref{RQ_CompareToolsProjMngmnt}.  Software tools are used to support
-the development, verification, maintenance, and evolution of software, software
-processes, and artifacts \citep{ghezzi1991fundamentals}. Many tools are used by
-LBM software packages, as summarized in Table~\ref{tbl_tools}.  The tools are
-subdivided into development tools, dependencies, and project management tools.
+Software tools are used to support the development, verification, maintenance,
+and evolution of software, software processes, and artifacts
+\citep{ghezzi1991fundamentals}. Many tools are used by LBM software packages, as
+summarized in Table~\ref{tbl_tools}.  The tools are subdivided into development
+tools, dependencies, and project management tools.  In this section we answers
+\rqref{RQ_CompareToolsProjMngmnt} by comparing the tool usage in LBM software to
+the utilization in the research software community in general.
 
@@
 Section \ref{SecSurfCorrectAndVerifiab} shows that continuous integration is
-rare in LBM.
+rare in LBM.  Recommended by many \citep{vanGompelEtAl2016} etc.
 
@@
 interviewed noted that an ad hoc peer review process is used to assess major
-changes and additions.
+changes and additions.  Peer review (also known as code review and technical
+review) is a frequently recommended component of developing research software
+\citep{HerouxEtAl2008, Givler2020, OrvizEtAl2017, USGS2019}.
 
@@
 from peers and through self-study, not from formal training
-\citep{HannayEtAl2009}.
-The thinking of the developer on the role of formal education matches the trend \citet{PintoEtAl2018} observes, where in their replication of a previous study \citep{HannayEtAl2009}, they show a growing interest in formal training (From 13\% of respondants in 2009 to 22\% in 2018).}
+\citep{HannayEtAl2009}.  The thinking of the developer on the role of formal
+education matches the trend \citet{PintoEtAl2018} observes, where in their
+replication of a previous study \citep{HannayEtAl2009}, they show a growing
+interest in formal training (From 13\% of respondants in 2009 to 22\% in 2018).}
 
@@
 
-``The questions of what constitutes an adequate testing oracle for scientific software and what constitutes adequate test data are left open.
-Some of the validation testing we researched was running up against the practical limits; they were using whatever oracle they could scrape together from industry data, benchmarks, and expert judgment.'' \citep{KellyAndSanders2008}
+``The questions of what constitutes an adequate testing oracle for scientific
+software and what constitutes adequate test data are left open. Some of the
+validation testing we researched was running up against the practical limits;
+they were using whatever oracle they could scrape together from industry data,
+benchmarks, and expert judgment.'' \citep{KellyAndSanders2008}
 
@@
 
-practices by the best measured.
+practices by the best measured.  Point to existing checklist and development
+guides.  This section should only mention things that are new.
 
 \begin{itemize}
-	\item The top projects provide great examples of achieving installability,
-	by taking such steps as: providing complete installation instructions,
-	writing instructions as if the user does not have any of the dependencies
-	installed, automation of the installation process (for instance using make),
-	including descriptive error message, and including steps for uninstallation.
-	\item Surface Correctness and Verifiability. Make public (on package website
-	or GitHub) the requirements specification document, or explicitly reference
-	the domain theory that the software is designed from.  This was done by
-	several top ranked packages, including ESPResSo, Ludwig, LUMA, and OpenLB.
-	\item Surface Correctness and Verifiability. Ensure the above information is
-	easy to find. Consider adding it to the user manual. The information is easy
-	to find in most top ranked packages.
 	\item Surface Correctness and Verifiability. Development teams should
@@
 	suggestion is based on developer comments.
-	\item Surface Correctness and Verifiability. Traceability. Visibility and
-	Transparency. Use and make public (on package website or GitHub) detailed
-	documentation. Consider using automatic document generation tools like
-	Doxygen, Drasil, or Sphinx. This was done by all of the top five ranked
-	packages.
-	\item Surface Correctness and Verifiability. Provide detailed tutorials that
-	include expected output, like the
-	\href{https://www.walberla.net/doxygen/index.html}{waLBerla tutorials}.
-	\item Surface Correctness and Verifiability. Use unit tests during
-	development and make them public (on package website or GitHub). This was
-	available for Ludwig.
-	\item Surface Correctness and Verifiability. Modularize the source code,
+	\item Surface Correctness and Verifiability. Surface Understandability.
+	Maintainability.  Modifiability. Reusability.  Modularize the source code,
 	separate components, hide information behind well defined interfaces. This
@@
 	Wilson's best practices.}
-	\item Surface Understandability. Maintainability.  Modifiability.
-	Reusability. Modularize the source code, separate components, hide
-	information behind well defined interfaces. This is suggested in SCS
-	literature, and in developer comments. \wss{add citation - Wilson mentions -
-	this points seems repetitive}
-	\item Surface Correctness and Verifiability. Use continuous integration
-	tools (Bamboo, Jenkins, and Travis CI) and processes during development.
-	This was done by top ranked packages ESPResSo and Ludwig.  \wss{Common in
-	LBM - more common than in other domains}
-	\item Surface Usability. State appropriate fluid dynamics problems that the
-	software is designed to model in the documentation, and explicitly state the
-	limits of the software. This is done in the ESPResSo user guide.
+	\item Surface Usability. Surface Understandability. Unambiguity. State
+	appropriate fluid dynamics problems that the software is designed to model
+	in the documentation, and explicitly state the limits of the software. This
+	is done in the ESPResSo user guide.
 	\item Surface Usability. Provide documentation that details the background
@@
 	done by top ranked packages.
-	\item Surface Usability. Maintain a user support model (Git, email, forum,
-	FAQ). This was done by all of the top five ranked packages.
-	\item Surface Usability. If possible, consider using popular user-friendly
-	software languages like Python. Especially consider this for parts of the
-	source code that is likely to be modified or reviewed by users. ESPResSo and
-	Sailfish use Python to shorten development time and improve usability.
-	\item Maintainability. Include version numbers and release notes for all
-	major source code and artifact releases. The top five ranked software
-	packages include release notes.
 	\item Maintainability. Have a defined process for accepting contributions,
@@
 	top five ranked software packages include information on how to contribute. 
-	\item Maintainability. Use an issue tracker (Git, email, SourceForge, other)
-	to manage bugs and changes. Issues should be regularly reviewed and closed.
-	All but three (laboetie, lettuce, Sailfish) of the software packages that
-	use Git have most of their issues closed.
-	\item Maintainability. Use a version control system (GitHub, CVS). Four
-	(ESPREsSo, Ludwig, LUMA, Palabos) of the top five ranked packages use a
-	version control system.
-	\item Maintainability. Source code needs to be well commented. Typically,
-	more than 10 percent of LBM package source code is comments, as presented in
-	Table~\ref{gitrepodata}. Four of the top five overall ranked packages
-	(Ludwig, ESPResSo, Palabos, OpenLB) have about 20 percent of their source
-	code as comments.
-	\item Maintainability. If possible, consider using popular user-friendly
-	software languages like Python. Especially consider this for parts of the
-	source code that is likely to be modified or reviewed by users. ESPResSo and
-	Sailfish use Python to address several software qualities, including
-	maintainability.
 	\item Modifiability. Consider flexibility of data structures and data
@@
 	coding standards. 
-	\item Surface Understandability. Add meaningful comments. Indicate what is
-	being done in each section of source code. All of the top ranked packages
-	have meaningful comments in their source code.
-	\item Surface Understandability. State appropriate fluid dynamics problems
-	that the software is designed to model in the documentation, and explicitly
-	state the limits of the software. This is done in the ESPResSo user guide. 
-	\item Surface Understandability. Consider adding a FAQ section to the
-	documentation. This helped resolve some usability issues for OpenLB.
 	\item Traceability. Provide a developer's guide to help orient developers to
@@
 	development process. Eleven packages have information on how to contribute. 
-	\item Visibility and Transparency. Identify the development environment.
-	Eight of the 24 software packages identify the development environment.
-	\item Visibility and Transparency. Include notes with all releases. Nine of
-	the packages include release notes.
-	\item Visibility and Transparency. Communicate all changes within the
-	development team. This suggestion is based on developer comments.
-	\item Visibility and Transparency. Use continuous integration processes and
-	tools (Bamboo, Jenkins, and Travis CI). This was done by top ranked packages
-	ESPResSo and Ludwig.
 	\item Visibility and Transparency. Consider peer review processes to assess
@@
 	inspection to assess contributions. High ranked package ESPResSo does this.
-	\item Unambiguity. State appropriate fluid dynamics problems that the
-	software is designed to model in the documentation, and explicitly state the
-	limits of the software. This is done in the ESPResSo user guide.
-	\item Unambiguity. The documentation should either explain the underlying
-	CFD theories or provide a reference to appropriate resources. This was done
-	by several top ranked packages, including ESPResSo, Ludwig, LUMA, and
-	OpenLB.
 	\item Unambiguity. Consider asking users for feedback on the documentation.
@@
 
+linters \citep{Zadka2018} were rarely observed in LBM projects.  Do not see much evidence of their presence in best practice guides.
+
 \section{Threats To Validity} \label{threats}
@@
 
+Our comparison may point out areas where some LBM software packages fall short
+of current best practices.  This is not intended to be a criticism of any
+existing packages, especially since in practice not every project needs to
+achieve the highest possible quality.  However, rather than delve into the
+nuances of which software can justify compromising which practice we will write
+our comparison under the ideal assumption that every project can have sufficient
+resources to match best practices.
+
 \subsection{Highlighted Recommendations} \label{highlightedrecommendations}
@@
 	pyLBM.
-	\item Keep all documentation in one location. This was done by top ranked packages.
+	\item Keep all documentation in one location. This was done by top ranked
+	packages.
 	\item If possible, consider using popular user-friendly software languages
@@
 governance and communication processes
+
+- should look at ReinerEtAl2022 on Domain of Ocean Modelling software to
+contrast to the state of the practice of LBM software

### 2022-02-08 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Completed the edits of Section 5 (Comparison Between LBM and Research Software for Artifacts)

@@
 
-All of the top four AHP ranked packages, ESPREsSo, Ludwig, Palabos, and OpenLB,
-have each of the commonly found artifacts, except Palabos does not have any
-theory notes.  The top four AHP ranked packages have most of the less common
-artifacts. At the time of data collection, only one of the four packages
-(Palabos) did not have a user manual or guide, but there was a broken link on
-the package website indicating that such an artifact might exist. This broken
-link was later fixed, but this is not reflected in our data because it was not
-present at the time of initial data collection (mid 2020). Despite the broken
-link, Palabos does have a detailed and informative website. Another one (Ludwig)
-of the top four packages does not appear to have publicly visible design
-documentation. A third package (OpenLB) from the list does not appear to use a
-version control system. It is possible that such a system is used by the
-package, since its website notes package version numbers, but the artifacts do
-not explicitly state the use of such a system.
+The top four packages, ESPREsSo, Ludwig, Palabos, and OpenLB, have all of the
+common artifacts, except Palabos does not have theory notes.  The top four
+packages also have most of the less common artifacts. At the time of data
+collection, only one of the four packages (Palabos) did not have a user manual
+or guide, but there was a broken link on the package website indicating that
+such an artifact might exist. This broken link was later fixed, but this is not
+reflected in our data because it was not present at the time of data collection
+(mid 2020). Despite the broken link, Palabos does have a detailed and
+informative website. As far as we could observe, Ludwig does not appear to have
+publicly visible design and OpenLB does not appear to use a version control
+system. However, OpenLB's use of version numbers suggests that version control
+may be used privately by the developers.
 
@@
 Although the LBM community participates in most of the practices we found listed
-in general research software guidelines, some recommended practices were not observed or rarely observed.
-In addition to the items in the last column of Table~\ref{artifactspresent}, we can add the following community recommended items that we rarely, if ever, observed:
+in the general research software guidelines, some recommended practices were not
+observed or rarely observed. In addition to the items in the last column of
+Table~\ref{artifactspresent}, we can add the following community recommended
+items that we rarely, if ever, observed:
 
@@
 vanGompelEtAl2016, WilsonEtAl2014}.  We saw this information as part of
-developer guides, but only rarely.
+some developer guides, but only rarely.
 \item Checklists can be used in projects to ensure that best practices are
@@
 releases \citep{HerouxEtAl2008} and checklists for overall software quality
-\citep{ThielEtAl2020, SSI2022}.  We did see evidence of checklists in the LBM
-projects, but they were used only rarely.
-\item Although not observed for LBM projects, having instructions to uninstall
-the software is part of best practices \citep{vanGompelEtAl2016}.
+\citep{ThielEtAl2020, SSI2022}.  For LBM software, ESPResSo has a checklist for
+managing releases, but otherwise they are used only rarely for LBM.
+\item Uninstall instructions \citep{vanGompelEtAl2016} were not observed for any
+of the LBM projects.
 \end{itemize}
@@
 The above discussion shows that, taken together, LBM projects fall short of
-recommended practices for research software.  However, LBM software is not alone
-in this.  Info on req, design and testing.  \citet{Nguyen-HoanEtAl2010} showed
-requirements are the least commonly produced type of documentation for
-scientific software in general.  Carver points out do not do requirements.  Not
-everyone recommends requirements - can find citations.  Feasible to do
-requirements.  Theory documents are already common (17 of 24? packages).  Need
-to add just a little more information.
-
-\wss{Document theory (17 of the packages), but not requirements.  Opportunity to
- go a little further and have a more complete requirements specificaiton. Add
- traceability to the code.  For instance, what are the consequences of changing
- an assumption?  adding an abstract transformation from theorty to code - when
- and how.  Example have a PDE, sate use finite difference. Consequence of change
- - likely and unlikely changes - give example of 2D vs 3D.  Basic idea is to
- give information on some simple things to add that would make a theory document
- like a requirements document. }
-
-\wss{More on requirements as a pain point is given in Weiss, and other papers
-cited by Carver (2021).  Isn't generally requirements in the SE sense of
-following a template or similar.}
-
-\wss{Weiss mentions pain points related to design documentation.  Carver (2021) 
-also mentions design documentation in the literature review.}
-
-In Table~\ref{artifactspresent} a rigorous requirements specification is rare,
-but parts of a full specification do occur, with theory notes being common and
-portions of the functional specification (statements about what the software
-does) being less common.  Discuss requirements stuff.
+recommended best practices for research software.  However, LBM software is not
+alone in this.  Many, if not most, research projects fall short of best
+practices.  Software requirements documentation provides a good example of the
+common deviation between recommendations and practice.  Although requirements
+documentation is recommended by some \citep{TobiasEtAl2018, HerouxEtAl2008,
+SmithAndKoothoor2016}, in practice scientific software developers often do not
+produce a proper requirements specification \citep{HeatonAndCarver2015}.
+\citet{SandersAndKelly2008} interviewed 16 scientists from 10 disciplines and
+found that none of the scientists created requirements specifications, unless
+regulations in their field mandated such a document. \citet{Nguyen-HoanEtAl2010}
+showed requirements are the least commonly produced type of documentation for
+scientific software in general. When looking at the pain points for research
+software developers, \citet{WieseEtAl2019} found that software requirements and
+management is the software engineering discipline that most hurts scientific
+developers, accounting for 23 \% of the technical problems reported by study
+participants.  The lack of support for requirements is likely due to the
+perception that up-front requirements are impossible for research software
+\citep{CarverEtAl2007,SegalAndMorris2008}, but when the instance on ``up-front''
+requirements is dropped, allowing the requirements to be written iteratively and
+incrementally, requirements are feasible \citep{Smith2016}.
+
+Requirements documentation is feasible, as shown by the prevalence of theory
+documents in LBM software.  Table~\ref{artifactspresent} shows that theory
+documentation is common.  We observed 17 of 24 packages had at least some theory
+documentation. Templates for scientific requirements, like \citet{SmithEtAl2007,
+SmithAndLai2005}, show theory is a significant part of the requirements
+documentation. With the addition of some extra information, the theory documents
+can be transformed into requirements specification.  The key extra information
+includes explicitly stating user characteristics, explicitly stating how the
+user interacts with the software in terms of input data requirements, and
+listing likely and unlikely changes.  Explicit statements about likely changes
+are invaluable in the design stage, since they provide developers guidance on
+how general the software needs to be.  For those wishing to maximize the value
+of a requirement specification, information could be added like prioritizing the
+nonfunctional requirements (to show the relative importance between qualities
+like portability, reliability and performance) and traceability information (to
+show the consequences of any changes to the assumptions).
 
@@
 
+\wss{Beyond software carpenty has data on the increasing use of version control.}
+
 Section \ref{SecSurfCorrectAndVerifiab} shows that continuous integration is
 rare in LBM.  Recommended by many \citep{vanGompelEtAl2016} etc.
-
-\wss{Also compare to recommended practices for research software via the
-checklists and other articles.}
 

### 2022-02-09 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Updated the number of packages that use version control to be consistent throughout the doc

@@
 
-\subsection{Maintainability}
+\subsection{Maintainability} \label{Sec_Maintainability}
 
@@
 (Palabos, waLBerla) use GitLab. Of the other packages, one package (SunlightLB)
-uses CVS for issue tracking, and seven packages do not appear to use any issue
-tracking system.
+uses CVS for issue tracking and version control, and seven packages do not
+appear to use any issue tracking system.
 
@@
 List of Related Publications & User Manual / Guide & Requirements Spec.\\
-Makefile / Build File & Version Control & \\
+Makefile / Build File &  & \\
 README File & & \\
@@
 Tutorial & & \\
+Version Control & & \\
 \bottomrule
@@
 The top four packages, ESPREsSo, Ludwig, Palabos, and OpenLB, have all of the
-common artifacts, except Palabos does not have theory notes.  The top four
-packages also have most of the less common artifacts. At the time of data
-collection, only one of the four packages (Palabos) did not have a user manual
-or guide, but there was a broken link on the package website indicating that
-such an artifact might exist. This broken link was later fixed, but this is not
-reflected in our data because it was not present at the time of data collection
-(mid 2020). Despite the broken link, Palabos does have a detailed and
+common artifacts, except Palabos does not have theory notes and only three of
+them (ESPREsSo, Ludwig, Palabos) appear to use a version control system.  The
+top four packages also have most of the less common artifacts. At the time of
+data collection, only one of the four packages (Palabos) did not have a user
+manual or guide, but there was a broken link on the package website indicating
+that such an artifact might exist. This broken link was later fixed, but this is
+not reflected in our data because it was not present at the time of data
+collection (mid 2020). Despite the broken link, Palabos does have a detailed and
 informative website. As far as we could observe, Ludwig does not appear to have
@@
 
-\section{Comparison of Tool Usage to Research Software in General}
+\section{Comparison of Tool Usage Between LBM and Other Research Software}
 \label{Sec_CompareTools}
@@
 Software tools are used to support the development, verification, maintenance,
-and evolution of software, software processes, and artifacts
-\citep{ghezzi1991fundamentals}. Many tools are used by LBM software packages, as
+and evolution of software, software processes, and artifacts \citep[p.\
+501]{GhezziEtAl2003}. Many tools are used by LBM software packages, as
 summarized in Table~\ref{tbl_tools}.  The tools are subdivided into development
 tools, dependencies, and project management tools.  In this section we answers
-\rqref{RQ_CompareToolsProjMngmnt} by comparing the tool usage in LBM software to
-the utilization in the research software community in general.
+\rqref{RQ_CompareToolsProjMngmnt} by comparing the tool usage in LBM to
+their utilization in the research software community in general.
 
@@
 part of them, unlike dependencies that remain in the application once it is
-released \citep{ghezzi1991fundamentals}. Although not shown in
+released \citep[p.\ 506]{GhezziEtAl2003}. Although not shown in
 Table~\ref{tbl_tools}, debuggers were also likely used.  Only three (ESPResSo,
@@
 Open MPI, PyZMQ). Domain specific libraries include scientific computing
-libraries (e.g.\ SciPy). Some used libraries will be missing, if they are not
-explicitly stated in the artifacts we observed.
+libraries (e.g.\ SciPy).
 
@@
 project management tools are used, but they were not visible in the artifacts to
-which we had access.  Collaboration tools are noted as being used when
-developing the software projects. Most often email and video conferencing is
-used. Project management software was not explicitly mentioned, but it is
-possible that some of the projects use such software. Many of the projects are
-located on GitHub, and its developers use the platform to help manage their
-projects, especially bug related issues. Most of the projects appear to use
-change tracking and version control tools. \wss{**Version Control**} \wss{This
-contrasts with work from a little over 10 years ago that shows version control
-was only used in 50\% of research software projects \citep{Nguyen-HoanEtAl2010}.
-At that time \citet{Nguyen-HoanEtAl2010} noted an increase from previous levels
-of version control software use.  Anecdotally this confirms our observations.}
+which we had access.  Collaboration tools, most often email and video, are noted
+as being used when developing the software projects. Project management software
+was not explicitly mentioned, but it is possible that some of the projects use
+such software. Many of the projects are located on GitHub, where the developers
+use the platform to help manage their projects, especially bug related issues.
+Most of the projects appear to use change tracking and version control tools.
 They often use GitHub or GitLab for this. One package (SunlightLB) uses CVS.
-Document generation tools are mentioned in the artifacts of 12 of the projects.
-The tools Sphinx and Doxygen are explicitly used in this capacity. \wss{look at
-what the other papers say are commonly used tools and bring that information
-into this section.}
-
-\wss{Beyond software carpenty has data on the increasing use of version control.}
+Document generation tools are mentioned in 12 of the 24 projects. The tools
+Sphinx and Doxygen are explicitly used in this capacity.
+
+The poor adoption of version control tools that Wilson lamented in 2006
+\citep{Wilson2006} has greatly improved in the intervening years.  From
+Section~\ref{Sec_Maintainability} 67\% of LBM packages use version control
+(GitHub, GitLab and CVS).  The proliferation of version control tools for LBM
+matches the increase in the broader research software community.  A little over
+10 years ago (in 2010) version control was estimated to be used in only 50\% of
+research software projects \citep{Nguyen-HoanEtAl2010}, but even at that time
+\citet{Nguyen-HoanEtAl2010} noted an increase from previous usage levels.  A
+survey in 2018 shows 81\% of developers use a version control system
+\citep{AlNoamanyAndBorghi2018}.  This matches with \citet{Smith2018} shows
+active projects in mesh generation, geographic information systems and
+statistical software for psychiatry increasing from 75\%, 89\% and 17\%
+(respectively) to 100\%, 95\% and 100\% (respectively) over a four year period
+ending in 2018.  (For completeness the same study showed a decrease in version
+control tool usage for seismology software over the same time period (from 41\%
+down to 36\%).  Almost every software guide cited in
+Section~\ref{Sec_CompareArtifacts} mentions the advice to use version control.
 

### 2022-02-10 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Revisions up to and including Section 7: Comparison of Principles, Process and Methodologies to Research Software in General

@@
 documentation and user guides. The observation of a video guide artifact is more
-novel.  Other projects likely use videos, but the practice has apparently not
-yet made it into community guidelines.  Putting aside the debate of the
+novel.  \citet{Fogel2005} recommends videos for open source projects, and other
+projects likely use them, but the practice has apparently not yet made it into
+research software community's guidelines.  Putting aside the debate of the
 effectiveness of learning from print versus video, video documentation is
@@
 in the general research software guidelines, some recommended practices were not
-observed or rarely observed. In addition to the items in the last column of
+observed or rarely observed. For instance API documentation was rarely observed
+for LBM software, but it is a frequently recommended artifact
+\citep{SmithAndRoscoe2018, ThielEtAl2020, vanGompelEtAl2016, OrvizEtAl2017,
+SSI2022, Zadka2018}.  In addition to the items in the last column of
 Table~\ref{artifactspresent}, we can add the following community recommended
@@
 summarized in Table~\ref{tbl_tools}.  The tools are subdivided into development
-tools, dependencies, and project management tools.  In this section we answers
-\rqref{RQ_CompareToolsProjMngmnt} by comparing the tool usage in LBM to
-their utilization in the research software community in general.
+tools, dependencies, and project management tools.  In this section we answer
+\rqref{RQ_CompareToolsProjMngmnt} by comparing aspects of the tool usage in LBM
+to their utilization in the research software community in general.
 
@@
 Most of the projects appear to use change tracking and version control tools.
-They often use GitHub or GitLab for this. One package (SunlightLB) uses CVS.
 Document generation tools are mentioned in 12 of the 24 projects. The tools
@@
 \citep{Wilson2006} has greatly improved in the intervening years.  From
-Section~\ref{Sec_Maintainability} 67\% of LBM packages use version control
-(GitHub, GitLab and CVS).  The proliferation of version control tools for LBM
-matches the increase in the broader research software community.  A little over
-10 years ago (in 2010) version control was estimated to be used in only 50\% of
+Section~\ref{Sec_Maintainability}, 67\% of LBM packages use version control
+(GitHub, GitLab or CVS).  (From Table~\ref{gitrepodata}, 11/15 or 73\% of alive
+packages use version control.)  The proliferation of version control tools for
+LBM matches the increase in the broader research software community.  A little
+over 10 years ago version control was estimated to be used in only 50\% of
 research software projects \citep{Nguyen-HoanEtAl2010}, but even at that time
@@
 survey in 2018 shows 81\% of developers use a version control system
-\citep{AlNoamanyAndBorghi2018}.  This matches with \citet{Smith2018} shows
-active projects in mesh generation, geographic information systems and
-statistical software for psychiatry increasing from 75\%, 89\% and 17\%
-(respectively) to 100\%, 95\% and 100\% (respectively) over a four year period
-ending in 2018.  (For completeness the same study showed a decrease in version
-control tool usage for seismology software over the same time period (from 41\%
-down to 36\%).  Almost every software guide cited in
-Section~\ref{Sec_CompareArtifacts} mentions the advice to use version control.
-
-Section \ref{SecSurfCorrectAndVerifiab} shows that continuous integration is
-rare in LBM.  Recommended by many \citep{vanGompelEtAl2016} etc.
-
-\section{Comparison of Principles, Process and Methodologies to Research
-Software in General} \label{Sec_CompareMethodologies}
-
-Answers \rqref{RQ_CompareMethodologies}.  Most of the software packages do not
-explicitly state in their artifacts the motivations or design principles that
-were considered when developing the software. One package, Sailfish, indicates
-in its artifacts that shortening the development time was considered in early
-stages of design, with the developers opting for using Python and CUDA/OpenCL to
-achieve this without sacrificing any computational performance. The goals of
-that project are explicitly listed as performance, scalability, agility and
-extendability, maintenance, and ease of use. The project scored well in these
-categories during our assessment.
-
-Processes, like methods, are ways of doing things, especially in an orderly way;
-while methodologies are defined as systems of methods
-\citep{ghezzi1991fundamentals}. It is not explicitly indicated in the artifacts
-of most of the packages that development involved following any specific model,
-like a waterfall or agile development model. One developer (ESPResSo) noted that
-while no formal model is used, their development model is something similar to a
-combination of agile and waterfall development models. The developer teams of
-the LBM packages are fairly small, so it is feasible for them to be organized
-without the need for such models.  \wss{some repetition from the previous
-section - should try to figure out how to remove the repetition.}
-
-Seven of the software packages contain artifacts recording the general
-development process and status. Eleven of the packages explicitly convey that
-they would accept outside contributors, but generally the teams are centralized,
-often working at the same institution.
-
-The developers that were interviewed all noted similar project management
-processes. In teams of only a couple of developers, additions of new features or
-major changes are discussed with the entire team. Projects with more than a
-couple developers have lead developer roles. These lead developers review
-potential additions to the software. One of the developers (ESPResSo) that was
-interviewed noted that an ad hoc peer review process is used to assess major
-changes and additions.  Peer review (also known as code review and technical
-review) is a frequently recommended component of developing research software
-\citep{HerouxEtAl2008, Givler2020, OrvizEtAl2017, USGS2019}.
-
-Thirteen (54\%) of the 24 software packages use GitHub for managing the project,
-including nine (60\%) of the 15 alive packages, and four of the nine (44\%) dead
-packages. Two projects (Palabos, WaLBerla) use GitLab. This could be indicative
-of a transition to such software development and version control tools for SCS.
-Typically there are several simultaneous development branches in these projects.
-
-Documentation was also noted as playing a significant role in the development
-process, specifically with on-boarding new developers. A goal of documentation
-is to lower the entry barrier for these new contributors. The documentation
-provides information on how to get started, orients the user to artifacts and
-the source code, and explains how the system works, including the so-called
-simulation engine and interface. The use of document generation tools is
-mentioned in the artifacts of 13 software packages, and was noted during
-interviews with developers. Sphinx and Doxygen are the tools that were
-mentioned.
+\citep{AlNoamanyAndBorghi2018}.  \citet{Smith2018} has similar results, showing
+version control usage for alive projects in mesh generation, geographic
+information systems and statistical software for psychiatry increasing from
+75\%, 89\% and 17\% (respectively) to 100\%, 95\% and 100\% (respectively) over
+a four year period ending in 2018.  (For completeness the same study showed a
+decrease in version control usage for seismology software over the same time
+period, from 41\% down to 36\%).  Almost every software guide cited in
+Section~\ref{Sec_CompareArtifacts} includes the advice to use version control.
+The high usage of version control tools in LBM software matches the trend in
+research software in general.
+
+As mentioned in Section~\ref{SecSurfCorrectAndVerifiab}, continuous integration
+is rarely used in LBM (3 of 24 packages or 12.5\%). This contrasts with the
+frequency with which continuous integration is recommended in research software
+development guidelines \citep{BrettEtAl2021, Brown2015, ThielEtAl2020,
+Zadka2018, vanGompelEtAl2016}.  We could not find published data on the
+frequency with which continuous integration is used in general research
+software.  Our impression is that like LBM software, other research software
+packages lag behind the recommended best practice of employing continuous
+integration tools.
+
+Interviews with the developers revealed a potentially more frequent use of both
+unit testing and continuous integration, compared to what was observed from
+studying the repository artifacts.
+
+\section{Comparison of Principles, Process and Methodologies to Research Software in General} \label{Sec_CompareMethodologies}
+
+This section answers research question \rqref{RQ_CompareMethodologies} by
+comparing the principles, processes and methodologies used for LBM software to
+what can be gleaned from the literature on research software in general. In our
+data collection for LBM software, the software development process is not
+explicitly indicated in the artifacts for most of the packages. However, during
+our interviews one developer (ESPResSo) told us their non-rigorous development
+model is like a combination of agile and waterfall. Employing a loosely defined
+process makes sense for LBM software, given that the teams are generally small
+and self-contained.  Although eleven of the packages explicitly convey that they
+would accept outside contributors, generally the teams are centralized, often
+working at the same institution.  Working at the same institution means that an
+informal process can show success, since informal conversations are relatively
+easy to have.
+
+Interviews with developers confirmed a similar project management processes. In
+teams of only a couple of developers, additions of new features or major changes
+are discussed with the entire team. Projects with more than a couple developers
+have lead developer roles. These lead developers review potential additions to
+the software. One of the developers (ESPResSo) that was interviewed noted that
+an ad hoc peer review process is used to assess major changes and additions.
+Using peer review (also called technical review) matches with recommended
+practice for research software \citep{HerouxEtAl2008, Givler2020, OrvizEtAl2017,
+USGS2019}.
 
@@
 former changes. Software bugs are typically addressed in a similar fashion as
-code refactoring, and issue tracking is commonly used to manage these changes.
-
-Interviews with the developers of software packages also revealed a more
-frequent use of both unit testing and continuous integration in the development
-process than was found by only assessing the artifacts. The use of automatic
-installation processes is also common. Most often this involved a Make script.
+code refactoring.  Issue tracking is commonly used to manage these changes.
+
+Our observations of a informally defined process, with elements of agile
+methods, matches what has been observed for research software in general.
+Scientific developers naturally use an agile philosophy \citep{AckroydEtAl2008,
+CarverEtAl2007, EasterbrookAndJohns2009, Segal2005, HeatonAndCarver2015}, or an
+amethododical process \citep{Kelly2013}, or a knowledge acquisition driven
+process \citep{Kelly2015}.
+
+Most of the software packages do not explicitly state the motivations or design
+principles that were considered when developing the software. One package,
+Sailfish, indicates in its artifacts that shortening the development time was
+considered in early stages of design, with the developers using Python and
+CUDA/OpenCL to achieve this without sacrificing performance. The Sailfish goals
+are explicitly listed as performance, scalability, agility and extendability,
+maintenance, and ease of use. The project scored well in these categories during
+our assessment.  The quality priorities for Sailfish roughly match the
+priorities observed for research software in general.
+\citet{Nguyen-HoanEtAl2010} surveyed developers to find the following list of
+qualities, in decreasing order of importance: reliability, functionality,
+maintainability, availability, performance, flexibility, testability, usability,
+reusability, traceability, and portability. The Sailfish list does not list
+reliability or functionality, but we can safely assume those are implicitly high
+priorities for any scientific project.  In earlier studies
+\citet{KellyAndSanders2008} and \citet{CarverEtAl2007} both highlight how
+important correctness is for research software.
+
+During our interviews, documentation was noted as playing a significant role in
+the development process, specifically with on-boarding new developers. A goal of
+documentation is to lower the entry barrier for these new contributors. The
+documentation provides information on how to get started, orients the user to
+artifacts and the source code, and explains how the system works, including the
+so-called simulation engine and interface.  This emphasis on documentation,
+especially for new developers, is echoed in research software guidelines.
+Multiple guidelines recommend a document explaining how to contribute to a
+project, often named CONTRIBUTING \citep{Yo2021, BrettEtAl2021, WilsonEtAl2016,
+ThielEtAl2020, vanGompelEtAl2016, OrvizEtAl2017, FLOSS2022, JimenezEtAl2017}.
+Tutorials \citep{ThielEtAl2020}, trouble shooting guides \citep{OrvizEtAl2017,
+SSI2022} and quick start examples \citep{ThielEtAl2020, vanGompelEtAl2016} are
+also recommended.  \citep{SmithAndRoscoe2018} suggests including instructions
+specifically for on-boarding new developers. For open source software in
+general, \citep{Fogel2005} recommends providing tutorial style examples,
+developer guidelines, demos and screenshots.
 
@@
 
+From Section~\ref{Sec_CompareTools}: ``Interviews with the developers revealed a
+potentially more frequent use of both unit testing and continuous integration,
+compared to what was observed from studying the repository artifacts.''
+
 \section{Conclusion} \label{conclusion}

### 2022-02-10 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Addition of some notes for planning the revisions of the next sections

@@
 also recommended.  \citep{SmithAndRoscoe2018} suggests including instructions
-specifically for on-boarding new developers. For open source software in
-general, \citep{Fogel2005} recommends providing tutorial style examples,
-developer guidelines, demos and screenshots.
+specifically for on-boarding new developers. For open source software in general
+(not just research software), \citep{Fogel2005} recommends providing tutorial
+style examples, developer guidelines, demos and screenshots.
 
@@
 
-Answers \rqref{RQ_PainPoints} and \rqref{RQ_ComparePainPoints}.  This subsection
-answers the research question: What are the pain points for developers working
-on research software projects? What aspects of the existing processes,
-methodologies and tools do they consider as potentially needing improvement? How
-should processes, methodologies and tools be changed to improve software
-development and software quality? Developers were asked to comment on obstacles
-in their development process, obstacles encountered by users, and potential
-future obstacles.
+This section aims to answer the research questions: What are the pain points for
+developers working on research software projects (\rqref{RQ_PainPoints})?; and,
+How do the pain points of developers from LBM compare to the pain points for
+research software in general (\rqref{RQ_ComparePainPoints})?
+
+Roadmap.  Go through pain points, link to existing literature, then section on
+pain points not mentioned.  Ways to address pain points in the next section
+
+Maybe just do a table of pain points?  and then contrast the pain points with
+the literature.  The mitigation strategies should be elsewhere.  Some of the
+mitigation strategies have already been mentioned, like documentation.
+
+Likely possible to combine pain points and developer concerns.  Emphasizes what
+the pain points are, and how they compare to other studies.  Don't propose
+solutions in this section.  Say that all of the information comes from the
+interviews (as mentioned at the start of the next section).
 
@@
 know each other from other projects. External contributions are rare despite
-many of the projects accepting them. \wss{Other projects cite the same point
+many of the projects accepting them. \wss{Other projects cite the same pain
 point - Wiese paper}  Aside from on-boarding new developers, time constraints
@@
 interest in formal training (From 13\% of respondants in 2009 to 22\% in 2018).}
-
-Poor software documentation is also mentioned by \citet{PintoEtAl2018} and
-\citet{LethbridgeEtAl2003}.  The survey from \citet{LethbridgeEtAl2003} is for
-software in general, not just research software.
 
@@
 underlying theory are answered in the documentation. OpenLB has an artifact for
-such questions.  \wss{Weise mentions documentation.}
+such questions.  \wss{Weise mentions documentation.}  \wss{Partly solve the
+problem by documentation user characteristics, to explicitly state what the user
+is assumed to already know.}
 
@@
 that is further discussed in Section~\ref{highlightedrecommendations}.
+
+Poor software documentation is also mentioned by \citet{PintoEtAl2018} and
+\citet{LethbridgeEtAl2003}.  The survey from \citet{LethbridgeEtAl2003} is for
+software in general, not just research software.
 
@@
 projects use it, a reasonable next step for projects that have not yet
-incorporated CI.}
+incorporated CI.}  \wss{There is definitely a great deal of repetition.}
 
@@
 tutorials.  Maybe highlight the example of Musubi?  Should trace back to this
-for the recommendations part.}
+for the recommendations part.  \citep{Fogel2005} gives advice on having a
+successful project, also the web article (if needed).}
 
@@
 nightmare''. Furthermore, the package was designed with flexible data structures
-and storage in mind.
+and storage in mind.  \wss{example suggests identifying likely changes.}
 

### 2022-02-14 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Revisions do Section 8 - Developer pain points

@@
 \newcommand{\rqref}[1]{RQ\ref{#1}}
+
+\newcounter{pnum} %pain point number
+\newcommand{\ppthepnum}{P`'\thepnum}
+\newcommand{\ppref}[1]{P\ref{#1}}
 
@@
 gain additional insight, we need to interview developers.  We need to learn
-their concerns, how they deal with these concerns, and what pain points exist
-for them.  We wish to know what practices are used by the top LBM projects, so
-that others can potentially emulate these practices. We also wish to identify
-new practices by borrowing successful ideas from other domains. For LBM
-developers interested in future cutting edge practices, ideas that are proposed,
-but not yet in practice, should also be identified.  The above points are
-covered by the questions outlined below:
+their concerns, how they deal with these concerns; we need to learn what pain
+points exist for them.  We wish to know what practices are used by the top LBM
+projects, so that others can potentially emulate these practices. We also wish
+to identify new practices by borrowing successful ideas from other domains. For
+LBM developers interested in future cutting edge practices, ideas that are
+proposed, but not yet in practice, should also be identified.  The above points
+are covered by the questions outlined below:
 
@@
 	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Concerns}:] For LBM
-	developers:
-	\begin{enumerate} 
-		\item [\rqref{RQ_Concerns}.a] what software development concerns exist; and, (Section~\ref{Sec_Concerns})
-		\item [\rqref{RQ_Concerns}.b] what specific practices are taken to
-		address these concerns? (Section~\ref{Sec_AddressConcerns})
-	\end{enumerate}
+	developers what specific practices are taken to address the pain point
+	concerns? (Section~\ref{Sec_AddressConcerns})
 	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Recommend}:]
@@
 	in~\rqref{RQ_PainPoints} and the concerns identified
-	in~\rqref{RQ_Concerns}.a? (Section~\ref{Sec_AddressConcerns})
+	in~\rqref{RQ_Concerns}? (Section~\ref{Sec_AddressConcerns})
 
@@
 meeting software (either Zoom or Teams), which facilitated recording and
-automatic transcription of the meetings.  The interview process was approved by
-the McMaster University Research Ethics Board under the application number 
+automatic transcription of the meetings.  Each meeting lasted about an hour. The
+interview process was approved by the McMaster University Research Ethics Board
+under the application number 
 \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/MACREM/Application.pdf}
@@
 
-This section aims to answer the research questions: What are the pain points for
-developers working on research software projects (\rqref{RQ_PainPoints})?; and,
-How do the pain points of developers from LBM compare to the pain points for
-research software in general (\rqref{RQ_ComparePainPoints})?
-
-Roadmap.  Go through pain points, link to existing literature, then section on
-pain points not mentioned.  Ways to address pain points in the next section
-
-Maybe just do a table of pain points?  and then contrast the pain points with
-the literature.  The mitigation strategies should be elsewhere.  Some of the
-mitigation strategies have already been mentioned, like documentation.
-
-Likely possible to combine pain points and developer concerns.  Emphasizes what
-the pain points are, and how they compare to other studies.  Don't propose
-solutions in this section.  Say that all of the information comes from the
-interviews (as mentioned at the start of the next section).
-
-Pain points that others, but not LBM developers, mentioned in surveys include
-\citep{PintoEtAl2018}: Cross-platform compatibility, interruptions while coding,
-scope bloat, lack of user feedback, mismatch between coding skills and
-subject-matter skills (for instance, ``I am a mathematician, not an expert on
-software''), lack of formal reward system, hard to collaborate on software
-projects, aloneness.
-
-\subsection{Lack of Development Time}
-
-A developer of pyLBM noted that their small development team has a lack of time
-to implement new features. Small development teams are common for LBM software
-packages. Team members are almost always part of the same institute or already
-know each other from other projects. External contributions are rare despite
-many of the projects accepting them. \wss{Other projects cite the same pain
-point - Wiese paper}  Aside from on-boarding new developers, time constraints
-could be mitigated by increasing developer efficiency, which could be addressed
-in several ways, including by improving the quality of documentation, or
-incorporating automatic code generation. \wss{Potential ways to address the pain
-points might go into a separate section?  Maybe propose the strategy, like code
-generation, and then list the pain points that are addressed?}
-
-Lack of time was also mentioned by \citet{PintoEtAl2018} and
-\citet{PintoEtAl2016}.
-
-\subsection{Lack of Software Development Experience}
-
-A lack of software development experience was noted by the developer of TCLB.
-Many of the team members on their project are domain experts and there can be a
-steep learning curve before these members contribute good quality source code.
-This problem has been mitigated by re-writing the code to more easily enable
-future contributions.  On a related point, improving the software engineering
-education or experience of developers is also an idea that was brought up by
-several developers.  \wss{The Wiese paper and others talk about the background
-of developers. The knowledge for developing research software primarily comes
-from peers and through self-study, not from formal training
-\citep{HannayEtAl2009}.  The thinking of the developer on the role of formal
-education matches the trend \citet{PintoEtAl2018} observes, where in their
-replication of a previous study \citep{HannayEtAl2009}, they show a growing
-interest in formal training (From 13\% of respondants in 2009 to 22\% in 2018).}
-
-\subsection{Lack of Incentive and Funding}
-
-The TCLB developer noted a lack of incentive and funding in academia for
-developing widely used scientific software. The importance of funding for
-scientific software has been discussed in \cite{gewaltig2012quality}, which
-notes that ``software tools are developed and maintained only for as long as
-there is explicit or implicit funding''. The developer further commented that
-there are no journals that publish such scientific software source code.
-However, there are ways to get such source code cited. Work has been done to
-address this in \citep{smith2016software}, which presents a set of software
-citation principles and discusses ``how they could be used to implement software
-citation in the scholarly community'' \citep{katz2019software}.  \wss{other
-papers also mention this pain point.  Solutions proposed by Katz, journal of
-open source etc.}
-
-\subsection{Lack of External Support}
-
-Another raised concern was that there are no organizations helping with the
-development of good quality software; but some do exist, including
-\href{https://bssw.io/}{Better Scientific Software (BSSw)},
-\href{https://www.software.ac.uk/}{Software Sustainability Institute}
+Based on interviews with 4 developers, this section aims to answer the research
+questions: i) What are the pain points for developers working on research
+software projects (\rqref{RQ_PainPoints})?; and, ii) How do the pain points of
+developers from LBM compare to the pain points for research software in general
+(\rqref{RQ_ComparePainPoints})?  Below we go through each of the identified pain
+points and include citations that contrast the LBM experience with observations
+from researchers in other domains.  Potential ways to address the pain points
+are covered in Section~\ref{Sec_AddressConcerns}. The full interview questions
+are found in \citet{SmithEtAl2021}.
+
+\citep{PintoEtAl2018} lists some pain points that did not come up in our
+conversations with LBM developers: Cross-platform compatibility, interruptions
+while coding, scope bloat, lack of user feedback, hard to collaborate on
+software projects, and aloneness. \citep{WieseEtAl2019} repeat some of the
+previous pain points and add the following: dependency management, data handling
+concerns (like data quality, data management and data privacy), reproducibility,
+and software scope determination. Although LBM developers did not mention these
+pain points, we cannot conclude that they are not relevant for LBM software
+development, since we only interviewed 4 LBM developers for about an hour each.
+
+\begin{enumerate}
+
+	\item[P\refstepcounter{pnum}\thepnum \label{P_LackDevTime}:] \textbf{Lack of
+	Development Time} A developer of pyLBM noted that their small development
+	team has a lack of time to implement new features. Small development teams
+	are common for LBM software packages (as shown in the measurement table
+	excerpt in Figure~\ref{measurement_template_image}). Lack of time is also
+	highlighted as a pain point by other research software developers
+	\citet{PintoEtAl2018, PintoEtAl2016, WieseEtAl2019}.
+
+	\item[P\refstepcounter{pnum}\thepnum \label{P_LackSoftDevExp}:] \textbf{Lack
+	of Software Development Experience} A lack of software development
+	experience was noted by the developer of TCLB, and others noted a need for
+	improving software engineering education. Many of the team members on their
+	project are domain experts, not computer scientists or software engineers.
+	This same trend is noted by \citet{Nguyen-HoanEtAl2010}, which showed only
+	23\% of research software survey respondents having a computing-related
+	background. Similarly, \citep{UditAndKatz2017} show that the majority (54\%)
+	of postdocs have not received training in software development.  The LBM
+	developer suggesting an increasing role for formal software education
+	matches the trend observed by \citet{PintoEtAl2018}, where their replication
+	of a previous study \citep{HannayEtAl2009}, shows a growing interest in
+	formal training (From 13\% of respondents in 2009 to 22\% in 2018).
+	\citep{PintoEtAl2018} found that some developers feel there is a mismatch
+	between coding skills and subject-matter skills.
+
+	\item[P\refstepcounter{pnum}\thepnum \label{P_LackFunding}:] \textbf{Lack of
+	Incentive and Funding} The TCLB developer noted a lack of incentives and
+	funding in academia for developing widely used scientific software. This
+	problem has also been noted others \citep{gewaltig2012quality, Goble2014,
+	KaterbowAndFeulner2018}.  \citet{WieseEtAl2019} reported developer pains
+	related to publicity, since historically publishing norms make it difficult
+	to get credit for creating software.  As studied by
+	\citet{HowisonAndBullard2016}, research software (specifically biology
+	software, but the trend likely applies to other research software domains)
+	is infrequently cited.  \citep{PintoEtAl2018} also mentions the lack of
+	formal reward system for research software.
+
+	\item[P\refstepcounter{pnum}\thepnum \label{P_LackExtSupport}:] \textbf{Lack
+	of External Support} A concern was raised that there are no organizations
+	helping with the development of good quality software.  This concern is not
+	echoed in the literature because there are such organizations, including
+	\href{https://bssw.io/} {Better Scientific Software (BSSw)},
+	\href{https://www.software.ac.uk/} {Software Sustainability Institute}
 	\citep{CrouchEtAl2013}, and \href{https://software-carpentry.org/}{Software
-Carpentry}. Some SCS developers may not be familiar with these organizations.
-\citep{WilsonAndLumsdaine2006}
-
-Scientific software is often developed in-house by the very researchers that
-temporarily use it in their own research. Empirical studies of such
-``professional end-user development'' of SCS is noted in \citep{segal2007end}.
-This kind of software has a defined user and purpose, and often does not meet
-the standards that would be required by external users. It has been categorized
-as a ``private tool'' by \citep{gewaltig2012quality}, which notes that despite
-often being made freely available, ``it is not always clear that it is
-sufficiently mature in terms of domain coverage, validity, documentation or
-usability, to be useful to other researchers''. It is less common for such ad
-hoc scientific software to become ``user-ready software'', which ``is not only
-research-ready, but should have most of the attributes commonly expected of
-commercial software products including broadness of scope, robustness,
-demonstrable correctness and adequate documentation''
-\citep{gewaltig2012quality}. As software becomes user-ready, it can become
-commercialized and closed-source.  \wss{remove this paragraph?}
-
-\wss{Remove this section?  The external support organizations will likely come
-up as the mitigation for some of the other pain points.}
-
-\subsection{Parallelization and Continuous Integration}
-
-Setting up parallelization was also noted as a technical pain point by one of
-the developers, and the introduction of continuous integration by another.
-Software development knowledge, and automatic code generation, could mitigate
-such pain points. As already noted, many of the developers are domain experts
-and not professional software developers. The developer of TCLB noted that
-eliminating equivalent statements using macros had helped improve the quality of
-their source code, specifically helping with reusing code to run on both the CPU
-and GPU. \wss{The point about macros isn't really related to parallelization or
-continuous integration.}
-
-\subsection{Ensuring Correctness}
-
-Difficulties with ensuring correctness were also noted by several developers.
-They indicated that tests are run on all new source code additions, testing both
-individual modules and the system to verify correctness. These tests compare the
-package output to known correct output using test cases. \wss{oracle}  The
-developer of TCLB commented that the amount of testing data that is needed for
-some cases is a problem as free testing services do not offer capabilities to
-store and process such large amounts of data, and in-house testing solutions
-needed to be created to address this limitation. The solution for this has been
-to limit the size of the testing problems, and to run tests in small batches
-with few iterations. \wss{The Weiss paper alludes this this as well.  I think
-the Kelly and Sanders paper likely talks about testing problems and correctness.
-``while many scientists believe that software testing is important, a smaller
-number believe they have sufficient understanding about testing concepts''
-\citep{HannayEtAl2009}.}
-
-\wss{** BEGIN Testing and Verification **}
-
-\citet{Nguyen-HoanEtAl2010} says the following:
-\begin{itemize}
-	\item 38 of 60 respondents performed unit tests
-	\item 23 of 60 respondents performed integration tests
-	\item ``In our results, the four most common activities involve comparisons
-	with known, trusted, or prior data. Verification and peer review were three
-	of the four least commonly performed activities.'' \wss{testing is not the
-	only way to verify software.  This addresses the oracle problem.}
-\end{itemize}
-
-May do more testing than artifacts indicate, but that is still a problem.
-Artifacts should be available so that others can verify the results, verify
-their installation.  Part of an assurance case means that a third party can
-verify the work.
-
-\wss{** END Testing and Verification **}
-
-``The questions of what constitutes an adequate testing oracle for scientific
-software and what constitutes adequate test data are left open. Some of the
-validation testing we researched was running up against the practical limits;
-they were using whatever oracle they could scrape together from industry data,
-benchmarks, and expert judgment.'' \citep{KellyAndSanders2008}
-
-\wss{Seems that LBM not concerned with the oracle problem?}
-
-\subsection{Usability}
-
-A few obstacles related to users were found. Several developers noted that users
-sometimes try to use incorrect LBM method combinations to solve their problems.
-Furthermore, some users think that the packages will work out of the box to
-solve their cases, while the packages require both a good understanding of CFD
-and an understanding of the requirements for formulating problems in the
-individual packages, which can be a significant endeavour. These software
-packages are not like commercial software packages. They are generally set up to
-solve specific research problems, and are often primarily used by their
-developers. While they are modifiable to solve similar problems, these
-modifications are not trivial. Better documentation, with attention on
-traceability, and automatic code generation are suggested when designing
-software for change, and would help with these modifications. \wss{Should decide
-if want to introduce ideas for improving within the presentation of the pain
-point, or separately?}  So far this problem of on-boarding new users has been
-addressed by updating the documentation to better inform users of the underlying
-LBM theory and package requirements. Similar issues with LBM parameters were
-noted by another developer. Updating the user interface to better explain
-theoretical principles, as well as test user input for compatibility, was the
-implemented solution. As noted above, sometimes frequently asked questions on
-the underlying theory and on how to use the software are answered in the
-documentation.  The interviewed developer of ESPResSo commented that parts of
-their package's source code had been refactored to Python to help address
-usability issues. Python was perceived as a much more usable language, and it
-would be easy for future users and developers to learn and understand the source
-code.
-
-\subsection{Technical Debt}
-
-A few potential future obstacles were noted. The developer of ESPResSo noted
-that their source code had been written with a specific application in mind and
-that due to this there was too much coupling between components of the source
-code. This results in technical debt, having an impact on future modifiability
-and reusability when trying to extend the software, and the code would need to
-be refactored.  As noted above, difficulties with ensuring future correctness
-could also arise. As new methods and functionality is added into the software,
-new test cases and test data will need to be developed.  \wss{Weiss mentions
-mainainability related issues under pain points.}
-
-\subsection{Quality of Documentation}
-
-Interviewees commented that documentation is important and that its quality
-could be improved. As already noted, there is often no time or funding for
-maintaining quality documentation for software that is rarely used outside of
-the development team. Furthermore, the documentation generally only provides a
-shallow overview of the underlying CFD theory. Users would be well advised to
-already be familiar with these topics, or they should spend significant time
-referencing theory resources. The documentation instead generally focuses on
-explaining how to use the software. It is of course not feasible for package
-documentation to address the underlying physics topics in detail, so it is
-advised that the package documentation links to resources that better explain
-the underlying theory. Sometimes, frequently asked questions about the
-underlying theory are answered in the documentation. OpenLB has an artifact for
-such questions.  \wss{Weise mentions documentation.}  \wss{Partly solve the
-problem by documentation user characteristics, to explicitly state what the user
-is assumed to already know.}
-
-The developer comments emphasized an importance on source code, while
-documentation seems to be of secondary importance. It must be stressed that
-improving documentation could benefit development and help eliminate some of the
-developer concerns that were raised. The use of automatic document generation
-tools that capture scientific and computing knowledge, and transform it into
-software artifacts, is advised. Drasil is an automatic document generation tool
-that is further discussed in Section~\ref{highlightedrecommendations}.
-
-Poor software documentation is also mentioned by \citet{PintoEtAl2018} and
-\citet{LethbridgeEtAl2003}.  The survey from \citet{LethbridgeEtAl2003} is for
-software in general, not just research software.
+	Carpentry} \wss{WilsonAndLumsdaine2006, Wilson2016}. Over time awareness of
+	these groups will grow, so this pain point is likely to disappear in the
+	future for LBM and other research software developers.
+
+	\item[P\refstepcounter{pnum}\thepnum \label{P_TechnologyHurdles}:]
+	\textbf{Technology Hurdles} Technology pain points include setting up
+	parallelization and continuous integration.
+
+	\item[P\refstepcounter{pnum}\thepnum \label{P_Correctness}:]
+	\textbf{Ensuring Correctness} Difficulties with ensuring correctness were
+	noted by several developers. The TCLB developer commented that the amount of
+	testing data needed was sometimes problematic, since free testing services
+	do not offer adequate facilities for large amounts of data, which means
+	in-house testing solutions are needed.  Other research software domains
+	point to different problems with testing, as follows: i)
+	\citep{PintoEtAl2018} mention the problem of insufficient testing; ii) the
+	survey of \citep{HannayEtAl2009} shows that more developers think testing is
+	important than the number that believe they have a sufficient understanding
+	of testing concepts; and, iii) \citep{HannayEtAl2009, KanewalaAndBieman2013,
+	KellyEtAl2011, WieseEtAl2019} point to the oracle problem, which
+	occurs in research software when we do not have a means to judge the
+	correctness of the calculated solutions.  Since LBM developers did not
+	allude to the oracle problem in our conversation, we believe that they have
+	developed techniques to work around the oracle problem.
+
+	\item[P\refstepcounter{pnum}\thepnum \label{P_Usability}:]
+	\textbf{Usability} Several developers noted that users sometimes try to use
+	incorrect LBM method combinations to solve their problems. Furthermore, some
+	users think that the packages will work out of the box to solve their cases,
+	while in reality CFD knowledge needs to be applied to correctly modify the
+	packages for the new endeavour.  \citep{WieseEtAl2019} also interacted with
+	developers that mentioned that users do not always have the expertise
+	required to install or use the software.
+
+	\item[P\refstepcounter{pnum}\thepnum \label{P_TechDebt}:] \textbf{Technical
+	Debt} The developer of ESPResSo said that their source code was written with
+	a specific application in mind, which later caused too much coupling between
+	components in the source code. This results in technical debt
+	\citet{KruchtenEtAl2012}, which has an impact on future modifiability and
+	reusability. Concern with technical debt is likely why researchers in the
+	survey of \citep{Nguyen-HoanEtAl2010} rated maintainability as the third
+	most important software quality. More recently the push for sustainable
+	software \citet{deSouzaEtAl2019} is motivated by the pain that past
+	developers have had with accumulating too much technical debt.
+
+	\item[P\refstepcounter{pnum}\thepnum \label{P_Documentation}:]
+	\textbf{Quality of Documentation} One of the consequences of pain point
+	\ppref{P_LackDevTime} is sometimes inadequate documentation. The
+	documentation for LBM is challenging because of the complexity of the
+	underlying CFD theory and the differing backgrounds of users. Poor research
+	software documentation is also mentioned by others \citep{PintoEtAl2018,
+	WieseEtAl2019} and the problem also arises with non-research software
+	\citep{LethbridgeEtAl2003}.  Recommendations on the development of research
+	software state that developers should critically evaluate their own
+	development processes in terms of quality assurance and comply with
+	international standards for software documentation
+	\citep{KaterbowAndFeulner2018}.
+
+\end{enumerate}
 
 \section{LBM Developer Identified Concerns} \label{Sec_Concerns}
-
-Answers \rqref{RQ_Concerns}.  This section presents qualitative findings on
-software qualities as gleaned from interviewing 4 developers. The full interview
-questions are found in \citet{SmithEtAl2021}.
 
@@
 and artifacts, and domain expert feedback.
+
+\wss{Address lack of development time (\ppref{P_LackDevTime}): Aside from
+on-boarding new developers, time constraints could be mitigated by increasing
+developer efficiency, which could be addressed in several ways, including by
+improving the quality of documentation, or incorporating automatic code
+generation.}
+
+\wss{Address lack of funding (\ppref{P_LackFunding}): The developer further
+commented that there are no journals that publish such scientific software
+source code. However, there are ways to get such source code cited. Work has
+been done to address this in \citep{SmithEtAl2016-softcite}, which presents a
+set of software citation principles and discusses ``how they could be used to
+implement software citation in the scholarly community''
+\citep{katz2019software}. other papers also mention this pain point.
+Solutions proposed by Katz, journal of open source etc. Journal of Open Source
+Software \citet{SmithEtAl2018-joss}.  Calling for direct software citations
+\citep{SmithEtAl2016-softcite}.  Guidance on software citation from
+\citet{KatzEtAl2021}.  Discussed in CarverEtAl2021 paper.}
+
+\wss{Address Usability (\ppref{P_Usability}): Better documentation, with
+attention on traceability, and automatic code generation are suggested when
+designing software for change, and would help with these modifications.  So far
+this problem of on-boarding new users has been addressed by updating the
+documentation to better inform users of the underlying LBM theory and package
+requirements. Similar issues with LBM parameters were noted by another
+developer. Updating the user interface to better explain theoretical principles,
+as well as test user input for compatibility, was the implemented solution. As
+noted above, sometimes frequently asked questions on the underlying theory and
+on how to use the software are answered in the documentation.  The interviewed
+developer of ESPResSo commented that parts of their package's source code had
+been refactored to Python to help address usability issues. Python was perceived
+as a much more usable language, and it would be easy for future users and
+developers to learn and understand the source code.}
+
+\wss{Address Correctness (\ppref{P_Correctness}): Testing is not the only way to
+build confidence in correctness.  Peer review.  }
+
+\wss{Address quality of documentation (\ppref{P_Documentation}): Make required
+background of users explicit, possibly in a requirements document.}
 

### 2022-02-15 StateOfPractice/Papers/ICCS-2022-SOP_MethodologyPlusExamples/DigDeep.tex
Updates to SOP for LBM, start of digging deeper paper

--- 2021-10-22 StateOfPractice/Methodology/Methodology.tex
@@
-\documentclass[letterpaper,cleveref]{lipics-v2019}
-
-\usepackage[round]{natbib}
+% This is samplepaper.tex, a sample chapter demonstrating the
+% LLNCS macro package for Springer Computer Science proceedings;
+% Version 2.21 of 2022/01/12
+%
+\documentclass[runningheads]{llncs}
+%
+\usepackage[T1]{fontenc}
+% T1 fonts will be used to generate the final print and online PDFs,
+% so please use T1 fonts in your manuscript whenever possible.
+% Other font encondings may result in incorrect characters.
+%
+\usepackage{graphicx}
+% Used for displaying a sample figure. If possible, figure files should
+% be included in EPS format.
+%
+% If you use the hyperref package, please uncomment the following two lines
+% to display URLs in blue roman font according to Springer's eBook style:
+% \usepackage{color}
+% \renewcommand\UrlFont{\color{blue}\rmfamily}
+%
+\usepackage{hyperref}
+\usepackage{paralist}
 \usepackage{booktabs}
-\usepackage{amsmath,amsthm}
-\usepackage{hyperref}
-\usepackage{xcolor}
-\usepackage{longtable}
-\usepackage{paralist}
 \usepackage{totcount}
-\usepackage{booktabs}
-
-\hypersetup{
-colorlinks=true,     % false: boxed links; true: colored links
-linkcolor=blue,      % color of internal links (change box color with linkbordercolor)
-citecolor=blue,      % color of links to bibliography
-filecolor=magenta,   % color of file links
-urlcolor=cyan        % color of external links
-}
-
-%% Comments
-\newif\ifcomments\commentstrue
-
-\ifcomments
-\newcommand{\authornote}[3]{\textcolor{#1}{[#3 ---#2]}}
-\newcommand{\todo}[1]{\textcolor{red}{[TODO: #1]}}
-\else
-\newcommand{\authornote}[3]{}
-\newcommand{\todo}[1]{}
-\fi
-
-\newcommand{\wss}[1]{\authornote{blue}{SS}{#1}} %Spencer Smith
-\newcommand{\jc}[1]{\authornote{red}{JC}{#1}} %Jacques Carette
-\newcommand{\oo}[1]{\authornote{magenta}{OO}{#1}} %Olu Owojaiye
-\newcommand{\pmi}[1]{\authornote{green}{PM}{#1}} %Peter Michalski
-\newcommand{\ad}[1]{\authornote{brown}{AD}{#1}} %Ao Dong
-
-\newcommand{\notdone}[1]{\textcolor{red}{#1}}
-\newcommand{\done}[1]{\textcolor{black}{#1}}
-
-%\oddsidemargin 0mm
-%\evensidemargin 0mm
-%\textwidth 160mm
-%\textheight 200mm
-
-\newcommand{\CC}{C\nolinebreak\hspace{-.05em}\raisebox{.4ex}{\small\bf
-+}\nolinebreak\hspace{-.10em}\raisebox{.4ex}{\small\bf +}}
 
@@
 
-\title{Methodology for Assessing the State of the Practice for Domain X} 
-\author{Spencer Smith}{McMaster University, Canada}{smiths@mcmaster.ca}{}{}
-\author{Jacques Carette}{McMaster University, Canada}{carette@mcmaster.ca}{}{}
-\author{Peter Michalski}{McMaster University, Canada}{michap@mcmaster.ca}{}{}
-\author{Ao Dong}{McMaster University, Canada}{donga9@mcmaster.ca}{}{}
-\author{Olu Owojaiye}{McMaster University, Canada}{owojaiyo@mcmaster.ca}{}{}
-
-\authorrunning{Smith et al.}  \Copyright{Spencer Smith and Jacques Carette and
- Peter Michalski and Ao Dong and Olu Owojaiye}
-
-\ccsdesc{Software and its engineering}
-\ccsdesc{Software and its engineering~Software product lines}
-\ccsdesc{General and reference~Empirical studies}
-
-\keywords{software quality, domain analysis, scientific computing software,
-  research software, empirical measures, analytic hierarchy process}
-
-\date{\today}
-
-\hideLIPIcs
-\nolinenumbers
-
 \begin{document}
-
-\maketitle
-
+%
+\title{Digging Deep to Assess the State of the Practice for Different Research Software Domains}
+%
+%\titlerunning{Abbreviated paper title}
+% If the paper title is too long for the running head, you can set
+% an abbreviated paper title here
+%
+\author{Spencer Smith\inst{1}\orcidID{0000-0002-0760-0987}} %\and
+%Second Author\inst{2,3}\orcidID{1111-2222-3333-4444} \and
+%Third Author\inst{3}\orcidID{2222--3333-4444-5555}}
+%
+\authorrunning{S.\ Smith}
+% First names are abbreviated in the running head.
+% If there are more than two authors, 'et al.' is used.
+%
+\institute{McMaster University, 1280 Main Street West, Hamilton ON L8S 4K1, Canada
+\email{smiths@mcmaster.ca}\\
+\url{http://www.cas.mcmaster.ca/~smiths/}}
+%
+\maketitle              % typeset the header of the contribution
+%
 \begin{abstract}
+
 	To improve software development methods and tools for research software, we
@@
   tool, we estimate (based on our experience with using the process) the time to
-  complete an assessment for a given domain at \total{totHours} person hours.
+  complete an assessment for a given domain at 173 person hours.
+
+The abstract should briefly summarize the contents of the paper in
+150--250 words.
+
+\keywords{First keyword  \and Second keyword \and Another keyword.}
 \end{abstract}
-
-~\newpage
-
-\tableofcontents
-
-~\newpage
-
+%
+%
+%
 \section{Introduction} \label{SecIntroduction}
+
+- research software is vital for making progress in science and engineering, but it is difficult to develop
+- this has promoted many studies to understand the challenges of research software development
+- the previous studies have usually used interviews, surveys and case studies
+- the interviews and surveys recruit participants from all domains of research software
+- the case studies, by their nature, focus on specific examples
+- what is missing is a methodology that:
+	- goes beyond surveys to look at the actual artifacts in the software repos
+	- focuses on one domain at a time
+- by focusing on one domain at a time we can:
+	- provide useful knowledge/advice to the developers in that domain
+	- build up a set of measures of different domains to compare and contrast them
 
@@
 practice'' for research software often does not incorporate ``state of the art''
-Software Engineering (SE) tools and methods \citep{JohansonAndHasselbring2018}.
+Software Engineering (SE) tools and methods \cite{JohansonAndHasselbring2018}.
 The lack of SE tools and methods contributes to sustainability and reliability
-problems \citep{FaulkEtAl2009}.  Problems with the current state of the practice
+problems \cite{FaulkEtAl2009}.  Problems with the current state of the practice
 are evident from embarrassing failures, like a retraction of derived molecular
-protein structures \citep{Miller2006}, false reproduction of sonoluminescent
-fusion \citep{PostAndVotta2005}, and fixing and then reintroducing the same
+protein structures \cite{Miller2006}, false reproduction of sonoluminescent
+fusion \cite{PostAndVotta2005}, and fixing and then reintroducing the same
 error in a large code base three times in 20 year
-\citep{MilewiczAndRaybourn2018}.  To improve this situation, we need to first
+\cite{MilewiczAndRaybourn2018}.  To improve this situation, we need to first
 fully understand the current state of the practice for research software
@@
 research domains.  We have updated the work that was done previously for domains
-such as Geographic Information Systems \citep{SmithEtAl2018_arXivGIS}, Mesh
-Generators \citep{SmithEtAl2016}, Seismology software \citep{SmithEtAl2018}, and
-Statistical software for psychology \citep{SmithEtAl2018_StatSoft}.  Initial
+such as Geographic Information Systems \cite{SmithEtAl2018_arXivGIS}, Mesh
+Generators \cite{SmithEtAl2016}, Seismology software \cite{SmithEtAl2018}, and
+Statistical software for psychology \cite{SmithEtAl2018_StatSoft}.  Initial
 tests of the new methodology have been done for medical image analysis software
-\citep{Dong2021} and for Lattice Boltzmann Method (LBM) software
-\citep{Michalski2021}.
+\cite{Dong2021} and for Lattice Boltzmann Method (LBM) software
+\cite{Michalski2021}.
 
@@
 In the proposed methodology, the collected data is combined to rank the software
-within each domain using the Analytic Hierarchy Process (AHP) \citep{Saaty1980}.
+within each domain using the Analytic Hierarchy Process (AHP) \cite{Saaty1980}.
 As in the previous measurement exercise, we use AHP to develop a list of
@@
 using benchmarks, to other projects, such as the work of
-\citet{kaagstrom1998gemm}.  Currently we are also leaving experiments to measure
+\cite{kaagstrom1998gemm}.  Currently we are also leaving experiments to measure
 usability and modifiability as future work, as discussed in
 Section~\ref{SecFutureWork}.
-
-Within this document the following typographic conventions have been adopted:
-\begin{inparaenum}[i)] 
-  \item \textcolor{red}{Blue} text denotes a link to sections of this document,
-  including citations 
-  \item \textcolor{cyan}{Cyan} text denotes an external URL link.
-\end{inparaenum}
 
@@
 generate, process or analyze results that [are intended] to appear in a
-publication'' \citep{hettrick2014uk} in a scientific or engineering context.
+publication'' \cite{hettrick2014uk} in a scientific or engineering context.
 Research software is a more general term for what is often called scientific
@@
 Some examples of domains that fit these criteria are finite element analysis
-\citep{szabo1996finite}, quantum chemistry \citep{veryazov20042molcas},
-seismology \citep{SmithEtAl2018}, as well as mesh generators
-\citep{smith2016state}.
+\cite{szabo1996finite}, quantum chemistry \cite{veryazov20042molcas},
+seismology \cite{SmithEtAl2018}, as well as mesh generators
+\cite{smith2016state}.
 
@@
 able to view the software as constituting a program family.  The concept of a
-program family is defined by \citet{parnas1976design} as ``a set of programs
+program family is defined by \cite{parnas1976design} as ``a set of programs
 whose common properties are so extensive that it is advantageous to study the
@@
 	\begin{center}
-	  \includegraphics[width=1.0\textwidth]{measurement_template}
+	  %\includegraphics[width=1.0\textwidth]{measurement_template}
 	  \caption{Excerpt of the Top Section of the Measurement Template (Summary Information)}
@@
 of developers, etc.  We follow the definitions given by
-\citet{GewaltigAndCannon2012} for the software categories.  Public means software
+\cite{GewaltigAndCannon2012} for the software categories.  Public means software
 intended for public use.  Private means software aimed only at a specific
@@
 
-As in \citet{SmithEtAl2016}, Virtual machines (VMs) are used to provide an
+As in \cite{SmithEtAl2016}, Virtual machines (VMs) are used to provide an
 optimal testing environments for each package. VMs were used because it is
@@
 In the previous state of the practice measurement process
-\citep{SmithEtAl2018_arXivGIS, SmithEtAl2016, SmithEtAl2018}, we only based our
+\cite{SmithEtAl2018_arXivGIS, SmithEtAl2016, SmithEtAl2018}, we only based our
 assessment on information available in on-line software repos.  However, this
@@
 a rough estimate, based on our experience completing assessments for medical
-image analysis software \citep{Dong2021} and for Lattice Boltzmann Method (LBM)
-software \citep{Michalski2021}.  These two domains were assessed at the same
+image analysis software \cite{Dong2021} and for Lattice Boltzmann Method (LBM)
+software \cite{Michalski2021}.  These two domains were assessed at the same
 time as designing the methodology presented in this document.  We did our best
@@
 practice for the following domains: Geographic Information Systems
-\citep{SmithEtAl2018_arXivGIS}, Mesh Generators \citep{SmithEtAl2016},
-Seismology software \citep{SmithEtAl2018}, Statistical software for psychology
-\citep{SmithEtAl2018_StatSoft}, medical image analysis software \citep{Dong2021}
-and for Lattice Boltzmann Method software \citep{Michalski2021}.  Future domains
+\cite{SmithEtAl2018_arXivGIS}, Mesh Generators \cite{SmithEtAl2016},
+Seismology software \cite{SmithEtAl2018}, Statistical software for psychology
+\cite{SmithEtAl2018_StatSoft}, medical image analysis software \cite{Dong2021}
+and for Lattice Boltzmann Method software \cite{Michalski2021}.  Future domains
 for measurement could include finite element software, computational medicine,
@@
 \href{https://github.com/JacquesCarette/Drasil} {Drasil}
-\citep{SzymczakEtAl2016}), which facilitates a development process that focuses
+\cite{SzymczakEtAl2016}), which facilitates a development process that focuses
 on knowledge capture, followed by generation of code and other software
@@
 changes, instead of any changes because software cannot be designed so that
-everything is equally easy to change \citep{parnas1986rational}.  The procedure
+everything is equally easy to change \cite{parnas1986rational}.  The procedure
 could be along the following lines:
@@
 templates and AHP tool, we estimate (based on our experience with using the
-process) the time to complete an assessment for a given domain at
-\total{totHours} person hours.
-
-\newpage
-
-\appendix
-
-\section{A Guide to Repository-Based Measurement Tools} \label{SecRepoTools}
-
-This appendix covers the tools used for collecting repository based data
-(Section~\ref{SecEmpiricalMeasures}).  The two tools covered are git\_stats and
-scc.  The tools do not have to be used in any particular order.
-
-\subsection{git\_stats}
-
-\subsubsection{Introduction}
-
-Source Code: \href{https://github.com/tomgi/git_stats}{GitHub repo}
-
-\subsubsection{User Manual} \label{git_stats_manual}
-
-Official Manual: \href{https://github.com/tomgi/git_stats}{GitHub repo}
-
-\subsubsection{Demo of Installation and Running the Tool}
-
-The installation steps on your machine may be different from this section.
-Please refer to the user manual mentioned in Section \ref{git_stats_manual}, if
-necessary.  The steps shown here were executed on a virtual machine with 8 cores
-and 16 GB RAM running Debian GNU/Linux 9.11.
-
-\begin{enumerate}
-\item Install ruby/gem environment
-\begin{lstlisting}
-apt-get install ruby ruby-nokogiri ruby-nokogiri-diff ruby-nokogumbo
-\end{lstlisting}
-
-Check the installation:
-\begin{lstlisting}
-gem --version
-\end{lstlisting}
-
-\item Install the tool
-\begin{lstlisting}
-sudo gem install git_stats
-\end{lstlisting}
-
-\item Prepare the target repo
-
-Make sure the target repo (the repo to be analyzed, not the repo of this tool)
-is on your machine.
-In this demo, the target repo is downloaded from a GitHub repo:
-\begin{lstlisting}
-# change [git path] to the url of your target repo
-git clone [git path]
-# e.g. git clone https://github.com/nroduit/Weasis.git
-\end{lstlisting}
-
-\item Generate analytics
-\begin{lstlisting}
-# make sure [repo path] is the target repo path
-# the [output path] can be anywhere you desire
-git_stats generate -p [repo path] -o [output path]
-# e.g. git_stats generate -p /home/user/git-stats/Weasis -o 
-# /home/user/git-stats/Weasis-analytics
-\end{lstlisting}
-
-\item View the analytics
-
-View the analytic results by open [output path]/index.html with any browser or
-other software supporting HTML web page format.
-
-\begin{figure}[!ht]
-\includegraphics[scale=0.4]{git-stats-demo.JPG}
-\end{figure}
-
-\item Download the data
-
-On most of the taps of this web page, the data can be downloaded for more
-analytics by clicking the menu button beside the data-range section.
-
-\end{enumerate}
-
-\subsection{scc}
-
-\subsubsection{Introduction}
-
-Source Code: \href{https://github.com/boyter/scc}{GitHub repo}
-
-\subsubsection{User Manual} \label{scc_manual}
-
-Official Manual: \href{https://github.com/boyter/scc}{GitHub repo}
-
-\subsubsection{Demo of Installation and Running the Tool}
-
-The installation steps on your machine may be different from this section.
-Please refer to the user manual mentioned in Section \ref{scc_manual}, if
-necessary.  These steps were executed on a virtual machine with 8 cores and 16
-GB RAM running Debian GNU/Linux 9.11
-
-\begin{enumerate}
-\item Install Golang
-
-Follow the \href{https://golang.org/doc/install}{official instructions}, or the
-following demo,
-
-download the installation package:
-\begin{lstlisting}
-wget https://dl.google.com/go/go1.14.3.linux-amd64.tar.gz
-\end{lstlisting}
-
-unpack to /usr/local:
-\begin{lstlisting}
-sudo tar -C /usr/local -xzf go1.14.3.linux-amd64.tar.gz
-\end{lstlisting}
-
-use a text editor to open \textasciitilde/.profile, e.g.:
-\begin{lstlisting}
-nano ~/.profile
-\end{lstlisting}
-
-add the following lines to the end of this file:
-\begin{lstlisting}
-export GOPATH=$HOME/go
-export PATH=$PATH:/usr/local/go/bin:$GOPATH/bin
-\end{lstlisting}
-
-save the file, and load the commands into the current shell instance:
-\begin{lstlisting}
-source ~/.profile
-\end{lstlisting}
-
-check the installation:
-\begin{lstlisting}
-go version
-\end{lstlisting}
-
-\item Install the tool
-\begin{lstlisting}
-go get -u github.com/boyter/scc/
-\end{lstlisting}
-
-\item Prepare the target repo
-
-Make sure the target repo (the repo to be analyzed, not the repo of this tool)
-is on your machine.
-In this demo, the target repo is downloaded from a GitHub repo:
-\begin{lstlisting}
-# change [project path] to your desired folder
-cd [project path]
-git clone https://github.com/nroduit/Weasis.git
-\end{lstlisting}
-
-\item Generate analytics
-\begin{lstlisting}
-# make sure [repo path] is the target repo path
-cd [repo path]
-# use scc to generate analytics
-scc
-\end{lstlisting}
-
-\item View the analytics
-
-The results will be appear something like the following.
-
-\begin{figure}[!ht]
-\includegraphics[scale=0.5]{scc-demo.JPG}
-\end{figure}
-
-\end{enumerate}
-
-\newpage
-
-\section{Measurement Template} \label{SecGradingTemplate}
-
-The table below lists the full set of measures that are assessed for each
-software product.  The measures are grouped under headings for each quality, and
-one for summary information.  Following each measure, the type for a valid
-result is given in brackets.  Many of the types are given as enumerated sets.
-For instance, the response on many of the questions is one of ``yes,'' ``no,''
-or ``unclear.''  The type ``number'' means natural number, a positive integer.  The
-types for date and url are not explicitly defined, but they are what one would
-expect from their names.  In some cases the response for a given question is not
-necessarily limited to one answer, such as the question on what platforms are
-supported by the software product.  Case like this are indicated by ``set of''
-preceding the type of an individual answer.  The type in these cases are then
-the power set of the individual response type.  In some cases a superscript $^*$
-is used to indicate that a response of this type should be accompanied by
-explanatory text.  For instance, if problems were caused by uninstall, the
-reviewer should note what problems were caused.  % An (I) precedes the question
-% description when its measurement requires a successful installation.
-The template also include 3 sections at the end for summarizing the repository
-based metrics.  A
-\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Combined_MeasurementTemplate_EmpiricalMeasures.xlsx}
-{blank measurement template spreadsheet} is available to save time with data
-entry.
-
-\newpage
-
-\captionof{table}{Measurement Template}\label{measurementtemplate}
-\def\arraystretch{1.22}
-\begin{tabular}{p{14cm}}
-	\hline
-	\textbf{Summary Information}\\
-	\hline
-	Software name? (string)\\
-	URL? (URL)\\
-	Affiliation (institution(s)) (string or {N/A})\\
-	Software purpose (string)\\
-	Number of developers (all developers that have contributed at least one commit
-	to the project) (use repo commit logs) (number)\\
-	How is the project funded? (unfunded, unclear, funded$^*$) where $^*$ requires a
-	string to say the source of funding\\
-	Initial release date? (date)\\
-	Last commit date? (date)\\
-	Status? (alive is defined as presence of commits in the last 18 months)
-	({alive, dead, unclear})\\
-	License? ({GNU GPL, BSD, MIT, terms of use, trial, none, unclear, other$^*$}) $^*$
-	given via a string \\
-	Platforms? (set of {Windows, Linux, OS X, Android, other$^*$}) $^*$ given via
-	string\\
-	Software Category? The concept category includes software that does not have
-	an officially released version. Public software has a released version in the
-	public domain. Private software has a released version available to authorized
-	users only. ({concept, public, private})\\
-	Development model? ({open source, freeware, commercial, unclear})\\
-	Publications about the software? Refers to publications that have used or
-	mentioned the software. (number or {unknown})\\
-	Source code URL? ({set of url, n/a, unclear})\\
-	Programming language(s)? (set of {FORTRAN, Matlab, C, C++, Java, R, Ruby,
-	Python, Cython, BASIC, Pascal, IDL, unclear, other$^*$}) $^*$ given via string \\
-	Is there evidence that performance was considered? Performance refers to
-	either speed, storage, or throughput. ({yes$^*$, no})\\
-	Additional comments? (can cover any metrics you feel are missing, or any other
-	thoughts you have) \\
-	\hline
-\end{tabular}
-
-\def\arraystretch{1.5}
-\begin{tabular}{p{14cm}}
-	\hline		
-	\textbf{Installability (Measured via installation on a virtual machine.) }\\
-	\hline
-	Are there installation instructions? ({yes, no})\\
-	Are the installation instructions in one place? Place referring to a single
-	document or web-page. ({yes, no, n/a})\\
-	Are the installation instructions linear? Linear meaning progressing  in a
-	single series of steps. ({yes, no, n/a})\\
-	Are the instructions written as if the person doing the installation has none
-	of the dependent packages installed? ({yes, no, unclear})\\
-	Are compatible operating system versions listed? ({yes, no})\\
-	Is there something in place to automate the installation (makefile, script,
-	installer, etc)? ({yes$^*$, no})\\
-	If the software installation broke, was a descriptive error message displayed?
-	({yes, no, n/a})\\
-	Is there a specified way to validate the installation? ({yes$^*$, no})\\
-	How many steps were involved in the installation? (Includes manual steps like
-	unzipping files) Specify OS. (number, OS)\\
-	What OS was used for the installation? ({Windows, Linux, OS X, Android, other$^*$
-	}) $^*$given via string\\
-	How many extra software packages need to be installed before or during
-	installation? (number)\\
-	Are required package versions listed? ({yes, no, n/a})\\
-	Are there instructions for the installation of required packages /
-	dependencies? ({yes, no, n/a})\\
-	Run uninstall, if available. Were any obvious problems caused? ({yes$^*$ , no,
-	unavail})\\
-	Overall impression? ({1 .. 10})\\
-	Additional comments? (can cover any metrics you feel are missing, or any other
-	thoughts you have)\\
-	\hline
-\end{tabular}
-
-\def\arraystretch{1.33}
-\begin{tabular}{p{14cm}}
-	\hline		
-	\textbf{Correctness and Verifiability}\\
-	\hline
-	Any reference to the requirements specifications of the program or theory
-	manuals? ({yes$^*$ , no, unclear})\\
-	What tools or techniques are used to build confidence of correctness?
-	({literate programming, automated testing, symbolic execution, model checking,
-	assertions used in the code, Sphinx, Doxygen, Javadoc, confluence, unclear,
-	other$^*$}) $^*$ given via string\\
-	If there is a getting started tutorial? ({yes, no})\\
-	Are the tutorial instructions linear? ({yes, no, n/a})\\
-	Does the getting started tutorial provide an expected output? ({yes, no$^*$,
-	n/a})\\
-	Does your tutorial output match the expected output? ({yes, no, n/a})\\
-	Are unit tests available?  ({yes, no, unclear})\\
-	Is there evidence of continuous integration? (for example mentioned in
-	documentation, Jenkins, Travis CI, Bamboo, other) ({yes$^*$, no, unclear})\\
-	Overall impression? ({1 .. 10})\\
-	Additional comments? (can cover any metrics you feel are missing, or any other
-	thoughts you have) \\
-	\hline	
-	\textbf{Surface Reliability}\\
-	\hline
-	Did the software “break” during installation? ({yes$^*$ , no})\\
-	If the software installation broke, was the installation instance recoverable?
-	({yes, no, n/a})\\
-	Did the software “break” during the initial tutorial testing? ({yes$^*$, no,
-	n/a})\\
-	If the tutorial testing broke, was a descriptive error message displayed?
-	({yes, no, n/a})\\
-	If the tutorial testing broke, was the tutorial testing instance recoverable?
-	({yes, no, n/a})\\
-	Overall impression? ({1 .. 10})\\
-	Additional comments? (can cover any metrics you feel are missing, or any other
-	thoughts you have)\\
-	\hline		
-\end{tabular}
-
-\def\arraystretch{1.4}
-\begin{tabular}{p{14cm}}
-	\hline		
-	\textbf{Surface Robustness}\\
-	\hline
-	Does the software handle unexpected/unanticipated input (like data of the
-	wrong type, empty input, missing files or links) reasonably? (a reasonable
-	response can include an appropriate error message.) ({yes, no$^*$ })\\
-	For any plain text input files, if all new lines are replaced with new lines
-	and carriage returns, will the software handle this gracefully? ({yes, no$^*$,
-	n/a})\\
-	Overall impression? ({1 .. 10})\\
-	Additional comments? (can cover any metrics you feel are missing, or any other
-	thoughts you have)\\
-	\hline		
-	\textbf{Surface Usability}\\
-	\hline
-	Is there a getting started tutorial? ({yes, no})\\
-	Is there a user manual? ({yes, no})\\
-	Are expected user characteristics documented? ({yes, no})\\
-	What is the user support model? FAQ? User forum? E-mail address to direct
-	questions? Etc. (string)\\
-	Overall impression? ({1 .. 10})\\
-	Additional comments? (can cover any metrics you feel are missing, or any other
-	thoughts you have)\\
-	\hline
-\end{tabular}
-
-\def\arraystretch{1.4}
-\begin{tabular}{p{14cm}}
-	\hline	
-	\textbf{Maintainability}\\
-	\hline
-	What is the current version number? (number)\\
-	Is there any information on how code is reviewed, or how to contribute?
-	({yes$^*$, no})\\
-	Are artifacts available? (List every type of file that is not a code file –
-	for examples please look at the ‘Artifact Name’ column of
-	https://gitlab.cas.mcmaster.ca/SEforSC/se4sc/-/blob/git-svn/GradStudents/Olu/ResearchProposal/Artifacts\_MiningV3.xlsx)
-	({yes$^*$, no, unclear}) $^*$list via string\\
-	What issue tracking tool is employed? (set of {Trac, JIRA, Redmine, e-mail,
-	discussion board, sourceforge, google code, git, BitBucket, none, unclear,
-	other$^*$}) $^*$ given via string\\
-	What is the percentage of identified issues that are closed? (percentage)\\
-	What percentage of code is comments? (percentage)\\
-	Which version control system is in use? ({svn, cvs, git, Github, unclear,
-	other$^*$}) $^*$ given via string\\
-	Overall impression? ({1 .. 10})\\
-	Additional comments? (can cover any metrics you feel are missing, or any other
-	thoughts you have)\\
-	\hline		
-	\textbf{Reusability}\\
-	\hline
-	How many code files are there? (number)\\
-	Is API documented? ({yes, no, n/a})\\
-	Overall impression? ({1 .. 10})\\
-	Additional comments? (can cover any metrics you feel are missing, or any other
-	thoughts you have)\\
-	\hline		
-\end{tabular}
-
-\def\arraystretch{1.4}
-\begin{tabular}{p{14cm}}
-	\hline		
-	\textbf{Surface Understandability (Based on 10 random source files)}\\
-	\hline
-	Consistent indentation and formatting style? ({yes, no, n/a})\\
-	Explicit identification of a coding standard? ({yes$^*$, no, n/a})\\
-	Are the code identifiers consistent, distinctive, and meaningful? ({yes, no$^*$ ,
-	n/a})\\
-	Are constants (other than 0 and 1) hard coded into the program? ({yes, no$^*$ ,
-	n/a})\\
-	Comments are clear, indicate what is being done, not how? ({yes, no$^*$ , n/a})\\
-	Parameters are in the same order for all functions? ({yes, no$^*$ , n/a})\\
-	Is the name/URL of any algorithms used mentioned? ({yes, no$^*$ , n/a})\\
-	Is code modularized? ({yes, no$^*$ , n/a})\\
-	Overall impression? ({1 .. 10})\\
-	Additional comments? (can cover any metrics you feel are missing, or any other
-	thoughts you have)\\
-	\hline		
-	\textbf{Visibility/Transparency}\\
-	\hline
-	Is the development process defined? If yes, what process is used. ({yes$^*$, no,
-	n/a})\\
-	Are there any documents recording the development process and status?  ({yes$^*$,
-	no}))\\
-	Is the development environment documented? ({yes$^*$, no})\\
-	Are there release notes? ({yes$^*$, no})\\
-	Overall impression? ({1 .. 10})\\
-	Additional comments? (can cover any metrics you feel are missing, or any other
-	thoughts you have)\\
-	\hline		
-	\textbf{Raw Metrics (Measured via git\_stats)}\\
-	\hline
-	Number of text-based files. (number)\\
-	Number of binary files. (number)\\
-	Number of total lines in text-based files. (number)\\
-	Number of total lines added to text-based files. (number)\\
-	Number of total lines deleted from text-based files. (number)\\
-	Number of total commits. (number)\\
-	Numbers of commits by year in the last 5 years. (Count from as early as
-	possible if the project is younger than 5 years.) (list of numbers)\\
-	Numbers of commits by month in the last 12 months. (list of numbers)\\
-	\hline
-\end{tabular}
-
-\def\arraystretch{1.4}
-\begin{tabular}{p{14cm}}
-\hline		
-\textbf{Raw Metrics (Measured via scc)}\\
-\hline
-Number of text-based files. (number)\\
-Number of total lines in text-based files. (number)\\
-Number of code lines in text-based files. (number)\\
-Number of comment lines in text-based files. (number)\\
-Number of blank lines in text-based files. (number)\\
-\hline
-\textbf{Repo Metrics (Measured via GitHub)}\\
-\hline
-Number of stars. (number)\\
-Number of forks. (number)\\
-Number of people watching this repo. (number)\\
-Number of open pull requests. (number)\\
-Number of closed pull requests. (number)\\
-\hline
-\end{tabular}
-
-\newpage
-
-\section{Measurement Template Impression Calculator} \label{SecImpressionCalculator}
-
-The table below lists how each quality measure of the measurement template is
-used to calculate an overall impression in each software quality set.
-
-\newpage
-
-\captionof{table}{Measurement Template}
-\def\arraystretch{1.5}
-\begin{tabular}{p{14cm}}
-	\hline		
-	\textbf{Installability (Measured via installation on a virtual machine.) }\\
-	\hline
-	Are there installation instructions? ({yes=1, no=-1})\\
-	Are the installation instructions in one place? Place referring to a single
-	document or web-page. ({yes=1, no=0, n/a=0})\\
-	Are the installation instructions linear? Linear meaning progressing  in a
-	single series of steps. ({yes=1, no=0, n/a=0})\\
-	Are the instructions written as if the person doing the installation has none
-	of the dependent packages installed? ({yes=1, no=0, unclear=0})\\
-	Are compatible operating system versions listed? ({yes=1, no=0})\\
-	Is there something in place to automate the installation (makefile, script,
-	installer, etc)? ({yes$^*$=1, no=-1})\\
-	If the software installation broke, was a descriptive error message displayed?
-	({yes=0, no=-2, n/a=1})\\
-	Is there a specified way to validate the installation? ({yes$^*$=1, no=0})\\
-	How many steps were involved in the installation? (Includes manual steps like
-	unzipping files) Specify OS. ($<$10 = 1)\\
-	What OS was used for the installation? (does not count)\\
-	How many extra software packages need to be installed before or during
-	installation? ($<$10 = 1)\\
-	Are required package versions listed? ({yes=1, no=0, n/a=1})\\
-	Are there instructions for the installation of required packages /
-	dependencies? ({yes=1, no=0, n/a=1})\\
-	Run uninstall, if available. Were any obvious problems caused? ({yes$^*$=0, no=1,
-	unavail=1})\\
-	Overall impression? (a sum of $>$10 is rounded down to 10)\\
-	\hline
-\end{tabular}
-
-\def\arraystretch{1.33}
-\begin{tabular}{p{14cm}}
-	\hline		
-	\textbf{Correctness and Verifiability}\\
-	\hline
-	Any reference to the requirements specifications of the program or theory
-	manuals? ({yes$^*$=2, no=0, unclear=0})\\
-	What tools or techniques are used to build confidence of correctness? (any=1,
-	unclear=0)\\
-	If there is a getting started tutorial? ({yes=2, no=0})\\
-	Are the tutorial instructions linear? ({yes=1, no=0, n/a=0})\\
-	Does the getting started tutorial provide an expected output? ({yes=1, no$^*$=0,
-	n/a=0})\\
-	Does your tutorial output match the expected output? ({yes=1, no=0, n/a=0})\\
-	Are unit tests available?  ({yes=1, no=0, unclear=0})\\
-	Is there evidence of continuous integration? (for example mentioned in
-	documentation, Jenkins, Travis CI, Bamboo, other) ({yes$^*$=1, no=0,
-	unclear=0})\\
-	\hline	
-	\textbf{Surface Reliability}\\
-	\hline
-	Did the software “break” during installation? ({yes$^*$=0, no=5})\\
-	If the software installation broke, was the installation instance recoverable?
-	({yes=5, no=0, n/a=0})\\
-	Did the software “break” during the initial tutorial testing? ({yes$^*$=0, no=5,
-	n/a=0})\\
-	If the tutorial testing broke, was a descriptive error message displayed?
-	({yes=2, no=0, n/a=0})\\
-	If the tutorial testing broke, was the tutorial testing instance recoverable?
-	({yes=3, no=0, n/a=0})\\
-	\hline		
-	\textbf{Surface Robustness}\\
-	\hline
-	Does the software handle unexpected/unanticipated input (like data of the
-	wrong type, empty input, missing files or links) reasonably? (a reasonable
-	response can include an appropriate error message.) ({yes=5, no$^*$=0})\\
-	For any plain text input files, if all new lines are replaced with new lines
-	and carriage returns, will the software handle this gracefully? ({yes=5,
-	no$^*$=0, n/a=5})\\
-	\hline		
-\end{tabular}
-
-\def\arraystretch{1.4}
-\begin{tabular}{p{14cm}}
-		\hline		
-	\textbf{Surface Usability}\\
-	\hline
-	Is there a getting started tutorial? ({yes=3, no=0})\\
-	Is there a user manual? ({yes=4, no=0})\\
-	Are expected user characteristics documented? ({yes=1, no=0})\\
-	What is the user support model? FAQ? User forum? E-mail address to direct
-	questions? Etc. (one=1, two+=2, none=0)\\
-	\hline
-	\textbf{Maintainability}\\
-	\hline
-	What is the current version number? (provided=1, nothing=0)\\
-	Is there any information on how code is reviewed, or how to contribute?
-	({yes$^*$=1, no=0})\\
-	Are artifacts available? (List every type of file that is not a code file –
-	for examples please look at the ‘Artifact Name’ column of
-	https://gitlab.cas.mcmaster.ca/SEforSC/se4sc/-/blob/git-svn/GradStudents/Olu/ResearchProposal/Artifacts\_MiningV3.xlsx)
-	(Rate 0 – 2 depending on how many and perceived quality)\\
-	What issue tracking tool is employed? (nothing=0, email of other private=1,
-	anything public or accessible by all devs (eg git) = 2)\\
-	What is the percentage of identified issues that are closed? (50$\%$+=1,
-	$<$50$\%$=0)\\
-	What percentage of code is comments? (10$\%$+=1, $<$10$\%$=0)\\
-	Which version control system is in use? (anything=2, nothing=0)\\
-	\hline		
-	\textbf{Reusability}\\
-	\hline
-	How many code files are there? (0-9=0, 10-49=1, 50-99=3, 100-299=4, 300-599=5,
-	600-999=6, 1000+=8)\\
-	Is API documented? ({yes=2, no=0, n/a=0})\\
-	\hline
-\end{tabular}
-
-\def\arraystretch{1.4}
-\begin{tabular}{p{14cm}}
-\hline	
-\textbf{Surface Understandability (Based on 10 random source files)}\\
-\hline
-Consistent indentation and formatting style? ({yes=1, no=0, n/a=0})\\
-Explicit identification of a coding standard? ({yes$^*$=1, no=0, n/a=0})\\
-Are the code identifiers consistent, distinctive, and meaningful? ({yes=2,
-no$^*$=0, n/a=0})\\
-Are constants (other than 0 and 1) hard coded into the program? ({yes=1, no$^*$=0,
-n/a=0})\\
-Comments are clear, indicate what is being done, not how? ({yes=2, no$^*$=0,
-n/a=0})\\
-Is the name/URL of any algorithms used mentioned? ({yes=1, no$^*$=0, n/a=0})\\
-Parameters are in the same order for all functions? ({yes=1, no$^*$=0, n/a=0})\\
-Is code modularized? ({yes=1, no$^*$=0, n/a=0})\\
-\hline		
-\textbf{Visibility/Transparency}\\
-\hline
-Is the development process defined? If yes, what process is used. ({yes$^*$=3,
-no=0, n/a=0})\\
-Are there any documents recording the development process and status?  ({yes$^*$=3,
-no=0}))\\
-Is the development environment documented? ({yes$^*$=2, no=0})\\
-Are there release notes? ({yes$^*$=2, no=0})\\
-\hline
-\end{tabular}
-
-\newpage
-
-\section{Survey Questions for Developers} \label{SecSurveyQuestions}
-
-Interviews will be one-to-one and will be open-ended (not just “yes or no”
-answers). Because of this, the exact wording of the questions may change a
-little. If more information, or clarification, is needed during the
-conversation, follow up questions will be asked, such as: ``So, you are saying
-that ...?'', ``Please tell me more?'', or ``Why do you think that is?''
-
-\subsection{Information about the developers and users}
-
-These questions are based on questions used in \citet{Jegatheesan2016}
-
-\begin{enumerate}
-\item Interviewees' current position/title? degrees?
-\item Interviewees' contribution to/relationship with the software?
-\item Length of time the interviewee has been involved with this software?
-\item How large is the development group?
-\item Do you have a defined process for accepting new contributions into your team?
-\item What is the typical background of a developer?
-\item What is your estimated number of users? How did you come up with that estimate?
-\item What is the typical background of a user?
-\end{enumerate}
-
-\subsection{Information about the software}
-
-Square brackets are used for traceability to the  relevant research
-questions found in Section~\ref{SecResearchQuestions}.
-
-\begin{enumerate}
-\item Currently, what are the most significant obstacles in your development
-process?
-\item How might you change your development process to remove or reduce these
-obstacles?
-\item How does documentation fit into your development process? Would improved
-documentation help with the obstacles you typically face? [research question 5b (traceability),
-research question 5i (visibility/transparency)]
-\item In the past, is there any major obstacle to your development process that
-has been solved? How did you solve it?
-\item What is your software development model? For example, waterfall, agile,
-etc.
-\item What is your project management process? Do you think improve this process
-can tackle the current problem? Were any project management tools used?
-\item Was it hard to ensure the correctness of the software? If there were any
- obstacles, what methods have been considered or practiced to improve the
- situation? If practiced, did it work? [research question 5e (correctness)]
-\item When designing the software, did you consider the ease of future changes?
- For example, will it be hard to change the structure of the system, modules or
- code blocks? What measures have been taken to ensure the ease of future changes
- and maintains? [research question 5d (maintainability), 
- research question 5c (modifiability)]
-\item Provide instances where users have misunderstood the software. What, if
-any, actions were taken to address understandability issues? [research question
-5f (understandability)]
-\item What, if any, actions were taken to address usability issues? [
-research question 5a (usability)]
-\item Do you think the current documentation can clearly convey all necessary
-knowledge to the users? If yes, how did you successfully achieve it? If no, what
-improvements are needed? [research question 5g (unambiguity)]
-\item Do you have any concern that your computational results won't be
-reproducible in the future? Have you taken any steps to ensure reproducibility?
-[research question 5h (reproducibility)]
-\end{enumerate}
-
-\newpage
-
-\bibliographystyle {plainnat}
-\bibliography {../../CommonFiles/ResearchProposal}
-
+process) the time to complete an assessment for a given domain at 173 person
+hours.
+
+\subsubsection{Acknowledgements} Please place your acknowledgments at
+the end of the paper, preceded by an unnumbered run-in heading (i.e.
+3rd-level heading).
+
+%
+% ---- Bibliography ----
+%
+% BibTeX users should specify bibliography style 'splncs04'.
+% References will then be sorted and formatted in the correct style.
+%
+% \bibliographystyle{splncs04}
+% \bibliography{mybibliography}
+%
+\begin{thebibliography}{8}
+\bibitem{ref_article1}
+Author, F.: Article title. Journal \textbf{2}(5), 99--110 (2016)
+
+\bibitem{ref_lncs1}
+Author, F., Author, S.: Title of a proceedings paper. In: Editor,
+F., Editor, S. (eds.) CONFERENCE 2016, LNCS, vol. 9999, pp. 1--13.
+Springer, Heidelberg (2016). \doi{10.10007/1234567890}
+
+\bibitem{ref_book1}
+Author, F., Author, S., Author, T.: Book title. 2nd edn. Publisher,
+Location (1999)
+
+\bibitem{ref_proc1}
+Author, A.-B.: Contribution title. In: 9th International Proceedings
+on Proceedings, pp. 1--2. Publisher, Location (2010)
+
+\bibitem{ref_url1}
+LNCS Homepage, \url{http://www.springer.com/lncs}. Last accessed 4
+Oct 2017
+\end{thebibliography}
 \end{document}

### 2022-02-15 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Updates to SOP for LBM, start of digging deeper paper

@@
 \newcommand{\ppref}[1]{P\ref{#1}}
+
+\newcounter{bpnum} %best practice number
+\newcommand{\bpthepnum}{BP`'\thebpnum}
+\newcommand{\bpref}[1]{BP\ref{#1}}
 
@@
 	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Concerns}:] For LBM
-	developers what specific practices are taken to address the pain point
-	concerns? (Section~\ref{Sec_AddressConcerns})
+	developers what specific best practices are taken to address the pain points
+	and software quality concerns? (Section~\ref{Sec_AddressConcerns})
 	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Recommend}:]
-	What, if any, existing research software best practice, not currently used
-	by LBM software, could address the concerns identified in
-	\rqref{RQ_Concerns}). (Section~\ref{Sec_AddressConcerns})
-	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Future}:] What practices
-	could be employed by future LBM developers to address the pain points found
-	in~\rqref{RQ_PainPoints} and the concerns identified
-	in~\rqref{RQ_Concerns}? (Section~\ref{Sec_AddressConcerns})
-
+	What research software development practice could potentially address the
+	pain point concerns identified in \rqref{RQ_PainPoints}).
+	(Section~\ref{Sec_Recommendations})
 \end{enumerate}
@@
 	\textbf{Ensuring Correctness} Difficulties with ensuring correctness were
-	noted by several developers. The TCLB developer commented that the amount of
-	testing data needed was sometimes problematic, since free testing services
-	do not offer adequate facilities for large amounts of data, which means
-	in-house testing solutions are needed.  Other research software domains
-	point to different problems with testing, as follows: i)
-	\citep{PintoEtAl2018} mention the problem of insufficient testing; ii) the
-	survey of \citep{HannayEtAl2009} shows that more developers think testing is
-	important than the number that believe they have a sufficient understanding
-	of testing concepts; and, iii) \citep{HannayEtAl2009, KanewalaAndBieman2013,
-	KellyEtAl2011, WieseEtAl2019} point to the oracle problem, which
-	occurs in research software when we do not have a means to judge the
-	correctness of the calculated solutions.  Since LBM developers did not
-	allude to the oracle problem in our conversation, we believe that they have
-	developed techniques to work around the oracle problem.
+	noted by several developers. Several developers alluded to difficulty with
+	testing the correctness of large numbers of features, and some manually
+	tested program output, as opposed to using automated testing. The TCLB
+	developer commented that the amount of testing data needed was sometimes
+	problematic, since free testing services do not offer adequate facilities
+	for large amounts of data, which means in-house testing solutions are
+	needed.  Other research software domains point to the following problems
+	with testing: i) \citep{PintoEtAl2018} mention the problem of insufficient
+	testing; ii) the survey of \citep{HannayEtAl2009} shows that more developers
+	think testing is important than the number that believe they have a
+	sufficient understanding of testing concepts; and, iii)
+	\citep{HannayEtAl2009, KanewalaAndBieman2013, KellyEtAl2011, WieseEtAl2019}
+	point to the oracle problem, which occurs in research software when we do
+	not have a means to judge the correctness of the calculated solutions.  The
+	LBM experience seems to overlap with i and ii, but the LBM developers did
+	not allude to problem iii (the oracle problem) in our conversation; we
+	believe that they have developed techniques to work around the oracle
+	problem.
 
@@
 	\item[P\refstepcounter{pnum}\thepnum \label{P_Documentation}:]
-	\textbf{Quality of Documentation} One of the consequences of pain point
-	\ppref{P_LackDevTime} is sometimes inadequate documentation. The
-	documentation for LBM is challenging because of the complexity of the
-	underlying CFD theory and the differing backgrounds of users. Poor research
-	software documentation is also mentioned by others \citep{PintoEtAl2018,
-	WieseEtAl2019} and the problem also arises with non-research software
-	\citep{LethbridgeEtAl2003}.  Recommendations on the development of research
-	software state that developers should critically evaluate their own
-	development processes in terms of quality assurance and comply with
-	international standards for software documentation
-	\citep{KaterbowAndFeulner2018}.
+	\textbf{Quality of Documentation}  The importance of documentation for both
+	users and developers was stressed throughout the interviews. However, it was
+	noted several times that a lack of time and funding (\ppref{P_LackDevTime})
+	has a negative affect on the documentation. Most of the developers are
+	scientific researchers evaluated on the scientific papers that they produce.
+	Writing and updating documentation is something that is done in their free
+	time, if that time arises. Inadequate research software documentation is
+	also mentioned by others \citep{PintoEtAl2018, WieseEtAl2019} and the
+	problem also arises with non-research software \citep{LethbridgeEtAl2003}.
+	Recommendations on the development of research software state that
+	developers should critically evaluate their own development processes in
+	terms of quality assurance and comply with international standards for
+	software documentation \citep{KaterbowAndFeulner2018}.
 
@@
 
-\section{LBM Developer Identified Concerns} \label{Sec_Concerns}
-
-\subsection{Correctness and Verifiability}
-
-Interviews with developers confirmed that these software packages are developed
-by domain experts with backgrounds in physics, mathematics and mechanical
-engineering. \wss{Could point to surveys that have shown the same thing.}  It
-was noted in these interviews that some of the developers do not have formal
-software engineering education. \wss{Can point to other studies to support
-this.} Some of the development teams include computer scientists. Despite a lack
-of visible domain documentation and a resulting lower surface correctness and
-verifiability score, it is clear that some of the software packages were
-developed by teams with significant domain knowledge.
-
-Interviews suggested a more frequent use of both unit testing and continuous
-integration in the development processes than what was observed from the initial
-survey. For example, OpenLB, pyLBM, and TCLB use such methods during
-development, despite this not being explicitly clear from an analysis of the
-material available online. \wss{Add this to threats to validity.}  The
-correctness and verifiability of such packages is not measured well using
-surface analysis.
-
-Several developers alluded to difficulty with testing the correctness of large
-numbers of features, and some even manually tested program output, as opposed to
-using automated testing. The use of well defined unit testing tools could
-decrease the time spent testing some feature.  \wss{link this to
-recommendations}
-
-\subsection{Usability}
-
-Interviews with developers revealed several usability issues. Some users have
-misunderstood the boundaries of LBM and CFD, and have combined or applied
-methods that are not physically sound. Sometimes users have applied LBM to
-poorly defined or inappropriate fluid dynamics problems. For example, they may
-wish to model flow through or around a structure despite having limited
-information about the structure or its environment, and having little previous
-knowledge of CFD. The users do not realize the limitations of the methods, of
-the software, and do not understand the requirements to properly model a problem
-with the software. As the developer of TCLB noted, such software packages are
-not designed to be used ``out of the box'' in a plug and play fashion, and it
-could take months or more to set up the CFD problems correctly. Developers of
-some software packages, including ESPResSo, mitigated this by editing the source
-code to prevent users from ``combining methods that are not physically sound
-together'', and by updating the documentation to better inform users of LBM
-limitations, and of the requirements to properly model appropriate problems,
-including what algorithms and parameters to use.  \wss{Could point to this for
-recommendations related to documentation.  Could point to this for a generative
-approach that won't allow inadmissible combinations, or at least includes an
-explicit warning.}
-
-Usability concerns were also raised for installation, understanding how to
-maneuver the interfaces and understanding how to set up and run models. These
-issues are addressed by various user support models, including frequently asked
-questions sections on the software package websites, user guides, and hardware
-and software requirements specifications.  \wss{Could point to there being room
-for improvement here, especially for requirements specification.}
-
-One software package (ESPResSo) changed some of its scripting language to Python
-to make it more usable. The developer commented that this was ``the biggest step
-in terms of usability over the years'', further commenting that ``most people in
-the field know [Python]'' and that ``it's easy to learn''.  \wss{A generative
-approach facilitates switching to whatever language is currently popular.}
-
-\subsection{Maintainability}
-
-Although not always visible from the publicly available artifacts, interviews
-with developers revealed that most projects have a defined process for accepting
-contributions from team members. The packages rarely get contributions from
-outside developers, but the process would be similar as for the aforementioned
-group.  In general, contributions are made through GitHub, and are then reviewed
-and pulled by lead developers, often with consultation with a group of core
-developers depending on the organizational model. Continuous integration is part
-of the process for some packages. \wss{Recommendation to increase.  Successful
-projects use it, a reasonable next step for projects that have not yet
-incorporated CI.}  \wss{There is definitely a great deal of repetition.}
-
-Some developers noted that their software package does not have well defined
-contributing guide in the repository, but it might be a good idea to add one in
-the near future. They would be happy to see contribution from outside of their
-organization, but currently this does not happen.  \wss{A comment was made at
-the recent DOE workshop that you get new users through documentation and
-tutorials.  Maybe highlight the example of Musubi?  Should trace back to this
-for the recommendations part.  \citep{Fogel2005} gives advice on having a
-successful project, also the web article (if needed).}
-
-Furthermore, maintainability has been addressed by increasing source code
-modularity, reducing duplicate information, and improving abstraction by
-developing well defined interfaces. This was noted by the developers of ESPResSo
-and pyLBM. Several software packages have had sections of their code base
-redeveloped with languages that the developers felt are more understandable and
-readable, and that are better supported, such as Python. Data structures have
-also been redeveloped and storage has been improved. A developer of pyLBM
-mentioned that the geometries and models of their system had been ``decoupled'',
-using abstraction and modularization of the source code, to make it ``very easy
-to add [new] features''.  \wss{Highlights the advantage of design.  Other
-projects can look to these examples.  Maybe cite the Anshu paper about Flash?
-Dune?  and point to the advantages of design.}
-
-The developer of TCLB mentioned that their package had two sets of code, for
-executing the models on the CPU and GPU, and that maintenance was decreased by
-introducing macros, a practice which became a common part of the development
-process.
-
-\subsection{Modifiability}
-
-Software packages were not quantitatively measured for modifiability. In this
-project we asked developers to comment on modifiability when we interviewed
-them. Specifically, we asked if ease of future changes to the system, modules,
-and code blocks was considered when designing the software. We also asked if any
-measures had been taken to ensure the ease of future changes. All of the
-developers that were interviewed noted that the ease of future changes was
-considered and that measures to ensure it had been taken, including requiring
-the separation of software components in the source code architecture.  \wss{A
-good approach for other developers to adopt - part of recommendations.  Trace
-back to this comment.}  A high degree of code modularity and abstraction was
-noted by developers as a measure to ensure the ease of future changes. This can
-be ensured by separating components and hiding information behind well defined
-interfaces. The developer of ESPResSo also noted that some of the code base was
-transitioned from C to C++, which could ease modifiability of that software
-package. The developer of TCLB noted that their software package was designed to
-allow for the addition of some LBM features, but changes to major aspects of the
-system would be difficult. For example, ``implementing a new model will be an
-easy contribution'', but changes to the ``Cartesian mesh...will be a
-nightmare''. Furthermore, the package was designed with flexible data structures
-and storage in mind.  \wss{example suggests identifying likely changes.}
+\section{Lessons from LBM Developers} \label{Sec_AddressConcerns}
+
+This section summarizes the best practices from LBM developers that are taken to
+address the pain points mentioned in Section~\ref{painpoints}.  The main source
+of information is the qualitative data from developer interviews
+(Section~\ref{SecSurvey}).  The practices summarized in this section can
+potentially be emulated by the LBM software packages that do not currently
+follow them.  Moreover, these practices may also provide examples that can be
+followed by other research software domains.
+
+\subsection{Design For Change} \label{Sec_DesForChange}
+
+Increasing source code modularity, reducing duplicate information,
+and improving abstraction by developing well defined interfaces. This was noted
+by the developers of ESPResSo and pyLBM.  A developer of pyLBM mentioned that
+the geometries and models of their system had been ``decoupled'', using
+abstraction and modularization of the source code, to make it ``very easy to add
+[new] features''.  \wss{Highlights the advantage of design.  Other projects can
+look to these examples.  Maybe cite the Anshu paper about Flash? Dune? and point
+to the advantages of design.} Addresses maintainability.  The developer of TCLB
+mentioned that their package had two sets of code, for executing the models on
+the CPU and GPU, and that maintenance was decreased by introducing macros, a
+practice which became a common part of the development process.
+
+A high degree of code modularity and abstraction was noted by developers as a
+measure to ensure the ease of future changes. This can be ensured by separating
+components and hiding information behind well defined interfaces. The developer
+of ESPResSo also noted that some of the code base was transitioned from C to
+C++, which could ease modifiability of that software package. The developer of
+TCLB noted that their software package was designed to allow for the addition of
+some LBM features, but changes to major aspects of the system would be
+difficult. For example, ``implementing a new model will be an easy
+contribution'', but changes to the ``Cartesian mesh...will be a nightmare''.
+\wss{example suggests identifying likely changes.}
+
+Modularize the source code, separate components, hide information behind well
+defined interfaces. This is suggested in SCS literature, and in developer
+comments.  \wss{one of Wilson's best practices.}
+
+\begin{itemize}
+	\item Modifiability. Consider flexibility of data structures and data
+	storage in the design stage. The package pyLBM redeveloped data structures
+	to ease future changes.
+	\item Reusability. Provide API documentation, if applicable. Only one
+	(ESPResSo) of the top five ranked packages provided API documentation.
+\end{itemize}
+
+\subsection{Verification and Validation} \label{Sec_VnV}
+
+\wss{Address Correctness (\ppref{P_Correctness}): Testing is not the only way to
+build confidence in correctness.  Peer review.  }
+
+Surface Correctness and Verifiability. Development teams should
+include both domain experts and experienced software developers. This
+suggestion is based on developer comments.
+
+Visibility and Transparency. Consider peer review processes to assess
+contributions and ensure the tracking of information. High ranked package
+ESPResSo uses a peer review process for contributions.
+
+\subsection{Reproducibility} \label{SecReproducibility}
+
+Software packages were not quantitatively measured for reproducibility. Instead,
+developers were asked if they had concerns that their computational results
+would not be reproducible in the future, and if they had taken any steps to
+ensure reproducibility.  The developers all agreed that reproducibility is
+important, but they differ in how far they go to support reproducibility.
+ESPResSo does the most to support reproducibility.  For instance, they compare
+the results of their methods against manually calculated results. These
+comparison tests are automatically run for all source code changes. The tests
+are run when a pull request is opened on GitHub. Even once these tests are
+complete, a peer review process is undertaken before changes are fully committed
+to the appropriate branch. The results for all of the LBM schemes on the
+software package development branch are also frequently compared for
+correctness, ensuring that the system output reflects the expected output.
+
+For other teams, the priority is verifying the mathematical foundations of the
+models, but the output of the software package is not compared to other output.
+Depending on the package and how it outputs solutions, this may not be practical
+or feasible. A correct output may not be exactly reproducible, as it may be
+dependent on a probability distribution, so strictly comparing results may not
+be appropriate. \wss{Ideas for dealing with reproducibility in these situations
+will be explored in the recommendations section (like properties of the solution,
+such as conservation of mass).}
 
@@
 
-Developers were asked to comment on the obstacles in their development process.
-The developer of ESPResSo noted that a lot of their source code had been written
-with a specific application in mind, and that there is too much coupling between
-components. Addressing this issue, possibly by making the design review process
-more transparent (Section~\ref{secVisib}), would help with code modifiability
-and reusability.  \wss{Traceback for recommendations.}
-
-\subsection{Understandability}
-
-Software developers noted that they believe users have generally found their
-packages to be understandable. \wss{Confusion between understandability and
-usability - threat to validity example.}  The interviewed developer of ESPResSo
-commented that some users have attempted to run physically incompatible LBM
-methods, and the solution was to edit the code to prevent such combinations, as
-well as to update the documentation to prevent misunderstanding the methods.
-Similarly, a developer of pyLBM noted that some users had issues setting up
-parameters for LBM schemes. The solution to this was to update the interface
-where these parameters are set, as well as to add functionality to test the
-stability of the parameters. A developer of OpenLB noted that some users lack
-the background knowledge to easily model fluid dynamics problems using their
-software. A frequently asked questions section was added to their package
-website to help users find answers to common questions. The package also has
-detailed documentation, including guides and usage requirements specification,
-to better help users understand the software. \wss{Peter - the discussion here
-is more about user usability than the understandability of the code for
-developers. Did any of the developers discuss developer understandability?  The
-confusion between understandability and usability was also a confusion during
-the interviews Ao conducted.} \pmm{The developer of ESPResSo did comment on
-developer understandability, specifically for developers that are/were part of
-their peer review process. He commented that code architecture is documented,
-and that code is made to be readable, and that these help peer reviews of the
-code.}
-
-\wss{I moved this from the visibility section, since it is more about
-understandability.}  Developers were also asked how documentation fits into their
-development process. Several developers noted that developer documentation plays
-an important role in familiarizing potential contributors to the software system
-architecture. Without the guidance that the documentation provides it would be
-unlikely that contributions would pass the peer review process.
-
-\subsection{Traceability}
+\begin{itemize}
+\item Reproducibility. Test output against automatically calculated or known
+correct results. High ranked package ESPResSo does this.
+\item Reproducibility. Automate the testing of output. Run tests after all code
+changes. High ranked package ESPResSo does this.
+\item Reproducibility. Consider peer review processes and task based inspection
+to assess contributions. High ranked package ESPResSo does this.
+\end{itemize}
+
+\subsection{Documentation} \label{Sec_RecommendDoc}
 
@@
 
-Developers noted the importance of documentation for both the users and
-developers of their software. New features are always added to the
-documentation. The developers use documentation to stay up to date on the status
-of the software package, and to help expand features, like computational models
-or algorithms. This is necessary so that the coding standard for these models is
-kept consistent for new developers. \wss{Peter - this also sounds like
-discussion that is relevant for understandability.} \pmm{The points here sound
-like they could work for both. The definition of traceability notes that `This
-quality refers to keeping track of information as it changes forms or relates
-between artifacts'. So the point `New features are always added to the
-documentation' could be reworded to fit that part of the definition of
-traceability. As noted above, this is similar to the point about some packages
-having including detailed change logs. Same with `The developers use
-documentation to stay up to date on the status of the software package, and to
-help expand features'. I agree that at least some of this paragraph could be in
-understandability, especially the first and second sentences.}
-
-The importance of documentation for both users and developers was stressed
-throughout the interviews. However, it was noted several times that a lack of
-time and funding has a negative affect on the documentation. Most of the
-developers are scientific researchers evaluated on the scientific papers that
-they produce. Writing and updating documentation is something that is done in
-their free time, if that time arises. Sometimes it is a last priority for the
-developers. Finding ways to hasten updating documentation would increase the
+Finding ways to hasten updating documentation would increase the
 frequency of such updates and benefit both users and developers.  \wss{Same
@@
 \wss{We can refer back to this for the recommendations.}
-
-\subsection{Visibility and Transparency} \label{secVisib}
-
-For the developers interviewed, visibility is not currently a priority.
-Visibility is not particularly necessary because the teams are fairly small.
-Although they accept outside contributors, these are rare and the teams remain
-tight-knit.  Most of the teams interviewed have all members at the same
-institution, although one team is spread out internationally. \wss{Peter, can
-you say which team is international, so that we can add this detail to the
-paper?} \pmm{The interviewed developer of OpenLB noted that there were tasks for
-local and international developers when discussing their software development
-model. He also noted that they have frequent local meetings and annual
-international meetings.}  For teams of only a few developers, new features or
-major changes are discussed with the entire team. Projects with more than a few
-developers have lead developer roles. These lead developers review potential
-additions to the software.
-
-Although visibility is not a priority, positive steps are being undertaken. For
-instance, all interviewed teams used GitHub, with its visibility enhancing
-features of branching, issue tracking, insights on contributors, the blame
-feature, wikis, pull requests, etc. Continuous integration, as used by ESPResSo,
-makes projects even more visible, since contributors can explicitly see the bar
-that must be crossed for code to be incorporated into the master branch.
-ESPResSo also uses peer review, which improves visibility by explicitly asking
-for feedback from other developers.
-
-Previously mentioned concerns with maintainability, modifiability, and
-understandability  could be improved by further improving visibility.  For
-future sustainability, projects will likely have to expand their development
-teams by recruiting more external developers. One component of recruiting
-developers is transparency in the development process, including the process for
-making a contribution.  \wss{in recommendations could point out what teams could
-do to get more external contributors, and potentially a longer life for their
-project.}
-
-\subsection{Reproducibility} \label{SecReproducibility}
-
-Software packages were not quantitatively measured for reproducibility. Instead,
-developers were asked if they had concerns that their computational results
-would not be reproducible in the future, and if they had taken any steps to
-ensure reproducibility.  The developers all agreed that reproducibility is
-important, but they differ in how far they go to support reproducibility.
-ESPResSo does the most to support reproducibility.  For instance, they compare
-the results of their methods against manually calculated results. These
-comparison tests are automatically run for all source code changes. The tests
-are run when a pull request is opened on GitHub. Even once these tests are
-complete, a peer review process is undertaken before changes are fully committed
-to the appropriate branch. The results for all of the LBM schemes on the
-software package development branch are also frequently compared for
-correctness, ensuring that the system output reflects the expected output.
-
-For other teams, the priority is verifying the mathematical foundations of the
-models, but the output of the software package is not compared to other output.
-Depending on the package and how it outputs solutions, this may not be practical
-or feasible. A correct output may not be exactly reproducible, as it may be
-dependent on a probability distribution, so strictly comparing results may not
-be appropriate. \wss{Ideas for dealing with reproducibility in these situations
-will be explored in the recommendations section (like properties of the solution,
-such as conservation of mass).}
-
-For the current project, we did not assess the stronger form of
-``reproducibility'' known as replicability.  A software project is considered
-replicable if the documentation alone (without the code) is sufficiently precise
-and complete to allow a third-part to re-obtain the results of the original code
-\citep{BenureauAndRougier2017}.  Replicability is difficult to achieve with
-current documentation practices.  This question should be considered in the next
-iteration of state of the practice assessments.
-
-\subsection{Unambiguity}
-
-Software packages were not quantitatively measured for unambiguity. We asked
-developers if they thought that the current documentation can clearly convey all
-necessary knowledge to the users, and if they had taken any steps to ensure
-clarity. The developer of ESPResSo noted that their documentation was meant for
-users that are already familiar with the underlying physics and CFD methods.
-These concepts are not explained in detail within the documentation. Users
-should acquire this knowledge from suitable external sources. The documentation
-focuses on how to technically use the software package, and includes a user
-guide and tutorial walk through of how to set up and run a simulation. With this
-in mind, the developer believes that their documentation is in reasonable shape
-for users with a minimum knowledge of the underlying physics. If new users have
-technical questions these can be addressed in further revisions of the
-documentation. New developers should find that the documentation is reasonably
-clear and useful. Information that is missing, like detailed explanations of
-dependencies, is referenced in the documents.
-
-The developer of pyLBM also noted that their documentation was in reasonable
-shape, but that they ``need more [user] feedback to improve [it]''. They also
-noted that they believe a lack of knowledge of the underlying physics and CFD
-concepts can be an issue for some users. This information can be referenced in
-the documentation, but it is not something that the documentation needs to
-detail.
-
-\section{Lessons from LBM Developers} \label{Sec_AddressConcerns}
-
-Answers \rqref{RQ_Recommend}.  Recommendations for Addressing Developer
-Concerns.  This section answers the research questions listed in
-Section~\ref{secIntro} using the quantitative data presented in
-Section~\ref{AHPresults}, qualitative data from developer interviews
-(Section~\ref{SecSurvey}), additional data from software package repositories
-and artifacts, and domain expert feedback.
-
-\wss{Address lack of development time (\ppref{P_LackDevTime}): Aside from
-on-boarding new developers, time constraints could be mitigated by increasing
-developer efficiency, which could be addressed in several ways, including by
-improving the quality of documentation, or incorporating automatic code
-generation.}
-
-\wss{Address lack of funding (\ppref{P_LackFunding}): The developer further
-commented that there are no journals that publish such scientific software
-source code. However, there are ways to get such source code cited. Work has
-been done to address this in \citep{SmithEtAl2016-softcite}, which presents a
-set of software citation principles and discusses ``how they could be used to
-implement software citation in the scholarly community''
-\citep{katz2019software}. other papers also mention this pain point.
-Solutions proposed by Katz, journal of open source etc. Journal of Open Source
-Software \citet{SmithEtAl2018-joss}.  Calling for direct software citations
-\citep{SmithEtAl2016-softcite}.  Guidance on software citation from
-\citet{KatzEtAl2021}.  Discussed in CarverEtAl2021 paper.}
 
@@
 
-\wss{Address Correctness (\ppref{P_Correctness}): Testing is not the only way to
-build confidence in correctness.  Peer review.  }
-
-\wss{Address quality of documentation (\ppref{P_Documentation}): Make required
-background of users explicit, possibly in a requirements document.}
-
-\subsection{Common Practices} \label{currentPractices}
-
-The practices that arose frequently.  Compare and contrast to what is done with
-research software in general.  The following points regarding software quality
-should be considered when developing LBM software packages. These points are
-based on developer interviews, SCS literature, what was found to have worked for
-packages that were designated as high quality in this assessment.  Our
+One software package (ESPResSo) changed some of its scripting language to Python
+to make it more usable. The developer commented that this was ``the biggest step
+in terms of usability over the years'', further commenting that ``most people in
+the field know [Python]'' and that ``it's easy to learn''.  \wss{A generative
+approach facilitates switching to whatever language is currently popular.}
+
+\begin{itemize}
+\item Usability. Understandability. Unambiguity. State appropriate fluid
+dynamics problems that the software is designed to model in the documentation,
+and explicitly state the limits of the software. This is done in the ESPResSo
+user guide.
+\item Usability. Provide documentation that details the background theory
+information, or provide a reference to such information. This was done by
+several top ranked packages, including ESPResSo, Ludwig, LUMA, and OpenLB.
+\item Usability. Identify expected user characteristics. LIMBES, Ludwig,
+laboetie, and Palabos did this. The importance of specifying user
+characteristics is discussed in \citep{smith2007requirements}.
+\item Usability. Keep all documentation in one location. This was done by top
+ranked packages.
+\item Usability. Do not duplicate artifact information. This is suggested as
+good software development practice. Duplicate information is difficult to
+maintain. 
+\end{itemize}
+
+\subsection{Visibility and Transparency} \label{secVisib}
+
+\wss{Will likely remove this section, but should find a way to say the good
+things that ESPResSo does.} Although visibility is not a priority, positive
+steps are being undertaken. For instance, all interviewed teams used GitHub,
+with its visibility enhancing features of branching, issue tracking, insights on
+contributors, the blame feature, wikis, pull requests, etc. Continuous
+integration, as used by ESPResSo, makes projects even more visible, since
+contributors can explicitly see the bar that must be crossed for code to be
+incorporated into the master branch. ESPResSo also uses peer review, which
+improves visibility by explicitly asking for feedback from other developers.
+
+\wss{in recommendations could point out what teams could do to get more external
+contributors, and potentially a longer life for their project.}
+
+\section{Recommendations for Future Practices} \label{Sec_Recommendations}
+
+Answers \rqref{RQ_Recommend}.  practices for the future.  Probably the easiest
+approach is to list the practice and then the PPs that it helps.  Our
 recommendations are not lists of what should have been done in the past, or what
 should be done now; they are just suggestions for consideration in the future.
-\wss{Rework this section to highlight what the top projects are doing.  Point to
-checklists for other advice?}  Highlights of common practices.  Put in context.
-
-Conclusion that LBM software common practice is similar to the common practice
-elsewhere.  Maybe point out where LBM is ahead.  Use results from common
-artifacts, common process etc. to support conclusion.  (Continuous integration,
-version control, README, Req etc. - point to Table 8.)
-
-\begin{itemize}
-	\item Surface Reliability. Surface Usability. The packages should include
-	detailed tutorials, including dependencies, expected output, and any
-	additional supplementary documentation that may be required. This was done
-	by many packages, including waLBerla, Palabos, MechSys, LUMA, and pyLBM.
-	\wss{Comment from ASCR attendee that success with open source requires good
-	tutorials and documentation - is this published anywhere?  I think I can
-	find this recommendation in other papers.}
-	\item Surface Usability. Unambiguity. Surface Understandability. Provide a
-	detailed user manual. It should identify elements of user interfaces, and
-	identify all requirements to model a system. Thirteen packages have a user
-	manual. ESPResSo has a well detailed manual. \wss{Include this with
-	tutorial?  Find papers that recommend this kind of documentation?}
-	\item Surface Usability. Include user hardware and software requirements
-	documentation. Hardware requirements are rarely listed in the software
-	packages that were assessed. Only those offering GPU processing (ESPResSo,
-	lbmpy, lettuce, pyLBM, Sailfish, TCLB) mentioned any hardware requirements.
-	On the other hand, some sort of software requirements were available for all
-	packages, even if only some dependencies or a compatible operating system
-	were mentioned.  \wss{This is a key part of reproducibility.}
-    \item Not much of a process - find papers that support this - papers
-    recommend "Define clear and transparent contribution, governance and
-    communication processes" JimenezEtAl2017
-\end{itemize}
-
-\subsection{Best Practices} \label{bestPractices}
-
-practices by the best measured.  Point to existing checklist and development
-guides.  This section should only mention things that are new.
-
-\begin{itemize}
-	\item Surface Correctness and Verifiability. Development teams should
-	include both domain experts and experienced software developers. This
-	suggestion is based on developer comments.
-	\item Surface Correctness and Verifiability. Surface Understandability.
-	Maintainability.  Modifiability. Reusability.  Modularize the source code,
-	separate components, hide information behind well defined interfaces. This
-	is suggested in SCS literature, and in developer comments.  \wss{one of
-	Wilson's best practices.}
-	\item Surface Usability. Surface Understandability. Unambiguity. State
-	appropriate fluid dynamics problems that the software is designed to model
-	in the documentation, and explicitly state the limits of the software. This
-	is done in the ESPResSo user guide.
-	\item Surface Usability. Provide documentation that details the background
-	theory information, or provide a reference to such information. This was
-	done by several top ranked packages, including ESPResSo, Ludwig, LUMA, and
-	OpenLB.
-	\item Surface Usability. Identify expected user characteristics. LIMBES,
-	Ludwig, laboetie, and Palabos did this. The importance of specifying user
-	characteristics is discussed in \citep{smith2007requirements}.
-	\item Surface Usability. Keep all documentation in one location. This was
-	done by top ranked packages.
-	\item Maintainability. Have a defined process for accepting contributions,
-	and make public documentation for making contributions to the project. The
-	top five ranked software packages include information on how to contribute. 
-	\item Modifiability. Consider flexibility of data structures and data
-	storage in the design stage. The package pyLBM redeveloped data structures
-	to ease future changes.
-	\item Reusability. Provide API documentation, if applicable. Only one
-	(ESPResSo) of the top five ranked packages provided API documentation.
-	\item Surface Understandability. Adopt a coding standard and document it in
-	artifacts, including some examples. A coding standard needs to be part of
-	continuous integration. Only LUMA, HemeLB, and Musubi explicitly identify
-	coding standards. 
-	\item Traceability. Provide a developer's guide to help orient developers to
-	the artifacts, source code, and system architecture so that they can better
-	document changes. Top ranked package OpenLB has a developer guide. \wss{How
-	is this related to traceability?}
-	\item Visibility and Transparency. Summarize the development process that is
-	used. Provide information on how new users can contribute. Identify the
-	development model by name (waterfall, agile, etc.), if appropriate.  Seven
-	of the software packages have some artifacts outlining the general
-	development process. Eleven packages have information on how to contribute. 
-	\item Visibility and Transparency. Consider peer review processes to assess
-	contributions and ensure the tracking of information. High ranked package
-	ESPResSo uses a peer review process for contributions. 
-	\item Reproducibility. Test output against automatically calculated or known
-	correct results. High ranked package ESPResSo does this.
-	\item Reproducibility. Automate the testing of output. Run tests after all
-	code changes. High ranked package ESPResSo does this.
-	\item Reproducibility. Consider peer review processes and task based
-	inspection to assess contributions. High ranked package ESPResSo does this.
-	\item Unambiguity. Consider asking users for feedback on the documentation.
-	The developer of pyLBM noted that such feedback would be appreciated.
-
-\end{itemize}
-
-practices mentioned in literature, but not practiced (if any).  Start with the
-practices that aren't derived from developer interviews.
-
-\begin{itemize}
-	\item Surface Correctness and Verifiability. Use a requirements
-	specification document. This is suggested in SCS literature, and several of
-	the top ranked packages had such a document or reference to theory manuals.
-	A potential template is presented in \citep{smith2005new}. 
-	\item Surface Usability. Do not duplicate artifact information. This is
-	suggested as good software development practice. Duplicate information is
-	difficult to maintain. 
-	\item Maintainability. Visibility and Transparency. Reproducibility. Keep
-	artifacts updated. This is suggested as good software maintenance practice.
-	\item Maintainability. Eliminate code duplication. This is suggested as good
-	software development practice. Duplicate code is difficult to maintain.
-	\wss{No support for this one or citation, either find or remove it.}
-	\item Modifiability. Consider future source code modifiability as early as
-	the design stage of development. This is suggested as good software
-	development practice. \wss{citation, support, or remove.}
-	\item Reusability. Document module interfaces in design and developer
-	documentation. This is suggested as good software development practice.
-	\wss{citation or support}
-	\item Traceability. Update all relevant documentation when a change to the
-	software is made. This is suggested as good software development practice.
-	\wss{support, cite or remove}
-	\item Unambiguity. Documentation must state all technical requirements and
-	software dependencies. A missing dependency was a frequent cause of fault
-	conditions during this assessment.
-	\item Unambiguity. Provide a table of symbols in the developer's guide that
-	maps to names used in the source code. This is suggested as good software
-	development practice.
-
-\end{itemize}
-
-\subsection{Future Practices} \label{futurePractices}
-
-Answers \rqref{RQ_Future}.  practices for the future.
 
@@
 linters \citep{Zadka2018} were rarely observed in LBM projects.  Do not see much evidence of their presence in best practice guides.
+
+\wss{Address lack of development time (\ppref{P_LackDevTime}): Aside from
+on-boarding new developers, time constraints could be mitigated by increasing
+developer efficiency, which could be addressed in several ways, including by
+improving the quality of documentation, or incorporating automatic code
+generation.}
+
+\wss{Address lack of funding (\ppref{P_LackFunding}): The developer further
+commented that there are no journals that publish such scientific software
+source code. However, there are ways to get such source code cited. Work has
+been done to address this in \citep{SmithEtAl2016-softcite}, which presents a
+set of software citation principles and discusses ``how they could be used to
+implement software citation in the scholarly community''
+\citep{katz2019software}. other papers also mention this pain point.
+Solutions proposed by Katz, journal of open source etc. Journal of Open Source
+Software \citet{SmithEtAl2018-joss}.  Calling for direct software citations
+\citep{SmithEtAl2016-softcite}.  Guidance on software citation from
+\citet{KatzEtAl2021}.  Discussed in CarverEtAl2021 paper.}
+
+\wss{Address quality of documentation (\ppref{P_Documentation}): Make required
+background of users explicit, possibly in a requirements document.}
+
+Surface Correctness and Verifiability. Use a requirements
+specification document. This is suggested in SCS literature, and several of
+the top ranked packages had such a document or reference to theory manuals.
+A potential template is presented in \citep{smith2005new}.
+
+\subsection{Grow the Number of Contributors}
+
+Maintainability. Have a defined process for accepting contributions, and make
+public documentation for making contributions to the project. The top five
+ranked software packages include information on how to contribute.
+
+Papers that talk about how to do open source.
 
@@
 compared to what was observed from studying the repository artifacts.''
+
+Interviews suggested a more frequent use of both unit testing and continuous
+integration in the development processes than what was observed from the initial
+survey. For example, OpenLB, pyLBM, and TCLB use such methods during
+development, despite this not being explicitly clear from an analysis of the
+material available online.  The correctness and verifiability of such packages
+is not measured well using surface analysis.
+
+\subsection{Future State Of The Practice Assessments} \label{futuresop}
+
+As mentioned in Section~\ref{threats}, the measures listed in our measurement
+template may not be broad enough to accurately capture some qualities. Adding or
+extending measures is worth considering. For example, usability experiments and
+performance benchmarks could be incorporated into the assessment.
+
+Section~\ref{SecReproducibility} noted that developers were not asked to comment
+on the reproducibility of their source code from their requirements
+specifications and design documentation. Adding this question to the interview
+guide should be considered in the next iteration of state of the practice
+assessments. Furthermore, as mentioned in Section~\ref{threats}, it was found
+that some pertinent information was not specified in public artifacts. The use
+of unit testing and continuous integration by several packages (OpenLB, pyLBM,
+TCLB) was only discovered during interviews with developers. Adding further
+questions to the interview guide regarding the measures that are on the
+measurement template could reduce instances of incorrect data being collected.
+This additional interview data could be analyzed and incorporated into the AHP
+ranking, ensuring quality designations more accurate represent the true quality
+of the software packages.  \wss{clarify the distinction between usability and
+understandability.}
+
+For the current project, we did not assess the stronger form of
+``reproducibility'' known as replicability.  A software project is considered
+replicable if the documentation alone (without the code) is sufficiently precise
+and complete to allow a third-part to re-obtain the results of the original code
+\citep{BenureauAndRougier2017}.  Replicability is difficult to achieve with
+current documentation practices.  This question should be considered in the next
+iteration of state of the practice assessments.
 
@@
 
-\subsection{Highlighted Recommendations} \label{highlightedrecommendations}
-
 Many software packages would benefit from a rewrite or reorganization of
@@
 
-The following recommendations improve software quality along multiple
-attributes, and provide the greatest return on investment:
-
-\begin{itemize}
-	\item Provide a detailed user manual. It should identify elements of user
-	interfaces, and identify all requirements to model a system. Thirteen
-	packages have a user manual. ESPResSo has a well detailed manual.
-	\item State appropriate fluid dynamics problems that the software is
-	designed to model in the documentation, and explicitly state the limits of
-	the software. This is done in the ESPResSo user guide.
-	\item Include detailed tutorials, including dependencies, expected output,
-	and any additional supplementary documentation that may be required. This
-	was done by many packages, including waLBerla, Palabos, MechSys, LUMA, and
-	pyLBM.
-	\item Keep all documentation in one location. This was done by top ranked
-	packages.
-	\item If possible, consider using popular user-friendly software languages
-	like Python. Especially consider this for parts of the source code that is
-	likely to be modified or reviewed by users. ESPResSo and Sailfish use Python
-	to address several software qualities, including maintainability,
-	modifiability and usability.
-	\item Modularize the source code, separate components, hide information
-	behind well defined interfaces. This is suggested in SCS literature, and in
-	developer comments.
-	\item Include descriptive error messages where appropriate. This was done by
-	most packages that encountered a fault.
-	\item Summarize the development process that is used. Provide information on
-	how new users can contribute. Seven of the software packages have some
-	artifacts outlining the general development process. Eleven packages have
-	information on how to contribute.
-	\item Consider peer review processes and task based inspection to assess
-	contributions. High ranked package ESPResSo does this.
-	\item Use continuous integration tools (Bamboo, Jenkins, and Travis CI) and
-	processes during development. This was done by top ranked packages ESPResSo
-	and Ludwig.
-	\item Use project management tools, including change and version control
-	tools (GitHub, GitLab, CVS), collaboration tools (GitHub, GitLab), and
-	document generation tools (Doxygen, Drasil, Sphinx). This was done by all of
-	the top five ranked packages.
-\end{itemize}
-
 Ensuring software quality can take significant time. An inability to develop
@@
 transforming this knowledge into relevant software artifacts without having to
-manually duplicate knowledge into multiple artifacts \citep{zhao2018}.
-\wss{Yuzhi is not the best reference to use for the Drasil work.}
-
-\subsection{Future State Of The Practice Assessments} \label{futuresop}
-
-As mentioned in Section~\ref{threats}, the measures listed in our measurement
-template may not be broad enough to accurately capture some qualities. Adding or
-extending measures is worth considering. For example, usability experiments and
-performance benchmarks could be incorporated into the assessment.
-
-Section~\ref{SecReproducibility} noted that developers were not asked to comment
-on the reproducibility of their source code from their requirements
-specifications and design documentation. Adding this question to the interview
-guide should be considered in the next iteration of state of the practice
-assessments. Furthermore, as mentioned in Section~\ref{threats}, it was found
-that some pertinent information was not specified in public artifacts. The use
-of unit testing and continuous integration by several packages (OpenLB, pyLBM,
-TCLB) was only discovered during interviews with developers. Adding further
-questions to the interview guide regarding the measures that are on the
-measurement template could reduce instances of incorrect data being collected.
-This additional interview data could be analyzed and incorporated into the AHP
-ranking, ensuring quality designations more accurate represent the true quality
-of the software packages.  \wss{clarify the distinction between usability and
-understandability.}
+manually duplicate knowledge into multiple artifacts \citep{SzymczakEtAl2016}.
 
@@
 
-- add table of features - include Mohammed columns
-- add new program that wasn't measured - means measuring 24, not 23
-- update simple stats for other projects
-- compare to existing survey results (do this in Ao's too)
-- read more of the existing survey studies, as highlighted by Carver paper
 - data on Mendeley
-
-- add a comparison to other studies
-- add conclusions related to how LBM compares to other fields, and the "average" of all SC
 
@@
 
-- value of tools to enforce good behavior of developers, like R, like MOOSE infrastructure
-
-- change RQs - make them more about understanding the range of development
-practices, and identifying projects that are successful.  Also identify room for
-improvement, as identified by the developers themselves, and as identified by
-us.  Compare the state of the practice of LBM to the state of practice of other
-domains.  Where does LBM stand relative to other domains.
-
-- read \citet{WieseEtAl2019} (Naming the Pain in Developing Scientific Software)
+- value of tools to enforce good behaviour of developers, like R, like MOOSE infrastructure
 
@@
 each instance, add page numbers).
-
-- add links or citations to all of the packages (could add them to the summary table)
 
@@
 
-- discuss Wilson papers ``Best Practices'' \citep{WilsonEtAl2014} and ``Good
-Enough Practices'' \citep{WilsonEtAl2016}
-
 - mention flake8 for Python?
@@
 - AlNoamanyAndBorghi2018 lists the most popular languages, Python is number 1
-
-- JimenezEtAl2017 talks about Define clear and transparent contribution,
-governance and communication processes
 

### 2022-02-17 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Minor edits to SOP for LBM paper

@@
 
-To focus our efforts, we devised 11 research questions.  The questions are used
+To focus our efforts, we devised 10 research questions.  The questions are used
 to structure the discussion in the paper, so for each research question below we
@@
 
+\wss{From Peter: The developer of ESPResSo mentioned ``tests which compare
+implementations against each other, like say if there's to solve us for
+electrostatic switch rely on different principles and they've been written by
+different people''.
+The developer of TCLB also mentioned automatic testing. They discussed the idea
+of using test cases, but concerns were raised with this, such as the required
+computing power to run such tests frequently and where to store large amounts of
+data. Also, ``constructing such cases would be very hard''.
+The developer of OpenLB commented that they have different kinds of tests, such
+as integration tests, and tests where they know the ``exact analytical solution''.
+They also mentioned comparing computations and ``check[ing] the misfits''.}
+
 \subsection{Reproducibility} \label{SecReproducibility}

### 2022-02-17 StateOfPractice/Papers/ICCS-2022-SOP_MethodologyPlusExamples/DigDeep.tex
Edits to digging deeper paper

@@
 \usepackage{totcount}
+\usepackage{amssymb}
 
@@
 
+\newcounter{rqnum} %research question number
+\newcommand{\rqtherqnum}{RQ`'\therqnum}
+\newcommand{\rqref}[1]{RQ\ref{#1}}
+
+\newcounter{pnum} %pain point number
+\newcommand{\ppthepnum}{P`'\thepnum}
+\newcommand{\ppref}[1]{P\ref{#1}}
+
+\newcommand{\CC}{C\nolinebreak\hspace{-.05em}\raisebox{.4ex}{\small\bf
++}\nolinebreak\hspace{-.10em}\raisebox{.4ex}{\small\bf +}}
+
 \begin{document}
 %
-\title{Digging Deep to Assess the State of the Practice for Different Research Software Domains}
-%
-%\titlerunning{Abbreviated paper title}
+% \title{Digging Deep to Assess the State of the Practice for Different Research Software Domains}
+\title{Digging Deeper Into the State of the Practice for Domain Specific Research Software}
+\titlerunning{Digging Deeper}
 % If the paper title is too long for the running head, you can set
@@
 %
-\author{Spencer Smith\inst{1}\orcidID{0000-0002-0760-0987}} %\and
-%Second Author\inst{2,3}\orcidID{1111-2222-3333-4444} \and
+\author{Spencer Smith\inst{1}\orcidID{0000-0002-0760-0987} \and
+Peter Michalski\inst{1}}%\orcidID{1111-2222-3333-4444} \and
 %Third Author\inst{3}\orcidID{2222--3333-4444-5555}}
 %
-\authorrunning{S.\ Smith}
+\authorrunning{S.\ Smith and P.\ Michalski}
 % First names are abbreviated in the running head.
@@
   To answer these questions, our methodology prescribes the following steps: 
-  \begin{inparaenum}[i)] 
-    \item Identify the domain;
-    \item Identify a list of candidate software packages;
-    \item Filter the list to a length of about 30 packages;
-    \item Gather source code and documentation for each package;
-    \item Collect repository related data on each software package, like number
-    of stars, number of open issues, number of lines of code;
-    \item Fill in the measurement template (the template consists of 108
-    questions to assess 9 qualities (including the qualities of installability,
-    usability and visibility));
-    \item Interview developers (the interview consists of 20 questions and takes
-    about an hour);
-    \item Rank the software using the Analytic Hierarchy Process (AHP); and,
-    \item Analyze the data to answer the questions posed above.
-  \end{inparaenum}
-  A domain expert should be engaged throughout the process, to ensure that
-  implicit information about the domain is properly represented and to assist
-  with conducting an analysis of the commonalities and variabilities between the
-  30 selected packages.  Using our methodology, spreadsheet templates and AHP
-  tool, we estimate (based on our experience with using the process) the time to
-  complete an assessment for a given domain at 173 person hours.
-
-The abstract should briefly summarize the contents of the paper in
-150--250 words.
+  The abstract should briefly summarize the contents of the paper in 150--250 words.
 
@@
 \end{abstract}
-%
-%
-%
+
 \section{Introduction} \label{SecIntroduction}
 
-- research software is vital for making progress in science and engineering, but it is difficult to develop
-- this has promoted many studies to understand the challenges of research software development
-- the previous studies have usually used interviews, surveys and case studies
-- the interviews and surveys recruit participants from all domains of research software
-- the case studies, by their nature, focus on specific examples
-- what is missing is a methodology that:
-	- goes beyond surveys to look at the actual artifacts in the software repos
-	- focuses on one domain at a time
-- by focusing on one domain at a time we can:
-	- provide useful knowledge/advice to the developers in that domain
-	- build up a set of measures of different domains to compare and contrast them
-
-Research software uses computing to simulate mathematical models of real world
-systems so that we can better understand and predict those systems' behaviour.  A
-small set of examples of important research software includes the
-following: designing new automotive parts, analyzing the flow of blood in the
-body, and determining the concentration of a pollutant released into the
-groundwater.  As these examples illustrate, research software can be used for
-tackling important problems that impact such areas as manufacturing, financial
-planning, environmental policy, and the health, welfare and safety of
-communities.
-
-Given the importance of research software, scientists and engineers are pushing
-for methods and tools to sustainably develop high quality software.  This is
-evident from the existence of such groups as the Software Sustainability
-Institute (\href{https://www.software.ac.uk/} {SEI}) and Better Scientific
-Software (\href{https://bssw.io/} {BSS}).  Sustainability promoting groups such
-as these are necessary because unfortunately the current ``state of the
-practice'' for research software often does not incorporate ``state of the art''
-Software Engineering (SE) tools and methods \cite{JohansonAndHasselbring2018}.
-The lack of SE tools and methods contributes to sustainability and reliability
-problems \cite{FaulkEtAl2009}.  Problems with the current state of the practice
-are evident from embarrassing failures, like a retraction of derived molecular
-protein structures \cite{Miller2006}, false reproduction of sonoluminescent
-fusion \cite{PostAndVotta2005}, and fixing and then reintroducing the same
-error in a large code base three times in 20 year
-\cite{MilewiczAndRaybourn2018}.  To improve this situation, we need to first
-fully understand the current state of the practice for research software
-development.
-
-The purpose of our proposed methodology is to understand how software quality is
-impacted by the software development principles, processes and tools currently
-used within research software communities.  Since research software is so broad
-a category, we will reduce the scope of our methodology to focus on one specific
-research software domain at a time.  To emphasize that the method is generic, we
-will label the specific domain as X within this document.
-
-This ``state of the practice for domain X'' exercise builds off of prior work on
-measuring/assessing the state of software development practice in several
-research domains.  We have updated the work that was done previously for domains
-such as Geographic Information Systems \cite{SmithEtAl2018_arXivGIS}, Mesh
-Generators \cite{SmithEtAl2016}, Seismology software \cite{SmithEtAl2018}, and
-Statistical software for psychology \cite{SmithEtAl2018_StatSoft}.  Initial
-tests of the new methodology have been done for medical image analysis software
-\cite{Dong2021} and for Lattice Boltzmann Method (LBM) software
-\cite{Michalski2021}.
-
-In the previous ``state of the practice'' project, we measured 30 software
-projects for each domain, but the measures were relatively shallow.  With this
-re-boot we still target about 30 software examples from each domain, but we are
-now collecting more data.  In keeping with the previous project, we still have
-the constraint that the work load for applying the methodology to a given domain
+Research software is critical for tackling problems in areas as diverse as
+manufacturing, financial planning, environmental policy and medical diagnosis
+and treatment.  However, developing reliable, reproducible, sustainable and fast
+research software to address new problems is challenging because of the
+complexity of the physical models and the nuances of floating point and parallel
+computation. The importance of research software and the difficulty with its
+development have prompted multiple researchers to investigate how developing
+this software differs from other classes of software.  Previous studies have
+focused on surveying developers \cite{Carveretc}, developer interviews \cite{}
+and case studies \cite{Segal}.  A valuable source of information that has
+received less attention is the data in the publicly available software
+repositories.  We propose a new methodology for digging deeper into how well
+projects are applying best practices by using a combination of manual and
+automated techniques to extract repository-based information.
+
+The surveys used in previous studies have tended to recruit participants from
+all domains of research software, possibly distinguishing them by programming
+language (for instance, R developers \cite{Rstudy}), or by the role of the
+developers (for instance postdoctoral researchers \cite{Katz}).  For their part,
+case studies have focused on a few specific examples at a time, since that is
+the nature of case studies.  We propose applying our new methodology between
+these two extremes.  Rather than focus on all research software, or just a few
+examples, we will focus on one domain of research software at a time. The
+practical reason for defining a domain scope is that digging deep into
+repository data takes time, making a broad scope infeasible.  We have the
+constraint that the work load for applying the methodology to a given domain
 needs to be feasible for a team as small as one individual, and for a time that
@@
 per week) at $8$ person hours per day, or $20 \cdot 8 = 160$ person hours.}
-
-To begin our re-boot of the previous methodology we critically assessed, and
-subsequently modified, our
-\href{https://github.com/adamlazz/DomainX/blob/master/TemplateToGradeSCSoft.pdf}
-{previous set of measures}.  In addition, the following data has been added to
-the new methodology:
-
-\begin{itemize}
-\item Characterization of the functionality provided by the software in the
-  domain via a commonality analysis.
-\item Project repository related data, such as the number of files, number of
-  lines of code, percentage of issues that are closed, etc. 
-\item Interviews with software developers in domain X.
-%\item Usability testing for a sample of software in the domain.  
-\end{itemize}
-
-Unlike the previous measurement process, the new methodology involves and
-engages a domain expert partner throughout.  We did not previously engage the
-domain expert with the rationale that we wished to eliminate potential bias.
-However, this advantage is not worth the inability to evaluate the functionality
-of the software.  Moreover, not having an expert makes navigating the on-line
-resources difficult, since on-line resources are often silent on information
-that is implicit to domain experts.  Furthermore, not every statement found
-on-line will necessarily be accurate.  The importance of the domain expert is
-particularly noteworthy when it comes time for publication and dissemination of
-the state of the practice assessment.  Throughout this document the person (or
-persons) that provide domain expertise will be designated as the \textit{Domain
-Expert}.
-
-In the proposed methodology, the collected data is combined to rank the software
-within each domain using the Analytic Hierarchy Process (AHP) \cite{Saaty1980}.
-As in the previous measurement exercise, we use AHP to develop a list of
-software ranked by quality.  However, in the new process we do not stop with
-this list.  The Domain Expert is consulted to verify the ordering, and to
-discuss the decisions that led to the ranking.  The AHP process is used to
-facilitate a conversation with the Domain Expert as a means to deepen our
-understanding of the software in the domain, and the needs of typical
-developers.
-
-So that the collected data for a given domain can benefit the scientific
-community, our recommendation is that all collected data be made public.  For
-instance, the data collection for each domain can be put on a GitHub repository.
-In addition to the project record left on GitHub, the final data can be exported
-to \href{https://data.mendeley.com/}{Mendeley Data}.  As an example, the
-measurements for the \href{https://data.mendeley.com/datasets/6kprpvv7r7/1}
-{state of the practice for GIS software} are available on Mendeley.  Ideally the
-full analysis of the state of the practice for domain X will also be published
-in a suitable journal, allowing for dissemination/feedback and communication.
-
-The scope of this methodology includes observations on product, artifact
-(documentation, test scripts, etc.) and process quality for research software.
-We leave the assessment of the performance of research software, for instance
-using benchmarks, to other projects, such as the work of
-\cite{kaagstrom1998gemm}.  Currently we are also leaving experiments to measure
-usability and modifiability as future work, as discussed in
-Section~\ref{SecFutureWork}.
-
-The full methodology is presented in the sections that follow.
-Section~\ref{SecResearchQuestions} highlights the research questions that are to
-be answered for each measured domain. These questions are answered by the data
-collected using the process outlined in Section~\ref{StepsAQDS}.  The major
-steps in the process are outlined in Sections~\ref{SecIdentifyDomain} --
-\ref{SecAHP}.  Following this, the time required for assessing a single domain
-is estimated in Section~\ref{SecEstTimeRequired}.
-
-\section{Research Questions} \label{SecResearchQuestions}
-
-The following are the research questions that we wish to answer for each of our
-selected domains.  In addition to answering the questions for each domain, we
-also wish to combine the data from multiple domains to answer these questions
-for research software in general.
+Focusing on one domain at a time has more than just practical advantages.  By
+restricting ourselves to a single domain we can bring domain knowledge and
+domain experts into the mix.  The domain customized insight provided by the
+assessment has the potential to help a specific domain as they adopt and develop
+new development practices.  Moreover, providing a means to measure different
+domains facilitates uncovering domain specific practices, which can later be
+compared and contrasted between different domains.
+
+Our methodology is built around 10 research questions for the domain being
+assessed.  The questions are used to structure the presentation of the
+methodology, so for each research question we point to the corresponding section
+of the paper.  Assuming we have identified the domain of interest, the first question is:
 
 \begin{enumerate}
-\item What artifacts are present in current software packages? 
-\item What tools (development, dependencies, project management) are used by
-current software packages?
-\item What principles, processes, and methodologies are used in the development
-  of current software packages?
-\item What are the pain points for developers working on research software
-  projects?  What aspects of the existing processes, methodologies and tools do
-  they consider as potentially needing improvement?  How should processes,
-  methodologies and tools be changed to improve software development and
-  software quality?
-\item For research software developers, what specific actions are taken to
-  address the following:
+	\item[RQ\refstepcounter{rqnum}\therqnum \label{RQ_WhatProjects}:] What
+	software projects exist in the domain, with the constraint that the source
+	code must be available for all identified projects?
+	(Section~\ref{methodology})
+\end{enumerate}
+
+We next wish to assess the representative software to determine how well they
+apply current software development best practices.  At this point in the
+process, to remove potential user/developer bias, we will base our assessment
+only on publicly available artifacts, where artifacts are the documents, scripts
+and code that we find in a project's public repository. Example artifacts
+include requirements, specifications, user manuals, unit test cases, system
+tests, usability tests, build scripts, API (Application Programming Interface)
+documentation, READMEs, license documents, process documents, and code.
+Following best practices does not guarantee popularity, so we will also compare
+our ranking to how the user community itself ranks the identified projects.
+
 \begin{enumerate}
-\item usability
-\item traceability
-\item modifiability
-\item maintainability
-\item correctness
-\item understandability
-\item unambiguity
-\item reproducibility
-\item visibility/transparency
+	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_HighestQuality}:] Which
+	of the projects identified in \rqref{RQ_WhatProjects} follow current best
+	practices, based on evidence found by experimenting with the software and
+	searching the artifacts available in each project's repository? (Section~\ref{AHPresults})
+	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareHQ2Popular}:] How
+	similar is the list of top projects identified in \rqref{RQ_HighestQuality}
+	to the most popular projects, as viewed by the scientific community? (Section~\ref{repmetrics})
 \end{enumerate}
-\item How does software designated as high quality by this methodology compare
-  with top rated software by the community?
+
+To understand the state of the practice we wish to learn the frequency with
+which different artifacts appear, the types of development tools used and the
+methodologies used for software development.  With this data, we can ask
+questions about how the domain software compares to other research software.
+
+\begin{enumerate}
+  \item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareArtifacts}:] How
+	do domain projects compare to research software in general with respect to the
+	artifacts present in their repositories? (Section~\ref{Sec_CompareArtifacts})
+	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareToolsProjMngmnt}:]
+	How do domain projects compare to research software in general with respect to
+	the use of tools (Section~\ref{Sec_CompareTools}) for:
+	\begin{enumerate} 
+		\item [\rqref{RQ_CompareToolsProjMngmnt}.a] development; and,
+		\item [\rqref{RQ_CompareToolsProjMngmnt}.b] project management?
 	\end{enumerate}
-
-\section{Overview of Steps in Assessing Quality of the Domain Software}
-\label{StepsAQDS}
-
-To answer the above research questions (Section~\ref{SecResearchQuestions}), we
-systematically measure the quality of the software through data collection.  An
-overview of the measurement process is given in the following steps, starting
-from determining a domain that is suitable for measurement:
+	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareMethodologies}:]
+	How do domain projects compare to research software in general with respect to
+	principles, processes and methodologies used? (Section~\ref{Sec_CompareMethodologies})
+\end{enumerate}
+
+Only so much information can be gleaned by digging into software repositories.
+To gain additional insight, we need to interview developers.  We need to learn
+their concerns and how they deal with these concerns; we need to learn their
+pain points.  We wish to know what practices are used by the top domain
+projects, so that others can potentially emulate these practices. We also wish
+to identify new practices by borrowing successful ideas from other domains. The
+above points are covered by the questions outlined below:
 
 \begin{enumerate}
-\item Identify the domain (X). (Section~\ref{SecIdentifyDomain})
-\item Ask the Domain Expert to create a top ten list of software packages in the
-  domain. (Section~\ref{SecDomainExpert})
-\item Meet with the Domain Expert to brief them on the overall objective,
-  research proposal, research questions, measurement template, and developer
-  interviews. (Section~\ref{SecDomainExpert}) \label{StepMeeting}
-\item Identify the broad list of candidate software packages in the domain.
-  (Section~\ref{SecIdentifyCandSoft})
-\item Preliminary filter of software packages list.
-  (Section~\ref{SecInitialFilter})
-\item Review software list with Domain Expert. (Section~\ref{SecDomainExpert})
-\item Domain Analysis (with the help of the Domain Expert).
-(Section~\ref{SecDomainAnalysis})
-\item Ask Domain Expert to vet domain analysis.
-  (Section~\ref{SecDomainExpert})
-\item Gather source code and documentation for each prospective software
-  package.
-\item Collect repository based information. (Section~\ref{SecEmpiricalMeasures})
-\item Measure using measurement template. (Section~\ref{SecMeasureTemplate})
-\item Survey developers. (Section~\ref{SecSurvey})
-\item Use AHP process to rank the software packages. (Section~\ref{SecAHP})
-% \item Identify a short list of top software packages, typically four to six, for
-%   deeper exploration according to the AHP rankings of the measurements.
-\item Ask Domain Expert to vet AHP ranking. (Section~\ref{SecDomainExpert})
-%\item Rank short list. (Section~\ref{SecRankShortList})
-\item Answer research questions (from Section~\ref{SecResearchQuestions}) and
-document answers.
+	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_PainPoints}:] What are
+	the pain points for developers working on domain software projects?
+	(Section~\ref{painpoints})
+	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_ComparePainPoints}:] How
+	do the pain points of domain developers compare to the pain points
+	for research software in general? (Section~\ref{painpoints})
+	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Concerns}:] For domain
+	developers what specific best practices are taken to address the pain points
+	and software quality concerns? (Section~\ref{Sec_AddressConcerns})
+	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Recommend}:]
+	What research software development practice could potentially address the
+	pain points identified in \rqref{RQ_PainPoints}).
+	(Section~\ref{Sec_Recommendations})
 \end{enumerate}
 
-%\wss{The domain expert is involved in multiple steps in the process.  How best
-%  to get their feedback?  The domain experts are busy and are unlikely to devote
-%  significant time to the project.  We need to quickly get to the point.  Maybe
-%  something around task based inspection?  Directed interview?}
-
-\section{How to Identify the Domain?} \label{SecIdentifyDomain}
-
-A domain of research software must be identified to begin the assessment.
-Research software is defined in this exercise as ``software that is used to
-generate, process or analyze results that [are intended] to appear in a
-publication'' \cite{hettrick2014uk} in a scientific or engineering context.
-Research software is a more general term for what is often called scientific
-computing.  To be applicable for the methodology described in this document, the
-chosen domain must have the following properties:
+Our methodology answers the research question through inspecting repositories,
+using the Analytic Hierarch Process (AHP) for ranking software, interviewing
+developers and interacting with at least one domain expert.  We leave the
+measurement of the performance, for instance using benchmarks, to other projects
+\cite{kaagstrom1998gemm}. The current methodology updates the approach used in
+prior assessments of domains like Geographic Information Systems
+\cite{SmithEtAl2018_arXivGIS}, Mesh Generators \cite{SmithEtAl2016}, Seismology
+software \cite{SmithEtAl2018}, and statistical software for psychology
+\cite{SmithEtAl2018_StatSoft}.  Initial tests of the new methodology have been
+done for medical image analysis software \cite{Dong2021} and for Lattice
+Boltzmann Method (LBM) software \cite{Michalski2021}.  The LBM example will be
+used throughout this paper to illustrate the steps in the methodology.  The
+paper ends with a summary of potential threats to validity
+(Section~\ref{threats}) and our final conclusions and recommendations
+(Section~\ref{conclusion}).
+
+\section{Methodology} \label{StepsAQDS}
+
+The assessment is conducted via the following steps.  The steps depend on
+interaction with a Domain Expert partner, as discussed in
+Section~\ref{sec_vet_software_list}.
+
+\begin{enumerate}
+  \item Identify the domain of interest. (Section~\ref{SecIdentifyDomain})
+	\item List candidate software packages for the domain.
+	(Section~\ref{identifysoftware})
+	\item Filter the software package list. (Section~\ref{filtersoftware})
+	\item Gather the source code and documentation for each software package.
+	\item Collect quantitative measures from the project repositories.
+	(Section~\ref{empiricalmeasures})
+	\item Measure using the measurement template.  The full measurement template
+	can be found in \cite{SmithEtAl2021}. (Section~\ref{empiricalmeasures})
+	\item Interview the developers. (Section~\ref{SecSurvey})
+	\item Use AHP to rank the software packages. (Section~\ref{AHP})
+	\item Analyze the results and answer the research questions. (Section~\ref{})
+\end{enumerate}
+
+Table~\ref{TabPersonHours} estimates the time required (in person hours) to
+complete a state of the practice assessment for a single domain.  The table
+assumes that the domain has already been decided and the Domain Expert has been
+recruited.  The time spent by the Domain Expert is not included in the numbers
+shown in the table, since the amount of time that the domain expert will work
+independently of the rest of the assessment team will be small.  Moreover, this
+amount of time will vary greatly depending on the preferred work habits of the
+Domain Expert.  The table follows the steps outlined above. It is assumed that
+the template spreadsheets linked in this document, and the developed AHP tool,
+will be employed, rather than developing new tools.  The person hours given are
+a rough estimate, based on our experience completing assessments for medical
+image analysis software \cite{Dong2021} and for Lattice Boltzmann Method (LBM)
+software \cite{Michalski2021}.  The estimate assumes 30 software packages will
+be measured; the numbers will need to be adjusted if the total packages changes.
+The total number of person hours is \total{totHours} hours.  This is close to
+our goal of 1 month of person hours (160 hours).
+
+\begin{table}[h]
+  \caption{Estimated Person Hours for Assessing the State of Practice for Domain
+  X} \label{TabPersonHours}
+  \centering
+  \begin{tabular}{p{12.5cm} l}
+    \toprule
+    \textbf{Task} & \textbf{Hours} \\
+    \midrule
+
+    Initial 1 hour meeting with the Domain Expert plus meeting prep & 5
+    \addtocounter{totHours}{5} \\
+
+    Identify broad list of candidate software
+    (Section~\ref{SecIdentifyCandSoft}) & 12 \addtocounter{totHours}{12} \\
+
+    Filter software list (Section~\ref{SecInitialFilter}) (10 minutes per
+    package) & 5 \addtocounter{totHours}{5} \\
+
+	Review software list with Domain Expert (Section~\ref{SecDomainExpert}) & 2
+	\addtocounter{totHours}{2} \\
+
+    Domain analysis (with help of Domain Expert)
+    (Section~\ref{SecDomainAnalysis}) & 20 \addtocounter{totHours}{20} \\
+
+	Vet domain analysis with Domain Expert (Section~\ref{SecDomainExpert}) & 3
+	\addtocounter{totHours}{3} \\
+
+	Gather source code and documentation for each package (10 minutes per
+	package)) & 5 \addtocounter{totHours}{5} \\
+
+    Collect repository based data (Section~\ref{SecEmpiricalMeasures}) (10
+    minutes per package) & 5 \addtocounter{totHours}{5} \\
+
+	Measure using measurement template (Section~\ref{SecMeasureTemplate}) (2.5
+    hours per repo) & 75 \addtocounter{totHours}{75} \\
+
+    Solicit developers for interviews & 2 \addtocounter{totHours}{2} \\
+
+    Conduct interviews (1.5 hour interviews with 10 developers (30\% recruitment
+    rate)) & 15 \addtocounter{totHours}{15} \\
+
+    AHP ranking & 2 \addtocounter{totHours}{2} \\
+
+	Work with Domain Expert to vet AHP ranking & 2 \addtocounter{totHours}{2} \\
+
+    Analyze data and answer research questions & 20 \addtocounter{totHours}{20}\\
+
+    \midrule
+    \textbf{Total} & \textbf{\thetotHours} \\
+    \bottomrule
+  \end{tabular}
+
+\end{table}
+
+\subsection{How to Identify the Domain?} \label{SecIdentifyDomain}
+
+A domain of research software must be identified to begin the assessment. To be
+applicable for the methodology described in this document, the chosen domain
+must have the following properties:
 
@@
 
-Some examples of domains that fit these criteria are finite element analysis
-\cite{szabo1996finite}, quantum chemistry \cite{veryazov20042molcas},
-seismology \cite{SmithEtAl2018}, as well as mesh generators
-\cite{smith2016state}.
-
-\section{Interaction With Domain Expert} \label{SecDomainExpert}
-
-As mentioned in the introduction (Section~\ref{SecIntroduction}), the Domain
-Expert is an important member of the state of the practice assessment team.
-Pitfalls exist if non-domain experts attempt to acquire an authoritative list of
-software, or perform a commonality analysis, or try to definitively rank the
-software. The main source of problems for non-domain experts is that they can
-only rely on information that is available on-line, but on-line data has two
-potential problems:
-\begin{inparaenum}[i)]
-  \item the on-line resources could have false or inaccurate information; and,
-  \item the on-line resources could leave out relevant information that is so
-in-grained with experts that nobody thinks to explicitly record the information.
-\end{inparaenum}
-
-Domain experts may be recruited from academia or industry.  The only
-requirements are knowledge of the domain and a willingness to be engaged in the
-assessment process.  The Domain Expert does not have to be a software developer,
-but they should be a user of domain software.  Given that the domain experts are
-likely to be busy people, the measurement process cannot put to much of a burden
-on their time.
-
-In advance of the first meeting with the Domain Expert (Step~\ref{StepsAQDS} in
-Section~\ref{StepsAQDS}) the expert is asked to create a top ten list of
-software packages in the domain.  This is done to help the expert get in the
-right mind set in advance of the first meeting.  Moreover, by doing the exercise
-in advance, we avoid the potential pitfall of the expert approving the
-discovered list of software without giving it adequate thought.  The emphasis
-during the first meeting is for the Domain Expert to learn what is expected of
-them.  The discussion should also cover avenues for publication and
-dissemination.
-
-The Domain Experts are asked to vet the collected data and analysis.  In
-particular, they are asked to vet the proposed list of software packages, the
-domain analysis and the AHP ranking.  These interactions can be done either
-electronically or with in-person (or virtual) meetings.
-
-\section{How to Identify Candidate Software?} \label{SecIdentifyCandSoft}
-
-Once the domain of interest is identified, the candidate software for measuring
-can be found through search engine queries targeting authoritative lists of
-software.  Potential places to search include \href{https://github.com/}
-{GitHub}, \href{https://swmath.org/} {swMATH} and domain related publications,
-such as review articles. Domain Experts are also asked for their suggestions and
-are asked to review the initial draft of the software list.
-
-When forming the list and reviewing the candidate software the following
-properties should be considered:
+\subsection{How to Identify Candidate Software from the Domain?}
+\label{identifysoftware}
+
+To answer~\rqref{RQ_WhatProjects} we needed to identify existing projects. The
+candidate software should be found through search engine queries targeting
+authoritative lists of software.  Candidate places to search include
+\href{https://github.com/} {GitHub} and \href{https://swmath.org/} {swMATH}, as
+well as through scholarly articles. The Domain Expert
+(Section~\ref{sec_vet_software_list}) should also be engaged in selecting the
+candidate software.
+
+The following properties were considered when creating the list and reviewing
+the candidate software:
 
@@
 	\item The source code must be viewable.
-\item The empirical measures listed in Section \ref{SecEmpiricalMeasures} should
-  ideally be available, which implies a preference for GitHub-style
-  repositories.
-\item The software cannot be marked as incomplete, or in an initial development
-  phase.
+	\item The repository based measures should be available, which implies a
+	preference for GitHub-style repositories.
+	\item The software cannot be marked as incomplete or in an initial state of
+	development.
 \end{enumerate}
 
-\section{How to Initially Filter the Software List?} \label{SecInitialFilter}
+For the LBM example the initial list had 46 packages \cite{Michalski2021}.
+
+\subsection{How to Filter the Software List?} \label{filtersoftware}
 
 If the list of software is too long (over around 30 packages), then steps need
-to be taken to create a more manageable list. To reduce the length of the list,
-the following filters are applied.  The filters are applied in the priority
-order listed, with the filtering process stopped once the list size is
-manageable.
+to be taken to create a more manageable list to answer~\rqref{RQ_WhatProjects}.
+The following filters were applied in the priority order listed. Copies of both
+the initial and filtered lists, along with the rationale for shortening the
+list, should be kept for traceability purposes.
 
 \begin{enumerate}
-\item Scope: Software is removed by narrowing what functionality is considered
-  to be within the scope of the domain.
-\item Usage: Software packages are eliminated if their installation procedure is
-  not clear and easy to follow.
-\item Age: The older software packages (age being measured by the last date when
-  a change was made) are eliminated, except in the cases where an older software
-  package appears to be highly recommended and currently in use.  (The Domain
-  Expert can be consulted on this question, if necessary.)
+	\item Scope: Software is removed by narrowing what functionality is
+	considered to be within the scope of the domain.
+	\item Usage: Software packages were eliminated if their installation
+	procedure was missing or not clear and easy to follow.
+	\item Age: The older software packages (age being measured by the last date
+	when a change was made) were eliminated, except in the cases where an older
+	software package appears to be highly recommended and currently in use. (The
+	Domain Expert should be consulted on this question as necessary.)
 \end{enumerate}
 
-Copies of both the initial and filtered lists, along with the rationale for
-shortening the list, should be kept for traceability purposes.
-
-\section{Domain Analysis} \label{SecDomainAnalysis}
-
-Since each domain we will study will have a reasonably small scope, we will be
-able to view the software as constituting a program family.  The concept of a
-program family is defined by \cite{parnas1976design} as ``a set of programs
-whose common properties are so extensive that it is advantageous to study the
-common properties of the programs before analyzing individual members''.
-Studying the common properties within a family of related programs is termed a
-domain analysis.
-
-The domain analysis consists of a commonality analysis of the family of software
-packages. Its purpose is to show the relationships between these packages, and
-to facilitate an understanding of the informal specification and development of
-them. \cite{weiss1997defining} defines commonality analysis as an approach to
-defining a family by identifying commonalities, variabilities, and common
-terminology for the family. Commonalities are goals, theories, models,
-definitions and assumptions that are common between family members.
-Variabilities are goals, theories, models, definitions and assumptions that
-differ between family members. Associated with each variability are its
-parameters of variation, which summarize the possible values for that
-variability, along with their potential binding time.  The binding time is when
-the value of the variability is set.  It could be set as specification time,
-build time (when the program is being compiled) or run time (when the code is
-executing).
-
-The final result of the domain analysis will be tables of commonalities,
-variabilities, and parameters of variation of a program family.
-\cite{smith2008commonality} present a template for conducting a commonality
-analysis, which was referred to when conducting this work.
-\cite{weiss1998commonality} describes another commonality analysis technique for
-deciding the members of a program family. \cite{SmithAndChen2004} and
-\cite{SmithMcCutchanAndCarette2017} are examples of a commonality analysis for a
-family of mesh generating software and a family of material models,
-respectively. The steps to produce a commonality analysis are:
-
-\begin{enumerate}
-\item Write an Introduction
-\item Write an Overview of the Domain Knowledge
-\item List Commonalities
-\item List Variabilities
-\item List Parameters of Variation
-\item Add Terminology, Definitions, Acronyms
-\end{enumerate}
-
-% A sample commonality analysis for Lattice Boltzmann Solvers can be found
-% \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Peter-Notes/Commonality-Analysis-LB-Systems.pdf}
-% {here}.
-
-\section{Repository Based Measures} \label{SecEmpiricalMeasures}
-
-Some quality measurements rely on gathering raw and processed data from software
-repositories. We focus on data that is reasonably easy to collect, which we
-combine and analyze. The measures that are collected relate to the research
-questions (Section~\ref{SecResearchQuestions}). For instance, we collect data to
-see how large a project is, to ascertain a project’s popularity, and to
-determine whether to project is being actively developed.
-
-Section~\ref{rawdata} lists the raw data that is collected.  Some of this data
-can be observed from GitHub repository metrics. The rest can be collected using
-freeware tools. \href{https://github.com/tomgi/git_stats}{GitStats} is used to
-measure the number of binary files as well as the number of added and deleted
-lines in a repository. This tool is also used to measure the number of commits
-over different intervals of time. \href{https://github.com/boyter/scc}{Sloc Cloc
-and Code (scc)} is used to measure the number of text based files as well as the
-number of total, code, comment, and blank lines in a repository. These tools
-were selected due to their installability, usability, and ability to gather the
-empirical measures listed below. Details on installing and running the tools can
-be found in Appendix~\ref{SecRepoTools}.  Section \ref{processeddata} introduces
-the required processed data, which is calculated using the raw data.
-
-\subsection{Raw Data}\label{rawdata}
-
-The following raw data measures are extracted from repositories:
-
-\begin{itemize}
-\item Number of stars.
-\item Number of forks.
-\item Number of people watching the repository.
-\item Number of open pull requests.
-\item Number of closed pull requests.	
-\item Number of developers.	
-\item Number of open issues.
-\item Number of closed issues.
-\item Initial release date.
-\item Last commit date.
-\item Programming languages used.
-\item Number of text-based files.
-\item Number of total lines in text-based files.
-\item Number of code lines in text-based files.
-\item Number of comment lines in text-based files.
-\item Number of blank lines in text-based files.
-\item Number of binary files.  
-\item Number of total lines added to text-based files.
-\item Number of total lines deleted from text-based files.
-\item Number of total commits.
-\item Numbers of commits by year in the last 5 years. (Count from as early as
-  possible if the project is younger than 5 years.)
-\item Numbers of commits by month in the last 12 months.
-\end{itemize}
-
-\subsection{Processed Data}\label{processeddata}
-
-The following measures are calculated from the raw data:
-
-\begin{itemize}
-\item Status of software package as either dead or alive, where alive is defined
-  as the presence of repository commits or software package version releases in
-  the last 18 months.
-\item Percentage of identified issues that are closed.
-\item Percentage of code that is comments.
-\end{itemize}
-
-\noindent The time frame of 18 months was selected as the separating point
-between alive and dead projects because this is the usual timeframe for
-operating system updates.
-
-\section{Measure Using Measurement Template} \label{SecMeasureTemplate}
-
-The Measurement Template is found in Appendix~\ref{SecGradingTemplate}.  This
-template is used to track measurements and quality scores for all of the
-software packages in the domain. For each software package, we fill-in the
-template questions. This process can take between 1 to 4 hours for each package.
-Project developers can be contacted for help regarding installation, if
-necessary, but a cap of about 2 hours should be imposed on the entire
-installation process, to keep the overall measurement time feasible.  To save
-time, a
-\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Combined_MeasurementTemplate_EmpiricalMeasures.xlsx}{blank
-measurement template spreadsheet} has been prepared, with the measures as rows.
-An excerpt of the spreadsheet is shown in
-Figure~\ref{measurement_template_image}.  A column should be added to this
-template for each software package to be measured.
+For the third item in the above filter, software packages are characterized as
+`alive' if their related documentation had been updated within the last 18
+month. Otherwise a package is categorized as `dead'.
+
+For the running example of LBM filtering by scope, usage, and age decreased the
+initial list of 46 to a list of 24. Many of the removed packages could not be
+tested as there was no installation guide, they were incomplete, source code was
+not publicly available, a license was needed, or the project was out of scope or
+not up to a standard that would support incorporating them into this
+study~\cite{Michalski2021}. Of the remaining packages, some were kept on the
+list despite being marked as dead due to their prevalence on authoritative lists
+on LBM software and due to their surface excellence.
+
+\subsection{Quantitative Measures} \label{empiricalmeasures}
+
+We rank the projects by how well they follow best practices
+(\rqref{RQ_HighestQuality}) via a measurement template, as described in
+\cite{SmithEtAl2021}.  For each software package (each column in the template),
+we fill-in the rows of the template. This process takes about 2 hours per
+package, with a cap of 4 hours. The time constraint is necessary so that the
+work load is feasible for a team as small as one, given our aim to cap the
+measurement phase at 160 person hours \cite{SmithEtAl2021}.  An excerpt of the
+template, in spreadsheet form, is shown in
+Figure~\ref{measurement_template_image}.
 
@@
 	\begin{center}
-	  %\includegraphics[width=1.0\textwidth]{measurement_template}
-	  \caption{Excerpt of the Top Section of the Measurement Template (Summary Information)}
+	  %\includegraphics[width=1.0\textwidth]{./figures/measurement_template.pdf}
+	  \caption{Excerpt of the Top Sections of the Measurement Template}
 	  \label{measurement_template_image}
@@
 
-The full template consists of 108 questions categorized under 9 qualities.  The
-questions were designed to be unambiguous, quantifiable and measurable with
-limited time and domain knowledge. The measures are grouped under headings for
-each quality, and one for summary information. The summary information (shown in
-Figure~\ref{measurement_template_image}) is the first section of the template.
-This section summarizes general information, such as the software name, number
-of developers, etc.  We follow the definitions given by
-\cite{GewaltigAndCannon2012} for the software categories.  Public means software
-intended for public use.  Private means software aimed only at a specific
-group, while the concept category is used for software written simply to demonstrate
-algorithms or concepts. The three categories of development models are: open
-source, where source code is freely available under an open source license;
-free-ware, where a binary or executable is provided for free; and, commercial,
-where the user must pay for the software product.
-
-Following the summary section are sections to measure 9 qualities: 
-\begin{inparaenum}
+The full template consists of 108 questions categorized under 9 qualities:
+\begin{inparaenum}[(i)]
 	\item installability;
@@
 \end{inparaenum}
-Definitions of these qualities are available in a
-\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/QDefOfQualities/QDefOfQualities.pdf}{working
-document on software quality}.  Several of the qualities use the word
-``surface''.  This is to highlight that, for these qualities in particular, the
-best that we can do is a shallow measure of the quality.  For instance, we are
-not currently doing any experiments to measure usability.  Instead, we are
-looking for an indication that usability was considered by the developers.  We
-do this by looking for cues in the documentation, like a getting started manual,
-a user manual and documentation of expected user characteristics.
+
+The questions were designed to be unambiguous, quantifiable and measurable with
+limited time and domain knowledge. The measures are grouped under headings for
+each quality, and one for summary information
+(Figure~\ref{measurement_template_image}).   The summary section provides
+general information, such as the software name, number of developers, etc.  We
+follow the definitions given by \cite{gewaltig2012quality} for the software
+categories.  Public means software intended for public use.  Private means
+software aimed only at a specific group, while the concept category is used for
+software written simply to demonstrate algorithms or concepts. The three
+categories of development models are: open source, where source code is freely
+available under an open source license; free-ware, where a binary or executable
+is provided for free; and, commercial, where the user must pay for the software
+product.
+
+Several of the qualities use the word ``surface''.  This is to highlight that,
+for these qualities in particular, the best that we can do is a shallow measure.
+For instance, we do not conduct experiments to measure usability. Instead, we
+are looking for an indication that usability was considered by the developers by
+looking for cues in the documentation, such as getting started instructions, a
+user manual or documentation of expected user characteristics.
 
@@
 \begin{enumerate}
-\item Initial release date: Mark the release year if an exact date is not
-  available.
 \item Publications about the software: A list of publications can be found
@@
   Scholar or a similar index.
-\item Is there evidence that performance was considered?: Search the software
-  artifacts for any mention of speed, storage, throughput, performance
-  optimization, parallelism, multi-core processing, or similar
+\item To answer whether there evidence that performance was considered, search
+  the software artifacts for any mention of speed, storage, throughput,
+  performance optimization, parallelism, multi-core processing, or similar
   considerations. The search function on GitHub can help.
-\item Getting started tutorial: Sometimes this is found within another artifact,
-  like the user manual.
-\item Continuous integration: Search the software artifacts for any mention of
-  continuous integration. The search function on GitHub can help.  In some
-  cases, \texttt{yaml} files will provide a hint that continuous integration is
-  employed.
+\item Evidence of a getting started tutorial may be found within another
+  artifact, like the user manual.
+\item To find evidence of continuous integration, search the software artifacts
+  for any mention of continuous integration. The search function on GitHub can
+  help.  In some cases, \texttt{yaml} files will provide a hint that continuous
+  integration is employed.
 \end{enumerate}
 
-\noindent To fill-in the spreadsheet template, the following steps should be followed:
-
-\begin{enumerate}
-\item Gather the summary information into the top section of the document
-  (Figure \ref{measurement_template_image}).
-\item Using the GitStats tool that is described in
-  Section~\ref{SecEmpiricalMeasures} gather the measurements for the Repo
-  Metrics - GitStats section found near the bottom of the spreadsheet.
-\item Using the SCC tool that is also described in
-  Section~\ref{SecEmpiricalMeasures} gather the measurements for the Repo
-  Metrics -- SCC section found near the bottom of the spreadsheet.
-\item If the software package is found on git, gather the measurements for the
-  Repo Metrics - the GitHub section found near the bottom of the spreadsheet.
-\item Review installation documentation and attempt to install the software
-  package on a virtual machine.
-\item Gather the measurements for installability
-\item Gather the measurements for correctness and verifiability
-\item Gather the measurements for surface reliability
-\item Gather the measurements for surface robustness
-\item Gather the measurements for surface usability
-\item Gather the measurements for maintainability
-\item Gather the measurements for reusability
-\item Gather the measurements for surface understandability
-\item Gather the measurements for visibility and transparency
-\item Assign a score out of ten for each quality. The score can be measured
-  using the Measurement Template Impression Calculator, found in
-  Appendix~\ref{SecImpressionCalculator}. For each quality measurement, the file
-  indicates the appropriate score to assign the measurement based on possible
-  measurement values.
-\end{enumerate}
+Tools were use to find some of the measurements, such as the number of files,
+number of lines of code (LOC), percentage of issues that are closed, etc. The
+tool \href{https://github.com/tomgi/git_stats}{GitStats} was used to measure
+each software package's GitHub repository for the number of binary files, the
+number of added and deleted lines, and the number of commits over varying time
+intervals. The tool \href{https://github.com/boyter/scc}{Sloc Cloc and Code
+(scc)} was used to measure the number of text based files as well as the number
+of total, code, comment, and blank lines in each GitHub repository.
 
@@
 advantage of using VMs is to level the playing field. Every software install
-starts from a clean slate, which removes ``works-on-my-computer'' errors. When
-filling in the measurement template spreadsheet, the the details for each VM
-should be noted, including hypervisor and operating system version.
-
-\section{Survey Developers} \label{SecSurvey}
-
-In the previous state of the practice measurement process
-\cite{SmithEtAl2018_arXivGIS, SmithEtAl2016, SmithEtAl2018}, we only based our
-assessment on information available in on-line software repos.  However, this
-approach meant we weren't able to learn about the development process, the
-attitudes of the developers, the pain points for developers and how the
-developers handle modifiability, reproducibility and usability.  Therefore, in
-the reboot of the measurement process, we have explicitly added a stage for
-interviewing research software developers.
-
-We designed a list of 20 questions to guide our interviews, which can be found
-in Appendix~\ref{SecSurveyQuestions}. Some questions are about the background of
-the software, the development teams, the interviewees, and how they organize the
-projects. We also ask about the developer's understandings of the users. Some
-questions focus on the current and past difficulties, and the solutions the team
-has found, or will try. We also discuss the importance and current situations of
-documentation. A few questions are about specific software qualities, such as
-maintainability, understandability, usability, and reproducibility. The
-interviews are semi-structured based on the question list; we ask follow-up
-questions when necessary. Based on our experience, the interviewees usually
-bring up some exciting ideas that we did not expect, and it is worth expanding
-on these topics.
+starts from a clean slate, which removes ``works-on-my-computer'' errors.
+
+\subsection{Analytical Hierarchy Process} \label{AHP}
+
+Once we have measured each package, we still need to rank them to answer
+\rqref{RQ_HighestQuality}.  To do this, we used the Analytical Hierarchy Process
+(AHP), a decision-making technique that is used to compare multiple options by
+multiple criteria \cite{Saaty}. AHP performs a pairwise analysis using a matrix
+and generates an overall score as well as individual quality scores for each
+software package. The advantage of pair-wise comparisons is that they
+facilitates a separation of concerns.  Rather than worry about the entire
+problem, the decision maker can focus on one comparison at a time.  In our work
+AHP is used for comparing and ranking the software packages of a domain using
+the quality scores that are gathered in the Measurement Template
+(Appendix~\ref{SecGradingTemplate}). AHP performs a pairwise analysis between
+each of the 9 quality options for each of the 30 software packages.  This
+results in a matrix, which is used to generate an overall score for each
+software package for the given criteria. \cite{SmithEtAl2016} shows how AHP is
+applied to ranking software based on quality measures.
+
+Example of LBM measurement - maybe show the summary of all qualities?
+
+\subsection{Interview Developers} \label{SecSurvey}
+
+Several of the research question (\rqref{RQ_CompareToolsProjMngmnt},
+\rqref{RQ_CompareMethodologies}, \rqref{RQ_PainPoints}, \rqref{RQ_Concerns} and
+\rqref{RQ_Recommend}) require going beyond the quantitative data from the
+measurement template. To gain the required insight, we interview developers
+using a list of 20 questions \cite{SmithEtAl2021}. The questions cover the
+background of the development teams, the interviewees, and the software itself.
+We ask the developers how they organize their projects. We also ask them about
+their understanding of the users. Some questions focus on the current and past
+difficulties, and the solutions the team has found, or will try. We also discuss
+the importance of, and the current situation for, documentation. A few questions
+are about specific software qualities, such as maintainability,
+understandability, usability, and reproducibility. The interviews are
+semi-structured based on the question list; we ask follow-up questions when
+necessary.  Each interview takes an 1 hour on average.
 
@@
 package.  Meeting will typically be held using on-line meeting software, like
-Zoom or Teams.  This facilitates recording and automatic transcription of the
+Zoom or Teams, facilitates recording and automatic transcription of the
 meetings.
@@
 
-\section{Analytic Hierarchy Process} \label{SecAHP}
-
-The Analytical Hierarchy Process (AHP) is a decision-making technique that can
-be used when comparing multiple options by multiple criteria. AHP focuses on
-pair-wise comparisons between all options for all criteria.  The advantage of
-pair-wise comparisons is that they facilitates a separation of concerns.  Rather
-than worry about the entire problem, the decision maker can focus on one
-comparison at a time.  In our work AHP is used for comparing and ranking the
-software packages of a domain using the quality scores that are gathered in the
-Measurement Template (Appendix~\ref{SecGradingTemplate}). AHP performs a
-pairwise analysis between each of the 9 quality options for each of the 30
-software packages.  This results in a matrix, which is used to generate an
-overall score for each software package for the given criteria.
-\cite{SmithEtAl2016} shows how AHP is applied to ranking software based on
-quality measures. We have developed a
-\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/AHP2020/LBM/README.txt}{tool}
-for conducting this process. The tool includes an AHP JAR script and a
-sensitivity analysis JAR script that is used to ensure that the software package
-rankings are appropriate with respect to the uncertainty of the quality scores.
-The
-\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/AHP2020/LBM/README.txt}{README
-file} outlines the requirements for, and configuration and usage of, the JAR
-scripts. The JAR scripts, source code, and required libraries are located in the
-same folder as the README file.
-
-% \section{Rank Short List} \label{SecRankShortList}
-
-% An AHP pairwise comparison between the short list packages with respect to
-% usability and modifiability experiment results is conducted, and then reviewed
-% by the domain expert. The tool can be found here. The usability and
-% modifiability survey can be found
-% \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Experiments.pdf}
-% {here}. This step is intended to start a conversation about the relative quality
-% of short-listed software packages.
-
-\section{Estimate of Time Required} \label{SecEstTimeRequired}
-
-Table~\ref{TabPersonHours} estimates the time required (in person hours) to
-complete a state of the practice assessment for domain X.  The table assumes
-that the domain has already been decided and the Domain Expert has been
-recruited.  The time spent by the Domain Expert is not included in the numbers
-shown in the table, since the amount of time that the domain expert will work
-independently of the rest of the assessment team will be small.  Moreover, this
-amount of time will vary greatly depending on the preferred work habits of the
-Domain Expert.  The table follows the steps outlined in Section~\ref{StepsAQDS}.
-Time is not included for reviewing the methodology. Moreover, it is assumed that
-the template spreadsheets linked in this document, and the developed AHP tool,
-will be employed, rather than developing new tools.  The person hours given are
-a rough estimate, based on our experience completing assessments for medical
-image analysis software \cite{Dong2021} and for Lattice Boltzmann Method (LBM)
-software \cite{Michalski2021}.  These two domains were assessed at the same
-time as designing the methodology presented in this document.  We did our best
-to estimate the time spent on measurement and separate it from the time spend on
-design and development.  The estimate assumes 30 software packages will be
-measured; the numbers will need to be adjusted if the total packages changes.
-
-\begin{table}[h]
-  \caption{Estimated Person Hours for Assessing the State of Practice for Domain
-  X} \label{TabPersonHours}
-  \centering
-  \begin{tabular}{p{10cm} l}
+For LBM we were able to recruit 4 developers to participate in our study.
+Results in next sections.
+
+\subsection{Interaction With Domain Expert} \label{sec_vet_software_list}
+
+We partnered with a Domain Expert to vet our list of projects
+(\rqref{RQ_WhatProjects}) and our ranking (\rqref{RQ_HighestQuality}).  The
+Domain Expert is an important member of the state of the practice assessment
+team. Pitfalls exist if non-experts attempt to acquire an authoritative list of
+software, or try to definitively rank the software. Non-experts have the problem
+that they can only rely on information available on-line, which has the
+following drawbacks:
+\begin{inparaenum}[i)]
+  \item the on-line resources could have false or inaccurate information; and,
+  \item the on-line resources could leave out relevant information that is so
+in-grained with experts that nobody thinks to explicitly record it.
+\end{inparaenum}
+
+Domain experts may be recruited from academia or industry.  The only
+requirements are knowledge of the domain and a willingness to be engaged in the
+assessment process.  The Domain Expert does not have to be a software developer,
+but they should be a user of domain software.  Given that the domain experts are
+likely to be busy people, the measurement process cannot put to much of a burden
+on their time.
+
+The Domain Expert has an important role with verifying the list of LBM packages.
+In advance of the first meeting with the Domain Expert, they were asked to
+create a list of top software packages in the domain.  This is done to help the
+expert get in the right mind set in advance of the meeting.  Moreover, by doing
+the exercise in advance, we avoid the potential pitfall of the expert approving
+the discovered list of software without giving it adequate thought.  The Domain
+Expert was also asked to vet the collected data and analysis.  In particular,
+they were asked to vet the proposed list of software packages and the AHP
+ranking.
+
+For the LBM project, our Domain Expert was Dr.\
+Zahra Motamed, Assistant Professor of Mechanical Engineering at McMaster
+University, Hamilton, Ontario, Canada.
+
+\subsection{Domain Analysis} \label{SecDomainAnalysis}
+
+Since each domain we will study will have a reasonably small scope, we will be
+able to view the software as constituting a program family.  The concept of a
+program family is defined by \cite{parnas1976design} as ``a set of programs
+whose common properties are so extensive that it is advantageous to study the
+common properties of the programs before analyzing individual members''.
+Studying the common properties within a family of related programs is termed a
+domain analysis.
+For the current methodology time constraints necessitate a shallow domain analysis.
+A table should be constructed that distinguishes the programs under study by the variabilities that distinguish them.
+In research software the variabilities are often related to assumptions.
+Table~\ref{} shows the variabilities for the LBM software example.
+
+\section{Comparison to Community Ranking} \label{repmetrics}
+
+To address \rqref{RQ_CompareHQ2Popular} we need to compare the ranking by best
+practices to the communities ranking.  Our best practices ranking comes from the
+results of the AHP ranking (Section~\ref{AHP}).  We estimate the communities
+ranking by repository stars and watches.  The comparison will provide insight on
+whether best practices are rewarded by popularity.  However, inconsistencies
+between the AHP ranking and the communities ranking are in inevitable for the
+following reasons: 
+\begin{inparaenum}[i)]
+	\item the overall quality ranking via AHP makes the unrealistic assumption
+	of equal weighting between the different quality factors;
+	\item stars are known to not be a particularly good measure of popularity,
+	since young projects have less time to accumulate stars \cite{}; 
+	\item and, as for consumer products, there are more factors that determine
+	popularity than quality alone.
+\end{inparaenum}
+
+Table~\ref{repometrics} compares the AHP ranking of the LBM package versus their
+popularity in the research community.  Nine packages do not use GitHub, so they
+do not have a measure of repository stars. Looking at the repository stars of
+the other 15 packages, we can observe a pattern where packages that have been
+highly ranked by our assessment tend to have more stars than lower ranked
+packages. The best ranked package by AHP (ESPResSo) has the second most stars,
+while the ninth ranked package (Sailfish) has the highest number of stars. The
+same correlation is observed in the repository watch column, although this
+column contains less data, since two of the packages (Palabos, waLBerla) use
+GitLab, which does not track watches. Packages designated as lower quality often
+do not use GitHub or GitLab, or have only a few stars/watches. Although the AHP
+ranking and the community popularity estimate are not perfect measures, they do
+suggest a correlation between best practices and popularity. We do not know
+which comes first, the use of best practices or popularity, but we do know that
+the top ranked packages tend to incorporate best practices.
+
+\begin{table}[!h]
+	\begin{center}
+		\begin{tabular}{ p{3cm}p{1.25cm}p{1.75cm}p{1.75cm}p{1.75cm}p{1.75cm} }
 			\toprule
-    \textbf{Task} & \textbf{Hours} \\
+			Name & Our Ranking & Repository Stars & Repository Star Rank &
+			Repository Watches & Repository Watch Rank\\
 			\midrule
-
-    Initial 1 hour meeting with the Domain Expert plus meeting prep & 5
-    \addtocounter{totHours}{5} \\
-
-    Identify broad list of candidate software
-    (Section~\ref{SecIdentifyCandSoft}) & 12 \addtocounter{totHours}{12} \\
-
-    Filter software list (Section~\ref{SecInitialFilter}) (10 minutes per
-    package for 30 packages) & 5 \addtocounter{totHours}{5} \\
-
-	Review software list with Domain Expert (Section~\ref{SecDomainExpert}) & 2
-	\addtocounter{totHours}{2} \\
-
-    Domain analysis (with help of Domain Expert)
-    (Section~\ref{SecDomainAnalysis}) & 20 \addtocounter{totHours}{20} \\
-
-	Vet domain analysis with Domain Expert (Section~\ref{SecDomainExpert}) & 3
-	\addtocounter{totHours}{3} \\
-
-	Gather source code and documentation for each package (10 minutes per
-	package for 30 packages)) & 5 \addtocounter{totHours}{5} \\
-
-    Collect repository based data (Section~\ref{SecEmpiricalMeasures}) (10
-    minutes per package for 30 packages) & 5 \addtocounter{totHours}{5} \\
-
-	Measure using measurement template (Section~\ref{SecMeasureTemplate}) (2.5
-    hours per repo for 30 repos) & 75 \addtocounter{totHours}{75} \\
-
-    Solicit developers for interviews & 2 \addtocounter{totHours}{2} \\
-
-    Conduct interviews (1.5 hour interviews with 10 developers (assuming 1 in 3
-    developers agree to an interview)) & 15 \addtocounter{totHours}{15} \\
-
-    AHP ranking & 2 \addtocounter{totHours}{2} \\
-
-	Work with Domain Expert to vet AHP ranking & 2 \addtocounter{totHours}{2} \\
-
-    Analyze data and answer research questions & 20 \addtocounter{totHours}{20}\\
-
-    \midrule
-    \textbf{Total} & \textbf{\thetotHours} \\
+			ESPResSo & 1 & 145 & 2 & 19& 2\\
+			Ludwig & 2 & 27 & 8 & 6& 7\\
+			Palabos & 3 & 34 & 6 & GitLab& GitLab\\
+			OpenLB & 4 & No Git & No Git & No Git& No Git\\
+			LUMA & 5 & 33 & 7 & 12& 4\\
+			pyLBM & 6 & 95 & 3 & 10& 5\\
+			DL\_MESO (LBE) & 7 & No Git & No Git & No Git & No Git\\
+			Musubi & 8 & No Git & No Git & No Git & No Git\\
+			Sailfish & 9 & 186 & 1 & 41& 1\\
+			waLBerla & 10 & 20 & 9 & GitLab& GitLab\\
+			laboetie & 11 & 4 & 13 & 5& 8\\
+			TCLB & 12 & 95 & 3 & 16& 3\\
+			MechSys & 13 & No Git & No Git & No Git& No Git\\
+			lettuce & 14 & 48 & 4 & 5& 8\\
+			ESPResSo++ & 15 & 35 & 5 & 12& 4\\
+			MP-LABS & 16 & 12 & 11 & 2& 9\\			
+			SunlightLB & 17 & No Git & No Git & No Git& No Git\\
+			LB3D & 18 & No Git & No Git & No Git& No Git\\			
+			LIMBES & 19 & No Git & No Git & No Git& No Git\\
+			LB2D-Prime & 20 & No Git & No Git & No Git& No Git\\		
+			HemeLB & 21 & 12 & 11 & 12& 4\\
+			lbmpy & 22 &  11 & 12 & 2& 9\\	
+			LB3D-Prime & 23 & No Git & No Git & No Git& No Git\\	
+			LatBo.jl & 24 & 17 & 10 & 8& 6\\			
 			\bottomrule
 		\end{tabular}
-
+		\caption{Repository Ranking Metrics} \label{repometrics}
+	\end{center}
 	\end{table}
 
-The total number of person hours is \total{totHours} hours.  This is close to
-our goal of 1 month of person hours (160 hours).  The amount of time spent by
-the Domain Expert can be estimated by summing the Domain Expert items in
-Table~\ref{TabPersonHours} and adding an estimate of the time that they will
-independently spend on their assigned tasks.  If we assume that the Domain
-Expert will spend 2 hours on the domain analysis and another 2 hours with
-answering questions, the Domain Expert time will be about 12 person hours.
-
-\section{Future Work} \label{SecFutureWork}
-
-As explained in the introduction (Section~\ref{SecIntroduction}), our eventual
-goal is to improve the state of practice of software development, which requires
-a baseline means for measuring the current state of the practice.  Now that we
-have a methodology for assessing the state of practice for a given domain, the
-next task is to complete the measurements for multiple domains.  Using the
-previous and updated methodologies, we have measured the state of
-practice for the following domains: Geographic Information Systems
-\cite{SmithEtAl2018_arXivGIS}, Mesh Generators \cite{SmithEtAl2016},
-Seismology software \cite{SmithEtAl2018}, Statistical software for psychology
-\cite{SmithEtAl2018_StatSoft}, medical image analysis software \cite{Dong2021}
-and for Lattice Boltzmann Method software \cite{Michalski2021}.  Future domains
-for measurement could include finite element software, computational medicine,
-machine learning, ordinary differential equation solvers, computer graphics,
-stoichiometry, etc.
-
-With the wealth of data from assessing the state of practice for multiple
-domains, the next step is a meta-analysis.  We would look at how the different
-domains compare. What lessons from one domain could be applied in other domains?
-What (if any) differences exist in the pain points between domains?  Are there
-differences in the tools, processes, and documentation between domains?
-
-The current methodology is constrained by limited resources.  A 4 hour cap on
-the measurement time for each software package limits what can be assessed.
-Within this limit, we can't measure some important qualities, like usability and
-modifiability.  In the future, we propose a more time-consuming process that
-would capture these other quality measures.  To improve the feasibility, the
-more time consuming measurements would not have to be completed for all 30
-packages. Instead, a short list could be identified using the output of the AHP
-ranking to select the top projects, or to select a sample of interesting
-projects across the quality spectrum.
-
-If we can add measures for modifiability and usability, we can start to measure
-the quality impact of software development processes, tools and techniques.  For
-instance, we have a project (called
-\href{https://github.com/JacquesCarette/Drasil} {Drasil}
-\cite{SzymczakEtAl2016}), which facilitates a development process that focuses
-on knowledge capture, followed by generation of code and other software
-artifacts from the captured knowledge.  To understand the advantages and
-disadvantages of Drasil, we could measure its quality.  In particular, we would
-like to see the impact of a generative process on the quality of usability and
-modifiability.
-
-\subsection{Measuring Usability in the Future}
-
-In the future, we propose an experiment for assessing the usability of a given
-software package.  Some initial thoughts on how this might be done are recorded
-in this section.  To do the experiment we need an experimental subject, who will
-be required to complete tasks with the software being studied.  The interaction
-with the software will allow the study subject to experience the software's
-usability.  The tasks for the subject to complete would vary by domain;
-therefore, the tasks would be selected with the help of the Domain Expert.
-Criteria for selecting candidate tasks are as follows:
-
-\begin {enumerate}
-\item Tasks should be executable for subjects with novice to intermediate
-experience.
-\item All tasks should take no more than one hour.
-\item Tasks should include the basic/common use cases of the software package.
-\item Include tasks that require sequential or hierarchical steps for
-completion.
-\end {enumerate}
-
-Once the study subject has experience with the software, they will be in a
-position to judge its usability.  We will measure the usability using a
-standardized usability questionnaire, like the
-\href{https://www.usabilitest.com/sus-pdf-generator} {System Usability Scale
-questionnaire} or the
-\href{https://uiuxtrend.com/pssuq-post-study-system-usability-questionnaire/}
-{Post-Study System Usability Questionnaire}.
-
-As a starting point for the experiment design, the procedure could be something
-like the following:
-
-\begin {enumerate}
-\item Survey participants to collect pre-experiment data (background, experience
-of subjects (especially with domain software)).
-\item Participants perform tasks based on task defined by Domain Experts.
-\item Observe the study subjects (take notes, record sessions (screen
-recorder), watch for body languages and verbal cues).
-\item Survey the study subjects to collect feedback (post-experiment interview),
-complete usability questionnaire.
-\item Prepare a summary report of the experimental results.
-\end {enumerate}
-
-\subsection{Measuring Modifiability in the Future}
-
-The next experiment is designed to gather qualitative data regarding the
-modifiability of each software package. This proposed experiment also involves
-experimental subjects/participants, who in this case are asked to modify a set
-of software packages.  The specific modifications requested will again depend on
-the software domain.  In advance of the experiment the Domain Expert will be
-asked the likely changes for software in this domain.  We emphasize likely
-changes, instead of any changes because software cannot be designed so that
-everything is equally easy to change \cite{parnas1986rational}.  The procedure
-could be along the following lines:
+\section{Comparison Between LBM and Research Software for Artifacts}
+\label{Sec_CompareArtifacts}
+
+In this step of the methodology we answer \rqref{RQ_CompareArtifacts} by
+comparing the artifacts that we observed in LBM repositories to those observed
+and recommended for research software in general.  As part of filling in the
+measurement template (Section~\ref{}), the domain software is examined for the
+presence of artifacts, which are then categorized by frequency. We suggest
+grouping them into the categories: common, less common, and rare.  Common
+artifacts are those found in more than two thirds of the examples, less common
+artifacts are between 1/3 and 2/3, while rare are in less than 1/3 of the
+software packages.
+
+The observed frequency of artifacts should then be compared to the artifacts
+recommended by research software guidelines.  Table~\ref{} summarizes the
+artifacts recommended by some existing guidelines.
+
+\begin{table}[!h]
+\begin{center}
+\begin{tabular}{ p{3cm}p{1cm}p{1cm}p{1cm}p{1cm}p{1cm}p{1cm}p{1cm}p{1cm} }
+\toprule
+~ \ & \cite{USGS2019} & \cite{TobiasEtAl2018} & \cite{BrettEtAl2021} & \cite{WilsonEtAl2016} & \cite{SmithAndRoscoe2018} & \cite{HerouxEtAl2008} & \cite{ThielEtAl2020} & \cite{vanGompelEtAl2016}\\
+\midrule
+LICENSE & \checkmark &  & \checkmark & \checkmark & \checkmark & & \checkmark & \checkmark\\
+README &  &  &  &  & & & & \\
+CONTRIBUTING &  &  &  &  & & & & \\
+CITATION &  &  &  &  & & & & \\
+CHANGELOG &  &  &  &  & & & & \\
+INSTALL &  &  &  &  & & & & \\
+Uninstall &  &  &  &  & & & & \\
+Contributor Guide &  &  &  &  & & & & \\
+Getting started &  &  &  &  & & & & \\
+Tutorials &  &  &  &  & & & & \\
+FAQ &  &  &  &  & & & & \\
+Dependency List &  &  &  &  & & & & \\
+Issue Track &  &  &  &  & & & & \\
+Version Control &  &  &  &  & & & &\\ 
+Requirements &  &  &  &  & & & & \\
+Design &  &  &  &  & & & & \\
+API Doc. &  &  &  &  & & & & \\
+Build Scripts &  &  &  &  & & & & \\
+Unit Tests &  &  &  &  & & & & \\
+Test Plan &  &  &  &  & & & & \\
+Integ. Tests &  &  &  &  & & & & \\
+System Tests &  &  &  &  & & & & \\
+Acceptance Tests &  &  &  &  & & & & \\
+Regression Tests &  &  &  &  & & & & \\
+Code Style Guide &  &  &  &  & & & & \\
+Release Info. &  &  &  &  & & & & \\
+Product Roadmap &  &  &  &  & & & & \\
+\midrule
+\end{tabular}
+\caption{Commonly Recommended Artifacts in Software Development Guidelines} \label{Tbl_Guidelines}
+\end{center}
+\end{table}
+
+Another point of comparison is to look at the use of requirements related
+artifacts.  Software requirements documentation provides a good example of the
+common deviation between recommendations and practice.  Although requirements
+documentation is recommended by some
+\cite{TobiasEtAl2018,HerouxEtAl2008,SmithAndKoothoor2016}, in practice
+scientific software developers often do not produce a proper requirements
+specification \cite{HeatonAndCarver2015}. Sanders and Kelly \cite{SandersAndKelly2008} interviewed
+16 scientists from 10 disciplines and found that none of the scientists created
+requirements specifications, unless regulations in their field mandated such a
+document. \cite{Nguyen-HoanEtAl2010} showed requirements are the least commonly
+produced type of documentation for scientific software in general. When looking
+at the pain points for research software developers, \cite{WieseEtAl2019} found
+that software requirements and management is the software engineering discipline
+that most hurts scientific developers, accounting for 23 \% of the technical
+problems reported by study participants.  The lack of support for requirements
+is likely due to the perception that up-front requirements are impossible for
+research software \cite{CarverEtAl2007,SegalAndMorris2008}, but when the
+instance on ``up-front'' requirements is dropped, allowing the requirements to
+be written iteratively and incrementally, requirements are feasible
+\cite{Smith2016}.
+
+The results for the LBM example are shown in Table~\ref{artifactspresent}.  The
+majority of LBM generated artifacts correspond to general recommendations from
+research software developers.  A union of the three columns in
+Table~\ref{artifactspresent} mostly corresponds to recommendations made by the
+research software community, as shown in Table~\ref{}.
+
+\begin{table}
+\begin{center}
+\begin{tabular}{ p{12 cm}}
+%\toprule
+\textbf{Common}\\
+\midrule
+Developer List, Issue Tracker, Dependency List, Installation Guide, Theory
+Notes, Related Publications, Build Files, README File, License, Tutorial,
+Version Control\\
+%\midrule
+\textbf{Less Common}\\
+\midrule
+Change Log, Design Doc., Functional Spec., Performance Info., Test Cases, User
+Manual\\
+%\midrule
+\textbf{Rare}\\
+\midrule
+API Doc., Developer Manual, FAQ, Verification Plan, Video Guide, Requirements
+Spec.\\
+%\bottomrule
+\end{tabular}
+\caption{Artifacts Present in LBM Packages, Classified by Frequency}
+\label{artifactspresent}
+\end{center}
+\end{table}
+
+Although the LBM community participates in most of the practices we found listed
+in the general research software guidelines, some recommended practices were not
+observed or rarely observed. For instance API documentation was rarely observed
+for LBM software, but it is a frequently recommended artifact
+\cite{SmithAndRoscoe2018,ThielEtAl2020,vanGompelEtAl2016,OrvizEtAl2017,SSI2022,Zadka2018}.
+In addition to the items in the last column of Table~\ref{artifactspresent}, we
+can add the following community recommended items that we rarely, if ever,
+observed:
+
+\begin{itemize}
+\item A roadmap showing what is planned for the future
+\cite{Yo2021,ThielEtAl2020,vanGompelEtAl2016}.  We did see this information
+mentioned in some repos, but we did not observe a specific artifact devoted to
+this purpose.
+\item A code of conduct to explicitly say how developers should treat one
+another \cite{Yo2021,ThielEtAl2020}.  We did not observe a policy of this
+sort, possibly because the number of developers for each LBM project is small.
+\item Programming style guidelines so that programming labels and formatting are
+consistent \cite{TobiasEtAl2018,ThielEtAl2020,OrvizEtAl2017,Zadka2018,vanGompelEtAl2016,WilsonEtAl2014}.  We saw this information as part of
+some developer guides, but only rarely.
+\item Checklists can be used in projects to ensure that best practices are
+followed by all developers.  Some examples include checklists merging branches
+into master \cite{Brown2015}, checklists for saving and sharing changes to the
+project \cite{WilsonEtAl2016}, checklists for new and departing team members
+\cite{HerouxAndBernholdt2018}, checklists for processes related to commits and
+releases \cite{HerouxEtAl2008} and checklists for overall software quality
+\cite{ThielEtAl2020,SSI2022}.  For LBM software, ESPResSo has a checklist for
+managing releases, but otherwise they are used only rarely for LBM.
+\item Uninstall instructions \cite{vanGompelEtAl2016} were not observed for any
+of the LBM projects.
+\end{itemize}
+
+\section{Comparison of Tool Usage Between LBM and Other Research Software}
+\label{Sec_CompareTools}
+
+Software tools are used to support the development, verification, maintenance,
+and evolution of software, software processes, and artifacts \cite[p.\
+501]{GhezziEtAl2003}. Many tools are used by LBM software packages, as
+summarized in Table~\ref{tbl_tools}.  The tools are subdivided into development
+tools, dependencies, and project management tools.  In this section we answer
+\rqref{RQ_CompareToolsProjMngmnt} by comparing aspects of the tool usage in LBM
+to their utilization in the research software community in general.
+
+\begin{table}
+	\begin{center}
+	\begin{tabular}{ p{5.3 cm} p{4.9 cm} p{5 cm}}
+	\toprule
+	\textbf{Development Tools} & \textbf{Dependencies} & \textbf{Project
+	Management Tools} \\
+	\midrule
+	Continuous Integration & Build Automation Tools & Collaboration Tools\\
+	Code Editors & Technical Libraries & Email\\
+	Development Environments & Domain Specific Libraries & Change Tracking Tools\\
+	Runtime Environments &  & Version Control Tools\\
+	Compilers &  & Document Generation Tools\\
+	Unit Testing Tools &  & \\
+	Correctness Verification Tools &  & \\
+	\bottomrule
+	\end{tabular}
+	\caption{Observed development tools, dependencies and project management
+	tools} \label{tbl_tools}
+	\end{center}
+\end{table}
+
+Development tools support the development of end products, but do not become
+part of them, unlike dependencies that remain in the application once it is
+released \cite[p.\ 506]{GhezziEtAl2003}. Although not shown in
+Table~\ref{tbl_tools}, debuggers were also likely used.  Only three (ESPResSo,
+Ludwig and Musubi) packages mentioned continuous integration tools, like Travis
+CI. Code editors and compilers were explicitly noted to have been used by
+several packages, and were likely used by all of them. One of the packages
+(Ludwig) explicitly noted the use of proprietary unit testing code written in C.
+Likewise, the use of proprietary code for verifying the correctness of output
+was noted by one of the developers (pyLBM). Similar tools were likely used when
+developing the other software packages.
+
+For the dependency tools (Table~\ref{tbl_tools}), we observed that most of the
+software packages use some sort of build automation tools, most commonly Make.
+They all use various technical and domain specific libraries. Technical
+libraries include visualization (e.g.\ Matplotlib, ParaView, Pygame, VTK), data
+analysis (e.g.\ Anaconda, Torch), and message passing libraries (e.g.\ MPICH,
+Open MPI, PyZMQ). Domain specific libraries include scientific computing
+libraries (e.g.\ SciPy).
+
+Many of the software packages that were assessed were developed by teams of two
+or more people. Their work needed to be coordinated and managed.
+Table~\ref{tbl_tools} shows the types of project management tools that were
+explicitly noted in the artifacts, web-pages, or interviews with the developers.
+As with development tools and dependencies, it is possible that other types of
+project management tools are used, but they were not visible in the artifacts to
+which we had access.  Collaboration tools, most often email and video, are noted
+as being used when developing the software projects. Project management software
+was not explicitly mentioned, but it is possible that some of the projects use
+such software. Many of the projects are located on GitHub, where the developers
+use the platform to help manage their projects, especially bug related issues.
+Most of the projects appear to use change tracking and version control tools.
+Document generation tools are mentioned in 12 of the 24 projects. The tools
+Sphinx and Doxygen are explicitly used in this capacity.
+
+The poor adoption of version control tools that Wilson lamented in 2006
+\cite{Wilson2006} has greatly improved in the intervening years.  From
+Section~\ref{Sec_Maintainability}, 67\% of LBM packages use version control
+(GitHub, GitLab or CVS).  (From Table~\ref{gitrepodata}, 11/15 or 73\% of alive
+packages use version control.)  The proliferation of version control tools for
+LBM matches the increase in the broader research software community.  A little
+over 10 years ago version control was estimated to be used in only 50\% of
+research software projects \cite{Nguyen-HoanEtAl2010}, but even at that time
+\cite{Nguyen-HoanEtAl2010} noted an increase from previous usage levels. A
+survey in 2018 shows 81\% of developers use a version control system
+\cite{AlNoamanyAndBorghi2018}.  \cite{Smith2018} has similar results, showing
+version control usage for alive projects in mesh generation, geographic
+information systems and statistical software for psychiatry increasing from
+75\%, 89\% and 17\% (respectively) to 100\%, 95\% and 100\% (respectively) over
+a four year period ending in 2018.  (For completeness the same study showed a
+decrease in version control usage for seismology software over the same time
+period, from 41\% down to 36\%).  Almost every software guide cited in
+Section~\ref{Sec_CompareArtifacts} includes the advice to use version control.
+The high usage of version control tools in LBM software matches the trend in
+research software in general.
+
+As mentioned in Section~\ref{SecSurfCorrectAndVerifiab}, continuous integration
+is rarely used in LBM (3 of 24 packages or 12.5\%). This contrasts with the
+frequency with which continuous integration is recommended in research software
+development guidelines \cite{BrettEtAl2021,Brown2015,ThielEtAl2020,Zadka2018,vanGompelEtAl2016}.  We could not find published data on the
+frequency with which continuous integration is used in general research
+software.  Our impression is that like LBM software, other research software
+packages lag behind the recommended best practice of employing continuous
+integration tools.
+
+Interviews with the developers revealed a potentially more frequent use of both
+unit testing and continuous integration, compared to what was observed from
+studying the repository artifacts.
+
+\section{Comparison of Principles, Process and Methodologies to Research Software in General} \label{Sec_CompareMethodologies}
+
+This section answers research question \rqref{RQ_CompareMethodologies} by
+comparing the principles, processes and methodologies used for LBM software to
+what can be gleaned from the literature on research software in general. In our
+data collection for LBM software, the software development process is not
+explicitly indicated in the artifacts for most of the packages. However, during
+our interviews one developer (ESPResSo) told us their non-rigorous development
+model is like a combination of agile and waterfall. Employing a loosely defined
+process makes sense for LBM software, given that the teams are generally small
+and self-contained.  Although eleven of the packages explicitly convey that they
+would accept outside contributors, generally the teams are centralized, often
+working at the same institution.  Working at the same institution means that an
+informal process can show success, since informal conversations are relatively
+easy to have.
+
+Interviews with developers confirmed a similar project management processes. In
+teams of only a couple of developers, additions of new features or major changes
+are discussed with the entire team. Projects with more than a couple developers
+have lead developer roles. These lead developers review potential additions to
+the software. One of the developers (ESPResSo) that was interviewed noted that
+an ad hoc peer review process is used to assess major changes and additions.
+Using peer review (also called technical review) matches with recommended
+practice for research software \cite{HerouxEtAl2008,Givler2020,OrvizEtAl2017,USGS2019}.
+
+Two types of software changes were discussed during interviews with developers.
+One is feature additions, which arise from a scientific or functional need.
+These changes involve formal discussions within the development team, and lead
+developer participation is mandatory. The other change type is code refactoring,
+which only sometimes involves formal discussions with the development team. New
+developers were noted to play an increased role in these changes compared to the
+former changes. Software bugs are typically addressed in a similar fashion as
+code refactoring.  Issue tracking is commonly used to manage these changes.
+
+Our observations of a informally defined process, with elements of agile
+methods, matches what has been observed for research software in general.
+Scientific developers naturally use an agile philosophy \cite{AckroydEtAl2008,CarverEtAl2007,EasterbrookAndJohns2009,Segal2005,HeatonAndCarver2015}, or an
+amethododical process \cite{Kelly2013}, or a knowledge acquisition driven
+process \cite{Kelly2015}.
+
+Most of the software packages do not explicitly state the motivations or design
+principles that were considered when developing the software. One package,
+Sailfish, indicates in its artifacts that shortening the development time was
+considered in early stages of design, with the developers using Python and
+CUDA/OpenCL to achieve this without sacrificing performance. The Sailfish goals
+are explicitly listed as performance, scalability, agility and extendability,
+maintenance, and ease of use. The project scored well in these categories during
+our assessment.  The quality priorities for Sailfish roughly match the
+priorities observed for research software in general.
+\cite{Nguyen-HoanEtAl2010} surveyed developers to find the following list of
+qualities, in decreasing order of importance: reliability, functionality,
+maintainability, availability, performance, flexibility, testability, usability,
+reusability, traceability, and portability. The Sailfish list does not list
+reliability or functionality, but we can safely assume those are implicitly high
+priorities for any scientific project.  In earlier studies
+\cite{KellyAndSanders2008} and \cite{CarverEtAl2007} both highlight how
+important correctness is for research software.
+
+During our interviews, documentation was noted as playing a significant role in
+the development process, specifically with on-boarding new developers. A goal of
+documentation is to lower the entry barrier for these new contributors. The
+documentation provides information on how to get started, orients the user to
+artifacts and the source code, and explains how the system works, including the
+so-called simulation engine and interface.  This emphasis on documentation,
+especially for new developers, is echoed in research software guidelines.
+Multiple guidelines recommend a document explaining how to contribute to a
+project, often named CONTRIBUTING \cite{Yo2021,BrettEtAl2021,WilsonEtAl2016,ThielEtAl2020,vanGompelEtAl2016,OrvizEtAl2017,FLOSS2022,JimenezEtAl2017}.
+Tutorials \cite{ThielEtAl2020}, trouble shooting guides \cite{OrvizEtAl2017,SSI2022} and quick start examples \cite{ThielEtAl2020,vanGompelEtAl2016} are
+also recommended.  \cite{SmithAndRoscoe2018} suggests including instructions
+specifically for on-boarding new developers. For open source software in general
+(not just research software), \cite{Fogel2005} recommends providing tutorial
+style examples, developer guidelines, demos and screenshots.
+
+\section{Developer Pain Points} \label{painpoints}
+
+Based on interviews with 4 developers, this section aims to answer the research
+questions: i) What are the pain points for developers working on research
+software projects (\rqref{RQ_PainPoints})?; and, ii) How do the pain points of
+developers from LBM compare to the pain points for research software in general
+(\rqref{RQ_ComparePainPoints})?  Below we go through each of the identified pain
+points and include citations that contrast the LBM experience with observations
+from researchers in other domains.  Potential ways to address the pain points
+are covered in Section~\ref{Sec_AddressConcerns}. The full interview questions
+are found in \cite{SmithEtAl2021}.
+
+\cite{PintoEtAl2018} lists some pain points that did not come up in our
+conversations with LBM developers: Cross-platform compatibility, interruptions
+while coding, scope bloat, lack of user feedback, hard to collaborate on
+software projects, and aloneness. \cite{WieseEtAl2019} repeat some of the
+previous pain points and add the following: dependency management, data handling
+concerns (like data quality, data management and data privacy), reproducibility,
+and software scope determination. Although LBM developers did not mention these
+pain points, we cannot conclude that they are not relevant for LBM software
+development, since we only interviewed 4 LBM developers for about an hour each.
 
 \begin{enumerate}
-	\item Domain Expert lists all likely changes that a developer
-	might be asked to make in a software package in the domain.
-	\item Survey study participants to collect pre-experiment data (background,
-	experience of subjects (especially with domain software)).
-	\item Participants perform modification tasks for likely changes on each
-	software package being studied.
-	\item Observe the study subjects (take notes, record sessions (screen
-	recorder), watch for body languages and verbal cues).
-	\item Record time needed to make the changes.	
-	\item Confirm through testing that the modified software has the correct
-	behaviour.
-	\item Survey the study subjects to collect feedback (post-experiment interview).
-	\item Prepare a summary report of the experimental results.
+
+	\item[P\refstepcounter{pnum}\thepnum \label{P_LackDevTime}:] \textbf{Lack of
+	Development Time} A developer of pyLBM noted that their small development
+	team has a lack of time to implement new features. Small development teams
+	are common for LBM software packages (as shown in the measurement table
+	excerpt in Figure~\ref{measurement_template_image}). Lack of time is also
+	highlighted as a pain point by other research software developers
+	\cite{PintoEtAl2018,PintoEtAl2016,WieseEtAl2019}.
+
+	\item[P\refstepcounter{pnum}\thepnum \label{P_LackSoftDevExp}:] \textbf{Lack
+	of Software Development Experience} A lack of software development
+	experience was noted by the developer of TCLB, and others noted a need for
+	improving software engineering education. Many of the team members on their
+	project are domain experts, not computer scientists or software engineers.
+	This same trend is noted by \cite{Nguyen-HoanEtAl2010}, which showed only
+	23\% of research software survey respondents having a computing-related
+	background. Similarly, \cite{UditAndKatz2017} show that the majority (54\%)
+	of postdocs have not received training in software development.  The LBM
+	developer suggesting an increasing role for formal software education
+	matches the trend observed by \cite{PintoEtAl2018}, where their replication
+	of a previous study \cite{HannayEtAl2009}, shows a growing interest in
+	formal training (From 13\% of respondents in 2009 to 22\% in 2018).
+	\cite{PintoEtAl2018} found that some developers feel there is a mismatch
+	between coding skills and subject-matter skills.
+
+	\item[P\refstepcounter{pnum}\thepnum \label{P_LackFunding}:] \textbf{Lack of
+	Incentive and Funding} The TCLB developer noted a lack of incentives and
+	funding in academia for developing widely used scientific software. This
+	problem has also been noted others \cite{gewaltig2012quality,Goble2014,KaterbowAndFeulner2018}.  \cite{WieseEtAl2019} reported developer pains
+	related to publicity, since historically publishing norms make it difficult
+	to get credit for creating software.  As studied by
+	\cite{HowisonAndBullard2016}, research software (specifically biology
+	software, but the trend likely applies to other research software domains)
+	is infrequently cited.  \cite{PintoEtAl2018} also mentions the lack of
+	formal reward system for research software.
+
+	\item[P\refstepcounter{pnum}\thepnum \label{P_LackExtSupport}:] \textbf{Lack
+	of External Support} A concern was raised that there are no organizations
+	helping with the development of good quality software.  This concern is not
+	echoed in the literature because there are such organizations, including
+	\href{https://bssw.io/} {Better Scientific Software (BSSw)},
+	\href{https://www.software.ac.uk/} {Software Sustainability Institute}
+	\cite{CrouchEtAl2013}, and \href{https://software-carpentry.org/}{Software
+	Carpentry} [* {WilsonAndLumsdaine2006, Wilson2016} *]. Over time awareness of
+	these groups will grow, so this pain point is likely to disappear in the
+	future for LBM and other research software developers.
+
+	\item[P\refstepcounter{pnum}\thepnum \label{P_TechnologyHurdles}:]
+	\textbf{Technology Hurdles} Technology pain points include setting up
+	parallelization and continuous integration.
+
+	\item[P\refstepcounter{pnum}\thepnum \label{P_Correctness}:]
+	\textbf{Ensuring Correctness} Difficulties with ensuring correctness were
+	noted by several developers. Several developers alluded to difficulty with
+	testing the correctness of large numbers of features, and some manually
+	tested program output, as opposed to using automated testing. The TCLB
+	developer commented that the amount of testing data needed was sometimes
+	problematic, since free testing services do not offer adequate facilities
+	for large amounts of data, which means in-house testing solutions are
+	needed.  Other research software domains point to the following problems
+	with testing: i) \cite{PintoEtAl2018} mention the problem of insufficient
+	testing; ii) the survey of \cite{HannayEtAl2009} shows that more developers
+	think testing is important than the number that believe they have a
+	sufficient understanding of testing concepts; and, iii)
+	\cite{HannayEtAl2009,KanewalaAndBieman2013,KellyEtAl2011,WieseEtAl2019}
+	point to the oracle problem, which occurs in research software when we do
+	not have a means to judge the correctness of the calculated solutions.  The
+	LBM experience seems to overlap with i and ii, but the LBM developers did
+	not allude to problem iii (the oracle problem) in our conversation; we
+	believe that they have developed techniques to work around the oracle
+	problem.
+
+	\item[P\refstepcounter{pnum}\thepnum \label{P_Usability}:]
+	\textbf{Usability} Several developers noted that users sometimes try to use
+	incorrect LBM method combinations to solve their problems. Furthermore, some
+	users think that the packages will work out of the box to solve their cases,
+	while in reality CFD knowledge needs to be applied to correctly modify the
+	packages for the new endeavour.  \cite{WieseEtAl2019} also interacted with
+	developers that mentioned that users do not always have the expertise
+	required to install or use the software.
+
+	\item[P\refstepcounter{pnum}\thepnum \label{P_TechDebt}:] \textbf{Technical
+	Debt} The developer of ESPResSo said that their source code was written with
+	a specific application in mind, which later caused too much coupling between
+	components in the source code. This results in technical debt
+	\cite{KruchtenEtAl2012}, which has an impact on future modifiability and
+	reusability. Concern with technical debt is likely why researchers in the
+	survey of \cite{Nguyen-HoanEtAl2010} rated maintainability as the third
+	most important software quality. More recently the push for sustainable
+	software \cite{deSouzaEtAl2019} is motivated by the pain that past
+	developers have had with accumulating too much technical debt.
+
+	\item[P\refstepcounter{pnum}\thepnum \label{P_Documentation}:]
+	\textbf{Quality of Documentation}  The importance of documentation for both
+	users and developers was stressed throughout the interviews. However, it was
+	noted several times that a lack of time and funding (\ppref{P_LackDevTime})
+	has a negative affect on the documentation. Most of the developers are
+	scientific researchers evaluated on the scientific papers that they produce.
+	Writing and updating documentation is something that is done in their free
+	time, if that time arises. Inadequate research software documentation is
+	also mentioned by others \cite{PintoEtAl2018,WieseEtAl2019} and the
+	problem also arises with non-research software \cite{LethbridgeEtAl2003}.
+	Recommendations on the development of research software state that
+	developers should critically evaluate their own development processes in
+	terms of quality assurance and comply with international standards for
+	software documentation \cite{KaterbowAndFeulner2018}.
+
 \end{enumerate}
 
-The study subjects should make the same changes in multiple pieces of software.
-The reporting of the results will focus more on the relative time differences
-between the set of software packages, rather than the absolute time to make any
-given change. To remove biases caused by the participants experience, the
-different study subjects should use a different order as they go through the
-list of software packages.  Details for this experiment still need to be
-resolved, such as how to take the participants prior knowledge into account,
-especially with respect to their programming experience.
+\section{Lessons from LBM Developers} \label{Sec_AddressConcerns}
+
+This section summarizes the best practices from LBM developers that are taken to
+address the pain points mentioned in Section~\ref{painpoints}.  The main source
+of information is the qualitative data from developer interviews
+(Section~\ref{SecSurvey}).  The practices summarized in this section can
+potentially be emulated by the LBM software packages that do not currently
+follow them.  Moreover, these practices may also provide examples that can be
+followed by other research software domains.
+
+\subsection{Design For Change} \label{Sec_DesForChange}
+
+Increasing source code modularity, reducing duplicate information,
+and improving abstraction by developing well defined interfaces. This was noted
+by the developers of ESPResSo and pyLBM.  A developer of pyLBM mentioned that
+the geometries and models of their system had been ``decoupled'', using
+abstraction and modularization of the source code, to make it ``very easy to add
+[new] features''.  [*{Highlights the advantage of design.  Other projects can
+look to these examples.  Maybe cite the Anshu paper about Flash? Dune? and point
+to the advantages of design.}*] Addresses maintainability.  The developer of TCLB
+mentioned that their package had two sets of code, for executing the models on
+the CPU and GPU, and that maintenance was decreased by introducing macros, a
+practice which became a common part of the development process.
+
+A high degree of code modularity and abstraction was noted by developers as a
+measure to ensure the ease of future changes. This can be ensured by separating
+components and hiding information behind well defined interfaces. The developer
+of ESPResSo also noted that some of the code base was transitioned from C to
+C++, which could ease modifiability of that software package. The developer of
+TCLB noted that their software package was designed to allow for the addition of
+some LBM features, but changes to major aspects of the system would be
+difficult. For example, ``implementing a new model will be an easy
+contribution'', but changes to the ``Cartesian mesh...will be a nightmare''.
+[*{example suggests identifying likely changes.}*]
+
+Modularize the source code, separate components, hide information behind well
+defined interfaces. This is suggested in SCS literature, and in developer
+comments.  [*{one of Wilson's best practices.}*]
+
+\begin{itemize}
+	\item Modifiability. Consider flexibility of data structures and data
+	storage in the design stage. The package pyLBM redeveloped data structures
+	to ease future changes.
+	\item Reusability. Provide API documentation, if applicable. Only one
+	(ESPResSo) of the top five ranked packages provided API documentation.
+\end{itemize}
+
+\section{Recommendations for Future Practices} \label{Sec_Recommendations}
+
+Answers \rqref{RQ_Recommend}.  practices for the future.  Probably the easiest
+approach is to list the practice and then the PPs that it helps.  Our
+recommendations are not lists of what should have been done in the past, or what
+should be done now; they are just suggestions for consideration in the future.
+
+``unique challenges of scientific research (e.g., the frequent and unforeseen
+changes in requirements ...)'' \cite{PintoEtAl2018}. We can point out that in
+the future this viewpoint could change and why.  Not really a unique challenge.
+Cite Smith 2016?
+
+linters \cite{Zadka2018} were rarely observed in LBM projects.  Do not see much
+evidence of their presence in best practice guides.
+
+[*{Address lack of development time (\ppref{P_LackDevTime}): Aside from
+on-boarding new developers, time constraints could be mitigated by increasing
+developer efficiency, which could be addressed in several ways, including by
+improving the quality of documentation, or incorporating automatic code
+generation.}
+
+[*{Address lack of funding (\ppref{P_LackFunding}): The developer further
+commented that there are no journals that publish such scientific software
+source code. However, there are ways to get such source code cited. Work has
+been done to address this in \cite{SmithEtAl2016-softcite}, which presents a
+set of software citation principles and discusses ``how they could be used to
+implement software citation in the scholarly community''
+\cite{katz2019software}. other papers also mention this pain point.
+Solutions proposed by Katz, journal of open source etc. Journal of Open Source
+Software \cite{SmithEtAl2018-joss}.  Calling for direct software citations
+\cite{SmithEtAl2016-softcite}.  Guidance on software citation from
+\cite{KatzEtAl2021}.  Discussed in CarverEtAl2021 paper.}
+
+[*{Address quality of documentation (\ppref{P_Documentation}): Make required
+background of users explicit, possibly in a requirements document.}
+
+Surface Correctness and Verifiability. Use a requirements
+specification document. This is suggested in SCS literature, and several of
+the top ranked packages had such a document or reference to theory manuals.
+A potential template is presented in \cite{smith2005new}.
+
+Papers that talk about how to do open source.
+
+Requirements documentation is feasible, as shown by the prevalence of theory
+documents in LBM software.  Table~\ref{artifactspresent} shows that theory
+documentation is common.  We observed 17 of 24 packages had at least some theory
+documentation. Templates for scientific requirements, like
+\cite{SmithEtAl2007,SmithAndLai2005}, show theory is a significant part of the
+requirements documentation. With the addition of some extra information, the
+theory documents can be transformed into requirements specification.  The key
+extra information includes explicitly stating user characteristics, explicitly
+stating how the user interacts with the software in terms of input data
+requirements, and listing likely and unlikely changes.  Explicit statements
+about likely changes are invaluable in the design stage, since they provide
+developers guidance on how general the software needs to be.  For those wishing
+to maximize the value of a requirement specification, information could be added
+like prioritizing the nonfunctional requirements (to show the relative
+importance between qualities like portability, reliability and performance) and
+traceability information (to show the consequences of any changes to the
+assumptions).
+
+\section{Threats To Validity} \label{threats}
+
+This section examines potential threats to the validity of this state of the
+practice assessment. These can be categorized into methodology and data
+collection issues. The goal of this assessment isn't to rank the software, but
+to use the ranking exercise as a means to understand the state of the practice
+of LBM software development.
+
+The measures listed in our measurement template may not be broad enough to
+accurately capture some qualities. For example, there are only two measures of
+surface robustness. The measurement of robustness could be expanded, as it
+currently only measures unexpected input. Other faults could be introduced, but
+could require a large investment of time to develop, and might not be a fair
+measure for all packages. Similarly, reusability is assessed along the number of
+code files and LOC per file. While this measure is indicative of modularity, it
+is possible that some packages have many files, with few LOC, but the files do
+not contain source code that is easily reusable. The files may be poorly
+formatted, or the source code may be vague and have ambiguous identifiers.
+Furthermore, the measurement of understandability relies on 10 random source
+code files. It is possible that the 10 files that were chosen to represent a
+software package may not be a good representation of the understandability of
+that package.
+
+Regarding data collection, a risk to the validity of this assessment is missing
+or incorrect data. Some software package data may not have been measured due to
+technology issues like broken links. This issue arose with the measurement of
+Palabos, which had a broken link to its user manual, as noted in
+Section~\ref{Sec_CompareArtifacts}.
+
+Some pertinent data may not have been specified in public artifacts, or may be
+obscure within an artifact or web-page. The use of unit testing and continuous
+integration was mentioned in the artifacts of only three (ESPResSo, Ludwig,
+Musubi) packages. However, interviews suggested a more frequent use of both unit
+testing and continuous integration in the development processes than what was
+observed from the initial survey of the artifacts. For example, OpenLB, pyLBM,
+and TCLB use such methods during development despite this not being explicitly
+clear from an analysis of the material available online.
+
+Furthermore, design documentation was measured to be a ``less common'' artifact
+in this assessment, but it is probable that such documentation is part of all
+LBM packages. After all, developing SCS is not a trivial endeavour. It is likely
+that many packages have such documentation but did not make it public, and due
+to this the measured data is not a true reflection of software package quality.
+
+[*{There are more threats to validity.  Brainstorm.  Watch for them while 
+editing.  Look at Ao paper.}
+
+Problems with the measurement of qualities:
+
+\begin{itemize}
+\item The assumption that more code files is an indicator of reusability
+(Section~\ref{reusabilityresults}) may not be a valid assumption.
+\item Understandability was measured using only 10 random files
+(Section~\ref{Sec_SurfUnderstandability}).  This could be improved.
+\end{itemize}
+
+Overall ranking with equally weighted qualities isn't realistic
+(Section~\ref{Sec_OverallQuality}).
+
+Threat to validity - the gap between measures of everything else and Musubi
+(Section~\ref{identifysoftware}), and then the gap between the manual measures
+and the automated measures (Section~\ref{repmetrics}).
+
+From Section~\ref{Sec_CompareTools}: ``Interviews with the developers revealed a
+potentially more frequent use of both unit testing and continuous integration,
+compared to what was observed from studying the repository artifacts.''
+
+Interviews suggested a more frequent use of both unit testing and continuous
+integration in the development processes than what was observed from the initial
+survey. For example, OpenLB, pyLBM, and TCLB use such methods during
+development, despite this not being explicitly clear from an analysis of the
+material available online.  The correctness and verifiability of such packages
+is not measured well using surface analysis.
 
 \section{Concluding Remarks} \label{SecConcludingRemarks}
+
+Our comparison may point out areas where some LBM software
+packages fall short of current best practices. This is not intended to be a
+criticism of any existing packages, especially since in practice not every
+project needs to achieve the highest possible quality.  However, rather than
+delve into the nuances of which software can justify compromising which
+practices we will write our comparison under the ideal assumption that every
+project has sufficient resources to match best practices.
+
+For each domain we wish to highlight success stories that can be shared amongst
+domain community, and with the broader research software community, while at the
+same time watching for areas for potential future improvement.
 
@@
 
+With the wealth of data from assessing the state of practice for multiple
+domains, the next step is a meta-analysis.  We would look at how the different
+domains compare. What lessons from one domain could be applied in other domains?
+What (if any) differences exist in the pain points between domains?  Are there
+differences in the tools, processes, and documentation between domains?
+
+The current methodology is constrained by limited resources.  A 4 hour cap on
+the measurement time for each software package limits what can be assessed.
+Within this limit, we can't measure some important qualities, like usability and
+modifiability.  In the future, we propose a more time-consuming process that
+would capture these other quality measures.  To improve the feasibility, the
+more time consuming measurements would not have to be completed for all 30
+packages. Instead, a short list could be identified using the output of the AHP
+ranking to select the top projects, or to select a sample of interesting
+projects across the quality spectrum.
+
 \subsubsection{Acknowledgements} Please place your acknowledgments at
@@
 %
-% \bibliographystyle{splncs04}
-% \bibliography{mybibliography}
+\bibliographystyle{splncs04}
+\bibliography{DiggingDeeper}
 %
-\begin{thebibliography}{8}
-\bibitem{ref_article1}
-Author, F.: Article title. Journal \textbf{2}(5), 99--110 (2016)
-
-\bibitem{ref_lncs1}
-Author, F., Author, S.: Title of a proceedings paper. In: Editor,
-F., Editor, S. (eds.) CONFERENCE 2016, LNCS, vol. 9999, pp. 1--13.
-Springer, Heidelberg (2016). \doi{10.10007/1234567890}
-
-\bibitem{ref_book1}
-Author, F., Author, S., Author, T.: Book title. 2nd edn. Publisher,
-Location (1999)
-
-\bibitem{ref_proc1}
-Author, A.-B.: Contribution title. In: 9th International Proceedings
-on Proceedings, pp. 1--2. Publisher, Location (2010)
-
-\bibitem{ref_url1}
-LNCS Homepage, \url{http://www.springer.com/lncs}. Last accessed 4
-Oct 2017
-\end{thebibliography}
+% \begin{thebibliography}{8}
+% \bibitem{ref_article1}
+% Author, F.: Article title. Journal \textbf{2}(5), 99--110 (2016)
+
+% \bibitem{ref_lncs1}
+% Author, F., Author, S.: Title of a proceedings paper. In: Editor,
+% F., Editor, S. (eds.) CONFERENCE 2016, LNCS, vol. 9999, pp. 1--13.
+% Springer, Heidelberg (2016). \doi{10.10007/1234567890}
+
+% \bibitem{ref_book1}
+% Author, F., Author, S., Author, T.: Book title. 2nd edn. Publisher,
+% Location (1999)
+
+% \bibitem{ref_proc1}
+% Author, A.-B.: Contribution title. In: 9th International Proceedings
+% on Proceedings, pp. 1--2. Publisher, Location (2010)
+
+% \bibitem{ref_url1}
+% LNCS Homepage, \url{http://www.springer.com/lncs}. Last accessed 4
+% Oct 2017
+% \end{thebibliography}
 \end{document}

### 2022-02-17 StateOfPractice/Papers/ICCS-2022-SOP_MethodologyPlusExamples/DigDeep.tex
Added another column to Table 3

@@
 \begin{center}
-\begin{tabular}{ p{3cm}p{1cm}p{1cm}p{1cm}p{1cm}p{1cm}p{1cm}p{1cm}p{1cm} }
+\begin{tabular}{ p{3cm}p{1cm}p{1cm}p{1cm}p{1cm}p{1cm}p{1cm}p{1cm}p{1cm}p{1cm} }
 \toprule
-~ \ & \cite{USGS2019} & \cite{TobiasEtAl2018} & \cite{BrettEtAl2021} & \cite{WilsonEtAl2016} & \cite{SmithAndRoscoe2018} & \cite{HerouxEtAl2008} & \cite{ThielEtAl2020} & \cite{vanGompelEtAl2016}\\
+~ \ & \cite{USGS2019} & \cite{TobiasEtAl2018} & \cite{BrettEtAl2021} & \cite{WilsonEtAl2016} & \cite{SmithAndRoscoe2018} & \cite{HerouxEtAl2008} & \cite{ThielEtAl2020} & \cite{vanGompelEtAl2016} & \cite{OrvizEtAl2017}\\
 \midrule
-LICENSE & \checkmark &  & \checkmark & \checkmark & \checkmark & & \checkmark & \checkmark\\
+LICENSE & \checkmark &  & \checkmark & \checkmark & \checkmark & & \checkmark & \checkmark & \checkmark\\
 README &  &  &  &  & & & & \\
-CONTRIBUTING &  &  &  &  & & & & \\
-CITATION &  &  &  &  & & & & \\
-CHANGELOG &  &  &  &  & & & & \\
-INSTALL &  &  &  &  & & & & \\
-Uninstall &  &  &  &  & & & & \\
-Contributor Guide &  &  &  &  & & & & \\
-Getting started &  &  &  &  & & & & \\
-Tutorials &  &  &  &  & & & & \\
-FAQ &  &  &  &  & & & & \\
-Dependency List &  &  &  &  & & & & \\
-Issue Track &  &  &  &  & & & & \\
-Version Control &  &  &  &  & & & &\\ 
-Requirements &  &  &  &  & & & & \\
-Design &  &  &  &  & & & & \\
-API Doc. &  &  &  &  & & & & \\
-Build Scripts &  &  &  &  & & & & \\
-Unit Tests &  &  &  &  & & & & \\
-Test Plan &  &  &  &  & & & & \\
-Integ. Tests &  &  &  &  & & & & \\
-System Tests &  &  &  &  & & & & \\
-Acceptance Tests &  &  &  &  & & & & \\
-Regression Tests &  &  &  &  & & & & \\
-Code Style Guide &  &  &  &  & & & & \\
-Release Info. &  &  &  &  & & & & \\
-Product Roadmap &  &  &  &  & & & & \\
+CONTRIBUTING &  &  &  &  & & & & & \\
+CITATION &  &  &  &  & & & & & \\
+CHANGELOG &  &  &  &  & & & & & \\
+INSTALL &  &  &  &  & & & & & \\
+Uninstall &  &  &  &  & & & & & \\
+Contributor Guide &  &  &  &  & & & & & \\
+Getting started &  &  &  &  & & & & & \\
+Tutorials &  &  &  &  & & & & & \\
+FAQ &  &  &  &  & & & & & \\
+Dependency List &  &  &  & & & & & & \\
+Issue Track &  &  &  &  & & & & & \\
+Version Control &  &  &  &  & & & & &\\ 
+Requirements &  &  &  &  & & & & & \\
+Design &  &  &  &  & & & & & \\
+API Doc. &  &  &  &  & & & & & \\
+Build Scripts &  &  &  &  & & & & & \\
+Unit Tests &  &  &  &  & & & & & \\
+Test Plan &  &  &  &  & & & & & \\
+Integ. Tests &  &  &  &  & & & & &  \\
+System Tests &  &  &  &  & & & & & \\
+Acceptance Tests &  &  &  &  & & & & &  \\
+Regression Tests &  &  &  &  & & & & & \\
+Code Style Guide &  &  &  &  & & & & & \\
+Release Info. &  &  &  &  & & & & & \\
+Product Roadmap &  &  &  &  & & & & & \\
 \midrule

### 2022-02-17 StateOfPractice/Papers/ICCS-2022-SOP_MethodologyPlusExamples/DigDeep.tex
Reformatted tools table, added an additional row in Table 3

@@
 Studying the common properties within a family of related programs is termed a
-domain analysis.
-For the current methodology time constraints necessitate a shallow domain analysis.
-A table should be constructed that distinguishes the programs under study by the variabilities that distinguish them.
-In research software the variabilities are often related to assumptions.
-Table~\ref{} shows the variabilities for the LBM software example.
+domain analysis.  For the current methodology time constraints necessitate a
+shallow domain analysis.  A table should be constructed that distinguishes the
+programs under study by the variabilities that distinguish them.  In research
+software the variabilities are often related to assumptions.  Table~\ref{} shows
+the variabilities for the LBM software example.
 
@@
 Getting started &  &  &  &  & & & & & \\
+User manual &  &  &  &  & & & & & \\
 Tutorials &  &  &  &  & & & & & \\
@@
 \begin{center}
-	\begin{tabular}{ p{5.3 cm} p{4.9 cm} p{5 cm}}
-	\toprule
-	\textbf{Development Tools} & \textbf{Dependencies} & \textbf{Project
-	Management Tools} \\
+\begin{tabular}{ p{12 cm}}
+%\toprule
+\textbf{Development Tools}\\
 \midrule
-	Continuous Integration & Build Automation Tools & Collaboration Tools\\
-	Code Editors & Technical Libraries & Email\\
-	Development Environments & Domain Specific Libraries & Change Tracking Tools\\
-	Runtime Environments &  & Version Control Tools\\
-	Compilers &  & Document Generation Tools\\
-	Unit Testing Tools &  & \\
-	Correctness Verification Tools &  & \\
-	\bottomrule
+Continuous Integration, Code Editors, Development Environments, Runtime
+Environments, Compilers, Unit Testing Tools, Correctness Verification Tools\\
+%\midrule
+\textbf{Dependencies}\\
+\midrule
+Build Automation Tools, Technical Libraries, Domain Specific Libraries\\
+%\midrule
+\textbf{Project Management Tools}\\
+\midrule
+Collaboration Tools, Email, Change Tracking Tools, Version Control Tools, Document Generation Tools\\
+%\bottomrule
 \end{tabular}
-	\caption{Observed development tools, dependencies and project management
-	tools} \label{tbl_tools}
+\caption{Observed development tools, dependencies and project management tools} 
+\label{tbl_tools}
 \end{center}

### 2022-02-17 StateOfPractice/Papers/ICCS-2022-SOP_MethodologyPlusExamples/DigDeep.tex
Additional row for Table 3

@@
 Product Roadmap &  &  &  &  & & & & & \\
+Code of Conduct &  &  &  &  & & & & & \\
 \midrule
@@
 
-Another point of comparison is to look at the use of requirements related
-artifacts.  Software requirements documentation provides a good example of the
-common deviation between recommendations and practice.  Although requirements
-documentation is recommended by some
-\cite{TobiasEtAl2018,HerouxEtAl2008,SmithAndKoothoor2016}, in practice
-scientific software developers often do not produce a proper requirements
-specification \cite{HeatonAndCarver2015}. Sanders and Kelly \cite{SandersAndKelly2008} interviewed
-16 scientists from 10 disciplines and found that none of the scientists created
-requirements specifications, unless regulations in their field mandated such a
-document. \cite{Nguyen-HoanEtAl2010} showed requirements are the least commonly
-produced type of documentation for scientific software in general. When looking
-at the pain points for research software developers, \cite{WieseEtAl2019} found
-that software requirements and management is the software engineering discipline
-that most hurts scientific developers, accounting for 23 \% of the technical
-problems reported by study participants.  The lack of support for requirements
-is likely due to the perception that up-front requirements are impossible for
-research software \cite{CarverEtAl2007,SegalAndMorris2008}, but when the
-instance on ``up-front'' requirements is dropped, allowing the requirements to
-be written iteratively and incrementally, requirements are feasible
-\cite{Smith2016}.
-
 The results for the LBM example are shown in Table~\ref{artifactspresent}.  The
@@
 Table~\ref{artifactspresent} mostly corresponds to recommendations made by the
-research software community, as shown in Table~\ref{}.
+research software community, as shown in Table~\ref{Tbl_Guidelines}.
 
@@
 observed or rarely observed. For instance API documentation was rarely observed
-for LBM software, but it is a frequently recommended artifact
-\cite{SmithAndRoscoe2018,ThielEtAl2020,vanGompelEtAl2016,OrvizEtAl2017,SSI2022,Zadka2018}.
+for LBM software, but it is a frequently recommended (Table~\ref{Tbl_Guidelines}).
 In addition to the items in the last column of Table~\ref{artifactspresent}, we
@@
 \begin{itemize}
-\item A roadmap showing what is planned for the future
-\cite{Yo2021,ThielEtAl2020,vanGompelEtAl2016}.  We did see this information
-mentioned in some repos, but we did not observe a specific artifact devoted to
-this purpose.
+\item A roadmap showing what is planned for the future.
 \item A code of conduct to explicitly say how developers should treat one
-another \cite{Yo2021,ThielEtAl2020}.  We did not observe a policy of this
-sort, possibly because the number of developers for each LBM project is small.
+another.
 \item Programming style guidelines so that programming labels and formatting are

### 2022-02-18 StateOfPractice/Papers/ICCS-2022-SOP_MethodologyPlusExamples/DigDeep.tex
Edits to ICCS paper

@@
 this software differs from other classes of software.  Previous studies have
-focused on surveying developers \cite{Carveretc}, developer interviews \cite{}
-and case studies \cite{Segal}.  A valuable source of information that has
-received less attention is the data in the publicly available software
-repositories.  We propose a new methodology for digging deeper into how well
-projects are applying best practices by using a combination of manual and
-automated techniques to extract repository-based information.
+focused on surveying developers
+\cite{HannayEtAl2009,Nguyen-HoanEtAl2010,PintoEtAl2018}, developer interviews
+\cite{SandersAndKelly2008} and case studies \cite{CarverEtAl2007,Segal2005}.  A
+valuable source of information that has received less attention is the data in
+publicly available software repositories.  We propose digging deeper into how
+well projects are applying best practices by using a new methodology that
+combines manual and automated techniques to extract repository-based
+information.
 
@@
 all domains of research software, possibly distinguishing them by programming
-language (for instance, R developers \cite{Rstudy}), or by the role of the
-developers (for instance postdoctoral researchers \cite{Katz}).  For their part,
-case studies have focused on a few specific examples at a time, since that is
-the nature of case studies.  We propose applying our new methodology between
-these two extremes.  Rather than focus on all research software, or just a few
-examples, we will focus on one domain of research software at a time. The
-practical reason for defining a domain scope is that digging deep into
-repository data takes time, making a broad scope infeasible.  We have the
-constraint that the work load for applying the methodology to a given domain
-needs to be feasible for a team as small as one individual, and for a time that
-is short, ideally around a person month per domain.\footnote{A person month is
-considered to be $20$ working days ($4$ weeks in a month, with $5$ days of work
-per week) at $8$ person hours per day, or $20 \cdot 8 = 160$ person hours.}
-Focusing on one domain at a time has more than just practical advantages.  By
-restricting ourselves to a single domain we can bring domain knowledge and
-domain experts into the mix.  The domain customized insight provided by the
-assessment has the potential to help a specific domain as they adopt and develop
-new development practices.  Moreover, providing a means to measure different
+language (for instance, R developers \cite{PintoEtAl2018}), or by the role of
+the developers (for instance postdoctoral researchers \cite{UditAndKatz2017}).
+Case studies, on the other hand, have focused on a few specific examples at a
+time, as is the nature of case studies.  For our new methodology, we propose a
+scope between these two extremes.  Rather than focus on all research software,
+or just a few examples, we will focus on one domain of research software at a
+time. The practical reason for defining a domain scope is that digging deep into
+repository data takes time, making a broad scope infeasible. We have determined
+that a practical constraint for the work load is for a team as small as one
+individual doing one person month worth of work per domain.\footnote{A person
+month is considered to be $20$ working days ($4$ weeks in a month, with $5$ days
+of work per week) at $8$ person hours per day, or $20 \cdot 8 = 160$ person
+hours.} Focusing on one domain at a time has more than just practical
+advantages.  By restricting ourselves to a single domain we can bring domain
+knowledge and domain experts into the mix.  The domain customized insight
+provided by the assessment has the potential to help a specific domain as they
+adopt and develop new development practices.  Moreover, measuring different
 domains facilitates uncovering domain specific practices, which can later be
-compared and contrasted between different domains.
-
-Our methodology is built around 10 research questions for the domain being
-assessed.  The questions are used to structure the presentation of the
-methodology, so for each research question we point to the corresponding section
-of the paper.  Assuming we have identified the domain of interest, the first question is:
+compared and contrasted.
+
+Our methodology is built around 10 research questions.  Assuming that the domain
+has been identified (Section~\ref{SecIdentifyDomain}), the first question is:
 
@@
 	code must be available for all identified projects?
-	(Section~\ref{methodology})
+	(Sections~\ref{identifysoftware},~\ref{filtersoftware})
 \end{enumerate}
@@
 	practices, based on evidence found by experimenting with the software and
-	searching the artifacts available in each project's repository? (Section~\ref{AHPresults})
+	searching the artifacts available in each project's repository? (Sections~\ref{empiricalmeasures},~\ref{AHP})
 	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareHQ2Popular}:] How
@@
 	How do domain projects compare to research software in general with respect to
-	principles, processes and methodologies used? (Section~\ref{Sec_CompareMethodologies})
+	the processes used? (Section~\ref{Sec_CompareMethodologies})
 \end{enumerate}
@@
 Only so much information can be gleaned by digging into software repositories.
-To gain additional insight, we need to interview developers.  We need to learn
-their concerns and how they deal with these concerns; we need to learn their
-pain points.  We wish to know what practices are used by the top domain
-projects, so that others can potentially emulate these practices. We also wish
-to identify new practices by borrowing successful ideas from other domains. The
-above points are covered by the questions outlined below:
+To gain additional insight, we need to interview developers
+(Section~\ref{SecSurvey}).  We need to learn their concerns and how they deal
+with these concerns; we need to learn their pain points.  We wish to know what
+practices are used by the top domain projects, so that others can potentially
+emulate these practices. We also wish to identify new practices by borrowing
+successful ideas from other domains. The above points are covered by the
+questions outlined below:
 
@@
 measurement of the performance, for instance using benchmarks, to other projects
-\cite{kaagstrom1998gemm}. The current methodology updates the approach used in
+\cite{KaagstromEtAl1998}. The current methodology updates the approach used in
 prior assessments of domains like Geographic Information Systems
@@
 (Section~\ref{threats}) and our final conclusions and recommendations
-(Section~\ref{conclusion}).
-
-\section{Methodology} \label{StepsAQDS}
+(Section~\ref{SecConcludingRemarks}).
+
+\section{Methodology} \label{methodology}
 
@@
 	\item Use AHP to rank the software packages. (Section~\ref{AHP})
-	\item Analyze the results and answer the research questions. (Section~\ref{})
+	\item Analyze the results and answer the research questions.
+(Sections~\ref{repmetrics}---\ref{Sec_Recommendations}) %,~\ref{Sec_CompareArtifacts},~\ref{Sec_CompareTools},~\ref{Sec_CompareMethodologies},~\ref{painpoints},~\ref{Sec_AddressConcerns},
 \end{enumerate}
 
-Table~\ref{TabPersonHours} estimates the time required (in person hours) to
-complete a state of the practice assessment for a single domain.  The table
-assumes that the domain has already been decided and the Domain Expert has been
-recruited.  The time spent by the Domain Expert is not included in the numbers
-shown in the table, since the amount of time that the domain expert will work
-independently of the rest of the assessment team will be small.  Moreover, this
-amount of time will vary greatly depending on the preferred work habits of the
-Domain Expert.  The table follows the steps outlined above. It is assumed that
-the template spreadsheets linked in this document, and the developed AHP tool,
-will be employed, rather than developing new tools.  The person hours given are
-a rough estimate, based on our experience completing assessments for medical
-image analysis software \cite{Dong2021} and for Lattice Boltzmann Method (LBM)
-software \cite{Michalski2021}.  The estimate assumes 30 software packages will
-be measured; the numbers will need to be adjusted if the total packages changes.
-The total number of person hours is \total{totHours} hours.  This is close to
-our goal of 1 month of person hours (160 hours).
-
-\begin{table}[h]
-  \caption{Estimated Person Hours for Assessing the State of Practice for Domain
-  X} \label{TabPersonHours}
-  \centering
-  \begin{tabular}{p{12.5cm} l}
-    \toprule
-    \textbf{Task} & \textbf{Hours} \\
-    \midrule
-
-    Initial 1 hour meeting with the Domain Expert plus meeting prep & 5
-    \addtocounter{totHours}{5} \\
-
-    Identify broad list of candidate software
-    (Section~\ref{SecIdentifyCandSoft}) & 12 \addtocounter{totHours}{12} \\
-
-    Filter software list (Section~\ref{SecInitialFilter}) (10 minutes per
-    package) & 5 \addtocounter{totHours}{5} \\
-
-	Review software list with Domain Expert (Section~\ref{SecDomainExpert}) & 2
-	\addtocounter{totHours}{2} \\
-
-    Domain analysis (with help of Domain Expert)
-    (Section~\ref{SecDomainAnalysis}) & 20 \addtocounter{totHours}{20} \\
-
-	Vet domain analysis with Domain Expert (Section~\ref{SecDomainExpert}) & 3
-	\addtocounter{totHours}{3} \\
-
-	Gather source code and documentation for each package (10 minutes per
-	package)) & 5 \addtocounter{totHours}{5} \\
-
-    Collect repository based data (Section~\ref{SecEmpiricalMeasures}) (10
-    minutes per package) & 5 \addtocounter{totHours}{5} \\
-
-	Measure using measurement template (Section~\ref{SecMeasureTemplate}) (2.5
-    hours per repo) & 75 \addtocounter{totHours}{75} \\
-
-    Solicit developers for interviews & 2 \addtocounter{totHours}{2} \\
-
-    Conduct interviews (1.5 hour interviews with 10 developers (30\% recruitment
-    rate)) & 15 \addtocounter{totHours}{15} \\
-
-    AHP ranking & 2 \addtocounter{totHours}{2} \\
-
-	Work with Domain Expert to vet AHP ranking & 2 \addtocounter{totHours}{2} \\
-
-    Analyze data and answer research questions & 20 \addtocounter{totHours}{20}\\
-
-    \midrule
-    \textbf{Total} & \textbf{\thetotHours} \\
-    \bottomrule
-  \end{tabular}
-
-\end{table}
+Close to our goal of 160 person hours, we estimate 173 hours to complete the
+assessment of a given domain~\cite{SmithEtAl2021}.  This estimate assumes that
+the domain has already been decided and the Domain Expert has been recruited.
+The estimate assumes that the template spreadsheets and developed AHP
+tool~\cite{SmithEtAl2021}, will be employed, rather than developing new tools.
+The person hours given are a rough estimate, based on our experience completing
+assessments for medical image analysis software \cite{Dong2021} and for Lattice
+Boltzmann Method (LBM) software \cite{Michalski2021}.  The estimate assumes 30
+software packages will be measured; the numbers will need to be adjusted if the
+total packages changes.
 
@@
 
-For the LBM example the initial list had 46 packages \cite{Michalski2021}.
-
 \subsection{How to Filter the Software List?} \label{filtersoftware}
@@
 
-For the running example of LBM filtering by scope, usage, and age decreased the
+For the LBM example the initial list had 46 packages~\cite{Michalski2021}. For
+the running example of LBM filtering by scope, usage, and age decreased the
 initial list of 46 to a list of 24. Many of the removed packages could not be
@@
 
-Most of the data to be collected should be straightforward from reviewing the
-measurement template.  However, in a few cases extra guidance is necessary
-to eliminate ambiguity, as follows:
-
-\begin{enumerate}
-\item Publications about the software: A list of publications can be found
-  directly on the website of some software packages. For others use Google
-  Scholar or a similar index.
-\item To answer whether there evidence that performance was considered, search
-  the software artifacts for any mention of speed, storage, throughput,
-  performance optimization, parallelism, multi-core processing, or similar
-  considerations. The search function on GitHub can help.
-\item Evidence of a getting started tutorial may be found within another
-  artifact, like the user manual.
-\item To find evidence of continuous integration, search the software artifacts
-  for any mention of continuous integration. The search function on GitHub can
-  help.  In some cases, \texttt{yaml} files will provide a hint that continuous
-  integration is employed.
-\end{enumerate}
-
-Tools were use to find some of the measurements, such as the number of files,
+Tools are used to find some of the measurements, such as the number of files,
 number of lines of code (LOC), percentage of issues that are closed, etc. The
@@
 As in \cite{SmithEtAl2016}, Virtual machines (VMs) are used to provide an
-optimal testing environments for each package. VMs were used because it is
+optimal testing environments for each package. VMs are used because it is
 easier to start with a fresh environment without having to worry about existing
@@
 
-For the LBM project, our Domain Expert was Dr.\
-Zahra Motamed, Assistant Professor of Mechanical Engineering at McMaster
-University, Hamilton, Ontario, Canada.
-
 \subsection{Domain Analysis} \label{SecDomainAnalysis}
@@
 able to view the software as constituting a program family.  The concept of a
-program family is defined by \cite{parnas1976design} as ``a set of programs
+program family is defined by~\cite{parnas1976design} as ``a set of programs
 whose common properties are so extensive that it is advantageous to study the
@@
 
-\section{Comparison Between LBM and Research Software for Artifacts}
+\section{Comparison of Artifacts to Other Research Software}
 \label{Sec_CompareArtifacts}
@@
 
-\begin{table}
+\begin{table}[h!]
 \begin{center}
@@
 in the general research software guidelines, some recommended practices were not
-observed or rarely observed. For instance API documentation was rarely observed
-for LBM software, but it is a frequently recommended (Table~\ref{Tbl_Guidelines}).
-In addition to the items in the last column of Table~\ref{artifactspresent}, we
-can add the following community recommended items that we rarely, if ever,
-observed:
-
-\begin{itemize}
-\item A roadmap showing what is planned for the future.
-\item A code of conduct to explicitly say how developers should treat one
-another.
-\item Programming style guidelines so that programming labels and formatting are
-consistent \cite{TobiasEtAl2018,ThielEtAl2020,OrvizEtAl2017,Zadka2018,vanGompelEtAl2016,WilsonEtAl2014}.  We saw this information as part of
-some developer guides, but only rarely.
-\item Checklists can be used in projects to ensure that best practices are
-followed by all developers.  Some examples include checklists merging branches
-into master \cite{Brown2015}, checklists for saving and sharing changes to the
-project \cite{WilsonEtAl2016}, checklists for new and departing team members
-\cite{HerouxAndBernholdt2018}, checklists for processes related to commits and
-releases \cite{HerouxEtAl2008} and checklists for overall software quality
-\cite{ThielEtAl2020,SSI2022}.  For LBM software, ESPResSo has a checklist for
-managing releases, but otherwise they are used only rarely for LBM.
-\item Uninstall instructions \cite{vanGompelEtAl2016} were not observed for any
-of the LBM projects.
-\end{itemize}
-
-\section{Comparison of Tool Usage Between LBM and Other Research Software}
+observed or rarely observed. The following community recommended items were
+rarely, if ever, observed: API documentation, a roadmap, a code of conduct,
+programming style guide, uninstall instructions.
+
+\section{Comparison of Tools to Other Research Software}
 \label{Sec_CompareTools}
@@
 
-\section{Comparison of Principles, Process and Methodologies to Research Software in General} \label{Sec_CompareMethodologies}
+\section{Comparison of Processes to Other Research Software} \label{Sec_CompareMethodologies}
 

### 2022-02-18 StateOfPractice/Papers/ICCS-2022-SOP_MethodologyPlusExamples/DigDeep.tex
Reorg paper, remove some sections

@@
 	practices, based on evidence found by experimenting with the software and
-	searching the artifacts available in each project's repository? (Sections~\ref{empiricalmeasures},~\ref{AHP})
+	searching the artifacts available in each project's repository? (Sections~\ref{empiricalmeasures})
 	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareHQ2Popular}:] How
@@
 (Section~\ref{SecSurvey}).  We need to learn their concerns and how they deal
-with these concerns; we need to learn their pain points.  We wish to know what
-practices are used by the top domain projects, so that others can potentially
-emulate these practices. We also wish to identify new practices by borrowing
-successful ideas from other domains. The above points are covered by the
-questions outlined below:
+with these concerns; we need to learn their pain points.
 
@@
 	for research software in general? (Section~\ref{painpoints})
-	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Concerns}:] For domain
-	developers what specific best practices are taken to address the pain points
-	and software quality concerns? (Section~\ref{Sec_AddressConcerns})
-	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Recommend}:]
-	What research software development practice could potentially address the
-	pain points identified in \rqref{RQ_PainPoints}).
-	(Section~\ref{Sec_Recommendations})
 \end{enumerate}
@@
 	\item Interview the developers. (Section~\ref{SecSurvey})
-	\item Use AHP to rank the software packages. (Section~\ref{AHP})
+	\item Use AHP to rank the software packages. (Section~\ref{empiricalmeasures})
 	\item Analyze the results and answer the research questions.
@@
 (Section~\ref{sec_vet_software_list}) should also be engaged in selecting the
-candidate software.
-
-The following properties were considered when creating the list and reviewing
-the candidate software:
+candidate software.  The following properties were considered when creating the
+list and reviewing the candidate software:
 
@@
 
-For the third item in the above filter, software packages are characterized as
-`alive' if their related documentation had been updated within the last 18
-month. Otherwise a package is categorized as `dead'.
-
-For the LBM example the initial list had 46 packages~\cite{Michalski2021}. For
-the running example of LBM filtering by scope, usage, and age decreased the
-initial list of 46 to a list of 24. Many of the removed packages could not be
-tested as there was no installation guide, they were incomplete, source code was
-not publicly available, a license was needed, or the project was out of scope or
-not up to a standard that would support incorporating them into this
-study~\cite{Michalski2021}. Of the remaining packages, some were kept on the
-list despite being marked as dead due to their prevalence on authoritative lists
-on LBM software and due to their surface excellence.
+For the LBM example the initial list had 46 packages~\cite{Michalski2021}. This
+list was filtered by scope, usage, and age to decrease the length to 24. Many of
+the removed packages could not be tested as there was no installation guide,
+they were incomplete, source code was not publicly available, a license was
+needed, or the project was out of scope or not up to a standard that would
+support incorporating them into this study~\cite{Michalski2021}.
 
@@
 We rank the projects by how well they follow best practices
-(\rqref{RQ_HighestQuality}) via a measurement template, as described in
-\cite{SmithEtAl2021}.  For each software package (each column in the template),
-we fill-in the rows of the template. This process takes about 2 hours per
-package, with a cap of 4 hours. The time constraint is necessary so that the
-work load is feasible for a team as small as one, given our aim to cap the
-measurement phase at 160 person hours \cite{SmithEtAl2021}.  An excerpt of the
-template, in spreadsheet form, is shown in
-Figure~\ref{measurement_template_image}.
+(\rqref{RQ_HighestQuality}) via a measurement template~\cite{SmithEtAl2021}. For
+each software package (each column in the template), we fill-in the rows. This
+process takes about 2 hours per package, with a cap of 4 hours.  An excerpt of
+the template is shown in Figure~\ref{measurement_template_image}.
 
@@
 	\begin{center}
-	  %\includegraphics[width=1.0\textwidth]{./figures/measurement_template.pdf}
+	  \includegraphics[width=1.0\textwidth]{./figures/measurement_template.pdf}
 	  \caption{Excerpt of the Top Sections of the Measurement Template}
@@
 
-The full template consists of 108 questions categorized under 9 qualities:
+The full template~\cite{SmithEtAl2021} consists of 108 questions categorized
+under 9 qualities:
 \begin{inparaenum}[(i)]
@@
 (Figure~\ref{measurement_template_image}).   The summary section provides
-general information, such as the software name, number of developers, etc.  We
-follow the definitions given by \cite{gewaltig2012quality} for the software
-categories.  Public means software intended for public use.  Private means
-software aimed only at a specific group, while the concept category is used for
-software written simply to demonstrate algorithms or concepts. The three
-categories of development models are: open source, where source code is freely
-available under an open source license; free-ware, where a binary or executable
-is provided for free; and, commercial, where the user must pay for the software
-product.
-
+general information, such as the software name, number of developers, etc.
 Several of the qualities use the word ``surface''.  This is to highlight that,
@@
 number of lines of code (LOC), percentage of issues that are closed, etc. The
-tool \href{https://github.com/tomgi/git_stats}{GitStats} was used to measure
+tool \href{https://github.com/tomgi/git_stats}{GitStats} is used to measure
 each software package's GitHub repository for the number of binary files, the
@@
 intervals. The tool \href{https://github.com/boyter/scc}{Sloc Cloc and Code
-(scc)} was used to measure the number of text based files as well as the number
+(scc)} is used to measure the number of text based files as well as the number
 of total, code, comment, and blank lines in each GitHub repository.
 
-As in \cite{SmithEtAl2016}, Virtual machines (VMs) are used to provide an
-optimal testing environments for each package. VMs are used because it is
-easier to start with a fresh environment without having to worry about existing
-libraries and conflicts. Moreover, when the tests are complete the VM can be
-deleted, without any impact on the host operating system. The most significant
-advantage of using VMs is to level the playing field. Every software install
-starts from a clean slate, which removes ``works-on-my-computer'' errors.
-
-\subsection{Analytical Hierarchy Process} \label{AHP}
+Virtual machines (VMs) are used to provide an optimal testing environments for
+each package~\cite{SmithEtAl2016}. VMs are used because it is easier to start
+with a fresh environment so as to not worry about existing libraries and
+conflicts. Moreover, when the tests are complete the VM can be deleted, without
+any impact on the host operating system. The most significant advantage of using
+VMs is to level the playing field. Every software install starts from a clean
+slate, which removes ``works-on-my-computer'' errors.
 
@@
 (AHP), a decision-making technique that is used to compare multiple options by
-multiple criteria \cite{Saaty}. AHP performs a pairwise analysis using a matrix
-and generates an overall score as well as individual quality scores for each
-software package. The advantage of pair-wise comparisons is that they
-facilitates a separation of concerns.  Rather than worry about the entire
-problem, the decision maker can focus on one comparison at a time.  In our work
-AHP is used for comparing and ranking the software packages of a domain using
-the quality scores that are gathered in the Measurement Template
-(Appendix~\ref{SecGradingTemplate}). AHP performs a pairwise analysis between
-each of the 9 quality options for each of the 30 software packages.  This
-results in a matrix, which is used to generate an overall score for each
-software package for the given criteria. \cite{SmithEtAl2016} shows how AHP is
-applied to ranking software based on quality measures.
-
-Example of LBM measurement - maybe show the summary of all qualities?
+multiple criteria \cite{Saaty1980}. AHP performs a pairwise analysis using a
+matrix and generates an overall score as well as individual quality scores for
+each software package. The advantage of pair-wise comparisons is they facilitate
+a separation of concerns.  Rather than worry about the entire problem, the
+decision maker can focus on one comparison at a time.  In our work AHP performs
+a pairwise analysis between each of the 9 quality options for each of the
+(approximately) 30 software packages.  This results in a matrix, which is used
+to generate an overall score for each software package for the given
+criteria~\cite{SmithEtAl2016}.
+
+\begin{figure}[h!]
+	\centering
+		\includegraphics[width=1.0\textwidth]{./figures/finalscore_chart.pdf}
+		\caption{AHP Overall Score}
+		\label{Fig_OverallScore}
+\end{figure}
+
+Figure~\ref{Fig_OverallScore} shows the overall ranking of the LBM software
+packages. In the absence of other information on priorities, the overall ranking
+was calculated by assuming an equal weighting between all qualities. If the
+qualities were weighed differently, the overall software package ranking would
+change.
 
@@
 semi-structured based on the question list; we ask follow-up questions when
-necessary.  Each interview takes an 1 hour on average.
-
-Our methodology suggests requesting interviews with a developer from each of the
-30 software package.  Requests for interviews are sent to all packages so as to
-not cause a potential bias by singling out any subset of the list. Moreover,
-since not every developer will agree to the interview request, asking 30 times
-will typically yield a reasonable number of responses. In our experience, the
-response rate is between 15\% and 30\%.  In some cases multiple developers from
-the same project will agree to be interviewed. When sending out interview
-requests, we recommend finding the contacts on the projects’ website, or code
-repository, or publications, or the biographic pages of the teams’ institutions.
-We send at most two interview request emails to a contact for each software
-package.  Meeting will typically be held using on-line meeting software, like
-Zoom or Teams, facilitates recording and automatic transcription of the
-meetings.
+necessary.  Each interview should take about 1 hour.
 
@@
 \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/MACREM/Application.pdf}
-{MREB\#: 5219}.
-
-For LBM we were able to recruit 4 developers to participate in our study.
-Results in next sections.
+{MREB\#: 5219}.  For LBM we were able to recruit 4 developers to participate in
+our study. Results in next sections.
 
@@
 
-We partnered with a Domain Expert to vet our list of projects
-(\rqref{RQ_WhatProjects}) and our ranking (\rqref{RQ_HighestQuality}).  The
-Domain Expert is an important member of the state of the practice assessment
+Our methodology relies on engaging a Domain Expert to vet the list of projects
+(\rqref{RQ_WhatProjects}) and the AHP ranking (\rqref{RQ_HighestQuality}).  The
+Domain Expert is an important member of the assessment
 team. Pitfalls exist if non-experts attempt to acquire an authoritative list of
@@
 \end{inparaenum}
-
 Domain experts may be recruited from academia or industry.  The only
@@
 
-The Domain Expert has an important role with verifying the list of LBM packages.
-In advance of the first meeting with the Domain Expert, they were asked to
-create a list of top software packages in the domain.  This is done to help the
-expert get in the right mind set in advance of the meeting.  Moreover, by doing
-the exercise in advance, we avoid the potential pitfall of the expert approving
-the discovered list of software without giving it adequate thought.  The Domain
-Expert was also asked to vet the collected data and analysis.  In particular,
-they were asked to vet the proposed list of software packages and the AHP
-ranking.
-
 \subsection{Domain Analysis} \label{SecDomainAnalysis}
@@
 Since each domain we will study will have a reasonably small scope, we will be
-able to view the software as constituting a program family.  The concept of a
-program family is defined by~\cite{parnas1976design} as ``a set of programs
-whose common properties are so extensive that it is advantageous to study the
-common properties of the programs before analyzing individual members''.
-Studying the common properties within a family of related programs is termed a
-domain analysis.  For the current methodology time constraints necessitate a
-shallow domain analysis.  A table should be constructed that distinguishes the
-programs under study by the variabilities that distinguish them.  In research
-software the variabilities are often related to assumptions.  Table~\ref{} shows
-the variabilities for the LBM software example.
+able to view the software as constituting a program family. Studying the common
+properties within a family of related programs is termed a domain analysis.  For
+the current methodology time constraints necessitate a shallow domain analysis.
+A table should be constructed that distinguishes the programs under study by the
+variabilities that distinguish them.  In research software the variabilities are
+often related to assumptions.  Table~\ref{tbl_features} shows the variabilities
+for the LBM software example.
+
+we will
+contrast the different packages based on their features, as shown in
+Table~\ref{tbl_features}. The set of LBM programs form a program family, or a
+software product line.  The packages all have the commonality of using LBM
+techniques to simulate the motion of fluids, but there are also variabilities
+between them.  LBM packages can be distinguished by the dimension of problems
+they simulate, whether or not (and how) they support parallel computation,
+whether or not compressibility can be considered, whether reflexive boundaries
+can be employed, whether the simulation handles multi-fluids, whether the
+simulations handles turbulence, and whether complex geometries can be simulated.
+All of these variabilities are listed as columns in Table~\ref{tbl_features}
+with the software listed in alphabetical order. A more complete analysis of the
+commonalities and variabilities between LBM packages is provided in
+\cite{Michalski2021}.
+
+\begin{table}[h!]
+	\begin{center}
+		\begin{tabular}{ p{3cm}llllllll}
+			\toprule
+			Name & Dim & Pll & Com & Rflx & MFl & Turb & CGE & OS\\
+			\midrule
+			DL\_MESO (LBE) & 2, 3 & MPI/OMP & Y & Y & Y & Y & Y & W, M, L\\
+			ESPResSo & 1, 2, 3 & CUDA/MPI & Y & Y & Y & Y & Y & M, L\\
+			ESPResSo++ & 1, 2, 3 & MPI & Y & Y & Y & Y & Y & L\\
+			HemeLB & 3 & MPI & Y & Y & Y & Y & Y & L\\
+			laboetie & 2, 3 & MPI & Y & Y & Y & Y & Y & L\\
+			LatBo.jl & 2, 3 & - & Y & Y & Y & N & Y & L\\
+			LB2D-Prime & 2 & MPI & Y & Y & Y & Y & Y & W, L\\
+			LB3D & 3 & MPI & N & Y & Y & Y & Y & L\\
+			LB3D-Prime & 3 & MPI & Y & Y & Y & Y & Y & W, L\\
+			lbmpy & 2, 3 & CUDA & Y & Y & Y & Y & Y & L\\
+			lettuce & 2, 3 & CUDA & Y & Y & Y & Y & Y & W, M, L\\
+			LIMBES & 2 & OMP & Y & Y & N & N & Y & L\\
+			Ludwig & 2, 3 & MPI & Y & Y & Y & Y & Y & L\\
+			LUMA & 2, 3 & MPI & Y & Y & Y & Y & Y & W, M, L\\
+			MechSys & 2, 3 & - & Y & Y & Y & Y & Y & L\\
+			MP-LABS & 2, 3 & MPI/OMP & N & Y & Y & N & N & L\\
+			Musubi & 2, 3 & MPI & Y & Y & Y & Y & Y & W, L\\
+			OpenLB & 1, 2, 3 & MPI & Y & Y & Y & Y & Y & W, M, L\\
+			Palabos & 2, 3 & MPI & Y & Y & Y & Y & Y & W, L\\
+			pyLBM & 1, 2, 3 & MPI & Y & Y & N & Y & Y & W, M, L\\
+			Sailfish & 2, 3 & CUDA & Y & Y & Y & Y & Y & M, L\\
+			SunlightLB & 3 & - & Y & Y & N & N & Y & L\\
+			TCLB & 2, 3 & CUDA/MPI & Y & Y & Y & Y & Y & L\\
+			waLBerla & 2, 3 & MPI & Y & Y & Y & Y & Y & L\\
+			\bottomrule
+		\end{tabular}
+		\caption{Features of Software Packages (Dim for Dimension (1, 2, 3), Pll
+			for Parallel (CUDA, MPI, OpenMP (OMP)), Com for Compressible (Yes or
+			No), Rflx for Reflexive Boundary Condition (Yes or No), MFl for
+			Multi-fluid (Yes or No), Turb for Turbulent (Yes or No), CGE for
+			Complex Geometries (Yes or No), OS for Operating System (Windows
+			(W), macOS (M), Linux (L)))} \label{tbl_features}
+	\end{center}
+\end{table}
 
@@
 practices to the communities ranking.  Our best practices ranking comes from the
-results of the AHP ranking (Section~\ref{AHP}).  We estimate the communities
-ranking by repository stars and watches.  The comparison will provide insight on
-whether best practices are rewarded by popularity.  However, inconsistencies
-between the AHP ranking and the communities ranking are in inevitable for the
-following reasons: 
+results of the AHP ranking (Section~\ref{empiricalmeasures}).  We estimate the
+communities ranking by repository stars and watches.  The comparison will
+provide insight on whether best practices are rewarded by popularity.  However,
+inconsistencies between the AHP ranking and the communities ranking are in
+inevitable for the following reasons: 
 \begin{inparaenum}[i)]
@@
 
-The results for the LBM example are shown in Table~\ref{artifactspresent}.  The
-majority of LBM generated artifacts correspond to general recommendations from
-research software developers.  A union of the three columns in
-Table~\ref{artifactspresent} mostly corresponds to recommendations made by the
-research software community, as shown in Table~\ref{Tbl_Guidelines}.
-
-\begin{table}[h!]
-\begin{center}
-\begin{tabular}{ p{12 cm}}
-%\toprule
-\textbf{Common}\\
-\midrule
-Developer List, Issue Tracker, Dependency List, Installation Guide, Theory
-Notes, Related Publications, Build Files, README File, License, Tutorial,
-Version Control\\
-%\midrule
-\textbf{Less Common}\\
-\midrule
-Change Log, Design Doc., Functional Spec., Performance Info., Test Cases, User
-Manual\\
-%\midrule
-\textbf{Rare}\\
-\midrule
-API Doc., Developer Manual, FAQ, Verification Plan, Video Guide, Requirements
-Spec.\\
-%\bottomrule
-\end{tabular}
-\caption{Artifacts Present in LBM Packages, Classified by Frequency}
-\label{artifactspresent}
-\end{center}
-\end{table}
+The results for the LBM example are shown below.  The majority of LBM generated
+artifacts correspond to general recommendations from research software
+developers.  For LBM, a union of the three categories mostly corresponds to
+recommendations made by the research software community, as shown in
+Table~\ref{Tbl_Guidelines}.
+
+\begin{description}
+	\item[Common:] Developer List, Issue Tracker, Dependency List, Installation
+	Guide, Theory Notes, Related Publications, Build Files, README File,
+	License, Tutorial, Version Control
+	\item[Less Common:] Change Log, Design Doc., Functional Spec., Performance
+	Info., Test Cases, User Manual
+	\item[Rare:] API Doc., Developer Manual, FAQ, Verification Plan, Video Guide,
+	Requirements Spec.
+\end{description}
+
+% \begin{table}[h!]
+% \begin{center}
+% \begin{tabular}{ p{12 cm}}
+% %\toprule
+% \textbf{Common}\\
+% \midrule
+% Developer List, Issue Tracker, Dependency List, Installation Guide, Theory
+% Notes, Related Publications, Build Files, README File, License, Tutorial,
+% Version Control\\
+% %\midrule
+% \textbf{Less Common}\\
+% \midrule
+% Change Log, Design Doc., Functional Spec., Performance Info., Test Cases, User
+% Manual\\
+% %\midrule
+% \textbf{Rare}\\
+% \midrule
+% API Doc., Developer Manual, FAQ, Verification Plan, Video Guide, Requirements
+% Spec.\\
+% %\bottomrule
+% \end{tabular}
+% \caption{Artifacts Present in LBM Packages, Classified by Frequency}
+% \label{artifactspresent}
+% \end{center}
+% \end{table}
 
@@
 in the general research software guidelines, some recommended practices were not
-observed or rarely observed. The following community recommended items were
-rarely, if ever, observed: API documentation, a roadmap, a code of conduct,
-programming style guide, uninstall instructions.
+observed or rarely observed as follows: API documentation, a roadmap, a code of
+conduct, programming style guide, uninstall instructions.
 
@@
 
-	\item[P\refstepcounter{pnum}\thepnum \label{P_LackFunding}:] \textbf{Lack of
-	Incentive and Funding} The TCLB developer noted a lack of incentives and
-	funding in academia for developing widely used scientific software. This
-	problem has also been noted others \cite{gewaltig2012quality,Goble2014,KaterbowAndFeulner2018}.  \cite{WieseEtAl2019} reported developer pains
-	related to publicity, since historically publishing norms make it difficult
-	to get credit for creating software.  As studied by
-	\cite{HowisonAndBullard2016}, research software (specifically biology
-	software, but the trend likely applies to other research software domains)
-	is infrequently cited.  \cite{PintoEtAl2018} also mentions the lack of
-	formal reward system for research software.
-
 	\item[P\refstepcounter{pnum}\thepnum \label{P_LackExtSupport}:] \textbf{Lack
@@
 
-	\item[P\refstepcounter{pnum}\thepnum \label{P_TechnologyHurdles}:]
-	\textbf{Technology Hurdles} Technology pain points include setting up
-	parallelization and continuous integration.
-
-	\item[P\refstepcounter{pnum}\thepnum \label{P_Correctness}:]
-	\textbf{Ensuring Correctness} Difficulties with ensuring correctness were
-	noted by several developers. Several developers alluded to difficulty with
-	testing the correctness of large numbers of features, and some manually
-	tested program output, as opposed to using automated testing. The TCLB
-	developer commented that the amount of testing data needed was sometimes
-	problematic, since free testing services do not offer adequate facilities
-	for large amounts of data, which means in-house testing solutions are
-	needed.  Other research software domains point to the following problems
-	with testing: i) \cite{PintoEtAl2018} mention the problem of insufficient
-	testing; ii) the survey of \cite{HannayEtAl2009} shows that more developers
-	think testing is important than the number that believe they have a
-	sufficient understanding of testing concepts; and, iii)
-	\cite{HannayEtAl2009,KanewalaAndBieman2013,KellyEtAl2011,WieseEtAl2019}
-	point to the oracle problem, which occurs in research software when we do
-	not have a means to judge the correctness of the calculated solutions.  The
-	LBM experience seems to overlap with i and ii, but the LBM developers did
-	not allude to problem iii (the oracle problem) in our conversation; we
-	believe that they have developed techniques to work around the oracle
-	problem.
-
-	\item[P\refstepcounter{pnum}\thepnum \label{P_Usability}:]
-	\textbf{Usability} Several developers noted that users sometimes try to use
-	incorrect LBM method combinations to solve their problems. Furthermore, some
-	users think that the packages will work out of the box to solve their cases,
-	while in reality CFD knowledge needs to be applied to correctly modify the
-	packages for the new endeavour.  \cite{WieseEtAl2019} also interacted with
-	developers that mentioned that users do not always have the expertise
-	required to install or use the software.
-
-	\item[P\refstepcounter{pnum}\thepnum \label{P_TechDebt}:] \textbf{Technical
-	Debt} The developer of ESPResSo said that their source code was written with
-	a specific application in mind, which later caused too much coupling between
-	components in the source code. This results in technical debt
-	\cite{KruchtenEtAl2012}, which has an impact on future modifiability and
-	reusability. Concern with technical debt is likely why researchers in the
-	survey of \cite{Nguyen-HoanEtAl2010} rated maintainability as the third
-	most important software quality. More recently the push for sustainable
-	software \cite{deSouzaEtAl2019} is motivated by the pain that past
-	developers have had with accumulating too much technical debt.
-
-	\item[P\refstepcounter{pnum}\thepnum \label{P_Documentation}:]
-	\textbf{Quality of Documentation}  The importance of documentation for both
-	users and developers was stressed throughout the interviews. However, it was
-	noted several times that a lack of time and funding (\ppref{P_LackDevTime})
-	has a negative affect on the documentation. Most of the developers are
-	scientific researchers evaluated on the scientific papers that they produce.
-	Writing and updating documentation is something that is done in their free
-	time, if that time arises. Inadequate research software documentation is
-	also mentioned by others \cite{PintoEtAl2018,WieseEtAl2019} and the
-	problem also arises with non-research software \cite{LethbridgeEtAl2003}.
-	Recommendations on the development of research software state that
-	developers should critically evaluate their own development processes in
-	terms of quality assurance and comply with international standards for
-	software documentation \cite{KaterbowAndFeulner2018}.
-
 \end{enumerate}
-
-\section{Lessons from LBM Developers} \label{Sec_AddressConcerns}
-
-This section summarizes the best practices from LBM developers that are taken to
-address the pain points mentioned in Section~\ref{painpoints}.  The main source
-of information is the qualitative data from developer interviews
-(Section~\ref{SecSurvey}).  The practices summarized in this section can
-potentially be emulated by the LBM software packages that do not currently
-follow them.  Moreover, these practices may also provide examples that can be
-followed by other research software domains.
-
-\subsection{Design For Change} \label{Sec_DesForChange}
-
-Increasing source code modularity, reducing duplicate information,
-and improving abstraction by developing well defined interfaces. This was noted
-by the developers of ESPResSo and pyLBM.  A developer of pyLBM mentioned that
-the geometries and models of their system had been ``decoupled'', using
-abstraction and modularization of the source code, to make it ``very easy to add
-[new] features''.  [*{Highlights the advantage of design.  Other projects can
-look to these examples.  Maybe cite the Anshu paper about Flash? Dune? and point
-to the advantages of design.}*] Addresses maintainability.  The developer of TCLB
-mentioned that their package had two sets of code, for executing the models on
-the CPU and GPU, and that maintenance was decreased by introducing macros, a
-practice which became a common part of the development process.
-
-A high degree of code modularity and abstraction was noted by developers as a
-measure to ensure the ease of future changes. This can be ensured by separating
-components and hiding information behind well defined interfaces. The developer
-of ESPResSo also noted that some of the code base was transitioned from C to
-C++, which could ease modifiability of that software package. The developer of
-TCLB noted that their software package was designed to allow for the addition of
-some LBM features, but changes to major aspects of the system would be
-difficult. For example, ``implementing a new model will be an easy
-contribution'', but changes to the ``Cartesian mesh...will be a nightmare''.
-[*{example suggests identifying likely changes.}*]
-
-Modularize the source code, separate components, hide information behind well
-defined interfaces. This is suggested in SCS literature, and in developer
-comments.  [*{one of Wilson's best practices.}*]
-
-\begin{itemize}
-	\item Modifiability. Consider flexibility of data structures and data
-	storage in the design stage. The package pyLBM redeveloped data structures
-	to ease future changes.
-	\item Reusability. Provide API documentation, if applicable. Only one
-	(ESPResSo) of the top five ranked packages provided API documentation.
-\end{itemize}
-
-\section{Recommendations for Future Practices} \label{Sec_Recommendations}
-
-Answers \rqref{RQ_Recommend}.  practices for the future.  Probably the easiest
-approach is to list the practice and then the PPs that it helps.  Our
-recommendations are not lists of what should have been done in the past, or what
-should be done now; they are just suggestions for consideration in the future.
-
-``unique challenges of scientific research (e.g., the frequent and unforeseen
-changes in requirements ...)'' \cite{PintoEtAl2018}. We can point out that in
-the future this viewpoint could change and why.  Not really a unique challenge.
-Cite Smith 2016?
-
-linters \cite{Zadka2018} were rarely observed in LBM projects.  Do not see much
-evidence of their presence in best practice guides.
-
-[*{Address lack of development time (\ppref{P_LackDevTime}): Aside from
-on-boarding new developers, time constraints could be mitigated by increasing
-developer efficiency, which could be addressed in several ways, including by
-improving the quality of documentation, or incorporating automatic code
-generation.}
-
-[*{Address lack of funding (\ppref{P_LackFunding}): The developer further
-commented that there are no journals that publish such scientific software
-source code. However, there are ways to get such source code cited. Work has
-been done to address this in \cite{SmithEtAl2016-softcite}, which presents a
-set of software citation principles and discusses ``how they could be used to
-implement software citation in the scholarly community''
-\cite{katz2019software}. other papers also mention this pain point.
-Solutions proposed by Katz, journal of open source etc. Journal of Open Source
-Software \cite{SmithEtAl2018-joss}.  Calling for direct software citations
-\cite{SmithEtAl2016-softcite}.  Guidance on software citation from
-\cite{KatzEtAl2021}.  Discussed in CarverEtAl2021 paper.}
-
-[*{Address quality of documentation (\ppref{P_Documentation}): Make required
-background of users explicit, possibly in a requirements document.}
-
-Surface Correctness and Verifiability. Use a requirements
-specification document. This is suggested in SCS literature, and several of
-the top ranked packages had such a document or reference to theory manuals.
-A potential template is presented in \cite{smith2005new}.
-
-Papers that talk about how to do open source.
-
-Requirements documentation is feasible, as shown by the prevalence of theory
-documents in LBM software.  Table~\ref{artifactspresent} shows that theory
-documentation is common.  We observed 17 of 24 packages had at least some theory
-documentation. Templates for scientific requirements, like
-\cite{SmithEtAl2007,SmithAndLai2005}, show theory is a significant part of the
-requirements documentation. With the addition of some extra information, the
-theory documents can be transformed into requirements specification.  The key
-extra information includes explicitly stating user characteristics, explicitly
-stating how the user interacts with the software in terms of input data
-requirements, and listing likely and unlikely changes.  Explicit statements
-about likely changes are invaluable in the design stage, since they provide
-developers guidance on how general the software needs to be.  For those wishing
-to maximize the value of a requirement specification, information could be added
-like prioritizing the nonfunctional requirements (to show the relative
-importance between qualities like portability, reliability and performance) and
-traceability information (to show the consequences of any changes to the
-assumptions).
 
@@
 \section{Concluding Remarks} \label{SecConcludingRemarks}
+
+Next step to recommend ways to improve software (previously section 9).  We wish
+to know what practices are used by the top domain projects, so that others can
+potentially emulate these practices. We also wish to identify new practices by
+borrowing successful ideas from other domains.
 

### 2022-02-18 StateOfPractice/Papers/ICCS-2022-SOP_MethodologyPlusExamples/DigDeep.tex
Updates to paper, now includes figures

@@
 	can be found in \cite{SmithEtAl2021}. (Section~\ref{empiricalmeasures})
+	\item Use AHP to rank the software packages. (Section~\ref{empiricalmeasures})
 	\item Interview the developers. (Section~\ref{SecSurvey})
-	\item Use AHP to rank the software packages. (Section~\ref{empiricalmeasures})
+    \item Domain analysis.  (Section~\ref{SecDomainAnalysis})
 	\item Analyze the results and answer the research questions.
-(Sections~\ref{repmetrics}---\ref{Sec_Recommendations}) %,~\ref{Sec_CompareArtifacts},~\ref{Sec_CompareTools},~\ref{Sec_CompareMethodologies},~\ref{painpoints},~\ref{Sec_AddressConcerns},
+(Sections~\ref{repmetrics}---\ref{painpoints}) %,~\ref{Sec_CompareArtifacts},~\ref{Sec_CompareTools},~\ref{Sec_CompareMethodologies},~\ref{painpoints},~\ref{Sec_AddressConcerns},
 \end{enumerate}
@@
 Several of the research question (\rqref{RQ_CompareToolsProjMngmnt},
-\rqref{RQ_CompareMethodologies}, \rqref{RQ_PainPoints}, \rqref{RQ_Concerns} and
-\rqref{RQ_Recommend}) require going beyond the quantitative data from the
-measurement template. To gain the required insight, we interview developers
-using a list of 20 questions \cite{SmithEtAl2021}. The questions cover the
-background of the development teams, the interviewees, and the software itself.
-We ask the developers how they organize their projects. We also ask them about
-their understanding of the users. Some questions focus on the current and past
-difficulties, and the solutions the team has found, or will try. We also discuss
-the importance of, and the current situation for, documentation. A few questions
-are about specific software qualities, such as maintainability,
-understandability, usability, and reproducibility. The interviews are
-semi-structured based on the question list; we ask follow-up questions when
-necessary.  Each interview should take about 1 hour.
+\rqref{RQ_CompareMethodologies} and \rqref{RQ_PainPoints}) require going beyond
+the quantitative data from the measurement template. To gain the required
+insight, we interview developers using a list of 20 questions
+\cite{SmithEtAl2021}. The questions cover the background of the development
+teams, the interviewees, and the software itself. We ask the developers how they
+organize their projects. We also ask them about their understanding of the
+users. Some questions focus on the current and past difficulties, and the
+solutions the team has found, or will try. We also discuss the importance of,
+and the current situation for, documentation. A few questions are about specific
+software qualities, such as maintainability, understandability, usability, and
+reproducibility. The interviews are semi-structured based on the question list;
+we ask follow-up questions when necessary.  Each interview should take about 1
+hour.
 
@@
 {MREB\#: 5219}.  For LBM we were able to recruit 4 developers to participate in
-our study. Results in next sections.
+our study.
 
@@
 properties within a family of related programs is termed a domain analysis.  For
-the current methodology time constraints necessitate a shallow domain analysis.
-A table should be constructed that distinguishes the programs under study by the
-variabilities that distinguish them.  In research software the variabilities are
-often related to assumptions.  Table~\ref{tbl_features} shows the variabilities
-for the LBM software example.
-
-we will
-contrast the different packages based on their features, as shown in
-Table~\ref{tbl_features}. The set of LBM programs form a program family, or a
-software product line.  The packages all have the commonality of using LBM
-techniques to simulate the motion of fluids, but there are also variabilities
-between them.  LBM packages can be distinguished by the dimension of problems
-they simulate, whether or not (and how) they support parallel computation,
-whether or not compressibility can be considered, whether reflexive boundaries
-can be employed, whether the simulation handles multi-fluids, whether the
-simulations handles turbulence, and whether complex geometries can be simulated.
-All of these variabilities are listed as columns in Table~\ref{tbl_features}
-with the software listed in alphabetical order. A more complete analysis of the
-commonalities and variabilities between LBM packages is provided in
-\cite{Michalski2021}.
+the current methodology time constraints necessitate a shallow analysis. For
+each domain a table should be constructed that distinguishes the programs under
+study by their variabilities.  In research software the variabilities are often
+related to assumptions.  Table~\ref{tbl_features} shows the variabilities for
+the LBM software example~\cite{Michalski2021}.
 
@@
 	\begin{center}
-		\begin{tabular}{ p{3cm}llllllll}
+		\begin{tabular}{ p{3cm}p{1.25cm}p{2.25cm}llllll}
 			\toprule
@@
 	\item stars are known to not be a particularly good measure of popularity,
-	since young projects have less time to accumulate stars \cite{}; 
+	because of how people use stars and since young projects have less time to
+	accumulate stars~\cite{Szulik2017}; 
 	\item and, as for consumer products, there are more factors that determine
@@
 column contains less data, since two of the packages (Palabos, waLBerla) use
-GitLab, which does not track watches. Packages designated as lower quality often
-do not use GitHub or GitLab, or have only a few stars/watches. Although the AHP
-ranking and the community popularity estimate are not perfect measures, they do
-suggest a correlation between best practices and popularity. We do not know
-which comes first, the use of best practices or popularity, but we do know that
-the top ranked packages tend to incorporate best practices.
+GitLab, which does not track watches. Although the AHP ranking and the community
+popularity estimate are not perfect measures, they do suggest a correlation
+between best practices and popularity.
 
@@
 In this step of the methodology we answer \rqref{RQ_CompareArtifacts} by
-comparing the artifacts that we observed in LBM repositories to those observed
-and recommended for research software in general.  As part of filling in the
-measurement template (Section~\ref{}), the domain software is examined for the
-presence of artifacts, which are then categorized by frequency. We suggest
-grouping them into the categories: common, less common, and rare.  Common
-artifacts are those found in more than two thirds of the examples, less common
-artifacts are between 1/3 and 2/3, while rare are in less than 1/3 of the
-software packages.
-
-The observed frequency of artifacts should then be compared to the artifacts
-recommended by research software guidelines.  Table~\ref{} summarizes the
-artifacts recommended by some existing guidelines.
+comparing the artifacts that we observed in the domain repositories to those
+observed and recommended for research software in general.  While filling in the
+measurement template (Section~\ref{empiricalmeasures}), the domain software is
+examined for the presence of artifacts, which are then categorized by frequency.
+We suggest grouping them into the categories: common (more than 2/3 of
+projects), less common (between 1/3 and 2/3), and rare (less than 1/3 of
+projects). The observed frequency of artifacts should then be compared to the
+artifacts recommended by research software guidelines, as summarized in
+Table~\ref{Tbl_Guidelines}.
 
@@
 recommendations made by the research software community, as shown in
-Table~\ref{Tbl_Guidelines}.
+Table~\ref{Tbl_Guidelines}. Although the LBM community participates in most of
+the practices we found listed in the general research software guidelines, some
+recommended practices were not observed or rarely observed as follows: API
+documentation, a roadmap, a code of conduct, programming style guide, uninstall
+instructions.
 
@@
 
-Although the LBM community participates in most of the practices we found listed
-in the general research software guidelines, some recommended practices were not
-observed or rarely observed as follows: API documentation, a roadmap, a code of
-conduct, programming style guide, uninstall instructions.
-
 \section{Comparison of Tools to Other Research Software}
@@
 
-Software tools are used to support the development, verification, maintenance,
-and evolution of software, software processes, and artifacts \cite[p.\
-501]{GhezziEtAl2003}. Many tools are used by LBM software packages, as
-summarized in Table~\ref{tbl_tools}.  The tools are subdivided into development
-tools, dependencies, and project management tools.  In this section we answer
-\rqref{RQ_CompareToolsProjMngmnt} by comparing aspects of the tool usage in LBM
-to their utilization in the research software community in general.
-
-\begin{table}[h!]
-\begin{center}
-\begin{tabular}{ p{12 cm}}
-%\toprule
-\textbf{Development Tools}\\
-\midrule
-Continuous Integration, Code Editors, Development Environments, Runtime
-Environments, Compilers, Unit Testing Tools, Correctness Verification Tools\\
-%\midrule
-\textbf{Dependencies}\\
-\midrule
-Build Automation Tools, Technical Libraries, Domain Specific Libraries\\
-%\midrule
-\textbf{Project Management Tools}\\
-\midrule
-Collaboration Tools, Email, Change Tracking Tools, Version Control Tools, Document Generation Tools\\
-%\bottomrule
-\end{tabular}
-\caption{Observed development tools, dependencies and project management tools} 
-\label{tbl_tools}
-\end{center}
-\end{table}
+To answer \rqref{RQ_CompareToolsProjMngmnt} we need to summarize tools that are
+visible in the repositories and that are mentioned during the developer
+interviews. Software tools are used to support the development, verification,
+maintenance, and evolution of software, software processes, and
+artifacts~\cite[p.\ 501]{GhezziEtAl2003}. Development tools support the
+development of end products, but do not become part of them, unlike dependencies
+that remain in the application once it is released \cite[p.\
+506]{GhezziEtAl2003}. The tools found for LBM software packages are as follows:
+
+\begin{description}
+	\item[Development Tools:] Continuous Integration, Code Editors, Development
+	Environments, Runtime Environments, Compilers, Unit Testing Tools,
+	Correctness Verification Tools
+	\item[Dependencies:] Build Automation Tools, Technical Libraries, Domain
+	Specific Libraries
+	\item[Project Management Tools:] Collaboration Tools, Email, Change Tracking
+	Tools, Version Control Tools, Document Generation Tools
+\end{description}
+
+% \begin{table}[h!]
+% \begin{center}
+% \begin{tabular}{ p{12 cm}}
+% %\toprule
+% \textbf{Development Tools}\\
+% \midrule
+% Continuous Integration, Code Editors, Development Environments, Runtime
+% Environments, Compilers, Unit Testing Tools, Correctness Verification Tools\\
+% %\midrule
+% \textbf{Dependencies}\\
+% \midrule
+% Build Automation Tools, Technical Libraries, Domain Specific Libraries\\
+% %\midrule
+% \textbf{Project Management Tools}\\
+% \midrule
+% Collaboration Tools, Email, Change Tracking Tools, Version Control Tools, Document Generation Tools\\
+% %\bottomrule
+% \end{tabular}
+% \caption{Observed development tools, dependencies and project management tools} 
+% \label{tbl_tools}
+% \end{center}
+% \end{table}
+
+Once the data on tools is collected, two specific tools that can be compared to
+research software norms are version control and continuous integration.
 
@@
 
+Ask developers during the interviews about their pain points.  Compare the identified pain points to the literature that discusses pain points \cite{Wieseet}.
 Based on interviews with 4 developers, this section aims to answer the research

### 2022-02-18 StateOfPractice/Papers/ICCS-2022-SOP_MethodologyPlusExamples/DigDeep.tex
Edits to Sections 5 and 6.

@@
 that remain in the application once it is released \cite[p.\
-506]{GhezziEtAl2003}. The tools found for LBM software packages are as follows:
+506]{GhezziEtAl2003}. As an example, the tools found for LBM software packages
+are as follows:
 
@@
 
-% \begin{table}[h!]
-% \begin{center}
-% \begin{tabular}{ p{12 cm}}
-% %\toprule
-% \textbf{Development Tools}\\
-% \midrule
-% Continuous Integration, Code Editors, Development Environments, Runtime
-% Environments, Compilers, Unit Testing Tools, Correctness Verification Tools\\
-% %\midrule
-% \textbf{Dependencies}\\
-% \midrule
-% Build Automation Tools, Technical Libraries, Domain Specific Libraries\\
-% %\midrule
-% \textbf{Project Management Tools}\\
-% \midrule
-% Collaboration Tools, Email, Change Tracking Tools, Version Control Tools, Document Generation Tools\\
-% %\bottomrule
-% \end{tabular}
-% \caption{Observed development tools, dependencies and project management tools} 
-% \label{tbl_tools}
-% \end{center}
-% \end{table}
-
-Once the data on tools is collected, two specific tools that can be compared to
-research software norms are version control and continuous integration.
-
-Development tools support the development of end products, but do not become
-part of them, unlike dependencies that remain in the application once it is
-released \cite[p.\ 506]{GhezziEtAl2003}. Although not shown in
-Table~\ref{tbl_tools}, debuggers were also likely used.  Only three (ESPResSo,
-Ludwig and Musubi) packages mentioned continuous integration tools, like Travis
-CI. Code editors and compilers were explicitly noted to have been used by
-several packages, and were likely used by all of them. One of the packages
-(Ludwig) explicitly noted the use of proprietary unit testing code written in C.
-Likewise, the use of proprietary code for verifying the correctness of output
-was noted by one of the developers (pyLBM). Similar tools were likely used when
-developing the other software packages.
-
-For the dependency tools (Table~\ref{tbl_tools}), we observed that most of the
-software packages use some sort of build automation tools, most commonly Make.
-They all use various technical and domain specific libraries. Technical
-libraries include visualization (e.g.\ Matplotlib, ParaView, Pygame, VTK), data
-analysis (e.g.\ Anaconda, Torch), and message passing libraries (e.g.\ MPICH,
-Open MPI, PyZMQ). Domain specific libraries include scientific computing
-libraries (e.g.\ SciPy).
-
-Many of the software packages that were assessed were developed by teams of two
-or more people. Their work needed to be coordinated and managed.
-Table~\ref{tbl_tools} shows the types of project management tools that were
-explicitly noted in the artifacts, web-pages, or interviews with the developers.
-As with development tools and dependencies, it is possible that other types of
-project management tools are used, but they were not visible in the artifacts to
-which we had access.  Collaboration tools, most often email and video, are noted
-as being used when developing the software projects. Project management software
-was not explicitly mentioned, but it is possible that some of the projects use
-such software. Many of the projects are located on GitHub, where the developers
-use the platform to help manage their projects, especially bug related issues.
-Most of the projects appear to use change tracking and version control tools.
-Document generation tools are mentioned in 12 of the 24 projects. The tools
-Sphinx and Doxygen are explicitly used in this capacity.
-
+Once the data on tools is collected, the use of two specific tools should be
+compared to research software norms: version control and continuous integration.
 The poor adoption of version control tools that Wilson lamented in 2006
-\cite{Wilson2006} has greatly improved in the intervening years.  From
-Section~\ref{Sec_Maintainability}, 67\% of LBM packages use version control
-(GitHub, GitLab or CVS).  (From Table~\ref{gitrepodata}, 11/15 or 73\% of alive
-packages use version control.)  The proliferation of version control tools for
-LBM matches the increase in the broader research software community.  A little
-over 10 years ago version control was estimated to be used in only 50\% of
-research software projects \cite{Nguyen-HoanEtAl2010}, but even at that time
+\cite{Wilson2006} has greatly improved in the intervening years.  A little over
+10 years ago version control was estimated to be used in only 50\% of research
+software projects \cite{Nguyen-HoanEtAl2010}, but even at that time
 \cite{Nguyen-HoanEtAl2010} noted an increase from previous usage levels. A
@@
 Section~\ref{Sec_CompareArtifacts} includes the advice to use version control.
-The high usage of version control tools in LBM software matches the trend in
-research software in general.
-
-As mentioned in Section~\ref{SecSurfCorrectAndVerifiab}, continuous integration
-is rarely used in LBM (3 of 24 packages or 12.5\%). This contrasts with the
-frequency with which continuous integration is recommended in research software
-development guidelines \cite{BrettEtAl2021,Brown2015,ThielEtAl2020,Zadka2018,vanGompelEtAl2016}.  We could not find published data on the
-frequency with which continuous integration is used in general research
-software.  Our impression is that like LBM software, other research software
-packages lag behind the recommended best practice of employing continuous
-integration tools.
-
-Interviews with the developers revealed a potentially more frequent use of both
-unit testing and continuous integration, compared to what was observed from
-studying the repository artifacts.
+For LBM packages 67\% of LBM packages use version control (GitHub, GitLab or
+CVS). The high usage of version control tools in LBM software matches the trend
+in research software in general.
+
+Continuous integration is rarely used in LBM (3 of 24 packages or 12.5\%). This
+contrasts with the frequency with which continuous integration is recommended in
+research software development guidelines
+\cite{BrettEtAl2021,Brown2015,ThielEtAl2020,Zadka2018,vanGompelEtAl2016}. In
+this case, it seems likely that the recommendations are ahead of common
+practice.
 
@@
 to this the measured data is not a true reflection of software package quality.
+
+Interviews with the developers revealed a potentially more frequent use of both
+unit testing and continuous integration, compared to what was observed from
+studying the repository artifacts.
 

### 2022-02-18 StateOfPractice/Papers/ICCS-2022-SOP_MethodologyPlusExamples/DigDeep.tex
Edits to Section 6.

@@
 \cite{HannayEtAl2009,Nguyen-HoanEtAl2010,PintoEtAl2018}, developer interviews
-\cite{SandersAndKelly2008} and case studies \cite{CarverEtAl2007,Segal2005}.  A
+\cite{Kelly2013} and case studies \cite{CarverEtAl2007,Segal2005}.  A
 valuable source of information that has received less attention is the data in
@@
 contrasts with the frequency with which continuous integration is recommended in
-research software development guidelines
-\cite{BrettEtAl2021,Brown2015,ThielEtAl2020,Zadka2018,vanGompelEtAl2016}. In
-this case, it seems likely that the recommendations are ahead of common
-practice.
+research software development
+guidelines~\cite{BrettEtAl2021,vanGompelEtAl2016,ThielEtAl2020}. In this case,
+it seems likely that the recommendations are ahead of common practice.
 
@@
 
-This section answers research question \rqref{RQ_CompareMethodologies} by
-comparing the principles, processes and methodologies used for LBM software to
-what can be gleaned from the literature on research software in general. In our
-data collection for LBM software, the software development process is not
-explicitly indicated in the artifacts for most of the packages. However, during
-our interviews one developer (ESPResSo) told us their non-rigorous development
-model is like a combination of agile and waterfall. Employing a loosely defined
-process makes sense for LBM software, given that the teams are generally small
-and self-contained.  Although eleven of the packages explicitly convey that they
-would accept outside contributors, generally the teams are centralized, often
-working at the same institution.  Working at the same institution means that an
-informal process can show success, since informal conversations are relatively
-easy to have.
-
-Interviews with developers confirmed a similar project management processes. In
-teams of only a couple of developers, additions of new features or major changes
-are discussed with the entire team. Projects with more than a couple developers
-have lead developer roles. These lead developers review potential additions to
-the software. One of the developers (ESPResSo) that was interviewed noted that
-an ad hoc peer review process is used to assess major changes and additions.
-Using peer review (also called technical review) matches with recommended
-practice for research software \cite{HerouxEtAl2008,Givler2020,OrvizEtAl2017,USGS2019}.
-
-Two types of software changes were discussed during interviews with developers.
-One is feature additions, which arise from a scientific or functional need.
-These changes involve formal discussions within the development team, and lead
-developer participation is mandatory. The other change type is code refactoring,
-which only sometimes involves formal discussions with the development team. New
-developers were noted to play an increased role in these changes compared to the
-former changes. Software bugs are typically addressed in a similar fashion as
-code refactoring.  Issue tracking is commonly used to manage these changes.
-
-Our observations of a informally defined process, with elements of agile
-methods, matches what has been observed for research software in general.
-Scientific developers naturally use an agile philosophy \cite{AckroydEtAl2008,CarverEtAl2007,EasterbrookAndJohns2009,Segal2005,HeatonAndCarver2015}, or an
-amethododical process \cite{Kelly2013}, or a knowledge acquisition driven
-process \cite{Kelly2015}.
-
-Most of the software packages do not explicitly state the motivations or design
-principles that were considered when developing the software. One package,
-Sailfish, indicates in its artifacts that shortening the development time was
-considered in early stages of design, with the developers using Python and
-CUDA/OpenCL to achieve this without sacrificing performance. The Sailfish goals
-are explicitly listed as performance, scalability, agility and extendability,
-maintenance, and ease of use. The project scored well in these categories during
-our assessment.  The quality priorities for Sailfish roughly match the
-priorities observed for research software in general.
-\cite{Nguyen-HoanEtAl2010} surveyed developers to find the following list of
-qualities, in decreasing order of importance: reliability, functionality,
-maintainability, availability, performance, flexibility, testability, usability,
-reusability, traceability, and portability. The Sailfish list does not list
-reliability or functionality, but we can safely assume those are implicitly high
-priorities for any scientific project.  In earlier studies
-\cite{KellyAndSanders2008} and \cite{CarverEtAl2007} both highlight how
-important correctness is for research software.
-
-During our interviews, documentation was noted as playing a significant role in
-the development process, specifically with on-boarding new developers. A goal of
-documentation is to lower the entry barrier for these new contributors. The
-documentation provides information on how to get started, orients the user to
-artifacts and the source code, and explains how the system works, including the
-so-called simulation engine and interface.  This emphasis on documentation,
-especially for new developers, is echoed in research software guidelines.
-Multiple guidelines recommend a document explaining how to contribute to a
-project, often named CONTRIBUTING \cite{Yo2021,BrettEtAl2021,WilsonEtAl2016,ThielEtAl2020,vanGompelEtAl2016,OrvizEtAl2017,FLOSS2022,JimenezEtAl2017}.
-Tutorials \cite{ThielEtAl2020}, trouble shooting guides \cite{OrvizEtAl2017,SSI2022} and quick start examples \cite{ThielEtAl2020,vanGompelEtAl2016} are
-also recommended.  \cite{SmithAndRoscoe2018} suggests including instructions
-specifically for on-boarding new developers. For open source software in general
-(not just research software), \cite{Fogel2005} recommends providing tutorial
-style examples, developer guidelines, demos and screenshots.
+The interview data on development processes is used to answer research question
+\rqref{RQ_CompareMethodologies}.  This data should be contrasted with the
+development process used by research software in general. The literature
+suggests that scientific developers naturally use an agile philosophy
+\cite{CarverEtAl2007,Segal2005}, or an amethododical process \cite{Kelly2013}.
+Another point of comparison should be on the use of peer review, since peer
+review is frequently recommended in software development
+guidelines~\cite{HerouxEtAl2008,OrvizEtAl2017,USGS2019}.
+
+The running example of LBM confirms an informal process, with elements of agile
+methods. The software development process is not explicitly indicated in the
+artifacts for most of the packages. However, during interviews one developer
+(ESPResSo) told us their non-rigorous development model is like a combination of
+agile and waterfall. Employing a loosely defined process makes sense for LBM
+software, given that the teams are generally small and self-contained. One of
+the developers (ESPResSo) that was interviewed also noted that an ad hoc peer
+review process is used to assess major changes and additions, matching the
+common recommendation.
 

### 2022-02-18 StateOfPractice/Papers/ICCS-2022-SOP_MethodologyPlusExamples/DigDeep.tex
Edits to Section 7 (Pain Points)

@@
 
-Ask developers during the interviews about their pain points.  Compare the identified pain points to the literature that discusses pain points \cite{Wieseet}.
-Based on interviews with 4 developers, this section aims to answer the research
-questions: i) What are the pain points for developers working on research
-software projects (\rqref{RQ_PainPoints})?; and, ii) How do the pain points of
-developers from LBM compare to the pain points for research software in general
-(\rqref{RQ_ComparePainPoints})?  Below we go through each of the identified pain
-points and include citations that contrast the LBM experience with observations
-from researchers in other domains.  Potential ways to address the pain points
-are covered in Section~\ref{Sec_AddressConcerns}. The full interview questions
-are found in \cite{SmithEtAl2021}.
-
-\cite{PintoEtAl2018} lists some pain points that did not come up in our
-conversations with LBM developers: Cross-platform compatibility, interruptions
+To answer \rqref{RQ_PainPoints}, we ask developers about their pain points and
+compare their responses to pain points identified in the
+literature~\cite{WieseEtAl2019,PintoEtAl2018}.  Pain points to watch for
+include: lack of development time, Cross-platform compatibility, interruptions
 while coding, scope bloat, lack of user feedback, hard to collaborate on
-software projects, and aloneness. \cite{WieseEtAl2019} repeat some of the
-previous pain points and add the following: dependency management, data handling
-concerns (like data quality, data management and data privacy), reproducibility,
-and software scope determination. Although LBM developers did not mention these
-pain points, we cannot conclude that they are not relevant for LBM software
-development, since we only interviewed 4 LBM developers for about an hour each.
+software projects, aloneness, dependency management, data handling concerns
+(like data quality, data management and data privacy), reproducibility, and
+software scope determination. For the LBM example, two sample pain points are:
 
@@
 	highlighted as a pain point by other research software developers
-	\cite{PintoEtAl2018,PintoEtAl2016,WieseEtAl2019}.
+	\cite{PintoEtAl2018,WieseEtAl2019}.
 
@@
 	between coding skills and subject-matter skills.
-
-	\item[P\refstepcounter{pnum}\thepnum \label{P_LackExtSupport}:] \textbf{Lack
-	of External Support} A concern was raised that there are no organizations
-	helping with the development of good quality software.  This concern is not
-	echoed in the literature because there are such organizations, including
-	\href{https://bssw.io/} {Better Scientific Software (BSSw)},
-	\href{https://www.software.ac.uk/} {Software Sustainability Institute}
-	\cite{CrouchEtAl2013}, and \href{https://software-carpentry.org/}{Software
-	Carpentry} [* {WilsonAndLumsdaine2006, Wilson2016} *]. Over time awareness of
-	these groups will grow, so this pain point is likely to disappear in the
-	future for LBM and other research software developers.
 

### 2022-02-18 StateOfPractice/Papers/ICCS-2022-SOP_MethodologyPlusExamples/DigDeep.tex
updated table 3

@@
 \midrule
-LICENSE & \checkmark &  & \checkmark & \checkmark & \checkmark & & \checkmark & \checkmark & \checkmark\\
-README &  &  &  &  & & & & \\
-CONTRIBUTING &  &  &  &  & & & & & \\
-CITATION &  &  &  &  & & & & & \\
-CHANGELOG &  &  &  &  & & & & & \\
-INSTALL &  &  &  &  & & & & & \\
-Uninstall &  &  &  &  & & & & & \\
-Contributor Guide &  &  &  &  & & & & & \\
-Getting started &  &  &  &  & & & & & \\
-User manual &  &  &  &  & & & & & \\
-Tutorials &  &  &  &  & & & & & \\
-FAQ &  &  &  &  & & & & & \\
-Dependency List &  &  &  & & & & & & \\
-Issue Track &  &  &  &  & & & & & \\
-Version Control &  &  &  &  & & & & &\\ 
-Requirements &  &  &  &  & & & & & \\
+LICENSE & \checkmark &  & \checkmark & \checkmark & \checkmark & & \checkmark & \checkmark & \checkmark \\
+README &  &  & \checkmark & \checkmark & \checkmark & & \checkmark & \checkmark & \checkmark\\
+CONTRIBUTING &  &  & \checkmark & \checkmark & \checkmark & & \checkmark & \checkmark & \checkmark \\
+CITATION &  &  &  & \checkmark & & & & \checkmark & \checkmark \\
+CHANGELOG &  &  &  & \checkmark & \checkmark & & \checkmark & & \\
+INSTALL &  &  &  &  & \checkmark & & \checkmark & \checkmark & \checkmark\\
+Uninstall &  &  &  &  & & & & \checkmark & \\
+Getting started &  &  &  &  & \checkmark & & \checkmark & \checkmark & \checkmark\\
+User manual &  &  & \checkmark &  & & & \checkmark & & \\
+Tutorials &  &  &  &  & & & \checkmark & & \\
+FAQ &  &  &  &  & & & \checkmark & \checkmark & \checkmark\\
+Dependency List &  &  & \checkmark & & \checkmark & & & \checkmark & \\
+Issue Track &  & \checkmark & \checkmark & & \checkmark & \checkmark & \checkmark & & \checkmark \\
+Version Control &  & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark\\ 
+Requirements &  & \checkmark &  & \checkmark & & \checkmark & \checkmark & \checkmark & \checkmark \\
 Design &  &  &  &  & & & & & \\
-API Doc. &  &  &  &  & & & & & \\
-Build Scripts &  &  &  &  & & & & & \\
-Unit Tests &  &  &  &  & & & & & \\
-Test Plan &  &  &  &  & & & & & \\
-Integ. Tests &  &  &  &  & & & & &  \\
-System Tests &  &  &  &  & & & & & \\
-Acceptance Tests &  &  &  &  & & & & &  \\
-Regression Tests &  &  &  &  & & & & & \\
-Code Style Guide &  &  &  &  & & & & & \\
-Release Info. &  &  &  &  & & & & & \\
-Product Roadmap &  &  &  &  & & & & & \\
-Code of Conduct &  &  &  &  & & & & & \\
+API Doc. &  &  &  &  & \checkmark & & \checkmark & \checkmark & \checkmark\\
+Build Scripts &  & \checkmark &  & \checkmark & \checkmark & \checkmark & \checkmark & & \checkmark \\
+Unit Tests & \checkmark & \checkmark & \checkmark &  & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
+Test Plan &  & \checkmark &  &  & & \checkmark & & & \\
+Integ. Tests &  & \checkmark & \checkmark &  & & & & \checkmark & \checkmark \\
+System Tests &  & \checkmark & \checkmark &  & & \checkmark & & \checkmark & \checkmark \\
+Acceptance Tests &  & \checkmark &  &  & & & & &  \\
+Regression Tests &  &  & \checkmark &  & & \checkmark & & & \checkmark \\
+Code Style Guide &  & \checkmark &  &  & & & \checkmark & \checkmark & \checkmark \\
+Release Info. &  & \checkmark &  &  & & \checkmark & \checkmark & & \\
+Product Roadmap &  &  &  &  & & \checkmark & \checkmark & \checkmark & \\
+Code of Conduct &  &  &  &  & & & \checkmark & & \\
 \midrule

### 2022-02-18 StateOfPractice/Papers/ICCS-2022-SOP_MethodologyPlusExamples/DigDeep.tex
Edits to paper

@@
 
-This section examines potential threats to the validity of this state of the
-practice assessment. These can be categorized into methodology and data
-collection issues. The goal of this assessment isn't to rank the software, but
-to use the ranking exercise as a means to understand the state of the practice
-of LBM software development.
+The measures in the measurement template~\cite{SmithEtAl2021} may not be broad
+enough to accurately capture some qualities. For example, there are only two
+measures of surface robustness. Similarly, reusability is assessed by the number
+of code files and LOC per file, assuming that a large number of relatively small
+files implies modularity. Furthermore, the measurement of understandability
+relies on 10 random source code files. It is possible that the 10 files that
+were chosen to represent a software package may not be a good representation.
 
-The measures listed in our measurement template may not be broad enough to
-accurately capture some qualities. For example, there are only two measures of
-surface robustness. The measurement of robustness could be expanded, as it
-currently only measures unexpected input. Other faults could be introduced, but
-could require a large investment of time to develop, and might not be a fair
-measure for all packages. Similarly, reusability is assessed along the number of
-code files and LOC per file. While this measure is indicative of modularity, it
-is possible that some packages have many files, with few LOC, but the files do
-not contain source code that is easily reusable. The files may be poorly
-formatted, or the source code may be vague and have ambiguous identifiers.
-Furthermore, the measurement of understandability relies on 10 random source
-code files. It is possible that the 10 files that were chosen to represent a
-software package may not be a good representation of the understandability of
-that package.
-
-Regarding data collection, a risk to the validity of this assessment is missing
-or incorrect data. Some software package data may not have been measured due to
-technology issues like broken links. This issue arose with the measurement of
-Palabos, which had a broken link to its user manual, as noted in
-Section~\ref{Sec_CompareArtifacts}.
-
-Some pertinent data may not have been specified in public artifacts, or may be
-obscure within an artifact or web-page. The use of unit testing and continuous
-integration was mentioned in the artifacts of only three (ESPResSo, Ludwig,
-Musubi) packages. However, interviews suggested a more frequent use of both unit
-testing and continuous integration in the development processes than what was
-observed from the initial survey of the artifacts. For example, OpenLB, pyLBM,
-and TCLB use such methods during development despite this not being explicitly
-clear from an analysis of the material available online.
-
-Furthermore, design documentation was measured to be a ``less common'' artifact
-in this assessment, but it is probable that such documentation is part of all
-LBM packages. After all, developing SCS is not a trivial endeavour. It is likely
-that many packages have such documentation but did not make it public, and due
-to this the measured data is not a true reflection of software package quality.
-
-Interviews with the developers revealed a potentially more frequent use of both
-unit testing and continuous integration, compared to what was observed from
-studying the repository artifacts.
-
-[*{There are more threats to validity.  Brainstorm.  Watch for them while 
-editing.  Look at Ao paper.}
-
-Problems with the measurement of qualities:
-
-\begin{itemize}
-\item The assumption that more code files is an indicator of reusability
-(Section~\ref{reusabilityresults}) may not be a valid assumption.
-\item Understandability was measured using only 10 random files
-(Section~\ref{Sec_SurfUnderstandability}).  This could be improved.
-\end{itemize}
-
-Overall ranking with equally weighted qualities isn't realistic
-(Section~\ref{Sec_OverallQuality}).
-
-Threat to validity - the gap between measures of everything else and Musubi
-(Section~\ref{identifysoftware}), and then the gap between the manual measures
-and the automated measures (Section~\ref{repmetrics}).
-
-From Section~\ref{Sec_CompareTools}: ``Interviews with the developers revealed a
-potentially more frequent use of both unit testing and continuous integration,
-compared to what was observed from studying the repository artifacts.''
-
-Interviews suggested a more frequent use of both unit testing and continuous
-integration in the development processes than what was observed from the initial
-survey. For example, OpenLB, pyLBM, and TCLB use such methods during
-development, despite this not being explicitly clear from an analysis of the
-material available online.  The correctness and verifiability of such packages
-is not measured well using surface analysis.
+Another risk to the validity of the proposed approach is missing or incorrect
+data. Some software package data may not have been measured due to technology
+issues like broken links. For the LBM example, this issue arose with the
+measurement of Palabos, which had a broken link to its user manual. Some
+pertinent data may not have been specified in public artifacts, or may be
+obscure within an artifact or web-page. For the LBM example, the use of unit
+testing and continuous integration was mentioned in the artifacts of only three
+(ESPResSo, Ludwig, Musubi) packages. However, interviews suggested a more
+frequent use of both unit testing and continuous integration in the development
+processes.
 
@@
 We have outlined a methodology for assessing the state of the practice for any
-given research software domain.  (Although the scope of the current work has been
-on research software, there is little in the methodology that is specific to
-research software, except for the interview question related to the quality of
-reproducibility.)  When applying the methodology to a given domain, we provide a
+given research software domain.  When applying the methodology to a given domain, we provide a
 means to answer the following questions:
@@
 
-The current methodology is constrained by limited resources.  A 4 hour cap on
-the measurement time for each software package limits what can be assessed.
-Within this limit, we can't measure some important qualities, like usability and
-modifiability.  In the future, we propose a more time-consuming process that
-would capture these other quality measures.  To improve the feasibility, the
-more time consuming measurements would not have to be completed for all 30
-packages. Instead, a short list could be identified using the output of the AHP
-ranking to select the top projects, or to select a sample of interesting
-projects across the quality spectrum.
-
-\subsubsection{Acknowledgements} Please place your acknowledgments at
-the end of the paper, preceded by an unnumbered run-in heading (i.e.
-3rd-level heading).
+The goal of this assessment isn't to rank the software, but to use the ranking
+exercise as a means to understand the state of the practice of LBM software
+development.
 

### 2022-02-18 StateOfPractice/Papers/ICCS-2022-SOP_MethodologyPlusExamples/DigDeep.tex
updated table 3, design

@@
 Requirements &  & \checkmark &  & \checkmark & & \checkmark & \checkmark & \checkmark & \checkmark \\
-Design &  &  &  &  & & & & & \\
+Design &  & \checkmark  & \checkmark &  & \checkmark & & \checkmark & \checkmark& \checkmark\\
 API Doc. &  &  &  &  & \checkmark & & \checkmark & \checkmark & \checkmark\\

### 2022-02-18 StateOfPractice/Papers/ICCS-2022-SOP_MethodologyPlusExamples/DigDeep.tex
First complete draft - 1.5 pages too long

@@
 
-	To improve software development methods and tools for research software, we
-	first need to understand the current state of the practice.  Therefore, we
-	have developed a methodology for assessing the state of the software
-	development practices for a given research software domain.  The methodology
-	is applied to one domain at a time in recognition that software development in
-	different domains is likely to have adopted different best practices.
-	Moreover, providing a means to measure different domains facilitates
-	comparison of development practices between domains.  For each domain we wish
-	to answer questions such as: 
-  \begin{inparaenum}[i)]
-    \item What artifacts (documents, code, test cases, etc.) are present?
-    \item What tools are used?
-    \item What principles, process and methodologies are used?
-    \item What are the pain points for developers?
-    \item What actions are used to improve qualities like maintainability and
-    reproducibility?
-  \end{inparaenum} 
-  To answer these questions, our methodology prescribes the following steps: 
-  The abstract should briefly summarize the contents of the paper in 150--250 words.
+	We have developed a methodology for assessing the state of the software
+	development practices for a given research software domain. Our methodology
+	prescribes the following steps: i) Identify the domain; ii) Identify a list
+	of candidate software packages; iii) Filter the list to a length of about 30
+	packages; iv) Collect repository related data on each software package, like
+	number of stars, number of open issues, number of lines of code; v) Fill in
+	the measurement template (the template consists of 108 questions to assess 9
+	qualities (including the qualities of installability, usability and
+	visibility)); vi) Interview developers (the interview consists of 20
+	questions and takes about an hour); and, vii) Rank the software using the
+	Analytic Hierarchy Process (AHP). The data is analyzed to: i) comparing the
+	ranking by best practices against an estimate of the communities ranking of
+	popularity; ii) comparing artifacts, tools and processes to current research
+	software development guidelines; and, iii) exploring pain points. We
+	estimate the time to complete an assessment for a given domain at 173 person
+	hours. We illustrate our method via the example of Lattice Boltzmann
+	Solvers, where we find that the top packages engaged in most of recommended
+	best practices, but show room for improvement with respect to providing API
+	documentation, a roadmap, a code of conduct, programming style guide,
+	uninstall instructions and continuous integration.
 
-\keywords{First keyword  \and Second keyword \and Another keyword.}
+\keywords{research software \and software quality \and empirical measures \and analytic hierarchy process \and software artifacts \and developer pain points \and Lattice Boltzmann Method}
+
 \end{abstract}
@@
 Close to our goal of 160 person hours, we estimate 173 hours to complete the
-assessment of a given domain~\cite{SmithEtAl2021}.  This estimate assumes that
-the domain has already been decided and the Domain Expert has been recruited.
-The estimate assumes that the template spreadsheets and developed AHP
-tool~\cite{SmithEtAl2021}, will be employed, rather than developing new tools.
-The person hours given are a rough estimate, based on our experience completing
-assessments for medical image analysis software \cite{Dong2021} and for Lattice
-Boltzmann Method (LBM) software \cite{Michalski2021}.  The estimate assumes 30
-software packages will be measured; the numbers will need to be adjusted if the
-total packages changes.
+assessment of a given domain~\cite{SmithEtAl2021}.  This estimate assumes the
+domain has been decided, the Domain Expert has been recruited, and the template
+spreadsheets and AHP tool~\cite{SmithEtAl2021}, will be utilized. The hours are
+a rough estimate, based on our experience completing assessments for medical
+image analysis~\cite{Dong2021} and the Lattice Boltzmann Method
+(LBM)~\cite{Michalski2021}.  The estimate should be adjusted if the number of
+packages measured differs from 30.
 
@@
 	preference for GitHub-style repositories.
-	\item The software cannot be marked as incomplete or in an initial state of
-	development.
+	\item The software cannot be marked as incomplete.
 \end{enumerate}
@@
 
-If the list of software is too long (over around 30 packages), then steps need
-to be taken to create a more manageable list to answer~\rqref{RQ_WhatProjects}.
-The following filters were applied in the priority order listed. Copies of both
-the initial and filtered lists, along with the rationale for shortening the
-list, should be kept for traceability purposes.
+If the list of software is too long (over around 30 packages)
+for~\rqref{RQ_WhatProjects}, then filters are applied in the priority order
+listed. Copies of both the initial and filtered lists, along with the rationale
+for shortening the list, should be kept for traceability purposes.
 
@@
 (like data quality, data management and data privacy), reproducibility, and
-software scope determination. For the LBM example, two sample pain points are:
+software scope determination.
 
-\begin{enumerate}
-
-	\item[P\refstepcounter{pnum}\thepnum \label{P_LackDevTime}:] \textbf{Lack of
-	Development Time} A developer of pyLBM noted that their small development
-	team has a lack of time to implement new features. Small development teams
-	are common for LBM software packages (as shown in the measurement table
-	excerpt in Figure~\ref{measurement_template_image}). Lack of time is also
-	highlighted as a pain point by other research software developers
-	\cite{PintoEtAl2018,WieseEtAl2019}.
-
-	\item[P\refstepcounter{pnum}\thepnum \label{P_LackSoftDevExp}:] \textbf{Lack
-	of Software Development Experience} A lack of software development
-	experience was noted by the developer of TCLB, and others noted a need for
-	improving software engineering education. Many of the team members on their
-	project are domain experts, not computer scientists or software engineers.
-	This same trend is noted by \cite{Nguyen-HoanEtAl2010}, which showed only
-	23\% of research software survey respondents having a computing-related
-	background. Similarly, \cite{UditAndKatz2017} show that the majority (54\%)
-	of postdocs have not received training in software development.  The LBM
-	developer suggesting an increasing role for formal software education
-	matches the trend observed by \cite{PintoEtAl2018}, where their replication
-	of a previous study \cite{HannayEtAl2009}, shows a growing interest in
-	formal training (From 13\% of respondents in 2009 to 22\% in 2018).
-	\cite{PintoEtAl2018} found that some developers feel there is a mismatch
-	between coding skills and subject-matter skills.
-
-\end{enumerate}
+An example pain point noted by LBM developers is a lack of development time. A
+developer of pyLBM noted that their small development team did not have enough
+time to implement new features. Small development teams are common for LBM
+software packages (as shown in the measurement table excerpt in
+Figure~\ref{measurement_template_image}).
 
@@
 
-Next step to recommend ways to improve software (previously section 9).  We wish
-to know what practices are used by the top domain projects, so that others can
-potentially emulate these practices. We also wish to identify new practices by
-borrowing successful ideas from other domains.
-
-Our comparison may point out areas where some LBM software
-packages fall short of current best practices. This is not intended to be a
-criticism of any existing packages, especially since in practice not every
-project needs to achieve the highest possible quality.  However, rather than
-delve into the nuances of which software can justify compromising which
-practices we will write our comparison under the ideal assumption that every
-project has sufficient resources to match best practices.
-
-For each domain we wish to highlight success stories that can be shared amongst
-domain community, and with the broader research software community, while at the
-same time watching for areas for potential future improvement.
+To improve the development of research software, both in terms of productivity,
+and the resulting software quality, we need to understand the current state of
+the practice. An exciting strategy to approach this goal is to assess one domain
+at a time, collecting data from developer interviews and digging deeply into
+their code repositories. By providing feedback specific to their domain, the
+developer community can be drawn into a dialogue on how to best make
+improvements going forward. Moreover, they can be encouraged to share their best
+practices between one another, and with other research software domains.
 
 We have outlined a methodology for assessing the state of the practice for any
-given research software domain.  When applying the methodology to a given domain, we provide a
-means to answer the following questions:
-\begin{inparaenum}[i)]
-\item What artifacts (documents, code, test cases, etc.) are present?
-\item What tools are used?
-\item What principles, process and methodologies are used?
-\item What are the pain points for developers?
-\item What actions are used to improve qualities like maintainability and
-reproducibility?
-\item What specific actions are taken to achieve the qualities of usability,
-traceability, modifiability, maintainability, correctness, understandability,
-unambiguity, reproducibility and visibility/transparency?
-\item How does software designated as high quality by this methodology compare
-  with top rated software by the community?
-\end{inparaenum}
+given research software domain based on a list of about 30 representative
+projects.  In addition to interviews, we use manual and automated inspection of
+the artifacts in each project's repositories.  The repository data is collected
+by filling in a 108 question measurement template, which requires installing the
+software on a VM, running simple tests (like completing the getting started
+instructions (if present)), and searching the code, documentation and test
+files. Using AHP the projects are ranked for 9 individual qualities
+(installability, correctness and verifiability, reliability, robustness,
+usability, maintainability, reusability, surface understandability,
+visibility/transparency) and for overall quality.  Perspective and insight is
+shared with the user community via the following: i) comparing the ranking by
+best practices against an estimate of the communities ranking of popularity; ii)
+comparing artifacts, tools and processes to current research software
+development guidelines; and, iii) exploring pain points via developer
+interviews. Using our methodology, spreadsheet templates and AHP tool, we
+estimate (based on our experience with using the process) the time to complete
+an assessment for a given domain at 173 person hours.
 
-The methodology depends on the engagement of a Domain Expert.  The Domain
-Expert's role is to ensure that the assessment is consistent with the culture of
-the community of practitioners in the domain.  The Domain Expert also has an
-important role to play with the domain analysis.  For each domain we 
-conduct a domain analysis to look at the commonalities, variabilities and
-parameters of variation, for the family of software in the domain.  The domain
-analysis means that software can be compared not just based on its quality, but
-also based on its functionality.
+For the running example of LBM we found that the top packages engaged in most of
+recommended best practices, including examples of practising peer review.
+However, we did find room for improvement with respect to providing API
+documentation, a roadmap, a code of conduct, programming style guide, and
+uninstall instructions.  In addition, the community could likely benefit by
+increased use of continuous integration.
 
-The methodology follows a systematic procedure that begins with identifying the
-domain and ends with answering the research questions posed above.  In between
-we collect an authoritative list of about 30 software packages.  For each
-package in the list we fill in our measurement template.  The template consists
-of repository related data (like number of open issues, number of lines of code,
-etc.) and 108 measures/questions related to 9 qualities: installability,
-correctness/verifiability, reliability, robustness, usability, maintainability,
-reusability, understandability and visibility/transparency. Filling in the
-template requires installing the software, running simple tests (like completing
-the getting started instructions (if present)), and searching the code,
-documentation and test files.
-
-The data for each domain is used to rank the software package according to each
-quality dimension using AHP.  The ranking is not intended to identify a single
-best software package.  Instead the ranking is intended to provide insights on
-the top set of software for each quality.  The top performers can be contrasted
-with the lesser performers to gain insight into what practices in the domain are
-working.  Deeper insight can be obtained by combining this data with the
-interview data from asking each recruited developer 20 questions.
-
-Combining the quantitative data from the measurement template with the interview
-results, along with the domain experts knowledge, we can determine the current
-state of the practice for domain X.  Using our methodology, spreadsheet
-templates and AHP tool, we estimate (based on our experience with using the
-process) the time to complete an assessment for a given domain at 173 person
-hours.
-
-With the wealth of data from assessing the state of practice for multiple
-domains, the next step is a meta-analysis.  We would look at how the different
-domains compare. What lessons from one domain could be applied in other domains?
-What (if any) differences exist in the pain points between domains?  Are there
-differences in the tools, processes, and documentation between domains?
-
-The goal of this assessment isn't to rank the software, but to use the ranking
-exercise as a means to understand the state of the practice of LBM software
-development.
+Building from the LBM example we can create a wealth of data on multiple
+domains.  The next step is a meta-analysis, where we look at how the different
+domains compare to answer new research questions like: What lessons from one
+domain could be applied in other domains? What (if any) differences exist in the
+pain points between domains?  Are there differences in the tools, processes, and
+documentation between domains?
 
@@
 % \end{thebibliography}
+
 \end{document}

### 2022-02-18 StateOfPractice/Papers/ICCS-2022-SOP_MethodologyPlusExamples/DigDeep.tex
Shortened paper to just over 14 pages.

@@
 	qualities (including the qualities of installability, usability and
-	visibility)); vi) Interview developers (the interview consists of 20
-	questions and takes about an hour); and, vii) Rank the software using the
-	Analytic Hierarchy Process (AHP). The data is analyzed to: i) comparing the
-	ranking by best practices against an estimate of the communities ranking of
-	popularity; ii) comparing artifacts, tools and processes to current research
-	software development guidelines; and, iii) exploring pain points. We
-	estimate the time to complete an assessment for a given domain at 173 person
-	hours. We illustrate our method via the example of Lattice Boltzmann
-	Solvers, where we find that the top packages engaged in most of recommended
-	best practices, but show room for improvement with respect to providing API
-	documentation, a roadmap, a code of conduct, programming style guide,
-	uninstall instructions and continuous integration.
+	visibility)); vi) Rank the software using the Analytic Hierarchy Process
+	(AHP); vii) Interview developers (the interview consists of 20 questions and
+	takes about an hour); and, viii) conduct a domain analysis. The collected
+	data is analyzed by: i) comparing the ranking by best practices against an
+	the ranking by popularity; ii) comparing artifacts, tools and processes to
+	current research software development guidelines; and, iii) exploring pain
+	points. We estimate the time to complete an assessment for a given domain at
+	173 person hours. The method is illustrated via the example of Lattice
+	Boltzmann Solvers, where we find that the top packages engaged in most of
+	recommended best practices, but still show room for improvement with respect
+	to providing API documentation, a roadmap, a code of conduct, programming
+	style guide, uninstall instructions and continuous integration.
 
@@
 all domains of research software, possibly distinguishing them by programming
-language (for instance, R developers \cite{PintoEtAl2018}), or by the role of
+language (for instance, R developers~\cite{PintoEtAl2018}), or by the role of
 the developers (for instance postdoctoral researchers \cite{UditAndKatz2017}).
@@
 or just a few examples, we will focus on one domain of research software at a
-time. The practical reason for defining a domain scope is that digging deep into
-repository data takes time, making a broad scope infeasible. We have determined
-that a practical constraint for the work load is for a team as small as one
-individual doing one person month worth of work per domain.\footnote{A person
+time. The practical reason for this scope is that digging deep into
+repository data takes time, making a broad scope infeasible. We have imposed
+a practical constraint of one person month worth of work per domain.\footnote{A person
 month is considered to be $20$ working days ($4$ weeks in a month, with $5$ days
@@
 
-Only so much information can be gleaned by digging into software repositories.
+Only so much information can be gleaned by digging into software repos.
 To gain additional insight, we need to interview developers
-(Section~\ref{SecSurvey}).  We need to learn their concerns and how they deal
-with these concerns; we need to learn their pain points.
+(Section~\ref{SecSurvey}) to learn:
 
@@
 domain has been decided, the Domain Expert has been recruited, and the template
-spreadsheets and AHP tool~\cite{SmithEtAl2021}, will be utilized. The hours are
-a rough estimate, based on our experience completing assessments for medical
-image analysis~\cite{Dong2021} and the Lattice Boltzmann Method
-(LBM)~\cite{Michalski2021}.  The estimate should be adjusted if the number of
-packages measured differs from 30.
+spreadsheets and AHP tool~\cite{SmithEtAl2021}, will be utilized.
 
@@
 
-A domain of research software must be identified to begin the assessment. To be
-applicable for the methodology described in this document, the chosen domain
-must have the following properties:
+To be applicable for the methodology described in this document, the chosen
+domain must have the following properties:
 
 \begin{enumerate}
-\item The domain must have well-defined and stable theoretical underpinning.  A
-  good sign of this is the existence of standard textbooks, preferably
-  understandable by an upper year undergraduate student.
+\item The domain must have well-defined and stable theoretical underpinning.
 \item There must be a community of people studying the domain.
@@
 \item A preliminary search, or discussion with experts, suggests that there will
-  be numerous, close to 30, candidate software packages in the domain that
-  qualify as `research software'.
+  be around 30 or more candidate software packages.
 \end{enumerate}
@@
 
-To answer~\rqref{RQ_WhatProjects} we needed to identify existing projects. The
-candidate software should be found through search engine queries targeting
-authoritative lists of software.  Candidate places to search include
-\href{https://github.com/} {GitHub} and \href{https://swmath.org/} {swMATH}, as
-well as through scholarly articles. The Domain Expert
-(Section~\ref{sec_vet_software_list}) should also be engaged in selecting the
-candidate software.  The following properties were considered when creating the
-list and reviewing the candidate software:
+The candidate software to answer~\rqref{RQ_WhatProjects} should be found through
+search engine queries,  
+\href{https://github.com/} {GitHub}, \href{https://swmath.org/} {swMATH}
+and scholarly articles. The Domain Expert (Section~\ref{sec_vet_software_list})
+should also be engaged in selecting the candidate software.  The following
+properties are considered when creating the list:
 
@@
 	\item The source code must be viewable.
-	\item The repository based measures should be available, which implies a
-	preference for GitHub-style repositories.
-	\item The software cannot be marked as incomplete.
+	\item The repository based measures should be available.
+	\item The software must be complete.
 \end{enumerate}
@@
 
-If the list of software is too long (over around 30 packages)
-for~\rqref{RQ_WhatProjects}, then filters are applied in the priority order
-listed. Copies of both the initial and filtered lists, along with the rationale
-for shortening the list, should be kept for traceability purposes.
+If the list of software is too long (over around 30 packages), then filters are
+applied in the priority order listed. Copies of both lists, along with the
+rationale for shortening the list, should be kept for traceability purposes.
 
@@
 For the LBM example the initial list had 46 packages~\cite{Michalski2021}. This
-list was filtered by scope, usage, and age to decrease the length to 24. Many of
-the removed packages could not be tested as there was no installation guide,
-they were incomplete, source code was not publicly available, a license was
-needed, or the project was out of scope or not up to a standard that would
-support incorporating them into this study~\cite{Michalski2021}.
+list was filtered by scope, usage, and age to decrease the length to 24.
 
@@
 \rqref{RQ_HighestQuality}.  To do this, we used the Analytical Hierarchy Process
-(AHP), a decision-making technique that is used to compare multiple options by
-multiple criteria \cite{Saaty1980}. AHP performs a pairwise analysis using a
-matrix and generates an overall score as well as individual quality scores for
-each software package. The advantage of pair-wise comparisons is they facilitate
-a separation of concerns.  Rather than worry about the entire problem, the
-decision maker can focus on one comparison at a time.  In our work AHP performs
-a pairwise analysis between each of the 9 quality options for each of the
-(approximately) 30 software packages.  This results in a matrix, which is used
-to generate an overall score for each software package for the given
+(AHP), a decision-making technique that uses pair-wise comparisons used to
+compare multiple options by multiple criteria \cite{Saaty1980}. In our work AHP
+performs a pairwise analysis between each of the 9 quality options for each of
+the (approximately) 30 software packages.  This results in a matrix, which is
+used to generate an overall score for each software package for the given
 criteria~\cite{SmithEtAl2016}.
@@
 packages. In the absence of other information on priorities, the overall ranking
-was calculated by assuming an equal weighting between all qualities. If the
-qualities were weighed differently, the overall software package ranking would
-change.
+was calculated by assuming an equal weighting between all qualities.
 
@@
 Domain experts may be recruited from academia or industry.  The only
-requirements are knowledge of the domain and a willingness to be engaged in the
-assessment process.  The Domain Expert does not have to be a software developer,
-but they should be a user of domain software.  Given that the domain experts are
-likely to be busy people, the measurement process cannot put to much of a burden
-on their time.
+requirements are knowledge of the domain and a willingness to be involved.
 
@@
 
-Since each domain we will study will have a reasonably small scope, we will be
-able to view the software as constituting a program family. Studying the common
-properties within a family of related programs is termed a domain analysis.  For
-the current methodology time constraints necessitate a shallow analysis. For
-each domain a table should be constructed that distinguishes the programs under
-study by their variabilities.  In research software the variabilities are often
-related to assumptions.  Table~\ref{tbl_features} shows the variabilities for
-the LBM software example~\cite{Michalski2021}.
+Since each domain will have a reasonably small scope, we will be able to view
+the software as a program family. Studying the common properties within a family
+via a domain analysis.  For the current methodology time constraints necessitate
+a shallow analysis. For each domain a table should be constructed that
+distinguishes the programs under study by their variabilities.  In research
+software the variabilities are often related to assumptions.
+Table~\ref{tbl_features} shows the variabilities for the LBM
+software~\cite{Michalski2021}.
 
@@
 
-Table~\ref{repometrics} compares the AHP ranking of the LBM package versus their
+Table~\ref{repometrics} compares the AHP ranking of the LBM package to their
 popularity in the research community.  Nine packages do not use GitHub, so they
@@
 packages. The best ranked package by AHP (ESPResSo) has the second most stars,
-while the ninth ranked package (Sailfish) has the highest number of stars. The
-same correlation is observed in the repository watch column, although this
-column contains less data, since two of the packages (Palabos, waLBerla) use
-GitLab, which does not track watches. Although the AHP ranking and the community
-popularity estimate are not perfect measures, they do suggest a correlation
-between best practices and popularity.
+while the ninth ranked package (Sailfish) has the highest number of stars.
+Although the AHP ranking and the community popularity estimate are not perfect
+measures, they do suggest a correlation between best practices and popularity.
 
@@
 			TCLB & 12 & 95 & 3 & 16& 3\\
-			MechSys & 13 & No Git & No Git & No Git& No Git\\
-			lettuce & 14 & 48 & 4 & 5& 8\\
-			ESPResSo++ & 15 & 35 & 5 & 12& 4\\
-			MP-LABS & 16 & 12 & 11 & 2& 9\\			
-			SunlightLB & 17 & No Git & No Git & No Git& No Git\\
-			LB3D & 18 & No Git & No Git & No Git& No Git\\			
-			LIMBES & 19 & No Git & No Git & No Git& No Git\\
-			LB2D-Prime & 20 & No Git & No Git & No Git& No Git\\		
-			HemeLB & 21 & 12 & 11 & 12& 4\\
-			lbmpy & 22 &  11 & 12 & 2& 9\\	
+			% MechSys & 13 & No Git & No Git & No Git& No Git\\
+			% lettuce & 14 & 48 & 4 & 5& 8\\
+			% ESPResSo++ & 15 & 35 & 5 & 12& 4\\
+			% MP-LABS & 16 & 12 & 11 & 2& 9\\			
+			% SunlightLB & 17 & No Git & No Git & No Git& No Git\\
+			% LB3D & 18 & No Git & No Git & No Git& No Git\\			
+			% LIMBES & 19 & No Git & No Git & No Git& No Git\\
+			% LB2D-Prime & 20 & No Git & No Git & No Git& No Git\\		
+			% HemeLB & 21 & 12 & 11 & 12& 4\\
+			% lbmpy & 22 &  11 & 12 & 2& 9\\	
+			... & ... & ... & ... & ... & ...\\	
 			LB3D-Prime & 23 & No Git & No Git & No Git& No Git\\	
@@
 		\end{tabular}
-		\caption{Repository Ranking Metrics} \label{repometrics}
+		\caption{Excerpt from Repository Ranking Metrics~\cite{Michalski2021}}
+		\label{repometrics}
 	\end{center}
@@
 
-In this step of the methodology we answer \rqref{RQ_CompareArtifacts} by
-comparing the artifacts that we observed in the domain repositories to those
-observed and recommended for research software in general.  While filling in the
-measurement template (Section~\ref{empiricalmeasures}), the domain software is
-examined for the presence of artifacts, which are then categorized by frequency.
-We suggest grouping them into the categories: common (more than 2/3 of
-projects), less common (between 1/3 and 2/3), and rare (less than 1/3 of
-projects). The observed frequency of artifacts should then be compared to the
-artifacts recommended by research software guidelines, as summarized in
-Table~\ref{Tbl_Guidelines}.
+We answer \rqref{RQ_CompareArtifacts} by comparing the artifacts that we observe
+to those observed and recommended for research software in general.  While
+filling in the measurement template (Section~\ref{empiricalmeasures}), the
+domain software is examined for the presence of artifacts, which are then
+categorized by frequency as: common (more than 2/3 of projects), less common
+(between 1/3 and 2/3), and rare (less than 1/3). The observed frequency of
+artifacts should then be compared to the artifacts recommended by research
+software guidelines, as summarized in Table~\ref{Tbl_Guidelines}.
 
@@
 compared to research software norms: version control and continuous integration.
-The poor adoption of version control tools that Wilson lamented in 2006
-\cite{Wilson2006} has greatly improved in the intervening years.  A little over
-10 years ago version control was estimated to be used in only 50\% of research
-software projects \cite{Nguyen-HoanEtAl2010}, but even at that time
-\cite{Nguyen-HoanEtAl2010} noted an increase from previous usage levels. A
-survey in 2018 shows 81\% of developers use a version control system
-\cite{AlNoamanyAndBorghi2018}.  \cite{Smith2018} has similar results, showing
-version control usage for alive projects in mesh generation, geographic
-information systems and statistical software for psychiatry increasing from
-75\%, 89\% and 17\% (respectively) to 100\%, 95\% and 100\% (respectively) over
-a four year period ending in 2018.  (For completeness the same study showed a
-decrease in version control usage for seismology software over the same time
-period, from 41\% down to 36\%).  Almost every software guide cited in
+A little over 10 years ago version control was estimated to be used in only 50\%
+of research software projects \cite{Nguyen-HoanEtAl2010}, but even at that time
+\cite{Nguyen-HoanEtAl2010} noted an increase from previous usage levels. More
+recently, version control usage rates for active mesh generation, geographic
+information system and statistical software packages were close to
+100~\cite{Smith2018}.  Almost every software guide cited in
 Section~\ref{Sec_CompareArtifacts} includes the advice to use version control.
@@
 To answer \rqref{RQ_PainPoints}, we ask developers about their pain points and
-compare their responses to pain points identified in the
+then compare their responses to pain points identified in the
 literature~\cite{WieseEtAl2019,PintoEtAl2018}.  Pain points to watch for
-include: lack of development time, Cross-platform compatibility, interruptions
-while coding, scope bloat, lack of user feedback, hard to collaborate on
-software projects, aloneness, dependency management, data handling concerns
-(like data quality, data management and data privacy), reproducibility, and
-software scope determination.
+include: lack of development time, cross-platform compatibility, scope bloat,
+lack of user feedback, dependency management, data handling concerns,
+reproducibility, and software scope determination.
 
@@
 time to implement new features. Small development teams are common for LBM
-software packages (as shown in the measurement table excerpt in
-Figure~\ref{measurement_template_image}).
+software packages (as shown in Figure~\ref{measurement_template_image}).
 

### 2022-02-18 StateOfPractice/Papers/ICCS-2022-SOP_MethodologyPlusExamples/DigDeep.tex
Paper shortened to 14 pages

@@
 Virtual machines (VMs) are used to provide an optimal testing environments for
-each package~\cite{SmithEtAl2016}. VMs are used because it is easier to start
-with a fresh environment so as to not worry about existing libraries and
-conflicts. Moreover, when the tests are complete the VM can be deleted, without
-any impact on the host operating system. The most significant advantage of using
-VMs is to level the playing field. Every software install starts from a clean
-slate, which removes ``works-on-my-computer'' errors.
+each package~\cite{SmithEtAl2016} because with a fresh VM there are no worries
+about conflicts with existing libraries. Moreover, when the tests are complete
+the VM can be deleted, without any impact on the host operating system. The most
+significant advantage of using VMs is that every software install starts from a
+clean slate, which removes ``works-on-my-computer'' errors.
 
@@
 teams, the interviewees, and the software itself. We ask the developers how they
-organize their projects. We also ask them about their understanding of the
-users. Some questions focus on the current and past difficulties, and the
-solutions the team has found, or will try. We also discuss the importance of,
-and the current situation for, documentation. A few questions are about specific
-software qualities, such as maintainability, understandability, usability, and
-reproducibility. The interviews are semi-structured based on the question list;
-we ask follow-up questions when necessary.  Each interview should take about 1
-hour.
+organize their projects and about their understanding of the users. Some
+questions focus on the current and past difficulties, and the solutions the team
+has found, or will try. We also discuss the importance of, and the current
+situation for, documentation. A few questions are about specific software
+qualities, such as maintainability, usability, and reproducibility. The
+interviews are semi-structured based on the question list.  Each interview
+should take about 1 hour.
 
@@
 
-Since each domain will have a reasonably small scope, we will be able to view
-the software as a program family. Studying the common properties within a family
-via a domain analysis.  For the current methodology time constraints necessitate
-a shallow analysis. For each domain a table should be constructed that
-distinguishes the programs under study by their variabilities.  In research
-software the variabilities are often related to assumptions.
-Table~\ref{tbl_features} shows the variabilities for the LBM
-software~\cite{Michalski2021}.
+For the current methodology time constraints necessitate a shallow domain
+analysis. For each domain a table should be constructed that distinguishes the
+programs under study by their variabilities.  In research software the
+variabilities are often assumptions. Table~\ref{tbl_features} shows the
+variabilities for the LBM software~\cite{Michalski2021}.
 
@@
 developers.  For LBM, a union of the three categories mostly corresponds to
-recommendations made by the research software community, as shown in
-Table~\ref{Tbl_Guidelines}. Although the LBM community participates in most of
-the practices we found listed in the general research software guidelines, some
-recommended practices were not observed or rarely observed as follows: API
-documentation, a roadmap, a code of conduct, programming style guide, uninstall
-instructions.
+general research software recommendations (Table~\ref{Tbl_Guidelines}). Areas
+where LBM developers could improve include providing: API documentation, a
+roadmap, a code of conduct, programming style guide, and uninstall instructions.
 
@@
 
-The running example of LBM confirms an informal process, with elements of agile
-methods. The software development process is not explicitly indicated in the
-artifacts for most of the packages. However, during interviews one developer
-(ESPResSo) told us their non-rigorous development model is like a combination of
-agile and waterfall. Employing a loosely defined process makes sense for LBM
-software, given that the teams are generally small and self-contained. One of
-the developers (ESPResSo) that was interviewed also noted that an ad hoc peer
-review process is used to assess major changes and additions, matching the
-common recommendation.
+The LBM example confirms an informal, agile-like, process.
+The development process is not explicitly indicated in the artifacts
+for most of the packages. However, during interviews one developer (ESPResSo)
+told us their non-rigorous development model is like a combination of agile and
+waterfall. Employing a loosely defined process makes sense for LBM software,
+given that the teams are generally small and self-contained. One of the
+developers (ESPResSo) that was interviewed also noted that an ad hoc peer review
+process is used.
 
@@
 To answer \rqref{RQ_PainPoints}, we ask developers about their pain points and
-then compare their responses to pain points identified in the
-literature~\cite{WieseEtAl2019,PintoEtAl2018}.  Pain points to watch for
-include: lack of development time, cross-platform compatibility, scope bloat,
+compare their responses to the literature~\cite{WieseEtAl2019,PintoEtAl2018}.
+Pain points to watch for include: cross-platform compatibility, scope bloat,
 lack of user feedback, dependency management, data handling concerns,
-reproducibility, and software scope determination.
+reproducibility, and software scope.
 
-An example pain point noted by LBM developers is a lack of development time. A
-developer of pyLBM noted that their small development team did not have enough
-time to implement new features. Small development teams are common for LBM
-software packages (as shown in Figure~\ref{measurement_template_image}).
+An example pain point noted for LBM is a lack of development time. A developer
+of pyLBM noted that their small development team did not have enough time to
+implement new features. Small development teams are common for LBM software
+packages (as shown in Figure~\ref{measurement_template_image}).
 

### 2022-02-18 StateOfPractice/Papers/ICCS-2022-SOP_MethodologyPlusExamples/DigDeep.tex
Added mention of publication of data and a link to LBM data (now over 14 pages)

@@
 process takes about 2 hours per package, with a cap of 4 hours.  An excerpt of
-the template is shown in Figure~\ref{measurement_template_image}.
+the template is shown in Figure~\ref{measurement_template_image}.  In keeping
+with scientific transparency, all data should be made available in a public
+repository.
 
@@
 was calculated by assuming an equal weighting between all qualities.
+The LBM data is available \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/DomainMeasurements/LBM-Peter/Completed_LBS_Combined_MeasurementTemplate_EmpiricalMeasures_ADDED_MUSUBI_UPDATED_STATS.xlsx} {on-line}.
 

### 2022-02-18 StateOfPractice/Papers/ICCS-2022-SOP_MethodologyPlusExamples/DigDeep.tex
Revisions to abstract

@@
 	We have developed a methodology for assessing the state of the software
-	development practices for a given research software domain. Our methodology
+	development practice for a given research software domain. Our methodology
 	prescribes the following steps: i) Identify the domain; ii) Identify a list
@@
 	(AHP); vii) Interview developers (the interview consists of 20 questions and
-	takes about an hour); and, viii) conduct a domain analysis. The collected
-	data is analyzed by: i) comparing the ranking by best practices against an
+	takes about an hour); and, viii) Conduct a domain analysis. The collected
+	data is analyzed by: i) comparing the ranking by best practices against 
 	the ranking by popularity; ii) comparing artifacts, tools and processes to

### 2022-02-18 StateOfPractice/Papers/ICCS-2022-SOP_MethodologyPlusExamples/DigDeep.tex
Modified introduction section

@@
 scope between these two extremes.  Rather than focus on all research software,
-or just a few examples, we will focus on one domain of research software at a
-time. The practical reason for this scope is that digging deep into
-repository data takes time, making a broad scope infeasible. We have imposed
-a practical constraint of one person month worth of work per domain.\footnote{A person
-month is considered to be $20$ working days ($4$ weeks in a month, with $5$ days
-of work per week) at $8$ person hours per day, or $20 \cdot 8 = 160$ person
-hours.} Focusing on one domain at a time has more than just practical
-advantages.  By restricting ourselves to a single domain we can bring domain
-knowledge and domain experts into the mix.  The domain customized insight
-provided by the assessment has the potential to help a specific domain as they
-adopt and develop new development practices.  Moreover, measuring different
-domains facilitates uncovering domain specific practices, which can later be
-compared and contrasted.
+or just a few examples, we will focus on one domain at a time. The practical
+reason for this scope is that digging deep takes time, making a broad scope
+infeasible. We have imposed a practical constraint of one person month of effort
+per domain.\footnote{A person month is considered to be $20$ working days ($4$
+weeks in a month, with $5$ days of work per week) at $8$ person hours per day,
+or $20 \cdot 8 = 160$ person hours.} Focusing on one domain at a time has more
+than just practical advantages.  By restricting ourselves to a single domain we
+can bring domain knowledge and domain experts into the mix.  The domain
+customized insight provided by the assessment has the potential to help a
+specific domain as they adopt and develop new practices.  Moreover, measuring
+multiple different domains facilitates comparing and contrasting domain specific
+practices.
 
@@
 paper ends with a summary of potential threats to validity
-(Section~\ref{threats}) and our final conclusions and recommendations
+(Section~\ref{threats}) and our conclusions
 (Section~\ref{SecConcludingRemarks}).

### 2022-02-18 StateOfPractice/Papers/ICCS-2022-SOP_MethodologyPlusExamples/DigDeep.tex
Revisions to intro blurb for Methodology section

@@
 	(Section~\ref{empiricalmeasures})
-	\item Measure using the measurement template.  The full measurement template
-	can be found in \cite{SmithEtAl2021}. (Section~\ref{empiricalmeasures})
+	\item Measure using the measurement template. (Section~\ref{empiricalmeasures})
 	\item Use AHP to rank the software packages. (Section~\ref{empiricalmeasures})
@@
 
-Close to our goal of 160 person hours, we estimate 173 hours to complete the
-assessment of a given domain~\cite{SmithEtAl2021}.  This estimate assumes the
-domain has been decided, the Domain Expert has been recruited, and the template
-spreadsheets and AHP tool~\cite{SmithEtAl2021}, will be utilized.
+We estimate 173 hours to complete the assessment of a given
+domain~\cite{SmithEtAl2021}, which is close our goal of 160 person hours. This
+estimate assumes the domain has been decided, the Domain Expert has been
+recruited, and the pre-existing template spreadsheet and AHP
+tool~\cite{SmithEtAl2021}, will be utilized.
 

### 2022-02-18 StateOfPractice/Papers/ICCS-2022-SOP_MethodologyPlusExamples/DigDeep.tex
Edits to How to Filter the Software List?

@@
 	considered to be within the scope of the domain.
-	\item Usage: Software packages were eliminated if their installation
-	procedure was missing or not clear and easy to follow.
-	\item Age: The older software packages (age being measured by the last date
-	when a change was made) were eliminated, except in the cases where an older
+	\item Usage: Software packages are eliminated if their installation
+	procedure is missing or not clear and easy to follow.
+	\item Age: Older software packages (age being measured by the last date
+	when a change was made) are eliminated, except in the cases where an older
 	software package appears to be highly recommended and currently in use. (The

### 2022-02-18 StateOfPractice/Papers/ICCS-2022-SOP_MethodologyPlusExamples/DigDeep.tex
Revisions to Section 2.4 Quantitative Measures

@@
 Several of the qualities use the word ``surface''.  This is to highlight that,
-for these qualities in particular, the best that we can do is a shallow measure.
-For instance, we do not conduct experiments to measure usability. Instead, we
-are looking for an indication that usability was considered by the developers by
-looking for cues in the documentation, such as getting started instructions, a
-user manual or documentation of expected user characteristics.
+for these qualities, the best that we can do is a shallow measure. For instance,
+we do not conduct experiments to measure usability. Instead, we are looking for
+an indication that usability was considered by looking for cues in the
+documentation, such as getting started instructions, a user manual or
+documentation of expected user characteristics.
 
@@
 \rqref{RQ_HighestQuality}.  To do this, we used the Analytical Hierarchy Process
-(AHP), a decision-making technique that uses pair-wise comparisons used to
+(AHP), a decision-making technique that uses pair-wise comparisons to
 compare multiple options by multiple criteria \cite{Saaty1980}. In our work AHP

### 2022-02-18 StateOfPractice/Papers/ICCS-2022-SOP_MethodologyPlusExamples/DigDeep.tex
Revisions to Section 2.5 "Interview Developers"

@@
 
-The interviewees should follow a process where they can make informed consent.
-The interviews should follow standard ethics guideline of asking for consent
-before interviewing, recording, and including participant details in the report.
-The interview process presented here was approved by the McMaster University
-Research Ethics Board under the application number 
+The interviewees should follow standard ethics guidelines for consent,
+recording, and including participant details in the report. The interview
+process presented here was approved by the McMaster University Research Ethics
+Board under the application number 
 \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/MACREM/Application.pdf}

### 2022-02-18 StateOfPractice/Papers/ICCS-2022-SOP_MethodologyPlusExamples/DigDeep.tex
Revisions to Section 2.6: Interaction with Domain Expert
Revisions to Section 2.7: Domain Analysis

@@
 team. Pitfalls exist if non-experts attempt to acquire an authoritative list of
-software, or try to definitively rank the software. Non-experts have the problem
+software, or try to definitively rank software. Non-experts have the problem
 that they can only rely on information available on-line, which has the
@@
 variabilities are often assumptions. Table~\ref{tbl_features} shows the
-variabilities for the LBM software~\cite{Michalski2021}.
+variabilities for LBM software~\cite{Michalski2021}.
 

### 2022-02-18 StateOfPractice/Papers/ICCS-2022-SOP_MethodologyPlusExamples/DigDeep.tex
Revisions to Section 3: Comparison to Community Ranking

@@
 To address \rqref{RQ_CompareHQ2Popular} we need to compare the ranking by best
-practices to the communities ranking.  Our best practices ranking comes from the
-results of the AHP ranking (Section~\ref{empiricalmeasures}).  We estimate the
-communities ranking by repository stars and watches.  The comparison will
-provide insight on whether best practices are rewarded by popularity.  However,
-inconsistencies between the AHP ranking and the communities ranking are in
-inevitable for the following reasons: 
+practices to the community's ranking.  Our best practices ranking comes from the
+AHP ranking (Section~\ref{empiricalmeasures}).  We estimate the community's
+ranking by repository stars and watches.  The comparison will provide insight on
+whether best practices are rewarded by popularity.  However, inconsistencies
+between the AHP ranking and the communities ranking are inevitable for the
+following reasons: 
 \begin{inparaenum}[i)]
 	\item the overall quality ranking via AHP makes the unrealistic assumption
-	of equal weighting between the different quality factors;
+	of equal weighting between quality factors;
 	\item stars are known to not be a particularly good measure of popularity,
-	because of how people use stars and since young projects have less time to
+	because of how people use stars and because young projects have less time to
 	accumulate stars~\cite{Szulik2017}; 
-	\item and, as for consumer products, there are more factors that determine
-	popularity than quality alone.
+	\item and, as for consumer products, there are more factors influencing
+	popularity than just quality.
 \end{inparaenum}
@@
 popularity in the research community.  Nine packages do not use GitHub, so they
-do not have a measure of repository stars. Looking at the repository stars of
-the other 15 packages, we can observe a pattern where packages that have been
-highly ranked by our assessment tend to have more stars than lower ranked
-packages. The best ranked package by AHP (ESPResSo) has the second most stars,
-while the ninth ranked package (Sailfish) has the highest number of stars.
-Although the AHP ranking and the community popularity estimate are not perfect
-measures, they do suggest a correlation between best practices and popularity.
+do not have a measure of stars. Looking at the stars of the other 15 packages,
+we can observe a pattern where packages that have been highly ranked by our
+assessment tend to have more stars than lower ranked packages. The best ranked
+package by AHP (ESPResSo) has the second most stars, while the ninth ranked
+package (Sailfish) has the highest number of stars. Although the AHP ranking and
+the community popularity estimate are not perfect measures, they do suggest a
+correlation between best practices and popularity.
 
@@
 			% lbmpy & 22 &  11 & 12 & 2& 9\\	
+			% LB3D-Prime & 23 & No Git & No Git & No Git& No Git\\	
 			... & ... & ... & ... & ... & ...\\				
-			LB3D-Prime & 23 & No Git & No Git & No Git& No Git\\	
 			LatBo.jl & 24 & 17 & 10 & 8& 6\\			

### 2022-02-18 StateOfPractice/Papers/ICCS-2022-SOP_MethodologyPlusExamples/DigDeep.tex
Revisions to Section 4: Comparison of Artifacts to Other Research Software
Revisions to Section 5: Comparison of Tools to Other Research Software

@@
 
-The results for the LBM example are shown below.  The majority of LBM generated
-artifacts correspond to general recommendations from research software
-developers.  For LBM, a union of the three categories mostly corresponds to
-general research software recommendations (Table~\ref{Tbl_Guidelines}). Areas
-where LBM developers could improve include providing: API documentation, a
-roadmap, a code of conduct, programming style guide, and uninstall instructions.
+The majority of LBM generated artifacts (summarized below) correspond to general
+recommendations from research software developers.  For LBM, a union of the
+three categories mostly corresponds to general research software recommendations
+(Table~\ref{Tbl_Guidelines}). Areas where LBM developers could improve include
+providing: API documentation, a roadmap, a code of conduct, programming style
+guide, and uninstall instructions.
 
@@
 
-% \begin{table}[h!]
-% \begin{center}
-% \begin{tabular}{ p{12 cm}}
-% %\toprule
-% \textbf{Common}\\
-% \midrule
-% Developer List, Issue Tracker, Dependency List, Installation Guide, Theory
-% Notes, Related Publications, Build Files, README File, License, Tutorial,
-% Version Control\\
-% %\midrule
-% \textbf{Less Common}\\
-% \midrule
-% Change Log, Design Doc., Functional Spec., Performance Info., Test Cases, User
-% Manual\\
-% %\midrule
-% \textbf{Rare}\\
-% \midrule
-% API Doc., Developer Manual, FAQ, Verification Plan, Video Guide, Requirements
-% Spec.\\
-% %\bottomrule
-% \end{tabular}
-% \caption{Artifacts Present in LBM Packages, Classified by Frequency}
-% \label{artifactspresent}
-% \end{center}
-% \end{table}
-
 \section{Comparison of Tools to Other Research Software}
@@
 
-To answer \rqref{RQ_CompareToolsProjMngmnt} we need to summarize tools that are
-visible in the repositories and that are mentioned during the developer
-interviews. Software tools are used to support the development, verification,
-maintenance, and evolution of software, software processes, and
-artifacts~\cite[p.\ 501]{GhezziEtAl2003}. Development tools support the
-development of end products, but do not become part of them, unlike dependencies
-that remain in the application once it is released \cite[p.\
-506]{GhezziEtAl2003}. As an example, the tools found for LBM software packages
-are as follows:
+Software tools are used to support the development, verification, maintenance,
+and evolution of software, software processes, and artifacts~\cite[p.\
+501]{GhezziEtAl2003}. Development tools support the development of end products,
+but do not become part of them, unlike dependencies that remain in the
+application once it is released \cite[p.\ 506]{GhezziEtAl2003}. To answer
+\rqref{RQ_CompareToolsProjMngmnt} we summarize tools that are visible in the
+repositories and that are mentioned during the developer interviews.  As an
+example, the tools found for LBM software packages are as follows:
 
@@
 information system and statistical software packages were close to
-100~\cite{Smith2018}.  Almost every software guide cited in
-Section~\ref{Sec_CompareArtifacts} includes the advice to use version control.
-For LBM packages 67\% of LBM packages use version control (GitHub, GitLab or
-CVS). The high usage of version control tools in LBM software matches the trend
-in research software in general.
+100\%~\cite{Smith2018}.  Almost every software guide cited in
+Table~\ref{Tbl_Guidelines} includes the advice to use version control. For LBM
+packages 67\% use version control (GitHub, GitLab or CVS). The high usage of
+version control tools in LBM software matches the trend in research software in
+general.
 

### 2022-02-18 StateOfPractice/Papers/ICCS-2022-SOP_MethodologyPlusExamples/DigDeep.tex
Revisions to Section 6: Comparison of Processes to Other Research Software

@@
 Another point of comparison should be on the use of peer review, since peer
-review is frequently recommended in software development
-guidelines~\cite{HerouxEtAl2008,OrvizEtAl2017,USGS2019}.
+review is frequently recommended~\cite{HerouxEtAl2008,OrvizEtAl2017,USGS2019}.
 
-The LBM example confirms an informal, agile-like, process.
-The development process is not explicitly indicated in the artifacts
-for most of the packages. However, during interviews one developer (ESPResSo)
-told us their non-rigorous development model is like a combination of agile and
-waterfall. Employing a loosely defined process makes sense for LBM software,
-given that the teams are generally small and self-contained. One of the
-developers (ESPResSo) that was interviewed also noted that an ad hoc peer review
-process is used.
+The LBM example confirms an informal, agile-like, process. The development
+process is not explicitly indicated in the artifacts. However, during interviews
+one developer (ESPResSo) told us their non-rigorous development model is like a
+combination of agile and waterfall. Employing a loosely defined process makes
+sense for LBM software, given that the teams are generally small and
+self-contained. One of the developers (ESPResSo) also noted that they use an ad
+hoc peer review process.
 

### 2022-02-18 StateOfPractice/Papers/ICCS-2022-SOP_MethodologyPlusExamples/DigDeep.tex
Revisions to Section 7: Threats To Validity

@@
 relies on 10 random source code files. It is possible that the 10 files that
-were chosen to represent a software package may not be a good representation.
+were chosen to represent a software package may not be representative.
 
 Another risk to the validity of the proposed approach is missing or incorrect
-data. Some software package data may not have been measured due to technology
+data. Some software package data may have been missed due to technology
 issues like broken links. For the LBM example, this issue arose with the

### 2022-02-18 StateOfPractice/Papers/ICCS-2022-SOP_MethodologyPlusExamples/DigDeep.tex
Revisions to Conclusions section, fixed communities to community's.  Done for now.

@@
 whether best practices are rewarded by popularity.  However, inconsistencies
-between the AHP ranking and the communities ranking are inevitable for the
+between the AHP ranking and the community's ranking are inevitable for the
 following reasons: 
@@
 shared with the user community via the following: i) comparing the ranking by
-best practices against an estimate of the communities ranking of popularity; ii)
+best practices against an estimate of the community's ranking of popularity; ii)
 comparing artifacts, tools and processes to current research software

### 2022-02-18 StateOfPractice/Papers/ICCS-2022-SOP_MethodologyPlusExamples/DigDeep.tex
Changed link to data to Memdeley dataset

@@
 packages. In the absence of other information on priorities, the overall ranking
-was calculated by assuming an equal weighting between all qualities.
-The LBM data is available \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/DomainMeasurements/LBM-Peter/Completed_LBS_Combined_MeasurementTemplate_EmpiricalMeasures_ADDED_MUSUBI_UPDATED_STATS.xlsx} {on-line}.
+was calculated by assuming an equal weighting between all qualities.  The LBM data is available \href{https://data.mendeley.com/datasets/5dym63wn6z/1} {on-line}.
 

### 2022-02-18 StateOfPractice/Papers/ICCS-2022-SOP_MethodologyPlusExamples/DigDeep.tex
The version submitted to ICCS on Feb 18, 2022, around 3:30 pm

@@
-% This is samplepaper.tex, a sample chapter demonstrating the
-% LLNCS macro package for Springer Computer Science proceedings;
-% Version 2.21 of 2022/01/12
-%
 \documentclass[runningheads]{llncs}
@@
 
-\begin{table}[!h]
+\begin{table}[!ht]
 	\begin{center}
@@
 
-\begin{table}[!h]
+\begin{table}[!ht]
 \begin{center}
@@
 Version Control &  & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark\\ 
-Requirements &  & \checkmark &  & \checkmark & & \checkmark & \checkmark & \checkmark & \checkmark \\
+Requirements &  & \checkmark &  &  & & \checkmark &  &  & \checkmark \\
 Design &  & \checkmark  & \checkmark &  & \checkmark & & \checkmark & \checkmark& \checkmark\\

### 2022-02-25 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Minor edits throughout, sections 1 to 8 are mostly done

@@
 \begin{keyword}
-	Lattice Boltzmann Method (LBM), scientific computing, software engineering,
+	Lattice Boltzmann Method (LBM), research software, software engineering,
 	software quality, Analytic Hierarchy Process
@@
 the development practices employed for research software in general.  We want to
-highlight success stories that LBM developers can share amongst their
-community, and with the broader research software community, while at the same time watching for areas for potential future improvement.
+highlight success stories that LBM developers can share amongst their community,
+and with the broader research software community, while at the same time
+looking for areas for potential future improvement.
 
@@
 	practices, based on evidence found by experimenting with the software and
-	searching the artifacts available in each project's repository? (Section~\ref{AHPresults})
+	searching the artifacts available in each project's repository?
+	(Section~\ref{AHPresults})
 	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareHQ2Popular}:] How
 	similar is the list of top projects identified in \rqref{RQ_HighestQuality}
-	to the most popular projects, as viewed by the scientific community? (Section~\ref{repmetrics})
+	to the most popular projects, as viewed by the scientific community?
+	(Section~\ref{repmetrics})
 \end{enumerate}
@@
 	How do LBM projects compare to research software in general with respect to
-	principles, processes and methodologies used? (Section~\ref{Sec_CompareMethodologies})
+	principles, processes and methodologies used?
+	(Section~\ref{Sec_CompareMethodologies})
 \end{enumerate}
@@
 projects, so that others can potentially emulate these practices. We also wish
-to identify new practices by borrowing successful ideas from other domains. For
-LBM developers interested in future cutting edge practices, ideas that are
-proposed, but not yet in practice, should also be identified.  The above points
-are covered by the questions outlined below:
+to identify new practices that LBM developers can adopt to improve their
+software in the future. The above points are covered by the questions outlined
+below:
 
@@
 We developed a methodology for evaluating the state of the practice of
-scientific software \citep{SmithEtAl2021}. This methodology can be instantiated
-for a specific domain of scientific software, which in the current case is LBM
+research software \citep{SmithEtAl2021}. This methodology can be instantiated
+for a specific domain of research software, which in the current case is LBM
 software.  Since we group best practices around conventional software qualities,
@@
 
-We adopt software quality definitions from various researchers and subject
-matter experts. The following are the software quality definitions used in this
-exercise, along with comments regarding their measurement.
+The following are the software quality definitions used in this exercise, along
+with comments regarding their measurement.
 
@@
 	the installation, uninstallation or reinstallation of a software product in
-	a specified environment \citep{ISO/IEC25010, lenhard2013measuring}. A good
-	measure of installability correlates with scenarios when low or moderate
-	effort is required to gather and prepare software for its general use on a
-	system for which it was designed. In our case effort includes the time spent
-	finding and understanding the installation instructions, the person-time and
-	resources spent performing the installation procedure, and the absence or
-	ease of overcoming system compatibility issues. An installability score
-	improves if the software has a means to validate the installation and
-	instructions for uninstallation.
+	a specified environment \citep{ISO/IEC25010, lenhard2013measuring}. In our
+	case effort includes the time spent finding and understanding the
+	installation instructions, the person-time and resources spent performing
+	the installation procedure, and the absence or ease of overcoming system
+	compatibility issues. An installability score improves if the software has a
+	means to validate the installation and instructions for uninstallation.
 
 	\item[Correctness] A software program is correct if it behaves according to
-	its stated specifications \citep{GhezziEtAl2003}. This requires that the
-	specification is available. Scientific software is unlikely to have an
+	its stated specifications \citep[p.\ 17]{GhezziEtAl2003}. This requires that
+	the specification is available. Research software is unlikely to have an
 	explicit specification, since this is not common practice in the field. As a
@@
 	failure-free operation of a computer program in a specified environment for
-	a specified time \citep{GhezziEtAl2003, musa1987software}. Reliability is
-	thus positively correlated with the absence of errors during installation
-	and use. Recoverability from errors also improves reliability.
+	a specified time \citep[p.\ 357]{GhezziEtAl2003}, \citep{musa1987software}.
+	Reliability is thus positively correlated with the absence of errors during
+	installation and use. Recoverability from errors also improves reliability.
 
@@
 	when the assumptions in its requirements specification are violated
-	\citep{boehm2007software, ghezzi1991fundamentals}. A good measure of
+	\citep{boehm2007software}, \citep[p.\ 19]{GhezziEtAl2003}. A good measure of
 	robustness correlates with a reasonable reaction to unexpected input,
 	including data of the wrong type, empty input, or missing files or links.
-	Examples of reasonable reaction include an appropriate error message and the
+	Examples of reasonable reactions include an appropriate error message and the
 	ability to recover the system.
@@
 	\citep{IEEEStdGlossarySET1990, boehm2007software}. In the current work
-	maintainability is measured by the quality of documentation artifacts, and
-	the presence of version control and issue tracking. These artifacts can
+	maintainability is measured by the quality of the documentation artifacts,
+	and the presence of version control and issue tracking. These artifacts can
 	greatly decrease the effort needed to modify software. There are many
@@
 	to the structure of the system, modules and code blocks. A follow up
-	question is asked if any measures had been taken.
+	question is asked if any measures have been taken.
 
@@
 	software package can be used with or without adaptation in other software
-	packages \citep{kalagiakos2003non}. A good measure of reusability results
-	from a large number of easily reusable components. Increased software
-	modularization, defined as the presence of smaller components with well
-	defined interfaces, is important. For this state of the practice assessment,
-	we assumed that a good measure of reusability correlates with an increased
+	packages \citep{kalagiakos2003non}. A good indicator of reusability is a
+	large number of easily reusable components. Therefore, we look for high
+	modularization, which is defined as the presence of smaller components with
+	well defined interfaces. For this state of the practice assessment, we
+	assume that a good measure of reusability correlates with a high
 	number of code files, and the availability of API documentation.
@@
 	specification \citep{McCallEtAl1977}. Similar to the quality of correctness,
-	this requires the availability of some form of specification. This quality
-	refers to keeping track of information as it changes forms or relates
-	between artifacts. We did not quantitatively measure traceability. Instead,
-	developers were asked in interviews how documentation fits into their
-	development process.
+	this requires the availability of some form of requirements specification.
+	This quality refers to keeping track of information as it changes forms or
+	relates between artifacts. We did not quantitatively measure traceability.
+	Instead, developers were asked in interviews how documentation fits into
+	their development process.
 
@@
 	extent to which all of the steps of a software development process, and the
-	current status of it, are conveyed clearly \citep{ghezzi1991fundamentals}.
+	current status of it, are conveyed clearly \citep[p.\ 32]{GhezziEtAl2003}.
 	In this state of the practice assessment a good measure of visibility and
 	transparency correlates with a well defined development process,
-	documentation of the development process and environment and software
-	package version release notes.
+	documentation of the development process and environment, and software
+	version release notes.
 
@@
 	not quantitatively measure unambiguity, but we did ask developers if they
-	think that the current documentation can clearly convey all necessary
-	knowledge to the users, and how they achieved this or what improvements are
-	needed to achieve it.
+	believe that their current documentation clearly conveys all necessary
+	knowledge to the users.  Follow-up questions ask how they achieved
+	unambiguity, or what improvements are needed to achieve it.
 
@@
 	can be found in \citet{SmithEtAl2021}. (Section~\ref{empiricalmeasures})
-	\item Survey the developers. (Section~\ref{SecSurvey})
 	\item Use AHP to rank the software packages. (Section~\ref{AHP})
+	\item Interview the developers. (Section~\ref{SecSurvey})
+	\item Domain analysis. (Section~\ref{SecDomainAnalysis})
 	\item Analyze the results and answer the research questions.
@@
 
-The above steps depend on interaction with a Domain Expert partner, as discussed
-in Section~\ref{sec_vet_software_list}.
+\noindent The above steps depend on interaction with a Domain Expert partner, as
+discussed in Section~\ref{sec_vet_software_list}.
 
@@
 	for GitHub-style repositories.
-	\item The software cannot be marked as incomplete or in an initial state of
+	\item The software cannot be marked as incomplete, or in an initial state of
 	development.
@@
 `alive' if their related documentation had been updated within the last 18
-month. Packages were categorized as `dead' if the last update of this
-information was more than 18 month ago.
+months. Packages were categorized as `dead' if the last update was more than 18
+month ago.
 
@@
 2022: Musubi.  Musubi did not come up in our initial search, but once we learned
-of it, we felt that this report is incomplete without including it.  With the
+of it, we felt that this report was incomplete without including it.  With the
 addition of Musubi we did not update the manual collection of data for the
@@
 each quality, and one for summary information
-(Figure~\ref{measurement_template_image}).   The summary section provides
-general information, such as the software name, number of developers, etc.  We
-follow the definitions given by \citet{gewaltig2012quality} for the software
+(Figure~\ref{measurement_template_image}). The summary section provides general
+information, such as the software name, number of developers, etc.  We follow
+the definitions given by \citet{gewaltig2012quality} for the software
 categories.  Public means software intended for public use.  Private means
@@
 
-Tools were use to find some of the measurements, such as the number of files,
+Tools were used to find some of the measurements, such as the number of files,
 number of lines of code (LOC), percentage of issues that are closed, etc. The
@@
 As in \citet{SmithEtAl2016}, Virtual machines (VMs) were used to provide an
-optimal testing environments for each package. VMs were used because it is
+optimal testing environment for each package. VMs were used because it is
 easier to start with a fresh environment without having to worry about existing
@@
 
-This project used a tool for conducting this process. The tool includes a
-sensitivity analysis that was used to ensure that the software package rankings
-are appropriate with respect to the uncertainty of the quality scores. For the
+This project used a tool we developed for conducting the AHP analysis. The tool
+includes a sensitivity analysis to ensure that the software package rankings are
+appropriate with respect to the uncertainty of the quality scores. For the
 sensitivity analysis we modified the score by 10\% for each package and verified
@@
 background of the development teams, the interviewees, and the software itself.
-We ask the developers how they organize their projects. We also ask them about
-their understanding of the users. Some questions focus on the current and past
+We ask the developers how they organize their projects. We also ask about their
+understanding of the users. Some questions focus on the current and past
 difficulties, and the solutions the team has found, or will try. We also discuss
@@
 meeting software (either Zoom or Teams), which facilitated recording and
-automatic transcription of the meetings.  Each meeting lasted about an hour. The
-interview process was approved by the McMaster University Research Ethics Board
-under the application number 
+automatic transcription.  Each meeting lasted about an hour. The interview
+process was approved by the McMaster University Research Ethics Board under the
+application number 
 \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/MACREM/Application.pdf}
@@
 
-\subsection{Interaction With Domain Expert} \label{sec_vet_software_list}
-
-We partnered with a Domain Expert to vet our list of projects
-(\rqref{RQ_WhatProjects}) and our ranking (\rqref{RQ_HighestQuality}).  The
-Domain Expert is an important member of the state of the practice assessment
-team. Pitfalls exist if non-experts attempt to acquire an authoritative list of
-software, or try to definitively rank the software. Non-experts have the problem
-that they can only rely on information available on-line, which has the
-following drawbacks:
-\begin{inparaenum}[i)]
-  \item the on-line resources could have false or inaccurate information; and,
-  \item the on-line resources could leave out relevant information that is so
-in-grained with experts that nobody thinks to explicitly record it.
-\end{inparaenum}
-For the current assessment, our Domain Expert (and paper co-author) is Dr.\
-Zahra Motamed, Assistant Professor of Mechanical Engineering at McMaster
-University, Hamilton, Ontario, Canada.
-
-The Domain Expert has an important role with verifying the list of LBM packages.
-In advance of the first meeting with the Domain Expert, they were asked to
-create a list of top software packages in the domain.  This is done to help the
-expert get in the right mind set in advance of the meeting.  Moreover, by doing
-the exercise in advance, we avoid the potential pitfall of the expert approving
-the discovered list of software without giving it adequate thought.  The Domain
-Expert was also asked to vet the collected data and analysis.  In particular,
-they were asked to vet the proposed list of software packages and the AHP
-ranking.  These interactions were done via virtual meetings.
-
-\section{Ranking Projects Based on Quality Measures} \label{AHPresults}
-
-In this section we answer \rqref{RQ_HighestQuality} by using AHP to rank the
-identified LBM software against best practices.  Before discussing quality
-differences between the packages and the results of the AHP ranking, we will
-contrast the different packages based on their features, as shown in
-Table~\ref{tbl_features}. The set of LBM programs form a program family, or a
-software product line.  The packages all have the commonality of using LBM
-techniques to simulate the motion of fluids, but there are also variabilities
-between them.  LBM packages can be distinguished by the dimension of problems
-they simulate, whether or not (and how) they support parallel computation,
-whether or not compressibility can be considered, whether reflexive boundaries
-can be employed, whether the simulation handles multi-fluids, whether the
-simulations handles turbulence, and whether complex geometries can be simulated.
-All of these variabilities are listed as columns in Table~\ref{tbl_features}
-with the software listed in alphabetical order. A more complete analysis of the
+\subsection{Domain Analysis} \label{SecDomainAnalysis}
+
+Since the LBM domain has a reasonably small scope, we are able to view the
+software as constituting a program family.  The concept of a program family is
+defined by \citet{parnas1976design} as ``a set of programs whose common
+properties are so extensive that it is advantageous to study the common
+properties of the programs before analyzing individual members''. Studying the
+common properties within a family of related programs is termed a domain
+analysis.
+
+The domain analysis (also called a commonality analysis) consists of
+understanding the commonalities, variabilities and common terminology for a
+program family \citep{weiss1997defining}. Commonalities are goals, theories,
+models, definitions and assumptions that are common between family members.
+Variabilities are goals, theories, models, definitions and assumptions that
+differ between family members. Associated with each variability are its
+parameters of variation, which summarize the possible values for that
+variability, along with their potential binding time.  The binding time is when
+the value of the variability is set.  It could be set as specification time,
+build time (when the program is being compiled) or run time (when the code is
+executing). As an example, for the current assessment, the packages all have the
+commonality of using LBM techniques to simulate the motion of fluids.  For
+variabilities, the LBM packages can be distinguished by the dimension of
+problems they simulate, whether or not (and how) they support parallel
+computation, whether or not compressibility can be considered, whether reflexive
+boundaries can be employed, whether the simulation handles multi-fluids, whether
+the simulations handles turbulence, and whether complex geometries can be
+simulated.
+
+To keep the current presentation focused, we only present the results of our
+shallow domain analysis. For the LBM domain a table was constructed that
+distinguishes the programs by their variabilities.  In research software the
+variabilities are often assumptions, as shown in Table~\ref{tbl_features}. The
+variabilities are listed as columns in Table~\ref{tbl_features} with the
+software listed in alphabetical order. A more complete analysis of the
 commonalities and variabilities between LBM packages is provided in
@@
 
-The sections below summarize how the software packages compare based on the
-qualities discussed in Section~\ref{softwarequalities}, as measured via the
-measurement template (Section~\ref{empiricalmeasures}).  For the purpose of
-discussion, the last subsection combines all the qualities into one ``global''
-ranking.
+\subsection{Interaction With Domain Expert} \label{sec_vet_software_list}
+
+We partnered with a Domain Expert to vet our list of projects
+(\rqref{RQ_WhatProjects}) and our ranking (\rqref{RQ_HighestQuality}).  The
+Domain Expert is an important member of the state of the practice assessment
+team. Pitfalls exist if non-experts attempt to acquire an authoritative list of
+software, or try to definitively rank the software. Non-experts have the problem
+that they can only rely on information available on-line, which has the
+following drawbacks:
+\begin{inparaenum}[i)]
+  \item the on-line resources could have false or inaccurate information; and,
+  \item the on-line resources could leave out relevant information that is so
+in-grained with experts that nobody thinks to explicitly record it.
+\end{inparaenum}
+For the current assessment, our Domain Expert (and paper co-author) is Dr.\
+Zahra Motamed, Assistant Professor of Mechanical Engineering at McMaster
+University, Hamilton, Ontario, Canada.
+
+The Domain Expert has an important role with verifying the list of LBM packages.
+In advance of the first meeting with the Domain Expert, they were asked to
+create a list of top software packages in the domain.  This is done to help the
+expert get in the right mind set in advance of the meeting.  Moreover, by doing
+the exercise in advance, we avoid the potential pitfall of the expert approving
+the discovered list of software without giving it adequate thought.  The Domain
+Expert was also asked to vet the collected data and analysis.  In particular,
+they were asked to vet the proposed list of software packages and the AHP
+ranking.  These interactions were done via virtual meetings.
+
+\section{Ranking Projects Based on Quality Measures} \label{AHPresults}
+
+In this section we answer \rqref{RQ_HighestQuality} by using AHP to rank the
+identified LBM software against best practices.  The sections below summarize
+how the software packages compare based on the qualities discussed in
+Section~\ref{softwarequalities}, as measured via the measurement template
+(Section~\ref{empiricalmeasures}).  For the purpose of discussion, the last
+subsection combines all the qualities into one ``global'' ranking.  The full LBM
+data is available publicly on Mendeley Data \citep{Smith2022}.
 
@@
 particularly  detailed tutorial that include expected output is provided by
-\href{https://www.walberla.net/doxygen/index.html}{waLBerla}.
+\href{https://www.walberla.net/doxygen/index.html} {waLBerla}.
 
 Unit tests are only explicitly available for two of the software packages,
-Ludwig and Musubi. Code modularization of most packages allow for users to
+Ludwig and Musubi. Code modularization of most packages allows for users to
 create tests with varying degrees of effort. These tests allow developers and
@@
 dependencies, and syntax errors in code files. In some instances the error
-messages were vague. Several automatic installation processes could not find and
+messages were vague. Several automatic installation processes could not find or
 load dependencies. In these instances the installation tried to access outdated
@@
 and one of the installations (LB3D-Prime) was assumed to be recovered due to the
-absence of any way to verify it. The installation of eight of the software
+absence of any way to verify otherwise. The installation of eight of the software
 packages could not be recovered. Most of these broken installations could not
@@
 command despite following the tutorial, and the final two errors were vague
-execution errors. Of the four broken tutorial instances, only the one that was
-missing a dependency was recoverable.
+execution errors. Of the four broken tutorial instances, only the missing
+dependency one was recoverable.
 
@@
 correctly passed this test. They output descriptive error messages or at least
-do not crash. Software packages with a lower surface robustness score did not
+did not crash. Software packages with a lower surface robustness score did not
 install correctly, so their robustness score may not be a true reflection of
@@
 Software packages were reviewed for the presence of artifacts. Every type of
-artifact or file that is not a code file was recorded. The software packages
-were also reviewed for software release and documentation version numbers. This
-information can be used to troubleshoot issues and organize documentation. All
-but three software packages (LatBo.jl, LB3D-Prime, MechSys) have source code
-release and documentation version numbers.
+artifact or file that is not a code file was recorded. (The list of artifacts
+will be discussed further in Section~\ref{Sec_CompareArtifacts}.)  The software
+packages were also reviewed for software release and documentation version
+numbers. This information can be used to troubleshoot issues and organize
+documentation. All but three software packages (LatBo.jl, LB3D-Prime, MechSys)
+have source code release and documentation version numbers.
 
@@
 to host their git repository, seven use email, and one (SunlightLB) uses
-SourceForge. Most software packages that use GitHub/GitLab have most of their
-issues closed, and only three (laboetie, lettuce, Sailfish) have less than 50
-percent of their issues closed. All of the top five overall ranked packages have
-most of their issues closed. Alive packages (11 use issue tracking) have 64\% of
-their issues closed, while dead packages (3 use issue tracking) have 71\% of
-their issues closed. This information is presented in Table~\ref{gitrepodata}.
-With respect to the host for the git repository, 13 packages use GitHub and two
-(Palabos, waLBerla) use GitLab. Of the other packages, one package (SunlightLB)
-uses CVS for issue tracking and version control, and seven packages do not
-appear to use any issue tracking system.
+SourceForge. Most software packages that use GitHub/GitLab have the majority of
+their issues closed, and only three (laboetie, lettuce, Sailfish) have less than
+50 percent of their issues closed. All of the top five overall ranked packages
+have most of their issues closed. Alive packages (11 use issue tracking) have
+64\% of their issues closed, while dead packages (3 use issue tracking) have
+71\% of their issues closed. This information is presented in
+Table~\ref{gitrepodata}. With respect to the host for the git repository, 13
+packages use GitHub and two (Palabos, waLBerla) use GitLab. Of the other
+packages, one package (SunlightLB) uses CVS for issue tracking and version
+control, and seven packages do not appear to use any issue tracking system.
 
@@
 that is comments. The findings are presented in Table~\ref{gitrepodata}, sorted
-by the percentage. Packages with a higher percentage of comments were assumed to be more maintainable. Comments represent more than 10 percent of code files in
+by the percentage. Packages with a higher percentage of comments were assumed to
+be more maintainable. Comments represent more than 10 percent of code files in
 16 packages, and the average percentage of code comments is about 14 percent.
@@
 
-Figure~\ref{Fig_Understandability} shows the surface understandability ranking
-of the software packages. Software packages with a high score have a consistent
-indentation and formatting style, and consistent, distinctive, and meaningful
-code identifiers. They also have symbolic constants, and explicitly identify
-mathematical and LBM algorithms. Their comments are clear and indicate what is
-being done in the source code. The source code is well modularized and
-structured. All of the top five ranked packages are noted as being alive.
+Figure~\ref{Fig_Understandability} shows the surface understandability ranking.
+Software packages with a high score have a consistent indentation and formatting
+style, and consistent, distinctive, and meaningful code identifiers. They also
+have symbolic constants, and explicitly identify mathematical and LBM
+algorithms. Their comments are clear and indicate what is being done in the
+source code. The source code is well modularized and structured. All of the top
+five ranked packages are noted as being alive.
 
@@
 development model, like a waterfall of agile development model, and the presence
-of documentation recording the development process and standard. They were also
+of documentation recording the development process and standards. They were also
 reviewed for the identification of the development environment, and the presence
@@
 ranking of the software packages. Software packages with a high score have an
-explicit development model and defined development process. They also had
+explicit development model and defined development process. They also have
 detailed and easy to access notes accompanying software releases. Four of the
@@
 between our measures and the star measures may also be due to inaccuracy with
-using stars to approximate popularity, since stars and watches tend to increase
-over time. Older packages (like ESPResSo++ and TCLB) have had more time to
-accumulate stars and watches. Young projects have less time to accumulate stars,
-like HemeLB, which only 12 stars (at the time of measurement) having recently
-moved from a different GitHub repository.  Lettuce is a young project (released
-in 2019) with a relatively high number of stars.  The discrepancy in rankings
-here may be due to a mismatch between the time when lettuce was manually
-measured by our template and when the stars were counted.  The late addition of
-Musubi meant a large gap between when projects were manually measured and the
-time when our automated data (like stars) was collected.
+using stars to approximate popularity, because of how people use stars, because
+young projects have less time to accumulate stars, and because stars represent
+the community's feeling in the past more than current preferences
+\citep{Szulik2017}. Older packages (like ESPResSo++ and TCLB) have had more time
+to accumulate stars and watches. Young projects have less time to accumulate
+stars, like HemeLB, which only has 12 stars (at the time of measurement) having
+recently moved from a different GitHub repository.  Lettuce is a young project
+(released in 2019) with a relatively high number of stars. The discrepancy in
+rankings here may be due to a mismatch between the time when lettuce was
+manually measured by our template and when the stars were counted. The late
+addition of Musubi meant a large gap between when projects were manually
+measured and the time when our automated data (like stars) was collected. A
+final reason for inconsistencies between our ranking and the community's ranking
+is that, as for consumer products, more factors influence popularity than just
+quality.
 
@@
 informative website. As far as we could observe, Ludwig does not appear to have
-publicly visible design and OpenLB does not appear to use a version control
+a publicly visible design and OpenLB does not appear to use a version control
 system. However, OpenLB's use of version numbers suggests that version control
@@
 (ESPREsSo, Ludwig, Palabos) have information on contributing to the project. Two
-of them (OpenLB, Palabos) have a FAQ section or forum. One (OpenLB) has
+of them (OpenLB, Palabos) have an FAQ section or forum. One (OpenLB) has
 verification and validation notes, and a video guide for the software.
@@
 Arts and Humanities) Guidelines for Software Quality \citep{vanGompelEtAl2016}.
-As an example, LBM developers list the external libraries on which the software
-depends as recommended by \citet{BrettEtAl2021}, \citet{SmithAndRoscoe2018} and
-\citet{vanGompelEtAl2016}. Although in the rare column, some LBM developers
-include an FAQ (Frequently Asked Questions) document as recommended by
-\citep{OrvizEtAl2017}, \citep{ThielEtAl2020} and \citep{vanGompelEtAl2016}. LBM
-developers also follow the ubiquitous recommendation to have a README file,
-although, as for other software, the content of README files is not standardized
-\citep{PranaEtAl2018}.
+As an example, LBM developers commonly list the external libraries on which the
+software depends, as recommended by \citet{BrettEtAl2021},
+\citet{SmithAndRoscoe2018} and \citet{vanGompelEtAl2016}. Although in the rare
+column, some LBM developers include an FAQ (Frequently Asked Questions) document
+as recommended by \citep{OrvizEtAl2017}, \citep{ThielEtAl2020} and
+\citep{vanGompelEtAl2016}. LBM developers also follow the ubiquitous
+recommendation to have a README file, although, as for other software, the
+content of README files is not standardized \citep{PranaEtAl2018}.
 
@@
 documentation is recommended by some \citep{TobiasEtAl2018, HerouxEtAl2008,
-SmithAndKoothoor2016}, in practice scientific software developers often do not
+SmithAndKoothoor2016}, in practice research software developers often do not
 produce a proper requirements specification \citep{HeatonAndCarver2015}.
@@
 showed requirements are the least commonly produced type of documentation for
-scientific software in general. When looking at the pain points for research
+research software in general. When looking at the pain points for research
 software developers, \citet{WieseEtAl2019} found that software requirements and
@@
 analysis (e.g.\ Anaconda, Torch), and message passing libraries (e.g.\ MPICH,
-Open MPI, PyZMQ). Domain specific libraries include scientific computing
-libraries (e.g.\ SciPy).
+Open MPI, PyZMQ). Domain specific libraries include research software libraries
+(e.g.\ SciPy).
 
@@
 which we had access.  Collaboration tools, most often email and video, are noted
-as being used when developing the software projects. Project management software
-was not explicitly mentioned, but it is possible that some of the projects use
-such software. Many of the projects are located on GitHub, where the developers
-use the platform to help manage their projects, especially bug related issues.
-Most of the projects appear to use change tracking and version control tools.
-Document generation tools are mentioned in 12 of the 24 projects. The tools
-Sphinx and Doxygen are explicitly used in this capacity.
+as being used when developing software projects. Project management software was
+not explicitly mentioned during our interviews, but it is possible that some of
+the projects use such software. Many of the projects are located on GitHub,
+where the developers use the platform to help manage their projects, especially
+bug related issues. Most of the projects appear to use change tracking and
+version control tools. Document generation tools are mentioned in 12 of the 24
+projects. The tools Sphinx and Doxygen are explicitly used in this capacity.
 
@@
 packages lag behind the recommended best practice of employing continuous
-integration tools.
+integration.
 
@@
 
-Our observations of a informally defined process, with elements of agile
+Our observations of an informally defined process, with elements of agile
 methods, matches what has been observed for research software in general.
@@
 priorities for any scientific project.  In earlier studies
-\citet{KellyAndSanders2008} and \citet{CarverEtAl2007} both highlight how
-important correctness is for research software.
+\citet{KellyAndSanders2008} and \citet{CarverEtAl2007} highlight how important
+correctness is for research software.
 
@@
 	Incentive and Funding} The TCLB developer noted a lack of incentives and
-	funding in academia for developing widely used scientific software. This
-	problem has also been noted others \citep{gewaltig2012quality, Goble2014,
+	funding in academia for developing widely used research software. This
+	problem has also been noted by others \citep{gewaltig2012quality, Goble2014,
 	KaterbowAndFeulner2018}.  \citet{WieseEtAl2019} reported developer pains
@@
 	\citep{CrouchEtAl2013}, and \href{https://software-carpentry.org/}{Software
-	Carpentry} \wss{WilsonAndLumsdaine2006, Wilson2016}. Over time awareness of
+	Carpentry} \citep{WilsonAndLumsdaine2006, Wilson2016}. Over time awareness of
 	these groups will grow, so this pain point is likely to disappear in the
@@
 	\item[P\refstepcounter{pnum}\thepnum \label{P_TechnologyHurdles}:]
-	\textbf{Technology Hurdles} Technology pain points include setting up
-	parallelization and continuous integration.
+	\textbf{Technology Hurdles} Technology pain points for LBM developers
+	include setting up parallelization and continuous integration. The pain
+	point survey of \citep{WieseEtAl2019} also highlighted technical-related
+	problems like dependency management, cross-platform compatibility,
+	continuous integration, hardware issues and operating system issues.
 
@@
 	\textbf{Ensuring Correctness} Difficulties with ensuring correctness were
-	noted by several developers. Several developers alluded to difficulty with
-	testing the correctness of large numbers of features, and some manually
-	tested program output, as opposed to using automated testing. The TCLB
-	developer commented that the amount of testing data needed was sometimes
-	problematic, since free testing services do not offer adequate facilities
-	for large amounts of data, which means in-house testing solutions are
-	needed.  Other research software domains point to the following problems
-	with testing: i) \citep{PintoEtAl2018} mention the problem of insufficient
-	testing; ii) the survey of \citep{HannayEtAl2009} shows that more developers
-	think testing is important than the number that believe they have a
-	sufficient understanding of testing concepts; and, iii)
+	noted by several developers. They alluded to difficulty with testing the
+	correctness of large numbers of features, and challenges with automated
+	automated testing. The TCLB developer commented that the amount of testing
+	data needed was sometimes problematic, since free testing services do not
+	offer adequate facilities for large amounts of data, which means in-house
+	testing solutions are needed.  Other research software domains point to the
+	following problems with testing: i) \citep{PintoEtAl2018} mention the
+	problem of insufficient testing; ii) the survey of \citep{HannayEtAl2009}
+	shows that more developers think testing is important than the number that
+	believe they have a sufficient understanding of testing concepts; and, iii)
 	\citep{HannayEtAl2009, KanewalaAndBieman2013, KellyEtAl2011, WieseEtAl2019}
@@
 	LBM experience seems to overlap with i and ii, but the LBM developers did
-	not allude to problem iii (the oracle problem) in our conversation; we
-	believe that they have developed techniques to work around the oracle
-	problem.
+	not allude to problem iii (the oracle problem) in our conversations.  As
+	discussed in Section~\ref{SecReproducibility}, the oracle problem likely did
+	not come up because the LBM developers we interviewed have strategies for
+	verifying the reproducibility of their work.
 
@@
 	while in reality CFD knowledge needs to be applied to correctly modify the
-	packages for the new endeavour.  \citep{WieseEtAl2019} also interacted with
-	developers that mentioned that users do not always have the expertise
-	required to install or use the software.
+	packages for the new endeavour.  Some of the respondents in the survey of
+	\citep{WieseEtAl2019} also mentioned that users do not always have the
+	expertise required to install or use the software.
 
@@
 defined interfaces. This is suggested in SCS literature, and in developer
-comments.  \wss{one of Wilson's best practices.}
-
-\begin{itemize}
-	\item Modifiability. Consider flexibility of data structures and data
+comments.  \wss{one of Wilson's best practices, although Wilson (like so many
+others) is silent on what criteria to use for modularization.}
+
+Modifiability. Consider flexibility of data structures and data
 storage in the design stage. The package pyLBM redeveloped data structures
 to ease future changes.
-	\item Reusability. Provide API documentation, if applicable. Only one
-	(ESPResSo) of the top five ranked packages provided API documentation.
-\end{itemize}
-
-\subsection{Verification and Validation} \label{Sec_VnV}
-
-\wss{Address Correctness (\ppref{P_Correctness}): Testing is not the only way to
-build confidence in correctness.  Peer review.  }
-
-Surface Correctness and Verifiability. Development teams should
-include both domain experts and experienced software developers. This
-suggestion is based on developer comments.
-
-Visibility and Transparency. Consider peer review processes to assess
-contributions and ensure the tracking of information. High ranked package
-ESPResSo uses a peer review process for contributions.
+
+Contrast with SE literature.  Can point to references, starting with Parnas, on
+the importance of modularization.  Also they seem to have a basis for the
+modularization that matches with Parnas - likely changes.
+
+\subsection{Circumvent the Oracle Problem to Achieve Reproducibility}
+\label{SecReproducibility}
+
+Software packages were not quantitatively measured for reproducibility. Instead,
+developers were asked if they had concerns that their computational results
+would not be reproducible in the future, and if they had taken any steps to
+ensure reproducibility.  The developers all agreed that reproducibility is
+important, but they differ in how far they go to support reproducibility.
+ESPResSo does the most to support reproducibility.  For instance, they compare
+the results of their methods against manually calculated results. These
+comparison tests are automatically run for all source code changes. Continuous
+integration is used, so these tests are run when a pull request is opened on
+GitHub. Even once these tests are complete, a peer review process is undertaken
+before changes are fully committed to the appropriate branch. The results for
+all of the LBM schemes on the software package development branch are also
+frequently compared for correctness, ensuring that the system output reflects
+the expected output.
+
+For other teams, the priority is verifying the mathematical foundations of the
+models, but the output of the software package is not compared to other output.
+Depending on the package and how it outputs solutions, this may not be practical
+or feasible. A correct output may not be exactly reproducible, as it may be
+dependent on a probability distribution, so strictly comparing results may not
+be appropriate. \wss{Ideas for dealing with reproducibility in these situations
+will be explored in the recommendations section (like properties of the solution,
+such as conservation of mass).}
+
+Some software packages, like Palabos, provide validation benchmarks for their
+core fundamental algorithmic ingredients \citep{latt2021palabos}. The stated
+intent of these benchmarks is to showcase the validity and usefulness of the
+package to stimulate the development of third-party extensions. The Palabos
+package identifies as a development framework for modelling problems in various
+CFD areas.  \wss{Add validation benchmarks as a ``best practice'' used by
+Palabos - others could emmulate.  Goal of this exercise is for people in the
+community to learn from each other, and potentially from our advice too.}
+
+Consider peer review processes to assess contributions and ensure the tracking
+of information. High ranked package ESPResSo uses a peer review process for
+contributions.
 
@@
 
-\subsection{Reproducibility} \label{SecReproducibility}
-
-Software packages were not quantitatively measured for reproducibility. Instead,
-developers were asked if they had concerns that their computational results
-would not be reproducible in the future, and if they had taken any steps to
-ensure reproducibility.  The developers all agreed that reproducibility is
-important, but they differ in how far they go to support reproducibility.
-ESPResSo does the most to support reproducibility.  For instance, they compare
-the results of their methods against manually calculated results. These
-comparison tests are automatically run for all source code changes. The tests
-are run when a pull request is opened on GitHub. Even once these tests are
-complete, a peer review process is undertaken before changes are fully committed
-to the appropriate branch. The results for all of the LBM schemes on the
-software package development branch are also frequently compared for
-correctness, ensuring that the system output reflects the expected output.
-
-For other teams, the priority is verifying the mathematical foundations of the
-models, but the output of the software package is not compared to other output.
-Depending on the package and how it outputs solutions, this may not be practical
-or feasible. A correct output may not be exactly reproducible, as it may be
-dependent on a probability distribution, so strictly comparing results may not
-be appropriate. \wss{Ideas for dealing with reproducibility in these situations
-will be explored in the recommendations section (like properties of the solution,
-such as conservation of mass).}
-
-Some software packages, like Palabos, provide validation benchmarks for their
-core fundamental algorithmic ingredients \citep{latt2021palabos}. The stated
-intent of these benchmarks is to showcase the validity and usefulness of the
-package to stimulate the development of third-user extensions. The Palabos
-package identifies as a development framework for modelling problems in various
-CFD areas.  \wss{Add validation benchmarks as a ``best practice'' used by
-Palabos - others could emmulate.  Goal of this exercise is for people in the
-community to learn from each other, and potentially from our advice too.}
-
-\begin{itemize}
-\item Reproducibility. Test output against automatically calculated or known
-correct results. High ranked package ESPResSo does this.
-\item Reproducibility. Automate the testing of output. Run tests after all code
-changes. High ranked package ESPResSo does this.
-\item Reproducibility. Consider peer review processes and task based inspection
-to assess contributions. High ranked package ESPResSo does this.
-\end{itemize}
-
-\subsection{Documentation} \label{Sec_RecommendDoc}
+\subsection{Prioritize Documentation} \label{Sec_RecommendDoc}
+
+Despite it being a pain point, spending time on documentation provides a
+justifiable return on investment.  Addresses \ref{P_Documentation} on
+documentation.
 
@@
 package build system works, and how the coupling between the simulation engine
-and the interface works.  \wss{Peter - these points about developer
-documentation seem like they would fit better under understandability, don't you
-think?}  \pmm{After reviewing the definitions of the two qualities, I think you
-are right. I think that the point presented in the first sentence of this
-paragraph belongs here in traceability (but could be reworded to note that
-ESPResSo also has detailed change logs in their docs - change logs are also
-available for Palabos and waLBerla), but the rest of the paragraph can be moved
-to understandability.} \wss{If ESPresSo scored high (I think it did), we can use
-that to further promote the value of documentation.  Also a threat to validity
-is that we have implicitly assumed documentation is valuable when measuring
-quality.}
-
-Finding ways to hasten updating documentation would increase the
-frequency of such updates and benefit both users and developers.  \wss{Same
-thing has come up in papers.  I'll try to find a reference for this.  This is
-also something we can point to for recommendations related to code generation.}
-\wss{Peter - this also sounds like understandability related discussion?  I
-wonder if we should drop Traceability as a quality that we assessed?  There
-doesn't seem to be much content specific to traceability.}  \pmm{Based on the
-comments I just added above, I think we can keep at least a small part of what
-is here in traceability, rewording parts to highlight the importance of change
-logs. Some of the content in this section should indeed be moved to
-understandability.} The developer of OpenLB noted the use of documentation
-generators like Doxygen. It would be advisable for more projects to use such
-automatic document generation tools, since some projects do not do this.
-\wss{We can refer back to this for the recommendations.}
+and the interface works. \wss{If ESPresSo scored high (I think it did), we can
+use that to further promote the value of documentation.  Also a threat to
+validity is that we have implicitly assumed documentation is valuable when
+measuring quality.}
+
+Finding ways to hasten updating documentation would increase the frequency of
+such updates and benefit both users and developers.  \wss{Same thing has come up
+in papers.  I'll try to find a reference for this.  This is also something we
+can point to for recommendations related to code generation.} The developer of
+OpenLB noted the use of documentation generators like Doxygen. It would be
+advisable for more projects to use such automatic document generation tools,
+since some projects do not do this. \wss{We can refer back to this for the
+recommendations.}
+
+Reusability. Provide API documentation, if applicable. Only one (ESPResSo) of
+the top five ranked packages provided API documentation.
 
@@
 
-\subsection{Visibility and Transparency} \label{secVisib}
-
-\wss{Will likely remove this section, but should find a way to say the good
-things that ESPResSo does.} Although visibility is not a priority, positive
-steps are being undertaken. For instance, all interviewed teams used GitHub,
-with its visibility enhancing features of branching, issue tracking, insights on
-contributors, the blame feature, wikis, pull requests, etc. Continuous
-integration, as used by ESPResSo, makes projects even more visible, since
-contributors can explicitly see the bar that must be crossed for code to be
-incorporated into the master branch. ESPResSo also uses peer review, which
-improves visibility by explicitly asking for feedback from other developers.
-
-\wss{in recommendations could point out what teams could do to get more external
-contributors, and potentially a longer life for their project.}
-
 \section{Recommendations for Future Practices} \label{Sec_Recommendations}
@@
 
+\wss{Address Correctness (\ppref{P_Correctness}): Testing is not the only way to
+build confidence in correctness.  Peer review.  }
+
 \wss{Address lack of funding (\ppref{P_LackFunding}): The developer further
-commented that there are no journals that publish such scientific software
+commented that there are no journals that publish such research software
 source code. However, there are ways to get such source code cited. Work has
@@
 
+Test the installation and tutorials.  Assign a junior team member the task?
+
 \subsection{Grow the Number of Contributors}
@@
 
+\wss{in recommendations could point out what teams could do to get more external
+contributors, and potentially a longer life for their project.}
+
+\subsection{Augment Theory Manual to Include Requirements}
+
+- previous section discusses challenges with requirements and not popular (maybe
+move that section here) - say what information to incorporate, low hanging fruit
+- return on investment
+
 \section{Threats To Validity} \label{threats}
+
+Maintainability is assumed to be improved by a higher ratio of comments to
+source code (Section~\ref{Sec_Maintainability}), but this is not necessarily
+true.
+
+We assume that a large number of source files is associated with increased
+reusability (Section~\ref{reusabilityresults}), but this is not necessarily the
+case.
 
@@
 
+Section~\ref{repmetrics} compares ranking by best practices to the community's
+ranking.  Using stars to estimate the community's ranking is not necessarily
+accurate.
+
+Tool usage: Interviews with the developers revealed a potentially more frequent
+use of both unit testing and continuous integration, compared to what was
+observed from studying the repository artifacts.
+
 \subsection{Future State Of The Practice Assessments} \label{futuresop}
@@
 
+The final result of the domain analysis will be tables of commonalities,
+variabilities, and parameters of variation of a program family.
+\cite{smith2008commonality} present a template for conducting a commonality
+analysis, which was referred to when conducting this work.
+\cite{weiss1998commonality} describes another commonality analysis technique for
+deciding the members of a program family. \cite{SmithAndChen2004} and
+\cite{SmithMcCutchanAndCarette2017} are examples of a commonality analysis for a
+family of mesh generating software and a family of material models,
+respectively. The steps to produce a commonality analysis are:
+
 \section{Conclusion} \label{conclusion}
@@
 top 10 ranked packages are ranked highly by a domain expert. Recommendations for
-improving software along quality metrics were made, and highlights are presented
-in Section~\ref{highlightedrecommendations} of this conclusion. The findings of
-this report can be used to guide future development of SCS, specifically along
+improving software along quality metrics were made... The findings of this
+report can be used to guide future development of SCS, specifically along
 quality attributes, and to reduce software quality failures.
@@
 
+Include a table that summarizes the top 4 projects for each quality, but an
+asterisk if still alive.
+
 \bibliographystyle{ACM-Reference-Format}
@@
 - value of tools to enforce good behaviour of developers, like R, like MOOSE infrastructure
-
-- change \citet{ghezzi1991fundamentals} to \citet{GhezziEtAl2003} (need to check
-each instance, add page numbers).
 

### 2022-02-25 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
added top ranking table

@@
 \usepackage{booktabs}
+\usepackage{pbox}
 
@@
 
+\begin{table}
+	\begin{center}
+		\begin{tabular}{ p{3cm}p{1.9cm}p{1.9cm}p{1.9cm}p{1.9cm}p{1.9cm} }
+			\toprule
+			Quality & Ranked 1st & Ranked 2nd & Ranked 3rd & Ranked 4th & Ranked 5th\\
+			\midrule
+			Installability & ESPResSo & MechSys & OpenLB & DL\_MESO (LBE) & Sailfish\\
+			\addlinespace[0.4cm]
+			\pbox{3.0cm}{Surface Correctness \\ and Verifiability} & Ludwig & Musubi & pyLBM & OpenLB & waLBerla\\
+			\addlinespace[0.4cm]
+			Surface Reliability & DL\_MESO (LBE) & laboetie & lettuce & MechSys & Musubi \\
+			\addlinespace[0.4cm]
+			Surface Robustness & DL\_MESO (LBE) & ESPResSo++ & lettuce & LUMA & OpenLB \\
+			\addlinespace[0.4cm]
+			Surface Usability & laboetie & ESPResSo & OpenLB & Ludwig & Sailfish\\
+			\addlinespace[0.4cm]
+			Maintainability & Ludwig & MP-LABS & ESPResSo++ & ESPResSo & Palabos\\
+			\addlinespace[0.4cm]
+			Reusability & ESPResSo & waLBerla & ESPResSo++ & Sailfish & HemeLB\\
+			\addlinespace[0.4cm]
+			\pbox{3.0cm}{Surface \\Understandability} & LUMA & Palabos & pyLBM & DL\_MESO (LBE)& MechSys\\
+			\addlinespace[0.4cm]
+			\pbox{3.0cm}{Visibility and \\Transparency} & Ludwig & Palabos & LUMA & ESPResSo & MP-LABS\\
+			\addlinespace[0.4cm]
+			Overall Quality & ESPResSo & Ludwig & Palabos & OpenLB & LUMA\\
+			\bottomrule
+		\end{tabular}
+		\caption{Top performers for each quality} \label{topperformerstable}
+	\end{center}
+\end{table}
+
 \wss{More on the lessons to learn from the top packages.  The comparison to

### 2022-02-25 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
added top ranking table - dead marked

@@
 			\addlinespace[0.4cm]
-			Surface Reliability & DL\_MESO (LBE) & laboetie & lettuce & MechSys & Musubi \\
+			Surface Reliability & DL\_MESO (LBE) & laboetie* & lettuce & MechSys & Musubi \\
 			\addlinespace[0.4cm]
@@
 			\addlinespace[0.4cm]
-			Surface Usability & laboetie & ESPResSo & OpenLB & Ludwig & Sailfish\\
+			Surface Usability & laboetie* & ESPResSo & OpenLB & Ludwig & Sailfish\\
 			\addlinespace[0.4cm]
-			Maintainability & Ludwig & MP-LABS & ESPResSo++ & ESPResSo & Palabos\\
+			Maintainability & Ludwig & MP-LABS* & ESPResSo++ & ESPResSo & Palabos\\
 			\addlinespace[0.4cm]
-			Reusability & ESPResSo & waLBerla & ESPResSo++ & Sailfish & HemeLB\\
+			Reusability & ESPResSo & waLBerla & ESPResSo++ & Sailfish & HemeLB*\\
 			\addlinespace[0.4cm]
@@
 			\addlinespace[0.4cm]
-			\pbox{3.0cm}{Visibility and \\Transparency} & Ludwig & Palabos & LUMA & ESPResSo & MP-LABS\\
+			\pbox{3.0cm}{Visibility and \\Transparency} & Ludwig & Palabos & LUMA & ESPResSo & MP-LABS*\\
 			\addlinespace[0.4cm]

### 2022-02-25 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
Udpates to SOP for MI up to Section 3.2 (List of Representative Software)
Minor notes added to qdef paper
Addition of a reference on sustainability to use for the quality definitions paper
Minor typo corrections

@@
 
+\wss{Add definitions from ``Continuous quality control of long lived software systems''}
+
 \begin{defn}
@@
 
-It would be worthwhile to look for some additional definitions.
+It would be worthwhile to look for some additional definitions.  Definitions are
+found here: \href{https://www.ncbi.nlm.nih.gov/books/NBK547546/}{Understsanding
+Reproducibility and Replicability}
 
@@
 Includes examples where replicability was found to be lacking.  Has a good
-litereature review in Section 2.}
-
-\wss{Needs to be completed}
+literature review in Section 2.}
 
@@
 \subsection{\notdone{Sustainability (internal, product)}} %\notdone \wss{owner}
+
+Additional references on sustainability can be found in \citet{FerrariEtAl2021}.
 
@@
 \begin{defn}
-Sustainable Software is software whose direct and indirect negative impacts on economy, society, human beings, and the environment resulting from development, deployment, and usage of the software is minimal and/or has a positive effect on sustainable development \citep{dick2010model}.
-\end{defn}
-
-\begin{defn}
-The art of developing sustainable software with a sustainable software engineering process so that negative and positive impacts result in and/or are expected to result from the software product over its whole life cycle are continuously assessed, documented, and used for further optimization of the software product \citep{venters2014software}.
+Sustainable Software is software whose direct and indirect negative impacts on
+economy, society, human beings, and the environment resulting from development,
+deployment, and usage of the software is minimal and/or has a positive effect on
+sustainable development \citep{dick2010model}.
+\end{defn}
+
+\begin{defn}
+The art of developing sustainable software with a sustainable software
+engineering process so that negative and positive impacts result in and/or are
+expected to result from the software product over its whole life cycle are
+continuously assessed, documented, and used for further optimization of the
+software product \citep{venters2014software}.
 \end{defn}

### 2022-02-25 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Updates to section 9, far from done

@@
 
-\subsection{Circumvent the Oracle Problem to Achieve Reproducibility}
+\subsection{Circumventing the Oracle Problem to Verify Reproducibility}
 \label{SecReproducibility}
@@
 		\end{tabular}
-		\caption{Top performers for each quality} \label{topperformerstable}
+		\caption{Top performers for each quality (all packages were alive at the time of measurement, except for those marked by an asterisk (*))} \label{topperformerstable}
 	\end{center}

### 2022-02-26 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
Minor edits

@@
 \begin{defn}
-Closeness of the agreement between the results of measurements of the same measurand carried out under changed conditions of measurement \citep{iso2017iec}.
-\end{defn}
-
-\begin{defn}
-Obtaining consistent results using the same input data, computational steps,  methods, and code, and conditions of analysis \citep{national2019reproducibility}.
-\end{defn}
-
-\begin{defn}
-Running the same software on the same input data and obtaining the same results \citep{rougier2017sustainable}.
+Closeness of the agreement between the results of measurements of the same
+measurand carried out under changed conditions of measurement
+\citep{iso2017iec}.
+\end{defn}
+
+\begin{defn}
+Obtaining consistent results using the same input data, computational steps,
+methods, and code, and conditions of analysis
+\citep{national2019reproducibility}.
+\end{defn}
+
+\begin{defn}
+Running the same software on the same input data and obtaining the same results
+\citep{rougier2017sustainable}.
 \end{defn}

### 2022-02-26 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
First draft of Section 9.1 (Design for change)

@@
 	\item[Reproducibility] Software achieves reproducibility if another
-	developer can take the requirements documentation and re-obtain the same
-	software artifacts \citep{BenureauAndRougier2017}. This includes the output
-	of the software, where the scientific results are compared between software
-	implementations, or between software implementations and manually calculated
-	results. We measured reproducibility qualitatively by asking developers if
-	they have any concern that their computational results won't be
-	reproducible, and if they have taken any steps to ensure reproducibility.
+	developer can take the original code and input data, run it, and obtain the
+	original output \citep{BenureauAndRougier2017}. We measured reproducibility
+	qualitatively by asking developers if they have any concern that their
+	computational results won't be reproducible, and if they have taken any
+	steps to ensure reproducibility.
 
@@
 
-Increasing source code modularity, reducing duplicate information,
-and improving abstraction by developing well defined interfaces. This was noted
-by the developers of ESPResSo and pyLBM.  A developer of pyLBM mentioned that
-the geometries and models of their system had been ``decoupled'', using
-abstraction and modularization of the source code, to make it ``very easy to add
-[new] features''.  \wss{Highlights the advantage of design.  Other projects can
-look to these examples.  Maybe cite the Anshu paper about Flash? Dune? and point
-to the advantages of design.} Addresses maintainability.  The developer of TCLB
-mentioned that their package had two sets of code, for executing the models on
-the CPU and GPU, and that maintenance was decreased by introducing macros, a
-practice which became a common part of the development process.
-
-A high degree of code modularity and abstraction was noted by developers as a
-measure to ensure the ease of future changes. This can be ensured by separating
-components and hiding information behind well defined interfaces. The developer
-of ESPResSo also noted that some of the code base was transitioned from C to
-C++, which could ease modifiability of that software package. The developer of
-TCLB noted that their software package was designed to allow for the addition of
-some LBM features, but changes to major aspects of the system would be
-difficult. For example, ``implementing a new model will be an easy
-contribution'', but changes to the ``Cartesian mesh...will be a nightmare''.
-\wss{example suggests identifying likely changes.}
-
-Modularize the source code, separate components, hide information behind well
-defined interfaces. This is suggested in SCS literature, and in developer
-comments.  \wss{one of Wilson's best practices, although Wilson (like so many
-others) is silent on what criteria to use for modularization.}
-
-Modifiability. Consider flexibility of data structures and data
-storage in the design stage. The package pyLBM redeveloped data structures
-to ease future changes.
-
-Contrast with SE literature.  Can point to references, starting with Parnas, on
-the importance of modularization.  Also they seem to have a basis for the
-modularization that matches with Parnas - likely changes.
+The top LBM developers show how modularization can be used to design research
+software for future change.  Although the advice to modularize research software
+to handle complexity is common \citep{WilsonEtAl2014, StewartEtAl2017,
+Storer2017}, specific guidelines on how to divide the software into modules is
+less prevalent.  Not every decomposition is a good design for supporting change,
+as shown by \citet{Parns1972a}.  A design with low cohesion and
+high coupling \citep[p.\ 48]{GhezziEtAl2003} will make change difficult.
+Especially in research software, where change is inevitable, designers need to
+produce a modularization that supports change.
+
+LBM developers highlighted in our interviews cases where their modularizations
+anticipated future changes.  The developer of pyLBM mentioned that the
+geometries and models of their system had been ``decoupled'', using abstraction
+and modularization of the source code, to make it ``very easy to add [new]
+features''.  The pyLBM design allows for independent changes to the geometry and
+the model.  We also learned that the package pyLBM redeveloped data structures
+to ease future changes. The developer of TCLB noted that their software package
+was designed to allow for the addition of some LBM features, but changes to
+major aspects of the system would be difficult. For example, ``implementing a
+new model will be an easy contribution'', but changes to the ``Cartesian mesh
+... will be a nightmare''.  The design of TCLB highlights that not every conceivable change needs to be supported, only the likely changes.  Likely changes.
+
+As the LBM developers illustrate, design for changes is accomplished by first
+identifying likely changes, either implicitly or explicitly, and second by
+hiding those likely changes behind well-defined module interfaces.
 
@@
 - return on investment
+
+\subsection{Reproducibility}
+
+- record versions of everything
+- Docker, Virtual machines
 

### 2022-02-26 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Minor edits to Section 9.1

@@
 new model will be an easy contribution'', but changes to the ``Cartesian mesh
-... will be a nightmare''.  The design of TCLB highlights that not every conceivable change needs to be supported, only the likely changes.  Likely changes.
+... will be a nightmare''.  The design of TCLB highlights that not every
+conceivable change needs to be supported, only the likely changes.  Likely
+changes.
 
@@
 identifying likely changes, either implicitly or explicitly, and second by
-hiding those likely changes behind well-defined module interfaces.
+hiding each likely change behind a well-defined module interface.  Although it
+is unclear whether the developers are aware of Parnas's work, the approach
+mirrors his recommendations \citep{Parns1972a}.
 

### 2022-02-28 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Completed first draft of section 9.2 (Circumventing the Oracle Problem and Addressing Reproducibility)

@@
 	not come up because the LBM developers we interviewed have strategies for
-	verifying the reproducibility of their work.
+	verifying their work.
 
@@
 
-The top LBM developers show how modularization can be used to design research
-software for future change.  Although the advice to modularize research software
-to handle complexity is common \citep{WilsonEtAl2014, StewartEtAl2017,
-Storer2017}, specific guidelines on how to divide the software into modules is
-less prevalent.  Not every decomposition is a good design for supporting change,
-as shown by \citet{Parns1972a}.  A design with low cohesion and
-high coupling \citep[p.\ 48]{GhezziEtAl2003} will make change difficult.
-Especially in research software, where change is inevitable, designers need to
-produce a modularization that supports change.
+To address technical debt (\pref{P_TechDebt}), the top LBM developers show how
+modularization can be used to design research software for future change.
+Although the advice to modularize research software to handle complexity is
+common \citep{WilsonEtAl2014, StewartEtAl2017, Storer2017}, specific guidelines
+on how to divide the software into modules is less prevalent.  Not every
+decomposition is a good design for supporting change, as shown by
+\citet{Parnas1972a}.  A design with low cohesion and high coupling \citep[p.\
+48]{GhezziEtAl2003} will make change difficult. Especially in research software,
+where change is inevitable, designers need to produce a modularization that
+supports change.
 
@@
 ... will be a nightmare''.  The design of TCLB highlights that not every
-conceivable change needs to be supported, only the likely changes.  Likely
-changes.
+conceivable change needs to be supported, only the likely changes.
 
@@
 is unclear whether the developers are aware of Parnas's work, the approach
-mirrors his recommendations \citep{Parns1972a}.
-
-\subsection{Circumventing the Oracle Problem to Verify Reproducibility}
+mirrors his recommendations \citep{Parnas1972a}.
+
+\subsection{Circumventing the Oracle Problem and Addressing Reproducibility}
 \label{SecReproducibility}
 
-Software packages were not quantitatively measured for reproducibility. Instead,
-developers were asked if they had concerns that their computational results
-would not be reproducible in the future, and if they had taken any steps to
-ensure reproducibility.  The developers all agreed that reproducibility is
-important, but they differ in how far they go to support reproducibility.
-ESPResSo does the most to support reproducibility.  For instance, they compare
-the results of their methods against manually calculated results. These
-comparison tests are automatically run for all source code changes. Continuous
-integration is used, so these tests are run when a pull request is opened on
-GitHub. Even once these tests are complete, a peer review process is undertaken
-before changes are fully committed to the appropriate branch. The results for
-all of the LBM schemes on the software package development branch are also
-frequently compared for correctness, ensuring that the system output reflects
-the expected output.
-
-For other teams, the priority is verifying the mathematical foundations of the
-models, but the output of the software package is not compared to other output.
-Depending on the package and how it outputs solutions, this may not be practical
-or feasible. A correct output may not be exactly reproducible, as it may be
-dependent on a probability distribution, so strictly comparing results may not
-be appropriate. \wss{Ideas for dealing with reproducibility in these situations
-will be explored in the recommendations section (like properties of the solution,
-such as conservation of mass).}
-
-Some software packages, like Palabos, provide validation benchmarks for their
-core fundamental algorithmic ingredients \citep{latt2021palabos}. The stated
-intent of these benchmarks is to showcase the validity and usefulness of the
-package to stimulate the development of third-party extensions. The Palabos
-package identifies as a development framework for modelling problems in various
-CFD areas.  \wss{Add validation benchmarks as a ``best practice'' used by
-Palabos - others could emmulate.  Goal of this exercise is for people in the
-community to learn from each other, and potentially from our advice too.}
-
-Consider peer review processes to assess contributions and ensure the tracking
-of information. High ranked package ESPResSo uses a peer review process for
-contributions.
-
-\wss{From Peter: The developer of ESPResSo mentioned ``tests which compare
-implementations against each other, like say if there's to solve us for
-electrostatic switch rely on different principles and they've been written by
-different people''.
-The developer of TCLB also mentioned automatic testing. They discussed the idea
-of using test cases, but concerns were raised with this, such as the required
-computing power to run such tests frequently and where to store large amounts of
-data. Also, ``constructing such cases would be very hard''.
-The developer of OpenLB commented that they have different kinds of tests, such
-as integration tests, and tests where they know the ``exact analytical solution''.
-They also mentioned comparing computations and ``check[ing] the misfits''.}
+The top LBM projects have adopted techniques for software verification that get
+around the oracle problem (mentioned earlier in Section~\ref{P_Correctness}). In
+the absence of a general test oracle, the following techniques have been adopted
+by LBM developers:
+
+\begin{itemize}
+	\item Compare the calculated results against manually calculated results in
+	the cases where manual solutions can be determined.  Although it is not
+	always feasible to calculate results a priori, developers can find sub-sets
+	of the general problem where answers can be determined.  In special cases
+	that allow this, the results can be compared against exact analytical
+	solutions, as done by OpenLB.
+	\item Related to peer review, verify the mathematical foundations of the
+	models to build confidence in the software's theoretical underpinnings.
+	\item Compare the calculated results against pseudo-oracle solutions for
+	verification benchmarks. The developer of ESPResSo mentioned the value of
+	``tests which compare implementations against each other'' where the tests
+	``rely on different principles and they've been written by different
+	people''. \citet{latt2021palabos} shows several benchmark tests comparing
+	Palabos solutions to other software solutions, such as solutions calculated
+	using AnSys.
+	\item Use peer review of the code for major changes and additions, as
+	mentioned in Section~\ref{Sec_CompareMethodologies} for ESPResSo. 
+\end{itemize}
+
+\noindent Additional strategies for getting around the oracle problem are
+provided in Section~\ref{Sec_Recommendations}.
+
+Some LBM developers improve reproducibility via automating the verification
+tests and making them part of a continuous integration process (mentioned in
+Section~\ref{Sec_CompareTools}).  Providing unit tests improves reproducibility
+because future developers will require a means to verify that they can obtain
+the same computational results as today.  Unit tests are a good way to embed
+self-diagnostics of reproducibility in the code \citep{BenureauAndRougier2017}.
 
@@
 
+Borrow from SE for HEP paper.
+
 Answers \rqref{RQ_Recommend}.  practices for the future.  Probably the easiest
@@
 
+\subsection{Verifiability}
+
+- further steps to address the oracle problem - borrow from SE for HEP paper.
+- anything in Smith 2016?
+
 \subsection{Reproducibility}
@@
 - Docker, Virtual machines
+
+Repeatability - A correct output may not be exactly reproducible, as it may be
+dependent on a probability distribution, so strictly comparing results may not
+be appropriate. \wss{Ideas for dealing with reproducibility in these situations
+will be explored in the recommendations section (like properties of the
+solution, such as conservation of mass).  5R paper mentions repeatability -
+design it so that the random number seed can be set.}
 

### 2022-03-01 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Updates to section 10 on recommendations

@@
 
-Requirements documentation is feasible, as shown by the prevalence of theory
-documents in LBM software.  Table~\ref{artifactspresent} shows that theory
-documentation is common.  We observed 17 of 24 packages had at least some theory
-documentation. Templates for scientific requirements, like \citet{SmithEtAl2007,
-SmithAndLai2005}, show theory is a significant part of the requirements
-documentation. With the addition of some extra information, the theory documents
-can be transformed into requirements specification.  The key extra information
-includes explicitly stating user characteristics, explicitly stating how the
-user interacts with the software in terms of input data requirements, and
-listing likely and unlikely changes.  Explicit statements about likely changes
-are invaluable in the design stage, since they provide developers guidance on
-how general the software needs to be.  For those wishing to maximize the value
-of a requirement specification, information could be added like prioritizing the
-nonfunctional requirements (to show the relative importance between qualities
-like portability, reliability and performance) and traceability information (to
-show the consequences of any changes to the assumptions).
-
 \section{Comparison of Tool Usage Between LBM and Other Research Software}
@@
 
-\subsection{Prioritize Documentation} \label{Sec_RecommendDoc}
-
-Despite it being a pain point, spending time on documentation provides a
-justifiable return on investment.  Addresses \ref{P_Documentation} on
-documentation.
-
-Software packages were not quantitatively measured for traceability. In this
-project we asked developers to comment on traceability, specifically on their
-software package's documentation and how it fit into their development process.
-
-The interviewed developer of ESPResSo noted that all major additions to their
+\subsection{Prioritize Documentation and Usability} \label{Sec_RecommendDoc}
+
+Although documentation is difficult (as mentioned under pain point
+\ref{P_Documentation}), some LBM developers have realized that prioritizing time
+on documentation provides a justifiable return on investment. For instance, the
+interviewed developer of ESPResSo noted that all major additions to their
 package had accompanying changes to artifacts and documentation. They noted that
-considerable effort had been put into the documentation. They further commented
-that they want to lower the entry barrier for new developers, and because of
-that their package has a considerable amount of developer documentation. This
-documentation informs developers on how to get started, and orients them to the
-artifacts, source code, and system architecture, as well as how the software
-package build system works, and how the coupling between the simulation engine
-and the interface works. \wss{If ESPresSo scored high (I think it did), we can
-use that to further promote the value of documentation.  Also a threat to
-validity is that we have implicitly assumed documentation is valuable when
-measuring quality.}
-
-Finding ways to hasten updating documentation would increase the frequency of
-such updates and benefit both users and developers.  \wss{Same thing has come up
-in papers.  I'll try to find a reference for this.  This is also something we
-can point to for recommendations related to code generation.} The developer of
-OpenLB noted the use of documentation generators like Doxygen. It would be
-advisable for more projects to use such automatic document generation tools,
-since some projects do not do this. \wss{We can refer back to this for the
-recommendations.}
-
-Reusability. Provide API documentation, if applicable. Only one (ESPResSo) of
-the top five ranked packages provided API documentation.
-
-\wss{Address Usability (\ppref{P_Usability}): Better documentation, with
-attention on traceability, and automatic code generation are suggested when
-designing software for change, and would help with these modifications.  So far
-this problem of on-boarding new users has been addressed by updating the
-documentation to better inform users of the underlying LBM theory and package
-requirements. Similar issues with LBM parameters were noted by another
-developer. Updating the user interface to better explain theoretical principles,
-as well as test user input for compatibility, was the implemented solution. As
-noted above, sometimes frequently asked questions on the underlying theory and
-on how to use the software are answered in the documentation.  The interviewed
-developer of ESPResSo commented that parts of their package's source code had
-been refactored to Python to help address usability issues. Python was perceived
-as a much more usable language, and it would be easy for future users and
-developers to learn and understand the source code.}
-
-One software package (ESPResSo) changed some of its scripting language to Python
-to make it more usable. The developer commented that this was ``the biggest step
-in terms of usability over the years'', further commenting that ``most people in
-the field know [Python]'' and that ``it's easy to learn''.  \wss{A generative
-approach facilitates switching to whatever language is currently popular.}
+considerable effort was put into their documentation. They further commented
+that their goal was to lower the entry barrier for new developers, which meant
+providing considerable developer documentation. This documentation informs
+developers on how to get started; orients them to the artifacts, source code,
+system architecture, and build system; and, explains the coupling between
+the simulation engine and the interface. In addition, ESPResSo developers
+included API documentation, although they are the only one of the top five
+packages to do so.
+
+Furthermore, documentation is used by LBM developers to address usability
+challenges (\ref{P_Usability}), as follows:
 
 \begin{itemize}
-\item Usability. Understandability. Unambiguity. State appropriate fluid
-dynamics problems that the software is designed to model in the documentation,
-and explicitly state the limits of the software. This is done in the ESPResSo
-user guide.
-\item Usability. Provide documentation that details the background theory
-information, or provide a reference to such information. This was done by
-several top ranked packages, including ESPResSo, Ludwig, LUMA, and OpenLB.
-\item Usability. Identify expected user characteristics. LIMBES, Ludwig,
-laboetie, and Palabos did this. The importance of specifying user
-characteristics is discussed in \citep{smith2007requirements}.
-\item Usability. Keep all documentation in one location. This was done by top
+	\item Explicitly state appropriate fluid dynamics problems that the software
+	is designed to model and explicitly state the limits of the software. This
+	is done in the ESPResSo user guide.
+	\item Provide documentation that details the background theory, or provide a
+	reference to such information. This was done by several top ranked packages,
+	including ESPResSo, Ludwig, LUMA, and OpenLB.
+	\item Identify expected user characteristics. LIMBES, Ludwig, laboetie, and
+	Palabos all did this. Explicit user characteristics are important so the
+	application and documentation can be pitched toward the right user
+	background \citep{smith2007requirements}.
+	\item Keep all documentation in one location. This was done by all top
 	ranked packages.
-\item Usability. Do not duplicate artifact information. This is suggested as
-good software development practice. Duplicate information is difficult to
-maintain. 
+	\item Use documentation generators, like Doxygen, as done (for instance) by
+	OpenLB.  More details on the packages that use generators is given in
+	Sections~\ref{SecSurfCorrectAndVerifiab} and~\ref{Sec_CompareTools}.
+	\item Provide an explanation of theoretical principles through the user interface.
+	\item Add guards in the code to test user inputs for compatibility before
+	proceeding with the calculations.
+	\item Include documentation on frequently asked questions (Section~\ref{Sec_CompareArtifacts}).
 \end{itemize}
@@
 
-Borrow from SE for HEP paper.
-
-Answers \rqref{RQ_Recommend}.  practices for the future.  Probably the easiest
-approach is to list the practice and then the PPs that it helps.  Our
-recommendations are not lists of what should have been done in the past, or what
+In this section we provide recommendations to address the pain points from
+Section~\ref{painpoints} to answer~\rqref{RQ_Recommend}.  Our recommendations
+are not lists of criticisms what should have been done in the past, or what
 should be done now; they are just suggestions for consideration in the future.
-
-``unique challenges of scientific research (e.g., the frequent and unforeseen
-changes in requirements ...)'' \citep{PintoEtAl2018}. We can point out that in
-the future this viewpoint could change and why.  Not really a unique challenge.
-Cite Smith 2016?
-
-linters \citep{Zadka2018} were rarely observed in LBM projects.  Do not see much evidence of their presence in best practice guides.
-
-\wss{Address lack of development time (\ppref{P_LackDevTime}): Aside from
-on-boarding new developers, time constraints could be mitigated by increasing
-developer efficiency, which could be addressed in several ways, including by
-improving the quality of documentation, or incorporating automatic code
-generation.}
-
-\wss{Address Correctness (\ppref{P_Correctness}): Testing is not the only way to
-build confidence in correctness.  Peer review.  }
-
-\wss{Address lack of funding (\ppref{P_LackFunding}): The developer further
-commented that there are no journals that publish such research software
-source code. However, there are ways to get such source code cited. Work has
-been done to address this in \citep{SmithEtAl2016-softcite}, which presents a
-set of software citation principles and discusses ``how they could be used to
-implement software citation in the scholarly community''
-\citep{katz2019software}. other papers also mention this pain point.
-Solutions proposed by Katz, journal of open source etc. Journal of Open Source
-Software \citet{SmithEtAl2018-joss}.  Calling for direct software citations
-\citep{SmithEtAl2016-softcite}.  Guidance on software citation from
-\citet{KatzEtAl2021}.  Discussed in CarverEtAl2021 paper.}
-
-\wss{Address quality of documentation (\ppref{P_Documentation}): Make required
-background of users explicit, possibly in a requirements document.}
+We will not be repeating ideas here that have been discussed in previous
+section, such as continuous integration, documentation of APIs, etc.  Our aim is
+to mention ideas that are at least somewhat beyond conventional best practices.
+The ideas listed have the potential to become best practices in the medium to
+long-term.  The ideas are provided roughly in the order of increasing novelty.
+
+\subsection{Employ Linters} \label{Sec_Linters}
+
+With the exception of \citet{ThielEtAl2020}, the research software guidelines
+that we consulted do not mention linters.  However, we feel linters have the
+potential to improve code quality at a relatively low cost.  A linter is a tool
+that statically analyzes code to find programming errors, stylistic
+inconsistencies and suspicious constructs \citep{Wikipedia2022_Lint}. Linters
+can be used to spot check code files, or even better as part of a continuous
+integration system.
+
+A sample of the potential benefits of linters include: finding memory leaks,
+finding potential bugs, standardizing code with respect to formatting, improving
+performance, removing silly errors before code reviews, and catching potential
+security issues \citep{SourceLevel2022_Lint}.  Linters are available for most
+popular programming languages.  For instance Python has the options of PyLint,
+flake8 and Black \citep{Zadka2018}.
+
+Linters can address the pain points for LBM developers.  For instance, a linter
+can increase the amount of development time (\ppref{P_LackDevTime}) by
+decreasing the number of mundane mistakes programmers have to catch.  Since the
+linter can include rules that capture the wisdom of senior programmers, it can
+help newer developers avoid common mistakes (\ref{P_LackSoftDevExp}).  Although
+there do not appear to be many linters for checking parallel programming
+mistakes (\ppref{P_TechnologyHurdles}), Parallel Lint is an option for OpenMP.
+If a linter is used consistently this can guard against some technical debt
+(\ppref{P_TechDebt}). Although linters are thought of as tools for code
+analysis, similar ideas can be applied to documentation to ensure that at least
+the basic documentation rules are followed.  The use of tools to check
+documentation is a partial explanation for the relatively higher quality of
+statistical tools that are part of the Comprehensive R Archive Network (CRAN)
+\citep{SmithEtAl2018_StatSoft}.
+
+\subsection{Conduct Rigorous Peer Reviews}
+
+Peer review is already a strategy adopted by some LBM developers to improve
+confidence in their software (Section~\ref{SecReproducibility}), but the
+benefits of peer review can be improved by making its application more rigorous.
+\citet{Jones2008} shows that rigorous inspection finds 60-65\% of latent defects
+on average, and often tops 85\% in defect removal efficiency.  By way of
+comparison, most forms of testing average between 30 and 35\% for defect removal
+efficiency \citep{EbertAndJones2009, Jones2008}.  A formal code inspection
+involves asking the reviewer to either follow a review checklist (check
+consistency of variable names, look for terminating loops, etc.), or perform
+specific review tasks (like summarize the purpose of the code, create a data
+dictionary for the module assigned, cross reference the code to the technical
+manual, etc.) The task based inspection approach has been effectively used for
+research software, as described by \citet{KellyAndShepard2000}. Task based
+inspection is an ideal fit with an issue tracking system, like GitHub.  The
+review tasks can be issues, so that they can be easily assigned, monitored and
+recorded.  Other potential issues for the tracker include assigning junior team
+members to test installation instructions and getting-started tutorials.
+
+Rigorous peer review addresses the same pain points as linters
+(Section~\ref{Sec_Linters}): \ppref{P_LackDevTime}, \ref{P_LackSoftDevExp},
+\ppref{P_TechnologyHurdles}), \ppref{P_TechDebt} and \ppref{P_Correctness}. Peer
+review addresses these pain points by efficiently searching for defects and
+problems.
+
+\subsection{Write and Submit More Papers on Software}
+
+To address the pain point of a lack of funding (\ppref{P_LackFunding}), LBM
+developers should watch for new opportunities to publish their research software
+source code. For instance, the Journal of Open Source Software (JOSS) reviews
+and publishes articles that credit the scholarship contained in the software
+itself \citep{SmithEtAl2018-joss}.  \citep{SmithEtAl2016-softcite} presents a
+set of software citation principles that may encourage broad adoption of a
+consistent policy for software citation across disciplines and venues.
+\citep{ChueHongEtAl2019} provides a software citation checklist for developers
+to make a release of their software citable.  Further guidance on software
+citation is provided by \citet{KatzEtAl2021}.  Another option is to use GitHub
+for citation, since it is relatively easy to add metadata to repositories and to
+generate citations for those repositories \citet{Smith2021-CitationsOnGitHub}.
+
+\subsection{Grow the Number of Contributors}
+
+Addresses lack of development time (\ppref{P_LackDevTime}) by increasing the
+number of developers.
+
+Maintainability. Have a defined process for accepting contributions, and make
+public documentation for making contributions to the project. The top five
+ranked software packages include information on how to contribute.
+
+Papers that talk about how to do open source. \citep{Jalan2016}
+
+Get more developers by better documentation, tutorials, etc.
+
+Combine existing teams?
+
+\subsection{Further Address the Oracle Problem} \label{Sec_FurtherOracle}
+
+- say what PPs are addressed
+- further steps to address the oracle problem - borrow from SE for HEP paper.
+- anything in Smith 2016?
+- introduce list on page 11 of HEP paper, add repeatability point
+
+In the absence of a test oracle, the following
+techniques can be used to build system tests \citep{Smith2016}:
+
+\begin{itemize}
+
+\item Test cases can be selected that are a subset of the real problem for which
+  a closed-form solution exists.  When using this approach, confidence in the
+  actual production code can only be built if it is the same code used for
+  testing; that is, nothing is gained if a separate, simpler, program is written
+  for testing the special cases.
+\item Verification test cases can be created by assuming a solution and using
+  this to calculate the inputs.  For instance, for a linear solver, if $A$ and
+  $x$ are assumed, $b$ can be calculated as $b = A x$.  Following this, $A x^* =
+  b$ can be solved and then $x$ and $x^*$, which should theoretically be equal,
+  can be compared.  In the case of solving Partial Differential Equations
+  (PDEs), this approach is called the Method of Manufactured Solutions
+  \citep{Roache1998}.
+\item Most scientific software uses floating point arithmetic, but for testing
+  purposes, the slower, but guaranteed correct, interval arithmetic
+  \citep{Hickey2001} can be employed. The faster floating point algorithm can
+  then be verified by ensuring that the calculated answers lie within the
+  guaranteed bounds.
+\item Verification tests should include plans for convergence studies.  The
+  discretization used in the numerical algorithm should be decreased (usually
+  halved) and the change in the solution assessed.
+\item Confidence can be built in a numerical algorithm by comparing the results
+  to another program that overlaps in functionality.  If the test results do not
+  agree, then one, possibly both of the programs is incorrect.
+\item The verification plan should also include test plans for nonfunctional
+  requirements, like accuracy, performance and portability, if these are
+  important implementation goals.  Performance tests can be planned to describe
+  how the software responds to changing inputs, such as problem size, condition
+  number etc.  Verification plans can include relative comparisons between
+  the new implementation and competing products \citep{Smith2006}.
+\item In cases where a test oracle is not available, as is usually the case for
+  scientific software, \emph{metamorphic} testing can be used.  ``This technique
+  operates by checking whether the program behaves according to a certain set of
+  properties called metamorphic relations.  A metamorphic relation is a
+  relationship between multiple inputs and output pairs of the program.''
+  \citep{KanewalaAndLundgren2016}.
+\end{itemize}
+
+\subsection{Augment Theory Manual to Include Requirements}
+
+Point to Section~\ref{Sec_CompareArtifacts}. Requirements documentation is
+feasible, as shown by the prevalence of theory documents in LBM software.
+Table~\ref{artifactspresent} shows that theory documentation is common.  We
+observed 17 of 24 packages had at least some theory documentation. Templates for
+scientific requirements, like \citet{SmithEtAl2007, SmithAndLai2005}, show
+theory is a significant part of the requirements documentation. With the
+addition of some extra information, the theory documents can be transformed into
+requirements specification.  The key extra information includes explicitly
+stating user characteristics, explicitly stating how the user interacts with the
+software in terms of input data requirements, and listing likely and unlikely
+changes.  Explicit statements about likely changes are invaluable in the design
+stage, since they provide developers guidance on how general the software needs
+to be.  For those wishing to maximize the value of a requirement specification,
+information could be added like prioritizing the nonfunctional requirements (to
+show the relative importance between qualities like portability, reliability and
+performance) and traceability information (to show the consequences of any
+changes to the assumptions).
+
+- say what PPs are addressed
+- previous section discusses challenges with requirements and not popular (maybe
+move that section here) - say what information to incorporate, low hanging fruit
+- return on investment
+- maybe borrow from SE for HEP
+- toward goal of replicability \citep{BenureauAndRougier2017}.
+
+Addresses lack of development time (\ppref{P_LackDevTime}) by getting developers
+up to speed more quickly.  Reducing the amount of time needed to train new
+developers.
 
@@
 
-Test the installation and tutorials.  Assign a junior team member the task?
-
-\subsection{Grow the Number of Contributors}
-
-Maintainability. Have a defined process for accepting contributions, and make
-public documentation for making contributions to the project. The top five
-ranked software packages include information on how to contribute.
-
-Papers that talk about how to do open source.
-
-\wss{in recommendations could point out what teams could do to get more external
-contributors, and potentially a longer life for their project.}
-
-\subsection{Augment Theory Manual to Include Requirements}
-
-- previous section discusses challenges with requirements and not popular (maybe
-move that section here) - say what information to incorporate, low hanging fruit
-- return on investment
-
-\subsection{Verifiability}
-
-- further steps to address the oracle problem - borrow from SE for HEP paper.
-- anything in Smith 2016?
-
-\subsection{Reproducibility}
-
+\subsection{Improve Repeatability, Reproducibility and Replicability}
+
+- say what PPs are addressed
 - record versions of everything
@@
 
+\subsection{Generate All Things}
+
+- say what PPs are addressed Addresses lack of development time
+(\ppref{P_LackDevTime}) by automating the writing of documentation and code.
+
 \section{Threats To Validity} \label{threats}
@@
 true.
+
+a threat to validity is that we have implicitly assumed documentation is
+valuable when measuring quality.
+
+threat to validity is the person doing the measuring.  In the previous exercise
+we did do an experiment to measure this and convince ourselves that it isn't
+that significant a risk.
 
@@
 observed from studying the repository artifacts.
-
-\subsection{Future State Of The Practice Assessments} \label{futuresop}
-
-As mentioned in Section~\ref{threats}, the measures listed in our measurement
-template may not be broad enough to accurately capture some qualities. Adding or
-extending measures is worth considering. For example, usability experiments and
-performance benchmarks could be incorporated into the assessment.
-
-Section~\ref{SecReproducibility} noted that developers were not asked to comment
-on the reproducibility of their source code from their requirements
-specifications and design documentation. Adding this question to the interview
-guide should be considered in the next iteration of state of the practice
-assessments. Furthermore, as mentioned in Section~\ref{threats}, it was found
-that some pertinent information was not specified in public artifacts. The use
-of unit testing and continuous integration by several packages (OpenLB, pyLBM,
-TCLB) was only discovered during interviews with developers. Adding further
-questions to the interview guide regarding the measures that are on the
-measurement template could reduce instances of incorrect data being collected.
-This additional interview data could be analyzed and incorporated into the AHP
-ranking, ensuring quality designations more accurate represent the true quality
-of the software packages.  \wss{clarify the distinction between usability and
-understandability.}
-
-For the current project, we did not assess the stronger form of
-``reproducibility'' known as replicability.  A software project is considered
-replicable if the documentation alone (without the code) is sufficiently precise
-and complete to allow a third-part to re-obtain the results of the original code
-\citep{BenureauAndRougier2017}.  Replicability is difficult to achieve with
-current documentation practices.  This question should be considered in the next
-iteration of state of the practice assessments.
-
-The final result of the domain analysis will be tables of commonalities,
-variabilities, and parameters of variation of a program family.
-\cite{smith2008commonality} present a template for conducting a commonality
-analysis, which was referred to when conducting this work.
-\cite{weiss1998commonality} describes another commonality analysis technique for
-deciding the members of a program family. \cite{SmithAndChen2004} and
-\cite{SmithMcCutchanAndCarette2017} are examples of a commonality analysis for a
-family of mesh generating software and a family of material models,
-respectively. The steps to produce a commonality analysis are:
 
@@
 quality attributes, and to reduce software quality failures.
+
+Recommendations, starting from current best practices.
+
+\begin{enumerate}
+	\item \textbf{Follow the example of LBM packages that already follow best practices.}  Point to the table summarizing the winners for each quality.
+Review the practices from the previous section.
+Addresses PP?.
+	\item \textbf{Continuous integration.} Addresses PP?.
+	\item \textbf{Document API.} Addresses PP?.
+\end{enumerate}
 
@@
 
-Recommendations for future state of the practice assessments are made in Section
-~\ref{futuresop}.
-
 Our comparison may point out areas where some LBM software packages fall short
@@
 
+Future State Of The Practice Assessments
+
+As mentioned in Section~\ref{threats}, the measures listed in our measurement
+template may not be broad enough to accurately capture some qualities. Adding or
+extending measures is worth considering. For example, usability experiments and
+performance benchmarks could be incorporated into the assessment.
+
+Section~\ref{SecReproducibility} noted that developers were not asked to comment
+on the reproducibility of their source code from their requirements
+specifications and design documentation. Adding this question to the interview
+guide should be considered in the next iteration of state of the practice
+assessments. Furthermore, as mentioned in Section~\ref{threats}, it was found
+that some pertinent information was not specified in public artifacts. The use
+of unit testing and continuous integration by several packages (OpenLB, pyLBM,
+TCLB) was only discovered during interviews with developers. Adding further
+questions to the interview guide regarding the measures that are on the
+measurement template could reduce instances of incorrect data being collected.
+This additional interview data could be analyzed and incorporated into the AHP
+ranking, ensuring quality designations more accurate represent the true quality
+of the software packages.  \wss{clarify the distinction between usability and
+understandability.}
+
+For the current project, we did not assess the stronger form of
+``reproducibility'' known as replicability.  A software project is considered
+replicable if the documentation alone (without the code) is sufficiently precise
+and complete to allow a third-part to re-obtain the results of the original code
+\citep{BenureauAndRougier2017}.  Replicability is difficult to achieve with
+current documentation practices.  This question should be considered in the next
+iteration of state of the practice assessments.
+
+The final result of the domain analysis will be tables of commonalities,
+variabilities, and parameters of variation of a program family.
+\cite{smith2008commonality} present a template for conducting a commonality
+analysis, which was referred to when conducting this work.
+\cite{weiss1998commonality} describes another commonality analysis technique for
+deciding the members of a program family. \cite{SmithAndChen2004} and
+\cite{SmithMcCutchanAndCarette2017} are examples of a commonality analysis for a
+family of mesh generating software and a family of material models,
+respectively. The steps to produce a commonality analysis are:
+
+\section*{Acknowledgements}
+
+We would like to thank Ao Dong and Oluwaseun Owojaiye for many fruitful
+discussions on topics relevant to this paper.
+
 \bibliographystyle{ACM-Reference-Format}
@@
 
-- data on Mendeley
-
-- make recommendations in order of conservative to ``bleeding edge''
-- add pointer to project checklists (as sent to Jacques, as promoted during workshop)
-
-- value of tools to enforce good behaviour of developers, like R, like MOOSE infrastructure
+- value of tools to enforce good behaviour of developers, like R, like MOOSE
+infrastructure
 
@@
 contrast to the state of the practice of LBM software
+
+- ``unique challenges of scientific research (e.g., the frequent and unforeseen
+changes in requirements ...)'' \citep{PintoEtAl2018}. We can point out that in
+the future this viewpoint could change and why.  Not really a unique challenge.
+Cite Smith 2016?

### 2022-03-02 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Edits up to and including Section 10.6 Augment Theory Manual to Include Requirements Information

@@
 	application and documentation can be pitched toward the right user
-	background \citep{smith2007requirements}.
+	background \citep{SmithEtAl2007}.
 	\item Keep all documentation in one location. This was done by all top
@@
 
-Addresses lack of development time (\ppref{P_LackDevTime}) by increasing the
-number of developers.
-
-Maintainability. Have a defined process for accepting contributions, and make
-public documentation for making contributions to the project. The top five
-ranked software packages include information on how to contribute.
-
-Papers that talk about how to do open source. \citep{Jalan2016}
-
-Get more developers by better documentation, tutorials, etc.
-
-Combine existing teams?
+In our interviews with developers we heard they would like to have additional
+software contributors.  More developers would help with addressing the lack of
+development time pain point (\ppref{P_LackDevTime}).  In investigating advice on
+increasing the number of contributors, we found that the advice usually starts
+with following best practices, including providing artifacts like those
+discussed in Section~\ref{Sec_CompareArtifacts} such as a contributor
+guidelines, a clear code of conduct and high quality code and documentation.
+This advice is logical, but in our assessment of LBM software we found examples
+of projects that already follow many best practices, but still have small
+development teams. Attracting developers apparently requires more than just
+building a good product.
+
+Some potential additional ideas for growing the number of contributors are
+as follows:
+
+\begin{itemize}
+	\item Clearly identify issues that are an appropriate starting points for
+	new developers \citep{Garcia2016, Jalan2016, Proffitt2017}.
+	\item Given that in most projects developers were once users
+	\citep{McQuaid2018}, recruit future developers from the current set of
+	users.
+	\item Create templates for pull requests and issues \citep{Jalan2016}.
+    \item Welcome all kinds of contributions, not just code.  Non-code
+    contributions include documentation, fixing typos, issue reporting and test
+    cases \citep{Jalan2016, Proffitt2017}.
+    \item Reward and recognize new contributors, via small rewards like stickers
+    or shirts, or even just a simple shoutout or mention in a blogpost or on
+    social media \citep{Jalan2016, Proffitt2017}.
+    \item Recognize that open source is more about people than it will ever be
+    about code \citep{Jalan2016}.
+    \item Look beyond just recruiting online and seek new developers at
+    conferences or other user meet ups \citep{Garcia2016}.
+    \item Follow the advice of \citet{Kuchner2012} to adapt ideas from marketing
+    to promote science.  Specific ideas that could be applied to open source
+    projects include the following: 
+    \begin{itemize}
+		\item Storytelling to motivate interest in the project, where the story
+		is a sequence of events and pauses for reflection \citep[p.\
+		21--22]{Kuchner2012}.
+		\item Invest effort in building relationships with users and with potential future developers.
+		\item Brand your project with what makes it unique and special.
+	\end{itemize}
+	\item Consider combining existing projects to pool collective resources, and
+	reduce competition between different open source projects.
+\end{itemize}
 
@@
 
-- say what PPs are addressed
-- further steps to address the oracle problem - borrow from SE for HEP paper.
-- anything in Smith 2016?
-- introduce list on page 11 of HEP paper, add repeatability point
-
-In the absence of a test oracle, the following
-techniques can be used to build system tests \citep{Smith2016}:
+To address the pain point of ensuring correctness (\ref{P_Correctness}),
+Section~\ref{SecReproducibility} outlined techniques that LBM developers use to
+circumvent the oracle problem for testing.  Additional techniques to build
+system tests are as follows \citep{Smith2016}:
 
@@
 
-\item Test cases can be selected that are a subset of the real problem for which
-  a closed-form solution exists.  When using this approach, confidence in the
-  actual production code can only be built if it is the same code used for
-  testing; that is, nothing is gained if a separate, simpler, program is written
-  for testing the special cases.
-\item Verification test cases can be created by assuming a solution and using
-  this to calculate the inputs.  For instance, for a linear solver, if $A$ and
+\item Create test cases by assuming a solution and using this solution to
+  calculate the necessary inputs.  For instance, for a linear solver, if $A$ and
   $x$ are assumed, $b$ can be calculated as $b = A x$.  Following this, $A x^* =
@@
   \citep{Roache1998}.
-\item Most scientific software uses floating point arithmetic, but for testing
-  purposes, the slower, but guaranteed correct, interval arithmetic
-  \citep{Hickey2001} can be employed. The faster floating point algorithm can
-  then be verified by ensuring that the calculated answers lie within the
-  guaranteed bounds.
-\item Verification tests should include plans for convergence studies.  The
-  discretization used in the numerical algorithm should be decreased (usually
-  halved) and the change in the solution assessed.
-\item Confidence can be built in a numerical algorithm by comparing the results
-  to another program that overlaps in functionality.  If the test results do not
-  agree, then one, possibly both of the programs is incorrect.
-\item The verification plan should also include test plans for nonfunctional
-  requirements, like accuracy, performance and portability, if these are
-  important implementation goals.  Performance tests can be planned to describe
-  how the software responds to changing inputs, such as problem size, condition
-  number etc.  Verification plans can include relative comparisons between
-  the new implementation and competing products \citep{Smith2006}.
-\item In cases where a test oracle is not available, as is usually the case for
-  scientific software, \emph{metamorphic} testing can be used.  ``This technique
-  operates by checking whether the program behaves according to a certain set of
-  properties called metamorphic relations.  A metamorphic relation is a
-  relationship between multiple inputs and output pairs of the program.''
-  \citep{KanewalaAndLundgren2016}.
+\item Use interval arithmetic to build test cases.  For testing purposes, the
+  slow, but guaranteed correct, interval arithmetic \citep{Hickey2001} can be
+  employed. The faster floating point algorithm can then be verified by ensuring
+  that the calculated answers lie within the guaranteed bounds.
+\item Include convergence studies.  The discretization used in the numerical
+  algorithm should be decreased (usually halved) and the change in the solution
+  assessed.  This is very likely already done by LBM developers, but it did not
+  come up during our interviews.
+\item Use \emph{metamorphic} testing, where the program is checked to see if it
+  satisfies a set of metamorphic relations, where a metamorphic relation is a
+  relationship between multiple inputs and output pairs
+  \citep{KanewalaAndLundgren2016}.  An example could be the relationship that
+  the output flow rate increases as the input flow rate increases.
 \end{itemize}
@@
 
-Point to Section~\ref{Sec_CompareArtifacts}. Requirements documentation is
-feasible, as shown by the prevalence of theory documents in LBM software.
-Table~\ref{artifactspresent} shows that theory documentation is common.  We
-observed 17 of 24 packages had at least some theory documentation. Templates for
-scientific requirements, like \citet{SmithEtAl2007, SmithAndLai2005}, show
-theory is a significant part of the requirements documentation. With the
-addition of some extra information, the theory documents can be transformed into
-requirements specification.  The key extra information includes explicitly
-stating user characteristics, explicitly stating how the user interacts with the
-software in terms of input data requirements, and listing likely and unlikely
-changes.  Explicit statements about likely changes are invaluable in the design
-stage, since they provide developers guidance on how general the software needs
-to be.  For those wishing to maximize the value of a requirement specification,
-information could be added like prioritizing the nonfunctional requirements (to
-show the relative importance between qualities like portability, reliability and
-performance) and traceability information (to show the consequences of any
-changes to the assumptions).
-
-- say what PPs are addressed
-- previous section discusses challenges with requirements and not popular (maybe
-move that section here) - say what information to incorporate, low hanging fruit
-- return on investment
-- maybe borrow from SE for HEP
-- toward goal of replicability \citep{BenureauAndRougier2017}.
-
-Addresses lack of development time (\ppref{P_LackDevTime}) by getting developers
-up to speed more quickly.  Reducing the amount of time needed to train new
-developers.
-
-Surface Correctness and Verifiability. Use a requirements
-specification document. This is suggested in SCS literature, and several of
-the top ranked packages had such a document or reference to theory manuals.
-A potential template is presented in \citep{smith2005new}.
+As discussed in Section~\ref{Sec_CompareArtifacts}, requirements specifications
+are rarely written for research software. Although a full requirements
+specification has advantages \citep{SmithEtAl2007, SmithAndLai2005},
+given the lack of development time (\ref{P_LackDevTime}), we recommend the
+intermediate step of augmenting the existing theory manuals with some
+requirements information.  Table~\ref{artifactspresent} shows that theory
+documentation is common.  We observed 17 of 24 packages had at least some theory
+documentation. Templates for scientific requirements, like \citet{SmithEtAl2007,
+SmithAndLai2005}, show theory is a significant part of the requirements
+documentation. With the addition of some extra information, the theory documents
+can be transformed into requirements specification.  The key extra information
+includes explicitly stating user characteristics, explicitly stating how the
+user interacts with the software in terms of input data requirements, and
+listing likely and unlikely changes.  Explicit statements about likely changes
+are invaluable in the design stage, since they provide developers guidance on
+how general the software needs to be.  For those wishing to maximize the value
+of a requirement specification, information could be added like prioritizing the
+nonfunctional requirements (to show the relative importance between qualities
+like portability, reliability and performance) and traceability information (to
+show the consequences of changes to the assumptions). Spending time to
+incorporate additional information in the theory manuals should help address the
+quality of existing documentation (\ref{P_Documentation})  and improve the
+software's usability (\ref{P_Usability}).
 
 \subsection{Improve Repeatability, Reproducibility and Replicability}
+
+- requirements are needed for the goal of replicability
+\citep{BenureauAndRougier2017}.
 

### 2022-03-02 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
updates

@@
 
-\subsection{Augment Theory Manual to Include Requirements}
+\subsection{Augment Theory Manual to Include Requirements Information}
 
@@
 specification has advantages \citep{SmithEtAl2007, SmithAndLai2005},
-given the lack of development time (\ref{P_LackDevTime}), we recommend the
+given the lack of development time (\ppref{P_LackDevTime}), we recommend the
 intermediate step of augmenting the existing theory manuals with some
@@
 
-\subsection{Improve Repeatability, Reproducibility and Replicability}
-
+\subsection{Improve Re-runnability, Repeatability, Reproducibility and Replicability}
+
+- part of correctness (\ppref{P_Correctness})
 - requirements are needed for the goal of replicability
@@
 
-- say what PPs are addressed
-- record versions of everything
-- Docker, Virtual machines
-
-Repeatability - A correct output may not be exactly reproducible, as it may be
-dependent on a probability distribution, so strictly comparing results may not
-be appropriate. \wss{Ideas for dealing with reproducibility in these situations
-will be explored in the recommendations section (like properties of the
-solution, such as conservation of mass).  5R paper mentions repeatability -
-design it so that the random number seed can be set.}
+\begin{description}
+
+	\item[Re-runnability] ``A re-runnable code is one that can be run again when
+	needed, and in particular more than the one time that was needed to produce
+	the results. It is important to notice that the re-runnability of a code is
+	not an intrinsic property. Rather, it depends on the context, and becomes
+	increasingly difficult as the code ages. Therefore, to be and remain
+	re-runnable on other researchers' computers, a re-runnable code should
+	describe---with enough details to be recreated---an execution environment in
+	which it is executable.''  - record versions of everything - Docker, Virtual
+	machines
+
+	\item[Repeatability] Repeatability - A correct output may not be exactly
+	reproducible, as it may be dependent on a probability distribution, so
+	strictly comparing results may not be appropriate.  ``The code is running
+	and producing the expected results. The next step is to make sure that you
+	can produce the same output over successive runs of your program. In other
+	words, the next step is to make your program deterministic, producing
+	repeatable output. Repeatability is valuable. If a run of the program
+	produces a particularly puzzling result, repeatability allows you to
+	scrutinize any step of the execution of the program by re-running it again
+	with extraneous prints, or inside a debugger. Repeatability is also useful
+	to prove that the program did indeed produce the published results.
+	Repeatability is not always possible or easy. But for sequential programs
+	not depending on analog inputs, it often comes down to controlling the
+	initialization of the pseudo- random number generators (RNG).''
+
+	\item[Reproducibility] ``To recap, reproducibility implies re-runnability
+	and repeatability and availability, yet imposes additional conditions.
+	Dependencies and platforms must be described as precisely and as
+	specifically as possible. Parameters values and inputs should accompany the
+	result files. The data and scripts behind the graphs must be published. Unit
+	tests are a good way to embed self-diagnostics of reproducibility in the
+	code. Reproducibility is hard, yet tremendously necessary.''
+
+	\item[Replicability] ``Replicability is the implicit assumption that any
+	article that does not provide the code source makes: that the description it
+	provides of the algorithms is sufficiently precise and complete to re-obtain
+	the results it presents.'' Requires requirements.  For the current project,
+	we did not assess the stronger form of ``reproducibility'' known as
+	replicability.  A software project is considered replicable if the
+	documentation alone (without the code) is sufficiently precise and complete
+	to allow a third-part to re-obtain the results of the original code
+	\citep{BenureauAndRougier2017}.  Replicability is difficult to achieve with
+	current documentation practices.  This question should be considered in the
+	next iteration of state of the practice assessments.
+
+	As an example, \citet{IonescuAndJansson2013} summarize the problem that
+	researchers had in reproducing the economic theory results obtained by Gintis in
+	2006.  The scientific method requires that another researcher be able to
+	independently replicate previous research results.
+
+\end{description}
 

### 2022-03-03 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
First draft complete of Section 10 (Recommendations)

--- 2022-03-02 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
@@
 
-To address technical debt (\pref{P_TechDebt}), the top LBM developers show how
+To address technical debt (\ppref{P_TechDebt}), the top LBM developers show how
 modularization can be used to design research software for future change.
@@
 The top LBM projects have adopted techniques for software verification that get
-around the oracle problem (mentioned earlier in Section~\ref{P_Correctness}). In
+around the oracle problem (mentioned earlier in Section~\ppref{P_Correctness}). In
 the absence of a general test oracle, the following techniques have been adopted
@@
 Although documentation is difficult (as mentioned under pain point
-\ref{P_Documentation}), some LBM developers have realized that prioritizing time
+\ppref{P_Documentation}), some LBM developers have realized that prioritizing time
 on documentation provides a justifiable return on investment. For instance, the
@@
 Furthermore, documentation is used by LBM developers to address usability
-challenges (\ref{P_Usability}), as follows:
+challenges (\ppref{P_Usability}), as follows:
 
@@
 linter can include rules that capture the wisdom of senior programmers, it can
-help newer developers avoid common mistakes (\ref{P_LackSoftDevExp}).  Although
+help newer developers avoid common mistakes (\ppref{P_LackSoftDevExp}).  Although
 there do not appear to be many linters for checking parallel programming
@@
 Rigorous peer review addresses the same pain points as linters
-(Section~\ref{Sec_Linters}): \ppref{P_LackDevTime}, \ref{P_LackSoftDevExp},
+(Section~\ref{Sec_Linters}): \ppref{P_LackDevTime}, \ppref{P_LackSoftDevExp},
 \ppref{P_TechnologyHurdles}), \ppref{P_TechDebt} and \ppref{P_Correctness}. Peer
@@
 
-To address the pain point of ensuring correctness (\ref{P_Correctness}),
+To address the pain point of ensuring correctness (\ppref{P_Correctness}),
 Section~\ref{SecReproducibility} outlined techniques that LBM developers use to
@@
 incorporate additional information in the theory manuals should help address the
-quality of existing documentation (\ref{P_Documentation})  and improve the
-software's usability (\ref{P_Usability}).
-
-\subsection{Improve Repeatability, Reproducibility and Replicability}
-
-- requirements are needed for the goal of replicability
+quality of existing documentation (\ppref{P_Documentation})  and improve the
+software's usability (\ppref{P_Usability}).
+
+\subsection{Improve Re-runnability, Repeatability, Reproducibility and Replicability}
+
+Ensuring correctness was identified as a pain point by developers
+(\ppref{P_Correctness}).  Inspired by \citet{BenureauAndRougier2017}, we make
+recommendations to ensure that correctness can be maintained into the future by
+prioritizing 4 increasingly sophisticated levels of replication.
+
+\begin{description}
+
+	\item[Re-runnability] means that today's code can be run again in the future
+	when needed.  Re-runnability becomes more difficult as code ages and the
+	software and hardware infrastructure around it changes, like changes to
+	compilers, GPUs, libraries etc.  To be re-runnable on other researchers'
+	computers, a code should provide a detailed description of the execution
+	environment in which it is executable.  Besides recording details, virtual
+	machines, docker and environment managers, like Conda, can be used to
+	improve re-runnability.
+
+	\item[Repeatability] means producing the same output over successive runs of
+	a program.  As was pointed out during our developer interviews, this can be
+	difficult for LBM software because of dependence on probability
+	distributions.  However, deterministic output is important for testing
+	purposes and for debugging problematic code.  For LBM software repeatability
+	will likely require running a serial version of the code with full control
+	of the initialization of the pseudo-random number generators.
+
+	\item[Reproducibility] Once re-runnability and repeatability are achieved,
+	the next step is reproducibility.  Current work by LBM developers on
+	reproducibility is discussed in Section~\ref{SecReproducibility}.  The
+	proposed goal for the future is for all results to be reproducible, which
+	means documenting dependencies, platforms, parameter values and input files.
+	The data and scripts behind the graphs should be published
 	\citep{BenureauAndRougier2017}.
 
-- say what PPs are addressed
-- record versions of everything
-- Docker, Virtual machines
-
-Repeatability - A correct output may not be exactly reproducible, as it may be
-dependent on a probability distribution, so strictly comparing results may not
-be appropriate. \wss{Ideas for dealing with reproducibility in these situations
-will be explored in the recommendations section (like properties of the
-solution, such as conservation of mass).  5R paper mentions repeatability -
-design it so that the random number seed can be set.}
+	\item[Replicability] means that the documentation provided for the software,
+	not the software itself, is sufficient for an independent third party to
+	reproduce the computational results.  Reproducibility does not help us if
+	our trust in the original code is in doubt.  What if we need to reproduce
+	the results not starting from the code, but starting from the original
+	theory? Unfortunately, multiple examples exist \citep{CrickAndHall2014,
+	IonescuAndJansson2013} where results from research software could not be
+	independently reproduced, due to a need for local knowledge missing from
+	published documents.  Examples of missing knowledge includes missing
+	assumptions, derivations, and undocumented modifications to the code.  In
+	the future, we wish for it to be easier to independently replicate the work
+	of others, starting from their theory documentation, augmented with
+	requirements information as discussed above, without needing to rely on
+	their code.
+
+\end{description}
 
@@
 
-- say what PPs are addressed Addresses lack of development time
-(\ppref{P_LackDevTime}) by automating the writing of documentation and code.
+We propose automatically generating LBM code and its documentation, using a
+Domain Specific Modelling (DSM) approach. DSM means creating a knowledge base of
+models for physics, computing, mathematics, documentation and certification and
+then writing explicit ``recipes'' that weave together this knowledge to generate
+the desired code, documentation, test cases, inspection reports and build
+scripts. This definition of DSM is more general than usual; in this
+recommendation DSM implies generation of all software artifacts, not just the
+code. DSM moves development to a higher level of abstraction; domain experts can
+work without concern for low-level implementation details. Using DSM, code and
+documentation can be optimally generated, redundancy is eliminated, the
+likelihood of errors is reduced and maintenance is automated. Moreover, a
+generative approach facilitates the inevitable exploration of LBM modelling
+assumptions.
+
+DSM provide a transformative technology for documentation, design and
+verification \citep{JohansonAndHasselbring2018, Smith2018}. DSM allows
+scientists to focus on their science, not software.  A generative approach
+removes the maintenance nightmare of documentation duplicates and near
+duplicates \citep{LucivEtAl2018}, since knowledge is only captured once and
+transformed as needed.  Code generation has previously been applied to improve
+research software.  For instance, ATLAS (Automatically Tuned Linear Algebra
+Software) \citep{WhaleyEtAl2001} and Blitz++ \citep{Veldhuizen1998} produces
+efficient and portable linear algebra software.  Spiral \citep{Pueschel2001}
+uses software/hardware generation for digital signal processing.
+\citet{Carette2008} shows how to generate a family of efficient, type-safe
+Gaussian elimination algorithms.  FEniCS (Finite Element and Computational
+Software) uses code generation when solving differential equations
+\citep{LoggEtAl2012}. \citet{OberEtAl2018} and \citet{MatkerimEtAl2013} apply
+DSM to High Performance Computing (HPC), using UML (Unified Modelling Language)
+for their domain models. Unlike previous DSM work, the current recommendation
+focuses on generating all software artifacts (requirements, design, etc.), not
+just code.  Initial work on this ``generate all things'' approach is discussed
+in \citep{SzymczakEtAl2016} with a prototype available
+\citep{CaretteEtAl2021-Drasil}.
+
+A DSM approach addresses multiple pain points.  For instance, once the
+infrastructure is in place, a DSM can decrease the amount of development time
+(\ppref{P_LackDevTime}) by automation.  Since a DSM approach allows scientists
+to focus on their science, rather than software, the lack of software
+development experience (\ppref{P_LackSoftDevExp}) will be less of an issue. The
+DSM approach can capture computing knowledge to mitigate the technology related
+pain points (\ppref{P_TechnologyHurdles}).  For ensuring correctness
+(\ppref{P_Correctness}) a generative approach can aim to be correct by
+construction.  In cases where mistakes are made, DSM has the advantage that the
+mistake is propagated throughout the generated artifacts, which greatly
+increases the chance that the mistake is noted.  Technical debt
+(\ppref{P_TechDebt}) is not a concern with a DSM approach since the recipes that
+are used for generation are written at a high level and are thus relatively easy
+to change.  With respect to pain points on usability (\ppref{P_Usability}) and
+documentation (\ppref{P_Documentation}), the generative approach we are
+proposing addresses these directly, since documentation is a primary concern,
+rather than an afterthought. and quality of documentation
 
 \section{Threats To Validity} \label{threats}
-
-Maintainability is assumed to be improved by a higher ratio of comments to
-source code (Section~\ref{Sec_Maintainability}), but this is not necessarily
-true.
-
-a threat to validity is that we have implicitly assumed documentation is
-valuable when measuring quality.
-
-threat to validity is the person doing the measuring.  In the previous exercise
-we did do an experiment to measure this and convince ourselves that it isn't
-that significant a risk.
-
-We assume that a large number of source files is associated with increased
-reusability (Section~\ref{reusabilityresults}), but this is not necessarily the
-case.
 
@@
 
-Overall ranking with equally weighted qualities isn't realistic (Section~\ref{Sec_OverallQuality}).
+Overall ranking with equally weighted qualities isn't realistic
+(Section~\ref{Sec_OverallQuality}).
 
@@
 observed from studying the repository artifacts.
+
+Maintainability is assumed to be improved by a higher ratio of comments to
+source code (Section~\ref{Sec_Maintainability}), but this is not necessarily
+true.
+
+a threat to validity is that we have implicitly assumed documentation is
+valuable when measuring quality.
+
+threat to validity is the person doing the measuring.  In the previous exercise
+we did do an experiment to measure this and convince ourselves that it isn't
+that significant a risk.
+
+We assume that a large number of source files is associated with increased
+reusability (Section~\ref{reusabilityresults}), but this is not necessarily the
+case.
 
@@
 Cite Smith 2016?
+
+- mention MOOSE when mention CRAN for improving quality

### 2022-03-03 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Completed revision of Section 11 (Threats to Validity)

@@
 the methodology.
-% of the next sections provide an overview of the specifics of identifying
-% (Section~\ref{identifysoftware}) and filtering (Section~\ref{filtersoftware})
-% the candidate software, our quantitative measurements
-% (Section~\ref{empiricalmeasures}), the AHP ranking process
-% (Section~\ref{AHP}), developer interviews (Section~\ref{SecSurvey}), and
-% interactions with one of the key members of the assessment team -- the Domain
-% Expert (Section~\ref{sec_vet_software_list}).
 
@@
 
-This section examines potential threats to the validity of this state of the
-practice assessment. These can be categorized into methodology and data
-collection issues. The goal of this assessment isn't to rank the software, but
-to use the ranking exercise as a means to understand the state of the practice
-of LBM software development.
-
-The measures listed in our measurement template may not be broad enough to
-accurately capture some qualities. For example, there are only two measures of
-surface robustness. The measurement of robustness could be expanded, as it
-currently only measures unexpected input. Other faults could be introduced, but
-could require a large investment of time to develop, and might not be a fair
-measure for all packages. Similarly, reusability is assessed along the number of
-code files and LOC per file. While this measure is indicative of modularity, it
-is possible that some packages have many files, with few LOC, but the files do
-not contain source code that is easily reusable. The files may be poorly
-formatted, or the source code may be vague and have ambiguous identifiers.
-Furthermore, the measurement of understandability relies on 10 random source
-code files. It is possible that the 10 files that were chosen to represent a
-software package may not be a good representation of the understandability of
-that package.
+The threats to validity can be categorized into methodology and data collection
+issues.  With respect to our methodology, the measurement template may not be
+detailed enough to accurately capture some qualities. For example, there are
+only two measures of surface robustness. The measurement of robustness could be
+expanded, as it currently only measures unexpected input. Other intentional
+faults could be introduced to assess robustness, but this would require a larger
+investment of measurement time. Similarly, reusability
+(Section~\ref{reusabilityresults}) is assessed via the number of code files and
+LOC per file. While this measure is indicative of modularity, it is possible
+that some packages have many files, with few LOC, but the files do not contain
+source code that is easily reusable. The files may be poorly formatted, or the
+source code may be vague and have ambiguous identifiers. Maintainability may
+also be incorrectly estimated, since it is assumed to be improved by a higher
+ratio of comments to source code (Section~\ref{Sec_Maintainability}), but this
+is not necessarily true. Furthermore, the measurement of understandability
+(Section~\ref{Sec_SurfUnderstandability}) relies on 10 random source code files.
+By random chance, the 10 files that were chosen may not be a good representation
+of the understandability of that package.
+
+In our methodology one individual measures each package and the beginning and
+ending of the measurement phase is separated by several months.  There is a
+threat that the measurement may include unconscious biases, or the reviewers
+judgement may drift over time.  In previous work \citep{SmithEtAl2016}, we
+demonstrated that the measurement process is reasonably reproducible by having 5
+products graded by a second reviewer. The ranking via this independent review
+was almost identical to the original ranking. As long as each grader uses
+consistent definitions, the relative comparisons in the AHP results will be
+consistent between graders.
 
 Regarding data collection, a risk to the validity of this assessment is missing
-or incorrect data. Some software package data may not have been measured due to
-technology issues like broken links. This issue arose with the measurement of
-Palabos, which had a broken link to its user manual, as noted in
-Section~\ref{Sec_CompareArtifacts}.
-
+or incorrect data. Some data may not have been measured due to technology issues
+like broken links. This issue arose with the measurement of Palabos, which had a
+broken link to its user manual, as noted in Section~\ref{Sec_CompareArtifacts}.
 Some pertinent data may not have been specified in public artifacts, or may be
-obscure within an artifact or web-page. The use of unit testing and continuous
-integration was mentioned in the artifacts of only three (ESPResSo, Ludwig,
-Musubi) packages. However, interviews suggested a more frequent use of both unit
-testing and continuous integration in the development processes than what was
-observed from the initial survey of the artifacts. For example, OpenLB, pyLBM,
-and TCLB use such methods during development despite this not being explicitly
-clear from an analysis of the material available online.
+obscure within an artifact or web-page. For instance, the use of unit testing
+and continuous integration was mentioned in the artifacts of only three
+(ESPResSo, Ludwig, Musubi) packages (Section~\ref{SecSurfCorrectAndVerifiab}).
+However, interviews suggested a more frequent use of both unit testing and
+continuous integration. For example, OpenLB, pyLBM, and TCLB use such methods
+during development despite this not being explicit from an analysis of the
+material available online.
 
@@
 in this assessment, but it is probable that such documentation is part of all
-LBM packages. After all, developing SCS is not a trivial endeavour. It is likely
-that many packages have such documentation but did not make it public, and due
-to this the measured data is not a true reflection of software package quality.
-
-\wss{There are more threats to validity.  Brainstorm.  Watch for them while 
-editing.  Look at Ao paper.}
-
-Problems with the measurement of qualities:
-
-\begin{itemize}
-\item The assumption that more code files is an indicator of reusability
-(Section~\ref{reusabilityresults}) may not be a valid assumption.
-\item Understandability was measured using only 10 random files
-(Section~\ref{Sec_SurfUnderstandability}).  This could be improved.
-\end{itemize}
-
-Overall ranking with equally weighted qualities isn't realistic
+LBM packages. After all, developing research software is not a trivial
+endeavour. It is likely that many packages have such documentation but did not
+make it public, and due to this the measured data may not be a true reflection
+of software package quality.
+
+Another data collection concern is the gap between measurement of Musubi and the
+measurement of all other packages (Section~\ref{identifysoftware}).
+Unfortunately we were not aware of the existence of Musubi until after the
+primary data collection period was completed. This gap has a secondary effect
+because it caused a mismatch between the time of manual and automated data
+collection (Section~\ref{repmetrics}).  With time and resource constraints we
+could not redo the manual measurement, but we did update the automated measures
+(like the number of files, LOC, etc.) when Musubi was added.
+
+Our comparison between our ranking and the community's ranking
+(Section~\ref{repmetrics}) may not be valid.  Using stars to estimate the
+community's ranking is not necessarily accurate.  Moreover, our overall AHP
+ranking makes the unrealistic assumption of equal weighting between qualities
 (Section~\ref{Sec_OverallQuality}).
 
-Threat to validity - the gap between measures of everything else and Musubi
-(Section~\ref{identifysoftware}), and then the gap between the manual measures
-and the automated measures (Section~\ref{repmetrics}).
-
-From Section~\ref{Sec_CompareTools}: ``Interviews with the developers revealed a
-potentially more frequent use of both unit testing and continuous integration,
-compared to what was observed from studying the repository artifacts.''
-
-Interviews suggested a more frequent use of both unit testing and continuous
-integration in the development processes than what was observed from the initial
-survey. For example, OpenLB, pyLBM, and TCLB use such methods during
-development, despite this not being explicitly clear from an analysis of the
-material available online.  The correctness and verifiability of such packages
-is not measured well using surface analysis.
-
-Section~\ref{repmetrics} compares ranking by best practices to the community's
-ranking.  Using stars to estimate the community's ranking is not necessarily
-accurate.
-
-Tool usage: Interviews with the developers revealed a potentially more frequent
-use of both unit testing and continuous integration, compared to what was
-observed from studying the repository artifacts.
-
-Maintainability is assumed to be improved by a higher ratio of comments to
-source code (Section~\ref{Sec_Maintainability}), but this is not necessarily
-true.
-
-a threat to validity is that we have implicitly assumed documentation is
-valuable when measuring quality.
-
-threat to validity is the person doing the measuring.  In the previous exercise
-we did do an experiment to measure this and convince ourselves that it isn't
-that significant a risk.
-
-We assume that a large number of source files is associated with increased
-reusability (Section~\ref{reusabilityresults}), but this is not necessarily the
-case.
+A final threat to validity is conceptual.  Our work assumes that following best
+practices leads to improved software quality.  As the advice in software
+engineering guidelines show, we are not alone in this assumption.  Ideally
+software qualities should be measured directly and then the impact of best
+practices on those qualities should be assessed.  However, researchers in this
+domain always face the challenge that many qualities, like maintainability,
+reusability and verifiability, are difficult to measure directly.  Therefore, we
+follow the usual approach of assuming that following procedures and adhering to
+standards yields a higher-quality product \citep[p.\ 112]{VanVliet2000}.
 
@@
 Section~\ref{softwarequalities}. A methodology for assessing the state of the
-practice of software development in SCS domains was presented. Domain packages
-were assessed to answer the software development related research questions
-listed in Section~\ref{secIntro} to understand how software quality is impacted
-by software development choices, including principles, processes, and tools.
-Software developers were interviewed to identify development pain points, and to
-identify how software quality is ensured. Quantitative data was used to rank the
-software packages using the AHP. The ranking designations were compared with
-rankings from the software development community, and we found that many of our
-top 10 ranked packages are ranked highly by a domain expert. Recommendations for
-improving software along quality metrics were made... The findings of this
-report can be used to guide future development of SCS, specifically along
-quality attributes, and to reduce software quality failures.
+practice of software development in scientific domains was presented. Domain
+packages were assessed to answer the software development related research
+questions listed in Section~\ref{secIntro} to understand how software quality is
+impacted by software development choices, including principles, processes, and
+tools. Software developers were interviewed to identify development pain points,
+and to identify how software quality is ensured. Quantitative data was used to
+rank the software packages using the AHP. The ranking designations were compared
+with rankings from the software development community, and we found that many of
+our top 10 ranked packages are ranked highly by a domain expert. Recommendations
+for improving software along quality metrics were made... The findings of this
+report can be used to guide future development of research software,
+specifically along quality attributes, and to reduce software quality failures.
 
@@
 
+The goal of this assessment isn't to rank the software, but to use the ranking
+exercise as a means to understand the state of the practice of LBM software
+development.
+
 We understand that software packages vary in goals, developers, funding, and

### 2022-03-03 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Edits to concluding remarks

@@
 	The data and scripts behind the graphs should be published
-	\citep{BenureauAndRougier2017}.
+	\citep{BenureauAndRougier2017}.  Some tools and techniques for computational
+	reproducibility are summarized in \citet{PiccoloAndFrampton2016}.
 
@@
 
-We analyzed the state of the practice of software development in the Lattice
-Boltzmann Methods software domain by quantitatively and qualitatively measuring,
-and comparing, 24 software packages along quality attributes. The software
-qualities that were assessed in this report are listed in
-Section~\ref{softwarequalities}. A methodology for assessing the state of the
-practice of software development in scientific domains was presented. Domain
-packages were assessed to answer the software development related research
-questions listed in Section~\ref{secIntro} to understand how software quality is
-impacted by software development choices, including principles, processes, and
-tools. Software developers were interviewed to identify development pain points,
-and to identify how software quality is ensured. Quantitative data was used to
-rank the software packages using the AHP. The ranking designations were compared
-with rankings from the software development community, and we found that many of
-our top 10 ranked packages are ranked highly by a domain expert. Recommendations
-for improving software along quality metrics were made... The findings of this
-report can be used to guide future development of research software,
-specifically along quality attributes, and to reduce software quality failures.
-
-Recommendations, starting from current best practices.
-
-\begin{enumerate}
-	\item \textbf{Follow the example of LBM packages that already follow best practices.}  Point to the table summarizing the winners for each quality.
-Review the practices from the previous section.
-Addresses PP?.
-	\item \textbf{Continuous integration.} Addresses PP?.
-	\item \textbf{Document API.} Addresses PP?.
-\end{enumerate}
+To potentially improve the quality of LBM software, and help future users select
+the right package, our goal was to analyze the current state of LBM software
+development, including determining how well current best practices are followed
+and finding pain points for developers.  We compared the state of the practice
+for LBM software to the development practices recommended and employed for
+research software in general. To improve the quality of LBM software we looked
+for examples where developers have been successful so that we can share these
+with the LBM community, and with the broader research software community.  We
+also aimed to identify potential approaches for mitigating the identified pain
+points in the future.
+
+Our methodology involved the following steps: i) Identify a list of 45 candidate
+software packages; ii) Filter the list to a length of 24 packages; iii) Collect
+repository related data on each software package, like number of stars, number
+of open issues, number of lines of code; iv) Fill in the measurement template
+(the template consists of 108 questions to assess 9 qualities (installability,
+correctness/verifiability, reliability, robustness, usability, maintainability,
+reusability, understandability, and visibility/transparency)   The repository
+data is collected by filling in a 108 question measurement template, which
+requires installing the software on a VM, running simple tests (like completing
+the getting started instructions (if present)), and searching the code,
+documentation and test files.; v) Rank the software using the Analytic Hierarchy
+Process (AHP); vi) Interview developers (the interview consists of 20 questions
+and takes about an hour); and, vii) Conduct a domain analysis. The collected
+data is analyzed by: a) comparing the ranking by best practices against the
+ranking by popularity; b) comparing artifacts, tools and processes to current
+research software development guidelines; and, c) exploring pain points.
+
+Table~\ref{topperformerstable} summarizes the top 5 performers for each of the qualities that we assessed.
 
@@
 
+Comparison to community.
+
+Comparison of artifacts.
+
+Comparison of tools.
+
+Comparison of processes.
+
+Developer pain Points
+
+Lessons from developers
+
+Recommendations, starting from current best practices.
+
+\begin{enumerate}
+	\item \textbf{Follow the example of LBM packages that already follow best practices.}  Point to the table summarizing the winners for each quality.
+Review the practices from the previous section.
+Addresses PP?.
+	\item \textbf{Continuous integration.} Addresses PP?.
+	\item \textbf{Document API.} Addresses PP?.
+\end{enumerate}
+
+Future work
+
 \wss{More on the lessons to learn from the top packages.  The comparison to
@@
 manually duplicate knowledge into multiple artifacts \citep{SzymczakEtAl2016}.
-
-Include a table that summarizes the top 4 projects for each quality, but an
-asterisk if still alive.
 
@@
 
-- value of tools to enforce good behaviour of developers, like R, like MOOSE
-infrastructure
-
-- introduce the idea of using our measurement template as a checklist, instead
-of, or in addition to other checklists - cite the other checklists
-\citep{SchlauchEtAl2018, BrettEtAl2021, vanGompelEtAl2016}, \href{Software
-Sustainability Institute Online
-Form}{https://docs.google.com/forms/d/e/1FAIpQLSf0ccsVdN-nXJCHLluJ-hANZlp8rDKgprJa0oTYiLZSDxh3DA/viewform}
-
-- mention flake8 for Python?
-
-- mention \href{https://joss.theoj.org/}{The Journal of Open Source Software}
-
-- reproducibility - can point to the PiccoloAndFrampton2016 paper
-
-- AlNoamanyAndBorghi2018 lists the most popular languages, Python is number 1
-
 - should look at ReinerEtAl2022 on Domain of Ocean Modelling software to
@@
 
-- ``unique challenges of scientific research (e.g., the frequent and unforeseen
-changes in requirements ...)'' \citep{PintoEtAl2018}. We can point out that in
-the future this viewpoint could change and why.  Not really a unique challenge.
-Cite Smith 2016?
-
 - mention MOOSE when mention CRAN for improving quality

### 2022-03-04 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Complete full draft of SOP for LBM paper

@@
     We analyze the state of software development practice for Lattice Boltzmann
-    solvers by quantitatively and qualitatively measuring and comparing 24
-    software packages for 10 software qualities (installability, correctness/
+    solvers and find by quantitatively and qualitatively measuring and comparing
+    24 software packages for 10 software qualities (installability, correctness/
     verifiability, reliability, robustness, usability, maintainability,
@@
     reproducibility). Our reproducible analysis method employs a measurement
-    template (containing 103 measures that are manually and automatically
+    template (containing 108 measures that are manually and automatically
     extracted from the project repository) and developer interviews (consisting
@@
     although the number one project by star count (Sailfish) is ranked 9th by
-    our methodology. Our top three packages were ESPResSo, Ludwig, and Palabos
-    (all in the top 10 by star count). We interviewed 4 developers to gain
-    insight into their current pain points. Identified challenges include lack
-    of development time, lack of funding, and difficulty with ensuring
-    correctness.  Based on our assessment, we make the following recommendations
-    for current and future LBM projects: provide a detailed user manual and
-    tutorials, explicitly state the limits of the software, use user-friendly
-    programming languages, consider a peer review process, communicate
-    development and contribution information, and use continuous integration and
-    project management tools.  \wss{Change the abstract to discuss how LBM
-    compares to common practices and best practices. Highlight some future
-    practices.  \wss{Follow the general research software trend of increasing
-    use of version control systems and project management on GitHub or
-    similar.}}
+    our methodology. Our finding is that the state of the practice is healthy
+    with 67\% of the measured packages ranking in the top 5 for at least one
+    quality, the majority of LBM generated artifacts corresponding to general
+    recommendations from research software developers, common utilization (67\%
+    of packages) of version control and adoption of a quasi-agile development
+    process.Areas of best practice to potentially improve include adoption of
+    continuous integration, API documentation and enforcement of programming
+    style guides.  We interviewed 4 developers to gain insight into their
+    current pain points. Identified challenges include lack of development time,
+    lack of funding, and difficulty with ensuring correctness.  Developers are
+    addressing these pain points by designing for change, circumventing the
+    oracle problem and prioritizing documentation and usability.  For future the
+    best practices of the future, we suggest the community consider: employing
+    linters, conducing rigorous peer reviews, writing and submitting more papers
+    on software, growing the number of contributors and augmenting the theory
+    manuals to include more requirements relevant information.
 
@@
 
-To potentially improve the quality of LBM software, and help future users select
-the right package, our goal was to analyze the current state of LBM software
-development, including determining how well current best practices are followed
-and finding pain points for developers.  We compared the state of the practice
-for LBM software to the development practices recommended and employed for
-research software in general. To improve the quality of LBM software we looked
-for examples where developers have been successful so that we can share these
-with the LBM community, and with the broader research software community.  We
-also aimed to identify potential approaches for mitigating the identified pain
-points in the future.
+To help users select the right LBM package for their needs, and to potentially
+provide direction for future LBM software development, our goal was to analyze
+the current state of LBM software development.  We aimed to determine how well
+best practices are followed and to understand current developer pain points.  We
+compared the state of the practice for LBM software to the development practices
+recommended and employed for research software in general. To improve the
+quality of LBM software we looked for examples where developers have been
+successful so that we can share these with the LBM community, and with the
+broader research software community.  We also aimed to identify potential
+approaches for mitigating the identified pain points in the future.
+
+Our analysis points to areas where some LBM software packages fall short of
+current best practices.  This is not intended to be a criticism of any existing
+packages, especially since in practice not every project needs to achieve the
+highest possible quality.  However, rather than delve into the nuances of which
+software can justify compromising which practice we conducted our comparison
+under the ideal assumption that every project can have sufficient resources to
+match best practices.
 
@@
 correctness/verifiability, reliability, robustness, usability, maintainability,
-reusability, understandability, and visibility/transparency)   The repository
-data is collected by filling in a 108 question measurement template, which
-requires installing the software on a VM, running simple tests (like completing
-the getting started instructions (if present)), and searching the code,
-documentation and test files.; v) Rank the software using the Analytic Hierarchy
-Process (AHP); vi) Interview developers (the interview consists of 20 questions
-and takes about an hour); and, vii) Conduct a domain analysis. The collected
-data is analyzed by: a) comparing the ranking by best practices against the
-ranking by popularity; b) comparing artifacts, tools and processes to current
-research software development guidelines; and, c) exploring pain points.
-
-Table~\ref{topperformerstable} summarizes the top 5 performers for each of the qualities that we assessed.
+reusability, understandability, and visibility/transparency).  Filling in the
+template requires installing the software on a VM, running simple tests (like
+completing the getting started instructions (if present)), and searching the
+code, documentation and test files; v) Rank the software using the Analytic
+Hierarchy Process (AHP); vi) Interview developers (the interview consists of 20
+questions and takes about an hour); and, vii) Conduct a domain analysis. The
+collected data was analyzed by: a) comparing the ranking by best practices
+against the ranking by popularity; b) comparing artifacts, tools and processes
+to current research software development guidelines; and, c) exploring pain
+points.  Our ranking was roughly consistent with ranking the repositories by the
+number of GitHub stars, although the number one project by star count (Sailfish)
+is ranked 9th by our methodology.
+
+Current LBM software provides many examples of following best practices, to the
+benefit of users and developers.  This is illustrated by
+Table~\ref{topperformerstable}, which summarizes the top 5 performers for each of
+the identified qualities and for the overall quality. Taking the union of all
+entries in the table, we have 16 top performers: ESPREsSo, MechSys, OpenLB,
+DL\_MESO, Sailfish, Ludwig, Musubi, pyLBM, waLBerla, laboetie, lettuce,
+ESPREsSo++, LUMA, MP-LABS, Palabos, and HemeLB.  Therefore, two thirds (67\%) of
+the 24 measured packages appear in the top 5 for at least one quality.  Of these
+entries almost all of them are classified as alive, except for three packages:
+laboetie, MP-LABS, and HemeLB.  This suggests that alive packages are in a
+healthy state with respect to adopting best practices.
 
@@
 			\toprule
-			Quality & Ranked 1st & Ranked 2nd & Ranked 3rd & Ranked 4th & Ranked 5th\\
+			Quality & Ranked 1st & Ranked 2nd & Ranked 3rd & Ranked 4th & Ranked
+			5th\\
 			\midrule
-			Installability & ESPResSo & MechSys & OpenLB & DL\_MESO (LBE) & Sailfish\\
+			Installability & ESPResSo & MechSys & OpenLB & DL\_MESO (LBE) &
+			Sailfish\\
 			\addlinespace[0.4cm]
-			\pbox{3.0cm}{Surface Correctness \\ and Verifiability} & Ludwig & Musubi & pyLBM & OpenLB & waLBerla\\
+			\pbox{3.0cm}{Surface Correctness \\ and Verifiability} & Ludwig &
+			Musubi & pyLBM & OpenLB & waLBerla\\
 			\addlinespace[0.4cm]
-			Surface Reliability & DL\_MESO (LBE) & laboetie* & lettuce & MechSys & Musubi \\
+			Surface Reliability & DL\_MESO (LBE) & laboetie* & lettuce & MechSys
+			& Musubi \\
 			\addlinespace[0.4cm]
-			Surface Robustness & DL\_MESO (LBE) & ESPResSo++ & lettuce & LUMA & OpenLB \\
+			Surface Robustness & DL\_MESO (LBE) & ESPResSo++ & lettuce & LUMA &
+			OpenLB \\
 			\addlinespace[0.4cm]
@@
 			\addlinespace[0.4cm]
-			\pbox{3.0cm}{Surface \\Understandability} & LUMA & Palabos & pyLBM & DL\_MESO (LBE)& MechSys\\
+			\pbox{3.0cm}{Surface \\Understandability} & LUMA & Palabos & pyLBM &
+			DL\_MESO (LBE)& MechSys\\
 			\addlinespace[0.4cm]
-			\pbox{3.0cm}{Visibility and \\Transparency} & Ludwig & Palabos & LUMA & ESPResSo & MP-LABS*\\
+			\pbox{3.0cm}{Visibility and \\Transparency} & Ludwig & Palabos &
+			LUMA & ESPResSo & MP-LABS*\\
 			\addlinespace[0.4cm]
@@
 
-Comparison to community.
-
-Comparison of artifacts.
-
-Comparison of tools.
-
-Comparison of processes.
-
-Developer pain Points
-
-Lessons from developers
-
-Recommendations, starting from current best practices.
+Further evidence of the healthy state of the practice can be seen by comparing
+the observed software artifacts versus those recommended by software engineering
+guidelines.  As Section~\ref{Sec_CompareArtifacts} shows, the majority of LBM
+generated artifacts correspond to general recommendations from research software
+developers.  LBM software even seems to be ahead of best practices with an
+example of a video user guide for one of the packages.  Comparing LBM with
+respect to tool usage (Section~\ref{Sec_CompareTools}) shows that LBM is keeping
+pace with other research software, with 67\% of the surveyed tools using version
+control.  The state of the practice for the software development process of LBM
+software matches the quasi-agile process generally adopted for research software
+(Section~\ref{Sec_CompareMethodologies}).  The best practice process of peer
+review is employed by some LBM developers.
+
+Although the state of the practice for LBM software is healthy, we have noted
+areas for potential improvement in the adoption of best practices, as follows:
+
+\begin{itemize}
+	\item Increase the utilization of continuous integration.
+	\item Increase the amount of API documentation.
+	\item Add roadmap documentation showing what is planned for the future.
+	\item Add a code of conduct to explicitly show how developers should treat
+	one another.
+	\item Increase the use and enforcement of programming style guides.
+	\item Use checklists to ensure best practices are followed by developers.
+	\item Include uninstall instructions.
+	\item Explicitly document the software's motivation and design principles.
+\end{itemize}
+
+To advance the state of the practice beyond current best practices toward future
+best practices, we interviewed 4 developers to learn about their pain points
+(Section~\ref{painpoints}).  The identified pain points include lack of
+development time, lack of software development experience, lack of incentive and
+funding, lack of external support, technology hurdles, ensuring correctness,
+usability, technical debt and documentation quality.  We found
+(Section~\ref{Sec_AddressConcerns} that developers are currently creating
+practices to address these pain points, including: designing for change,
+circumventing the oracle testing problem, and prioritizing documentation and
+usability.  To further plan for future best practices, we identified eight
+potential directions (Section~\ref{Sec_Recommendations}), in order of increasing
+novelty:
 
 \begin{enumerate}
-	\item \textbf{Follow the example of LBM packages that already follow best practices.}  Point to the table summarizing the winners for each quality.
-Review the practices from the previous section.
-Addresses PP?.
-	\item \textbf{Continuous integration.} Addresses PP?.
-	\item \textbf{Document API.} Addresses PP?.
+	\item Employ linters.
+	\item Conduct rigorous peer reviews.
+	\item Write and submit papers on software.
+	\item Grow the number of contributors.
+	\item Further address the oracle problem for testing.
+	\item Augment theory manuals to include requirements information.
+	\item Improve re-runnability, repeatability, reproducibility and
+	replicability.
+	\item Generate all things.
 \end{enumerate}
 
-Future work
-
-\wss{More on the lessons to learn from the top packages.  The comparison to
-common practice - is LBM ahead or behind?  More on best practices comparison.
-Future practices.}
-
-The goal of this assessment isn't to rank the software, but to use the ranking
-exercise as a means to understand the state of the practice of LBM software
-development.
-
-We understand that software packages vary in goals, developers, funding, and
-other aspects. Our goal was to highlight software quality successes through an
-evaluation of the entire domain software family. Furthermore, threats to the
-validity of the findings are highlighted in Section~\ref{threats}.
-
-Our comparison may point out areas where some LBM software packages fall short
-of current best practices.  This is not intended to be a criticism of any
-existing packages, especially since in practice not every project needs to
-achieve the highest possible quality.  However, rather than delve into the
-nuances of which software can justify compromising which practice we will write
-our comparison under the ideal assumption that every project can have sufficient
-resources to match best practices.
-
-Many software packages would benefit from a rewrite or reorganization of
-installation instructions. A single location for installation instructions would
-improve their maintainability and correctness. Listing compatible operating
-system and dependency versions would decrease installation time and errors, as
-would adding instructions on installing dependencies. Installation process
-errors should prompt the system to display detailed messages. Once a software
-package is installed, either an automatic validation needs to be performed or
-the user needs to be able to perform a manual validation using test examples
-that include expected output. Finally, uninstallation instructions should be
-included in the documentation. \wss{Have to decide how much to emphasizse
-installability.}  \wss{Maybe at some point should mention more modern
-installation support, since most of the projects use make, and there are other
-options.}
-
-Ensuring software quality can take significant time. An inability to develop
-high quality documentation due to time constraints could be mitigated in the
-future by the use of automatic document generation tools like Drasil. The Drasil
-Framework consists of a collection of Domain Specific Languages (DSL) for
-capturing scientific documents, structures, and computing knowledge, and then
-transforming this knowledge into relevant software artifacts without having to
-manually duplicate knowledge into multiple artifacts \citep{SzymczakEtAl2016}.
-
-Future State Of The Practice Assessments
-
-As mentioned in Section~\ref{threats}, the measures listed in our measurement
-template may not be broad enough to accurately capture some qualities. Adding or
-extending measures is worth considering. For example, usability experiments and
-performance benchmarks could be incorporated into the assessment.
-
-Section~\ref{SecReproducibility} noted that developers were not asked to comment
-on the reproducibility of their source code from their requirements
-specifications and design documentation. Adding this question to the interview
-guide should be considered in the next iteration of state of the practice
-assessments. Furthermore, as mentioned in Section~\ref{threats}, it was found
-that some pertinent information was not specified in public artifacts. The use
-of unit testing and continuous integration by several packages (OpenLB, pyLBM,
-TCLB) was only discovered during interviews with developers. Adding further
-questions to the interview guide regarding the measures that are on the
-measurement template could reduce instances of incorrect data being collected.
-This additional interview data could be analyzed and incorporated into the AHP
-ranking, ensuring quality designations more accurate represent the true quality
-of the software packages.  \wss{clarify the distinction between usability and
-understandability.}
-
-For the current project, we did not assess the stronger form of
-``reproducibility'' known as replicability.  A software project is considered
-replicable if the documentation alone (without the code) is sufficiently precise
-and complete to allow a third-part to re-obtain the results of the original code
-\citep{BenureauAndRougier2017}.  Replicability is difficult to achieve with
-current documentation practices.  This question should be considered in the next
-iteration of state of the practice assessments.
-
-The final result of the domain analysis will be tables of commonalities,
-variabilities, and parameters of variation of a program family.
-\cite{smith2008commonality} present a template for conducting a commonality
-analysis, which was referred to when conducting this work.
-\cite{weiss1998commonality} describes another commonality analysis technique for
-deciding the members of a program family. \cite{SmithAndChen2004} and
+Future work can be done to address the threats to validity mentioned in
+Section~\ref{threats}.  For instance, the measures listed in our measurement
+template could be broadened to better capture some qualities. For example,
+usability experiments and performance benchmarks could be incorporated into the
+assessment.  Section~\ref{threats} also mentions that some pertinent information
+on continuous integration and unit testing was not specified in public
+artifacts. Adding further questions to the interview guide regarding the
+measures that are on the measurement template could reduce instances of
+incomplete data. This additional interview data could be analyzed and
+incorporated into the AHP ranking, ensuring quality designations more accurate
+represent the true quality of the software packages. In the future we could also
+expand the interview questions to go beyond reproducibility to also discuss
+replicability. Furthermore, in future interviews we would like to clarify the
+distinction between usability and understandability.
+
+Our domain analysis (Section~\ref{SecDomainAnalysis} was relatively shallow.  In
+the future we would like to dig more deeply into this analysis.  We would like
+to produce tables of commonalities, variabilities, and parameters of variation
+for the family of LBM solvers. We could follow the template from
+\citep{smith2008commonality}, or adopt the analysis techniques from
+\citep{weiss1998commonality}. \cite{SmithAndChen2004} and
 \cite{SmithMcCutchanAndCarette2017} are examples of a commonality analysis for a
 family of mesh generating software and a family of material models,
-respectively. The steps to produce a commonality analysis are:
+respectively.
 
@@
 
-We would like to thank Ao Dong and Oluwaseun Owojaiye for many fruitful
-discussions on topics relevant to this paper.
+We would like to thank Ao Dong, Oluwaseun Owojaiye, Mohammad Ali Daeian and Reza
+Sadeghi for fruitful discussions on topics relevant to this paper.
 

### 2022-03-04 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Edits to abstract and conclusions

@@
     We analyze the state of software development practice for Lattice Boltzmann
-    solvers and find by quantitatively and qualitatively measuring and comparing
-    24 software packages for 10 software qualities (installability, correctness/
+    solvers by quantitatively and qualitatively measuring and comparing 24
+    software packages for 10 software qualities (installability, correctness/
     verifiability, reliability, robustness, usability, maintainability,
@@
     quality, the majority of LBM generated artifacts corresponding to general
-    recommendations from research software developers, common utilization (67\%
-    of packages) of version control and adoption of a quasi-agile development
-    process.Areas of best practice to potentially improve include adoption of
+    recommendations from research software developers, common utilization of
+    version control (67\% of packages) and adoption of a quasi-agile development
+    process.  Areas of best practice to potentially improve include adoption of
     continuous integration, API documentation and enforcement of programming
@@
     oracle problem and prioritizing documentation and usability.  For future the
-    best practices of the future, we suggest the community consider: employing
+    best practices, we suggest the community consider the following: employing
     linters, conducing rigorous peer reviews, writing and submitting more papers
@@
 provide direction for future LBM software development, our goal was to analyze
-the current state of LBM software development.  We aimed to determine how well
-best practices are followed and to understand current developer pain points.  We
+the current state of LBM software development.  We determined how well best
+practices are followed and we investigated current developer pain points.  We
 compared the state of the practice for LBM software to the development practices
@@
 circumventing the oracle testing problem, and prioritizing documentation and
-usability.  To further plan for future best practices, we identified eight
+usability.  To plan for future best practices, we identified eight
 potential directions (Section~\ref{Sec_Recommendations}), in order of increasing
@@
 incomplete data. This additional interview data could be analyzed and
-incorporated into the AHP ranking, ensuring quality designations more accurate
+incorporated into the AHP ranking, ensuring quality designations more accurately
 represent the true quality of the software packages. In the future we could also
 expand the interview questions to go beyond reproducibility to also discuss
-replicability. Furthermore, in future interviews we would like to clarify the
-distinction between usability and understandability.
+replicability. Furthermore, in future interviews we would look to clarify the
+distinction between usability and understandability
+(Section~\ref{softwarequalities}), since the distinction confused some
+participants.
 
@@
 \citep{weiss1998commonality}. \cite{SmithAndChen2004} and
-\cite{SmithMcCutchanAndCarette2017} are examples of a commonality analysis for a
-family of mesh generating software and a family of material models,
-respectively.
+\cite{SmithMcCutchanAndCarette2017} provide examples that could be followed for
+a commonality analysis of a family of mesh generating software and a family of
+material models, respectively.
 

### 2022-03-06 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
reviewed abstract and section 1

@@
     addressing these pain points by designing for change, circumventing the
-    oracle problem and prioritizing documentation and usability.  For future the
+    oracle problem and prioritizing documentation and usability.  For future
     best practices, we suggest the community consider the following: employing
-    linters, conducing rigorous peer reviews, writing and submitting more papers
-    on software, growing the number of contributors and augmenting the theory
-    manuals to include more requirements relevant information.
+    linters, conducting rigorous peer reviews, writing and submitting more papers on software, growing the number of contributors and augmenting the theory manuals to include more requirements relevant information.
 

### 2022-03-06 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
reviewed abstract and section 2 - methodology

@@
 We developed a methodology for evaluating the state of the practice of
-research software \citep{SmithEtAl2021}. This methodology can be instantiated
-for a specific domain of research software, which in the current case is LBM
+research software \citep{SmithEtAl2021}. This methodology can be applied
+to a specific domain of research software, which in the current case is LBM
 software.  Since we group best practices around conventional software qualities,
@@
 provides an overview of the steps we used to select, measure and compare LBM
-software.  The remaining sections provide details on each step of the steps in
-the methodology.
+software.  The remaining sections provide details on each step in the methodology.
 
@@
 	a specified environment \citep{ISO/IEC25010, lenhard2013measuring}. In our
-	case effort includes the time spent finding and understanding the
+	case, effort includes the time spent finding and understanding the
 	installation instructions, the person-time and resources spent performing
@@
 	explicit specification, since this is not common practice in the field. As a
-	consequence, the correctness of software cannot often be measured directly.
+	consequence, the correctness of software often cannot be measured directly.
 	We assess correctness indirectly by looking for the following: availability
@@
 	performance or other attributes, and satisfy new requirements
-	\citep{IEEEStdGlossarySET1990, boehm2007software}. In the current work
+	\citep{IEEEStdGlossarySET1990, boehm2007software}. In the current work,
 	maintainability is measured by the quality of the documentation artifacts,
@@
 	to the structure of the system, modules and code blocks. A follow up
-	question is asked if any measures have been taken.
+	question asked if any measures have been taken to ensure the ease of future changes.
 
@@
 To reduce the length of the list of software packages and to
-answer~\rqref{RQ_WhatProjects}, we filter the initial list of 45.  The following
-filters were applied in the priority order listed.
+answer~\rqref{RQ_WhatProjects}, we filter the initial list of 45 packages.  The following filters were applied in the priority order listed.
 
@@
 	considered to be within the scope of the domain.
-	\item Usage: Software packages were eliminated if their installation
+	\item Usage: Software packages are eliminated if their installation
 	procedure was missing or not clear and easy to follow.
 	\item Age: The older software packages (age being measured by the last date
-	when a change was made) were eliminated, except in the cases where an older
+	when a change was made) are eliminated, except in the cases where an older
 	software package appears to be highly recommended and currently in use. 
@@
 are looking for an indication that usability was considered by the developers by
-looking for cues in the documentation, such as getting started instructions, a
-user manual or documentation of expected user characteristics.
+looking for cues in the documentation, such as the presence of getting started instructions, a user manual or documentation of expected user characteristics.
 
@@
 
-\subsection{Analytical Hierarchy Process} \label{AHP}
+\subsection{Analytic Hierarchy Process} \label{AHP}
 
 Once we have measured each package, we still need to rank them to answer
-\rqref{RQ_HighestQuality}.  To do this, we used the Analytical Hierarchy Process
+\rqref{RQ_HighestQuality}.  To do this, we used the Analytic Hierarchy Process
 (AHP), a decision-making technique that is used to compare multiple options by
@@
 the importance of, and the current situation for, documentation. A few questions
-are about specific software qualities, such as maintainability,
+focus on specific software qualities, such as maintainability,
 understandability, usability, and reproducibility. The interviews are
@@
 subjects were found from project websites, code repositories, publications, and
-biographic information on institution web-pages. Meeting were held using on-line
-meeting software (either Zoom or Teams), which facilitated recording and
+biographic information on institution web-pages. Meetings were held using on-line meeting software (either Zoom or Teams), which facilitated recording and
 automatic transcription.  Each meeting lasted about an hour. The interview

### 2022-03-06 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
reviewed abstract and section 3 - ranking projects..

@@
 instruction manual or on a web-page. Sometimes, like with Ludwig, incomplete
-installation instructions are found on a home page, with more detailed
+installation instructions are found on a homepage, with more detailed
 instructions located on another web-page, or within the documentation.
@@
 user may need to assume correctness if there are no visible errors.  A
-particularly  detailed tutorial that include expected output is provided by
+particularly  detailed tutorial that includes expected output is provided by
 \href{https://www.walberla.net/doxygen/index.html} {waLBerla}.
@@
 command despite following the tutorial, and the final two errors were vague
-execution errors. Of the four broken tutorial instances, only the missing
-dependency one was recoverable.
+execution errors. Of the four broken tutorial instances, only the one due to a missing dependency was recoverable.
 
@@
 to host their git repository, seven use email, and one (SunlightLB) uses
-SourceForge. Most software packages that use GitHub/GitLab have the majority of
-their issues closed, and only three (laboetie, lettuce, Sailfish) have less than
-50 percent of their issues closed. All of the top five overall ranked packages
-have most of their issues closed. Alive packages (11 use issue tracking) have
-64\% of their issues closed, while dead packages (3 use issue tracking) have
-71\% of their issues closed. This information is presented in
+SourceForge.
+Most software packages that use GitHub or GitLab have the majority of their issues closed, and only three (laboetie, lettuce, Sailfish) have less than 50 percent of their issues closed.
+All of the top five overall ranked packages have most of their issues closed.
+Alive packages (11 use issue tracking) have 64\% of their issues closed, while dead packages (3 use issue tracking) have 71\% of their issues closed.
+This information is presented in
 Table~\ref{gitrepodata}. With respect to the host for the git repository, 13
@@
 Software package artifacts were reviewed for the identification of a specific
-development model, like a waterfall of agile development model, and the presence
+development model, like a waterfall or agile development model, and the presence
 of documentation recording the development process and standards. They were also
@@
 transparency, the small close-knit nature of the development teams make
-explicitly publicly specifying this information practically unnecessary. Version
+explicitly specifying this information practically unnecessary. Version
 release notes were found in nine of the software packages.

### 2022-03-06 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
reviewed abstract and section 4 - comparison to community ranking
reviewed abstract and section 5 - comparison bw LBM and research...

@@
 lower quality often do not use GitHub or GitLab, or have only a few
-stars/watches. Although the details of the rankings differ, our measures show
+stars and watches. Although the details of the rankings differ, our measures show
 that following best practices tends to correlate with increased popularity in
@@
 of the software packages. Rare artifacts were found in 1 to 7 ($<$30\%) of the
-software packages.  In this section we answers \rqref{RQ_CompareArtifacts} by
+software packages.  In this section we answer \rqref{RQ_CompareArtifacts} by
 comparing the artifacts that we observed in LBM repositories to those observed
@@
 The top four packages, ESPREsSo, Ludwig, Palabos, and OpenLB, have all of the
-common artifacts, except Palabos does not have theory notes and only three of
-them (ESPREsSo, Ludwig, Palabos) appear to use a version control system.  The
-top four packages also have most of the less common artifacts. At the time of
-data collection, only one of the four packages (Palabos) did not have a user
-manual or guide, but there was a broken link on the package website indicating
-that such an artifact might exist. This broken link was later fixed, but this is
-not reflected in our data because it was not present at the time of data
-collection (mid 2020). Despite the broken link, Palabos does have a detailed and
-informative website. As far as we could observe, Ludwig does not appear to have
-a publicly visible design and OpenLB does not appear to use a version control
-system. However, OpenLB's use of version numbers suggests that version control
-may be used privately by the developers.
+common artifacts, except Palabos does not have theory notes and only three (ESPREsSo, Ludwig, Palabos) of the packages appear to use a version control system.
+The top four packages also have most of the less common artifacts.
+At the time of data collection, only one of the four packages (Palabos) did not have a user manual or guide, but there was a broken link on the package website indicating that such an artifact might exist.
+This broken link was later fixed, but this is not reflected in our data because it was not present at the time of data collection (mid 2020).
+Despite the broken link, Palabos does have a detailed and informative website.
+As far as we could observe, Ludwig does not appear to have a publicly visible design and OpenLB does not appear to use a version control system.
+However, OpenLB's use of version numbers suggests that version control may be used privately by the developers.
 
@@
 management is the software engineering discipline that most hurts scientific
-developers, accounting for 23 \% of the technical problems reported by study
+developers, accounting for 23\% of the technical problems reported by study
 participants.  The lack of support for requirements is likely due to the

### 2022-03-06 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
reviewed abstract and section 6 - comparison of tool..

@@
 tools, dependencies, and project management tools.  In this section we answer
-\rqref{RQ_CompareToolsProjMngmnt} by comparing aspects of the tool usage in LBM
-to their utilization in the research software community in general.
+\rqref{RQ_CompareToolsProjMngmnt} by comparing aspects of tool usage in LBM software packages to their utilization in the research software community in general.
 
@@
 As mentioned in Section~\ref{SecSurfCorrectAndVerifiab}, continuous integration
-is rarely used in LBM (3 of 24 packages or 12.5\%). This contrasts with the
-frequency with which continuous integration is recommended in research software
-development guidelines \citep{BrettEtAl2021, Brown2015, ThielEtAl2020,
+is rarely used in LBM software (3 of 24 packages or 12.5\%).
+This contrasts with the frequency with which continuous integration is recommended in research software development guidelines \citep{BrettEtAl2021, Brown2015, ThielEtAl2020,
 Zadka2018, vanGompelEtAl2016}.  We could not find published data on the

### 2022-03-06 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
reviewed abstract and section 7 - comparison of principles...

@@
 
-Interviews with developers confirmed a similar project management processes. In
+Interviews with developers confirmed a similar project management process. In
 teams of only a couple of developers, additions of new features or major changes
@@
 SSI2022} and quick start examples \citep{ThielEtAl2020, vanGompelEtAl2016} are
-also recommended.  \citep{SmithAndRoscoe2018} suggests including instructions
+also recommended.  \citet{SmithAndRoscoe2018} suggests including instructions
 specifically for on-boarding new developers. For open source software in general
-(not just research software), \citep{Fogel2005} recommends providing tutorial
+(not just research software), \citet{Fogel2005} recommends providing tutorial
 style examples, developer guidelines, demos and screenshots.

### 2022-03-06 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
reviewed abstract and section 8 - dev pain points; 4 to four

@@
     continuous integration, API documentation and enforcement of programming
-    style guides.  We interviewed 4 developers to gain insight into their
+    style guides.  We interviewed four developers to gain insight into their
     current pain points. Identified challenges include lack of development time,
@@
 
-Based on interviews with 4 developers, this section aims to answer the research
-questions: i) What are the pain points for developers working on research
-software projects (\rqref{RQ_PainPoints})?; and, ii) How do the pain points of
-developers from LBM compare to the pain points for research software in general
-(\rqref{RQ_ComparePainPoints})?  Below we go through each of the identified pain
-points and include citations that contrast the LBM experience with observations
-from researchers in other domains.  Potential ways to address the pain points
-are covered in Section~\ref{Sec_AddressConcerns}. The full interview questions
-are found in \citet{SmithEtAl2021}.
-
-\citep{PintoEtAl2018} lists some pain points that did not come up in our
+Based on interviews with four developers, this section aims to answer the research questions: i) What are the pain points for developers working on research software projects (\rqref{RQ_PainPoints})?; and, ii) How do the pain points of developers from LBM compare to the pain points for research software in general (\rqref{RQ_ComparePainPoints})?
+Below we go through each of the identified pain points and include citations that contrast the LBM experience with observations from researchers in other domains.
+Potential ways to address the pain points are covered in Section~\ref{Sec_AddressConcerns}.
+The full interview questions are found in \citet{SmithEtAl2021}.
+
+\citet{PintoEtAl2018} lists some pain points that did not come up in our
 conversations with LBM developers: Cross-platform compatibility, interruptions
 while coding, scope bloat, lack of user feedback, hard to collaborate on
-software projects, and aloneness. \citep{WieseEtAl2019} repeat some of the
+software projects, and aloneness. \citet{WieseEtAl2019} repeat some of the
 previous pain points and add the following: dependency management, data handling
@@
 pain points, we cannot conclude that they are not relevant for LBM software
-development, since we only interviewed 4 LBM developers for about an hour each.
+development, since we only interviewed four LBM developers for about an hour each.
 
@@
 	project are domain experts, not computer scientists or software engineers.
-	This same trend is noted by \citet{Nguyen-HoanEtAl2010}, which showed only
+	This same trend is noted by \citep{Nguyen-HoanEtAl2010}, which showed only
 	23\% of research software survey respondents having a computing-related
@@
 	\citep{CrouchEtAl2013}, and \href{https://software-carpentry.org/}{Software
-	Carpentry} \citep{WilsonAndLumsdaine2006, Wilson2016}. Over time awareness of
-	these groups will grow, so this pain point is likely to disappear in the
+	Carpentry} \citep{WilsonAndLumsdaine2006, Wilson2016}. Over time awareness of these groups will grow, so this pain point is likely to disappear in the
 	future for LBM and other research software developers.
@@
 To advance the state of the practice beyond current best practices toward future
-best practices, we interviewed 4 developers to learn about their pain points
+best practices, we interviewed four developers to learn about their pain points
 (Section~\ref{painpoints}).  The identified pain points include lack of

### 2022-03-07 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
reviewed abstract and section 9 - lessons from devs

@@
 
-LBM developers highlighted in our interviews cases where their modularizations
+Interviewed LBM developers highlighted cases where their modularizations
 anticipated future changes.  The developer of pyLBM mentioned that the
@@
 	proceeding with the calculations.
-	\item Include documentation on frequently asked questions (Section~\ref{Sec_CompareArtifacts}).
+	\item Include documentation on FAQ (Section~\ref{Sec_CompareArtifacts}).
 \end{itemize}

### 2022-03-07 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
reviewed abstract and section 10 - recommendations

@@
 Section~\ref{painpoints} to answer~\rqref{RQ_Recommend}.  Our recommendations
-are not lists of criticisms what should have been done in the past, or what
+are not lists of criticisms for what should have been done in the past, or what
 should be done now; they are just suggestions for consideration in the future.
 We will not be repeating ideas here that have been discussed in previous
-section, such as continuous integration, documentation of APIs, etc.  Our aim is
-to mention ideas that are at least somewhat beyond conventional best practices.
-The ideas listed have the potential to become best practices in the medium to
-long-term.  The ideas are provided roughly in the order of increasing novelty.
+sections, such as continuous integration, documentation of APIs, etc.
+Our aim is to mention ideas that are at least somewhat beyond conventional best practices.
+The ideas listed here have the potential to become best practices in the medium to long-term.
+The ideas are presented roughly in the order of increasing novelty.
 
@@
 Linters can address the pain points for LBM developers.  For instance, a linter
-can increase the amount of development time (\ppref{P_LackDevTime}) by
+can decrease the amount of development time (\ppref{P_LackDevTime}) by
 decreasing the number of mundane mistakes programmers have to catch.  Since the
@@
 
-Peer review is already a strategy adopted by some LBM developers to improve
+Peer review is a strategy already adopted by some LBM developers to improve
 confidence in their software (Section~\ref{SecReproducibility}), but the
@@
 
-\subsection{Augment Theory Manual to Include Requirements Information}
+\subsection{Augment Theory Manuals to Include Requirements Information}
 
@@
 recommendations to ensure that correctness can be maintained into the future by
-prioritizing 4 increasingly sophisticated levels of replication.
+prioritizing four increasingly sophisticated levels of replication.
 

### 2022-03-07 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
digits under

@@
     our methodology. Our finding is that the state of the practice is healthy
-    with 67\% of the measured packages ranking in the top 5 for at least one
+    with 67\% of the measured packages ranking in the top five for at least one
     quality, the majority of LBM generated artifacts corresponding to general
@@
 \citet{SmithEtAl2021}.  For each software package (each column in the template),
-we fill-in the rows of the template. This process takes about 2 hours per
-package, with a cap of 4 hours. The time constraint is necessary so that the
+we fill-in the rows of the template. This process takes about two hours per
+package, with a cap of four hours. The time constraint is necessary so that the
 work load is feasible for a team as small as one, given our aim to cap the
@@
 
-The full template consists of 108 questions categorized under 9 qualities:
+The full template consists of 108 questions categorized under nine qualities:
 \begin{inparaenum}[(i)]
@@
 since Munubi was added after our ethics approved had expired.  We were able to
-recruit 4 developers to participate in our study.  Our potential interview
+recruit four developers to participate in our study.  Our potential interview
 subjects were found from project websites, code repositories, publications, and
@@
 mentioned parallelization. This included GPU processing and the CUDA parallel
-computing platform, which were mentioned in the artifacts of 6 packages
+computing platform, which were mentioned in the artifacts of six packages
 (ESPResSo, lbmpy, lettuce, pyLBM, Sailfish, TCLB). When a high degree of
@@
 closed. Their code files are well commented with more than 10 percent of the
-code being comments. Four of the top 5 ranked packages are noted as being alive.
+code being comments. Four of the top five ranked packages are noted as being alive.
 
@@
 of open issues, number of lines of code; iv) Fill in the measurement template
-(the template consists of 108 questions to assess 9 qualities (installability,
-correctness/verifiability, reliability, robustness, usability, maintainability,
-reusability, understandability, and visibility/transparency).  Filling in the
-template requires installing the software on a VM, running simple tests (like
-completing the getting started instructions (if present)), and searching the
-code, documentation and test files; v) Rank the software using the Analytic
-Hierarchy Process (AHP); vi) Interview developers (the interview consists of 20
-questions and takes about an hour); and, vii) Conduct a domain analysis. The
-collected data was analyzed by: a) comparing the ranking by best practices
-against the ranking by popularity; b) comparing artifacts, tools and processes
-to current research software development guidelines; and, c) exploring pain
-points.  Our ranking was roughly consistent with ranking the repositories by the
-number of GitHub stars, although the number one project by star count (Sailfish)
-is ranked 9th by our methodology.
+(the template consists of 108 questions to assess nine qualities (installability, correctness/verifiability, reliability, robustness, usability, maintainability, reusability, understandability, and visibility/transparency).
+Filling in the template requires installing the software on a VM, running simple tests (like completing the getting started instructions (if present)), and searching the code, documentation and test files; v) Rank the software using the Analytic Hierarchy Process (AHP); vi) Interview developers (the interview consists of 20 questions and takes about an hour); and, vii) Conduct a domain analysis.
+The collected data was analyzed by: a) comparing the ranking by best practices against the ranking by popularity; b) comparing artifacts, tools and processes to current research software development guidelines; and, c) exploring pain points.
+Our ranking was roughly consistent with ranking the repositories by the number of GitHub stars, although the number one project by star count (Sailfish) is ranked 9th by our methodology.
 
@@
 benefit of users and developers.  This is illustrated by
-Table~\ref{topperformerstable}, which summarizes the top 5 performers for each of
-the identified qualities and for the overall quality. Taking the union of all
-entries in the table, we have 16 top performers: ESPREsSo, MechSys, OpenLB,
-DL\_MESO, Sailfish, Ludwig, Musubi, pyLBM, waLBerla, laboetie, lettuce,
+Table~\ref{topperformerstable}, which summarizes the top five performers for each of the identified qualities and for the overall quality.
+Taking the union of all entries in the table, we have 16 top performers: ESPREsSo, MechSys, OpenLB, DL\_MESO, Sailfish, Ludwig, Musubi, pyLBM, waLBerla, laboetie, lettuce,
 ESPREsSo++, LUMA, MP-LABS, Palabos, and HemeLB.  Therefore, two thirds (67\%) of
-the 24 measured packages appear in the top 5 for at least one quality.  Of these
-entries almost all of them are classified as alive, except for three packages:
-laboetie, MP-LABS, and HemeLB.  This suggests that alive packages are in a
-healthy state with respect to adopting best practices.
+the 24 measured packages appear in the top five for at least one quality.
+Of these entries almost all of them are classified as alive, except for three packages: laboetie, MP-LABS, and HemeLB.
+This suggests that alive packages are in a healthy state with respect to adopting best practices.
 

### 2022-03-07 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
reviewed threats to validity

@@
 
-In our methodology one individual measures each package and the beginning and
+In our methodology, one individual measures each package and the beginning and
 ending of the measurement phase is separated by several months.  There is a
-threat that the measurement may include unconscious biases, or the reviewers
-judgement may drift over time.  In previous work \citep{SmithEtAl2016}, we
-demonstrated that the measurement process is reasonably reproducible by having 5
-products graded by a second reviewer. The ranking via this independent review
-was almost identical to the original ranking. As long as each grader uses
+threat that the measurement may be affected by unconscious biases, or the reviewers judgement may drift over time.
+In previous work \citep{SmithEtAl2016}, we demonstrated that the measurement process is reasonably reproducible by having five products graded by a second reviewer.
+The ranking via this independent review was almost identical to the original ranking.
+As long as each grader uses
 consistent definitions, the relative comparisons in the AHP results will be
@@
 
-Another data collection concern is the gap between measurement of Musubi and the
-measurement of all other packages (Section~\ref{identifysoftware}).
+Another data collection concern is the gap between the measurement of Musubi and the measurement of all other packages (Section~\ref{identifysoftware}).
 Unfortunately we were not aware of the existence of Musubi until after the

### 2022-03-07 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Reformatted to 80 column width

@@
     best practices, we suggest the community consider the following: employing
-    linters, conducting rigorous peer reviews, writing and submitting more papers on software, growing the number of contributors and augmenting the theory manuals to include more requirements relevant information.
+    linters, conducting rigorous peer reviews, writing and submitting more
+    papers on software, growing the number of contributors and augmenting the
+    theory manuals to include more requirements relevant information.
 
@@
 	to the structure of the system, modules and code blocks. A follow up
-	question asked if any measures have been taken to ensure the ease of future changes.
+	question is asked if any measures have been taken to ensure the ease of
+	future changes.
 
@@
 To reduce the length of the list of software packages and to
-answer~\rqref{RQ_WhatProjects}, we filter the initial list of 45 packages.  The following filters were applied in the priority order listed.
+answer~\rqref{RQ_WhatProjects}, we filter the initial list of 45 packages.  The
+following filters were applied in the priority order listed.
 
@@
 are looking for an indication that usability was considered by the developers by
-looking for cues in the documentation, such as the presence of getting started instructions, a user manual or documentation of expected user characteristics.
+looking for cues in the documentation, such as the presence of getting started
+instructions, a user manual or documentation of expected user characteristics.
 
@@
 subjects were found from project websites, code repositories, publications, and
-biographic information on institution web-pages. Meetings were held using on-line meeting software (either Zoom or Teams), which facilitated recording and
+biographic information on institution web-pages. Meetings were held using
+on-line meeting software (either Zoom or Teams), which facilitated recording and
 automatic transcription.  Each meeting lasted about an hour. The interview
@@
 command despite following the tutorial, and the final two errors were vague
-execution errors. Of the four broken tutorial instances, only the one due to a missing dependency was recoverable.
+execution errors. Of the four broken tutorial instances, only the one due to a
+missing dependency was recoverable.
 
@@
 to host their git repository, seven use email, and one (SunlightLB) uses
-SourceForge.
-Most software packages that use GitHub or GitLab have the majority of their issues closed, and only three (laboetie, lettuce, Sailfish) have less than 50 percent of their issues closed.
-All of the top five overall ranked packages have most of their issues closed.
-Alive packages (11 use issue tracking) have 64\% of their issues closed, while dead packages (3 use issue tracking) have 71\% of their issues closed.
-This information is presented in
+SourceForge. Most software packages that use GitHub or GitLab have the majority
+of their issues closed, and only three (laboetie, lettuce, Sailfish) have less
+than 50 percent of their issues closed. All of the top five overall ranked
+packages have most of their issues closed. Alive packages (11 use issue
+tracking) have 64\% of their issues closed, while dead packages (3 use issue
+tracking) have 71\% of their issues closed. This information is presented in
 Table~\ref{gitrepodata}. With respect to the host for the git repository, 13
@@
 The top four packages, ESPREsSo, Ludwig, Palabos, and OpenLB, have all of the
-common artifacts, except Palabos does not have theory notes and only three (ESPREsSo, Ludwig, Palabos) of the packages appear to use a version control system.
-The top four packages also have most of the less common artifacts.
-At the time of data collection, only one of the four packages (Palabos) did not have a user manual or guide, but there was a broken link on the package website indicating that such an artifact might exist.
-This broken link was later fixed, but this is not reflected in our data because it was not present at the time of data collection (mid 2020).
-Despite the broken link, Palabos does have a detailed and informative website.
-As far as we could observe, Ludwig does not appear to have a publicly visible design and OpenLB does not appear to use a version control system.
-However, OpenLB's use of version numbers suggests that version control may be used privately by the developers.
+common artifacts, except Palabos does not have theory notes and only three
+(ESPREsSo, Ludwig, Palabos) of the packages appear to use a version control
+system.  The top four packages also have most of the less common artifacts. At
+the time of data collection, only one of the four packages (Palabos) did not
+have a user manual or guide, but there was a broken link on the package website
+indicating that such an artifact might exist. This broken link was later fixed,
+but this is not reflected in our data because it was not present at the time of
+data collection (mid 2020). Despite the broken link, Palabos does have a
+detailed and informative website. As far as we could observe, Ludwig does not
+appear to have a publicly visible design and OpenLB does not appear to use a
+version control system. However, OpenLB's use of version numbers suggests that
+version control may be used privately by the developers.
 
@@
 tools, dependencies, and project management tools.  In this section we answer
-\rqref{RQ_CompareToolsProjMngmnt} by comparing aspects of tool usage in LBM software packages to their utilization in the research software community in general.
+\rqref{RQ_CompareToolsProjMngmnt} by comparing aspects of tool usage in LBM
+software packages to their utilization in the research software community in
+general.
 
@@
 As mentioned in Section~\ref{SecSurfCorrectAndVerifiab}, continuous integration
-is rarely used in LBM software (3 of 24 packages or 12.5\%).
-This contrasts with the frequency with which continuous integration is recommended in research software development guidelines \citep{BrettEtAl2021, Brown2015, ThielEtAl2020,
+is rarely used in LBM software (3 of 24 packages or 12.5\%). This contrasts with
+the frequency with which continuous integration is recommended in research
+software development guidelines \citep{BrettEtAl2021, Brown2015, ThielEtAl2020,
 Zadka2018, vanGompelEtAl2016}.  We could not find published data on the
@@
 
-Based on interviews with four developers, this section aims to answer the research questions: i) What are the pain points for developers working on research software projects (\rqref{RQ_PainPoints})?; and, ii) How do the pain points of developers from LBM compare to the pain points for research software in general (\rqref{RQ_ComparePainPoints})?
-Below we go through each of the identified pain points and include citations that contrast the LBM experience with observations from researchers in other domains.
-Potential ways to address the pain points are covered in Section~\ref{Sec_AddressConcerns}.
-The full interview questions are found in \citet{SmithEtAl2021}.
+Based on interviews with four developers, this section aims to answer the
+research questions: i) What are the pain points for developers working on
+research software projects (\rqref{RQ_PainPoints})?; and, ii) How do the pain
+points of developers from LBM compare to the pain points for research software
+in general (\rqref{RQ_ComparePainPoints})?  Below we go through each of the
+identified pain points and include citations that contrast the LBM experience
+with observations from researchers in other domains.  Potential ways to address
+the pain points are covered in Section~\ref{Sec_AddressConcerns}. The full
+interview questions are found in \citet{SmithEtAl2021}.
 
@@
 	\citep{CrouchEtAl2013}, and \href{https://software-carpentry.org/}{Software
-	Carpentry} \citep{WilsonAndLumsdaine2006, Wilson2016}. Over time awareness of these groups will grow, so this pain point is likely to disappear in the
+	Carpentry} \citep{WilsonAndLumsdaine2006, Wilson2016}. Over time awareness
+	of these groups will grow, so this pain point is likely to disappear in the
 	future for LBM and other research software developers.
@@
 We will not be repeating ideas here that have been discussed in previous
-sections, such as continuous integration, documentation of APIs, etc.
-Our aim is to mention ideas that are at least somewhat beyond conventional best practices.
-The ideas listed here have the potential to become best practices in the medium to long-term.
-The ideas are presented roughly in the order of increasing novelty.
+sections, such as continuous integration, documentation of APIs, etc.  Our aim
+is to mention ideas that are at least somewhat beyond conventional best
+practices. The ideas listed here have the potential to become best practices in
+the medium to long-term. The ideas are presented roughly in the order of
+increasing novelty.
 
@@
 ending of the measurement phase is separated by several months.  There is a
-threat that the measurement may be affected by unconscious biases, or the reviewers judgement may drift over time.
-In previous work \citep{SmithEtAl2016}, we demonstrated that the measurement process is reasonably reproducible by having five products graded by a second reviewer.
-The ranking via this independent review was almost identical to the original ranking.
-As long as each grader uses
-consistent definitions, the relative comparisons in the AHP results will be
-consistent between graders.
+threat that the measurement may be affected by unconscious biases, or the
+reviewers judgement may drift over time.  In previous work
+\citep{SmithEtAl2016}, we demonstrated that the measurement process is
+reasonably reproducible by having five products graded by a second reviewer. The
+ranking via this independent review was almost identical to the original
+ranking. As long as each grader uses consistent definitions, the relative
+comparisons in the AHP results will be consistent between graders.
 
@@
 
-Another data collection concern is the gap between the measurement of Musubi and the measurement of all other packages (Section~\ref{identifysoftware}).
+Another data collection concern is the gap between the measurement of Musubi and
+the measurement of all other packages (Section~\ref{identifysoftware}).
 Unfortunately we were not aware of the existence of Musubi until after the
@@
 of open issues, number of lines of code; iv) Fill in the measurement template
-(the template consists of 108 questions to assess nine qualities (installability, correctness/verifiability, reliability, robustness, usability, maintainability, reusability, understandability, and visibility/transparency).
-Filling in the template requires installing the software on a VM, running simple tests (like completing the getting started instructions (if present)), and searching the code, documentation and test files; v) Rank the software using the Analytic Hierarchy Process (AHP); vi) Interview developers (the interview consists of 20 questions and takes about an hour); and, vii) Conduct a domain analysis.
-The collected data was analyzed by: a) comparing the ranking by best practices against the ranking by popularity; b) comparing artifacts, tools and processes to current research software development guidelines; and, c) exploring pain points.
-Our ranking was roughly consistent with ranking the repositories by the number of GitHub stars, although the number one project by star count (Sailfish) is ranked 9th by our methodology.
+(the template consists of 108 questions to assess nine qualities
+(installability, correctness/verifiability, reliability, robustness, usability,
+maintainability, reusability, understandability, and visibility/transparency).
+Filling in the template requires installing the software on a VM, running simple
+tests (like completing the getting started instructions (if present)), and
+searching the code, documentation and test files; v) Rank the software using the
+Analytic Hierarchy Process (AHP); vi) Interview developers (the interview
+consists of 20 questions and takes about an hour); and, vii) Conduct a domain
+analysis. The collected data was analyzed by: a) comparing the ranking by best
+practices against the ranking by popularity; b) comparing artifacts, tools and
+processes to current research software development guidelines; and, c) exploring
+pain points.  Our ranking was roughly consistent with ranking the repositories
+by the number of GitHub stars, although the number one project by star count
+(Sailfish) is ranked 9th by our methodology.
 
@@
 benefit of users and developers.  This is illustrated by
-Table~\ref{topperformerstable}, which summarizes the top five performers for each of the identified qualities and for the overall quality.
-Taking the union of all entries in the table, we have 16 top performers: ESPREsSo, MechSys, OpenLB, DL\_MESO, Sailfish, Ludwig, Musubi, pyLBM, waLBerla, laboetie, lettuce,
+Table~\ref{topperformerstable}, which summarizes the top five performers for
+each of the identified qualities and for the overall quality. Taking the union
+of all entries in the table, we have 16 top performers: ESPREsSo, MechSys,
+OpenLB, DL\_MESO, Sailfish, Ludwig, Musubi, pyLBM, waLBerla, laboetie, lettuce,
 ESPREsSo++, LUMA, MP-LABS, Palabos, and HemeLB.  Therefore, two thirds (67\%) of
-the 24 measured packages appear in the top five for at least one quality.
-Of these entries almost all of them are classified as alive, except for three packages: laboetie, MP-LABS, and HemeLB.
-This suggests that alive packages are in a healthy state with respect to adopting best practices.
+the 24 measured packages appear in the top five for at least one quality.  Of
+these entries almost all of them are classified as alive, except for three
+packages: laboetie, MP-LABS, and HemeLB.  This suggests that alive packages are
+in a healthy state with respect to adopting best practices.
 
@@
 		\end{tabular}
-		\caption{Top performers for each quality (all packages were alive at the time of measurement, except for those marked by an asterisk (*))} \label{topperformerstable}
+		\caption{Top performers for each quality (all packages were alive at the
+		time of measurement, except for those marked by an asterisk (*))}
+		\label{topperformerstable}
 	\end{center}

### 2022-03-16 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Fixed Motamed to Keshavarz-Motamed

@@
 \newcommand{\jc}[1]{\authornote{red}{JC}{#1}} %Jacques Carette
-\newcommand{\zm}[1]{\authornote{magenta}{MN}{#1}} %Zahra Motamed
+\newcommand{\zm}[1]{\authornote{magenta}{MN}{#1}} %Zahra Keshavarz-Motamed
 \newcommand{\pmm}[1]{\authornote{cyan}{PM}{#1}} %Peter Michalski
@@
 \author[CAS]{Jacques Carette}
-\author[ME]{Zahra Motamed}
+\author[ME]{Zahra Keshavarz-Motamed}
 
@@
 For the current assessment, our Domain Expert (and paper co-author) is Dr.\
-Zahra Motamed, Assistant Professor of Mechanical Engineering at McMaster
-University, Hamilton, Ontario, Canada.
+Zahra Keshavarz-Motamed, Assistant Professor of Mechanical Engineering at
+McMaster University, Hamilton, Ontario, Canada.
 
@@
 contrast to the state of the practice of LBM software
-
-- mention MOOSE when mention CRAN for improving quality

### 2022-03-17 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Addition of Sadeghi et al references, as suggested by Zahra
Addition of comparison to ocean modelling software

@@
 \citep{AnwarAndSukop2009}, and analyzing the flow of blood in the cardiovascular
-system \citep{SadeghiEtAl2020}.  As these examples illustrate, LBM can be used
-for tackling problems that impact such areas as environmental policy,
-manufacturing, and health and safety.  Given the important applications of LBM,
-users of the libraries will be concerned with software qualities like
-reliability, robustness, reproducibility, performance, correctness and
-verifiability.  Since their time is valuable, developers that create or modify
-LBM libraries will have additional quality concerns, including maintainability,
-reusability, and understandability.  With the quality considerations in mind,
-and the potentially overwhelming number of choices for existing LBM libraries,
-the time is right to assess the state of the practice.  The goal of this report
-is to analyze the current state of LBM software development and compare it to
-the development practices employed for research software in general.  We want to
-highlight success stories that LBM developers can share amongst their community,
-and with the broader research software community, while at the same time
-looking for areas for potential future improvement.
+system \citep{SadeghiEtAl2022b, SadeghiEtAl2022, SadeghiEtAl2020}.  As these
+examples illustrate, LBM can be used for tackling problems that impact such
+areas as environmental policy, manufacturing, and health and safety.  Given the
+important applications of LBM, users of the libraries will be concerned with
+software qualities like reliability, robustness, reproducibility, performance,
+correctness and verifiability.  Since their time is valuable, developers that
+create or modify LBM libraries will have additional quality concerns, including
+maintainability, reusability, and understandability.  With the quality
+considerations in mind, and the potentially overwhelming number of choices for
+existing LBM libraries, the time is right to assess the state of the practice.
+The goal of this report is to analyze the current state of LBM software
+development and compare it to the development practices employed for research
+software in general.  We want to highlight success stories that LBM developers
+can share amongst their community, and with the broader research software
+community, while at the same time looking for areas for potential future
+improvement.
 
@@
 recommendation to have a README file, although, as for other software, the
-content of README files is not standardized \citep{PranaEtAl2018}.
+content of README files is not standardized \citep{PranaEtAl2018}.  Compared to
+ocean modelling software \citep{JungEtAl2022}, LBM software pays more
+attention to artifacts related to testing.
 
@@
 
+Based on information provided by \citet{JungEtAl2022}, tool utilization for
+LBM software has much in common with tool utilization for ocean modelling
+software.  Both use tools for editing, compiling, code management, testing,
+building and project management.  From the data available, ocean modelling
+differs from LBM software in its use of Kanban boards for project management.
+LBM software difference from the ocean domain through the use of continuous
+integration, and document generation tools.
+
 The poor adoption of version control tools that Wilson lamented in 2006
@@
 where change is inevitable, designers need to produce a modularization that
-supports change.
+supports change. Ocean modelling software is currently feeling the pain of
+not emphasizing modularization in legacy code \citep{Reiner2022}.
 
@@
 (Section~\ref{softwarequalities}), since the distinction confused some
-participants.
+participants.  Analysis of future interview data would likely benefit from a
+more rigorous process, like a thematic domain analysis \citep{JungEtAl2022}.
 
@@
 \end{document}
-
-TO DO
-
-- should look at ReinerEtAl2022 on Domain of Ocean Modelling software to
-contrast to the state of the practice of LBM software

### 2022-03-18 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Addition of figure of LBM for aorta artery flow

@@
 transferred among particles that collide \citep{bao2011lattice}. As an example
-of the output of LBM, Figure~\ref{circularflow} presents the streamlines of
-converged flow past a stationary circular cylinder with varying Reynolds
-numbers. \wss{Zahra can likely provide a better example.}  LBM has several
+of an important application of LBM, Figure~\ref{Fig_coarctation} shows the flow
+in a defective aorta pre and post medical intervention. LBM has several
 advantages over conventional CFD methods, including a simple calculation
@@
 	\begin{center}
-		\includegraphics[width=0.7\textwidth]{./figures/circularflow}
-		\caption{Streamlines of flow past a stationary circular cylinder at
-		Reynolds number = 10, 20, and 40 \citep{chen2021phase}}
-		\label{circularflow}
+		\includegraphics[width=0.7\textwidth]{./figures/FlowModellingCoarctation.pdf}
+		\caption{LBM (and Lattice Poisson-Boltzmann) computed flow (velocity
+		magnitude, turbulent kinetic energy and time averaged wall shear stress)
+		in an aorta with a coarctation (narrowing of the artery) and a major
+		aneurysm (a ballooning of the artery wall) downstream of the
+		coarctation, pre and post intravascular stent intervention
+		\citep{SadeghiEtAl2020}}.
+		\label{Fig_coarctation}
 	\end{center}
@@
 \citep{SmithEtAl2018_StatSoft} and medical image analysis software
-\citep{Dong2021}.  \wss{add citation to medical imaging software, if it is an
-option}
+\citep{Dong2021}.
 

### 2022-03-22 StateOfPractice/Papers/MedImageSoft_SOP-arXiv/MedImageSoft_SOP_AcademicRadiology.tex
Revisions to Introduction section, updated RQs

@@
 \newcommand{\done}[1]{\textcolor{black}{#1}}
+
+\newcounter{rqnum} %research question number
+\newcommand{\rqtherqnum}{RQ`'\therqnum}
+\newcommand{\rqref}[1]{RQ\ref{#1}}
+
+\newcounter{pnum} %pain point number
+\newcommand{\ppthepnum}{P`'\thepnum}
+\newcommand{\ppref}[1]{P\ref{#1}}
 
@@
 
-We present a general method to assess the state of the practice for Scientific
-Computing (SC) software and apply this method to Medical Imaging (MI) software.
-We selected 29 MI software projects from 48 candidates, assessed 10 software
-qualities (installability, correctness/ verifiability, reliability, robustness,
-usability, maintainability, reusability, understandability,
-visibility/transparency and reproducibility) by answering 103 questions for each
-software project, and interviewed 8 of the 29 development teams. Based on the
-quantitative data for the first 9 qualities, we ranked the MI software with the
-Analytic Hierarchy Process (AHP). The top three software products were
-\textit{3D Slicer}, \textit{ImageJ}, and \textit{OHIF Viewer}. By interviewing
-the developers, we identified three major pain points: i) lack of resources; ii)
-difficulty balancing between compatibility, maintainability, performance, and
-security; and, iii) lack of access to real-world datasets for testing. For
-future MI software projects, we propose adopting test-driven development, using
-continuous integration and continuous delivery (CI/CD), using git and GitHub,
-maintaining good documentation, supporting third-party plugins or extensions,
-considering web application solutions, and improving collaboration between
-different MI software projects.
+We present the state of the practice for Medical Imaging software. We selected
+29 medical imaging projects from 48 candidates, assessed 10 software qualities
+(installability, correctness/ verifiability, reliability, robustness, usability,
+maintainability, reusability, understandability, visibility/transparency and
+reproducibility) by answering 108 questions for each software project, and
+interviewed 8 of the 29 development teams. Based on the quantitative data for
+the first 9 qualities, we ranked the MI software with the Analytic Hierarchy
+Process (AHP). The top three software products were \textit{3D Slicer},
+\textit{ImageJ}, and \textit{OHIF Viewer}. By interviewing the developers, we
+identified three major pain points: i) lack of resources; ii) difficulty
+balancing between compatibility, maintainability, performance, and security;
+and, iii) lack of access to real-world datasets for testing. For future MI
+software projects, we propose adopting test-driven development, using continuous
+integration and continuous delivery (CI/CD), using git and GitHub, maintaining
+good documentation, supporting third-party plugins or extensions, considering
+web application solutions, and improving collaboration between different MI
+software projects. \wss{Update after the paper has been revised.}
 
@@
 \begin{keyword}
-	medical imaging, scientific computing, software engineering, software
+	medical imaging, research software, software engineering, software
 	quality, Analytic Hierarchy Process, developer interviews
@@
 analytic, and medical applications \citep{FDA2021, enwiki:1034887445,
-Zhang2008}.  Given the importance of MI software and the high number of
-competing software projects, we wish to understand the merits and drawbacks of
-the current development processes, tools and methodologies.  We aim to assess
-through a software engineering lens the quality of the existing software with
-the hope of highlighting standout examples, understanding current pain points
-and providing guidelines and recommendations for future development.
-
-\subsection{Purpose} \label{sec_motivation}
+Zhang2008}.  An example medical image is provided in Figure~\ref{Fig_Example}.
+Given the importance of MI software and the high number of competing software
+projects, we wish to understand the merits and drawbacks of the current
+development processes, tools and methodologies.  We aim to assess through a
+software engineering lens the quality of the existing software with the hope of
+highlighting standout examples, understanding current pain points and providing
+guidelines and recommendations for future development.
+
+\begin{figure}[!ht]
+    \begin{center}
+        \includegraphics[scale=0.42]{figures/Example_Aorta.png}        
+    \end{center}
+    \caption{Example medical image showing the segmentation of an aorta}
+    \label{Fig_Example}
+\end{figure}
+
+\subsection{Research Questions} \label{sec_motivation}
 
@@
 general. We wish to understand the impact of the often cited gap, or chasm,
-between software engineering and scientific programming \citep{Storer2017}.
-Although scientists spend a substantial proportion of their working hours on
-software development \citep{Hannay2009, Prabhu2011}, many developers learn
-software engineering skills by themselves or from their peers, instead of from
-proper training \citep{Hannay2009}. \citet{Hannay2009} observe that many
+between software engineering and scientific programming \citep{Kelly2007,
+Storer2017}. Although scientists spend a substantial proportion of their working
+hours on software development \citep{Hannay2009, Prabhu2011}, many developers
+learn software engineering skills by themselves or from their peers, instead of
+from proper training \citep{Hannay2009}. \citet{Hannay2009} observe that many
 scientists showed ignorance and indifference to standard software engineering
 concepts. For instance, according to a survey by \citet{Prabhu2011}, more than
-half of their 114 subjects did not use any proper debugger for their software.
-
-To gain insights, we devised five research questions, which can be applied to
-MI, as well as to other domains, of research software \citep{SmithEtAl2021}. The
+half of their 114 subjects did not use a proper debugger when coding.
+
+To gain insights, we devised 10 research questions, which can be applied to MI,
+as well as to other domains, of research software \citep{SmithEtAl2021}.  The
 questions are designed to learn about the community's interest in, and
 experience with, artifacts, tools, principles, processes, methodologies and
-qualities.  When we mention artifacts we mean documents, scripts and code that
-constitutes a software development project. Example artifacts include
+qualities.  When we mention artifacts we mean the documents, scripts and code
+that constitutes a software development project. Example artifacts include
 requirements, specifications, user manuals, unit tests, system tests, usability
 tests, build scripts, API (Application Programming Interface) documentation,
-READMEs, license documents, process documents, and code.  We aimed to collect
-enough information to understand the current pain points experienced by the MI
-software development community so that we can make some preliminary
-recommendations for future improvements.
-
-\begin{description}
-
-\item[RQ1.] \hypertarget{rq1}What artifacts are present in current MI software
-projects? What role does documentation play in the projects? What are the
-developers' attitude toward it?
-\item[RQ2.] \hypertarget{rq2}What tools are used in the development of MI packages?
-\item[RQ3.] \hypertarget{rq3}What principles, processes, and methodologies are
-used in the development of current MI packages?
-\item[RQ4.] \hypertarget{rq4}What are the pain points for developers working on
-MI software projects? What aspects of the existing processes, methodologies, and
-tools do they consider as potentially needing improvement? What changes to
-processes, methodologies, and tools can improve software development and
-software quality?
-\item[RQ5.] \hypertarget{rq5}What is the current status of the following
-software qualities for LBM projects? What actions have the developers taken to
-address them?  (Section~\ref{ch_background} provides definitions of the software
-qualities.)
-\begin{itemize}
-	\item Installability
-	\item Correctness \& Verifiability
-	\item Reliability
-	\item Robustness
-	\item Usability
-	\item Maintainability
-	\item Reusability
-	\item Understandability
-	\item Visibility/Transparency
-	\item Reproducibility
-\end{itemize}
-\item[RQ6.] \hypertarget{rq6}How does software designated as high quality by
-this methodology compare with top-rated software by the community?
-\end{description}
+READMEs, license documents, process documents, and code.  Once we have learned
+what MI developers do, we then put this information in context by contrasting MI
+software against the trends shown by developers in other research software
+communities.  Our aim is to collect enough information to understand the current
+pain points experienced by the MI software development community so that we can
+make some preliminary recommendations for future improvements.
+
+The questions are used to structure the discussion in the paper, so for each
+research question below we point to the section that contains our answer.  We
+start with identifying the relevant examples of MI software:
+
+\begin{enumerate}
+	\item[RQ\refstepcounter{rqnum}\therqnum \label{RQ_WhatProjects}:] What MI
+	software projects exist, with the constraint that the source code must be
+	available for all identified projects? (Section~\ref{ch_results})
+	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_HighestQuality}:] Which
+	of the projects identified in \rqref{RQ_WhatProjects} follow current best
+	practices, based on evidence found by experimenting with the software and
+	searching the artifacts available in each project's repository?
+	(Section~\ref{ch_results})
+	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareHQ2Popular}:] How
+	similar is the list of top projects identified in \rqref{RQ_HighestQuality}
+	to the most popular projects, as viewed by the scientific community?
+	(Section~\ref{Sec_VsCommunityRanking})
+    \item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareArtifacts}:] How
+	do MI projects compare to research software in general with respect to the
+	artifacts present in their repositories?
+	(Section~\ref{Sec_CompareArtifacts})
+	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareToolsProjMngmnt}:]
+	How do MI projects compare to research software in general with respect to
+	the use of tools (Section~\ref{Sec_CompareTools}) for:
+	\begin{enumerate} 
+		\item [\rqref{RQ_CompareToolsProjMngmnt}.a] development; and,
+		\item [\rqref{RQ_CompareToolsProjMngmnt}.b] project management?
+	\end{enumerate}
+	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareMethodologies}:]
+	How do MI projects compare to research software in general with respect to
+	principles, processes and methodologies used?
+	(Section~\ref{Sec_CompareMethodologies})
+	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_PainPoints}:] What are
+	the pain points for developers working on MI software projects?
+	(Section~\ref{ch_interview})
+	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_ComparePainPoints}:] How
+	do the pain points of developers from MI compare to the pain points
+	for research software in general? (Section~\ref{ch_interview})
+	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Concerns}:] For MI
+	developers what specific best practices are taken to address the pain points
+	and software quality concerns? (Section~\ref{ch_interview})
+	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Recommend}:]
+	What research software development practice could potentially address the
+	pain point concerns identified in \rqref{RQ_PainPoints}).
+	(Section~\ref{ch_recommendations})
+
+\end{enumerate}
 
@@
 
-% Based on our literature survey, we divided MI software into five sub-groups and
-% several sub-sub-groups by their major functions, as shown in
-% Figure~\ref{fig_mi_functions}.
-
-% \begin{figure}[!ht]
-% \centering
-% \includegraphics[scale=0.5]{figures/FunctionsOfMiSoftware.pdf}
-% \caption{Major functions of MI Software} \label{fig_mi_functions}
-% \end{figure}
-
-% \begin{figure}[!ht]
-% \centering
-% \begin{tikzpicture}[mindmap, grow cyclic, every node/.style=concept, concept
-% color=orange!40,
-% level 1/.append style={level distance=5cm,sibling angle=72},
-% level 2/.append style={level distance=3cm,sibling angle=90},
-% level 3/.style={level distance=2.3cm,sibling angle=90}]
-% \node{MI software}
-% child[concept color=teal!40] { node {Enhancement}}
-% child[concept color=teal!40] { node {Analysis}
-%     child[concept color=blue!30] { node {Registration}}
-%     child[concept color=blue!30,] { node {Segmentation}}
-%     child[concept color=blue!30] { node {Statistical Analysis}
-%     child[concept color=blue!20] { node {Feature Extraction}}
-%     child[concept color=blue!20] { node {Classification}}
-%     child[concept color=blue!20] { node {Interpretation}}
-%     }}
-% child[concept color=teal!40] { node {Simulation/\\Modeling}}
-% child[concept color=teal!40] { node {Supporting}
-%     child[concept color=blue!30] { node {Tool Kit}}
-%     child[concept color=blue!30] { node {PACS}}
-% }
-% child[concept color=teal!40] { node {Visualization}
-%     child[concept color=blue!30] { node {2D Display}}
-%     child[concept color=blue!30] { node {3D Rendering}}
-%     child[concept color=blue!30] { node {Reformatted Views}}
-% };
-% \end{tikzpicture}
-% \caption{Major functions of MI software}
-% \label{fig_mi_functions}
-% \end{figure}
-
 \subsection{Methodology Overview}
@@
 We designed a general methodology to assess the state of the practice for SC
-software \citep{SmithEtAl2021}. Details can be found in Section~\ref{ch_methods}
-and in \citep{Dong2021}.  This methodology builds off prior work to assess the
-state of the practice for such domains as Geographic Information Systems
-\citep{smith2018state}, Mesh Generators \citep{smith2016state}, Seismology
-software \citep{Smith2018Seismology}, and Statistical software for psychology
+software \citep{SmithEtAl2021}. Details can be found in
+Section~\ref{ch_methods}.  Our methodology has been applied to MI software
+\citet{Dong2021} and Lattice Boltzmann Solvers \citet{Michalski2021}.  This
+methodology builds off prior work to assess the state of the practice for such
+domains as Geographic Information Systems \citep{smith2018state}, Mesh
+Generators \citep{smith2016state}, Seismology software
+\citep{Smith2018Seismology}, and Statistical software for psychology
 \citep{smith2018statistical}.  In keeping with the previous methodology, we have
@@
 methodology, we interview some of the development teams to further understand
-the status of their development process. Finally, we summarize the measurements
-(Section~\ref{ch_results}), summarize the interview answers
-(Section~\ref{ch_interview}), answer the research questions
-(Section~\ref{ch_answers}) and propose recommendations
-(Section~\ref{ch_recommendations}) for future SC software development.
+the status of their development process.
 
@@
 assessing the software packages.  We also review the definitions of software
-qualities (Section \ref{sec_software_quality}) and provide an overview of the
+qualities (Section \ref{sec_software_quality}) and provide an overview of 
 AHP (Section \ref{sec_AHP}).
@@
 as the order qualities are considered in practice.  \wss{To be completed - not
-currently in sync with QDef paper, QDef paper not completely done.}
+currently in sync with QDef paper, QDef paper not completely done.  Look at LBM
+paper too.}
 
@@
 for $A$. However, rather than comparing the importance between criteria, we
-pairwise decide how much we favor one alternative over the other. We use the
+pairwise decide how much we favour one alternative over the other. We use the
 same method to calculate the local priority vector for each $B_i$.  The local
@@
 Biomedical Engineering, and Director of Medical Imaging Physics and Engineering
-at St. Joseph’s Healthcare, Hamilton, Ontario, Canada.
+at St. Joseph's Healthcare, Hamilton, Ontario, Canada.
 
@@
     \item whether the software breaks during installation;
-    \item the operation of the software following the getting started tutorial (if present);
+    \item the operation of the software following the getting started tutorial
+    (if present);
     \item whether the error messages are descriptive; and,
@@
 
-The repository measures help us in wo ways. Firstly, they help us with get a
-fast and accurate project overview. For example, the number of commits over the
-last 12 months shows how active a project has been, and the number of stars and
-forks may reveal its popularity. Secondly, the results may affect our decisions
+The repository measures help us in many ways. Firstly, they help us get a fast
+and accurate project overview. For example, the number of commits over the last
+12 months shows how active a project has been, and the number of stars and forks
+may reveal its popularity. Secondly, the results may affect our decisions
 regarding the grading scores for some software qualities. For example, if the
-percentage of comment lines is low, we double-check the
-understandability of the code; if the ratio of open versus closed pull
-requests is high, we pay more attention to maintainability.
+percentage of comment lines is low, we double-check the understandability of the
+code; if the ratio of open versus closed pull requests is high, we pay more
+attention to maintainability.
 
@@
 medInria \citep{Fillard2012} & 2009 & 2020-11 &  & 21 & 148924 & X & X & X &  \\
-BioImage Suite Web \citep{Papademetris2005} & 2018 & 2020-10 & X & 13 & 139699 & X & X & X & X \\
+BioImage Suite Web \citep{Papademetris2005} & 2018 & 2020-10 & X & 13 & 139699 &
+X & X & X & X \\
 Weasis \citep{Roduit2021} & 2010 & 2020-08 &  & 8 & 123272 & X & X & X &  \\
@@
 
-\section{Interviews with Developers} \label{ch_interview}
-
-The measurement results in Section \ref{ch_results} are based on information we
-collected from on-line resources. This information is incomplete because it
-doesn't generally capture the development process, the developer pain points,
-the perceived threats to software quality, and the developers' strategies to
-address these threats. To collect this information we interviewed nine
-developers from eight of the 29 MI software projects. The eight projects are
-\textit{3D Slicer}, \textit{INVESALIUS 3}, \textit{dwv}, \textit{BioImage Suite
-Web}, \textit{ITK-SNAP}, \textit{MRIcroGL}, \textit{Weasis}, and \textit{OHIF}.
-We spent about 90 minutes for each interview and asked 20 prepared questions
-(summarized in~\citet{SmithEtAl2021}). We also asked following-up questions as
-the conversation dictated. One participant was too busy to have an interview, so
-they wrote down their answers. The interviewees may have provided multiple
-answers to each question. Thus, when we summarize the frequency of different
-responses, the total is sometimes larger than nine.
-
-The full interview answers can be found in \citet{Dong2021}.  Selected questions
-and answers are highlighted and analyzed below.  The question numbers refer to
-the numbering in \citet{SmithEtAl2021}.
-
-\subsection{Current and Past Pain Points} \label{sec_interview_pain_points}
-
-The pain points include current and past obstacles, as explored via the
-following questions:
-
-\begin{description}
-\item[Q9.] \hypertarget{q9}Currently, what are the most significant obstacles in
-your development process?
-\item[Q10.] \hypertarget{q10}How might you change your development process to
-remove or reduce these obstacles?
-\item[Q12.] \hypertarget{q12}In the past, is there any major obstacle to your
-development process that has been solved? How did you solve it?
-\end{description}
-
-Table \ref{tab_obstacles} shows the number of times the interviewees mentioned
-the current and past obstacles in their projects.   We group these pain points
-into three major categories of obstacles: \textit{resource}, \textit{balance},
-and \textit{testing}. We put the less mentioned ones into the category
-\textit{Others}.
-
-\begin{table}[!ht]
-\centering
-\begin{tabular}{llll}
-\hline
-\multirow{2}{*}{Category} & \multirow{2}{*}{Obstacle} & \multicolumn{2}{c}{Num ans.} \\ \cline{3-4} 
- &  & current & past \\ \hline
-\multirow{2}{*}{Resource} & Lack of fundings & 3 &  \\
- & Lack of time to devote to the project & 2 & 1 \\ \hdashline
-\multirow{3}{*}{Balance} & Hard to keep up with changes in OS and libraries & 1 &  \\
- & Hard to support multiple OS & 2 &  \\
- & Hard to support lower-end computers & 1 & 2 \\ \hdashline
-Testing & Lack of access to real-world datasets for testing & 3 & 2 \\ \hdashline
-\multirow{7}{*}{Others}
- & Hard to have a high level roadmap from the start & 1 &  \\
- & Not enough participants for usability tests & 1 &  \\
- & Only a few people fully understand the large codebase & 1 &  \\
- & Hard to transfer to new technologies & & 2 \\
- & Hard to understand users' needs & & 1 \\
- & Hard to maintain good documentations & & 1 \\ \hline
-\end{tabular}
-\caption{\label{tab_obstacles}Current and past obstacles by the numbers of interviewees with the answers}
-\end{table}
-
-Developers identified a resource related pain point caused by a lack of funding
-and time.  Potential and proven solutions suggested by the interviewees include:
-
-\begin{itemize}
-\item Shifting from development to maintenance when the team does not have
-enough developers for building new features and fixing bugs at the same time;
-\item Licensing the software to commercial companies that integrate it into
-their products;
-\item Improving documentation to save time answering users' and developers'
-questions;
-\item Supporting third-party plugins and extensions; and,
-\item Using GitHub Actions for CI/CD to save time.
-\end{itemize}
-
-Many interviewees thought lack of fundings and lack of time were their most
-significant obstacles. The interviewees from \textit{3D Slicer} team and
-\textit{OHIF} team pointed out that it was more challenging to get fundings for
-software maintenance as opposed to research. The interviewee from the
-\textit{ITK-SNAP} team thought more fundings was a way to solve the lack of time
-problem, because they could hire more dedicated developers. On the other hand,
-the interviewee from the \textit{Weasis} team did not feel that fundings could
-solve the same problem, since he still would need a lot of time to supervise the
-project.  No interviewee suggested any solution to bring extra funding to the
-project. However, they provided ideas to save time, such as better
-documentation, third-party plugins, and good CI/CD tools.
-
-With respect to balance, developers expressed difficulty balancing between four
-factors: cross-platform compatibility, convenience to development \&
-maintenance, performance, and security.  The potential and proven solutions are:
-
-\begin{itemize}
-\item Adopting a web-based approach with backend servers, to better support
-lower-end computers;
-\item Using memory-mapped files to consume less computer memory, to better
-support lower-end computers, 
-\item Using computing power from the computers GPU for web applications;
-\item Increasing funding;
-\item Maintaining better documentations to ease the development \& maintenance
-processes;
-\item Improving performance via more powerful computers, which one interviewee
-pointed out has already happened to reduce the balance problem.
-\end{itemize}
-
-As the above list shows, web-based applications are perceived to help the
-balance problem.  Table~\ref{tab_native_vs_web} shows the teams' choices between
-native application and web application. In all the 29 teams on our list, most of
-them chose to develop native applications. For the eight teams we interviewed,
-three of them were building web applications, and the \textit{MRIcroGL} team was
-considering web-based solutions. So we had a good chance to discuss the
-differences between the two choices with the interviewees.
-
-\begin{table}[!ht]
-\centering
-\begin{tabular}{lll}
-\hline
-Software team & Native application & Web application \\ \hline
-3D Slicer & X & \\
-INVESALIUS 3 & X & \\
-dwv & & X \\
-BioImage Suite Web & & X \\
-ITK-SNAP & X & \\
-MRIcroGL & X & \\
-Weasis & X & \\
-OHIF & & X \\ \hdashline
-Total number among the eight teams & 5 & 3 \\
-Total number among the 29 teams & 24 & 5 \\ \hline
-\end{tabular}
-\caption{\label{tab_native_vs_web}Teams' choices between native application and
-web application}
-\end{table}
-
-The advantage for native applications is higher performance, while web
-applications have the advantage of cross-platform compatibility and a simpler
-build process.  These web advantages are mirrored by the native application
-disadvantages of difficulty with cross-platform compatibility and a complex
-build process.  The lower performance disadvantage of web-applications can be
-improved with a server backend, but in this case there are disadvantages for
-privacy protection and the cost of the servers.
-
-For the pain point category of testing developers identified the the lack of access to
-real-world datasets.  Some proposed and proven solutions are as follows:
-
-\begin{itemize}
-\item Using open datasets
-\item Asking the users to provide de-identified copies of medical images if they
-have problems loading the images;
-\item Sending the beta versions of software to medical workers who can access
-the data and complete the tests;
-\item If (part of) the team belongs to a medical school or a hospital, using the
-datasets they can access;
-\item If the team has access to MRI scanners, self-building sample images for
-testing;
-\item If the team has connections with MI equipment manufacturers, asking for
-their help on data format problems;
-\item Storing all images that cause special problems, and maintaining this
-special dataset over time.
-\end{itemize}
-
-No interviewee provided a perfect solution to the testing problem. However,
-increased connections between the development team and medical
-professionals/institutions could ease the pain.
-
-\subsection{Documents in the Projects} \label{sec_interview_documents}
-
-We tried to understand the interviewees' opinions on documentation and the
-quality of documentations with two questions:
-
-\begin{description}
-\item[Q11.] How does documentation fit into your development process? Would
-improved documentation help with the obstacles you typically face?
-\item[Q19.] Do you think the current documentation can clearly convey all
-necessary knowledge to the users? If yes, how did you successfully achieve it?
-If no, what improvements are needed?
-\end{description}
-
-Table~\ref{tab_opinion_doc} summarizes interviewees' opinions on documentation.
-Interviewees from each of the eight projects thought that documentation was
-important to their projects, and most of them said that it could save their time
-to answer questions from users and developers. Most of them saw the need to
-improve their documentation, and only three of them thought that their
-documentations conveyed information clearly enough.
-
-\begin{table}[!ht]
-\centering
-\begin{tabular}{ll}
-\hline
-Opinion on documentation & Num ans. \\ \hline
-Documentation is vital to the project& 8 \\
-Documentation of the project needs improvements & 7 \\
-Referring to documentation saves time to answer questions & 6 \\
-Lack of time to maintain good documentation & 4 \\
-Documentation of the project conveys information clearly & 3 \\
-Coding is more fun than documentation & 2 \\
-Users help each other by referring to documentation & 1 \\ \hline
-\end{tabular}
-\caption{\label{tab_opinion_doc}Opinions on documentation by the numbers of
-interviewees with the answers}
-\end{table}
-
-For documentation tools and methods mentioned by the interviewees, the most
-popular (mentioned by about 30\% of developers) were forum discussions and
-videos.  The second most popular options (mentioned by about 20\% of developers)
-were GitHub, wiki pages, workshops, and social media.  The least frequently
-mentioned options (about 10\% of developers) included writing books, google
-forms and state management.
-
-\subsection{Project Management} \label{sec_contribution_pm}
-
-To investigate project management we asked three questions:
-
-\begin{description}
-\item[Q5.] Do you have a defined process for accepting new contributions into
-your team?
-\item[Q13.] What is your software development model? For example, waterfall,
-agile, etc.
-\item[Q14.] What is your project management process? Do you
-think improving this process can tackle the current problem? Were any project
-management tools used?
-\end{description}
-
-Although teams may have a process for accepting new contributions, no one
-discussed this during their interview. However, most teams (8 of 9) mentioned
-using GitHub and pull requests to manage contributions from the community. The
-interviewees generally gave very positive feedback on using GitHub. Some teams
-previously used a different approach to version control and eventually
-transferred to git and GitHub.  The past approaches included contributions
-from e-mail (3 teams), contributions from forums (1 team) and sharing the
-git repository via e-mail (1 team).  \wss{I do not know what this last one means.
-Should ask Ao.} \ad{The last one happened when cloud-based repositories (Github,
-Gitlab, Bitbucket, etc.) were not widely used. The project git repository was
-stored locally on the project owner's local computer. The owner shared the whole
-git repo to team members and other individuals who wanted to contribute. The
-contributors made changes to the repo, and sent it back. They transferred the
-repo back and forth by email.}
-
-For managing contributions, the \textit{3D Slicer} team encouraged users to
-develop their extensions for specific use cases, while the \textit{OHIF} team
-enabled the use of plug-ins.  The interviewee from the \textit{ITK-SNAP} team
-said one way of accepting new team members was through funded academic projects.
-
-For the most part, the responses about development model were vague, with only
-two interviewees following a definite development model. In some cases the
-interviewees felt their process was similar to an existing development model.
-Three teams (about 38\%) either followed agile, or something similar to agile.
-Two teams (25\%) either followed a waterfall process, or something similar.
-Three teams (about 38\%) explicitly stated that their process was undefined or
-self-directed.
-
-Some interviewees mentioned the project management tools they used. Generally
-speaking, the interviewees talked about two types of tools:
-\begin{inparaenum}[i)]
-\item trackers, including GitHub, issue trackers, bug trackers and Jira; and,
-\item documentation tools, including GitHub, Wiki page, Google Doc, and Confluence.
-\end{inparaenum}
-Of the specifically named tools in the above lists, GitHub was mentioned 3
-times, while each of the other tools was only mentioned once.
-
-No interviewee introduced any strictly defined project management process. The
-most common approach was following the issues, such as bugs and feature requests.
-Additionally, the \textit{3D Slicer} team had weekly meetings to discuss the
-goals for the project; the \textit{INVESALIUS 3} team relied on the GitHub
-process for their project management; the \textit{ITK-SNAP} team had a fixed
-six-month release pace; only the interviewee from the \textit{OHIF} team
-mentioned that the team has a project manager; the \textit{3D Slicer} team and
-\textit{BioImage Suite Web} team do nightly builds and tests.
-
-Most interviewees skipped the part of the last question asking ``Do you think
-improving this process can tackle the current problem?''. In retrospect, we
-should not have asked a yes-or-no question, since a binary answer is not very
-informative. The interviewee from the \textit{OHIF} team gave a positive answer
-to this question. They believed that a better project management process can
-improve the efficiency of junior developers. They also improved the project
-management tools (from public Jira to public GitHub repository plus private
-Jira), so they could better communicate externally and internally.
-
-\subsection{Discussions on Software Qualities} \label{sec_interview_software_qualities}
-
-Questions 15--18, and 20 cover the software qualities of correctness,
-maintainability, understandability, usability, and reproducibility,
-respectively. We discuss each quality below.
-
-\begin{description}
-\item[Q15.] Was it hard to ensure the correctness of the software? If there were
-any obstacles, what methods have been considered or practiced to improve the
-situation? If practiced, did it work?
-\end{description}
-
-The interviewees identified multiple threats to correctness.  The most
-frequently mentioned threat was complexity.  Complexity enters the software by
-various means, including a variety of data formats, complicated data standards,
-differing outputs between medical imaging machines, and the addition of
-(non-viewing related) functionality.  Other threats to correctness identified by at least one time
-include the following:
-\begin{itemize}
-\item Lack of real world image data for testing;
-\item The team cannot use private data for debugging even when the data causes
-problems; \wss{Why not?  Should ask Ao} \ad{For example, when a physician faces
-some problems to use the software to display a medical image, he/she might reach
-to the team for help. However, the image is private, so the physician can only
-describe the problems, but not share the image with the team to reproduce the
-problems for debugging.}
-\item Expense and time consumed by tests because of the huge datasets;
-\item Difficulty managing releases;
-\item No unit tests; and,
-\item No dedicated quality assurance team.
-\end{itemize}
-
-Testing was the most often mentioned strategy for ensuring correctness.  Seven
-teams mentioned test related activities, including test-driven development,
-component tests, integration tests, smoke tests, regression tests, self tests
-and automated tests.  Another approach frequently adopted (mentioned by 3
-interviewees) is a two state development process with stable releases and
-nightly builds.  Other strategies for ensuring correctness that came up during
-the interviews include CI/CD, using de-identified copies of medical images for
-debugging, sending beta versions to medical workers who can access the data to
-do the tests, and collecting/maintaining a dataset of problematic images.
-
-\begin{description}
-\item[Q16.] When designing the software, did you consider the ease of future
-changes? For example, will it be hard to change the system’s structure, modules,
-or code blocks? What measures have been taken to ensure the ease of future
-changes and maintains?
-\end{description}
-
-The most popular, with five out of nine interviewees mentioning it, strategy for
-ensuring maintainability was to use a modular approach, with often repeated
-functions in a library.  Other strategies that were mentioned for improving
-maintainability include supporting third-party extensions, an easy-to-understand
-architecture, a dedicated architect, starting from simple solutions, and
-documentation.  The \textit{3D Slicer} team used a well-defined structure for
-the software, which they named as an ``event-driven MVC pattern''. Moreover,
-\textit{3D Slicer} discovers and loads necessary modules at runtime, according
-to the configuration and installed extensions. The \textit{BioImage Suite Web}
-team had designed and re-designed their software multiple times in the last 10+
-years. They found that their modular approach effectively supports
-maintainability \citep{Joshi2011}.
-
-\begin{description}
-\item[Q17.] Provide instances where users have misunderstood the software. What,
-if any, actions were taken to address understandability issues?
-\end{description}
-
-The discussion with the developers focused on understandability issues for two
-classes of users: the end users and other developers.  The threats to
-understandability for end users include users not understanding how to use
-features (\wss{Because the users don't have sufficient background?  Ask Ao.})
-\ad{Did you mean medical or medical imaging background? No, it was majorly
-caused by the software usability. Based on the context of our interview
-conversations, it means that users don't understand how to complete certain
-operations or what functions a feature provides. It's due to unclear description,
-lack of documentation, poorly designed UI or nonintuitive process},
-the team not having a dedicated user experience (UX) designer, some important
-indicators are not noticeable (e.g. a progress bar), not all users understand
-the purpose of the software, not all users know if the software includes certain
-features, not all users understand how to use the command line tool, not all
-users understand that the software is a web application. For developers the
-threats to understandability include developers not understanding how to deploy
-the software (\wss{This sounds like a consequence of a lack of
-understandability, rather than threat to it.  Or is the point that developers
-lack sufficient background?  Ask Ao}) \ad{Yes, this should be a consequence of a
-lack of understandability. I think the direct cause is that the project doesn't
-clearly explain how to deploy. The background of the developers wasn't discussed
-at that point.}, and the architecture is difficult for new
-developers to understand.
-
-The most common strategy, cited by 4 developers, for ensuring understandability
-was to use documentation (user manuals, mailing lists, forums).  Other suggested
-and practiced strategies include a graphical user interface, testing every
-release with active users, making simple things simple and complicated things
-possible, icons with clear visual expressions, designing the software to be
-intuitive, having a UX designer with the right experience, dialog windows for
-important notifications, and providing an example for users to follow.
-
-\begin{description}
-\item[Q18.] What, if any, actions were taken to address usability issues?
-\end{description}
-
-The two most frequently mentioned strategies (mentioned 3 times each) for
-improving usability are:
-\begin{inparaenum}[i)]
-    \item usability tests and interviews with end users; and,
-    \item adjusting the software according to user feedback.
-\end{inparaenum}
-Other ideas that came up include a straightforward and intuitively designed
-interface / professional UX designer, providing step-by-step processes, making
-the basic functions easy to use without reading the documentation, focusing on
-limited number of functions, making the software more streamlined, downsampling
-images to consume less memory, and an option to load only part of the data to
-boost performance.
-
-\begin{description}
-\item[Q20.] Do you have any concern that your computational results won’t be
-reproducible in the future? Have you taken any steps to ensure reproducibility?
-\end{description}
-
-We discussed threats to reproducibility and strategies for improving it.  The
-threats that were mentioned include closed-source software, no user interaction
-tests, no unit tests, using different versions of some common libraries
-\wss{because the interface and/or behaviour of the library could have changed?
-Ask Ao} \ad{Yes. Behaviour change could be caused by a lib changing to use
-different algorithms or different dependencies}, variability
-between CPUs, and misinterpretation of how manufacturers
-create medical images. The most commonly cited (by 6 teams) strategy to improve
-reproducibility was testing (regression tests, unit tests, having good tests).
-The second most common strategy (mentioned by 5 teams) is making code, data, and
-documentation available, possibly by creating open-source libraries.  Other
-ideas that were mentioned include running the same tests on all platforms, a
-dockerized version of the software to insulate it from the OS environment, using
-standard libraries, monitoring the upgrades of the library dependencies, clearly
-documenting the version information, bringing along the exact versions of all
-the dependencies with the software, providing checksums of the data, and
-benchmarking the software against other software that overlaps in functionality.
-Specifically one interviewee suggested using \textit{3D Slicer} as the benchmark
-to test their reproducibility.
-
-\section{Answers to Research Questions} \label{ch_answers}
-
-This section answers the research questions from
-Section~\ref{sec_motivation}.  We will go through each of the questions
-in turn.
-
-\begin{description}
-\item[RQ1.] What artifacts are present in current software projects?
-\end{description}
-
-We answered this question by examining and analyzing the repositories for each
-of the 29 projects.  In particular, as part of assessing maintainability, we
-listed the artifacts (documents, test scripts etc.) that appear in each project.
-The full measurement results are summarized in \citet{Dong2021-Data}.  The
-artifacts found were as follows (with the number of occurrences in brackets):
-Source code (29), README (29), Version Control (29), License (28), Bug tracker
-(28), Change request (28), User Manual (22), Release notes (22), Build file
-(18), Tutorials (18), Installation Guide (16), Test cases (15), Authors (14),
-FAQ (14), Acknowledgements (12), Executable files (10), Developer's Manual (8),
-API documentation (7), Troubleshooting guide (6), Project Plan (5).  The
-traceability of some document types to the projects in which they occur was
-previously presented in Table~\ref{tab_maintainability_docs} for maintainability
-related documents, and Table~\ref{tab_Visibility/Transparency_docs} for
-visibility/transparency related documents.
-
-The list in the previous paragraph shows that some types of software engineering
-recommended artifacts are missing in MI projects.  For instance, we did not find
-evidence of user stories (an Agile software development tool for capturing
-software features from an end-user perspective), user requirements, test plans,
-test code coverage, etc.  Our preliminary work suggests that this is not unique
-to MI software.  A gap exists in scientific computing development practices and
-software engineering recommendations \citep{OwojaiyeEtAl2021_CSE}.
-
-\begin{description}
-\item[RQ2.] What tools are used in the development of current software packages?
-\end{description}
-
-We answer this question by measuring the qualities and interviewing the
-developers. We learned that developers use tools for CI/CD
-(Section~\ref{sec_result_correctness_verifiability}), user support
-(Table~\ref{tab_user_support_model}), version control
-(Section~\ref{sec_score_maintainability}), documentation
-(Section~\ref{sec_contribution_pm}), contribution management
-(Section~\ref{sec_contribution_pm}), and project management
-(Section~\ref{sec_contribution_pm}).
-
-As mentioned in Section \ref{sec_result_correctness_verifiability}, we
-identified five projects using CI/CD tools: \textit{3D Slicer} and \textit{OHIF
-Viewer} use CircleCI; \textit{ImageJ}, \textit{Fiji}, and \textit{dwv} use
-Travis. We identified the above projects and tools by examining the
-documentation and source code of all projects. Not all tools leave identifiable
-artifacts in the repositories; therefore, we may have missed the use of some
-tools. According to the interviews with developers, \textit{dwv} and
-\textit{Weasis} used GitHub Actions.
-
-\begin{description}
-\item[RQ3.] What principles, processes, and methodologies are used in the
-development of current software packages?
-\end{description}
-
-We answer this question by measuring the qualities and interviewing the
-developers. We previously summarized principles, processes, and methodologies in
-software testing (Section~\ref{sec_interview_pain_points} and
-Section~\ref{sec_interview_software_qualities}), documentation
-(Section~\ref{sec_interview_documents}, Table~\ref{tab_maintainability_docs},
-Table~\ref{tab_Visibility/Transparency_docs}, and Section~\ref{ch_answers}),
-contribution management (Section~\ref{sec_contribution_pm}), and project
-management (Section~\ref{sec_contribution_pm}).
-
-We identified the use of unit testing in less than half of the 29 projects. On
-the other hand, the interviewees believed that testing (including usability
-tests with users) was the top solution to improve correctness, usability, and
-reproducibility. One pain point in the development process is the lack of access
-to real-world datasets for testing. The developers' strategies to address this
-are summarized in Section~\ref{sec_pain_points_3} \ad{to be fixed}. One threat to correctness is:
-with huge datasets for testing, the tests are expensive and time-consuming.
-Three interviewees endorsed self tests / automated tests, which may save time
-for testing.
-
-Documentation is apparently not emphasized in any of the 29 projects.  None of
-them had theory manuals, although we did identify a road map in the \textit{3D
-Slicer} project.  No requirements specifications were found. Eight of the nine
-interviewees thought that documentation was essential to their projects.
-However, they hold the common opinion that their documentation needed
-improvements. Nearly half of them also believed that the lack of time prevented
-them from improving the documentation.
-
-\begin{description}
-\item[RQ4.] What are the pain points for developers working on research software
-projects? What aspects of the existing processes, methodologies, and tools do
-they consider as potentially needing improvement? What changes to processes,
-methodologies, and tools can improve software development and software quality?
-\end{description}
-
-We answer this question by interviewing the developers. The answers to this
-question, including the pain points and the solutions proposed by the
-developers, are in Section~\ref{sec_interview_pain_points}.
-
-\begin{description}
-\item[RQ5.] What is the current status of the following
-software qualities for the projects? What actions have the developers taken to
-address them?
-\end{description}
-
-The current status of the software qualities was assessed through filling in the
-measurement template \citep{Dong2021-Data}.  A summary of the comparison between
-all projects can be found in Section~\ref{ch_results}.  We explored some of the
-qualities more deeply, especially with respect to techniques addressing quality
-concerns, through the interview questions.  Specifically, the qualities of
-correctness \& verifiability, usability, maintainability, understandability and
-reproducibility are explored in Section~\ref{ch_interview}.
-
-\begin{description}
-\item[RQ6.] How does software designated as high quality by this methodology
-compare with top-rated software by the community?
-\end{description}
-
-We answer this question by grading the qualities of the software, collecting
-ratings from the MI community, then conducting two comparisons:
+\section{Comparison to Community Ranking} \label{Sec_VsCommunityRanking}
+
+To address \rqref{RQ_CompareHQ2Popular}, Table~\ref{tab_ranking_vs_GitHub}
+compares our LBM software package rankings against their popularity in the
+research community, as estimated by repository stars and watches. We answer this
+question by grading the qualities of the software, collecting ratings from the
+MI community, then conducting two comparisons:
 \begin{itemize}
@@
 
-\wss{In the shorter journal paper version, the focus could be just on answering the 
-research questions and deleting the previous two sections.}
+\section{Comparison Between LBM and Research Software for Artifacts}
+\label{Sec_CompareArtifacts}
+
+As part of filling in the measurement template the LBM packages were examined
+for the presence of artifacts, which were then categorized by frequency. We have
+grouped them into common, less common, and rare artifacts in
+Table~?. Common artifacts were found in 16 to 24 ($>$63\%)
+of the software packages. Less common artifacts were found in 8 to 15 (30-63\%)
+of the software packages. Rare artifacts were found in 1 to 7 ($<$30\%) of the
+software packages.  In this section we answer \rqref{RQ_CompareArtifacts} by
+comparing the artifacts that we observed in LBM repositories to those observed
+and recommended for research software in general. Our comparison may point out
+areas where some LBM software packages fall short of current best practices.
+This is not intended to be a criticism of any existing packages, especially
+since in practice not every project needs to achieve the highest possible
+quality.  However, rather than delve into the nuances of which software can
+justify compromising which practices we will write our comparison under the
+ideal assumption that every project has sufficient resources to match best
+practices.
+
+\section{Comparison of Tool Usage Between LBM and Other Research Software}
+\label{Sec_CompareTools}
+
+Software tools are used to support the development, verification, maintenance,
+and evolution of software, software processes, and artifacts \citep[p.\
+501]{GhezziEtAl2003}. Many tools are used by LBM software packages, as
+summarized in Table~?.  The tools are subdivided into development
+tools, dependencies, and project management tools.  In this section we answer
+\rqref{RQ_CompareToolsProjMngmnt} by comparing aspects of tool usage in LBM
+software packages to their utilization in the research software community in
+general.
+
+\section{Comparison of Principles, Process and Methodologies to Research Software in General} \label{Sec_CompareMethodologies}
+
+This section answers research question \rqref{RQ_CompareMethodologies} by
+comparing the principles, processes and methodologies used for LBM software to
+what can be gleaned from the literature on research software in general. In our
+data collection for LBM software, the software development process is not
+explicitly indicated in the artifacts for most of the packages. However, during
+our interviews one developer (ESPResSo) told us their non-rigorous development
+model is like a combination of agile and waterfall. Employing a loosely defined
+process makes sense for LBM software, given that the teams are generally small
+and self-contained.  Although eleven of the packages explicitly convey that they
+would accept outside contributors, generally the teams are centralized, often
+working at the same institution.  Working at the same institution means that an
+informal process can show success, since informal conversations are relatively
+easy to have.
+
+\section{Interviews with Developers} \label{ch_interview}
+
+The measurement results in Section \ref{ch_results} are based on information we
+collected from on-line resources. This information is incomplete because it
+doesn't generally capture the development process, the developer pain points,
+the perceived threats to software quality, and the developers' strategies to
+address these threats. To collect this information we interviewed nine
+developers from eight of the 29 MI software projects. The eight projects are
+\textit{3D Slicer}, \textit{INVESALIUS 3}, \textit{dwv}, \textit{BioImage Suite
+Web}, \textit{ITK-SNAP}, \textit{MRIcroGL}, \textit{Weasis}, and \textit{OHIF}.
+We spent about 90 minutes for each interview and asked 20 prepared questions
+(summarized in~\citet{SmithEtAl2021}). We also asked following-up questions as
+the conversation dictated. One participant was too busy to have an interview, so
+they wrote down their answers. The interviewees may have provided multiple
+answers to each question. Thus, when we summarize the frequency of different
+responses, the total is sometimes larger than nine.
+
+The full interview answers can be found in \citet{Dong2021}.  Selected questions
+and answers are highlighted and analyzed below.  The question numbers refer to
+the numbering in \citet{SmithEtAl2021}.
+
+\subsection{Current and Past Pain Points} \label{sec_interview_pain_points}
+
+The pain points include current and past obstacles, as explored via the
+following questions:
+
+\begin{description}
+\item[Q9.] \hypertarget{q9}Currently, what are the most significant obstacles in
+your development process?
+\item[Q10.] \hypertarget{q10}How might you change your development process to
+remove or reduce these obstacles?
+\item[Q12.] \hypertarget{q12}In the past, is there any major obstacle to your
+development process that has been solved? How did you solve it?
+\end{description}
+
+Table \ref{tab_obstacles} shows the number of times the interviewees mentioned
+the current and past obstacles in their projects.   We group these pain points
+into three major categories of obstacles: \textit{resource}, \textit{balance},
+and \textit{testing}. We put the less mentioned ones into the category
+\textit{Others}.
+
+\begin{table}[!ht]
+\centering
+\begin{tabular}{llll}
+\hline
+\multirow{2}{*}{Category} & \multirow{2}{*}{Obstacle} & \multicolumn{2}{c}{Num ans.} \\ \cline{3-4} 
+ &  & current & past \\ \hline
+\multirow{2}{*}{Resource} & Lack of fundings & 3 &  \\
+ & Lack of time to devote to the project & 2 & 1 \\ \hdashline
+\multirow{3}{*}{Balance} & Hard to keep up with changes in OS and libraries & 1 &  \\
+ & Hard to support multiple OS & 2 &  \\
+ & Hard to support lower-end computers & 1 & 2 \\ \hdashline
+Testing & Lack of access to real-world datasets for testing & 3 & 2 \\ \hdashline
+\multirow{7}{*}{Others}
+ & Hard to have a high level roadmap from the start & 1 &  \\
+ & Not enough participants for usability tests & 1 &  \\
+ & Only a few people fully understand the large codebase & 1 &  \\
+ & Hard to transfer to new technologies & & 2 \\
+ & Hard to understand users' needs & & 1 \\
+ & Hard to maintain good documentations & & 1 \\ \hline
+\end{tabular}
+\caption{\label{tab_obstacles}Current and past obstacles by the numbers of interviewees with the answers}
+\end{table}
+
+Developers identified a resource related pain point caused by a lack of funding
+and time.  Potential and proven solutions suggested by the interviewees include:
+
+\begin{itemize}
+\item Shifting from development to maintenance when the team does not have
+enough developers for building new features and fixing bugs at the same time;
+\item Licensing the software to commercial companies that integrate it into
+their products;
+\item Improving documentation to save time answering users' and developers'
+questions;
+\item Supporting third-party plugins and extensions; and,
+\item Using GitHub Actions for CI/CD to save time.
+\end{itemize}
+
+Many interviewees thought lack of fundings and lack of time were their most
+significant obstacles. The interviewees from \textit{3D Slicer} team and
+\textit{OHIF} team pointed out that it was more challenging to get fundings for
+software maintenance as opposed to research. The interviewee from the
+\textit{ITK-SNAP} team thought more fundings was a way to solve the lack of time
+problem, because they could hire more dedicated developers. On the other hand,
+the interviewee from the \textit{Weasis} team did not feel that fundings could
+solve the same problem, since he still would need a lot of time to supervise the
+project.  No interviewee suggested any solution to bring extra funding to the
+project. However, they provided ideas to save time, such as better
+documentation, third-party plugins, and good CI/CD tools.
+
+With respect to balance, developers expressed difficulty balancing between four
+factors: cross-platform compatibility, convenience to development \&
+maintenance, performance, and security.  The potential and proven solutions are:
+
+\begin{itemize}
+\item Adopting a web-based approach with backend servers, to better support
+lower-end computers;
+\item Using memory-mapped files to consume less computer memory, to better
+support lower-end computers, 
+\item Using computing power from the computers GPU for web applications;
+\item Increasing funding;
+\item Maintaining better documentations to ease the development \& maintenance
+processes;
+\item Improving performance via more powerful computers, which one interviewee
+pointed out has already happened to reduce the balance problem.
+\end{itemize}
+
+As the above list shows, web-based applications are perceived to help the
+balance problem.  Table~\ref{tab_native_vs_web} shows the teams' choices between
+native application and web application. In all the 29 teams on our list, most of
+them chose to develop native applications. For the eight teams we interviewed,
+three of them were building web applications, and the \textit{MRIcroGL} team was
+considering web-based solutions. So we had a good chance to discuss the
+differences between the two choices with the interviewees.
+
+\begin{table}[!ht]
+\centering
+\begin{tabular}{lll}
+\hline
+Software team & Native application & Web application \\ \hline
+3D Slicer & X & \\
+INVESALIUS 3 & X & \\
+dwv & & X \\
+BioImage Suite Web & & X \\
+ITK-SNAP & X & \\
+MRIcroGL & X & \\
+Weasis & X & \\
+OHIF & & X \\ \hdashline
+Total number among the eight teams & 5 & 3 \\
+Total number among the 29 teams & 24 & 5 \\ \hline
+\end{tabular}
+\caption{\label{tab_native_vs_web}Teams' choices between native application and
+web application}
+\end{table}
+
+The advantage for native applications is higher performance, while web
+applications have the advantage of cross-platform compatibility and a simpler
+build process.  These web advantages are mirrored by the native application
+disadvantages of difficulty with cross-platform compatibility and a complex
+build process.  The lower performance disadvantage of web-applications can be
+improved with a server backend, but in this case there are disadvantages for
+privacy protection and the cost of the servers.
+
+For the pain point category of testing developers identified the the lack of access to
+real-world datasets.  Some proposed and proven solutions are as follows:
+
+\begin{itemize}
+\item Using open datasets
+\item Asking the users to provide de-identified copies of medical images if they
+have problems loading the images;
+\item Sending the beta versions of software to medical workers who can access
+the data and complete the tests;
+\item If (part of) the team belongs to a medical school or a hospital, using the
+datasets they can access;
+\item If the team has access to MRI scanners, self-building sample images for
+testing;
+\item If the team has connections with MI equipment manufacturers, asking for
+their help on data format problems;
+\item Storing all images that cause special problems, and maintaining this
+special dataset over time.
+\end{itemize}
+
+No interviewee provided a perfect solution to the testing problem. However,
+increased connections between the development team and medical
+professionals/institutions could ease the pain.
+
+\subsection{Documents in the Projects} \label{sec_interview_documents}
+
+We tried to understand the interviewees' opinions on documentation and the
+quality of documentations with two questions:
+
+\begin{description}
+\item[Q11.] How does documentation fit into your development process? Would
+improved documentation help with the obstacles you typically face?
+\item[Q19.] Do you think the current documentation can clearly convey all
+necessary knowledge to the users? If yes, how did you successfully achieve it?
+If no, what improvements are needed?
+\end{description}
+
+Table~\ref{tab_opinion_doc} summarizes interviewees' opinions on documentation.
+Interviewees from each of the eight projects thought that documentation was
+important to their projects, and most of them said that it could save their time
+to answer questions from users and developers. Most of them saw the need to
+improve their documentation, and only three of them thought that their
+documentations conveyed information clearly enough.
+
+\begin{table}[!ht]
+\centering
+\begin{tabular}{ll}
+\hline
+Opinion on documentation & Num ans. \\ \hline
+Documentation is vital to the project& 8 \\
+Documentation of the project needs improvements & 7 \\
+Referring to documentation saves time to answer questions & 6 \\
+Lack of time to maintain good documentation & 4 \\
+Documentation of the project conveys information clearly & 3 \\
+Coding is more fun than documentation & 2 \\
+Users help each other by referring to documentation & 1 \\ \hline
+\end{tabular}
+\caption{\label{tab_opinion_doc}Opinions on documentation by the numbers of
+interviewees with the answers}
+\end{table}
+
+For documentation tools and methods mentioned by the interviewees, the most
+popular (mentioned by about 30\% of developers) were forum discussions and
+videos.  The second most popular options (mentioned by about 20\% of developers)
+were GitHub, wiki pages, workshops, and social media.  The least frequently
+mentioned options (about 10\% of developers) included writing books, google
+forms and state management.
+
+\subsection{Project Management} \label{sec_contribution_pm}
+
+To investigate project management we asked three questions:
+
+\begin{description}
+\item[Q5.] Do you have a defined process for accepting new contributions into
+your team?
+\item[Q13.] What is your software development model? For example, waterfall,
+agile, etc.
+\item[Q14.] What is your project management process? Do you
+think improving this process can tackle the current problem? Were any project
+management tools used?
+\end{description}
+
+Although teams may have a process for accepting new contributions, no one
+discussed this during their interview. However, most teams (8 of 9) mentioned
+using GitHub and pull requests to manage contributions from the community. The
+interviewees generally gave very positive feedback on using GitHub. Some teams
+previously used a different approach to version control and eventually
+transferred to git and GitHub.  The past approaches included contributions
+from e-mail (3 teams), contributions from forums (1 team) and sharing the
+git repository via e-mail (1 team).  \wss{I do not know what this last one means.
+Should ask Ao.} \ad{The last one happened when cloud-based repositories (Github,
+Gitlab, Bitbucket, etc.) were not widely used. The project git repository was
+stored locally on the project owner's local computer. The owner shared the whole
+git repo to team members and other individuals who wanted to contribute. The
+contributors made changes to the repo, and sent it back. They transferred the
+repo back and forth by email.}
+
+For managing contributions, the \textit{3D Slicer} team encouraged users to
+develop their extensions for specific use cases, while the \textit{OHIF} team
+enabled the use of plug-ins.  The interviewee from the \textit{ITK-SNAP} team
+said one way of accepting new team members was through funded academic projects.
+
+For the most part, the responses about development model were vague, with only
+two interviewees following a definite development model. In some cases the
+interviewees felt their process was similar to an existing development model.
+Three teams (about 38\%) either followed agile, or something similar to agile.
+Two teams (25\%) either followed a waterfall process, or something similar.
+Three teams (about 38\%) explicitly stated that their process was undefined or
+self-directed.
+
+Some interviewees mentioned the project management tools they used. Generally
+speaking, the interviewees talked about two types of tools:
+\begin{inparaenum}[i)]
+\item trackers, including GitHub, issue trackers, bug trackers and Jira; and,
+\item documentation tools, including GitHub, Wiki page, Google Doc, and Confluence.
+\end{inparaenum}
+Of the specifically named tools in the above lists, GitHub was mentioned 3
+times, while each of the other tools was only mentioned once.
+
+No interviewee introduced any strictly defined project management process. The
+most common approach was following the issues, such as bugs and feature requests.
+Additionally, the \textit{3D Slicer} team had weekly meetings to discuss the
+goals for the project; the \textit{INVESALIUS 3} team relied on the GitHub
+process for their project management; the \textit{ITK-SNAP} team had a fixed
+six-month release pace; only the interviewee from the \textit{OHIF} team
+mentioned that the team has a project manager; the \textit{3D Slicer} team and
+\textit{BioImage Suite Web} team do nightly builds and tests.
+
+Most interviewees skipped the part of the last question asking ``Do you think
+improving this process can tackle the current problem?''. In retrospect, we
+should not have asked a yes-or-no question, since a binary answer is not very
+informative. The interviewee from the \textit{OHIF} team gave a positive answer
+to this question. They believed that a better project management process can
+improve the efficiency of junior developers. They also improved the project
+management tools (from public Jira to public GitHub repository plus private
+Jira), so they could better communicate externally and internally.
+
+\subsection{Discussions on Software Qualities} \label{sec_interview_software_qualities}
+
+Questions 15--18, and 20 cover the software qualities of correctness,
+maintainability, understandability, usability, and reproducibility,
+respectively. We discuss each quality below.
+
+\begin{description}
+\item[Q15.] Was it hard to ensure the correctness of the software? If there were
+any obstacles, what methods have been considered or practiced to improve the
+situation? If practiced, did it work?
+\end{description}
+
+The interviewees identified multiple threats to correctness.  The most
+frequently mentioned threat was complexity.  Complexity enters the software by
+various means, including a variety of data formats, complicated data standards,
+differing outputs between medical imaging machines, and the addition of
+(non-viewing related) functionality.  Other threats to correctness identified by at least one time
+include the following:
+\begin{itemize}
+\item Lack of real world image data for testing;
+\item The team cannot use private data for debugging even when the data causes
+problems; \wss{Why not?  Should ask Ao} \ad{For example, when a physician faces
+some problems to use the software to display a medical image, he/she might reach
+to the team for help. However, the image is private, so the physician can only
+describe the problems, but not share the image with the team to reproduce the
+problems for debugging.}
+\item Expense and time consumed by tests because of the huge datasets;
+\item Difficulty managing releases;
+\item No unit tests; and,
+\item No dedicated quality assurance team.
+\end{itemize}
+
+Testing was the most often mentioned strategy for ensuring correctness.  Seven
+teams mentioned test related activities, including test-driven development,
+component tests, integration tests, smoke tests, regression tests, self tests
+and automated tests.  Another approach frequently adopted (mentioned by 3
+interviewees) is a two state development process with stable releases and
+nightly builds.  Other strategies for ensuring correctness that came up during
+the interviews include CI/CD, using de-identified copies of medical images for
+debugging, sending beta versions to medical workers who can access the data to
+do the tests, and collecting/maintaining a dataset of problematic images.
+
+\begin{description}
+\item[Q16.] When designing the software, did you consider the ease of future
+changes? For example, will it be hard to change the system’s structure, modules,
+or code blocks? What measures have been taken to ensure the ease of future
+changes and maintains?
+\end{description}
+
+The most popular, with five out of nine interviewees mentioning it, strategy for
+ensuring maintainability was to use a modular approach, with often repeated
+functions in a library.  Other strategies that were mentioned for improving
+maintainability include supporting third-party extensions, an easy-to-understand
+architecture, a dedicated architect, starting from simple solutions, and
+documentation.  The \textit{3D Slicer} team used a well-defined structure for
+the software, which they named as an ``event-driven MVC pattern''. Moreover,
+\textit{3D Slicer} discovers and loads necessary modules at runtime, according
+to the configuration and installed extensions. The \textit{BioImage Suite Web}
+team had designed and re-designed their software multiple times in the last 10+
+years. They found that their modular approach effectively supports
+maintainability \citep{Joshi2011}.
+
+\begin{description}
+\item[Q17.] Provide instances where users have misunderstood the software. What,
+if any, actions were taken to address understandability issues?
+\end{description}
+
+The discussion with the developers focused on understandability issues for two
+classes of users: the end users and other developers.  The threats to
+understandability for end users include users not understanding how to use
+features (\wss{Because the users don't have sufficient background?  Ask Ao.})
+\ad{Did you mean medical or medical imaging background? No, it was majorly
+caused by the software usability. Based on the context of our interview
+conversations, it means that users don't understand how to complete certain
+operations or what functions a feature provides. It's due to unclear description,
+lack of documentation, poorly designed UI or nonintuitive process},
+the team not having a dedicated user experience (UX) designer, some important
+indicators are not noticeable (e.g. a progress bar), not all users understand
+the purpose of the software, not all users know if the software includes certain
+features, not all users understand how to use the command line tool, not all
+users understand that the software is a web application. For developers the
+threats to understandability include developers not understanding how to deploy
+the software (\wss{This sounds like a consequence of a lack of
+understandability, rather than threat to it.  Or is the point that developers
+lack sufficient background?  Ask Ao}) \ad{Yes, this should be a consequence of a
+lack of understandability. I think the direct cause is that the project doesn't
+clearly explain how to deploy. The background of the developers wasn't discussed
+at that point.}, and the architecture is difficult for new
+developers to understand.
+
+The most common strategy, cited by 4 developers, for ensuring understandability
+was to use documentation (user manuals, mailing lists, forums).  Other suggested
+and practiced strategies include a graphical user interface, testing every
+release with active users, making simple things simple and complicated things
+possible, icons with clear visual expressions, designing the software to be
+intuitive, having a UX designer with the right experience, dialog windows for
+important notifications, and providing an example for users to follow.
+
+\begin{description}
+\item[Q18.] What, if any, actions were taken to address usability issues?
+\end{description}
+
+The two most frequently mentioned strategies (mentioned 3 times each) for
+improving usability are:
+\begin{inparaenum}[i)]
+    \item usability tests and interviews with end users; and,
+    \item adjusting the software according to user feedback.
+\end{inparaenum}
+Other ideas that came up include a straightforward and intuitively designed
+interface / professional UX designer, providing step-by-step processes, making
+the basic functions easy to use without reading the documentation, focusing on
+limited number of functions, making the software more streamlined, downsampling
+images to consume less memory, and an option to load only part of the data to
+boost performance.
+
+\begin{description}
+\item[Q20.] Do you have any concern that your computational results won’t be
+reproducible in the future? Have you taken any steps to ensure reproducibility?
+\end{description}
+
+We discussed threats to reproducibility and strategies for improving it.  The
+threats that were mentioned include closed-source software, no user interaction
+tests, no unit tests, using different versions of some common libraries
+\wss{because the interface and/or behaviour of the library could have changed?
+Ask Ao} \ad{Yes. Behaviour change could be caused by a lib changing to use
+different algorithms or different dependencies}, variability
+between CPUs, and misinterpretation of how manufacturers
+create medical images. The most commonly cited (by 6 teams) strategy to improve
+reproducibility was testing (regression tests, unit tests, having good tests).
+The second most common strategy (mentioned by 5 teams) is making code, data, and
+documentation available, possibly by creating open-source libraries.  Other
+ideas that were mentioned include running the same tests on all platforms, a
+dockerized version of the software to insulate it from the OS environment, using
+standard libraries, monitoring the upgrades of the library dependencies, clearly
+documenting the version information, bringing along the exact versions of all
+the dependencies with the software, providing checksums of the data, and
+benchmarking the software against other software that overlaps in functionality.
+Specifically one interviewee suggested using \textit{3D Slicer} as the benchmark
+to test their reproducibility.
+
+\section{Answers to Research Questions} \label{ch_answers}
+
+This section answers the research questions from
+Section~\ref{sec_motivation}.  We will go through each of the questions
+in turn.
+
+\begin{description}
+\item[RQ1.] What artifacts are present in current software projects?
+\end{description}
+
+We answered this question by examining and analyzing the repositories for each
+of the 29 projects.  In particular, as part of assessing maintainability, we
+listed the artifacts (documents, test scripts etc.) that appear in each project.
+The full measurement results are summarized in \citet{Dong2021-Data}.  The
+artifacts found were as follows (with the number of occurrences in brackets):
+Source code (29), README (29), Version Control (29), License (28), Bug tracker
+(28), Change request (28), User Manual (22), Release notes (22), Build file
+(18), Tutorials (18), Installation Guide (16), Test cases (15), Authors (14),
+FAQ (14), Acknowledgements (12), Executable files (10), Developer's Manual (8),
+API documentation (7), Troubleshooting guide (6), Project Plan (5).  The
+traceability of some document types to the projects in which they occur was
+previously presented in Table~\ref{tab_maintainability_docs} for maintainability
+related documents, and Table~\ref{tab_Visibility/Transparency_docs} for
+visibility/transparency related documents.
+
+The list in the previous paragraph shows that some types of software engineering
+recommended artifacts are missing in MI projects.  For instance, we did not find
+evidence of user stories (an Agile software development tool for capturing
+software features from an end-user perspective), user requirements, test plans,
+test code coverage, etc.  Our preliminary work suggests that this is not unique
+to MI software.  A gap exists in scientific computing development practices and
+software engineering recommendations \citep{OwojaiyeEtAl2021_CSE}.
+
+\begin{description}
+\item[RQ2.] What tools are used in the development of current software packages?
+\end{description}
+
+We answer this question by measuring the qualities and interviewing the
+developers. We learned that developers use tools for CI/CD
+(Section~\ref{sec_result_correctness_verifiability}), user support
+(Table~\ref{tab_user_support_model}), version control
+(Section~\ref{sec_score_maintainability}), documentation
+(Section~\ref{sec_contribution_pm}), contribution management
+(Section~\ref{sec_contribution_pm}), and project management
+(Section~\ref{sec_contribution_pm}).
+
+As mentioned in Section \ref{sec_result_correctness_verifiability}, we
+identified five projects using CI/CD tools: \textit{3D Slicer} and \textit{OHIF
+Viewer} use CircleCI; \textit{ImageJ}, \textit{Fiji}, and \textit{dwv} use
+Travis. We identified the above projects and tools by examining the
+documentation and source code of all projects. Not all tools leave identifiable
+artifacts in the repositories; therefore, we may have missed the use of some
+tools. According to the interviews with developers, \textit{dwv} and
+\textit{Weasis} used GitHub Actions.
+
+\begin{description}
+\item[RQ3.] What principles, processes, and methodologies are used in the
+development of current software packages?
+\end{description}
+
+We answer this question by measuring the qualities and interviewing the
+developers. We previously summarized principles, processes, and methodologies in
+software testing (Section~\ref{sec_interview_pain_points} and
+Section~\ref{sec_interview_software_qualities}), documentation
+(Section~\ref{sec_interview_documents}, Table~\ref{tab_maintainability_docs},
+Table~\ref{tab_Visibility/Transparency_docs}, and Section~\ref{ch_answers}),
+contribution management (Section~\ref{sec_contribution_pm}), and project
+management (Section~\ref{sec_contribution_pm}).
+
+We identified the use of unit testing in less than half of the 29 projects. On
+the other hand, the interviewees believed that testing (including usability
+tests with users) was the top solution to improve correctness, usability, and
+reproducibility. One pain point in the development process is the lack of access
+to real-world datasets for testing. The developers' strategies to address this
+are summarized in Section~\ref{ch_interview} \ad{to be fixed}. One threat to correctness is:
+with huge datasets for testing, the tests are expensive and time-consuming.
+Three interviewees endorsed self tests / automated tests, which may save time
+for testing.
+
+Documentation is apparently not emphasized in any of the 29 projects.  None of
+them had theory manuals, although we did identify a road map in the \textit{3D
+Slicer} project.  No requirements specifications were found. Eight of the nine
+interviewees thought that documentation was essential to their projects.
+However, they hold the common opinion that their documentation needed
+improvements. Nearly half of them also believed that the lack of time prevented
+them from improving the documentation.
+
+\begin{description}
+\item[RQ4.] What are the pain points for developers working on research software
+projects? What aspects of the existing processes, methodologies, and tools do
+they consider as potentially needing improvement? What changes to processes,
+methodologies, and tools can improve software development and software quality?
+\end{description}
+
+We answer this question by interviewing the developers. The answers to this
+question, including the pain points and the solutions proposed by the
+developers, are in Section~\ref{sec_interview_pain_points}.
+
+\begin{description}
+\item[RQ5.] What is the current status of the following
+software qualities for the projects? What actions have the developers taken to
+address them?
+\end{description}
+
+The current status of the software qualities was assessed through filling in the
+measurement template \citep{Dong2021-Data}.  A summary of the comparison between
+all projects can be found in Section~\ref{ch_results}.  We explored some of the
+qualities more deeply, especially with respect to techniques addressing quality
+concerns, through the interview questions.  Specifically, the qualities of
+correctness \& verifiability, usability, maintainability, understandability and
+reproducibility are explored in Section~\ref{ch_interview}.
+
+\begin{description}
+\item[RQ6.] How does software designated as high quality by this methodology
+compare with top-rated software by the community?
+\end{description}
 
@@
 information, a team can ask the data owner to remove such information or add
-noise to protect privacy. One example is using deidentified copies of medical
+noise to protect privacy. One example is using de-identified copies of medical
 images for testing.
@@
 and engineering, along with theory and experiment \citep{Landau2005}. Almost all
-areas in science and engineering use computers for modeling \citep{Golub2014},
+areas in science and engineering use computers for modelling \citep{Golub2014},
 and software plays an essential role in modern scientific research

### 2022-03-22 StateOfPractice/Papers/MedImageSoft_SOP-arXiv/MedImageSoft_SOP_AcademicRadiology.tex
Updates to Section 2 (Background)

@@
 analytic, and medical applications \citep{FDA2021, enwiki:1034887445,
-Zhang2008}.  An example medical image is provided in Figure~\ref{Fig_Example}.
-Given the importance of MI software and the high number of competing software
-projects, we wish to understand the merits and drawbacks of the current
-development processes, tools and methodologies.  We aim to assess through a
-software engineering lens the quality of the existing software with the hope of
-highlighting standout examples, understanding current pain points and providing
-guidelines and recommendations for future development.
+Zhang2008}.  An example medical image of the brain is provided in
+Figure~\ref{Fig_Example}. Given the importance of MI software and the high
+number of competing software projects, we wish to understand the merits and
+drawbacks of the current development processes, tools and methodologies.  We aim
+to assess through a software engineering lens the quality of the existing
+software with the hope of highlighting standout examples, understanding current
+pain points and providing guidelines and recommendations for future development.
 
@@
     \begin{center}
-        \includegraphics[scale=0.42]{figures/Example_Aorta.png}        
+        \includegraphics[scale=0.25]{figures/MPR.png}        
     \end{center}
-    \caption{Example medical image showing the segmentation of an aorta}
+    \caption{Example brain image showing a multi-planar reformat using Horos (free open-source medical imaging/DICOM viewer for OSX, based on OsiriX)}
     \label{Fig_Example}
@@
 Section~\ref{ch_methods}.  Our methodology has been applied to MI software
-\citet{Dong2021} and Lattice Boltzmann Solvers \citet{Michalski2021}.  This
+\citep{Dong2021} and Lattice Boltzmann Solvers \citep{Michalski2021}.  This
 methodology builds off prior work to assess the state of the practice for such
@@
 
-In this section, we introduce the relevant software categories (Section
-\ref{sec_software_categories}). The categories will be used when collecting and
-assessing the software packages.  We also review the definitions of software
-qualities (Section \ref{sec_software_quality}) and provide an overview of 
-AHP (Section \ref{sec_AHP}).
+To measure the existing MI software we need two sets of definitions: i) the
+definitions of relevant software license models (Section
+\ref{sec_software_categories}); and, ii) the definitions of the software
+qualities that we will be assessing (Section \ref{sec_software_quality}). In our
+assessment we rank the software packages for each quality; therefore, this
+section also provides the background on our ranking process -- the Analytic
+Hierarchy Process (Section \ref{sec_AHP}).
 
@@
 be available to us or not.  In this section, we present three common software
-categories:
+categories.  We will only assess software that fits under the Open Source
+Software license.
 
@@
 \item \textbf{Freeware} Freeware is software that can be used free of charge.
-Unlike with OSS, the authors of do not allow users to access or modify the
-source code \citep{LINFO2006}. To many end-users, the differences between
-freeware and OSS may not be relevant. However, software developers, end-users
-who wish to modify the source code, and researchers looking for insight into
-software development process may find the inaccessible source code a problem.
+Unlike OSS, the authors of do not allow access or modify the source code
+\citep{LINFO2006}. To many end-users, the differences between freeware and OSS
+may not be relevant. However, software developers who wish to modify the source
+code, and researchers looking for insight into software development process may
+find the inaccessible source code a problem.
 
@@
 set of measures.  That is, quality is a collection of different qualities, often
-called ``ilities.''  Below we list the 10 qualities of interest for this study,
-as summarized in \citet{SmithEtAl2021-QDef}.  The order of the qualities follows
-the order used in \cite{GhezziEtAl2003}, which puts related qualities (like
-correctness and reliability) together.  Moreover, the order is roughly the same
-as the order qualities are considered in practice.  \wss{To be completed - not
-currently in sync with QDef paper, QDef paper not completely done.  Look at LBM
-paper too.}
+called ``ilities.''  Below we list the 10 qualities of interest for this study.
+The order of the qualities follows the order used in \citet{GhezziEtAl2003},
+which puts related qualities (like correctness and reliability) together.
+Moreover, the order is roughly the same as the order qualities are considered in
+practice.
 
@@
 	\item \textbf{Installability} The effort required for the installation
-    and/or uninstallation of software in a specified environment.
+    and/or uninstallation of software in a specified environment
+    \citep{ISO/IEC25010, lenhard2013measuring}.
 
 	\item \textbf{Correctness \& Verifiability} A program is correct if it
-    matches its specification.  The specification can either be explicitly or
-    implicitly stated.  Verifiability is the extent to which the software
-    components or the integrated product can be evaluated to demonstrate whether
-    the system functions as expected.
+    matches its specification \citep[p.\ 17]{GhezziEtAl2003}.  The specification
+    can either be explicitly or implicitly stated.  The related quality of
+    verifiability is the ease with which the software components or the
+    integrated product can be checked to demonstrate its correctness.
 
 	\item \textbf{Reliability} The probability of failure-free operation of a
-	computer program in a specified environment for a specified time, i.e., the
-	average time interval between two failures also known as the mean time to
-	failure (MTTF) \citep{musa1987software} \citep{GhezziEtAl2003}.
+	computer program in a specified environment for a specified time \citep{musa1987software}, \citep[p.\ 357]{GhezziEtAl2003}.
 
@@
 	and ii) when the assumptions in its requirements specification are violated
-	\citep{ghezzi1991fundamentals} \citep{boehm2007software}.
+	\citep[p.\ 19]{GhezziEtAl2003}, \citep{boehm2007software}.
 
@@
 	other attributes; iii) satisfy new requirements
-	\citep{IEEEStdGlossarySET1990} \citep{boehm2007software}.
+	\citep{IEEEStdGlossarySET1990}, \citep{boehm2007software}.
 
@@
 	of a software development process and the current status of it are conveyed
-	clearly \citep{ghezzi1991fundamentals}.
+	clearly \citep[p.\ 32]{GhezziEtAl2003}.
 
@@
 
-\subsection{Analytic Hierarchy Process} \label{sec_AHP}
-
-AHP is used to generate a ranking for a set of software packages. Thomas L.\
-Saaty developed AHP in the 1970s, and people have widely used it since to make
-and analyze multiple criteria decisions \citep{VaidyaEtAl2006}. AHP organizes
-multiple criteria factors in a hierarchical structure and uses pairwise
-comparisons between alternatives to calculate relative ratios \citep{Saaty1990}.
-
-For a project with $m$ criteria, we can use an $m\times m$ matrix $A$ to record
+\subsection{Analytic Hierarchy Process (AHP)} \label{sec_AHP}
+
+Thomas L.\ Saaty developed AHP in the 1970s, and people have widely used it
+since to make and analyze multiple criteria decisions \citep{VaidyaEtAl2006}.
+AHP organizes multiple criteria factors in a hierarchical structure and uses
+pairwise comparisons between alternatives to calculate relative ratios
+\citep{Saaty1990}. We use AHP to generate a ranking for a set of software
+packages.
+
+For a project with $m$ criteria, we can use an $m \times m$ matrix $A$ to record
 the relative importance between factors. When comparing criterion $i$ and
 criterion $j$, the value of $A_{ij}$ is decided as follows, with the value of
-$A_{ji}$ generally as $1/A_{ij}$ \citep{Saaty1990}.  $A_{ij} = 1$ if criterion
-$i$ and criterion $j$ are equally important, while $A_{ij} = 9$ if criterion $i$
-is extremely more important than criterion $j$.  The natural numbers between 1
-and 9 are used to show the different levels of relative importance between these
-two extremes. The above assumes that criterion $i$ is not less important than
-criterion $j$.  If that is not the case, we reverse $i$ and $j$ and determine
-$A_{ji}$ first, then $A_{ij} = 1/A_{ji}$.
+$A_{ji}$ generally equal to $1/A_{ij}$ \citep{Saaty1990}: $A_{ij} = 1$ if
+criterion $i$ and criterion $j$ are equally important, while $A_{ij} = 9$ if
+criterion $i$ is extremely more important than criterion $j$.  The natural
+numbers between 1 and 9 are used to show the different levels of relative
+importance between these two extremes. The above assumes that criterion $i$ is
+not less important than criterion $j$.  If that is not the case, we reverse $i$
+and $j$ and determine $A_{ji}$ first, then $A_{ij} = 1/A_{ji}$.
 
@@
 \begin{equation} 
-    A w = \lambda_{max} w,
+    A w = \lambda_{\text{max}} w,
 \end{equation}
-where $\lambda_{max}$ is the maximal eigenvalue of $A$.  In this project, $w$ is
-approximated with the classic \textit{mean of normalized values} approach
+where $\lambda_{\text{max}}$ is the maximal eigenvalue of $A$.  In this project,
+$w$ is approximated with the classic \textit{mean of normalized values} approach
 \citep{AlessioEtAl2006}:
@@
 an $n\times n$ matrix $B_i$ to record the relative preferences between these
-choices for each criterion. The way of generating $B_i$ is similar to the one
-for $A$. However, rather than comparing the importance between criteria, we
-pairwise decide how much we favour one alternative over the other. We use the
+choices for each of the $m$ criterion. The way of generating $B_i$ is similar to
+the one for $A$. However, rather than comparing the importance between criteria,
+we pairwise decide how much we favour one alternative over the other. We use the
 same method to calculate the local priority vector for each $B_i$.  The local
@@
 \ref{sec_software_quality} are the criteria ($m = 9$), while 29 software
-packages ($n = 29$) are compared for each of the $m$ criteria. The software are
+packages ($n = 29$) are compared for each of the $m$ criteria. The packages are
 evaluated with the grading template in Section~\ref{sec_grading_software}, which
@@
 
-We developed a methodology for evaluating the state of the practice of SC
+We developed a methodology for evaluating the state of the practice of research
 software \citep{SmithEtAl2021}.  The  methodology can be instantiated for a
@@
 domain expert partner throughout, as discussed in
-Section~\ref{sec_vet_software_list}.  The four main steps of the methodology are:
+Section~\ref{sec_vet_software_list}.  The four main steps of the methodology
+are:
 

### 2022-03-23 StateOfPractice/Papers/MedImageSoft_SOP-arXiv/MedImageSoft_SOP_AcademicRadiology.tex
Revisions to SOP for MI paper, mostly in Methodology section

@@
 (Section~\ref{sec_software_selection});
-\item Measure (or grade) the selected software (Section~\ref{sec_grading_software});
-\item Interview developers (Section~\ref{sec_interview_methods}).
-\item Answer the research questions (as given in Section~\ref{sec_motivation})
+\item Measure (or grade) the selected software
+(Section~\ref{sec_grading_software});
+\item Interview developers (Section~\ref{sec_interview_methods});
+\item Answer the research questions (as given in Section~\ref{sec_motivation}).
 \end{enumerate}
@@
 Biomedical Engineering, and Director of Medical Imaging Physics and Engineering
-at St. Joseph's Healthcare, Hamilton, Ontario, Canada.
+at St.\ Joseph's Healthcare, Hamilton, Ontario, Canada.
 
@@
 48 packages is available in \citet{Dong2021}.  To reduce the length of the list
-to a manageable number (29 in this case), we filtered the original list as
-follows:
+to a manageable number (29 in this case, as given in Section~\ref{ch_results}),
+we filtered the original list as follows:
 
@@
 \item We removed \textit{Open Dicom Viewer}, since it has not received any
-updates for a long time (since 2011).
+updates in a long time (since 2011).
 
@@
 We grade the selected software using the measurement template summarized
-in~\citet{SmithEtAl2021}.  The template summarizes measures of the qualities
+in~\citet{SmithEtAl2021}.  The template provides measures of the qualities
 listed in Section~\ref{sec_software_quality}, except for reproducibility, which
@@
 Project developers can be contacted for help regarding installation, if
-necessary, but a cap of about 2 hours is imposed on the
-installation process, to keep the overall measurement time feasible. An excerpt
-of the spreadsheet is shown in Figure~\ref{fg_grading_template_example}.
-\wss{The Figure should be updated to be easier to read.}  A column is included
-to for each measured software package.
+necessary, but a cap of about 2 hours is imposed on the installation process, to
+keep the overall measurement time feasible. An excerpt of the spreadsheet is
+shown in Figure~\ref{fg_grading_template_example}.  A column is included for
+each measured software package.
 
 \begin{figure}[!ht]
-\includegraphics[scale=0.42]{figures/template.png}
+\includegraphics[scale=0.67]{figures/template.pdf}
 \caption{Grading template example}

### 2022-03-23 StateOfPractice/Papers/MedImageSoft_SOP-arXiv/MedImageSoft_SOP_AcademicRadiology.tex
Complete revisions to Section 3 (Methodology)

@@
 
-We grade the selected software using the measurement template summarized
-in~\citet{SmithEtAl2021}.  The template provides measures of the qualities
-listed in Section~\ref{sec_software_quality}, except for reproducibility, which
-is assessed through the developer interviews
-(Section~\ref{sec_interview_methods}). For each software package, we fill-in the
-template questions. This process can take between 1 to 4 hours for each package.
-Project developers can be contacted for help regarding installation, if
-necessary, but a cap of about 2 hours is imposed on the installation process, to
-keep the overall measurement time feasible. An excerpt of the spreadsheet is
-shown in Figure~\ref{fg_grading_template_example}.  A column is included for
-each measured software package.
+We grade the selected software using the measurement template summarized in
+\citet{SmithEtAl2021}.  The template provides measures of the qualities listed
+in Section~\ref{sec_software_quality}, except for reproducibility, which is
+assessed through the developer interviews (Section~\ref{sec_interview_methods}).
+For each software package, we fill-in the template questions. To stay within the
+target of 160 person hours to measure the domain, we allocated between 1 to 4
+hours for each package. Project developers can be contacted for help regarding
+installation, if necessary, but a cap of about 2 hours is imposed on the
+installation process, to keep the overall measurement time feasible. An excerpt
+of the spreadsheet is shown in Figure~\ref{fg_grading_template_example}.  A
+column is included for each measured software package.
 
@@
 started manual, a user manual and documentation of expected user
-characteristics.  Below is a summary of how each quality is measured.
+characteristics.  Below is a summary of how we assess adoption of best practices
+by measuring each quality.
 
@@
 interaction with the software during testing.  In general, an easy-to-use
-graphical user interface scores higher.
+graphical user interface will score high.
 
@@
 artifacts as API documentation, bug tracker information, release notes, test
-cases, and build files. We also check for the use of tools supporting issue
+cases, and build scripts. We also check for the use of tools supporting issue
 tracking and version control, the percentages of closed issues, and the
@@
 choices for reuse. Furthermore, well-modularized code, which tends to have
-smaller parts in separate files, is typically easier to reuse. Thus, we consider
-the projects with more code files and less Lines of Code (LOC) per file to be
+smaller parts in separate files, is typically easier to reuse. Thus, we assume
+that projects with more code files and less Lines of Code (LOC) per file are
 more reusable. We also consider projects with API documentation as delivering
@@
 cannot look at all code files for each project; therefore, we randomly examine
-10 code files for their understandability. We check the code’s style within each
+10 code files for their understandability. We check the code's style within each
 file, such as whether the identifiers, parameters, indentation, and formatting
@@
 For projects on GitHub we manually collect additional information, such as the
-numbers of stars, forks, people watching this repository, and open pull requests
-on GitHub. A git repository can have a creation date much earlier than the first
-pull, closed pull requests, and the number of months a repository has been on
-GitHub. For example, the developers created the git repository for \textit{3D
-Slicer} in 2002, but did not upload a copy of it to GitHub until 2020. Some
-GitHub data can be found using its the GitHub Application Program Interface
-(API) via the following url:
+numbers of stars, forks, people watching this repository, open pull requests,
+closed pull requests, and the number of months a repository has been on GitHub.
+We need to take care with the project creation date, since a repository can have
+a creation date much earlier than the first day on GitHub.  For example, the
+developers created the git repository for \textit{3D Slicer} in 2002, but did
+not upload a copy of it to GitHub until 2020. Some GitHub data can be found
+using its the GitHub Application Program Interface (API) via the following url:
 \textit{https://api.github.com/repos/[owner]/[repository]} where [owner] and
@@
 12 months shows how active a project has been, and the number of stars and forks
-may reveal its popularity. Secondly, the results may affect our decisions
-regarding the grading scores for some software qualities. For example, if the
-percentage of comment lines is low, we double-check the understandability of the
-code; if the ratio of open versus closed pull requests is high, we pay more
-attention to maintainability.
+may reveal its popularity (used to assess \rqref{RQ_CompareHQ2Popular}).
+Secondly, the results may affect our decisions regarding the grading scores for
+some software qualities. For example, if the percentage of comment lines is low,
+we double-check the understandability of the code; if the ratio of open versus
+closed pull requests is high, we pay more attention to maintainability.
 
@@
 filling in the measurement template spreadsheet, the details for each VM
-should be noted, including hypervisor and operating system version.
+are noted, including hypervisor and operating system version.
 
 When grading the software, we found 27 out of the 29 packages are compatible
-with two or three different OS such as Windows, macOS, and Linux, and 5 of them
-are browser-based, making them platform-independent. However, in the interest of
-time, we only performed the measurements for each project by installing it on
-one of the platforms.  When it was an option, we selected Windows as the host
-OS.
+with two or three different OSes, such as Windows, macOS, and Linux, and 5 of
+them are browser-based, making them platform-independent. However, in the
+interest of time, we only performed the measurements for each project by
+installing it on one of the platforms.  When it was an option, we selected
+Windows as the host OS.
 
@@
 
-We designed a list of 20 questions to guide our interviews, which can be found
-in \citet{SmithEtAl2021}. Some questions are about the background of
-the software, the development teams, the interviewees, and how they organize the
-projects. We also ask about the developer's understandings of the users. Some
-questions focus on the current and past difficulties, and the solutions the team
-has found, or will try. We also discuss the importance and current situations of
+Our interviews were guided by a list of 20 questions, which can be found in
+\citet{SmithEtAl2021}. Some questions are about the background of the software,
+the development teams, the interviewees, and how they organize the projects. We
+also ask about the developer's understandings of the users. Some questions focus
+on the current and past difficulties, and the solutions the team has found, or
+will try. We also discuss the importance and current situations of
 documentation. A few questions are about specific software qualities, such as
@@
 questions when necessary. Based on our experience, the interviewees usually
-bring up some exciting ideas that we did not expect, and it is worth expanding
-on these topics.
+brought up unexpected and exciting ideas.
 
@@
 interviewed. We found contact information from projects websites, code
-repository, publications, and from biographic pages at the teams’ institutions.
-Meetings were held on-line using either Zoom or Teams, which facilitates
+repository, publications, and from biographic pages at the teams's institutions.
+Meetings were held on-line using either Zoom or Teams, which facilitated
 recording and automatic transcription of the meetings.  The interview process
@@
 We spent about 90 minutes for each interview and asked 20 prepared questions
-(summarized in~\citet{SmithEtAl2021}). We also asked following-up questions as
+(summarized in \citet{SmithEtAl2021}). We also asked following-up questions as
 the conversation dictated. One participant was too busy to have an interview, so

### 2022-03-23 StateOfPractice/Papers/MedImageSoft_SOP-arXiv/MedImageSoft_SOP_AcademicRadiology.tex
Updated captions for figures where the order of the entries needed to be clarified.  Closes #115

@@
 project members, since many projects received change requests and code from the
-community.  For the supported OS, 25 packages work on all three OSs: Windows
+community.  With respect to the OS, 25 packages work on all three OSs: Windows
 (W), macOS (M), and Linux (L). Although the usual approach to cross-platform
@@
 \end{tabular}
-\caption{\label{tab_final_list}Final software list}
+\caption{Final software list (sorted in descending order of the number of Lines
+Of Code (LOC))}
+\label{tab_final_list}
 \end{table}
@@
 \centering
-\includegraphics[scale=0.7]{figures/PrimaryLanguages.pdf}
+\includegraphics[scale=0.5]{figures/PrimaryLanguages.pdf}
 \centering
@@
 \end{tabular}
-\caption{\label{tab_loc_per_file}Number of files and lines}
+\caption{Number of files and lines (entries are sorted in descending order of
+their reusability scores)}
+\label{tab_loc_per_file}
 \end{table}
@@
 \end{tabular}
-\caption{\label{tab_ranking_vs_GitHub}Software ranking versus GitHub metrics}
+\caption{\label{tab_ranking_vs_GitHub}Software ranking versus GitHub metrics
+(Sorted in descending order of GitHub popularity, as estimated by the number of
+new stars per year)}
 \end{table}

### 2022-03-24 StateOfPractice/Papers/MedImageSoft_SOP-arXiv/MedImageSoft_SOP_AcademicRadiology.tex
Updates to SOP for MI paper, especially updated figures on AHP results

@@
 \usepackage{paralist}
+\usepackage{booktabs}
 
@@
 \begin{tabular}{p{6cm}lllllllll}
-\hline
+\toprule
 \multirow{2}{*}{Software} & \multirow{2}{*}{Rlsd} & \multirow{2}{*}{Updated} & \multirow{2}{*}{Fnd} & \multirow{2}{*}{NOC} & \multirow{2}{*}{LOC} & \multicolumn{3}{c}{OS} & \multirow{2}{*}{Web} \\ \cline{7-9}
- &  &  &  &  &  & W & M & L &  \\ \hline
+ &  &  &  &  &  & W & M & L &  \\ \midrule
 ParaView \citep{Ahrens2005} & 2002 & 2020-10 & X & 100 & 886326 & X & X & X & X \\
@@
 MRIcroGL \citep{Rorden2021} & 2015 & 2020-08 &  & 2 & 8493 & X & X & X &  \\
-DicomBrowser \citep{Archie2012} & 2012 & 2020-08 &  & 3 & 5505 & X & X & X &  \\ \hline
+DicomBrowser \citep{Archie2012} & 2012 & 2020-08 &  & 3 & 5505 & X & X & X &  \\ \bottomrule
 \end{tabular}
@@
 since they are web applications. Installing 10 of the projects required extra
-dependencies. Five of them are the web applications (as shown in
-Table~\ref{tab_final_list}) and depended on a browser; \textit{dwv},
-\textit{OHIF Viewer}, and \textit{GATE} needed extra dependencies to build;
-\textit{ImageJ} and	\textit{Fiji} needed an unzip tool; \textit{MatrixUser} was
-based on Matlab; \textit{DICOM Viewer} needed to work on a Nextcloud platform.
+dependencies. Five of them are web applications (as shown in
+Table~\ref{tab_final_list}) and depend on a browser; \textit{dwv}, \textit{OHIF
+Viewer}, and \textit{GATE} needs extra dependencies to build; \textit{ImageJ}
+and	\textit{Fiji} need an unzip tool; \textit{MatrixUser} is based on Matlab;
+\textit{DICOM Viewer} needs to work on a Nextcloud platform.
 
 \begin{figure}[!ht]
-\includegraphics[scale=0.38]{figures/installability_scores.png}
+\includegraphics[scale=0.48]{figures/installability_scores.pdf}
 \caption{AHP installability scores}
@@
 no trouble installing them, such as \textit{INVESALIUS 3}, \textit{Gwyddion},
-\textit{XMedCon}, and \textit{MicroView}. We gave them various scores based on
+\textit{XMedCon}, and \textit{MicroView}. We determined their scores based on
 the understandability of the instructions, installation steps, and user
 experience. Since \textit{BioImage Suite Web} and \textit{Slice:Drop} needed no
-installation, we gave them higher scores. \textit{BioImage Suite Web} also
+installation, we gave them high scores. \textit{BioImage Suite Web} also
 provided an option to download cache for offline usage, which was easy
@@
 acquire a license.  Of course, for users that already work within Matlab, the
-installability score would be higher.
+installability score should be higher.
 
@@
 packages with higher scores adopted more techniques to improve correctness, and
-had better documentation for us to verify this.  For instance, we looked for
+had better documentation for us to verify against.  For instance, we looked for
 evidence of unit testing, since it benefits most parts of the software's life
@@
 \begin{figure}[!ht]
-\includegraphics[scale=0.38]{figures/correctness_verifiability_scores.png}
+\includegraphics[scale=0.48]{figures/correctness_verifiability_scores.pdf}
 \caption{AHP correctness \& verifiability scores}
@@
 Slicer}, which contained design requirements for upcoming changes.
-
-% In this section, the information about CI/CD tools is part of the answers to
-% \hyperlink{rq2}{RQ2}, and the information about software testing and
-% documentation is part of the answers to \hyperlink{rq3}{RQ3}.
 
@@
 \begin{figure}[!ht]
-\includegraphics[scale=0.38]{figures/reliability_scores.png}
+\includegraphics[scale=0.48]{figures/reliability_scores.pdf}
 \caption{AHP surface reliability scores}
@@
 \begin{figure}[!ht]
-\includegraphics[scale=0.38]{figures/robustness_scores.png}
+\includegraphics[scale=0.48]{figures/robustness_scores.pdf}
 \caption{AHP surface robustness scores}
@@
 software packages should support the DICOM standard. To test robustness, we
-prepared two types of image files: correct format and broken format (created by
-relabelled a text file to have the ``.dcm'' extension).  All software packages
-loaded the correct format image, except for \textit{GATE}, which failed for
-unknown reasons.  For the broken format, \textit{MatrixUser}, \textit{dwv}, and
-\textit{Slice:Drop} ignored the incorrect format of the file and loaded it
-regardless. They did not show any error message and displayed a blank image.
-\textit{MRIcroGL} behaved similarly except that it showed a meaningless image
-with noise pixels. \textit{Drishti} successfully detected the broken format of
-the file, but the software crashed as a result.
+prepared two types of image files: correct and incorrect formats (with the
+incorrect format created by relabelled a text file to have the ``.dcm''
+extension).  All software packages loaded the correct format image, except for
+\textit{GATE}, which failed for unknown reasons.  For the broken format,
+\textit{MatrixUser}, \textit{dwv}, and \textit{Slice:Drop} ignored the incorrect
+format of the file and loaded it regardless. They did not show any error message
+and displayed a blank image. \textit{MRIcroGL} behaved similarly except that it
+showed a meaningless image. \textit{Drishti} successfully detected the broken
+format of the file, but the software crashed as a result.
 
@@
 \begin{figure}[!ht]
-\includegraphics[scale=0.38]{figures/usability_scores.png}
+\includegraphics[scale=0.48]{figures/usability_scores.pdf}
 \caption{AHP surface usability scores}
@@
 Table~\ref{tab_user_support_model} summarizes the user support models by the
-number of projects using them. We do not know whether the prevalent use of
-GitHub issues for user support is by design, or whether this just naturally
-happened as users sought help.
-%  This table contains part of the answers to \hyperlink{rq2}{RQ2}.
+number of projects using them (projects may use more than one support model). We
+do not know whether the prevalent use of GitHub issues for user support is by
+design, or whether this just naturally happened as users sought help.
 
@@
 \centering
-\begin{tabular}{ll}
-\hline
-\multicolumn{1}{c}{User support model} & Number of projects \\ \hline
+\begin{tabular}{lc}
+\toprule
+\multicolumn{1}{c}{User support model} & Num.\ projects \\
+\midrule
 GitHub issue & 24 \\
@@
 Troubleshooting & 2 \\
-Contact form & 1 \\ \hline
+Contact form & 1 \\ 
+\bottomrule
 \end{tabular}
@@
 Figure~\ref{fg_maintainability_scores} shows the ranking results for
-maintainability. We marked \textit{3D Slicer} with a higher score because
+maintainability. We marked \textit{3D Slicer} with the highest score because
 we found it to have the most comprehensive artifacts. For example, as far as we
@@
 maintainability scores.
-% This table contains part of the answers to \hyperlink{rq1}{RQ1} and \hyperlink{rq3}{RQ3}.
 
 \begin{figure}[!ht]
-\includegraphics[scale=0.38]{figures/maintainability_scores.png}
+\includegraphics[scale=0.48]{figures/maintainability_scores.pdf}
 \caption{AHP maintainability scores}
@@
 \centering
-\begin{tabular}{llll}
-\hline
-\multicolumn{1}{c}{Software} & Proj plan & Dev manual & API doc \\ \hline
+\begin{tabular}{lccc}
+\toprule
+\multicolumn{1}{c}{Software} & Proj.\ plan & Dev.\ manual & API doc. \\ 
+\midrule
 3D Slicer & X & X & X \\
@@
 BioImage Suite Web &  & X &  \\
-Gwyddion &  & X & X \\ \hline
+Gwyddion &  & X & X \\ 
+\bottomrule
 \end{tabular}
-\caption{\label{tab_maintainability_docs}Software with the maintainability documents}
+\caption{Software with the maintainability documents (listed in descending order of 
+maintainability score)}
+\label{tab_maintainability_docs}
 \end{table}
@@
 SourceForge. \textit{DicomBrowser} and \textit{3DimViewer} used BitBucket.
-% The information about version control tools is part of the answers to
-% \hyperlink{rq2}{RQ2}.
 
@@
 \begin{figure}[!ht]
-\includegraphics[scale=0.38]{figures/reusability_scores.png}
+\includegraphics[scale=0.48]{figures/reusability_scores.pdf}
 \caption{AHP reusability scores}
@@
 \begin{tabular}{lllll}
-\hline
+\toprule
 \multirow{2}{*}{Software} & \multirow{2}{*}{Text files} & \multirow{2}{*}{Total lines} & \multirow{2}{*}{LOC} & \multirow{2}{*}{LOC/file} \\
- &  &  &  &  \\ \hline
+ &  &  &  &  \\ 
+\midrule
 OHIF Viewer & 1162 & 86306 & 63951 & 55 \\
@@
 DicomBrowser & 54 & 7375 & 5505 & 102 \\
-dicompyler & 48 & 19201 & 15941 & 332 \\ \hline
+dicompyler & 48 & 19201 & 15941 & 332 \\ 
+\bottomrule
 \end{tabular}
@@
 \begin{figure}[!ht]
-\includegraphics[scale=0.38]{figures/understandability_scores.png}
+\includegraphics[scale=0.48]{figures/understandability_scores.pdf}
 \caption{AHP surface understandability scores}
@@
 \begin{figure}[!ht]
-\includegraphics[scale=0.38]{figures/visibility_transparency_scores.png}
+\includegraphics[scale=0.48]{figures/visibility_transparency_scores.pdf}
 \caption{AHP visibility/transparency scores}
@@
 \begin{tabular}{lllll}
-\hline
-Software & Dev process & Proj status & Dev env & Rls notes \\ \hline
+\toprule
+Software & Dev process & Proj status & Dev env & Rls notes \\ 
+\midrule
 3D Slicer & X & X & X & X \\
@@
 dwv &  &  &  & X \\
-Gwyddion &  &  &  & X \\ \hline
+Gwyddion &  &  &  & X \\ 
+\bottomrule
 \end{tabular}
@@
 \begin{figure}[!ht]
-\includegraphics[scale=0.38]{figures/overall_scores.png}
+\includegraphics[scale=0.48]{figures/overall_scores.pdf}
 \caption{Overall AHP scores for all 9 software qualities}
@@
 \begin{tabular}{lllll}
-\hline
-Software & Our ranking & Stars/yr & Watches/yr & Forks/yr \\ \hline
+\toprule
+Software & Our ranking & Stars/yr & Watches/yr & Forks/yr \\ 
+\midrule
 3D Slicer & 1 & 284 & 19 & 128 \\
@@
 AMIDE & 22 & n/a & n/a & n/a \\
-3DimViewer & 24 & n/a & n/a & n/a \\ \hline
+3DimViewer & 24 & n/a & n/a & n/a \\ 
+\bottomrule
 \end{tabular}
@@
 \begin{tabular}{llll}
-\hline
+\toprule
 \multirow{2}{*}{Category} & \multirow{2}{*}{Obstacle} & \multicolumn{2}{c}{Num ans.} \\ \cline{3-4} 
- &  & current & past \\ \hline
+ &  & current & past \\ 
+\midrule
 \multirow{2}{*}{Resource} & Lack of fundings & 3 &  \\
- & Lack of time to devote to the project & 2 & 1 \\ \hdashline
+ & Lack of time to devote to the project & 2 & 1 \\ 
+\midrule
 \multirow{3}{*}{Balance} & Hard to keep up with changes in OS and libraries & 1 &  \\
  & Hard to support multiple OS & 2 &  \\
- & Hard to support lower-end computers & 1 & 2 \\ \hdashline
-Testing & Lack of access to real-world datasets for testing & 3 & 2 \\ \hdashline
+ & Hard to support lower-end computers & 1 & 2 \\ 
+\midrule
+Testing & Lack of access to real-world datasets for testing & 3 & 2 \\ 
+\midrule
 \multirow{7}{*}{Others}
@@
  & Hard to understand users' needs & & 1 \\
- & Hard to maintain good documentations & & 1 \\ \hline
+ & Hard to maintain good documentations & & 1 \\ 
+\bottomrule
 \end{tabular}
@@
 \begin{tabular}{lll}
-\hline
-Software team & Native application & Web application \\ \hline
+\toprule
+Software team & Native application & Web application \\ 
+\midrule
 3D Slicer & X & \\
@@
 Weasis & X & \\
-OHIF & & X \\ \hdashline
+OHIF & & X \\ 
+\midrule
 Total number among the eight teams & 5 & 3 \\
-Total number among the 29 teams & 24 & 5 \\ \hline
+Total number among the 29 teams & 24 & 5 \\ 
+\bottomrule
 \end{tabular}
-\caption{\label{tab_native_vs_web}Teams' choices between native application and
-web application}
+\caption{Teams' choices between native application and web application}
+\label{tab_native_vs_web}
 \end{table}
@@
 \begin{tabular}{ll}
-\hline
-Opinion on documentation & Num ans. \\ \hline
-Documentation is vital to the project& 8 \\
+\toprule
+Opinion on documentation & Num ans. \\ 
+\midrule
+Documentation is vital to the project & 8 \\
 Documentation of the project needs improvements & 7 \\
@@
 Coding is more fun than documentation & 2 \\
-Users help each other by referring to documentation & 1 \\ \hline
+Users help each other by referring to documentation & 1 \\ 
+\bottomrule
 \end{tabular}

### 2022-03-24 StateOfPractice/Papers/MedImageSoft_SOP-arXiv/MedImageSoft_SOP_AcademicRadiology.tex
Completed edits of Section 4 (Measurement results)

@@
 
-Figure 4.7 shows the AHP results for reusability. As described in
-Section~\ref{sec_grading_software}, we gave higher scores to the projects with
-API documentation. As shown in Table \ref{tab_maintainability_docs}, seven
-projects had API documents. We also considered projects with more code files and
-less LOC per code file as more reusable. Table \ref{tab_loc_per_file} shows the
-number of text-based files by projects, which we used to approximate the number
-of code files. The table also lists the total number of lines (including
-comments and blanks), LOC, and average LOC per file. We arranged the items in
-descending order of their reusability scores.
+Figure~\ref{fg_reusability_scores} shows the AHP results for reusability. As
+described in Section~\ref{sec_grading_software}, we gave higher scores to the
+projects with API documentation. As shown in
+Table~\ref{tab_maintainability_docs}, seven projects had API documents. We also
+assumed that projects with more code files and less LOC per code file as more
+reusable. Table \ref{tab_loc_per_file} shows the number of text-based files by
+projects, which we used to approximate the number of code files. The table also
+lists the total number of lines (including comments and blanks), LOC, and
+average LOC per file. We arranged the items in descending order of their
+reusability scores.
 
@@
 Figure~\ref{fg_surface_understandability_scores} shows the scores for surface
-understandability. All projects had a consistent code style with parameters in
+understandability. All projects had a consistent coding style with parameters in
 the same order for all functions; the code was modularized; the comments were
@@
 identification of a coding standard for 3 out of the 29: \textit{3D Slicer},
-\textit{Weasis}, and \textit{ImageJ}. We also found hard-coded constants in
-\textit{medInria}, \textit{dicompyler}, \textit{MicroView}, and \textit{Papaya}.
-We did not find any reference to the used algorithms in projects
-\textit{XMedCon}, \textit{DicomBrowser}, \textit{3DimViewer}, \textit{BioImage
-Suite Web}, \textit{Slice:Drop}, \textit{MatrixUser}, \textit{DICOM Viewer},
-\textit{dicompyler}, and \textit{Papaya}.
+\textit{Weasis}, and \textit{ImageJ}. We also found hard-coded constants (rather
+than symbolic constants) in \textit{medInria}, \textit{dicompyler},
+\textit{MicroView}, and \textit{Papaya}. We did not find any reference to the
+algorithms used in projects \textit{XMedCon}, \textit{DicomBrowser},
+\textit{3DimViewer}, \textit{BioImage Suite Web}, \textit{Slice:Drop},
+\textit{MatrixUser}, \textit{DICOM Viewer}, \textit{dicompyler}, and
+\textit{Papaya}.
 
@@
 
-Figure \ref{fg_visibility_transparency_scores} shows the AHP scores for
-visibility/transparency. Generally speaking, the teams that actively
-documented their development process and plans scored higher. Table
-\ref{tab_Visibility/Transparency_docs} shows the projects that had documents for
-the development process, project status, development environment, and release
-notes, in the descending order of their visibility/transparency scores.
-%This table contains part of the answers to \hyperlink{rq1}{RQ1} and \hyperlink{rq3}{RQ3}.
+Figure~\ref{fg_visibility_transparency_scores} shows the AHP scores for
+visibility/transparency. Generally speaking, the teams that actively documented
+their development process and plans scored higher.
+Table~\ref{tab_Visibility/Transparency_docs} shows the projects that had
+documents for the development process, project status, development environment,
+and release notes, in descending order of their visibility/transparency
+scores.
 
@@
 \end{tabular}
-\caption{\label{tab_Visibility/Transparency_docs}Software with the visibility/transparency documents}
+\caption{Software with the visibility/transparency documents (software is listed
+in descending order of visibility/transparency score)}
+\label{tab_Visibility/Transparency_docs}
 \end{table}
@@
 a specific real world context, we assumed all nine qualities are equally
-important. Figure \ref{fg_overall_scores} shows the overall scores in descending
+important. Figure~\ref{fg_overall_scores} shows the overall scores in descending
 order. Since we produced the scores from the AHP process, the total sum of the
@@
 \includegraphics[scale=0.48]{figures/overall_scores.pdf}
-\caption{Overall AHP scores for all 9 software qualities}
+\caption{Overall AHP scores with an equal weighting for all 9 software qualities}
 

### 2022-03-24 StateOfPractice/Papers/MedImageSoft_SOP-arXiv/MedImageSoft_SOP_AcademicRadiology.tex
Edits completed for Section 5 on Comparison to Community Ranking

@@
 
-To address \rqref{RQ_CompareHQ2Popular}, Table~\ref{tab_ranking_vs_GitHub}
-compares our LBM software package rankings against their popularity in the
-research community, as estimated by repository stars and watches. We answer this
-question by grading the qualities of the software, collecting ratings from the
-MI community, then conducting two comparisons:
-\begin{itemize}
-\item comparing our ranking with the community ratings on GitHub, such as GitHub
-stars, number of forks, and number of people watching the projects (Section~\ref{sec_grading_software});
-\item comparing top-rated software designated by our methodology with the ones
-recommended by our domain experts (Section \ref{sec_software_selection}).
+To address~\rqref{RQ_CompareHQ2Popular} about how our ranking compares to the
+popularity of projects as judged by the scientific community, we make two
+comparisons:
+\begin{itemize}
+\item A comparison of our ranking (from Section~\ref{ch_results}) with the
+community ratings on GitHub, as shown by GitHub stars, number of forks, and
+number of people watching the projects; and,
+\item A comparison of top-rated software from our methodology with the top
+recommendations from our domain experts (as mentioned in
+Section~\ref{sec_software_selection}).
 \end{itemize}
@@
 watching, and forks per 12 months. The method of getting the creation date is
-described in Section~\ref{sec_grading_software}.  We obtained these
-measures in July, 2021.
-
-In Table~\ref{tab_ranking_vs_GitHub}, we used the average number of new stars
-per 12 months as an indicator of the GitHub popularity, and listed the items in
-the descending order of this number. We ordered the non-GitHub items by our
-ranking. Generally speaking, most of the top-ranking MI software projects also
-received greater attention and popularity on GitHub. Between our ranking and the
-GitHub stars-per-year ranking, four of the top five software projects are the
-same.
-
-Project \textit{dwv} was popular on GitHub, but we ranked it low. As mentioned
-in Section~\ref{sec_result_installability}, we failed to build it locally, and
-used the test version on its websites for the measurements. We followed the
-instructions and tried to run the command ``yarn run test'' locally, which did
-not work. In addition, the test version did not detect a broken DICOM file and
-displayed a blank image as described in Section \ref{sec_result_robustness}. We
-might underestimate the scores for \textit{dwv} due to uncommon technical
-issues. We also ranked \textit{DICOM Viewer} much lower than its popularity. As
-mentioned in Section \ref{sec_result_installability} it depended on the
-NextCloud platform that we could not successfully install. Thus, we might
-underestimate the scores of its surface reliability and surface
-robustness. We weighted all qualities equally, which is not likely to be the
-case with all users. As a result, some projects with high popularity may not
-perform well in all qualities.
+described in Section~\ref{sec_grading_software}.  The items in
+Table~\ref{tab_ranking_vs_GitHub} are listed in descending order of the average
+number of new stars per year.  The non-GitHub items are listed in the order of
+our ranking.  All GitHub statistics were collected in July, 2021.
+
+Generally speaking, most of the top-ranking MI software projects also received
+greater attention and popularity on GitHub. Between our ranking and the GitHub
+stars-per-year ranking, four of the top five software projects are the same.
+Any more stats?
+
+In some cases projects are popular in the community, but our methodology ranked
+them low.  This is the case for \textit{dwv}. The reason for the low ranking is
+that, as mentioned in Section~\ref{sec_result_installability}, we failed to
+build it locally, and used the test version on its websites for the
+measurements. We followed the instructions and tried to run the command ``yarn
+run test'' locally, which did not work. In addition, the test version did not
+detect a broken DICOM file and displayed a blank image as described in
+Section~\ref{sec_result_robustness}. We might underestimate the scores for
+\textit{dwv} due to uncommon technical issues. We also ranked \textit{DICOM
+Viewer} much lower than its popularity. As mentioned in
+Section~\ref{sec_result_installability} it depended on the NextCloud platform
+that we could not successfully install. Thus, we might underestimate the scores
+of its surface reliability and surface robustness. In addition, we weighted all
+qualities equally, which is not likely to be the case with all users. As a
+result, some projects with high community popularity may have scored lower with
+our method because of poor scores for some qualities.  A further explanation for
+discrepancies between our measures and the star measures may also be due to
+inaccuracy with using stars to approximate popularity.  Stars are not an idea
+measure because stars represent the community's feeling in the past more than
+current preferences \citep{Szulik2017}, since stars tend only to be added, not
+removed.
 
@@
 \end{tabular}
-\caption{\label{tab_ranking_vs_GitHub}Software ranking versus GitHub metrics
-(Sorted in descending order of GitHub popularity, as estimated by the number of
-new stars per year)}
+\caption{Software ranking versus GitHub metrics (Sorted in descending order of
+GitHub popularity, as estimated by the number of new stars per year)}
+\label{tab_ranking_vs_GitHub}
 \end{table}
@@
 As shown in Section~\ref{sec_software_selection}, our domain experts recommended
-a list of top software with 12 software products.  All of the top 4 software
-from the Domain Expert's list are among the top 12 ranked by our methodology. 3
-of the top 4 on both lists are the same: \textit{3D Slicer}, \textit{ImageJ},
+a list of top software with 12 software products.  All of the top 4 entries from
+the Domain Expert's list are among the top 12 ranked by our methodology. Three
+of the top four on both lists are the same: \textit{3D Slicer}, \textit{ImageJ},
 and \textit{Fiji}. \textit{3D Slicer} is top project by both rankings (and by
@@
 Expert has significant experience with MI software, he has not used every one of
-the 29 packages.
+the 29 packages that were measured.
+
+Although our ranking and the estimate of the community's ranking are not perfect
+measures, they do suggest a correlation between best practices and popularity.
+We do not know which comes first, the use of best practices or popularity, but
+we do know that the top ranked packages tend to incorporate best practices. The
+next sections will explore how the practices of the MI community compare to the
+broader research software community. We will also investigate the practices from
+the top projects that others within the MI community, and within the broader
+research software community, can potentially adopt.
 

### 2022-03-25 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Update abstract in SOP for LBM to address Jacques's feedback (#135).

@@
     software using the Analytic Hierarchy Process (AHP). Our ranking was roughly
-    consistent with ranking the repositories by the number of GitHub stars,
-    although the number one project by star count (Sailfish) is ranked 9th by
-    our methodology. Our finding is that the state of the practice is healthy
-    with 67\% of the measured packages ranking in the top five for at least one
-    quality, the majority of LBM generated artifacts corresponding to general
-    recommendations from research software developers, common utilization of
-    version control (67\% of packages) and adoption of a quasi-agile development
-    process.  Areas of best practice to potentially improve include adoption of
-    continuous integration, API documentation and enforcement of programming
-    style guides.  We interviewed four developers to gain insight into their
-    current pain points. Identified challenges include lack of development time,
-    lack of funding, and difficulty with ensuring correctness.  Developers are
-    addressing these pain points by designing for change, circumventing the
-    oracle problem and prioritizing documentation and usability.  For future
-    best practices, we suggest the community consider the following: employing
-    linters, conducting rigorous peer reviews, writing and submitting more
-    papers on software, growing the number of contributors and augmenting the
-    theory manuals to include more requirements relevant information.
+    consistent with ranking by GitHub stars, suggesting at least a partial
+    correlation between the use of best practices and popularity. We find the
+    state of the practice to be healthy, with 67\% of the measured packages
+    ranking in the top five for at least one quality, the majority of LBM
+    generated artifacts corresponding to general recommendations from research
+    software developers, common utilization of version control (67\% of
+    packages) and adoption of a quasi-agile development process.  Areas of best
+    practice to potentially improve include adoption of continuous integration,
+    API documentation and enforcement of programming style guides.  We
+    interviewed four developers to gain insight into their current pain points.
+    Identified challenges include lack of development time, lack of funding, and
+    difficulty with ensuring correctness.  Developers are addressing these pain
+    points by designing for change, circumventing the oracle problem and
+    prioritizing documentation and usability. For future improvements we suggest
+    the following: employing linters, conducting rigorous peer reviews, writing
+    and submitting more papers on software, growing the number of contributors
+    by following current recommendations for open source projects, and
+    augmenting the theory manuals to include more requirements relevant
+    information.
 

### 2022-03-25 StateOfPractice/Papers/MedImageSoft_SOP-arXiv/MedImageSoft_SOP_AcademicRadiology.tex
Updates to SOP for MI paper

@@
 
-\section{Comparison Between LBM and Research Software for Artifacts}
+\section{Comparison Between MI and Research Software for Artifacts}
 \label{Sec_CompareArtifacts}
@@
 
-\section{Comparison of Tool Usage Between LBM and Other Research Software}
+We answered this question by examining and analyzing the repositories for each
+of the 29 projects.  In particular, as part of assessing maintainability, we
+listed the artifacts (documents, test scripts etc.) that appear in each project.
+The full measurement results are summarized in \citet{Dong2021-Data}.  The
+artifacts found were as follows (with the number of occurrences in brackets):
+Source code (29), README (29), Version Control (29), License (28), Bug tracker
+(28), Change request (28), User Manual (22), Release notes (22), Build file
+(18), Tutorials (18), Installation Guide (16), Test cases (15), Authors (14),
+FAQ (14), Acknowledgements (12), Executable files (10), Developer's Manual (8),
+API documentation (7), Troubleshooting guide (6), Project Plan (5).  The
+traceability of some document types to the projects in which they occur was
+previously presented in Table~\ref{tab_maintainability_docs} for maintainability
+related documents, and Table~\ref{tab_Visibility/Transparency_docs} for
+visibility/transparency related documents.
+
+The list in the previous paragraph shows that some types of software engineering
+recommended artifacts are missing in MI projects.  For instance, we did not find
+evidence of user stories (an Agile software development tool for capturing
+software features from an end-user perspective), user requirements, test plans,
+test code coverage, etc.  Our preliminary work suggests that this is not unique
+to MI software.  A gap exists in scientific computing development practices and
+software engineering recommendations \citep{OwojaiyeEtAl2021_CSE}.
+
+\section{Comparison of Tool Usage Between MI and Other Research Software}
 \label{Sec_CompareTools}
@@
 
-\section{Interviews with Developers} \label{ch_interview}
+\section{Developer Pain Points} \label{painpoints}
+
+Based on interviews with four developers, this section aims to answer the
+research questions: i) What are the pain points for developers working on
+research software projects (\rqref{RQ_PainPoints})?; and, ii) How do the pain
+points of developers from LBM compare to the pain points for research software
+in general (\rqref{RQ_ComparePainPoints})?  Below we go through each of the
+identified pain points and include citations that contrast the LBM experience
+with observations from researchers in other domains.  Potential ways to address
+the pain points are covered in Section~\ref{Sec_AddressConcerns}. The full
+interview questions are found in \citet{SmithEtAl2021}.
 
@@
 
-\subsection{Current and Past Pain Points} \label{sec_interview_pain_points}
+%\label{sec_interview_pain_points}
 
@@
 \toprule
-\multirow{2}{*}{Category} & \multirow{2}{*}{Obstacle} & \multicolumn{2}{c}{Num ans.} \\ \cline{3-4} 
- &  & current & past \\ 
+\multirow{2}{*}{Category} & \multirow{2}{*}{Obstacle} & \multicolumn{2}{c}{Num
+ ans.} \\ \cline{3-4} &  & current & past \\ 
 \midrule
@@
 
+\section{Interviews with Developers} \label{ch_interview}
+
 \subsection{Documents in the Projects} \label{sec_interview_documents}
@@
 in turn.
-
-\begin{description}
-\item[RQ1.] What artifacts are present in current software projects?
-\end{description}
-
-We answered this question by examining and analyzing the repositories for each
-of the 29 projects.  In particular, as part of assessing maintainability, we
-listed the artifacts (documents, test scripts etc.) that appear in each project.
-The full measurement results are summarized in \citet{Dong2021-Data}.  The
-artifacts found were as follows (with the number of occurrences in brackets):
-Source code (29), README (29), Version Control (29), License (28), Bug tracker
-(28), Change request (28), User Manual (22), Release notes (22), Build file
-(18), Tutorials (18), Installation Guide (16), Test cases (15), Authors (14),
-FAQ (14), Acknowledgements (12), Executable files (10), Developer's Manual (8),
-API documentation (7), Troubleshooting guide (6), Project Plan (5).  The
-traceability of some document types to the projects in which they occur was
-previously presented in Table~\ref{tab_maintainability_docs} for maintainability
-related documents, and Table~\ref{tab_Visibility/Transparency_docs} for
-visibility/transparency related documents.
-
-The list in the previous paragraph shows that some types of software engineering
-recommended artifacts are missing in MI projects.  For instance, we did not find
-evidence of user stories (an Agile software development tool for capturing
-software features from an end-user perspective), user requirements, test plans,
-test code coverage, etc.  Our preliminary work suggests that this is not unique
-to MI software.  A gap exists in scientific computing development practices and
-software engineering recommendations \citep{OwojaiyeEtAl2021_CSE}.
 

### 2022-03-26 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
no solo tables; table 4 column reordering

@@
 
-\begin{table}
+\begin{table}[ht!]
 	\begin{center}
@@
 			\toprule
-			Name & $\%$ Issues Closed & $\%$ Code Comments & Status\\
+			Name & $\%$ Code Comments & $\%$ Issues Closed & Status\\
 			\midrule
-			MP-LABS & 100.00 & 26.67 & Dead\\
-			Musubi & Not Git & 24.19 & Alive\\
-			waLBerla & 72.90 & 22.62 & Alive\\
-			OpenLB & Not Git & 22.43 & Alive\\
-			ESPResSo & 89.26 & 21.78 & Alive\\
-			Ludwig& 60.00 & 20.70 & Alive\\
-			Palabos & 89.47 & 17.76 & Alive\\
-			SunlightLB & Not Git & 17.67 & Dead\\
-			LIMBES & Not Git & 17.39 & Dead\\
-			ESPResSo++ & 66.28 & 17.10 & Alive\\
-			HemeLB & No Issues & 16.68 & Dead\\
-			pyLBM & 66.67& 16.12 & Alive\\
-			MechSys & Not Git & 15.11 & Alive\\
-			LB3D-Prime & Not Git & 14.34 & Dead\\
-			LB3D & Not Git & 13.76 & Dead\\
-			LB2D-Prime & Not Git & 13.61 & Dead\\
-			Sailfish & 22.22 & 9.26 & Alive\\
-			lettuce & 33.33 & 8.19 & Alive\\
-			DL\_MESO (LBE) & Not Git & 8.06& Alive\\
-			TCLB & 60.32 & 6.02 & Alive\\
-			laboetie & 18.75 & 2.47 & Dead\\	
-			lbmpy& 58.33  & 2.03 & Alive\\	
-			LatBo.jl & 93.33 & 0.40 & Dead\\
-			LUMA& 85.71   & 0.20 & Alive\\
+			MP-LABS & 26.67 & 100.00 & Dead\\
+			Musubi & 24.19 & Not Git & Alive\\
+			waLBerla & 22.62 & 72.90 & Alive\\
+			OpenLB & 22.43 & Not Git & Alive\\
+			ESPResSo & 21.78 & 89.26 & Alive\\
+			Ludwig& 20.70 & 60.00 & Alive\\
+			Palabos & 17.76 & 89.47 & Alive\\
+			SunlightLB & 17.67 & Not Git & Dead\\
+			LIMBES & 17.39 & Not Git & Dead\\
+			ESPResSo++ & 17.10 & 66.28 & Alive\\
+			HemeLB & 16.68 & No Issues & Dead\\
+			pyLBM & 16.12 & 66.67 & Alive\\
+			MechSys & 15.11 & Not Git & Alive\\
+			LB3D-Prime & 14.34 & Not Git & Dead\\
+			LB3D & 13.76 & Not Git & Dead\\
+			LB2D-Prime & 13.61 & Not Git & Dead\\
+			Sailfish & 9.26 & 22.22 & Alive\\
+			lettuce & 8.19 & 33.33 & Alive\\
+			DL\_MESO (LBE) & 8.06 & Not Git & Alive\\
+			TCLB & 6.02 & 60.32 & Alive\\
+			laboetie & 2.47 & 18.75 & Dead\\	
+			lbmpy& 2.03 & 58.33 & Alive\\	
+			LatBo.jl & 0.40 & 93.33 & Dead\\
+			LUMA& 0.20 & 85.71 & Alive\\
 			\bottomrule
@@
 
-\begin{table}
+\begin{table}[ht!]
 \begin{center}

### 2022-03-26 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
replaced 'Not Git' with 'N/A' in Table 6

@@
 			Palabos & 3 & 34 & 6 & GitLab& GitLab\\
-			OpenLB & 4 & No Git & No Git & No Git& No Git\\
+			OpenLB & 4 & N/A & N/A & N/A& N/A\\
 			LUMA & 5 & 33 & 7 & 12& 4\\
 			pyLBM & 6 & 95 & 3 & 10& 5\\
-			DL\_MESO (LBE) & 7 & No Git & No Git & No Git & No Git\\
-			Musubi & 8 & No Git & No Git & No Git & No Git\\
+			DL\_MESO (LBE) & 7 & N/A & N/A & N/A & N/A\\
+			Musubi & 8 & N/A & N/A & N/A & N/A\\
 			Sailfish & 9 & 186 & 1 & 41& 1\\
@@
 			TCLB & 12 & 95 & 3 & 16& 3\\
-			MechSys & 13 & No Git & No Git & No Git& No Git\\
+			MechSys & 13 & N/A & N/A & N/A& N/A\\
 			lettuce & 14 & 48 & 4 & 5& 8\\
@@
 			MP-LABS & 16 & 12 & 11 & 2& 9\\			
-			SunlightLB & 17 & No Git & No Git & No Git& No Git\\
-			LB3D & 18 & No Git & No Git & No Git& No Git\\			
-			LIMBES & 19 & No Git & No Git & No Git& No Git\\
-			LB2D-Prime & 20 & No Git & No Git & No Git& No Git\\		
+			SunlightLB & 17 & N/A & N/A & N/A& N/A\\
+			LB3D & 18 & N/A & N/A & N/A& N/A\\			
+			LIMBES & 19 & N/A & N/A & N/A& N/A\\
+			LB2D-Prime & 20 & N/A & N/A & N/A& N/A\\		
 			HemeLB & 21 & 12 & 11 & 12& 4\\
 			lbmpy & 22 &  11 & 12 & 2& 9\\	
-			LB3D-Prime & 23 & No Git & No Git & No Git& No Git\\	
+			LB3D-Prime & 23 & N/A & N/A & N/A& N/A\\	
 			LatBo.jl & 24 & 17 & 10 & 8& 6\\			

### 2022-03-26 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
table 5 column reordering

@@
 	\begin{center}
-		\begin{tabular}{ p{3.5cm}p{2cm}p{2.5cm}p{2cm}p{2.5cm} }
+		\begin{tabular}{ p{3cm}p{3cm}p{2cm}p{2cm}p{2cm} }
 			\toprule
-			Name & Text Files & Binary Files & LOC & Avg.\ LOC / Text File\\
+			Name & Avg.\ LOC / Text File & LOC & Text Files & Binary Files \\
 			\midrule
-			LUMA & 314 & 19 & 4399723 & 14011\\
-			LatBo.jl & 41& 0 & 42172 & 1029\\
-			LB2D-Prime & 82& 19 & 54755 & 668\\
-			LB3D-Prime & 23& 6 & 12944 & 563\\
-			DL\_MESO (LBE) & 310 & 51 & 170223 & 549\\
-			LB3D & 99 & 76 & 39766 & 402\\
-			laboetie & 133& 1 & 48403 & 364\\
-			waLBerla & 2643 & 69 & 873988 & 331\\
-			MechSys & 333 & 3 & 95707 & 287\\
-			Palabos & 1974 & 71 & 563841 & 286\\
-			lbmpy& 250 & 29 & 61632 & 247\\
-			SunlightLB & 36 & 1 & 7646 & 212\\
-			Musubi & 1347 & 1839 & 281879 & 209\\
-			LIMBES & 26 & 1 & 4872 & 187\\
-			Ludwig & 954 & 32 & 162518 & 170\\
-			OpenLB & 1438 & 7 & 218406 & 152\\
-			ESPResSo & 1390& 83 & 195083 & 140\\
-			MP-LABS & 307 & 3 & 43124 & 140\\
-			pyLBM & 272 & 108 & 37234 & 137\\
-			ESPResSo++ & 1406& 30 & 165194 & 118\\
-			HemeLB & 1102& 44 & 123806 & 112\\
-			Sailfish & 632 & 11 & 69398 & 110\\
-			lettuce & 73 & 1 & 7660 & 105\\
-			TCLB & 594 & 7 & 49156 & 83\\
+			LUMA & 14011 & 4399723 & 314 & 19 \\
+			LatBo.jl & 1029 & 42172 & 41& 0 \\
+			LB2D-Prime & 668 & 54755 & 82& 19 \\
+			LB3D-Prime & 563 & 12944 & 23& 6 \\
+			DL\_MESO (LBE) & 549 & 170223 & 310 & 51 \\
+			LB3D & 402 & 39766 & 99 & 76 \\
+			laboetie & 364 & 48403 & 133& 1 \\
+			waLBerla & 331 & 873988 & 2643 & 69 \\
+			MechSys & 287 & 95707 & 333 & 3 \\
+			Palabos & 286 & 563841 & 1974 & 71 \\
+			lbmpy & 247 & 61632 & 250 & 29 \\
+			SunlightLB & 212 & 7646 & 36 & 1 \\
+			Musubi & 209 & 281879 & 1347 & 1839 \\
+			LIMBES & 187 & 4872 & 26 & 1 \\
+			Ludwig & 170 & 162518 & 954 & 32 \\
+			OpenLB & 152 & 218406 & 1438 & 7 \\
+			ESPResSo & 140 & 195083 & 1390& 83 \\
+			MP-LABS & 140 & 43124 & 307 & 3 \\
+			pyLBM & 137 & 37234 & 272 & 108 \\
+			ESPResSo++ & 118 & 165194 & 1406& 30 \\
+			HemeLB & 112 & 123806 & 1102& 44 \\
+			Sailfish & 110 & 69398 & 632 & 11 \\
+			lettuce & 105 & 7660 & 73 & 1 \\
+			TCLB & 83 & 49156 & 594 & 7 \\
 			\bottomrule

### 2022-03-26 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
added comment to tables about sorting, sorted some tables

@@
 		\end{tabular}
-		\caption{Alive Software Packages} \label{alivepackages}
+		\caption{Alive Software Packages (Sorted Alphabetically)} \label{alivepackages}
 	\end{center}
@@
 		\end{tabular}
-		\caption{Dead Software Packages} \label{deadpackages}
+		\caption{Dead Software Packages (Sorted Alphabetically)} \label{deadpackages}
 	\end{center}
@@
 			Complex Geometries (Yes or No), OS for Operating System (Windows
-			(W), macOS (M), Linux (L)))} \label{tbl_features}
+			(W), macOS (M), Linux (L))) (Sorted Alphabetically)} \label{tbl_features}
 	\end{center}
@@
 		\caption{Git Repository Data, Sorted by Percentage of Code that is
-			Comments} \label{gitrepodata}
+			Comments (Sorted by $\%$ Code Comments)} \label{gitrepodata}
 	\end{center}
@@
 		\end{tabular}
-		\caption{Module Data} \label{moduledata}
+		\caption{Module Data (Sorted by Avg.\ LOC / Text File)} \label{moduledata}
 	\end{center}
@@
 		\end{tabular}
-		\caption{Repository Ranking Metrics} \label{repometrics}
+		\caption{Repository Ranking Metrics (Sorted by Our Ranking)} \label{repometrics}
 	\end{center}
@@
 Authors / Developers List & Change Log / Release Notes & API Documentation\\
-Issue Tracker & Design Documentation & Developer / Contributor Manual\\
-Library Dependency List & Functional Spec. / Notes & FAQ / Forum\\
-Installation Guide / Instructions & Performance Information & Verification and
+Installation Guide / Instructions & Design Documentation & Developer / Contributor Manual\\
+Issue Tracker & Functional Spec. / Notes & FAQ / Forum\\
+Library Dependency List & Performance Information & Requirements Spec.\\
+License & Test Plan / Script / Cases & Verification and
 Validation Plan\\
-Theory Notes & Test Plan / Script / Cases & Video Guide
+List of Related Publications & User Manual / Guide & Video Guide
 (including YouTube)\\
-List of Related Publications & User Manual / Guide & Requirements Spec.\\
 Makefile / Build File &  & \\
 README File & & \\
-License & & \\
+Theory Notes & & \\
 Tutorial & & \\
@@
 \end{tabular}
-\caption{Artifacts Present in LBM Packages, Classified by Frequency}
+\caption{Artifacts Present in LBM Packages, Classified by Frequency (Sorted Alphabetically)}
 \label{artifactspresent}
@@
 	\midrule
-	Continuous Integration & Build Automation Tools & Collaboration Tools\\
-	Code Editors & Technical Libraries & Email\\
-	Development Environments & Domain Specific Libraries & Change Tracking Tools\\
-	Runtime Environments &  & Version Control Tools\\
-	Compilers &  & Document Generation Tools\\
+	Code Editors & Build Automation Tools & Change Tracking Tools\\
+	Compilers & Domain Specific Libraries & Collaboration Tools\\
+	Continuous Integration & Technical Libraries & Document Generation Tools\\
+	Correctness Verification Tools &  & Email\\
+	Development Environments &  & Version Control Tools\\
+	Runtime Environments &  & \\
 	Unit Testing Tools &  & \\
-	Correctness Verification Tools &  & \\
 	\bottomrule
@@
 	\caption{Observed development tools, dependencies and project management
-	tools} \label{tbl_tools}
+	tools (Sorted Alphabetically)} \label{tbl_tools}
 	\end{center}
@@
 		\caption{Top performers for each quality (all packages were alive at the
-		time of measurement, except for those marked by an asterisk (*))}
+		time of measurement, except for those marked by an asterisk (*)) (Sorted by Order of Quality Measurement)}
 		\label{topperformerstable}

### 2022-03-26 StateOfPractice/Papers/MedImageSoft_SOP-arXiv/MedImageSoft_SOP_AcademicRadiology.tex
Updated AHP figures
Edits to paper

@@
 	the pain points for developers working on MI software projects?
-	(Section~\ref{ch_interview})
+	(Section~\ref{painpoints})
 	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_ComparePainPoints}:] How
 	do the pain points of developers from MI compare to the pain points
-	for research software in general? (Section~\ref{ch_interview})
+	for research software in general? (Section~\ref{painpoints})
 	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Concerns}:] For MI
 	developers what specific best practices are taken to address the pain points
-	and software quality concerns? (Section~\ref{ch_interview})
+	and software quality concerns? (Section~\ref{Sec_AddressConcerns})
 	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Recommend}:]
@@
 
+We answer this question by measuring the qualities and interviewing the
+developers. We learned that developers use tools for CI/CD
+(Section~\ref{sec_result_correctness_verifiability}), user support
+(Table~\ref{tab_user_support_model}), version control
+(Section~\ref{sec_score_maintainability}), documentation
+(Section~\ref{sec_contribution_pm}), contribution management
+(Section~\ref{sec_contribution_pm}), and project management
+(Section~\ref{sec_contribution_pm}).
+
+As mentioned in Section \ref{sec_result_correctness_verifiability}, we
+identified five projects using CI/CD tools: \textit{3D Slicer} and \textit{OHIF
+Viewer} use CircleCI; \textit{ImageJ}, \textit{Fiji}, and \textit{dwv} use
+Travis. We identified the above projects and tools by examining the
+documentation and source code of all projects. Not all tools leave identifiable
+artifacts in the repositories; therefore, we may have missed the use of some
+tools. According to the interviews with developers, \textit{dwv} and
+\textit{Weasis} used GitHub Actions.
+
 \section{Comparison of Principles, Process and Methodologies to Research Software in General} \label{Sec_CompareMethodologies}
@@
 easy to have.
+
+We answer this question by measuring the qualities and interviewing the
+developers. We previously summarized principles, processes, and methodologies in
+software testing (Section~\ref{Sec_CompareMethodologies}), documentation
+(Section~\ref{sec_interview_documents}, Table~\ref{tab_maintainability_docs},
+Table~\ref{tab_Visibility/Transparency_docs}, contribution management
+(Section~\ref{sec_contribution_pm}), and project management
+(Section~\ref{sec_contribution_pm}).
+
+We identified the use of unit testing in less than half of the 29 projects. On
+the other hand, the interviewees believed that testing (including usability
+tests with users) was the top solution to improve correctness, usability, and
+reproducibility. One pain point in the development process is the lack of access
+to real-world datasets for testing. The developers' strategies to address this
+are summarized in Section~\ref{Sec_AddressConcerns} \ad{to be fixed}. One threat to
+correctness is: with huge datasets for testing, the tests are expensive and
+time-consuming. Three interviewees endorsed self tests / automated tests, which
+may save time for testing.
+
+Documentation is apparently not emphasized in any of the 29 projects.  None of
+them had theory manuals, although we did identify a road map in the \textit{3D
+Slicer} project.  No requirements specifications were found. Eight of the nine
+interviewees thought that documentation was essential to their projects.
+However, they hold the common opinion that their documentation needed
+improvements. Nearly half of them also believed that the lack of time prevented
+them from improving the documentation.
+
+\subsection{Documents in the Projects} \label{sec_interview_documents}
+
+We tried to understand the interviewees' opinions on documentation and the
+quality of documentations with two questions:
+
+\begin{description}
+\item[Q11.] How does documentation fit into your development process? Would
+improved documentation help with the obstacles you typically face?
+\item[Q19.] Do you think the current documentation can clearly convey all
+necessary knowledge to the users? If yes, how did you successfully achieve it?
+If no, what improvements are needed?
+\end{description}
+
+Table~\ref{tab_opinion_doc} summarizes interviewees' opinions on documentation.
+Interviewees from each of the eight projects thought that documentation was
+important to their projects, and most of them said that it could save their time
+to answer questions from users and developers. Most of them saw the need to
+improve their documentation, and only three of them thought that their
+documentations conveyed information clearly enough.
+
+\begin{table}[!ht]
+\centering
+\begin{tabular}{ll}
+\toprule
+Opinion on documentation & Num ans. \\ 
+\midrule
+Documentation is vital to the project & 8 \\
+Documentation of the project needs improvements & 7 \\
+Referring to documentation saves time to answer questions & 6 \\
+Lack of time to maintain good documentation & 4 \\
+Documentation of the project conveys information clearly & 3 \\
+Coding is more fun than documentation & 2 \\
+Users help each other by referring to documentation & 1 \\ 
+\bottomrule
+\end{tabular}
+\caption{\label{tab_opinion_doc}Opinions on documentation by the numbers of
+interviewees with the answers}
+\end{table}
+
+For documentation tools and methods mentioned by the interviewees, the most
+popular (mentioned by about 30\% of developers) were forum discussions and
+videos.  The second most popular options (mentioned by about 20\% of developers)
+were GitHub, wiki pages, workshops, and social media.  The least frequently
+mentioned options (about 10\% of developers) included writing books, google
+forms and state management.
+
+\subsection{Project Management} \label{sec_contribution_pm}
+
+To investigate project management we asked three questions:
+
+\begin{description}
+\item[Q5.] Do you have a defined process for accepting new contributions into
+your team?
+\item[Q13.] What is your software development model? For example, waterfall,
+agile, etc.
+\item[Q14.] What is your project management process? Do you
+think improving this process can tackle the current problem? Were any project
+management tools used?
+\end{description}
+
+Although teams may have a process for accepting new contributions, no one
+discussed this during their interview. However, most teams (8 of 9) mentioned
+using GitHub and pull requests to manage contributions from the community. The
+interviewees generally gave very positive feedback on using GitHub. Some teams
+previously used a different approach to version control and eventually
+transferred to git and GitHub.  The past approaches included contributions
+from e-mail (3 teams), contributions from forums (1 team) and sharing the
+git repository via e-mail (1 team).  \wss{I do not know what this last one means.
+Should ask Ao.} \ad{The last one happened when cloud-based repositories (Github,
+Gitlab, Bitbucket, etc.) were not widely used. The project git repository was
+stored locally on the project owner's local computer. The owner shared the whole
+git repo to team members and other individuals who wanted to contribute. The
+contributors made changes to the repo, and sent it back. They transferred the
+repo back and forth by email.}
+
+For managing contributions, the \textit{3D Slicer} team encouraged users to
+develop their extensions for specific use cases, while the \textit{OHIF} team
+enabled the use of plug-ins.  The interviewee from the \textit{ITK-SNAP} team
+said one way of accepting new team members was through funded academic projects.
+
+For the most part, the responses about development model were vague, with only
+two interviewees following a definite development model. In some cases the
+interviewees felt their process was similar to an existing development model.
+Three teams (about 38\%) either followed agile, or something similar to agile.
+Two teams (25\%) either followed a waterfall process, or something similar.
+Three teams (about 38\%) explicitly stated that their process was undefined or
+self-directed.
+
+Some interviewees mentioned the project management tools they used. Generally
+speaking, the interviewees talked about two types of tools:
+\begin{inparaenum}[i)]
+\item trackers, including GitHub, issue trackers, bug trackers and Jira; and,
+\item documentation tools, including GitHub, Wiki page, Google Doc, and Confluence.
+\end{inparaenum}
+Of the specifically named tools in the above lists, GitHub was mentioned 3
+times, while each of the other tools was only mentioned once.
+
+No interviewee introduced any strictly defined project management process. The
+most common approach was following the issues, such as bugs and feature requests.
+Additionally, the \textit{3D Slicer} team had weekly meetings to discuss the
+goals for the project; the \textit{INVESALIUS 3} team relied on the GitHub
+process for their project management; the \textit{ITK-SNAP} team had a fixed
+six-month release pace; only the interviewee from the \textit{OHIF} team
+mentioned that the team has a project manager; the \textit{3D Slicer} team and
+\textit{BioImage Suite Web} team do nightly builds and tests.
+
+Most interviewees skipped the part of the last question asking ``Do you think
+improving this process can tackle the current problem?''. In retrospect, we
+should not have asked a yes-or-no question, since a binary answer is not very
+informative. The interviewee from the \textit{OHIF} team gave a positive answer
+to this question. They believed that a better project management process can
+improve the efficiency of junior developers. They also improved the project
+management tools (from public Jira to public GitHub repository plus private
+Jira), so they could better communicate externally and internally.
 
@@
 
-\section{Interviews with Developers} \label{ch_interview}
-
-\subsection{Documents in the Projects} \label{sec_interview_documents}
-
-We tried to understand the interviewees' opinions on documentation and the
-quality of documentations with two questions:
-
-\begin{description}
-\item[Q11.] How does documentation fit into your development process? Would
-improved documentation help with the obstacles you typically face?
-\item[Q19.] Do you think the current documentation can clearly convey all
-necessary knowledge to the users? If yes, how did you successfully achieve it?
-If no, what improvements are needed?
-\end{description}
-
-Table~\ref{tab_opinion_doc} summarizes interviewees' opinions on documentation.
-Interviewees from each of the eight projects thought that documentation was
-important to their projects, and most of them said that it could save their time
-to answer questions from users and developers. Most of them saw the need to
-improve their documentation, and only three of them thought that their
-documentations conveyed information clearly enough.
-
-\begin{table}[!ht]
-\centering
-\begin{tabular}{ll}
-\toprule
-Opinion on documentation & Num ans. \\ 
-\midrule
-Documentation is vital to the project & 8 \\
-Documentation of the project needs improvements & 7 \\
-Referring to documentation saves time to answer questions & 6 \\
-Lack of time to maintain good documentation & 4 \\
-Documentation of the project conveys information clearly & 3 \\
-Coding is more fun than documentation & 2 \\
-Users help each other by referring to documentation & 1 \\ 
-\bottomrule
-\end{tabular}
-\caption{\label{tab_opinion_doc}Opinions on documentation by the numbers of
-interviewees with the answers}
-\end{table}
-
-For documentation tools and methods mentioned by the interviewees, the most
-popular (mentioned by about 30\% of developers) were forum discussions and
-videos.  The second most popular options (mentioned by about 20\% of developers)
-were GitHub, wiki pages, workshops, and social media.  The least frequently
-mentioned options (about 10\% of developers) included writing books, google
-forms and state management.
-
-\subsection{Project Management} \label{sec_contribution_pm}
-
-To investigate project management we asked three questions:
-
-\begin{description}
-\item[Q5.] Do you have a defined process for accepting new contributions into
-your team?
-\item[Q13.] What is your software development model? For example, waterfall,
-agile, etc.
-\item[Q14.] What is your project management process? Do you
-think improving this process can tackle the current problem? Were any project
-management tools used?
-\end{description}
-
-Although teams may have a process for accepting new contributions, no one
-discussed this during their interview. However, most teams (8 of 9) mentioned
-using GitHub and pull requests to manage contributions from the community. The
-interviewees generally gave very positive feedback on using GitHub. Some teams
-previously used a different approach to version control and eventually
-transferred to git and GitHub.  The past approaches included contributions
-from e-mail (3 teams), contributions from forums (1 team) and sharing the
-git repository via e-mail (1 team).  \wss{I do not know what this last one means.
-Should ask Ao.} \ad{The last one happened when cloud-based repositories (Github,
-Gitlab, Bitbucket, etc.) were not widely used. The project git repository was
-stored locally on the project owner's local computer. The owner shared the whole
-git repo to team members and other individuals who wanted to contribute. The
-contributors made changes to the repo, and sent it back. They transferred the
-repo back and forth by email.}
-
-For managing contributions, the \textit{3D Slicer} team encouraged users to
-develop their extensions for specific use cases, while the \textit{OHIF} team
-enabled the use of plug-ins.  The interviewee from the \textit{ITK-SNAP} team
-said one way of accepting new team members was through funded academic projects.
-
-For the most part, the responses about development model were vague, with only
-two interviewees following a definite development model. In some cases the
-interviewees felt their process was similar to an existing development model.
-Three teams (about 38\%) either followed agile, or something similar to agile.
-Two teams (25\%) either followed a waterfall process, or something similar.
-Three teams (about 38\%) explicitly stated that their process was undefined or
-self-directed.
-
-Some interviewees mentioned the project management tools they used. Generally
-speaking, the interviewees talked about two types of tools:
-\begin{inparaenum}[i)]
-\item trackers, including GitHub, issue trackers, bug trackers and Jira; and,
-\item documentation tools, including GitHub, Wiki page, Google Doc, and Confluence.
-\end{inparaenum}
-Of the specifically named tools in the above lists, GitHub was mentioned 3
-times, while each of the other tools was only mentioned once.
-
-No interviewee introduced any strictly defined project management process. The
-most common approach was following the issues, such as bugs and feature requests.
-Additionally, the \textit{3D Slicer} team had weekly meetings to discuss the
-goals for the project; the \textit{INVESALIUS 3} team relied on the GitHub
-process for their project management; the \textit{ITK-SNAP} team had a fixed
-six-month release pace; only the interviewee from the \textit{OHIF} team
-mentioned that the team has a project manager; the \textit{3D Slicer} team and
-\textit{BioImage Suite Web} team do nightly builds and tests.
-
-Most interviewees skipped the part of the last question asking ``Do you think
-improving this process can tackle the current problem?''. In retrospect, we
-should not have asked a yes-or-no question, since a binary answer is not very
-informative. The interviewee from the \textit{OHIF} team gave a positive answer
-to this question. They believed that a better project management process can
-improve the efficiency of junior developers. They also improved the project
-management tools (from public Jira to public GitHub repository plus private
-Jira), so they could better communicate externally and internally.
-
 \subsection{Discussions on Software Qualities} \label{sec_interview_software_qualities}
@@
 \begin{description}
-\item[Q20.] Do you have any concern that your computational results won’t be
+\item[Q20.] Do you have any concern that your computational results won't be
 reproducible in the future? Have you taken any steps to ensure reproducibility?
@@
 
-\section{Answers to Research Questions} \label{ch_answers}
-
-This section answers the research questions from
-Section~\ref{sec_motivation}.  We will go through each of the questions
-in turn.
-
-\begin{description}
-\item[RQ2.] What tools are used in the development of current software packages?
-\end{description}
-
-We answer this question by measuring the qualities and interviewing the
-developers. We learned that developers use tools for CI/CD
-(Section~\ref{sec_result_correctness_verifiability}), user support
-(Table~\ref{tab_user_support_model}), version control
-(Section~\ref{sec_score_maintainability}), documentation
-(Section~\ref{sec_contribution_pm}), contribution management
-(Section~\ref{sec_contribution_pm}), and project management
-(Section~\ref{sec_contribution_pm}).
-
-As mentioned in Section \ref{sec_result_correctness_verifiability}, we
-identified five projects using CI/CD tools: \textit{3D Slicer} and \textit{OHIF
-Viewer} use CircleCI; \textit{ImageJ}, \textit{Fiji}, and \textit{dwv} use
-Travis. We identified the above projects and tools by examining the
-documentation and source code of all projects. Not all tools leave identifiable
-artifacts in the repositories; therefore, we may have missed the use of some
-tools. According to the interviews with developers, \textit{dwv} and
-\textit{Weasis} used GitHub Actions.
-
-\begin{description}
-\item[RQ3.] What principles, processes, and methodologies are used in the
-development of current software packages?
-\end{description}
-
-We answer this question by measuring the qualities and interviewing the
-developers. We previously summarized principles, processes, and methodologies in
-software testing (Section~\ref{sec_interview_pain_points} and
-Section~\ref{sec_interview_software_qualities}), documentation
-(Section~\ref{sec_interview_documents}, Table~\ref{tab_maintainability_docs},
-Table~\ref{tab_Visibility/Transparency_docs}, and Section~\ref{ch_answers}),
-contribution management (Section~\ref{sec_contribution_pm}), and project
-management (Section~\ref{sec_contribution_pm}).
-
-We identified the use of unit testing in less than half of the 29 projects. On
-the other hand, the interviewees believed that testing (including usability
-tests with users) was the top solution to improve correctness, usability, and
-reproducibility. One pain point in the development process is the lack of access
-to real-world datasets for testing. The developers' strategies to address this
-are summarized in Section~\ref{ch_interview} \ad{to be fixed}. One threat to correctness is:
-with huge datasets for testing, the tests are expensive and time-consuming.
-Three interviewees endorsed self tests / automated tests, which may save time
-for testing.
-
-Documentation is apparently not emphasized in any of the 29 projects.  None of
-them had theory manuals, although we did identify a road map in the \textit{3D
-Slicer} project.  No requirements specifications were found. Eight of the nine
-interviewees thought that documentation was essential to their projects.
-However, they hold the common opinion that their documentation needed
-improvements. Nearly half of them also believed that the lack of time prevented
-them from improving the documentation.
-
-\begin{description}
-\item[RQ4.] What are the pain points for developers working on research software
-projects? What aspects of the existing processes, methodologies, and tools do
-they consider as potentially needing improvement? What changes to processes,
-methodologies, and tools can improve software development and software quality?
-\end{description}
-
-We answer this question by interviewing the developers. The answers to this
-question, including the pain points and the solutions proposed by the
-developers, are in Section~\ref{sec_interview_pain_points}.
-
-\begin{description}
-\item[RQ5.] What is the current status of the following
-software qualities for the projects? What actions have the developers taken to
-address them?
-\end{description}
-
-The current status of the software qualities was assessed through filling in the
-measurement template \citep{Dong2021-Data}.  A summary of the comparison between
-all projects can be found in Section~\ref{ch_results}.  We explored some of the
-qualities more deeply, especially with respect to techniques addressing quality
-concerns, through the interview questions.  Specifically, the qualities of
-correctness \& verifiability, usability, maintainability, understandability and
-reproducibility are explored in Section~\ref{ch_interview}.
-
-\begin{description}
-\item[RQ6.] How does software designated as high quality by this methodology
-compare with top-rated software by the community?
-\end{description}
+\section{Lessons from MI Developers} \label{Sec_AddressConcerns}
+
+This section summarizes the best practices from LBM developers that are taken to
+address the pain points mentioned in Section~\ref{painpoints}.  The main source
+of information is the qualitative data from developer interviews
+(Section~\ref{sec_interview_methods}).  The practices summarized in this section
+can potentially be emulated by the LBM software packages that do not currently
+follow them.  Moreover, these practices may also provide examples that can be
+followed by other research software domains.
 
@@
 no artifacts of this activity remained in the publicly available repository.
-\item As mentioned in Section~\ref{ch_interview}, one interviewee was too busy
+\item As mentioned in Section~\ref{painpoints}, one interviewee was too busy
 to participate in a full interview, so they provided a version of written answers
@@
 \ref{sec_recommendations_testing_dataset} are based on the primary pain points
-(from Section~\ref{sec_interview_pain_points}) collected from developers in the
+(from Section~\ref{painpoints}) collected from developers in the
 MI domain.
@@
 Based on our quality measurements in Section~\ref{ch_results} and discussions
-with the developers in Section~\ref{ch_interview}, we collected key points for
+with the developers in Section~\ref{painpoints}, we collected key points for
 improving software qualities. These points should be considered for new and
@@
 lack of fundings, time, and developers. We summarized our discussion with the MI
-software developers in Section~\ref{sec_interview_pain_points}.  Many of our
-recommendations involve investing more effort into processes, tools and methods,
-so that the limited resources can be used more effectively.  Our recommendations
-are as follows:
+software developers in Section~\ref{painpoints}.  Many of our recommendations
+involve investing more effort into processes, tools and methods, so that the
+limited resources can be used more effectively.  Our recommendations are as
+follows:
 
@@
 A tech stack refers to a set of technologies used by a team to build software
-and manage the project. Section~\ref{sec_interview_pain_points} lists the
-advantages and disadvantages between native and web applications. In this
-section, we give further suggestions on the choice of a tech stack to improve
-the four priority factors identified by developers: compatibility,
-maintainability, performance and security.  The suggestions are intended to
-provide ideas and avenues for exploration; not all of the suggestions will be
-the right fit for all projects and all teams.
+and manage the project. Section~\ref{painpoints} lists the advantages and
+disadvantages between native and web applications. In this section, we give
+further suggestions on the choice of a tech stack to improve the four priority
+factors identified by developers: compatibility, maintainability, performance
+and security.  The suggestions are intended to provide ideas and avenues for
+exploration; not all of the suggestions will be the right fit for all projects
+and all teams.
 
@@
 
-As described in Section~\ref{sec_interview_pain_points}, it is difficult for
+As described in Section~\ref{painpoints}, it is difficult for
 software development teams to access real-world medical imaging datasets. This
@@
 understanding current practice, answering our six research questions
-(Section~\ref{sec_motivation}) and providing recommendations for current
-and future projects.  Our methods in Section~\ref{ch_methods} form a general
-process to evaluate domain-specific software, that we apply to the specific
-domain of MI software. We identified 48 MI software candidates, then, with the
-help of the Domain Expert selected 29 of them to our final list.
-Section~\ref{ch_results} lists our measurements to nine software qualities for
-the 29 projects, and Section~\ref{ch_interview} contains our interviews with
-eight of the 29 teams, discussing their development process and five software
-qualities.  We answered our six research questions in Section~\ref{ch_answers}.
-In addition, Section~\ref{ch_recommendations} presents our recommendations on SC
-software development.
+(Section~\ref{sec_motivation}) and providing recommendations for current and
+future projects.  Our methods in Section~\ref{ch_methods} form a general process
+to evaluate domain-specific software, that we apply to the specific domain of MI
+software. We identified 48 MI software candidates, then, with the help of the
+Domain Expert selected 29 of them to our final list. Section~\ref{ch_results}
+lists our measurements to nine software qualities for the 29 projects, and
+Section~\ref{painpoints} contains our interviews with eight of the 29 teams,
+discussing their development process and five software qualities.  We answered
+our research questions. In addition, Section~\ref{ch_recommendations} presents
+our recommendations on SC software development.
 
@@
 
-The interview results in Section~\ref{ch_interview} show some merits, drawbacks,
+The interview results in Section~\ref{painpoints} show some merits, drawbacks,
 and pain points within the development process. The three primary categories of
@@
 
-Our answers to the research questions (Section~\ref{ch_answers}) are based on
-the above findings. We identified the existing artifacts, tools, principles,
-processes, and methodologies in the 29 projects. By comparisons with the implied
-popularity of existing projects we found: 1) four of the top five software
-projects in our ranking were also among the top five ones receiving the most
-GitHub stars per year (Table~\ref{tab_ranking_vs_GitHub}); 2) three of the top
-four in our ranking were among the top four provided by the domain experts.
+Our answers to the research questions are based on the above findings. We
+identified the existing artifacts, tools, principles, processes, and
+methodologies in the 29 projects. By comparisons with the implied popularity of
+existing projects we found: 1) four of the top five software projects in our
+ranking were also among the top five ones receiving the most GitHub stars per
+year (Table~\ref{tab_ranking_vs_GitHub}); 2) three of the top four in our
+ranking were among the top four provided by the domain experts.
 
@@
 - Radiology - should not exceed 6000 words, 50 refs, 8 figures and 4 tables
+
+- add discussion of sensitivity analysis (borrow from LBM)

### 2022-03-27 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
sorted table 5

@@
 	\begin{center}
-		\begin{tabular}{ p{3cm}p{3cm}p{2cm}p{2cm}p{2cm} }
+		\begin{tabular}{ p{3cm}p{2cm}p{2cm}p{2cm}p{3cm} }
 			\toprule
-			Name & Avg.\ LOC / Text File & LOC & Text Files & Binary Files \\
+			Name & Binary Files & LOC & Text Files & Avg.\ LOC / Text File \\
 			\midrule
-			LUMA & 14011 & 4399723 & 314 & 19 \\
-			LatBo.jl & 1029 & 42172 & 41& 0 \\
-			LB2D-Prime & 668 & 54755 & 82& 19 \\
-			LB3D-Prime & 563 & 12944 & 23& 6 \\
-			DL\_MESO (LBE) & 549 & 170223 & 310 & 51 \\
-			LB3D & 402 & 39766 & 99 & 76 \\
-			laboetie & 364 & 48403 & 133& 1 \\
-			waLBerla & 331 & 873988 & 2643 & 69 \\
-			MechSys & 287 & 95707 & 333 & 3 \\
-			Palabos & 286 & 563841 & 1974 & 71 \\
-			lbmpy & 247 & 61632 & 250 & 29 \\
-			SunlightLB & 212 & 7646 & 36 & 1 \\
-			Musubi & 209 & 281879 & 1347 & 1839 \\
-			LIMBES & 187 & 4872 & 26 & 1 \\
-			Ludwig & 170 & 162518 & 954 & 32 \\
-			OpenLB & 152 & 218406 & 1438 & 7 \\
-			ESPResSo & 140 & 195083 & 1390& 83 \\
-			MP-LABS & 140 & 43124 & 307 & 3 \\
-			pyLBM & 137 & 37234 & 272 & 108 \\
-			ESPResSo++ & 118 & 165194 & 1406& 30 \\
-			HemeLB & 112 & 123806 & 1102& 44 \\
-			Sailfish & 110 & 69398 & 632 & 11 \\
-			lettuce & 105 & 7660 & 73 & 1 \\
-			TCLB & 83 & 49156 & 594 & 7 \\
+			LUMA & 19 & 4399723 & 314 & 14011 \\
+			LatBo.jl & 0 & 42172 & 41& 1029 \\
+			LB2D-Prime & 19 & 54755 & 82& 668 \\
+			LB3D-Prime & 6 & 12944 & 23& 563 \\
+			DL\_MESO (LBE) & 51 & 170223 & 310 & 549 \\
+			LB3D & 76 & 39766 & 99 & 402 \\
+			laboetie & 1 & 48403 & 133& 364 \\
+			waLBerla & 69 & 873988 & 2643 & 331 \\
+			MechSys & 3 & 95707 & 333 & 287 \\
+			Palabos & 71 & 563841 & 1974 & 286 \\
+			lbmpy & 29 & 61632 & 250 & 247 \\
+			SunlightLB & 1 & 7646 & 36 & 212 \\
+			Musubi & 1839 & 281879 & 1347 & 209 \\
+			LIMBES & 1 & 4872 & 26 & 187 \\
+			Ludwig & 32 & 162518 & 954 & 170 \\
+			OpenLB & 7 & 218406 & 1438 & 152 \\
+			ESPResSo & 83 & 195083 & 1390& 140 \\
+			MP-LABS & 3 & 43124 & 307 & 140 \\
+			pyLBM & 108 & 37234 & 272 & 137 \\
+			ESPResSo++ & 30 & 165194 & 1406& 118 \\
+			HemeLB & 44 & 123806 & 1102& 112 \\
+			Sailfish & 11 & 69398 & 632 & 110 \\
+			lettuce & 1 & 7660 & 73 & 105 \\
+			TCLB & 7 & 49156 & 594 & 83 \\
 			\bottomrule

### 2022-03-28 StateOfPractice/Papers/MedImageSoft_SOP-arXiv/MedImageSoft_SOP_AcademicRadiology.tex
Revisions to Section 5 (Comparison to community ranking)

@@
 greater attention and popularity on GitHub. Between our ranking and the GitHub
-stars-per-year ranking, four of the top five software projects are the same.
-Any more stats?
-
-In some cases projects are popular in the community, but our methodology ranked
-them low.  This is the case for \textit{dwv}. The reason for the low ranking is
-that, as mentioned in Section~\ref{sec_result_installability}, we failed to
-build it locally, and used the test version on its websites for the
+stars-per-year ranking, four of the top five software projects appear in both
+lists.Our top 5 packages are scattered among the first 8 positions on the GitHub
+list. However, as discussed below there are discrepancies between the two lists.
+
+In some cases projects are popular in the community, but were assigned a low
+rank by our methodology.  This is the case for \textit{dwv}. The reason for the
+low ranking is that, as mentioned in Section~\ref{sec_result_installability}, we
+failed to build it locally, and used the test version on its websites for the
 measurements. We followed the instructions and tried to run the command ``yarn
@@
 result, some projects with high community popularity may have scored lower with
-our method because of poor scores for some qualities.  A further explanation for
-discrepancies between our measures and the star measures may also be due to
-inaccuracy with using stars to approximate popularity.  Stars are not an idea
-measure because stars represent the community's feeling in the past more than
-current preferences \citep{Szulik2017}, since stars tend only to be added, not
-removed.
+our method because of a relatively higher (compared to the scientific
+community's implicit ranking) weighting of the poor scores for some qualities. A
+further explanation for discrepancies between our measures and the star measures
+may also be due to inaccuracy with using stars to approximate popularity.  Stars
+are not an ideal measure because stars represent the community's feeling in the
+past more than they measure current preferences \citep{Szulik2017}.  The issue
+with stars is that they tend only to be added, not removed.  A final reason for
+inconsistencies between our ranking and the community's ranking is that, as for
+consumer products, more factors influence popularity than just quality.
 
@@
 \centering
-\begin{tabular}{lllll}
+\begin{tabular}{llllll}
 \toprule
-Software & Our ranking & Stars/yr & Watches/yr & Forks/yr \\ 
+Software & Comm.\ rank & Our rank & Stars/yr & Watches/yr & Forks/yr \\ 
 \midrule
-3D Slicer & 1 & 284 & 19 & 128 \\
-OHIF Viewer & 3 & 277 & 19 & 224 \\
-dwv & 23 & 124 & 12 & 51 \\
-ImageJ & 2 & 84 & 9 & 30 \\
-ParaView & 5 & 67 & 7 & 28 \\
-Horos & 12 & 49 & 9 & 18 \\
-Papaya & 17 & 45 & 5 & 20 \\
-Fiji & 4 & 44 & 5 & 21 \\
-DICOM Viewer & 29 & 43 & 6 & 9 \\
-INVESALIUS 3 & 10 & 40 & 4 & 17 \\
-Weasis & 6 & 36 & 5 & 19 \\
-dicompyler & 26 & 35 & 5 & 14 \\
-OsiriX Lite & 9 & 34 & 9 & 24 \\
-MRIcroGL & 18 & 24 & 3 & 3 \\
-GATE & 25 & 19 & 6 & 26 \\
-Ginkgo CADx & 14 & 19 & 4 & 6 \\
-BioImage Suite Web & 8 & 18 & 5 & 7 \\
-Drishti & 27 & 16 & 4 & 4 \\
-Slice:Drop & 20 & 10 & 2 & 5 \\
-ITK-SNAP & 15 & 9 & 1 & 4 \\
-medInria & 7 & 7 & 3 & 6 \\
-SMILI & 13 & 3 & 1 & 2 \\
-MatrixUser & 28 & 2 & 0 & 0 \\
-MicroView & 16 & 1 & 1 & 1 \\
-Gwyddion & 11 & n/a & n/a & n/a \\
-XMedCon & 19 & n/a & n/a & n/a \\
-DicomBrower & 21 & n/a & n/a & n/a \\
-AMIDE & 22 & n/a & n/a & n/a \\
-3DimViewer & 24 & n/a & n/a & n/a \\ 
+3D Slicer & 1 & 1 & 284 & 19 & 128 \\
+OHIF Viewer & 2 & 3 & 277 & 19 & 224 \\
+dwv & 3 & 23 & 124 & 12 & 51 \\
+ImageJ & 4 & 2 & 84 & 9 & 30 \\
+ParaView & 5 & 5 & 67 & 7 & 28 \\
+Horos & 6 & 12 & 49 & 9 & 18 \\
+Papaya & 7 & 17 & 45 & 5 & 20 \\
+Fiji & 8 & 4 & 44 & 5 & 21 \\
+DICOM Viewer & 9 & 29 & 43 & 6 & 9 \\
+INVESALIUS 3 & 10 & 10 & 40 & 4 & 17 \\
+Weasis & 11 & 6 & 36 & 5 & 19 \\
+dicompyler & 12 & 26 & 35 & 5 & 14 \\
+OsiriX Lite & 13 & 9 & 34 & 9 & 24 \\
+MRIcroGL & 14 & 18 & 24 & 3 & 3 \\
+GATE & 15 & 25 & 19 & 6 & 26 \\
+Ginkgo CADx & 16 & 14 & 19 & 4 & 6 \\
+BioImage Suite Web & 17 & 8 & 18 & 5 & 7 \\
+Drishti & 18 & 27 & 16 & 4 & 4 \\
+Slice:Drop & 19 & 20 & 10 & 2 & 5 \\
+ITK-SNAP & 20 & 15 & 9 & 1 & 4 \\
+medInria & 21 & 7 & 7 & 3 & 6 \\
+SMILI & 22 & 13 & 3 & 1 & 2 \\
+MatrixUser & 23 & 28 & 2 & 0 & 0 \\
+MicroView & 24 & 16 & 1 & 1 & 1 \\
+Gwyddion & 25 & 11 & n/a & n/a & n/a \\
+XMedCon & 26 & 19 & n/a & n/a & n/a \\
+DicomBrower & 27 & 21 & n/a & n/a & n/a \\
+AMIDE & 28 & 22 & n/a & n/a & n/a \\
+3DimViewer & 29 & 24 & n/a & n/a & n/a \\ 
 \bottomrule
 \end{tabular}
-\caption{Software ranking versus GitHub metrics (Sorted in descending order of
-GitHub popularity, as estimated by the number of new stars per year)}
+\caption{Software ranking by our methodology versus the community (Comm.)
+ranking using GitHub metrics (Sorted in descending order of community
+popularity, as estimated by the number of new stars per year)}
 \label{tab_ranking_vs_GitHub}

### 2022-03-30 StateOfPractice/Papers/ICCS-2022-SOP_MethodologyPlusExamples/DigDeep.tex
Minor updates to Digging Deeper paper, fixes to Table 3

@@
 \midrule
-LICENSE & \checkmark &  & \checkmark & \checkmark & \checkmark & & \checkmark & \checkmark & \checkmark \\
-README &  &  & \checkmark & \checkmark & \checkmark & & \checkmark & \checkmark & \checkmark\\
-CONTRIBUTING &  &  & \checkmark & \checkmark & \checkmark & & \checkmark & \checkmark & \checkmark \\
+LICENSE & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & & \checkmark & \checkmark & \checkmark \\
+README &  & \checkmark & \checkmark & \checkmark & \checkmark & & \checkmark & \checkmark & \checkmark\\
+CONTRIBUTING &  & \checkmark & \checkmark & \checkmark & \checkmark & & \checkmark & \checkmark & \checkmark \\
 CITATION &  &  &  & \checkmark & & & & \checkmark & \checkmark \\
-CHANGELOG &  &  &  & \checkmark & \checkmark & & \checkmark & & \\
+CHANGELOG &  & \checkmark &  & \checkmark & \checkmark & & \checkmark & & \\
 INSTALL &  &  &  &  & \checkmark & & \checkmark & \checkmark & \checkmark\\
 Uninstall &  &  &  &  & & & & \checkmark & \\
+Authors &  &  &  &  &  &  & \checkmark & \checkmark & \checkmark\\
 Getting started &  &  &  &  & \checkmark & & \checkmark & \checkmark & \checkmark\\
@@
 Build Scripts &  & \checkmark &  & \checkmark & \checkmark & \checkmark & \checkmark & & \checkmark \\
+Test Plan &  & \checkmark &  &  & & \checkmark & & & \\
 Unit Tests & \checkmark & \checkmark & \checkmark &  & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
-Test Plan &  & \checkmark &  &  & & \checkmark & & & \\
 Integ. Tests &  & \checkmark & \checkmark &  & & & & \checkmark & \checkmark \\
@@
 Code of Conduct &  &  &  &  & & & \checkmark & & \\
+Acknowledgements &  &  &  &  &  &  & \checkmark & \checkmark & \checkmark\\
 \midrule

### 2022-03-30 StateOfPractice/Papers/MedImageSoft_SOP-arXiv/MedImageSoft_SOP_AcademicRadiology.tex
Edits to SOP for MI in artifacts observed versus recommended section

@@
 distributed under.  In particular, we need to know whether the source code will
-be available to us or not.  In this section, we present three common software
-categories.  We will only assess software that fits under the Open Source
-Software license.
+be available to us or not.  We define three common software categories.  We will
+only assess software that fits under the Open Source Software license.
 
@@
 
-As part of filling in the measurement template the LBM packages were examined
-for the presence of artifacts, which were then categorized by frequency. We have
+As part of filling in the measurement template the MI packages were examined for
+the presence of artifacts, which were then categorized by frequency. We have
 grouped them into common, less common, and rare artifacts in
-Table~?. Common artifacts were found in 16 to 24 ($>$63\%)
-of the software packages. Less common artifacts were found in 8 to 15 (30-63\%)
-of the software packages. Rare artifacts were found in 1 to 7 ($<$30\%) of the
-software packages.  In this section we answer \rqref{RQ_CompareArtifacts} by
-comparing the artifacts that we observed in LBM repositories to those observed
-and recommended for research software in general. Our comparison may point out
-areas where some LBM software packages fall short of current best practices.
-This is not intended to be a criticism of any existing packages, especially
-since in practice not every project needs to achieve the highest possible
-quality.  However, rather than delve into the nuances of which software can
-justify compromising which practices we will write our comparison under the
-ideal assumption that every project has sufficient resources to match best
-practices.
-
-We answered this question by examining and analyzing the repositories for each
-of the 29 projects.  In particular, as part of assessing maintainability, we
-listed the artifacts (documents, test scripts etc.) that appear in each project.
-The full measurement results are summarized in \citet{Dong2021-Data}.  The
-artifacts found were as follows (with the number of occurrences in brackets):
-Source code (29), README (29), Version Control (29), License (28), Bug tracker
-(28), Change request (28), User Manual (22), Release notes (22), Build file
-(18), Tutorials (18), Installation Guide (16), Test cases (15), Authors (14),
-FAQ (14), Acknowledgements (12), Executable files (10), Developer's Manual (8),
-API documentation (7), Troubleshooting guide (6), Project Plan (5).  The
-traceability of some document types to the projects in which they occur was
-previously presented in Table~\ref{tab_maintainability_docs} for maintainability
-related documents, and Table~\ref{tab_Visibility/Transparency_docs} for
-visibility/transparency related documents.
-
-The list in the previous paragraph shows that some types of software engineering
-recommended artifacts are missing in MI projects.  For instance, we did not find
-evidence of user stories (an Agile software development tool for capturing
-software features from an end-user perspective), user requirements, test plans,
-test code coverage, etc.  Our preliminary work suggests that this is not unique
-to MI software.  A gap exists in scientific computing development practices and
-software engineering recommendations \citep{OwojaiyeEtAl2021_CSE}.
+Table~\ref{artifactspresent}. Common artifacts were found in 20 to 29 ($>$67\%)
+of the software packages. Uncommon and rare artifacts were found in
+10 to 19 (33-67\%) and 1 to 9 ($<$33\%) of the software packages, respectively.
+The full measurements are summarized in \citet{Dong2021-Data}.  The details on
+which projects use which types of artifacts is summarized in
+Tables~\ref{tab_maintainability_docs} and~\ref{tab_Visibility/Transparency_docs}
+for documents related to maintainability and visibility, respectively.
+
+\begin{table}[ht!]
+    \begin{center}
+    \begin{tabular}{ p{4.6 cm} p{5.6 cm} p{5 cm}}
+    \toprule
+    \textbf{Common} & \textbf{Uncommon} & \textbf{Rare} \\
+    \midrule
+    README (29) & Build scripts (18) & Developer's manual (8)\\
+    Version control (29) & Tutorials (18) & API documentation (7)\\
+    License (28) & Installation guide (16) & Troubleshooting guide (6)\\
+    Issue tracker (28) & Test cases (15) & Product roadmap (5)\\
+    Change request (28) & Authors (14) & \\
+    User manual (22) & Frequently Asked Questions (FAQ) (14) & \\
+    Release info. (22) & Acknowledgements (12) & \\
+    \bottomrule
+    \end{tabular}
+    \caption{Artifacts Present in MI Packages, Classified by Frequency (The number in brackets is the number of occurrences)}
+    \label{artifactspresent}
+    \end{center}
+\end{table}
+
+We answer \rqref{RQ_CompareArtifacts} by comparing the artifacts that we
+observed in MI repositories to those observed and recommended for research
+software in general. Our comparison may point out areas where some MI software
+packages fall short of current best practices. This is not intended to be a
+criticism of any existing packages, especially since in practice not every
+project needs to achieve the highest possible quality. However, rather than
+delve into the nuances of which software can justify compromising which
+practices we will write our comparison under the ideal assumption that every
+project has sufficient resources to match best practices.
+
+Table~\ref{artifactspresent} shows that some types of software engineering
+recommended artifacts are missing in MI projects.  
+  A gap exists in scientific
+computing development practices and software engineering recommendations
+\citep{OwojaiyeEtAl2021_CSE}.
+
+\begin{table}[!ht]
+\begin{center}
+\begin{tabular}{ p{2.5cm}p{1cm}p{1cm}p{1cm}p{1cm}p{1cm}p{1cm}p{1cm}p{1.2cm}p{1cm}p{0.8cm} }
+\toprule
+~ \ & \cite{USGS2019} & \cite{TobiasEtAl2018} & \cite{BrettEtAl2021} & \cite{WilsonEtAl2016} & \cite{SmithAndRoscoe2018} & \cite{HerouxEtAl2008} & \cite{ThielEtAl2020} & \cite{vanGompelEtAl2016} & \cite{OrvizEtAl2017} & MI\\
+\midrule
+LICENSE & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & & \checkmark & \checkmark & \checkmark & C\\
+README &  & \checkmark & \checkmark & \checkmark & \checkmark & & \checkmark & \checkmark & \checkmark & C\\
+CONTRIBUTING &  & \checkmark & \checkmark & \checkmark & \checkmark & & \checkmark & \checkmark & \checkmark \\
+CITATION &  &  &  & \checkmark & & & & \checkmark & \checkmark &  \\
+CHANGELOG &  & \checkmark &  & \checkmark & \checkmark & & \checkmark &  &  &  \\
+INSTALL &  &  &  &  & \checkmark & & \checkmark & \checkmark & \checkmark & U\\
+\midrule
+Uninstall &  &  &  &  & & & & \checkmark & &  \\
+Dependency List &  &  & \checkmark & & \checkmark & & & \checkmark &  &  \\
+Authors &  &  &  &  &  &  & \checkmark & \checkmark & \checkmark & U\\
+Code of Conduct &  &  &  &  & & & \checkmark & & &  \\
+Acknowledgements &  &  &  &  &  &  & \checkmark & \checkmark & \checkmark & U\\
+Code Style Guide &  & \checkmark &  &  & & & \checkmark & \checkmark & \checkmark &\\
+Release Info. &  & \checkmark &  &  & & \checkmark & \checkmark & & & C\\
+Prod.\ Roadmap &  &  &  &  & & \checkmark & \checkmark & \checkmark & & R\\
+\midrule
+Getting started &  &  &  &  & \checkmark & & \checkmark & \checkmark & \checkmark &  \\
+User manual &  &  & \checkmark &  & & & \checkmark & & & C\\
+Tutorials &  &  &  &  & & & \checkmark & & & U\\
+FAQ &  &  &  &  & & & \checkmark & \checkmark & \checkmark & U\\
+\midrule
+Issue Track &  & \checkmark & \checkmark & & \checkmark & \checkmark & \checkmark & & \checkmark & C\\
+Version Control &  & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & C\\ 
+Build Scripts &  & \checkmark &  & \checkmark & \checkmark & \checkmark & \checkmark & & \checkmark & U\\
+\midrule
+Requirements &  & \checkmark &  &  & & \checkmark &  &  & \checkmark &  \\
+Design Doc.\ &  & \checkmark  & \checkmark &  & \checkmark & & \checkmark & \checkmark& \checkmark & \\
+API Doc. &  &  &  &  & \checkmark & & \checkmark & \checkmark & \checkmark & R\\
+Test Plan &  & \checkmark &  &  & & \checkmark & & & &  \\
+Test Cases & \checkmark & \checkmark & \checkmark &  & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & U\\
+\bottomrule
+\end{tabular}
+\caption{Comparison of Recommended Artifacts in Software Development Guidelines
+to Artifacts in MI Projects (C for Common, U for Uncommon and R for Rare)}
+\label{Tbl_Guidelines}
+    \end{center}
+\end{table}
+
+The majority of MI generated artifacts correspond to general recommendations
+from research software developers.  A union of the three columns in
+Table~\ref{artifactspresent} mostly corresponds to recommendations made by the
+research software community.  Most of the items in Table~\ref{artifactspresent}
+are included in guidelines for writing research software, such as the DLR
+(German Aerospace Centre) Software Engineering Guidelines
+\citep{TobiasEtAl2018}, xSDK (Extreme-scale Scientific Software Development Kit)
+Community Package Policies \citep{SmithAndRoscoe2018}, the EURISE (European
+Research Infrastructure Software Engineers') Network Technical Reference
+\cite{ThielEtAl2020} and CLARIAH (Common Lab Research Infrastructure for the
+Arts and Humanities) Guidelines for Software Quality \citep{vanGompelEtAl2016}.
+As an example, MI developers commonly list the external libraries on which the
+software depends, as recommended by \citet{BrettEtAl2021},
+\citet{SmithAndRoscoe2018} and \citet{vanGompelEtAl2016}. Although in the rare
+column, some MI developers include an FAQ (Frequently Asked Questions) document
+as recommended by \citep{OrvizEtAl2017}, \citep{ThielEtAl2020} and
+\citep{vanGompelEtAl2016}. MI developers also follow the ubiquitous
+recommendation to have a README file, although, as for other software, the
+content of README files is not standardized \citep{PranaEtAl2018}.  Compared to
+ocean modelling software \citep{JungEtAl2022}, MI software pays more
+attention to artifacts related to testing.
+
+Three of the items that appear in Table~\ref{artifactspresent} do not explicitly
+appear in the software development guidelines: i) list of related publications,
+ii) performance information and iii) video guides.  The occurrence of the first
+two items in this list are likely more a consequence of how we collected our
+list of MI artifacts, rather than being a unique practice of MI developers. We
+highlighted publications and performance information because we were explicitly
+looking for both as part of filling in our measurement template. Related
+publications and performance information certainly appear in other research
+software projects; the practices just aren't highlighted in guidelines because
+they are implicit in other documentation recommendations, such as theory
+documentation and user guides. The observation of a video guide artifact is more
+novel.  \citet{Fogel2005} recommends videos for open source projects, and other
+projects likely use them, but the practice has apparently not yet made it into
+research software community's guidelines.  Putting aside the debate of the
+effectiveness of learning from print versus video, video documentation is
+growing in popularity, with a majority of Generation Z learners preferring video
+to print \citep{Genota2018}.
+
+Although the MI community participates in most of the practices we found listed
+in the general research software guidelines, some recommended practices were not
+observed or rarely observed. For instance API documentation was rarely observed
+for MI software, but it is a frequently recommended artifact
+\citep{SmithAndRoscoe2018, ThielEtAl2020, vanGompelEtAl2016, OrvizEtAl2017,
+SSI2022, Zadka2018}.  In addition to the items in the last column of
+Table~\ref{artifactspresent}, we can add the following community recommended
+items that we rarely, if ever, observed:
+
+\begin{itemize}
+\item A roadmap showing what is planned for the future \citep{Yo2021,
+ThielEtAl2020, vanGompelEtAl2016}.  We did see this information mentioned in
+some repos, but we did not observe a specific artifact devoted to this purpose.
+\item A code of conduct to explicitly say how developers should treat one
+another \citep{Yo2021, ThielEtAl2020}.  We did not observe a policy of this
+sort, possibly because the number of developers for each MI project is small.
+\item Programming style guidelines so that programming labels and formatting are
+consistent \citep{TobiasEtAl2018, ThielEtAl2020, OrvizEtAl2017, Zadka2018,
+vanGompelEtAl2016, WilsonEtAl2014}.  We saw this information as part of
+some developer guides, but only rarely.
+\item Checklists can be used in projects to ensure that best practices are
+followed by all developers.  Some examples include checklists merging branches
+into master \citep{Brown2015}, checklists for saving and sharing changes to the
+project \citep{WilsonEtAl2016}, checklists for new and departing team members
+\citep{HerouxAndBernholdt2018}, checklists for processes related to commits and
+releases \citep{HerouxEtAl2008} and checklists for overall software quality
+\citep{ThielEtAl2020, SSI2022}.  For MI software, ESPResSo has a checklist for
+managing releases, but otherwise they are used only rarely for MI.
+\item Uninstall instructions \citep{vanGompelEtAl2016} were not observed for any
+of the MI projects.
+\end{itemize}
+
+The above discussion shows that, taken together, MI projects fall short of
+recommended best practices for research software.  However, MI software is not
+alone in this.  Many, if not most, research projects fall short of best
+practices.  Software requirements documentation provides a good example of the
+common deviation between recommendations and practice.  Although requirements
+documentation is recommended by some \citep{TobiasEtAl2018, HerouxEtAl2008,
+SmithAndKoothoor2016}, in practice research software developers often do not
+produce a proper requirements specification \citep{HeatonAndCarver2015}.
+\citet{SandersAndKelly2008} interviewed 16 scientists from 10 disciplines and
+found that none of the scientists created requirements specifications, unless
+regulations in their field mandated such a document. \citet{Nguyen-HoanEtAl2010}
+showed requirements are the least commonly produced type of documentation for
+research software in general. When looking at the pain points for research
+software developers, \citet{WieseEtAl2019} found that software requirements and
+management is the software engineering discipline that most hurts scientific
+developers, accounting for 23\% of the technical problems reported by study
+participants.  The lack of support for requirements is likely due to the
+perception that up-front requirements are impossible for research software
+\citep{CarverEtAl2007,SegalAndMorris2008}, but when the instance on ``up-front''
+requirements is dropped, allowing the requirements to be written iteratively and
+incrementally, requirements are feasible \citep{Smith2016}.
 
@@
 and evolution of software, software processes, and artifacts \citep[p.\
-501]{GhezziEtAl2003}. Many tools are used by LBM software packages, as
-summarized in Table~?.  The tools are subdivided into development
-tools, dependencies, and project management tools.  In this section we answer
-\rqref{RQ_CompareToolsProjMngmnt} by comparing aspects of tool usage in LBM
-software packages to their utilization in the research software community in
-general.
+501]{GhezziEtAl2003}. Many tools are used by MI software packages, as summarized
+in Table~?.  The tools are subdivided into development tools, dependencies, and
+project management tools.  We answer \rqref{RQ_CompareToolsProjMngmnt} by
+comparing aspects of tool usage in MI software packages to their utilization in
+the research software community in general.
 
@@
 This section answers research question \rqref{RQ_CompareMethodologies} by
-comparing the principles, processes and methodologies used for LBM software to
+comparing the principles, processes and methodologies used for MI software to
 what can be gleaned from the literature on research software in general. In our
-data collection for LBM software, the software development process is not
+data collection for MI software, the software development process is not
 explicitly indicated in the artifacts for most of the packages. However, during
@@
 model is like a combination of agile and waterfall. Employing a loosely defined
-process makes sense for LBM software, given that the teams are generally small
+process makes sense for MI software, given that the teams are generally small
 and self-contained.  Although eleven of the packages explicitly convey that they
@@
 research software projects (\rqref{RQ_PainPoints})?; and, ii) How do the pain
-points of developers from LBM compare to the pain points for research software
+points of developers from MI compare to the pain points for research software
 in general (\rqref{RQ_ComparePainPoints})?  Below we go through each of the
-identified pain points and include citations that contrast the LBM experience
+identified pain points and include citations that contrast the MI experience
 with observations from researchers in other domains.  Potential ways to address
@@
 
-This section summarizes the best practices from LBM developers that are taken to
-address the pain points mentioned in Section~\ref{painpoints}.  The main source
-of information is the qualitative data from developer interviews
-(Section~\ref{sec_interview_methods}).  The practices summarized in this section
-can potentially be emulated by the LBM software packages that do not currently
-follow them.  Moreover, these practices may also provide examples that can be
-followed by other research software domains.
+The best practices from MI developers are taken to address the pain points
+mentioned in Section~\ref{painpoints}.  We found the practices as part of the
+qualitative data from developer interviews
+(Section~\ref{sec_interview_methods}).  The practices we summarize can
+potentially be emulated by the MI software packages that do not currently follow
+them.  Moreover, these practices may also provide examples that can be followed
+by other research software domains.
 
@@
 and manage the project. Section~\ref{painpoints} lists the advantages and
-disadvantages between native and web applications. In this section, we give
-further suggestions on the choice of a tech stack to improve the four priority
-factors identified by developers: compatibility, maintainability, performance
-and security.  The suggestions are intended to provide ideas and avenues for
-exploration; not all of the suggestions will be the right fit for all projects
-and all teams.
+disadvantages between native and web applications. We give further suggestions
+on the choice of a tech stack to improve the four priority factors identified by
+developers: compatibility, maintainability, performance and security.  The
+suggestions are intended to provide ideas and avenues for exploration; not all
+of the suggestions will be the right fit for all projects and all teams.
 

### 2022-04-01 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Updates to SOP for LBM noticed while working on SOP for MI
Updates to SOP for LBM

@@
 for the presence of artifacts, which were then categorized by frequency. We have
-grouped them into common, less common, and rare artifacts in
+grouped them into common, uncommon, and rare artifacts in
 Table~\ref{artifactspresent}. Common artifacts were found in 16 to 24 ($>$63\%)
-of the software packages. Less common artifacts were found in 8 to 15 (30-63\%)
-of the software packages. Rare artifacts were found in 1 to 7 ($<$30\%) of the
-software packages.  In this section we answer \rqref{RQ_CompareArtifacts} by
-comparing the artifacts that we observed in LBM repositories to those observed
-and recommended for research software in general. Our comparison may point out
-areas where some LBM software packages fall short of current best practices.
-This is not intended to be a criticism of any existing packages, especially
-since in practice not every project needs to achieve the highest possible
-quality.  However, rather than delve into the nuances of which software can
-justify compromising which practices we will write our comparison under the
-ideal assumption that every project has sufficient resources to match best
-practices.
+of the software packages. Uncommon artifacts were found in 8 to 15 (30-63\%) of
+the software packages. Rare artifacts were found in 1 to 7 ($<$30\%) of the
+software packages.  We answer \rqref{RQ_CompareArtifacts} by comparing the
+artifacts that we observed in LBM repositories to those observed and recommended
+for research software in general. Our comparison may point out areas where some
+LBM software packages fall short of current best practices. This is not intended
+to be a criticism of any existing packages, especially since in practice not
+every project needs to achieve the highest possible quality.  However, rather
+than delve into the nuances of which software can justify compromising which
+practices we will write our comparison under the ideal assumption that every
+project has sufficient resources to match best practices.
 
@@
 \toprule
-\textbf{Common} & \textbf{Less Common} & \textbf{Rare} \\
+\textbf{Common} & \textbf{Uncommon} & \textbf{Rare} \\
 \midrule
@@
 (ESPREsSo, Ludwig, Palabos) of the packages appear to use a version control
-system.  The top four packages also have most of the less common artifacts. At
+system.  The top four packages also have most of the uncommon artifacts. At
 the time of data collection, only one of the four packages (Palabos) did not

### 2022-04-01 StateOfPractice/Papers/MedImageSoft_SOP-arXiv/MedImageSoft_SOP_AcademicRadiology.tex
Updates to SOP for MI for the artifacts section

@@
 As part of filling in the measurement template the MI packages were examined for
-the presence of artifacts, which were then categorized by frequency. We have
-grouped them into common, less common, and rare artifacts in
-Table~\ref{artifactspresent}. Common artifacts were found in 20 to 29 ($>$67\%)
-of the software packages. Uncommon and rare artifacts were found in
-10 to 19 (33-67\%) and 1 to 9 ($<$33\%) of the software packages, respectively.
-The full measurements are summarized in \citet{Dong2021-Data}.  The details on
-which projects use which types of artifacts is summarized in
-Tables~\ref{tab_maintainability_docs} and~\ref{tab_Visibility/Transparency_docs}
-for documents related to maintainability and visibility, respectively.
+the presence of artifacts, which were then categorized by frequency.
+Table~\ref{artifactspresent} groups them into categories of common (20 to 29
+($>$67\%) packages), uncommon (10 to 19 (33-67\%) packages), and rare (1 to 9
+($<$33\%) packages). The full measurements are summarized in
+\citet{Dong2021-Data}.  The details on which projects use which types of
+artifacts are summarized in Tables~\ref{tab_maintainability_docs}
+and~\ref{tab_Visibility/Transparency_docs} for documents related to
+maintainability and visibility, respectively.
 
@@
     \midrule
-    README (29) & Build scripts (18) & Developer's manual (8)\\
-    Version control (29) & Tutorials (18) & API documentation (7)\\
-    License (28) & Installation guide (16) & Troubleshooting guide (6)\\
-    Issue tracker (28) & Test cases (15) & Product roadmap (5)\\
-    Change request (28) & Authors (14) & \\
-    User manual (22) & Frequently Asked Questions (FAQ) (14) & \\
-    Release info. (22) & Acknowledgements (12) & \\
+    README (29) & Build scripts (18) & Getting Started (9)\\
+    Version control (29) & Tutorials (18) & Developer's manual (8)\\
+    License (28) & Installation guide (16) & Contributing (8)\\
+    Issue tracker (28) & Test cases (15) & API documentation (7)\\
+    Change request (28) & Authors (14) & Dependency List (7)\\
+    User manual (22) & Frequently Asked Questions (FAQ) (14) & Troubleshooting guide (6)\\
+    Release info. (22) & Acknowledgements (12) & Product roadmap (5)\\
+     & Changelog (12) & Design Documentation (5)\\
+     & Citation (11) & Code Style Guide (3)\\
+     & & Code of Conduct (1)\\
     \bottomrule
@@
 
-Table~\ref{artifactspresent} shows that some types of software engineering
-recommended artifacts are missing in MI projects.  
-  A gap exists in scientific
-computing development practices and software engineering recommendations
-\citep{OwojaiyeEtAl2021_CSE}.
+Table~\ref{Tbl_Guidelines} shows that the MI artifacts generally match the recommendations found in nine current research software development guidelines:
+\begin{inparaenum}[i)]
+\item United States Geological Survey Software Planning Checklist
+\citep{USGS2019},
+\item DLR (German Aerospace Centre) Software Engineering Guidelines
+\citep{TobiasEtAl2018}, 
+\item Scottish Covid-19 Response Consortium Software Checklist
+\citep{BrettEtAl2021},
+\item Good Enough Practices in Scientific Computing \citep{WilsonEtAl2016},
+\item xSDK (Extreme-scale Scientific Software Development Kit) Community Package
+Policies \citep{SmithAndRoscoe2018},
+\item Trilinos Developers Guide \citep{HerouxEtAl2008},
+\item EURISE (European Research Infrastructure Software Engineers') Network
+Technical Reference \citep{ThielEtAl2020},
+\item CLARIAH (Common Lab Research Infrastructure for the Arts and Humanities)
+Guidelines for Software Quality \citep{vanGompelEtAl2016}, and
+\item A Set of Common Software Quality Assurance Baseline Criteria for Research
+Projects \citep{OrvizEtAl2017}.
+\end{inparaenum}
+As examples of compliance with most guidelines, MI developers commonly include
+the almost ubiquitous advice to have LICENSE files and README files. , although,
+as for other software, the content of README files is not standardized
+\citep{PranaEtAl2018}. 15 of 26 artifacts (58\%).  Although in the rare column, some MI developers include
+an FAQ (Frequently Asked Questions) document as recommended by
+\citep{OrvizEtAl2017}, \citep{ThielEtAl2020} and \citep{vanGompelEtAl2016}. MI
+developers also follow the ubiquitous recommendation to have a README file
+Compared to ocean modelling software \citep{JungEtAl2022}, MI software pays more
+attention to artifacts related to testing.
 
@@
 README &  & \checkmark & \checkmark & \checkmark & \checkmark & & \checkmark & \checkmark & \checkmark & C\\
-CONTRIBUTING &  & \checkmark & \checkmark & \checkmark & \checkmark & & \checkmark & \checkmark & \checkmark \\
-CITATION &  &  &  & \checkmark & & & & \checkmark & \checkmark &  \\
-CHANGELOG &  & \checkmark &  & \checkmark & \checkmark & & \checkmark &  &  &  \\
+CONTRIBUTING &  & \checkmark & \checkmark & \checkmark & \checkmark & & \checkmark & \checkmark & \checkmark & R\\
+CITATION &  &  &  & \checkmark & & & & \checkmark & \checkmark & U\\
+CHANGELOG &  & \checkmark &  & \checkmark & \checkmark & & \checkmark &  &  & U\\
 INSTALL &  &  &  &  & \checkmark & & \checkmark & \checkmark & \checkmark & U\\
@@
 Uninstall &  &  &  &  & & & & \checkmark & &  \\
-Dependency List &  &  & \checkmark & & \checkmark & & & \checkmark &  &  \\
+Dependency List &  &  & \checkmark & & \checkmark & & & \checkmark &  & R\\
 Authors &  &  &  &  &  &  & \checkmark & \checkmark & \checkmark & U\\
-Code of Conduct &  &  &  &  & & & \checkmark & & &  \\
+Code of Conduct &  &  &  &  & & & \checkmark & & & R\\
 Acknowledgements &  &  &  &  &  &  & \checkmark & \checkmark & \checkmark & U\\
-Code Style Guide &  & \checkmark &  &  & & & \checkmark & \checkmark & \checkmark &\\
+Code Style Guide &  & \checkmark &  &  & & & \checkmark & \checkmark & \checkmark & R\\
 Release Info. &  & \checkmark &  &  & & \checkmark & \checkmark & & & C\\
@@
 \midrule
-Getting started &  &  &  &  & \checkmark & & \checkmark & \checkmark & \checkmark &  \\
+Getting started &  &  &  &  & \checkmark & & \checkmark & \checkmark & \checkmark & R\\
 User manual &  &  & \checkmark &  & & & \checkmark & & & C\\
@@
 Requirements &  & \checkmark &  &  & & \checkmark &  &  & \checkmark &  \\
-Design Doc.\ &  & \checkmark  & \checkmark &  & \checkmark & & \checkmark & \checkmark& \checkmark & \\
+Design Doc.\ &  & \checkmark  & \checkmark &  & \checkmark & & \checkmark & \checkmark& \checkmark & R\\
 API Doc. &  &  &  &  & \checkmark & & \checkmark & \checkmark & \checkmark & R\\
@@
 
-The majority of MI generated artifacts correspond to general recommendations
-from research software developers.  A union of the three columns in
-Table~\ref{artifactspresent} mostly corresponds to recommendations made by the
-research software community.  Most of the items in Table~\ref{artifactspresent}
-are included in guidelines for writing research software, such as the DLR
-(German Aerospace Centre) Software Engineering Guidelines
-\citep{TobiasEtAl2018}, xSDK (Extreme-scale Scientific Software Development Kit)
-Community Package Policies \citep{SmithAndRoscoe2018}, the EURISE (European
-Research Infrastructure Software Engineers') Network Technical Reference
-\cite{ThielEtAl2020} and CLARIAH (Common Lab Research Infrastructure for the
-Arts and Humanities) Guidelines for Software Quality \citep{vanGompelEtAl2016}.
-As an example, MI developers commonly list the external libraries on which the
-software depends, as recommended by \citet{BrettEtAl2021},
-\citet{SmithAndRoscoe2018} and \citet{vanGompelEtAl2016}. Although in the rare
-column, some MI developers include an FAQ (Frequently Asked Questions) document
-as recommended by \citep{OrvizEtAl2017}, \citep{ThielEtAl2020} and
-\citep{vanGompelEtAl2016}. MI developers also follow the ubiquitous
-recommendation to have a README file, although, as for other software, the
-content of README files is not standardized \citep{PranaEtAl2018}.  Compared to
-ocean modelling software \citep{JungEtAl2022}, MI software pays more
-attention to artifacts related to testing.
+A gap exists in scientific
+computing development practices and software engineering recommendations
+\citep{OwojaiyeEtAl2021_CSE}.
 

### 2022-04-04 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Addition of a table mapping artifacts to guidelines and LBM software

@@
 
+Table~\ref{Tbl_Guidelines} (based on data from \citep{SmithAndMichalski2022})
+shows that MI artifacts generally match the recommendations found in nine
+current research software development guidelines:
+\begin{itemize}
+\item United States Geological Survey Software Planning Checklist
+\citep{USGS2019},
+\item DLR (German Aerospace Centre) Software Engineering Guidelines
+\citep{TobiasEtAl2018}, 
+\item Scottish Covid-19 Response Consortium Software Checklist
+\citep{BrettEtAl2021},
+\item Good Enough Practices in Scientific Computing \citep{WilsonEtAl2016},
+\item xSDK (Extreme-scale Scientific Software Development Kit) Community Package
+Policies \citep{SmithAndRoscoe2018},
+\item Trilinos Developers Guide \citep{HerouxEtAl2008},
+\item EURISE (European Research Infrastructure Software Engineers') Network
+Technical Reference \citep{ThielEtAl2020},
+\item CLARIAH (Common Lab Research Infrastructure for the Arts and Humanities)
+Guidelines for Software Quality \citep{vanGompelEtAl2016}, and
+\item A Set of Common Software Quality Assurance Baseline Criteria for Research
+Projects \citep{OrvizEtAl2017}.
+\end{itemize}
+
+In Table~\ref{Tbl_Guidelines} each row corresponds to an artifact.  For a given
+row, a checkmark in one of the columns means that the corresponding guideline
+recommends this artifact.  The last column shows whether the artifact appears in
+the measured set of MI software, either not at all (blank), commonly (C),
+uncommonly (U) or rarely (R).  We did our best to interpret the meaning of each
+artifact consistently between guidelines and specific MI software, but the
+terminology and the contents of artifacts are not standardized.  The challenge
+even exists for the ubiquitous README file.  As illustrated by
+\citet{PranaEtAl2018}, the content of README files shows significant variation
+between projects.  Although some content is reasonably consistent, with 97\% of
+README files contain at least one section describing the `What` of the
+repository and 88.5 \% offering some `How` content, other categories are more
+variable.  For instance, information on `Contribution`, `Why`', and `Who`,
+appear in 27.8\%, 25.7\% and 52.9\% of the analyzed files, respectively
+\citep{PranaEtAl2018}.
+
+The frequency of checkmarks in Table~\ref{Tbl_Guidelines} indicates the
+popularity of recommending a given artifact, but it does not imply that the most
+commonly recommended artifacts are the most important artifacts. Just because an
+artifact is not explicitly recommended in a given guidelines, does not mean that
+the artifact is not valued by the guideline authors.  They may have exluded it
+because it is out of the scope of their recommendations, or outside of their
+experience.  For instance, an artifact related to uninstall is only explicitly
+mentioned by \citep{vanGompelEtAl2016}), but other guideline authors would
+likely see its value.  They may simply feel that uninstall is implied by
+install, or they may have never asked themselves whether separate uninstall
+instructions are needed.
+
+\begin{table}[!ht]
+\begin{center}
+\begin{tabular}{ p{2.5cm}p{1cm}p{1cm}p{1cm}p{1cm}p{1cm}p{1cm}p{1cm}p{1.2cm}p{1cm}p{0.8cm} }
+\toprule
+~ \ & \cite{USGS2019} & \cite{TobiasEtAl2018} & \cite{BrettEtAl2021} &
+\cite{WilsonEtAl2016} & \cite{SmithAndRoscoe2018} & \cite{HerouxEtAl2008} &
+\cite{ThielEtAl2020} & \cite{vanGompelEtAl2016} & \cite{OrvizEtAl2017} & LBM\\
+\midrule
+LICENSE & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & &
+\checkmark & \checkmark & \checkmark & C\\
+README &  & \checkmark & \checkmark & \checkmark & \checkmark & & \checkmark &
+\checkmark & \checkmark & C\\
+CONTRIBUTING &  & \checkmark & \checkmark & \checkmark & \checkmark & &
+\checkmark & \checkmark & \checkmark & R\\
+CITATION &  &  &  & \checkmark & & & & \checkmark & \checkmark & U\\
+CHANGELOG &  & \checkmark &  & \checkmark & \checkmark & & \checkmark &  &  & U\\
+INSTALL &  &  &  &  & \checkmark & & \checkmark & \checkmark & \checkmark & U\\
+\midrule
+Uninstall &  &  &  &  & & & & \checkmark & &  \\
+Dependency List &  &  & \checkmark & & \checkmark & & & \checkmark &  & R\\
+Authors &  &  &  &  &  &  & \checkmark & \checkmark & \checkmark & U\\
+Code of Conduct &  &  &  &  & & & \checkmark & & & R\\
+Acknowledgements &  &  &  &  &  &  & \checkmark & \checkmark & \checkmark & U\\
+Code Style Guide &  & \checkmark &  &  & & & \checkmark & \checkmark & \checkmark & R\\
+Release Info. &  & \checkmark &  &  & & \checkmark & \checkmark & & & C\\
+Prod.\ Roadmap &  &  &  &  & & \checkmark & \checkmark & \checkmark & & R\\
+\midrule
+Getting started &  &  &  &  & \checkmark & & \checkmark & \checkmark & \checkmark & R\\
+User manual &  &  & \checkmark &  & & & \checkmark & & & C\\
+Tutorials &  &  &  &  & & & \checkmark & & & U\\
+FAQ &  &  &  &  & & & \checkmark & \checkmark & \checkmark & U\\
+\midrule
+Issue Track &  & \checkmark & \checkmark & & \checkmark & \checkmark &
+\checkmark & & \checkmark & C\\
+Version Control &  & \checkmark & \checkmark & \checkmark & \checkmark &
+\checkmark & \checkmark & \checkmark & \checkmark & C\\ 
+Build Scripts &  & \checkmark &  & \checkmark & \checkmark & \checkmark &
+\checkmark & & \checkmark & U\\
+\midrule
+Requirements &  & \checkmark &  &  & & \checkmark &  &  & \checkmark &  \\
+Design Doc.\ &  & \checkmark  & \checkmark &  & \checkmark & & \checkmark &
+\checkmark& \checkmark & R\\
+API Doc. &  &  &  &  & \checkmark & & \checkmark & \checkmark & \checkmark & R\\
+Test Plan &  & \checkmark &  &  & & \checkmark & & & &  \\
+Test Cases & \checkmark & \checkmark & \checkmark &  & \checkmark & \checkmark &
+\checkmark & \checkmark & \checkmark & U\\
+\bottomrule
+\end{tabular}
+\caption{Comparison of Recommended Artifacts in Software Development Guidelines
+to Artifacts in MI Projects (C for Common, U for Uncommon and R for Rare)}
+\label{Tbl_Guidelines}
+\end{center}
+\end{table}
+
 The top four packages, ESPREsSo, Ludwig, Palabos, and OpenLB, have all of the

### 2022-04-04 StateOfPractice/Papers/MedImageSoft_SOP-arXiv/MedImageSoft_SOP_AcademicRadiology.tex
Complete draft of edits to Section 6 (Comparison of Artifacts)

@@
 
-As part of filling in the measurement template the MI packages were examined for
-the presence of artifacts, which were then categorized by frequency.
-Table~\ref{artifactspresent} groups them into categories of common (20 to 29
-($>$67\%) packages), uncommon (10 to 19 (33-67\%) packages), and rare (1 to 9
-($<$33\%) packages). The full measurements are summarized in
-\citet{Dong2021-Data}.  The details on which projects use which types of
-artifacts are summarized in Tables~\ref{tab_maintainability_docs}
+As part of filling in the measurement template (from
+Section~\ref{sec_grading_software}) we summarized the artifacts observed in each
+MI package. Table~\ref{artifactspresent} groups the artifacts by frequency into
+categories of common (20 to 29 ($>$67\%) packages), uncommon (10 to 19 (33-67\%)
+packages), and rare (1 to 9 ($<$33\%) packages). The full measurements are
+summarized in \citet{Dong2021-Data}.  The details on which projects use which
+types of artifacts are summarized in Tables~\ref{tab_maintainability_docs}
 and~\ref{tab_Visibility/Transparency_docs} for documents related to
@@
     Issue tracker (28) & Test cases (15) & API documentation (7)\\
-    Change request (28) & Authors (14) & Dependency List (7)\\
-    User manual (22) & Frequently Asked Questions (FAQ) (14) & Troubleshooting guide (6)\\
-    Release info. (22) & Acknowledgements (12) & Product roadmap (5)\\
-     & Changelog (12) & Design Documentation (5)\\
-     & Citation (11) & Code Style Guide (3)\\
-     & & Code of Conduct (1)\\
+    User manual (22) & Authors (14) & Dependency list (7)\\
+    Release info. (22) & Frequently Asked Questions (FAQ) (14) & Troubleshooting guide (6)\\
+     & Acknowledgements (12) & Product roadmap (5)\\
+     & Changelog (12) & Design documentation (5)\\
+     & Citation (11) & Code style guide (3)\\
+     & & Code of conduct (1)\\
     \bottomrule
     \end{tabular}
-    \caption{Artifacts Present in MI Packages, Classified by Frequency (The number in brackets is the number of occurrences)}
+    \caption{Artifacts Present in MI Packages, Classified by Frequency (The number 
+    in brackets is the number of occurrences)}
     \label{artifactspresent}
@@
 
-We answer \rqref{RQ_CompareArtifacts} by comparing the artifacts that we
+We answer~\rqref{RQ_CompareArtifacts} by comparing the artifacts that we
 observed in MI repositories to those observed and recommended for research
@@
 
-Table~\ref{Tbl_Guidelines} shows that the MI artifacts generally match the recommendations found in nine current research software development guidelines:
-\begin{inparaenum}[i)]
+Table~\ref{Tbl_Guidelines} (based on data from \citep{SmithAndMichalski2022})
+shows that MI artifacts generally match the recommendations found in nine
+current research software development guidelines:
+\begin{itemize}
 \item United States Geological Survey Software Planning Checklist
@@
 Projects \citep{OrvizEtAl2017}.
-\end{inparaenum}
-As examples of compliance with most guidelines, MI developers commonly include
-the almost ubiquitous advice to have LICENSE files and README files. , although,
-as for other software, the content of README files is not standardized
-\citep{PranaEtAl2018}. 15 of 26 artifacts (58\%).  Although in the rare column, some MI developers include
-an FAQ (Frequently Asked Questions) document as recommended by
-\citep{OrvizEtAl2017}, \citep{ThielEtAl2020} and \citep{vanGompelEtAl2016}. MI
-developers also follow the ubiquitous recommendation to have a README file
-Compared to ocean modelling software \citep{JungEtAl2022}, MI software pays more
-attention to artifacts related to testing.
+\end{itemize}
+
+In Table~\ref{Tbl_Guidelines} each row corresponds to an artifact.  For a given
+row, a checkmark in one of the columns means that the corresponding guideline
+recommends this artifact.  The last column shows whether the artifact appears in
+the measured set of MI software, either not at all (blank), commonly (C),
+uncommonly (U) or rarely (R).  We did our best to interpret the meaning of each
+artifact consistently between guidelines and specific MI software, but the
+terminology and the contents of artifacts are not standardized.  The challenge
+even exists for the ubiquitous README file.  As illustrated by
+\citet{PranaEtAl2018}, the content of README files shows significant variation
+between projects.  Although some content is reasonably consistent, with 97\% of
+README files contain at least one section describing the `What` of the
+repository and 89\% offering some `How` content, other categories are more
+variable.  For instance, information on `Contribution`, `Why`', and `Who`,
+appear in 28\%, 26\% and 53\% of the analyzed files, respectively
+\citep{PranaEtAl2018}.
+
+The frequency of checkmarks in Table~\ref{Tbl_Guidelines} indicates the
+popularity of recommending a given artifact, but it does not imply that the most
+commonly recommended artifacts are the most important artifacts. Just because an
+artifact is not explicitly recommended in a given guidelines, does not mean that
+the artifact is not valued by the guideline authors.  They may have exluded it
+because it is out of the scope of their recommendations, or outside of their
+experience.  For instance, an artifact related to uninstall is only explicitly
+mentioned by \citep{vanGompelEtAl2016}), but other guideline authors would
+likely see its value.  They may simply feel that uninstall is implied by
+install, or they may have never asked themselves whether separate uninstall
+instructions are needed.
 
@@
 \toprule
-~ \ & \cite{USGS2019} & \cite{TobiasEtAl2018} & \cite{BrettEtAl2021} & \cite{WilsonEtAl2016} & \cite{SmithAndRoscoe2018} & \cite{HerouxEtAl2008} & \cite{ThielEtAl2020} & \cite{vanGompelEtAl2016} & \cite{OrvizEtAl2017} & MI\\
+~ \ & \cite{USGS2019} & \cite{TobiasEtAl2018} & \cite{BrettEtAl2021} &
+\cite{WilsonEtAl2016} & \cite{SmithAndRoscoe2018} & \cite{HerouxEtAl2008} &
+\cite{ThielEtAl2020} & \cite{vanGompelEtAl2016} & \cite{OrvizEtAl2017} & MI\\
 \midrule
-LICENSE & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & & \checkmark & \checkmark & \checkmark & C\\
-README &  & \checkmark & \checkmark & \checkmark & \checkmark & & \checkmark & \checkmark & \checkmark & C\\
-CONTRIBUTING &  & \checkmark & \checkmark & \checkmark & \checkmark & & \checkmark & \checkmark & \checkmark & R\\
+LICENSE & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & &
+\checkmark & \checkmark & \checkmark & C\\
+README &  & \checkmark & \checkmark & \checkmark & \checkmark & & \checkmark &
+\checkmark & \checkmark & C\\
+CONTRIBUTING &  & \checkmark & \checkmark & \checkmark & \checkmark & &
+\checkmark & \checkmark & \checkmark & R\\
 CITATION &  &  &  & \checkmark & & & & \checkmark & \checkmark & U\\
@@
 \midrule
-Issue Track &  & \checkmark & \checkmark & & \checkmark & \checkmark & \checkmark & & \checkmark & C\\
-Version Control &  & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & C\\ 
-Build Scripts &  & \checkmark &  & \checkmark & \checkmark & \checkmark & \checkmark & & \checkmark & U\\
+Issue Track &  & \checkmark & \checkmark & & \checkmark & \checkmark &
+\checkmark & & \checkmark & C\\
+Version Control &  & \checkmark & \checkmark & \checkmark & \checkmark &
+\checkmark & \checkmark & \checkmark & \checkmark & C\\ 
+Build Scripts &  & \checkmark &  & \checkmark & \checkmark & \checkmark &
+\checkmark & & \checkmark & U\\
 \midrule
 Requirements &  & \checkmark &  &  & & \checkmark &  &  & \checkmark &  \\
-Design Doc.\ &  & \checkmark  & \checkmark &  & \checkmark & & \checkmark & \checkmark& \checkmark & R\\
+Design Doc.\ &  & \checkmark  & \checkmark &  & \checkmark & & \checkmark &
+\checkmark& \checkmark & R\\
 API Doc. &  &  &  &  & \checkmark & & \checkmark & \checkmark & \checkmark & R\\
 Test Plan &  & \checkmark &  &  & & \checkmark & & & &  \\
-Test Cases & \checkmark & \checkmark & \checkmark &  & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & U\\
+Test Cases & \checkmark & \checkmark & \checkmark &  & \checkmark & \checkmark &
+\checkmark & \checkmark & \checkmark & U\\
 \bottomrule
@@
 
-A gap exists in scientific
-computing development practices and software engineering recommendations
-\citep{OwojaiyeEtAl2021_CSE}.
-
-Three of the items that appear in Table~\ref{artifactspresent} do not explicitly
-appear in the software development guidelines: i) list of related publications,
-ii) performance information and iii) video guides.  The occurrence of the first
-two items in this list are likely more a consequence of how we collected our
-list of MI artifacts, rather than being a unique practice of MI developers. We
-highlighted publications and performance information because we were explicitly
-looking for both as part of filling in our measurement template. Related
-publications and performance information certainly appear in other research
-software projects; the practices just aren't highlighted in guidelines because
-they are implicit in other documentation recommendations, such as theory
-documentation and user guides. The observation of a video guide artifact is more
-novel.  \citet{Fogel2005} recommends videos for open source projects, and other
-projects likely use them, but the practice has apparently not yet made it into
-research software community's guidelines.  Putting aside the debate of the
-effectiveness of learning from print versus video, video documentation is
-growing in popularity, with a majority of Generation Z learners preferring video
-to print \citep{Genota2018}.
-
-Although the MI community participates in most of the practices we found listed
-in the general research software guidelines, some recommended practices were not
-observed or rarely observed. For instance API documentation was rarely observed
-for MI software, but it is a frequently recommended artifact
-\citep{SmithAndRoscoe2018, ThielEtAl2020, vanGompelEtAl2016, OrvizEtAl2017,
-SSI2022, Zadka2018}.  In addition to the items in the last column of
-Table~\ref{artifactspresent}, we can add the following community recommended
-items that we rarely, if ever, observed:
-
-\begin{itemize}
-\item A roadmap showing what is planned for the future \citep{Yo2021,
-ThielEtAl2020, vanGompelEtAl2016}.  We did see this information mentioned in
-some repos, but we did not observe a specific artifact devoted to this purpose.
-\item A code of conduct to explicitly say how developers should treat one
-another \citep{Yo2021, ThielEtAl2020}.  We did not observe a policy of this
-sort, possibly because the number of developers for each MI project is small.
-\item Programming style guidelines so that programming labels and formatting are
-consistent \citep{TobiasEtAl2018, ThielEtAl2020, OrvizEtAl2017, Zadka2018,
-vanGompelEtAl2016, WilsonEtAl2014}.  We saw this information as part of
-some developer guides, but only rarely.
-\item Checklists can be used in projects to ensure that best practices are
-followed by all developers.  Some examples include checklists merging branches
-into master \citep{Brown2015}, checklists for saving and sharing changes to the
-project \citep{WilsonEtAl2016}, checklists for new and departing team members
-\citep{HerouxAndBernholdt2018}, checklists for processes related to commits and
-releases \citep{HerouxEtAl2008} and checklists for overall software quality
-\citep{ThielEtAl2020, SSI2022}.  For MI software, ESPResSo has a checklist for
-managing releases, but otherwise they are used only rarely for MI.
-\item Uninstall instructions \citep{vanGompelEtAl2016} were not observed for any
-of the MI projects.
-\end{itemize}
-
-The above discussion shows that, taken together, MI projects fall short of
-recommended best practices for research software.  However, MI software is not
-alone in this.  Many, if not most, research projects fall short of best
-practices.  Software requirements documentation provides a good example of the
-common deviation between recommendations and practice.  Although requirements
-documentation is recommended by some \citep{TobiasEtAl2018, HerouxEtAl2008,
-SmithAndKoothoor2016}, in practice research software developers often do not
-produce a proper requirements specification \citep{HeatonAndCarver2015}.
-\citet{SandersAndKelly2008} interviewed 16 scientists from 10 disciplines and
-found that none of the scientists created requirements specifications, unless
-regulations in their field mandated such a document. \citet{Nguyen-HoanEtAl2010}
-showed requirements are the least commonly produced type of documentation for
-research software in general. When looking at the pain points for research
-software developers, \citet{WieseEtAl2019} found that software requirements and
-management is the software engineering discipline that most hurts scientific
-developers, accounting for 23\% of the technical problems reported by study
-participants.  The lack of support for requirements is likely due to the
-perception that up-front requirements are impossible for research software
+Two of the items that appear in Table~\ref{artifactspresent} do not appear in
+the software development guidelines shown in Table~\ref{Tbl_Guidelines}:
+Troubleshooting guide and Developer's manual.  Although these two artifacts
+aren't specifically named in the guidelines that we found, the information
+contained within them overlaps with the recommended artifacts.  A
+Troubleshooting guideline contains information that would typically be found in
+a User manual.  A Developer's guide overlaps with information from the README,
+INSTALL, Uninstall, Dependency List, Release Information, API documentation and
+Design documentation.  In our current analysis, we have identified artifacts by
+the names given by the software guidelines and MI examples.  In the future, a
+more in-depth analysis would look at the knowledge fragments that are captured
+in the artifacts, rather than focusing on the names of the files that collect
+these fragments together.
+
+Although the MI community shows examples of 88\% (23 of 26) of the practices we
+found in research software guidelines (Table~\ref{Tbl_Guidelines}), three
+recommended artifacts were not observed: i) Unistall, ii) Test plans, and iii)
+Requirements.  Uninstall is likely an omission caused by the focus on installing
+software. Given the storage capacity of current hardware, developers and users
+do not generally concerned with uninstall. Moreover, as mentioned above,
+uninstall is not particularly emphasized in existing recommendations.  Test
+plans were not observed for MI software, but that doesn't mean they weren't
+created; it means that the plans are not under version control.  Test plans
+would have to at least be implicitly created, since test cases were observed
+with reasonable frequency for MI software (uncommon).
+
+MI software is like other research software in its neglect of requirements
+documentation.  Although requirements documentation is recommended by some
+\citep{TobiasEtAl2018, HerouxEtAl2008, SmithAndKoothoor2016}, in practice
+research software developers often do not produce a proper requirements
+specification \citep{HeatonAndCarver2015}. \citet{SandersAndKelly2008}
+interviewed 16 scientists from 10 disciplines and found that none of the
+scientists created requirements specifications, unless regulations in their
+field mandated such a document. \citet{Nguyen-HoanEtAl2010} showed requirements
+are the least commonly produced type of documentation for research software in
+general. When looking at the pain points for research software developers,
+\citet{WieseEtAl2019} found that software requirements and management is the
+software engineering discipline that most hurts scientific developers,
+accounting for 23\% of the technical problems reported by study participants.
+The lack of support for requirements is likely due to the perception that
+up-front requirements are impossible for research software
 \citep{CarverEtAl2007,SegalAndMorris2008}, but when the instance on ``up-front''
@@
 incrementally, requirements are feasible \citep{Smith2016}.
+
+Table~\ref{Tbl_Guidelines} shows several artifacts that are rarely observed in
+practice.  A theme amount these rare artifacts is that many of them are related
+to developers more than users.  For instance, the rare artifacts include
+Contributing, developer Code of conduct, Code Style Guide, Product roadmaps,
+Requirements, Design documentation, API documentation and a Test plan.  The rare
+artifacts for MI software are similar to the rare artifacts for Lattice
+Boltzmann solvers \citep{Michalski2021}.  However, MI software differs from
+ocean modelling software, since, according to \citep{JungEtAl2022}, the later
+pays relatively less attention to testing.
+
+To improve MI software in the future, an increased use of checklists could help.
+Checklists can be used in projects to ensure that best practices are followed by
+all developers.  Some examples include checklists merging branches into master
+\citep{Brown2015}, checklists for saving and sharing changes to the project
+\citep{WilsonEtAl2016}, checklists for new and departing team members
+\citep{HerouxAndBernholdt2018}, checklists for processes related to commits and
+releases \citep{HerouxEtAl2008} and checklists for overall software quality
+\citep{ThielEtAl2020, SSI2022}.  For instace, for Lattice Boltzmann solver
+software, ESPResSo has a checklist for managing releases \citep{Michalski2021}.
+
+The above discussion shows that, taken together, MI projects fall somewhat short
+of recommended best practices for research software.  However, MI software is
+not alone in this.  Many, if not most, research projects fall short of best
+practices.  A gap exists in scientific computing development practices and
+software engineering recommendations \citep{Storer2017, Kelly2007,
+OwojaiyeEtAl2021_CSE}. \citet{JohansonAndHasselbring2018} observe that the
+state-of-the-practice for SCS in industry and academia does not incorporate
+state-of-the-art SE tools and methods.  This causes sustainability and
+reliability problems \citep{FaulkEtAl2009}. Rather than benefit from capturing
+and reusing previous knowledge, projects waste time and energy ``reinventing the
+wheel'' \citep{deSouzaEtAl2019}.
 
@@
 
+\wss{observed artifacts (Section~\ref{Sec_CompareArtifacts}) - human judgement,
+not always given the expected name, but the content is there.}
+
 \begin{description}
@@
 
+\wss{Talk about increasing the number of developers.  CONTRIBUTING is rare
+(Table~\ref{artifactspresent}), but it doesn't have to be.  Related to this
+information for new developers is rare: Code Style Guide, Requirements, Design
+Doc, API.}
+
 \begin{itemize}

### 2022-04-05 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Added updated artifact information from Peter.

@@
 \midrule
-Authors / Developers List & Change Log / Release Notes & API Documentation\\
-Installation Guide / Instructions & Design Documentation & Developer / Contributor Manual\\
-Issue Tracker & Functional Spec. / Notes & FAQ / Forum\\
-Library Dependency List & Performance Information & Requirements Spec.\\
-License & Test Plan / Script / Cases & Verification and
-Validation Plan\\
-List of Related Publications & User Manual / Guide & Video Guide
-(including YouTube)\\
-Makefile / Build File &  & \\
-README File & & \\
-Theory Notes & & \\
+Authors / Developers List & Change Log / Release Notes & Acknowledgements\\
+Getting Started & Citation & API Documentation\\
+Installation Guide / Instructions & Contributing & Code Style Guide\\
+Issue Tracker & Design Documentation & Developer / Contributor Manual\\
+Library Dependency List & Functional Spec. / Notes & FAQ / Forum\\
+License & Performance Information & Requirements Spec.\\
+List of Related Publications & Test Plan / Script / Cases & Product Roadmap\\
+Makefile / Build File & User Manual / Guide & Unistall\\
+README File & & Verification and Validation Plan\\
+Theory Notes & & Video Guide (including YouTube)\\
 Tutorial & & \\
@@
 \midrule
-LICENSE & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & &
-\checkmark & \checkmark & \checkmark & C\\
-README &  & \checkmark & \checkmark & \checkmark & \checkmark & & \checkmark &
-\checkmark & \checkmark & C\\
-CONTRIBUTING &  & \checkmark & \checkmark & \checkmark & \checkmark & &
-\checkmark & \checkmark & \checkmark & R\\
+LICENSE & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & & \checkmark & \checkmark & \checkmark & C\\
+README &  & \checkmark & \checkmark & \checkmark & \checkmark & & \checkmark & \checkmark & \checkmark & C\\
+CONTRIBUTING &  & \checkmark & \checkmark & \checkmark & \checkmark & & \checkmark & \checkmark & \checkmark & U\\
 CITATION &  &  &  & \checkmark & & & & \checkmark & \checkmark & U\\
 CHANGELOG &  & \checkmark &  & \checkmark & \checkmark & & \checkmark &  &  & U\\
-INSTALL &  &  &  &  & \checkmark & & \checkmark & \checkmark & \checkmark & U\\
+INSTALL &  &  &  &  & \checkmark & & \checkmark & \checkmark & \checkmark & C\\
 \midrule
-Uninstall &  &  &  &  & & & & \checkmark & &  \\
-Dependency List &  &  & \checkmark & & \checkmark & & & \checkmark &  & R\\
-Authors &  &  &  &  &  &  & \checkmark & \checkmark & \checkmark & U\\
-Code of Conduct &  &  &  &  & & & \checkmark & & & R\\
-Acknowledgements &  &  &  &  &  &  & \checkmark & \checkmark & \checkmark & U\\
+Uninstall &  &  &  &  & & & & \checkmark & & R\\
+Dependency List &  &  & \checkmark & & \checkmark & & & \checkmark &  & C\\
+Authors &  &  &  &  &  &  & \checkmark & \checkmark & \checkmark & C\\
+Code of Conduct &  &  &  &  & & & \checkmark & & & \\
+Acknowledgements &  &  &  &  &  &  & \checkmark & \checkmark & \checkmark & R\\
 Code Style Guide &  & \checkmark &  &  & & & \checkmark & \checkmark & \checkmark & R\\
-Release Info. &  & \checkmark &  &  & & \checkmark & \checkmark & & & C\\
+Release Info. &  & \checkmark &  &  & & \checkmark & \checkmark & & & U\\
 Prod.\ Roadmap &  &  &  &  & & \checkmark & \checkmark & \checkmark & & R\\
 \midrule
-Getting started &  &  &  &  & \checkmark & & \checkmark & \checkmark & \checkmark & R\\
-User manual &  &  & \checkmark &  & & & \checkmark & & & C\\
-Tutorials &  &  &  &  & & & \checkmark & & & U\\
-FAQ &  &  &  &  & & & \checkmark & \checkmark & \checkmark & U\\
+Getting started &  &  &  &  & \checkmark & & \checkmark & \checkmark & \checkmark & C\\
+User manual &  &  & \checkmark &  & & & \checkmark & & & U\\
+Tutorials &  &  &  &  & & & \checkmark & & & C\\
+FAQ &  &  &  &  & & & \checkmark & \checkmark & \checkmark & R\\
 \midrule
-Issue Track &  & \checkmark & \checkmark & & \checkmark & \checkmark &
+Issue Track &  & \checkmark & \checkmark & & \checkmark & \checkmark & \checkmark & & \checkmark & C\\
+Version Control &  & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & C\\ 
+Build Scripts &  & \checkmark &  & \checkmark & \checkmark & \checkmark &
 \checkmark & & \checkmark & C\\
-Version Control &  & \checkmark & \checkmark & \checkmark & \checkmark &
-\checkmark & \checkmark & \checkmark & \checkmark & C\\ 
-Build Scripts &  & \checkmark &  & \checkmark & \checkmark & \checkmark &
-\checkmark & & \checkmark & U\\
 \midrule
@@
 Design Doc.\ &  & \checkmark  & \checkmark &  & \checkmark & & \checkmark &
-\checkmark& \checkmark & R\\
+\checkmark& \checkmark & U\\
 API Doc. &  &  &  &  & \checkmark & & \checkmark & \checkmark & \checkmark & R\\
-Test Plan &  & \checkmark &  &  & & \checkmark & & & &  \\
+Test Plan &  & \checkmark &  &  & & \checkmark & & & &  U\\
 Test Cases & \checkmark & \checkmark & \checkmark &  & \checkmark & \checkmark &

### 2022-04-05 StateOfPractice/Papers/MedImageSoft_SOP-arXiv/MedImageSoft_SOP_AcademicRadiology.tex
Updates to Section 6 of SOP for MI

@@
 
-Table~\ref{tab_user_support_model} summarizes the user support models by the
-number of projects using them (projects may use more than one support model). We
-do not know whether the prevalent use of GitHub issues for user support is by
-design, or whether this just naturally happened as users sought help.
-
-\begin{table}[!ht]
-\centering
-\begin{tabular}{lc}
-\toprule
-\multicolumn{1}{c}{User support model} & Num.\ projects \\
-\midrule
-GitHub issue & 24 \\
-Frequently Asked Questions (FAQ) & 12 \\
-Forum & 10 \\
-E-mail address & 9 \\
-GitLab issue, SourceForge discussions & 2 \\
-Troubleshooting & 2 \\
-Contact form & 1 \\ 
-\bottomrule
-\end{tabular}
-\caption{\label{tab_user_support_model}User support models by number of projects}
-\end{table}
-
 \subsection{Maintainability} \label{sec_score_maintainability}
@@
 we found it to have the most comprehensive artifacts. For example, as far as we
-could find, only a few of the 29 projects had a project plan, developer's
+could find, only a few of the 29 projects had a product, developer's
 manual, or API documentation, and only \textit{3D Slicer}, \textit{ImageJ},
@@
 \toprule
-\multicolumn{1}{c}{Software} & Proj.\ plan & Dev.\ manual & API doc. \\ 
+\multicolumn{1}{c}{Software} & Prod.\ roadmap & Dev.\ manual & API doc. \\ 
 \midrule
@@
      & & Code of conduct (1)\\
+     & & Requirements (1)\\
     \bottomrule
@@
 \midrule
-Requirements &  & \checkmark &  &  & & \checkmark &  &  & \checkmark &  \\
+Requirements &  & \checkmark &  &  & & \checkmark &  &  & \checkmark & R\\
 Design Doc.\ &  & \checkmark  & \checkmark &  & \checkmark & & \checkmark &
@@
 and evolution of software, software processes, and artifacts \citep[p.\
-501]{GhezziEtAl2003}. Many tools are used by MI software packages, as summarized
-in Table~?.  The tools are subdivided into development tools, dependencies, and
-project management tools.  We answer \rqref{RQ_CompareToolsProjMngmnt} by
-comparing aspects of tool usage in MI software packages to their utilization in
-the research software community in general.
-
-We answer this question by measuring the qualities and interviewing the
-developers. We learned that developers use tools for CI/CD
-(Section~\ref{sec_result_correctness_verifiability}), user support
-(Table~\ref{tab_user_support_model}), version control
-(Section~\ref{sec_score_maintainability}), documentation
-(Section~\ref{sec_contribution_pm}), contribution management
-(Section~\ref{sec_contribution_pm}), and project management
-(Section~\ref{sec_contribution_pm}).
-
-As mentioned in Section \ref{sec_result_correctness_verifiability}, we
-identified five projects using CI/CD tools: \textit{3D Slicer} and \textit{OHIF
-Viewer} use CircleCI; \textit{ImageJ}, \textit{Fiji}, and \textit{dwv} use
-Travis. We identified the above projects and tools by examining the
-documentation and source code of all projects. Not all tools leave identifiable
-artifacts in the repositories; therefore, we may have missed the use of some
-tools. According to the interviews with developers, \textit{dwv} and
-\textit{Weasis} used GitHub Actions.
+501]{GhezziEtAl2003}. To tools used in MI software include tools for CI/CD, user
+support, version control, documentation and project management.  We answer
+\rqref{RQ_CompareToolsProjMngmnt} by comparing aspects of tool usage in MI
+software packages to their utilization in the research software community in
+general.
+
+CI/CD paragraph.  As mentioned in
+Section~\ref{sec_result_correctness_verifiability}, we identified five projects
+using CI/CD tools: \textit{3D Slicer}, \textit{ImageJ}, \textit{Fiji},
+\textit{dwv}, and \textit{OHIF Viewer}. We identified the above projects and
+tools by examining the documentation and source code of all projects.
+Potentially not all tools leave identifiable artifacts in the repositories;
+therefore, we may have missed the use of some tools.
+
+user support paragraph. Table~\ref{tab_user_support_model} summarizes the user
+support models by the number of projects using them (projects may use more than
+one support model). We do not know whether the prevalent use of GitHub issues
+for user support is by design, or whether this just naturally happened as users
+sought help.
+
+\begin{table}[!ht]
+\centering
+\begin{tabular}{lc}
+\toprule
+\multicolumn{1}{c}{User support model} & Num.\ projects \\
+\midrule
+GitHub issue & 24 \\
+Frequently Asked Questions (FAQ) & 12 \\
+Forum & 10 \\
+E-mail address & 9 \\
+GitLab issue, SourceForge discussions & 2 \\
+Troubleshooting & 2 \\
+Contact form & 1 \\ 
+\bottomrule
+\end{tabular}
+\caption{\label{tab_user_support_model}User support models by number of projects}
+\end{table}
+
+version control paragraph.  From Section~\ref{sec_score_maintainability}, 27 of
+the 29 projects used git as the version control tool, with 24 of these using
+GitHub. \textit{AMIDE} used Mercurial and \textit{Gwyddion} used Subversion.
+\textit{XMedCon}, \textit{AMIDE}, and \textit{Gwyddion} used SourceForge.
+\textit{DicomBrowser} and \textit{3DimViewer} used BitBucket.  Although teams
+may have a process for accepting new contributions, no one discussed this during
+their interview. However, most teams (8 of 9) mentioned using GitHub and pull
+requests to manage contributions from the community. The interviewees generally
+gave very positive feedback on using GitHub. Some teams previously used a
+different approach to version control and eventually transferred to git and
+GitHub.  The past approaches included contributions from e-mail (3 teams),
+contributions from forums (1 team) and sharing the git repository via e-mail (1
+team).  \wss{I do not know what this last one means. Should ask Ao.} \ad{The
+last one happened when cloud-based repositories (Github, Gitlab, Bitbucket,
+etc.) were not widely used. The project git repository was stored locally on the
+project owner's local computer. The owner shared the whole git repo to team
+members and other individuals who wanted to contribute. The contributors made
+changes to the repo, and sent it back. They transferred the repo back and forth
+by email.}
+
+documentation paragraph. For documentation tools and methods mentioned by the
+interviewees, the most popular (mentioned by about 30\% of developers) were
+forum discussions and videos.  The second most popular options (mentioned by
+about 20\% of developers) were GitHub, wiki pages, workshops, and social media.
+The least frequently mentioned options (about 10\% of developers) included
+writing books, google forms and state management.
+
+project management paragraph. Some interviewees mentioned the project management
+tools they used. Generally speaking, the interviewees talked about two types of
+tools:
+\begin{inparaenum}[i)]
+\item trackers, including GitHub, issue trackers, bug trackers and Jira; and,
+\item documentation tools, including GitHub, Wiki page, Google Doc, and Confluence.
+\end{inparaenum}
+Of the specifically named tools in the above lists, GitHub was mentioned 3
+times, while each of the other tools was only mentioned once.
 
@@
 
-For documentation tools and methods mentioned by the interviewees, the most
-popular (mentioned by about 30\% of developers) were forum discussions and
-videos.  The second most popular options (mentioned by about 20\% of developers)
-were GitHub, wiki pages, workshops, and social media.  The least frequently
-mentioned options (about 10\% of developers) included writing books, google
-forms and state management.
-
 \subsection{Project Management} \label{sec_contribution_pm}
@@
 
-Although teams may have a process for accepting new contributions, no one
-discussed this during their interview. However, most teams (8 of 9) mentioned
-using GitHub and pull requests to manage contributions from the community. The
-interviewees generally gave very positive feedback on using GitHub. Some teams
-previously used a different approach to version control and eventually
-transferred to git and GitHub.  The past approaches included contributions
-from e-mail (3 teams), contributions from forums (1 team) and sharing the
-git repository via e-mail (1 team).  \wss{I do not know what this last one means.
-Should ask Ao.} \ad{The last one happened when cloud-based repositories (Github,
-Gitlab, Bitbucket, etc.) were not widely used. The project git repository was
-stored locally on the project owner's local computer. The owner shared the whole
-git repo to team members and other individuals who wanted to contribute. The
-contributors made changes to the repo, and sent it back. They transferred the
-repo back and forth by email.}
-
 For managing contributions, the \textit{3D Slicer} team encouraged users to
@@
 self-directed.
-
-Some interviewees mentioned the project management tools they used. Generally
-speaking, the interviewees talked about two types of tools:
-\begin{inparaenum}[i)]
-\item trackers, including GitHub, issue trackers, bug trackers and Jira; and,
-\item documentation tools, including GitHub, Wiki page, Google Doc, and Confluence.
-\end{inparaenum}
-Of the specifically named tools in the above lists, GitHub was mentioned 3
-times, while each of the other tools was only mentioned once.
 
@@
     \item usability tests and interviews with end users;
-    \item adjusting according to users’ feedbacks;
+    \item adjusting according to users' feedbacks;
     \item getting started tutorials;

### 2022-04-07 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Updates to SOP for LBM

@@
 
-This section answers research question \rqref{RQ_CompareMethodologies} by
-comparing the principles, processes and methodologies used for LBM software to
-what can be gleaned from the literature on research software in general. In our
-data collection for LBM software, the software development process is not
-explicitly indicated in the artifacts for most of the packages. However, during
-our interviews one developer (ESPResSo) told us their non-rigorous development
-model is like a combination of agile and waterfall. Employing a loosely defined
+We answer research question \rqref{RQ_CompareMethodologies} by comparing the
+principles, processes and methodologies used for LBM software to what can be
+gleaned from the literature on research software in general. In our data
+collection for LBM software, the software development process is not explicitly
+indicated in the artifacts for most of the packages. However, during our
+interviews one developer (ESPResSo) told us their non-rigorous development model
+is like a combination of agile and waterfall. Employing a loosely defined
 process makes sense for LBM software, given that the teams are generally small
@@
 supports change. Ocean modelling software is currently feeling the pain of
-not emphasizing modularization in legacy code \citep{Reiner2022}.
+not emphasizing modularization in legacy code \citep{JungEtAl2022}.
 

### 2022-04-07 StateOfPractice/Papers/MedImageSoft_SOP-arXiv/MedImageSoft_SOP_AcademicRadiology.tex
Updates to SOP for MI - working on pain points section

@@
 
+The measurement results summarize to this point are based on information we
+collect from on-line resources. This information is incomplete because it
+doesn't generally capture the development process, the developer pain points,
+the perceived threats to software quality, and the developers' strategies to
+address these threats.  Therefore, part of our methodology is to interview
+developers.
+
 Our interviews were guided by a list of 20 questions, which can be found in
@@
 interviews are semi-structured based on the question list; we ask follow-up
-questions when necessary. Based on our experience, the interviewees usually
-brought up unexpected and exciting ideas.
-
-We sent out interview requests to all 29 projects, with 9 projects responding.
-In some cases multiple developers from the same project agreed to be
-interviewed. We found contact information from projects websites, code
-repository, publications, and from biographic pages at the teams's institutions.
-Meetings were held on-line using either Zoom or Teams, which facilitated
-recording and automatic transcription of the meetings.  The interview process
-presented here was approved by the McMaster University Research Ethics Board
-under the application number 
+questions when necessary.  The interview process presented here was approved by
+the McMaster University Research Ethics Board under the application number 
 \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/MACREM/Application.pdf}
 {MREB\#: 5219}.
+
+We sent interview requests to all 29 projects using contact information from
+projects websites, code repository, publications, and from biographic pages at
+the teams's institutions.  In the end nine developers from eight of the projects
+agreed to participate: \textit{3D Slicer}, \textit{INVESALIUS 3}, \textit{dwv},
+\textit{BioImage Suite Web}, \textit{ITK-SNAP}, \textit{MRIcroGL},
+\textit{Weasis}, and \textit{OHIF}. We spent about 90 minutes for each
+interview. One participant was too busy to have an interview, so they wrote down
+their answers. In one case two developers from the same project agreed to be
+interviewed. Meetings were held on-line using either Zoom or Teams, which
+facilitated recording and automatic transcription of the meetings. The full
+interview answers can be found in \citet{Dong2021}.
 
@@
 \end{tabular}
-\caption{Software with the visibility/transparency documents (software is listed
-in descending order of visibility/transparency score)}
+\caption{Software with visibility/transparency related documents (software is
+listed in descending order of visibility/transparency score)}
 \label{tab_Visibility/Transparency_docs}
@@
 software. Given the storage capacity of current hardware, developers and users
-do not generally concerned with uninstall. Moreover, as mentioned above,
+are not generally concerned with uninstall.  Moreover, as mentioned above,
 uninstall is not particularly emphasized in existing recommendations.  Test
@@
 would have to at least be implicitly created, since test cases were observed
-with reasonable frequency for MI software (uncommon).
+with reasonable frequency for MI software (test cases are are categorized as
+uncommon).
 
@@
 up-front requirements are impossible for research software
-\citep{CarverEtAl2007,SegalAndMorris2008}, but when the instance on ``up-front''
-requirements is dropped, allowing the requirements to be written iteratively and
-incrementally, requirements are feasible \citep{Smith2016}.
+\citep{CarverEtAl2007, SegalAndMorris2008}, but when the instance on
+``up-front'' requirements is dropped, allowing the requirements to be written
+iteratively and incrementally, requirements are feasible \citep{Smith2016}.
 
 Table~\ref{Tbl_Guidelines} shows several artifacts that are rarely observed in
-practice.  A theme amount these rare artifacts is that many of them are related
+practice.  A theme among these rare artifacts is that many of them are related
 to developers more than users.  For instance, the rare artifacts include
-Contributing, developer Code of conduct, Code Style Guide, Product roadmaps,
+Contributing, Developer code of conduct, Code style guide, Product roadmap,
 Requirements, Design documentation, API documentation and a Test plan.  The rare
 artifacts for MI software are similar to the rare artifacts for Lattice
-Boltzmann solvers \citep{Michalski2021}.  However, MI software differs from
-ocean modelling software, since, according to \citep{JungEtAl2022}, the later
-pays relatively less attention to testing.
+Boltzmann solvers \citep{Michalski2021}, except LBM software is more likely to
+have developer related artifacts, like Contributing, Dependency list, and Design
+documentation.
 
@@
 and evolution of software, software processes, and artifacts \citep[p.\
-501]{GhezziEtAl2003}. To tools used in MI software include tools for CI/CD, user
-support, version control, documentation and project management.  We answer
-\rqref{RQ_CompareToolsProjMngmnt} by comparing aspects of tool usage in MI
-software packages to their utilization in the research software community in
-general.
-
-CI/CD paragraph.  As mentioned in
-Section~\ref{sec_result_correctness_verifiability}, we identified five projects
-using CI/CD tools: \textit{3D Slicer}, \textit{ImageJ}, \textit{Fiji},
-\textit{dwv}, and \textit{OHIF Viewer}. We identified the above projects and
-tools by examining the documentation and source code of all projects.
-Potentially not all tools leave identifiable artifacts in the repositories;
-therefore, we may have missed the use of some tools.
-
-user support paragraph. Table~\ref{tab_user_support_model} summarizes the user
-support models by the number of projects using them (projects may use more than
-one support model). We do not know whether the prevalent use of GitHub issues
-for user support is by design, or whether this just naturally happened as users
-sought help.
+501]{GhezziEtAl2003}. MI software uses tools for CI/CD, user support, version
+control, documentation and project management.  To answer
+\rqref{RQ_CompareToolsProjMngmnt} we summarize the tool usage in these
+categories, and compare this to the usage by the research software community.
+
+As mentioned in Section~\ref{sec_result_correctness_verifiability}, we
+identified five projects using CI/CD tools (about 17\% of the assessed
+projects). CI/CD using projects were found by examining the documentation and
+source code of all projects. The count of CI/CD usage may actually be higher,
+since traces of CI/CD usage may not always appear in a repository.  This was the
+case for a study of LBM software, where interviews with developers showed that
+more projects used CI/CD than was evident from repository artifacts alone
+\citep{Michalski2021}.  The 17\% utilization for MI software contrasts with the
+high frequency with which continuous integration is recommended in research
+software development guidelines \citep{BrettEtAl2021, Brown2015, ThielEtAl2020,
+Zadka2018, vanGompelEtAl2016}. Although there is currently little data available
+on CI/CD utilization for research software, our impression is that CI/CD is not
+yet common practice, despite its recommendation.  For LBM software at least the
+situation is similar to MI software, with only 12.5\% of 24 LBM packages showing
+evidence of CI/CD in their repositories \citep{Michalski2021}.  \wss{add
+reference to CarverEtAl2022, once it is published.}
+
+Table~\ref{tab_user_support_model} summarizes the user support models by the
+number of projects using each model (projects may use more than one support
+model). We do not know whether the prevalent use of GitHub issues for user
+support is by design, or whether this just naturally happenes as users seek
+help. The common use of GitHub by MI developers is not surprising, given that
+GitHub is the largest code host in the world, with over 128 million public
+repositories and over 23 million users (as of roughly February 26, 2020)
+\citep{Kashyap2020}.
 
@@
 
-version control paragraph.  From Section~\ref{sec_score_maintainability}, 27 of
-the 29 projects used git as the version control tool, with 24 of these using
-GitHub. \textit{AMIDE} used Mercurial and \textit{Gwyddion} used Subversion.
-\textit{XMedCon}, \textit{AMIDE}, and \textit{Gwyddion} used SourceForge.
-\textit{DicomBrowser} and \textit{3DimViewer} used BitBucket.  Although teams
-may have a process for accepting new contributions, no one discussed this during
-their interview. However, most teams (8 of 9) mentioned using GitHub and pull
-requests to manage contributions from the community. The interviewees generally
-gave very positive feedback on using GitHub. Some teams previously used a
-different approach to version control and eventually transferred to git and
-GitHub.  The past approaches included contributions from e-mail (3 teams),
-contributions from forums (1 team) and sharing the git repository via e-mail (1
-team).  \wss{I do not know what this last one means. Should ask Ao.} \ad{The
-last one happened when cloud-based repositories (Github, Gitlab, Bitbucket,
-etc.) were not widely used. The project git repository was stored locally on the
-project owner's local computer. The owner shared the whole git repo to team
-members and other individuals who wanted to contribute. The contributors made
-changes to the repo, and sent it back. They transferred the repo back and forth
-by email.}
-
-documentation paragraph. For documentation tools and methods mentioned by the
-interviewees, the most popular (mentioned by about 30\% of developers) were
-forum discussions and videos.  The second most popular options (mentioned by
-about 20\% of developers) were GitHub, wiki pages, workshops, and social media.
-The least frequently mentioned options (about 10\% of developers) included
-writing books, google forms and state management.
-
-project management paragraph. Some interviewees mentioned the project management
-tools they used. Generally speaking, the interviewees talked about two types of
-tools:
+From Section~\ref{sec_score_maintainability}, 27 of the 29 projects used git as
+the version control tool, 1 used Mercurial and 1 used Subversion.  The hosting
+is on GitHub for 24 packages, SourceForge for 3 and BitBucket for 2.  Although
+teams may have a process for accepting new contributions, no one discussed this
+during their interviews. However, most teams (8 of 9) mentioned using GitHub and
+pull requests to manage contributions from the community. The interviewees
+generally gave very positive feedback on using GitHub. Some teams previously
+used a different approach to version control and eventually transferred to git
+and GitHub.  The past approaches included contributions from e-mail (3 teams),
+contributions from forums (1 team) and e-mailing the git repository back and
+forth between developers (1 team).
+
+The common use of version control for MI software illustrates considerable
+improvement from the poor adoption of version control tools that Wilson lamented
+in 2006 \citep{Wilson2006}.  The proliferation of version control tools for MI
+matches the increase in the broader research software community.  A little over
+10 years ago version control was estimated to be used in only 50\% of research
+software projects \citep{Nguyen-HoanEtAl2010}, but even at that time
+\citet{Nguyen-HoanEtAl2010} noted an increase from previous usage levels. A
+survey in 2018 shows 81\% of developers use a version control system
+\citep{AlNoamanyAndBorghi2018}. \citet{Smith2018} has similar results, showing
+version control usage for alive projects in mesh generation, geographic
+information systems and statistical software for psychiatry increasing from
+75\%, 89\% and 17\% (respectively) to 100\%, 95\% and 100\% (respectively) over
+a four year period ending in 2018. (For completeness the same study showed a
+decrease in version control usage for seismology software over the same time
+period, from 41\% down to 36\%).  Almost every software guide cited in
+Section~\ref{Sec_CompareArtifacts} includes the advice to use version control.
+The high usage of version control tools in MI software matches the trend in
+research software in general.
+
+For documentation tools and methods mentioned by the interviewees, the most
+popular (mentioned by about 30\% of developers) were forum discussions and
+videos.  The second most popular options (mentioned by about 20\% of developers)
+were GitHub, wiki pages, workshops, and social media. The least frequently
+mentioned options (about 10\% of developers) included writing books, google
+forms and state management.  In contrasting MI software with LBM software, the
+most significant documentation tool difference is that LBM software often uses
+document generation tools, like doxygen and sphinx \citep{Michalski2021}, while
+MI does not appear to use these tools.
+
+Some interviewees mentioned the project management tools they used. Generally
+speaking, the interviewees talked about two types of tools:
 \begin{inparaenum}[i)]
 \item trackers, including GitHub, issue trackers, bug trackers and Jira; and,
-\item documentation tools, including GitHub, Wiki page, Google Doc, and Confluence.
+\item documentation tools, including GitHub, Wiki page, Google Doc, and
+Confluence.
 \end{inparaenum}
@@
 
+Based on information provided by \citet{JungEtAl2022}, tool utilization for MI
+software has much in common with tool utilization for ocean modelling software.
+Both use tools for editing, compiling, code management, testing, building and
+project management.  From the data available, ocean modelling differs from MI
+software in its use of Kanban boards for project management.
+
 \section{Comparison of Principles, Process and Methodologies to Research Software in General} \label{Sec_CompareMethodologies}
 
-This section answers research question \rqref{RQ_CompareMethodologies} by
-comparing the principles, processes and methodologies used for MI software to
-what can be gleaned from the literature on research software in general. In our
-data collection for MI software, the software development process is not
-explicitly indicated in the artifacts for most of the packages. However, during
-our interviews one developer (ESPResSo) told us their non-rigorous development
-model is like a combination of agile and waterfall. Employing a loosely defined
-process makes sense for MI software, given that the teams are generally small
-and self-contained.  Although eleven of the packages explicitly convey that they
-would accept outside contributors, generally the teams are centralized, often
-working at the same institution.  Working at the same institution means that an
-informal process can show success, since informal conversations are relatively
-easy to have.
-
-We answer this question by measuring the qualities and interviewing the
-developers. We previously summarized principles, processes, and methodologies in
-software testing (Section~\ref{Sec_CompareMethodologies}), documentation
-(Section~\ref{sec_interview_documents}, Table~\ref{tab_maintainability_docs},
-Table~\ref{tab_Visibility/Transparency_docs}, contribution management
-(Section~\ref{sec_contribution_pm}), and project management
-(Section~\ref{sec_contribution_pm}).
+We answer research question \rqref{RQ_CompareMethodologies} by comparing the
+principles, processes and methodologies used for MI software to what can be
+gleaned from the literature on research software in general. In our interviews
+with developers the responses about development model were vague, with only two
+interviewees following a definite development model. In some cases the
+interviewees felt their process was similar to an existing development model.
+Three teams (about 38\%) either followed agile, or something similar to agile.
+Two teams (25\%) either followed a waterfall process, or something similar.
+Three teams (about 38\%) explicitly stated that their process was undefined or
+self-directed.
+
+Our observations of an informally defined process, with elements of agile
+methods, matches what has been observed for research software in general.
+Scientific developers naturally use an agile philosophy \citep{AckroydEtAl2008,
+CarverEtAl2007, EasterbrookAndJohns2009, Segal2005, HeatonAndCarver2015}, or an
+amethododical process \citep{Kelly2013}, or a knowledge acquisition driven
+process \citep{Kelly2015}.  A waterfall like process can work for research
+software \citep{Smith2016}, especially if the developers work iteratively and
+incrementally, but externally document their work as if a rationale design
+process were followed \citep{parnas1986rational}.
+
+No interviewee introduced any strictly defined project management process. The
+most common approach was following the issues, such as bugs and feature
+requests. Additionally, the \textit{3D Slicer} team had weekly meetings to
+discuss the goals for the project; the \textit{INVESALIUS 3} team relied on the
+GitHub process for their project management; the \textit{ITK-SNAP} team had a
+fixed six-month release pace; only the interviewee from the \textit{OHIF} team
+mentioned that the team has a project manager; the \textit{3D Slicer} team and
+\textit{BioImage Suite Web} team do nightly builds and tests. The \textit{OHIF}
+developer believes that a better project management process can improve junior
+developer efficiency while also improving internal and external communication.
 
@@
 tests with users) was the top solution to improve correctness, usability, and
-reproducibility. One pain point in the development process is the lack of access
-to real-world datasets for testing. The developers' strategies to address this
-are summarized in Section~\ref{Sec_AddressConcerns} \ad{to be fixed}. One threat to
-correctness is: with huge datasets for testing, the tests are expensive and
-time-consuming. Three interviewees endorsed self tests / automated tests, which
-may save time for testing.
-
-Documentation is apparently not emphasized in any of the 29 projects.  None of
-them had theory manuals, although we did identify a road map in the \textit{3D
-Slicer} project.  No requirements specifications were found. Eight of the nine
-interviewees thought that documentation was essential to their projects.
-However, they hold the common opinion that their documentation needed
-improvements. Nearly half of them also believed that the lack of time prevented
-them from improving the documentation.
-
-\subsection{Documents in the Projects} \label{sec_interview_documents}
-
-We tried to understand the interviewees' opinions on documentation and the
-quality of documentations with two questions:
-
-\begin{description}
-\item[Q11.] How does documentation fit into your development process? Would
-improved documentation help with the obstacles you typically face?
-\item[Q19.] Do you think the current documentation can clearly convey all
-necessary knowledge to the users? If yes, how did you successfully achieve it?
-If no, what improvements are needed?
-\end{description}
-
+reproducibility.  This level of testing matches what was observed for LBM
+software \citep{Michalski2021} and is apparently greater than the level of
+testing for ocean modelling software.  \citet{JungEtAl2022} reports that testing
+is underemphasized in ocean modelling.
+
+As the observed artifacts in Table~\ref{artifactspresent} show, documentation is
+not emphasized as part of the development process in any of the 29 projects.
+None of them had theory manuals, although we did identify a road map in the
+\textit{3D Slicer} project.  No requirements specifications were found.
 Table~\ref{tab_opinion_doc} summarizes interviewees' opinions on documentation.
 Interviewees from each of the eight projects thought that documentation was
-important to their projects, and most of them said that it could save their time
+essential to their projects, and most of them said that it could save their time
 to answer questions from users and developers. Most of them saw the need to
 improve their documentation, and only three of them thought that their
-documentations conveyed information clearly enough.
+documentations conveyed information clearly enough. Nearly half of developers
+also believed that the lack of time prevented them from improving documentation.
 
@@
 \end{tabular}
-\caption{\label{tab_opinion_doc}Opinions on documentation by the numbers of
-interviewees with the answers}
+\caption{Opinions on documentation by the numbers of interviewees with the
+answers}
+\label{tab_opinion_doc}
 \end{table}
 
-\subsection{Project Management} \label{sec_contribution_pm}
-
-To investigate project management we asked three questions:
-
-\begin{description}
-\item[Q5.] Do you have a defined process for accepting new contributions into
-your team?
-\item[Q13.] What is your software development model? For example, waterfall,
-agile, etc.
-\item[Q14.] What is your project management process? Do you
-think improving this process can tackle the current problem? Were any project
-management tools used?
-\end{description}
-
-For managing contributions, the \textit{3D Slicer} team encouraged users to
-develop their extensions for specific use cases, while the \textit{OHIF} team
-enabled the use of plug-ins.  The interviewee from the \textit{ITK-SNAP} team
-said one way of accepting new team members was through funded academic projects.
-
-For the most part, the responses about development model were vague, with only
-two interviewees following a definite development model. In some cases the
-interviewees felt their process was similar to an existing development model.
-Three teams (about 38\%) either followed agile, or something similar to agile.
-Two teams (25\%) either followed a waterfall process, or something similar.
-Three teams (about 38\%) explicitly stated that their process was undefined or
-self-directed.
-
-No interviewee introduced any strictly defined project management process. The
-most common approach was following the issues, such as bugs and feature requests.
-Additionally, the \textit{3D Slicer} team had weekly meetings to discuss the
-goals for the project; the \textit{INVESALIUS 3} team relied on the GitHub
-process for their project management; the \textit{ITK-SNAP} team had a fixed
-six-month release pace; only the interviewee from the \textit{OHIF} team
-mentioned that the team has a project manager; the \textit{3D Slicer} team and
-\textit{BioImage Suite Web} team do nightly builds and tests.
-
-Most interviewees skipped the part of the last question asking ``Do you think
-improving this process can tackle the current problem?''. In retrospect, we
-should not have asked a yes-or-no question, since a binary answer is not very
-informative. The interviewee from the \textit{OHIF} team gave a positive answer
-to this question. They believed that a better project management process can
-improve the efficiency of junior developers. They also improved the project
-management tools (from public Jira to public GitHub repository plus private
-Jira), so they could better communicate externally and internally.
+As Table~\ref{Tbl_Guidelines} suggests, an emphasis on documentation, especially
+for new developers, is echoed in research software guidelines. Multiple
+guidelines recommend a document explaining how to contribute to a project, often
+named CONTRIBUTING. Tutorials, user guides and quick start examples are also
+recommended. \citet{SmithAndRoscoe2018} suggests including instructions
+specifically for on-boarding new developers. For open source software in general
+(not just research software), \citet{Fogel2005} recommends providing tutorial
+style examples, developer guidelines, demos and screenshots.
 
@@
 
-Based on interviews with four developers, this section aims to answer the
-research questions: i) What are the pain points for developers working on
-research software projects (\rqref{RQ_PainPoints})?; and, ii) How do the pain
-points of developers from MI compare to the pain points for research software
-in general (\rqref{RQ_ComparePainPoints})?  Below we go through each of the
-identified pain points and include citations that contrast the MI experience
-with observations from researchers in other domains.  Potential ways to address
-the pain points are covered in Section~\ref{Sec_AddressConcerns}. The full
-interview questions are found in \citet{SmithEtAl2021}.
-
-The measurement results in Section \ref{ch_results} are based on information we
-collected from on-line resources. This information is incomplete because it
-doesn't generally capture the development process, the developer pain points,
-the perceived threats to software quality, and the developers' strategies to
-address these threats. To collect this information we interviewed nine
-developers from eight of the 29 MI software projects. The eight projects are
-\textit{3D Slicer}, \textit{INVESALIUS 3}, \textit{dwv}, \textit{BioImage Suite
-Web}, \textit{ITK-SNAP}, \textit{MRIcroGL}, \textit{Weasis}, and \textit{OHIF}.
-We spent about 90 minutes for each interview and asked 20 prepared questions
-(summarized in \citet{SmithEtAl2021}). We also asked following-up questions as
-the conversation dictated. One participant was too busy to have an interview, so
-they wrote down their answers. The interviewees may have provided multiple
-answers to each question. Thus, when we summarize the frequency of different
-responses, the total is sometimes larger than nine.
-
-The full interview answers can be found in \citet{Dong2021}.  Selected questions
-and answers are highlighted and analyzed below.  The question numbers refer to
-the numbering in \citet{SmithEtAl2021}.
-
-%\label{sec_interview_pain_points}
-
-The pain points include current and past obstacles, as explored via the
-following questions:
-
-\begin{description}
-\item[Q9.] \hypertarget{q9}Currently, what are the most significant obstacles in
-your development process?
-\item[Q10.] \hypertarget{q10}How might you change your development process to
-remove or reduce these obstacles?
-\item[Q12.] \hypertarget{q12}In the past, is there any major obstacle to your
-development process that has been solved? How did you solve it?
-\end{description}
+Based on interviews with nine developers (described in
+Section~\ref{sec_interview_methods}), we answer two research questions: i) What
+are the pain points for developers working on research software projects?
+(\rqref{RQ_PainPoints}); and, ii) How do the pain points of developers from MI
+compare to the pain points for research software in general?
+(\rqref{RQ_ComparePainPoints}).  We go through each of the identified pain
+points and include citations that contrast the MI experience with observations
+from researchers in other domains.  Potential ways to address the pain points
+are covered in Sections~\ref{Sec_AddressConcerns} and~\ref{ch_recommendations}.
 
@@
 and \textit{testing}. We put the less mentioned ones into the category
-\textit{Others}.
+\textit{Others}. The interviewees may have provided multiple
+answers to each question. Thus, when we summarize the frequency of different
+responses, the total is sometimes larger than nine.
 
@@
 \midrule
-\multirow{3}{*}{Balance} & Hard to keep up with changes in OS and libraries & 1 &  \\
+\multirow{3}{*}{Balance} & Hard to keep up with changes in OS and libraries & 1
+&  \\
  & Hard to support multiple OS & 2 &  \\
@@
 
-For the pain point category of testing developers identified the the lack of access to
-real-world datasets.  Some proposed and proven solutions are as follows:
+For the pain point category of testing developers identified the the lack of
+access to real-world datasets.  One pain point in the development process is the
+lack of access to real-world datasets for testing. The developers' strategies to
+address this are summarized in Section~\ref{Sec_AddressConcerns} \ad{to be
+fixed}. One threat to correctness is: with huge datasets for testing, the tests
+are expensive and time-consuming. Three interviewees endorsed self tests /
+automated tests, which may save time for testing. Some proposed and proven
+solutions are as follows:
 
@@
 by other research software domains.
+
+\section{Recommendations} \label{ch_recommendations}
+
+This section presents our recommendations on MI software development. Although
+our focus is on MI software, unless noted otherwise, our recommendations apply
+to any scientific computing software.
+Section~\ref{sec_recommendations_qualities} discusses the actions that can
+potentially improve the ten software qualities.
+Sections~\ref{sec_recommendations_limited_resources},
+\ref{sec_recommendations_tech_stack}, and
+\ref{sec_recommendations_testing_dataset} are based on the primary pain points
+(from Section~\ref{painpoints}) collected from developers in the
+MI domain.
+
+\subsection{Recommendations on Improving Software Qualities}
+\label{sec_recommendations_qualities}
+
+Peer review --  One of the developers (ESPResSo) that was interviewed noted that
+an ad hoc peer review process is used to assess major changes and additions.
+Using peer review (also called technical review) matches with recommended
+practice for research software \citep{HerouxEtAl2008, Givler2020, OrvizEtAl2017,
+USGS2019}.
+
+Based on our quality measurements in Section~\ref{ch_results} and discussions
+with the developers in Section~\ref{painpoints}, we collected key points for
+improving software qualities. These points should be considered for new and
+existing projects, but we are not saying that every project should do everything
+on our list.  All projects will have finite resources, so strategies will have
+to be selected that provide the greatest return on investment.  Our specific
+recommendations for each quality are as follows:
+
+\begin{itemize}
+\item \textbf{Installability} (Section~\ref{sec_result_installability})
+\begin{itemize}
+    \item providing clear instructions;
+    \item automating installation;
+    \item including all dependencies in the installer;
+    \item avoiding heavily depending on other commercial products (e.g.\ Matlab);
+    \item potentially building a web application that needs no installation.
+\end{itemize}
+\item \textbf{Correctness \& Verifiability}
+(Section~\ref{sec_result_correctness_verifiability} and
+Section~\ref{sec_interview_software_qualities})
+\begin{itemize}
+    \item test-driven development with unit tests, integration tests, and
+    nightly tests;
+    \item two stage development process with stable release \& nightly builds;
+    \item CI/CD;
+    \item requirements specifications and theory manuals \citep{Smith2016, SmithAndLai2005}.
+    \item static code analysis tools (e.g. Lint and SonarQube)
+\end{itemize}
+\item \textbf{Reliability} (Section~\ref{sec_result_reliability})
+\begin{itemize}
+    \item test-driven development with unit tests, integration tests, and nightly tests.
+    \item two stage development process with stable release \& nightly builds;
+    \item descriptive error messages.
+\end{itemize}
+\item \textbf{Robustness} (Section~\ref{sec_result_robustness})
+\begin{itemize}
+    \item designing with exception handling so the software can fail gracefully;
+    \item descriptive error messages.
+\end{itemize}
+\item \textbf{Usability} (Section~\ref{sec_result_usability} and Section~\ref{sec_interview_software_qualities})
+\begin{itemize}
+    \item usability tests and interviews with end users;
+    \item adjusting according to users' feedbacks;
+    \item getting started tutorials;
+    \item user manuals;
+    \item professional UX (User eXperience) designs;
+    \item graphical user interface;
+    \item active supports to users.
+\end{itemize}
+\item \textbf{Maintainability} (Section~\ref{sec_score_maintainability} and Section~\ref{sec_interview_software_qualities})
+\begin{itemize}
+    \item use GitHub (or equivalent version control and project management environment);
+    \item use a modular approach to design (we advocate basing the design on
+    Parnas's principle of information hiding: ``system details that are likely
+    to change independently should be the secrets of separate modules; the only
+    assumptions that should appear in the interfaces between modules are those
+    that are considered unlikely to change.'' \citep{ParnasEtAl2000})
+    \item documentation for developers: project plan, developer’s manual, and API documentation.
+\end{itemize}
+\item \textbf{Reusability} (Section~\ref{sec_result_reusability})
+\begin{itemize}
+    \item modular approach;
+    \item API documentation;
+    \item tools that generate software documentation for developers (e.g.\
+    Doxygen, Javadoc, and Sphinx).
+\end{itemize}
+\item \textbf{Understandability} (Section~\ref{sec_result_understandability} and
+Section~\ref{sec_interview_software_qualities})
+\begin{itemize}
+    \item modular approach;
+    \item good coding style: consistent indentation and formatting style;
+    consistent, distinctive, and meaningful code identifiers; keeping parameters
+    in the same order for all functions; avoiding hard-coded constants (other
+    than 0 and 1);
+    \item clear comments, indicating what is being done, not how;
+    \item description of algorithms used;
+    \item documentation of explicit requirements on coding standards;
+    \item communication between developers and users via GitHub issues, mailing
+    lists, and forums.
+\end{itemize}
+\item \textbf{Visibility/Transparency} (Section~\ref{sec_result_visibility_transparency})
+\begin{itemize}
+    \item documents for the development process, contributor's guide, project
+    status, development environment, and release notes.
+\end{itemize}
+\item \textbf{Reproducibility} (Section~\ref{sec_interview_software_qualities})
+\begin{itemize}
+    \item test-driven development with unit tests, integration tests, and nightly tests.
+    \item open-source;
+    \item making data and documentation available;
+    \item using open-source libraries.
+\end{itemize}
+\end{itemize}
+
+\wss{Could ``reverse'' the list and list the recommendations and beside each the qualities 
+that that recommendation helps with - this would reduce repetition.}
+
+\subsection{Recommendations on Dealing With Limited Resources}
+\label{sec_recommendations_limited_resources}
+
+The limitation of resources has many faces, with the key manifestations being
+lack of fundings, time, and developers. We summarized our discussion with the MI
+software developers in Section~\ref{painpoints}.  Many of our recommendations
+involve investing more effort into processes, tools and methods, so that the
+limited resources can be used more effectively.  Our recommendations are as
+follows:
+
+\wss{Talk about increasing the number of developers.  CONTRIBUTING is rare
+(Table~\ref{artifactspresent}), but it doesn't have to be.  Related to this
+information for new developers is rare: Code Style Guide, Requirements, Design
+Doc, API.}
+
+\begin{itemize}
+\item \textbf{Identify the root cause.} More fundings or developers may not
+solve problems called by a lack of time. It is beneficial to identify the
+underlying obstacles to the team.  \wss{Is there something from the mythical man
+month that should be cited here?}
+
+\item \textbf{Maintain good documentation.} Creating and updating documentation
+consumes time, but can save much more time in the long term. If the users and
+developers can find answers to their questions themselves, they are less likely
+to abuse the team's issue tracker.  \wss{If it is an option, we could cite one
+of our productivity related papers.}
+
+\item \textbf{Adopt time-saving tools.} A good CI/CD tool (e.g.\ GitHub Actions)
+saves time for building and deploying the product, and automated tests can work
+in the background while developers are focusing on other tasks.
+
+\item \textbf{Use test-driven development process.} Many people think writing
+test cases is less rewarding than writing code, but without testing identifying
+and fixing bugs can consume substantial resources. Setting up the test cases
+costs time, but generates more benefits in the long run.
+
+\item \textbf{Consider supporting third-party plugins or extensions.} Why not
+let users share the burden? No software product can deliver every user's needs,
+and the large quantity of features leads to more bugs and maintenance problems.
+So it may be a good idea to shift some development and maintenance
+responsibilities to the users. The users may also be happy with the extra
+flexibility.
+
+\item \textbf{Consider ``hibernating'' for a while.} When there aren't enough
+developers, the team can shift from development mode toward maintenance mode.
+The team can stop building new features for a while and instead fix bugs and
+design problems from the past. If the development team can repay some of its
+technical debt \citep{KruchtenEtAl2012}, the software qualities will likely
+improve as a result.
+
+\item \textbf{Commercialization is not always toxic.} Licensing the software to
+commercial companies to use as internal modules of their products may bring
+financial support to the team. Meanwhile, the project can stay open-source for
+the community.
+\end{itemize}
+
+\subsection{Recommendations on Choosing A Tech Stack}
+\label{sec_recommendations_tech_stack}
+
+A tech stack refers to a set of technologies used by a team to build software
+and manage the project. Section~\ref{painpoints} lists the advantages and
+disadvantages between native and web applications. We give further suggestions
+on the choice of a tech stack to improve the four priority factors identified by
+developers: compatibility, maintainability, performance and security.  The
+suggestions are intended to provide ideas and avenues for exploration; not all
+of the suggestions will be the right fit for all projects and all teams.
+
+\begin{itemize}
+\item \textbf{Identify the priorities between the factors.} Simultaneously
+achieving high levels for all four factors (compatibility, maintainability,
+performance and security) is difficult. A team needs to prioritize its
+objectives according to its resource and experience.
+
+\item \textbf{Be open-minded about new technologies.} Web applications with only
+a frontend are known for worse performance than native applications. However,
+new technologies may ease this difference. For example, some JavaScript
+libraries can help the frontend harness the power of the computer's GPU and
+accelerate graphical computing. In addition, there are new frameworks helping
+developers with cross-platform compatibility. For example, the
+\href{https://flutter.dev/}{Flutter} project enables support for web, mobile,
+and desktop OS with one codebase.
+
+\item \textbf{Use git and GitHub.} As mentioned in
+Section~\ref{sec_score_maintainability}, almost all of the 29 MI software
+projects used git, and the majority of them used GitHub. We found from the
+projects' websites and our interviews with developers that, some projects moved
+from other version control tools to git and GitHub. GitHub provides convenient
+repository and project management, and OSS projects tend to receive more
+attention and contributions on GitHub.
+
+\item \textbf{Web applications can also deliver high performance.} Web
+applications with backend servers may perform even better than native
+applications. If a team needs to support lower-end computers, it is good to use
+back-end servers for heavy computing tasks.
+
+\item \textbf{Backend servers can have low costs.} Serverless solutions from
+major cloud service providers may be worth exploring. Serverless still uses a
+server, but the team is only charged when they use it. The solution is
+event-driven, and costs the team by the number of requests it processes. Thus,
+serverless can be very cost-effective for less intensively used functions.
+
+\item \textbf{Web transmission may diminish security.} Transferring sensitive
+data online can be a problem for projects requiring high security. Regulations
+for some MI applications may forbid doing web transmissions. In this case, a web
+application with a backend may not be an option.
+
+\item \textbf{Maintain good documentation.} No matter what tech stack a team
+uses, a well-maintained project plan, developer's manual, and API documentation
+generally help team members to contribute more and make fewer mistakes.
+\end{itemize}
+
+\subsection{Recommendations on Enriching the Testing Datasets} 
+\label{sec_recommendations_testing_dataset}
+
+As described in Section~\ref{painpoints}, it is difficult for software
+development teams to access real-world medical imaging datasets. This problem
+restricts their capability and flexibility for testing. We provide some
+suggestions as follows:
+
+\begin{itemize}
+\item \textbf{Build and maintain good connections to datasets.} A team can build
+connections with professionals working in the medical domain, who may have
+access to private datasets and can perform tests for the team. If a team has
+such professionals as internal members, the process can be simplified.
+
+\item \textbf{Collect and maintain datasets over time.} A team may face problems
+caused by various unique inputs over the years of software development. This
+data should be collected and maintained over time to form a good, comprehensive,
+dataset for testing.
+
+\item \textbf{Search for open data sources.} In general, there are many open MI
+datasets.  For instance, there are
+\href{https://nihcc.app.box.com/v/ChestXray-NIHCC}{Chest X-ray Datasets} by
+National Institute of Health \citep{WangEtAl2017},
+\href{https://www.cancerimagingarchive.net/}{Cancer Imaging Archive}
+\citep{PriorEtAl2017}, and \href{https://medpix.nlm.nih.gov/home}{MedPix} by
+National Library of Medicine \citep{Smirniotopoulos2014}. A team developing MI
+software should be able to find more open datasets according to their needs.
+
+\item \textbf{Create sample data for testing.} If a team can access tools
+creating sample data, they may also self-build datasets for testing. For
+example, an MI software development team can use an MRI scanner to create images
+of objects, animals, and volunteers. The team can build the images based on
+specific testing requirements.
+
+\item \textbf{Remove privacy from sensitive data.} For data with sensitive
+information, a team can ask the data owner to remove such information or add
+noise to protect privacy. One example is using de-identified copies of medical
+images for testing.
+
+\item \textbf{Establish community collaboration in the domain.} During our
+interviews with developers in the MI domain, we heard many stories of asking for
+supports from other professionals or equipment manufacturers. However, we
+believe that broader collaboration between development teams can address this
+problem better. Some datasets are too sensitive to share, but if the community
+has some kind of ``group discussion'', teams can better express their needs, and
+professionals can better offer voluntary support for testing. Ultimately, the
+community can establish a nonprofit organization as a third-party, which
+maintains large datasets, tests OSS in the domain, and protects privacy.
+
+\end{itemize}
 
@@
 
-\section{Recommendations} \label{ch_recommendations}
-
-This section presents our recommendations on MI software development. Although
-our focus is on MI software, unless noted otherwise, our recommendations apply
-to any scientific computing software.
-Section~\ref{sec_recommendations_qualities} discusses the actions that can
-potentially improve the ten software qualities.
-Sections~\ref{sec_recommendations_limited_resources},
-\ref{sec_recommendations_tech_stack}, and
-\ref{sec_recommendations_testing_dataset} are based on the primary pain points
-(from Section~\ref{painpoints}) collected from developers in the
-MI domain.
-
-\subsection{Recommendations on Improving Software Qualities}
-\label{sec_recommendations_qualities}
-
-Based on our quality measurements in Section~\ref{ch_results} and discussions
-with the developers in Section~\ref{painpoints}, we collected key points for
-improving software qualities. These points should be considered for new and
-existing projects, but we are not saying that every project should do everything
-on our list.  All projects will have finite resources, so strategies will have
-to be selected that provide the greatest return on investment.  Our specific
-recommendations for each quality are as follows:
-
-\begin{itemize}
-\item \textbf{Installability} (Section~\ref{sec_result_installability})
-\begin{itemize}
-    \item providing clear instructions;
-    \item automating installation;
-    \item including all dependencies in the installer;
-    \item avoiding heavily depending on other commercial products (e.g.\ Matlab);
-    \item potentially building a web application that needs no installation.
-\end{itemize}
-\item \textbf{Correctness \& Verifiability}
-(Section~\ref{sec_result_correctness_verifiability} and
-Section~\ref{sec_interview_software_qualities})
-\begin{itemize}
-    \item test-driven development with unit tests, integration tests, and
-    nightly tests;
-    \item two stage development process with stable release \& nightly builds;
-    \item CI/CD;
-    \item requirements specifications and theory manuals \citep{Smith2016, SmithAndLai2005}.
-    \item static code analysis tools (e.g. Lint and SonarQube)
-\end{itemize}
-\item \textbf{Reliability} (Section~\ref{sec_result_reliability})
-\begin{itemize}
-    \item test-driven development with unit tests, integration tests, and nightly tests.
-    \item two stage development process with stable release \& nightly builds;
-    \item descriptive error messages.
-\end{itemize}
-\item \textbf{Robustness} (Section~\ref{sec_result_robustness})
-\begin{itemize}
-    \item designing with exception handling so the software can fail gracefully;
-    \item descriptive error messages.
-\end{itemize}
-\item \textbf{Usability} (Section~\ref{sec_result_usability} and Section~\ref{sec_interview_software_qualities})
-\begin{itemize}
-    \item usability tests and interviews with end users;
-    \item adjusting according to users' feedbacks;
-    \item getting started tutorials;
-    \item user manuals;
-    \item professional UX (User eXperience) designs;
-    \item graphical user interface;
-    \item active supports to users.
-\end{itemize}
-\item \textbf{Maintainability} (Section~\ref{sec_score_maintainability} and Section~\ref{sec_interview_software_qualities})
-\begin{itemize}
-    \item use GitHub (or equivalent version control and project management environment);
-    \item use a modular approach to design (we advocate basing the design on
-    Parnas's principle of information hiding: ``system details that are likely
-    to change independently should be the secrets of separate modules; the only
-    assumptions that should appear in the interfaces between modules are those
-    that are considered unlikely to change.'' \citep{ParnasEtAl2000})
-    \item documentation for developers: project plan, developer’s manual, and API documentation.
-\end{itemize}
-\item \textbf{Reusability} (Section~\ref{sec_result_reusability})
-\begin{itemize}
-    \item modular approach;
-    \item API documentation;
-    \item tools that generate software documentation for developers (e.g.\
-    Doxygen, Javadoc, and Sphinx).
-\end{itemize}
-\item \textbf{Understandability} (Section~\ref{sec_result_understandability} and
-Section~\ref{sec_interview_software_qualities})
-\begin{itemize}
-    \item modular approach;
-    \item good coding style: consistent indentation and formatting style;
-    consistent, distinctive, and meaningful code identifiers; keeping parameters
-    in the same order for all functions; avoiding hard-coded constants (other
-    than 0 and 1);
-    \item clear comments, indicating what is being done, not how;
-    \item description of algorithms used;
-    \item documentation of explicit requirements on coding standards;
-    \item communication between developers and users via GitHub issues, mailing
-    lists, and forums.
-\end{itemize}
-\item \textbf{Visibility/Transparency} (Section~\ref{sec_result_visibility_transparency})
-\begin{itemize}
-    \item documents for the development process, contributor's guide, project
-    status, development environment, and release notes.
-\end{itemize}
-\item \textbf{Reproducibility} (Section~\ref{sec_interview_software_qualities})
-\begin{itemize}
-    \item test-driven development with unit tests, integration tests, and nightly tests.
-    \item open-source;
-    \item making data and documentation available;
-    \item using open-source libraries.
-\end{itemize}
-\end{itemize}
-
-\wss{Could ``reverse'' the list and list the recommendations and beside each the qualities 
-that that recommendation helps with - this would reduce repetition.}
-
-\subsection{Recommendations on Dealing With Limited Resources}
-\label{sec_recommendations_limited_resources}
-
-The limitation of resources has many faces, with the key manifestations being
-lack of fundings, time, and developers. We summarized our discussion with the MI
-software developers in Section~\ref{painpoints}.  Many of our recommendations
-involve investing more effort into processes, tools and methods, so that the
-limited resources can be used more effectively.  Our recommendations are as
-follows:
-
-\wss{Talk about increasing the number of developers.  CONTRIBUTING is rare
-(Table~\ref{artifactspresent}), but it doesn't have to be.  Related to this
-information for new developers is rare: Code Style Guide, Requirements, Design
-Doc, API.}
-
-\begin{itemize}
-\item \textbf{Identify the root cause.} More fundings or developers may not
-solve problems called by a lack of time. It is beneficial to identify the
-underlying obstacles to the team.  \wss{Is there something from the mythical man
-month that should be cited here?}
-
-\item \textbf{Maintain good documentation.} Creating and updating documentation
-consumes time, but can save much more time in the long term. If the users and
-developers can find answers to their questions themselves, they are less likely
-to abuse the team's issue tracker.  \wss{If it is an option, we could cite one
-of our productivity related papers.}
-
-\item \textbf{Adopt time-saving tools.} A good CI/CD tool (e.g.\ GitHub Actions)
-saves time for building and deploying the product, and automated tests can work
-in the background while developers are focusing on other tasks.
-
-\item \textbf{Use test-driven development process.} Many people think writing
-test cases is less rewarding than writing code, but without testing identifying
-and fixing bugs can consume substantial resources. Setting up the test cases
-costs time, but generates more benefits in the long run.
-
-\item \textbf{Consider supporting third-party plugins or extensions.} Why not
-let users share the burden? No software product can deliver every user's needs,
-and the large quantity of features leads to more bugs and maintenance problems.
-So it may be a good idea to shift some development and maintenance
-responsibilities to the users. The users may also be happy with the extra
-flexibility.
-
-\item \textbf{Consider ``hibernating'' for a while.} When there aren't enough
-developers, the team can shift from development mode toward maintenance mode.
-The team can stop building new features for a while and instead fix bugs and
-design problems from the past. If the development team can repay some of its
-technical debt \citep{KruchtenEtAl2012}, the software qualities will likely
-improve as a result.
-
-\item \textbf{Commercialization is not always toxic.} Licensing the software to
-commercial companies to use as internal modules of their products may bring
-financial support to the team. Meanwhile, the project can stay open-source for
-the community.
-\end{itemize}
-
-\subsection{Recommendations on Choosing A Tech Stack}
-\label{sec_recommendations_tech_stack}
-
-A tech stack refers to a set of technologies used by a team to build software
-and manage the project. Section~\ref{painpoints} lists the advantages and
-disadvantages between native and web applications. We give further suggestions
-on the choice of a tech stack to improve the four priority factors identified by
-developers: compatibility, maintainability, performance and security.  The
-suggestions are intended to provide ideas and avenues for exploration; not all
-of the suggestions will be the right fit for all projects and all teams.
-
-\begin{itemize}
-\item \textbf{Identify the priorities between the factors.} Simultaneously
-achieving high levels for all four factors (compatibility, maintainability,
-performance and security) is difficult. A team needs to prioritize its
-objectives according to its resource and experience.
-
-\item \textbf{Be open-minded about new technologies.} Web applications with only
-a frontend are known for worse performance than native applications. However,
-new technologies may ease this difference. For example, some JavaScript
-libraries can help the frontend harness the power of the computer's GPU and
-accelerate graphical computing. In addition, there are new frameworks helping
-developers with cross-platform compatibility. For example, the
-\href{https://flutter.dev/}{Flutter} project enables support for web, mobile,
-and desktop OS with one codebase.
-
-\item \textbf{Use git and GitHub.} As mentioned in
-Section~\ref{sec_score_maintainability}, almost all of the 29 MI software
-projects used git, and the majority of them used GitHub. We found from the
-projects' websites and our interviews with developers that, some projects moved
-from other version control tools to git and GitHub. GitHub provides convenient
-repository and project management, and OSS projects tend to receive more
-attention and contributions on GitHub.
-
-\item \textbf{Web applications can also deliver high performance.} Web
-applications with backend servers may perform even better than native
-applications. If a team needs to support lower-end computers, it is good to use
-back-end servers for heavy computing tasks.
-
-\item \textbf{Backend servers can have low costs.} Serverless solutions from
-major cloud service providers may be worth exploring. Serverless still uses a
-server, but the team is only charged when they use it. The solution is
-event-driven, and costs the team by the number of requests it processes. Thus,
-serverless can be very cost-effective for less intensively used functions.
-
-\item \textbf{Web transmission may diminish security.} Transferring sensitive
-data online can be a problem for projects requiring high security. Regulations
-for some MI applications may forbid doing web transmissions. In this case, a web
-application with a backend may not be an option.
-
-\item \textbf{Maintain good documentation.} No matter what tech stack a team
-uses, a well-maintained project plan, developer's manual, and API documentation
-generally help team members to contribute more and make fewer mistakes.
-\end{itemize}
-
-\subsection{Recommendations on Enriching the Testing Datasets} 
-\label{sec_recommendations_testing_dataset}
-
-As described in Section~\ref{painpoints}, it is difficult for
-software development teams to access real-world medical imaging datasets. This
-problem restricts their capability and flexibility for testing. We provide some
-suggestions as follows:
-
-\begin{itemize}
-\item \textbf{Build and maintain good connections to datasets.} A team can build
-connections with professionals working in the medical domain, who may have
-access to private datasets and can perform tests for the team. If a team has
-such professionals as internal members, the process can be simplified.
-
-\item \textbf{Collect and maintain datasets over time.} A team may face problems
-caused by various unique inputs over the years of software development. This
-data should be collected and maintained over time to form a good, comprehensive,
-dataset for testing.
-
-\item \textbf{Search for open data sources.} In general, there are many open MI
-datasets.  For instance, there are
-\href{https://nihcc.app.box.com/v/ChestXray-NIHCC}{Chest X-ray Datasets} by
-National Institute of Health \citep{WangEtAl2017},
-\href{https://www.cancerimagingarchive.net/}{Cancer Imaging Archive}
-\citep{PriorEtAl2017}, and \href{https://medpix.nlm.nih.gov/home}{MedPix} by
-National Library of Medicine \citep{Smirniotopoulos2014}. A team developing MI
-software should be able to find more open datasets according to their needs.
-
-\item \textbf{Create sample data for testing.} If a team can access tools
-creating sample data, they may also self-build datasets for testing. For
-example, an MI software development team can use an MRI scanner to create images
-of objects, animals, and volunteers. The team can build the images based on
-specific testing requirements.
-
-\item \textbf{Remove privacy from sensitive data.} For data with sensitive
-information, a team can ask the data owner to remove such information or add
-noise to protect privacy. One example is using de-identified copies of medical
-images for testing.
-
-\item \textbf{Establish community collaboration in the domain.} During our
-interviews with developers in the MI domain, we heard many stories of asking for
-supports from other professionals or equipment manufacturers. However, we
-believe that broader collaboration between development teams can address this
-problem better. Some datasets are too sensitive to share, but if the community
-has some kind of ``group discussion'', teams can better express their needs, and
-professionals can better offer voluntary support for testing. Ultimately, the
-community can establish a nonprofit organization as a third-party, which
-maintains large datasets, tests OSS in the domain, and protects privacy.
-
-\end{itemize}
-
 \section{Conclusions} \label{ch_conclusions}
@@
 \item using git and GitHub;
-\item modular approach with the design principle proposed by \citet{ParnasEtAl2000};
+\item modular approach with the design principle proposed by
+\citet{ParnasEtAl2000};
 \item considering newer technologies (e.g.\ web application and serverless
@@
         and domain experts.
-        \item surface understandability: our current method does not
-        require understanding the source code. As software engineers, perhaps we
-        can select a small module of each project, read the source code and
+        \item surface understandability: our current method does not require
+        understanding the source code. As software engineers, perhaps we can
+        select a small module of each project, read the source code and
         documentation, try to understand the logic, and score the ease of the
-        process.  Ideas for getting started are available in \citet{SmithEtAl2021}.
+        process.  Ideas for getting started are available in
+        \citet{SmithEtAl2021}.
         \item measure modifiability as part of the measurement of
@@
 	\begin{itemize}
-	    \item in {Q14}, ``Do you think improving this process can
-	    tackle the current problem?'' is a yes-or-no question, which is not
-	    informative enough. As mentioned in Section~\ref{sec_contribution_pm},
-	    most interviewees ignored it. We can change it to ``By improving this
+	    \item in {Q14}, ``Do you think improving this process can tackle the
+	    current problem?'' is a yes-or-no question, which is not informative
+	    enough. As mentioned in Section~\ref{Sec_CompareMethodologies}, most
+	    interviewees ignored it. We can change it to ``By improving this
 	    process, what current problems can be tackled?''; 
@@
 simulate mathematical models of real world systems of engineering or scientific
-importance so that we can better understand and predict the system’s behaviour''
+importance so that we can better understand and predict the system's behaviour''
 \citep{Smith2006}. Many researchers consider SC as the third pillar of science

### 2022-04-13 StateOfPractice/Papers/ICCS-2022-SOP_MethodologyPlusExamples/DigDeep.tex
Final version of digging deeper paper and conference registration and copyright transfer

--- 2022-02-18 StateOfPractice/Papers/ICCS-2022-SOP_MethodologyPlusExamples/DigDeep.tex
@@
 
-	We have developed a methodology for assessing the state of the software
-	development practice for a given research software domain. Our methodology
+	To improve the productivity of research software developers we need to first
+	understand their development practices.  Previous studies on this topic have
+	collected data by surveying as many developers as possible, across a broad
+	range of application domains.  We propose to dig deeper into the state of
+	the practice by instead looking at what developers in specific domains
+	create, as evidenced by the contents of their repositories.  Our methodology
 	prescribes the following steps: i) Identify the domain; ii) Identify a list
-	of candidate software packages; iii) Filter the list to a length of about 30
-	packages; iv) Collect repository related data on each software package, like
-	number of stars, number of open issues, number of lines of code; v) Fill in
-	the measurement template (the template consists of 108 questions to assess 9
+	of candidate software; iii) Filter the list to a length of about 30
+	packages; iv) Collect repository related data on each package, like number
+	of stars, number of open issues, number of lines of code; v) Fill in the
+	measurement template (the template consists of 108 questions to assess 9
 	qualities (including the qualities of installability, usability and
@@
 	takes about an hour); and, viii) Conduct a domain analysis. The collected
-	data is analyzed by: i) comparing the ranking by best practices against 
-	the ranking by popularity; ii) comparing artifacts, tools and processes to
+	data is analyzed by: i) comparing the ranking by best practices against the
+	ranking by popularity; ii) comparing artifacts, tools and processes to
 	current research software development guidelines; and, iii) exploring pain
-	points. We estimate the time to complete an assessment for a given domain at
-	173 person hours. The method is illustrated via the example of Lattice
-	Boltzmann Solvers, where we find that the top packages engaged in most of
-	recommended best practices, but still show room for improvement with respect
-	to providing API documentation, a roadmap, a code of conduct, programming
-	style guide, uninstall instructions and continuous integration.
+	points. We estimate the time to complete an assessment at 173 person hours.
+	The method is illustrated via the example of Lattice Boltzmann Solvers,
+	where we find that the top packages engaged in most of recommended best
+	practices, but still show room for improvement with respect to providing API
+	documentation, a roadmap, a code of conduct, programming style guide and
+	continuous integration.
 
-\keywords{research software \and software quality \and empirical measures \and analytic hierarchy process \and software artifacts \and developer pain points \and Lattice Boltzmann Method}
+\keywords{research software \and software quality \and empirical measures \and  software artifacts \and developer pain points \and Lattice Boltzmann Method}
 
@@
 and treatment.  However, developing reliable, reproducible, sustainable and fast
-research software to address new problems is challenging because of the
-complexity of the physical models and the nuances of floating point and parallel
-computation. The importance of research software and the difficulty with its
-development have prompted multiple researchers to investigate how developing
-this software differs from other classes of software.  Previous studies have
-focused on surveying developers
-\cite{HannayEtAl2009,Nguyen-HoanEtAl2010,PintoEtAl2018}, developer interviews
-\cite{Kelly2013} and case studies \cite{CarverEtAl2007,Segal2005}.  A
-valuable source of information that has received less attention is the data in
-publicly available software repositories.  We propose digging deeper into how
-well projects are applying best practices by using a new methodology that
-combines manual and automated techniques to extract repository-based
-information.
+research software is challenging because of the complexity of physical models
+and the nuances of floating point and parallel computation. The critical
+importance of research software, and the challenges with its development, have
+prompted multiple researchers to investigate the state of development practice.
+Understanding the creation of research software is critical for devising future
+methods and tools for reducing development time and improving software quality.
+
+Previous studies on the state of the practice for research software have often
+focused on surveying
+developers~\cite{HannayEtAl2009,Nguyen-HoanEtAl2010,PintoEtAl2018}. Although
+surveys provide valuable information, they are limited by their reliance on what
+developers say they do, rather than directly measuring what they actually do.
+Therefore, we propose a state of the practice assessment methodology that
+investigates the work products developers create by digging into the
+software repositories they create.
+
+Although other studies~\cite{GrannanEtAl2020,SoodEtAl2019} mine research
+software repositories to estimate productivity, code quality and project
+popularity, the other studies focus on automation and on code related artifacts.
+To gain deeper insight, at the expense of taking more time, we will relax the
+automation requirement and use manual investigation where necessary.  We will
+also expand our assessment beyond code to include other artifacts, where
+artifacts are the documents, scripts and code that we find in a project's public
+repository. Example artifacts include requirements, specifications, user
+manuals, unit test cases, system tests, usability tests, build scripts, API
+(Application Programming Interface) documentation, READMEs, license documents,
+and process documents.
 
 The surveys used in previous studies have tended to recruit participants from
-all domains of research software, possibly distinguishing them by programming
-language (for instance, R developers~\cite{PintoEtAl2018}), or by the role of
-the developers (for instance postdoctoral researchers \cite{UditAndKatz2017}).
-Case studies, on the other hand, have focused on a few specific examples at a
-time, as is the nature of case studies.  For our new methodology, we propose a
-scope between these two extremes.  Rather than focus on all research software,
-or just a few examples, we will focus on one domain at a time. The practical
-reason for this scope is that digging deep takes time, making a broad scope
-infeasible. We have imposed a practical constraint of one person month of effort
-per domain.\footnote{A person month is considered to be $20$ working days ($4$
-weeks in a month, with $5$ days of work per week) at $8$ person hours per day,
-or $20 \cdot 8 = 160$ person hours.} Focusing on one domain at a time has more
-than just practical advantages.  By restricting ourselves to a single domain we
-can bring domain knowledge and domain experts into the mix.  The domain
-customized insight provided by the assessment has the potential to help a
-specific domain as they adopt and develop new practices.  Moreover, measuring
-multiple different domains facilitates comparing and contrasting domain specific
-practices.
+all domains of research software.  This is what we will call \emph{research
+software in general}, as opposed to specific domain software. The surveys may
+distinguish participants by programming language (for instance, R
+developers~\cite{PintoEtAl2018}), or by the role of the developers (for instance
+postdoctoral researchers \cite{UditAndKatz2017}), but the usual goal is to cast
+as wide a net as possible. Case studies \cite{CarverEtAl2007,Segal2005}, on the
+other hand, go more in depth by focusing on a few specific examples at a time.
+For our new methodology, we propose a scope between these two extremes.  Rather
+than focus on assessing the state of the practice for research software in
+general, or just a few examples, we will focus on one scientific domain at a
+time. The practical reason for this scope is that digging deep takes time,
+making a broad scope infeasible. We have imposed a practical constraint of one
+person month of effort per domain.\footnote{A person month is considered to be
+$20$ working days ($4$ weeks in a month, with $5$ days of work per week) at $8$
+person hours per day, or $20 \cdot 8 = 160$ person hours.}  Focusing on one
+domain at a time has more than just practical advantages.  By restricting
+ourselves to a single domain we can bring domain knowledge and domain experts
+into the mix.  The domain customized insight provided by the assessment has the
+potential to help a specific domain as they adopt and develop new practices.
+Moreover, measuring multiple different domains facilitates comparing and
+contrasting domain specific practices.
 
-Our methodology is built around 10 research questions.  Assuming that the domain
+Our methodology is built around 8 research questions.  Assuming that the domain
 has been identified (Section~\ref{SecIdentifyDomain}), the first question is:
@@
 We next wish to assess the representative software to determine how well they
-apply current software development best practices.  At this point in the
-process, to remove potential user/developer bias, we will base our assessment
-only on publicly available artifacts, where artifacts are the documents, scripts
-and code that we find in a project's public repository. Example artifacts
-include requirements, specifications, user manuals, unit test cases, system
-tests, usability tests, build scripts, API (Application Programming Interface)
-documentation, READMEs, license documents, process documents, and code.
-Following best practices does not guarantee popularity, so we will also compare
-our ranking to how the user community itself ranks the identified projects.
+apply current software development best practices.  By \emph{best practices} we
+mean methods, techniques and tools that are generally believed to improve
+software development, like testing and documentation.  As we will discuss in
+Section~\ref{empiricalmeasures}, we will categorize our best practices around
+software qualities. Following best practices does not guarantee popularity, so
+we will also compare our ranking to how the user community itself ranks the
+identified projects.
 
@@
 	practices, based on evidence found by experimenting with the software and
-	searching the artifacts available in each project's repository? (Sections~\ref{empiricalmeasures})
+	searching the artifacts available in each project's repository?
+	(Sections~\ref{empiricalmeasures})
 	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareHQ2Popular}:] How
 	similar is the list of top projects identified in \rqref{RQ_HighestQuality}
-	to the most popular projects, as viewed by the scientific community? (Section~\ref{repmetrics})
+	to the most popular projects, as viewed by the scientific community?
+	(Section~\ref{repmetrics})
 \end{enumerate}
@@
 	How do domain projects compare to research software in general with respect to
-	the use of tools (Section~\ref{Sec_CompareTools}) for:
-	\begin{enumerate} 
-		\item [\rqref{RQ_CompareToolsProjMngmnt}.a] development; and,
-		\item [\rqref{RQ_CompareToolsProjMngmnt}.b] project management?
-	\end{enumerate}
+	the use of tools? (Section~\ref{Sec_CompareTools})
 	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareMethodologies}:]
@@
 
-Only so much information can be gleaned by digging into software repos.
-To gain additional insight, we need to interview developers
-(Section~\ref{SecSurvey}) to learn:
+Only so much information can be gleaned by digging into software repos. To gain
+additional insight, we need to interview developers (Section~\ref{SecSurvey}),
+as done in other state of the practice assessments \cite{Kelly2013}, to learn:
 
@@
 	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_PainPoints}:] What are
-	the pain points for developers working on domain software projects?
+	the pain points for developers?
 	(Section~\ref{painpoints})
@@
 Boltzmann Method (LBM) software \cite{Michalski2021}.  The LBM example will be
-used throughout this paper to illustrate the steps in the methodology.  The
-paper ends with a summary of potential threats to validity
-(Section~\ref{threats}) and our conclusions
-(Section~\ref{SecConcludingRemarks}).
+used to illustrate the steps in the methodology.
 
@@
 
-The assessment is conducted via the following steps.  The steps depend on
+The assessment is conducted via the following steps, which depend on
 interaction with a Domain Expert partner, as discussed in
@@
 	\item Gather the source code and documentation for each software package.
-	\item Collect quantitative measures from the project repositories.
-	(Section~\ref{empiricalmeasures})
 	\item Measure using the measurement template. (Section~\ref{empiricalmeasures})
@@
 	\item Analyze the results and answer the research questions.
-(Sections~\ref{repmetrics}---\ref{painpoints}) %,~\ref{Sec_CompareArtifacts},~\ref{Sec_CompareTools},~\ref{Sec_CompareMethodologies},~\ref{painpoints},~\ref{Sec_AddressConcerns},
+	(Sections~\ref{SecRanking}---\ref{painpoints})
 \end{enumerate}
@@
 We estimate 173 hours to complete the assessment of a given
-domain~\cite{SmithEtAl2021}, which is close our goal of 160 person hours. This
-estimate assumes the domain has been decided, the Domain Expert has been
-recruited, and the pre-existing template spreadsheet and AHP
-tool~\cite{SmithEtAl2021}, will be utilized.
+domain~\cite{SmithEtAl2021}, which is close our goal of 160 person hours.
+% This estimate assumes the domain has been decided, the Domain Expert has been
+% recruited, and the pre-existing template spreadsheet and AHP
+% tool~\cite{SmithEtAl2021}, will be utilized.
 
@@
 
-To be applicable for the methodology described in this document, the chosen
-domain must have the following properties:
+To be applicable the chosen domain must have the following properties:
 
 \begin{enumerate}
-\item The domain must have well-defined and stable theoretical underpinning.
+\item The domain must have well-defined and stable theoretical underpinnings.
 \item There must be a community of people studying the domain.
 \item The software packages must have open source options.
-\item A preliminary search, or discussion with experts, suggests that there will
-  be around 30 or more candidate software packages.
+\item The domain expert says there will be at least 30 candidate packages.
 \end{enumerate}
@@
 
-The candidate software to answer~\rqref{RQ_WhatProjects} should be found through
+The candidate software to answer \rqref{RQ_WhatProjects} should be found through
 search engine queries,  
@@
 
-For the LBM example the initial list had 46 packages~\cite{Michalski2021}. This
-list was filtered by scope, usage, and age to decrease the length to 24.
-
 \subsection{Quantitative Measures} \label{empiricalmeasures}
@@
 the template is shown in Figure~\ref{measurement_template_image}.  In keeping
-with scientific transparency, all data should be made available in a public
-repository.
+with scientific transparency, all data should be made publicly available.
 
@@
 number of lines of code (LOC), percentage of issues that are closed, etc. The
-tool \href{https://github.com/tomgi/git_stats}{GitStats} is used to measure
-each software package's GitHub repository for the number of binary files, the
-number of added and deleted lines, and the number of commits over varying time
+tool \href{https://github.com/tomgi/git_stats}{GitStats} is used to measure each
+software package's GitHub repository for the number of binary files, the number
+of added and deleted lines, and the number of commits over varying time
 intervals. The tool \href{https://github.com/boyter/scc}{Sloc Cloc and Code
 (scc)} is used to measure the number of text based files as well as the number
-of total, code, comment, and blank lines in each GitHub repository.
+of total, code, comment, and blank lines in each GitHub repository.  At this
+time our focus is on simple metrics, so we do no need the static analysis
+capabilities of tools like
+\href{https://www.hello2morrow.com/products/sonargraph} {Sonargraph} or
+\href{https://www.sonarcloud.io}{SonarCloud}.
 
@@
 criteria~\cite{SmithEtAl2016}.
-
-\begin{figure}[h!]
-	\centering
-		\includegraphics[width=1.0\textwidth]{./figures/finalscore_chart.pdf}
-		\caption{AHP Overall Score}
-		\label{Fig_OverallScore}
-\end{figure}
-
-Figure~\ref{Fig_OverallScore} shows the overall ranking of the LBM software
-packages. In the absence of other information on priorities, the overall ranking
-was calculated by assuming an equal weighting between all qualities.  The LBM data is available \href{https://data.mendeley.com/datasets/5dym63wn6z/1} {on-line}.
 
@@
 
-For the current methodology time constraints necessitate a shallow domain
-analysis. For each domain a table should be constructed that distinguishes the
-programs under study by their variabilities.  In research software the
-variabilities are often assumptions. Table~\ref{tbl_features} shows the
-variabilities for LBM software~\cite{Michalski2021}.
+For each domain a table should be constructed that distinguishes the programs by
+their variabilities.  In research software the variabilities are often
+assumptions.
+
+\section{Measuring and Ranking the LBM Domain} \label{SecRanking}
+
+For the LBM example the initial list had 46 packages~\cite{Michalski2021}. To
+answer \rqref{RQ_WhatProjects}, this list was filtered by scope, usage, and age
+to decrease the length to 24 packages, as shown in Table~\ref{tbl_features}.
+This table also shows the domain analysis, in terms of variabilities, for LBM
+software~\cite{Michalski2021}.
 
@@
 
+To answer \rqref{RQ_HighestQuality} the 24 LBM packages were ranked for the 9
+qualities~\cite{Michalski2021}.  For space reasons we will only show the overall
+ranking in Figure~\ref{Fig_OverallScore}. In the absence of information on
+priorities, the overall ranking was calculated with an equal weight between all
+qualities.  The LBM data is available on
+\href{https://data.mendeley.com/datasets/5dym63wn6z/1} {Mendeley}.
+
+\begin{figure}[h!]
+	\centering
+		\includegraphics[width=1.0\textwidth]{./figures/finalscore_chart.pdf}
+		\caption{AHP Overall Score}
+		\label{Fig_OverallScore}
+\end{figure}
+
 \section{Comparison to Community Ranking} \label{repmetrics}
@@
 	of equal weighting between quality factors;
-	\item stars are known to not be a particularly good measure of popularity,
+	\item stars are not a particularly good measure of popularity,
 	because of how people use stars and because young projects have less time to
-	accumulate stars~\cite{Szulik2017}; 
+	accumulate them~\cite{Szulik2017}; 
 	\item and, as for consumer products, there are more factors influencing
@@
 popularity in the research community.  Nine packages do not use GitHub, so they
-do not have a measure of stars. Looking at the stars of the other 15 packages,
+do not have a star measure. Looking at the stars of the other 15 packages,
 we can observe a pattern where packages that have been highly ranked by our
@@
 			Palabos & 3 & 34 & 6 & GitLab& GitLab\\
-			OpenLB & 4 & No Git & No Git & No Git& No Git\\
+			OpenLB & 4 & N/A & N/A & N/A& N/A\\
 			LUMA & 5 & 33 & 7 & 12& 4\\
 			pyLBM & 6 & 95 & 3 & 10& 5\\
-			DL\_MESO (LBE) & 7 & No Git & No Git & No Git & No Git\\
-			Musubi & 8 & No Git & No Git & No Git & No Git\\
+			DL\_MESO (LBE) & 7 & N/A & N/A & N/A & N/A\\
+			Musubi & 8 & N/A & N/A & N/A & N/A\\
 			Sailfish & 9 & 186 & 1 & 41& 1\\
-			waLBerla & 10 & 20 & 9 & GitLab& GitLab\\
-			laboetie & 11 & 4 & 13 & 5& 8\\
-			TCLB & 12 & 95 & 3 & 16& 3\\
-			% MechSys & 13 & No Git & No Git & No Git& No Git\\
+			% waLBerla & 10 & 20 & 9 & GitLab& GitLab\\
+			% laboetie & 11 & 4 & 13 & 5& 8\\
+			% TCLB & 12 & 95 & 3 & 16& 3\\
+			% % MechSys & 13 & N/A & N/A & N/A& N/A\\
 			% lettuce & 14 & 48 & 4 & 5& 8\\
@@
 			% MP-LABS & 16 & 12 & 11 & 2& 9\\			
-			% SunlightLB & 17 & No Git & No Git & No Git& No Git\\
-			% LB3D & 18 & No Git & No Git & No Git& No Git\\			
-			% LIMBES & 19 & No Git & No Git & No Git& No Git\\
-			% LB2D-Prime & 20 & No Git & No Git & No Git& No Git\\		
+			% SunlightLB & 17 & N/A & N/A & N/A& N/A\\
+			% LB3D & 18 & N/A & N/A & N/A& N/A\\			
+			% LIMBES & 19 & N/A & N/A & N/A& N/A\\
+			% LB2D-Prime & 20 & N/A & N/A & N/A& N/A\\		
 			% HemeLB & 21 & 12 & 11 & 12& 4\\
 			% lbmpy & 22 &  11 & 12 & 2& 9\\	
-			% LB3D-Prime & 23 & No Git & No Git & No Git& No Git\\	
+			% LB3D-Prime & 23 & N/A & N/A & N/A& N/A\\	
 			... & ... & ... & ... & ... & ...\\				
@@
 domain software is examined for the presence of artifacts, which are then
-categorized by frequency as: common (more than 2/3 of projects), less common
+categorized by frequency as: common (more than 2/3 of projects), uncommon
 (between 1/3 and 2/3), and rare (less than 1/3). The observed frequency of
@@
 
-\begin{table}[!ht]
+As Table~\ref{Tbl_Guidelines} shows, the majority of LBM generated artifacts
+correspond to general recommendations from research software developers.  Areas
+where LBM developers could improve include providing: API documentation, a
+product roadmap, a code of conduct, code style guide, and uninstall
+instructions.
+
+\begin{table}[!h]
 	\begin{center}
-\begin{tabular}{ p{3cm}p{1cm}p{1cm}p{1cm}p{1cm}p{1cm}p{1cm}p{1cm}p{1cm}p{1cm} }
+	\begin{tabular}{ p{2.5cm}p{1cm}p{1cm}p{1cm}p{1cm}p{1cm}p{1cm}p{1cm}p{1cm}p{1cm}p{0.8cm} }
 	\toprule
-~ \ & \cite{USGS2019} & \cite{TobiasEtAl2018} & \cite{BrettEtAl2021} & \cite{WilsonEtAl2016} & \cite{SmithAndRoscoe2018} & \cite{HerouxEtAl2008} & \cite{ThielEtAl2020} & \cite{vanGompelEtAl2016} & \cite{OrvizEtAl2017}\\
+	~ \ & \cite{USGS2019} & \cite{TobiasEtAl2018} & \cite{BrettEtAl2021} & \cite{WilsonEtAl2016} & \cite{SmithAndRoscoe2018} & \cite{HerouxEtAl2008} & \cite{ThielEtAl2020} & \cite{vanGompelEtAl2016} & \cite{OrvizEtAl2017} & LBM\\
 	\midrule
-LICENSE & \checkmark &  & \checkmark & \checkmark & \checkmark & & \checkmark & \checkmark & \checkmark \\
-README &  &  & \checkmark & \checkmark & \checkmark & & \checkmark & \checkmark & \checkmark\\
-CONTRIBUTING &  &  & \checkmark & \checkmark & \checkmark & & \checkmark & \checkmark & \checkmark \\
-CITATION &  &  &  & \checkmark & & & & \checkmark & \checkmark \\
-CHANGELOG &  &  &  & \checkmark & \checkmark & & \checkmark & & \\
-INSTALL &  &  &  &  & \checkmark & & \checkmark & \checkmark & \checkmark\\
-Uninstall &  &  &  &  & & & & \checkmark & \\
-Getting started &  &  &  &  & \checkmark & & \checkmark & \checkmark & \checkmark\\
-User manual &  &  & \checkmark &  & & & \checkmark & & \\
-Tutorials &  &  &  &  & & & \checkmark & & \\
-FAQ &  &  &  &  & & & \checkmark & \checkmark & \checkmark\\
-Dependency List &  &  & \checkmark & & \checkmark & & & \checkmark & \\
-Issue Track &  & \checkmark & \checkmark & & \checkmark & \checkmark & \checkmark & & \checkmark \\
-Version Control &  & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark\\ 
-Requirements &  & \checkmark &  &  & & \checkmark &  &  & \checkmark \\
-Design &  & \checkmark  & \checkmark &  & \checkmark & & \checkmark & \checkmark& \checkmark\\
-API Doc. &  &  &  &  & \checkmark & & \checkmark & \checkmark & \checkmark\\
-Build Scripts &  & \checkmark &  & \checkmark & \checkmark & \checkmark & \checkmark & & \checkmark \\
-Unit Tests & \checkmark & \checkmark & \checkmark &  & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
-Test Plan &  & \checkmark &  &  & & \checkmark & & & \\
-Integ. Tests &  & \checkmark & \checkmark &  & & & & \checkmark & \checkmark \\
-System Tests &  & \checkmark & \checkmark &  & & \checkmark & & \checkmark & \checkmark \\
-Acceptance Tests &  & \checkmark &  &  & & & & &  \\
-Regression Tests &  &  & \checkmark &  & & \checkmark & & & \checkmark \\
-Code Style Guide &  & \checkmark &  &  & & & \checkmark & \checkmark & \checkmark \\
-Release Info. &  & \checkmark &  &  & & \checkmark & \checkmark & & \\
-Product Roadmap &  &  &  &  & & \checkmark & \checkmark & \checkmark & \\
-Code of Conduct &  &  &  &  & & & \checkmark & & \\
+	LICENSE & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & & \checkmark & \checkmark & \checkmark & C\\
+	README &  & \checkmark & \checkmark & \checkmark & \checkmark & & \checkmark & \checkmark & \checkmark & C\\
+	CONTRIBUTING &  & \checkmark & \checkmark & \checkmark & \checkmark & & \checkmark & \checkmark & \checkmark & U\\
+	CITATION &  &  &  & \checkmark & & & & \checkmark & \checkmark & U\\
+	CHANGELOG &  & \checkmark &  & \checkmark & \checkmark & & \checkmark &  &  & U\\
+	INSTALL &  &  &  &  & \checkmark & & \checkmark & \checkmark & \checkmark & C\\
+	\midrule
+	Uninstall &  &  &  &  & & & & \checkmark & & R\\
+	Dependency List &  &  & \checkmark & & \checkmark & & & \checkmark &  & C\\
+	Authors &  &  &  &  &  &  & \checkmark & \checkmark & \checkmark & C\\
+	Code of Conduct &  &  &  &  & & & \checkmark & & & \\
+	Acknowledgements &  &  &  &  &  &  & \checkmark & \checkmark & \checkmark & R\\
+	Code Style Guide &  & \checkmark &  &  & & & \checkmark & \checkmark & \checkmark & R\\
+	Release Info. &  & \checkmark &  &  & & \checkmark & \checkmark & & & U\\
+	Prod.\ Roadmap &  &  &  &  & & \checkmark & \checkmark & \checkmark & & R\\
+	\midrule
+	Getting started &  &  &  &  & \checkmark & & \checkmark & \checkmark & \checkmark & C\\
+	User manual &  &  & \checkmark &  & & & \checkmark & & & U\\
+	Tutorials &  &  &  &  & & & \checkmark & & & C\\
+	FAQ &  &  &  &  & & & \checkmark & \checkmark & \checkmark & R\\
+	\midrule
+	Issue Track &  & \checkmark & \checkmark & & \checkmark & \checkmark & \checkmark & & \checkmark & C\\
+	Version Control &  & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & C\\ 
+	Build Scripts &  & \checkmark &  & \checkmark & \checkmark & \checkmark &
+	\checkmark & & \checkmark & C\\
+	\midrule
+	Requirements &  & \checkmark &  &  & & \checkmark &  &  & \checkmark &  \\
+	Design Doc.\ &  & \checkmark  & \checkmark &  & \checkmark & & \checkmark &
+	\checkmark& \checkmark & U\\
+	API Doc. &  &  &  &  & \checkmark & & \checkmark & \checkmark & \checkmark & R\\
+	Test Plan &  & \checkmark &  &  & & \checkmark & & & &  U\\
+	Test Cases & \checkmark & \checkmark & \checkmark &  & \checkmark & \checkmark &
+	\checkmark & \checkmark & \checkmark & U\\
 	\midrule
 	\end{tabular}
-\caption{Commonly Recommended Artifacts in Software Development Guidelines} \label{Tbl_Guidelines}
+	\caption{Comparison of Recommended Artifacts in Software Development Guidelines
+	to Artifacts in LBM Projects (C for Common, U for Uncommon and R for Rare)}
+	\label{Tbl_Guidelines}
 	\end{center}
 	\end{table}
-
-The majority of LBM generated artifacts (summarized below) correspond to general
-recommendations from research software developers.  For LBM, a union of the
-three categories mostly corresponds to general research software recommendations
-(Table~\ref{Tbl_Guidelines}). Areas where LBM developers could improve include
-providing: API documentation, a roadmap, a code of conduct, programming style
-guide, and uninstall instructions.
-
-\begin{description}
-	\item[Common:] Developer List, Issue Tracker, Dependency List, Installation
-	Guide, Theory Notes, Related Publications, Build Files, README File,
-	License, Tutorial, Version Control
-	\item[Less Common:] Change Log, Design Doc., Functional Spec., Performance
-	Info., Test Cases, User Manual
-	\item[Rare:] API Doc., Developer Manual, FAQ, Verification Plan, Video Guide,
-	Requirements Spec.
-\end{description}
 
@@
 and evolution of software, software processes, and artifacts~\cite[p.\
-501]{GhezziEtAl2003}. Development tools support the development of end products,
-but do not become part of them, unlike dependencies that remain in the
-application once it is released \cite[p.\ 506]{GhezziEtAl2003}. To answer
-\rqref{RQ_CompareToolsProjMngmnt} we summarize tools that are visible in the
-repositories and that are mentioned during the developer interviews.  As an
-example, the tools found for LBM software packages are as follows:
+501]{GhezziEtAl2003}. To answer \rqref{RQ_CompareToolsProjMngmnt} we summarize
+tools that are visible in the repositories and that are mentioned during the
+developer interviews.  The tools for LBM are as follows:
 
 \begin{description}
-	\item[Development Tools:] Continuous Integration, Code Editors, Development
-	Environments, Runtime Environments, Compilers, Unit Testing Tools,
-	Correctness Verification Tools
+	\item[Development Tools:] Continuous Integration (CI), Code Editors,
+	Development Environments, Runtime Environments, Compilers, Unit Testing
+	Tools, Correctness Verification Tools
 	\item[Dependencies:] Build Automation Tools, Technical Libraries, Domain
@@
 Once the data on tools is collected, the use of two specific tools should be
-compared to research software norms: version control and continuous integration.
-A little over 10 years ago version control was estimated to be used in only 50\%
-of research software projects \cite{Nguyen-HoanEtAl2010}, but even at that time
-\cite{Nguyen-HoanEtAl2010} noted an increase from previous usage levels. More
-recently, version control usage rates for active mesh generation, geographic
-information system and statistical software packages were close to
-100\%~\cite{Smith2018}.  Almost every software guide cited in
-Table~\ref{Tbl_Guidelines} includes the advice to use version control. For LBM
-packages 67\% use version control (GitHub, GitLab or CVS). The high usage of
-version control tools in LBM software matches the trend in research software in
-general.
-
-Continuous integration is rarely used in LBM (3 of 24 packages or 12.5\%). This
-contrasts with the frequency with which continuous integration is recommended in
-research software development
-guidelines~\cite{BrettEtAl2021,vanGompelEtAl2016,ThielEtAl2020}. In this case,
-it seems likely that the recommendations are ahead of common practice.
+compared to research software norms: version control and CI. A little over 10
+years ago version control was estimated to be used in only 50\% of research
+software projects \cite{Nguyen-HoanEtAl2010}. More recently, version control
+usage rates for active mesh generation, geographic information system and
+statistical software packages were close to 100\%~\cite{Smith2018}.  Almost
+every software guide cited in Table~\ref{Tbl_Guidelines} includes the advice to
+use version control. For LBM packages 67\% use version control. The high usage
+in LBM software matches the trend in research software in general. CI is rarely
+used in LBM (3 of 24 packages or 12.5\%). This contrasts with the frequency with
+which CI is recommended in
+guidelines~\cite{BrettEtAl2021,vanGompelEtAl2016,ThielEtAl2020}.
 
@@
 
-The interview data on development processes is used to answer research question
+The interview data on development processes is used to answer
 \rqref{RQ_CompareMethodologies}.  This data should be contrasted with the
@@
 \cite{CarverEtAl2007,Segal2005}, or an amethododical process \cite{Kelly2013}.
-Another point of comparison should be on the use of peer review, since peer
-review is frequently recommended~\cite{HerouxEtAl2008,OrvizEtAl2017,USGS2019}.
+Another point of comparison is on the use of the frequently recommended peer
+review approach~\cite{HerouxEtAl2008,OrvizEtAl2017,USGS2019}.
 
@@
 one developer (ESPResSo) told us their non-rigorous development model is like a
-combination of agile and waterfall. Employing a loosely defined process makes
-sense for LBM software, given that the teams are generally small and
-self-contained. One of the developers (ESPResSo) also noted that they use an ad
-hoc peer review process.
+combination of agile and waterfall. A loosely defined process makes sense for
+LBM software, given the small self-contained teams. One of the developers
+(ESPResSo) also noted that they use an ad hoc peer review process.
 
@@
 
-To answer \rqref{RQ_PainPoints}, we ask developers about their pain points and
-compare their responses to the literature~\cite{WieseEtAl2019,PintoEtAl2018}.
-Pain points to watch for include: cross-platform compatibility, scope bloat,
-lack of user feedback, dependency management, data handling concerns,
-reproducibility, and software scope.
+To answer \rqref{RQ_PainPoints} and \rqref{RQ_ComparePainPoints}, we ask
+developers about their pain points and compare their responses to the
+literature~\cite{WieseEtAl2019,PintoEtAl2018}. Pain points to watch for include:
+cross-platform compatibility, scope bloat, lack of user feedback, dependency
+management, data handling concerns, reproducibility, and software scope.
 
-An example pain point noted for LBM is a lack of development time. A developer
-of pyLBM noted that their small development team did not have enough time to
-implement new features. Small development teams are common for LBM software
-packages (as shown in Figure~\ref{measurement_template_image}).
+An example pain point noted for LBM is a lack of development time. Small
+development teams are common for LBM software packages
+(Figure~\ref{measurement_template_image}).  Lack of time is also highlighted by
+other research software developers
+\cite{PintoEtAl2018,PintoEtAl2016,WieseEtAl2019}.
 
-\section{Threats To Validity} \label{threats}
+\section{Threats To Validity} \label{SecThreats}
 
@@
 
-Another risk to the validity of the proposed approach is missing or incorrect
-data. Some software package data may have been missed due to technology
-issues like broken links. For the LBM example, this issue arose with the
-measurement of Palabos, which had a broken link to its user manual. Some
-pertinent data may not have been specified in public artifacts, or may be
-obscure within an artifact or web-page. For the LBM example, the use of unit
-testing and continuous integration was mentioned in the artifacts of only three
-(ESPResSo, Ludwig, Musubi) packages. However, interviews suggested a more
-frequent use of both unit testing and continuous integration in the development
-processes.
+Another risk is missing or incorrect data. Some software package data may have
+been missed due to technology issues like broken links. Some pertinent data may
+not have been specified in public artifacts, or may be obscure within an
+artifact or web-page. For the LBM example, the use of unit testing and CI was
+mentioned in the artifacts of only three (ESPResSo, Ludwig, Musubi) packages.
+However, interviews suggested a more frequent use of both unit testing and CI in
+the development processes.  Given that we could not interview a representative
+for every project, to keep the measures consistent we did not modify our initial
+repository inspection based CI count.
 
@@
 However, we did find room for improvement with respect to providing API
-documentation, a roadmap, a code of conduct, programming style guide, and
-uninstall instructions.  In addition, the community could likely benefit by
-increased use of continuous integration.
+documentation, a roadmap, a code of conduct, programming style guide, uninstall
+instructions and CI.
 
-Building from the LBM example we can create a wealth of data on multiple
-domains.  The next step is a meta-analysis, where we look at how the different
-domains compare to answer new research questions like: What lessons from one
-domain could be applied in other domains? What (if any) differences exist in the
+In the future, we would like reduce measurement time down by increased
+automation.  We would also like to conduct a meta-analysis, where we look at how
+different domains compare to answer new research questions like: What lessons
+from one domain could be applied in other domains? What differences exist in the
 pain points between domains?  Are there differences in the tools, processes, and

### 2022-04-13 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Edits to add more background on recommendations

@@
     generated artifacts corresponding to general recommendations from research
-    software developers, common utilization of version control (67\% of
+    software developers, common use of version control (67\% of
     packages) and adoption of a quasi-agile development process.  Areas of best
@@
 the research questions that underpin our data collection methodology
-\citep{SmithEtAl2021}. We start with identifying the examples of LBM software
-where we have access to the source code:
+\citep{SmithEtAl2021, SmithAndMichalski2022}. We start with identifying the
+examples of LBM software where we have access to the source code:
 
@@
 We next wish to assess the representative software to determine how well they
-apply current software development best practices.  At this point in the
-process, to remove potential user/developer bias, we will base our assessment
-only on publicly available artifacts, where artifacts are the documents, scripts
-and code that we find in a project's public repository. Example artifacts
-include requirements, specifications, user manuals, unit test cases, system
-tests, usability tests, build scripts, API (Application Programming Interface)
-documentation, READMEs, license documents, process documents, and code.
-Following best practices does not guarantee popularity, so we will also compare
-our ranking to how the user community itself ranks the identified projects.
+apply current software development best practices.  By \emph{best practices} we
+mean methods, techniques and tools that are generally believed to improve
+software development, like testing and documentation.  As we will discuss in
+Section~\ref{empiricalmeasures}, we will categorize our best practices around
+software qualities.  At this point in the process, to remove potential
+user/developer bias, we will base our assessment only on publicly available
+artifacts, where artifacts are the documents, scripts and code that we find in a
+project's public repository. Example artifacts include requirements,
+specifications, user manuals, unit test cases, system tests, usability tests,
+build scripts, API (Application Programming Interface) documentation, READMEs,
+license documents, process documents, and code. Following best practices does
+not guarantee popularity, so we will also compare our ranking to how the user
+community itself ranks the identified projects.
 
@@
 methodologies used for software development.  With this data, we can ask
-questions about how LBM software compares to other research software.  That is,
-in what ways does LBM follow the trends from other developer communities, and in
-which ways is it different?
+questions about how LBM software compares to other research software.  For cases
+where data is available, we will compare to other specific domains of research
+software, like medical imaging \citep{Dong2021} and ocean modelling
+\citep{JungEtAl2022}.  However, most of the literature on the state of the
+practice does not focus on specific domains, but rather on all research software
+\citep{SmithAndMichalski2022}. We will refer to studies and guidelines that do
+not distinguish an application domain as \emph{research software in general}. By
+comparing LBM software to research software in general we can determine the ways
+LBM follows the trends from other developer communities, and ways in which it is
+different.
 
@@
 We investigated the research questions by applying the general methodology
-summarized in \citet{SmithEtAl2021}.  The specific application of the
-methodology to LBM is reviewed in Section~\ref{methodology}, with the full
-details in \citep{Michalski2021}.  The current methodology updates the approach
-used in prior assessments of domains like Geographic Information Systems
-\citep{SmithEtAl2018_arXivGIS}, Mesh Generators \citep{SmithEtAl2016},
+summarized in \citet{SmithEtAl2021, SmithAndMichalski2022}.  The specific
+application of the methodology to LBM is reviewed in Section~\ref{methodology},
+with the full details in \citep{Michalski2021}.  The current methodology updates
+the approach used in prior assessments of domains like Geographic Information
+Systems \citep{SmithEtAl2018_arXivGIS}, Mesh Generators \citep{SmithEtAl2016},
 Oceanographic Software \citep{smith2015state}, Seismology software
@@
 approximately measure the application of best practices for each package by
-filling in a grading template. Compared with our previous methodology (as used
-in \citep{SmithEtAl2016} for instance), the new methodology also includes
+filling in a grading template. Our analysis points to areas where some LBM
+software packages fall short of current best practices.  This is not intended to
+be a criticism of any existing packages, especially since not every project is
+actively maintained.  Moreover, we recognize that not every project needs to
+achieve the highest possible quality.  However, rather than delve into the
+nuances of which software can justify compromising which practice we conducted
+our comparison under the ideal assumption that every project can have sufficient
+resources to match best practices.
+
+Compared with our previous methodology (as used in \citep{SmithEtAl2016} for
+instance), measuring the application of best practices now also includes
 repository based metrics, such as the number of files, number of lines of code,
@@
 
-We developed a methodology for evaluating the state of the practice of
-research software \citep{SmithEtAl2021}. This methodology can be applied
-to a specific domain of research software, which in the current case is LBM
-software.  Since we group best practices around conventional software qualities,
-the first section below provides definitions for the qualities of interest
-(Section~\ref{softwarequalities}), along with information on how we assessed
-these qualities. The following section (Section~\ref{Sec_OverallProcess})
+We developed a methodology for evaluating the state of the practice of research
+software \citep{SmithEtAl2021, SmithAndMichalski2022}. This methodology can be
+applied to a specific domain of research software, which in the current case is
+LBM software.  Since we group best practices around conventional software
+qualities, the first section below provides definitions for the qualities of
+interest (Section~\ref{softwarequalities}), along with information on how to
+measure these qualities. The following section (Section~\ref{Sec_OverallProcess})
 provides an overview of the steps we used to select, measure and compare LBM
-software.  The remaining sections provide details on each step in the methodology.
-
-\subsection{Software Qualities} \label{softwarequalities}
-
-The following are the software quality definitions used in this exercise, along
-with comments regarding their measurement.
+software.  The remaining sections provide details on each step in the
+methodology.
+
+\subsection{Measuring Software Qualities} \label{softwarequalities}
+
+To ensure our assessment of best practices is organized and complete, our
+measurements are centred around software qualities.  For completeness we used
+the the list of software engineering qualities from \citet[p.\
+17--33]{GhezziEtAl2003}, excluding interoperability, productivity and
+timeliness.  We excluded interoperability because we are focusing on LBM
+software alone, not its interaction with other software.  The other two
+qualities were excluded because they are process qualities, and we are measuring
+product qualities.  To the list from \citet{GhezziEtAl2003} we added
+installability, since it is so critical to the success of any practical project,
+and reproducibility, since it is so important for research software. Below we
+provide information on each quality of interest, including information on how
+they are measured.
 
@@
 	developer manuals, specifications, README files, change logs, release notes,
-	publications, forums, and instructional websites.
-
-	\item[Modifiability] Modifiability refers to the ease with which stable
-	changes can be made to a system and the flexibility of the system to adopt
-	such changes \citep{8016712}. We did not directly measure modifiability.
-	Instead, developers were asked in interviews if they considered the ease of
-	future changes when developing the software packages, specifically changes
-	to the structure of the system, modules and code blocks. A follow up
-	question is asked if any measures have been taken to ensure the ease of
-	future changes.
+	publications, forums, and instructional websites. We also asked developers
+	if they considered the ease of future changes when developing the software
+	packages.
 
@@
 
-	\item[Traceability] Traceability refers to the ability to link the software
-	implementation and the software artifacts, especially the requirement
-	specification \citep{McCallEtAl1977}. Similar to the quality of correctness,
-	this requires the availability of some form of requirements specification.
-	This quality refers to keeping track of information as it changes forms or
-	relates between artifacts. We did not quantitatively measure traceability.
-	Instead, developers were asked in interviews how documentation fits into
-	their development process.
-
 	\item[Visibility and Transparency] Visibility and transparency refer to the
@@
 	steps to ensure reproducibility.
-
-	\item[Unambiguity] Unambiguity refers to the extent to which two readers
-	have similar interpretations when reading software artifacts. In other
-	words, artifacts are unambiguous if, and only if, they only have one
-	interpretation \citep{IEEE1998}. This state of the practice assessment did
-	not quantitatively measure unambiguity, but we did ask developers if they
-	believe that their current documentation clearly conveys all necessary
-	knowledge to the users.  Follow-up questions ask how they achieved
-	unambiguity, or what improvements are needed to achieve it.
 
@@
 
-The full template consists of 108 questions categorized under nine qualities:
+The full template consists of 108 questions categorized under nine qualities
+(Section~\ref{softwarequalities}):
 \begin{inparaenum}[(i)]
@@
 
-\section{Comparison Between LBM and Research Software for Artifacts}
+\section{Comparison Between LBM and Other Research Software for Artifacts}
 \label{Sec_CompareArtifacts}
@@
 for the presence of artifacts, which were then categorized by frequency. We have
-grouped them into common, uncommon, and rare artifacts in
+grouped the artifacts into common, uncommon, and rare categories in
 Table~\ref{artifactspresent}. Common artifacts were found in 16 to 24 ($>$63\%)
-of the software packages. Uncommon artifacts were found in 8 to 15 (30-63\%) of
-the software packages. Rare artifacts were found in 1 to 7 ($<$30\%) of the
-software packages.  We answer \rqref{RQ_CompareArtifacts} by comparing the
-artifacts that we observed in LBM repositories to those observed and recommended
-for research software in general. Our comparison may point out areas where some
-LBM software packages fall short of current best practices. This is not intended
-to be a criticism of any existing packages, especially since in practice not
-every project needs to achieve the highest possible quality.  However, rather
-than delve into the nuances of which software can justify compromising which
-practices we will write our comparison under the ideal assumption that every
-project has sufficient resources to match best practices.
+of the software packages. Uncommon artifacts were found in 8 to 15 (30-63\%) and
+rare artifacts were found in 1 to 7 ($<$30\%) of the packages.  We answer
+\rqref{RQ_CompareArtifacts} by comparing the artifacts that we observed in LBM
+repositories to those observed and recommended for research software in general.
+Our comparison may point out areas where some LBM software packages fall short
+of current best practices. This is not intended to be a criticism of any
+existing packages, especially since in practice not every project needs to
+achieve the highest possible quality.  However, rather than delve into the
+nuances of which software can justify compromising which practices we will write
+our comparison under the ideal assumption that every project has sufficient
+resources to match best practices.
 
@@
 List of Related Publications & Test Plan / Script / Cases & Product Roadmap\\
-Makefile / Build File & User Manual / Guide & Unistall\\
+Makefile / Build File & User Manual / Guide & Uninstall\\
 README File & & Verification and Validation Plan\\
@@
 recommends this artifact.  The last column shows whether the artifact appears in
-the measured set of MI software, either not at all (blank), commonly (C),
+the measured set of LBM software, either not at all (blank), commonly (C),
 uncommonly (U) or rarely (R).  We did our best to interpret the meaning of each
@@
 commonly recommended artifacts are the most important artifacts. Just because an
-artifact is not explicitly recommended in a given guidelines, does not mean that
-the artifact is not valued by the guideline authors.  They may have exluded it
-because it is out of the scope of their recommendations, or outside of their
-experience.  For instance, an artifact related to uninstall is only explicitly
-mentioned by \citep{vanGompelEtAl2016}), but other guideline authors would
-likely see its value.  They may simply feel that uninstall is implied by
-install, or they may have never asked themselves whether separate uninstall
-instructions are needed.
+artifact is not explicitly recommended in a given guideline, does not mean that
+the artifact is not valued.  The guideline authors may have excluded it because
+it is out of the scope of their recommendations, or outside of their experience.
+For instance, an artifact related to uninstall is only explicitly mentioned by
+\citep{vanGompelEtAl2016}), but other guideline authors would likely see its
+value.  They may simply feel that uninstall is implied by install, or they may
+have never asked themselves whether separate uninstall instructions are needed.
 
@@
 \caption{Comparison of Recommended Artifacts in Software Development Guidelines
-to Artifacts in MI Projects (C for Common, U for Uncommon and R for Rare)}
+to Artifacts in LBM Projects (C for Common, U for Uncommon and R for Rare)}
 \label{Tbl_Guidelines}
@@
 verification and validation notes, and a video guide for the software.
-
-The majority of LBM generated artifacts correspond to general recommendations
-from research software developers.  A union of the three columns in
-Table~\ref{artifactspresent} mostly corresponds to recommendations made by the
-research software community.  Most of the items in Table~\ref{artifactspresent}
-are included in guidelines for writing research software, such as the DLR
-(German Aerospace Centre) Software Engineering Guidelines
-\citep{TobiasEtAl2018}, xSDK (Extreme-scale Scientific Software Development Kit)
-Community Package Policies \citep{SmithAndRoscoe2018}, the EURISE (European
-Research Infrastructure Software Engineers') Network Technical Reference
-\cite{ThielEtAl2020} and CLARIAH (Common Lab Research Infrastructure for the
-Arts and Humanities) Guidelines for Software Quality \citep{vanGompelEtAl2016}.
-As an example, LBM developers commonly list the external libraries on which the
-software depends, as recommended by \citet{BrettEtAl2021},
-\citet{SmithAndRoscoe2018} and \citet{vanGompelEtAl2016}. Although in the rare
-column, some LBM developers include an FAQ (Frequently Asked Questions) document
-as recommended by \citep{OrvizEtAl2017}, \citep{ThielEtAl2020} and
-\citep{vanGompelEtAl2016}. LBM developers also follow the ubiquitous
-recommendation to have a README file, although, as for other software, the
-content of README files is not standardized \citep{PranaEtAl2018}.  Compared to
-ocean modelling software \citep{JungEtAl2022}, LBM software pays more
-attention to artifacts related to testing.
 
@@
 
-Although the LBM community participates in most of the practices we found listed
-in the general research software guidelines, some recommended practices were not
-observed or rarely observed. For instance API documentation was rarely observed
-for LBM software, but it is a frequently recommended artifact
-\citep{SmithAndRoscoe2018, ThielEtAl2020, vanGompelEtAl2016, OrvizEtAl2017,
-SSI2022, Zadka2018}.  In addition to the items in the last column of
-Table~\ref{artifactspresent}, we can add the following community recommended
-items that we rarely, if ever, observed:
+As Table~\ref{Tbl_Guidelines} shows the LBM community participates in most of
+the practices we found listed in the general research software guidelines;
+however, some recommended practices were not observed or rarely observed. For
+instance API documentation was rarely observed for LBM software, but it is a
+frequently recommended artifact \citep{SmithAndRoscoe2018, ThielEtAl2020,
+vanGompelEtAl2016, OrvizEtAl2017, SSI2022, Zadka2018}.  In addition to the items
+in the last column of Table~\ref{artifactspresent}, we can add the following
+community recommended items that we rarely, if ever, observed:
 
 \begin{itemize}
-\item A roadmap showing what is planned for the future \citep{Yo2021,
-ThielEtAl2020, vanGompelEtAl2016}.  We did see this information mentioned in
-some repos, but we did not observe a specific artifact devoted to this purpose.
-\item A code of conduct to explicitly say how developers should treat one
-another \citep{Yo2021, ThielEtAl2020}.  We did not observe a policy of this
-sort, possibly because the number of developers for each LBM project is small.
-\item Programming style guidelines so that programming labels and formatting are
-consistent \citep{TobiasEtAl2018, ThielEtAl2020, OrvizEtAl2017, Zadka2018,
-vanGompelEtAl2016, WilsonEtAl2014}.  We saw this information as part of
-some developer guides, but only rarely.
-\item Checklists can be used in projects to ensure that best practices are
-followed by all developers.  Some examples include checklists merging branches
-into master \citep{Brown2015}, checklists for saving and sharing changes to the
-project \citep{WilsonEtAl2016}, checklists for new and departing team members
-\citep{HerouxAndBernholdt2018}, checklists for processes related to commits and
-releases \citep{HerouxEtAl2008} and checklists for overall software quality
-\citep{ThielEtAl2020, SSI2022}.  For LBM software, ESPResSo has a checklist for
-managing releases, but otherwise they are used only rarely for LBM.
-\item Uninstall instructions \citep{vanGompelEtAl2016} were not observed for any
-of the LBM projects.
+
+\item \textbf{A roadmap} is considered by some to be a critical document for an
+organization to map out the vision and direction for a product offering
+\citep{MunchEtAl2019}.  Roadmaps take different forms but they all aim to
+address the following: \begin{inparaenum}[i)]
+	\item Where are we now?
+	\item Where are we going? and
+	\item How can we get there? \end{inparaenum} \citep{PhaalEtAl2005}.
+Developers can create a roadmap by the following steps:
+\begin{inparaenum}[i)]
+	\item define and outline a strategic mission and product vision, 
+	\item scan the environment, 
+	\item revise and distil the product vision to write the product roadmap, 
+	\item estimate the product life cycle and evaluate the mix of planned
+development efforts. \end{inparaenum} \citep{VahaniittyEtAl2002}.  A product
+roadmap has the benefits of providing continuity of purpose, facilitating
+stakeholder collaboration and assisting with prioritization \citep{Pichler2012}.
+Although we saw some elements of a roadmap in the LBM projects, we did not
+observe a specific artifact devoted to this purpose.
+
+\item \textbf{A code of conduct} documents expectations and standards of ethical
+behaviour and explicitly states how developers should treat one another
+\citep{TouraniEtAl2017}. A code of conduct outlines rules for communication,
+establishes enforcement mechanisms for violations and codifies the spirit of a
+community, such that anyone can contribute comfortably regardless of, for
+example, ethnicity, gender, or sexual orientation \citep{TouraniEtAl2017}. The
+top three most popular code's of conduct from \citet{TouraniEtAl2017} are:
+\href{https://ubuntu.com/community/code-of-conduct} {Ubuntu Code of Conduct},
+\href{https://www.contributor-covenant.org/version/2/1/code_of_conduct/}
+{Contributor Covenant}, \href{https://www.djangoproject.com/conduct/} {Django
+Code of Conduct}. \citet{SinghEtAl2021} shows the value of adopting a code of
+conduct on the participation of women. Improved inclusivity means a wider pool
+of potential contributors. Moreover, a code of conduct facilitates meeting
+ethical obligations for developers that abide by a code of ethics, such as the
+IEEE Code of Ethics \citep{IEEE1999}, or the Professional Engineers of Ontario
+code of ethics \citep[p.\ 23--24]{PEO2021}. For the LBM software, we did not
+observe any codes of conduct, possibly because the number of developers for each
+LBM project is small and because the idea hasn't yet occurred to any of the
+participants.
+
+\item \textbf{Code style guides} provide rules and guidance on writing code.
+Style guides standardize such elements as naming identifiers, formatting, code
+comments, best practices and dangers to avoid \citep{Carty2020}. An example of a
+ubiquitous coding standard is to use ALLCAPS for symbolic constants.
+Standardization improves understandability (Section~\ref{softwarequalities}),
+since developers spend more time on the content of the code, and less time
+distracted by its style.  Three example style guides include:
+\href{http://cnl.sogang.ac.kr/cnlab/lectures/programming/python/PEP8_Style_Guide.pdf}
+{PEP8 Style Guide for Python},
+\href{https://google.github.io/styleguide/javaguide.html} {Google Java Style
+Guide} and \href{https://google.github.io/styleguide/cppguide.html} {Google \CC
+Style Guide}.  Tools like \href{https://pypi.org/project/flake8/}{flake8} can be
+used to automatically check that coding styles, in this case PEP8, are enforced.
+For LBM software we saw code style advice as part of some developer guides, but
+only rarely.
+
+\item \textbf{Checklists} can be used in projects to ensure that best practices
+are followed by all developers.  Some examples include checklists merging
+branches into master \citep{Brown2015}, checklists for saving and sharing
+changes to the project \citep{WilsonEtAl2016}, checklists for new and departing
+team members \citep{HerouxAndBernholdt2018}, checklists for processes related to
+commits and releases \citep{HerouxEtAl2008} and checklists for overall software
+quality \citep{ThielEtAl2020, SSI2022}.  For LBM software, ESPResSo has a
+checklist for managing releases, but we did not find any other examples.
+
+\item \textbf{Uninstall instructions} tell a user how to remove the software and
+associated libraries from their system.  The advantage of uninstall is removing
+files that consume space and potentially pollute the users build system by
+causing conflicts when installing and using other software packages.  We did not
+observe uninstall instructions for any of the LBM projects.
+
+\item \textbf{API documentation} is \wss{fill in definition, examples and
+benefits.  talk about doxygen to support.}
+
 \end{itemize}
 
-The above discussion shows that, taken together, LBM projects fall short of
+Where data is available, LBM software artifacts can be compared to other
+research software domains.  Compared to ocean modelling software
+\citep{JungEtAl2022}, LBM software pays more attention to artifacts related to
+testing.  Compared to Medical Imaging Software \citep{Dong2021}, the artifact
+frequency is similar, except LBM software is more likely to have developer
+related artifacts, like Contributing, Dependency list, and Design documentation.
+
+Table~\ref{Tbl_Guidelines} shows that many LBM projects fall short of
 recommended best practices for research software.  However, LBM software is not
 alone in this.  Many, if not most, research projects fall short of best
-practices.  Software requirements documentation provides a good example of the
-common deviation between recommendations and practice.  Although requirements
+practices.   A gap exists in scientific computing development practices and
+software engineering recommendations \citep{Storer2017, Kelly2007,
+OwojaiyeEtAl2021_CSE}. \citet{JohansonAndHasselbring2018} observe that the
+state-of-the-practice for SCS in industry and academia does not incorporate
+state-of-the-art SE tools and methods.  This causes sustainability and
+reliability problems \citep{FaulkEtAl2009}. Rather than benefit from capturing
+and reusing previous knowledge, projects waste time and energy ``reinventing the
+wheel'' \citep{deSouzaEtAl2019}.
+
+Software requirements documentation provides a good example of the common
+deviation between recommendations and practice.  Although requirements
 documentation is recommended by some \citep{TobiasEtAl2018, HerouxEtAl2008,
@@
 \rqref{RQ_CompareToolsProjMngmnt} by comparing aspects of tool usage in LBM
-software packages to their utilization in the research software community in
+software packages to their use in the research software community in
 general.
@@
 
-Based on information provided by \citet{JungEtAl2022}, tool utilization for
-LBM software has much in common with tool utilization for ocean modelling
+Based on information provided by \citet{JungEtAl2022}, tool usage for
+LBM software has much in common with tool usage for ocean modelling
 software.  Both use tools for editing, compiling, code management, testing,
@@
 software development guidelines \citep{BrettEtAl2021, Brown2015, ThielEtAl2020,
-Zadka2018, vanGompelEtAl2016}.  We could not find published data on the
-frequency with which continuous integration is used in general research
-software.  Our impression is that like LBM software, other research software
-packages lag behind the recommended best practice of employing continuous
-integration.
+Zadka2018, vanGompelEtAl2016}.  \wss{why continuous integration.}  We could not
+find published data on the frequency with which continuous integration is used
+in general research software.  \wss{add reference to Carver paper.}  Our
+impression is that like LBM software, other research software packages lag
+behind the recommended best practice of employing continuous integration.
 
@@
 correctness is for research software.
+
+\wss{how to doc motivations and design principles. examples. advantages.}
 
@@
 	components in the source code. This results in technical debt
-	\citet{KruchtenEtAl2012}, which has an impact on future modifiability and
+	\citet{KruchtenEtAl2012}, which has an impact on future maintainability and
 	reusability. Concern with technical debt is likely why researchers in the
@@
 
-\subsection{Conduct Rigorous Peer Reviews}
+\subsection{Conduct Rigorous Peer Reviews} \label{Sec_PeerReview}
 
@@
 
-\subsection{Write and Submit More Papers on Software}
+\subsection{Write and Submit More Papers on Software} \label{Sec_MorePapers}
 
@@
 
-\subsection{Grow the Number of Contributors}
+\subsection{Grow the Number of Contributors} \label{Sec_GrowContributors}
 
@@
 discussed in Section~\ref{Sec_CompareArtifacts} such as a contributor
-guidelines, a clear code of conduct and high quality code and documentation.
-This advice is logical, but in our assessment of LBM software we found examples
-of projects that already follow many best practices, but still have small
-development teams. Attracting developers apparently requires more than just
-building a good product.
+guidelines, code style guidelines, a clear code of conduct and high quality code
+and documentation. This advice is logical, but in our assessment of LBM software
+we found examples of projects that already follow many best practices, but still
+have small development teams. Attracting developers apparently requires more
+than just building a good product.
 
@@
 
-\subsection{Augment Theory Manuals to Include Requirements Information}
+\subsection{Augment Theory Manuals to Include Requirements Information} \label{Sec_AugTheoryMans}
 
@@
 \subsection{Improve Re-runnability, Repeatability, Reproducibility and Replicability} 
+\label{Sec_Reproducibility}
 
@@
 
-\subsection{Generate All Things}
+\subsection{Generate All Things} \label{Sec_GenAllThings}
 
@@
 during development despite this not being explicit from an analysis of the
-material available online.
+material available online. Given that we could not interview a representative
+for every project, to keep the measures consistent we did not modify our initial
+repository inspection based CI count.
 
@@
 
-To help users select the right LBM package for their needs, and to potentially
-provide direction for future LBM software development, our goal was to analyze
-the current state of LBM software development.  We determined how well best
-practices are followed and we investigated current developer pain points.  We
-compared the state of the practice for LBM software to the development practices
-recommended and employed for research software in general. To improve the
-quality of LBM software we looked for examples where developers have been
-successful so that we can share these with the LBM community, and with the
-broader research software community.  We also aimed to identify potential
-approaches for mitigating the identified pain points in the future.
-
-Our analysis points to areas where some LBM software packages fall short of
-current best practices.  This is not intended to be a criticism of any existing
-packages, especially since in practice not every project needs to achieve the
-highest possible quality.  However, rather than delve into the nuances of which
-software can justify compromising which practice we conducted our comparison
-under the ideal assumption that every project can have sufficient resources to
-match best practices.
-
-Our methodology involved the following steps: i) Identify a list of 45 candidate
-software packages; ii) Filter the list to a length of 24 packages; iii) Collect
-repository related data on each software package, like number of stars, number
-of open issues, number of lines of code; iv) Fill in the measurement template
-(the template consists of 108 questions to assess nine qualities
-(installability, correctness/verifiability, reliability, robustness, usability,
-maintainability, reusability, understandability, and visibility/transparency).
-Filling in the template requires installing the software on a VM, running simple
-tests (like completing the getting started instructions (if present)), and
-searching the code, documentation and test files; v) Rank the software using the
-Analytic Hierarchy Process (AHP); vi) Interview developers (the interview
-consists of 20 questions and takes about an hour); and, vii) Conduct a domain
-analysis. The collected data was analyzed by: a) comparing the ranking by best
-practices against the ranking by popularity; b) comparing artifacts, tools and
-processes to current research software development guidelines; and, c) exploring
-pain points.  Our ranking was roughly consistent with ranking the repositories
-by the number of GitHub stars, although the number one project by star count
-(Sailfish) is ranked 9th by our methodology.
-
-Current LBM software provides many examples of following best practices, to the
-benefit of users and developers.  This is illustrated by
+This paper will help scientists select a suitable LBM package for their needs.
+However, its more fundamental purpose is to deepen our understanding of the
+state of the practice for developing LBM software. Understanding what artifacts
+are produced, what tools are used, what process are employed and what pain
+points are experienced, is the starting point for devising future methods and
+tools for reducing development time and improving software quality.
+
+To assess the state of the practice, we followed five steps:
+\begin{inparaenum}[i)]
+	\item Identified a list of 24 LBM packages;
+	\item Measured the packages using a measurement template (the template
+	consists of 108 questions to assess 9 qualities (including the qualities of
+	installability, usability and visibility))
+	\item Ranked the software using the Analytic Hierarchy Process (AHP);
+	\item Interviewed developers (each interview consisted of 20 questions and took
+	about an hour); and, 
+	\item Conducted a domain analysis. 
+\end{inparaenum}
+The collected data was analyzed by: 
+\begin{inparaenum}[i)]
+	\item Comparing our ranking by best practices against the ranking by
+	popularity;
+	\item Comparing artifacts, tools and processes to current research software
+	development guidelines; and, 
+	\item Exploring pain points.
+\end{inparaenum}
+
+Current LBM software mostly follows best practices as illustrated by
 Table~\ref{topperformerstable}, which summarizes the top five performers for
-each of the identified qualities and for the overall quality. Taking the union
-of all entries in the table, we have 16 top performers: ESPREsSo, MechSys,
-OpenLB, DL\_MESO, Sailfish, Ludwig, Musubi, pyLBM, waLBerla, laboetie, lettuce,
-ESPREsSo++, LUMA, MP-LABS, Palabos, and HemeLB.  Therefore, two thirds (67\%) of
-the 24 measured packages appear in the top five for at least one quality.  Of
-these entries almost all of them are classified as alive, except for three
-packages: laboetie, MP-LABS, and HemeLB.  This suggests that alive packages are
-in a healthy state with respect to adopting best practices.
+each quality and for the overall quality. Taking the union of all entries in the
+table, we have 16 top performers: ESPREsSo, MechSys, OpenLB, DL\_MESO, Sailfish,
+Ludwig, Musubi, pyLBM, waLBerla, laboetie, lettuce, ESPREsSo++, LUMA, MP-LABS,
+Palabos, and HemeLB.  Therefore, two thirds (67\%) of the 24 measured packages
+appear in the top five for at least one quality.  Of these entries almost all
+are classified as alive, except for three packages: laboetie, MP-LABS, and
+HemeLB.  Alive packages are in a healthy state with respect to adopting best
+practices.
 
@@
 		\caption{Top performers for each quality (all packages were alive at the
-		time of measurement, except for those marked by an asterisk (*)) (Sorted by Order of Quality Measurement)}
+		time of measurement, except for those marked by an asterisk (*)) (Sorted
+		by Order of Quality Measurement)}
 		\label{topperformerstable}
@@
 
-Further evidence of the healthy state of the practice can be seen by comparing
-the observed software artifacts versus those recommended by software engineering
-guidelines.  As Section~\ref{Sec_CompareArtifacts} shows, the majority of LBM
-generated artifacts correspond to general recommendations from research software
-developers.  LBM software even seems to be ahead of best practices with an
-example of a video user guide for one of the packages.  Comparing LBM with
-respect to tool usage (Section~\ref{Sec_CompareTools}) shows that LBM is keeping
-pace with other research software, with 67\% of the surveyed tools using version
-control.  The state of the practice for the software development process of LBM
-software matches the quasi-agile process generally adopted for research software
+LBM software is shown to be in a healthy state when comparing artifacts, tools
+and processes, to research software guidelines and other research software.  The
+majority of LBM generated artifacts correspond to general recommendations from
+research software developers (Section~\ref{Sec_CompareArtifacts}).  LBM software
+even seems to be ahead of best practices with an example of a video user guide
+for one of the packages. LBM is mostly keeping pace with other research software
+with respect to tool usage (Section~\ref{Sec_CompareTools}), with 67\% of the
+surveyed tools using version control.  The state of the practice for the
+software development process of LBM software matches the quasi-agile process
+generally adopted for research software
 (Section~\ref{Sec_CompareMethodologies}).  The best practice process of peer
@@
 Although the state of the practice for LBM software is healthy, we have noted
-areas for potential improvement in the adoption of best practices, as follows:
+areas where the state of the practice does not match existing research software
+development guidelines, as follows:
 
 \begin{itemize}
-	\item Increase the utilization of continuous integration.
-	\item Increase the amount of API documentation.
 	\item Add roadmap documentation showing what is planned for the future.
+	\wss{reason} (Section~\ref{Sec_CompareArtifacts})
 	\item Add a code of conduct to explicitly show how developers should treat
-	one another.
+	one another. \wss{reason} (Section~\ref{Sec_CompareArtifacts})
 	\item Increase the use and enforcement of programming style guides.
+	\wss{reason} (Section~\ref{Sec_CompareArtifacts})
 	\item Use checklists to ensure best practices are followed by developers.
-	\item Include uninstall instructions.
+	\wss{reason} (Section~\ref{Sec_CompareArtifacts})
+	\item Include uninstall instructions. \wss{reason} (Section~\ref{Sec_CompareArtifacts})
+	\item Increase the amount of API documentation. \wss{reason} (Section~\ref{Sec_CompareArtifacts})
+	\item Increase the use of continuous integration. \wss{reason} (Section~\ref{Sec_CompareTools})
 	\item Explicitly document the software's motivation and design principles.
+	\wss{reason - where is this discussed the first time - should add more
+	information there.} (Section~\ref{Sec_CompareMethodologies})
 \end{itemize}
 
-To advance the state of the practice beyond current best practices toward future
-best practices, we interviewed four developers to learn about their pain points
-(Section~\ref{painpoints}).  The identified pain points include lack of
-development time, lack of software development experience, lack of incentive and
-funding, lack of external support, technology hurdles, ensuring correctness,
-usability, technical debt and documentation quality.  We found
-(Section~\ref{Sec_AddressConcerns} that developers are currently creating
-practices to address these pain points, including: designing for change,
-circumventing the oracle testing problem, and prioritizing documentation and
-usability.  To plan for future best practices, we identified eight
-potential directions (Section~\ref{Sec_Recommendations}), in order of increasing
-novelty:
+For insight into devising future methods and tools, we interviewed four
+developers to learn about their pain points (Section~\ref{painpoints}).  The
+identified pain points include lack of development time, lack of software
+development experience, lack of incentive and funding, lack of external support,
+technology hurdles, ensuring correctness, usability, technical debt and
+documentation quality.  We found (Section~\ref{Sec_AddressConcerns} that
+developers are currently creating practices to address these pain points,
+including: designing for change, circumventing the oracle testing problem, and
+prioritizing documentation and usability.  To plan for future best practices, we
+identified eight potential directions (Section~\ref{Sec_Recommendations}), in
+order of increasing development effort:
 
 \begin{enumerate}
-	\item Employ linters.
-	\item Conduct rigorous peer reviews.
-	\item Write and submit papers on software.
-	\item Grow the number of contributors.
-	\item Further address the oracle problem for testing.
+	\item Employ linters. \wss{reason}  (Section~\ref{Sec_Linters})
+	\item Conduct rigorous peer reviews. \wss{reason}
+	(Section~\ref{Sec_PeerReview})
+	\item Write and submit papers on software. \wss{reason}
+	(Section~\ref{Sec_MorePapers})
+	\item Grow the number of contributors. \wss{reason - change this to
+	documentation?} (Section~\ref{Sec_GrowContributors})
+	\item Further address the oracle problem for testing.  \wss{reason}
+	(Section~\ref{Sec_FurtherOracle})
 	\item Augment theory manuals to include requirements information.
+	\wss{reason} (Section~\ref{Sec_AugTheoryMans})
 	\item Improve re-runnability, repeatability, reproducibility and
-	replicability.
-	\item Generate all things.
+	replicability. \wss{reason}  (Section~\ref{Sec_Reproducibility})
+	\item Generate all things. \wss{reason}  (Section~\ref{Sec_GenAllThings})
 \end{enumerate}
 
-Future work can be done to address the threats to validity mentioned in
-Section~\ref{threats}.  For instance, the measures listed in our measurement
-template could be broadened to better capture some qualities. For example,
-usability experiments and performance benchmarks could be incorporated into the
-assessment.  Section~\ref{threats} also mentions that some pertinent information
-on continuous integration and unit testing was not specified in public
-artifacts. Adding further questions to the interview guide regarding the
-measures that are on the measurement template could reduce instances of
-incomplete data. This additional interview data could be analyzed and
-incorporated into the AHP ranking, ensuring quality designations more accurately
-represent the true quality of the software packages. In the future we could also
-expand the interview questions to go beyond reproducibility to also discuss
-replicability. Furthermore, in future interviews we would look to clarify the
-distinction between usability and understandability
-(Section~\ref{softwarequalities}), since the distinction confused some
-participants.  Analysis of future interview data would likely benefit from a
-more rigorous process, like a thematic domain analysis \citep{JungEtAl2022}.
+Future work can be done to address the threats to validity
+(Section~\ref{threats}).  For instance, our measures could be broadened to
+better capture shallow qualities. For example, usability experiments and
+performance benchmarks could be incorporated into the assessment. The accuracy
+of our estimate of participation level for continuous integration and unit
+testing, which is not always visible in public artifacts, can be improved by
+adding interview questions. This additional data could be incorporated into the
+AHP ranking, ensuring quality designations that are more representative of the
+truth. In the future we could also expand the interview questions to go beyond
+reproducibility to also discuss replicability. Furthermore, future interview
+guidelines should clarify the differences between usability and
+understandability (Section~\ref{softwarequalities}), since their answers suggest
+that not all participants understood the distinction.  Analysis of future
+interview data would likely benefit from a more rigorous process, like a
+thematic domain analysis \citep{JungEtAl2022}.
 
 Our domain analysis (Section~\ref{SecDomainAnalysis} was relatively shallow.  In
-the future we would like to dig more deeply into this analysis.  We would like
-to produce tables of commonalities, variabilities, and parameters of variation
-for the family of LBM solvers. We could follow the template from
-\citep{smith2008commonality}, or adopt the analysis techniques from
-\citep{weiss1998commonality}. \cite{SmithAndChen2004} and
-\cite{SmithMcCutchanAndCarette2017} provide examples that could be followed for
-a commonality analysis of a family of mesh generating software and a family of
-material models, respectively.
+the future we would like to dig more deeply.  We would like to produce tables of
+commonalities, variabilities, and parameters of variation for the family of LBM
+solvers. We could follow the template from \citep{smith2008commonality}, or
+adopt the analysis techniques from \citep{weiss1998commonality}.
+\cite{SmithAndChen2004} and \cite{SmithMcCutchanAndCarette2017} provide examples
+that could be followed for a commonality analysis of a family of mesh generating
+software and a family of material models, respectively.
 
@@
 \end{document}
+
+TO do
+- check references
+
+UNUSED Text
+
+Conclusions Related
+
+To help users select the right LBM package for their needs, and to potentially
+provide direction for future LBM software development, our goal was to analyze
+the current state of LBM software development.  We determined how well best
+practices are followed and we investigated current developer pain points.  We
+compared the state of the practice for LBM software to the development practices
+recommended and employed for research software in general. To improve the
+quality of LBM software we looked for examples where developers have been
+successful so that we can share these with the LBM community, and with the
+broader research software community.  We also aimed to identify potential
+approaches for mitigating the identified pain points in the future.

### 2022-04-13 StateOfPractice/Papers/MedImageSoft_SOP-arXiv/MedImageSoft_SOP_AcademicRadiology.tex
Edits to SOP for MI paper

@@
 artifact is not explicitly recommended in a given guidelines, does not mean that
-the artifact is not valued by the guideline authors.  They may have exluded it
+the artifact is not valued by the guideline authors.  They may have excluded it
 because it is out of the scope of their recommendations, or outside of their
@@
 found in research software guidelines (Table~\ref{Tbl_Guidelines}), three
-recommended artifacts were not observed: i) Unistall, ii) Test plans, and iii)
+recommended artifacts were not observed: i) Uninstall, ii) Test plans, and iii)
 Requirements.  Uninstall is likely an omission caused by the focus on installing
@@
 Contributing, Developer code of conduct, Code style guide, Product roadmap,
-Requirements, Design documentation, API documentation and a Test plan.  The rare
-artifacts for MI software are similar to the rare artifacts for Lattice
-Boltzmann solvers \citep{Michalski2021}, except LBM software is more likely to
-have developer related artifacts, like Contributing, Dependency list, and Design
-documentation.
+Requirements, Design documentation, API documentation and a Test plan.
+\wss{additional details are available in the LBM document for other missing
+parts, like code style guides.}  The rare artifacts for MI software are similar
+to the rare artifacts for Lattice Boltzmann solvers \citep{Michalski2021},
+except LBM software is more likely to have developer related artifacts, like
+Contributing, Dependency list, and Design documentation.
 
@@
 releases \citep{HerouxEtAl2008} and checklists for overall software quality
-\citep{ThielEtAl2020, SSI2022}.  For instace, for Lattice Boltzmann solver
+\citep{ThielEtAl2020, SSI2022}.  For instance, for Lattice Boltzmann solver
 software, ESPResSo has a checklist for managing releases \citep{Michalski2021}.
@@
 
+\citet{PintoEtAl2018} lists some pain points that did not come up in our
+conversations with LBM developers: Cross-platform compatibility, interruptions
+while coding, scope bloat, lack of user feedback, hard to collaborate on
+software projects, and aloneness. \citet{WieseEtAl2019} repeat some of the
+previous pain points and add the following: dependency management, data handling
+concerns (like data quality, data management and data privacy), reproducibility,
+and software scope determination. Although LBM developers did not mention these
+pain points, we cannot conclude that they are not relevant for LBM software
+development, since we only interviewed four LBM developers for about an hour each.
+
+\begin{enumerate}
+
+	\item[P\refstepcounter{pnum}\thepnum \label{P_LackDevTime}:] \textbf{Lack of
+	Development Time} A developer of pyLBM noted that their small development
+	team has a lack of time to implement new features. Small development teams
+	are common for LBM software packages (as shown in the measurement table
+	excerpt in Figure~\ref{measurement_template_image}). Lack of time is also
+	highlighted as a pain point by other research software developers
+	\citet{PintoEtAl2018, PintoEtAl2016, WieseEtAl2019}.
+
+\end{enumerate}
+
 Developers identified a resource related pain point caused by a lack of funding

### 2022-04-20 StateOfPractice/Papers/MedImageSoft_SOP-arXiv/MedImageSoft_SOP_AcademicRadiology.tex
Fixed some typos

@@
 XMedCon & 26 & 19 & n/a & n/a & n/a \\
-DicomBrower & 27 & 21 & n/a & n/a & n/a \\
+DicomBrowser & 27 & 21 & n/a & n/a & n/a \\
 AMIDE & 28 & 22 & n/a & n/a & n/a \\
@@
 model). We do not know whether the prevalent use of GitHub issues for user
-support is by design, or whether this just naturally happenes as users seek
+support is by design, or whether this just naturally happens as users seek
 help. The common use of GitHub by MI developers is not surprising, given that

### 2022-04-20 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Edits up to and including Section 11 (Threats to Validity)

@@
 	the availability of well written tutorials that include expected output,
-	with software unit testing, and with evidence of continuous integration. In
-	our process we measure correctness and verifiability together.
+	with software unit testing, and with evidence of Continuous Integration
+	(CI). In our process we measure correctness and verifiability together.
 
@@
 
-The use of continuous integration tools and techniques alludes to a more refined
-development process where faults are isolated and better recognized. Only three
-of the packages (ESPResSo, Ludwig, Musubi) mentioned applying the practice of
-continuous integration in their development process.
+The use of Continuous Integration (CI) tools and techniques alludes to a more
+refined development process where faults are isolated and better recognized.
+Only three of the packages (ESPResSo, Ludwig, Musubi) mentioned applying the
+practice of CI in their development process.
 
@@
 ranked packages, Ludwig and Musubi, indicate the use of unit testing in their
-documentation. They also incorporated continuous integration in the development
+documentation. They also incorporated CI in the development
 process. Furthermore, eight of the top 10 ranked packages are noted as being
@@
 \toprule
-~ \ & \cite{USGS2019} & \cite{TobiasEtAl2018} & \cite{BrettEtAl2021} &
-\cite{WilsonEtAl2016} & \cite{SmithAndRoscoe2018} & \cite{HerouxEtAl2008} &
-\cite{ThielEtAl2020} & \cite{vanGompelEtAl2016} & \cite{OrvizEtAl2017} & LBM\\
+~ \ & \citet{USGS2019} & \citet{TobiasEtAl2018} & \citet{BrettEtAl2021} &
+\citet{WilsonEtAl2016} & \citet{SmithAndRoscoe2018} & \citet{HerouxEtAl2008} &
+\citet{ThielEtAl2020} & \citet{vanGompelEtAl2016} & \citet{OrvizEtAl2017} &
+LBM\\
 \midrule
@@
 
-\item \textbf{API documentation} is \wss{fill in definition, examples and
-benefits.  talk about doxygen to support.}
+\item \textbf{API documentation} shows programmers the services or data provided
+by the software application (or library) through such resources as its methods
+or objects \citep{MengEtAl2018}.  API documentation has been found to be a
+critical factor for API understandability (Section~\ref{softwarequalities})
+\citep{MengEtAl2018}. Some API documentation can be generated using tools like
+Doxygen, but the when learning a new API developers mostly use code examples and
+tutorials \citep{MengEtAl2018}.  API documentation was rare within our sample of
+LBM packages.
 
@@
 Table~\ref{tbl_tools}, debuggers were also likely used.  Only three (ESPResSo,
-Ludwig and Musubi) packages mentioned continuous integration tools, like Travis
+Ludwig and Musubi) packages mentioned CI tools, like Travis
 CI. Code editors and compilers were explicitly noted to have been used by
@@
 integration, and document generation tools.
+
+\subsection{Version Control Usage Compared to Research Software in General}
 
@@
 
-As mentioned in Section~\ref{SecSurfCorrectAndVerifiab}, continuous integration
-is rarely used in LBM software (3 of 24 packages or 12.5\%). This contrasts with
-the frequency with which continuous integration is recommended in research
-software development guidelines \citep{BrettEtAl2021, Brown2015, ThielEtAl2020,
-Zadka2018, vanGompelEtAl2016}.  \wss{why continuous integration.}  We could not
-find published data on the frequency with which continuous integration is used
-in general research software.  \wss{add reference to Carver paper.}  Our
-impression is that like LBM software, other research software packages lag
-behind the recommended best practice of employing continuous integration.
-
-Interviews with the developers revealed a potentially more frequent use of both
-unit testing and continuous integration, compared to what was observed from
-studying the repository artifacts.
+\subsection{CI Usage Compared to Research Software in General}
+
+As mentioned in Section~\ref{SecSurfCorrectAndVerifiab}, CI is rarely used in
+LBM software (3 of 24 packages or 12.5\%), although interviews with developers
+suggest that the actual rate might be higher. Low use of CI for LBM contrasts
+with how frequently it is recommended in research software development
+guidelines
+\citep{BrettEtAl2021, Brown2015, ThielEtAl2020, Zadka2018, vanGompelEtAl2016}
+and its popularity with non-research software. \citet{HiltonEtAl2016} show CI
+usage across a large sample of GitHub projects (34,544 projects) is at least
+40\%, with more popular projects (as measured by stars), having even higher
+usage rates. Moreover, \citet{HiltonEtAl2016} predict that CI adoption rates for
+open-source projects will increase even further in the future.  We could not
+find published data on the CI usage rates for research software in general.
+However, we saw 17\% utilization in the specific domain of imaging software
+\citep{Dong2021}, which is comparable to the LBM usage rate. Our impression is
+that medical imaging software and LBM software are not unique -- most research
+software domains likely lag behind the best practice recommendations, and
+non-research software usage rates, for continuous integration.
+
+CI is the process of building and testing software on every push to the code
+repository, with pushes done very frequently \citep[p.\ 13]
+{HumbleAndFarley2010}, \citep{ShahinEtAl2017, Fowler2006}.  CI consists of the
+following elements:
+\begin{itemize}
+	\item A version control system \citep{Fowler2006}. To be effective, all
+	files should be under version control, not just code files.  Anything that
+	is needed to build, install and run the software should be under version
+	control, including configuration files, build scripts, test harnesses, and
+	operating system configurations \citep[p. 19]{HumbleAndFarley2010}.
+	Fortunately for the LBM domain, as discussed above, most alive projects
+	(73\% of those assessed) use version control.
+	\item A fully automated build system \citep{Fowler2006}.  As \citet[p.\
+	5]{HumbleAndFarley2010} point out, deploying software manually is an
+	anti-pattern.  Fortunately, as discussed above, most LBM projects use build
+	automation.
+	\item An automated test system \citep{Fowler2006}. Building quality software
+	involves creating automated tests at the unit, component, and acceptance
+	test level, and executing these tests whenever a change is made to the code,
+	its configuration, the environment, or the software stack that it runs on
+	\citep[p.\ 83]{HumbleAndFarley2010}. As Table~\ref{artifactspresent} shows
+	test cases are in the uncommon category for LBM software artifacts, which
+	means that some LBM projects will need to increase their testing automation
+	if they wish to pursue CI.  The usual advice for CI is to keep the build and
+	test process short \citep[p.\ 60]{HumbleAndFarley2010}. Given that LBM can
+	be computationally expensive, the tests run with every check-in may need to
+	focus on simple code interface tests, saving larger more complex tests for
+	less frequent runs.  A more sophisticated option for when tests create a
+	bottleneck for merges is CIVET (Continuous Integration,
+	Verification, Enhancement, and Testing), which solves this problem by
+	intelligently pinning, canceling, and if necessary, restarting jobs as
+	merges occur, so that two merges are being tested at all times
+	\citep{SlaughterEtAl2021}.
+	\item An automated system for other tasks, such as code checking,
+	documentation building and web-site updating.  These other tasks are not
+	essential parts of CI, but they can be incorporated to improve the quality
+	of the code and the communication between developers and users. For
+	instance, a static analysis (possibly via linters) of the code may find poor
+	programming practice or lack of adherence to adopted coding standards.
+	\citet{SlaughterEtAl2021} provides another example of an automated task --
+	checking that a test specification includes the test's motivation, a test
+	description, and a design description for all changes. 
+	\item An integration build system to pull everything together.  Every time
+	there is a check-in (for instance a pull request), the integration server
+	automatically checks out the sources onto the integration machine, starts a
+	build, runs tests, and informs the committer of the results. 
+\end{itemize}
+
+Although CI can take some time and effort to set-up and integrate into a team's
+workflow, the benefits can be significant, as follows:
+
+\begin{itemize}
+	\item Headaches associated with a separate integration phase are eliminated
+	\citep{Fowler2006}, \citep[p.\ 20]{HumbleAndFarley2010}. If integration of
+	the work of different developers, or even separate chunks of work by the
+	same developer, is postponed, integration problems are inevitable.  By
+	continuously integrating problems are immediately obvious, and the source of
+	the problem can be isolated to the small increment that was just committed.
+	\item Bugs can be quickly detected and removed \citep{Fowler2006} via
+	automated testing.  To improve productivity, defects are best discovered and
+	fixed at the point where they are introduced \citep[p.\
+	23]{HumbleAndFarley2010}.  Errors are not just found in the code, but also
+	in the files and scripts related to configuration management \citep[p.\
+	18]{HumbleAndFarley2010}.
+	\item Developers are always working on a stable base, since the master
+	branch will always be working.  A stable base passes all tests and, if the
+	CI system is configured with generators and linters, it will also have up to
+	date documentation and coding standard compliant code.  A stable base
+	improves developer productivity, allowing them to focus on coding, testing
+	and documentation.
+\end{itemize}
+
+Although there are still challenges, setting up a CI system has never been
+easier than it is today.  The option of installing a dedicated CI server (either
+physically or virtually) exists with tools such as
+\href{https://www.jenkins.io/} {Jenkins}, \href{http://buildbot.net/}
+{Buildbot}, \href{https://www.gocd.org/} {Go}, and
+\href{http://integrity.github.io/} {Integrity}. However, installation on your
+own server is not necessary since there are many hosted CI tool solutions
+available, such as: \href{https://travis-ci.org/} {Travis CI},
+\href{https://github.com/features/actions} {GitHub Actions} and
+\href{https://circleci.com/} {CircleCI}.  Getting started with a hosted CI is
+straightforward.  All that is required to begin is editing a few lines of a
+simple YAML configuration file in the project's root directory.
+
+Challenges that exist for adopting CI include lack of awareness and
+transparency, coordination and collaboration challenges, lack of expertise and
+skills, more pressure and workload for team members, general resistance to
+change, scepticism and distrust on continuous practices \citep{ShahinEtAl2017}.
+The most common reason given for not adopting CI is that developers on my
+project are not familiar enough with CI \citep{HiltonEtAl2016}.  These problems
+can be mitigated via planning and documentation, promoting a team mindset,
+adopting new rules and policies, improving testing activities, and decomposing
+development into smaller units \citep{ShahinEtAl2017}.
 
@@
 have lead developer roles. These lead developers review potential additions to
-the software. One of the developers (ESPResSo) that was interviewed noted that
-an ad hoc peer review process is used to assess major changes and additions.
-Using peer review (also called technical review) matches with recommended
-practice for research software \citep{HerouxEtAl2008, Givler2020, OrvizEtAl2017,
-USGS2019}.
+the software. One of the developers (ESPResSo) noted that an ad hoc peer review
+process is used to assess major changes and additions. Using peer review (also
+called technical review) matches with recommended practice for research software
+\citep{HerouxEtAl2008, Givler2020, OrvizEtAl2017, USGS2019}.
 
@@
 process \citep{Kelly2015}.
+
+\subsection{Design Principles and Priorities} \label{Sec_DesPrincipAndPriorities}
 
@@
 
-\wss{how to doc motivations and design principles. examples. advantages.}
+\subsection{Documentation as Part of the Development Process}
 
@@
 
+Although Table~\ref{Tbl_Guidelines} shows that design documentation is
+frequently suggested, the recommendations are often short on details. For
+instance, the guidelines usually encourage modular design, but are generally
+silent on what criteria to use for the modularization.  Little information is
+given on how to document the design itself. Some non-mutually exclusive ideas
+for documenting the design, roughly in order of increasing effort, are as
+follows:
+
+\begin{itemize}
+	\item Explicitly state the design goals and priorities, so that those using
+	and developing your software know part of the rationale for the design
+	decisions. Sailfish is an example where this has been done, as discussed in
+	Section~\ref{Sec_DesPrincipAndPriorities}.
+	\item Write down the likely changes that the design is intended to support.
+	Since most designs are implicitly (or explicitly) motivated to facilitate
+	future change, recording the likely changes can help readers understand the
+	design.
+	\item Explicitly record the philosophy or principle behind the
+	decomposition. One potential principle, related to the previously mentioned
+	list of likely changes, is the principle of information hiding
+	\citep{Parnas1972a}.  This principle supports design for change through the
+	``secrets'' of each module. A text blurb can be given for each module or
+	class to describe its secrets and the services it provides.
+	\item Rigorously document the design following the template for a Module
+	Guide (MG) \citep{ParnasEtAl1984}.  An MG organizes the modules in a
+	hierarchy by their secrets. The application of the Parnas approach to
+	scientific software has been illustrated by applying it to the example of a
+	mesh generator \citep{SmithAndYu2009}.
+	\item Graphically represent the design using Unified Modelling Language
+	class diagrams.  This approach is suited to object oriented design and
+	designs that use patterns \citep{Gamma1995}.
+	\item Explain the design using data flow diagrams to show typical use cases
+	for how inputs are transformed.
+	\item Using a textual description, list for each module (or class), the
+	state variables (if any), exported constants and all exported access
+	programs.  This shows the interface that can used to access each module's
+	services.
+	\item Document the Module Interface Specification (MIS)
+	\citep{HoffmanAndStrooper1995}. An MIS is an abstract model that 
+	formally shows the state variables, inputs and outputs
+	\citep{ElSheikhEtAl2004, SmithAndYu2009}.
+\end{itemize}
+
 \section{Developer Pain Points} \label{painpoints}
@@
 	\textbf{Technology Hurdles} Technology pain points for LBM developers
-	include setting up parallelization and continuous integration. The pain
+	include setting up parallelization and CI. The pain
 	point survey of \citep{WieseEtAl2019} also highlighted technical-related
 	problems like dependency management, cross-platform compatibility,
-	continuous integration, hardware issues and operating system issues.
+	CI, hardware issues and operating system issues.
 
@@
 Some LBM developers improve reproducibility via automating the verification
-tests and making them part of a continuous integration process (mentioned in
+tests and making them part of a CI process (mentioned in
 Section~\ref{Sec_CompareTools}).  Providing unit tests improves reproducibility
@@
 We will not be repeating ideas here that have been discussed in previous
-sections, such as continuous integration, documentation of APIs, etc.  Our aim
+sections, such as CI, documentation of APIs, etc.  Our aim
 is to mention ideas that are at least somewhat beyond conventional best
@@
 can be used to spot check code files, or even better as part of a continuous
-integration system.
+integration system, as discussed in Section~\ref{Sec_CompareTools}.
 
@@
 
-The threats to validity can be categorized into methodology and data collection
-issues.  With respect to our methodology, the measurement template may not be
-detailed enough to accurately capture some qualities. For example, there are
-only two measures of surface robustness. The measurement of robustness could be
-expanded, as it currently only measures unexpected input. Other intentional
-faults could be introduced to assess robustness, but this would require a larger
-investment of measurement time. Similarly, reusability
-(Section~\ref{reusabilityresults}) is assessed via the number of code files and
-LOC per file. While this measure is indicative of modularity, it is possible
-that some packages have many files, with few LOC, but the files do not contain
-source code that is easily reusable. The files may be poorly formatted, or the
-source code may be vague and have ambiguous identifiers. Maintainability may
-also be incorrectly estimated, since it is assumed to be improved by a higher
-ratio of comments to source code (Section~\ref{Sec_Maintainability}), but this
-is not necessarily true. Furthermore, the measurement of understandability
-(Section~\ref{Sec_SurfUnderstandability}) relies on 10 random source code files.
-By random chance, the 10 files that were chosen may not be a good representation
-of the understandability of that package.
-
-In our methodology, one individual measures each package and the beginning and
-ending of the measurement phase is separated by several months.  There is a
-threat that the measurement may be affected by unconscious biases, or the
-reviewers judgement may drift over time.  In previous work
-\citep{SmithEtAl2016}, we demonstrated that the measurement process is
+Our threats to validity categories come from an analysis of software engineering
+secondary studies by \citet{AmpatzoglouEtAl2019}.  We use the classification of
+\citep{AmpatzoglouEtAl2019} because studying a collection of software is a form
+of secondary study, where a secondary study analyses the data from a set of
+primary studies.  A common example of a secondary study is a systematic
+literature review.  Our methodology is essentially a systematic software review
+-- the primary studies are the software, and our work collects and analyzes
+these primary studies.
+
+\subsection{Reliability}
+
+A study is reliable if its data and analysis are independent of the specific
+researcher(s) doing the study.  That is, repetition of the study by a different
+researcher should achieve the same results, as long as the same methodology is
+used \citep{AmpatzoglouEtAl2019}.  For our study the identified reliability
+related threats include the following:
+
+\begin{itemize}
+\item In our methodology the manual measures are done by one individual for all
+packages. The measures may be influenced by the graders abilities, experience
+and biases.  
+\item Completing the manual measures of all packages took place over several
+months. During this time the contents of the software repositories may change
+and the reviewer's judgment may drift.
+\end{itemize}
+
+We investigated the risk of reviewer dependence in previous work
+\citep{SmithEtAl2016}, where we demonstrated that the measurement process is
 reasonably reproducible by having five products graded by a second reviewer. The
@@
 
-Regarding data collection, a risk to the validity of this assessment is missing
-or incorrect data. Some data may not have been measured due to technology issues
-like broken links. This issue arose with the measurement of Palabos, which had a
-broken link to its user manual, as noted in Section~\ref{Sec_CompareArtifacts}.
-Some pertinent data may not have been specified in public artifacts, or may be
-obscure within an artifact or web-page. For instance, the use of unit testing
-and continuous integration was mentioned in the artifacts of only three
-(ESPResSo, Ludwig, Musubi) packages (Section~\ref{SecSurfCorrectAndVerifiab}).
-However, interviews suggested a more frequent use of both unit testing and
-continuous integration. For example, OpenLB, pyLBM, and TCLB use such methods
-during development despite this not being explicit from an analysis of the
-material available online. Given that we could not interview a representative
-for every project, to keep the measures consistent we did not modify our initial
-repository inspection based CI count.
-
-Furthermore, design documentation was measured to be a ``less common'' artifact
-in this assessment, but it is probable that such documentation is part of all
-LBM packages. After all, developing research software is not a trivial
-endeavour. It is likely that many packages have such documentation but did not
-make it public, and due to this the measured data may not be a true reflection
-of software package quality.
-
-Another data collection concern is the gap between the measurement of Musubi and
-the measurement of all other packages (Section~\ref{identifysoftware}).
-Unfortunately we were not aware of the existence of Musubi until after the
-primary data collection period was completed. This gap has a secondary effect
-because it caused a mismatch between the time of manual and automated data
-collection (Section~\ref{repmetrics}).  With time and resource constraints we
-could not redo the manual measurement, but we did update the automated measures
-(like the number of files, LOC, etc.) when Musubi was added.
-
-Our comparison between our ranking and the community's ranking
-(Section~\ref{repmetrics}) may not be valid.  Using stars to estimate the
-community's ranking is not necessarily accurate.  Moreover, our overall AHP
-ranking makes the unrealistic assumption of equal weighting between qualities
-(Section~\ref{Sec_OverallQuality}).
-
-A final threat to validity is conceptual.  Our work assumes that following best
-practices leads to improved software quality.  As the advice in software
-engineering guidelines show, we are not alone in this assumption.  Ideally
-software qualities should be measured directly and then the impact of best
-practices on those qualities should be assessed.  However, researchers in this
-domain always face the challenge that many qualities, like maintainability,
-reusability and verifiability, are difficult to measure directly.  Therefore, we
-follow the usual approach of assuming that following procedures and adhering to
-standards yields a higher-quality product \citep[p.\ 112]{VanVliet2000}.
+\subsection{Construct Validity}
+
+Construct validity is threatened if metrics are used that do not measure what
+they are intended to measure \citep{AmpatzoglouEtAl2019}. The construct
+validity related threats for the LBM study include the following:
+
+\begin{itemize}
+\item Our approximation of popularity by stars and watches
+(Section~\ref{repmetrics}) may not be valid. 
+\item Our overall AHP ranking makes the unrealistic assumption of equal
+weighting between qualities (Section~\ref{Sec_OverallQuality}).
+\item Our measure of surface robustness in only based on two pieces of data per
+project. The measurement of robustness could be expanded, as it currently only
+measures the impact of unexpected input. Other intentional faults could be
+introduced, but this would require a larger investment of measurement time. 
+\item Reusability (Section~\ref{reusabilityresults}) is assessed via the number
+of code files and LOC per file. While this measure is indicative of modularity,
+it is possible that some packages have many files, with few LOC, but the files
+do not contain source code that is easily reusable. The files may be poorly
+formatted, or the source code may be vague and have ambiguous identifiers. 
+\item Maintainability may be incorrectly estimated, since it is assumed to be
+improved by a higher ratio of comments to source code
+(Section~\ref{Sec_Maintainability}). 
+\item The measurement of understandability
+(Section~\ref{Sec_SurfUnderstandability}) relies on 10 random source code files.
+By random chance, the 10 files that were chosen may not be representative. 
+\item We measure our software qualities indirectly, rather than directly.
+Unfortunately meaningful direct measures for qualities like maintainability,
+reusability and verifiability, are not available.  Therefore, we follow the
+usual approach of assuming that following procedures and adhering to standards
+yields a higher-quality product \citep[p.\ 112]{VanVliet2000}.
+\end{itemize}
+
+\subsection{Internal Validity} \label{Sec_InternalValidity}
+
+A study is internally valid if the discovered causal relations are trustworthy
+and cannot be explained by other factors \citep{AmpatzoglouEtAl2019}. The
+internal validity threats include the following:
+
+\begin{itemize}
+\item We may have missed a relevant software package in our search for LBM
+software packages (Section~\ref{identifysoftware}).
+\item A temporal gap exists between the manual measurement of Musubi and the
+measurement of all other packages (Section~\ref{identifysoftware}). This gap
+caused a mismatch between the time of manual and automated data collection
+(Section~\ref{repmetrics}), since we updated the automated measures (like the
+number of files, LOC, etc.) when Musubi was added.
+\item Our interviews were with a relatively small sample of four LBM developers.
+The pain points they expressed may not be representative of others in the
+community.
+\item In our manual measures of the software repositories some data may not have
+been measured due to technology issues like broken links. This issue arose with
+the measurement of Palabos, which had a broken link to its user manual, as noted
+in Section~\ref{Sec_CompareArtifacts}. 
+\item Our methodology assumes that artifacts will be present in the repositories
+to approach used for software development.  However, not all activities will
+necessarily leave a trace in the repositories. For instance, the use of unit
+testing and CI was mentioned in the artifacts of only three (ESPResSo, Ludwig,
+Musubi) packages (Section~\ref{SecSurfCorrectAndVerifiab}). However, interviews
+suggested a more frequent use of both unit testing and CI. For example, OpenLB,
+pyLBM, and TCLB use such methods during development despite this not being
+explicit from an analysis of the material available online.
+\end{itemize}
+
+Although we found that unit testing and CI were used more often than the
+repository artifacts suggested, we did not change the manual measurement
+results.  Given that we could not interview a representative for every project,
+we left the data as collected to keep our measures consistent.  Moreover, CI and
+unit testing done in private should not be counted as equivalent to the public
+versions we detected.  When CI and unit testing are done outside of the public
+repository, they are not as effective, since they are not available for others
+to inspect and reproduce.
+
+\subsection{External Validity}
+
+External validity concerns whether the results of the study can be generalized
+(applied) to other situations/cases \citep{AmpatzoglouEtAl2019}.  The scope of
+our work is LBM solvers.  We are confident that our search was exhaustive and
+that we did not miss any highly popular examples.  Therefore, our validity
+concerns about LBM software observations are internal
+(Section~\ref{Sec_InternalValidity}), not external.  However, our hope is that
+the trends observed and the lessons learned for LBM software can be applied to
+other research software.  With that in mind we identified the following threat
+to external validity:
+
+\begin{itemize}
+\item The development of LBM software is fundamentally different from other
+research software.
+\end{itemize}
 
@@
 	\item Increase the amount of API documentation. \wss{reason} (Section~\ref{Sec_CompareArtifacts})
-	\item Increase the use of continuous integration. \wss{reason} (Section~\ref{Sec_CompareTools})
+	\item Increase the use of CI. \wss{reason} (Section~\ref{Sec_CompareTools})
 	\item Explicitly document the software's motivation and design principles.
@@
 developers to learn about their pain points (Section~\ref{painpoints}).  The
-identified pain points include lack of development time, lack of software
-development experience, lack of incentive and funding, lack of external support,
-technology hurdles, ensuring correctness, usability, technical debt and
-documentation quality.  We found (Section~\ref{Sec_AddressConcerns} that
-developers are currently creating practices to address these pain points,
-including: designing for change, circumventing the oracle testing problem, and
-prioritizing documentation and usability.  To plan for future best practices, we
-identified eight potential directions (Section~\ref{Sec_Recommendations}), in
-order of increasing development effort:
+identified pain points \wss{expand into bullet list} include lack of development
+time, lack of software development experience, lack of incentive and funding,
+lack of external support, technology hurdles, ensuring correctness, usability,
+technical debt and documentation quality.  We found
+(Section~\ref{Sec_AddressConcerns} that developers are currently creating
+practices to address these pain points, including \wss{expand into list}:
+designing for change \wss{what pain point? why? do the same for the others in
+the list}, circumventing the oracle testing problem, and prioritizing
+documentation and usability.  To plan for future best practices, we identified
+eight potential directions (Section~\ref{Sec_Recommendations}), in order of
+increasing development effort:
 
@@
 	(Section~\ref{Sec_MorePapers})
-	\item Grow the number of contributors. \wss{reason - change this to
-	documentation?} (Section~\ref{Sec_GrowContributors})
+	\item Apply advice from the open-source community on how to grow the number
+	of contributors. \wss{reason - change this to documentation?}
+	(Section~\ref{Sec_GrowContributors})
 	\item Further address the oracle problem for testing.  \wss{reason}
@@
 performance benchmarks could be incorporated into the assessment. The accuracy
-of our estimate of participation level for continuous integration and unit
+of our estimate of participation level for CI and unit
 testing, which is not always visible in public artifacts, can be improved by
@@
 adopt the analysis techniques from \citep{weiss1998commonality}.
-\cite{SmithAndChen2004} and \cite{SmithMcCutchanAndCarette2017} provide examples
-that could be followed for a commonality analysis of a family of mesh generating
-software and a family of material models, respectively.
+\citet{SmithAndChen2004} and \citet{SmithMcCutchanAndCarette2017} provide
+examples that could be followed for a commonality analysis of a family of mesh
+generating software and a family of material models, respectively.
 
@@
 - check references
+- maybe add a continuous integration diagram?
 

### 2022-04-21 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Revision complete, now requires proof reading

@@
-%% 
 %% Copyright 2007-2020 Elsevier Ltd
@@
 \usepackage{hyperref}
+\usepackage{xcolor}
+\hypersetup{
+    colorlinks,
+    linkcolor={red!50!black},
+    citecolor={blue!50!black},
+    urlcolor={blue!80!black}
+}
+
 \usepackage{multirow}
@@
 The top LBM projects have adopted techniques for software verification that get
-around the oracle problem (mentioned earlier in Section~\ppref{P_Correctness}). In
-the absence of a general test oracle, the following techniques have been adopted
-by LBM developers:
+around the oracle problem (mentioned earlier in \ppref{P_Correctness}). In the
+absence of a general test oracle, the following techniques have been adopted by
+LBM developers:
 
@@
 
-\subsection{Grow the Number of Contributors} \label{Sec_GrowContributors}
+\subsection{Follow Advice From the Open-Source Community to Grow the Number of Contributors} \label{Sec_GrowContributors}
 
@@
 \begin{itemize}
-	\item Add roadmap documentation showing what is planned for the future.
-	\wss{reason} (Section~\ref{Sec_CompareArtifacts})
-	\item Add a code of conduct to explicitly show how developers should treat
-	one another. \wss{reason} (Section~\ref{Sec_CompareArtifacts})
-	\item Increase the use and enforcement of programming style guides.
-	\wss{reason} (Section~\ref{Sec_CompareArtifacts})
-	\item Use checklists to ensure best practices are followed by developers.
-	\wss{reason} (Section~\ref{Sec_CompareArtifacts})
-	\item Include uninstall instructions. \wss{reason} (Section~\ref{Sec_CompareArtifacts})
-	\item Increase the amount of API documentation. \wss{reason} (Section~\ref{Sec_CompareArtifacts})
-	\item Increase the use of CI. \wss{reason} (Section~\ref{Sec_CompareTools})
-	\item Explicitly document the software's motivation and design principles.
-	\wss{reason - where is this discussed the first time - should add more
-	information there.} (Section~\ref{Sec_CompareMethodologies})
+	\item Add roadmap documentation showing what is planned for the future to
+	provide continuity of purpose, facilitate stakeholder collaboration and
+	assist with prioritization. (Section~\ref{Sec_CompareArtifacts})
+	\item Add a code of conduct to improve inclusivity and attract a wider pool
+	of collaborators. (Section~\ref{Sec_CompareArtifacts})
+	\item Increase the use and enforcement of programming style guides to improve understandability and productivity. (Section~\ref{Sec_CompareArtifacts})
+	\item Use checklists to ensure best practices are followed by developers. (Section~\ref{Sec_CompareArtifacts})
+	\item Include uninstall instructions as a courtesy to users that wish to
+	save space and avoid polluting their build system.
+	(Section~\ref{Sec_CompareArtifacts})
+	\item Increase the amount of API documentation to improve API
+	understandability. (Section~\ref{Sec_CompareArtifacts})
+	\item Increase the use of CI to remove headaches associated with a separate
+	integration phase, quickly detect and remove bugs and improve productivity,
+	since developers will always working from a stable base.
+	(Section~\ref{Sec_CompareTools})
+	\item Explicitly document the software's motivation and design principles to
+	assist with on-boarding new developers.
+	(Section~\ref{Sec_CompareMethodologies})
 \end{itemize}
@@
 developers to learn about their pain points (Section~\ref{painpoints}).  The
-identified pain points \wss{expand into bullet list} include lack of development
-time, lack of software development experience, lack of incentive and funding,
-lack of external support, technology hurdles, ensuring correctness, usability,
-technical debt and documentation quality.  We found
-(Section~\ref{Sec_AddressConcerns} that developers are currently creating
-practices to address these pain points, including \wss{expand into list}:
-designing for change \wss{what pain point? why? do the same for the others in
-the list}, circumventing the oracle testing problem, and prioritizing
-documentation and usability.  To plan for future best practices, we identified
-eight potential directions (Section~\ref{Sec_Recommendations}), in order of
-increasing development effort:
+identified pain points include:
+
+\begin{description}
+\item [\ppref{P_LackDevTime}] Lack of development time, %1
+\item [\ppref{P_LackSoftDevExp}] Lack of software development experience, %2
+\item [\ppref{P_LackFunding}] Lack of incentive and funding, %3
+\item [\ppref{P_LackExtSupport}] Lack of external support, %4
+\item [\ppref{P_TechnologyHurdles}] Technology hurdles, %5
+\item [\ppref{P_Correctness}] Ensuring correctness, %6
+\item [\ppref{P_Usability}] Usability, %7
+\item [\ppref{P_TechDebt}] Technical debt, and %8
+\item [\ppref{P_Documentation}] Documentation quality. %9
+\end{description}
+
+We found (Section~\ref{Sec_AddressConcerns}) that developers are currently
+creating practices to address these pain points.  To address technical debt
+(\ppref{P_TechDebt}) some developers have committed to designing for change
+(Section~\ref{Sec_DesForChange}). To improve correctness (\ppref{P_Correctness})
+developers are finding ways to circumvent the oracle testing problem, such as
+using peer review and pseudo-oracles (Section~\ref{SecReproducibility}).
+Developers have prioritized documentation and usability
+(Section~\ref{Sec_RecommendDoc} to address the pain points of documentation and
+usability challenges (\ppref{P_Usability} and \ppref{P_Documentation}).  For the
+pain point related to external support (\ppref{P_LackExtSupport}), we found that
+this is currently addressed by groups like \href{https://bssw.io/} {Better
+Scientific Software (BSSw)},
+\href{https://www.software.ac.uk/} {Software Sustainability Institute}
+\citep{CrouchEtAl2013}, and \href{https://software-carpentry.org/}{Software
+Carpentry} \citep{WilsonAndLumsdaine2006, Wilson2016}. Other pain points do not
+seem to be addressed by current practices.  This motivates our eight
+recommendations for potential future directions
+(Section~\ref{Sec_Recommendations}), in order of increasing development effort:
 
 \begin{enumerate}
-	\item Employ linters. \wss{reason}  (Section~\ref{Sec_Linters})
-	\item Conduct rigorous peer reviews. \wss{reason}
-	(Section~\ref{Sec_PeerReview})
-	\item Write and submit papers on software. \wss{reason}
+	\item Employ linters to address \ppref{P_LackDevTime}, \ppref{P_LackSoftDevExp}, \ppref{P_TechnologyHurdles} and \ppref{P_TechDebt}. (Section~\ref{Sec_Linters})
+	\item Conduct rigorous peer reviews to address \ppref{P_LackDevTime},
+	\ppref{P_LackSoftDevExp}, \ppref{P_TechnologyHurdles} and
+	\ppref{P_TechDebt}.	(Section~\ref{Sec_PeerReview})
+	\item Write and submit papers on software to address \ppref{P_LackFunding}.
 	(Section~\ref{Sec_MorePapers})
 	\item Apply advice from the open-source community on how to grow the number
-	of contributors. \wss{reason - change this to documentation?}
+	of contributors to address \ppref{P_LackDevTime}.
 	(Section~\ref{Sec_GrowContributors})
-	\item Further address the oracle problem for testing.  \wss{reason}
-	(Section~\ref{Sec_FurtherOracle})
-	\item Augment theory manuals to include requirements information.
-	\wss{reason} (Section~\ref{Sec_AugTheoryMans})
+	\item Further address the oracle problem for testing, with techniques such as the method of manufactured solutions and metamorphic testing, to address
+	\ppref{P_Correctness}. (Section~\ref{Sec_FurtherOracle})
+	\item Augment theory manuals to include requirements information to address
+	\ppref{P_LackDevTime}, \ppref{P_Usability} and \ppref{P_Documentation}.
+	(Section~\ref{Sec_AugTheoryMans})
 	\item Improve re-runnability, repeatability, reproducibility and
-	replicability. \wss{reason}  (Section~\ref{Sec_Reproducibility})
-	\item Generate all things. \wss{reason}  (Section~\ref{Sec_GenAllThings})
+	replicability to improve \ppref{P_Correctness}.
+	(Section~\ref{Sec_Reproducibility})
+	\item Use Domain Specific Modelling to generate all LBM code, build scripts,
+	documentation, test cases from a stable knowledge base to address
+	\ppref{P_LackDevTime}, \ppref{P_LackSoftDevExp},
+	\ppref{P_TechnologyHurdles}, \ppref{P_Correctness}, \ppref{P_Usability},
+	\ppref{P_TechDebt} and \ppref{P_Documentation}.
+	(Section~\ref{Sec_GenAllThings})
 \end{enumerate}

### 2022-04-21 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Revisions to abstract, first two paragraphs of intro

@@
     extracted from the project repository) and developer interviews (consisting
-    of a set of 20 questions). From the measurement results, we ranked the
-    software using the Analytic Hierarchy Process (AHP). Our ranking was roughly
-    consistent with ranking by GitHub stars, suggesting at least a partial
-    correlation between the use of best practices and popularity. We find the
-    state of the practice to be healthy, with 67\% of the measured packages
-    ranking in the top five for at least one quality, the majority of LBM
-    generated artifacts corresponding to general recommendations from research
-    software developers, common use of version control (67\% of
-    packages) and adoption of a quasi-agile development process.  Areas of best
-    practice to potentially improve include adoption of continuous integration,
-    API documentation and enforcement of programming style guides.  We
-    interviewed four developers to gain insight into their current pain points.
-    Identified challenges include lack of development time, lack of funding, and
-    difficulty with ensuring correctness.  Developers are addressing these pain
-    points by designing for change, circumventing the oracle problem and
-    prioritizing documentation and usability. For future improvements we suggest
-    the following: employing linters, conducting rigorous peer reviews, writing
-    and submitting more papers on software, growing the number of contributors
-    by following current recommendations for open source projects, and
-    augmenting the theory manuals to include more requirements relevant
-    information.
+    of 20 questions). After measuring we ranked the software using
+    the Analytic Hierarchy Process (AHP). Our ranking was roughly consistent
+    with GitHub stars ranking, suggesting at least a partial correlation
+    between the use of best practices and popularity. We find the state of the
+    practice to be healthy, with 67\% of the measured packages ranking in the
+    top five for at least one quality, the majority of LBM generated artifacts
+    corresponding to general recommendations from research software developers,
+    common use of version control (67\% of packages) and the adoption of a
+    quasi-agile development process.  Areas of best practice to potentially
+    improve include adoption of continuous integration, API documentation and
+    enforcement of programming style guides.  We interviewed four developers to
+    gain insight into their current pain points. Identified challenges include
+    lack of development time, lack of funding, and difficulty with ensuring
+    correctness.  Developers are addressing these pain points by designing for
+    change, circumventing the oracle problem and prioritizing documentation and
+    usability. For future improvements we suggest the following: employing
+    linters, conducting rigorous peer reviews, writing and submitting more
+    papers on software, growing the number of contributors by following current
+    recommendations for open source projects, and augmenting the theory manuals
+    to include more requirements relevant information.
 
@@
 positional probability of a collection of particles moving through a lattice
-structure following a two step process: i) streaming, where the particles move
-along the lattice via links; and, ii) colliding, where energy and momentum is
+structure following a two-step process: i) streaming, where the particles move
+along the lattice via links; and, ii) colliding, where energy and momentum are
 transferred among particles that collide \citep{bao2011lattice}. As an example
 of an important application of LBM, Figure~\ref{Fig_coarctation} shows the flow
-in a defective aorta pre and post medical intervention. LBM has several
+in a defective aorta pre and post-medical intervention. LBM has several
 advantages over conventional CFD methods, including a simple calculation
@@
 
-A small sample of important applications of LBM include the following: designing
-fuel cells \citep{ZhangEtAl2018}, modelling groundwater flow
+A small sample of important applications of LBM includes the following:
+designing fuel cells \citep{ZhangEtAl2018}, modelling groundwater flow
 \citep{AnwarAndSukop2009}, and analyzing the flow of blood in the cardiovascular

### 2022-04-21 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Edits to introduction section

@@
 
-A small sample of important applications of LBM includes the following:
-designing fuel cells \citep{ZhangEtAl2018}, modelling groundwater flow
-\citep{AnwarAndSukop2009}, and analyzing the flow of blood in the cardiovascular
-system \citep{SadeghiEtAl2022b, SadeghiEtAl2022, SadeghiEtAl2020}.  As these
-examples illustrate, LBM can be used for tackling problems that impact such
-areas as environmental policy, manufacturing, and health and safety.  Given the
-important applications of LBM, users of the libraries will be concerned with
-software qualities like reliability, robustness, reproducibility, performance,
-correctness and verifiability.  Since their time is valuable, developers that
-create or modify LBM libraries will have additional quality concerns, including
-maintainability, reusability, and understandability.  With the quality
-considerations in mind, and the potentially overwhelming number of choices for
-existing LBM libraries, the time is right to assess the state of the practice.
-The goal of this report is to analyze the current state of LBM software
-development and compare it to the development practices employed for research
-software in general.  We want to highlight success stories that LBM developers
-can share amongst their community, and with the broader research software
-community, while at the same time looking for areas for potential future
-improvement.
-
-To focus our efforts, we devised 10 research questions.  The questions are used
-to structure the discussion in the paper, so for each research question below we
-point to the section that contains our answer.  The questions are inspired by
-the research questions that underpin our data collection methodology
-\citep{SmithEtAl2021, SmithAndMichalski2022}. We start with identifying the
-examples of LBM software where we have access to the source code:
+A small sample of safety and mission-critical applications of LBM includes the
+following: designing fuel cells \citep{ZhangEtAl2018}, modelling groundwater
+flow \citep{AnwarAndSukop2009}, and analyzing the flow of blood in the
+cardiovascular system \citep{SadeghiEtAl2022b, SadeghiEtAl2022,
+SadeghiEtAl2020}.  As these examples illustrate, engineers and scientists use
+LBM for tackling problems that impact such areas as environmental policy,
+manufacturing, and health and safety. Given the critical applications of LBM,
+users of the libraries are concerned with software qualities like reliability,
+robustness, reproducibility, performance, correctness and verifiability.  Since
+their time is valuable, developers of LBM libraries have additional quality
+concerns, including maintainability, reusability, and understandability.  With
+the quality considerations in mind, and the potentially overwhelming number of
+choices for existing LBM libraries, the time is right to assess the state of the
+practice. Therefore, we analyze the current state of LBM
+software development and compare it to the development practices employed for
+research software in general.  We want to highlight success stories that LBM
+developers can share amongst their community, and with the broader research
+software community, while at the same time looking for areas for potential
+future improvement.
+
+We use ten research questions to structure and focus our discussion. Our data
+collection methodology \citep{SmithEtAl2021, SmithAndMichalski2022} inspires the
+questions.  For each question below, we point to the section that contains our
+answer.  We start with identifying the examples of LBM software where we have
+access to the source code:
 
@@
 We next wish to assess the representative software to determine how well they
-apply current software development best practices.  By \emph{best practices} we
+apply current software development best practices.  By \emph{best practices}, we
 mean methods, techniques and tools that are generally believed to improve
-software development, like testing and documentation.  As we will discuss in
-Section~\ref{empiricalmeasures}, we will categorize our best practices around
-software qualities.  At this point in the process, to remove potential
-user/developer bias, we will base our assessment only on publicly available
-artifacts, where artifacts are the documents, scripts and code that we find in a
-project's public repository. Example artifacts include requirements,
+software development, like testing and documentation.  We will categorize our
+best practices around software qualities, as we will discuss in
+Section~\ref{empiricalmeasures}.  At this point in the process, to remove
+potential user/developer bias, we will base our assessment only on publicly
+available artifacts, where artifacts are the documents, scripts and code that we
+find in a project's public repository. Example artifacts include requirements,
 specifications, user manuals, unit test cases, system tests, usability tests,
@@
 not guarantee popularity, so we will also compare our ranking to how the user
-community itself ranks the identified projects.
+community ranks the identified projects.
 
@@
 
-To understand the state of the practice we wish to learn the frequency with
+To understand the state of the practice, we wish to learn the frequency with
 which different artifacts appear, the types of development tools used and the
@@
 questions about how LBM software compares to other research software.  For cases
-where data is available, we will compare to other specific domains of research
+where data is available, we can compare to other specific domains of research
 software, like medical imaging \citep{Dong2021} and ocean modelling
@@
 not distinguish an application domain as \emph{research software in general}. By
-comparing LBM software to research software in general we can determine the ways
-LBM follows the trends from other developer communities, and ways in which it is
-different.
+comparing LBM software to research software in general we can determine how LBM
+follows the trends from other developer communities and how it is different.
 
@@
 gain additional insight, we need to interview developers.  We need to learn
-their concerns, how they deal with these concerns; we need to learn what pain
-points exist for them.  We wish to know what practices are used by the top LBM
-projects, so that others can potentially emulate these practices. We also wish
-to identify new practices that LBM developers can adopt to improve their
-software in the future. The above points are covered by the questions outlined
-below:
+their concerns and how they deal with them; we need to understand their pain
+points. We wish to know what practices are used by the top LBM projects, so that
+others can potentially emulate these practices. We also wish to identify new
+practices that LBM developers can adopt to improve their software in the future.
+The question below cover these points:
 

### 2022-04-21 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Completion of edits to section 1

@@
 
-We investigated the research questions by applying the general methodology
+We investigate the research questions by applying the general methodology
 summarized in \citet{SmithEtAl2021, SmithAndMichalski2022}.  The specific
 application of the methodology to LBM is reviewed in Section~\ref{methodology},
-with the full details in \citep{Michalski2021}.  The current methodology updates
+with the full details in \citet{Michalski2021}.  Our methodology updates
 the approach used in prior assessments of domains like Geographic Information
@@
 
-With our methodology we start by identifying 24 LBM software packages.  We then
-approximately measure the application of best practices for each package by
-filling in a grading template. Our analysis points to areas where some LBM
-software packages fall short of current best practices.  This is not intended to
-be a criticism of any existing packages, especially since not every project is
-actively maintained.  Moreover, we recognize that not every project needs to
-achieve the highest possible quality.  However, rather than delve into the
-nuances of which software can justify compromising which practice we conducted
-our comparison under the ideal assumption that every project can have sufficient
-resources to match best practices.
-
-Compared with our previous methodology (as used in \citep{SmithEtAl2016} for
+We start by identifying 24 LBM software packages.  We then approximately measure
+the application of best practices for each package by filling in a grading
+template. Our analysis points to areas where some LBM software packages fall
+short of current best practices.  Our intention is not to criticize any existing
+packages, especially since not every project is actively maintained.  Moreover,
+we recognize that not every project needs to achieve the highest possible
+quality.  However, rather than delve into the nuances of which software can
+justify compromising which practice we conducted our comparison under the ideal
+assumption that every project can have sufficient resources to match best
+practices.
+
+Compared with our previous methodology (as used in \citet{SmithEtAl2016} for
 instance), measuring the application of best practices now also includes
 repository based metrics, such as the number of files, number of lines of code,
-percentage of issues that are closed, etc.  With the quantitative data in the
-grading template, we rank the software with the Analytic Hierarchy Process
-(AHP). After this, as another addition to our previous methodology, we interview
-some of the development teams to further understand the status of their
-development process. The quantitative and qualitative data is then used to
-answer the research questions (Sections~\ref{methodology}
+percentage of issues that are closed, etc.  We rank the software with the
+Analytic Hierarchy Process (AHP), using the quantitative data from the grading
+template. After this, as another addition to our previous methodology, we
+interview some of the development teams to understand the status of their
+development processes. We next answer the research questions using the collected
+quantitative and qualitative data (Sections~\ref{methodology}
 to~\ref{Sec_AddressConcerns}).  Finally, we summarize the threats to validity

### 2022-04-22 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Edits up to, but not including, Section 2.5

@@
 We developed a methodology for evaluating the state of the practice of research
-software \citep{SmithEtAl2021, SmithAndMichalski2022}. This methodology can be
-applied to a specific domain of research software, which in the current case is
-LBM software.  Since we group best practices around conventional software
-qualities, the first section below provides definitions for the qualities of
-interest (Section~\ref{softwarequalities}), along with information on how to
-measure these qualities. The following section (Section~\ref{Sec_OverallProcess})
-provides an overview of the steps we used to select, measure and compare LBM
-software.  The remaining sections provide details on each step in the
-methodology.
+software \citep{SmithEtAl2021, SmithAndMichalski2022} and applied it to the
+specific domain of LBM software.  Since we group best practices around
+conventional software qualities, the first section below provides definitions
+for the qualities of interest (Section~\ref{softwarequalities}), along with
+information on how to measure these qualities. The following section
+(Section~\ref{Sec_OverallProcess}) provides an overview of the steps we used to
+select, measure and compare LBM software.  The remaining sections provide
+details on each step in the methodology.
 
@@
 
-To ensure our assessment of best practices is organized and complete, our
-measurements are centred around software qualities.  For completeness we used
-the the list of software engineering qualities from \citet[p.\
-17--33]{GhezziEtAl2003}, excluding interoperability, productivity and
-timeliness.  We excluded interoperability because we are focusing on LBM
-software alone, not its interaction with other software.  The other two
-qualities were excluded because they are process qualities, and we are measuring
-product qualities.  To the list from \citet{GhezziEtAl2003} we added
-installability, since it is so critical to the success of any practical project,
-and reproducibility, since it is so important for research software. Below we
-provide information on each quality of interest, including information on how
-they are measured.
+To ensure our assessment of best practices is organized and complete, we centre
+our measurements around software qualities. For completeness, we use the list of
+software engineering qualities from \citet[p.\ 17--33]{GhezziEtAl2003},
+excluding interoperability, productivity and timeliness. We excluded
+interoperability because our focus is on LBM software alone, not its interaction
+with other software. Our exclusion of the other two qualities is because they
+are process qualities, and our focus is on product qualities. We added
+installability to our list since potential users will ignore software they
+cannot install.  Since the scientific method depends on reproducibility, we also
+added this quality. Below we provide information on each quality, including
+details on how they are measured.
 
@@
 	a specified environment \citep{ISO/IEC25010, lenhard2013measuring}. In our
-	case, effort includes the time spent finding and understanding the
-	installation instructions, the person-time and resources spent performing
-	the installation procedure, and the absence or ease of overcoming system
-	compatibility issues. An installability score improves if the software has a
-	means to validate the installation and instructions for uninstallation.
+	case, the effort includes the following: time spent finding and
+	understanding the installation instructions, the person-time and resources
+	spent performing the installation procedure, and the absence or ease of
+	overcoming system compatibility issues. An installability score improves if
+	the software has means to validate the installation and instructions for
+	uninstallation.
 
 	\item[Correctness] A software program is correct if it behaves according to
-	its stated specifications \citep[p.\ 17]{GhezziEtAl2003}. This requires that
-	the specification is available. Research software is unlikely to have an
-	explicit specification, since this is not common practice in the field. As a
-	consequence, the correctness of software often cannot be measured directly.
-	We assess correctness indirectly by looking for the following: availability
-	of a requirements specification, reference to domain theory, and explicit
-	use of tools and/or techniques for building confidence, such as
-	documentation generators and software analysis tools.
+	its stated specifications \citep[p.\ 17]{GhezziEtAl2003}. Therefore,
+	measuring correctness requires that a specification is available. However,
+	research software is unlikely to have an explicit specification. As a
+	consequence, correctness often cannot be measured directly. We assess
+	correctness indirectly by looking for the following: availability of a
+	requirements specification, reference to domain theory, and explicit use of
+	tools and/or techniques for building confidence, such as documentation
+	generators and software analysis tools.
 
@@
 	domain knowledge. A good measure of verifiability is further correlated with
-	the availability of well written tutorials that include expected output,
-	with software unit testing, and with evidence of Continuous Integration
-	(CI). In our process we measure correctness and verifiability together.
+	the following: availability of well-written tutorials that include expected
+	output, software unit tests, and evidence of Continuous Integration (CI).
+	Our process measures correctness and verifiability together.
 
@@
 	(hard real-time demands) \citep{IEEEStdGlossarySET1990, wiegers2003softreq}.
-	In this state of the practice assessment performance was not directly
-	quantitatively measured. Instead the documentation of each software package
-	was observed for information alluding to consideration of performance, such
-	testing results, or parallelization instructions.
+	In this state of the practice assessment performance is not directly
+	quantitatively measured. Instead we check the documentation for each
+	software package looking for information alluding to consideration of
+	performance, such as testing results or parallelization instructions.
 
@@
 	performance or other attributes, and satisfy new requirements
-	\citep{IEEEStdGlossarySET1990, boehm2007software}. In the current work,
-	maintainability is measured by the quality of the documentation artifacts,
-	and the presence of version control and issue tracking. These artifacts can
-	greatly decrease the effort needed to modify software. There are many
-	documentation artifacts that can improve maintainability, including user and
-	developer manuals, specifications, README files, change logs, release notes,
+	\citep{IEEEStdGlossarySET1990, boehm2007software}. In the current work, we
+	measure maintainability by the quality of the documentation and the presence
+	of version control and issue tracking. There are many documentation
+	artifacts that can improve maintainability, including user and developer
+	manuals, specifications, README files, change logs, release notes,
 	publications, forums, and instructional websites. We also asked developers
@@
 	large number of easily reusable components. Therefore, we look for high
-	modularization, which is defined as the presence of smaller components with
+	modularization, which we define as the presence of small components with
 	well defined interfaces. For this state of the practice assessment, we
-	assume that a good measure of reusability correlates with a high
-	number of code files, and the availability of API documentation.
+	assume that a good measure of reusability correlates with a high number of
+	code files, and the availability of API documentation.
 
@@
 	function \citep{ISO9126}. It is an artifact-dependent quality.
-	Understandability is different for the user-interface, source code, and
+	Understandability is different for the user interface, source code, and
 	documentation. In this state of the practice analysis, understandability
-	focuses on the source code. It is measured by the consistency of a
-	formatting style, the extend of modularization, the explicit identification
-	of coding standards, the presence of meaningful identifiers, and the clarity
-	of comments.
+	focuses on the source code. We  measure it by the consistency of  formatting
+	style, the extent of modularization, the explicit identification of coding
+	standards, the presence of meaningful identifiers, and the clarity of
+	comments.
 
 	\item[Visibility and Transparency] Visibility and transparency refer to the
-	extent to which all of the steps of a software development process, and the
-	current status of it, are conveyed clearly \citep[p.\ 32]{GhezziEtAl2003}.
-	In this state of the practice assessment a good measure of visibility and
-	transparency correlates with a well defined development process,
+	extent to which all of the steps of a software development process, and its
+	current status, are conveyed clearly \citep[p.\ 32]{GhezziEtAl2003}. In this
+	state of the practice assessment a good measure of visibility and
+	transparency correlates with a well-defined development process,
 	documentation of the development process and environment, and software
@@
 	original output \citep{BenureauAndRougier2017}. We measured reproducibility
-	qualitatively by asking developers if they have any concern that their
-	computational results won't be reproducible, and if they have taken any
+	qualitatively by asking developers if they have any concerns that their
+	computational results won't be reproducible and if they have taken any
 	steps to ensure reproducibility.
@@
 
-To answer~\rqref{RQ_WhatProjects} we needed to identify existing LBM projects.
-The candidate software was found through search engine queries targeting
-authoritative lists of software. We found LBM software listed on
-\href{https://github.com/} {GitHub} and \href{https://swmath.org/} {swMATH}, as
-well as through scholarly articles. The Domain Expert
-(Section~\ref{sec_vet_software_list}) was also engaged in selecting the
-candidate software.
-
-The following properties were considered when creating the list and reviewing
-the candidate software:
+To answer~\rqref{RQ_WhatProjects} we identify existing LBM projects.
+We find the candidate software through search engine queries targeting
+authoritative lists. LBM software is listed on \href{https://github.com/}
+{GitHub} and \href{https://swmath.org/} {swMATH} and found in scholarly
+articles. The Domain Expert (Section~\ref{sec_vet_software_list}) helps with
+selecting the candidate software.
+
+We consider the following properties when creating the list and reviewing the
+candidate software:
 
@@
 
-The initial list had 45 packages, including a few that were later found to not
-have publicly available source code, or to be incomplete projects.
+The initial list had 45 packages, including some that we later found we could
+not use because of incompleteness or an absence of publicly available source
+code.
 
@@
 To reduce the length of the list of software packages and to
-answer~\rqref{RQ_WhatProjects}, we filter the initial list of 45 packages.  The
-following filters were applied in the priority order listed.
+answer~\rqref{RQ_WhatProjects}, we filter the initial list of 45 packages. We
+apply the following filters in the priority order listed.
 
 \begin{enumerate}
-	\item Scope: Software is removed by narrowing what functionality is
-	considered to be within the scope of the domain.
-	\item Usage: Software packages are eliminated if their installation
-	procedure was missing or not clear and easy to follow.
-	\item Age: The older software packages (age being measured by the last date
-	when a change was made) are eliminated, except in the cases where an older
-	software package appears to be highly recommended and currently in use. 
+	\item Scope: We remove software by narrowing what functionality we
+    consider within the scope of the domain.
+	\item Usage: We eliminate software if its installation procedure is
+	missing or unclear.
+	\item Age: We eliminate the older software packages (with age measured by
+	the last date of a change), except when an older software package appears to
+	be highly recommended and currently in use. 
 \end{enumerate}
 
-For the third item in the above filter, software packages were characterized as
-`alive' if their related documentation had been updated within the last 18
-months. Packages were categorized as `dead' if the last update was more than 18
-month ago.
+For the third item in the above filter, we categorized software packages as
+`alive' if documentation updates happened within the last 18 months. Packages
+were categorized as `dead' if their last update were more than 18 month ago.
 
@@
 Filtering by scope, usage, and age decreased the size of the list to 23
-packages. Many of the 22 packages that were removed could not be tested as there
-was no installation guide, they were incomplete, source code was not publicly
-available, a license was needed, or the project was out of scope or not up to a
-standard that would support incorporating them into this study. These eliminated
-software packages are listed in the Appendix of \citet{Michalski2021}. Of the
-remaining 23 packages that were studied, some were kept on the list despite
-being marked as dead due to their prevalence on authoritative lists on LBM
-software and due to their surface excellence.  After the initial measurement
-exercise, which took place in mid 2020, we added one more package in January
-2022: Musubi.  Musubi did not come up in our initial search, but once we learned
-of it, we felt that this report was incomplete without including it.  With the
-addition of Musubi we did not update the manual collection of data for the
-previous packages, due to time constraints, but we did redo all data collection
-that was automated by tools.
-
-The final list of 24 software packages that were analyzed in this project can be
-found in two tables. Table~\ref{alivepackages} lists packages that fell into the
-`alive' category as of mid 2020 (January 2022 for Musubi), and
-Table~\ref{deadpackages} lists packages that were `dead' at that time. The
-tables include a hyperlink to each project and, when available, citations for
-relevant publications.  The final list of software packages shows considerable
-variation in purpose, size, user interfaces, and programming languages. For
-example, the OpenLB software package is predominantly a \CC{} package that makes
-use of hybrid parallelization and was designed to address a range of CFD
-problems \citep{heuveline2009towards}. The software package pyLBM is an
-all-in-one Python language package for numerical simulations
-\citep{graille2017pylbm}. ESPResSo is an extensible simulation package that is
-specifically for research on soft matter, and is written in \CC{} and Python
-\citep{weik2019espresso}. The HemeLB package is used for efficient simulation of
-fluid flow in several medical domains, and is written predominantly in C, \CC,
-and Python \citep{mazzeo2008hemelb}.
+packages. We removed 22 packages for the following reasons: we cannot install
+them, they are incomplete, source code is not publicly available, a license is
+needed, or the project is out of scope. \citet{Michalski2021} lists the
+eliminated software packages. Of the remaining 23 packages, we kept some despite
+their `dead` status because of their presence on authoritative LBM software
+lists and because of their surface excellence. After the initial measurement
+exercise, which took place in mid-2020, we added one more package in January
+2022: Musubi. We found Musubi after our initial search, but we felt obliged to
+include it since it clearly fits our criteria. Time constraints meant we didn't
+update the previous manual data collection, but we did redo the automated data
+collection.
+
+The final list of 24 software packages is in two tables.
+Table~\ref{alivepackages} lists packages that fell into the `alive' category as
+of mid-2020 (January 2022 for Musubi), and Table~\ref{deadpackages} lists
+packages that were `dead'. The tables include a hyperlink to each project and,
+when available, citations for relevant publications.  The final list of software
+packages shows considerable variation in purpose, size, user interfaces, and
+programming languages. For example, the OpenLB software package is predominantly
+a \CC{} package that uses hybrid parallelization and addresses many CFD problems
+\citep{heuveline2009towards}. The software package pyLBM is an all-in-one Python
+language package for numerical simulations \citep{graille2017pylbm}. ESPResSo
+consists of extensible \CC{} and Python code specifically for research on soft
+matter \citep{weik2019espresso}. The HemeLB package simulates fluid flow in
+several medical domains and is written predominantly in C, \CC, and Python
+\citep{mazzeo2008hemelb}.
 
@@
 but this is not reflected in our data because it was not present at the time of
-data collection (mid 2020). Despite the broken link, Palabos does have a
+data collection (mid-2020). Despite the broken link, Palabos does have a
 detailed and informative website. As far as we could observe, Ludwig does not

### 2022-04-22 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Edits completed for Section 2

@@
 \citet{SmithEtAl2021}.  For each software package (each column in the template),
-we fill-in the rows of the template. This process takes about two hours per
+we fill in the rows of the template. This process takes about two hours per
 package, with a cap of four hours. The time constraint is necessary so that the
-work load is feasible for a team as small as one, given our aim to cap the
-measurement phase at 160 person hours \citep{SmithEtAl2021}.  An excerpt of the
+workload is feasible for a team as small as one, given our aim to cap the
+measurement phase at 160 person-hours \citep{SmithEtAl2021}.  An excerpt of the
 template, in spreadsheet form, is shown in
@@
 
-The questions were designed to be unambiguous, quantifiable and measurable with
-limited time and domain knowledge. The measures are grouped under headings for
-each quality, and one for summary information
-(Figure~\ref{measurement_template_image}). The summary section provides general
-information, such as the software name, number of developers, etc.  We follow
-the definitions given by \citet{gewaltig2012quality} for the software
-categories.  Public means software intended for public use.  Private means
-software aimed only at a specific group, while the concept category is used for
-software written simply to demonstrate algorithms or concepts. The three
-categories of development models are: open source, where source code is freely
-available under an open source license; free-ware, where a binary or executable
-is provided for free; and, commercial, where the user must pay for the software
-product.
-
-Several of the qualities use the word ``surface''.  This is to highlight that,
-for these qualities in particular, the best that we can do is a shallow measure.
-For instance, we do not conduct experiments to measure usability. Instead, we
-are looking for an indication that usability was considered by the developers by
-looking for cues in the documentation, such as the presence of getting started
-instructions, a user manual or documentation of expected user characteristics.
-
-Tools were used to find some of the measurements, such as the number of files,
-number of lines of code (LOC), percentage of issues that are closed, etc. The
-tool \href{https://github.com/tomgi/git_stats}{GitStats} was used to measure
-each software package's GitHub repository for the number of binary files, the
-number of added and deleted lines, and the number of commits over varying time
-intervals. The tool \href{https://github.com/boyter/scc}{Sloc Cloc and Code
-(scc)} was used to measure the number of text based files as well as the number
-of total, code, comment, and blank lines in each GitHub repository.
-
-As in \citet{SmithEtAl2016}, Virtual machines (VMs) were used to provide an
-optimal testing environment for each package. VMs were used because it is
-easier to start with a fresh environment without having to worry about existing
-libraries and conflicts. Moreover, when the tests are complete the VM can be
-deleted, without any impact on the host operating system. The most significant
-advantage of using VMs is to level the playing field. Every software install
-starts from a clean slate, which removes ``works-on-my-computer'' errors.
+We designed the questions to be unambiguous, quantifiable and measurable with
+limited time and domain knowledge. We categorize them by quality, with an extra
+category for summary information (Figure~\ref{measurement_template_image}). The
+summary section provides general data, such as the software name, number of
+developers, etc. \citet{gewaltig2012quality} provide us with our software
+categories. Public means software available to all, while private means software
+available only to a specific group.  The concept category defines software
+written to demonstrate algorithms or concepts. The three categories of
+development models are open source, where source code is freely available under
+an open-source license; free-ware, where a binary or executable is provided for
+free; and commercial, where the user must pay.
+
+Several of the qualities use the word ``surface'' to highlight that, for these
+qualities, the best that we can do is a shallow measure. For instance, we do not
+conduct experiments to measure usability directly. Instead, we look for an
+indication that usability is considered by searching for cues in the
+documentation, such as the presence of getting started instructions, a user
+manual or documentation of expected user characteristics.
+
+We used tools to automatically measure the number of files, number of lines of
+code (LOC), percentage of closed issues, etc. GitStats measures each package's
+GitHub repository for the number of binary files, the number of added and
+deleted lines, and the number of commits over varying time intervals. Sloc Cloc
+and Code (scc) measures the number of text files and the number of total, code,
+comment, and blank lines in each GitHub repository.
+
+As in \citet{SmithEtAl2016}, we used Virtual machines (VMs) to provide an
+optimal testing environment for each package. We used VMs to create fresh
+environments, eliminating worry about existing libraries and conflicts.
+Moreover, when the tests are complete, the VM can be deleted, without impacting
+the host operating system. The most significant advantage of using VMs is to
+level the playing field. Every software install starts from a clean slate, which
+removes ``works-on-my-computer'' errors.
 
@@
 
-Once we have measured each package, we still need to rank them to answer
-\rqref{RQ_HighestQuality}.  To do this, we used the Analytic Hierarchy Process
-(AHP), a decision-making technique that is used to compare multiple options by
-multiple criteria. AHP was used for comparing and ranking the LBM software
-packages using the overall impression quality scores that were gathered in the
-measurement template.  AHP performs a pairwise analysis using a matrix and
-generates an overall score as well as individual quality scores for each
-software package. \citep{SmithEtAl2016} shows how AHP is applied to ranking
+Once we measure each package, we still need to rank them to answer
+\rqref{RQ_HighestQuality}. We do this via the Analytic Hierarchy Process (AHP),
+a decision-making technique that compares multiple options by multiple criteria.
+Our AHP algorithm uses the overall impression score, gathered via the
+measurement template, to compare and rank the LBM software. AHP performs a
+pairwise analysis using a matrix and generates an individual quality scores for
+each software package. \citet{SmithEtAl2016} shows the details on how AHP ranks
 software based on quality measures.
 
-This project used a tool we developed for conducting the AHP analysis. The tool
+Our ranking uses a tool we developed for conducting the AHP analysis. The tool
 includes a sensitivity analysis to ensure that the software package rankings are
 appropriate with respect to the uncertainty of the quality scores. For the
-sensitivity analysis we modified the score by 10\% for each package and verified
-that the overall ranking was stable.  The
+sensitivity analysis, we modified the score by 10\% for each package and
+verified that the overall ranking was stable.  The
 \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/AHP2020/LBM/README.txt}{README}
-file of the tool includes requirements and usage information.
+file for the tool includes requirements and usage information.
 
@@
 \rqref{RQ_Recommend}) require going beyond the quantitative data from the
-measurement template. To gain the required insight, we interviewed developers
-using a list of 20 questions from \citet{SmithEtAl2021}. The questions cover the
-background of the development teams, the interviewees, and the software itself.
-We ask the developers how they organize their projects. We also ask about their
+measurement template. To gain insight, we interviewed developers using a list of
+20 questions from \citet{SmithEtAl2021}. The questions cover the background of
+the development teams, the interviewees, and the software itself. We ask the
+developers how they organize their projects. We also ask about their
 understanding of the users. Some questions focus on the current and past
-difficulties, and the solutions the team has found, or will try. We also discuss
+difficulties, and the solutions the team has found or will try. We also discuss
 the importance of, and the current situation for, documentation. A few questions
@@
 semi-structured based on the question list; we ask follow-up questions when
-necessary. Based on our experience, the interviewees often bring up exciting and
+necessary. Based on our experience, the interviewees often raise exciting and
 unexpected ideas.
@@
 We requested interviews with developers for all 24 packages, except Munubi,
-since Munubi was added after our ethics approved had expired.  We were able to
-recruit four developers to participate in our study.  Our potential interview
-subjects were found from project websites, code repositories, publications, and
-biographic information on institution web-pages. Meetings were held using
-on-line meeting software (either Zoom or Teams), which facilitated recording and
-automatic transcription.  Each meeting lasted about an hour. The interview
-process was approved by the McMaster University Research Ethics Board under the
+since we added Munubi after our ethics approval expired. We found the developers
+from project websites, code repositories, publications, and biographic
+information on institution webpages. Four developers agreed to participate in
+our study. We met with them online, using either Zoom or Teams to facilitate
+recording and automatic transcription. Each meeting lasted about an hour. The
+McMaster University Research Ethics Board approved the interview process under
 application number
@@
 
-Since the LBM domain has a reasonably small scope, we are able to view the
-software as constituting a program family.  The concept of a program family is
-defined by \citet{parnas1976design} as ``a set of programs whose common
-properties are so extensive that it is advantageous to study the common
-properties of the programs before analyzing individual members''. Studying the
-common properties within a family of related programs is termed a domain
-analysis.
+Since the LBM domain has a reasonably small scope, we can view the software as
+constituting a program family.  \citet{parnas1976design} defines a program
+family as ``a set of programs whose common properties are so extensive that it
+is advantageous to study the common properties of the programs before analyzing
+individual members.'' Studying the common properties within a family of related
+programs is termed a domain analysis.
 
@@
 Variabilities are goals, theories, models, definitions and assumptions that
-differ between family members. Associated with each variability are its
-parameters of variation, which summarize the possible values for that
-variability, along with their potential binding time.  The binding time is when
-the value of the variability is set.  It could be set as specification time,
-build time (when the program is being compiled) or run time (when the code is
-executing). As an example, for the current assessment, the packages all have the
-commonality of using LBM techniques to simulate the motion of fluids.  For
-variabilities, the LBM packages can be distinguished by the dimension of
-problems they simulate, whether or not (and how) they support parallel
-computation, whether or not compressibility can be considered, whether reflexive
-boundaries can be employed, whether the simulation handles multi-fluids, whether
-the simulations handles turbulence, and whether complex geometries can be
-simulated.
+differ between family members. Associated with each variability are parameters
+of variation, which summarize the possible values for that variability, along
+with their potential binding time. The binding time is when the value of a
+variability is set. Variabilities are set at specification time, build time
+(when the program compiles) or run time (when the code executes). For the
+current assessment, the packages all have the commonality of using LBM
+techniques to simulate the motion of fluids. Variabilities for LBM packages
+include the following: the dimension of problems they simulate, whether or not
+(and how) they support parallel computation, whether or not compressibility can
+be considered, whether reflexive boundaries can be employed, whether the
+simulation handles multi-fluids, whether the simulations handles turbulence, and
+whether complex geometries can be simulated.
 
 To keep the current presentation focused, we only present the results of our
-shallow domain analysis. For the LBM domain a table was constructed that
-distinguishes the programs by their variabilities.  In research software the
-variabilities are often assumptions, as shown in Table~\ref{tbl_features}. The
-variabilities are listed as columns in Table~\ref{tbl_features} with the
-software listed in alphabetical order. A more complete analysis of the
-commonalities and variabilities between LBM packages is provided in
-\citet{Michalski2021}.
+shallow domain analysis. For the LBM domain, a table was constructed that
+distinguishes the programs by their variabilities. In research software, the
+variabilities are often assumptions. Table~\ref{tbl_features} lists the
+variabilities as columns, with the software alphabetically ordered.
+\citet{Michalski2021} provides a detailed analysis of the commonalities and
+variabilities between LBM packages.
 
@@
 (\rqref{RQ_WhatProjects}) and our ranking (\rqref{RQ_HighestQuality}). The
-Domain Expert is an important member of the state of the practice assessment
-team. Pitfalls exist if non-experts attempt to acquire an authoritative list of
-software, or try to definitively rank the software. Non-experts have the problem
-that they can only rely on information available on-line, which has the
-following drawbacks:
+Domain Expert is a crucial member of the state of the practice assessment team.
+Pitfalls exist if non-experts attempt to acquire an authoritative list of
+software or try to rank the software. Non-experts rely on online information,
+which has the following drawbacks:
 \begin{inparaenum}[i)]
-  \item the on-line resources could have false or inaccurate information; and,
-  \item the on-line resources could leave out relevant information that is so
-in-grained with experts that nobody thinks to explicitly record it.
+  \item the online resources could have false or inaccurate information, and 
+  \item the online resources could leave out relevant information that is too
+  ingrained with experts for them to explicitly record it.
 \end{inparaenum}
@@
 
-The Domain Expert has an important role with verifying the list of LBM packages.
-In advance of the first meeting with the Domain Expert, they were asked to
-create a list of top software packages in the domain.  This is done to help the
-expert get in the right mind set in advance of the meeting.  Moreover, by doing
-the exercise in advance, we avoid the potential pitfall of the expert approving
-the discovered list of software without giving it adequate thought.  The Domain
-Expert was also asked to vet the collected data and analysis.  In particular,
-they were asked to vet the proposed list of software packages and the AHP
-ranking.  These interactions were done via virtual meetings.
+The Domain Expert has a central role in verifying the list of LBM packages. In
+advance of the first meeting with the Domain Expert, we asked them to create a
+list of top software packages in the domain.  We did this to help the expert get
+in the right mindset for our meeting.  Moreover, by doing the exercise in
+advance, we avoid the potential pitfall of the expert approving the discovered
+list of software without giving it adequate thought. We asked the Domain Expert,
+via virtual meetings, to vet the collected data and analysis. In particular, we
+asked them to vet the proposed list of software packages and the AHP ranking.
 

### 2022-04-22 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Edits up to and including Section 3.1

@@
 \newcommand{\CC}{C\nolinebreak\hspace{-.05em}\raisebox{.4ex}{\small\bf
++}\nolinebreak\hspace{-.10em}\raisebox{.4ex}{\small\bf +}}
+
+\newcommand{\esp}{ESPResSo\nolinebreak\hspace{-.05em}\raisebox{.4ex}{\small\bf
 +}\nolinebreak\hspace{-.10em}\raisebox{.4ex}{\small\bf +}}
@@
 			ESPResSo & 1, 2, 3 & CUDA/MPI & Y & Y & Y & Y & Y & M, L\\
-			ESPResSo++ & 1, 2, 3 & MPI & Y & Y & Y & Y & Y & L\\
+			\esp{} & 1, 2, 3 & MPI & Y & Y & Y & Y & Y & L\\
 			HemeLB & 3 & MPI & Y & Y & Y & Y & Y & L\\
@@
 
-In this section we answer \rqref{RQ_HighestQuality} by using AHP to rank the
-identified LBM software against best practices.  The sections below summarize
-how the software packages compare based on the qualities discussed in
+We answer \rqref{RQ_HighestQuality} by using AHP to rank the identified LBM
+software against best practices.  The sections below summarize how the software
+packages compare based on the qualities discussed in
 Section~\ref{softwarequalities}, as measured via the measurement template
-(Section~\ref{empiricalmeasures}).  For the purpose of discussion, the last
-subsection combines all the qualities into one ``global'' ranking.  The full LBM
-data is available publicly on Mendeley Data \citep{Smith2022}.
+(Section~\ref{empiricalmeasures}).  The last subsection combines all the
+qualities into one ``global'' ranking.  The full LBM data is available publicly
+on Mendeley Data \citep{Smith2022}.
 
@@
 
-All of the 24 software packages that were tested have installation instructions.
-Most have their installation instructions located in one place, often in an
-instruction manual or on a web-page. Sometimes, like with Ludwig, incomplete
-installation instructions are found on a homepage, with more detailed
-instructions located on another web-page, or within the documentation.
-Installability of the software, and maintainability and correctness of the
-instructions themselves, are improved if all the instructions are in one
-location.
-
-All 24 packages on the short list can be installed on a Unix-like system. Eight
-packages could be installed on Windows, and five on macOS. Operating system
-requirements are listed in the documentation of 20 software packages, but
-compatible operating system versions are only specified in four packages
-(ESPResSo, LUMA, Mechsys, Sailfish). All but one of the software packages (TCLB)
-were tested on Ubuntu for this state of the practice assessment. TCLB was tested
-on CentOS, since CentOS is mentioned in its installation instructions.
+All 24 software packages have installation instructions. Most have their
+installation instructions located in one place, often in an instruction manual
+or a webpage. Sometimes, like with Ludwig, we found incomplete installation
+instructions in one place, but complete instructions elsewhere. Installability
+of the software, and maintainability and correctness of the instructions
+themselves, are improved if all the instructions are in one location.
+
+All 24 packages are installable on a Unix-like system. Eight packages are also
+installable on Windows, and five on macOS. Twenty packages list their operating
+system requirements, but only four (ESPResSo, LUMA, Mechsys, Sailfish) list
+compatible operating system versions. Ubuntu was used to test all but one of the
+packages. TCLB was tested on CentOS, as suggested by its installation
+instructions.
 
@@
 
-Errors encountered during the installation process were often quickly fixed
+Errors encountered during the installation process were often quickly fixed,
 thanks to descriptive error messages. Systems that provided vague error
 messages, such as messages that did not specify which action or file was at
-fault, were more difficult to troubleshoot. Only three software packages
-(HemeLB, LB3D, lbmpy) that displayed a descriptive error message were not
-recoverable, and most of these instances were due to hardware and operating
-system incompatibility, such as the requirement of CUDA. Fourteen software
-packages broke during installation. Some packages, such as LB2D-Prime and
-LB3D-Prime, did not provide a definitive message of the success or failure of
-installation. In these instances, validating the installation required
-performing a tutorial or running a script, as described below, if these were
-available.
-
-About half of the installation instructions are written as if the person doing
-the installation has none of the dependent packages installed. Unfortunately,
-software packages, like ESPResSo++, Ludwig, and LUMA, frequently don't list all
-of their dependencies, or provide only a partial list. Sometimes only an error
-message during the installation process informs the user of the requirement of
-these additional packages. A detailed rewrite of the installation instructions
-from the point of view of installation on a clean operating system is suggested.
-A clean environment can be achieved for testing purposes by using a virtual
-machine.
-
-Seventeen software packages require less than 10 dependencies to be installed.
+fault, were harder to troubleshoot. Only three software packages (HemeLB, LB3D,
+lbmpy) that displayed a descriptive error message were not recoverable, and most
+of these instances were due to hardware and operating system incompatibility,
+such as the requirement of CUDA. Fourteen software packages broke during
+installation. Some packages, such as LB2D-Prime and LB3D-Prime did not provide a
+definitive message of the success or failure of the installation. In these
+instances, validating the installation required performing a tutorial or running
+a script, as described below, if these were available.
+
+About half of the installation instructions assume the installation environment
+has not already installed dependent packages. Unfortunately, software packages,
+like \esp, Ludwig and LUMA, frequently don't list all their dependencies.
+Sometimes only an error message during the installation process informs the user
+of the requirement for additional packages. In case like this, we suggest a
+detailed rewrite of the installation instructions to accommodate a clean
+operating system. A virtual machine provides such an environment for testing
+purposes.
+
+Seventeen software packages require installation of less than ten dependencies.
 All but one software package (LatBo.jl) require less than 20 dependencies. Some
@@
 software dependencies. Some software package installation issues, specifically
-those occurring when manual installation of dependencies is required, may be
-avoided if versions of dependencies are specified. Sixteen software packages do
-not have detailed instructions for installing dependencies. Sixteen software
-packages have less than 10 manual installation steps. If dependencies are
-installed in one command then none of the software packages take more than 20
-steps to install. The average number of steps is about eight, and the fewest is
-two (LB3D-Prime).
+those occurring when the manual installation of dependencies is required, might
+be avoided if versions of dependencies are specified. Sixteen software packages
+do not have detailed instructions for installing dependencies. Sixteen software
+packages have less than ten manual installation steps. If one command installs
+dependencies, then none of the software packages take more than 20 steps to
+install. The average number of steps is eight, and the fewest is two
+(LB3D-Prime).
 
 All but six (ESPResSo, HemeLB, laboetie, LB3D-Prime, lbmpy, waLBerla) of the
-software packages have a way to verify the installation. Most have some sort of
-tutorial examples that can be run by the user. Some other ways of installation
-validation include validation scripts (LB2D-Prime, lettuce, Ludwig, LUMA),
-automatic validation after the installation (LatBo.jl), and instructions to
-manually review the file system (LIMBES).  Uninstallation instructions were
-found for only one of the software packages: pyLBM.
+software packages have a way to verify the installation. Most have tutorial
+examples for the user to run. Some other approaches to installation validation
+include validation scripts (LB2D-Prime, lettuce, Ludwig, LUMA), automatic
+validation after the installation (LatBo.jl), and instructions to manually
+review the file system (LIMBES). Only one software package (pyLBM) has
+uninstallation instructions.
 
@@
 Figure~\ref{Fig_Installability} shows the installability ranking of the software
-packages. Software packages with a higher score (ESPResSo, MechSys, OpenLB,
-DL\_MESO, Palabos) tend to have one set of linear installation instructions
-written as if the person doing the installation has none of the dependencies
-installed. The instructions often list compatible operating system versions and
-include instructions for the installation of dependencies. The top ranked
-packages often incorporate some sort of automation of the installation process
-and have fewer manual installation steps. The number of dependencies a package
-has does not correlate with a higher score. The ability to validate the
-installation process, often through tutorials or test examples that include
-expected output, is correlated with a higher score. Active work on a project
-seems to improve installability, since the top seven ranked packages are all
-classified as alive.
+packages. Software packages with a higher score (ESPResSo, MechSys, OpenLB, DL
+MESO, Palabos) tend to have one set of linear installation instructions written
+as if the person doing the installation has none of the dependencies installed.
+The instructions often list compatible operating system versions and include
+instructions for installing dependencies. The top-ranked packages often
+incorporate some automation of the installation process. The number of
+dependencies a package has does not correlate with a higher score. The ability
+to validate the installation process, say through tutorials or test cases, leads
+to a higher score. Apparently active work on a project improves installability
+since the top seven ranked packages are all classified as alive.
 
@@
 within other documents. When present, theory related information is generally
-found within a user manual, on a web-page, or in a cited publication. In the
+found within a user manual, on a webpage, or in a cited publication. In the
 latter case significant time may be needed to track down the information.
@@
 Of the 14 software packages that installed correctly and also have tutorials,
-four (pyLBM, ESPResSo++, LIMBES, Ludwig) broke during tutorial testing. All of
+four (pyLBM, \esp, LIMBES, Ludwig) broke during tutorial testing. All of
 these instances resulted in an error message being displayed. One error (pyLBM)
@@
 user manuals are in the form of a file that can be downloaded, while some are
-rendered on a web-page. Some packages (waLBerla) do not have a user manual, but
-do have useful documentation distributed throughout their web-pages. Expected
+rendered on a webpage. Some packages (waLBerla) do not have a user manual, but
+do have useful documentation distributed throughout their webpages. Expected
 user characteristics are documented in five software packages (laboetie, LIMBES,
@@
 which was found in various artifacts, including in developer guides, contributor
-guides, user guides, developer web-pages, and README files.
+guides, user guides, developer webpages, and README files.
 
@@
 			LIMBES & 17.39 & Not Git & Dead\\
-			ESPResSo++ & 17.10 & 66.28 & Alive\\
+			\esp & 17.10 & 66.28 & Alive\\
 			HemeLB & 16.68 & No Issues & Dead\\
@@
 lines of source code, with over four million. The next largest package is
-ESPResSo++ with one million.
+\esp{} with one million.
 
@@
 			pyLBM & 108 & 37234 & 272 & 137 \\
-			ESPResSo++ & 30 & 165194 & 1406& 118 \\
+			\esp & 30 & 165194 & 1406& 118 \\
 			HemeLB & 44 & 123806 & 1102& 112 \\
@@
 count, but relatively low in the ``follow best practices'' ranking: TCLB (star
-rank 3, our rank 12), lettuce (star rank 4, our rank 14) and ESPResSo++ (star
+rank 3, our rank 12), lettuce (star rank 4, our rank 14) and \esp{} (star
 rank 5, our rank 15).  Part of the discrepancy is because all projects are
@@
 the community's feeling in the past more than current preferences
-\citep{Szulik2017}. Older packages (like ESPResSo++ and TCLB) have had more time
+\citep{Szulik2017}. Older packages (like \esp{} and TCLB) have had more time
 to accumulate stars and watches. Young projects have less time to accumulate
@@
 			lettuce & 14 & 48 & 4 & 5& 8\\
-			ESPResSo++ & 15 & 35 & 5 & 12& 4\\
+			\esp & 15 & 35 & 5 & 12& 4\\
 			MP-LABS & 16 & 12 & 11 & 2& 9\\			
@@
 Table~\ref{tbl_tools} shows the types of project management tools that were
-explicitly noted in the artifacts, web-pages, or interviews with the developers.
+explicitly noted in the artifacts, webpages, or interviews with the developers.
 As with development tools and dependencies, it is possible that other types of
@@
 table, we have 16 top performers: ESPREsSo, MechSys, OpenLB, DL\_MESO, Sailfish,
-Ludwig, Musubi, pyLBM, waLBerla, laboetie, lettuce, ESPREsSo++, LUMA, MP-LABS,
+Ludwig, Musubi, pyLBM, waLBerla, laboetie, lettuce, \esp, LUMA, MP-LABS,
 Palabos, and HemeLB.  Therefore, two thirds (67\%) of the 24 measured packages
@@
 			\addlinespace[0.4cm]
-			Surface Robustness & DL\_MESO (LBE) & ESPResSo++ & lettuce & LUMA &
+			Surface Robustness & DL\_MESO (LBE) & \esp & lettuce & LUMA &
 			OpenLB \\
@@
 			\addlinespace[0.4cm]
-			Maintainability & Ludwig & MP-LABS* & ESPResSo++ & ESPResSo & Palabos\\
+			Maintainability & Ludwig & MP-LABS* & \esp & ESPResSo & Palabos\\
 			\addlinespace[0.4cm]
-			Reusability & ESPResSo & waLBerla & ESPResSo++ & Sailfish & HemeLB*\\
+			Reusability & ESPResSo & waLBerla & \esp & Sailfish & HemeLB*\\
 			\addlinespace[0.4cm]

### 2022-04-23 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Edits from LanguageTool corrections

@@
 \affiliation[CAS]{organization={McMaster University, Computing and Software
-Department},%Department and Organization
+Department}, %Department and Organization
             addressline={1280 Main Street West}, 
@@
 
-\affiliation[ME]{organization={McMaster University, Mechanical Engineering},%Department and Organization
+\affiliation[ME]{organization={McMaster University, Mechanical Engineering}, %Department and Organization
             addressline={1280 Main Street West}, 
@@
     extracted from the project repository) and developer interviews (consisting
-    of 20 questions). After measuring we ranked the software using
+    of 20 questions). After measuring, we ranked the software using
     the Analytic Hierarchy Process (AHP). Our ranking was roughly consistent
@@
 incorporating additional physical complexities \citep{chen1998lattice}, such as
-reflective and non-reflective boundaries. LBM considers the behaviour of a
+reflective and non-reflective boundaries. LBM solvers consider the behaviour of a
 collection of particles as a single unit at the mesoscopic scale, which lies
-between the nanoscopic and microscopic scales. LBM solvers predict the
+between the nanoscopic and microscopic scales. These solvers predict the
 positional probability of a collection of particles moving through a lattice
 structure following a two-step process: i) streaming, where the particles move
-along the lattice via links; and, ii) colliding, where energy and momentum are
-transferred among particles that collide \citep{bao2011lattice}. As an example
-of an important application of LBM, Figure~\ref{Fig_coarctation} shows the flow
-in a defective aorta pre and post-medical intervention. LBM has several
-advantages over conventional CFD methods, including a simple calculation
-procedure, improved parallelization, and robust handling of complex geometries
+along the lattice via links; and, ii) colliding, where particles transfer energy
+and momentum when they collide \citep{bao2011lattice}. As an example of an
+important application of LBM, Figure~\ref{Fig_coarctation} shows the flow in a
+defective aorta pre- and post-medical intervention. LBM has several advantages
+over conventional CFD methods, including a simple calculation procedure,
+improved parallelization, and robust handling of complex geometries
 \citep{ganji2015application}.
@@
 manufacturing, and health and safety. Given the critical applications of LBM,
-users of the libraries are concerned with software qualities like reliability,
-robustness, reproducibility, performance, correctness and verifiability.  Since
+users worry about with software qualities like reliability,
+robustness, reproducibility, performance, correctness, and verifiability.  Since
 their time is valuable, developers of LBM libraries have additional quality
@@
 
-We use ten research questions to structure and focus our discussion. Our data
-collection methodology \citep{SmithEtAl2021, SmithAndMichalski2022} inspires the
+We use ten research questions to structure and focus our discussion. Our
+methodology \citep{SmithEtAl2021, SmithAndMichalski2022} is driven by the
 questions. For each question below, we point to the section that contains our
@@
 apply current software development best practices.  By \emph{best practices}, we
-mean methods, techniques and tools that are generally believed to improve
+mean methods, techniques, and tools that are generally believed to improve
 software development, like testing and documentation.  We will categorize our
@@
 potential user/developer bias, we will base our assessment only on publicly
-available artifacts, where artifacts are the documents, scripts and code that we
+available artifacts, where artifacts are the documents, scripts, and code that we
 find in a project's public repository. Example artifacts include requirements,
@@
 To understand the state of the practice, we wish to learn the frequency with
-which different artifacts appear, the types of development tools used and the
+which different artifacts appear, the types of development tools used, and the
 methodologies used for software development.  With this data, we can ask
@@
 	How do LBM projects compare to research software in general with respect to
-	principles, processes and methodologies used?
+	principles, processes, and methodologies used?
 	(Section~\ref{Sec_CompareMethodologies})
@@
 their concerns and how they deal with them; we need to understand their pain
-points. We wish to know what practices are used by the top LBM projects, so that
-others can potentially emulate these practices. We also wish to identify new
+points. We wish to know what practices top LBM projects use, so that
+others can potentially emulate these practices. Part of our goal is to identify new
 practices that LBM developers can adopt to improve their software in the future.
@@
 	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Concerns}:] For LBM
-	developers what specific best practices are taken to address the pain points
+	developers what specific best practices address the pain points
 	and software quality concerns? (Section~\ref{Sec_AddressConcerns})
 	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Recommend}:]
-	What research software development practice could potentially address the
-	pain point concerns identified in \rqref{RQ_PainPoints}).
+	What research software development practices could potentially address the
+	pain point concerns identified in \rqref{RQ_PainPoints}.
 	(Section~\ref{Sec_Recommendations})
@@
 template. After this, as another addition to our previous methodology, we
-interview some of the development teams to understand the status of their
+interview some development teams to understand the status of their
 development processes. We next answer the research questions using the collected
@@
 software engineering qualities from \citet[p.\ 17--33]{GhezziEtAl2003},
-excluding interoperability, productivity and timeliness. We excluded
+excluding interoperability, productivity, and timeliness. We excluded
 interoperability because our focus is on LBM software alone, not its interaction
@@
 	\item[Installability] Installability is measured by the effort required for
-	the installation, uninstallation or reinstallation of a software product in
+	the installation, uninstallation, or reinstallation of a software product in
 	a specified environment \citep{ISO/IEC25010, lenhard2013measuring}. In our
@@
 	In this state of the practice assessment performance is not directly
-	quantitatively measured. Instead we check the documentation for each
+	quantitatively measured. Instead, we check the documentation for each
 	software package looking for information alluding to consideration of
@@
 	software package can be used with or without adaptation in other software
-	packages \citep{kalagiakos2003non}. A good indicator of reusability is a
-	large number of easily reusable components. Therefore, we look for high
-	modularization, which we define as the presence of small components with
-	well defined interfaces. For this state of the practice assessment, we
-	assume that a good measure of reusability correlates with a high number of
-	code files, and the availability of API documentation.
+	packages \citep{kalagiakos2003non}. A good indicator of reusability is many
+	easily reusable components. Therefore, we look for high modularization,
+	which we define as the presence of small components with well-defined
+	interfaces. For this state of the practice assessment, we assume that a good
+	measure of reusability correlates with a high number of code files, and the
+	availability of API documentation.
 
@@
 	\item[Visibility and Transparency] Visibility and transparency refer to the
-	extent to which all of the steps of a software development process, and its
+	extent to which all the steps of a software development process, and its
 	current status, are conveyed clearly \citep[p.\ 32]{GhezziEtAl2003}. In this
@@
 `alive' if documentation updates happened within the last 18 months. Packages
-were categorized as `dead' if their last update were more than 18 month ago.
+were categorized as `dead' if their last update were more than 18 months ago.
 
@@
 
-We designed the questions to be unambiguous, quantifiable and measurable with
+We designed the questions to be unambiguous, quantifiable, and measurable with
 limited time and domain knowledge. We categorize them by quality, with an extra
@@
 
-We requested interviews with developers for all 24 packages, except Munubi,
-since we added Munubi after our ethics approval expired. We found the developers
+We requested interviews with developers for all 24 packages, except Musubi,
+since we added Musubi after our ethics approval expired. We found the developers
 from project websites, code repositories, publications, and biographic
@@
 The domain analysis (also called a commonality analysis) consists of
-understanding the commonalities, variabilities and common terminology for a
+understanding the commonalities, variabilities, and common terminology for a
 program family \citep{weiss1997defining}. Commonalities are goals, theories,
-models, definitions and assumptions that are common between family members.
-Variabilities are goals, theories, models, definitions and assumptions that
+models, definitions, and assumptions that are common between family members.
+Variabilities are goals, theories, models, definitions, and assumptions that
 differ between family members. Associated with each variability are parameters
@@
 techniques to simulate the motion of fluids. Variabilities for LBM packages
-include the following: the dimension of problems they simulate, whether or not
-(and how) they support parallel computation, whether or not compressibility can
+include the following: the dimension of problems they simulate, whether 
+(and how) they support parallel computation, whether compressibility can
 be considered, whether reflexive boundaries can be employed, whether the
-simulation handles multi-fluids, whether the simulations handles turbulence, and
+simulation handles multi-fluids, whether the simulation handles turbulence, and
 whether complex geometries can be simulated.
@@
 			laboetie & 2, 3 & MPI & Y & Y & Y & Y & Y & L\\
-			LatBo.jl & 2, 3 & - & Y & Y & Y & N & Y & L\\
+			LatBo.jl & 2, 3 & — & Y & Y & Y & N & Y & L\\
 			LB2D-Prime & 2 & MPI & Y & Y & Y & Y & Y & W, L\\
@@
 			LUMA & 2, 3 & MPI & Y & Y & Y & Y & Y & W, M, L\\
-			MechSys & 2, 3 & - & Y & Y & Y & Y & Y & L\\
+			MechSys & 2, 3 & — & Y & Y & Y & Y & Y & L\\
 			MP-LABS & 2, 3 & MPI/OMP & N & Y & Y & N & N & L\\
@@
 			Sailfish & 2, 3 & CUDA & Y & Y & Y & Y & Y & M, L\\
-			SunlightLB & 3 & - & Y & Y & N & N & Y & L\\
+			SunlightLB & 3 & — & Y & Y & N & N & Y & L\\
 			TCLB & 2, 3 & CUDA/MPI & Y & Y & Y & Y & Y & L\\
@@
 
-All but one of the software packages (LatBo.jl) have automated at least some of
+All but one of the software packages (LatBo.jl) have automated at least part of
 the installation process. Most of these packages, such as waLBerla and
@@
 Figure~\ref{Fig_Installability} shows the installability ranking of the software
-packages. Software packages with a higher score (ESPResSo, MechSys, OpenLB, DL
-MESO, Palabos) tend to have one set of linear installation instructions written
-as if the person doing the installation has none of the dependencies installed.
-The instructions often list compatible operating system versions and include
-instructions for installing dependencies. The top-ranked packages often
+packages. Software packages with a higher score (ESPResSo, MechSys, OpenLB,
+DL\_MESO, Palabos) tend to have one set of linear installation instructions
+written as if the person doing the installation has none of the dependencies
+installed. The instructions often list compatible operating system versions and
+include instructions for installing dependencies. The top-ranked packages often
 incorporate some automation of the installation process. The number of
@@
 
-Seventeen of the software packages explicitly reference domain theory, but a
-full and rigorous requirements specification, in the software engineering sense
-\citep{IEEE1998, RobertsonAndRobertson1999Vol, ESA1991}, is absent from all of
-the projects. Software packages that include a subset of a requirements
+Seventeen software packages explicitly reference domain theory, but a complete
+and rigorous requirements specification, in the software engineering sense
+\citep{IEEE1998, RobertsonAndRobertson1999Vol, ESA1991},  is absent from all
+projects. Software packages that include a subset of a requirements
 specification, such as DL\_MESO (LBE), keep the information brief and include it
@@
 found within a user manual, on a webpage, or in a cited publication. In the
-latter case significant time may be needed to track down the information.
+latter case tracking down the knowledge may take a significant time.
 
@@
 generation tool that helps with verification, thus building confidence in
-correctness. The top ranked software packages all include an easy to follow
-getting started tutorial, and most of these include expected output. The two top
-ranked packages, Ludwig and Musubi, indicate the use of unit testing in their
-documentation. They also incorporated CI in the development
-process. Furthermore, eight of the top 10 ranked packages are noted as being
-alive.
+correctness. The top ranked software packages all include getting started
+tutorials, and most of these include expected output. The two top ranked
+packages, Ludwig and Musubi, indicate the use of unit testing in their
+documentation. They also incorporated CI in the development process.
+Furthermore, eight of the top 10 ranked packages are noted as being alive.
 
@@
 software packages. Software packages with a high score either did not break
-during installation, or the broken installation was recoverable. All of the top
+during installation, or the broken installation was recoverable. All the top
 five ranked packages have tutorials. The package pyLBM broke during tutorial
@@
 packages. Software packages with a high score behaved reasonably in response to
-unexpected input as described above. All of the software packages that installed
+unexpected input as described above. All the software packages that installed
 correctly passed this test. They output descriptive error messages or at least
@@
 Table~\ref{tbl_features}). Due to this, a comparison of run-time performance is
-not appropriate. Instead we looked through each software package's artifacts for
+not appropriate. Instead, we looked through each software package's artifacts for
 evidence that performance was considered. The artifacts of 18 software packages
@@
 of their issues closed, and only three (laboetie, lettuce, Sailfish) have less
-than 50 percent of their issues closed. All of the top five overall ranked
+than 50 percent of their issues closed. All the top five overall ranked
 packages have most of their issues closed. Alive packages (11 use issue
@@
 16 packages, and the average percentage of code comments is about 14 percent.
-All of the top five overall ranked packages have more than the average. LUMA has
+All the top five overall ranked packages have more than the average. LUMA has
 only 0.2 percent comments, the fewest of any package. This package has the most
@@
 We measured the total number of source code files for each project. We assume
-that a large number of source files is associated with increased reusability,
+that numerous source files is associated with increased reusability,
 due to our assumption that this indicates increased modularization. Some
@@
 packages. Packages with a high reusability score do not have as many LOC per
-text file, generally having a few hundred lines or less. This suggests that the
+text file, generally having a few hundred lines or fewer. This suggests that the
 source code of these packages is likely functionally modularized, and modules
@@
 
-All of the packages appear to have consistent indentation and formatting. Only
+All the packages appear to have consistent indentation and formatting. Only
 HemeLB, LUMA, and Musubi explicitly identify coding standards that are used
@@
 packages. The constants are used for various parameters, mathematical constants,
-and matrix definitions. All of the packages are reasonably well commented, with
+and matrix definitions. All the packages are reasonably well commented, with
 the comments clearly indicating what is being done (as opposed to how it is
@@
 algorithms. Their comments are clear and indicate what is being done in the
-source code. The source code is well modularized and structured. All of the top
+source code. The source code is well modularized and structured. All the top
 five ranked packages are noted as being alive.
@@
 			ESPResSo & 1 & 145 & 2 & 19& 2\\
-			Ludwig & 2 & 27 & 8 & 6& 7\\
-			Palabos & 3 & 34 & 6 & GitLab& GitLab\\
+			Ludwig & 2 & 27 & 8 & 6 & 7\\
+			Palabos & 3 & 34 & 6 & GitLab & GitLab\\
 			OpenLB & 4 & N/A & N/A & N/A& N/A\\
@@
 			Sailfish & 9 & 186 & 1 & 41& 1\\
-			waLBerla & 10 & 20 & 9 & GitLab& GitLab\\
-			laboetie & 11 & 4 & 13 & 5& 8\\
-			TCLB & 12 & 95 & 3 & 16& 3\\
-			MechSys & 13 & N/A & N/A & N/A& N/A\\
-			lettuce & 14 & 48 & 4 & 5& 8\\
-			\esp & 15 & 35 & 5 & 12& 4\\
-			MP-LABS & 16 & 12 & 11 & 2& 9\\			
-			SunlightLB & 17 & N/A & N/A & N/A& N/A\\
-			LB3D & 18 & N/A & N/A & N/A& N/A\\			
-			LIMBES & 19 & N/A & N/A & N/A& N/A\\
-			LB2D-Prime & 20 & N/A & N/A & N/A& N/A\\		
-			HemeLB & 21 & 12 & 11 & 12& 4\\
-			lbmpy & 22 &  11 & 12 & 2& 9\\	
-			LB3D-Prime & 23 & N/A & N/A & N/A& N/A\\	
-			LatBo.jl & 24 & 17 & 10 & 8& 6\\			
+			waLBerla & 10 & 20 & 9 & GitLab & GitLab\\
+			laboetie & 11 & 4 & 13 & 5 & 8\\
+			TCLB & 12 & 95 & 3 & 16 & 3\\
+			MechSys & 13 & N/A & N/A & N/A & N/A\\
+			lettuce & 14 & 48 & 4 & 5 & 8\\
+			\esp & 15 & 35 & 5 & 12 & 4\\
+			MP-LABS & 16 & 12 & 11 & 2 & 9\\			
+			SunlightLB & 17 & N/A & N/A & N/A & N/A\\
+			LB3D & 18 & N/A & N/A & N/A & N/A\\			
+			LIMBES & 19 & N/A & N/A & N/A & N/A\\
+			LB2D-Prime & 20 & N/A & N/A & N/A & N/A\\		
+			HemeLB & 21 & 12 & 11 & 12 & 4\\
+			lbmpy & 22 &  11 & 12 & 2 & 9\\	
+			LB3D-Prime & 23 & N/A & N/A & N/A & N/A\\	
+			LatBo.jl & 24 & 17 & 10 & 8 & 6\\			
 			\bottomrule
@@
 the artifact is not valued.  The guideline authors may have excluded it because
-it is out of the scope of their recommendations, or outside of their experience.
+it is out of the scope of their recommendations, or outside their experience.
 For instance, an artifact related to uninstall is only explicitly mentioned by
-\citep{vanGompelEtAl2016}), but other guideline authors would likely see its
+\citet{vanGompelEtAl2016}, but other guideline authors would likely see its
 value.  They may simply feel that uninstall is implied by install, or they may
@@
 
-The top four packages, ESPREsSo, Ludwig, Palabos, and OpenLB, have all of the
+The top four packages, ESPREsSo, Ludwig, Palabos, and OpenLB, have all the
 common artifacts, except Palabos does not have theory notes and only three
@@
 however, some recommended practices were not observed or rarely observed. For
-instance API documentation was rarely observed for LBM software, but it is a
+instance, API documentation was rarely observed for LBM software, but it is a
 frequently recommended artifact \citep{SmithAndRoscoe2018, ThielEtAl2020,
@@
 organization to map out the vision and direction for a product offering
-\citep{MunchEtAl2019}.  Roadmaps take different forms but they all aim to
+\citep{MunchEtAl2019}.  Roadmaps take different forms, but they all aim to
 address the following: \begin{inparaenum}[i)]
@@
 	\item scan the environment, 
-	\item revise and distil the product vision to write the product roadmap, 
+	\item revise and distill the product vision to write the product roadmap, 
 	\item estimate the product life cycle and evaluate the mix of planned
@@
 \citep{MengEtAl2018}. Some API documentation can be generated using tools like
-Doxygen, but the when learning a new API developers mostly use code examples and
+Doxygen, but when learning a new API developers mostly use code examples and
 tutorials \citep{MengEtAl2018}.  API documentation was rare within our sample of
@@
 They all use various technical and domain specific libraries. Technical
-libraries include visualization (e.g.\ Matplotlib, ParaView, Pygame, VTK), data
-analysis (e.g.\ Anaconda, Torch), and message passing libraries (e.g.\ MPICH,
+libraries include visualization (e.g.,\ Matplotlib, ParaView, Pygame, VTK), data
+analysis (e.g.,\ Anaconda, Torch), and message passing libraries (e.g.,\ MPICH,
 Open MPI, PyZMQ). Domain specific libraries include research software libraries
-(e.g.\ SciPy).
+(e.g.,\ SciPy).
 
@@
 as being used when developing software projects. Project management software was
-not explicitly mentioned during our interviews, but it is possible that some of
-the projects use such software. Many of the projects are located on GitHub,
-where the developers use the platform to help manage their projects, especially
-bug related issues. Most of the projects appear to use change tracking and
-version control tools. Document generation tools are mentioned in 12 of the 24
-projects. The tools Sphinx and Doxygen are explicitly used in this capacity.
+not explicitly mentioned during our interviews, but it is possible that some
+projects use such software. Many of the projects are located on GitHub, where
+the developers use the platform to help manage their projects, especially bug
+related issues. Most of the projects appear to use change tracking and version
+control tools. Document generation tools are mentioned in 12 of the 24 projects.
+The tools Sphinx and Doxygen are explicitly used in this capacity.
 
@@
 software.  Both use tools for editing, compiling, code management, testing,
-building and project management.  From the data available, ocean modelling
+building, and project management.  From the data available, ocean modelling
 differs from LBM software in its use of Kanban boards for project management.
@@
 Section~\ref{Sec_Maintainability}, 67\% of LBM packages use version control
-(GitHub, GitLab or CVS).  (From Table~\ref{gitrepodata}, 11/15 or 73\% of alive
+(GitHub, GitLab, or CVS).  (From Table~\ref{gitrepodata}, 11/15 or 73\% of alive
 packages use version control.)  The proliferation of version control tools for
@@
 75\%, 89\% and 17\% (respectively) to 100\%, 95\% and 100\% (respectively) over
-a four year period ending in 2018.  (For completeness the same study showed a
+a 4-year period ending in 2018.  (For completeness the same study showed a
 decrease in version control usage for seismology software over the same time
@@
 	Verification, Enhancement, and Testing), which solves this problem by
-	intelligently pinning, canceling, and if necessary, restarting jobs as
+	intelligently pinning, cancelling, and if necessary, restarting jobs as
 	merges occur, so that two merges are being tested at all times
@@
 
-Although CI can take some time and effort to set-up and integrate into a team's
+Although CI can take some time and effort to set up and integrate into a team's
 workflow, the benefits can be significant, as follows:
@@
 	branch will always be working.  A stable base passes all tests and, if the
-	CI system is configured with generators and linters, it will also have up to
-	date documentation and coding standard compliant code.  A stable base
+	CI system is configured with generators and linters, it will also have current documentation and coding standard compliant code.  A stable base
 	improves developer productivity, allowing them to focus on coding, testing
@@
 developers were noted to play an increased role in these changes compared to the
-former changes. Software bugs are typically addressed in a similar fashion as
+former changes. Software bugs are typically addressed similarly to
 code refactoring.  Issue tracking is commonly used to manage these changes.
@@
 ThielEtAl2020, vanGompelEtAl2016, OrvizEtAl2017, FLOSS2022, JimenezEtAl2017}.
-Tutorials \citep{ThielEtAl2020}, trouble shooting guides \citep{OrvizEtAl2017,
+Tutorials \citep{ThielEtAl2020}, troubleshooting guides \citep{OrvizEtAl2017,
 SSI2022} and quick start examples \citep{ThielEtAl2020, vanGompelEtAl2016} are
@@
 	\item Graphically represent the design using Unified Modelling Language
-	class diagrams.  This approach is suited to object oriented design and
+	class diagrams.  This approach is suited to object-oriented design and
 	designs that use patterns \citep{Gamma1995}.
@@
 while coding, scope bloat, lack of user feedback, hard to collaborate on
-software projects, and aloneness. \citet{WieseEtAl2019} repeat some of the
-previous pain points and add the following: dependency management, data handling
-concerns (like data quality, data management and data privacy), reproducibility,
-and software scope determination. Although LBM developers did not mention these
-pain points, we cannot conclude that they are not relevant for LBM software
-development, since we only interviewed four LBM developers for about an hour each.
+software projects, and aloneness. \citet{WieseEtAl2019} repeat some previous
+pain points and add the following: dependency management, data handling concerns
+(like data quality, data management and data privacy), reproducibility, and
+software scope determination. Although LBM developers did not mention these pain
+points, we cannot conclude that they are not relevant for LBM software
+development, since we only interviewed four LBM developers for about an hour
+each.
 
@@
 	correctness of large numbers of features, and challenges with automated
-	automated testing. The TCLB developer commented that the amount of testing
+	testing. The TCLB developer commented that the amount of testing
 	data needed was sometimes problematic, since free testing services do not
@@
 	while in reality CFD knowledge needs to be applied to correctly modify the
-	packages for the new endeavour.  Some of the respondents in the survey of
+	packages for the new endeavour.  Some respondents in the survey of
 	\citep{WieseEtAl2019} also mentioned that users do not always have the
@@
 	noted several times that a lack of time and funding (\ppref{P_LackDevTime})
-	has a negative affect on the documentation. Most of the developers are
+	has a negative effect on the documentation. Most of the developers are
 	scientific researchers evaluated on the scientific papers that they produce.
@@
 	``tests which compare implementations against each other'' where the tests
-	``rely on different principles and they've been written by different
+	``rely on different principles, and they've been written by different
 	people''. \citet{latt2021palabos} shows several benchmark tests comparing
@@
 
-With the exception of \citet{ThielEtAl2020}, the research software guidelines
-that we consulted do not mention linters.  However, we feel linters have the
-potential to improve code quality at a relatively low cost.  A linter is a tool
-that statically analyzes code to find programming errors, stylistic
-inconsistencies and suspicious constructs \citep{Wikipedia2022_Lint}. Linters
-can be used to spot check code files, or even better as part of a continuous
-integration system, as discussed in Section~\ref{Sec_CompareTools}.
+Except for \citet{ThielEtAl2020}, the research software guidelines that we
+consulted do not mention linters.  However, we feel linters have the potential
+to improve code quality at a relatively low cost.  A linter is a tool that
+statically analyzes code to find programming errors, stylistic inconsistencies
+and suspicious constructs \citep{Wikipedia2022_Lint}. Linters can be used to
+spot check code files, or even better as part of a continuous integration
+system, as discussed in Section~\ref{Sec_CompareTools}.
 
@@
 specific review tasks (like summarize the purpose of the code, create a data
-dictionary for the module assigned, cross reference the code to the technical
+dictionary for the module assigned, cross-reference the code to the technical
 manual, etc.) The task based inspection approach has been effectively used for
@@
 (Section~\ref{Sec_Linters}): \ppref{P_LackDevTime}, \ppref{P_LackSoftDevExp},
-\ppref{P_TechnologyHurdles}), \ppref{P_TechDebt} and \ppref{P_Correctness}. Peer
+\ppref{P_TechnologyHurdles}, \ppref{P_TechDebt} and \ppref{P_Correctness}. Peer
 review addresses these pain points by efficiently searching for defects and
@@
     \item Reward and recognize new contributors, via small rewards like stickers
-    or shirts, or even just a simple shoutout or mention in a blogpost or on
+    or shirts, or even just a simple shoutout or mention in a blog post or on
     social media \citep{Jalan2016, Proffitt2017}.
@@
 
-DSM provide a transformative technology for documentation, design and
+DSM provides a transformative technology for documentation, design and
 verification \citep{JohansonAndHasselbring2018, Smith2018}. DSM allows
@@
 proposing addresses these directly, since documentation is a primary concern,
-rather than an afterthought. and quality of documentation
+rather than an afterthought.
 
@@
 \citep{AmpatzoglouEtAl2019} because studying a collection of software is a form
-of secondary study, where a secondary study analyses the data from a set of
+of secondary study, where a secondary study analyzes the data from a set of
 primary studies.  A common example of a secondary study is a systematic
@@
 weighting between qualities (Section~\ref{Sec_OverallQuality}).
-\item Our measure of surface robustness in only based on two pieces of data per
+\item The measure of surface robustness in only based on two pieces of data per
 project. The measurement of robustness could be expanded, as it currently only
@@
 unit testing done in private should not be counted as equivalent to the public
-versions we detected.  When CI and unit testing are done outside of the public
+versions we detected.  When CI and unit testing are done outside the public
 repository, they are not as effective, since they are not available for others
@@
 (Section~\ref{Sec_InternalValidity}), not external.  However, our hope is that
-the trends observed and the lessons learned for LBM software can be applied to
+the trends observed, and the lessons learned for LBM software can be applied to
 other research software.  With that in mind we identified the following threat
@@
 	integration phase, quickly detect and remove bugs and improve productivity,
-	since developers will always working from a stable base.
+	since developers will always be working from a stable base.
 	(Section~\ref{Sec_CompareTools})
@@
 We found (Section~\ref{Sec_AddressConcerns}) that developers are currently
-creating practices to address these pain points.  To address technical debt
+creating practices that address these pain points.  To address technical debt
 (\ppref{P_TechDebt}) some developers have committed to designing for change
@@
 Developers have prioritized documentation and usability
-(Section~\ref{Sec_RecommendDoc} to address the pain points of documentation and
+(Section~\ref{Sec_RecommendDoc}) to address the pain points of documentation and
 usability challenges (\ppref{P_Usability} and \ppref{P_Documentation}).  For the
@@
 performance benchmarks could be incorporated into the assessment. The accuracy
-of our estimate of participation level for CI and unit
-testing, which is not always visible in public artifacts, can be improved by
-adding interview questions. This additional data could be incorporated into the
-AHP ranking, ensuring quality designations that are more representative of the
-truth. In the future we could also expand the interview questions to go beyond
-reproducibility to also discuss replicability. Furthermore, future interview
-guidelines should clarify the differences between usability and
-understandability (Section~\ref{softwarequalities}), since their answers suggest
-that not all participants understood the distinction.  Analysis of future
-interview data would likely benefit from a more rigorous process, like a
-thematic domain analysis \citep{JungEtAl2022}.
-
-Our domain analysis (Section~\ref{SecDomainAnalysis} was relatively shallow.  In
+of our estimate of participation level for CI and unit testing, which is not
+always visible in public artifacts, can be improved by adding interview
+questions. This additional data could be incorporated into the AHP ranking,
+ensuring quality designations that are more representative of the truth. In the
+future we could also expand the interview questions to go beyond reproducibility
+to also discuss replicability. Furthermore, future interview guidelines should
+clarify the differences between usability and understandability
+(Section~\ref{softwarequalities}), since their answers suggest that not all
+participants understood the distinction.  Analysis of future interview data
+would likely benefit from a more rigorous process, like a thematic domain
+analysis \citep{JungEtAl2022}.
+
+Our domain analysis (Section~\ref{SecDomainAnalysis}) was relatively shallow.  In
 the future we would like to dig more deeply.  We would like to produce tables of
@@
 
-We would like to thank Ao Dong, Oluwaseun Owojaiye, Mohammad Ali Daeian and Reza
+We would like to thank Ao Dong, Oluwaseun Owojaiye, Mohammad Ali Daeian, and Reza
 Sadeghi for fruitful discussions on topics relevant to this paper.
@@
 TO do
-- check references
-- maybe add a continuous integration diagram?
+— check references
+— maybe add a continuous integration diagram?
 
@@
 the current state of LBM software development.  We determined how well best
-practices are followed and we investigated current developer pain points.  We
+practices are followed, and we investigated current developer pain points.  We
 compared the state of the practice for LBM software to the development practices

### 2022-04-23 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Fixing some of the incidents of passive voice

@@
 	statement is correct (adapted from recommended practice for requirements
-	specifications \citep{IEEE1998}). Similarly to correctness, verifiability is
-	correlated with the availability of a specification and with reference to
-	domain knowledge. A good measure of verifiability is further correlated with
+	specifications \citep{IEEE1998}). Similar to correctness, verifiability
+	increases with the presence of a specification and with references and
+	pointers to domain knowledge. A good measure of verifiability results from
 	the following: availability of well-written tutorials that include expected
@@
 	a specified time \citep[p.\ 357]{GhezziEtAl2003}, \citep{musa1987software}.
-	Reliability is thus positively correlated with the absence of errors during
-	installation and use. Recoverability from errors also improves reliability.
+	The absence of errors during installation and use increases the reliability
+	score. Recoverability from errors also improves reliability.
 
@@
 	circumstances not anticipated in the requirements specification; and ii)
-	when the assumptions in its requirements specification are violated
+	when the software inputs violate the requirements specification assumptions
 	\citep{boehm2007software}, \citep[p.\ 19]{GhezziEtAl2003}. A good measure of
@@
 	(hard real-time demands) \citep{IEEEStdGlossarySET1990, wiegers2003softreq}.
-	In this state of the practice assessment performance is not directly
-	quantitatively measured. Instead, we check the documentation for each
-	software package looking for information alluding to consideration of
-	performance, such as testing results or parallelization instructions.
+	Our state of the practice assessment does not directly quantitatively
+	measure performance. Instead, we check the documentation for each software
+	package looking for information alluding to consideration of performance,
+	such as testing results or parallelization instructions.
 
@@
 
-The assessment was conducted via the following steps:
+Our assessment process consists of the following steps:
 
@@
 
-To answer~\rqref{RQ_WhatProjects} we identify existing LBM projects.
-We find the candidate software through search engine queries targeting
-authoritative lists. LBM software is listed on \href{https://github.com/}
-{GitHub} and \href{https://swmath.org/} {swMATH} and found in scholarly
-articles. The Domain Expert (Section~\ref{sec_vet_software_list}) helps with
-selecting the candidate software.
+To answer~\rqref{RQ_WhatProjects} we identify existing LBM projects. We find the
+candidate software through search engine queries targeting authoritative lists.
+Scholarly articles, \href{https://github.com/} {GitHub} and
+\href{https://swmath.org/} {swMATH} provide candidate LBM software. The Domain
+Expert (Section~\ref{sec_vet_software_list}) helps with selecting the candidate
+software.
 
@@
 For the third item in the above filter, we categorized software packages as
-`alive' if documentation updates happened within the last 18 months. Packages
-were categorized as `dead' if their last update were more than 18 months ago.
+`alive' if documentation updates happened within the last 18 months, and `dead'
+otherwise.
 
@@
 with following best practices, including providing artifacts like those
-discussed in Section~\ref{Sec_CompareArtifacts} such as a contributor
+discussed in Section~\ref{Sec_CompareArtifacts}, such as a contributor
 guidelines, code style guidelines, a clear code of conduct and high quality code
@@
 nonfunctional requirements (to show the relative importance between qualities
-like portability, reliability and performance) and traceability information (to
+like portability, reliability, and performance) and traceability information (to
 show the consequences of changes to the assumptions). Spending time to
@@
 
-\subsection{Improve Re-runnability, Repeatability, Reproducibility and Replicability} 
+\subsection{Improve Re-runnability, Repeatability, Reproducibility, and Replicability} 
 \label{Sec_Reproducibility}
@@
 Domain Specific Modelling (DSM) approach. DSM means creating a knowledge base of
-models for physics, computing, mathematics, documentation and certification and
+models for physics, computing, mathematics, documentation, and certification and
 then writing explicit ``recipes'' that weave together this knowledge to generate
@@
 
-DSM provides a transformative technology for documentation, design and
+DSM provides a transformative technology for documentation, design, and
 verification \citep{JohansonAndHasselbring2018, Smith2018}. DSM allows
@@
 \item In our methodology the manual measures are done by one individual for all
-packages. The measures may be influenced by the graders abilities, experience
+packages. The measures may be influenced by the graders abilities, experience,
 and biases.  
@@
 	popularity;
-	\item Comparing artifacts, tools and processes to current research software
+	\item Comparing artifacts, tools, and processes to current research software
 	development guidelines; and, 
@@
 surveyed tools using version control.  The state of the practice for the
-software development process of LBM software matches the quasi-agile process
-generally adopted for research software
-(Section~\ref{Sec_CompareMethodologies}).  The best practice process of peer
-review is employed by some LBM developers.
+development of LBM software matches the quasi-agile process generally adopted
+for research software (Section~\ref{Sec_CompareMethodologies}).  The best
+practice process of peer review is employed by some LBM developers.
 
@@
 	(Section~\ref{Sec_AugTheoryMans})
-	\item Improve re-runnability, repeatability, reproducibility and
+	\item Improve re-runnability, repeatability, reproducibility, and
 	replicability to improve \ppref{P_Correctness}.

### 2022-04-23 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Addressing more passive voice problems

@@
 packages. We removed 22 packages for the following reasons: we cannot install
-them, they are incomplete, source code is not publicly available, a license is
-needed, or the project is out of scope. \citet{Michalski2021} lists the
-eliminated software packages. Of the remaining 23 packages, we kept some despite
-their `dead` status because of their presence on authoritative LBM software
-lists and because of their surface excellence. After the initial measurement
-exercise, which took place in mid-2020, we added one more package in January
-2022: Musubi. We found Musubi after our initial search, but we felt obliged to
-include it since it clearly fits our criteria. Time constraints meant we didn't
-update the previous manual data collection, but we did redo the automated data
-collection.
+them, they are incomplete, source code is not publicly available, we are not
+licensed to use them, or the project is out of scope. \citet{Michalski2021}
+lists the eliminated software packages. Of the remaining 23 packages, we kept
+some despite their `dead` status because of their presence on authoritative LBM
+software lists and because of their surface excellence. After the initial
+measurement exercise, which took place in mid-2020, we added one more package in
+January 2022: Musubi. We found Musubi after our initial search, but we felt
+obliged to include it since it clearly fits our criteria. Time constraints meant
+we didn't update the previous manual data collection, but we did redo the
+automated data collection.
 
@@
 development models are open source, where source code is freely available under
-an open-source license; free-ware, where a binary or executable is provided for
-free; and commercial, where the user must pay.
+an open-source license; free-ware, which freely provides a binary or executable,
+but not source code; and commercial, where the user must pay for the software.
 
@@
 qualities, the best that we can do is a shallow measure. For instance, we do not
-conduct experiments to measure usability directly. Instead, we look for an
-indication that usability is considered by searching for cues in the
-documentation, such as the presence of getting started instructions, a user
-manual or documentation of expected user characteristics.
+conduct experiments to measure usability directly. Instead, we look for
+usability considerations by searching for cues in the documentation, such as the
+presence of getting started instructions, a user manual or documentation of
+expected user characteristics.
 
@@
 is advantageous to study the common properties of the programs before analyzing
-individual members.'' Studying the common properties within a family of related
-programs is termed a domain analysis.
+individual members.'' Domain analysis is the term for studying the common
+properties within a family of related programs.
 
@@
 of variation, which summarize the possible values for that variability, along
-with their potential binding time. The binding time is when the value of a
-variability is set. Variabilities are set at specification time, build time
-(when the program compiles) or run time (when the code executes). For the
+with their potential binding time. The binding time is when we specify a value
+for a variability. Binding time options include specification time, build time
+(when the program compiles) and run time (when the code executes). For the
 current assessment, the packages all have the commonality of using LBM
 techniques to simulate the motion of fluids. Variabilities for LBM packages
-include the following: the dimension of problems they simulate, whether 
-(and how) they support parallel computation, whether compressibility can
-be considered, whether reflexive boundaries can be employed, whether the
-simulation handles multi-fluids, whether the simulation handles turbulence, and
-whether complex geometries can be simulated.
+include the following: the dimension of problems they simulate, whether (and
+how) they support parallel computation, whether compressibility can be
+considered, whether reflexive boundaries can be employed, whether the simulation
+handles multi-fluids, whether the simulation handles turbulence, and whether
+complex geometries can be simulated.
 
 To keep the current presentation focused, we only present the results of our
-shallow domain analysis. For the LBM domain, a table was constructed that
-distinguishes the programs by their variabilities. In research software, the
-variabilities are often assumptions. Table~\ref{tbl_features} lists the
+shallow domain analysis. We distinguish programs by their variabilities, which
+for research software are often assumptions. Table~\ref{tbl_features} lists the
 variabilities as columns, with the software alphabetically ordered.
@@
 system requirements, but only four (ESPResSo, LUMA, Mechsys, Sailfish) list
-compatible operating system versions. Ubuntu was used to test all but one of the
-packages. TCLB was tested on CentOS, as suggested by its installation
-instructions.
+compatible operating system versions. Ubuntu provided the testing platform for
+all but one of the packages. TCLB was tested on CentOS, as suggested by its
+installation instructions.
 
@@
 
-Errors encountered during the installation process were often quickly fixed,
-thanks to descriptive error messages. Systems that provided vague error
-messages, such as messages that did not specify which action or file was at
-fault, were harder to troubleshoot. Only three software packages (HemeLB, LB3D,
-lbmpy) that displayed a descriptive error message were not recoverable, and most
-of these instances were due to hardware and operating system incompatibility,
-such as the requirement of CUDA. Fourteen software packages broke during
-installation. Some packages, such as LB2D-Prime and LB3D-Prime did not provide a
-definitive message of the success or failure of the installation. In these
-instances, validating the installation required performing a tutorial or running
-a script, as described below, if these were available.
+We quickly fixed errors encountered during the installation process, thanks to
+descriptive error messages. Systems that provided vague error messages, such as
+messages that did not specify which action or file was at fault, were harder to
+troubleshoot. Only three software packages (HemeLB, LB3D, lbmpy) that displayed
+a descriptive error message were not recoverable, and most of these instances
+were due to hardware and operating system incompatibility, such as the
+requirement of CUDA. Fourteen software packages broke during installation. Some
+packages, such as LB2D-Prime and LB3D-Prime did not provide a definitive message
+of the success or failure of the installation. In these instances, validating
+the installation required performing a tutorial or running a script, as
+described below, if these were available.
 
@@
 Twenty of the software packages do not explicitly indicate the versions of
-software dependencies. Some software package installation issues, specifically
-those occurring when the manual installation of dependencies is required, might
-be avoided if versions of dependencies are specified. Sixteen software packages
-do not have detailed instructions for installing dependencies. Sixteen software
+software dependencies. Some software package installation issues could have been
+avoided by providing a list of dependencies. Sixteen software packages do not
+have detailed instructions for installing dependencies. Sixteen software
 packages have less than ten manual installation steps. If one command installs
@@
 specification, such as DL\_MESO (LBE), keep the information brief and include it
-within other documents. When present, theory related information is generally
-found within a user manual, on a webpage, or in a cited publication. In the
-latter case tracking down the knowledge may take a significant time.
-
-Document generation tools are explicitly used by 13 software packages. Sphinx is
-used by eight of them, and Doxygen is used by eight. Several of the packages use
-both.
+within other documents, such as a user manual, webpage, or a publication. In the
+latter case tracking down the knowledge may take significant time.
+
+Thirteen software packages explicitly use document generation tools. Eight of
+them use Sphinx and the same number use Doxygen. Several packages use both.
 
@@
 the software packages that are missing this key information. In these cases the
-user may need to assume correctness if there are no visible errors.  A
-particularly  detailed tutorial that includes expected output is provided by
-\href{https://www.walberla.net/doxygen/index.html} {waLBerla}.
+user may need to assume correctness if there are no visible errors.
+\href{https://www.walberla.net/doxygen/index.html} {waLBerla} provides a
+particularly detailed tutorial, including expected output.
 
@@
 
-The use of Continuous Integration (CI) tools and techniques alludes to a more
-refined development process where faults are isolated and better recognized.
-Only three of the packages (ESPResSo, Ludwig, Musubi) mentioned applying the
-practice of CI in their development process.
+The use of Continuous Integration (CI) tools and techniques alludes to a refined
+development process that quickly isolates and fixes faults. Only three of the
+packages (ESPResSo, Ludwig, Musubi) mentioned applying the practice of CI in
+their development process.
 
@@
 documentation. They also incorporated CI in the development process.
-Furthermore, eight of the top 10 ranked packages are noted as being alive.
+Furthermore, eight of the top 10 ranked packages are alive.
 
@@
 load dependencies. In these instances the installation tried to access outdated
-external repositories. Seven of the installations were recovered and verified,
-and one of the installations (LB3D-Prime) was assumed to be recovered due to the
+external repositories. We recovered and verified seven of the installations,
+and we assumed that one of the installations (LB3D-Prime) was recovered due to the
 absence of any way to verify otherwise. The installation of eight of the software
@@
 Of the 14 software packages that installed correctly and also have tutorials,
-four (pyLBM, \esp, LIMBES, Ludwig) broke during tutorial testing. All of
-these instances resulted in an error message being displayed. One error (pyLBM)
-was due to a missing tutorial dependency, another (Ludwig) was due to an invalid
-command despite following the tutorial, and the final two errors were vague
-execution errors. Of the four broken tutorial instances, only the one due to a
-missing dependency was recoverable.
+four (pyLBM, \esp, LIMBES, Ludwig) broke during tutorial testing. All of these
+instances resulted in the display of an error message. One error (pyLBM) was due
+to a missing tutorial dependency, another (Ludwig) was due to an invalid command
+despite following the tutorial, and the final two errors were vague execution
+errors. Of the four broken tutorial instances, only the one due to a missing
+dependency was recoverable.
 
@@
 testing, but a descriptive error message helped in recovery. Nine of the top 10
-ranked packages are noted as being alive.
+ranked packages are alive.
 
@@
 
-The software packages were tested for handling unexpected input, including
-incorrect data types, empty input, and missing files or links. Success is
-predicated on a reasonable response from the system, including appropriate error
-messages and an absence of unrecoverable system failures.
+Testing includes handling unexpected input, such as incorrect data types, empty
+input, and missing files or links. A reasonable response, including appropriate
+error messages and an absence of unrecoverable system failures, means success.
 
@@
 files, including a replacement of new lines with carriage returns. Furthermore,
-nine of the top 10 ranked packages are noted as being alive.
+nine of the top 10 ranked packages are alive.
 
@@
 Table~\ref{tbl_features}). Due to this, a comparison of run-time performance is
-not appropriate. Instead, we looked through each software package's artifacts for
-evidence that performance was considered. The artifacts of 18 software packages
+not appropriate. Instead, we looked through each software package's artifacts
+for evidence of considering performance. The artifacts of 18 software packages
 mentioned parallelization. This included GPU processing and the CUDA parallel
@@
 (ESPResSo, lbmpy, lettuce, pyLBM, Sailfish, TCLB). When a high degree of
-parallelization is possible GPUs provide superior speed compared to CPUs. The
+parallelization is possible, GPUs provide superior speed compared to CPUs. The
 software package TCLB is implemented in a highly efficient multi-GPU code to
 achieve performance suitable for model optimization \citep{rutkowski2020open}.
-In the Ludwig package, a so-called mixed mode approach is used where
-fine-grained parallelism is implemented on the GPU, and MPI is used for even
-larger scale parallelism \citep{gray2013ludwig}. While one software package
-(Sailfish) required CUDA and GPU processing, some (ESPResSo, lbmpy, lettuce,
-pyLBM, TCLB) have the option of using either the GPU or the CPU. In general, the
-packages that require GPU and CUDA have better performance at the expense of
+In the Ludwig package uses a so-called mixed mode approach implementing
+fine-grained parallelism on the GPU.  MPI is used for even larger scale
+parallelism \citep{gray2013ludwig}. While one software package (Sailfish)
+required CUDA and GPU processing, some (ESPResSo, lbmpy, lettuce, pyLBM, TCLB)
+have the option of using either the GPU or the CPU. In general, the packages
+that require GPU and CUDA have better performance at the expense of
 installability and surface reliability.
@@
 
-Software package artifacts were reviewed for the presence of a tutorial, a user
+We reviewed software package artifacts for the presence of a tutorial, a user
 manual, documented user characteristics, and a user support model. In total 19
@@
 rendered on a webpage. Some packages (waLBerla) do not have a user manual, but
-do have useful documentation distributed throughout their webpages. Expected
-user characteristics are documented in five software packages (laboetie, LIMBES,
-Ludwig, Musubi, Palabos). Users are typically scientists or engineers. Their
+do have useful documentation distributed throughout their webpages. Five
+software packages (laboetie, LIMBES, Ludwig, Musubi, Palabos) document expected
+user characteristics. Users are typically scientists or engineers. Their
 background is often physics, chemistry, biophysics, or mathematics. All but one
@@
 support model. Many packages have several user support models. Furthermore, four
-of the top five packages are noted as being alive.
+of the top five packages are alive.
 
@@
 
-Software packages were reviewed for the presence of artifacts. Every type of
-artifact or file that is not a code file was recorded. (The list of artifacts
-will be discussed further in Section~\ref{Sec_CompareArtifacts}.)  The software
-packages were also reviewed for software release and documentation version
-numbers. This information can be used to troubleshoot issues and organize
-documentation. All but three software packages (LatBo.jl, LB3D-Prime, MechSys)
-have source code release and documentation version numbers.
-
-When present, information on how code is reviewed, or how to contribute to the
-project was also noted. In total, 11 software packages have this information,
-which was found in various artifacts, including in developer guides, contributor
-guides, user guides, developer webpages, and README files.
-
-Issue tracking is used in 23 software packages, 15 of which use GitHub or GitLab
-to host their git repository, seven use email, and one (SunlightLB) uses
+We reviewed software packages for the presence of artifacts and recorded every
+type of artifact that is not a code file. (The list of artifacts will be
+discussed further in Section~\ref{Sec_CompareArtifacts}.)  We also reviewed the
+software packages for software release and documentation version numbers. This
+information can be used to troubleshoot issues and organize documentation. All
+but three software packages (LatBo.jl, LB3D-Prime, MechSys) have source code
+release and documentation version numbers.
+
+When present, we noted how the team reviews code, and how users can contribute
+to the project. In total, 11 software packages have this information, which was
+found in various artifacts, including in developer guides, contributor guides,
+user guides, developer webpages, and README files.
+
+Twenty-three packages use issue tracking, 15 of which use GitHub or GitLab to
+host their git repository, seven use email, and one (SunlightLB) uses
 SourceForge. Most software packages that use GitHub or GitLab have the majority
 of their issues closed, and only three (laboetie, lettuce, Sailfish) have less
-than 50 percent of their issues closed. All the top five overall ranked
-packages have most of their issues closed. Alive packages (11 use issue
-tracking) have 64\% of their issues closed, while dead packages (3 use issue
-tracking) have 71\% of their issues closed. This information is presented in
-Table~\ref{gitrepodata}. With respect to the host for the git repository, 13
-packages use GitHub and two (Palabos, waLBerla) use GitLab. Of the other
-packages, one package (SunlightLB) uses CVS for issue tracking and version
-control, and seven packages do not appear to use any issue tracking system.
+than 50 percent of their issues closed. All the top five overall ranked packages
+have most of their issues closed. Alive packages (11 use issue tracking) have
+64\% of their issues closed, while dead packages (3 use issue tracking) have
+71\% of their issues closed. Table~\ref{gitrepodata} presents this information.
+With respect to the host for the git repository, 13 packages use GitHub and two
+(Palabos, waLBerla) use GitLab. Of the other packages, one package (SunlightLB)
+uses CVS for issue tracking and version control, and seven packages do not
+appear to use any issue tracking system.
 
@@
 
-Software package code files were further measured for the percentage of code
-that is comments. The findings are presented in Table~\ref{gitrepodata}, sorted
-by the percentage. Packages with a higher percentage of comments were assumed to
-be more maintainable. Comments represent more than 10 percent of code files in
-16 packages, and the average percentage of code comments is about 14 percent.
-All the top five overall ranked packages have more than the average. LUMA has
-only 0.2 percent comments, the fewest of any package. This package has the most
-lines of source code, with over four million. The next largest package is
-\esp{} with one million.
+Table~\ref{gitrepodata} shows the percentage of code that is comments, in
+decreasing order. Maintainability improves for packages with a
+higher percentage of comments. Comments represent more than 10 percent of code
+files in 16 packages, and the average percentage of code comments is about 14
+percent. All the top five overall ranked packages have more than the average.
+LUMA has only 0.2 percent comments, the fewest of any package. This package has
+the most lines of source code, with over four million. The next largest package
+is \esp{} with one million.
 
@@
 packages also appear to manage their issue tracking, having most of their issues
-closed. Their code files are well commented with more than 10 percent of the
-code being comments. Four of the top five ranked packages are noted as being alive.
+closed. Maintainable packages by our ranking have more than 10 percent of the
+code as comments. Four of the top five ranked packages are alive.
 
@@
 not on a level field. However, these packages do have an abundance of reusable
-components. Nine of the top 10 ranked packages are noted as being alive.
+components. Nine of the top 10 ranked packages are alive.
 
@@
 source code. The source code is well modularized and structured. All the top
-five ranked packages are noted as being alive.
+five ranked packages are alive.
 
@@
 detailed and easy to access notes accompanying software releases. Four of the
-top five ranked packages are noted as being alive.
+top five ranked packages are alive.
 

### 2022-04-23 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
reviewed Section 5

@@
 concerns, including maintainability, reusability, and understandability.  With
-the quality considerations in mind, and the potentially overwhelming number of
-choices for existing LBM libraries, the time is right to assess the state of the
-practice. Therefore, we analyze the current state of LBM
+these quality considerations in mind, and the potentially overwhelming number of choices for existing LBM libraries, the time is right to assess the state of the practice.
+Therefore, we analyze the current state of LBM
 software development and compare it to the development practices employed for
@@
 achieve the highest possible quality.  However, rather than delve into the
-nuances of which software can justify compromising which practices we will write
-our comparison under the ideal assumption that every project has sufficient
-resources to match best practices.
+nuances of which software can justify compromising which practices, we will write our comparison under the ideal assumption that every project has sufficient resources to match best practices.
 
@@
 
-In Table~\ref{Tbl_Guidelines} each row corresponds to an artifact.  For a given
+In Table~\ref{Tbl_Guidelines}, each row corresponds to an artifact.  For a given
 row, a checkmark in one of the columns means that the corresponding guideline
@@
 between projects.  Although some content is reasonably consistent, with 97\% of
-README files contain at least one section describing the `What` of the
+README files containing at least one section describing the `What` of the
 repository and 88.5 \% offering some `How` content, other categories are more
-variable.  For instance, information on `Contribution`, `Why`', and `Who`,
+variable.  For instance, information on `Contribution`, `Why`, and `Who`,
 appear in 27.8\%, 25.7\% and 52.9\% of the analyzed files, respectively
@@
 commonly recommended artifacts are the most important artifacts. Just because an
-artifact is not explicitly recommended in a given guideline, does not mean that
+artifact is not explicitly recommended in a given guideline does not mean that
 the artifact is not valued.  The guideline authors may have excluded it because
-it is out of the scope of their recommendations, or outside their experience.
+it is out of the scope of their recommendations or outside their experience.
 For instance, an artifact related to uninstall is only explicitly mentioned by
@@
 
-The top four packages, ESPREsSo, Ludwig, Palabos, and OpenLB, have all the
+The top four packages (ESPREsSo, Ludwig, Palabos, and OpenLB) have all the
 common artifacts, except Palabos does not have theory notes and only three
@@
 publications and performance information certainly appear in other research
-software projects; the practices just aren't highlighted in guidelines because
+software projects; the practices just are not highlighted in guidelines because
 they are implicit in other documentation recommendations, such as theory
@@
 
-As Table~\ref{Tbl_Guidelines} shows the LBM community participates in most of
+As Table~\ref{Tbl_Guidelines} shows, the LBM community participates in most of
 the practices we found listed in the general research software guidelines;
@@
 address the following: \begin{inparaenum}[i)]
-	\item Where are we now?
-	\item Where are we going? and
+	\item Where are we now?,
+	\item Where are we going?, and
 	\item How can we get there? \end{inparaenum} \citep{PhaalEtAl2005}.
@@
 	\item scan the environment, 
-	\item revise and distill the product vision to write the product roadmap, 
+	\item revise and distill the product vision to write the product roadmap, and 
 	\item estimate the product life cycle and evaluate the mix of planned
@@
 \href{https://www.contributor-covenant.org/version/2/1/code_of_conduct/}
-{Contributor Covenant}, \href{https://www.djangoproject.com/conduct/} {Django
+{Contributor Covenant}, and \href{https://www.djangoproject.com/conduct/} {Django
 Code of Conduct}. \citet{SinghEtAl2021} shows the value of adopting a code of
@@
 observe any codes of conduct, possibly because the number of developers for each
-LBM project is small and because the idea hasn't yet occurred to any of the
+LBM project is small and because the idea has not yet occurred to any of the
 participants.

### 2022-04-23 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
reviewed Section 6

@@
 
-For the dependency tools (Table~\ref{tbl_tools}), we observed that most of the
+For the dependency tools (Table~\ref{tbl_tools}), we observe that most of the
 software packages use some sort of build automation tools, most commonly Make.
@@
 	its configuration, the environment, or the software stack that it runs on
-	\citep[p.\ 83]{HumbleAndFarley2010}. As Table~\ref{artifactspresent} shows
+	\citep[p.\ 83]{HumbleAndFarley2010}. As Table~\ref{artifactspresent} shows,
 	test cases are in the uncommon category for LBM software artifacts, which
@@
 	same developer, is postponed, integration problems are inevitable.  By
-	continuously integrating problems are immediately obvious, and the source of
+	continuously integrating, problems are immediately obvious and the source of
 	the problem can be isolated to the small increment that was just committed.

### 2022-04-23 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
reviewed Section 7

@@
 Scientific developers naturally use an agile philosophy \citep{AckroydEtAl2008,
-CarverEtAl2007, EasterbrookAndJohns2009, Segal2005, HeatonAndCarver2015}, or an
+CarverEtAl2007, EasterbrookAndJohns2009, Segal2005, HeatonAndCarver2015}, an
 amethododical process \citep{Kelly2013}, or a knowledge acquisition driven

### 2022-04-23 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
reviewed Section 8

@@
 	highlighted as a pain point by other research software developers
-	\citet{PintoEtAl2018, PintoEtAl2016, WieseEtAl2019}.
+	\citep{PintoEtAl2018, PintoEtAl2016, WieseEtAl2019}.
 
@@
 	project are domain experts, not computer scientists or software engineers.
-	This same trend is noted by \citep{Nguyen-HoanEtAl2010}, which showed only
+	This same trend is noted by \citet{Nguyen-HoanEtAl2010}, which showed only
 	23\% of research software survey respondents having a computing-related
-	background. Similarly, \citep{UditAndKatz2017} show that the majority (54\%)
+	background. Similarly, \citet{UditAndKatz2017} show that the majority (54\%)
 	of postdocs have not received training in software development.  The LBM
@@
 	formal training (From 13\% of respondents in 2009 to 22\% in 2018).
-	\citep{PintoEtAl2018} found that some developers feel there is a mismatch
+	\citet{PintoEtAl2018} found that some developers feel there is a mismatch
 	between coding skills and subject-matter skills.
@@
 	software, but the trend likely applies to other research software domains)
-	is infrequently cited.  \citep{PintoEtAl2018} also mentions the lack of
+	is infrequently cited.  \citet{PintoEtAl2018} also mentions the lack of
 	formal reward system for research software.
@@
 	include setting up parallelization and CI. The pain
-	point survey of \citep{WieseEtAl2019} also highlighted technical-related
+	point survey of \citet{WieseEtAl2019} also highlighted technical-related
 	problems like dependency management, cross-platform compatibility,
@@
 	shows that more developers think testing is important than the number that
-	believe they have a sufficient understanding of testing concepts; and, iii)
+	believe they have a sufficient understanding of testing concepts; and, iii) previous studies
 	\citep{HannayEtAl2009, KanewalaAndBieman2013, KellyEtAl2011, WieseEtAl2019}
@@
 	packages for the new endeavour.  Some respondents in the survey of
-	\citep{WieseEtAl2019} also mentioned that users do not always have the
+	\citet{WieseEtAl2019} also mentioned that users do not always have the
 	expertise required to install or use the software.
@@
 	components in the source code. This results in technical debt
-	\citet{KruchtenEtAl2012}, which has an impact on future maintainability and
+	\citep{KruchtenEtAl2012}, which has an impact on future maintainability and
 	reusability. Concern with technical debt is likely why researchers in the
-	survey of \citep{Nguyen-HoanEtAl2010} rated maintainability as the third
+	survey of \citet{Nguyen-HoanEtAl2010} rated maintainability as the third
 	most important software quality. More recently the push for sustainable
-	software \citet{deSouzaEtAl2019} is motivated by the pain that past
+	software \citep{deSouzaEtAl2019} is motivated by the pain that past
 	developers have had with accumulating too much technical debt.

### 2022-04-23 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
reviewed Section 10

@@
 are not lists of criticisms for what should have been done in the past, or what
-should be done now; they are just suggestions for consideration in the future.
+should be done now; they are suggestions for consideration in the future.
 We will not be repeating ideas here that have been discussed in previous
@@
 Except for \citet{ThielEtAl2020}, the research software guidelines that we
-consulted do not mention linters.  However, we feel linters have the potential
+consulted do not mention linters.  However, we believe that linters have the potential
 to improve code quality at a relatively low cost.  A linter is a tool that
@@
 and publishes articles that credit the scholarship contained in the software
-itself \citep{SmithEtAl2018-joss}.  \citep{SmithEtAl2016-softcite} presents a
+itself \citep{SmithEtAl2018-joss}.  \citet{SmithEtAl2016-softcite} presents a
 set of software citation principles that may encourage broad adoption of a
 consistent policy for software citation across disciplines and venues.
-\citep{ChueHongEtAl2019} provides a software citation checklist for developers
+\citet{ChueHongEtAl2019} provides a software citation checklist for developers
 to make a release of their software citable.  Further guidance on software

### 2022-04-23 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
reviewed Section 11

@@
 secondary studies by \citet{AmpatzoglouEtAl2019}.  We use the classification of
-\citep{AmpatzoglouEtAl2019} because studying a collection of software is a form
+\citet{AmpatzoglouEtAl2019} because studying a collection of software is a form
 of secondary study, where a secondary study analyzes the data from a set of
@@
 \begin{itemize}
-\item In our methodology the manual measures are done by one individual for all
-packages. The measures may be influenced by the graders abilities, experience,
+\item In our methodology, the manual measures are done by one individual for all
+packages. The measures may be influenced by the grader's abilities, experience,
 and biases.  
@@
 community.
-\item In our manual measures of the software repositories some data may not have
+\item In our manual measures of the software repositories, some data may not have
 been measured due to technology issues like broken links. This issue arose with
@@
 versions we detected.  When CI and unit testing are done outside the public
-repository, they are not as effective, since they are not available for others
+repository, they are not as effective since they are not available for others
 to inspect and reproduce.

### 2022-04-23 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
reviewed Section 12

@@
 state of the practice for developing LBM software. Understanding what artifacts
-are produced, what tools are used, what process are employed and what pain
-points are experienced, is the starting point for devising future methods and
+are produced, what tools are used, what processes are employed and what pain
+points are experienced is the starting point for devising future methods and
 tools for reducing development time and improving software quality.
@@
 	consists of 108 questions to assess 9 qualities (including the qualities of
-	installability, usability and visibility))
+	installability, usability and visibility);
 	\item Ranked the software using the Analytic Hierarchy Process (AHP);
@@
 commonalities, variabilities, and parameters of variation for the family of LBM
-solvers. We could follow the template from \citep{smith2008commonality}, or
-adopt the analysis techniques from \citep{weiss1998commonality}.
+solvers. We could follow the template from \citet{smith2008commonality}, or
+adopt the analysis techniques from \citet{weiss1998commonality}.
 \citet{SmithAndChen2004} and \citet{SmithMcCutchanAndCarette2017} provide

### 2022-04-23 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Further reduction of passive voice

@@
 We measured the total number of source code files for each project. We assume
-that numerous source files is associated with increased reusability,
-due to our assumption that this indicates increased modularization. Some
-packages have more features than others.  This is assumed to contribute to
-reusability, since they have more source code for potential reuse. The software
-packages were also reviewed for the presence of API documentation, which
-indicates that a software package was developed with interaction between other
-software applications in mind.
+that numerous source files increases reusability, since more files implies
+increased modularization. Some packages have more features than others.  This is
+assumed to contribute to reusability, since they have more source code for
+potential reuse. We reviewed the software packages for the presence of API
+documentation, which suggests that the developers considered potential future
+reuse.
 
@@
 packages. Packages with a high reusability score do not have as many LOC per
-text file, generally having a few hundred lines or fewer. This suggests that the
-source code of these packages is likely functionally modularized, and modules
-could be reused in other projects.
+text file, generally having a few hundred lines or fewer. We assume that
+relatively small files implies modularization.
 
 The package waLBerla scored high on reusability because of its focus on
-modularity.  The modularity in the waLBerla framework was included to enhance
-productivity, reusability, and maintainability \citep{bauer2021walberla}. The
-design of waLBerla has enabled it to be successfully applied in several projects
-as a basis for various extensions \citep{bauer2021walberla}.
+modularity.  The modularity in the waLBerla framework enhances productivity,
+reusability, and maintainability \citep{bauer2021walberla}. The design of
+waLBerla has enabled it to be successfully applied in several projects as a
+basis for various extensions \citep{bauer2021walberla}.
 
@@
 
-Ten random source code files of each software package were reviewed for several
-measures. This assessment of surface understandability may not perfectly reflect
-each package, due to the practical limitation of only examining 10 files.
+To measure understandability, we reviewed ten random source code files per
+package.  By using only 10 files we may incorrectly estimate understandability,
+but for practical consideration our time measuring each package has to be
+limited.
 
@@
 LB3D-Prime, LIMBES, MP-LABS) appear to use vague identifiers, such as single
-letters for variables. Symbolic constants were observed in the source code of 13
-packages. The constants are used for various parameters, mathematical constants,
-and matrix definitions. All the packages are reasonably well commented, with
-the comments clearly indicating what is being done (as opposed to how it is
-being done). Domain algorithms are noted in the source code of 11 packages.
+letters for variables. Thirteen source code files used symbolic constants. The
+constants are used for various parameters, mathematical constants, and matrix
+definitions. All the packages are reasonably well commented, with the comments
+clearly indicating what is being done (as opposed to how it is being done).
+Domain algorithms are noted in the source code of 11 packages.
 Table~\ref{moduledata} suggests that the software packages are modularized to

### 2022-04-24 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Revised abstract - fixed verb tense

@@
     extracted from the project repository) and developer interviews (consisting
-    of 20 questions). After measuring, we ranked the software using
-    the Analytic Hierarchy Process (AHP). Our ranking was roughly consistent
-    with GitHub stars ranking, suggesting at least a partial correlation
-    between the use of best practices and popularity. We find the state of the
-    practice to be healthy, with 67\% of the measured packages ranking in the
-    top five for at least one quality, the majority of LBM generated artifacts
-    corresponding to general recommendations from research software developers,
-    common use of version control (67\% of packages) and the adoption of a
-    quasi-agile development process.  Areas of best practice to potentially
-    improve include adoption of continuous integration, API documentation and
-    enforcement of programming style guides.  We interviewed four developers to
-    gain insight into their current pain points. Identified challenges include
-    lack of development time, lack of funding, and difficulty with ensuring
-    correctness.  Developers are addressing these pain points by designing for
-    change, circumventing the oracle problem and prioritizing documentation and
-    usability. For future improvements we suggest the following: employing
-    linters, conducting rigorous peer reviews, writing and submitting more
-    papers on software, growing the number of contributors by following current
-    recommendations for open source projects, and augmenting the theory manuals
-    to include more requirements relevant information.
+    of 20 questions). After measuring, we rank the software using the Analytic
+    Hierarchy Process (AHP). Our ranking is roughly consistent with GitHub stars
+    ranking, suggesting at least a partial correlation between the use of best
+    practices and popularity. We find the state of the practice to be healthy,
+    with 67\% of the measured packages ranking in the top five for at least one
+    quality, the majority of LBM generated artifacts corresponding to general
+    recommendations from research software developers, common use of version
+    control (67\% of packages) and the adoption of a quasi-agile development
+    process.  Areas of best practice to potentially improve include adoption of
+    continuous integration, API documentation and enforcement of programming
+    style guides.  We interviewed four developers to gain insight into their
+    current pain points. Identified challenges include lack of development time,
+    lack of funding, and difficulty with ensuring correctness.  Developers are
+    addressing these pain points by designing for change, circumventing the
+    oracle problem and prioritizing documentation and usability. For future
+    improvements we suggest the following: employing linters, conducting
+    rigorous peer reviews, writing and submitting more papers on software,
+    growing the number of contributors by following current recommendations for
+    open source projects, and augmenting the theory manuals to include more
+    requirements specification relevant information.
 
@@
 LB3D-Prime, LIMBES, MP-LABS) appear to use vague identifiers, such as single
-letters for variables. Thirteen source code files used symbolic constants. The
-constants are used for various parameters, mathematical constants, and matrix
-definitions. All the packages are reasonably well commented, with the comments
-clearly indicating what is being done (as opposed to how it is being done).
-Domain algorithms are noted in the source code of 11 packages.
-Table~\ref{moduledata} suggests that the software packages are modularized to
-various degrees. When observing the source code files, it was found that 14 of
-the packages have a consistent style and order of function parameters.
+letters for variables. Thirteen source code files used symbolic constants for
+various parameters, mathematical constants, and matrix definitions. The code
+includes appropriate comments for all packages, with the comments clearly
+indicating what is being done (as opposed to how it is being done). The source
+code of 11 packages includes notes on domain algorithms. Table~\ref{moduledata}
+suggests that the software packages are modular to various degrees. When
+observing the source code files, 14 of the packages have a consistent style and
+order of function parameters.
 
@@
 
-Software package artifacts were reviewed for the identification of a specific
+We reviewed the software package artifacts for the identification of a specific
 development model, like a waterfall or agile development model, and the presence
@@
 transparency, the small close-knit nature of the development teams make
-explicitly specifying this information practically unnecessary. Version
-release notes were found in nine of the software packages.
+explicitly specifying this information practically unnecessary. Nine of the
+software packages included version release notes.
 
@@
 Figure~\ref{Fig_OverallScore} shows the overall ranking of the software
-packages. In the absence of other information on priorities, the overall ranking
-was calculated by assuming an equal weighting between all qualities. If the
-qualities were weighed differently, the overall software package ranking would
+packages. In the absence of other information on priorities, we calculated the
+overall ranking by assuming an equal weighting between all qualities. If we
+weigh the qualities differently, the overall software package ranking will
 change. Software packages with an overall high score ranked high in at least
@@
 most stars, while our ninth ranked package (Sailfish) has the highest number of
-stars. The same correlation is observed in the repository watch column, although
+stars. The repository watch column shows the same correlation, although
 this column contains less data, since two of the packages (Palabos, waLBerla)
@@
 count, but relatively low in the ``follow best practices'' ranking: TCLB (star
-rank 3, our rank 12), lettuce (star rank 4, our rank 14) and \esp{} (star
-rank 5, our rank 15).  Part of the discrepancy is because all projects are
-ranked by our process, but not all projects are measured by stars.  Also, our
+rank 3, our rank 12), lettuce (star rank 4, our rank 14) and \esp{} (star rank
+5, our rank 15).  Part of the discrepancy is because our process ranks all
+projects, but the star measure does not apply to all projects.  Also, our
 process equally weights all qualities (Section~\ref{Sec_OverallQuality}), but
@@
 the community's feeling in the past more than current preferences
-\citep{Szulik2017}. Older packages (like \esp{} and TCLB) have had more time
-to accumulate stars and watches. Young projects have less time to accumulate
-stars, like HemeLB, which only has 12 stars (at the time of measurement) having
+\citep{Szulik2017}. Older packages (like \esp{} and TCLB) have had more time to
+accumulate stars and watches. Young projects have less time to accumulate stars,
+like HemeLB, which only has 12 stars (at the time of measurement) having
 recently moved from a different GitHub repository.  Lettuce is a young project
 (released in 2019) with a relatively high number of stars. The discrepancy in
-rankings here may be due to a mismatch between the time when lettuce was
-manually measured by our template and when the stars were counted. The late
-addition of Musubi meant a large gap between when projects were manually
-measured and the time when our automated data (like stars) was collected. A
-final reason for inconsistencies between our ranking and the community's ranking
-is that, as for consumer products, more factors influence popularity than just
-quality.
+rankings may be due to a mismatch between the time we manually measured lettuce
+and when we counted the stars. The late addition of Musubi meant a large gap
+between when we measured projects and the time when our automated data (like
+stars) was collected. A final reason for inconsistencies between our ranking and
+the community's ranking is that, as for consumer products, more factors
+influence popularity than just quality.
 
@@
 broader research software community.  We will also investigate the practices
-from the top projects that others within the LBM community, and within the
-broader research software community, can potentially adopt.
+from the top projects that others within the LBM community and beyond can
+potentially adopt.
 
@@
 
-As part of filling in the measurement template the LBM packages were examined
-for the presence of artifacts, which were then categorized by frequency. We have
-grouped the artifacts into common, uncommon, and rare categories in
-Table~\ref{artifactspresent}. Common artifacts were found in 16 to 24 ($>$63\%)
-of the software packages. Uncommon artifacts were found in 8 to 15 (30-63\%) and
-rare artifacts were found in 1 to 7 ($<$30\%) of the packages.  We answer
-\rqref{RQ_CompareArtifacts} by comparing the artifacts that we observed in LBM
-repositories to those observed and recommended for research software in general.
-Our comparison may point out areas where some LBM software packages fall short
-of current best practices. This is not intended to be a criticism of any
-existing packages, especially since in practice not every project needs to
-achieve the highest possible quality.  However, rather than delve into the
-nuances of which software can justify compromising which practices, we will write our comparison under the ideal assumption that every project has sufficient resources to match best practices.
+As part of filling in the measurement template, we listed the artifacts for each
+package, and we then categorized them by frequency. We group the artifacts into
+common, uncommon, and rare categories in Table~\ref{artifactspresent}. Common
+artifacts are in 16 to 24 ($>$63\%) of the software packages. Uncommon artifacts
+are in 8 to 15 (30-63\%) and rare artifacts are in 1 to 7 ($<$30\%) of the
+packages.  We answer \rqref{RQ_CompareArtifacts} by comparing the artifacts that
+we observed in LBM repositories to those observed and recommended for research
+software in general. Our comparison may point out areas where some LBM software
+packages fall short of current best practices. This is not intended to be a
+criticism of any existing packages, especially since in practice not every
+project needs to achieve the highest possible quality. However, rather than
+delve into the nuances of which software can justify compromising which
+practices, we will write our comparison under the ideal assumption that every
+project has sufficient resources to match best practices.
 
@@
 popularity of recommending a given artifact, but it does not imply that the most
-commonly recommended artifacts are the most important artifacts. Just because an
-artifact is not explicitly recommended in a given guideline does not mean that
-the artifact is not valued.  The guideline authors may have excluded it because
-it is out of the scope of their recommendations or outside their experience.
-For instance, an artifact related to uninstall is only explicitly mentioned by
+commonly recommended artifacts are the most important artifacts. Just because a
+guideline does not explicitly recommend an artifact does not mean that the
+artifact has no value.  The guideline authors may have excluded it because it is
+out of the scope of their recommendations or outside their experience. For
+instance, an artifact related to uninstall is only explicitly mentioned by
 \citet{vanGompelEtAl2016}, but other guideline authors would likely see its
 value.  They may simply feel that uninstall is implied by install, or they may
-have never asked themselves whether separate uninstall instructions are needed.
+have never asked themselves about the need for separate uninstall instructions.
 
@@
 (ESPREsSo, Ludwig, Palabos) of the packages appear to use a version control
-system.  The top four packages also have most of the uncommon artifacts. At
-the time of data collection, only one of the four packages (Palabos) did not
-have a user manual or guide, but there was a broken link on the package website
-indicating that such an artifact might exist. This broken link was later fixed,
-but this is not reflected in our data because it was not present at the time of
-data collection (mid-2020). Despite the broken link, Palabos does have a
-detailed and informative website. As far as we could observe, Ludwig does not
-appear to have a publicly visible design and OpenLB does not appear to use a
+system.  The top four packages also have most of the uncommon artifacts. At the
+time of data collection, only one of the four packages (Palabos) did not have a
+user manual or guide, but there was a broken link on the package website
+indicating that such an artifact might exist. The developers later fixed this
+broken link, but this is not reflected in our data because it was not present at
+the time of data collection (mid-2020). Despite the broken link, Palabos does
+have a detailed and informative website. As far as we could observe, Ludwig does
+not appear to have a publicly visible design and OpenLB does not appear to use a
 version control system. However, OpenLB's use of version numbers suggests that
-version control may be used privately by the developers.
+its developers may privately use version control.
 
@@
 the practices we found listed in the general research software guidelines;
-however, some recommended practices were not observed or rarely observed. For
-instance, API documentation was rarely observed for LBM software, but it is a
+however, we did not observe, or rarely observed, some recommended practices. For
+instance, we rarely saw API documentation for LBM software, but it is a
 frequently recommended artifact \citep{SmithAndRoscoe2018, ThielEtAl2020,
@@
 
-\item \textbf{Checklists} can be used in projects to ensure that best practices
-are followed by all developers.  Some examples include checklists merging
-branches into master \citep{Brown2015}, checklists for saving and sharing
-changes to the project \citep{WilsonEtAl2016}, checklists for new and departing
-team members \citep{HerouxAndBernholdt2018}, checklists for processes related to
-commits and releases \citep{HerouxEtAl2008} and checklists for overall software
-quality \citep{ThielEtAl2020, SSI2022}.  For LBM software, ESPResSo has a
-checklist for managing releases, but we did not find any other examples.
+\item \textbf{Checklists} can be used to ensure that developers follow best
+practices.  Some examples include checklists merging branches into master
+\citep{Brown2015}, checklists for saving and sharing changes to the project
+\citep{WilsonEtAl2016}, checklists for new and departing team members
+\citep{HerouxAndBernholdt2018}, checklists for processes related to commits and
+releases \citep{HerouxEtAl2008} and checklists for overall software quality
+\citep{ThielEtAl2020, SSI2022}.  For LBM software, ESPResSo has a checklist for
+managing releases, but we did not find any other examples.
 
@@
 by the software application (or library) through such resources as its methods
-or objects \citep{MengEtAl2018}.  API documentation has been found to be a
-critical factor for API understandability (Section~\ref{softwarequalities})
-\citep{MengEtAl2018}. Some API documentation can be generated using tools like
-Doxygen, but when learning a new API developers mostly use code examples and
-tutorials \citep{MengEtAl2018}.  API documentation was rare within our sample of
-LBM packages.
+or objects \citep{MengEtAl2018}.  API documentation is a critical factor for API
+understandability \citep{MengEtAl2018} (Section~\ref{softwarequalities}). Some
+API documentation can be generated using tools like Doxygen, but when learning a
+new API developers mostly use code examples and tutorials \citep{MengEtAl2018}.
+API documentation was rare within our sample of LBM packages.
 
@@
 Software requirements documentation provides a good example of the common
-deviation between recommendations and practice.  Although requirements
-documentation is recommended by some \citep{TobiasEtAl2018, HerouxEtAl2008,
+deviation between recommendations and practice.  Although some guidelines
+recommend requirements documentation \citep{TobiasEtAl2018, HerouxEtAl2008,
 SmithAndKoothoor2016}, in practice research software developers often do not
@@
 perception that up-front requirements are impossible for research software
-\citep{CarverEtAl2007,SegalAndMorris2008}, but when the instance on ``up-front''
-requirements is dropped, allowing the requirements to be written iteratively and
-incrementally, requirements are feasible \citep{Smith2016}.
+\citep{CarverEtAl2007,SegalAndMorris2008}, but when we drop the insistence on
+``up-front'' requirements, allowing the requirements to be written
+iteratively and incrementally, requirements are feasible \citep{Smith2016}.
 
@@
 
-Software tools are used to support the development, verification, maintenance,
-and evolution of software, software processes, and artifacts \citep[p.\
-501]{GhezziEtAl2003}. Many tools are used by LBM software packages, as
-summarized in Table~\ref{tbl_tools}.  The tools are subdivided into development
-tools, dependencies, and project management tools.  In this section we answer
-\rqref{RQ_CompareToolsProjMngmnt} by comparing aspects of tool usage in LBM
-software packages to their use in the research software community in
-general.
+Developers use software tools to support the development, verification,
+maintenance, and evolution of software, software processes, and artifacts
+\citep[p.\ 501]{GhezziEtAl2003}. Table~\ref{tbl_tools} summarizes the tools,
+subdivided into development tools, dependencies, and project management tools.
+In this section we answer \rqref{RQ_CompareToolsProjMngmnt} by comparing aspects
+of tool usage in LBM software packages to their use in the research software
+community in general.
 
@@
 released \citep[p.\ 506]{GhezziEtAl2003}. Although not shown in
-Table~\ref{tbl_tools}, debuggers were also likely used.  Only three (ESPResSo,
-Ludwig and Musubi) packages mentioned CI tools, like Travis
-CI. Code editors and compilers were explicitly noted to have been used by
-several packages, and were likely used by all of them. One of the packages
-(Ludwig) explicitly noted the use of proprietary unit testing code written in C.
-Likewise, the use of proprietary code for verifying the correctness of output
-was noted by one of the developers (pyLBM). Similar tools were likely used when
-developing the other software packages.
+Table~\ref{tbl_tools}, developers also likely use debuggers.  Only three
+(ESPResSo, Ludwig and Musubi) packages mentioned CI tools, like Travis CI.
+Several packages explicitly noted code editors and compilers, but likely all
+packages use them. One of the packages (Ludwig) explicitly noted the use of
+proprietary unit testing code written in C. Likewise, one of the developers
+(pyLBM) noted the use of proprietary code for verifying the correctness of
+output. Developers of other packages likely used similar tools.
 
@@
 explicitly noted in the artifacts, webpages, or interviews with the developers.
-As with development tools and dependencies, it is possible that other types of
-project management tools are used, but they were not visible in the artifacts to
-which we had access.  Collaboration tools, most often email and video, are noted
-as being used when developing software projects. Project management software was
-not explicitly mentioned during our interviews, but it is possible that some
-projects use such software. Many of the projects are located on GitHub, where
-the developers use the platform to help manage their projects, especially bug
+As with development tools and dependencies, developers may use other types of
+project management tools, but they were not visible in the artifacts to which we
+had access.  Collaboration tools, most often email and video, are noted as being
+used when developing software projects. Our interviewees did not mention project
+management software, but the sample is not large enough to rule out other
+projects using such software. GitHub is the home for many of the projects.
+Developers use the platform to help manage their projects, especially bug
 related issues. Most of the projects appear to use change tracking and version
-control tools. Document generation tools are mentioned in 12 of the 24 projects.
-The tools Sphinx and Doxygen are explicitly used in this capacity.
+control tools. Half of the packages mention document generation tools, like
+Sphinx and Doxygen.
 
@@
 packages use version control.)  The proliferation of version control tools for
-LBM matches the increase in the broader research software community.  A little
-over 10 years ago version control was estimated to be used in only 50\% of
-research software projects \citep{Nguyen-HoanEtAl2010}, but even at that time
+LBM matches the increase in the broader research software community.
+\citet{Nguyen-HoanEtAl2010} estimated a little over 10 years ago that only 50\%
+of research software projects use version control, but even at that time
 \citet{Nguyen-HoanEtAl2010} noted an increase from previous usage levels. A
@@
 
-As mentioned in Section~\ref{SecSurfCorrectAndVerifiab}, CI is rarely used in
-LBM software (3 of 24 packages or 12.5\%), although interviews with developers
+As mentioned in Section~\ref{SecSurfCorrectAndVerifiab}, LBM developers rarely
+uses CI (3 of 24 packages or 12.5\%), although interviews with developers
 suggest that the actual rate might be higher. Low use of CI for LBM contrasts
@@
 \citep{Dong2021}, which is comparable to the LBM usage rate. Our impression is
-that medical imaging software and LBM software are not unique -- most research
+that medical imaging software and LBM software are not unique --- most research
 software domains likely lag behind the best practice recommendations, and
@@
 	involves creating automated tests at the unit, component, and acceptance
-	test level, and executing these tests whenever a change is made to the code,
-	its configuration, the environment, or the software stack that it runs on
-	\citep[p.\ 83]{HumbleAndFarley2010}. As Table~\ref{artifactspresent} shows,
-	test cases are in the uncommon category for LBM software artifacts, which
-	means that some LBM projects will need to increase their testing automation
-	if they wish to pursue CI.  The usual advice for CI is to keep the build and
-	test process short \citep[p.\ 60]{HumbleAndFarley2010}. Given that LBM can
-	be computationally expensive, the tests run with every check-in may need to
-	focus on simple code interface tests, saving larger more complex tests for
-	less frequent runs.  A more sophisticated option for when tests create a
-	bottleneck for merges is CIVET (Continuous Integration,
+	test level, and executing these tests whenever someone makes a change to the
+	code, its configuration, the environment, or the software stack that it runs
+	on \citep[p.\ 83]{HumbleAndFarley2010}. As Table~\ref{artifactspresent}
+	shows, test cases are in the uncommon category for LBM software artifacts,
+	which means that some LBM projects will need to increase their testing
+	automation if they wish to pursue CI.  The usual advice for CI is to keep
+	the build and test process short \citep[p.\ 60]{HumbleAndFarley2010}. Given
+	that LBM can be computationally expensive, the tests run with every check-in
+	may need to focus on simple code interface tests, saving larger more complex
+	tests for less frequent runs.  A more sophisticated option for when tests
+	create a bottleneck for merges is CIVET (Continuous Integration,
 	Verification, Enhancement, and Testing), which solves this problem by
@@
 	programming practice or lack of adherence to adopted coding standards.
-	\citet{SlaughterEtAl2021} provides another example of an automated task --
+	\citet{SlaughterEtAl2021} provides another example of an automated task ---
 	checking that a test specification includes the test's motivation, a test
@@
 \begin{itemize}
-	\item Headaches associated with a separate integration phase are eliminated
+	\item Elimination of headaches associated with a separate integration phase
 	\citep{Fowler2006}, \citep[p.\ 20]{HumbleAndFarley2010}. If integration of
@@
 	fixed at the point where they are introduced \citep[p.\
-	23]{HumbleAndFarley2010}.  Errors are not just found in the code, but also
-	in the files and scripts related to configuration management \citep[p.\
-	18]{HumbleAndFarley2010}.
+	23]{HumbleAndFarley2010}.  Code is not the only source of errors; they are
+	also found in the files and scripts related to configuration management
+	\citep[p.\ 18]{HumbleAndFarley2010}.
 	\item Developers are always working on a stable base, since the master
 	branch will always be working.  A stable base passes all tests and, if the
-	CI system is configured with generators and linters, it will also have current documentation and coding standard compliant code.  A stable base
-	improves developer productivity, allowing them to focus on coding, testing
-	and documentation.
+	CI system uses generators and linters, it will also have current
+	documentation and coding standard compliant code.  A stable base improves
+	developer productivity, allowing them to focus on coding, testing, and
+	documentation.
 \end{itemize}
@@
 
-\section{Comparison of Principles, Process and Methodologies to Research Software in General} \label{Sec_CompareMethodologies}
+\section{Comparison of Principles, Process, and Methodologies to Research Software in General} \label{Sec_CompareMethodologies}
 
 We answer research question \rqref{RQ_CompareMethodologies} by comparing the
-principles, processes and methodologies used for LBM software to what can be
-gleaned from the literature on research software in general. In our data
-collection for LBM software, the software development process is not explicitly
-indicated in the artifacts for most of the packages. However, during our
-interviews one developer (ESPResSo) told us their non-rigorous development model
-is like a combination of agile and waterfall. Employing a loosely defined
-process makes sense for LBM software, given that the teams are generally small
-and self-contained.  Although eleven of the packages explicitly convey that they
-would accept outside contributors, generally the teams are centralized, often
-working at the same institution.  Working at the same institution means that an
-informal process can show success, since informal conversations are relatively
-easy to have.
+principles, processes, and methodologies used for LBM software to what can be
+gleaned from the literature on research software in general. Our measured data
+on LBM software artifacts does not explicitly indicate the development process.
+However, during our interviews one developer (ESPResSo) told us their
+non-rigorous development model is like a combination of agile and waterfall.
+Employing a loosely defined process makes sense for LBM software, given that the
+teams are generally small and self-contained.  Although eleven of the packages
+explicitly convey that they would accept outside contributors, teams are
+generally centralized, often working at the same institution. Working at the
+same institution means that an informal process can show success, since informal
+conversations are relatively easy to have.
 
 Interviews with developers confirmed a similar project management process. In
-teams of only a couple of developers, additions of new features or major changes
-are discussed with the entire team. Projects with more than a couple developers
-have lead developer roles. These lead developers review potential additions to
-the software. One of the developers (ESPResSo) noted that an ad hoc peer review
-process is used to assess major changes and additions. Using peer review (also
+teams of only a couple of developers, the entire team discusses additions of new
+features or major changes. Projects with more than a couple developers have lead
+developer roles. These lead developers review potential additions to the
+software. One of the developers (ESPResSo) noted that they use an ad hoc peer
+review process to assess major changes and additions. Using peer review (also
 called technical review) matches with recommended practice for research software
@@
 
-Two types of software changes were discussed during interviews with developers.
-One is feature additions, which arise from a scientific or functional need.
-These changes involve formal discussions within the development team, and lead
+During our interviews, we discussed two types of software changes. One is
+feature additions, which arise from a scientific or functional need. These
+changes involve formal discussions within the development team, and lead
 developer participation is mandatory. The other change type is code refactoring,
-which only sometimes involves formal discussions with the development team. New
-developers were noted to play an increased role in these changes compared to the
-former changes. Software bugs are typically addressed similarly to
-code refactoring.  Issue tracking is commonly used to manage these changes.
+which only sometimes involves formal discussions with the development team. We
+learned that new developers play a larger role with these changes compared to
+the former changes. Developers address software bugs similarly to code
+refactoring, and they manage change via issue tracking.
 
@@
 principles that were considered when developing the software. One package,
-Sailfish, indicates in its artifacts that shortening the development time was
-considered in early stages of design, with the developers using Python and
-CUDA/OpenCL to achieve this without sacrificing performance. The Sailfish goals
-are explicitly listed as performance, scalability, agility and extendability,
-maintenance, and ease of use. The project scored well in these categories during
-our assessment.  The quality priorities for Sailfish roughly match the
-priorities observed for research software in general.
-\citet{Nguyen-HoanEtAl2010} surveyed developers to find the following list of
-qualities, in decreasing order of importance: reliability, functionality,
-maintainability, availability, performance, flexibility, testability, usability,
-reusability, traceability, and portability. The Sailfish list does not list
-reliability or functionality, but we can safely assume those are implicitly high
-priorities for any scientific project.  In earlier studies
-\citet{KellyAndSanders2008} and \citet{CarverEtAl2007} highlight how important
-correctness is for research software.
+Sailfish, indicates in its artifacts the explicit goal of shortening development
+time, with the developers using Python and CUDA/OpenCL to achieve this without
+sacrificing performance. The Sailfish documentation explicitly lists the goals
+of performance, scalability, agility and extendability, maintenance, and ease of
+use. The project scored well in these categories during our assessment.  The
+quality priorities for Sailfish roughly match the priorities observed for
+research software in general. \citet{Nguyen-HoanEtAl2010} surveyed developers to
+find the following list of qualities, in decreasing order of importance:
+reliability, functionality, maintainability, availability, performance,
+flexibility, testability, usability, reusability, traceability, and portability.
+The Sailfish list does not list reliability or functionality, but we can safely
+assume those are implicitly high priorities for any scientific project.  In
+earlier studies \citet{KellyAndSanders2008} and \citet{CarverEtAl2007} highlight
+how important correctness is for research software.
 
@@
 
-During our interviews, documentation was noted as playing a significant role in
-the development process, specifically with on-boarding new developers. A goal of
-documentation is to lower the entry barrier for these new contributors. The
-documentation provides information on how to get started, orients the user to
-artifacts and the source code, and explains how the system works, including the
-so-called simulation engine and interface.  This emphasis on documentation,
+During our interviews, developers noted that documentation plays a significant
+role in the development process, specifically with on-boarding new developers. A
+goal of documentation is to lower the entry barrier for these new contributors.
+The documentation provides information on how to get started, orients the user
+to artifacts and the source code, and explains how the system works, including
+the so-called simulation engine and interface.  This emphasis on documentation,
 especially for new developers, is echoed in research software guidelines.
@@
 (not just research software), \citet{Fogel2005} recommends providing tutorial
-style examples, developer guidelines, demos and screenshots.
+style examples, developer guidelines, demos, and screenshots.
 
@@
 instance, the guidelines usually encourage modular design, but are generally
-silent on what criteria to use for the modularization.  Little information is
-given on how to document the design itself. Some non-mutually exclusive ideas
-for documenting the design, roughly in order of increasing effort, are as
-follows:
+silent on what criteria to use for the modularization.  The guidelines give
+little information on how to document the design itself. Some non-mutually
+exclusive ideas for documenting the design, roughly in order of increasing
+effort, are as follows:
 
@@
 	\item Explain the design using data flow diagrams to show typical use cases
-	for how inputs are transformed.
+	for input transformation.
 	\item Using a textual description, list for each module (or class), the
@@
 identified pain points and include citations that contrast the LBM experience
-with observations from researchers in other domains.  Potential ways to address
-the pain points are covered in Section~\ref{Sec_AddressConcerns}. The full
-interview questions are found in \citet{SmithEtAl2021}.
+with observations from researchers in other domains.
+Section~\ref{Sec_AddressConcerns} covers potential ways to address the pain
+points and \citet{SmithEtAl2021} provides the full interview questions.
 
@@
 	are common for LBM software packages (as shown in the measurement table
-	excerpt in Figure~\ref{measurement_template_image}). Lack of time is also
-	highlighted as a pain point by other research software developers
+	excerpt in Figure~\ref{measurement_template_image}). Other domains of
+	research software also experience the lack of time pain point
 	\citep{PintoEtAl2018, PintoEtAl2016, WieseEtAl2019}.
@@
 	\item[P\refstepcounter{pnum}\thepnum \label{P_LackSoftDevExp}:] \textbf{Lack
-	of Software Development Experience} A lack of software development
-	experience was noted by the developer of TCLB, and others noted a need for
-	improving software engineering education. Many of the team members on their
-	project are domain experts, not computer scientists or software engineers.
-	This same trend is noted by \citet{Nguyen-HoanEtAl2010}, which showed only
-	23\% of research software survey respondents having a computing-related
-	background. Similarly, \citet{UditAndKatz2017} show that the majority (54\%)
-	of postdocs have not received training in software development.  The LBM
-	developer suggesting an increasing role for formal software education
-	matches the trend observed by \citet{PintoEtAl2018}, where their replication
-	of a previous study \citep{HannayEtAl2009}, shows a growing interest in
-	formal training (From 13\% of respondents in 2009 to 22\% in 2018).
-	\citet{PintoEtAl2018} found that some developers feel there is a mismatch
-	between coding skills and subject-matter skills.
+	of Software Development Experience} The developer of TCLB noted a lack of
+	software development experience, and others noted a need for improving
+	software engineering education. Many of the team members are domain experts,
+	not computer scientists or software engineers. \citet{Nguyen-HoanEtAl2010}
+	noted this same trend, with only 23\% of research software survey
+	respondents having a computing-related background. Similarly,
+	\citet{UditAndKatz2017} show that the majority (54\%) of postdocs have not
+	received training in software development.  The LBM developer suggesting an
+	increasing role for formal software education matches the trend observed by
+	\citet{PintoEtAl2018}, where their replication of a previous study
+	\citep{HannayEtAl2009}, shows a growing interest in formal training (From
+	13\% of respondents in 2009 to 22\% in 2018). \citet{PintoEtAl2018} found
+	that some developers feel there is a mismatch between coding skills and
+	subject-matter skills.
 
@@
 	\item[P\refstepcounter{pnum}\thepnum \label{P_LackExtSupport}:] \textbf{Lack
-	of External Support} A concern was raised that there are no organizations
-	helping with the development of good quality software.  This concern is not
-	echoed in the literature because there are such organizations, including
-	\href{https://bssw.io/} {Better Scientific Software (BSSw)},
+	of External Support} Interviewees raised a concern that there are no
+	organizations helping with the development of good quality software.  The
+	literature on research software does not tend to echo this concern because
+	there are such organizations, including \href{https://bssw.io/} {Better
+	Scientific Software (BSSw)},
 	\href{https://www.software.ac.uk/} {Software Sustainability Institute}
@@
 	\item[P\refstepcounter{pnum}\thepnum \label{P_Correctness}:]
-	\textbf{Ensuring Correctness} Difficulties with ensuring correctness were
-	noted by several developers. They alluded to difficulty with testing the
+	\textbf{Ensuring Correctness} Several developers noted difficulties with
+	ensuring correctness. They alluded to difficulty with testing the
 	correctness of large numbers of features, and challenges with automated
-	testing. The TCLB developer commented that the amount of testing
-	data needed was sometimes problematic, since free testing services do not
-	offer adequate facilities for large amounts of data, which means in-house
-	testing solutions are needed.  Other research software domains point to the
-	following problems with testing: i) \citep{PintoEtAl2018} mention the
-	problem of insufficient testing; ii) the survey of \citep{HannayEtAl2009}
-	shows that more developers think testing is important than the number that
-	believe they have a sufficient understanding of testing concepts; and, iii) previous studies
+	testing. The TCLB developer commented that the amount of testing data needed
+	was sometimes problematic, since free testing services do not offer adequate
+	facilities for large amounts of data, requiring in-house testing solutions.
+	Other research software domains point to the following problems with
+	testing: i) \citep{PintoEtAl2018} mention the problem of insufficient
+	testing; ii) the survey of \citep{HannayEtAl2009} shows that more developers
+	think testing is important than the number that believe they have a
+	sufficient understanding of testing concepts; and, iii) previous studies
 	\citep{HannayEtAl2009, KanewalaAndBieman2013, KellyEtAl2011, WieseEtAl2019}
@@
 	\item[P\refstepcounter{pnum}\thepnum \label{P_TechDebt}:] \textbf{Technical
-	Debt} The developer of ESPResSo said that their source code was written with
-	a specific application in mind, which later caused too much coupling between
-	components in the source code. This results in technical debt
-	\citep{KruchtenEtAl2012}, which has an impact on future maintainability and
-	reusability. Concern with technical debt is likely why researchers in the
-	survey of \citet{Nguyen-HoanEtAl2010} rated maintainability as the third
+	Debt} The developer of ESPResSo said that they originally wrote their source
+	code with a specific application in mind, which later caused too much
+	coupling between components in the source code. This results in technical
+	debt \citep{KruchtenEtAl2012}, which has an impact on future maintainability
+	and reusability. Concern with technical debt is likely why researchers in
+	the survey of \citet{Nguyen-HoanEtAl2010} rated maintainability as the third
 	most important software quality. More recently the push for sustainable
@@
 	\item[P\refstepcounter{pnum}\thepnum \label{P_Documentation}:]
-	\textbf{Quality of Documentation}  The importance of documentation for both
-	users and developers was stressed throughout the interviews. However, it was
-	noted several times that a lack of time and funding (\ppref{P_LackDevTime})
-	has a negative effect on the documentation. Most of the developers are
-	scientific researchers evaluated on the scientific papers that they produce.
-	Writing and updating documentation is something that is done in their free
-	time, if that time arises. Inadequate research software documentation is
-	also mentioned by others \citep{PintoEtAl2018, WieseEtAl2019} and the
+	\textbf{Quality of Documentation}  Interviewees stressed the importance of
+	documentation for both users and developers throughout the interviews.
+	However, it was noted several times that a lack of time and funding
+	(\ppref{P_LackDevTime}) has a negative effect on the documentation. Most of
+	the developers are scientific researchers evaluated on the scientific papers
+	that they produce. Writing and updating documentation is something that is
+	done in their free time, if that time arises. Others also mention inadequate
+	research software documentation \citep{PintoEtAl2018, WieseEtAl2019}.  The
 	problem also arises with non-research software \citep{LethbridgeEtAl2003}.
@@
 the model.  We also learned that the package pyLBM redeveloped data structures
-to ease future changes. The developer of TCLB noted that their software package
-was designed to allow for the addition of some LBM features, but changes to
-major aspects of the system would be difficult. For example, ``implementing a
-new model will be an easy contribution'', but changes to the ``Cartesian mesh
-... will be a nightmare''.  The design of TCLB highlights that not every
-conceivable change needs to be supported, only the likely changes.
-
-As the LBM developers illustrate, design for changes is accomplished by first
+to ease future changes. The developer of TCLB noted that their design allows for
+the addition of some LBM features, but changes to major aspects of the system
+would be difficult. For example, ``implementing a new model will be an easy
+contribution'', but changes to the ``Cartesian mesh … will be a nightmare''.
+The design of TCLB highlights that not every conceivable change needs to be
+supported, only the likely changes.
+
+As the LBM developers illustrate, they accomplish design for change by first
 identifying likely changes, either implicitly or explicitly, and second by
@@
 
-\noindent Additional strategies for getting around the oracle problem are
-provided in Section~\ref{Sec_Recommendations}.
+\noindent Section~\ref{Sec_Recommendations} provides additional strategies for
+getting around the oracle problem.
 
@@
 Although documentation is difficult (as mentioned under pain point
-\ppref{P_Documentation}), some LBM developers have realized that prioritizing time
-on documentation provides a justifiable return on investment. For instance, the
-interviewed developer of ESPResSo noted that all major additions to their
-package had accompanying changes to artifacts and documentation. They noted that
-considerable effort was put into their documentation. They further commented
-that their goal was to lower the entry barrier for new developers, which meant
-providing considerable developer documentation. This documentation informs
-developers on how to get started; orients them to the artifacts, source code,
-system architecture, and build system; and, explains the coupling between
-the simulation engine and the interface. In addition, ESPResSo developers
-included API documentation, although they are the only one of the top five
-packages to do so.
-
-Furthermore, documentation is used by LBM developers to address usability
-challenges (\ppref{P_Usability}), as follows:
+\ppref{P_Documentation}), some LBM developers have realized that prioritizing
+time on documentation provides a justifiable return on investment. For instance,
+the interviewed developer of ESPResSo noted that all major additions to their
+package had accompanying changes to artifacts and documentation. They put
+considerable time into their documentation. They further commented that their
+goal was to lower the entry barrier for new developers, which meant providing
+considerable developer documentation. This documentation informs developers on
+how to get started; orients them to the artifacts, source code, system
+architecture, and build system; and, explains the coupling between the
+simulation engine and the interface. In addition, ESPResSo developers included
+API documentation, although they are the only one of the top five packages to do
+so.
+
+Furthermore, LBM developers use documentation to address usability challenges
+(\ppref{P_Usability}), as follows:
 
@@
 	\item Explicitly state appropriate fluid dynamics problems that the software
-	is designed to model and explicitly state the limits of the software. This
-	is done in the ESPResSo user guide.
+	uses in its models and explicitly state the software's limitations. This is
+	done in the ESPResSo user guide.
 	\item Provide documentation that details the background theory, or provide a
@@
 	\item Use documentation generators, like Doxygen, as done (for instance) by
-	OpenLB.  More details on the packages that use generators is given in
-	Sections~\ref{SecSurfCorrectAndVerifiab} and~\ref{Sec_CompareTools}.
+	OpenLB.  Sections~\ref{SecSurfCorrectAndVerifiab} and~\ref{Sec_CompareTools}
+	give more details on the packages that use generators.
 	\item Provide an explanation of theoretical principles through the user interface.
@@
 are not lists of criticisms for what should have been done in the past, or what
-should be done now; they are suggestions for consideration in the future.
-We will not be repeating ideas here that have been discussed in previous
-sections, such as CI, documentation of APIs, etc.  Our aim
-is to mention ideas that are at least somewhat beyond conventional best
-practices. The ideas listed here have the potential to become best practices in
-the medium to long-term. The ideas are presented roughly in the order of
-increasing novelty.
+should be done now; they are suggestions for consideration in the future. We
+will not be repeating previously discussed ideas here, such as CI, documentation
+of APIs, etc.  Our aim is to mention ideas that are at least somewhat beyond
+conventional best practices. The ideas listed here have the potential to become
+best practices in the medium to long-term. The ideas below are roughly in the
+order of increasing novelty.
 
@@
 linter can include rules that capture the wisdom of senior programmers, it can
-help newer developers avoid common mistakes (\ppref{P_LackSoftDevExp}).  Although
-there do not appear to be many linters for checking parallel programming
-mistakes (\ppref{P_TechnologyHurdles}), Parallel Lint is an option for OpenMP.
-If a linter is used consistently this can guard against some technical debt
-(\ppref{P_TechDebt}). Although linters are thought of as tools for code
-analysis, similar ideas can be applied to documentation to ensure that at least
-the basic documentation rules are followed.  The use of tools to check
-documentation is a partial explanation for the relatively higher quality of
-statistical tools that are part of the Comprehensive R Archive Network (CRAN)
-\citep{SmithEtAl2018_StatSoft}.
+help newer developers avoid common mistakes (\ppref{P_LackSoftDevExp}).
+Although there do not appear to be many linters for checking parallel
+programming mistakes (\ppref{P_TechnologyHurdles}), Parallel Lint is an option
+for OpenMP. Consistently using a linter can guard against some technical debt
+(\ppref{P_TechDebt}). Although linters are tools for code analysis, similar
+ideas can be applied to ``lint'' documentation to ensure adherence to basic
+rules.  The use of tools to check documentation is a partial explanation for the
+relatively higher quality of statistical tools that are part of the
+Comprehensive R Archive Network (CRAN) \citep{SmithEtAl2018_StatSoft}.
 
@@
 \citet{ChueHongEtAl2019} provides a software citation checklist for developers
-to make a release of their software citable.  Further guidance on software
-citation is provided by \citet{KatzEtAl2021}.  Another option is to use GitHub
-for citation, since it is relatively easy to add metadata to repositories and to
+to make a release of their software citable.  \citet{KatzEtAl2021} provides
+further guidance on software citation.  Another option is to use GitHub for
+citation, since it is relatively easy to add metadata to repositories and to
 generate citations for those repositories \citet{Smith2021-CitationsOnGitHub}.
@@
   can be compared.  In the case of solving Partial Differential Equations
-  (PDEs), this approach is called the Method of Manufactured Solutions
+  (PDEs), the name for this approach is the Method of Manufactured Solutions
   \citep{Roache1998}.
@@
   slow, but guaranteed correct, interval arithmetic \citep{Hickey2001} can be
-  employed. The faster floating point algorithm can then be verified by ensuring
-  that the calculated answers lie within the guaranteed bounds.
+  employed. Analysts can then verify the faster floating point algorithm by
+  ensuring that the calculated answers lie within the guaranteed bounds.
 \item Include convergence studies.  The discretization used in the numerical
@@
   come up during our interviews.
-\item Use \emph{metamorphic} testing, where the program is checked to see if it
-  satisfies a set of metamorphic relations, where a metamorphic relation is a
-  relationship between multiple inputs and output pairs
-  \citep{KanewalaAndLundgren2016}.  An example could be the relationship that
-  the output flow rate increases as the input flow rate increases.
+\item Use \emph{metamorphic} testing, which checks a program to see if it
+  satisfies a set of metamorphic relations, where a metamorphic relation relates
+  multiple inputs and output pairs \citep{KanewalaAndLundgren2016}.  An example
+  could be the relationship that the output flow rate increases as the input
+  flow rate increases.
 \end{itemize}
@@
 
-As discussed in Section~\ref{Sec_CompareArtifacts}, requirements specifications
-are rarely written for research software. Although a full requirements
+As discussed in Section~\ref{Sec_CompareArtifacts}, developers rarely write requirements specifications for research software. Although a full requirements
 specification has advantages \citep{SmithEtAl2007, SmithAndLai2005},
@@
 
-Ensuring correctness was identified as a pain point by developers
+Developers identified ensuring correctness as a pain point
 (\ppref{P_Correctness}).  Inspired by \citet{BenureauAndRougier2017}, we make
@@
 
-	\item[Reproducibility] Once re-runnability and repeatability are achieved,
-	the next step is reproducibility.  Current work by LBM developers on
-	reproducibility is discussed in Section~\ref{SecReproducibility}.  The
-	proposed goal for the future is for all results to be reproducible, which
-	means documenting dependencies, platforms, parameter values and input files.
-	The data and scripts behind the graphs should be published
-	\citep{BenureauAndRougier2017}.  Some tools and techniques for computational
-	reproducibility are summarized in \citet{PiccoloAndFrampton2016}.
+	\item[Reproducibility] Once we achieve re-runnability and repeatability, the
+	next step is reproducibility.  Section~\ref{SecReproducibility} discusses
+	current work by LBM developers on reproducibility.  The proposed goal for
+	the future is for all results to be reproducible, which means documenting
+	dependencies, platforms, parameter values and input files. The data and
+	scripts behind the graphs should be published
+	\citep{BenureauAndRougier2017}.  \citet{PiccoloAndFrampton2016} summarizes
+	some tools and techniques for computational reproducibility.
 
@@
 code. DSM moves development to a higher level of abstraction; domain experts can
-work without concern for low-level implementation details. Using DSM, code and
-documentation can be optimally generated, redundancy is eliminated, the
-likelihood of errors is reduced and maintenance is automated. Moreover, a
-generative approach facilitates the inevitable exploration of LBM modelling
-assumptions.
+work without concern for low-level implementation details. Using DSM, we
+optimally generate code and documentation, eliminate redundancy, reduce the
+likelihood of errors, and automate maintenance. Moreover, a generative approach
+facilitates the inevitable exploration of LBM modelling assumptions.
 
@@
 removes the maintenance nightmare of documentation duplicates and near
-duplicates \citep{LucivEtAl2018}, since knowledge is only captured once and
-transformed as needed.  Code generation has previously been applied to improve
+duplicates \citep{LucivEtAl2018}, since developers capture knowledge once and
+transform it as needed.  Code generation has previously been applied to improve
 research software.  For instance, ATLAS (Automatically Tuned Linear Algebra
@@
 focuses on generating all software artifacts (requirements, design, etc.), not
-just code.  Initial work on this ``generate all things'' approach is discussed
-in \citep{SzymczakEtAl2016} with a prototype available
-\citep{CaretteEtAl2021-Drasil}.
+just code.  \citet{SzymczakEtAl2016} presents initial work on this ``generate
+all things'' approach and \citep{CaretteEtAl2021-Drasil} provides a prototype.
 
@@
 (\ppref{P_Correctness}) a generative approach can aim to be correct by
-construction.  In cases where mistakes are made, DSM has the advantage that the
-mistake is propagated throughout the generated artifacts, which greatly
-increases the chance that the mistake is noted.  Technical debt
-(\ppref{P_TechDebt}) is not a concern with a DSM approach since the recipes that
-are used for generation are written at a high level and are thus relatively easy
-to change.  With respect to pain points on usability (\ppref{P_Usability}) and
+construction.  If there are mistakes, DSM has the advantage that they are
+propagated throughout the generated artifacts, which greatly increases the
+chance that someone will notice the mistake.  Technical debt
+(\ppref{P_TechDebt}) is not a concern with a DSM approach since developers write
+the recipes used for generation at a high level making them relatively easy to
+change.  With respect to pain points on usability (\ppref{P_Usability}) and
 documentation (\ppref{P_Documentation}), the generative approach we are
@@
 literature review.  Our methodology is essentially a systematic software review
--- the primary studies are the software, and our work collects and analyzes
+--- the primary studies are the software, and our work collects and analyzes
 these primary studies.
@@
 researcher(s) doing the study.  That is, repetition of the study by a different
-researcher should achieve the same results, as long as the same methodology is
-used \citep{AmpatzoglouEtAl2019}.  For our study the identified reliability
-related threats include the following:
+researcher should achieve the same results, as long as they use the same
+methodology \citep{AmpatzoglouEtAl2019}.  For our study the identified
+reliability related threats include the following:
 
 \begin{itemize}
-\item In our methodology, the manual measures are done by one individual for all
-packages. The measures may be influenced by the grader's abilities, experience,
-and biases.  
+\item In our methodology, one individual does the manual measures for all
+packages. The grader's abilities, experiences, and biases may influence the
+results.  
 \item Completing the manual measures of all packages took place over several
@@
 
-Construct validity is threatened if metrics are used that do not measure what
-they are intended to measure \citep{AmpatzoglouEtAl2019}. The construct
-validity related threats for the LBM study include the following:
+Construct validity means that adopted metrics represent what they are intended
+to measure \citep{AmpatzoglouEtAl2019}. The construct validity related threats
+for the LBM study include the following:
 
@@
 it is possible that some packages have many files, with few LOC, but the files
-do not contain source code that is easily reusable. The files may be poorly
-formatted, or the source code may be vague and have ambiguous identifiers. 
-\item Maintainability may be incorrectly estimated, since it is assumed to be
-improved by a higher ratio of comments to source code
+do not contain source code that is easily reusable. The formatting may be poor,
+or the source code may be vague and have ambiguous identifiers. 
+\item We may have inaccurately estimated maintainability by assuming a higher
+ratio of comments to source code improves maintainability
 (Section~\ref{Sec_Maintainability}). 
@@
 community.
-\item In our manual measures of the software repositories, some data may not have
-been measured due to technology issues like broken links. This issue arose with
-the measurement of Palabos, which had a broken link to its user manual, as noted
-in Section~\ref{Sec_CompareArtifacts}. 
+\item In our manual measures of the software repositories, we may have missed
+some data due to technology issues like broken links. This issue arose with the
+measurement of Palabos, which had a broken link to its user manual, as noted in
+Section~\ref{Sec_CompareArtifacts}. 
 \item Our methodology assumes that artifacts will be present in the repositories
-to approach used for software development.  However, not all activities will
-necessarily leave a trace in the repositories. For instance, the use of unit
-testing and CI was mentioned in the artifacts of only three (ESPResSo, Ludwig,
-Musubi) packages (Section~\ref{SecSurfCorrectAndVerifiab}). However, interviews
-suggested a more frequent use of both unit testing and CI. For example, OpenLB,
+to show the approach used for software development.  However, not all activities
+will necessarily leave a trace in the repositories. For instance, only three
+packages (ESPResSo, Ludwig, Musubi) show evidence in their artifacts of unit
+testing and CI (Section~\ref{SecSurfCorrectAndVerifiab}), even though interviews
+suggest a more frequent use of both unit testing and CI. For example, OpenLB,
 pyLBM, and TCLB use such methods during development despite this not being
@@
 
-Although we found that unit testing and CI were used more often than the
-repository artifacts suggested, we did not change the manual measurement
-results.  Given that we could not interview a representative for every project,
-we left the data as collected to keep our measures consistent.  Moreover, CI and
-unit testing done in private should not be counted as equivalent to the public
-versions we detected.  When CI and unit testing are done outside the public
-repository, they are not as effective since they are not available for others
-to inspect and reproduce.
+Although we found that unit testing and CI usage is likely higher than the
+repository artifacts suggest, we did not change the manual measurement results.
+Given that we could not interview a representative for every project, we left
+the data as collected to keep our measures consistent.  Moreover, CI and unit
+testing done in private should not be counted as equivalent to the public
+versions we detected.  When CI and unit testing are private, they are not as
+effective since they are not available for others to inspect and reproduce.
 
@@
 state of the practice for developing LBM software. Understanding what artifacts
-are produced, what tools are used, what processes are employed and what pain
-points are experienced is the starting point for devising future methods and
-tools for reducing development time and improving software quality.
+developers produce, what tools they use, what processes they employ, and what
+pain points they experience is the starting point for devising future methods
+and tools for reducing development time and improving software quality.
 
@@
 	consists of 108 questions to assess 9 qualities (including the qualities of
-	installability, usability and visibility);
+	installability, usability and visibility));
 	\item Ranked the software using the Analytic Hierarchy Process (AHP);
@@
 \end{inparaenum}
-The collected data was analyzed by: 
+We analyzed the collected data by: 
 \begin{inparaenum}[i)]
@@
 
-Current LBM software mostly follows best practices as illustrated by
+Current LBM software mostly follows best practices, as illustrated by
 Table~\ref{topperformerstable}, which summarizes the top five performers for
@@
 Palabos, and HemeLB.  Therefore, two thirds (67\%) of the 24 measured packages
-appear in the top five for at least one quality.  Of these entries almost all
-are classified as alive, except for three packages: laboetie, MP-LABS, and
-HemeLB.  Alive packages are in a healthy state with respect to adopting best
-practices.
+appear in the top five for at least one quality.  Almost all entries are alive,
+except for three packages: laboetie, MP-LABS, and HemeLB.  Alive packages are in
+a healthy state with respect to adopting best practices.
 
@@
 
-LBM software is shown to be in a healthy state when comparing artifacts, tools
-and processes, to research software guidelines and other research software.  The
-majority of LBM generated artifacts correspond to general recommendations from
-research software developers (Section~\ref{Sec_CompareArtifacts}).  LBM software
-even seems to be ahead of best practices with an example of a video user guide
-for one of the packages. LBM is mostly keeping pace with other research software
-with respect to tool usage (Section~\ref{Sec_CompareTools}), with 67\% of the
-surveyed tools using version control.  The state of the practice for the
-development of LBM software matches the quasi-agile process generally adopted
-for research software (Section~\ref{Sec_CompareMethodologies}).  The best
-practice process of peer review is employed by some LBM developers.
+LBM software is in a healthy state, as shown by comparing LBM artifacts, tools
+and processes, to current software guidelines and to other research software.
+The majority of LBM generated artifacts correspond to general recommendations
+from research software developers (Section~\ref{Sec_CompareArtifacts}).  LBM
+software even seems to be ahead of best practices with an example of a video
+user guide for one of the packages. LBM is mostly keeping pace with other
+research software with respect to tool usage (Section~\ref{Sec_CompareTools}),
+with 67\% of the surveyed tools using version control.  The state of the
+practice for the development of LBM software matches the quasi-agile process
+generally adopted for research software
+(Section~\ref{Sec_CompareMethodologies}).  Some LBM developers employ the best
+practice process of peer review.
 
@@
 	\item Increase the use and enforcement of programming style guides to improve understandability and productivity. (Section~\ref{Sec_CompareArtifacts})
-	\item Use checklists to ensure best practices are followed by developers. (Section~\ref{Sec_CompareArtifacts})
+	\item Use checklists to ensure developers follow best practices.
+	(Section~\ref{Sec_CompareArtifacts})
 	\item Include uninstall instructions as a courtesy to users that wish to
@@
 — maybe add a continuous integration diagram?
-
-UNUSED Text
-
-Conclusions Related
-
-To help users select the right LBM package for their needs, and to potentially
-provide direction for future LBM software development, our goal was to analyze
-the current state of LBM software development.  We determined how well best
-practices are followed, and we investigated current developer pain points.  We
-compared the state of the practice for LBM software to the development practices
-recommended and employed for research software in general. To improve the
-quality of LBM software we looked for examples where developers have been
-successful so that we can share these with the LBM community, and with the
-broader research software community.  We also aimed to identify potential
-approaches for mitigating the identified pain points in the future.

### 2022-04-24 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Edits to introduction

@@
 	Lattice Boltzmann Method (LBM), research software, software engineering,
-	software quality, Analytic Hierarchy Process
+	software quality, Analytic Hierarchy Process (AHP)
 \end{keyword}
@@
 manufacturing, and health and safety. Given the critical applications of LBM,
-users worry about with software qualities like reliability,
-robustness, reproducibility, performance, correctness, and verifiability.  Since
-their time is valuable, developers of LBM libraries have additional quality
-concerns, including maintainability, reusability, and understandability.  With
-these quality considerations in mind, and the potentially overwhelming number of choices for existing LBM libraries, the time is right to assess the state of the practice.
-Therefore, we analyze the current state of LBM
-software development and compare it to the development practices employed for
-research software in general.  We want to highlight success stories that LBM
-developers can share amongst their community, and with the broader research
-software community, while at the same time looking for areas for potential
-future improvement.
+users worry about software qualities like reliability, robustness,
+reproducibility, performance, correctness, and verifiability.  Since their time
+is valuable, developers of LBM libraries have additional quality concerns,
+including maintainability, reusability, and understandability.  With these
+quality considerations in mind, and the potentially overwhelming number of
+choices for existing LBM libraries, the time is right to assess the state of the
+practice. Therefore, we analyze the current state of LBM software development
+and compare it to the development practices employed for research software in
+general.  We want to highlight success stories that LBM developers can share
+amongst their community, and with the broader research software community, while
+at the same time looking for areas for potential future improvement.
 
 We use ten research questions to structure and focus our discussion. Our
-methodology \citep{SmithEtAl2021, SmithAndMichalski2022} is driven by the
+methodology \citep{SmithEtAl2021, SmithAndMichalski2022} is driven by these
 questions. For each question below, we point to the section that contains our
@@
 mean methods, techniques, and tools that are generally believed to improve
-software development, like testing and documentation.  We will categorize our
-best practices around software qualities, as we will discuss in
-Section~\ref{empiricalmeasures}.  At this point in the process, to remove
-potential user/developer bias, we will base our assessment only on publicly
-available artifacts, where artifacts are the documents, scripts, and code that we
-find in a project's public repository. Example artifacts include requirements,
-specifications, user manuals, unit test cases, system tests, usability tests,
-build scripts, API (Application Programming Interface) documentation, READMEs,
-license documents, process documents, and code. Following best practices does
-not guarantee popularity, so we will also compare our ranking to how the user
-community ranks the identified projects.
+software development, like testing and documentation.  We categorize our best
+practices around software qualities (Section~\ref{softwarequalities}).  At this
+point in the process, to remove potential user/developer bias, we base our
+assessment only on publicly available artifacts, where artifacts are the
+documents, scripts, and code that we find in a project's public repository.
+Example artifacts include requirements, specifications, user manuals, unit test
+cases, system tests, usability tests, build scripts, API (Application
+Programming Interface) documentation, READMEs, license documents, process
+documents, and code. Following best practices does not guarantee popularity, so
+we also compare our ranking to how the user community ranks the identified
+projects.
 
@@
 	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Concerns}:] For LBM
-	developers what specific best practices address the pain points
+	developers what specific best practices address their pain points
 	and software quality concerns? (Section~\ref{Sec_AddressConcerns})
@@
 quantitative and qualitative data (Sections~\ref{methodology}
-to~\ref{Sec_AddressConcerns}).  Finally, we summarize the threats to validity
+to~\ref{Sec_Recommendations}).  Finally, we summarize the threats to validity
 (Section~\ref{threats}) and the final conclusions and recommendations

### 2022-04-25 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Revisions to Section 2.1, fixed verb tense

@@
 
-We developed a methodology for evaluating the state of the practice of research
-software \citep{SmithEtAl2021, SmithAndMichalski2022} and applied it to the
-specific domain of LBM software.  Since we group best practices around
-conventional software qualities, the first section below provides definitions
-for the qualities of interest (Section~\ref{softwarequalities}), along with
-information on how to measure these qualities. The following section
+We apply our methodology for evaluating the state of the practice of research
+software \citep{SmithEtAl2021, SmithAndMichalski2022} to the specific domain of
+LBM software.  Since we group best practices around conventional software
+qualities, the first section below provides definitions for the qualities of
+interest (Section~\ref{softwarequalities}), along with information on how to
+measure these qualities. The following section
 (Section~\ref{Sec_OverallProcess}) provides an overview of the steps we used to
-select, measure and compare LBM software.  The remaining sections provide
-details on each step in the methodology.
+select, measure and compare LBM software.  The remaining sections
+(Sections~\ref{identifysoftware} to~\ref{SecDomainAnalysis}) provide details on
+each step in the methodology.
 
@@
 	effectiveness, efficiency, and satisfaction in a specified context of use
-	\citep{nielsonusability}. We assumed that high usability correlates with
+	\citep{nielsonusability}. We assume that high usability correlates with
 	the presence of documentation, including tutorials, manuals, defined
@@
 	manuals, specifications, README files, change logs, release notes,
-	publications, forums, and instructional websites. We also asked developers
-	if they considered the ease of future changes when developing the software
-	packages.
+	publications, forums, and instructional websites. We also ask developers
+	if they considered the ease of future changes when developing their
+	software.
 
@@
 	developer can take the original code and input data, run it, and obtain the
-	original output \citep{BenureauAndRougier2017}. We measured reproducibility
+	original output \citep{BenureauAndRougier2017}. We measure reproducibility
 	qualitatively by asking developers if they have any concerns that their
-	computational results won't be reproducible and if they have taken any
+	computational results are not reproducible, and if they have taken any
 	steps to ensure reproducibility.

### 2022-04-25 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Revisions completed for Section 2

@@
 select, measure and compare LBM software.  The remaining sections
-(Sections~\ref{identifysoftware} to~\ref{SecDomainAnalysis}) provide details on
-each step in the methodology.
+(Sections~\ref{identifysoftware} to~\ref{sec_vet_software_list}) provide details
+on each step in the methodology.
 
@@
 	can be found in \citet{SmithEtAl2021}. (Section~\ref{empiricalmeasures})
-	\item Use AHP to rank the software packages. (Section~\ref{AHP})
+	\item Rank the software using AHP. (Section~\ref{AHP})
 	\item Interview the developers. (Section~\ref{SecSurvey})
-	\item Domain analysis. (Section~\ref{SecDomainAnalysis})
+	\item Analyze the domain. (Section~\ref{SecDomainAnalysis})
 	\item Analyze the results and answer the research questions.
@@
 
-To answer~\rqref{RQ_WhatProjects} we identify existing LBM projects. We find the
-candidate software through search engine queries targeting authoritative lists.
-Scholarly articles, \href{https://github.com/} {GitHub} and
-\href{https://swmath.org/} {swMATH} provide candidate LBM software. The Domain
-Expert (Section~\ref{sec_vet_software_list}) helps with selecting the candidate
-software.
-
-We consider the following properties when creating the list and reviewing the
-candidate software:
+To answer~\rqref{RQ_WhatProjects} we identify existing LBM projects through
+search engine queries targeting authoritative lists. We find the software in
+scholarly articles, \href{https://github.com/} {GitHub} and
+\href{https://swmath.org/} {swMATH}. The Domain Expert
+(Section~\ref{sec_vet_software_list}) also helps with selecting the candidate
+software.  We build our list based on the following criteria:
 
@@
 
-The initial list had 45 packages, including some that we later found we could
-not use because of incompleteness or an absence of publicly available source
-code.
+\noindent Our initial list had 45 packages, including some that we later found
+to be unusable because of incompleteness, or an absence of publicly available
+source code.
 
@@
 
-To reduce the length of the list of software packages and to
-answer~\rqref{RQ_WhatProjects}, we filter the initial list of 45 packages. We
-apply the following filters in the priority order listed.
+To reduce the length of the list and to answer~\rqref{RQ_WhatProjects}, we
+filter the initial list of 45 packages. We apply the following filters in the
+priority order listed.
 
@@
 
-For the third item in the above filter, we categorized software packages as
+For the third item in the above filter, we categorize software packages as
 `alive' if documentation updates happened within the last 18 months, and `dead'
@@
 obliged to include it since it clearly fits our criteria. Time constraints meant
-we didn't update the previous manual data collection, but we did redo the
+we could not update the previous manual data collection, but we did redo the
 automated data collection.
@@
 The final list of 24 software packages is in two tables.
-Table~\ref{alivepackages} lists packages that fell into the `alive' category as
+Table~\ref{alivepackages} lists packages that fall into the `alive' category as
 of mid-2020 (January 2022 for Musubi), and Table~\ref{deadpackages} lists
-packages that were `dead'. The tables include a hyperlink to each project and,
+packages that are `dead'. The tables include a hyperlink to each project and,
 when available, citations for relevant publications.  The final list of software
@@
 
-We rank the projects by how well they follow best practices
-(\rqref{RQ_HighestQuality}) via a measurement template, as described in
-\citet{SmithEtAl2021}.  For each software package (each column in the template),
-we fill in the rows of the template. This process takes about two hours per
-package, with a cap of four hours. The time constraint is necessary so that the
-workload is feasible for a team as small as one, given our aim to cap the
-measurement phase at 160 person-hours \citep{SmithEtAl2021}.  An excerpt of the
-template, in spreadsheet form, is shown in
-Figure~\ref{measurement_template_image}.
+We use a measurement template to rank the projects by how well they follow best
+practices (\rqref{RQ_HighestQuality}), as described in \citet{SmithEtAl2021}.
+For each software package (each column in the template), we fill in the rows of
+the template. This process takes about two hours per package, with a cap of four
+hours. The time constraint is necessary so that the workload is feasible for a
+team as small as one, given our aim to cap the measurement phase at 160
+person-hours \citep{SmithEtAl2021}.  An excerpt of the template, in spreadsheet
+form, is shown in Figure~\ref{measurement_template_image}.
 
@@
 an open-source license; free-ware, which freely provides a binary or executable,
-but not source code; and commercial, where the user must pay for the software.
+but not source code; and commercial, where the user must pay.
 
@@
 usability considerations by searching for cues in the documentation, such as the
-presence of getting started instructions, a user manual or documentation of
+presence of getting started instructions, a user manual, or documentation of
 expected user characteristics.
@@
 We used tools to automatically measure the number of files, number of lines of
-code (LOC), percentage of closed issues, etc. GitStats measures each package's
-GitHub repository for the number of binary files, the number of added and
-deleted lines, and the number of commits over varying time intervals. Sloc Cloc
-and Code (scc) measures the number of text files and the number of total, code,
-comment, and blank lines in each GitHub repository.
-
-As in \citet{SmithEtAl2016}, we used Virtual machines (VMs) to provide an
+code (LOC), percentage of closed issues, etc.
+\href{https://github.com/tomgi/git_stats}{GitStats} \citep{Gieniusz2019}
+measures each package's GitHub repository for the number of binary files, the
+number of added and deleted lines, and the number of commits over varying time
+intervals. \href{https://github.com/boyter/scc}{Sloc Cloc and Code (scc)}
+\citep{Boyter2021} measures the number of text files and the number of total,
+code, comment, and blank lines in each GitHub repository.
+
+As in \citet{SmithEtAl2016}, we used Virtual Machines (VMs) to provide an
 optimal testing environment for each package. We used VMs to create fresh
-environments, eliminating worry about existing libraries and conflicts.
+environments, eliminating any worries about existing libraries and conflicts.
 Moreover, when the tests are complete, the VM can be deleted, without impacting
-the host operating system. The most significant advantage of using VMs is to
-level the playing field. Every software install starts from a clean slate, which
-removes ``works-on-my-computer'' errors.
+the host operating system. The most significant advantage of using VMs is that
+they level the playing field. Every software install starts from a clean slate,
+which removes ``works-on-my-computer'' errors.
 
@@
 
-Our ranking uses a tool we developed for conducting the AHP analysis. The tool
-includes a sensitivity analysis to ensure that the software package rankings are
+We use our own simple tool for conducting the AHP analysis. The tool includes a
+sensitivity analysis to ensure that the software package rankings are
 appropriate with respect to the uncertainty of the quality scores. For the
@@
 
-Several of the research question (\rqref{RQ_CompareToolsProjMngmnt},
-\rqref{RQ_CompareMethodologies}, \rqref{RQ_PainPoints}, \rqref{RQ_Concerns} and
-\rqref{RQ_Recommend}) require going beyond the quantitative data from the
-measurement template. To gain insight, we interviewed developers using a list of
-20 questions from \citet{SmithEtAl2021}. The questions cover the background of
-the development teams, the interviewees, and the software itself. We ask the
-developers how they organize their projects. We also ask about their
-understanding of the users. Some questions focus on the current and past
-difficulties, and the solutions the team has found or will try. We also discuss
-the importance of, and the current situation for, documentation. A few questions
-focus on specific software qualities, such as maintainability,
-understandability, usability, and reproducibility. The interviews are
-semi-structured based on the question list; we ask follow-up questions when
-necessary. Based on our experience, the interviewees often raise exciting and
-unexpected ideas.
+Two of the research question (\rqref{RQ_PainPoints} and \rqref{RQ_Concerns})
+require going beyond the quantitative data from the measurement template. To
+gain insight, we interview developers using a list of 20 questions from
+\citet{SmithEtAl2021}. The questions cover the background of the development
+teams, the interviewees, and the software itself. We ask the developers how they
+organize their projects and about their understanding of the users. Some
+questions focus on the current and past difficulties, and the solutions the team
+has found or will try. We also discuss the importance of, and the current
+situation for, documentation. A few questions focus on specific software
+qualities, such as maintainability, understandability, usability, and
+reproducibility. The interviews are semi-structured based on the question list;
+we ask follow-up questions when necessary. Based on our experience, the
+interviewees often raise exciting and unexpected ideas.
 
@@
 of variation, which summarize the possible values for that variability, along
-with their potential binding time. The binding time is when we specify a value
-for a variability. Binding time options include specification time, build time
+with their potential binding time. The binding time is when a variability is
+assigned a value. Binding time options include specification time, build time
 (when the program compiles) and run time (when the code executes). For the
@@
 
-To keep the current presentation focused, we only present the results of our
-shallow domain analysis. We distinguish programs by their variabilities, which
-for research software are often assumptions. Table~\ref{tbl_features} lists the
-variabilities as columns, with the software alphabetically ordered.
-\citet{Michalski2021} provides a detailed analysis of the commonalities and
-variabilities between LBM packages.
+To focus the current presentation, we only present the results of our shallow
+domain analysis. We distinguish programs by their variabilities, which for
+research software are often assumptions. Table~\ref{tbl_features}, which is
+sorted alphabetically, lists the variabilities as columns. \citet{Michalski2021}
+provides a more detailed analysis of the commonalities and variabilities between
+LBM packages.
 
@@
 
-The Domain Expert has a central role in verifying the list of LBM packages. In
-advance of the first meeting with the Domain Expert, we asked them to create a
-list of top software packages in the domain.  We did this to help the expert get
+The Domain Expert has a central role in verifying the list of packages. In
+advance of the first meeting with the Domain Expert, we ask them to create a
+list of top software packages in the domain.  We do this to help the expert get
 in the right mindset for our meeting.  Moreover, by doing the exercise in
 advance, we avoid the potential pitfall of the expert approving the discovered
-list of software without giving it adequate thought. We asked the Domain Expert,
-via virtual meetings, to vet the collected data and analysis. In particular, we
-asked them to vet the proposed list of software packages and the AHP ranking.
+list of software without giving it adequate thought. We ask the Domain Expert to
+vet the collected data and analysis. In particular, we ask them to vet the
+proposed list of software packages and the AHP ranking.
 

### 2022-04-25 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Revisions to Section 3

@@
 
-We use ten research questions to structure and focus our discussion. Our
+We use 10research questions to structure and focus our discussion. Our
 methodology \citep{SmithEtAl2021, SmithAndMichalski2022} is driven by these
@@
 compatible operating system versions. Ubuntu provided the testing platform for
-all but one of the packages. TCLB was tested on CentOS, as suggested by its
+all packages, except for TCLB, which we tested on CentOS, as suggested by its
 installation instructions.
@@
 All but one of the software packages (LatBo.jl) have automated at least part of
-the installation process. Most of these packages, such as waLBerla and
-SunlightLB, use Make to automate the installation, and a few of them, like
-lbmpy, use custom scripts.
-
-We quickly fixed errors encountered during the installation process, thanks to
-descriptive error messages. Systems that provided vague error messages, such as
+the installation process. Most packages, such as waLBerla and SunlightLB, use
+Make to automate the installation, and a few of them, like lbmpy, use custom
+scripts.
+
+In most of the cases with installation errors, descriptive error messages
+allowed us to proceed. Systems that provided vague error messages, such as
 messages that did not specify which action or file was at fault, were harder to
@@
 of the success or failure of the installation. In these instances, validating
-the installation required performing a tutorial or running a script, as
-described below, if these were available.
-
-About half of the installation instructions assume the installation environment
-has not already installed dependent packages. Unfortunately, software packages,
-like \esp, Ludwig and LUMA, frequently don't list all their dependencies.
-Sometimes only an error message during the installation process informs the user
-of the requirement for additional packages. In case like this, we suggest a
-detailed rewrite of the installation instructions to accommodate a clean
-operating system. A virtual machine provides such an environment for testing
-purposes.
-
-Seventeen software packages require installation of less than ten dependencies.
-All but one software package (LatBo.jl) require less than 20 dependencies. Some
-packages may automatically install additional dependencies in the background.
-Twenty of the software packages do not explicitly indicate the versions of
-software dependencies. Some software package installation issues could have been
-avoided by providing a list of dependencies. Sixteen software packages do not
-have detailed instructions for installing dependencies. Sixteen software
-packages have less than ten manual installation steps. If one command installs
-dependencies, then none of the software packages take more than 20 steps to
-install. The average number of steps is eight, and the fewest is two
+the installation required performing a tutorial or running a script, if these
+were available.
+
+About half of the installation instructions assume the user is already aware of,
+and has already installed, dependent packages. Many packages, like \esp, Ludwig
+and LUMA, do not list all their dependencies. Sometimes only an error message
+during the installation process informs the user of a required dependency. In
+cases like this, we suggest a detailed rewrite of the installation instructions
+to accommodate a clean operating system. A virtual machine can provide an
+environment for installability testing.
+
+Seventeen software packages require installation of less than 10 dependencies.
+All software package (except for LatBo.jl) require less than 20 dependencies.
+Some packages may automatically install additional dependencies in the
+background. Twenty of the software packages do not explicitly indicate the
+versions of software dependencies. Some software package installation issues
+could have been avoided by providing a list of dependencies. Sixteen software
+packages do not have detailed instructions for installing dependencies. Sixteen
+software packages have less than 10 manual installation steps. If one command
+installs dependencies, then none of the software packages take more than 20
+steps to install. The average number of steps is eight, and the fewest is two
 (LB3D-Prime).
@@
 packages. Software packages with a higher score (ESPResSo, MechSys, OpenLB,
-DL\_MESO, Palabos) tend to have one set of linear installation instructions
+DL\_MESO, Palabos) tend to have one set of linear installation instructions,
 written as if the person doing the installation has none of the dependencies
@@
 
-Tutorials are available for 19 of the software packages. Generally they are
-linearly written and easy to follow. However, only eight tutorials provide an
-expected output. It is not possible to verify the correctness of the output of
-the software packages that are missing this key information. In these cases the
-user may need to assume correctness if there are no visible errors.
+Tutorials are available for 19 of the packages. Generally they are linearly
+written and easy to follow. However, only eight tutorials provide expected
+outputs, making it difficult to verify the correctness of their output. In these
+cases the user may need to assume correctness if there are no visible errors.
 \href{https://www.walberla.net/doxygen/index.html} {waLBerla} provides a
@@
 
-The analysis of surface reliability focused on package installation and
+The analysis of surface reliability focuses on package installation and
 tutorials. Errors occurred when installing 16 of the software packages. Every
@@
 load dependencies. In these instances the installation tried to access outdated
-external repositories. We recovered and verified seven of the installations,
-and we assumed that one of the installations (LB3D-Prime) was recovered due to the
-absence of any way to verify otherwise. The installation of eight of the software
+external repositories. We recovered and verified seven of the installations, and
+we assumed that one of the installations (LB3D-Prime) was recovered, since there
+was no way to test the installation. The installation of eight of the software
 packages could not be recovered. Most of these broken installations could not
@@
 testing, but a descriptive error message helped in recovery. Nine of the top 10
-ranked packages are alive.
+packages are alive.
 
@@
 
-Testing includes handling unexpected input, such as incorrect data types, empty
-input, and missing files or links. A reasonable response, including appropriate
-error messages and an absence of unrecoverable system failures, means success.
+Robustness testing includes handling unexpected input, such as incorrect data
+types, empty input, and missing files or links. A reasonable response, including
+appropriate error messages and an absence of unrecoverable system failures,
+means success.
 
@@
 unexpected input as described above. All the software packages that installed
-correctly passed this test. They output descriptive error messages or at least
-did not crash. Software packages with a lower surface robustness score did not
+correctly passed this test. They did not crash and usually output descriptive
+error messages. Software packages with a lower surface robustness score did not
 install correctly, so their robustness score may not be a true reflection of
-runtime robustness. All successfully installed software packages that require a
-plain text input file correctly handled an unexpected change to these input
-files, including a replacement of new lines with carriage returns. Furthermore,
-nine of the top 10 ranked packages are alive.
+their runtime robustness. All successfully installed software packages that
+require a plain text input file correctly handled an unexpected change to these
+input files, including a replacement of new lines with carriage returns.
+Following the trend of the other qualities, nine of the top 10 ranked packages
+are alive.
 
@@
 Although the software packages all apply LBM to solve scientific computing
-problems, the packages focus on varied CFD problems, with varying parameters,
-and are technically different from each other (as shown in
-Table~\ref{tbl_features}). Due to this, a comparison of run-time performance is
-not appropriate. Instead, we looked through each software package's artifacts
+problems, they differ in scope and capabilities, as shown in
+Table~\ref{tbl_features}. Therefore, tests to compare run-time performance are
+not appropriate. Instead, we look through each software package's artifacts
 for evidence of considering performance. The artifacts of 18 software packages
@@
 achieve performance suitable for model optimization \citep{rutkowski2020open}.
-In the Ludwig package uses a so-called mixed mode approach implementing
+The Ludwig package uses a so-called mixed mode approach implementing
 fine-grained parallelism on the GPU.  MPI is used for even larger scale
@@
 
-We reviewed software package artifacts for the presence of a tutorial, a user
-manual, documented user characteristics, and a user support model. In total 19
-software packages have a tutorial, 13 have a user manual, and 11 have both. The
-tutorials vary in scope and substance, and eight include expected output. Most
-user manuals are in the form of a file that can be downloaded, while some are
+We review software repositories for the presence of a tutorial, a user manual,
+documented user characteristics, and a user support model. In total 19 software
+packages have a tutorial, 13 have a user manual, and 11 have both. The tutorials
+vary in scope and substance, and eight include expected output. Most user
+manuals are in the form of a file that can be downloaded, while some are
 rendered on a webpage. Some packages (waLBerla) do not have a user manual, but
@@
 user characteristics. Users are typically scientists or engineers. Their
-background is often physics, chemistry, biophysics, or mathematics. All but one
-of the packages (LIMBES) have a user support model, and many of them have
-multiple avenues of user support. The most popular avenue of support is Git,
-followed by email and forums. One software package (OpenLB) has an FAQ page.
+background is often physics, chemistry, biophysics, or mathematics. All projects
+(except LIMBES) have a user support model, and many of them have multiple
+avenues of user support. The most popular avenue of support is Git, followed by
+email and forums. One software package (OpenLB) has an FAQ page.
 
@@
 
-We reviewed software packages for the presence of artifacts and recorded every
-type of artifact that is not a code file. (The list of artifacts will be
-discussed further in Section~\ref{Sec_CompareArtifacts}.)  We also reviewed the
-software packages for software release and documentation version numbers. This
+We review software packages for the presence of artifacts and recorded every
+type of artifact that is not a code file. (The artifacts will be discussed
+further in Section~\ref{Sec_CompareArtifacts}.)  We also reviewed the software
+packages for software release and documentation version numbers. This
 information can be used to troubleshoot issues and organize documentation. All
@@
 
-When present, we noted how the team reviews code, and how users can contribute
-to the project. In total, 11 software packages have this information, which was
-found in various artifacts, including in developer guides, contributor guides,
-user guides, developer webpages, and README files.
+We look for evidence that teams review code and have information on how users
+can contribute. In total, 11 software packages have this information, found in
+various artifacts, including in developer guides, contributor guides, user
+guides, developer webpages, and README files.
 
@@
 of their issues closed, and only three (laboetie, lettuce, Sailfish) have less
-than 50 percent of their issues closed. All the top five overall ranked packages
-have most of their issues closed. Alive packages (11 use issue tracking) have
-64\% of their issues closed, while dead packages (3 use issue tracking) have
-71\% of their issues closed. Table~\ref{gitrepodata} presents this information.
-With respect to the host for the git repository, 13 packages use GitHub and two
-(Palabos, waLBerla) use GitLab. Of the other packages, one package (SunlightLB)
-uses CVS for issue tracking and version control, and seven packages do not
-appear to use any issue tracking system.
+than 50\% of their issues closed. All the top five ranked packages have most of
+their issues closed. Alive packages (11 use issue tracking) have 64\% of their
+issues closed, while dead packages (3 use issue tracking) have 71\% of their
+issues closed. Table~\ref{gitrepodata} presents this information. With respect
+to the host for the git repository, 13 packages use GitHub and two (Palabos,
+waLBerla) use GitLab. Of the other packages, one package (SunlightLB) uses CVS
+for issue tracking and version control, and seven do not appear to use any issue
+tracking system.
 
@@
 Table~\ref{gitrepodata} shows the percentage of code that is comments, in
-decreasing order. Maintainability improves for packages with a
-higher percentage of comments. Comments represent more than 10 percent of code
-files in 16 packages, and the average percentage of code comments is about 14
-percent. All the top five overall ranked packages have more than the average.
-LUMA has only 0.2 percent comments, the fewest of any package. This package has
-the most lines of source code, with over four million. The next largest package
-is \esp{} with one million.
+decreasing order. We assume maintainability improves for packages with a higher
+percentage of comments. Comments represent more than 10\% of code files in 16
+packages, and the average percentage of code comments is about 14\%. All the top
+five overall ranked packages have more than the average. LUMA has only 0.2\%
+comments, the fewest of any package. However, this package also has the most
+lines of source code, with over four million. The next largest package is \esp{}
+with one million.
 
@@
 packages also appear to manage their issue tracking, having most of their issues
-closed. Maintainable packages by our ranking have more than 10 percent of the
-code as comments. Four of the top five ranked packages are alive.
+closed. Maintainable packages by our ranking have more than 10\% of the code as
+comments. Four of the top five ranked packages are alive.
 
@@
 
-We measured the total number of source code files for each project. We assume
+We measure the total number of source code files for each project. We assume
 that numerous source files increases reusability, since more files implies
-increased modularization. Some packages have more features than others.  This is
-assumed to contribute to reusability, since they have more source code for
+increased modularization. Some packages have more features than others.  We
+assume this contributes to reusability, since they have more source code for
 potential reuse. We reviewed the software packages for the presence of API
@@
 non-LBM modelling. For this reason, a comparison with other software packages is
-not on a level field. However, these packages do have an abundance of reusable
-components. Nine of the top 10 ranked packages are alive.
+not on a level field. Nine of the top 10 ranked packages are alive.
 
@@
 reusability, and maintainability \citep{bauer2021walberla}. The design of
-waLBerla has enabled it to be successfully applied in several projects as a
-basis for various extensions \citep{bauer2021walberla}.
+waLBerla has enabled it to be successfully applied as a basis for extension in
+several projects \citep{bauer2021walberla}.
 
@@
 
-To measure understandability, we reviewed ten random source code files per
-package.  By using only 10 files we may incorrectly estimate understandability,
-but for practical consideration our time measuring each package has to be
-limited.
+To measure understandability, we review 10 random source code files per package.
+Using only 10 files may incorrectly estimate understandability, but for
+practical considerations we have to limit our time measuring each package.
 
@@
 LB3D-Prime, LIMBES, MP-LABS) appear to use vague identifiers, such as single
-letters for variables. Thirteen source code files used symbolic constants for
+letters for variables. Thirteen source code files use symbolic constants for
 various parameters, mathematical constants, and matrix definitions. The code
@@
 indicating what is being done (as opposed to how it is being done). The source
-code of 11 packages includes notes on domain algorithms. Table~\ref{moduledata}
+code of 11 packages include notes on domain algorithms. Table~\ref{moduledata}
 suggests that the software packages are modular to various degrees. When
@@
 
-We reviewed the software package artifacts for the identification of a specific
-development model, like a waterfall or agile development model, and the presence
-of documentation recording the development process and standards. They were also
-reviewed for the identification of the development environment, and the presence
-of release notes. The packages tended to not explicitly use well-known
-development models. This was also noted in the interviews with developers
-(Section~\ref{Sec_CompareMethodologies}). The development teams of these
-packages are fairly small and easily organized without the need for such
-processes. Seven of the software packages did have some artifacts outlining the
-general development process, how to contribute, and the status of the package or
-its components. Eight of the packages have artifacts that note the development
+We review the software artifacts for process documentation, a development
+environment description, and release notes.  The packages tend to not explicitly
+use well-known development models (see also
+Section~\ref{Sec_CompareMethodologies}). The development teams of these packages
+are fairly small and easily organized without the need for such processes. Seven
+of the software packages did have some artifacts outlining the general
+development process, how to contribute, and the status of the package or its
+components. Eight of the packages have artifacts that note the development
 environment. While this information could help developers, and would improve
@@
 explicit development model and defined development process. They also have
-detailed and easy to access notes accompanying software releases. Four of the
-top five ranked packages are alive.
+detailed and easy to access software release notes. Four of the top five ranked
+packages are alive.
 
@@
 
-Figure~\ref{Fig_OverallScore} shows the overall ranking of the software
-packages. In the absence of other information on priorities, we calculated the
-overall ranking by assuming an equal weighting between all qualities. If we
-weigh the qualities differently, the overall software package ranking will
-change. Software packages with an overall high score ranked high in at least
-several of the individual qualities that were quantitatively measured.
+Figure~\ref{Fig_OverallScore} shows the overall ranking. In the absence of other
+information on priorities, we calculated the ranking by assuming an equal
+weighting between all qualities. Software packages with an overall high score
+ranked high in at least several of the individual qualities.
 
@@
 			OpenLB & 4 & N/A & N/A & N/A& N/A\\
-			LUMA & 5 & 33 & 7 & 12& 4\\
-			pyLBM & 6 & 95 & 3 & 10& 5\\
+			LUMA & 5 & 33 & 7 & 12 & 4\\
+			pyLBM & 6 & 95 & 3 & 10 & 5\\
 			DL\_MESO (LBE) & 7 & N/A & N/A & N/A & N/A\\

### 2022-04-25 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Revisions to Section 4

@@
 To address \rqref{RQ_CompareHQ2Popular}, Table~\ref{repometrics} compares our
-LBM software package rankings against their popularity in the research
-community, as estimated by repository stars and watches. Nine packages do not
-use GitHub, so they do not have a measure of repository stars. Looking at the
-repository stars of the other 15 packages, we can observe a pattern where
-packages that have been highly ranked by our assessment tend to have more stars
-than lower ranked packages. Our best ranked package (ESPResSo) has the second
-most stars, while our ninth ranked package (Sailfish) has the highest number of
-stars. The repository watch column shows the same correlation, although
-this column contains less data, since two of the packages (Palabos, waLBerla)
-use GitLab, which does not track watches. Consistent with the star measure, our
-ninth ranked package (Sailfish) has the highest number of watches and our best
-ranked package (ESPResSo) has the second most watches. Packages designated as
-lower quality often do not use GitHub or GitLab, or have only a few
-stars and watches. Although the details of the rankings differ, our measures show
-that following best practices tends to correlate with increased popularity in
-the scientific community.
+LBM software rankings against their popularity in the research community, as
+estimated by repository stars and watches. Nine packages do not use GitHub, so
+they do not have a measure of repository stars. Looking at the repository stars
+of the other 15 packages, we observe a pattern where packages that have been
+highly ranked by our assessment tend to have more stars than lower ranked
+packages. Our best ranked package (ESPResSo) has the second most stars, while
+our ninth ranked package (Sailfish) has the highest number of stars. The
+repository watch column shows the same correlation, although this column
+contains less data, since two of the packages (Palabos, waLBerla) use GitLab,
+which does not track watches. Consistent with the star measure, our ninth ranked
+package (Sailfish) has the highest number of watches and our best ranked package
+(ESPResSo) has the second most watches. Packages designated as lower quality
+often do not use GitHub or GitLab, or have only a few stars and watches.
+Although the details of the rankings differ, our measures show that following
+best practices tends to correlate with increased popularity in the scientific
+community.
 
@@
 rankings may be due to a mismatch between the time we manually measured lettuce
-and when we counted the stars. The late addition of Musubi meant a large gap
-between when we measured projects and the time when our automated data (like
-stars) was collected. A final reason for inconsistencies between our ranking and
-the community's ranking is that, as for consumer products, more factors
-influence popularity than just quality.
+and when we counted the stars. The late addition of Musubi means a gap between
+when we measured projects and the time when our automated data (like stars) was
+collected. A final reason for inconsistencies between our ranking and the
+community's ranking is that, as for consumer products, more factors influence
+popularity than just quality.
 

### 2022-04-25 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Revisions to Section 6

@@
 As part of filling in the measurement template, we listed the artifacts for each
-package, and we then categorized them by frequency. We group the artifacts into
+package, and then categorized them by frequency. We group the artifacts into
 common, uncommon, and rare categories in Table~\ref{artifactspresent}. Common
@@
 we observed in LBM repositories to those observed and recommended for research
-software in general. Our comparison may point out areas where some LBM software
-packages fall short of current best practices. This is not intended to be a
-criticism of any existing packages, especially since in practice not every
-project needs to achieve the highest possible quality. However, rather than
-delve into the nuances of which software can justify compromising which
-practices, we will write our comparison under the ideal assumption that every
-project has sufficient resources to match best practices.
+software in general.
 
@@
 artifact consistently between guidelines and specific MI software, but the
-terminology and the contents of artifacts are not standardized.  The challenge
+terminology and the contents of artifacts are not standardized.  This challenge
 even exists for the ubiquitous README file.  As illustrated by
@@
 popularity of recommending a given artifact, but it does not imply that the most
-commonly recommended artifacts are the most important artifacts. Just because a
-guideline does not explicitly recommend an artifact does not mean that the
-artifact has no value.  The guideline authors may have excluded it because it is
-out of the scope of their recommendations or outside their experience. For
+commonly recommended artifacts are the most important artifacts. Moreover, just
+because a guideline does not explicitly recommend an artifact does not mean that
+the artifact has no value.  The guideline authors may have excluded it because
+it is out of the scope of their recommendations or outside their experience. For
 instance, an artifact related to uninstall is only explicitly mentioned by
@@
 
-The top four packages (ESPREsSo, Ludwig, Palabos, and OpenLB) have all the
-common artifacts, except Palabos does not have theory notes and only three
-(ESPREsSo, Ludwig, Palabos) of the packages appear to use a version control
-system.  The top four packages also have most of the uncommon artifacts. At the
-time of data collection, only one of the four packages (Palabos) did not have a
-user manual or guide, but there was a broken link on the package website
-indicating that such an artifact might exist. The developers later fixed this
-broken link, but this is not reflected in our data because it was not present at
-the time of data collection (mid-2020). Despite the broken link, Palabos does
-have a detailed and informative website. As far as we could observe, Ludwig does
-not appear to have a publicly visible design and OpenLB does not appear to use a
-version control system. However, OpenLB's use of version numbers suggests that
-its developers may privately use version control.
-
-The top four ranked packages do not have many of the rare artifacts. None of the
-top four packages have any explicit API documentation. Three of these packages
-(ESPREsSo, Ludwig, Palabos) have information on contributing to the project. Two
-of them (OpenLB, Palabos) have an FAQ section or forum. One (OpenLB) has
+The top four packages (ESPREsSo, Ludwig, Palabos, and OpenLB) include most of
+the common and uncommon artifacts, but not many of the rare artifacts.  They
+collectively have all the common artifacts, except Palabos does not have theory
+notes, and OpenLB does not appear to use version control.  The top four packages
+have most of the uncommon artifacts. Only one them (Ludwig) is missing design
+information, and only one (Palabos) does not have a user manual.  However, for
+Palabos there was a broken link on the package website indicating that such an
+artifact might exist. The developers later fixed this broken link, but this is
+not reflected in our data because it was not present at the time of data
+collection (mid-2020). Despite the broken link, Palabos does have a detailed and
+informative website. The top four ranked packages do not have many of the rare
+artifacts. None of them have any explicit API documentation. Three of them
+(ESPREsSo, Ludwig, Palabos) have information on contributing to the project and
+two (OpenLB, Palabos) have an FAQ section or forum. One (OpenLB) has
 verification and validation notes, and a video guide for the software.
@@
 highlighted publications and performance information because we were explicitly
-looking for both as part of filling in our measurement template. Related
+looking for both as part of our measurement template. Related
 publications and performance information certainly appear in other research
@@
 
-As Table~\ref{Tbl_Guidelines} shows, the LBM community participates in most of
-the practices we found listed in the general research software guidelines;
-however, we did not observe, or rarely observed, some recommended practices. For
+As Table~\ref{Tbl_Guidelines} shows, the LBM community participates in most
+practices we found listed in the general research software guidelines; however,
+we did not observe, or rarely observed, some recommended practices. For
 instance, we rarely saw API documentation for LBM software, but it is a
@@
 	\item estimate the product life cycle and evaluate the mix of planned
-development efforts. \end{inparaenum} \citep{VahaniittyEtAl2002}.  A product
+development efforts \end{inparaenum} \citep{VahaniittyEtAl2002}.  A product
 roadmap has the benefits of providing continuity of purpose, facilitating
@@
 \href{https://www.contributor-covenant.org/version/2/1/code_of_conduct/}
-{Contributor Covenant}, and \href{https://www.djangoproject.com/conduct/} {Django
-Code of Conduct}. \citet{SinghEtAl2021} shows the value of adopting a code of
-conduct on the participation of women. Improved inclusivity means a wider pool
-of potential contributors. Moreover, a code of conduct facilitates meeting
-ethical obligations for developers that abide by a code of ethics, such as the
-IEEE Code of Ethics \citep{IEEE1999}, or the Professional Engineers of Ontario
-code of ethics \citep[p.\ 23--24]{PEO2021}. For the LBM software, we did not
-observe any codes of conduct, possibly because the number of developers for each
-LBM project is small and because the idea has not yet occurred to any of the
-participants.
+{Contributor Covenant}, and \href{https://www.djangoproject.com/conduct/}
+{Django Code of Conduct}. \citet{SinghEtAl2021} shows the value of adopting a
+code of conduct on the participation of women. Improved inclusivity means a
+wider pool of potential contributors. Moreover, a code of conduct facilitates
+meeting ethical obligations for developers that abide by a code of ethics, such
+as the IEEE Code of Ethics \citep{IEEE1999}, or the Professional Engineers of
+Ontario code of ethics \citep[p.\ 23--24]{PEO2021}. For the LBM software, we did
+not observe any codes of conduct, possibly because the number of developers for
+each project is small and because the idea has not yet occurred to them.
 
@@
 \citep{ThielEtAl2020, SSI2022}.  For LBM software, ESPResSo has a checklist for
-managing releases, but we did not find any other examples.
+managing releases, but we did not see any other examples.
 
 \item \textbf{Uninstall instructions} tell a user how to remove the software and
-associated libraries from their system.  The advantage of uninstall is removing
-files that consume space and potentially pollute the users build system by
-causing conflicts when installing and using other software packages.  We did not
-observe uninstall instructions for any of the LBM projects.
+associated libraries from their system.  Uninstall removes files that consume
+space and potentially pollute the user's build system with potential future
+compilation and linking conflicts.  Only one software package (pyLBM) has
+uninstallation instructions.
 
@@
 new API developers mostly use code examples and tutorials \citep{MengEtAl2018}.
-API documentation was rare within our sample of LBM packages.
+API documentation is rare within our sample of LBM packages.
 
@@
 research software domains.  Compared to ocean modelling software
-\citep{JungEtAl2022}, LBM software pays more attention to artifacts related to
-testing.  Compared to Medical Imaging Software \citep{Dong2021}, the artifact
-frequency is similar, except LBM software is more likely to have developer
-related artifacts, like Contributing, Dependency list, and Design documentation.
+\citep{JungEtAl2022}, LBM software seems to pay more attention to artifacts
+related to testing.  Compared to Medical Imaging Software \citep{Dong2021}, the
+artifact frequency is similar, except LBM software is more likely to have
+developer related artifacts, like Contributing, Dependency list, and Design
+documentation.
 
@@
 alone in this.  Many, if not most, research projects fall short of best
-practices.   A gap exists in scientific computing development practices and
-software engineering recommendations \citep{Storer2017, Kelly2007,
-OwojaiyeEtAl2021_CSE}. \citet{JohansonAndHasselbring2018} observe that the
-state-of-the-practice for SCS in industry and academia does not incorporate
-state-of-the-art SE tools and methods.  This causes sustainability and
-reliability problems \citep{FaulkEtAl2009}. Rather than benefit from capturing
-and reusing previous knowledge, projects waste time and energy ``reinventing the
-wheel'' \citep{deSouzaEtAl2019}.
+practices.  A gap exists in research software development practices and software
+engineering recommendations \citep{Storer2017, Kelly2007, OwojaiyeEtAl2021_CSE}.
+\citet{JohansonAndHasselbring2018} observe that the state-of-the-practice for
+research software in industry and academia does not incorporate state-of-the-art
+SE tools and methods.  This causes sustainability and reliability problems
+\citep{FaulkEtAl2009}. Rather than benefit from capturing and reusing previous
+knowledge, projects waste time and energy ``reinventing the wheel''
+\citep{deSouzaEtAl2019}.
 
@@
 regulations in their field mandated such a document. \citet{Nguyen-HoanEtAl2010}
-showed requirements are the least commonly produced type of documentation for
+shows requirements are the least commonly produced type of documentation for
 research software in general. When looking at the pain points for research
-software developers, \citet{WieseEtAl2019} found that software requirements and
+software developers, \citet{WieseEtAl2019} state that software requirements and
 management is the software engineering discipline that most hurts scientific
@@
 perception that up-front requirements are impossible for research software
-\citep{CarverEtAl2007,SegalAndMorris2008}, but when we drop the insistence on
+\citep{CarverEtAl2007, SegalAndMorris2008}, but when we drop the insistence on
 ``up-front'' requirements, allowing the requirements to be written

### 2022-04-25 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Revisions to Section 7

@@
 for issue tracking and version control, and seven do not appear to use any issue
-tracking system.
+tracking system.  Assuming the use of GitHub, GitLab or CVS implies the use of
+version control, 16 of 24 (67\%) of projects use version control.
 
@@
 maintenance, and evolution of software, software processes, and artifacts
-\citep[p.\ 501]{GhezziEtAl2003}. Table~\ref{tbl_tools} summarizes the tools,
-subdivided into development tools, dependencies, and project management tools.
-In this section we answer \rqref{RQ_CompareToolsProjMngmnt} by comparing aspects
-of tool usage in LBM software packages to their use in the research software
-community in general.
+\citep[p.\ 501]{GhezziEtAl2003}. Table~\ref{tbl_tools} summarizes the observed
+tools, subdivided into development tools, dependencies, and project management
+tools. In this section we answer \rqref{RQ_CompareToolsProjMngmnt} by comparing
+aspects of tool usage in LBM software packages to their use in the research
+software community in general.
 
@@
 Table~\ref{tbl_tools}, developers also likely use debuggers.  Only three
-(ESPResSo, Ludwig and Musubi) packages mentioned CI tools, like Travis CI.
-Several packages explicitly noted code editors and compilers, but likely all
-packages use them. One of the packages (Ludwig) explicitly noted the use of
-proprietary unit testing code written in C. Likewise, one of the developers
-(pyLBM) noted the use of proprietary code for verifying the correctness of
-output. Developers of other packages likely used similar tools.
+(ESPResSo, Ludwig and Musubi) packages mention CI tools, like Travis CI. Several
+packages explicitly note code editors and compilers, but likely all packages use
+them. One of the packages (Ludwig) explicitly notes the use of proprietary unit
+testing code written in C. Likewise, one of the developers of pyLBM mentions the
+use of proprietary code for verifying the correctness of output. Developers of
+other packages likely use similar tools.
 
@@
 
-Many of the software packages that were assessed were developed by teams of two
-or more people. Their work needed to be coordinated and managed.
-Table~\ref{tbl_tools} shows the types of project management tools that were
-explicitly noted in the artifacts, webpages, or interviews with the developers.
-As with development tools and dependencies, developers may use other types of
-project management tools, but they were not visible in the artifacts to which we
-had access.  Collaboration tools, most often email and video, are noted as being
-used when developing software projects. Our interviewees did not mention project
-management software, but the sample is not large enough to rule out other
-projects using such software. GitHub is the home for many of the projects.
+Many of the software packages have teams of two or more. Their work needs to be
+coordinated and managed. Table~\ref{tbl_tools} shows the types of project
+management tools that were explicitly noted in the artifacts, webpages, or
+interviews with the developers. As with development tools and dependencies,
+developers may use other types of project management tools, but they were not
+visible in the artifacts to which we had access.  For collaboration tools the
+most often observed are email and video. Our interviewees did not mention
+project management software, but our sample is not large enough to rule out
+other projects using such software. GitHub is the home for many of the projects.
 Developers use the platform to help manage their projects, especially bug
@@
 
-Based on information provided by \citet{JungEtAl2022}, tool usage for
-LBM software has much in common with tool usage for ocean modelling
-software.  Both use tools for editing, compiling, code management, testing,
-building, and project management.  From the data available, ocean modelling
-differs from LBM software in its use of Kanban boards for project management.
-LBM software difference from the ocean domain through the use of continuous
-integration, and document generation tools.
+Based on information provided by \citet{JungEtAl2022}, tool usage for LBM
+software has much in common with tool usage for ocean modelling software.  Both
+use tools for editing, compiling, code management, testing, building, and
+project management.  From the data available, ocean modelling differs from LBM
+in its use of Kanban boards for project management. LBM software differs from
+the ocean modelling through the use of continuous integration, and document
+generation.
 
@@
 Section~\ref{Sec_Maintainability}, 67\% of LBM packages use version control
-(GitHub, GitLab, or CVS).  (From Table~\ref{gitrepodata}, 11/15 or 73\% of alive
-packages use version control.)  The proliferation of version control tools for
-LBM matches the increase in the broader research software community.
-\citet{Nguyen-HoanEtAl2010} estimated a little over 10 years ago that only 50\%
-of research software projects use version control, but even at that time
-\citet{Nguyen-HoanEtAl2010} noted an increase from previous usage levels. A
-survey in 2018 shows 81\% of developers use a version control system
-\citep{AlNoamanyAndBorghi2018}.  \citet{Smith2018} has similar results, showing
-version control usage for alive projects in mesh generation, geographic
-information systems and statistical software for psychiatry increasing from
-75\%, 89\% and 17\% (respectively) to 100\%, 95\% and 100\% (respectively) over
-a 4-year period ending in 2018.  (For completeness the same study showed a
-decrease in version control usage for seismology software over the same time
-period, from 41\% down to 36\%).  Almost every software guide cited in
-Section~\ref{Sec_CompareArtifacts} includes the advice to use version control.
-The high usage of version control tools in LBM software matches the trend in
-research software in general.
+(GitHub, GitLab, or CVS). If we restrict our attention to alive packages,
+Table~\ref{gitrepodata}, shows 11/15 or 73\% of alive packages use version
+control. The proliferation of version control tools for LBM matches the increase
+in the broader research software community. \citet{Nguyen-HoanEtAl2010}
+estimated a little over 10 years ago that only 50\% of research software
+projects use version control, but even at that time they noted an increase from
+previous usage levels. A survey in 2018 shows 81\% of developers use a version
+control system \citep{AlNoamanyAndBorghi2018}. \citet{Smith2018} has similar
+results, showing version control usage for alive projects in mesh generation,
+geographic information systems and statistical software for psychiatry
+increasing from 75\%, 89\% and 17\% (respectively) to 100\%, 95\% and 100\%
+(respectively) over a 4-year period ending in 2018.  (For completeness the same
+study showed a decrease in version control usage for seismology software over
+the same time period, from 41\% down to 36\%).  Almost every software guide
+cited in Section~\ref{Sec_CompareArtifacts} includes the advice to use version
+control. The high usage of version control tools in LBM software matches the
+trend in research software in general.
 
@@
 usage across a large sample of GitHub projects (34,544 projects) is at least
-40\%, with more popular projects (as measured by stars), having even higher
-usage rates. Moreover, \citet{HiltonEtAl2016} predict that CI adoption rates for
+40\%, with more popular projects (as measured by stars), having rates up to
+70\%. Moreover, \citet{HiltonEtAl2016} predict that CI adoption rates for
 open-source projects will increase even further in the future.  We could not
@@
 software domains likely lag behind the best practice recommendations, and
-non-research software usage rates, for continuous integration.
-
-CI is the process of building and testing software on every push to the code
-repository, with pushes done very frequently \citep[p.\ 13]
+non-research software usage rates, for CI.
+
+Continuous integration is the process of building and testing software on every
+push to the code repository, with pushes done very frequently \citep[p.\ 13]
 {HumbleAndFarley2010}, \citep{ShahinEtAl2017, Fowler2006}.  CI consists of the
@@
 	that LBM can be computationally expensive, the tests run with every check-in
-	may need to focus on simple code interface tests, saving larger more complex
-	tests for less frequent runs.  A more sophisticated option for when tests
-	create a bottleneck for merges is CIVET (Continuous Integration,
-	Verification, Enhancement, and Testing), which solves this problem by
-	intelligently pinning, cancelling, and if necessary, restarting jobs as
-	merges occur, so that two merges are being tested at all times
-	\citep{SlaughterEtAl2021}.
+	may need to focus on simple code interface tests, saving large tests for
+	less frequent runs.  A more sophisticated option to address the bottleneck
+	for merges is CIVET (Continuous Integration, Verification, Enhancement, and
+	Testing), which solves this problem by intelligently pinning, cancelling,
+	and if necessary, restarting jobs as merges occur \citep{SlaughterEtAl2021}.
 	\item An automated system for other tasks, such as code checking,
 	documentation building and web-site updating.  These other tasks are not
-	essential parts of CI, but they can be incorporated to improve the quality
+	essential to CI, but they can be incorporated to improve the quality
 	of the code and the communication between developers and users. For
@@
 \href{http://integrity.github.io/} {Integrity}. However, installation on your
-own server is not necessary since there are many hosted CI tool solutions
+own server is unnecessary since there are many hosted CI tool solutions
 available, such as: \href{https://travis-ci.org/} {Travis CI},
@@
 \href{https://circleci.com/} {CircleCI}.  Getting started with a hosted CI is
-straightforward.  All that is required to begin is editing a few lines of a
-simple YAML configuration file in the project's root directory.
+straightforward.  All that is required to begin is editing a few lines of a YAML
+configuration file in the project's root directory.
 
@@
 We answer research question \rqref{RQ_CompareMethodologies} by comparing the
-principles, processes, and methodologies used for LBM software to what can be
-gleaned from the literature on research software in general. Our measured data
-on LBM software artifacts does not explicitly indicate the development process.
-However, during our interviews one developer (ESPResSo) told us their
-non-rigorous development model is like a combination of agile and waterfall.
-Employing a loosely defined process makes sense for LBM software, given that the
-teams are generally small and self-contained.  Although eleven of the packages
-explicitly convey that they would accept outside contributors, teams are
-generally centralized, often working at the same institution. Working at the
-same institution means that an informal process can show success, since informal
-conversations are relatively easy to have.
-
-Interviews with developers confirmed a similar project management process. In
-teams of only a couple of developers, the entire team discusses additions of new
-features or major changes. Projects with more than a couple developers have lead
-developer roles. These lead developers review potential additions to the
-software. One of the developers (ESPResSo) noted that they use an ad hoc peer
-review process to assess major changes and additions. Using peer review (also
-called technical review) matches with recommended practice for research software
-\citep{HerouxEtAl2008, Givler2020, OrvizEtAl2017, USGS2019}.
+principles, processes, and methodologies used for developing LBM software to
+those used for research software in general. Our measured data on LBM software
+artifacts does not explicitly indicate the development process. However, during
+our interviews one developer (ESPResSo) told us their non-rigorous development
+model is like a combination of agile and waterfall. Employing a loosely defined
+process makes sense for LBM software, given that the teams are generally small
+and self-contained.  Although eleven of the packages explicitly convey that they
+would accept outside contributors, teams are generally centralized, often
+working at the same institution. Working at the same institution means that an
+informal process can show success, since informal conversations are relatively
+easy to have.
+
+Our interviews with LBM developers confirmed a similar project management
+process. In teams of only a couple of developers, the entire team discusses
+additions of new features or major changes. Projects with more than a couple
+developers have lead developer roles. These lead developers review potential
+additions to the software. One of the developers (ESPResSo) noted that they use
+an ad hoc peer review process to assess major changes and additions. Using peer
+review (also called technical review) matches with recommended practice for
+research software \citep{HerouxEtAl2008, Givler2020, OrvizEtAl2017, USGS2019}.
 
@@
 quality priorities for Sailfish roughly match the priorities observed for
-research software in general. \citet{Nguyen-HoanEtAl2010} surveyed developers to
+research software in general. \citet{Nguyen-HoanEtAl2010} surveys developers to
 find the following list of qualities, in decreasing order of importance:
@@
 the so-called simulation engine and interface.  This emphasis on documentation,
-especially for new developers, is echoed in research software guidelines.
-Multiple guidelines recommend a document explaining how to contribute to a
-project, often named CONTRIBUTING \citep{Yo2021, BrettEtAl2021, WilsonEtAl2016,
-ThielEtAl2020, vanGompelEtAl2016, OrvizEtAl2017, FLOSS2022, JimenezEtAl2017}.
-Tutorials \citep{ThielEtAl2020}, troubleshooting guides \citep{OrvizEtAl2017,
-SSI2022} and quick start examples \citep{ThielEtAl2020, vanGompelEtAl2016} are
-also recommended.  \citet{SmithAndRoscoe2018} suggests including instructions
-specifically for on-boarding new developers. For open source software in general
+especially for new developers, is echoed in research software guidelines
+summarized in Table~\ref{Tbl_Guidelines}. For open source software in general
 (not just research software), \citet{Fogel2005} recommends providing tutorial
@@
 	hierarchy by their secrets. The application of the Parnas approach to
-	scientific software has been illustrated by applying it to the example of a
+	research software has been illustrated by applying it to the example of a
 	mesh generator \citep{SmithAndYu2009}.
@@
 	\item Document the Module Interface Specification (MIS)
-	\citep{HoffmanAndStrooper1995}. An MIS is an abstract model that 
-	formally shows the state variables, inputs and outputs
+	\citep{HoffmanAndStrooper1995}. An MIS is an abstract model that formally
+	shows each module's access programs and the associated transitions and
+	outputs based on their state, environment, and input variables
 	\citep{ElSheikhEtAl2004, SmithAndYu2009}.
@@
 
-This section summarizes the best practices from LBM developers that are taken to
-address the pain points mentioned in Section~\ref{painpoints}.  The main source
-of information is the qualitative data from developer interviews
-(Section~\ref{SecSurvey}).  The practices summarized in this section can
-potentially be emulated by the LBM software packages that do not currently
-follow them.  Moreover, these practices may also provide examples that can be
-followed by other research software domains.
+This section answers \rqref{RQ_Concerns} by summarizing the best practices LBM
+developers take to address the pain points mentioned in
+Section~\ref{painpoints}.  The main source of information is the qualitative
+data from developer interviews (Section~\ref{SecSurvey}).  The practices
+summarized in this section can potentially be emulated by the LBM software
+packages that do not currently follow them.  Moreover, these practices may also
+provide examples that can be followed by other research software domains.
 

### 2022-04-25 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Revisions to Section 8

@@
 	\item[P\refstepcounter{pnum}\thepnum \label{P_LackDevTime}:] \textbf{Lack of
-	Development Time} A developer of pyLBM noted that their small development
+	Development Time:} A developer of pyLBM noted that their small development
 	team has a lack of time to implement new features. Small development teams
@@
 	\item[P\refstepcounter{pnum}\thepnum \label{P_LackSoftDevExp}:] \textbf{Lack
-	of Software Development Experience} The developer of TCLB noted a lack of
+	of Software Development Experience:} The developer of TCLB noted a lack of
 	software development experience, and others noted a need for improving
@@
 	not computer scientists or software engineers. \citet{Nguyen-HoanEtAl2010}
-	noted this same trend, with only 23\% of research software survey
-	respondents having a computing-related background. Similarly,
+	noted this same trend with research software developers, with only 23\% of
+	survey respondents having a computing-related background. Similarly,
 	\citet{UditAndKatz2017} show that the majority (54\%) of postdocs have not
@@
 	\item[P\refstepcounter{pnum}\thepnum \label{P_LackFunding}:] \textbf{Lack of
-	Incentive and Funding} The TCLB developer noted a lack of incentives and
+	Incentive and Funding:} The TCLB developer noted a lack of incentives and
 	funding in academia for developing widely used research software. This
@@
 	KaterbowAndFeulner2018}.  \citet{WieseEtAl2019} reported developer pains
-	related to publicity, since historically publishing norms make it difficult
-	to get credit for creating software.  As studied by
+	related to publicity, since publishing norms have historically made it
+	difficult to get credit for creating software.  As studied by
 	\citet{HowisonAndBullard2016}, research software (specifically biology
@@
 	\item[P\refstepcounter{pnum}\thepnum \label{P_LackExtSupport}:] \textbf{Lack
-	of External Support} Interviewees raised a concern that there are no
+	of External Support:} Interviewees raised a concern that there are no
 	organizations helping with the development of good quality software.  The
-	literature on research software does not tend to echo this concern because
+	literature on research software does not echo this concern because
 	there are such organizations, including \href{https://bssw.io/} {Better
@@
 	\item[P\refstepcounter{pnum}\thepnum \label{P_TechnologyHurdles}:]
-	\textbf{Technology Hurdles} Technology pain points for LBM developers
+	\textbf{Technology Hurdles:} Technology pain points for LBM developers
 	include setting up parallelization and CI. The pain
@@
 	\item[P\refstepcounter{pnum}\thepnum \label{P_Correctness}:]
-	\textbf{Ensuring Correctness} Several developers noted difficulties with
+	\textbf{Ensuring Correctness:} Several developers noted difficulties with
 	ensuring correctness. They alluded to difficulty with testing the
@@
 	Other research software domains point to the following problems with
-	testing: i) \citep{PintoEtAl2018} mention the problem of insufficient
-	testing; ii) the survey of \citep{HannayEtAl2009} shows that more developers
+	testing: i) \citet{PintoEtAl2018} mention the problem of insufficient
+	testing; ii) \citet{HannayEtAl2009} show that more developers
 	think testing is important than the number that believe they have a
-	sufficient understanding of testing concepts; and, iii) previous studies
-	\citep{HannayEtAl2009, KanewalaAndBieman2013, KellyEtAl2011, WieseEtAl2019}
+	sufficient understanding of testing concepts; and, iii) 
+	\citet{HannayEtAl2009, KanewalaAndBieman2013, KellyEtAl2011, WieseEtAl2019}
 	point to the oracle problem, which occurs in research software when we do
-	not have a means to judge the correctness of the calculated solutions.  The
-	LBM experience seems to overlap with i and ii, but the LBM developers did
-	not allude to problem iii (the oracle problem) in our conversations.  As
-	discussed in Section~\ref{SecReproducibility}, the oracle problem likely did
-	not come up because the LBM developers we interviewed have strategies for
-	verifying their work.
+	not have a means to judge the correctness of the calculated solutions. Based
+	on the interviews, the LBM experience seems to overlap with i and ii; the
+	developers did not allude to problem iii (the oracle problem).  As discussed
+	in Section~\ref{SecReproducibility}, the oracle problem likely did not come
+	up because the LBM developers have strategies for verifying their work.
 
 	\item[P\refstepcounter{pnum}\thepnum \label{P_Usability}:]
-	\textbf{Usability} Several developers noted that users sometimes try to use
+	\textbf{Usability:} Several developers noted that users sometimes try to use
 	incorrect LBM method combinations to solve their problems. Furthermore, some
@@
 	\item[P\refstepcounter{pnum}\thepnum \label{P_TechDebt}:] \textbf{Technical
-	Debt} The developer of ESPResSo said that they originally wrote their source
+	Debt:} The developer of ESPResSo said that they originally wrote their source
 	code with a specific application in mind, which later caused too much
-	coupling between components in the source code. This results in technical
+	coupling between components in the source code. This resulted in technical
 	debt \citep{KruchtenEtAl2012}, which has an impact on future maintainability
 	and reusability. Concern with technical debt is likely why researchers in
-	the survey of \citet{Nguyen-HoanEtAl2010} rated maintainability as the third
+	the survey of \citet{Nguyen-HoanEtAl2010} rate maintainability as the third
 	most important software quality. More recently the push for sustainable
@@
 	\item[P\refstepcounter{pnum}\thepnum \label{P_Documentation}:]
-	\textbf{Quality of Documentation}  Interviewees stressed the importance of
-	documentation for both users and developers throughout the interviews.
-	However, it was noted several times that a lack of time and funding
-	(\ppref{P_LackDevTime}) has a negative effect on the documentation. Most of
-	the developers are scientific researchers evaluated on the scientific papers
-	that they produce. Writing and updating documentation is something that is
-	done in their free time, if that time arises. Others also mention inadequate
-	research software documentation \citep{PintoEtAl2018, WieseEtAl2019}.  The
-	problem also arises with non-research software \citep{LethbridgeEtAl2003}.
-	Recommendations on the development of research software state that
-	developers should critically evaluate their own development processes in
-	terms of quality assurance and comply with international standards for
-	software documentation \citep{KaterbowAndFeulner2018}.
+	\textbf{Quality of Documentation:}  Interviewees stressed the importance of
+	documentation for both users and developers throughout the interviews. They
+	emphasize that a lack of time and funding (\ppref{P_LackDevTime}) has a
+	negative effect on the documentation. Most of the developers are scientific
+	researchers evaluated on the scientific papers that they produce. Writing
+	and updating documentation is something that is done in their free time, if
+	that time arises. Others also mention inadequate research software
+	documentation \citep{PintoEtAl2018, WieseEtAl2019}.  The problem also arises
+	with non-research software \citep{LethbridgeEtAl2003}. Recommendations on
+	the development of research software state that developers should critically
+	evaluate their own development processes in terms of quality assurance and
+	comply with international standards for software documentation
+	\citep{KaterbowAndFeulner2018}.
 

### 2022-04-25 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Revise section 9

@@
 avenues of user support. The most popular avenue of support is Git, followed by
-email and forums. One software package (OpenLB) has an FAQ page.
+email and forums. One software package (OpenLB) has an FAQ (Frequently Asked
+Questions) page.
 
@@
 
-\subsection{Documentation as Part of the Development Process}
+\subsection{Documentation as Part of the Development Process} \label{Sec_DocInDevProc}
 
@@
 	\citet{PintoEtAl2018}, where their replication of a previous study
-	\citep{HannayEtAl2009}, shows a growing interest in formal training (From
+	\citep{HannayEtAl2009}, shows a growing interest in formal training (from
 	13\% of respondents in 2009 to 22\% in 2018). \citet{PintoEtAl2018} found
@@
 would be difficult. For example, ``implementing a new model will be an easy
-contribution'', but changes to the ``Cartesian mesh … will be a nightmare''.
-The design of TCLB highlights that not every conceivable change needs to be
+contribution'', but changes to the ``Cartesian mesh … will be a nightmare''. The
+design of TCLB highlights that not every conceivable change needs to be
 supported, only the likely changes.
@@
 is unclear whether the developers are aware of Parnas's work, the approach
-mirrors his recommendations \citep{Parnas1972a}.
+mirrors his recommendations \citep{Parnas1972a}. Section~\ref{Sec_DocInDevProc}
+lists ideas for how to document the design, including the likely changes, so
+that they are more visible to others.
 
@@
 time on documentation provides a justifiable return on investment. For instance,
-the interviewed developer of ESPResSo noted that all major additions to their
+the interviewed developer of ESPResSo said that all major additions to their
 package had accompanying changes to artifacts and documentation. They put
@@
 	give more details on the packages that use generators.
-	\item Provide an explanation of theoretical principles through the user interface.
+	\item Provide an explanation of theoretical principles through the user
+	interface.
 	\item Add guards in the code to test user inputs for compatibility before
 	proceeding with the calculations.
-	\item Include documentation on FAQ (Section~\ref{Sec_CompareArtifacts}).
+	\item Include an FAQ (Section~\ref{Sec_CompareArtifacts}).
 \end{itemize}

### 2022-04-26 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Revisions to Section 10.

@@
 
-\subsection{CI Usage Compared to Research Software in General}
+\subsection{CI Usage Compared to Research Software in General} \label{Sec_CI}
 
@@
 conventional best practices. The ideas listed here have the potential to become
-best practices in the medium to long-term. The ideas below are roughly in the
-order of increasing novelty.
+best practices in the medium to long-term. The ideas in the following
+subsections are roughly in the order of increasing implementation effort.
 
@@
 Except for \citet{ThielEtAl2020}, the research software guidelines that we
-consulted do not mention linters.  However, we believe that linters have the potential
-to improve code quality at a relatively low cost.  A linter is a tool that
-statically analyzes code to find programming errors, stylistic inconsistencies
-and suspicious constructs \citep{Wikipedia2022_Lint}. Linters can be used to
-spot check code files, or even better as part of a continuous integration
-system, as discussed in Section~\ref{Sec_CompareTools}.
+consulted do not mention linters.  However, we believe that linters have the
+potential to improve code quality at a relatively low cost.  A linter is a tool
+that statically analyzes code to find programming errors, stylistic
+inconsistencies and suspicious constructs \citep{Wikipedia2022_Lint}. Linters
+can be used to spot check code files, or even better as part of a continuous
+integration system, as discussed in Section~\ref{Sec_CI}.
 
@@
 linter can include rules that capture the wisdom of senior programmers, it can
-help newer developers avoid common mistakes (\ppref{P_LackSoftDevExp}).
-Although there do not appear to be many linters for checking parallel
-programming mistakes (\ppref{P_TechnologyHurdles}), Parallel Lint is an option
-for OpenMP. Consistently using a linter can guard against some technical debt
+help newer developers avoid common mistakes (\ppref{P_LackSoftDevExp}). Although
+there are not many linters for checking parallel programming mistakes
+(\ppref{P_TechnologyHurdles}), Parallel Lint is an option for OpenMP.
+Consistently using a linter can guard against some technical debt
 (\ppref{P_TechDebt}). Although linters are tools for code analysis, similar
@@
 specific review tasks (like summarize the purpose of the code, create a data
-dictionary for the module assigned, cross-reference the code to the technical
+dictionary for a given module, cross-reference the code to the technical
 manual, etc.) The task based inspection approach has been effectively used for
@@
 (Section~\ref{Sec_Linters}): \ppref{P_LackDevTime}, \ppref{P_LackSoftDevExp},
-\ppref{P_TechnologyHurdles}, \ppref{P_TechDebt} and \ppref{P_Correctness}. Peer
-review addresses these pain points by efficiently searching for defects and
-problems.
+\ppref{P_TechnologyHurdles}, and \ppref{P_TechDebt}. Peer review addresses these
+points by efficiently searching for defects and problems.  Finding
+misunderstandings in of how the code implements the required theory improves the
+software's correctness (\ppref{P_Correctness}).
 
@@
 developers should watch for new opportunities to publish their research software
-source code. For instance, the Journal of Open Source Software (JOSS) reviews
-and publishes articles that credit the scholarship contained in the software
-itself \citep{SmithEtAl2018-joss}.  \citet{SmithEtAl2016-softcite} presents a
-set of software citation principles that may encourage broad adoption of a
-consistent policy for software citation across disciplines and venues.
+source code. For instance, the Journal of Open Source Software (JOSS)
+\citep{SmithEtAl2018-joss} reviews and publishes articles that credit the
+scholarship contained in the software itself.  \citet{SmithEtAl2016-softcite}
+presents a set of software citation principles that may encourage broad adoption
+of a consistent policy for software citation across disciplines and venues.
 \citet{ChueHongEtAl2019} provides a software citation checklist for developers
@@
 citation, since it is relatively easy to add metadata to repositories and to
-generate citations for those repositories \citet{Smith2021-CitationsOnGitHub}.
+generate citations for those repositories \citep{Smith2021-CitationsOnGitHub}.
 
@@
 circumvent the oracle problem for testing.  Additional techniques to build
-system tests are as follows \citep{Smith2016}:
+confidence in correctness are as follows \citep{Smith2016}:
 
@@
 
-As discussed in Section~\ref{Sec_CompareArtifacts}, developers rarely write requirements specifications for research software. Although a full requirements
-specification has advantages \citep{SmithEtAl2007, SmithAndLai2005},
-given the lack of development time (\ppref{P_LackDevTime}), we recommend the
-intermediate step of augmenting the existing theory manuals with some
-requirements information.  Table~\ref{artifactspresent} shows that theory
-documentation is common.  We observed 17 of 24 packages had at least some theory
-documentation. Templates for scientific requirements, like \citet{SmithEtAl2007,
+As discussed in Section~\ref{Sec_CompareArtifacts}, developers rarely write
+requirements specifications for research software. Although a full requirements
+specification has advantages \citep{SmithEtAl2007, SmithAndLai2005}, given the
+lack of development time (\ppref{P_LackDevTime}), we recommend the intermediate
+step of augmenting the existing theory manuals with some requirements
+information.  Table~\ref{artifactspresent} shows that theory documentation is
+common.  We observed 17 of 24 packages had at least some theory documentation.
+Templates for scientific requirements, like \citet{SmithEtAl2007,
 SmithAndLai2005}, show theory is a significant part of the requirements
@@
 show the consequences of changes to the assumptions). Spending time to
-incorporate additional information in the theory manuals should help address the
-quality of existing documentation (\ppref{P_Documentation})  and improve the
-software's usability (\ppref{P_Usability}).
+incorporate additional information in the theory manuals should improve the
+software's usability (\ppref{P_Usability}) and help address the quality of
+existing documentation (\ppref{P_Documentation}).
 
@@
 	\item[Repeatability] means producing the same output over successive runs of
-	a program.  As was pointed out during our developer interviews, this can be
+	a program.  As pointed out during our developer interviews, this can be
 	difficult for LBM software because of dependence on probability
@@
 just code.  \citet{SzymczakEtAl2016} presents initial work on this ``generate
-all things'' approach and \citep{CaretteEtAl2021-Drasil} provides a prototype.
+all things'' approach, \citet{SmithAndCarette2021-BRIC} presents a motivating
+example, and \citep{CaretteEtAl2021-Drasil} provides a prototype.
 
@@
 	\ppref{P_TechDebt}.	(Section~\ref{Sec_PeerReview})
-	\item Write and submit papers on software to address \ppref{P_LackFunding}.
-	(Section~\ref{Sec_MorePapers})
+	\item Write and submit papers on software, to journals like JOSS, to address
+	\ppref{P_LackFunding}. (Section~\ref{Sec_MorePapers})
 	\item Apply advice from the open-source community on how to grow the number

### 2022-04-26 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Revisions to Section 11.

@@
 
-\subsection{Surface Robustness}
+\subsection{Surface Robustness} \label{Sec_Robustness}
 
@@
 
-Our threats to validity categories come from an analysis of software engineering
-secondary studies by \citet{AmpatzoglouEtAl2019}.  We use the classification of
-\citet{AmpatzoglouEtAl2019} because studying a collection of software is a form
-of secondary study, where a secondary study analyzes the data from a set of
-primary studies.  A common example of a secondary study is a systematic
-literature review.  Our methodology is essentially a systematic software review
---- the primary studies are the software, and our work collects and analyzes
-these primary studies.
+Our categories for threats to validity come from an analysis of software
+engineering secondary studies by \citet{AmpatzoglouEtAl2019}, where a secondary
+study analyzes the data from a set of primary studies.  We use the
+classification of \citet{AmpatzoglouEtAl2019} because a common example of a
+secondary study is a systematic literature review. Our methodology is
+essentially a systematic software review --- the primary studies are the
+software packages, and our work collects and analyzes these primary studies.
 
@@
 researcher should achieve the same results, as long as they use the same
-methodology \citep{AmpatzoglouEtAl2019}.  For our study the identified
+methodology \citep{RunesonAndHost2009}.  For our study the identified
 reliability related threats include the following:
@@
 Construct validity means that adopted metrics represent what they are intended
-to measure \citep{AmpatzoglouEtAl2019}. The construct validity related threats
+to measure \citep{RunesonAndHost2009}. The construct validity related threats
 for the LBM study include the following:
@@
 \begin{itemize}
-\item Our approximation of popularity by stars and watches
-(Section~\ref{repmetrics}) may not be valid. 
-\item Our overall AHP ranking makes the unrealistic assumption of equal
-weighting between qualities (Section~\ref{Sec_OverallQuality}).
-\item The measure of surface robustness in only based on two pieces of data per
-project. The measurement of robustness could be expanded, as it currently only
-measures the impact of unexpected input. Other intentional faults could be
-introduced, but this would require a larger investment of measurement time. 
+\item We measure our software qualities indirectly, rather than directly.
+Unfortunately meaningful direct measures for qualities like maintainability,
+reusability and verifiability, are not available.  Therefore, we follow the
+usual approach of assuming that following procedures and adhering to standards
+yields a higher-quality product \citep[p.\ 112]{VanVliet2000}.
+\item The measure of surface robustness (Section~\ref{Sec_Robustness}) is only
+based on two pieces of data per project. The measurement of robustness could be
+expanded, as it currently only measures the impact of unexpected input. Other
+intentional faults could be introduced, but this would require a larger
+investment of measurement time. 
+\item We may have inaccurately estimated maintainability by assuming a higher
+ratio of comments to source code improves maintainability
+(Section~\ref{Sec_Maintainability}). 
 \item Reusability (Section~\ref{reusabilityresults}) is assessed via the number
@@
 or the source code may be vague and have ambiguous identifiers. 
-\item We may have inaccurately estimated maintainability by assuming a higher
-ratio of comments to source code improves maintainability
-(Section~\ref{Sec_Maintainability}). 
 \item The measurement of understandability
@@
 By random chance, the 10 files that were chosen may not be representative. 
-\item We measure our software qualities indirectly, rather than directly.
-Unfortunately meaningful direct measures for qualities like maintainability,
-reusability and verifiability, are not available.  Therefore, we follow the
-usual approach of assuming that following procedures and adhering to standards
-yields a higher-quality product \citep[p.\ 112]{VanVliet2000}.
+\item Our overall AHP ranking makes the unrealistic assumption of equal
+weighting between qualities (Section~\ref{Sec_OverallQuality}).
+\item Our approximation of popularity by stars and watches
+(Section~\ref{repmetrics}) may not be valid. 
 \end{itemize}
@@
 A study is internally valid if the discovered causal relations are trustworthy
-and cannot be explained by other factors \citep{AmpatzoglouEtAl2019}. The
+and cannot be explained by other factors \citep{RunesonAndHost2009}. The
 internal validity threats include the following:
@@
 \item A temporal gap exists between the manual measurement of Musubi and the
-measurement of all other packages (Section~\ref{identifysoftware}). This gap
-caused a mismatch between the time of manual and automated data collection
-(Section~\ref{repmetrics}), since we updated the automated measures (like the
-number of files, LOC, etc.) when Musubi was added.
-\item Our interviews were with a relatively small sample of four LBM developers.
-The pain points they expressed may not be representative of others in the
-community.
+measurement of all other packages (Section~\ref{filtersoftware}). This gap
+caused a mismatch between the time of manual and automated data collection,
+since we updated the automated measures (like the number of files, LOC, etc.)
+when Musubi was added.
+\item Our methodology assumes that artifacts in the repositories will show us
+the approach used for software development.  However, not all activities will
+necessarily leave a trace in the repositories. For instance, only three packages
+(ESPResSo, Ludwig, Musubi) show evidence in their artifacts of unit testing and
+CI (Section~\ref{SecSurfCorrectAndVerifiab}), even though interviews suggest a
+more frequent use of both unit testing and CI. For example, OpenLB, pyLBM, and
+TCLB use such methods during development despite this not being obvious from an
+analysis of the online material.
 \item In our manual measures of the software repositories, we may have missed
@@
 Section~\ref{Sec_CompareArtifacts}. 
-\item Our methodology assumes that artifacts will be present in the repositories
-to show the approach used for software development.  However, not all activities
-will necessarily leave a trace in the repositories. For instance, only three
-packages (ESPResSo, Ludwig, Musubi) show evidence in their artifacts of unit
-testing and CI (Section~\ref{SecSurfCorrectAndVerifiab}), even though interviews
-suggest a more frequent use of both unit testing and CI. For example, OpenLB,
-pyLBM, and TCLB use such methods during development despite this not being
-explicit from an analysis of the material available online.
+\item Our interviews were with a relatively small sample of four LBM developers.
+The pain points (Section~\ref{painpoints}) they expressed may not be
+representative of the rest of their community.
 \end{itemize}
@@
 External validity concerns whether the results of the study can be generalized
-(applied) to other situations/cases \citep{AmpatzoglouEtAl2019}.  The scope of
+(applied) to other situations/cases \citep{RunesonAndHost2009}.  The scope of
 our work is LBM solvers.  We are confident that our search was exhaustive and
-that we did not miss any highly popular examples.  Therefore, our validity
-concerns about LBM software observations are internal
+that we did not miss any highly popular examples.  Therefore, the bulk of our
+validity concerns about LBM software observations are internal
 (Section~\ref{Sec_InternalValidity}), not external.  However, our hope is that
@@
 \begin{itemize}
-\item The development of LBM software is fundamentally different from other
-research software.
+\item Generalization of our results will not be possible if the development of
+LBM software is fundamentally different from other research software.
 \end{itemize}

### 2022-04-29 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Anonymized SOP for LBM for double blind review

@@
 
-\journal{???}
+\journal{Archives of Computational Methods in Engineering}
 
@@
 
-\author[CAS]{Spencer Smith}
-\author[CAS]{Peter Michalski}
-\author[CAS]{Jacques Carette}
-\author[ME]{Zahra Keshavarz-Motamed}
-
-\affiliation[CAS]{organization={McMaster University, Computing and Software
-Department}, %Department and Organization
-            addressline={1280 Main Street West}, 
-            city={Hamilton},
-            postcode={L8S 4K1}, 
-            state={Ontario},
-            country={Canada}}
-
-\affiliation[ME]{organization={McMaster University, Mechanical Engineering}, %Department and Organization
-            addressline={1280 Main Street West}, 
-            city={Hamilton},
-            postcode={L8S 4K1}, 
-            state={Ontario},
-            country={Canada}}
+\author{Anonymous}
+% \author[CAS]{Spencer Smith}
+% \author[CAS]{Peter Michalski}
+% \author[CAS]{Jacques Carette}
+% \author[ME]{Zahra Keshavarz-Motamed}
+
+% \affiliation[CAS]{organization={McMaster University, Computing and Software
+% Department}, %Department and Organization
+%             addressline={1280 Main Street West}, 
+%             city={Hamilton},
+%             postcode={L8S 4K1}, 
+%             state={Ontario},
+%             country={Canada}}
+
+% \affiliation[ME]{organization={McMaster University, Mechanical Engineering}, %Department and Organization
+%             addressline={1280 Main Street West}, 
+%             city={Hamilton},
+%             postcode={L8S 4K1}, 
+%             state={Ontario},
+%             country={Canada}}
 
@@
 
-We use 10research questions to structure and focus our discussion. Our
-methodology \citep{SmithEtAl2021, SmithAndMichalski2022} is driven by these
-questions. For each question below, we point to the section that contains our
-answer.  We start with identifying the examples of LBM software where we have
-access to the source code:
+We use 10 research questions to structure and focus our discussion. These
+questions drive our methodology. 
+% \citep{SmithEtAl2021, SmithAndMichalski2022} % DOUBLE BLIND
+For each question below, we point to the section that contains our answer.  We
+start with identifying the examples of LBM software where we have access to the
+source code:
 
@@
 \citep{JungEtAl2022}.  However, most of the literature on the state of the
-practice does not focus on specific domains, but rather on all research software
-\citep{SmithAndMichalski2022}. We will refer to studies and guidelines that do
+practice does not focus on specific domains, but rather on all research software.
+%\citep{SmithAndMichalski2022}. % DOUBLE BLIND
+We will refer to studies and guidelines that do
 not distinguish an application domain as \emph{research software in general}. By
@@
 We investigate the research questions by applying the general methodology
-summarized in \citet{SmithEtAl2021, SmithAndMichalski2022}.  The specific
+summarized in Anonymous (xxxx)\footnote{Anonymous for double-blind
+review}.  %\citet{SmithEtAl2021, SmithAndMichalski2022}.  %DOUBLE BLIND
+The specific
 application of the methodology to LBM is reviewed in Section~\ref{methodology},
-with the full details in \citet{Michalski2021}.  Our methodology updates
-the approach used in prior assessments of domains like Geographic Information
+with the full details in Anonymous (xxxx)\footnotemark[1].  %\citet{Michalski2021}.  
+Our methodology updates
+an approach used in prior assessments of domains like Geographic Information
 Systems \citep{SmithEtAl2018_arXivGIS}, Mesh Generators \citep{SmithEtAl2016},
@@
 
-Compared with our previous methodology (as used in \citet{SmithEtAl2016} for
-instance), measuring the application of best practices now also includes
+Compared with our previous methodology (Anonymous, xxxx)\footnotemark[1],
+%(as used in \citet{SmithEtAl2016} for instance) %DOUBLE BLIND
+measuring the application of best practices now also includes
 repository based metrics, such as the number of files, number of lines of code,
@@
 We apply our methodology for evaluating the state of the practice of research
-software \citep{SmithEtAl2021, SmithAndMichalski2022} to the specific domain of
+software %\citep{SmithEtAl2021, SmithAndMichalski2022} % DOUBLE BLIND
+to the specific domain of
 LBM software.  Since we group best practices around conventional software
@@
 	\item Measure using the measurement template.  The full measurement template
-	can be found in \citet{SmithEtAl2021}. (Section~\ref{empiricalmeasures})
+	can be found in Anonymous (xxxx)\footnotemark[1]. %\citet{SmithEtAl2021}. %DOUBLE BLIND
+	(Section~\ref{empiricalmeasures})
 	\item Rank the software using AHP. (Section~\ref{AHP})
@@
 We use a measurement template to rank the projects by how well they follow best
-practices (\rqref{RQ_HighestQuality}), as described in \citet{SmithEtAl2021}.
+practices (\rqref{RQ_HighestQuality}), as described in Anonymous (xxxx)\footnotemark[1].
+% \citet{SmithEtAl2021}. % DOUBLE BLIND
 For each software package (each column in the template), we fill in the rows of
@@
 team as small as one, given our aim to cap the measurement phase at 160
-person-hours \citep{SmithEtAl2021}.  An excerpt of the template, in spreadsheet
+person-hours (Anonymous, xxxx)\footnotemark[1]. %\citep{SmithEtAl2021}. %DOUBLE BLIND
+An excerpt of the template, in spreadsheet
 form, is shown in Figure~\ref{measurement_template_image}.
@@
 
-As in \citet{SmithEtAl2016}, we used Virtual Machines (VMs) to provide an
+%As in \citet{SmithEtAl2016}, %DOUBLE BLIND
+We used Virtual Machines (VMs) to provide an
 optimal testing environment for each package. We used VMs to create fresh
@@
 measurement template, to compare and rank the LBM software. AHP performs a
-pairwise analysis using a matrix and generates an individual quality scores for
-each software package. \citet{SmithEtAl2016} shows the details on how AHP ranks
+pairwise analysis using a matrix and generates an individual quality score for
+each software package. Anonymous (xxxx)\footnotemark[1] %\citet{SmithEtAl2016} %DOUBLE BLIND
+shows the details on how AHP ranks
 software based on quality measures.
@@
 sensitivity analysis, we modified the score by 10\% for each package and
-verified that the overall ranking was stable.  The
-\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/AHP2020/LBM/README.txt}{README}
-file for the tool includes requirements and usage information.
+verified that the overall ranking was stable.  
+% The
+% \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/AHP2020/LBM/README.txt}{README}
+% file for the tool includes requirements and usage information. %DOUBLE BLIND
 
@@
 require going beyond the quantitative data from the measurement template. To
-gain insight, we interview developers using a list of 20 questions from
-\citet{SmithEtAl2021}. The questions cover the background of the development
+gain insight, we interview developers using a list of 20 questions from Anonymous (xxxx)\footnotemark[1].
+%\citet{SmithEtAl2021}. %DOUBLE BLIND
+The questions cover the background of the development
 teams, the interviewees, and the software itself. We ask the developers how they
@@
 our study. We met with them online, using either Zoom or Teams to facilitate
-recording and automatic transcription. Each meeting lasted about an hour. The
-McMaster University Research Ethics Board approved the interview process under
-application number
-\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/MACREM/Application.pdf}
-{MREB\#: 5219}.
+recording and automatic transcription. Each meeting lasted about an hour. 
+% The
+% McMaster University Research Ethics Board approved the interview process under
+% application number
+% \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/MACREM/Application.pdf}
+% {MREB\#: 5219}. %DOUBLE BLIND
 
@@
 qualities into one ``global'' ranking.  The full LBM data is available publicly
-on Mendeley Data \citep{Smith2022}.
+on Mendeley Data (Anonymous, xxxx)\footnotemark[1]. %\citep{Smith2022}. %DOUBLE BLIND
 
@@
 
-Table~\ref{Tbl_Guidelines} (based on data from \citep{SmithAndMichalski2022})
+Table~\ref{Tbl_Guidelines} %(based on data from \citep{SmithAndMichalski2022}) %DOUBLE BLIND
 shows that MI artifacts generally match the recommendations found in nine
@@
 Section~\ref{Sec_AddressConcerns} covers potential ways to address the pain
-points and \citet{SmithEtAl2021} provides the full interview questions.
+points and Anonymous (xxxx)\footnotemark[1] %\citet{SmithEtAl2021} 
+provides the full interview questions.
 
@@
 
-We investigated the risk of reviewer dependence in previous work
-\citep{SmithEtAl2016}, where we demonstrated that the measurement process is
+We investigated the risk of reviewer dependence in previous work (Anonymous, xxxx)\footnotemark[1],
+%\citep{SmithEtAl2016} % DOUBLE BLIND
+where we demonstrated that the measurement process is
 reasonably reproducible by having five products graded by a second reviewer. The
@@
 
-To assess the state of the practice, we followed five steps:
+To assess the state of the practice, we follow five steps:
 \begin{inparaenum}[i)]
-	\item Identified a list of 24 LBM packages;
-	\item Measured the packages using a measurement template (the template
+	\item Identify a list of software packages;
+	\item Measure the packages using a measurement template (the template
 	consists of 108 questions to assess 9 qualities (including the qualities of
 	installability, usability and visibility));
-	\item Ranked the software using the Analytic Hierarchy Process (AHP);
-	\item Interviewed developers (each interview consisted of 20 questions and took
+	\item Rank the software using the Analytic Hierarchy Process (AHP);
+	\item Interview developers (each interview consists of 20 questions and takes
 	about an hour); and, 
-	\item Conducted a domain analysis. 
+	\item Conduct a domain analysis. 
 \end{inparaenum}
-We analyzed the collected data by: 
+We analyze the collected data by: 
 \begin{inparaenum}[i)]
@@
 Ludwig, Musubi, pyLBM, waLBerla, laboetie, lettuce, \esp, LUMA, MP-LABS,
-Palabos, and HemeLB.  Therefore, two thirds (67\%) of the 24 measured packages
+Palabos, and HemeLB.  Therefore, two thirds (67\%) of the 24 packages
 appear in the top five for at least one quality.  Almost all entries are alive,
@@
 
-Although the state of the practice for LBM software is healthy, we have noted
-areas where the state of the practice does not match existing research software
-development guidelines, as follows:
+Although the state of the practice for LBM software is healthy, we note areas
+where the state does not match existing research software development
+guidelines, as follows:
 
 \begin{itemize}
-	\item Add roadmap documentation showing what is planned for the future to
-	provide continuity of purpose, facilitate stakeholder collaboration and
+	\item Add a roadmap document showing what is planned for the project's future
+	to provide continuity of purpose, facilitate stakeholder collaboration and
 	assist with prioritization. (Section~\ref{Sec_CompareArtifacts})
@@
 
-We found (Section~\ref{Sec_AddressConcerns}) that developers are currently
-creating practices that address these pain points.  To address technical debt
+We found that developers are currently creating practices that address these
+pain points (Section~\ref{Sec_AddressConcerns}).  To address technical debt
 (\ppref{P_TechDebt}) some developers have committed to designing for change
@@
 	\item Conduct rigorous peer reviews to address \ppref{P_LackDevTime},
-	\ppref{P_LackSoftDevExp}, \ppref{P_TechnologyHurdles} and
-	\ppref{P_TechDebt}.	(Section~\ref{Sec_PeerReview})
+	\ppref{P_LackSoftDevExp}, \ppref{P_TechnologyHurdles},
+	\ppref{P_Correctness}, and \ppref{P_TechDebt}.
+	(Section~\ref{Sec_PeerReview}) 
 	\item Write and submit papers on software, to journals like JOSS, to address
@@
 future we could also expand the interview questions to go beyond reproducibility
-to also discuss replicability. Furthermore, future interview guidelines should
-clarify the differences between usability and understandability
-(Section~\ref{softwarequalities}), since their answers suggest that not all
-participants understood the distinction.  Analysis of future interview data
-would likely benefit from a more rigorous process, like a thematic domain
-analysis \citep{JungEtAl2022}.
+to also discuss replicability (Section~\ref{Sec_Reproducibility}). Furthermore,
+future interview guidelines should clarify the differences between usability and
+understandability (Section~\ref{softwarequalities}), since their answers suggest
+that not all participants understood the distinction.  Analysis of future
+interview data would likely benefit from a more rigorous process, like a
+thematic domain analysis \citep{JungEtAl2022}.
 

### 2022-07-11 StateOfPractice/Papers/MedImageSoft_SOP-arXiv/MedImageSoft_SOP_AcademicRadiology.tex
Fixed an instance of passive voice
Updates to include CarverEtAl2022 and SmithAndMichalski2022

@@
 \usepackage{hyperref}
+\usepackage{xcolor}
+\hypersetup{
+    colorlinks,
+    linkcolor={red!50!black},
+    citecolor={blue!50!black},
+    urlcolor={blue!80!black}
+}
+
 %\usepackage{tikz}
@@
 To gain insights, we devised 10 research questions, which can be applied to MI,
-as well as to other domains, of research software \citep{SmithEtAl2021}.  The
-questions are designed to learn about the community's interest in, and
-experience with, artifacts, tools, principles, processes, methodologies and
-qualities.  When we mention artifacts we mean the documents, scripts and code
-that constitutes a software development project. Example artifacts include
-requirements, specifications, user manuals, unit tests, system tests, usability
-tests, build scripts, API (Application Programming Interface) documentation,
-READMEs, license documents, process documents, and code.  Once we have learned
-what MI developers do, we then put this information in context by contrasting MI
-software against the trends shown by developers in other research software
-communities.  Our aim is to collect enough information to understand the current
-pain points experienced by the MI software development community so that we can
-make some preliminary recommendations for future improvements.
+as well as to other domains, of research software \citep{SmithEtAl2021,
+SmithAndMichalski2022}.  The questions are designed to learn about the
+community's interest in, and experience with, artifacts, tools, principles,
+processes, methodologies and qualities.  When we mention artifacts we mean the
+documents, scripts and code that constitutes a software development project.
+Example artifacts include requirements, specifications, user manuals, unit
+tests, system tests, usability tests, build scripts, API (Application
+Programming Interface) documentation, READMEs, license documents, process
+documents, and code.  Once we have learned what MI developers do, we then put
+this information in context by contrasting MI software against the trends shown
+by developers in other research software communities.  Our aim is to collect
+enough information to understand the current pain points experienced by the MI
+software development community so that we can make some preliminary
+recommendations for future improvements.
 
@@
 	developers what specific best practices are taken to address the pain points
-	and software quality concerns? (Section~\ref{Sec_AddressConcerns})
+	and software quality concerns? (Section~\ref{painpoints})
 	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Recommend}:]
@@
 We designed a general methodology to assess the state of the practice for SC
-software \citep{SmithEtAl2021}. Details can be found in
+software \citep{SmithEtAl2021, SmithAndMichalski2022}. Details can be found in
 Section~\ref{ch_methods}.  Our methodology has been applied to MI software
@@
 We developed a methodology for evaluating the state of the practice of research
-software \citep{SmithEtAl2021}.  The methodology can be instantiated for a
-specific domain of scientific software, which in the current case is medical
-imaging software for visualization.  Our methodology involves and engages a
-domain expert partner throughout, as discussed in
+software \citep{SmithEtAl2021, SmithAndMichalski2022}.  The methodology can be
+instantiated for a specific domain of scientific software, which in the current
+case is medical imaging software for visualization.  Our methodology involves
+and engages a domain expert partner throughout, as discussed in
 Section~\ref{sec_vet_software_list}.  The four main steps of the methodology
@@
 \citep{Gieniusz2019} is used to measure the number of binary files as well as
-the number of added and deleted lines in a repository. This tool is also used to
+the number of added and deleted lines in a repository. We also use this tool to
 measure the number of commits over different intervals of time.
@@
 
-Our interviews were guided by a list of 20 questions, which can be found in
+We based uur interviews on a list of 20 questions, which can be found in
 \citet{SmithEtAl2021}. Some questions are about the background of the software,
@@
 501]{GhezziEtAl2003}. MI software uses tools for CI/CD, user support, version
-control, documentation and project management.  To answer
+control, documentation, and project management.  To answer
 \rqref{RQ_CompareToolsProjMngmnt} we summarize the tool usage in these
 categories, and compare this to the usage by the research software community.
-
-As mentioned in Section~\ref{sec_result_correctness_verifiability}, we
-identified five projects using CI/CD tools (about 17\% of the assessed
-projects). CI/CD using projects were found by examining the documentation and
-source code of all projects. The count of CI/CD usage may actually be higher,
-since traces of CI/CD usage may not always appear in a repository.  This was the
-case for a study of LBM software, where interviews with developers showed that
-more projects used CI/CD than was evident from repository artifacts alone
-\citep{Michalski2021}.  The 17\% utilization for MI software contrasts with the
-high frequency with which continuous integration is recommended in research
-software development guidelines \citep{BrettEtAl2021, Brown2015, ThielEtAl2020,
-Zadka2018, vanGompelEtAl2016}. Although there is currently little data available
-on CI/CD utilization for research software, our impression is that CI/CD is not
-yet common practice, despite its recommendation.  For LBM software at least the
-situation is similar to MI software, with only 12.5\% of 24 LBM packages showing
-evidence of CI/CD in their repositories \citep{Michalski2021}.  \wss{add
-reference to CarverEtAl2022, once it is published.}
 
@@
 matches the increase in the broader research software community.  A little over
-10 years ago version control was estimated to be used in only 50\% of research
-software projects \citep{Nguyen-HoanEtAl2010}, but even at that time
+10 years ago \citet{Nguyen-HoanEtAl2010} estimated that version control was used
+in only 50\% of research software projects, but even at that time
 \citet{Nguyen-HoanEtAl2010} noted an increase from previous usage levels. A
@@
 75\%, 89\% and 17\% (respectively) to 100\%, 95\% and 100\% (respectively) over
-a four year period ending in 2018. (For completeness the same study showed a
+a four-year period ending in 2018. (For completeness the same study showed a
 decrease in version control usage for seismology software over the same time
-period, from 41\% down to 36\%).  Almost every software guide cited in
-Section~\ref{Sec_CompareArtifacts} includes the advice to use version control.
-The high usage of version control tools in MI software matches the trend in
-research software in general.
+period, from 41\% down to 36\%).  A recent survey by \citet{CarverEtAl2022}
+shows version control use among practitioners at over 95\%, with 83/87 survey
+respondents indicating that they use it. All but one of the software guides
+cited in Section~\ref{Sec_CompareArtifacts} includes the advice to use version
+control. (The USGS guide \citep{USGS2019} was the only set of recommendations to
+not mention version control.) The high usage of version control tools in MI
+software matches the trend in research software in general.
+
+As mentioned in Section~\ref{sec_result_correctness_verifiability}, we
+identified five projects using CI/CD tools (about 17\% of the assessed
+projects). We found CI/CD using projects by examining the documentation and
+source code of all projects. The count of CI/CD usage may actually be higher,
+since traces of CI/CD usage may not always appear in a repository.  This was the
+case for a study of LBM software, where interviews with developers showed that
+more projects used CI/CD than was evident from repository artifacts alone
+\citep{Michalski2021}.  The 17\% utilization for MI software contrasts with the
+high frequency with which research software development guidelines recommend
+continuous integration \citep{BrettEtAl2021, Brown2015, ThielEtAl2020,
+Zadka2018, vanGompelEtAl2016}. Although there is currently little data available
+on CI/CD utilization for research software, our impression is that CI/CD is not
+yet common practice, despite its recommendation.  For LBM software at least the
+situation is similar to MI software, with only 12.5\% of 24 LBM packages showing
+evidence of CI/CD in their repositories \citep{Michalski2021}.  The survey of
+\citet{CarverEtAl2022} suggests higher use of CI/CD with 54\% (54/100)
+respondents indicating that they use it.
 
@@
 guidelines recommend a document explaining how to contribute to a project, often
-named CONTRIBUTING. Tutorials, user guides and quick start examples are also
-recommended. \citet{SmithAndRoscoe2018} suggests including instructions
+named CONTRIBUTING. Guidelins also recommend tutorials, user guides and quick start examples. \citet{SmithAndRoscoe2018} suggests including instructions
 specifically for on-boarding new developers. For open source software in general
 (not just research software), \citet{Fogel2005} recommends providing tutorial
-style examples, developer guidelines, demos and screenshots.
+style examples, developer guidelines, demos, and screenshots.
 
@@
 Based on interviews with nine developers (described in
-Section~\ref{sec_interview_methods}), we answer two research questions: i) What
-are the pain points for developers working on research software projects?
-(\rqref{RQ_PainPoints}); and, ii) How do the pain points of developers from MI
-compare to the pain points for research software in general?
-(\rqref{RQ_ComparePainPoints}).  We go through each of the identified pain
-points and include citations that contrast the MI experience with observations
-from researchers in other domains.  Potential ways to address the pain points
-are covered in Sections~\ref{Sec_AddressConcerns} and~\ref{ch_recommendations}.
-
-Table \ref{tab_obstacles} shows the number of times the interviewees mentioned
-the current and past obstacles in their projects.   We group these pain points
-into three major categories of obstacles: \textit{resource}, \textit{balance},
-and \textit{testing}. We put the less mentioned ones into the category
-\textit{Others}. The interviewees may have provided multiple
-answers to each question. Thus, when we summarize the frequency of different
-responses, the total is sometimes larger than nine.
-
-\begin{table}[!ht]
-\centering
-\begin{tabular}{llll}
-\toprule
-\multirow{2}{*}{Category} & \multirow{2}{*}{Obstacle} & \multicolumn{2}{c}{Num
- ans.} \\ \cline{3-4} &  & current & past \\ 
-\midrule
-\multirow{2}{*}{Resource} & Lack of fundings & 3 &  \\
- & Lack of time to devote to the project & 2 & 1 \\ 
-\midrule
-\multirow{3}{*}{Balance} & Hard to keep up with changes in OS and libraries & 1
-&  \\
- & Hard to support multiple OS & 2 &  \\
- & Hard to support lower-end computers & 1 & 2 \\ 
-\midrule
-Testing & Lack of access to real-world datasets for testing & 3 & 2 \\ 
-\midrule
-\multirow{7}{*}{Others}
- & Hard to have a high level roadmap from the start & 1 &  \\
- & Not enough participants for usability tests & 1 &  \\
- & Only a few people fully understand the large codebase & 1 &  \\
- & Hard to transfer to new technologies & & 2 \\
- & Hard to understand users' needs & & 1 \\
- & Hard to maintain good documentations & & 1 \\ 
-\bottomrule
-\end{tabular}
-\caption{\label{tab_obstacles}Current and past obstacles by the numbers of interviewees with the answers}
-\end{table}
+Section~\ref{sec_interview_methods}), we answer three research questions:
+\rqref{RQ_PainPoints}) What are the pain points for developers working on
+research software projects?; \rqref{RQ_ComparePainPoints}) How do the pain
+points of developers from MI compare to the pain points for research software in
+general?; and \rqref{RQ_Concerns}) For MI developers what specific best practices
+are taken to address the pain points and software quality concerns?  We go
+through each of the identified pain points and include citations that contrast
+the MI experience with observations from researchers in other domains.  We also
+cover potential ways to address the pain points, as promoted by the community in
+this section, and based on our experience in Section~\ref{ch_recommendations}.
+
+Ways to address pain points found by community as part of the qualitative data
+from developer interviews (Section~\ref{sec_interview_methods}).  The practices
+we summarize can potentially be emulated by the MI software packages that do not
+currently follow them.  Moreover, these practices may also provide examples that
+can be followed by other research software domains.
+
+\begin{itemize}
+\item Hard to have a high level roadmap from the start
+\item Not enough participants for usability tests
+\item Only a few people fully understand the large codebase
+\item Hard to transfer to new technologies
+\item Hard to understand users' needs
+\item Hard to maintain good documentations
+\end{itemize}
 
@@
 \item[P\refstepcounter{pnum}\thepnum \label{P_LackDevTime}:] \textbf{Lack of
-	Development Time} A developer of pyLBM noted that their small development
-	team has a lack of time to implement new features. Small development teams
-	are common for LBM software packages (as shown in the measurement table
-	excerpt in Figure~\ref{measurement_template_image}). Lack of time is also
-	highlighted as a pain point by other research software developers
-	\citet{PintoEtAl2018, PintoEtAl2016, WieseEtAl2019}.
-
-\end{enumerate}
-
-Developers identified a resource related pain point caused by a lack of funding
-and time.  Potential and proven solutions suggested by the interviewees include:
+Development Time:} A developer of pyLBM noted that their small development team
+has a lack of time to implement new features. Small development teams are common
+for LBM software packages (as shown in the measurement table excerpt in
+Figure~\ref{measurement_template_image}). Other domains of research software
+also experience the lack of time pain point \citep{PintoEtAl2018, PintoEtAl2016,
+WieseEtAl2019}. Developers identified a resource related pain point caused by a
+lack of funding and time.  Potential and proven solutions suggested by the
+interviewees include:
 
@@
 enough developers for building new features and fixing bugs at the same time;
-\item Licensing the software to commercial companies that integrate it into
-their products;
 \item Improving documentation to save time answering users' and developers'
@@
 
-With respect to balance, developers expressed difficulty balancing between four
+\item[P\refstepcounter{pnum}\thepnum \label{P_LackFunding}:] \textbf{Lack of
+Incentive and Funding:} The TCLB developer noted a lack of incentives and
+funding in academia for developing widely used research software. This problem
+has also been noted by others \citep{gewaltig2012quality, Goble2014,
+KaterbowAndFeulner2018}.  \citet{WieseEtAl2019} reported developer pains related
+to publicity, since publishing norms have historically made it difficult to get
+credit for creating software.  As studied by \citet{HowisonAndBullard2016},
+research software (specifically biology software, but the trend likely applies
+to other research software domains) is infrequently cited. \citet{PintoEtAl2018}
+also mentions the lack of formal reward system for research software.
+
+\begin{itemize}
+\item Licensing the software to commercial companies that integrate it into
+their products;
+\end{itemize}
+
+\item[P\refstepcounter{pnum}\thepnum \label{P_TechnologyHurdles}:]
+\textbf{Technology Hurdles:} Technology pain points for LBM developers include
+setting up parallelization and CI. The pain point survey of
+\citet{WieseEtAl2019} also highlighted technical-related problems like
+dependency management, cross-platform compatibility, CI, hardware issues and
+operating system issues. Things mentioned by developers in interviews: Hard to
+keep up with changes in OS and libraries, Hard to support multiple OS, and Hard
+to support lower-end computers. With respect to balance, developers expressed difficulty balancing between four
 factors: cross-platform compatibility, convenience to development \&
@@
 
-For the pain point category of testing developers identified the the lack of
-access to real-world datasets.  One pain point in the development process is the
-lack of access to real-world datasets for testing. The developers' strategies to
-address this are summarized in Section~\ref{Sec_AddressConcerns} \ad{to be
-fixed}. One threat to correctness is: with huge datasets for testing, the tests
-are expensive and time-consuming. Three interviewees endorsed self tests /
-automated tests, which may save time for testing. Some proposed and proven
-solutions are as follows:
+\item[P\refstepcounter{pnum}\thepnum \label{P_Correctness}:]
+\textbf{Ensuring Correctness:} Several developers noted difficulties with
+ensuring correctness. They alluded to difficulty with testing the correctness of
+large numbers of features, and challenges with automated testing. The TCLB
+developer commented that the amount of testing data needed was sometimes
+problematic, since free testing services do not offer adequate facilities for
+large amounts of data, requiring in-house testing solutions. Other research
+software domains point to the following problems with testing: i)
+\citet{PintoEtAl2018} mention the problem of insufficient testing; ii)
+\citet{HannayEtAl2009} show that more developers think testing is important than
+the number that believe they have a sufficient understanding of testing
+concepts; and, iii) \citet{HannayEtAl2009, KanewalaAndBieman2013, KellyEtAl2011,
+WieseEtAl2019} point to the oracle problem, which occurs in research software
+when we do not have a means to judge the correctness of the calculated
+solutions. Based on the interviews, the LBM experience seems to overlap with i
+and ii; the developers did not allude to problem iii (the oracle problem).  As
+discussed in Section~\ref{SecReproducibility}, the oracle problem likely did not
+come up because the LBM developers have strategies for verifying their work.
+
+For the pain point category of testing developers identified the lack of access
+to real-world datasets.  One pain point in the development process is the lack
+of access to real-world datasets for testing. The developers' strategies to
+address this are summarized in this section. One threat to correctness is: with
+huge datasets for testing, the tests are expensive and time-consuming. Three
+interviewees endorsed self tests / automated tests, which may save time for
+testing. Some proposed and proven solutions are as follows:
 
@@
 
-\subsection{Discussions on Software Qualities} \label{sec_interview_software_qualities}
-
-Questions 15--18, and 20 cover the software qualities of correctness,
-maintainability, understandability, usability, and reproducibility,
-respectively. We discuss each quality below.
-
-\begin{description}
-\item[Q15.] Was it hard to ensure the correctness of the software? If there were
-any obstacles, what methods have been considered or practiced to improve the
-situation? If practiced, did it work?
-\end{description}
-
 The interviewees identified multiple threats to correctness.  The most
@@
 
-\begin{description}
-\item[Q16.] When designing the software, did you consider the ease of future
-changes? For example, will it be hard to change the system’s structure, modules,
-or code blocks? What measures have been taken to ensure the ease of future
-changes and maintains?
-\end{description}
-
-The most popular, with five out of nine interviewees mentioning it, strategy for
-ensuring maintainability was to use a modular approach, with often repeated
-functions in a library.  Other strategies that were mentioned for improving
-maintainability include supporting third-party extensions, an easy-to-understand
-architecture, a dedicated architect, starting from simple solutions, and
-documentation.  The \textit{3D Slicer} team used a well-defined structure for
-the software, which they named as an ``event-driven MVC pattern''. Moreover,
-\textit{3D Slicer} discovers and loads necessary modules at runtime, according
-to the configuration and installed extensions. The \textit{BioImage Suite Web}
-team had designed and re-designed their software multiple times in the last 10+
-years. They found that their modular approach effectively supports
-maintainability \citep{Joshi2011}.
-
-\begin{description}
-\item[Q17.] Provide instances where users have misunderstood the software. What,
-if any, actions were taken to address understandability issues?
-\end{description}
+\item[P\refstepcounter{pnum}\thepnum \label{P_Usability}:]
+\textbf{Usability:} Several developers noted that users sometimes try to use
+incorrect LBM method combinations to solve their problems. Furthermore, some
+users think that the packages will work out of the box to solve their cases,
+while in reality CFD knowledge needs to be applied to correctly modify the
+packages for the new endeavour.  Some respondents in the survey of
+\citet{WieseEtAl2019} also mentioned that users do not always have the expertise
+required to install or use the software.
 
@@
 
-\begin{description}
-\item[Q18.] What, if any, actions were taken to address usability issues?
-\end{description}
-
 The two most frequently mentioned strategies (mentioned 3 times each) for
@@
 
-\begin{description}
-\item[Q20.] Do you have any concern that your computational results won't be
-reproducible in the future? Have you taken any steps to ensure reproducibility?
-\end{description}
+\end{enumerate}
+
+blurb to explain the ones that are coming were not explicitly identified as pain
+points.
+
+\begin{enumerate}
+\item \textbf{Maintainability:} The developer of ESPResSo said that they
+originally wrote their source code with a specific application in mind, which
+later caused too much coupling between components in the source code. This
+resulted in technical debt \citep{KruchtenEtAl2012}, which has an impact on
+future maintainability and reusability. Concern with technical debt is likely
+why researchers in the survey of \citet{Nguyen-HoanEtAl2010} rate
+maintainability as the third most important software quality. More recently the
+push for sustainable software \citep{deSouzaEtAl2019} is motivated by the pain
+that past developers have had with accumulating too much technical debt.
+
+The most popular, with five out of nine interviewees mentioning it, strategy for
+ensuring maintainability was to use a modular approach, with often repeated
+functions in a library.  Other strategies that were mentioned for improving
+maintainability include supporting third-party extensions, an easy-to-understand
+architecture, a dedicated architect, starting from simple solutions, and
+documentation.  The \textit{3D Slicer} team used a well-defined structure for
+the software, which they named as an ``event-driven MVC pattern''. Moreover,
+\textit{3D Slicer} discovers and loads necessary modules at runtime, according
+to the configuration and installed extensions. The \textit{BioImage Suite Web}
+team had designed and re-designed their software multiple times in the last 10+
+years. They found that their modular approach effectively supports
+maintainability \citep{Joshi2011}.
+
+\item \textbf{Reproducibility:}  Interviewees stressed the importance of
+documentation for both users and developers throughout the interviews. They
+emphasize that a lack of time and funding (\ppref{P_LackDevTime}) has a negative
+effect on the documentation. Most of the developers are scientific researchers
+evaluated on the scientific papers that they produce. Writing and updating
+documentation is something that is done in their free time, if that time arises.
+Others also mention inadequate research software documentation
+\citep{PintoEtAl2018, WieseEtAl2019}.  The problem also arises with non-research
+software \citep{LethbridgeEtAl2003}. Recommendations on the development of
+research software state that developers should critically evaluate their own
+development processes in terms of quality assurance and comply with
+international standards for software documentation
+\citep{KaterbowAndFeulner2018}.
 
@@
 
-\section{Lessons from MI Developers} \label{Sec_AddressConcerns}
-
-The best practices from MI developers are taken to address the pain points
-mentioned in Section~\ref{painpoints}.  We found the practices as part of the
-qualitative data from developer interviews
-(Section~\ref{sec_interview_methods}).  The practices we summarize can
-potentially be emulated by the MI software packages that do not currently follow
-them.  Moreover, these practices may also provide examples that can be followed
-by other research software domains.
+\end{enumerate}
 
 \section{Recommendations} \label{ch_recommendations}
+
+- add in ACs, borrow from previous paper.
 
@@
 methodology for assessing the state of the practice for SC software
-\citep{SmithEtAl2021}. We apply this process to Medical Imaging (MI) software
-that belongs to a specific domain of SC.
+\citep{SmithEtAl2021, SmithAndMichalski2022}. We apply this process to Medical
+Imaging (MI) software that belongs to a specific domain of SC.
 
@@
 
+- CarverEtAl2022 has discussion of requirements etc. that we could also contrast (seems that people overestimate their use of requirements.)
 - add more on comparison to existing literature surveys

### 2022-07-12 StateOfPractice/Papers/MedImageSoft_SOP-arXiv/MedImageSoft_SOP_AcademicRadiology.tex
Edits to introduction to pain points section

@@
 points of developers from MI compare to the pain points for research software in
-general?; and \rqref{RQ_Concerns}) For MI developers what specific best practices
-are taken to address the pain points and software quality concerns?  We go
-through each of the identified pain points and include citations that contrast
-the MI experience with observations from researchers in other domains.  We also
-cover potential ways to address the pain points, as promoted by the community in
-this section, and based on our experience in Section~\ref{ch_recommendations}.
-
-Ways to address pain points found by community as part of the qualitative data
-from developer interviews (Section~\ref{sec_interview_methods}).  The practices
-we summarize can potentially be emulated by the MI software packages that do not
-currently follow them.  Moreover, these practices may also provide examples that
-can be followed by other research software domains.
-
-\begin{itemize}
-\item Hard to have a high level roadmap from the start
-\item Not enough participants for usability tests
-\item Only a few people fully understand the large codebase
-\item Hard to transfer to new technologies
-\item Hard to understand users' needs
-\item Hard to maintain good documentations
-\end{itemize}
+general?; and \rqref{RQ_Concerns}) For MI developers what specific best
+practices are taken to address the pain points and software quality concerns?
+The identified pain points relate to lack of time and funding, technology
+hurdles, improving correctness, and improving usability.  We go through each
+pain points and include citations that contrast the MI experience with
+observations from other domains.  We also cover potential ways to address the
+pain points, as promoted by the community in this section, and based on our
+experience in Section~\ref{ch_recommendations}.  In this section we also discuss
+some MI developer strategies for improving maintainability and reproducibility.
+These two qualities were not explicitly idenfied as pain points, but ways to
+improve them were still part of our interview discussions. The MI interviewee
+practices to address pain points and improve quality can potentially be emulated
+by other the MI software packages.  Moreover, these practices may provide
+examples that can be followed by other research software domains.
 
 \citet{PintoEtAl2018} lists some pain points that did not come up in our
-conversations with LBM developers: Cross-platform compatibility, interruptions
+conversations with LBM developers [CHECK  THESE]: Cross-platform compatibility, interruptions
 while coding, scope bloat, lack of user feedback, hard to collaborate on
@@
 WieseEtAl2019}. Developers identified a resource related pain point caused by a
-lack of funding and time.  Potential and proven solutions suggested by the
+lack of funding and time.
+
+Potential and proven solutions suggested by the
 interviewees include:
@@
 maintenance, performance, and security.  The potential and proven solutions are:
+
+% Hard to transfer to new technologies %from Ao
 
@@
 
+% Hard to maintain good documentations % from Ao
+% \begin{itemize}
+%     \item Hard to have a high level roadmap from the start
+%     \item Only a few people fully understand the large codebase
+% \end{itemize}
+
 For the pain point category of testing developers identified the lack of access
@@
 required to install or use the software.
+
+%\item Not enough participants for usability tests %from Ao
+%Hard to understand users' needs % from Ao
 

### 2022-07-18 StateOfPractice/Papers/MedImageSoft_SOP-arXiv/MedImageSoft_SOP_AcademicRadiology.tex
Edits to pain points section, specifically P1 and P2

@@
 Based on interviews with nine developers (described in
-Section~\ref{sec_interview_methods}), we answer three research questions:
-\rqref{RQ_PainPoints}) What are the pain points for developers working on
-research software projects?; \rqref{RQ_ComparePainPoints}) How do the pain
-points of developers from MI compare to the pain points for research software in
-general?; and \rqref{RQ_Concerns}) For MI developers what specific best
-practices are taken to address the pain points and software quality concerns?
-The identified pain points relate to lack of time and funding, technology
-hurdles, improving correctness, and improving usability.  We go through each
-pain points and include citations that contrast the MI experience with
+Section~\ref{sec_interview_methods}), we answer three research questions (first
+mentioned in Section~\ref{sec_motivation}): \rqref{RQ_PainPoints}) What are the
+pain points for developers working on research software projects?;
+\rqref{RQ_ComparePainPoints}) How do the pain points of developers from MI
+compare to the pain points for research software in general?; and
+\rqref{RQ_Concerns}) For MI developers what specific best practices are taken to
+address the pain points and software quality concerns?
+
+Our interviews identify pain points that relate to lack of time and funding,
+technology hurdles, improving correctness, and improving usability.  In this
+section, we go through each pain point and contrast the MI experience with
 observations from other domains.  We also cover potential ways to address the
-pain points, as promoted by the community in this section, and based on our
-experience in Section~\ref{ch_recommendations}.  In this section we also discuss
-some MI developer strategies for improving maintainability and reproducibility.
-These two qualities were not explicitly idenfied as pain points, but ways to
-improve them were still part of our interview discussions. The MI interviewee
-practices to address pain points and improve quality can potentially be emulated
-by other the MI software packages.  Moreover, these practices may provide
-examples that can be followed by other research software domains.
+pain points, as promoted by the community.  (Later, in
+Section~\ref{ch_recommendations}, we propose additional pain mitigation
+strategies based on our experience.)  In addition to pain points, we summarize
+MI developer strategies for improving maintainability and reproducibility.
+Although these two qualities were not explicitly idenfied as pain points, they
+were discussed during our interviews.  The interviewee's practices for
+addressing pain points and improving quality can potentially be emulated by
+other MI developers. Moreover, these practices may provide examples that can be
+followed by other research software domains.
 
 \citet{PintoEtAl2018} lists some pain points that did not come up in our
-conversations with LBM developers [CHECK  THESE]: Cross-platform compatibility, interruptions
-while coding, scope bloat, lack of user feedback, hard to collaborate on
-software projects, and aloneness. \citet{WieseEtAl2019} repeat some of the
-previous pain points and add the following: dependency management, data handling
-concerns (like data quality, data management and data privacy), reproducibility,
-and software scope determination. Although LBM developers did not mention these
-pain points, we cannot conclude that they are not relevant for LBM software
-development, since we only interviewed four LBM developers for about an hour each.
+conversations with MI developers [CHECK  THESE, after writing all the PPs]:
+Cross-platform compatibility, interruptions while coding, scope bloat, lack of
+user feedback, hard to collaborate on software projects, and aloneness.
+\citet{WieseEtAl2019} repeat some previous pain points and add the following:
+dependency management, data handling concerns (like data quality, data
+management and data privacy) [Actually mentioned by MI software],
+reproducibility, and software scope determination. Although LBM developers did
+not mention these pain points, we cannot conclude that they are not relevant for
+LBM software development, since we only interviewed four LBM developers for
+about an hour each.
 
@@
 \item[P\refstepcounter{pnum}\thepnum \label{P_LackDevTime}:] \textbf{Lack of
-Development Time:} A developer of pyLBM noted that their small development team
-has a lack of time to implement new features. Small development teams are common
-for LBM software packages (as shown in the measurement table excerpt in
-Figure~\ref{measurement_template_image}). Other domains of research software
-also experience the lack of time pain point \citep{PintoEtAl2018, PintoEtAl2016,
-WieseEtAl2019}. Developers identified a resource related pain point caused by a
-lack of funding and time.
-
-Potential and proven solutions suggested by the
-interviewees include:
+Development Time:} Many interviewees thought lack of time and lack of funding
+(discussed next) were their most significant obstacles. Other domains of
+research software also experience the lack of time pain point
+\citep{PintoEtAl2018, PintoEtAl2016, WieseEtAl2019}. Our study of LBM software
+\citep{SmithEtAl2022} also highlighted lack of time as a frequently mentioned
+pain point.
+
+Potential and proven solutions suggested by the interviewees include:
 
@@
 \item Supporting third-party plugins and extensions; and,
-\item Using GitHub Actions for CI/CD to save time.
-\end{itemize}
-
-Many interviewees thought lack of fundings and lack of time were their most
-significant obstacles. The interviewees from \textit{3D Slicer} team and
-\textit{OHIF} team pointed out that it was more challenging to get fundings for
-software maintenance as opposed to research. The interviewee from the
-\textit{ITK-SNAP} team thought more fundings was a way to solve the lack of time
-problem, because they could hire more dedicated developers. On the other hand,
-the interviewee from the \textit{Weasis} team did not feel that fundings could
-solve the same problem, since he still would need a lot of time to supervise the
-project.  No interviewee suggested any solution to bring extra funding to the
-project. However, they provided ideas to save time, such as better
-documentation, third-party plugins, and good CI/CD tools.
+\item Using GitHub Actions for continunous integration and continuous
+development.
+\end{itemize}
 
 \item[P\refstepcounter{pnum}\thepnum \label{P_LackFunding}:] \textbf{Lack of
-Incentive and Funding:} The TCLB developer noted a lack of incentives and
-funding in academia for developing widely used research software. This problem
-has also been noted by others \citep{gewaltig2012quality, Goble2014,
-KaterbowAndFeulner2018}.  \citet{WieseEtAl2019} reported developer pains related
-to publicity, since publishing norms have historically made it difficult to get
-credit for creating software.  As studied by \citet{HowisonAndBullard2016},
-research software (specifically biology software, but the trend likely applies
-to other research software domains) is infrequently cited. \citet{PintoEtAl2018}
-also mentions the lack of formal reward system for research software.
-
-\begin{itemize}
-\item Licensing the software to commercial companies that integrate it into
-their products;
-\end{itemize}
+Incentive and Funding:} As mentioned above, developers felt the pain of having
+to attract funding the development and maintenance of their software projects.
+For instance, the interviewees from \textit{3D Slicer} and \textit{OHIF} pointed
+out that it was more challenging to get fundings for software maintenance as
+opposed to research. The interviewee from the \textit{ITK-SNAP} team thought
+more fundings was a way to solve the lack of time problem, because they could
+hire more dedicated developers. On the other hand, the interviewee from the
+\textit{Weasis} team did not feel that fundings could solve the same problem,
+since he still would need a lot of time to supervise the project.
+
+Funding challenges have also been noted by others \citep{GewaltigAndCannon2012,
+Goble2014, KaterbowAndFeulner2018, SmithEtAl2022}. \citet{WieseEtAl2019}
+reported developer pains related to publicity, since publishing norms have
+historically made it difficult to get credit for creating software.  As studied
+by \citet{HowisonAndBullard2016}, research software (specifically biology
+software, but the trend likely applies to other research software domains) is
+infrequently cited. \citet{PintoEtAl2018} also mentions the lack of formal
+reward system for research software.
+
+An interviewee proposed an idea for increasing funding: Licensing the software
+to commercial companies that integrate it into their products.
 
@@
 
+%\citet{WieseEtAl2019} mention data handling concerns (like data quality, data management and data privacy)
+
 For the pain point category of testing developers identified the lack of access
@@
 - add in ACs, borrow from previous paper.
+- mention the solutions proposed by LBM developers, like design for change
 

### 2022-07-21 StateOfPractice/Papers/MedImageSoft_SOP-arXiv/MedImageSoft_SOP_AcademicRadiology.tex
Further edits to pain points section

@@
 quantitative data in the grading template, we rank the software with the
-Analytic Hierarchy Process (AHP) (Details are found in
-Section~\ref{ch_background}). After this, as another addition to our previous
-methodology, we interview some of the development teams to further understand
-the status of their development process.
+Analytic Hierarchy Process (AHP) (Section~\ref{ch_background} provides details).
+After this, as another addition to our previous methodology, we interview some
+development teams to further understand the status of their development process.
 
@@
 
-\citet{PintoEtAl2018} lists some pain points that did not come up in our
-conversations with MI developers [CHECK  THESE, after writing all the PPs]:
+[CHECK  THESE, after writing all the PPs] \citet{PintoEtAl2018} lists some pain
+points that did not come up in our conversations with MI developers :
 Cross-platform compatibility, interruptions while coding, scope bloat, lack of
@@
 \citep{PintoEtAl2018, PintoEtAl2016, WieseEtAl2019}. Our study of LBM software
-\citep{SmithEtAl2022} also highlighted lack of time as a frequently mentioned
-pain point.
+\citep{SmithEtAl2022} also highlighted lack of time as a significant pain point.
 
@@
 \item[P\refstepcounter{pnum}\thepnum \label{P_LackFunding}:] \textbf{Lack of
-Incentive and Funding:} As mentioned above, developers felt the pain of having
-to attract funding the development and maintenance of their software projects.
-For instance, the interviewees from \textit{3D Slicer} and \textit{OHIF} pointed
-out that it was more challenging to get fundings for software maintenance as
-opposed to research. The interviewee from the \textit{ITK-SNAP} team thought
-more fundings was a way to solve the lack of time problem, because they could
-hire more dedicated developers. On the other hand, the interviewee from the
-\textit{Weasis} team did not feel that fundings could solve the same problem,
-since he still would need a lot of time to supervise the project.
+Funding:} As mentioned above, developers felt the pain of having to attract
+funding to develop and maintain their software. For instance, the interviewees
+from \textit{3D Slicer} and \textit{OHIF} said getting funding for software
+maintenance is more challenging than for funding for research. The interviewee
+from the \textit{ITK-SNAP} team thought more fundings was a way to solve the
+lack of time problem, because they could hire more dedicated developers. On the
+other hand, the interviewee from the \textit{Weasis} team did not feel that
+fundings could solve the same problem, since he still would need time to
+supervise the project.
 
 Funding challenges have also been noted by others \citep{GewaltigAndCannon2012,
-Goble2014, KaterbowAndFeulner2018, SmithEtAl2022}. \citet{WieseEtAl2019}
-reported developer pains related to publicity, since publishing norms have
-historically made it difficult to get credit for creating software.  As studied
-by \citet{HowisonAndBullard2016}, research software (specifically biology
-software, but the trend likely applies to other research software domains) is
-infrequently cited. \citet{PintoEtAl2018} also mentions the lack of formal
-reward system for research software.
+Goble2014, KaterbowAndFeulner2018, SmithEtAl2022}. Researchers that devote time
+to software have the additional challenge that funding agencies do not always
+count software when they are judging the academic excellence of the applicant.
+\citet{WieseEtAl2019} reported developer pains related to publicity, since
+publishing norms have historically made it difficult to get credit for creating
+software.  As studied by \citet{HowisonAndBullard2016}, research software
+(specifically biology software, but the trend likely applies to other research
+software domains) is infrequently cited. \citet{PintoEtAl2018} also mentions the
+lack of formal reward system for research software.
 
 An interviewee proposed an idea for increasing funding: Licensing the software
-to commercial companies that integrate it into their products.
+to commercial companies to integrate it into their products.
 
 \item[P\refstepcounter{pnum}\thepnum \label{P_TechnologyHurdles}:]
-\textbf{Technology Hurdles:} Technology pain points for LBM developers include
-setting up parallelization and CI. The pain point survey of
-\citet{WieseEtAl2019} also highlighted technical-related problems like
-dependency management, cross-platform compatibility, CI, hardware issues and
-operating system issues. Things mentioned by developers in interviews: Hard to
-keep up with changes in OS and libraries, Hard to support multiple OS, and Hard
-to support lower-end computers. With respect to balance, developers expressed difficulty balancing between four
-factors: cross-platform compatibility, convenience to development \&
-maintenance, performance, and security.  The potential and proven solutions are:
-
-% Hard to transfer to new technologies %from Ao
+\textbf{Technology Hurdles:} The technology hurdles mentioned by MI develoers
+include: hard to keep up with changes in OS and libraries, difficult to transer
+to new technologies, hard to support multiple OS, and hard to support lower-end
+computers. Developers expressed difficulty balancing between four factors:
+cross-platform compatibility, convenience to development and maintenance,
+performance, and security.
+
+The pain point survey of \citet{WieseEtAl2019} also highlighted
+technical-related problems like dependency management, cross-platform
+compatibility, CI, hardware issues and operating system issues. From
+\citep{SmithEtAl2022} technology pain points for LBM developers include setting
+up parallelization and CI.
+
+The solutions proposed by the MI developers include the following:
 
@@
 \item Increasing funding;
-\item Maintaining better documentations to ease the development \& maintenance
+\item Maintaining better documentations to ease the development and maintenance
 processes;
@@
 
-As the above list shows, web-based applications are perceived to help the
+As the above list shows, developers perceive web-based applications to help the
 balance problem.  Table~\ref{tab_native_vs_web} shows the teams' choices between
-native application and web application. In all the 29 teams on our list, most of
-them chose to develop native applications. For the eight teams we interviewed,
-three of them were building web applications, and the \textit{MRIcroGL} team was
-considering web-based solutions. So we had a good chance to discuss the
-differences between the two choices with the interviewees.
+native application and web application. In all the 29 teams on our list, most
+chose to develop native applications. For the eight teams we interviewed, three
+of them were building web applications, and the \textit{MRIcroGL} team was
+considering web-based solutions.
 
@@
 applications have the advantage of cross-platform compatibility and a simpler
-build process.  These web advantages are mirrored by the native application
-disadvantages of difficulty with cross-platform compatibility and a complex
-build process.  The lower performance disadvantage of web-applications can be
-improved with a server backend, but in this case there are disadvantages for
-privacy protection and the cost of the servers.
+build process.  These web advantages mirror the native disadvantages of
+difficulty with cross-platform compatibility and a complex build process.  The
+lower performance disadvantage of web applications can be improved with a server
+backend, but in this case there are disadvantages for privacy protection and
+server costs.
 
@@
 \textbf{Ensuring Correctness:} Several developers noted difficulties with
-ensuring correctness. They alluded to difficulty with testing the correctness of
-large numbers of features, and challenges with automated testing. The TCLB
-developer commented that the amount of testing data needed was sometimes
-problematic, since free testing services do not offer adequate facilities for
-large amounts of data, requiring in-house testing solutions. Other research
-software domains point to the following problems with testing: i)
+ensuring correctness.
+
+Other research software domains point to the following problems with testing: i)
 \citet{PintoEtAl2018} mention the problem of insufficient testing; ii)
@@
 when we do not have a means to judge the correctness of the calculated
-solutions. Based on the interviews, the LBM experience seems to overlap with i
-and ii; the developers did not allude to problem iii (the oracle problem).  As
-discussed in Section~\ref{SecReproducibility}, the oracle problem likely did not
-come up because the LBM developers have strategies for verifying their work.
+solutions. LBM developers also found difficult with testing including difficulty
+with testing the correctness of large numbers of features, and challenges with
+automated testing, since free testing services do not offer adequate facilities
+for large amounts of data.
 
@@
 to real-world datasets.  One pain point in the development process is the lack
-of access to real-world datasets for testing. The developers' strategies to
-address this are summarized in this section. One threat to correctness is: with
+of access to real-world datasets for testing. One threat to correctness is: with
 huge datasets for testing, the tests are expensive and time-consuming. Three
@@
 differing outputs between medical imaging machines, and the addition of
-(non-viewing related) functionality.  Other threats to correctness identified by at least one time
-include the following:
+(non-viewing related) functionality.  Other threats to correctness identified by
+at least one time include the following:
 \begin{itemize}
@@
 conversations, it means that users don't understand how to complete certain
-operations or what functions a feature provides. It's due to unclear description,
-lack of documentation, poorly designed UI or nonintuitive process},
+operations or what functions a feature provides. It's due to unclear
+description, lack of documentation, poorly designed UI or nonintuitive process},
 the team not having a dedicated user experience (UX) designer, some important
@@
 lack of understandability. I think the direct cause is that the project doesn't
-clearly explain how to deploy. The background of the developers wasn't discussed
-at that point.}, and the architecture is difficult for new
-developers to understand.
+clearly explain how to deploy. We did not discuss the background of the
+developers at that point.}, and the architecture is difficult for new developers
+to understand.
 

### 2022-09-30 StateOfPractice/Papers/MedImageSoft_SOP-arXiv/MedImageSoft_SOP_AcademicRadiology.tex
Updates to MI SOP paper

@@
 analytic, and medical applications \citep{FDA2021, enwiki:1034887445,
-Zhang2008}.  An example medical image of the brain is provided in
-Figure~\ref{Fig_Example}. Given the importance of MI software and the high
-number of competing software projects, we wish to understand the merits and
-drawbacks of the current development processes, tools and methodologies.  We aim
-to assess through a software engineering lens the quality of the existing
-software with the hope of highlighting standout examples, understanding current
-pain points and providing guidelines and recommendations for future development.
+Zhang2008}.  Figure~\ref{Fig_Example} shows an example image of the brain. Given
+the importance of MI software and the high number of competing software
+projects, we wish to understand the merits and drawbacks of the current
+development processes, tools, and methodologies.  We aim to assess through a
+software engineering lens the quality of the existing software with the hope of
+highlighting standout examples, understanding current pain points and providing
+guidelines and recommendations for future development.
 
@@
 as well as to other domains, of research software \citep{SmithEtAl2021,
-SmithAndMichalski2022}.  The questions are designed to learn about the
+SmithAndMichalski2022}.  We designed the questions to learn about the
 community's interest in, and experience with, artifacts, tools, principles,
-processes, methodologies and qualities.  When we mention artifacts we mean the
+processes, methodologies, and qualities.  When we mention artifacts we mean the
 documents, scripts and code that constitutes a software development project.
@@
 
-The questions are used to structure the discussion in the paper, so for each
+We based the structure of the paper on the research questions, so for each
 research question below we point to the section that contains our answer.  We
@@
 	How do MI projects compare to research software in general with respect to
-	principles, processes and methodologies used?
+	principles, processes, and methodologies used?
 	(Section~\ref{Sec_CompareMethodologies})
@@
 be feasible for a team as small as one person, and for a short time, ideally
-around a person month of effort. A person month is considered to be $20$ working
+around a person month of effort. We consider a person month as $20$ working
 days ($4$ weeks in a month, with $5$ days of work per week) at $8$ person hours
@@
 assessment we rank the software packages for each quality; therefore, this
-section also provides the background on our ranking process -- the Analytic
+section also provides the background on our ranking process --- the Analytic
 Hierarchy Process (Section \ref{sec_AHP}).
@@
 
-When assessing software packages, we need to know what license the software is
-distributed under.  In particular, we need to know whether the source code will
-be available to us or not.  We define three common software categories.  We will
-only assess software that fits under the Open Source Software license.
+When assessing software packages, we need to know the software's license.  In
+particular, we need to know whether the source code will be available to us or
+not.  We define three common software categories.  We will only assess software
+that fits under the Open Source Software license.
 
@@
 \item \textbf{Commercial Software} ``Commercial software is software developed
-by a business as part of its business'' \citep{GNU2019}. Typically speaking, the
-users are required to pay to access all of the features of commercial software,
+by a business as part of its business'' \citep{GNU2019}. Typically speaking, commercial software requires users to pay to access all of its features,
 excluding access to the source code. However, some commercial software is also
@@
 which puts related qualities (like correctness and reliability) together.
-Moreover, the order is roughly the same as the order qualities are considered in
-practice.
+Moreover, the order is roughly the same as the order developers consider
+qualities in practice.
 
@@
 	encounters circumstances not anticipated in the requirements specification,
-	and ii) when the assumptions in its requirements specification are violated
+	and ii) when users violate the assumptions in its requirements specification 
 	\citep[p.\ 19]{GhezziEtAl2003}, \citep{boehm2007software}.
@@
 
-	\item \textbf{Visibility/Transparency} The extent to which all of the steps
+	\item \textbf{Visibility/Transparency} The extent to which all the steps
 	of a software development process and the current status of it are conveyed
@@
 \ref{sec_software_quality} are the criteria ($m = 9$), while 29 software
-packages ($n = 29$) are compared for each of the $m$ criteria. The packages are
-evaluated with the grading template in Section~\ref{sec_grading_software}, which
-includes a subjective score from $1$ to $10$ for each quality for each package.
-For each quality, for a pair of packages $i$ and $j$, with $\mathit{score}_i >=
-\mathit{score}_j$, the difference between the two scores is $\mathit{diff_{ij}}
-= \mathit{score}_i - \mathit{score}_j$. The mapping between $\mathit{diff_{ij}}$
-(which can vary between 0 and 9) and the values in $A_{ij}$ (which can vary
-between 1 and 9) is as follows:
+packages ($n = 29$) are compared for each of the $m$ criteria.
+Section~\ref{sec_grading_software} shows the evaluation via the grading
+template, including a subjective score from $1$ to $10$ for each quality for
+each package. For each quality, for a pair of packages $i$ and $j$, with
+$\mathit{score}_i >= \mathit{score}_j$, the difference between the two scores is
+$\mathit{diff_{ij}} = \mathit{score}_i - \mathit{score}_j$. The mapping between
+$\mathit{diff_{ij}}$ (which can vary between 0 and 9) and the values in $A_{ij}$
+(which can vary between 1 and 9) is as follows:
 
@@
 
-The measurement results summarize to this point are based on information we
-collect from on-line resources. This information is incomplete because it
-doesn't generally capture the development process, the developer pain points,
-the perceived threats to software quality, and the developers' strategies to
-address these threats.  Therefore, part of our methodology is to interview
-developers.
+The repository-based measurements summarize the information we can collect from
+on-line resources. This information is incomplete because it doesn't generally
+capture the development process, the developer pain points, the perceived
+threats to software quality, and the developers' strategies to address these
+threats.  Therefore, part of our methodology is to also interview developers.
 
@@
 
-Our interviews identify pain points that relate to lack of time and funding,
+Our interviews identified pain points related to lack of time and funding,
 technology hurdles, improving correctness, and improving usability.  In this
@@
 Although these two qualities were not explicitly idenfied as pain points, they
-were discussed during our interviews.  The interviewee's practices for
-addressing pain points and improving quality can potentially be emulated by
-other MI developers. Moreover, these practices may provide examples that can be
-followed by other research software domains.
+were discussed as part of our interview process \citep{SmithEtAl2021}.  The
+interviewee's practices for addressing pain points and improving quality can
+potentially be emulated by other MI developers. Moreover, these practices may
+provide examples that can be followed by other research software domains.
 
@@
 management and data privacy) [Actually mentioned by MI software],
-reproducibility, and software scope determination. Although LBM developers did
+reproducibility, and software scope determination. Although MI developers did
 not mention these pain points, we cannot conclude that they are not relevant for
-LBM software development, since we only interviewed four LBM developers for
-about an hour each.
+MI software development, since we only interviewed nine MI developers for about
+an hour each.
 
@@
 \item[P\refstepcounter{pnum}\thepnum \label{P_LackDevTime}:] \textbf{Lack of
-Development Time:} Many interviewees thought lack of time and lack of funding
-(discussed next) were their most significant obstacles. Other domains of
-research software also experience the lack of time pain point
+Development Time:} Many interviewees thought lack of time, along with lack of
+funding (discussed next), were their most significant obstacles. Other domains
+of research software also experience the lack of time pain point
 \citep{PintoEtAl2018, PintoEtAl2016, WieseEtAl2019}. Our study of LBM software
@@
 \item[P\refstepcounter{pnum}\thepnum \label{P_LackFunding}:] \textbf{Lack of
-Funding:} As mentioned above, developers felt the pain of having to attract
-funding to develop and maintain their software. For instance, the interviewees
-from \textit{3D Slicer} and \textit{OHIF} said getting funding for software
-maintenance is more challenging than for funding for research. The interviewee
-from the \textit{ITK-SNAP} team thought more fundings was a way to solve the
-lack of time problem, because they could hire more dedicated developers. On the
-other hand, the interviewee from the \textit{Weasis} team did not feel that
-fundings could solve the same problem, since he still would need time to
-supervise the project.
+Funding:} Developers felt the pain of having to attract funding to develop and
+maintain their software. For instance, the interviewees from \textit{3D Slicer}
+and \textit{OHIF} said getting funding for software maintenance is more
+challenging than finding funding for research. The interviewee from the
+\textit{ITK-SNAP} team thought more fundings was a way to solve the lack of time
+problem, because they could hire more dedicated developers. On the other hand,
+the interviewee from the \textit{Weasis} team did not feel that fundings could
+solve the same problem, since he would still need time to supervise the project.
 
@@
 \item[P\refstepcounter{pnum}\thepnum \label{P_TechnologyHurdles}:]
-\textbf{Technology Hurdles:} The technology hurdles mentioned by MI develoers
-include: hard to keep up with changes in OS and libraries, difficult to transer
+\textbf{Technology Hurdles:} The technology hurdles mentioned by MI developers
+include: hard to keep up with changes in OS and libraries, difficult to transfer
 to new technologies, hard to support multiple OS, and hard to support lower-end
@@
 \item Using memory-mapped files to consume less computer memory, to better
-support lower-end computers, 
+support lower-end computers; 
 \item Using computing power from the computers GPU for web applications;
@@
 
-As the above list shows, developers perceive web-based applications to help the
-balance problem.  Table~\ref{tab_native_vs_web} shows the teams' choices between
-native application and web application. In all the 29 teams on our list, most
-chose to develop native applications. For the eight teams we interviewed, three
-of them were building web applications, and the \textit{MRIcroGL} team was
-considering web-based solutions.
+As the above list shows, developers perceive that web-based applications will
+address the technology hurdle.  Table~\ref{tab_native_vs_web} shows the teams'
+choices between native application and web application. Most of the 29 teams (24
+of 29, or 83\%) chose to develop native applications. For the eight teams we
+interviewed, three of them were building web applications, and the
+\textit{MRIcroGL} team was considering web-based solutions.
 
@@
 \item[P\refstepcounter{pnum}\thepnum \label{P_Correctness}:]
-\textbf{Ensuring Correctness:} Several developers noted difficulties with
-ensuring correctness.
-
-Other research software domains point to the following problems with testing: i)
-\citet{PintoEtAl2018} mention the problem of insufficient testing; ii)
-\citet{HannayEtAl2009} show that more developers think testing is important than
-the number that believe they have a sufficient understanding of testing
-concepts; and, iii) \citet{HannayEtAl2009, KanewalaAndBieman2013, KellyEtAl2011,
-WieseEtAl2019} point to the oracle problem, which occurs in research software
-when we do not have a means to judge the correctness of the calculated
-solutions. LBM developers also found difficult with testing including difficulty
-with testing the correctness of large numbers of features, and challenges with
+\textbf{Ensuring Correctness:} Interviewees identified multiple threats to
+correctness.  The most frequently mentioned threat was complexity.  Complexity
+enters the software by various means, including the large variety of data formats,
+complicated data standards, differing outputs between medical imaging machines,
+and the addition of (non-viewing related) functionality.  Other threats to
+correctness identified include the following:
+
+\begin{itemize}
+\item Lack of real world image data for testing, in part because of patient
+privacy concerns;
+\item Tests are expensive and time-consuming because of the need for huge datasets;
+\item Software releases are difficult to manage;
+\item No systematic unit testing; and,
+\item No dedicated quality assurance team.
+\end{itemize}
+
+**start editing here** Change to others have trouble getting developers to test.
+That isn't the problem for MI.  Testing is also a pain point for other research
+software domains point. For instance, \citet{PintoEtAl2018} mention the problem
+of insufficient testing and \citet{HannayEtAl2009} show that more developers
+think testing is important than the number that believe they have a sufficient
+understanding of testing concepts. \citet{WieseEtAl2019} mention data handling
+concerns (like data quality, data management and data privacy).
+\citet{HannayEtAl2009, KanewalaAndBieman2013, KellyEtAl2011, WieseEtAl2019}
+point to the oracle problem, which occurs in research software when we do not
+have a means to judge the correctness of the calculated solutions. For the
+specific case of LBM developers, testing challenges include difficulty with
+testing the correctness of large numbers of features, and challenges with
 automated testing, since free testing services do not offer adequate facilities
-for large amounts of data.
-
-% Hard to maintain good documentations % from Ao
-% \begin{itemize}
-%     \item Hard to have a high level roadmap from the start
-%     \item Only a few people fully understand the large codebase
-% \end{itemize}
-
-%\citet{WieseEtAl2019} mention data handling concerns (like data quality, data management and data privacy)
-
-For the pain point category of testing developers identified the lack of access
-to real-world datasets.  One pain point in the development process is the lack
-of access to real-world datasets for testing. One threat to correctness is: with
-huge datasets for testing, the tests are expensive and time-consuming. Three
-interviewees endorsed self tests / automated tests, which may save time for
-testing. Some proposed and proven solutions are as follows:
+for large amounts of data \citep{SmithEtAl2022}.
+
+Testing was the most often mentioned strategy for MI developers for ensuring
+correctness.  Seven teams mentioned test related activities, including
+test-driven development, component tests, integration tests, smoke tests,
+regression tests, self tests and automated tests.  Another approach frequently
+adopted (mentioned by 3 interviewees) is a two state development process with
+stable releases and nightly builds.  Other strategies for ensuring correctness
+that came up during the interviews include CI/CD, using de-identified copies of
+medical images for debugging, sending beta versions to medical workers who can
+access the data to do the tests, and collecting/maintaining a dataset of
+problematic images.  Some additional strategies used by MI developers include:
 
@@
 
-No interviewee provided a perfect solution to the testing problem. However,
-increased connections between the development team and medical
-professionals/institutions could ease the pain.
-
-The interviewees identified multiple threats to correctness.  The most
-frequently mentioned threat was complexity.  Complexity enters the software by
-various means, including a variety of data formats, complicated data standards,
-differing outputs between medical imaging machines, and the addition of
-(non-viewing related) functionality.  Other threats to correctness identified by
-at least one time include the following:
-\begin{itemize}
-\item Lack of real world image data for testing;
-\item The team cannot use private data for debugging even when the data causes
-problems; \wss{Why not?  Should ask Ao} \ad{For example, when a physician faces
-some problems to use the software to display a medical image, he/she might reach
-to the team for help. However, the image is private, so the physician can only
-describe the problems, but not share the image with the team to reproduce the
-problems for debugging.}
-\item Expense and time consumed by tests because of the huge datasets;
-\item Difficulty managing releases;
-\item No unit tests; and,
-\item No dedicated quality assurance team.
-\end{itemize}
-
-Testing was the most often mentioned strategy for ensuring correctness.  Seven
-teams mentioned test related activities, including test-driven development,
-component tests, integration tests, smoke tests, regression tests, self tests
-and automated tests.  Another approach frequently adopted (mentioned by 3
-interviewees) is a two state development process with stable releases and
-nightly builds.  Other strategies for ensuring correctness that came up during
-the interviews include CI/CD, using de-identified copies of medical images for
-debugging, sending beta versions to medical workers who can access the data to
-do the tests, and collecting/maintaining a dataset of problematic images.
+The feedback from the interviewees makes it clear that increased connections
+between the development team and medical professionals/institutions could ease
+the pain of ensuring correctness via testing.
+
+% Hard to maintain good documentations % from Ao
+% \begin{itemize}
+%     \item Hard to have a high level roadmap from the start
+%     \item Only a few people fully understand the large codebase
+% \end{itemize}
 

### 2022-10-02 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Submitted version of SOP for LBM
Edits to LBM SOP paper to bring more detail on AHP

@@
 
-\author{Anonymous}
-% \author[CAS]{Spencer Smith}
-% \author[CAS]{Peter Michalski}
-% \author[CAS]{Jacques Carette}
-% \author[ME]{Zahra Keshavarz-Motamed}
-
-% \affiliation[CAS]{organization={McMaster University, Computing and Software
-% Department}, %Department and Organization
-%             addressline={1280 Main Street West}, 
-%             city={Hamilton},
-%             postcode={L8S 4K1}, 
-%             state={Ontario},
-%             country={Canada}}
-
-% \affiliation[ME]{organization={McMaster University, Mechanical Engineering}, %Department and Organization
-%             addressline={1280 Main Street West}, 
-%             city={Hamilton},
-%             postcode={L8S 4K1}, 
-%             state={Ontario},
-%             country={Canada}}
+%\author{Anonymous}
+\author[CAS]{Spencer Smith}
+\author[CAS]{Peter Michalski}
+\author[CAS]{Jacques Carette}
+\author[ME]{Zahra Keshavarz-Motamed}
+
+\affiliation[CAS]{organization={McMaster University, Computing and Software
+Department}, %Department and Organization
+            addressline={1280 Main Street West}, 
+            city={Hamilton},
+            postcode={L8S 4K1}, 
+            state={Ontario},
+            country={Canada}}
+
+\affiliation[ME]{organization={McMaster University, Mechanical Engineering}, %Department and Organization
+            addressline={1280 Main Street West}, 
+            city={Hamilton},
+            postcode={L8S 4K1}, 
+            state={Ontario},
+            country={Canada}}
 
@@
 We use 10 research questions to structure and focus our discussion. These
-questions drive our methodology. 
-% \citep{SmithEtAl2021, SmithAndMichalski2022} % DOUBLE BLIND
+questions drive our methodology \citep{SmithEtAl2021, SmithAndMichalski2022}.
 For each question below, we point to the section that contains our answer.  We
@@
 \citep{JungEtAl2022}.  However, most of the literature on the state of the
-practice does not focus on specific domains, but rather on all research software.
-%\citep{SmithAndMichalski2022}. % DOUBLE BLIND
-We will refer to studies and guidelines that do
+practice does not focus on specific domains, but rather on all research software
+\citep{SmithAndMichalski2022}. We will refer to studies and guidelines that do
 not distinguish an application domain as \emph{research software in general}. By
@@
 We investigate the research questions by applying the general methodology
-summarized in Anonymous (xxxx)\footnote{Anonymous for double-blind
-review}.  %\citet{SmithEtAl2021, SmithAndMichalski2022}.  %DOUBLE BLIND
-The specific
+summarized in \citet{SmithEtAl2021, SmithAndMichalski2022}. The specific
 application of the methodology to LBM is reviewed in Section~\ref{methodology},
-with the full details in Anonymous (xxxx)\footnotemark[1].  %\citet{Michalski2021}.  
+with the full details in Anonymous (xxxx)\footnotemark[1].
+%\citet{Michalski2021}.  
 Our methodology updates
@@
 
-Compared with our previous methodology (Anonymous, xxxx)\footnotemark[1],
-%(as used in \citet{SmithEtAl2016} for instance) %DOUBLE BLIND
-measuring the application of best practices now also includes
-repository based metrics, such as the number of files, number of lines of code,
-percentage of issues that are closed, etc.  We rank the software with the
-Analytic Hierarchy Process (AHP), using the quantitative data from the grading
-template. After this, as another addition to our previous methodology, we
-interview some development teams to understand the status of their
-development processes. We next answer the research questions using the collected
-quantitative and qualitative data (Sections~\ref{methodology}
-to~\ref{Sec_Recommendations}).  Finally, we summarize the threats to validity
-(Section~\ref{threats}) and the final conclusions and recommendations
-(Section~\ref{conclusion}).
+Compared with our previous methodology (as used in \citet{SmithEtAl2016} for
+instance) %(Anonymous, xxxx)\footnotemark[1] %DOUBLE BLIND
+measuring the application of best practices now also includes repository based
+metrics, such as the number of files, number of lines of code, percentage of
+issues that are closed, etc.  We rank the software with the Analytic Hierarchy
+Process (AHP), using the quantitative data from the grading template. After
+this, as another addition to our previous methodology, we interview some
+development teams to understand the status of their development processes. We
+next answer the research questions using the collected quantitative and
+qualitative data (Sections~\ref{methodology} to~\ref{Sec_Recommendations}).
+Finally, we summarize the threats to validity (Section~\ref{threats}) and the
+final conclusions and recommendations (Section~\ref{conclusion}).
 
@@
 We apply our methodology for evaluating the state of the practice of research
-software %\citep{SmithEtAl2021, SmithAndMichalski2022} % DOUBLE BLIND
+software \citep{SmithEtAl2021, SmithAndMichalski2022} % DOUBLE BLIND
 to the specific domain of
@@
 	\item Measure using the measurement template.  The full measurement template
-	can be found in Anonymous (xxxx)\footnotemark[1]. %\citet{SmithEtAl2021}. %DOUBLE BLIND
+	can be found in \citet{SmithEtAl2021}. %DOUBLE BLIND
 	(Section~\ref{empiricalmeasures})
@@
 We use a measurement template to rank the projects by how well they follow best
-practices (\rqref{RQ_HighestQuality}), as described in Anonymous (xxxx)\footnotemark[1].
-% \citet{SmithEtAl2021}. % DOUBLE BLIND
+practices (\rqref{RQ_HighestQuality}), as described in \citet{SmithEtAl2021}. % DOUBLE BLIND
 For each software package (each column in the template), we fill in the rows of
@@
 team as small as one, given our aim to cap the measurement phase at 160
-person-hours (Anonymous, xxxx)\footnotemark[1]. %\citep{SmithEtAl2021}. %DOUBLE BLIND
+person-hours \citep{SmithEtAl2021}. %DOUBLE BLIND
 An excerpt of the template, in spreadsheet
@@
 
-%As in \citet{SmithEtAl2016}, %DOUBLE BLIND
-We used Virtual Machines (VMs) to provide an
+As in \citet{SmithEtAl2016}, %DOUBLE BLIND
+we used Virtual Machines (VMs) to provide an
 optimal testing environment for each package. We used VMs to create fresh
@@
 pairwise analysis using a matrix and generates an individual quality score for
-each software package. Anonymous (xxxx)\footnotemark[1] %\citet{SmithEtAl2016} %DOUBLE BLIND
-shows the details on how AHP ranks
-software based on quality measures.
+each software package.  We have previously used this technique to rank software
+in different domains, as in \citet{SmithEtAl2016}.
+
+AHP focuses on relative comparisons, rather than requiring an unattainable
+unified scale for measuring quality.  AHP starts with sets of $n$
+\textit{options} and $m$ \textit{criteria}.  In our project there are 24
+software products ($n=24$) and 9 criteria ($m = 9$).  To get a ranking for each
+criterion, we perform a pairwise analysis between each of the options, in the
+form of an $n \times n$ matrix $A$.  The value of $A_{jk}$ ranges from 1 to 9,
+using the following definitions \citep{Saaty1980}:
+
+\begin{itemize}
+\item 1 means equal importance. Two activities contribute equally to the
+  objective.
+\item 3 means weak importance of one over another. Experience and judgment
+  slightly favor one activity over another.
+\item 5 means essential or strong importance. Experience and judgment strongly
+  favor one activity over another.
+\item 7 means demonstrated importance. An activity is strongly favored and its
+  dominance demonstrated in practice.
+\item 9 means absolute importance. The evidence favoring one activity over
+  another is of the highest possible order of affirmation.
+\item 2, 4, 6, 8 are defined as intermediate values between the two adjacent
+  judgments.
+\end{itemize}
+
+During our grading process we capture a subjective score, from one to ten, for
+each package for each criterion.  To turn these into Saaty's pair-wise scores
+for option $x$ versus option $y$, we use the following calculation for the
+respective subjective scores $x_{\text{sub}}$ and $y_{\text{sub}}$:
+\[
+\begin{cases}
+\min\{9, x_{\text{sub}} - y_{\text{sub}} + 1\} & x_{\text{sub}} \geq y_{\text{sub}} \\
+1 / \min\{9, y_{\text{sub}} - x_{\text{sub}} + 1\} & x_{\text{sub}} < y_{\text{sub}}
+\end{cases}
+\]
+
+For example, the installability for DL\_MESO and waLBerla is measured as $9$ and
+$4$, respectively, so the score for DL\_MESO compared to waLBerla is 6 and
+waLBerla compared to DL\_MESO is 1/6, as shown in Figure~\ref{}.
+
+In our work, we make the common assumption of a reciprocal relationship between
+options $k$ and $j$; that is, $A_{kj}=1/A_{jk}$.  A partial view of $A$ for the
+quality of installability is found in Figure~\ref{}.  To obtain the overall
+score for each option for a given criterion we use matrix $B$, where $B_{jk} =
+A_{jk}/\sum_{i=1}^{i=n}(A_{i k})$.  That is, to get $B$ we normalize the columns
+of $A$, as shown in Figure~\ref{}.  For a given quality, the score ($w_j$) for
+software package $j$ is found by averaging the entries in row $j$ of $B$.  That
+is, $w_j = \frac{1}{m} \sum_{i=1}^{i=m}(A_{j i})$. Besides providing a ranking,
+the $w_j$ values can be combined for all qualities for an overall ranking, as
+discussed in Section~\ref{Sec_OverallQuality}.
+
+Figure
 
@@
 verified that the overall ranking was stable.  
-% The
-% \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/AHP2020/LBM/README.txt}{README}
-% file for the tool includes requirements and usage information. %DOUBLE BLIND
+The
+\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/AHP2020/LBM/README.txt}{README}
+file for the tool includes requirements and usage information. %DOUBLE BLIND
 
@@
 require going beyond the quantitative data from the measurement template. To
-gain insight, we interview developers using a list of 20 questions from Anonymous (xxxx)\footnotemark[1].
-%\citet{SmithEtAl2021}. %DOUBLE BLIND
+gain insight, we interview developers using a list of 20 questions from 
+\citet{SmithEtAl2021}. %DOUBLE BLIND
 The questions cover the background of the development
@@
 recording and automatic transcription. Each meeting lasted about an hour. 
-% The
-% McMaster University Research Ethics Board approved the interview process under
-% application number
-% \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/MACREM/Application.pdf}
-% {MREB\#: 5219}. %DOUBLE BLIND
+The
+McMaster University Research Ethics Board approved the interview process under
+application number
+\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/MACREM/Application.pdf}
+{MREB\#: 5219}. %DOUBLE BLIND
 
@@
 qualities into one ``global'' ranking.  The full LBM data is available publicly
-on Mendeley Data (Anonymous, xxxx)\footnotemark[1]. %\citep{Smith2022}. %DOUBLE BLIND
+on Mendeley Data \citep{Smith2022}. %DOUBLE BLIND
 
@@
 weighting between all qualities. Software packages with an overall high score
-ranked high in at least several of the individual qualities.
+ranked high in at least several of the individual qualities. ****TODO: Explain how
+to combine rankings.****
 
@@
 
-Table~\ref{Tbl_Guidelines} %(based on data from \citep{SmithAndMichalski2022}) %DOUBLE BLIND
-shows that MI artifacts generally match the recommendations found in nine
+Table~\ref{Tbl_Guidelines} (based on data from \citep{SmithAndMichalski2022}) %DOUBLE BLIND
+shows that LBM artifacts generally match the recommendations found in nine
 current research software development guidelines:
@@
 
-We investigated the risk of reviewer dependence in previous work (Anonymous, xxxx)\footnotemark[1],
-%\citep{SmithEtAl2016} % DOUBLE BLIND
+We investigated the risk of reviewer dependence in previous work \citep{SmithEtAl2016}, % DOUBLE BLIND
 where we demonstrated that the measurement process is
@@
 
+\section*{Conflict of Interest}
+
+On behalf of all authors, the corresponding author states that there is no
+conflict of interest.
+
 \bibliographystyle{ACM-Reference-Format}

### 2022-10-02 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Revisions to sections 2.6 and 3.11 to explain AHP calculations in response to reviewer comments

@@
 \usepackage{pbox}
+\usepackage{makecell}
 
@@
 Our AHP algorithm uses the overall impression score, gathered via the
-measurement template, to compare and rank the LBM software. AHP performs a
-pairwise analysis using a matrix and generates an individual quality score for
-each software package.  We have previously used this technique to rank software
-in different domains, as in \citet{SmithEtAl2016}.
+measurement template, along with pairwise comparisons between packages to rank
+the LBM software. We have previously used this technique to rank scientific
+software \citep{SmithEtAl2016}.
 
@@
 \textit{options} and $m$ \textit{criteria}.  In our project there are 24
-software products ($n=24$) and 9 criteria ($m = 9$).  To get a ranking for each
+software products ($n=24$) and 9 qualities ($m = 9$).  To get a ranking for each
 criterion, we perform a pairwise analysis between each of the options, in the
@@
 each package for each criterion.  To turn these into Saaty's pair-wise scores
-for option $x$ versus option $y$, we use the following calculation for the
+for option $x$ versus option $y$, we use the following calculation from the
 respective subjective scores $x_{\text{sub}}$ and $y_{\text{sub}}$:
@@
 
-For example, the installability for DL\_MESO and waLBerla is measured as $9$ and
-$4$, respectively, so the score for DL\_MESO compared to waLBerla is 6 and
-waLBerla compared to DL\_MESO is 1/6, as shown in Figure~\ref{}.
+\noindent For example, we measured the installability for DL\_MESO and waLBerla 
+as $9$ and $4$, respectively; therefore, on the 9-point scale, DL\_MESO compared to
+waLBerla is 6 and waLBerla compared to DL\_MESO is 1/6, as shown in the sample
+AHP calculations (Table~\ref{Tbl_SampleAHP}).
 
@@
 options $k$ and $j$; that is, $A_{kj}=1/A_{jk}$.  A partial view of $A$ for the
-quality of installability is found in Figure~\ref{}.  To obtain the overall
-score for each option for a given criterion we use matrix $B$, where $B_{jk} =
-A_{jk}/\sum_{i=1}^{i=n}(A_{i k})$.  That is, to get $B$ we normalize the columns
-of $A$, as shown in Figure~\ref{}.  For a given quality, the score ($w_j$) for
+quality of installability is found in Table~\ref{Tbl_SampleAHP}.  To obtain the
+overall score for each option for a given criterion we use matrix $B$, where
+$B_{jk} = A_{jk}/\sum_{i=1}^{i=n}(A_{i k})$.  That is, to get $B$ we normalize
+the columns of $A$ by dividing by the sum of the entries in the column, as shown
+in Table~\ref{Tbl_SampleAHP}.  For a given quality, the score ($s_j$) for
 software package $j$ is found by averaging the entries in row $j$ of $B$.  That
-is, $w_j = \frac{1}{m} \sum_{i=1}^{i=m}(A_{j i})$. Besides providing a ranking,
-the $w_j$ values can be combined for all qualities for an overall ranking, as
-discussed in Section~\ref{Sec_OverallQuality}.
-
-Figure
+is, $s_j = \frac{1}{n} \sum_{i=1}^{i=n}(B_{j i})$, as shown in the final column
+of Table~\ref{Tbl_SampleAHP}. Sorting the $s_j$ values provides a ranking for
+the software packages for the quality under investigation.  The $s_j$ values can
+be combined for all qualities for an overall ranking, as discussed in
+Section~\ref{Sec_OverallQuality}.
+
+\begin{table}[h!]
+\begin{center}
+\begin{tabular}{ l c c c c c | c c c c c | c }
+ %\toprule
+ ~ & \multicolumn{5}{l}{A Matrix} & \multicolumn{5}{l}{BMatrix} & ~\\
+ \midrule
+ ~ & \rotatebox{90}{DL\_MESO} & \rotatebox{90}{waLBerla} & \rotatebox{90}{SunlightLB} & $\cdots$ & \rotatebox{90}{Musubi} & \rotatebox{90}{DL\_MESO} & \rotatebox{90}{waLBerla} & \rotatebox{90}{SunlightLB} & $\cdots$ & \rotatebox{90}{Musubi} & AVG \\
+ \midrule
+ DL\_MESO (LBE) & 1 & 6 & 3 & $\cdots$ & 2 & 0.0691 & 0.0622 & 0.0808 & $\cdots$ & 0.0808 & 0.072\\
+ waLBerla & 1/6 & 1 & 1/4 & $\cdots$ & 1/5 & 0.0115 & 0.0104 & 0.0067 & $\cdots$ & 0.0067 & 0.009\\
+ SunlightLB & 1/3 & 4 & 1 & $\cdots$ & 1/2 & 0.0230 & 0.0415 & 0.0269 & $\cdots$ & 0.0269 & 0.031\\
+ $\vdots$ & $\vdots$ & $\vdots$ & $\vdots$ & $\ddots$ & $\vdots$ & $\vdots$ & $\vdots$ & $\vdots$ & $\ddots$ & $\vdots$ & $\vdots$\\
+ Musubi & 1/2 & 5 & 2 & $\cdots$ & 1 & 0.0345 & 0.0518 & 0.0539 & $\cdots$ & 0.0539 & 0.048\\  
+ \midrule
+ SUM = & 14.48 & 96.50 & 37.12 & $\cdots$ & 23.78 & 1.000 & 1.000 & 1.000 & $\cdots$ & 1.000 & 1.000\\
+ \bottomrule
+\end{tabular}
+\end{center}
+\caption{Sample AHP Calculations for the Quality of Installability} \label{Tbl_SampleAHP}
+\end{table}
 
@@
 
-Figure~\ref{Fig_OverallScore} shows the overall ranking. In the absence of other
-information on priorities, we calculated the ranking by assuming an equal
-weighting between all qualities. Software packages with an overall high score
-ranked high in at least several of the individual qualities. ****TODO: Explain how
-to combine rankings.****
+Figure~\ref{Fig_OverallScore} shows the overall ranking. The AHP method finds
+the weighting of quality $i$ ($w_i$) in the same way that the score ($s_j$) was
+found for software package $j$ evaluated for a given quality
+(Section~\ref{AHP}).  That is, we conduct a pairwise comparison between the
+priority of different qualities to construct the $m \times m$ matrices $A$ and
+$B$ and then we average row $i$ of matrix $B$ to find the priority value $w_i$
+for quality $i$.  If we introduce the notation that $s^i_j$ is the score for
+quality $i$ for package $j$, then the overall score $S_j$ for package $j$ is
+found via:
+
+$$S_j = \sum_{i=1}^m s^i_j w_i$$
+
+In the absence of other information on priorities, we calculated the ranking by
+assuming an equal weighting between all qualities. Software packages with an
+overall high score ranked high in at least several of the individual qualities.
 

### 2022-10-02 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/OrigSubmissionLatticeBoltz_SOP.tex
Addition of infrastructure to compare new version to old version of paper

--- 2022-10-02 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
@@
 pairwise analysis using a matrix and generates an individual quality score for
-each software package.  We have previously used this technique to rank software
-in different domains, as in \citet{SmithEtAl2016}.
-
-AHP focuses on relative comparisons, rather than requiring an unattainable
-unified scale for measuring quality.  AHP starts with sets of $n$
-\textit{options} and $m$ \textit{criteria}.  In our project there are 24
-software products ($n=24$) and 9 criteria ($m = 9$).  To get a ranking for each
-criterion, we perform a pairwise analysis between each of the options, in the
-form of an $n \times n$ matrix $A$.  The value of $A_{jk}$ ranges from 1 to 9,
-using the following definitions \citep{Saaty1980}:
-
-\begin{itemize}
-\item 1 means equal importance. Two activities contribute equally to the
-  objective.
-\item 3 means weak importance of one over another. Experience and judgment
-  slightly favor one activity over another.
-\item 5 means essential or strong importance. Experience and judgment strongly
-  favor one activity over another.
-\item 7 means demonstrated importance. An activity is strongly favored and its
-  dominance demonstrated in practice.
-\item 9 means absolute importance. The evidence favoring one activity over
-  another is of the highest possible order of affirmation.
-\item 2, 4, 6, 8 are defined as intermediate values between the two adjacent
-  judgments.
-\end{itemize}
-
-During our grading process we capture a subjective score, from one to ten, for
-each package for each criterion.  To turn these into Saaty's pair-wise scores
-for option $x$ versus option $y$, we use the following calculation for the
-respective subjective scores $x_{\text{sub}}$ and $y_{\text{sub}}$:
-\[
-\begin{cases}
-\min\{9, x_{\text{sub}} - y_{\text{sub}} + 1\} & x_{\text{sub}} \geq y_{\text{sub}} \\
-1 / \min\{9, y_{\text{sub}} - x_{\text{sub}} + 1\} & x_{\text{sub}} < y_{\text{sub}}
-\end{cases}
-\]
-
-For example, the installability for DL\_MESO and waLBerla is measured as $9$ and
-$4$, respectively, so the score for DL\_MESO compared to waLBerla is 6 and
-waLBerla compared to DL\_MESO is 1/6, as shown in Figure~\ref{}.
-
-In our work, we make the common assumption of a reciprocal relationship between
-options $k$ and $j$; that is, $A_{kj}=1/A_{jk}$.  A partial view of $A$ for the
-quality of installability is found in Figure~\ref{}.  To obtain the overall
-score for each option for a given criterion we use matrix $B$, where $B_{jk} =
-A_{jk}/\sum_{i=1}^{i=n}(A_{i k})$.  That is, to get $B$ we normalize the columns
-of $A$, as shown in Figure~\ref{}.  For a given quality, the score ($w_j$) for
-software package $j$ is found by averaging the entries in row $j$ of $B$.  That
-is, $w_j = \frac{1}{m} \sum_{i=1}^{i=m}(A_{j i})$. Besides providing a ranking,
-the $w_j$ values can be combined for all qualities for an overall ranking, as
-discussed in Section~\ref{Sec_OverallQuality}.
-
-Figure
+each software package. \citet{SmithEtAl2016} %DOUBLE BLIND
+shows the details on how AHP ranks
+software based on quality measures.
 
@@
 weighting between all qualities. Software packages with an overall high score
-ranked high in at least several of the individual qualities. ****TODO: Explain how
-to combine rankings.****
+ranked high in at least several of the individual qualities.
 

### 2022-10-02 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/ResubmissionResponseToReviewer.tex
Start of resubmission letter to the editor

\documentclass[12pt]{casletter}

\oddsidemargin -10mm
\evensidemargin -10mm
\textwidth 180mm
\textheight 200mm
%\renewcommand\baselinestretch{0.85}

\usepackage{xcolor}

\pagestyle {plain}
\pagenumbering {arabic}

\email{smiths@mcmaster.ca}
\telephone{(905) 525-9140 ext.\ 24131}
\www{{http://}www.cas.mcmaster.ca/\~\@smiths}

\signature{Spencer Smith}

\address{Dr.\ Spencer Smith, Associate Professor\\
Computing and Software Department\\
Faculty of Engineering\\
McMaster University\\
1280 Main Street West\\
Hamilton, Ontario, Canada L8S 4K1}%  \\ Phone: 525-9140 Ext.~27929 \\E-mail:
                                  %  smiths@mcmaster.ca}

\begin{document}

\begin{letter}{~\\
    ~\\
    ~\\
    ~\\
    Dr.\ Eugenio O\~{n}ate\\
    Editor-in-Chief\\
    Archives of Computational Methods in Engineering\\
    ~\\
    {\bf Re: Revisions to ARCO-D-22-00308R1, State of the Practice for Lattice
    Boltzmann Method Software}}

  \opening {Dear Dr.\ O\~{n}ate:}

  Thank you and the reviewer for the feedback on our submission.  In response to
  your e-mail, dated August 26, 2022, we have revised the paper to incorporate
  the requested revisions.  We provide a summary below.  We have also included
  with the revised submission a version of the paper showing all additions and
  deletions.

  We would like to thank you and the reviewer for their thoughtful and
  constructive comments.

  \textbf{Reviewer \#1}

  \begin{enumerate}
  \item The anonymization of the authors poses some challenges in evaluating
  the paper because the citations for the methodology and the raw data
  underlying the study have been anonymized away. \medskip

    \emph{We have removed the anonymization in the resubmission.  \smallskip}

  \item I think it would be useful to have slightly more by way of explanation
  here -- to provide a bit more understanding of the process followed for
  readers who choose not to pursue the details provided in the reference.  Two
  key steps to put a little more flesh around would be (1) how are the entries
  of the measurement template (Fig 2) turned into a single numerical score for
  the section, and (2) how is the aggregate AHP score obtained?  As to the first
  point, for example, I look at the Installability Overall Impression scores,
  which I initially took to be a numerical score for installability, but in the
  AHP Installability scores shown in Fig 3, there is no correlation between the
  number in the measurement template and the final score. \medskip

    \emph{Response. \smallskip}

  \item [I]n Fig 2, I see responses for "Descriptive error messages?" of yes, no,
  and n/a.  I have a hard time imagining what n/a could mean in this context.
  Presumably the installer either provides descriptive error messages or it does
  not?  Is this explained in one of the anonymized references or in the raw
  dataset, which I can't access either due to anonymization. \medskip

    \emph{Response. \smallskip}

  \item AHP is left completely to outside references, and I think it would be
  beneficial to have at least a brief description here. \medskip

    \emph{Response. \smallskip}

  \item I think there was a statement that AHP weights all of the qualities
  equally, but in the graphs of the scores for various qualities (e.g., Figs
  3-6), there are dramatic differences in the ranges of the scores given to the
  packages -- anywhere from 0.06 as the top score to nearly 0.16.  The highest
  overall score is ~0.075 (Fig 12). So what is the *theoretical* top score in
  any category and overall?  Are there some categories in which even the
  highest-scoring packages are quite low on the absolute scale?  If so, I would
  think such situations deserve some discussion. \medskip

    \emph{Response. \smallskip}

  \end{enumerate}

  \closing{Best regards,~\newline} \vspace{-29mm}
  \includegraphics[scale=0.75]{Signature.pdf}

\end {letter}

\end{document}

### 2022-10-03 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
updated paper to reflect updated figures

@@
     practices and popularity. We find the state of the practice to be healthy,
-    with 67\% of the measured packages ranking in the top five for at least one
+    with 87.5\% of the measured packages ranking in the top five for at least one
     quality, the majority of LBM generated artifacts corresponding to general
@@
 to a higher score. Apparently active work on a project improves installability
-since the top seven ranked packages are all classified as alive.
+since the top six ranked packages are all classified as alive.
 
@@
 documentation. They also incorporated CI in the development process.
-Furthermore, eight of the top 10 ranked packages are alive.
+Furthermore, all of the top 5 ranked packages are alive.
 
@@
 software packages. Software packages with a high score either did not break
-during installation, or the broken installation was recoverable. All the top
-five ranked packages have tutorials. The package pyLBM broke during tutorial
-testing, but a descriptive error message helped in recovery. Nine of the top 10
-packages are alive.
+during installation, or the broken installation was recoverable.
+The package pyLBM broke during tutorial testing, but a descriptive error message helped in recovery.
+Ten of the top 11 packages are alive.
 
@@
 input files, including a replacement of new lines with carriage returns.
-Following the trend of the other qualities, nine of the top 10 ranked packages
-are alive.
 
@@
 of their issues closed, and only three (laboetie, lettuce, Sailfish) have less
-than 50\% of their issues closed. All the top five ranked packages have most of
+than 50\% of their issues closed. All the top four ranked packages have most of
 their issues closed. Alive packages (11 use issue tracking) have 64\% of their
@@
 packages, and the average percentage of code comments is about 14\%. All the top
-five overall ranked packages have more than the average. LUMA has only 0.2\%
+four overall ranked packages have more than the average. LUMA has only 0.2\%
 comments, the fewest of any package. However, this package also has the most
@@
 closed. Maintainable packages by our ranking have more than 10\% of the code as
-comments. Four of the top five ranked packages are alive.
+comments. All of the top four ranked packages are alive.
 
@@
 non-LBM modelling. For this reason, a comparison with other software packages is
-not on a level field. Nine of the top 10 ranked packages are alive.
+not on a level field. Seven of the top 8 ranked packages are alive.
 
@@
 explicit development model and defined development process. They also have
-detailed and easy to access software release notes. Four of the top five ranked
+detailed and easy to access software release notes. Five of the top six ranked
 packages are alive.
@@
 count, but relatively low in the ``follow best practices'' ranking: TCLB (star
-rank 3, our rank 12), lettuce (star rank 4, our rank 14) and \esp{} (star rank
+rank 3, our rank 10), lettuce (star rank 4, our rank 12) and \esp{} (star rank
 5, our rank 15).  Part of the discrepancy is because our process ranks all
@@
 			Palabos & 3 & 34 & 6 & GitLab & GitLab\\
-			OpenLB & 4 & N/A & N/A & N/A& N/A\\
-			LUMA & 5 & 33 & 7 & 12 & 4\\
+			LUMA & 4 & 33 & 7 & 12 & 4\\
+			OpenLB & 5 & N/A & N/A & N/A& N/A\\
 			pyLBM & 6 & 95 & 3 & 10 & 5\\
-			DL\_MESO (LBE) & 7 & N/A & N/A & N/A & N/A\\
-			Musubi & 8 & N/A & N/A & N/A & N/A\\
+			Musubi & 7 & N/A & N/A & N/A & N/A\\
+			waLBerla & 8 & 20 & 9 & GitLab & GitLab\\
+			DL\_MESO (LBE) & 9 & N/A & N/A & N/A & N/A\\
 			Sailfish & 9 & 186 & 1 & 41& 1\\
-			waLBerla & 10 & 20 & 9 & GitLab & GitLab\\
+			TCLB & 10 & 95 & 3 & 16 & 3\\
 			laboetie & 11 & 4 & 13 & 5 & 8\\
-			TCLB & 12 & 95 & 3 & 16 & 3\\
+			lettuce & 12 & 48 & 4 & 5 & 8\\
 			MechSys & 13 & N/A & N/A & N/A & N/A\\
-			lettuce & 14 & 48 & 4 & 5 & 8\\
-			\esp & 15 & 35 & 5 & 12 & 4\\
-			MP-LABS & 16 & 12 & 11 & 2 & 9\\			
-			SunlightLB & 17 & N/A & N/A & N/A & N/A\\
-			LB3D & 18 & N/A & N/A & N/A & N/A\\			
-			LIMBES & 19 & N/A & N/A & N/A & N/A\\
-			LB2D-Prime & 20 & N/A & N/A & N/A & N/A\\		
-			HemeLB & 21 & 12 & 11 & 12 & 4\\
-			lbmpy & 22 &  11 & 12 & 2 & 9\\	
-			LB3D-Prime & 23 & N/A & N/A & N/A & N/A\\	
-			LatBo.jl & 24 & 17 & 10 & 8 & 6\\			
+			\esp & 14 & 35 & 5 & 12 & 4\\
+			LB2D-Prime & 15 & N/A & N/A & N/A & N/A\\
+			LB3D & 16 & N/A & N/A & N/A & N/A\\
+			SunlightLB & 16 & N/A & N/A & N/A & N/A\\
+			HemeLB & 17 & 12 & 11 & 12 & 4\\
+			MP-LABS & 17 & 12 & 11 & 2 & 9\\
+			LIMBES & 18 & N/A & N/A & N/A & N/A\\
+			lbmpy & 19 &  11 & 12 & 2 & 9\\
+			LB3D-Prime & 20 & N/A & N/A & N/A & N/A\\
+			LatBo.jl & 21 & 17 & 10 & 8 & 6\\
 			\bottomrule
@@
 
-The top four packages (ESPREsSo, Ludwig, Palabos, and OpenLB) include most of
+The top three packages (ESPREsSo, Ludwig, Palabos) include most of
 the common and uncommon artifacts, but not many of the rare artifacts.  They
 collectively have all the common artifacts, except Palabos does not have theory
-notes, and OpenLB does not appear to use version control.  The top four packages
+notes.  The top three packages
 have most of the uncommon artifacts. Only one them (Ludwig) is missing design
@@
 not reflected in our data because it was not present at the time of data
-collection (mid-2020). Despite the broken link, Palabos does have a detailed and
-informative website. The top four ranked packages do not have many of the rare
-artifacts. None of them have any explicit API documentation. Three of them
-(ESPREsSo, Ludwig, Palabos) have information on contributing to the project and
-two (OpenLB, Palabos) have an FAQ section or forum. One (OpenLB) has
-verification and validation notes, and a video guide for the software.
+collection (mid-2020).
+Despite the broken link, Palabos does have a detailed and informative website.
+The top three ranked packages do not have many of the rare artifacts.
+None of them have any explicit API documentation.
+All of them have information on contributing to the project and
+one (Palabos) has an FAQ section or forum.
 
@@
 each quality and for the overall quality. Taking the union of all entries in the
-table, we have 16 top performers: ESPREsSo, MechSys, OpenLB, DL\_MESO, Sailfish,
-Ludwig, Musubi, pyLBM, waLBerla, laboetie, lettuce, \esp, LUMA, MP-LABS,
-Palabos, and HemeLB.  Therefore, two thirds (67\%) of the 24 packages
-appear in the top five for at least one quality.  Almost all entries are alive,
-except for three packages: laboetie, MP-LABS, and HemeLB.  Alive packages are in
+table, we have 21 top performers: ESPREsSo, MechSys, OpenLB, DL\_MESO, Palabos, Sailfish, Ludwig, Musubi, pyLBM, waLBerla, laboetie, lettuce, LUMA, TCLB, \esp, LB2D-Prime, LB3D-Prime, LIMBES, SunlightLB, HemeLB, MP-LABS.
+Therefore, seven eights (87.5\%) of the 24 packages
+appear in the top five for at least one quality.  Most entries are alive.  Alive packages are in
 a healthy state with respect to adopting best practices.
@@
 	\begin{center}
-		\begin{tabular}{ p{3cm}p{1.9cm}p{1.9cm}p{1.9cm}p{1.9cm}p{1.9cm} }
+		\begin{tabular}{ p{3cm}p{3cm}p{2.1cm}p{2.1cm}p{2.1cm}p{2.1cm} }
 			\toprule
@@
 			\midrule
-			Installability & ESPResSo & MechSys & OpenLB & DL\_MESO (LBE) &
-			Sailfish\\
+			Installability & ESPResSo, MechSys, OpenLB & & & DL\_MESO (LBE),
+			Palabos, Sailfish &\\
 			\addlinespace[0.4cm]
 			\pbox{3.0cm}{Surface Correctness \\ and Verifiability} & Ludwig &
-			Musubi & pyLBM & OpenLB & waLBerla\\
+			Musubi, pyLBM & & ESPResSo, waLBerla &\\
 			\addlinespace[0.4cm]
-			Surface Reliability & DL\_MESO (LBE) & laboetie* & lettuce & MechSys
-			& Musubi \\
+			Surface Reliability & Musubi, DL\_MESO (LBE), ESPResSo, laboetie*, lettuce, LUMA, MechSys, OpenLB, Palabos, pyLBM, TCLB, & & & & \\
 			\addlinespace[0.4cm]
-			Surface Robustness & DL\_MESO (LBE) & \esp & lettuce & LUMA &
-			OpenLB \\
+			Surface Robustness & DL\_MESO, ESPResSo, \esp, laboetie*, LB2D-Prime*, LB3D-Prime*, lettuce, LIMBES*, Ludwig, LUMA, MechSys, OpenLB, Palabos, pyLBM, Sailfish, SunlightLB*, TCLB, waLBerla & & & & \\
 			\addlinespace[0.4cm]
-			Surface Usability & laboetie* & ESPResSo & OpenLB & Ludwig & Sailfish\\
+			Surface Usability & ESPResSo, laboetie*, Ludwig, OpenLB, Sailfish & & & & \\
 			\addlinespace[0.4cm]
-			Maintainability & Ludwig & MP-LABS* & \esp & ESPResSo & Palabos\\
+			Maintainability & ESPResSo, Ludwig, Palabos, waLBerla & & & & \esp \\
 			\addlinespace[0.4cm]
-			Reusability & ESPResSo & waLBerla & \esp & Sailfish & HemeLB*\\
+			Reusability & ESPResSo, waLBerla & & \esp, HemeLB*, Musubi, OpenLB, Palabos, Sailfish & &\\
 			\addlinespace[0.4cm]
-			\pbox{3.0cm}{Surface \\Understandability} & LUMA & Palabos & pyLBM &
-			DL\_MESO (LBE)& MechSys\\
+			\pbox{3.0cm}{Surface \\Understandability} & LUMA & DL\_MESO (LBE), lettuce, Musubi, Palabos & & & \\
 			\addlinespace[0.4cm]
-			\pbox{3.0cm}{Visibility and \\Transparency} & Ludwig & Palabos &
-			LUMA & ESPResSo & MP-LABS*\\
+			\pbox{3.0cm}{Visibility and \\Transparency} & Ludwig & ESPResSo, LUMA, Palabos & & & MP-LABS*\\
 			\addlinespace[0.4cm]
-			Overall Quality & ESPResSo & Ludwig & Palabos & OpenLB & LUMA\\
+			Overall Quality & ESPResSo & Ludwig & Palabos & LUMA & OpenLB \\
 			\bottomrule

### 2022-10-03 StateOfPractice/AHP/Source code/DataProcess/src/Main.java
renamed original AHP folder to AHP2014

@@ -1,143 +1,0 @@
-import java.io.File; 
-import java.io.FileInputStream;
-import java.io.FileOutputStream;
-import java.util.Iterator;
-
-import javax.swing.JOptionPane;
-
-import org.apache.poi.ss.usermodel.Cell;
-import org.apache.poi.ss.usermodel.Row;
-import org.apache.poi.xssf.usermodel.XSSFSheet;
-import org.apache.poi.xssf.usermodel.XSSFWorkbook;
-
-public class Main {
-	
-	public static void main(String[] args)
-    {
-		Cell[][] table=new Cell[13][30];
-		Double[][][] ahpTable= new Double[13][30][29];
-        try
-        {
-            FileInputStream file = new FileInputStream(new File("AHP_Template.xlsx"));
- 
-            //Create Workbook instance holding reference to .xlsx file
-            XSSFWorkbook workbook = new XSSFWorkbook(file);
- 
-            //Get second sheet from the workbook
-            XSSFSheet sheet = workbook.getSheetAt(1);
- 
-            //Iterate through each rows one by one
-            Iterator<Row> rowIterator = sheet.iterator();            
-            int r=0;
-            int col=0;
-            while (rowIterator.hasNext())
-            {
-                Row row = rowIterator.next();
-                //For each row, iterate through all the columns
-                Iterator<Cell> cellIterator = row.cellIterator();
-                 
-                while (cellIterator.hasNext())
-                {
-                    Cell cell = cellIterator.next();
-                    //Check the cell type and format accordingly
-                    switch (cell.getCellType())
-                    {
-                        case Cell.CELL_TYPE_NUMERIC:
-                        	table[r][col]=cell;    
-                        	col++;
-                            break;
-                        case Cell.CELL_TYPE_STRING:
-                            break;
-                        default:
-                           	break;
-                    }                    
-                }
-                r++;
-                col=0;
-            }
-            XSSFSheet sheetnew;
-            if(workbook.getSheet("AHP Example")==null){
-            	sheetnew= workbook.createSheet("AHPDATA");
-            }
-            else{
-            	sheetnew=workbook.getSheet("AHP Example");
-            }
-            double x=0;
-            double y=0;
-            for(int i=0;i<30;i++)
-            {
-            	for(int j=i;j<29;j++)
-            	{   
-            		for(int k=0;k<13;k++)
-            		{
[…]

### 2022-10-03 StateOfPractice/AHP2014/Source code/DataProcess/src/Main.java

@@ -1,0 +1,143 @@
+import java.io.File; 
+import java.io.FileInputStream;
+import java.io.FileOutputStream;
+import java.util.Iterator;
+
+import javax.swing.JOptionPane;
+
+import org.apache.poi.ss.usermodel.Cell;
+import org.apache.poi.ss.usermodel.Row;
+import org.apache.poi.xssf.usermodel.XSSFSheet;
+import org.apache.poi.xssf.usermodel.XSSFWorkbook;
+
+public class Main {
+	
+	public static void main(String[] args)
+    {
+		Cell[][] table=new Cell[13][30];
+		Double[][][] ahpTable= new Double[13][30][29];
+        try
+        {
+            FileInputStream file = new FileInputStream(new File("AHP_Template.xlsx"));
+ 
+            //Create Workbook instance holding reference to .xlsx file
+            XSSFWorkbook workbook = new XSSFWorkbook(file);
+ 
+            //Get second sheet from the workbook
+            XSSFSheet sheet = workbook.getSheetAt(1);
+ 
+            //Iterate through each rows one by one
+            Iterator<Row> rowIterator = sheet.iterator();            
+            int r=0;
+            int col=0;
+            while (rowIterator.hasNext())
+            {
+                Row row = rowIterator.next();
+                //For each row, iterate through all the columns
+                Iterator<Cell> cellIterator = row.cellIterator();
+                 
+                while (cellIterator.hasNext())
+                {
+                    Cell cell = cellIterator.next();
+                    //Check the cell type and format accordingly
+                    switch (cell.getCellType())
+                    {
+                        case Cell.CELL_TYPE_NUMERIC:
+                        	table[r][col]=cell;    
+                        	col++;
+                            break;
+                        case Cell.CELL_TYPE_STRING:
+                            break;
+                        default:
+                           	break;
+                    }                    
+                }
+                r++;
+                col=0;
+            }
+            XSSFSheet sheetnew;
+            if(workbook.getSheet("AHP Example")==null){
+            	sheetnew= workbook.createSheet("AHPDATA");
+            }
+            else{
+            	sheetnew=workbook.getSheet("AHP Example");
+            }
+            double x=0;
+            double y=0;
+            for(int i=0;i<30;i++)
+            {
+            	for(int j=i;j<29;j++)
+            	{   
+            		for(int k=0;k<13;k++)
+            		{
[…]

### 2022-10-06 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Edits to SOP for LBM to update Table 11 to show top two packages for each quality

@@
 	\begin{center}
-		\begin{tabular}{ p{3cm}p{3cm}p{2.1cm}p{2.1cm}p{2.1cm}p{2.1cm} }
+		\begin{tabular}{ p{3cm}p{13cm} }
 			\toprule
-			Quality & Ranked 1st & Ranked 2nd & Ranked 3rd & Ranked 4th & Ranked
-			5th\\
+			Quality & Ranked 1st or 2nd\\
 			\midrule
-			Installability & ESPResSo, MechSys, OpenLB & & & DL\_MESO (LBE),
-			Palabos, Sailfish &\\
+			Installability & ESPResSo, MechSys, OpenLB, DL\_MESO (LBE),
+			Palabos, Sailfish\\
 			\addlinespace[0.4cm]
-			\pbox{3.0cm}{Surface Correctness \\ and Verifiability} & Ludwig &
-			Musubi, pyLBM & & ESPResSo, waLBerla &\\
+			\pbox{3.0cm}{Surface Correctness \\ and Verifiability} & Ludwig,
+			Musubi, pyLBM\\%, ESPResSo, waLBerla\\
 			\addlinespace[0.4cm]
-			Surface Reliability & Musubi, DL\_MESO (LBE), ESPResSo, laboetie*, lettuce, LUMA, MechSys, OpenLB, Palabos, pyLBM, TCLB, & & & & \\
+			Surface Reliability & Musubi, DL\_MESO (LBE), ESPResSo, laboetie*, lettuce, LUMA, MechSys, OpenLB, Palabos, pyLBM, TCLB\\
 			\addlinespace[0.4cm]
-			Surface Robustness & DL\_MESO, ESPResSo, \esp, laboetie*, LB2D-Prime*, LB3D-Prime*, lettuce, LIMBES*, Ludwig, LUMA, MechSys, OpenLB, Palabos, pyLBM, Sailfish, SunlightLB*, TCLB, waLBerla & & & & \\
+			Surface Robustness & DL\_MESO, ESPResSo, \esp, laboetie*, LB2D-Prime*, LB3D-Prime*, lettuce, LIMBES*, Ludwig, LUMA, MechSys, OpenLB, Palabos, pyLBM, Sailfish, SunlightLB*, TCLB, waLBerla, Musubi\\
 			\addlinespace[0.4cm]
-			Surface Usability & ESPResSo, laboetie*, Ludwig, OpenLB, Sailfish & & & & \\
+			Surface Usability & ESPResSo, laboetie*, Ludwig, OpenLB, Sailfish, DL\_MESO, \esp LB2D-Prime*, LB3D, LUMA, TCLB\\
 			\addlinespace[0.4cm]
-			Maintainability & ESPResSo, Ludwig, Palabos, waLBerla & & & & \esp \\
+			Maintainability & ESPResSo, Ludwig, Palabos, waLBerla, \esp, LUMA, MP-LABS, pyLBM, TCLB\\
 			\addlinespace[0.4cm]
-			Reusability & ESPResSo, waLBerla & & \esp, HemeLB*, Musubi, OpenLB, Palabos, Sailfish & &\\
+			Reusability & ESPResSo, waLBerla, \esp, HemeLB*, Musubi, OpenLB, Palabos, Sailfish\\
 			\addlinespace[0.4cm]
-			\pbox{3.0cm}{Surface \\Understandability} & LUMA & DL\_MESO (LBE), lettuce, Musubi, Palabos & & & \\
+			\pbox{3.0cm}{Surface \\Understandability} & LUMA, DL\_MESO (LBE), lettuce, Musubi, Palabos\\
 			\addlinespace[0.4cm]
-			\pbox{3.0cm}{Visibility and \\Transparency} & Ludwig & ESPResSo, LUMA, Palabos & & & MP-LABS*\\
+			\pbox{3.0cm}{Visibility and \\Transparency} & Ludwig, ESPResSo, LUMA, Palabos\\
 			\addlinespace[0.4cm]
-			Overall Quality & ESPResSo & Ludwig & Palabos & LUMA & OpenLB \\
+			Overall Quality & ESPResSo, Ludwig\\%, Palabos, LUMA, OpenLB\\
 			\bottomrule

### 2022-10-08 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Updates to paper
Update paper to the plots and discussion for the actual measured data (not the sensitivity analysis data)

@@
     practices and popularity. We find the state of the practice to be healthy,
-    with 87.5\% of the measured packages ranking in the top five for at least one
-    quality, the majority of LBM generated artifacts corresponding to general
+    with 67\% of the measured packages ranking in the top two for at least two
+    qualities, the majority of LBM generated artifacts corresponding to general
     recommendations from research software developers, common use of version
@@
 application of the methodology to LBM is reviewed in Section~\ref{methodology},
-with the full details in Anonymous (xxxx)\footnotemark[1].
-%\citet{Michalski2021}.  
+with the full details in \citet{Michalski2021}.  
 Our methodology updates
@@
 Once we measure each package, we still need to rank them to answer
-\rqref{RQ_HighestQuality}. We do this via the Analytic Hierarchy Process (AHP),
-a decision-making technique that compares multiple options by multiple criteria.
-Our AHP algorithm uses the overall impression score, gathered via the
-measurement template, along with pairwise comparisons between packages to rank
-the LBM software. We have previously used this technique to rank scientific
-software \citep{SmithEtAl2016}.
+\rqref{RQ_HighestQuality}. We do this via the Analytic Hierarchy Process (AHP)
+\citep{Saaty1980}, a decision-making technique that compares multiple options by
+multiple criteria. Our AHP algorithm uses the overall impression score, gathered
+via the measurement template, along with pairwise comparisons between packages
+to rank the LBM software. We have previously used this technique to rank
+scientific software \citep{SmithEtAl2016}.
 
@@
 packages. Software packages with a higher score (ESPResSo, MechSys, OpenLB,
-DL\_MESO, Palabos) tend to have one set of linear installation instructions,
-written as if the person doing the installation has none of the dependencies
-installed. The instructions often list compatible operating system versions and
-include instructions for installing dependencies. The top-ranked packages often
-incorporate some automation of the installation process. The number of
-dependencies a package has does not correlate with a higher score. The ability
-to validate the installation process, say through tutorials or test cases, leads
-to a higher score. Apparently active work on a project improves installability
-since the top six ranked packages are all classified as alive.
+DL\_MESO, Palabos, Sailfish) tend to have one set of linear installation
+instructions, written as if the person doing the installation has none of the
+dependencies installed. The instructions often list compatible operating system
+versions and include instructions for installing dependencies. The top-ranked
+packages often incorporate some automation of the installation process. The
+number of dependencies a package has does not correlate with a higher score. The
+ability to validate the installation process, say through tutorials or test
+cases, leads to a higher score. Apparently active work on a project improves
+installability since the top six ranked packages are all classified as alive.
 
@@
 correctness. The top ranked software packages all include getting started
-tutorials, and most of these include expected output. The two top ranked
+tutorials, and most of these include expected output. Two of the top ranked
 packages, Ludwig and Musubi, indicate the use of unit testing in their
 documentation. They also incorporated CI in the development process.
-Furthermore, all of the top 5 ranked packages are alive.
+Furthermore, all the top 5 ranked packages are alive.
 
@@
 software packages. Software packages with a high score either did not break
-during installation, or the broken installation was recoverable.
-The package pyLBM broke during tutorial testing, but a descriptive error message helped in recovery.
-Ten of the top 11 packages are alive.
+during installation, or the broken installation was recoverable. The package
+pyLBM broke during tutorial testing, but a descriptive error message helped in
+recovery. Ten of the top 11 packages are alive.
 
@@
 explicit development model and defined development process. They also have
-detailed and easy to access software release notes. Five of the top six ranked
+detailed and easy to access software release notes. All the top five ranked
 packages are alive.
@@
 priority of different qualities to construct the $m \times m$ matrices $A$ and
-$B$ and then we average row $i$ of matrix $B$ to find the priority value $w_i$
+$B$, and then we average row $i$ of matrix $B$ to find the priority value $w_i$
 for quality $i$.  If we introduce the notation that $s^i_j$ is the score for
@@
 rank 3, our rank 10), lettuce (star rank 4, our rank 12) and \esp{} (star rank
-5, our rank 15).  Part of the discrepancy is because our process ranks all
+5, our rank 14).  Part of the discrepancy is because our process ranks all
 projects, but the star measure does not apply to all projects.  Also, our
@@
 Section~\ref{Sec_AddressConcerns} covers potential ways to address the pain
-points and Anonymous (xxxx)\footnotemark[1] %\citet{SmithEtAl2021} 
-provides the full interview questions.
+points and \citet{SmithEtAl2021} provides the full interview questions.
 
@@
 Current LBM software mostly follows best practices, as illustrated by
-Table~\ref{topperformerstable}, which summarizes the top five performers for
-each quality and for the overall quality. Taking the union of all entries in the
-table, we have 21 top performers: ESPREsSo, MechSys, OpenLB, DL\_MESO, Palabos, Sailfish, Ludwig, Musubi, pyLBM, waLBerla, laboetie, lettuce, LUMA, TCLB, \esp, LB2D-Prime, LB3D-Prime, LIMBES, SunlightLB, HemeLB, MP-LABS.
-Therefore, seven eights (87.5\%) of the 24 packages
-appear in the top five for at least one quality.  Most entries are alive.  Alive packages are in
-a healthy state with respect to adopting best practices.
+Table~\ref{topperformerstable}, which summarizes the top two packages for each
+quality and for the overall quality. Taking the union of all entries in the
+table that appear under 2 or more qualities, we have 16 top performers:
+ESPREsSo, MechSys, OpenLB, DL\_MESO (LBE), Sailfish, Ludwig, Musubi, pyLBM,
+waLBerla, laboetie, lettuce, \esp, LUMA, Palabos, TCLB and LB2D-Prime.
+Therefore, two thirds (67\%) of the 24 packages appear in the top two for at
+least two qualities.  Almost all the top performers are alive, except for two
+packages: laboetie and LB2D-Prime.  Alive packages are in a healthy state with
+respect to adopting best practices.
 
@@
 			\pbox{3.0cm}{Surface Correctness \\ and Verifiability} & Ludwig,
-			Musubi, pyLBM\\%, ESPResSo, waLBerla\\
+			Musubi, pyLBM\\
 			\addlinespace[0.4cm]
-			Surface Reliability & Musubi, DL\_MESO (LBE), ESPResSo, laboetie*, lettuce, LUMA, MechSys, OpenLB, Palabos, pyLBM, TCLB\\
+			Surface Reliability & Musubi, DL\_MESO (LBE), ESPResSo, laboetie*,
+			lettuce, LUMA, MechSys, OpenLB, Palabos, pyLBM, TCLB\\
 			\addlinespace[0.4cm]
-			Surface Robustness & DL\_MESO, ESPResSo, \esp, laboetie*, LB2D-Prime*, LB3D-Prime*, lettuce, LIMBES*, Ludwig, LUMA, MechSys, OpenLB, Palabos, pyLBM, Sailfish, SunlightLB*, TCLB, waLBerla, Musubi\\
+			Surface Robustness & DL\_MESO (LBE), ESPResSo, \esp, laboetie*,
+			LB2D-Prime*, LB3D-Prime*, lettuce, LIMBES*, Ludwig, LUMA, MechSys,
+			OpenLB, Palabos, pyLBM, Sailfish, SunlightLB*, TCLB, waLBerla,
+			Musubi\\
 			\addlinespace[0.4cm]
-			Surface Usability & ESPResSo, laboetie*, Ludwig, OpenLB, Sailfish, DL\_MESO, \esp LB2D-Prime*, LB3D, LUMA, TCLB\\
+			Surface Usability & ESPResSo, laboetie*, Ludwig, OpenLB, Sailfish,
+			DL\_MESO (LBE), \esp, LB2D-Prime*, LB3D*, LUMA, TCLB\\
 			\addlinespace[0.4cm]
-			Maintainability & ESPResSo, Ludwig, Palabos, waLBerla, \esp, LUMA, MP-LABS, pyLBM, TCLB\\
+			Maintainability & ESPResSo, Ludwig, Palabos, waLBerla, \esp, LUMA,
+			MP-LABS*, pyLBM, TCLB\\
 			\addlinespace[0.4cm]
-			Reusability & ESPResSo, waLBerla, \esp, HemeLB*, Musubi, OpenLB, Palabos, Sailfish\\
+			Reusability & ESPResSo, waLBerla, \esp, HemeLB*, Musubi, OpenLB,
+			Palabos, Sailfish\\
 			\addlinespace[0.4cm]
-			\pbox{3.0cm}{Surface \\Understandability} & LUMA, DL\_MESO (LBE), lettuce, Musubi, Palabos\\
+			\pbox{3.0cm}{Surface \\Understandability} & LUMA, DL\_MESO (LBE),
+			lettuce, Musubi, Palabos\\
 			\addlinespace[0.4cm]
-			\pbox{3.0cm}{Visibility and \\Transparency} & Ludwig, ESPResSo, LUMA, Palabos\\
+			\pbox{3.0cm}{Visibility and \\Transparency} & Ludwig, ESPResSo,
+			LUMA, Palabos\\
 			\addlinespace[0.4cm]
@@
 		time of measurement, except for those marked by an asterisk (*)) (Sorted
-		by Order of Quality Measurement)}
-		\label{topperformerstable}
+		by Order of Quality Measurement)} \label{topperformerstable}
 	\end{center}
@@
 	of collaborators. (Section~\ref{Sec_CompareArtifacts})
-	\item Increase the use and enforcement of programming style guides to improve understandability and productivity. (Section~\ref{Sec_CompareArtifacts})
+	\item Increase the use and enforcement of programming style guides to
+	improve understandability and productivity.
+	(Section~\ref{Sec_CompareArtifacts})
 	\item Use checklists to ensure developers follow best practices.

### 2022-10-09 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/ResubmissionResponseToReviewer.tex
Changes for the resubmitted version of the SOP of LBM

@@
 \email{smiths@mcmaster.ca}
-\telephone{(905) 525-9140 ext.\ 24131}
+\telephone{(905) 525-9140 ext.\ 27929}
 \www{{http://}www.cas.mcmaster.ca/\~\@smiths}
@@
 
-  Thank you and the reviewer for the feedback on our submission.  In response to
-  your e-mail, dated August 26, 2022, we have revised the paper to incorporate
-  the requested revisions.  We provide a summary below.  We have also included
-  with the revised submission a version of the paper showing all additions and
-  deletions.
+  Thank you and the reviewer for the feedback on our submission.  The reviewer
+  provided thoughtful and constructive comments.  In response to your e-mail,
+  dated August 26, 2022, we have revised the paper to incorporate the requested
+  revisions.  We provide a summary below.  The revised submission also includes
+  a ``diff'' version of the paper showing the additions and deletions.
 
-  We would like to thank you and the reviewer for their thoughtful and
-  constructive comments.
+  While addressing the reviewer's comments, we noticed a minor inconsistency in
+  several of our figures.  In the original submission we inadvertently used the
+  AHP scores from our 10\% sensitivity analysis, rather than using the collected
+  data directly. Correcting this in the resubmission does not alter any of our
+  conclusions, but it does change the appearance of the figures for the AHP
+  scores (Figures 3--12). Some summary tables also had to be modified, since
+  with less artificial noise in the data there are more ties in the scores and
+  some minor shifting in the rankings.  The ``diff'' version of the paper shows
+  the minor changes in the discussion that arose from correcting the pairwise
+  comparison data.  In the diff versions of Tables 7 and 11 show many additions
+  and deletions, but this is because the difference tool bases its calculations
+  on the syntax.  Although there are more ties in the corrected scores and some
+  minor shifts in position, the general trends in the rankings are unchanged.
 
@@
   \item I think it would be useful to have slightly more by way of explanation
-  here -- to provide a bit more understanding of the process followed for
+  here --- to provide a bit more understanding of the process followed for
   readers who choose not to pursue the details provided in the reference.  Two
@@
 
-    \emph{Response. \smallskip}
+    \emph{In addition to removing the anonymization that hid some details, we have expanded Section 2.6 where we explain how we used Saaty's Analytic Hierarchy Process (AHP).
+Specifically, we explain AHP's central idea of pairwise comparison, the process for performing an AHP ranking, and how we map from our subjective scores to Saaty's scores.
+We also include, in Table 3, sample calculations.
+We also added text to Section 3.11 (Overall Quality), to explain how we use the priority ranking of qualities to combine the separate quality rankings into an overall ranking.\smallskip}
 
-  \item [I]n Fig 2, I see responses for "Descriptive error messages?" of yes, no,
+  \item In Fig 2, I see responses for ``Descriptive error messages?'' of yes, no,
   and n/a.  I have a hard time imagining what n/a could mean in this context.
@@
 
-    \emph{Response. \smallskip}
+    \emph{The source of the confusion here is a space-saving edit to the question we used for collecting data.
+The original question for grading the software was: ``If install fails were there descriptive error messages?'' To make the text fit in the limited space for the paper, we changed this to ``Descriptive error messages?''.
+In retrospect, we can see how this is confusing.
+The case of n/a applies when there were no errors in the installation, but the short-form version removed this key information.
+The revised Figure 2 text removes the confusion by restoring the original text: ``If install fails were there descriptive error messages?''\smallskip}
 
@@
 
-    \emph{Response. \smallskip}
+    \emph{Yes, we agree with the reviewer.  As mentioned above, Section 2.6 and 3.11 were expanded to give details on AHP. \smallskip}
 
@@
   equally, but in the graphs of the scores for various qualities (e.g., Figs
-  3-6), there are dramatic differences in the ranges of the scores given to the
-  packages -- anywhere from 0.06 as the top score to nearly 0.16.  The highest
+  3--6), there are dramatic differences in the ranges of the scores given to the
+  packages --- anywhere from 0.06 as the top score to nearly 0.16.  The highest
   overall score is ~0.075 (Fig 12). So what is the *theoretical* top score in
@@
 
-    \emph{Response. \smallskip}
+    \emph{The meaning of the AHP scores should be clearer now that we have
+    explained the details of the method.  In our revisions we added the
+    following sentence (in Section 2.6) to address the reviewer's concern: ``For
+    the AHP method, the sum of the final $s_j$ values for a given quality is
+    1.0.''  As a consequence of this, a low top score occurs when there is
+    little to distinguish the packages from one another.  The reviewer is
+    correct that we weigh each of the qualities equally, but that is only
+    relevant when we are combining the quality rankings into an overall ranking,
+    as described in Section 3.11. \smallskip}
 

### 2022-10-09 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Changes for the resubmitted version of the SOP of LBM

@@
 is, $s_j = \frac{1}{n} \sum_{i=1}^{i=n}(B_{j i})$, as shown in the final column
-of Table~\ref{Tbl_SampleAHP}. Sorting the $s_j$ values provides a ranking for
+of Table~\ref{Tbl_SampleAHP}. For the AHP method, the sum of the final $s_j$ values for a given quality is 1.0. Sorting the $s_j$ values provides a ranking for
 the software packages for the quality under investigation.  The $s_j$ values can
@@
 Current LBM software mostly follows best practices, as illustrated by
-Table~\ref{topperformerstable}, which summarizes the top two packages for each
-quality and for the overall quality. Taking the union of all entries in the
-table that appear under 2 or more qualities, we have 16 top performers:
-ESPREsSo, MechSys, OpenLB, DL\_MESO (LBE), Sailfish, Ludwig, Musubi, pyLBM,
-waLBerla, laboetie, lettuce, \esp, LUMA, Palabos, TCLB and LB2D-Prime.
+Table~\ref{topperformerstable}, which summarizes the packages ranked first and
+second for each quality and for the overall quality. Taking the union of all
+entries in the table that appear under 2 or more qualities, we have 16 top
+performers: ESPREsSo, MechSys, OpenLB, DL\_MESO (LBE), Sailfish, Ludwig, Musubi,
+pyLBM, waLBerla, laboetie, lettuce, \esp, LUMA, Palabos, TCLB and LB2D-Prime.
 Therefore, two thirds (67\%) of the 24 packages appear in the top two for at

### 2022-10-12 StateOfPractice/Papers/MedImageSoft_SOP-arXiv/MedImageSoft_SOP_AcademicRadiology.tex
Edits to SOP for MI, mainly section 9 (pain points), some parts removed from Section 10 (recommendations)

@@
 
-We based uur interviews on a list of 20 questions, which can be found in
+We based our interviews on a list of 20 questions, which can be found in
 \citet{SmithEtAl2021}. Some questions are about the background of the software,
@@
 software \citep{Michalski2021} and is apparently greater than the level of
-testing for ocean modelling software.  \citet{JungEtAl2022} reports that testing
-is underemphasized in ocean modelling.
-
-As the observed artifacts in Table~\ref{artifactspresent} show, documentation is
-not emphasized as part of the development process in any of the 29 projects.
-None of them had theory manuals, although we did identify a road map in the
-\textit{3D Slicer} project.  No requirements specifications were found.
-Table~\ref{tab_opinion_doc} summarizes interviewees' opinions on documentation.
-Interviewees from each of the eight projects thought that documentation was
-essential to their projects, and most of them said that it could save their time
-to answer questions from users and developers. Most of them saw the need to
-improve their documentation, and only three of them thought that their
-documentations conveyed information clearly enough. Nearly half of developers
-also believed that the lack of time prevented them from improving documentation.
+testing for ocean modelling software.  \citet{JungEtAl2022} reports that ocean
+modellers underemphasize testing.
+
+As the observed artifacts in Table~\ref{artifactspresent} show, none of the 29
+projects emphasize documentation. None of them had theory manuals, although we
+did identify a road map in the \textit{3D Slicer} project.  We did not find
+requirements specifications. Table~\ref{tab_opinion_doc} summarizes
+interviewees' opinions on documentation. Interviewees from each of the eight
+projects thought that documentation was essential to their projects, and most of
+them said that it could save their time to answer questions from users and
+developers. Most of them saw the need to improve their documentation, and only
+three of them thought that their documentations conveyed information clearly
+enough. Nearly half of developers also believed that the lack of time prevented
+them from improving documentation.
 
@@
 guidelines recommend a document explaining how to contribute to a project, often
-named CONTRIBUTING. Guidelins also recommend tutorials, user guides and quick start examples. \citet{SmithAndRoscoe2018} suggests including instructions
+named CONTRIBUTING. Guidelines also recommend tutorials, user guides and quick
+start examples. \citet{SmithAndRoscoe2018} suggests including instructions
 specifically for on-boarding new developers. For open source software in general
@@
 
-Our interviews identified pain points related to lack of time and funding,
+Our interviews identified pain points related to a lack of time and funding,
 technology hurdles, improving correctness, and improving usability.  In this
@@
 MI developer strategies for improving maintainability and reproducibility.
-Although these two qualities were not explicitly idenfied as pain points, they
-were discussed as part of our interview process \citep{SmithEtAl2021}.  The
-interviewee's practices for addressing pain points and improving quality can
-potentially be emulated by other MI developers. Moreover, these practices may
-provide examples that can be followed by other research software domains.
-
-[CHECK  THESE, after writing all the PPs] \citet{PintoEtAl2018} lists some pain
-points that did not come up in our conversations with MI developers :
+Although the interviewees did not explicitly identify these two qualities as
+pain points, they were discussed as part of our interview process
+\citep{SmithEtAl2021}.  The interviewee's practices for addressing pain points
+and improving quality can potentially be emulated by other MI developers.
+Moreover, these practices may provide examples that can be followed by other
+research software domains.
+
+\wss{CHECK  THESE, after writing all the PPs} \citet{PintoEtAl2018} lists some pain
+points that did not come up in our conversations with MI developers:
 Cross-platform compatibility, interruptions while coding, scope bloat, lack of
@@
 dependency management, data handling concerns (like data quality, data
-management and data privacy) [Actually mentioned by MI software],
+management and data privacy) \wss{Actually mentioned by MI software},
 reproducibility, and software scope determination. Although MI developers did
 not mention these pain points, we cannot conclude that they are not relevant for
-MI software development, since we only interviewed nine MI developers for about
+MI software development, since we only interviewed nine developers for about
 an hour each.
@@
 \item Supporting third-party plugins and extensions; and,
-\item Using GitHub Actions for continunous integration and continuous
+\item Using GitHub Actions for continuous integration and continuous
 development.
@@
 challenging than finding funding for research. The interviewee from the
-\textit{ITK-SNAP} team thought more fundings was a way to solve the lack of time
+\textit{ITK-SNAP} team thought more funding was a way to solve the lack of time
 problem, because they could hire more dedicated developers. On the other hand,
-the interviewee from the \textit{Weasis} team did not feel that fundings could
-solve the same problem, since he would still need time to supervise the project.
+the interviewee from the \textit{Weasis} team did not feel that funding could
+solve the same problem, since they would still need time to supervise the project.
 
@@
 include: hard to keep up with changes in OS and libraries, difficult to transfer
-to new technologies, hard to support multiple OS, and hard to support lower-end
+to new technologies, hard to support multiple OSes, and hard to support lower-end
 computers. Developers expressed difficulty balancing between four factors:
@@
 \item Using computing power from the computers GPU for web applications;
-\item Increasing funding;
+%\item Increasing funding;
 \item Maintaining better documentations to ease the development and maintenance
@@
 interviewed, three of them were building web applications, and the
-\textit{MRIcroGL} team was considering web-based solutions.
+\textit{MRIcroGL} team was considering a web-based solution.
 
@@
 
-**start editing here** Change to others have trouble getting developers to test.
-That isn't the problem for MI.  Testing is also a pain point for other research
-software domains point. For instance, \citet{PintoEtAl2018} mention the problem
-of insufficient testing and \citet{HannayEtAl2009} show that more developers
-think testing is important than the number that believe they have a sufficient
-understanding of testing concepts. \citet{WieseEtAl2019} mention data handling
-concerns (like data quality, data management and data privacy).
-\citet{HannayEtAl2009, KanewalaAndBieman2013, KellyEtAl2011, WieseEtAl2019}
-point to the oracle problem, which occurs in research software when we do not
-have a means to judge the correctness of the calculated solutions. For the
-specific case of LBM developers, testing challenges include difficulty with
-testing the correctness of large numbers of features, and challenges with
-automated testing, since free testing services do not offer adequate facilities
-for large amounts of data \citep{SmithEtAl2022}.
-
-Testing was the most often mentioned strategy for MI developers for ensuring
-correctness.  Seven teams mentioned test related activities, including
-test-driven development, component tests, integration tests, smoke tests,
-regression tests, self tests and automated tests.  Another approach frequently
-adopted (mentioned by 3 interviewees) is a two state development process with
-stable releases and nightly builds.  Other strategies for ensuring correctness
-that came up during the interviews include CI/CD, using de-identified copies of
-medical images for debugging, sending beta versions to medical workers who can
-access the data to do the tests, and collecting/maintaining a dataset of
-problematic images.  Some additional strategies used by MI developers include:
+As implied by the above threats to correctness, testing was the most often
+mentioned strategy for MI developers for ensuring correctness.  Seven teams
+mentioned test related activities, including test-driven development, component
+tests, integration tests, smoke tests, regression tests, self tests and
+automated tests.  With the common emphasis on testing to improve correctness, MI
+software is ahead of some other scientific domains.  For scientific software in
+general \citet{PintoEtAl2018} mention the problem of insufficient testing and
+\citet{HannayEtAl2009} show that more developers think testing is important than
+the number that believe they have a sufficient understanding of testing
+concepts.  Our study of LBM software suggests that this domain shares the
+challenges of insufficient testing and insufficient understanding of testing
+concepts \citep{SmithEtAl2022}. Automated testing is a specific challenge for
+LBM software since free testing services do not offer adequate facilities for
+large amounts of data \citep{SmithEtAl2022}. Although not specifically mentioned
+during our interviews, the large data sets for MI likely cause a challenge for
+using free testing services, like GitHub Actions.
+
+Research software in general often struggles with the oracle problem for testing
+because for many potential test cases the developer doesn't have a means to
+judge the correctness of their calculated solutions \citep{HannayEtAl2009,
+KanewalaAndBieman2013, KellyEtAl2011, WieseEtAl2019}.  The MI developers did not
+allude to this challenge, likely because for a give image (test case) it is
+possible to determine, possibly using other software, the expected analysis
+results.
+
+A frequently cited strategy for building confidence in correctness (mentioned by
+3 interviewees) is a two state development process with stable releases and
+nightly builds.  Other strategies for ensuring correctness that came up during
+the interviews include CI/CD, using de-identified copies of medical images for
+debugging, sending beta versions to medical workers who can access the data to
+do the tests, and collecting/maintaining a dataset of problematic images.  Some
+additional strategies used by MI developers include:
 
 \begin{itemize}
-\item Using open datasets
-\item Asking the users to provide de-identified copies of medical images if they
-have problems loading the images;
-\item Sending the beta versions of software to medical workers who can access
-the data and complete the tests;
+\item Using open datasets, like the datasets for the liver \citep{BilicEtAl2019}
+and brain \citep{MenzeEtAl2015} tumor segmentation benchmarks.
 \item If (part of) the team belongs to a medical school or a hospital, using the
@@
 their help on data format problems;
-\item Storing all images that cause special problems, and maintaining this
-special dataset over time.
 \end{itemize}
@@
 
-% Hard to maintain good documentations % from Ao
-% \begin{itemize}
-%     \item Hard to have a high level roadmap from the start
-%     \item Only a few people fully understand the large codebase
-% \end{itemize}
-
 \item[P\refstepcounter{pnum}\thepnum \label{P_Usability}:]
-\textbf{Usability:} Several developers noted that users sometimes try to use
-incorrect LBM method combinations to solve their problems. Furthermore, some
-users think that the packages will work out of the box to solve their cases,
-while in reality CFD knowledge needs to be applied to correctly modify the
-packages for the new endeavour.  Some respondents in the survey of
-\citet{WieseEtAl2019} also mentioned that users do not always have the expertise
-required to install or use the software.
-
-%\item Not enough participants for usability tests %from Ao
-%Hard to understand users' needs % from Ao
-
-The discussion with the developers focused on understandability issues for two
-classes of users: the end users and other developers.  The threats to
-understandability for end users include users not understanding how to use
-features (\wss{Because the users don't have sufficient background?  Ask Ao.})
-\ad{Did you mean medical or medical imaging background? No, it was majorly
-caused by the software usability. Based on the context of our interview
-conversations, it means that users don't understand how to complete certain
-operations or what functions a feature provides. It's due to unclear
-description, lack of documentation, poorly designed UI or nonintuitive process},
-the team not having a dedicated user experience (UX) designer, some important
-indicators are not noticeable (e.g. a progress bar), not all users understand
-the purpose of the software, not all users know if the software includes certain
-features, not all users understand how to use the command line tool, not all
-users understand that the software is a web application. For developers the
-threats to understandability include developers not understanding how to deploy
-the software (\wss{This sounds like a consequence of a lack of
-understandability, rather than threat to it.  Or is the point that developers
-lack sufficient background?  Ask Ao}) \ad{Yes, this should be a consequence of a
-lack of understandability. I think the direct cause is that the project doesn't
-clearly explain how to deploy. We did not discuss the background of the
-developers at that point.}, and the architecture is difficult for new developers
-to understand.
-
-The most common strategy, cited by 4 developers, for ensuring understandability
-was to use documentation (user manuals, mailing lists, forums).  Other suggested
-and practiced strategies include a graphical user interface, testing every
-release with active users, making simple things simple and complicated things
-possible, icons with clear visual expressions, designing the software to be
-intuitive, having a UX designer with the right experience, dialog windows for
-important notifications, and providing an example for users to follow.
-
-The two most frequently mentioned strategies (mentioned 3 times each) for
-improving usability are:
-\begin{inparaenum}[i)]
-    \item usability tests and interviews with end users; and,
-    \item adjusting the software according to user feedback.
-\end{inparaenum}
-Other ideas that came up include a straightforward and intuitively designed
-interface / professional UX designer, providing step-by-step processes, making
-the basic functions easy to use without reading the documentation, focusing on
-limited number of functions, making the software more streamlined, downsampling
-images to consume less memory, and an option to load only part of the data to
-boost performance.
+\textbf{Usability:}
+
+The discussion with the developers focused on usability issues for two classes
+of users: the end users and other developers.  The threats to usability for end
+users include an unintuitive user interface, inadequate feedback from the
+interface (such as lack of a progress bar), users being unable to determine the purpose of
+the software, not all users knowing if the software includes certain features, not
+all users understanding how to use the command line tool, and not all users
+understanding that the software is a web application. For developers the threats to
+usability include not being able to find clear instructions on how to deploy the
+software, and the architecture being difficult for new developers to understand.
+
+At least to some extent the problems for MI software users are due to holes in
+their background knowledge.  The survey of \citet{WieseEtAl2019} for research
+software in general also mentioned that users do not always have the expertise
+required to install or use the software. \citet{SmithEtAl2022} observes a
+similar pattern for LBM software, with several LBM developers noting that users
+sometimes try to use incorrect method combinations. Furthermore, some LBM users
+think that the packages will work out of the box to solve their cases, while in
+reality CFD knowledge needs to be applied to correctly modify the packages for a
+new endeavour.
+
+To improve the usability of MI software, the most common strategies mentioned by
+developers are as follows:
+
+\begin{itemize}[i)]
+    \item use documentation (user manuals, mailing lists, forums) (mentioned by
+    4 developers)
+    \item usability tests and interviews with end users; and, (mentioned by 3
+    developers)
+    \item adjusting the software according to user feedback. (mentioned by 3
+    developers)
+\end{itemize}
+
+Other suggested and practiced strategies include a graphical user interface,
+testing every release with active users, making simple things simple and
+complicated things possible, focusing on limited number of functions, icons with
+clear visual expressions, designing the software to be intuitive, having a UX
+designer, dialog windows for important notifications, providing an example for
+users to follow, downsampling images to consume less memory, and an option to
+load only part of the data to boost performance.  The last two points recognize
+that an important component of usability is performance, since poor performance
+frustrates users.
 
@@
 
-blurb to explain the ones that are coming were not explicitly identified as pain
-points.
+Up to this point, we have covered the pain points that came up in interviews
+with MI developers, along with a summary of the techniques that are currently
+used to address these pain points.  Although the developers did not explicitly
+identify the qualities of maintainability and reproducibility as pain points in
+our interviews, as part of our interview questions
+(Section~\ref{sec_interview_methods}) they did share their approaches for
+improving these qualities, as discussed below.
 
 \begin{enumerate}
-\item \textbf{Maintainability:} The developer of ESPResSo said that they
-originally wrote their source code with a specific application in mind, which
-later caused too much coupling between components in the source code. This
-resulted in technical debt \citep{KruchtenEtAl2012}, which has an impact on
-future maintainability and reusability. Concern with technical debt is likely
-why researchers in the survey of \citet{Nguyen-HoanEtAl2010} rate
-maintainability as the third most important software quality. More recently the
-push for sustainable software \citep{deSouzaEtAl2019} is motivated by the pain
-that past developers have had with accumulating too much technical debt.
-
-The most popular, with five out of nine interviewees mentioning it, strategy for
-ensuring maintainability was to use a modular approach, with often repeated
+\item \textbf{Maintainability:} Although not explicitly highlighted as a pain
+point during our interviews, \citet{Nguyen-HoanEtAl2010} rate maintainability as
+the third most important software quality for research software in general. The push for
+sustainable software \citep{deSouzaEtAl2019} is motivated by the pain that past
+developers have had with accumulating too much technical debt
+\citep{KruchtenEtAl2012}.  For LBM software, \citet{SmithEtAl2022} identifies
+technical debt as one of the developer pain points.
+
+To improve maintainability, the most popular (with five out of nine interviewees
+mentioning it) strategy is to use a modular approach, with often repeated
 functions in a library.  Other strategies that were mentioned for improving
@@
 
-\item \textbf{Reproducibility:}  Interviewees stressed the importance of
-documentation for both users and developers throughout the interviews. They
-emphasize that a lack of time and funding (\ppref{P_LackDevTime}) has a negative
-effect on the documentation. Most of the developers are scientific researchers
-evaluated on the scientific papers that they produce. Writing and updating
-documentation is something that is done in their free time, if that time arises.
-Others also mention inadequate research software documentation
-\citep{PintoEtAl2018, WieseEtAl2019}.  The problem also arises with non-research
-software \citep{LethbridgeEtAl2003}. Recommendations on the development of
-research software state that developers should critically evaluate their own
-development processes in terms of quality assurance and comply with
-international standards for software documentation
-\citep{KaterbowAndFeulner2018}.
-
-We discussed threats to reproducibility and strategies for improving it.  The
-threats that were mentioned include closed-source software, no user interaction
-tests, no unit tests, using different versions of some common libraries
-\wss{because the interface and/or behaviour of the library could have changed?
-Ask Ao} \ad{Yes. Behaviour change could be caused by a lib changing to use
-different algorithms or different dependencies}, variability
-between CPUs, and misinterpretation of how manufacturers
-create medical images. The most commonly cited (by 6 teams) strategy to improve
-reproducibility was testing (regression tests, unit tests, having good tests).
-The second most common strategy (mentioned by 5 teams) is making code, data, and
-documentation available, possibly by creating open-source libraries.  Other
-ideas that were mentioned include running the same tests on all platforms, a
-dockerized version of the software to insulate it from the OS environment, using
-standard libraries, monitoring the upgrades of the library dependencies, clearly
+\item \textbf{Reproducibility:}  Although the MI developers did not mention
+reproducibility explicitly as a pain point, they did mention the need to improve
+documentation.  Good documentation doesn't just address the pain points of lack
+of developer time (\ppref{P_LackDevTime}), technology hurdles
+(\ppref{P_TechnologyHurdles}), usability \ppref{P_Usability}, and
+Maintainability.  Documentation is also necessary for reproducibility. The
+challenges of inadequate documentation are a known problem for research software
+\citep{PintoEtAl2018, WieseEtAl2019} and for non-research software
+\citep{LethbridgeEtAl2003}.
+
+In our interviews, we discussed threats to reproducibility and strategies for
+improving it.  The threats that were mentioned include closed-source software,
+no user interaction tests, no unit tests, the need to change versions of some
+common libraries, variability between CPUs, and misinterpretation of how
+manufacturers create medical images.
+
+The most commonly cited (by 6 teams) strategy to improve reproducibility was
+testing (regression tests, unit tests, having good tests). The second most
+common strategy (mentioned by 5 teams) is making code, data, and documentation
+available, possibly by creating open-source libraries.  Other ideas that were
+mentioned include running the same tests on all platforms, a dockerized version
+of the software to insulate it from the OS environment, using standard
+libraries, monitoring the upgrades of the library dependencies, clearly
 documenting the version information, bringing along the exact versions of all
@@
 
+- not repeating best practices mentioned previously, like CI/CD, version
+control, testing, etc.
+
 Based on our quality measurements in Section~\ref{ch_results} and discussions
@@
 recommendations for each quality are as follows:
-
-\begin{itemize}
-\item \textbf{Installability} (Section~\ref{sec_result_installability})
-\begin{itemize}
-    \item providing clear instructions;
-    \item automating installation;
-    \item including all dependencies in the installer;
-    \item avoiding heavily depending on other commercial products (e.g.\ Matlab);
-    \item potentially building a web application that needs no installation.
-\end{itemize}
-\item \textbf{Correctness \& Verifiability}
-(Section~\ref{sec_result_correctness_verifiability} and
-Section~\ref{sec_interview_software_qualities})
-\begin{itemize}
-    \item test-driven development with unit tests, integration tests, and
-    nightly tests;
-    \item two stage development process with stable release \& nightly builds;
-    \item CI/CD;
-    \item requirements specifications and theory manuals \citep{Smith2016, SmithAndLai2005}.
-    \item static code analysis tools (e.g. Lint and SonarQube)
-\end{itemize}
-\item \textbf{Reliability} (Section~\ref{sec_result_reliability})
-\begin{itemize}
-    \item test-driven development with unit tests, integration tests, and nightly tests.
-    \item two stage development process with stable release \& nightly builds;
-    \item descriptive error messages.
-\end{itemize}
-\item \textbf{Robustness} (Section~\ref{sec_result_robustness})
-\begin{itemize}
-    \item designing with exception handling so the software can fail gracefully;
-    \item descriptive error messages.
-\end{itemize}
-\item \textbf{Usability} (Section~\ref{sec_result_usability} and Section~\ref{sec_interview_software_qualities})
-\begin{itemize}
-    \item usability tests and interviews with end users;
-    \item adjusting according to users' feedbacks;
-    \item getting started tutorials;
-    \item user manuals;
-    \item professional UX (User eXperience) designs;
-    \item graphical user interface;
-    \item active supports to users.
-\end{itemize}
-\item \textbf{Maintainability} (Section~\ref{sec_score_maintainability} and Section~\ref{sec_interview_software_qualities})
-\begin{itemize}
-    \item use GitHub (or equivalent version control and project management environment);
-    \item use a modular approach to design (we advocate basing the design on
-    Parnas's principle of information hiding: ``system details that are likely
-    to change independently should be the secrets of separate modules; the only
-    assumptions that should appear in the interfaces between modules are those
-    that are considered unlikely to change.'' \citep{ParnasEtAl2000})
-    \item documentation for developers: project plan, developer’s manual, and API documentation.
-\end{itemize}
-\item \textbf{Reusability} (Section~\ref{sec_result_reusability})
-\begin{itemize}
-    \item modular approach;
-    \item API documentation;
-    \item tools that generate software documentation for developers (e.g.\
-    Doxygen, Javadoc, and Sphinx).
-\end{itemize}
-\item \textbf{Understandability} (Section~\ref{sec_result_understandability} and
-Section~\ref{sec_interview_software_qualities})
-\begin{itemize}
-    \item modular approach;
-    \item good coding style: consistent indentation and formatting style;
-    consistent, distinctive, and meaningful code identifiers; keeping parameters
-    in the same order for all functions; avoiding hard-coded constants (other
-    than 0 and 1);
-    \item clear comments, indicating what is being done, not how;
-    \item description of algorithms used;
-    \item documentation of explicit requirements on coding standards;
-    \item communication between developers and users via GitHub issues, mailing
-    lists, and forums.
-\end{itemize}
-\item \textbf{Visibility/Transparency} (Section~\ref{sec_result_visibility_transparency})
-\begin{itemize}
-    \item documents for the development process, contributor's guide, project
-    status, development environment, and release notes.
-\end{itemize}
-\item \textbf{Reproducibility} (Section~\ref{sec_interview_software_qualities})
-\begin{itemize}
-    \item test-driven development with unit tests, integration tests, and nightly tests.
-    \item open-source;
-    \item making data and documentation available;
-    \item using open-source libraries.
-\end{itemize}
-\end{itemize}
-
-\wss{Could ``reverse'' the list and list the recommendations and beside each the qualities 
-that that recommendation helps with - this would reduce repetition.}
-
-\subsection{Recommendations on Dealing With Limited Resources}
-\label{sec_recommendations_limited_resources}
-
-The limitation of resources has many faces, with the key manifestations being
-lack of fundings, time, and developers. We summarized our discussion with the MI
-software developers in Section~\ref{painpoints}.  Many of our recommendations
-involve investing more effort into processes, tools and methods, so that the
-limited resources can be used more effectively.  Our recommendations are as
-follows:
-
-\wss{Talk about increasing the number of developers.  CONTRIBUTING is rare
-(Table~\ref{artifactspresent}), but it doesn't have to be.  Related to this
-information for new developers is rare: Code Style Guide, Requirements, Design
-Doc, API.}
-
-\begin{itemize}
-\item \textbf{Identify the root cause.} More fundings or developers may not
-solve problems called by a lack of time. It is beneficial to identify the
-underlying obstacles to the team.  \wss{Is there something from the mythical man
-month that should be cited here?}
-
-\item \textbf{Maintain good documentation.} Creating and updating documentation
-consumes time, but can save much more time in the long term. If the users and
-developers can find answers to their questions themselves, they are less likely
-to abuse the team's issue tracker.  \wss{If it is an option, we could cite one
-of our productivity related papers.}
-
-\item \textbf{Adopt time-saving tools.} A good CI/CD tool (e.g.\ GitHub Actions)
-saves time for building and deploying the product, and automated tests can work
-in the background while developers are focusing on other tasks.
-
-\item \textbf{Use test-driven development process.} Many people think writing
-test cases is less rewarding than writing code, but without testing identifying
-and fixing bugs can consume substantial resources. Setting up the test cases
-costs time, but generates more benefits in the long run.
-
-\item \textbf{Consider supporting third-party plugins or extensions.} Why not
-let users share the burden? No software product can deliver every user's needs,
-and the large quantity of features leads to more bugs and maintenance problems.
-So it may be a good idea to shift some development and maintenance
-responsibilities to the users. The users may also be happy with the extra
-flexibility.
-
-\item \textbf{Consider ``hibernating'' for a while.} When there aren't enough
-developers, the team can shift from development mode toward maintenance mode.
-The team can stop building new features for a while and instead fix bugs and
-design problems from the past. If the development team can repay some of its
-technical debt \citep{KruchtenEtAl2012}, the software qualities will likely
-improve as a result.
-
-\item \textbf{Commercialization is not always toxic.} Licensing the software to
-commercial companies to use as internal modules of their products may bring
-financial support to the team. Meanwhile, the project can stay open-source for
-the community.
-\end{itemize}
 
@@
 
-\item \textbf{Maintain good documentation.} No matter what tech stack a team
-uses, a well-maintained project plan, developer's manual, and API documentation
-generally help team members to contribute more and make fewer mistakes.
 \end{itemize}

### 2022-10-12 StateOfPractice/Papers/MedImageSoft_SOP-arXiv/MedImageSoft_SOP_AcademicRadiology.tex
Minor edit to SOP for MI

@@
 
-\subsection{Recommendations on Choosing A Tech Stack}
+\subsection{Move To Web Applications}
 \label{sec_recommendations_tech_stack}
@@
 on the choice of a tech stack to improve the four priority factors identified by
-developers: compatibility, maintainability, performance and security.  The
+developers: compatibility, maintainability, performance, and security.  The
 suggestions are intended to provide ideas and avenues for exploration; not all
@@
 
-\item \textbf{Use git and GitHub.} As mentioned in
-Section~\ref{sec_score_maintainability}, almost all of the 29 MI software
-projects used git, and the majority of them used GitHub. We found from the
-projects' websites and our interviews with developers that, some projects moved
-from other version control tools to git and GitHub. GitHub provides convenient
-repository and project management, and OSS projects tend to receive more
-attention and contributions on GitHub.
-
 \item \textbf{Web applications can also deliver high performance.} Web
@@
 \item \textbf{Web transmission may diminish security.} Transferring sensitive
-data online can be a problem for projects requiring high security. Regulations
+data on-line can be a problem for projects requiring high security. Regulations
 for some MI applications may forbid doing web transmissions. In this case, a web

### 2022-10-13 StateOfPractice/Papers/MedImageSoft_SOP-arXiv/MedImageSoft_SOP_AcademicRadiology.tex
Revisions to MI for SOP, mainly in recommendationss

@@
 \newcommand{\ppref}[1]{P\ref{#1}}
+
+\newcounter{qnum} %quality number
+\newcommand{\qthepnum}{Q`'\theqnum}
+\newcommand{\qref}[1]{Q\ref{#1}}
 
@@
 MI software development, since we only interviewed nine developers for about
-an hour each.
+an hour each.  \wss{Add LBM pain points not covered.}
 
@@
 \begin{enumerate}
-\item \textbf{Maintainability:} Although not explicitly highlighted as a pain
-point during our interviews, \citet{Nguyen-HoanEtAl2010} rate maintainability as
-the third most important software quality for research software in general. The push for
-sustainable software \citep{deSouzaEtAl2019} is motivated by the pain that past
-developers have had with accumulating too much technical debt
+\item[Q\refstepcounter{qnum}\theqnum \label{Q_Maintainability}:]
+\textbf{Maintainability:} Although not explicitly highlighted as a pain point
+during our interviews, \citet{Nguyen-HoanEtAl2010} rate maintainability as the
+third most important software quality for research software in general. The push
+for sustainable software \citep{deSouzaEtAl2019} is motivated by the pain that
+past developers have had with accumulating too much technical debt
 \citep{KruchtenEtAl2012}.  For LBM software, \citet{SmithEtAl2022} identifies
@@
 
-\item \textbf{Reproducibility:}  Although the MI developers did not mention
+\item[Q\refstepcounter{qnum}\theqnum \label{Q_Reproducibility}:]
+\textbf{Reproducibility:}  Although the MI developers did not mention
 reproducibility explicitly as a pain point, they did mention the need to improve
@@
 
+- not repeating best practices mentioned previously, like CI/CD, version
+control, testing, etc.
+
 This section presents our recommendations on MI software development. Although
@@
 to any scientific computing software.
-Section~\ref{sec_recommendations_qualities} discusses the actions that can
-potentially improve the ten software qualities.
-Sections~\ref{sec_recommendations_limited_resources},
-\ref{sec_recommendations_tech_stack}, and
-\ref{sec_recommendations_testing_dataset} are based on the primary pain points
-(from Section~\ref{painpoints}) collected from developers in the
-MI domain.
-
-\subsection{Recommendations on Improving Software Qualities}
-\label{sec_recommendations_qualities}
+
+In this section we provide recommendations to address the pain points from
+Section~\ref{painpoints} to answer~\rqref{RQ_Recommend}.  Our recommendations
+are not lists of criticisms for what should have been done in the past, or what
+should be done now; they are suggestions for consideration in the future. We
+will not be repeating previously discussed ideas here, such as CI, documentation
+of APIs, etc.  Our aim is to mention ideas that are at least somewhat beyond
+conventional best practices. The ideas listed here have the potential to become
+best practices in the medium to long-term. The ideas in the following
+subsections are roughly in the order of increasing implementation effort.
+
+\subsection{Continuous Integration} \label{Sec_ContinuousIntegration}
+
+Continuous integration is the process of building and testing software on every
+push to the code repository, with pushes done very frequently \citep[p.\ 13]
+{HumbleAndFarley2010}, \citep{ShahinEtAl2017, Fowler2006}.  CI consists of the
+following elements:
+\begin{itemize}
+	\item A version control system \citep{Fowler2006}. To be effective, all
+	files should be under version control, not just code files.  Anything that
+	is needed to build, install and run the software should be under version
+	control, including configuration files, build scripts, test harnesses, and
+	operating system configurations \citep[p. 19]{HumbleAndFarley2010}.
+	Fortunately for the LBM domain, as discussed above, most alive projects
+	(73\% of those assessed) use version control.
+	\item A fully automated build system \citep{Fowler2006}.  As \citet[p.\
+	5]{HumbleAndFarley2010} point out, deploying software manually is an
+	anti-pattern.  Fortunately, as discussed above, most LBM projects use build
+	automation.
+	\item An automated test system \citep{Fowler2006}. Building quality software
+	involves creating automated tests at the unit, component, and acceptance
+	test level, and executing these tests whenever someone makes a change to the
+	code, its configuration, the environment, or the software stack that it runs
+	on \citep[p.\ 83]{HumbleAndFarley2010}. As Table~\ref{artifactspresent}
+	shows, test cases are in the uncommon category for LBM software artifacts,
+	which means that some LBM projects will need to increase their testing
+	automation if they wish to pursue CI.  The usual advice for CI is to keep
+	the build and test process short \citep[p.\ 60]{HumbleAndFarley2010}. Given
+	that LBM can be computationally expensive, the tests run with every check-in
+	may need to focus on simple code interface tests, saving large tests for
+	less frequent runs.  A more sophisticated option to address the bottleneck
+	for merges is CIVET (Continuous Integration, Verification, Enhancement, and
+	Testing), which solves this problem by intelligently pinning, cancelling,
+	and if necessary, restarting jobs as merges occur \citep{SlaughterEtAl2021}.
+	\item An automated system for other tasks, such as code checking,
+	documentation building and web-site updating.  These other tasks are not
+	essential to CI, but they can be incorporated to improve the quality
+	of the code and the communication between developers and users. For
+	instance, a static analysis (possibly via linters) of the code may find poor
+	programming practice or lack of adherence to adopted coding standards.
+	\citet{SlaughterEtAl2021} provides another example of an automated task ---
+	checking that a test specification includes the test's motivation, a test
+	description, and a design description for all changes. 
+	\item An integration build system to pull everything together.  Every time
+	there is a check-in (for instance a pull request), the integration server
+	automatically checks out the sources onto the integration machine, starts a
+	build, runs tests, and informs the committer of the results. 
+\end{itemize}
+
+Although CI can take some time and effort to set up and integrate into a team's
+workflow, the benefits can be significant, as follows:
+
+\begin{itemize}
+	\item Elimination of headaches associated with a separate integration phase
+	\citep{Fowler2006}, \citep[p.\ 20]{HumbleAndFarley2010}. If integration of
+	the work of different developers, or even separate chunks of work by the
+	same developer, is postponed, integration problems are inevitable.  By
+	continuously integrating, problems are immediately obvious and the source of
+	the problem can be isolated to the small increment that was just committed.
+	\item Bugs can be quickly detected and removed \citep{Fowler2006} via
+	automated testing.  To improve productivity, defects are best discovered and
+	fixed at the point where they are introduced \citep[p.\
+	23]{HumbleAndFarley2010}.  Code is not the only source of errors; they are
+	also found in the files and scripts related to configuration management
+	\citep[p.\ 18]{HumbleAndFarley2010}.
+	\item Developers are always working on a stable base, since the master
+	branch will always be working.  A stable base passes all tests and, if the
+	CI system uses generators and linters, it will also have current
+	documentation and coding standard compliant code.  A stable base improves
+	developer productivity, allowing them to focus on coding, testing, and
+	documentation.
+\end{itemize}
+
+Although there are still challenges, setting up a CI system has never been
+easier than it is today.  The option of installing a dedicated CI server (either
+physically or virtually) exists with tools such as
+\href{https://www.jenkins.io/} {Jenkins}, \href{http://buildbot.net/}
+{Buildbot}, \href{https://www.gocd.org/} {Go}, and
+\href{http://integrity.github.io/} {Integrity}. However, installation on your
+own server is unnecessary since there are many hosted CI tool solutions
+available, such as: \href{https://travis-ci.org/} {Travis CI},
+\href{https://github.com/features/actions} {GitHub Actions} and
+\href{https://circleci.com/} {CircleCI}.  Getting started with a hosted CI is
+straightforward.  All that is required to begin is editing a few lines of a YAML
+configuration file in the project's root directory.
+
+Challenges that exist for adopting CI include lack of awareness and
+transparency, coordination and collaboration challenges, lack of expertise and
+skills, more pressure and workload for team members, general resistance to
+change, scepticism and distrust on continuous practices \citep{ShahinEtAl2017}.
+The most common reason given for not adopting CI is that developers on my
+project are not familiar enough with CI \citep{HiltonEtAl2016}.  These problems
+can be mitigated via planning and documentation, promoting a team mindset,
+adopting new rules and policies, improving testing activities, and decomposing
+development into smaller units \citep{ShahinEtAl2017}.
+
+\subsection{Employ Linters} \label{Sec_Linters}
+
+Except for \citet{ThielEtAl2020}, the research software guidelines that we
+consulted do not mention linters.  However, we believe that linters have the
+potential to improve code quality at a relatively low cost.  A linter is a tool
+that statically analyzes code to find programming errors, stylistic
+inconsistencies and suspicious constructs \citep{Wikipedia2022_Lint}. Linters
+can be used to spot check code files, or even better as part of a continuous
+integration system, as discussed in Section~\ref{Sec_ContinuousIntegration}.
+
+A sample of the potential benefits of linters include: finding memory leaks,
+finding potential bugs, standardizing code with respect to formatting, improving
+performance, removing silly errors before code reviews, and catching potential
+security issues \citep{SourceLevel2022_Lint}.  Linters are available for most
+popular programming languages.  For instance Python has the options of PyLint,
+flake8 and Black \citep{Zadka2018}.
+
+Linters can address the pain points for LBM developers.  For instance, a linter
+can decrease the amount of development time (\ppref{P_LackDevTime}) by
+decreasing the number of mundane mistakes programmers have to catch.  Since the
+linter can include rules that capture the wisdom of senior programmers, it can
+help newer developers avoid common mistakes. Although there are not many linters
+for checking parallel programming mistakes (\ppref{P_TechnologyHurdles}),
+Parallel Lint is an option for OpenMP. Consistently using a linter can guard
+against some technical debt (\ppref{Q_Maintainability}). Although linters are
+tools for code analysis, similar ideas can be applied to ``lint'' documentation
+to ensure adherence to basic rules.  The use of tools to check documentation is
+a partial explanation for the relatively higher quality of statistical tools
+that are part of the Comprehensive R Archive Network (CRAN)
+\citep{SmithEtAl2018_StatSoft}.
+
+\subsection{Conduct Rigorous Peer Reviews} \label{Sec_PeerReview}
+
+Peer review is a strategy already adopted by some LBM developers to improve
+confidence in their software, but the benefits of peer review can be improved by
+making its application more rigorous. \citet{Jones2008} shows that rigorous
+inspection finds 60-65\% of latent defects on average, and often tops 85\% in
+defect removal efficiency.  By way of comparison, most forms of testing average
+between 30 and 35\% for defect removal efficiency \citep{EbertAndJones2009,
+Jones2008}.  A formal code inspection involves asking the reviewer to either
+follow a review checklist (check consistency of variable names, look for
+terminating loops, etc.), or perform specific review tasks (like summarize the
+purpose of the code, create a data dictionary for a given module,
+cross-reference the code to the technical manual, etc.) The task based
+inspection approach has been effectively used for research software, as
+described by \citet{KellyAndShepard2000}. Task based inspection is an ideal fit
+with an issue tracking system, like GitHub.  The review tasks can be issues, so
+that they can be easily assigned, monitored and recorded.  Other potential
+issues for the tracker include assigning junior team members to test
+installation instructions and getting-started tutorials.
+
+Rigorous peer review addresses the same pain points as linters
+(Section~\ref{Sec_Linters}): \ppref{P_LackDevTime}, \ppref{P_TechnologyHurdles},
+and \ppref{Q_Maintainability}. Peer review addresses these points by efficiently
+searching for defects and problems.  Finding misunderstandings in of how the
+code implements the required theory improves the software's correctness
+(\ppref{P_Correctness}).
 
@@
 
-- not repeating best practices mentioned previously, like CI/CD, version
-control, testing, etc.
-
-Based on our quality measurements in Section~\ref{ch_results} and discussions
-with the developers in Section~\ref{painpoints}, we collected key points for
-improving software qualities. These points should be considered for new and
-existing projects, but we are not saying that every project should do everything
-on our list.  All projects will have finite resources, so strategies will have
-to be selected that provide the greatest return on investment.  Our specific
-recommendations for each quality are as follows:
-
-\subsection{Move To Web Applications}
-\label{sec_recommendations_tech_stack}
+\subsection{Move To Web Applications} \label{sec_recommendations_tech_stack}
 
@@
 \end{itemize}
+
+\subsection{Follow Advice From the Open-Source Community to Grow the Number of Contributors} \label{Sec_GrowContributors}
+
+In our interviews with developers we heard they would like to have additional
+software contributors.  More developers would help with addressing the lack of
+development time pain point (\ppref{P_LackDevTime}).  In investigating advice on
+increasing the number of contributors, we found that the advice usually starts
+with following best practices, including providing artifacts like those
+discussed in Section~\ref{Sec_CompareArtifacts}, such as a contributor
+guidelines, code style guidelines, a clear code of conduct and high quality code
+and documentation. This advice is logical, but in our assessment of LBM software
+we found examples of projects that already follow many best practices, but still
+have small development teams. Attracting developers apparently requires more
+than just building a good product.
+
+Some potential additional ideas for growing the number of contributors are
+as follows:
+
+\begin{itemize}
+	\item Clearly identify issues that are an appropriate starting points for
+	new developers \citep{Garcia2016, Jalan2016, Proffitt2017}.
+	\item Given that in most projects developers were once users
+	\citep{McQuaid2018}, recruit future developers from the current set of
+	users.
+	\item Create templates for pull requests and issues \citep{Jalan2016}.
+    \item Welcome all kinds of contributions, not just code.  Non-code
+    contributions include documentation, fixing typos, issue reporting and test
+    cases \citep{Jalan2016, Proffitt2017}.
+    \item Reward and recognize new contributors, via small rewards like stickers
+    or shirts, or even just a simple shoutout or mention in a blog post or on
+    social media \citep{Jalan2016, Proffitt2017}.
+    \item Recognize that open source is more about people than it will ever be
+    about code \citep{Jalan2016}.
+    \item Look beyond just recruiting online and seek new developers at
+    conferences or other user meet ups \citep{Garcia2016}.
+    \item Follow the advice of \citet{Kuchner2012} to adapt ideas from marketing
+    to promote science.  Specific ideas that could be applied to open source
+    projects include the following: 
+    \begin{itemize}
+		\item Storytelling to motivate interest in the project, where the story
+		is a sequence of events and pauses for reflection \citep[p.\
+		21--22]{Kuchner2012}.
+		\item Invest effort in building relationships with users and with potential future developers.
+		\item Brand your project with what makes it unique and special.
+	\end{itemize}
+	\item Consider combining existing projects to pool collective resources, and
+	reduce competition between different open source projects.
+\end{itemize}
+
+\subsection{Design For Change} \label{Sec_DesForChange}
+
+To address technical debt (\ppref{Q_Maintainability}), the top LBM developers show how
+modularization can be used to design research software for future change.
+Although the advice to modularize research software to handle complexity is
+common \citep{WilsonEtAl2014, StewartEtAl2017, Storer2017}, specific guidelines
+on how to divide the software into modules is less prevalent.  Not every
+decomposition is a good design for supporting change, as shown by
+\citet{Parnas1972a}.  A design with low cohesion and high coupling \citep[p.\
+48]{GhezziEtAl2003} will make change difficult. Especially in research software,
+where change is inevitable, designers need to produce a modularization that
+supports change. Ocean modelling software is currently feeling the pain of
+not emphasizing modularization in legacy code \citep{JungEtAl2022}.
+
+Interviewed LBM developers highlighted cases where their modularizations
+anticipated future changes.  The developer of pyLBM mentioned that the
+geometries and models of their system had been ``decoupled'', using abstraction
+and modularization of the source code, to make it ``very easy to add [new]
+features''.  The pyLBM design allows for independent changes to the geometry and
+the model.  We also learned that the package pyLBM redeveloped data structures
+to ease future changes. The developer of TCLB noted that their design allows for
+the addition of some LBM features, but changes to major aspects of the system
+would be difficult. For example, ``implementing a new model will be an easy
+contribution'', but changes to the ``Cartesian mesh … will be a nightmare''. The
+design of TCLB highlights that not every conceivable change needs to be
+supported, only the likely changes.
+
+As the LBM developers illustrate, they accomplish design for change by first
+identifying likely changes, either implicitly or explicitly, and second by
+hiding each likely change behind a well-defined module interface.  Although it
+is unclear whether the developers are aware of Parnas's work, the approach
+mirrors his recommendations \citep{Parnas1972a}. \citet{SmithEtAl2022} (Section
+Documentation as Part of the Development Process) lists ideas for how to
+document the design, including the likely changes, so that they are more visible
+to others.
+
+\subsection{Assurance Case} \label{AssuranceCases}
+
+From \cite{RinehartEtAl2015}, an assurance case is ``[a] reasoned and compelling
+argument, supported by a body of evidence, that a system, service or
+organization will operate as intended for a defined application in a defined
+environment.''  Assurance cases have been successfully employed for safety
+critical systems, but using this technique for SC software is a new idea.  The
+Goal Structuring Notation (GSN) \cite{Spriggs2012} seems like a good framework
+for initial investigation.
+
+Scientific software, such as medical software, is often subject to
+standardization and regulatory approval. While applying such approvals and
+standards has had a beneficial effect on system quality, it does not provide
+good tracking of the development stages, as the compliance with the standards
+is mostly checked after the system development. Once a system is implemented,
+its documentations must be approved by the regulators. This process is lengthy
+and expensive. In contrast, assurance case development usually occurs in
+parallel with the system construction, resulting in a traceable, detailed
+argument for the desired property.  Moreover, since the assurance case is
+integrated into the system construction, the domain experts are involved from
+the start.
+
+Putting the argument in the hands of the experts means that they will work to
+convince themselves, along with the regulators.  They will use the expertise
+that the regulators do not have and they will see the value of documentation (as
+mentioned in section on pain points with methods to mitigate); they will be
+engaged.  This engagement will hopefully help bridge the current chasm between
+software engineering and scientific computing~\cite{Kelly2007}, by motivating
+scientists toward documentation and correcting the problem of software engineers
+failing to meet scientists' expectations~\cite{Segal2008}.  Significant
+documentation will still likely be necessary, but now the developers control the
+documentation.  What is created will be relevant and necessary.
+
+For SCS a top level claim could read ``Program X delivers correct outputs when
+used for its intended use/purpose in its intended environment.''  The next step
+would be to decompose this claim into sub-claims that will be easier to prove.
+The sub-claims will likely also be further divided until the bottom of the
+graph, where the measurable evidence is provided.  Typical evidence will consist
+of documents, expert reviews, test case results etc.
+
+Using the specific example of preparing an assurance case for pre-existing
+medical image analysis software (3dfim+)~\cite{SmithEtAl2018_ICSEPoster}, the
+top level can be decomposed into four sub-goals, as shown in
+Figure~\ref{TopGoal}.  This example follows the same pattern as used for medical
+devices~\cite{Wassyng2015}.  The first sub-goal (GR) argues for the quality of
+the documentation of the requirements (SRS).  The second sub-goal (GD) says that
+the design complies with the requirements and the third proposes that the
+implementation also complies with the requirements.  The fourth sub-goal (GA)
+claims that the inputs to 3dfim+ will satisfy the operational assumptions, since
+we need valid input to make an argument for the correctness of the output.
+
+\begin{figure*}[!h]
+\centering
+\includegraphics[width=1.0\textwidth]{./figures/TopGoal.pdf}
+\caption{Top Goal of the assurance case and its sub-goals}
+\label{TopGoal}
+\end{figure*}
+
+Preparing an assurance case for the pre-existing 3dfim+ software justifies the
+value of assurance cases for the certification of SCS~\cite{SmithEtAl2018}.
+Although no errors were found in the output of the existing software, the rigour
+of the proposed approach did lead to finding ambiguities and omissions in the
+existing documentation, such as missing information on the coordinate system
+convention.  In addition, a potential concern for the software itself was
+identified from the GA argument: running the software does not produce any
+warning about the obligation of the user to provide data that matches the
+parametric statistical model employed for the correlation calculations.
+
+Potential objectives for research on assurance cases applied to SCS are as
+follows:
+
+\begin{enumerate}
+\item Creation of an assurance case template will require looking at a variety
+  of SCS projects, which will require engaged project partners; therefore, the
+  initial approach should be to target domains where certification is required,
+  or domains where verification is critical, but challenging, like the High
+  Energy Physics (HEP). \wss{reference the paper about domain specific
+  methodology - maybe I already did?}
+\item Evidence is provided at the bottom of an assurance case, but what is
+  the best form for this evidence?  As mentioned for the SRS, we need an
+  approach for verification.  We also need an approach to verify the other
+  software artifacts, the traceability between them, the team of developers, the
+  team of reviewers, etc.  The question for the future is what combination of
+  tools and processes will provide convincing evidence?
+\item Creation of assurance cases is challenging with existing tools.  In the
+  future, the goal should be automatic generation (see
+  \citet{Smith2018}, Section SecAutoGenerate) of the visual assurance case documentation from
+  the arguments and evidence
+\item Creating a revelation on the value of documentation \citep{Smith2018} may
+   require a pseudo adversarial approach.  SCS developers should be explicitly
+   challenged to present their verification efforts so that their work can be
+   independently verified.  Although SCS developers have developed many
+   successful theories, techniques, and testing procedures for verification, the
+   evidence is generally presented in an ad hoc manner.  Developers should be
+   asked to connect all the pieces of their evidence in a coherent argument.  As
+   a bonus, the act of creating the assurance case may also lead developers to
+   discover subtle edge cases, which would not have been noticed with a less
+   rigorous and systematic approach.
+
+\end{enumerate}
+
+\subsection{Generate All Things} \label{Sec_GenAllThings}
+
+We propose automatically generating LBM code and its documentation, using a
+Domain Specific Modelling (DSM) approach. DSM means creating a knowledge base of
+models for physics, computing, mathematics, documentation, and certification and
+then writing explicit ``recipes'' that weave together this knowledge to generate
+the desired code, documentation, test cases, inspection reports and build
+scripts. This definition of DSM is more general than usual; in this
+recommendation DSM implies generation of all software artifacts, not just the
+code. DSM moves development to a higher level of abstraction; domain experts can
+work without concern for low-level implementation details. Using DSM, we
+optimally generate code and documentation, eliminate redundancy, reduce the
+likelihood of errors, and automate maintenance. Moreover, a generative approach
+facilitates the inevitable exploration of LBM modelling assumptions.
+
+DSM provides a transformative technology for documentation, design, and
+verification \citep{JohansonAndHasselbring2018, Smith2018}. DSM allows
+scientists to focus on their science, not software.  A generative approach
+removes the maintenance nightmare of documentation duplicates and near
+duplicates \citep{LucivEtAl2018}, since developers capture knowledge once and
+transform it as needed.  Code generation has previously been applied to improve
+research software.  For instance, ATLAS (Automatically Tuned Linear Algebra
+Software) \citep{WhaleyEtAl2001} and Blitz++ \citep{Veldhuizen1998} produces
+efficient and portable linear algebra software.  Spiral \citep{Pueschel2001}
+uses software/hardware generation for digital signal processing.
+\citet{Carette2008} shows how to generate a family of efficient, type-safe
+Gaussian elimination algorithms.  FEniCS (Finite Element and Computational
+Software) uses code generation when solving differential equations
+\citep{LoggEtAl2012}. \citet{OberEtAl2018} and \citet{MatkerimEtAl2013} apply
+DSM to High Performance Computing (HPC), using UML (Unified Modelling Language)
+for their domain models. Unlike previous DSM work, the current recommendation
+focuses on generating all software artifacts (requirements, design, etc.), not
+just code.  \citet{SzymczakEtAl2016} presents initial work on this ``generate
+all things'' approach, \citet{SmithAndCarette2021-BRIC} presents a motivating
+example, and \citep{CaretteEtAl2021-Drasil} provides a prototype.
+
+A DSM approach addresses multiple pain points.  For instance, once the
+infrastructure is in place, a DSM can decrease the amount of development time
+(\ppref{P_LackDevTime}) by automation.  Since a DSM approach allows scientists
+to focus on their science, rather than software, the lack of software
+development experience will be less of an issue. The DSM approach can capture
+computing knowledge to mitigate the technology related pain points
+(\ppref{P_TechnologyHurdles}).  For ensuring correctness (\ppref{P_Correctness})
+a generative approach can aim to be correct by construction.  If there are
+mistakes, DSM has the advantage that they are propagated throughout the
+generated artifacts, which greatly increases the chance that someone will notice
+the mistake.  Technical debt (\ppref{Q_Maintainability}) is not a concern with a
+DSM approach since developers write the recipes used for generation at a high
+level making them relatively easy to change.  With respect to pain points on
+usability (\ppref{P_Usability}), the generative approach we are proposing
+addresses these directly, since documentation is a primary concern, rather than
+an afterthought.
 

### 2022-10-13 StateOfPractice/Papers/MedImageSoft_SOP-arXiv/MedImageSoft_SOP_AcademicRadiology.tex
Updates to first two subsections sections of the recommendations section

@@
 
-\wss{CHECK  THESE, after writing all the PPs} \citet{PintoEtAl2018} lists some pain
-points that did not come up in our conversations with MI developers:
-Cross-platform compatibility, interruptions while coding, scope bloat, lack of
-user feedback, hard to collaborate on software projects, and aloneness.
-\citet{WieseEtAl2019} repeat some previous pain points and add the following:
-dependency management, data handling concerns (like data quality, data
-management and data privacy) \wss{Actually mentioned by MI software},
-reproducibility, and software scope determination. Although MI developers did
-not mention these pain points, we cannot conclude that they are not relevant for
-MI software development, since we only interviewed nine developers for about
-an hour each.  \wss{Add LBM pain points not covered.}
+\citet{PintoEtAl2018} lists some pain points that did not come up in our
+conversations with MI developers: interruptions while coding, scope bloat, lack
+of user feedback, hard to collaborate on software projects, and aloneness.
+\citet{WieseEtAl2019} also mention two research software pain points that did
+not explicitly arise in our interviews: reproducibility, and software scope
+determination.  To the list of pain points not discussed for MI, our study of
+LBM software \citep{SmithEtAl2022} adds lack of software experience for the
+developers, technical debt, and documentation. We did not observe any pain
+points for MI that were not also observed for LBM. From the pain points
+mentioned above, although the topics of reproducibility and technical debt did
+not come up in our MI interviews, these two topics were discussed as part of the
+discussion of software qualities, as covered at the end of this section.
+Although previous studies show pain points that were not mentioned by MI
+developers, we cannot conclude that these pain points are not relevant for MI
+software development, since we only interviewed nine developers for about an
+hour each.
 
@@
 
-The pain point survey of \citet{WieseEtAl2019} also highlighted
-technical-related problems like dependency management, cross-platform
-compatibility, CI, hardware issues and operating system issues. From
+The pain point survey of \citet{WieseEtAl2019} highlights that technology
+hurdles are an issue for research software in general.  Some of the
+technical-related problems mentioned by \citet{WieseEtAl2019} include dependency
+management, cross-platform compatibility (also mentioned by
+\citet{PintoEtAl2018}), CI, hardware issues and operating system issues. From
 \citep{SmithEtAl2022} technology pain points for LBM developers include setting
@@
 \item Using computing power from the computers GPU for web applications;
-%\item Increasing funding;
 \item Maintaining better documentations to ease the development and maintenance
@@
 \item Lack of real world image data for testing, in part because of patient
-privacy concerns;
+privacy concerns (\citet{WieseEtAl2019} mentions that the pain point of privacy
+concerns also arises for research software in general);
 \item Tests are expensive and time-consuming because of the need for huge datasets;
@@
 
-\begin{itemize}[i)]
-    \item use documentation (user manuals, mailing lists, forums) (mentioned by
+\begin{itemize}
+    \item Use documentation (user manuals, mailing lists, forums) (mentioned by
     4 developers)
-    \item usability tests and interviews with end users; and, (mentioned by 3
+    \item Usability tests and interviews with end users; and, (mentioned by 3
     developers)
-    \item adjusting the software according to user feedback. (mentioned by 3
+    \item Adjusting the software according to user feedback. (mentioned by 3
     developers)
@@
 
-- add in ACs, borrow from previous paper.
-- mention the solutions proposed by LBM developers, like design for change
-
-- not repeating best practices mentioned previously, like CI/CD, version
-control, testing, etc.
-
-This section presents our recommendations on MI software development. Although
-our focus is on MI software, unless noted otherwise, our recommendations apply
-to any scientific computing software.
-
 In this section we provide recommendations to address the pain points from
@@
 should be done now; they are suggestions for consideration in the future. We
-will not be repeating previously discussed ideas here, such as CI, documentation
-of APIs, etc.  Our aim is to mention ideas that are at least somewhat beyond
+expand on some of the ideas that came out of our interviews with developers
+(Section~\ref{painpoints}), including continuous integration, moving to web
+applications, and enriching the test data sets. We also bring in new ideas from
+our experience like employing linters, peer review, design for change and
+assurance cases.  Our aim is to mention ideas that are at least somewhat beyond
 conventional best practices. The ideas listed here have the potential to become
-best practices in the medium to long-term. The ideas in the following
-subsections are roughly in the order of increasing implementation effort.
+best practices in the medium to long-term. We list the ideas roughly in the
+order of increasing implementation effort.
 
@@
 
-Continuous integration is the process of building and testing software on every
-push to the code repository, with pushes done very frequently \citep[p.\ 13]
-{HumbleAndFarley2010}, \citep{ShahinEtAl2017, Fowler2006}.  CI consists of the
-following elements:
+Continuous integration involves frequent pushes to a code repository.  With
+every push the software is built and tested \citep[p.\ 13]
+{HumbleAndFarley2010}, \citep{ShahinEtAl2017, Fowler2006}.  CI can take
+significant time and effort to set up and integrate into a team's workflow, but
+the benefits are significant, as follows:
+
+\begin{itemize}
+	\item Elimination of headaches associated with a separate integration phase
+	\citep{Fowler2006}, \citep[p.\ 20]{HumbleAndFarley2010}. If integration is
+	postponed, integration problems are inevitable.  Continuous integration
+	means that problems are immediately obvious and the source of the problem
+	can be isolated to the small increment that was just committed.
+	\item Detection and removal of bugs \citep{Fowler2006} via
+	automated testing.  To improve productivity, defects are best discovered and
+	fixed at the point where they are introduced \citep[p.\
+	23]{HumbleAndFarley2010}.  Code is not the only source of errors; they are
+	also found in the files and scripts related to configuration management
+	\citep[p.\ 18]{HumbleAndFarley2010}.
+	\item Everyone is always working on a stable base, since the rejection of
+	inadequate commits means that the main branch will always be working.  A
+	stable base will always pass all tests.  If the CI system uses generators
+	and linters, it will also have current documentation and coding standard
+	compliant code.  A stable base improves developer productivity, allowing
+	them to focus on coding, testing, and documentation.
+\end{itemize}
+
+CI consists of the following elements:
+
 \begin{itemize}
@@
 	operating system configurations \citep[p. 19]{HumbleAndFarley2010}.
-	Fortunately for the LBM domain, as discussed above, most alive projects
-	(73\% of those assessed) use version control.
+	Fortunately for the MI, as shown in Section~\ref{sec_score_maintainability}
+	all our measured projects use version control.
 	\item A fully automated build system \citep{Fowler2006}.  As \citet[p.\
 	5]{HumbleAndFarley2010} point out, deploying software manually is an
-	anti-pattern.  Fortunately, as discussed above, most LBM projects use build
-	automation.
+	anti-pattern.  For MI software, Table~\ref{artifactspresent} shows 18 of 29
+	packages (62\%) were observed to include build scripts.  Projects without a
+	build system will need to add one to pursue using CI.
 	\item An automated test system \citep{Fowler2006}. Building quality software
@@
 	on \citep[p.\ 83]{HumbleAndFarley2010}. As Table~\ref{artifactspresent}
-	shows, test cases are in the uncommon category for LBM software artifacts,
-	which means that some LBM projects will need to increase their testing
-	automation if they wish to pursue CI.  The usual advice for CI is to keep
-	the build and test process short \citep[p.\ 60]{HumbleAndFarley2010}. Given
-	that LBM can be computationally expensive, the tests run with every check-in
-	may need to focus on simple code interface tests, saving large tests for
-	less frequent runs.  A more sophisticated option to address the bottleneck
-	for merges is CIVET (Continuous Integration, Verification, Enhancement, and
-	Testing), which solves this problem by intelligently pinning, cancelling,
-	and if necessary, restarting jobs as merges occur \citep{SlaughterEtAl2021}.
+	shows, test cases are in the uncommon category for MI software artifacts,
+	which means that some MI projects will need to increase their testing
+	automation if they wish to pursue CI.
 	\item An automated system for other tasks, such as code checking,
@@
 	programming practice or lack of adherence to adopted coding standards.
-	\citet{SlaughterEtAl2021} provides another example of an automated task ---
-	checking that a test specification includes the test's motivation, a test
-	description, and a design description for all changes. 
 	\item An integration build system to pull everything together.  Every time
@@
 
-Although CI can take some time and effort to set up and integrate into a team's
-workflow, the benefits can be significant, as follows:
-
-\begin{itemize}
-	\item Elimination of headaches associated with a separate integration phase
-	\citep{Fowler2006}, \citep[p.\ 20]{HumbleAndFarley2010}. If integration of
-	the work of different developers, or even separate chunks of work by the
-	same developer, is postponed, integration problems are inevitable.  By
-	continuously integrating, problems are immediately obvious and the source of
-	the problem can be isolated to the small increment that was just committed.
-	\item Bugs can be quickly detected and removed \citep{Fowler2006} via
-	automated testing.  To improve productivity, defects are best discovered and
-	fixed at the point where they are introduced \citep[p.\
-	23]{HumbleAndFarley2010}.  Code is not the only source of errors; they are
-	also found in the files and scripts related to configuration management
-	\citep[p.\ 18]{HumbleAndFarley2010}.
-	\item Developers are always working on a stable base, since the master
-	branch will always be working.  A stable base passes all tests and, if the
-	CI system uses generators and linters, it will also have current
-	documentation and coding standard compliant code.  A stable base improves
-	developer productivity, allowing them to focus on coding, testing, and
-	documentation.
-\end{itemize}
-
-Although there are still challenges, setting up a CI system has never been
-easier than it is today.  The option of installing a dedicated CI server (either
-physically or virtually) exists with tools such as
+To enable incorporation into a team's workflow, \citet[p.\
+60]{HumbleAndFarley2010} explain that the usual approach for CI is to keep the
+build and test process short. Since MI files are large, the tests run with every
+check-in may need to focus on simple code interface tests, saving large tests
+for less frequent execution.  A more sophisticated option to address the
+bottleneck for merges is CIVET (Continuous Integration, Verification,
+Enhancement, and Testing), which solves this problem by intelligently pinning,
+cancelling, and if necessary, restarting jobs as merges occur
+\citep{SlaughterEtAl2021}. A more sophisticated process management system can
+also enforce rules for pull requests, like checking that a test specification
+includes the test's motivation, a test description, and a design description for
+all changes \citet{SlaughterEtAl2021}.
+
+Setting up a CI system has never been easier than it is today.  A dedicated CI
+server (either physically or virtually) can be installed with tools such as
 \href{https://www.jenkins.io/} {Jenkins}, \href{http://buildbot.net/}
@@
 \href{http://integrity.github.io/} {Integrity}. However, installation on your
-own server is unnecessary since there are many hosted CI tool solutions
-available, such as: \href{https://travis-ci.org/} {Travis CI},
+own server is often unnecessary since there are many hosted CI solutions, such
+as: \href{https://travis-ci.org/} {Travis CI},
 \href{https://github.com/features/actions} {GitHub Actions} and
-\href{https://circleci.com/} {CircleCI}.  Getting started with a hosted CI is
-straightforward.  All that is required to begin is editing a few lines of a YAML
+\href{https://circleci.com/} {CircleCI}.  All that is required to begin using a
+hosted CI is to select the service and then edit a few lines of a YAML
 configuration file in the project's root directory.
 
-Challenges that exist for adopting CI include lack of awareness and
-transparency, coordination and collaboration challenges, lack of expertise and
-skills, more pressure and workload for team members, general resistance to
-change, scepticism and distrust on continuous practices \citep{ShahinEtAl2017}.
-The most common reason given for not adopting CI is that developers on my
-project are not familiar enough with CI \citep{HiltonEtAl2016}.  These problems
-can be mitigated via planning and documentation, promoting a team mindset,
-adopting new rules and policies, improving testing activities, and decomposing
-development into smaller units \citep{ShahinEtAl2017}.
+\citet{ShahinEtAl2017} highlights the following challenges for adopting CI: lack
+of awareness and transparency, lack of expertise and skills, coordination and
+collaboration challenges, more pressure and workload for team members, general
+resistance to change, scepticism and distrust on continuous practices. The most
+common reason given for not adopting CI is that developers are not familiar
+enough with CI \citep{HiltonEtAl2016}.  \citet{ShahinEtAl2017} observes that
+these problems can be mitigated via improving testing activities, planning and
+documentation, promoting a team mindset, adopting new rules and policies, and
+decomposing development into smaller units.
+
+\subsection{Move To Web Applications} \label{sec_recommendations_tech_stack}
+
+Section~\ref{painpoints} lists the advantages and disadvantages between native
+and web applications. Here we give further suggestions deciding whether to adopt
+a web application tech stack.  The goal is to improve the four priority factors
+identified by developers: compatibility, maintainability, performance, and
+security.  The suggestions are intended to provide ideas and avenues for
+exploration; a web application might not be the right fit for all projects and
+all teams.
+
+\begin{itemize}
+\item \textbf{Identify the priorities between the factors.} Simultaneously
+achieving high levels for all four factors (compatibility, maintainability,
+performance and security) is difficult. To enable decision making, a team needs
+to prioritize its objectives according to its resource and experience.
+
+\item \textbf{Be open-minded about new technologies.} Web applications with only
+a frontend are known for worse performance than native applications. However,
+new technologies may ease this difference. For example, some JavaScript
+libraries can help the frontend harness the power of the computer's GPU and
+accelerate graphical computing. In addition, there are new frameworks helping
+developers with cross-platform compatibility. For example, the
+\href{https://flutter.dev/}{Flutter} project enables support for web, mobile,
+and desktop OS with one codebase.
+
+\item \textbf{Web applications can also deliver high performance.} Web
+applications with backend servers may perform even better than native
+applications. If a team needs to support lower-end computers, it is good to use
+back-end servers for heavy computing tasks.
+
+\item \textbf{Backend servers can have low costs.} Serverless solutions from
+major cloud service providers may be worth exploring. Serverless solutions still
+use a server, but the team is only charged when they use it. The solution is
+event-driven, and costs the team by the number of requests it processes. Thus,
+serverless can be very cost-effective for less intensively used functions.
+
+\item \textbf{Web transmission may diminish security.} Transferring sensitive
+data on-line can be a problem for projects requiring high security. Regulations
+for some MI applications may forbid doing web transmissions. In this case, a web
+application with a backend may not be an option.
+
+\end{itemize}
+
+\subsection{Recommendations on Enriching the Testing Datasets} 
+\label{sec_recommendations_testing_dataset}
+
+As described in Section~\ref{painpoints}, it is difficult for software
+development teams to access real-world medical imaging datasets. This problem
+restricts their capability and flexibility for testing. We provide some
+suggestions as follows:
+
+\begin{itemize}
+\item \textbf{Build and maintain good connections to datasets.} A team can build
+connections with professionals working in the medical domain, who may have
+access to private datasets and can perform tests for the team. If a team has
+such professionals as internal members, the process can be simplified.
+
+\item \textbf{Collect and maintain datasets over time.} A team may face problems
+caused by various unique inputs over the years of software development. This
+data should be collected and maintained over time to form a good, comprehensive,
+dataset for testing.
+
+\item \textbf{Search for open data sources.} In general, there are many open MI
+datasets.  For instance, there are
+\href{https://nihcc.app.box.com/v/ChestXray-NIHCC}{Chest X-ray Datasets} by
+National Institute of Health \citep{WangEtAl2017},
+\href{https://www.cancerimagingarchive.net/}{Cancer Imaging Archive}
+\citep{PriorEtAl2017}, and \href{https://medpix.nlm.nih.gov/home}{MedPix} by
+National Library of Medicine \citep{Smirniotopoulos2014}. A team developing MI
+software should be able to find more open datasets according to their needs.
+
+\item \textbf{Create sample data for testing.} If a team can access tools
+creating sample data, they may also self-build datasets for testing. For
+example, an MI software development team can use an MRI scanner to create images
+of objects, animals, and volunteers. The team can build the images based on
+specific testing requirements.
+
+\item \textbf{Remove privacy from sensitive data.} For data with sensitive
+information, a team can ask the data owner to remove such information or add
+noise to protect privacy. One example is using de-identified copies of medical
+images for testing.
+
+\item \textbf{Establish community collaboration in the domain.} During our
+interviews with developers in the MI domain, we heard many stories of asking for
+supports from other professionals or equipment manufacturers. However, we
+believe that broader collaboration between development teams can address this
+problem better. Some datasets are too sensitive to share, but if the community
+has some kind of ``group discussion'', teams can better express their needs, and
+professionals can better offer voluntary support for testing. Ultimately, the
+community can establish a nonprofit organization as a third-party, which
+maintains large datasets, tests OSS in the domain, and protects privacy.
+
+\end{itemize}
 
@@
 USGS2019}.
-
-\subsection{Move To Web Applications} \label{sec_recommendations_tech_stack}
-
-A tech stack refers to a set of technologies used by a team to build software
-and manage the project. Section~\ref{painpoints} lists the advantages and
-disadvantages between native and web applications. We give further suggestions
-on the choice of a tech stack to improve the four priority factors identified by
-developers: compatibility, maintainability, performance, and security.  The
-suggestions are intended to provide ideas and avenues for exploration; not all
-of the suggestions will be the right fit for all projects and all teams.
-
-\begin{itemize}
-\item \textbf{Identify the priorities between the factors.} Simultaneously
-achieving high levels for all four factors (compatibility, maintainability,
-performance and security) is difficult. A team needs to prioritize its
-objectives according to its resource and experience.
-
-\item \textbf{Be open-minded about new technologies.} Web applications with only
-a frontend are known for worse performance than native applications. However,
-new technologies may ease this difference. For example, some JavaScript
-libraries can help the frontend harness the power of the computer's GPU and
-accelerate graphical computing. In addition, there are new frameworks helping
-developers with cross-platform compatibility. For example, the
-\href{https://flutter.dev/}{Flutter} project enables support for web, mobile,
-and desktop OS with one codebase.
-
-\item \textbf{Web applications can also deliver high performance.} Web
-applications with backend servers may perform even better than native
-applications. If a team needs to support lower-end computers, it is good to use
-back-end servers for heavy computing tasks.
-
-\item \textbf{Backend servers can have low costs.} Serverless solutions from
-major cloud service providers may be worth exploring. Serverless still uses a
-server, but the team is only charged when they use it. The solution is
-event-driven, and costs the team by the number of requests it processes. Thus,
-serverless can be very cost-effective for less intensively used functions.
-
-\item \textbf{Web transmission may diminish security.} Transferring sensitive
-data on-line can be a problem for projects requiring high security. Regulations
-for some MI applications may forbid doing web transmissions. In this case, a web
-application with a backend may not be an option.
-
-\end{itemize}
-
-\subsection{Recommendations on Enriching the Testing Datasets} 
-\label{sec_recommendations_testing_dataset}
-
-As described in Section~\ref{painpoints}, it is difficult for software
-development teams to access real-world medical imaging datasets. This problem
-restricts their capability and flexibility for testing. We provide some
-suggestions as follows:
-
-\begin{itemize}
-\item \textbf{Build and maintain good connections to datasets.} A team can build
-connections with professionals working in the medical domain, who may have
-access to private datasets and can perform tests for the team. If a team has
-such professionals as internal members, the process can be simplified.
-
-\item \textbf{Collect and maintain datasets over time.} A team may face problems
-caused by various unique inputs over the years of software development. This
-data should be collected and maintained over time to form a good, comprehensive,
-dataset for testing.
-
-\item \textbf{Search for open data sources.} In general, there are many open MI
-datasets.  For instance, there are
-\href{https://nihcc.app.box.com/v/ChestXray-NIHCC}{Chest X-ray Datasets} by
-National Institute of Health \citep{WangEtAl2017},
-\href{https://www.cancerimagingarchive.net/}{Cancer Imaging Archive}
-\citep{PriorEtAl2017}, and \href{https://medpix.nlm.nih.gov/home}{MedPix} by
-National Library of Medicine \citep{Smirniotopoulos2014}. A team developing MI
-software should be able to find more open datasets according to their needs.
-
-\item \textbf{Create sample data for testing.} If a team can access tools
-creating sample data, they may also self-build datasets for testing. For
-example, an MI software development team can use an MRI scanner to create images
-of objects, animals, and volunteers. The team can build the images based on
-specific testing requirements.
-
-\item \textbf{Remove privacy from sensitive data.} For data with sensitive
-information, a team can ask the data owner to remove such information or add
-noise to protect privacy. One example is using de-identified copies of medical
-images for testing.
-
-\item \textbf{Establish community collaboration in the domain.} During our
-interviews with developers in the MI domain, we heard many stories of asking for
-supports from other professionals or equipment manufacturers. However, we
-believe that broader collaboration between development teams can address this
-problem better. Some datasets are too sensitive to share, but if the community
-has some kind of ``group discussion'', teams can better express their needs, and
-professionals can better offer voluntary support for testing. Ultimately, the
-community can establish a nonprofit organization as a third-party, which
-maintains large datasets, tests OSS in the domain, and protects privacy.
-
-\end{itemize}
 

### 2022-10-14 StateOfPractice/Papers/MedImageSoft_SOP-arXiv/MedImageSoft_SOP_AcademicRadiology.tex
Update to MI for SOP for AHP Calculations description and web app recommendation

@@
 
-Thomas L.\ Saaty developed AHP in the 1970s, and people have widely used it
-since to make and analyze multiple criteria decisions \citep{VaidyaEtAl2006}.
-AHP organizes multiple criteria factors in a hierarchical structure and uses
-pairwise comparisons between alternatives to calculate relative ratios
-\citep{Saaty1990}. We use AHP to generate a ranking for a set of software
-packages.
-
-For a project with $m$ criteria, we can use an $m \times m$ matrix $A$ to record
-the relative importance between factors. When comparing criterion $i$ and
-criterion $j$, the value of $A_{ij}$ is decided as follows, with the value of
-$A_{ji}$ generally equal to $1/A_{ij}$ \citep{Saaty1990}: $A_{ij} = 1$ if
-criterion $i$ and criterion $j$ are equally important, while $A_{ij} = 9$ if
+Saaty developed AHP in the 1970s, and people have widely used it since to make
+and analyze multiple criteria decisions \citep{VaidyaEtAl2006}. AHP organizes
+multiple criteria in a hierarchical structure and uses pairwise comparisons
+between alternatives to calculate relative ratios \citep{Saaty1990}. AHP works
+with sets of $n$ \textit{options} and $m$ \textit{criteria}.  In our project
+$n=29$ and $m=9$ since there are 29 options (software products) and 9 criteria
+(qualities). We rank the software for each of the qualities, and then we combine
+the quality rankings into an overall ranking based on the relative priorities
+between qualities.
+
+The first step for ranking the software choices for a given quality involves a
+pairwise comparison between each of the $n$ software options for that quality.
+AHP expresses the comparison through an $n \times n$ matrix $A$. When comparing
+option $i$ and option $j$, the value of $A_{ij}$ is decided as follows, with the
+value of $A_{ji}$ generally equal to $1/A_{ij}$ \citep{Saaty1990}: $A_{ij} = 1$
+if criterion $i$ and criterion $j$ are equally important, while $A_{ij} = 9$ if
 criterion $i$ is extremely more important than criterion $j$.  The natural
 numbers between 1 and 9 are used to show the different levels of relative
-importance between these two extremes. The above assumes that criterion $i$ is
-not less important than criterion $j$.  If that is not the case, we reverse $i$
-and $j$ and determine $A_{ji}$ first, then $A_{ij} = 1/A_{ji}$.
-
-The priority vector $w$, which ranks the criteria by their importance, can be
-calculated by solving the equation \citep{Saaty1990}:
+importance between these two extremes. The above assumes that option $i$ is of
+equal, or more, importance compared to option $j$ ($i \geq j$).  If that is not
+the case, we reverse $i$ and $j$ and determine $A_{ji}$ first, then $A_{ij} =
+1/A_{ji}$.
+
+Section~\ref{sec_grading_software} shows how we measure the software via a
+grading template.  For the AHP process, the relevant measure is the subjective
+score from $1$ to $10$ for each quality for each package. To turn these
+subjective measures $x_{\text{sub}}$ and $y_{\text{sub}}$ into Saaty's
+pair-wise scores for option $x$ versus option $y$, respectively, we use the
+following calculation:
+\[
+\begin{cases}
+\min\{9, x_{\text{sub}} - y_{\text{sub}} + 1\} & x_{\text{sub}} \geq y_{\text{sub}} \\
+1 / \min\{9, y_{\text{sub}} - x_{\text{sub}} + 1\} & x_{\text{sub}} < y_{\text{sub}}
+\end{cases}
+\]
+
+\noindent For example, we measured the usability for 3D Slicer and Ginkgo CADx
+as $8$ and $7$, respectively; therefore, on the 9-point scale, 3D Slicer compared
+to Ginkgo CADx is 2 and Ginkgo CADX versus 3D Slicer is 1/2, as shown in the
+sample AHP calculations (Table~\ref{Tbl_SampleAHP}).
+
+The second step is to calculate the priority vector $w$ from $A$.  The
+vector $w$ ranks the software options by how well they achieve the given
+quality.  The priority vector can be calculated by solving the equation
+\citep{Saaty1990}:
 \begin{equation} 
@@
 \begin{equation}
-w_i = \frac{1}{m}\sum_{j=1}^{m}\frac{A_{ij}}{\sum_{k=1}^{m}A_{kj}}
+w_i = \frac{1}{n}\sum_{j=1}^{n}\frac{A_{ij}}{\sum_{k=1}^{n}A_{kj}}
 \end{equation}
 
-If there are $n$ alternatives, for criterion $i = 1, 2, ... , m$, we can create
-an $n\times n$ matrix $B_i$ to record the relative preferences between these
-choices for each of the $m$ criterion. The way of generating $B_i$ is similar to
-the one for $A$. However, rather than comparing the importance between criteria,
-we pairwise decide how much we favour one alternative over the other. We use the
-same method to calculate the local priority vector for each $B_i$.  The local
-priority vector in this case ranks the $n$ alternatives for criterion $i$.
-
-In this project, the first nine software qualities mentioned in Section
-\ref{sec_software_quality} are the criteria ($m = 9$), while 29 software
-packages ($n = 29$) are compared for each of the $m$ criteria.
-Section~\ref{sec_grading_software} shows the evaluation via the grading
-template, including a subjective score from $1$ to $10$ for each quality for
-each package. For each quality, for a pair of packages $i$ and $j$, with
-$\mathit{score}_i >= \mathit{score}_j$, the difference between the two scores is
-$\mathit{diff_{ij}} = \mathit{score}_i - \mathit{score}_j$. The mapping between
-$\mathit{diff_{ij}}$ (which can vary between 0 and 9) and the values in $A_{ij}$
-(which can vary between 1 and 9) is as follows:
-
-\begin{itemize}
-\item $A_{ij} = 1$ and $\mathit{diff_{ij}} = 0$ when criterion $i$ and criterion
-$j$ are equally important;
-\item $A_{ij}$ increases when $\mathit{diff_{ij}}$ increases;
-\item $A_{ij} = 9$ and $\mathit{diff_{ij}} = 9$ when criterion $i$ is extremely
-more important than criterion $j$.
-\end{itemize}
-
-\noindent Thus, we approximate the pairwise comparison result of $i$ versus $j$
-by the following equation:
-
-\begin{equation}
-A_{ij} = \text{min}(\mathit{score}_i - \mathit{score}_j + 1, 9)
-\end{equation}
+Table~\ref{Tbl_SampleAHP} summarizes the above two steps for the quality of
+installability.  The matrix $A$ is shown in the first set of columns, then the
+normalized version of $A$ and finally the average of the normalized values to
+form the vector $w$ in the last column.
+
+\begin{table}[h!]
+\begin{center}
+\begin{tabular}{ l c c c c c | c c c c c | c }
+ \toprule
+ ~ & \multicolumn{5}{c|}{$A_{ij}$} & \multicolumn{5}{c|}{${A_{ij}}/{\sum_{k=1}^{n}A_{kj}}$} & ~\\
+ \midrule
+ ~ & \rotatebox{90}{3D Slicer} & \rotatebox{90}{Ginkgo} & \rotatebox{90}{XMedCon} & $\cdots$ & \rotatebox{90}{Gwyddion} & \rotatebox{90}{3D Slicer} & \rotatebox{90}{Ginkgo} & \rotatebox{90}{XMedCon} & $\cdots$ & \rotatebox{90}{Gwyddion} & AVG \\
+ \midrule
+ 3D Slicer & 1 & 2 & 4 & $\cdots$ & 2 & 0.071 & 0.078 & 0.060 & $\cdots$ & 0.078 & 0.068\\
+ Ginkgo & 1/2 & 1 & 3 & $\cdots$ & 1 & 0.036 & 0.039 & 0.045 & $\cdots$ & 0.039 & 0.041\\
+ XMedCon & 1/4 & 1/3 & 1 & $\cdots$ & 1/3 & 0.018 & 0.013 & 0.015 & $\cdots$ & 0.013 & 0.015\\
+ $\vdots$ & $\vdots$ & $\vdots$ & $\vdots$ & $\ddots$ & $\vdots$ & $\vdots$ & $\vdots$ & $\vdots$ & $\ddots$ & $\vdots$ & $\vdots$\\
+ Gwyddion & 1/2 & 1 & 3 & $\cdots$ & 1 & 0.036 & 0.039 & 0.045 & $\cdots$ & 0.039 & 0.041\\  
+ \midrule
+ SUM = & 14.01 & 25.58 & 66.75 & $\cdots$ & 25.58 & 1.000 & 1.000 & 1.000 & $\cdots$ & 1.000 & 1.000\\
+ \bottomrule
+\end{tabular}
+\end{center}
+\caption{Sample AHP Calculations for the Quality of Usability} \label{Tbl_SampleAHP}
+\end{table}
+
+We repeat the first and second steps for each of the qualities.  The third step
+combines the quality rankings into an overall ranking.  Following AHP, we need
+to first prioritize the qualities.  The AHP method finds the priority of quality
+$i$ ($p_i$) in the same way that the score ($w_j$) was found for software
+package $j$ evaluated for a given quality (as shown above).  That is, we
+conduct a pairwise comparison between the priority of different qualities to
+construct the $m \times m$ matrix $A$, and then we take the mean of normalized
+values for row $i$ to find the priority value $p_i$ for quality $i$.  If we
+introduce the notation that $w^i_j$ is the score for quality $i$ for package
+$j$, then the overall score $S_j$ for package $j$ is found via:
+
+$$S_j = \sum_{i=1}^m w^i_j p_i$$
 
@@
 
-\subsection{Continuous Integration} \label{Sec_ContinuousIntegration}
+\subsection{Use Continuous Integration} \label{Sec_ContinuousIntegration}
 
@@
 Section~\ref{painpoints} lists the advantages and disadvantages between native
-and web applications. Here we give further suggestions deciding whether to adopt
-a web application tech stack.  The goal is to improve the four priority factors
-identified by developers: compatibility, maintainability, performance, and
-security.  The suggestions are intended to provide ideas and avenues for
-exploration; a web application might not be the right fit for all projects and
-all teams.
+and web applications. Here we give further suggestions to help with deciding
+whether to adopt a web application.  The decision will be based on whether, on
+balance, the web application improves the four factors identified by developers:
+compatibility, maintainability, performance, and security. To enable
+decision-making, a team will need to prioritize between these factors, based on
+the team's objectives and experience. The suggestions are intended to provide
+ideas and avenues for exploration; a web application might not be the right fit
+for all projects and all teams.
 
 \begin{itemize}
-\item \textbf{Identify the priorities between the factors.} Simultaneously
-achieving high levels for all four factors (compatibility, maintainability,
-performance and security) is difficult. To enable decision making, a team needs
-to prioritize its objectives according to its resource and experience.
-
-\item \textbf{Be open-minded about new technologies.} Web applications with only
-a frontend are known for worse performance than native applications. However,
-new technologies may ease this difference. For example, some JavaScript
-libraries can help the frontend harness the power of the computer's GPU and
-accelerate graphical computing. In addition, there are new frameworks helping
-developers with cross-platform compatibility. For example, the
-\href{https://flutter.dev/}{Flutter} project enables support for web, mobile,
-and desktop OS with one codebase.
-
-\item \textbf{Web applications can also deliver high performance.} Web
+
+\item \textbf{Modern technologies may improve frontend performance.} Web
+applications with only a frontend usually perform worse than native
+applications. However, new technologies may ease this difference. For example,
+some JavaScript libraries can help the frontend harness the power of the
+computer's GPU and accelerate graphical computing. In addition, there are new
+frameworks helping developers with cross-platform compatibility. For example,
+the \href{https://flutter.dev/}{Flutter} project enables support for web,
+mobile, and desktop OS with one codebase.
+
+\item \textbf{Web applications can potentially deliver high performance.} Web
 applications with backend servers may perform even better than native
@@
 major cloud service providers may be worth exploring. Serverless solutions still
-use a server, but the team is only charged when they use it. The solution is
-event-driven, and costs the team by the number of requests it processes. Thus,
-serverless can be very cost-effective for less intensively used functions.
+use a server, but the server provider only charges the team when they use the
+server. The solution is event-driven, and costs the team by the number of
+requests processed. Thus, serverless can be very cost-effective for less
+intensively used functions.
 

### 2022-10-14 StateOfPractice/Papers/MedImageSoft_SOP-arXiv/MedImageSoft_SOP_AcademicRadiology.tex
Added a link to the technology hurdles pain point

@@
 
-Section~\ref{painpoints} lists the advantages and disadvantages between native
-and web applications. Here we give further suggestions to help with deciding
-whether to adopt a web application.  The decision will be based on whether, on
-balance, the web application improves the four factors identified by developers:
-compatibility, maintainability, performance, and security. To enable
-decision-making, a team will need to prioritize between these factors, based on
-the team's objectives and experience. The suggestions are intended to provide
-ideas and avenues for exploration; a web application might not be the right fit
-for all projects and all teams.
+Section~\ref{painpoints} describes technology hurdles
+(\ppref{P_TechnologyHurdles}) that motivate considering the use of web
+applications. Here we give further advice to help with deciding whether to adopt
+a web application. The decision will be based on whether, on balance, the web
+application improves the four factors identified by developers: compatibility,
+maintainability, performance, and security. To enable decision-making, a team
+will need to prioritize between these factors, based on the team's objectives
+and experience. The suggestions are intended to provide ideas and avenues for
+exploration; a web application might not be the right fit for all projects and
+all teams.
 
@@
 
-As described in Section~\ref{painpoints}, it is difficult for software
-development teams to access real-world medical imaging datasets. This problem
-restricts their capability and flexibility for testing. We provide some
-suggestions as follows:
+As described in Section~\ref{painpoints} technology hurdles
+(\ppref{P_TechnologyHurdles}) make it difficult for software development teams
+to access real-world medical imaging datasets. This problem restricts their
+capability and flexibility for testing. We provide some suggestions as follows:
 

### 2022-10-18 StateOfPractice/Papers/MedImageSoft_SOP-arXiv/MedImageSoft_SOP_AcademicRadiology.tex
Edits, mostly to the recommendations section

@@
 OwojaiyeEtAl2021_CSE}. \citet{JohansonAndHasselbring2018} observe that the
-state-of-the-practice for SCS in industry and academia does not incorporate
+state-of-the-practice for research software in industry and academia does not incorporate
 state-of-the-art SE tools and methods.  This causes sustainability and
@@
 We answer research question \rqref{RQ_CompareMethodologies} by comparing the
-principles, processes and methodologies used for MI software to what can be
+principles, processes, and methodologies used for MI software to what can be
 gleaned from the literature on research software in general. In our interviews
@@
 software \citep{Smith2016}, especially if the developers work iteratively and
-incrementally, but externally document their work as if a rationale design
-process were followed \citep{parnas1986rational}.
+incrementally, but externally document their work as if they followed a
+rationale design process \citep{parnas1986rational}.
 
@@
 mentioned above, although the topics of reproducibility and technical debt did
-not come up in our MI interviews, these two topics were discussed as part of the
-discussion of software qualities, as covered at the end of this section.
+not come up in our MI interviews, we covered these two topics as part of the
+discussion of software qualities, as summarized at the end of this section.
 Although previous studies show pain points that were not mentioned by MI
@@
 \begin{itemize}
-\item Using open datasets, like the datasets for the liver \citep{BilicEtAl2019}
-and brain \citep{MenzeEtAl2015} tumor segmentation benchmarks.
+\item Using open datasets.
 \item If (part of) the team belongs to a medical school or a hospital, using the
@@
 reproducibility explicitly as a pain point, they did mention the need to improve
-documentation.  Good documentation doesn't just address the pain points of lack
+documentation.  Good documentation does not just address the pain points of lack
 of developer time (\ppref{P_LackDevTime}), technology hurdles
@@
 
-\subsection{Move To Web Applications} \label{sec_recommendations_tech_stack}
-
-Section~\ref{painpoints} describes technology hurdles
-(\ppref{P_TechnologyHurdles}) that motivate considering the use of web
+\subsection{Move To Web Applications} \label{sec_webapps}
+
+Section~\ref{painpoints} describes the pain point of technology hurdles
+(\ppref{P_TechnologyHurdles}), which motivates considering the use of web
 applications. Here we give further advice to help with deciding whether to adopt
@@
 maintainability, performance, and security. To enable decision-making, a team
-will need to prioritize between these factors, based on the team's objectives
-and experience. The suggestions are intended to provide ideas and avenues for
-exploration; a web application might not be the right fit for all projects and
+will need to prioritize between these factors, based on their objectives and
+experience. The suggestions are intended to provide ideas and avenues for
+exploration; a web application will not be the right fit for all projects and
 all teams.
@@
 the \href{https://flutter.dev/}{Flutter} project enables support for web,
-mobile, and desktop OS with one codebase.
-
-\item \textbf{Web applications can potentially deliver high performance.} Web
+mobile, and desktop OS with one codebase.  Other options include
+\href{https://vuejs.org/} {Vue}, \href{https://angular.io/} {Angular} and
+\href{https://reactjs.org/} {React} and \href{https://elm-lang.org/}{Elm}.
+
+\item \textbf{Backend servers can potentially deliver high performance.} Web
 applications with backend servers may perform even better than native
 applications. If a team needs to support lower-end computers, it is good to use
-back-end servers for heavy computing tasks.
+back-end servers for heavy computing tasks.  For backend servers where traffic
+and latency is not an issue, options include
+\href{https://www.django-rest-framework.org/} {Django}, \href{https://laravel.com/} {Laravel} and
+\href{https://nodejs.org/en/} {Node.js}.  The advantage of Django is that it provides access to Python
+libraries.  For backend servers where traffic and latency is an issue,
+\href{https://github.com/gin-gonic/gin} {Gin} is an option.
 
 \item \textbf{Backend servers can have low costs.} Serverless solutions from
-major cloud service providers may be worth exploring. Serverless solutions still
-use a server, but the server provider only charges the team when they use the
-server. The solution is event-driven, and costs the team by the number of
-requests processed. Thus, serverless can be very cost-effective for less
-intensively used functions.
+major cloud service providers (like Amazon Web Services (AWS) and Google Cloud
+Platform) may be worth exploring. Serverless solutions still use a server, but
+the server provider only charges the team when they use the server. The solution
+is event-driven, and costs the team by the number of requests processed. Thus,
+serverless can be very cost-effective for less intensively used functions.
 
@@
 
-\subsection{Recommendations on Enriching the Testing Datasets} 
+\subsection{Enrich the Testing Datasets} 
 \label{sec_recommendations_testing_dataset}
 
-As described in Section~\ref{painpoints} technology hurdles
-(\ppref{P_TechnologyHurdles}) make it difficult for software development teams
-to access real-world medical imaging datasets. This problem restricts their
-capability and flexibility for testing. We provide some suggestions as follows:
+As described in Section~\ref{painpoints}, ensuring correctness
+(\ppref{P_Correctness}) via testing can be problematic because of limited access
+to real-world medical imaging datasets.  We build on the suggestions we heard
+from our interviewees as follows:
 
@@
 \href{https://www.cancerimagingarchive.net/}{Cancer Imaging Archive}
-\citep{PriorEtAl2017}, and \href{https://medpix.nlm.nih.gov/home}{MedPix} by
-National Library of Medicine \citep{Smirniotopoulos2014}. A team developing MI
-software should be able to find more open datasets according to their needs.
+\citep{PriorEtAl2017}, \href{https://medpix.nlm.nih.gov/home}{MedPix} by
+National Library of Medicine \citep{Smirniotopoulos2014}, and datasets for liver
+\citep{BilicEtAl2019} and brain \citep{MenzeEtAl2015} tumor segmentation
+benchmarks.  A team developing MI software should be able to find more open
+datasets according to their needs.
 
@@
 professionals can better offer voluntary support for testing. Ultimately, the
-community can establish a nonprofit organization as a third-party, which
-maintains large datasets, tests OSS in the domain, and protects privacy.
+community can establish a nonprofit organization as a third party, which
+maintains large datasets, tests Open Source Software (OSS) in the domain, and
+protects privacy.
 
@@
 
-Except for \citet{ThielEtAl2020}, the research software guidelines that we
-consulted do not mention linters.  However, we believe that linters have the
-potential to improve code quality at a relatively low cost.  A linter is a tool
-that statically analyzes code to find programming errors, stylistic
-inconsistencies and suspicious constructs \citep{Wikipedia2022_Lint}. Linters
-can be used to spot check code files, or even better as part of a continuous
-integration system, as discussed in Section~\ref{Sec_ContinuousIntegration}.
-
-A sample of the potential benefits of linters include: finding memory leaks,
-finding potential bugs, standardizing code with respect to formatting, improving
-performance, removing silly errors before code reviews, and catching potential
-security issues \citep{SourceLevel2022_Lint}.  Linters are available for most
-popular programming languages.  For instance Python has the options of PyLint,
-flake8 and Black \citep{Zadka2018}.
-
-Linters can address the pain points for LBM developers.  For instance, a linter
-can decrease the amount of development time (\ppref{P_LackDevTime}) by
-decreasing the number of mundane mistakes programmers have to catch.  Since the
-linter can include rules that capture the wisdom of senior programmers, it can
-help newer developers avoid common mistakes. Although there are not many linters
-for checking parallel programming mistakes (\ppref{P_TechnologyHurdles}),
-Parallel Lint is an option for OpenMP. Consistently using a linter can guard
-against some technical debt (\ppref{Q_Maintainability}). Although linters are
-tools for code analysis, similar ideas can be applied to ``lint'' documentation
-to ensure adherence to basic rules.  The use of tools to check documentation is
-a partial explanation for the relatively higher quality of statistical tools
-that are part of the Comprehensive R Archive Network (CRAN)
-\citep{SmithEtAl2018_StatSoft}.
-
-\subsection{Conduct Rigorous Peer Reviews} \label{Sec_PeerReview}
-
-Peer review is a strategy already adopted by some LBM developers to improve
-confidence in their software, but the benefits of peer review can be improved by
-making its application more rigorous. \citet{Jones2008} shows that rigorous
-inspection finds 60-65\% of latent defects on average, and often tops 85\% in
-defect removal efficiency.  By way of comparison, most forms of testing average
+A linter is a tool that statically analyzes code to find programming errors,
+suspicious constructs, and stylistic inconsistencies \citep{Wikipedia2022_Lint}.
+Linters can be used as an ad hoc check for code files, but they really come into
+their own when used as part of a continuous integration system, as discussed in
+Section~\ref{Sec_ContinuousIntegration}. Almost none of the research software
+guidelines that we consulted, summarized in Section~\ref{Sec_CompareArtifacts},
+mention linters.  The one exception is \citet{ThielEtAl2020}.  Despite the lack
+of mention in the guidelines, we believe that linters have the potential to
+improve code quality at a relatively low cost.
+
+Linters have the following benefits: finding potential bugs, finding memory
+leaks, improving performance, standardizing code with respect to formatting,
+removing silly errors before code reviews, and catching potential security
+issues \citep{SourceLevel2022_Lint}. Most popular programming languages have an
+accompanying linter.  For example, Python has the options of PyLint, flake8 and
+Black \citep{Zadka2018}.
+
+We recommend the use of linters because they are relatively easy to incorporate
+into a developer's workflow, and they address several MI pain points
+(Section~\ref{painpoints}).  For instance, linters address the lack of
+development time (\ppref{P_LackDevTime}) by increasing the developer's
+productive time via guarding against making frustrating, time-consuming, mundane
+mistakes.  Moreover, since a linter can include rules that capture the wisdom of
+senior programmers, it can help newer developers avoid common mistakes. With
+respect to the technology hurdle pain point (\ppref{P_TechnologyHurdles}),
+linters can assist with the move toward web applications
+(Section~\ref{sec_webapps}).  For instance, ESLint in React is a pluggable
+linter that lets the developer know if they have imported something and not used
+it, if a function could be short-handed, if there are indentation
+inconsistencies, etc. \citep{Whitehouse2018}. By insisting on code
+standarization linters can reduce technical debt and thus improve
+maintainability (\qref{Q_Maintainability}). Although linters are tools for code
+analysis, the idea of statically checking for adherence to basic rules can be
+extended to check documentation. \citet{SmithEtAl2018_StatSoft} shows how the
+use of tools to enforce documentation standards partially explains the
+relatively higher quality of statistical tools that are part of the
+Comprehensive R Archive Network (CRAN).
+
+\subsection{Conduct a Mix of Rigorous and Informal Peer Reviews} \label{Sec_PeerReview}
+
+We advocate incorporating peer review into the development process, as
+frequently recommended for research software \citep{HerouxEtAl2008, Givler2020,
+OrvizEtAl2017, USGS2019}. In most cases a modern, lightweight review, should be
+adequate.  Modern code review is informal, tool-based, asynchronous, and focused
+on reviewing code changes \citep{SadowskiEtAl2018}. Managing a project via
+GitHub pull requests is an example of a modern approach to reviewing code.
+Software development organizations have moved to this lightweight style of code
+review because of the inefficiencies of rigorous inspections
+\citep{RigbyAndBird2013}.  However, for important parts of the code, developers
+may benefit from mixing in a more rigorous approach.
+
+\citet{Fagan1976} began work on rigorous review via code inspection.  Elements
+of a typical inspection include reviewing the code against a checklist (checking
+the consistency of variable names, look for terminating loops, etc.), performing
+specific review tasks (such as summarizing the code's purpose, cross-referencing
+the code to the technical manual, creating a data dictionary for a given module,
+etc.) Rigorous inspection finds 60-65\% of latent defects on average, and often
+tops 85\% in defect removal efficiency \citep{Jones2008}. The success rate of
+code inspection is generally higher than most forms of testing, which average
 between 30 and 35\% for defect removal efficiency \citep{EbertAndJones2009,
-Jones2008}.  A formal code inspection involves asking the reviewer to either
-follow a review checklist (check consistency of variable names, look for
-terminating loops, etc.), or perform specific review tasks (like summarize the
-purpose of the code, create a data dictionary for a given module,
-cross-reference the code to the technical manual, etc.) The task based
-inspection approach has been effectively used for research software, as
-described by \citet{KellyAndShepard2000}. Task based inspection is an ideal fit
-with an issue tracking system, like GitHub.  The review tasks can be issues, so
-that they can be easily assigned, monitored and recorded.  Other potential
-issues for the tracker include assigning junior team members to test
-installation instructions and getting-started tutorials.
-
-Rigorous peer review addresses the same pain points as linters
+Jones2008}. For research software, \citep{KellyAndShepard2000} show a task based
+inspection approach can be effective. Task based inspection is an ideal fit with
+an issue tracking system, like GitHub.  The review tasks can be issues, so that
+they can be easily assigned, monitored and recorded. Potential issues include
+assigning junior developers to test getting-started tutorial and installation
+instructions.
+
+As indicated in Section~\ref{Sec_CompareMethodologies} some MI projects use
+modern code review, via issue tracking and the use of GitHub.  Those MI projects
+not incorporating modern code review would likely benefit by adopting it.
+Although a rigorous code inspection is likely not worth the required resources,
+for critical parts of the code, developers may want to adopt a more rigorous
+approach. For instance, developers may drop the modern requirement for
+asynchronous review and instead occasionally use synchronous review to help
+uncover errors and disseminate best practices throughout the team.  For
+instance, teams could periodically meet, either in-person or virtually, and have
+junior members walk through their code.  In-person reviews will likely help
+realize the benefits of modern code review noticed by
+\citet{BirdAndBacchelli2013}: defect detection, knowledge transfer, increased
+team awareness, and creation of alternative solutions to problems.
+
+Due to improving code quality and increasing knowledge transfer, peer review
+addresses the same pain points and qualities as linters
 (Section~\ref{Sec_Linters}): \ppref{P_LackDevTime}, \ppref{P_TechnologyHurdles},
-and \ppref{Q_Maintainability}. Peer review addresses these points by efficiently
-searching for defects and problems.  Finding misunderstandings in of how the
-code implements the required theory improves the software's correctness
-(\ppref{P_Correctness}).
-
-Peer review --  One of the developers (ESPResSo) that was interviewed noted that
-an ad hoc peer review process is used to assess major changes and additions.
-Using peer review (also called technical review) matches with recommended
-practice for research software \citep{HerouxEtAl2008, Givler2020, OrvizEtAl2017,
-USGS2019}.
-
-\subsection{Follow Advice From the Open-Source Community to Grow the Number of Contributors} \label{Sec_GrowContributors}
-
-In our interviews with developers we heard they would like to have additional
-software contributors.  More developers would help with addressing the lack of
-development time pain point (\ppref{P_LackDevTime}).  In investigating advice on
-increasing the number of contributors, we found that the advice usually starts
-with following best practices, including providing artifacts like those
-discussed in Section~\ref{Sec_CompareArtifacts}, such as a contributor
-guidelines, code style guidelines, a clear code of conduct and high quality code
-and documentation. This advice is logical, but in our assessment of LBM software
-we found examples of projects that already follow many best practices, but still
-have small development teams. Attracting developers apparently requires more
-than just building a good product.
-
-Some potential additional ideas for growing the number of contributors are
-as follows:
+and \qref{Q_Maintainability}. Peer review can potentially find misunderstandings
+in of how the code implements the required theory, which will improve the
+software's correctness (\ppref{P_Correctness}). The benefits of peer review for
+addressing pain points can be increased by extending the review from just code,
+to also reviewing all software artifacts, including documentation, build
+scripts, test cases and the development process itself.
+
+\subsection{Design For Change} \label{Sec_DesForChange}
+
+In our ``state of the practice'' assessment exercise for LBM software
+\citep{SmithEtAl2022}, we noticed that LBM developers implicitly used
+modularization based on the principle of design for change to improve
+maintainability (\qref{Q_Maintainability}).  We recommend that MI developers use
+the same principle for their modularizations.  Although the advice to modularize
+research software to handle complexity is common \citep{WilsonEtAl2014,
+StewartEtAl2017, Storer2017}, specific guidelines on how to divide the software
+into modules is less prevalent.  Not every decomposition is a good design for
+supporting change, as shown by \citet{Parnas1972a}.  A design with low cohesion
+and high coupling \citep[p.\ 48]{GhezziEtAl2003} will make change difficult.
+Especially in research software, where change is inevitable, designers need to
+produce a modularization that supports change. \citep{JungEtAl2022} points out
+that ocean modelling software is currently feeling the pain of not emphasizing
+modularization in legacy code.
+
+Specific examples of design for change for LBM software \citep{SmithEtAl2022}
+include the following:
 
 \begin{itemize}
-	\item Clearly identify issues that are an appropriate starting points for
-	new developers \citep{Garcia2016, Jalan2016, Proffitt2017}.
-	\item Given that in most projects developers were once users
-	\citep{McQuaid2018}, recruit future developers from the current set of
-	users.
-	\item Create templates for pull requests and issues \citep{Jalan2016}.
-    \item Welcome all kinds of contributions, not just code.  Non-code
-    contributions include documentation, fixing typos, issue reporting and test
-    cases \citep{Jalan2016, Proffitt2017}.
-    \item Reward and recognize new contributors, via small rewards like stickers
-    or shirts, or even just a simple shoutout or mention in a blog post or on
-    social media \citep{Jalan2016, Proffitt2017}.
-    \item Recognize that open source is more about people than it will ever be
-    about code \citep{Jalan2016}.
-    \item Look beyond just recruiting online and seek new developers at
-    conferences or other user meet ups \citep{Garcia2016}.
-    \item Follow the advice of \citet{Kuchner2012} to adapt ideas from marketing
-    to promote science.  Specific ideas that could be applied to open source
-    projects include the following: 
-    \begin{itemize}
-		\item Storytelling to motivate interest in the project, where the story
-		is a sequence of events and pauses for reflection \citep[p.\
-		21--22]{Kuchner2012}.
-		\item Invest effort in building relationships with users and with potential future developers.
-		\item Brand your project with what makes it unique and special.
+\item \href{https://github.com/pylbm/pylbm}{pyLBM} has decoupled geometries and
+models of their system using abstraction and modularization of the source code,
+to make it easy to add new features.  The pyLBM design allows for independent
+changes to the geometry and the model.  pyLBM also redeveloped data structures
+to ease future change. 
+\item \href{https://github.com/CFD-GO/TCLB}{TCLB} \citep{rokicki2016adjoint} is
+designed to allow for the addition of some LBM features, but changes to major
+aspects of the system would be difficult. For example, ``implementing a new
+model will be an easy contribution'', but changes to the ``Cartesian mesh … will
+be a nightmare'' \citep{SmithEtAl2022}. The design of TCLB highlights that not
+every conceivable change needs to be supported, only the likely changes.  
 \end{itemize}
-	\item Consider combining existing projects to pool collective resources, and
-	reduce competition between different open source projects.
-\end{itemize}
-
-\subsection{Design For Change} \label{Sec_DesForChange}
-
-To address technical debt (\ppref{Q_Maintainability}), the top LBM developers show how
-modularization can be used to design research software for future change.
-Although the advice to modularize research software to handle complexity is
-common \citep{WilsonEtAl2014, StewartEtAl2017, Storer2017}, specific guidelines
-on how to divide the software into modules is less prevalent.  Not every
-decomposition is a good design for supporting change, as shown by
-\citet{Parnas1972a}.  A design with low cohesion and high coupling \citep[p.\
-48]{GhezziEtAl2003} will make change difficult. Especially in research software,
-where change is inevitable, designers need to produce a modularization that
-supports change. Ocean modelling software is currently feeling the pain of
-not emphasizing modularization in legacy code \citep{JungEtAl2022}.
-
-Interviewed LBM developers highlighted cases where their modularizations
-anticipated future changes.  The developer of pyLBM mentioned that the
-geometries and models of their system had been ``decoupled'', using abstraction
-and modularization of the source code, to make it ``very easy to add [new]
-features''.  The pyLBM design allows for independent changes to the geometry and
-the model.  We also learned that the package pyLBM redeveloped data structures
-to ease future changes. The developer of TCLB noted that their design allows for
-the addition of some LBM features, but changes to major aspects of the system
-would be difficult. For example, ``implementing a new model will be an easy
-contribution'', but changes to the ``Cartesian mesh … will be a nightmare''. The
-design of TCLB highlights that not every conceivable change needs to be
-supported, only the likely changes.
-
-As the LBM developers illustrate, they accomplish design for change by first
-identifying likely changes, either implicitly or explicitly, and second by
-hiding each likely change behind a well-defined module interface.  Although it
-is unclear whether the developers are aware of Parnas's work, the approach
-mirrors his recommendations \citep{Parnas1972a}. \citet{SmithEtAl2022} (Section
-Documentation as Part of the Development Process) lists ideas for how to
-document the design, including the likely changes, so that they are more visible
-to others.
+
+As the LBM examples above illustrate, developers can accomplish design for
+change by first identifying likely changes, either implicitly or explicitly, and
+second by hiding each likely change behind a well-defined module interface.
+This approach mirrors the recommendations from \citet{Parnas1972a}.
+\citet{SmithEtAl2022} (under Section ``Documentation as Part of the Development
+Process'') lists ideas for how to document the design, including the likely
+changes, so that they are more visible to others.
 
@@
 
-From \cite{RinehartEtAl2015}, an assurance case is ``[a] reasoned and compelling
-argument, supported by a body of evidence, that a system, service or
-organization will operate as intended for a defined application in a defined
-environment.''  Assurance cases have been successfully employed for safety
-critical systems, but using this technique for SC software is a new idea.  The
-Goal Structuring Notation (GSN) \cite{Spriggs2012} seems like a good framework
-for initial investigation.
-
-Scientific software, such as medical software, is often subject to
-standardization and regulatory approval. While applying such approvals and
-standards has had a beneficial effect on system quality, it does not provide
-good tracking of the development stages, as the compliance with the standards
-is mostly checked after the system development. Once a system is implemented,
-its documentations must be approved by the regulators. This process is lengthy
-and expensive. In contrast, assurance case development usually occurs in
-parallel with the system construction, resulting in a traceable, detailed
-argument for the desired property.  Moreover, since the assurance case is
-integrated into the system construction, the domain experts are involved from
-the start.
-
-Putting the argument in the hands of the experts means that they will work to
-convince themselves, along with the regulators.  They will use the expertise
-that the regulators do not have and they will see the value of documentation (as
-mentioned in section on pain points with methods to mitigate); they will be
-engaged.  This engagement will hopefully help bridge the current chasm between
-software engineering and scientific computing~\cite{Kelly2007}, by motivating
-scientists toward documentation and correcting the problem of software engineers
-failing to meet scientists' expectations~\cite{Segal2008}.  Significant
-documentation will still likely be necessary, but now the developers control the
-documentation.  What is created will be relevant and necessary.
-
-For SCS a top level claim could read ``Program X delivers correct outputs when
-used for its intended use/purpose in its intended environment.''  The next step
-would be to decompose this claim into sub-claims that will be easier to prove.
-The sub-claims will likely also be further divided until the bottom of the
-graph, where the measurable evidence is provided.  Typical evidence will consist
-of documents, expert reviews, test case results etc.
-
-Using the specific example of preparing an assurance case for pre-existing
-medical image analysis software (3dfim+)~\cite{SmithEtAl2018_ICSEPoster}, the
-top level can be decomposed into four sub-goals, as shown in
-Figure~\ref{TopGoal}.  This example follows the same pattern as used for medical
-devices~\cite{Wassyng2015}.  The first sub-goal (GR) argues for the quality of
-the documentation of the requirements (SRS).  The second sub-goal (GD) says that
-the design complies with the requirements and the third proposes that the
-implementation also complies with the requirements.  The fourth sub-goal (GA)
-claims that the inputs to 3dfim+ will satisfy the operational assumptions, since
-we need valid input to make an argument for the correctness of the output.
-
-\begin{figure*}[!h]
+To ensure correctness (\ppref{P_Correctness} and to achieve the quality of
+reproducibility (\qref{Q_Reproducibility}), we recommend considering the use of
+assurance cases.  \citet{RinehartEtAl2015} defines an assurance case as ``[a]
+reasoned and compelling argument, supported by a body of evidence, that a
+system, service, or organization will operate as intended for a defined
+application in a defined environment.''  An assurance cases provide an organized
+and explicit argument that the software and its documentation achieves desired
+qualities, such as correctness and reproducibility.  Although assurance cases
+have been successfully employed for safety critical systems
+\citep{RinehartEtAl2015}, this technique is relatively new for research software
+\citep{SmithEtAl2020_AC, Smith2018}.
+
+One way to present an assurance case is through the Goal Structuring Notation
+(GSN) \cite{Spriggs2012}, which make arguments clear, easy to read and, hence,
+easy to challenge. GSN starts with a Top Goal (Claim), like ``Program X delivers
+correct outputs when used for its intended use/purpose in its intended
+environment.''  We then decompose this top goal into Sub-Goals, which themselves
+may be further decomposed.  The purpose of the decomposition is to take the
+abstract higher level goals and bring them down to something concrete that can
+be proven.  The decomposition ends with the terminal Sub-Goals that are
+supported by Solutions (Evidence). Typical evidence will consist of documents,
+expert reviews, test case results, peer review, etc.  Within the GSN framework,
+there are also strategy blocks, which describe the rationale for decomposing a
+Goal or Sub-Goal into more detailed Sub-Goals. A common tool for creating,
+editing, and presenting, a GSN argument is \href{https://astah.net/} {Astah}.
+
+\citet{SmithEtAl2020_AC} shows the example of arguing for the correctness of the
+Analysis of Functional NeuroImages (AFNI) package 3dfim+ \citep{Ward2000}.
+3dfim+ analyzes the activity of the brain by computing the correlation between
+an ideal signal and the measured brain signal for each voxel. The assurance case
+for the correctness of 3dfim+ has the top level decomposed into four sub-goals,
+as shown in Figure~\ref{TopGoal}.  This example follows the same pattern as used
+for medical devices~\cite{Wassyng2015}.  The first sub-goal (GR) argues for the
+quality of the documentation of the requirements (SRS).  The second sub-goal
+(GD) says that the design complies with the requirements and the third proposes
+that the implementation also complies with the requirements.  The fourth
+sub-goal (GA) claims that the inputs to 3dfim+ will satisfy the operational
+assumptions, since we need valid input to make an argument for the correctness
+of the output.
+
+\begin{figure}[!h]
 \centering
@@
 \label{TopGoal}
-\end{figure*}
-
-Preparing an assurance case for the pre-existing 3dfim+ software justifies the
-value of assurance cases for the certification of SCS~\cite{SmithEtAl2018}.
-Although no errors were found in the output of the existing software, the rigour
-of the proposed approach did lead to finding ambiguities and omissions in the
-existing documentation, such as missing information on the coordinate system
-convention.  In addition, a potential concern for the software itself was
-identified from the GA argument: running the software does not produce any
-warning about the obligation of the user to provide data that matches the
-parametric statistical model employed for the correlation calculations.
-
-Potential objectives for research on assurance cases applied to SCS are as
-follows:
-
-\begin{enumerate}
-\item Creation of an assurance case template will require looking at a variety
-  of SCS projects, which will require engaged project partners; therefore, the
-  initial approach should be to target domains where certification is required,
-  or domains where verification is critical, but challenging, like the High
-  Energy Physics (HEP). \wss{reference the paper about domain specific
-  methodology - maybe I already did?}
-\item Evidence is provided at the bottom of an assurance case, but what is
-  the best form for this evidence?  As mentioned for the SRS, we need an
-  approach for verification.  We also need an approach to verify the other
-  software artifacts, the traceability between them, the team of developers, the
-  team of reviewers, etc.  The question for the future is what combination of
-  tools and processes will provide convincing evidence?
-\item Creation of assurance cases is challenging with existing tools.  In the
-  future, the goal should be automatic generation (see
-  \citet{Smith2018}, Section SecAutoGenerate) of the visual assurance case documentation from
-  the arguments and evidence
-\item Creating a revelation on the value of documentation \citep{Smith2018} may
-   require a pseudo adversarial approach.  SCS developers should be explicitly
-   challenged to present their verification efforts so that their work can be
-   independently verified.  Although SCS developers have developed many
-   successful theories, techniques, and testing procedures for verification, the
-   evidence is generally presented in an ad hoc manner.  Developers should be
-   asked to connect all the pieces of their evidence in a coherent argument.  As
-   a bonus, the act of creating the assurance case may also lead developers to
-   discover subtle edge cases, which would not have been noticed with a less
-   rigorous and systematic approach.
-
-\end{enumerate}
+\end{figure}
+
+Preparing an assurance case for the pre-existing 3dfim+ software shows the value
+of an assurance cases for research software. Although \citet{SmithEtAl2020}
+found no errors in the output of the existing software, the rigour of the
+proposed approach did lead to finding ambiguities and omissions in the existing
+documentation, such as missing information on the coordinate system convention.
+In addition, a potential concern for the software itself was identified from the
+GA argument: running the software does not produce any warning about the
+obligation of the user to provide data that matches the parametric statistical
+model employed for the correlation calculations.
 
@@
 
-We propose automatically generating LBM code and its documentation, using a
-Domain Specific Modelling (DSM) approach. DSM means creating a knowledge base of
-models for physics, computing, mathematics, documentation, and certification and
-then writing explicit ``recipes'' that weave together this knowledge to generate
-the desired code, documentation, test cases, inspection reports and build
-scripts. This definition of DSM is more general than usual; in this
-recommendation DSM implies generation of all software artifacts, not just the
-code. DSM moves development to a higher level of abstraction; domain experts can
-work without concern for low-level implementation details. Using DSM, we
-optimally generate code and documentation, eliminate redundancy, reduce the
-likelihood of errors, and automate maintenance. Moreover, a generative approach
-facilitates the inevitable exploration of LBM modelling assumptions.
-
-DSM provides a transformative technology for documentation, design, and
-verification \citep{JohansonAndHasselbring2018, Smith2018}. DSM allows
-scientists to focus on their science, not software.  A generative approach
-removes the maintenance nightmare of documentation duplicates and near
+To address developer pain points, we propose automatically generating MI code
+and its documentation via a Domain Specific Modelling (DSM) approach. A DSM
+approach uses models to capture knowledge in domains such as physics, computing,
+mathematics, documentation, and certification.  Developers combine and transform
+knowledge via explicit ``recipes'', which weave it together to generate the
+desired code, documentation, test cases, inspection reports and build scripts. A
+recipe can even potentially be written to generate an assurance case
+(Section~\ref{AssuranceCases}). Our definition here of DSM is more general than
+usual; in this recommendation DSM implies generation of all software artifacts,
+not just the code. DSM moves development to a higher level of abstraction so
+that domain experts can work without concern for low-level implementation
+details. DSM allows developers to optimally generate code and documentation,
+reduce the likelihood of errors, eliminate redundancy, and automate maintenance.
+With a DSM approach MI developers can experiment with different algorithm
+choices, different input formats, different outputs formats, etc.
+
+DSMs are not typically used for research software.  In the future, some believe
+that they will transform coding, documentation, design, and verification
+\citep{JohansonAndHasselbring2018, Smith2018}. DSMs remove the distraction of
+writing software, allowing developers to focus on their science.  A DSM approach
+removes the maintenance headaches of documentation duplicates and near
 duplicates \citep{LucivEtAl2018}, since developers capture knowledge once and
 transform it as needed.  Code generation has previously been applied to improve
-research software.  For instance, ATLAS (Automatically Tuned Linear Algebra
-Software) \citep{WhaleyEtAl2001} and Blitz++ \citep{Veldhuizen1998} produces
-efficient and portable linear algebra software.  Spiral \citep{Pueschel2001}
-uses software/hardware generation for digital signal processing.
-\citet{Carette2008} shows how to generate a family of efficient, type-safe
-Gaussian elimination algorithms.  FEniCS (Finite Element and Computational
-Software) uses code generation when solving differential equations
-\citep{LoggEtAl2012}. \citet{OberEtAl2018} and \citet{MatkerimEtAl2013} apply
-DSM to High Performance Computing (HPC), using UML (Unified Modelling Language)
-for their domain models. Unlike previous DSM work, the current recommendation
-focuses on generating all software artifacts (requirements, design, etc.), not
-just code.  \citet{SzymczakEtAl2016} presents initial work on this ``generate
-all things'' approach, \citet{SmithAndCarette2021-BRIC} presents a motivating
-example, and \citep{CaretteEtAl2021-Drasil} provides a prototype.
-
-A DSM approach addresses multiple pain points.  For instance, once the
-infrastructure is in place, a DSM can decrease the amount of development time
-(\ppref{P_LackDevTime}) by automation.  Since a DSM approach allows scientists
-to focus on their science, rather than software, the lack of software
-development experience will be less of an issue. The DSM approach can capture
-computing knowledge to mitigate the technology related pain points
-(\ppref{P_TechnologyHurdles}).  For ensuring correctness (\ppref{P_Correctness})
-a generative approach can aim to be correct by construction.  If there are
-mistakes, DSM has the advantage that they are propagated throughout the
-generated artifacts, which greatly increases the chance that someone will notice
-the mistake.  Technical debt (\ppref{Q_Maintainability}) is not a concern with a
-DSM approach since developers write the recipes used for generation at a high
-level making them relatively easy to change.  With respect to pain points on
-usability (\ppref{P_Usability}), the generative approach we are proposing
-addresses these directly, since documentation is a primary concern, rather than
-an afterthought.
+research software, such as linear algebra software packages like Blitz++
+\citep{Veldhuizen1998}, and ATLAS (Automatically Tuned Linear Algebra Software)
+\citep{WhaleyEtAl2001}.  Software/hardware generation has been applied for
+digital signal processing in Spiral \citep{Pueschel2001}. A generative approach
+has also been used for a family of efficient, type-safe Gaussian elimination
+algorithms \citet{Carette2008}. \citep{LoggEtAl2012} use code generation when
+solving partial differential equations in FEniCS (Finite Element and
+Computational Software). \citet{MatkerimEtAl2013} and \citet{OberEtAl2018} use
+DSMs for High Performance Computing (HPC), using UML (Unified Modelling
+Language) for their domain models. Unlike previous DSM work, the current
+recommendation focuses on generating all software artifacts (requirements,
+design, etc.), not just code. \citet{SzymczakEtAl2016} presents initial work on
+this approach, \citet{SmithAndCarette2021-BRIC} presents a motivating example,
+and \citep{CaretteEtAl2021-Drasil} provides a prototype.
+
+A ``generate all things'' approach addresses multiple pain points.  For example,
+a generative approach can decrease development time (\ppref{P_LackDevTime}) by
+automation, once the necessary infrastructure is in place.  The DSM approach
+addresses the technology related pain point (\ppref{P_TechnologyHurdles})
+because technology information can be captured in the models and transformed as
+needed. To ensure correctness (\ppref{P_Correctness}), a DSM approach should be
+correct by construction.  If there are mistakes, DSM has the advantage that they
+are propagated throughout the generated artifacts, which greatly increases the
+chance that someone will notice the mistake.  Maintainability
+(\qref{Q_Maintainability}) is addressed because developers write the recipes
+used for generation at a high level making them relatively easy to change.
+Usability (\ppref{P_Usability}) is addressed because of the emphasis on
+generating up-to-date documentation.  Reproducibility.
 
@@
 our recommendations on SC software development.
+
+Compare to LBM.
 
@@
 
+\section*{Acknowledgements}
+
+We would like to thank Peter Michalski and Oluwaseun Owojaiye for fruitful
+discussions on topics relevant to this paper.  We would also like to thank Jason
+Balaci for advice on web applications.
+
 %% The Appendices part is started with the command \appendix;

### 2022-10-27 StateOfPractice/Papers/MedImageSoft_SOP-arXiv/MedImageSoft_SOP.tex
Updates to the threats to validity section

--- 2022-10-18 StateOfPractice/Papers/MedImageSoft_SOP-arXiv/MedImageSoft_SOP_AcademicRadiology.tex
@@
 Below we categorize and list the threats to validity that we have identified.
-The categories come from \citet{AmpatzoglouEtAl2019} and \citet{ZhouEtAl2016}.
-
-\wss{observed artifacts (Section~\ref{Sec_CompareArtifacts}) - human judgement,
-not always given the expected name, but the content is there.}
-
-\begin{description}
-    \item[Construct Validity:] ``Defines how effectively a test or experiment measures
-    up to its claims. This aspect deals with whether or not the researcher measures
-    what is intended to be measured'' \citep{AmpatzoglouEtAl2019}.
-\end{description}
+Our categories come from an analysis of software engineering secondary studies
+by \citet{AmpatzoglouEtAl2019}, where a secondary study analyzes the data from a
+set of primary studies.  \citet{AmpatzoglouEtAl2019} is appropriate because a
+common example of a secondary study is a systematic literature review. Our
+methodology is a systematic software review --- the primary studies
+are the software packages, and our work collects and analyzes these primary
+studies.
+
+\subsection{Reliability}
+
+\subsection{Reliability}
+
+A study is reliable if its data and analysis are independent of the specific
+researcher(s) doing the study.  That is, repetition of the study by a different
+researcher should achieve the same results, as long as they use the same
+methodology \citep{RunesonAndHost2009}.  For our study the identified
+reliability related threats include the following:
+
+\begin{itemize}
+\item In our methodology, one individual does the manual measures for all
+packages. The grader's abilities, experiences, and biases may influence the
+results.  
+\item Completing the manual measures of all packages took place over several
+months. During this time the contents of the software repositories may change
+and the reviewer's judgment may drift.
+\end{itemize}
+
+\begin{itemize}
+\item The grading template included an entry for the reviewer's impression.  We
+aimed for objectivity, but there is a risk that some scores may be subjective
+and biased.
+\item The measurements are made at an instant in time, and the instant differs
+by a few weeks between projects, due to the time needed to measure them. For the
+most part, the projects are living and continually changing, which means if we
+measured at a different time, our results may change.
+\end{itemize}
+
+We investigated the risk of reviewer dependence in previous work \citep{SmithEtAl2016}, % DOUBLE BLIND
+where we demonstrated that the measurement process is
+reasonably reproducible by having five products graded by a second reviewer. The
+ranking via this independent review was almost identical to the original
+ranking. As long as each grader uses consistent definitions, the relative
+comparisons in the AHP results will be consistent between graders.
+
+\subsection{Construct Validity}
+
+Construct validity means that adopted metrics represent what they are intended
+to measure \citep{RunesonAndHost2009}. The construct validity related threats
+for the LBM study include the following:
+
+\begin{itemize}
+\item We measure our software qualities indirectly, rather than directly.
+Unfortunately meaningful direct measures for qualities like maintainability,
+reusability and verifiability, are not available.  Therefore, we follow the
+usual approach of assuming that following procedures and adhering to standards
+yields a higher-quality product \citep[p.\ 112]{VanVliet2000}.
+\item The measure of surface robustness (Section~\ref{Sec_Robustness}) is only
+based on two pieces of data per project. The measurement of robustness could be
+expanded, as it currently only measures the impact of unexpected input. Other
+intentional faults could be introduced, but this would require a larger
+investment of measurement time. 
+\item We may have inaccurately estimated maintainability by assuming a higher
+ratio of comments to source code improves maintainability
+(Section~\ref{Sec_Maintainability}). 
+\item Reusability (Section~\ref{reusabilityresults}) is assessed via the number
+of code files and LOC per file. While this measure is indicative of modularity,
+it is possible that some packages have many files, with few LOC, but the files
+do not contain source code that is easily reusable. The formatting may be poor,
+or the source code may be vague and have ambiguous identifiers. 
+\item The measurement of understandability
+(Section~\ref{Sec_SurfUnderstandability}) relies on 10 random source code files.
+By random chance, the 10 files that were chosen may not be representative. 
+\item Our overall AHP ranking makes the unrealistic assumption of equal
+weighting between qualities (Section~\ref{Sec_OverallQuality}).
+\item Our approximation of popularity by stars and watches
+(Section~\ref{repmetrics}) may not be valid. 
+\end{itemize}
 
@@
 
-\begin{description}
-    \item[Internal Validity:] ``This aspect relates to the examination of causal
-    relations. Internal validity examines whether an experimental
-    treatment/condition makes a difference or not, and whether there is evidence
-    to support the claim'' \citep{AmpatzoglouEtAl2019}.
-\end{description}
+\subsection{Internal Validity} \label{Sec_InternalValidity}
+
+A study is internally valid if the discovered causal relations are trustworthy
+and cannot be explained by other factors \citep{RunesonAndHost2009}. The
+internal validity threats include the following:
+
+\begin{itemize}
+\item We may have missed a relevant software package in our search for LBM
+software packages (Section~\ref{identifysoftware}).
+\item A temporal gap exists between the manual measurement of Musubi and the
+measurement of all other packages (Section~\ref{filtersoftware}). This gap
+caused a mismatch between the time of manual and automated data collection,
+since we updated the automated measures (like the number of files, LOC, etc.)
+when Musubi was added.
+\item Our methodology assumes that artifacts in the repositories will show us
+the approach used for software development.  However, not all activities will
+necessarily leave a trace in the repositories. For instance, only three packages
+(ESPResSo, Ludwig, Musubi) show evidence in their artifacts of unit testing and
+CI (Section~\ref{SecSurfCorrectAndVerifiab}), even though interviews suggest a
+more frequent use of both unit testing and CI. For example, OpenLB, pyLBM, and
+TCLB use such methods during development despite this not being obvious from an
+analysis of the online material.
+\item In our manual measures of the software repositories, we may have missed
+some data due to technology issues like broken links. This issue arose with the
+measurement of Palabos, which had a broken link to its user manual, as noted in
+Section~\ref{Sec_CompareArtifacts}. 
+\item Our interviews were with a relatively small sample of four LBM developers.
+The pain points (Section~\ref{painpoints}) they expressed may not be
+representative of the rest of their community.
+\end{itemize}
 
@@
 
-\begin{description}
-    \item[External Validity:] ``Define the domain to which a study's findings
-    can be generalized'' \citep{ZhouEtAl2016}.
-\end{description}
+Although we found that unit testing and CI usage is likely higher than the
+repository artifacts suggest, we did not change the manual measurement results.
+Given that we could not interview a representative for every project, we left
+the data as collected to keep our measures consistent.  Moreover, CI and unit
+testing done in private should not be counted as equivalent to the public
+versions we detected.  When CI and unit testing are private, they are not as
+effective since they are not available for others to inspect and reproduce.
+
+\subsection{External Validity}
+
+External validity concerns whether the results of the study can be generalized
+(applied) to other situations/cases \citep{RunesonAndHost2009}.  The scope of
+our work is LBM solvers.  We are confident that our search was exhaustive and
+that we did not miss any highly popular examples.  Therefore, the bulk of our
+validity concerns about LBM software observations are internal
+(Section~\ref{Sec_InternalValidity}), not external.  However, our hope is that
+the trends observed, and the lessons learned for LBM software can be applied to
+other research software.  With that in mind we identified the following threat
+to external validity:
+
+\begin{itemize}
+\item Generalization of our results will not be possible if the development of
+LBM software is fundamentally different from other research software.
+\end{itemize}
 
@@
 
-\begin{description}
-    \item[Conclusion Validity:] ``Demonstrate that the operations of a study
-    such as the data collection procedure can be repeated, with the same
-    results'' \citep{ZhouEtAl2016}.
-\end{description}
-
-\begin{itemize}
-\item The grading template included an entry for the reviewer's impression.  We
-aimed for objectivity, but there is a risk that some scores may be subjective
-and biased.
-\item The measurements are made at an instant in time, and the instant differs
-by a few weeks between projects, due to the time needed to measure them. For the
-most part, the projects are living and continually changing, which means if we
-measured at a different time, our results may change.
-\end{itemize}
+\wss{observed artifacts (Section~\ref{Sec_CompareArtifacts}) - human judgement,
+not always given the expected name, but the content is there.}
 

### 2022-10-27 StateOfPractice/Papers/MedImageSoft_SOP-arXiv/MedImageSoft_SOP.tex
Minor edits to threats to validity section

@@
 are the software packages, and our work collects and analyzes these primary
-studies.
+studies.  \citep{SmithEtAl2022}.  \wss{add citation to LBM paper}
 
@@
 
-\subsection{Reliability}
-
-A study is reliable if its data and analysis are independent of the specific
-researcher(s) doing the study.  That is, repetition of the study by a different
-researcher should achieve the same results, as long as they use the same
-methodology \citep{RunesonAndHost2009}.  For our study the identified
-reliability related threats include the following:
+A study is reliable if repetition of the study by different researchers using
+the original study's methodology would lead to the same results
+\citep{RunesonAndHost2009}. Reliability means that data and analysis are
+independent of the specific researcher(s) doing the study.  For the current
+study the identified reliability related threats are as follows:
 
 \begin{itemize}
-\item In our methodology, one individual does the manual measures for all
-packages. The grader's abilities, experiences, and biases may influence the
-results.  
-\item Completing the manual measures of all packages took place over several
-months. During this time the contents of the software repositories may change
-and the reviewer's judgment may drift.
+\item One individual does the manual measures for all packages. A different
+evaluator might yield different results, due to differences in abilities,
+experiences, and biases.
+\item The manual measurements for the full set of packages took several months.
+Over this time the software repositories may have changed and the reviewer's
+judgment may have drifted.
 \end{itemize}
 
-\begin{itemize}
-\item The grading template included an entry for the reviewer's impression.  We
-aimed for objectivity, but there is a risk that some scores may be subjective
-and biased.
-\item The measurements are made at an instant in time, and the instant differs
-by a few weeks between projects, due to the time needed to measure them. For the
-most part, the projects are living and continually changing, which means if we
-measured at a different time, our results may change.
-\end{itemize}
-
-We investigated the risk of reviewer dependence in previous work \citep{SmithEtAl2016}, % DOUBLE BLIND
+We investigated the reliability risk associated with the reviewer's judgement was investigated in previous work
+\citep{SmithEtAl2016}, % DOUBLE BLIND
 where we demonstrated that the measurement process is

### 2022-10-28 StateOfPractice/Papers/MedImageSoft_SOP-arXiv/MedImageSoft_SOP.tex
Updates to threats to validity section

--- 2022-10-18 StateOfPractice/Papers/MedImageSoft_SOP-arXiv/MedImageSoft_SOP_AcademicRadiology.tex
@@
 
-We have a two step process for selecting software packages: i) identify software
+We have a two-step process for selecting software packages: i) identify software
 candidates in the chosen domain; and, ii) filter the list to remove less
@@
 
-\subsection{Overall Scores}
+\subsection{Overall Scores} \label{Sec_OverallQ}
 
@@
 Below we categorize and list the threats to validity that we have identified.
-The categories come from \citet{AmpatzoglouEtAl2019} and \citet{ZhouEtAl2016}.
-
-\wss{observed artifacts (Section~\ref{Sec_CompareArtifacts}) - human judgement,
-not always given the expected name, but the content is there.}
-
-\begin{description}
-    \item[Construct Validity:] ``Defines how effectively a test or experiment measures
-    up to its claims. This aspect deals with whether or not the researcher measures
-    what is intended to be measured'' \citep{AmpatzoglouEtAl2019}.
-\end{description}
+Our categories come from an analysis of software engineering secondary studies
+by \citet{AmpatzoglouEtAl2019}, where a secondary study analyzes the data from a
+set of primary studies.  \citet{AmpatzoglouEtAl2019} is appropriate because a
+common example of a secondary study is a systematic literature review. Our
+methodology is a systematic software review --- the primary studies are the
+software packages, and our work collects and analyzes these primary studies.  We
+identified similar threats to validity in our assessment of the state of the
+practice of Lattice Boltzmann Solvers \citep{SmithEtAl2022}.
+
+\subsection{Reliability}
+
+A study is reliable if repetition of the study by different researchers using
+the original study's methodology would lead to the same results
+\citep{RunesonAndHost2009}. Reliability means that data and analysis are
+independent of the specific researcher(s) doing the study.  For the current
+study the identified reliability related threats are as follows:
 
 \begin{itemize}
-\item For practical considerations the time spent measuring each package had to
-be limited.  The time limit may have caused an assessment to have missed
-something something relevant.
-\item Our ranking is partly based on surface (shallow) measurement, which may
-not fully reveal the underlying qualities.
-\item The questions in the measurement template may not actually measure the
-qualities they are associated with.  For instance, we have assumed that
-maintainability is improved if a high percentage of identified issues are
-closed, but it is possible for a project with a wealth of ideas to have many
-open issues, and still be maintainable.
-\item With the exception of the interview data for 8 projects, we collected all
-of the information for each project from the artifacts available on the
-Internet. In some cases this source of data may mean we did not find evidence of
-something, like unit testing, not because the project didn't do it, but because
-no artifacts of this activity remained in the publicly available repository.
-\item As mentioned in Section~\ref{painpoints}, one interviewee was too busy
-to participate in a full interview, so they provided a version of written answers
-to us. Since we did not have the chance to explain our questions or ask them
-follow-up questions, there is a possibility of misinterpretation of the
-questions or answers.
+\item One individual does the manual measures for all packages. A different
+evaluator might find different results, due to differences in abilities,
+experiences, and biases.
+\item The manual measurements for the full set of packages took several months.
+Over this time the software repositories may have changed and the reviewer's
+judgement may have drifted.
+\end{itemize}
+
+In \citet{SmithEtAl2016} we reduced concern over the reliability risk associated
+with the reviewer's judgement  % BLIND REVIEW
+by demonstrating that the measurement process is reasonably reproducible.  In
+\citet{SmithEtAl2016} we graded five software products by two reviewers. Their
+rankings were almost identical. As long as each grader uses consistent
+definitions, the relative comparisons in the AHP results will be consistent
+between graders.
+
+\subsection{Construct Validity}
+
+\citet{RunesonAndHost2009} defines construct validity as when the adopted
+metrics represent what they are intended to measure. Our construct threats are
+often related to how we assume our measurements influences the various software
+qualities, as summarized in Section~\ref{sec_grading_software}. Specifically,
+our construct validity related threats include the following:
+
+\begin{itemize}
+\item We make indirect measurement of software qualities since meaningful direct
+measures for qualities like maintainability, reusability and verifiability, are
+unavailable.  We follow the usual assumption that developers achieve higher
+quality by following procedures and adhering to standards \citep[p.\
+112]{VanVliet2000}.
 \item As mentioned in Section~\ref{sec_result_installability}, we could not
@@
 no alternative for \textit{DICOM Viewer}. We might underestimate their rank due
-to an uncommon technical issue.
+to this uncommon technical issue.
+\item We only use two pieces of data to measure surface robustness. This is
+likely part of the reason for limited variation in the robustness scores
+(Figure~\ref{fg_robustness_scores}). We could add more robustness data by
+pushing the software to deal with more unexpected situations, like a broken
+Internet connection, but this would require a larger investment of measurement
+time. 
+\item We may have inaccurately estimated maintainability by assuming a higher
+ratio of comments to source code improves maintainability. Moreover, we assumed
+that maintainability is improved if a high percentage of issues are closed, but
+a project may have a wealth of open issues, and still be maintainable.
+\item We assess reusability by the number of code files and LOC per file. This
+measure is indicative of modularity, but it does not necessarily mean a good
+modularization. The modules may not be general enough to be easily reused, or
+the formatting may be poor, or the understandability of the code may be low.
+\item The understandability measure relies on 10 random source code files, but
+the 10 files will not necessarily be representative. 
+\item As discussed in Section~\ref{Sec_OverallQ}, our overall AHP ranking makes
+the unrealistic assumption of equal weighting.
+\item We approximated popularity by stars and watches
+(Section~\ref{Sec_VsCommunityRanking}), but this assumption may not be valid. 
+\item As mentioned in Section~\ref{sec_interview_methods}, one interviewee was
+too busy to participate in a full interview, so they provided written answers
+instead. Since we did not have the chance to explain our questions or ask them
+follow-up questions, there is a possibility of misinterpretation of the
+questions or answers.
+\item In building Table~\ref{Tbl_Guidelines} some judgement was necessary on our
+part, since not all guidelines use the same names for artifacts that contain
+essentially the same information.
 \end{itemize}
 
-\begin{description}
-    \item[Internal Validity:] ``This aspect relates to the examination of causal
-    relations. Internal validity examines whether an experimental
-    treatment/condition makes a difference or not, and whether there is evidence
-    to support the claim'' \citep{AmpatzoglouEtAl2019}.
-\end{description}
+\subsection{Internal Validity} \label{Sec_InternalValidity}
+
+Internal validity means that discovered causal relations are trustworthy and
+cannot be explained by other factors \citep{RunesonAndHost2009}. In our
+methodology the internal validity threats include the following:
 
 \begin{itemize}
-    \item Many of the measurement template questions look for the presence of
-    certain artifacts, like a user manual, or a getting started tutorial.  We
-    have implicitly assumed that the presence of these artifacts is an
-    indication of a given quality, but this is an indirect measure.  It may be
-    possible to achieve qualities without the artifacts we sought.  A direct
-    measure of quality, like an usability experiment, would not have this
-    problem.
-    \item We compared our rankings to the rankings by the community and the
-    Domain Expert, but we have assumed all qualities are equally weighted.  The
-    community and the Domain Expert likely have a more complex weighting between
-    qualities.
-    \item We assumed there was a casual relationship between the number of files
-    and reusability, since we assumed multiple files implies modularity.  Of
-    course the code can be modular, and not divided into separate files.
-    Moreover, having many files does not immediately in itself imply a
-    decomposition based on sound design principles, like information hiding.
+\item In our search for software packages
+(Section~\ref{sec_software_selection}), we may have missed a relevant package.
+\item Our methodology assumes that all relevant software development activities
+will leave a trace in the repositories, but this is not necessarily true. For
+instance, the possibility exists that CI usage was higher than what we observed
+through the artifacts (Section~\ref{Sec_CompareTools}). As another example,
+although we saw little evidence of requirements
+(Section~\ref{Sec_CompareTools}), maybe teams keep this kind of information
+outside their repos, possibly in journal papers or technical reports.
+\item We interviewed a relatively small sample of 8 teams.  Their pain points
+(Section~\ref{painpoints}) may not be representative of the rest of their
+community.
 \end{itemize}
 
-\begin{description}
-    \item[External Validity:] ``Define the domain to which a study's findings
-    can be generalized'' \citep{ZhouEtAl2016}.
-\end{description}
+\subsection{External Validity}
+
+If the results of a study can be generalized (applied) to other
+situations/cases, then the study is externally valid \citep{RunesonAndHost2009}.
+We are confident that our search was exhaustive.  We do not believe that we
+missed any highly popular examples.  Therefore, the bulk of our validity
+concerns are internal (Section~\ref{Sec_InternalValidity}).
+However, our hope is that the trends observed, and the lessons learned for MI
+software can be applied to other research software.  With that in mind we
+identified the following threat to external validity:
 
 \begin{itemize}
-\item We interviewed eight teams, which is a good proportion of the 29. However,
-there is still a risk that the subset of the entire group does not represent the
-whole MI software community.
-\item We identified qualities that we believe the community will be interested
-in, and we weighted these qualities equally, but the qualities we chose, and
-their weighting, may not match the external reality.
-\item The number of GitHub stars, watches, and forks are not ideal measures of
-popularity.  \wss{Can we find a reference for this?}
+\item We cannot generalize our results if the development of MI software is
+fundamentally different from other research software.
 \end{itemize}
 
-\begin{description}
-    \item[Conclusion Validity:] ``Demonstrate that the operations of a study
-    such as the data collection procedure can be repeated, with the same
-    results'' \citep{ZhouEtAl2016}.
-\end{description}
-
-\begin{itemize}
-\item The grading template included an entry for the reviewer's impression.  We
-aimed for objectivity, but there is a risk that some scores may be subjective
-and biased.
-\item The measurements are made at an instant in time, and the instant differs
-by a few weeks between projects, due to the time needed to measure them. For the
-most part, the projects are living and continually changing, which means if we
-measured at a different time, our results may change.
-\end{itemize}
+Although there are differences, like the importance of data privacy for MI data,
+we found the approach to developing LBM software \citep{SmithEtAl2022} and MI
+software to be similar.  Except for the domain specific aspects, we believe that
+the trends observed in the current study are externally valid for other research
+software.
 

### 2022-10-31 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Edits to SOP for LBM paper
Edits to SOP for MI paper

@@
 
-The top three packages (ESPREsSo, Ludwig, Palabos) include most of
-the common and uncommon artifacts, but not many of the rare artifacts.  They
-collectively have all the common artifacts, except Palabos does not have theory
-notes.  The top three packages
-have most of the uncommon artifacts. Only one them (Ludwig) is missing design
-information, and only one (Palabos) does not have a user manual.  However, for
-Palabos there was a broken link on the package website indicating that such an
-artifact might exist. The developers later fixed this broken link, but this is
-not reflected in our data because it was not present at the time of data
-collection (mid-2020).
-Despite the broken link, Palabos does have a detailed and informative website.
-The top three ranked packages do not have many of the rare artifacts.
-None of them have any explicit API documentation.
-All of them have information on contributing to the project and
+The top three packages (ESPREsSo, Ludwig, Palabos) include most of the common
+and uncommon artifacts, but not many of the rare artifacts.  They collectively
+have all the common artifacts, except Palabos does not have theory notes.  The
+top three packages have most of the uncommon artifacts. Only one them (Ludwig)
+is missing design information, and only one (Palabos) does not have a user
+manual.  However, for Palabos there was a broken link on the package website
+indicating that such an artifact might exist. The developers later fixed this
+broken link, but this is not reflected in our data because it was not present at
+the time of data collection (mid-2020). Despite the broken link, Palabos does
+have a detailed and informative website. The top three ranked packages do not
+have many of the rare artifacts. None of them have any explicit API
+documentation. All of them have information on contributing to the project and
 one (Palabos) has an FAQ section or forum.
@@
 \href{https://google.github.io/styleguide/javaguide.html} {Google Java Style
-Guide} and \href{https://google.github.io/styleguide/cppguide.html} {Google \CC
+Guide}, and \href{https://google.github.io/styleguide/cppguide.html} {Google \CC
 Style Guide}.  Tools like \href{https://pypi.org/project/flake8/}{flake8} can be

### 2022-10-31 StateOfPractice/Papers/MedImageSoft_SOP-arXiv/MedImageSoft_SOP.tex
Edits to SOP for MI paper

@@
 \usepackage{amssymb}
-%% The amsthm package provides extended theorem environments
 \usepackage{amsmath,amsthm}
@@
 
+\newcommand{\CC}{C\nolinebreak\hspace{-.05em}\raisebox{.4ex}{\small\bf
++}\nolinebreak\hspace{-.10em}\raisebox{.4ex}{\small\bf +}}
+
 \journal{Academic Radiology}
 
-%\pgfplotsset{compat=1.13}
+\pgfplotsset{compat=1.13}
 
@@
 the first 9 qualities, we ranked the MI software with the Analytic Hierarchy
-Process (AHP). The top three software products were \textit{3D Slicer},
-\textit{ImageJ}, and \textit{OHIF Viewer}. By interviewing the developers, we
-identified three major pain points: i) lack of resources; ii) difficulty
-balancing between compatibility, maintainability, performance, and security;
-and, iii) lack of access to real-world datasets for testing. For future MI
-software projects, we propose adopting test-driven development, using continuous
-integration and continuous delivery (CI/CD), using git and GitHub, maintaining
-good documentation, supporting third-party plugins or extensions, considering
-web application solutions, and improving collaboration between different MI
-software projects. \wss{Update after the paper has been revised.}
+Process (AHP). The four top ranked software products are: \textit{3D Slicer},
+\textit{ImageJ}, \textit{Fiji}, and \textit{OHIF Viewer}. Our ranking is mostly
+consistent with the community's ranking, with four of our top five projects also
+appearing in the top five of a list ordered by stars-per-year. \textit{DICOM
+Viewer} and \textit{dmv} are exceptions; we ranked them lower than their
+estimated popularity because of difficulty with their installation.  Comparison
+for tool usage. Comparison for process. From interviewing the developers, we
+identified five pain points and two qualities of potential concern:: i) lack of
+resources; ii) difficulty balancing between compatibility, maintainability,
+performance, and security; and, iii) lack of access to real-world datasets for
+testing. For future MI software projects, we propose adopting test-driven
+development, using continuous integration and continuous delivery (CI/CD), using
+git and GitHub, maintaining good documentation, supporting third-party plugins
+or extensions, considering web application solutions, and improving
+collaboration between different MI software projects. \wss{Update after the
+paper has been revised.}
 
@@
 analytic, and medical applications \citep{FDA2021, enwiki:1034887445,
-Zhang2008}.  Figure~\ref{Fig_Example} shows an example image of the brain. Given
-the importance of MI software and the high number of competing software
-projects, we wish to understand the merits and drawbacks of the current
-development processes, tools, and methodologies.  We aim to assess through a
-software engineering lens the quality of the existing software with the hope of
-highlighting standout examples, understanding current pain points and providing
-guidelines and recommendations for future development.
+Zhang2008}.  Figure~\ref{Fig_Example}, which shows an image of the brain,
+highlights the importance and value of MI. Through MI medical practitioners and
+researchers can noninvasively gain insights into the human body, including
+information on injuries and illnesses. Given the importance of MI software and
+the high number of competing software projects, we wish to understand the merits
+and drawbacks of the current development processes, tools, and methodologies. We
+aim to assess through a software engineering lens the quality of the existing
+software with the hope of highlighting standout examples, understanding current
+pain points and providing guidelines and recommendations for future development.
 
@@
     \end{center}
-    \caption{Example brain image showing a multi-planar reformat using Horos (free open-source medical imaging/DICOM viewer for OSX, based on OsiriX)}
+    \caption{Example brain image showing a multi-planar reformat using Horos
+	(free open-source medical imaging/DICOM viewer for OSX, based on OsiriX)}
     \label{Fig_Example}
@@
 general. We wish to understand the impact of the often cited gap, or chasm,
-between software engineering and scientific programming \citep{Kelly2007,
+between software engineering and research software \citep{Kelly2007,
 Storer2017}. Although scientists spend a substantial proportion of their working
@@
 SmithAndMichalski2022}.  We designed the questions to learn about the
-community's interest in, and experience with, artifacts, tools, principles,
-processes, methodologies, and qualities.  When we mention artifacts we mean the
-documents, scripts and code that constitutes a software development project.
-Example artifacts include requirements, specifications, user manuals, unit
-tests, system tests, usability tests, build scripts, API (Application
+community's interest in, and experience with, software artifacts, tools,
+principles, processes, methodologies, and qualities.  When we mention artifacts
+we mean the documents, scripts and code that constitutes a software development
+project. Example artifacts include requirements, specifications, user manuals,
+unit tests, system tests, usability tests, build scripts, API (Application
 Programming Interface) documentation, READMEs, license documents, process
@@
 research question below we point to the section that contains our answer.  We
-start with identifying the relevant examples of MI software:
+start with identifying the relevant examples of MI software for the assessment
+exercise:
 
@@
 	What research software development practice could potentially address the
-	pain point concerns identified in \rqref{RQ_PainPoints}).
+	pain point concerns identified in \rqref{RQ_PainPoints}?
 	(Section~\ref{ch_recommendations})
@@
 
-We designed a general methodology to assess the state of the practice for SC
-software \citep{SmithEtAl2021, SmithAndMichalski2022}. Details can be found in
-Section~\ref{ch_methods}.  Our methodology has been applied to MI software
-\citep{Dong2021} and Lattice Boltzmann Solvers \citep{Michalski2021}.  This
-methodology builds off prior work to assess the state of the practice for such
-domains as Geographic Information Systems \citep{smith2018state}, Mesh
-Generators \citep{smith2016state}, Seismology software
-\citep{Smith2018Seismology}, and Statistical software for psychology
+We designed a general methodology to assess the state of the practice for
+research software \citep{SmithEtAl2021, SmithAndMichalski2022}. Details can be
+found in Section~\ref{ch_methods}.  Our methodology has been applied to MI
+software \citep{Dong2021} and Lattice Boltzmann Solvers \citep{SmithEtAl2022,
+Michalski2021}.  This methodology builds off prior work to assess the state of
+the practice for such domains as Geographic Information Systems
+\citep{smith2018state}, Mesh Generators \citep{smith2016state}, Seismology
+software \citep{Smith2018Seismology}, and Statistical software for psychology
 \citep{smith2018statistical}.  In keeping with the previous methodology, we have
@@
 be feasible for a team as small as one person, and for a short time, ideally
-around a person month of effort. We consider a person month as $20$ working
-days ($4$ weeks in a month, with $5$ days of work per week) at $8$ person hours
-per day, or $20 \times 8 = 160$ person hours.
-
-With our methodology, we first choose an SC domain (in the current case MI) and
-identify a list of about 30 software packages. (For measuring MI we used 29
-software packages.)  We approximately measure the qualities of each package by
-filling in a grading template. Compared with our previous methodology, the new
-methodology also includes repository based metrics, such as the number of files,
-number of lines of code, percentage of issues that are closed, etc.  With the
-quantitative data in the grading template, we rank the software with the
-Analytic Hierarchy Process (AHP) (Section~\ref{ch_background} provides details).
-After this, as another addition to our previous methodology, we interview some
-development teams to further understand the status of their development process.
+around a person month of effort. We consider a person month as $20$ working days
+($4$ weeks in a month, with $5$ days of work per week) at $8$ person hours per
+day, or $20 \times 8 = 160$ person hours.
+
+With our methodology, we first choose a research software domain (in the current
+case MI) and identify a list of about 30 software packages. (For measuring MI we
+used 29 software packages.)  We approximately measure the qualities of each
+package by filling in a grading template. Compared with our previous
+methodology, the new methodology also includes repository based metrics, such as
+the number of files, number of lines of code, percentage of issues that are
+closed, etc.  With the quantitative data in the grading template, we rank the
+software with the Analytic Hierarchy Process (AHP) (Section~\ref{ch_background}
+provides details). After this, as another addition to our previous methodology,
+we interview some development teams to further understand the status of their
+development process.
 
@@
 developers achieve their works, and the flaws and potential risks in the final
-product. Thus, it brings much more convenience to researchers analyzing the
-qualities of the project.
+product. Thus, OSS is suitable for researchers analyzing the qualities of a
+project.
 
@@
 \item \textbf{Commercial Software} ``Commercial software is software developed
-by a business as part of its business'' \citep{GNU2019}. Typically speaking, commercial software requires users to pay to access all of its features,
+by a business as part of its business'' \citep{GNU2019}. Typically speaking,
+commercial software requires users to pay to access all of its features,
 excluding access to the source code. However, some commercial software is also
@@
 as $8$ and $7$, respectively; therefore, on the 9-point scale, 3D Slicer compared
-to Ginkgo CADx is 2 and Ginkgo CADX versus 3D Slicer is 1/2, as shown in the
+to Ginkgo CADx is 2 and Ginkgo CADx versus 3D Slicer is 1/2, as shown in the
 sample AHP calculations (Table~\ref{Tbl_SampleAHP}).
@@
 but they should be a user of domain software.  Given that the domain experts are
-likely to be busy people, the measurement process cannot put to much of a burden
+likely to be busy people, the measurement process cannot put too much of a burden
 on their time.  For the current assessment, our Domain Expert (and paper
@@
 
-In advance of the first meeting with the Domain Expert, the expert is asked to
+In advance of the first meeting with the Domain Expert, they are asked to
 create a list of top software packages in the domain.  This is done to help
@@
 We initially identified 48 MI candidate software projects from the literature
-\citep{Bjorn2017, Bruhschwein2019, Haak2015}, online articles \citep{Emms2019,
+\citep{Bjorn2017, Bruhschwein2019, Haak2015}, on-line articles \citep{Emms2019,
 Hasan2020, Mu2019}, and forum discussions \citep{Samala2014}.  The full list of
@@
 \item We focused on the MI software that provides visualization functions, as
-described in Section~\ref{sec_scope}. We removed seven packages that were
+described in Section~\ref{sec_scope}. Furthermore, we removed seven packages that were
 toolkits or libraries, such as \textit{VTK}, \textit{ITK}, and \textit{dcm4che}.
@@
 assessed through the developer interviews (Section~\ref{sec_interview_methods}).
-For each software package, we fill-in the template questions. To stay within the
-target of 160 person hours to measure the domain, we allocated between 1 to 4
-hours for each package. Project developers can be contacted for help regarding
-installation, if necessary, but a cap of about 2 hours is imposed on the
-installation process, to keep the overall measurement time feasible. An excerpt
-of the spreadsheet is shown in Figure~\ref{fg_grading_template_example}.  A
-column is included for each measured software package.
+For each software package, we fill in the template questions. To stay within the
+target of 160 person hours to measure the domain, we allocated between one and
+four hours for each package. Project developers can be contacted for help
+regarding installation, if necessary, but we impose a cap of about two hours on
+the installation process, to keep the overall measurement time feasible.
+Figure~\ref{fg_grading_template_example} shows an excerpt of the spreadsheet.
+The spreadsheet includes a column for each measured software package.
 
 \begin{figure}[!ht]
-\includegraphics[scale=0.67]{figures/template.pdf}
+\includegraphics[scale=0.66]{figures/template.pdf}
 \caption{Grading template example}
@@
 
-The full template consists of 108 questions categorized under 9 qualities.  The
-questions were designed to be unambiguous, quantifiable and measurable with
-limited time and domain knowledge. The measures are grouped under headings for
-each quality, and one for summary information. The summary information (shown in
+The full template consists of 108 questions categorized under 9 qualities.  We
+designed the questions to be unambiguous, quantifiable, and measurable with
+limited time and domain knowledge. We group the measures under headings for each
+quality, and one for summary information. The summary information (shown in
 Figure~\ref{fg_grading_template_example}) is the first section of the template.
@@
 software intended for public use.  Private means software aimed only at a
-specific group, while the concept category is used for software written simply
-to demonstrate algorithms or concepts. The three categories of development
-models are (open source, free-ware and commercial) are discussed in
+specific group, while the concept category is for software written simply to
+demonstrate algorithms or concepts. The three categories of development models
+are (open source, free-ware and commercial) are discussed in
 Section~\ref{sec_software_categories}.  Information in the summary section sets
@@
 among the collection of ``yes'', ``no'', ``n/a'', ``unclear'', a number, a
-string, a date, a set of strings, etc. Each quality is assigned an overall
-score, between 1 and 10, based on all the previous questions.  Several of the
-qualities use the word ``surface''.  This is to highlight that, for these
+string, a date, a set of strings, etc. The grader assigns each quality an
+overall score, between 1 and 10, based on all the previous questions.  Several
+of the qualities use the word ``surface''.  This is to highlight that, for these
 qualities in particular, the best that we can do is a shallow measure.  For
 instance, we are not currently doing any experiments to measure usability.
-Instead, we are looking for an indication that usability was considered by the
-developers.  We do this by looking for cues in the documentation, like a getting
-started manual, a user manual and documentation of expected user
-characteristics.  Below is a summary of how we assess adoption of best practices
-by measuring each quality.
+Instead, we are looking for an indication that the developers considered
+usability.  We do this by looking for cues in the documentation, like a getting
+started manual, a user manual and a statement of expected user characteristics.
+Below is a summary of how we assess adoption of best practices by measuring each
+quality.
 
@@
 including source code, documents, and building scripts, significantly influence
-its maintainability. Thus we check each project for the presence of such
+its maintainability. Thus, we check each project for the presence of such
 artifacts as API documentation, bug tracker information, release notes, test
@@
 \item \textbf{Reusability} We count the total number of code files for each
-project. Projects with a large number of components potentially provide more
-choices for reuse. Furthermore, well-modularized code, which tends to have
-smaller parts in separate files, is typically easier to reuse. Thus, we assume
-that projects with more code files and less Lines of Code (LOC) per file are
-more reusable. We also consider projects with API documentation as delivering
-better reusability.
+project. Projects with numerous components potentially provide more choices for
+reuse. Furthermore, well-modularized code, which tends to have smaller parts in
+separate files, is typically easier to reuse. Thus, we assume that projects with
+more code files and fewer Lines of Code (LOC) per file are more reusable. We also
+consider projects with API documentation as delivering better reusability.
 
@@
 file, such as whether the identifiers, parameters, indentation, and formatting
-are consistent, whether the constants (other than 0 and 1) are hardcoded, and
+are consistent, whether the constants (other than 0 and 1) are not hardcoded, and
 whether the code is modularized. We also check the descriptive information for
@@
 not upload a copy of it to GitHub until 2020. Some GitHub data can be found
-using its the GitHub Application Program Interface (API) via the following url:
+using its GitHub Application Program Interface (API) via the following url:
 \textit{https://api.github.com/repos/[owner]/[repository]} where [owner] and
@@
 As in \citet{SmithEtAl2016}, Virtual machines (VMs) were used to provide an
-optimal testing environments for each package. VMs were used because it is
-easier to start with a fresh environment without having to worry about existing
+optimal testing environment for each package. We used VMs because it is easier
+to start with a fresh environment, without having to worry about existing
 libraries and conflicts. Moreover, when the tests are complete the VM can be
@@
 starts from a clean slate, which removes ``works-on-my-computer'' errors. When
-filling in the measurement template spreadsheet, the details for each VM
-are noted, including hypervisor and operating system version.
+filling in the measurement template, the grader notes the details for each VM,
+including hypervisor and operating system version.
 
@@
 threats to software quality, and the developers' strategies to address these
-threats.  Therefore, part of our methodology is to also interview developers.
+threats.  Therefore, part of our methodology involves interviewing developers.
 
@@
 \citet{SmithEtAl2021}. Some questions are about the background of the software,
-the development teams, the interviewees, and how they organize the projects. We
-also ask about the developer's understandings of the users. Some questions focus
-on the current and past difficulties, and the solutions the team has found, or
-will try. We also discuss the importance and current situations of
-documentation. A few questions are about specific software qualities, such as
-maintainability, understandability, usability, and reproducibility. The
-interviews are semi-structured based on the question list; we ask follow-up
-questions when necessary.  The interview process presented here was approved by
-the McMaster University Research Ethics Board under the application number 
+the development teams, the interviewees, and how they organize their projects.
+We also ask about the developer's understanding of the users. Some questions
+focus on the current and past difficulties, and the solutions the team has
+found, or plan to try. We also discuss documentation, both with respect to how
+it is currently done, and how it is perceived. A few questions are about
+specific software qualities, such as maintainability, understandability,
+usability, and reproducibility. The interviews are semi-structured based on the
+question list; we ask follow-up questions when necessary.  The interview process
+presented here was approved by the McMaster University Research Ethics Board
+under the application number 
 \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/MACREM/Application.pdf}
@@
 projects websites, code repository, publications, and from biographic pages at
-the teams's institutions.  In the end nine developers from eight of the projects
+the teams' institutions.  In the end nine developers from eight of the projects
 agreed to participate: \textit{3D Slicer}, \textit{INVESALIUS 3}, \textit{dwv},
@@
 their answers. In one case two developers from the same project agreed to be
-interviewed. Meetings were held on-line using either Zoom or Teams, which
-facilitated recording and automatic transcription of the meetings. The full
-interview answers can be found in \citet{Dong2021}.
+interviewed. We held the meetings on-line using either Zoom or Teams, which
+facilitated recording and automatic transcription. The full interview answers
+can be found in \citet{Dong2021}.
 
@@
 
-Table \ref{tab_final_list} shows the 29 software packages that we measured,
+Table~\ref{tab_final_list} shows the 29 software packages that we measured,
 along with summary data collected in the year 2020. We arrange the items in
 descending order of LOC. We found the initial release dates (Rlsd) for most
-projects and marked the two unknown dates with ``?''. We used the date of the
-latest change to each code repository to decide the latest update. We found
-funding information (Fnd) for only eight projects.  For the number of
-contributors (NOC) we considered anyone who made at least one accepted commit as
+projects and marked the two unknown dates with ``?''. The date of the last
+update is the date of the latest update, at the time of measurement. We found
+funding information (Fnd) for only eight projects.  For the Number Of
+Contributors (NOC) we considered anyone who made at least one accepted commit as
 a contributor. The NOC is not usually the same as the number of long-term
@@
  &  &  &  &  &  & W & M & L &  \\ \midrule
-ParaView \citep{Ahrens2005} & 2002 & 2020-10 & X & 100 & 886326 & X & X & X & X \\
-Gwyddion \citep{Nevcas2012} & 2004 & 2020-11 &  & 38 & 643427 & X & X & X &  \\
-Horos \citep{horosproject2020} & ? & 2020-04 &  & 21 & 561617 &  & X &  &  \\
-OsiriX Lite \citep{PixmeoSARL2019} & 2004 & 2019-11 &  & 9 & 544304 &  & X &  &  \\
-3D Slicer \citep{Kikinis2014} & 1998 & 2020-08 & X & 100 & 501451 & X & X & X &  \\
-Drishti \citep{Limaye2012} & 2012 & 2020-08 &  & 1 & 268168 & X & X & X &  \\
-Ginkgo CADx \citep{Wollny2020} & 2010 & 2019-05 &  & 3 & 257144 & X & X & X &  \\
-GATE \citep{Jan2004} & 2011 & 2020-10 &  & 45 & 207122 &  & X & X &  \\
-3DimViewer \citep{TESCAN2020} & ? & 2020-03 & X & 3 & 178065 & X & X &  &  \\
-medInria \citep{Fillard2012} & 2009 & 2020-11 &  & 21 & 148924 & X & X & X &  \\
-BioImage Suite Web \citep{Papademetris2005} & 2018 & 2020-10 & X & 13 & 139699 &
-X & X & X & X \\
-Weasis \citep{Roduit2021} & 2010 & 2020-08 &  & 8 & 123272 & X & X & X &  \\
-AMIDE \citep{Loening2017} & 2006 & 2017-01 &  & 4 & 102827 & X & X & X &  \\
-XMedCon \citep{Nolf2003} & 2000 & 2020-08 &  & 2 & 96767 & X & X & X &  \\
-ITK-SNAP \citep{Yushkevich2006} & 2006 & 2020-06 & X & 13 & 88530 & X & X & X &  \\
-Papaya \citep{UTHSCSA2019} & 2012 & 2019-05 &  & 9 & 71831 & X & X & X &  \\
-OHIF Viewer \citep{Ziegler2020} & 2015 & 2020-10 &  & 76 & 63951 & X & X & X & X \\
-SMILI \citep{Chandra2018} & 2014 & 2020-06 &  & 9 & 62626 & X & X & X &  \\
-INVESALIUS 3 \citep{Amorim2015} & 2009 & 2020-09 &  & 10 & 48605 & X & X & X &  \\
-dwv \citep{Martelli2021} & 2012 & 2020-09 &  & 22 & 47815 & X & X & X & X \\
-DICOM Viewer \citep{Afsar2021} & 2018 & 2020-04 & X & 5 & 30761 & X & X & X &  \\
-MicroView \citep{ParallaxInnovations2020} & 2015 & 2020-08 &  & 2 & 27470 & X & X & X &  \\
-MatrixUser \citep{Liu2016} & 2013 & 2018-07 &  & 1 & 23121 & X & X & X &  \\
-Slice:Drop \citep{Haehn2013} & 2012 & 2020-04 &  & 3 & 19020 & X & X & X & X \\
-dicompyler \citep{Panchal2010} & 2009 & 2020-01 &  & 2 & 15941 & X & X &  &  \\
-Fiji \citep{Schindelin2012} & 2011 & 2020-08 & X & 55 & 10833 & X & X & X &  \\
-ImageJ \citep{Rueden2017} & 1997 & 2020-08 & X & 18 & 9681 & X & X & X &  \\
-MRIcroGL \citep{Rorden2021} & 2015 & 2020-08 &  & 2 & 8493 & X & X & X &  \\
-DicomBrowser \citep{Archie2012} & 2012 & 2020-08 &  & 3 & 5505 & X & X & X &  \\ \bottomrule
+ParaView \citep{Ahrens2005} & 2002 & 2020-10 & \checkmark & 100 & 886326 & \checkmark & \checkmark & \checkmark & \checkmark \\
+Gwyddion \citep{Nevcas2012} & 2004 & 2020-11 &  & 38 & 643427 & \checkmark & \checkmark & \checkmark &  \\
+Horos \citep{horosproject2020} & ? & 2020-04 &  & 21 & 561617 &  & \checkmark &  &  \\
+OsiriX Lite \citep{PixmeoSARL2019} & 2004 & 2019-11 &  & 9 & 544304 &  & \checkmark &  &  \\
+3D Slicer \citep{Kikinis2014} & 1998 & 2020-08 & \checkmark & 100 & 501451 & \checkmark & \checkmark & \checkmark &  \\
+Drishti \citep{Limaye2012} & 2012 & 2020-08 &  & 1 & 268168 & \checkmark & \checkmark & \checkmark &  \\
+Ginkgo CADx \citep{Wollny2020} & 2010 & 2019-05 &  & 3 & 257144 & \checkmark & \checkmark & \checkmark &  \\
+GATE \citep{Jan2004} & 2011 & 2020-10 &  & 45 & 207122 &  & \checkmark & \checkmark &  \\
+3DimViewer \citep{TESCAN2020} & ? & 2020-03 & \checkmark & 3 & 178065 & \checkmark & \checkmark &  &  \\
+medInria \citep{Fillard2012} & 2009 & 2020-11 &  & 21 & 148924 & \checkmark & \checkmark & \checkmark &  \\
+BioImage Suite Web \citep{Papademetris2005} & 2018 & 2020-10 & \checkmark & 13 & 139699 &
+\checkmark & \checkmark & \checkmark & \checkmark \\
+Weasis \citep{Roduit2021} & 2010 & 2020-08 &  & 8 & 123272 & \checkmark & \checkmark & \checkmark &  \\
+AMIDE \citep{Loening2017} & 2006 & 2017-01 &  & 4 & 102827 & \checkmark & \checkmark & \checkmark &  \\
+XMedCon \citep{Nolf2003} & 2000 & 2020-08 &  & 2 & 96767 & \checkmark & \checkmark & \checkmark &  \\
+ITK-SNAP \citep{Yushkevich2006} & 2006 & 2020-06 & \checkmark & 13 & 88530 & \checkmark & \checkmark & \checkmark &  \\
+Papaya \citep{UTHSCSA2019} & 2012 & 2019-05 &  & 9 & 71831 & \checkmark & \checkmark & \checkmark &  \\
+OHIF Viewer \citep{Ziegler2020} & 2015 & 2020-10 &  & 76 & 63951 & \checkmark & \checkmark & \checkmark & \checkmark \\
+SMILI \citep{Chandra2018} & 2014 & 2020-06 &  & 9 & 62626 & \checkmark & \checkmark & \checkmark &  \\
+INVESALIUS 3 \citep{Amorim2015} & 2009 & 2020-09 &  & 10 & 48605 & \checkmark & \checkmark & \checkmark &  \\
+dwv \citep{Martelli2021} & 2012 & 2020-09 &  & 22 & 47815 & \checkmark & \checkmark & \checkmark & \checkmark \\
+DICOM Viewer \citep{Afsar2021} & 2018 & 2020-04 & \checkmark & 5 & 30761 & \checkmark & \checkmark & \checkmark &  \\
+MicroView \citep{ParallaxInnovations2020} & 2015 & 2020-08 &  & 2 & 27470 & \checkmark & \checkmark & \checkmark &  \\
+MatrixUser \citep{Liu2016} & 2013 & 2018-07 &  & 1 & 23121 & \checkmark & \checkmark & \checkmark &  \\
+Slice:Drop \citep{Haehn2013} & 2012 & 2020-04 &  & 3 & 19020 & \checkmark & \checkmark & \checkmark & \checkmark \\
+dicompyler \citep{Panchal2010} & 2009 & 2020-01 &  & 2 & 15941 & \checkmark & \checkmark &  &  \\
+Fiji \citep{Schindelin2012} & 2011 & 2020-08 & \checkmark & 55 & 10833 & \checkmark & \checkmark & \checkmark &  \\
+ImageJ \citep{Rueden2017} & 1997 & 2020-08 & \checkmark & 18 & 9681 & \checkmark & \checkmark & \checkmark &  \\
+MRIcroGL \citep{Rorden2021} & 2015 & 2020-08 &  & 2 & 8493 & \checkmark & \checkmark & \checkmark &  \\
+DicomBrowser \citep{Archie2012} & 2012 & 2020-08 &  & 3 & 5505 & \checkmark & \checkmark & \checkmark &  \\ \bottomrule
 \end{tabular}
@@
 of the project's code; in most cases projects also use other languages.  The
-most popular language is C++, with almost 40\% of projects (11 of 29).  The two
+most popular language is \CC, with almost 40\% of projects (11 of 29).  The two
 least popular choices are Pascal and Matlab, with around 3\% of projects each
@@
 \begin{figure}[!ht]
-\includegraphics[scale=0.48]{figures/installability_scores.pdf}
+\includegraphics[scale=0.47]{figures/installability_scores.pdf}
 \caption{AHP installability scores}
@@
 \textit{3D Slicer} has the highest score because it had easy to follow
-installation instructions, and the installation processes were automated, fast,
-and frustration-free, with all dependencies automatically added. There were also
-no errors during the installation and uninstallation steps. Many other software
-packages also had installation instructions and automated installers, and we had
-no trouble installing them, such as \textit{INVESALIUS 3}, \textit{Gwyddion},
-\textit{XMedCon}, and \textit{MicroView}. We determined their scores based on
-the understandability of the instructions, installation steps, and user
-experience. Since \textit{BioImage Suite Web} and \textit{Slice:Drop} needed no
-installation, we gave them high scores. \textit{BioImage Suite Web} also
-provided an option to download cache for offline usage, which was easy
-to apply.
-
-\textit{dwv}, \textit{GATE}, and \textit{DICOM Viewer} showed severe
+installation instructions, and an automated, fast, frustration-free installation
+process. The installer added all dependencies automatically and no errors
+occurred during the installation and uninstallation steps. Many other software
+packages also had installation instructions and automated installers.  We had no
+trouble installing the following packages: \textit{INVESALIUS 3},
+\textit{Gwyddion}, \textit{XMedCon}, and \textit{MicroView}. We determined their
+scores based on the understandability of the instructions, installation steps,
+and user experience. Since \textit{BioImage Suite Web} and \textit{Slice:Drop}
+needed no installation, we gave them high scores. \textit{BioImage Suite Web}
+also provided an option to download cache for offline usage, which was easy to
+apply.
+
+\textit{GATE}, \textit{dwv}, and \textit{DICOM Viewer} showed severe
 installation problems. We were not able to install them, even after a reasonable
@@
 from the source code, but we were able to proceed with measuring other qualities
-using a deployed online version for \textit{dwv}, and a VM version for
+using a deployed on-line version for \textit{dwv}, and a VM version for
 \textit{GATE}. For \textit{DICOM Viewer} we could not install the NextCloud
@@
 
-\textit{MatrixUser} has a lower score because it depended on Matlab. We assessed
+\textit{MatrixUser} has a lower score because it depends on Matlab. We assessed
 the score from the point of view of a user that would have to install Matlab and
@@
 
-The scores of correctness \& verifiability are shown in
-Figure~\ref{fg_correctness_verifiability_scores}. Generally speaking, the
-packages with higher scores adopted more techniques to improve correctness, and
-had better documentation for us to verify against.  For instance, we looked for
-evidence of unit testing, since it benefits most parts of the software's life
-cycle, such as designing, coding, debugging, and optimization
-\citep{Hamill2004}.  We only found evidence of unit testing for about half of
-the projects. We identified five projects using CI/CD tools: \textit{3D Slicer},
-\textit{ImageJ}, \textit{Fiji}, \textit{dwv}, and \textit{OHIF Viewer}.
+Figure~\ref{fg_correctness_verifiability_scores} shows the scores of correctness
+and verifiability. Generally speaking, the packages with higher scores adopted
+more techniques to improve correctness, and had better documentation for us to
+verify against.  For instance, we looked for evidence of unit testing, since it
+benefits most parts of the software's life cycle, such as designing, coding,
+debugging, and optimization \citep{Hamill2004}.  We only found evidence of unit
+testing for about half of the projects. We identified five projects using CI/CD
+tools: \textit{3D Slicer}, \textit{ImageJ}, \textit{Fiji}, \textit{dwv}, and
+\textit{OHIF Viewer}.
 
 \begin{figure}[!ht]
-\includegraphics[scale=0.48]{figures/correctness_verifiability_scores.pdf}
+\includegraphics[scale=0.47]{figures/correctness_verifiability_scores.pdf}
 \caption{AHP correctness \& verifiability scores}
@@
 
-Even for some projects with well-organized documents, requirements
+Even for some projects with well-organized documentation, requirements
 specifications and theory manuals were still missing.  We could not identify
-theory manuals for all projects and we did not find requirements specifications
-for most projects. The only document we found was a road map of \textit{3D
-Slicer}, which contained design requirements for upcoming changes.
+theory manuals for all projects, and we did not find requirements specifications
+for most projects. The only requirements-related document we found was a road
+map of \textit{3D Slicer}, which contained design requirements for upcoming
+changes.
 
@@
 when loading damaged image files, without showing any descriptive error message.
-On the other hand, we did not find any problems with the online version of
+We did not find any problems with the on-line version of
 \textit{dwv}.
@@
 \begin{figure}[!ht]
-\includegraphics[scale=0.48]{figures/reliability_scores.pdf}
+\includegraphics[scale=0.47]{figures/reliability_scores.pdf}
 \caption{AHP surface reliability scores}
@@
 \begin{figure}[!ht]
-\includegraphics[scale=0.48]{figures/robustness_scores.pdf}
+\includegraphics[scale=0.47]{figures/robustness_scores.pdf}
 \caption{AHP surface robustness scores}
@@
 
-Figure \ref{fg_usability_scores} shows the AHP scores for surface usability. The
+Figure~\ref{fg_usability_scores} shows the AHP scores for surface usability. The
 software with higher scores usually provided both comprehensive documented
 guidance and a good user experience. \textit{INVESALIUS 3} provided an excellent
-example of a detailed and precise user manual. \textit{GATE} also provided a
-large number of documents, but unfortunately we had difficulty understanding and
-using them. We found getting started tutorials for only 11 projects, but a user
-manual for 22 projects. \textit{MRIcroGL} was the only project that explicitly
+example of a detailed and precise user manual. \textit{GATE} also provided
+numerous documents, but unfortunately we had difficulty understanding and using
+them. We found getting started tutorials for only 11 projects, but a user manual
+for 22 projects. \textit{MRIcroGL} was the only project that explicitly
 documented expected user characteristics.
@@
 \begin{figure}[!ht]
-\includegraphics[scale=0.48]{figures/usability_scores.pdf}
+\includegraphics[scale=0.47]{figures/usability_scores.pdf}
 \caption{AHP surface usability scores}
@@
 Figure~\ref{fg_maintainability_scores} shows the ranking results for
-maintainability. We marked \textit{3D Slicer} with the highest score because
-we found it to have the most comprehensive artifacts. For example, as far as we
-could find, only a few of the 29 projects had a product, developer's
-manual, or API documentation, and only \textit{3D Slicer}, \textit{ImageJ},
-\textit{Fiji} included all three documents. Moreover, \textit{3D Slicer} has a
-much higher percentage of closed issues (91.65\%) compared to \textit{ImageJ}
-(52.49\%) and \textit{Fiji} (63.79\%). Table~\ref{tab_maintainability_docs}
-shows which projects had these documents, in the descending order of their
-maintainability scores.
+maintainability. We gave \textit{3D Slicer} the highest score because we found
+it to have the most comprehensive artifacts. For example, as far as we could
+find, only a few of the 29 projects had a product, developer's manual, or API
+documentation, and only \textit{3D Slicer}, \textit{ImageJ}, \textit{Fiji}
+included all three documents. Moreover, \textit{3D Slicer} has a much higher
+percentage of closed issues (91.65\%) compared to \textit{ImageJ} (52.49\%) and
+\textit{Fiji} (63.79\%). Table~\ref{tab_maintainability_docs} shows which
+projects had these documents, in descending order of their maintainability
+scores.
 
 \begin{figure}[!ht]
-\includegraphics[scale=0.48]{figures/maintainability_scores.pdf}
+\includegraphics[scale=0.47]{figures/maintainability_scores.pdf}
 \caption{AHP maintainability scores}
@@
 \toprule
-\multicolumn{1}{c}{Software} & Prod.\ roadmap & Dev.\ manual & API doc. \\ 
+\multicolumn{1}{c}{Software} & Prod.\ Roadmap & Dev.\ Manual & API Doc. \\ 
 \midrule
-3D Slicer & X & X & X \\
-ImageJ & X & X & X \\
-Weasis &  & X &  \\
-OHIF Viewer &  & X & X \\
-Fiji & X & X & X \\
-ParaView & X &  &  \\
-SMILI &  &  & X \\
-medInria &  & X &  \\
-INVESALIUS 3 & X &  &  \\
-dwv &  &  & X \\
-BioImage Suite Web &  & X &  \\
-Gwyddion &  & X & X \\ 
+3D Slicer & \checkmark & \checkmark & \checkmark \\
+ImageJ & \checkmark & \checkmark & \checkmark \\
+Weasis &  & \checkmark &  \\
+OHIF Viewer &  & \checkmark & \checkmark \\
+Fiji & \checkmark & \checkmark & \checkmark \\
+ParaView & \checkmark &  &  \\
+SMILI &  &  & \checkmark \\
+medInria &  & \checkmark &  \\
+INVESALIUS 3 & \checkmark &  &  \\
+dwv &  &  & \checkmark \\
+BioImage Suite Web &  & \checkmark &  \\
+Gwyddion &  & \checkmark & \checkmark \\ 
 \bottomrule
@@
 
-27 of the 29 projects used git as the version control tool, with 24 of these
+Twenty-seven of the 29 projects used git as the version control tool, with 24 of these
 using GitHub. \textit{AMIDE} used Mercurial and \textit{Gwyddion} used
@@
 Table~\ref{tab_maintainability_docs}, seven projects had API documents. We also
-assumed that projects with more code files and less LOC per code file as more
+assumed that projects with more code files and less LOC per code file are more
 reusable. Table \ref{tab_loc_per_file} shows the number of text-based files by
-projects, which we used to approximate the number of code files. The table also
+project, which we used to approximate the number of code files. The table also
 lists the total number of lines (including comments and blanks), LOC, and
@@
 \begin{figure}[!ht]
-\includegraphics[scale=0.48]{figures/reusability_scores.pdf}
+\includegraphics[scale=0.47]{figures/reusability_scores.pdf}
 \caption{AHP reusability scores}
@@
 \toprule
-\multirow{2}{*}{Software} & \multirow{2}{*}{Text files} & \multirow{2}{*}{Total lines} & \multirow{2}{*}{LOC} & \multirow{2}{*}{LOC/file} \\
+\multirow{2}{*}{Software} & \multirow{2}{*}{Text Files} & \multirow{2}{*}{Total Lines} & \multirow{2}{*}{LOC} & \multirow{2}{*}{LOC/file} \\
  &  &  &  &  \\ 
@@
 \end{tabular}
-\caption{Number of files and lines (entries are sorted in descending order of
-their reusability scores)}
+\caption{Number of files and lines (sorted in descending order of reusability
+scores)}
 \label{tab_loc_per_file}
@@
 understandability. All projects had a consistent coding style with parameters in
-the same order for all functions; the code was modularized; the comments were
-clear, indicating what is being done, not how. However, we only found explicit
-identification of a coding standard for 3 out of the 29: \textit{3D Slicer},
-\textit{Weasis}, and \textit{ImageJ}. We also found hard-coded constants (rather
-than symbolic constants) in \textit{medInria}, \textit{dicompyler},
-\textit{MicroView}, and \textit{Papaya}. We did not find any reference to the
-algorithms used in projects \textit{XMedCon}, \textit{DicomBrowser},
-\textit{3DimViewer}, \textit{BioImage Suite Web}, \textit{Slice:Drop},
-\textit{MatrixUser}, \textit{DICOM Viewer}, \textit{dicompyler}, and
-\textit{Papaya}.
+the same order for all functions; modularized code; and, clear comments, indicating
+what is done, not how. However, we only found explicit identification of a
+coding standard for 3 out of the 29: \textit{3D Slicer}, \textit{Weasis}, and
+\textit{ImageJ}. We also found hard-coded constants (rather than symbolic
+constants) in \textit{medInria}, \textit{dicompyler}, \textit{MicroView}, and
+\textit{Papaya}. We did not find any reference to the algorithms used in
+projects \textit{XMedCon}, \textit{DicomBrowser}, \textit{3DimViewer},
+\textit{BioImage Suite Web}, \textit{Slice:Drop}, \textit{MatrixUser},
+\textit{DICOM Viewer}, \textit{dicompyler}, and \textit{Papaya}.
 
 \begin{figure}[!ht]
-\includegraphics[scale=0.48]{figures/understandability_scores.pdf}
+\includegraphics[scale=0.47]{figures/understandability_scores.pdf}
 \caption{AHP surface understandability scores}
@@
 \begin{figure}[!ht]
-\includegraphics[scale=0.48]{figures/visibility_transparency_scores.pdf}
+\includegraphics[scale=0.47]{figures/visibility_transparency_scores.pdf}
 \caption{AHP visibility/transparency scores}
@@
 \toprule
-Software & Dev process & Proj status & Dev env & Rls notes \\ 
+Software & Dev.\ Process & Proj.\ Status & Dev.\ Env. & Rls.\ Notes \\ 
 \midrule
-3D Slicer & X & X & X & X \\
-ImageJ & X & X & X & X \\
-Fiji & X & X & X &  \\
-MRIcroGL &  &  &  & X \\
-Weasis &  &  & X & X \\
-ParaView &  & X &  &  \\
-OHIF Viewer &  &  & X & X \\
-DICOM Viewer &  &  & X & X \\
-medInria &  &  & X & X \\
-SMILI &  &  &  & X \\
-Drishti &  &  &  & X \\
-INVESALIUS 3 &  &  &  & X \\
-OsiriX Lite &  &  &  & X \\
-GATE &  &  &  & X \\
-MicroView &  &  &  & X \\
-MatrixUser &  &  &  & X \\
-BioImage Suite Web &  &  & X &  \\
-ITK-SNAP &  &  &  & X \\
-Horos &  &  &  & X \\
-dwv &  &  &  & X \\
-Gwyddion &  &  &  & X \\ 
+3D Slicer & \checkmark & \checkmark & \checkmark & \checkmark \\
+ImageJ & \checkmark & \checkmark & \checkmark & \checkmark \\
+Fiji & \checkmark & \checkmark & \checkmark &  \\
+MRIcroGL &  &  &  & \checkmark \\
+Weasis &  &  & \checkmark & \checkmark \\
+ParaView &  & \checkmark &  &  \\
+OHIF Viewer &  &  & \checkmark & \checkmark \\
+DICOM Viewer &  &  & \checkmark & \checkmark \\
+medInria &  &  & \checkmark & \checkmark \\
+SMILI &  &  &  & \checkmark \\
+Drishti &  &  &  & \checkmark \\
+INVESALIUS 3 &  &  &  & \checkmark \\
+OsiriX Lite &  &  &  & \checkmark \\
+GATE &  &  &  & \checkmark \\
+MicroView &  &  &  & \checkmark \\
+MatrixUser &  &  &  & \checkmark \\
+BioImage Suite Web &  &  & \checkmark &  \\
+ITK-SNAP &  &  &  & \checkmark \\
+Horos &  &  &  & \checkmark \\
+dwv &  &  &  & \checkmark \\
+Gwyddion &  &  &  & \checkmark \\ 
 \bottomrule
 \end{tabular}
-\caption{Software with visibility/transparency related documents (software is
-listed in descending order of visibility/transparency score)}
+\caption{Software with visibility/transparency related documents (listed in
+descending order of visibility/transparency score)}
 \label{tab_Visibility/Transparency_docs}
@@
 order. Since we produced the scores from the AHP process, the total sum of the
-29 scores is precisely 1.
+29 scores is precisely 1.0.
 
 \begin{figure}[!ht]
-\includegraphics[scale=0.48]{figures/overall_scores.pdf}
+\includegraphics[scale=0.47]{figures/overall_scores.pdf}
 \caption{Overall AHP scores with an equal weighting for all 9 software qualities}
@@
 
-The top three software products \textit{3D Slicer}, \textit{ImageJ}, and
-\textit{OHIF Viewer} had higher scores in most criteria. \textit{3D Slicer}
-ranked in the top two software products for all qualities except \textit{surface
-robustness}; \textit{ImageJ} ranked in the top three for correctness \&
-verifiability, surface reliability, surface usability, maintainability, surface
-understandability, and visibility/transparency. \textit{OHIF Viewer} ranked in
-the top five products for correctness \& verifiability, surface reliability,
-surface usability, maintainability, and reusability. Given the installation
-problems, we may have underestimated the scores on reliability and robustness
-for \textit{DICOM Viewer}, but we compared it equally for the other seven
-qualities.
+The top four software products \textit{3D Slicer}, \textit{ImageJ},
+\textit{Fiji}, and \textit{OHIF Viewer} have higher scores in most criteria.
+\textit{3D Slicer} has a score in the top two for all qualities; \textit{ImageJ}
+ranks near the top for all qualities, except for correctness \& verifiability.
+\textit{OHIF Viewer} and \textit{Fiji} have similar overall scores, with
+\textit{Fiji} doing better in installability and \textit{OHIF Viewer} doing
+better in correctness \& verifiability.  Given the installation problems, we may
+have underestimated the scores on reliability and robustness for \textit{DICOM
+Viewer}, but we compared it equally for the other seven qualities.
 
@@
 stayed on GitHub, and calculate the average number of new stars, people
-watching, and forks per 12 months. The method of getting the creation date is
-described in Section~\ref{sec_grading_software}.  The items in
+watching, and forks per 12 months. Section~\ref{sec_grading_software} describes
+the method of getting the creation date.  The items in
 Table~\ref{tab_ranking_vs_GitHub} are listed in descending order of the average
 number of new stars per year.  The non-GitHub items are listed in the order of
-our ranking.  All GitHub statistics were collected in July, 2021.
+our ranking.  We collected all GitHub statistics in July 2021.
 
@@
 stars-per-year ranking, four of the top five software projects appear in both
-lists.Our top 5 packages are scattered among the first 8 positions on the GitHub
-list. However, as discussed below there are discrepancies between the two lists.
+lists. Our top five packages are scattered among the first eight positions on the
+GitHub list. However, as discussed below there are discrepancies between the two
+lists.
 
@@
 Section~\ref{sec_result_robustness}. We might underestimate the scores for
-\textit{dwv} due to uncommon technical issues. We also ranked \textit{DICOM
-Viewer} much lower than its popularity. As mentioned in
-Section~\ref{sec_result_installability} it depended on the NextCloud platform
-that we could not successfully install. Thus, we might underestimate the scores
-of its surface reliability and surface robustness. In addition, we weighted all
-qualities equally, which is not likely to be the case with all users. As a
-result, some projects with high community popularity may have scored lower with
-our method because of a relatively higher (compared to the scientific
-community's implicit ranking) weighting of the poor scores for some qualities. A
-further explanation for discrepancies between our measures and the star measures
-may also be due to inaccuracy with using stars to approximate popularity.  Stars
-are not an ideal measure because stars represent the community's feeling in the
-past more than they measure current preferences \citep{Szulik2017}.  The issue
-with stars is that they tend only to be added, not removed.  A final reason for
+\textit{dwv} due to uncommon technical issues.
+
+We also ranked \textit{DICOM Viewer} much lower than its popularity. As
+mentioned in Section~\ref{sec_result_installability}, it depended on the
+NextCloud platform that we could not successfully install. Thus, we might
+underestimate the scores of its surface reliability and surface robustness.
+
+Further reason for discrepancies between our ranking and the community's ranking
+is that we weighted all qualities equally. This is not likely how users
+implicitly rank the different qualities. As a result, some projects with high
+community popularity may have scored lower with our method because of a
+relatively higher (compared to the scientific community's implicit ranking)
+weighting of the poor scores for some qualities. A further explanation for
+discrepancies between our measures and the star measures may also be due to
+inaccuracy with using stars to approximate popularity.  Stars are not an ideal
+measure because stars represent the community's feeling in the past more than
+they measure current preferences \citep{Szulik2017}.  The issue with stars is
+that they tend only to be added, not removed.  A final reason for
 inconsistencies between our ranking and the community's ranking is that, as for
@@
 \toprule
-Software & Comm.\ rank & Our rank & Stars/yr & Watches/yr & Forks/yr \\ 
+Software & Comm.\ Rank & Our Rank & Stars/yr & Watches/yr & Forks/yr \\ 
 \midrule
 3D Slicer & 1 & 1 & 284 & 19 & 128 \\
-OHIF Viewer & 2 & 3 & 277 & 19 & 224 \\
-dwv & 3 & 23 & 124 & 12 & 51 \\
+OHIF Viewer & 2 & 4 & 277 & 19 & 224 \\
+dwv & 3 & 19 & 124 & 12 & 51 \\
 ImageJ & 4 & 2 & 84 & 9 & 30 \\
@@
 Papaya & 7 & 17 & 45 & 5 & 20 \\
-Fiji & 8 & 4 & 44 & 5 & 21 \\
+Fiji & 8 & 3 & 44 & 5 & 21 \\
 DICOM Viewer & 9 & 29 & 43 & 6 & 9 \\
-INVESALIUS 3 & 10 & 10 & 40 & 4 & 17 \\
-Weasis & 11 & 6 & 36 & 5 & 19 \\
+INVESALIUS 3 & 10 & 8 & 40 & 4 & 17 \\
+Weasis & 11 & 7 & 36 & 5 & 19 \\
 dicompyler & 12 & 26 & 35 & 5 & 14 \\
-OsiriX Lite & 13 & 9 & 34 & 9 & 24 \\
+OsiriX Lite & 13 & 11 & 34 & 9 & 24 \\
 MRIcroGL & 14 & 18 & 24 & 3 & 3 \\
-GATE & 15 & 25 & 19 & 6 & 26 \\
+GATE & 15 & 24 & 19 & 6 & 26 \\
 Ginkgo CADx & 16 & 14 & 19 & 4 & 6 \\
-BioImage Suite Web & 17 & 8 & 18 & 5 & 7 \\
+BioImage Suite Web & 17 & 6 & 18 & 5 & 7 \\
 Drishti & 18 & 27 & 16 & 4 & 4 \\
-Slice:Drop & 19 & 20 & 10 & 2 & 5 \\
-ITK-SNAP & 20 & 15 & 9 & 1 & 4 \\
-medInria & 21 & 7 & 7 & 3 & 6 \\
-SMILI & 22 & 13 & 3 & 1 & 2 \\
+Slice:Drop & 19 & 21 & 10 & 2 & 5 \\
+ITK-SNAP & 20 & 13 & 9 & 1 & 4 \\
+medInria & 21 & 9 & 7 & 3 & 6 \\
+SMILI & 22 & 10 & 3 & 1 & 2 \\
 MatrixUser & 23 & 28 & 2 & 0 & 0 \\
-MicroView & 24 & 16 & 1 & 1 & 1 \\
-Gwyddion & 25 & 11 & n/a & n/a & n/a \\
-XMedCon & 26 & 19 & n/a & n/a & n/a \\
-DicomBrowser & 27 & 21 & n/a & n/a & n/a \\
-AMIDE & 28 & 22 & n/a & n/a & n/a \\
-3DimViewer & 29 & 24 & n/a & n/a & n/a \\ 
+MicroView & 24 & 15 & 1 & 1 & 1 \\
+Gwyddion & 25 & 16 & n/a & n/a & n/a \\
+XMedCon & 26 & 20 & n/a & n/a & n/a \\
+DicomBrowser & 27 & 22 & n/a & n/a & n/a \\
+AMIDE & 28 & 23 & n/a & n/a & n/a \\
+3DimViewer & 29 & 25 & n/a & n/a & n/a \\ 
 \bottomrule
 \end{tabular}
-\caption{Software ranking by our methodology versus the community (Comm.)
+\caption{Software ranking by our methodology versus the community (Comm.)\
 ranking using GitHub metrics (Sorted in descending order of community
@@
 As shown in Section~\ref{sec_software_selection}, our domain experts recommended
-a list of top software with 12 software products.  All of the top 4 entries from
+a list of top software with 12 software products.  All the top 4 entries from
 the Domain Expert's list are among the top 12 ranked by our methodology. Three
@@
 and \textit{Fiji}. \textit{3D Slicer} is top project by both rankings (and by
-the GithHub stars measure as well).  The Domain Expert ranked \textit{Horos} as
+the GitHub stars measure as well).  The Domain Expert ranked \textit{Horos} as
 their second choice, while we ranked it twelfth.  Our third ranked project,
@@
 recommended by our methodology and the Domain Expert are not surprising.  As
-mentioned above, the methodology weights all qualities equally, but that may not
+mentioned above, our methodology weights all qualities equally, but that may not
 be the case for the Domain Expert's impressions.  Moreover, although the Domain
-Expert has significant experience with MI software, he has not used every one of
-the 29 packages that were measured.
+Expert has significant experience with MI software, they have not used all 29
+packages that were measured.
 
@@
 As part of filling in the measurement template (from
-Section~\ref{sec_grading_software}) we summarized the artifacts observed in each
-MI package. Table~\ref{artifactspresent} groups the artifacts by frequency into
-categories of common (20 to 29 ($>$67\%) packages), uncommon (10 to 19 (33-67\%)
-packages), and rare (1 to 9 ($<$33\%) packages). The full measurements are
-summarized in \citet{Dong2021-Data}.  The details on which projects use which
-types of artifacts are summarized in Tables~\ref{tab_maintainability_docs}
-and~\ref{tab_Visibility/Transparency_docs} for documents related to
-maintainability and visibility, respectively.
+Section~\ref{sec_grading_software}), we summarized the artifacts observed in
+each MI package. Table~\ref{artifactspresent} groups the artifacts by frequency
+into categories of common (20 to 29 ($>$67\%) packages), uncommon (10 to 19
+(33-67\%) packages), and rare (1 to 9 ($<$33\%) packages). \citet{Dong2021-Data}
+summarizes the full measurements.  Tables~\ref{tab_maintainability_docs}
+and~\ref{tab_Visibility/Transparency_docs} show the details on which projects
+use which types of artifacts for documents related to maintainability and
+visibility, respectively.
 
@@
     \toprule
-    \textbf{Common} & \textbf{Uncommon} & \textbf{Rare} \\
+    Common & Uncommon & Rare \\
     \midrule
@@
 between projects.  Although some content is reasonably consistent, with 97\% of
-README files contain at least one section describing the `What` of the
-repository and 89\% offering some `How` content, other categories are more
-variable.  For instance, information on `Contribution`, `Why`', and `Who`,
+README files contain at least one section describing the `What' of the
+repository and 89\% offering some `How' content, other categories are more
+variable.  For instance, information on `Contribution', `Why', and `Who',
 appear in 28\%, 26\% and 53\% of the analyzed files, respectively
@@
 popularity of recommending a given artifact, but it does not imply that the most
-commonly recommended artifacts are the most important artifacts. Just because an
-artifact is not explicitly recommended in a given guidelines, does not mean that
-the artifact is not valued by the guideline authors.  They may have excluded it
-because it is out of the scope of their recommendations, or outside of their
-experience.  For instance, an artifact related to uninstall is only explicitly
-mentioned by \citep{vanGompelEtAl2016}), but other guideline authors would
-likely see its value.  They may simply feel that uninstall is implied by
-install, or they may have never asked themselves whether separate uninstall
-instructions are needed.
+commonly recommended artifacts are the most important artifacts. Just because a
+guideline does not explicitly recommend an artifact, does not mean the guideline
+authors do not value that artifact.  They may have excluded it because it is out
+of the scope of their recommendations, or outside their experience.  For
+instance, an artifact related to uninstall is only explicitly mentioned by
+\citet{vanGompelEtAl2016}, but other guideline authors would likely see its
+value.  They may simply feel that uninstall is implied by install, or they may
+have never asked themselves whether they need separate uninstall instructions.
 
@@
 the software development guidelines shown in Table~\ref{Tbl_Guidelines}:
-Troubleshooting guide and Developer's manual.  Although these two artifacts
-aren't specifically named in the guidelines that we found, the information
-contained within them overlaps with the recommended artifacts.  A
-Troubleshooting guideline contains information that would typically be found in
-a User manual.  A Developer's guide overlaps with information from the README,
-INSTALL, Uninstall, Dependency List, Release Information, API documentation and
-Design documentation.  In our current analysis, we have identified artifacts by
-the names given by the software guidelines and MI examples.  In the future, a
-more in-depth analysis would look at the knowledge fragments that are captured
-in the artifacts, rather than focusing on the names of the files that collect
-these fragments together.
+Troubleshooting guide and Developer's manual.  Although the guidelines do not
+name these two artifacts, the information contained within them overlaps with
+the recommended artifacts.  A Troubleshooting guideline contains information
+that would typically be found in a User manual.  A Developer's guide overlaps
+with information from the README, INSTALL, Uninstall, Dependency List, Release
+Information, API documentation and Design documentation.  In our current
+analysis, we have identified artifacts by the names given by the software
+guidelines and MI examples.  In the future, a more in-depth analysis would look
+at the knowledge fragments that are captured in the artifacts, rather than
+focusing on the names of the files that collect these fragments together.
 
 Although the MI community shows examples of 88\% (23 of 26) of the practices we
-found in research software guidelines (Table~\ref{Tbl_Guidelines}), three
-recommended artifacts were not observed: i) Uninstall, ii) Test plans, and iii)
+found in research software guidelines (Table~\ref{Tbl_Guidelines}), we did not
+observe three recommended artifacts: i) Uninstall, ii) Test plans, and iii)
 Requirements.  Uninstall is likely an omission caused by the focus on installing
@@
 are not generally concerned with uninstall.  Moreover, as mentioned above,
-uninstall is not particularly emphasized in existing recommendations.  Test
-plans were not observed for MI software, but that doesn't mean they weren't
+uninstall is not particularly emphasized in existing recommendations.  We did
+not observe test plans for MI software, but that doesn't mean plans weren't
 created; it means that the plans are not under version control.  Test plans
-would have to at least be implicitly created, since test cases were observed
-with reasonable frequency for MI software (test cases are are categorized as
-uncommon).
+would have to at least be implicitly created, since we observed test cases with
+reasonable frequency for MI software (test cases are categorized as uncommon).
 
@@
 up-front requirements are impossible for research software
-\citep{CarverEtAl2007, SegalAndMorris2008}, but when the instance on
-``up-front'' requirements is dropped, allowing the requirements to be written
+\citep{CarverEtAl2007, SegalAndMorris2008}, but if we drop the insistence on
+``up-front'' requirements, allowing instead for the requirements to be written
 iteratively and incrementally, requirements are feasible \citep{Smith2016}.
@@
 practice.  A theme among these rare artifacts is that many of them are related
-to developers more than users.  For instance, the rare artifacts include
-Contributing, Developer code of conduct, Code style guide, Product roadmap,
-Requirements, Design documentation, API documentation and a Test plan.
-\wss{additional details are available in the LBM document for other missing
-parts, like code style guides.}  The rare artifacts for MI software are similar
-to the rare artifacts for Lattice Boltzmann solvers \citep{Michalski2021},
-except LBM software is more likely to have developer related artifacts, like
-Contributing, Dependency list, and Design documentation.
+to developers more than users.  For instance, the developer-focused, rare, and
+in some cases non-observed, artifacts include the following:
+
+\begin{itemize}
+
+\item \textbf{A Contributing file} provides new contributors with the information
+that they need to start adding/modifying the repository's files.
+\citet{Abdalla2016} provides a simple template for creating an open-source
+contributor guideline.
+
+\item \textbf{A Developer Code of Conduct} explicitly states the expectations
+for developers on how they should treat one another \citep{TouraniEtAl2017}. The
+code outlines rules for communication and establishes enforcement mechanisms for
+violations.  As \citet{TouraniEtAl2017} states, the code of conduct documents
+the spirit of a community so that anyone can comfortably contribute regardless
+of ethnicity, gender, or sexual orientation. Three popular code's of conduct are
+\citep{TouraniEtAl2017}:
+\href{https://www.contributor-covenant.org/version/2/1/code_of_conduct/}
+{Contributor Covenant}, \href{https://ubuntu.com/community/code-of-conduct}
+{Ubuntu Code of Conduct}, and \href{https://www.djangoproject.com/conduct/}
+{Django Code of Conduct}. A code of conduct can improve inclusivity, which
+brings the benefit of a wider pool of contributors.  For example, a code of
+conduct can improve the participation of women \citep{SinghEtAl2021}. A standard
+of ethical behaviour can be captured in the code, for projects that are looking
+to abide by a code of ethics, such as the IEEE Code of Ethics \citep{IEEE1999},
+or the Professional Engineers of Ontario code of ethics \citep[p.\
+23--24]{PEO2021}.
+
+\item \textbf{Code Style Guidelines} present standards for writing code. Style
+guides codify such elements as formatting, commenting, naming identifiers, best
+practices and dangers to avoid \citep{Carty2020}. For instance, most coding
+style guides will specify using ALLCAPS when naming symbolic constants.
+Understandability improves under standardization, since developers spend
+more time on the content of the code, and less time distracted by its style.
+Three sample style guides are:
+\href{https://google.github.io/styleguide/javaguide.html} {Google Java Style
+Guide},
+\href{http://cnl.sogang.ac.kr/cnlab/lectures/programming/python/PEP8_Style_Guide.pdf}
+{PEP8 Style Guide for Python}, and
+\href{https://google.github.io/styleguide/cppguide.html} {Google \CC Style
+Guide}.  Linting tools, like \href{https://pypi.org/project/flake8/}{flake8} for
+Python, can be used to enforced coding styles, like the PEP8 standard.
+
+\item \textbf{A Product Roadmap} explains the vision and direction for a product
+offering \citep{MunchEtAl2019}.  Although they have different forms, all
+roadmaps cover the following: \begin{inparaenum}[i)]
+	\item Where are we now?,
+	\item Where are we going?, and
+	\item How can we get there? \end{inparaenum} \citep{PhaalEtAl2005}. As
+stated by \citet{Pichler2012}, a product roadmap provides the following
+benefits: continuity of purpose, facilitation of collaboration, and assistance
+with prioritization. Creating a roadmap involves the following steps:
+\begin{inparaenum}[i)]
+	\item define and outline a strategic mission and product vision, 
+	\item scan the environment, 
+	\item revise and distill the product vision to write the product roadmap, and 
+	\item estimate the product life cycle and evaluate the mix of planned
+development efforts \end{inparaenum} \citep{VahaniittyEtAl2002}.
+
+\item \textbf{Requirements Documentation} records the functionalities, expected
+performance, goals, context, design constraints, external interfaces and other
+quality attributes of the software \citep{IEEE1998}.  Templates, which provide
+structure, guidelines and rules, are often used for writing requirements
+documents. Example templates include \citet{ESA1991}, \citet{IEEE1998},
+\citet{NASA1989}, and \citet{RobertsonAndRobertson1999Vol}. There is no
+universally accepted template. \citet{SmithEtAl2007} provides a template
+tailored to research software.
+
+\item \textbf{Design Documentation} explicitly states the design goals and
+priorities, records the likely changes, decomposes the complex system into
+separate modules, and specifies the relationship between the modules. Design
+documentation shows the syntax of the interface for the modules, and in some
+cases also documents the semantics.  Some potential elements of design
+documentation include the following:
+
+\begin{itemize}
+	\item Represent the design using Unified Modelling Language class diagrams.
+	This approach is suited to object-oriented design and designs that use
+	patterns \citep{Gamma1995}.
+	\item Rigorously document the design following the template for a Module
+	Guide (MG) \citep{ParnasEtAl1984}.  An MG organizes the modules in a
+	hierarchy by their secrets.
+	\item Explain the design using data flow diagrams to show typical use cases
+	for input transformation.
+	\item Traceability
+	\item Using a textual description, list for each module (or class), the
+	state variables (if any), exported constants and all exported access
+	programs.  This shows the interface that can used to access each module's
+	services.
+	\item Document the Module Interface Specification (MIS)
+	\citep{HoffmanAndStrooper1995}. An MIS is an abstract model that formally
+	shows each module's access programs and the associated transitions and
+	outputs based on their state, environment, and input variables
+	\citep{ElSheikhEtAl2004, SmithAndYu2009}.
+\end{itemize}
+
+\item \textbf{API Documentation} shows developers the services or data provided
+by the software application (or library) through such resources as its methods
+or objects \citep{MengEtAl2018}.  Understandability is improved by API
+documentation \citep{MengEtAl2018}. API documentation can be generated using
+tools like Doxygen, pydoc, and javadoc.
+
+\item \textbf{A test plan} Verification involves checking that the governing
+equations for the model, together with other definitions, including boundary
+and/or initial conditions, are solved correctly.  Validation, on the other hand,
+involves checking that the model is close enough to reality for whatever
+scientific or engineering problem is being addressed by the software.
+Verification is an exercise in mathematics, while validation is an exercise in
+engineering and science~\cite[p.~24]{Roache1998}.  Verification is not just
+important for the code.  As shown by the IEEE Standard for Software Verification
+and Validation Plans~\cite[p.\ 412]{VanVliet2000}, V\&V activities are
+recommended for each phase of the software development lifecycle.
+
+\end{itemize}
+
+The rare artifacts for MI software are similar to the rare artifacts for Lattice
+Boltzmann solvers \citep{Michalski2021}, except LBM software is more likely to
+have developer related artifacts, like Contributing, Dependency list, and Design
+documentation.
 
 To improve MI software in the future, an increased use of checklists could help.
-Checklists can be used in projects to ensure that best practices are followed by
-all developers.  Some examples include checklists merging branches into master
-\citep{Brown2015}, checklists for saving and sharing changes to the project
-\citep{WilsonEtAl2016}, checklists for new and departing team members
-\citep{HerouxAndBernholdt2018}, checklists for processes related to commits and
-releases \citep{HerouxEtAl2008} and checklists for overall software quality
-\citep{ThielEtAl2020, SSI2022}.  For instance, for Lattice Boltzmann solver
-software, ESPResSo has a checklist for managing releases \citep{Michalski2021}.
+Developers can use checklists to ensure they follow best practices.  Some
+examples include checklists merging branches into master \citep{Brown2015},
+checklists for saving and sharing changes to the project \citep{WilsonEtAl2016},
+checklists for new and departing team members \citep{HerouxAndBernholdt2018},
+checklists for processes related to commits and releases \citep{HerouxEtAl2008}
+and checklists for overall software quality \citep{ThielEtAl2020, SSI2022}.  For
+instance, for Lattice Boltzmann solver software, ESPResSo has a checklist for
+managing releases \citep{Michalski2021}.
 
@@
 
-Software tools are used to support the development, verification, maintenance,
-and evolution of software, software processes, and artifacts \citep[p.\
-501]{GhezziEtAl2003}. MI software uses tools for CI/CD, user support, version
-control, documentation, and project management.  To answer
+Developers use software tools to support the development, verification,
+maintenance, and evolution of software, software processes, and artifacts
+\citep[p.\ 501]{GhezziEtAl2003}. MI software uses tools for CI/CD, user support,
+version control, documentation, and project management.  To answer
 \rqref{RQ_CompareToolsProjMngmnt} we summarize the tool usage in these
@@
 \toprule
-\multicolumn{1}{c}{User support model} & Num.\ projects \\
+\multicolumn{1}{c}{User Support Model} & Num.\ Projects \\
 \midrule
@@
 From Section~\ref{sec_score_maintainability}, 27 of the 29 projects used git as
-the version control tool, 1 used Mercurial and 1 used Subversion.  The hosting
-is on GitHub for 24 packages, SourceForge for 3 and BitBucket for 2.  Although
-teams may have a process for accepting new contributions, no one discussed this
-during their interviews. However, most teams (8 of 9) mentioned using GitHub and
-pull requests to manage contributions from the community. The interviewees
-generally gave very positive feedback on using GitHub. Some teams previously
-used a different approach to version control and eventually transferred to git
-and GitHub.  The past approaches included contributions from e-mail (3 teams),
-contributions from forums (1 team) and e-mailing the git repository back and
-forth between developers (1 team).
+the version control tool, one used Mercurial and one used Subversion.  The
+hosting is on GitHub for 24 packages, SourceForge for three and BitBucket for
+two.  Although teams may have a process for accepting new contributions, no one
+discussed this during their interviews. However, most teams (eight of nine)
+mentioned using GitHub and pull requests to manage contributions from the
+community. The interviewees generally gave very positive feedback on using
+GitHub. Some teams previously used a different approach to version control and
+eventually transferred to git and GitHub.  The past approaches included
+contributions from e-mail (three teams), contributions from forums (one team)
+and e-mailing the git repository back and forth between developers (one team).
 
@@
 matches the increase in the broader research software community.  A little over
-10 years ago \citet{Nguyen-HoanEtAl2010} estimated that version control was used
-in only 50\% of research software projects, but even at that time
+10 years ago \citet{Nguyen-HoanEtAl2010} estimated that only 50\% of research
+software projects use version control, but even at that time
 \citet{Nguyen-HoanEtAl2010} noted an increase from previous usage levels. A
@@
 \end{inparaenum}
-Of the specifically named tools in the above lists, GitHub was mentioned 3
-times, while each of the other tools was only mentioned once.
+Of the specifically named tools in the above lists, interviewees mentioned
+GitHub 3 times, and each of the other tools once each.
 
@@
 software has much in common with tool utilization for ocean modelling software.
-Both use tools for editing, compiling, code management, testing, building and
+Both use tools for editing, compiling, code management, testing, building, and
 project management.  From the data available, ocean modelling differs from MI
@@
 
-\section{Comparison of Principles, Process and Methodologies to Research Software in General} \label{Sec_CompareMethodologies}
+\section{Comparison of Principles, Process, and Methodologies to Research Software in General} \label{Sec_CompareMethodologies}
 
@@
 \toprule
-Opinion on documentation & Num ans. \\ 
+Opinion on Documentation & Num.\ Ans. \\ 
 \midrule
@@
 start examples. \citet{SmithAndRoscoe2018} suggests including instructions
-specifically for on-boarding new developers. For open source software in general
+specifically for on-boarding new developers. For open-source software in general
 (not just research software), \citet{Fogel2005} recommends providing tutorial
@@
 \toprule
-Software team & Native application & Web application \\ 
+Software Team & Native Application & Web Application \\ 
 \midrule
-3D Slicer & X & \\
-INVESALIUS 3 & X & \\
-dwv & & X \\
-BioImage Suite Web & & X \\
-ITK-SNAP & X & \\
-MRIcroGL & X & \\
-Weasis & X & \\
-OHIF & & X \\ 
+3D Slicer & \checkmark & \\
+INVESALIUS 3 & \checkmark & \\
+dwv & & \checkmark \\
+BioImage Suite Web & & \checkmark \\
+ITK-SNAP & \checkmark & \\
+MRIcroGL & \checkmark & \\
+Weasis & \checkmark & \\
+OHIF & & \checkmark \\ 
 \midrule
@@
 think that the packages will work out of the box to solve their cases, while in
-reality CFD knowledge needs to be applied to correctly modify the packages for a
-new endeavour.
+reality computational fluid dynamics knowledge needs to be applied to correctly
+modify the packages for a new endeavour.
 
@@
 	\item Elimination of headaches associated with a separate integration phase
-	\citep{Fowler2006}, \citep[p.\ 20]{HumbleAndFarley2010}. If integration is
-	postponed, integration problems are inevitable.  Continuous integration
-	means that problems are immediately obvious and the source of the problem
-	can be isolated to the small increment that was just committed.
+	\citep{Fowler2006}, \citep[p.\ 20]{HumbleAndFarley2010}. If developers
+	postpone integration, integration problems are inevitable.  Continuous
+	integration means that problems are immediately obvious and the source of
+	the problem can be isolated to the small increment that was just committed.
 	\item Detection and removal of bugs \citep{Fowler2006} via
@@
 inconsistencies, etc. \citep{Whitehouse2018}. By insisting on code
-standarization linters can reduce technical debt and thus improve
+standardization linters can reduce technical debt and thus improve
 maintainability (\qref{Q_Maintainability}). Although linters are tools for code
@@
 for medical devices~\cite{Wassyng2015}.  The first sub-goal (GR) argues for the
-quality of the documentation of the requirements (SRS).  The second sub-goal
-(GD) says that the design complies with the requirements and the third proposes
-that the implementation also complies with the requirements.  The fourth
-sub-goal (GA) claims that the inputs to 3dfim+ will satisfy the operational
-assumptions, since we need valid input to make an argument for the correctness
-of the output.
-
-\begin{figure}[!h]
+quality of the documentation of the requirements.  The second sub-goal (GD) says
+that the design complies with the requirements and the third proposes that the
+implementation also complies with the requirements.  The fourth sub-goal (GA)
+claims that the inputs to 3dfim+ will satisfy the operational assumptions, since
+we need valid input to make an argument for the correctness of the output.
+
+\begin{figure}[!ht]
 \centering
@@
 install or build \textit{dwv}, \textit{GATE}, and \textit{DICOM Viewer}. We used
-a deployed online version for \textit{dwv}, a VM version for \textit{GATE}, but
+a deployed on-line version for \textit{dwv}, a VM version for \textit{GATE}, but
 no alternative for \textit{DICOM Viewer}. We might underestimate their rank due
 to this uncommon technical issue.
-\item We only use two pieces of data to measure surface robustness. This is
+\item Measuring software robustness only involved two pieces of data. This is
 likely part of the reason for limited variation in the robustness scores
@@
 
+\wss{Go through the RQs?}
+
 We analyzed the state of the practice for the MI domain with the goal of
@@
 our research questions. In addition, Section~\ref{ch_recommendations} presents
-our recommendations on SC software development.
+our recommendations on research software development.
 
@@
 current status of MI software development. We ranked the 29 software projects in
-nine qualities.  Based on on the grading scores \textit{3D Slicer},
+nine qualities.  Based on the grading scores \textit{3D Slicer},
 \textit{ImageJ}, and \textit{OHIF Viewer} are the top three software packages.
@@
 \begin{itemize}
-\item the lack of fundings and time;
+\item the lack of funding and time;
 \item the difficulty to balance between four factors: cross-platform
@@
 including developing a web-based approach with backend servers and maintaining
-better documentation. We also collected the status of documentation.  We found
-that for all 8 interviewed teams that documentation is felt to be vital to a
-project, with the most popular form of documentation being forum discussions and
-videos.  With respect to project management almost all teams used GitHub and
-pull requests to manage contributions.  Very few teams used a specific
-development model.  It appears that the development process is more ad hoc than
-planned for the majority of projects.
-
-Our answers to the research questions are based on the above findings. We
+better documentation. We also collected the status of documentation.  In our
+interviews, we found that for all 8 interviewed teams felt documentation is
+vital to a project, with the most popular form of documentation being forum
+discussions and videos.  With respect to project management almost all teams
+used GitHub and pull requests to manage contributions.  Very few teams used a
+specific development model.  It appears that the development process is more ad
+hoc than planned for the majority of projects.
+
+The above findings are the basis for our answers to the research questions. We
 identified the existing artifacts, tools, principles, processes, and
@@
 requirements specifications, theory manuals, getting started tutorials, user
-manuals, project plan, developer’s manual, API documentation, requirements on
+manuals, project plan, developer's manual, API documentation, requirements on
 coding standards, development process, project status, development environment,
@@
 \citet{ParnasEtAl2000};
-\item considering newer technologies (e.g.\ web application and serverless
+\item considering newer technologies (e.g.,\ web application and serverless
 solution);
@@
     \begin{itemize}
-        \item surface reliability: our current measurement relies on
+        \item Surface reliability: our current measurement relies on
         the processes of installation and getting started tutorials. However,
@@
         with each software, and record any errors.
-        \item surface robustness: we used damaged images as inputs for
+        \item Surface robustness: we used damaged images as inputs for
         this measuring MI software. This process is similar to fuzz testing
@@
         identify tools and libraries to automate this process.
-        \item surface usability: we can design usability tests and test
+        \item Surface usability: we can design usability tests and test
         all software projects with end-users. The end-users can be volunteers
         and domain experts.
-        \item surface understandability: our current method does not require
+        \item Surface understandability: our current method does not require
         understanding the source code. As software engineers, perhaps we can
@@
         \citet{SmithEtAl2021}.
-        \item measure modifiability as part of the measurement of
+        \item Measure modifiability as part of the measurement of
         maintainability.  An experiment could be conducted asking participants
@@
     \end{itemize}
-	\item we can further automate the measurements on the grading template. For
+	\item We can further automate the measurements on the grading template. For
 	example, with automation scripts and the GitHub API, we may save significant
@@
 	collect metrics from the GitHub API, and record the results.
-	\item the rubric for the grading standard can be made more explicit.
-	\item we can improve some interview questions. Some examples are:
+	\item The rubric for the grading standard can be made more explicit.
+	\item We can improve some interview questions. Some examples are:
 	\begin{itemize}
@@
 	    developers and usability to end-users, since there was confusion during
-	    the interviews as to which group was being discussed.
+	    the interviews on the meaning of the terminology.
 	\end{itemize}
-	\item we can better organize the interview questions. Since we use audio
+	\item We can better organize the interview questions. Since we use audio
 	conversion tools to transcribe the answers, we should make the transcription
@@
 	the five software qualities and compose a similar structure for each.
-	\item we can mark the follow-up interview questions with keywords. For
+	\item We can mark the follow-up interview questions with keywords. For
 	example, say ``this is a follow-up question'' every time asking one. Thus, we
@@
 %%%%%%%%%%%
-
-% Overview of scientific computing
-We define Scientific Computing (SC) as ``the use of computer tools to analyze or
-simulate mathematical models of real world systems of engineering or scientific
-importance so that we can better understand and predict the system's behaviour''
-\citep{Smith2006}. Many researchers consider SC as the third pillar of science
-and engineering, along with theory and experiment \citep{Landau2005}. Almost all
-areas in science and engineering use computers for modelling \citep{Golub2014},
-and software plays an essential role in modern scientific research
-\citep{Hannay2009, Wilson2014}.  Software development in SC depends on three
-fields of knowledge: engineering or scientific domain knowledge, mathematical
-algorithm knowledge, and computational algorithm knowledge \citep{Landau2005,
-Mehta2015}. Thus, most SC software developers are scientists in SC domains
-\citep{Wilson2014}. However, they do not always use modern software development
-techniques, tools, and methods \citep{Wilson2014}. Therefore, we developed a
-methodology for assessing the state of the practice for SC software
-\citep{SmithEtAl2021, SmithAndMichalski2022}. We apply this process to Medical
-Imaging (MI) software that belongs to a specific domain of SC.
-
-TO DO
-
-- CarverEtAl2022 has discussion of requirements etc. that we could also contrast (seems that people overestimate their use of requirements.)
-- add more on comparison to existing literature surveys
-- ask Mike:
-    - for journal version ask for more figures?
-    - for journal double blind solution
-    - can he submit
-    - potential reviewers
-    - feedback
-- Radiology - should not exceed 6000 words, 50 refs, 8 figures and 4 tables
-
-- add discussion of sensitivity analysis (borrow from LBM)

### 2022-10-31 StateOfPractice/Papers/MedImageSoft_SOP-arXiv/MedImageSoft_SOP.tex
Edits to the Comparision to Artifacts section

@@
 code outlines rules for communication and establishes enforcement mechanisms for
-violations.  As \citet{TouraniEtAl2017} states, the code of conduct documents
+violations.  As \citet{TouraniEtAl2017} states, the developer code documents
 the spirit of a community so that anyone can comfortably contribute regardless
-of ethnicity, gender, or sexual orientation. Three popular code's of conduct are
+of ethnicity, gender, or sexual orientation. Three popular codes of conduct are
 \citep{TouraniEtAl2017}:
@@
 \begin{itemize}
-	\item Represent the design using Unified Modelling Language class diagrams.
-	This approach is suited to object-oriented design and designs that use
-	patterns \citep{Gamma1995}.
-	\item Rigorously document the design following the template for a Module
-	Guide (MG) \citep{ParnasEtAl1984}.  An MG organizes the modules in a
-	hierarchy by their secrets.
-	\item Explain the design using data flow diagrams to show typical use cases
+	\item Representation of the system design and class design using Unified
+	Modelling Language class diagrams. This approach is suited to
+	object-oriented design and designs that use patterns \citep{Gamma1995}.
+	\item Rigorous documentation of the system design following the template for
+	a Module Guide (MG) \citep{ParnasEtAl1984}.  An MG organizes the modules in
+	a hierarchy by their secrets.
+	\item An explanation of the design using data flow diagrams to show typical use cases
 	for input transformation.
-	\item Traceability
-	\item Using a textual description, list for each module (or class), the
-	state variables (if any), exported constants and all exported access
-	programs.  This shows the interface that can used to access each module's
-	services.
-	\item Document the Module Interface Specification (MIS)
-	\citep{HoffmanAndStrooper1995}. An MIS is an abstract model that formally
-	shows each module's access programs and the associated transitions and
-	outputs based on their state, environment, and input variables
-	\citep{ElSheikhEtAl2004, SmithAndYu2009}.
+	\item A table or graph showing the traceability between the requirements and
+	the modules (or classes)
+	\item The syntax of the modules (or classes) by providing lists of the state
+	variables, exported constants and all exported access programs for each
+	module (or class).  This shows the interface that can used to access each
+	module's services.
+	\item A formal specification of the semantics of input/output relationships
+	and state transitions for each module using a Module Interface Specification
+	(MIS) \citep{HoffmanAndStrooper1995}. An MIS is an abstract model that
+	formally shows each module's access programs and the associated transitions
+	and outputs based on their state, environment, and input variables.
+	\citet{ElSheikhEtAl2004, SmithAndYu2009} show the example of an MIS for a
+	mesh generator.
 \end{itemize}
@@
 
-\item \textbf{A test plan} Verification involves checking that the governing
-equations for the model, together with other definitions, including boundary
-and/or initial conditions, are solved correctly.  Validation, on the other hand,
-involves checking that the model is close enough to reality for whatever
-scientific or engineering problem is being addressed by the software.
-Verification is an exercise in mathematics, while validation is an exercise in
-engineering and science~\cite[p.~24]{Roache1998}.  Verification is not just
-important for the code.  As shown by the IEEE Standard for Software Verification
-and Validation Plans~\cite[p.\ 412]{VanVliet2000}, V\&V activities are
-recommended for each phase of the software development lifecycle.
+\item \textbf{A Test Plan} describes the scope, approach, resources, and the
+schedule of planned test activities \citep{VanVliet2000}.  The plan should cover
+details such as the makeup of the testing team, automated testing tools and
+technology to employ, the testing process, system tests, integration tests and
+unit tests.  A test plan can also include the specific plans for peer code (and
+documentation) review, as discussed further in Section~\ref{Sec_PeerReview}.
 
@@
 not alone in this.  Many, if not most, research projects fall short of best
-practices.  A gap exists in scientific computing development practices and
-software engineering recommendations \citep{Storer2017, Kelly2007,
-OwojaiyeEtAl2021_CSE}. \citet{JohansonAndHasselbring2018} observe that the
-state-of-the-practice for research software in industry and academia does not incorporate
-state-of-the-art SE tools and methods.  This causes sustainability and
-reliability problems \citep{FaulkEtAl2009}. Rather than benefit from capturing
-and reusing previous knowledge, projects waste time and energy ``reinventing the
-wheel'' \citep{deSouzaEtAl2019}.
+practices.  A gap exists in research software development practices and software
+engineering recommendations \citep{Storer2017, Kelly2007, OwojaiyeEtAl2021_CSE}.
+\citet{JohansonAndHasselbring2018} observe that the state-of-the-practice for
+research software in industry and academia does not incorporate state-of-the-art
+SE tools and methods.  This causes sustainability and reliability problems
+\citep{FaulkEtAl2009}. Rather than benefit from capturing and reusing previous
+knowledge, projects waste time and energy ``reinventing the wheel''
+\citep{deSouzaEtAl2019}.
 

### 2022-10-31 StateOfPractice/Papers/MedImageSoft_SOP-arXiv/MedImageSoft_SOP.tex
Updates to comparison of tools section

@@
 
-We present the state of the practice for Medical Imaging software. We selected
-29 medical imaging projects from 48 candidates, assessed 10 software qualities
-(installability, correctness/ verifiability, reliability, robustness, usability,
-maintainability, reusability, understandability, visibility/transparency and
-reproducibility) by answering 108 questions for each software project, and
-interviewed 8 of the 29 development teams. Based on the quantitative data for
-the first 9 qualities, we ranked the MI software with the Analytic Hierarchy
-Process (AHP). The four top ranked software products are: \textit{3D Slicer},
-\textit{ImageJ}, \textit{Fiji}, and \textit{OHIF Viewer}. Our ranking is mostly
-consistent with the community's ranking, with four of our top five projects also
-appearing in the top five of a list ordered by stars-per-year. \textit{DICOM
-Viewer} and \textit{dmv} are exceptions; we ranked them lower than their
-estimated popularity because of difficulty with their installation.  Comparison
-for tool usage. Comparison for process. From interviewing the developers, we
-identified five pain points and two qualities of potential concern:: i) lack of
-resources; ii) difficulty balancing between compatibility, maintainability,
-performance, and security; and, iii) lack of access to real-world datasets for
-testing. For future MI software projects, we propose adopting test-driven
-development, using continuous integration and continuous delivery (CI/CD), using
-git and GitHub, maintaining good documentation, supporting third-party plugins
-or extensions, considering web application solutions, and improving
-collaboration between different MI software projects. \wss{Update after the
-paper has been revised.}
+We present the state of the practice for Medical Imaging (MI) software. We
+selected 29 medical imaging projects from 48 candidates, assessed 10 software
+qualities (installability, correctness/ verifiability, reliability, robustness,
+usability, maintainability, reusability, understandability,
+visibility/transparency and reproducibility) by answering 108 questions for each
+software project, and interviewed 8 of the 29 development teams. Based on the
+quantitative data for the first 9 qualities, we ranked the MI software with the
+Analytic Hierarchy Process (AHP). The four top ranked software products are:
+\textit{3D Slicer}, \textit{ImageJ}, \textit{Fiji}, and \textit{OHIF Viewer}.
+Our ranking is mostly consistent with the community's ranking, with four of our
+top five projects also appearing in the top five of a list ordered by
+stars-per-year. \textit{DICOM Viewer} and \textit{dmv} are exceptions; we ranked
+them lower than their estimated popularity because of difficulty with their
+installation.  Generally, MI software is in a healthy state.  The documentation
+artifacts created, and development tools employed, typically match the
+recommendations given in research software development guidelines. A few places
+in the current state of the practice that deviate from the recommendations
+include some underemphasized artifacts (like a developer's code of conduct, code
+style guidelines, product roadmaps, and Application Program Interface (API)
+documentation) and ... Exceptions include ... tools With respect to software
+artifacts, MI software generally Comparison for tool usage. Comparison for
+process. From interviewing the developers, we identified five pain points and
+two qualities of potential concern:: i) lack of resources; ii) difficulty
+balancing between compatibility, maintainability, performance, and security;
+and, iii) lack of access to real-world datasets for testing. For future MI
+software projects, we propose adopting test-driven development, using continuous
+integration and continuous delivery (CI/CD), using git and GitHub, maintaining
+good documentation, supporting third-party plugins or extensions, considering
+web application solutions, and improving collaboration between different MI
+software projects.
+\wss{Update after the paper has been revised.}
 
@@
 a four-year period ending in 2018. (For completeness the same study showed a
-decrease in version control usage for seismology software over the same time
-period, from 41\% down to 36\%).  A recent survey by \citet{CarverEtAl2022}
+small decrease in version control usage for seismology software over the same
+time period, from 41\% down to 36\%).  A recent survey by \citet{CarverEtAl2022}
 shows version control use among practitioners at over 95\%, with 83/87 survey
@@
 identified five projects using CI/CD tools (about 17\% of the assessed
-projects). We found CI/CD using projects by examining the documentation and
+projects). We found which projects used CI/CD by examining the documentation and
 source code of all projects. The count of CI/CD usage may actually be higher,
@@
 on CI/CD utilization for research software, our impression is that CI/CD is not
-yet common practice, despite its recommendation.  For LBM software at least the
-situation is similar to MI software, with only 12.5\% of 24 LBM packages showing
-evidence of CI/CD in their repositories \citep{Michalski2021}.  The survey of
-\citet{CarverEtAl2022} suggests higher use of CI/CD with 54\% (54/100)
-respondents indicating that they use it.
+yet common practice, despite its recommendation.  This is certainly the case for
+LBM software, where usage numbers are similar to MI software, with only 12.5\%
+of 24 LBM packages showing evidence of CI/CD in their repositories
+\citep{Michalski2021}.  The survey of \citet{CarverEtAl2022} suggests higher use
+of CI/CD with 54\% (54/100) of respondents indicating that they use it. However,
+that survey measures something different from the current one by surveying
+practitioners, rather than assessing projects.
 
@@
 were GitHub, wiki pages, workshops, and social media. The least frequently
-mentioned options (about 10\% of developers) included writing books, google
-forms and state management.  In contrasting MI software with LBM software, the
-most significant documentation tool difference is that LBM software often uses
-document generation tools, like doxygen and sphinx \citep{Michalski2021}, while
-MI does not appear to use these tools.
+mentioned options (about 10\% of developers) included writing books, and google
+forms.  In contrasting MI software with LBM software, the most significant
+documentation tool difference is that LBM software often uses document
+generation tools, like doxygen and sphinx \citep{Michalski2021}, while MI does
+not appear to use these tools.
 

### 2022-10-31 StateOfPractice/Papers/MedImageSoft_SOP-arXiv/MedImageSoft_SOP.tex
Revisions to Pain points section

@@
 and, iii) lack of access to real-world datasets for testing. For future MI
-software projects, we propose adopting test-driven development, using continuous
-integration and continuous delivery (CI/CD), using git and GitHub, maintaining
+software projects, we propose adopting test-driven development, using Continuous
+Integration and Continuous Delivery (CI/CD), using git and GitHub, maintaining
 good documentation, supporting third-party plugins or extensions, considering
@@
 amethododical process \citep{Kelly2013}, or a knowledge acquisition driven
-process \citep{Kelly2015}.  A waterfall like process can work for research
+process \citep{Kelly2015}.  A waterfall-like process can work for research
 software \citep{Smith2016}, especially if the developers work iteratively and
@@
 Although the interviewees did not explicitly identify these two qualities as
-pain points, they were discussed as part of our interview process
-\citep{SmithEtAl2021}.  The interviewee's practices for addressing pain points
-and improving quality can potentially be emulated by other MI developers.
-Moreover, these practices may provide examples that can be followed by other
-research software domains.
+pain points, we did discuss threats to these qualities and ways to improve them
+as part of our interview process \citep{SmithEtAl2021}.  The interviewee's
+practices for addressing pain points and improving quality can potentially be
+emulated by other MI developers. Moreover, these practices may provide examples
+that can be followed by other research software domains.
 
@@
 \item Supporting third-party plugins and extensions; and,
-\item Using GitHub Actions for continuous integration and continuous
-development.
+\item Using GitHub Actions for CI/CD (Continuous Integration and Continuous
+Delivery.)
 \end{itemize}
@@
 and \textit{OHIF} said getting funding for software maintenance is more
-challenging than finding funding for research. The interviewee from the
-\textit{ITK-SNAP} team thought more funding was a way to solve the lack of time
+challenging than finding funding for research. The interviewee from 
+\textit{ITK-SNAP} thought more funding was a way to solve the lack of time
 problem, because they could hire more dedicated developers. On the other hand,
-the interviewee from the \textit{Weasis} team did not feel that funding could
+the interviewee from \textit{Weasis} did not feel that funding could
 solve the same problem, since they would still need time to supervise the project.
@@
 The pain point survey of \citet{WieseEtAl2019} highlights that technology
-hurdles are an issue for research software in general.  Some of the
-technical-related problems mentioned by \citet{WieseEtAl2019} include dependency
-management, cross-platform compatibility (also mentioned by
-\citet{PintoEtAl2018}), CI, hardware issues and operating system issues. From
-\citep{SmithEtAl2022} technology pain points for LBM developers include setting
-up parallelization and CI.
+hurdles are an issue for research software in general.  Some technical-related
+problems mentioned by \citet{WieseEtAl2019} include dependency management,
+cross-platform compatibility (also mentioned by \citet{PintoEtAl2018}), CI,
+hardware issues and operating system issues. From \citep{SmithEtAl2022}
+technology pain points for LBM developers include setting up parallelization and
+CI.
 
@@
 \item Improving performance via more powerful computers, which one interviewee
-pointed out has already happened to reduce the balance problem.
+pointed out has already happened.
 \end{itemize}
@@
 backend, but in this case there are disadvantages for privacy protection and
-server costs.
+server costs.  These issues are discussed further in the recommendations
+(Section~\ref{sec_webapps}).
 
@@
 large amounts of data \citep{SmithEtAl2022}. Although not specifically mentioned
-during our interviews, the large data sets for MI likely cause a challenge for
-using free testing services, like GitHub Actions.
+during our interviews, the large data sets for MI likely also cause a challenge
+for using free testing services, like GitHub Actions.
 
@@
 allude to this challenge, likely because for a give image (test case) it is
-possible to determine, possibly using other software, the expected analysis
-results.
+possible to determine, potentially by using other software, the expected
+analysis results.
 
@@
 clear visual expressions, designing the software to be intuitive, having a UX
-designer, dialog windows for important notifications, providing an example for
-users to follow, downsampling images to consume less memory, and an option to
-load only part of the data to boost performance.  The last two points recognize
-that an important component of usability is performance, since poor performance
-frustrates users.
+(User eXperience) designer, dialog windows for important notifications,
+providing an example for users to follow, downsampling images to consume less
+memory, and providing an option to load only part of the data to boost
+performance.  The last two points recognize that an important component of
+usability is performance, since poor performance frustrates users.
 
@@
 \item[Q\refstepcounter{qnum}\theqnum \label{Q_Maintainability}:]
-\textbf{Maintainability:} Although not explicitly highlighted as a pain point
-during our interviews, \citet{Nguyen-HoanEtAl2010} rate maintainability as the
+\textbf{Maintainability:} \citet{Nguyen-HoanEtAl2010} rate maintainability as the
 third most important software quality for research software in general. The push
@@
 (\ppref{P_TechnologyHurdles}), usability \ppref{P_Usability}, and
-Maintainability.  Documentation is also necessary for reproducibility. The
+maintainability.  Documentation is also necessary for reproducibility. The
 challenges of inadequate documentation are a known problem for research software

### 2022-11-02 StateOfPractice/Papers/MedImageSoft_SOP-arXiv/MedImageSoft_SOP.tex
Final draft for submission, next step to anonymize

@@
 top five projects also appearing in the top five of a list ordered by
-stars-per-year. \textit{DICOM Viewer} and \textit{dmv} are exceptions; we ranked
-them lower than their estimated popularity because of difficulty with their
-installation.  Generally, MI software is in a healthy state.  The documentation
-artifacts created, and development tools employed, typically match the
-recommendations given in research software development guidelines. A few places
-in the current state of the practice that deviate from the recommendations
-include some underemphasized artifacts (like a developer's code of conduct, code
-style guidelines, product roadmaps, and Application Program Interface (API)
-documentation) and ... Exceptions include ... tools With respect to software
-artifacts, MI software generally Comparison for tool usage. Comparison for
-process. From interviewing the developers, we identified five pain points and
-two qualities of potential concern:: i) lack of resources; ii) difficulty
-balancing between compatibility, maintainability, performance, and security;
-and, iii) lack of access to real-world datasets for testing. For future MI
-software projects, we propose adopting test-driven development, using Continuous
-Integration and Continuous Delivery (CI/CD), using git and GitHub, maintaining
-good documentation, supporting third-party plugins or extensions, considering
-web application solutions, and improving collaboration between different MI
-software projects.
-\wss{Update after the paper has been revised.}
+stars-per-year. Generally, MI software is in a healthy state as shown by the
+following: in the repositories we observed 88\% of the documentation artifacts
+recommended by research software development guidelines, 100\% of MI projects
+use version control tools, and developers appear to use the common quasi-agile
+research software development process. However, the current state of the
+practice deviates from the existing guidelines because of the rarity of some
+recommended artifacts (like test plans, requirements specification, code of
+conduct, code style guidelines, product roadmaps, and Application Program
+Interface (API) documentation), low usage of continuous integration (17\% of the
+projects), low use of unit testing (about 50\% of projects), and room for
+improvement with documentation (six of nine developers felt their documentation
+wasn't clear enough). From interviewing the developers, we identified five pain
+points and two qualities of potential concern: lack of development time, lack of
+funding, technology hurdles, ensuring correctness, usability, maintainability,
+and reproducibility. The interviewees proposed strategies to improve the state
+of the practice, to address the identified pain points, and to improve software
+quality. Combining their ideas with ours, we have the following list of
+recommendations: increase documentation, increase testing by enriching datasets,
+increase continuous integration usage, move to web applications, employ linters,
+use peer reviews, design for change, add assurance cases, and incorporate a
+``Generate All Things'' approach.
 
@@
 \toprule
-~ \ & \cite{USGS2019} & \cite{TobiasEtAl2018} & \cite{BrettEtAl2021} &
-\cite{WilsonEtAl2016} & \cite{SmithAndRoscoe2018} & \cite{HerouxEtAl2008} &
-\cite{ThielEtAl2020} & \cite{vanGompelEtAl2016} & \cite{OrvizEtAl2017} & MI\\
+~ \ & \citet{USGS2019} & \citet{TobiasEtAl2018} & \citet{BrettEtAl2021} &
+\citet{WilsonEtAl2016} & \citet{SmithAndRoscoe2018} & \citet{HerouxEtAl2008} &
+\citet{ThielEtAl2020} & \citet{vanGompelEtAl2016} & \citet{OrvizEtAl2017} & MI\\
 \midrule
@@
 are not generally concerned with uninstall.  Moreover, as mentioned above,
-uninstall is not particularly emphasized in existing recommendations.  We did
-not observe test plans for MI software, but that doesn't mean plans weren't
-created; it means that the plans are not under version control.  Test plans
-would have to at least be implicitly created, since we observed test cases with
-reasonable frequency for MI software (test cases are categorized as uncommon).
+uninstall is not particularly emphasized in existing recommendations.
+
+Test plans describe the scope, approach, resources, and the schedule of planned
+test activities \citep{VanVliet2000}.  The plan should cover details such as the
+makeup of the testing team, automated testing tools and technology to employ,
+the testing process, system tests, integration tests and unit tests. We did not
+observe test plans for MI software, but that doesn't mean plans weren't created;
+it means that the plans are not under version control. Test plans would have to
+at least be implicitly created, since we observed test cases with reasonable
+frequency for MI software (test cases are categorized as uncommon).
+
+The other apparently neglected document is the requirements specification, which
+records the functionalities, expected performance, goals, context, design
+constraints, external interfaces and other quality attributes of the software
+\citep{IEEE1998}.  If developers write requirements they typically are based on
+a template, which provide documentation structure, guidelines, and rules.
+Although there is no universally accepted template, examples include
+\citet{ESA1991}, \citet{IEEE1998}, \citet{NASA1989}, and
+\citet{RobertsonAndRobertson1999Vol}.
 
@@
 iteratively and incrementally, requirements are feasible \citep{Smith2016}.
-
-Table~\ref{Tbl_Guidelines} shows several artifacts that are rarely observed in
-practice.  A theme among these rare artifacts is that many of them are related
-to developers more than users.  For instance, the developer-focused, rare, and
-in some cases non-observed, artifacts include the following:
+\citet{SmithEtAl2007} provides a requirements template tailored to research
+software.
+
+Table~\ref{Tbl_Guidelines} shows several recommended artifacts that are rarely
+observed in practice.  A theme among these rare artifacts is that, except for
+the user-focused getting started manual, they are developer-focused.  The
+dependency list, which is a list of software library dependencies, was rarely
+observed, but this information is still likely present, just embedded in build
+scripts. The other developer-focused and rare artifacts are as follows:
 
@@
 development efforts \end{inparaenum} \citep{VahaniittyEtAl2002}.
-
-\item \textbf{Requirements Documentation} records the functionalities, expected
-performance, goals, context, design constraints, external interfaces and other
-quality attributes of the software \citep{IEEE1998}.  Templates, which provide
-structure, guidelines and rules, are often used for writing requirements
-documents. Example templates include \citet{ESA1991}, \citet{IEEE1998},
-\citet{NASA1989}, and \citet{RobertsonAndRobertson1999Vol}. There is no
-universally accepted template. \citet{SmithEtAl2007} provides a template
-tailored to research software.
 
@@
 
-\item \textbf{A Test Plan} describes the scope, approach, resources, and the
-schedule of planned test activities \citep{VanVliet2000}.  The plan should cover
-details such as the makeup of the testing team, automated testing tools and
-technology to employ, the testing process, system tests, integration tests and
-unit tests.  A test plan can also include the specific plans for peer code (and
-documentation) review, as discussed further in Section~\ref{Sec_PeerReview}.
-
 \end{itemize}
@@
 that survey measures something different from the current one by surveying
-practitioners, rather than assessing projects.
+practitioners, rather than assessing projects.  Additional information on CI/CD
+is given in the recommendations (Section~\ref{Sec_ContinuousIntegration}).
 
@@
 	stable base will always pass all tests.  If the CI system uses generators
-	and linters, it will also have current documentation and coding standard
-	compliant code.  A stable base improves developer productivity, allowing
-	them to focus on coding, testing, and documentation.
+	and linters, it will also have current documentation and standard compliant
+	code.  A stable base improves developer productivity, allowing them to focus
+	on coding, testing, and documentation.
 \end{itemize}
@@
 	control, including configuration files, build scripts, test harnesses, and
-	operating system configurations \citep[p. 19]{HumbleAndFarley2010}.
+	operating system configuration files \citep[p. 19]{HumbleAndFarley2010}.
 	Fortunately for the MI, as shown in Section~\ref{sec_score_maintainability}
@@
 	programming practice or lack of adherence to adopted coding standards.
-	\item An integration build system to pull everything together.  Every time
+	\item An integrated build system to pull everything together.  Every time
 	there is a check-in (for instance a pull request), the integration server
@@
 
+Continuous integration and delivery helps with addressing several pain points.
+For instance, CI/CD helps reduce development time (\ppref{P_LackDevTime}) by
+removing the need for a time-consuming integration stage and by automating
+regression testing.  Automated regression tests also help with ensuring
+correctness (\ppref{P_Correctness}) and the quality of reproducibility
+(\qref{Q_Reproducibility}).
+
 \subsection{Move To Web Applications} \label{sec_webapps}
@@
 \href{https://vuejs.org/} {Vue}, \href{https://angular.io/} {Angular} and
-\href{https://reactjs.org/} {React} and \href{https://elm-lang.org/}{Elm}.
+\href{https://reactjs.org/} {React}, and \href{https://elm-lang.org/}{Elm}.
 
@@
 Linters can be used as an ad hoc check for code files, but they really come into
-their own when used as part of a continuous integration system, as discussed in
+their own when used as part of a CI system, as discussed in
 Section~\ref{Sec_ContinuousIntegration}. Almost none of the research software
@@
 code inspection is generally higher than most forms of testing, which average
-between 30 and 35\% for defect removal efficiency \citep{EbertAndJones2009,
+between 30 --- 35\% for defect removal efficiency \citep{EbertAndJones2009,
 Jones2008}. For research software, \citep{KellyAndShepard2000} show a task based
@@
 for critical parts of the code, developers may want to adopt a more rigorous
-approach. For instance, developers may drop the modern requirement for
+approach. For instance, developers may drop the modern trend of
 asynchronous review and instead occasionally use synchronous review to help
@@
 and \qref{Q_Maintainability}. Peer review can potentially find misunderstandings
-in of how the code implements the required theory, which will improve the
+in how the code implements the required theory, which will improve the
 software's correctness (\ppref{P_Correctness}). The benefits of peer review for
@@
 into modules is less prevalent.  Not every decomposition is a good design for
-supporting change, as shown by \citet{Parnas1972a}.  A design with low cohesion
-and high coupling \citep[p.\ 48]{GhezziEtAl2003} will make change difficult.
-Especially in research software, where change is inevitable, designers need to
-produce a modularization that supports change. \citep{JungEtAl2022} points out
-that ocean modelling software is currently feeling the pain of not emphasizing
-modularization in legacy code.
+supporting change, as shown by \citet{Parnas1972a}.  For instance, a design with
+low cohesion and high coupling \citep[p.\ 48]{GhezziEtAl2003} will make change
+difficult. Especially in research software, where change is inevitable,
+designers need to produce a modularization that supports change.
+\citet{JungEtAl2022} points out that ocean modelling software is currently
+feeling the pain of not emphasizing modularization in legacy code.
 
@@
 change by first identifying likely changes, either implicitly or explicitly, and
-second by hiding each likely change behind a well-defined module interface.
-This approach mirrors the recommendations from \citet{Parnas1972a}.
-\citet{SmithEtAl2022} (under Section ``Documentation as Part of the Development
-Process'') lists ideas for how to document the design, including the likely
-changes, so that they are more visible to others.
+second by hiding each likely change behind a well-defined module interface. This
+approach mirrors the recommendations from \citet{Parnas1972a}.
+Section~\ref{Sec_CompareArtifacts} lists ideas for how to document the design,
+including the likely changes, so that they are more visible to others.
 
@@
 One way to present an assurance case is through the Goal Structuring Notation
-(GSN) \cite{Spriggs2012}, which make arguments clear, easy to read and, hence,
+(GSN) \citep{Spriggs2012}, which make arguments clear, easy to read and, hence,
 easy to challenge. GSN starts with a Top Goal (Claim), like ``Program X delivers
@@
 as shown in Figure~\ref{TopGoal}.  This example follows the same pattern as used
-for medical devices~\cite{Wassyng2015}.  The first sub-goal (GR) argues for the
+for medical devices \citep{Wassyng2015}.  The first sub-goal (GR) argues for the
 quality of the documentation of the requirements.  The second sub-goal (GD) says
@@
 To address developer pain points, we propose automatically generating MI code
-and its documentation via a Domain Specific Modelling (DSM) approach. A DSM
+and its documentation via a ``Generate All Things'' (GAT) approach. A GAT
 approach uses models to capture knowledge in domains such as physics, computing,
@@
 recipe can even potentially be written to generate an assurance case
-(Section~\ref{AssuranceCases}). Our definition here of DSM is more general than
-usual; in this recommendation DSM implies generation of all software artifacts,
-not just the code. DSM moves development to a higher level of abstraction so
-that domain experts can work without concern for low-level implementation
-details. DSM allows developers to optimally generate code and documentation,
-reduce the likelihood of errors, eliminate redundancy, and automate maintenance.
-With a DSM approach MI developers can experiment with different algorithm
-choices, different input formats, different outputs formats, etc.
-
-DSMs are not typically used for research software.  In the future, some believe
-that they will transform coding, documentation, design, and verification
-\citep{JohansonAndHasselbring2018, Smith2018}. DSMs remove the distraction of
-writing software, allowing developers to focus on their science.  A DSM approach
+(Section~\ref{AssuranceCases}). Our definition of GAT implies generation of all
+software artifacts, not just the code. GAT moves development to a higher level
+of abstraction so that domain experts can work without concern for low-level
+implementation details. GAT allows developers to optimally generate code and
+documentation, reduce the likelihood of errors, eliminate redundancy, and
+automate maintenance. With a GAT approach MI developers can experiment with
+different algorithm choices, different input formats, different outputs formats,
+etc.
+
+Part of GAT is code generation.  In the future, some believe that code
+generation will transform coding, documentation, design, and verification
+\citep{JohansonAndHasselbring2018, Smith2018}. GAT removes the distraction of
+writing software, allowing developers to focus on their science. A GAT approach
 removes the maintenance headaches of documentation duplicates and near
@@
 Computational Software). \citet{MatkerimEtAl2013} and \citet{OberEtAl2018} use
-DSMs for High Performance Computing (HPC), using UML (Unified Modelling
-Language) for their domain models. Unlike previous DSM work, the current
-recommendation focuses on generating all software artifacts (requirements,
-design, etc.), not just code. \citet{SzymczakEtAl2016} presents initial work on
-this approach, \citet{SmithAndCarette2021-BRIC} presents a motivating example,
+code generation for High Performance Computing (HPC), using UML (Unified Modelling
+Language) for their domain models. \citet{SzymczakEtAl2016} presents initial work on
+the GAT approach, \citet{SmithAndCarette2021-BRIC} presents a motivating example,
 and \citep{CaretteEtAl2021-Drasil} provides a prototype.
 
-A ``generate all things'' approach addresses multiple pain points.  For example,
-a generative approach can decrease development time (\ppref{P_LackDevTime}) by
-automation, once the necessary infrastructure is in place.  The DSM approach
-addresses the technology related pain point (\ppref{P_TechnologyHurdles})
-because technology information can be captured in the models and transformed as
-needed. To ensure correctness (\ppref{P_Correctness}), a DSM approach should be
-correct by construction.  If there are mistakes, DSM has the advantage that they
-are propagated throughout the generated artifacts, which greatly increases the
+A GAT approach addresses multiple pain points.  For example, a generative
+approach can decrease development time (\ppref{P_LackDevTime}) by automation,
+once the necessary infrastructure is in place.  The GAT approach addresses the
+technology related pain point (\ppref{P_TechnologyHurdles}) because technology
+information can be captured in the models and transformed as needed. To ensure
+correctness (\ppref{P_Correctness}), a GAT approach should be correct by
+construction.  If there are mistakes, GAT has the advantage that they are
+propagated throughout the generated artifacts, which greatly increases the
 chance that someone will notice the mistake.  Maintainability
@@
 Usability (\ppref{P_Usability}) is addressed because of the emphasis on
-generating up-to-date documentation.  Reproducibility.
+generating up-to-date documentation.  GAT facilitates reproducibility
+(\qref{Q_Reproducibility}) because at any time all the code and documentation
+can be regenerated.  The generator can include explicit traceability to show the
+dependence of the software on specific versions of software libraries.
 
@@
 
-\citet{RunesonAndHost2009} defines construct validity as when the adopted
-metrics represent what they are intended to measure. Our construct threats are
+\citet{RunesonAndHost2009} defines construct validity as the adopted
+metrics representing what they are intended to measure. Our construct threats are
 often related to how we assume our measurements influences the various software
@@
 no alternative for \textit{DICOM Viewer}. We might underestimate their rank due
-to this uncommon technical issue.
+to these technical issues.
 \item Measuring software robustness only involved two pieces of data. This is
@@
 
-\section{Conclusions} \label{ch_conclusions}
-
-\wss{Go through the RQs?}
-
-We analyzed the state of the practice for the MI domain with the goal of
-understanding current practice, answering our six research questions
-(Section~\ref{sec_motivation}) and providing recommendations for current and
-future projects.  Our methods in Section~\ref{ch_methods} form a general process
-to evaluate domain-specific software, that we apply to the specific domain of MI
-software. We identified 48 MI software candidates, then, with the help of the
-Domain Expert selected 29 of them to our final list. Section~\ref{ch_results}
-lists our measurements to nine software qualities for the 29 projects, and
-Section~\ref{painpoints} contains our interviews with eight of the 29 teams,
-discussing their development process and five software qualities.  We answered
-our research questions. In addition, Section~\ref{ch_recommendations} presents
-our recommendations on research software development.
-
-Compare to LBM.
-
-\subsection{Key Findings}
-
-With the measurement results in Section~\ref{ch_results}, we summarized the
-current status of MI software development. We ranked the 29 software projects in
-nine qualities.  Based on the grading scores \textit{3D Slicer},
-\textit{ImageJ}, and \textit{OHIF Viewer} are the top three software packages.
-
-The interview results in Section~\ref{painpoints} show some merits, drawbacks,
-and pain points within the development process. The three primary categories of
-pain points are:
+\section{Future Work} \label{Sec_FutureWork}
+
+The following recommendations for future state of the practice measurement
+exercises, for MI or for other domains, could address some threats to validity
+mentioned above.  Moreover, some ideas may make the data collection more
+efficient.
+
 \begin{itemize}
-\item the lack of funding and time;
-\item the difficulty to balance between four factors: cross-platform
-compatibility, convenience to development \& maintenance, performance, and
-security;
-\item the lack of access to real-world datasets for testing.
-\end{itemize}
-We summarized the solutions from the developers to address these problems,
-including developing a web-based approach with backend servers and maintaining
-better documentation. We also collected the status of documentation.  In our
-interviews, we found that for all 8 interviewed teams felt documentation is
-vital to a project, with the most popular form of documentation being forum
-discussions and videos.  With respect to project management almost all teams
-used GitHub and pull requests to manage contributions.  Very few teams used a
-specific development model.  It appears that the development process is more ad
-hoc than planned for the majority of projects.
-
-The above findings are the basis for our answers to the research questions. We
-identified the existing artifacts, tools, principles, processes, and
-methodologies in the 29 projects. By comparisons with the implied popularity of
-existing projects we found: 1) four of the top five software projects in our
-ranking were also among the top five ones receiving the most GitHub stars per
-year (Table~\ref{tab_ranking_vs_GitHub}); 2) three of the top four in our
-ranking were among the top four provided by the domain experts.
-
-Section~\ref{ch_recommendations} presents our recommendations on improving
-software qualities and easing pain points during development. Some highlighted
-recommendations are as follows:
+    \item We would like to make surface measurements less shallow. For example:
     \begin{itemize}
-\item adopting test-driven development with unit tests, integration tests, and
-nightly tests;
-\item maintaining good documentation (e.g., installation instructions,
-requirements specifications, theory manuals, getting started tutorials, user
-manuals, project plan, developer's manual, API documentation, requirements on
-coding standards, development process, project status, development environment,
-and release notes);
-\item using CI/CD;
-\item using git and GitHub;
-\item modular approach with the design principle proposed by
-\citet{ParnasEtAl2000};
-\item considering newer technologies (e.g.,\ web application and serverless
-solution);
-\item various ways of enriching the testing datasets, such as using existing
-open data sources and establishing greater community collaboration in the MI
-domain (Section~\ref{sec_recommendations_testing_dataset}).
-\end{itemize}
-
-\subsection{Future Works}
-
-With learnings from this project, we summarized recommendations for the future
-state of the practice assessments:
-\begin{itemize}
-    \item we can make the surface measurements less shallow. For example:
-    \begin{itemize}
-        \item Surface reliability: our current measurement relies on
-        the processes of installation and getting started tutorials. However,
-        not all software needs installation or has a getting started tutorial.
-        We can design a list of operation steps, perform the same operations
-        with each software, and record any errors.
-        \item Surface robustness: we used damaged images as inputs for
-        this measuring MI software. This process is similar to fuzz testing
+        \item Surface reliability: our current measurement relies on the
+        processes of installation and getting started tutorials. However, not
+        all software needs installation or has a getting started tutorial. We
+        could devise a list of operation steps (with the help of the Domain
+        Expert), perform the same operations with each software, and record any
+        errors.
+        \item Surface robustness: we used damaged images as inputs for this
+        measuring MI software. This process is similar to fuzz testing
         \citep{enwiki:1039424308}, which is one type of fault injection
-        \citep{enwiki:1039005082}. We may adopt more fault injection methods, and
-        identify tools and libraries to automate this process.
-        \item Surface usability: we can design usability tests and test
-        all software projects with end-users. The end-users can be volunteers
-        and domain experts.
+        \citep{enwiki:1039005082}. We may adopt more fault injection methods,
+        and identify tools and libraries to automate this process.
+        \item Surface usability: we can design usability tests and test all
+        software projects with end-users. The end-users can be volunteers and
+        domain experts. Ideas for getting started are available in
+        \citet{SmithEtAl2021}.
         \item Surface understandability: our current method does not require
@@
         documentation, try to understand the logic, and score the ease of the
-        process.  Ideas for getting started are available in
-        \citet{SmithEtAl2021}.
-        \item Measure modifiability as part of the measurement of
-        maintainability.  An experiment could be conducted asking participants
-        to make modifications, observing the study subjects during the
-        modifications, testing the resulting software and surveying the
+        process.  
+        \item Maintainability: we can add a measure modifiability as part of the
+        measurement of maintainability.  An experiment could be conducted asking
+        participants to make modifications, observing the study subjects during
+        the modifications, testing the resulting software and surveying the
         participants \citep{SmithEtAl2021}.
@@
 	collect metrics from the GitHub API, and record the results.
-	\item The rubric for the grading standard can be made more explicit.
 	\item We can improve some interview questions. Some examples are:
 	\begin{itemize}
-	    \item in {Q14}, ``Do you think improving this process can tackle the
-	    current problem?'' is a yes-or-no question, which is not informative
-	    enough. As mentioned in Section~\ref{Sec_CompareMethodologies}, most
-	    interviewees ignored it. We can change it to ``By improving this
-	    process, what current problems can be tackled?''; 
-	    \item in {Q16}, we can ask for more details about the modular
-	    approach, such as ``What principles did you use to divide code into
-	    modules? Can you describe an example of using your principles?'';
-	    \item {Q17} and {Q18} should respectively ask understandability to
-	    developers and usability to end-users, since there was confusion during
-	    the interviews on the meaning of the terminology.
+	    \item In one question we ask, ``Do you think improving this process can
+	    tackle the current problem?''  The problem is that this is a yes-or-no
+	    question, which is not informative. We could change the question to ``By
+	    improving this process, what current problems can be tackled?''; 
+	    \item We can ask for more details about the modular approach, such as
+	    ``What principles did you use to divide code into modules? Can you
+	    describe an example of using your principles?''.
 	\end{itemize}
@@
 	\item We can mark the follow-up interview questions with keywords. For
-	example, say ``this is a follow-up question'' every time asking one. Thus, we
-	record this sentence in the transcription, and it will be much easier to
+	example, say ``this is a follow-up question'' every time asking one. Thus,
+	we record this sentence in the transcription, and it will be much easier to
 	distinguish the follow-up questions from the 20 designed questions.
 \end{itemize}
+
+\section{Conclusions} \label{ch_conclusions}
+
+We analyzed the state of the practice for the MI domain with the goal of
+understanding current practice, answering our ten research questions
+(Section~\ref{sec_motivation}) and providing recommendations for current and
+future projects.  Our methods in Section~\ref{ch_methods} form a general process
+to evaluate domain-specific software, that we apply to the specific domain of MI
+software. We identified 48 MI software candidates, then, with the help of the
+Domain Expert selected 29 of them to our final list.
+
+Section~\ref{ch_results} lists our measurement results for ranking the 29
+projects for nine software qualities. Our ranking results appear credible since
+they are mostly consistent with the ranking from the scientific community
+implied by the GitHub stars-per-year metric. As discussed in
+Section~\ref{Sec_VsCommunityRanking}, four of the top five software projects
+appear in both our list and in the GitHub popularity list.  Moreover, our top
+five packages appear among the first eight positions on the GitHub list.  The
+noteworthy discrepancies between the two lists are for the packages that we were
+unable to install (\textit{dwv} and \textit{Dicom Viewer}).
+
+Based on our grading scores \textit{3D Slicer}, \textit{ImageJ}, \textit{Fiji}
+and \textit{OHIF Viewer} are the top four software performers.  However, the
+separation between the top performers and the others is not extreme.  Almost all
+packages do well on at least a few qualities, as shown in
+Table~\ref{topperformerstable}, which summarizes the packages ranked first and
+second for each quality. Almost 70\% (20 of 29) of the software packages appear
+in the top two for at least two qualities.  The only packages that do not appear
+in Table~\ref{topperformerstable}, or only appear once, are \textit{Papaya},
+\textit{MatrixUser}, \textit{MRIcroGL}, \textit{XMedCon}, \textit{dicompyler},
+\textit{DicomBrowser}, \textit{AMIDE}, \textit{3DimViewer}, and
+\textit{Drishti}. The shortness of this list suggests parity with respect to
+adoption of best practices for MI software overall.
+
+% software packages not in the top two for anything
+% , , ,  , , , ,
+% ,   , , , ,
+% Papaya,  ,  , , 
+% MatrixUser, , , , , MRIcroGL, , , , , , ,
+
+% Installability: 3D Slicer, BioImage Suite Web, Slice:Drop, INVESALIUS,
+% Correctness (first appearance): OHIF Viewer, ImageJ, medInria,
+% Reliability (first appearance): SMILI, Fiji, 
+% Maintainability (first appearance): Weasis, ParaView
+% Reusability (first appearance): dwv, GATE
+
+% Robustness: XMedCon, OsiriX Lite, MicroView, ITK-SNAP, Horos, Gwyddion,
+% dicompyler, DicomBrowser, AMIDE, 3DimViewer, DICOM Viewer
+
+% Usability: Ginkgo CADx
+
+% Understandability: Drishti
+
+\begin{table}[ht!]
+	\begin{center}
+	\renewcommand{\arraystretch}{1.8}
+	\begin{tabular}{ p{3cm}p{13cm} }
+		\toprule
+
+		Quality & Ranked 1st or 2nd\\
+
+		\midrule
+
+		Installability & 3D Slicer, BioImage Suite Web, Slice:Drop, INVESALIUS\\
+
+		\pbox{3.0cm}{Correctness \\ and Verifiability} & OHIF Viewer, 3D
+		Slicer, ImageJ\\
+
+		Reliability & SMILI, ImageJ, Fiji, 3D Slicer, Slice:Drop, OHIF
+		Viewer\\
+
+		Robustness & XMedCon, Weasis, SMILI, ParaView, OsiriX Lite,
+		MicroView, medInria, ITK-SNAP, INVESALIUS, ImageJ, Horos, Gwyddion,
+		Fiji, dicompyler, DicomBrowser, BioImage Suite Web, AMIDE, 3DimViewer,
+		3D Slicer, OHIF Viewer, DICOM Viewer\\
+
+		Usability & 3D Slicer, ImageJ, Fiji, OHIF Viewer, ParaView,
+		INVESALIUS, Ginkgo CADx, SMILI, OsiriX Lite, BioImage Suite Web,
+		ITK-SNAP, medInria, MicroView, Gwyddion\\
+
+		Maintainability & 3D Slicer, Weasis, ImageJ, OHIF Viewer, ParaView\\
+
+		Reusability & 3D Slicer, ImageJ, Fiji, OHIF Viewer, SMILI, dwv, BioImage
+		Suite Web, GATE, ParaView\\
+
+		\pbox{3.0cm}{Understandability} & 3D Slicer, ImageJ, Weasis,
+		Fiji, Horos, OsiriX Lite, dwv, Drishti, OHIF Viewer, GATE, ITK-SNAP,
+		ParaView, INVESALIUS\\
+
+		\pbox{3.0cm}{Visibility and \\Transparency} & ImageJ, 3D Slicer, Fiji\\
+
+		Overall Quality & 3D Slicer, ImageJ\\
+
+		\bottomrule		
+	\end{tabular}
+	\caption{Top performers for each quality (sorted by order of quality
+	measurement)} \label{topperformerstable}
+	\end{center}
+\end{table}
+
+For insight into devising future methods and tools, we interviewed nine
+developers (from eight teams) to learn about their pain points
+(Section~\ref{painpoints}).  We also discussed qualities of potential concern.
+The identified pain points and qualities of concern include:
+
+\begin{description}
+\item [\ppref{P_LackDevTime}] Lack of development time, %P1
+\item [\ppref{P_LackFunding}] Lack of funding, %P2
+\item [\ppref{P_TechnologyHurdles}] Technology hurdles, %P3
+\item [\ppref{P_Correctness}] Ensuring correctness, %P4
+\item [\ppref{P_Usability}] Usability, %P5
+\item [\qref{Q_Maintainability}] Quality of maintainability, and %Q1
+\item [\qref{Q_Reproducibility}] Quality of reproducibility. %Q2
+\end{description}
+
+Despite the pain points, overall MI software is in a healthy state for software
+development practices.  In our survey of the selected projects we observed 88\%
+of the documentation artifacts recommended by research software development
+guidelines (Section~\ref{Sec_CompareArtifacts}).  With respect to tools, MI is
+keeping pace with other research software with 100\% of the projects using
+version control (with 93\% specifically using git)
+(Section~\ref{Sec_CompareTools}).  We observed that the MI developers tend to
+follow the typical research software trend of using a quasi-agile software
+development process (Section~\ref{Sec_CompareMethodologies}).
+
+Although the state of the practice for MI software is healthy, we did notice
+areas where practice seems to lag behind the research software development
+guidelines.  For instance, the guidelines recommend three artifacts that were
+not observed: uninstall instructions, test plans, and requirements
+documentation. We observed the following recommended artifacts, but only rarely:
+contributing file, developer code of conduct, code style guidelines, product
+roadmap, design documentation, and API documentation
+(Section~\ref{Sec_CompareArtifacts}). Although software development tool use
+seems healthy, we found the use of CI/CD behind typical usage rates (17\% of the
+projects used CI/CD) (Section~\ref{Sec_CompareTools}).  With respect to the
+development process, developer identified areas for improvement included testing
+(only 50\% of projects were identified to have unit testing) and documentation
+(only three out of nine developers felt their documentation was clear enough)
+(Section~\ref{Sec_CompareMethodologies}).
+
+Our interviewees proposed strategies to improve the state of the practice, to
+address the identified pain points, and to improve software quality.  To their
+list (Section~\ref{painpoints}) we added some of our own recommended strategies
+(Section~\ref{ch_recommendations}).  Below we summarize the proposed strategies,
+with traceability to where we discuss the strategy, and to the relevant pain
+points.
+
+\begin{enumerate}
+
+\item Increase documentation to address \ppref{P_LackDevTime},
+\ppref{P_TechnologyHurdles}, \ppref{P_Usability}, \qref{Q_Maintainability},
+\qref{Q_Reproducibility} (Section~\ref{painpoints})
+
+\item Increase testing by enriching datasets to address \ppref{P_Correctness},
+\qref{Q_Reproducibility}
+(Section~\ref{painpoints},~\ref{sec_recommendations_testing_dataset})
+
+\item Increase modularity to address \qref{Q_Maintainability} (Section~\ref{painpoints})
+
+\item Use continuous integration to address \ppref{P_LackDevTime},
+\ppref{P_Correctness}, \qref{Q_Reproducibility}
+(Section~\ref{painpoints},~\ref{Sec_ContinuousIntegration})
+
+\item Move to web applications to address \ppref{P_TechnologyHurdles}
+(Section~\ref{painpoints},~\ref{sec_webapps})
+
+\item Employ linters to address \ppref{P_LackDevTime},
+\ppref{P_TechnologyHurdles}, \qref{Q_Maintainability}
+(Section~\ref{Sec_Linters})
+
+\item Peer reviews to address \ppref{P_LackDevTime},
+\ppref{P_TechnologyHurdles}, \ppref{P_Correctness}, \qref{Q_Maintainability}
+(Section~\ref{Sec_PeerReview})
+
+\item Design for change to address \qref{Q_Maintainability} (Section~\ref{Sec_DesForChange})
+
+\item Assurance case to address \ppref{P_Correctness}, \qref{Q_Reproducibility}
+(Section~\ref{AssuranceCases})
+
+\item Generate all things to address \ppref{P_LackDevTime},
+\ppref{P_TechnologyHurdles}, \ppref{P_Correctness}, \ppref{P_Usability},
+\qref{Q_Maintainability} and \qref{Q_Reproducibility}.
+(Section~\ref{Sec_GenAllThings})
+
+\end{enumerate}
 
@@
 Balaci for advice on web applications.
+
+\section*{Conflict of Interest}
+
+On behalf of all authors, the corresponding author states that there is no
+conflict of interest.
 

### 2022-11-02 StateOfPractice/Papers/MedImageSoft_SOP/MedImageSoft_SOP.tex
Anonymized version of the SOP for MI paper

--- 2022-11-02 StateOfPractice/Papers/MedImageSoft_SOP-arXiv/MedImageSoft_SOP.tex
@@
 %\documentclass[3p, 12pt,authoryear]{elsarticle}
-\documentclass[final, 3p, times, authoryear]{elsarticle}
+\documentclass[final, 3p, times, authoryear, doubleblind]{elsarticle}
 
@@
 
-\journal{Academic Radiology}
+\journal{Archives of Computational Methods in Engineering}
 
@@
 	\item \textbf{Reliability} The probability of failure-free operation of a
-	computer program in a specified environment for a specified time \citep{musa1987software}, \citep[p.\ 357]{GhezziEtAl2003}.
+	computer program in a specified environment for a specified time
+	\citep{musa1987software}, \citep[p.\ 357]{GhezziEtAl2003}.
 
@@
 	specified users to achieve specified goals with effectiveness, efficiency,
-	and satisfaction in a specified context of use'' \citep{ISO/TR16982:2002}
-	\citep{ISO9241-11:2018}.
+	and satisfaction in a specified context of use'' \citep{ISO/TR16982:2002,
+	ISO9241-11:2018}.
 
@@
 	other attributes; iii) satisfy new requirements
-	\citep{IEEEStdGlossarySET1990}, \citep{boehm2007software}.
+	\citep{IEEEStdGlossarySET1990, boehm2007software}.
 
@@
 on their time.  For the current assessment, our Domain Expert (and paper
-co-author) is Dr.\ Michael Noseworthy, Professor of Electrical and Computer
-Engineering at McMaster University, Co-Director of the McMaster School of
-Biomedical Engineering, and Director of Medical Imaging Physics and Engineering
-at St.\ Joseph's Healthcare, Hamilton, Ontario, Canada.
+co-author) is Anonymous % DOUBLE BLIND
+% Dr.\ Michael Noseworthy, Professor of Electrical and Computer
+% Engineering at McMaster University, Co-Director of the McMaster School of
+% Biomedical Engineering, and Director of Medical Imaging Physics and Engineering
+% at St.\ Joseph's Healthcare, Hamilton, Ontario, Canada.
 
@@
 question list; we ask follow-up questions when necessary.  The interview process
-presented here was approved by the McMaster University Research Ethics Board
-under the application number 
-\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/MACREM/Application.pdf}
-{MREB\#: 5219}.
+presented here was approved by the Anonymous Ethics Board. %% DOUBLE BLIND
+% McMaster University Research Ethics Board
+% under the application number 
+% \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/MACREM/Application.pdf}
+% {MREB\#: 5219}.
 
@@
 included all three documents. Moreover, \textit{3D Slicer} has a much higher
-percentage of closed issues (91.65\%) compared to \textit{ImageJ} (52.49\%) and
-\textit{Fiji} (63.79\%). Table~\ref{tab_maintainability_docs} shows which
+percentage of closed issues (92\%) compared to \textit{ImageJ} (52\%) and
+\textit{Fiji} (64\%). Table~\ref{tab_maintainability_docs} shows which
 projects had these documents, in descending order of their maintainability
@@
 \bibliographystyle{ACM-Reference-Format}
-\bibliography{MedImageSoft_SOP}
+%\bibliography{MedImageSoft_SOP}
+\bibliography{MedImageSoft_SOP_Anonymous}
 

### 2023-07-29 StateOfPractice/Papers/LatticeBoltz_SOP/LatticeBoltz_SOP.tex
Corrected postal code for Mechanical Engineering to match JHE building postal code

--- 2022-10-31 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
@@
             city={Hamilton},
-            postcode={L8S 4K1}, 
+            postcode={L8S 4L7}, 
             state={Ontario},

### 2023-07-29 StateOfPractice/Papers/LatticeBoltz_SOP/LatticeBoltz_SOP.tex
Removed confusing comma from abstract

@@
     ranking, suggesting at least a partial correlation between the use of best
-    practices and popularity. We find the state of the practice to be healthy,
+    practices and popularity. We find the state of the practice to be healthy
     with 67\% of the measured packages ranking in the top two for at least two

### 2023-07-31 StateOfPractice/Papers/LatticeBoltz_SOP/LatticeBoltz_SOP.tex
Fixed additional typos for published version

@@
 lists the eliminated software packages. Of the remaining 23 packages, we kept
-some despite their `dead` status because of their presence on authoritative LBM
+some despite their `dead' status because of their presence on authoritative LBM
 software lists and because of their surface excellence. After the initial
@@
 between projects.  Although some content is reasonably consistent, with 97\% of
-README files containing at least one section describing the `What` of the
-repository and 88.5 \% offering some `How` content, other categories are more
-variable.  For instance, information on `Contribution`, `Why`, and `Who`,
+README files containing at least one section describing the `What' of the
+repository and 88.5 \% offering some `How' content, other categories are more
+variable.  For instance, information on `Contribution', `Why', and `Who',
 appear in 27.8\%, 25.7\% and 52.9\% of the analyzed files, respectively

### 2024-05-27 StateOfPractice/Papers/PainPoints_OS_MI/PainPoints_OS_MI.tex
Split versions of SOP papers

--- 2022-11-02 StateOfPractice/Papers/MedImageSoft_SOP-arXiv/MedImageSoft_SOP.tex
@@
-%% 
-%% Copyright 2007-2020 Elsevier Ltd
-%% 
-%% This file is part of the 'Elsarticle Bundle'.
-%% ---------------------------------------------
-%% 
-%% It may be distributed under the conditions of the LaTeX Project Public
-%% License, either version 1.2 of this license or (at your option) any
-%% later version.  The latest version of this license is in
-%%    http://www.latex-project.org/lppl.txt
-%% and version 1.2 or later is part of all distributions of LaTeX
-%% version 1999/12/01 or later.
-%% 
-%% The list of all files belonging to the 'Elsarticle Bundle' is
-%% given in the file `manifest.txt'.
-%% 
-%% Template article for Elsevier's document class `elsarticle'
-%% with Harvard style bibliographic references
-
-%\documentclass[preprint,12pt,authoryear]{elsarticle}
-%\documentclass[3p, 12pt,authoryear]{elsarticle}
-\documentclass[final, 3p, times, authoryear]{elsarticle}
-
-%% Use the option review to obtain double line spacing
-%% \documentclass[authoryear,preprint,review,12pt]{elsarticle}
-
-%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
-%% for a journal layout:
-%% \documentclass[final,1p,times,authoryear]{elsarticle}
-%% \documentclass[final,1p,times,twocolumn,authoryear]{elsarticle}
-%% \documentclass[final,3p,times,authoryear]{elsarticle}
-%% \documentclass[final,3p,times,twocolumn,authoryear]{elsarticle}
-%% \documentclass[final,5p,times,authoryear]{elsarticle}
-%% \documentclass[final,5p,times,twocolumn,authoryear]{elsarticle}
-
-%% For including figures, graphicx.sty has been loaded in
-%% elsarticle.cls. If you prefer to use the old commands
-%% please give \usepackage{epsfig}
-
-%% The amssymb package provides various useful mathematical symbols
+\documentclass[11pt]{article}
+
+\usepackage{authblk}
+\usepackage{fancyhdr}
+\usepackage{float}
+
+\usepackage[round, sort]{natbib}
+
 \usepackage{amssymb}
 \usepackage{amsmath,amsthm}
-
-%% The lineno packages adds line numbers. Start line numbering with
-%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
-%% for the whole article with \linenumbers.
-%% \usepackage{lineno}
 
@@
 
-%\usepackage{tikz}
-%\usetikzlibrary{mindmap}
 \usepackage{multirow}
@@
 
-%% Comments
-\newif\ifcomments\commentstrue
-
-\ifcomments
-\newcommand{\authornote}[3]{\textcolor{#1}{[#3 ---#2]}}
-\newcommand{\todo}[1]{\textcolor{red}{[TODO: #1]}}
-\else
-\newcommand{\authornote}[3]{}
-\newcommand{\todo}[1]{}
-\fi
-
-\newcommand{\wss}[1]{\authornote{blue}{SS}{#1}} %Spencer Smith
-\newcommand{\jc}[1]{\authornote{red}{JC}{#1}} %Jacques Carette
-\newcommand{\mn}[1]{\authornote{magenta}{MN}{#1}} %Mike Noseworthy
-\newcommand{\ad}[1]{\authornote{cyan}{AD}{#1}} %Ao Dong
-
-\newcommand{\notdone}[1]{\textcolor{red}{#1}}
-\newcommand{\done}[1]{\textcolor{black}{#1}}
+\oddsidemargin 0mm
+\evensidemargin 0mm
+\textwidth 165mm
+\textheight 205mm
+
+\pagestyle{fancy}
+\lhead{\leftmark}
+\chead{}
+\rhead{\rightmark}
+\cfoot{\thepage}
 
@@
 
-\journal{Academic Radiology}
-
 \pgfplotsset{compat=1.13}
@@
 
-\begin{frontmatter}
-
-%% Title, authors and addresses
-
-%% use the tnoteref command within \title for footnotes;
-%% use the tnotetext command for theassociated footnote;
-%% use the fnref command within \author or \affiliation for footnotes;
-%% use the fntext command for theassociated footnote;
-%% use the corref command within \author for corresponding author footnotes;
-%% use the cortext command for theassociated footnote;
-%% use the ead command for the email address,
-%% and the form \ead[url] for the home page:
-%% \title{Title\tnoteref{label1}}
-%% \tnotetext[label1]{}
-%% \author{Name\corref{cor1}\fnref{label2}}
-%% \ead{email address}
-%% \ead[url]{home page}
-%% \fntext[label2]{}
-%% \cortext[cor1]{}
-%% \affiliation{organization={},
-%%            addressline={}, 
-%%            city={},
-%%            postcode={}, 
-%%            state={},
-%%            country={}}
-%% \fntext[label3]{}
-
-\title{State of the Practice for Medical Imaging Software}
-
-%% use optional labels to link authors explicitly to addresses:
-%% \author[label1,label2]{}
-%% \affiliation[label1]{organization={},
-%%             addressline={},
-%%             city={},
-%%             postcode={},
-%%             state={},
-%%             country={}}
-%%
-%% \affiliation[label2]{organization={},
-%%             addressline={},
-%%             city={},
-%%             postcode={},
-%%             state={},
-%%             country={}}
-
-\author[CAS]{Spencer Smith}
-\author[CAS]{Ao Dong}
-\author[CAS]{Jacques Carette}
-\author[ECE]{Mike Noseworthy}
-
-\affiliation[CAS]{organization={McMaster University, Computing and Software Department},%Department and Organization
-            addressline={1280 Main Street West}, 
-            city={Hamilton},
-            postcode={L8S 4K1}, 
-            state={Ontario},
-            country={Canada}}
-
-\affiliation[ECE]{organization={McMaster University, Electrical Engineering},%Department and Organization
-            addressline={1280 Main Street West}, 
-            city={Hamilton},
-            postcode={L8S 4K1}, 
-            state={Ontario},
-            country={Canada}}
+\title{State of the Practice for Medical Imaging Software Based on Open Source Repositories}
+
+\author[1,*]{W.\ Spencer Smith}
+\author[1]{Ao Dong}
+\author[1]{Jacques Carette}
+\author[2]{Michael D.\ Noseworthy}
+
+\affil[1]{McMaster University, Computing and Software Department, {Canada}}
+\affil[2]{McMaster University, Electrical \& Computer Engineering Department,
+{Canada}}
+\affil[*]{Corresponding Author}
+
+\maketitle
+
+% RQ5 to RQ10
+% Some of Section 1 (Introduction)
+% Sections 7 to 10 (includes use of tools, some of which was based on interviews with developers)
+% Some of Sections 11 to 13 (Threats to Validity, Future Work, Conclusions)
 
@@
 
-%%Graphical abstract
-%\begin{graphicalabstract}
-%\includegraphics{grabs}
-%\end{graphicalabstract}
-
-%%Research highlights
-%\begin{highlights}
-%\item Research highlight 1
-%\item Research highlight 2
-%\end{highlights}
-
-\begin{keyword}
+\noindent \emph{Keywords:}
 	medical imaging, research software, software engineering, software
-	quality, Analytic Hierarchy Process, developer interviews
-\end{keyword}
-
-\end{frontmatter}
-
-%% \linenumbers
+	quality, analytic hierarchy process, developer interviews
 
@@
     \begin{center}
-        \includegraphics[scale=0.25]{figures/MPR.png}        
+        \includegraphics[scale=0.25]{MPR.png}        
     \end{center}
@@
 found in Section~\ref{ch_methods}.  Our methodology has been applied to MI
-software \citep{Dong2021} and Lattice Boltzmann Solvers \citep{SmithEtAl2022,
-Michalski2021}.  This methodology builds off prior work to assess the state of
+software \citep{Dong2021} and Lattice Boltzmann Solvers \citep{Michalski2021,
+SmithEtAl2024}.  This methodology builds off prior work to assess the state of
 the practice for such domains as Geographic Information Systems
@@
 \begin{figure}[!ht]
-\includegraphics[scale=0.66]{figures/template.pdf}
+\includegraphics[scale=0.66]{template.pdf}
 \caption{Grading template example}
@@
 can be found in \citet{Dong2021}.
-
-\section{Measurement Results} \label{ch_results}
-
-Table~\ref{tab_final_list} shows the 29 software packages that we measured,
-along with summary data collected in the year 2020. We arrange the items in
-descending order of LOC. We found the initial release dates (Rlsd) for most
-projects and marked the two unknown dates with ``?''. The date of the last
-update is the date of the latest update, at the time of measurement. We found
-funding information (Fnd) for only eight projects.  For the Number Of
-Contributors (NOC) we considered anyone who made at least one accepted commit as
-a contributor. The NOC is not usually the same as the number of long-term
-project members, since many projects received change requests and code from the
-community.  With respect to the OS, 25 packages work on all three OSs: Windows
-(W), macOS (M), and Linux (L). Although the usual approach to cross-platform
-compatibility was to work natively on multiple OSes, five projects achieved
-platform-independence via web applications. The full measurement data for all
-packages is available in \citet{Dong2021-Data}.
-
-\begin{table}[!ht]
-\centering
-\begin{tabular}{p{6cm}lllllllll}
-\toprule
-\multirow{2}{*}{Software} & \multirow{2}{*}{Rlsd} & \multirow{2}{*}{Updated} & \multirow{2}{*}{Fnd} & \multirow{2}{*}{NOC} & \multirow{2}{*}{LOC} & \multicolumn{3}{c}{OS} & \multirow{2}{*}{Web} \\ \cline{7-9}
- &  &  &  &  &  & W & M & L &  \\ \midrule
-ParaView \citep{Ahrens2005} & 2002 & 2020-10 & \checkmark & 100 & 886326 & \checkmark & \checkmark & \checkmark & \checkmark \\
-Gwyddion \citep{Nevcas2012} & 2004 & 2020-11 &  & 38 & 643427 & \checkmark & \checkmark & \checkmark &  \\
-Horos \citep{horosproject2020} & ? & 2020-04 &  & 21 & 561617 &  & \checkmark &  &  \\
-OsiriX Lite \citep{PixmeoSARL2019} & 2004 & 2019-11 &  & 9 & 544304 &  & \checkmark &  &  \\
-3D Slicer \citep{Kikinis2014} & 1998 & 2020-08 & \checkmark & 100 & 501451 & \checkmark & \checkmark & \checkmark &  \\
-Drishti \citep{Limaye2012} & 2012 & 2020-08 &  & 1 & 268168 & \checkmark & \checkmark & \checkmark &  \\
-Ginkgo CADx \citep{Wollny2020} & 2010 & 2019-05 &  & 3 & 257144 & \checkmark & \checkmark & \checkmark &  \\
-GATE \citep{Jan2004} & 2011 & 2020-10 &  & 45 & 207122 &  & \checkmark & \checkmark &  \\
-3DimViewer \citep{TESCAN2020} & ? & 2020-03 & \checkmark & 3 & 178065 & \checkmark & \checkmark &  &  \\
-medInria \citep{Fillard2012} & 2009 & 2020-11 &  & 21 & 148924 & \checkmark & \checkmark & \checkmark &  \\
-BioImage Suite Web \citep{Papademetris2005} & 2018 & 2020-10 & \checkmark & 13 & 139699 &
-\checkmark & \checkmark & \checkmark & \checkmark \\
-Weasis \citep{Roduit2021} & 2010 & 2020-08 &  & 8 & 123272 & \checkmark & \checkmark & \checkmark &  \\
-AMIDE \citep{Loening2017} & 2006 & 2017-01 &  & 4 & 102827 & \checkmark & \checkmark & \checkmark &  \\
-XMedCon \citep{Nolf2003} & 2000 & 2020-08 &  & 2 & 96767 & \checkmark & \checkmark & \checkmark &  \\
-ITK-SNAP \citep{Yushkevich2006} & 2006 & 2020-06 & \checkmark & 13 & 88530 & \checkmark & \checkmark & \checkmark &  \\
-Papaya \citep{UTHSCSA2019} & 2012 & 2019-05 &  & 9 & 71831 & \checkmark & \checkmark & \checkmark &  \\
-OHIF Viewer \citep{Ziegler2020} & 2015 & 2020-10 &  & 76 & 63951 & \checkmark & \checkmark & \checkmark & \checkmark \\
-SMILI \citep{Chandra2018} & 2014 & 2020-06 &  & 9 & 62626 & \checkmark & \checkmark & \checkmark &  \\
-INVESALIUS 3 \citep{Amorim2015} & 2009 & 2020-09 &  & 10 & 48605 & \checkmark & \checkmark & \checkmark &  \\
-dwv \citep{Martelli2021} & 2012 & 2020-09 &  & 22 & 47815 & \checkmark & \checkmark & \checkmark & \checkmark \\
-DICOM Viewer \citep{Afsar2021} & 2018 & 2020-04 & \checkmark & 5 & 30761 & \checkmark & \checkmark & \checkmark &  \\
-MicroView \citep{ParallaxInnovations2020} & 2015 & 2020-08 &  & 2 & 27470 & \checkmark & \checkmark & \checkmark &  \\
-MatrixUser \citep{Liu2016} & 2013 & 2018-07 &  & 1 & 23121 & \checkmark & \checkmark & \checkmark &  \\
-Slice:Drop \citep{Haehn2013} & 2012 & 2020-04 &  & 3 & 19020 & \checkmark & \checkmark & \checkmark & \checkmark \\
-dicompyler \citep{Panchal2010} & 2009 & 2020-01 &  & 2 & 15941 & \checkmark & \checkmark &  &  \\
-Fiji \citep{Schindelin2012} & 2011 & 2020-08 & \checkmark & 55 & 10833 & \checkmark & \checkmark & \checkmark &  \\
-ImageJ \citep{Rueden2017} & 1997 & 2020-08 & \checkmark & 18 & 9681 & \checkmark & \checkmark & \checkmark &  \\
-MRIcroGL \citep{Rorden2021} & 2015 & 2020-08 &  & 2 & 8493 & \checkmark & \checkmark & \checkmark &  \\
-DicomBrowser \citep{Archie2012} & 2012 & 2020-08 &  & 3 & 5505 & \checkmark & \checkmark & \checkmark &  \\ \bottomrule
-\end{tabular}
-\caption{Final software list (sorted in descending order of the number of Lines
-Of Code (LOC))}
-\label{tab_final_list}
-\end{table}
-
-Figure \ref{fig_language} shows the primary languages versus the number of
-projects using them.  The primary language is the language used for the majority
-of the project's code; in most cases projects also use other languages.  The
-most popular language is \CC, with almost 40\% of projects (11 of 29).  The two
-least popular choices are Pascal and Matlab, with around 3\% of projects each
-(1 of 29).
-
-\begin{figure}[!ht]
-\centering
-\includegraphics[scale=0.5]{figures/PrimaryLanguages.pdf}
-\centering
-\caption{\label{fig_language}Primary languages versus number of projects}
-\end{figure}
-
-\subsection{Installability} \label{sec_result_installability}
-
-Figure \ref{fg_installability_scores} lists the installability scores.  We found
-installation instructions for 16 projects. Among the ones without instructions,
-\textit{BioImage Suite Web} and \textit{Slice:Drop} do not need installation,
-since they are web applications. Installing 10 of the projects required extra
-dependencies. Five of them are web applications (as shown in
-Table~\ref{tab_final_list}) and depend on a browser; \textit{dwv}, \textit{OHIF
-Viewer}, and \textit{GATE} needs extra dependencies to build; \textit{ImageJ}
-and	\textit{Fiji} need an unzip tool; \textit{MatrixUser} is based on Matlab;
-\textit{DICOM Viewer} needs to work on a Nextcloud platform.
-
-\begin{figure}[!ht]
-\includegraphics[scale=0.47]{figures/installability_scores.pdf}
-\caption{AHP installability scores}
-\label{fg_installability_scores}
-\end{figure}
-
-\textit{3D Slicer} has the highest score because it had easy to follow
-installation instructions, and an automated, fast, frustration-free installation
-process. The installer added all dependencies automatically and no errors
-occurred during the installation and uninstallation steps. Many other software
-packages also had installation instructions and automated installers.  We had no
-trouble installing the following packages: \textit{INVESALIUS 3},
-\textit{Gwyddion}, \textit{XMedCon}, and \textit{MicroView}. We determined their
-scores based on the understandability of the instructions, installation steps,
-and user experience. Since \textit{BioImage Suite Web} and \textit{Slice:Drop}
-needed no installation, we gave them high scores. \textit{BioImage Suite Web}
-also provided an option to download cache for offline usage, which was easy to
-apply.
-
-\textit{GATE}, \textit{dwv}, and \textit{DICOM Viewer} showed severe
-installation problems. We were not able to install them, even after a reasonable
-amount of time (2 hours).  For \textit{dwv} and \textit{GATE} we failed to build
-from the source code, but we were able to proceed with measuring other qualities
-using a deployed on-line version for \textit{dwv}, and a VM version for
-\textit{GATE}. For \textit{DICOM Viewer} we could not install the NextCloud
-dependency, and we did not have another option for running the software.
-Therefore, for \textit{DICOM Viewer} we could not measure reliability or
-robustness.  The other seven qualities could be measured, since they do not
-require installation.
-
-\textit{MatrixUser} has a lower score because it depends on Matlab. We assessed
-the score from the point of view of a user that would have to install Matlab and
-acquire a license.  Of course, for users that already work within Matlab, the
-installability score should be higher.
-
-\subsection{Correctness \& Verifiability} \label{sec_result_correctness_verifiability}
-
-Figure~\ref{fg_correctness_verifiability_scores} shows the scores of correctness
-and verifiability. Generally speaking, the packages with higher scores adopted
-more techniques to improve correctness, and had better documentation for us to
-verify against.  For instance, we looked for evidence of unit testing, since it
-benefits most parts of the software's life cycle, such as designing, coding,
-debugging, and optimization \citep{Hamill2004}.  We only found evidence of unit
-testing for about half of the projects. We identified five projects using CI/CD
-tools: \textit{3D Slicer}, \textit{ImageJ}, \textit{Fiji}, \textit{dwv}, and
-\textit{OHIF Viewer}.
-
-\begin{figure}[!ht]
-\includegraphics[scale=0.47]{figures/correctness_verifiability_scores.pdf}
-\caption{AHP correctness \& verifiability scores}
-\label{fg_correctness_verifiability_scores}
-\end{figure}
-
-Even for some projects with well-organized documentation, requirements
-specifications and theory manuals were still missing.  We could not identify
-theory manuals for all projects, and we did not find requirements specifications
-for most projects. The only requirements-related document we found was a road
-map of \textit{3D Slicer}, which contained design requirements for upcoming
-changes.
-
-\subsection{Surface Reliability} \label{sec_result_reliability}
-
-Figure~\ref{fg_reliability_scores} shows the AHP results.  As shown in
-Section~\ref{sec_result_installability}, most of the software products did not
-``break'' during installation, or did not need installation; \textit{dwv} and
-\textit{GATE} broke in the building stage, and the processes were not
-recoverable; we could not install the dependency for \textit{DICOM Viewer}. Of
-the seven software packages with a getting started tutorial and operation steps
-in the tutorial, most showed no error when we followed the steps. However,
-\textit{GATE} could not open macro files and became unresponsive several times,
-without any descriptive error message. When assessing robustness
-(Section~\ref{sec_result_robustness}), we found that \textit{Drishti} crashed
-when loading damaged image files, without showing any descriptive error message.
-We did not find any problems with the on-line version of
-\textit{dwv}.
-
-\begin{figure}[!ht]
-\includegraphics[scale=0.47]{figures/reliability_scores.pdf}
-\caption{AHP surface reliability scores}
-\label{fg_reliability_scores}
-\end{figure}
-
-\subsection{Surface Robustness} \label{sec_result_robustness}
-
-Figure \ref{fg_robustness_scores} presents the scores for surface robustness.
-The packages with higher scores elegantly handled unexpected/unanticipated
-inputs, typically showing a clear error message. We may have underestimated the
-score of \textit{OHIF Viewer}, since we needed further customization to load
-data.
-
-\begin{figure}[!ht]
-\includegraphics[scale=0.47]{figures/robustness_scores.pdf}
-\caption{AHP surface robustness scores}
-\label{fg_robustness_scores}
-\end{figure}
-
-Digital Imaging and Communications in Medicine (DICOM) ``defines the formats for
-medical images that can be exchanged with the data and quality necessary for
-clinical use'' \citep{MITA2021}. According to their documentation, all 29
-software packages should support the DICOM standard. To test robustness, we
-prepared two types of image files: correct and incorrect formats (with the
-incorrect format created by relabelled a text file to have the ``.dcm''
-extension).  All software packages loaded the correct format image, except for
-\textit{GATE}, which failed for unknown reasons.  For the broken format,
-\textit{MatrixUser}, \textit{dwv}, and \textit{Slice:Drop} ignored the incorrect
-format of the file and loaded it regardless. They did not show any error message
-and displayed a blank image. \textit{MRIcroGL} behaved similarly except that it
-showed a meaningless image. \textit{Drishti} successfully detected the broken
-format of the file, but the software crashed as a result.
-
-\subsection{Surface Usability} \label{sec_result_usability}
-
-Figure~\ref{fg_usability_scores} shows the AHP scores for surface usability. The
-software with higher scores usually provided both comprehensive documented
-guidance and a good user experience. \textit{INVESALIUS 3} provided an excellent
-example of a detailed and precise user manual. \textit{GATE} also provided
-numerous documents, but unfortunately we had difficulty understanding and using
-them. We found getting started tutorials for only 11 projects, but a user manual
-for 22 projects. \textit{MRIcroGL} was the only project that explicitly
-documented expected user characteristics.
-
-\begin{figure}[!ht]
-\includegraphics[scale=0.47]{figures/usability_scores.pdf}
-\caption{AHP surface usability scores}
-\label{fg_usability_scores}
-\end{figure}
-
-\subsection{Maintainability} \label{sec_score_maintainability}
-
-Figure~\ref{fg_maintainability_scores} shows the ranking results for
-maintainability. We gave \textit{3D Slicer} the highest score because we found
-it to have the most comprehensive artifacts. For example, as far as we could
-find, only a few of the 29 projects had a product, developer's manual, or API
-documentation, and only \textit{3D Slicer}, \textit{ImageJ}, \textit{Fiji}
-included all three documents. Moreover, \textit{3D Slicer} has a much higher
-percentage of closed issues (91.65\%) compared to \textit{ImageJ} (52.49\%) and
-\textit{Fiji} (63.79\%). Table~\ref{tab_maintainability_docs} shows which
-projects had these documents, in descending order of their maintainability
-scores.
-
-\begin{figure}[!ht]
-\includegraphics[scale=0.47]{figures/maintainability_scores.pdf}
-\caption{AHP maintainability scores}
-\label{fg_maintainability_scores}
-\end{figure}
-
-\begin{table}[!ht]
-\centering
-\begin{tabular}{lccc}
-\toprule
-\multicolumn{1}{c}{Software} & Prod.\ Roadmap & Dev.\ Manual & API Doc. \\ 
-\midrule
-3D Slicer & \checkmark & \checkmark & \checkmark \\
-ImageJ & \checkmark & \checkmark & \checkmark \\
-Weasis &  & \checkmark &  \\
-OHIF Viewer &  & \checkmark & \checkmark \\
-Fiji & \checkmark & \checkmark & \checkmark \\
-ParaView & \checkmark &  &  \\
-SMILI &  &  & \checkmark \\
-medInria &  & \checkmark &  \\
-INVESALIUS 3 & \checkmark &  &  \\
-dwv &  &  & \checkmark \\
-BioImage Suite Web &  & \checkmark &  \\
-Gwyddion &  & \checkmark & \checkmark \\ 
-\bottomrule
-\end{tabular}
-\caption{Software with the maintainability documents (listed in descending order of 
-maintainability score)}
-\label{tab_maintainability_docs}
-\end{table}
-
-Twenty-seven of the 29 projects used git as the version control tool, with 24 of these
-using GitHub. \textit{AMIDE} used Mercurial and \textit{Gwyddion} used
-Subversion. \textit{XMedCon}, \textit{AMIDE}, and \textit{Gwyddion} used
-SourceForge. \textit{DicomBrowser} and \textit{3DimViewer} used BitBucket.
-
-\subsection{Reusability} \label{sec_result_reusability}
-
-Figure~\ref{fg_reusability_scores} shows the AHP results for reusability. As
-described in Section~\ref{sec_grading_software}, we gave higher scores to the
-projects with API documentation. As shown in
-Table~\ref{tab_maintainability_docs}, seven projects had API documents. We also
-assumed that projects with more code files and less LOC per code file are more
-reusable. Table \ref{tab_loc_per_file} shows the number of text-based files by
-project, which we used to approximate the number of code files. The table also
-lists the total number of lines (including comments and blanks), LOC, and
-average LOC per file. We arranged the items in descending order of their
-reusability scores.
-
-\begin{figure}[!ht]
-\includegraphics[scale=0.47]{figures/reusability_scores.pdf}
-\caption{AHP reusability scores}
-\label{fg_reusability_scores}
-\end{figure}
-
-\begin{table}[!ht]
-\centering
-\begin{tabular}{lllll}
-\toprule
-\multirow{2}{*}{Software} & \multirow{2}{*}{Text Files} & \multirow{2}{*}{Total Lines} & \multirow{2}{*}{LOC} & \multirow{2}{*}{LOC/file} \\
- &  &  &  &  \\ 
-\midrule
-OHIF Viewer & 1162 & 86306 & 63951 & 55 \\
-3D Slicer & 3386 & 709143 & 501451 & 148 \\
-Gwyddion & 2060 & 787966 & 643427 & 312 \\
-ParaView & 5556 & 1276863 & 886326 & 160 \\
-OsiriX Lite & 2270 & 873025 & 544304 & 240 \\
-Horos & 2346 & 912496 & 561617 & 239 \\
-medInria & 1678 & 214607 & 148924 & 89 \\
-Weasis & 1027 & 156551 & 123272 & 120 \\
-BioImage Suite Web & 931 & 203810 & 139699 & 150 \\
-GATE & 1720 & 311703 & 207122 & 120 \\
-Ginkgo CADx & 974 & 361207 & 257144 & 264 \\
-SMILI & 275 & 90146 & 62626 & 228 \\
-Fiji & 136 & 13764 & 10833 & 80 \\
-Drishti & 757 & 345225 & 268168 & 354 \\
-ITK-SNAP & 677 & 139880 & 88530 & 131 \\
-3DimViewer & 730 & 240627 & 178065 & 244 \\
-DICOM Viewer & 302 & 34701 & 30761 & 102 \\
-ImageJ & 40 & 10740 & 9681 & 242 \\
-dwv & 188 & 71099 & 47815 & 254 \\
-MatrixUser & 216 & 31336 & 23121 & 107 \\
-INVESALIUS 3 & 156 & 59328 & 48605 & 312 \\
-AMIDE & 183 & 139658 & 102827 & 562 \\
-Papaya & 110 & 95594 & 71831 & 653 \\
-MicroView & 137 & 36173 & 27470 & 201 \\
-XMedCon & 202 & 129991 & 96767 & 479 \\
-MRIcroGL & 97 & 50445 & 8493 & 88 \\
-Slice:Drop & 77 & 25720 & 19020 & 247 \\
-DicomBrowser & 54 & 7375 & 5505 & 102 \\
-dicompyler & 48 & 19201 & 15941 & 332 \\ 
-\bottomrule
-\end{tabular}
-\caption{Number of files and lines (sorted in descending order of reusability
-scores)}
-\label{tab_loc_per_file}
-\end{table}
-
-\subsection{Surface Understandability} \label{sec_result_understandability}
-
-Figure~\ref{fg_surface_understandability_scores} shows the scores for surface
-understandability. All projects had a consistent coding style with parameters in
-the same order for all functions; modularized code; and, clear comments, indicating
-what is done, not how. However, we only found explicit identification of a
-coding standard for 3 out of the 29: \textit{3D Slicer}, \textit{Weasis}, and
-\textit{ImageJ}. We also found hard-coded constants (rather than symbolic
-constants) in \textit{medInria}, \textit{dicompyler}, \textit{MicroView}, and
-\textit{Papaya}. We did not find any reference to the algorithms used in
-projects \textit{XMedCon}, \textit{DicomBrowser}, \textit{3DimViewer},
-\textit{BioImage Suite Web}, \textit{Slice:Drop}, \textit{MatrixUser},
-\textit{DICOM Viewer}, \textit{dicompyler}, and \textit{Papaya}.
-
-\begin{figure}[!ht]
-\includegraphics[scale=0.47]{figures/understandability_scores.pdf}
-\caption{AHP surface understandability scores}
-\label{fg_surface_understandability_scores}
-\end{figure}
-
-\subsection{Visibility/Transparency} \label{sec_result_visibility_transparency}
-
-Figure~\ref{fg_visibility_transparency_scores} shows the AHP scores for
-visibility/transparency. Generally speaking, the teams that actively documented
-their development process and plans scored higher.
-Table~\ref{tab_Visibility/Transparency_docs} shows the projects that had
-documents for the development process, project status, development environment,
-and release notes, in descending order of their visibility/transparency
-scores.
-
-\begin{figure}[!ht]
-\includegraphics[scale=0.47]{figures/visibility_transparency_scores.pdf}
-\caption{AHP visibility/transparency scores}
-\label{fg_visibility_transparency_scores}
-\end{figure}
-
-\begin{table}[!ht]
-\centering
-\begin{tabular}{lllll}
-\toprule
-Software & Dev.\ Process & Proj.\ Status & Dev.\ Env. & Rls.\ Notes \\ 
-\midrule
-3D Slicer & \checkmark & \checkmark & \checkmark & \checkmark \\
-ImageJ & \checkmark & \checkmark & \checkmark & \checkmark \\
-Fiji & \checkmark & \checkmark & \checkmark &  \\
-MRIcroGL &  &  &  & \checkmark \\
-Weasis &  &  & \checkmark & \checkmark \\
-ParaView &  & \checkmark &  &  \\
-OHIF Viewer &  &  & \checkmark & \checkmark \\
-DICOM Viewer &  &  & \checkmark & \checkmark \\
-medInria &  &  & \checkmark & \checkmark \\
-SMILI &  &  &  & \checkmark \\
-Drishti &  &  &  & \checkmark \\
-INVESALIUS 3 &  &  &  & \checkmark \\
-OsiriX Lite &  &  &  & \checkmark \\
-GATE &  &  &  & \checkmark \\
-MicroView &  &  &  & \checkmark \\
-MatrixUser &  &  &  & \checkmark \\
-BioImage Suite Web &  &  & \checkmark &  \\
-ITK-SNAP &  &  &  & \checkmark \\
-Horos &  &  &  & \checkmark \\
-dwv &  &  &  & \checkmark \\
-Gwyddion &  &  &  & \checkmark \\ 
-\bottomrule
-\end{tabular}
-\caption{Software with visibility/transparency related documents (listed in
-descending order of visibility/transparency score)}
-\label{tab_Visibility/Transparency_docs}
-\end{table}
-
-\subsection{Overall Scores} \label{Sec_OverallQ}
-
-As described in Section~\ref{sec_AHP}, for our AHP measurements, we have nine
-criteria (qualities) and 29 alternatives (software packages). In the absence of
-a specific real world context, we assumed all nine qualities are equally
-important. Figure~\ref{fg_overall_scores} shows the overall scores in descending
-order. Since we produced the scores from the AHP process, the total sum of the
-29 scores is precisely 1.0.
-
-\begin{figure}[!ht]
-\includegraphics[scale=0.47]{figures/overall_scores.pdf}
-\caption{Overall AHP scores with an equal weighting for all 9 software qualities}
-
-\label{fg_overall_scores}
-\end{figure}
-
-The top four software products \textit{3D Slicer}, \textit{ImageJ},
-\textit{Fiji}, and \textit{OHIF Viewer} have higher scores in most criteria.
-\textit{3D Slicer} has a score in the top two for all qualities; \textit{ImageJ}
-ranks near the top for all qualities, except for correctness \& verifiability.
-\textit{OHIF Viewer} and \textit{Fiji} have similar overall scores, with
-\textit{Fiji} doing better in installability and \textit{OHIF Viewer} doing
-better in correctness \& verifiability.  Given the installation problems, we may
-have underestimated the scores on reliability and robustness for \textit{DICOM
-Viewer}, but we compared it equally for the other seven qualities.
-
-\section{Comparison to Community Ranking} \label{Sec_VsCommunityRanking}
-
-To address~\rqref{RQ_CompareHQ2Popular} about how our ranking compares to the
-popularity of projects as judged by the scientific community, we make two
-comparisons:
-\begin{itemize}
-\item A comparison of our ranking (from Section~\ref{ch_results}) with the
-community ratings on GitHub, as shown by GitHub stars, number of forks, and
-number of people watching the projects; and,
-\item A comparison of top-rated software from our methodology with the top
-recommendations from our domain experts (as mentioned in
-Section~\ref{sec_software_selection}).
-\end{itemize}
-
-Table~\ref{tab_ranking_vs_GitHub} shows our ranking of the 29 MI projects, and
-their GitHub metrics, if applicable. As mentioned in
-Section~\ref{sec_score_maintainability}, 24 projects used GitHub. Since GitHub
-repositories have different creation dates, we collect the number of months each
-stayed on GitHub, and calculate the average number of new stars, people
-watching, and forks per 12 months. Section~\ref{sec_grading_software} describes
-the method of getting the creation date.  The items in
-Table~\ref{tab_ranking_vs_GitHub} are listed in descending order of the average
-number of new stars per year.  The non-GitHub items are listed in the order of
-our ranking.  We collected all GitHub statistics in July 2021.
-
-Generally speaking, most of the top-ranking MI software projects also received
-greater attention and popularity on GitHub. Between our ranking and the GitHub
-stars-per-year ranking, four of the top five software projects appear in both
-lists. Our top five packages are scattered among the first eight positions on the
-GitHub list. However, as discussed below there are discrepancies between the two
-lists.
-
-In some cases projects are popular in the community, but were assigned a low
-rank by our methodology.  This is the case for \textit{dwv}. The reason for the
-low ranking is that, as mentioned in Section~\ref{sec_result_installability}, we
-failed to build it locally, and used the test version on its websites for the
-measurements. We followed the instructions and tried to run the command ``yarn
-run test'' locally, which did not work. In addition, the test version did not
-detect a broken DICOM file and displayed a blank image as described in
-Section~\ref{sec_result_robustness}. We might underestimate the scores for
-\textit{dwv} due to uncommon technical issues.
-
-We also ranked \textit{DICOM Viewer} much lower than its popularity. As
-mentioned in Section~\ref{sec_result_installability}, it depended on the
-NextCloud platform that we could not successfully install. Thus, we might
-underestimate the scores of its surface reliability and surface robustness.
-
-Further reason for discrepancies between our ranking and the community's ranking
-is that we weighted all qualities equally. This is not likely how users
-implicitly rank the different qualities. As a result, some projects with high
-community popularity may have scored lower with our method because of a
-relatively higher (compared to the scientific community's implicit ranking)
-weighting of the poor scores for some qualities. A further explanation for
-discrepancies between our measures and the star measures may also be due to
-inaccuracy with using stars to approximate popularity.  Stars are not an ideal
-measure because stars represent the community's feeling in the past more than
-they measure current preferences \citep{Szulik2017}.  The issue with stars is
-that they tend only to be added, not removed.  A final reason for
-inconsistencies between our ranking and the community's ranking is that, as for
-consumer products, more factors influence popularity than just quality.
-
-\begingroup
-\renewcommand{\arraystretch}{0.85}
-\begin{table}[!ht]
-\centering
-\begin{tabular}{llllll}
-\toprule
-Software & Comm.\ Rank & Our Rank & Stars/yr & Watches/yr & Forks/yr \\ 
-\midrule
-3D Slicer & 1 & 1 & 284 & 19 & 128 \\
-OHIF Viewer & 2 & 4 & 277 & 19 & 224 \\
-dwv & 3 & 19 & 124 & 12 & 51 \\
-ImageJ & 4 & 2 & 84 & 9 & 30 \\
-ParaView & 5 & 5 & 67 & 7 & 28 \\
-Horos & 6 & 12 & 49 & 9 & 18 \\
-Papaya & 7 & 17 & 45 & 5 & 20 \\
-Fiji & 8 & 3 & 44 & 5 & 21 \\
-DICOM Viewer & 9 & 29 & 43 & 6 & 9 \\
-INVESALIUS 3 & 10 & 8 & 40 & 4 & 17 \\
-Weasis & 11 & 7 & 36 & 5 & 19 \\
-dicompyler & 12 & 26 & 35 & 5 & 14 \\
-OsiriX Lite & 13 & 11 & 34 & 9 & 24 \\
-MRIcroGL & 14 & 18 & 24 & 3 & 3 \\
-GATE & 15 & 24 & 19 & 6 & 26 \\
-Ginkgo CADx & 16 & 14 & 19 & 4 & 6 \\
-BioImage Suite Web & 17 & 6 & 18 & 5 & 7 \\
-Drishti & 18 & 27 & 16 & 4 & 4 \\
-Slice:Drop & 19 & 21 & 10 & 2 & 5 \\
-ITK-SNAP & 20 & 13 & 9 & 1 & 4 \\
-medInria & 21 & 9 & 7 & 3 & 6 \\
-SMILI & 22 & 10 & 3 & 1 & 2 \\
-MatrixUser & 23 & 28 & 2 & 0 & 0 \\
-MicroView & 24 & 15 & 1 & 1 & 1 \\
-Gwyddion & 25 & 16 & n/a & n/a & n/a \\
-XMedCon & 26 & 20 & n/a & n/a & n/a \\
-DicomBrowser & 27 & 22 & n/a & n/a & n/a \\
-AMIDE & 28 & 23 & n/a & n/a & n/a \\
-3DimViewer & 29 & 25 & n/a & n/a & n/a \\ 
-\bottomrule
-\end{tabular}
-\caption{Software ranking by our methodology versus the community (Comm.)\
-ranking using GitHub metrics (Sorted in descending order of community
-popularity, as estimated by the number of new stars per year)}
-\label{tab_ranking_vs_GitHub}
-\end{table}
-\endgroup
-
-As shown in Section~\ref{sec_software_selection}, our domain experts recommended
-a list of top software with 12 software products.  All the top 4 entries from
-the Domain Expert's list are among the top 12 ranked by our methodology. Three
-of the top four on both lists are the same: \textit{3D Slicer}, \textit{ImageJ},
-and \textit{Fiji}. \textit{3D Slicer} is top project by both rankings (and by
-the GitHub stars measure as well).  The Domain Expert ranked \textit{Horos} as
-their second choice, while we ranked it twelfth.  Our third ranked project,
-\textit{OHIF Viewer} was not listed by the Domain Expert.  Neither were the
-software packages that we ranked from fifth to eleventh (\textit{ParaView},
-\textit{Weasis}, \textit{medInria}, \textit{BioImage Suite Web}, \textit{OsiriX
-Lite}, \textit{INVESALIUS}, and \textit{Gwyddion}).  The software mentioned by
-the Domain Expert that we did not rank were the six recommended packages that
-did not have visualization as the primary function (as discussed in
-Section~\ref{sec_software_selection}).  The differences between the list
-recommended by our methodology and the Domain Expert are not surprising.  As
-mentioned above, our methodology weights all qualities equally, but that may not
-be the case for the Domain Expert's impressions.  Moreover, although the Domain
-Expert has significant experience with MI software, they have not used all 29
-packages that were measured.
-
-Although our ranking and the estimate of the community's ranking are not perfect
-measures, they do suggest a correlation between best practices and popularity.
-We do not know which comes first, the use of best practices or popularity, but
-we do know that the top ranked packages tend to incorporate best practices. The
-next sections will explore how the practices of the MI community compare to the
-broader research software community. We will also investigate the practices from
-the top projects that others within the MI community, and within the broader
-research software community, can potentially adopt.
-
-\section{Comparison Between MI and Research Software for Artifacts}
-\label{Sec_CompareArtifacts}
-
-As part of filling in the measurement template (from
-Section~\ref{sec_grading_software}), we summarized the artifacts observed in
-each MI package. Table~\ref{artifactspresent} groups the artifacts by frequency
-into categories of common (20 to 29 ($>$67\%) packages), uncommon (10 to 19
-(33-67\%) packages), and rare (1 to 9 ($<$33\%) packages). \citet{Dong2021-Data}
-summarizes the full measurements.  Tables~\ref{tab_maintainability_docs}
-and~\ref{tab_Visibility/Transparency_docs} show the details on which projects
-use which types of artifacts for documents related to maintainability and
-visibility, respectively.
-
-\begin{table}[ht!]
-    \begin{center}
-    \begin{tabular}{ p{4.6 cm} p{5.6 cm} p{5 cm}}
-    \toprule
-    Common & Uncommon & Rare \\
-    \midrule
-    README (29) & Build scripts (18) & Getting Started (9)\\
-    Version control (29) & Tutorials (18) & Developer's manual (8)\\
-    License (28) & Installation guide (16) & Contributing (8)\\
-    Issue tracker (28) & Test cases (15) & API documentation (7)\\
-    User manual (22) & Authors (14) & Dependency list (7)\\
-    Release info. (22) & Frequently Asked Questions (FAQ) (14) & Troubleshooting guide (6)\\
-     & Acknowledgements (12) & Product roadmap (5)\\
-     & Changelog (12) & Design documentation (5)\\
-     & Citation (11) & Code style guide (3)\\
-     & & Code of conduct (1)\\
-     & & Requirements (1)\\
-    \bottomrule
-    \end{tabular}
-    \caption{Artifacts Present in MI Packages, Classified by Frequency (The number 
-    in brackets is the number of occurrences)}
-    \label{artifactspresent}
-    \end{center}
-\end{table}
-
-We answer~\rqref{RQ_CompareArtifacts} by comparing the artifacts that we
-observed in MI repositories to those observed and recommended for research
-software in general. Our comparison may point out areas where some MI software
-packages fall short of current best practices. This is not intended to be a
-criticism of any existing packages, especially since in practice not every
-project needs to achieve the highest possible quality. However, rather than
-delve into the nuances of which software can justify compromising which
-practices we will write our comparison under the ideal assumption that every
-project has sufficient resources to match best practices.
-
-Table~\ref{Tbl_Guidelines} (based on data from \citep{SmithAndMichalski2022})
-shows that MI artifacts generally match the recommendations found in nine
-current research software development guidelines:
-\begin{itemize}
-\item United States Geological Survey Software Planning Checklist
-\citep{USGS2019},
-\item DLR (German Aerospace Centre) Software Engineering Guidelines
-\citep{TobiasEtAl2018}, 
-\item Scottish Covid-19 Response Consortium Software Checklist
-\citep{BrettEtAl2021},
-\item Good Enough Practices in Scientific Computing \citep{WilsonEtAl2016},
-\item xSDK (Extreme-scale Scientific Software Development Kit) Community Package
-Policies \citep{SmithAndRoscoe2018},
-\item Trilinos Developers Guide \citep{HerouxEtAl2008},
-\item EURISE (European Research Infrastructure Software Engineers') Network
-Technical Reference \citep{ThielEtAl2020},
-\item CLARIAH (Common Lab Research Infrastructure for the Arts and Humanities)
-Guidelines for Software Quality \citep{vanGompelEtAl2016}, and
-\item A Set of Common Software Quality Assurance Baseline Criteria for Research
-Projects \citep{OrvizEtAl2017}.
-\end{itemize}
-
-In Table~\ref{Tbl_Guidelines} each row corresponds to an artifact.  For a given
-row, a checkmark in one of the columns means that the corresponding guideline
-recommends this artifact.  The last column shows whether the artifact appears in
-the measured set of MI software, either not at all (blank), commonly (C),
-uncommonly (U) or rarely (R).  We did our best to interpret the meaning of each
-artifact consistently between guidelines and specific MI software, but the
-terminology and the contents of artifacts are not standardized.  The challenge
-even exists for the ubiquitous README file.  As illustrated by
-\citet{PranaEtAl2018}, the content of README files shows significant variation
-between projects.  Although some content is reasonably consistent, with 97\% of
-README files contain at least one section describing the `What' of the
-repository and 89\% offering some `How' content, other categories are more
-variable.  For instance, information on `Contribution', `Why', and `Who',
-appear in 28\%, 26\% and 53\% of the analyzed files, respectively
-\citep{PranaEtAl2018}.
-
-The frequency of checkmarks in Table~\ref{Tbl_Guidelines} indicates the
-popularity of recommending a given artifact, but it does not imply that the most
-commonly recommended artifacts are the most important artifacts. Just because a
-guideline does not explicitly recommend an artifact, does not mean the guideline
-authors do not value that artifact.  They may have excluded it because it is out
-of the scope of their recommendations, or outside their experience.  For
-instance, an artifact related to uninstall is only explicitly mentioned by
-\citet{vanGompelEtAl2016}, but other guideline authors would likely see its
-value.  They may simply feel that uninstall is implied by install, or they may
-have never asked themselves whether they need separate uninstall instructions.
-
-\begin{table}[!ht]
-\begin{center}
-\begin{tabular}{ p{2.5cm}p{1cm}p{1cm}p{1cm}p{1cm}p{1cm}p{1cm}p{1cm}p{1.2cm}p{1cm}p{0.8cm} }
-\toprule
-~ \ & \citet{USGS2019} & \citet{TobiasEtAl2018} & \citet{BrettEtAl2021} &
-\citet{WilsonEtAl2016} & \citet{SmithAndRoscoe2018} & \citet{HerouxEtAl2008} &
-\citet{ThielEtAl2020} & \citet{vanGompelEtAl2016} & \citet{OrvizEtAl2017} & MI\\
-\midrule
-LICENSE & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & &
-\checkmark & \checkmark & \checkmark & C\\
-README &  & \checkmark & \checkmark & \checkmark & \checkmark & & \checkmark &
-\checkmark & \checkmark & C\\
-CONTRIBUTING &  & \checkmark & \checkmark & \checkmark & \checkmark & &
-\checkmark & \checkmark & \checkmark & R\\
-CITATION &  &  &  & \checkmark & & & & \checkmark & \checkmark & U\\
-CHANGELOG &  & \checkmark &  & \checkmark & \checkmark & & \checkmark &  &  & U\\
-INSTALL &  &  &  &  & \checkmark & & \checkmark & \checkmark & \checkmark & U\\
-\midrule
-Uninstall &  &  &  &  & & & & \checkmark & &  \\
-Dependency List &  &  & \checkmark & & \checkmark & & & \checkmark &  & R\\
-Authors &  &  &  &  &  &  & \checkmark & \checkmark & \checkmark & U\\
-Code of Conduct &  &  &  &  & & & \checkmark & & & R\\
-Acknowledgements &  &  &  &  &  &  & \checkmark & \checkmark & \checkmark & U\\
-Code Style Guide &  & \checkmark &  &  & & & \checkmark & \checkmark & \checkmark & R\\
-Release Info. &  & \checkmark &  &  & & \checkmark & \checkmark & & & C\\
-Prod.\ Roadmap &  &  &  &  & & \checkmark & \checkmark & \checkmark & & R\\
-\midrule
-Getting started &  &  &  &  & \checkmark & & \checkmark & \checkmark & \checkmark & R\\
-User manual &  &  & \checkmark &  & & & \checkmark & & & C\\
-Tutorials &  &  &  &  & & & \checkmark & & & U\\
-FAQ &  &  &  &  & & & \checkmark & \checkmark & \checkmark & U\\
-\midrule
-Issue Track &  & \checkmark & \checkmark & & \checkmark & \checkmark &
-\checkmark & & \checkmark & C\\
-Version Control &  & \checkmark & \checkmark & \checkmark & \checkmark &
-\checkmark & \checkmark & \checkmark & \checkmark & C\\ 
-Build Scripts &  & \checkmark &  & \checkmark & \checkmark & \checkmark &
-\checkmark & & \checkmark & U\\
-\midrule
-Requirements &  & \checkmark &  &  & & \checkmark &  &  & \checkmark & R\\
-Design Doc.\ &  & \checkmark  & \checkmark &  & \checkmark & & \checkmark &
-\checkmark& \checkmark & R\\
-API Doc. &  &  &  &  & \checkmark & & \checkmark & \checkmark & \checkmark & R\\
-Test Plan &  & \checkmark &  &  & & \checkmark & & & &  \\
-Test Cases & \checkmark & \checkmark & \checkmark &  & \checkmark & \checkmark &
-\checkmark & \checkmark & \checkmark & U\\
-\bottomrule
-\end{tabular}
-\caption{Comparison of Recommended Artifacts in Software Development Guidelines
-to Artifacts in MI Projects (C for Common, U for Uncommon and R for Rare)}
-\label{Tbl_Guidelines}
-\end{center}
-\end{table}
-
-Two of the items that appear in Table~\ref{artifactspresent} do not appear in
-the software development guidelines shown in Table~\ref{Tbl_Guidelines}:
-Troubleshooting guide and Developer's manual.  Although the guidelines do not
-name these two artifacts, the information contained within them overlaps with
-the recommended artifacts.  A Troubleshooting guideline contains information
-that would typically be found in a User manual.  A Developer's guide overlaps
-with information from the README, INSTALL, Uninstall, Dependency List, Release
-Information, API documentation and Design documentation.  In our current
-analysis, we have identified artifacts by the names given by the software
-guidelines and MI examples.  In the future, a more in-depth analysis would look
-at the knowledge fragments that are captured in the artifacts, rather than
-focusing on the names of the files that collect these fragments together.
-
-Although the MI community shows examples of 88\% (23 of 26) of the practices we
-found in research software guidelines (Table~\ref{Tbl_Guidelines}), we did not
-observe three recommended artifacts: i) Uninstall, ii) Test plans, and iii)
-Requirements.  Uninstall is likely an omission caused by the focus on installing
-software. Given the storage capacity of current hardware, developers and users
-are not generally concerned with uninstall.  Moreover, as mentioned above,
-uninstall is not particularly emphasized in existing recommendations.
-
-Test plans describe the scope, approach, resources, and the schedule of planned
-test activities \citep{VanVliet2000}.  The plan should cover details such as the
-makeup of the testing team, automated testing tools and technology to employ,
-the testing process, system tests, integration tests and unit tests. We did not
-observe test plans for MI software, but that doesn't mean plans weren't created;
-it means that the plans are not under version control. Test plans would have to
-at least be implicitly created, since we observed test cases with reasonable
-frequency for MI software (test cases are categorized as uncommon).
-
-The other apparently neglected document is the requirements specification, which
-records the functionalities, expected performance, goals, context, design
-constraints, external interfaces and other quality attributes of the software
-\citep{IEEE1998}.  If developers write requirements they typically are based on
-a template, which provide documentation structure, guidelines, and rules.
-Although there is no universally accepted template, examples include
-\citet{ESA1991}, \citet{IEEE1998}, \citet{NASA1989}, and
-\citet{RobertsonAndRobertson1999Vol}.
-
-MI software is like other research software in its neglect of requirements
-documentation.  Although requirements documentation is recommended by some
-\citep{TobiasEtAl2018, HerouxEtAl2008, SmithAndKoothoor2016}, in practice
-research software developers often do not produce a proper requirements
-specification \citep{HeatonAndCarver2015}. \citet{SandersAndKelly2008}
-interviewed 16 scientists from 10 disciplines and found that none of the
-scientists created requirements specifications, unless regulations in their
-field mandated such a document. \citet{Nguyen-HoanEtAl2010} showed requirements
-are the least commonly produced type of documentation for research software in
-general. When looking at the pain points for research software developers,
-\citet{WieseEtAl2019} found that software requirements and management is the
-software engineering discipline that most hurts scientific developers,
-accounting for 23\% of the technical problems reported by study participants.
-The lack of support for requirements is likely due to the perception that
-up-front requirements are impossible for research software
-\citep{CarverEtAl2007, SegalAndMorris2008}, but if we drop the insistence on
-``up-front'' requirements, allowing instead for the requirements to be written
-iteratively and incrementally, requirements are feasible \citep{Smith2016}.
-\citet{SmithEtAl2007} provides a requirements template tailored to research
-software.
-
-Table~\ref{Tbl_Guidelines} shows several recommended artifacts that are rarely
-observed in practice.  A theme among these rare artifacts is that, except for
-the user-focused getting started manual, they are developer-focused.  The
-dependency list, which is a list of software library dependencies, was rarely
-observed, but this information is still likely present, just embedded in build
-scripts. The other developer-focused and rare artifacts are as follows:
-
-\begin{itemize}
-
-\item \textbf{A Contributing file} provides new contributors with the information
-that they need to start adding/modifying the repository's files.
-\citet{Abdalla2016} provides a simple template for creating an open-source
-contributor guideline.
-
-\item \textbf{A Developer Code of Conduct} explicitly states the expectations
-for developers on how they should treat one another \citep{TouraniEtAl2017}. The
-code outlines rules for communication and establishes enforcement mechanisms for
-violations.  As \citet{TouraniEtAl2017} states, the developer code documents
-the spirit of a community so that anyone can comfortably contribute regardless
-of ethnicity, gender, or sexual orientation. Three popular codes of conduct are
-\citep{TouraniEtAl2017}:
-\href{https://www.contributor-covenant.org/version/2/1/code_of_conduct/}
-{Contributor Covenant}, \href{https://ubuntu.com/community/code-of-conduct}
-{Ubuntu Code of Conduct}, and \href{https://www.djangoproject.com/conduct/}
-{Django Code of Conduct}. A code of conduct can improve inclusivity, which
-brings the benefit of a wider pool of contributors.  For example, a code of
-conduct can improve the participation of women \citep{SinghEtAl2021}. A standard
-of ethical behaviour can be captured in the code, for projects that are looking
-to abide by a code of ethics, such as the IEEE Code of Ethics \citep{IEEE1999},
-or the Professional Engineers of Ontario code of ethics \citep[p.\
-23--24]{PEO2021}.
-
-\item \textbf{Code Style Guidelines} present standards for writing code. Style
-guides codify such elements as formatting, commenting, naming identifiers, best
-practices and dangers to avoid \citep{Carty2020}. For instance, most coding
-style guides will specify using ALLCAPS when naming symbolic constants.
-Understandability improves under standardization, since developers spend
-more time on the content of the code, and less time distracted by its style.
-Three sample style guides are:
-\href{https://google.github.io/styleguide/javaguide.html} {Google Java Style
-Guide},
-\href{http://cnl.sogang.ac.kr/cnlab/lectures/programming/python/PEP8_Style_Guide.pdf}
-{PEP8 Style Guide for Python}, and
-\href{https://google.github.io/styleguide/cppguide.html} {Google \CC Style
-Guide}.  Linting tools, like \href{https://pypi.org/project/flake8/}{flake8} for
-Python, can be used to enforced coding styles, like the PEP8 standard.
-
-\item \textbf{A Product Roadmap} explains the vision and direction for a product
-offering \citep{MunchEtAl2019}.  Although they have different forms, all
-roadmaps cover the following: \begin{inparaenum}[i)]
-	\item Where are we now?,
-	\item Where are we going?, and
-	\item How can we get there? \end{inparaenum} \citep{PhaalEtAl2005}. As
-stated by \citet{Pichler2012}, a product roadmap provides the following
-benefits: continuity of purpose, facilitation of collaboration, and assistance
-with prioritization. Creating a roadmap involves the following steps:
-\begin{inparaenum}[i)]
-	\item define and outline a strategic mission and product vision, 
-	\item scan the environment, 
-	\item revise and distill the product vision to write the product roadmap, and 
-	\item estimate the product life cycle and evaluate the mix of planned
-development efforts \end{inparaenum} \citep{VahaniittyEtAl2002}.
-
-\item \textbf{Design Documentation} explicitly states the design goals and
-priorities, records the likely changes, decomposes the complex system into
-separate modules, and specifies the relationship between the modules. Design
-documentation shows the syntax of the interface for the modules, and in some
-cases also documents the semantics.  Some potential elements of design
-documentation include the following:
-
-\begin{itemize}
-	\item Representation of the system design and class design using Unified
-	Modelling Language class diagrams. This approach is suited to
-	object-oriented design and designs that use patterns \citep{Gamma1995}.
-	\item Rigorous documentation of the system design following the template for
-	a Module Guide (MG) \citep{ParnasEtAl1984}.  An MG organizes the modules in
-	a hierarchy by their secrets.
-	\item An explanation of the design using data flow diagrams to show typical use cases
-	for input transformation.
-	\item A table or graph showing the traceability between the requirements and
-	the modules (or classes)
-	\item The syntax of the modules (or classes) by providing lists of the state
-	variables, exported constants and all exported access programs for each
-	module (or class).  This shows the interface that can used to access each
-	module's services.
-	\item A formal specification of the semantics of input/output relationships
-	and state transitions for each module using a Module Interface Specification
-	(MIS) \citep{HoffmanAndStrooper1995}. An MIS is an abstract model that
-	formally shows each module's access programs and the associated transitions
-	and outputs based on their state, environment, and input variables.
-	\citet{ElSheikhEtAl2004, SmithAndYu2009} show the example of an MIS for a
-	mesh generator.
-\end{itemize}
-
-\item \textbf{API Documentation} shows developers the services or data provided
-by the software application (or library) through such resources as its methods
-or objects \citep{MengEtAl2018}.  Understandability is improved by API
-documentation \citep{MengEtAl2018}. API documentation can be generated using
-tools like Doxygen, pydoc, and javadoc.
-
-\end{itemize}
-
-The rare artifacts for MI software are similar to the rare artifacts for Lattice
-Boltzmann solvers \citep{Michalski2021}, except LBM software is more likely to
-have developer related artifacts, like Contributing, Dependency list, and Design
-documentation.
-
-To improve MI software in the future, an increased use of checklists could help.
-Developers can use checklists to ensure they follow best practices.  Some
-examples include checklists merging branches into master \citep{Brown2015},
-checklists for saving and sharing changes to the project \citep{WilsonEtAl2016},
-checklists for new and departing team members \citep{HerouxAndBernholdt2018},
-checklists for processes related to commits and releases \citep{HerouxEtAl2008}
-and checklists for overall software quality \citep{ThielEtAl2020, SSI2022}.  For
-instance, for Lattice Boltzmann solver software, ESPResSo has a checklist for
-managing releases \citep{Michalski2021}.
-
-The above discussion shows that, taken together, MI projects fall somewhat short
-of recommended best practices for research software.  However, MI software is
-not alone in this.  Many, if not most, research projects fall short of best
-practices.  A gap exists in research software development practices and software
-engineering recommendations \citep{Storer2017, Kelly2007, OwojaiyeEtAl2021_CSE}.
-\citet{JohansonAndHasselbring2018} observe that the state-of-the-practice for
-research software in industry and academia does not incorporate state-of-the-art
-SE tools and methods.  This causes sustainability and reliability problems
-\citep{FaulkEtAl2009}. Rather than benefit from capturing and reusing previous
-knowledge, projects waste time and energy ``reinventing the wheel''
-\citep{deSouzaEtAl2019}.
 
@@
 
-\section{Comparison of Principles, Process, and Methodologies to Research Software in General} \label{Sec_CompareMethodologies}
+\section[Comparison to Other Research Software]{Comparison of Principles, Process, and
+Methodologies to Research Software in General} \label{Sec_CompareMethodologies}
 
@@
 determination.  To the list of pain points not discussed for MI, our study of
-LBM software \citep{SmithEtAl2022} adds lack of software experience for the
+LBM software \citep{SmithEtAl2024} adds lack of software experience for the
 developers, technical debt, and documentation. We did not observe any pain
@@
 \citep{PintoEtAl2018, PintoEtAl2016, WieseEtAl2019}. Our study of LBM software
-\citep{SmithEtAl2022} also highlighted lack of time as a significant pain point.
+\citep{SmithEtAl2024} also highlighted lack of time as a significant pain point.
 
@@
 Funding challenges have also been noted by others \citep{GewaltigAndCannon2012,
-Goble2014, KaterbowAndFeulner2018, SmithEtAl2022}. Researchers that devote time
+Goble2014, KaterbowAndFeulner2018, SmithEtAl2024}. Researchers that devote time
 to software have the additional challenge that funding agencies do not always
@@
 cross-platform compatibility (also mentioned by \citet{PintoEtAl2018}), CI,
-hardware issues and operating system issues. From \citep{SmithEtAl2022}
+hardware issues and operating system issues. From \citep{SmithEtAl2024}
 technology pain points for LBM developers include setting up parallelization and
@@
 challenges of insufficient testing and insufficient understanding of testing
-concepts \citep{SmithEtAl2022}. Automated testing is a specific challenge for
+concepts \citep{SmithEtAl2024}. Automated testing is a specific challenge for
 LBM software since free testing services do not offer adequate facilities for
-large amounts of data \citep{SmithEtAl2022}. Although not specifically mentioned
+large amounts of data \citep{SmithEtAl2024}. Although not specifically mentioned
 during our interviews, the large data sets for MI likely also cause a challenge
@@
 software in general also mentioned that users do not always have the expertise
-required to install or use the software. \citet{SmithEtAl2022} observes a
+required to install or use the software. \citet{SmithEtAl2024} observes a
 similar pattern for LBM software, with several LBM developers noting that users
@@
 past developers have had with accumulating too much technical debt
-\citep{KruchtenEtAl2012}.  For LBM software, \citet{SmithEtAl2022} identifies
+\citep{KruchtenEtAl2012}.  For LBM software, \citet{SmithEtAl2024} identifies
 technical debt as one of the developer pain points.
@@
 In our ``state of the practice'' assessment exercise for LBM software
-\citep{SmithEtAl2022}, we noticed that LBM developers implicitly used
+\citep{SmithEtAl2024}, we noticed that LBM developers implicitly used
 modularization based on the principle of design for change to improve
@@
 
-Specific examples of design for change for LBM software \citep{SmithEtAl2022}
+Specific examples of design for change for LBM software \citep{SmithEtAl2024}
 include the following:
@@
 model will be an easy contribution'', but changes to the ``Cartesian mesh … will
-be a nightmare'' \citep{SmithEtAl2022}. The design of TCLB highlights that not
+be a nightmare'' \citep{SmithEtAl2024}. The design of TCLB highlights that not
 every conceivable change needs to be supported, only the likely changes.  
@@
 \centering
-\includegraphics[width=1.0\textwidth]{./figures/TopGoal.pdf}
+\includegraphics[width=1.0\textwidth]{TopGoal.pdf}
 \caption{Top Goal of the assurance case and its sub-goals}
@@
 identified similar threats to validity in our assessment of the state of the
-practice of Lattice Boltzmann Solvers \citep{SmithEtAl2022}.
+practice of Lattice Boltzmann Solvers \citep{SmithEtAl2024}.
 
@@
 In \citet{SmithEtAl2016} we reduced concern over the reliability risk associated
-with the reviewer's judgement  % BLIND REVIEW
-by demonstrating that the measurement process is reasonably reproducible.  In
-\citet{SmithEtAl2016} we graded five software products by two reviewers. Their
-rankings were almost identical. As long as each grader uses consistent
-definitions, the relative comparisons in the AHP results will be consistent
-between graders.
+with the reviewer's judgement by demonstrating that the measurement process is
+reasonably reproducible.  In \citet{SmithEtAl2016} we graded five software
+products by two reviewers. Their rankings were almost identical. As long as each
+grader uses consistent definitions, the relative comparisons in the AHP results
+will be consistent between graders.
 
@@
 Although there are differences, like the importance of data privacy for MI data,
-we found the approach to developing LBM software \citep{SmithEtAl2022} and MI
+we found the approach to developing LBM software \citep{SmithEtAl2024} and MI
 software to be similar.  Except for the domain specific aspects, we believe that
@@
 
-% software packages not in the top two for anything
-% , , ,  , , , ,
-% ,   , , , ,
-% Papaya,  ,  , , 
-% MatrixUser, , , , , MRIcroGL, , , , , , ,
-
-% Installability: 3D Slicer, BioImage Suite Web, Slice:Drop, INVESALIUS,
-% Correctness (first appearance): OHIF Viewer, ImageJ, medInria,
-% Reliability (first appearance): SMILI, Fiji, 
-% Maintainability (first appearance): Weasis, ParaView
-% Reusability (first appearance): dwv, GATE
-
-% Robustness: XMedCon, OsiriX Lite, MicroView, ITK-SNAP, Horos, Gwyddion,
-% dicompyler, DicomBrowser, AMIDE, 3DimViewer, DICOM Viewer
-
-% Usability: Ginkgo CADx
-
-% Understandability: Drishti
-
 \begin{table}[ht!]
@@
 
-%% The Appendices part is started with the command \appendix;
-%% appendix sections are then done as normal sections
-%% \appendix
-
-%% \section{}
-%% \label{}
-
-%% If you have bibdatabase file and want bibtex to generate the
-%% bibitems, please use
-%%
-%%  \bibliographystyle{elsarticle-harv} 
-%%  \bibliography{<your bibdatabase>}
-
-%% else use the following coding to input the bibitems directly in the
-%% TeX file.
-
-%% Bibliography
 \bibliographystyle{ACM-Reference-Format}
-\bibliography{MedImageSoft_SOP}
-
-% \begin{thebibliography}{00}
-
-% %% \bibitem[Author(year)]{label}
-% %% Text of bibliographic item
-
-% \bibitem[ ()]{}
-
-% \end{thebibliography}
+\bibliography{PainPoints_OS_MI}
 
 \end{document}
-
-\endinput
-%%
-%% End of file `elsarticle-template-harv.tex'.
-
-%%%%%%%%%%%

### 2024-05-27 StateOfPractice/Papers/SOP_MI_OSR/SOP_MI_OSR.tex
Split versions of SOP papers

--- 2022-11-02 StateOfPractice/Papers/MedImageSoft_SOP-arXiv/MedImageSoft_SOP.tex
@@
 
-\begin{frontmatter}
-
-%% Title, authors and addresses
-
-%% use the tnoteref command within \title for footnotes;
-%% use the tnotetext command for theassociated footnote;
-%% use the fnref command within \author or \affiliation for footnotes;
-%% use the fntext command for theassociated footnote;
-%% use the corref command within \author for corresponding author footnotes;
-%% use the cortext command for theassociated footnote;
-%% use the ead command for the email address,
-%% and the form \ead[url] for the home page:
-%% \title{Title\tnoteref{label1}}
-%% \tnotetext[label1]{}
-%% \author{Name\corref{cor1}\fnref{label2}}
-%% \ead{email address}
-%% \ead[url]{home page}
-%% \fntext[label2]{}
-%% \cortext[cor1]{}
-%% \affiliation{organization={},
-%%            addressline={}, 
-%%            city={},
-%%            postcode={}, 
-%%            state={},
-%%            country={}}
-%% \fntext[label3]{}
-
-\title{State of the Practice for Medical Imaging Software}
-
-%% use optional labels to link authors explicitly to addresses:
-%% \author[label1,label2]{}
-%% \affiliation[label1]{organization={},
-%%             addressline={},
-%%             city={},
-%%             postcode={},
-%%             state={},
-%%             country={}}
-%%
-%% \affiliation[label2]{organization={},
-%%             addressline={},
-%%             city={},
-%%             postcode={},
-%%             state={},
-%%             country={}}
-
-\author[CAS]{Spencer Smith}
-\author[CAS]{Ao Dong}
-\author[CAS]{Jacques Carette}
-\author[ECE]{Mike Noseworthy}
-
-\affiliation[CAS]{organization={McMaster University, Computing and Software Department},%Department and Organization
-            addressline={1280 Main Street West}, 
-            city={Hamilton},
-            postcode={L8S 4K1}, 
-            state={Ontario},
-            country={Canada}}
-
-\affiliation[ECE]{organization={McMaster University, Electrical Engineering},%Department and Organization
-            addressline={1280 Main Street West}, 
-            city={Hamilton},
-            postcode={L8S 4K1}, 
-            state={Ontario},
-            country={Canada}}
+\title{State of the Practice for Medical Imaging Software Based on Open Source Repositories}
+
+\author[1,*]{W.\ Spencer Smith}
+\author[1]{Ao Dong}
+\author[1]{Jacques Carette}
+\author[2]{Michael D.\ Noseworthy}
+
+\affil[1]{McMaster University, Computing and Software Department, {Canada}}
+\affil[2]{McMaster University, Electrical \& Computer Engineering Department,
+{Canada}}
+\affil[*]{Corresponding Author}
+
+\maketitle
+
+% RQ1 to RQ4
+% Some of Section 1 (Introduction)
+% Sections 2 to 6 (includes looking at artifacts observed in repos)
+% Some of Sections 11 to 13 (Threats to Validity, Future Work, Conclusions)
 
@@
 
-We present the state of the practice for Medical Imaging (MI) software. We
-selected 29 medical imaging projects from 48 candidates, assessed 10 software
-qualities (installability, correctness/ verifiability, reliability, robustness,
-usability, maintainability, reusability, understandability,
-visibility/transparency and reproducibility) by answering 108 questions for each
-software project, and interviewed 8 of the 29 development teams. Based on the
-quantitative data for the first 9 qualities, we ranked the MI software with the
-Analytic Hierarchy Process (AHP). The four top ranked software products are:
-\textit{3D Slicer}, \textit{ImageJ}, \textit{Fiji}, and \textit{OHIF Viewer}.
-Our ranking is mostly consistent with the community's ranking, with four of our
-top five projects also appearing in the top five of a list ordered by
-stars-per-year. Generally, MI software is in a healthy state as shown by the
-following: in the repositories we observed 88\% of the documentation artifacts
-recommended by research software development guidelines, 100\% of MI projects
-use version control tools, and developers appear to use the common quasi-agile
-research software development process. However, the current state of the
-practice deviates from the existing guidelines because of the rarity of some
-recommended artifacts (like test plans, requirements specification, code of
-conduct, code style guidelines, product roadmaps, and Application Program
-Interface (API) documentation), low usage of continuous integration (17\% of the
-projects), low use of unit testing (about 50\% of projects), and room for
-improvement with documentation (six of nine developers felt their documentation
-wasn't clear enough). From interviewing the developers, we identified five pain
-points and two qualities of potential concern: lack of development time, lack of
-funding, technology hurdles, ensuring correctness, usability, maintainability,
-and reproducibility. The interviewees proposed strategies to improve the state
-of the practice, to address the identified pain points, and to improve software
-quality. Combining their ideas with ours, we have the following list of
-recommendations: increase documentation, increase testing by enriching datasets,
-increase continuous integration usage, move to web applications, employ linters,
-use peer reviews, design for change, add assurance cases, and incorporate a
-``Generate All Things'' approach.
+We present the state of the practice for Medical Imaging (MI) software based on
+data available in open source repositories. We selected 29 medical imaging
+projects from 48 candidates and assessed 10 software qualities (installability,
+correctness/ verifiability, reliability, robustness, usability, maintainability,
+reusability, understandability, visibility/transparency and reproducibility) by
+answering 108 questions for each software project. Based on the quantitative
+data for the first 9 qualities, we ranked the MI software with the Analytic
+Hierarchy Process (AHP). The four top ranked software products are \textit{3D
+Slicer}, \textit{ImageJ}, \textit{Fiji}, and \textit{OHIF Viewer}.  Our ranking
+is mostly consistent with the community's ranking, with four of our top five
+projects also appearing in the top five of a list ordered by stars-per-year.  We
+observed 88\% of the documentation artifacts recommended by research software
+development guidelines.  However, the current state of the practice deviates
+from the existing guidelines because of the rarity of some recommended artifacts
+(like test plans, requirements specification, code of conduct, code style
+guidelines, product roadmaps, and Application Program Interface (API)
+documentation). Open source MI software projects may wish to consider increasing
+their documentation.
 
@@
 We aim to study the state of software development practice for Medical Imaging
-(MI) software.  MI tools use images of the interior of the body (from sources
-such as Magnetic Resonance Imaging (MRI), Computed Tomography (CT), Positron
-Emission Tomography (PET) and Ultrasound) to provide information for diagnostic,
-analytic, and medical applications \citep{FDA2021, enwiki:1034887445,
-Zhang2008}.  Figure~\ref{Fig_Example}, which shows an image of the brain,
-highlights the importance and value of MI. Through MI medical practitioners and
-researchers can noninvasively gain insights into the human body, including
-information on injuries and illnesses. Given the importance of MI software and
-the high number of competing software projects, we wish to understand the merits
-and drawbacks of the current development processes, tools, and methodologies. We
-aim to assess through a software engineering lens the quality of the existing
-software with the hope of highlighting standout examples, understanding current
-pain points and providing guidelines and recommendations for future development.
+(MI) software using data available in open source repositories.  MI tools use
+images of the interior of the body (from sources such as Magnetic Resonance
+Imaging (MRI), Computed Tomography (CT), Positron Emission Tomography (PET) and
+Ultrasound) to provide information for diagnostic, analytic, and medical
+applications \citep{FDA2021, enwiki:1034887445, Zhang2008}.
+Figure~\ref{Fig_Example}, which shows an image of the brain, highlights the
+importance and value of MI. Through MI medical practitioners and researchers can
+noninvasively gain insights into the human body, including information on
+injuries and illnesses. Given the importance of MI software and the high number
+of competing software projects, we wish to understand the merits and drawbacks
+of the current development processes, tools, and methodologies. We aim to assess
+through a software engineering lens the quality of the existing software with
+the hope of highlighting standout examples, understanding current pain points
+and providing guidelines and recommendations for future development.
 
@@
 \centering
-\includegraphics[scale=0.5]{figures/PrimaryLanguages.pdf}
+\includegraphics[scale=0.5]{PrimaryLanguages.pdf}
 \centering
@@
 \begin{figure}[!ht]
-\includegraphics[scale=0.47]{figures/installability_scores.pdf}
+\includegraphics[scale=0.47]{installability_scores.pdf}
 \caption{AHP installability scores}
@@
 \begin{figure}[!ht]
-\includegraphics[scale=0.47]{figures/correctness_verifiability_scores.pdf}
+\includegraphics[scale=0.47]{correctness_verifiability_scores.pdf}
 \caption{AHP correctness \& verifiability scores}
@@
 \begin{figure}[!ht]
-\includegraphics[scale=0.47]{figures/reliability_scores.pdf}
+\includegraphics[scale=0.47]{reliability_scores.pdf}
 \caption{AHP surface reliability scores}
@@
 \begin{figure}[!ht]
-\includegraphics[scale=0.47]{figures/robustness_scores.pdf}
+\includegraphics[scale=0.47]{robustness_scores.pdf}
 \caption{AHP surface robustness scores}
@@
 \begin{figure}[!ht]
-\includegraphics[scale=0.47]{figures/usability_scores.pdf}
+\includegraphics[scale=0.47]{usability_scores.pdf}
 \caption{AHP surface usability scores}
@@
 \begin{figure}[!ht]
-\includegraphics[scale=0.47]{figures/maintainability_scores.pdf}
+\includegraphics[scale=0.47]{maintainability_scores.pdf}
 \caption{AHP maintainability scores}
@@
 \begin{figure}[!ht]
-\includegraphics[scale=0.47]{figures/reusability_scores.pdf}
+\includegraphics[scale=0.47]{reusability_scores.pdf}
 \caption{AHP reusability scores}
@@
 \begin{figure}[!ht]
-\includegraphics[scale=0.47]{figures/understandability_scores.pdf}
+\includegraphics[scale=0.47]{understandability_scores.pdf}
 \caption{AHP surface understandability scores}
@@
 \begin{figure}[!ht]
-\includegraphics[scale=0.47]{figures/visibility_transparency_scores.pdf}
+\includegraphics[scale=0.47]{visibility_transparency_scores.pdf}
 \caption{AHP visibility/transparency scores}
@@
 \begin{figure}[!ht]
-\includegraphics[scale=0.47]{figures/overall_scores.pdf}
+\includegraphics[scale=0.47]{overall_scores.pdf}
 \caption{Overall AHP scores with an equal weighting for all 9 software qualities}
@@
 
-\begin{table}[!ht]
+\begin{table}[H]
 \begin{center}
@@
 
-\section{Comparison of Tool Usage Between MI and Other Research Software}
-\label{Sec_CompareTools}
-
-Developers use software tools to support the development, verification,
-maintenance, and evolution of software, software processes, and artifacts
-\citep[p.\ 501]{GhezziEtAl2003}. MI software uses tools for CI/CD, user support,
-version control, documentation, and project management.  To answer
-\rqref{RQ_CompareToolsProjMngmnt} we summarize the tool usage in these
-categories, and compare this to the usage by the research software community.
-
-Table~\ref{tab_user_support_model} summarizes the user support models by the
-number of projects using each model (projects may use more than one support
-model). We do not know whether the prevalent use of GitHub issues for user
-support is by design, or whether this just naturally happens as users seek
-help. The common use of GitHub by MI developers is not surprising, given that
-GitHub is the largest code host in the world, with over 128 million public
-repositories and over 23 million users (as of roughly February 26, 2020)
-\citep{Kashyap2020}.
-
-\begin{table}[!ht]
-\centering
-\begin{tabular}{lc}
-\toprule
-\multicolumn{1}{c}{User Support Model} & Num.\ Projects \\
-\midrule
-GitHub issue & 24 \\
-Frequently Asked Questions (FAQ) & 12 \\
-Forum & 10 \\
-E-mail address & 9 \\
-GitLab issue, SourceForge discussions & 2 \\
-Troubleshooting & 2 \\
-Contact form & 1 \\ 
-\bottomrule
-\end{tabular}
-\caption{\label{tab_user_support_model}User support models by number of projects}
-\end{table}
-
-From Section~\ref{sec_score_maintainability}, 27 of the 29 projects used git as
-the version control tool, one used Mercurial and one used Subversion.  The
-hosting is on GitHub for 24 packages, SourceForge for three and BitBucket for
-two.  Although teams may have a process for accepting new contributions, no one
-discussed this during their interviews. However, most teams (eight of nine)
-mentioned using GitHub and pull requests to manage contributions from the
-community. The interviewees generally gave very positive feedback on using
-GitHub. Some teams previously used a different approach to version control and
-eventually transferred to git and GitHub.  The past approaches included
-contributions from e-mail (three teams), contributions from forums (one team)
-and e-mailing the git repository back and forth between developers (one team).
-
-The common use of version control for MI software illustrates considerable
-improvement from the poor adoption of version control tools that Wilson lamented
-in 2006 \citep{Wilson2006}.  The proliferation of version control tools for MI
-matches the increase in the broader research software community.  A little over
-10 years ago \citet{Nguyen-HoanEtAl2010} estimated that only 50\% of research
-software projects use version control, but even at that time
-\citet{Nguyen-HoanEtAl2010} noted an increase from previous usage levels. A
-survey in 2018 shows 81\% of developers use a version control system
-\citep{AlNoamanyAndBorghi2018}. \citet{Smith2018} has similar results, showing
-version control usage for alive projects in mesh generation, geographic
-information systems and statistical software for psychiatry increasing from
-75\%, 89\% and 17\% (respectively) to 100\%, 95\% and 100\% (respectively) over
-a four-year period ending in 2018. (For completeness the same study showed a
-small decrease in version control usage for seismology software over the same
-time period, from 41\% down to 36\%).  A recent survey by \citet{CarverEtAl2022}
-shows version control use among practitioners at over 95\%, with 83/87 survey
-respondents indicating that they use it. All but one of the software guides
-cited in Section~\ref{Sec_CompareArtifacts} includes the advice to use version
-control. (The USGS guide \citep{USGS2019} was the only set of recommendations to
-not mention version control.) The high usage of version control tools in MI
-software matches the trend in research software in general.
-
-As mentioned in Section~\ref{sec_result_correctness_verifiability}, we
-identified five projects using CI/CD tools (about 17\% of the assessed
-projects). We found which projects used CI/CD by examining the documentation and
-source code of all projects. The count of CI/CD usage may actually be higher,
-since traces of CI/CD usage may not always appear in a repository.  This was the
-case for a study of LBM software, where interviews with developers showed that
-more projects used CI/CD than was evident from repository artifacts alone
-\citep{Michalski2021}.  The 17\% utilization for MI software contrasts with the
-high frequency with which research software development guidelines recommend
-continuous integration \citep{BrettEtAl2021, Brown2015, ThielEtAl2020,
-Zadka2018, vanGompelEtAl2016}. Although there is currently little data available
-on CI/CD utilization for research software, our impression is that CI/CD is not
-yet common practice, despite its recommendation.  This is certainly the case for
-LBM software, where usage numbers are similar to MI software, with only 12.5\%
-of 24 LBM packages showing evidence of CI/CD in their repositories
-\citep{Michalski2021}.  The survey of \citet{CarverEtAl2022} suggests higher use
-of CI/CD with 54\% (54/100) of respondents indicating that they use it. However,
-that survey measures something different from the current one by surveying
-practitioners, rather than assessing projects.  Additional information on CI/CD
-is given in the recommendations (Section~\ref{Sec_ContinuousIntegration}).
-
-For documentation tools and methods mentioned by the interviewees, the most
-popular (mentioned by about 30\% of developers) were forum discussions and
-videos.  The second most popular options (mentioned by about 20\% of developers)
-were GitHub, wiki pages, workshops, and social media. The least frequently
-mentioned options (about 10\% of developers) included writing books, and google
-forms.  In contrasting MI software with LBM software, the most significant
-documentation tool difference is that LBM software often uses document
-generation tools, like doxygen and sphinx \citep{Michalski2021}, while MI does
-not appear to use these tools.
-
-Some interviewees mentioned the project management tools they used. Generally
-speaking, the interviewees talked about two types of tools:
-\begin{inparaenum}[i)]
-\item trackers, including GitHub, issue trackers, bug trackers and Jira; and,
-\item documentation tools, including GitHub, Wiki page, Google Doc, and
-Confluence.
-\end{inparaenum}
-Of the specifically named tools in the above lists, interviewees mentioned
-GitHub 3 times, and each of the other tools once each.
-
-Based on information provided by \citet{JungEtAl2022}, tool utilization for MI
-software has much in common with tool utilization for ocean modelling software.
-Both use tools for editing, compiling, code management, testing, building, and
-project management.  From the data available, ocean modelling differs from MI
-software in its use of Kanban boards for project management.
-
-\section{Comparison of Principles, Process, and Methodologies to Research Software in General} \label{Sec_CompareMethodologies}
-
-We answer research question \rqref{RQ_CompareMethodologies} by comparing the
-principles, processes, and methodologies used for MI software to what can be
-gleaned from the literature on research software in general. In our interviews
-with developers the responses about development model were vague, with only two
-interviewees following a definite development model. In some cases the
-interviewees felt their process was similar to an existing development model.
-Three teams (about 38\%) either followed agile, or something similar to agile.
-Two teams (25\%) either followed a waterfall process, or something similar.
-Three teams (about 38\%) explicitly stated that their process was undefined or
-self-directed.
-
-Our observations of an informally defined process, with elements of agile
-methods, matches what has been observed for research software in general.
-Scientific developers naturally use an agile philosophy \citep{AckroydEtAl2008,
-CarverEtAl2007, EasterbrookAndJohns2009, Segal2005, HeatonAndCarver2015}, or an
-amethododical process \citep{Kelly2013}, or a knowledge acquisition driven
-process \citep{Kelly2015}.  A waterfall-like process can work for research
-software \citep{Smith2016}, especially if the developers work iteratively and
-incrementally, but externally document their work as if they followed a
-rationale design process \citep{parnas1986rational}.
-
-No interviewee introduced any strictly defined project management process. The
-most common approach was following the issues, such as bugs and feature
-requests. Additionally, the \textit{3D Slicer} team had weekly meetings to
-discuss the goals for the project; the \textit{INVESALIUS 3} team relied on the
-GitHub process for their project management; the \textit{ITK-SNAP} team had a
-fixed six-month release pace; only the interviewee from the \textit{OHIF} team
-mentioned that the team has a project manager; the \textit{3D Slicer} team and
-\textit{BioImage Suite Web} team do nightly builds and tests. The \textit{OHIF}
-developer believes that a better project management process can improve junior
-developer efficiency while also improving internal and external communication.
-
-We identified the use of unit testing in less than half of the 29 projects. On
-the other hand, the interviewees believed that testing (including usability
-tests with users) was the top solution to improve correctness, usability, and
-reproducibility.  This level of testing matches what was observed for LBM
-software \citep{Michalski2021} and is apparently greater than the level of
-testing for ocean modelling software.  \citet{JungEtAl2022} reports that ocean
-modellers underemphasize testing.
-
-As the observed artifacts in Table~\ref{artifactspresent} show, none of the 29
-projects emphasize documentation. None of them had theory manuals, although we
-did identify a road map in the \textit{3D Slicer} project.  We did not find
-requirements specifications. Table~\ref{tab_opinion_doc} summarizes
-interviewees' opinions on documentation. Interviewees from each of the eight
-projects thought that documentation was essential to their projects, and most of
-them said that it could save their time to answer questions from users and
-developers. Most of them saw the need to improve their documentation, and only
-three of them thought that their documentations conveyed information clearly
-enough. Nearly half of developers also believed that the lack of time prevented
-them from improving documentation.
-
-\begin{table}[!ht]
-\centering
-\begin{tabular}{ll}
-\toprule
-Opinion on Documentation & Num.\ Ans. \\ 
-\midrule
-Documentation is vital to the project & 8 \\
-Documentation of the project needs improvements & 7 \\
-Referring to documentation saves time to answer questions & 6 \\
-Lack of time to maintain good documentation & 4 \\
-Documentation of the project conveys information clearly & 3 \\
-Coding is more fun than documentation & 2 \\
-Users help each other by referring to documentation & 1 \\ 
-\bottomrule
-\end{tabular}
-\caption{Opinions on documentation by the numbers of interviewees with the
-answers}
-\label{tab_opinion_doc}
-\end{table}
-
-As Table~\ref{Tbl_Guidelines} suggests, an emphasis on documentation, especially
-for new developers, is echoed in research software guidelines. Multiple
-guidelines recommend a document explaining how to contribute to a project, often
-named CONTRIBUTING. Guidelines also recommend tutorials, user guides and quick
-start examples. \citet{SmithAndRoscoe2018} suggests including instructions
-specifically for on-boarding new developers. For open-source software in general
-(not just research software), \citet{Fogel2005} recommends providing tutorial
-style examples, developer guidelines, demos, and screenshots.
-
-\section{Developer Pain Points} \label{painpoints}
-
-Based on interviews with nine developers (described in
-Section~\ref{sec_interview_methods}), we answer three research questions (first
-mentioned in Section~\ref{sec_motivation}): \rqref{RQ_PainPoints}) What are the
-pain points for developers working on research software projects?;
-\rqref{RQ_ComparePainPoints}) How do the pain points of developers from MI
-compare to the pain points for research software in general?; and
-\rqref{RQ_Concerns}) For MI developers what specific best practices are taken to
-address the pain points and software quality concerns?
-
-Our interviews identified pain points related to a lack of time and funding,
-technology hurdles, improving correctness, and improving usability.  In this
-section, we go through each pain point and contrast the MI experience with
-observations from other domains.  We also cover potential ways to address the
-pain points, as promoted by the community.  (Later, in
-Section~\ref{ch_recommendations}, we propose additional pain mitigation
-strategies based on our experience.)  In addition to pain points, we summarize
-MI developer strategies for improving maintainability and reproducibility.
-Although the interviewees did not explicitly identify these two qualities as
-pain points, we did discuss threats to these qualities and ways to improve them
-as part of our interview process \citep{SmithEtAl2021}.  The interviewee's
-practices for addressing pain points and improving quality can potentially be
-emulated by other MI developers. Moreover, these practices may provide examples
-that can be followed by other research software domains.
-
-\citet{PintoEtAl2018} lists some pain points that did not come up in our
-conversations with MI developers: interruptions while coding, scope bloat, lack
-of user feedback, hard to collaborate on software projects, and aloneness.
-\citet{WieseEtAl2019} also mention two research software pain points that did
-not explicitly arise in our interviews: reproducibility, and software scope
-determination.  To the list of pain points not discussed for MI, our study of
-LBM software \citep{SmithEtAl2022} adds lack of software experience for the
-developers, technical debt, and documentation. We did not observe any pain
-points for MI that were not also observed for LBM. From the pain points
-mentioned above, although the topics of reproducibility and technical debt did
-not come up in our MI interviews, we covered these two topics as part of the
-discussion of software qualities, as summarized at the end of this section.
-Although previous studies show pain points that were not mentioned by MI
-developers, we cannot conclude that these pain points are not relevant for MI
-software development, since we only interviewed nine developers for about an
-hour each.
-
-\begin{enumerate}
-
-\item[P\refstepcounter{pnum}\thepnum \label{P_LackDevTime}:] \textbf{Lack of
-Development Time:} Many interviewees thought lack of time, along with lack of
-funding (discussed next), were their most significant obstacles. Other domains
-of research software also experience the lack of time pain point
-\citep{PintoEtAl2018, PintoEtAl2016, WieseEtAl2019}. Our study of LBM software
-\citep{SmithEtAl2022} also highlighted lack of time as a significant pain point.
-
-Potential and proven solutions suggested by the interviewees include:
-
-\begin{itemize}
-\item Shifting from development to maintenance when the team does not have
-enough developers for building new features and fixing bugs at the same time;
-\item Improving documentation to save time answering users' and developers'
-questions;
-\item Supporting third-party plugins and extensions; and,
-\item Using GitHub Actions for CI/CD (Continuous Integration and Continuous
-Delivery.)
-\end{itemize}
-
-\item[P\refstepcounter{pnum}\thepnum \label{P_LackFunding}:] \textbf{Lack of
-Funding:} Developers felt the pain of having to attract funding to develop and
-maintain their software. For instance, the interviewees from \textit{3D Slicer}
-and \textit{OHIF} said getting funding for software maintenance is more
-challenging than finding funding for research. The interviewee from 
-\textit{ITK-SNAP} thought more funding was a way to solve the lack of time
-problem, because they could hire more dedicated developers. On the other hand,
-the interviewee from \textit{Weasis} did not feel that funding could
-solve the same problem, since they would still need time to supervise the project.
-
-Funding challenges have also been noted by others \citep{GewaltigAndCannon2012,
-Goble2014, KaterbowAndFeulner2018, SmithEtAl2022}. Researchers that devote time
-to software have the additional challenge that funding agencies do not always
-count software when they are judging the academic excellence of the applicant.
-\citet{WieseEtAl2019} reported developer pains related to publicity, since
-publishing norms have historically made it difficult to get credit for creating
-software.  As studied by \citet{HowisonAndBullard2016}, research software
-(specifically biology software, but the trend likely applies to other research
-software domains) is infrequently cited. \citet{PintoEtAl2018} also mentions the
-lack of formal reward system for research software.
-
-An interviewee proposed an idea for increasing funding: Licensing the software
-to commercial companies to integrate it into their products.
-
-\item[P\refstepcounter{pnum}\thepnum \label{P_TechnologyHurdles}:]
-\textbf{Technology Hurdles:} The technology hurdles mentioned by MI developers
-include: hard to keep up with changes in OS and libraries, difficult to transfer
-to new technologies, hard to support multiple OSes, and hard to support lower-end
-computers. Developers expressed difficulty balancing between four factors:
-cross-platform compatibility, convenience to development and maintenance,
-performance, and security.
-
-The pain point survey of \citet{WieseEtAl2019} highlights that technology
-hurdles are an issue for research software in general.  Some technical-related
-problems mentioned by \citet{WieseEtAl2019} include dependency management,
-cross-platform compatibility (also mentioned by \citet{PintoEtAl2018}), CI,
-hardware issues and operating system issues. From \citep{SmithEtAl2022}
-technology pain points for LBM developers include setting up parallelization and
-CI.
-
-The solutions proposed by the MI developers include the following:
-
-\begin{itemize}
-\item Adopting a web-based approach with backend servers, to better support
-lower-end computers;
-\item Using memory-mapped files to consume less computer memory, to better
-support lower-end computers; 
-\item Using computing power from the computers GPU for web applications;
-\item Maintaining better documentations to ease the development and maintenance
-processes;
-\item Improving performance via more powerful computers, which one interviewee
-pointed out has already happened.
-\end{itemize}
-
-As the above list shows, developers perceive that web-based applications will
-address the technology hurdle.  Table~\ref{tab_native_vs_web} shows the teams'
-choices between native application and web application. Most of the 29 teams (24
-of 29, or 83\%) chose to develop native applications. For the eight teams we
-interviewed, three of them were building web applications, and the
-\textit{MRIcroGL} team was considering a web-based solution.
-
-\begin{table}[!ht]
-\centering
-\begin{tabular}{lll}
-\toprule
-Software Team & Native Application & Web Application \\ 
-\midrule
-3D Slicer & \checkmark & \\
-INVESALIUS 3 & \checkmark & \\
-dwv & & \checkmark \\
-BioImage Suite Web & & \checkmark \\
-ITK-SNAP & \checkmark & \\
-MRIcroGL & \checkmark & \\
-Weasis & \checkmark & \\
-OHIF & & \checkmark \\ 
-\midrule
-Total number among the eight teams & 5 & 3 \\
-Total number among the 29 teams & 24 & 5 \\ 
-\bottomrule
-\end{tabular}
-\caption{Teams' choices between native application and web application}
-\label{tab_native_vs_web}
-\end{table}
-
-The advantage for native applications is higher performance, while web
-applications have the advantage of cross-platform compatibility and a simpler
-build process.  These web advantages mirror the native disadvantages of
-difficulty with cross-platform compatibility and a complex build process.  The
-lower performance disadvantage of web applications can be improved with a server
-backend, but in this case there are disadvantages for privacy protection and
-server costs.  These issues are discussed further in the recommendations
-(Section~\ref{sec_webapps}).
-
-\item[P\refstepcounter{pnum}\thepnum \label{P_Correctness}:]
-\textbf{Ensuring Correctness:} Interviewees identified multiple threats to
-correctness.  The most frequently mentioned threat was complexity.  Complexity
-enters the software by various means, including the large variety of data formats,
-complicated data standards, differing outputs between medical imaging machines,
-and the addition of (non-viewing related) functionality.  Other threats to
-correctness identified include the following:
-
-\begin{itemize}
-\item Lack of real world image data for testing, in part because of patient
-privacy concerns (\citet{WieseEtAl2019} mentions that the pain point of privacy
-concerns also arises for research software in general);
-\item Tests are expensive and time-consuming because of the need for huge datasets;
-\item Software releases are difficult to manage;
-\item No systematic unit testing; and,
-\item No dedicated quality assurance team.
-\end{itemize}
-
-As implied by the above threats to correctness, testing was the most often
-mentioned strategy for MI developers for ensuring correctness.  Seven teams
-mentioned test related activities, including test-driven development, component
-tests, integration tests, smoke tests, regression tests, self tests and
-automated tests.  With the common emphasis on testing to improve correctness, MI
-software is ahead of some other scientific domains.  For scientific software in
-general \citet{PintoEtAl2018} mention the problem of insufficient testing and
-\citet{HannayEtAl2009} show that more developers think testing is important than
-the number that believe they have a sufficient understanding of testing
-concepts.  Our study of LBM software suggests that this domain shares the
-challenges of insufficient testing and insufficient understanding of testing
-concepts \citep{SmithEtAl2022}. Automated testing is a specific challenge for
-LBM software since free testing services do not offer adequate facilities for
-large amounts of data \citep{SmithEtAl2022}. Although not specifically mentioned
-during our interviews, the large data sets for MI likely also cause a challenge
-for using free testing services, like GitHub Actions.
-
-Research software in general often struggles with the oracle problem for testing
-because for many potential test cases the developer doesn't have a means to
-judge the correctness of their calculated solutions \citep{HannayEtAl2009,
-KanewalaAndBieman2013, KellyEtAl2011, WieseEtAl2019}.  The MI developers did not
-allude to this challenge, likely because for a give image (test case) it is
-possible to determine, potentially by using other software, the expected
-analysis results.
-
-A frequently cited strategy for building confidence in correctness (mentioned by
-3 interviewees) is a two state development process with stable releases and
-nightly builds.  Other strategies for ensuring correctness that came up during
-the interviews include CI/CD, using de-identified copies of medical images for
-debugging, sending beta versions to medical workers who can access the data to
-do the tests, and collecting/maintaining a dataset of problematic images.  Some
-additional strategies used by MI developers include:
-
-\begin{itemize}
-\item Using open datasets.
-\item If (part of) the team belongs to a medical school or a hospital, using the
-datasets they can access;
-\item If the team has access to MRI scanners, self-building sample images for
-testing;
-\item If the team has connections with MI equipment manufacturers, asking for
-their help on data format problems;
-\end{itemize}
-
-The feedback from the interviewees makes it clear that increased connections
-between the development team and medical professionals/institutions could ease
-the pain of ensuring correctness via testing.
-
-\item[P\refstepcounter{pnum}\thepnum \label{P_Usability}:]
-\textbf{Usability:}
-
-The discussion with the developers focused on usability issues for two classes
-of users: the end users and other developers.  The threats to usability for end
-users include an unintuitive user interface, inadequate feedback from the
-interface (such as lack of a progress bar), users being unable to determine the purpose of
-the software, not all users knowing if the software includes certain features, not
-all users understanding how to use the command line tool, and not all users
-understanding that the software is a web application. For developers the threats to
-usability include not being able to find clear instructions on how to deploy the
-software, and the architecture being difficult for new developers to understand.
-
-At least to some extent the problems for MI software users are due to holes in
-their background knowledge.  The survey of \citet{WieseEtAl2019} for research
-software in general also mentioned that users do not always have the expertise
-required to install or use the software. \citet{SmithEtAl2022} observes a
-similar pattern for LBM software, with several LBM developers noting that users
-sometimes try to use incorrect method combinations. Furthermore, some LBM users
-think that the packages will work out of the box to solve their cases, while in
-reality computational fluid dynamics knowledge needs to be applied to correctly
-modify the packages for a new endeavour.
-
-To improve the usability of MI software, the most common strategies mentioned by
-developers are as follows:
-
-\begin{itemize}
-    \item Use documentation (user manuals, mailing lists, forums) (mentioned by
-    4 developers)
-    \item Usability tests and interviews with end users; and, (mentioned by 3
-    developers)
-    \item Adjusting the software according to user feedback. (mentioned by 3
-    developers)
-\end{itemize}
-
-Other suggested and practiced strategies include a graphical user interface,
-testing every release with active users, making simple things simple and
-complicated things possible, focusing on limited number of functions, icons with
-clear visual expressions, designing the software to be intuitive, having a UX
-(User eXperience) designer, dialog windows for important notifications,
-providing an example for users to follow, downsampling images to consume less
-memory, and providing an option to load only part of the data to boost
-performance.  The last two points recognize that an important component of
-usability is performance, since poor performance frustrates users.
-
-\end{enumerate}
-
-Up to this point, we have covered the pain points that came up in interviews
-with MI developers, along with a summary of the techniques that are currently
-used to address these pain points.  Although the developers did not explicitly
-identify the qualities of maintainability and reproducibility as pain points in
-our interviews, as part of our interview questions
-(Section~\ref{sec_interview_methods}) they did share their approaches for
-improving these qualities, as discussed below.
-
-\begin{enumerate}
-\item[Q\refstepcounter{qnum}\theqnum \label{Q_Maintainability}:]
-\textbf{Maintainability:} \citet{Nguyen-HoanEtAl2010} rate maintainability as the
-third most important software quality for research software in general. The push
-for sustainable software \citep{deSouzaEtAl2019} is motivated by the pain that
-past developers have had with accumulating too much technical debt
-\citep{KruchtenEtAl2012}.  For LBM software, \citet{SmithEtAl2022} identifies
-technical debt as one of the developer pain points.
-
-To improve maintainability, the most popular (with five out of nine interviewees
-mentioning it) strategy is to use a modular approach, with often repeated
-functions in a library.  Other strategies that were mentioned for improving
-maintainability include supporting third-party extensions, an easy-to-understand
-architecture, a dedicated architect, starting from simple solutions, and
-documentation.  The \textit{3D Slicer} team used a well-defined structure for
-the software, which they named as an ``event-driven MVC pattern''. Moreover,
-\textit{3D Slicer} discovers and loads necessary modules at runtime, according
-to the configuration and installed extensions. The \textit{BioImage Suite Web}
-team had designed and re-designed their software multiple times in the last 10+
-years. They found that their modular approach effectively supports
-maintainability \citep{Joshi2011}.
-
-\item[Q\refstepcounter{qnum}\theqnum \label{Q_Reproducibility}:]
-\textbf{Reproducibility:}  Although the MI developers did not mention
-reproducibility explicitly as a pain point, they did mention the need to improve
-documentation.  Good documentation does not just address the pain points of lack
-of developer time (\ppref{P_LackDevTime}), technology hurdles
-(\ppref{P_TechnologyHurdles}), usability \ppref{P_Usability}, and
-maintainability.  Documentation is also necessary for reproducibility. The
-challenges of inadequate documentation are a known problem for research software
-\citep{PintoEtAl2018, WieseEtAl2019} and for non-research software
-\citep{LethbridgeEtAl2003}.
-
-In our interviews, we discussed threats to reproducibility and strategies for
-improving it.  The threats that were mentioned include closed-source software,
-no user interaction tests, no unit tests, the need to change versions of some
-common libraries, variability between CPUs, and misinterpretation of how
-manufacturers create medical images.
-
-The most commonly cited (by 6 teams) strategy to improve reproducibility was
-testing (regression tests, unit tests, having good tests). The second most
-common strategy (mentioned by 5 teams) is making code, data, and documentation
-available, possibly by creating open-source libraries.  Other ideas that were
-mentioned include running the same tests on all platforms, a dockerized version
-of the software to insulate it from the OS environment, using standard
-libraries, monitoring the upgrades of the library dependencies, clearly
-documenting the version information, bringing along the exact versions of all
-the dependencies with the software, providing checksums of the data, and
-benchmarking the software against other software that overlaps in functionality.
-Specifically one interviewee suggested using \textit{3D Slicer} as the benchmark
-to test their reproducibility.
-
-\end{enumerate}
-
-\section{Recommendations} \label{ch_recommendations}
-
-In this section we provide recommendations to address the pain points from
-Section~\ref{painpoints} to answer~\rqref{RQ_Recommend}.  Our recommendations
-are not lists of criticisms for what should have been done in the past, or what
-should be done now; they are suggestions for consideration in the future. We
-expand on some of the ideas that came out of our interviews with developers
-(Section~\ref{painpoints}), including continuous integration, moving to web
-applications, and enriching the test data sets. We also bring in new ideas from
-our experience like employing linters, peer review, design for change and
-assurance cases.  Our aim is to mention ideas that are at least somewhat beyond
-conventional best practices. The ideas listed here have the potential to become
-best practices in the medium to long-term. We list the ideas roughly in the
-order of increasing implementation effort.
-
-\subsection{Use Continuous Integration} \label{Sec_ContinuousIntegration}
-
-Continuous integration involves frequent pushes to a code repository.  With
-every push the software is built and tested \citep[p.\ 13]
-{HumbleAndFarley2010}, \citep{ShahinEtAl2017, Fowler2006}.  CI can take
-significant time and effort to set up and integrate into a team's workflow, but
-the benefits are significant, as follows:
-
-\begin{itemize}
-	\item Elimination of headaches associated with a separate integration phase
-	\citep{Fowler2006}, \citep[p.\ 20]{HumbleAndFarley2010}. If developers
-	postpone integration, integration problems are inevitable.  Continuous
-	integration means that problems are immediately obvious and the source of
-	the problem can be isolated to the small increment that was just committed.
-	\item Detection and removal of bugs \citep{Fowler2006} via
-	automated testing.  To improve productivity, defects are best discovered and
-	fixed at the point where they are introduced \citep[p.\
-	23]{HumbleAndFarley2010}.  Code is not the only source of errors; they are
-	also found in the files and scripts related to configuration management
-	\citep[p.\ 18]{HumbleAndFarley2010}.
-	\item Everyone is always working on a stable base, since the rejection of
-	inadequate commits means that the main branch will always be working.  A
-	stable base will always pass all tests.  If the CI system uses generators
-	and linters, it will also have current documentation and standard compliant
-	code.  A stable base improves developer productivity, allowing them to focus
-	on coding, testing, and documentation.
-\end{itemize}
-
-CI consists of the following elements:
-
-\begin{itemize}
-	\item A version control system \citep{Fowler2006}. To be effective, all
-	files should be under version control, not just code files.  Anything that
-	is needed to build, install and run the software should be under version
-	control, including configuration files, build scripts, test harnesses, and
-	operating system configuration files \citep[p. 19]{HumbleAndFarley2010}.
-	Fortunately for the MI, as shown in Section~\ref{sec_score_maintainability}
-	all our measured projects use version control.
-	\item A fully automated build system \citep{Fowler2006}.  As \citet[p.\
-	5]{HumbleAndFarley2010} point out, deploying software manually is an
-	anti-pattern.  For MI software, Table~\ref{artifactspresent} shows 18 of 29
-	packages (62\%) were observed to include build scripts.  Projects without a
-	build system will need to add one to pursue using CI.
-	\item An automated test system \citep{Fowler2006}. Building quality software
-	involves creating automated tests at the unit, component, and acceptance
-	test level, and executing these tests whenever someone makes a change to the
-	code, its configuration, the environment, or the software stack that it runs
-	on \citep[p.\ 83]{HumbleAndFarley2010}. As Table~\ref{artifactspresent}
-	shows, test cases are in the uncommon category for MI software artifacts,
-	which means that some MI projects will need to increase their testing
-	automation if they wish to pursue CI.
-	\item An automated system for other tasks, such as code checking,
-	documentation building and web-site updating.  These other tasks are not
-	essential to CI, but they can be incorporated to improve the quality
-	of the code and the communication between developers and users. For
-	instance, a static analysis (possibly via linters) of the code may find poor
-	programming practice or lack of adherence to adopted coding standards.
-	\item An integrated build system to pull everything together.  Every time
-	there is a check-in (for instance a pull request), the integration server
-	automatically checks out the sources onto the integration machine, starts a
-	build, runs tests, and informs the committer of the results. 
-\end{itemize}
-
-To enable incorporation into a team's workflow, \citet[p.\
-60]{HumbleAndFarley2010} explain that the usual approach for CI is to keep the
-build and test process short. Since MI files are large, the tests run with every
-check-in may need to focus on simple code interface tests, saving large tests
-for less frequent execution.  A more sophisticated option to address the
-bottleneck for merges is CIVET (Continuous Integration, Verification,
-Enhancement, and Testing), which solves this problem by intelligently pinning,
-cancelling, and if necessary, restarting jobs as merges occur
-\citep{SlaughterEtAl2021}. A more sophisticated process management system can
-also enforce rules for pull requests, like checking that a test specification
-includes the test's motivation, a test description, and a design description for
-all changes \citet{SlaughterEtAl2021}.
-
-Setting up a CI system has never been easier than it is today.  A dedicated CI
-server (either physically or virtually) can be installed with tools such as
-\href{https://www.jenkins.io/} {Jenkins}, \href{http://buildbot.net/}
-{Buildbot}, \href{https://www.gocd.org/} {Go}, and
-\href{http://integrity.github.io/} {Integrity}. However, installation on your
-own server is often unnecessary since there are many hosted CI solutions, such
-as: \href{https://travis-ci.org/} {Travis CI},
-\href{https://github.com/features/actions} {GitHub Actions} and
-\href{https://circleci.com/} {CircleCI}.  All that is required to begin using a
-hosted CI is to select the service and then edit a few lines of a YAML
-configuration file in the project's root directory.
-
-\citet{ShahinEtAl2017} highlights the following challenges for adopting CI: lack
-of awareness and transparency, lack of expertise and skills, coordination and
-collaboration challenges, more pressure and workload for team members, general
-resistance to change, scepticism and distrust on continuous practices. The most
-common reason given for not adopting CI is that developers are not familiar
-enough with CI \citep{HiltonEtAl2016}.  \citet{ShahinEtAl2017} observes that
-these problems can be mitigated via improving testing activities, planning and
-documentation, promoting a team mindset, adopting new rules and policies, and
-decomposing development into smaller units.
-
-Continuous integration and delivery helps with addressing several pain points.
-For instance, CI/CD helps reduce development time (\ppref{P_LackDevTime}) by
-removing the need for a time-consuming integration stage and by automating
-regression testing.  Automated regression tests also help with ensuring
-correctness (\ppref{P_Correctness}) and the quality of reproducibility
-(\qref{Q_Reproducibility}).
-
-\subsection{Move To Web Applications} \label{sec_webapps}
-
-Section~\ref{painpoints} describes the pain point of technology hurdles
-(\ppref{P_TechnologyHurdles}), which motivates considering the use of web
-applications. Here we give further advice to help with deciding whether to adopt
-a web application. The decision will be based on whether, on balance, the web
-application improves the four factors identified by developers: compatibility,
-maintainability, performance, and security. To enable decision-making, a team
-will need to prioritize between these factors, based on their objectives and
-experience. The suggestions are intended to provide ideas and avenues for
-exploration; a web application will not be the right fit for all projects and
-all teams.
-
-\begin{itemize}
-
-\item \textbf{Modern technologies may improve frontend performance.} Web
-applications with only a frontend usually perform worse than native
-applications. However, new technologies may ease this difference. For example,
-some JavaScript libraries can help the frontend harness the power of the
-computer's GPU and accelerate graphical computing. In addition, there are new
-frameworks helping developers with cross-platform compatibility. For example,
-the \href{https://flutter.dev/}{Flutter} project enables support for web,
-mobile, and desktop OS with one codebase.  Other options include
-\href{https://vuejs.org/} {Vue}, \href{https://angular.io/} {Angular} and
-\href{https://reactjs.org/} {React}, and \href{https://elm-lang.org/}{Elm}.
-
-\item \textbf{Backend servers can potentially deliver high performance.} Web
-applications with backend servers may perform even better than native
-applications. If a team needs to support lower-end computers, it is good to use
-back-end servers for heavy computing tasks.  For backend servers where traffic
-and latency is not an issue, options include
-\href{https://www.django-rest-framework.org/} {Django}, \href{https://laravel.com/} {Laravel} and
-\href{https://nodejs.org/en/} {Node.js}.  The advantage of Django is that it provides access to Python
-libraries.  For backend servers where traffic and latency is an issue,
-\href{https://github.com/gin-gonic/gin} {Gin} is an option.
-
-\item \textbf{Backend servers can have low costs.} Serverless solutions from
-major cloud service providers (like Amazon Web Services (AWS) and Google Cloud
-Platform) may be worth exploring. Serverless solutions still use a server, but
-the server provider only charges the team when they use the server. The solution
-is event-driven, and costs the team by the number of requests processed. Thus,
-serverless can be very cost-effective for less intensively used functions.
-
-\item \textbf{Web transmission may diminish security.} Transferring sensitive
-data on-line can be a problem for projects requiring high security. Regulations
-for some MI applications may forbid doing web transmissions. In this case, a web
-application with a backend may not be an option.
-
-\end{itemize}
-
-\subsection{Enrich the Testing Datasets} 
-\label{sec_recommendations_testing_dataset}
-
-As described in Section~\ref{painpoints}, ensuring correctness
-(\ppref{P_Correctness}) via testing can be problematic because of limited access
-to real-world medical imaging datasets.  We build on the suggestions we heard
-from our interviewees as follows:
-
-\begin{itemize}
-\item \textbf{Build and maintain good connections to datasets.} A team can build
-connections with professionals working in the medical domain, who may have
-access to private datasets and can perform tests for the team. If a team has
-such professionals as internal members, the process can be simplified.
-
-\item \textbf{Collect and maintain datasets over time.} A team may face problems
-caused by various unique inputs over the years of software development. This
-data should be collected and maintained over time to form a good, comprehensive,
-dataset for testing.
-
-\item \textbf{Search for open data sources.} In general, there are many open MI
-datasets.  For instance, there are
-\href{https://nihcc.app.box.com/v/ChestXray-NIHCC}{Chest X-ray Datasets} by
-National Institute of Health \citep{WangEtAl2017},
-\href{https://www.cancerimagingarchive.net/}{Cancer Imaging Archive}
-\citep{PriorEtAl2017}, \href{https://medpix.nlm.nih.gov/home}{MedPix} by
-National Library of Medicine \citep{Smirniotopoulos2014}, and datasets for liver
-\citep{BilicEtAl2019} and brain \citep{MenzeEtAl2015} tumor segmentation
-benchmarks.  A team developing MI software should be able to find more open
-datasets according to their needs.
-
-\item \textbf{Create sample data for testing.} If a team can access tools
-creating sample data, they may also self-build datasets for testing. For
-example, an MI software development team can use an MRI scanner to create images
-of objects, animals, and volunteers. The team can build the images based on
-specific testing requirements.
-
-\item \textbf{Remove privacy from sensitive data.} For data with sensitive
-information, a team can ask the data owner to remove such information or add
-noise to protect privacy. One example is using de-identified copies of medical
-images for testing.
-
-\item \textbf{Establish community collaboration in the domain.} During our
-interviews with developers in the MI domain, we heard many stories of asking for
-supports from other professionals or equipment manufacturers. However, we
-believe that broader collaboration between development teams can address this
-problem better. Some datasets are too sensitive to share, but if the community
-has some kind of ``group discussion'', teams can better express their needs, and
-professionals can better offer voluntary support for testing. Ultimately, the
-community can establish a nonprofit organization as a third party, which
-maintains large datasets, tests Open Source Software (OSS) in the domain, and
-protects privacy.
-
-\end{itemize}
-
-\subsection{Employ Linters} \label{Sec_Linters}
-
-A linter is a tool that statically analyzes code to find programming errors,
-suspicious constructs, and stylistic inconsistencies \citep{Wikipedia2022_Lint}.
-Linters can be used as an ad hoc check for code files, but they really come into
-their own when used as part of a CI system, as discussed in
-Section~\ref{Sec_ContinuousIntegration}. Almost none of the research software
-guidelines that we consulted, summarized in Section~\ref{Sec_CompareArtifacts},
-mention linters.  The one exception is \citet{ThielEtAl2020}.  Despite the lack
-of mention in the guidelines, we believe that linters have the potential to
-improve code quality at a relatively low cost.
-
-Linters have the following benefits: finding potential bugs, finding memory
-leaks, improving performance, standardizing code with respect to formatting,
-removing silly errors before code reviews, and catching potential security
-issues \citep{SourceLevel2022_Lint}. Most popular programming languages have an
-accompanying linter.  For example, Python has the options of PyLint, flake8 and
-Black \citep{Zadka2018}.
-
-We recommend the use of linters because they are relatively easy to incorporate
-into a developer's workflow, and they address several MI pain points
-(Section~\ref{painpoints}).  For instance, linters address the lack of
-development time (\ppref{P_LackDevTime}) by increasing the developer's
-productive time via guarding against making frustrating, time-consuming, mundane
-mistakes.  Moreover, since a linter can include rules that capture the wisdom of
-senior programmers, it can help newer developers avoid common mistakes. With
-respect to the technology hurdle pain point (\ppref{P_TechnologyHurdles}),
-linters can assist with the move toward web applications
-(Section~\ref{sec_webapps}).  For instance, ESLint in React is a pluggable
-linter that lets the developer know if they have imported something and not used
-it, if a function could be short-handed, if there are indentation
-inconsistencies, etc. \citep{Whitehouse2018}. By insisting on code
-standardization linters can reduce technical debt and thus improve
-maintainability (\qref{Q_Maintainability}). Although linters are tools for code
-analysis, the idea of statically checking for adherence to basic rules can be
-extended to check documentation. \citet{SmithEtAl2018_StatSoft} shows how the
-use of tools to enforce documentation standards partially explains the
-relatively higher quality of statistical tools that are part of the
-Comprehensive R Archive Network (CRAN).
-
-\subsection{Conduct a Mix of Rigorous and Informal Peer Reviews} \label{Sec_PeerReview}
-
-We advocate incorporating peer review into the development process, as
-frequently recommended for research software \citep{HerouxEtAl2008, Givler2020,
-OrvizEtAl2017, USGS2019}. In most cases a modern, lightweight review, should be
-adequate.  Modern code review is informal, tool-based, asynchronous, and focused
-on reviewing code changes \citep{SadowskiEtAl2018}. Managing a project via
-GitHub pull requests is an example of a modern approach to reviewing code.
-Software development organizations have moved to this lightweight style of code
-review because of the inefficiencies of rigorous inspections
-\citep{RigbyAndBird2013}.  However, for important parts of the code, developers
-may benefit from mixing in a more rigorous approach.
-
-\citet{Fagan1976} began work on rigorous review via code inspection.  Elements
-of a typical inspection include reviewing the code against a checklist (checking
-the consistency of variable names, look for terminating loops, etc.), performing
-specific review tasks (such as summarizing the code's purpose, cross-referencing
-the code to the technical manual, creating a data dictionary for a given module,
-etc.) Rigorous inspection finds 60-65\% of latent defects on average, and often
-tops 85\% in defect removal efficiency \citep{Jones2008}. The success rate of
-code inspection is generally higher than most forms of testing, which average
-between 30 --- 35\% for defect removal efficiency \citep{EbertAndJones2009,
-Jones2008}. For research software, \citep{KellyAndShepard2000} show a task based
-inspection approach can be effective. Task based inspection is an ideal fit with
-an issue tracking system, like GitHub.  The review tasks can be issues, so that
-they can be easily assigned, monitored and recorded. Potential issues include
-assigning junior developers to test getting-started tutorial and installation
-instructions.
-
-As indicated in Section~\ref{Sec_CompareMethodologies} some MI projects use
-modern code review, via issue tracking and the use of GitHub.  Those MI projects
-not incorporating modern code review would likely benefit by adopting it.
-Although a rigorous code inspection is likely not worth the required resources,
-for critical parts of the code, developers may want to adopt a more rigorous
-approach. For instance, developers may drop the modern trend of
-asynchronous review and instead occasionally use synchronous review to help
-uncover errors and disseminate best practices throughout the team.  For
-instance, teams could periodically meet, either in-person or virtually, and have
-junior members walk through their code.  In-person reviews will likely help
-realize the benefits of modern code review noticed by
-\citet{BirdAndBacchelli2013}: defect detection, knowledge transfer, increased
-team awareness, and creation of alternative solutions to problems.
-
-Due to improving code quality and increasing knowledge transfer, peer review
-addresses the same pain points and qualities as linters
-(Section~\ref{Sec_Linters}): \ppref{P_LackDevTime}, \ppref{P_TechnologyHurdles},
-and \qref{Q_Maintainability}. Peer review can potentially find misunderstandings
-in how the code implements the required theory, which will improve the
-software's correctness (\ppref{P_Correctness}). The benefits of peer review for
-addressing pain points can be increased by extending the review from just code,
-to also reviewing all software artifacts, including documentation, build
-scripts, test cases and the development process itself.
-
-\subsection{Design For Change} \label{Sec_DesForChange}
-
-In our ``state of the practice'' assessment exercise for LBM software
-\citep{SmithEtAl2022}, we noticed that LBM developers implicitly used
-modularization based on the principle of design for change to improve
-maintainability (\qref{Q_Maintainability}).  We recommend that MI developers use
-the same principle for their modularizations.  Although the advice to modularize
-research software to handle complexity is common \citep{WilsonEtAl2014,
-StewartEtAl2017, Storer2017}, specific guidelines on how to divide the software
-into modules is less prevalent.  Not every decomposition is a good design for
-supporting change, as shown by \citet{Parnas1972a}.  For instance, a design with
-low cohesion and high coupling \citep[p.\ 48]{GhezziEtAl2003} will make change
-difficult. Especially in research software, where change is inevitable,
-designers need to produce a modularization that supports change.
-\citet{JungEtAl2022} points out that ocean modelling software is currently
-feeling the pain of not emphasizing modularization in legacy code.
-
-Specific examples of design for change for LBM software \citep{SmithEtAl2022}
-include the following:
-
-\begin{itemize}
-\item \href{https://github.com/pylbm/pylbm}{pyLBM} has decoupled geometries and
-models of their system using abstraction and modularization of the source code,
-to make it easy to add new features.  The pyLBM design allows for independent
-changes to the geometry and the model.  pyLBM also redeveloped data structures
-to ease future change. 
-\item \href{https://github.com/CFD-GO/TCLB}{TCLB} \citep{rokicki2016adjoint} is
-designed to allow for the addition of some LBM features, but changes to major
-aspects of the system would be difficult. For example, ``implementing a new
-model will be an easy contribution'', but changes to the ``Cartesian mesh … will
-be a nightmare'' \citep{SmithEtAl2022}. The design of TCLB highlights that not
-every conceivable change needs to be supported, only the likely changes.  
-\end{itemize}
-
-As the LBM examples above illustrate, developers can accomplish design for
-change by first identifying likely changes, either implicitly or explicitly, and
-second by hiding each likely change behind a well-defined module interface. This
-approach mirrors the recommendations from \citet{Parnas1972a}.
-Section~\ref{Sec_CompareArtifacts} lists ideas for how to document the design,
-including the likely changes, so that they are more visible to others.
-
-\subsection{Assurance Case} \label{AssuranceCases}
-
-To ensure correctness (\ppref{P_Correctness} and to achieve the quality of
-reproducibility (\qref{Q_Reproducibility}), we recommend considering the use of
-assurance cases.  \citet{RinehartEtAl2015} defines an assurance case as ``[a]
-reasoned and compelling argument, supported by a body of evidence, that a
-system, service, or organization will operate as intended for a defined
-application in a defined environment.''  An assurance cases provide an organized
-and explicit argument that the software and its documentation achieves desired
-qualities, such as correctness and reproducibility.  Although assurance cases
-have been successfully employed for safety critical systems
-\citep{RinehartEtAl2015}, this technique is relatively new for research software
-\citep{SmithEtAl2020_AC, Smith2018}.
-
-One way to present an assurance case is through the Goal Structuring Notation
-(GSN) \citep{Spriggs2012}, which make arguments clear, easy to read and, hence,
-easy to challenge. GSN starts with a Top Goal (Claim), like ``Program X delivers
-correct outputs when used for its intended use/purpose in its intended
-environment.''  We then decompose this top goal into Sub-Goals, which themselves
-may be further decomposed.  The purpose of the decomposition is to take the
-abstract higher level goals and bring them down to something concrete that can
-be proven.  The decomposition ends with the terminal Sub-Goals that are
-supported by Solutions (Evidence). Typical evidence will consist of documents,
-expert reviews, test case results, peer review, etc.  Within the GSN framework,
-there are also strategy blocks, which describe the rationale for decomposing a
-Goal or Sub-Goal into more detailed Sub-Goals. A common tool for creating,
-editing, and presenting, a GSN argument is \href{https://astah.net/} {Astah}.
-
-\citet{SmithEtAl2020_AC} shows the example of arguing for the correctness of the
-Analysis of Functional NeuroImages (AFNI) package 3dfim+ \citep{Ward2000}.
-3dfim+ analyzes the activity of the brain by computing the correlation between
-an ideal signal and the measured brain signal for each voxel. The assurance case
-for the correctness of 3dfim+ has the top level decomposed into four sub-goals,
-as shown in Figure~\ref{TopGoal}.  This example follows the same pattern as used
-for medical devices \citep{Wassyng2015}.  The first sub-goal (GR) argues for the
-quality of the documentation of the requirements.  The second sub-goal (GD) says
-that the design complies with the requirements and the third proposes that the
-implementation also complies with the requirements.  The fourth sub-goal (GA)
-claims that the inputs to 3dfim+ will satisfy the operational assumptions, since
-we need valid input to make an argument for the correctness of the output.
-
-\begin{figure}[!ht]
-\centering
-\includegraphics[width=1.0\textwidth]{./figures/TopGoal.pdf}
-\caption{Top Goal of the assurance case and its sub-goals}
-\label{TopGoal}
-\end{figure}
-
-Preparing an assurance case for the pre-existing 3dfim+ software shows the value
-of an assurance cases for research software. Although \citet{SmithEtAl2020}
-found no errors in the output of the existing software, the rigour of the
-proposed approach did lead to finding ambiguities and omissions in the existing
-documentation, such as missing information on the coordinate system convention.
-In addition, a potential concern for the software itself was identified from the
-GA argument: running the software does not produce any warning about the
-obligation of the user to provide data that matches the parametric statistical
-model employed for the correlation calculations.
-
-\subsection{Generate All Things} \label{Sec_GenAllThings}
-
-To address developer pain points, we propose automatically generating MI code
-and its documentation via a ``Generate All Things'' (GAT) approach. A GAT
-approach uses models to capture knowledge in domains such as physics, computing,
-mathematics, documentation, and certification.  Developers combine and transform
-knowledge via explicit ``recipes'', which weave it together to generate the
-desired code, documentation, test cases, inspection reports and build scripts. A
-recipe can even potentially be written to generate an assurance case
-(Section~\ref{AssuranceCases}). Our definition of GAT implies generation of all
-software artifacts, not just the code. GAT moves development to a higher level
-of abstraction so that domain experts can work without concern for low-level
-implementation details. GAT allows developers to optimally generate code and
-documentation, reduce the likelihood of errors, eliminate redundancy, and
-automate maintenance. With a GAT approach MI developers can experiment with
-different algorithm choices, different input formats, different outputs formats,
-etc.
-
-Part of GAT is code generation.  In the future, some believe that code
-generation will transform coding, documentation, design, and verification
-\citep{JohansonAndHasselbring2018, Smith2018}. GAT removes the distraction of
-writing software, allowing developers to focus on their science. A GAT approach
-removes the maintenance headaches of documentation duplicates and near
-duplicates \citep{LucivEtAl2018}, since developers capture knowledge once and
-transform it as needed.  Code generation has previously been applied to improve
-research software, such as linear algebra software packages like Blitz++
-\citep{Veldhuizen1998}, and ATLAS (Automatically Tuned Linear Algebra Software)
-\citep{WhaleyEtAl2001}.  Software/hardware generation has been applied for
-digital signal processing in Spiral \citep{Pueschel2001}. A generative approach
-has also been used for a family of efficient, type-safe Gaussian elimination
-algorithms \citet{Carette2008}. \citep{LoggEtAl2012} use code generation when
-solving partial differential equations in FEniCS (Finite Element and
-Computational Software). \citet{MatkerimEtAl2013} and \citet{OberEtAl2018} use
-code generation for High Performance Computing (HPC), using UML (Unified Modelling
-Language) for their domain models. \citet{SzymczakEtAl2016} presents initial work on
-the GAT approach, \citet{SmithAndCarette2021-BRIC} presents a motivating example,
-and \citep{CaretteEtAl2021-Drasil} provides a prototype.
-
-A GAT approach addresses multiple pain points.  For example, a generative
-approach can decrease development time (\ppref{P_LackDevTime}) by automation,
-once the necessary infrastructure is in place.  The GAT approach addresses the
-technology related pain point (\ppref{P_TechnologyHurdles}) because technology
-information can be captured in the models and transformed as needed. To ensure
-correctness (\ppref{P_Correctness}), a GAT approach should be correct by
-construction.  If there are mistakes, GAT has the advantage that they are
-propagated throughout the generated artifacts, which greatly increases the
-chance that someone will notice the mistake.  Maintainability
-(\qref{Q_Maintainability}) is addressed because developers write the recipes
-used for generation at a high level making them relatively easy to change.
-Usability (\ppref{P_Usability}) is addressed because of the emphasis on
-generating up-to-date documentation.  GAT facilitates reproducibility
-(\qref{Q_Reproducibility}) because at any time all the code and documentation
-can be regenerated.  The generator can include explicit traceability to show the
-dependence of the software on specific versions of software libraries.
-
-\section{Threats to Validity} \label{sec_threats_to_validity}
-
-Below we categorize and list the threats to validity that we have identified.
-Our categories come from an analysis of software engineering secondary studies
-by \citet{AmpatzoglouEtAl2019}, where a secondary study analyzes the data from a
-set of primary studies.  \citet{AmpatzoglouEtAl2019} is appropriate because a
-common example of a secondary study is a systematic literature review. Our
-methodology is a systematic software review --- the primary studies are the
-software packages, and our work collects and analyzes these primary studies.  We
-identified similar threats to validity in our assessment of the state of the
-practice of Lattice Boltzmann Solvers \citep{SmithEtAl2022}.
-
-\subsection{Reliability}
-
-A study is reliable if repetition of the study by different researchers using
-the original study's methodology would lead to the same results
-\citep{RunesonAndHost2009}. Reliability means that data and analysis are
-independent of the specific researcher(s) doing the study.  For the current
-study the identified reliability related threats are as follows:
-
-\begin{itemize}
-\item One individual does the manual measures for all packages. A different
-evaluator might find different results, due to differences in abilities,
-experiences, and biases.
-\item The manual measurements for the full set of packages took several months.
-Over this time the software repositories may have changed and the reviewer's
-judgement may have drifted.
-\end{itemize}
-
-In \citet{SmithEtAl2016} we reduced concern over the reliability risk associated
-with the reviewer's judgement  % BLIND REVIEW
-by demonstrating that the measurement process is reasonably reproducible.  In
-\citet{SmithEtAl2016} we graded five software products by two reviewers. Their
-rankings were almost identical. As long as each grader uses consistent
-definitions, the relative comparisons in the AHP results will be consistent
-between graders.
-
-\subsection{Construct Validity}
-
-\citet{RunesonAndHost2009} defines construct validity as the adopted
-metrics representing what they are intended to measure. Our construct threats are
-often related to how we assume our measurements influences the various software
-qualities, as summarized in Section~\ref{sec_grading_software}. Specifically,
-our construct validity related threats include the following:
-
-\begin{itemize}
-\item We make indirect measurement of software qualities since meaningful direct
-measures for qualities like maintainability, reusability and verifiability, are
-unavailable.  We follow the usual assumption that developers achieve higher
-quality by following procedures and adhering to standards \citep[p.\
-112]{VanVliet2000}.
-\item As mentioned in Section~\ref{sec_result_installability}, we could not
-install or build \textit{dwv}, \textit{GATE}, and \textit{DICOM Viewer}. We used
-a deployed on-line version for \textit{dwv}, a VM version for \textit{GATE}, but
-no alternative for \textit{DICOM Viewer}. We might underestimate their rank due
-to these technical issues.
-\item Measuring software robustness only involved two pieces of data. This is
-likely part of the reason for limited variation in the robustness scores
-(Figure~\ref{fg_robustness_scores}). We could add more robustness data by
-pushing the software to deal with more unexpected situations, like a broken
-Internet connection, but this would require a larger investment of measurement
-time. 
-\item We may have inaccurately estimated maintainability by assuming a higher
-ratio of comments to source code improves maintainability. Moreover, we assumed
-that maintainability is improved if a high percentage of issues are closed, but
-a project may have a wealth of open issues, and still be maintainable.
-\item We assess reusability by the number of code files and LOC per file. This
-measure is indicative of modularity, but it does not necessarily mean a good
-modularization. The modules may not be general enough to be easily reused, or
-the formatting may be poor, or the understandability of the code may be low.
-\item The understandability measure relies on 10 random source code files, but
-the 10 files will not necessarily be representative. 
-\item As discussed in Section~\ref{Sec_OverallQ}, our overall AHP ranking makes
-the unrealistic assumption of equal weighting.
-\item We approximated popularity by stars and watches
-(Section~\ref{Sec_VsCommunityRanking}), but this assumption may not be valid. 
-\item As mentioned in Section~\ref{sec_interview_methods}, one interviewee was
-too busy to participate in a full interview, so they provided written answers
-instead. Since we did not have the chance to explain our questions or ask them
-follow-up questions, there is a possibility of misinterpretation of the
-questions or answers.
-\item In building Table~\ref{Tbl_Guidelines} some judgement was necessary on our
-part, since not all guidelines use the same names for artifacts that contain
-essentially the same information.
-\end{itemize}
-
-\subsection{Internal Validity} \label{Sec_InternalValidity}
-
-Internal validity means that discovered causal relations are trustworthy and
-cannot be explained by other factors \citep{RunesonAndHost2009}. In our
-methodology the internal validity threats include the following:
-
-\begin{itemize}
-\item In our search for software packages
-(Section~\ref{sec_software_selection}), we may have missed a relevant package.
-\item Our methodology assumes that all relevant software development activities
-will leave a trace in the repositories, but this is not necessarily true. For
-instance, the possibility exists that CI usage was higher than what we observed
-through the artifacts (Section~\ref{Sec_CompareTools}). As another example,
-although we saw little evidence of requirements
-(Section~\ref{Sec_CompareTools}), maybe teams keep this kind of information
-outside their repos, possibly in journal papers or technical reports.
-\item We interviewed a relatively small sample of 8 teams.  Their pain points
-(Section~\ref{painpoints}) may not be representative of the rest of their
-community.
-\end{itemize}
-
-\subsection{External Validity}
-
-If the results of a study can be generalized (applied) to other
-situations/cases, then the study is externally valid \citep{RunesonAndHost2009}.
-We are confident that our search was exhaustive.  We do not believe that we
-missed any highly popular examples.  Therefore, the bulk of our validity
-concerns are internal (Section~\ref{Sec_InternalValidity}).
-However, our hope is that the trends observed, and the lessons learned for MI
-software can be applied to other research software.  With that in mind we
-identified the following threat to external validity:
-
-\begin{itemize}
-\item We cannot generalize our results if the development of MI software is
-fundamentally different from other research software.
-\end{itemize}
-
-Although there are differences, like the importance of data privacy for MI data,
-we found the approach to developing LBM software \citep{SmithEtAl2022} and MI
-software to be similar.  Except for the domain specific aspects, we believe that
-the trends observed in the current study are externally valid for other research
-software.
-
-\section{Future Work} \label{Sec_FutureWork}
-
-The following recommendations for future state of the practice measurement
-exercises, for MI or for other domains, could address some threats to validity
-mentioned above.  Moreover, some ideas may make the data collection more
-efficient.
-
-\begin{itemize}
-    \item We would like to make surface measurements less shallow. For example:
-    \begin{itemize}
-        \item Surface reliability: our current measurement relies on the
-        processes of installation and getting started tutorials. However, not
-        all software needs installation or has a getting started tutorial. We
-        could devise a list of operation steps (with the help of the Domain
-        Expert), perform the same operations with each software, and record any
-        errors.
-        \item Surface robustness: we used damaged images as inputs for this
-        measuring MI software. This process is similar to fuzz testing
-        \citep{enwiki:1039424308}, which is one type of fault injection
-        \citep{enwiki:1039005082}. We may adopt more fault injection methods,
-        and identify tools and libraries to automate this process.
-        \item Surface usability: we can design usability tests and test all
-        software projects with end-users. The end-users can be volunteers and
-        domain experts. Ideas for getting started are available in
-        \citet{SmithEtAl2021}.
-        \item Surface understandability: our current method does not require
-        understanding the source code. As software engineers, perhaps we can
-        select a small module of each project, read the source code and
-        documentation, try to understand the logic, and score the ease of the
-        process.  
-        \item Maintainability: we can add a measure modifiability as part of the
-        measurement of maintainability.  An experiment could be conducted asking
-        participants to make modifications, observing the study subjects during
-        the modifications, testing the resulting software and surveying the
-        participants \citep{SmithEtAl2021}.
-    \end{itemize}
-	\item We can further automate the measurements on the grading template. For
-	example, with automation scripts and the GitHub API, we may save significant
-	time on retrieving the GitHub metrics through a GitHub Metric Collector.
-	This Collector can take GitHub repository links as input, automatically
-	collect metrics from the GitHub API, and record the results.
-	\item We can improve some interview questions. Some examples are:
-	\begin{itemize}
-	    \item In one question we ask, ``Do you think improving this process can
-	    tackle the current problem?''  The problem is that this is a yes-or-no
-	    question, which is not informative. We could change the question to ``By
-	    improving this process, what current problems can be tackled?''; 
-	    \item We can ask for more details about the modular approach, such as
-	    ``What principles did you use to divide code into modules? Can you
-	    describe an example of using your principles?''.
-	\end{itemize}
-	\item We can better organize the interview questions. Since we use audio
-	conversion tools to transcribe the answers, we should make the transcription
-	easier to read. For example, we can order them together for questions about
-	the five software qualities and compose a similar structure for each.
-	\item We can mark the follow-up interview questions with keywords. For
-	example, say ``this is a follow-up question'' every time asking one. Thus,
-	we record this sentence in the transcription, and it will be much easier to
-	distinguish the follow-up questions from the 20 designed questions.
-\end{itemize}
-
 \section{Conclusions} \label{ch_conclusions}
@@
 adoption of best practices for MI software overall.
-
-% software packages not in the top two for anything
-% , , ,  , , , ,
-% ,   , , , ,
-% Papaya,  ,  , , 
-% MatrixUser, , , , , MRIcroGL, , , , , , ,
-
-% Installability: 3D Slicer, BioImage Suite Web, Slice:Drop, INVESALIUS,
-% Correctness (first appearance): OHIF Viewer, ImageJ, medInria,
-% Reliability (first appearance): SMILI, Fiji, 
-% Maintainability (first appearance): Weasis, ParaView
-% Reusability (first appearance): dwv, GATE
-
-% Robustness: XMedCon, OsiriX Lite, MicroView, ITK-SNAP, Horos, Gwyddion,
-% dicompyler, DicomBrowser, AMIDE, 3DimViewer, DICOM Viewer
-
-% Usability: Ginkgo CADx
-
-% Understandability: Drishti
 
@@
 
-%% The Appendices part is started with the command \appendix;
-%% appendix sections are then done as normal sections
-%% \appendix
-
-%% \section{}
-%% \label{}
-
-%% If you have bibdatabase file and want bibtex to generate the
-%% bibitems, please use
-%%
-%%  \bibliographystyle{elsarticle-harv} 
-%%  \bibliography{<your bibdatabase>}
-
-%% else use the following coding to input the bibitems directly in the
-%% TeX file.
-
-%% Bibliography
 \bibliographystyle{ACM-Reference-Format}
-\bibliography{MedImageSoft_SOP}
-
-% \begin{thebibliography}{00}
-
-% %% \bibitem[Author(year)]{label}
-% %% Text of bibliographic item
-
-% \bibitem[ ()]{}
-
-% \end{thebibliography}
+\bibliography{SOP_MI_OSR}
 
 \end{document}
-
-\endinput
-%%
-%% End of file `elsarticle-template-harv.tex'.
-
-%%%%%%%%%%%

### 2024-05-27 StateOfPractice/Papers/SOP_MI_OSR/SOP_MI_OSR.tex
Revision to SOP MI paper

@@
 data available in open source repositories. We selected 29 medical imaging
-projects from 48 candidates and assessed 10 software qualities (installability,
+projects from 48 candidates and assessed 9 software qualities (installability,
 correctness/ verifiability, reliability, robustness, usability, maintainability,
-reusability, understandability, visibility/transparency and reproducibility) by
-answering 108 questions for each software project. Based on the quantitative
-data for the first 9 qualities, we ranked the MI software with the Analytic
-Hierarchy Process (AHP). The four top ranked software products are \textit{3D
-Slicer}, \textit{ImageJ}, \textit{Fiji}, and \textit{OHIF Viewer}.  Our ranking
-is mostly consistent with the community's ranking, with four of our top five
-projects also appearing in the top five of a list ordered by stars-per-year.  We
-observed 88\% of the documentation artifacts recommended by research software
-development guidelines.  However, the current state of the practice deviates
-from the existing guidelines because of the rarity of some recommended artifacts
-(like test plans, requirements specification, code of conduct, code style
-guidelines, product roadmaps, and Application Program Interface (API)
-documentation). Open source MI software projects may wish to consider increasing
-their documentation.
+reusability, understandability, and visibility/transparency) by answering 108
+questions for each software project. Based on the quantitative data, we ranked
+the MI software with the Analytic Hierarchy Process (AHP). The top four software
+products are \textit{3D Slicer}, \textit{ImageJ}, \textit{Fiji}, and
+\textit{OHIF Viewer}.  Our ranking is mostly consistent with the community's
+ranking, with four of our top five projects also appearing in the top five of a
+list ordered by stars-per-year.  We observed 88\% of the documentation artifacts
+recommended by research software development guidelines.  However, the current
+state of the practice deviates from the existing guidelines because of the
+rarity of some recommended artifacts (like test plans, requirements
+specification, code of conduct, code style guidelines, product roadmaps, and
+Application Program Interface (API) documentation).
 
@@
 	medical imaging, research software, software engineering, software
-	quality, analytic hierarchy process, developer interviews
+	quality, analytic hierarchy process
 
@@
 through a software engineering lens the quality of the existing software with
-the hope of highlighting standout examples, understanding current pain points
-and providing guidelines and recommendations for future development.
+the hope of highlighting standout examples, and providing guidelines and
+recommendations for future development.
 
@@
 
-To gain insights, we devised 10 research questions, which can be applied to MI,
+To gain insight, we devised 10 research questions, which can be applied to MI,
 as well as to other domains, of research software \citep{SmithEtAl2021,
@@
 this information in context by contrasting MI software against the trends shown
-by developers in other research software communities.  Our aim is to collect
-enough information to understand the current pain points experienced by the MI
-software development community so that we can make some preliminary
-recommendations for future improvements.
+by developers in other research software communities.
 
@@
 	(Section~\ref{Sec_CompareArtifacts})
-	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareToolsProjMngmnt}:]
-	How do MI projects compare to research software in general with respect to
-	the use of tools (Section~\ref{Sec_CompareTools}) for:
-	\begin{enumerate} 
-		\item [\rqref{RQ_CompareToolsProjMngmnt}.a] development; and,
-		\item [\rqref{RQ_CompareToolsProjMngmnt}.b] project management?
-	\end{enumerate}
-	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareMethodologies}:]
-	How do MI projects compare to research software in general with respect to
-	principles, processes, and methodologies used?
-	(Section~\ref{Sec_CompareMethodologies})
-	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_PainPoints}:] What are
-	the pain points for developers working on MI software projects?
-	(Section~\ref{painpoints})
-	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_ComparePainPoints}:] How
-	do the pain points of developers from MI compare to the pain points
-	for research software in general? (Section~\ref{painpoints})
-	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Concerns}:] For MI
-	developers what specific best practices are taken to address the pain points
-	and software quality concerns? (Section~\ref{painpoints})
-	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Recommend}:]
-	What research software development practice could potentially address the
-	pain point concerns identified in \rqref{RQ_PainPoints}?
-	(Section~\ref{ch_recommendations})
-
 \end{enumerate}
@@
 software with the Analytic Hierarchy Process (AHP) (Section~\ref{ch_background}
-provides details). After this, as another addition to our previous methodology,
-we interview some development teams to further understand the status of their 
-development process.
+provides details).
 
@@
 
-To measure the existing MI software we need two sets of definitions: i) the
-definitions of relevant software license models (Section
-\ref{sec_software_categories}); and, ii) the definitions of the software
+To measure the existing MI software we need the definitions of the software
 qualities that we will be assessing (Section \ref{sec_software_quality}). In our
@@
 Hierarchy Process (Section \ref{sec_AHP}).
-
-\subsection{Software Categories} \label{sec_software_categories}
-
-When assessing software packages, we need to know the software's license.  In
-particular, we need to know whether the source code will be available to us or
-not.  We define three common software categories.  We will only assess software
-that fits under the Open Source Software license.
-
-\begin{itemize}
-
-\item \textbf{Open Source Software (OSS)} For OSS, the source code is openly
-accessible. Users have the right to study, change and distribute it under a
-license granted by the copyright holder. For many OSS projects, the development
-process relies on the collaboration of different contributors worldwide
-\citep{Corbly2014}. Accessible source code usually exposes more ``secrets'' of a
-software project, such as the underlying logic of software functions, how
-developers achieve their works, and the flaws and potential risks in the final
-product. Thus, OSS is suitable for researchers analyzing the qualities of a
-project.
-
-\item \textbf{Freeware} Freeware is software that can be used free of charge.
-Unlike OSS, the authors of do not allow access or modify the source code
-\citep{LINFO2006}. To many end-users, the differences between freeware and OSS
-may not be relevant. However, software developers who wish to modify the source
-code, and researchers looking for insight into software development process may
-find the inaccessible source code a problem.
-
-\item \textbf{Commercial Software} ``Commercial software is software developed
-by a business as part of its business'' \citep{GNU2019}. Typically speaking,
-commercial software requires users to pay to access all of its features,
-excluding access to the source code. However, some commercial software is also
-free of charge \citep{GNU2019}. Based on our experience, most commercial
-software products are not OSS.
-
-\end{itemize}
 
@@
 
-	\item \textbf{Reproducibility} ``A result is said to be reproducible if
-	another researcher can take the original code and input data, execute it,
-	and re-obtain the same result'' \citep{BenureauAndRougier2017}.
 \end{itemize}
@@
 and engages a domain expert partner throughout, as discussed in
-Section~\ref{sec_vet_software_list}.  The four main steps of the methodology
+Section~\ref{sec_vet_software_list}.  The three main steps of the methodology
 are:
@@
 (Section~\ref{sec_grading_software});
-\item Interview developers (Section~\ref{sec_interview_methods});
 \item Answer the research questions (as given in Section~\ref{sec_motivation}).
@@
 \citet{SmithEtAl2021}.  The template provides measures of the qualities listed
-in Section~\ref{sec_software_quality}, except for reproducibility, which is
-assessed through the developer interviews (Section~\ref{sec_interview_methods}).
+in Section~\ref{sec_software_quality}.
 For each software package, we fill in the template questions. To stay within the
@@
 
+% move the definitions here; it seems repetitive the way it is now - maybe
+% remove the definitions?
 \item \textbf{Correctness \& Verifiability} We check each project to identify
@@
 Windows as the host OS.
-
-\subsection{Interview Methods} \label{sec_interview_methods}
-
-The repository-based measurements summarize the information we can collect from
-on-line resources. This information is incomplete because it doesn't generally
-capture the development process, the developer pain points, the perceived
-threats to software quality, and the developers' strategies to address these
-threats.  Therefore, part of our methodology involves interviewing developers.
-
-We based our interviews on a list of 20 questions, which can be found in
-\citet{SmithEtAl2021}. Some questions are about the background of the software,
-the development teams, the interviewees, and how they organize their projects.
-We also ask about the developer's understanding of the users. Some questions
-focus on the current and past difficulties, and the solutions the team has
-found, or plan to try. We also discuss documentation, both with respect to how
-it is currently done, and how it is perceived. A few questions are about
-specific software qualities, such as maintainability, understandability,
-usability, and reproducibility. The interviews are semi-structured based on the
-question list; we ask follow-up questions when necessary.  The interview process
-presented here was approved by the McMaster University Research Ethics Board
-under the application number
-\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/MACREM/Application.pdf}
-{MREB\#: 5219}.
-
-We sent interview requests to all 29 projects using contact information from
-projects websites, code repository, publications, and from biographic pages at
-the teams' institutions.  In the end nine developers from eight of the projects
-agreed to participate: \textit{3D Slicer}, \textit{INVESALIUS 3}, \textit{dwv},
-\textit{BioImage Suite Web}, \textit{ITK-SNAP}, \textit{MRIcroGL},
-\textit{Weasis}, and \textit{OHIF}. We spent about 90 minutes for each
-interview. One participant was too busy to have an interview, so they wrote down
-their answers. In one case two developers from the same project agreed to be
-interviewed. We held the meetings on-line using either Zoom or Teams, which
-facilitated recording and automatic transcription. The full interview answers
-can be found in \citet{Dong2021}.
 
@@
 
+\section{Threats to Validity} \label{sec_threats_to_validity}
+
+Below we categorize and list the threats to validity that we have identified.
+Our categories come from an analysis of software engineering secondary studies
+by \citet{AmpatzoglouEtAl2019}, where a secondary study analyzes the data from a
+set of primary studies.  \citet{AmpatzoglouEtAl2019} is appropriate because a
+common example of a secondary study is a systematic literature review. Our
+methodology is a systematic software review --- the primary studies are the
+software packages, and our work collects and analyzes these primary studies.  We
+identified similar threats to validity in our assessment of the state of the
+practice of Lattice Boltzmann Solvers \citep{SmithEtAl2024}.
+
+\subsection{Reliability}
+
+A study is reliable if repetition of the study by different researchers using
+the original study's methodology would lead to the same results
+\citep{RunesonAndHost2009}. Reliability means that data and analysis are
+independent of the specific researcher(s) doing the study.  For the current
+study the identified reliability related threats are as follows:
+
+\begin{itemize}
+\item One individual does the manual measures for all packages. A different
+evaluator might find different results, due to differences in abilities,
+experiences, and biases.
+\item The manual measurements for the full set of packages took several months.
+Over this time the software repositories may have changed and the reviewer's
+judgement may have drifted.
+\end{itemize}
+
+In \citet{SmithEtAl2016} we reduced concern over the reliability risk associated
+with the reviewer's judgement by demonstrating that the measurement process is
+reasonably reproducible.  In \citet{SmithEtAl2016} we graded five software
+products by two reviewers. Their rankings were almost identical. As long as each
+grader uses consistent definitions, the relative comparisons in the AHP results
+will be consistent between graders.
+
+\subsection{Construct Validity}
+
+\citet{RunesonAndHost2009} defines construct validity as the adopted
+metrics representing what they are intended to measure. Our construct threats are
+often related to how we assume our measurements influences the various software
+qualities, as summarized in Section~\ref{sec_grading_software}. Specifically,
+our construct validity related threats include the following:
+
+\begin{itemize}
+\item We make indirect measurement of software qualities since meaningful direct
+measures for qualities like maintainability, reusability and verifiability, are
+unavailable.  We follow the usual assumption that developers achieve higher
+quality by following procedures and adhering to standards \citep[p.\
+112]{VanVliet2000}.
+\item As mentioned in Section~\ref{sec_result_installability}, we could not
+install or build \textit{dwv}, \textit{GATE}, and \textit{DICOM Viewer}. We used
+a deployed on-line version for \textit{dwv}, a VM version for \textit{GATE}, but
+no alternative for \textit{DICOM Viewer}. We might underestimate their rank due
+to these technical issues.
+\item Measuring software robustness only involved two pieces of data. This is
+likely part of the reason for limited variation in the robustness scores
+(Figure~\ref{fg_robustness_scores}). We could add more robustness data by
+pushing the software to deal with more unexpected situations, like a broken
+Internet connection, but this would require a larger investment of measurement
+time. 
+\item We may have inaccurately estimated maintainability by assuming a higher
+ratio of comments to source code improves maintainability. Moreover, we assumed
+that maintainability is improved if a high percentage of issues are closed, but
+a project may have a wealth of open issues, and still be maintainable.
+\item We assess reusability by the number of code files and LOC per file. This
+measure is indicative of modularity, but it does not necessarily mean a good
+modularization. The modules may not be general enough to be easily reused, or
+the formatting may be poor, or the understandability of the code may be low.
+\item The understandability measure relies on 10 random source code files, but
+the 10 files will not necessarily be representative. 
+\item As discussed in Section~\ref{Sec_OverallQ}, our overall AHP ranking makes
+the unrealistic assumption of equal weighting.
+\item We approximated popularity by stars and watches
+(Section~\ref{Sec_VsCommunityRanking}), but this assumption may not be valid. 
+\item In building Table~\ref{Tbl_Guidelines} some judgement was necessary on our
+part, since not all guidelines use the same names for artifacts that contain
+essentially the same information.
+\end{itemize}
+
+\subsection{Internal Validity} \label{Sec_InternalValidity}
+
+Internal validity means that discovered causal relations are trustworthy and
+cannot be explained by other factors \citep{RunesonAndHost2009}. In our
+methodology the internal validity threats include the following:
+
+\begin{itemize}
+\item In our search for software packages
+(Section~\ref{sec_software_selection}), we may have missed a relevant package.
+\item Our methodology assumes that all relevant software development activities
+will leave a trace in the repositories, but this is not necessarily true. For
+instance, the possibility exists that CI usage was higher than what we observed
+through the artifacts (Section~\ref{Sec_CompareTools}). As another example,
+although we saw little evidence of requirements
+(Section~\ref{Sec_CompareTools}), maybe teams keep this kind of information
+outside their repos, possibly in journal papers or technical reports.
+\end{itemize}
+
+\subsection{External Validity}
+
+If the results of a study can be generalized (applied) to other
+situations/cases, then the study is externally valid \citep{RunesonAndHost2009}.
+We are confident that our search was exhaustive.  We do not believe that we
+missed any highly popular examples.  Therefore, the bulk of our validity
+concerns are internal (Section~\ref{Sec_InternalValidity}).
+However, our hope is that the trends observed, and the lessons learned for MI
+software can be applied to other research software.  With that in mind we
+identified the following threat to external validity:
+
+\begin{itemize}
+\item We cannot generalize our results if the development of MI software is
+fundamentally different from other research software.
+\end{itemize}
+
+Although there are differences, like the importance of data privacy for MI data,
+we found the approach to developing LBM software \citep{SmithEtAl2024} and MI
+software to be similar.  Except for the domain specific aspects, we believe that
+the trends observed in the current study are externally valid for other research
+software.
+
 \section{Conclusions} \label{ch_conclusions}
@@
 We analyzed the state of the practice for the MI domain with the goal of
-understanding current practice, answering our ten research questions
-(Section~\ref{sec_motivation}) and providing recommendations for current and
-future projects.  Our methods in Section~\ref{ch_methods} form a general process
-to evaluate domain-specific software, that we apply to the specific domain of MI
-software. We identified 48 MI software candidates, then, with the help of the
-Domain Expert selected 29 of them to our final list.
+understanding current practice by answering our four research questions
+(Section~\ref{sec_motivation}).  Our methods in Section~\ref{ch_methods} form a
+general process to evaluate domain-specific software, that we apply to the
+specific domain of MI software. We identified 48 MI software candidates, then,
+with the help of the Domain Expert selected 29 of them to our final list.
 
@@
 
-For insight into devising future methods and tools, we interviewed nine
-developers (from eight teams) to learn about their pain points
-(Section~\ref{painpoints}).  We also discussed qualities of potential concern.
-The identified pain points and qualities of concern include:
-
-\begin{description}
-\item [\ppref{P_LackDevTime}] Lack of development time, %P1
-\item [\ppref{P_LackFunding}] Lack of funding, %P2
-\item [\ppref{P_TechnologyHurdles}] Technology hurdles, %P3
-\item [\ppref{P_Correctness}] Ensuring correctness, %P4
-\item [\ppref{P_Usability}] Usability, %P5
-\item [\qref{Q_Maintainability}] Quality of maintainability, and %Q1
-\item [\qref{Q_Reproducibility}] Quality of reproducibility. %Q2
-\end{description}
-
-Despite the pain points, overall MI software is in a healthy state for software
-development practices.  In our survey of the selected projects we observed 88\%
+When it comes to recommended software artifacts, the state of open source MI projects is healthy, with our surveyed examples showing 88\%
 of the documentation artifacts recommended by research software development
-guidelines (Section~\ref{Sec_CompareArtifacts}).  With respect to tools, MI is
-keeping pace with other research software with 100\% of the projects using
-version control (with 93\% specifically using git)
-(Section~\ref{Sec_CompareTools}).  We observed that the MI developers tend to
-follow the typical research software trend of using a quasi-agile software
-development process (Section~\ref{Sec_CompareMethodologies}).
-
-Although the state of the practice for MI software is healthy, we did notice
+guidelines (Section~\ref{Sec_CompareArtifacts}).  However, we did notice
 areas where practice seems to lag behind the research software development
@@
 roadmap, design documentation, and API documentation
-(Section~\ref{Sec_CompareArtifacts}). Although software development tool use
-seems healthy, we found the use of CI/CD behind typical usage rates (17\% of the
-projects used CI/CD) (Section~\ref{Sec_CompareTools}).  With respect to the
-development process, developer identified areas for improvement included testing
-(only 50\% of projects were identified to have unit testing) and documentation
-(only three out of nine developers felt their documentation was clear enough)
-(Section~\ref{Sec_CompareMethodologies}).
-
-Our interviewees proposed strategies to improve the state of the practice, to
-address the identified pain points, and to improve software quality.  To their
-list (Section~\ref{painpoints}) we added some of our own recommended strategies
-(Section~\ref{ch_recommendations}).  Below we summarize the proposed strategies,
-with traceability to where we discuss the strategy, and to the relevant pain
-points.
-
-\begin{enumerate}
-
-\item Increase documentation to address \ppref{P_LackDevTime},
-\ppref{P_TechnologyHurdles}, \ppref{P_Usability}, \qref{Q_Maintainability},
-\qref{Q_Reproducibility} (Section~\ref{painpoints})
-
-\item Increase testing by enriching datasets to address \ppref{P_Correctness},
-\qref{Q_Reproducibility}
-(Section~\ref{painpoints},~\ref{sec_recommendations_testing_dataset})
-
-\item Increase modularity to address \qref{Q_Maintainability} (Section~\ref{painpoints})
-
-\item Use continuous integration to address \ppref{P_LackDevTime},
-\ppref{P_Correctness}, \qref{Q_Reproducibility}
-(Section~\ref{painpoints},~\ref{Sec_ContinuousIntegration})
-
-\item Move to web applications to address \ppref{P_TechnologyHurdles}
-(Section~\ref{painpoints},~\ref{sec_webapps})
-
-\item Employ linters to address \ppref{P_LackDevTime},
-\ppref{P_TechnologyHurdles}, \qref{Q_Maintainability}
-(Section~\ref{Sec_Linters})
-
-\item Peer reviews to address \ppref{P_LackDevTime},
-\ppref{P_TechnologyHurdles}, \ppref{P_Correctness}, \qref{Q_Maintainability}
-(Section~\ref{Sec_PeerReview})
-
-\item Design for change to address \qref{Q_Maintainability} (Section~\ref{Sec_DesForChange})
-
-\item Assurance case to address \ppref{P_Correctness}, \qref{Q_Reproducibility}
-(Section~\ref{AssuranceCases})
-
-\item Generate all things to address \ppref{P_LackDevTime},
-\ppref{P_TechnologyHurdles}, \ppref{P_Correctness}, \ppref{P_Usability},
-\qref{Q_Maintainability} and \qref{Q_Reproducibility}.
-(Section~\ref{Sec_GenAllThings})
-
-\end{enumerate}
+(Section~\ref{Sec_CompareArtifacts}).  Future MI projects may wish to consider incorporating more artifacts to potentially improve software quality.
+
+Next steps.  Mention threats to validity?  Future work on pain points.  Point to
+LBM methodology in that SOP paper?
 

### 2024-05-28 StateOfPractice/Papers/SOP_MI_OSR/SOP_MI_OSR.tex
Some edits to the SOP paper, reorg to match required section titles

@@
 products are \textit{3D Slicer}, \textit{ImageJ}, \textit{Fiji}, and
-\textit{OHIF Viewer}.  Our ranking is mostly consistent with the community's
+\textit{OHIF Viewer}.  Our ranking is consistent with the community's
 ranking, with four of our top five projects also appearing in the top five of a
@@
 Not only do we wish to gain insight into the state of the practice for MI
-software, we also wish to understand the development of research software in
-general. We wish to understand the impact of the often cited gap, or chasm,
+software, we also wish to understand the impact of the often cited gap, or chasm,
 between software engineering and research software \citep{Kelly2007,
@@
 
-\subsection{Methodology Overview}
+\subsection{Methodology}
 
@@
 
-\section{Background} \label{ch_background}
-
-To measure the existing MI software we need the definitions of the software
-qualities that we will be assessing (Section \ref{sec_software_quality}). In our
-assessment we rank the software packages for each quality; therefore, this
-section also provides the background on our ranking process --- the Analytic
-Hierarchy Process (Section \ref{sec_AHP}).
-
-\subsection{Software Quality Definitions} \label{sec_software_quality}
+We developed a methodology for evaluating the state of the practice of research
+software \citep{SmithEtAl2021, SmithAndMichalski2022}.  The methodology can be
+instantiated for a specific domain of scientific software, which in the current
+case is medical imaging software for visualization.  Our methodology involves
+and engages a domain expert partner throughout, as discussed in
+Section~\ref{sec_vet_software_list}.  The three main steps of the methodology
+are:
+
+\begin{enumerate}
+\item Identify list of representative software packages
+(Section~\ref{sec_software_selection});
+\item Measure (or grade) the selected software
+(Section~\ref{sec_grading_software}) on qualities using AHP;
+\end{enumerate}
+
+define software qualities and ahp.
+
+\subsubsection{Interaction With Domain Expert} \label{sec_vet_software_list}
+
+The Domain Expert is an important member of the state of the practice assessment
+team. Pitfalls exist if non-experts attempt to acquire an authoritative list of
+software, or try to definitively rank the software. Non-experts have the problem
+that they can only rely on information available on-line, which has the
+following drawbacks:
+\begin{inparaenum}[i)]
+  \item the on-line resources could have false or inaccurate information; and,
+  \item the on-line resources could leave out relevant information that is so
+in-grained with experts that nobody thinks to explicitly record it.
+\end{inparaenum}
+
+Domain experts may be recruited from academia or industry.  The only
+requirements are knowledge of the domain and a willingness to be engaged in the
+assessment process.  The Domain Expert does not have to be a software developer,
+but they should be a user of domain software.  Given that the domain experts are
+likely to be busy people, the measurement process cannot put too much of a burden
+on their time.  For the current assessment, our Domain Expert (and paper
+co-author) is Dr.\ Michael Noseworthy, Professor of Electrical and Computer
+Engineering at McMaster University, Co-Director of the McMaster School of
+Biomedical Engineering, and Director of Medical Imaging Physics and Engineering
+at St.\ Joseph's Healthcare, Hamilton, Ontario, Canada.
+
+In advance of the first meeting with the Domain Expert, they are asked to
+create a list of top software packages in the domain.  This is done to help
+the expert get in the right mind set in advance of the meeting.  Moreover,
+by doing the exercise in advance, we avoid the potential pitfall of the expert
+approving the discovered list of software without giving it adequate thought.
+
+The Domain Experts are asked to vet the collected data and analysis.  In
+particular, they are asked to vet the proposed list of software packages and the
+AHP ranking.  These interactions can be done either electronically or with
+in-person (or virtual) meetings.
+
+\subsubsection{Grading Software} \label{sec_grading_software}
+
+We grade the selected software using the measurement template summarized in
+\citet{SmithEtAl2021}.  The template provides measures of the qualities listed
+in Section~\ref{sec_software_quality}.
+For each software package, we fill in the template questions. To stay within the
+target of 160 person hours to measure the domain, we allocated between one and
+four hours for each package. Project developers can be contacted for help
+regarding installation, if necessary, but we impose a cap of about two hours on
+the installation process, to keep the overall measurement time feasible.
+Figure~\ref{fg_grading_template_example} shows an excerpt of the spreadsheet.
+The spreadsheet includes a column for each measured software package.
+
+\begin{figure}[!ht]
+\includegraphics[scale=0.66]{template.pdf}
+\caption{Grading template example}
+\label{fg_grading_template_example}
+\end{figure}
+
+The full template consists of 108 questions categorized under 9 qualities.  We
+designed the questions to be unambiguous, quantifiable, and measurable with
+limited time and domain knowledge. We group the measures under headings for each
+quality, and one for summary information. The summary information (shown in
+Figure~\ref{fg_grading_template_example}) is the first section of the template.
+This section summarizes general information, such as the software name, purpose,
+platform, programming language, publications about the software, the first
+release and the most recent change date, website, source code repository of the
+product, number of developers, etc.  We follow the definitions given by
+\citet{GewaltigAndCannon2012} for the software categories.  Public means
+software intended for public use.  Private means software aimed only at a
+specific group, while the concept category is for software written simply to
+demonstrate algorithms or concepts. The three categories of development models
+are (open source, free-ware and commercial) are discussed in
+Section~\ref{sec_software_categories}.  Information in the summary section sets
+the context for the project, but it does not directly affect the grading scores.
+
+For measuring each quality, we ask several questions and the typical answers are
+among the collection of ``yes'', ``no'', ``n/a'', ``unclear'', a number, a
+string, a date, a set of strings, etc. The grader assigns each quality an
+overall score, between 1 and 10, based on all the previous questions.  Several
+of the qualities use the word ``surface''.  This is to highlight that, for these
+qualities in particular, the best that we can do is a shallow measure.  For
+instance, we are not currently doing any experiments to measure usability.
+Instead, we are looking for an indication that the developers considered
+usability.  We do this by looking for cues in the documentation, like a getting
+started manual, a user manual and a statement of expected user characteristics.
+Details on how we measure each quality are provided in \citet{SmithEtAl2024}.
+
+As part of filling in the measurement template, we use freeware tools to collect
+repository related data. \href{https://github.com/tomgi/git_stats}{GitStats}
+\citep{Gieniusz2019} is used to measure the number of binary files as well as
+the number of added and deleted lines in a repository. We also use this tool to
+measure the number of commits over different intervals of time.
+\href{https://github.com/boyter/scc}{Sloc Cloc and Code (scc)}
+\citep{Boyter2021} is used to measure the number of text based files as well as
+the number of total, code, comment, and blank lines in a repository.
+
+Both tools measure the number of text-based files in a git repository and lines
+of text in these files. Based on our experience, most text-based files in a
+repository contain programming source code, and developers use them to compile
+and build software products. A minority of these files are instructions and
+other documents. So we roughly regard the lines of text in text-based files as
+lines of programming code. The two tools usually generate similar but not
+identical results. From our understanding, this minor difference is due to the
+different techniques to detect if a file is text-based or binary.
+
+For projects on GitHub we manually collect additional information, such as the
+numbers of stars, forks, people watching this repository, open pull requests,
+closed pull requests, and the number of months a repository has been on GitHub.
+We need to take care with the project creation date, since a repository can have
+a creation date much earlier than the first day on GitHub.  For example, the
+developers created the git repository for \textit{3D Slicer} in 2002, but did
+not upload a copy of it to GitHub until 2020. Some GitHub data can be found
+using its GitHub Application Program Interface (API) via the following url:
+\textit{https://api.github.com/repos/[owner]/[repository]} where [owner] and
+[repository] are replaced by the repo specific values. The number of months a
+repository has been on GitHub helps us understand the average change of metrics
+over time, like the average new stars per month.
+
+The repository measures help us in many ways. Firstly, they help us get a fast
+and accurate project overview. For example, the number of commits over the last
+12 months shows how active a project has been, and the number of stars and forks
+may reveal its popularity (used to assess \rqref{RQ_CompareHQ2Popular}).
+Secondly, the results may affect our decisions regarding the grading scores for
+some software qualities. For example, if the percentage of comment lines is low,
+we double-check the understandability of the code; if the ratio of open versus
+closed pull requests is high, we pay more attention to maintainability.
+
+As in \citet{SmithEtAl2016}, Virtual machines (VMs) were used to provide an
+optimal testing environment for each package. We used VMs because it is easier
+to start with a fresh environment, without having to worry about existing
+libraries and conflicts. Moreover, when the tests are complete the VM can be
+deleted, without any impact on the host operating system. The most significant
+advantage of using VMs is to level the playing field. Every software install
+starts from a clean slate, which removes ``works-on-my-computer'' errors. When
+filling in the measurement template, the grader notes the details for each VM,
+including hypervisor and operating system version.
+
+When grading the software, we found 27 out of the 29 packages are compatible
+with two or three different OSes, such as Windows, macOS, and Linux, and 5 of
+them are browser-based, making them platform-independent. However, in the
+interest of time, we only performed the measurements for each project by
+installing it on one of the platforms.  When it was an option, we selected
+Windows as the host OS.
+
+\subsubsection{Software Quality Definitions} \label{sec_software_quality}
 
@@
 set of measures.  That is, quality is a collection of different qualities, often
-called ``ilities.''  Below we list the 10 qualities of interest for this study.
-The order of the qualities follows the order used in \citet{GhezziEtAl2003},
-which puts related qualities (like correctness and reliability) together.
-Moreover, the order is roughly the same as the order developers consider
-qualities in practice.
-
-\begin{itemize}
-	\item \textbf{Installability} The effort required for the installation
-    and/or uninstallation of software in a specified environment
-    \citep{ISO/IEC25010, lenhard2013measuring}.
-
-	\item \textbf{Correctness \& Verifiability} A program is correct if it
-    matches its specification \citep[p.\ 17]{GhezziEtAl2003}.  The specification
-    can either be explicitly or implicitly stated.  The related quality of
-    verifiability is the ease with which the software components or the
-    integrated product can be checked to demonstrate its correctness.
-
-	\item \textbf{Reliability} The probability of failure-free operation of a
-	computer program in a specified environment for a specified time
-	\citep{musa1987software}, \citep[p.\ 357]{GhezziEtAl2003}.
-
-	\item \textbf{Robustness} Software possesses the characteristic of
-	robustness if it behaves ``reasonably'' in two situations: i) when it
-	encounters circumstances not anticipated in the requirements specification,
-	and ii) when users violate the assumptions in its requirements specification 
-	\citep[p.\ 19]{GhezziEtAl2003}, \citep{boehm2007software}.
-
-	\item \textbf{Usability} ``The extent to which a product can be used by
-	specified users to achieve specified goals with effectiveness, efficiency,
-	and satisfaction in a specified context of use'' \citep{ISO/TR16982:2002,
-	ISO9241-11:2018}.
-
-	\item \textbf{Maintainability} The effort with which a software system or
-	component can be modified to i) correct faults; ii) improve performance or
-	other attributes; iii) satisfy new requirements
-	\citep{IEEEStdGlossarySET1990, boehm2007software}.
-
-	\item \textbf{Reusability} ``The extent to which a software component can be
-	used with or without adaptation in a problem solution other than the one for
-	which it was originally developed'' \citep{kalagiakos2003non}.
-
-	\item \textbf{Understandability} ``The capability of the software product to
-	enable the user to understand whether the software is suitable, and how it
-	can be used for particular tasks and conditions of use'' \citep{iso2001iec}.
-
-	\item \textbf{Visibility/Transparency} The extent to which all the steps
-	of a software development process and the current status of it are conveyed
-	clearly \citep[p.\ 32]{GhezziEtAl2003}.
-
-\end{itemize}
-
-\subsection{Analytic Hierarchy Process (AHP)} \label{sec_AHP}
+called ``ilities.''  For this study we selected 9 qualities to measure: .....
+With the exception of installability, all the qualities are defined in
+\citet{GhezziEtAl2003}. Installability is defined as the effort required for the
+installation and/or uninstallation of software in a specified environment
+\citep{ISO/IEC25010, lenhard2013measuring}.  Further details on the software
+qualities can be found in \citet{SmithEtAl2024}.
+
+\subsubsection{Analytic Hierarchy Process (AHP)} \label{sec_AHP}
 
@@
 
-\section{Methodology} \label{ch_methods}
-
-We developed a methodology for evaluating the state of the practice of research
-software \citep{SmithEtAl2021, SmithAndMichalski2022}.  The methodology can be
-instantiated for a specific domain of scientific software, which in the current
-case is medical imaging software for visualization.  Our methodology involves
-and engages a domain expert partner throughout, as discussed in
-Section~\ref{sec_vet_software_list}.  The three main steps of the methodology
-are:
-
-\begin{enumerate}
-\item Identify list of representative software packages
-(Section~\ref{sec_software_selection});
-\item Measure (or grade) the selected software
-(Section~\ref{sec_grading_software});
-\item Answer the research questions (as given in Section~\ref{sec_motivation}).
-\end{enumerate}
-
-In the sections below we provide additional detail on the above steps, while
-concurrently giving examples of how we applied the methodology to the MI domain.
-
-\subsection{Interaction With Domain Expert} \label{sec_vet_software_list}
-
-The Domain Expert is an important member of the state of the practice assessment
-team. Pitfalls exist if non-experts attempt to acquire an authoritative list of
-software, or try to definitively rank the software. Non-experts have the problem
-that they can only rely on information available on-line, which has the
-following drawbacks:
-\begin{inparaenum}[i)]
-  \item the on-line resources could have false or inaccurate information; and,
-  \item the on-line resources could leave out relevant information that is so
-in-grained with experts that nobody thinks to explicitly record it.
-\end{inparaenum}
-
-Domain experts may be recruited from academia or industry.  The only
-requirements are knowledge of the domain and a willingness to be engaged in the
-assessment process.  The Domain Expert does not have to be a software developer,
-but they should be a user of domain software.  Given that the domain experts are
-likely to be busy people, the measurement process cannot put too much of a burden
-on their time.  For the current assessment, our Domain Expert (and paper
-co-author) is Dr.\ Michael Noseworthy, Professor of Electrical and Computer
-Engineering at McMaster University, Co-Director of the McMaster School of
-Biomedical Engineering, and Director of Medical Imaging Physics and Engineering
-at St.\ Joseph's Healthcare, Hamilton, Ontario, Canada.
-
-In advance of the first meeting with the Domain Expert, they are asked to
-create a list of top software packages in the domain.  This is done to help
-the expert get in the right mind set in advance of the meeting.  Moreover,
-by doing the exercise in advance, we avoid the potential pitfall of the expert
-approving the discovered list of software without giving it adequate thought.
-
-The Domain Experts are asked to vet the collected data and analysis.  In
-particular, they are asked to vet the proposed list of software packages and the
-AHP ranking.  These interactions can be done either electronically or with
-in-person (or virtual) meetings.
-
-\subsection{List of Representative Software} \label{sec_software_selection}
-
-We have a two-step process for selecting software packages: i) identify software
-candidates in the chosen domain; and, ii) filter the list to remove less
-relevant members \citep{SmithEtAl2021}.
+\section{Review} \label{ch_results}
+
+In this section we review the selected software packages.  We used a two-step
+process for selecting the software packages: i) identify software candidates in
+the chosen domain; and, ii) filter the list to remove less relevant members
+\citep{SmithEtAl2021}.
 
@@
 visualization.  The Domain Expert agreed with our final choice of 29 packages.
-
-\subsection{Grading Software} \label{sec_grading_software}
-
-We grade the selected software using the measurement template summarized in
-\citet{SmithEtAl2021}.  The template provides measures of the qualities listed
-in Section~\ref{sec_software_quality}.
-For each software package, we fill in the template questions. To stay within the
-target of 160 person hours to measure the domain, we allocated between one and
-four hours for each package. Project developers can be contacted for help
-regarding installation, if necessary, but we impose a cap of about two hours on
-the installation process, to keep the overall measurement time feasible.
-Figure~\ref{fg_grading_template_example} shows an excerpt of the spreadsheet.
-The spreadsheet includes a column for each measured software package.
-
-\begin{figure}[!ht]
-\includegraphics[scale=0.66]{template.pdf}
-\caption{Grading template example}
-\label{fg_grading_template_example}
-\end{figure}
-
-The full template consists of 108 questions categorized under 9 qualities.  We
-designed the questions to be unambiguous, quantifiable, and measurable with
-limited time and domain knowledge. We group the measures under headings for each
-quality, and one for summary information. The summary information (shown in
-Figure~\ref{fg_grading_template_example}) is the first section of the template.
-This section summarizes general information, such as the software name, purpose,
-platform, programming language, publications about the software, the first
-release and the most recent change date, website, source code repository of the
-product, number of developers, etc.  We follow the definitions given by
-\citet{GewaltigAndCannon2012} for the software categories.  Public means
-software intended for public use.  Private means software aimed only at a
-specific group, while the concept category is for software written simply to
-demonstrate algorithms or concepts. The three categories of development models
-are (open source, free-ware and commercial) are discussed in
-Section~\ref{sec_software_categories}.  Information in the summary section sets
-the context for the project, but it does not directly affect the grading scores.
-
-For measuring each quality, we ask several questions and the typical answers are
-among the collection of ``yes'', ``no'', ``n/a'', ``unclear'', a number, a
-string, a date, a set of strings, etc. The grader assigns each quality an
-overall score, between 1 and 10, based on all the previous questions.  Several
-of the qualities use the word ``surface''.  This is to highlight that, for these
-qualities in particular, the best that we can do is a shallow measure.  For
-instance, we are not currently doing any experiments to measure usability.
-Instead, we are looking for an indication that the developers considered
-usability.  We do this by looking for cues in the documentation, like a getting
-started manual, a user manual and a statement of expected user characteristics.
-Below is a summary of how we assess adoption of best practices by measuring each
-quality.
-
-\begin{itemize}
-
-\item \textbf{Installability} We assess the following: 
-\begin{inparaenum}[i)]
-    \item existence and quality of installation instructions;
-    \item the quality of the user experience via the ease of following
-    instructions, number of steps, automation tools; and,
-    \item whether there is a means to verify the installation.
-\end{inparaenum}
-If any problem interrupts the process of installation or uninstallation, we give
-a lower score. We also record the Operating System (OS) used for the
-installation test.
-
-% move the definitions here; it seems repetitive the way it is now - maybe
-% remove the definitions?
-\item \textbf{Correctness \& Verifiability} We check each project to identify
-any techniques used to ensure this quality, such as literate programming,
-automated testing, symbolic execution, model checking, unit tests, etc. We also
-examine whether the projects use Continuous Integration and Continuous Delivery
-(CI/CD). For verifiability, we go through the documents of the projects to check
-for the presence of requirements specifications, theory manuals, and getting
-started tutorials. If a getting started tutorial exists and provides expected
-results, we follow it to check the correctness of the output.
-
-\item \textbf{Surface Reliability} We check the following: 
-\begin{inparaenum}[i)]
-    \item whether the software breaks during installation;
-    \item the operation of the software following the getting started tutorial
-    (if present);
-    \item whether the error messages are descriptive; and,
-    \item whether we can recover the process after an error.
-\end{inparaenum}
-
-\item \textbf{Surface Robustness} We check how the software handles
-unexpected/unanticipated input. For example, we prepare broken image files for
-MI software packages that load image files. We use a text file (.txt) with a
-modified extension name (.dcm) as an unexpected/unanticipated input. We load a
-few correct input files to ensure the function is working correctly before
-testing the unexpected/unanticipated ones.
-
-\item \textbf{Surface Usability} We examine the project's documentation,
-checking for the presence of getting started tutorials and/or a user manual. We
-also check whether users have channels to request support, such as an e-mail
-address, or issue tracker. Our impressions of usability are based on our
-interaction with the software during testing.  In general, an easy-to-use
-graphical user interface will score high.
-
-\item \textbf{Maintainability} We believe that the artifacts of a project,
-including source code, documents, and building scripts, significantly influence
-its maintainability. Thus, we check each project for the presence of such
-artifacts as API documentation, bug tracker information, release notes, test
-cases, and build scripts. We also check for the use of tools supporting issue
-tracking and version control, the percentages of closed issues, and the
-proportion of comment lines in the code.
-
-\item \textbf{Reusability} We count the total number of code files for each
-project. Projects with numerous components potentially provide more choices for
-reuse. Furthermore, well-modularized code, which tends to have smaller parts in
-separate files, is typically easier to reuse. Thus, we assume that projects with
-more code files and fewer Lines of Code (LOC) per file are more reusable. We also
-consider projects with API documentation as delivering better reusability.
-
-\item \textbf{Surface Understandability} Given that time is a constraint, we
-cannot look at all code files for each project; therefore, we randomly examine
-10 code files for their understandability. We check the code's style within each
-file, such as whether the identifiers, parameters, indentation, and formatting
-are consistent, whether the constants (other than 0 and 1) are not hardcoded, and
-whether the code is modularized. We also check the descriptive information for
-the code, such as documents mentioning the coding standard, the comments in the
-code, and the descriptions or links for details on algorithms in the code.
-
-\item \textbf{Visibility/Transparency} To measure this quality, we check the
-existing documents to find whether the software development process and
-current status of a project are visible and transparent. We examine the
-development process, current status, development environment, and release notes
-for each project.
-\end{itemize}
-
-As part of filling in the measurement template, we use freeware tools to collect
-repository related data. \href{https://github.com/tomgi/git_stats}{GitStats}
-\citep{Gieniusz2019} is used to measure the number of binary files as well as
-the number of added and deleted lines in a repository. We also use this tool to
-measure the number of commits over different intervals of time.
-\href{https://github.com/boyter/scc}{Sloc Cloc and Code (scc)}
-\citep{Boyter2021} is used to measure the number of text based files as well as
-the number of total, code, comment, and blank lines in a repository.
-
-Both tools measure the number of text-based files in a git repository and lines
-of text in these files. Based on our experience, most text-based files in a
-repository contain programming source code, and developers use them to compile
-and build software products. A minority of these files are instructions and
-other documents. So we roughly regard the lines of text in text-based files as
-lines of programming code. The two tools usually generate similar but not
-identical results. From our understanding, this minor difference is due to the
-different techniques to detect if a file is text-based or binary.
-
-For projects on GitHub we manually collect additional information, such as the
-numbers of stars, forks, people watching this repository, open pull requests,
-closed pull requests, and the number of months a repository has been on GitHub.
-We need to take care with the project creation date, since a repository can have
-a creation date much earlier than the first day on GitHub.  For example, the
-developers created the git repository for \textit{3D Slicer} in 2002, but did
-not upload a copy of it to GitHub until 2020. Some GitHub data can be found
-using its GitHub Application Program Interface (API) via the following url:
-\textit{https://api.github.com/repos/[owner]/[repository]} where [owner] and
-[repository] are replaced by the repo specific values. The number of months a
-repository has been on GitHub helps us understand the average change of metrics
-over time, like the average new stars per month.
-
-The repository measures help us in many ways. Firstly, they help us get a fast
-and accurate project overview. For example, the number of commits over the last
-12 months shows how active a project has been, and the number of stars and forks
-may reveal its popularity (used to assess \rqref{RQ_CompareHQ2Popular}).
-Secondly, the results may affect our decisions regarding the grading scores for
-some software qualities. For example, if the percentage of comment lines is low,
-we double-check the understandability of the code; if the ratio of open versus
-closed pull requests is high, we pay more attention to maintainability.
-
-As in \citet{SmithEtAl2016}, Virtual machines (VMs) were used to provide an
-optimal testing environment for each package. We used VMs because it is easier
-to start with a fresh environment, without having to worry about existing
-libraries and conflicts. Moreover, when the tests are complete the VM can be
-deleted, without any impact on the host operating system. The most significant
-advantage of using VMs is to level the playing field. Every software install
-starts from a clean slate, which removes ``works-on-my-computer'' errors. When
-filling in the measurement template, the grader notes the details for each VM,
-including hypervisor and operating system version.
-
-When grading the software, we found 27 out of the 29 packages are compatible
-with two or three different OSes, such as Windows, macOS, and Linux, and 5 of
-them are browser-based, making them platform-independent. However, in the
-interest of time, we only performed the measurements for each project by
-installing it on one of the platforms.  When it was an option, we selected
-Windows as the host OS.
-
-\section{Measurement Results} \label{ch_results}
 
@@
 
-\section{Comparison to Community Ranking} \label{Sec_VsCommunityRanking}
+\section{Discussion}
+
+The review section shows us how the ranking between the existing open source MI
+software, but does our ranking match how practitioners in the field rank the
+software.  how does our ranking compare to the software compare to other
+compares compare to community ranking and look at artifacts.
+
+\subsection{Comparison to Community Ranking} \label{Sec_VsCommunityRanking}
 
@@
 
-\section{Comparison Between MI and Research Software for Artifacts}
+\subsection{Comparison Between MI and Research Software for Artifacts}
 \label{Sec_CompareArtifacts}
@@
 
-\section{Threats to Validity} \label{sec_threats_to_validity}
+\subsection{Threats to Validity} \label{sec_threats_to_validity}
 
@@
 
-\subsection{Reliability}
+\subsubsection{Reliability}
 
@@
 
-\subsection{Construct Validity}
+\subsubsection{Construct Validity}
 
@@
 
-\subsection{Internal Validity} \label{Sec_InternalValidity}
+\subsubsection{Internal Validity} \label{Sec_InternalValidity}
 
@@
 
-\subsection{External Validity}
+\subsubsection{External Validity}
 

### 2024-06-01 StateOfPractice/Papers/SOP_MI_OSR/SOP_MI_OSR.tex
Updates up to and including methodology section

@@
 We designed a general methodology to assess the state of the practice for
-research software \citep{SmithEtAl2021, SmithAndMichalski2022}. Details can be
-found in Section~\ref{ch_methods}.  Our methodology has been applied to MI
-software \citep{Dong2021} and Lattice Boltzmann Solvers \citep{Michalski2021,
-SmithEtAl2024}.  This methodology builds off prior work to assess the state of
-the practice for such domains as Geographic Information Systems
-\citep{smith2018state}, Mesh Generators \citep{smith2016state}, Seismology
-software \citep{Smith2018Seismology}, and Statistical software for psychology
-\citep{smith2018statistical}.  In keeping with the previous methodology, we have
-maintained the constraint that the work load for measuring a given domain should
-be feasible for a team as small as one person, and for a short time, ideally
-around a person month of effort. We consider a person month as $20$ working days
-($4$ weeks in a month, with $5$ days of work per week) at $8$ person hours per
-day, or $20 \times 8 = 160$ person hours.
+research software \citep{SmithEtAl2021, SmithAndMichalski2022}.  Our methodology
+has been applied to MI software \citep{Dong2021} and Lattice Boltzmann Solvers
+\citep{Michalski2021, SmithEtAl2024}.  This methodology builds off prior work to
+assess the state of the practice for such domains as Geographic Information
+Systems \citep{smith2018state}, Mesh Generators \citep{smith2016state},
+Seismology software \citep{Smith2018Seismology}, and Statistical software for
+psychology \citep{smith2018statistical}.  In keeping with the previous
+methodology, we have maintained the constraint that the work load for measuring
+a given domain should be feasible for a team as small as one person, and for a
+short time, ideally around a person month of effort. We consider a person month
+as $20$ working days ($4$ weeks in a month, with $5$ days of work per week) at
+$8$ person hours per day, or $20 \times 8 = 160$ person hours.
 
@@
 case MI) and identify a list of about 30 software packages. (For measuring MI we
-used 29 software packages.)  We approximately measure the qualities of each
-package by filling in a grading template. Compared with our previous
-methodology, the new methodology also includes repository based metrics, such as
-the number of files, number of lines of code, percentage of issues that are
-closed, etc.  With the quantitative data in the grading template, we rank the
-software with the Analytic Hierarchy Process (AHP) (Section~\ref{ch_background}
-provides details).
-
-We developed a methodology for evaluating the state of the practice of research
-software \citep{SmithEtAl2021, SmithAndMichalski2022}.  The methodology can be
-instantiated for a specific domain of scientific software, which in the current
-case is medical imaging software for visualization.  Our methodology involves
-and engages a domain expert partner throughout, as discussed in
-Section~\ref{sec_vet_software_list}.  The three main steps of the methodology
-are:
-
-\begin{enumerate}
-\item Identify list of representative software packages
-(Section~\ref{sec_software_selection});
-\item Measure (or grade) the selected software
-(Section~\ref{sec_grading_software}) on qualities using AHP;
-\end{enumerate}
-
-define software qualities and ahp.
+used 29 software packages.)  The selected packages are vetted by a domain
+expert. We approximately measure the qualities of each package by filling in a
+grading template. The grader looks at the contents of the open source repository
+and repository based metrics, such as the number of files, number of lines of
+code, percentage of issues that are closed, etc.  With the quantitative data in
+the grading template, we rank the software with the Analytic Hierarchy Process
+(AHP). Further details on interaction with the domain expert, software
+qualities, grading the software and AHP are found below and in
+\citep{SmithEtAl2024_MI_SOP}.
 
@@
 
-The Domain Expert is an important member of the state of the practice assessment
-team. Pitfalls exist if non-experts attempt to acquire an authoritative list of
-software, or try to definitively rank the software. Non-experts have the problem
-that they can only rely on information available on-line, which has the
-following drawbacks:
-\begin{inparaenum}[i)]
-  \item the on-line resources could have false or inaccurate information; and,
-  \item the on-line resources could leave out relevant information that is so
-in-grained with experts that nobody thinks to explicitly record it.
-\end{inparaenum}
-
-Domain experts may be recruited from academia or industry.  The only
-requirements are knowledge of the domain and a willingness to be engaged in the
-assessment process.  The Domain Expert does not have to be a software developer,
-but they should be a user of domain software.  Given that the domain experts are
-likely to be busy people, the measurement process cannot put too much of a burden
-on their time.  For the current assessment, our Domain Expert (and paper
-co-author) is Dr.\ Michael Noseworthy, Professor of Electrical and Computer
-Engineering at McMaster University, Co-Director of the McMaster School of
-Biomedical Engineering, and Director of Medical Imaging Physics and Engineering
-at St.\ Joseph's Healthcare, Hamilton, Ontario, Canada.
+The Domain Experts vet the proposed list of software packages and the AHP
+ranking.  Doing these tasks with non-experts runs the risk of trusting
+inaccurate resources and overlooking the tacit knowledge of an expert. For the
+current assessment, our Domain Expert (and paper co-author) is Dr.\ Michael
+Noseworthy, Professor of Electrical and Computer Engineering at McMaster
+University, Co-Director of the McMaster School of Biomedical Engineering, and
+Director of Medical Imaging Physics and Engineering at St.\ Joseph's Healthcare,
+Hamilton, Ontario, Canada.
 
@@
 
-The Domain Experts are asked to vet the collected data and analysis.  In
-particular, they are asked to vet the proposed list of software packages and the
-AHP ranking.  These interactions can be done either electronically or with
-in-person (or virtual) meetings.
+\subsubsection{Software Qualities} \label{sec_software_quality}
+
+Quality is defined as a measure of the excellence or worth of an entity.  As is
+common practice, we do not think of quality as a single measure, but rather as a
+set of measures.  That is, quality is a collection of different qualities, often
+called ``ilities.''  For this study we selected 9 qualities to measure:
+installability, correctness/ verifiability, reliability, robustness, usability,
+maintainability, reusability, understandability, and visibility/transparency.
+With the exception of installability, all the qualities are defined in
+\citet{GhezziEtAl2003}. Installability is defined as the effort required for the
+installation and/or uninstallation of software in a specified environment
+\citep{ISO/IEC25010, lenhard2013measuring}.
 
@@
 Figure~\ref{fg_grading_template_example} shows an excerpt of the spreadsheet.
-The spreadsheet includes a column for each measured software package.
+The spreadsheet includes a column for each measured software package. TODO - point to full data online.
 
@@
 release and the most recent change date, website, source code repository of the
-product, number of developers, etc.  We follow the definitions given by
-\citet{GewaltigAndCannon2012} for the software categories.  Public means
-software intended for public use.  Private means software aimed only at a
-specific group, while the concept category is for software written simply to
-demonstrate algorithms or concepts. The three categories of development models
-are (open source, free-ware and commercial) are discussed in
-Section~\ref{sec_software_categories}.  Information in the summary section sets
-the context for the project, but it does not directly affect the grading scores.
+product, number of developers, etc.  We only measure open source software that
+is ready for public use by other researchers (following the definition of the
+public category from \citep{GewaltigAndCannon2012}). Information in the summary
+section sets the context for the project, but it does not directly affect the
+grading scores.
 
@@
 started manual, a user manual and a statement of expected user characteristics.
-Details on how we measure each quality are provided in \citet{SmithEtAl2024}.
-
-As part of filling in the measurement template, we use freeware tools to collect
-repository related data. \href{https://github.com/tomgi/git_stats}{GitStats}
-\citep{Gieniusz2019} is used to measure the number of binary files as well as
-the number of added and deleted lines in a repository. We also use this tool to
-measure the number of commits over different intervals of time.
+We use two freeware tools to collect repository related data:
+\href{https://github.com/tomgi/git_stats}{GitStats} \citep{Gieniusz2019} and
 \href{https://github.com/boyter/scc}{Sloc Cloc and Code (scc)}
-\citep{Boyter2021} is used to measure the number of text based files as well as
-the number of total, code, comment, and blank lines in a repository.
-
-Both tools measure the number of text-based files in a git repository and lines
-of text in these files. Based on our experience, most text-based files in a
-repository contain programming source code, and developers use them to compile
-and build software products. A minority of these files are instructions and
-other documents. So we roughly regard the lines of text in text-based files as
-lines of programming code. The two tools usually generate similar but not
-identical results. From our understanding, this minor difference is due to the
-different techniques to detect if a file is text-based or binary.
-
-For projects on GitHub we manually collect additional information, such as the
-numbers of stars, forks, people watching this repository, open pull requests,
-closed pull requests, and the number of months a repository has been on GitHub.
-We need to take care with the project creation date, since a repository can have
-a creation date much earlier than the first day on GitHub.  For example, the
-developers created the git repository for \textit{3D Slicer} in 2002, but did
-not upload a copy of it to GitHub until 2020. Some GitHub data can be found
-using its GitHub Application Program Interface (API) via the following url:
-\textit{https://api.github.com/repos/[owner]/[repository]} where [owner] and
-[repository] are replaced by the repo specific values. The number of months a
-repository has been on GitHub helps us understand the average change of metrics
-over time, like the average new stars per month.
-
-The repository measures help us in many ways. Firstly, they help us get a fast
-and accurate project overview. For example, the number of commits over the last
-12 months shows how active a project has been, and the number of stars and forks
-may reveal its popularity (used to assess \rqref{RQ_CompareHQ2Popular}).
-Secondly, the results may affect our decisions regarding the grading scores for
-some software qualities. For example, if the percentage of comment lines is low,
-we double-check the understandability of the code; if the ratio of open versus
-closed pull requests is high, we pay more attention to maintainability.
-
-As in \citet{SmithEtAl2016}, Virtual machines (VMs) were used to provide an
-optimal testing environment for each package. We used VMs because it is easier
-to start with a fresh environment, without having to worry about existing
-libraries and conflicts. Moreover, when the tests are complete the VM can be
-deleted, without any impact on the host operating system. The most significant
-advantage of using VMs is to level the playing field. Every software install
-starts from a clean slate, which removes ``works-on-my-computer'' errors. When
-filling in the measurement template, the grader notes the details for each VM,
-including hypervisor and operating system version.
-
-When grading the software, we found 27 out of the 29 packages are compatible
-with two or three different OSes, such as Windows, macOS, and Linux, and 5 of
-them are browser-based, making them platform-independent. However, in the
-interest of time, we only performed the measurements for each project by
-installing it on one of the platforms.  When it was an option, we selected
-Windows as the host OS.
-
-\subsubsection{Software Quality Definitions} \label{sec_software_quality}
-
-Quality is defined as a measure of the excellence or worth of an entity.  As is
-common practice, we do not think of quality as a single measure, but rather as a
-set of measures.  That is, quality is a collection of different qualities, often
-called ``ilities.''  For this study we selected 9 qualities to measure: .....
-With the exception of installability, all the qualities are defined in
-\citet{GhezziEtAl2003}. Installability is defined as the effort required for the
-installation and/or uninstallation of software in a specified environment
-\citep{ISO/IEC25010, lenhard2013measuring}.  Further details on the software
-qualities can be found in \citet{SmithEtAl2024}.
+\citep{Boyter2021}.  Details on how we measure each quality are provided in
+\citet{SmithEtAl2024_MI_SOP}.
 
@@
 between qualities.
-
-The first step for ranking the software choices for a given quality involves a
-pairwise comparison between each of the $n$ software options for that quality.
-AHP expresses the comparison through an $n \times n$ matrix $A$. When comparing
-option $i$ and option $j$, the value of $A_{ij}$ is decided as follows, with the
-value of $A_{ji}$ generally equal to $1/A_{ij}$ \citep{Saaty1990}: $A_{ij} = 1$
-if criterion $i$ and criterion $j$ are equally important, while $A_{ij} = 9$ if
-criterion $i$ is extremely more important than criterion $j$.  The natural
-numbers between 1 and 9 are used to show the different levels of relative
-importance between these two extremes. The above assumes that option $i$ is of
-equal, or more, importance compared to option $j$ ($i \geq j$).  If that is not
-the case, we reverse $i$ and $j$ and determine $A_{ji}$ first, then $A_{ij} =
-1/A_{ji}$.
-
-Section~\ref{sec_grading_software} shows how we measure the software via a
-grading template.  For the AHP process, the relevant measure is the subjective
-score from $1$ to $10$ for each quality for each package. To turn these
-subjective measures $x_{\text{sub}}$ and $y_{\text{sub}}$ into Saaty's
-pair-wise scores for option $x$ versus option $y$, respectively, we use the
-following calculation:
-\[
-\begin{cases}
-\min\{9, x_{\text{sub}} - y_{\text{sub}} + 1\} & x_{\text{sub}} \geq y_{\text{sub}} \\
-1 / \min\{9, y_{\text{sub}} - x_{\text{sub}} + 1\} & x_{\text{sub}} < y_{\text{sub}}
-\end{cases}
-\]
-
-\noindent For example, we measured the usability for 3D Slicer and Ginkgo CADx
-as $8$ and $7$, respectively; therefore, on the 9-point scale, 3D Slicer compared
-to Ginkgo CADx is 2 and Ginkgo CADx versus 3D Slicer is 1/2, as shown in the
-sample AHP calculations (Table~\ref{Tbl_SampleAHP}).
-
-The second step is to calculate the priority vector $w$ from $A$.  The
-vector $w$ ranks the software options by how well they achieve the given
-quality.  The priority vector can be calculated by solving the equation
-\citep{Saaty1990}:
-\begin{equation} 
-    A w = \lambda_{\text{max}} w,
-\end{equation}
-where $\lambda_{\text{max}}$ is the maximal eigenvalue of $A$.  In this project,
-$w$ is approximated with the classic \textit{mean of normalized values} approach
-\citep{AlessioEtAl2006}:
-
-\begin{equation}
-w_i = \frac{1}{n}\sum_{j=1}^{n}\frac{A_{ij}}{\sum_{k=1}^{n}A_{kj}}
-\end{equation}
-
-Table~\ref{Tbl_SampleAHP} summarizes the above two steps for the quality of
-installability.  The matrix $A$ is shown in the first set of columns, then the
-normalized version of $A$ and finally the average of the normalized values to
-form the vector $w$ in the last column.
-
-\begin{table}[h!]
-\begin{center}
-\begin{tabular}{ l c c c c c | c c c c c | c }
- \toprule
- ~ & \multicolumn{5}{c|}{$A_{ij}$} & \multicolumn{5}{c|}{${A_{ij}}/{\sum_{k=1}^{n}A_{kj}}$} & ~\\
- \midrule
- ~ & \rotatebox{90}{3D Slicer} & \rotatebox{90}{Ginkgo} & \rotatebox{90}{XMedCon} & $\cdots$ & \rotatebox{90}{Gwyddion} & \rotatebox{90}{3D Slicer} & \rotatebox{90}{Ginkgo} & \rotatebox{90}{XMedCon} & $\cdots$ & \rotatebox{90}{Gwyddion} & AVG \\
- \midrule
- 3D Slicer & 1 & 2 & 4 & $\cdots$ & 2 & 0.071 & 0.078 & 0.060 & $\cdots$ & 0.078 & 0.068\\
- Ginkgo & 1/2 & 1 & 3 & $\cdots$ & 1 & 0.036 & 0.039 & 0.045 & $\cdots$ & 0.039 & 0.041\\
- XMedCon & 1/4 & 1/3 & 1 & $\cdots$ & 1/3 & 0.018 & 0.013 & 0.015 & $\cdots$ & 0.013 & 0.015\\
- $\vdots$ & $\vdots$ & $\vdots$ & $\vdots$ & $\ddots$ & $\vdots$ & $\vdots$ & $\vdots$ & $\vdots$ & $\ddots$ & $\vdots$ & $\vdots$\\
- Gwyddion & 1/2 & 1 & 3 & $\cdots$ & 1 & 0.036 & 0.039 & 0.045 & $\cdots$ & 0.039 & 0.041\\  
- \midrule
- SUM = & 14.01 & 25.58 & 66.75 & $\cdots$ & 25.58 & 1.000 & 1.000 & 1.000 & $\cdots$ & 1.000 & 1.000\\
- \bottomrule
-\end{tabular}
-\end{center}
-\caption{Sample AHP Calculations for the Quality of Usability} \label{Tbl_SampleAHP}
-\end{table}
-
-We repeat the first and second steps for each of the qualities.  The third step
-combines the quality rankings into an overall ranking.  Following AHP, we need
-to first prioritize the qualities.  The AHP method finds the priority of quality
-$i$ ($p_i$) in the same way that the score ($w_j$) was found for software
-package $j$ evaluated for a given quality (as shown above).  That is, we
-conduct a pairwise comparison between the priority of different qualities to
-construct the $m \times m$ matrix $A$, and then we take the mean of normalized
-values for row $i$ to find the priority value $p_i$ for quality $i$.  If we
-introduce the notation that $w^i_j$ is the score for quality $i$ for package
-$j$, then the overall score $S_j$ for package $j$ is found via:
-
-$$S_j = \sum_{i=1}^m w^i_j p_i$$
 

### 2024-06-01 StateOfPractice/Papers/SOP_MI_OSR/SOP_MI_OSR.tex
Edits up to and including the Results section

@@
 Figure~\ref{fg_grading_template_example} shows an excerpt of the spreadsheet.
-The spreadsheet includes a column for each measured software package. TODO - point to full data online.
+The spreadsheet includes a column for each measured software package. \citep{Dong2021-Data} provides the full set of measurement data.
 
@@
 $n=29$ and $m=9$ since there are 29 options (software products) and 9 criteria
-(qualities). We rank the software for each of the qualities, and then we combine
-the quality rankings into an overall ranking based on the relative priorities
-between qualities.
+(qualities). With AHP the sum of the grades (scores) for all products for a
+given quality will be 1.0.  We rank the software for each of the qualities, and
+then we combine the quality rankings into an overall ranking based on the
+relative priorities between qualities.
 
@@
 48 packages is available in \citet{Dong2021}.  To reduce the length of the list
-to a manageable number (29 in this case, as given in Section~\ref{ch_results}),
-we filtered the original list as follows:
+to a manageable number (29 in this case), we filtered the original list as
+follows:
 
@@
 
-\item We removed the packages that did not have source code available, such as
+\item We removed the packages with no source code available, such as
 \textit{MicroDicom}, \textit{Aliza}, and \textit{jivex}.
@@
 \item We focused on the MI software that provides visualization functions, as
-described in Section~\ref{sec_scope}. Furthermore, we removed seven packages that were
-toolkits or libraries, such as \textit{VTK}, \textit{ITK}, and \textit{dcm4che}.
-We removed another three that were for PACS.
+described in Section~\ref{sec_scope}. Furthermore, we removed seven packages
+that were toolkits or libraries, such as \textit{VTK}, \textit{ITK}, and
+\textit{dcm4che}.  We removed another three that were for PACS.
 
@@
 \textit{Diffusion Toolkit}, and \textit{MRItrix}) were on the Domain Expert
-list, but not on our filtered list.  However, when we examined those packages,
+list, but not on our filtered list.  When we examined those packages,
 we found they were out of scope, since their primary function was not
@@
 \textit{GATE} could not open macro files and became unresponsive several times,
-without any descriptive error message. When assessing robustness
-(Section~\ref{sec_result_robustness}), we found that \textit{Drishti} crashed
+without any descriptive error message. We found that \textit{Drishti} crashed
 when loading damaged image files, without showing any descriptive error message.
-We did not find any problems with the on-line version of
-\textit{dwv}.
+We did not find any problems with the on-line version of \textit{dwv}.
 
@@
 maintainability. We gave \textit{3D Slicer} the highest score because we found
-it to have the most comprehensive artifacts. For example, as far as we could
-find, only a few of the 29 projects had a product, developer's manual, or API
-documentation, and only \textit{3D Slicer}, \textit{ImageJ}, \textit{Fiji}
-included all three documents. Moreover, \textit{3D Slicer} has a much higher
-percentage of closed issues (92\%) compared to \textit{ImageJ} (52\%) and
-\textit{Fiji} (64\%). Table~\ref{tab_maintainability_docs} shows which
-projects had these documents, in descending order of their maintainability
-scores.
+it to have the most comprehensive artifacts. Only a few of the 29 projects had a
+product, developer's manual, or API documentation, and only \textit{3D Slicer},
+\textit{ImageJ}, \textit{Fiji} included all three documents. Moreover,
+\textit{3D Slicer} has a much higher percentage of closed issues (92\%) compared
+to \textit{ImageJ} (52\%) and \textit{Fiji} (64\%).
+Table~\ref{tab_maintainability_docs} shows which projects had these documents,
+in descending order of their maintainability scores.
 
@@
 
-Twenty-seven of the 29 projects used git as the version control tool, with 24 of these
-using GitHub. \textit{AMIDE} used Mercurial and \textit{Gwyddion} used
+Twenty-seven of the 29 projects used git as the version control tool, with 24 of
+these using GitHub. \textit{AMIDE} used Mercurial and \textit{Gwyddion} used
 Subversion. \textit{XMedCon}, \textit{AMIDE}, and \textit{Gwyddion} used
@@
 understandability. All projects had a consistent coding style with parameters in
-the same order for all functions; modularized code; and, clear comments, indicating
-what is done, not how. However, we only found explicit identification of a
-coding standard for 3 out of the 29: \textit{3D Slicer}, \textit{Weasis}, and
-\textit{ImageJ}. We also found hard-coded constants (rather than symbolic
+the same order for all functions, modularized code, and, clear comments that
+indicate what is done, not how. However, we only found explicit identification
+of a coding standard for 3 out of the 29: \textit{3D Slicer}, \textit{Weasis},
+and \textit{ImageJ}. We also found hard-coded constants (rather than symbolic
 constants) in \textit{medInria}, \textit{dicompyler}, \textit{MicroView}, and

### 2024-06-05 StateOfPractice/Papers/SOP_MI_OSR/SOP_MI_OSR.tex
Add roadmap for subsections of Section 3 (Discussion)

@@
 
-The review section shows us how the ranking between the existing open source MI
-software, but does our ranking match how practitioners in the field rank the
-software.  how does our ranking compare to the software compare to other
-compares compare to community ranking and look at artifacts.
+In Section~\ref{Sec_VsCommunityRanking}, we compare our ranking to the MI
+software community's ranking, as a way to judge our results, and a way to
+understand what the MI community currently values. To understand the state of
+the practice for MI software relative to other software, we compare MI software
+to other research software in Section~\ref{Sec_CompareArtifacts}.  The
+comparison is based on the artifacts our measurement exercise found in the MI
+repositories, versus what is typically recommended for research software.  We
+also discuss potential reasons to doubt our results in the threats to validity
+section (Section~\ref{sec_threats_to_validity}).
 

### 2024-06-07 StateOfPractice/Papers/SOP_MI_OSR/SOP_MI_OSR.tex
Edits to SOP MI paper in Section 3

@@
 In Section~\ref{Sec_VsCommunityRanking}, we compare our ranking to the MI
-software community's ranking, as a way to judge our results, and a way to
+software community's ranking, as a way to judge our results, and as a way to
 understand what the MI community currently values. To understand the state of
@@
 repositories, versus what is typically recommended for research software.  We
-also discuss potential reasons to doubt our results in the threats to validity
-section (Section~\ref{sec_threats_to_validity}).
+discuss threats to the validity of our data and conclusions in section
+(Section~\ref{sec_threats_to_validity}).
 
@@
 
-To address~\rqref{RQ_CompareHQ2Popular} about how our ranking compares to the
+To answer~\rqref{RQ_CompareHQ2Popular} about how our ranking compares to the
 popularity of projects as judged by the scientific community, we make two
 comparisons:
+
 \begin{itemize} 
-\item A comparison of our ranking (from Section~\ref{ch_results}) with the
-community ratings on GitHub, as shown by GitHub stars, number of forks, and
-number of people watching the projects; and,
+\item A comparison of our ranking (from
+Section~\ref{ch_results}) with the community ratings on GitHub, as shown by
+GitHub stars, number of forks, and number of people watching the projects; and,
 \item A comparison of top-rated software from our methodology with the top
-recommendations from our domain experts (as mentioned in
-Section~\ref{sec_software_selection}).
+recommendations from our domain expert (as mentioned in
+Section~\ref{sec_vet_software_list}).
 \end{itemize}
@@
 stayed on GitHub, and calculate the average number of new stars, people
-watching, and forks per 12 months. Section~\ref{sec_grading_software} describes
-the method of getting the creation date.  The items in
+watching, and forks per 12 months. The items in
 Table~\ref{tab_ranking_vs_GitHub} are listed in descending order of the average
@@
 lists. Our top five packages are scattered among the first eight positions on the
-GitHub list. However, as discussed below there are discrepancies between the two
+GitHub list. However, as discussed below, there are discrepancies between the two
 lists.

### 2024-06-10 StateOfPractice/Papers/SOP_MI_OSR/SOP_MI_OSR.tex
Fixed reference labels that were out of date

@@
 \begin{itemize} 
-\item A comparison of our ranking (from
-Section~\ref{ch_results}) with the community ratings on GitHub, as shown by
-GitHub stars, number of forks, and number of people watching the projects; and,
+\item A comparison of our ranking (from Section~\ref{ch_results}) with the
+community ratings on GitHub, as shown by GitHub stars, number of forks, and
+number of people watching the projects; and,
 \item A comparison of top-rated software from our methodology with the top
@@
 failed to build it locally, and used the test version on its websites for the
-measurements. We followed the instructions and tried to run the command ``yarn
-run test'' locally, which did not work. In addition, the test version did not
-detect a broken DICOM file and displayed a blank image as described in
-Section~\ref{sec_result_robustness}. We might underestimate the scores for
-\textit{dwv} due to uncommon technical issues.
+measurements. We followed the instructions and tried to run the command
+\texttt{yarn run test} locally, which did not work. In addition, the test
+version did not detect a broken DICOM file and displayed a blank image as
+described in Section~\ref{sec_result_robustness}. We might underestimate the
+scores for \textit{dwv} due to uncommon technical issues.
 
@@
 
-As shown in Section~\ref{sec_software_selection}, our domain experts recommended
+% The Domain Expert part can be removed if space is needed
+As shown in Section~\ref{ch_results}, our domain experts recommended
 a list of top software with 12 software products.  All the top 4 entries from
@@
 did not have visualization as the primary function (as discussed in
-Section~\ref{sec_software_selection}).  The differences between the list
+Section~\ref{ch_results}).  The differences between the list
 recommended by our methodology and the Domain Expert are not surprising.  As
@@
 next sections will explore how the practices of the MI community compare to the
-broader research software community. We will also investigate the practices from
-the top projects that others within the MI community, and within the broader
-research software community, can potentially adopt.
+broader research software community.
 
@@
 \begin{itemize}
-\item In our search for software packages
-(Section~\ref{sec_software_selection}), we may have missed a relevant package.
+\item In our search for software packages (Section~\ref{ch_results}), we may
+have missed a relevant package.
 \item Our methodology assumes that all relevant software development activities
@@
 understanding current practice by answering our four research questions
-(Section~\ref{sec_motivation}).  Our methods in Section~\ref{ch_methods} form a
+(Section~\ref{sec_motivation}).  Our methods in Section~\ref{SecMethodology} form a
 general process to evaluate domain-specific software, that we apply to the

### 2024-06-10 StateOfPractice/Papers/SOP_MI_OSR/SOP_MI_OSR.tex
Edits to SOP for MI paper

@@
 
-\usepackage[round, sort]{natbib}
+%\usepackage[round, sort]{natbib}
 
@@
 \pgfplotsset{compat=1.13}
+
+% prepared for: Journal of Imaging Informatics in Medicine
+% https://link.springer.com/journal/10278
 
@@
 Ultrasound) to provide information for diagnostic, analytic, and medical
-applications \citep{FDA2021, enwiki:1034887445, Zhang2008}.
+applications \cite{FDA2021, enwiki:1034887445, Zhang2008}.
 Figure~\ref{Fig_Example}, which shows an image of the brain, highlights the
@@
 Not only do we wish to gain insight into the state of the practice for MI
-software, we also wish to understand the impact of the often cited gap, or chasm,
-between software engineering and research software \citep{Kelly2007,
+software, we also wish to understand the impact of the often cited gap, or
+chasm, between software engineering and research software \cite{Kelly2007,
 Storer2017}. Although scientists spend a substantial proportion of their working
-hours on software development \citep{Hannay2009, Prabhu2011}, many developers
+hours on software development \cite{Hannay2009, Prabhu2011}, many developers
 learn software engineering skills by themselves or from their peers, instead of
-from proper training \citep{Hannay2009}. \citet{Hannay2009} observe that many
-scientists showed ignorance and indifference to standard software engineering
-concepts. For instance, according to a survey by \citet{Prabhu2011}, more than
-half of their 114 subjects did not use a proper debugger when coding.
+from proper training \cite{Hannay2009}. In the past many scientists have shown
+ignorance and indifference to standard software engineering concepts
+\cite{Hannay2009}. For instance, according to a survey by Prabhu in 2011
+\cite{Prabhu2011}, more than half of their 114 subjects did not use a proper
+debugger when coding.
 
 To gain insight, we devised 10 research questions, which can be applied to MI,
-as well as to other domains, of research software \citep{SmithEtAl2021,
+as well as to other domains, of research software \cite{SmithEtAl2021,
 SmithAndMichalski2022}.  We designed the questions to learn about the
@@
 Simulation, plus MI archiving and telemedicine systems (Compression, Storage,
-and Communication) (as summarized by \citet{Bankman2000} and
-\citet{Angenent2006}).  We also exclude Statistical Analysis and Image-based
-Physiological Modelling \citep{enwiki:1034877594} and Feature Extraction,
-Classification, and Interpretation \citep{Kim2011}. Software that provides MI
-support functions is also out of scope; therefore, we have not assessed the
-toolkit libraries VTK \citep{SchroederEtAl2006} and ITK \citep{McCormick2014}.
-Finally, Picture Archiving and Communication System (PACS), which helps users to
-economically store and conveniently access images \citep{Choplin1992}, are
-considered out of scope.
+and Communication) \cite{Bankman2000, Angenent2006}.  We also exclude
+Statistical Analysis and Image-based Physiological Modelling
+\cite{enwiki:1034877594} and Feature Extraction, Classification, and
+Interpretation \cite{Kim2011}. Software that provides MI support functions is
+also out of scope; therefore, we have not assessed the toolkit libraries VTK
+\cite{SchroederEtAl2006} and ITK \cite{McCormick2014}.  Finally, Picture
+Archiving and Communication System (PACS), which helps users to economically
+store and conveniently access images \cite{Choplin1992}, are considered out of
+scope.
 
@@
 We designed a general methodology to assess the state of the practice for
-research software \citep{SmithEtAl2021, SmithAndMichalski2022}.  Our methodology
-has been applied to MI software \citep{Dong2021} and Lattice Boltzmann Solvers
-\citep{Michalski2021, SmithEtAl2024}.  This methodology builds off prior work to
+research software \cite{SmithEtAl2021, SmithAndMichalski2022}.  Our methodology
+has been applied to MI software \cite{Dong2021} and Lattice Boltzmann Solvers
+\cite{Michalski2021, SmithEtAl2024}.  This methodology builds off prior work to
 assess the state of the practice for such domains as Geographic Information
-Systems \citep{smith2018state}, Mesh Generators \citep{smith2016state},
-Seismology software \citep{Smith2018Seismology}, and Statistical software for
-psychology \citep{smith2018statistical}.  In keeping with the previous
+Systems \cite{smith2018state}, Mesh Generators \cite{smith2016state},
+Seismology software \cite{Smith2018Seismology}, and Statistical software for
+psychology \cite{smith2018statistical}.  In keeping with the previous
 methodology, we have maintained the constraint that the work load for measuring
@@
 qualities, grading the software and AHP are found below and in
-\citep{SmithEtAl2024_MI_SOP}.
+\cite{SmithEtAl2024_MI_SOP}.
 
@@
 maintainability, reusability, understandability, and visibility/transparency.
-With the exception of installability, all the qualities are defined in
-\citet{GhezziEtAl2003}. Installability is defined as the effort required for the
-installation and/or uninstallation of software in a specified environment
-\citep{ISO/IEC25010, lenhard2013measuring}.
+With the exception of installability, all the qualities are defined in Ghezzi et
+al. (2003) \cite{GhezziEtAl2003}. Installability is defined as the effort
+required for the installation and/or uninstallation of software in a specified
+environment \cite{ISO/IEC25010, lenhard2013measuring}.
 
@@
 
-We grade the selected software using the measurement template summarized in
-\citet{SmithEtAl2021}.  The template provides measures of the qualities listed
-in Section~\ref{sec_software_quality}.
-For each software package, we fill in the template questions. To stay within the
-target of 160 person hours to measure the domain, we allocated between one and
-four hours for each package. Project developers can be contacted for help
-regarding installation, if necessary, but we impose a cap of about two hours on
-the installation process, to keep the overall measurement time feasible.
-Figure~\ref{fg_grading_template_example} shows an excerpt of the spreadsheet.
-The spreadsheet includes a column for each measured software package. \citep{Dong2021-Data} provides the full set of measurement data.
+We grade the selected software using an existing measurement template
+\cite{SmithEtAl2021}.  The template provides measures of the qualities listed in
+Section~\ref{sec_software_quality}.  For each software package, we fill in the
+template questions. To stay within the target of 160 person hours to measure the
+domain, we allocated between one and four hours for each package. Project
+developers can be contacted for help regarding installation, if necessary, but
+we impose a cap of about two hours on the installation process, to keep the
+overall measurement time feasible.  Figure~\ref{fg_grading_template_example}
+shows an excerpt of the spreadsheet.  The spreadsheet includes a column for each
+measured software package. \cite{Dong2021-Data} provides the full set of
+measurement data.
 
@@
 is ready for public use by other researchers (following the definition of the
-public category from \citep{GewaltigAndCannon2012}). Information in the summary
+public category from \cite{GewaltigAndCannon2012}). Information in the summary
 section sets the context for the project, but it does not directly affect the
@@
 We use two freeware tools to collect repository related data:
-\href{https://github.com/tomgi/git_stats}{GitStats} \citep{Gieniusz2019} and
+\href{https://github.com/tomgi/git_stats}{GitStats} \cite{Gieniusz2019} and
 \href{https://github.com/boyter/scc}{Sloc Cloc and Code (scc)}
-\citep{Boyter2021}.  Details on how we measure each quality are provided in
-\citet{SmithEtAl2024_MI_SOP}.
+\cite{Boyter2021}.  Details on how we measure each quality are provided in
+Smith et al.\ (2024) \cite{SmithEtAl2024_MI_SOP}.
 
@@
 Saaty developed AHP in the 1970s, and people have widely used it since to make
-and analyze multiple criteria decisions \citep{VaidyaEtAl2006}. AHP organizes
+and analyze multiple criteria decisions \cite{VaidyaEtAl2006}. AHP organizes
 multiple criteria in a hierarchical structure and uses pairwise comparisons
-between alternatives to calculate relative ratios \citep{Saaty1990}. AHP works
+between alternatives to calculate relative ratios \cite{Saaty1990}. AHP works
 with sets of $n$ \textit{options} and $m$ \textit{criteria}.  In our project
@@
 the chosen domain; and, ii) filter the list to remove less relevant members
-\citep{SmithEtAl2021}.
-
-We initially identified 48 MI candidate software projects from the literature
-\citep{Bjorn2017, Bruhschwein2019, Haak2015}, on-line articles \citep{Emms2019,
-Hasan2020, Mu2019}, and forum discussions \citep{Samala2014}.  The full list of
-48 packages is available in \citet{Dong2021}.  To reduce the length of the list
-to a manageable number (29 in this case), we filtered the original list as
-follows:
+\cite{SmithEtAl2021}.
+
+We initially identified 48 MI candidate software projects \cite{Dong2021} from
+the literature \cite{Bjorn2017, Bruhschwein2019, Haak2015}, on-line articles
+\cite{Emms2019, Hasan2020, Mu2019}, and forum discussions \cite{Samala2014}.  To
+reduce the length of the list to a manageable number (29 in this case), we
+filtered the original list as follows:
 
@@
 platform-independence via web applications. The full measurement data for all
-packages is available in \citet{Dong2021-Data}.
+packages is available on \href{https://data.mendeley.com/datasets/k3pcdvdzj2/1}
+{Mendeley Data}.
 
@@
  &  &  &  &  &  & W & M & L &  \\ \midrule
-ParaView \citep{Ahrens2005} & 2002 & 2020-10 & \checkmark & 100 & 886326 & \checkmark & \checkmark & \checkmark & \checkmark \\
-Gwyddion \citep{Nevcas2012} & 2004 & 2020-11 &  & 38 & 643427 & \checkmark & \checkmark & \checkmark &  \\
-Horos \citep{horosproject2020} & ? & 2020-04 &  & 21 & 561617 &  & \checkmark &  &  \\
-OsiriX Lite \citep{PixmeoSARL2019} & 2004 & 2019-11 &  & 9 & 544304 &  & \checkmark &  &  \\
-3D Slicer \citep{Kikinis2014} & 1998 & 2020-08 & \checkmark & 100 & 501451 & \checkmark & \checkmark & \checkmark &  \\
-Drishti \citep{Limaye2012} & 2012 & 2020-08 &  & 1 & 268168 & \checkmark & \checkmark & \checkmark &  \\
-Ginkgo CADx \citep{Wollny2020} & 2010 & 2019-05 &  & 3 & 257144 & \checkmark & \checkmark & \checkmark &  \\
-GATE \citep{Jan2004} & 2011 & 2020-10 &  & 45 & 207122 &  & \checkmark & \checkmark &  \\
-3DimViewer \citep{TESCAN2020} & ? & 2020-03 & \checkmark & 3 & 178065 & \checkmark & \checkmark &  &  \\
-medInria \citep{Fillard2012} & 2009 & 2020-11 &  & 21 & 148924 & \checkmark & \checkmark & \checkmark &  \\
-BioImage Suite Web \citep{Papademetris2005} & 2018 & 2020-10 & \checkmark & 13 & 139699 &
+ParaView \cite{Ahrens2005} & 2002 & 2020-10 & \checkmark & 100 & 886326 & \checkmark & \checkmark & \checkmark & \checkmark \\
+Gwyddion \cite{Nevcas2012} & 2004 & 2020-11 &  & 38 & 643427 & \checkmark & \checkmark & \checkmark &  \\
+Horos \cite{horosproject2020} & ? & 2020-04 &  & 21 & 561617 &  & \checkmark &  &  \\
+OsiriX Lite \cite{PixmeoSARL2019} & 2004 & 2019-11 &  & 9 & 544304 &  & \checkmark &  &  \\
+3D Slicer \cite{Kikinis2014} & 1998 & 2020-08 & \checkmark & 100 & 501451 & \checkmark & \checkmark & \checkmark &  \\
+Drishti \cite{Limaye2012} & 2012 & 2020-08 &  & 1 & 268168 & \checkmark & \checkmark & \checkmark &  \\
+Ginkgo CADx \cite{Wollny2020} & 2010 & 2019-05 &  & 3 & 257144 & \checkmark & \checkmark & \checkmark &  \\
+GATE \cite{Jan2004} & 2011 & 2020-10 &  & 45 & 207122 &  & \checkmark & \checkmark &  \\
+3DimViewer \cite{TESCAN2020} & ? & 2020-03 & \checkmark & 3 & 178065 & \checkmark & \checkmark &  &  \\
+medInria \cite{Fillard2012} & 2009 & 2020-11 &  & 21 & 148924 & \checkmark & \checkmark & \checkmark &  \\
+BioImage Suite Web \cite{Papademetris2005} & 2018 & 2020-10 & \checkmark & 13 & 139699 &
 \checkmark & \checkmark & \checkmark & \checkmark \\
-Weasis \citep{Roduit2021} & 2010 & 2020-08 &  & 8 & 123272 & \checkmark & \checkmark & \checkmark &  \\
-AMIDE \citep{Loening2017} & 2006 & 2017-01 &  & 4 & 102827 & \checkmark & \checkmark & \checkmark &  \\
-XMedCon \citep{Nolf2003} & 2000 & 2020-08 &  & 2 & 96767 & \checkmark & \checkmark & \checkmark &  \\
-ITK-SNAP \citep{Yushkevich2006} & 2006 & 2020-06 & \checkmark & 13 & 88530 & \checkmark & \checkmark & \checkmark &  \\
-Papaya \citep{UTHSCSA2019} & 2012 & 2019-05 &  & 9 & 71831 & \checkmark & \checkmark & \checkmark &  \\
-OHIF Viewer \citep{Ziegler2020} & 2015 & 2020-10 &  & 76 & 63951 & \checkmark & \checkmark & \checkmark & \checkmark \\
-SMILI \citep{Chandra2018} & 2014 & 2020-06 &  & 9 & 62626 & \checkmark & \checkmark & \checkmark &  \\
-INVESALIUS 3 \citep{Amorim2015} & 2009 & 2020-09 &  & 10 & 48605 & \checkmark & \checkmark & \checkmark &  \\
-dwv \citep{Martelli2021} & 2012 & 2020-09 &  & 22 & 47815 & \checkmark & \checkmark & \checkmark & \checkmark \\
-DICOM Viewer \citep{Afsar2021} & 2018 & 2020-04 & \checkmark & 5 & 30761 & \checkmark & \checkmark & \checkmark &  \\
-MicroView \citep{ParallaxInnovations2020} & 2015 & 2020-08 &  & 2 & 27470 & \checkmark & \checkmark & \checkmark &  \\
-MatrixUser \citep{Liu2016} & 2013 & 2018-07 &  & 1 & 23121 & \checkmark & \checkmark & \checkmark &  \\
-Slice:Drop \citep{Haehn2013} & 2012 & 2020-04 &  & 3 & 19020 & \checkmark & \checkmark & \checkmark & \checkmark \\
-dicompyler \citep{Panchal2010} & 2009 & 2020-01 &  & 2 & 15941 & \checkmark & \checkmark &  &  \\
-Fiji \citep{Schindelin2012} & 2011 & 2020-08 & \checkmark & 55 & 10833 & \checkmark & \checkmark & \checkmark &  \\
-ImageJ \citep{Rueden2017} & 1997 & 2020-08 & \checkmark & 18 & 9681 & \checkmark & \checkmark & \checkmark &  \\
-MRIcroGL \citep{Rorden2021} & 2015 & 2020-08 &  & 2 & 8493 & \checkmark & \checkmark & \checkmark &  \\
-DicomBrowser \citep{Archie2012} & 2012 & 2020-08 &  & 3 & 5505 & \checkmark & \checkmark & \checkmark &  \\ \bottomrule
+Weasis \cite{Roduit2021} & 2010 & 2020-08 &  & 8 & 123272 & \checkmark & \checkmark & \checkmark &  \\
+AMIDE \cite{Loening2017} & 2006 & 2017-01 &  & 4 & 102827 & \checkmark & \checkmark & \checkmark &  \\
+XMedCon \cite{Nolf2003} & 2000 & 2020-08 &  & 2 & 96767 & \checkmark & \checkmark & \checkmark &  \\
+ITK-SNAP \cite{Yushkevich2006} & 2006 & 2020-06 & \checkmark & 13 & 88530 & \checkmark & \checkmark & \checkmark &  \\
+Papaya \cite{UTHSCSA2019} & 2012 & 2019-05 &  & 9 & 71831 & \checkmark & \checkmark & \checkmark &  \\
+OHIF Viewer \cite{Ziegler2020} & 2015 & 2020-10 &  & 76 & 63951 & \checkmark & \checkmark & \checkmark & \checkmark \\
+SMILI \cite{Chandra2018} & 2014 & 2020-06 &  & 9 & 62626 & \checkmark & \checkmark & \checkmark &  \\
+INVESALIUS 3 \cite{Amorim2015} & 2009 & 2020-09 &  & 10 & 48605 & \checkmark & \checkmark & \checkmark &  \\
+dwv \cite{Martelli2021} & 2012 & 2020-09 &  & 22 & 47815 & \checkmark & \checkmark & \checkmark & \checkmark \\
+DICOM Viewer \cite{Afsar2021} & 2018 & 2020-04 & \checkmark & 5 & 30761 & \checkmark & \checkmark & \checkmark &  \\
+MicroView \cite{ParallaxInnovations2020} & 2015 & 2020-08 &  & 2 & 27470 & \checkmark & \checkmark & \checkmark &  \\
+MatrixUser \cite{Liu2016} & 2013 & 2018-07 &  & 1 & 23121 & \checkmark & \checkmark & \checkmark &  \\
+Slice:Drop \cite{Haehn2013} & 2012 & 2020-04 &  & 3 & 19020 & \checkmark & \checkmark & \checkmark & \checkmark \\
+dicompyler \cite{Panchal2010} & 2009 & 2020-01 &  & 2 & 15941 & \checkmark & \checkmark &  &  \\
+Fiji \cite{Schindelin2012} & 2011 & 2020-08 & \checkmark & 55 & 10833 & \checkmark & \checkmark & \checkmark &  \\
+ImageJ \cite{Rueden2017} & 1997 & 2020-08 & \checkmark & 18 & 9681 & \checkmark & \checkmark & \checkmark &  \\
+MRIcroGL \cite{Rorden2021} & 2015 & 2020-08 &  & 2 & 8493 & \checkmark & \checkmark & \checkmark &  \\
+DicomBrowser \cite{Archie2012} & 2012 & 2020-08 &  & 3 & 5505 & \checkmark & \checkmark & \checkmark &  \\ \bottomrule
 \end{tabular}
@@
 benefits most parts of the software's life cycle, such as designing, coding,
-debugging, and optimization \citep{Hamill2004}.  We only found evidence of unit
+debugging, and optimization \cite{Hamill2004}.  We only found evidence of unit
 testing for about half of the projects. We identified five projects using CI/CD
@@
 medical images that can be exchanged with the data and quality necessary for
-clinical use'' \citep{MITA2021}. According to their documentation, all 29
+clinical use'' \cite{MITA2021}. According to their documentation, all 29
 software packages should support the DICOM standard. To test robustness, we
@@
 repositories, versus what is typically recommended for research software.  We
-discuss threats to the validity of our data and conclusions in section
-(Section~\ref{sec_threats_to_validity}).
+provide details on artifacts that are recommended but rarely observed for MI
+software. Section~\ref{sec_threats_to_validity} presents threats to the validity
+of our data and conclusions.
 
@@
 measure because stars represent the community's feeling in the past more than
-they measure current preferences \citep{Szulik2017}.  The issue with stars is
+they measure current preferences \cite{Szulik2017}.  The issue with stars is
 that they tend only to be added, not removed.  A final reason for
@@
 into categories of common (20 to 29 ($>$67\%) packages), uncommon (10 to 19
-(33-67\%) packages), and rare (1 to 9 ($<$33\%) packages). \citet{Dong2021-Data}
-summarizes the full measurements.  Tables~\ref{tab_maintainability_docs}
-and~\ref{tab_Visibility/Transparency_docs} show the details on which projects
-use which types of artifacts for documents related to maintainability and
-visibility, respectively.
+(33-67\%) packages), and rare (1 to 9 ($<$33\%) packages).
+Tables~\ref{tab_maintainability_docs} and~\ref{tab_Visibility/Transparency_docs}
+show the details on which projects use which types of artifacts for documents
+related to maintainability and visibility, respectively.
 
@@
 packages fall short of current best practices. This is not intended to be a
-criticism of any existing packages, especially since in practice not every
-project needs to achieve the highest possible quality. However, rather than
-delve into the nuances of which software can justify compromising which
-practices we will write our comparison under the ideal assumption that every
-project has sufficient resources to match best practices.
-
-Table~\ref{Tbl_Guidelines} (based on data from \citep{SmithAndMichalski2022})
+criticism of any existing packages, especially since for practical reasons not
+every project has the resources to achieve the highest possible quality.
+However, rather than delve into the nuances of which software can justify
+compromising which practices, we will write our comparison under the ideal
+assumption that every project has sufficient resources to match best practices.
+
+Table~\ref{Tbl_Guidelines} (based on data from \cite{SmithAndMichalski2022})
 shows that MI artifacts generally match the recommendations found in nine
@@
 \item United States Geological Survey Software Planning Checklist
-\citep{USGS2019},
+\cite{USGS2019},
 \item DLR (German Aerospace Centre) Software Engineering Guidelines
-\citep{TobiasEtAl2018}, 
+\cite{TobiasEtAl2018}, 
 \item Scottish Covid-19 Response Consortium Software Checklist
-\citep{BrettEtAl2021},
-\item Good Enough Practices in Scientific Computing \citep{WilsonEtAl2016},
+\cite{BrettEtAl2021},
+\item Good Enough Practices in Scientific Computing \cite{WilsonEtAl2016},
 \item xSDK (Extreme-scale Scientific Software Development Kit) Community Package
-Policies \citep{SmithAndRoscoe2018},
-\item Trilinos Developers Guide \citep{HerouxEtAl2008},
+Policies \cite{SmithAndRoscoe2018},
+\item Trilinos Developers Guide \cite{HerouxEtAl2008},
 \item EURISE (European Research Infrastructure Software Engineers') Network
-Technical Reference \citep{ThielEtAl2020},
+Technical Reference \cite{ThielEtAl2020},
 \item CLARIAH (Common Lab Research Infrastructure for the Arts and Humanities)
-Guidelines for Software Quality \citep{vanGompelEtAl2016}, and
+Guidelines for Software Quality \cite{vanGompelEtAl2016}, and
 \item A Set of Common Software Quality Assurance Baseline Criteria for Research
-Projects \citep{OrvizEtAl2017}.
+Projects \cite{OrvizEtAl2017}.
 \end{itemize}
@@
 terminology and the contents of artifacts are not standardized.  The challenge
-even exists for the ubiquitous README file.  As illustrated by
-\citet{PranaEtAl2018}, the content of README files shows significant variation
-between projects.  Although some content is reasonably consistent, with 97\% of
-README files contain at least one section describing the `What' of the
-repository and 89\% offering some `How' content, other categories are more
-variable.  For instance, information on `Contribution', `Why', and `Who',
-appear in 28\%, 26\% and 53\% of the analyzed files, respectively
-\citep{PranaEtAl2018}.
+even exists for the ubiquitous README file.  The content of README files shows
+significant variation between projects \cite{PranaEtAl2018}.  Although some
+content is reasonably consistent, with 97\% of README files contain at least one
+section describing the `What' of the repository and 89\% offering some `How'
+content, other categories are more variable.  For instance, information on
+`Contribution', `Why', and `Who', appear in 28\%, 26\% and 53\% of the analyzed
+files, respectively \cite{PranaEtAl2018}.
 
@@
 guideline does not explicitly recommend an artifact, does not mean the guideline
-authors do not value that artifact.  They may have excluded it because it is out
-of the scope of their recommendations, or outside their experience.  For
-instance, an artifact related to uninstall is only explicitly mentioned by
-\citet{vanGompelEtAl2016}, but other guideline authors would likely see its
-value.  They may simply feel that uninstall is implied by install, or they may
-have never asked themselves whether they need separate uninstall instructions.
+authors do not value it.  They may have excluded it because it is out of the
+scope of their recommendations, or outside their experience.  For instance, an
+artifact related to uninstall is only explicitly mentioned by one set of
+guidelines \cite{vanGompelEtAl2016}, but other guideline authors would likely
+see its value.  They may simply feel that uninstall is implied by install, or
+they may have never asked themselves whether they need separate uninstall
+instructions.
 
@@
 \toprule
-~ \ & \citet{USGS2019} & \citet{TobiasEtAl2018} & \citet{BrettEtAl2021} &
-\citet{WilsonEtAl2016} & \citet{SmithAndRoscoe2018} & \citet{HerouxEtAl2008} &
-\citet{ThielEtAl2020} & \citet{vanGompelEtAl2016} & \citet{OrvizEtAl2017} & MI\\
+~ \ & \cite{USGS2019} & \cite{TobiasEtAl2018} & \cite{BrettEtAl2021} &
+\cite{WilsonEtAl2016} & \cite{SmithAndRoscoe2018} & \cite{HerouxEtAl2008} &
+\cite{ThielEtAl2020} & \cite{vanGompelEtAl2016} & \cite{OrvizEtAl2017} & MI\\
 \midrule
@@
 Test plans describe the scope, approach, resources, and the schedule of planned
-test activities \citep{VanVliet2000}.  The plan should cover details such as the
+test activities \cite{VanVliet2000}.  The plan should cover details such as the
 makeup of the testing team, automated testing tools and technology to employ,
@@
 constraints, external interfaces and other quality attributes of the software
-\citep{IEEE1998}.  If developers write requirements they typically are based on
-a template, which provide documentation structure, guidelines, and rules.
-Although there is no universally accepted template, examples include
-\citet{ESA1991}, \citet{IEEE1998}, \citet{NASA1989}, and
-\citet{RobertsonAndRobertson1999Vol}.
+\cite{IEEE1998}.  If developers write requirements they typically are based on a
+template, which provide documentation structure, guidelines, and rules.
+Although there is no universally accepted template, examples exist
+\cite{ESA1991, IEEE1998, NASA1989, RobertsonAndRobertson1999Vol}.
 
@@
 documentation.  Although requirements documentation is recommended by some
-\citep{TobiasEtAl2018, HerouxEtAl2008, SmithAndKoothoor2016}, in practice
+\cite{TobiasEtAl2018, HerouxEtAl2008, SmithAndKoothoor2016}, in practice
 research software developers often do not produce a proper requirements
-specification \citep{HeatonAndCarver2015}. \citet{SandersAndKelly2008}
-interviewed 16 scientists from 10 disciplines and found that none of the
-scientists created requirements specifications, unless regulations in their
-field mandated such a document. \citet{Nguyen-HoanEtAl2010} showed requirements
-are the least commonly produced type of documentation for research software in
-general. When looking at the pain points for research software developers,
-\citet{WieseEtAl2019} found that software requirements and management is the
-software engineering discipline that most hurts scientific developers,
-accounting for 23\% of the technical problems reported by study participants.
-The lack of support for requirements is likely due to the perception that
-up-front requirements are impossible for research software
-\citep{CarverEtAl2007, SegalAndMorris2008}, but if we drop the insistence on
-``up-front'' requirements, allowing instead for the requirements to be written
-iteratively and incrementally, requirements are feasible \citep{Smith2016}.
-\citet{SmithEtAl2007} provides a requirements template tailored to research
-software.
+specification \cite{HeatonAndCarver2015}. Sanders and Kelly
+\cite{SandersAndKelly2008} interviewed 16 scientists from 10 disciplines and
+found that none of the scientists created requirements specifications, unless
+regulations in their field mandated such a document.  Requirements are the least
+commonly produced type of documentation for research software in general
+\cite{Nguyen-HoanEtAl2010}. When looking at the pain points for research
+software developers, software requirements and management is the software
+engineering discipline that most hurts scientific developers, accounting for
+23\% of the technical problems \cite{WieseEtAl2019}.  The lack of support for
+requirements is likely due to the perception that up-front requirements are
+impossible for research software \cite{CarverEtAl2007, SegalAndMorris2008}, but
+if we drop the insistence on ``up-front'' requirements, allowing instead for the
+requirements to be written iteratively and incrementally, requirements are
+feasible \cite{Smith2016}.  Smith et al. provide a requirements template
+tailored to research software \cite{SmithEtAl2007}.
 
@@
 
-\item \textbf{A Contributing file} provides new contributors with the information
-that they need to start adding/modifying the repository's files.
-\citet{Abdalla2016} provides a simple template for creating an open-source
-contributor guideline.
+\item \textbf{A Contributing file} provides new contributors with the
+information that they need to start adding/modifying the repository's files.
+Abdalla \cite{Abdalla2016} provides a simple template for creating an
+open-source contributor guideline.
 
 \item \textbf{A Developer Code of Conduct} explicitly states the expectations
-for developers on how they should treat one another \citep{TouraniEtAl2017}. The
+for developers on how they should treat one another \cite{TouraniEtAl2017}. The
 code outlines rules for communication and establishes enforcement mechanisms for
-violations.  As \citet{TouraniEtAl2017} states, the developer code documents
-the spirit of a community so that anyone can comfortably contribute regardless
-of ethnicity, gender, or sexual orientation. Three popular codes of conduct are
-\citep{TouraniEtAl2017}:
+violations.  As Tourani et al.\ state, the developer code documents the spirit
+of a community so that anyone can comfortably contribute regardless of
+ethnicity, gender, or sexual orientation. Three popular codes of conduct are
+\cite{TouraniEtAl2017}:
 \href{https://www.contributor-covenant.org/version/2/1/code_of_conduct/}
@@
 brings the benefit of a wider pool of contributors.  For example, a code of
-conduct can improve the participation of women \citep{SinghEtAl2021}. A standard
+conduct can improve the participation of women \cite{SinghEtAl2021}. A standard
 of ethical behaviour can be captured in the code, for projects that are looking
-to abide by a code of ethics, such as the IEEE Code of Ethics \citep{IEEE1999},
-or the Professional Engineers of Ontario code of ethics \citep[p.\
+to abide by a code of ethics, such as the IEEE Code of Ethics \cite{IEEE1999},
+or the Professional Engineers of Ontario code of ethics \cite[p.\
 23--24]{PEO2021}.
@@
 guides codify such elements as formatting, commenting, naming identifiers, best
-practices and dangers to avoid \citep{Carty2020}. For instance, most coding
+practices and dangers to avoid \cite{Carty2020}. For instance, most coding
 style guides will specify using ALLCAPS when naming symbolic constants.
@@
 \item \textbf{A Product Roadmap} explains the vision and direction for a product
-offering \citep{MunchEtAl2019}.  Although they have different forms, all
+offering \cite{MunchEtAl2019}.  Although they have different forms, all
 roadmaps cover the following: \begin{inparaenum}[i)]
@@
 	\item Where are we going?, and
-	\item How can we get there? \end{inparaenum} \citep{PhaalEtAl2005}. As
-stated by \citet{Pichler2012}, a product roadmap provides the following
-benefits: continuity of purpose, facilitation of collaboration, and assistance
-with prioritization. Creating a roadmap involves the following steps:
+	\item How can we get there? \end{inparaenum} \cite{PhaalEtAl2005}. A product
+roadmap provides the following benefits: continuity of purpose, facilitation of
+collaboration, and assistance with prioritization \cite{Pichler2012}. Creating a
+roadmap involves the following steps:
 \begin{inparaenum}[i)]
@@
 	\item estimate the product life cycle and evaluate the mix of planned
-development efforts \end{inparaenum} \citep{VahaniittyEtAl2002}.
+development efforts \end{inparaenum} \cite{VahaniittyEtAl2002}.
 
@@
 	Modelling Language class diagrams. This approach is suited to
-	object-oriented design and designs that use patterns \citep{Gamma1995}.
+	object-oriented design and designs that use patterns \cite{Gamma1995}.
 	\item Rigorous documentation of the system design following the template for
-	a Module Guide (MG) \citep{ParnasEtAl1984}.  An MG organizes the modules in
+	a Module Guide (MG) \cite{ParnasEtAl1984}.  An MG organizes the modules in
 	a hierarchy by their secrets.
@@
 	and state transitions for each module using a Module Interface Specification
-	(MIS) \citep{HoffmanAndStrooper1995}. An MIS is an abstract model that
+	(MIS) \cite{HoffmanAndStrooper1995}. An MIS is an abstract model that
 	formally shows each module's access programs and the associated transitions
 	and outputs based on their state, environment, and input variables.
-	\citet{ElSheikhEtAl2004, SmithAndYu2009} show the example of an MIS for a
-	mesh generator.
+	Previous work shows the example of an MIS for a mesh generator
+	\cite{ElSheikhEtAl2004, SmithAndYu2009}.
 \end{itemize}
@@
 by the software application (or library) through such resources as its methods
-or objects \citep{MengEtAl2018}.  Understandability is improved by API
-documentation \citep{MengEtAl2018}. API documentation can be generated using
+or objects \cite{MengEtAl2018}.  Understandability is improved by API
+documentation \cite{MengEtAl2018}. API documentation can be generated using
 tools like Doxygen, pydoc, and javadoc.
@@
 The rare artifacts for MI software are similar to the rare artifacts for Lattice
-Boltzmann solvers \citep{Michalski2021}, except LBM software is more likely to
+Boltzmann solvers \cite{Michalski2021}, except LBM software is more likely to
 have developer related artifacts, like Contributing, Dependency list, and Design
@@
 Developers can use checklists to ensure they follow best practices.  Some
-examples include checklists merging branches into master \citep{Brown2015},
-checklists for saving and sharing changes to the project \citep{WilsonEtAl2016},
-checklists for new and departing team members \citep{HerouxAndBernholdt2018},
-checklists for processes related to commits and releases \citep{HerouxEtAl2008}
-and checklists for overall software quality \citep{ThielEtAl2020, SSI2022}.  For
-instance, for Lattice Boltzmann solver software, ESPResSo has a checklist for
-managing releases \citep{Michalski2021}.
+examples include checklists merging branches into master \cite{Brown2015},
+checklists for saving and sharing changes to the project \cite{WilsonEtAl2016},
+checklists for new and departing team members \cite{HerouxAndBernholdt2018},
+checklists for processes related to commits and releases \cite{HerouxEtAl2008}
+and checklists for overall software quality \cite{ThielEtAl2020, SSI2022}.
 
@@
 practices.  A gap exists in research software development practices and software
-engineering recommendations \citep{Storer2017, Kelly2007, OwojaiyeEtAl2021_CSE}.
-\citet{JohansonAndHasselbring2018} observe that the state-of-the-practice for
-research software in industry and academia does not incorporate state-of-the-art
-SE tools and methods.  This causes sustainability and reliability problems
-\citep{FaulkEtAl2009}. Rather than benefit from capturing and reusing previous
-knowledge, projects waste time and energy ``reinventing the wheel''
-\citep{deSouzaEtAl2019}.
+engineering recommendations \cite{Storer2017, Kelly2007, OwojaiyeEtAl2021_CSE}.
+Johanson and Hasselbring observe that the state-of-the-practice for research
+software in industry and academia does not incorporate state-of-the-art SE tools
+and methods \cite{JohansonAndHasselbring2018}.  This causes sustainability and
+reliability problems \cite{FaulkEtAl2009}. Rather than benefit from capturing
+and reusing previous knowledge, projects waste time and energy ``reinventing the
+wheel'' \cite{deSouzaEtAl2019}.
 
@@
 Our categories come from an analysis of software engineering secondary studies
-by \citet{AmpatzoglouEtAl2019}, where a secondary study analyzes the data from a
-set of primary studies.  \citet{AmpatzoglouEtAl2019} is appropriate because a
-common example of a secondary study is a systematic literature review. Our
-methodology is a systematic software review --- the primary studies are the
-software packages, and our work collects and analyzes these primary studies.  We
-identified similar threats to validity in our assessment of the state of the
-practice of Lattice Boltzmann Solvers \citep{SmithEtAl2024}.
+by Ampatzoglou et al.\ \cite{AmpatzoglouEtAl2019}, where a secondary study
+analyzes the data from a set of primary studies.  Ampatzoglou et al.\ is
+appropriate because a common example of a secondary study is a systematic
+literature review. Our methodology is a systematic software review --- the
+primary studies are the software packages, and our work collects and analyzes
+these primary studies.  We identified similar threats to validity in our
+assessment of the state of the practice of Lattice Boltzmann Solvers
+\cite{SmithEtAl2024}.
 
@@
 the original study's methodology would lead to the same results
-\citep{RunesonAndHost2009}. Reliability means that data and analysis are
+\cite{RunesonAndHost2009}. Reliability means that data and analysis are
 independent of the specific researcher(s) doing the study.  For the current
@@
 
-In \citet{SmithEtAl2016} we reduced concern over the reliability risk associated
-with the reviewer's judgement by demonstrating that the measurement process is
-reasonably reproducible.  In \citet{SmithEtAl2016} we graded five software
-products by two reviewers. Their rankings were almost identical. As long as each
-grader uses consistent definitions, the relative comparisons in the AHP results
-will be consistent between graders.
+We reduced concern over the reliability risk associated with the reviewer's
+judgement by demonstrating that the measurement process is reasonably
+reproducible \cite{SmithEtAl2016}.  In previous work \cite{SmithEtAl2016} we
+graded five software products by two reviewers. Their rankings were almost
+identical. As long as each grader uses consistent definitions, the relative
+comparisons in the AHP results will be consistent between graders.
 
@@
 
-\citet{RunesonAndHost2009} defines construct validity as the adopted
-metrics representing what they are intended to measure. Our construct threats are
-often related to how we assume our measurements influences the various software
-qualities, as summarized in Section~\ref{sec_grading_software}. Specifically,
-our construct validity related threats include the following:
+Runeson and Host \cite{RunesonAndHost2009} define construct validity as the
+adopted metrics representing what they are intended to measure. Our construct
+threats are often related to how we assume our measurements influences the
+various software qualities, as summarized in Section~\ref{sec_grading_software}.
+Specifically, our construct validity related threats include the following:
 
@@
 unavailable.  We follow the usual assumption that developers achieve higher
-quality by following procedures and adhering to standards \citep[p.\
+quality by following procedures and adhering to standards \cite[p.\
 112]{VanVliet2000}.
@@
 Internal validity means that discovered causal relations are trustworthy and
-cannot be explained by other factors \citep{RunesonAndHost2009}. In our
+cannot be explained by other factors \cite{RunesonAndHost2009}. In our
 methodology the internal validity threats include the following:
@@
 If the results of a study can be generalized (applied) to other
-situations/cases, then the study is externally valid \citep{RunesonAndHost2009}.
+situations/cases, then the study is externally valid \cite{RunesonAndHost2009}.
 We are confident that our search was exhaustive.  We do not believe that we
@@
 Although there are differences, like the importance of data privacy for MI data,
-we found the approach to developing LBM software \citep{SmithEtAl2024} and MI
+we found the approach to developing LBM software \cite{SmithEtAl2024} and MI
 software to be similar.  Except for the domain specific aspects, we believe that
@@
 
-\bibliographystyle{ACM-Reference-Format}
+\bibliographystyle{plain}
 \bibliography{SOP_MI_OSR}

### 2024-06-12 StateOfPractice/Papers/SOP_MI_OSR/SOP_MI_OSR.tex
Move to Springer LaTeX, shortened paper by removing unnecessary citations, unnecessary text

@@
-\documentclass[11pt]{article}
-
-\usepackage{authblk}
-\usepackage{fancyhdr}
-\usepackage{float}
-
-%\usepackage[round, sort]{natbib}
+%\documentclass[11pt]{article}
+\documentclass[sn-mathphys-num]{sn-jnl}
+
+\usepackage{booktabs}
 
@@
 
-\oddsidemargin 0mm
-\evensidemargin 0mm
-\textwidth 165mm
-\textheight 205mm
-
-\pagestyle{fancy}
-\lhead{\leftmark}
-\chead{}
-\rhead{\rightmark}
-\cfoot{\thepage}
-
 \newcounter{rqnum} %research question number
@@
 
-\title{State of the Practice for Medical Imaging Software Based on Open Source Repositories}
-
-\author[1,*]{W.\ Spencer Smith}
-\author[1]{Ao Dong}
-\author[1]{Jacques Carette}
-\author[2]{Michael D.\ Noseworthy}
-
-\affil[1]{McMaster University, Computing and Software Department, {Canada}}
-\affil[2]{McMaster University, Electrical \& Computer Engineering Department,
-{Canada}}
-\affil[*]{Corresponding Author}
-
-\maketitle
-
-% RQ1 to RQ4
-% Some of Section 1 (Introduction)
-% Sections 2 to 6 (includes looking at artifacts observed in repos)
-% Some of Sections 11 to 13 (Threats to Validity, Future Work, Conclusions)
-
-\begin{abstract}
-
+\title[SOP for MI Software]{State of the Practice for Medical Imaging Software
+Based on Open Source Repositories}
+
+\author*[1]{\fnm{W.\ Spencer} \sur{Smith}}\email{smiths@mcmaster.ca}
+
+\author[1]{\fnm{Ao} \sur{Dong}}
+
+\author[1]{\fnm{Jacques} \sur{Carette}}\email{carette@mcmaster.ca}
+
+\author[2]{\fnm{Michael} \sur{Noseworthy}}\email{nosewor@mcmaster.ca}
+
+\affil*[1]{\orgdiv{Computing and Software Department}, \orgname{McMaster University}, \orgaddress{\street{1280 Main Street West}, \city{Hamilton}, \postcode{L8S 4K1}, \state{Ontario}, \country{Canada}}}
+
+\affil[2]{\orgdiv{Electrical and Computer Engineering Department}, \orgname{McMaster University}, \orgaddress{\street{1280 Main Street West}, \city{Hamilton}, \postcode{L8S 4K1}, \state{Ontario}, \country{Canada}}}
+
+% \author[1,*]{W.\ Spencer Smith}
+% \author[1]{Ao Dong}
+% \author[1]{Jacques Carette}
+% \author[2]{Michael D.\ Noseworthy}
+
+% \affil[1]{McMaster University, Computing and Software Department, {Canada}}
+% \affil[2]{McMaster University, Electrical \& Computer Engineering Department,
+% {Canada}}
+% \affil[*]{Corresponding Author}
+
+\abstract{
 We present the state of the practice for Medical Imaging (MI) software based on
@@
 Application Program Interface (API) documentation).
-
-\end{abstract}
-
-\noindent \emph{Keywords:}
+}
+
+\keywords{
 	medical imaging, research software, software engineering, software
-	quality, analytic hierarchy process
+	quality, analytic hierarchy process}
+
+\maketitle
+
+% RQ1 to RQ4
+% Some of Section 1 (Introduction)
+% Sections 2 to 6 (includes looking at artifacts observed in repos)
+% Some of Sections 11 to 13 (Threats to Validity, Future Work, Conclusions)
 
@@
 Ultrasound) to provide information for diagnostic, analytic, and medical
-applications \cite{FDA2021, enwiki:1034887445, Zhang2008}.
-Figure~\ref{Fig_Example}, which shows an image of the brain, highlights the
-importance and value of MI. Through MI medical practitioners and researchers can
-noninvasively gain insights into the human body, including information on
-injuries and illnesses. Given the importance of MI software and the high number
-of competing software projects, we wish to understand the merits and drawbacks
-of the current development processes, tools, and methodologies. We aim to assess
-through a software engineering lens the quality of the existing software with
-the hope of highlighting standout examples, and providing guidelines and
-recommendations for future development.
-
-\begin{figure}[!ht]
-    \begin{center}
-        \includegraphics[scale=0.25]{MPR.png}        
-    \end{center}
-    \caption{Example brain image showing a multi-planar reformat using Horos
-	(free open-source medical imaging/DICOM viewer for OSX, based on OsiriX)}
-    \label{Fig_Example}
-\end{figure}
+applications. Through MI medical practitioners and
+researchers can noninvasively gain insights into the human body, including
+information on injuries and illnesses. Given the importance of MI software and
+the high number of competing software projects, we wish to understand the merits
+and drawbacks of the current development processes, tools, and methodologies. We
+aim to assess through a software engineering lens the quality of the existing
+software with the hope of highlighting standout examples, and providing
+guidelines and recommendations for future development.
 
@@
 software, we also wish to understand the impact of the often cited gap, or
-chasm, between software engineering and research software \cite{Kelly2007,
-Storer2017}. Although scientists spend a substantial proportion of their working
+chasm, between software engineering and research software \cite{Storer2017}. Although scientists spend a substantial proportion of their working
 hours on software development \cite{Hannay2009, Prabhu2011}, many developers
@@
 Simulation, plus MI archiving and telemedicine systems (Compression, Storage,
-and Communication) \cite{Bankman2000, Angenent2006}.  We also exclude
-Statistical Analysis and Image-based Physiological Modelling
-\cite{enwiki:1034877594} and Feature Extraction, Classification, and
-Interpretation \cite{Kim2011}. Software that provides MI support functions is
-also out of scope; therefore, we have not assessed the toolkit libraries VTK
-\cite{SchroederEtAl2006} and ITK \cite{McCormick2014}.  Finally, Picture
-Archiving and Communication System (PACS), which helps users to economically
-store and conveniently access images \cite{Choplin1992}, are considered out of
-scope.
+and Communication).  We also exclude Statistical Analysis and Image-based
+Physiological Modelling and Feature Extraction, Classification, and
+Interpretation. Software that provides MI support functions is also out of
+scope; therefore, we have not assessed the toolkit libraries VTK and ITK.
+Finally, Picture Archiving and Communication System (PACS), which helps users to
+economically store and conveniently access images, are considered out of scope.
 
@@
 research software \cite{SmithEtAl2021, SmithAndMichalski2022}.  Our methodology
-has been applied to MI software \cite{Dong2021} and Lattice Boltzmann Solvers
-\cite{Michalski2021, SmithEtAl2024}.  This methodology builds off prior work to
-assess the state of the practice for such domains as Geographic Information
-Systems \cite{smith2018state}, Mesh Generators \cite{smith2016state},
-Seismology software \cite{Smith2018Seismology}, and Statistical software for
-psychology \cite{smith2018statistical}.  In keeping with the previous
-methodology, we have maintained the constraint that the work load for measuring
-a given domain should be feasible for a team as small as one person, and for a
-short time, ideally around a person month of effort. We consider a person month
-as $20$ working days ($4$ weeks in a month, with $5$ days of work per week) at
-$8$ person hours per day, or $20 \times 8 = 160$ person hours.
+has been applied to MI software and Lattice Boltzmann Solvers
+\cite{SmithEtAl2024}.  This methodology builds off prior work to assess the
+state of the practice for such domains as Geographic Information Systems
+\cite{smith2018state}, Mesh Generators \cite{smith2016state}, Seismology
+software \cite{Smith2018Seismology}, and Statistical software for psychology
+\cite{smith2018statistical}.  In keeping with the previous methodology, we have
+maintained the constraint that the work load for measuring a given domain should
+be feasible for a team as small as one person, and for a short time, ideally
+around a person month of effort. We consider a person month as $20$ working days
+($4$ weeks in a month, with $5$ days of work per week) at $8$ person hours per
+day, or $20 \times 8 = 160$ person hours.
 
@@
 required for the installation and/or uninstallation of software in a specified
-environment \cite{ISO/IEC25010, lenhard2013measuring}.
+environment \cite{ISO/IEC25010}.
 
@@
 \begin{figure}[!ht]
-\includegraphics[scale=0.66]{template.pdf}
+\includegraphics[scale=0.53]{template.pdf}
 \caption{Grading template example}
@@
 We use two freeware tools to collect repository related data:
-\href{https://github.com/tomgi/git_stats}{GitStats} \cite{Gieniusz2019} and
-\href{https://github.com/boyter/scc}{Sloc Cloc and Code (scc)}
-\cite{Boyter2021}.  Details on how we measure each quality are provided in
-Smith et al.\ (2024) \cite{SmithEtAl2024_MI_SOP}.
+\href{https://github.com/tomgi/git_stats}{GitStats} and
+\href{https://github.com/boyter/scc}{Sloc Cloc and Code (scc)}.  Details on how
+we measure each quality are provided in Smith et al.\ (2024)
+\cite{SmithEtAl2024_MI_SOP}.
 
@@
 \centering
-\begin{tabular}{p{6cm}lllllllll}
+\begin{tabular}{p{3.7cm}lllllllll}
 \toprule
-\multirow{2}{*}{Software} & \multirow{2}{*}{Rlsd} & \multirow{2}{*}{Updated} & \multirow{2}{*}{Fnd} & \multirow{2}{*}{NOC} & \multirow{2}{*}{LOC} & \multicolumn{3}{c}{OS} & \multirow{2}{*}{Web} \\ \cline{7-9}
+\multirow{2}{*}{Software} & \multirow{2}{*}{Rlsd} & \multirow{2}{*}{Updated} & \multirow{2}{*}{Fnd} & \multirow{2}{*}{NOC} & \multirow{2}{*}{LOC} & \multicolumn{3}{c}{OS} & \multirow{2}{*}{Web} \\ \cmidrule{7-9}
  &  &  &  &  &  & W & M & L &  \\ \midrule
@@
 
-Figure \ref{fig_language} shows the primary languages versus the number of
-projects using them.  The primary language is the language used for the majority
-of the project's code; in most cases projects also use other languages.  The
-most popular language is \CC, with almost 40\% of projects (11 of 29).  The two
-least popular choices are Pascal and Matlab, with around 3\% of projects each
-(1 of 29).
-
-\begin{figure}[!ht]
-\centering
-\includegraphics[scale=0.5]{PrimaryLanguages.pdf}
-\centering
-\caption{\label{fig_language}Primary languages versus number of projects}
-\end{figure}
+The programming languages used in order of decreasing popularity are \CC,
+JavaScript, Java, C, Python, Pascal, Matlab.  The most popular language is \CC,
+with almost 40\% of projects (11 of 29).  The two least popular choices are
+Pascal and Matlab, with around 3\% of projects each (1 of 29).
 
@@
 \begin{figure}[!ht]
-\includegraphics[scale=0.47]{installability_scores.pdf}
+\includegraphics[scale=0.38]{installability_scores.pdf}
 \caption{AHP installability scores}
@@
 \begin{figure}[!ht]
-\includegraphics[scale=0.47]{correctness_verifiability_scores.pdf}
+\includegraphics[scale=0.38]{correctness_verifiability_scores.pdf}
 \caption{AHP correctness \& verifiability scores}
@@
 \begin{figure}[!ht]
-\includegraphics[scale=0.47]{reliability_scores.pdf}
+\includegraphics[scale=0.38]{reliability_scores.pdf}
 \caption{AHP surface reliability scores}
@@
 \begin{figure}[!ht]
-\includegraphics[scale=0.47]{robustness_scores.pdf}
+\includegraphics[scale=0.38]{robustness_scores.pdf}
 \caption{AHP surface robustness scores}
@@
 
-Digital Imaging and Communications in Medicine (DICOM) ``defines the formats for
-medical images that can be exchanged with the data and quality necessary for
-clinical use'' \cite{MITA2021}. According to their documentation, all 29
-software packages should support the DICOM standard. To test robustness, we
-prepared two types of image files: correct and incorrect formats (with the
-incorrect format created by relabelled a text file to have the ``.dcm''
-extension).  All software packages loaded the correct format image, except for
-\textit{GATE}, which failed for unknown reasons.  For the broken format,
-\textit{MatrixUser}, \textit{dwv}, and \textit{Slice:Drop} ignored the incorrect
-format of the file and loaded it regardless. They did not show any error message
-and displayed a blank image. \textit{MRIcroGL} behaved similarly except that it
-showed a meaningless image. \textit{Drishti} successfully detected the broken
-format of the file, but the software crashed as a result.
+According to their documentation, all 29 software packages should support the
+DICOM standard. To test robustness, we prepared two types of image files:
+correct and incorrect formats (with the incorrect format created by relabelled a
+text file to have the ``.dcm'' extension).  All software packages loaded the
+correct format image, except for \textit{GATE}, which failed for unknown
+reasons.  For the broken format, \textit{MatrixUser}, \textit{dwv}, and
+\textit{Slice:Drop} ignored the incorrect format of the file and loaded it
+regardless. They did not show any error message and displayed a blank image.
+\textit{MRIcroGL} behaved similarly except that it showed a meaningless image.
+\textit{Drishti} successfully detected the broken format of the file, but the
+software crashed as a result.
 
@@
 \begin{figure}[!ht]
-\includegraphics[scale=0.47]{usability_scores.pdf}
+\includegraphics[scale=0.38]{usability_scores.pdf}
 \caption{AHP surface usability scores}
@@
 \begin{figure}[!ht]
-\includegraphics[scale=0.47]{maintainability_scores.pdf}
+\includegraphics[scale=0.38]{maintainability_scores.pdf}
 \caption{AHP maintainability scores}
@@
 \begin{figure}[!ht]
-\includegraphics[scale=0.47]{reusability_scores.pdf}
+\includegraphics[scale=0.38]{reusability_scores.pdf}
 \caption{AHP reusability scores}
@@
 \begin{figure}[!ht]
-\includegraphics[scale=0.47]{understandability_scores.pdf}
+\includegraphics[scale=0.38]{understandability_scores.pdf}
 \caption{AHP surface understandability scores}
@@
 \begin{figure}[!ht]
-\includegraphics[scale=0.47]{visibility_transparency_scores.pdf}
+\includegraphics[scale=0.38]{visibility_transparency_scores.pdf}
 \caption{AHP visibility/transparency scores}
@@
 \begin{figure}[!ht]
-\includegraphics[scale=0.47]{overall_scores.pdf}
+\includegraphics[scale=0.38]{overall_scores.pdf}
 \caption{Overall AHP scores with an equal weighting for all 9 software qualities}
@@
 To answer~\rqref{RQ_CompareHQ2Popular} about how our ranking compares to the
-popularity of projects as judged by the scientific community, we make two
-comparisons:
-
-\begin{itemize} 
-\item A comparison of our ranking (from Section~\ref{ch_results}) with the
-community ratings on GitHub, as shown by GitHub stars, number of forks, and
-number of people watching the projects; and,
-\item A comparison of top-rated software from our methodology with the top
-recommendations from our domain expert (as mentioned in
-Section~\ref{sec_vet_software_list}).
-\end{itemize}
+popularity of projects as judged by the scientific community, we compare our
+ranking (from Section~\ref{ch_results}) with the community ratings on GitHub, as
+shown by GitHub stars, number of forks, and number of people watching the
+projects.
 
@@
 
-% The Domain Expert part can be removed if space is needed
-As shown in Section~\ref{ch_results}, our domain experts recommended
-a list of top software with 12 software products.  All the top 4 entries from
-the Domain Expert's list are among the top 12 ranked by our methodology. Three
-of the top four on both lists are the same: \textit{3D Slicer}, \textit{ImageJ},
-and \textit{Fiji}. \textit{3D Slicer} is top project by both rankings (and by
-the GitHub stars measure as well).  The Domain Expert ranked \textit{Horos} as
-their second choice, while we ranked it twelfth.  Our third ranked project,
-\textit{OHIF Viewer} was not listed by the Domain Expert.  Neither were the
-software packages that we ranked from fifth to eleventh (\textit{ParaView},
-\textit{Weasis}, \textit{medInria}, \textit{BioImage Suite Web}, \textit{OsiriX
-Lite}, \textit{INVESALIUS}, and \textit{Gwyddion}).  The software mentioned by
-the Domain Expert that we did not rank were the six recommended packages that
-did not have visualization as the primary function (as discussed in
-Section~\ref{ch_results}).  The differences between the list
-recommended by our methodology and the Domain Expert are not surprising.  As
-mentioned above, our methodology weights all qualities equally, but that may not
-be the case for the Domain Expert's impressions.  Moreover, although the Domain
-Expert has significant experience with MI software, they have not used all 29
-packages that were measured.
-
 Although our ranking and the estimate of the community's ranking are not perfect
@@
 \begin{table}[ht!]
-    \begin{center}
-    \begin{tabular}{ p{4.6 cm} p{5.6 cm} p{5 cm}}
+    %\begin{center}
+    \begin{tabular}{ p{3.1 cm} p{5.4 cm} p{3.5 cm}}
     \toprule
@@
     \label{artifactspresent}
-    \end{center}
+    %\end{center}
 \end{table}
@@
 
-\begin{table}[H]
-\begin{center}
-\begin{tabular}{ p{2.5cm}p{1cm}p{1cm}p{1cm}p{1cm}p{1cm}p{1cm}p{1cm}p{1.2cm}p{1cm}p{0.8cm} }
+\begin{table}[h]
+%\begin{center}
+\begin{tabular}{ p{2.5cm}p{0.5cm}p{0.5cm}p{0.5cm}p{0.5cm}p{0.5cm}p{0.5cm}p{0.5cm}p{0.5cm}p{0.5cm}p{0.5cm} }
 \toprule
@@
 \label{Tbl_Guidelines}
-\end{center}
+%\end{center}
 \end{table}
@@
 Although there is no universally accepted template, examples exist
-\cite{ESA1991, IEEE1998, NASA1989, RobertsonAndRobertson1999Vol}.
+\cite{IEEE1998, NASA1989, RobertsonAndRobertson1999Vol}.
 
@@
 \item \textbf{A Developer Code of Conduct} explicitly states the expectations
-for developers on how they should treat one another \cite{TouraniEtAl2017}. The
-code outlines rules for communication and establishes enforcement mechanisms for
-violations.  As Tourani et al.\ state, the developer code documents the spirit
-of a community so that anyone can comfortably contribute regardless of
-ethnicity, gender, or sexual orientation. Three popular codes of conduct are
-\cite{TouraniEtAl2017}:
+for developers on how they should treat one another \cite{TouraniEtAl2017}.
+Three popular codes of conduct are \cite{TouraniEtAl2017}:
 \href{https://www.contributor-covenant.org/version/2/1/code_of_conduct/}
@@
 brings the benefit of a wider pool of contributors.  For example, a code of
-conduct can improve the participation of women \cite{SinghEtAl2021}. A standard
-of ethical behaviour can be captured in the code, for projects that are looking
-to abide by a code of ethics, such as the IEEE Code of Ethics \cite{IEEE1999},
-or the Professional Engineers of Ontario code of ethics \cite[p.\
-23--24]{PEO2021}.
+conduct can improve the participation of women \cite{SinghEtAl2021}.
 
@@
 guides codify such elements as formatting, commenting, naming identifiers, best
-practices and dangers to avoid \cite{Carty2020}. For instance, most coding
-style guides will specify using ALLCAPS when naming symbolic constants.
-Understandability improves under standardization, since developers spend
-more time on the content of the code, and less time distracted by its style.
-Three sample style guides are:
+practices and dangers to avoid \cite{Carty2020}.  Three sample style guides are:
 \href{https://google.github.io/styleguide/javaguide.html} {Google Java Style
@@
 roadmap provides the following benefits: continuity of purpose, facilitation of
-collaboration, and assistance with prioritization \cite{Pichler2012}. Creating a
-roadmap involves the following steps:
-\begin{inparaenum}[i)]
-	\item define and outline a strategic mission and product vision, 
-	\item scan the environment, 
-	\item revise and distill the product vision to write the product roadmap, and 
-	\item estimate the product life cycle and evaluate the mix of planned
-development efforts \end{inparaenum} \cite{VahaniittyEtAl2002}.
+collaboration, and assistance with prioritization \cite{Pichler2012}.
 
@@
 cases also documents the semantics.  Some potential elements of design
-documentation include the following:
-
-\begin{itemize}
-	\item Representation of the system design and class design using Unified
-	Modelling Language class diagrams. This approach is suited to
-	object-oriented design and designs that use patterns \cite{Gamma1995}.
-	\item Rigorous documentation of the system design following the template for
-	a Module Guide (MG) \cite{ParnasEtAl1984}.  An MG organizes the modules in
-	a hierarchy by their secrets.
-	\item An explanation of the design using data flow diagrams to show typical use cases
-	for input transformation.
-	\item A table or graph showing the traceability between the requirements and
-	the modules (or classes)
-	\item The syntax of the modules (or classes) by providing lists of the state
-	variables, exported constants and all exported access programs for each
-	module (or class).  This shows the interface that can used to access each
-	module's services.
-	\item A formal specification of the semantics of input/output relationships
-	and state transitions for each module using a Module Interface Specification
-	(MIS) \cite{HoffmanAndStrooper1995}. An MIS is an abstract model that
-	formally shows each module's access programs and the associated transitions
-	and outputs based on their state, environment, and input variables.
-	Previous work shows the example of an MIS for a mesh generator
-	\cite{ElSheikhEtAl2004, SmithAndYu2009}.
-\end{itemize}
+documentation include representation of the system design and class design using
+Unified Modelling Language class diagrams \cite{Gamma1995}, or rigorous
+documentation of the system design following the template for a Module Guide
+(MG) \cite{ParnasEtAl1984}.  An MG organizes the modules in a hierarchy by their
+secrets.
 
@@
 The rare artifacts for MI software are similar to the rare artifacts for Lattice
-Boltzmann solvers \cite{Michalski2021}, except LBM software is more likely to
+Boltzmann solvers \cite{SmithEtAl2024}, except LBM software is more likely to
 have developer related artifacts, like Contributing, Dependency list, and Design
@@
 practices.  A gap exists in research software development practices and software
-engineering recommendations \cite{Storer2017, Kelly2007, OwojaiyeEtAl2021_CSE}.
-Johanson and Hasselbring observe that the state-of-the-practice for research
-software in industry and academia does not incorporate state-of-the-art SE tools
-and methods \cite{JohansonAndHasselbring2018}.  This causes sustainability and
-reliability problems \cite{FaulkEtAl2009}. Rather than benefit from capturing
-and reusing previous knowledge, projects waste time and energy ``reinventing the
-wheel'' \cite{deSouzaEtAl2019}.
+engineering recommendations \cite{Storer2017}.  Johanson and Hasselbring observe
+that the state-of-the-practice for research software in industry and academia
+does not incorporate state-of-the-art SE tools and methods
+\cite{JohansonAndHasselbring2018}.  This causes sustainability and reliability
+problems \cite{FaulkEtAl2009}. Rather than benefit from capturing and reusing
+previous knowledge, projects waste time and energy ``reinventing the wheel''
+\cite{deSouzaEtAl2019}.
 
@@
 
-Below we categorize and list the threats to validity that we have identified.
-Our categories come from an analysis of software engineering secondary studies
-by Ampatzoglou et al.\ \cite{AmpatzoglouEtAl2019}, where a secondary study
-analyzes the data from a set of primary studies.  Ampatzoglou et al.\ is
-appropriate because a common example of a secondary study is a systematic
-literature review. Our methodology is a systematic software review --- the
-primary studies are the software packages, and our work collects and analyzes
-these primary studies.  We identified similar threats to validity in our
-assessment of the state of the practice of Lattice Boltzmann Solvers
+Our categories for threat to validity come from an analysis of software
+engineering secondary studies by Ampatzoglou et al.\ \cite{AmpatzoglouEtAl2019},
+where a secondary study analyzes the data from a set of primary studies.
+Ampatzoglou et al.\ is appropriate because a common example of a secondary study
+is a systematic literature review. Our methodology is a systematic software
+review --- the primary studies are the software packages, and our work collects
+and analyzes these primary studies.  We identified similar threats to validity
+in our assessment of the state of the practice of Lattice Boltzmann Solvers
 \cite{SmithEtAl2024}.
@@
 judgement by demonstrating that the measurement process is reasonably
-reproducible \cite{SmithEtAl2016}.  In previous work \cite{SmithEtAl2016} we
-graded five software products by two reviewers. Their rankings were almost
-identical. As long as each grader uses consistent definitions, the relative
-comparisons in the AHP results will be consistent between graders.
+reproducible.  In previous work \cite{SmithEtAl2016} we graded five software
+products by two reviewers. Their rankings were almost identical. As long as each
+grader uses consistent definitions, the relative comparisons in the AHP results
+will be consistent between graders.
 
@@
 
-Runeson and Host \cite{RunesonAndHost2009} define construct validity as the
-adopted metrics representing what they are intended to measure. Our construct
-threats are often related to how we assume our measurements influences the
-various software qualities, as summarized in Section~\ref{sec_grading_software}.
-Specifically, our construct validity related threats include the following:
+Construct validity is when the adopted metrics represent what they are intended
+to measure \cite{RunesonAndHost2009}. Our construct threats are often related to
+how we assume our measurements influences the various software qualities, as
+summarized in Section~\ref{sec_grading_software}.  Specifically, our construct
+validity related threats include the following:
 
@@
 will leave a trace in the repositories, but this is not necessarily true. For
-instance, the possibility exists that CI usage was higher than what we observed
-through the artifacts (Section~\ref{Sec_CompareTools}). As another example,
-although we saw little evidence of requirements
-(Section~\ref{Sec_CompareTools}), maybe teams keep this kind of information
-outside their repos, possibly in journal papers or technical reports.
+instance, we saw little evidence of requirements
+(Section~\ref{Sec_CompareArtifacts}), but teams might keep this kind of
+information outside their repos, possibly in journal papers or technical
+reports.
 \end{itemize}
@@
 We are confident that our search was exhaustive.  We do not believe that we
-missed any highly popular examples.  Therefore, the bulk of our validity
-concerns are internal (Section~\ref{Sec_InternalValidity}).
-However, our hope is that the trends observed, and the lessons learned for MI
-software can be applied to other research software.  With that in mind we
-identified the following threat to external validity:
-
-\begin{itemize}
-\item We cannot generalize our results if the development of MI software is
-fundamentally different from other research software.
-\end{itemize}
+missed any highly popular MI examples.  Therefore, the bulk of our validity
+concerns are internal (Section~\ref{Sec_InternalValidity}).  However, our hope
+is that the trends observed, and the lessons learned for MI software can be
+applied to other research software.  With that in mind we identified the
+following threat to external validity: We cannot generalize our results if the
+development of MI software is fundamentally different from other research
+software.
 
@@
 We analyzed the state of the practice for the MI domain with the goal of
-understanding current practice by answering our four research questions
+understanding current practice by answering four research questions
 (Section~\ref{sec_motivation}).  Our methods in Section~\ref{SecMethodology} form a
@@
 specific domain of MI software. We identified 48 MI software candidates, then,
-with the help of the Domain Expert selected 29 of them to our final list.
+with the help of the Domain Expert selected 29 of them for our final list.
 
@@
 noteworthy discrepancies between the two lists are for the packages that we were
-unable to install (\textit{dwv} and \textit{Dicom Viewer}).
+unable to install (\textit{dwv} and \textit{Dicom Viewer}). Our findings suggest
+that the community's top ranked packages tend to be those that incorporate best
+practices.
 
@@
 \begin{table}[ht!]
-	\begin{center}
+	%\begin{center}
 	\renewcommand{\arraystretch}{1.8}
-	\begin{tabular}{ p{3cm}p{13cm} }
+	\begin{tabular}{ p{2.3cm}p{9.9cm} }
 		\toprule
@@
 	measurement)} \label{topperformerstable}
-	\end{center}
+	%\end{center}
 \end{table}
 
-When it comes to recommended software artifacts, the state of open source MI projects is healthy, with our surveyed examples showing 88\%
-of the documentation artifacts recommended by research software development
-guidelines (Section~\ref{Sec_CompareArtifacts}).  However, we did notice
-areas where practice seems to lag behind the research software development
-guidelines.  For instance, the guidelines recommend three artifacts that were
-not observed: uninstall instructions, test plans, and requirements
-documentation. We observed the following recommended artifacts, but only rarely:
-contributing file, developer code of conduct, code style guidelines, product
-roadmap, design documentation, and API documentation
-(Section~\ref{Sec_CompareArtifacts}).  Future MI projects may wish to consider incorporating more artifacts to potentially improve software quality.
-
-Next steps.  Mention threats to validity?  Future work on pain points.  Point to
-LBM methodology in that SOP paper?
+When it comes to recommended software artifacts, the state of open source MI
+projects is healthy, with our surveyed examples showing 88\% of the
+documentation artifacts recommended by research software development guidelines
+(Section~\ref{Sec_CompareArtifacts}).  However, we did notice areas where
+practice seems to lag behind the research software development guidelines.  For
+instance, the guidelines recommend three artifacts that were not observed:
+uninstall instructions, test plans, and requirements documentation. We observed
+the following recommended artifacts, but only rarely: contributing file,
+developer code of conduct, code style guidelines, product roadmap, design
+documentation, and API documentation (Section~\ref{Sec_CompareArtifacts}).
+Future MI projects may wish to consider incorporating more artifacts to
+potentially improve software quality.  A deeper understanding of the needs of
+the MI community will require data beyond what is available in open source
+repositories.  Future work should involve interviewing MI developers to more
+deeply understand their software development pain points.
 
@@
 We would like to thank Peter Michalski and Oluwaseun Owojaiye for fruitful
-discussions on topics relevant to this paper.  We would also like to thank Jason
-Balaci for advice on web applications.
+discussions on topics relevant to this paper.
 
@@
 
-\bibliographystyle{plain}
 \bibliography{SOP_MI_OSR}

### 2024-06-13 StateOfPractice/Papers/SOP_MI_OSR/SOP_MI_OSR.tex
done up to section 2.3. Basically down 2 pages.

@@
 We present the state of the practice for Medical Imaging (MI) software based on
-data available in open source repositories. We selected 29 medical imaging
+data available in open source repositories. We selected 29 
 projects from 48 candidates and assessed 9 software qualities (installability,
@@
 reusability, understandability, and visibility/transparency) by answering 108
-questions for each software project. Based on the quantitative data, we ranked
-the MI software with the Analytic Hierarchy Process (AHP). The top four software
-products are \textit{3D Slicer}, \textit{ImageJ}, \textit{Fiji}, and
-\textit{OHIF Viewer}.  Our ranking is consistent with the community's
-ranking, with four of our top five projects also appearing in the top five of a
-list ordered by stars-per-year.  We observed 88\% of the documentation artifacts
-recommended by research software development guidelines.  However, the current
-state of the practice deviates from the existing guidelines because of the
-rarity of some recommended artifacts (like test plans, requirements
-specification, code of conduct, code style guidelines, product roadmaps, and
-Application Program Interface (API) documentation).
+questions for each. Using the Analytic Hierarchy Process (AHP) on the
+quantitive data, we ranked the MI software.  The top five are \textit{3D
+Slicer}, \textit{ImageJ}, \textit{Fiji}, \textit{OHIF Viewer}, and
+\textit{ParaView}.  This is consistent with the community's, with four of
+these also appearing in the top five using GitHub metrics (stars-per-year).  
+In general, the quality and quantity of documentation present in the
+project correlates quite well with its popularity.
+%We observed 88\% of the documentation artifacts
+%recommended by research software development guidelines.  However, the current
+%state of the practice deviates from the existing guidelines because of the
+%rarity of some recommended artifacts (like test plans, requirements
+%specification, code of conduct, code style guidelines, product roadmaps, and
+%Application Program Interface (API) documentation).
 }
@@
 
-We aim to study the state of software development practice for Medical Imaging
+We study the state of software development practice for Medical Imaging
 (MI) software using data available in open source repositories.  MI tools use
@@
 Ultrasound) to provide information for diagnostic, analytic, and medical
-applications. Through MI medical practitioners and
-researchers can noninvasively gain insights into the human body, including
-information on injuries and illnesses. Given the importance of MI software and
-the high number of competing software projects, we wish to understand the merits
-and drawbacks of the current development processes, tools, and methodologies. We
-aim to assess through a software engineering lens the quality of the existing
-software with the hope of highlighting standout examples, and providing
-guidelines and recommendations for future development.
+applications. Given its importance, we want to understand the merits and
+drawbacks of the current development processes, tools, and methodologies. We
+use a software engineering lens to assess the quality of existing MI software
+and provide recommendations for future development.
 
@@
 
-Not only do we wish to gain insight into the state of the practice for MI
-software, we also wish to understand the impact of the often cited gap, or
-chasm, between software engineering and research software \cite{Storer2017}. Although scientists spend a substantial proportion of their working
-hours on software development \cite{Hannay2009, Prabhu2011}, many developers
-learn software engineering skills by themselves or from their peers, instead of
-from proper training \cite{Hannay2009}. In the past many scientists have shown
-ignorance and indifference to standard software engineering concepts
-\cite{Hannay2009}. For instance, according to a survey by Prabhu in 2011
-\cite{Prabhu2011}, more than half of their 114 subjects did not use a proper
-debugger when coding.
-
-To gain insight, we devised 10 research questions, which can be applied to MI,
-as well as to other domains, of research software \cite{SmithEtAl2021,
-SmithAndMichalski2022}.  We designed the questions to learn about the
-community's interest in, and experience with, software artifacts, tools,
-principles, processes, methodologies, and qualities.  When we mention artifacts
-we mean the documents, scripts and code that constitutes a software development
-project. Example artifacts include requirements, specifications, user manuals,
-unit tests, system tests, usability tests, build scripts, API (Application
-Programming Interface) documentation, READMEs, license documents, process
-documents, and code.  Once we have learned what MI developers do, we then put
-this information in context by contrasting MI software against the trends shown
-by developers in other research software communities.
-
-We based the structure of the paper on the research questions, so for each
-research question below we point to the section that contains our answer.  We
-start with identifying the relevant examples of MI software for the assessment
-exercise:
-
+As well as state of the practice (SOP) for MI software, we would like to
+understand the impact of the often cited gap (chasm!) between recommended
+software engineering practices and that used for most research 
+software~\cite{Storer2017}. Although scientists spend
+a substantial proportion of their working times on software development
+\cite{Hannay2009, Prabhu2011}, few are formally trained~\cite{Hannay2009}.
+
+%We have a standard set of questions designed to assess the qualities of any
+%research software project \cite{SmithEtAl2021, SmithAndMichalski2022}. We
+%particularly focus on \emph{software artifactcs}, i.e.
+%the documents, scripts and code that constitutes a software development
+%project. Example artifacts include requirements, specifications, user manuals,
+%unit tests, system tests, usability tests, build scripts, API (Application
+%Programming Interface) documentation, READMEs, license documents, process
+%documents, and code.  %Once we have learned what MI developers do, we then put
+%% this information in context by contrasting MI software against the trends shown
+%% by developers in other research software communities.
+
+Our investigation is based on the following four research questions:
 \begin{enumerate}
 \item[RQ\refstepcounter{rqnum}\therqnum \label{RQ_WhatProjects}:] What MI
-	software projects exist, with the constraint that the source code must be
-	available for all identified projects? (Section~\ref{ch_results})
-	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_HighestQuality}:] Which
-	of the projects identified in \rqref{RQ_WhatProjects} follow current best
-	practices, based on evidence found by experimenting with the software and
-	searching the artifacts available in each project's repository?
+\textbf{open source} software projects exist? (Section~\ref{ch_results})
+\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_HighestQuality}:] Establish
+quantitative measurements of those project's development practices. From
+these measures, which of those projects follow best practices?
 (Section~\ref{ch_results})
 \item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareHQ2Popular}:] How
-	similar is the list of top projects identified in \rqref{RQ_HighestQuality}
-	to the most popular projects, as viewed by the scientific community?
+similar are the top projects identified in \rqref{RQ_HighestQuality}
+to the most popular projects as viewed by the community?
 (Section~\ref{Sec_VsCommunityRanking})
 \item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareArtifacts}:] How
-	do MI projects compare to research software in general with respect to the
+do MI projects compare to general research software with respect to the
 artifacts present in their repositories?
@@
 
-To make the project feasible, we only cover MI visualization software.  As a
-consequence we are excluding many other categories of MI software, including
+We only cover MI visualization software.  We exclude
+other categories of MI software, including
 Segmentation, Registration, Visualization, Enhancement, Quantification,
@@
 
-We designed a general methodology to assess the state of the practice for
-research software \cite{SmithEtAl2021, SmithAndMichalski2022}.  Our methodology
-has been applied to MI software and Lattice Boltzmann Solvers
-\cite{SmithEtAl2024}.  This methodology builds off prior work to assess the
+We have a standard set of questions designed to assess the qualities of any
+research software project~\cite{SmithEtAl2021, SmithAndMichalski2022}.
+This has been applied to MI software and Lattice Boltzmann Solvers
+\cite{SmithEtAl2024}.  This builds off prior work to assess the
 state of the practice for such domains as Geographic Information Systems
@@
 software \cite{Smith2018Seismology}, and Statistical software for psychology
-\cite{smith2018statistical}.  In keeping with the previous methodology, we have
-maintained the constraint that the work load for measuring a given domain should
-be feasible for a team as small as one person, and for a short time, ideally
-around a person month of effort. We consider a person month as $20$ working days
-($4$ weeks in a month, with $5$ days of work per week) at $8$ person hours per
-day, or $20 \times 8 = 160$ person hours.
-
-With our methodology, we first choose a research software domain (in the current
-case MI) and identify a list of about 30 software packages. (For measuring MI we
-used 29 software packages.)  The selected packages are vetted by a domain
-expert. We approximately measure the qualities of each package by filling in a
-grading template. The grader looks at the contents of the open source repository
-and repository based metrics, such as the number of files, number of lines of
-code, percentage of issues that are closed, etc.  With the quantitative data in
-the grading template, we rank the software with the Analytic Hierarchy Process
-(AHP). Further details on interaction with the domain expert, software
-qualities, grading the software and AHP are found below and in
-\cite{SmithEtAl2024_MI_SOP}.
-
-\subsubsection{Interaction With Domain Expert} \label{sec_vet_software_list}
-
-The Domain Experts vet the proposed list of software packages and the AHP
-ranking.  Doing these tasks with non-experts runs the risk of trusting
-inaccurate resources and overlooking the tacit knowledge of an expert. For the
-current assessment, our Domain Expert (and paper co-author) is Dr.\ Michael
-Noseworthy, Professor of Electrical and Computer Engineering at McMaster
-University, Co-Director of the McMaster School of Biomedical Engineering, and
-Director of Medical Imaging Physics and Engineering at St.\ Joseph's Healthcare,
-Hamilton, Ontario, Canada.
+\cite{smith2018statistical}.  We maintain the previous constraint that the
+work load for measuring a given domain should take around one person-month's
+worth of effort ($20$ working days at $8$ person-hours per day).
+
+We identify a list of potential packages (through online searches) which is
+then filtered and vetted by a domain expert. We aim for roughly $30$
+packages. For each remaining package, we measure its qualities by filling in a
+grading template~\cite{SmithEtAl2021}.  This data is used to rank the projects
+with AHP.  Further details on interaction with the domain expert, software
+qualities, grading the software and AHP are summarizes below and in longer form
+in~\cite{SmithEtAl2024_MI_SOP}.
+
+\subsubsection{Domain Expert} \label{sec_vet_software_list}
+
+The Domain Expert vets the proposed list. For too many domains, there are
+online resources that are wildly inaccurate, which only an expert can correct.
+(Details of our domain expert removed for double-blind).
+%The Domain Experts vet the proposed list of software packages and the AHP
+%ranking.  Doing these tasks with non-experts runs the risk of trusting
+%inaccurate resources and overlooking the tacit knowledge of an expert. For the
+%current assessment, our Domain Expert (and paper co-author) is Dr.\ Michael
+%Noseworthy, Professor of Electrical and Computer Engineering at McMaster
+%University, Co-Director of the McMaster School of Biomedical Engineering, and
+%Director of Medical Imaging Physics and Engineering at St.\ Joseph's Healthcare,
+%Hamilton, Ontario, Canada.
 
 In advance of the first meeting with the Domain Expert, they are asked to
-create a list of top software packages in the domain.  This is done to help
-the expert get in the right mind set in advance of the meeting.  Moreover,
-by doing the exercise in advance, we avoid the potential pitfall of the expert
-approving the discovered list of software without giving it adequate thought.
+independently create a list of top software packages in the domain.  This helps
+get the expert's knowledge refreshed in advance of the meeting.
 
@@
 Quality is defined as a measure of the excellence or worth of an entity.  As is
-common practice, we do not think of quality as a single measure, but rather as a
-set of measures.  That is, quality is a collection of different qualities, often
-called ``ilities.''  For this study we selected 9 qualities to measure:
+common practice, we do not think of quality as a single measure, but rather as
+a set of measures.  That is, quality is a collection of different qualities,
+often called ``ilities.''  For this study we selected 9 qualities to measure:
 installability, correctness/ verifiability, reliability, robustness, usability,
 maintainability, reusability, understandability, and visibility/transparency.
-With the exception of installability, all the qualities are defined in Ghezzi et
-al. (2003) \cite{GhezziEtAl2003}. Installability is defined as the effort
+With the exception of installability, all the qualities are defined in Ghezzi
+et al. (2003) \cite{GhezziEtAl2003}. Installability is defined as the effort
 required for the installation and/or uninstallation of software in a specified
@@
 
-\subsubsection{Grading Software} \label{sec_grading_software}
-
-We grade the selected software using an existing measurement template
-\cite{SmithEtAl2021}.  The template provides measures of the qualities listed in
-Section~\ref{sec_software_quality}.  For each software package, we fill in the
-template questions. To stay within the target of 160 person hours to measure the
-domain, we allocated between one and four hours for each package. Project
-developers can be contacted for help regarding installation, if necessary, but
-we impose a cap of about two hours on the installation process, to keep the
-overall measurement time feasible.  Figure~\ref{fg_grading_template_example}
-shows an excerpt of the spreadsheet.  The spreadsheet includes a column for each
-measured software package. \cite{Dong2021-Data} provides the full set of
-measurement data.
+\subsubsection{Grading} \label{sec_grading_software}
+
+We use an existing template~\cite{SmithEtAl2021} that is designed to measure
+the aforementioned qualities. To stay within our given measurement time frame,
+each package gets up to five hours of time.  Project developers can be
+contacted for help regarding installation, if necessary, but we impose a cap of
+about two hours on the installation process.
+Figure~\ref{fg_grading_template_example} shows an excerpt of the measurement
+spreadsheet.  It includes a column for each measured software package.
+% \cite{Dong2021-Data} provides the full set of measurement data.
+(The full data is available on Mendeley; link will be provided after
+refereeing.)
 
@@
 
-The full template consists of 108 questions categorized under 9 qualities.  We
-designed the questions to be unambiguous, quantifiable, and measurable with
-limited time and domain knowledge. We group the measures under headings for each
-quality, and one for summary information. The summary information (shown in
-Figure~\ref{fg_grading_template_example}) is the first section of the template.
-This section summarizes general information, such as the software name, purpose,
-platform, programming language, publications about the software, the first
-release and the most recent change date, website, source code repository of the
-product, number of developers, etc.  We only measure open source software that
-is ready for public use by other researchers (following the definition of the
-public category from \cite{GewaltigAndCannon2012}). Information in the summary
-section sets the context for the project, but it does not directly affect the
-grading scores.
-
-For measuring each quality, we ask several questions and the typical answers are
-among the collection of ``yes'', ``no'', ``n/a'', ``unclear'', a number, a
-string, a date, a set of strings, etc. The grader assigns each quality an
-overall score, between 1 and 10, based on all the previous questions.  Several
-of the qualities use the word ``surface''.  This is to highlight that, for these
-qualities in particular, the best that we can do is a shallow measure.  For
-instance, we are not currently doing any experiments to measure usability.
-Instead, we are looking for an indication that the developers considered
-usability.  We do this by looking for cues in the documentation, like a getting
-started manual, a user manual and a statement of expected user characteristics.
+The full template consists of 108 questions over 9 qualities.  These
+questions are designed to be unambiguous, quantifiable, and measurable
+with constrained time and domain knowledge.
+%We
+%designed the questions to be unambiguous, quantifiable, and measurable with
+%limited time and domain knowledge. We group the measures under headings for each
+%quality, and one for summary information. The summary information (shown in
+%Figure~\ref{fg_grading_template_example}) is the first section of the template.
+%This section summarizes general information, such as the software name, purpose,
+%platform, programming language, publications about the software, the first
+%release and the most recent change date, website, source code repository of the
+%product, number of developers, etc.  We only measure open source software that
+%is ready for public use by other researchers (following the definition of the
+%public category from \cite{GewaltigAndCannon2012}). Information in the summary
+%section sets the context for the project, but it does not directly affect the
+%grading scores.
+
+%For measuring each quality, we ask several questions and the typical answers are
+%among the collection of ``yes'', ``no'', ``n/a'', ``unclear'', a number, a
+%string, a date, a set of strings, etc. 
+The grader, after answering questions for each quality assigned an overall
+score (between 1 and 10) based on the answers.  Several of the qualities use
+the word ``surface'' to highlight that these particular qualities are a shallow
+measure.  For example, we do not do user studies for measuring usability.
+Instead, we are look signs that the developers considered usability.
 We use two freeware tools to collect repository related data:
 \href{https://github.com/tomgi/git_stats}{GitStats} and
-\href{https://github.com/boyter/scc}{Sloc Cloc and Code (scc)}.  Details on how
-we measure each quality are provided in Smith et al.\ (2024)
+\href{https://github.com/boyter/scc}{Sloc Cloc and Code (scc)}.  Further
+details on quality measurement are provided in~% Smith et al.\ (2024)
 \cite{SmithEtAl2024_MI_SOP}.
@@
 
-Saaty developed AHP in the 1970s, and people have widely used it since to make
-and analyze multiple criteria decisions \cite{VaidyaEtAl2006}. AHP organizes
+Developed by Saaty in the 1970s, AHP is widely used to 
+analyze multiple criteria decisions~\cite{VaidyaEtAl2006}. AHP organizes
 multiple criteria in a hierarchical structure and uses pairwise comparisons
-between alternatives to calculate relative ratios \cite{Saaty1990}. AHP works
+between alternatives to calculate relative ratios~\cite{Saaty1990}. AHP works
 with sets of $n$ \textit{options} and $m$ \textit{criteria}.  In our project
@@
 
-In this section we review the selected software packages.  We used a two-step
-process for selecting the software packages: i) identify software candidates in
-the chosen domain; and, ii) filter the list to remove less relevant members
-\cite{SmithEtAl2021}.
-
-We initially identified 48 MI candidate software projects \cite{Dong2021} from
+We initially identified 48 candidate software projects from
 the literature \cite{Bjorn2017, Bruhschwein2019, Haak2015}, on-line articles
-\cite{Emms2019, Hasan2020, Mu2019}, and forum discussions \cite{Samala2014}.  To
-reduce the length of the list to a manageable number (29 in this case), we
-filtered the original list as follows:
+\cite{Emms2019, Hasan2020, Mu2019}, and forum discussions \cite{Samala2014}.
+Then we filtered as follows:
 
@@
 
-\item We removed the packages with no source code available, such as
+\item Removed the packages with no source code available, such as
 \textit{MicroDicom}, \textit{Aliza}, and \textit{jivex}.
 
-\item We focused on the MI software that provides visualization functions, as
-described in Section~\ref{sec_scope}. Furthermore, we removed seven packages
+\item Focused on MI software that provides visualization functions.
+We removed seven packages
 that were toolkits or libraries, such as \textit{VTK}, \textit{ITK}, and
-\textit{dcm4che}.  We removed another three that were for PACS.
-
-\item We removed \textit{Open Dicom Viewer}, since it has not received any
-updates in a long time (since 2011).
+\textit{dcm4che}, and another three that were for PACS.
+
+\item Removed \textit{Open Dicom Viewer} as it has not received any
+updates since 2011.
 
@@
 
-The Domain Expert provided a list of his top 12 software packages.  We compared
-his list to our list of 29.  We found 6 packages were on both lists: \textit{3D
-Slicer}, \textit{Horos}, \textit{ImageJ}, \textit{Fiji}, \textit{MRIcron} (we
-actually use the update version \textit{MRIcroGL}) and \textit{Mango} (we
-actually use the web version \textit{Papaya}).  Six software packages
-(\textit{AFNI}, \textit{FSL}, \textit{Freesurfer}, \textit{Tarquin},
-\textit{Diffusion Toolkit}, and \textit{MRItrix}) were on the Domain Expert
-list, but not on our filtered list.  When we examined those packages,
-we found they were out of scope, since their primary function was not
-visualization.  The Domain Expert agreed with our final choice of 29 packages.
-
-Table~\ref{tab_final_list} shows the 29 software packages that we measured,
-along with summary data collected in the year 2020. We arrange the items in
-descending order of LOC. We found the initial release dates (Rlsd) for most
+The Domain Expert provided a list 12 software packages.
+We found 6 packages were on both lists: \textit{3D
+Slicer}, \textit{Horos}, \textit{ImageJ}, \textit{Fiji}, \textit{MRIcron} 
+(we use its descendant \textit{MRIcroGL}) and \textit{Mango} (we
+use the web version \textit{Papaya}).  The remaining six packages were on our
+out-of-scope list. The Domain Expert agreed with our final choice of 29
+packages -- see Table~\ref{tab_final_list}. This contains summary data
+collected in the year 2020.
+
+The projects are sorted in descending order of lines of code.
+We found the initial release dates (Rlsd) for most
 projects and marked the two unknown dates with ``?''. The date of the last
@@
 JavaScript, Java, C, Python, Pascal, Matlab.  The most popular language is \CC,
-with almost 40\% of projects (11 of 29).  The two least popular choices are
-Pascal and Matlab, with around 3\% of projects each (1 of 29).
+for 11 of 29 projects; Pascal and Matlab were each used for a single project.
 
@@
 Figure \ref{fg_installability_scores} lists the installability scores.  We found
-installation instructions for 16 projects. Among the ones without instructions,
-\textit{BioImage Suite Web} and \textit{Slice:Drop} do not need installation,
-since they are web applications. Installing 10 of the projects required extra
-dependencies. Five of them are web applications (as shown in
-Table~\ref{tab_final_list}) and depend on a browser; \textit{dwv}, \textit{OHIF
-Viewer}, and \textit{GATE} needs extra dependencies to build; \textit{ImageJ}
-and	\textit{Fiji} need an unzip tool; \textit{MatrixUser} is based on Matlab;
-\textit{DICOM Viewer} needs to work on a Nextcloud platform.
+installation instructions for 16 projects, but two did not need them
+(\textit{BioImage Suite Web} and \textit{Slice:Drop})
+as they are web applications. 10 of the projects required extra
+dependencies: Five depend on a specific browser; \textit{dwv}, \textit{OHIF
+Viewer}, and \textit{GATE} needs extra libraries to build; \textit{ImageJ} and
+\textit{Fiji} need an unzip tool; \textit{MatrixUser} needs Matlab;
+\textit{DICOM Viewer} needs a Nextcloud platform.
 
@@
 
-\textit{3D Slicer} has the highest score because it had easy to follow
-installation instructions, and an automated, fast, frustration-free installation
-process. The installer added all dependencies automatically and no errors
-occurred during the installation and uninstallation steps. Many other software
-packages also had installation instructions and automated installers.  We had no
-trouble installing the following packages: \textit{INVESALIUS 3},
-\textit{Gwyddion}, \textit{XMedCon}, and \textit{MicroView}. We determined their
-scores based on the understandability of the instructions, installation steps,
-and user experience. Since \textit{BioImage Suite Web} and \textit{Slice:Drop}
-needed no installation, we gave them high scores. \textit{BioImage Suite Web}
-also provided an option to download cache for offline usage, which was easy to
-apply.
+The scores are based on the ease of following the installation instructions,
+and automated installation and uninstallation process. There were no issues
+for all but the bottom four, just various degrees of ease and automation.
 
@@
 \textit{GATE}. For \textit{DICOM Viewer} we could not install the NextCloud
-dependency, and we did not have another option for running the software.
-Therefore, for \textit{DICOM Viewer} we could not measure reliability or
-robustness.  The other seven qualities could be measured, since they do not
-require installation.
-
-\textit{MatrixUser} has a lower score because it depends on Matlab. We assessed
-the score from the point of view of a user that would have to install Matlab and
-acquire a license.  Of course, for users that already work within Matlab, the
-installability score should be higher.
+dependency, and thus could not measure reliability nor robustness for it.
+
+\textit{MatrixUser} depends on Matlab, whose installation is not easy.
+For users who already have Matlab, this score should be higher.
 
@@
 
-Figure~\ref{fg_correctness_verifiability_scores} shows the scores of correctness
-and verifiability. Generally speaking, the packages with higher scores adopted
-more techniques to improve correctness, and had better documentation for us to
-verify against.  For instance, we looked for evidence of unit testing, since it
-benefits most parts of the software's life cycle, such as designing, coding,
-debugging, and optimization \cite{Hamill2004}.  We only found evidence of unit
-testing for about half of the projects. We identified five projects using CI/CD
+The packages with higher scores for correctness and verifiability (see
+Figure~\ref{fg_correctness_verifiability_scores}) used a wider array of
+techniques to improve correctness, and had better documentation to witness
+this. For instance, we looked for evidence of unit testing, and only evidence
+for about half of the projects. We identified five projects using CI/CD
 tools: \textit{3D Slicer}, \textit{ImageJ}, \textit{Fiji}, \textit{dwv}, and
@@
 Even for some projects with well-organized documentation, requirements
-specifications and theory manuals were still missing.  We could not identify
-theory manuals for all projects, and we did not find requirements specifications
-for most projects. The only requirements-related document we found was a road
+specifications and theory manuals were still missing.
+The only requirements-related document we found was a road
 map of \textit{3D Slicer}, which contained design requirements for upcoming
@@
 
-Figure~\ref{fg_reliability_scores} shows the AHP results.  As shown in
-Section~\ref{sec_result_installability}, most of the software products did not
-``break'' during installation, or did not need installation; \textit{dwv} and
-\textit{GATE} broke in the building stage, and the processes were not
-recoverable; we could not install the dependency for \textit{DICOM Viewer}. Of
-the seven software packages with a getting started tutorial and operation steps
-in the tutorial, most showed no error when we followed the steps. However,
+Figure~\ref{fg_reliability_scores} shows the our results.  We were able to
+follow the steps in the tutorials that existed (seven packages had them.)
+However
 \textit{GATE} could not open macro files and became unresponsive several times,
@@
 when loading damaged image files, without showing any descriptive error message.
-We did not find any problems with the on-line version of \textit{dwv}.
 
@@
 The packages with higher scores elegantly handled unexpected/unanticipated
-inputs, typically showing a clear error message. We may have underestimated the
-score of \textit{OHIF Viewer}, since we needed further customization to load
+inputs, typically showing a clear error message. We may have underscored
+\textit{OHIF Viewer}, since we needed further customization to load
 data.

### 2024-06-14 StateOfPractice/Papers/SOP_MI_OSR/SOP_MI_OSR.tex
more shorten

@@
 Figure \ref{fg_robustness_scores} presents the scores for surface robustness.
-The packages with higher scores elegantly handled unexpected/unanticipated
+The packages with higher scores elegantly handled unexpected or unanticipated
 inputs, typically showing a clear error message. We may have underscored
@@
 DICOM standard. To test robustness, we prepared two types of image files:
-correct and incorrect formats (with the incorrect format created by relabelled a
+correct and incorrect formats (with the incorrect format created by relabelling a
 text file to have the ``.dcm'' extension).  All software packages loaded the
@@
 reasons.  For the broken format, \textit{MatrixUser}, \textit{dwv}, and
-\textit{Slice:Drop} ignored the incorrect format of the file and loaded it
-regardless. They did not show any error message and displayed a blank image.
+\textit{Slice:Drop} ignored the incorrect format,
+did not show any error message and displayed a blank image.
 \textit{MRIcroGL} behaved similarly except that it showed a meaningless image.
@@
 
-\begin{figure}[!ht]
+\begin{figure}[ht]
 \includegraphics[scale=0.38]{usability_scores.pdf}
@@
 product, developer's manual, or API documentation, and only \textit{3D Slicer},
-\textit{ImageJ}, \textit{Fiji} included all three documents. Moreover,
+\textit{ImageJ}, \textit{Fiji} included all three documents 
+(see Table~\ref{tab_maintainability_docs} the full data).  Moreover,
 \textit{3D Slicer} has a much higher percentage of closed issues (92\%) compared
 to \textit{ImageJ} (52\%) and \textit{Fiji} (64\%).
-Table~\ref{tab_maintainability_docs} shows which projects had these documents,
-in descending order of their maintainability scores.
-
-\begin{figure}[!ht]
+
+\begin{figure}[ht]
 \includegraphics[scale=0.38]{maintainability_scores.pdf}
@@
 
-\begin{table}[!ht]
+\begin{table}[ht]
 \centering
@@
 
-Twenty-seven of the 29 projects used git as the version control tool, with 24 of
+Twenty-seven of the 29 projects used git for version control, with 24 of
 these using GitHub. \textit{AMIDE} used Mercurial and \textit{Gwyddion} used
@@
 
-Figure~\ref{fg_reusability_scores} shows the AHP results for reusability. As
-described in Section~\ref{sec_grading_software}, we gave higher scores to the
-projects with API documentation. As shown in
-Table~\ref{tab_maintainability_docs}, seven projects had API documents. We also
-assumed that projects with more code files and less LOC per code file are more
-reusable. Table \ref{tab_loc_per_file} shows the number of text-based files by
-project, which we used to approximate the number of code files. The table also
-lists the total number of lines (including comments and blanks), LOC, and
-average LOC per file. We arranged the items in descending order of their
-reusability scores.
-
-\begin{figure}[!ht]
+API documents (see Table~\ref{tab_maintainability_docs}) also help maintainability,
+which explains some of the scores shown in
+Figure~\ref{fg_reusability_scores}. We have assumed that smaller code files are
+likely more reusable -- see
+Table \ref{tab_loc_per_file} for the details.
+
+\begin{figure}[ht]
 \includegraphics[scale=0.38]{reusability_scores.pdf}
@@
 \end{tabular}
-\caption{Number of files and lines (sorted in descending order of reusability
-scores)}
+\caption{Number of files and lines (by reusability scores)}
 \label{tab_loc_per_file}
@@
 Figure~\ref{fg_surface_understandability_scores} shows the scores for surface
-understandability. All projects had a consistent coding style with parameters in
+understandability. 
+All projects had a consistent coding style with parameters in
 the same order for all functions, modularized code, and, clear comments that
@@
 
-\begin{figure}[!ht]
+\begin{figure}[ht]
 \includegraphics[scale=0.38]{understandability_scores.pdf}
@@
 documents for the development process, project status, development environment,
-and release notes, in descending order of their visibility/transparency
-scores.
-
-\begin{figure}[!ht]
+and release notes.
+
+\begin{figure}[ht]
 \includegraphics[scale=0.38]{visibility_transparency_scores.pdf}
@@
 
-As described in Section~\ref{sec_AHP}, for our AHP measurements, we have nine
-criteria (qualities) and 29 alternatives (software packages). In the absence of
+In the absence of
 a specific real world context, we assumed all nine qualities are equally
 important. Figure~\ref{fg_overall_scores} shows the overall scores in descending
-order. Since we produced the scores from the AHP process, the total sum of the
-29 scores is precisely 1.0.
+order.
 

### 2024-06-14 StateOfPractice/Papers/SOP_MI_OSR/SOP_MI_OSR.tex
down 5 pages. main body just over 18.

@@
 
-In Section~\ref{Sec_VsCommunityRanking}, we compare our ranking to the MI
-software community's ranking, as a way to judge our results, and as a way to
-understand what the MI community currently values. To understand the state of
-the practice for MI software relative to other software, we compare MI software
-to other research software in Section~\ref{Sec_CompareArtifacts}.  The
-comparison is based on the artifacts our measurement exercise found in the MI
-repositories, versus what is typically recommended for research software.  We
-provide details on artifacts that are recommended but rarely observed for MI
-software. Section~\ref{sec_threats_to_validity} presents threats to the validity
-of our data and conclusions.
+We first compare our ranking to a (proxy for) the community's ranking. We then
+compare the state of the practice for MI with that of other research software.
+In particular we provide details on recommended artifacts but rarely observed
+for MI software. Section~\ref{sec_threats_to_validity} presents threats to the
+validity of our data and conclusions.
 
@@
 
-To answer~\rqref{RQ_CompareHQ2Popular} about how our ranking compares to the
-popularity of projects as judged by the scientific community, we compare our
-ranking (from Section~\ref{ch_results}) with the community ratings on GitHub, as
-shown by GitHub stars, number of forks, and number of people watching the
-projects.
-
-Table~\ref{tab_ranking_vs_GitHub} shows our ranking of the 29 MI projects, and
-their GitHub metrics, if applicable. As mentioned in
-Section~\ref{sec_score_maintainability}, 24 projects used GitHub. Since GitHub
-repositories have different creation dates, we collect the number of months each
-stayed on GitHub, and calculate the average number of new stars, people
-watching, and forks per 12 months. The items in
-Table~\ref{tab_ranking_vs_GitHub} are listed in descending order of the average
-number of new stars per year.  The non-GitHub items are listed in the order of
-our ranking.  We collected all GitHub statistics in July 2021.
-
-Generally speaking, most of the top-ranking MI software projects also received
-greater attention and popularity on GitHub. Between our ranking and the GitHub
-stars-per-year ranking, four of the top five software projects appear in both
-lists. Our top five packages are scattered among the first eight positions on the
-GitHub list. However, as discussed below, there are discrepancies between the two
-lists.
-
-In some cases projects are popular in the community, but were assigned a low
-rank by our methodology.  This is the case for \textit{dwv}. The reason for the
-low ranking is that, as mentioned in Section~\ref{sec_result_installability}, we
-failed to build it locally, and used the test version on its websites for the
-measurements. We followed the instructions and tried to run the command
-\texttt{yarn run test} locally, which did not work. In addition, the test
-version did not detect a broken DICOM file and displayed a blank image as
-described in Section~\ref{sec_result_robustness}. We might underestimate the
-scores for \textit{dwv} due to uncommon technical issues.
-
-We also ranked \textit{DICOM Viewer} much lower than its popularity. As
-mentioned in Section~\ref{sec_result_installability}, it depended on the
-NextCloud platform that we could not successfully install. Thus, we might
-underestimate the scores of its surface reliability and surface robustness.
-
-Further reason for discrepancies between our ranking and the community's ranking
-is that we weighted all qualities equally. This is not likely how users
-implicitly rank the different qualities. As a result, some projects with high
-community popularity may have scored lower with our method because of a
-relatively higher (compared to the scientific community's implicit ranking)
-weighting of the poor scores for some qualities. A further explanation for
-discrepancies between our measures and the star measures may also be due to
-inaccuracy with using stars to approximate popularity.  Stars are not an ideal
-measure because stars represent the community's feeling in the past more than
-they measure current preferences \cite{Szulik2017}.  The issue with stars is
-that they tend only to be added, not removed.  A final reason for
-inconsistencies between our ranking and the community's ranking is that, as for
-consumer products, more factors influence popularity than just quality.
+We use GitHub stars, number of forks and number of people watching the projects
+are proxies for community ranking -- see Table~\ref{tab_ranking_vs_GitHub}
+for statistics collected in July 2021.  Recall that 24 projects use GitHub.
+Our ranking and GitHub popularity, at least for the top five projects,
+seems to line up fairly well.
+
+We ranked some popular packages fairly low, such as \textit{dwv}. This is
+because we were unable to build it locally, even though we followed its
+installation instructions. But we were able to use its web
+version for the rest of the measurements. Additionally, this version 
+did not detect a broken DICOM file and instead displayed a blank image
+(Section~\ref{sec_result_robustness}).
+\textit{DICOM Viewer} ranked low as we were unable to install the
+NextCloud platform.
+
+Another likely reason for discrepancies is that we weighted all qualities
+equally. This is not likely how users implicitly rank the different qualities.
+This would require a broader user study to properly assess. Furthermore our
+measures of popularity are only \emph{proxies} which are biased towards
+past rather than current preferences~\cite{Szulik2017}, as these are
+monotonically increasing quantities. Finally there are often more factors than
+just quality that influence the popularity of ``consumer'' products.
 
@@
 
-Although our ranking and the estimate of the community's ranking are not perfect
-measures, they do suggest a correlation between best practices and popularity.
-We do not know which comes first, the use of best practices or popularity, but
-we do know that the top ranked packages tend to incorporate best practices. The
-next sections will explore how the practices of the MI community compare to the
-broader research software community.
-
-\subsection{Comparison Between MI and Research Software for Artifacts}
+Although both rankings are imperfect measures, they nevertheless suggest a
+correlation between best practices and popularity. We don't know if this is
+causal, in either direction (i.e. if best practices enable popularity or if
+popularity increases the need for using more software development best
+practices).
+
+\subsection{Software Artifacts}
 \label{Sec_CompareArtifacts}
 
-As part of filling in the measurement template (from
-Section~\ref{sec_grading_software}), we summarized the artifacts observed in
-each MI package. Table~\ref{artifactspresent} groups the artifacts by frequency
-into categories of common (20 to 29 ($>$67\%) packages), uncommon (10 to 19
-(33-67\%) packages), and rare (1 to 9 ($<$33\%) packages).
-Tables~\ref{tab_maintainability_docs} and~\ref{tab_Visibility/Transparency_docs}
-show the details on which projects use which types of artifacts for documents
-related to maintainability and visibility, respectively.
-
-\begin{table}[ht!]
-    %\begin{center}
-    \begin{tabular}{ p{3.1 cm} p{5.4 cm} p{3.5 cm}}
-    \toprule
-    Common & Uncommon & Rare \\
-    \midrule
-    README (29) & Build scripts (18) & Getting Started (9)\\
-    Version control (29) & Tutorials (18) & Developer's manual (8)\\
-    License (28) & Installation guide (16) & Contributing (8)\\
-    Issue tracker (28) & Test cases (15) & API documentation (7)\\
-    User manual (22) & Authors (14) & Dependency list (7)\\
-    Release info. (22) & Frequently Asked Questions (FAQ) (14) & Troubleshooting guide (6)\\
-     & Acknowledgements (12) & Product roadmap (5)\\
-     & Changelog (12) & Design documentation (5)\\
-     & Citation (11) & Code style guide (3)\\
-     & & Code of conduct (1)\\
-     & & Requirements (1)\\
-    \bottomrule
-    \end{tabular}
-    \caption{Artifacts Present in MI Packages, Classified by Frequency (The number 
-    in brackets is the number of occurrences)}
-    \label{artifactspresent}
-    %\end{center}
-\end{table}
-
-We answer~\rqref{RQ_CompareArtifacts} by comparing the artifacts that we
-observed in MI repositories to those observed and recommended for research
-software in general. Our comparison may point out areas where some MI software
-packages fall short of current best practices. This is not intended to be a
-criticism of any existing packages, especially since for practical reasons not
-every project has the resources to achieve the highest possible quality.
-However, rather than delve into the nuances of which software can justify
-compromising which practices, we will write our comparison under the ideal
-assumption that every project has sufficient resources to match best practices.
-
-Table~\ref{Tbl_Guidelines} (based on data from \cite{SmithAndMichalski2022})
-shows that MI artifacts generally match the recommendations found in nine
-current research software development guidelines:
+We use nine research software development guidelines to measure what
+software artifacts are present in MI software, versus those recommended.
+These guidelines are:
 \begin{itemize}
@@
 
-In Table~\ref{Tbl_Guidelines} each row corresponds to an artifact.  For a given
-row, a checkmark in one of the columns means that the corresponding guideline
-recommends this artifact.  The last column shows whether the artifact appears in
-the measured set of MI software, either not at all (blank), commonly (C),
-uncommonly (U) or rarely (R).  We did our best to interpret the meaning of each
-artifact consistently between guidelines and specific MI software, but the
-terminology and the contents of artifacts are not standardized.  The challenge
-even exists for the ubiquitous README file.  The content of README files shows
-significant variation between projects \cite{PranaEtAl2018}.  Although some
-content is reasonably consistent, with 97\% of README files contain at least one
-section describing the `What' of the repository and 89\% offering some `How'
-content, other categories are more variable.  For instance, information on
-`Contribution', `Why', and `Who', appear in 28\%, 26\% and 53\% of the analyzed
-files, respectively \cite{PranaEtAl2018}.
-
-The frequency of checkmarks in Table~\ref{Tbl_Guidelines} indicates the
-popularity of recommending a given artifact, but it does not imply that the most
-commonly recommended artifacts are the most important artifacts. Just because a
-guideline does not explicitly recommend an artifact, does not mean the guideline
-authors do not value it.  They may have excluded it because it is out of the
-scope of their recommendations, or outside their experience.  For instance, an
-artifact related to uninstall is only explicitly mentioned by one set of
-guidelines \cite{vanGompelEtAl2016}, but other guideline authors would likely
-see its value.  They may simply feel that uninstall is implied by install, or
-they may have never asked themselves whether they need separate uninstall
-instructions.
-
-\begin{table}[h]
+\begin{table}[ht]
 %\begin{center}
@@
 
-Two of the items that appear in Table~\ref{artifactspresent} do not appear in
-the software development guidelines shown in Table~\ref{Tbl_Guidelines}:
-Troubleshooting guide and Developer's manual.  Although the guidelines do not
-name these two artifacts, the information contained within them overlaps with
-the recommended artifacts.  A Troubleshooting guideline contains information
-that would typically be found in a User manual.  A Developer's guide overlaps
-with information from the README, INSTALL, Uninstall, Dependency List, Release
-Information, API documentation and Design documentation.  In our current
-analysis, we have identified artifacts by the names given by the software
-guidelines and MI examples.  In the future, a more in-depth analysis would look
-at the knowledge fragments that are captured in the artifacts, rather than
-focusing on the names of the files that collect these fragments together.
-
-Although the MI community shows examples of 88\% (23 of 26) of the practices we
-found in research software guidelines (Table~\ref{Tbl_Guidelines}), we did not
-observe three recommended artifacts: i) Uninstall, ii) Test plans, and iii)
-Requirements.  Uninstall is likely an omission caused by the focus on installing
-software. Given the storage capacity of current hardware, developers and users
-are not generally concerned with uninstall.  Moreover, as mentioned above,
-uninstall is not particularly emphasized in existing recommendations.
-
-Test plans describe the scope, approach, resources, and the schedule of planned
-test activities \cite{VanVliet2000}.  The plan should cover details such as the
-makeup of the testing team, automated testing tools and technology to employ,
-the testing process, system tests, integration tests and unit tests. We did not
-observe test plans for MI software, but that doesn't mean plans weren't created;
-it means that the plans are not under version control. Test plans would have to
-at least be implicitly created, since we observed test cases with reasonable
-frequency for MI software (test cases are categorized as uncommon).
-
-The other apparently neglected document is the requirements specification, which
-records the functionalities, expected performance, goals, context, design
-constraints, external interfaces and other quality attributes of the software
-\cite{IEEE1998}.  If developers write requirements they typically are based on a
-template, which provide documentation structure, guidelines, and rules.
-Although there is no universally accepted template, examples exist
-\cite{IEEE1998, NASA1989, RobertsonAndRobertson1999Vol}.
-
-MI software is like other research software in its neglect of requirements
-documentation.  Although requirements documentation is recommended by some
+In Table~\ref{Tbl_Guidelines} each row corresponds to an artifact.  For a given
+row, a checkmark in one of the columns means that the corresponding guideline
+recommends this artifact.  The last column shows whether the artifact appears in
+the measured set of MI software, either not at all (blank), commonly (C),
+uncommonly (U) or rarely (R).  We did our best to interpret the meaning of each
+artifact consistently between guidelines and specific MI software, but the
+terminology and the contents of artifacts are not standardized.  The challenge
+even exists for the ubiquitous README file.  The content of README files shows
+significant variation between projects \cite{PranaEtAl2018}.  Although some
+content is reasonably consistent, with 97\% of README files contain at least one
+section describing the `What' of the repository and 89\% offering some `How'
+content, other categories are more variable.  For instance, information on
+`Contribution', `Why', and `Who', appear in 28\%, 26\% and 53\% of the analyzed
+files, respectively \cite{PranaEtAl2018}.
+
+\begin{table}[ht!]
+    %\begin{center}
+    \begin{tabular}{ p{3.1 cm} p{5.4 cm} p{3.5 cm}}
+    \toprule
+    Common & Uncommon & Rare \\
+    \midrule
+    README (29) & Build scripts (18) & Getting Started (9)\\
+    Version control (29) & Tutorials (18) & Developer's manual (8)\\
+    License (28) & Installation guide (16) & Contributing (8)\\
+    Issue tracker (28) & Test cases (15) & API documentation (7)\\
+    User manual (22) & Authors (14) & Dependency list (7)\\
+    Release info. (22) & Frequently Asked Questions (FAQ) (14) & Troubleshooting guide (6)\\
+     & Acknowledgements (12) & Product roadmap (5)\\
+     & Changelog (12) & Design documentation (5)\\
+     & Citation (11) & Code style guide (3)\\
+     & & Code of conduct (1)\\
+     & & Requirements (1)\\
+    \bottomrule
+    \end{tabular}
+    \caption{Artifacts Present in MI Packages, Classified by Frequency (The number 
+    in brackets is the number of occurrences)}
+    \label{artifactspresent}
+    %\end{center}
+\end{table}
+
+Table~\ref{artifactspresent} presents our measurements for MI software.
+It groups the artifacts by frequency
+into categories of common (20 to 29 ($>$67\%) packages), uncommon (10 to 19
+(33-67\%) packages), and rare (1 to 9 ($<$33\%) packages).
+Tables~\ref{tab_maintainability_docs} and~\ref{tab_Visibility/Transparency_docs}
+show the details on which projects use which types of artifacts for documents
+related to maintainability and visibility, respectively.
+
+Note that ``popularity'' in Table~\ref{Tbl_Guidelines} does no imply that these
+oft recommended artifacts are the most important. Guidelines are often brief,
+to encourage adoption, and thus even guidelines that mention the need for
+installation instructions rarely mention uninstallation instructions.
+Two items in Table~\ref{artifactspresent} do not appear in any guidelines:
+\emph{Troubleshooting guide} and \emph{Developer's manual}.  However the
+information within these documents overlaps with
+the recommended artifacts.  Troubleshooting information often can be found in
+a User Manual, while the information in a ``Developer's Manual'' is often
+scattered amongst many other documents.
+Three of the 26 recommended artifacts were never observed:
+i) Uninstall, ii) Test plans, and iii) Requirements. It is possible that
+some of these were created but never put under version control.
+
+It is unfortunately common for research software to neglect requirements
+documentation, and MI software is no exception.
+Although such documentation is recommended by some
 \cite{TobiasEtAl2018, HerouxEtAl2008, SmithAndKoothoor2016}, in practice
-research software developers often do not produce a proper requirements
-specification \cite{HeatonAndCarver2015}. Sanders and Kelly
+this is rare~\cite{HeatonAndCarver2015}. Sanders and Kelly
 \cite{SandersAndKelly2008} interviewed 16 scientists from 10 disciplines and
@@
 commonly produced type of documentation for research software in general
-\cite{Nguyen-HoanEtAl2010}. When looking at the pain points for research
-software developers, software requirements and management is the software
-engineering discipline that most hurts scientific developers, accounting for
-23\% of the technical problems \cite{WieseEtAl2019}.  The lack of support for
-requirements is likely due to the perception that up-front requirements are
-impossible for research software \cite{CarverEtAl2007, SegalAndMorris2008}, but
-if we drop the insistence on ``up-front'' requirements, allowing instead for the
-requirements to be written iteratively and incrementally, requirements are
-feasible \cite{Smith2016}.  Smith et al. provide a requirements template
-tailored to research software \cite{SmithEtAl2007}.
-
-Table~\ref{Tbl_Guidelines} shows several recommended artifacts that are rarely
-observed in practice.  A theme among these rare artifacts is that, except for
-the user-focused getting started manual, they are developer-focused.  The
-dependency list, which is a list of software library dependencies, was rarely
-observed, but this information is still likely present, just embedded in build
-scripts. The other developer-focused and rare artifacts are as follows:
+\cite{Nguyen-HoanEtAl2010}.
+
+This is quite unfortunate as when surveyed, research software developers
+identify the lack of requirements and requirements management as accounting
+gets mentioned the most often and accounts for 23\% of the technical problems
+\cite{WieseEtAl2019}.  There is unfortunately widespread perception that
+up-front requirements are impossible for research software
+\cite{CarverEtAl2007, SegalAndMorris2008}. Fortunately a more agile approach
+to requirements is feasible~\cite{Smith2016}, and research-software specific
+templates exist~\cite{SmithEtAl2007}.
+
+A theme emerges amongst the artifacts rarely observed in practice: they
+are developer-focused (a list of library dependencies, a contributor's guide,
+a developer Code of Conduct, coding style guidelines, product roadmap, design
+documentation and API documentation).
+
+Other communities use checklists to help with best practices. Examples
+include checklists for
+merging branches~\cite{Brown2015},
+for saving and sharing changes to the project~\cite{WilsonEtAl2016},
+for new and departing team members~\cite{HerouxAndBernholdt2018},
+for processes related to commits and releases~\cite{HerouxEtAl2008}
+and for overall software quality~\cite{ThielEtAl2020, SSI2022}.
+
+MI projects fall somewhat short of recommended best practices, but are not
+alone amongst research software projects. This gap has been
+documented before~\cite{Storer2017,JohansonAndHasselbring2018}, is known
+to cause sustainability and reliability
+problems \cite{FaulkEtAl2009}, and to waste development effort%
+~\cite{deSouzaEtAl2019}.
+
+\subsection{Threats to Validity} \label{sec_threats_to_validity}
+
+We follow Ampatzoglou et al.\ \cite{AmpatzoglouEtAl2019}'s analysis of
+threats to validity in software engineering secondary studies.
+
+\subsubsection{Reliability}
+
+A study is reliable if repeated by another researcher, using the same
+methodology, would lead to the same results~\cite{RunesonAndHost2009}.
+We identify the following threats:
 
 \begin{itemize}
-
-\item \textbf{A Contributing file} provides new contributors with the
-information that they need to start adding/modifying the repository's files.
-Abdalla \cite{Abdalla2016} provides a simple template for creating an
-open-source contributor guideline.
-
-\item \textbf{A Developer Code of Conduct} explicitly states the expectations
-for developers on how they should treat one another \cite{TouraniEtAl2017}.
-Three popular codes of conduct are \cite{TouraniEtAl2017}:
-\href{https://www.contributor-covenant.org/version/2/1/code_of_conduct/}
-{Contributor Covenant}, \href{https://ubuntu.com/community/code-of-conduct}
-{Ubuntu Code of Conduct}, and \href{https://www.djangoproject.com/conduct/}
-{Django Code of Conduct}. A code of conduct can improve inclusivity, which
-brings the benefit of a wider pool of contributors.  For example, a code of
-conduct can improve the participation of women \cite{SinghEtAl2021}.
-
-\item \textbf{Code Style Guidelines} present standards for writing code. Style
-guides codify such elements as formatting, commenting, naming identifiers, best
-practices and dangers to avoid \cite{Carty2020}.  Three sample style guides are:
-\href{https://google.github.io/styleguide/javaguide.html} {Google Java Style
-Guide},
-\href{http://cnl.sogang.ac.kr/cnlab/lectures/programming/python/PEP8_Style_Guide.pdf}
-{PEP8 Style Guide for Python}, and
-\href{https://google.github.io/styleguide/cppguide.html} {Google \CC Style
-Guide}.  Linting tools, like \href{https://pypi.org/project/flake8/}{flake8} for
-Python, can be used to enforced coding styles, like the PEP8 standard.
-
-\item \textbf{A Product Roadmap} explains the vision and direction for a product
-offering \cite{MunchEtAl2019}.  Although they have different forms, all
-roadmaps cover the following: \begin{inparaenum}[i)]
-	\item Where are we now?,
-	\item Where are we going?, and
-	\item How can we get there? \end{inparaenum} \cite{PhaalEtAl2005}. A product
-roadmap provides the following benefits: continuity of purpose, facilitation of
-collaboration, and assistance with prioritization \cite{Pichler2012}.
-
-\item \textbf{Design Documentation} explicitly states the design goals and
-priorities, records the likely changes, decomposes the complex system into
-separate modules, and specifies the relationship between the modules. Design
-documentation shows the syntax of the interface for the modules, and in some
-cases also documents the semantics.  Some potential elements of design
-documentation include representation of the system design and class design using
-Unified Modelling Language class diagrams \cite{Gamma1995}, or rigorous
-documentation of the system design following the template for a Module Guide
-(MG) \cite{ParnasEtAl1984}.  An MG organizes the modules in a hierarchy by their
-secrets.
-
-\item \textbf{API Documentation} shows developers the services or data provided
-by the software application (or library) through such resources as its methods
-or objects \cite{MengEtAl2018}.  Understandability is improved by API
-documentation \cite{MengEtAl2018}. API documentation can be generated using
-tools like Doxygen, pydoc, and javadoc.
-
-\end{itemize}
-
-The rare artifacts for MI software are similar to the rare artifacts for Lattice
-Boltzmann solvers \cite{SmithEtAl2024}, except LBM software is more likely to
-have developer related artifacts, like Contributing, Dependency list, and Design
-documentation.
-
-To improve MI software in the future, an increased use of checklists could help.
-Developers can use checklists to ensure they follow best practices.  Some
-examples include checklists merging branches into master \cite{Brown2015},
-checklists for saving and sharing changes to the project \cite{WilsonEtAl2016},
-checklists for new and departing team members \cite{HerouxAndBernholdt2018},
-checklists for processes related to commits and releases \cite{HerouxEtAl2008}
-and checklists for overall software quality \cite{ThielEtAl2020, SSI2022}.
-
-The above discussion shows that, taken together, MI projects fall somewhat short
-of recommended best practices for research software.  However, MI software is
-not alone in this.  Many, if not most, research projects fall short of best
-practices.  A gap exists in research software development practices and software
-engineering recommendations \cite{Storer2017}.  Johanson and Hasselbring observe
-that the state-of-the-practice for research software in industry and academia
-does not incorporate state-of-the-art SE tools and methods
-\cite{JohansonAndHasselbring2018}.  This causes sustainability and reliability
-problems \cite{FaulkEtAl2009}. Rather than benefit from capturing and reusing
-previous knowledge, projects waste time and energy ``reinventing the wheel''
-\cite{deSouzaEtAl2019}.
-
-\subsection{Threats to Validity} \label{sec_threats_to_validity}
-
-Our categories for threat to validity come from an analysis of software
-engineering secondary studies by Ampatzoglou et al.\ \cite{AmpatzoglouEtAl2019},
-where a secondary study analyzes the data from a set of primary studies.
-Ampatzoglou et al.\ is appropriate because a common example of a secondary study
-is a systematic literature review. Our methodology is a systematic software
-review --- the primary studies are the software packages, and our work collects
-and analyzes these primary studies.  We identified similar threats to validity
-in our assessment of the state of the practice of Lattice Boltzmann Solvers
-\cite{SmithEtAl2024}.
-
-\subsubsection{Reliability}
-
-A study is reliable if repetition of the study by different researchers using
-the original study's methodology would lead to the same results
-\cite{RunesonAndHost2009}. Reliability means that data and analysis are
-independent of the specific researcher(s) doing the study.  For the current
-study the identified reliability related threats are as follows:
-
-\begin{itemize}
-\item One individual does the manual measures for all packages. A different
+\item One individual does the measures for all packages. A different
 evaluator might find different results, due to differences in abilities,
-experiences, and biases.
-\item The manual measurements for the full set of packages took several months.
+experience, and biases.
+\item The measurements for the full set of packages took several months
+(of elapsed time).
 Over this time the software repositories may have changed and the reviewer's
@@
 
-We reduced concern over the reliability risk associated with the reviewer's
-judgement by demonstrating that the measurement process is reasonably
-reproducible.  In previous work \cite{SmithEtAl2016} we graded five software
+The measurement process used has previously been shown to be reasonably
+reproducible.  \cite{SmithEtAl2016} reports grading five software
 products by two reviewers. Their rankings were almost identical. As long as each
@@
 Construct validity is when the adopted metrics represent what they are intended
-to measure \cite{RunesonAndHost2009}. Our construct threats are often related to
-how we assume our measurements influences the various software qualities, as
-summarized in Section~\ref{sec_grading_software}.  Specifically, our construct
-validity related threats include the following:
-
+to measure~\cite{RunesonAndHost2009}. We have identified the following
+potential issues:
 \begin{itemize}
@@
 112]{VanVliet2000}.
-\item As mentioned in Section~\ref{sec_result_installability}, we could not
+\item We could not
 install or build \textit{dwv}, \textit{GATE}, and \textit{DICOM Viewer}. We used
 a deployed on-line version for \textit{dwv}, a VM version for \textit{GATE}, but
-no alternative for \textit{DICOM Viewer}. We might underestimate their rank due
-to these technical issues.
-\item Measuring software robustness only involved two pieces of data. This is
-likely part of the reason for limited variation in the robustness scores
-(Figure~\ref{fg_robustness_scores}). We could add more robustness data by
-pushing the software to deal with more unexpected situations, like a broken
-Internet connection, but this would require a larger investment of measurement
-time. 
-\item We may have inaccurately estimated maintainability by assuming a higher
-ratio of comments to source code improves maintainability. Moreover, we assumed
-that maintainability is improved if a high percentage of issues are closed, but
-a project may have a wealth of open issues, and still be maintainable.
-\item We assess reusability by the number of code files and LOC per file. This
-measure is indicative of modularity, but it does not necessarily mean a good
-modularization. The modules may not be general enough to be easily reused, or
-the formatting may be poor, or the understandability of the code may be low.
+no alternative for \textit{DICOM Viewer}.
+\item Robustness measurements involve only two pieces of data, leading
+to limited variation in the robustness scores
+(Figure~\ref{fg_robustness_scores}). Our measurement-time budget limited
+what we could achieve her.
+\item Our maintainability proxies (higher ratio of comments to source,
+high percentage of closed issues) have not been validated.
+\item While smaller modules tend to be easier to reuse, small modules are
+not necessarily good modules, nor understandable modules.
 \item The understandability measure relies on 10 random source code files, but
 the 10 files will not necessarily be representative. 
-\item As discussed in Section~\ref{Sec_OverallQ}, our overall AHP ranking makes
-the unrealistic assumption of equal weighting.
-\item We approximated popularity by stars and watches
-(Section~\ref{Sec_VsCommunityRanking}), but this assumption may not be valid. 
-\item In building Table~\ref{Tbl_Guidelines} some judgement was necessary on our
-part, since not all guidelines use the same names for artifacts that contain
+\item Our overall AHP ranking makes the unrealistic assumption of equal
+weighting.
+\item We approximated popularity by stars and watches.
+\item Table~\ref{Tbl_Guidelines} required judgement as
+not all guidelines use the same names for artifacts that contain
 essentially the same information.
@@
 Internal validity means that discovered causal relations are trustworthy and
-cannot be explained by other factors \cite{RunesonAndHost2009}. In our
-methodology the internal validity threats include the following:
+cannot be explained by other factors~\cite{RunesonAndHost2009}. We identify
+the following:
 
 \begin{itemize}
-\item In our search for software packages (Section~\ref{ch_results}), we may
+\item Our search (Section~\ref{ch_results}) may
 have missed a relevant package.
-\item Our methodology assumes that all relevant software development activities
+\item Our methodology assumes that development activities
 will leave a trace in the repositories, but this is not necessarily true. For
@@
 (Section~\ref{Sec_CompareArtifacts}), but teams might keep this kind of
-information outside their repos, possibly in journal papers or technical
-reports.
+information outside their repos.
 \end{itemize}
@@
 
-If the results of a study can be generalized (applied) to other
-situations/cases, then the study is externally valid \cite{RunesonAndHost2009}.
-We are confident that our search was exhaustive.  We do not believe that we
-missed any highly popular MI examples.  Therefore, the bulk of our validity
-concerns are internal (Section~\ref{Sec_InternalValidity}).  However, our hope
-is that the trends observed, and the lessons learned for MI software can be
-applied to other research software.  With that in mind we identified the
-following threat to external validity: We cannot generalize our results if the
-development of MI software is fundamentally different from other research
-software.
-
-Although there are differences, like the importance of data privacy for MI data,
-we found the approach to developing LBM software \cite{SmithEtAl2024} and MI
-software to be similar.  Except for the domain specific aspects, we believe that
-the trends observed in the current study are externally valid for other research
-software.
+If the results of a study can be generalized to other situations, then the
+study is externally valid \cite{RunesonAndHost2009}.  In other words, cannot
+generalize our results if the development of MI software is fundamentally
+different from other research software.  Although there are differences, like
+the importance of data privacy for MI data, we found the approach to developing
+LBM software \cite{SmithEtAl2024} and MI software to be similar.  Except for
+the domain specific aspects, the trends observed in the current study are
+similar to that for other research software.
 
@@
 
-We analyzed the state of the practice for the MI domain with the goal of
-understanding current practice by answering four research questions
-(Section~\ref{sec_motivation}).  Our methods in Section~\ref{SecMethodology} form a
-general process to evaluate domain-specific software, that we apply to the
-specific domain of MI software. We identified 48 MI software candidates, then,
-with the help of the Domain Expert selected 29 of them for our final list.
-
-Section~\ref{ch_results} lists our measurement results for ranking the 29
-projects for nine software qualities. Our ranking results appear credible since
-they are mostly consistent with the ranking from the scientific community
-implied by the GitHub stars-per-year metric. As discussed in
-Section~\ref{Sec_VsCommunityRanking}, four of the top five software projects
-appear in both our list and in the GitHub popularity list.  Moreover, our top
-five packages appear among the first eight positions on the GitHub list.  The
-noteworthy discrepancies between the two lists are for the packages that we were
-unable to install (\textit{dwv} and \textit{Dicom Viewer}). Our findings suggest
-that the community's top ranked packages tend to be those that incorporate best
-practices.
-
-Based on our grading scores \textit{3D Slicer}, \textit{ImageJ}, \textit{Fiji}
-and \textit{OHIF Viewer} are the top four software performers.  However, the
-separation between the top performers and the others is not extreme.  Almost all
-packages do well on at least a few qualities, as shown in
+Our analysis of the state of the practice for the MI domain along nine
+software qualities strongly indicates that ``higher quality'' is
+consistent with community ranking proxies. It is important to note that
+our quality measures are, on purpose, rather shallow. This undescores the
+importance that \emph{first impressions} of any software has on its
+adoption.
+
+\begin{table}[ht!]
+	%\begin{center}
+	\renewcommand{\arraystretch}{1.8}
+	\begin{tabular}{ p{2.3cm}p{9.9cm} }
+		\toprule
+
+		Quality & Ranked 1st or 2nd\\
+
+		\midrule
+
+		Installability & 3D Slicer, BioImage Suite Web, Slice:Drop, INVESALIUS\\
+
+		\pbox{3.0cm}{Correctness \\ and Verifiability} & OHIF Viewer, 3D
+		Slicer, ImageJ\\
+
+		Reliability & SMILI, ImageJ, Fiji, 3D Slicer, Slice:Drop, OHIF
+		Viewer\\
+
+		Robustness & XMedCon, Weasis, SMILI, ParaView, OsiriX Lite,
+		MicroView, medInria, ITK-SNAP, INVESALIUS, ImageJ, Horos, Gwyddion,
+		Fiji, dicompyler, DicomBrowser, BioImage Suite Web, AMIDE, 3DimViewer,
+		3D Slicer, OHIF Viewer, DICOM Viewer\\
+
+		Usability & 3D Slicer, ImageJ, Fiji, OHIF Viewer, ParaView,
+		INVESALIUS, Ginkgo CADx, SMILI, OsiriX Lite, BioImage Suite Web,
+		ITK-SNAP, medInria, MicroView, Gwyddion\\
+
+		Maintainability & 3D Slicer, Weasis, ImageJ, OHIF Viewer, ParaView\\
+
+		Reusability & 3D Slicer, ImageJ, Fiji, OHIF Viewer, SMILI, dwv, BioImage
+		Suite Web, GATE, ParaView\\
+
+		\pbox{3.0cm}{Understandability} & 3D Slicer, ImageJ, Weasis,
+		Fiji, Horos, OsiriX Lite, dwv, Drishti, OHIF Viewer, GATE, ITK-SNAP,
+		ParaView, INVESALIUS\\
+
+		\pbox{3.0cm}{Visibility and \\Transparency} & ImageJ, 3D Slicer, Fiji\\
+
+		Overall Quality & 3D Slicer, ImageJ\\
+
+		\bottomrule		
+	\end{tabular}
+	\caption{Top performers for each quality (sorted by order of quality
+	measurement)} \label{topperformerstable}
+	%\end{center}
+\end{table}
+
+Our grading scores indicate that
+\textit{3D Slicer}, \textit{ImageJ}, \textit{Fiji}
+and \textit{OHIF Viewer} are the overall top four.  However, the
+separation between the top performers and the others is not extreme.  Almost
+all packages do well on at least a few qualities, as shown in
 Table~\ref{topperformerstable}, which summarizes the packages ranked first and
@@
 \textit{DicomBrowser}, \textit{AMIDE}, \textit{3DimViewer}, and
-\textit{Drishti}. The shortness of this list suggests parity with respect to
-adoption of best practices for MI software overall.
-
-\begin{table}[ht!]
-	%\begin{center}
-	\renewcommand{\arraystretch}{1.8}
-	\begin{tabular}{ p{2.3cm}p{9.9cm} }
-		\toprule
-
-		Quality & Ranked 1st or 2nd\\
-
-		\midrule
-
-		Installability & 3D Slicer, BioImage Suite Web, Slice:Drop, INVESALIUS\\
-
-		\pbox{3.0cm}{Correctness \\ and Verifiability} & OHIF Viewer, 3D
-		Slicer, ImageJ\\
-
-		Reliability & SMILI, ImageJ, Fiji, 3D Slicer, Slice:Drop, OHIF
-		Viewer\\
-
-		Robustness & XMedCon, Weasis, SMILI, ParaView, OsiriX Lite,
-		MicroView, medInria, ITK-SNAP, INVESALIUS, ImageJ, Horos, Gwyddion,
-		Fiji, dicompyler, DicomBrowser, BioImage Suite Web, AMIDE, 3DimViewer,
-		3D Slicer, OHIF Viewer, DICOM Viewer\\
-
-		Usability & 3D Slicer, ImageJ, Fiji, OHIF Viewer, ParaView,
-		INVESALIUS, Ginkgo CADx, SMILI, OsiriX Lite, BioImage Suite Web,
-		ITK-SNAP, medInria, MicroView, Gwyddion\\
-
-		Maintainability & 3D Slicer, Weasis, ImageJ, OHIF Viewer, ParaView\\
-
-		Reusability & 3D Slicer, ImageJ, Fiji, OHIF Viewer, SMILI, dwv, BioImage
-		Suite Web, GATE, ParaView\\
-
-		\pbox{3.0cm}{Understandability} & 3D Slicer, ImageJ, Weasis,
-		Fiji, Horos, OsiriX Lite, dwv, Drishti, OHIF Viewer, GATE, ITK-SNAP,
-		ParaView, INVESALIUS\\
-
-		\pbox{3.0cm}{Visibility and \\Transparency} & ImageJ, 3D Slicer, Fiji\\
-
-		Overall Quality & 3D Slicer, ImageJ\\
-
-		\bottomrule		
-	\end{tabular}
-	\caption{Top performers for each quality (sorted by order of quality
-	measurement)} \label{topperformerstable}
-	%\end{center}
-\end{table}
-
-When it comes to recommended software artifacts, the state of open source MI
-projects is healthy, with our surveyed examples showing 88\% of the
-documentation artifacts recommended by research software development guidelines
-(Section~\ref{Sec_CompareArtifacts}).  However, we did notice areas where
-practice seems to lag behind the research software development guidelines.  For
-instance, the guidelines recommend three artifacts that were not observed:
-uninstall instructions, test plans, and requirements documentation. We observed
-the following recommended artifacts, but only rarely: contributing file,
-developer code of conduct, code style guidelines, product roadmap, design
-documentation, and API documentation (Section~\ref{Sec_CompareArtifacts}).
-Future MI projects may wish to consider incorporating more artifacts to
-potentially improve software quality.  A deeper understanding of the needs of
-the MI community will require data beyond what is available in open source
-repositories.  Future work should involve interviewing MI developers to more
-deeply understand their software development pain points.
+\textit{Drishti}.
+
+While we did find a reasonable amount of documentation, especially when
+consider all MI projects, there were definitely some holes. Some important
+documentation (test plans and requirements documentation) was missing,
+and other (contributors' guide, code of conduct, code style guidelines, product
+roadmap, design documentation, and API documentation
+(Section~\ref{Sec_CompareArtifacts}) were rare.
+
+A deeper understanding of the needs of
+the MI community will require data beyond what is available in
+repositories.  Future work should involve interviewing MI developers to
+better understand their ``pain points''.
 

### 2024-07-12 StateOfPractice/Papers/SOP_MI_OSR/SOP_MI_OSR.tex
Revise document to smooth out changes made by JC.  Ready for double blind next.

@@
 
-\affil*[1]{\orgdiv{Computing and Software Department}, \orgname{McMaster University}, \orgaddress{\street{1280 Main Street West}, \city{Hamilton}, \postcode{L8S 4K1}, \state{Ontario}, \country{Canada}}}
-
-\affil[2]{\orgdiv{Electrical and Computer Engineering Department}, \orgname{McMaster University}, \orgaddress{\street{1280 Main Street West}, \city{Hamilton}, \postcode{L8S 4K1}, \state{Ontario}, \country{Canada}}}
-
-% \author[1,*]{W.\ Spencer Smith}
-% \author[1]{Ao Dong}
-% \author[1]{Jacques Carette}
-% \author[2]{Michael D.\ Noseworthy}
-
-% \affil[1]{McMaster University, Computing and Software Department, {Canada}}
-% \affil[2]{McMaster University, Electrical \& Computer Engineering Department,
-% {Canada}}
-% \affil[*]{Corresponding Author}
+\affil*[1]{\orgdiv{Computing and Software Department}, \orgname{McMaster
+University}, \orgaddress{\street{1280 Main Street West}, \city{Hamilton},
+\postcode{L8S 4K1}, \state{Ontario}, \country{Canada}}}
+
+\affil[2]{\orgdiv{Electrical and Computer Engineering Department},
+\orgname{McMaster University}, \orgaddress{\street{1280 Main Street West},
+\city{Hamilton}, \postcode{L8S 4K1}, \state{Ontario}, \country{Canada}}}
 
@@
 questions for each. Using the Analytic Hierarchy Process (AHP) on the
-quantitive data, we ranked the MI software.  The top five are \textit{3D
+quantitative data, we ranked the MI software.  The top five are \textit{3D
 Slicer}, \textit{ImageJ}, \textit{Fiji}, \textit{OHIF Viewer}, and
-\textit{ParaView}.  This is consistent with the community's, with four of
+\textit{ParaView}.  This is consistent with the community's view, with four of
 these also appearing in the top five using GitHub metrics (stars-per-year).  
-In general, the quality and quantity of documentation present in the
+In general, the quality and quantity of documentation present in a
 project correlates quite well with its popularity.
-%We observed 88\% of the documentation artifacts
-%recommended by research software development guidelines.  However, the current
-%state of the practice deviates from the existing guidelines because of the
-%rarity of some recommended artifacts (like test plans, requirements
-%specification, code of conduct, code style guidelines, product roadmaps, and
-%Application Program Interface (API) documentation).
 }
@@
 \maketitle
-
-% RQ1 to RQ4
-% Some of Section 1 (Introduction)
-% Sections 2 to 6 (includes looking at artifacts observed in repos)
-% Some of Sections 11 to 13 (Threats to Validity, Future Work, Conclusions)
 
@@
 Imaging (MRI), Computed Tomography (CT), Positron Emission Tomography (PET) and
-Ultrasound) to provide information for diagnostic, analytic, and medical
+Ultrasound) to provide critical information for diagnostic, analytic, and medical
 applications. Given its importance, we want to understand the merits and
 drawbacks of the current development processes, tools, and methodologies. We
-use a software engineering lens to assess the quality of existing MI software
-and provide recommendations for future development.
+use a software engineering lens to assess the quality of existing MI software.
 
@@
 
-%We have a standard set of questions designed to assess the qualities of any
-%research software project \cite{SmithEtAl2021, SmithAndMichalski2022}. We
-%particularly focus on \emph{software artifactcs}, i.e.
-%the documents, scripts and code that constitutes a software development
-%project. Example artifacts include requirements, specifications, user manuals,
-%unit tests, system tests, usability tests, build scripts, API (Application
-%Programming Interface) documentation, READMEs, license documents, process
-%documents, and code.  %Once we have learned what MI developers do, we then put
-%% this information in context by contrasting MI software against the trends shown
-%% by developers in other research software communities.
-
 Our investigation is based on the following four research questions:
@@
 \textbf{open source} software projects exist? (Section~\ref{ch_results})
-\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_HighestQuality}:] Establish
-quantitative measurements of those project's development practices. From
-these measures, which of those projects follow best practices?
-(Section~\ref{ch_results})
+\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_HighestQuality}:] Based on
+quantitative measurements of each project's development practices, which
+projects follow best practices? (Section~\ref{ch_results})
 \item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareHQ2Popular}:] How
@@
 do MI projects compare to general research software with respect to the
-artifacts present in their repositories?
+artifacts (documents, scripts and code) present in their repositories?
 (Section~\ref{Sec_CompareArtifacts})
@@
 
-We only cover MI visualization software.  We exclude
-other categories of MI software, including
-Segmentation, Registration, Visualization, Enhancement, Quantification,
-Simulation, plus MI archiving and telemedicine systems (Compression, Storage,
-and Communication).  We also exclude Statistical Analysis and Image-based
-Physiological Modelling and Feature Extraction, Classification, and
-Interpretation. Software that provides MI support functions is also out of
+We only cover MI visualization software.  We exclude other categories of MI
+software, including Segmentation, Registration, Visualization, Enhancement,
+Quantification, Simulation, plus MI archiving and telemedicine systems
+(Compression, Storage, and Communication).  We also exclude Statistical Analysis
+and Image-based Physiological Modelling and Feature Extraction, Classification,
+and Interpretation. Software that provides MI support functions is also out of
 scope; therefore, we have not assessed the toolkit libraries VTK and ITK.
@@
 We have a standard set of questions designed to assess the qualities of any
-research software project~\cite{SmithEtAl2021, SmithAndMichalski2022}.
-This has been applied to MI software and Lattice Boltzmann Solvers
-\cite{SmithEtAl2024}.  This builds off prior work to assess the
-state of the practice for such domains as Geographic Information Systems
-\cite{smith2018state}, Mesh Generators \cite{smith2016state}, Seismology
-software \cite{Smith2018Seismology}, and Statistical software for psychology
-\cite{smith2018statistical}.  We maintain the previous constraint that the
-work load for measuring a given domain should take around one person-month's
-worth of effort ($20$ working days at $8$ person-hours per day).
-
-We identify a list of potential packages (through online searches) which is
-then filtered and vetted by a domain expert. We aim for roughly $30$
-packages. For each remaining package, we measure its qualities by filling in a
-grading template~\cite{SmithEtAl2021}.  This data is used to rank the projects
-with AHP.  Further details on interaction with the domain expert, software
-qualities, grading the software and AHP are summarizes below and in longer form
-in~\cite{SmithEtAl2024_MI_SOP}.
+research software project~\cite{SmithEtAl2021, SmithAndMichalski2022}.  This has
+been applied to MI software and Lattice Boltzmann Solvers \cite{SmithEtAl2024}.
+This builds off prior work to assess the state of the practice for such domains
+as Geographic Information Systems \cite{smith2018state}, Mesh Generators
+\cite{smith2016state}, Seismology software \cite{Smith2018Seismology}, and
+Statistical software for psychology \cite{smith2018statistical}.  We maintain
+the previous constraint that the work load for measuring a given domain should
+take around one person-month's worth of effort ($20$ working days at $8$
+person-hours per day).
+
+We identify a list of potential packages (through online searches) which is then
+filtered and vetted by a domain expert. We aim for roughly $30$ packages. For
+each remaining package, we measure its qualities by filling in a grading
+template~\cite{SmithEtAl2021}.  This data is used to rank the projects with the
+Analytic Hierarchy Process (AHP).  We summarize further details on the
+interaction with the domain expert, software qualities, grading the software and
+AHP below and in longer form in Smith et al.\
+(2024)~\cite{SmithEtAl2024_MI_SOP}.
 
@@
 
-The Domain Expert vets the proposed list. For too many domains, there are
-online resources that are wildly inaccurate, which only an expert can correct.
-(Details of our domain expert removed for double-blind).
-%The Domain Experts vet the proposed list of software packages and the AHP
-%ranking.  Doing these tasks with non-experts runs the risk of trusting
-%inaccurate resources and overlooking the tacit knowledge of an expert. For the
-%current assessment, our Domain Expert (and paper co-author) is Dr.\ Michael
+The Domain Expert vets the proposed list because online resources can be
+inaccurate.  The expert also vets the AHP ranking.  For the current assessment,
+our Domain Expert is [Details of our domain expert removed for double-blind].
+%%DOUBLE BLIND %our Domain Expert (and paper co-author) is Dr.\ Michael
 %Noseworthy, Professor of Electrical and Computer Engineering at McMaster
 %University, Co-Director of the McMaster School of Biomedical Engineering, and
-%Director of Medical Imaging Physics and Engineering at St.\ Joseph's Healthcare,
-%Hamilton, Ontario, Canada.
+%Director of Medical Imaging Physics and Engineering at St.\ Joseph's
+%Healthcare, Hamilton, Ontario, Canada.
 
@@
 
-We use an existing template~\cite{SmithEtAl2021} that is designed to measure
-the aforementioned qualities. To stay within our given measurement time frame,
-each package gets up to five hours of time.  Project developers can be
-contacted for help regarding installation, if necessary, but we impose a cap of
-about two hours on the installation process.
-Figure~\ref{fg_grading_template_example} shows an excerpt of the measurement
-spreadsheet.  It includes a column for each measured software package.
-% \cite{Dong2021-Data} provides the full set of measurement data.
-(The full data is available on Mendeley; link will be provided after
-refereeing.)
+We use an existing template~\cite{SmithEtAl2021} that is designed to measure the
+aforementioned qualities. To stay within our given measurement time frame, each
+package gets up to five hours of time.  Project developers can be contacted for
+help regarding installation, if necessary, but we impose a cap of about two
+hours on the installation process.  Figure~\ref{fg_grading_template_example}
+shows an excerpt of the measurement spreadsheet.  The rows are the measures and
+the columns correspond to the software packages.  [The full data is available on
+Mendeley; link will be provided after refereeing.] 
+% \cite{Dong2021-Data}
+%provides the full set of measurement data.   %%DOUBLE BLIND
 
@@
 with constrained time and domain knowledge.
-%We
-%designed the questions to be unambiguous, quantifiable, and measurable with
-%limited time and domain knowledge. We group the measures under headings for each
-%quality, and one for summary information. The summary information (shown in
-%Figure~\ref{fg_grading_template_example}) is the first section of the template.
-%This section summarizes general information, such as the software name, purpose,
-%platform, programming language, publications about the software, the first
-%release and the most recent change date, website, source code repository of the
-%product, number of developers, etc.  We only measure open source software that
-%is ready for public use by other researchers (following the definition of the
-%public category from \cite{GewaltigAndCannon2012}). Information in the summary
-%section sets the context for the project, but it does not directly affect the
-%grading scores.
-
-%For measuring each quality, we ask several questions and the typical answers are
-%among the collection of ``yes'', ``no'', ``n/a'', ``unclear'', a number, a
-%string, a date, a set of strings, etc. 
-The grader, after answering questions for each quality assigned an overall
+
+The grader, after answering questions for each quality assigns an overall
 score (between 1 and 10) based on the answers.  Several of the qualities use
 the word ``surface'' to highlight that these particular qualities are a shallow
-measure.  For example, we do not do user studies for measuring usability.
-Instead, we are look signs that the developers considered usability.
+measure.  For example, usability is not measured using user studies.
+Instead, we look for signs that the developers considered usability.
 We use two freeware tools to collect repository related data:
@@
 \href{https://github.com/boyter/scc}{Sloc Cloc and Code (scc)}.  Further
-details on quality measurement are provided in~% Smith et al.\ (2024)
+details on quality measurement are provided in Smith et al.\ (2024)
 \cite{SmithEtAl2024_MI_SOP}.
@@
 
-Developed by Saaty in the 1970s, AHP is widely used to 
-analyze multiple criteria decisions~\cite{VaidyaEtAl2006}. AHP organizes
-multiple criteria in a hierarchical structure and uses pairwise comparisons
-between alternatives to calculate relative ratios~\cite{Saaty1990}. AHP works
-with sets of $n$ \textit{options} and $m$ \textit{criteria}.  In our project
-$n=29$ and $m=9$ since there are 29 options (software products) and 9 criteria
-(qualities). With AHP the sum of the grades (scores) for all products for a
-given quality will be 1.0.  We rank the software for each of the qualities, and
-then we combine the quality rankings into an overall ranking based on the
-relative priorities between qualities.
+Developed by Saaty in the 1970s, AHP is widely used to analyze multiple criteria
+decisions~\cite{VaidyaEtAl2006}. AHP organizes multiple criteria in a
+hierarchical structure and uses pairwise comparisons between alternatives to
+calculate relative ratios~\cite{Saaty1990}. AHP works with sets of $n$
+\textit{options} and $m$ \textit{criteria}.  In our project $n=29$ and $m=9$
+since there are 29 options (software products) and 9 criteria (qualities). With
+AHP the sum of the grades (scores) for all products for a given quality will be
+1.0.  We rank the software for each of the qualities, and then we combine the
+quality rankings into an overall ranking based on the relative priorities
+between qualities.
 
@@
 
-We initially identified 48 candidate software projects from
-the literature \cite{Bjorn2017, Bruhschwein2019, Haak2015}, on-line articles
-\cite{Emms2019, Hasan2020, Mu2019}, and forum discussions \cite{Samala2014}.
-Then we filtered as follows:
+We initially identified 48 candidate software projects from the literature
+\cite{Bjorn2017, Bruhschwein2019, Haak2015}, on-line articles \cite{Emms2019,
+Hasan2020, Mu2019}, and forum discussions \cite{Samala2014}.  Then we filtered
+as follows:
 
@@
 
-\item Focused on MI software that provides visualization functions.
-We removed seven packages
-that were toolkits or libraries, such as \textit{VTK}, \textit{ITK}, and
-\textit{dcm4che}, and another three that were for PACS.
+\item Focused on MI software that provides visualization functions.  We removed
+seven packages that were toolkits or libraries, such as \textit{VTK},
+\textit{ITK}, and \textit{dcm4che}, and another three that were for PACS.
 
@@
 
-The Domain Expert provided a list 12 software packages.
-We found 6 packages were on both lists: \textit{3D
-Slicer}, \textit{Horos}, \textit{ImageJ}, \textit{Fiji}, \textit{MRIcron} 
-(we use its descendant \textit{MRIcroGL}) and \textit{Mango} (we
-use the web version \textit{Papaya}).  The remaining six packages were on our
-out-of-scope list. The Domain Expert agreed with our final choice of 29
-packages -- see Table~\ref{tab_final_list}. This contains summary data
-collected in the year 2020.
-
-The projects are sorted in descending order of lines of code.
-We found the initial release dates (Rlsd) for most
-projects and marked the two unknown dates with ``?''. The date of the last
-update is the date of the latest update, at the time of measurement. We found
-funding information (Fnd) for only eight projects.  For the Number Of
-Contributors (NOC) we considered anyone who made at least one accepted commit as
-a contributor. The NOC is not usually the same as the number of long-term
-project members, since many projects received change requests and code from the
-community.  With respect to the OS, 25 packages work on all three OSs: Windows
-(W), macOS (M), and Linux (L). Although the usual approach to cross-platform
-compatibility was to work natively on multiple OSes, five projects achieved
-platform-independence via web applications. The full measurement data for all
-packages is available on \href{https://data.mendeley.com/datasets/k3pcdvdzj2/1}
-{Mendeley Data}.
+The Domain Expert provided a list 12 software packages.  We found 6 packages
+were on both lists: \textit{3D Slicer}, \textit{Horos}, \textit{ImageJ},
+\textit{Fiji}, \textit{MRIcron} (we use its descendant \textit{MRIcroGL}) and
+\textit{Mango} (we use the web version \textit{Papaya}).  The remaining six
+packages were on our out-of-scope list. The Domain Expert agreed with our final
+choice of 29 packages -- see Table~\ref{tab_final_list}. This contains summary
+data collected in the year 2020.
+
+The projects are sorted in descending order of lines of code.  We found the
+initial release dates (Rlsd) for most projects and marked the two unknown dates
+with ``?''. The date of the last update is the date of the latest update, at the
+time of measurement. We found funding information (Fnd) for only eight projects.
+For the Number Of Contributors (NOC) we considered anyone who made at least one
+accepted commit as a contributor. The NOC is not usually the same as the number
+of long-term project members, since many projects received change requests and
+code from the community.  With respect to the OS, 25 packages work on all three
+OSs: Windows (W), macOS (M), and Linux (L). Although the usual approach to
+cross-platform compatibility was to work natively on multiple OSes, five
+projects achieved platform-independence via web applications. The full
+measurement data for all packages is available on [removed for blind review]
+%\href{https://data.mendeley.com/datasets/k3pcdvdzj2/1} {Mendeley Data}.  
+%% DOUBLE BLIND
 
@@
 for all but the bottom four, just various degrees of ease and automation.
-
 \textit{GATE}, \textit{dwv}, and \textit{DICOM Viewer} showed severe
@@
 from the source code, but we were able to proceed with measuring other qualities
-using a deployed on-line version for \textit{dwv}, and a VM version for
+using a deployed on-line version for \textit{dwv}, and a virtual machine version for
 \textit{GATE}. For \textit{DICOM Viewer} we could not install the NextCloud
@@
 Figure~\ref{fg_correctness_verifiability_scores}) used a wider array of
-techniques to improve correctness, and had better documentation to witness
-this. For instance, we looked for evidence of unit testing, and only evidence
-for about half of the projects. We identified five projects using CI/CD
-tools: \textit{3D Slicer}, \textit{ImageJ}, \textit{Fiji}, \textit{dwv}, and
-\textit{OHIF Viewer}.
+techniques to improve correctness, and had better documentation to witness this.
+For instance, we looked for evidence of unit testing, and found evidence for
+only about half of the projects. We identified five projects using continuous
+integration tools: \textit{3D Slicer}, \textit{ImageJ}, \textit{Fiji},
+\textit{dwv}, and \textit{OHIF Viewer}.
 
@@
 
-Even for some projects with well-organized documentation, requirements
+Even for projects with well-organized documentation, requirements
 specifications and theory manuals were still missing.
@@
 
-Figure~\ref{fg_reliability_scores} shows the our results.  We were able to
-follow the steps in the tutorials that existed (seven packages had them.)
-However
+Figure~\ref{fg_reliability_scores} shows our results.  We were able to follow
+the steps in the tutorials that existed (seven packages had them.) However,
 \textit{GATE} could not open macro files and became unresponsive several times,
@@
 Figure \ref{fg_robustness_scores} presents the scores for surface robustness.
-The packages with higher scores elegantly handled unexpected or unanticipated
+The packages with higher scores gracefully handled unexpected or unanticipated
 inputs, typically showing a clear error message. We may have underscored
@@
 maintainability. We gave \textit{3D Slicer} the highest score because we found
-it to have the most comprehensive artifacts. Only a few of the 29 projects had a
-product, developer's manual, or API documentation, and only \textit{3D Slicer},
-\textit{ImageJ}, \textit{Fiji} included all three documents 
-(see Table~\ref{tab_maintainability_docs} the full data).  Moreover,
-\textit{3D Slicer} has a much higher percentage of closed issues (92\%) compared
-to \textit{ImageJ} (52\%) and \textit{Fiji} (64\%).
+it had the most comprehensive artifacts. Only a few of the 29 projects had a
+product, developer's manual, or API (Application Programming Interface)
+documentation, and only \textit{3D Slicer}, \textit{ImageJ}, \textit{Fiji}
+included all three documents (see Table~\ref{tab_maintainability_docs} for the
+full data).  Moreover, \textit{3D Slicer} has a much higher percentage of closed
+issues (92\%) compared to \textit{ImageJ} (52\%) and \textit{Fiji} (64\%).
 
@@
 
-API documents (see Table~\ref{tab_maintainability_docs}) also help maintainability,
-which explains some of the scores shown in
+API documents (see Table~\ref{tab_maintainability_docs}) also help
+maintainability, which explains some of the scores shown in
 Figure~\ref{fg_reusability_scores}. We have assumed that smaller code files are
-likely more reusable -- see
-Table \ref{tab_loc_per_file} for the details.
+likely more reusable -- see Table \ref{tab_loc_per_file} for the details.
 
@@
 compare the state of the practice for MI with that of other research software.
-In particular we provide details on recommended artifacts but rarely observed
+In particular we provide details on recommended artifacts that are rarely observed
 for MI software. Section~\ref{sec_threats_to_validity} presents threats to the
@@
 because we were unable to build it locally, even though we followed its
-installation instructions. But we were able to use its web
+installation instructions. However, we were able to use its web
 version for the rest of the measurements. Additionally, this version 
@@
 
-We use nine research software development guidelines to measure what
-software artifacts are present in MI software, versus those recommended.
+We use nine research software development guidelines to compare recommended 
+software artifacts versus those present in MI software.
 These guidelines are:
@@
 Table~\ref{artifactspresent} presents our measurements for MI software.
-It groups the artifacts by frequency
+The table groups the artifacts by frequency
 into categories of common (20 to 29 ($>$67\%) packages), uncommon (10 to 19
@@
 scattered amongst many other documents.
-Three of the 26 recommended artifacts were never observed:
+
+Three of the 26 recommended artifacts were never observed in the MI software:
 i) Uninstall, ii) Test plans, and iii) Requirements. It is possible that
@@
 
-It is unfortunately common for research software to neglect requirements
-documentation, and MI software is no exception.
-Although such documentation is recommended by some
-\cite{TobiasEtAl2018, HerouxEtAl2008, SmithAndKoothoor2016}, in practice
-this is rare~\cite{HeatonAndCarver2015}. Sanders and Kelly
-\cite{SandersAndKelly2008} interviewed 16 scientists from 10 disciplines and
-found that none of the scientists created requirements specifications, unless
-regulations in their field mandated such a document.  Requirements are the least
-commonly produced type of documentation for research software in general
-\cite{Nguyen-HoanEtAl2010}.
-
-This is quite unfortunate as when surveyed, research software developers
-identify the lack of requirements and requirements management as accounting
-gets mentioned the most often and accounts for 23\% of the technical problems
-\cite{WieseEtAl2019}.  There is unfortunately widespread perception that
-up-front requirements are impossible for research software
-\cite{CarverEtAl2007, SegalAndMorris2008}. Fortunately a more agile approach
-to requirements is feasible~\cite{Smith2016}, and research-software specific
-templates exist~\cite{SmithEtAl2007}.
+Neglecting requirements documentation is unfortunately common for research
+software, and MI software is no exception to this trend.  Although such
+documentation is recommended by some \cite{TobiasEtAl2018, HerouxEtAl2008,
+SmithAndKoothoor2016}, in practice this is rare~\cite{HeatonAndCarver2015}.
+Sanders and Kelly \cite{SandersAndKelly2008} interviewed 16 scientists from 10
+disciplines and found that none of the scientists created requirements
+specifications, unless regulations in their field mandated such a document.
+Requirements are the least commonly produced type of documentation for research
+software in general \cite{Nguyen-HoanEtAl2010}.
+
+This is unfortunate as when scientific developers are surveyed on their pain
+points, Wiese et al.~\cite{WieseEtAl2019} found that software requirements and
+management is the software engineering discipline that most hurts them,
+accounting for 23\% of the technical problems reported by study participants.
+Further adding to the misfortune, there is a widespread perception that up-front
+requirements are impossible for research software \cite{CarverEtAl2007,
+SegalAndMorris2008}. Fortunately a more agile approach to requirements is
+feasible~\cite{Smith2016}, and research-software specific templates
+exist~\cite{SmithEtAl2007}.
 
@@
 
-Other communities use checklists to help with best practices. Examples
-include checklists for
-merging branches~\cite{Brown2015},
-for saving and sharing changes to the project~\cite{WilsonEtAl2016},
-for new and departing team members~\cite{HerouxAndBernholdt2018},
-for processes related to commits and releases~\cite{HerouxEtAl2008}
-and for overall software quality~\cite{ThielEtAl2020, SSI2022}.
-
-MI projects fall somewhat short of recommended best practices, but are not
-alone amongst research software projects. This gap has been
-documented before~\cite{Storer2017,JohansonAndHasselbring2018}, is known
-to cause sustainability and reliability
-problems \cite{FaulkEtAl2009}, and to waste development effort%
-~\cite{deSouzaEtAl2019}.
+Other communities use checklists to help with best practices. Examples include
+checklists for merging branches~\cite{Brown2015}, for saving and sharing changes
+to the project~\cite{WilsonEtAl2016}, for new and departing team
+members~\cite{HerouxAndBernholdt2018}, for processes related to commits and
+releases~\cite{HerouxEtAl2008} and for overall software
+quality~\cite{ThielEtAl2020, SSI2022}.
+
+MI projects fall somewhat short of recommended best practices, but are not alone
+amongst research software projects. This gap has been documented
+before~\cite{Storer2017,JohansonAndHasselbring2018}, and is known to cause
+sustainability and reliability problems \cite{FaulkEtAl2009}, and to waste
+development effort~\cite{deSouzaEtAl2019}.
 
@@
 
-A study is reliable if repeated by another researcher, using the same
+A study is reliable if repeating it by another researcher, using the same
 methodology, would lead to the same results~\cite{RunesonAndHost2009}.
@@
 experience, and biases.
-\item The measurements for the full set of packages took several months
-(of elapsed time).
-Over this time the software repositories may have changed and the reviewer's
-judgement may have drifted.
+\item The measurements for the full set of packages took several months (of
+elapsed time).  Over this time the software repositories may have changed and
+the reviewer's judgement may have drifted.
 \end{itemize}
@@
 112]{VanVliet2000}.
-\item We could not
-install or build \textit{dwv}, \textit{GATE}, and \textit{DICOM Viewer}. We used
-a deployed on-line version for \textit{dwv}, a VM version for \textit{GATE}, but
-no alternative for \textit{DICOM Viewer}.
+\item We could not install or build \textit{dwv}, \textit{GATE}, and
+\textit{DICOM Viewer}. We used a deployed on-line version for \textit{dwv}, a
+virtual machine version for \textit{GATE}, but no alternative for \textit{DICOM
+Viewer}.
 \item Robustness measurements involve only two pieces of data, leading
@@
 (Figure~\ref{fg_robustness_scores}). Our measurement-time budget limited
-what we could achieve her.
+what we could achieve here.
 \item Our maintainability proxies (higher ratio of comments to source,
@@
 \begin{itemize}
-\item Our search (Section~\ref{ch_results}) may
+\item Our search (Section~\ref{ch_results}) could
 have missed a relevant package.
@@
 If the results of a study can be generalized to other situations, then the
-study is externally valid \cite{RunesonAndHost2009}.  In other words, cannot
+study is externally valid \cite{RunesonAndHost2009}.  In other words, we cannot
 generalize our results if the development of MI software is fundamentally
@@
 
-Our analysis of the state of the practice for the MI domain along nine
-software qualities strongly indicates that ``higher quality'' is
-consistent with community ranking proxies. It is important to note that
-our quality measures are, on purpose, rather shallow. This undescores the
-importance that \emph{first impressions} of any software has on its
-adoption.
+Our analysis of the state of the practice for MI domain along nine software
+qualities strongly indicates that ``higher quality'' is consistent with
+community ranking proxies. Although our quality measures are rather shallow, we
+see this as an advantage. The shallow measures are a proxy for the importance of
+\emph{first impressions} for software adoption.
 
@@
 and other (contributors' guide, code of conduct, code style guidelines, product
-roadmap, design documentation, and API documentation
+roadmap, design documentation, and API documentation)
 (Section~\ref{Sec_CompareArtifacts}) were rare.

### 2024-07-12 StateOfPractice/Papers/SOP_MI_OSR/SOP_MI_OSR.tex
Ready for submission

@@
 
-\usepackage{booktabs}
-
 \usepackage{amssymb}
-\usepackage{amsmath,amsthm}
+\usepackage{amsmath}
 
@@
 \usepackage{multirow}
-\usepackage{pgfplots}
-\usepackage{arydshln}
-\usepackage{paralist}
 \usepackage{booktabs}
@@
 
-\newcounter{pnum} %pain point number
-\newcommand{\ppthepnum}{P`'\thepnum}
-\newcommand{\ppref}[1]{P\ref{#1}}
-
-\newcounter{qnum} %quality number
-\newcommand{\qthepnum}{Q`'\theqnum}
-\newcommand{\qref}[1]{Q\ref{#1}}
-
 \newcommand{\CC}{C\nolinebreak\hspace{-.05em}\raisebox{.4ex}{\small\bf
@@
 
-%\pgfplotsset{compat=1.13}
-
 % prepared for: Journal of Imaging Informatics in Medicine
@@
 
-\author*[1]{\fnm{W.\ Spencer} \sur{Smith}}\email{smiths@mcmaster.ca}
-
-\author[1]{\fnm{Ao} \sur{Dong}}
-
-\author[1]{\fnm{Jacques} \sur{Carette}}\email{carette@mcmaster.ca}
-
-\author[2]{\fnm{Michael} \sur{Noseworthy}}\email{nosewor@mcmaster.ca}
-
-\affil*[1]{\orgdiv{Computing and Software Department}, \orgname{McMaster
-University}, \orgaddress{\street{1280 Main Street West}, \city{Hamilton},
-\postcode{L8S 4K1}, \state{Ontario}, \country{Canada}}}
-
-\affil[2]{\orgdiv{Electrical and Computer Engineering Department},
-\orgname{McMaster University}, \orgaddress{\street{1280 Main Street West},
-\city{Hamilton}, \postcode{L8S 4K1}, \state{Ontario}, \country{Canada}}}
+%% DOUBLE BLIND
+% \author*[1]{\fnm{W.\ Spencer} \sur{Smith}}\email{smiths@mcmaster.ca}
+
+% \author[1]{\fnm{Ao} \sur{Dong}}
+
+% \author[1]{\fnm{Jacques} \sur{Carette}}\email{carette@mcmaster.ca}
+
+% \author[2]{\fnm{Michael} \sur{Noseworthy}}\email{nosewor@mcmaster.ca}
+
+% \affil*[1]{\orgdiv{Computing and Software Department}, \orgname{McMaster
+% University}, \orgaddress{\street{1280 Main Street West}, \city{Hamilton},
+% \postcode{L8S 4K1}, \state{Ontario}, \country{Canada}}}
+
+% \affil[2]{\orgdiv{Electrical and Computer Engineering Department},
+% \orgname{McMaster University}, \orgaddress{\street{1280 Main Street West},
+% \city{Hamilton}, \postcode{L8S 4K1}, \state{Ontario}, \country{Canada}}}
 
@@
 We have a standard set of questions designed to assess the qualities of any
-research software project~\cite{SmithEtAl2021, SmithAndMichalski2022}.  This has
-been applied to MI software and Lattice Boltzmann Solvers \cite{SmithEtAl2024}.
-This builds off prior work to assess the state of the practice for such domains
-as Geographic Information Systems \cite{smith2018state}, Mesh Generators
-\cite{smith2016state}, Seismology software \cite{Smith2018Seismology}, and
-Statistical software for psychology \cite{smith2018statistical}.  We maintain
+research software project.  [blind review redacted details on the author's
+previous studies using the same methodology.]
+%% DOUBLE BLIND
+% ~\cite{SmithEtAl2021, SmithAndMichalski2022}.  This has
+% been applied to MI software and Lattice Boltzmann Solvers \cite{SmithEtAl2024}.
+% This builds off prior work to assess the state of the practice for such domains
+% as Geographic Information Systems \cite{smith2018state}, Mesh Generators
+% \cite{smith2016state}, Seismology software \cite{Smith2018Seismology}, and
+% Statistical software for psychology \cite{smith2018statistical}.  
+We maintain
 the previous constraint that the work load for measuring a given domain should
@@
 each remaining package, we measure its qualities by filling in a grading
-template~\cite{SmithEtAl2021}.  This data is used to rank the projects with the
+template [citation redacted for double blind]. %%DOUBLE BLIND
+%~\cite{SmithEtAl2021}.  
+This data is used to rank the projects with the
 Analytic Hierarchy Process (AHP).  We summarize further details on the
 interaction with the domain expert, software qualities, grading the software and
-AHP below and in longer form in Smith et al.\
-(2024)~\cite{SmithEtAl2024_MI_SOP}.
+AHP below and in longer form in [redacted for double blind].
+%% DOUBLE BLIND
+%Smith et al.\ (2024)~\cite{SmithEtAl2024_MI_SOP}.
 
@@
 inaccurate.  The expert also vets the AHP ranking.  For the current assessment,
-our Domain Expert is [Details of our domain expert removed for double-blind].
+our Domain Expert is [details of our domain expert removed for double-blind].
 %%DOUBLE BLIND %our Domain Expert (and paper co-author) is Dr.\ Michael
@@
 
-We use an existing template~\cite{SmithEtAl2021} that is designed to measure the
+We use an existing template [citation redacted]
+%% DOUBLE BLIND
+%~\cite{SmithEtAl2021} 
+that is designed to measure the
 aforementioned qualities. To stay within our given measurement time frame, each
@@
 \href{https://github.com/boyter/scc}{Sloc Cloc and Code (scc)}.  Further
-details on quality measurement are provided in Smith et al.\ (2024)
-\cite{SmithEtAl2024_MI_SOP}.
+details on quality measurement are provided in [redacted for double blind].
+%% DOUBLE BLIND
+%Smith et al.\ (2024) \cite{SmithEtAl2024_MI_SOP}.
 
@@
 The measurement process used has previously been shown to be reasonably
-reproducible.  \cite{SmithEtAl2016} reports grading five software
+reproducible.  [citation redacted] 
+%% DOUBLE BLIND
+%Smith et al.~\cite{SmithEtAl2016} 
+reports grading five software
 products by two reviewers. Their rankings were almost identical. As long as each
@@
 the importance of data privacy for MI data, we found the approach to developing
-LBM software \cite{SmithEtAl2024} and MI software to be similar.  Except for
+[other software (citation redacted)]
+%% DOUBLE BLIND
+%LBM software \cite{SmithEtAl2024} 
+and MI software to be similar.  Except for
 the domain specific aspects, the trends observed in the current study are
@@
 
-We would like to thank Peter Michalski and Oluwaseun Owojaiye for fruitful
-discussions on topics relevant to this paper.
+[redacted for blind review]
+%% DOUBLE BLIND
+% We would like to thank Peter Michalski and Oluwaseun Owojaiye for fruitful
+% discussions on topics relevant to this paper.
 

### 2024-10-28 StateOfPractice/Papers/SOP_MI_CritReviewsBioMedEng/SOP_MI_OSR.tex
Initial draft of Critical Reviews in BioMed Eng paper

--- 2022-11-02 StateOfPractice/Papers/MedImageSoft_SOP-arXiv/MedImageSoft_SOP.tex
@@
-%% 
-%% Copyright 2007-2020 Elsevier Ltd
-%% 
-%% This file is part of the 'Elsarticle Bundle'.
-%% ---------------------------------------------
-%% 
-%% It may be distributed under the conditions of the LaTeX Project Public
-%% License, either version 1.2 of this license or (at your option) any
-%% later version.  The latest version of this license is in
-%%    http://www.latex-project.org/lppl.txt
-%% and version 1.2 or later is part of all distributions of LaTeX
-%% version 1999/12/01 or later.
-%% 
-%% The list of all files belonging to the 'Elsarticle Bundle' is
-%% given in the file `manifest.txt'.
-%% 
-%% Template article for Elsevier's document class `elsarticle'
-%% with Harvard style bibliographic references
-
-%\documentclass[preprint,12pt,authoryear]{elsarticle}
-%\documentclass[3p, 12pt,authoryear]{elsarticle}
-\documentclass[final, 3p, times, authoryear]{elsarticle}
-
-%% Use the option review to obtain double line spacing
-%% \documentclass[authoryear,preprint,review,12pt]{elsarticle}
-
-%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
-%% for a journal layout:
-%% \documentclass[final,1p,times,authoryear]{elsarticle}
-%% \documentclass[final,1p,times,twocolumn,authoryear]{elsarticle}
-%% \documentclass[final,3p,times,authoryear]{elsarticle}
-%% \documentclass[final,3p,times,twocolumn,authoryear]{elsarticle}
-%% \documentclass[final,5p,times,authoryear]{elsarticle}
-%% \documentclass[final,5p,times,twocolumn,authoryear]{elsarticle}
-
-%% For including figures, graphicx.sty has been loaded in
-%% elsarticle.cls. If you prefer to use the old commands
-%% please give \usepackage{epsfig}
-
-%% The amssymb package provides various useful mathematical symbols
+\documentclass[final, 12pt, 3p, times]{elsarticle}
+
 \usepackage{amssymb}
-\usepackage{amsmath,amsthm}
-
-%% The lineno packages adds line numbers. Start line numbering with
-%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
-%% for the whole article with \linenumbers.
-%% \usepackage{lineno}
+\usepackage{amsmath}
 
@@
 
-%\usepackage{tikz}
-%\usetikzlibrary{mindmap}
 \usepackage{multirow}
-\usepackage{pgfplots}
-\usepackage{arydshln}
-\usepackage{paralist}
 \usepackage{booktabs}
 \usepackage{pbox}
-
-%% Comments
-\newif\ifcomments\commentstrue
-
-\ifcomments
-\newcommand{\authornote}[3]{\textcolor{#1}{[#3 ---#2]}}
-\newcommand{\todo}[1]{\textcolor{red}{[TODO: #1]}}
-\else
-\newcommand{\authornote}[3]{}
-\newcommand{\todo}[1]{}
-\fi
-
-\newcommand{\wss}[1]{\authornote{blue}{SS}{#1}} %Spencer Smith
-\newcommand{\jc}[1]{\authornote{red}{JC}{#1}} %Jacques Carette
-\newcommand{\mn}[1]{\authornote{magenta}{MN}{#1}} %Mike Noseworthy
-\newcommand{\ad}[1]{\authornote{cyan}{AD}{#1}} %Ao Dong
-
-\newcommand{\notdone}[1]{\textcolor{red}{#1}}
-\newcommand{\done}[1]{\textcolor{black}{#1}}
+\usepackage{paralist}
 
@@
 
-\journal{Academic Radiology}
-
-\pgfplotsset{compat=1.13}
+% prepared for: Journal of Imaging Informatics in Medicine
+% https://link.springer.com/journal/10278
+
+\journal{Critical Reviews in Biomedical Engineering}
 
@@
 
-\title{State of the Practice for Medical Imaging Software}
+\title{State of the Practice for Medical Imaging Software Based on Open Source
+Repositories}
 
@@
 
+%\author{Anonymous}
 \author[CAS]{Spencer Smith}
@@
 \author[CAS]{Jacques Carette}
-\author[ECE]{Mike Noseworthy}
-
-\affiliation[CAS]{organization={McMaster University, Computing and Software Department},%Department and Organization
+\author[EE]{Michael Noseworthy}
+
+\affiliation[CAS]{organization={McMaster University, Computing and Software
+Department}, %Department and Organization
             addressline={1280 Main Street West}, 
@@
 
-\affiliation[ECE]{organization={McMaster University, Electrical Engineering},%Department and Organization
+\affiliation[ME]{organization={McMaster University, Electrical and Computer Engineering}, %Department and Organization
             addressline={1280 Main Street West}, 
@@
 
-We present the state of the practice for Medical Imaging (MI) software. We
-selected 29 medical imaging projects from 48 candidates, assessed 10 software
-qualities (installability, correctness/ verifiability, reliability, robustness,
-usability, maintainability, reusability, understandability,
-visibility/transparency and reproducibility) by answering 108 questions for each
-software project, and interviewed 8 of the 29 development teams. Based on the
-quantitative data for the first 9 qualities, we ranked the MI software with the
-Analytic Hierarchy Process (AHP). The four top ranked software products are:
-\textit{3D Slicer}, \textit{ImageJ}, \textit{Fiji}, and \textit{OHIF Viewer}.
-Our ranking is mostly consistent with the community's ranking, with four of our
-top five projects also appearing in the top five of a list ordered by
-stars-per-year. Generally, MI software is in a healthy state as shown by the
-following: in the repositories we observed 88\% of the documentation artifacts
-recommended by research software development guidelines, 100\% of MI projects
-use version control tools, and developers appear to use the common quasi-agile
-research software development process. However, the current state of the
-practice deviates from the existing guidelines because of the rarity of some
-recommended artifacts (like test plans, requirements specification, code of
-conduct, code style guidelines, product roadmaps, and Application Program
-Interface (API) documentation), low usage of continuous integration (17\% of the
-projects), low use of unit testing (about 50\% of projects), and room for
-improvement with documentation (six of nine developers felt their documentation
-wasn't clear enough). From interviewing the developers, we identified five pain
-points and two qualities of potential concern: lack of development time, lack of
-funding, technology hurdles, ensuring correctness, usability, maintainability,
-and reproducibility. The interviewees proposed strategies to improve the state
-of the practice, to address the identified pain points, and to improve software
-quality. Combining their ideas with ours, we have the following list of
-recommendations: increase documentation, increase testing by enriching datasets,
-increase continuous integration usage, move to web applications, employ linters,
-use peer reviews, design for change, add assurance cases, and incorporate a
-``Generate All Things'' approach.
+    We present the state of the practice for Medical Imaging (MI) software based
+on data available in open source repositories. We selected 29 projects from 48
+candidates and assessed 9 software qualities (installability, correctness/
+verifiability, reliability, robustness, usability, maintainability, reusability,
+understandability, and visibility/transparency) by answering 108 questions for
+each. Using the Analytic Hierarchy Process (AHP) on the quantitative data, we
+ranked the MI software.  The top five are \textit{3D Slicer}, \textit{ImageJ},
+\textit{Fiji}, \textit{OHIF Viewer}, and \textit{ParaView}.  This is consistent
+with the community's view, with four of these also appearing in the top five
+using GitHub metrics (stars-per-year).  
+In general, the quality and quantity of documentation present in a project
+correlates quite well with its popularity.
 
@@
 	medical imaging, research software, software engineering, software
-	quality, Analytic Hierarchy Process, developer interviews
+	quality, analytic hierarchy process
 \end{keyword}
@@
 
-We aim to study the state of software development practice for Medical Imaging
-(MI) software.  MI tools use images of the interior of the body (from sources
-such as Magnetic Resonance Imaging (MRI), Computed Tomography (CT), Positron
-Emission Tomography (PET) and Ultrasound) to provide information for diagnostic,
-analytic, and medical applications \citep{FDA2021, enwiki:1034887445,
-Zhang2008}.  Figure~\ref{Fig_Example}, which shows an image of the brain,
-highlights the importance and value of MI. Through MI medical practitioners and
-researchers can noninvasively gain insights into the human body, including
-information on injuries and illnesses. Given the importance of MI software and
-the high number of competing software projects, we wish to understand the merits
-and drawbacks of the current development processes, tools, and methodologies. We
-aim to assess through a software engineering lens the quality of the existing
-software with the hope of highlighting standout examples, understanding current
-pain points and providing guidelines and recommendations for future development.
-
-\begin{figure}[!ht]
-    \begin{center}
-        \includegraphics[scale=0.25]{figures/MPR.png}        
-    \end{center}
-    \caption{Example brain image showing a multi-planar reformat using Horos
-	(free open-source medical imaging/DICOM viewer for OSX, based on OsiriX)}
-    \label{Fig_Example}
-\end{figure}
+We study the state of software development practice for Medical Imaging
+(MI) software using data available in open source repositories.  MI tools use
+images of the interior of the body (from sources such as Magnetic Resonance
+Imaging (MRI), Computed Tomography (CT), Positron Emission Tomography (PET) and
+Ultrasound) to provide critical information for diagnostic, analytic, and medical
+applications. Given its importance, we want to understand the merits and
+drawbacks of the current development processes, tools, and methodologies. We
+use a software engineering lens to assess the quality of existing MI software.
 
@@
 
-Not only do we wish to gain insight into the state of the practice for MI
-software, we also wish to understand the development of research software in
-general. We wish to understand the impact of the often cited gap, or chasm,
-between software engineering and research software \citep{Kelly2007,
-Storer2017}. Although scientists spend a substantial proportion of their working
-hours on software development \citep{Hannay2009, Prabhu2011}, many developers
-learn software engineering skills by themselves or from their peers, instead of
-from proper training \citep{Hannay2009}. \citet{Hannay2009} observe that many
-scientists showed ignorance and indifference to standard software engineering
-concepts. For instance, according to a survey by \citet{Prabhu2011}, more than
-half of their 114 subjects did not use a proper debugger when coding.
-
-To gain insights, we devised 10 research questions, which can be applied to MI,
-as well as to other domains, of research software \citep{SmithEtAl2021,
-SmithAndMichalski2022}.  We designed the questions to learn about the
-community's interest in, and experience with, software artifacts, tools,
-principles, processes, methodologies, and qualities.  When we mention artifacts
-we mean the documents, scripts and code that constitutes a software development
-project. Example artifacts include requirements, specifications, user manuals,
-unit tests, system tests, usability tests, build scripts, API (Application
-Programming Interface) documentation, READMEs, license documents, process
-documents, and code.  Once we have learned what MI developers do, we then put
-this information in context by contrasting MI software against the trends shown
-by developers in other research software communities.  Our aim is to collect
-enough information to understand the current pain points experienced by the MI
-software development community so that we can make some preliminary
-recommendations for future improvements.
-
-We based the structure of the paper on the research questions, so for each
-research question below we point to the section that contains our answer.  We
-start with identifying the relevant examples of MI software for the assessment
-exercise:
-
+As well as state of the practice (SOP) for MI software, we would like to
+understand the impact of the often cited gap (chasm!) between recommended
+software engineering practices and that used for most research 
+software~\cite{Storer2017}. Although scientists spend
+a substantial proportion of their working times on software development
+\cite{Hannay2009, Prabhu2011}, few are formally trained~\cite{Hannay2009}.
+
+Our investigation is based on the following four research questions:
 \begin{enumerate}
 \item[RQ\refstepcounter{rqnum}\therqnum \label{RQ_WhatProjects}:] What MI
-	software projects exist, with the constraint that the source code must be
-	available for all identified projects? (Section~\ref{ch_results})
-	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_HighestQuality}:] Which
-	of the projects identified in \rqref{RQ_WhatProjects} follow current best
-	practices, based on evidence found by experimenting with the software and
-	searching the artifacts available in each project's repository?
-	(Section~\ref{ch_results})
+\textbf{open source} software projects exist? (Section~\ref{ch_results})
+\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_HighestQuality}:] Based on
+quantitative measurements of each project's development practices, which
+projects follow best practices? (Section~\ref{ch_results})
 \item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareHQ2Popular}:] How
-	similar is the list of top projects identified in \rqref{RQ_HighestQuality}
-	to the most popular projects, as viewed by the scientific community?
+similar are the top projects identified in \rqref{RQ_HighestQuality}
+to the most popular projects as viewed by the community?
 (Section~\ref{Sec_VsCommunityRanking})
 \item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareArtifacts}:] How
-	do MI projects compare to research software in general with respect to the
-	artifacts present in their repositories?
+do MI projects compare to general research software with respect to the
+artifacts (documents, scripts and code) present in their repositories?
 (Section~\ref{Sec_CompareArtifacts})
-	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareToolsProjMngmnt}:]
-	How do MI projects compare to research software in general with respect to
-	the use of tools (Section~\ref{Sec_CompareTools}) for:
-	\begin{enumerate} 
-		\item [\rqref{RQ_CompareToolsProjMngmnt}.a] development; and,
-		\item [\rqref{RQ_CompareToolsProjMngmnt}.b] project management?
 \end{enumerate}
-	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareMethodologies}:]
-	How do MI projects compare to research software in general with respect to
-	principles, processes, and methodologies used?
-	(Section~\ref{Sec_CompareMethodologies})
-	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_PainPoints}:] What are
-	the pain points for developers working on MI software projects?
-	(Section~\ref{painpoints})
-	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_ComparePainPoints}:] How
-	do the pain points of developers from MI compare to the pain points
-	for research software in general? (Section~\ref{painpoints})
-	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Concerns}:] For MI
-	developers what specific best practices are taken to address the pain points
-	and software quality concerns? (Section~\ref{painpoints})
-	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Recommend}:]
-	What research software development practice could potentially address the
-	pain point concerns identified in \rqref{RQ_PainPoints}?
-	(Section~\ref{ch_recommendations})
-
-\end{enumerate}
 
@@
 
-To make the project feasible, we only cover MI visualization software.  As a
-consequence we are excluding many other categories of MI software, including
-Segmentation, Registration, Visualization, Enhancement, Quantification,
-Simulation, plus MI archiving and telemedicine systems (Compression, Storage,
-and Communication) (as summarized by \citet{Bankman2000} and
-\citet{Angenent2006}).  We also exclude Statistical Analysis and Image-based
-Physiological Modelling \citep{enwiki:1034877594} and Feature Extraction,
-Classification, and Interpretation \citep{Kim2011}. Software that provides MI
-support functions is also out of scope; therefore, we have not assessed the
-toolkit libraries VTK \citep{SchroederEtAl2006} and ITK \citep{McCormick2014}.
+We only cover MI visualization software.  We exclude other categories of MI
+software, including Segmentation, Registration, Visualization, Enhancement,
+Quantification, Simulation, plus MI archiving and telemedicine systems
+(Compression, Storage, and Communication).  We also exclude Statistical Analysis
+and Image-based Physiological Modelling and Feature Extraction, Classification,
+and Interpretation. Software that provides MI support functions is also out of
+scope; therefore, we have not assessed the toolkit libraries VTK and ITK.
 Finally, Picture Archiving and Communication System (PACS), which helps users to
-economically store and conveniently access images \citep{Choplin1992}, are
-considered out of scope.
-
-\subsection{Methodology Overview}
-
-We designed a general methodology to assess the state of the practice for
-research software \citep{SmithEtAl2021, SmithAndMichalski2022}. Details can be
-found in Section~\ref{ch_methods}.  Our methodology has been applied to MI
-software \citep{Dong2021} and Lattice Boltzmann Solvers \citep{SmithEtAl2022,
-Michalski2021}.  This methodology builds off prior work to assess the state of
-the practice for such domains as Geographic Information Systems
-\citep{smith2018state}, Mesh Generators \citep{smith2016state}, Seismology
-software \citep{Smith2018Seismology}, and Statistical software for psychology
-\citep{smith2018statistical}.  In keeping with the previous methodology, we have
-maintained the constraint that the work load for measuring a given domain should
-be feasible for a team as small as one person, and for a short time, ideally
-around a person month of effort. We consider a person month as $20$ working days
-($4$ weeks in a month, with $5$ days of work per week) at $8$ person hours per
-day, or $20 \times 8 = 160$ person hours.
-
-With our methodology, we first choose a research software domain (in the current
-case MI) and identify a list of about 30 software packages. (For measuring MI we
-used 29 software packages.)  We approximately measure the qualities of each
-package by filling in a grading template. Compared with our previous
-methodology, the new methodology also includes repository based metrics, such as
-the number of files, number of lines of code, percentage of issues that are
-closed, etc.  With the quantitative data in the grading template, we rank the
-software with the Analytic Hierarchy Process (AHP) (Section~\ref{ch_background}
-provides details). After this, as another addition to our previous methodology,
-we interview some development teams to further understand the status of their
-development process.
-
-\section{Background} \label{ch_background}
-
-To measure the existing MI software we need two sets of definitions: i) the
-definitions of relevant software license models (Section
-\ref{sec_software_categories}); and, ii) the definitions of the software
-qualities that we will be assessing (Section \ref{sec_software_quality}). In our
-assessment we rank the software packages for each quality; therefore, this
-section also provides the background on our ranking process --- the Analytic
-Hierarchy Process (Section \ref{sec_AHP}).
-
-\subsection{Software Categories} \label{sec_software_categories}
-
-When assessing software packages, we need to know the software's license.  In
-particular, we need to know whether the source code will be available to us or
-not.  We define three common software categories.  We will only assess software
-that fits under the Open Source Software license.
-
-\begin{itemize}
-
-\item \textbf{Open Source Software (OSS)} For OSS, the source code is openly
-accessible. Users have the right to study, change and distribute it under a
-license granted by the copyright holder. For many OSS projects, the development
-process relies on the collaboration of different contributors worldwide
-\citep{Corbly2014}. Accessible source code usually exposes more ``secrets'' of a
-software project, such as the underlying logic of software functions, how
-developers achieve their works, and the flaws and potential risks in the final
-product. Thus, OSS is suitable for researchers analyzing the qualities of a
-project.
-
-\item \textbf{Freeware} Freeware is software that can be used free of charge.
-Unlike OSS, the authors of do not allow access or modify the source code
-\citep{LINFO2006}. To many end-users, the differences between freeware and OSS
-may not be relevant. However, software developers who wish to modify the source
-code, and researchers looking for insight into software development process may
-find the inaccessible source code a problem.
-
-\item \textbf{Commercial Software} ``Commercial software is software developed
-by a business as part of its business'' \citep{GNU2019}. Typically speaking,
-commercial software requires users to pay to access all of its features,
-excluding access to the source code. However, some commercial software is also
-free of charge \citep{GNU2019}. Based on our experience, most commercial
-software products are not OSS.
-
-\end{itemize}
-
-\subsection{Software Quality Definitions} \label{sec_software_quality}
+economically store and conveniently access images, are considered out of scope.
+
+\subsection{Methodology} \label{SecMethodology}
+
+We have a standard set of questions designed to assess the qualities of any
+research software project.  [blind review redacted details on the author's
+previous studies using the same methodology.]
+%% DOUBLE BLIND
+% ~\cite{SmithEtAl2021, SmithAndMichalski2022}.  This has
+% been applied to MI software and Lattice Boltzmann Solvers \cite{SmithEtAl2024}.
+% This builds off prior work to assess the state of the practice for such domains
+% as Geographic Information Systems \cite{smith2018state}, Mesh Generators
+% \cite{smith2016state}, Seismology software \cite{Smith2018Seismology}, and
+% Statistical software for psychology \cite{smith2018statistical}.  
+We maintain
+the previous constraint that the work load for measuring a given domain should
+take around one person-month's worth of effort ($20$ working days at $8$
+person-hours per day).
+
+We identify a list of potential packages (through online searches) which is then
+filtered and vetted by a domain expert. We aim for roughly $30$ packages. For
+each remaining package, we measure its qualities by filling in a grading
+template [citation redacted for double blind]. %%DOUBLE BLIND
+%~\cite{SmithEtAl2021}.  
+This data is used to rank the projects with the
+Analytic Hierarchy Process (AHP).  We summarize further details on the
+interaction with the domain expert, software qualities, grading the software and
+AHP below and in longer form in [redacted for double blind].
+%% DOUBLE BLIND
+%Smith et al.\ (2024)~\cite{SmithEtAl2024_MI_SOP}.
+
+\subsubsection{Domain Expert} \label{sec_vet_software_list}
+
+The Domain Expert vets the proposed list because online resources can be
+inaccurate.  The expert also vets the AHP ranking.  For the current assessment,
+our Domain Expert is [details of our domain expert removed for double-blind].
+%%DOUBLE BLIND %our Domain Expert (and paper co-author) is Dr.\ Michael
+%Noseworthy, Professor of Electrical and Computer Engineering at McMaster
+%University, Co-Director of the McMaster School of Biomedical Engineering, and
+%Director of Medical Imaging Physics and Engineering at St.\ Joseph's
+%Healthcare, Hamilton, Ontario, Canada.
+
+In advance of the first meeting with the Domain Expert, they are asked to
+independently create a list of top software packages in the domain.  This helps
+get the expert's knowledge refreshed in advance of the meeting.
+
+\subsubsection{Software Qualities} \label{sec_software_quality}
 
 Quality is defined as a measure of the excellence or worth of an entity.  As is
-common practice, we do not think of quality as a single measure, but rather as a
-set of measures.  That is, quality is a collection of different qualities, often
-called ``ilities.''  Below we list the 10 qualities of interest for this study.
-The order of the qualities follows the order used in \citet{GhezziEtAl2003},
-which puts related qualities (like correctness and reliability) together.
-Moreover, the order is roughly the same as the order developers consider
-qualities in practice.
-
-\begin{itemize}
-	\item \textbf{Installability} The effort required for the installation
-    and/or uninstallation of software in a specified environment
-    \citep{ISO/IEC25010, lenhard2013measuring}.
-
-	\item \textbf{Correctness \& Verifiability} A program is correct if it
-    matches its specification \citep[p.\ 17]{GhezziEtAl2003}.  The specification
-    can either be explicitly or implicitly stated.  The related quality of
-    verifiability is the ease with which the software components or the
-    integrated product can be checked to demonstrate its correctness.
-
-	\item \textbf{Reliability} The probability of failure-free operation of a
-	computer program in a specified environment for a specified time \citep{musa1987software}, \citep[p.\ 357]{GhezziEtAl2003}.
-
-	\item \textbf{Robustness} Software possesses the characteristic of
-	robustness if it behaves ``reasonably'' in two situations: i) when it
-	encounters circumstances not anticipated in the requirements specification,
-	and ii) when users violate the assumptions in its requirements specification 
-	\citep[p.\ 19]{GhezziEtAl2003}, \citep{boehm2007software}.
-
-	\item \textbf{Usability} ``The extent to which a product can be used by
-	specified users to achieve specified goals with effectiveness, efficiency,
-	and satisfaction in a specified context of use'' \citep{ISO/TR16982:2002}
-	\citep{ISO9241-11:2018}.
-
-	\item \textbf{Maintainability} The effort with which a software system or
-	component can be modified to i) correct faults; ii) improve performance or
-	other attributes; iii) satisfy new requirements
-	\citep{IEEEStdGlossarySET1990}, \citep{boehm2007software}.
-
-	\item \textbf{Reusability} ``The extent to which a software component can be
-	used with or without adaptation in a problem solution other than the one for
-	which it was originally developed'' \citep{kalagiakos2003non}.
-
-	\item \textbf{Understandability} ``The capability of the software product to
-	enable the user to understand whether the software is suitable, and how it
-	can be used for particular tasks and conditions of use'' \citep{iso2001iec}.
-
-	\item \textbf{Visibility/Transparency} The extent to which all the steps
-	of a software development process and the current status of it are conveyed
-	clearly \citep[p.\ 32]{GhezziEtAl2003}.
-
-	\item \textbf{Reproducibility} ``A result is said to be reproducible if
-	another researcher can take the original code and input data, execute it,
-	and re-obtain the same result'' \citep{BenureauAndRougier2017}.
-\end{itemize}
-
-\subsection{Analytic Hierarchy Process (AHP)} \label{sec_AHP}
-
-Saaty developed AHP in the 1970s, and people have widely used it since to make
-and analyze multiple criteria decisions \citep{VaidyaEtAl2006}. AHP organizes
-multiple criteria in a hierarchical structure and uses pairwise comparisons
-between alternatives to calculate relative ratios \citep{Saaty1990}. AHP works
-with sets of $n$ \textit{options} and $m$ \textit{criteria}.  In our project
-$n=29$ and $m=9$ since there are 29 options (software products) and 9 criteria
-(qualities). We rank the software for each of the qualities, and then we combine
-the quality rankings into an overall ranking based on the relative priorities
-between qualities.
-
-The first step for ranking the software choices for a given quality involves a
-pairwise comparison between each of the $n$ software options for that quality.
-AHP expresses the comparison through an $n \times n$ matrix $A$. When comparing
-option $i$ and option $j$, the value of $A_{ij}$ is decided as follows, with the
-value of $A_{ji}$ generally equal to $1/A_{ij}$ \citep{Saaty1990}: $A_{ij} = 1$
-if criterion $i$ and criterion $j$ are equally important, while $A_{ij} = 9$ if
-criterion $i$ is extremely more important than criterion $j$.  The natural
-numbers between 1 and 9 are used to show the different levels of relative
-importance between these two extremes. The above assumes that option $i$ is of
-equal, or more, importance compared to option $j$ ($i \geq j$).  If that is not
-the case, we reverse $i$ and $j$ and determine $A_{ji}$ first, then $A_{ij} =
-1/A_{ji}$.
-
-Section~\ref{sec_grading_software} shows how we measure the software via a
-grading template.  For the AHP process, the relevant measure is the subjective
-score from $1$ to $10$ for each quality for each package. To turn these
-subjective measures $x_{\text{sub}}$ and $y_{\text{sub}}$ into Saaty's
-pair-wise scores for option $x$ versus option $y$, respectively, we use the
-following calculation:
-\[
-\begin{cases}
-\min\{9, x_{\text{sub}} - y_{\text{sub}} + 1\} & x_{\text{sub}} \geq y_{\text{sub}} \\
-1 / \min\{9, y_{\text{sub}} - x_{\text{sub}} + 1\} & x_{\text{sub}} < y_{\text{sub}}
-\end{cases}
-\]
-
-\noindent For example, we measured the usability for 3D Slicer and Ginkgo CADx
-as $8$ and $7$, respectively; therefore, on the 9-point scale, 3D Slicer compared
-to Ginkgo CADx is 2 and Ginkgo CADx versus 3D Slicer is 1/2, as shown in the
-sample AHP calculations (Table~\ref{Tbl_SampleAHP}).
-
-The second step is to calculate the priority vector $w$ from $A$.  The
-vector $w$ ranks the software options by how well they achieve the given
-quality.  The priority vector can be calculated by solving the equation
-\citep{Saaty1990}:
-\begin{equation} 
-    A w = \lambda_{\text{max}} w,
-\end{equation}
-where $\lambda_{\text{max}}$ is the maximal eigenvalue of $A$.  In this project,
-$w$ is approximated with the classic \textit{mean of normalized values} approach
-\citep{AlessioEtAl2006}:
-
-\begin{equation}
-w_i = \frac{1}{n}\sum_{j=1}^{n}\frac{A_{ij}}{\sum_{k=1}^{n}A_{kj}}
-\end{equation}
-
-Table~\ref{Tbl_SampleAHP} summarizes the above two steps for the quality of
-installability.  The matrix $A$ is shown in the first set of columns, then the
-normalized version of $A$ and finally the average of the normalized values to
-form the vector $w$ in the last column.
-
-\begin{table}[h!]
-\begin{center}
-\begin{tabular}{ l c c c c c | c c c c c | c }
- \toprule
- ~ & \multicolumn{5}{c|}{$A_{ij}$} & \multicolumn{5}{c|}{${A_{ij}}/{\sum_{k=1}^{n}A_{kj}}$} & ~\\
- \midrule
- ~ & \rotatebox{90}{3D Slicer} & \rotatebox{90}{Ginkgo} & \rotatebox{90}{XMedCon} & $\cdots$ & \rotatebox{90}{Gwyddion} & \rotatebox{90}{3D Slicer} & \rotatebox{90}{Ginkgo} & \rotatebox{90}{XMedCon} & $\cdots$ & \rotatebox{90}{Gwyddion} & AVG \\
- \midrule
- 3D Slicer & 1 & 2 & 4 & $\cdots$ & 2 & 0.071 & 0.078 & 0.060 & $\cdots$ & 0.078 & 0.068\\
- Ginkgo & 1/2 & 1 & 3 & $\cdots$ & 1 & 0.036 & 0.039 & 0.045 & $\cdots$ & 0.039 & 0.041\\
- XMedCon & 1/4 & 1/3 & 1 & $\cdots$ & 1/3 & 0.018 & 0.013 & 0.015 & $\cdots$ & 0.013 & 0.015\\
- $\vdots$ & $\vdots$ & $\vdots$ & $\vdots$ & $\ddots$ & $\vdots$ & $\vdots$ & $\vdots$ & $\vdots$ & $\ddots$ & $\vdots$ & $\vdots$\\
- Gwyddion & 1/2 & 1 & 3 & $\cdots$ & 1 & 0.036 & 0.039 & 0.045 & $\cdots$ & 0.039 & 0.041\\  
- \midrule
- SUM = & 14.01 & 25.58 & 66.75 & $\cdots$ & 25.58 & 1.000 & 1.000 & 1.000 & $\cdots$ & 1.000 & 1.000\\
- \bottomrule
-\end{tabular}
-\end{center}
-\caption{Sample AHP Calculations for the Quality of Usability} \label{Tbl_SampleAHP}
-\end{table}
-
-We repeat the first and second steps for each of the qualities.  The third step
-combines the quality rankings into an overall ranking.  Following AHP, we need
-to first prioritize the qualities.  The AHP method finds the priority of quality
-$i$ ($p_i$) in the same way that the score ($w_j$) was found for software
-package $j$ evaluated for a given quality (as shown above).  That is, we
-conduct a pairwise comparison between the priority of different qualities to
-construct the $m \times m$ matrix $A$, and then we take the mean of normalized
-values for row $i$ to find the priority value $p_i$ for quality $i$.  If we
-introduce the notation that $w^i_j$ is the score for quality $i$ for package
-$j$, then the overall score $S_j$ for package $j$ is found via:
-
-$$S_j = \sum_{i=1}^m w^i_j p_i$$
-
-\section{Methodology} \label{ch_methods}
-
-We developed a methodology for evaluating the state of the practice of research
-software \citep{SmithEtAl2021, SmithAndMichalski2022}.  The methodology can be
-instantiated for a specific domain of scientific software, which in the current
-case is medical imaging software for visualization.  Our methodology involves
-and engages a domain expert partner throughout, as discussed in
-Section~\ref{sec_vet_software_list}.  The four main steps of the methodology
-are:
-
-\begin{enumerate}
-\item Identify list of representative software packages
-(Section~\ref{sec_software_selection});
-\item Measure (or grade) the selected software
-(Section~\ref{sec_grading_software});
-\item Interview developers (Section~\ref{sec_interview_methods});
-\item Answer the research questions (as given in Section~\ref{sec_motivation}).
-\end{enumerate}
-
-In the sections below we provide additional detail on the above steps, while
-concurrently giving examples of how we applied the methodology to the MI domain.
-
-\subsection{Interaction With Domain Expert} \label{sec_vet_software_list}
-
-The Domain Expert is an important member of the state of the practice assessment
-team. Pitfalls exist if non-experts attempt to acquire an authoritative list of
-software, or try to definitively rank the software. Non-experts have the problem
-that they can only rely on information available on-line, which has the
-following drawbacks:
-\begin{inparaenum}[i)]
-  \item the on-line resources could have false or inaccurate information; and,
-  \item the on-line resources could leave out relevant information that is so
-in-grained with experts that nobody thinks to explicitly record it.
-\end{inparaenum}
-
-Domain experts may be recruited from academia or industry.  The only
-requirements are knowledge of the domain and a willingness to be engaged in the
-assessment process.  The Domain Expert does not have to be a software developer,
-but they should be a user of domain software.  Given that the domain experts are
-likely to be busy people, the measurement process cannot put too much of a burden
-on their time.  For the current assessment, our Domain Expert (and paper
-co-author) is Dr.\ Michael Noseworthy, Professor of Electrical and Computer
-Engineering at McMaster University, Co-Director of the McMaster School of
-Biomedical Engineering, and Director of Medical Imaging Physics and Engineering
-at St.\ Joseph's Healthcare, Hamilton, Ontario, Canada.
-
-In advance of the first meeting with the Domain Expert, they are asked to
-create a list of top software packages in the domain.  This is done to help
-the expert get in the right mind set in advance of the meeting.  Moreover,
-by doing the exercise in advance, we avoid the potential pitfall of the expert
-approving the discovered list of software without giving it adequate thought.
-
-The Domain Experts are asked to vet the collected data and analysis.  In
-particular, they are asked to vet the proposed list of software packages and the
-AHP ranking.  These interactions can be done either electronically or with
-in-person (or virtual) meetings.
-
-\subsection{List of Representative Software} \label{sec_software_selection}
-
-We have a two-step process for selecting software packages: i) identify software
-candidates in the chosen domain; and, ii) filter the list to remove less
-relevant members \citep{SmithEtAl2021}.
-
-We initially identified 48 MI candidate software projects from the literature
-\citep{Bjorn2017, Bruhschwein2019, Haak2015}, on-line articles \citep{Emms2019,
-Hasan2020, Mu2019}, and forum discussions \citep{Samala2014}.  The full list of
-48 packages is available in \citet{Dong2021}.  To reduce the length of the list
-to a manageable number (29 in this case, as given in Section~\ref{ch_results}),
-we filtered the original list as follows:
-
-\begin{enumerate}
-
-\item We removed the packages that did not have source code available, such as
-\textit{MicroDicom}, \textit{Aliza}, and \textit{jivex}.
-
-\item We focused on the MI software that provides visualization functions, as
-described in Section~\ref{sec_scope}. Furthermore, we removed seven packages that were
-toolkits or libraries, such as \textit{VTK}, \textit{ITK}, and \textit{dcm4che}.
-We removed another three that were for PACS.
-
-\item We removed \textit{Open Dicom Viewer}, since it has not received any
-updates in a long time (since 2011).
-
-\end{enumerate}
-
-The Domain Expert provided a list of his top 12 software packages.  We compared
-his list to our list of 29.  We found 6 packages were on both lists: \textit{3D
-Slicer}, \textit{Horos}, \textit{ImageJ}, \textit{Fiji}, \textit{MRIcron} (we
-actually use the update version \textit{MRIcroGL}) and \textit{Mango} (we
-actually use the web version \textit{Papaya}).  Six software packages
-(\textit{AFNI}, \textit{FSL}, \textit{Freesurfer}, \textit{Tarquin},
-\textit{Diffusion Toolkit}, and \textit{MRItrix}) were on the Domain Expert
-list, but not on our filtered list.  However, when we examined those packages,
-we found they were out of scope, since their primary function was not
-visualization.  The Domain Expert agreed with our final choice of 29 packages.
-
-\subsection{Grading Software} \label{sec_grading_software}
-
-We grade the selected software using the measurement template summarized in
-\citet{SmithEtAl2021}.  The template provides measures of the qualities listed
-in Section~\ref{sec_software_quality}, except for reproducibility, which is
-assessed through the developer interviews (Section~\ref{sec_interview_methods}).
-For each software package, we fill in the template questions. To stay within the
-target of 160 person hours to measure the domain, we allocated between one and
-four hours for each package. Project developers can be contacted for help
-regarding installation, if necessary, but we impose a cap of about two hours on
-the installation process, to keep the overall measurement time feasible.
-Figure~\ref{fg_grading_template_example} shows an excerpt of the spreadsheet.
-The spreadsheet includes a column for each measured software package.
-
-\begin{figure}[!ht]
-\includegraphics[scale=0.66]{figures/template.pdf}
+common practice, we do not think of quality as a single measure, but rather as
+a set of measures.  That is, quality is a collection of different qualities,
+often called ``ilities.''  For this study we selected 9 qualities to measure:
+installability, correctness/ verifiability, reliability, robustness, usability,
+maintainability, reusability, understandability, and visibility/transparency.
+With the exception of installability, all the qualities are defined in Ghezzi
+et al. (2003) \cite{GhezziEtAl2003}. Installability is defined as the effort
+required for the installation and/or uninstallation of software in a specified
+environment \cite{ISO/IEC25010}.
+
+\subsubsection{Grading} \label{sec_grading_software}
+
+We use an existing template [citation redacted]
+%% DOUBLE BLIND
+%~\cite{SmithEtAl2021} 
+that is designed to measure the
+aforementioned qualities. To stay within our given measurement time frame, each
+package gets up to five hours of time.  Project developers can be contacted for
+help regarding installation, if necessary, but we impose a cap of about two
+hours on the installation process.  Figure~\ref{fg_grading_template_example}
+shows an excerpt of the measurement spreadsheet.  The rows are the measures and
+the columns correspond to the software packages.  [The full data is available on
+Mendeley; link will be provided after refereeing.] 
+% \cite{Dong2021-Data}
+%provides the full set of measurement data.   %%DOUBLE BLIND
+
+\begin{figure*}[!ht]
+\includegraphics[scale=0.53]{template.pdf}
 \caption{Grading template example}
 \label{fg_grading_template_example}
-\end{figure}
-
-The full template consists of 108 questions categorized under 9 qualities.  We
-designed the questions to be unambiguous, quantifiable, and measurable with
-limited time and domain knowledge. We group the measures under headings for each
-quality, and one for summary information. The summary information (shown in
-Figure~\ref{fg_grading_template_example}) is the first section of the template.
-This section summarizes general information, such as the software name, purpose,
-platform, programming language, publications about the software, the first
-release and the most recent change date, website, source code repository of the
-product, number of developers, etc.  We follow the definitions given by
-\citet{GewaltigAndCannon2012} for the software categories.  Public means
-software intended for public use.  Private means software aimed only at a
-specific group, while the concept category is for software written simply to
-demonstrate algorithms or concepts. The three categories of development models
-are (open source, free-ware and commercial) are discussed in
-Section~\ref{sec_software_categories}.  Information in the summary section sets
-the context for the project, but it does not directly affect the grading scores.
-
-For measuring each quality, we ask several questions and the typical answers are
-among the collection of ``yes'', ``no'', ``n/a'', ``unclear'', a number, a
-string, a date, a set of strings, etc. The grader assigns each quality an
-overall score, between 1 and 10, based on all the previous questions.  Several
-of the qualities use the word ``surface''.  This is to highlight that, for these
-qualities in particular, the best that we can do is a shallow measure.  For
-instance, we are not currently doing any experiments to measure usability.
-Instead, we are looking for an indication that the developers considered
-usability.  We do this by looking for cues in the documentation, like a getting
-started manual, a user manual and a statement of expected user characteristics.
-Below is a summary of how we assess adoption of best practices by measuring each
-quality.
-
-\begin{itemize}
-
-\item \textbf{Installability} We assess the following: 
-\begin{inparaenum}[i)]
-    \item existence and quality of installation instructions;
-    \item the quality of the user experience via the ease of following
-    instructions, number of steps, automation tools; and,
-    \item whether there is a means to verify the installation.
-\end{inparaenum}
-If any problem interrupts the process of installation or uninstallation, we give
-a lower score. We also record the Operating System (OS) used for the
-installation test.
-
-\item \textbf{Correctness \& Verifiability} We check each project to identify
-any techniques used to ensure this quality, such as literate programming,
-automated testing, symbolic execution, model checking, unit tests, etc. We also
-examine whether the projects use Continuous Integration and Continuous Delivery
-(CI/CD). For verifiability, we go through the documents of the projects to check
-for the presence of requirements specifications, theory manuals, and getting
-started tutorials. If a getting started tutorial exists and provides expected
-results, we follow it to check the correctness of the output.
-
-\item \textbf{Surface Reliability} We check the following: 
-\begin{inparaenum}[i)]
-    \item whether the software breaks during installation;
-    \item the operation of the software following the getting started tutorial
-    (if present);
-    \item whether the error messages are descriptive; and,
-    \item whether we can recover the process after an error.
-\end{inparaenum}
-
-\item \textbf{Surface Robustness} We check how the software handles
-unexpected/unanticipated input. For example, we prepare broken image files for
-MI software packages that load image files. We use a text file (.txt) with a
-modified extension name (.dcm) as an unexpected/unanticipated input. We load a
-few correct input files to ensure the function is working correctly before
-testing the unexpected/unanticipated ones.
-
-\item \textbf{Surface Usability} We examine the project's documentation,
-checking for the presence of getting started tutorials and/or a user manual. We
-also check whether users have channels to request support, such as an e-mail
-address, or issue tracker. Our impressions of usability are based on our
-interaction with the software during testing.  In general, an easy-to-use
-graphical user interface will score high.
-
-\item \textbf{Maintainability} We believe that the artifacts of a project,
-including source code, documents, and building scripts, significantly influence
-its maintainability. Thus, we check each project for the presence of such
-artifacts as API documentation, bug tracker information, release notes, test
-cases, and build scripts. We also check for the use of tools supporting issue
-tracking and version control, the percentages of closed issues, and the
-proportion of comment lines in the code.
-
-\item \textbf{Reusability} We count the total number of code files for each
-project. Projects with numerous components potentially provide more choices for
-reuse. Furthermore, well-modularized code, which tends to have smaller parts in
-separate files, is typically easier to reuse. Thus, we assume that projects with
-more code files and fewer Lines of Code (LOC) per file are more reusable. We also
-consider projects with API documentation as delivering better reusability.
-
-\item \textbf{Surface Understandability} Given that time is a constraint, we
-cannot look at all code files for each project; therefore, we randomly examine
-10 code files for their understandability. We check the code's style within each
-file, such as whether the identifiers, parameters, indentation, and formatting
-are consistent, whether the constants (other than 0 and 1) are not hardcoded, and
-whether the code is modularized. We also check the descriptive information for
-the code, such as documents mentioning the coding standard, the comments in the
-code, and the descriptions or links for details on algorithms in the code.
-
-\item \textbf{Visibility/Transparency} To measure this quality, we check the
-existing documents to find whether the software development process and
-current status of a project are visible and transparent. We examine the
-development process, current status, development environment, and release notes
-for each project.
-\end{itemize}
-
-As part of filling in the measurement template, we use freeware tools to collect
-repository related data. \href{https://github.com/tomgi/git_stats}{GitStats}
-\citep{Gieniusz2019} is used to measure the number of binary files as well as
-the number of added and deleted lines in a repository. We also use this tool to
-measure the number of commits over different intervals of time.
-\href{https://github.com/boyter/scc}{Sloc Cloc and Code (scc)}
-\citep{Boyter2021} is used to measure the number of text based files as well as
-the number of total, code, comment, and blank lines in a repository.
-
-Both tools measure the number of text-based files in a git repository and lines
-of text in these files. Based on our experience, most text-based files in a
-repository contain programming source code, and developers use them to compile
-and build software products. A minority of these files are instructions and
-other documents. So we roughly regard the lines of text in text-based files as
-lines of programming code. The two tools usually generate similar but not
-identical results. From our understanding, this minor difference is due to the
-different techniques to detect if a file is text-based or binary.
-
-For projects on GitHub we manually collect additional information, such as the
-numbers of stars, forks, people watching this repository, open pull requests,
-closed pull requests, and the number of months a repository has been on GitHub.
-We need to take care with the project creation date, since a repository can have
-a creation date much earlier than the first day on GitHub.  For example, the
-developers created the git repository for \textit{3D Slicer} in 2002, but did
-not upload a copy of it to GitHub until 2020. Some GitHub data can be found
-using its GitHub Application Program Interface (API) via the following url:
-\textit{https://api.github.com/repos/[owner]/[repository]} where [owner] and
-[repository] are replaced by the repo specific values. The number of months a
-repository has been on GitHub helps us understand the average change of metrics
-over time, like the average new stars per month.
-
-The repository measures help us in many ways. Firstly, they help us get a fast
-and accurate project overview. For example, the number of commits over the last
-12 months shows how active a project has been, and the number of stars and forks
-may reveal its popularity (used to assess \rqref{RQ_CompareHQ2Popular}).
-Secondly, the results may affect our decisions regarding the grading scores for
-some software qualities. For example, if the percentage of comment lines is low,
-we double-check the understandability of the code; if the ratio of open versus
-closed pull requests is high, we pay more attention to maintainability.
-
-As in \citet{SmithEtAl2016}, Virtual machines (VMs) were used to provide an
-optimal testing environment for each package. We used VMs because it is easier
-to start with a fresh environment, without having to worry about existing
-libraries and conflicts. Moreover, when the tests are complete the VM can be
-deleted, without any impact on the host operating system. The most significant
-advantage of using VMs is to level the playing field. Every software install
-starts from a clean slate, which removes ``works-on-my-computer'' errors. When
-filling in the measurement template, the grader notes the details for each VM,
-including hypervisor and operating system version.
-
-When grading the software, we found 27 out of the 29 packages are compatible
-with two or three different OSes, such as Windows, macOS, and Linux, and 5 of
-them are browser-based, making them platform-independent. However, in the
-interest of time, we only performed the measurements for each project by
-installing it on one of the platforms.  When it was an option, we selected
-Windows as the host OS.
-
-\subsection{Interview Methods} \label{sec_interview_methods}
-
-The repository-based measurements summarize the information we can collect from
-on-line resources. This information is incomplete because it doesn't generally
-capture the development process, the developer pain points, the perceived
-threats to software quality, and the developers' strategies to address these
-threats.  Therefore, part of our methodology involves interviewing developers.
-
-We based our interviews on a list of 20 questions, which can be found in
-\citet{SmithEtAl2021}. Some questions are about the background of the software,
-the development teams, the interviewees, and how they organize their projects.
-We also ask about the developer's understanding of the users. Some questions
-focus on the current and past difficulties, and the solutions the team has
-found, or plan to try. We also discuss documentation, both with respect to how
-it is currently done, and how it is perceived. A few questions are about
-specific software qualities, such as maintainability, understandability,
-usability, and reproducibility. The interviews are semi-structured based on the
-question list; we ask follow-up questions when necessary.  The interview process
-presented here was approved by the McMaster University Research Ethics Board
-under the application number 
-\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/MACREM/Application.pdf}
-{MREB\#: 5219}.
-
-We sent interview requests to all 29 projects using contact information from
-projects websites, code repository, publications, and from biographic pages at
-the teams' institutions.  In the end nine developers from eight of the projects
-agreed to participate: \textit{3D Slicer}, \textit{INVESALIUS 3}, \textit{dwv},
-\textit{BioImage Suite Web}, \textit{ITK-SNAP}, \textit{MRIcroGL},
-\textit{Weasis}, and \textit{OHIF}. We spent about 90 minutes for each
-interview. One participant was too busy to have an interview, so they wrote down
-their answers. In one case two developers from the same project agreed to be
-interviewed. We held the meetings on-line using either Zoom or Teams, which
-facilitated recording and automatic transcription. The full interview answers
-can be found in \citet{Dong2021}.
-
-\section{Measurement Results} \label{ch_results}
-
-Table~\ref{tab_final_list} shows the 29 software packages that we measured,
-along with summary data collected in the year 2020. We arrange the items in
-descending order of LOC. We found the initial release dates (Rlsd) for most
-projects and marked the two unknown dates with ``?''. The date of the last
-update is the date of the latest update, at the time of measurement. We found
-funding information (Fnd) for only eight projects.  For the Number Of
-Contributors (NOC) we considered anyone who made at least one accepted commit as
-a contributor. The NOC is not usually the same as the number of long-term
-project members, since many projects received change requests and code from the
-community.  With respect to the OS, 25 packages work on all three OSs: Windows
-(W), macOS (M), and Linux (L). Although the usual approach to cross-platform
-compatibility was to work natively on multiple OSes, five projects achieved
-platform-independence via web applications. The full measurement data for all
-packages is available in \citet{Dong2021-Data}.
-
-\begin{table}[!ht]
+\end{figure*}
+
+The full template consists of 108 questions over 9 qualities.  These
+questions are designed to be unambiguous, quantifiable, and measurable
+with constrained time and domain knowledge.
+
+The grader, after answering questions for each quality assigns an overall
+score (between 1 and 10) based on the answers.  Several of the qualities use
+the word ``surface'' to highlight that these particular qualities are a shallow
+measure.  For example, usability is not measured using user studies.
+Instead, we look for signs that the developers considered usability.
+We use two freeware tools to collect repository related data:
+\href{https://github.com/tomgi/git_stats}{GitStats} and
+\href{https://github.com/boyter/scc}{Sloc Cloc and Code (scc)}.  Further
+details on quality measurement are provided in [redacted for double blind].
+%% DOUBLE BLIND
+%Smith et al.\ (2024) \cite{SmithEtAl2024_MI_SOP}.
+
+\subsubsection{Analytic Hierarchy Process (AHP)} \label{sec_AHP}
+
+Developed by Saaty in the 1970s, AHP is widely used to analyze multiple criteria
+decisions~\cite{VaidyaEtAl2006}. AHP organizes multiple criteria in a
+hierarchical structure and uses pairwise comparisons between alternatives to
+calculate relative ratios~\cite{Saaty1990}. AHP works with sets of $n$
+\textit{options} and $m$ \textit{criteria}.  In our project $n=29$ and $m=9$
+since there are 29 options (software products) and 9 criteria (qualities). With
+AHP the sum of the grades (scores) for all products for a given quality will be
+1.0.  We rank the software for each of the qualities, and then we combine the
+quality rankings into an overall ranking based on the relative priorities
+between qualities.
+
+\section{Results} \label{ch_results}
+
+We initially identified 48 candidate software projects from the literature
+\cite{Bjorn2017, Bruhschwein2019, Haak2015}, on-line articles \cite{Emms2019,
+Hasan2020, Mu2019}, and forum discussions \cite{Samala2014}.  Then we filtered
+as follows:
+
+\begin{enumerate}
+
+\item Removed the packages with no source code available, such as
+\textit{MicroDicom}, \textit{Aliza}, and \textit{jivex}.
+
+\item Focused on MI software that provides visualization functions.  We removed
+seven packages that were toolkits or libraries, such as \textit{VTK},
+\textit{ITK}, and \textit{dcm4che}, and another three that were for PACS.
+
+\item Removed \textit{Open Dicom Viewer} as it has not received any
+updates since 2011.
+
+\end{enumerate}
+
+The Domain Expert provided a list 12 software packages.  We found 6 packages
+were on both lists: \textit{3D Slicer}, \textit{Horos}, \textit{ImageJ},
+\textit{Fiji}, \textit{MRIcron} (we use its descendant \textit{MRIcroGL}) and
+\textit{Mango} (we use the web version \textit{Papaya}).  The remaining six
+packages were on our out-of-scope list. The Domain Expert agreed with our final
+choice of 29 packages -- see Table~\ref{tab_final_list}. This contains summary
+data collected in the year 2020.
+
+The projects are sorted in descending order of lines of code.  We found the
+initial release dates (Rlsd) for most projects and marked the two unknown dates
+with ``?''. The date of the last update is the date of the latest update, at the
+time of measurement. We found funding information (Fnd) for only eight projects.
+For the Number Of Contributors (NOC) we considered anyone who made at least one
+accepted commit as a contributor. The NOC is not usually the same as the number
+of long-term project members, since many projects received change requests and
+code from the community.  With respect to the OS, 25 packages work on all three
+OSs: Windows (W), macOS (M), and Linux (L). Although the usual approach to
+cross-platform compatibility was to work natively on multiple OSes, five
+projects achieved platform-independence via web applications. The full
+measurement data for all packages is available on [removed for blind review]
+%\href{https://data.mendeley.com/datasets/k3pcdvdzj2/1} {Mendeley Data}.  
+%% DOUBLE BLIND
+
+\begin{table*}[!ht]
 \centering
-\begin{tabular}{p{6cm}lllllllll}
+\begin{tabular}{p{3.7cm}lllllllll}
 \toprule
-\multirow{2}{*}{Software} & \multirow{2}{*}{Rlsd} & \multirow{2}{*}{Updated} & \multirow{2}{*}{Fnd} & \multirow{2}{*}{NOC} & \multirow{2}{*}{LOC} & \multicolumn{3}{c}{OS} & \multirow{2}{*}{Web} \\ \cline{7-9}
+\multirow{2}{*}{Software} & \multirow{2}{*}{Rlsd} & \multirow{2}{*}{Updated} & \multirow{2}{*}{Fnd} & \multirow{2}{*}{NOC} & \multirow{2}{*}{LOC} & \multicolumn{3}{c}{OS} & \multirow{2}{*}{Web} \\ \cmidrule{7-9}
  &  &  &  &  &  & W & M & L &  \\ \midrule
-ParaView \citep{Ahrens2005} & 2002 & 2020-10 & \checkmark & 100 & 886326 & \checkmark & \checkmark & \checkmark & \checkmark \\
-Gwyddion \citep{Nevcas2012} & 2004 & 2020-11 &  & 38 & 643427 & \checkmark & \checkmark & \checkmark &  \\
-Horos \citep{horosproject2020} & ? & 2020-04 &  & 21 & 561617 &  & \checkmark &  &  \\
-OsiriX Lite \citep{PixmeoSARL2019} & 2004 & 2019-11 &  & 9 & 544304 &  & \checkmark &  &  \\
-3D Slicer \citep{Kikinis2014} & 1998 & 2020-08 & \checkmark & 100 & 501451 & \checkmark & \checkmark & \checkmark &  \\
-Drishti \citep{Limaye2012} & 2012 & 2020-08 &  & 1 & 268168 & \checkmark & \checkmark & \checkmark &  \\
-Ginkgo CADx \citep{Wollny2020} & 2010 & 2019-05 &  & 3 & 257144 & \checkmark & \checkmark & \checkmark &  \\
-GATE \citep{Jan2004} & 2011 & 2020-10 &  & 45 & 207122 &  & \checkmark & \checkmark &  \\
-3DimViewer \citep{TESCAN2020} & ? & 2020-03 & \checkmark & 3 & 178065 & \checkmark & \checkmark &  &  \\
-medInria \citep{Fillard2012} & 2009 & 2020-11 &  & 21 & 148924 & \checkmark & \checkmark & \checkmark &  \\
-BioImage Suite Web \citep{Papademetris2005} & 2018 & 2020-10 & \checkmark & 13 & 139699 &
+ParaView \cite{Ahrens2005} & 2002 & 2020-10 & \checkmark & 100 & 886326 & \checkmark & \checkmark & \checkmark & \checkmark \\
+Gwyddion \cite{Nevcas2012} & 2004 & 2020-11 &  & 38 & 643427 & \checkmark & \checkmark & \checkmark &  \\
+Horos \cite{horosproject2020} & ? & 2020-04 &  & 21 & 561617 &  & \checkmark &  &  \\
+OsiriX Lite \cite{PixmeoSARL2019} & 2004 & 2019-11 &  & 9 & 544304 &  & \checkmark &  &  \\
+3D Slicer \cite{Kikinis2014} & 1998 & 2020-08 & \checkmark & 100 & 501451 & \checkmark & \checkmark & \checkmark &  \\
+Drishti \cite{Limaye2012} & 2012 & 2020-08 &  & 1 & 268168 & \checkmark & \checkmark & \checkmark &  \\
+Ginkgo CADx \cite{Wollny2020} & 2010 & 2019-05 &  & 3 & 257144 & \checkmark & \checkmark & \checkmark &  \\
+GATE \cite{Jan2004} & 2011 & 2020-10 &  & 45 & 207122 &  & \checkmark & \checkmark &  \\
+3DimViewer \cite{TESCAN2020} & ? & 2020-03 & \checkmark & 3 & 178065 & \checkmark & \checkmark &  &  \\
+medInria \cite{Fillard2012} & 2009 & 2020-11 &  & 21 & 148924 & \checkmark & \checkmark & \checkmark &  \\
+BioImage Suite Web \cite{Papademetris2005} & 2018 & 2020-10 & \checkmark & 13 & 139699 &
 \checkmark & \checkmark & \checkmark & \checkmark \\
-Weasis \citep{Roduit2021} & 2010 & 2020-08 &  & 8 & 123272 & \checkmark & \checkmark & \checkmark &  \\
-AMIDE \citep{Loening2017} & 2006 & 2017-01 &  & 4 & 102827 & \checkmark & \checkmark & \checkmark &  \\
-XMedCon \citep{Nolf2003} & 2000 & 2020-08 &  & 2 & 96767 & \checkmark & \checkmark & \checkmark &  \\
-ITK-SNAP \citep{Yushkevich2006} & 2006 & 2020-06 & \checkmark & 13 & 88530 & \checkmark & \checkmark & \checkmark &  \\
-Papaya \citep{UTHSCSA2019} & 2012 & 2019-05 &  & 9 & 71831 & \checkmark & \checkmark & \checkmark &  \\
-OHIF Viewer \citep{Ziegler2020} & 2015 & 2020-10 &  & 76 & 63951 & \checkmark & \checkmark & \checkmark & \checkmark \\
-SMILI \citep{Chandra2018} & 2014 & 2020-06 &  & 9 & 62626 & \checkmark & \checkmark & \checkmark &  \\
-INVESALIUS 3 \citep{Amorim2015} & 2009 & 2020-09 &  & 10 & 48605 & \checkmark & \checkmark & \checkmark &  \\
-dwv \citep{Martelli2021} & 2012 & 2020-09 &  & 22 & 47815 & \checkmark & \checkmark & \checkmark & \checkmark \\
-DICOM Viewer \citep{Afsar2021} & 2018 & 2020-04 & \checkmark & 5 & 30761 & \checkmark & \checkmark & \checkmark &  \\
-MicroView \citep{ParallaxInnovations2020} & 2015 & 2020-08 &  & 2 & 27470 & \checkmark & \checkmark & \checkmark &  \\
-MatrixUser \citep{Liu2016} & 2013 & 2018-07 &  & 1 & 23121 & \checkmark & \checkmark & \checkmark &  \\
-Slice:Drop \citep{Haehn2013} & 2012 & 2020-04 &  & 3 & 19020 & \checkmark & \checkmark & \checkmark & \checkmark \\
-dicompyler \citep{Panchal2010} & 2009 & 2020-01 &  & 2 & 15941 & \checkmark & \checkmark &  &  \\
-Fiji \citep{Schindelin2012} & 2011 & 2020-08 & \checkmark & 55 & 10833 & \checkmark & \checkmark & \checkmark &  \\
-ImageJ \citep{Rueden2017} & 1997 & 2020-08 & \checkmark & 18 & 9681 & \checkmark & \checkmark & \checkmark &  \\
-MRIcroGL \citep{Rorden2021} & 2015 & 2020-08 &  & 2 & 8493 & \checkmark & \checkmark & \checkmark &  \\
-DicomBrowser \citep{Archie2012} & 2012 & 2020-08 &  & 3 & 5505 & \checkmark & \checkmark & \checkmark &  \\ \bottomrule
+Weasis \cite{Roduit2021} & 2010 & 2020-08 &  & 8 & 123272 & \checkmark & \checkmark & \checkmark &  \\
+AMIDE \cite{Loening2017} & 2006 & 2017-01 &  & 4 & 102827 & \checkmark & \checkmark & \checkmark &  \\
+XMedCon \cite{Nolf2003} & 2000 & 2020-08 &  & 2 & 96767 & \checkmark & \checkmark & \checkmark &  \\
+ITK-SNAP \cite{Yushkevich2006} & 2006 & 2020-06 & \checkmark & 13 & 88530 & \checkmark & \checkmark & \checkmark &  \\
+Papaya \cite{UTHSCSA2019} & 2012 & 2019-05 &  & 9 & 71831 & \checkmark & \checkmark & \checkmark &  \\
+OHIF Viewer \cite{Ziegler2020} & 2015 & 2020-10 &  & 76 & 63951 & \checkmark & \checkmark & \checkmark & \checkmark \\
+SMILI \cite{Chandra2018} & 2014 & 2020-06 &  & 9 & 62626 & \checkmark & \checkmark & \checkmark &  \\
+INVESALIUS 3 \cite{Amorim2015} & 2009 & 2020-09 &  & 10 & 48605 & \checkmark & \checkmark & \checkmark &  \\
+dwv \cite{Martelli2021} & 2012 & 2020-09 &  & 22 & 47815 & \checkmark & \checkmark & \checkmark & \checkmark \\
+DICOM Viewer \cite{Afsar2021} & 2018 & 2020-04 & \checkmark & 5 & 30761 & \checkmark & \checkmark & \checkmark &  \\
+MicroView \cite{ParallaxInnovations2020} & 2015 & 2020-08 &  & 2 & 27470 & \checkmark & \checkmark & \checkmark &  \\
+MatrixUser \cite{Liu2016} & 2013 & 2018-07 &  & 1 & 23121 & \checkmark & \checkmark & \checkmark &  \\
+Slice:Drop \cite{Haehn2013} & 2012 & 2020-04 &  & 3 & 19020 & \checkmark & \checkmark & \checkmark & \checkmark \\
+dicompyler \cite{Panchal2010} & 2009 & 2020-01 &  & 2 & 15941 & \checkmark & \checkmark &  &  \\
+Fiji \cite{Schindelin2012} & 2011 & 2020-08 & \checkmark & 55 & 10833 & \checkmark & \checkmark & \checkmark &  \\
+ImageJ \cite{Rueden2017} & 1997 & 2020-08 & \checkmark & 18 & 9681 & \checkmark & \checkmark & \checkmark &  \\
+MRIcroGL \cite{Rorden2021} & 2015 & 2020-08 &  & 2 & 8493 & \checkmark & \checkmark & \checkmark &  \\
+DicomBrowser \cite{Archie2012} & 2012 & 2020-08 &  & 3 & 5505 & \checkmark & \checkmark & \checkmark &  \\ \bottomrule
 \end{tabular}
@@
 \label{tab_final_list}
-\end{table}
-
-Figure \ref{fig_language} shows the primary languages versus the number of
-projects using them.  The primary language is the language used for the majority
-of the project's code; in most cases projects also use other languages.  The
-most popular language is \CC, with almost 40\% of projects (11 of 29).  The two
-least popular choices are Pascal and Matlab, with around 3\% of projects each
-(1 of 29).
-
-\begin{figure}[!ht]
-\centering
-\includegraphics[scale=0.5]{figures/PrimaryLanguages.pdf}
-\centering
-\caption{\label{fig_language}Primary languages versus number of projects}
-\end{figure}
+\end{table*}
+
+The programming languages used in order of decreasing popularity are \CC,
+JavaScript, Java, C, Python, Pascal, Matlab.  The most popular language is \CC,
+for 11 of 29 projects; Pascal and Matlab were each used for a single project.
+
+In the absence of
+a specific real world context, we assumed all nine qualities are equally
+important. Figure~\ref{fg_overall_scores} shows the overall scores in descending
+order.
+
+\begin{figure*}[!ht]
+\includegraphics[scale=0.38]{overall_scores.pdf}
+\caption{Overall AHP scores with an equal weighting for all 9 software qualities}
+
+\label{fg_overall_scores}
+\end{figure*}
+
+The top four software products \textit{3D Slicer}, \textit{ImageJ},
+\textit{Fiji}, and \textit{OHIF Viewer} have higher scores in most criteria.
+\textit{3D Slicer} has a score in the top two for all qualities; \textit{ImageJ}
+ranks near the top for all qualities, except for correctness \& verifiability.
+\textit{OHIF Viewer} and \textit{Fiji} have similar overall scores, with
+\textit{Fiji} doing better in installability and \textit{OHIF Viewer} doing
+better in correctness \& verifiability.  Given the installation problems, we may
+have underestimated the scores on reliability and robustness for \textit{DICOM
+Viewer}, but we compared it equally for the other seven qualities.
+
+[Summary description list]
 
@@
 
-Figure \ref{fg_installability_scores} lists the installability scores.  We found
-installation instructions for 16 projects. Among the ones without instructions,
-\textit{BioImage Suite Web} and \textit{Slice:Drop} do not need installation,
-since they are web applications. Installing 10 of the projects required extra
-dependencies. Five of them are web applications (as shown in
-Table~\ref{tab_final_list}) and depend on a browser; \textit{dwv}, \textit{OHIF
-Viewer}, and \textit{GATE} needs extra dependencies to build; \textit{ImageJ}
-and	\textit{Fiji} need an unzip tool; \textit{MatrixUser} is based on Matlab;
-\textit{DICOM Viewer} needs to work on a Nextcloud platform.
-
-\begin{figure}[!ht]
-\includegraphics[scale=0.47]{figures/installability_scores.pdf}
-\caption{AHP installability scores}
-\label{fg_installability_scores}
-\end{figure}
-
-\textit{3D Slicer} has the highest score because it had easy to follow
-installation instructions, and an automated, fast, frustration-free installation
-process. The installer added all dependencies automatically and no errors
-occurred during the installation and uninstallation steps. Many other software
-packages also had installation instructions and automated installers.  We had no
-trouble installing the following packages: \textit{INVESALIUS 3},
-\textit{Gwyddion}, \textit{XMedCon}, and \textit{MicroView}. We determined their
-scores based on the understandability of the instructions, installation steps,
-and user experience. Since \textit{BioImage Suite Web} and \textit{Slice:Drop}
-needed no installation, we gave them high scores. \textit{BioImage Suite Web}
-also provided an option to download cache for offline usage, which was easy to
-apply.
-
+We found
+installation instructions for 16 projects, but two did not need them
+(\textit{BioImage Suite Web} and \textit{Slice:Drop})
+as they are web applications. 10 of the projects required extra
+dependencies: Five depend on a specific browser; \textit{dwv}, \textit{OHIF
+Viewer}, and \textit{GATE} needs extra libraries to build; \textit{ImageJ} and
+\textit{Fiji} need an unzip tool; \textit{MatrixUser} needs Matlab;
+\textit{DICOM Viewer} needs a Nextcloud platform.
+
+The scores are based on the ease of following the installation instructions,
+and automated installation and uninstallation process. There were no issues
+for all but the bottom four, just various degrees of ease and automation.
 \textit{GATE}, \textit{dwv}, and \textit{DICOM Viewer} showed severe
@@
 from the source code, but we were able to proceed with measuring other qualities
-using a deployed on-line version for \textit{dwv}, and a VM version for
+using a deployed on-line version for \textit{dwv}, and a virtual machine version for
 \textit{GATE}. For \textit{DICOM Viewer} we could not install the NextCloud
-dependency, and we did not have another option for running the software.
-Therefore, for \textit{DICOM Viewer} we could not measure reliability or
-robustness.  The other seven qualities could be measured, since they do not
-require installation.
-
-\textit{MatrixUser} has a lower score because it depends on Matlab. We assessed
-the score from the point of view of a user that would have to install Matlab and
-acquire a license.  Of course, for users that already work within Matlab, the
-installability score should be higher.
+dependency, and thus could not measure reliability nor robustness for it.
+
+\textit{MatrixUser} depends on Matlab, whose installation is not easy.
+For users who already have Matlab, this score should be higher.
 
@@
 
-Figure~\ref{fg_correctness_verifiability_scores} shows the scores of correctness
-and verifiability. Generally speaking, the packages with higher scores adopted
-more techniques to improve correctness, and had better documentation for us to
-verify against.  For instance, we looked for evidence of unit testing, since it
-benefits most parts of the software's life cycle, such as designing, coding,
-debugging, and optimization \citep{Hamill2004}.  We only found evidence of unit
-testing for about half of the projects. We identified five projects using CI/CD
-tools: \textit{3D Slicer}, \textit{ImageJ}, \textit{Fiji}, \textit{dwv}, and
-\textit{OHIF Viewer}.
-
-\begin{figure}[!ht]
-\includegraphics[scale=0.47]{figures/correctness_verifiability_scores.pdf}
-\caption{AHP correctness \& verifiability scores}
-\label{fg_correctness_verifiability_scores}
-\end{figure}
-
-Even for some projects with well-organized documentation, requirements
-specifications and theory manuals were still missing.  We could not identify
-theory manuals for all projects, and we did not find requirements specifications
-for most projects. The only requirements-related document we found was a road
+The packages with higher scores for correctness and verifiability used a wider array of
+techniques to improve correctness, and had better documentation to witness this.
+For instance, we looked for evidence of unit testing, and found evidence for
+only about half of the projects. We identified five projects using continuous
+integration tools: \textit{3D Slicer}, \textit{ImageJ}, \textit{Fiji},
+\textit{dwv}, and \textit{OHIF Viewer}.
+
+Even for projects with well-organized documentation, requirements
+specifications and theory manuals were still missing.
+The only requirements-related document we found was a road
 map of \textit{3D Slicer}, which contained design requirements for upcoming
@@
 
-Figure~\ref{fg_reliability_scores} shows the AHP results.  As shown in
-Section~\ref{sec_result_installability}, most of the software products did not
-``break'' during installation, or did not need installation; \textit{dwv} and
-\textit{GATE} broke in the building stage, and the processes were not
-recoverable; we could not install the dependency for \textit{DICOM Viewer}. Of
-the seven software packages with a getting started tutorial and operation steps
-in the tutorial, most showed no error when we followed the steps. However,
-\textit{GATE} could not open macro files and became unresponsive several times,
-without any descriptive error message. When assessing robustness
-(Section~\ref{sec_result_robustness}), we found that \textit{Drishti} crashed
-when loading damaged image files, without showing any descriptive error message.
-We did not find any problems with the on-line version of
-\textit{dwv}.
-
-\begin{figure}[!ht]
-\includegraphics[scale=0.47]{figures/reliability_scores.pdf}
-\caption{AHP surface reliability scores}
-\label{fg_reliability_scores}
-\end{figure}
+We were able to follow the steps in the tutorials that existed (seven packages
+had them.) However, \textit{GATE} could not open macro files and became
+unresponsive several times, without any descriptive error message. We found that
+\textit{Drishti} crashed when loading damaged image files, without showing any
+descriptive error message.
 
@@
 
-Figure \ref{fg_robustness_scores} presents the scores for surface robustness.
-The packages with higher scores elegantly handled unexpected/unanticipated
-inputs, typically showing a clear error message. We may have underestimated the
-score of \textit{OHIF Viewer}, since we needed further customization to load
+The packages with higher robustness scores gracefully handled unexpected or unanticipated
+inputs, typically showing a clear error message. We may have underscored
+\textit{OHIF Viewer}, since we needed further customization to load
 data.
 
-\begin{figure}[!ht]
-\includegraphics[scale=0.47]{figures/robustness_scores.pdf}
-\caption{AHP surface robustness scores}
-\label{fg_robustness_scores}
-\end{figure}
-
-Digital Imaging and Communications in Medicine (DICOM) ``defines the formats for
-medical images that can be exchanged with the data and quality necessary for
-clinical use'' \citep{MITA2021}. According to their documentation, all 29
-software packages should support the DICOM standard. To test robustness, we
-prepared two types of image files: correct and incorrect formats (with the
-incorrect format created by relabelled a text file to have the ``.dcm''
-extension).  All software packages loaded the correct format image, except for
-\textit{GATE}, which failed for unknown reasons.  For the broken format,
-\textit{MatrixUser}, \textit{dwv}, and \textit{Slice:Drop} ignored the incorrect
-format of the file and loaded it regardless. They did not show any error message
-and displayed a blank image. \textit{MRIcroGL} behaved similarly except that it
-showed a meaningless image. \textit{Drishti} successfully detected the broken
-format of the file, but the software crashed as a result.
+According to their documentation, all 29 software packages should support the
+DICOM standard. To test robustness, we prepared two types of image files:
+correct and incorrect formats (with the incorrect format created by relabelling a
+text file to have the ``.dcm'' extension).  All software packages loaded the
+correct format image, except for \textit{GATE}, which failed for unknown
+reasons.  For the broken format, \textit{MatrixUser}, \textit{dwv}, and
+\textit{Slice:Drop} ignored the incorrect format,
+did not show any error message and displayed a blank image.
+\textit{MRIcroGL} behaved similarly except that it showed a meaningless image.
+\textit{Drishti} successfully detected the broken format of the file, but the
+software crashed as a result.
 
@@
 
-Figure~\ref{fg_usability_scores} shows the AHP scores for surface usability. The
-software with higher scores usually provided both comprehensive documented
-guidance and a good user experience. \textit{INVESALIUS 3} provided an excellent
-example of a detailed and precise user manual. \textit{GATE} also provided
-numerous documents, but unfortunately we had difficulty understanding and using
-them. We found getting started tutorials for only 11 projects, but a user manual
-for 22 projects. \textit{MRIcroGL} was the only project that explicitly
-documented expected user characteristics.
-
-\begin{figure}[!ht]
-\includegraphics[scale=0.47]{figures/usability_scores.pdf}
-\caption{AHP surface usability scores}
-\label{fg_usability_scores}
-\end{figure}
+The software with higher usability scores usually provided both comprehensive
+documented guidance and a good user experience. \textit{INVESALIUS 3} provided
+an excellent example of a detailed and precise user manual. \textit{GATE} also
+provided numerous documents, but unfortunately we had difficulty understanding
+and using them. We found getting started tutorials for only 11 projects, but a
+user manual for 22 projects. \textit{MRIcroGL} was the only project that
+explicitly documented expected user characteristics.
 
@@
 
-Figure~\ref{fg_maintainability_scores} shows the ranking results for
-maintainability. We gave \textit{3D Slicer} the highest score because we found
-it to have the most comprehensive artifacts. For example, as far as we could
-find, only a few of the 29 projects had a product, developer's manual, or API
+We gave \textit{3D Slicer} the highest score for maintainability because we
+found it had the most comprehensive artifacts. Only a few of the 29 projects had
+a product, developer's manual, or API (Application Programming Interface)
 documentation, and only \textit{3D Slicer}, \textit{ImageJ}, \textit{Fiji}
-included all three documents. Moreover, \textit{3D Slicer} has a much higher
-percentage of closed issues (91.65\%) compared to \textit{ImageJ} (52.49\%) and
-\textit{Fiji} (63.79\%). Table~\ref{tab_maintainability_docs} shows which
-projects had these documents, in descending order of their maintainability
-scores.
-
-\begin{figure}[!ht]
-\includegraphics[scale=0.47]{figures/maintainability_scores.pdf}
-\caption{AHP maintainability scores}
-\label{fg_maintainability_scores}
-\end{figure}
-
-\begin{table}[!ht]
+included all three documents (see Table~\ref{tab_maintainability_docs} for the
+full data).  Moreover, \textit{3D Slicer} has a much higher percentage of closed
+issues (92\%) compared to \textit{ImageJ} (52\%) and \textit{Fiji} (64\%).
+
+\begin{table*}[ht]
 \centering
@@
 \label{tab_maintainability_docs}
-\end{table}
-
-Twenty-seven of the 29 projects used git as the version control tool, with 24 of these
-using GitHub. \textit{AMIDE} used Mercurial and \textit{Gwyddion} used
+\end{table*}
+
+Twenty-seven of the 29 projects used git for version control, with 24 of
+these using GitHub. \textit{AMIDE} used Mercurial and \textit{Gwyddion} used
 Subversion. \textit{XMedCon}, \textit{AMIDE}, and \textit{Gwyddion} used
@@
 
-Figure~\ref{fg_reusability_scores} shows the AHP results for reusability. As
-described in Section~\ref{sec_grading_software}, we gave higher scores to the
-projects with API documentation. As shown in
-Table~\ref{tab_maintainability_docs}, seven projects had API documents. We also
-assumed that projects with more code files and less LOC per code file are more
-reusable. Table \ref{tab_loc_per_file} shows the number of text-based files by
-project, which we used to approximate the number of code files. The table also
-lists the total number of lines (including comments and blanks), LOC, and
-average LOC per file. We arranged the items in descending order of their
-reusability scores.
-
-\begin{figure}[!ht]
-\includegraphics[scale=0.47]{figures/reusability_scores.pdf}
-\caption{AHP reusability scores}
-\label{fg_reusability_scores}
-\end{figure}
-
-\begin{table}[!ht]
+We have assumed that smaller code files are
+likely more reusable -- see Table \ref{tab_loc_per_file} for the details.
+
+\begin{table*}[ht]
 \centering
@@
 \end{tabular}
-\caption{Number of files and lines (sorted in descending order of reusability
-scores)}
+\caption{Number of files and lines (by reusability scores)}
 \label{tab_loc_per_file}
-\end{table}
+\end{table*}
 
@@
 
-Figure~\ref{fg_surface_understandability_scores} shows the scores for surface
-understandability. All projects had a consistent coding style with parameters in
-the same order for all functions; modularized code; and, clear comments, indicating
-what is done, not how. However, we only found explicit identification of a
-coding standard for 3 out of the 29: \textit{3D Slicer}, \textit{Weasis}, and
-\textit{ImageJ}. We also found hard-coded constants (rather than symbolic
+All projects had a consistent coding style with parameters in
+the same order for all functions, modularized code, and, clear comments that
+indicate what is done, not how. However, we only found explicit identification
+of a coding standard for 3 out of the 29: \textit{3D Slicer}, \textit{Weasis},
+and \textit{ImageJ}. We also found hard-coded constants (rather than symbolic
 constants) in \textit{medInria}, \textit{dicompyler}, \textit{MicroView}, and
@@
 
-\begin{figure}[!ht]
-\includegraphics[scale=0.47]{figures/understandability_scores.pdf}
-\caption{AHP surface understandability scores}
-\label{fg_surface_understandability_scores}
-\end{figure}
-
 \subsection{Visibility/Transparency} \label{sec_result_visibility_transparency}
 
-Figure~\ref{fg_visibility_transparency_scores} shows the AHP scores for
-visibility/transparency. Generally speaking, the teams that actively documented
+Generally speaking, the teams that actively documented
 their development process and plans scored higher.
@@
 documents for the development process, project status, development environment,
-and release notes, in descending order of their visibility/transparency
-scores.
-
-\begin{figure}[!ht]
-\includegraphics[scale=0.47]{figures/visibility_transparency_scores.pdf}
-\caption{AHP visibility/transparency scores}
-\label{fg_visibility_transparency_scores}
-\end{figure}
-
-\begin{table}[!ht]
+and release notes.
+
+\begin{table*}[!ht]
 \centering
@@
 \label{tab_Visibility/Transparency_docs}
-\end{table}
-
-\subsection{Overall Scores} \label{Sec_OverallQ}
-
-As described in Section~\ref{sec_AHP}, for our AHP measurements, we have nine
-criteria (qualities) and 29 alternatives (software packages). In the absence of
-a specific real world context, we assumed all nine qualities are equally
-important. Figure~\ref{fg_overall_scores} shows the overall scores in descending
-order. Since we produced the scores from the AHP process, the total sum of the
-29 scores is precisely 1.0.
-
-\begin{figure}[!ht]
-\includegraphics[scale=0.47]{figures/overall_scores.pdf}
-\caption{Overall AHP scores with an equal weighting for all 9 software qualities}
-
-\label{fg_overall_scores}
-\end{figure}
-
-The top four software products \textit{3D Slicer}, \textit{ImageJ},
-\textit{Fiji}, and \textit{OHIF Viewer} have higher scores in most criteria.
-\textit{3D Slicer} has a score in the top two for all qualities; \textit{ImageJ}
-ranks near the top for all qualities, except for correctness \& verifiability.
-\textit{OHIF Viewer} and \textit{Fiji} have similar overall scores, with
-\textit{Fiji} doing better in installability and \textit{OHIF Viewer} doing
-better in correctness \& verifiability.  Given the installation problems, we may
-have underestimated the scores on reliability and robustness for \textit{DICOM
-Viewer}, but we compared it equally for the other seven qualities.
-
-\section{Comparison to Community Ranking} \label{Sec_VsCommunityRanking}
-
-To address~\rqref{RQ_CompareHQ2Popular} about how our ranking compares to the
-popularity of projects as judged by the scientific community, we make two
-comparisons:
-\begin{itemize}
-\item A comparison of our ranking (from Section~\ref{ch_results}) with the
-community ratings on GitHub, as shown by GitHub stars, number of forks, and
-number of people watching the projects; and,
-\item A comparison of top-rated software from our methodology with the top
-recommendations from our domain experts (as mentioned in
-Section~\ref{sec_software_selection}).
-\end{itemize}
-
-Table~\ref{tab_ranking_vs_GitHub} shows our ranking of the 29 MI projects, and
-their GitHub metrics, if applicable. As mentioned in
-Section~\ref{sec_score_maintainability}, 24 projects used GitHub. Since GitHub
-repositories have different creation dates, we collect the number of months each
-stayed on GitHub, and calculate the average number of new stars, people
-watching, and forks per 12 months. Section~\ref{sec_grading_software} describes
-the method of getting the creation date.  The items in
-Table~\ref{tab_ranking_vs_GitHub} are listed in descending order of the average
-number of new stars per year.  The non-GitHub items are listed in the order of
-our ranking.  We collected all GitHub statistics in July 2021.
-
-Generally speaking, most of the top-ranking MI software projects also received
-greater attention and popularity on GitHub. Between our ranking and the GitHub
-stars-per-year ranking, four of the top five software projects appear in both
-lists. Our top five packages are scattered among the first eight positions on the
-GitHub list. However, as discussed below there are discrepancies between the two
-lists.
-
-In some cases projects are popular in the community, but were assigned a low
-rank by our methodology.  This is the case for \textit{dwv}. The reason for the
-low ranking is that, as mentioned in Section~\ref{sec_result_installability}, we
-failed to build it locally, and used the test version on its websites for the
-measurements. We followed the instructions and tried to run the command ``yarn
-run test'' locally, which did not work. In addition, the test version did not
-detect a broken DICOM file and displayed a blank image as described in
-Section~\ref{sec_result_robustness}. We might underestimate the scores for
-\textit{dwv} due to uncommon technical issues.
-
-We also ranked \textit{DICOM Viewer} much lower than its popularity. As
-mentioned in Section~\ref{sec_result_installability}, it depended on the
-NextCloud platform that we could not successfully install. Thus, we might
-underestimate the scores of its surface reliability and surface robustness.
-
-Further reason for discrepancies between our ranking and the community's ranking
-is that we weighted all qualities equally. This is not likely how users
-implicitly rank the different qualities. As a result, some projects with high
-community popularity may have scored lower with our method because of a
-relatively higher (compared to the scientific community's implicit ranking)
-weighting of the poor scores for some qualities. A further explanation for
-discrepancies between our measures and the star measures may also be due to
-inaccuracy with using stars to approximate popularity.  Stars are not an ideal
-measure because stars represent the community's feeling in the past more than
-they measure current preferences \citep{Szulik2017}.  The issue with stars is
-that they tend only to be added, not removed.  A final reason for
-inconsistencies between our ranking and the community's ranking is that, as for
-consumer products, more factors influence popularity than just quality.
+\end{table*}
+
+\section{Discussion}
+
+We first compare our ranking to a (proxy for) the community's ranking. We then
+compare the state of the practice for MI with that of other research software.
+In particular we provide details on recommended artifacts that are rarely observed
+for MI software. [other new sections] Section~\ref{sec_threats_to_validity} presents threats to the
+validity of our data and conclusions.
+
+\subsection{Comparison to Community Ranking} \label{Sec_VsCommunityRanking}
+
+We use GitHub stars, number of forks and number of people watching the projects
+are proxies for community ranking -- see Table~\ref{tab_ranking_vs_GitHub}
+for statistics collected in July 2021.  Recall that 24 projects use GitHub.
+Our ranking and GitHub popularity, at least for the top five projects,
+seems to line up fairly well.
+
+We ranked some popular packages fairly low, such as \textit{dwv}. This is
+because we were unable to build it locally, even though we followed its
+installation instructions. However, we were able to use its web
+version for the rest of the measurements. Additionally, this version 
+did not detect a broken DICOM file and instead displayed a blank image
+(Section~\ref{sec_result_robustness}).
+\textit{DICOM Viewer} ranked low as we were unable to install the
+NextCloud platform.
+
+Another likely reason for discrepancies is that we weighted all qualities
+equally. This is not likely how users implicitly rank the different qualities.
+This would require a broader user study to properly assess. Furthermore our
+measures of popularity are only \emph{proxies} which are biased towards
+past rather than current preferences~\cite{Szulik2017}, as these are
+monotonically increasing quantities. Finally there are often more factors than
+just quality that influence the popularity of ``consumer'' products.
 
@@
 \renewcommand{\arraystretch}{0.85}
-\begin{table}[!ht]
+\begin{table*}[!ht]
 \centering
@@
 \label{tab_ranking_vs_GitHub}
-\end{table}
+\end{table*}
 \endgroup
 
-As shown in Section~\ref{sec_software_selection}, our domain experts recommended
-a list of top software with 12 software products.  All the top 4 entries from
-the Domain Expert's list are among the top 12 ranked by our methodology. Three
-of the top four on both lists are the same: \textit{3D Slicer}, \textit{ImageJ},
-and \textit{Fiji}. \textit{3D Slicer} is top project by both rankings (and by
-the GitHub stars measure as well).  The Domain Expert ranked \textit{Horos} as
-their second choice, while we ranked it twelfth.  Our third ranked project,
-\textit{OHIF Viewer} was not listed by the Domain Expert.  Neither were the
-software packages that we ranked from fifth to eleventh (\textit{ParaView},
-\textit{Weasis}, \textit{medInria}, \textit{BioImage Suite Web}, \textit{OsiriX
-Lite}, \textit{INVESALIUS}, and \textit{Gwyddion}).  The software mentioned by
-the Domain Expert that we did not rank were the six recommended packages that
-did not have visualization as the primary function (as discussed in
-Section~\ref{sec_software_selection}).  The differences between the list
-recommended by our methodology and the Domain Expert are not surprising.  As
-mentioned above, our methodology weights all qualities equally, but that may not
-be the case for the Domain Expert's impressions.  Moreover, although the Domain
-Expert has significant experience with MI software, they have not used all 29
-packages that were measured.
-
-Although our ranking and the estimate of the community's ranking are not perfect
-measures, they do suggest a correlation between best practices and popularity.
-We do not know which comes first, the use of best practices or popularity, but
-we do know that the top ranked packages tend to incorporate best practices. The
-next sections will explore how the practices of the MI community compare to the
-broader research software community. We will also investigate the practices from
-the top projects that others within the MI community, and within the broader
-research software community, can potentially adopt.
-
-\section{Comparison Between MI and Research Software for Artifacts}
+Although both rankings are imperfect measures, they nevertheless suggest a
+correlation between best practices and popularity. We don't know if this is
+causal, in either direction (i.e. if best practices enable popularity or if
+popularity increases the need for using more software development best
+practices).
+
+\subsection{Software Artifacts}
 \label{Sec_CompareArtifacts}
 
-As part of filling in the measurement template (from
-Section~\ref{sec_grading_software}), we summarized the artifacts observed in
-each MI package. Table~\ref{artifactspresent} groups the artifacts by frequency
-into categories of common (20 to 29 ($>$67\%) packages), uncommon (10 to 19
-(33-67\%) packages), and rare (1 to 9 ($<$33\%) packages). \citet{Dong2021-Data}
-summarizes the full measurements.  Tables~\ref{tab_maintainability_docs}
-and~\ref{tab_Visibility/Transparency_docs} show the details on which projects
-use which types of artifacts for documents related to maintainability and
-visibility, respectively.
-
-\begin{table}[ht!]
-    \begin{center}
-    \begin{tabular}{ p{4.6 cm} p{5.6 cm} p{5 cm}}
-    \toprule
-    Common & Uncommon & Rare \\
-    \midrule
-    README (29) & Build scripts (18) & Getting Started (9)\\
-    Version control (29) & Tutorials (18) & Developer's manual (8)\\
-    License (28) & Installation guide (16) & Contributing (8)\\
-    Issue tracker (28) & Test cases (15) & API documentation (7)\\
-    User manual (22) & Authors (14) & Dependency list (7)\\
-    Release info. (22) & Frequently Asked Questions (FAQ) (14) & Troubleshooting guide (6)\\
-     & Acknowledgements (12) & Product roadmap (5)\\
-     & Changelog (12) & Design documentation (5)\\
-     & Citation (11) & Code style guide (3)\\
-     & & Code of conduct (1)\\
-     & & Requirements (1)\\
-    \bottomrule
-    \end{tabular}
-    \caption{Artifacts Present in MI Packages, Classified by Frequency (The number 
-    in brackets is the number of occurrences)}
-    \label{artifactspresent}
-    \end{center}
-\end{table}
-
-We answer~\rqref{RQ_CompareArtifacts} by comparing the artifacts that we
-observed in MI repositories to those observed and recommended for research
-software in general. Our comparison may point out areas where some MI software
-packages fall short of current best practices. This is not intended to be a
-criticism of any existing packages, especially since in practice not every
-project needs to achieve the highest possible quality. However, rather than
-delve into the nuances of which software can justify compromising which
-practices we will write our comparison under the ideal assumption that every
-project has sufficient resources to match best practices.
-
-Table~\ref{Tbl_Guidelines} (based on data from \citep{SmithAndMichalski2022})
-shows that MI artifacts generally match the recommendations found in nine
-current research software development guidelines:
+We use nine research software development guidelines to compare recommended 
+software artifacts versus those present in MI software.
+These guidelines are:
 \begin{itemize}
 \item United States Geological Survey Software Planning Checklist
-\citep{USGS2019},
+\cite{USGS2019},
 \item DLR (German Aerospace Centre) Software Engineering Guidelines
-\citep{TobiasEtAl2018}, 
+\cite{TobiasEtAl2018}, 
 \item Scottish Covid-19 Response Consortium Software Checklist
-\citep{BrettEtAl2021},
-\item Good Enough Practices in Scientific Computing \citep{WilsonEtAl2016},
+\cite{BrettEtAl2021},
+\item Good Enough Practices in Scientific Computing \cite{WilsonEtAl2016},
 \item xSDK (Extreme-scale Scientific Software Development Kit) Community Package
-Policies \citep{SmithAndRoscoe2018},
-\item Trilinos Developers Guide \citep{HerouxEtAl2008},
+Policies \cite{SmithAndRoscoe2018},
+\item Trilinos Developers Guide \cite{HerouxEtAl2008},
 \item EURISE (European Research Infrastructure Software Engineers') Network
-Technical Reference \citep{ThielEtAl2020},
+Technical Reference \cite{ThielEtAl2020},
 \item CLARIAH (Common Lab Research Infrastructure for the Arts and Humanities)
-Guidelines for Software Quality \citep{vanGompelEtAl2016}, and
+Guidelines for Software Quality \cite{vanGompelEtAl2016}, and
 \item A Set of Common Software Quality Assurance Baseline Criteria for Research
-Projects \citep{OrvizEtAl2017}.
+Projects \cite{OrvizEtAl2017}.
 \end{itemize}
 
-In Table~\ref{Tbl_Guidelines} each row corresponds to an artifact.  For a given
-row, a checkmark in one of the columns means that the corresponding guideline
-recommends this artifact.  The last column shows whether the artifact appears in
-the measured set of MI software, either not at all (blank), commonly (C),
-uncommonly (U) or rarely (R).  We did our best to interpret the meaning of each
-artifact consistently between guidelines and specific MI software, but the
-terminology and the contents of artifacts are not standardized.  The challenge
-even exists for the ubiquitous README file.  As illustrated by
-\citet{PranaEtAl2018}, the content of README files shows significant variation
-between projects.  Although some content is reasonably consistent, with 97\% of
-README files contain at least one section describing the `What' of the
-repository and 89\% offering some `How' content, other categories are more
-variable.  For instance, information on `Contribution', `Why', and `Who',
-appear in 28\%, 26\% and 53\% of the analyzed files, respectively
-\citep{PranaEtAl2018}.
-
-The frequency of checkmarks in Table~\ref{Tbl_Guidelines} indicates the
-popularity of recommending a given artifact, but it does not imply that the most
-commonly recommended artifacts are the most important artifacts. Just because a
-guideline does not explicitly recommend an artifact, does not mean the guideline
-authors do not value that artifact.  They may have excluded it because it is out
-of the scope of their recommendations, or outside their experience.  For
-instance, an artifact related to uninstall is only explicitly mentioned by
-\citet{vanGompelEtAl2016}, but other guideline authors would likely see its
-value.  They may simply feel that uninstall is implied by install, or they may
-have never asked themselves whether they need separate uninstall instructions.
-
-\begin{table}[!ht]
-\begin{center}
-\begin{tabular}{ p{2.5cm}p{1cm}p{1cm}p{1cm}p{1cm}p{1cm}p{1cm}p{1cm}p{1.2cm}p{1cm}p{0.8cm} }
+\begin{table*}[ht]
+%\begin{center}
+\begin{tabular}{ p{2.5cm}p{0.5cm}p{0.5cm}p{0.5cm}p{0.5cm}p{0.5cm}p{0.5cm}p{0.5cm}p{0.5cm}p{0.5cm}p{0.5cm} }
 \toprule
-~ \ & \citet{USGS2019} & \citet{TobiasEtAl2018} & \citet{BrettEtAl2021} &
-\citet{WilsonEtAl2016} & \citet{SmithAndRoscoe2018} & \citet{HerouxEtAl2008} &
-\citet{ThielEtAl2020} & \citet{vanGompelEtAl2016} & \citet{OrvizEtAl2017} & MI\\
+~ \ & \cite{USGS2019} & \cite{TobiasEtAl2018} & \cite{BrettEtAl2021} &
+\cite{WilsonEtAl2016} & \cite{SmithAndRoscoe2018} & \cite{HerouxEtAl2008} &
+\cite{ThielEtAl2020} & \cite{vanGompelEtAl2016} & \cite{OrvizEtAl2017} & MI\\
 \midrule
@@
 \label{Tbl_Guidelines}
-\end{center}
-\end{table}
-
-Two of the items that appear in Table~\ref{artifactspresent} do not appear in
-the software development guidelines shown in Table~\ref{Tbl_Guidelines}:
-Troubleshooting guide and Developer's manual.  Although the guidelines do not
-name these two artifacts, the information contained within them overlaps with
-the recommended artifacts.  A Troubleshooting guideline contains information
-that would typically be found in a User manual.  A Developer's guide overlaps
-with information from the README, INSTALL, Uninstall, Dependency List, Release
-Information, API documentation and Design documentation.  In our current
-analysis, we have identified artifacts by the names given by the software
-guidelines and MI examples.  In the future, a more in-depth analysis would look
-at the knowledge fragments that are captured in the artifacts, rather than
-focusing on the names of the files that collect these fragments together.
-
-Although the MI community shows examples of 88\% (23 of 26) of the practices we
-found in research software guidelines (Table~\ref{Tbl_Guidelines}), we did not
-observe three recommended artifacts: i) Uninstall, ii) Test plans, and iii)
-Requirements.  Uninstall is likely an omission caused by the focus on installing
-software. Given the storage capacity of current hardware, developers and users
-are not generally concerned with uninstall.  Moreover, as mentioned above,
-uninstall is not particularly emphasized in existing recommendations.
-
-Test plans describe the scope, approach, resources, and the schedule of planned
-test activities \citep{VanVliet2000}.  The plan should cover details such as the
-makeup of the testing team, automated testing tools and technology to employ,
-the testing process, system tests, integration tests and unit tests. We did not
-observe test plans for MI software, but that doesn't mean plans weren't created;
-it means that the plans are not under version control. Test plans would have to
-at least be implicitly created, since we observed test cases with reasonable
-frequency for MI software (test cases are categorized as uncommon).
-
-The other apparently neglected document is the requirements specification, which
-records the functionalities, expected performance, goals, context, design
-constraints, external interfaces and other quality attributes of the software
-\citep{IEEE1998}.  If developers write requirements they typically are based on
-a template, which provide documentation structure, guidelines, and rules.
-Although there is no universally accepted template, examples include
-\citet{ESA1991}, \citet{IEEE1998}, \citet{NASA1989}, and
-\citet{RobertsonAndRobertson1999Vol}.
-
-MI software is like other research software in its neglect of requirements
-documentation.  Although requirements documentation is recommended by some
-\citep{TobiasEtAl2018, HerouxEtAl2008, SmithAndKoothoor2016}, in practice
-research software developers often do not produce a proper requirements
-specification \citep{HeatonAndCarver2015}. \citet{SandersAndKelly2008}
-interviewed 16 scientists from 10 disciplines and found that none of the
-scientists created requirements specifications, unless regulations in their
-field mandated such a document. \citet{Nguyen-HoanEtAl2010} showed requirements
-are the least commonly produced type of documentation for research software in
-general. When looking at the pain points for research software developers,
-\citet{WieseEtAl2019} found that software requirements and management is the
-software engineering discipline that most hurts scientific developers,
+%\end{center}
+\end{table*}
+
+In Table~\ref{Tbl_Guidelines} each row corresponds to an artifact.  For a given
+row, a checkmark in one of the columns means that the corresponding guideline
+recommends this artifact.  The last column shows whether the artifact appears in
+the measured set of MI software, either not at all (blank), commonly (C),
+uncommonly (U) or rarely (R).  We did our best to interpret the meaning of each
+artifact consistently between guidelines and specific MI software, but the
+terminology and the contents of artifacts are not standardized.  The challenge
+even exists for the ubiquitous README file.  The content of README files shows
+significant variation between projects \cite{PranaEtAl2018}.  Although some
+content is reasonably consistent, with 97\% of README files contain at least one
+section describing the `What' of the repository and 89\% offering some `How'
+content, other categories are more variable.  For instance, information on
+`Contribution', `Why', and `Who', appear in 28\%, 26\% and 53\% of the analyzed
+files, respectively \cite{PranaEtAl2018}.
+
+\begin{table*}[ht!]
+    %\begin{center}
+    \begin{tabular}{ p{3.1 cm} p{5.4 cm} p{3.5 cm}}
+    \toprule
+    Common & Uncommon & Rare \\
+    \midrule
+    README (29) & Build scripts (18) & Getting Started (9)\\
+    Version control (29) & Tutorials (18) & Developer's manual (8)\\
+    License (28) & Installation guide (16) & Contributing (8)\\
+    Issue tracker (28) & Test cases (15) & API documentation (7)\\
+    User manual (22) & Authors (14) & Dependency list (7)\\
+    Release info. (22) & Frequently Asked Questions (FAQ) (14) & Troubleshooting guide (6)\\
+     & Acknowledgements (12) & Product roadmap (5)\\
+     & Changelog (12) & Design documentation (5)\\
+     & Citation (11) & Code style guide (3)\\
+     & & Code of conduct (1)\\
+     & & Requirements (1)\\
+    \bottomrule
+    \end{tabular}
+    \caption{Artifacts Present in MI Packages, Classified by Frequency (The number 
+    in brackets is the number of occurrences)}
+    \label{artifactspresent}
+    %\end{center}
+\end{table*}
+
+Table~\ref{artifactspresent} presents our measurements for MI software.
+The table groups the artifacts by frequency
+into categories of common (20 to 29 ($>$67\%) packages), uncommon (10 to 19
+(33-67\%) packages), and rare (1 to 9 ($<$33\%) packages).
+Tables~\ref{tab_maintainability_docs} and~\ref{tab_Visibility/Transparency_docs}
+show the details on which projects use which types of artifacts for documents
+related to maintainability and visibility, respectively.
+
+Note that ``popularity'' in Table~\ref{Tbl_Guidelines} does no imply that these
+oft recommended artifacts are the most important. Guidelines are often brief,
+to encourage adoption, and thus even guidelines that mention the need for
+installation instructions rarely mention uninstallation instructions.
+Two items in Table~\ref{artifactspresent} do not appear in any guidelines:
+\emph{Troubleshooting guide} and \emph{Developer's manual}.  However the
+information within these documents overlaps with
+the recommended artifacts.  Troubleshooting information often can be found in
+a User Manual, while the information in a ``Developer's Manual'' is often
+scattered amongst many other documents.
+
+Three of the 26 recommended artifacts were never observed in the MI software:
+i) Uninstall, ii) Test plans, and iii) Requirements. It is possible that
+some of these were created but never put under version control.
+
+Neglecting requirements documentation is unfortunately common for research
+software, and MI software is no exception to this trend.  Although such
+documentation is recommended by some \cite{TobiasEtAl2018, HerouxEtAl2008,
+SmithAndKoothoor2016}, in practice this is rare~\cite{HeatonAndCarver2015}.
+Sanders and Kelly \cite{SandersAndKelly2008} interviewed 16 scientists from 10
+disciplines and found that none of the scientists created requirements
+specifications, unless regulations in their field mandated such a document.
+Requirements are the least commonly produced type of documentation for research
+software in general \cite{Nguyen-HoanEtAl2010}.
+
+This is unfortunate as when scientific developers are surveyed on their pain
+points, Wiese et al.~\cite{WieseEtAl2019} found that software requirements and
+management is the software engineering discipline that most hurts them,
 accounting for 23\% of the technical problems reported by study participants.
-The lack of support for requirements is likely due to the perception that
-up-front requirements are impossible for research software
-\citep{CarverEtAl2007, SegalAndMorris2008}, but if we drop the insistence on
-``up-front'' requirements, allowing instead for the requirements to be written
-iteratively and incrementally, requirements are feasible \citep{Smith2016}.
-\citet{SmithEtAl2007} provides a requirements template tailored to research
-software.
-
-Table~\ref{Tbl_Guidelines} shows several recommended artifacts that are rarely
-observed in practice.  A theme among these rare artifacts is that, except for
-the user-focused getting started manual, they are developer-focused.  The
-dependency list, which is a list of software library dependencies, was rarely
-observed, but this information is still likely present, just embedded in build
-scripts. The other developer-focused and rare artifacts are as follows:
-
-\begin{itemize}
-
-\item \textbf{A Contributing file} provides new contributors with the information
-that they need to start adding/modifying the repository's files.
-\citet{Abdalla2016} provides a simple template for creating an open-source
-contributor guideline.
-
-\item \textbf{A Developer Code of Conduct} explicitly states the expectations
-for developers on how they should treat one another \citep{TouraniEtAl2017}. The
-code outlines rules for communication and establishes enforcement mechanisms for
-violations.  As \citet{TouraniEtAl2017} states, the developer code documents
-the spirit of a community so that anyone can comfortably contribute regardless
-of ethnicity, gender, or sexual orientation. Three popular codes of conduct are
-\citep{TouraniEtAl2017}:
-\href{https://www.contributor-covenant.org/version/2/1/code_of_conduct/}
-{Contributor Covenant}, \href{https://ubuntu.com/community/code-of-conduct}
-{Ubuntu Code of Conduct}, and \href{https://www.djangoproject.com/conduct/}
-{Django Code of Conduct}. A code of conduct can improve inclusivity, which
-brings the benefit of a wider pool of contributors.  For example, a code of
-conduct can improve the participation of women \citep{SinghEtAl2021}. A standard
-of ethical behaviour can be captured in the code, for projects that are looking
-to abide by a code of ethics, such as the IEEE Code of Ethics \citep{IEEE1999},
-or the Professional Engineers of Ontario code of ethics \citep[p.\
-23--24]{PEO2021}.
-
-\item \textbf{Code Style Guidelines} present standards for writing code. Style
-guides codify such elements as formatting, commenting, naming identifiers, best
-practices and dangers to avoid \citep{Carty2020}. For instance, most coding
-style guides will specify using ALLCAPS when naming symbolic constants.
-Understandability improves under standardization, since developers spend
-more time on the content of the code, and less time distracted by its style.
-Three sample style guides are:
-\href{https://google.github.io/styleguide/javaguide.html} {Google Java Style
-Guide},
-\href{http://cnl.sogang.ac.kr/cnlab/lectures/programming/python/PEP8_Style_Guide.pdf}
-{PEP8 Style Guide for Python}, and
-\href{https://google.github.io/styleguide/cppguide.html} {Google \CC Style
-Guide}.  Linting tools, like \href{https://pypi.org/project/flake8/}{flake8} for
-Python, can be used to enforced coding styles, like the PEP8 standard.
-
-\item \textbf{A Product Roadmap} explains the vision and direction for a product
-offering \citep{MunchEtAl2019}.  Although they have different forms, all
-roadmaps cover the following: \begin{inparaenum}[i)]
-	\item Where are we now?,
-	\item Where are we going?, and
-	\item How can we get there? \end{inparaenum} \citep{PhaalEtAl2005}. As
-stated by \citet{Pichler2012}, a product roadmap provides the following
-benefits: continuity of purpose, facilitation of collaboration, and assistance
-with prioritization. Creating a roadmap involves the following steps:
-\begin{inparaenum}[i)]
-	\item define and outline a strategic mission and product vision, 
-	\item scan the environment, 
-	\item revise and distill the product vision to write the product roadmap, and 
-	\item estimate the product life cycle and evaluate the mix of planned
-development efforts \end{inparaenum} \citep{VahaniittyEtAl2002}.
-
-\item \textbf{Design Documentation} explicitly states the design goals and
-priorities, records the likely changes, decomposes the complex system into
-separate modules, and specifies the relationship between the modules. Design
-documentation shows the syntax of the interface for the modules, and in some
-cases also documents the semantics.  Some potential elements of design
-documentation include the following:
-
-\begin{itemize}
-	\item Representation of the system design and class design using Unified
-	Modelling Language class diagrams. This approach is suited to
-	object-oriented design and designs that use patterns \citep{Gamma1995}.
-	\item Rigorous documentation of the system design following the template for
-	a Module Guide (MG) \citep{ParnasEtAl1984}.  An MG organizes the modules in
-	a hierarchy by their secrets.
-	\item An explanation of the design using data flow diagrams to show typical use cases
-	for input transformation.
-	\item A table or graph showing the traceability between the requirements and
-	the modules (or classes)
-	\item The syntax of the modules (or classes) by providing lists of the state
-	variables, exported constants and all exported access programs for each
-	module (or class).  This shows the interface that can used to access each
-	module's services.
-	\item A formal specification of the semantics of input/output relationships
-	and state transitions for each module using a Module Interface Specification
-	(MIS) \citep{HoffmanAndStrooper1995}. An MIS is an abstract model that
-	formally shows each module's access programs and the associated transitions
-	and outputs based on their state, environment, and input variables.
-	\citet{ElSheikhEtAl2004, SmithAndYu2009} show the example of an MIS for a
-	mesh generator.
-\end{itemize}
-
-\item \textbf{API Documentation} shows developers the services or data provided
-by the software application (or library) through such resources as its methods
-or objects \citep{MengEtAl2018}.  Understandability is improved by API
-documentation \citep{MengEtAl2018}. API documentation can be generated using
-tools like Doxygen, pydoc, and javadoc.
-
-\end{itemize}
-
-The rare artifacts for MI software are similar to the rare artifacts for Lattice
-Boltzmann solvers \citep{Michalski2021}, except LBM software is more likely to
-have developer related artifacts, like Contributing, Dependency list, and Design
-documentation.
-
-To improve MI software in the future, an increased use of checklists could help.
-Developers can use checklists to ensure they follow best practices.  Some
-examples include checklists merging branches into master \citep{Brown2015},
-checklists for saving and sharing changes to the project \citep{WilsonEtAl2016},
-checklists for new and departing team members \citep{HerouxAndBernholdt2018},
-checklists for processes related to commits and releases \citep{HerouxEtAl2008}
-and checklists for overall software quality \citep{ThielEtAl2020, SSI2022}.  For
-instance, for Lattice Boltzmann solver software, ESPResSo has a checklist for
-managing releases \citep{Michalski2021}.
-
-The above discussion shows that, taken together, MI projects fall somewhat short
-of recommended best practices for research software.  However, MI software is
-not alone in this.  Many, if not most, research projects fall short of best
-practices.  A gap exists in research software development practices and software
-engineering recommendations \citep{Storer2017, Kelly2007, OwojaiyeEtAl2021_CSE}.
-\citet{JohansonAndHasselbring2018} observe that the state-of-the-practice for
-research software in industry and academia does not incorporate state-of-the-art
-SE tools and methods.  This causes sustainability and reliability problems
-\citep{FaulkEtAl2009}. Rather than benefit from capturing and reusing previous
-knowledge, projects waste time and energy ``reinventing the wheel''
-\citep{deSouzaEtAl2019}.
+Further adding to the misfortune, there is a widespread perception that up-front
+requirements are impossible for research software \cite{CarverEtAl2007,
+SegalAndMorris2008}. Fortunately a more agile approach to requirements is
+feasible~\cite{Smith2016}, and research-software specific templates
+exist~\cite{SmithEtAl2007}.
+
+A theme emerges amongst the artifacts rarely observed in practice: they
+are developer-focused (a list of library dependencies, a contributor's guide,
+a developer Code of Conduct, coding style guidelines, product roadmap, design
+documentation and API documentation).
+
+Other communities use checklists to help with best practices. Examples include
+checklists for merging branches~\cite{Brown2015}, for saving and sharing changes
+to the project~\cite{WilsonEtAl2016}, for new and departing team
+members~\cite{HerouxAndBernholdt2018}, for processes related to commits and
+releases~\cite{HerouxEtAl2008} and for overall software
+quality~\cite{ThielEtAl2020, SSI2022}.
+
+MI projects fall somewhat short of recommended best practices, but are not alone
+amongst research software projects. This gap has been documented
+before~\cite{Storer2017,JohansonAndHasselbring2018}, and is known to cause
+sustainability and reliability problems \cite{FaulkEtAl2009}, and to waste
+development effort~\cite{deSouzaEtAl2019}.
+
+\subsection{Interview Methods} \label{sec_interview_methods}
+
+The repository-based measurements summarize the information we can collect from
+on-line resources. This information is incomplete because it doesn't generally
+capture the development process, the developer pain points, the perceived
+threats to software quality, and the developers' strategies to address these
+threats.  Therefore, part of our methodology involves interviewing developers.
+
+We based our interviews on a list of 20 questions, which can be found in
+\cite{SmithEtAl2021}. Some questions are about the background of the software,
+the development teams, the interviewees, and how they organize their projects.
+We also ask about the developer's understanding of the users. Some questions
+focus on the current and past difficulties, and the solutions the team has
+found, or plan to try. We also discuss documentation, both with respect to how
+it is currently done, and how it is perceived. A few questions are about
+specific software qualities, such as maintainability, understandability,
+usability, and reproducibility. The interviews are semi-structured based on the
+question list; we ask follow-up questions when necessary.  The interview process
+presented here was approved by the McMaster University Research Ethics Board
+under the application number
+\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/MACREM/Application.pdf}
+{MREB\#: 5219}.
+
+We sent interview requests to all 29 projects using contact information from
+projects websites, code repository, publications, and from biographic pages at
+the teams' institutions.  In the end nine developers from eight of the projects
+agreed to participate: \textit{3D Slicer}, \textit{INVESALIUS 3}, \textit{dwv},
+\textit{BioImage Suite Web}, \textit{ITK-SNAP}, \textit{MRIcroGL},
+\textit{Weasis}, and \textit{OHIF}. We spent about 90 minutes for each
+interview. One participant was too busy to have an interview, so they wrote down
+their answers. In one case two developers from the same project agreed to be
+interviewed. We held the meetings on-line using either Zoom or Teams, which
+facilitated recording and automatic transcription. The full interview answers
+can be found in \cite{Dong2021}.
 
@@
 maintenance, and evolution of software, software processes, and artifacts
-\citep[p.\ 501]{GhezziEtAl2003}. MI software uses tools for CI/CD, user support,
+\cite[p.\ 501]{GhezziEtAl2003}. MI software uses tools for CI/CD, user support,
 version control, documentation, and project management.  To answer
@@
 repositories and over 23 million users (as of roughly February 26, 2020)
-\citep{Kashyap2020}.
+\cite{Kashyap2020}.
 
@@
 improvement from the poor adoption of version control tools that Wilson lamented
-in 2006 \citep{Wilson2006}.  The proliferation of version control tools for MI
+in 2006 \cite{Wilson2006}.  The proliferation of version control tools for MI
 matches the increase in the broader research software community.  A little over
-10 years ago \citet{Nguyen-HoanEtAl2010} estimated that only 50\% of research
+10 years ago \cite{Nguyen-HoanEtAl2010} estimated that only 50\% of research
 software projects use version control, but even at that time
-\citet{Nguyen-HoanEtAl2010} noted an increase from previous usage levels. A
+\cite{Nguyen-HoanEtAl2010} noted an increase from previous usage levels. A
 survey in 2018 shows 81\% of developers use a version control system
-\citep{AlNoamanyAndBorghi2018}. \citet{Smith2018} has similar results, showing
+\cite{AlNoamanyAndBorghi2018}. \cite{Smith2018} has similar results, showing
 version control usage for alive projects in mesh generation, geographic
@@
 small decrease in version control usage for seismology software over the same
-time period, from 41\% down to 36\%).  A recent survey by \citet{CarverEtAl2022}
+time period, from 41\% down to 36\%).  A recent survey by \cite{CarverEtAl2022}
 shows version control use among practitioners at over 95\%, with 83/87 survey
@@
 cited in Section~\ref{Sec_CompareArtifacts} includes the advice to use version
-control. (The USGS guide \citep{USGS2019} was the only set of recommendations to
+control. (The USGS guide \cite{USGS2019} was the only set of recommendations to
 not mention version control.) The high usage of version control tools in MI
@@
 more projects used CI/CD than was evident from repository artifacts alone
-\citep{Michalski2021}.  The 17\% utilization for MI software contrasts with the
+\cite{Michalski2021}.  The 17\% utilization for MI software contrasts with the
 high frequency with which research software development guidelines recommend
-continuous integration \citep{BrettEtAl2021, Brown2015, ThielEtAl2020,
+continuous integration \cite{BrettEtAl2021, Brown2015, ThielEtAl2020,
 Zadka2018, vanGompelEtAl2016}. Although there is currently little data available
@@
 of 24 LBM packages showing evidence of CI/CD in their repositories
-\citep{Michalski2021}.  The survey of \citet{CarverEtAl2022} suggests higher use
+\cite{Michalski2021}.  The survey of \cite{CarverEtAl2022} suggests higher use
 of CI/CD with 54\% (54/100) of respondents indicating that they use it. However,
@@
 documentation tool difference is that LBM software often uses document
-generation tools, like doxygen and sphinx \citep{Michalski2021}, while MI does
+generation tools, like doxygen and sphinx \cite{Michalski2021}, while MI does
 not appear to use these tools.
@@
 
-Based on information provided by \citet{JungEtAl2022}, tool utilization for MI
+Based on information provided by \cite{JungEtAl2022}, tool utilization for MI
 software has much in common with tool utilization for ocean modelling software.
@@
 methods, matches what has been observed for research software in general.
-Scientific developers naturally use an agile philosophy \citep{AckroydEtAl2008,
+Scientific developers naturally use an agile philosophy \cite{AckroydEtAl2008,
 CarverEtAl2007, EasterbrookAndJohns2009, Segal2005, HeatonAndCarver2015}, or an
-amethododical process \citep{Kelly2013}, or a knowledge acquisition driven
-process \citep{Kelly2015}.  A waterfall-like process can work for research
-software \citep{Smith2016}, especially if the developers work iteratively and
+amethododical process \cite{Kelly2013}, or a knowledge acquisition driven
+process \cite{Kelly2015}.  A waterfall-like process can work for research
+software \cite{Smith2016}, especially if the developers work iteratively and
 incrementally, but externally document their work as if they followed a
-rationale design process \citep{parnas1986rational}.
+rationale design process \cite{parnas1986rational}.
 
@@
 reproducibility.  This level of testing matches what was observed for LBM
-software \citep{Michalski2021} and is apparently greater than the level of
-testing for ocean modelling software.  \citet{JungEtAl2022} reports that ocean
+software \cite{Michalski2021} and is apparently greater than the level of
+testing for ocean modelling software.  \cite{JungEtAl2022} reports that ocean
 modellers underemphasize testing.
@@
 named CONTRIBUTING. Guidelines also recommend tutorials, user guides and quick
-start examples. \citet{SmithAndRoscoe2018} suggests including instructions
+start examples. \cite{SmithAndRoscoe2018} suggests including instructions
 specifically for on-boarding new developers. For open-source software in general
-(not just research software), \citet{Fogel2005} recommends providing tutorial
+(not just research software), \cite{Fogel2005} recommends providing tutorial
 style examples, developer guidelines, demos, and screenshots.
@@
 pain points, we did discuss threats to these qualities and ways to improve them
-as part of our interview process \citep{SmithEtAl2021}.  The interviewee's
+as part of our interview process \cite{SmithEtAl2021}.  The interviewee's
 practices for addressing pain points and improving quality can potentially be
@@
 
-\citet{PintoEtAl2018} lists some pain points that did not come up in our
+\cite{PintoEtAl2018} lists some pain points that did not come up in our
 conversations with MI developers: interruptions while coding, scope bloat, lack
 of user feedback, hard to collaborate on software projects, and aloneness.
-\citet{WieseEtAl2019} also mention two research software pain points that did
+\cite{WieseEtAl2019} also mention two research software pain points that did
 not explicitly arise in our interviews: reproducibility, and software scope
 determination.  To the list of pain points not discussed for MI, our study of
-LBM software \citep{SmithEtAl2022} adds lack of software experience for the
+LBM software \cite{SmithEtAl2024} adds lack of software experience for the
 developers, technical debt, and documentation. We did not observe any pain
@@
 of research software also experience the lack of time pain point
-\citep{PintoEtAl2018, PintoEtAl2016, WieseEtAl2019}. Our study of LBM software
-\citep{SmithEtAl2022} also highlighted lack of time as a significant pain point.
+\cite{PintoEtAl2018, PintoEtAl2016, WieseEtAl2019}. Our study of LBM software
+\cite{SmithEtAl2024} also highlighted lack of time as a significant pain point.
 
@@
 
-Funding challenges have also been noted by others \citep{GewaltigAndCannon2012,
-Goble2014, KaterbowAndFeulner2018, SmithEtAl2022}. Researchers that devote time
+Funding challenges have also been noted by others \cite{GewaltigAndCannon2012,
+Goble2014, KaterbowAndFeulner2018, SmithEtAl2024}. Researchers that devote time
 to software have the additional challenge that funding agencies do not always
 count software when they are judging the academic excellence of the applicant.
-\citet{WieseEtAl2019} reported developer pains related to publicity, since
+\cite{WieseEtAl2019} reported developer pains related to publicity, since
 publishing norms have historically made it difficult to get credit for creating
-software.  As studied by \citet{HowisonAndBullard2016}, research software
+software.  As studied by \cite{HowisonAndBullard2016}, research software
 (specifically biology software, but the trend likely applies to other research
-software domains) is infrequently cited. \citet{PintoEtAl2018} also mentions the
+software domains) is infrequently cited. \cite{PintoEtAl2018} also mentions the
 lack of formal reward system for research software.
@@
 
-The pain point survey of \citet{WieseEtAl2019} highlights that technology
+The pain point survey of \cite{WieseEtAl2019} highlights that technology
 hurdles are an issue for research software in general.  Some technical-related
-problems mentioned by \citet{WieseEtAl2019} include dependency management,
-cross-platform compatibility (also mentioned by \citet{PintoEtAl2018}), CI,
-hardware issues and operating system issues. From \citep{SmithEtAl2022}
+problems mentioned by \cite{WieseEtAl2019} include dependency management,
+cross-platform compatibility (also mentioned by \cite{PintoEtAl2018}), CI,
+hardware issues and operating system issues. From \cite{SmithEtAl2024}
 technology pain points for LBM developers include setting up parallelization and
@@
 \item Lack of real world image data for testing, in part because of patient
-privacy concerns (\citet{WieseEtAl2019} mentions that the pain point of privacy
+privacy concerns (\cite{WieseEtAl2019} mentions that the pain point of privacy
 concerns also arises for research software in general);
@@
 software is ahead of some other scientific domains.  For scientific software in
-general \citet{PintoEtAl2018} mention the problem of insufficient testing and
-\citet{HannayEtAl2009} show that more developers think testing is important than
+general \cite{PintoEtAl2018} mention the problem of insufficient testing and
+\cite{HannayEtAl2009} show that more developers think testing is important than
 the number that believe they have a sufficient understanding of testing
@@
 challenges of insufficient testing and insufficient understanding of testing
-concepts \citep{SmithEtAl2022}. Automated testing is a specific challenge for
+concepts \cite{SmithEtAl2024}. Automated testing is a specific challenge for
 LBM software since free testing services do not offer adequate facilities for
-large amounts of data \citep{SmithEtAl2022}. Although not specifically mentioned
+large amounts of data \cite{SmithEtAl2024}. Although not specifically mentioned
 during our interviews, the large data sets for MI likely also cause a challenge
@@
 because for many potential test cases the developer doesn't have a means to
-judge the correctness of their calculated solutions \citep{HannayEtAl2009,
+judge the correctness of their calculated solutions \cite{HannayEtAl2009,
 KanewalaAndBieman2013, KellyEtAl2011, WieseEtAl2019}.  The MI developers did not
@@
 At least to some extent the problems for MI software users are due to holes in
-their background knowledge.  The survey of \citet{WieseEtAl2019} for research
+their background knowledge.  The survey of \cite{WieseEtAl2019} for research
 software in general also mentioned that users do not always have the expertise
-required to install or use the software. \citet{SmithEtAl2022} observes a
+required to install or use the software. \cite{SmithEtAl2024} observes a
 similar pattern for LBM software, with several LBM developers noting that users
@@
 \item[Q\refstepcounter{qnum}\theqnum \label{Q_Maintainability}:]
-\textbf{Maintainability:} \citet{Nguyen-HoanEtAl2010} rate maintainability as the
+\textbf{Maintainability:} \cite{Nguyen-HoanEtAl2010} rate maintainability as the
 third most important software quality for research software in general. The push
-for sustainable software \citep{deSouzaEtAl2019} is motivated by the pain that
+for sustainable software \cite{deSouzaEtAl2019} is motivated by the pain that
 past developers have had with accumulating too much technical debt
-\citep{KruchtenEtAl2012}.  For LBM software, \citet{SmithEtAl2022} identifies
+\cite{KruchtenEtAl2012}.  For LBM software, \cite{SmithEtAl2024} identifies
 technical debt as one of the developer pain points.
@@
 years. They found that their modular approach effectively supports
-maintainability \citep{Joshi2011}.
+maintainability \cite{Joshi2011}.
 
@@
 challenges of inadequate documentation are a known problem for research software
-\citep{PintoEtAl2018, WieseEtAl2019} and for non-research software
-\citep{LethbridgeEtAl2003}.
+\cite{PintoEtAl2018, WieseEtAl2019} and for non-research software
+\cite{LethbridgeEtAl2003}.
 
@@
 Continuous integration involves frequent pushes to a code repository.  With
-every push the software is built and tested \citep[p.\ 13]
-{HumbleAndFarley2010}, \citep{ShahinEtAl2017, Fowler2006}.  CI can take
+every push the software is built and tested \cite[p.\ 13]
+{HumbleAndFarley2010}, \cite{ShahinEtAl2017, Fowler2006}.  CI can take
 significant time and effort to set up and integrate into a team's workflow, but
@@
 	\item Elimination of headaches associated with a separate integration phase
-	\citep{Fowler2006}, \citep[p.\ 20]{HumbleAndFarley2010}. If developers
+	\cite{Fowler2006}, \cite[p.\ 20]{HumbleAndFarley2010}. If developers
 	postpone integration, integration problems are inevitable.  Continuous
@@
 	the problem can be isolated to the small increment that was just committed.
-	\item Detection and removal of bugs \citep{Fowler2006} via
+	\item Detection and removal of bugs \cite{Fowler2006} via
 	automated testing.  To improve productivity, defects are best discovered and
-	fixed at the point where they are introduced \citep[p.\
+	fixed at the point where they are introduced \cite[p.\
 	23]{HumbleAndFarley2010}.  Code is not the only source of errors; they are
 	also found in the files and scripts related to configuration management
-	\citep[p.\ 18]{HumbleAndFarley2010}.
+	\cite[p.\ 18]{HumbleAndFarley2010}.
 	\item Everyone is always working on a stable base, since the rejection of
@@
 \begin{itemize}
-	\item A version control system \citep{Fowler2006}. To be effective, all
+	\item A version control system \cite{Fowler2006}. To be effective, all
 	files should be under version control, not just code files.  Anything that
@@
 	control, including configuration files, build scripts, test harnesses, and
-	operating system configuration files \citep[p. 19]{HumbleAndFarley2010}.
+	operating system configuration files \cite[p. 19]{HumbleAndFarley2010}.
 	Fortunately for the MI, as shown in Section~\ref{sec_score_maintainability}
 	all our measured projects use version control.
-	\item A fully automated build system \citep{Fowler2006}.  As \citet[p.\
+	\item A fully automated build system \cite{Fowler2006}.  As \cite[p.\
 	5]{HumbleAndFarley2010} point out, deploying software manually is an
@@
 	build system will need to add one to pursue using CI.
-	\item An automated test system \citep{Fowler2006}. Building quality software
+	\item An automated test system \cite{Fowler2006}. Building quality software
 	involves creating automated tests at the unit, component, and acceptance
@@
 	code, its configuration, the environment, or the software stack that it runs
-	on \citep[p.\ 83]{HumbleAndFarley2010}. As Table~\ref{artifactspresent}
+	on \cite[p.\ 83]{HumbleAndFarley2010}. As Table~\ref{artifactspresent}
 	shows, test cases are in the uncommon category for MI software artifacts,
@@
 
-To enable incorporation into a team's workflow, \citet[p.\
+To enable incorporation into a team's workflow, \cite[p.\
 60]{HumbleAndFarley2010} explain that the usual approach for CI is to keep the
@@
 cancelling, and if necessary, restarting jobs as merges occur
-\citep{SlaughterEtAl2021}. A more sophisticated process management system can
+\cite{SlaughterEtAl2021}. A more sophisticated process management system can
 also enforce rules for pull requests, like checking that a test specification
 includes the test's motivation, a test description, and a design description for
-all changes \citet{SlaughterEtAl2021}.
+all changes \cite{SlaughterEtAl2021}.
 
@@
 
-\citet{ShahinEtAl2017} highlights the following challenges for adopting CI: lack
+\cite{ShahinEtAl2017} highlights the following challenges for adopting CI: lack
 of awareness and transparency, lack of expertise and skills, coordination and
@@
 common reason given for not adopting CI is that developers are not familiar
-enough with CI \citep{HiltonEtAl2016}.  \citet{ShahinEtAl2017} observes that
+enough with CI \cite{HiltonEtAl2016}.  \cite{ShahinEtAl2017} observes that
 these problems can be mitigated via improving testing activities, planning and
@@
 \href{https://nihcc.app.box.com/v/ChestXray-NIHCC}{Chest X-ray Datasets} by
-National Institute of Health \citep{WangEtAl2017},
+National Institute of Health \cite{WangEtAl2017},
 \href{https://www.cancerimagingarchive.net/}{Cancer Imaging Archive}
-\citep{PriorEtAl2017}, \href{https://medpix.nlm.nih.gov/home}{MedPix} by
-National Library of Medicine \citep{Smirniotopoulos2014}, and datasets for liver
-\citep{BilicEtAl2019} and brain \citep{MenzeEtAl2015} tumor segmentation
+\cite{PriorEtAl2017}, \href{https://medpix.nlm.nih.gov/home}{MedPix} by
+National Library of Medicine \cite{Smirniotopoulos2014}, and datasets for liver
+\cite{BilicEtAl2019} and brain \cite{MenzeEtAl2015} tumor segmentation
 benchmarks.  A team developing MI software should be able to find more open
@@
 A linter is a tool that statically analyzes code to find programming errors,
-suspicious constructs, and stylistic inconsistencies \citep{Wikipedia2022_Lint}.
+suspicious constructs, and stylistic inconsistencies \cite{Wikipedia2022_Lint}.
 Linters can be used as an ad hoc check for code files, but they really come into
@@
 guidelines that we consulted, summarized in Section~\ref{Sec_CompareArtifacts},
-mention linters.  The one exception is \citet{ThielEtAl2020}.  Despite the lack
+mention linters.  The one exception is \cite{ThielEtAl2020}.  Despite the lack
 of mention in the guidelines, we believe that linters have the potential to
@@
 removing silly errors before code reviews, and catching potential security
-issues \citep{SourceLevel2022_Lint}. Most popular programming languages have an
+issues \cite{SourceLevel2022_Lint}. Most popular programming languages have an
 accompanying linter.  For example, Python has the options of PyLint, flake8 and
-Black \citep{Zadka2018}.
+Black \cite{Zadka2018}.
 
@@
 it, if a function could be short-handed, if there are indentation
-inconsistencies, etc. \citep{Whitehouse2018}. By insisting on code
+inconsistencies, etc. \cite{Whitehouse2018}. By insisting on code
 standardization linters can reduce technical debt and thus improve
@@
 analysis, the idea of statically checking for adherence to basic rules can be
-extended to check documentation. \citet{SmithEtAl2018_StatSoft} shows how the
+extended to check documentation. \cite{SmithEtAl2018_StatSoft} shows how the
 use of tools to enforce documentation standards partially explains the
@@
 We advocate incorporating peer review into the development process, as
-frequently recommended for research software \citep{HerouxEtAl2008, Givler2020,
+frequently recommended for research software \cite{HerouxEtAl2008, Givler2020,
 OrvizEtAl2017, USGS2019}. In most cases a modern, lightweight review, should be
 adequate.  Modern code review is informal, tool-based, asynchronous, and focused
-on reviewing code changes \citep{SadowskiEtAl2018}. Managing a project via
+on reviewing code changes \cite{SadowskiEtAl2018}. Managing a project via
 GitHub pull requests is an example of a modern approach to reviewing code.
@@
 review because of the inefficiencies of rigorous inspections
-\citep{RigbyAndBird2013}.  However, for important parts of the code, developers
+\cite{RigbyAndBird2013}.  However, for important parts of the code, developers
 may benefit from mixing in a more rigorous approach.
 
-\citet{Fagan1976} began work on rigorous review via code inspection.  Elements
+\cite{Fagan1976} began work on rigorous review via code inspection.  Elements
 of a typical inspection include reviewing the code against a checklist (checking
@@
 etc.) Rigorous inspection finds 60-65\% of latent defects on average, and often
-tops 85\% in defect removal efficiency \citep{Jones2008}. The success rate of
+tops 85\% in defect removal efficiency \cite{Jones2008}. The success rate of
 code inspection is generally higher than most forms of testing, which average
-between 30 --- 35\% for defect removal efficiency \citep{EbertAndJones2009,
-Jones2008}. For research software, \citep{KellyAndShepard2000} show a task based
+between 30 --- 35\% for defect removal efficiency \cite{EbertAndJones2009,
+Jones2008}. For research software, \cite{KellyAndShepard2000} show a task based
 inspection approach can be effective. Task based inspection is an ideal fit with
@@
 realize the benefits of modern code review noticed by
-\citet{BirdAndBacchelli2013}: defect detection, knowledge transfer, increased
+\cite{BirdAndBacchelli2013}: defect detection, knowledge transfer, increased
 team awareness, and creation of alternative solutions to problems.
@@
 In our ``state of the practice'' assessment exercise for LBM software
-\citep{SmithEtAl2022}, we noticed that LBM developers implicitly used
+\cite{SmithEtAl2024}, we noticed that LBM developers implicitly used
 modularization based on the principle of design for change to improve
@@
 the same principle for their modularizations.  Although the advice to modularize
-research software to handle complexity is common \citep{WilsonEtAl2014,
+research software to handle complexity is common \cite{WilsonEtAl2014,
 StewartEtAl2017, Storer2017}, specific guidelines on how to divide the software
 into modules is less prevalent.  Not every decomposition is a good design for
-supporting change, as shown by \citet{Parnas1972a}.  For instance, a design with
-low cohesion and high coupling \citep[p.\ 48]{GhezziEtAl2003} will make change
+supporting change, as shown by \cite{Parnas1972a}.  For instance, a design with
+low cohesion and high coupling \cite[p.\ 48]{GhezziEtAl2003} will make change
 difficult. Especially in research software, where change is inevitable,
 designers need to produce a modularization that supports change.
-\citet{JungEtAl2022} points out that ocean modelling software is currently
+\cite{JungEtAl2022} points out that ocean modelling software is currently
 feeling the pain of not emphasizing modularization in legacy code.
 
-Specific examples of design for change for LBM software \citep{SmithEtAl2022}
+Specific examples of design for change for LBM software \cite{SmithEtAl2024}
 include the following:
@@
 to ease future change. 
-\item \href{https://github.com/CFD-GO/TCLB}{TCLB} \citep{rokicki2016adjoint} is
+\item \href{https://github.com/CFD-GO/TCLB}{TCLB} \cite{rokicki2016adjoint} is
 designed to allow for the addition of some LBM features, but changes to major
@@
 model will be an easy contribution'', but changes to the ``Cartesian mesh … will
-be a nightmare'' \citep{SmithEtAl2022}. The design of TCLB highlights that not
+be a nightmare'' \cite{SmithEtAl2024}. The design of TCLB highlights that not
 every conceivable change needs to be supported, only the likely changes.  
@@
 second by hiding each likely change behind a well-defined module interface. This
-approach mirrors the recommendations from \citet{Parnas1972a}.
+approach mirrors the recommendations from \cite{Parnas1972a}.
 Section~\ref{Sec_CompareArtifacts} lists ideas for how to document the design,
@@
 
-\subsection{Assurance Case} \label{AssuranceCases}
-
-To ensure correctness (\ppref{P_Correctness} and to achieve the quality of
-reproducibility (\qref{Q_Reproducibility}), we recommend considering the use of
-assurance cases.  \citet{RinehartEtAl2015} defines an assurance case as ``[a]
-reasoned and compelling argument, supported by a body of evidence, that a
-system, service, or organization will operate as intended for a defined
-application in a defined environment.''  An assurance cases provide an organized
-and explicit argument that the software and its documentation achieves desired
-qualities, such as correctness and reproducibility.  Although assurance cases
-have been successfully employed for safety critical systems
-\citep{RinehartEtAl2015}, this technique is relatively new for research software
-\citep{SmithEtAl2020_AC, Smith2018}.
-
-One way to present an assurance case is through the Goal Structuring Notation
-(GSN) \citep{Spriggs2012}, which make arguments clear, easy to read and, hence,
-easy to challenge. GSN starts with a Top Goal (Claim), like ``Program X delivers
-correct outputs when used for its intended use/purpose in its intended
-environment.''  We then decompose this top goal into Sub-Goals, which themselves
-may be further decomposed.  The purpose of the decomposition is to take the
-abstract higher level goals and bring them down to something concrete that can
-be proven.  The decomposition ends with the terminal Sub-Goals that are
-supported by Solutions (Evidence). Typical evidence will consist of documents,
-expert reviews, test case results, peer review, etc.  Within the GSN framework,
-there are also strategy blocks, which describe the rationale for decomposing a
-Goal or Sub-Goal into more detailed Sub-Goals. A common tool for creating,
-editing, and presenting, a GSN argument is \href{https://astah.net/} {Astah}.
-
-\citet{SmithEtAl2020_AC} shows the example of arguing for the correctness of the
-Analysis of Functional NeuroImages (AFNI) package 3dfim+ \citep{Ward2000}.
-3dfim+ analyzes the activity of the brain by computing the correlation between
-an ideal signal and the measured brain signal for each voxel. The assurance case
-for the correctness of 3dfim+ has the top level decomposed into four sub-goals,
-as shown in Figure~\ref{TopGoal}.  This example follows the same pattern as used
-for medical devices \citep{Wassyng2015}.  The first sub-goal (GR) argues for the
-quality of the documentation of the requirements.  The second sub-goal (GD) says
-that the design complies with the requirements and the third proposes that the
-implementation also complies with the requirements.  The fourth sub-goal (GA)
-claims that the inputs to 3dfim+ will satisfy the operational assumptions, since
-we need valid input to make an argument for the correctness of the output.
-
-\begin{figure}[!ht]
-\centering
-\includegraphics[width=1.0\textwidth]{./figures/TopGoal.pdf}
-\caption{Top Goal of the assurance case and its sub-goals}
-\label{TopGoal}
-\end{figure}
-
-Preparing an assurance case for the pre-existing 3dfim+ software shows the value
-of an assurance cases for research software. Although \citet{SmithEtAl2020}
-found no errors in the output of the existing software, the rigour of the
-proposed approach did lead to finding ambiguities and omissions in the existing
-documentation, such as missing information on the coordinate system convention.
-In addition, a potential concern for the software itself was identified from the
-GA argument: running the software does not produce any warning about the
-obligation of the user to provide data that matches the parametric statistical
-model employed for the correlation calculations.
-
-\subsection{Generate All Things} \label{Sec_GenAllThings}
-
-To address developer pain points, we propose automatically generating MI code
-and its documentation via a ``Generate All Things'' (GAT) approach. A GAT
-approach uses models to capture knowledge in domains such as physics, computing,
-mathematics, documentation, and certification.  Developers combine and transform
-knowledge via explicit ``recipes'', which weave it together to generate the
-desired code, documentation, test cases, inspection reports and build scripts. A
-recipe can even potentially be written to generate an assurance case
-(Section~\ref{AssuranceCases}). Our definition of GAT implies generation of all
-software artifacts, not just the code. GAT moves development to a higher level
-of abstraction so that domain experts can work without concern for low-level
-implementation details. GAT allows developers to optimally generate code and
-documentation, reduce the likelihood of errors, eliminate redundancy, and
-automate maintenance. With a GAT approach MI developers can experiment with
-different algorithm choices, different input formats, different outputs formats,
-etc.
-
-Part of GAT is code generation.  In the future, some believe that code
-generation will transform coding, documentation, design, and verification
-\citep{JohansonAndHasselbring2018, Smith2018}. GAT removes the distraction of
-writing software, allowing developers to focus on their science. A GAT approach
-removes the maintenance headaches of documentation duplicates and near
-duplicates \citep{LucivEtAl2018}, since developers capture knowledge once and
-transform it as needed.  Code generation has previously been applied to improve
-research software, such as linear algebra software packages like Blitz++
-\citep{Veldhuizen1998}, and ATLAS (Automatically Tuned Linear Algebra Software)
-\citep{WhaleyEtAl2001}.  Software/hardware generation has been applied for
-digital signal processing in Spiral \citep{Pueschel2001}. A generative approach
-has also been used for a family of efficient, type-safe Gaussian elimination
-algorithms \citet{Carette2008}. \citep{LoggEtAl2012} use code generation when
-solving partial differential equations in FEniCS (Finite Element and
-Computational Software). \citet{MatkerimEtAl2013} and \citet{OberEtAl2018} use
-code generation for High Performance Computing (HPC), using UML (Unified Modelling
-Language) for their domain models. \citet{SzymczakEtAl2016} presents initial work on
-the GAT approach, \citet{SmithAndCarette2021-BRIC} presents a motivating example,
-and \citep{CaretteEtAl2021-Drasil} provides a prototype.
-
-A GAT approach addresses multiple pain points.  For example, a generative
-approach can decrease development time (\ppref{P_LackDevTime}) by automation,
-once the necessary infrastructure is in place.  The GAT approach addresses the
-technology related pain point (\ppref{P_TechnologyHurdles}) because technology
-information can be captured in the models and transformed as needed. To ensure
-correctness (\ppref{P_Correctness}), a GAT approach should be correct by
-construction.  If there are mistakes, GAT has the advantage that they are
-propagated throughout the generated artifacts, which greatly increases the
-chance that someone will notice the mistake.  Maintainability
-(\qref{Q_Maintainability}) is addressed because developers write the recipes
-used for generation at a high level making them relatively easy to change.
-Usability (\ppref{P_Usability}) is addressed because of the emphasis on
-generating up-to-date documentation.  GAT facilitates reproducibility
-(\qref{Q_Reproducibility}) because at any time all the code and documentation
-can be regenerated.  The generator can include explicit traceability to show the
-dependence of the software on specific versions of software libraries.
-
 \section{Threats to Validity} \label{sec_threats_to_validity}
@@
 Our categories come from an analysis of software engineering secondary studies
-by \citet{AmpatzoglouEtAl2019}, where a secondary study analyzes the data from a
-set of primary studies.  \citet{AmpatzoglouEtAl2019} is appropriate because a
+by \cite{AmpatzoglouEtAl2019}, where a secondary study analyzes the data from a
+set of primary studies.  \cite{AmpatzoglouEtAl2019} is appropriate because a
 common example of a secondary study is a systematic literature review. Our
@@
 identified similar threats to validity in our assessment of the state of the
-practice of Lattice Boltzmann Solvers \citep{SmithEtAl2022}.
+practice of Lattice Boltzmann Solvers \cite{SmithEtAl2024}.
 
@@
 
-In \citet{SmithEtAl2016} we reduced concern over the reliability risk associated
-with the reviewer's judgement  % BLIND REVIEW
-by demonstrating that the measurement process is reasonably reproducible.  In
-\citet{SmithEtAl2016} we graded five software products by two reviewers. Their
-rankings were almost identical. As long as each grader uses consistent
-definitions, the relative comparisons in the AHP results will be consistent
-between graders.
+In \cite{SmithEtAl2016} we reduced concern over the reliability risk associated
+with the reviewer's judgement by demonstrating that the measurement process is
+reasonably reproducible.  In \cite{SmithEtAl2016} we graded five software
+products by two reviewers. Their rankings were almost identical. As long as each
+grader uses consistent definitions, the relative comparisons in the AHP results
+will be consistent between graders.
 
@@
 
-\citet{RunesonAndHost2009} defines construct validity as the adopted
+\cite{RunesonAndHost2009} defines construct validity as the adopted
 metrics representing what they are intended to measure. Our construct threats are
@@
 Although there are differences, like the importance of data privacy for MI data,
-we found the approach to developing LBM software \citep{SmithEtAl2022} and MI
+we found the approach to developing LBM software \cite{SmithEtAl2024} and MI
 software to be similar.  Except for the domain specific aspects, we believe that
@@
         measuring MI software. This process is similar to fuzz testing
-        \citep{enwiki:1039424308}, which is one type of fault injection
-        \citep{enwiki:1039005082}. We may adopt more fault injection methods,
+        \cite{enwiki:1039424308}, which is one type of fault injection
+        \cite{enwiki:1039005082}. We may adopt more fault injection methods,
         and identify tools and libraries to automate this process.
@@
         domain experts. Ideas for getting started are available in
-        \citet{SmithEtAl2021}.
+        \cite{SmithEtAl2021}.
         \item Surface understandability: our current method does not require
@@
         the modifications, testing the resulting software and surveying the
-        participants \citep{SmithEtAl2021}.
+        participants \cite{SmithEtAl2021}.
     \end{itemize}
@@
 
+\subsection{Threats to Validity} \label{sec_threats_to_validity}
+
+We follow Ampatzoglou et al.\ \cite{AmpatzoglouEtAl2019}'s analysis of
+threats to validity in software engineering secondary studies.
+
+\subsubsection{Reliability}
+
+A study is reliable if repeating it by another researcher, using the same
+methodology, would lead to the same results~\cite{RunesonAndHost2009}.
+We identify the following threats:
+
+\begin{itemize}
+\item One individual does the measures for all packages. A different
+evaluator might find different results, due to differences in abilities,
+experience, and biases.
+\item The measurements for the full set of packages took several months (of
+elapsed time).  Over this time the software repositories may have changed and
+the reviewer's judgement may have drifted.
+\end{itemize}
+
+The measurement process used has previously been shown to be reasonably
+reproducible.  [citation redacted] 
+%% DOUBLE BLIND
+%Smith et al.~\cite{SmithEtAl2016} 
+reports grading five software
+products by two reviewers. Their rankings were almost identical. As long as each
+grader uses consistent definitions, the relative comparisons in the AHP results
+will be consistent between graders.
+
+\subsubsection{Construct Validity}
+
+Construct validity is when the adopted metrics represent what they are intended
+to measure~\cite{RunesonAndHost2009}. We have identified the following
+potential issues:
+\begin{itemize}
+\item We make indirect measurement of software qualities since meaningful direct
+measures for qualities like maintainability, reusability and verifiability, are
+unavailable.  We follow the usual assumption that developers achieve higher
+quality by following procedures and adhering to standards \cite[p.\
+112]{VanVliet2000}.
+\item We could not install or build \textit{dwv}, \textit{GATE}, and
+\textit{DICOM Viewer}. We used a deployed on-line version for \textit{dwv}, a
+virtual machine version for \textit{GATE}, but no alternative for \textit{DICOM
+Viewer}.
+\item Robustness measurements involve only two pieces of data, leading
+to limited variation in the robustness scores
+(Figure~\ref{fg_robustness_scores}). Our measurement-time budget limited
+what we could achieve here.
+\item Our maintainability proxies (higher ratio of comments to source,
+high percentage of closed issues) have not been validated.
+\item While smaller modules tend to be easier to reuse, small modules are
+not necessarily good modules, nor understandable modules.
+\item The understandability measure relies on 10 random source code files, but
+the 10 files will not necessarily be representative. 
+\item Our overall AHP ranking makes the unrealistic assumption of equal
+weighting.
+\item We approximated popularity by stars and watches.
+\item Table~\ref{Tbl_Guidelines} required judgement as
+not all guidelines use the same names for artifacts that contain
+essentially the same information.
+\end{itemize}
+
+\subsubsection{Internal Validity} \label{Sec_InternalValidity}
+
+Internal validity means that discovered causal relations are trustworthy and
+cannot be explained by other factors~\cite{RunesonAndHost2009}. We identify
+the following:
+
+\begin{itemize}
+\item Our search (Section~\ref{ch_results}) could
+have missed a relevant package.
+\item Our methodology assumes that development activities
+will leave a trace in the repositories, but this is not necessarily true. For
+instance, we saw little evidence of requirements
+(Section~\ref{Sec_CompareArtifacts}), but teams might keep this kind of
+information outside their repos.
+\end{itemize}
+
+\subsubsection{External Validity}
+
+If the results of a study can be generalized to other situations, then the
+study is externally valid \cite{RunesonAndHost2009}.  In other words, we cannot
+generalize our results if the development of MI software is fundamentally
+different from other research software.  Although there are differences, like
+the importance of data privacy for MI data, we found the approach to developing
+[other software (citation redacted)]
+%% DOUBLE BLIND
+%LBM software \cite{SmithEtAl2024} 
+and MI software to be similar.  Except for
+the domain specific aspects, the trends observed in the current study are
+similar to that for other research software.
+
+\section{Conclusions} \label{ch_conclusions}
+
+Our analysis of the state of the practice for MI domain along nine software
+qualities strongly indicates that ``higher quality'' is consistent with
+community ranking proxies. Although our quality measures are rather shallow, we
+see this as an advantage. The shallow measures are a proxy for the importance of
+\emph{first impressions} for software adoption.
+
+\begin{table*}[ht!]
+	%\begin{center}
+	\renewcommand{\arraystretch}{1.8}
+	\begin{tabular}{ p{2.3cm}p{9.9cm} }
+		\toprule
+
+		Quality & Ranked 1st or 2nd\\
+
+		\midrule
+
+		Installability & 3D Slicer, BioImage Suite Web, Slice:Drop, INVESALIUS\\
+
+		\pbox{3.0cm}{Correctness \\ and Verifiability} & OHIF Viewer, 3D
+		Slicer, ImageJ\\
+
+		Reliability & SMILI, ImageJ, Fiji, 3D Slicer, Slice:Drop, OHIF
+		Viewer\\
+
+		Robustness & XMedCon, Weasis, SMILI, ParaView, OsiriX Lite,
+		MicroView, medInria, ITK-SNAP, INVESALIUS, ImageJ, Horos, Gwyddion,
+		Fiji, dicompyler, DicomBrowser, BioImage Suite Web, AMIDE, 3DimViewer,
+		3D Slicer, OHIF Viewer, DICOM Viewer\\
+
+		Usability & 3D Slicer, ImageJ, Fiji, OHIF Viewer, ParaView,
+		INVESALIUS, Ginkgo CADx, SMILI, OsiriX Lite, BioImage Suite Web,
+		ITK-SNAP, medInria, MicroView, Gwyddion\\
+
+		Maintainability & 3D Slicer, Weasis, ImageJ, OHIF Viewer, ParaView\\
+
+		Reusability & 3D Slicer, ImageJ, Fiji, OHIF Viewer, SMILI, dwv, BioImage
+		Suite Web, GATE, ParaView\\
+
+		\pbox{3.0cm}{Understandability} & 3D Slicer, ImageJ, Weasis,
+		Fiji, Horos, OsiriX Lite, dwv, Drishti, OHIF Viewer, GATE, ITK-SNAP,
+		ParaView, INVESALIUS\\
+
+		\pbox{3.0cm}{Visibility and \\Transparency} & ImageJ, 3D Slicer, Fiji\\
+
+		Overall Quality & 3D Slicer, ImageJ\\
+
+		\bottomrule		
+	\end{tabular}
+	\caption{Top performers for each quality (sorted by order of quality
+	measurement)} \label{topperformerstable}
+	%\end{center}
+\end{table*}
+
+Our grading scores indicate that
+\textit{3D Slicer}, \textit{ImageJ}, \textit{Fiji}
+and \textit{OHIF Viewer} are the overall top four.  However, the
+separation between the top performers and the others is not extreme.  Almost
+all packages do well on at least a few qualities, as shown in
+Table~\ref{topperformerstable}, which summarizes the packages ranked first and
+second for each quality. Almost 70\% (20 of 29) of the software packages appear
+in the top two for at least two qualities.  The only packages that do not appear
+in Table~\ref{topperformerstable}, or only appear once, are \textit{Papaya},
+\textit{MatrixUser}, \textit{MRIcroGL}, \textit{XMedCon}, \textit{dicompyler},
+\textit{DicomBrowser}, \textit{AMIDE}, \textit{3DimViewer}, and
+\textit{Drishti}.
+
+While we did find a reasonable amount of documentation, especially when
+consider all MI projects, there were definitely some holes. Some important
+documentation (test plans and requirements documentation) was missing,
+and other (contributors' guide, code of conduct, code style guidelines, product
+roadmap, design documentation, and API documentation)
+(Section~\ref{Sec_CompareArtifacts}) were rare.
+
+A deeper understanding of the needs of
+the MI community will require data beyond what is available in
+repositories.  Future work should involve interviewing MI developers to
+better understand their ``pain points''.
+
 \section*{Acknowledgements}
 
-We would like to thank Peter Michalski and Oluwaseun Owojaiye for fruitful
-discussions on topics relevant to this paper.  We would also like to thank Jason
-Balaci for advice on web applications.
+[redacted for blind review]
+%% DOUBLE BLIND
+% We would like to thank Peter Michalski and Oluwaseun Owojaiye for fruitful
+% discussions on topics relevant to this paper.
 
@@
 
-%% The Appendices part is started with the command \appendix;
-%% appendix sections are then done as normal sections
-%% \appendix
-
-%% \section{}
-%% \label{}
-
-%% If you have bibdatabase file and want bibtex to generate the
-%% bibitems, please use
-%%
-%%  \bibliographystyle{elsarticle-harv} 
-%%  \bibliography{<your bibdatabase>}
-
-%% else use the following coding to input the bibitems directly in the
-%% TeX file.
-
-%% Bibliography
-\bibliographystyle{ACM-Reference-Format}
-\bibliography{MedImageSoft_SOP}
-
-% \begin{thebibliography}{00}
-
-% %% \bibitem[Author(year)]{label}
-% %% Text of bibliographic item
-
-% \bibitem[ ()]{}
-
-% \end{thebibliography}
+\bibliographystyle{Med_Bibliography_Style}
+\bibliography{SOP_MI_CritRev}
 
 \end{document}
-
-\endinput
-%%
-%% End of file `elsarticle-template-harv.tex'.
-
-%%%%%%%%%%%

### 2024-10-29 StateOfPractice/Papers/SOP_MI_CritReviewsBioMedEng/SOP_MI_OSR.tex
Revised abstract, now for full paper

@@
 \author[CAS]{Jacques Carette}
-\author[EE]{Michael Noseworthy}
+\author[ECE]{Michael Noseworthy}
 
@@
 
-\affiliation[ME]{organization={McMaster University, Electrical and Computer Engineering}, %Department and Organization
+\affiliation[ECE]{organization={McMaster University, Electrical and Computer Engineering}, %Department and Organization
             addressline={1280 Main Street West}, 
@@
 
-    We present the state of the practice for Medical Imaging (MI) software based
-on data available in open source repositories. We selected 29 projects from 48
+We present the state of the practice for Medical Imaging (MI) software based on
+data available in open source repositories. We selected 29 projects from 48
 candidates and assessed 9 software qualities (installability, correctness/
@@
 with the community's view, with four of these also appearing in the top five
-using GitHub metrics (stars-per-year).  
-In general, the quality and quantity of documentation present in a project
-correlates quite well with its popularity.
+using GitHub metrics (stars-per-year).  The quality and quantity of
+documentation present in a project correlates quite well with its popularity.
+Generally, MI software is in a healthy state as shown by the following: in the
+repositories we observed 88\% of the documentation artifacts recommended by
+research software development guidelines, 100\% of MI projects use version
+control tools, and developers appear to use the common quasi-agile research
+software development process. However, the current state of the practice
+deviates from the existing guidelines because of the rarity of some recommended
+artifacts (like a test plan, requirements specification, code of conduct, and
+code style guidelines), low usage of continuous integration (17\% of the
+projects), low use of unit testing (about 50\% of projects), and room for
+improvement with documentation. From interviewing the developers, we identified
+6 concerns: lack of development time, lack of funding, technology hurdles,
+correctness, usability, maintainability, and reproducibility. Recommendations
+for improving the state of the practice include the following: increase
+documentation, increase testing by enriching datasets, increase continuous
+integration usage, move to web applications, employ linters, use peer reviews,
+and design for change.
 

### 2024-10-29 StateOfPractice/Papers/SOP_MI_CritReviewsBioMedEng/SOP_MI_OSR.tex
Updates up to and including Section 2

@@
 \item[RQ\refstepcounter{rqnum}\therqnum \label{RQ_WhatProjects}:] What MI
-\textbf{open source} software projects exist? (Section~\ref{ch_results})
+{open source} software projects exist? (Section~\ref{ch_results})
 \item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_HighestQuality}:] Based on
@@
 (Section~\ref{Sec_CompareArtifacts})
+\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareToolsProjMngmnt}:]
+How do MI projects compare to research software in general with respect to
+the use of tools (Section~\ref{Sec_CompareTools}) for:
+\begin{enumerate} 
+	\item [\rqref{RQ_CompareToolsProjMngmnt}.a] development; and,
+	\item [\rqref{RQ_CompareToolsProjMngmnt}.b] project management?
 \end{enumerate}
+\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareMethodologies}:]
+How do MI projects compare to research software in general with respect to
+principles, processes, and methodologies used?
+(Section~\ref{Sec_CompareMethodologies})
+\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_PainPoints}:] What are
+the pain points for developers working on MI software projects?
+(Section~\ref{painpoints})
+\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_ComparePainPoints}:] How
+do the pain points of developers from MI compare to the pain points
+for research software in general? (Section~\ref{painpoints})
+\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Concerns}:] For MI
+developers what specific best practices are taken to address the pain points
+and software quality concerns? (Section~\ref{painpoints})
+\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Recommend}:]
+What research software development practice could potentially address the
+pain point concerns identified in \rqref{RQ_PainPoints}?
+(Section~\ref{ch_recommendations})
+\end{enumerate}
 
@@
 We only cover MI visualization software.  We exclude other categories of MI
-software, including Segmentation, Registration, Visualization, Enhancement,
-Quantification, Simulation, plus MI archiving and telemedicine systems
-(Compression, Storage, and Communication).  We also exclude Statistical Analysis
-and Image-based Physiological Modelling and Feature Extraction, Classification,
-and Interpretation. Software that provides MI support functions is also out of
+software, including segmentation, registration, visualization, enhancement,
+quantification, simulation, plus MI archiving and telemedicine systems
+(compression, storage, and communication).  We also exclude statistical analysis
+and image-based physiological modelling and feature extraction, classification,
+and interpretation. Software that provides MI support functions is also out of
 scope; therefore, we have not assessed the toolkit libraries VTK and ITK.
@@
 \begin{figure*}[!ht]
-\includegraphics[scale=0.53]{template.pdf}
+\includegraphics[scale=0.67]{template.pdf}
 \caption{Grading template example}
@@
 
+In this section we answer \rqref{RQ_WhatProjects} and \rqref{RQ_HighestQuality}.
+
+\subsection{In-Scope Open-Source MI Software} \label{SecWhatProjects}
+
 We initially identified 48 candidate software projects from the literature
@@
 packages were on our out-of-scope list. The Domain Expert agreed with our final
-choice of 29 packages -- see Table~\ref{tab_final_list}. This contains summary
-data collected in the year 2020.
-
-The projects are sorted in descending order of lines of code.  We found the
-initial release dates (Rlsd) for most projects and marked the two unknown dates
-with ``?''. The date of the last update is the date of the latest update, at the
-time of measurement. We found funding information (Fnd) for only eight projects.
-For the Number Of Contributors (NOC) we considered anyone who made at least one
-accepted commit as a contributor. The NOC is not usually the same as the number
-of long-term project members, since many projects received change requests and
-code from the community.  With respect to the OS, 25 packages work on all three
-OSs: Windows (W), macOS (M), and Linux (L). Although the usual approach to
-cross-platform compatibility was to work natively on multiple OSes, five
-projects achieved platform-independence via web applications. The full
-measurement data for all packages is available on [removed for blind review]
-%\href{https://data.mendeley.com/datasets/k3pcdvdzj2/1} {Mendeley Data}.  
-%% DOUBLE BLIND
+choice of 29 packages.  Table~\ref{tab_final_list} summarizes the in-scope
+open-source MI software that is available at the time of measurement (the year
+2020), thus answering \rqref{RQ_WhatProjects}.
 
@@
 
+In Table~\ref{tab_final_list} the projects are sorted in descending order of
+lines of code.  We found the initial release dates (Rlsd) for most projects and
+marked the two unknown dates with ``?''. The date of the last update is the date
+of the latest update, at the time of measurement. We found funding information
+(Fnd) for only eight projects.  For the Number Of Contributors (NOC) we
+considered anyone who made at least one accepted commit as a contributor. The
+NOC is not usually the same as the number of long-term project members, since
+many projects received change requests and code from the community.  With
+respect to the OS, 25 packages work on all three OSs: Windows (W), macOS (M),
+and Linux (L). Although the usual approach to cross-platform compatibility was
+to work natively on multiple OSes, five projects achieved platform-independence
+via web applications. The full measurement data for all packages is available on
+[removed for blind review]
+%\href{https://data.mendeley.com/datasets/k3pcdvdzj2/1} {Mendeley Data}.  
+%% DOUBLE BLIND
+
 The programming languages used in order of decreasing popularity are \CC,
@@
 
-In the absence of
-a specific real world context, we assumed all nine qualities are equally
-important. Figure~\ref{fg_overall_scores} shows the overall scores in descending
-order.
+\subsection{Which Projects Follow Best Practices?}
+
+To answer \rqref{RQ_HighestQuality} we measured the software as described in
+Section~\ref{SecMethodology}.  In the absence of a specific real world context,
+we assumed all nine qualities are equally important.
+Figure~\ref{fg_overall_scores} shows the overall AHP scores in descending order.
 
 \begin{figure*}[!ht]
-\includegraphics[scale=0.38]{overall_scores.pdf}
+\includegraphics[scale=0.47]{overall_scores.pdf}
 \caption{Overall AHP scores with an equal weighting for all 9 software qualities}

### 2024-10-29 StateOfPractice/Papers/SOP_MI_CritReviewsBioMedEng/SOP_MI_OSR.tex
Edits up to Section 6 (inclusive)

@@
 \item[RQ\refstepcounter{rqnum}\therqnum \label{RQ_WhatProjects}:] What MI
-{open source} software projects exist? (Section~\ref{ch_results})
+{open source} software projects exist? (Section~\ref{SecWhatProjects})
 \item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_HighestQuality}:] Based on
 quantitative measurements of each project's development practices, which
-projects follow best practices? (Section~\ref{ch_results})
+projects follow best practices? (Section~\ref{SecMeasurementResults})
 \item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareHQ2Popular}:] How
@@
 (Section~\ref{Sec_CompareArtifacts})
-\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareToolsProjMngmnt}:]
-How do MI projects compare to research software in general with respect to
-the use of tools (Section~\ref{Sec_CompareTools}) for:
-\begin{enumerate} 
-	\item [\rqref{RQ_CompareToolsProjMngmnt}.a] development; and,
-	\item [\rqref{RQ_CompareToolsProjMngmnt}.b] project management?
-\end{enumerate}
+\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareToolsProjMngmnt}:] How
+do MI projects compare to research software in general with respect to the use
+of tools (Section~\ref{Sec_CompareTools}) for development
+(\rqref{RQ_CompareToolsProjMngmnt}.a); and, project management
+(\rqref{RQ_CompareToolsProjMngmnt}.b)?
 \item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareMethodologies}:]
@@
 % Statistical software for psychology \cite{smith2018statistical}.  
-We maintain
-the previous constraint that the work load for measuring a given domain should
-take around one person-month's worth of effort ($20$ working days at $8$
-person-hours per day).
+We maintain the previous constraint that the work load for measuring a given
+domain should take around one person-month's worth of effort ($20$ working days
+at $8$ person-hours per day).
 
@@
 
-\section{Results} \label{ch_results}
-
-In this section we answer \rqref{RQ_WhatProjects} and \rqref{RQ_HighestQuality}.
-
-\subsection{In-Scope Open-Source MI Software} \label{SecWhatProjects}
+\subsubsection{Interview Methods} \label{sec_interview_methods}
+
+The repository-based measurements are incomplete because they don't generally
+capture the development process, the developer pain points, the perceived
+threats to software quality, and the developers' strategies to address these
+threats.  Therefore, part of our methodology involves interviewing developers.
+We based our interviews on a list of 20 questions, which can be found in
+[citation redacted for double blind].
+%Smith et al. \cite{SmithEtAl2021}. 
+Some questions are about the background of the software, the development teams,
+the interviewees, and how they organize their projects.  We also ask about the
+developer's understanding of the users. Some questions focus on the current and
+past difficulties, and the solutions the team has found, or plan to try. We also
+discuss documentation, both with respect to how it is currently done, and how it
+is perceived. A few questions are about specific software qualities, such as
+maintainability, understandability, usability, and reproducibility. The
+interviews are semi-structured based on the question list.
+
+\section{In-Scope Open-Source MI Software} \label{SecWhatProjects}
 
@@
 
-\subsection{Which Projects Follow Best Practices?}
+\section{Which Projects Follow Best Practices?} \label{SecMeasurementResults}
 
@@
 
-[Summary description list]
-
-\subsection{Installability} \label{sec_result_installability}
-
-We found
-installation instructions for 16 projects, but two did not need them
-(\textit{BioImage Suite Web} and \textit{Slice:Drop})
-as they are web applications. 10 of the projects required extra
-dependencies: Five depend on a specific browser; \textit{dwv}, \textit{OHIF
-Viewer}, and \textit{GATE} needs extra libraries to build; \textit{ImageJ} and
-\textit{Fiji} need an unzip tool; \textit{MatrixUser} needs Matlab;
-\textit{DICOM Viewer} needs a Nextcloud platform.
-
-The scores are based on the ease of following the installation instructions,
-and automated installation and uninstallation process. There were no issues
-for all but the bottom four, just various degrees of ease and automation.
-\textit{GATE}, \textit{dwv}, and \textit{DICOM Viewer} showed severe
-installation problems. We were not able to install them, even after a reasonable
-amount of time (2 hours).  For \textit{dwv} and \textit{GATE} we failed to build
-from the source code, but we were able to proceed with measuring other qualities
-using a deployed on-line version for \textit{dwv}, and a virtual machine version for
-\textit{GATE}. For \textit{DICOM Viewer} we could not install the NextCloud
-dependency, and thus could not measure reliability nor robustness for it.
-
-\textit{MatrixUser} depends on Matlab, whose installation is not easy.
-For users who already have Matlab, this score should be higher.
-
-\subsection{Correctness \& Verifiability} \label{sec_result_correctness_verifiability}
-
-The packages with higher scores for correctness and verifiability used a wider array of
-techniques to improve correctness, and had better documentation to witness this.
-For instance, we looked for evidence of unit testing, and found evidence for
-only about half of the projects. We identified five projects using continuous
-integration tools: \textit{3D Slicer}, \textit{ImageJ}, \textit{Fiji},
-\textit{dwv}, and \textit{OHIF Viewer}.
-
-Even for projects with well-organized documentation, requirements
-specifications and theory manuals were still missing.
-The only requirements-related document we found was a road
+The overall score is based on the measurement of the identified qualities. Full
+details of how the projects compare for each quality can be found in [citation
+redacted for blind review].
+%% DOUBLE BLIND
+%Smith et al.~\cite{SmithEtAl2024_MI_SOP}
+The highlights are as follows:
+
+\begin{description}
+
+\item [Installability] We found installation instructions for 16 projects, but
+two did not need them (\textit{BioImage Suite Web} and \textit{Slice:Drop}) as
+they are web applications. 10 of the projects required extra dependencies: Five
+depend on a specific browser; \textit{dwv}, \textit{OHIF Viewer}, and
+\textit{GATE} needs extra libraries to build; \textit{ImageJ} and \textit{Fiji}
+need an unzip tool; \textit{MatrixUser} needs Matlab; \textit{DICOM Viewer}
+needs a Nextcloud platform. \textit{GATE}, \textit{dwv}, and \textit{DICOM
+Viewer} showed severe installation problems. We were not able to install them,
+even after a reasonable amount of time (2 hours).
+
+\item [Correctness \& Verifiability] The packages with higher scores for
+correctness and verifiability used a wider array of techniques to improve
+correctness, and had better documentation to witness this.  For instance, we
+looked for evidence of unit testing, and found evidence for only about half of
+the projects. We identified five projects using continuous integration tools:
+\textit{3D Slicer}, \textit{ImageJ}, \textit{Fiji}, \textit{dwv}, and
+\textit{OHIF Viewer}. The only requirements-related document we found was a road
 map of \textit{3D Slicer}, which contained design requirements for upcoming
@@
 
-\subsection{Surface Reliability} \label{sec_result_reliability}
-
-We were able to follow the steps in the tutorials that existed (seven packages
-had them.) However, \textit{GATE} could not open macro files and became
-unresponsive several times, without any descriptive error message. We found that
-\textit{Drishti} crashed when loading damaged image files, without showing any
-descriptive error message.
-
-\subsection{Surface Robustness} \label{sec_result_robustness}
-
-The packages with higher robustness scores gracefully handled unexpected or unanticipated
-inputs, typically showing a clear error message. We may have underscored
-\textit{OHIF Viewer}, since we needed further customization to load
-data.
-
-According to their documentation, all 29 software packages should support the
-DICOM standard. To test robustness, we prepared two types of image files:
-correct and incorrect formats (with the incorrect format created by relabelling a
-text file to have the ``.dcm'' extension).  All software packages loaded the
-correct format image, except for \textit{GATE}, which failed for unknown
-reasons.  For the broken format, \textit{MatrixUser}, \textit{dwv}, and
-\textit{Slice:Drop} ignored the incorrect format,
-did not show any error message and displayed a blank image.
-\textit{MRIcroGL} behaved similarly except that it showed a meaningless image.
-\textit{Drishti} successfully detected the broken format of the file, but the
-software crashed as a result.
-
-\subsection{Surface Usability} \label{sec_result_usability}
-
-The software with higher usability scores usually provided both comprehensive
-documented guidance and a good user experience. \textit{INVESALIUS 3} provided
-an excellent example of a detailed and precise user manual. \textit{GATE} also
-provided numerous documents, but unfortunately we had difficulty understanding
-and using them. We found getting started tutorials for only 11 projects, but a
-user manual for 22 projects. \textit{MRIcroGL} was the only project that
-explicitly documented expected user characteristics.
-
-\subsection{Maintainability} \label{sec_score_maintainability}
-
-We gave \textit{3D Slicer} the highest score for maintainability because we
-found it had the most comprehensive artifacts. Only a few of the 29 projects had
-a product, developer's manual, or API (Application Programming Interface)
-documentation, and only \textit{3D Slicer}, \textit{ImageJ}, \textit{Fiji}
-included all three documents (see Table~\ref{tab_maintainability_docs} for the
-full data).  Moreover, \textit{3D Slicer} has a much higher percentage of closed
-issues (92\%) compared to \textit{ImageJ} (52\%) and \textit{Fiji} (64\%).
-
-\begin{table*}[ht]
-\centering
-\begin{tabular}{lccc}
-\toprule
-\multicolumn{1}{c}{Software} & Prod.\ Roadmap & Dev.\ Manual & API Doc. \\ 
-\midrule
-3D Slicer & \checkmark & \checkmark & \checkmark \\
-ImageJ & \checkmark & \checkmark & \checkmark \\
-Weasis &  & \checkmark &  \\
-OHIF Viewer &  & \checkmark & \checkmark \\
-Fiji & \checkmark & \checkmark & \checkmark \\
-ParaView & \checkmark &  &  \\
-SMILI &  &  & \checkmark \\
-medInria &  & \checkmark &  \\
-INVESALIUS 3 & \checkmark &  &  \\
-dwv &  &  & \checkmark \\
-BioImage Suite Web &  & \checkmark &  \\
-Gwyddion &  & \checkmark & \checkmark \\ 
-\bottomrule
-\end{tabular}
-\caption{Software with the maintainability documents (listed in descending order of 
-maintainability score)}
-\label{tab_maintainability_docs}
-\end{table*}
-
-Twenty-seven of the 29 projects used git for version control, with 24 of
-these using GitHub. \textit{AMIDE} used Mercurial and \textit{Gwyddion} used
-Subversion. \textit{XMedCon}, \textit{AMIDE}, and \textit{Gwyddion} used
-SourceForge. \textit{DicomBrowser} and \textit{3DimViewer} used BitBucket.
-
-\subsection{Reusability} \label{sec_result_reusability}
-
-We have assumed that smaller code files are
+\item [Surface Reliability] We were able to follow the steps in the tutorials
+that existed (seven packages had them.) However, \textit{GATE} could not open
+macro files and became unresponsive several times, without any descriptive error
+message. We found that \textit{Drishti} crashed when loading damaged image
+files, without showing any descriptive error message.
+
+\item [Surface Robustness] According to their documentation, all 29 software
+packages should support the DICOM standard. To test robustness, we prepared two
+types of image files: correct and incorrect formats (with the incorrect format
+created by relabelling a text file to have the ``.dcm'' extension).  All
+software packages loaded the correct format image, except for \textit{GATE},
+which failed for unknown reasons.  For the broken format, \textit{MatrixUser},
+\textit{dwv}, and \textit{Slice:Drop} ignored the incorrect format, did not show
+any error message and displayed a blank image.  \textit{MRIcroGL} behaved
+similarly except that it showed a meaningless image.  \textit{Drishti}
+successfully detected the broken format of the file, but the software crashed as
+a result.
+
+\item [Surface Usability] The software with higher usability scores usually
+provided both comprehensive documented guidance and a good user experience.
+\textit{INVESALIUS 3} provided an excellent example of a detailed and precise
+user manual. \textit{GATE} also provided numerous documents, but unfortunately
+we had difficulty understanding and using them. We found getting started
+tutorials for only 11 projects, but a user manual for 22 projects.
+\textit{MRIcroGL} was the only project that explicitly documented expected user
+characteristics.
+
+\item [Maintainability] We gave \textit{3D Slicer} the highest score for
+maintainability because we found it had the most comprehensive artifacts. Only a
+few of the 29 projects had a product, developer's manual, or API (Application
+Programming Interface) documentation, and only \textit{3D Slicer},
+\textit{ImageJ}, \textit{Fiji} included all three documents.  Moreover,
+\textit{3D Slicer} has a much higher percentage of closed issues (92\%) compared
+to \textit{ImageJ} (52\%) and \textit{Fiji} (64\%). Twenty-seven of the 29
+projects used git for version control, with 24 of these using GitHub.
+\textit{AMIDE} used Mercurial and \textit{Gwyddion} used Subversion.
+\textit{XMedCon}, \textit{AMIDE}, and \textit{Gwyddion} used SourceForge.
+\textit{DicomBrowser} and \textit{3DimViewer} used BitBucket.
+
+\item [Reusability] We have assumed that smaller code files are
 likely more reusable -- see Table \ref{tab_loc_per_file} for the details.
 
-\begin{table*}[ht]
+\item [Surface Understandability] All projects had a consistent coding style
+with parameters in the same order for all functions, modularized code, and,
+clear comments that indicate what is done, not how. However, we only found
+explicit identification of a coding standard for 3 out of the 29: \textit{3D
+Slicer}, \textit{Weasis}, and \textit{ImageJ}. We also found hard-coded
+constants (rather than symbolic constants) in \textit{medInria},
+\textit{dicompyler}, \textit{MicroView}, and \textit{Papaya}. We did not find
+any reference to the algorithms used in projects \textit{XMedCon},
+\textit{DicomBrowser}, \textit{3DimViewer}, \textit{BioImage Suite Web},
+\textit{Slice:Drop}, \textit{MatrixUser}, \textit{DICOM Viewer},
+\textit{dicompyler}, and \textit{Papaya}.
+
+\item [Visibility/Transparency] Generally speaking, the teams that actively
+documented their development process and plans scored higher.  \textit{3D
+Slicer} and \textit{ImageJ} were the only projects to include documentation for
+all of the following: development process, project status, development
+environment and release notes.
+
+\end{description}
+
+\begin{table*}[!ht]
 \centering
@@
 
-\subsection{Surface Understandability} \label{sec_result_understandability}
-
-All projects had a consistent coding style with parameters in
-the same order for all functions, modularized code, and, clear comments that
-indicate what is done, not how. However, we only found explicit identification
-of a coding standard for 3 out of the 29: \textit{3D Slicer}, \textit{Weasis},
-and \textit{ImageJ}. We also found hard-coded constants (rather than symbolic
-constants) in \textit{medInria}, \textit{dicompyler}, \textit{MicroView}, and
-\textit{Papaya}. We did not find any reference to the algorithms used in
-projects \textit{XMedCon}, \textit{DicomBrowser}, \textit{3DimViewer},
-\textit{BioImage Suite Web}, \textit{Slice:Drop}, \textit{MatrixUser},
-\textit{DICOM Viewer}, \textit{dicompyler}, and \textit{Papaya}.
-
-\subsection{Visibility/Transparency} \label{sec_result_visibility_transparency}
-
-Generally speaking, the teams that actively documented
-their development process and plans scored higher.
-Table~\ref{tab_Visibility/Transparency_docs} shows the projects that had
-documents for the development process, project status, development environment,
-and release notes.
-
-\begin{table*}[!ht]
-\centering
-\begin{tabular}{lllll}
-\toprule
-Software & Dev.\ Process & Proj.\ Status & Dev.\ Env. & Rls.\ Notes \\ 
-\midrule
-3D Slicer & \checkmark & \checkmark & \checkmark & \checkmark \\
-ImageJ & \checkmark & \checkmark & \checkmark & \checkmark \\
-Fiji & \checkmark & \checkmark & \checkmark &  \\
-MRIcroGL &  &  &  & \checkmark \\
-Weasis &  &  & \checkmark & \checkmark \\
-ParaView &  & \checkmark &  &  \\
-OHIF Viewer &  &  & \checkmark & \checkmark \\
-DICOM Viewer &  &  & \checkmark & \checkmark \\
-medInria &  &  & \checkmark & \checkmark \\
-SMILI &  &  &  & \checkmark \\
-Drishti &  &  &  & \checkmark \\
-INVESALIUS 3 &  &  &  & \checkmark \\
-OsiriX Lite &  &  &  & \checkmark \\
-GATE &  &  &  & \checkmark \\
-MicroView &  &  &  & \checkmark \\
-MatrixUser &  &  &  & \checkmark \\
-BioImage Suite Web &  &  & \checkmark &  \\
-ITK-SNAP &  &  &  & \checkmark \\
-Horos &  &  &  & \checkmark \\
-dwv &  &  &  & \checkmark \\
-Gwyddion &  &  &  & \checkmark \\ 
-\bottomrule
-\end{tabular}
-\caption{Software with visibility/transparency related documents (listed in
-descending order of visibility/transparency score)}
-\label{tab_Visibility/Transparency_docs}
-\end{table*}
-
-\section{Discussion}
-
-We first compare our ranking to a (proxy for) the community's ranking. We then
-compare the state of the practice for MI with that of other research software.
-In particular we provide details on recommended artifacts that are rarely observed
-for MI software. [other new sections] Section~\ref{sec_threats_to_validity} presents threats to the
-validity of our data and conclusions.
-
-\subsection{Comparison to Community Ranking} \label{Sec_VsCommunityRanking}
-
-We use GitHub stars, number of forks and number of people watching the projects
-are proxies for community ranking -- see Table~\ref{tab_ranking_vs_GitHub}
-for statistics collected in July 2021.  Recall that 24 projects use GitHub.
-Our ranking and GitHub popularity, at least for the top five projects,
-seems to line up fairly well.
-
-We ranked some popular packages fairly low, such as \textit{dwv}. This is
-because we were unable to build it locally, even though we followed its
-installation instructions. However, we were able to use its web
-version for the rest of the measurements. Additionally, this version 
-did not detect a broken DICOM file and instead displayed a blank image
-(Section~\ref{sec_result_robustness}).
-\textit{DICOM Viewer} ranked low as we were unable to install the
-NextCloud platform.
+\section{Comparison to Community Ranking} \label{Sec_VsCommunityRanking}
+
+To answer \rqref{RQ_CompareHQ2Popular} we use GitHub stars, number of forks and
+number of people watching the projects are proxies for community ranking.
+Table~\ref{tab_ranking_vs_GitHub} show the statistics for data collected in July
+2021.  Recall that 24 projects use GitHub (Section~\ref{SecMeasurementResults}).
+
+Our ranking and GitHub popularity, at least for the top five projects, seems to
+line up fairly well. However, we ranked some popular packages fairly low, such
+as \textit{dwv}. This is because we were unable to build it locally, even though
+we followed its installation instructions. However, we were able to use its web
+version for the rest of the measurements. Additionally, this version did not
+detect a broken DICOM file and instead displayed a blank image
+(Section~\ref{SecMeasurementResults}).  \textit{DICOM Viewer} ranked low as we
+were unable to install the NextCloud platform.
 
@@
 
-\subsection{Software Artifacts}
-\label{Sec_CompareArtifacts}
-
-We use nine research software development guidelines to compare recommended 
-software artifacts versus those present in MI software.
-These guidelines are:
+\section{Compare to Recommended Software Artifacts} \label{Sec_CompareArtifacts}
+
+For \rqref{RQ_CompareArtifacts} we use nine research software development
+guidelines to compare recommended software artifacts versus those present in MI
+software.  These guidelines are:
 \begin{itemize}
@@
 
-\begin{table*}[ht]
-%\begin{center}
-\begin{tabular}{ p{2.5cm}p{0.5cm}p{0.5cm}p{0.5cm}p{0.5cm}p{0.5cm}p{0.5cm}p{0.5cm}p{0.5cm}p{0.5cm}p{0.5cm} }
+\begin{table*}[!ht]
+\begin{center}
+\begin{tabular}{ p{3.5cm}p{0.5cm}p{0.5cm}p{0.5cm}p{0.5cm}p{0.5cm}p{0.5cm}p{0.5cm}p{0.5cm}p{0.5cm}p{0.5cm} }
 \toprule
@@
 \label{Tbl_Guidelines}
-%\end{center}
+\end{center}
 \end{table*}
@@
 
-\begin{table*}[ht!]
-    %\begin{center}
-    \begin{tabular}{ p{3.1 cm} p{5.4 cm} p{3.5 cm}}
+\begin{table*}[!ht]
+    \begin{center}
+    \begin{tabular}{ p{3.8 cm} p{6.4 cm} p{4.5 cm}}
     \toprule
@@
     User manual (22) & Authors (14) & Dependency list (7)\\
-    Release info. (22) & Frequently Asked Questions (FAQ) (14) & Troubleshooting guide (6)\\
+    Release info. (22) & Frequently Asked Questions (14) & Troubleshooting guide (6)\\
      & Acknowledgements (12) & Product roadmap (5)\\
@@
     \label{artifactspresent}
-    %\end{center}
+    \end{center}
 \end{table*}
@@
 (33-67\%) packages), and rare (1 to 9 ($<$33\%) packages).
-Tables~\ref{tab_maintainability_docs} and~\ref{tab_Visibility/Transparency_docs}
-show the details on which projects use which types of artifacts for documents
-related to maintainability and visibility, respectively.
 
@@
 
-\subsection{Interview Methods} \label{sec_interview_methods}
-
-The repository-based measurements summarize the information we can collect from
-on-line resources. This information is incomplete because it doesn't generally
-capture the development process, the developer pain points, the perceived
-threats to software quality, and the developers' strategies to address these
-threats.  Therefore, part of our methodology involves interviewing developers.
-
-We based our interviews on a list of 20 questions, which can be found in
-\cite{SmithEtAl2021}. Some questions are about the background of the software,
-the development teams, the interviewees, and how they organize their projects.
-We also ask about the developer's understanding of the users. Some questions
-focus on the current and past difficulties, and the solutions the team has
-found, or plan to try. We also discuss documentation, both with respect to how
-it is currently done, and how it is perceived. A few questions are about
-specific software qualities, such as maintainability, understandability,
-usability, and reproducibility. The interviews are semi-structured based on the
-question list; we ask follow-up questions when necessary.  The interview process
-presented here was approved by the McMaster University Research Ethics Board
-under the application number
-\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/MACREM/Application.pdf}
-{MREB\#: 5219}.
-
-We sent interview requests to all 29 projects using contact information from
-projects websites, code repository, publications, and from biographic pages at
-the teams' institutions.  In the end nine developers from eight of the projects
-agreed to participate: \textit{3D Slicer}, \textit{INVESALIUS 3}, \textit{dwv},
-\textit{BioImage Suite Web}, \textit{ITK-SNAP}, \textit{MRIcroGL},
-\textit{Weasis}, and \textit{OHIF}. We spent about 90 minutes for each
-interview. One participant was too busy to have an interview, so they wrote down
-their answers. In one case two developers from the same project agreed to be
-interviewed. We held the meetings on-line using either Zoom or Teams, which
-facilitated recording and automatic transcription. The full interview answers
-can be found in \cite{Dong2021}.
-
 \section{Comparison of Tool Usage Between MI and Other Research Software}
@@
 
-Developers use software tools to support the development, verification,
-maintenance, and evolution of software, software processes, and artifacts
-\cite[p.\ 501]{GhezziEtAl2003}. MI software uses tools for CI/CD, user support,
-version control, documentation, and project management.  To answer
-\rqref{RQ_CompareToolsProjMngmnt} we summarize the tool usage in these
+We use data from interviews, as described in Section~\ref{sec_interview_methods}
+to answer \rqref{RQ_CompareToolsProjMngmnt}, \rqref{RQ_CompareMethodologies}
+and \rqref{RQ_PainPoints}. Requests were sent to all 29 projects.  Nine
+developers from eight of the projects agreed to participate: \textit{3D Slicer},
+\textit{INVESALIUS 3}, \textit{dwv}, \textit{BioImage Suite Web},
+\textit{ITK-SNAP}, \textit{MRIcroGL}, \textit{Weasis}, and \textit{OHIF}. We
+spent about 90 minutes for each interview. The full interview answers can be
+found in [citation redacted for blind review].
+%\cite{Dong2021}.
+
+Developers use software tools for user support, version control, Continuous
+Integration and Deployment (CI/CD), documentation and project management. To
+answer \rqref{RQ_CompareToolsProjMngmnt}, we summarize the tool usage in these
 categories, and compare this to the usage by the research software community.
@@
 number of projects using each model (projects may use more than one support
-model). We do not know whether the prevalent use of GitHub issues for user
-support is by design, or whether this just naturally happens as users seek
-help. The common use of GitHub by MI developers is not surprising, given that
-GitHub is the largest code host in the world, with over 128 million public
-repositories and over 23 million users (as of roughly February 26, 2020)
-\cite{Kashyap2020}.
+model). The prevalent usage of GitHub issues may be by design, or it might just
+naturally arise as users seek help.
 
@@
 
-From Section~\ref{sec_score_maintainability}, 27 of the 29 projects used git as
-the version control tool, one used Mercurial and one used Subversion.  The
-hosting is on GitHub for 24 packages, SourceForge for three and BitBucket for
-two.  Although teams may have a process for accepting new contributions, no one
-discussed this during their interviews. However, most teams (eight of nine)
-mentioned using GitHub and pull requests to manage contributions from the
-community. The interviewees generally gave very positive feedback on using
-GitHub. Some teams previously used a different approach to version control and
-eventually transferred to git and GitHub.  The past approaches included
-contributions from e-mail (three teams), contributions from forums (one team)
-and e-mailing the git repository back and forth between developers (one team).
+Most teams interview (eight of nine) mentioned using GitHub and pull requests to manage
+contributions from the community. The interviewees generally gave very positive
+feedback on using GitHub. Some teams previously used a different approach to
+version control and eventually transferred to git and GitHub.
 
@@
 matches the increase in the broader research software community.  A little over
-10 years ago \cite{Nguyen-HoanEtAl2010} estimated that only 50\% of research
-software projects use version control, but even at that time
-\cite{Nguyen-HoanEtAl2010} noted an increase from previous usage levels. A
-survey in 2018 shows 81\% of developers use a version control system
-\cite{AlNoamanyAndBorghi2018}. \cite{Smith2018} has similar results, showing
-version control usage for alive projects in mesh generation, geographic
-information systems and statistical software for psychiatry increasing from
-75\%, 89\% and 17\% (respectively) to 100\%, 95\% and 100\% (respectively) over
-a four-year period ending in 2018. (For completeness the same study showed a
-small decrease in version control usage for seismology software over the same
-time period, from 41\% down to 36\%).  A recent survey by \cite{CarverEtAl2022}
-shows version control use among practitioners at over 95\%, with 83/87 survey
-respondents indicating that they use it. All but one of the software guides
-cited in Section~\ref{Sec_CompareArtifacts} includes the advice to use version
-control. (The USGS guide \cite{USGS2019} was the only set of recommendations to
-not mention version control.) The high usage of version control tools in MI
-software matches the trend in research software in general.
-
-As mentioned in Section~\ref{sec_result_correctness_verifiability}, we
-identified five projects using CI/CD tools (about 17\% of the assessed
-projects). We found which projects used CI/CD by examining the documentation and
-source code of all projects. The count of CI/CD usage may actually be higher,
-since traces of CI/CD usage may not always appear in a repository.  This was the
-case for a study of LBM software, where interviews with developers showed that
-more projects used CI/CD than was evident from repository artifacts alone
-\cite{Michalski2021}.  The 17\% utilization for MI software contrasts with the
-high frequency with which research software development guidelines recommend
-continuous integration \cite{BrettEtAl2021, Brown2015, ThielEtAl2020,
-Zadka2018, vanGompelEtAl2016}. Although there is currently little data available
-on CI/CD utilization for research software, our impression is that CI/CD is not
-yet common practice, despite its recommendation.  This is certainly the case for
-LBM software, where usage numbers are similar to MI software, with only 12.5\%
-of 24 LBM packages showing evidence of CI/CD in their repositories
-\cite{Michalski2021}.  The survey of \cite{CarverEtAl2022} suggests higher use
-of CI/CD with 54\% (54/100) of respondents indicating that they use it. However,
-that survey measures something different from the current one by surveying
+10 years ago Nguyen-Hoan et al.\ \cite{Nguyen-HoanEtAl2010} estimated that only
+50\% of research software projects used version control, but even at that time
+there was an increase from previous usage levels. A survey in 2018 shows 81\% of
+developers use a version control system \cite{AlNoamanyAndBorghi2018}.
+\cite{Smith2018} has similar results, showing version control usage for alive
+projects in mesh generation, geographic information systems and statistical
+software for psychiatry increasing from 75\%, 89\% and 17\% (respectively) to
+100\%, 95\% and 100\% (respectively) over a four-year period ending in 2018. A
+recent survey by \cite{CarverEtAl2022} shows version control use among
+practitioners at over 95\%. All but one of the software guides cited in
+Section~\ref{Sec_CompareArtifacts} recommends version control. The high usage of
+version control tools in MI matches the general trend in research software.
+
+As mentioned in Section~\ref{SecMeasurementResults}, we identified 17\% of MI
+projects using CI/CD tools. We found which projects used CI/CD by examining the
+documentation and source code. The count of CI/CD usage may be higher, since
+traces of CI/CD may not always appear in a repository.  This was the case for a
+study of Lattice Boltzmann Method (LBM) software, where interviews with
+developers showed that more projects used CI/CD than was evident from repository
+artifacts alone [citation redacted for blind review].  %\cite{Michalski2021}.  
+The 17\% utilization for MI software contrasts with the high frequency with
+which research software development guidelines recommend continuous integration
+\cite{BrettEtAl2021, Brown2015, ThielEtAl2020, Zadka2018, vanGompelEtAl2016}.
+Although there is currently little data available on CI/CD utilization for
+research software, our impression is that CI/CD is not yet common practice,
+despite its recommendation.  This is certainly the case for LBM software, where
+usage numbers sit around 12.5\% [citation redacted for blind review]
+%\cite{Michalski2021}.  
+The survey of \cite{CarverEtAl2022} suggests higher use of CI/CD with 54\%
+(54/100) of respondents indicating that they use it. However, that study surveys
 practitioners, rather than assessing projects.  Additional information on CI/CD
@@
 documentation tool difference is that LBM software often uses document
-generation tools, like doxygen and sphinx \cite{Michalski2021}, while MI does
-not appear to use these tools.
-
-Some interviewees mentioned the project management tools they used. Generally
-speaking, the interviewees talked about two types of tools:
+generation tools, like doxygen and sphinx [citation redacted for blind review].
+%\cite{Michalski2021}
+
+The interviewees talked about two types of project management tools:
 \begin{inparaenum}[i)]
@@
 \end{inparaenum}
-Of the specifically named tools in the above lists, interviewees mentioned
-GitHub 3 times, and each of the other tools once each.
+Of the specifically named tools, interviewees mentioned GitHub 3 times, and each
+of the other tools once each.
 
@@
 software \cite{Michalski2021} and is apparently greater than the level of
-testing for ocean modelling software.  \cite{JungEtAl2022} reports that ocean
-modellers underemphasize testing.
+testing for ocean modelling software \cite{JungEtAl2022}.
 
@@
 
-\begin{table*}[ht!]
+\begin{table*}[!ht]
 	%\begin{center}

### 2024-10-29 StateOfPractice/Papers/SOP_MI_CritReviewsBioMedEng/SOP_MI_OSR.tex
Updates to Section 7

@@
 Our observations of an informally defined process, with elements of agile
-methods, matches what has been observed for research software in general.
+methods, matches observations for research software in general.
 Scientific developers naturally use an agile philosophy \cite{AckroydEtAl2008,
@@
 most common approach was following the issues, such as bugs and feature
-requests. Additionally, the \textit{3D Slicer} team had weekly meetings to
-discuss the goals for the project; the \textit{INVESALIUS 3} team relied on the
-GitHub process for their project management; the \textit{ITK-SNAP} team had a
-fixed six-month release pace; only the interviewee from the \textit{OHIF} team
-mentioned that the team has a project manager; the \textit{3D Slicer} team and
+requests. The \textit{3D Slicer} team had weekly meetings to
+discuss the goals for the project. The \textit{INVESALIUS 3} team relied on the
+GitHub process for their project management while the \textit{ITK-SNAP} team had a
+fixed six-month release pace. Only the interviewee from \textit{OHIF} 
+mentioned the team having a project manager. The \textit{3D Slicer} team and
 \textit{BioImage Suite Web} team do nightly builds and tests. The \textit{OHIF}
@@
 
-We identified the use of unit testing in less than half of the 29 projects. On
-the other hand, the interviewees believed that testing (including usability
-tests with users) was the top solution to improve correctness, usability, and
+We identified the use of unit testing in less than half of the 29 projects even
+though the interviewees believed that testing (including usability tests with
+users) was the top approach to improve correctness, usability, and
 reproducibility.  This level of testing matches what was observed for LBM
-software \cite{Michalski2021} and is apparently greater than the level of
+software [citation redacted for blind review]
+%\cite{Michalski2021} 
+and is apparently greater than the level of
 testing for ocean modelling software \cite{JungEtAl2022}.
 
-As the observed artifacts in Table~\ref{artifactspresent} show, none of the 29
-projects emphasize documentation. None of them had theory manuals, although we
-did identify a road map in the \textit{3D Slicer} project.  We did not find
-requirements specifications. Table~\ref{tab_opinion_doc} summarizes
-interviewees' opinions on documentation. Interviewees from each of the eight
-projects thought that documentation was essential to their projects, and most of
-them said that it could save their time to answer questions from users and
-developers. Most of them saw the need to improve their documentation, and only
-three of them thought that their documentations conveyed information clearly
-enough. Nearly half of developers also believed that the lack of time prevented
-them from improving documentation.
-
-\begin{table}[!ht]
-\centering
-\begin{tabular}{ll}
-\toprule
-Opinion on Documentation & Num.\ Ans. \\ 
-\midrule
-Documentation is vital to the project & 8 \\
-Documentation of the project needs improvements & 7 \\
-Referring to documentation saves time to answer questions & 6 \\
-Lack of time to maintain good documentation & 4 \\
-Documentation of the project conveys information clearly & 3 \\
-Coding is more fun than documentation & 2 \\
-Users help each other by referring to documentation & 1 \\ 
-\bottomrule
-\end{tabular}
-\caption{Opinions on documentation by the numbers of interviewees with the
-answers}
-\label{tab_opinion_doc}
-\end{table}
+Interviewees from all of the eight projects thought that documentation was
+essential to their projects, and most of them (7 of 8) said that it could save
+their time to answer questions from users and developers. Half of them (4 of 8)
+saw the need to improve their documentation, and only three of them thought that
+their documentations conveyed information clearly enough.
 

### 2024-10-31 StateOfPractice/Papers/SOP_MI_CritReviewsBioMedEng/SOP_MI_OSR.tex
Complete draft of the paper

@@
 named CONTRIBUTING. Guidelines also recommend tutorials, user guides and quick
-start examples. \cite{SmithAndRoscoe2018} suggests including instructions
+start examples. Smith and Roscoe \cite{SmithAndRoscoe2018} suggest including instructions
 specifically for on-boarding new developers. For open-source software in general
@@
 observations from other domains.  We also cover potential ways to address the
-pain points, as promoted by the community.  (Later, in
-Section~\ref{ch_recommendations}, we propose additional pain mitigation
-strategies based on our experience.)  In addition to pain points, we summarize
+pain points, as promoted by the community.  In addition to pain points, we summarize
 MI developer strategies for improving maintainability and reproducibility.
@@
 
-\cite{PintoEtAl2018} lists some pain points that did not come up in our
-conversations with MI developers: interruptions while coding, scope bloat, lack
-of user feedback, hard to collaborate on software projects, and aloneness.
-\cite{WieseEtAl2019} also mention two research software pain points that did
+Pinto et al.\ \cite{PintoEtAl2018} list some pain points that did not come up in
+our conversations with MI developers: interruptions while coding, scope bloat,
+lack of user feedback, hard to collaborate on software projects, and aloneness.
+Wiese et al.\ \cite{WieseEtAl2019} mention two additional pain points that did
 not explicitly arise in our interviews: reproducibility, and software scope
 determination.  To the list of pain points not discussed for MI, our study of
-LBM software \cite{SmithEtAl2024} adds lack of software experience for the
-developers, technical debt, and documentation. We did not observe any pain
-points for MI that were not also observed for LBM. From the pain points
-mentioned above, although the topics of reproducibility and technical debt did
-not come up in our MI interviews, we covered these two topics as part of the
-discussion of software qualities, as summarized at the end of this section.
-Although previous studies show pain points that were not mentioned by MI
-developers, we cannot conclude that these pain points are not relevant for MI
-software development, since we only interviewed nine developers for about an
-hour each.
+LBM software [citation redacted for blind review] %\cite{SmithEtAl2024} 
+adds lack of software experience for the developers, technical debt, and
+documentation. We did not observe any pain points for MI that were not also
+observed for LBM. From the pain points mentioned above, although the topics of
+reproducibility and technical debt did not come up in our MI interviews, we
+covered these two topics as part of the discussion of software qualities, as
+summarized at the end of this section.  Although previous studies show pain
+points that were not mentioned by MI developers, we cannot conclude that these
+pain points are not relevant for MI software development, since we only
+interviewed nine developers.
 
@@
 of research software also experience the lack of time pain point
-\cite{PintoEtAl2018, PintoEtAl2016, WieseEtAl2019}. Our study of LBM software
-\cite{SmithEtAl2024} also highlighted lack of time as a significant pain point.
+\cite{PintoEtAl2018, PintoEtAl2016, WieseEtAl2019}. Our study of LBM software [citation redacted for blind review]
+%\cite{SmithEtAl2024} 
+also highlighted lack of time as a significant pain point.
 
@@
 count software when they are judging the academic excellence of the applicant.
-\cite{WieseEtAl2019} reported developer pains related to publicity, since
-publishing norms have historically made it difficult to get credit for creating
-software.  As studied by \cite{HowisonAndBullard2016}, research software
-(specifically biology software, but the trend likely applies to other research
-software domains) is infrequently cited. \cite{PintoEtAl2018} also mentions the
-lack of formal reward system for research software.
+Wiese et al.\ \cite{WieseEtAl2019} reported developer pains related to
+publicity, since publishing norms have historically made it difficult to get
+credit for creating software.  As studied by \cite{HowisonAndBullard2016},
+research software (specifically biology software, but the trend likely applies
+to other research software domains) is infrequently cited. There is a lack of a
+formal reward system for research software \cite{PintoEtAl2018}.
 
@@
 
-The pain point survey of \cite{WieseEtAl2019} highlights that technology
+Another pain point survey \cite{WieseEtAl2019} highlights that technology
 hurdles are an issue for research software in general.  Some technical-related
-problems mentioned by \cite{WieseEtAl2019} include dependency management,
-cross-platform compatibility (also mentioned by \cite{PintoEtAl2018}), CI,
-hardware issues and operating system issues. From \cite{SmithEtAl2024}
-technology pain points for LBM developers include setting up parallelization and
-CI.
-
-The solutions proposed by the MI developers include the following:
+problems mentioned include dependency management, cross-platform compatibility,
+CI, hardware issues and operating system issues \cite{WieseEtAl2019,
+PintoEtAl2018}. Technology pain points for LBM developers include setting up
+parallelization and CI [citation redacted for blind review].
+%\cite{SmithEtAl2024}.
+
+Five solutions proposed by the MI developers are:
 
@@
 As the above list shows, developers perceive that web-based applications will
-address the technology hurdle.  Table~\ref{tab_native_vs_web} shows the teams'
-choices between native application and web application. Most of the 29 teams (24
-of 29, or 83\%) chose to develop native applications. For the eight teams we
-interviewed, three of them were building web applications, and the
-\textit{MRIcroGL} team was considering a web-based solution.
-
-\begin{table}[!ht]
-\centering
-\begin{tabular}{lll}
-\toprule
-Software Team & Native Application & Web Application \\ 
-\midrule
-3D Slicer & \checkmark & \\
-INVESALIUS 3 & \checkmark & \\
-dwv & & \checkmark \\
-BioImage Suite Web & & \checkmark \\
-ITK-SNAP & \checkmark & \\
-MRIcroGL & \checkmark & \\
-Weasis & \checkmark & \\
-OHIF & & \checkmark \\ 
-\midrule
-Total number among the eight teams & 5 & 3 \\
-Total number among the 29 teams & 24 & 5 \\ 
-\bottomrule
-\end{tabular}
-\caption{Teams' choices between native application and web application}
-\label{tab_native_vs_web}
-\end{table}
+address the technology hurdle.  Most of the 29 teams (83\%) chose to develop
+native applications. For the eight teams we interviewed, three of them were
+building web applications, and the \textit{MRIcroGL} team was considering a
+web-based solution.
 
@@
 \item Lack of real world image data for testing, in part because of patient
-privacy concerns (\cite{WieseEtAl2019} mentions that the pain point of privacy
-concerns also arises for research software in general);
+privacy concerns (the pain point of privacy concerns also arises for research
+software in general \cite{WieseEtAl2019});
 \item Tests are expensive and time-consuming because of the need for huge datasets;
@@
 automated tests.  With the common emphasis on testing to improve correctness, MI
-software is ahead of some other scientific domains.  For scientific software in
-general \cite{PintoEtAl2018} mention the problem of insufficient testing and
-\cite{HannayEtAl2009} show that more developers think testing is important than
-the number that believe they have a sufficient understanding of testing
-concepts.  Our study of LBM software suggests that this domain shares the
-challenges of insufficient testing and insufficient understanding of testing
-concepts \cite{SmithEtAl2024}. Automated testing is a specific challenge for
-LBM software since free testing services do not offer adequate facilities for
-large amounts of data \cite{SmithEtAl2024}. Although not specifically mentioned
-during our interviews, the large data sets for MI likely also cause a challenge
-for using free testing services, like GitHub Actions.
+software is ahead of some other scientific domains.  For research software in
+general, there is a problem of insufficient testing \cite{PintoEtAl2018} and
+more developers think testing is important than the number that believe they
+have a sufficient understanding of testing concepts \cite{HannayEtAl2009}.  Our
+study of LBM software suggests that this domain shares the challenges of
+insufficient testing and insufficient understanding of testing concepts
+[citation redacted for double blind]. %\cite{SmithEtAl2024}.  
+Automated testing
+is a specific challenge for LBM software since free testing services do not
+offer adequate facilities for large amounts of data.
+Although not specifically mentioned during our interviews, the large data sets
+for MI likely also cause a challenge for using free testing services, like
+GitHub Actions.
 
@@
 
-\item[P\refstepcounter{pnum}\thepnum \label{P_Usability}:]
-\textbf{Usability:}
-
+\item[P\refstepcounter{pnum}\thepnum \label{P_Usability}:] \textbf{Usability:}
 The discussion with the developers focused on usability issues for two classes
@@
 users include an unintuitive user interface, inadequate feedback from the
-interface (such as lack of a progress bar), users being unable to determine the purpose of
-the software, not all users knowing if the software includes certain features, not
-all users understanding how to use the command line tool, and not all users
-understanding that the software is a web application. For developers the threats to
-usability include not being able to find clear instructions on how to deploy the
-software, and the architecture being difficult for new developers to understand.
+interface (such as lack of a progress bar), users being unable to determine the
+purpose of the software, not all users knowing if the software includes certain
+features, not all users understanding how to use the command line tool, and not
+all users understanding that the software is a web application. For developers
+the threats to usability include not being able to find clear instructions on
+how to deploy the software, and the architecture being difficult for new
+developers to understand.
 
 At least to some extent the problems for MI software users are due to holes in
-their background knowledge.  The survey of \cite{WieseEtAl2019} for research
-software in general also mentioned that users do not always have the expertise
-required to install or use the software. \cite{SmithEtAl2024} observes a
-similar pattern for LBM software, with several LBM developers noting that users
-sometimes try to use incorrect method combinations. Furthermore, some LBM users
-think that the packages will work out of the box to solve their cases, while in
-reality computational fluid dynamics knowledge needs to be applied to correctly
-modify the packages for a new endeavour.
+their background knowledge.  For research software in general, users do not
+always have the expertise required to install or use the software
+\cite{WieseEtAl2019}. A similar pattern has been observed during interviews with
+LBM software developers [citation redacted for blind review].
+%\cite{SmithEtAl2024}.
 
@@
 
-\begin{enumerate}
-\item[Q\refstepcounter{qnum}\theqnum \label{Q_Maintainability}:]
-\textbf{Maintainability:} \cite{Nguyen-HoanEtAl2010} rate maintainability as the
-third most important software quality for research software in general. The push
-for sustainable software \cite{deSouzaEtAl2019} is motivated by the pain that
-past developers have had with accumulating too much technical debt
-\cite{KruchtenEtAl2012}.  For LBM software, \cite{SmithEtAl2024} identifies
-technical debt as one of the developer pain points.
+\begin{enumerate} \item[Q\refstepcounter{qnum}\theqnum
+\label{Q_Maintainability}:] \textbf{Maintainability:}  Maintainability has been
+rated as the third most important software quality for research software
+\cite{Nguyen-HoanEtAl2010}. The push for sustainable software
+\cite{deSouzaEtAl2019} is motivated by the pain that past developers have had
+with accumulating too much technical debt \cite{KruchtenEtAl2012}.  For LBM
+software, technical debt has been identified as one of the developer pain points
+[citation redacted for blind review]. %\cite{SmithEtAl2024}.
 
@@
 documentation.  The \textit{3D Slicer} team used a well-defined structure for
-the software, which they named as an ``event-driven MVC pattern''. Moreover,
-\textit{3D Slicer} discovers and loads necessary modules at runtime, according
-to the configuration and installed extensions. The \textit{BioImage Suite Web}
-team had designed and re-designed their software multiple times in the last 10+
-years. They found that their modular approach effectively supports
+the software, which they named as an ``event-driven MVC pattern''.  \textit{3D
+Slicer} discovers and loads necessary modules at runtime. The \textit{BioImage
+Suite Web} team had designed and re-designed their software multiple times in
+the last 10+ years. They found that their modular approach effectively supports
 maintainability \cite{Joshi2011}.
@@
 (Section~\ref{painpoints}), including continuous integration, moving to web
-applications, and enriching the test data sets. We also bring in new ideas from
-our experience like employing linters, peer review, design for change and
-assurance cases.  Our aim is to mention ideas that are at least somewhat beyond
-conventional best practices. The ideas listed here have the potential to become
-best practices in the medium to long-term. We list the ideas roughly in the
-order of increasing implementation effort.
+applications, and enriching the test data sets. We also bring in new like
+employing linters, peer review, and design for change.  Our aim is to mention
+ideas that are at least somewhat beyond conventional best practices. The ideas
+listed here have the potential to become best practices in the medium to
+long-term. We list the ideas roughly in the order of increasing implementation
+effort.
 
@@
 significant time and effort to set up and integrate into a team's workflow, but
-the benefits are significant, as follows:
+the benefits are impressive, as follows:
 
@@
 	\item Elimination of headaches associated with a separate integration phase
-	\cite{Fowler2006}, \cite[p.\ 20]{HumbleAndFarley2010}. If developers
-	postpone integration, integration problems are inevitable.  Continuous
-	integration means that problems are immediately obvious and the source of
-	the problem can be isolated to the small increment that was just committed.
+	\cite{Fowler2006}, \cite[p.\ 20]{HumbleAndFarley2010}.
 	\item Detection and removal of bugs \cite{Fowler2006} via
-	automated testing.  To improve productivity, defects are best discovered and
-	fixed at the point where they are introduced \cite[p.\
-	23]{HumbleAndFarley2010}.  Code is not the only source of errors; they are
-	also found in the files and scripts related to configuration management
-	\cite[p.\ 18]{HumbleAndFarley2010}.
+	automated testing.  
 	\item Everyone is always working on a stable base, since the rejection of
@@
 
-CI consists of the following elements:
-
-\begin{itemize}
-	\item A version control system \cite{Fowler2006}. To be effective, all
-	files should be under version control, not just code files.  Anything that
-	is needed to build, install and run the software should be under version
-	control, including configuration files, build scripts, test harnesses, and
-	operating system configuration files \cite[p. 19]{HumbleAndFarley2010}.
-	Fortunately for the MI, as shown in Section~\ref{sec_score_maintainability}
-	all our measured projects use version control.
-	\item A fully automated build system \cite{Fowler2006}.  As \cite[p.\
-	5]{HumbleAndFarley2010} point out, deploying software manually is an
-	anti-pattern.  For MI software, Table~\ref{artifactspresent} shows 18 of 29
-	packages (62\%) were observed to include build scripts.  Projects without a
-	build system will need to add one to pursue using CI.
-	\item An automated test system \cite{Fowler2006}. Building quality software
-	involves creating automated tests at the unit, component, and acceptance
-	test level, and executing these tests whenever someone makes a change to the
-	code, its configuration, the environment, or the software stack that it runs
-	on \cite[p.\ 83]{HumbleAndFarley2010}. As Table~\ref{artifactspresent}
-	shows, test cases are in the uncommon category for MI software artifacts,
-	which means that some MI projects will need to increase their testing
-	automation if they wish to pursue CI.
-	\item An automated system for other tasks, such as code checking,
-	documentation building and web-site updating.  These other tasks are not
-	essential to CI, but they can be incorporated to improve the quality
-	of the code and the communication between developers and users. For
-	instance, a static analysis (possibly via linters) of the code may find poor
-	programming practice or lack of adherence to adopted coding standards.
-	\item An integrated build system to pull everything together.  Every time
-	there is a check-in (for instance a pull request), the integration server
-	automatically checks out the sources onto the integration machine, starts a
-	build, runs tests, and informs the committer of the results. 
-\end{itemize}
-
-To enable incorporation into a team's workflow, \cite[p.\
-60]{HumbleAndFarley2010} explain that the usual approach for CI is to keep the
-build and test process short. Since MI files are large, the tests run with every
-check-in may need to focus on simple code interface tests, saving large tests
-for less frequent execution.  A more sophisticated option to address the
-bottleneck for merges is CIVET (Continuous Integration, Verification,
-Enhancement, and Testing), which solves this problem by intelligently pinning,
-cancelling, and if necessary, restarting jobs as merges occur
-\cite{SlaughterEtAl2021}. A more sophisticated process management system can
-also enforce rules for pull requests, like checking that a test specification
-includes the test's motivation, a test description, and a design description for
-all changes \cite{SlaughterEtAl2021}.
-
-Setting up a CI system has never been easier than it is today.  A dedicated CI
-server (either physically or virtually) can be installed with tools such as
-\href{https://www.jenkins.io/} {Jenkins}, \href{http://buildbot.net/}
-{Buildbot}, \href{https://www.gocd.org/} {Go}, and
-\href{http://integrity.github.io/} {Integrity}. However, installation on your
-own server is often unnecessary since there are many hosted CI solutions, such
-as: \href{https://travis-ci.org/} {Travis CI},
-\href{https://github.com/features/actions} {GitHub Actions} and
-\href{https://circleci.com/} {CircleCI}.  All that is required to begin using a
-hosted CI is to select the service and then edit a few lines of a YAML
-configuration file in the project's root directory.
-
-\cite{ShahinEtAl2017} highlights the following challenges for adopting CI: lack
-of awareness and transparency, lack of expertise and skills, coordination and
-collaboration challenges, more pressure and workload for team members, general
-resistance to change, scepticism and distrust on continuous practices. The most
-common reason given for not adopting CI is that developers are not familiar
-enough with CI \cite{HiltonEtAl2016}.  \cite{ShahinEtAl2017} observes that
-these problems can be mitigated via improving testing activities, planning and
-documentation, promoting a team mindset, adopting new rules and policies, and
-decomposing development into smaller units.
-
 Continuous integration and delivery helps with addressing several pain points.
@@
 correctness (\ppref{P_Correctness}) and the quality of reproducibility
-(\qref{Q_Reproducibility}).
+(\qref{Q_Reproducibility}). Details on CI for MI can be found in [citation
+removed for blind review] %Smith et al. \cite{SmithEtAl2024_MI_SOP} 
+, including an overview of the elements of CI, how to enable incorporation of CI
+into a team's workflow, how to set-up a CI system, and challenges for CI.
 
@@
 connections with professionals working in the medical domain, who may have
-access to private datasets and can perform tests for the team. If a team has
-such professionals as internal members, the process can be simplified.
+access to private datasets and can perform tests for the team.
 
@@
 \cite{BilicEtAl2019} and brain \cite{MenzeEtAl2015} tumor segmentation
-benchmarks.  A team developing MI software should be able to find more open
-datasets according to their needs.
+benchmarks.
 
@@
 images for testing.
-
-\item \textbf{Establish community collaboration in the domain.} During our
-interviews with developers in the MI domain, we heard many stories of asking for
-supports from other professionals or equipment manufacturers. However, we
-believe that broader collaboration between development teams can address this
-problem better. Some datasets are too sensitive to share, but if the community
-has some kind of ``group discussion'', teams can better express their needs, and
-professionals can better offer voluntary support for testing. Ultimately, the
-community can establish a nonprofit organization as a third party, which
-maintains large datasets, tests Open Source Software (OSS) in the domain, and
-protects privacy.
 
@@
 guidelines that we consulted, summarized in Section~\ref{Sec_CompareArtifacts},
-mention linters.  The one exception is \cite{ThielEtAl2020}.  Despite the lack
-of mention in the guidelines, we believe that linters have the potential to
-improve code quality at a relatively low cost.
+mention linters.  Despite the lack of mention in the guidelines, we believe that
+linters have the potential to improve code quality at a relatively low cost.
 
@@
 
-We recommend the use of linters because they are relatively easy to incorporate
-into a developer's workflow, and they address several MI pain points
-(Section~\ref{painpoints}).  For instance, linters address the lack of
-development time (\ppref{P_LackDevTime}) by increasing the developer's
-productive time via guarding against making frustrating, time-consuming, mundane
+Linters address several MI pain points.  For instance, linters address the lack
+of development time (\ppref{P_LackDevTime}) by increasing the developer's
+productive time via guarding against frustrating, time-consuming, and mundane
 mistakes.  Moreover, since a linter can include rules that capture the wisdom of
@@
 it, if a function could be short-handed, if there are indentation
-inconsistencies, etc. \cite{Whitehouse2018}. By insisting on code
-standardization linters can reduce technical debt and thus improve
+inconsistencies, etc. By insisting on code
+standardization, linters can reduce technical debt and thus improve
 maintainability (\qref{Q_Maintainability}). Although linters are tools for code
 analysis, the idea of statically checking for adherence to basic rules can be
-extended to check documentation. \cite{SmithEtAl2018_StatSoft} shows how the
-use of tools to enforce documentation standards partially explains the
-relatively higher quality of statistical tools that are part of the
-Comprehensive R Archive Network (CRAN).
+extended to check documentation. Tools to enforce documentation standards
+partially explains the relatively higher quality of statistical tools that are
+part of the Comprehensive R Archive Network (CRAN) [citation redacted for blind
+review].  %\cite{SmithEtAl2018_StatSoft}.
 
@@
 We advocate incorporating peer review into the development process, as
-frequently recommended for research software \cite{HerouxEtAl2008, Givler2020,
+frequently recommended for research software \cite{HerouxEtAl2008,
 OrvizEtAl2017, USGS2019}. In most cases a modern, lightweight review, should be
 adequate.  Modern code review is informal, tool-based, asynchronous, and focused
-on reviewing code changes \cite{SadowskiEtAl2018}. Managing a project via
-GitHub pull requests is an example of a modern approach to reviewing code.
-Software development organizations have moved to this lightweight style of code
-review because of the inefficiencies of rigorous inspections
-\cite{RigbyAndBird2013}.  However, for important parts of the code, developers
-may benefit from mixing in a more rigorous approach.
-
-\cite{Fagan1976} began work on rigorous review via code inspection.  Elements
-of a typical inspection include reviewing the code against a checklist (checking
-the consistency of variable names, look for terminating loops, etc.), performing
-specific review tasks (such as summarizing the code's purpose, cross-referencing
-the code to the technical manual, creating a data dictionary for a given module,
-etc.) Rigorous inspection finds 60-65\% of latent defects on average, and often
-tops 85\% in defect removal efficiency \cite{Jones2008}. The success rate of
-code inspection is generally higher than most forms of testing, which average
-between 30 --- 35\% for defect removal efficiency \cite{EbertAndJones2009,
-Jones2008}. For research software, \cite{KellyAndShepard2000} show a task based
-inspection approach can be effective. Task based inspection is an ideal fit with
-an issue tracking system, like GitHub.  The review tasks can be issues, so that
-they can be easily assigned, monitored and recorded. Potential issues include
-assigning junior developers to test getting-started tutorial and installation
-instructions.
-
-As indicated in Section~\ref{Sec_CompareMethodologies} some MI projects use
-modern code review, via issue tracking and the use of GitHub.  Those MI projects
-not incorporating modern code review would likely benefit by adopting it.
-Although a rigorous code inspection is likely not worth the required resources,
-for critical parts of the code, developers may want to adopt a more rigorous
-approach. For instance, developers may drop the modern trend of
-asynchronous review and instead occasionally use synchronous review to help
-uncover errors and disseminate best practices throughout the team.  For
-instance, teams could periodically meet, either in-person or virtually, and have
-junior members walk through their code.  In-person reviews will likely help
-realize the benefits of modern code review noticed by
-\cite{BirdAndBacchelli2013}: defect detection, knowledge transfer, increased
-team awareness, and creation of alternative solutions to problems.
+on reviewing code changes \cite{SadowskiEtAl2018}. Managing a project via GitHub
+pull requests is an example of a modern approach to reviewing code.  Software
+development organizations have moved to this lightweight style of code review
+because of the inefficiencies of rigorous inspections \cite{RigbyAndBird2013}.
+However, for important parts of the code, developers may benefit from mixing in
+a more rigorous approach.
+
+Fagan \cite{Fagan1976} began work on rigorous review via code inspection.
+Elements of a typical inspection include reviewing the code against a checklist
+(checking the consistency of variable names, look for terminating loops, etc.),
+performing specific review tasks (such as summarizing the code's purpose,
+cross-referencing the code to the technical manual, creating a data dictionary
+for a given module, etc.) Rigorous inspection finds 60-65\% of latent defects on
+average, and often tops 85\% in defect removal efficiency \cite{Jones2008}. The
+success rate of code inspection is generally higher than most forms of testing,
+which average between 30 --- 35\% for defect removal efficiency
+\cite{EbertAndJones2009, Jones2008}. For research software, Kelly and Shepard
+\cite{KellyAndShepard2000} show a task based inspection approach can be
+effective.
 
@@
 
-In our ``state of the practice'' assessment exercise for LBM software
-\cite{SmithEtAl2024}, we noticed that LBM developers implicitly used
-modularization based on the principle of design for change to improve
-maintainability (\qref{Q_Maintainability}).  We recommend that MI developers use
-the same principle for their modularizations.  Although the advice to modularize
-research software to handle complexity is common \cite{WilsonEtAl2014,
-StewartEtAl2017, Storer2017}, specific guidelines on how to divide the software
-into modules is less prevalent.  Not every decomposition is a good design for
-supporting change, as shown by \cite{Parnas1972a}.  For instance, a design with
-low cohesion and high coupling \cite[p.\ 48]{GhezziEtAl2003} will make change
-difficult. Especially in research software, where change is inevitable,
-designers need to produce a modularization that supports change.
-\cite{JungEtAl2022} points out that ocean modelling software is currently
-feeling the pain of not emphasizing modularization in legacy code.
-
-Specific examples of design for change for LBM software \cite{SmithEtAl2024}
-include the following:
-
-\begin{itemize}
-\item \href{https://github.com/pylbm/pylbm}{pyLBM} has decoupled geometries and
-models of their system using abstraction and modularization of the source code,
-to make it easy to add new features.  The pyLBM design allows for independent
-changes to the geometry and the model.  pyLBM also redeveloped data structures
-to ease future change. 
-\item \href{https://github.com/CFD-GO/TCLB}{TCLB} \cite{rokicki2016adjoint} is
-designed to allow for the addition of some LBM features, but changes to major
-aspects of the system would be difficult. For example, ``implementing a new
-model will be an easy contribution'', but changes to the ``Cartesian mesh … will
-be a nightmare'' \cite{SmithEtAl2024}. The design of TCLB highlights that not
-every conceivable change needs to be supported, only the likely changes.  
-\end{itemize}
-
-As the LBM examples above illustrate, developers can accomplish design for
-change by first identifying likely changes, either implicitly or explicitly, and
-second by hiding each likely change behind a well-defined module interface. This
-approach mirrors the recommendations from \cite{Parnas1972a}.
-Section~\ref{Sec_CompareArtifacts} lists ideas for how to document the design,
-including the likely changes, so that they are more visible to others.
+We recommend that MI developers use the principle of design for change for their
+modularizations.  Although the advice to modularize research software to handle
+complexity is common \cite{WilsonEtAl2014, StewartEtAl2017, Storer2017},
+specific guidelines on how to divide the software into modules is less
+prevalent.  Not every decomposition is a good design for supporting change, as
+shown by \cite{Parnas1972a}.  For instance, a design with low cohesion and high
+coupling \cite[p.\ 48]{GhezziEtAl2003} will make change difficult. Especially in
+research software, where change is inevitable, designers need to produce a
+modularization that supports change.  Ocean modelling software is currently
+feeling the pain of not emphasizing modularization in legacy code
+\cite{JungEtAl2022}.
+
+Developers can accomplish design for change by first identifying likely changes,
+either implicitly or explicitly, and second by hiding each likely change behind
+a well-defined module interface \cite{Parnas1972a}.
 
@@
 
-Below we categorize and list the threats to validity that we have identified.
-Our categories come from an analysis of software engineering secondary studies
-by \cite{AmpatzoglouEtAl2019}, where a secondary study analyzes the data from a
-set of primary studies.  \cite{AmpatzoglouEtAl2019} is appropriate because a
-common example of a secondary study is a systematic literature review. Our
-methodology is a systematic software review --- the primary studies are the
-software packages, and our work collects and analyzes these primary studies.  We
-identified similar threats to validity in our assessment of the state of the
-practice of Lattice Boltzmann Solvers \cite{SmithEtAl2024}.
+We follow Ampatzoglou et al.'s \cite{AmpatzoglouEtAl2019} analysis of
+threats to validity in software engineering secondary studies.
 
 \subsection{Reliability}
-
-A study is reliable if repetition of the study by different researchers using
-the original study's methodology would lead to the same results
-\cite{RunesonAndHost2009}. Reliability means that data and analysis are
-independent of the specific researcher(s) doing the study.  For the current
-study the identified reliability related threats are as follows:
-
-\begin{itemize}
-\item One individual does the manual measures for all packages. A different
-evaluator might find different results, due to differences in abilities,
-experiences, and biases.
-\item The manual measurements for the full set of packages took several months.
-Over this time the software repositories may have changed and the reviewer's
-judgement may have drifted.
-\end{itemize}
-
-In \cite{SmithEtAl2016} we reduced concern over the reliability risk associated
-with the reviewer's judgement by demonstrating that the measurement process is
-reasonably reproducible.  In \cite{SmithEtAl2016} we graded five software
-products by two reviewers. Their rankings were almost identical. As long as each
-grader uses consistent definitions, the relative comparisons in the AHP results
-will be consistent between graders.
-
-\subsection{Construct Validity}
-
-\cite{RunesonAndHost2009} defines construct validity as the adopted
-metrics representing what they are intended to measure. Our construct threats are
-often related to how we assume our measurements influences the various software
-qualities, as summarized in Section~\ref{sec_grading_software}. Specifically,
-our construct validity related threats include the following:
-
-\begin{itemize}
-\item We make indirect measurement of software qualities since meaningful direct
-measures for qualities like maintainability, reusability and verifiability, are
-unavailable.  We follow the usual assumption that developers achieve higher
-quality by following procedures and adhering to standards \cite[p.\
-112]{VanVliet2000}.
-\item As mentioned in Section~\ref{sec_result_installability}, we could not
-install or build \textit{dwv}, \textit{GATE}, and \textit{DICOM Viewer}. We used
-a deployed on-line version for \textit{dwv}, a VM version for \textit{GATE}, but
-no alternative for \textit{DICOM Viewer}. We might underestimate their rank due
-to these technical issues.
-\item Measuring software robustness only involved two pieces of data. This is
-likely part of the reason for limited variation in the robustness scores
-(Figure~\ref{fg_robustness_scores}). We could add more robustness data by
-pushing the software to deal with more unexpected situations, like a broken
-Internet connection, but this would require a larger investment of measurement
-time. 
-\item We may have inaccurately estimated maintainability by assuming a higher
-ratio of comments to source code improves maintainability. Moreover, we assumed
-that maintainability is improved if a high percentage of issues are closed, but
-a project may have a wealth of open issues, and still be maintainable.
-\item We assess reusability by the number of code files and LOC per file. This
-measure is indicative of modularity, but it does not necessarily mean a good
-modularization. The modules may not be general enough to be easily reused, or
-the formatting may be poor, or the understandability of the code may be low.
-\item The understandability measure relies on 10 random source code files, but
-the 10 files will not necessarily be representative. 
-\item As discussed in Section~\ref{Sec_OverallQ}, our overall AHP ranking makes
-the unrealistic assumption of equal weighting.
-\item We approximated popularity by stars and watches
-(Section~\ref{Sec_VsCommunityRanking}), but this assumption may not be valid. 
-\item As mentioned in Section~\ref{sec_interview_methods}, one interviewee was
-too busy to participate in a full interview, so they provided written answers
-instead. Since we did not have the chance to explain our questions or ask them
-follow-up questions, there is a possibility of misinterpretation of the
-questions or answers.
-\item In building Table~\ref{Tbl_Guidelines} some judgement was necessary on our
-part, since not all guidelines use the same names for artifacts that contain
-essentially the same information.
-\end{itemize}
-
-\subsection{Internal Validity} \label{Sec_InternalValidity}
-
-Internal validity means that discovered causal relations are trustworthy and
-cannot be explained by other factors \cite{RunesonAndHost2009}. In our
-methodology the internal validity threats include the following:
-
-\begin{itemize}
-\item In our search for software packages
-(Section~\ref{sec_software_selection}), we may have missed a relevant package.
-\item Our methodology assumes that all relevant software development activities
-will leave a trace in the repositories, but this is not necessarily true. For
-instance, the possibility exists that CI usage was higher than what we observed
-through the artifacts (Section~\ref{Sec_CompareTools}). As another example,
-although we saw little evidence of requirements
-(Section~\ref{Sec_CompareTools}), maybe teams keep this kind of information
-outside their repos, possibly in journal papers or technical reports.
-\item We interviewed a relatively small sample of 8 teams.  Their pain points
-(Section~\ref{painpoints}) may not be representative of the rest of their
-community.
-\end{itemize}
-
-\subsection{External Validity}
-
-If the results of a study can be generalized (applied) to other
-situations/cases, then the study is externally valid \cite{RunesonAndHost2009}.
-We are confident that our search was exhaustive.  We do not believe that we
-missed any highly popular examples.  Therefore, the bulk of our validity
-concerns are internal (Section~\ref{Sec_InternalValidity}).
-However, our hope is that the trends observed, and the lessons learned for MI
-software can be applied to other research software.  With that in mind we
-identified the following threat to external validity:
-
-\begin{itemize}
-\item We cannot generalize our results if the development of MI software is
-fundamentally different from other research software.
-\end{itemize}
-
-Although there are differences, like the importance of data privacy for MI data,
-we found the approach to developing LBM software \cite{SmithEtAl2024} and MI
-software to be similar.  Except for the domain specific aspects, we believe that
-the trends observed in the current study are externally valid for other research
-software.
-
-\section{Future Work} \label{Sec_FutureWork}
-
-The following recommendations for future state of the practice measurement
-exercises, for MI or for other domains, could address some threats to validity
-mentioned above.  Moreover, some ideas may make the data collection more
-efficient.
-
-\begin{itemize}
-    \item We would like to make surface measurements less shallow. For example:
-    \begin{itemize}
-        \item Surface reliability: our current measurement relies on the
-        processes of installation and getting started tutorials. However, not
-        all software needs installation or has a getting started tutorial. We
-        could devise a list of operation steps (with the help of the Domain
-        Expert), perform the same operations with each software, and record any
-        errors.
-        \item Surface robustness: we used damaged images as inputs for this
-        measuring MI software. This process is similar to fuzz testing
-        \cite{enwiki:1039424308}, which is one type of fault injection
-        \cite{enwiki:1039005082}. We may adopt more fault injection methods,
-        and identify tools and libraries to automate this process.
-        \item Surface usability: we can design usability tests and test all
-        software projects with end-users. The end-users can be volunteers and
-        domain experts. Ideas for getting started are available in
-        \cite{SmithEtAl2021}.
-        \item Surface understandability: our current method does not require
-        understanding the source code. As software engineers, perhaps we can
-        select a small module of each project, read the source code and
-        documentation, try to understand the logic, and score the ease of the
-        process.  
-        \item Maintainability: we can add a measure modifiability as part of the
-        measurement of maintainability.  An experiment could be conducted asking
-        participants to make modifications, observing the study subjects during
-        the modifications, testing the resulting software and surveying the
-        participants \cite{SmithEtAl2021}.
-    \end{itemize}
-	\item We can further automate the measurements on the grading template. For
-	example, with automation scripts and the GitHub API, we may save significant
-	time on retrieving the GitHub metrics through a GitHub Metric Collector.
-	This Collector can take GitHub repository links as input, automatically
-	collect metrics from the GitHub API, and record the results.
-	\item We can improve some interview questions. Some examples are:
-	\begin{itemize}
-	    \item In one question we ask, ``Do you think improving this process can
-	    tackle the current problem?''  The problem is that this is a yes-or-no
-	    question, which is not informative. We could change the question to ``By
-	    improving this process, what current problems can be tackled?''; 
-	    \item We can ask for more details about the modular approach, such as
-	    ``What principles did you use to divide code into modules? Can you
-	    describe an example of using your principles?''.
-	\end{itemize}
-	\item We can better organize the interview questions. Since we use audio
-	conversion tools to transcribe the answers, we should make the transcription
-	easier to read. For example, we can order them together for questions about
-	the five software qualities and compose a similar structure for each.
-	\item We can mark the follow-up interview questions with keywords. For
-	example, say ``this is a follow-up question'' every time asking one. Thus,
-	we record this sentence in the transcription, and it will be much easier to
-	distinguish the follow-up questions from the 20 designed questions.
-\end{itemize}
-
-\section{Conclusions} \label{ch_conclusions}
-
-We analyzed the state of the practice for the MI domain with the goal of
-understanding current practice, answering our ten research questions
-(Section~\ref{sec_motivation}) and providing recommendations for current and
-future projects.  Our methods in Section~\ref{ch_methods} form a general process
-to evaluate domain-specific software, that we apply to the specific domain of MI
-software. We identified 48 MI software candidates, then, with the help of the
-Domain Expert selected 29 of them to our final list.
-
-Section~\ref{ch_results} lists our measurement results for ranking the 29
-projects for nine software qualities. Our ranking results appear credible since
-they are mostly consistent with the ranking from the scientific community
-implied by the GitHub stars-per-year metric. As discussed in
-Section~\ref{Sec_VsCommunityRanking}, four of the top five software projects
-appear in both our list and in the GitHub popularity list.  Moreover, our top
-five packages appear among the first eight positions on the GitHub list.  The
-noteworthy discrepancies between the two lists are for the packages that we were
-unable to install (\textit{dwv} and \textit{Dicom Viewer}).
-
-Based on our grading scores \textit{3D Slicer}, \textit{ImageJ}, \textit{Fiji}
-and \textit{OHIF Viewer} are the top four software performers.  However, the
-separation between the top performers and the others is not extreme.  Almost all
-packages do well on at least a few qualities, as shown in
-Table~\ref{topperformerstable}, which summarizes the packages ranked first and
-second for each quality. Almost 70\% (20 of 29) of the software packages appear
-in the top two for at least two qualities.  The only packages that do not appear
-in Table~\ref{topperformerstable}, or only appear once, are \textit{Papaya},
-\textit{MatrixUser}, \textit{MRIcroGL}, \textit{XMedCon}, \textit{dicompyler},
-\textit{DicomBrowser}, \textit{AMIDE}, \textit{3DimViewer}, and
-\textit{Drishti}. The shortness of this list suggests parity with respect to
-adoption of best practices for MI software overall.
-
-\begin{table}[ht!]
-	\begin{center}
-	\renewcommand{\arraystretch}{1.8}
-	\begin{tabular}{ p{3cm}p{13cm} }
-		\toprule
-
-		Quality & Ranked 1st or 2nd\\
-
-		\midrule
-
-		Installability & 3D Slicer, BioImage Suite Web, Slice:Drop, INVESALIUS\\
-
-		\pbox{3.0cm}{Correctness \\ and Verifiability} & OHIF Viewer, 3D
-		Slicer, ImageJ\\
-
-		Reliability & SMILI, ImageJ, Fiji, 3D Slicer, Slice:Drop, OHIF
-		Viewer\\
-
-		Robustness & XMedCon, Weasis, SMILI, ParaView, OsiriX Lite,
-		MicroView, medInria, ITK-SNAP, INVESALIUS, ImageJ, Horos, Gwyddion,
-		Fiji, dicompyler, DicomBrowser, BioImage Suite Web, AMIDE, 3DimViewer,
-		3D Slicer, OHIF Viewer, DICOM Viewer\\
-
-		Usability & 3D Slicer, ImageJ, Fiji, OHIF Viewer, ParaView,
-		INVESALIUS, Ginkgo CADx, SMILI, OsiriX Lite, BioImage Suite Web,
-		ITK-SNAP, medInria, MicroView, Gwyddion\\
-
-		Maintainability & 3D Slicer, Weasis, ImageJ, OHIF Viewer, ParaView\\
-
-		Reusability & 3D Slicer, ImageJ, Fiji, OHIF Viewer, SMILI, dwv, BioImage
-		Suite Web, GATE, ParaView\\
-
-		\pbox{3.0cm}{Understandability} & 3D Slicer, ImageJ, Weasis,
-		Fiji, Horos, OsiriX Lite, dwv, Drishti, OHIF Viewer, GATE, ITK-SNAP,
-		ParaView, INVESALIUS\\
-
-		\pbox{3.0cm}{Visibility and \\Transparency} & ImageJ, 3D Slicer, Fiji\\
-
-		Overall Quality & 3D Slicer, ImageJ\\
-
-		\bottomrule		
-	\end{tabular}
-	\caption{Top performers for each quality (sorted by order of quality
-	measurement)} \label{topperformerstable}
-	\end{center}
-\end{table}
-
-For insight into devising future methods and tools, we interviewed nine
-developers (from eight teams) to learn about their pain points
-(Section~\ref{painpoints}).  We also discussed qualities of potential concern.
-The identified pain points and qualities of concern include:
-
-\begin{description}
-\item [\ppref{P_LackDevTime}] Lack of development time, %P1
-\item [\ppref{P_LackFunding}] Lack of funding, %P2
-\item [\ppref{P_TechnologyHurdles}] Technology hurdles, %P3
-\item [\ppref{P_Correctness}] Ensuring correctness, %P4
-\item [\ppref{P_Usability}] Usability, %P5
-\item [\qref{Q_Maintainability}] Quality of maintainability, and %Q1
-\item [\qref{Q_Reproducibility}] Quality of reproducibility. %Q2
-\end{description}
-
-Despite the pain points, overall MI software is in a healthy state for software
-development practices.  In our survey of the selected projects we observed 88\%
-of the documentation artifacts recommended by research software development
-guidelines (Section~\ref{Sec_CompareArtifacts}).  With respect to tools, MI is
-keeping pace with other research software with 100\% of the projects using
-version control (with 93\% specifically using git)
-(Section~\ref{Sec_CompareTools}).  We observed that the MI developers tend to
-follow the typical research software trend of using a quasi-agile software
-development process (Section~\ref{Sec_CompareMethodologies}).
-
-Although the state of the practice for MI software is healthy, we did notice
-areas where practice seems to lag behind the research software development
-guidelines.  For instance, the guidelines recommend three artifacts that were
-not observed: uninstall instructions, test plans, and requirements
-documentation. We observed the following recommended artifacts, but only rarely:
-contributing file, developer code of conduct, code style guidelines, product
-roadmap, design documentation, and API documentation
-(Section~\ref{Sec_CompareArtifacts}). Although software development tool use
-seems healthy, we found the use of CI/CD behind typical usage rates (17\% of the
-projects used CI/CD) (Section~\ref{Sec_CompareTools}).  With respect to the
-development process, developer identified areas for improvement included testing
-(only 50\% of projects were identified to have unit testing) and documentation
-(only three out of nine developers felt their documentation was clear enough)
-(Section~\ref{Sec_CompareMethodologies}).
-
-Our interviewees proposed strategies to improve the state of the practice, to
-address the identified pain points, and to improve software quality.  To their
-list (Section~\ref{painpoints}) we added some of our own recommended strategies
-(Section~\ref{ch_recommendations}).  Below we summarize the proposed strategies,
-with traceability to where we discuss the strategy, and to the relevant pain
-points.
-
-\begin{enumerate}
-
-\item Increase documentation to address \ppref{P_LackDevTime},
-\ppref{P_TechnologyHurdles}, \ppref{P_Usability}, \qref{Q_Maintainability},
-\qref{Q_Reproducibility} (Section~\ref{painpoints})
-
-\item Increase testing by enriching datasets to address \ppref{P_Correctness},
-\qref{Q_Reproducibility}
-(Section~\ref{painpoints},~\ref{sec_recommendations_testing_dataset})
-
-\item Increase modularity to address \qref{Q_Maintainability} (Section~\ref{painpoints})
-
-\item Use continuous integration to address \ppref{P_LackDevTime},
-\ppref{P_Correctness}, \qref{Q_Reproducibility}
-(Section~\ref{painpoints},~\ref{Sec_ContinuousIntegration})
-
-\item Move to web applications to address \ppref{P_TechnologyHurdles}
-(Section~\ref{painpoints},~\ref{sec_webapps})
-
-\item Employ linters to address \ppref{P_LackDevTime},
-\ppref{P_TechnologyHurdles}, \qref{Q_Maintainability}
-(Section~\ref{Sec_Linters})
-
-\item Peer reviews to address \ppref{P_LackDevTime},
-\ppref{P_TechnologyHurdles}, \ppref{P_Correctness}, \qref{Q_Maintainability}
-(Section~\ref{Sec_PeerReview})
-
-\item Design for change to address \qref{Q_Maintainability} (Section~\ref{Sec_DesForChange})
-
-\item Assurance case to address \ppref{P_Correctness}, \qref{Q_Reproducibility}
-(Section~\ref{AssuranceCases})
-
-\item Generate all things to address \ppref{P_LackDevTime},
-\ppref{P_TechnologyHurdles}, \ppref{P_Correctness}, \ppref{P_Usability},
-\qref{Q_Maintainability} and \qref{Q_Reproducibility}.
-(Section~\ref{Sec_GenAllThings})
-
-\end{enumerate}
-
-\subsection{Threats to Validity} \label{sec_threats_to_validity}
-
-We follow Ampatzoglou et al.\ \cite{AmpatzoglouEtAl2019}'s analysis of
-threats to validity in software engineering secondary studies.
-
-\subsubsection{Reliability}
 
@@
 
-\subsubsection{Construct Validity}
+\subsection{Construct Validity}
 
@@
 Viewer}.
-\item Robustness measurements involve only two pieces of data, leading
-to limited variation in the robustness scores
-(Figure~\ref{fg_robustness_scores}). Our measurement-time budget limited
+\item Robustness measurements involve only two pieces of data, leading to
+limited variation in the robustness scores. Our measurement-time budget limited
 what we could achieve here.
@@
 
-\subsubsection{Internal Validity} \label{Sec_InternalValidity}
+\subsection{Internal Validity} \label{Sec_InternalValidity}
 
@@
 \begin{itemize}
-\item Our search (Section~\ref{ch_results}) could
-have missed a relevant package.
+\item Our search (Section~\ref{SecWhatProjects}) could have missed a relevant
+package.
 \item Our methodology assumes that development activities
@@
 information outside their repos.
+\item We interviewed a relatively small sample of 8 teams. Their pain points
+(Section~\ref{painpoints}) may not be representative of the rest of their
+community.
 \end{itemize}
 
-\subsubsection{External Validity}
+\subsection{External Validity}
 
@@
 	\renewcommand{\arraystretch}{1.8}
-	\begin{tabular}{ p{2.3cm}p{9.9cm} }
+	\begin{tabular}{ p{3cm}p{12.5cm} }
 		\toprule
@@
 
-Our grading scores indicate that
-\textit{3D Slicer}, \textit{ImageJ}, \textit{Fiji}
-and \textit{OHIF Viewer} are the overall top four.  However, the
-separation between the top performers and the others is not extreme.  Almost
-all packages do well on at least a few qualities, as shown in
+Our grading scores indicate that \textit{3D Slicer}, \textit{ImageJ},
+\textit{Fiji} and \textit{OHIF Viewer} are the overall top four.  However, the
+separation between the top performers and the others is not extreme.  Almost all
+packages do well on at least a few qualities, as shown in
 Table~\ref{topperformerstable}, which summarizes the packages ranked first and
@@
 
-While we did find a reasonable amount of documentation, especially when
-consider all MI projects, there were definitely some holes. Some important
-documentation (test plans and requirements documentation) was missing,
-and other (contributors' guide, code of conduct, code style guidelines, product
-roadmap, design documentation, and API documentation)
+While we did find a reasonable amount of documentation, especially when consider
+all MI projects, there were definitely some holes. Some important documentation
+(test plans and requirements documentation) was missing, and other
+(contributors' guide, code of conduct, code style guidelines, product roadmap,
+design documentation, and API documentation)
 (Section~\ref{Sec_CompareArtifacts}) were rare.
 
-A deeper understanding of the needs of
-the MI community will require data beyond what is available in
-repositories.  Future work should involve interviewing MI developers to
-better understand their ``pain points''.
+Our interviewees proposed strategies to improve the state of the practice, to
+address the identified pain points, and to improve software quality. To their
+list (Section~\ref{painpoints}) we added some of our own recommended strategies
+(Section~\ref{ch_recommendations}). The strategies that emerged include
+increasing documentation, increasing testing by enriching datasets, increasing
+modularity, using continuous integration, moving to web applications, employing
+linters, increasing peer review, and using design for change to guide
+modularization.
+
+A deeper understanding of the needs of the MI community will require data beyond
+what is available in repositories.  Future work should involve interviewing MI
+developers to better understand their ``pain points''.
 

### 2024-11-06 StateOfPractice/Papers/SOP_MI_CritReviewsBioMedEng/SOP_MI_OSR.tex
edit start of abstract and remove all tabs.
start editing

@@
 
-We present the state of the practice for Medical Imaging (MI) software based on
-data available in open source repositories. We selected 29 projects from 48
-candidates and assessed 9 software qualities (installability, correctness/
-verifiability, reliability, robustness, usability, maintainability, reusability,
-understandability, and visibility/transparency) by answering 108 questions for
-each. Using the Analytic Hierarchy Process (AHP) on the quantitative data, we
-ranked the MI software.  The top five are \textit{3D Slicer}, \textit{ImageJ},
-\textit{Fiji}, \textit{OHIF Viewer}, and \textit{ParaView}.  This is consistent
-with the community's view, with four of these also appearing in the top five
-using GitHub metrics (stars-per-year).  The quality and quantity of
-documentation present in a project correlates quite well with its popularity.
-Generally, MI software is in a healthy state as shown by the following: in the
-repositories we observed 88\% of the documentation artifacts recommended by
-research software development guidelines, 100\% of MI projects use version
-control tools, and developers appear to use the common quasi-agile research
-software development process. However, the current state of the practice
-deviates from the existing guidelines because of the rarity of some recommended
-artifacts (like a test plan, requirements specification, code of conduct, and
-code style guidelines), low usage of continuous integration (17\% of the
-projects), low use of unit testing (about 50\% of projects), and room for
-improvement with documentation. From interviewing the developers, we identified
-6 concerns: lack of development time, lack of funding, technology hurdles,
-correctness, usability, maintainability, and reproducibility. Recommendations
-for improving the state of the practice include the following: increase
-documentation, increase testing by enriching datasets, increase continuous
-integration usage, move to web applications, employ linters, use peer reviews,
-and design for change.
+We present the state of the practice for the development of Medical Imaging
+(MI) software based on data available in open source repositories. We selected
+29 projects from 48 candidates and assessed 9 software qualities
+(installability, correctness/ verifiability, reliability, robustness,
+usability, maintainability, reusability, understandability, and
+visibility/transparency) by answering 108 questions for each. Using the
+Analytic Hierarchy Process (AHP) on the quantitative data, we ranked the MI
+software.  The top five are \textit{3D Slicer}, \textit{ImageJ}, \textit{Fiji},
+\textit{OHIF Viewer}, and \textit{ParaView}.  This is consistent with the
+community's view, with four of these also appearing in the top five using
+GitHub metrics (stars-per-year).  The quality and quantity of documentation
+present in a project correlates quite well with its popularity.  Generally, MI
+software is in a healthy state: in the repositories
+we observed 88\% of the documentation artifacts recommended by research
+software development guidelines, 100\% of MI projects use version control
+tools, and developers appear to use the common quasi-agile research software
+development process. However, the current state of the practice deviates from
+existing guidelines as some recommended artifacts are rarely present
+(like a test plan, requirements specification, code of conduct, and code style
+guidelines), low usage of continuous integration (17\% of the projects), low
+use of unit testing (about 50\% of projects), and room for improvement with
+documentation. From developper interviews , we identified 7 concerns: lack
+of development time, lack of funding, technology hurdles, correctness,
+usability, maintainability, and reproducibility. We recommend:
+increase effort on documentation,
+increase testing by enriching datasets, increase continuous integration usage,
+move to web applications, employ linters, use peer reviews, and design for
+change.
 
@@
 As well as state of the practice (SOP) for MI software, we would like to
-understand the impact of the often cited gap (chasm!) between recommended
+understand the impact of the often cited gap between recommended
 software engineering practices and that used for most research 
@@
 
-Our investigation is based on the following four research questions:
+Our investigation is based on the following research questions:
 \begin{enumerate}
@@
 need an unzip tool; \textit{MatrixUser} needs Matlab; \textit{DICOM Viewer}
-needs a Nextcloud platform. \textit{GATE}, \textit{dwv}, and \textit{DICOM
-Viewer} showed severe installation problems. We were not able to install them,
-even after a reasonable amount of time (2 hours).
+needs a Nextcloud platform. We were not able to install \textit{GATE},
+\textit{dwv}, \textit{DICOM Viewer} even after trying for 2 hours (each).
 
@@
 similarly except that it showed a meaningless image.  \textit{Drishti}
-successfully detected the broken format of the file, but the software crashed as
-a result.
+successfully detected the broken format, but the software then crashed.
 
@@
 To answer \rqref{RQ_CompareHQ2Popular} we use GitHub stars, number of forks and
-number of people watching the projects are proxies for community ranking.
+number of people watching the projects as proxies for community ranking.
 Table~\ref{tab_ranking_vs_GitHub} show the statistics for data collected in July
-2021.  Recall that 24 projects use GitHub (Section~\ref{SecMeasurementResults}).
+2021.
 
@@
 
-Another likely reason for discrepancies is that we weighted all qualities
-equally. This is not likely how users implicitly rank the different qualities.
-This would require a broader user study to properly assess. Furthermore our
-measures of popularity are only \emph{proxies} which are biased towards
-past rather than current preferences~\cite{Szulik2017}, as these are
-monotonically increasing quantities. Finally there are often more factors than
-just quality that influence the popularity of ``consumer'' products.
+We weighted all qualities equally, which is not the likely (implicit) weight
+that users implicitly use. To properly assess this would require a broad user
+study.  Furthermore our measures of popularity are only \emph{proxies} which
+are biased towards past rather than current preferences~\cite{Szulik2017}, as
+these are monotonically increasing quantities. Finally there are often more
+factors than just quality that influence the popularity of ``consumer''
+products.
 
@@
 causal, in either direction (i.e. if best practices enable popularity or if
-popularity increases the need for using more software development best
-practices).
+popularity increases the need for best practices).
 

### 2024-11-08 StateOfPractice/Papers/SOP_MI_CritReviewsBioMedEng/SOP_MI_OSR.tex
section 6

@@
-\documentclass[final, 12pt, 3p, times]{elsarticle}
+\documentclass[doubleblind,12pt, 3p, times]{elsarticle}
 
@@
 
-\begin{figure*}[!ht]
+\begin{figure*}[ht!]
 \includegraphics[scale=0.67]{template.pdf}
@@
 
-\begin{table*}[!ht]
+\begin{table*}[ht!]
 \centering
@@
 
-To answer \rqref{RQ_HighestQuality} we measured the software as described in
+We measured the software as described in
 Section~\ref{SecMethodology}.  In the absence of a specific real world context,
@@
 
-\begin{figure*}[!ht]
+\begin{figure*}[ht!]
 \includegraphics[scale=0.47]{overall_scores.pdf}
@@
 
-To answer \rqref{RQ_CompareHQ2Popular} we use GitHub stars, number of forks and
+We use GitHub stars, number of forks and
 number of people watching the projects as proxies for community ranking.
@@
 \renewcommand{\arraystretch}{0.85}
-\begin{table*}[!ht]
+\begin{table*}[ht!]
 \centering
@@
 
-\section{Compare to Recommended Software Artifacts} \label{Sec_CompareArtifacts}
-
-For \rqref{RQ_CompareArtifacts} we use nine research software development
+\section{Comparing with Recommended Software Artifacts} \label{Sec_CompareArtifacts}
+
+We use a set of nine research software development
 guidelines to compare recommended software artifacts versus those present in MI
-software.  These guidelines are:
+software:
 \begin{itemize}
@@
 
-\begin{table*}[!ht]
+\begin{table*}[ht!]
 \begin{center}
@@
 
-In Table~\ref{Tbl_Guidelines} each row corresponds to an artifact.  For a given
-row, a checkmark in one of the columns means that the corresponding guideline
-recommends this artifact.  The last column shows whether the artifact appears in
-the measured set of MI software, either not at all (blank), commonly (C),
-uncommonly (U) or rarely (R).  We did our best to interpret the meaning of each
-artifact consistently between guidelines and specific MI software, but the
+We show the recommended artifacts in Table~\ref{Tbl_Guidelines}, one per row.
+For a given row, a checkmark in one of the columns means that the corresponding
+guideline recommends this artifact.  The last column shows whether the artifact
+appears in the measured set of MI software, either not at all (blank), commonly
+(C), uncommonly (U) or rarely (R).  We did our best to interpret the meaning of
+each artifact consistently between guidelines and specific MI software, but the
 terminology and the contents of artifacts are not standardized.  The challenge
 even exists for the ubiquitous README file.  The content of README files shows
-significant variation between projects \cite{PranaEtAl2018}.  Although some
-content is reasonably consistent, with 97\% of README files contain at least one
-section describing the `What' of the repository and 89\% offering some `How'
-content, other categories are more variable.  For instance, information on
+significant variation between projects~\cite{PranaEtAl2018}.  Some content is
+reasonably consistent, with 97\% of README files contain at least one section
+describing the `What' of the repository and 89\% offering some `How' content,
+other categories are more variable.  For instance, information on
 `Contribution', `Why', and `Who', appear in 28\%, 26\% and 53\% of the analyzed
-files, respectively \cite{PranaEtAl2018}.
-
-\begin{table*}[!ht]
+files, respectively~\cite{PranaEtAl2018}.
+
+\begin{table*}[ht!]
     \begin{center}
@@
     \end{tabular}
-    \caption{Artifacts Present in MI Packages, Classified by Frequency (The number 
-    in brackets is the number of occurrences)}
+    \caption{Artifacts Present in MI Packages, Classified by Frequency
+    (number of occurrences)}
     \label{artifactspresent}
@@
 
-Note that ``popularity'' in Table~\ref{Tbl_Guidelines} does no imply that these
+Note that ``popularity'' in Table~\ref{Tbl_Guidelines} does not imply that these
 oft recommended artifacts are the most important. Guidelines are often brief,
@@
 We use data from interviews, as described in Section~\ref{sec_interview_methods}
-to answer \rqref{RQ_CompareToolsProjMngmnt}, \rqref{RQ_CompareMethodologies}
-and \rqref{RQ_PainPoints}. Requests were sent to all 29 projects.  Nine
-developers from eight of the projects agreed to participate: \textit{3D Slicer},
-\textit{INVESALIUS 3}, \textit{dwv}, \textit{BioImage Suite Web},
-\textit{ITK-SNAP}, \textit{MRIcroGL}, \textit{Weasis}, and \textit{OHIF}. We
-spent about 90 minutes for each interview. The full interview answers can be
+to answer \rqref{RQ_CompareToolsProjMngmnt} (tool usage),
+\rqref{RQ_CompareMethodologies} (methodology) and \rqref{RQ_PainPoints} (pain
+points). Requests were sent to all 29 projects.  Nine developers from eight of
+the projects agreed to participate: \textit{3D Slicer}, \textit{INVESALIUS 3},
+\textit{dwv}, \textit{BioImage Suite Web}, \textit{ITK-SNAP},
+\textit{MRIcroGL}, \textit{Weasis}, and \textit{OHIF}. We spent about 90
+minutes for each interview. The full interview answers can be
 found in [citation redacted for blind review].
@@
 
-Developers use software tools for user support, version control, Continuous
-Integration and Deployment (CI/CD), documentation and project management. To
-answer \rqref{RQ_CompareToolsProjMngmnt}, we summarize the tool usage in these
-categories, and compare this to the usage by the research software community.
-
-Table~\ref{tab_user_support_model} summarizes the user support models by the
-number of projects using each model (projects may use more than one support
-model). The prevalent usage of GitHub issues may be by design, or it might just
-naturally arise as users seek help.
-
-\begin{table}[!ht]
+Developers use software tools for user support, version control (VCS),
+Continuous Integration and Deployment (CI/CD), documentation and project
+management.  We summarize the tool usage in each of these categories, and
+compare this to the usage by the research software community.
+
+\paragraph{User support} Table~\ref{tab_user_support_model} summarizes the user
+support models by the number of projects using each model (projects may use
+more than one support model).
+
+\begin{table}[ht!]
 \centering
@@
 
-Most teams interview (eight of nine) mentioned using GitHub and pull requests to manage
-contributions from the community. The interviewees generally gave very positive
-feedback on using GitHub. Some teams previously used a different approach to
-version control and eventually transferred to git and GitHub.
-
-The common use of version control for MI software illustrates considerable
-improvement from the poor adoption of version control tools that Wilson lamented
-in 2006 \cite{Wilson2006}.  The proliferation of version control tools for MI
-matches the increase in the broader research software community.  A little over
-10 years ago Nguyen-Hoan et al.\ \cite{Nguyen-HoanEtAl2010} estimated that only
-50\% of research software projects used version control, but even at that time
-there was an increase from previous usage levels. A survey in 2018 shows 81\% of
-developers use a version control system \cite{AlNoamanyAndBorghi2018}.
-\cite{Smith2018} has similar results, showing version control usage for alive
-projects in mesh generation, geographic information systems and statistical
-software for psychiatry increasing from 75\%, 89\% and 17\% (respectively) to
-100\%, 95\% and 100\% (respectively) over a four-year period ending in 2018. A
-recent survey by \cite{CarverEtAl2022} shows version control use among
-practitioners at over 95\%. All but one of the software guides cited in
-Section~\ref{Sec_CompareArtifacts} recommends version control. The high usage of
-version control tools in MI matches the general trend in research software.
-
-As mentioned in Section~\ref{SecMeasurementResults}, we identified 17\% of MI
-projects using CI/CD tools. We found which projects used CI/CD by examining the
-documentation and source code. The count of CI/CD usage may be higher, since
-traces of CI/CD may not always appear in a repository.  This was the case for a
-study of Lattice Boltzmann Method (LBM) software, where interviews with
-developers showed that more projects used CI/CD than was evident from repository
-artifacts alone [citation redacted for blind review].  %\cite{Michalski2021}.  
-The 17\% utilization for MI software contrasts with the high frequency with
-which research software development guidelines recommend continuous integration
+\paragraph{Version control} Most teams interviewed (eight of nine) use GitHub
+for version control, which makes it unsurprising that they also use pull
+requests for managing community contributions. The experience of using GitHub
+was reported as generally positive; some teams had actively migrated to GitHub
+from other VCS.  This represents a considerable uptake since its rare use in
+2006~\cite{Wilson2006}, which matches the uptake in other communities: 10 years
+ago Nguyen et al~\cite{Nguyen-HoanEtAl2010} estimated that only 50\% of
+research software projects used version control, and by 2018 this had jumped to
+over 80\%~\cite{AlNoamanyAndBorghi2018} with many communities~\cite{Smith2018}
+reaching 100\% already in 2018. In 2022, version control use sits at over
+95\%~\cite{CarverEtAl2022}.
+
+\paragraph{CI/CD} 
+Only 17\% of MI projects use CI/CD tools. We estimated this using artifacts
+present in the repository -- actual usage may be higher as some tools are
+entirely external. This was the case for a study of Lattice Boltzmann Method
+(LBM) software [citation redacted for blind review].  %\cite{Michalski2021}.
+This contrasts with the high frequency of guidelines recommending continuous
+integration
 \cite{BrettEtAl2021, Brown2015, ThielEtAl2020, Zadka2018, vanGompelEtAl2016}.
-Although there is currently little data available on CI/CD utilization for
-research software, our impression is that CI/CD is not yet common practice,
-despite its recommendation.  This is certainly the case for LBM software, where
-usage numbers sit around 12.5\% [citation redacted for blind review]
-%\cite{Michalski2021}.  
-The survey of \cite{CarverEtAl2022} suggests higher use of CI/CD with 54\%
-(54/100) of respondents indicating that they use it. However, that study surveys
-practitioners, rather than assessing projects.  Additional information on CI/CD
-is given in the recommendations (Section~\ref{Sec_ContinuousIntegration}).
-
-For documentation tools and methods mentioned by the interviewees, the most
-popular (mentioned by about 30\% of developers) were forum discussions and
-videos.  The second most popular options (mentioned by about 20\% of developers)
-were GitHub, wiki pages, workshops, and social media. The least frequently
-mentioned options (about 10\% of developers) included writing books, and google
-forms.  In contrasting MI software with LBM software, the most significant
-documentation tool difference is that LBM software often uses document
-generation tools, like doxygen and sphinx [citation redacted for blind review].
+A recent survey~\cite{CarverEtAl2022} of practioners (rather than projects)
+suggests higher use of CI/CD with 54\% of respondents indicating that they use
+it.
+
+\paragraph{Documentation} The most popular (mentioned by about 30\% of
+developers) were forum discussions and videos.  The second most popular options
+(20\%) were GitHub, wiki pages, workshops, and social media. The least
+frequently mentioned options (about 10\% of developers) included writing books,
+and Google forms.  As a point of contrast, LBM software most often uses API
+document generation tools, like doxygen and sphinx [citation redacted for blind
+review].
 %\cite{Michalski2021}
 
-The interviewees talked about two types of project management tools:
+\paragraph{Project management} Two types of tools were mentionned:
 \begin{inparaenum}[i)]
@@
 \end{inparaenum}
-Of the specifically named tools, interviewees mentioned GitHub 3 times, and each
+Of the named tools, interviewees mentioned GitHub 3 times, and each
 of the other tools once each.
 
-Based on information provided by \cite{JungEtAl2022}, tool utilization for MI
+Based on~\cite{JungEtAl2022}, tool utilization for MI
 software has much in common with tool utilization for ocean modelling software.
@@
 
-\begin{table*}[!ht]
+\begin{table*}[ht!]
     %\begin{center}

### 2024-11-08 StateOfPractice/Papers/SOP_MI_CritReviewsBioMedEng/SOP_MI_OSR.tex
Section 7 edited, and first full page saved.

@@
 
-Based on~\cite{JungEtAl2022}, tool utilization for MI
-software has much in common with tool utilization for ocean modelling software.
+Based on~\cite{JungEtAl2022}, tool use for MI
+software is similar to tool use for ocean modelling software.
 Both use tools for editing, compiling, code management, testing, building, and
-project management.  From the data available, ocean modelling differs from MI
-software in its use of Kanban boards for project management.
+project management, but differ
+in the latter's use of Kanban boards for project management.
 
@@
 
-We answer research question \rqref{RQ_CompareMethodologies} by comparing the
-principles, processes, and methodologies used for MI software to what can be
-gleaned from the literature on research software in general. In our interviews
-with developers the responses about development model were vague, with only two
-interviewees following a definite development model. In some cases the
-interviewees felt their process was similar to an existing development model.
-Three teams (about 38\%) either followed agile, or something similar to agile.
-Two teams (25\%) either followed a waterfall process, or something similar.
-Three teams (about 38\%) explicitly stated that their process was undefined or
-self-directed.
-
-Our observations of an informally defined process, with elements of agile
-methods, matches observations for research software in general.
+In our interviews, developers' responses to questions about development model
+were vague, with only two interviewees following a definite model, with others
+feeling their process was ``similar'' to an existing model.  Three teams
+followed agile or alike, two teams  followed waterfall or alike, and three
+teams explicitly stated that their process was undefined or self-directed.
+
+This matches observations for research software in general.
 Scientific developers naturally use an agile philosophy \cite{AckroydEtAl2008,
@@
 amethododical process \cite{Kelly2013}, or a knowledge acquisition driven
-process \cite{Kelly2015}.  A waterfall-like process can work for research
-software \cite{Smith2016}, especially if the developers work iteratively and
+process~\cite{Kelly2015}.  A waterfall-like process can work for research
+software~\cite{Smith2016}, especially if the developers work iteratively and
 incrementally, but externally document their work as if they followed a
-rationale design process \cite{parnas1986rational}.
-
-No interviewee introduced any strictly defined project management process. The
-most common approach was following the issues, such as bugs and feature
-requests. The \textit{3D Slicer} team had weekly meetings to
-discuss the goals for the project. The \textit{INVESALIUS 3} team relied on the
-GitHub process for their project management while the \textit{ITK-SNAP} team had a
-fixed six-month release pace. Only the interviewee from \textit{OHIF} 
-mentioned the team having a project manager. The \textit{3D Slicer} team and
-\textit{BioImage Suite Web} team do nightly builds and tests. The \textit{OHIF}
-developer believes that a better project management process can improve junior
-developer efficiency while also improving internal and external communication.
-
-We identified the use of unit testing in less than half of the 29 projects even
-though the interviewees believed that testing (including usability tests with
-users) was the top approach to improve correctness, usability, and
-reproducibility.  This level of testing matches what was observed for LBM
+rational design process~\cite{parnas1986rational}.
+
+No interviewee mentioned a strictly defined project management process. 
+% JC: I don't think this level of per-project detail makes sense in this
+% review article - it should be about stats and trends, not reporting on
+% minute details. We could refer the reader to Ao's thesis for the details.
+% 
+% The
+% most common approach was following the issues, such as bugs and feature
+% requests. The \textit{3D Slicer} team had weekly meetings to discuss
+% the goals for the project. The \textit{INVESALIUS 3} team relied on the
+% GitHub process for their project management while the \textit{ITK-SNAP} 
+% team had a fixed six-month release pace. Only the interviewee from 
+% \textit{OHIF} 
+% mentioned the team having a project manager. The \textit{3D Slicer} team and
+% \textit{BioImage Suite Web} team do nightly builds and tests. The \textit{OHIF}
+% developer believes that a better project management process can improve junior
+% developer efficiency while also improving internal and external communication.
+
+Less than half of the projects used unit testing, even
+though the interviewees believed that testing (including usability tests) was
+the top approach to improve correctness, usability, and reproducibility.  This
+level of testing matches what was observed for LBM
 software [citation redacted for blind review]
@@
 and is apparently greater than the level of
-testing for ocean modelling software \cite{JungEtAl2022}.
-
-Interviewees from all of the eight projects thought that documentation was
+testing for ocean modelling software~\cite{JungEtAl2022}.
+
+All Interviewees thought that documentation was
 essential to their projects, and most of them (7 of 8) said that it could save
@@
 their documentations conveyed information clearly enough.
-
-As Table~\ref{Tbl_Guidelines} suggests, an emphasis on documentation, especially
-for new developers, is echoed in research software guidelines. Multiple
-guidelines recommend a document explaining how to contribute to a project, often
-named CONTRIBUTING. Guidelines also recommend tutorials, user guides and quick
-start examples. Smith and Roscoe \cite{SmithAndRoscoe2018} suggest including instructions
-specifically for on-boarding new developers. For open-source software in general
-(not just research software), \cite{Fogel2005} recommends providing tutorial
-style examples, developer guidelines, demos, and screenshots.
 

### 2024-11-08 StateOfPractice/Papers/SOP_MI_CritReviewsBioMedEng/SOP_MI_OSR.tex
partway done section 8

@@
 \usepackage{paralist}
+\usepackage[inline]{enumitem}
 
@@
 developers what specific best practices are taken to address the pain points
-and software quality concerns? (Section~\ref{painpoints})
+(Section~\ref{painpoints})
+and software quality concerns? (Section~\ref{sec:sq})
 \item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Recommend}:]
@@
 
-Based on interviews with nine developers (described in
-Section~\ref{sec_interview_methods}), we answer three research questions (first
-mentioned in Section~\ref{sec_motivation}): \rqref{RQ_PainPoints}) What are the
-pain points for developers working on research software projects?;
-\rqref{RQ_ComparePainPoints}) How do the pain points of developers from MI
-compare to the pain points for research software in general?; and
-\rqref{RQ_Concerns}) For MI developers what specific best practices are taken to
-address the pain points and software quality concerns?
-
-Our interviews identified pain points related to a lack of time and funding,
-technology hurdles, improving correctness, and improving usability.  In this
-section, we go through each pain point and contrast the MI experience with
-observations from other domains.  We also cover potential ways to address the
-pain points, as promoted by the community.  In addition to pain points, we summarize
-MI developer strategies for improving maintainability and reproducibility.
-Although the interviewees did not explicitly identify these two qualities as
-pain points, we did discuss threats to these qualities and ways to improve them
-as part of our interview process \cite{SmithEtAl2021}.  The interviewee's
-practices for addressing pain points and improving quality can potentially be
-emulated by other MI developers. Moreover, these practices may provide examples
-that can be followed by other research software domains.
-
-Pinto et al.\ \cite{PintoEtAl2018} list some pain points that did not come up in
-our conversations with MI developers: interruptions while coding, scope bloat,
-lack of user feedback, hard to collaborate on software projects, and aloneness.
-Wiese et al.\ \cite{WieseEtAl2019} mention two additional pain points that did
-not explicitly arise in our interviews: reproducibility, and software scope
-determination.  To the list of pain points not discussed for MI, our study of
-LBM software [citation redacted for blind review] %\cite{SmithEtAl2024} 
-adds lack of software experience for the developers, technical debt, and
-documentation. We did not observe any pain points for MI that were not also
-observed for LBM. From the pain points mentioned above, although the topics of
-reproducibility and technical debt did not come up in our MI interviews, we
-covered these two topics as part of the discussion of software qualities, as
-summarized at the end of this section.  Although previous studies show pain
-points that were not mentioned by MI developers, we cannot conclude that these
-pain points are not relevant for MI software development, since we only
-interviewed nine developers.
+We draw out the point points the points experienced by MI
+software developers, compare them to those of general research software
+developers, and what practices are used to address these pain points.
+
+Interviews identified the following issues:
+\begin{enumerate*}
+\item lack of time and funding,
+\item technology hurdles, 
+\item correctness, and 
+\item usability.  
+\end{enumerate*}
+We first detail each pain point, and contrast this experience with that
+from other domains. We then separately cover potential way to mitigate these
+issues.
+
+Pinto et al.~\cite{PintoEtAl2018} list some pain points that did not come up in
+our conversations: interruptions while coding, scope bloat, lack of user
+feedback, hard to collaborate on software projects, and aloneness.  Wiese et
+al.~\cite{WieseEtAl2019} mentions two more: reproducibility, and software scope
+determination.  [citation redacted for blind review] %\cite{SmithEtAl2024} 
+adds lack of software experience, technical debt, and documentation.
+We cannot conclude that these are not also issues with MI software due to our
+small response size.
 
@@
 \item[P\refstepcounter{pnum}\thepnum \label{P_LackDevTime}:] \textbf{Lack of
-Development Time:} Many interviewees thought lack of time, along with lack of
-funding (discussed next), were their most significant obstacles. Other domains
-of research software also experience the lack of time pain point
-\cite{PintoEtAl2018, PintoEtAl2016, WieseEtAl2019}. Our study of LBM software [citation redacted for blind review]
+Development Time:} This was felt to be the most significant obstacle, and
+is not uncommon
+\cite{PintoEtAl2018, PintoEtAl2016, WieseEtAl2019}.
 %\cite{SmithEtAl2024}
-also highlighted lack of time as a significant pain point.
-
-Potential and proven solutions suggested by the interviewees include:
-
+
+\item[P\refstepcounter{pnum}\thepnum \label{P_LackFunding}:] \textbf{Lack of
+Funding:} As research software is not yet seen as essential infrastructure,
+there is an ongoing struggle to attract funding and, even harder, maintain
+research software. This is a common complaint%
+\cite{GewaltigAndCannon2012, Goble2014, KaterbowAndFeulner2018, SmithEtAl2024}.
+As software output is not always counted when judging the academic excellence
+of applicants, which forces researchers who write software to spend extra
+time on publicity~\cite{WieseEtAl2019}. In some domains, like biology%
+\cite{HowisonAndBullard2016}, software is infrequently cited even when used
+pervasively. In other words, there is a lack of a formal reward system for
+research software~\cite{PintoEtAl2018}.
+
+\item[P\refstepcounter{pnum}\thepnum \label{P_TechnologyHurdles}:]
+\textbf{Technology Hurdles:} Developers mentioned: hard to keep up with changes
+in OS and libraries, difficult to transfer to new technologies, hard to support
+multiple OSes, and hard to support lower-end
+computers. Developers expressed difficulty balancing between four factors:
+cross-platform compatibility, convenience of development and maintenance,
+performance, and security.
+
+\item[P\refstepcounter{pnum}\thepnum \label{P_Correctness}:]
+\textbf{Correctness:} The most mentioned thread to correctness was complexity.
+Example sources of complexity are: Interviewees identified multiple threats to
+the large variety of data formats, complicated data standards, differing
+outputs between medical imaging machines, and the addition of (non-viewing
+related) functionality.  Other identified threats to correctness:
+
+\begin{itemize}
+\item Lack of real world image data for testing, in part because of patient
+privacy concerns;
+\item Tests are expensive and time-consuming because of the need for huge datasets;
+\item Software releases are difficult to manage;
+\item No systematic unit testing; and,
+\item No dedicated quality assurance team.
+\end{itemize}
+
+\item[P\refstepcounter{pnum}\thepnum \label{P_Usability}:] \textbf{Usability:}
+The discussion with the developers focused on usability issues for two classes
+of users: the end users and other developers.  The threats to usability for end
+users include an unintuitive user interface, inadequate feedback from the
+interface (such as lack of a progress bar), users being unable to determine the
+purpose of the software, not all users knowing if the software includes certain
+features, not all users understanding how to use the command line tool, and not
+all users understanding that the software is a web application. For developers
+the threats to usability include not being able to find clear instructions on
+how to deploy the software, and the architecture being difficult for new
+developers to understand.
+
+At least to some extent the problems for MI software users are due to holes in
+their background knowledge.  For research software in general, users do not
+always have the expertise required to install or use the software
+\cite{WieseEtAl2019}. A similar pattern has been observed during interviews with
+LBM software developers [citation redacted for blind review].
+%\cite{SmithEtAl2024}.
+
+\end{enumerate}
+
+For each of the above, the MI developers suggested various potential
+solutions, some of which have proven their worth in other domains.
+
+\begin{enumerate}
+\item[P\ref{P_LackDevTime}:] \textbf{Lack of Development Time:}
 \begin{itemize}
@@
 
-\item[P\refstepcounter{pnum}\thepnum \label{P_LackFunding}:] \textbf{Lack of
-Funding:} Developers felt the pain of having to attract funding to develop and
-maintain their software. For instance, the interviewees from \textit{3D Slicer}
-and \textit{OHIF} said getting funding for software maintenance is more
-challenging than finding funding for research. The interviewee from 
-\textit{ITK-SNAP} thought more funding was a way to solve the lack of time
-problem, because they could hire more dedicated developers. On the other hand,
-the interviewee from \textit{Weasis} did not feel that funding could
-solve the same problem, since they would still need time to supervise the project.
-
-Funding challenges have also been noted by others \cite{GewaltigAndCannon2012,
-Goble2014, KaterbowAndFeulner2018, SmithEtAl2024}. Researchers that devote time
-to software have the additional challenge that funding agencies do not always
-count software when they are judging the academic excellence of the applicant.
-Wiese et al.\ \cite{WieseEtAl2019} reported developer pains related to
-publicity, since publishing norms have historically made it difficult to get
-credit for creating software.  As studied by \cite{HowisonAndBullard2016},
-research software (specifically biology software, but the trend likely applies
-to other research software domains) is infrequently cited. There is a lack of a
-formal reward system for research software \cite{PintoEtAl2018}.
-
-An interviewee proposed an idea for increasing funding: Licensing the software
+\item[P\ref{P_LackFunding}:] \textbf{Lack of Funding:}
+One interviewee proposed an interesting idea: Licensing the software
 to commercial companies to integrate it into their products.
-
-\item[P\refstepcounter{pnum}\thepnum \label{P_TechnologyHurdles}:]
-\textbf{Technology Hurdles:} The technology hurdles mentioned by MI developers
-include: hard to keep up with changes in OS and libraries, difficult to transfer
-to new technologies, hard to support multiple OSes, and hard to support lower-end
-computers. Developers expressed difficulty balancing between four factors:
-cross-platform compatibility, convenience to development and maintenance,
-performance, and security.
-
-Another pain point survey \cite{WieseEtAl2019} highlights that technology
-hurdles are an issue for research software in general.  Some technical-related
-problems mentioned include dependency management, cross-platform compatibility,
-CI, hardware issues and operating system issues \cite{WieseEtAl2019,
-PintoEtAl2018}. Technology pain points for LBM developers include setting up
-parallelization and CI [citation redacted for blind review].
-%\cite{SmithEtAl2024}.
-
-Five solutions proposed by the MI developers are:
-
+In general, solutions need to be more systemic, and we are seeing the rise
+of ``research software engineers'' in various jurisdictions.
+
+\item[P\ref{P_TechnologyHurdles}:] \textbf{Technology Hurdles:}
 \begin{itemize}
@@
 
-As the above list shows, developers perceive that web-based applications will
-address the technology hurdle.  Most of the 29 teams (83\%) chose to develop
-native applications. For the eight teams we interviewed, three of them were
-building web applications, and the \textit{MRIcroGL} team was considering a
-web-based solution.
-
-The advantage for native applications is higher performance, while web
-applications have the advantage of cross-platform compatibility and a simpler
-build process.  These web advantages mirror the native disadvantages of
-difficulty with cross-platform compatibility and a complex build process.  The
-lower performance disadvantage of web applications can be improved with a server
-backend, but in this case there are disadvantages for privacy protection and
-server costs.  These issues are discussed further in the recommendations
-(Section~\ref{sec_webapps}).
-
-\item[P\refstepcounter{pnum}\thepnum \label{P_Correctness}:]
-\textbf{Ensuring Correctness:} Interviewees identified multiple threats to
-correctness.  The most frequently mentioned threat was complexity.  Complexity
-enters the software by various means, including the large variety of data formats,
-complicated data standards, differing outputs between medical imaging machines,
-and the addition of (non-viewing related) functionality.  Other threats to
-correctness identified include the following:
-
-\begin{itemize}
-\item Lack of real world image data for testing, in part because of patient
-privacy concerns (the pain point of privacy concerns also arises for research
-software in general \cite{WieseEtAl2019});
-\item Tests are expensive and time-consuming because of the need for huge datasets;
-\item Software releases are difficult to manage;
-\item No systematic unit testing; and,
-\item No dedicated quality assurance team.
-\end{itemize}
-
-As implied by the above threats to correctness, testing was the most often
-mentioned strategy for MI developers for ensuring correctness.  Seven teams
+A number of developers perceive that some of the ``shifting technologies''
+problems could be alleviated by moving to web-based applications.
+Most of the teams (83\%) chose to develop
+native applications. Three of the eight we interviewed were
+building web applications, and another team was considering it too.
+
+\item[P\ref{P_Correctness}:] \textbf{Correctness:}
+Testing was the most often strategy for ensuring correctness, and teams
 mentioned test related activities, including test-driven development, component
 tests, integration tests, smoke tests, regression tests, self tests and
-automated tests.  With the common emphasis on testing to improve correctness, MI
-software is ahead of some other scientific domains.  For research software in
-general, there is a problem of insufficient testing \cite{PintoEtAl2018} and
-more developers think testing is important than the number that believe they
-have a sufficient understanding of testing concepts \cite{HannayEtAl2009}.  Our
-study of LBM software suggests that this domain shares the challenges of
-insufficient testing and insufficient understanding of testing concepts
-[citation redacted for double blind]. %\cite{SmithEtAl2024}.  
-Automated testing
-is a specific challenge for LBM software since free testing services do not
-offer adequate facilities for large amounts of data.
-Although not specifically mentioned during our interviews, the large data sets
-for MI likely also cause a challenge for using free testing services, like
-GitHub Actions.
-
-Research software in general often struggles with the oracle problem for testing
-because for many potential test cases the developer doesn't have a means to
-judge the correctness of their calculated solutions \cite{HannayEtAl2009,
-KanewalaAndBieman2013, KellyEtAl2011, WieseEtAl2019}.  The MI developers did not
-allude to this challenge, likely because for a give image (test case) it is
-possible to determine, potentially by using other software, the expected
-analysis results.
-
-A frequently cited strategy for building confidence in correctness (mentioned by
-3 interviewees) is a two state development process with stable releases and
-nightly builds.  Other strategies for ensuring correctness that came up during
-the interviews include CI/CD, using de-identified copies of medical images for
-debugging, sending beta versions to medical workers who can access the data to
-do the tests, and collecting/maintaining a dataset of problematic images.  Some
-additional strategies used by MI developers include:
-
-\begin{itemize}
-\item Using open datasets.
-\item If (part of) the team belongs to a medical school or a hospital, using the
-datasets they can access;
-\item If the team has access to MRI scanners, self-building sample images for
-testing;
-\item If the team has connections with MI equipment manufacturers, asking for
-their help on data format problems;
-\end{itemize}
-
-The feedback from the interviewees makes it clear that increased connections
-between the development team and medical professionals/institutions could ease
-the pain of ensuring correctness via testing.
-
-\item[P\refstepcounter{pnum}\thepnum \label{P_Usability}:] \textbf{Usability:}
-The discussion with the developers focused on usability issues for two classes
-of users: the end users and other developers.  The threats to usability for end
-users include an unintuitive user interface, inadequate feedback from the
-interface (such as lack of a progress bar), users being unable to determine the
-purpose of the software, not all users knowing if the software includes certain
-features, not all users understanding how to use the command line tool, and not
-all users understanding that the software is a web application. For developers
-the threats to usability include not being able to find clear instructions on
-how to deploy the software, and the architecture being difficult for new
-developers to understand.
-
-At least to some extent the problems for MI software users are due to holes in
-their background knowledge.  For research software in general, users do not
-always have the expertise required to install or use the software
-\cite{WieseEtAl2019}. A similar pattern has been observed during interviews with
-LBM software developers [citation redacted for blind review].
-%\cite{SmithEtAl2024}.
-
-To improve the usability of MI software, the most common strategies mentioned by
-developers are as follows:
+automated tests.  This puts MI software ahead of other domains where
+insufficient testing is a problem~\cite{PintoEtAl2018} and
+insufficiently well-understood~\cite{HannayEtAl2009}.
+
+Another correctness strategy mentioned multiple times is a development process
+that involves stable releases and nightly builds.  Other strategies that were
+mentioned:
+\begin{enumerate*}
+\item using CI/CD, 
+\item using de-identified copies of medical images for
+debugging, 
+\item sending beta versions to medical workers who can access the data to
+do the tests, 
+\item collecting/maintaining a dataset of problematic images,
+\item using open datasets,
+\item if (part of) the team belongs to a medical school or a hospital, using the
+datasets they can access,
+\item if the team has access to MRI scanners, self-building sample images for
+testing, and
+\item if the team has connections with MI equipment manufacturers, asking for
+their help on data format problems.
+\end{enumerate*}
+
+Synthesizing the feedback from the interviewees, it is clear that increased
+connections between the development team and medical professionals/institutions
+could improve testing effectiveness and thus increase correctness.
+
+\item[P\ref{P_Usability}:] \textbf{Usability:}
 
@@
 
+\section{Software Qualities}\label{sec:sq}
+In addition to pain points, we summarize
+MI developer strategies for improving maintainability and reproducibility.
+Although the interviewees did not explicitly identify these two qualities as
+pain points, we did discuss threats to these qualities and ways to improve them
+as part of our interview process \cite{SmithEtAl2021}.  The interviewee's
+practices for addressing pain points and improving quality can potentially be
+emulated by other MI developers. Moreover, these practices may provide examples
+%HERE
+that can be followed by other research software domains.
 Up to this point, we have covered the pain points that came up in interviews

### 2024-11-08 StateOfPractice/Papers/SOP_MI_CritReviewsBioMedEng/SOP_MI_OSR.tex
to the end of the paper

@@
 At least to some extent the problems for MI software users are due to holes in
-their background knowledge.  For research software in general, users do not
-always have the expertise required to install or use the software
-\cite{WieseEtAl2019}. A similar pattern has been observed during interviews with
-LBM software developers [citation redacted for blind review].
+their knowledge of software and computing technology, and sometimes also
+lack expertize in their domain to be able to adequately use the software
+\cite{WieseEtAl2019}. 
+%A similar pattern has been observed during interviews with
+%LBM software developers [citation redacted for blind review].
 %\cite{SmithEtAl2024}.
@@
 \begin{itemize}
-    \item Use documentation (user manuals, mailing lists, forums) (mentioned by
-    4 developers)
-    \item Usability tests and interviews with end users; and, (mentioned by 3
-    developers)
-    \item Adjusting the software according to user feedback. (mentioned by 3
-    developers)
+    \item Use documentation (user manuals, mailing lists, forums) 
+    \item Usability tests and interviews with end users; and, 
+    \item Adjusting the software according to user feedback.
 \end{itemize}
@@
 \section{Software Qualities}\label{sec:sq}
-In addition to pain points, we summarize
-MI developer strategies for improving maintainability and reproducibility.
-Although the interviewees did not explicitly identify these two qualities as
-pain points, we did discuss threats to these qualities and ways to improve them
-as part of our interview process \cite{SmithEtAl2021}.  The interviewee's
-practices for addressing pain points and improving quality can potentially be
-emulated by other MI developers. Moreover, these practices may provide examples
-%HERE
-that can be followed by other research software domains.
-Up to this point, we have covered the pain points that came up in interviews
-with MI developers, along with a summary of the techniques that are currently
-used to address these pain points.  Although the developers did not explicitly
-identify the qualities of maintainability and reproducibility as pain points in
-our interviews, as part of our interview questions
-(Section~\ref{sec_interview_methods}) they did share their approaches for
-improving these qualities, as discussed below.
+
+The structured interviews also included an active discussion of software
+qualities, and in particular maintainability and reproducibility. Note that
+these did not come up as ``pain points''.
+%\cite{SmithEtAl2021}.  
+The developers discussed their approaches for improving these qualities, which
+we now relay.
 
@@
 rated as the third most important software quality for research software
-\cite{Nguyen-HoanEtAl2010}. The push for sustainable software
-\cite{deSouzaEtAl2019} is motivated by the pain that past developers have had
-with accumulating too much technical debt \cite{KruchtenEtAl2012}.  For LBM
-software, technical debt has been identified as one of the developer pain points
-[citation redacted for blind review]. %\cite{SmithEtAl2024}.
-
-To improve maintainability, the most popular (with five out of nine interviewees
-mentioning it) strategy is to use a modular approach, with often repeated
-functions in a library.  Other strategies that were mentioned for improving
-maintainability include supporting third-party extensions, an easy-to-understand
-architecture, a dedicated architect, starting from simple solutions, and
-documentation.  The \textit{3D Slicer} team used a well-defined structure for
-the software, which they named as an ``event-driven MVC pattern''.  \textit{3D
-Slicer} discovers and loads necessary modules at runtime. The \textit{BioImage
-Suite Web} team had designed and re-designed their software multiple times in
-the last 10+ years. They found that their modular approach effectively supports
-maintainability \cite{Joshi2011}.
+\cite{Nguyen-HoanEtAl2010}, and developers complain about 
+accumulating too much technical debt \cite{KruchtenEtAl2012}.
+
+The main strategy is to reduce code duplication by using properly structured
+libraries.  Other strategies mentioned include supporting third-party
+extensions, an easy-to-understand architecture, a dedicated architect, starting
+from simple solutions, and documentation.
 
 \item[Q\refstepcounter{qnum}\theqnum \label{Q_Reproducibility}:]
-\textbf{Reproducibility:}  Although the MI developers did not mention
-reproducibility explicitly as a pain point, they did mention the need to improve
-documentation.  Good documentation does not just address the pain points of lack
-of developer time (\ppref{P_LackDevTime}), technology hurdles
-(\ppref{P_TechnologyHurdles}), usability \ppref{P_Usability}, and
-maintainability.  Documentation is also necessary for reproducibility. The
-challenges of inadequate documentation are a known problem for research software
-\cite{PintoEtAl2018, WieseEtAl2019} and for non-research software
-\cite{LethbridgeEtAl2003}.
-
-In our interviews, we discussed threats to reproducibility and strategies for
-improving it.  The threats that were mentioned include closed-source software,
-no user interaction tests, no unit tests, the need to change versions of some
-common libraries, variability between CPUs, and misinterpretation of how
-manufacturers create medical images.
-
-The most commonly cited (by 6 teams) strategy to improve reproducibility was
+\textbf{Reproducibility:}  
+An important precondition for reproducibity is increased and better
+documentation (which also helps with many of the afore mentioned pain points).
+The challenges of inadequate documentation are a known problem for research
+software \cite{PintoEtAl2018, WieseEtAl2019} and for non-research software
+\cite{LethbridgeEtAl2003} alike.
+
+The threats to reproducibility that were mentioned include closed-source
+software, no user interaction tests, no unit tests, the need to change versions
+of some common libraries, variability between CPUs, and misinterpretation of
+how manufacturers create medical images.
+
+The most common strategy proposed was
 testing (regression tests, unit tests, having good tests). The second most
-common strategy (mentioned by 5 teams) is making code, data, and documentation
+common strategy is making code, data, and documentation
 available, possibly by creating open-source libraries.  Other ideas that were
@@
 
-In this section we provide recommendations to address the pain points from
-Section~\ref{painpoints} to answer~\rqref{RQ_Recommend}.  Our recommendations
-are not lists of criticisms for what should have been done in the past, or what
-should be done now; they are suggestions for consideration in the future. We
-expand on some of the ideas that came out of our interviews with developers
-(Section~\ref{painpoints}), including continuous integration, moving to web
-applications, and enriching the test data sets. We also bring in new like
-employing linters, peer review, and design for change.  Our aim is to mention
-ideas that are at least somewhat beyond conventional best practices. The ideas
-listed here have the potential to become best practices in the medium to
-long-term. We list the ideas roughly in the order of increasing implementation
-effort.
+We provide specific recommendations for future consideration -- these are
+not criticisms of past and/or current development practices. We
+expand on some of the ideas from our interviews, and bring in further ideas
+that could have a noticable impact. The ``new'' ideas are not novel to this
+paper but rather from the software engineering and software carpentry domains
+that have yet to have seen wide adoption.  We list the ideas roughly in the
+order of increasing implementation effort.
 
@@
 
-Continuous integration involves frequent pushes to a code repository.  With
-every push the software is built and tested \cite[p.\ 13]
-{HumbleAndFarley2010}, \cite{ShahinEtAl2017, Fowler2006}.  CI can take
-significant time and effort to set up and integrate into a team's workflow, but
-the benefits are impressive, as follows:
-
+Continuous integration involves building and testing the software every
+time a coherent change is made to the code (i.e. via a Pull Request)%
+\cite[p.\ 13] {HumbleAndFarley2010}, \cite{ShahinEtAl2017, Fowler2006}. Initial
+setup can be cumbersome, but
+the benefits are impressive:
 \begin{itemize}
@@
     automated testing.  
-    \item Everyone is always working on a stable base, since the rejection of
-    inadequate commits means that the main branch will always be working.  A
-    stable base will always pass all tests.  If the CI system uses generators
-    and linters, it will also have current documentation and standard compliant
-    code.  A stable base improves developer productivity, allowing them to focus
-    on coding, testing, and documentation.
+    \item Everyone is always working on a stable base that always passes all
+    the tests.
+    \item If the CI system uses generators and linters, it will also have
+    current documentation and clean code.
 \end{itemize}
 
-Continuous integration and delivery helps with addressing several pain points.
-For instance, CI/CD helps reduce development time (\ppref{P_LackDevTime}) by
-removing the need for a time-consuming integration stage and by automating
-regression testing.  Automated regression tests also help with ensuring
-correctness (\ppref{P_Correctness}) and the quality of reproducibility
-(\qref{Q_Reproducibility}). Details on CI for MI can be found in [citation
-removed for blind review] %Smith et al. \cite{SmithEtAl2024_MI_SOP} 
-, including an overview of the elements of CI, how to enable incorporation of CI
-into a team's workflow, how to set-up a CI system, and challenges for CI.
-
-\subsection{Move To Web Applications} \label{sec_webapps}
-
-Section~\ref{painpoints} describes the pain point of technology hurdles
-(\ppref{P_TechnologyHurdles}), which motivates considering the use of web
-applications. Here we give further advice to help with deciding whether to adopt
-a web application. The decision will be based on whether, on balance, the web
-application improves the four factors identified by developers: compatibility,
-maintainability, performance, and security. To enable decision-making, a team
-will need to prioritize between these factors, based on their objectives and
-experience. The suggestions are intended to provide ideas and avenues for
-exploration; a web application will not be the right fit for all projects and
-all teams.
+CI helps relieve several point points:
+\begin{enumerate*}
+\item by eliminating the time-consuming integration phase, 
+freeing more time for development (\ppref{P_LackDevTime}),
+\item Automated tests help with ensuring
+correctness (\ppref{P_Correctness}) and reproducibility
+(\qref{Q_Reproducibility}). 
+\end{enumerate*}
+
+\subsection{Perhaps Move To Web Applications} \label{sec_webapps}
+
+Multiple pain points encountered, especially technology hurdles
+(\ppref{P_TechnologyHurdles}), could be alleviated by moving to a web
+application. But this must be an active decision as it may not be a good
+fit in all cases.  The decision needs to be based on whether, on balance, this
+would improve the four factors identified earlier: compatibility,
+maintainability, performance, and security. We outline the factors to consider
+to enable better decision making, as a final decision crucially depends on
+each teams' objectives.
 
@@
 
-\item \textbf{Modern technologies may improve frontend performance.} Web
-applications with only a frontend usually perform worse than native
-applications. However, new technologies may ease this difference. For example,
+\item \textbf{Modern technologies may improve frontend performance.} 
+Complex, graphics-heavy user-interfaces often do not respond as quickly
+as native applications.  However, new technologies may help. For example,
 some JavaScript libraries can help the frontend harness the power of the
@@
 
-\item \textbf{Backend servers can potentially deliver high performance.} Web
-applications with backend servers may perform even better than native
-applications. If a team needs to support lower-end computers, it is good to use
-back-end servers for heavy computing tasks.  For backend servers where traffic
-and latency is not an issue, options include
+\item \textbf{Backend servers can potentially deliver high performance.} If
+the principal bottleneck is computational, backend servers may outperform
+native applications, as long as latency is not an issue.
+Options include
 \href{https://www.django-rest-framework.org/} {Django}, \href{https://laravel.com/} {Laravel} and
-\href{https://nodejs.org/en/} {Node.js}.  The advantage of Django is that it provides access to Python
-libraries.  For backend servers where traffic and latency is an issue,
+\href{https://nodejs.org/en/} {Node.js}.
+When traffic and latency are an issue,
 \href{https://github.com/gin-gonic/gin} {Gin} is an option.
 
-\item \textbf{Backend servers can have low costs.} Serverless solutions from
-major cloud service providers (like Amazon Web Services (AWS) and Google Cloud
-Platform) may be worth exploring. Serverless solutions still use a server, but
-the server provider only charges the team when they use the server. The solution
-is event-driven, and costs the team by the number of requests processed. Thus,
-serverless can be very cost-effective for less intensively used functions.
+\item \textbf{Some backend servers can be inexpensive.} Serverless solutions
+from cloud service providers (like Amazon Web Services (AWS) and Google Cloud
+Platform) may be worth exploring. These still use a server, but only actual use
+is charged.
 
 \item \textbf{Web transmission may diminish security.} Transferring sensitive
-data on-line can be a problem for projects requiring high security. Regulations
-for some MI applications may forbid doing web transmissions. In this case, a web
-application with a backend may not be an option.
+data on-line can be a problem for projects requiring high security, even
+if using encryption. Some jurisdiction may not allow transmission of some
+health data at all.
 
@@
 
-As described in Section~\ref{painpoints}, ensuring correctness
-(\ppref{P_Correctness}) via testing can be problematic because of limited access
-to real-world medical imaging datasets.  We build on the suggestions we heard
-from our interviewees as follows:
+Access to real-world imaging datasets can be a testing bottleneck,
+jeopardising correctness. Building on developers' suggestions:
 
 \begin{itemize}
-\item \textbf{Build and maintain good connections to datasets.} A team can build
-connections with professionals working in the medical domain, who may have
+\item \textbf{Build and maintain good connections to datasets.} 
+Teams can build
+connections with medical professionals, who may have
 access to private datasets and can perform tests for the team.
 
-\item \textbf{Collect and maintain datasets over time.} A team may face problems
-caused by various unique inputs over the years of software development. This
-data should be collected and maintained over time to form a good, comprehensive,
-dataset for testing.
-
-\item \textbf{Search for open data sources.} In general, there are many open MI
-datasets.  For instance, there are
+\item \textbf{Collect and maintain datasets over time.} It is especially
+important to have access to ``unique'' inputs that were outliers in
+historical testing scenarios.
+
+\item \textbf{Search for open data sources.} There are many open MI
+datasets:
 \href{https://nihcc.app.box.com/v/ChestXray-NIHCC}{Chest X-ray Datasets} by
@@
 
-\item \textbf{Create sample data for testing.} If a team can access tools
-creating sample data, they may also self-build datasets for testing. For
-example, an MI software development team can use an MRI scanner to create images
-of objects, animals, and volunteers. The team can build the images based on
-specific testing requirements.
-
-\item \textbf{Remove privacy from sensitive data.} For data with sensitive
-information, a team can ask the data owner to remove such information or add
-noise to protect privacy. One example is using de-identified copies of medical
-images for testing.
+\item \textbf{Create sample data for testing.} If a team is able to create
+baseline and/or synthetic data sets with known characteristics, this is a
+worthwhile investment (and should hopefully be made public). For example, using
+an MRI scanner to create images of objects, animals, and volunteers who waive
+the privacy rights to their image (i.e. not patients).
+
+%JC: I'm aware of too many caveats wrt this recommendation, so I am taking
+% it out.
+%
+%\item \textbf{Remove privacy from sensitive data.} For data with sensitive
+%information, a team can ask the data owner to remove such information or add
+%noise to protect privacy. One example is using de-identified copies of medical
+%images for testing.
 
@@
 
-\subsection{Employ Linters} \label{Sec_Linters}
-
-A linter is a tool that statically analyzes code to find programming errors,
+\subsection{Use Linters and other Static Analysis Tools} \label{Sec_Linters}
+
+Linters are simple tools that analyzes code to find programming errors,
 suspicious constructs, and stylistic inconsistencies \cite{Wikipedia2022_Lint}.
-Linters can be used as an ad hoc check for code files, but they really come into
-their own when used as part of a CI system, as discussed in
-Section~\ref{Sec_ContinuousIntegration}. Almost none of the research software
-guidelines that we consulted, summarized in Section~\ref{Sec_CompareArtifacts},
-mention linters.  Despite the lack of mention in the guidelines, we believe that
-linters have the potential to improve code quality at a relatively low cost.
-
-Linters have the following benefits: finding potential bugs, finding memory
+While they can be used in an ad hoc fashion, coupling them with CI is where
+the largest gain happens. Automatic use of code quality tools is increasing
+commonplace for industrial software -- it is regretable that the
+research software guidlines have no caught up to this.
+
+These tools have many benefits: finding potential bugs, finding memory
 leaks, improving performance, standardizing code with respect to formatting,
 removing silly errors before code reviews, and catching potential security
-issues \cite{SourceLevel2022_Lint}. Most popular programming languages have an
-accompanying linter.  For example, Python has the options of PyLint, flake8 and
-Black \cite{Zadka2018}.
-
-Linters address several MI pain points.  For instance, linters address the lack
-of development time (\ppref{P_LackDevTime}) by increasing the developer's
-productive time via guarding against frustrating, time-consuming, and mundane
-mistakes.  Moreover, since a linter can include rules that capture the wisdom of
-senior programmers, it can help newer developers avoid common mistakes. With
-respect to the technology hurdle pain point (\ppref{P_TechnologyHurdles}),
-linters can assist with the move toward web applications
-(Section~\ref{sec_webapps}).  For instance, ESLint in React is a pluggable
-linter that lets the developer know if they have imported something and not used
-it, if a function could be short-handed, if there are indentation
-inconsistencies, etc. By insisting on code
-standardization, linters can reduce technical debt and thus improve
-maintainability (\qref{Q_Maintainability}). Although linters are tools for code
-analysis, the idea of statically checking for adherence to basic rules can be
-extended to check documentation. Tools to enforce documentation standards
+issues~\cite{SourceLevel2022_Lint}. Most popular programming languages possess
+such tools. This helps with \ppref{P_LackDevTime}, \qref{Q_Maintainability},
+and correctness (\ppref{P_Correctness}).
+
+The pervasive use of such tools to enforce documentation standards
 partially explains the relatively higher quality of statistical tools that are
@@
 
-\subsection{Conduct a Mix of Rigorous and Informal Peer Reviews} \label{Sec_PeerReview}
-
-We advocate incorporating peer review into the development process, as
-frequently recommended for research software \cite{HerouxEtAl2008,
-OrvizEtAl2017, USGS2019}. In most cases a modern, lightweight review, should be
-adequate.  Modern code review is informal, tool-based, asynchronous, and focused
-on reviewing code changes \cite{SadowskiEtAl2018}. Managing a project via GitHub
-pull requests is an example of a modern approach to reviewing code.  Software
-development organizations have moved to this lightweight style of code review
-because of the inefficiencies of rigorous inspections \cite{RigbyAndBird2013}.
-However, for important parts of the code, developers may benefit from mixing in
-a more rigorous approach.
-
-Fagan \cite{Fagan1976} began work on rigorous review via code inspection.
-Elements of a typical inspection include reviewing the code against a checklist
-(checking the consistency of variable names, look for terminating loops, etc.),
-performing specific review tasks (such as summarizing the code's purpose,
-cross-referencing the code to the technical manual, creating a data dictionary
-for a given module, etc.) Rigorous inspection finds 60-65\% of latent defects on
+\subsection{Conduct Peer Reviews} \label{Sec_PeerReview}
+
+Peer review is recommended by many \cite{HerouxEtAl2008,
+OrvizEtAl2017, USGS2019}. Modern peer review processes are lightweight,
+informal, tool-based, asynchronous, and focused
+on reviewing code changes~\cite{SadowskiEtAl2018}. GitHub, for example,
+provides such tools for its pull requests.
+
+Rigorous inspection finds 60-65\% of latent defects on
 average, and often tops 85\% in defect removal efficiency \cite{Jones2008}. The
@@
 Due to improving code quality and increasing knowledge transfer, peer review
-addresses the same pain points and qualities as linters
-(Section~\ref{Sec_Linters}): \ppref{P_LackDevTime}, \ppref{P_TechnologyHurdles},
-and \qref{Q_Maintainability}. Peer review can potentially find misunderstandings
-in how the code implements the required theory, which will improve the
-software's correctness (\ppref{P_Correctness}). The benefits of peer review for
-addressing pain points can be increased by extending the review from just code,
-to also reviewing all software artifacts, including documentation, build
+addresses the same pain points and qualities as linters \ppref{P_LackDevTime},
+\qref{Q_Maintainability} and \ppref{P_Correctness}. The benefits can be
+increased by extending reviews to all artifacts, including documentation, build
 scripts, test cases and the development process itself.
@@
 
-We recommend that MI developers use the principle of design for change for their
-modularizations.  Although the advice to modularize research software to handle
+ Although the advice to modularize research software to handle
 complexity is common \cite{WilsonEtAl2014, StewartEtAl2017, Storer2017},
@@
 
-Developers can accomplish design for change by first identifying likely changes,
-either implicitly or explicitly, and second by hiding each likely change behind
-a well-defined module interface \cite{Parnas1972a}.
-
 \section{Threats to Validity} \label{sec_threats_to_validity}
@@
 We identify the following threats:
-
-\begin{itemize}
+\begin{enumerate*}
 \item One individual does the measures for all packages. A different
@@
 the reviewer's judgement may have drifted.
-\end{itemize}
+\end{enumerate*}
 
 The measurement process used has previously been shown to be reasonably
-reproducible.  [citation redacted] 
+reproducible, [citation redacted]
 %% DOUBLE BLIND
 %Smith et al.~\cite{SmithEtAl2016} 
-reports grading five software
-products by two reviewers. Their rankings were almost identical. As long as each
-grader uses consistent definitions, the relative comparisons in the AHP results
-will be consistent between graders.
+reports grading five software products by two reviewers. Their rankings were
+almost identical. As long as each grader uses consistent definitions, the
+relative comparisons in the AHP results will be consistent between graders.
 
@@
 112]{VanVliet2000}.
-\item We could not install or build \textit{dwv}, \textit{GATE}, and
-\textit{DICOM Viewer}. We used a deployed on-line version for \textit{dwv}, a
-virtual machine version for \textit{GATE}, but no alternative for \textit{DICOM
-Viewer}.
+\item We could not do all measurements for some software as we could not
+install or build \textit{dwv}, \textit{GATE}, and \textit{DICOM Viewer}. We
+used a deployed on-line version for \textit{dwv}, a virtual machine version for
+\textit{GATE}, but no alternative for \textit{DICOM Viewer}.
 \item Robustness measurements involve only two pieces of data, leading to
@@
 (contributors' guide, code of conduct, code style guidelines, product roadmap,
-design documentation, and API documentation)
-(Section~\ref{Sec_CompareArtifacts}) were rare.
+design documentation, and API documentation) were rare.
 
@@
 increasing documentation, increasing testing by enriching datasets, increasing
-modularity, using continuous integration, moving to web applications, employing
+modularity, using continuous integration, moving to web applications, using
 linters, increasing peer review, and using design for change to guide
@@
 A deeper understanding of the needs of the MI community will require data beyond
-what is available in repositories.  Future work should involve interviewing MI
-developers to better understand their ``pain points''.
+what is available in repositories.
 

### 2024-11-09 StateOfPractice/Papers/SOP_MI_CritReviewsBioMedEng/SOP_MI_OSR.tex
Edited up to, but not including, Section 7

@@
 
-We present the state of the practice for the development of Medical Imaging
-(MI) software based on data available in open source repositories. We selected
-29 projects from 48 candidates and assessed 9 software qualities
-(installability, correctness/ verifiability, reliability, robustness,
-usability, maintainability, reusability, understandability, and
-visibility/transparency) by answering 108 questions for each. Using the
-Analytic Hierarchy Process (AHP) on the quantitative data, we ranked the MI
-software.  The top five are \textit{3D Slicer}, \textit{ImageJ}, \textit{Fiji},
-\textit{OHIF Viewer}, and \textit{ParaView}.  This is consistent with the
-community's view, with four of these also appearing in the top five using
-GitHub metrics (stars-per-year).  The quality and quantity of documentation
-present in a project correlates quite well with its popularity.  Generally, MI
-software is in a healthy state: in the repositories
-we observed 88\% of the documentation artifacts recommended by research
-software development guidelines, 100\% of MI projects use version control
-tools, and developers appear to use the common quasi-agile research software
-development process. However, the current state of the practice deviates from
-existing guidelines as some recommended artifacts are rarely present
-(like a test plan, requirements specification, code of conduct, and code style
-guidelines), low usage of continuous integration (17\% of the projects), low
-use of unit testing (about 50\% of projects), and room for improvement with
-documentation. From developper interviews , we identified 7 concerns: lack
-of development time, lack of funding, technology hurdles, correctness,
-usability, maintainability, and reproducibility. We recommend:
-increase effort on documentation,
-increase testing by enriching datasets, increase continuous integration usage,
-move to web applications, employ linters, use peer reviews, and design for
-change.
+We present the state of the practice for the development of Medical Imaging (MI)
+software based on data available in open source repositories. We selected 29
+projects from 48 candidates and assessed 9 software qualities (installability,
+correctness/verifiability, reliability, robustness, usability, maintainability,
+reusability, understandability, and visibility/transparency) by answering 108
+questions for each. Using the Analytic Hierarchy Process (AHP) on the
+quantitative data, we ranked the MI software.  The top five are \textit{3D
+Slicer}, \textit{ImageJ}, \textit{Fiji}, \textit{OHIF Viewer}, and
+\textit{ParaView}.  This is consistent with the community's view, with four of
+these also appearing in the top five using GitHub metrics (stars-per-year).  The
+quality and quantity of documentation present in a project correlates quite well
+with its popularity.  Generally, MI software is in a healthy state: in the
+repositories we observed 88\% of the documentation artifacts recommended by
+research software development guidelines, 100\% of MI projects use version
+control tools, and developers appear to use the common quasi-agile research
+software development process. However, the current state of the practice
+deviates from existing guidelines as some recommended artifacts are rarely
+present (like a test plan, requirements specification, code of conduct, and code
+style guidelines), low usage of continuous integration (17\% of the projects),
+low use of unit testing (about 50\% of projects), and room for improvement with
+documentation. From developer interviews, we identified 7 concerns: lack of
+development time, lack of funding, technology hurdles, correctness, usability,
+maintainability, and reproducibility. We recommend: increase effort on
+documentation, increase testing by enriching datasets, increase continuous
+integration usage, move to web applications, employ linters, use peer reviews,
+and design for change.
 
@@
 
-As well as state of the practice (SOP) for MI software, we would like to
-understand the impact of the often cited gap between recommended
-software engineering practices and that used for most research 
-software~\cite{Storer2017}. Although scientists spend
-a substantial proportion of their working times on software development
-\cite{Hannay2009, Prabhu2011}, few are formally trained~\cite{Hannay2009}.
-
-Our investigation is based on the following research questions:
+As well as state of the practice for MI software, we would like to understand
+the impact of the often cited gap between recommended software engineering
+practices and the practices used for most research software~\cite{Storer2017}. Although
+scientists spend a substantial proportion of their working times on software
+development \cite{Hannay2009, Prabhu2011}, few are formally
+trained~\cite{Hannay2009}.
+
+Our investigation is based on ten research questions:
 \begin{enumerate}
@@
 often called ``ilities.''  For this study we selected 9 qualities to measure:
-installability, correctness/ verifiability, reliability, robustness, usability,
+installability, correctness/verifiability, reliability, robustness, usability,
 maintainability, reusability, understandability, and visibility/transparency.
@@
 the interviewees, and how they organize their projects.  We also ask about the
-developer's understanding of the users. Some questions focus on the current and
+developer's understanding of the users. Additional questions focus on the current and
 past difficulties, and the solutions the team has found, or plan to try. We also
@@
 have underestimated the scores on reliability and robustness for \textit{DICOM
-Viewer}, but we compared it equally for the other seven qualities.
+Viewer}, but we compared it equally for the other qualities.
 
@@
 correctness and verifiability used a wider array of techniques to improve
-correctness, and had better documentation to witness this.  For instance, we
+correctness, and had better documentation to support this. For instance, we
 looked for evidence of unit testing, and found evidence for only about half of
@@
 Our ranking and GitHub popularity, at least for the top five projects, seems to
-line up fairly well. However, we ranked some popular packages fairly low, such
+line up well. However, we ranked some popular packages fairly low, such
 as \textit{dwv}. This is because we were unable to build it locally, even though
@@
 
-We weighted all qualities equally, which is not the likely (implicit) weight
+We weighted all qualities equally, which is not the likely the same weighting
 that users implicitly use. To properly assess this would require a broad user
-study.  Furthermore our measures of popularity are only \emph{proxies} which
+study.  Furthermore our measures of popularity (like stars) are only \emph{proxies} which
 are biased towards past rather than current preferences~\cite{Szulik2017}, as
 these are monotonically increasing quantities. Finally there are often more
-factors than just quality that influence the popularity of ``consumer''
+factors than just quality that influence the popularity of 
 products.
@@
 appears in the measured set of MI software, either not at all (blank), commonly
-(C), uncommonly (U) or rarely (R).  We did our best to interpret the meaning of
-each artifact consistently between guidelines and specific MI software, but the
-terminology and the contents of artifacts are not standardized.  The challenge
-even exists for the ubiquitous README file.  The content of README files shows
-significant variation between projects~\cite{PranaEtAl2018}.  Some content is
-reasonably consistent, with 97\% of README files contain at least one section
-describing the `What' of the repository and 89\% offering some `How' content,
-other categories are more variable.  For instance, information on
-`Contribution', `Why', and `Who', appear in 28\%, 26\% and 53\% of the analyzed
-files, respectively~\cite{PranaEtAl2018}.
+(C) (20 to 29 ($>$67\%) packages), uncommonly (U) (10 to 19 (33-67\%) packages)
+or rarely (R) (1 to 9 ($<$33\%) packages).  We did our best to interpret the
+meaning of each artifact consistently between guidelines and specific MI
+software, but the terminology and the contents of artifacts are not
+standardized.  The challenge even exists for the ubiquitous README file.  The
+content of README files shows significant variation between
+projects~\cite{PranaEtAl2018}.  Some content is reasonably consistent, with 97\%
+of README files contain at least one section describing the `What' of the
+repository and 89\% offering some `How' content, other categories are more
+variable.  For instance, information on `Contribution', `Why', and `Who', appear
+in 28\%, 26\% and 53\% of the analyzed files, respectively~\cite{PranaEtAl2018}.
 
@@
 
-Table~\ref{artifactspresent} presents our measurements for MI software.
-The table groups the artifacts by frequency
-into categories of common (20 to 29 ($>$67\%) packages), uncommon (10 to 19
-(33-67\%) packages), and rare (1 to 9 ($<$33\%) packages).
-
-Note that ``popularity'' in Table~\ref{Tbl_Guidelines} does not imply that these
-oft recommended artifacts are the most important. Guidelines are often brief,
-to encourage adoption, and thus even guidelines that mention the need for
-installation instructions rarely mention uninstallation instructions.
-Two items in Table~\ref{artifactspresent} do not appear in any guidelines:
-\emph{Troubleshooting guide} and \emph{Developer's manual}.  However the
-information within these documents overlaps with
-the recommended artifacts.  Troubleshooting information often can be found in
-a User Manual, while the information in a ``Developer's Manual'' is often
-scattered amongst many other documents.
+Popularity in Table~\ref{Tbl_Guidelines} does not imply that these oft
+recommended artifacts are the most important. Guidelines are often brief, to
+encourage adoption, and thus even guidelines that mention the need for
+installation instructions rarely mention uninstallation instructions.  Two items
+in Table~\ref{artifactspresent}, which presents our measurements for the MI
+software, do not appear in any guidelines: \emph{Troubleshooting guide} and
+\emph{Developer's manual}.  However the information within these documents
+overlaps with the recommended artifacts.  Troubleshooting information often can
+be found in a User Manual, while the information in a ``Developer's Manual'' is
+often scattered amongst many other documents.
 
@@
 requirements are impossible for research software \cite{CarverEtAl2007,
-SegalAndMorris2008}. Fortunately a more agile approach to requirements is
+SegalAndMorris2008}. Fortunately an iterative approach to requirements is
 feasible~\cite{Smith2016}, and research-software specific templates
@@
 
-Developers use software tools for user support, version control (VCS),
+Developers use software tools for user support, version control,
 Continuous Integration and Deployment (CI/CD), documentation and project
@@
 was reported as generally positive; some teams had actively migrated to GitHub
-from other VCS.  This represents a considerable uptake since its rare use in
-2006~\cite{Wilson2006}, which matches the uptake in other communities: 10 years
-ago Nguyen et al~\cite{Nguyen-HoanEtAl2010} estimated that only 50\% of
-research software projects used version control, and by 2018 this had jumped to
-over 80\%~\cite{AlNoamanyAndBorghi2018} with many communities~\cite{Smith2018}
-reaching 100\% already in 2018. In 2022, version control use sits at over
-95\%~\cite{CarverEtAl2022}.
+from other version control systems.  This represents a considerable uptake since
+its rare use in 2006~\cite{Wilson2006}, which matches the uptake in other
+communities: 10 years ago Nguyen et al~\cite{Nguyen-HoanEtAl2010} estimated that
+only 50\% of research software projects used version control, and by 2018 this
+had jumped to over 80\%~\cite{AlNoamanyAndBorghi2018} with many
+communities~\cite{Smith2018} reaching 100\% already in 2018. In 2022, version
+control use sits at over 95\%~\cite{CarverEtAl2022}.
 
@@
 \cite{BrettEtAl2021, Brown2015, ThielEtAl2020, Zadka2018, vanGompelEtAl2016}.
-A recent survey~\cite{CarverEtAl2022} of practioners (rather than projects)
+A recent survey~\cite{CarverEtAl2022} of practitioners (rather than projects)
 suggests higher use of CI/CD with 54\% of respondents indicating that they use
@@
 
-\paragraph{Project management} Two types of tools were mentionned:
+\paragraph{Project management} Two types of tools were mentioned:
 \begin{inparaenum}[i)]
@@
 Of the named tools, interviewees mentioned GitHub 3 times, and each
-of the other tools once each.
-
-Based on~\cite{JungEtAl2022}, tool use for MI
-software is similar to tool use for ocean modelling software.
-Both use tools for editing, compiling, code management, testing, building, and
-project management, but differ
-in the latter's use of Kanban boards for project management.
+of the other tools once.
+
+Tool use for MI software is similar to tool use for ocean modelling
+software~\cite{JungEtAl2022}.  Both use tools for editing, compiling, code
+management, testing, building, and project management, but differ in the
+latter's use of Kanban boards for project management.
 

### 2024-11-09 StateOfPractice/Papers/SOP_MI_CritReviewsBioMedEng/SOP_MI_OSR.tex
Revised Section 7

@@
 feeling their process was ``similar'' to an existing model.  Three teams
-followed agile or alike, two teams  followed waterfall or alike, and three
-teams explicitly stated that their process was undefined or self-directed.
+followed agile or agile-like, two teams followed waterfall or waterfall-like,
+and three teams explicitly stated that their process was undefined or
+self-directed.
 
@@
 
-No interviewee mentioned a strictly defined project management process. 
+No interviewee mentioned a strictly defined project management process. The most
+common approach was issue-driven via bug reports and feature requests.
+[Redacted for blind review] %Dong~\cite{Dong2021} 
+provides details on the specific approaches used by the interviewees.
 % JC: I don't think this level of per-project detail makes sense in this
@@
 % 
+% SS: I've revised the text to not include the details
 % The
@@
 
-All Interviewees thought that documentation was
+All interviewees thought that documentation was
 essential to their projects, and most of them (7 of 8) said that it could save
-their time to answer questions from users and developers. Half of them (4 of 8)
-saw the need to improve their documentation, and only three of them thought that
+time spent answering questions from users and developers. Half of them (4 of 8)
+saw the need to improve their documentation, and only three of them thought 
 their documentations conveyed information clearly enough.

### 2024-11-09 StateOfPractice/Papers/SOP_MI_CritReviewsBioMedEng/SOP_MI_OSR.tex
Updates to section 8

@@
 
-We draw out the point points the points experienced by MI
-software developers, compare them to those of general research software
-developers, and what practices are used to address these pain points.
-
 Interviews identified the following issues:
 \begin{enumerate*}
-\item lack of time and funding,
+\item lack of time,
+\item lack of funding,
 \item technology hurdles, 
@@
 \cite{PintoEtAl2018, PintoEtAl2016, WieseEtAl2019}.
-%\cite{SmithEtAl2024}
+%DOUBLE BLIND \cite{SmithEtAl2024} [redacted for blind review]
 
@@
 Funding:} As research software is not yet seen as essential infrastructure,
-there is an ongoing struggle to attract funding and, even harder, maintain
-research software. This is a common complaint%
-\cite{GewaltigAndCannon2012, Goble2014, KaterbowAndFeulner2018, SmithEtAl2024}.
-As software output is not always counted when judging the academic excellence
-of applicants, which forces researchers who write software to spend extra
-time on publicity~\cite{WieseEtAl2019}. In some domains, like biology%
-\cite{HowisonAndBullard2016}, software is infrequently cited even when used
-pervasively. In other words, there is a lack of a formal reward system for
+there is an ongoing struggle to attract funding to build and, even harder, maintain
+research software. This is a common complaint~\cite{GewaltigAndCannon2012, Goble2014, KaterbowAndFeulner2018}. %, SmithEtAl2024 %[redacted for double blind review]
+Software output is not always counted when judging the academic excellence of
+academics, which forces researchers who write software to spend extra time on
+publicity~\cite{WieseEtAl2019}. In some domains, like
+biology~\cite{HowisonAndBullard2016}, software is infrequently cited even when
+used pervasively. In other words, there is a lack of a formal reward system for
 research software~\cite{PintoEtAl2018}.
@@
 \item[P\refstepcounter{pnum}\thepnum \label{P_TechnologyHurdles}:]
-\textbf{Technology Hurdles:} Developers mentioned: hard to keep up with changes
-in OS and libraries, difficult to transfer to new technologies, hard to support
-multiple OSes, and hard to support lower-end
+\textbf{Technology Hurdles:} Developers mentioned the following challenges: hard
+to keep up with changes in OS and libraries, difficult to transfer to new
+technologies, hard to support multiple OSes, and hard to support lower-end
 computers. Developers expressed difficulty balancing between four factors:
@@
 \textbf{Correctness:} The most mentioned thread to correctness was complexity.
-Example sources of complexity are: Interviewees identified multiple threats to
-the large variety of data formats, complicated data standards, differing
-outputs between medical imaging machines, and the addition of (non-viewing
-related) functionality.  Other identified threats to correctness:
+Example sources of complexity include: a large variety of data formats,
+complicated data standards, differing outputs between medical imaging machines,
+and the addition of (non-viewing related) functionality.  Other identified
+threats to correctness:
 
@@
 their knowledge of software and computing technology, and sometimes also
-lack expertize in their domain to be able to adequately use the software
+lack of expertize in their domain to be able to adequately use the software
 \cite{WieseEtAl2019}. 
@@
 to commercial companies to integrate it into their products.
-In general, solutions need to be more systemic, and we are seeing the rise
+In general, solutions need to be more systemic, thus we are seeing the rise
 of ``research software engineers'' in various jurisdictions.
@@
 native applications. Three of the eight we interviewed were
-building web applications, and another team was considering it too.
+building web applications, and another team was considering it.
 
 \item[P\ref{P_Correctness}:] \textbf{Correctness:}
-Testing was the most often strategy for ensuring correctness, and teams
-mentioned test related activities, including test-driven development, component
+Testing was the most often mentioned strategy for ensuring correctness.  Teams
+mentioned several test-related activities, including test-driven development, component
 tests, integration tests, smoke tests, regression tests, self tests and
@@
 that involves stable releases and nightly builds.  Other strategies that were
-mentioned:
+mentioned include:
 \begin{enumerate*}

### 2024-11-10 StateOfPractice/Papers/SOP_MI_CritReviewsBioMedEng/SOP_MI_OSR.tex
Revisiong to Section 9 (Software Qualities)

@@
 
-Synthesizing the feedback from the interviewees, it is clear that increased
-connections between the development team and medical professionals/institutions
-could improve testing effectiveness and thus increase correctness.
-
 \item[P\ref{P_Usability}:] \textbf{Usability:}
@@
 The structured interviews also included an active discussion of software
-qualities, and in particular maintainability and reproducibility. Note that
-these did not come up as ``pain points''.
-%\cite{SmithEtAl2021}.  
-The developers discussed their approaches for improving these qualities, which
-we now relay.
+qualities, in particular maintainability and reproducibility. This
+discussion did not come up as ``pain points'', but rather through
+planned questions we asked for improving these qualities.
 
@@
 
-The main strategy is to reduce code duplication by using properly structured
+The main strategy mentioned to reduce code duplication is using properly structured
 libraries.  Other strategies mentioned include supporting third-party
@@
 \textbf{Reproducibility:}  
-An important precondition for reproducibity is increased and better
-documentation (which also helps with many of the afore mentioned pain points).
+An important precondition for reproducibility is increased and better
+documentation (which also helps with many of the aforementioned pain points).
 The challenges of inadequate documentation are a known problem for research

### 2024-11-12 StateOfPractice/Papers/SOP_MI_CritReviewsBioMedEng/SOP_MI_OSR.tex
Changed title of Section 9 to Improving Software Qualities
Revisions to section 10 (Recommendations)

@@
 
-\section{Software Qualities}\label{sec:sq}
+\section{Improving Software Qualities}\label{sec:sq}
 
@@
 
-We provide specific recommendations for future consideration -- these are
-not criticisms of past and/or current development practices. We
-expand on some of the ideas from our interviews, and bring in further ideas
-that could have a noticable impact. The ``new'' ideas are not novel to this
-paper but rather from the software engineering and software carpentry domains
-that have yet to have seen wide adoption.  We list the ideas roughly in the
-order of increasing implementation effort.
+We provide specific recommendations for future consideration -- these are not
+criticisms of past and/or current development practices. We expand on some of
+the ideas from our interviews, and bring in further ideas that could have a
+noticeable impact. The ``new'' ideas are not novel to this paper but rather from
+the software engineering and software carpentry domains, but have not yet seen
+wide adoption.  We list the ideas roughly in the order of increasing
+implementation effort.
 
@@
 
-Multiple pain points encountered, especially technology hurdles
+Multiple pain points, especially technology hurdles
 (\ppref{P_TechnologyHurdles}), could be alleviated by moving to a web
-application. But this must be an active decision as it may not be a good
+application. However, this must be an active decision as it may not be a good
 fit in all cases.  The decision needs to be based on whether, on balance, this
 would improve the four factors identified earlier: compatibility,
-maintainability, performance, and security. We outline the factors to consider
-to enable better decision making, as a final decision crucially depends on
-each teams' objectives.
+maintainability, performance, and security. The following considerations may
+help a team make a decision about the move to web applications:
 
@@
 
-\subsection{Enrich the Testing Datasets} 
-\label{sec_recommendations_testing_dataset}
+\subsection{Enrich the Testing Datasets} \label{sec_recommendations_testing_dataset}
 
@@
 % it out.
+%SS: okay, I agree.
 %
@@
 
-Linters are simple tools that analyzes code to find programming errors,
+Linters are simple tools that analyze code to find programming errors,
 suspicious constructs, and stylistic inconsistencies \cite{Wikipedia2022_Lint}.
@@
 the largest gain happens. Automatic use of code quality tools is increasing
-commonplace for industrial software -- it is regretable that the
-research software guidlines have no caught up to this.
+commonplace for industrial software -- it is regrettable that the
+research software guidelines have no caught up to this.
 
@@
 prevalent.  Not every decomposition is a good design for supporting change, as
-shown by \cite{Parnas1972a}.  For instance, a design with low cohesion and high
+shown by Parnas~\cite{Parnas1972a}.  For instance, a design with low cohesion and high
 coupling \cite[p.\ 48]{GhezziEtAl2003} will make change difficult. Especially in

### 2024-11-15 StateOfPractice/Papers/SOP_MI_CritReviewsBioMedEng/SOP_MI_OSR.tex
Edits to Section 10 on Recommendations

@@
 noticeable impact. The ``new'' ideas are not novel to this paper but rather from
-the software engineering and software carpentry domains, but have not yet seen
-wide adoption.  We list the ideas roughly in the order of increasing
-implementation effort.
+the software engineering and software carpentry domains, but these ideas have
+not yet seen wide adoption for MI software.  We list the ideas roughly in the
+order of increasing implementation effort.
 
@@
 
-Continuous integration involves building and testing the software every
-time a coherent change is made to the code (i.e. via a Pull Request)%
-\cite[p.\ 13] {HumbleAndFarley2010}, \cite{ShahinEtAl2017, Fowler2006}. Initial
-setup can be cumbersome, but
-the benefits are impressive:
+Continuous integration involves building and testing the software every time a
+coherent change is made to the code (i.e.\ via a pull request)~\cite[p.\ 13]
+{HumbleAndFarley2010}, \cite{ShahinEtAl2017, Fowler2006}. Initial setup can be
+cumbersome, but the benefits are impressive:
+
 \begin{itemize}
@@
     automated testing.  
-    \item Everyone is always working on a stable base that always passes all
-    the tests.
+    \item Everyone is always working on a stable base that always passes all tests.
     \item If the CI system uses generators and linters, it will also have
@@
 fit in all cases.  The decision needs to be based on whether, on balance, this
-would improve the four factors identified earlier: compatibility,
-maintainability, performance, and security. The following considerations may
-help a team make a decision about the move to web applications:
+would improve the four factors identified earlier for the technology hurdle:
+compatibility, maintainability, performance, and security. The following
+considerations may help a team make a decision about the move to web
+applications:
 
@@
 frameworks helping developers with cross-platform compatibility. For example,
-the \href{https://flutter.dev/}{Flutter} project enables support for web,
+the \href{https://flutter.dev/}{Flutter} project supports web,
 mobile, and desktop OS with one codebase.  Other options include
-\href{https://vuejs.org/} {Vue}, \href{https://angular.io/} {Angular} and
+\href{https://vuejs.org/} {Vue}, \href{https://angular.io/} {Angular},
 \href{https://reactjs.org/} {React}, and \href{https://elm-lang.org/}{Elm}.
@@
 % it out.
-%SS: okay, I agree.
+%SS: I agree with the removal.
 %
@@
 
-Rigorous inspection finds 60-65\% of latent defects on
+Rigorous inspection finds 60--65\% of latent defects on
 average, and often tops 85\% in defect removal efficiency \cite{Jones2008}. The
 success rate of code inspection is generally higher than most forms of testing,
-which average between 30 --- 35\% for defect removal efficiency
+which average between 30--35\% for defect removal efficiency
 \cite{EbertAndJones2009, Jones2008}. For research software, Kelly and Shepard
@@
 Due to improving code quality and increasing knowledge transfer, peer review
-addresses the same pain points and qualities as linters \ppref{P_LackDevTime},
-\qref{Q_Maintainability} and \ppref{P_Correctness}. The benefits can be
+addresses the same pain points and qualities as linters (\ppref{P_LackDevTime},
+\qref{Q_Maintainability} and \ppref{P_Correctness}). The benefits can be
 increased by extending reviews to all artifacts, including documentation, build

### 2024-11-15 StateOfPractice/Papers/SOP_MI_CritReviewsBioMedEng/SOP_MI_OSR.tex
Edits to Section 11 (Threats to Validity)

@@
 The measurement process used has previously been shown to be reasonably
-reproducible, [citation redacted]
+reproducible.
 %% DOUBLE BLIND
 %Smith et al.~\cite{SmithEtAl2016} 
-reports grading five software products by two reviewers. Their rankings were
-almost identical. As long as each grader uses consistent definitions, the
-relative comparisons in the AHP results will be consistent between graders.
+[citation redacted] reports grading five software products by two reviewers.
+Their rankings were almost identical. As long as each grader uses consistent
+definitions, the relative comparisons in the AHP results will be consistent
+between graders.
 

### 2024-11-15 StateOfPractice/Papers/SOP_MI_CritReviewsBioMedEng/LetterToEditor.tex
Submitted version of paper, letter to editor and copyright transfer

\documentclass[12pt]{casletter}

\usepackage{url}
\usepackage{hyperref}
\usepackage{xcolor}
\hypersetup{
    colorlinks,
    linkcolor={red!50!black},
    citecolor={blue!50!black},
    urlcolor={blue!80!black}
}

%\usepackage{fullpage}

% \oddsidemargin 0mm
% \evensidemargin 0mm
% \textwidth 160mm
% \textheight 180mm
%\renewcommand\baselinestretch{0.85}

\pagestyle {plain}
\pagenumbering {arabic}

\email{smiths@mcmaster.ca}
\telephone{(905) 525-9140 ext.\ 27929}
\www{{http://}www.cas.mcmaster.ca/\~\@smiths}

\signature{Spencer Smith}

\address{Dr.\ Spencer Smith, Associate Professor\\
Computing and Software Department\\
Faculty of Engineering\\
McMaster University\\
1280 Main Street West\\
Hamilton, Ontario, Canada L8S 4K1}%  \\ Phone: 525-9140 Ext.~27929 \\E-mail: smiths@mcmaster.ca}

\begin{document}

\begin{letter}{~\\
~\\
~\\
~\\
Dr.\ Anthony McGoron\\
Editor-in-Chief\\
Critical Reviews in Biomedical Engineering\\
~\\
{\bf Re: Review Article Submission}}

\opening {Dear Dr.\ Anthony McGoron:}

We are submitting for your consideration a review article titled ``State of the
Practice for Medical Imaging Software Based on Open Source Repositories''.  This
article has not been published elsewhere and it has not been simultaneously
submitted for publication elsewhere. We have published a significantly extended
paper on the same topic on the arXiv.

We review the state of the practice for the development of Medical Imaging (MI)
software based on data available in open source repositories. We identify the
open source packages available, rank them by their application of software
engineering best practices, compare MI software to other research software and
provide recommendations to potentially improve MI software quality going
forward.

We suggest the following reviewers:

\begin{itemize}
    \item \href{https://bme.ubc.ca/?directory=alexander-weber} {Dr.\ Alexander Weber}: \texttt{aweber@bcchr.ca}
    \item \href{https://www.robarts.ca/scholl_group/people/bio/akbari_alireza.html} {Dr.\ Alireza Akbari}: \texttt{aakbari@uwo.ca}
    \item \href{https://www.se.informatik.uni-kiel.de/en/team/prof.-dr.-wilhelm-willi-hasselbring} {Dr.\ Wilhelm Hasselbring}: \texttt{hasselbring@email.uni-kiel.de}
\end{itemize}

Thank you for considering our work for publication in your journal.  If you have
any questions or concerns, please let me know.

\closing{Best regards,~\newline}
\vspace{-29mm}
\includegraphics[scale=0.7]{Signature.pdf}

\end {letter}

\end{document}

### 2024-11-15 StateOfPractice/Papers/SOP_MI_CritReviewsBioMedEng/SOP_MI_OSR.tex
Revisions to conclusions
Submitted version of paper, letter to editor and copyright transfer

@@
 
-We present the state of the practice for the development of Medical Imaging (MI)
-software based on data available in open source repositories. We selected 29
-projects from 48 candidates and assessed 9 software qualities (installability,
-correctness/verifiability, reliability, robustness, usability, maintainability,
-reusability, understandability, and visibility/transparency) by answering 108
+We review the state of the practice for the development of Medical Imaging (MI)
+software based on data available in open-source repositories. We selected 29
+projects from 48 candidates and assessed 9 software qualities by answering 108
 questions for each. Using the Analytic Hierarchy Process (AHP) on the
@@
 these also appearing in the top five using GitHub metrics (stars-per-year).  The
-quality and quantity of documentation present in a project correlates quite well
+quality and quantity of documentation present in a project correlate quite well
 with its popularity.  Generally, MI software is in a healthy state: in the
-repositories we observed 88\% of the documentation artifacts recommended by
-research software development guidelines, 100\% of MI projects use version
-control tools, and developers appear to use the common quasi-agile research
-software development process. However, the current state of the practice
+repositories, we observed 88\% of the documentation artifacts recommended by
+research software development guidelines and 100\% of MI projects use version
+control tools. However, the current state of the practice
 deviates from existing guidelines as some recommended artifacts are rarely
-present (like a test plan, requirements specification, code of conduct, and code
+present (like a test plan, requirements specification, and code
 style guidelines), low usage of continuous integration (17\% of the projects),
@@
 development time, lack of funding, technology hurdles, correctness, usability,
-maintainability, and reproducibility. We recommend: increase effort on
-documentation, increase testing by enriching datasets, increase continuous
-integration usage, move to web applications, employ linters, use peer reviews,
-and design for change.
+maintainability, and reproducibility. We recommend: increasing effort on
+documentation, increasing testing by enriching datasets, increasing continuous
+integration, moving to web applications, employing linters, using peer reviews,
+and designing for change.
 
@@
 
-While we did find a reasonable amount of documentation, especially when consider
-all MI projects, there were definitely some holes. Some important documentation
-(test plans and requirements documentation) was missing, and other
+While we did find a reasonable amount of documentation, especially when we
+consider all MI projects, there were definitely some holes. Some important
+documentation (test plans and requirements documentation) was missing, and other
 (contributors' guide, code of conduct, code style guidelines, product roadmap,

### 2024-11-22 StateOfPractice/Papers/SOP_MI_CritReviewsBioMedEng/SOP_MI_OSR.tex
Revisions to remove overlap with another paper

@@
 (Section~\ref{Sec_VsCommunityRanking})
-\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareArtifacts}:] How
-do MI projects compare to general research software with respect to the
-artifacts (documents, scripts and code) present in their repositories?
+\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareArtifacts}:] How do
+the artifacts (documents, scripts and code) present in MI repositories compare
+to the artifacts used for research software in general?
 (Section~\ref{Sec_CompareArtifacts})
 \item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareToolsProjMngmnt}:] How
-do MI projects compare to research software in general with respect to the use
-of tools (Section~\ref{Sec_CompareTools}) for development
+does the use of development tools compare between MI software and research
+software in general (Section~\ref{Sec_CompareTools})
 (\rqref{RQ_CompareToolsProjMngmnt}.a); and, project management
@@
 \item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareMethodologies}:]
-How do MI projects compare to research software in general with respect to
-principles, processes, and methodologies used?
-(Section~\ref{Sec_CompareMethodologies})
-\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_PainPoints}:] What are
-the pain points for developers working on MI software projects?
+How does the use of principles, processes, and methodologies compare between MI software and research software in general? (Section~\ref{Sec_CompareMethodologies})
+\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_PainPoints}:] For MI software
+developers, what pain points do they experience?  (Section~\ref{painpoints})
+\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_ComparePainPoints}:] How do
+the pain points compare between MI software and research software in general?
 (Section~\ref{painpoints})
-\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_ComparePainPoints}:] How
-do the pain points of developers from MI compare to the pain points
-for research software in general? (Section~\ref{painpoints})
-\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Concerns}:] For MI
-developers what specific best practices are taken to address the pain points
-(Section~\ref{painpoints})
-and software quality concerns? (Section~\ref{sec:sq})
+\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Concerns}:] What best
+practices are taken by MI developers to address their pain points
+(Section~\ref{painpoints}) and quality concerns? (Section~\ref{sec:sq})
 \item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Recommend}:]
-What research software development practice could potentially address the
-pain point concerns identified in \rqref{RQ_PainPoints}?
+What processes, techniques and tools can potentially address the
+identified pain points from \rqref{RQ_PainPoints}?
 (Section~\ref{ch_recommendations})
@@
 
-We only cover MI visualization software.  We exclude other categories of MI
-software, including segmentation, registration, visualization, enhancement,
-quantification, simulation, plus MI archiving and telemedicine systems
-(compression, storage, and communication).  We also exclude statistical analysis
-and image-based physiological modelling and feature extraction, classification,
-and interpretation. Software that provides MI support functions is also out of
-scope; therefore, we have not assessed the toolkit libraries VTK and ITK.
-Finally, Picture Archiving and Communication System (PACS), which helps users to
-economically store and conveniently access images, are considered out of scope.
+Our scope is limited to MI visualization software.  We exclude other categories
+of MI software, including segmentation, registration, visualization,
+enhancement, quantification, simulation, plus MI archiving and telemedicine
+systems (compression, storage, and communication).  We also exclude statistical
+analysis and image-based physiological modelling and feature extraction,
+classification, and interpretation. Software that provides MI support functions
+is also out of scope; therefore, we have not assessed the toolkit libraries VTK
+and ITK.  Finally, Picture Archiving and Communication System (PACS), which
+helps users to economically store and conveniently access images, are considered
+out of scope.
 
@@
 filtered and vetted by a domain expert. We aim for roughly $30$ packages. For
-each remaining package, we measure its qualities by filling in a grading
+each remaining package, we measure its qualities via a grading
 template [citation redacted for double blind]. %%DOUBLE BLIND
@@
 The Domain Expert vets the proposed list because online resources can be
-inaccurate.  The expert also vets the AHP ranking.  For the current assessment,
-our Domain Expert is [details of our domain expert removed for double-blind].
+inaccurate.  The expert also vets the AHP ranking.  Our Domain Expert is [details of our domain expert removed for double-blind].
 %%DOUBLE BLIND %our Domain Expert (and paper co-author) is Dr.\ Michael
@@
 
-In advance of the first meeting with the Domain Expert, they are asked to
-independently create a list of top software packages in the domain.  This helps
-get the expert's knowledge refreshed in advance of the meeting.
+The first step for the Domain Expert is to independently create a list of their
+choices for the top software.  This refreshes the expert's knowledge in advance
+of the first meeting.
 
@@
 shows an excerpt of the measurement spreadsheet.  The rows are the measures and
-the columns correspond to the software packages.  [The full data is available on
+the columns correspond to the software packages.  [All data can be found on
 Mendeley; link will be provided after refereeing.] 
@@
 
-Although both rankings are imperfect measures, they nevertheless suggest a
-correlation between best practices and popularity. We don't know if this is
-causal, in either direction (i.e. if best practices enable popularity or if
-popularity increases the need for best practices).
+Although both rankings are imperfect, they still suggest a correlation between
+popularity and best practices. An open question remains on whether this
+correlation is causal in either direction. That is, do best practices enable
+popularity, or does popularity increases the need for best practices?
 
@@
 \begin{itemize}
-\item United States Geological Survey Software Planning Checklist
+\item Checklist for United States Geological Survey Software Planning
 \cite{USGS2019},
-\item DLR (German Aerospace Centre) Software Engineering Guidelines
+\item Software Engineering Guidelines for DLR (German Aerospace Centre)
 \cite{TobiasEtAl2018}, 
-\item Scottish Covid-19 Response Consortium Software Checklist
+\item Software Checklist for Scottish Covid-19 Response Consortium
 \cite{BrettEtAl2021},
 \item Good Enough Practices in Scientific Computing \cite{WilsonEtAl2016},
-\item xSDK (Extreme-scale Scientific Software Development Kit) Community Package
-Policies \cite{SmithAndRoscoe2018},
-\item Trilinos Developers Guide \cite{HerouxEtAl2008},
-\item EURISE (European Research Infrastructure Software Engineers') Network
-Technical Reference \cite{ThielEtAl2020},
-\item CLARIAH (Common Lab Research Infrastructure for the Arts and Humanities)
+\item Community Package Policies for xSDK (Extreme-scale Scientific Software
+Development Kit) \cite{SmithAndRoscoe2018},
+\item Developer's Guide for Trilinos \cite{HerouxEtAl2008},
+\item Network Technical Reference for EURISE (European Research Infrastructure
+Software Engineers') \cite{ThielEtAl2020},
+\item Common Lab Research Infrastructure for the Arts and Humanities (CLARIAH)
 Guidelines for Software Quality \cite{vanGompelEtAl2016}, and

### 2024-11-26 StateOfPractice/Papers/SOP_MI_CritReviewsBioMedEng/AddressConcerns_LetterToEditor.tex
Letter to editor revised submission

--- 2024-11-15 StateOfPractice/Papers/SOP_MI_CritReviewsBioMedEng/LetterToEditor.tex
@@
 ~\\
-{\bf Re: Review Article Submission}}
+{\bf Re: Revisions to Review Article}}
 
@@
 
-We are submitting for your consideration a review article titled ``State of the
-Practice for Medical Imaging Software Based on Open Source Repositories''.  This
-article has not been published elsewhere and it has not been simultaneously
-submitted for publication elsewhere. We have published a significantly extended
-paper on the same topic on the arXiv.
+Thank you for expressing your concerns. To address them, we are submitting a
+revised version of our paper ``State of the Practice for Medical Imaging
+Software Based on Open Source Repositories''.  The text has been modified to
+eliminate the textual similarity to our paper: Smith, S., Michalski, P.,
+Carette, J. et al.\ State of the Practice for Lattice Boltzmann Method Software.
+Archives of Computational Methods in Engineering 31, 313--350 (2024).
 
-We review the state of the practice for the development of Medical Imaging (MI)
-software based on data available in open source repositories. We identify the
-open source packages available, rank them by their application of software
-engineering best practices, compare MI software to other research software and
-provide recommendations to potentially improve MI software quality going
-forward.
-
-We suggest the following reviewers:
-
-\begin{itemize}
-    \item \href{https://bme.ubc.ca/?directory=alexander-weber} {Dr.\ Alexander Weber}: \texttt{aweber@bcchr.ca}
-    \item \href{https://www.robarts.ca/scholl_group/people/bio/akbari_alireza.html} {Dr.\ Alireza Akbari}: \texttt{aakbari@uwo.ca}
-    \item \href{https://www.se.informatik.uni-kiel.de/en/team/prof.-dr.-wilhelm-willi-hasselbring} {Dr.\ Wilhelm Hasselbring}: \texttt{hasselbring@email.uni-kiel.de}
-\end{itemize}
+The ``State of the Practice for Lattice Boltzmann Method Software'' paper is
+frequently cited in the non-redacted version of our submission. Those citations
+were redacted for the purpose of the double blind review.
 
 Thank you for considering our work for publication in your journal.  If you have
-any questions or concerns, please let me know.
+any further questions or concerns, please let me know.
 

### 2025-01-24 StateOfPractice/Papers/SOP_MI_CritReviewsBioMedEng/ResubmissionResponseToReviewer.tex
Preparing resubmission of SOP for MI for Crit Reviews in Bio Med Eng

\documentclass[12pt]{casletter}

\oddsidemargin -10mm
\evensidemargin -10mm
\textwidth 180mm
\textheight 200mm
%\renewcommand\baselinestretch{0.85}

\usepackage{xcolor}

\pagestyle {plain}
\pagenumbering {arabic}

\email{smiths@mcmaster.ca}
\telephone{(905) 525-9140 ext.\ 27929}
\www{{http://}www.cas.mcmaster.ca/\~\@smiths}

\signature{Spencer Smith}

\address{Dr.\ Spencer Smith, Associate Professor\\
Computing and Software Department\\
Faculty of Engineering\\
McMaster University\\
1280 Main Street West\\
Hamilton, Ontario, Canada L8S 4K1}%  \\ Phone: 525-9140 Ext.~27929 \\E-mail:
                                  %  smiths@mcmaster.ca}

\begin{document}

\begin{letter}{~\\
    ~\\
    ~\\
    ~\\
    ~\\
    Dr.\ Anthony McGoron\\
    Editor-in-Chief\\
    Critical Reviews in Biomedical Engineering\\
    ~\\
    {\bf Re: Revisions to CRB-57015, State of the Practice for Medical Imaging
    Software Based on Open Source Repositories}}

  \opening {Dear Dr.\ McGoron:}

  Thank you and the reviewers for the feedback on our submission.  The reviewers
  provided thoughtful and constructive comments.  In response to your e-mail,
  dated January 3, 2025, we have revised the paper to incorporate the requested
  revisions.  We provide a summary below.  The revised submission also includes
  a ``diff'' version of the paper showing the additions and deletions.

  Some of the reviewer's questions are best answered by citing our related work.
  Therefore, the new version undoes the redaction originally added for the
  double blind review.

  \textbf{Reviewer \#1}

  \begin{enumerate}
  \item The relative priorities of each software quality were set equal.
  Couldn't some different weighting have been considered? Perhaps one that comes
  from interviews with developers or cited in papers? \medskip

  \emph{Papers don't agree and papers are from many domains of application for
  research software.  \cite{Nguyen-HoanEtAl2010}  Not aware of a study focused
  on medical imaging practitioners.  Interviewing a statistically significant
  number of developers is beyond the scope of this paper.  However, developers
  can see the impact of a different weighting themselves.  Point to spreadsheet.
  Say what text is added to the paper.  Put it in blue?  \smallskip}

  \item In addition to github, could any other reference have been used, such as
  the number of citations in papers or relevant websites in the field? \medskip

  \emph{Citations and websites are a good idea, but don't have this data.
  Papers unreliable because people don't tend to cite scientific softwarwe.
  cite Katz to support this.  Changes to paper.\smallskip}

  \item Where are the quality ratings for each software? \medskip

  \emph{Double blind review meant didn't show citations, but citations have the
  full quality ratings.  The overall quality rating is given in Figure 2. The
  actual numbers are not as relevant as the ranking. \smallskip}

  \item Caption tables above them, not below them. \medskip

  \emph{This has been fixed in the revised manuscript.  \smallskip}

  \item What are the possible reasons for the differences between the ranking
  found and that of GitHub, in addition to the impossibility of installing some
  of them? \medskip

  \emph{Add content from the arxiv version.  ``Further reason for discrepancies
  between our ranking and the community's ranking is that we weighted all
  qualities equally. This is not likely how users implicitly rank the different
  qualities. As a result, some projects with high community popularity may have
  scored lower with our method because of a relatively higher (compared to the
  scientific community's implicit ranking) weighting of the poor scores for some
  qualities. A further explanation for discrepancies between our measures and
  the star measures may also be due to inaccuracy with using stars to
  approximate popularity.  Stars are not an ideal measure because stars
  represent the community's feeling in the past more than they measure current
  preferences \citep{Szulik2017}.  The issue with stars is that they tend only
  to be added, not removed.  A final reason for inconsistencies between our
  ranking and the community's ranking is that, as for consumer products, more
  factors influence popularity than just quality.'')\smallskip}

  \item Some software could not be installed. What problems were found? Why not
  remove it from the list? \medskip

  \emph{Say the problems from installability section.  Didn't remove because
  only affected a few qualities.  Favoured including with a warning.
  \smallskip}

  \end{enumerate}

  \textbf{Reviewer \#2}

  \begin{enumerate}

  \item The introduction section should contain the scope, significance of the
  research by summarizing current understanding and background information,
  stating the purpose of the work, and highlighting the potential outcomes.
  Also, there's no need to divide various subsection in the introduction.
  \medskip

  \emph{Response.\smallskip}

  \item The title of the table should be placed above the table. \medskip

  \emph{This has been fixed in the revised manuscript.  \smallskip}

  \item It is not recommended to include tables in the conclusion. \medskip

  \emph{Move it.  \smallskip}

  \end{enumerate}

  \closing{Best regards,~\newline} \vspace{-29mm}
  \includegraphics[scale=0.75]{Signature.pdf}

\end {letter}

\end{document}

### 2025-01-24 StateOfPractice/Papers/SOP_MI_CritReviewsBioMedEng/OrigSubmissionSOP_MI_OSR.tex
Preparing resubmission of SOP for MI for Crit Reviews in Bio Med Eng

--- 2024-11-22 StateOfPractice/Papers/SOP_MI_CritReviewsBioMedEng/SOP_MI_OSR.tex
@@
 We show the recommended artifacts in Table~\ref{Tbl_Guidelines}, one per row.
-For a given row, a checkmark in one of the columns means that the corresponding
-guideline recommends this artifact.  The last column shows whether the artifact
-appears in the measured set of MI software, either not at all (blank), commonly
-(C) (20 to 29 ($>$67\%) packages), uncommonly (U) (10 to 19 (33-67\%) packages)
-or rarely (R) (1 to 9 ($<$33\%) packages).  We did our best to interpret the
-meaning of each artifact consistently between guidelines and specific MI
-software, but the terminology and the contents of artifacts are not
-standardized.  The challenge even exists for the ubiquitous README file.  The
-content of README files shows significant variation between
-projects~\cite{PranaEtAl2018}.  Some content is reasonably consistent, with 97\%
-of README files contain at least one section describing the `What' of the
-repository and 89\% offering some `How' content, other categories are more
-variable.  For instance, information on `Contribution', `Why', and `Who', appear
-in 28\%, 26\% and 53\% of the analyzed files, respectively~\cite{PranaEtAl2018}.
+The columns correspond to guidelines. A checkmark in a column means that the
+guideline suggests the artifact in this row.  The final column shows the
+frequency of the artifact in the measured set of MI software.  The frequency
+ranges from not at all (blank), rarely (R) ($<$33\%), uncommonly (U) (33-67\%)
+to commonly (C) ($>$67\%).  Due to a lack of standardization, the mapping
+between specific MI software and guidelines was challenging, even for the
+ubiquitous README file.  Prana et al.~\cite{PranaEtAl2018} show significant
+variation in the content of README files between projects. Although 97\% of
+README files contain at least one section describing the `What' and 89\% offer
+content on `How', other content is more variable.  For example, information on
+`Who', `Contribution', and `Why', appear in 53\%, 28\%, 26\% of the analyzed
+files, respectively~\cite{PranaEtAl2018}.
 
@@
     \end{tabular}
-    \caption{Artifacts Present in MI Packages, Classified by Frequency
-    (number of occurrences)}
+    \caption{Software Artifacts Found in MI Packages, Classified by Their Number
+    of Occurrences}
     \label{artifactspresent}
@@
 
-Popularity in Table~\ref{Tbl_Guidelines} does not imply that these oft
-recommended artifacts are the most important. Guidelines are often brief, to
+Popularity in Table~\ref{Tbl_Guidelines} does not mean these 
+artifacts are more important than other artifacts. Guidelines are often brief, to
 encourage adoption, and thus even guidelines that mention the need for
@@
 documentation is recommended by some \cite{TobiasEtAl2018, HerouxEtAl2008,
-SmithAndKoothoor2016}, in practice this is rare~\cite{HeatonAndCarver2015}.
-Sanders and Kelly \cite{SandersAndKelly2008} interviewed 16 scientists from 10
-disciplines and found that none of the scientists created requirements
-specifications, unless regulations in their field mandated such a document.
-Requirements are the least commonly produced type of documentation for research
-software in general \cite{Nguyen-HoanEtAl2010}.
-
-This is unfortunate as when scientific developers are surveyed on their pain
-points, Wiese et al.~\cite{WieseEtAl2019} found that software requirements and
-management is the software engineering discipline that most hurts them,
-accounting for 23\% of the technical problems reported by study participants.
-Further adding to the misfortune, there is a widespread perception that up-front
-requirements are impossible for research software \cite{CarverEtAl2007,
-SegalAndMorris2008}. Fortunately an iterative approach to requirements is
-feasible~\cite{Smith2016}, and research-software specific templates
-exist~\cite{SmithEtAl2007}.
+SmithAndKoothoor2016}, in practice this is rare~\cite{HeatonAndCarver2015}.  In
+interviews with 16 scientists from 10 disciplines, the scientists admitted that
+they only wrote a requirements specification if the regulations in their field
+mandated it~\cite{SandersAndKelly2008}.  For research software in general,
+requirements are the least commonly written document~\cite{Nguyen-HoanEtAl2010}.
+
+This is unfortunate since when scientific developers are
+surveyed~\cite{WieseEtAl2019}, software requirements and management proves to be
+their greatest pain points, accounting for 23\% of their technical problems.
+Further adding to their misfortune, up-front requirements are often viewed to be
+impossible for research software \cite{CarverEtAl2007, SegalAndMorris2008}.
+Fortunately an iterative approach to requirements is feasible~\cite{Smith2016},
+and research-software specific templates exist~\cite{SmithEtAl2007}.
 
@@
 
-Other communities use checklists to help with best practices. Examples include
-checklists for merging branches~\cite{Brown2015}, for saving and sharing changes
-to the project~\cite{WilsonEtAl2016}, for new and departing team
-members~\cite{HerouxAndBernholdt2018}, for processes related to commits and
-releases~\cite{HerouxEtAl2008} and for overall software
+Other communities use checklists as part of best practices. For instance, there
+are checklists for commits and releases~\cite{HerouxEtAl2008}, for team
+turnover~\cite{HerouxAndBernholdt2018}, for branch merging~\cite{Brown2015}, for
+sharing and saving project changes~\cite{WilsonEtAl2016}, and for software
 quality~\cite{ThielEtAl2020, SSI2022}.
 
-MI projects fall somewhat short of recommended best practices, but are not alone
-amongst research software projects. This gap has been documented
+MI software does not follow all recommended best practices, but they are not
+alone amongst research software projects. This gap has been documented
 before~\cite{Storer2017,JohansonAndHasselbring2018}, and is known to cause
@@
 
-\section{Comparison of Tool Usage Between MI and Other Research Software}
+\section{Comparison Between MI and Other Research Software for Tool Usage}
 \label{Sec_CompareTools}
 
-We use data from interviews, as described in Section~\ref{sec_interview_methods}
+We interview data, as described in Section~\ref{sec_interview_methods}
 to answer \rqref{RQ_CompareToolsProjMngmnt} (tool usage),
@@
 its rare use in 2006~\cite{Wilson2006}, which matches the uptake in other
-communities: 10 years ago Nguyen et al~\cite{Nguyen-HoanEtAl2010} estimated that
-only 50\% of research software projects used version control, and by 2018 this
-had jumped to over 80\%~\cite{AlNoamanyAndBorghi2018} with many
-communities~\cite{Smith2018} reaching 100\% already in 2018. In 2022, version
-control use sits at over 95\%~\cite{CarverEtAl2022}.
+communities: 10 years ago only 50\% of research software used version
+control~\cite{Nguyen-HoanEtAl2010}, but by 2018 this jumped to over
+80\%~\cite{AlNoamanyAndBorghi2018} with many communities~\cite{Smith2018}
+reaching 100\% at this time. In 2022, version control use sits at over
+95\%~\cite{CarverEtAl2022}.
 
@@
 
-Tool use for MI software is similar to tool use for ocean modelling
-software~\cite{JungEtAl2022}.  Both use tools for editing, compiling, code
-management, testing, building, and project management, but differ in the
-latter's use of Kanban boards for project management.
+Tool use for MI software and ocean modelling software is
+similar~\cite{JungEtAl2022}.  Both use tools for code and project management,
+compiling and building, testing, and, editing. They differ in the more frequent
+use of project management via Kanban boards for ocean modelling.
 
@@
 
-This matches observations for research software in general.
-Scientific developers naturally use an agile philosophy \cite{AckroydEtAl2008,
+This matches previous observations for other research software.  Scientific
+software developers often use an agile philosophy \cite{AckroydEtAl2008,
 CarverEtAl2007, EasterbrookAndJohns2009, Segal2005, HeatonAndCarver2015}, or an
@@
 software~\cite{Smith2016}, especially if the developers work iteratively and
-incrementally, but externally document their work as if they followed a
-rational design process~\cite{parnas1986rational}.
+incrementally, but externally document their work as if they followed a rational
+design process~\cite{parnas1986rational}.
 
@@
 
-Pinto et al.~\cite{PintoEtAl2018} list some pain points that did not come up in
-our conversations: interruptions while coding, scope bloat, lack of user
-feedback, hard to collaborate on software projects, and aloneness.  Wiese et
-al.~\cite{WieseEtAl2019} mentions two more: reproducibility, and software scope
-determination.  [citation redacted for blind review] %\cite{SmithEtAl2024} 
-adds lack of software experience, technical debt, and documentation.
-We cannot conclude that these are not also issues with MI software due to our
-small response size.
+Pinto et al.~\cite{PintoEtAl2018} list other pain points not mentioned by our
+interviewees: scope bloat, aloneness, lack of user feedback, interruptions while
+coding, and collaboration challenges. Two more are~\cite{WieseEtAl2019}:
+reproducibility, and software scope determination.  [citation redacted for blind
+review] %\cite{SmithEtAl2024} 
+adds lack of software experience, documentation and technical debt.  Our small
+sample size means that we cannot conclude that these are not also issues with MI
+software.
 
@@
 the software engineering and software carpentry domains, but these ideas have
-not yet seen wide adoption for MI software.  We list the ideas roughly in the
-order of increasing implementation effort.
+not yet seen wide adoption for MI software.  We sorted the list of ideas roughly
+on increasing effort.
 
@@
 \begin{itemize}
-    \item Elimination of headaches associated with a separate integration phase
+    \item Remove headaches caused by a separate integration phase
     \cite{Fowler2006}, \cite[p.\ 20]{HumbleAndFarley2010}.
@@
     automated testing.  
-    \item Everyone is always working on a stable base that always passes all tests.
-    \item If the CI system uses generators and linters, it will also have
-    current documentation and clean code.
+    \item The base is constantly stable for everyone, always passing all tests.
+    \item The code will be standardized and the documentation current, if the CI
+    system uses generators and linters.
 \end{itemize}
@@
 These tools have many benefits: finding potential bugs, finding memory
-leaks, improving performance, standardizing code with respect to formatting,
-removing silly errors before code reviews, and catching potential security
+leaks, improving performance, creating standard compliant code,
+removing trivial errors before code reviews, and finding security
 issues~\cite{SourceLevel2022_Lint}. Most popular programming languages possess
@@
 
-The pervasive use of such tools to enforce documentation standards
-partially explains the relatively higher quality of statistical tools that are
-part of the Comprehensive R Archive Network (CRAN) [citation redacted for blind
-review].  %\cite{SmithEtAl2018_StatSoft}.
+The pervasive use of such tools to enforce documentation standards partially
+explains the relatively high quality of software in the Comprehensive R Archive
+Network (CRAN) [citation redacted for blind review].
+%\cite{SmithEtAl2018_StatSoft}.
 
@@
 
-Rigorous inspection finds 60--65\% of latent defects on
-average, and often tops 85\% in defect removal efficiency \cite{Jones2008}. The
-success rate of code inspection is generally higher than most forms of testing,
-which average between 30--35\% for defect removal efficiency
-\cite{EbertAndJones2009, Jones2008}. For research software, Kelly and Shepard
-\cite{KellyAndShepard2000} show a task based inspection approach can be
-effective.
-
-Due to improving code quality and increasing knowledge transfer, peer review
-addresses the same pain points and qualities as linters (\ppref{P_LackDevTime},
-\qref{Q_Maintainability} and \ppref{P_Correctness}). The benefits can be
-increased by extending reviews to all artifacts, including documentation, build
-scripts, test cases and the development process itself.
+Latent defects can found on average at a rate of 60--65\% by rigorous
+inspection, with this rate reaching as high as 85\% \cite{Jones2008}. The
+success rate of code inspection is generally higher than via testing, since testing
+averages around 30--35\% for the efficiency of defect removal \cite{EbertAndJones2009,
+Jones2008}. For research software, Kelly and Shepard \cite{KellyAndShepard2000}
+show a task based inspection approach can be effective.
+
+Peer review addresses the same pain points and qualities as linters
+(\ppref{P_LackDevTime}, \ppref{P_Correctness}, and \qref{Q_Maintainability}),
+since they both focus on improving code quality and increasing knowledge
+transfer.  The benefits can be increased by extending reviews to all artifacts,
+including documentation, build scripts, test cases and the development process
+itself.
 
@@
 
-Although the advice to modularize research software to handle
-complexity is common \cite{WilsonEtAl2014, StewartEtAl2017, Storer2017},
-specific guidelines on how to divide the software into modules is less
-prevalent.  Not every decomposition is a good design for supporting change, as
-shown by Parnas~\cite{Parnas1972a}.  For instance, a design with low cohesion and high
-coupling \cite[p.\ 48]{GhezziEtAl2003} will make change difficult. Especially in
-research software, where change is inevitable, designers need to produce a
-modularization that supports change.  Ocean modelling software is currently
-feeling the pain of not emphasizing modularization in legacy code
-\cite{JungEtAl2022}.
+The advice to modularize scientific software to handle complexity is common
+\cite{WilsonEtAl2014, StewartEtAl2017, Storer2017}; however, specific guidelines
+on the criteria to use for modularization is rare.  Parnas~\cite{Parnas1972a}
+shows that not every decomposition supports change.  For instance, a design with
+high coupling and low cohesion will make change challenging~\cite[p.\
+48]{GhezziEtAl2003}. Since change is frequent in research software, the best
+modularizations will support change. The pain of not previously designing for
+change is currently being felt by ocean modelling community~\cite{JungEtAl2022}.
 
@@
 We follow Ampatzoglou et al.'s \cite{AmpatzoglouEtAl2019} analysis of
-threats to validity in software engineering secondary studies.
+threats to validity in secondary studies of software engineering.
 
@@
 
-A study is reliable if repeating it by another researcher, using the same
-methodology, would lead to the same results~\cite{RunesonAndHost2009}.
+Reliability means a study can be repeated by another researcher, using the same
+methodology, and the same results will be achieved~\cite{RunesonAndHost2009}.
 We identify the following threats:
 \begin{enumerate*}
-\item One individual does the measures for all packages. A different
-evaluator might find different results, due to differences in abilities,
-experience, and biases.
+\item A single person measures all packages. A different evaluator might find
+different results, due to differences in abilities, experience, and biases.
 \item The measurements for the full set of packages took several months (of
@@
 [citation redacted] reports grading five software products by two reviewers.
-Their rankings were almost identical. As long as each grader uses consistent
-definitions, the relative comparisons in the AHP results will be consistent
-between graders.
+Their rankings were almost identical. The relative comparisons in the AHP
+results will be consistent between graders, as long as the evaluators are
+consistent in their judgement.
 
@@
 
-Construct validity is when the adopted metrics represent what they are intended
-to measure~\cite{RunesonAndHost2009}. We have identified the following
-potential issues:
+If the adopted metrics match their intended measure~\cite{RunesonAndHost2009},
+then construct validity is achieved. We have identified the following potential
+issues:
 \begin{itemize}
 \item We make indirect measurement of software qualities since meaningful direct
-measures for qualities like maintainability, reusability and verifiability, are
-unavailable.  We follow the usual assumption that developers achieve higher
-quality by following procedures and adhering to standards \cite[p.\
-112]{VanVliet2000}.
+measures are unavailable for reusability, maintainability, verifiability, etc.
+We assume that following procedures and adhering to standards leads to higher
+quality \cite[p.\ 112]{VanVliet2000}.
 \item We could not do all measurements for some software as we could not
@@
 not necessarily good modules, nor understandable modules.
-\item The understandability measure relies on 10 random source code files, but
-the 10 files will not necessarily be representative. 
-\item Our overall AHP ranking makes the unrealistic assumption of equal
-weighting.
-\item We approximated popularity by stars and watches.
-\item Table~\ref{Tbl_Guidelines} required judgement as
-not all guidelines use the same names for artifacts that contain
-essentially the same information.
+\item The understandability measure is based on a random sample of 10 code
+files, which many not be representative of the entire code base. 
+\item The overall AHP ranking makes the unrealistic assumption of equal
+weighting between qualities.
+\item We used stars and watches to approximate popularity.
+\item Subjective judgement was required for Table~\ref{Tbl_Guidelines} since not
+all guidelines use the same names for artifacts that contain essentially the
+same information.
 \end{itemize}
@@
 
-Internal validity means that discovered causal relations are trustworthy and
-cannot be explained by other factors~\cite{RunesonAndHost2009}. We identify
-the following:
+The study is internally valid if the suggested causal relations are trustworthy
+and not explainable by other factors~\cite{RunesonAndHost2009}. The identified threats to internal validity are:
 
 \begin{itemize}
-\item Our search (Section~\ref{SecWhatProjects}) could have missed a relevant
-package.
-\item Our methodology assumes that development activities
-will leave a trace in the repositories, but this is not necessarily true. For
+\item Relevant packages could have been missed in our search
+(Section~\ref{SecWhatProjects}).
+\item We assume that evidence of development activities
+will be visible in the repositories, but this is not necessarily true. For
 instance, we saw little evidence of requirements
@@
 information outside their repos.
-\item We interviewed a relatively small sample of 8 teams. Their pain points
-(Section~\ref{painpoints}) may not be representative of the rest of their
-community.
+\item We interviewed a relatively small sample of 8 teams, which means their
+identified pain points might not be representative (Section~\ref{painpoints}).
 \end{itemize}
@@
 
-If the results of a study can be generalized to other situations, then the
-study is externally valid \cite{RunesonAndHost2009}.  In other words, we cannot
-generalize our results if the development of MI software is fundamentally
-different from other research software.  Although there are differences, like
-the importance of data privacy for MI data, we found the approach to developing
-[other software (citation redacted)]
+If the results of a study can be generalized to other situations, then the study
+is externally valid \cite{RunesonAndHost2009}.  In other words, we cannot
+generalize our results if there is a fundamental difference between the
+development of MI software and other research software.  Although there are
+differences, like the importance of data privacy for MI data, we found the
+approach to developing [other software (citation redacted)]
 %% DOUBLE BLIND
@@
 packages do well on at least a few qualities, as shown in
-Table~\ref{topperformerstable}, which summarizes the packages ranked first and
-second for each quality. Almost 70\% (20 of 29) of the software packages appear
-in the top two for at least two qualities.  The only packages that do not appear
-in Table~\ref{topperformerstable}, or only appear once, are \textit{Papaya},
-\textit{MatrixUser}, \textit{MRIcroGL}, \textit{XMedCon}, \textit{dicompyler},
-\textit{DicomBrowser}, \textit{AMIDE}, \textit{3DimViewer}, and
-\textit{Drishti}.
+Table~\ref{topperformerstable}, where we summarize the first and second ranked
+packages for each quality. Almost 70\% (20 of 29) of the software 
+is ranked in the top two for at least two qualities.  Packages that do not
+appear in Table~\ref{topperformerstable}, or only appear once, are
+\textit{Papaya}, \textit{MatrixUser}, \textit{MRIcroGL}, \textit{XMedCon},
+\textit{dicompyler}, \textit{DicomBrowser}, \textit{AMIDE}, \textit{3DimViewer},
+and \textit{Drishti}.
 

### 2025-01-25 StateOfPractice/Papers/SOP_MI_CritReviewsBioMedEng/ResubmissionResponseToReviewer.tex
Updates to resubmission of critical reviews paper

@@
 
-  \emph{Papers don't agree and papers are from many domains of application for
-  research software.  \cite{Nguyen-HoanEtAl2010}  Not aware of a study focused
-  on medical imaging practitioners.  Interviewing a statistically significant
-  number of developers is beyond the scope of this paper.  However, developers
-  can see the impact of a different weighting themselves.  Point to spreadsheet.
-  Say what text is added to the paper.  Put it in blue?  \smallskip}
+  \emph{The reviewer is correct that other weightings could be considered.  We
+  left this out of the scope of the original submission because of space
+  considerations and because determining a meaningful alternative weighting is
+  difficult.  Although Nguyen-Hoan et al.\ (2010) (A Survey of Scientific
+  Software Development) provides survey data on the relative importance of
+  various software qualities, their list of qualities differs from ours.  We are
+  not aware of a recent survey that assesses the relative importance of
+  qualities for scientific software. Moreover, even if we have quality
+  priorities for all scientific software, our paper focuses only on medical
+  imaging software.  The priorities for medical imaging software likely differ
+  from the priorities for other scientific applications.  Furthermore, the
+  priorities likely differ between different medical imaging applications as
+  well.
+Rather than speculate on the relative priorities, we used the simplest case of equal priority.
+For any readers that would like to apply their own priorities, we can direct them to the spreadsheet used to generate the results in the paper.
+With some simple modifications the plots can regenerated with whatever priority ranking a user wishes to see.
+To make this clear, we have added the following text to the revised paper:
+
+  \smallskip}
 
@@
   represent the community's feeling in the past more than they measure current
-  preferences \citep{Szulik2017}.  The issue with stars is that they tend only
+  preferences (Szulik2017).  The issue with stars is that they tend only
   to be added, not removed.  A final reason for inconsistencies between our
   ranking and the community's ranking is that, as for consumer products, more
-  factors influence popularity than just quality.'')\smallskip}
+  factors influence popularity than just quality.'' \smallskip}
 

### 2025-01-25 StateOfPractice/Papers/SOP_MI_CritReviewsBioMedEng/SOP_MI_OSR.tex
Updates to resubmission of critical reviews paper

@@
-\documentclass[doubleblind,12pt, 3p, times]{elsarticle}
+%\documentclass[doubleblind,12pt, 3p, times]{elsarticle}
+\documentclass[12pt, 3p, times]{elsarticle}
 
@@
 \usepackage[inline]{enumitem}
+
+\usepackage[tableposition=top]{caption} %to space the caption properly
 
@@
 We have a standard set of questions designed to assess the qualities of any
-research software project.  [blind review redacted details on the author's
-previous studies using the same methodology.]
-%% DOUBLE BLIND
-% ~\cite{SmithEtAl2021, SmithAndMichalski2022}.  This has
-% been applied to MI software and Lattice Boltzmann Solvers \cite{SmithEtAl2024}.
-% This builds off prior work to assess the state of the practice for such domains
-% as Geographic Information Systems \cite{smith2018state}, Mesh Generators
-% \cite{smith2016state}, Seismology software \cite{Smith2018Seismology}, and
-% Statistical software for psychology \cite{smith2018statistical}.  
-We maintain the previous constraint that the work load for measuring a given
-domain should take around one person-month's worth of effort ($20$ working days
-at $8$ person-hours per day).
+research software project~\cite{SmithEtAl2021, SmithAndMichalski2022}. We build
+off prior work to assess the state of the practice for such domains as Lattice
+Boltzmann Solvers \cite{SmithEtAl2024}, Geographic Information Systems
+\cite{smith2018state}, Mesh Generators \cite{smith2016state}, Seismology
+software \cite{Smith2018Seismology}, and Statistical software for psychology
+\cite{smith2018statistical}.  %% DOUBLE BLIND 
+We maintain the previous
+constraint that the work load for measuring a given domain should take around
+one person-month's worth of effort ($20$ working days at $8$ person-hours per
+day).
 
@@
 each remaining package, we measure its qualities via a grading
-template [citation redacted for double blind]. %%DOUBLE BLIND
-%~\cite{SmithEtAl2021}.  
+template~\cite{SmithEtAl2021}.    %%DOUBLE BLIND
 This data is used to rank the projects with the
@@
 interaction with the domain expert, software qualities, grading the software and
-AHP below and in longer form in [redacted for double blind].
+AHP below and in longer form in Smith et al.\ (2024)~\cite{SmithEtAl2024_MI_SOP}.
 %% DOUBLE BLIND
-%Smith et al.\ (2024)~\cite{SmithEtAl2024_MI_SOP}.
 
@@
 The Domain Expert vets the proposed list because online resources can be
-inaccurate.  The expert also vets the AHP ranking.  Our Domain Expert is [details of our domain expert removed for double-blind].
-%%DOUBLE BLIND %our Domain Expert (and paper co-author) is Dr.\ Michael
-%Noseworthy, Professor of Electrical and Computer Engineering at McMaster
-%University, Co-Director of the McMaster School of Biomedical Engineering, and
-%Director of Medical Imaging Physics and Engineering at St.\ Joseph's
-%Healthcare, Hamilton, Ontario, Canada.
+inaccurate.  The expert also vets the AHP ranking.  Our Domain Expert (and
+paper co-author) is Dr.\ Michael Noseworthy, Professor of Electrical and
+Computer Engineering at McMaster University, Co-Director of the McMaster School
+of Biomedical Engineering, and Director of Medical Imaging Physics and
+Engineering at St.\ Joseph's Healthcare, Hamilton, Ontario, Canada.  %%DOUBLE BLIND
 
@@
 
-We use an existing template [citation redacted]
-%% DOUBLE BLIND
-%~\cite{SmithEtAl2021} 
+We use an existing template \cite{SmithEtAl2021} %% DOUBLE BLIND
 that is designed to measure the
@@
 shows an excerpt of the measurement spreadsheet.  The rows are the measures and
-the columns correspond to the software packages.  [All data can be found on
-Mendeley; link will be provided after refereeing.] 
-% \cite{Dong2021-Data}
-%provides the full set of measurement data.   %%DOUBLE BLIND
+the columns correspond to the software packages.  Dong (2021)~\cite{Dong2021-Data}
+provides the full set of measurement data.   %%DOUBLE BLIND
 
@@
 \href{https://github.com/boyter/scc}{Sloc Cloc and Code (scc)}.  Further
-details on quality measurement are provided in [redacted for double blind].
+details on quality measurement are provided in Smith et al.\ (2024) \cite{SmithEtAl2024_MI_SOP}. 
 %% DOUBLE BLIND
-%Smith et al.\ (2024) \cite{SmithEtAl2024_MI_SOP}.
 
@@
 We based our interviews on a list of 20 questions, which can be found in
-[citation redacted for double blind].
-%Smith et al. \cite{SmithEtAl2021}. 
+Smith et al. \cite{SmithEtAl2021}. %%DOUBLE BLIND
 Some questions are about the background of the software, the development teams,
@@
 \begin{table*}[ht!]
+\caption{Final software list (sorted in descending order of the number of Lines
+Of Code (LOC))}
 \centering
@@
 \end{tabular}
-\caption{Final software list (sorted in descending order of the number of Lines
-Of Code (LOC))}
 \label{tab_final_list}
@@
 via web applications. The full measurement data for all packages is available on
-[removed for blind review]
-%\href{https://data.mendeley.com/datasets/k3pcdvdzj2/1} {Mendeley Data}.  
+\href{https://data.mendeley.com/datasets/k3pcdvdzj2/1} {Mendeley Data}.  
 %% DOUBLE BLIND
@@
 The overall score is based on the measurement of the identified qualities. Full
-details of how the projects compare for each quality can be found in [citation
-redacted for blind review].
-%% DOUBLE BLIND
-%Smith et al.~\cite{SmithEtAl2024_MI_SOP}
+details of how the projects compare for each quality can be found in Smith et
+al.~\cite{SmithEtAl2024_MI_SOP}. %% DOUBLE BLIND
+
 The highlights are as follows:
@@
 \begin{table*}[ht!]
+\caption{Number of files and lines (by reusability scores)}
 \centering
@@
 \end{tabular}
-\caption{Number of files and lines (by reusability scores)}
 \label{tab_loc_per_file}
@@
 \begin{table*}[ht!]
+\caption{Software ranking by our methodology versus the community (Comm.)\
+ranking using GitHub metrics (Sorted in descending order of community
+popularity, as estimated by the number of new stars per year)}
 \centering
@@
 \end{tabular}
-\caption{Software ranking by our methodology versus the community (Comm.)\
-ranking using GitHub metrics (Sorted in descending order of community
-popularity, as estimated by the number of new stars per year)}
 \label{tab_ranking_vs_GitHub}
@@
 \begin{table*}[ht!]
+\caption{Comparison of Recommended Artifacts in Software Development Guidelines
+to Artifacts in MI Projects (C for Common, U for Uncommon and R for Rare)}    
 \begin{center}
@@
 \end{tabular}
-\caption{Comparison of Recommended Artifacts in Software Development Guidelines
-to Artifacts in MI Projects (C for Common, U for Uncommon and R for Rare)}
 \label{Tbl_Guidelines}
@@
 We show the recommended artifacts in Table~\ref{Tbl_Guidelines}, one per row.
-For a given row, a checkmark in one of the columns means that the corresponding
-guideline recommends this artifact.  The last column shows whether the artifact
-appears in the measured set of MI software, either not at all (blank), commonly
-(C) (20 to 29 ($>$67\%) packages), uncommonly (U) (10 to 19 (33-67\%) packages)
-or rarely (R) (1 to 9 ($<$33\%) packages).  We did our best to interpret the
-meaning of each artifact consistently between guidelines and specific MI
-software, but the terminology and the contents of artifacts are not
-standardized.  The challenge even exists for the ubiquitous README file.  The
-content of README files shows significant variation between
-projects~\cite{PranaEtAl2018}.  Some content is reasonably consistent, with 97\%
-of README files contain at least one section describing the `What' of the
-repository and 89\% offering some `How' content, other categories are more
-variable.  For instance, information on `Contribution', `Why', and `Who', appear
-in 28\%, 26\% and 53\% of the analyzed files, respectively~\cite{PranaEtAl2018}.
+The columns correspond to guidelines. A checkmark in a column means that the
+guideline suggests the artifact in this row.  The final column shows the
+frequency of the artifact in the measured set of MI software.  The frequency
+ranges from not at all (blank), rarely (R) ($<$33\%), uncommonly (U) (33-67\%)
+to commonly (C) ($>$67\%).  Due to a lack of standardization, the mapping
+between specific MI software and guidelines was challenging, even for the
+ubiquitous README file.  Prana et al.~\cite{PranaEtAl2018} show significant
+variation in the content of README files between projects. Although 97\% of
+README files contain at least one section describing the `What' and 89\% offer
+content on `How', other content is more variable.  For example, information on
+`Who', `Contribution', and `Why', appear in 53\%, 28\%, 26\% of the analyzed
+files, respectively~\cite{PranaEtAl2018}.
 
 \begin{table*}[ht!]
+    \caption{Software Artifacts Found in MI Packages, Classified by Their Number
+    of Occurrences}
     \begin{center}
@@
     \end{tabular}
-    \caption{Artifacts Present in MI Packages, Classified by Frequency
-    (number of occurrences)}
     \label{artifactspresent}
@@
 \begin{table}[ht!]
+\caption{\label{tab_user_support_model}User support models by number of projects}
 \centering
@@
 \end{tabular}
-\caption{\label{tab_user_support_model}User support models by number of projects}
 \end{table}
@@
 is not uncommon
-\cite{PintoEtAl2018, PintoEtAl2016, WieseEtAl2019}.
-%DOUBLE BLIND \cite{SmithEtAl2024} [redacted for blind review]
+\cite{SmithEtAl2024, WieseEtAl2019, PintoEtAl2018, PintoEtAl2016}.
+%DOUBLE BLIND  [redacted for blind review]
 
@@
 there is an ongoing struggle to attract funding to build and, even harder, maintain
-research software. This is a common complaint~\cite{GewaltigAndCannon2012, Goble2014, KaterbowAndFeulner2018}. %, SmithEtAl2024 %[redacted for double blind review]
+research software. This is a common complaint~\cite{GewaltigAndCannon2012, Goble2014, KaterbowAndFeulner2018, SmithEtAl2024}. %,  %DOUBLE BLIND
 Software output is not always counted when judging the academic excellence of
@@
 %% DOUBLE BLIND
-%Smith et al.~\cite{SmithEtAl2016} 
-[citation redacted] reports grading five software products by two reviewers.
-Their rankings were almost identical. As long as each grader uses consistent
-definitions, the relative comparisons in the AHP results will be consistent
-between graders.
+Smith et al.~\cite{SmithEtAl2016} reports grading five software products by two reviewers.
+Their rankings were almost identical. The relative comparisons in the AHP
+results will be consistent between graders, as long as the evaluators are
+consistent in their judgement.
 
@@
 
-If the results of a study can be generalized to other situations, then the
-study is externally valid \cite{RunesonAndHost2009}.  In other words, we cannot
-generalize our results if the development of MI software is fundamentally
-different from other research software.  Although there are differences, like
-the importance of data privacy for MI data, we found the approach to developing
-[other software (citation redacted)]
+If the results of a study can be generalized to other situations, then the study
+is externally valid \cite{RunesonAndHost2009}.  In other words, we cannot
+generalize our results if there is a fundamental difference between the
+development of MI software and other research software.  Although there are
+differences, like the importance of data privacy for MI data, we found the
+approach to developing LBM software \cite{SmithEtAl2024} 
 %% DOUBLE BLIND
-%LBM software \cite{SmithEtAl2024} 
 and MI software to be similar.  Except for
@@
     %\begin{center}
+    \caption{Top performers for each quality (sorted by order of quality
+    measurement)}
     \renewcommand{\arraystretch}{1.8}
@@
     \end{tabular}
-    \caption{Top performers for each quality (sorted by order of quality
-    measurement)} \label{topperformerstable}
+    \label{topperformerstable}
     %\end{center}
@@
 
-[redacted for blind review]
 %% DOUBLE BLIND
-% We would like to thank Peter Michalski and Oluwaseun Owojaiye for fruitful
-% discussions on topics relevant to this paper.
+We would like to thank Peter Michalski and Oluwaseun Owojaiye for fruitful
+discussions on topics relevant to this paper.
 

### 2025-01-26 StateOfPractice/Papers/SOP_MI_CritReviewsBioMedEng/ResubmissionResponseToReviewer.tex
Resubmission version with reviewers comments addressed.

@@
 \usepackage{xcolor}
+\usepackage{hyperref}
 
@@
   from the priorities for other scientific applications.  Furthermore, the
-  priorities likely differ between different medical imaging applications as
-  well.
-Rather than speculate on the relative priorities, we used the simplest case of equal priority.
-For any readers that would like to apply their own priorities, we can direct them to the spreadsheet used to generate the results in the paper.
-With some simple modifications the plots can regenerated with whatever priority ranking a user wishes to see.
-To make this clear, we have added the following text to the revised paper:
+  priorities likely differ between different medical imaging applications.
+  Rather than speculate on the relative priorities, we used the simplest case of
+  equal priority. For any readers that would like to apply their own priorities,
+  we can direct them to the spreadsheet used to generate the results shown in the
+  paper.  With some simple modifications the plots can regenerated with whatever
+  priority ranking a user wishes to see.  To make this clear, we have added the
+  following text to the revised paper:  } \smallskip
 
-  \smallskip}
+  \textcolor{blue}{(Alternative priority schemes can be investigated by changing
+the Prioritization Matrix in
+\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/AHP2021/MedicalImaging/AHP_Template.xlsx}
+{AHP\_Template.xlsx} available on GitHub.)}
 
@@
 
-  \emph{Citations and websites are a good idea, but don't have this data.
-  Papers unreliable because people don't tend to cite scientific softwarwe.
-  cite Katz to support this.  Changes to paper.\smallskip}
+  \emph{Although the idea of including citations in papers is a good one, we
+  have excluded this data because we feel it is unreliable.  Unfortunately,
+  software is infrequently and inconsistently cited even when it is used (as
+  observed by Smith et al (2016) (Software Citation Principles)).  Similar
+  inconsistency problems exist for data on websites.  For other readers who have
+  the same observation as the reviewer, we have added the following text to the
+  revised version: \smallskip}
+
+  \textcolor{blue}{We considered also comparing software citations, but this
+  measure of popularity would be unreliable since software is infrequently and
+  inconsistently cited (SmithEtAl2016).}
 
@@
 
-  \emph{Double blind review meant didn't show citations, but citations have the
-  full quality ratings.  The overall quality rating is given in Figure 2. The
-  actual numbers are not as relevant as the ranking. \smallskip}
+  \emph{The quality ratings for each software are summarized in Figure 2
+  (Overall AHP scores with an equal weighting for all 9 software qualities). To
+  keep the length of the paper reasonable, we do not have figures showing the
+  ranking of the software packages for each quality, but this information is
+  available through the citations given in the now unredacted revised version of the
+  paper.  Some of the relevant citations include Dong (2021) (Assessing the
+  state of the practice for medical imaging software, MEng thesis), and Smith et
+  al (2024) (State of the practice for medical imaging software, extended report
+  on arXiv). \smallskip}
 
@@
 
-  \emph{Add content from the arxiv version.  ``Further reason for discrepancies
-  between our ranking and the community's ranking is that we weighted all
-  qualities equally. This is not likely how users implicitly rank the different
-  qualities. As a result, some projects with high community popularity may have
-  scored lower with our method because of a relatively higher (compared to the
-  scientific community's implicit ranking) weighting of the poor scores for some
-  qualities. A further explanation for discrepancies between our measures and
-  the star measures may also be due to inaccuracy with using stars to
-  approximate popularity.  Stars are not an ideal measure because stars
-  represent the community's feeling in the past more than they measure current
-  preferences (Szulik2017).  The issue with stars is that they tend only
-  to be added, not removed.  A final reason for inconsistencies between our
-  ranking and the community's ranking is that, as for consumer products, more
-  factors influence popularity than just quality.'' \smallskip}
+  \emph{Other possible reasons for the differences in ranking are covered in the
+  original submitted paper in the following text: ``We weighted all qualities
+  equally, which is not the likely the same weighting that users implicitly use.
+  To properly assess this would require a broad user study.  Furthermore our
+  measures of popularity (like stars) are only \emph{proxies} which are biased
+  towards past rather than current preferences (Szulik2017), as these are
+  monotonically increasing quantities. Finally there are often more factors than
+  just quality that influence the popularity of products.''  Given that this
+  wasn't obvious to the reviewer, we strengthened the topic sentence that
+  introduced the reasons with the following text: \smallskip}
+
+  \textcolor{blue}{Besides the installation problem, another possible reason for
+  discrepancies between our ranking and GitHub popularity is that we weighted
+  all qualities equally, ....}
 
@@
 
-  \emph{Say the problems from installability section.  Didn't remove because
-  only affected a few qualities.  Favoured including with a warning.
-  \smallskip}
+  \emph{In the interest of space, we have chosen to not include the details in
+  on installation problems for \textit{GATE}, \textit{dwv}, and \textit{DICOM
+  Viewer} in the paper.  Instead, the additional details are given in the
+  extended version of the paper on arXiv (Smith et al (2024)) and in the
+  associated Masters thesis (Dong (2021)).  The software with installation
+  problems was not removed from our review because we aimed for as complete an
+  overview as possible.   Moreover, the installation problems only affected the
+  score on installability, reliability and robustness.  The other qualities are
+  based on the artifacts in the repository, not on running the software.  To
+  clarify this for readers of the paper, we have added text to the methodology
+  for grading in the revised manuscript: \smallskip}
+
+  \textcolor{blue}{We explicitly note any software that cannot be installed,
+  but for completeness we still measure it for the remaining qualities, since
+  only reliability and robustness require running the software.}
 
@@
 
-  \emph{Response.\smallskip}
+  \emph{We believe that our introduction covers the scope (in Section 1.2 of the
+  submitted version) the significance (in the first paragraph).  We also believe
+  that the current understanding, background information and potential outcomes
+  are covered by the research questions. There is not a considerable amount of
+  background information because there isn't a previous study (that we are aware
+  of) that assesses the state of the practice for medical imaging software. The
+  potential outcomes are the answers to the research questions. \smallskip}
+
+  \emph{The criticism of the introduction, including the comment about not
+  having subsections, is likely due to our covering the methodology in the
+  introduction section. In the revised version, we have made the methodology its
+  own section and removed the subsections from the introduction. \smallskip}
 
@@
 
-  \emph{Move it.  \smallskip}
+  \emph{Although including a table in the conclusions is rare, we think it is
+  appropriate for this paper.  A table is a quickest way to summarize the
+  software package quality.  Summarizing the same information in words would be
+  hard to follow.  A conclusion shouldn't introduce new information, but the
+  table is not new information; it is an alternate view of the information
+  presented in Section 4 of the paper.  This new view makes some conclusions
+  about the state of the practice clear, so it feels appropriate for the
+  conclusion section. \smallskip}
 

### 2025-01-26 StateOfPractice/Papers/SOP_MI_CritReviewsBioMedEng/SOP_MI_OSR.tex
Resubmission version with reviewers comments addressed.

@@
 
-\subsection{Research Questions} \label{sec_motivation}
+%\subsection{Research Questions} \label{sec_motivation}
 
@@
 
-\subsection{Scope} \label{sec_scope}
+%\subsection{Scope} \label{sec_scope}
 
@@
 
-\subsection{Methodology} \label{SecMethodology}
+\section{Methodology} \label{SecMethodology}
 
@@
 
-\subsubsection{Domain Expert} \label{sec_vet_software_list}
+\subsection{Domain Expert} \label{sec_vet_software_list}
 
@@
 
-\subsubsection{Software Qualities} \label{sec_software_quality}
+\subsection{Software Qualities} \label{sec_software_quality}
 
@@
 
-\subsubsection{Grading} \label{sec_grading_software}
+\subsection{Grading} \label{sec_grading_software}
 
@@
 help regarding installation, if necessary, but we impose a cap of about two
-hours on the installation process.  Figure~\ref{fg_grading_template_example}
+hours on the installation process.
+We explicitly note any software that cannot be installed, but for completeness we still measure it for the remaining qualities, since only reliability and robustness require running the software.
+Figure~\ref{fg_grading_template_example}
 shows an excerpt of the measurement spreadsheet.  The rows are the measures and
@@
 
-\subsubsection{Analytic Hierarchy Process (AHP)} \label{sec_AHP}
+\subsection{Analytic Hierarchy Process (AHP)} \label{sec_AHP}
 
@@
 
-\subsubsection{Interview Methods} \label{sec_interview_methods}
+\subsection{Interview Methods} \label{sec_interview_methods}
 
@@
 
-We measured the software as described in
-Section~\ref{SecMethodology}.  In the absence of a specific real world context,
-we assumed all nine qualities are equally important.
-Figure~\ref{fg_overall_scores} shows the overall AHP scores in descending order.
+We measured the software as described in Section~\ref{SecMethodology}.  In the
+absence of a specific real world context, we assumed all nine qualities are
+equally important.  (Alternative priority schemes can be investigated by changing the
+Prioritization Matrix in
+\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/AHP2021/MedicalImaging/AHP_Template.xlsx}
+{AHP\_Template.xlsx} available on GitHub.) Figure~\ref{fg_overall_scores} shows
+the overall AHP scores in descending order.
 
@@
 number of people watching the projects as proxies for community ranking.
-Table~\ref{tab_ranking_vs_GitHub} show the statistics for data collected in July
+We considered also comparing software citations, but this measure of popularity would be unreliable since software is infrequently and inconsistently cited~\cite{SmithEtAl2016-softcite}.
+Table~\ref{tab_ranking_vs_GitHub} show the statistics for GitHub data collected in July
 2021.
@@
 
-We weighted all qualities equally, which is not the likely the same weighting
+Besides the installation problem, another possible reason for discrepancies between our ranking and GitHub popularity is that we weighted all qualities equally, which is not the likely the same weighting
 that users implicitly use. To properly assess this would require a broad user
@@
 
-We interview data, as described in Section~\ref{sec_interview_methods}
-to answer \rqref{RQ_CompareToolsProjMngmnt} (tool usage),
+We interview developers, as described in Section~\ref{sec_interview_methods}, to answer \rqref{RQ_CompareToolsProjMngmnt} (tool usage),
 \rqref{RQ_CompareMethodologies} (methodology) and \rqref{RQ_PainPoints} (pain
@@
 minutes for each interview. The full interview answers can be
-found in [citation redacted for blind review].
-%\cite{Dong2021}.
+found in Dong (2021)~\cite{Dong2021}. %%DOUBLE BLIND
 
@@
 entirely external. This was the case for a study of Lattice Boltzmann Method
-(LBM) software [citation redacted for blind review].  %\cite{Michalski2021}.
+(LBM) software~\cite{Michalski2021}. %%DOUBLE BLIND
 This contrasts with the high frequency of guidelines recommending continuous
 integration
-\cite{BrettEtAl2021, Brown2015, ThielEtAl2020, Zadka2018, vanGompelEtAl2016}.
+\cite{BrettEtAl2021, ThielEtAl2020, vanGompelEtAl2016, Brown2015, Zadka2018}.
 A recent survey~\cite{CarverEtAl2022} of practitioners (rather than projects)
@@
 and Google forms.  As a point of contrast, LBM software most often uses API
-document generation tools, like doxygen and sphinx [citation redacted for blind
-review].
-%\cite{Michalski2021}
+document generation tools, like doxygen and sphinx~\cite{Michalski2021}. %%DOUBLE BLIND
 
@@
 This matches previous observations for other research software.  Scientific
-software developers often use an agile philosophy \cite{AckroydEtAl2008,
-CarverEtAl2007, EasterbrookAndJohns2009, Segal2005, HeatonAndCarver2015}, or an
+software developers often use an agile philosophy \cite{HeatonAndCarver2015, CarverEtAl2007, AckroydEtAl2008, EasterbrookAndJohns2009, Segal2005}, or an
 amethododical process \cite{Kelly2013}, or a knowledge acquisition driven
@@
 common approach was issue-driven via bug reports and feature requests.
-[Redacted for blind review] %Dong~\cite{Dong2021} 
+Dong (2021)~\cite{Dong2021} %%DOUBLE BLIND
 provides details on the specific approaches used by the interviewees.
@@
 level of testing matches what was observed for LBM
-software [citation redacted for blind review]
-%\cite{Michalski2021} 
+software~\cite{Michalski2021} %%DOUBLE BLIND
 and is apparently greater than the level of
@@
 coding, and collaboration challenges. Two more are~\cite{WieseEtAl2019}:
-reproducibility, and software scope determination.  [citation redacted for blind
-review] %\cite{SmithEtAl2024} 
+reproducibility, and software scope determination.  Smith et al.\ (2024)~\cite{SmithEtAl2024} 
 adds lack of software experience, documentation and technical debt.  Our small
@@
 \cite{SmithEtAl2024, WieseEtAl2019, PintoEtAl2018, PintoEtAl2016}.
-%DOUBLE BLIND  [redacted for blind review]
+%DOUBLE BLIND
 
@@
 there is an ongoing struggle to attract funding to build and, even harder, maintain
-research software. This is a common complaint~\cite{GewaltigAndCannon2012, Goble2014, KaterbowAndFeulner2018, SmithEtAl2024}. %,  %DOUBLE BLIND
+research software. This is a common complaint~\cite{SmithEtAl2024, GewaltigAndCannon2012, Goble2014, KaterbowAndFeulner2018}. %,  %DOUBLE BLIND
 Software output is not always counted when judging the academic excellence of
@@
 \cite{WieseEtAl2019}. 
-%A similar pattern has been observed during interviews with
-%LBM software developers [citation redacted for blind review].
-%\cite{SmithEtAl2024}.
+A similar pattern has been observed during interviews with
+LBM software developers \cite{SmithEtAl2024}. %%DOUBLE BLIND
 
@@
     \item Remove headaches caused by a separate integration phase
-    \cite{Fowler2006}, \cite[p.\ 20]{HumbleAndFarley2010}.
+    \cite[p.\ 20]{HumbleAndFarley2010}, \cite{Fowler2006}.
     \item Detection and removal of bugs \cite{Fowler2006} via
@@
 explains the relatively high quality of software in the Comprehensive R Archive
-Network (CRAN) [citation redacted for blind review].
-%\cite{SmithEtAl2018_StatSoft}.
+Network (CRAN)~\cite{SmithEtAl2018_StatSoft}. %%DOUBLE BLIND
 
@@
 success rate of code inspection is generally higher than via testing, since testing
-averages around 30--35\% for the efficiency of defect removal \cite{EbertAndJones2009,
-Jones2008}. For research software, Kelly and Shepard \cite{KellyAndShepard2000}
+averages around 30--35\% for the efficiency of defect removal \cite{Jones2008, EbertAndJones2009}. For research software, Kelly and Shepard \cite{KellyAndShepard2000}
 show a task based inspection approach can be effective.
@@
 The advice to modularize scientific software to handle complexity is common
-\cite{WilsonEtAl2014, StewartEtAl2017, Storer2017}; however, specific guidelines
+\cite{Storer2017, WilsonEtAl2014, StewartEtAl2017}; however, specific guidelines
 on the criteria to use for modularization is rare.  Parnas~\cite{Parnas1972a}

### 2025-01-28 StateOfPractice/Papers/SOP_MI_CritReviewsBioMedEng/SOP_MI_OSR.tex
Cleanup comments in tex file, add jpg versions of figures

@@
 %\documentclass[doubleblind,12pt, 3p, times]{elsarticle}
+%\documentclass[draft, review, 12pt, 3p, times]{elsarticle} %so figures are blank, double spaced
 \documentclass[12pt, 3p, times]{elsarticle}
@@
 
-% prepared for: Journal of Imaging Informatics in Medicine
-% https://link.springer.com/journal/10278
-
 \journal{Critical Reviews in Biomedical Engineering}
 
+%\DeclareGraphicsExtensions{.pdf} %use pdf version of figures
+\DeclareGraphicsExtensions{.jpg} %use jpg versions (high resolution won't actually fit)
+
 \begin{document}
@@
 \begin{frontmatter}
-
-%% Title, authors and addresses
-
-%% use the tnoteref command within \title for footnotes;
-%% use the tnotetext command for theassociated footnote;
-%% use the fnref command within \author or \affiliation for footnotes;
-%% use the fntext command for theassociated footnote;
-%% use the corref command within \author for corresponding author footnotes;
-%% use the cortext command for theassociated footnote;
-%% use the ead command for the email address,
-%% and the form \ead[url] for the home page:
-%% \title{Title\tnoteref{label1}}
-%% \tnotetext[label1]{}
-%% \author{Name\corref{cor1}\fnref{label2}}
-%% \ead{email address}
-%% \ead[url]{home page}
-%% \fntext[label2]{}
-%% \cortext[cor1]{}
-%% \affiliation{organization={},
-%%            addressline={}, 
-%%            city={},
-%%            postcode={}, 
-%%            state={},
-%%            country={}}
-%% \fntext[label3]{}
 
@@
 
-%% use optional labels to link authors explicitly to addresses:
-%% \author[label1,label2]{}
-%% \affiliation[label1]{organization={},
-%%             addressline={},
-%%             city={},
-%%             postcode={},
-%%             state={},
-%%             country={}}
-%%
-%% \affiliation[label2]{organization={},
-%%             addressline={},
-%%             city={},
-%%             postcode={},
-%%             state={},
-%%             country={}}
-
-%\author{Anonymous}
 \author[CAS]{Spencer Smith}
@@
 \affiliation[CAS]{organization={McMaster University, Computing and Software
-Department}, %Department and Organization
+Department}, 
             addressline={1280 Main Street West}, 
@@
 
-\affiliation[ECE]{organization={McMaster University, Electrical and Computer Engineering}, %Department and Organization
+\affiliation[ECE]{organization={McMaster University, Electrical and Computer Engineering}, 
             addressline={1280 Main Street West}, 
@@
 
-%%Graphical abstract
-%\begin{graphicalabstract}
-%\includegraphics{grabs}
-%\end{graphicalabstract}
-
-%%Research highlights
-%\begin{highlights}
-%\item Research highlight 1
-%\item Research highlight 2
-%\end{highlights}
-
 \begin{keyword}
@@
 \end{frontmatter}
-
-%% \linenumbers
 
@@
 use a software engineering lens to assess the quality of existing MI software.
-
-%\subsection{Research Questions} \label{sec_motivation}
 
@@
 \item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareMethodologies}:]
-How does the use of principles, processes, and methodologies compare between MI software and research software in general? (Section~\ref{Sec_CompareMethodologies})
+How does the use of principles, processes, and methodologies compare between MI
+software and research software in general?
+(Section~\ref{Sec_CompareMethodologies})
 \item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_PainPoints}:] For MI software
@@
 
-%\subsection{Scope} \label{sec_scope}
-
 Our scope is limited to MI visualization software.  We exclude other categories
@@
 software \cite{Smith2018Seismology}, and Statistical software for psychology
-\cite{smith2018statistical}.  %% DOUBLE BLIND 
-We maintain the previous
-constraint that the work load for measuring a given domain should take around
-one person-month's worth of effort ($20$ working days at $8$ person-hours per
-day).
+\cite{smith2018statistical}.  We maintain the previous constraint that the work
+load for measuring a given domain should take around one person-month's worth of
+effort ($20$ working days at $8$ person-hours per day).
 
@@
 each remaining package, we measure its qualities via a grading
-template~\cite{SmithEtAl2021}.    %%DOUBLE BLIND
-This data is used to rank the projects with the
+template~\cite{SmithEtAl2021}.  This data is used to rank the projects with the
 Analytic Hierarchy Process (AHP).  We summarize further details on the
 interaction with the domain expert, software qualities, grading the software and
-AHP below and in longer form in Smith et al.\ (2024)~\cite{SmithEtAl2024_MI_SOP}.
-%% DOUBLE BLIND
+AHP below and in longer form in Smith et al.\
+(2024)~\cite{SmithEtAl2024_MI_SOP}.
 
@@
 The Domain Expert vets the proposed list because online resources can be
-inaccurate.  The expert also vets the AHP ranking.  Our Domain Expert (and
-paper co-author) is Dr.\ Michael Noseworthy, Professor of Electrical and
-Computer Engineering at McMaster University, Co-Director of the McMaster School
-of Biomedical Engineering, and Director of Medical Imaging Physics and
-Engineering at St.\ Joseph's Healthcare, Hamilton, Ontario, Canada.  %%DOUBLE BLIND
+inaccurate.  The expert also vets the AHP ranking.  Our Domain Expert (and paper
+co-author) is Dr.\ Michael Noseworthy, Professor of Electrical and Computer
+Engineering at McMaster University, Co-Director of the McMaster School of
+Biomedical Engineering, and Director of Medical Imaging Physics and Engineering
+at St.\ Joseph's Healthcare, Hamilton, Ontario, Canada.
 
@@
 Quality is defined as a measure of the excellence or worth of an entity.  As is
-common practice, we do not think of quality as a single measure, but rather as
-a set of measures.  That is, quality is a collection of different qualities,
-often called ``ilities.''  For this study we selected 9 qualities to measure:
+common practice, we do not think of quality as a single measure, but rather as a
+set of measures.  That is, quality is a collection of different qualities, often
+called ``ilities.''  For this study we selected 9 qualities to measure:
 installability, correctness/verifiability, reliability, robustness, usability,
 maintainability, reusability, understandability, and visibility/transparency.
-With the exception of installability, all the qualities are defined in Ghezzi
-et al. (2003) \cite{GhezziEtAl2003}. Installability is defined as the effort
+With the exception of installability, all the qualities are defined in Ghezzi et
+al. (2003) \cite{GhezziEtAl2003}. Installability is defined as the effort
 required for the installation and/or uninstallation of software in a specified
@@
 
-We use an existing template \cite{SmithEtAl2021} %% DOUBLE BLIND
-that is designed to measure the
+We use an existing template \cite{SmithEtAl2021} that is designed to measure the
 aforementioned qualities. To stay within our given measurement time frame, each
@@
 help regarding installation, if necessary, but we impose a cap of about two
-hours on the installation process.
-We explicitly note any software that cannot be installed, but for completeness we still measure it for the remaining qualities, since only reliability and robustness require running the software.
-Figure~\ref{fg_grading_template_example}
-shows an excerpt of the measurement spreadsheet.  The rows are the measures and
-the columns correspond to the software packages.  Dong (2021)~\cite{Dong2021-Data}
-provides the full set of measurement data.   %%DOUBLE BLIND
+hours on the installation process.  We explicitly note any software that cannot
+be installed, but for completeness we still measure it for the remaining
+qualities, since only reliability and robustness require running the software.
+Figure~\ref{fg_grading_template_example} shows an excerpt of the measurement
+spreadsheet.  The rows are the measures and the columns correspond to the
+software packages.  Dong (2021)~\cite{Dong2021-Data} provides the full set of
+measurement data.
 
 \begin{figure*}[ht!]
-\includegraphics[scale=0.67]{template.pdf}
+\includegraphics[scale=0.67]{template}
 \caption{Grading template example}
@@
 details on quality measurement are provided in Smith et al.\ (2024) \cite{SmithEtAl2024_MI_SOP}.
-%% DOUBLE BLIND
 
@@
 threats.  Therefore, part of our methodology involves interviewing developers.
-We based our interviews on a list of 20 questions, which can be found in
-Smith et al. \cite{SmithEtAl2021}. %%DOUBLE BLIND
-Some questions are about the background of the software, the development teams,
-the interviewees, and how they organize their projects.  We also ask about the
-developer's understanding of the users. Additional questions focus on the current and
-past difficulties, and the solutions the team has found, or plan to try. We also
-discuss documentation, both with respect to how it is currently done, and how it
-is perceived. A few questions are about specific software qualities, such as
-maintainability, understandability, usability, and reproducibility. The
-interviews are semi-structured based on the question list.
+We based our interviews on a list of 20 questions, which can be found in Smith
+et al. \cite{SmithEtAl2021}.  Some questions are about the background of the
+software, the development teams, the interviewees, and how they organize their
+projects.  We also ask about the developer's understanding of the users.
+Additional questions focus on the current and past difficulties, and the
+solutions the team has found, or plan to try. We also discuss documentation,
+both with respect to how it is currently done, and how it is perceived. A few
+questions are about specific software qualities, such as maintainability,
+understandability, usability, and reproducibility. The interviews are
+semi-structured based on the question list.
 
@@
 \toprule
-\multirow{2}{*}{Software} & \multirow{2}{*}{Rlsd} & \multirow{2}{*}{Updated} & \multirow{2}{*}{Fnd} & \multirow{2}{*}{NOC} & \multirow{2}{*}{LOC} & \multicolumn{3}{c}{OS} & \multirow{2}{*}{Web} \\ \cmidrule{7-9}
+\multirow{2}{*}{Software} & \multirow{2}{*}{Rlsd} & \multirow{2}{*}{Updated} &
+\multirow{2}{*}{Fnd} & \multirow{2}{*}{NOC} & \multirow{2}{*}{LOC} &
+\multicolumn{3}{c}{OS} & \multirow{2}{*}{Web} \\ \cmidrule{7-9}
  &  &  &  &  &  & W & M & L &  \\ \midrule
-ParaView \cite{Ahrens2005} & 2002 & 2020-10 & \checkmark & 100 & 886326 & \checkmark & \checkmark & \checkmark & \checkmark \\
+ParaView \cite{Ahrens2005} & 2002 & 2020-10 & \checkmark & 100 & 886326 &
+\checkmark & \checkmark & \checkmark & \checkmark \\
 Gwyddion \cite{Nevcas2012} & 2004 & 2020-11 &  & 38 & 643427 & \checkmark & \checkmark & \checkmark &  \\
@@
 \href{https://data.mendeley.com/datasets/k3pcdvdzj2/1} {Mendeley Data}.
-%% DOUBLE BLIND
 
@@
 absence of a specific real world context, we assumed all nine qualities are
-equally important.  (Alternative priority schemes can be investigated by changing the
-Prioritization Matrix in
+equally important.  (Alternative priority schemes can be investigated by
+changing the Prioritization Matrix in
 \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/AHP2021/MedicalImaging/AHP_Template.xlsx}
@@
 \begin{figure*}[ht!]
-\includegraphics[scale=0.47]{overall_scores.pdf}
-\caption{Overall AHP scores with an equal weighting for all 9 software qualities}
+\includegraphics[scale=0.47]{overall_scores}
+\caption{Overall AHP scores with an equal weighting for all 9 software
+qualities}
 
@@
 details of how the projects compare for each quality can be found in Smith et
-al.~\cite{SmithEtAl2024_MI_SOP}. %% DOUBLE BLIND
+al.~\cite{SmithEtAl2024_MI_SOP}.
 
@@
 \toprule
-\multirow{2}{*}{Software} & \multirow{2}{*}{Text Files} & \multirow{2}{*}{Total Lines} & \multirow{2}{*}{LOC} & \multirow{2}{*}{LOC/file} \\
+\multirow{2}{*}{Software} & \multirow{2}{*}{Text Files} & \multirow{2}{*}{Total
+Lines} & \multirow{2}{*}{LOC} & \multirow{2}{*}{LOC/file} \\
  &  &  &  &  \\ 
@@
 
-We use GitHub stars, number of forks and
-number of people watching the projects as proxies for community ranking.
-We considered also comparing software citations, but this measure of popularity would be unreliable since software is infrequently and inconsistently cited~\cite{SmithEtAl2016-softcite}.
-Table~\ref{tab_ranking_vs_GitHub} show the statistics for GitHub data collected in July
-2021.
+We use GitHub stars, number of forks and number of people watching the projects
+as proxies for community ranking.  We considered also comparing software
+citations, but this measure of popularity would be unreliable since software is
+infrequently and inconsistently cited~\cite{SmithEtAl2016-softcite}.
+Table~\ref{tab_ranking_vs_GitHub} show the statistics for GitHub data collected
+in July 2021.
 
 Our ranking and GitHub popularity, at least for the top five projects, seems to
-line up well. However, we ranked some popular packages fairly low, such
-as \textit{dwv}. This is because we were unable to build it locally, even though
-we followed its installation instructions. However, we were able to use its web
+line up well. However, we ranked some popular packages fairly low, such as
+\textit{dwv}. This is because we were unable to build it locally, even though we
+followed its installation instructions. However, we were able to use its web
 version for the rest of the measurements. Additionally, this version did not
@@
 
-Besides the installation problem, another possible reason for discrepancies between our ranking and GitHub popularity is that we weighted all qualities equally, which is not the likely the same weighting
-that users implicitly use. To properly assess this would require a broad user
-study.  Furthermore our measures of popularity (like stars) are only \emph{proxies} which
-are biased towards past rather than current preferences~\cite{Szulik2017}, as
-these are monotonically increasing quantities. Finally there are often more
-factors than just quality that influence the popularity of 
-products.
+Besides the installation problem, another possible reason for discrepancies
+between our ranking and GitHub popularity is that we weighted all qualities
+equally, which is not the likely the same weighting that users implicitly use.
+To properly assess this would require a broad user study.  Furthermore our
+measures of popularity (like stars) are only \emph{proxies} which are biased
+towards past rather than current preferences~\cite{Szulik2017}, as these are
+monotonically increasing quantities. Finally there are often more factors than
+just quality that influence the popularity of products.
 
@@
 \begin{center}
-\begin{tabular}{ p{3.5cm}p{0.5cm}p{0.5cm}p{0.5cm}p{0.5cm}p{0.5cm}p{0.5cm}p{0.5cm}p{0.5cm}p{0.5cm}p{0.5cm} }
+\begin{tabular}{
+p{3.5cm}p{0.5cm}p{0.5cm}p{0.5cm}p{0.5cm}p{0.5cm}p{0.5cm}p{0.5cm}p{0.5cm}p{0.5cm}p{0.5cm}
+}
 \toprule
@@
 
-We interview developers, as described in Section~\ref{sec_interview_methods}, to answer \rqref{RQ_CompareToolsProjMngmnt} (tool usage),
+We interview developers, as described in Section~\ref{sec_interview_methods}, to
+answer \rqref{RQ_CompareToolsProjMngmnt} (tool usage),
 \rqref{RQ_CompareMethodologies} (methodology) and \rqref{RQ_PainPoints} (pain
@@
 the projects agreed to participate: \textit{3D Slicer}, \textit{INVESALIUS 3},
-\textit{dwv}, \textit{BioImage Suite Web}, \textit{ITK-SNAP},
-\textit{MRIcroGL}, \textit{Weasis}, and \textit{OHIF}. We spent about 90
-minutes for each interview. The full interview answers can be
-found in Dong (2021)~\cite{Dong2021}. %%DOUBLE BLIND
+\textit{dwv}, \textit{BioImage Suite Web}, \textit{ITK-SNAP}, \textit{MRIcroGL},
+\textit{Weasis}, and \textit{OHIF}. We spent about 90 minutes for each
+interview. The full interview answers can be found in Dong
+(2021)~\cite{Dong2021}.
 
@@
 
-\paragraph{CI/CD} 
-Only 17\% of MI projects use CI/CD tools. We estimated this using artifacts
-present in the repository -- actual usage may be higher as some tools are
-entirely external. This was the case for a study of Lattice Boltzmann Method
-(LBM) software~\cite{Michalski2021}. %%DOUBLE BLIND
-This contrasts with the high frequency of guidelines recommending continuous
-integration
-\cite{BrettEtAl2021, ThielEtAl2020, vanGompelEtAl2016, Brown2015, Zadka2018}.
-A recent survey~\cite{CarverEtAl2022} of practitioners (rather than projects)
-suggests higher use of CI/CD with 54\% of respondents indicating that they use
-it.
+\paragraph{CI/CD} Only 17\% of MI projects use CI/CD tools. We estimated this
+using artifacts present in the repository -- actual usage may be higher as some
+tools are entirely external. This was the case for a study of Lattice Boltzmann
+Method (LBM) software~\cite{Michalski2021}.  This contrasts with the high
+frequency of guidelines recommending continuous integration \cite{BrettEtAl2021,
+ThielEtAl2020, vanGompelEtAl2016, Brown2015, Zadka2018}.  A recent
+survey~\cite{CarverEtAl2022} of practitioners (rather than projects) suggests
+higher use of CI/CD with 54\% of respondents indicating that they use it.
 
@@
 and Google forms.  As a point of contrast, LBM software most often uses API
-document generation tools, like doxygen and sphinx~\cite{Michalski2021}. %%DOUBLE BLIND
+document generation tools, like doxygen and sphinx~\cite{Michalski2021}.
 
@@
 This matches previous observations for other research software.  Scientific
-software developers often use an agile philosophy \cite{HeatonAndCarver2015, CarverEtAl2007, AckroydEtAl2008, EasterbrookAndJohns2009, Segal2005}, or an
+software developers often use an agile philosophy \cite{HeatonAndCarver2015,
+CarverEtAl2007, AckroydEtAl2008, EasterbrookAndJohns2009, Segal2005}, or an
 amethododical process \cite{Kelly2013}, or a knowledge acquisition driven
@@
 No interviewee mentioned a strictly defined project management process. The most
-common approach was issue-driven via bug reports and feature requests.
-Dong (2021)~\cite{Dong2021} %%DOUBLE BLIND
-provides details on the specific approaches used by the interviewees.
-% JC: I don't think this level of per-project detail makes sense in this
-% review article - it should be about stats and trends, not reporting on
-% minute details. We could refer the reader to Ao's thesis for the details.
-% 
-% SS: I've revised the text to not include the details
-% The
-% most common approach was following the issues, such as bugs and feature
-% requests. The \textit{3D Slicer} team had weekly meetings to discuss
-% the goals for the project. The \textit{INVESALIUS 3} team relied on the
-% GitHub process for their project management while the \textit{ITK-SNAP} 
-% team had a fixed six-month release pace. Only the interviewee from 
-% \textit{OHIF} 
-% mentioned the team having a project manager. The \textit{3D Slicer} team and
-% \textit{BioImage Suite Web} team do nightly builds and tests. The \textit{OHIF}
-% developer believes that a better project management process can improve junior
-% developer efficiency while also improving internal and external communication.
-
-Less than half of the projects used unit testing, even
-though the interviewees believed that testing (including usability tests) was
-the top approach to improve correctness, usability, and reproducibility.  This
-level of testing matches what was observed for LBM
-software~\cite{Michalski2021} %%DOUBLE BLIND
-and is apparently greater than the level of
-testing for ocean modelling software~\cite{JungEtAl2022}.
-
-All interviewees thought that documentation was
-essential to their projects, and most of them (7 of 8) said that it could save
-time spent answering questions from users and developers. Half of them (4 of 8)
-saw the need to improve their documentation, and only three of them thought 
-their documentations conveyed information clearly enough.
+common approach was issue-driven via bug reports and feature requests.  Dong
+(2021)~\cite{Dong2021} provides details on the specific approaches used by the
+interviewees.
+
+Less than half of the projects used unit testing, even though the interviewees
+believed that testing (including usability tests) was the top approach to
+improve correctness, usability, and reproducibility.  This level of testing
+matches what was observed for LBM software~\cite{Michalski2021} and is
+apparently greater than the level of testing for ocean modelling
+software~\cite{JungEtAl2022}.
+
+All interviewees thought that documentation was essential to their projects, and
+most of them (7 of 8) said that it could save time spent answering questions
+from users and developers. Half of them (4 of 8) saw the need to improve their
+documentation, and only three of them thought their documentations conveyed
+information clearly enough.
 
@@
 \end{enumerate*}
-We first detail each pain point, and contrast this experience with that
-from other domains. We then separately cover potential way to mitigate these
-issues.
+We first detail each pain point, and contrast this experience with that from
+other domains. We then separately cover potential way to mitigate these issues.
 
@@
 coding, and collaboration challenges. Two more are~\cite{WieseEtAl2019}:
-reproducibility, and software scope determination.  Smith et al.\ (2024)~\cite{SmithEtAl2024} 
-adds lack of software experience, documentation and technical debt.  Our small
-sample size means that we cannot conclude that these are not also issues with MI
-software.
+reproducibility, and software scope determination.  Smith et al.\
+(2024)~\cite{SmithEtAl2024} adds lack of software experience, documentation and
+technical debt.  Our small sample size means that we cannot conclude that these
+are not also issues with MI software.
 
@@
 \item[P\refstepcounter{pnum}\thepnum \label{P_LackDevTime}:] \textbf{Lack of
-Development Time:} This was felt to be the most significant obstacle, and
-is not uncommon
-\cite{SmithEtAl2024, WieseEtAl2019, PintoEtAl2018, PintoEtAl2016}.
-%DOUBLE BLIND
+Development Time:} This was felt to be the most significant obstacle, and is not
+uncommon \cite{SmithEtAl2024, WieseEtAl2019, PintoEtAl2018, PintoEtAl2016}.
 
@@
 Funding:} As research software is not yet seen as essential infrastructure,
-there is an ongoing struggle to attract funding to build and, even harder, maintain
-research software. This is a common complaint~\cite{SmithEtAl2024, GewaltigAndCannon2012, Goble2014, KaterbowAndFeulner2018}. %,  %DOUBLE BLIND
-Software output is not always counted when judging the academic excellence of
-academics, which forces researchers who write software to spend extra time on
+there is an ongoing struggle to attract funding to build and, even harder,
+maintain research software. This is a common complaint~\cite{SmithEtAl2024,
+GewaltigAndCannon2012, Goble2014, KaterbowAndFeulner2018}.  Software output is
+not always counted when judging the academic excellence of academics, which
+forces researchers who write software to spend extra time on
 publicity~\cite{WieseEtAl2019}. In some domains, like
@@
 privacy concerns;
-\item Tests are expensive and time-consuming because of the need for huge datasets;
+\item Tests are expensive and time-consuming because of the need for huge
+datasets;
 \item Software releases are difficult to manage;
@@
 At least to some extent the problems for MI software users are due to holes in
-their knowledge of software and computing technology, and sometimes also
-lack of expertize in their domain to be able to adequately use the software
-\cite{WieseEtAl2019}. 
-A similar pattern has been observed during interviews with
-LBM software developers \cite{SmithEtAl2024}. %%DOUBLE BLIND
+their knowledge of software and computing technology, and sometimes also lack of
+expertize in their domain to be able to adequately use the software
+\cite{WieseEtAl2019}.  A similar pattern has been observed during interviews
+with LBM software developers \cite{SmithEtAl2024}.
 
@@
 A number of developers perceive that some of the ``shifting technologies''
-problems could be alleviated by moving to web-based applications.
-Most of the teams (83\%) chose to develop
-native applications. Three of the eight we interviewed were
-building web applications, and another team was considering it.
-
-\item[P\ref{P_Correctness}:] \textbf{Correctness:}
-Testing was the most often mentioned strategy for ensuring correctness.  Teams
-mentioned several test-related activities, including test-driven development, component
-tests, integration tests, smoke tests, regression tests, self tests and
-automated tests.  This puts MI software ahead of other domains where
-insufficient testing is a problem~\cite{PintoEtAl2018} and
-insufficiently well-understood~\cite{HannayEtAl2009}.
+problems could be alleviated by moving to web-based applications.  Most of the
+teams (83\%) chose to develop native applications. Three of the eight we
+interviewed were building web applications, and another team was considering it.
+
+\item[P\ref{P_Correctness}:] \textbf{Correctness:} Testing was the most often
+mentioned strategy for ensuring correctness.  Teams mentioned several
+test-related activities, including test-driven development, component tests,
+integration tests, smoke tests, regression tests, self tests and automated
+tests.  This puts MI software ahead of other domains where insufficient testing
+is a problem~\cite{PintoEtAl2018} and insufficiently
+well-understood~\cite{HannayEtAl2009}.
 
@@
 
-\item \textbf{Backend servers can potentially deliver high performance.} If
-the principal bottleneck is computational, backend servers may outperform
-native applications, as long as latency is not an issue.
-Options include
-\href{https://www.django-rest-framework.org/} {Django}, \href{https://laravel.com/} {Laravel} and
-\href{https://nodejs.org/en/} {Node.js}.
-When traffic and latency are an issue,
+\item \textbf{Backend servers can potentially deliver high performance.} If the
+principal bottleneck is computational, backend servers may outperform native
+applications, as long as latency is not an issue.  Options include
+\href{https://www.django-rest-framework.org/} {Django},
+\href{https://laravel.com/} {Laravel} and \href{https://nodejs.org/en/}
+{Node.js}.  When traffic and latency are an issue,
 \href{https://github.com/gin-gonic/gin} {Gin} is an option.
@@
 
-%JC: I'm aware of too many caveats wrt this recommendation, so I am taking
-% it out.
-%SS: I agree with the removal.
-%
-%\item \textbf{Remove privacy from sensitive data.} For data with sensitive
-%information, a team can ask the data owner to remove such information or add
-%noise to protect privacy. One example is using de-identified copies of medical
-%images for testing.
-
 \end{itemize}
@@
 explains the relatively high quality of software in the Comprehensive R Archive
-Network (CRAN)~\cite{SmithEtAl2018_StatSoft}. %%DOUBLE BLIND
+Network (CRAN)~\cite{SmithEtAl2018_StatSoft}.
 
@@
 inspection, with this rate reaching as high as 85\% \cite{Jones2008}. The
-success rate of code inspection is generally higher than via testing, since testing
-averages around 30--35\% for the efficiency of defect removal \cite{Jones2008, EbertAndJones2009}. For research software, Kelly and Shepard \cite{KellyAndShepard2000}
-show a task based inspection approach can be effective.
+success rate of code inspection is generally higher than via testing, since
+testing averages around 30--35\% for the efficiency of defect removal
+\cite{Jones2008, EbertAndJones2009}. For research software, Kelly and Shepard
+\cite{KellyAndShepard2000} show a task based inspection approach can be
+effective.
 
@@
 The measurement process used has previously been shown to be reasonably
-reproducible.
-%% DOUBLE BLIND
-Smith et al.~\cite{SmithEtAl2016} reports grading five software products by two reviewers.
-Their rankings were almost identical. The relative comparisons in the AHP
-results will be consistent between graders, as long as the evaluators are
-consistent in their judgement.
+reproducible.  Smith et al.~\cite{SmithEtAl2016} reports grading five software
+products by two reviewers.  Their rankings were almost identical. The relative
+comparisons in the AHP results will be consistent between graders, as long as
+the evaluators are consistent in their judgement.
 
@@
 The study is internally valid if the suggested causal relations are trustworthy
-and not explainable by other factors~\cite{RunesonAndHost2009}. The identified threats to internal validity are:
+and not explainable by other factors~\cite{RunesonAndHost2009}. The identified
+threats to internal validity are:
 
@@
 differences, like the importance of data privacy for MI data, we found the
-approach to developing LBM software \cite{SmithEtAl2024} 
-%% DOUBLE BLIND
-and MI software to be similar.  Except for
-the domain specific aspects, the trends observed in the current study are
-similar to that for other research software.
+approach to developing LBM software \cite{SmithEtAl2024} and MI software to be
+similar.  Except for the domain specific aspects, the trends observed in the
+current study are similar to that for other research software.
 
@@
 \begin{table*}[ht!]
-    %\begin{center}
     \caption{Top performers for each quality (sorted by order of quality
@@
     \label{topperformerstable}
-    %\end{center}
 \end{table*}
@@
 
-%% DOUBLE BLIND
 We would like to thank Peter Michalski and Oluwaseun Owojaiye for fruitful
